\biblebook{2 Chronicles}

\begin{biblechapter} % 2 Chronicles 1
\verseWithHeading{Solomon Worships at Gibeon} And Solomon the son of David strengthened himself concerning his kingdom, and Adonai his God was with him and made him exceedingly great.
\verse And Solomon spoke to all Israel, to the commanders of thousands and of hundreds, and to the judges, and to all the leaders for all Israel, the heads of the \textit{families}.
\verse And Solomon and the whole assembly with him went to the high place that was at Gibeon, for the tent of assembly of God that Moses the servant of Adonai had made in the desert was there.
\verse (But the ark of God David had brought up from Kiriath-Jearim to the place David had prepared for it, for David had pitched a tent for it in Jerusalem.)
\verse And the bronze altar that Bezalel the son of Uri, the son of Hur, had made was there before the tabernacle of Adonai. And Solomon and the assembly sought it out.
\verse And Solomon went up there to the bronze altar before Adonai, which was at the tent of assembly, and he offered upon it a thousand burnt offerings.
\verseWithHeading{Solomon’s Prayer for Wisdom} On that night God appeared to Solomon and said to him, “Ask what I shall give to you.”
\verse Then Solomon said to God, “You yourself have shown steadfast loyal love to David my father and have made me king in his place.
\verse Now, O Adonai, let your word to David my father be fulfilled, for you yourself have made me king over a people as numerous as the dust of the earth.
\verse Now, give to me wisdom and knowledge that I may go out and come in before this people, for who can judge this, your great people?”
\verse Then God said to Solomon, “Because this was with your heart and you did not ask for wealth, possessions, honor, and the lives of your enemies, and also did not ask for \textit{long life}, but have asked for wisdom and knowledge that you might judge my people over whom I have made you king,
\verse wisdom and knowledge is given to you. And I will also give to you wealth, possessions, and honor, the like of which was not had by the kings who were before you, nor will there be the like after you.”
\verse So Solomon went from the high place which was at Gibeon to Jerusalem before the tent of assembly. And he reigned over Israel.
\verseWithHeading{Solomon’s Wealth} And Solomon gathered chariots and horsemen. And he had one thousand four hundred chariots and twelve thousand horsemen, and he placed them in the chariot cities and with the king in Jerusalem.
\verse And the king made silver and gold in Jerusalem like the stones. And he made cedar as abundant as the sycamore fig trees that were in the Shephelah.
\verse And Solomon’s horses were imported from Egypt and from Kue. The traders of the king received them from Kue at a price.
\verse They went up and exported a chariot from Egypt for six hundred shekels of silver and a horse for one hundred and fifty shekels. And these were \textit{likewise exported} to all the kings of the Hittites and the kings of Aram.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 2
\verseWithHeading{Solomon Petitions the King of Tyre}  And Solomon planned to build a house for the name of Adonai and a \textit{royal palace for himself}.
\verse And Solomon assigned seventy thousand men \textit{to bear burdens} and eighty thousand \textit{quarriers} in the hill country, and three thousand six hundred men to supervise them.
\verse And Solomon sent word to Huram king of Tyre, saying, “As you have dealt with David my father and sent cedar to him to build for himself a house in which to live, please deal with me.
\verse Look, I am building a house for the name of Adonai my God to dedicate to him, to offer sweet spices of incense before him, and for the regular rows of bread, and burnt offerings for mornings, evenings, Sabbaths, and new moon festivals, and for appointed feasts of Adonai our God which are everlasting for Israel.
\verse And the house that I am building will be great, for our God is greater than all gods.
\verse Now who indeed has adequate strength to build a house for him? For the heavens and \textit{the highest heavens} are not able to contain him. Now who am I that I would build a house for him, except to burn incense before him?
\verse So then, send to me skilled men to work with gold, silver, bronze, and iron, and with purple, crimson, and blue fabric, knowledgeable in engraving, with the skilled men who are with me in Judah and in Jerusalem, whom David my father has established.
\verse Send me trees of cedar, cypress, and algum from Lebanon, for I myself know that your servants are knowledgeable in cutting the trees of Lebanon. Now see, my servants will be with your servants
\verse to prepare trees in abundance for me, for the house that I am building will be great and wonderful.
\verse Now see, I will provide twenty thousand dry measures of crushed wheat, twenty thousand dry measures of barley, twenty thousand baths of wine, and twenty thousand baths of oil to the woodcarvers and those who cut timber.”
\verse Then Huram king of Tyre answered in a letter, and he sent word to Solomon: “Because Adonai loves his people, he has made you king over them.”
\verse Then Huram said, “Blessed be Adonai, the God of Israel, who has made the heavens and the earth, who has given to King David a wise son knowing discretion and understanding, who will build for Adonai a house and a royal palace for himself.
\verse So now I have sent a skilled man, knowledgeable and with understanding: my master Huram,
\verse a son of a woman from the daughters of Dan, and his father was a man of Tyre, knowledgeable for working with gold, silver, bronze, iron, stone, and wood, and with purple, blue, and crimson fabric, and with fine linen, for engraving any engraving and devising any plan that is given to him, with your skilled men and the skilled men of my lord David your father.
\verse Now as for the wheat, barley, oil, and wine that my lord mentioned, let him send that to his servants.
\verse And we ourselves will cut trees from Lebanon according to all your need, and we will bring them to you on rafts over the sea to Joppa, so that you may bring them up to Jerusalem.”
\verse Then Solomon counted all the resident alien men who were in the land of Israel after the census that David his father had taken of them. And there were found one hundred and fifty-three thousand.
\verse And he appointed from them seventy thousand \textit{to bear burdens}, eighty thousand \textit{to quarry} in the hill country, and three thousand six hundred overseers to make the people work.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 3
\verseWithHeading{Solomon Builds the Temple} Then Solomon began to build the house of Adonai in Jerusalem, on Mount Moriah, where Adonai had appeared to David his father, at the place that David had established, on the threshing floor of Ornan the Jebusite.
\verse And he began to build in the second month of the fourth year of his reign.
\verse Now these were the measurements of Solomon for building the house of God: the length in cubits by the former measurement was sixty cubits, and the width was twenty cubits.
\verse \textit{The length of the portico that was in front was the same as the front of the width of the house}: twenty cubits. And its height was one hundred and twenty cubits. And he overlaid it on the inside with pure gold.
\verse And the great house itself he covered with cypress wood, then he overlaid it with pure gold. And he put on it palm tree images and ornate chains.
\verse Then he overlaid the house with precious stone as decoration. (Now the gold was the gold of Parvaim.)
\verse And he overlaid the house with gold—the beams, the thresholds, the walls, and the doors. And he carved cherubim upon the walls.
\verse Then he made \textit{the most holy place}. Its length \textit{was equal to} the width of the house: twenty cubits. And its breadth was twenty cubits. And he overlaid it with six hundred talents of fine gold.
\verse And the weight for the nails was fifty shekels of gold. And he overlaid its upper rooms with gold.
\verse And he made two sculpted wood cherubim in \textit{the most holy place}, and he overlaid them with gold.
\verse The length of the outstretched wings of the cherubim were twenty cubits; one wing of five cubits touched the wall of the house, and the other wing of five cubits was touching the outstretched wing of the other cherub.
\verse And the wing of five cubits of the other cherub was touching the wall of the house, and its other wing five cubits long touching the wing of the first cherub.
\verse The wings of these cherubim extended twenty cubits, and they were standing on their feet, their faces toward the house.
\verse And he made the curtain of blue and purple and crimson and fine linen, and put cherubim on it.
\verse And at the front of the house he made two columns, thirty-five cubits high, and the capital on their top was five cubits.
\verse And he made decorative chains in the inner sanctuary and put them on top of the columns. And he made one hundred pomegranate ornaments, and put them on the chains.
\verse And he erected the columns in front of the temple, one on the south and one on the north. He called the name of the southern one Jakin, and the name of the northern one Boaz.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 4
\verseWithHeading{The Furnishings of Solomon’s Temple} And he made an altar of bronze, twenty cubits was its length, and twenty cubits was its width, and ten cubits was its height.
\verse Then he made the sea of molten metal, from brim to brim it was ten cubits, completely round. And it was five cubits in height, and \textit{its circumference measured} thirty cubits.
\verse Under it were figures of oxen all around it, ten cubits high, encircling the sea all around. The oxen were in two rows \textit{cast as one piece with it}.
\verse It was standing upon twelve oxen, three facing north, and three facing west, and three facing south, and three facing east. The sea was set upon them from above, and all their hindquarters faced \textit{inward}.
\verse And its thickness was a handbreadth, and its brim was like the working of the lip of a cup, the blossom of a lily. And it held three thousand baths.
\verse And he made ten basins. And he set five on the south and five on the north in which to wash; they washed off the work of the burnt offering in them, but the sea was for the priests to wash therein.
\verse And he made ten golden lampstands according to their custom, and he set them in the temple, five on the south and five on the north.
\verse And he made ten tables and placed them in the temple, five on the south and five on the north. And he made a hundred drinking bowls of gold.
\verse And he made the courtyard of the priests and the great outer courtyard and the doors for the outer court. And he overlaid their doors with bronze.
\verse And he set the sea at the southeast corner of the temple.
\verse And Huram made the pots, the shovels, and the drinking bowls. So Hiram finished making the work that he made on the house of God for King Solomon:
\verse the two columns, the bowls, and the two capitals on top of the columns, and the two latticeworks to cover the two bowls of the capitals that were on top of the columns,
\verse and the four hundred pomegranates for the two latticeworks, two rows of pomegranates for the latticework, to cover the two bowls of the capitals that were before the columns.
\verse And he made the water carts, and he made the basins on the water carts,
\verse and the one sea and the twelve oxen underneath it.
\verse And Huram-abi made for King Solomon the pots, the shovels, the three-pronged meat forks, and all the utensils of polished bronze for the house of Adonai.
\verse The king cast them in the plain of the Jordan, in the clay soil between Succoth and Zeredah.
\verse Solomon made all these utensils in great abundance, for the weight of the bronze could not be determined.
\verse So Solomon made all the objects that were in the house of God: the altar of gold; the tables upon which was the bread of the presence;
\verse the lampstands, and the lamps for burning according to the custom before the inner sanctuary, of solid gold;
\verse and the blossoms, the lamps, and the tongs that were of solid gold;
\verse and the snuffers, the drinking bowls, the dishes, and the firepans, of solid gold; and the entrance to the house, the inner doors to \textit{the most holy place}, and the doors to the house of the temple were of gold.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 5
\verseWithHeading{The Ark Installed in the Temple} When all the work that Solomon did for the house of Adonai was finished, Solomon brought the holy objects of David his father: the silver, the gold, and all the objects he had put into the storehouses of the house of God.
\verse Then Solomon assembled all the elders of Israel and all the heads of the tribes, the leaders of the \textit{families} for the \textit{Israelites}, to Jerusalem in order to bring up the ark of the covenant of Adonai from the city of David, which is Zion.
\verse And all the men of Israel assembled before the king at the feast that is in the seventh month.
\verse And all the elders of Israel came, and the Levites took up the ark.
\verse And they brought up the ark, the tent of assembly, and all the holy objects that were in the tent; the Levitical priests brought them up.
\verse And King Solomon and the whole community of Israel that was assembled before him were before the ark, sacrificing sheep and cattle that could not be counted nor numbered for abundance.
\verse Then the priests brought up the ark of the covenant of Adonai to its place in the inner sanctuary of the house, into \textit{the most holy place}, underneath the wings of the cherubim.
\verse The cherubim were spreading their wings over the place of the ark, so that the cherubim were covering the ark and over its poles from above.
\verse But the poles were so long that the ends of the poles from the ark were seen from before the inner sanctuary, but they could not be seen from the outside. And they are there until this day.
\verse There was nothing in the ark but the two stone tablets that Moses had put inside at Horeb on which Adonai had \textit{made} a covenant with the \textit{Israelites} when they went out from Egypt.
\verse And when the priests came out of the sanctuary (for all the priests who were present sanctified themselves without keeping their divisions)
\verse and all the Levitical singers—Asaph, Heman, Jeduthun, their sons, their brothers, dressed in fine linen, with cymbals, harps, and stringed instruments—they were standing to the east of the altar. And with them outside were one hundred and twenty priests who were trumpet blowers.
\verse And it was the primary duty of the trumpeters and singers to make themselves heard with one voice, to praise and give thanks to Adonai. And when a sound from the trumpets, cymbals, and other instruments of song was raised to Adonai—for he is good, because his loyal love is everlasting—then the house, the house of Adonai, was filled with a cloud.
\verse And the priests were not able to stand to minister because of the cloud, for the glory of Adonai filled the house of God.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 6
\verseWithHeading{Solomon Blesses the People} Then Solomon said, “Adonai has decided to dwell in the deep gloom.
\verse But I have built for you an exalted house, even a place for you to dwell in forever.”
\verse Then the king turned around and blessed all the assembly of Israel, and all the assembly of Israel were standing.
\verse And he said, “Blessed be Adonai, the God of Israel, who spoke with his mouth with David my father, and with his hands he has fulfilled it, saying,
\verse ‘From the day that I brought my people out of the land of Egypt I did not choose a city among all the tribes of Israel to build a house in order for my name to be there. Nor did I choose a man to be leader over my people Israel.
\verse But I have chosen my name to be there in Jerusalem, and I have chosen David to be over my people Israel.’
\verse Now, it was in the heart of David my father to build a house for the name of Adonai, the God of Israel.
\verse But Adonai said to David my father, ‘Because it was in your heart to build a house for my name, you did well that it was in your heart,
\verse but you yourself will not build the house, but your son who will proceed from your loins; he himself will build the house for my name.’
\verse Now Adonai has fulfilled his word that he spoke. I have risen in the place of David my father and am seated on the throne of Israel as Adonai has spoken. I have built the house for the name of Adonai, the God of Israel.
\verse And I have there the ark in which is the covenant of Adonai that he \textit{made} with the \textit{Israelites}.”
\verseWithHeading{Solomon’s Prayer to Dedicate the Temple} Then Solomon stood before the altar of Adonai before all the assembly of Israel and spread out his hands.
\verse (For Solomon had made a platform of bronze five cubits long and five cubits wide and three cubits high, and placed it in the midst of the outer court.) And he stood upon it, then knelt down on his knees before the whole assembly of Israel and spread out his hands to heaven.
\verse And he said, “O Adonai, God of Israel, there is none like you, a God in the heavens and on the earth, keeping the covenant and loyal love with your servants who walk before you with all their heart,
\verse that you have kept for your servant David my father what you spoke to him. You spoke with your mouth, and by your hand you have fulfilled it on this day.
\verse And now, O Adonai, God of Israel, keep for your servant David, my father, what you have spoken to him, saying, ‘\textit{You shall not lack} a man to sit on the throne of Israel before me, if only your sons keep their way, to walk in my law as you have walked before me.’
\verse And now, O Adonai, God of Israel, let your word that you have spoken to your servant David be confirmed.
\verse “But will God indeed dwell with humankind upon the earth? Look, the heavens and \textit{the highest heavens} cannot contain you! Surely then this house that I have built will not contain you!
\verse But turn to the prayer of your servant and to his plea, O Adonai my God, to hearken to the pleading and to the prayer that your servant is praying before you,
\verse that your eyes may be open day and night to this house, to the place that you have promised to place your name there, that you may hear the prayer that your servant has prayed to this place.
\verse And listen to the pleas of your servant and your people Israel when they pray toward this place, that you yourself might hear from the place of your dwelling, from the heavens, that you might hear and forgive.
\verse “If a man sins against his neighbor and \textit{makes him swear an oath} and comes with an oath before your altar in this house,
\verse then may you yourself hear from the heavens and act and judge your servants, to repay the guilty by bringing his way upon his head and to vindicate the righteous by giving to him according to his righteousness.
\verse “And if your people Israel are defeated before an enemy because they have sinned against you and they repent and confess your name and pray and plead before you in this house,
\verse then may you yourself hear from the heavens and forgive the sin of your people Israel and again bring them back to the land that you gave to them and to their ancestors.
\verse “When the heavens are stopped up and there is no rain because they have sinned against you and they pray to this place and confess your name and turn back from their sin when you humble them,
\verse then may you yourself hear from the heavens and forgive the sin of your servants and your people Israel, when you teach them the good way in which they should walk, and give rain upon your land that you have given to your people as an inheritance.
\verse “If there is a famine in the land; if there is disease, blight, mildew, locust, and caterpillar; if it happens that his enemy lays siege to him in the land and his gates; any kind of plague, any kind of disease,
\verse then any prayer, any plea that is made by any person and by all your people Israel, each one who knows his own affliction and his own anguish and who spreads out his hands to this house,
\verse then may you yourself hear from the heavens, your dwelling place, and forgive and give to each according to all his ways, since you know the heart of the children of humankind,
\verse that they may fear you and walk in your ways all the days that they are alive upon the face of the land that you have given to our ancestors.
\verse “And also, with respect to the foreigner, he who is not of your people Israel, but comes from a distant land, for the sake of your great name and your powerful hand and outstretched arm, when he comes and prays to this house,
\verse then may you yourself hear from the heavens, from your dwelling place, and do according to all that the foreigner asks of you, so that all peoples of earth will know of your name and fear you, as do your people Israel, and that they may know that your name \textit{possesses} this house that I have built.
\verse “When your people go out to battle against their enemy according to the way that you have sent them, and they pray to you toward this city that you have chosen and the house that I have built for your name,
\verse then may you hear their prayer and their pleas from the heavens and uphold their case.
\verse When they sin against you—for there is no person who does not sin—and you become angry with them and place them before their enemy that they carry them away as captives to a land far or near,
\verse but \textit{if they repent} in the land where they were taken captive and \textit{repent} and beg you for mercy in the land of their captivity, saying, ‘We have sinned and acted perversely and done wickedly,
\verse and we return to you,’ with their whole heart and with all their inmost being in the land of their captivity where they took them captive, and if they pray toward their land that you have given to their ancestors and the city that you have chosen and to the house that I have built for your name,
\verse then may you hear their prayer and their pleas from the heavens, from your dwelling place, and uphold their case and forgive your people who sinned against you.
\verse Now, O my God, please let your eyes be open and your ears attentive to the prayer of this place!
\verse “And now arise, O Adonai, to your resting place, you and the ark of your strength! O Adonai, let the priests be clothed with salvation, and let your holy ones rejoice in your goodness!
\verse O Adonai, do not turn away the face of your anointed one! Remember the loyal love of your servant, David!”
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 7
\verseWithHeading{God Answers with Fire from Heaven} And when Solomon finished praying, then fire came down from heaven and consumed the burnt offering and the sacrifices, and the glory of Adonai filled the house.
\verse And the priests were not able to go into the house of Adonai, for the glory of Adonai had filled the house.
\verse When all the \textit{Israelites} saw the fire come down and the glory of Adonai upon the house, they knelt down with their faces to the ground on the pavement and worshiped and gave thanks to Adonai, for he is good, for his loyal love is everlasting.
\verseWithHeading{The Temple Dedication} Then the king and all the people were offering a sacrifice before Adonai.
\verse And King Solomon offered a sacrifice of twenty-two thousand cattle and one hundred and twenty thousand sheep. So the king and all the people dedicated the house of God.
\verse And the priests stood at their posts with the Levites, with the instruments of the song of Adonai that King David had made to give thanks to Adonai—for his loyal love is everlasting—whenever David offered praise by their hand. Now the priests sounded trumpets standing opposite them, and all Israel stood.
\verse And Solomon consecrated the middle of the courtyard that was before the house of Adonai, for there he had made the burnt offerings and the fat of the fellowship offerings, for the bronze altar that Solomon had made was not able to hold the burnt offerings, cereal offerings, and the fat.
\verse And Solomon held a feast at that time of seven days. And all Israel was with him, a very great assembly, from Lebo-Hamath to the river of Egypt.
\verse And on the eighth day they made a solemn assembly, for they held the dedication of the altar seven days and the feast seven days.
\verse And on the twenty-third day of the seventh month he sent the people away to their tents rejoicing and festive of heart on account of the goodness that Adonai had done for David, Solomon, and Israel his people.
\verse And Solomon finished the house of Adonai and the house of the king. And all that came into the heart of Solomon to do with respect to the house of Adonai and his own house he accomplished.
\verseWithHeading{God’s Charge to Solomon} Then Adonai appeared to Solomon at night and said to him, “I have heard your prayer and have chosen this place for myself as a house of sacrifice.
\verse When I hold back the heavens so that there is not rain, and when I command the locust to devour the earth, and if I send disease among my people,
\verse then if my people who are called by my name will humble themselves and will pray and will seek my face and will turn from their evil ways, then I myself shall hear from the heavens and will forgive their sins and heal their land.
\verse Now my eyes will be open and my ears attentive to the prayer of this place.
\verse And now I have chosen and consecrated this house for my name to be there forever. My eyes and my heart will be there \textit{for all time}.
\verse Now as for you, if you will walk before me as David your father walked and do according to all that I have commanded you and will keep my ordinances and judgments,
\verse then I will establish the throne of your kingdom as I promised to David your father, saying, ‘A man shall not be cut off for you from ruling over Israel.’
\verse “But if you turn yourselves away and forsake my ordinances and my commandments which I have given before you all and will go and serve other gods and bow down to them,
\verse then I will uproot them from upon my land that I have given to them, and this house that I have consecrated for my name I will send away from before my face, and I will make it a proverb and a taunt among all the nations.
\verse And as for this house, which was exalted, all who pass by it will be appalled and will say, ‘Why has Adonai done thus to this land and to this house?’
\verse Then they will say, ‘Because they forsook Adonai, the God of their ancestors, who brought them out from the land of Egypt, and they laid hold of other gods and worshiped them and served them. Therefore he has brought upon them all this evil.’ ”
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 8
\verseWithHeading{Solomon’s Accomplishments and Deeds} And it happened that at the end of twenty years in which Solomon had built the house of Adonai and his own house,
\verse Solomon also built the cities that Huram had given to him, and he settled the \textit{Israelites} in them.
\verse And Solomon went to Hamath-Zobah and laid siege upon it.
\verse And he built Tadmor in the desert, and he built all the storage cities in Hamath.
\verse He also built Upper Beth-Horon and Lower Beth-Horon, fortified cities with walls, gates, and bars;
\verse Baalath and all the storage cities that belonged to Solomon; all the cities for chariots and cities for horses; and any desire of Solomon that he desired to build in Jerusalem, in Lebanon, and in all the land of his dominion.
\verse All the people who remained from the Hittites, the Amorites, the Perizzites, the Hivites, and the Jebusites who were not from Israel,
\verse from the descendants who were left over after them in the land, whom the Israelites had not destroyed, Solomon conscripted them as forced labor until this day.
\verse But from the \textit{Israelites} Solomon did not give as servants for his work, but they were men of war, his commanders of his officers, and commanders of his chariots and his horses.
\verse And these were the commanders of the garrisons that belonged to King Solomon, two hundred and fifty who exercised authority over the people.
\verse And Solomon brought up the daughter of Pharaoh from the city of David to the house that he had built for her, for he said, “My wife shall not dwell in the house of David, king of Israel, for those places to which the ark of Adonai has come are holy.
\verse Then Solomon offered burnt offerings to Adonai on the altar of Adonai that he had built before the portico,
\verse \textit{as was the daily requirement}, to offer up according to the commandments of Moses for Sabbaths, new moon festivals, and the three \textit{annual} feasts: the Feast of Unleavened Bread, the Feast of Weeks, and the Feast of Booths.
\verse And in accord with the ordinance of David his father he appointed working groups of priests for their service, Levites for their responsibilities for praise and cultic service before the priests \textit{according to the daily requirement}, and the gatekeepers in their working groups, gate by gate. For thus David the man of God had commanded.
\verse And they did not turn aside from the commandment of the king concerning the priests and the Levites concerning anything, and concerning the treasury rooms.
\verse And Solomon accomplished all the work from the day the foundation stone was laid for the house of Adonai until its completion, and the house of Adonai was finished.
\verse Then Solomon went to Ezion-Geber, and to Elath, on the edge of the sea in the land of Edom.
\verse And Huram sent to him, by the hand of his servants, ships and servants knowledgeable of the sea. And they went with the servants of Solomon to Ophir, and from there they collected four hundred and fifty talents of gold and brought it to King Solomon.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 9
\verseWithHeading{Solomon and the Queen of Sheba} Now the queen of Sheba heard of the fame of Solomon, and she came to Jerusalem to test Solomon with hard questions, with much wealth and camels, bearing spices and much gold and precious stones. And she came to Solomon and told him all that was \textit{on her mind}.
\verse And Solomon \textit{answered all her questions}. Not a word was hidden from Solomon that he did not answer for her.
\verse When the queen of Sheba saw the wisdom of Solomon and the house he had built,
\verse and the food of his table, and the position of his officers, the duty of his servants and their clothing, his cupbearers and their clothing, and his burnt offerings that he offered at the house of Adonai, there was no longer any breath in her.
\verse And she said to the king, “The word which I heard in my land concerning your words and your wisdom is true,
\verse but I did not believe their words until I came and saw with my own eyes. And behold, half the greatness of your wisdom was not reported to me; you surpass the rumors that I had heard.
\verse Blessed are your men, and blessed are these your servants who stand before you regularly and hear your wisdom!
\verse Blessed be Adonai your God who took delight in you, to put you on his throne as king for Adonai your God! Because your God loved Israel, he established him forever and has put you over them as king to do justice and righteousness!”
\verse And she gave to the king one hundred and twenty talents of gold and a very great amount of spices and precious stones. And there were no spices such as those that the queen of Sheba gave to King Solomon.
\verse Moreover the servants of Hiram and the servants of Solomon who brought gold from Ophir brought algum wood and precious stones.
\verse And the king made steps for the house of Adonai and the house of the king, and lyres and harps for the singers from the algum wood. And there was not seen anything like them before in the land of Judah.
\verse Now King Solomon gave to the queen of Sheba every desired thing she asked, besides what she had brought to the king. Then she turned and went to her land, she and her servants.
\verseWithHeading{Solomon’s Great Wealth} And it happened that the weight of the gold that came to Solomon in one year was six hundred and sixty-six talents of gold,
\verse besides the gold and silver the traders and merchants brought, and all the kings of Arabia and the governors of the land brought.
\verse And King Solomon made two hundred shields of beaten gold; six hundred shekels of beaten gold went into each shield.
\verse And he made three hundred small shields of beaten gold; three hundred shekels went into each small shield. And the king put them in the House of the Forest of Lebanon.
\verse And the king made a large ivory throne and overlaid it with pure gold.
\verse And the throne had six steps, and a footstool of gold was attached to the throne. And there were armrests \textit{on each side}. And by \textit{the seat} were two lions standing beside the armrests.
\verse And twelve lions were standing there on each of the six steps \textit{on each end}. And there was nothing like it made in all the kingdom.
\verse And all the drinking vessels of King Solomon were of gold, and all the vessels of the House of the Forest of Lebanon were of pure gold. Silver was not valued as anything in the days of Solomon.
\verse For the ships of the king went to Tarshish with the servants of Huram. Once \textit{every three years} the ships of Tarshish came carrying gold, silver, ivory, apes, and peacocks.
\verse Thus King Solomon was greater than all the kings of the earth in wealth and wisdom.
\verse And all the kings of the earth sought the face of Solomon to hear his wisdom that God put in his heart.
\verse And each of them brought his gifts, objects of silver, objects of gold, cloaks, myrrh, spices, horses, and mules, as an amount year by year.
\verse And Solomon had four thousand stalls for horses and chariots, and twelve thousand horsemen. And he stationed them in the chariot cities and with the king in Jerusalem.
\verse And he ruled over all the kings from the River to the land of the Philistines, to the boundary of Egypt.
\verse And the king made silver in Jerusalem as stones, and he made cedar as the sycamore trees that are in the Shephelah for abundance.
\verse And the horses were imported from Egypt and from all lands for Solomon.
\verseWithHeading{The Death of Solomon} As for the remainder of the words of Solomon from the first to the last, are they not written in the chronicles of Nathan the prophet and in the prophecy of Ahijah the Shilonite, and in the visions of Iddo the seer, concerning Jeroboam the son of Nebat?
\verse And Solomon reigned in Jerusalem over all Israel forty years.
\verse And Solomon slept with his ancestors, and they buried him in the city of David his father. And Rehoboam his son reigned in his place.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 10
\verseWithHeading{Rehoboam’s Debacle} Then Rehoboam went to Shechem, for all Israel had come to Shechem to make him king.
\verse And it happened that when Jeroboam the son of Nebat heard it—now he was in Egypt, where he had fled from the presence of King Solomon—Jeroboam returned from Egypt.
\verse And they sent and called him. Then Jeroboam and all Israel went, and they spoke to Rehoboam, saying,
\verse “Your father made our yoke heavy. Now, therefore, lighten the hard service of your father and his heavy yoke which he put upon us, and we will serve you.”
\verse And he said to them, “In three days return to me again.” And the people went away.
\verse Then King Rehoboam took counsel with the elders who had been \textit{serving} before Solomon his father when he was alive, saying, “What word do you advise to answer this people?”
\verse And they said to him, “If you will be good to this people and please them, then speak good words to them. Then they will be your servants \textit{forever}.”
\verse But he forsook the advice of the elders that advised him and took counsel of the young men who had grown up with him who were \textit{serving} before him.
\verse And he said to them, “What do you advise that we should say in return to this people, who said to me, ‘Lighten the yoke that your father has put upon us’?”
\verse Then the young men who had grown up with him said to him, “Thus you should say to this people who have said to you, ‘Your father made our yoke heavy, so you yourself should lighten it for us.’ Thus you should say to them, ‘My little finger is thicker than the loins of my father.
\verse So now, my father laid upon you a heavy yoke, but I myself will add to the yoke. My father disciplined you with whips, but I myself will do so with scorpions.’ ”
\verse Then Jeroboam and all the people came to Rehoboam on the third day as the king had spoken, saying, “Return to me on the third day.”
\verse And the king answered them harshly, and King Rehoboam forsook the advice of the elders.
\verse And he spoke to them according to the advice of the young men, saying, “I will make your yoke heavy, and I myself will add to it. My father disciplined you with whips, but I will do so with scorpions.”
\verse So the king did not listen to the people, for it was a turning of events from God, so that Adonai might fulfill his word that he had spoken by the hand of Ahijah the Shilonite to Jeroboam son of Nebat.
\verse So all Israel saw that the king would not listen to them, and the people answered the king, saying, “What portion is there for us in David? We have no inheritance in the son of Jesse. Each to your tents, O Israel! Now look to your own house, David!” And all Israel went to their own tents.
\verse But as for the \textit{Israelites} who were living in the cities of Judah, Rehoboam ruled over them.
\verse Then King Rehoboam sent Hadoram, who was \textit{in charge of} the forced labor, but the \textit{Israelites} stoned him with stones, and he died. Then King Rehoboam hastily went up into the chariot to flee to Jerusalem.
\verse So Israel has rebelled against the house of David until this day.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 11
\verseWithHeading{Rehoboam’s Kingdom} When Rehoboam came to Jerusalem he assembled the house of Judah and Benjamin, one hundred and eighty thousand \textit{chosen warriors}, to fight against Israel to restore the kingdom to Rehoboam.
\verse But the word of Adonai came to Shemaiah, the man of God, saying,
\verse “Say to Rehoboam, son of Solomon, king of Judah, and to all Israel in Judah and Benjamin,
\verse ‘Thus says Adonai: “Do not go up, and do not fight against your brothers. Return, each man to his house, for this matter has come from me.” ’ ” So they obeyed the words of Adonai and turned back from going against Jeroboam.
\verse And Rehoboam lived in Jerusalem, and he built cities as strongholds in Judah.
\verse He built Bethlehem, Etah, Tekoa,
\verse Beth-Zur, Socoh, Adullam,
\verse Gath, Mareshah, Ziph,
\verse Adoraim, Lachish, Azekah,
\verse Zorah, Aijalon, and Hebron, fortified cities that are in Judah and in Benjamin.
\verse And he strengthened the fortifications and put commanders in them, along with stores of food, olive oil, and wine.
\verse And in all the cities he put shields and spears, and he greatly strengthened them. So he had Judah and Benjamin.
\verseWithHeading{The Priests and Levites Come to Jerusalem} Now the priests and the Levites who were throughout all Israel took their stand with him from all their territories.
\verse For the Levites left their pasturelands and their property and came to Judah and to Jerusalem, for Jeroboam and his sons had rejected them from serving as priests to Adonai.
\verse And he appointed his own priests for the high places and for the goat idols and for the bull calves that he had made.
\verse Then after them, from all the tribes of Israel, those who set their heart to seek Adonai, the God of Israel, came to Jerusalem to offer to Adonai, the God of their ancestors.
\verse And they strengthened the kingdom of Judah and made Rehoboam, the son of Solomon, secure for three years, for they walked in the way of David and Solomon for three years.
\verseWithHeading{Rehoboam’s Harem and Family} Then Rehoboam took to himself as a wife Mahalath, the daughter of Yerimot son of David, and of Abihail the daughter of Eliab the son of Jesse.
\verse And she bore to him sons: Jeush, Shemariah, and Zaham.
\verse And after her he took Maacah the daughter of Absalom, who bore to him Abijah, Attai, Ziza, and Shelomith.
\verse And Rehoboam loved Maacah the daughter of Absalom more than all his wives and concubines (for he took eighteen wives and sixty concubines, and he fathered twenty-eight sons and sixty daughters).
\verse And Rehoboam appointed Abijah the son of Maacah as chief and crown prince over his brothers, in order to make him king.
\verse And he dealt wisely and distributed some of his sons throughout all the land of Judah and Benjamin, through all the fortified cities, and gave abundant provisions to them and obtained many wives for them.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 12
\verseWithHeading{Shishak of Egypt Invades Jerusalem} And when the kingdom of Rehoboam was established and when he was strengthened, he forsook the law of Adonai, and all Israel with him.
\verse And it happened that in the fifth year of King Rehoboam (for they had acted unfaithfully against Adonai), Shishak the king of Egypt came up against Jerusalem
\verse with one thousand two hundred chariots and sixty thousand horsemen. And there was no number to the people who came up with him from Egypt—Libyans, Sukkites, and Cushites.
\verse And he took the fortified cities that belonged to Judah, and he came up to Jerusalem.
\verse Then Shemaiah the prophet came to Rehoboam and the princes of Judah who had gathered at Jerusalem from before Shishak. And he said to them, “Thus says Adonai: ‘You yourselves have abandoned me, and I myself have surely abandoned you into the hand of Shishak.’ ”
\verse Then the princes of Israel and the king humbled themselves and said, “Adonai is righteous.”
\verse And when Adonai saw that they humbled themselves, the word of Adonai came to Shemaiah, saying, “They have humbled themselves; I will not destroy. I will give to them some way of escape, that my anger not be poured out against Jerusalem by the hand of Shishak.
\verse However, they shall be his servants, that they might know my service and the service of the kingdoms of other countries.”
\verse Then Shishak king of Egypt came up against Jerusalem, and he took the treasures out of the house of Adonai and the treasures out of the king’s house. He took everything. He also took the small shields of gold that Solomon had made.
\verse And King Rehoboam made small shields of bronze in their place and committed them into the hand of the commanders of the guards who were keeping the entrance of the house of the king.
\verse And whenever the king went into the house of Adonai, the guards came and carried them, and then they returned them to the alcove of the guards.
\verse And when he humbled himself, the anger of Adonai was turned away from him, so that he did not destroy the city completely. Moreover, matters were well in Judah.
\verseWithHeading{The Death of Rehoboam} So King Rehoboam was strengthened in Jerusalem, and he reigned there. Now, Rehoboam was forty-one years old when he became king, and he reigned seventeen years in Jerusalem, the city where Adonai had chosen to put his name out of all the tribes of Israel. And the name of his mother was Naamah the Ammonite.
\verse And he did evil, for he did not set his heart to seek Adonai.
\verse Now the words of Rehoboam from the first to the last, are they not written in the chronicles of Shemaiah the prophet and Iddo the seer as a record? And there were battles between Rehoboam and Jeroboam all their days.
\verse And Rehoboam slept with his ancestors, and he was buried in the city of David, and Abijah his son reigned in his place.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 13
\verseWithHeading{Abijah and Civil War} In the eighteenth year of King Jeroboam, Abijah reigned over Judah.
\verse He reigned for three years in Jerusalem. And the name of his mother was Micaiah the daughter of Uriel from Gibeah.
\verse Now there was war between Abijah and between Jeroboam.
\verse Then Abijah stood up on top of Mount Zemaraim that is in the hill country of Ephraim and said, “Hear me, O Jeroboam and all Israel!
\verse Do you not all know that Adonai the God of Israel gave the kingdom to David over Israel forever, to him and to his sons by a covenant of salt?
\verse But Jeroboam the son of Nebat, a servant of Solomon the son of David, rose up and rebelled against his master,
\verse and worthless men, \textit{sons of wickedness}, were gathered about him and became hostile against Rehoboam the son of Solomon, when Rehoboam was a boy and gentle of heart and was not strong enough to stand before them.
\verse So now you yourselves are considering that you are strong before the kingdom of Adonai which is in the hand the sons of David, since you yourselves are a great multitude and you have bull calves of gold that Jeroboam has made for idols.
\verse Did you not drive out the priests of Adonai, the sons of Aaron, the Levites? Have you not made for yourselves priests like all the peoples of the lands? All who come \textit{to be consecrated for priestly service} with a young bull or seven rams becomes a priest of what is not a god.
\verse But as for us, Adonai is our God, and we have not forsaken him. The priests serving Adonai are the sons of Aaron, the Levites, according to their service.
\verse They offer burnt offerings to Adonai \textit{every morning} and \textit{every evening}, sweet spices of incense, and the rows of bread upon the pure table. And the golden lampstand with its lamps they keep kindled \textit{every evening}, for we are keeping the requirement of Adonai our God, but you yourselves have forsaken him.
\verse And look, God is with us at the head, and his priests have the trumpets for sounding blasts against you, O sons of Israel. Do not fight against Adonai, the God of your ancestors, for you will not succeed.”
\verse Then Jeroboam sent around an ambush to come behind them. So they were in front of Judah, but the ambush was behind them.
\verse When Judah turned, then behold, the battle against them was in front and behind, and they cried to Adonai, and the priests blew on the trumpets.
\verse Then the men of Judah shouted, and it happened that when the men of Judah were shouting, then God defeated Jeroboam and all Israel before Abijah and Judah.
\verse And the \textit{Israelites} fled from before Judah, and God gave them into their hand.
\verse And Abijah and his people struck a great blow against them, and the dead from Israel that fell were five hundred thousand chosen men.
\verse And the \textit{Israelites} were subdued at that time, and the people of Judah were victorious, for they relied upon Adonai the God of their ancestors.
\verse And Abijah pursued after Jeroboam and took cities from him: Bethel with its villages, Jeshanah with its villages, and Ephron with its villages.
\verse And Jeroboam did not regain strength again in the days of Abijah. And Adonai plagued him and he died.
\verse But Abijah became strong, and he took to himself fourteen wives and fathered twenty-two sons and sixteen daughters.
\verse Now the rest of the words of Abijah and his ways and his words are written in the story of the prophet Iddo.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 14
\verseWithHeading{The Reign of Asa in Judah}  And Abijah slept with his ancestors, and they buried him in the city of David. And Asa his son became king in his place. In his days the land was at rest ten years.
\verse Asa did that which was good and right in the eyes of Adonai his God.
\verse He removed the foreign altars and high places, broke down the stone pillars, and cut down the Asherahs.
\verse And he commanded Judah to seek Adonai, the God of their ancestors, and to keep the law and the commandment.
\verse And he removed the high places and the incense stands from all the cities of Judah, and the kingdom had rest under him.
\verse And he built fortified cities in Judah, for the land was at rest. Under him there was no war in those years, for Adonai had given rest to him.
\verse And he said to Judah, “Let us build these cities and surround them with walls, towers, gates, and bars while the land is before us, because we have sought Adonai our God. We have sought and he has given rest to us on every side.” So they built and had success.
\verse And Asa had an army of three hundred thousand from Judah, bearing shields and spears, and two hundred and eighty thousand from Benjamin who carried shields and drew bows. All of these were mighty warriors of strength.
\verse Zerah the Cushite came out against them with an army of a thousand thousands and three hundred chariots. And he came to Mareshah.
\verse And Asa went out before him, and a battle was put in order in the valley of Zepah at Mareshah.
\verse Then Asa cried out to Adonai his God. And he said, “O Adonai, there is none with you to help between the great and the powerless. Help us, O Adonai our God, for we rely upon you, and in your name we have come against this multitude. O Adonai, you are our God! Let no man prevail against you!”
\verse So Adonai defeated the Cushites before Asa and before Judah, and the Cushites fled.
\verse Then Asa and the army with him pursued them to Gerar. And many from the Cushites fell, for they had no deliverance, for they were shattered before Adonai and before his army. And they carried away much war booty.
\verse And they struck all the cities around Gerar, for the fear of Adonai was upon them. And they plundered all the cities, for there was much plunder in them.
\verse And they also struck the tents of the herdsmen and carried away many sheep and camels. Then they returned to Jerusalem.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 15
\verseWithHeading{Asa’s Religious Reforms} Now Azariah the son of Oded—the Spirit of God came upon him.
\verse And he went out before Asa and said to him, “Hear me, O Asa and all of Judah and Benjamin! Adonai is with you while you are with him. And if you will seek him he will be found by you. But if you forsake him he will forsake you.
\verse Now Israel has been without the true God many days, and without a teaching priest, and without law,
\verse but at its trouble he returned to Adonai, the God of Israel. They sought him, and he was found by them.
\verse And in those times there was no peace for the one going out and the one coming in, for great tumults were upon all the inhabitants of the lands.
\verse Nation was crushed by nation, and city was against city, for God threw them into confusion by all sorts of trouble.
\verse But as for you, be strong and let not your hands be weak, for there is reward for your labor.”
\verse And when Asa heard these words, the prophecy of Oded the prophet, he took courage and removed the vile idols from all the lands of Judah and Benjamin, and from the cities that he had taken in the hill country of Ephraim, and he repaired the altar of Adonai that was in front of the portico of Adonai.
\verse And he gathered all Judah and Benjamin and those sojourning with them, from Ephraim, Manasseh, and Simeon, for many had deserted to him from Israel when they saw that Adonai his God was with him.
\verse And they were gathered at Jerusalem in the third month of the fifteenth year of the reign of Asa.
\verse And they sacrificed to Adonai on that day from the war booty they brought back: seven hundred oxen and seven thousand sheep.
\verse And they entered into a covenant to seek Adonai, the God of their ancestors, with all their heart and with all their inmost being,
\verse but all who will not seek Adonai the God of Israel should be killed, from young to old, from men to women.
\verse And they took an oath to Adonai with a great voice, with shouting, with trumpets, and with horns.
\verse And all Judah rejoiced over the oath, for they swore with all their heart. And they sought him with their whole desire, and he was found by them, and Adonai gave rest to them all around.
\verse And also Maacah, the mother of Asa, the king removed her from being queen, because she had made a repulsive image for Asherah. And Asa cut down her repulsive image, and he crushed and burned it at the Wadi Kidron.
\verse But the high places were not removed from Israel. Nevertheless, the heart of Asa was fully devoted all his days.
\verse And he brought the holy objects of his father and his own holy objects into the house of God—silver and gold and vessels.
\verse And there was no war until the thirty-fifth year of the reign of Asa.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 16
\verseWithHeading{The End of Asa’s Reign} In the thirty-sixth year of the reign of Asa, Baasha the king of Israel went up against Judah. And he built Ramah in order not to allow anyone to go out or come in to Asa, king of Judah.
\verse Then Asa brought out silver and gold from the storehouses of the house of Adonai and the house of the king, and he sent them to Ben-Hadad, king of Aram, who lived in Damascus, saying,
\verse “There is a covenant between me and you, as there was between my father and your father. Look, I am sending you silver and gold. Go, break your covenant with Baasha, king of Israel, that he might withdraw from me.”
\verse And Ben-Hadad listened to King Asa, and he sent commanders of his troops that were with him against the cities of Israel. And they struck Ijon, Dan, Abel-Maim, and all the storage cities of Naphtali.
\verse And it happened that when Baasha heard this, he stopped building Ramah and ceased his work.
\verse Then King Asa took all of Judah, and they carried away the stones of Ramah and its timber with which Baasha had been building, and he built Geba and Mizpah with them.
\verse And at that time Hanani the seer came to Asa king of Judah and said to him, “On account of your reliance upon the king of Aram, and since you did not rely on Adonai your God, therefore the troops of the king of Aram escaped from your hand.
\verse Were not the Cushites and Libyans a mighty army with very abundant chariots and horsemen? And because of your reliance on Adonai he gave them into your hand.
\verse For the eyes of Adonai roam throughout all the earth to strengthen those whose heart is fully devoted to him. You have been foolish in this, for from now on you will have wars.”
\verse Then Asa was angry with the seer and put him in \textit{the prison}, for he was enraged with him concerning this. And Asa oppressed some of the people at that time.
\verse Now behold, the words of Asa from the first to the last, behold, they are written in the scroll of the kings of Judah and Israel.
\verse And in the thirty-ninth year of his reign, he fell \textit{severely} ill in his feet. But even in his illness he did not seek Adonai, but only among the healers.
\verse And Asa slept with his ancestors. And he died in the forty-first year of his reign.
\verse And they buried him in his burial site, which had been cut out for him in the city of David. And they laid him on the bier which they had filled with all kinds of spices made by the perfumers as a fragment ointment. And they made a great fire in his honor.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 17
\verseWithHeading{Jehoshaphat Reigns in Judah} And Jehoshaphat his son reigned in his place, and he strengthened himself against Israel.
\verse And he put troops in all the fortified cities of Judah, and put garrisons in the land of Judah and in the cities of Ephraim that Asa his father had taken.
\verse And Adonai was with Jehoshaphat, because he walked in the former ways of David his father and did not seek after the Baals,
\verse but he sought after the God of his ancestors and walked in his commandments and not according to the works of Israel.
\verse So Adonai established the kingdom in his hand. And all Judah gave tribute to Jehoshaphat, and he had much wealth and honor.
\verse And his heart was courageous in the ways of Adonai. Moreover, he removed the high places and the Asherahs from Judah.
\verse In the third year of his reign he sent his officials, Ben-Hail, Obadiah, Zechariah, Nethanel, and Micaiah, to teach in the cities of Judah;
\verse and with them the Levites Shemaiah, Nethaniah, Zebadiah, Asahel, Shemiramoth, Jehonathan, Adonijah, Tobijah, and Tob-Adonijah; and with them the priests Elisham and Joram.
\verse And they taught in Judah, and the scroll of the law of Adonai was with them when they went around in all the cities of Judah, and they taught the people.
\verse And the fear of Adonai was upon all the kingdoms of the lands surrounding Judah, and they did not make war against Jehoshaphat.
\verse And some of the Philistines brought a gift and silver to Jehoshaphat as tribute. The Arabians also brought him seven thousand seven hundred sheep and seven thousand seven hundred goats.
\verse And Jehoshaphat \textit{grew greater and greater}, and he built fortresses and storage cities in Judah.
\verse And he had many supplies in the cities of Judah and mighty warriors for battle in Jerusalem.
\verse Now these were their enrollment by the house of their ancestors: of Judah, the commanders of thousands: Adnah, the commander, and with him were three hundred thousand mighty armed warriors;
\verse and at his side, Jehohanan the commander, and with him were two hundred and eighty thousand;
\verse and at his side, Amasiah the son of Zicri, who had made a freewill offering to Adonai, and with him were two hundred thousand mighty armed warriors.
\verse And from Benjamin: Eliada, a powerful mighty warrior, and with him were two hundred thousand armed with bow and shield;
\verse and at his side, Jehozabad, and with him were one hundred and eighty thousand armed for war.
\verse These were serving the king, besides those whom the king had placed in the fortified cities throughout all Judah.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 18
\verseWithHeading{Jehoshaphat’s Alliance with Ahab} Now Jehoshaphat had much wealth and honor, and he became son-in-law to Ahab.
\verse And \textit{after some years} he went down to Ahab in Samaria. And Ahab slaughtered many sheep and cattle for him and for the people who were with him, and urged him to go up against Ramoth-Gilead.
\verse And Ahab the king of Israel said to Jehoshaphat, king of Judah, “Will you go up with me to Ramoth-Gilead?” And he answered him, “I am as you are, and my people are as your people with you in war.”
\verse Then Jehoshaphat said to the king of Israel, “Seek \textit{first of all} the word of Adonai.”
\verse Then the king of Israel gathered the prophets, four hundred men. And he said to them, “Shall we go to Ramoth-Gilead to battle, or shall we cease?” And they said, “Go up, since God will give it into the hand of the king.”
\verse Then Jehoshaphat said, “Is there not here another prophet of Adonai that we might inquire of him?”
\verse Then the king of Israel said to Jehoshaphat, “There is yet one man by whom to seek Adonai, but I hate him, because he never prophesies good concerning me, but \textit{always} disaster. He is Micaiah the son of Imlah.” Then Jehoshaphat said, “Let not the king say thus.”
\verse Then the king of Israel called to a eunuch and said, “Quickly bring Micaiah the son of Imlah.”
\verse Now the king of Israel and Jehoshaphat the king of Judah were each sitting on his throne, clothed in robes. And they were sitting at the threshing floor at the entrance of the gate of Samaria, and all the prophets were prophesying before them.
\verse And Zedekiah the son of Kenaanah made for himself horns of iron, and he said, “Thus says Adonai: ‘With these you will gore Aram to their destruction.’ ”
\verse And all the prophets prophesied thus, saying, “Go up to Ramoth-Gilead and triumph! Adonai will give it into the hand of the king.”
\verse Then the messenger who went to call Micaiah said to him, “Behold, the words of the prophets are good with one voice to the king. Please let your word be as one with them and speak good.”
\verse But Micaiah said, “As Adonai lives, only what my God has said, that will I speak.”
\verse And when he had come to the king, the king said to him, “Micaiah, shall we go to Ramoth-Gilead to war or shall I cease?” And he said, “Go up and triumph! They shall be given into your hand.”
\verse But the king said to him, “How many times shall I make you swear that you speak nothing except the truth in the name of Adonai?”
\verse Then he said, “I saw all Israel scattered upon the mountains like sheep that had no shepherd. And Adonai said, ‘These have no masters; let them return to his own house in peace.’ ”
\verse Then the king of Israel said to Jehoshaphat, “Did I not say to you he would not prophesy good concerning me, but only disaster?”
\verse Then Micaiah said, “Therefore hear the word of Adonai: I saw Adonai sitting upon his throne with all the host of heaven standing on his right and on his left.
\verse And Adonai said, ‘Who will entice Ahab the king of Israel that he would go up and fall at Ramoth-Gilead?’ And one said this, and another said that.
\verse Then a spirit came forth and stood before Adonai and said, ‘I will entice him.’ Then Adonai said to him, ‘By what means?’
\verse Then he said, ‘I will go forth and will be a lying spirit in the mouth of all his prophets.’ And he said, ‘You will entice him and will also succeed. Go out and do so.’
\verse So now, behold, Adonai has put a spirit of deception into the mouths of these your prophets. Adonai has spoken disaster against you.”
\verse Then Zedekiah the son of Kenaanah came near and struck Micaiah on the cheek and said, “\textit{Which way} did the Spirit of Adonai come from me to speak to you?”
\verse Then Micaiah said, “Behold, you will see on that day when you go into a private room to hide yourself.”
\verse Then the king of Israel said, “Take Micaiah and return him to Amon the commander of the city and to Jehoash the son of the king,
\verse and say, ‘Thus says the king: “Put this one into the prison house, and let him eat \textit{a meager ration of bread and water} until I return in peace.” ’ ”
\verse And Micaiah said, “If you surely return in peace Adonai has not spoken with me.” Then he said, “Hear, peoples, all of them!”
\verseWithHeading{The Death of Ahab} Then the king of Israel and Jehoshaphat king of Judah went up to Ramoth-Gilead.
\verse And the king of Israel said to Jehoshaphat, “I will disguise myself and go into battle, but you, wear your garments.” So the king of Israel disguised himself, and they went to war.
\verse Now the king of Aram had commanded the commanders of his chariots, saying, “Do not fight with the small or the great, but only with the king of Israel.”
\verse And it happened that when the commanders of the chariots saw Jehoshaphat, then they said, “It is the king of Israel!” And they turned against him to fight. Then Jehoshaphat cried out and Adonai helped him. God drove them away from him.
\verse And it happened that when the commanders of the chariots saw that it was not the king of Israel, they turned back from following him.
\verse Now a certain man drew the bow \textit{at random} and struck the king of Israel between the soldering of his breastplate armor. Then he said to the chariot driver, “\textit{Turn around}; get me away from the war camp, for I am wounded.”
\verse And the battle went on that day, and the king of Israel was propped up in the chariot facing Aram until the evening. And he died at the time of the going down of the sun.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 19
\verseWithHeading{Jehoshaphat’s Reforms} Then Jehoshaphat king of Judah returned to his house in peace to Jerusalem.
\verse And Jehu son of Hanani the seer went out to meet him. And he said to King Jehoshaphat, “Should you be a help to the wicked and love those who hate Adonai? On account of this, wrath has come against you from Adonai.
\verse Nevertheless some good things have been found in regard to you, for you burned the Asherahs out of the land and set your heart to seek God.”
\verse And Jehoshaphat lived in Jerusalem and returned and went out among the people from Beersheba to the hill country of Ephraim and brought them back to Adonai, the God of their ancestors.
\verse And he appointed judges in the land and in all the fortified cities of Judah, city by city.
\verse And he said to the judges, “Consider what you are doing, for you are not judging for a man but for Adonai. Now he is with you in speaking justice.
\verse So now, let the fear of Adonai be upon you all. \textit{Be careful what you do}, for there is neither wickedness nor \textit{showing partiality} and taking bribes with Adonai our God.”
\verse Moreover, Jehoshaphat appointed in Jerusalem some of the Levites and the priests and heads of the \textit{families} of Israel as judges and to decide legal disputes for Adonai. And they sat in Jerusalem.
\verse And he commanded them, saying, “Thus you must do in the fear of Adonai in faithfulness and with a whole heart.
\verse Any legal dispute that comes before you from your brothers who live in their cities concerning bloodguilt, law, commandment, statutes, and justice, you must warn them, so that they do not incur guilt before Adonai, and that wrath may not come upon you and your brothers. Thus you must do that you do not incur guilt.
\verse And behold, Amariah the chief priest is over you in all matters of the word of Adonai, and Zebadiah the son of Ishmael, the governor of the house of Judah, in all matters of the king, and the Levites shall be before you as officials. Be strong and do well! May Adonai be with the upright.”
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 20
\verseWithHeading{The Prayer of Jehoshaphat} And it happened that afterward, the \textit{Moabites}, the \textit{Ammonites}, and some of the Meunites came against Jehoshaphat for war.
\verse And they came and reported it to Jehoshaphat, saying, “A great multitude from beyond the sea, from Aram, is coming against you. Now behold, they are in Hazazon Tamar” (that is, En Gedi).
\verse Then Jehoshaphat was afraid and set his face to seek Adonai. And he called for a fast through all Judah.
\verse And Judah assembled to seek after Adonai; even from all the cities of Judah they came to seek Adonai.
\verse Then Jehoshaphat stood in the assembly of Judah and Jerusalem in the house of Adonai before the new courtyard.
\verse And he said, “O Adonai, God of our ancestors, are you not God in heaven? Now you rule in all the kingdoms of the nations, and in your hand are power and might, and there is none who can resist against you.
\verse O, our God, did you yourself not drive out the inhabitants of this land before your people Israel and give it to the descendants of Abraham your friend forever?
\verse And they lived in it and built for you a sanctuary for your name in it, saying,
\verse ‘If disaster, a sword, punishment, disease, or famine come upon us, we shall stand before this house and before you—for your name is in this house—and cry out to you out of our trouble, then you will hear and save us.’
\verse And now, look, the people of Moab and Ammon and Mount Seir whom you did not allow Israel to come against when they came from Egypt, from whom they turned aside and did not destroy,
\verse now behold, they are rewarding us by coming to dispossess us from your possession which you gave us as an inheritance.
\verse O our God, will you not judge them, for there is no power in us before this great multitude that is coming against us. Now we do not know what we shall do, for our eyes are upon you.”
\verse And all Judah was standing before Adonai, and also their little children, their wives, and their children.
\verse Then the Spirit of Adonai came upon Jahaziel, the son of Zechariah, the son of Benaiah, the son of Jeiel, the son of Mattaniah, the Levite from the descendants of Asaph, in the midst of the assembly.
\verse And he said, “Listen, all Judah and inhabitants of Jerusalem and King Jehoshaphat, thus says Adonai to you: ‘Do not be afraid and do not be dismayed before this great multitude, for the battle is not yours, but God’s.
\verse Tomorrow go down against them. Behold, they are coming up the ascent of Ziz. And you will find them at the end of the valley facing the desert of Jeruel.
\verse There will be no need for you to fight in this battle. Take your stand! Stand and see the deliverance of Adonai among you, O Judah and Jerusalem! Do not fear and do not be dismayed. Tomorrow go out before them and Adonai will be with you.’ ”
\verse Then Jehoshaphat bowed his face down to the ground, and all Judah and the inhabitants of Jerusalem fell before Adonai to worship Adonai.
\verse Then the Levites, from the \textit{Kohathites} and from the \textit{Korahites} stood up to praise Adonai the God of Israel with an exceedingly loud voice.
\verseWithHeading{God Delivers Judah and Jerusalem} And they rose up early in the morning and went out to the desert of Tekoa. And when they went out, Jehoshaphat stood and said, “Hear me, O Judah and inhabitants of Jerusalem! Believe in Adonai your God, and you shall be established! Believe in his prophets, and you shall succeed!”
\verse And when he had taken counsel with the people, he appointed singers to Adonai who were giving praise to the splendor of his holiness when they went out before the army. And they said, “Give thanks to Adonai, for his loyal love is everlasting!”
\verse And \textit{when} they began with singing and praise, Adonai set ambushes against the people of Ammon, Moab, and Mount Seir who were coming against Judah. And they were defeated
\verse when the people of Ammon and Moab rose up against the inhabitants of Mount Seir to destroy and demolish them. And when the inhabitants of Seir had made an end of them, each helped to destroy his comrade.
\verse And Judah came against the watchtower of the wilderness. And they looked at the crowd, and behold, corpses were lying on the ground. There was none who had escaped.
\verse When Jehoshaphat came with his people to plunder their booty, they found among them in abundance possessions, corpses, and precious items. And they took plunder for themselves \textit{until they could carry no more}. And they were three days in plundering the booty, for it was abundant.
\verse And on the fourth day, they assembled in the Valley of Berakah, for there they blessed Adonai. Therefore the name of that place is called the Valley of Berakah up to this day.
\verse And all the men of Judah and Jerusalem returned, and Jehoshaphat was at their head in returning to Jerusalem with joy, for Adonai caused them to rejoice over their enemies.
\verse And they came to Jerusalem with harps, stringed instruments, and trumpets to the house of Adonai.
\verse And the fear of God came upon all the kingdoms of the earth when they heard that Adonai had fought against the enemies of Israel.
\verse So the kingdom of Jehoshaphat was at rest, since his God gave rest to him all around.
\verseWithHeading{The End of Jehoshaphat’s Reign} So Jehoshaphat reigned over Judah. He was thirty-five years old when he began to reign. And he reigned twenty-five years in Jerusalem. The name of his mother was Azubah, the daughter of Shilhi.
\verse And he walked in the way of his father Asa and did not turn aside from it, to do what was right in the eyes of Adonai.
\verse Only the high places were not removed. The people still had not yet set their heart on the God of their ancestors.
\verse Now the rest of the words of Jehoshaphat, from the first to the last, behold, they are written in the chronicles of Jehu the son of Hanani, which are recorded in the scroll of the kings of Israel.
\verse And afterward Jehoshaphat the king of Judah joined with Ahaziah the king of Israel, who acted wickedly.
\verse He joined with him to build ships to go to Tarshish. And they built the ships in Ezion Geber.
\verse Then Eliezer the son of Dodavahu of Mareshah prophesied against Jehoshaphat, saying, “Because of your joining with Ahaziah, Adonai will break down your works.” And the ships were destroyed and were not able to go to Tarshish.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 21
\verseWithHeading{The Reign of Jehoram} And Jehoshaphat slept with his ancestors and was buried with his ancestors in the city of David, and Jehoram his son reigned in his place.
\verse Now he had brothers, the sons of Jehoshaphat: Azariah, Jehiel, Zechariah, Azariah, Michael, and Shephatiah. All these were the sons of Jehoshaphat, king of Israel.
\verse And their father gave to them many gifts of silver, gold, and valuables, with fortified cities in Judah, but he gave the kingdom to Jehoram, for he was the firstborn.
\verse When Jehoram ascended to the kingdom of his father, he strengthened himself and murdered all his brothers with the sword, and even some of the princes of Israel.
\verse Jehoram was thirty-two years old when he became king, and he reigned eight years in Jerusalem.
\verse And he walked in the way of the kings of Israel as the house of Ahab had done, for the daughter of Ahab was his wife. And he did evil in the sight of Adonai.
\verse But Adonai was not willing to destroy the house of David on account of the covenant that he had \textit{made} with David and since he had promised to give a lamp to him and to his descendants \textit{forever}.
\verse In his days Edom revolted from under the hand of Judah, and they set up a king of their own.
\verse Then Jehoram passed over with his commanders and all the chariots with him, and when night came he struck Edom who was all around him and the commanders of his chariots.
\verse So Edom revolted from under the hand of Judah until this day. Then Libnah also revolted at that time from under his hand, because he had forsaken Adonai, the God of his ancestors.
\verse Moreover, he made high places in the hill country of Judah, and he enticed the inhabitants of Jerusalem to be unfaithful, and he led Judah astray.
\verse And a letter from Elijah the prophet came to him, saying, “Thus says Adonai, the God of David your father: ‘Because you have not walked in the ways of Jehoshaphat your father or in the ways of Asa, the king of Judah,
\verse but have walked in the way of the kings of Israel and have enticed Judah and the inhabitants of Jerusalem to be unfaithful like the unfaithfulness of the house of Ahab, and have also murdered your brothers of the house of your father who were better than you,
\verse behold, Adonai is inflicting a great plague on your people, your children, your wives, and all your possessions,
\verse and you yourself will be afflicted with great illness, with sickness in your bowels, until your bowels come out on account of the illness, day by day.’ ”
\verse And Adonai stirred up the spirit of the Philistines and the Arabs who were \textit{near} the Cushites against Jehoram.
\verse And they came up against Judah and invaded it and carried away all the possessions found in the house of the king, and also his sons and his wives, so that no son was left to him except Jehoahaz his youngest son.
\verse And after all this Adonai afflicted him in his bowels with an illness for which there was no cure.
\verse And it happened that \textit{after many days, at the end of two years}, his bowels came out because of his illness, and he died in terrible agony. And his people did not make a fire for him like the fire for his ancestors.
\verse He was thirty-two years old when he began to reign, and he reigned eight years in Jerusalem. And he departed with no one’s regret. And they buried him in the city of David, but not in the burial sites of the kings.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 22
\verseWithHeading{The Reign of Ahaziah} And the inhabitants of Jerusalem made Ahaziah his youngest son king in his place, for the band of men who had come with the Arabs to the camp had murdered all the older sons. And Ahaziah the son of Jehoram, the king of Judah, reigned.
\verse Ahaziah was forty-two years old when he became king, and he reigned one year in Jerusalem. And the name of his mother was Athaliah, the daughter of Omri.
\verse He also walked in the ways of the house of Ahab, for his mother was his counselor for doing wickedness.
\verse And he did evil in the eyes of Adonai as the house of Ahab had done, for they were his counselors to his destruction after the death of his father.
\verse He also walked in their counsel and went with Jehoram the son of Ahab, king of Israel, to war against Hazael king of Aram at Ramoth-Gilead. And the Arameans wounded Joram.
\verse And he returned to be healed in Jezreel because of the wounds he had suffered at Ramah, when he made war against Hazael king of Aram. And Azariah the son of Jehoram, king of Judah, went down to see Jehoram son of Ahab in Jezreel for he was sick.
\verse But the downfall of Ahaziah was from God, and intended to come to Joram. And when he came, he went out with Jehoram to Jehu the son of Nimshi, whom Adonai had anointed to destroy the house of Ahab.
\verse And it happened that when Jehu was judging the house of Ahab, he found the princes of Judah and the sons of the brothers of Ahaziah who ministered to Ahaziah, and he killed them.
\verse And he searched for Ahaziah and captured him. And he was hiding in Samaria. And they brought him to Jehu and put him to death. And they buried him, for they said, “He is the descendant of Jehoshaphat, who sought Adonai with his whole heart.” And the house of Ahab had no one to exercise power over the kingdom.
\verseWithHeading{The Reign of Athaliah} Now when Athaliah the mother of Ahaziah saw that her son was dead, she arose and murdered all the \textit{royal descendants} of the house of Judah.
\verse But Jehosheba the daughter of the king took Joash, the son of Ahaziah, and kidnapped him away from the sons of the king who were to be put to death. And she put him and his nurse in a private bedroom. So Jehosheba the daughter of King Jehoram, the wife of Jehoiada the priest, because she was a sister of Ahaziah, hid him from before Athaliah so that she did not put him to death.
\verse And he was hidden with them in the house of God six years, but Athaliah reigned over the land.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 23
\verseWithHeading{Joash Is Made King} And in the seventh year Jehoiada strengthened himself and took with him into a covenant relationship the commanders of hundreds: Azariah the son of Jeroham, Ishmael the son of Jehohanan, Azariah the son of Obed, Maaseiah the son of Adaiah, and Elishaphat the son of Zicri.
\verse And they went around in Judah and gathered the Levites from all the cities of Judah, and the heads of the \textit{families} of Israel, and they came to Jerusalem.
\verse And all the assembly \textit{made} a covenant with the king in the house of God. And he said to them, “Behold, the son of the king shall reign as Adonai has spoken concerning the sons of David.
\verse This is the thing that you must do: one third of you priests and Levites \textit{coming off duty on the Sabbath} shall be gatekeepers at the entrances,
\verse and one third at the house of the king, and one third at the Gate of the Foundation. And all the people shall be in the courtyards of the house of Adonai.
\verse Let no one enter into the house of Adonai except the priests and the ministering Levites. They themselves may enter, for they are holy, but all the people shall keep the requirements of Adonai.
\verse And the Levites shall surround the king all around, each with his weapons in his hand. And whoever comes into the house shall be put to death. You shall be with the king when he comes in and when he goes out.”
\verse And the Levites and all Judah did according to all that Jehoiada the priest commanded. And they each took his men who were \textit{coming off duty on the Sabbath} with \textit{the ones going out to duty on the Sabbath}, for Jehoiada the priest did not dismiss the working groups.
\verse And Jehoiada the priest gave the spears and small shields that had belonged to King David that were in the house of God to the commanders of hundreds.
\verse And he appointed all the people, each with his weapon in his hand from the south side of the house to the north side of the house around the altar and the house.
\verse Then they brought out the son of the king and set upon him the crown and gave him the statute and made him king. And Jehoiada and his sons anointed him and said, “\textit{Long live the king}!”
\verse When Athaliah heard the sound of the people running and praising the king, she came to the people at the house of Adonai.
\verse And she looked and, behold, the king was standing by his column at the entrance, and the commanders and trumpeters beside the king, along with all the people of the land, rejoicing and blowing with trumpets, and the singers with the instruments of song leading the praise. And Athaliah tore her garments and cried, “Conspiracy! Conspiracy!”
\verse Then Jehoiada the priest brought out the commanders of the hundreds appointed over the troops, and he said to them, “Bring her out between the ranks, and whoever goes after her shall be put to death with the sword.” For the priest had said, “You must not kill her in the house of Adonai.”
\verse And they laid hands on her, and she went into the entrance of the Horse Gate of the house of the king, and they killed her there.
\verse And Jehoiada \textit{made} a covenant between himself and all the people and the king, that they should be Adonai’s people.
\verse Then all the people went to the house of Baal and broke it down. And they smashed his altars and his images, and they killed Mattan the priest of Baal before the altars.
\verse And Jehoiada placed appointees at the house of Adonai under the hand of the Levitical priests whom David had allotted to the house of Adonai to offer burnt offerings to Adonai as was written in the law of Moses, with joy and with song, according to the \textit{order of David}.
\verse And he set the gatekeepers at the gates of the house of Adonai so that no person unclean with respect to any matter could enter.
\verse And he took the commanders of hundreds, the noblemen, the governors over the people, and all the people of the land, and they brought the king down from the house of Adonai. And they came through the upper gate to the house of the king, and they set the king upon the throne of the kingdom.
\verse So all the people of the land rejoiced, and the city was quiet after Athaliah had been killed with the sword.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 24
\verseWithHeading{Joash Repairs the Temple} Joash was seven years old when he became king, and he reigned forty years in Jerusalem. And the name of his mother was Zibiah from Beersheba.
\verse And Joash did what was right in the eyes of Adonai all the days of Jehoiada the priest.
\verse And Jehoiada took two wives for him, and he fathered sons and daughters.
\verse And afterward it was in the heart of Joash to repair the house of Adonai.
\verse So he gathered the priests and the Levites and said to them, “Go out to the cities of Judah and gather money from all Israel to strengthen the house of your God sufficiently year by year. Now you yourselves must hasten to the matter.” But the Levites did not act with haste.
\verse So the king called Jehoiada the chief, and he said to him, “Why have you not required the Levites to bring from Judah and Jerusalem the tax of Moses, the servant of Adonai, and of the assembly of Israel for the tent of the testimony?”
\verse For the sons of the wicked Athaliah had broken into the house of God and had used all the holy vessels of the house of Adonai for the Baals.
\verse Then the king commanded, and they made a chest and put it at the gate outside the house of Adonai.
\verse And \textit{they issued a proclamation} in Judah and in Jerusalem to bring to Adonai the tax that Moses the servant of God had levied upon Israel in the wilderness.
\verse And all the princes and all the people rejoiced, and they brought their tax and threw it into the chest until it was finished.
\verse And whenever he brought the chest to the appointee of the king by the hand of the Levites, and when they saw that there was much money in it, then the secretary of the king and the officer of the chief priest came and emptied the chest, then they took it and returned it to its place. Thus they did day by day and gathered money in abundance.
\verse And the king and Jehoiada gave it to those who did the work of the service of the house of Adonai. And they hired stonemasons and skilled craftsmen to restore the house of Adonai, and also workers in iron and bronze to repair the house of Adonai.
\verse So those doing the work labored, and the restoration for the work made progress under their hand. And they restored the house of God to its position and strengthened it.
\verse And when they had finished, they brought the remainder of the money before the king and Jehoiada, and they used it for objects for the house of Adonai, objects for the service and the burnt offerings, dishes, and objects of gold and silver. And they were offering burnt offerings in the house of Adonai regularly, all the days of Jehoiada.
\verse And Jehoiada grew old and full of days, and he died; he was one hundred and thirty years old at his death.
\verse And they buried him in the city of David with the kings, for he did good in Israel and with respect to God and his house.
\verse Now after the death of Jehoiada, the princes of Judah came and bowed down to the king. Then the king listened to them.
\verse And they forsook the house of Adonai, the God of their ancestors and served the Asherahs and the idols. And wrath came upon Judah and Jerusalem on account of this guilt.
\verse But he sent prophets among them, to bring them back to Adonai. And they testified against them, and they did not obey.
\verse Then the Spirit of God clothed Zechariah the son of Jehoiada the priest, and he stood above the people and said to them, “Thus says God: ‘Why are you transgressing the commandments of Adonai so that you will not succeed? For you have forsaken Adonai, so he will forsake you.’ ”
\verse Then they conspired against him and stoned him with stones at the command of the king in the courtyard of the house of Adonai.
\verse So King Joash did not remember the loyal love that Jehoiada had shown to him, and he killed his son. And as he was dying he said, “May Adonai see and avenge!”
\verseWithHeading{The Death of Joash} And it happened at the turn of the year that the army of Aram went up against him, and they came to Judah and Jerusalem and destroyed all the commanders of the people from among the people, and all the war booty they sent to the king of Damascus.
\verse Though the army of Aram came with few men, Adonai gave a very large army into their hand, for they had forsaken Adonai, the God of their ancestors. And they inflicted punishment on Joash.
\verse And when they were going away from him (for they had left him with many wounds), his servants conspired against him on account of the blood of the sons of Jehoiada the priest. So they killed him on his bed and he died. And they buried him in the city of David, but they did not bury him at the burial sites of the kings.
\verse Now these are the ones who conspired against him: Zabad the son of Shimeath, the Ammonite, and Jehozabad the son of Shimrith, the Moabite.
\verse And as for his sons, the many oracles against him, and his repair of the foundation of the house of God, behold, they are written in the story of the scroll of the kings. And Amaziah his son reigned in his place.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 25
\verseWithHeading{The Reign of Amaziah} Amaziah was twenty-five years old when he became king. And he reigned twenty-nine years in Jerusalem, and the name of his mother was Jehoaddan from Jerusalem.
\verse And he did what was right in the eyes of Adonai, only not with a fully-devoted heart.
\verse And it happened that as his reign was strengthened, he killed his servants who had killed his father the king.
\verse But he did not put their children to death, according to what is written in the law, in the scroll of Moses, where Adonai commanded, saying, “Fathers shall not be put to death on account of the sons, and the sons shall not be put to death on account of the fathers, but each shall be put to death for his own sin.”
\verse Then Amaziah assembled the men of Judah and set them by \textit{families} under commanders of thousands and commanders of hundreds for all Judah and Benjamin. And he counted them from twenty years old and upward, and found them to be three hundred thousand chosen for going out to battle, able with spear and shield.
\verse And he hired from Israel one hundred thousand strong, mighty warriors for one hundred talents.
\verse But a man of God came to him, saying, “O king, do not let the army of Israel go with you, for Adonai is not with Israel, all the Ephraimites.
\verse But even still, you yourself go, do! Be strong for the battle else God will make you stumble before the enemy! For there is power with God to help and to cause stumbling.”
\verse And Amaziah said to the man of God, “Now what should I do with the one hundred talents that I have given to the troops of Israel?” And the man of God said, “\textit{Adonai is able to give} to you more than this.”
\verse Then Amaziah dismissed the troops that came to him from Ephraim to go to their home. And \textit{they became very angry} with Judah, and they returned to their home \textit{in great anger}.
\verse Then Amaziah strengthened himself, and leading his army, he went to the Valley of Salt. And he struck down ten thousand men of Seir.
\verse Now the \textit{troops} captured ten thousand others alive, and they brought them to the top of the rock and threw them from the top of the rock. So all of them were smashed to pieces.
\verse As for the \textit{troops} Amaziah had sent back from going with him to war, they raided the cities of Judah from Samaria to Beth-Horon. And they struck down three thousand of them and plundered much booty.
\verse And it happened that after Amaziah came back from killing the Edomites, he brought the gods of the sons of Seir, and he stood them up to himself as gods and bowed down before them and make smoke offerings for them.
\verse So \textit{Adonai became very angry} with Amaziah and sent a prophet to him, and he said to him, “Why have you sought the gods of the people who could not deliver their own people from your hand?”
\verse And it happened that while he was speaking to him, he said to him, “Have we appointed you as a counselor to the king? Stop—\textit{why should you be killed}?” So the prophet stopped and said, “I know that God has determined to destroy you, because you have done this and have not listened to my advice.”
\verse Then Amaziah king of Judah took counsel, and he sent to Joash the son of Jehoahaz the son of Jehu, king of Israel, saying, “Come, \textit{let us face one another}.
\verse And Joash the king of Israel sent to Amaziah king of Judah, saying, “The thorn bush that is in Lebanon has sent to the cedar which is in Lebanon, saying, ‘Give your daughter to my son as wife.’ But a wild animal of the field that was in Lebanon passed by and trampled the thorn bush.
\verse Look, you say that you have struck Edom, and your heart has lifted you up to boast. Now remain at your home. Why stir up disaster that you fall, you and Judah with you?”
\verse But Amaziah did not listen, for it was from God that he might give them into the hand of enemies, because they sought the gods of Edom.
\verse So Joash the king of Israel went up, and he and Amaziah the king of Judah \textit{faced one another} at Beth-Shemesh, which belongs to Judah.
\verse And Judah was defeated before Israel, and each man fled to his tent.
\verse And Joash the king of Israel captured Amaziah the king of Judah, the son of Joash, the son of Jehoahaz, at Beth-Shemesh. And they brought him to Jerusalem and broke down the wall of Jerusalem from the Gate of Ephraim to the Corner Gate, four hundred cubits.
\verse Then with all the gold and silver, all the objects found in the house of God with Obed-Edom, the storehouses of the house of the king, and \textit{the hostages}, he returned to Samaria.
\verse And Amaziah the son of Joash, king of Judah, lived after the death of Joash the son of Jehoahaz, the king of Israel, fifteen years.
\verse Now the remainder of the words of Amaziah, from the first to the last, behold, are they not written in the scroll of the kings of Judah and Israel?
\verse And from the time that Amaziah turned away from Adonai, they had plotted a conspiracy against him in Jerusalem. And he fled to Lachish, but they sent after him to Lachish and killed him there.
\verse And they carried him on the horses and buried him with his ancestors in the city of Judah.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 26
\verseWithHeading{The Reign of Uzziah} Then all the people of Judah took Uzziah (now he was sixteen years old) and made him king in the place of his father Amaziah.
\verse He built Elath and returned it to Judah after the king slept with his ancestors.
\verse Uzziah was sixteen years old when he became king, and he reigned in Jerusalem fifty-two years. And the name of his mother was Yecoliah of Jerusalem.
\verse And he did what was right in the eyes of Adonai, according to all that Amaziah his father had done.
\verse And he began to seek God in the days of Zechariah who was teaching in visions of God. And \textit{whenever} he sought Adonai made him have success.
\verse And he went out and made war against the Philistines, and he broke down the wall of Gath and the wall of Jabneh and the wall of Ashdod, and he built cities in Ashdod and among the Philistines.
\verse And God helped him against the Philistines and against the Arabs who were living in Gur-Baal and Meunim.
\verse And the Ammonites gave tribute to Uzziah, and his fame went out to the boundary of Egypt, for he became \textit{very strong}.
\verse And Uzziah built towers in Jerusalem at the Corner Gate, at the Valley Gate, and at the Angle, and he strengthened them.
\verse And he built towers in the desert and hewed many cisterns, for he had much livestock, both in the Shephelah and on the plain, along with farmers and vinedressers in the hills and in the fertile lands, for he loved the soil.
\verse And Uzziah had troops \textit{trained and ready for war} in divisions according to the number of their enrollment at the hand of Jeuel the scribe and Maaseiah the officer, under the hand of Hananiah of the king’s commanders.
\verse The whole number of the heads of the \textit{families} for mighty warriors of strength was two thousand six hundred.
\verse And under their hand were army troops numbering three hundred and seven thousand five hundred who could make war with power and strength to help the king against an enemy.
\verse And Uzziah prepared small shields, spears, helmets, breastplates, bows, and slingstones for the whole army.
\verse And he made siege machines in Jerusalem designed by skillful men to be set upon the towers and upon the corners to shoot arrows and large slingstones. And his fame went out far, for he helped marvelously, for he was strong.
\verse But on account of his strength his heart grew proud unto destruction. And he acted unfaithfully against Adonai his God and went into the temple of Adonai to offer an incense offering on the altar.
\verse And Azariah the priest, along with eighty strong priests with him, went in after him.
\verse And they stood against King Uzziah and said to him, “It is not for you, Uzziah, to burn incense to Adonai, but it is for the priests, the descendants of Aaron, who are consecrated to burn incense. Go out of the sanctuary, for you have acted unfaithfully. There will be no honor for you from Adonai.”
\verse Then Uzziah, his censer in hand to burn incense, became angry. And when he became angry with the priests, then leprosy appeared on his forehead in front of the priests in the house of Adonai at the altar of incense.
\verse When Azariah the chief priest and all the priests turned to him, behold, he was leprous in the forehead. So they rushed him away from there, and he also hastened to go out, for Adonai had smitten him.
\verse So King Uzziah was leprous until the day of his death, and he lived in a separate house as a leper, for he was cut off from the house of Adonai. And Jotham his son was over the house of the king, judging the people of the land.
\verse Now the remainder of the words of Uzziah, from the first to the last, Isaiah the prophet, the son of Amoz, wrote them.
\verse And Uzziah slept with his ancestors, and they buried him with his ancestors in the burial field which belonged to the kings, because, they said, “He was leprous.” And Jotham his son reigned in his place.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 27
\verseWithHeading{The Reign of Jotham} Jotham was twenty-five years old when he became king, and he reigned sixteen years in Jerusalem. And the name of his mother was Jerusha the daughter of Zadok.
\verse And he did what was right in the eyes of Adonai, according to all that his father Uzziah had done—only he did not go into the temple of Adonai. But the people still acted corruptly.
\verse He built the upper gate of the house of Adonai, and \textit{he did much restoration} on the wall of Ophel.
\verse And he built cities in the hill country of Judah, and he built forts and towers in the wooded places.
\verse Now he fought with the king of the \textit{Ammonites} and prevailed against them. And the \textit{Ammonites} gave to him in that year one hundred talents of silver and ten thousand measures of wheat and ten thousand measures of barley. This is what the \textit{Ammonites} rendered to him, as well as in the second and third year.
\verse And Jotham strengthened himself, for he established his ways before Adonai his God.
\verse Now the remainder of the words of Jotham and all his wars and his ways, behold, they are written upon the scroll of the kings of Israel and Judah.
\verse He was twenty-five years old when he became king, and he reigned sixteen years in Jerusalem.
\verse And Jotham slept with his ancestors, and they buried him in the city of David. And his son Ahaz became king in his place.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 28
\verseWithHeading{The Reign of Ahaz} Ahaz was twenty years old when he became king, and he reigned sixteen years in Jerusalem. And he did not do what is right in the eyes of Adonai like David his ancestor.
\verse But he walked in the ways of the kings of Israel and also made molten idols for the Baals.
\verse And he himself burned incense in the valley of Ben-Hinnom and burned his sons in the fire, according to the detestable practices of the nations whom Adonai drove out before the Israelites.
\verse And he sacrificed and burned incense at the high places, on the hilltops, and under every flourishing tree.
\verse Then Adonai his God gave him into the hand of the king of Aram who defeated him and captured many captives from him and brought them to Damascus. Moreover, he was also given into the hand of the king of Israel, and he defeated him by a great attack.
\verse And Pekah, the son of Remaliah, killed in one day one hundred and twenty thousand in Judah, all men of strength, because they forsook Adonai the God of their ancestors.
\verse And Zikri, a mighty warrior of Ephraim, killed Maaseiah the son of the king, Azrikam the commander of the house, and Elkanah the second to the king.
\verse And the Israelites captured two hundred thousand of their brothers, women, sons, and daughters. And they also plundered much booty from them and brought the booty to Samaria.
\verse Now a prophet of Adonai was there; Oded was his name. And he went out before the army that was coming to Samaria and said to them, “Look, it is on account of the anger of Adonai, the God of your ancestors, against Judah that he gave them into your hands, and you killed them in a rage reaching to the heavens.
\verse So now, you yourselves plan to subdue people of Judah and Jerusalem as slaves and female servants for yourselves. \textit{Is there not also guilt on your part} against Adonai your God?
\verse So now, listen to me and return the captives of your brothers whom you have taken captive, for the fierce anger of Adonai is against you.”
\verse Then men from the heads of Ephraim—Azariah the son of Jehohanan, Berekiah the son of Meshillemoth, Hezekiah the son of Shallum, and Amasa the son of Hadlai—stood against those returning from the fight.
\verse And they said to them, “You shall not bring the captives here, for it is as guilt against Adonai upon us. You yourselves plan to add to our sins and to our guilt, but our guilt is very great, and there is great anger against Israel.”
\verse So the soldiers left the captives and the plunder before the commanders and all the assembly.
\verse Then the men designated by name arose and took the captives and from the plunder clothed all their nakedness. So they clothed them, gave them sandals, gave them food to eat, gave them water to drink, anointed them, and guided them with the donkeys provided for all those who stumbled, and brought them to Jericho, the city of the palm trees, next to their brothers. Then they returned to Samaria.
\verse At that time King Ahaz sent to the king of Assyria to help him,
\verse since again the Edomites came and struck against Judah and took captives.
\verse And the Philistines sent raids into the cities of the Shephelah and the Negeb of Judah. And they captured Beth-Shemesh, Aijalon, Gederoth, Socoh with its villages, Timnah with its villages, and Gimzo with its villages. And they settled there.
\verse For Adonai subdued Judah for the sake of Ahaz, king of Israel, for he created disorder in Judah and acted unfaithfully against Adonai.
\verse Then Tiglath-Pileser the king of Assyria came against him and brought trouble for him, \textit{instead of helping him},
\verse for Ahaz took a portion from the house of Adonai and the house of the king and the princes, and he gave it to the king of Assyria, but it was no help to him.
\verse And in the time of his distress, King Ahaz continued to act unfaithfully against Adonai.
\verse He sacrificed to the gods of Damascus that had defeated him and said, “Because the gods of the king of Aram helped them, I will sacrifice to them so that they will help me.” But they were \textit{a disaster} to him and to all Israel.
\verse Then Ahaz gathered the objects of the house of God, and he cut the objects of the house of God to pieces. And he shut the doors of the house of Adonai and made for himself altars in every corner of Jerusalem.
\verse And \textit{in every city} of Judah he made high places to burn incense to other gods, and he provoked Adonai, the God of his ancestors, to anger.
\verse Now the remainder of his words and all his ways, from the first to the last, behold, they are written upon the scroll of the kings of Judah and Israel.
\verse And Ahaz slept with his ancestors, and they buried him in the city, in Jerusalem, for they did not bring him to the burial site of the kings of Israel. And Hezekiah his son became king in his place.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 29
\verseWithHeading{The Reign of Hezekiah} Hezekiah became king when he was twenty-five years old, and he reigned twenty-nine years in Jerusalem. And the name of his mother was Abijah the daughter of Zechariah.
\verse And he did that which was right in the eyes of Adonai, according to all that David his father had done.
\verse In the first year of his kingship, in the first month, he opened the doors of the house of Adonai and strengthened them.
\verse And he brought in the priests and the Levites, and he gathered them into the eastern square.
\verse And he said to them, “Hear me, O Levites! Sanctify yourselves now, and sanctify the house of Adonai, the God of your ancestors, and bring out the defilement from the sanctuary.
\verse For our ancestors acted unfaithfully and did evil in the sight of Adonai our God, and have forsaken him and turned away their faces from the dwelling place of Adonai. \textit{They have turned their backs} on it.
\verse They also shut the doors of the portico and extinguished the lamps. And they did not burn incense or offer burnt offerings in the sanctuary to the God of Israel.
\verse And the wrath of Adonai was upon Judah and Jerusalem. And he made them as an abhorrence, as a desolation and an object of scorn, as you see with your own eyes.
\verse Now behold, our fathers have fallen by the sword, and our sons and our daughters and our wives are in captivity on account of this.
\verse Now \textit{it is in my heart} to \textit{make} a covenant with Adonai, the God of Israel, so that his fierce anger might turn away from him.
\verse Now, my sons, do not be negligent, for Adonai has chosen you to stand before him to serve him and to be his servants and incense burners.”
\verse Then the Levites arose—Mahath the son of Amasai and Joel the son of Azariah, from the Kohathites; and Kish the son of Abdi and Azariah the son of Jehallelel from the Merarites; and Joah son of Zimmah and Eden son of Joah, from the Gershonites.
\verse And from the descendants of Elizaphan: Shimri and Jeiel; and from the descendants of Asaph: Zechariah and Mattaniah;
\verse and from the descendants of Heman: Jeiel and Shimei; and from the descendants of Jeduthun: Shemaiah and Uzziel—
\verse and they gathered their brothers and sanctified themselves, and they went in according to the command of the king because of the words of Adonai, to purify the house of Adonai.
\verse And the priests went into the inner part of the house of Adonai to purify it. And they brought out all the impurity that they found in the temple of Adonai out to the courtyard of the house of Adonai. And the Levites took them to bring them outside to the Wadi Kidron.
\verse And they began to sanctify themselves on the first day of the first month. And on the eighth day of the month they came to the portico of Adonai. And they sanctified the house of Adonai for eight days, and on the sixteenth day of the first month they finished.
\verse And they went into the inner part to King Hezekiah and said, “We have cleansed all the house of Adonai: the altar of burnt offering, all its objects, the table of the rows of bread and all its objects,
\verse and all the objects that King Ahaz discarded in his reign when he sinned, we have prepared and sanctified. Look, they are before the altar of Adonai.”
\verse Then King Hezekiah rose early and gathered the officials of the city, and he went up to the house of Adonai.
\verse And they brought seven bulls, seven rams, seven lambs, and seven male goats as a purification offering for the kingdom, for the sanctuary, and for Judah. And he commanded the descendants of Aaron, the priests, to offer them on the altar of Adonai.
\verse And they slaughtered the cattle, and the priests received the blood and threw it toward the altar. Then they slaughtered the rams and threw the blood toward the altar, and then they slaughtered the lambs and threw the blood toward the altar.
\verse Then they brought the goats of the sin offering before the king and the assembly, and they laid hands upon them.
\verse And the priest killed them and offered their blood against the altar to make atonement for all Israel, for the king had commanded the burnt offering and the sin offering for all Israel.
\verse And he stood the Levites in the house of Adonai with cymbals, harps, and lyres, according to the command of David and Gad, the seer of the king, and Nathan the prophet, for the command was by the hand of Adonai by the hand of his prophets.
\verse When the Levites stood with the objects of David and the priests with the trumpets,
\verse Hezekiah commanded to offer the burnt offering upon the altar. And at the time the burnt offering began, the song of Adonai began, and also the trumpets, at the hands of the instruments of David king of Israel.
\verse Then all the assembly bowed down, and the song was sung, and the trumpeters trumpeted—all together until the end of the burnt offering.
\verse And when the offering was finished, the king and all who were found with him knelt down and bowed themselves.
\verse Then King Hezekiah and the princes commanded the Levites to praise Adonai with the words of David and Asaph the seer. So they offered praise with joy, and they bowed down and worshiped.
\verse Then Hezekiah answered and said, “Now \textit{you have consecrated yourselves} to Adonai. Come near and bring sacrifices and thank offerings to the house of Adonai!” And the assembly brought sacrifices and thank offerings, and all who were willing in heart brought burnt offerings.
\verse And the number of burnt offerings that the assembly brought was seventy cattle, one hundred rams, two hundred lambs—all these were to offer to Adonai.
\verse And the consecrated offerings of cattle were six hundred cattle and three thousand sheep.
\verse But the priests were too few, and they were not able to offer all the burnt offerings, so their brothers, the Levites, strengthened them to finish the work, and until the rest of the priests consecrated themselves. (For the Levites were \textit{more conscientious} to sanctify themselves than the priests.)
\verse And beside the many burnt offerings was the fat of the peace offerings and the libations for the burnt offerings. So the service of the house of Adonai was established.
\verse And Hezekiah and all the people rejoiced over what God had established with the people, for the matter happened suddenly.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 30
\verseWithHeading{Hezekiah Celebrates Passover} Then Hezekiah sent to all Israel and Judah, and he also wrote letters to Ephraim and Manasseh, to come to the house of Adonai in Jerusalem to make a Passover feast to Adonai the God of Israel.
\verse Now the king and his princes, and all the assembly in Jerusalem had taken counsel to make the Passover feast in the second month—
\verse but they were not able to make it at that time, for the priests had not consecrated themselves \textit{in sufficient numbers}, and the people had not been assembled in Jerusalem—
\verse and the plan seemed right in the eyes of the king and in the eyes of all the assembly.
\verse So they let the decree stand, \textit{to make a proclamation} throughout all Israel, from Beersheba to Dan, to come to make a Passover feast to Adonai the God of Israel.
\verse And the runners went with the letter from the hand of the king and his princes throughout all Israel and Judah according to the command of the king, saying, “O sons of Israel, return to Adonai, the God of Abraham, Isaac, and Israel, that he may turn again to the remnant of you who escaped from the hand of the kings of Assyria.
\verse Do not be like your fathers and like your brothers, who acted unfaithfully before Adonai, the God of their ancestors, so that he made them as a desolation, as you see.
\verse Now, do not stiffen your neck as your fathers, but give a hand to Adonai and come to his sanctuary, which he has consecrated forever, and serve Adonai your God that his fierce anger may turn away from you.
\verse For when you return to Adonai, your brothers and your children will find compassion before their captors and return to this land, for Adonai your God is gracious and compassionate, and he will not turn away his face from you if you return to him.”
\verse And it happened that the runners were passing from city to city in the land of Ephraim and Manasseh, and up to Zebulun, but they were laughing at them and mocking them.
\verse Only men from Asher, Manasseh, and Zebulun humbled themselves and came to Jerusalem.
\verse The hand of God was also upon Judah, to give them one heart to obey the command of the king and the princes concerning the word of Adonai.
\verse So many people gathered in Jerusalem to hold the Feast of Unleavened Bread in the second month; it was a very great assembly.
\verse And they rose up and removed the altars that were in Jerusalem. And they removed all the incense altars and threw them away in the Wadi Kidron.
\verse And they slaughtered the Passover lamb on the fourteenth day of the second month. And the priests and the Levites were disgraced, so they consecrated themselves and brought burnt offerings into the house of Adonai.
\verse Then they stood at their positions according to custom, according to the law of Moses the man of God. The priests were sprinkling the blood from the hand of the Levites.
\verse For there were many in the assembly who did not consecrate themselves, so the Levites were over the killing of the Passover sacrifices for all who were not clean, to consecrate them to Adonai.
\verse For a majority of the people, many from Ephraim, Manasseh, Issachar, and Zebulun, had not cleansed themselves. But they ate the Passover sacrifice \textit{otherwise than prescribed}, but Hezekiah had prayed for them, saying, “May the good Adonai make atonement unto
\verse everyone whose heart is set to seek God, Adonai the God of his ancestors, though not according to the cleansing for the sanctuary.”
\verse And Adonai listened to Hezekiah, and he healed the people.
\verse And the \textit{Israelites} who were found in Jerusalem kept the Feast of Unleavened Bread seven days with great joy. And the Levites and the priests were offering praise to Adonai day by day, with powerful instruments to Adonai.
\verse And Hezekiah spoke to the heart of all the Levites \textit{who were skilled at service} to Adonai. So they ate the feast seven days, offering peace offerings and giving thanks to Adonai, the God of their ancestors.
\verse Then the whole assembly decided to make seven more days of feasting, so they made seven more days with joy,
\verse for Hezekiah the king of Judah had \textit{provided} one thousand bulls and seven thousand sheep for the assembly, and the princes \textit{provided} one thousand bulls and ten thousand sheep for the assembly. And many priests consecrated themselves.
\verse And the whole assembly of Judah, the priests and the Levites, all the assembly who had come from the land of Israel, and those living in Judah rejoiced.
\verse And there was great joy in Jerusalem, for from the days of Solomon the son of David, king of Israel, there was nothing such as this in Jerusalem.
\verse Then the priests and the Levites arose and blessed the people, and their voice was heard, and their prayer came to his holy dwelling place in the heavens.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 31
\verseWithHeading{Hezekiah Organizes the Priesthood} And when all this was finished, all Israel who were found in the cities of Judah went out and shattered the stone pillars, cut down the Asherahs, and destroyed the high places and the altars from all Judah, Benjamin, Ephraim, and Manasseh \textit{to the very last one}. Then all the Israelites returned, each to his own property and to their cities.
\verse And Hezekiah appointed working groups of the priests and the Levites, each according to their divisions, \textit{according to the service for priests and Levites}, for burnt offerings, for peace offerings, to serve, to give thanks, and to praise in the gates of the camps of Adonai.
\verse And the gift of the king was out of his own possessions for the burnt offerings—the burnt offerings for the morning and the evening and for Sabbaths, new moon festivals, and appointed feasts, as is written in the law of Adonai.
\verse Then he said to the people, to the inhabitants of Jerusalem, to give a gift for the priests and the Levites, so that they might be strong in the law of Adonai.
\verse And the word spread. The Israelites gave abundantly from the firstfruits of grain, new wine, olive oil, honey, and all kinds of grains from the field. And they brought tithes of everything in abundance.
\verse Then the people of Israel and Judah who were living in the cities of Judah, they also brought a tithe of cattle and sheep and a tithe of consecrated objects to Adonai their God. And they gave heaps upon heaps.
\verse In the third month the heaps began to pile up, and in the seventh month they ceased.
\verse When Hezekiah and the princes came and saw the heaps, they blessed Adonai and his people Israel.
\verse And Hezekiah questioned the priests and the Levites concerning the heaps.
\verse And Azariah the chief priest of the house of Zadok answered him, saying, “Since the offerings began to be brought to the house of Adonai, we have eaten and been satisfied, and have much to spare, for Adonai has blessed his people. Now see this abundant remainder!”
\verse And Hezekiah commanded to prepare chambers in the house of Adonai, and they were prepared.
\verse Then they brought the offerings and tithes and holy objects faithfully. And over them as leader was Conaniah the Levite. And Shimei his brother was second in rank.
\verse And Jehiel, Azaziah, Nathan, Asahel, Jerimoth, Jozabad, Eliel, Ismachiah, Mahath, and Benaiah were chief officers under the hand of Conaniah and Shimei, his brother, by appointment of King Hezekiah. And Azaraiah was leader of the house of God.
\verse Now Kore the son of Imlah, the Levite, the gatekeeper of the east gate was over the freewill offerings of God, to give offerings to Adonai and the most holy things.
\verse And at his hand were Eden, Miniamin, Jeshua, Shemaiah, Amariah, and Shecaniah in the cities of the priests and \textit{entrusted} to give to their brothers by working groups, according to the eldest to the youngest,
\verse besides their enrollment by genealogy, to males from three years old and upward, to all who came into the house of Adonai, \textit{as each day required} for their service in their posts according to their working groups.
\verse And the enrollment of the priests was according to the house of their fathers, and the Levites, from twenty years old and upward, according to their posts by their working groups.
\verse And they were enrolled with all their little ones, their wives, their sons, and their daughters for the whole assembly because of their faithfulness. And they sanctified themselves in holiness.
\verse And for the descendants of Aaron, the priests, who were in the fields of the pasturelands of their cities, \textit{in each and every one of their cities} were men who were designated by name to give shares to every male among the priests and to everyone enrolled among the Levites.
\verse And Hezekiah did according to this throughout all Judah. And he did what is good and what is right and what is faithful before Adonai his God.
\verse And in every work that he began in the service of the house of God or with respect to the law and the commandment to seek his God, he did with all his heart, and he prospered.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 32
\verseWithHeading{Sennacherib’s Invasion of Judah} After these things and these acts of faithfulness, Sennacherib the king of Assyria came, and he came against Judah. And he encamped against the fortified cities and planned to break them down for himself.
\verse When Hezekiah saw that Sennacherib had come and that his face was set for battle against Jerusalem,
\verse he took counsel with his commanders and his mighty warriors to block off the waters of the springs that came from outside the city, and they helped him.
\verse Then many people were gathered, and they blocked off all the springs and the river that flowed through the midst of the land, saying, “Why should the kings of Assyria come and find much water?”
\verse Then he strengthened himself and built up all the walls that were broken down, and raised towers upon them and another wall \textit{outside}. And he strengthened the Millo of the city of David and made much weaponry and small shields.
\verse And he appointed commanders for battle over the people and gathered them to himself into the public square of the gate of the city. And he spoke to their hearts, saying,
\verse “Be strong! Be courageous! Do not fear and do not be dismayed before the king of Assyria and before all the crowd that is with him, for there are more with us than with him.
\verse With him is the arm of flesh, and with us is Adonai our God, to help us and to fight our battles.” And the people took confidence with the words of Hezekiah, king of Judah.
\verse After this Sennacherib the king of Assyria sent his servants to Jerusalem (now he and all his armies with him were against Lachish) to Hezekiah king of Judah and to all of Judah that was in Jerusalem, saying,
\verse “Thus says Sennacherib the king of Assyria: ‘On what are you relying that you are dwelling in siege works in Jerusalem?
\verse Is not Hezekiah urging you to give you up to die by starvation and thirst, saying, “Adonai our God will save us from the hand of the king of Assyria”?
\verse Has not Hezekiah himself removed his high places and his altars and said to Judah and to Jerusalem, saying, “You must bow down before one altar and upon it you must make offerings”?
\verse Do you not know what I have done, I and my ancestors, to all the peoples of the lands? Were the gods of the nations of all the lands at all able to save their land from my hand?
\verse Who among all the gods of those nations whom my ancestors utterly destroyed was able to save his people from my hand, that your God will be able to save you from my hand?
\verse So now, do not let Hezekiah deceive you. Do not let him urge you according to this. Do not put trust in him, for no god of any nation and kingdom has been able to save his people from my hand and from the hand of my ancestors. Surely then your God will not save you from my hand!’ ”
\verse And still more his servants said against Adonai and against Hezekiah his servant.
\verse And he wrote letters to treat Adonai the God of Israel with contempt and spoke against him, saying, “As the gods of the nations of the earth who did not save their people from my hand, so likewise the God of Hezekiah will not save his people from my hand.”
\verse Then they called with a great voice in Judean to the people of Jerusalem who were upon the wall to frighten them and terrify them, so that they could take the city captive.
\verse And they spoke about the God of Jerusalem as about the gods of the peoples of the earth, the works of the hands of humankind.
\verseWithHeading{Adonai’s Deliverance of Jerusalem} Then King Hezekiah and Isaiah the son of Amoz, the prophet, prayed concerning this. And they cried to the heavens.
\verse Then Adonai sent an angel, and he destroyed every mighty warrior of strength, commander, and officer in the camp of the king of Assyria. And he returned with shamed face to his land and went into the house of his god. And some of the offspring of his loins fell upon him there with the sword.
\verse So Adonai saved Hezekiah and the inhabitants of Jerusalem from the hand of Sennacherib, king of Assyria, and from the all their enemies, and gave them rest all around.
\verse And many brought tribute to Adonai, to Jerusalem, and precious things to Hezekiah king of Judah, so that he was exalted in the eyes of all the nations thereafter.
\verseWithHeading{The Rest of Hezekiah’s Reign} In those days Hezekiah fell ill unto death, and he prayed to Adonai. And he answered him and gave him a sign.
\verse But Hezekiah did not reciprocate according to the benefit placed upon him, because his heart became proud. So wrath was upon him and upon Judah and Jerusalem.
\verse But Hezekiah humbled himself with respect to the arrogance of his heart, he and the inhabitants of Jerusalem, so that the wrath of Adonai did not come upon them in the days of Hezekiah.
\verse And Hezekiah had very much wealth and honor, and he made storehouses for himself for silver, gold, precious stones, spices, small shields, and all sorts of desirable objects;
\verse and storage buildings for the yield of grain, new wine, and olive oil; and animal stalls for all kinds of animals, and animals and herds for animal stalls.
\verse And he made cities for himself, and livestock of sheep and abundant cattle, for God had given to him very abundant possessions.
\verse And this same Hezekiah blocked off the flow of the waters of the upper Gihon, and directed them down the west side of the city of David. And Hezekiah prospered in all his works.
\verse And thus in the matter of the envoys of the commanders of Babylon who had been sent to him to seek the sign that had happened in the land, God forsook him, to test him and to know all that was in his heart.
\verse Now the remainder of the words of Hezekiah and his loyal love, behold, they are written in the visions of Isaiah the son of Amoz, the prophet, upon the scroll of the kings of Judah and Israel.
\verse And Hezekiah slept with his ancestors, and they buried him in the upper part of the burial sites of the descendants of David. And all Judah and the inhabitants of Jerusalem did him honor at his death. And Manasseh his son became king in his place.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 33
\verseWithHeading{The Reign of Manasseh} Manasseh was twelve years old when he began to reign, and he reigned fifty-five years in Jerusalem.
\verse And he did evil in the eyes of Adonai according to the detestable things of the nations whom Adonai drove out before the \textit{Israelites}.
\verse And \textit{he rebuilt} the high places that Hezekiah his father broke down, and he set up altars for the Baals, made Asherahs, and bowed down to the host of heaven and served them.
\verse And he built altars in the house of Adonai, of which Adonai had said, “In Jerusalem my name shall be forever.”
\verse And he built altars to all the host of heaven in the two courtyards of the house of Adonai.
\verse And he himself burned his sons in the fire in the Valley of Ben-Hinnom. And he practiced sorcery and divination, and he engaged in witchcraft and dealt with mediums and spiritists. And he did much evil in the eyes of Adonai, to provoke him to anger.
\verse And he placed the carved image of the idol that he had made in the house of God, of which God had said to David and to Solomon his son, “In this house and in Jerusalem, which I have chosen from all the tribes of Israel, I will put my name forever.
\verse And I will \textit{never} remove the foot of Israel from upon the land that I appointed to your ancestors, if only you will take care to do all that I have commanded them, all the law, the regulations, and the judgments by the hand of Moses.”
\verse And Manasseh seduced Judah and the inhabitants of Jerusalem to do evil more than the nations that Adonai destroyed before the \textit{Israelites}.
\verseWithHeading{Manasseh’s Repentance} And Adonai spoke to Manasseh and to his people, but they would not listen.
\verse So Adonai brought the commanders of the army of the king of Assyria upon them, and they took Manasseh captive with hooks, and they bound him with bronze fetters and brought him to Babylon.
\verse And when he was in distress he entreated Adonai his God and greatly humbled himself before the God of his ancestors
\verse and prayed to him. And God responded to him and heard his plea and let him return to Jerusalem to his kingdom. And Manasseh knew that Adonai was God.
\verse Then afterward he built an outer wall for the city of David west of the Gihon in the valley, and for the entrance into the Gate of the Fishes. And it encircled the Ophel and raised it very high. Then he placed strong commanders in all the fortified cities of Judah.
\verse And he removed the foreign gods and the carved image from the house of Adonai and all the altars which he built on the mountain of the house of Adonai and in Jerusalem, and he threw them outside the city.
\verse And he restored the altar of Adonai and offered sacrifices of peace offerings and thank offerings upon it. And he commanded Judah to serve Adonai, the God of Israel.
\verse Nevertheless, the people offered at the high places, but only to Adonai their God.
\verse Now the remainder of the words of Manasseh and his prayer to his God and the words of the seers who spoke to him in the name of Adonai, the God of Israel, behold, they are in the words of the kings of Israel.
\verse And his prayer, and how he responded to him, all his sin and his unfaithful acts, and the places where he built the high places and set up Asherahs and idols before he humbled himself, behold, they are written in the words of the seers.
\verseWithHeading{Amon’s Reign} And Manasseh slept with his ancestors, and they buried him in his house. And Amon his son became king in his place.
\verse Amon was twenty-two years old when he began to reign, and he reigned two years in Jerusalem.
\verse And he did what was evil in the eyes of Adonai as Manasseh his father had done. And Amon sacrificed to all the idols that Manasseh his father had made, and he served them.
\verse And he did not humble himself before Adonai, as Manasseh his father humbled himself, but Amon himself multiplied his guilt.
\verse And his servants conspired against him and killed him in his house.
\verse But the people of the land struck down all who conspired against King Amon. And the people of the land made his son Josiah king in his place.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 34
\verseWithHeading{The Reign of Josiah} Josiah was eight years old when he became king, and he reigned thirty-one years in Jerusalem.
\verse And he did what was right in the eyes of Adonai, and he walked in the ways of David his ancestor and did not deviate to the right or to the left.
\verse In the eighth year of his reign, while he was still a boy, he began to seek the God of David his ancestor. And in the twelfth year he began to cleanse Judah and Jerusalem of the high places, the Asherahs, the idols, and the images.
\verse And he destroyed before him the altars for the Baals, and he cut down the incense stands above them. And the Asherahs, the idols, and the images he smashed. And he ground them to powder and sprinkled the dust over the burial sites of those who sacrificed to them.
\verse And he burned the bones of the priests on the altars and cleansed Judah and Jerusalem.
\verse And in the cities of Manasseh, Ephraim, Simeon, and as far as Naphtali, in their ruins all around,
\verse he broke down the altars and the Asherahs and crushed the idols, grinding them to dust, and he cut down all the incense stands in all the land of Israel. Then he returned to Jerusalem.
\verseWithHeading{The Scroll of the Law Discovered} Now in the eighteenth year of his reign after he purged the land and the house, he sent Shaphan the son of Azaliah, Maaseiah the commander of the city, and Joah the son of Joahaz, the recorder, to repair the house of Adonai his God.
\verse And they came to Hilkiah the high priest and gave the money that was brought for the house of God, which the Levites, the guardians of the threshold, had gathered from the hand of Manasseh, Ephraim, and from the whole remnant of Israel, and from all Judah and Benjamin and the inhabitants of Jerusalem.
\verse And they delivered it into the hand of the workmen who were appointed in the house of Adonai, and they in turn gave it to those doing the work who were at work in the house of Adonai, to repair and restore the house.
\verse And they dispensed it to the skilled craftsmen and the builders to buy stone for hewing and timber for the seams and for binding the houses that the kings of Judah had destroyed.
\verse And the men did the work faithfully. Over them as supervisors were appointed Jahath and Obadiah, the Levites from the Merarites, and Zechariah and Meshullam from the Kohathites. And other Levites, all who were skilled in instruments of song,
\verse were over the porters and were directing all who worked \textit{on each task}. And some of the Levites were scribes, officials, and gatekeepers.
\verse And when they brought out the money that had been brought to the house of Adonai, Hilkiah the priest found a scroll of the law of Adonai by the hand of Moses.
\verse And Hilkiah answered and said to Shaphan the scribe, “I have found a scroll of the law in the house of Adonai!” Then Hilkiah gave the scroll to Shaphan.
\verse And Shaphan brought the scroll to the king and \textit{also informed} the king, saying, “All that was put into the hand of your servants they are doing.
\verse They have dispensed the money which was found in the house of Adonai and have put it into the hand of the overseers and those doing the work.”
\verse Then Shaphan the scribe reported to the king, “Hilkiah the priest gave to me a scroll.” Then Shaphan read from it before the king.
\verse And when the king heard the words of the law, he tore his garments,
\verse and the king commanded Hilkiah, Ahikam the son of Shaphan, Abdon the son of Micah, Shaphan the scribe, and Asaiah the servant of the king, saying,
\verse “Go! Seek Adonai on my behalf and on behalf of those remaining in Israel and Judah concerning the words of the scroll that was found, for the anger of Adonai that was poured out on us is great, because our ancestors have not kept the word of Adonai, to do all that is written in this scroll.”
\verse Then Hilkiah and those whom the king had sent went to Huldah the prophetess, the wife of Shallum, son of Tikvath, son of Hasrah, keeper of the garments. (Now she was living in Jerusalem in the second district.) And they spoke to her concerning this.
\verse And she said to them, “Thus says Adonai, the God of Israel: ‘Say to the man who sent you all to me:
\verse “Thus says Adonai: ‘Behold, I am bringing disaster upon this place and upon its inhabitants—all the curses written in the scroll that were read before the king of Judah—
\verse because they have abandoned me and burned incense to other gods so as to provoke me to anger with all the works of their hands. And my wrath will be poured out on this place and it will not be quenched.’
\verse But to the king of Judah who sent you to seek Adonai, thus shall you say to him: ‘Thus says Adonai, the God of Israel: As for the words which you heard,
\verse because your heart was responsive and you humbled yourself before God when you heard his words against this place and against its inhabitants and you humbled yourself before me and tore your garments and wept before me, now I also have heard,’ \textit{declares} Adonai.
\verse ‘Behold, I will gather you to your ancestors, and you will be gathered to your grave in peace. Your eyes will not see any of the disaster that I am bringing upon this place.’ ” ’ ”
\verse Then they sent word to the king, and he gathered all the elders of Judah and Jerusalem.
\verse And the king went up to the house of Adonai with all the men of Judah and the inhabitants of Jerusalem, with the priests and the Levites—all the people from great to small—and they read in their ears all the words of the scroll of the covenant that was found in the house of Adonai.
\verse And the king stood at his position. And he made the covenant before Adonai, to walk after Adonai and to keep his commands and statutes and regulations with all his heart and with all his soul, to do the words of the covenant that were written on this scroll.
\verse Then he obligated all who were found in Jerusalem and in Benjamin, that the inhabitants of Jerusalem would do according to the covenant of God, the God of their ancestors.
\verse And Josiah removed all the detestable things from the whole land that belonged to the \textit{Israelites} and obligated all who were found in Israel to serve Adonai their God. All his days they did not turn aside from following Adonai, the God of their ancestors.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 35
\verseWithHeading{Josiah Observes the Passover} Josiah kept the Passover to Adonai in Jerusalem. And they slaughtered the Passover sacrifice on the fourteenth day of the first month.
\verse And he set the priests at their posts and encouraged them in the service of the house of Adonai.
\verse And he said to the Levites who were teaching all Israel, who were sanctified to Adonai, “Put the holy ark in the house which Solomon the son of David, king of Israel, built. There is no need to carry it on your shoulders. Now serve Adonai your God and his people Israel.
\verse And prepare yourselves according to \textit{your families}, according to your working groups, according to the decree of David king of Israel and according to the decree of Solomon his son.
\verse And stand in the sanctuary according to the clans of \textit{your families} for your brothers, the \textit{lay people}, and a part of the \textit{family} of the Levites.
\verse And slaughter the Passover lamb and consecrate yourself and prepare for your brothers to do according to the word of Adonai by the hand of Moses.
\verse Then Josiah provided for the \textit{lay people} sheep, lambs, and young goats, all for the Passover for all who were found there, to the sum of thirty thousand, along with three thousand bulls. These were the possession of the king.
\verse And his officials contributed a freewill offering to the people, to the priests, and to the Levites. Hilkiah, Zechariah, and Jehiel, commanders of the house of God, gave to the priests two thousand six hundred Passover lambs and three hundred bulls.
\verse Conaniah also, and Shemaiah and Nethanel his brothers, and Hashabiah and Jeiel and Jehozabad, commanders of the Levites, contributed to the Levites five thousand Passover lambs and five hundred bulls.
\verse And when the work had been completed, the priests stood at their positions and the Levites in their working groups according to the command of the king.
\verse And they slaughtered the Passover lamb, and the priests sprinkled the blood from their hand, and the Levites flayed the sacrifices.
\verse And they set aside the burnt offerings to give them to the clans of the \textit{families} for the \textit{lay people}, to offer to Adonai according to what is written in the scroll of Moses. And thus they did with the bulls.
\verse Then they roasted the Passover lamb in the fire according to the ordinance. And they boiled the holy offerings in pots, in kettles, and in dishes and brought it quickly to the \textit{lay people}.
\verse And afterward they prepared for themselves and for the priests, since the priests, the descendants of Aaron, were busy with offering the burnt offerings and the fat until night. So the Levites prepared for themselves and for the priests, the descendants of Aaron.
\verse And the singers, the descendants of Asaph, were at their stations according to the decree of David and Asaph and Heman and Jeduthun, the seer of the king. And the gatekeepers were at \textit{each gate}. They did not need to depart from their service, for their brothers the Levites had made preparations for them.
\verse So all the service of Adonai was prepared on that day to keep the Passover and to sacrifice burnt offerings on the altar of Adonai, according to the command of King Josiah.
\verse And the Israelites who were present kept the Passover at that time, and the Feast of Unleavened Bread seven days.
\verse And there was no Passover like it kept in Israel since the days of Samuel the prophet. And none of the kings of Israel kept a Passover like that which Josiah and the priests and the Levites, and all of Judah and Israel who were present, and the inhabitants of Jerusalem, kept.
\verse In the eighteenth year of the reign of Josiah this Passover was kept.
\verseWithHeading{The Death of Josiah} After all of this that Josiah had prepared for the temple, Neco the king of Egypt went up to make war at Carchemish on the Euphrates, and Josiah went up to meet him.
\verse And he sent messengers to him, saying, “\textit{What is there between us}, O king of Judah? I am not against you yourself this day, but against the house making war against me. And God has commanded me to make haste. Stop opposing God, who is with me that he will not destroy you.”
\verse But Josiah did not turn his face from him, but he disguised himself to fight with him. He did not listen to the words of Neco from the mouth of God, and he went out to fight in the plain of Megiddo.
\verse And the archers shot King Josiah, and the king said to his servants, “Take me away, for I am badly wounded.”
\verse So his servants took him out of the chariot and transported him in his second chariot and brought him to Jerusalem. And he died and was buried in the burial sites of his ancestors. And all Judah and Jerusalem mourned over Josiah.
\verse And Jeremiah uttered a lament for Josiah, and all the male and female singers have spoken in their laments about Josiah to this day. And they made them a requirement in Israel, and behold, they are written in the laments.
\verse Now the remainder of the words of Josiah and his loyal love for what is written in the law of Adonai
\verse and his words, from the first to the last, behold, they are written in the scroll of the kings of Israel and Judah.
\end{biblechapter}

\begin{biblechapter} % 2 Chronicles 36
\verseWithHeading{The Decline of Judah} And the people of the land took Jehoahaz the son of Josiah and made him king in place of his father in Jerusalem.
\verse Jehoahaz was twenty-three years old when he began to reign, and he reigned three months in Jerusalem.
\verse Then the king of Egypt deposed him in Jerusalem and laid a tribute upon the land of one hundred talents of silver and one talent of gold.
\verse And the king of Egypt made Eliakim his brother king over Judah and Jerusalem, and he changed his name to Jehoiakim. But Neco took Jehoahaz his brother and brought him to Egypt.
\verse Jehoiakim was twenty-five years old when he began to reign, and he reigned eleven years in Jerusalem. And he did evil in the eyes of Adonai his God.
\verse Nebuchadnezzar the king of Babylon went up against him, and he bound him with bronze fetters to bring him to Babylon.
\verse And Nebuchadnezzar brought to Babylon the objects of the house of Adonai and put them into the temple in Babylon.
\verse Now the remainder of the words of Jehoiakim and the detestable things that he did and what was found against him, behold, they are written in the scroll of the kings of Israel and Judah. And Jehoiachin his son reigned in his place.
\verse Jehoiachin was eighteen years old when he became king, and he reigned three months and ten days in Jerusalem. And he did evil in the eyes of Adonai.
\verse And at the turn of the year King Nebuchadnezzar sent and brought him to Babylon, with objects of the treasure of the house of Adonai. And he made Zedekiah his brother king in Judah and Jerusalem.
\verse Zedekiah was twenty-one years old when he began to reign, and he reigned eleven years in Jerusalem.
\verse And he did evil in the eyes of Adonai his God. He did not humble himself before Jeremiah the prophet \textit{who spoke for} Adonai.
\verse And he also rebelled against King Nebuchadnezzar, who had made him swear by God. He stiffened his neck and hardened his heart against turning to Adonai, the God of Israel.
\verse All the leaders of the priests and the people likewise increased in unfaithfulness according to all the detestable things of the nations. And they polluted the house of Adonai that he had consecrated in Jerusalem.
\verse And Adonai, the God of their ancestors, had repeatedly and persistently sent to them by the hand of his messengers, because he had compassion on his people and on his dwelling place.
\verse But they were mocking the messengers of God and despising his words and scoffing at his prophets until the wrath of Adonai rose against his people until there was no remedy.
\verseWithHeading{The Destruction of Jerusalem} Therefore he brought against them the king of the Chaldeans, and he killed their young men with the sword in the house of their sanctuary. He showed no mercy on a young man, a virgin, the elderly, or decrepit. He delivered all into his hand.
\verse And all the vessels of the house of God, the great and the small, the storehouses of the house of Adonai, and the storehouses of the king and his princes, these all he brought to Babylon.
\verse And they burned the house of God. And they shattered the walls of Jerusalem and burned its citadels with fire and destroyed all the vessels of its treasuries.
\verse And he took \textit{those who escaped the sword} to Babylon. And they became servants to him and to his sons until the rule of the kingdom of Persia,
\verse to fulfill the word of Adonai by the mouth of Jeremiah, until the land has enjoyed its Sabbaths. All the days of desolation it kept Sabbath, to fulfill seventy years.
\verseWithHeading{The Decree of Cyrus} And in the first year of Cyrus, the king of Persia, in order to fulfill the word of Adonai by the mouth of Jeremiah, Adonai stirred up the spirit of Cyrus, king of Persia, so that he \textit{made a proclamation} throughout all his kingdom and also in writing, saying:
\verse “Thus says Cyrus, king of Persia: ‘Adonai the God of heaven has given to me all the kingdoms of the earth, and he has appointed me to build a house for him at Jerusalem, which is in Judah. Whoever is among you from all his people, may Adonai his God go up with him.’ ”
\end{biblechapter}

