\biblebook{2 Samuel}

\begin{biblechapter} % 2 Samuel 1
\verseWithHeading{The Report of Saul’s Death by the Amalekite}{\textit{After}\lebnote{“And it happened after”}the death of Saul, David returned from defeating the Amalekites and he stayed at Ziklag two days.}%
\verse{On the third day, a man came from the camp from being with Saul, with his clothes torn and with dirt on his head. \textit{When he came}\lebnote{“And it happened at his coming”}to David, he fell to the ground and bowed down.}%
\verse{David said to him, “Where did you come from?” He said to him, “I have escaped from the camp of Israel.”}%
\verse{Then David said to him, “\textit{How did things go}?\lebnote{“What was the thing”}Please tell me.” He answered, “\textit{When}\lebnote{“That”}the army fled from the battle, and many of the people fell; also, Saul and Jonathan his son died.”}%
\verse{Then David asked the young man who was reporting to him, “How do you know that Saul and his son Jonathan died?”}%
\verse{The young man who was reporting to him said, “I merely happened to be on Mount Gilboa. Here Saul was leaning on his spear, and look, the chariots and the horsemen were getting close to him.}%
\verse{When he turned around and saw me, he called to me, and I said, ‘Here I am.’}%
\verse{Then he said to me, ‘Who are you?’ And I said to him, ‘I am an Amalekite.’}%
\verse{He said to me, ‘Please stand over me and kill me, for convulsions have seized me, even though my life is still in me.’}%
\verse{So I stood over him and killed him, for I knew that he could not live after his falling; I took the crown that was on his head and the bracelet which was on his arm; and here, I have brought them to my lord.}%
\verse{David grabbed at his clothes and tore them, as did all of the men who were with him.}%
\verse{Then they mourned and wept and fasted over Saul and Jonathan his son until the evening, as well as over the people of Adonai and over the house of Israel because they had fallen by the sword.}%
\verse{Then David said to the young man who was reporting to him, “Where are you from?” And he said, “I am the son of an alien man. I am an Amalekite.”}%
\verse{David said to him, “How is it that you were not afraid to stretch out your hand to destroy Adonai’s anointed one?”}%
\verse{Then David called to one of the young men and said to him, “Come near; strike him.” So he struck him down and he died.}%
\verse{David said to him, “Your blood is on your head, for your mouth has testified against you by saying, ‘I killed Adonai’s anointed one!’ ”}%
\verseWithHeading{David Laments Jonathan with the “Song of the Bow”}{Then David sang this funeral song over Saul and over Jonathan his son.}%
\verse{And he ordered “The Bow” to be taught to the children of Judah. Look, it is written on the scroll of Jashar.\lebnote{The Book of Jashar (“the upright”) is also mentioned in Josh 10:13.}}%
\verse{“The glory of Israel is on your high places; how the mighty have fallen!}%
\verse{Do not tell it in Gath; do not proclaim it in the streets of Ashkelon, 
lest the daughters of the Philistines rejoice, 
lest the daughters of the uncircumcised exult.}%
\verse{O mountains of Gilboa, let there be no dew or rain upon you 
or on the fields of grain for offerings, 
for there the small shield of the mighty was defiled, 
the small shield of Saul was not anointed with oil.}%
\verse{From the blood of the slain, from the fat of the mighty, 
the bow of Jonathan did not turn back, 
and the sword of Saul did not return \textit{without effect}.\lebnote{“empty handedly”}}%
\verse{Saul and Jonathan were beloved and pleasant in their lives 
and were not separated in their death. 
They were swifter than eagles, stronger than lions.}%
\verse{O daughters of Israel, weep over Saul, 
the one who clothed you with crimson, 
the one who \textit{adorned your clothing with golden ornaments}.\lebnote{“put gold ornaments on your clothing”}}%
\verse{How the mighty have fallen in the midst of the battle; 
Jonathan lies slain on your high places.}%
\verse{\textit{I am distressed}\lebnote{“Distress is for me”}over you, my brother Jonathan. 
you were very dear to me; 
your love was more wonderful to me than the love of women.}%
\verse{How the mighty have fallen and the weapons of warfare perished.”}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 2
\verseWithHeading{David Moves to Hebron}{It happened after this that David inquired of Adonai, saying, “Shall I go up into one of the cities of Judah?” And Adonai said to him, “Go up.” David asked, “Where shall I go up?” And he said, “To Hebron.”}%
\verse{So David went up there \textit{along with}\lebnote{“and also”}his two wives, Ahinoam \textit{from Jezreel}\lebnote{“the Jezreelitess”}and Abigail the wife of Nabal the Carmelite.}%
\verse{Also, David brought up his men who were with him, each \textit{with}\lnFSM{}his household, and they settled in the towns of Hebron.}%
\verseWithHeading{David Anointed King over Judah at Hebron}{Then the men of Judah came, and they anointed David there as king over the house of Judah, and they told David, “The men of Jabesh-Gilead buried Saul.”}%
\verse{So David sent messengers to the men of Jabesh-Gilead and said to them, “May you be blessed by Adonai because you did this loyal love with your lord, with Saul, and you buried him.}%
\verse{Now may Adonai show loyal love and faithfulness with you. I will also show the good with you that you have done in this matter.}%
\verse{So then, let your hands be strong and \textit{be valiant},\lebnote{“as sons of ability”}for your lord Saul is dead, and the house of Judah has anointed me as king over them.”}%
\verseWithHeading{Ish-Bosheth over Israel}{But Abner the son of Ner, the commander of Saul’s army, had taken Ish-Bosheth the son of Saul and brought him over to Mahanaim.}%
\verse{He made him king over Gilead, over the Ashurites, over Jezreel, over Ephraim, over Benjamin, and over Israel, all of it.}%
\verse{Ish-Bosheth the son of Saul was forty years old when he became king over Israel and he reigned two years; however, the house of Judah \textit{followed}\lebnote{“were after”}David.}%
\verse{The number of days that David was king over Hebron, over the house of Judah, was seven years and six months.}%
\verseWithHeading{War between Judah and Israel}{Abner the son of Ner and the servants of Ish-Bosheth the son of Saul went out from Mahanaim to Gibeon.}%
\verse{Then Joab the son of Zeruiah and the servants of David went out, and they met at the pool of Gibeon. The one group sat on one side of the pool, and the other sat on the other side.}%
\verse{Abner said to Joab, “Let the young men come up and fight in our presence.” And Joab said, “Let them come up.”}%
\verse{So they came forward and passed by in number: twelve for Benjamin and for Ish-Bosheth the son of Saul, and twelve from the servants of David.}%
\verse{Then each seized his \textit{opponent}\lebnote{“neighbor”}by the head and each thrust his sword in the side of his opponent, so they fell together. So they\lnAQG{}called the name of that place Helkath Hazzurim,\lebnote{Helkath Hazzurim possibly means “the slippery slope”}which is in Gibeon.}%
\verse{Then the battle \textit{became increasingly fierce}\lebnote{“and it was/became strong/fierce until very”}on that day, and Abner and the men of Israel were defeated before the servants of David.}%
\verse{The three sons of Zeruiah were there, Joab and Abishai and Asahel. Now Asahel was swift with his feet as one of the gazelles which is in the open field.}%
\verse{So Asahel pursued\lnFTO{}Abner, and he did not turn aside to the right or to the left from going after Abner.}%
\verse{Abner turned around and said, “Are you this Asahel?” And he said, “Yes.”}%
\verse{Abner said to him, “Turn aside to your right or to your left; seize for yourself one of the young men, and take his belongings for yourself.” But Asahel was not willing to turn aside from him.}%
\verse{So Abner said to Asahel \textit{once again},\lebnote{“and he again did again”}“\textit{For your own sake},\lebnote{“for yourself”}turn aside \textit{from following me}.\lebnote{“from after me”}Why should I strike you down to the ground? How could I \textit{show my face}\lebnote{“lift up my face”}to Joab your brother?”}%
\verse{But he refused to turn away, so Abner struck him in the stomach with the butt of the spear, and the spear went out of his back. He fell there and he died \textit{on the spot}.\lebnote{“in his place”}\textit{Then}\lnDXN{}all who came to the place where Asahel fell and died just stood there.}%
\verse{So Joab and Abishai pursued\lnFTO{}Abner \textit{when}\lnFSM{}the sun went down. And they came to the hill country of Ammah, which is \textit{before Giah}\lebnote{“lies on the face of Giah”}on the way to the wilderness of Gibeon.}%
\verse{The descendants\lnAFK{}of Benjamin rallied after Abner, and they became as one fighting group and stood on the top of a certain hill.}%
\verse{Then Abner called to Joab and said, “Will the sword devour forever? Do you not know that there will be bitterness in the end? \textit{How long}\lebnote{“And up to when”}will you not tell the people to turn \textit{away from pursuing}\lnFTT{}their brothers?”}%
\verse{Joab said, “\textit{As God lives},\lebnote{“The life of God”}for if you had not spoken, the people would surely have gone up in the morning, each one of them \textit{from following after}\lnFTT{}his brother.”}%
\verse{Then Joab blew on the trumpet and all the people stopped, and they no longer pursued after Israel, and they did not fight with them again.}%
\verse{Then Abner and his men went through the Arabah all that night, and they crossed over the Jordan. They went all the forenoon and came to Mahanaim.}%
\verse{After Joab returned from pursuing\lebnote{Hebrew “after”}Abner, he gathered all the people; nineteen of the servants of David were missing along with Asahel.}%
\verse{The servants of David had killed some of the Benjaminites among the men of Abner; three hundred and sixty men had died.}%
\verse{Then they picked up Asahel and buried him in the grave of his father, which was at Bethlehem. Joab and his men went all that night \textit{arriving in Hebron at first light}.\lebnote{“it became light for them in Hebron”}}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 3
\verseWithHeading{The House of David Grows Stronger}{The battle was prolonged between the house of Saul and the house of David, but David \textit{was growing stronger and stronger}\lebnote{“was going and strong”}while the house of Saul \textit{was becoming weaker and weaker}.\lebnote{“was going and weak”}}%
\verse{And sons were born to David in Hebron; his firstborn was Amnon by Ahinoam \textit{from Jezreel}\lebnote{“the Jezreelitess”}.}%
\verse{His second was Kileab by Abigail the wife of Nabal the Carmelite; the third was Absalom the son of Maacah, who was the daughter of Talmai the king of Geshur.}%
\verse{The fourth was Adonijah the son of Haggith; the fifth was Shephatiah the son of Abital.}%
\verse{The sixth was Ithream by Eglah the wife of David. These were born to David in Hebron.}%
\verse{As the war between the house of Saul and the house of David was continuing, Abner was strengthening himself in the house of Saul.}%
\verse{Saul had had a concubine, and her name was Rizpah the daughter of Aiah. Then Ish-Bosheth\lebnote{Supported by some Hebrew and Greek manuscripts}said to Abner, “Why \textit{did you have sex with}\lebnote{“have you gone into”}my father’s concubine?”}%
\verse{\textit{Abner became very angry}\lebnote{“And it became hot for Abner very”}at the words of Ish-Bosheth, and he said, “Am I the head of a dog which is for Judah today? Do I not continue to show loyal love with the house of Saul your father, to his brothers, and to his friends? I have not let you fall into the hands of David, yet you have \textit{accused me of sin with this woman}\lebnote{“you have called me into account with the guilt of the woman”}today.}%
\verse{Thus \textit{may God punish Abner},\lebnote{“may God do to Abner, and thus may he add to it”}\textit{if I do not accomplish what Adonai has sworn to David};\lebnote{“for as that Adonai has sworn to David, so I will accomplish for him”}}%
\verse{specifically, to transfer the kingdom from the house of Saul and to establish the throne of David over Israel and over Judah from Dan to Beersheba!”}%
\verse{And Ish-Bosheth was no longer able to \textit{answer}\lebnote{“return a word to”}Abner \textit{because he feared him}.\lebnote{“because of his fearing him”}}%
\verseWithHeading{Abner Pledges Support for David}{Abner sent messengers to David \textit{where he was},\lebnote{“under him”}saying, “To whom does the land belong? \textit{Make your covenant with me}!\lebnote{“cut your covenant with me”}Look, my hand is with you to bring all of Israel over to you!”}%
\verse{He\lebnote{David}said, “Good. \textit{I will make a covenant with you}.\lebnote{“I will cut a covenant with you”}I am asking only one thing from you: You shall not see my face unless you bring Michal the daughter of Saul when you come to see \textit{me}.”\lebnote{“my face”}}%
\verse{Then David sent messengers to Ish-Bosheth the son of Saul, saying, “Give me my wife Michal whom I betrothed to myself\lnDMG{}for a hundred foreskins of the Philistines.”}%
\verse{So Ish-Bosheth sent and took her from her husband, from Paltiel the son of Laish.\lebnote{The Hebrew Masoretic text (\textit{Kethib}) reads “Lush”; \textit{Qere} reads “Laish”}}%
\verse{But her husband went with her, \textit{weeping all along}\lebnote{“going and weeping”}after her as far as Bahurim. Then Abner said to him, “Go and return.” So he returned.}%
\verse{The word of Abner \textit{came to}\lebnote{“was with”}the elders of Israel, saying, “\textit{For quite some time}\lebnote{“Both yesterday and the day before”}you were seeking David as king over you.}%
\verse{So then, bring it about, because Adonai had said to David, “Through the hand of David my servant I am about to save my people Israel from the hand of the Philistines and from the hand of all their enemies.”}%
\verse{Abner also spoke \textit{privately to Benjamin}.\lebnote{“in the ears of Benjamin”}Abner also went \textit{to speak to David}\lebnote{“to speak in the ears of David”}in Hebron, all that was good in the eyes of Israel and in the eyes of all the house of Benjamin.}%
\verse{Then Abner came to David in Hebron and with him were twenty men. David had prepared a feast for Abner and for the men who were with him.}%
\verse{And Abner said to David, “Let me get up and go and gather all of Israel to my lord the king, that they may \textit{make a covenant}\lebnote{“cut a covenant”}with you so you can reign over all which your soul desires.” So David dismissed Abner, and he left in peace.}%
\verse{And look, the servants of David and Joab came from the raid, and they brought much plunder with them. But Abner was not with David at Hebron, for he\lebnote{David}had dismissed him, and he had gone in peace.}%
\verse{When Joab and all the army that was with him came, they told Joab, “Abner the son of Ner came to the king, and he sent him away, and he left in peace.”}%
\verseWithHeading{Joab Assassinates Abner}{Then Joab came to the king and said, “What have you done? Abner came here to you? Why have you dismissed him that he \textit{actually went away}?\lebnote{“and he went going”}}%
\verse{You know that Abner the son of Ner came to deceive you. He came to learn about your going out and coming in and to know all which you are doing.”}%
\verse{Then Joab went out from David, and he sent messengers after Abner, and they brought him back from Bor Hasirah, but David did not know it.}%
\verse{When Abner returned to Hebron, Joab took him aside to the middle of the gate to speak with him in private. Then he struck him in the stomach there and he died, \textit{for he had shed the blood of Asahel his brother}.\lebnote{“for the blood of Asahel his brother.” See Num 35:16–25}}%
\verse{When David heard this afterwards, he said: “I and my kingdom are innocent \textit{before Adonai}\lebnote{“from with Adonai”}forever for the blood of Abner the son of Ner.}%
\verse{May the blood come down on the head of Joab and all the house of his father. \textit{May the house of Joab never lack}\lebnote{“May there not be cut off from the house of Joab”}one with a bodily discharge or one with leprosy or one who grasps the distaff or one who falls by the sword or one who is lacking food.”}%
\verse{So Joab and Abishai, his brother, killed Abner because he had killed Asahel, their brother, at Gibeon in the battle.}%
\verse{David said to Joab and to all the people who were with him, “Tear your clothing and put on sackcloth and mourn before Abner.” Now King David was following after the bier.}%
\verse{And they buried Abner at Hebron. And the king lifted up his voice and wept at the grave of Abner, and all the people wept.}%
\verse{The king sang a lament for Abner and said, “Should Abner have died the death of a fool?}%
\verse{Your hands were not tied and your feet were not in contact with bronze fetters. You have fallen as one who falls before sons of wickedness.” Then \textit{all the people wept over him again}.\lebnote{“and all the people did again to weep over him”}}%
\verse{Then all the people came to give David food. Still on that day, David swore, “\textit{May God punish me}\lebnote{“Thus may God do to me and thus may he add”}if I taste food or anything before the sun goes down.”}%
\verse{All the people noticed, and it was good in their eyes, as everything that the king did was good in the eyes of all the people.}%
\verse{Then all the people and all of Israel realized on that day that \textit{the king had not desired}\lebnote{“that it was not from the king”}to kill Abner the son of Ner.}%
\verse{Then the king said to his servants, “Did you not realize that a prince and a great man has fallen today in Israel?}%
\verse{I am weak today even though anointed king, and these men, the sons of Zeruiah, are crueler than I am. May Adonai pay them back for doing wickedness according to their own wickedness.”}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 4
\verseWithHeading{Ish-Bosheth is Assassinated}{When the son of Saul heard that Abner had died in Hebron, \textit{his courage failed}\lebnote{“his hands hung limp”}and all of Israel was horrified.}%
\verse{Two of the men, commanders of the raiding bands, were for the son of Saul. The name of one was Baanah, and the name of the other was Recab, the sons of Rimmon the Beerothite from the descendants\lnAFK{}of Benjamin; also, Beeroth was regarded as belonging to Benjamin.}%
\verse{The Beerothites fled to Gittaim, and they are resident aliens there until this day.}%
\verse{(Now Jonathan the son of Saul had a son who was crippled in the feet. He was five years old when the message of Saul and Jonathan came from Jezreel, and his nurse had picked him up and fled. It happened that as she was hurrying away to flee, he fell and became crippled. His name was Mephibosheth.)}%
\verse{When the sons of Rimmon the Beerothite, Recab and Baanah, set out, they came at the heat of the day to the house of Ish-Bosheth while \textit{he was taking a noontime rest}.\lebnote{“he was lying with the bed at noon”}}%
\verse{They came as far as the middle of the house as if takers of wheat, and they struck him in the stomach. Then Recab and Baanah his brother escaped.}%
\verse{When they had come into the house, he was lying on his couch \textit{in his bedchamber},\lebnote{“in the room of his lying”}and they attacked him and killed him. Then they \textit{beheaded him},\lebnote{“they took away his head”}and they took his head and went on the way of the Arabah all night.}%
\verse{They brought the head of Ish-Bosheth to David at Hebron, and they said to the king, “Here is the head of Ish-Bosheth, the son of Saul your enemy, who sought your life. Adonai has given to my lord the king vengeance this day on Saul and on his offspring.”}%
\verse{Then David answered Recab and Baanah his brother, the sons of Rimmon the Beerothite, and he said to them, “\textit{As Adonai lives},\lnFSP{}who redeemed my soul from all trouble,}%
\verse{when the one who told me, “Look, Saul is dead,” \textit{thought that he was bringing good news},\lebnote{“and he was as one bringing good news in his eyes”}I seized him and killed him at Ziklag, which was as my giving the news back to him.}%
\verse{\textit{How much more}\lebnote{“Also”}when wicked men kill a righteous man in his house, on his bed! So then, shall I not seek his lifeblood from your hand, so that I may destroy you from the earth?”}%
\verse{Then David commanded the young men, and they killed them, and they cut off their hands and their feet, and they hung them at the pool at Hebron, but the head of Ish-Bosheth they took and buried in the grave of Abner at Hebron.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 5
\verseWithHeading{David Anointed as King over All of Israel}{So all the tribes of Israel came to David at Hebron, and they said, “Here we are, \textit{we are your flesh and blood}.\lebnote{“We are your bone and your flesh”}}%
\verse{\textit{For some time},\lebnote{“Both yesterday and three days before”}when Saul was king over us, \textit{you were leading Israel in and out}.\lebnote{“you were going out and coming in, Israel.” (So \textit{Kethib}; \textit{Qere} reads “bringing out and bringing in Israel”)}Adonai had said to you, ‘You shall be the shepherd of my people Israel, and you will be the leader over Israel.’ ”}%
\verse{All the elders of Israel came to the king at Hebron, and King David \textit{made a covenant}\lebnote{“cut a covenant”}with them at Hebron in the presence of Adonai; then they anointed David as king over Israel.}%
\verse{David was thirty years old when he began to reign; he reigned forty years.}%
\verse{He reigned over Judah at Hebron for seven years and six months; and he reigned over all Israel and Judah at Jerusalem for thirty-three years.}%
\verseWithHeading{The Capture and Growth of Jerusalem}{The king and his men went to Jerusalem, to the Jebusites, the inhabitants of the land. They said to David, “You will not come here, for even the blind and the lame can turn you back, saying, ‘David cannot come here.’ ”}%
\verse{David captured the fortress of Zion, the city of David.}%
\verse{David had said, “On that day \textit{when we attack the Jebusites},\lebnote{“in all of the striking down the Jebusites”}one must attack the lame and the blind, those who hate the soul of David, by means of the water supply.” For thus the blind and the lame would say, “He cannot come into the house.”}%
\verse{David occupied the fortress and called it the city of David. And David built all around it from the Millo and \textit{inward}.\lebnote{“to the house”}}%
\verse{David \textit{continued growing stronger and stronger},\lebnote{“went going and great”}and Adonai the God of hosts was with him.}%
\verse{So Hiram, the king of Tyre, sent messengers to David, along with cedar trees and \textit{craftsmen skilled in wood and in stone masonry},\lebnote{“skilled craftsmen of wood and skilled craftsmen of stone wall”}and they built a house for David.}%
\verse{David realized that Adonai had established him as king over Israel and that he had exalted his kingdom because of his people Israel.}%
\verse{David took more concubines and wives from Jerusalem after he came from Hebron, and more sons and daughters were born to him.}%
\verse{These are the names of the ones born to him in Jerusalem: Shammua and Shobab and Nathan and Solomon;}%
\verse{Ibhar and Elishua and Nepheg and Japhia;}%
\verse{Elishama and Eliada and Eliphelet.}%
\verseWithHeading{War with the Philistines}{When the Philistines heard that they had anointed David as king over Israel, all the Philistines went up to seek David, but David heard and went down to the stronghold.\lebnote{Thought by many to be the stronghold of Masada}}%
\verse{Now the Philistines had come, and they spread out in the Valley of Rephaim.}%
\verse{And David inquired of Adonai, saying, “Shall I go up to the Philistines? Will you give them into my hands?” Adonai said to David, “Go up, for \textit{I will certainly give}\lebnote{“giving I will give”}the Philistines into your hand.”}%
\verse{So David came to Baal Perazim and defeated them there; and David said, “Adonai has burst through my enemies before me like the bursting of water.” Therefore he called the name of that place Baal Perazim.\lebnote{Baal Perazim may mean “the master of the bursting through”}}%
\verse{They had left their idols there, so David and his men carried them away.}%
\verse{\textit{Once again}\lebnote{“And they again did again”}the Philistines came up and spread out in the Valley of Rephaim.}%
\verse{So David inquired of Adonai, but he said, “You shall not go up. Rather, go around to their rear and come to them from opposite the balsam trees.}%
\verse{\textit{And it shall be}\lebnote{“And it shall happen”}that when you hear the sound of marching in the tops of the balsam trees, then pay attention, for then Adonai has gone out before you to strike down the army of the Philistines.”}%
\verse{So David did thus, just as Adonai had commanded him, and he struck down the Philistines from Geba \textit{all the way}\lebnote{“as you go up”}to Gezer.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 6
\verseWithHeading{David Brings the Ark of Adonai to Jerusalem}{David again gathered all the chosen men in Israel, thirty thousand.}%
\verse{David got up and went and all the people who were with him, from Baale-judah to bring up from there the ark of God which is called the name, the name of Adonai of hosts, \textit{upon which the cherubim sit}.\lebnote{“the cherubim are sitting on him”}}%
\verse{They loaded the ark of God on a new utility cart, and they carried it from the house of Abinadab, which was on the hill. Now Uzza and Ahio, the sons of Abinadab, were driving the new utility cart along.}%
\verse{So they brought it out from the house of Abinadab, which was on the hill with the ark of God, and Ahio was going before the ark.}%
\verse{Now David and all the house of Israel were dancing before Adonai, with all kinds of musical instruments made from ash trees, and with zithers, harps, tambourines, sistrums, and cymbals.}%
\verse{When they came to the threshing floor of Nakon, Uzza reached out to the ark of God and took hold of it, because the oxen had stumbled.}%
\verse{\textit{Then the anger of Adonai was kindled}\lebnote{“Then the nose of Adonai became hot”}against Uzza, and God struck him down there because of the indiscretion,\lebnote{Legal regulations offered by Moses did not permit an Israelite to handle the ark of the covenant in this manner (compare Exod 25:12–16)}and he died there beside the ark of God.}%
\verse{\textit{David was angry}\lebnote{“It was hot for David”}because Adonai had burst out against Uzza, and he called that place Perez-Uzza until this day.}%
\verse{But David feared Adonai on that day and said, “How can the ark of Adonai come to me?”}%
\verse{However, David was not willing to bring the ark of Adonai to himself, to the city of David, so David caused it to turn to the house of Obed-Edom the Gittite.}%
\verse{So the ark of Adonai remained in the house of Obed-Edom the Gittite for three months, and Adonai blessed Obed-Edom and all his household.}%
\verse{It was told to King David, “Adonai has blessed the household of Obed-Edom and all that is his because of the ark of God.” So David went and brought up the ark of God from the house of Obed-Edom to the city of David with jubilation.}%
\verse{It happened that when the carriers of the ark of Adonai had marched six steps that he sacrificed an ox and a fatling.}%
\verse{Now David was dancing with all his might before Adonai, and David was wearing\lebnote{Hebrew “girded with”}a linen ephod.}%
\verse{So David and all the house of Israel were bringing up the ark of Adonai with shouts of joyful acclaim and with the sound of the trumpet.}%
\verse{It happened that when the ark of Adonai came into the city of David, Michal the daughter of Saul looked down through the window and saw King David leaping and dancing before Adonai, and she despised him in her heart.}%
\verse{They brought the ark of Adonai and set it in its place in the middle of the tent which David had pitched for it. Then David offered up burnt offerings and fellowship offerings in the presence of Adonai.}%
\verse{When David had finished from the sacrificing of the burnt offerings and the fellowship offerings, he blessed the people in the name of Adonai of hosts.}%
\verse{Then he distributed to all the people, to all the multitude of Israel, from man to woman, to each: one ring-shaped piece of bread, one cake of dates, and one cake of raisins; and all the people went each to his house.}%
\verse{When David returned to bless his household, Michal the daughter of Saul came out to meet David. She said, “How the king of Israel honored himself today \textit{by uncovering himself}\lebnote{“who has uncovered himself”}before the eyes of the maids of his servants, \textit{as the total exposure of a worthless one}.”\lebnote{“as the total uncovering of one of the worthless”}}%
\verse{So David said to Michal, “In the presence of Adonai who chose me \textit{over}\lnFTY{}your father and \textit{over}\lnFTY{}his household, to appoint me as leader over the people of Adonai, over Israel, I have celebrated before Adonai.}%
\verse{I will demean myself again more than this, and I will be abased in my eyes, but with the maids whom you mentioned, with them I will be honored.”}%
\verse{So for Michal the daughter of Saul, she had no child until the day of her death.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 7
\verseWithHeading{Adonai Makes a Covenant with David}{It happened that the king settled in his house. (Now Adonai had given rest to him from all his enemies all around.)}%
\verse{And the king said to Nathan the prophet, “Look, please, I am living in a house of cedar, but the ark of God is staying in the middle of the tent.”}%
\verse{Nathan said to the king, “Go and do all that is in your heart, for Adonai is with you.”}%
\verse{But it happened \textit{that night},\lebnote{“in the night the that”}the word of Adonai came\lebnote{Hebrew “was”}to Nathan, saying,}%
\verse{“Go and tell my servant David, ‘Thus says Adonai: “Are you the one to build for me a house for my dwelling?}%
\verse{For I have not dwelt in a house from the day I brought up the \textit{Israelites}\lnAKM{}from Egypt until this day; rather, I was going about in a tent and in a tabernacle.}%
\verse{In all of my going about among all the \textit{Israelites},\lnAKM{}did I speak a word with one of the tribes of Israel whom I commanded to shepherd my people Israel, saying, ‘Why did you not build me a cedar house?’ ” ’}%
\verse{So then, thus you shall say to my servant David, ‘Thus says Adonai of hosts, “I took you from the pasture \textit{from following}\lebnote{“from behind”}the sheep to be a leader over my people, over Israel,}%
\verse{and I have been with you \textit{everywhere you went}.\lebnote{“in all that you went”}I have cut off all of your enemies from in front of you, and I will make a great name for you, as the name of the great ones who are on the earth.}%
\verse{I will make a place for my people Israel, and I will plant them\lnDJA{}so that they\lnAQG{}can dwell \textit{in their own place}.\lebnote{“under him”}They will not tremble any longer, and the children of wickedness will not afflict them again, as in the former days.}%
\verse{\textit{In the manner that}\lebnote{“And as far as” or “And from the day”}I appointed judges over my people Israel, I will give you rest from all your enemies. And Adonai declares to you that Adonai will build a house for you.}%
\verse{When your days are full and you lie down with your ancestors,\lnAGP{}I will raise up your offspring after you who will go out from your body, and I will establish his kingdom.}%
\verse{He will build a house for my name, and I will establish the throne of his kingdom forever.}%
\verse{I will be a father to him, and he will be a son for me, whom I will punish when he does wrong, with a rod of men and with blows of the \textit{human beings}.\lebnote{“children of men”}}%
\verse{But my loyal love shall not depart from him as I took it from Saul, whom I removed from before you.}%
\verse{Your house and your kingdom shall endure forever before you; your throne shall be established forever.” ’ ”}%
\verseWithHeading{David Responds to Adonai’s Covenant}{According to all these words and according to all this vision, this Nathan spoke to David.}%
\verse{Then King David went and sat before Adonai and said, “Who am I, my lord Adonai, and what is my house that you have brought me up to \textit{this place}?\lebnote{“here”}}%
\verse{Still, this was \textit{insignificant}\lebnote{“small”}in your eyes, my lord Adonai, and also you have spoken about the house of your servant from afar, and this may be the teaching of humans, my lord Adonai.}%
\verse{\textit{What more can David say to you}?\lebnote{“And what can David still add to say”}You know your servant, my lord Adonai.}%
\verse{Because of your word and according to your heart, you have done all of this great thing, in order to let your servant know.}%
\verse{Therefore you are great, my lord Adonai, for there is no one like you, and there is no god except you, in all that we have heard with our ears.}%
\verse{Who is like your people, like Israel? Israel is the one nation on earth whose God \textit{led them},\lebnote{“went”; LXX reads “led”}in order to redeem a people for himself, and to make a name for himself, and to do for you\lebnote{David switches pronouns from the third person to the second person}the great and awesome things for your land in the presence of your people whom he redeemed for himself from Egypt, from the nations and their gods.}%
\verse{You have established your people Israel for yourself as a people forever, and you, O Adonai, have become their God.}%
\verse{So then, Adonai, the word that you have spoken concerning your servant and concerning his house, confirm it forever, and do just as you have promised.}%
\verse{Your name shall be great forever, \textit{and they will say},\lebnote{“saying”}‘Adonai of hosts is God over Israel’; and the house of your servant David shall be established before you.}%
\verse{For you, O Adonai of hosts, are God of Israel! You have revealed to the ear of your servant, ‘I will build a house for you’; therefore your servant \textit{has found courage}\lebnote{“has found his heart”}to pray this prayer to you.}%
\verse{Now, my Lord Adonai, you alone are God, and your words are\lebnote{Hebrew “will be”}true. You have promised this good to your servant.}%
\verse{So then, be content and bless the house of your servant to be forever in your presence, for you, my Lord Adonai, have spoken, and because of your blessing, may the house of your servant be blessed forever.”}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 8
\verseWithHeading{David’s Military Victories Continue}{It happened afterwards that David attacked the Philistines and subdued them, and he took Metheg Ammah from the hands of the Philistines.}%
\verse{When he defeated Moab, he measured them with the cord, making them lie on the ground. He measured two cords to kill and \textit{the third cord}\lebnote{“the filling of a cord”}to let live.\lebnote{This method of selection must have been well known to the reader, but the exact procedure has been obscured over the millennia. It may be that one out of three lived through the process, only to become subservient}So Moab became servants of David, bringing tribute.}%
\verse{Then David struck down Hadadezer the son of Rehob, king of Zobah, when he went to restore his monument\lebnote{Hebrew “hand”}at the Euphrates River.}%
\verse{David captured from him one thousand and seven hundred horsemen and twenty thousand \textit{foot soldiers}.\lebnote{“men of infantry”}David hamstrung all the chariot horses, but \textit{from them}\lebnote{“from him”}he spared a hundred chariot horses.}%
\verse{When Aram of Damascus came to help Hadadezer, the king of Zobah, David killed twenty-two thousand men of Aram.}%
\verse{David placed garrisons in Aram of Damascus, so Aram became servants of David, bringing tribute. Adonai protected David everywhere he went.}%
\verse{David took the small round gold shields which had \textit{belonged}\lebnote{“been”}to the servants of Hadadezer, and he brought them to Jerusalem.}%
\verse{From Betah and from Berothai, the towns of Hadadezer, King David took very many bronze items.}%
\verse{When Toi, the king of Hamath, heard that David had defeated all the army of Hadadezer,}%
\verse{Toi sent Joram his son to King David \textit{to greet him}\lebnote{“to ask peace for himself”}and to congratulate him because he had fought against Hadadezer and defeated him; \textit{for Hadadezer had often been at war with Toi}.\lebnote{“for Hadadezer was a man of wars with Toi”}\textit{He brought with him}\lebnote{“In his hand were”}objects of silver and objects of gold and objects of bronze.}%
\verse{King David dedicated them also to Adonai along with the silver and the gold that he had dedicated from all of the nations which he had subdued;}%
\verse{from Aram and from Moab and from the \textit{Ammonites}\lnAIA{}and from the Philistines and from Amalek and from the plunder of Hadadezer the son of Rehob, the king of Zobah.}%
\verse{So David made a name for himself when he returned from defeating Aram in the Valley of Salt, eighteen thousand.}%
\verse{He put garrisons in Edom; all over Edom he put garrisons, and all of Edom became servants of David. Adonai protected David wherever he went.}%
\verse{David reigned over all of Israel, and he\lebnote{Hebrew “David”}was administering justice and righteousness for all his people.}%
\verse{Joab the son of Zeruiah was over the army, and Jehoshaphat the son of Ahilud was secretary.}%
\verse{Zadok the son of Ahitub and Ahimelech the son of Abiathar were priests and Seraiah was scribe.}%
\verse{Benaiah the son of Jehoiada was over the Kerethites and the Pelethites, and the sons of David were priests.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 9
\verseWithHeading{David Cares for Mephibosheth}{Then David said, “Is there still anyone who is left for the house of Saul that I may show loyal love to him for the sake of Jonathan?”}%
\verse{Now Saul’s household had a servant whose name was Ziba, so they summoned him to David, and the king asked him, “Are you Ziba?” He said, “\textit{At your service}!”\lebnote{“Your servant”}}%
\verse{Then the king said, “But is there still anyone of Saul’s household that I may show the loyal love of God with him?” And Ziba said to the king, “There is still a son of Jonathan who is lame in the feet.”}%
\verse{The king said to him, “Where is he?” And Ziba said to the king, “\textit{He is here}\lebnote{“Look, he”}in the house of Makir the son of Ammiel in Lo Debar.”}%
\verse{So King David sent and brought him from the house of Makir the son of Ammiel from Lo Debar.}%
\verse{So Mephibosheth the son of Jonathan, the son of Saul, came to David. And he fell on his face and did obeisance. David said, “Mephibosheth,” and he said, “\textit{I am ready to serve you}.”\lebnote{“Here is your servant”}}%
\verse{Then David said to him, “Don’t be afraid, for \textit{I will certainly show}\lebnote{“doing I will do” or “showing I will show”}loyal love to you for the sake of Jonathan your father, and I will restore to you all the lands of Saul your father. And you shall always eat food at my table.”}%
\verse{Then he\lebnote{Mephibosheth}did obeisance and said, “What is your servant that you have paid attention to the dead dog like me?”}%
\verse{David summoned Ziba the servant of Saul and said to him, “All that was Saul’s and all his household I have given to the son of your master.}%
\verse{You shall till the land for him, you and your sons and your servants; you shall bring in the produce and it shall be food for the son of your master that he may eat. But Mephibosheth the son of your master may always eat food at my table.” (Now Ziba had fifteen sons and twenty slaves.)}%
\verse{Ziba said to the king, “According to all that my master the king commands his servant, so your servant will do.” So Mephibosheth was eating at his table as one of the sons of the king.}%
\verse{Now Mephibosheth had a young son whose name was Micah, and all of the dwelling of the household of Ziba became servants for Mephibosheth.}%
\verse{And Mephibosheth was living in Jerusalem, because he was continually eating at the table of the king, even though he was lame in both of his feet.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 10
\verseWithHeading{The Ammonites Refuse David’s Loyal Love}{\textit{Afterwards the king of the Ammonites died},\lebnote{“And it happened afterwards that the king of the sons of Ammon died”}and his son Hanun ruled in his place.}%
\verse{David said, “I will show loyal love with Hanun, the son of Nahash, as his father showed loyal love with me.” So David sent to console him concerning his father, by the hand of his servants. And the servants of David came to the land of the \textit{Ammonites}.\lnAIA{}}%
\verse{But the commanders of the \textit{Ammonites}\lnAIA{}said to Hanun their master, “\textit{In your opinion},\lebnote{“In your eyes”}is David honoring your father because he has sent condolences to you? Is it not in order to search the city, to spy it out, and to overthrow it that David sent his servants to you?”}%
\verse{Then Hanun took the servants of David, and he shaved off half of their beards and cut their garments off in the middle up to their buttocks, then sent them away.}%
\verse{When they told David, he sent to meet them, for the men were greatly ashamed. And the king said, “Remain in Jericho until your beards have grown, and then you can return.”}%
\verseWithHeading{Israel Fights Ammon and Aram}{When the \textit{Ammonites}\lnAIA{}saw that they had become odious to David, the \textit{Ammonites}\lnAIA{}sent word and hired Aram Beth-Rehob and Aram-Zobah, twenty thousand infantry; and they also hired the king of Maacah, a thousand men, and the men of Tob, twelve thousand men.}%
\verse{When David heard, he sent Joab and all the army of mighty warriors.}%
\verse{The \textit{Ammonites}\lnAIA{}came out and \textit{drew up a battle formation}\lebnote{“drew up battle”}at the entrance of the gate, but Aram-Zobah and Rehob and the men of Tob and Maacah were alone in the open field.}%
\verse{When Joab saw that \textit{the battle was to be fought on two fronts},\lebnote{“the face of the battle was against him in front and in the rear”}he chose from all the members of the elite troops of Israel and \textit{lined them up for battle}\lnFUL{}to meet Aram.}%
\verse{The rest of the army he placed into the hand of his brother Abishai, who \textit{arranged them in battle lines}\lnFUL{}to meet the \textit{Ammonites}.\lnAIA{}}%
\verse{Then he said, “If Aram is stronger than I am, you must become my deliverer; but if the \textit{Ammonites}\lnAIA{}are too strong for you, then I will come to rescue you.}%
\verse{Be strong, and let us strengthen ourselves for the sake of the people and for the sake of the cities of our God. May Adonai do what is good in his eyes.}%
\verse{Joab and all the people who were with him moved forward into the battle against Aram, and they fled from before him.}%
\verse{When the \textit{Ammonites}\lnAIA{}saw that Aram had fled, they fled from before Abishai and entered the city. Then Joab returned from fighting against the \textit{Ammonites}\lnAIA{}and came to Jerusalem.}%
\verseWithHeading{The Arameans Regroup for Attack}{When the Arameans saw that they were defeated before Israel, they gathered themselves together.}%
\verse{Then Hadadezer sent and brought out the Arameans who were beyond the Euphrates, and they came to Helam. Now Shobach, the commander of the army of Hadadezer, \textit{was at their head}.\lebnote{“before the face of them”}}%
\verse{David was told, so he gathered all Israel and crossed over the Jordan and came to Helam. Aram \textit{arranged themselves in battle lines}\lnFUL{}to meet David, and they fought with him.}%
\verse{And Aram fled before Israel, and David killed from the Arameans seven hundred chariot teams and forty thousand horsemen. He struck down Shobach, the commander of his army, and he died there.}%
\verse{When all the kings, the servants of Hadadezer, saw that he had been defeated before Israel, they made peace with Israel and served them, and Aram was afraid to help the \textit{Ammonites}\lnAIA{}any longer.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 11
\verseWithHeading{David Commits Adultery with Bathsheba}{\textit{It came about in the spring},\lebnote{“And it happened at the turn of the year”}at the time \textit{kings}\lebnote{According to the reading tradition (\textit{Qere}); \textit{Kethib} has “angels” or “messengers”}go out, David sent Joab and his servants with him and all of Israel. They ravaged all of the \textit{Ammonites}\lnAIA{}and besieged Rabbah, but David was remaining in Jerusalem.}%
\verse{It happened \textit{late one afternoon}\lebnote{“at the time of the evening”}that David got up from his bed and walked about on the roof of the king’s house, and he saw a woman bathing on her\lnBBD{}roof. Now the woman \textit{was very beautiful}.\lebnote{“very good of appearance”}}%
\verse{David sent and inquired about the woman, and someone said, “Is this not Bathsheba the daughter of Eliam, the wife of Uriah the Hittite?”}%
\verse{Then David sent messengers and took her, and she came to him, and he slept with her. (Now she had been purifying herself from her uncleanness.) And she returned to her house.}%
\verse{The woman became pregnant, and she sent and told David, and she said, “I am pregnant.”}%
\verse{So David sent to Joab, “Send Uriah the Hittite to me.” So Joab sent Uriah to David.}%
\verse{Uriah came to him, and David asked \textit{how Joab and the army fared and how the war was going}.\lebnote{“as far as the peace of Joab, as far as the peace of the army, and as far as the peace of the battle”}}%
\verse{David said to Uriah, “Go down to your house, and wash your feet.” So Uriah went out from the king’s house, and a gift from the king went out after him.}%
\verse{But Uriah slept at the entrance of the king’s house with all the servants of his master and did not go down to his house.}%
\verse{They told David, “Uriah did not go down to his house.” David said to Uriah, “Are you not coming from a journey? Why did you not go down to your house?”}%
\verse{Uriah said to David, “The ark and Israel and Judah are living in the booths; and my lord Joab and the servants of my lord are camping on the surface of the open field; and I, shall I go to my house to eat and to drink and to sleep with my wife? By your life and the life of your soul, I surely will not do this thing.”}%
\verse{David said to Uriah, “Remain here \textit{today},\lebnote{“also the day”}and tomorrow I will send you away.” So Uriah remained in Jerusalem on that day and the next.}%
\verse{David invited him, and he ate and drank in his presence \textit{so that he became drunk},\lebnote{“and he made him drunk”}and he went out in the evening to lie on his bed with the servants of his lord, but he did not go down to his house.}%
\verse{And it happened in the morning, David wrote a letter to Joab, and he sent it by the hand of Uriah.}%
\verse{He had written in the letter, “Put Uriah in the front, in the face of the fiercest fighting, then draw back from behind him so that he may be struck down and die.”}%
\verse{\textit{When Joab was besieging}\lebnote{“And it happened at the besieging of Joab”}the city, he put Uriah toward the place which he knew \textit{there were valiant warriors}.\lebnote{“there were men of ability”}}%
\verse{The men of the city came out and fought with Joab. Some from the army from the servants of David fell; Uriah the Hittite also died.}%
\verse{Joab sent and told David all of the news of the battle.}%
\verse{He instructed the messenger, saying, “As you are finishing to speak all the news of the battle to the king,}%
\verse{if the anger of the king rises and he says to you, ‘Why did you go near the city to fight? Did you not know that they would shoot from atop the wall?}%
\verse{Who killed Abimelech the son of Jerub-bosheth,\lebnote{In putting words in David’s mouth, Joab alludes to the story of Abimelech the son of Gideon from Judg 9:52–55. Though Gideon was also known as Jerub-ba’al, Joab conventionally substitutes \textit{bosheth} (shame) for Ba’al to avoid naming the Canaanite deity}if not a woman who threw an upper millstone on him from atop the wall and he died at Thebez? Why did you go near the wall?’ Then you shall say, ‘Your servant Uriah the Hittite also died.’ ”}%
\verse{Then the messenger left, and he came and told David all that Joab had sent him to say.}%
\verse{The messenger said to David, “Because \textit{the men overpowered us},\lebnote{“the men were superior over us”}the men came out to us in the field, but \textit{we forced them back}\lebnote{“we were upon them”}to the entrance of the gate.}%
\verse{The archers shot at your servant from atop the wall, and some of the servants of the king died; your servant Uriah the Hittite also died.”}%
\verse{Then David said to the messenger, “Thus you shall say to Joab, ‘\textit{Do not feel badly about this matter};\lebnote{“Do not let his matter be evil in your eyes”}\textit{now one and then another}\lebnote{“for as this and as this”}the sword will devour. Intensify your attack on the city and overthrow it.’ ” And he encouraged him.}%
\verse{When the wife of Uriah heard that Uriah her husband was dead, she mourned over her husband.}%
\verse{When the mourning was over, David sent and brought her to his household, and she became his wife and bore him a son. But the thing which David had done was evil in the eyes of Adonai.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 12
\verseWithHeading{Nathan Reproves David}{So Adonai sent Nathan to David, and he came to him and said, “Two men were in a certain city; one was rich and the other was poor.}%
\verse{The rich man had very many flocks and herds,}%
\verse{but the poor man had nothing except for one small ewe lamb which he had bought. He had nurtured her, and she grew up with him and with his children together. She used to eat from his morsel and drink from his cup, and she used to lie in his lap and became like a daughter for him.}%
\verse{And a visitor came to the rich man, but he \textit{was reluctant}\lebnote{“felt sorry”}to take from his flocks or from his herds to prepare a meal for the traveler when he came to him. So he took the ewe lamb of the poor man and prepared it for the man who had come to him.”}%
\verse{Then \textit{the anger of David was kindled}\lebnote{“the nose of David became very hot”}against the man, and he said to Nathan, “\textit{As Adonai lives},\lnFSP{}the man who has done this \textit{deserves to die}!\lebnote{“is a son of death”}}%
\verse{He shall restore the ewe lamb fourfold because he has done this thing, and because he had no pity.”}%
\verse{Then Nathan said to him, “You are the man! Thus says Adonai the God of Israel: ‘I anointed you as king over Israel, and I delivered you from the hand of Saul.}%
\verse{I gave you the household of your master and the women of your master into your lap. I also gave you the house of Israel and Judah; if that had been too little, I would have added to you \textit{much more}.\lebnote{“as here and as here”}}%
\verse{Why have you despised the word of Adonai by doing evil in his eyes?\lebnote{Thus \textit{Kethib}; the reading tradition (\textit{Qere}) reads “my eyes”}Uriah the Hittite you have struck down with the sword, and his wife you have taken to yourself as wife! You have killed him with the sword of the \textit{Ammonites}!\lnAIA{}}%
\verse{So then, a sword will not turn away from your house forever, because you have despised me and have taken the wife of Uriah the Hittite to be your wife!’}%
\verse{Thus says Adonai, ‘Look, I am going to raise up evil against you from within your house, and I will take your women before your eyes, and I will give them to your neighbor, and he shall sleep with your wives \textit{in broad daylight}.\lebnote{“before the eyes of this sun”}}%
\verse{Though you did this in secret, I will do this thing before all of Israel \textit{in broad daylight}!’ ”\lebnote{“before the sun”}}%
\verseWithHeading{David Repents, But the Child Dies}{Then David said to Nathan, “I have sinned against Adonai!”\lebnote{See Pss 32:1–5; 51:1–13}Nathan said to David, “Adonai has also forgiven your sin; you shall not die.}%
\verse{But because you have \textit{utterly scorned}\lebnote{“scorning you have scorned.” The term “enemies of” in the MT is likely a textual corruption and is deleted here}Adonai in this matter, the son born for you \textit{will certainly die}.”\lebnote{“dying will die” (compare Gen 2:17)}}%
\verse{Then Nathan went to his house, and Adonai struck the child that the wife of Uriah bore for David, and he became ill.}%
\verse{David pleaded with God on behalf of the boy and David fasted. He went to spend the night and lay upon the ground.}%
\verse{The elders of his household stood over him to lift him up from the ground, but he was not willing, and he did not eat any food with them.}%
\verse{It happened on the seventh day that the child died, and the servants of David were afraid to tell him that the child was dead, for they said, “Look, when the child was alive, we spoke to him, but he would not listen to our voice. How can we tell him, ‘The child is dead’? He may do something evil.”}%
\verse{When David saw that his servants were whispering together, he realized that the child was dead. Then David said to his servants, “Is the child dead?” And they said, “He is dead.”}%
\verse{David stood up from the ground and washed and anointed himself and changed his clothing. Then he went to the house of Adonai and worshiped, and he went to his own house. He asked, so they served him food, and he ate.}%
\verse{Then his servants said to him, “What is this thing that you have done? While the child was alive, you fasted and wept; now that the child has died, you get up and eat food!”}%
\verse{He said, “When the child was still alive, I fasted and I wept because I thought, ‘Who knows? Adonai may have mercy on me that the child will live.’}%
\verse{But now he is dead. Why should I be fasting? Am I able to return him again? I am going to him, but he cannot return to me.”}%
\verse{David consoled Bathsheba his wife, and he went to her and slept with her. She bore a son, and he called\lebnote{So \textit{Kethib}; \textit{Qere} reads “she called”}him Solomon, and Adonai loved him.}%
\verse{He sent word by the hand of Nathan the prophet, so he called him Jedidiah\lebnote{Jedidiah means “beloved of Adonai”}because of Adonai.}%
\verseWithHeading{Battle with the Ammonites}{And Joab fought against Rabbah of the \textit{Ammonites},\lnAIA{}and he captured \textit{the royal city}.\lebnote{“city of the kingship”}}%
\verse{Then Joab sent messengers to David and said, “We have fought against Rabbah, and we captured the city of the waters.}%
\verse{So then, gather the remainder of the army and encamp against the city and capture it, lest I capture the city and my name be proclaimed over it.”}%
\verse{So David gathered all of the army, and he went to Rabbah and fought against it and captured it.}%
\verse{He took the crown of their king from his head. (Now its weight was a talent of gold, and there was a precious stone in it and it was put on David’s head.) He brought out the plunder of the city \textit{in great abundance}.\lebnote{“very many”}}%
\verse{He also brought out the people who were in it and put them to the saws and to the iron picks and to the iron axes, and he sent them to the place of the brickmakers. Thus he used to do to all the cities of the \textit{Ammonites},\lnAIA{}and he and all of the army returned to Jerusalem.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 13
\verseWithHeading{Amnon Assaults His Sister Tamar}{It happened afterwards that Absalom the son of David had a beautiful sister whose name was Tamar, and Amnon the son of David fell in love with her.}%
\verse{And Amnon \textit{was so frustrated that he felt ill}\lebnote{“was distressed to the point of sickness”}because of Tamar his sister, because she was a virgin, and it was too difficult in Amnon’s eyes to do anything with her.}%
\verse{Now Amnon had a friend whose name was Jonadab the son of Shimeah, the brother of David. (Now Jonadab was a very crafty man.)}%
\verse{And he said to him, “Why are you so sullen \textit{every morning},\lebnote{“in the morning, in the morning”}O son of the king? Will you not tell me?” And Amnon said to him, “I am in love with Tamar, the sister of my brother Absalom.”}%
\verse{Then Jonadab said to him, “Lie down on your bed and appear ill. If your father comes to see you, you shall say to him, ‘Please let Tamar my sister come and give me food to eat, and let her prepare the food before my eyes, in order that I may see it and eat from her hand.’ ”}%
\verse{So Amnon lay down and pretended to be ill, and the king came to see him. Amnon said to the king, “Please let Tamar my sister come, and let her bake two cakes before my eyes that I may eat from her hand.”}%
\verse{So David sent to the house for Tamar, saying, “Please go to the house of Amnon your brother and prepare food for him.”}%
\verse{Tamar went to the house of Amnon her brother. Now he was lying down, and she took the dough and kneaded it and made cakes before his eyes, and she baked the cakes.}%
\verse{Then she took the pan and poured it out before him, but he refused to eat. Then Amnon said, “Let all the men go out from me.” So all the men went out from him.}%
\verse{Then Amnon said to Tamar, “Bring the food to the private room that I may eat from your hand.” So Tamar took the cakes which she had made and brought them to Amnon her brother in\lnBYY{}the private room.}%
\verse{When she brought them near to him to eat, he took hold of her and said to her, “Come, lie with me, my sister!”}%
\verse{Then Tamar said to him, “No, my brother! Do not force me, for such a thing has not been done in Israel. Do not do this disgraceful thing!}%
\verse{As for me, where should I take my disgrace? You will be as one of the fools in Israel. So please, speak to the king, for he will not withhold me from you.”}%
\verse{But he was not willing to listen to her voice. He was stronger than she, and he forced her and lay with her.}%
\verse{Then Amnon hated her \textit{very deeply},\lebnote{“with a great hatred”}for the hatred with which he hated her was greater than the love with which he had loved her. So Amnon said to her, “Get up and go.”}%
\verse{She said to him, “No, because this evil in sending me away is greater than the other you have done to me.” But he was not willing to listen to her.}%
\verse{Then he called his young man who was serving him and said, “Please send this woman from me to the outside, and bolt the door behind her!”}%
\verse{Now there was a long-sleeved robe on her, for so they clothed the daughters of the king who were virgins, in robes. His servant put her outside, and he bolted the door behind her.}%
\verse{Tamar put ashes on her head, and she tore the long-sleeved robe which was on her. She put her hand on her head, and \textit{she went away, crying out as she went}.\lebnote{“she went, going and crying aloud”}}%
\verse{Absalom her brother said to her, “Was Amnon your brother with you? But now, my sister, be quiet; he is your brother. \textit{Do not take this matter to heart}.”\lebnote{“Do not place/set your heart to this matter”}So Tamar remained a desolate woman in the house of Absalom her brother.}%
\verseWithHeading{Absalom Kills Amnon}{Now King David heard all these things, and \textit{he became very angry}.\lebnote{“it became very hot for him”}}%
\verse{Absalom did not speak with Amnon \textit{either bad or good},\lebnote{“as far as from bad and up to good”}for Absalom hated Amnon \textit{over the matter when}\lebnote{“because of the thing that”}he raped Tamar his sister.}%
\verse{\textit{About two full years later},\lebnote{“And it happened at two years of days”}Absalom’s shearers were in Baal Hazor, which is near Ephraim, and Absalom summoned all the sons of the king.}%
\verse{Then Absalom went to the king and said, “Look, here are your servant’s shearers; please let the king and his servants go with your servant.}%
\verse{The king said to Absalom, “No my son, not all of us shall go, so that we not be a burden to you.” And he urged him, but he was not willing to go, but he blessed him.}%
\verse{So Absalom said, “But will you not let Amnon my brother go with us?” And the king said to him, “Why should he go with you?”}%
\verse{But Absalom pressed him, so he sent Amnon with him and all of the sons of the king.}%
\verse{Absalom commanded his servants, saying, “Please watch. At the moment the heart of Amnon is \textit{tipsy}\lebnote{“good”}with wine, then I shall say to you, ‘Strike Amnon down,’ and you shall kill him! Don’t be afraid. Is it not I myself who has commanded you? Be courageous and be \textit{valiant!}\lebnote{“as sons of ability”}}%
\verse{So Absalom’s servants did to Amnon just as Absalom commanded, and all the sons of the king got up, and each mounted his mule and fled.}%
\verse{\textit{While they were on the way},\lebnote{“And it happened they were on the way”}the message came to David, “Absalom has killed all the sons of the king, and not one of them was left.”}%
\verse{Then the king rose and tore his garments, and he lay on the ground, and all his servants standing by were tearing their garments.}%
\verse{Then Jonadab the son of Shimeah, the brother of David, responded and said, “My lord should not think that all the young men, the sons of the king, are dead, because only Amnon is dead. \textit{Absalom was talking about it},\lebnote{“For it was on the mouth of Absalom”}as it was being determined from the day he\lebnote{Amnon}raped Tamar his sister.}%
\verse{So then, let not my lord the king set his heart on this matter, thinking, ‘All the sons of the king are dead,’ for only Amnon alone is dead.”}%
\verseWithHeading{Absalom Flees}{So Absalom fled, and the young man who was keeping watch lifted up his eyes and saw, and there were many people coming from the road behind him from the side of the mountain.}%
\verse{Jonadab said to the king, “Look, the sons of the king have come. According to the word of your servant, so it has come about.”}%
\verse{When he finished speaking, look, the sons of the king came and they lifted up their voice and wept. Also, the king and all his servants wept a very great weeping.}%
\verse{Absalom fled and went to Talmai the son of Ammihur, the king of Geshur. David mourned over his son \textit{day after day}.\lnDZX{}}%
\verse{But Absalom had fled and went to Geshur, and he was there three years.}%
\verse{King David longed to go out to Absalom, for he was consoled that Amnon had died.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 14
\verseWithHeading{Joab Plots to Reconcile David with Absalom}{Joab the son of Zeruiah realized that the mind of the king was on Absalom.}%
\verse{So Joab sent to Tekoa and took from there a wise woman, and he said to her, “Please pretend to mourn and put on garments of mourning. You should not anoint yourself with oil, and you must act like this woman who has been mourning over the dead for \textit{a long time}.\lebnote{“many days”}}%
\verse{Then you must go to the king and speak to him according to this word.” Thus Joab put the words in her mouth.}%
\verse{So the Tekoite woman spoke to the king, and she fell on her face to the ground and did obeisance. She said, “Help me, O king!”}%
\verse{Then the king asked her, “\textit{What do you want}?”\lebnote{“What is for you”}And she said, “Truly I am a widow, and my husband is dead.}%
\verse{Your servant had two sons, and they both fought in the open field, and there was no one \textit{to part them}.\lebnote{“to save between them”}One struck the other and killed him.}%
\verse{And look, all of the family has risen up against your servant, and they said, ‘Give up the one who struck his brother, that we may kill him in exchange for the life of his brother whom he murdered. We will also wipe out the heir,’ and so they would put out my embers which remain, by not preserving for my husband a name and a remnant on the face of the earth.”}%
\verse{Then the king said to the woman, “Go to your house, and I myself will give the command concerning you.”}%
\verse{The Tekoite woman said to the king, “On me, my lord the king, is the guilt, and on the house of my father, but the king on\lnDZA{}his throne is innocent.”}%
\verse{The king said, “Whoever has spoken to you, bring him to me, and he will not touch you again.”}%
\verse{Then she said, “Please may the king remember Adonai your God, \textit{to prevent the increase of blood avengers who kill},\lebnote{“from making numerous the avenger of blood to kill”}so that they not wipe out my son.” He said, “\textit{As Adonai lives},\lnFSP{}surely not one hair shall fall from your son to the ground.”}%
\verse{The woman said, “Please let your servant speak a word to my lord the king.” And he said, “Speak.”}%
\verse{The woman said, “But why have you plotted like this against the people of God? By speaking this word, he is guilty not to bring back his banished one.}%
\verse{For \textit{we must certainly die},\lebnote{“dying we must die”}and we are as the waters spilled to the ground which cannot be gathered. God will not take a life but devises plans for a banished person not to be cast out from him.}%
\verse{Now I have come to speak this word to my lord the king, because the people made me afraid, and your servant thought, ‘I will speak to the king, perhaps the king will grant\lebnote{Hebrew “do/make”}the request of his servant.}%
\verse{For the king will listen, to deliver his servant from the hand of the man who seeks to destroy me and my son together from the inheritance of God.’}%
\verse{Your servant also thought, ‘May the word of my lord the king \textit{bring rest},\lebnote{“become a rest”}for as an angel of God, so is my lord the king, \textit{to sense what is good and what is bad}.’\lebnote{“to hear the good and the bad”}May Adonai your God be with you.”}%
\verse{The king answered and said to the woman, “Please do not withhold from me a thing which I am about to ask you.” The woman said, “Please let my lord the king speak.”}%
\verse{The king asked, “Was the hand of Joab with you in all of this?” The woman answered and said, “\textit{As your soul lives},\lebnote{“The life of your soul”}my lord the king, surely one cannot go to the right or to the left from all that my lord the king has spoken. Yes, your servant Joab himself commanded me, and he put all of these words in the mouth of your servant.}%
\verse{In order \textit{to change the situation},\lebnote{“to turn the face of the thing”}your servant Joab did this thing. But my lord has wisdom, as the wisdom of an angel of God, to know all that is on the earth.”}%
\verse{Then the king said to Joab, “Look, please, I will grant this thing. Go and bring back the young man Absalom.”}%
\verse{Joab fell with his face to the ground and did obeisance. And he blessed the king, and he\lebnote{Hebrew “Joab”}said, “Today your servant knows that I have found favor in your eyes, my lord the king, in that the king has granted the request of his servant.”}%
\verseWithHeading{Absalom Returns to Jerusalem}{Then Joab got up and went to Geshur and brought Absalom to Jerusalem.}%
\verse{The king said, “Let him go over to his house, and he may not see my face.” So Absalom went over to his house, and did not see the face of the king.}%
\verse{As far as Absalom, there was not a more handsome man in all of Israel to admire so much; from the sole of his foot up to his crown, there was no physical defect on him.}%
\verse{When he shaved his head, it would happen \textit{every year},\lebnote{“from the end of days for the days”}which he did because it was heavy on him, he would shave it off and weigh the hair of his head: two hundred shekels \textit{by the king’s weight}.\lebnote{“by the stone of the king”}}%
\verse{Three sons were born to Absalom and one daughter, whose name was Tamar. She was a woman beautiful of appearance.}%
\verse{Absalom lived in Jerusalem \textit{two full years},\lebnote{“two years of days”}but he did not see the face of the king.}%
\verse{So Absalom sent for Joab, in order that he send him to the king, but he was not willing to go to him. He sent again a second time, but he was not willing to go.}%
\verse{So he said to his servants, “Look at the tract of land of Joab \textit{next to mine},\lebnote{“toward my hand”}for he has barley plants there. Go, set it ablaze with fire.” So the servants of Absalom set the tract of land ablaze with fire.}%
\verse{Then Joab got up and went to Absalom, to the house, and said to him, “Why have your servants set my tract of land ablaze with fire?”}%
\verse{Absalom said to Joab, “Look, I have sent to you, saying, ‘Come here that I may send you to the king to say, “Why have I come from Geshur? It would be better for me if I were still there.” ’ So then, let me see the face of the king; if there is guilt in me, then let him kill me.”}%
\verse{So Joab went to the king and he told him. Then he summoned Absalom, and he came to the king, and he bowed down to him with his face to the ground before the king. Then the king kissed Absalom.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 15
\verseWithHeading{Absalom Leads a Rebellion}{It happened afterward that Absalom made himself a chariot with horses and fifty men running before him.}%
\verse{Absalom used to rise early in the morning, and he stood \textit{beside}\lebnote{“on hand of”}the road at the gate; \textit{anyone}\lnFVD{}who had a legal dispute to bring to the king for judgment Absalom would call to him and say, “\textit{Where are you from?}”\lebnote{“Where from this city are you?”}And he would say, “Your servant is from one of the tribes of Israel.”}%
\verse{Then Absalom would say to him, “Look, your case is good and right, but for you there is no hearing with the king.”}%
\verse{Then Absalom would say, “\textit{Oh, that someone would}\lebnote{“Who will …?”}appoint me as judge in the land, that \textit{anyone}\lnFVD{}might come to me who had a legal dispute or a case, and I would give him justice.”}%
\verse{It happened whenever anyone drew near to do obeisance to him, that he would stretch his hand and take hold of him and kiss him.}%
\verse{Absalom did like this to all of Israel who came to the king for judgment; so Absalom stole the hearts\lnEOY{}of the people of Israel.}%
\verse{It happened at the end of four\lebnote{So LXX, followed by Syriac and Vulgate; MT has “forty”}years that Absalom said to the king, “Please let me go and pay my vow which I have made to Adonai in Hebron,}%
\verse{for your servant made a vow while I was staying in Geshur in Aram, saying, ‘\textit{If Adonai will indeed let me return}\lebnote{(following \textit{Qere}) “If returning Adonai will return me”}to Jerusalem, then I will worship Adonai.’ ”}%
\verse{Then the king said to him, “Go in peace.” So he got up and went to Hebron.}%
\verse{Then Absalom sent scouts throughout all the tribes of Israel, saying, “When you hear the sound of the trumpet, you shall shout ‘Absalom has become king in Hebron!’ ”}%
\verse{Two hundred men from Jerusalem went with Absalom as invited guests, going in their innocence as they did not know anything.}%
\verse{Absalom sent for Ahithophel the Gilonite, the advisor of David, from his city Giloh, while he offered the sacrifices. The conspiracy \textit{grew in strength},\lebnote{“became strong”}and the people were going and \textit{increasing}\lebnote{“many”}with Absalom.}%
\verseWithHeading{David Flees from Jerusalem}{Then the messenger came to David, saying, “The hearts\lnEOY{}of the men of Israel have gone after Absalom.”}%
\verse{Then David said to all his servants who were with him in Jerusalem, “Get up and let us flee, for there will be no escape for us from Absalom! Hurry to go, otherwise he will come quickly and overtake us! And he will bring disaster on us and evil! He will attack the city with the edge of the sword!”}%
\verse{Then the servants of the king said to the king, “According to all that my lord the king chooses, \textit{your servants are ready to act}!”\lebnote{“here are your servants”}}%
\verse{The king went out with\lnDZA{}all his household \textit{following him},\lnFVI{}but the king left behind ten concubines to look after the house.}%
\verse{So the king went out with\lnDZA{}all the people \textit{following him},\lnFVI{}and they stopped \textit{at the last house}.\lebnote{“at the house of the distance”}}%
\verse{All his servants were passing \textit{by him}:\lebnote{“on his hand”}all of the Kerethites and all of the Pelethites and all of the Gittites—six hundred men \textit{who had followed him}\lebnote{“who had come at his feet”}from Gath—passing \textit{before the king}.\lebnote{“on the face of the king”}}%
\verse{The king said to Ittai the Gittite, “Why are you also coming with us? Return and stay with the king, for you are a foreigner; moreover, you are an exile. \textit{You are far from your place}.\lebnote{“You as far as your place”}}%
\verse{Yesterday when you came and \textit{today},\lnDXP{}I have caused you to wander by going with us. Now I am going to where I am going; return and let your brothers return. May loyal love and faithfulness be with you.”}%
\verse{But Ittai answered the king and said, “\textit{As Adonai lives}\lnFSP{}and \textit{my lord the king lives},\lebnote{“the life of my lord the king”}surely in the place wherever my lord the king shall be, if for death or if for life, surely there your servant will be.”}%
\verse{Then David said to Ittai, “Go and pass over.” So Ittai the Gittite passed over and all his men and all of the little children who were with him.}%
\verse{All of the land was weeping with a loud voice as all the people were passing by and the king was crossing through the Wadi Kidron, and all the people were passing \textit{on the road to the wilderness}.\lebnote{“on the face of the road with the wilderness”}}%
\verseWithHeading{The Priests Offer Sacrifices for David}{Suddenly Zadok was there, and all of the Levites with him, carrying the ark of the covenant of God. They set the ark of God down, and Abiathar offered sacrifices until all the people passed out of the city.}%
\verse{The king said to Zadok, “Let the ark of God return to the city. If I find favor in the eyes of Yawheh, he will let me return and let me see him in his dwelling place.}%
\verse{But if he says, ‘I take no pleasure in you,’ then \textit{I am ready}.\lebnote{“here I am”}Let him do to me that which is good in his eyes.}%
\verse{Then the king said to Zadok the priest, “Are you a seer? Return to the city in peace, with Ahimaaz your son and Jonathan the son of Abiathar, your two sons with you.}%
\verse{See, I am waiting at the fords of the wilderness until a word comes from you all to inform me.”}%
\verse{So Zadok and Abiathar returned the ark of God to Jerusalem and they remained there.}%
\verseWithHeading{Hushai Offers to Serve King David}{Now David was going up on the Ascent of the Olives, \textit{weeping as he went},\lnFVN{}with his head covered and going barefoot. All the people who were with him each covered their head \textit{and wept as they went}.\lnFVN{}}%
\verse{Now David was told, “Ahithophel was among the conspirators with Absalom.” Then David said, “Please frustrate the counsel of Ahithophel, O Adonai.”}%
\verse{It happened that as David was coming up to the summit where he used to worship God, suddenly Hushai the Arkite was there to meet him. His coat was torn and dirt was on his head.}%
\verse{David said to him, “If you move on with me, you will be a burden to me,}%
\verse{but if you want to return to the city and say to Absalom, ‘I am your servant, O king. I used to be a servant of your father, but from then and now I will be your servant,’ then you can frustrate the counsel of Ahithophel for me.}%
\verse{Will not Zadok and Abiathar the priests be with you there? It shall be that all the words you hear from the house of the king you shall tell Zadok and Abiathar the priests.}%
\verse{Look, there with them are their two sons, Ahimaaz of Zadok and Jonathan of Abiathar. You shall send to me \textit{by means of them}\lebnote{“by their hand”}all the words that you hear.”}%
\verse{So Hushai the friend of David came to the city as Absalom was entering Jerusalem.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 16
\verseWithHeading{Ziba Brings Provisions}{Now David passed a little from beyond the summit, and suddenly Ziba the servant of Mephibosheth was there to meet him with a pair of saddled donkeys; on them were two hundred loaves of bread and a hundred raisin cakes, with a hundred summer fruits and a skin of wine.}%
\verse{The king said to Ziba, “\textit{What do you want to accomplish by bringing these}?”\lebnote{“What are these to you?”}And Ziba said, “The donkeys are for the king’s household to ride on; the bread and the summer fruit are for the young men to eat, and the wine is for the faint in the wilderness to drink.}%
\verse{Then the king said, “Where is the son of your lord?” And Ziba said to the king, “He is living in Jerusalem for he said, ‘Today the house of Israel shall return the kingdom of my father to me.’ ”}%
\verse{The king said to Ziba, “Look, all that was Mephibosheth’s is yours.” Ziba said, “I hereby do obeisance; may I find favor in your eyes, my lord the king.”}%
\verseWithHeading{Shimei Curses David}{King David came up to Bahurim and suddenly a man from there was coming out from the family of the house of Saul, and his name was Shimei the son of Gera. \textit{He was cursing as he came out}.\lebnote{“He was coming out, coming out and cursing”}}%
\verse{And he threw stones at David and at all the servants of King David and at all the people and at all the mighty warriors on his right and on his left.}%
\verse{Shimei said while cursing him, “Go out, go out, you \textit{man of bloodshed},\lebnote{“man of bloods”}you \textit{man of wickedness}.\lebnote{“and man of the wickedness”}}%
\verse{Adonai has returned on you all the blood of the household of Saul \textit{whom you have supplanted as king},\lebnote{“who you have reigned in place of him”}and Adonai has given the kingship into the hand of Absalom your son. Look, you are in disaster for you are a man of blood.”}%
\verse{Then Abishai the son of Zeruiah said to the king, “Why should this dead dog curse my lord the king? Please let me go over and take off his head.”}%
\verse{The king said, “\textit{What do we have in common},\lebnote{“What is for me and what is for you”}sons of Zeruiah? If\lebnote{Hebrew “Because”}he curses because Adonai has said to him ‘Curse David,’ who can say, “Why have you done this?”}%
\verse{David said to Abishai and to all his servants, “Look, my son who came out of my loins\lebnote{Hebrew “intestines”}is seeking my life. Now \textit{as far as}\lebnote{“and even that now”}this Benjaminite, leave him alone and let him curse, for Adonai has spoken to him.}%
\verse{Perhaps Adonai will look \textit{in my eye}\lebnote{According to the reading tradition (\textit{Qere}); \textit{Kethib} has “at my guilt”}and repay good for me in place of his curse this day.}%
\verse{Then David and his men went on the road, with Shimei going on the side of the hill beside him, \textit{cursing as he went}.\lebnote{“going and cursing”}He threw stones beside him and threw\lebnote{Hebrew “flung”}dust in the air.}%
\verse{When the king and all of the people who were with him arrived, he was weary, so he recovered there.}%
\verseWithHeading{Hushai Comes to Absalom}{Now Absalom and all the people, the men of Israel, had come to Jerusalem, and Ahithophel was with him.}%
\verse{When Hushai the Arkite the friend of David came to Absalom, Hushai said to Absalom, “Long live the king, long live the king!”}%
\verse{Absalom said to Hushai, “This is your loyal love with your friend? Why did you not go with your friend?”}%
\verse{Then Hushai said to Absalom, “No, rather, whom Adonai and this people and all the men of Israel have chosen, to him\lebnote{According to the reading tradition (\textit{Qere}); \textit{Kethib} has “no”}I will be, and with him I will remain.}%
\verse{\textit{Furthermore},\lebnote{“And the second thing”}for whom have I served? Is it not in the presence of his son that I have served before your father? \textit{So shall I serve you}!”\lebnote{“So shall I be in your presence”}}%
\verse{Then Absalom said to Ahithophel, “Give your advice. What shall we do?”}%
\verse{And Ahithophel said to Absalom, “Go to the concubines of your father whom he left behind to watch over the house, then all of Israel will hear that you made yourself odious to your father, and \textit{all of your followers will be motivated}!”\lebnote{“the hands of all who are with you will be strengthened”}}%
\verse{Then they pitched a tent for Absalom on the roof, and he went in to the concubines of his father before the eyes of all Israel.}%
\verse{The counsel that Ahithophel gave in those days was regarded as when a man\lebnote{\textit{Qere} reads “a man inquired”; \textit{Kethib} omits “a man”}inquired of the word of God, so all the counsel of Ahithophel was esteemed both by David and by Absalom.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 17
\verseWithHeading{Hushai Frustrates the Counsel of Ahithophel}{Then Ahithophel said to Absalom, “Please let me choose twelve thousand men, and I will set out and pursue after David \textit{tonight}.\lebnote{“the night”}}%
\verse{I will come upon him while he is weary and \textit{weak}.\lebnote{“slack of hands”}I will startle him, and all of the people who are with him will flee. Then I will strike down the king while he is alone.}%
\verse{Then I will return all the people to you; when all have returned, the man whom you are seeking will be dead,\lebnote{According to the MT; the LXX has “and I will return all the people to you the way a bride returns to her husband, except the life of the one man whom you are seeking”}but all the people will be safe.”}%
\verse{The word was right in the eyes of Absalom and in the eyes of all the elders of Israel.}%
\verse{Then Absalom said, “Please call also for Hushai the Arkite, and let us hear \textit{what he has to say}\lebnote{“what is in his mouth”}also.”}%
\verse{So Hushai came to Absalom, who said to him, “Shall we do according to this word that Ahithophel has spoken? If not, then you tell us.”}%
\verse{Then Hushai said to Absalom, “The counsel that Ahithophel gave is not good \textit{at this time}.”\lebnote{“at this occurrence”}}%
\verse{Hushai continued, “You know your father and his men, that they are mighty warriors and they \textit{are enraged}\lebnote{“are bitter of soul”}as a bear robbed of her offspring in the field. Your father is a man of war, so he will not spend the night with the troops.}%
\verse{Now he has hidden himself in one of the caves or in one of the places. At the moment he falls on them\lebnote{I.e., Absalom’s troops}the first time, \textit{whoever hears}\lebnote{“and the one who hears hears”}the report will say, ‘There has been a defeat among the people who follow after Absalom.’}%
\verse{And he is also a \textit{valiant warrior},\lnFVP{}whose heart is like the heart of the lion. He\lebnote{i.e., one of Absalom’s troops}will \textit{certainly melt}\lebnote{“melting he will melt”}with fear, for all Israel knows that your father is a mighty warrior and those who are with him are \textit{valiant warriors}.\lnFVP{}}%
\verse{I give the advice that all of Israel from Dan to Beersheba should be completely gathered to you, as the sand which is on the seashore for abundance, with \textit{you personally}\lebnote{“your face”}going into the battle.}%
\verse{Then we will come to him in one of the places where he may be found, and we shall come upon him as the dew falls on the ground. He and all the men who are with him will not survive, not even one!}%
\verse{Even if he withdraws to a city, all Israel will bring up ropes to that city, and they will drag him away to the valley until there is not even a pebble to be found.”}%
\verse{Then Absalom and all the men of Israel said, “The advice of Hushai the Arkite is better than the advice of Ahithophel.” (Now Adonai had ordained to frustrate the good counsel of Ahithophel in order for Adonai to bring misery upon Absalom).}%
\verseWithHeading{Hushai Sends Word to David}{Hushai said to Zadok and to Abiathar the priests, “\textit{Thus and so}\lnFVR{}Ahithophel advised Absalom and the elders of Israel, but \textit{thus and so}\lnFVR{}I have advised.}%
\verse{So then, send quickly and tell David, ‘Don’t spend the night at the fords of the wilderness! Moreover, by all means cross over lest the king and all the people who are with him be swallowed up.’ ”}%
\verse{Now Jonathan and Ahimaaz were staying at En Rogel, so a servant girl used to go and tell them, then they would go and tell King David, for they were not able to be seen coming to the city.}%
\verse{But a young man saw them and told Absalom, so both of them went quickly and came to the house of a man at Bahurim. Now he had a well in his courtyard, so they went down there.}%
\verse{Then the woman took and spread a covering over the opening of the well; then she spread out dried grain on it, so nothing was discovered.}%
\verse{When the servants of Absalom came to the woman at the house, they said, “Where are Ahimaaz and Jonathan?” And she said to them, “They crossed over the brook of water.” So they searched but could not find them, so they returned to Jerusalem.}%
\verse{After they left, they came up from the well and went and told King David. They said to David, “Set out and cross over the water quickly, for thus Ahithophel has advised against you.”}%
\verse{So David and all the people who were with him set out, and they crossed over the Jordan until morning light until \textit{there was no one}\lebnote{“until one was not missing”}missing who had not crossed over the Jordan.}%
\verse{When Ahithophel saw that his advice was not followed, he saddled the donkey, and he set out and went up to his house in his city. \textit{After he set his house in order},\lebnote{“And he commanded to his household”}he hanged himself, and he died and was buried in the tomb of his ancestors.\lnAGP{}}%
\verse{Now David had come to Mahanaim, and Absalom had crossed over the Jordan, he and all the men of Israel with him.}%
\verse{Absalom had appointed Amasa in place of Joab over the army. Now Amasa was the son of a man whose name was Ithra the Israelite, \textit{who had married}\lebnote{“who went to”}Abigail the daughter of Nahash the sister of Zeruiah, the mother of Joab.}%
\verse{Israel and Absalom camped in the land of Gilead.}%
\verse{Just as David had arrived in Mahanaim, Shobi the son of Nahash from Rabbah of the \textit{Ammonites}\lnAIA{}and Makir the son of Ammiel from Lo Debar and Barzillai the Gileadite from Rogelim}%
\verse{brought beds and basins and objects of pottery, as well as wheat, barley, flour, roasted grain, beans, lentils,}%
\verse{honey, curds, sheep, cheese, and cattle for David and for the people who were with him to eat. For they had thought, “The troops are hungry and weary and thirsty in the wilderness.”}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 18
\verseWithHeading{Absalom Dies in Battle}{Then David mustered the people who were with him, and he appointed over them commanders of thousands and commanders of hundreds.}%
\verse{David sent forth a third of the troops \textit{under the command of Joab},\lebnote{“in the hand of Joab”}and a third \textit{under the command of Abishai},\lebnote{“in the hand of Abishai”}the brother of Joab, the son of Zeruiah, and the remaining third \textit{under the command of Ittai}\lebnote{“in the hand of Ittai”}the Gittite. And the king said to the troops, “I, even I, \textit{will certainly go out}\lebnote{“going out I will go out”}with you.”}%
\verse{Then the troops said, “You will not go out, for \textit{if we must flee},\lebnote{“fleeing we must flee”}then \textit{they will not care about us};\lnFVV{}even if half of us die, \textit{they will not care about us},\lnFVV{}but now, you are like ten thousand of us. And so then, it is better for us that you be in the city to help.”}%
\verse{The king said to them, “I will do what seems good in your eyes.” So the king stood at the side of the gate while all of the troops went out by hundreds and by thousands.}%
\verse{The king ordered Joab, Abishai, and Ittai, “With respect to the young man Absalom, deal gently for me.” And all the troops heard when the king ordered all of the commanders concerning the matter of Absalom.}%
\verse{The army went out to the field to meet Israel, and the battle was fought in the forest of Ephraim.}%
\verse{The army of Israel was defeated there before the servants of David, and the defeat there was great on that day: twenty thousand.}%
\verse{The battle there was spreading over the surface of all the land, and \textit{the forest devoured more among the army than}\lebnote{“the forest made numerous to eat among the army more than”}the sword did on that day.}%
\verse{Absalom was found in the presence of the servants of David as he was riding on the mule. The mule went under the thicket of the great oak tree, and his head was caught in the tree. He was left hanging between heaven and earth, and the mule which was under him went on.}%
\verse{When a certain man saw it, he told Joab, and he said, “Look, I saw Absalom hanging in the oak tree!”}%
\verse{Then Joab said to the man who was telling him, “Look, if you saw, why did you not strike him down to the ground there? \textit{I would have gladly given you}\lebnote{“And upon me to give to you”}ten pieces of silver and a leather belt.”}%
\verse{The man said to Joab, “\textit{Even if I felt the weight}\lebnote{“I am not weighing out on my palms”}of a thousand pieces of silver in my palms, I would not have sent my hand against the son of the king, for in our ears the king commanded you and Abishai and Ittai, ‘Whoever you may be, protect the young man Absalom.’}%
\verse{If I had dealt treacherously against his life, and there is not any matter hidden from the king, \textit{you would have presented yourself aloof}.”\lebnote{“you would have presented yourself from before”}}%
\verse{Joab said, “No longer will I wait in your presence.” Then he took three spears in his hand and thrust them into the heart of Absalom while he was still alive in the oak tree.}%
\verse{Then ten of the young men who bore the weapons of Joab surrounded him and struck and killed Absalom.}%
\verse{Then Joab blew on the trumpet and the troops returned from pursuing after Israel, for Joab kept back the troops.}%
\verse{They took Absalom and they threw him into the large pit in the forest and raised a very great heap of stones over him. Then all of Israel fled, each to his tent.}%
\verse{(Now Absalom had taken and set up for himself in his lifetime a stone pillar that is in the valley of the king, because he said, “I have no son in order to remember my name,” and he called the stone pillar by his name. It is called the monument of Absalom until this day).}%
\verse{Now Ahimaaz the son of Zadok said, “Please let me run and bring the good news to the king that Adonai has vindicated him from the hand of his enemies.”}%
\verse{Joab said to him, “You will not be a man \textit{bringing}\lebnote{“for upon”}good news this day! You may bring good news on another day, but today you will not be bringing good news because the king’s son is dead.”}%
\verse{Then Joab said to the Cushite, “Go, tell the king what you have seen”; then the Cushite bowed down to Joab and ran off.}%
\verse{Then Ahimaaz the son of Zadok again said to Joab, “\textit{Come what may},\lnFVX{}please let me also run after the Cushite.” Joab asked, “Why are you wanting to run, my son, when for you there is no \textit{messenger’s reward}?”\lebnote{“and for you there is no good news finding”}}%
\verse{“\textit{Come what may},\lnFVX{}I want to run.” He said to him, “Run,” so Ahimaaz ran on the road on the plain, and he passed the Cushite.}%
\verse{Now David was sitting between the two gates, and the sentinel went up to the roof of the gate by the wall and he lifted up his eyes and watched, and look, a man was running by himself.}%
\verse{The sentry called and told the king, and the king said, “If he is alone, good news is in his mouth.” \textit{He kept coming closer}.\lebnote{“And he came, coming and near”}}%
\verse{Then the sentinel saw another man running, so the sentinel called to the gatekeeper and said, “Look, a man running alone.” The king said, “This one also is bringing good news.”}%
\verse{The sentinel said, “I am seeing that the running of the first is like the running of Ahimaaz the son of Zakok.” The king said, “He is a good man; he will come, for good news.”}%
\verse{Then Ahimaaz called and said to the king, “Peace.” He bowed down to the king with his face to the ground, and he said, “May Adonai your God be blessed, who has delivered the men who raised their hand against my lord the king.”}%
\verse{The king said, “Is it peace for the young man Absalom?” Ahimaaz said, “I saw the great commotion when Joab the servant of the king sent your servant, but I do not know what it was all about.”}%
\verse{Then the king said, “Turn aside, take your place here,” so he turned aside and waited.}%
\verse{Suddenly the Cushite arrived and said, “May my lord the king receive the good news, for Adonai has vindicated you today from the power of all who stood up against you.”}%
\verse{The king said to the Cushite, “Is it peace for the young man Absalom?” Then the Cushite answered, “May the enemies of my lord the king and all who rise up against you \textit{to harm you}\lebnote{“for evil”}be like the young man!”}%
\verse{\lebnote{2 Samuel 18:33–19:43 in the English Bible is 19:1–44 in the Hebrew Bible}The king was upset, and he went up to the upper room of the gate and wept. He said as he went, “My son, Absalom, my son, my son, Absalom. \textit{If only}\lebnote{“Who would grant my dying in place of you”}I had died instead of you, Absalom, my son, my son.”}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 19
\verseWithHeading{King David Weeps for Absalom}{And it was told to Joab that the king was weeping and he mourned over Absalom,}%
\verse{so the victory turned to mourning on that day for all the army, because they had heard that day, “The king is grieving over his son.”}%
\verse{The army \textit{came secretly}\lebnote{“stole secretly to come”}into the city on that day because the army was disgraced when they fled in the battle,}%
\verse{and because the king had covered his face and cried with a loud voice, “My son, Absalom, Absalom, my son, my son.”}%
\verse{Then Joab came to the king’s house and said, “Today you have humiliated the faces of all of your servants who have saved your life this day, and the life of your sons and your daughters, the life of your wives, and the life of your concubines,}%
\verse{by loving those who hate you and hating those who love you. Indeed, you have made clear this day that you have no regard for your commanders or officers, for I have realized today that were Absalom alive, and all of us were dead, then that would have been right in your eyes!}%
\verse{So then, get up and go out and \textit{speak kindly to your servants},\lebnote{“speak to the heart of your servants”}for I swear by Adonai, if you do not go out, no man will lodge this night with you, and this disaster will be greater for you than any disaster that has come upon you from your childhood until now!”}%
\verse{So the king got up and he sat in the gate, and they told all the army, “Look, the king is sitting in the gate.” Then all the army came before the king; whereas all of Israel had fled, each to his tent.}%
\verseWithHeading{David Returns from Exile}{Then it happened that all the people were disputing among all the tribes of Israel, saying, “The king delivered us from the hand of our enemies, and he saved us from the hand of the Philistines, but now he has fled from the land because of Absalom.}%
\verse{Now Absalom whom we anointed as king over us has died in the battle; so then, why are you taking no action to restore the king?”}%
\verse{Then King David sent to Zadok and to Abiathar the priests, saying, “Speak to the elders of Judah: ‘Why are you last to bring back the king to his house? The talk of all Israel has come to the king in his house.}%
\verse{My brothers, you are my bones and you are my flesh. Why should you be the last to bring back the king?’}%
\verse{To Amasa you shall say: ‘Are you not my bones and my flesh? \textit{May God punish me}\lebnote{“Thus may God do to me and thus may he add”}if you are not the commander of my army before me forever, in place of Joab.’ ”}%
\verse{So he turned the heart of all the men of Judah as one man, and they sent word to the king, “Return, you and all your servants.”}%
\verse{Then the king returned and he came to the Jordan; Judah had come to Gilgal to come to meet the king, to bring the king over the Jordan.}%
\verse{Then Shimei the son of Gera, the son of the Benjaminite, who was from Bahurim quickly came down with the men of Judah to meet King David,}%
\verse{and a thousand men were with him from Benjamin. Too, Ziba the servant of the household of Saul and fifteen of his sons and twenty of his servants were with him, and they rushed to the Jordan before the king.}%
\verse{\textit{The crossing took place}\lebnote{“And the ford crossed”}to bring the household of the king over and to do good in his eyes. Then Shimei the son of Gera fell before the king when he crossed over the Jordan,}%
\verse{and he said to the king, “May not my lord hold me guilty, and may you not remember how your servant did wrong on the day that my lord the king went out from Jerusalem, by taking it to heart!}%
\verse{For your servant knows that I have sinned; look, I have come this day as the first of all the house of Joseph to come down to meet my lord the king.”}%
\verse{Then Abishai the son of Zeruiah responded and said, “Because of this, should not Shimei be put to death, for he cursed the anointed one of Adonai?”}%
\verse{Then David said, “What is it to me or to you, sons of Zeruiah, that you should be an adversary today? Should anyone be put to death in Israel? Do I not know today that I am king over Israel?”}%
\verse{Then the king said to Shimei, “You shall not die,” and so the king swore to him.}%
\verse{Now, Mephibosheth the son of Saul came down to meet the king; he had not taken care of his feet nor trimmed his moustache nor washed his clothes from the day the king left until the day he came back in peace.}%
\verse{It happened that when he came to Jerusalem to meet the king, the king said to him, “Why did you not come with me, Mephibosheth?”}%
\verse{Then he said, “My lord the king, my servant deceived me, for your servant had said, ‘Let me saddle the donkey that I may ride on her and go with the king,’ for your servant is lame.}%
\verse{But he slandered against your servant to my lord the king. My lord the king is like the angel of God; \textit{do as you see fit.}.”\lebnote{“do the good in your eyes”}}%
\verse{For there was no one in all the house of my father \textit{who were not doomed to death}\lebnote{“except men of death”}before my lord the king, but you set your servant among those who eat at your table. Do I have any righteousness any longer except to cry out to the king?”}%
\verse{Then the king said to him, “Why should you speak any more about the matter? I have decided: you and Ziba shall divide the land.”}%
\verse{Then Mephibosheth said to the king, “Let him take the whole thing, since my lord the king has come \textit{safely}\lebnote{“in peace”}to his house.”}%
\verse{Then Barzillai the Gileadite came down from Rogelim and crossed with the king over the Jordan to escort him through the Jordan.}%
\verse{Now Barzillai was very old, \textit{eighty years old}.\lebnote{“a son of eighty years”}Now he had provided the king with food while he was staying at Mahanaim, for he was a very wealthy man.}%
\verse{The king said to Barzillai, “You cross over with me, and I will provide for you to dwell with me in Jerusalem.”}%
\verse{Then Barzillai said to the king, “What are the days of the years of my life, that I should go with the king to Jerusalem?}%
\verse{I am eighty years old today. Can I discern between good and bad? Or can your servant taste what I eat or what I drink? Or can I still hear the voice of singing men and women? Why should your servant be a burden any longer to my lord the king?}%
\verse{Your servant shall go over the Jordan with the king a little way, but why should the king recompense me with this reward?}%
\verse{Please let your servant return, and let me die in my city in the tomb of my father and my mother. Here is your servant Kimham; let him cross over with my lord the king, and do for him that which is good in your eyes.”}%
\verse{The king said, “Let Kimham go over with me, and I will do for him the good in your eyes, and all that you desire of me I will do for you.”}%
\verse{Then all the people crossed over the Jordan, and the king crossed and kissed Barzillai and blessed him; then he returned to his place.}%
\verse{The king went over to Gilgal, and Kimham went over with him. All the people of Judah went over with the king, and half of the people of Israel too.}%
\verse{Suddenly, all the men of Israel were coming to the king. They said to the king, “Why have our brothers the men of Judah stolen away and brought the king and his household over the Jordan, with all the men of David?”}%
\verse{Then all the men of Judah answered the men of Israel, “Because the king is my close relative! Why are you this angry over this matter? \textit{Have we by any means eaten anything from the king? Did we take by any means anything that was not ours}?”\lebnote{“By any means did we eat from the king or by any means was it dragged away for us?”}}%
\verse{Then the people of Israel answered the men of Judah and said, “I have \textit{ten times as much}\lebnote{“ten hands”}in the king, moreover in David I have more than you. Why did you treat me with contempt \textit{by not giving me first chance}\lebnote{“and my word was not the first”}to bring back my king?” But the words of the men of Judah were fiercer than the word of the men of Israel.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 20
\verseWithHeading{Sheba Leads a Revolt}{Now a man of wickedness was found there whose name was Sheba the son of Bichri, a Benjaminite. He blew the horn and said, “There is no share for us in David, and there is no inheritance for us in the son of Jesse; each to his tents, O Israel!”}%
\verse{Then all the men of Israel went up from following after David, following instead after Sheba the son of Bicri, but the men of Judah stuck to their king from the Jordan up to Jerusalem.}%
\verse{David went up to his house in Jerusalem, then the king took the ten concubines whom he had left to look after the house, and he put them \textit{under confinement}.\lebnote{“a house of guard”}However, he provided for them, but \textit{he did not sleep with them}.\lebnote{“he did not come into them”}So they were confined until the day of their death, like a lifetime of widowhood.}%
\verseWithHeading{Joab Assassinates Amasa}{Then the king said to Amasa, “Summon for me the men of Judah within three days, \textit{and be here yourself}.”\lebnote{“and you stand here”}}%
\verse{So Amasa went to summon Judah, but he tarried more than the appointed time which he had set for him.}%
\verse{Then David said to Abishai, “Now Sheba the son of Bicri will do us more harm than Absalom. You take the servants of your lord and pursue after him, lest he find fortified cities for himself and escape from us.”}%
\verse{Then the men of Joab, the Kerethites and the Pelethites, and all the mighty warriors went out after him; they went out from Jerusalem to pursue after Sheba the son of Bicri.}%
\verse{They were near the big rock that is in Gibeon, and Amasa came before them. Joab \textit{was dressed in his military clothing},\lebnote{“was dressed his garment his clothing”}with a utility belt on him and a sword strapped to his waist in its scabbard. Now he went out, and it fell out.}%
\verse{Then Joab said to Amasa, “Is it peace, O you my brother?” Then the right hand of Joab took hold of the beard of Amasa as if to kiss him.}%
\verse{Now Amasa was not on his guard against the sword that was in Joab’s hand, and he struck him with it into the stomach, and his entrails poured out to the ground. He did not strike him again, and he died. Then Joab and Abishai his brother pursued after Sheba the son of Bicri.}%
\verse{A young man stood over him, from the young men of Joab, and he said, “Whoever takes delight in Joab and whoever is for David, follow after Joab.”}%
\verse{Now Amasa was wallowing in the blood in the middle of the highway; when the man saw that all the people stood there, he turned Amasa over from the highway into the field, and he threw a garment over him because he had seen that all who had come by him had stopped.}%
\verse{After he was removed from the highway, all the men passed by after Joab to pursue after Sheba the son of Bicri.}%
\verseWithHeading{Wisdom from a Woman under Siege}{(He\lebnote{i.e., “Sheba”}had passed through all the tribes of Israel to Abel and Beth Maacah; now all of the Berites had been treated badly, so they also followed after him.)}%
\verse{And they came and besieged him in Abel Beth Maacah. They threw up a siege ramp against the city, and they stood against the ramparts. And all the army who were with Joab were battering to cause the wall to fall.}%
\verse{Then a wise woman from the city called out, “Listen, listen! Please speak to Joab to come near here so that I may speak to you.”}%
\verse{Then he came near to her, and the woman asked, “Are you Joab?” And he said, “I am.” Then she said to him, “Listen to the words of your servant.” He said, “I am listening.”}%
\verse{Then she said, “In former times, \textit{they would always say},\lebnote{“speaking they would speak”}‘By all means, let them inquire in Abel,’ and so they settled things.}%
\verse{I am one of the faithful representatives of Israel. You are seeking to destroy a city and a mother in Israel. Why do you want to swallow the inheritance of Adonai?”}%
\verse{Then Joab answered and said, “Far be it, far be it from me that I should swallow or I should destroy.}%
\verse{That is not the matter. But a man from the mountains of Ephraim, whose name is Sheba the son Bicri, has lifted up his hand against the king, against David. Give only him to us, and I will depart from the city.” The woman said to Joab, “Look, his head is being thrown down to you over the wall.”}%
\verse{The woman went to all of the people with her wise plan, so they cut off the head of Sheba the son of Bicri and threw it to Joab. Then he blew the horn and dispersed from the city, each to his tent. Then Joab returned to Jerusalem to the king.}%
\verse{Now Joab was over all the army of Israel and Benaiah the son of Jehoiada was over the Carites and over the Pelethites.}%
\verse{Adoram was over the forced labor, and Jehoshaphat the son of Ahilud was the recorder.}%
\verse{Shiya was secretary, and Zadok and Abiathar were priests.}%
\verse{Also Ira the Jairite was priest for David.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 21
\verseWithHeading{The Famine Brings Justice}{There was a famine in the days of David for three years, year after year. And David \textit{inquired of Adonai},\lebnote{“sought the face of Adonai”}and Adonai said, “The bloodguilt is on Saul and on his household, because he killed the Gibeonites.”}%
\verse{So the king called the Gibeonites and spoke to them. Now the Gibeonites were not from the \textit{Israelites};\lnAKM{}they were from the remainder of the Amorites. Now the \textit{Israelites}\lnAKM{}had sworn to them,\lebnote{The treaty between Israel and the Gibeonites is found in Josh 9}but Saul tried to wipe them out in his zeal for the \textit{Israelites}\lnAKM{}and Judah.}%
\verse{So David asked the Gibeonites, “What can I do for you, and with what can I make amends that you may bless the inheritance of Adonai?”}%
\verse{Then the Gibeonites said to him, “\textit{It is not a matter for us of}\lebnote{“There is not for me (\textit{Qere} ‘us’)”}silver or gold with Saul or with his household. It is not for us to put to death anyone in Israel.” He asked, “What are you saying that I should do for you all?”}%
\verse{Then they said to the king, “The man who consumed us and who plotted against us so that we were destroyed from existing in all of the territory of Israel,}%
\verse{let seven men from his sons be given over to us, and we will execute them before Adonai in Gibeah of Saul, the chosen one of Adonai.” Then the king said, “I will give them over.”}%
\verse{But the king spared Mephibosheth the son of Jonathan, the son of Saul, because of the sworn oath of Adonai which was between them, between David and Jonathan the son of Saul.}%
\verse{So the king took two of the sons of Rizpah the daughter of Aiah, whom she bore to Saul, namely Armoni and Mephibosheth, and five of the sons of Michal the daughter of Saul whom she had borne to Adriel the son of Barzillai the Meholathite.}%
\verse{He gave them into the hands of the Gibeonites, and they executed them on the mountain in the presence of Adonai, and the seven fell together. Now they were put to death in the days of the harvest, at the beginning of the harvest of barley.}%
\verse{Rizpah the daughter of Aiah took the sackcloth, and she spread it for herself on the rock at the beginning of the harvest until water gushed forth on them from heaven, but she did not allow the birds of heaven to rest on them by day nor the animals of the field by night.}%
\verse{David was told about what Rizpah the daughter of Aiah, the concubine of Saul, had done.}%
\verse{So David left and took the bones of Saul and the bones of Jonathan his son from the rulers of Jabesh Gilead, who had stolen them from the public square of Beth Shan, where the Philistines hung them \textit{when}\lebnote{“on the day”}the Philistines killed Saul on Gilboa.}%
\verse{He brought up the bones of Saul and the bones of Jonathan his son from there, and they gathered the bones of the executed.}%
\verse{And they buried the bones of Saul and Jonathan his son in the land of Benjamin at Zela, in the tomb of Kish his father. They did all that the king had commanded, and afterward God was entreated for the land.}%
\verseWithHeading{Battles with the Philistines Recounted}{There was war again for the Philistines with Israel, and David and his servants with him went down, and they fought the Philistines, and David grew weary.}%
\verse{Now Yishbi in Nob, who was among the descendents of Raphah\lebnote{The descendants of Raphah (i.e., the Rephaim) were thought to be giants}(now the weight of his spearhead was three hundredweight of bronze, and he was newly armed), said that he would kill David.}%
\verse{But Abishai the son of Zeruiah helped him, and he attacked the Philistine and killed him. Then the men of David swore to him, saying, “You shall not go out with us any longer to the battle, so that you do not quench the lamp of Israel.”}%
\verse{It happened afterward that there was again battle at Gob with the Philistines. Then Sibbecai the Hushathite killed Saph who was among the descendants of the Raphah.}%
\verse{There was again a battle with the Philistines at Gob. And Elhanan the son of Jaare-Oregim, the Bethlehemite, killed Goliath\lebnote{In view of the account of David and Goliath in 1 Sam 17, it is likely that Elhanan actually killed the brother of Goliath, Lahmi, in 1 Chr 20:5}the Gittite, the shaft of whose spear was like the beam of a weaver.}%
\verse{Once again there was battle at Gath, and there was \textit{a man of great size}.\lebnote{“a man of measurement”}The fingers of his hand and the toes of his feet were six and six, twenty-four in number. He was also born to the Raphah.}%
\verse{He taunted Israel but Jonathan the son of Shimei, the brother of David, killed him.}%
\verse{These four were born for the Raphah in Gath, and they fell by the hand of David and by the hand of his servants.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 22
\verseWithHeading{The Victory Song of David}{Then David spoke to Adonai the words of this song,\lebnote{This same song is recorded in Psa 18, with minor differences}on the day Adonai delivered him from the hand of all his enemies and from the hand of Saul.}%
\verse{And he said: 
“Adonai is my rock, my fortress, and my deliverer.}%
\verse{I take refuge in God, my rock,\lebnote{So many LXX mss, Targum, Vulgate, and Psa 18:3. MT has “the God of my rock”}my shield, and the \textit{strength}\lebnote{“horn”}of my salvation. 
My stronghold and my refuge, O my savior, you will save me from violence!}%
\verse{I call upon Adonai who is praiseworthy, 
and I am saved from those who hate me.}%
\verse{For the breaker waves of death engulfed me; 
the currents of chaos overwhelmed me.}%
\verse{The ropes of Sheol entangled me; 
the snares of death confronted me.}%
\verse{In my distress I called upon Adonai, and to my God I called. 
He heard my voice from his temple, 
and my cry for help was to his ears.}%
\verse{The earth heaved and shook, 
the foundations of heaven trembled and heaved 
because \textit{he was angry}.\lebnote{“it was hot for him”}}%
\verse{Smoke went up from his nostrils and fire from his mouth. 
Burning coals devoured, they burned from him.}%
\verse{He bowed the heavens and came down; 
a very thick cloud was under his feet.}%
\verse{He rode upon a cherub and flew; 
he was seen on the wings of the wind.}%
\verse{He put darkness as a canopy all around him, 
a collection of \textit{thick rain clouds}.\lebnote{“clouds of thick clouds”}}%
\verse{From the brightness before him 
flamed burning coals of fire.}%
\verse{Adonai thundered from heaven, 
and the Most High \textit{uttered his voice}.\lebnote{“gave his voice”}}%
\verse{He sent arrows and he scattered them, 
lightning, and he directed them.}%
\verse{Then the channels of water of the sea were exposed, 
the foundations of the world, 
at the rebuke of Yawheh, 
from the blast of the breath of his nostrils.}%
\verse{He sent from a high position and took me; 
he drew me from mighty waters.}%
\verse{He delivered me from my strong enemies, 
from those who hate me, for they were mightier than I.}%
\verse{They approached me on the day of my disaster, 
but Adonai was my support.}%
\verse{He brought me out to a spacious place. 
He delivered me because he delighted in me.}%
\verse{Adonai rewarded me according to my righteousness, 
according to the cleanness of my hands he recompensed me.}%
\verse{For I have kept the ways of Adonai; 
I have not acted wickedly against my God.}%
\verse{For all of his ordinances were before me, 
and I did not turn aside from his statutes.}%
\verse{I was blameless before him, 
and I kept myself from my iniquity.}%
\verse{Adonai has recompensed me according to my righteousness, 
according to my cleanness before his eyes.}%
\verse{With the loyal, you act as loyal, 
and with the blameless, you show yourself blameless.}%
\verse{With the pure, you show yourself pure, 
but with the crooked, you appear as a fool.}%
\verse{Humble people you will deliver, 
but your eyes are on the haughty, whom you bring down.}%
\verse{For you, O Adonai are my lamp, 
and Adonai lightens my darkness.}%
\verse{For by you I can run a raid; 
by my God I can leap over a wall.}%
\verse{This God, his way is blameless; 
the promise of Adonai is flawless. 
He is a shield to all who take refuge in him.}%
\verse{For who is God apart from Adonai? 
And who is a rock apart from our God?}%
\verse{God is my strong refuge, 
he has fully opened my way.\lebnote{Thus \textit{Qere}; \textit{Kethib} has “his way”}}%
\verse{He makes my feet\lebnote{Thus \textit{Qere}; \textit{Kethib} has “his feet”}like a doe deer, 
and on my high places he has set me.}%
\verse{He trains my hands for the war, 
so that my arms can bend a bow of bronze.}%
\verse{You have given me the shield of your salvation; 
your willingness to help has made me great.}%
\verse{You have broadened my steps beneath me; 
my ankles have not wobbled.}%
\verse{I pursued those who hate me, and I destroyed them. 
I did not turn back until finishing them.}%
\verse{I wiped them out and I smashed them; 
they did not get up; they fell under my feet.}%
\verse{You have girded me with physical strength for the battle; 
you caused those who rose up against me to kneel under me.}%
\verse{\textit{My enemies you cause to retreat from me};\lebnote{“Those who hate me you give to me back”}
I destroy those who hate me.}%
\verse{They looked out, but there was no deliverer, 
even to Adonai, but he did not answer them.}%
\verse{I beat them fine, like the dust of the earth; 
like the mire of the streets, I crushed them, I stamped them down.}%
\verse{You delivered me from the strife of my people; 
you preserved me as the head of the nations. 
A people I had not known served me.}%
\verse{Children of a foreign land came cringing to me; 
\textit{when they heard of me},\lebnote{“at the hearing of an ear”}they became obedient to me.}%
\verse{Children of a foreign land lost heart 
and came trembling from their strongholds.}%
\verse{Adonai lives! Blessed be my rock! 
May God, the rock of my salvation, be exalted!}%
\verse{God does vengeance for me, 
bringing down peoples under me.}%
\verse{He brings me out from my enemies, 
and from those who rose up against me you lift me up, 
and from men of violence you rescue me.}%
\verse{Therefore I will extol you, Adonai, among the nations! 
I will sing praises to your name!}%
\verse{He makes great salvation for his king 
and shows loyal love to his anointed one, David 
and to his descendants forever.”}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 23
\verseWithHeading{David Extols Adonai}{These are the last words of David, the declaration of David the son of Jesse, and the declaration of the man exalted by the Most High, the anointed one of the God of Jacob and the darling of the songs of Israel.}%
\verse{“The spirit of Adonai speaks through me, and his word is upon my tongue.}%
\verse{The God of Israel said to me, the rock of Israel has spoken; ‘He who rules over mankind rules righteously, in the fear of God.}%
\verse{Like the light of the morning when the sun rises, shining with no clouds, bringing vegetation from the earth apart from rain.’\lebnote{In other words, even men who rule rightly pass like the dew of the morning}}%
\verse{Yet not so is my house with God, for he made an everlasting covenant for me, arranging everything. He has secured all my deliverance, and all my desire he will cause to happen.}%
\verse{But evil persons are like thorns cast aside; all of them, because they cannot be picked up in the hand.}%
\verse{And if a man wants to touch them, he must use an iron instrument or the shaft of a spear; then they are consumed entirely with fire on the spot.”}%
\verseWithHeading{David’s Faithful Soldiers}{These are the names of the mighty warriors who were David’s: Josheb-Basshebeth a Tahkemonite was chief of three officers; first Adino, whose spear was against eight hundred slain on one occurrence.}%
\verse{Next to him Eleazar, the son of Dodo the son of an Ahohite, was among the three mighty warriors with David when they defied the Philistines and they gathered there for the battle and the men of Israel withdrew.}%
\verse{He stood up and struck down the Philistines until his hand grew tired and his hand clung to the sword, and Adonai brought about a great victory on that day. Then the army returned back to him only for stripping the dead.}%
\verse{Next to him was Shamma, the son of Agee the Hararite. When the Philistines assembled at Lehi, a plot of the field was there filled with lentils, and the army fled there from the presence of the Philistines.}%
\verse{But he took a stand in the middle of the plot of land and defended it. He killed the Philistines, and Adonai brought about a great victory.}%
\verse{Then three\lebnote{So \textit{Qere}; \textit{Kethib} reads “thirty”}of the thirty leaders went down and came to David \textit{at the time of the harvest}\lebnote{“to the harvest”}to the cave of Adullam, while a group of the Philistines were camping in the valley of the Rephaim.}%
\verse{Now at that time, David was in the stronghold, and a garrison of the Philistines was in Bethlehem at that same time.}%
\verse{David \textit{said longingly},\lebnote{“desired and said”}“\textit{Oh that someone would bring me a drink}\lebnote{“Who will let me drink …?”}of water from the well of Bethlehem that is at the gate.”}%
\verse{So three of the mighty warriors broke into the camp of the Philistines, and they drew water from the well of Bethlehem that was at the gate, and they carried it and brought it to David. But he was not willing to drink it, but poured it out to Adonai.}%
\verse{He said, “Far be it from me before Adonai that I should do this. Is this not the blood of the men who went at the risk of their lives?” So he was not willing to drink it. These things the three mighty warriors did.}%
\verse{Now Abishai the brother of Joab the son of Zeruiah was himself the leader of the thirty. He was wielding his spear against three hundred slain and \textit{gained a name}\lebnote{“and to him a name”}among the thirty.}%
\verse{Among the thirty, is it not that he was honored and became a commander for them? But he did not come up to the three.}%
\verse{Benaiah the son of Jehoiada, the son of Ish-Hai, was a great man of deeds from Kabzeel. He struck down two sons of Ariel of Moab, and he went down and killed a lion in the middle of a pit on a snowy day.}%
\verse{He also killed a good-looking Egyptian man, in whose hand was a spear. He went down against him with the staff and snatched the spear from the hand of the Egyptian and killed him with his spear.}%
\verse{These things Benaiah the son of Jehoiada did and gained a name for himself among the three mighty warriors.}%
\verse{He was honored more than the thirty, but he did not come up to the three. David appointed him in charge of his bodyguard.}%
\verseWithHeading{The Mighty Men of David}{Among the thirty were Asahel the brother of Joab, Elhanan the son of Dodo of Bethlehem,}%
\verse{Shammah the Harodite, Elika the Harodite,}%
\verse{Helez the Paltite, Ira the son of Ikkesh the Tekoite,}%
\verse{Abiezer the Anathothite, Mebunnai the Hushathite,}%
\verse{Zalmon the Ahohite, Maharai the Netophathite,}%
\verse{Heleb the son of Baanah the Netophathite, Ittai the son of Ribai from Gibeah of the sons of Benjamin,}%
\verse{Benaiah the Pirathonite, Hiddai from the wadis of Gaash,}%
\verse{Abi-Albon the Arbathite, Azmaveth the Barhumite,}%
\verse{Eliahba the Shaalbonite, the sons of Jashen, Jonathan}%
\verse{the son of Shammah the Hararite, Ahiam the son of Sharar the Ararite,}%
\verse{Eliphelet the son of Ahasbai the son of the Maacathite, Eliam the son of Ahithophel the Gilonite,}%
\verse{Hezro the Carmelite, Paarai the Arbite,}%
\verse{Igal the son of Nathan from Zobah, Bani the Gadite,}%
\verse{Zelek the Ammonite, Naharai the Beerothite, the carriers of the weapons of Joab the son of Zeruiah,}%
\verse{Ira the Ithrite, Gareb the Ithrite,}%
\verse{Uriah the Hittite; in all, thirty-seven.}%
\end{biblechapter}

\begin{biblechapter} % 2 Samuel 24
\verseWithHeading{David and the Census of the People}{Again Adonai was angry with Israel, and he\lebnote{The parallel passage in 1 Chr 21 names the “he” as Satan}incited David against them, saying, “Go count Israel and Judah.”}%
\verse{The king said to Joab, the commander of the army who was with him: “Please go about through all the tribes of Israel from Dan to Beersheba, and count the people that I may know the number of the people.”}%
\verse{Then Joab said to the king, “May Adonai your God increase the people a hundred times \textit{what they are}\lebnote{“as them and as them”}as the eyes of my lord the king are seeing. But my lord the king, why does he desire this thing?”}%
\verse{But the word of the king prevailed over Joab and over the commanders of the army, so Joab and the commanders of the army went out from before the king to count the people of Israel.}%
\verse{They crossed over the Jordan and camped at Aroer to the south of the city, which was in the middle of the wadi of Gad, and up to Jazer.}%
\verse{Then they went to Gilead and to the land of Tahtim Hodshi. They came to Dan Jaan and around to Sidon}%
\verse{and came to the fortress of Tyre and all the cities of the Hivites and the Canaanites. Then they went out to the Negev of Judah at Beersheba.}%
\verse{They went about through all the land, and they came to Jerusalem at the end of nine months and twenty days.}%
\verse{Then Joab gave the number of the counting of the people to the king. Israel was eight hundred thousand \textit{valiant warriors}\lebnote{“men of ability”}wielding the sword, and the men of Judah were five hundred thousand.}%
\verse{The heart of David struck him after he had counted the people, and David said to Adonai, “I have sinned greatly in what I have done! So then, O Adonai, please forgive the guilt of your servant because I have acted very foolishly.”}%
\verse{When David got up in the morning, the word of Adonai came to Gad the prophet, the seer of David, saying,}%
\verse{“Go and speak to David, ‘Thus says Adonai, three things I am laying on you; choose for yourself one of them and I will do it to you.’ ”}%
\verse{Then Gad came to David, and he told him and said to him, “Shall seven years of famine in the land come to you? Or three months of your fleeing from your enemies while he is pursuing you? Or should there be three days of pestilence in your land? Now consider and decide what I must return to the one who sent me a word.”}%
\verse{Then David said to Gad, “I am greatly distressed. Please let us fall into the hand of Adonai, because he is great in his compassion; but into the hand of man don’t let me fall.”}%
\verse{Then Adonai sent a plague into Israel from the morning \textit{until the agreed time},\lebnote{“until the time of agreed time”}and from the people from Dan to Beersheba, seventy thousand men died.}%
\verse{When the angel stretched out his hand to destroy Jerusalem, Adonai regretted about the evil, and he said to the angel who brought destruction among the people, “Enough, now relax your hand.” Now the angel of Adonai was at the threshing floor of Araunah the Jebusite.}%
\verse{David spoke to Adonai when he saw the angel destroying among the people, and he said, “Look, I have sinned and I have done wrong, but these sheep, what did they do? Please let your hand be against me and against the house of my father.”}%
\verse{Then Gad came to David on that same day and said to him, “Go up and erect an altar to Adonai at the threshing floor of Araunah the Jebusite.”}%
\verse{So David went up according to the word of Gad, as Adonai had commanded.}%
\verse{Araunah looked down and saw the king and his servants coming over to him, so Araunah went out and bowed down before the king with his face to the ground.}%
\verse{Then Araunah said, “Why has my lord the king come to his servant?” David said, “To buy from you the threshing floor, to build an altar to Adonai who brought a halt to the plague on the people.”}%
\verse{Araunah said to David, “Let my lord the king take and offer what is good in his eyes. Look, here are the cattle for the burnt offering and the threshing sledge and the yokes of the oxen for the firewood.}%
\verse{All of this Araunah hereby gives to the king.” Then Araunah said to the king, “May Adonai your God respond favorably for you.”}%
\verse{Then the king said to Araunah, “No, but \textit{I will certainly buy}\lebnote{“buying I will buy”}it from you for a price; I don’t want to offer to Adonai my God burnt offerings that cost me nothing.” So David bought the threshing floor and the cattle for fifty shekels of silver.}%
\verse{David built an altar to Adonai there, and he offered burnt offerings and fellowship offerings. Then Adonai responded to his prayer for the land and brought the plague to a halt from upon Israel.}%
\end{biblechapter}

