\biblebook{Daniel}

\begin{biblechapter} % Daniel 1
\verseWithHeading{Daniel’s Development in the King’s Court}{In the third year of the reign of Jehoiakim king of Judah, Nebuchadnezzar the king of Babylon came to Jerusalem and besieged it.}%
\verse{And the Lord gave Jehoiakim the king of Judah into his hand and \textit{some of}\lebnote{Literally “from the end/extremity/limits”}of the utensils of the temple of God,\lebnote{Hebrew “the God”}and he brought them to the land of Shinar to the temple of his gods, and he brought the utensils to \textit{the treasury}\lebnote{Literally “the house of the treasury”}of his gods.}%
\verse{And the king \textit{ordered}\lebnote{Literally “said to”}Ashpenaz, the commander of his court officials, to bring \textit{some of the Israelites}\lebnote{Literally “from the sons/children of Israel”}from \textit{the royal family}\lebnote{Literally “from the seed of the kingship”}and from the lords,\lebnote{Or “nobles”}}%
\verse{youths who \textit{have no physical defect},\lebnote{Literally “there is not in them any physical defect”}and who are \textit{handsome},\lebnote{Literally “good/pleasing of appearance”}and who are prudent\lebnote{Or “insightful”}in all wisdom and \textit{endowed with knowledge},\lebnote{Literally “who have knowledge of knowledge”}and who understand insight, and who have the \textit{ability}\lebnote{Literally “strength”}in them to serve in the palace of the king. And the king ordered him to teach them the literature and the language of the Chaldeans.}%
\verse{And the king assigned to them \textit{his daily portion}\lebnote{Literally “the portion of the day on its day”}from the fine food of the king, and from \textit{the wine that he drank},\lebnote{Literally “the wine of his drink”}and instructed that \textit{they were to be educated for three years}.\lebnote{Literally “and to be educated them years three”}And at the end of their training, \textit{they were to be stationed}\lebnote{Literally “they would stand”}\textit{before}\lebnote{Literally “to the face of”}the king.}%
\verse{Now there was among them \textit{from the Judeans},\lebnote{Literally “from the children of Judah”}Daniel, Hananiah, Mishael and Azariah.}%
\verse{And the commander of the court officials gave them names, and he called Daniel, Belteshazzar; and Hananiah, Shadrach; and Mishael, Meshach; and Azariah, Abednego.}%
\verseWithHeading{Daniel’s Resolve}{Now \textit{Daniel resolved}\lebnote{Literally “Daniel placed upon his heart”}that he would not defile himself with the fine food of the king, and with \textit{the wine that he drank},\lebnote{Literally “the wine of his drink”}and so he requested from the commander of the court officials permission so that he would not defile himself.}%
\verse{And God\lebnote{Hebrew “the God”}gave Daniel favor and compassion \textit{before}\lebnote{Literally “to the face of”}the commander of the court officials,}%
\verse{and the commander of the court officials said to Daniel, “I am afraid of my lord, the king, who has determined your food and your drink, for \textit{why should}\lebnote{Literally “to what would”}he see your face having a worse appearance than the young men who \textit{are your age}?\lebnote{Literally “like your age”; “your” is plural}Then you will endanger my head with the king.”}%
\verse{Then\lebnote{Hebrew “And”}Daniel asked the guard\lebnote{Or “overseer”}whom the commander of the court officials had appointed over Daniel, Hananiah, Mishael, and Azariah,}%
\verse{“Please test your servants for ten days, and let them give us \textit{some of the vegetables},\lebnote{Literally “from the vegetables”}and let us eat and let us drink water.}%
\verse{Then\lebnote{Hebrew “And”}let our appearances and the appearance of the young men who are eating the fine food of the king be compared \textit{before you},\lebnote{Literally “in your presence”}and then deal with your servants \textit{according to what you see}.”\lebnote{Literally “and according to/that you see”}}%
\verse{So he agreed to this proposal with them, and he tested them for ten days.}%
\verse{And at the end of ten days their appearances appeared better and they were healthier of body than all the young men who were eating the fine food of the king.}%
\verse{\textit{So}\lebnote{Literally “And it happened”}the guard \textit{continued to withhold}\lebnote{Literally “he was … withdrawing”}their fine food and the wine of their drink, and he gave them\lebnote{Hebrew “to them”}vegetables.}%
\verse{\textit{And as for these four young men},\lebnote{Literally “And these young men, four of them”}God\lebnote{Hebrew “the God”}gave to them knowledge and insight into all literature and wisdom, and Daniel had insight into all visions\lebnote{Hebrew “vision”}and dreams.}%
\verse{And at the end of the time the king had set to bring them, the commander of the court officials brought them in \textit{before}\lebnote{Literally “to the face of”}Nebuchadnezzar.}%
\verse{And the king spoke with them, \textit{and among all of them no one was found like}\lebnote{Literally “and he was not found from all of them like”}Daniel, Hananiah, Mishael, and Azariah; then they stood \textit{before}\lebnote{Literally “to the face of”}the king.}%
\verse{And in every matter of wisdom and understanding about which the king inquired from them, he found them ten times better than all of the magicians\lebnote{Or “soothsayer-priests”}and conjurers\lebnote{Or “enchanters”}that \textit{were in his entire kingdom}.\lebnote{Literally “in the whole of his kingdom”}}%
\verse{And Daniel was there until the first year of Cyrus the king.}%
\end{biblechapter}

\begin{biblechapter} % Daniel 2
\verseWithHeading{The King’s Troubling Dream}{Now\lebnote{Hebrew “And”}in the second year of the reign of Nebuchadnezzar, Nebuchadnezzar dreamed dreams; and his spirit was troubled and his sleep left him.}%
\verse{So the king said to call the magicians\lebnote{Or “soothsayer-priests”}and the conjurers\lebnote{Or “enchanters”}and the sorcerers and the \textit{astrologers}\lebnote{Literally “Chaldeans”}to tell to the king his dreams. And they came in and they stood \textit{before}\lebnote{Literally “to the face of”}the king.}%
\verse{And the king said to them, “\textit{I have had a dream}\lebnote{Literally “A dream I have dreamed”}and my spirit is anxious to know the dream.”}%
\verse{And the \textit{astrologers}\lebnote{Literally “Chaldeans”}said to the king in Aramaic,\lebnote{The text of Daniel is in Aramaic from this point on through Daniel chapter 7. Then the text resumes in Hebrew to the end of the book}“O king, live forever! Tell the dream to your servants and we will reveal the explanation.”}%
\verse{The king answered and said to the \textit{astrologers},\lebnote{Literally “Chaldeans”}“The command from me is firm: if you do not make known to me the dream and its explanation,\lebnote{Or “interpretation”}then you will be broken into pieces and your houses will be laid in ruins.}%
\verse{But if you tell me the dream and its explanation, you will receive gifts and rewards\lebnote{Aramaic “reward”}and great honor from me. Therefore, tell me the dream and its explanation.”\lebnote{Or “interpretation”}}%
\verse{They answered once more and said, “Let the king tell the dream to his servants and we will make the explanation known.”}%
\verse{The king answered and said, “\textit{Certainly}\lebnote{Literally “From a certainty”}I know that \textit{you are trying to gain time}\lebnote{Literally “the time you are gaining”}\textit{because}\lebnote{Literally “all because of that”}you have seen \textit{that this matter is firmly decreed by me},\lebnote{Literally “that firm from me is the decree”}}%
\verse{for if you do not make the dream known to me, \textit{your verdict is fixed},\lebnote{Literally “one is your verdict”}and you have conspired to say a lying and deceitful word to me until the \textit{circumstances}\lebnote{Literally “time”}will change. Therefore, tell me the dream, and I will know that you can tell me its explanation.”}%
\verse{The \textit{astrologers}\lebnote{Literally “Chaldeans”}answered the king and said, “There is not a man on earth that is able to reveal the word of the king; \textit{in fact},\lebnote{Literally “all because that”}no great and powerful king has ever asked a thing like this of any magician\lebnote{Or “soothsayer-priest”}or conjurer\lebnote{Or “enchanter”}or \textit{astrologer}.\lebnote{Literally “Chaldean”}}%
\verse{And the thing that the king is asking is too difficult and \textit{there is no one who}\lebnote{Literally “there is not another”}can reveal it to the king except the gods \textit{whose dwelling is not with mortals}.”\lebnote{Literally “concerning their dwelling is not with flesh”}}%
\verse{Because of this the king became angry, and he became very much enraged, and he said that all the wise men of Babylon are to be destroyed.}%
\verse{And the decree was issued, and the wise men \textit{were on the verge of being executed},\lebnote{Literally “were in the process of being killed”}and they searched for Daniel and his companions to be executed.}%
\verse{Then Daniel responded prudently and discretely to Arioch, the commander of the imperial guard of the king, who had gone out to execute the wise men of Babylon.}%
\verse{\textit{He asked}\lebnote{Literally “he answered”}and said to Arioch, the royal official of the king, “\textit{Why}\lebnote{Literally “Over what”}is the decree from the king so severe?” Then Arioch explained the matter\lebnote{Or “situation”}to Daniel.}%
\verse{And Daniel went in and requested from the king that he would give him time, and \textit{he would tell the king}\lebnote{Literally “to tell/reveal to the king”}the explanation.\lebnote{Or “interpretation”}}%
\verse{Then Daniel went to his home, and he made the matter known to Hananiah, Mishael and Azariah, his companions,}%
\verse{and told them to seek mercy \textit{from the God of heaven}\lebnote{Literally “from before the God of the heaven”}concerning this mystery, so that Daniel and his companions, along with the remainder of the wise men of Babylon, would not be killed.}%
\verse{Then in a vision of the night the mystery was revealed to Daniel; then Daniel blessed the God of heaven.}%
\verse{Daniel \textit{said}:\lebnote{Literally “answered and he said”}}%
\verse{“Let his name, the name of God,\lebnote{Aramaic “the God”}be blessed \textit{throughout the ages},\lebnote{Literally “from the age unto the age”}
for the wisdom and the power \textit{are his}.\lebnote{Literally “of to him it is”}}%
\verse{And he changes the times and the seasons, 
and he deposes kings and he sets up kings; 
he gives wisdom\lebnote{Aramaic “the wisdom”}to wise men 
and knowledge to men who know understanding.}%
\verse{He reveals the deep and the hidden things; 
he knows what is in the darkness, 
and the light dwells with him.}%
\verseWithHeading{Daniel Praises God and Interprets the King’s Dream}{\textit{Therefore}\lebnote{Literally “Now because this”}Daniel went to Arioch, whom the king had appointed\lebnote{Aramaic “has appointed”}to destroy the wise men of Babylon; he went and thus he said to him: “You must not destroy the wise men of Babylon; take me in before the king and I will give the explanation\lebnote{Or “interpretation”}to the king.”}%
\verse{Then Arioch \textit{quickly}\lebnote{Literally “in haste”}brought Daniel in before the king and thus he said to him: “I have found a man among \textit{the exiles}\lebnote{Literally “the children of exiles”}of Judah\lebnote{Aramaic “Jehud”}who \textit{can relate}\lebnote{Literally “he will make known”}the explanation\lebnote{Or “interpretation”}to the king.}%
\verse{\textit{The king then asked}\lebnote{Literally “The king answered”}and said to Daniel, whose name was Belteshazzar, “Are you able to make known to me the dream that I have seen and its explanation?”\lebnote{Or “interpretation”}}%
\verse{Daniel answered the king and said, “The mystery that the king asks, no wise men, conjurers,\lebnote{Or “enchanters”}magicians,\lebnote{Or “soothsayer-priests”}or diviners are able to make known to the king.}%
\verse{But there is a God in heaven who reveals mysteries, and he has made known to King Nebuchadnezzar what it is that will be at the end of days. This is your dream and the visions of your head on your bed.}%
\verse{“As for you, king,\lebnote{Aramaic “the king”}your thoughts on your bed \textit{turned to}\lebnote{Literally “they went up”}what it was that would be \textit{in the future},\lebnote{Literally “after this”}and the revealer of mysteries has made known to you what that would be.}%
\verse{And as for me, it is not because of wisdom that is in me more than \textit{any other living person}\lebnote{Literally “all of living beings”}that this mystery is revealed to me, but \textit{in order that}\lebnote{Literally “concerning the matter of”}the explanation\lebnote{Or “interpretation”}may be made known to the king and you will \textit{understand}\lebnote{Literally “know”}the thoughts of your \textit{mind}.\lebnote{Literally “heart”}}%
\verse{“You, O king, were looking and, look, there was one great statue. This statue was huge and its brilliance extraordinary, standing there before you, and its appearance was frightening.\lebnote{Or “dreadful”}}%
\verse{The head of this statue was of fine gold, its chest and its arms of silver, its belly and its thighs of bronze,}%
\verse{its legs of iron, its feet, part of them of iron and part of them of clay.}%
\verse{You were looking on \textit{until}\lebnote{Literally “until that”}a stone was chiseled out\lebnote{Or “became loose”}—that not by hands—and it struck the statue on its feet of iron and clay, and it broke them in pieces.}%
\verse{Then the iron, the clay, the bronze, the silver and the gold \textit{all at once}\lebnote{Literally “like one”}broke into pieces\lebnote{Or “were crushed”}and they became like chaff from the summer threshing floor, and the wind carried them away and any trace of them could not be found. But the stone that struck the statue became a great mountain and it filled the whole earth.}%
\verse{“This was the dream, and now we will tell its interpretation to the king.}%
\verse{You, O king, the king of kings, to whom the God of heaven has given the kingdom, the power and the might and the glory,}%
\verse{and also \textit{human beings wherever they dwell},\lebnote{Literally “and into all that dwell sons of humankind”}the animals of the field and the birds\lebnote{Aramaic “bird”}of heaven\lebnote{Or “the sky”}—he has given into your hand and made you ruler over all of them—you are the head of gold.}%
\verse{And after you another kingdom inferior to yours will arise, and another third kingdom of bronze that will rule over the whole earth.}%
\verse{And a fourth kingdom will be strong as iron, and \textit{just as}\lebnote{Literally “all of because”}iron crushes and smashes \textit{everything},\lebnote{Literally “the whole thing”}and as iron shatters all of these other metals, so it will crush and it will shatter these nations.}%
\verse{\textit{And just as}\lebnote{Literally “And that”}you saw the feet and the toes, \textit{partly potter’s clay}\lebnote{Literally “from them clay of the potter”}and \textit{partly iron},\lebnote{Literally “from them iron”}it will be a divided kingdom; and \textit{some hardness}\lebnote{Literally “from the hardness”}of the iron will be in it, \textit{just as}\lebnote{Literally “all of because that”}you have seen the iron mixed with the wet clay.}%
\verse{And as the toes of the feet were partly iron and partly clay, so part of the kingdom will be strong \textit{and partly brittle}.\lebnote{Literally “and from it, it will be brittle”}}%
\verse{And in that you saw the iron was mixed with wet clay, so \textit{they will mix in marriage}\lebnote{Literally “mixing they will in the seed of man”}and they will not cling \textit{to one another},\lebnote{Literally “this one with that one”}just as iron does not mix with clay.}%
\verse{And \textit{in the day of those kings},\lebnote{Literally “in their days of those kings”}the God of heaven will set up a kingdom that will \textit{never}\lebnote{Literally “to forever not”}be destroyed, and the kingdom will not be left for another nation,\lebnote{Or “people”}and it will bring an end to all these kingdoms, but it will stand \textit{forever}.\lebnote{Literally “to eternity”}}%
\verse{\textit{Inasmuch as}\lebnote{Literally “All of because that”}you saw that a stone from the mountain was chiseled out\lebnote{Or “became loose”}but not by hands, and that it crushed the iron, the bronze, the clay, the silver and the gold, thereby the great God made known to the king what will be \textit{in the future},\lebnote{Literally “after this”}and the dream is certain and its explanation trustworthy.”}%
\verseWithHeading{The King’s Response to Daniel and His God}{Then the king, Nebuchadnezzar, fell on his face and he paid homage to Daniel; and he commanded them to offer a grain offering and incense offering to him.}%
\verse{The king answered Daniel and said, “\textit{Truly}\lebnote{Literally “From truth that”}your God is the God of gods and the Lord of kings, and he reveals mysteries, for you are able to reveal this mystery.”}%
\verse{Then the king promoted Daniel and gave him many great gifts, and made him ruler over the whole province of Babylon and the chief prefect over all the wise men of Babylon.}%
\verse{And Daniel made a request from the king, and he appointed Shadrach, Meshach and Abednego over the affairs of the province of Babylon, while Daniel remained in the court of the king.}%
\end{biblechapter}

\begin{biblechapter} % Daniel 3
\verseWithHeading{The Faithfulness of Three Young Israelites and God’s Deliverance}{Nebuchadnezzar the king made a statue of gold. Its height was sixty cubits and its width was six cubits; he set it up in the valley of Dura in the province of Babylon.}%
\verse{Then Nebuchadnezzar the king sent directions to assemble the satraps, the prefects, and the governors, the counselors, the treasurers, the judges, the magistrates and all of the officials of the provinces to come to the dedication of the statue that Nebuchadnezzar the king had set up.}%
\verse{Then the satraps, the prefects, the governors, the counselors, the treasurers, the judges, the magistrates, and all the officials of the province were assembled for the dedication of the statue that Nebuchadnezzar the king had set up, and were standing \textit{before}\lebnote{Literally “to the front”}the statue that Nebuchadnezzar had set up.}%
\verse{Then the herald proclaimed \textit{aloud},\lebnote{Literally “with power”}“To you \textit{it is commanded},\lebnote{Literally “they are saying”}O peoples, \textit{nations and people of all languages},\lebnote{Literally “the nations and the languages”}}%
\verse{that at the time that you hear the sound of the horn, the flute, the lyre, the trigon, the harp, the drum and all kinds of music, you must fall down and you must worship the statue of gold\lebnote{Aramaic “the gold”}that Nebuchadnezzar the king has set up.}%
\verse{And \textit{whoever}\lebnote{Literally “who that”}does not fall down and worship will be thrown \textit{immediately}\lebnote{Literally “the time”}into the midst of the furnace of blazing fire.”}%
\verse{\textit{Therefore},\lebnote{Literally “All of because this”}\textit{at that time},\lebnote{Literally “at the time”}\textit{as soon as}\lebnote{Literally “as that”}all the peoples heard the sound of the horn, the flute, the lyre, the trigon, the harp and all kinds of music, all the peoples, \textit{the nations and people of all languages}\lebnote{Literally “the nations and the languages”}were falling down and were worshiping the statue of gold\lebnote{Aramaic “the gold”}that Nebuchadnezzar the king had set up.}%
\verse{\textit{Therefore}\lebnote{Literally “All because of this”}\textit{at this time}\lebnote{Literally “at it the time”}some \textit{astrologers}\lebnote{Literally “Chaldean men”}came forward and \textit{they denounced the Jews}.\lebnote{Literally “they ate their pieces of the Jews”}}%
\verse{\textit{They responded and said}\lebnote{Literally “they answered and were saying”}to Nebuchadnezzar the king, “O king, may you live \textit{forever}!\lebnote{Literally “to eternity”}}%
\verse{You, O king, have made a decree that \textit{everyone}\lebnote{Literally “all of humanity”}who hears the sound of the horn, the flute, the lyre, the trigon, the harp and the drum and all kinds of music, he shall fall down and shall worship the statue of gold.\lebnote{Aramaic “the gold”}}%
\verse{\textit{And whoever}\lebnote{Literally “And who that”}\textit{does not fall down}\lebnote{Literally “not he falls down”}and worship shall be thrown into the midst of the furnace of blazing fire.}%
\verse{However there are Judean men whom you have appointed\lebnote{Aramaic “whom you have appointed them”}over the affairs of the province of Babylon—Shadrach, Meshach, and Abednego—these men, O king, \textit{pay you no heed}\lebnote{Literally “not they pay heed to you”}and were not serving your god, and the statue of gold\lebnote{Aramaic “the gold”}that you set up they are not worshiping.”}%
\verse{Then Nebuchadnezzar said in rage and anger to bring in Shadrach, Meshach and Abednego; then they brought in these men before the king.}%
\verse{Nebuchadnezzar answered and said to them, “Is it true, Shadrach, Meshach and Abednego, that you are not serving my god, and you are not worshiping the statue of gold\lebnote{Aramaic “the gold”}that I have set up?}%
\verse{Now if you are ready so that \textit{when}\lebnote{Literally “at that time”}you hear the sound of the horn, the flute, the lyre, the trigon, the harp and the drum and all kinds of music, you fall down and you worship the statue that I have made, that will be good. But if you do not worship it, \textit{immediately}\lebnote{Literally “at that very time”}you will be thrown into the midst of the furnace of blazing fire, and \textit{who is the god}\lebnote{Literally “who he god”}who\lebnote{Or “that”}will rescue you from my hands?”}%
\verse{Shadrach, Meshach and Abednego answered and said to the king, Nebuchadnezzar, “We have no need on this matter \textit{to present a defense to you}.\lebnote{Literally “to return to you”}}%
\verse{\textit{If it is so},\lebnote{Literally “If there is”}our God, whom we serve, is able to rescue us from the furnace of blazing fire. And from your hand, O king, let him rescue us.}%
\verse{And if not, let it be known to you, O king, that \textit{we will not serve your gods},\lebnote{Literally “your gods not there is us serving”}and the statue of gold\lebnote{Aramaic “the gold”}that you have set up we will not worship.”}%
\verseWithHeading{The Wrath of the King and His Fiery Furnace of Death}{Then Nebuchadnezzar was filled with anger and the image of his face was changed \textit{toward}\lebnote{Literally “against”}Shadrach, Meshach and Abednego, so he \textit{ordered and said}\lebnote{Literally “answering and saying”}to heat up the one furnace seven times what was usual to heat it up.}%
\verse{And he commanded \textit{the strongest men of the guards}\lebnote{Literally “to the strong men, men of strength”}who were in his army to bind Shadrach, Meshach and Abednego and to throw them into the furnace of blazing fire.}%
\verse{Then these men were bound with their garments, their trousers and their turbans and their other clothing, and they were thrown into the midst of the furnace of blazing fire.}%
\verse{\textit{Therefore}\lebnote{Literally “because of this”}\textit{because}\lebnote{Literally “from that”}the word of the king was severe and the furnace was exceedingly hot, the flame of the fire killed these men who lifted up Shadrach, Meshach and Abednego.}%
\verse{But these men, the three of them, Shadrach, Meshach and Abednego, fell down into the midst of the furnace of blazing fire, and they were bound.}%
\verse{Then Nebuchadnezzar the king was astonished, and he rose up in haste and he asked, saying to his advisers, “Did we not throw three men, bound, into the midst of the fire? They answered, saying to the king, “Certainly, O king!”}%
\verse{\textit{He answered},\lebnote{Literally “was answering”}saying, “Look, I see four men unbound walking in the midst of the fire and there is no damage to them, and the\lebnote{Aramaic “its”}appearance of the fourth man resembles the son of a god.”}%
\verse{Then Nebuchadnezzar approached the door of the furnace of blazing fire, and he called out, saying, “Shadrach, Meshach and Abednego, \textit{servants}\lebnote{Literally “his servant”}of the Most High God, come out\lebnote{Or “Go out”}and come here!”}%
\verse{And the satraps, the prefects, the governors and the advisors of the king were assembling, and they saw these men, that the fire had no power over their bodies, and the hair of their heads was\lebnote{Aramaic plural}not singed, and their garments were not harmed, and the smell of fire did not come from them.}%
\verseWithHeading{The King’s Praise for the Great Deliverance of the Three Young Israelites}{Nebuchadnezzar \textit{responded},\lebnote{Literally “answering”}saying, “Blessed be their God, the God of Shadrach, Meshach and Abednego, who sent his angel\lebnote{Or “messenger”}and rescued his servants who trusted in him, and the command of the king they disobeyed, and they gave their bodies so that they did not serve and did not worship any god except their God.}%
\verse{And from me \textit{is set forth}\lebnote{Literally “setting out”}a decree that any people, nation, or language that may utter criticism against their God—the God of Shadrach, Meshach and Abednego—will be broken into pieces and their\lebnote{Aramaic “his”}house will be made like ruins. \textit{For}\lebnote{Literally “All of because”}there is not another God who is able to rescue like this God.”}%
\verse{Then the king promoted Shadrach, Meshach and Abednego in the province of Babylon.}%
\end{biblechapter}

\begin{biblechapter} % Daniel 4
\verseWithHeading{Nebuchadnezzar Declares What the Most High God Has Done for Him}{\lebnote{Daniel 4:1–4:37 in the English Bible is 3:31–4:34 in the Hebrew Bible}Nebuchadnezzar the king to all of the nations, the peoples and languages that live in the whole earth: “May your prosperity increase!}%
\verse{It is pleasing to me to recount the signs and wonders that the Most High God worked for me.}%
\verse{How great are his signs and wonders, how strong is his kingdom, \textit{an everlasting kingdom};\lebnote{Literally “a kingdom of eternity”}and his sovereignty is from generation to generation.}%
\verseWithHeading{The Dream and the King’s Search for an Interpreter}{“I, Nebuchadnezzar, was content in my house and prospering in my palace.}%
\verse{And I saw a dream and a revelation on my bed and it frightened me, and the visions of my head terrified me.}%
\verse{And a decree was \textit{sent out},\lebnote{Literally “put/placed”}ordering that all the wise men of Babylon were to be brought in before me so that they may make known to me the explanation of the dream.}%
\verse{Then the magicians,\lebnote{Or “soothsayer-priests”}the conjurers,\lebnote{Or “enchanters”}the \textit{astrologers}\lebnote{Literally “Chaldeans”}and the diviners came in and I told them the dream, but they could not make known to me its explanation.\lebnote{Or “interpretation”}}%
\verse{Then \textit{at last}\lebnote{Literally “until another”}Daniel came before me \textit{whose name was}\lebnote{Literally “who name his”}Belteshazzar, \textit{according to}\lebnote{Literally “like”}the name of my god, and \textit{in whom was the spirit of the holy gods},\lebnote{Literally “and who the spirit of gods holy in him”}\textit{and I related the dream to him}.\lebnote{Literally “and the dream before him I told”}}%
\verse{“ ‘O Belteshazzar, chief of the magicians,\lebnote{Or “soothsayer-priests”}I myself know that the spirit\lebnote{Or “a spirit”}of the holy gods is in you, and \textit{no mystery is too difficult}\lebnote{Literally “every mystery is not difficult”}for you. Now tell me the visions of my dream that I saw, and its explanation.\lebnote{Or “interpretation”}}%
\verse{Now these were the visions of my head as I was lying on my bed: I was \textit{gazing}\lebnote{Literally “seeing”}and, look, a tree was in the midst of the earth, and its height was exalted.}%
\verse{The tree grew and it became strong, 
and its height reached to heaven,\lebnote{Or “the sky”}
and \textit{it was visible to the end of the whole earth}.\lebnote{Literally “and its appearance was to the end of the whole earth”}}%
\verse{Its foliage was beautiful, 
and its fruit abundant, 
and in it was provision for all. 
Under it the animals\lebnote{Aramaic “animal”}of the field sought shade, 
and in its branches the birds of heaven\lebnote{Or “the sky”}nested, 
and from it all the living beings were fed.}%
\verse{“ ‘I was looking in the vision of my head as I lay on my bed, and look, a watcher,\lebnote{Or “angelic being”}and a holy one, came down from heaven.}%
\verse{He cried \textit{aloud}\lebnote{Literally “in strength”}and so he said:}%
\verse{“Cut down the tree and chop off its branches; 
shake off its foliage and scatter its fruit. 
Let the animals flee from under it, 
and the birds from its branches.}%
\verse{But the stump of its roots 
leave in the earth, 
\textit{along with}\lebnote{Literally “and with”}a band of iron and bronze; 
leave it in the grass of the field. 
And in\lebnote{Or “with”}the dew of heaven 
let it\lebnote{Or “him,” that is, the king}be watered, 
and with the animals\lebnote{Aramaic “animal”}
let his lot be in the grass of the earth.}%
\verse{Let his mind be changed from that of a human,\lebnote{Or “the human”}
and let the mind of an animal be given to him, 
and let seven times pass over him.}%
\verse{“ ‘This is the dream that I, Nebuchadnezzar the king, saw. Now\lebnote{Or “And”}you, Belteshazzar, declare its explanation, \textit{for}\lebnote{Literally “all of because that”}all of the wise men of my kingdom were not able to make the explanation known to me, but you are able because the spirit of holy gods is in you.’}%
\verseWithHeading{Daniel Relates and Interprets the Dream}{“Then Daniel, whose name was Belteshazzar, was distressed \textit{for some time},\lebnote{Literally “for one hour”}and his thoughts disturbed him. The king answered and he said, ‘Belteshazzar, let the dream and its explanation not disturb you.’ Belteshazzar answered and said, ‘My lord, may the dream and its explanation be for those who hate you and for your enemies.}%
\verse{The tree that you saw, which grew and became strong and its height reached to heaven\lebnote{Or “the sky”}and \textit{it was visible to the end of the whole earth},\lebnote{Literally “its appearance to the all the earth”}}%
\verse{and its foliage was beautiful and its fruit abundant, and so there was provision for all in it, and the animals\lebnote{Aramaic “animal”}of the field lived under it and in its branches nest the birds of heaven,\lebnote{Or “the sky”}}%
\verse{—it is you, O king, who have grown great and you have grown strong, and so your greatness has increased and it has reached to heaven\lebnote{Or “the sky”}and your sovereignty to the end of the earth.}%
\verse{And inasmuch that the king saw the watcher, a holy one coming down from heaven\lebnote{Or “the sky”}and he said, “Cut down the tree and destroy it, but the stump of its root in the earth leave with a band of iron and bronze in the grass of the field, and let it be watered with the dew of heaven and let his lot be with the animals of the field \textit{until seven times have passed over him}.”\lebnote{Literally “until that seven times they pass over him”}}%
\verse{This is the explanation, O king, and it is a decree of the Most High that has come upon my lord the king:}%
\verse{you will be driven away \textit{from human society}\lebnote{Literally “from the humankind”}and \textit{you will dwell}\lebnote{Literally “your dwelling will be”}with the animals\lebnote{Aramaic “animal”}of the field, and you will be caused to graze grass like the oxen \textit{yourself},\lebnote{Literally “with respect to you”}and you will be watered with the dew of heaven, and seven periods of time will pass over you until that you have acknowledged that the Most High is sovereign over the kingdom of humankind, and \textit{to whom he wills}\lebnote{Literally “that he wills”}he gives it.}%
\verse{\textit{And in that}\lebnote{Literally “And for”}they said to leave alone the stump of the tree’s root, so your kingdom will be restored for you \textit{when}\lebnote{Literally “from that”}you acknowledge that heaven is sovereign.}%
\verse{Therefore, O king, let my advice be acceptable to you and your sin remove\lebnote{Or “replace”}with righteousness and your iniquity with having mercy on the oppressed, \textit{in case there might be a prolongation of your prosperity}.’ ”\lebnote{Literally “if will be prolongation for your prosperity”}}%
\verseWithHeading{God Judges the King for His Hubris}{\textit{All this happened}\lebnote{Literally “It came upon”}to Nebuchadnezzar the king.}%
\verse{At the end of twelve months he was walking on the roof of \textit{the royal palace}\lebnote{Literally “the place of the kingdom”}of Babylon.}%
\verse{And the king answered and said, “Is this not the great Babylon which I have built as a royal palace by the strength of my own power, and for the glory of my own majesty?”}%
\verse{While the words were still in the mouth of the king, a voice from heaven came, saying, ‘To you, King Nebuchadnezzar, it is declared that the kingdom has departed from you,}%
\verse{and you will be driven away from \textit{human society}\lebnote{Literally “the humankind”}and your dwelling will be with the animals\lebnote{Aramaic “animal”}of the field and they will cause you to graze the grass like oxen, and seven times will pass over you, until you acknowledge that the Most High is sovereign over the kingdom of humankind and that he gives it to whom he wills.’}%
\verse{\textit{Immediately}\lebnote{Literally “In that hour”}the word was fulfilled \textit{concerning}\lebnote{Literally “over”}Nebuchadnezzar, and he was expelled from \textit{human society}\lebnote{Literally “the humankind”}and he ate grass like oxen, and his body was bathed with the dew of heaven until his hair was like the hair of an eagle and his nails grew like a bird’s claws.}%
\verseWithHeading{The King Praises and Extols the Most High God}{“But at the end of \textit{that period},\lebnote{Literally “the day”}I, Nebuchadnezzar, lifted up my eyes to heaven, and then my reason returned to me; and I blessed the Most High and \textit{the one who lives forever}\lebnote{Literally “the one living forever”}I praised and I honored.}%
\verse{“For his sovereignty is an everlasting sovereignty, 
and his kingdom continues from generation to generation.}%
\verse{“\textit{At that time}\lebnote{Literally “In it the time”}my reason returned to me, and also the glory of my kingdom and my majesty and splendor returned to me, and my advisers and my lords\lebnote{Or “nobles”}searched me out, and I was established over my kingdom and abundant greatness was added to me.}%
\verse{Now I, Nebuchadnezzar, praise and exalt and honor the king of heaven, for all his works are truth,\lebnote{Or “true”}and his ways are justice and that he is able to humble those who walk in pride.”}%
\end{biblechapter}

\begin{biblechapter} % Daniel 5
\verseWithHeading{The Mysterious Writing on the Wall}{Belshazzar the king made a great festival for a thousand of his lords,\lebnote{Or “nobles”}and \textit{in the presence of}\lebnote{Literally “to over against”}the thousand lords he was drinking wine.\lebnote{Or “the wine”}}%
\verse{\textit{When he tasted the wine},\lebnote{Literally “At the taste of the wine”}Belshazzar commanded that they bring the vessels of gold and silver that Nebuchadnezzar his \textit{predecessor}\lebnote{Literally “father”}had taken from the temple that was in Jerusalem, so that the king and his lords,\lebnote{Or “nobles”}his wives and his concubines may drink from them.}%
\verse{Then they brought in the vessels of gold that they took from the temple, the house of God that was in Jerusalem, and the king and his lords,\lebnote{Or “nobles”}his wives and his concubines drank from them.}%
\verse{They drank the wine and praised the gods of gold and silver, bronze, iron, wood and stone.\lebnote{All these metals have the definite article that does not have to be translated into English to retain the meaning}}%
\verse{\textit{Immediately}\lebnote{Literally “In its the hour”}\textit{human fingers}\lebnote{Literally “fingers of the hand of a man”}appeared and they wrote \textit{opposite}\lebnote{Literally “to over against”}the lampstand on the plaster of the wall of the palace of the king, and the king was watching\lebnote{Or “gazing at”}the palm of the hand that was writing.}%
\verse{Then his face changed and his thoughts terrified him, and \textit{his hip joints gave way}\lebnote{Literally “the limbs of his hip became loose”}and \textit{his knees knocked together}.\lebnote{Literally “his knees this to this they knocked together”}}%
\verse{The king cried \textit{aloud}\lebnote{Literally “with strength”}to bring in the conjurers,\lebnote{Or “enchanters”}the \textit{astrologers}\lebnote{Literally “Chaldeans”}and the diviners; the king \textit{spoke}\lebnote{Literally “answered”}and said to the wise men of Babylon, “Any man that can read this writing and can tell me its explanation will be clothed in purple and will have a necklace of gold hung around his neck and he will rule as third in authority in the kingdom.”}%
\verse{Then all the wise men of the king came in, but they were not able to read the writing or to make known its explanation.\lebnote{Aramaic “the explanation/ interpretation”}}%
\verse{Then the king, Belshazzar, became greatly terrified, and \textit{his facial features}\lebnote{Literally “shining of his face”}changed upon him, and his lords\lebnote{Or “nobles”}were perplexed.}%
\verse{Because of the words of the king and his lords,\lebnote{Or “nobles”}the queen came into \textit{the banqueting hall}\lebnote{Literally “the house of the drinking”}and the queen \textit{spoke up}\lebnote{Literally “answered”}and said, “O king, live \textit{forever},\lebnote{Literally “to eternity”}and let not your thoughts terrify you and do not let your facial expressions \textit{grow pale}.\lebnote{Literally “change”}}%
\verse{There is a man in your kingdom \textit{who has the spirit of the holy gods in him}.\lebnote{Literally “who the spirit of holy gods is in him”}And in the days of your \textit{predecessor},\lebnote{Literally “father”}enlightenment and insight and wisdom like the wisdom of the gods was found in him; and, O king, Nebuchadnezzar your \textit{predecessor}\lebnote{Literally “father”}appointed him as chief of the magicians,\lebnote{Or “soothsayer-priests”}the conjurers,\lebnote{Or “enchanters”}the \textit{astrologers},\lebnote{Literally “Chaldeans”}and the diviners. Your \textit{predecessor}\lebnote{Literally “father”}the king did this}%
\verse{\textit{because}\lebnote{Literally “all of because that”}there was found in him an excellent spirit and understanding and insight for interpreting dreams and explaining riddles and solving riddles; that is, in Daniel \textit{whom the king named}\lebnote{Literally “whom the king gave his name”}Belteshazzar. Now, let Daniel be called and he will tell the explanation.”\lebnote{Or “interpretation”}}%
\verseWithHeading{Daniel Deciphers and Interprets the Writing}{Then Daniel was brought in before the king, and the king \textit{spoke}\lebnote{Literally “answered”}and said to Daniel, “You are Daniel \textit{who are one of the exiles}\lebnote{Literally “who from sons of the exile”}of Judah\lebnote{Aramaic “Jehud”}whom my \textit{predecessor},\lebnote{Literally “father”}the king, brought from Judah.\lebnote{Aramaic “Jehud”}}%
\verse{And I have heard that a spirit of the gods is in you and enlightenment and insight and excellent wisdom was found in you.}%
\verse{And now the wise men and the conjurers\lebnote{Or “enchanters”}were brought in before me so that they could read this writing in order to make its explanation\lebnote{Or “interpretation”}known to me, but they were not able to disclose the explanation\lebnote{Or “interpretation”}of the matter.}%
\verse{But I have heard concerning you that you are able \textit{to produce interpretations}\lebnote{Literally “to interpret interpretations”}and to solve riddles; now if you are able to read the writing and to make known its explanation\lebnote{Or “interpretation”}to me, you will be clothed in purple and a necklace of gold will be placed around your neck and you will rule as third in command in the kingdom.”}%
\verse{Then Daniel answered and said before the king, “Let your gifts be for yourself or your rewards give to another; nevertheless, I will read the writing to the king and I will make known to him the explanation.\lebnote{Or “interpretation”}}%
\verse{\textit{O king},\lebnote{Literally “You, O king”}the Most High God gave the kingdom and the greatness and the glory and the majesty to Nebuchadnezzar your \textit{predecessor}.\lebnote{Literally “father”}}%
\verse{And because of the greatness that he gave to him, all the peoples, the nations and languages trembled and feared before him; whomever he wanted he killed, and whomever he wanted he let live, and whomever he wanted he honored, and whomever he wanted he humbled.}%
\verse{But \textit{when}\lebnote{Literally “as that”}his heart became arrogant and his spirit became hard so as to act proudly, he was deposed from the throne of his kingdom and the\lebnote{Or “his”}glory \textit{was taken away from him}.\lebnote{Literally “they took away”}}%
\verse{And he was driven away \textit{from human society}\lebnote{Literally “from the sons of the humankind”}and his mind was made like the animals\lebnote{Aramaic “with the animals”}and his dwelling was with the wild asses; and \textit{he was given}\lebnote{Literally “they gave him”}grass like oxen to eat, and with\lebnote{Aramaic “from”}the dew of heaven his body was bathed, until he acknowledged that the Most High God is sovereign over the kingdom of humankind, and \textit{whoever}\lebnote{Literally “to from that”}he wants he sets over it.}%
\verse{“But you his \textit{successor},\lebnote{Literally “son”}Belshazzar, have not humbled your heart \textit{even though}\lebnote{Literally “all of because that”}you knew all this.}%
\verse{And now you have exalted yourself against the Lord of heaven, and the vessels of his temple you have brought in before you, and you and your lords,\lebnote{Or “nobles”}your wives and your concubines, have been drinking wine from\lebnote{Aramaic “with”}them, and you have praised the gods of silver, gold, bronze, iron, wood, and stone\lebnote{All the materials in this list have the definite article that does not have to be translated into English}that do not see and do not hear and do not know, but the God \textit{who holds your life in his hand}\lebnote{Literally “who your breath is in his hand”}and all of your ways \textit{come from him},\lebnote{Literally “belong to him”}you have not honored.}%
\verse{So then the palm of the hand was sent out from his presence and this writing was inscribed.}%
\verse{“Now this was the writing that was inscribed: ‘Mene, Mene, Tekel and Parsin.’}%
\verse{“This is the explanation of the matter:\lebnote{Or “word, message”}‘Mene’—God has numbered your kingdom and brought an end to it.}%
\verse{“ ‘Tekel’—you have been weighed on scales and you have been found wanting.\lebnote{Or “deficient”}}%
\verse{“ ‘Peres’—your kingdom has been divided and given to the Medes and Persians.’ ”}%
\verse{Then Belshazzar commanded, and they clothed Daniel with purple and placed a necklace of gold around his neck, and they made a proclamation concerning him that he would be the third ruler in authority in the kingdom.}%
\verse{\textit{That same night}\lebnote{Literally “In it, in the night”}Belshazzar, the Chaldean king, was killed.}%
\verse{\lebnote{Daniel 5:31–6:28 in the English Bible is 6:1–29 in the Hebrew Bible}And Darius the Mede received the kingdom \textit{when he was about sixty-two years old}.\lebnote{Literally “when a son of sixty and two years”}}%
\end{biblechapter}

\begin{biblechapter} % Daniel 6
\verseWithHeading{Daniel’s Integrity and His Entrapment by His Enemies}{\textit{It pleased Darius},\lebnote{Literally “It was pleasant before Darius”}and he set up one hundred and twenty satraps over the kingdom, that they were \textit{throughout the whole kingdom},\lebnote{Literally “in all the kingdom”}}%
\verse{and over them were three administrators, of whom Daniel was one, so that these satraps were giving account to them, and the king would not be suffering loss.}%
\verse{Then Daniel \textit{began distinguishing himself}\lebnote{Literally “was distinguishing himself”}above the administrators and the satraps \textit{because}\lebnote{Literally “all of because that”}an exceptional spirit was in him, and so the king planned to appoint him over the whole kingdom.}%
\verse{Then the other administrators and satraps \textit{began to seek}\lebnote{Literally “were seeking”}to find a pretext against Daniel \textit{in connection with the kingdom},\lebnote{Literally “from the side of the kingdom”}but they were not able to find any pretext and \textit{corruption}\lebnote{Literally “being corrupt”}\textit{because}\lebnote{Literally “all because that”}he was trustworthy, and no\lebnote{Aramaic “any”}negligence or \textit{corruption}\lebnote{Literally “being corrupt”}\textit{could be found in him}.\lebnote{Literally “not could be found”}}%
\verse{Then these men \textit{said},\lebnote{Literally “were saying”}“We will not find any pretext\lebnote{Or “basis for accusation”}against this Daniel unless we find it in connection with the law of his God.”}%
\verse{So the administrators and the satraps conspired \textit{with respect to}\lebnote{Literally “on/upon”}the king and so they said to him, “Darius, O king, live \textit{forever}!\lebnote{Literally “to eternity”}}%
\verse{All of the administrators of the kingdom, and the prefects, the satraps, the counselors and the governors took counsel and have agreed to establish an edict of the king and to enforce a decree that \textit{whoever}\lebnote{Literally “all that”}will seek a prayer from any god or human except from you for up to thirty days will be thrown into \textit{the lion pit}.\lebnote{Literally “pit of lions”}}%
\verse{Now, O king, establish the edict and you must sign the document so that it cannot be changed, according to the law of the Medes and Persians which cannot be revoked.”}%
\verse{\textit{So}\lebnote{Literally “Like to before this”}the king, Darius, signed the writing and the interdict.\lebnote{Or “the writing that is the interdict”}}%
\verse{Now \textit{when}\lebnote{Literally “as that”}Daniel realized that the document was signed, he went to his house (\textit{now he had windows in his upper room that were open}\lebnote{Literally “and windows were open to him in his upper room”}toward Jerusalem), and three times \textit{daily}\lebnote{Literally “in the day”}he knelt on his knees and prayed and gave praise before his God, \textit{just as}\lebnote{Literally “all because that”}he had been doing \textit{previously}.\lebnote{Literally “from before this”}}%
\verse{Then these men came as a group and they found Daniel praying and pleading for mercy before his God.}%
\verse{Then they approached \textit{and spoke with the king}\lebnote{Literally “and saying before the king”}concerning the edict of the king, “\textit{Did you not sign an edict}\lebnote{Literally “Not an edict you signed?”}that any person who would seek anything from any God or human within thirty days except from you, O king, would be thrown into \textit{the lion pit}?”\lebnote{Literally “pit of lions”}The king answered and said, “The matter as you have just stated is certain according to the law of the Medes and Persians which cannot be revoked.”}%
\verseWithHeading{God’s Miraculous Deliverance of Daniel Again}{Then \textit{they responded}\lebnote{Literally “they answered”}and said before the king, “Daniel, who is from \textit{the exiles}\lebnote{Literally “the sons of the exile”}of Judah,\lebnote{Aramaic “Jehud”}\textit{is not paying any attention}\lebnote{Literally “not he sets his mind”}to you, O king, or to the decree that you have signed, \textit{and three times daily}\lebnote{Literally “and three times on the day”}he says his prayer.”}%
\verse{Then the king, when\lebnote{Aramaic “as ”}he heard that \textit{report},\lebnote{Literally “word”}he was extremely distressed over it; and \textit{concerning Daniel}\lebnote{Literally “upon Daniel”}\textit{he was determined}\lebnote{Literally “he set his heart”}to rescue him. And until the setting of the sun he was making every effort to deliver him.}%
\verse{Then these men came as a group to the king and \textit{said},\lebnote{Literally “saying”}“Recall, O king, that with respect to the law of\lebnote{Aramaic “to”}the Medes and Persians that \textit{any}\lebnote{Literally “all/every”}decree or edict that the king establishes cannot be changed.”}%
\verse{Then the king gave the command, and Daniel was brought in and they threw him into \textit{the lion pit}.\lebnote{Literally “pit of lions”}\textit{The king said}\lebnote{Literally “The king answered and saying”}to Daniel, “Your God, whom you serve \textit{faithfully},\lebnote{Literally “in the loyalty”}may he rescue you!”}%
\verse{And a\lebnote{Aramaic “one”}stone was brought and it was put on the entrance of the pit, and the king sealed it with his signet ring and with the signet rings of his lords,\lebnote{Or “nobles”}so that nothing would be changed concerning Daniel.}%
\verse{Then the king went to his palace and spent the night in fasting, and no food was brought in before him and his sleep fled from him.}%
\verse{Then the king got up at daybreak, at first light, and he went \textit{in haste}\lebnote{Literally “to hasten”}to \textit{the lion pit}.\lebnote{Literally “pit of lions”}}%
\verse{And \textit{when he came near}\lebnote{Literally “as coming near him”}to the pit, he cried out to Daniel with a distressed voice, and the king \textit{spoke}\lebnote{Literally “answered”}and said to Daniel, “O Daniel, servant of the living God, your God whom you serve faithfully, was he able to rescue you from the lions?”}%
\verse{Then Daniel spoke to\lebnote{Aramaic “with”}the king, “O king, live \textit{forever}!\lebnote{Literally “unto endless ages”}}%
\verse{My God sent his angel and he shut the mouth of the lions and they did not\lebnote{Aramaic “would not”}hurt me, \textit{because}\lebnote{Literally “all of because that”}before him \textit{I was found}\lebnote{Literally “it was found for me”}blameless, and also before you, O king, I have not done any wrong.”}%
\verse{Then the king was exceedingly glad over it\lebnote{That is, the situation}and commanded that Daniel be lifted up from the pit; and there was not any wound found on him, because he had trusted in his God.}%
\verse{The king then commanded, and these men were brought \textit{who had accused Daniel},\lebnote{Literally “they ate his pieces of Daniel”}and they threw them and their children and their wives into \textit{the lion pit},\lebnote{Literally “pit of lions”}and they had not reached the floor of the pit \textit{before}\lebnote{Literally “until that”}the lions had overpowered them and they had crushed all of their bones.}%
\verseWithHeading{The King Praises the God of Daniel and Daniel Prospers}{Then Darius the king wrote to all the people, the nations, and the languages living in the whole earth, “May your prosperity become great!}%
\verse{\textit{I make a decree}\lebnote{Literally “From before me is put forth a decree”}that in all the dominion of my kingdom people will be trembling and fearing before the God of Daniel, for he is the living God and endures \textit{forever}\lebnote{Literally “to eternity/endless ages”}and his\lebnote{Literally “and whose”}kingdom is one that will not be destroyed and his \textit{dominion has no end}.\lebnote{Literally “dominion his to the end”}}%
\verse{He is rescuing, delivering, and working signs and wonders in the heavens and on earth, for he has rescued Daniel from the \textit{power}\lebnote{Literally “hand”}of the lions.”}%
\verse{So this Daniel prospered during the kingdom of Darius and during the kingdom of Cyrus the Persian.}%
\end{biblechapter}

\begin{biblechapter} % Daniel 7
\verseWithHeading{Daniel’s Vision of Four Beasts and the Son of Man}{In the first year of Belshazzar the king of Babylon, Daniel saw a dream and visions\lebnote{Or “the visions”}of his head as he lay on his bed; then he wrote down the dream and the \textit{summary}\lebnote{Literally “head”}of the words \textit{as follows}:\lebnote{Literally “saying”}}%
\verse{Daniel \textit{explained}\lebnote{Literally “answered”}and said, “I was looking in my vision in the night, and look, the four winds of heaven were stirring up the great sea.}%
\verse{And four great beasts were coming up from the sea, differing \textit{from one another}.\lebnote{Literally “this from this”}}%
\verse{The first was like a lion and \textit{had}\lebnote{Literally “for him/it”}the wings of an eagle. I was watching until its wings were plucked off, and it was lifted up from the earth and it was raised on its feet like a human, and a human heart was given to it.}%
\verse{And look, another beast, a second one, looking like a bear. And it was raised up on one side and three tusks\lebnote{Or “ribs”}were in its mouth between its teeth, and so \textit{it was told},\lebnote{Literally “they were saying”}“Arise, eat much flesh!”}%
\verse{After this I was watching and look, another beast like a leopard; \textit{it had four wings}\lebnote{Literally “and there were to it four wings”}of a bird on its back, and \textit{the beast had}\lebnote{Literally “to the beast”}four heads, and dominion was given to it.}%
\verse{After this in the visions of the night I was looking and there was a fourth beast, terrifying and frightful and exceedingly strong, and \textit{it had}\lebnote{Literally “were to it”}\textit{great iron teeth},\lebnote{Literally “teeth of iron for it great”}and it was devouring and crushing, and it stamped the remainder\lebnote{Or “residue”}with its feet; and it was different from all the other beasts \textit{that preceded it}\lebnote{Literally “that preceding her/it”}\textit{and it had ten horns}.\lebnote{Literally “and ten horns for her/it”}}%
\verse{I was considering the horns, and look, another little horn came up \textit{among}\lebnote{Literally “between”}them, and three of the earlier horns \textit{were rooted out}\lebnote{Literally “dehorned”}from before it, and there were eyes like the eyes of a human\lebnote{Or “humankind”}in this horn and also a mouth that was speaking boastfully.}%
\verse{“\textit{I continued watching}\lebnote{Literally “I was watching”}until thrones were placed and an Ancient of Days sat; his clothing was like white snow and the hair of his head was like pure wool and his throne was a flame\lebnote{Or “ablaze with”}of fire and its wheels\lebnote{Aramaic “wheel”}were burning fire.}%
\verse{A stream of fire issued forth and flowed \textit{from his presence};\lebnote{Literally “from before him”}thousands upon thousands served him and ten thousand upon ten thousand stood before him. The judge sat, and the books were opened.}%
\verse{“\textit{I continued watching}\lebnote{Literally “I was watching”}then\lebnote{Aramaic “in then”}because of the noise of the boastful words of the horn who was speaking; \textit{I continued watching}\lebnote{Literally “I was watching”}until the beast was slain and its body was destroyed, and it was given over to burning with fire.}%
\verse{And as for the remainder of the beasts, \textit{their dominion was taken away},\lebnote{Literally “they took away their dominion”}but a prolongation \textit{of their life}\lebnote{Literally “in life”}was given to them for a season and a time.}%
\verse{“\textit{I continued watching}\lebnote{Literally “I was watching”}in the visions of the night, and look, with the clouds of heaven one like a son of man\lebnote{Or “mortal,” or “son of humankind”}was coming, and he came to the Ancient of Days, and \textit{was presented}\lebnote{Literally “they presented him”}before him.}%
\verse{And to him was given dominion and glory and kingship that\lebnote{Aramaic “and”}all the peoples, the nations, and languages would serve him; his dominion is a dominion \textit{without end}\lebnote{Literally “of eternity”}that will not cease, and his kingdom is one that will not be destroyed.}%
\verseWithHeading{Daniel Explains and Interprets the Vision}{“\textit{As for me, Daniel},\lebnote{Literally “I Daniel”}my spirit was troubled \textit{within me},\lebnote{Literally “in the midst of sheath”}and the visions of my head terrified me.}%
\verse{So I approached one of the attendants and I asked him\lebnote{Aramaic “from him”}about the truth concerning all this; and he told me\lebnote{Aramaic “to me”}that he would make known to me the explanation of the matter.}%
\verse{‘These great beasts which are four in number are four kings who will arise from the earth.}%
\verse{But the holy ones\lebnote{Or “saints”}of the Most High will receive the kingdom, and they will take possession of the kingdom forever, \textit{forever and ever}.’\lebnote{Literally “forever and to forever”}}%
\verse{“Then I desired to make certain concerning the fourth beast that was different from \textit{all the others}\lebnote{Literally “all of them”}—exceedingly terrifying, with its iron teeth and its claws of bronze; it devoured and crushed and stamped the remainder with its feet—}%
\verse{and concerning the horns that were on its head, and concerning the other horn that came up and \textit{before which three horns fell},\lebnote{Literally “and fell from before it three”}and this horn \textit{had eyes}\lebnote{Literally “and eyes to it”}and a mouth speaking \textit{boastfully},\lebnote{Literally “abundantly”}and its appearance was larger than its companions.}%
\verse{\textit{I continued watching},\lebnote{Literally “I was watching”}and this horn made war with the holy ones\lebnote{Or “saints”}and it prevailed over them,}%
\verse{until the Ancient of Days came and gave judgment to the holy ones\lebnote{Or “saints”}of the Most High; and the time arrived and the holy ones\lebnote{Or “saints”}took possession of the kingdom.}%
\verse{“And he said, ‘The fourth beast is the fourth kingdom that will be on the earth that will be different from all the other kingdoms, and it will devour the whole earth and it will trample it and it will crush it.}%
\verse{And as for the ten horns coming from it, from this kingdom\lebnote{Or “the kingdom”}ten kings will arise, and another will arise after them. And he will be different from the earlier ones, and he will subdue three kings.}%
\verse{And he will speak words against the Most High, and he will wear out the holy ones\lebnote{Or “saints”}of the Most High, and he will attempt to change times and law, and they will be given into his hand for a time and two times and half a time.}%
\verse{Then the court will sit, and his dominion will be removed,\lebnote{Or “taken away”}to be eradicated and to be destroyed \textit{totally}.\lebnote{Literally “until the end”}}%
\verse{And the kingdom and the dominion and the greatness of the kingdoms under the whole heaven will be given to the nation of the holy ones\lebnote{Or “saints”}of the Most High; his kingdom \textit{is an everlasting kingdom},\lebnote{Literally “is a kingdom of eternity”}and all the dominions will serve and obey him.’}%
\verse{\textit{This is the end of the account}.\lebnote{Literally “Up to here is the end of the account”}\textit{As for me, Daniel}\lebnote{Literally “I Daniel”}—my thoughts terrified me greatly and my face changed over me, but I kept the matter in my heart.”\lebnote{Or “to myself”}}%
\end{biblechapter}

\begin{biblechapter} % Daniel 8
\verseWithHeading{The Vision of a Small Horn that Oppresses Israel and Threatens the Temple}{In the third year of the kingdom of Belshazzar the king, a vision appeared \textit{to me, Daniel},\lebnote{Literally “I Daniel”}after the one that appeared to me \textit{previously}.\lebnote{Literally “in the beginning”}}%
\verse{And I saw in the vision, \textit{and when I saw},\lebnote{Literally “and it was at my seeing”}I was in Susa, the citadel that was in the province of Elam. And I saw in the vision, and \textit{I myself}\lebnote{Literally “I, I was”}was at the stream of Ulai.\lebnote{Or “Ulai Canal” (NASB); “stream of the Ulai” (NEB)}}%
\verse{And I lifted up my eyes and I saw, and look! A ram\lebnote{Hebrew “One ram”}standing \textit{before}\lebnote{Literally “to the face of”}the stream, and \textit{it had}\lebnote{Literally “for it”}two horns, and the horns were long, but the one was longer than the second, and the longer one came up \textit{after the other one}.\lebnote{Literally “in the behind of the other”}}%
\verse{I saw the ram charging westward and northward and southward, and \textit{none of the beasts stood before it},\lebnote{Literally “all the beasts not they withstood to the face of him”}and there was no rescuing from \textit{its power},\lebnote{Literally “its hand”}and it did \textit{what it wanted}\lebnote{Literally “according to its will”}and it became strong.\lebnote{Or “it/he magnified itself/himself”}}%
\verse{And I was considering this, and look, a he-goat coming from the west across the face of the whole earth, and it was not touching the ground;\lebnote{Or “earth,” or “land”}and the he-goat had \textit{a conspicuous horn}\lebnote{Literally “a horn of vision”}between its eyes.}%
\verse{Then it came toward the ram \textit{that had the two horns}\lebnote{Literally “the owner of the two horns”}that I saw standing \textit{before}\lebnote{Literally “to the face of”}the stream, and it ran at it with\lebnote{Or “in”}the rage of its power.}%
\verse{And I saw it approaching the ram and it was furious at it, and it struck the ram, and it broke its two horns, and \textit{there was not strength in the ram to stand before him},\lebnote{Literally “and not was strength in the ram to withstand to the face of him”}and he threw it down to the ground and trampled it, \textit{and there was no one who could rescue the ram}\lebnote{Literally “and not he/it was one rescuing”}from its power.}%
\verse{And the he-goat grew \textit{exceedingly great},\lebnote{Literally “until much”}and \textit{at the height of its power}\lebnote{Literally “and at his being powerful”}the great horn was broken, and four conspicuous horns\lebnote{LXX reads “others” here for “conspicuous.” The words in Hebrew can be confused for each other}came up in place of it toward the four winds of heaven.}%
\verse{\textit{And from one of them}\lebnote{Literally “And from the one from them”}came forth a horn, \textit{a little one},\lebnote{Literally “one from little”}and it grew exceedingly\lebnote{Or “pre-eminently”}toward the south, and toward the east, and toward the beautiful land.}%
\verse{And it grew up to the host of heaven, and it threw down to the ground some of the host and some of the stars and trampled them.}%
\verse{Even against the prince of the hosts it acted arrogantly and took away from him the regular burnt offering, and the place of his sanctuary was overthrown.}%
\verse{And on account of transgression, the host was given over to the horn in addition to the regular burnt offering. And it cast down truth to the ground, and it \textit{acted},\lebnote{Literally “did”}and it had success.}%
\verse{And I heard a certain holy one speaking, and a certain other holy one said to the specific one \textit{who was speaking},\lebnote{Literally “the one speaking”}“\textit{For how long}\lebnote{Literally “Until when”}is the vision concerning the regular burnt offering, and the transgression that makes desolate, and the giving over of the sanctuary and the host to trampling?”}%
\verse{And he said to me, “For two thousand three hundred evenings and mornings, then the sanctuary will be restored.”}%
\verseWithHeading{Gabriel Gives Angelic Help and Interprets the Vision}{\textit{And then}\lebnote{Literally “And it happened”}when I, Daniel, saw the vision, and I was seeking understanding, there was one standing before me with the appearance of a man.}%
\verse{And I heard the voice of a human at the Ulai, and he called and said, “Gabriel, make this man understand the vision.”}%
\verse{And he came beside \textit{where I was standing},\lebnote{Literally “my standing”}and \textit{when he came}\lebnote{Literally “at his coming”}I became terrified and I fell prostrate on my face. And he said to me, “Understand, son of man,\lebnote{Or “mortal,” or “son of humankind”}that the vision is for the time of the end.”}%
\verse{And \textit{when he spoke}\lebnote{Literally “at his speaking”}with me I fell into a trance with my face to the ground, and he touched me and made me stand on by feet.}%
\verse{And he said, “Look, I am making known to you what will happen in the period of wrath,\lebnote{Or “the wrath”}for it refers to the appointed time of the end.}%
\verse{“The ram that you saw \textit{who had two horns}\lebnote{Literally “the owner of the two horns”}represents the kings of Media and Persia.}%
\verse{“And the hairy he-goat is the \textit{kingdom}\lebnote{Literally “king”}of Greece, and the great horn that is between his eyes—he is the first king.}%
\verse{And the horn that was broken, and then there arose four horns in place of it—these are four kingdoms that will arise from his nation, but not with his power.}%
\verse{And at the end of their kingdom, when the transgressions are completed, a king will arise, fierce in countenance and skilled in riddles.}%
\verse{And his power will grow, but not by his own power, and he will cause fearful destruction, and he will succeed and \textit{he will act},\lebnote{Literally “he will do”}and he will destroy the mighty and the people of the holy ones.\lebnote{Or “saints”}}%
\verse{And by his planning\lebnote{Or “cunning”}he will make a success of deceit by\lebnote{Hebrew “in”}his hand, and in his \textit{mind}\lebnote{Literally “heart”}he will \textit{boast},\lebnote{Literally “magnify himself”}and in their ease he will destroy many, and even against the prince of princes he will rise up, and he will be broken, but not by human hands.}%
\verse{And the vision of the evening and the morning that has been \textit{described},\lebnote{Literally “told”}it is true; and you, seal up the vision, for \textit{it refers to many days to come}.”\lebnote{Literally “for to days many”}}%
\verse{And I, Daniel, was overcome, and I became ill for some days, and I \textit{performed}\lebnote{Literally “did”}the business of the king, and I was dismayed over the vision and \textit{I did not understand it}.\lebnote{Literally “there was not understanding”}}%
\end{biblechapter}

\begin{biblechapter} % Daniel 9
\verseWithHeading{Daniel’s Prayer for His People}{In the first year of Darius, the son of Ahasuerus,\lebnote{That is, Xerxes}from the offspring of the Medes, who became king over the kingdom of the Chaldeans—}%
\verse{in the first year of his kingship I, Daniel, observed in the scrolls the number of the years that it was that were to be fulfilled according to the word of Adonai to Jeremiah the prophet for the devastation of Jerusalem—seventy years.}%
\verse{Then I turned my face to the Lord God to seek him by prayer and pleas for mercy,\lebnote{Or “supplications”}in fasting and in sackcloth and ashes.}%
\verse{And I prayed to Adonai my God, and I made confession and I said, “O Lord, the great and awesome God, keeping the covenant and loyal love with those who love him and with those who keep his commandments,}%
\verse{we have sinned and we have done wrong and we acted wickedly and we rebelled and have been turning aside from your commandments and from your ordinances.}%
\verse{We have not listened to your servants the prophets, who spoke in your name to our kings, our princes and our ancestors\lebnote{Or “fathers”}and to all the people of the land.}%
\verse{“\textit{Righteousness belongs to you, O Lord},\lebnote{Literally “for you Lord the righteousness”}and on us is open shame, just as it is this day to the people of Judah and to the inhabitants of Jerusalem and to all Israel, \textit{those who are near and those who are far off}\lebnote{Literally “those near and those far”}in all the lands to which you have driven them,\lebnote{Hebrew “to which you have driven them there”}because of their infidelity which they displayed against you.}%
\verse{Adonai, on us is open shame, on our kings, on our princes, and on our ancestors,\lebnote{Or “fathers”}because we have sinned against you.}%
\verse{\textit{Compassion and forgiveness belong to the Lord, our God},\lebnote{Literally “to the Lord our God the compassion and the forgiveness”}for we have rebelled against him,}%
\verse{and we have not listened to the voice of Adonai our God, by following his law \textit{which he placed before us}\lebnote{Literally “which he gave to the face of us”}by the hand of his servants the prophets.}%
\verse{“And all Israel transgressed your law and turned aside so as not to listen to your voice, and so the curse and the oath which was written in the law of Moses, the servant of God, has been poured out upon us, because we have sinned against him.}%
\verse{And so he has carried out his words which he spoke against us and against our rulers who ruled us, to bring upon us a great calamity which was not done under all of heaven as it was done in Jerusalem.}%
\verse{Just as it is written in the law of Moses, all of this calamity has come upon us, and \textit{we have not implored}\lebnote{Literally “we not implored”}the face of Adonai our God so as to turn from our iniquities and to attend closely to your faithfulness.\lebnote{Or “truth”}}%
\verse{So Adonai has kept watch over the\lebnote{Or “his”}calamity, and now he has brought it upon us. Indeed, Adonai our God is righteous concerning all his works\lebnote{Or “acts”}that he has done, but we have not listened to his voice.}%
\verse{“And now, Lord our God, who have brought your people out from the land of Egypt with a strong hand, and you have made for yourself a name until this day—we have sinned, we have acted wickedly.}%
\verse{Lord, according to all your righteousness, please let your anger and your rage turn away from your city Jerusalem, \textit{your holy mountain},\lebnote{Literally “the mountain of your holiness”}because through our sins and through the iniquities of our ancestors\lebnote{Or “fathers”}Jerusalem and your people have become an object of mockery among all of our neighbors.}%
\verse{“And now, listen to the prayer of your servant and to his pleas for mercy. Shine\lebnote{Or “let shine” or “cause to shine”}your face upon your desolate sanctuary \textit{for your sake, O Lord}.\lebnote{Literally “for the sake of the Lord”}}%
\verse{Incline your ear, my God, and listen; open your eyes and look at our desolation and the city that is called by your name, for we are not presenting our pleas for mercy \textit{before you}\lebnote{Literally “to the face of you”}because of our righteousness, but rather because of your great compassion.}%
\verse{Lord, listen! Lord, forgive! Lord, pay heed and act! You must not delay \textit{for your sake},\lebnote{Literally “for the sake of you”}my God; because \textit{your city and your people are called by your name}.”\lebnote{Literally “your name is called over your city and over your people”}}%
\verseWithHeading{Gabriel’s Announcement and Presentation of the Seventy Sets of Seven}{Now I was still speaking and praying, and confessing my sin and the sin of my people Israel, and presenting my plea \textit{before}\lebnote{Literally “to the face of”}Adonai my God, on behalf \textit{of the holy mountain of my God}.\lebnote{Literally “the mountain of the holiness of my God”}}%
\verse{And I was still speaking in prayer, and the man Gabriel, whom I had seen in the vision \textit{previously},\lebnote{Literally “in the beginning”}\textit{in my weariness touched me}\lebnote{Literally “my being wearied, touching me”}at the time of the evening offering.}%
\verse{And he instructed me and he spoke with me and he said, “Daniel I have now come out\lebnote{Or “forth”}to teach you understanding.}%
\verse{At the beginning of your pleas for mercy, a word went out, and now I have come to declare\lebnote{Or “deliver”}it, for you are highly esteemed, and so consider the word and understand the vision.}%
\verseWithHeading{The Seventy Sets of Seven Detailed}{“Seventy weeks is decreed for your people and \textit{for your holy city},\lebnote{Literally “for the city of your holiness”}to put an end to the transgression and to seal up sin\lebnote{Or “to make an end to sin”}and to make atonement for guilt and to bring in everlasting righteousness and to seal vision and prophet and to anoint \textit{the most holy place}.\lebnote{Literally “the holy place of holy places”}}%
\verse{And you must know and you must understand\lebnote{Or “so you are to know and discern” (NASB), or “know then and understand” (e.g. NEB)}that from the time of the going out of the word to restore and build Jerusalem until an anointed\lebnote{Or “messiah”}one—a leader—will be \textit{seven weeks and sixty-two weeks};\lebnote{Literally “weeks seven and weeks sixty and two”}it will be restored and will be built with streets and a moat, but \textit{in a time of oppression}.\lebnote{Literally “in distress/oppression of the times/time”}}%
\verse{“And after the sixty and two weeks an anointed one\lebnote{Or “messiah”}shall be cut off, and \textit{he shall have nothing},\lebnote{Literally “there is not for him”}and the people of the coming leader will destroy the city and the sanctuary,\lebnote{Or “holy place”}and its end will be with the flood and on to the end there shall be war; these desolations are determined.}%
\verse{And he will make a strong covenant with the many for one week, but in half of the week he will let cease sacrifice and offering \textit{and in its place}\lebnote{Literally “and on a wing of”}a desolating abomination comes even until \textit{the determined complete destruction}\lebnote{Literally “a complete destruction and being determined”}is poured out on the desolator.”}%
\end{biblechapter}

\begin{biblechapter} % Daniel 10
\verseWithHeading{Daniel’s Vision of a Dazzling Heavenly Messenger}{In the third year of Cyrus the king of the Persians, a word\lebnote{Or “matter,” or “message”}was revealed to Daniel, who was called by his name Belteshazzar, and the word\lebnote{Or “matter,” or “message”}was reliable and it concerned a great tribulation,\lebnote{Or “conflict”}and he understood the word\lebnote{Or “matter,” or “message”}and \textit{he received understanding}.\lebnote{Literally “and understanding to him”}}%
\verse{In those days, I, Daniel, I myself\lebnote{This is an emphatic use of the first person personal pronoun}was in mourning \textit{for three whole weeks}.\lebnote{Literally “three weeks of days”}}%
\verse{I had not eaten any choice food, and meat and wine did not enter my mouth, and \textit{I did not use any ointment}\lebnote{Literally “I have not anointed at all”}\textit{until the end of three whole weeks}.\lebnote{Literally “until being full/complete three weeks of days”}}%
\verse{And then on the twenty-fourth day of the first month, I myself\lebnote{This is an emphatic use of the first person personal pronoun}was on the bank of the great river; that is, the Tigris.}%
\verse{And I lifted up my eyes and I saw,\lebnote{Or “looked”}and there was a man, and he was dressed in linen, and his waist was girded with the gold of Uphaz.}%
\verse{Now his body was like turquoise,\lebnote{Or “yellow jasper” or other gold-colored stones; the exact identity of ancient precious stones is difficult}and his face was like the appearance of lightning, and his eyes were like torches of fire, and his arms and his legs were like the gleam of polished bronze, and the sound of his words was like the sound of a multitude.}%
\verse{And I saw, I, Daniel \textit{alone},\lebnote{Literally “to alone me”}the vision; and the people who were with me \textit{did not see}\lebnote{Literally “not they saw”}the vision; nevertheless, a great trembling fell upon them and they fled in order to hide themselves.}%
\verse{And I myself, Daniel, alone saw this great vision, and as a result \textit{no strength was left in me}\lebnote{Literally “not was left in me strength”}and \textit{my complexion grew deathly pale},\lebnote{Literally “and complexion my changed over me to destruction”}and I did not retain any strength.}%
\verse{And I heard the sound of his words, and \textit{when I heard}\lebnote{Literally “at my hearing”}the sound of his words I myself began falling into a trance on my face, \textit{with my face to the ground}.\lebnote{Literally “and my face toward the ground”}}%
\verse{And look, a hand touched me and it roused me to my knees\lebnote{Hebrew “on my knees”}and the palms\lebnote{Hebrew “palm”}of my hands.}%
\verse{And he said to me, “Daniel, a man beloved, pay attention to the words that I am speaking to you and \textit{stand upright where you are},\lebnote{Literally “stand on your position”}for I have now been sent to you.” And \textit{while he was speaking}\lebnote{Literally “at his speaking”}with me this word,\lebnote{Or “matter,” or “message”}I stood up trembling.}%
\verse{And he said to me, “You must not fear, Daniel, for from the first day that you set your heart to understand and to humble yourself \textit{before}\lebnote{Literally “to the face of”}your God, your words were heard, and I myself\lebnote{This is an emphatic use of the first person personal pronoun}have come because of your words.}%
\verse{But the prince of the kingdom of Persia stood \textit{before me}\lebnote{Literally “to front of me”}for twenty-one days. And look,\lebnote{Or “finally”}Michael, one of the chief princes, came to assist me, and I left him there beside the king of the Persians.}%
\verse{And I have come\lebnote{Hebrew “I came”}to instruct you about what\lebnote{Hebrew “that”}will happen to your people \textit{in the future},\lebnote{Literally “at the end of the days”}for there is a further vision here \textit{for the future}.\lebnote{Literally “for the days”}}%
\verse{And while he was speaking with me according to these words, I turned my face toward the ground and I was speechless.}%
\verse{Then look, there was \textit{one in the form of a human};\lebnote{Literally “as the likeness of a son of man”}he touched my lips and I opened my mouth and I spoke and I said to the one standing \textit{before me},\lebnote{Literally “to opposite me”}“My lord, because of the vision my anxieties fell upon me and I have not retained my strength.}%
\verse{So \textit{how am I, a servant of my lord}\lebnote{Literally “how is he able a servant of my lord this”}to speak with \textit{you, my lord},\lebnote{Literally “with my lord this”}and I \textit{just now}\lebnote{Literally “from now”}\textit{have no strength in me}?”\lebnote{Literally “not it remains in me strength”}}%
\verse{And \textit{he again touched}\lebnote{Literally “he added to and touched”}me, \textit{the one in the form of a human},\lebnote{Literally “like vision of a man”}and he strengthened me.}%
\verse{And he said, “You must not fear, O beloved man. \textit{Peace be to you};\lebnote{Literally “Peace to/for you”}be strong and be courageous!” And \textit{when he spoke}\lebnote{Literally “at his speaking”}with me, I was strengthened and I said, “Let my lord speak, for you have strengthened me.”}%
\verse{Then he asked, “Do you know why I have come to you? And now I return to fight against the prince of Persia and I myself\lebnote{This is an emphatic use of the first person personal pronoun}am going, and look,\lebnote{Or “listen”}the prince of Javan\lebnote{That is, Greece}will come.}%
\verse{However, I will tell you \textit{what is inscribed}\lebnote{Literally “the being inscribed”}in the book of truth, and there is not one who contends with me against these beings \textit{except}\lebnote{Literally “but if”}Michael, your prince.”}%
\end{biblechapter}

\begin{biblechapter} % Daniel 11
\verseWithHeading{A Survey of the Kings of the North and South}{“And I, in the first year of Darius the Mede, \textit{I stood}\lebnote{Literally “my standing”}as a support and as a protection for him.}%
\verse{And now I will reveal the truth to you. Look! Still three kings will arise in Persia, and the fourth will get abundance and great wealth, even more than all of them, and \textit{when he becomes strong}\lebnote{Literally “at his strength”}through his wealth, then he will stir up \textit{everyone}\lebnote{Literally “the all”}against the kingdom of Javan.\lebnote{That is, Greece}}%
\verse{And a mighty king will arise, and he will rule with great dominion, and he will do \textit{as he pleases}.\lebnote{Literally “as his pleasure”}}%
\verse{But \textit{as he rises in power},\lebnote{Literally “as his rising/to rise”}his kingdom will be divided toward the four winds of heaven, but not to his posterity, and not according to his dominion by which he ruled, for his kingdom will be uprooted and be given \textit{to others besides them}.\lebnote{Literally “to others from to alone these”}}%
\verse{“Then the king of the south will grow strong and also \textit{one of his officials},\lebnote{Literally “from among his officials”}and he will grow stronger \textit{than him}\lebnote{Literally “over him”}and he will rule \textit{a dominion greater than his dominion}.\lebnote{Literally “a dominion great dominion his”}}%
\verse{And at the end of some years they will make an alliance, and the daughter of the king of the south will come to the king of the north to make a peace treaty, but \textit{she will not retain her position of power},\lebnote{Literally “she will not retain the power of her arm”}and his offspring\lebnote{Or “power”}will not endure, and she will be given up, she and her attendants and her child supporting her, in those times.\lebnote{Hebrew “in the times”}}%
\verse{And a branch from her roots\lebnote{That is, a member of her family}will rise up in his place, and he will come against the army and he will enter the stronghold of the king of the north and he will take action against them and he will prevail.}%
\verse{And also their gods with their idols and with the \textit{precious vessels},\lebnote{Literally “vessels of their desire”}silver and gold he will take to Egypt into captivity, and for years \textit{he will leave the king of the north alone}.\lebnote{Literally “he will stand back from the king of the north”}}%
\verse{And then he\lebnote{That is, the king of the north}will come into the kingdom of the king of the south, but he will return to his land.}%
\verse{“But his sons will wage war and they will gather a multitude of great forces and \textit{he will advance with great force},\lebnote{Literally “he will come, to come”}and he will overflow like a flood and he will pass through and he will return, and they will wage war up to his fortress.}%
\verse{And the king of the south will become furious, and he will go and he will battle against him, against the king of the north; and he will muster a great multitude, and the multitude will be given into his hand.}%
\verse{When the multitude is carried off, his heart will be exalted and he will overthrow \textit{tens of thousands},\lebnote{Literally “myriads”}but he will not prevail.}%
\verse{And the king of the north will again raise a multitude, greater than the former, and \textit{at the end of some years}\lebnote{Literally “to the end of the times years”}he will surely come with a great army and with great supplies.\lebnote{Or “resources”}}%
\verse{“And in these times many will rise up against the king of the south, and \textit{the violent ones of your people}\lebnote{Literally “the sons of the lawless ones of your people”}will lift themselves \textit{to fulfill}\lebnote{Literally “to cause to stand”}the vision, but they will fall.}%
\verse{And the king of the north will come, and he will throw up siege ramps\lebnote{Hebrew “ramp”}and capture a city of fortifications and the military forces of the south and \textit{his choice troops}\lebnote{Literally “the army of his chosen troops”}will not stand, for\lebnote{Hebrew “and”}there is no strength left to resist.}%
\verse{And the one coming to him will act \textit{according to}\lebnote{Literally “as”}his pleasure, and there is no one who \textit{will stand}\lebnote{Literally “standing”}\textit{before him},\lebnote{Literally “to the face of him”}and he will stand in \textit{the beautiful land}\lebnote{Literally “in the land of the beauty”}and complete destruction will be in his power.}%
\verse{And he will set his face to come with the authority of his whole kingdom and will form an agreement;\lebnote{Or “treaty,” or “peace proposal”; the Hebrew is difficult}and he will \textit{act},\lebnote{Literally “do, perform”}and the daughter of women he will give to him to destroy it,\lebnote{Or “her”}but the ploy will not succeed \textit{and she will not support him}.\lebnote{Literally “and not for him will she be”}}%
\verse{And he will turn his face to the coastlands, and he will capture many, but\lebnote{Hebrew “and”}a commander will end his insults to him \textit{so that instead his insults will turn back upon him}.\lebnote{Literally “so that not his insults he will return against him”}}%
\verse{And he will turn back his face toward\lebnote{Hebrew “to”}the strongholds of his land, but he will stumble and he will fall and will not be found.}%
\verse{“Then in his place will arise one sending an official throughout the glory\lebnote{Or “splendor”}of his kingdom, and \textit{in a few days}\lebnote{Literally “in some days”}he will be broken, but not in anger and not in battle.}%
\verse{And in his place a despicable person will arise on whom\lebnote{Hebrew “him”}they have not conferred the majesty of the kingdom, and he will come in \textit{without warning}\lebnote{Literally “in ease”}and he will seize the kingdom by deceit.}%
\verse{And \textit{before him}\lebnote{Literally “to the face of him”}\textit{mighty military forces}\lebnote{Literally “military forces of the flood”}will be utterly swept away, and they will be broken, and also the leader of the covenant.}%
\verse{And after an alliance is made with him, he will act deceitfully, and he will rise and he will become powerful with few people backing him.}%
\verse{In a time of ease and in the rich parts of the province, he will come and he will do what \textit{his predecessors}\lebnote{Literally “his fathers and the father of his fathers”}did not do; he will distribute plunder and spoil and possessions to them, and he will devise his plans against fortifications, but only \textit{for a time}.\lebnote{Literally “until time”}}%
\verse{And he will stir up his power and his heart against the king of the south and \textit{with a much greater and stronger army};\lebnote{Literally “with an army great and numerous exceedingly”}but he will not succeed, for they will devise plans against him.}%
\verse{And those who eat of his royal rations will break him and his army will be overwhelmed, and many will fall, slain.}%
\verse{\textit{And two of the kings}\lebnote{Literally “And two of them, the kings”}\textit{will bend their hearts}\lebnote{Literally “their heart”}to evil. And at the \textit{same table}\lebnote{Literally “one table”}they will speak lies, but what is discussed will not succeed, for still an end is coming at the appointed time.}%
\verse{Then he will return to his land with many possessions, but his heart will be set against \textit{the holy covenant},\lebnote{Literally “covenant of holiness”}and he will take action and he will return to his land.}%
\verse{“At the appointed time he will return and he will come into the south, but it will not be \textit{as it was before}.\lebnote{Literally “as the former and as the last”}}%
\verse{And the ships of Kittim will come against him, and he will lose heart, and he will turn back, and he will be enraged \textit{against the holy covenant},\lebnote{Literally “against the covenant of holiness”}and he will take action, and he will turn back, and he will pay attention to those who forsake \textit{the holy covenant}.\lebnote{Literally “the covenant of holiness”}}%
\verse{And military forces from him \textit{will occupy}\lebnote{Literally “will stand forth”}and will profane\lebnote{Or “desecrate”}the \textit{sanctuary stronghold},\lebnote{Literally “the sanctuary, the stronghold”}and they will abolish the regular burnt offering, and they will set up the abomination that causes desolation.}%
\verse{“And those who violate the covenant he will seduce with flattery, but the persons who know their God\lebnote{Hebrew “the ones who know his God”}will stand firm and will take action.}%
\verse{And those who have insight \textit{will instruct}\lebnote{Literally “make understand”}the many, but they will fall by sword and by flame, by captivity and by plunder \textit{for some time}.\lebnote{Literally “during days”}}%
\verse{And \textit{when they fall}\lebnote{Literally “in/at their falling”}they will receive little help, and many will join with them in hypocrisy.}%
\verse{And even some of those \textit{who have insight}﻿\lebnote{Literally “the ones who produce insight”}will fall\lebnote{Or “stumble”}in order for them to be refined by it, and to be purified and cleansed until the time of the end, for the appointed time is still to come.}%
\verse{“Then the king will do \textit{as he pleases},\lebnote{Literally “as his pleasure”}and he will exalt himself and will consider himself above any god, and he will speak horrendous things \textit{against}\lebnote{Literally “over”}the God of gods, yet he will succeed until the period of anger is finished, for what is determined will be done.}%
\verse{He will not pay respect to the gods of his ancestors,\lebnote{Or “fathers”}or to the darling of women, and not to any god will he pay respect, for he will consider himself great over all gods.}%
\verse{But \textit{instead}\lebnote{Literally “in his place”}he will honor the god of fortresses, a god whom his ancestors\lebnote{Or “fathers”}did not know. He will honor him with gold, and with silver, and with precious stones\lebnote{Hebrew “precious stone”}and with costly gifts.}%
\verse{And he will deal with the \textit{fortified strongholds}\lebnote{Literally “the fortifications of strongholds”}with the help of \textit{a foreign god};\lebnote{Literally “the god of a foreign land”}and he will increase wealth for whoever will acknowledge him, and he will cause them to rule\lebnote{Or “appoint”}over the many, and he will distribute land for a price.}%
\verse{“And at the time of the end the king of the south will attack him, and the king of the north will storm against him with chariots\lebnote{Hebrew “chariot”}and with horsemen and with many ships, and he will advance against the countries and \textit{he will sweep through like a flood}.\lebnote{Literally “and he will overflow them and he will pass through”}}%
\verse{And he will come \textit{into the beautiful land}\lebnote{Literally “into the land of the beauty”}and many will fall victim, but these will escape from his power: Edom and Moab and the best part\lebnote{A difficult word to translate in this context: possibly it means “the foremost” (NASB), or it may mean “the remnant” (NEB)}of the \textit{Ammonites}.\lebnote{Literally “sons/children of Ammon”}}%
\verse{And he will stretch out his hand against countries and the land of Egypt \textit{will not escape}.\lebnote{Literally “not it will be to escape”}}%
\verse{And he will rule over the treasures of gold and the silver and over all the precious things of Egypt; and the Libyans and the Cushites will follow in his footsteps.}%
\verse{But reports will terrify him from the east and from the north, and he will go out with great fury to destroy and to exterminate many.}%
\verse{And then he will pitch the tents of his palace between the sea and \textit{the beautiful holy mountain},\lebnote{Literally “the mountain of the beauty holiness”}and he will come to his end, and there is no one\lebnote{Or “not one”}helping him.”}%
\end{biblechapter}

\begin{biblechapter} % Daniel 12
\verseWithHeading{The Time of the Judgment and the Resurrection of Two Groups}{“Now at that time, Michael, the great prince, will arise, \textit{the protector}\lebnote{Literally “the one who stands with”}over the sons of your people, and it will be a time of distress that has not been since your people \textit{have been}\lebnote{Literally “to be” or “being”}a nation until that time. And at that time your people will escape, \textit{everyone who is found}\lebnote{Literally “all of the being found”}written in the scroll.\lebnote{Or “book”}}%
\verse{And many from those sleeping \textit{in the dusty ground}\lebnote{Literally “of the ground of dust”}will awake, some to \textit{everlasting life}\lebnote{Literally “life of eternity”}and some to disgrace and \textit{everlasting contempt}.\lebnote{Literally “contempt of eternity”}}%
\verse{But the ones having insight will shine like the brightness of the expanse, and the ones providing justice for the many\lebnote{Or “leading the many in righteousness”}will be like the stars \textit{forever and ever}.\lebnote{Literally “to eternity and ever”}}%
\verse{But you, Daniel, keep the words secret and seal the scroll\lebnote{Or “book”}until the time of the end; many will run back and forth and knowledge will increase.”}%
\verse{Then I looked, I myself, Daniel, and look, there were two others standing: \textit{one on this bank of the stream and one on the other}.\lebnote{Literally “one here on the bank of the stream and one there on the bank of the stream”}}%
\verse{Then he said to the man who was clothed in linen who was above\lebnote{Hebrew “from above”}the water of the stream,\lebnote{Or “river”}“\textit{How long until}\lebnote{Literally “Until when”}the end of the wonders?”}%
\verse{And I heard the man who was clothed in linen who was above\lebnote{Hebrew “from above”}the water of the stream,\lebnote{Or “river”}and he raised his right hand and his left hand to heaven and he swore \textit{by the one who lives forever}\lebnote{Literally “by the life of the eternity”}that an appointed time, appointed times, and half an appointed time would pass when the shattering of \textit{the power of the holy people}\lebnote{Literally “the power of the people of holiness”}\textit{would be completed};\lebnote{Literally “to come to an end”}then all these things will be accomplished.}%
\verseWithHeading{Daniel Seeks Additional Insights}{Now I myself heard, but I did not understand, and I said, “My lord, what will be the outcome of these things?”}%
\verse{And he said, “Go, Daniel, for the words are secret and are sealed up until the time of the end.}%
\verse{Many will be purified and will be cleansed and will be refined, but the wicked will act wickedly \textit{and none of the wicked will understand},\lebnote{Literally “and not will understand all the wicked”}but \textit{those who have insight}\lebnote{Literally “the ones having insight”}will understand.}%
\verse{And from the time the regular burnt offering is removed and the abomination that causes desolation \textit{is set up}\lebnote{Literally “to be set up” or “to put in place”}there will be one thousand two hundred and ninety days.}%
\verse{Happy is the one who is persevering, and attains to the one thousand three hundred and thirty-five days.}%
\verse{But you, go on to the end and rest, and you will arise for your allotted inheritance at the end of the days.”}%
\end{biblechapter}

