\biblebook{Deuteronomy}

\begin{biblechapter} % Deuteronomy 1
\verseWithHeading{Preamble} These are the words that Moses spoke to all of Israel \textit{on the other side of}\lebnote{Literally “in the beyond of”} the Jordan in the desert,\lebnote{Or “the wilderness”} in the desert plateau\lebnote{Or “desert plain”; others render this as a named location, “Arabah”} opposite Suph, between Paran and between Tophel and Laban and Hazeroth and Dizahab.
\verse It is a journey of \textit{eleven days}\lebnote{Literally “one and ten days”} from Herob \textit{by the way of Mount Seir}\lebnote{Literally “by the mountain of Seir”} up to Kadesh Barnea.
\verse \textit{And it was}\lebnote{Literally “and it happened” or “and then”} in the fortieth year, on the eleventh month, on the first day of the month, Moses spoke to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} according to all that Adonai had instructed him to speak to them.
\verse This happened \textit{after defeating}\lebnote{Literally “after he had struck down”} Sihon king of the Amorites, who was reigning\lebnote{Or “dwelling,” and in this context “reigning”} in Heshbon, and Og the king of Bashan, who was reigning\lebnote{Or “dwelling,” and in this context “reigning”} in Ashtaroth in Edrei.
\verse On the other side of \lebnote{Literally “in the beyond of”} the Jordan in the land of Moab Moses began to explain this law,\lebnote{Hebrew \textit{torah}} \textit{saying}:\lebnote{Literally “to say”}
\verseWithHeading{Historical Prologue} “Adonai our God spoke to us at Horeb, \textit{saying},\lebnote{Literally “to say”} ‘You have stayed \textit{long enough}\lebnote{Literally “much time”} at this mountain.
\verse Turn now and \textit{move on},\lebnote{Literally “journey on with respect to yourselves” or “move yourselves out”} and go into the hill of the Amorites\lebnote{Hebrew “Amorite”} and to all of the neighboring regions\lebnote{Or “peoples”} in the Jordan \textit{Valley}\lebnote{Literally “Arabah,” arid stretches of land.} in the hill country and in the Negev and in the coastal area along the sea, into the land of the Canaanites\lebnote{Hebrew “Canaanite”} and into the Lebanon, as far as the great river Euphrates.
\verse Look, I have set the land \textit{before you};\lebnote{Literally “before your face”} go and take possession of the land that Adonai swore to your ancestors,\lebnote{Or “fathers”} to Abraham, to Isaac, and to Jacob, to give it to them and to their offspring after them.’
\verse “And I spoke to them at that time, \textit{saying},\lebnote{Literally “to say”} ‘I am not able to bear you alone.
\verse Adonai your God has multiplied you, and look; you are today as the stars of the heaven \textit{in number}.\lebnote{Literally “with respect to multitude/abundance”}
\verse Adonai, the God of your ancestors,\lebnote{Or “fathers”} may he add to you as you are now a thousand times, and may he bless you just as he \textit{promised you}.\lebnote{Literally “spoke to you”}
\verse How can I bear you \textit{by myself},\lebnote{Literally “to me alone”} your burden and your load and your strife?
\verse Choose for yourselves \textit{wise and discerning and knowledgeable men}\lebnote{Literally “men wise and discerning and knowledgeable”} for each of your tribes, and I will appoint them as your leaders.’
\verse “And you answered me, and you said, ‘The thing you have said to do is good.’
\verse And so I took the leaders of your tribes, wise and knowledgeable men, and then I appointed them as leaders over you as commanders of groups of thousands and commanders of groups of hundreds and commanders of groups of fifties and commanders of groups of tens as officials for your tribes.
\verse And at that time I instructed your judges, saying, ‘\textit{hear out your fellow men},\lebnote{Literally “hear between your brothers,” with the idea of listening carefully in this context} and then judge fairly\lebnote{Or “righteously”} between a man and between his brother and between \textit{his opponent who is a resident alien}.\lebnote{Literally “between his resident alien/non-Israelite who dwells temporarily among Israel”}
\verse You must not \textit{be partial}\lebnote{Literally “recognize faces”} \textit{in your judgment}; \lebnote{Literally “in the process of rendering a judgment”} hear out the small person as also the great person; \textit{do not be intimidated by any person},\lebnote{Literally “do not fear before the faces of any man”} because the judgment is God’s; and the case that is too difficult\lebnote{Or “hard”} for you, bring it to me, and I will hear it out.’
\verse And so I instructed you at that time concerning all of the things that you should do.
\verse “Then we set out from Horeb, and we went through the whole of that great and terrible desert that you saw on the way to the hill country of the Amorites\lebnote{Hebrew “Amorite”} as Adonai our God had commanded us, and so we came up to Kadesh Barnea.
\verse I said to you, ‘\textit{You have reached}\lebnote{Literally “You have come up to”} the hill country of the Amorites\lebnote{Hebrew “Amorite”} that Adonai our God is giving to us.
\verse See, Adonai your God has set before you the land; go up and possess it as Adonai the God of your ancestors\lebnote{Or “fathers”} said to you; do not fear and do not be dismayed.’
\verse “Then all of you approached me, and you said, ‘Let us send men \textit{before us},\lebnote{Literally “before our faces” or “ahead of us”} and let them explore the land for us, and let them bring back \textit{a report}\lebnote{Literally “a word”} to us concerning the way that \textit{we should take}\lebnote{Literally “we should go up by it”} and concerning the cities that we shall come to.’
\verse The plan was good \textit{in my opinion},\lebnote{Literally “in my eyes”} and so I took from among you twelve men, \textit{one from each tribe}.\lebnote{Literally “man one from the tribe”}
\verse And they \textit{set out}\lebnote{Literally “turned”} and \textit{went up into the hill country},\lebnote{Literally “went up toward the hill country mountain”} and they went up to the wadi\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season} of Eschol, and they spied out the land.
\verse They took in their hands\lebnote{Hebrew “hand”} \textit{some of the fruit}\lebnote{Literally “from the fruit”} of the land, and they brought it down to us, and they brought to us back \textit{a report},\lebnote{Literally “a word”} and they said, ‘The land that Adonai our God is giving to us is good.’
\verse But you were not willing to go up, and you rebelled against the \textit{command}\lebnote{Literally “mouth”} of Adonai your God.
\verse And you grumbled in your tents, and you said, ‘Because of the hatred of Adonai toward us he has brought us out from the land of Egypt to give us into the hand of the Amorites\lebnote{Hebrew “Amorite”} to destroy us.
\verse Where can we go up? Our brothers have \textit{made our hearts melt},\lebnote{Literally “caused to melt our hearts”} \textit{saying},\lebnote{Literally “to say”} “The people are greater\lebnote{Or “bigger”} and taller than we are,\lebnote{Hebrew “than us”} and there are great fortified cities reaching up to heaven, and we saw the sons of the Anakites living there.” ’
\verse “And so I said to you, ‘Do not be terrified, and do not fear them.
\verse Adonai your God, who is going \textit{before you},\lebnote{Literally “to your faces”} will himself\lebnote{The Hebrew pronoun indicates emphasis} fight for you, \textit{just as}\lebnote{Literally “like all that”} he did for you in Egypt before your eyes,
\verse and just as he did in the wilderness\lebnote{Or “desert”} when\lebnote{Or “where”} you saw that\lebnote{Or “how”} Adonai your God carried you, just as someone\lebnote{Or “a man”} carries his son, all along the way that you traveled until \textit{you reached}\lebnote{Literally “you came”} this place.’
\verse But through all of this you did not trust in Adonai your God,
\verse \textit{who goes}\lebnote{Literally “is the one going”} \textit{before you}\lebnote{Literally “before your faces”} on your\lebnote{Hebrew “the”} way, seeking a place for your encampment, in fire at night and in a cloud by day, to show you the way that \textit{you should go}.\lebnote{Literally “you should go in it”}
\verse “Then Adonai heard the sound of your words, and he was angry, and he swore, \textit{saying},\lebnote{Literally “to say”}
\verse ‘No one of these men\lebnote{The beginning of an oath in Hebrew; the text emphatically denies anyone of that evil generation the privilege of seeing the promised land} of this evil generation will see the good land that I swore to give to your ancestors,\lebnote{Or “fathers”}
\verse except Caleb, the son of Jephunneh; he himself\lebnote{Emphatic use of the pronoun} shall see it, and to him I will give the land upon which he has trodden and to his sons\lebnote{Or “descendants”} because \textit{he followed Adonai unreservedly}.’\lebnote{Literally “he filled his hands after Adonai”}
\verse Even with me Adonai was angry because of you, saying, ‘Not even you shall enter there.
\verse Joshua, the son of Nun, \textit{your assistant},\lebnote{Literally “the one standing before you”} will go there; encourage him because he will cause Israel to inherit it.
\verse And your little children, who you thought shall become plunder, and your sons, who do not today know good or bad, shall themselves\lebnote{The Hebrew pronoun is used for emphasis} go there, and I will give it to them, and they shall take possession of it.
\verse But you turn and set out in the direction of the wilderness by way of the \textit{Red Sea}.’\lebnote{Literally “sea of reeds”}
\verse “You replied and said to me, ‘We have sinned against Adonai, and now we will go up and fight according to all that Adonai our God commanded us’; and so each man fastened on \textit{his battle gear},\lebnote{Literally “his vessels of battle”} and you regarded it as easy to go up into the hill country.
\verse So Adonai said to me, ‘Say to them, “You shall not go up, and you shall not fight because I am not in your midst; you will be defeated \textit{before}\lebnote{Literally “in the faces of your enemies”} your enemies.” ’
\verse So I spoke to you, but you did not listen; you rebelled against the \textit{command of Adonai};\lebnote{Literally “mouth of Adonai”} you behaved presumptuously, and you went up into the hill country.
\verse The Amorites living in the hill country went out \textit{to oppose you}\lebnote{Literally “to meet you”} and chased you as a swarm of wild honey bees do; and so they \textit{beat}\lebnote{Literally “cut”} you down in Seir as far as Hormah.
\verse So you returned and wept \textit{before Adonai};\lebnote{Literally “before the faces of Adonai”} but Adonai did not listen to your voice and did not pay any attention to you.
\verse You stayed in Kadesh many days; such were the days that you stayed there.
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 2
\verse “Then we turned and set out toward the wilderness\lebnote{Or “desert”} in the direction of the \textit{Red Sea},\lebnote{Literally “sea of reeds”} as Adonai told me, and we went around Mount Seir for many days.
\verse Adonai spoke to me, \textit{saying},\lebnote{Literally “to say”}
\verse ‘Long enough you have been skirting this mountain; turn yourselves north,
\verse and instruct\lebnote{Or “command”} the people, \textit{saying},\lebnote{Literally “to say”} “You are about to cross through the territory of your brothers, the descendants\lebnote{Or “sons”} of Esau, who are living in Seir; they will be afraid of you, and so be very careful.
\verse Do not get involved in battle\lebnote{Or “provoke”} with them, for I will not give you any of their land, not even \textit{a foot’s breadth}\lebnote{Literally “a sole’s foot of space”} of it; since I have given Mount Seir as a possession for Esau.
\verse You shall buy food from them so that you may eat; and also you shall purchase water from them with money so that you may drink.
\verse The fact of the matter is, Adonai your God has blessed you in \textit{all the work you have done};\lebnote{Literally “the work of your hand”} he knows \textit{your travels}\lebnote{Literally “your journeying”} with respect to this great wilderness; forty years Adonai your God has been with you; you have not lacked a thing.” ’
\verse And so we passed by our brothers, the descendants\lebnote{Or “sons”} of Esau, who live in Seir, past the road of the Arabah,\lebnote{Or “Jordan Valley” since the Arabah is an extension of it} from Elath and Ezion Geber, and we turned and traveled along the route of the desert\lebnote{Or “wilderness”} of Moab.
\verse And Adonai said to me, ‘You shall not attack Moab, and you shall not engage in war\lebnote{Or “battle”} with them, for I will not give you any of his land as a possession; I have given Ar to the descendants\lebnote{Or “sons”} of Lot as a possession.’
\verse (The Emim previously lived in it, a people large,\lebnote{Or “great” in the sense of influential} numerous, and tall, like the Anakites.
\verse They were reckoned also as Rephaim as the Anakites were; but the Moabites called them Emim.
\verse The Horites previously lived in Seir, but the descendants\lebnote{Or “sons”} of Esau dispossessed them and destroyed them \textit{from among themselves},\lebnote{Literally “from their presence”} as Israel did with respect to the land of their\lebnote{Literally “its/his”} possession that Adonai gave to them.)
\verse So now arise and cross over the wadi\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season} of Zered yourselves; and so we crossed the wadi\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season} of Zered.
\verse Now the \textit{length of time}\lebnote{Literally “days”} that we had traveled from Kadesh Barnea until the time when we crossed the wadi\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season} of Zered was thirty-eight years, until the perishing of all of that generation; that is, the men of war from the midst of the camp as Adonai had sworn to them.
\verse The hand of Adonai was against them to root them out from the midst of the camp until they perished completely.
\verse “\textit{And then}\lebnote{Literally “And it happened”} when all the men of war\lebnote{Or “battle”} \textit{had died}\lebnote{Literally “had finished/completed to die”} from among the people,
\verse Adonai spoke to me, \textit{saying},\lebnote{Literally “to say”}
\verse ‘You are about to cross over the boundary of Moab \textit{today}\lebnote{Literally “the day”} at Ar.
\verse When you approach \textit{the border of}\lebnote{Literally “opposite”} the \textit{Ammonites},\lebnote{Literally “the sons/children of Ammon”} you shall not harass them, and you shall not get involved in battle with them, for I have not given the land of \textit{the Ammonites}\lebnote{Literally “the sons/children of Ammon”} to you as a possession; because I have given it to the descendants\lebnote{Or “sons”} of Lot as a possession.
\verse (It is also considered the land of Rephaim; Rephaim lived in it \textit{previously},\lebnote{Literally “before”} and the Ammonites called them Zamzummim,
\verse a people great and numerous and as tall as the Anakites; Adonai destroyed them from before them,\lebnote{That is, the Ammonites} and they dispossessed them and settled in place of them,
\verse just as he did for the descendants\lebnote{Or “sons”} of Esau, who live in Seir, when he destroyed the Horites from \textit{before them}\lebnote{Literally “from the face/presence of them”} and dispossessed them, and then they settled in their\lebnote{Hebrew “them”} place up to this day.
\verse And also the Avvites, who lived in villages as far as Gaza, and the Caphtorim, who came out from Caphtor, destroyed them and then settled in their place.
\verse Arise,\lebnote{Or “Get up”} set out and cross over the wadi\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season} of Arnon. Look! I have given Sihon the Amorite, the king of Heshbon, and his land into your hand; begin to take possession of it, and engage with him in battle.
\verse This day I will begin to place \textit{the dread of you}\lebnote{Literally “your dread”} and the fear of you \textit{before}\lebnote{Literally “upon the faces of”} the peoples under all the heavens. They will hear \textit{the report about you},\lebnote{Literally “your report”} and so they will shake and tremble \textit{because of you}.’ \lebnote{Literally “from before you”}
\verse “So I sent messengers from the wilderness\lebnote{Or “desert”} of Kedemoth to Sihon king of Heshbon; I sent terms of peace, \textit{saying},\lebnote{Literally “to say”}
\verse ‘Let me cross through your land and \textit{only along the road}\lebnote{Literally “in the road, in the road”} I will go; I will not turn aside to the right or to the left.
\verse Food for money you shall sell me, so that I may eat, and water for money you will give to me, so that I may drink; just let me cross on foot.
\verse Just as the descendants\lebnote{Or “the children/sons of”} of Esau did for me, who live in Seir, and the Moabites, who live in Ar, until I cross the Jordan into the land that Adonai our God is giving to us.’
\verse But Sihon king of Heshbon was not willing to let us cross through his territory because Adonai your God hardened his spirit and \textit{made him obstinate}\lebnote{Literally “made firm his heart”} \textit{in order to give him}\lebnote{Literally “to give him,” indicating purpose} into your hand, \textit{just as he has now done}.\lebnote{Literally “as it is this day”}
\verse Adonai said to me, ‘Look! I have begun to give \textit{over to you}\lebnote{Literally “to the face of you”} Sihon and his land; begin \textit{to take possession of his land}.’\lebnote{Literally, “take possession in order to possess his land”}
\verse Then Sihon and all his people came out to meet us for battle at Jahaz.
\verse And so Adonai our God gave him over \textit{to us},\lebnote{Literally “before us”} and we struck him down, and his sons\lebnote{Or “descendants”} and all of his people.
\verse So we captured all of his cities at that time, and we destroyed each town of males and the women and the children; we did not leave behind a survivor.
\verse We took only the livestock as spoil for ourselves,\lebnote{Hebrew “us”} and also the booty of the cities that we had captured.
\verse From Aroer, which is on the edge of the wadi\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season} of Arnon and the city that was in the wadi\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season} on up to Gilead, there was not a city that was inaccessible to us; Adonai our God gave \textit{everything}\lebnote{Literally “the whole”} \textit{to us}.\lebnote{Literally “before us”}
\verse Only the land of \textit{the Ammonites}\lebnote{Literally “children/sons of Ammon”} you did not approach, all along the whole upper region of the Jabbok \textit{River}\lebnote{Literally “wadi,” which here refers to a flowing river} and the towns of the hill country, according to all that Adonai our God had instructed.
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 3
\verse “Then we turned, and we went up the road to Bashan, and Og the king of Bashan came out to meet us, he and all of his army for the battle at Edrei.
\verse And Adonai said to me, ‘You should not fear him, for I have given him and all of his army\lebnote{Or “people”} and his land into your hand. And so you will do to him as you did to Sihon the king of the Amorites,\lebnote{Hebrew “Amorite”} who was reigning in Heshbon.’
\verse And so Adonai our God also gave Og the king of Bashan, and all of his army\lebnote{Or “people”} into our hand, and we struck him down until not a survivor remained to\lebnote{Or “for”} him.
\verse And we captured all of his towns\lebnote{Or “villages/cities small”} at that time; there was not a city that we did not take from them.
\verse All of these were fortified towns with high walls, gates, and bars,\lebnote{Hebrew “bar”} \textit{apart from}\lebnote{Literally “alone from”} very many of the villages of the open country.
\verse And so we destroyed them just as we had done to Sihon the king of Heshbon; we destroyed utterly each town of males, the women, and the little children.
\verse But all of the livestock and the booty of the towns we kept as spoil for ourselves.\lebnote{Hebrew “us”}
\verse “And so we took at that time the land from \textit{the control of}\lebnote{Literally “the hand of”} the two kings of the Amorites\lebnote{Hebrew “Amorite”} who were \textit{on the other side of the Jordan},\lebnote{Literally “in the beyond of the Jordan”} from the wadi\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season} of Arnon up to \textit{Mount Hermon}.\lebnote{Literally “the mountain of Hermon”}
\verse (The Sidonians called Hermon ‘Sirion,’ and the Amorites called it ‘Senir.’)
\verse All of the towns of the plateau and the whole of Gilead and all of Bashan up to Salecah and Edrei, the towns of the kingdom of Og in Bashan.
\verse (For only Og, king of Bashan, was left from the remnant of the Rephaim. Indeed, his bedstead—it was a bedstead of iron. It is in Rabbah of the \textit{Ammonites}.\lebnote{Literally “sons/children of Ammon”} Nine cubits is its length, and four cubits is its width according to the cubit of a man.)
\verse And so we took possession of this land at that time, from Aroer, which is on the edge of the wadi\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season} of Arnon, and also half of the hill country of Gilead and its towns I gave to the Reubenites\lebnote{Hebrew “Reubenite”} and to the Gadites.\lebnote{Hebrew “Gadite”}
\verse And the remainder of Gilead and all of Bashan, the kingdom of Og, I gave to the half-tribe of Manasseh, the whole region of Argo. All of that area of Bashan was called the land of the Rephaim.
\verse Jair the descendant\lebnote{Or “son”} of Manasseh acquired the whole region of Argob, up to the boundary of the Geshurites and the Maacathites, and he called it,\lebnote{Hebrew “them”} that is Bashan, after his own name, Havvoth Jair, \textit{as it still is today}.\lebnote{Literally “up to this day”}
\verse And also I gave Gilead to Makir.
\verse And to the Reubenites\lebnote{Hebrew “Reubenite”} and to the Gadites\lebnote{Hebrew “Gadite”} I gave, from Gilead up to the wadi\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season} of Arnon, the middle of the wadi\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season} as a boundary and up to the Jabbok \textit{River},\lebnote{Literally “wadi,” which here refers to a flowing river} the boundary of the \textit{Ammonites}.\lebnote{Literally “sons/children of Ammon”}
\verse And the \textit{Jordan Valley}\lebnote{Literally “Arabah”} with the Jordan River as its boundary, from Kinnereth\lebnote{Or “Chinnereth,” the Sea of Galilee} up to the Sea of the Arabah, the Salt Sea,\lebnote{Or the Dead Sea} with the slopes of Pisgah toward the east.
\verse “And I charged you all at that time when I said, “Adonai has given you—to all of you—this land to possess. All the \textit{warriors}\lebnote{Literally “men of valor”} shall cross over, ready to fight, before your brothers, the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse Only your wives and your little children and your livestock (I know that \textit{you have much livestock})\lebnote{Literally “livestock much there is to you”} must stay in your towns that I have given you,
\verse until Adonai shall give rest to your brothers as he did to you, and also they take possession of the land that Adonai your God is giving to them beyond the Jordan; then they may return, each one to his possession that I have given to them.
\verse And I commanded Joshua at that time, saying, ‘Your eyes see all that Adonai your God has done to these two kings; so Adonai will do to all of the kingdoms where you are about to cross over to.
\verse You shall not fear them, for Adonai your God is the one fighting for you.
\verse “And I pleaded with Adonai at that time, saying,
\verse ‘Lord Adonai, you have begun to show your servant your greatness and your strong hand, for what\lebnote{Hebrew “who”} god is there in the heaven or on the earth who can do according to your works and according to your mighty deeds?
\verse Let me cross over, please, and let me see the good land \textit{that is beyond the Jordan},\lebnote{Literally “in the beyond of the Jordan”} this good hill country and Lebanon.’
\verse But Adonai was very angry with me because of you, and he would not listen to me, and Adonai said, ‘\textit{Enough of that from you}!\lebnote{Literally “Much to you”} You shall not speak to me any longer about this matter!
\verse Go up to the top of Pisgah and \textit{look around you}\lebnote{Literally “lift up your eyes”} toward the west, toward the north, and toward the east, and \textit{view}\lebnote{Literally “look”} the land with your eyes, for you will not cross this Jordan.
\verse Now instruct Joshua and support him and encourage him because he himself\lebnote{Emphatic use of the pronoun} will cross over before this people and enable them to inherit the land that you will see.’
\verse So we remained in the valley opposite Beth Peor.
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 4
\verseWithHeading{Introduction to the Stipulations} “Now, Israel, listen to the rules and to the regulations that I am teaching you to do, in order that you may live and you may go in and you may take possession of the land that Adonai, the God of your ancestors,\lebnote{Or “fathers”} is giving to you.
\verse You must not add to the word that I am commanding you, and you shall not take away from it in order to keep\lebnote{Or “observe”} the commands of Adonai your God that I am commanding you to observe.
\verse Your eyes have seen\lebnote{Literally “your eyes the seeing what”} what Adonai did with\lebnote{Or “in”} the case of Baal Peor, for \textit{each}\lebnote{Literally “every one of”} man that followed after Baal Peor Adonai your God destroyed from your midst.
\verse But you, the ones holding fast to Adonai your God, are all alive \textit{today}.\lebnote{Literally “the day”}
\verse See, I now teach\lebnote{Or “I have taught” (compare NASB, NEB)} you rules and regulations just as Adonai my God has commanded me, to observe them just so in the midst of the land where you are going, to take possession of it.
\verse And \textit{you must observe them diligently},\lebnote{Literally “you must observe and you must do”} for that is your wisdom and your insight before the eyes of the people, who will hear all of these rules, and they will say, ‘Surely this great nation is a wise and discerning people.’
\verse For what great nation has for it a god near to it as Adonai our God, whenever\lebnote{Literally “in every matter/every time we call ”} we call upon him?
\verse And what other great nation has for it\lebnote{Literally “which/that is for him it”} just rules and regulations just like\lebnote{Literally “as”} \textit{this whole}\lebnote{Literally “all of”} law that I am setting \textit{before}\lebnote{Literally “to the face of”} you \textit{today}?\lebnote{Literally “the day”}
\verse “However, \textit{take care}\lebnote{Literally “watch”} for yourself and watch your inner self\lebnote{Or “soul”} closely, so that you do not forget the things that your eyes have seen, so that they do not slip from your mind all the days of your life; and you shall make them known to your children and to \textit{your grandchildren}.\lebnote{Literally “the children of your children”}
\verse Remember the day that you stood \textit{before}\lebnote{Literally “to the face of”} Adonai your God at Horeb \textit{when Adonai said to me},\lebnote{Literally “when to say Adonai”} ‘Summon for me the people so that I can tell them my words, that they may learn to fear me all of the days they are alive on the earth and so that they may teach their children.’
\verse And so you came near, and you stood under the mountain, and the mountain was burning with fire up to the heart of the heaven, \textit{dark with a very thick cloud}.\lebnote{Literally “darkness, cloud, and very thick darkness”}
\verse And Adonai spoke to you from the midst of the fire; \textit{you heard}\lebnote{Literally “you were hearing”} a sound of words, but \textit{you did not see}\lebnote{Literally “you were not seeing”} a form—only a voice.
\verse And he declared to you his covenant, \textit{the Ten Commandments},\lebnote{Literally “the ten words”} which he charged you \textit{to observe},\lebnote{Literally “to do”} and he wrote them on the two tablets of stone.
\verse And Adonai charged me at that time to teach you rules and regulations \textit{for your observation of them}\lebnote{Literally “for your doing them”} in the land that you are \textit{about to cross into}\lebnote{Literally “about to cross into there”} to take possession of it.
\verse “So \textit{you must be very careful for yourselves},\lebnote{Literally “watch yourselves diligently with respect to your souls”} because you did not see\lebnote{Or “have not seen”} any form on the day Adonai spoke to you at Horeb from the midst of the fire,
\verse so that you do not \textit{ruin yourselves}\lebnote{Literally “corrupt yourselves”} and make for yourselves a divine image in a form of any image, a replica of male or female,
\verse a replica of any animal that is upon the earth, a replica of any winged bird that flies in the air,
\verse a replica of any creeping thing on the ground, a replica of any fish that is in the water \textit{below}\lebnote{Literally “under”} the earth.
\verse \textit{And do this so that you do not lift}\lebnote{Literally “And lest you lift up”} your eyes toward heaven and \textit{observe}\lebnote{Literally “see”} the sun and the moon and the stars, all the host of the heaven, and be led astray and bow down to them and serve them, things that Adonai your God has allotted to all of the peoples under all of the heaven.
\verse But Adonai has taken you and brought you out from the furnace of iron, from Egypt, to be a people of inheritance to\lebnote{Or “for”; a people of inheritance = a people who is his very own possession} him, \textit{as it is this day}.\lebnote{Literally “as the day the this” = as you are now}
\verse “And Adonai was angry with me \textit{because of you},\lebnote{Literally “because of your matter”} and he swore \textit{that I would not cross the Jordan}\lebnote{Literally “not I to cross the Jordan”} and \textit{that I would not go to the good land}\lebnote{Literally “not I going to the land”} that Adonai your God is giving you as an inheritance.
\verse For I am going to die in this land; I am not going to cross the Jordan, but you are going to cross, and you are going to take possession of this good land.
\verse Watch out for yourselves so that you do not forget the covenant of Adonai your God that he had \textit{made}\lebnote{Literally “cut”} with you and make for yourselves a divine image of the form of anything that Adonai your God \textit{has forbidden},\lebnote{Literally “has commanded you about”}
\verse for Adonai your God is a devouring fire, a jealous\lebnote{Or “zealous” or “impassioned”} God.
\verse “When you have had children and \textit{grandchildren}\lebnote{Literally “children of children”} and you have grown old in the land and you act corruptly and you make a divine image of the form of anything and you do evil in the eyes of Adonai your God, thus provoking him to anger,
\verse I call\lebnote{Or “I shall call to witness”} to witness against you today the heaven and the earth, that you will perish soon and completely from the land that you are crossing the Jordan into it to take possession of it; \textit{you will not live long on it},\lebnote{Literally “you will not extend days”} but you will be completely destroyed.
\verse And Adonai will scatter you among the peoples, and you will be left \textit{few in number}\lebnote{Literally “people of number,” as opposed to people without number} among the nations \textit{to where Adonai will lead you}.\lebnote{Literally “he will lead Adonai you there”}
\verse And you will there serve gods \textit{made by human hands},\lebnote{Literally “the work of the hands of human”} of wood and stone, gods that cannot see and cannot hear and cannot eat and cannot smell.
\verse But from there you shall seek Adonai your God and will find him, if you seek him with all your heart and with all your soul.\lebnote{Or “inner self”}
\verse \textit{In your distress}\lebnote{Literally “In the distress for you”} when\lebnote{Hebrew “and”} all these things have found you in the \textit{latter days},\lebnote{Literally “in the last of the days”} then\lebnote{Hebrew “and”} you will return to Adonai your God, and you will listen to his voice.
\verse For Adonai your God is a compassionate God; he \textit{will not abandon you},\lebnote{Literally “will not leave you alone”} and he will not destroy you, and he will not forget the covenant of your ancestors\lebnote{Or “fathers”} that he swore to them.
\verse “Yes, ask, please, about former days that \textit{preceded you}\lebnote{Literally “that they were to the face of you”} from the day that God created humankind on the earth; ask even from one end of the heaven up to the other end of heaven \textit{whether anything ever happened}\lebnote{Literally “was it ever”} like this great thing or \textit{whether anything like it was ever heard}.\lebnote{Literally “was it ever heard as it”}
\verse Has a people ever heard the voice of God speaking from the midst of the fire, just as you heard it, and lived?
\verse Or has a god ever attempted to go to take for himself\lebnote{Hebrew “for him”} a nation from the midst of a nation, using trials and signs and wonders and war, with an outstretched arm and with great and awesome deeds, like all that Adonai your God did for you in Egypt before your eyes?
\verse You yourselves\lebnote{Emphatic use of pronoun; plural meaning implied} were shown this wonder in order for you to acknowledge that Adonai is the God;\lebnote{The definite article indicates that Israel’s God is alone the true God and the one who revealed himself to them} there is no other God \textit{besides him}.\lebnote{Literally “except him” or “to him alone”}
\verse From heaven he made you hear his voice to teach you, and on the earth he showed you his great fire, and you heard his words from the midst of the fire.
\verse And because he loved your ancestors\lebnote{Or “fathers”} he chose their \textit{descendants}\lebnote{Literally “seed”} after them. And he brought you forth from Egypt \textit{with his own presence},\lebnote{Literally “with his faces”} by his great strength,
\verse to drive out nations greater and more numerous than you \textit{from before you},\lebnote{Literally “from your face”} to bring you and to give to you their land as an inheritance, as it is this day.
\verse So you shall acknowledge \textit{today},\lebnote{Literally “the day”} and \textit{you must call to mind}\lebnote{Literally “you shall bring back to your heart”} that Adonai is God in heaven above and on the earth beneath. There is no other God.
\verse And you shall keep his rules and his commandments that I am commanding you \textit{today},\lebnote{Literally “the day”} so that \textit{it may go well}\lebnote{Literally “he/it is good”} for you and for your children\lebnote{Or “descendants”} after you, and so that \textit{you may remain a long time}\lebnote{Literally “you may make long/prolong days”} on the land that Adonai your God is giving to you during all of those days.”
\verse Then Moses set apart three cities \textit{on the other side of the Jordan},\lebnote{Literally “in the beyond of the Jordan”} \textit{toward the east},\lebnote{Literally “toward rising of the sun”}
\verse in order for \textit{a manslayer}\lebnote{Literally “a killer of a man”} to flee there who has killed his neighbor \textit{without intent}\lebnote{Literally “without previous knowledge”} and was not hating him \textit{previously},\lebnote{Literally “the day before yesterday”} and so he could flee to one of these cities \textit{and be safe}.\lebnote{Literally “and live”}
\verse He set apart Bezer in the wilderness\lebnote{Or “desert”} in the land of the plateau of the Reubenites;\lebnote{Hebrew “Reubenite”} Ramoth in Gilead of the Gadites,\lebnote{Hebrew “Gadite”} and Golan in Bashan of the Manassites.\lebnote{Hebrew “Manassite”}
\verse Now this is the law\lebnote{Hebrew “the \textit{torah}” = teaching, instruction, law} that Moses set \textit{before}\lebnote{Literally “before the faces of”} the \textit{Israelites};\lebnote{Literally “sons/children of Israel”}
\verse these are the legal provisions and the rules and the regulations that Moses spoke to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} \textit{when they left Egypt},\lebnote{Literally “at their going out from Egypt”}
\verse beyond the Jordan in the valley opposite Beth Peor in the land of Sihon the king of the Amorites,\lebnote{Hebrew “Amorite”} who was reigning in Heshbon and whom Moses and the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} defeated \textit{when they came out of Egypt}.\lebnote{Literally “at their going out from Egypt”}
\verse And so they took possession of his land and the land of Og king of Bashan, the two kings of the Amorites\lebnote{Hebrew “Amorite”} who were beyond the Jordan, \textit{eastward},\lebnote{Literally “toward the rising of the sun”}
\verse from Aroer, which is on the bank of the wadi\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season} of Arnon and as far as Mount Sirion; that is, Hermon,
\verse and all of the Arabah\lebnote{Or “valley” in this instance} beyond the Jordan, eastward, and as far as the Sea of the Arabah\lebnote{Commonly known today as the Dead Sea} under the slopes of Pisgah.
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 5
\verseWithHeading{Basic Stipulations} And then Moses summoned all of Israel and said to them, “Hear, Israel, the rules and the regulations that I am speaking in your ears \textit{today},\lebnote{Literally “the day”} and you shall learn them, and \textit{you must observe them diligently}.\lebnote{Literally “you shall observe them carefully to do them”}
\verse Adonai our God made a covenant with us at Horeb.
\verse It was not with our ancestors\lebnote{Or “fathers”} that Adonai made this covenant, but with these of us who are here alive today.
\verse \textit{Face to face}\lebnote{Literally “Faces to faces”} Adonai spoke with you at the mountain from the midst of the fire.
\verse I was standing \textit{between Adonai and you}\lebnote{Literally “between Adonai and between you”} at that time to report\lebnote{Or “declare”} to you the word of Adonai, for you were afraid because of \textit{the presence of}\lebnote{Literally “the faces of”} the fire, and so you did not go up the mountain. \textit{He said},\lebnote{Literally “To say”}
\verse ‘I am Adonai your God, who brought you out from the land of Egypt, from the house of slavery.
\verse There shall not be for you other gods \textit{besides me}.\lebnote{Literally “before my face”}
\verse ‘You shall not make for yourself a divine image of any type of form that is in the heaven above or that is on the earth beneath or that is in the water under the earth.
\verse ‘You shall not bow down to them, and you shall not serve them, for I, Adonai your God, am a jealous God, punishing the guilt of fathers upon their children and upon the third and upon the fourth generation of those hating me,
\verse but showing loyal love to thousands of those who love me and of those who keep my commandments.
\verse ‘You shall not take up the name of Adonai your God for a worthless purpose, for Adonai will not leave unpunished anyone who uses his name for a worthless purpose.
\verse ‘Observe the \textit{Sabbath day}\lebnote{Literally “the day of the Sabbath”} to make it holy,\lebnote{Or “to sanctify it”} just as Adonai your God has commanded you.
\verse Six days you shall work, and you shall do all of your work,
\verse but the seventh day is a Sabbath unto Adonai your God; you shall not do any work, or your son, or your daughter, or your slave, or your slave woman, or your ox, or your donkey, or any of your domestic animals, or your resident alien who is in your \textit{towns},\lebnote{Literally “gates”} so that your slave and your slave woman may rest as you rest.
\verse And you shall remember that you were a slave in the land of Egypt, and Adonai your God brought you out with a strong hand and with an outstretched arm; therefore, Adonai your God commanded you to keep \textit{the Sabbath}.\lebnote{Literally “the day of the Sabbath”}
\verse ‘Honor your father and your mother, as Adonai your God commanded you, so that it will be good for you\lebnote{Or “it may go well for you”} \textit{in the land}\lebnote{Literally “on the ground”} that Adonai your God is giving to you.
\verse ‘You shall not murder.
\verse ‘And you shall not commit adultery.
\verse ‘And you shall not steal.
\verse ‘And you shall not falsely bear evidence against your neighbor.
\verse ‘And you shall not covet the wife of your neighbor, and you shall not crave the house of your neighbor, his field or his slave or his slave woman or his ox and his donkey or anything \textit{that belongs to your neighbor}.’\lebnote{Literally “that is for your neighbor”}
\verse “These words Adonai spoke to your whole assembly at the mountain from the midst of the fire and the very thick cloud with a loud voice, and \textit{he did not add anything},\lebnote{Literally “and not he added”} and then he wrote them on two tablets of stone and gave them to me.
\verse \textit{And then}\lebnote{Literally “And it happened”} \textit{when you heard}\lebnote{Literally “and as/when you heard” or “at the moment of your hearing”} the voice from the midst of the darkness, and as the mountain was burning with fire, and and all the heads of your tribes and your elders approached me,
\verse you said, ‘Look, Adonai our God has shown us his glory and his greatness, and we have heard his voice from the midst of the fire; this day we have seen\lebnote{Or “saw”} that God can speak with a human being, \textit{yet he remains alive}.\lebnote{Literally “and he lives”}
\verse And so then why shall we die, for this great fire will consume us if \textit{we continue}\lebnote{Literally “we do again to hear”; or “we are doing again to hear”} to hear the voice of Adonai our God any longer, and so we shall die?
\verse For who is there of all flesh who has heard the voice of the living God speaking from the midst of the fire as we have heard it \textit{and remained alive}?\lebnote{Literally “and lives”}
\verse You go near and hear \textit{everything}\lebnote{Literally “all of that which”} that Adonai our God will say; and then you tell us all that Adonai our God tells you, and we will listen, and we will do it.’
\verse “And Adonai heard the sound of your words \textit{when you spoke to me},\lebnote{Literally “in/at you to speak to me”} and Adonai said to me, ‘I have heard the sound of the words of this people that they have spoken to you; they are right with respect to all that they have spoken.
\verse \textit{If only}\lebnote{Literally “Who gives/will give”} \textit{they had such a mind}’;\lebnote{Literally “it would be their heart this to them”} that is, to fear me and to keep all my commandments \textit{at all times},\lebnote{Literally “all the days”} so that \textit{it will go well}\lebnote{Literally “he/it is good”} for them and for their children \textit{forever}.\lebnote{Literally “to eternity”}
\verse Go! Say to them, “Return to your tents.”
\verse But you stand here with me, and let me tell you all of the commandments and the rules and the regulations that you shall teach them, so that they may do them in the land that I am giving to them to take possession of it.’
\verse “So you must be careful to do just as Adonai your God commanded you; you shall not turn to the right or to the left.
\verse \textit{In exactly the path}\lebnote{Literally “In all the way/path”} that Adonai your God has commanded, you must go, so that you may live and \textit{it will go well}\lebnote{Literally “and good it is”} for you and \textit{you may live long}\lebnote{Literally “you may make long”} in the land that you will take possession of.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 6
\verseWithHeading{Detailed Stipulations} “Now this is the commandment, the rules and the regulations, that Adonai your God charged to teach to you for you \textit{to observe}\lebnote{Literally “to do”} in the land that you are about to cross over into\lebnote{Or “there”} to take possession of it,
\verse so that you may revere Adonai your God by keeping all his statutes and his commandments that I am commanding you, you and your children and \textit{grandchildren},\lebnote{Literally “the children of your children”} all the days of your life and so you may live long\lebnote{Literally “they may be long your days”} lives.
\verse And you shall hear, Israel, and be careful to observe these instructions, so that \textit{it may go well for you}\lebnote{Literally “it/he may be good”} and that you may multiply greatly, just as Adonai, the God of your ancestors,\lebnote{Or “fathers”} \textit{promised}\lebnote{Literally “spoke”} you, in a land with milk and honey.
\verse “Hear, Israel, Adonai our God, Adonai is unique.\lebnote{Or “one”; or possibly “one Lord”}
\verse And you shall love Adonai your God with all of your heart\lebnote{Or “mind”} and with all of your soul\lebnote{Or “inner self”} and with all of your might.
\verse And these words that I am commanding you \textit{today}\lebnote{Literally “the day”} shall be on\lebnote{Or “in”} your heart.\lebnote{Or “mind”}
\verse And you shall recite them to your children, and you shall talk about them at the time of your living in your house and at the time of your going on the road and at the time of your lying down and at the time of your rising up.
\verse And you shall bind them as a sign on your hand, and they shall be as an emblem between your eyes.
\verse And you shall write them on the doorframe of your house and on your gates.
\verse “And then it will happen that when Adonai your God will bring you to the land that he swore to your ancestors,\lebnote{Or “fathers”} to Abraham, to Isaac, and to Jacob, to give to you large and fine cities that you did not build,
\verse and houses full of all sorts of good things\lebnote{Hebrew “thing”} that you did not fill, and hewn cisterns that you did not hew, vineyards and olive groves that you did not plant, and \textit{you have eaten your fill},\lebnote{Literally “you have eaten and you are satisfied”}
\verse then take care for yourself, so that you do not forget Adonai, who brought you out from the land of Egypt from the house of slavery.
\verse “You shall fear Adonai your God, and you shall serve him, and by his name you shall swear.
\verse You shall not go after other gods from the gods of the peoples who are all around you,
\verse for Adonai your God is a jealous god in your midst, \textit{so that the anger of Adonai your God would be kindled},\lebnote{Literally “so that not/lest the nose of Adonai your God would become hot”} and he would destroy you from the face of the earth.
\verse You shall not put Adonai your God to the test, as you tested him at Massah.
\verse You shall diligently keep the commandments of Adonai your God and his legal provisions and his rules that he has commanded you.
\verse And you shall do what is right and good in the eyes of Adonai, so that \textit{it shall go well}\lebnote{Literally “he/it shall be good”} for you and so that you may go and you may take possession of the good land that Adonai swore for your ancestors,\lebnote{Or “your fathers”}
\verse by driving out all of your enemies \textit{before you},\lebnote{Literally “the face of you”} just as Adonai \textit{has promised}.\lebnote{Literally “had spoken”}
\verse “When your child\lebnote{Or “son”} asks you in the future, \textit{saying},\lebnote{Literally “to say”} ‘What is the meaning of the legal provisions and the rules and the regulations that Adonai our God commanded for you?’
\verse Then you shall say to your child,\lebnote{Or “son”} ‘We were slaves of Pharaoh in Egypt, and Adonai brought us out from Egypt with a strong hand.
\verse And Adonai gave great and awesome signs and wonders in Egypt against Pharaoh and against his entire household \textit{in our presence}.\lebnote{Literally “before our eyes”}
\verse But he brought us out from there in order to bring us here to give us the land that he swore to our ancestors.\lebnote{Or “fathers”}
\verse And so Adonai commanded us to observe all of these rules and to revere Adonai our God \textit{for our benefit}\lebnote{Literally “for good for us”} \textit{all the days that we live},\lebnote{Literally “all the days for our living”} \textit{as it is today}.\lebnote{Literally “as the day the this”}
\verse And it shall be righteousness for us if we diligently observe and do all of this commandment \textit{before}\lebnote{Literally “in the faces of”} Adonai our God, as he has commanded us.’
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 7
\verse “When Adonai your God brings you into the land that you are about to enter \textit{into it}\lebnote{Literally “there”} to take possession of it, and he drives\lebnote{Or “he will drive out”} out many nations \textit{before you},\lebnote{Literally “from your faces”} the Hittites\lebnote{Hebrew “Hittite”} and the Girgashites\lebnote{Hebrew “Girgashite”} and the Amorites\lebnote{Hebrew “Amorite”} and the Canaanites\lebnote{Hebrew “Canaanite”} and the Hivites\lebnote{Hebrew “Hivite”} and the Jebusites,\lebnote{Hebrew “Jebusite”} seven nations mightier and more numerous than you,
\verse and Adonai your God will give them \textit{over to you}\lebnote{Literally “to your faces” or “before you”} and you defeat them, you must \textit{utterly destroy them};\lebnote{Literally “set them aside for destruction”} you shall not make a covenant with them, and you shall not show mercy to them.
\verse And you shall not intermarry with them; you shall not give your daughter to their son; and you shall not take his daughter for your son.\lebnote{Grammatically singular pronominal suffixes here can also be rendered “their”}
\verse For their\lebnote{Or “his”} sons and daughters will cause your son\lebnote{Or “sons”} to turn away \textit{from following me},\lebnote{Literally from “behind/after me”} and so they will serve other gods, and \textit{the anger of Adonai would be kindled}\lebnote{Literally “and it would become hot the nose of Adonai”} against you, and he would quickly destroy you.
\verse But this is what you must do to them: you shall break down their altars, and their stone pillars you shall smash, and their Asherah poles you shall hew down, and you shall burn their idols with fire.
\verse For you are a holy people for Adonai your God; Adonai your God has chosen you to be for him a people, a treasured possession from among all the peoples that are on the face\lebnote{Or “surface”} of the earth.
\verse “Adonai loved you and chose you not \textit{because of your great number}\lebnote{Literally “from your multitude/abundance”} exceeding all other peoples, for you are fewer than all of the peoples,
\verse but\lebnote{Or “for”} \textit{because of}\lebnote{Literally “from”} the love of Adonai for you and because of his keeping of the sworn oath that he swore to your ancestors,\lebnote{Or “fathers”} Adonai brought you out with a strong hand and redeemed you from the house of slavery, from the hand of Pharaoh, the king of Egypt.
\verse So know that Adonai your God, he is God, the trustworthy God, maintaining his\lebnote{Hebrew “the”} covenant and his\lebnote{Hebrew “the”} loyal love with those who love him and with those who keep his commandments to a thousand generations,\lebnote{Hebrew “generation”}
\verse but repaying those\lebnote{Hebrew “the one”} who hate him \textit{in their own person}\lebnote{Literally “to their faces”} to destroy them;\lebnote{Hebrew “him”; “them” in sense} he is not slow with those who hate him \textit{in their own person};\lebnote{Literally “to their faces”} he repays them.\lebnote{Hebrew “him”}
\verse And so you shall keep the commandment and the rules and the regulations that I am commanding you \textit{today}\lebnote{Literally “the day”} to observe them.
\verse “\textit{And then}\lebnote{Literally “And it will happen”} because you listen to these regulations and you diligently keep and you do them, then Adonai your God will maintain his\lebnote{Hebrew “the”} covenant and his\lebnote{Or “fathers”} loyal love that he swore to your ancestors.\lebnote{Or “fathers”}
\verse And he will love you, and he will bless you, and he will multiply you, and he will bless the fruit of your womb and the fruit of your soil, your grain, your wine, and your olive oil, and newborn calves of your cattle, and the newborn lambs\lebnote{Or “increase of”} of your flocks in the land that he swore to your ancestors\lebnote{Or “fathers”} to give you.
\verse You shall be blessed more than all of the peoples; among you there shall not be sterility and bareness, even\lebnote{Hebrew “and”} among domestic animals.\lebnote{Hebrew “animal”}
\verse And Adonai will turn away from you all the illness and all the harmful diseases of Egypt that you experienced; he will not lay them on you, but he will lay them on all of those who hate you.
\verse And you shall devour all of the peoples that Adonai your God is giving to you; \textit{you shall not pity them},\lebnote{Literally “your eye shall not take pity on them”} and you shall not serve their gods, which will be a snare for you.
\verse “If you think in your heart, ‘These nations are more numerous than I, so how can I dispossess them?’
\verse then remember you must not be afraid of them; you must well remember what Adonai your God did to Pharaoh and to all of Egypt:
\verse the great trials that your eyes saw and the signs and the wonders and the workings of the strong hand and the outstretched arm by which Adonai your God brought you out; so Adonai your God will do to all of the peoples \textit{because of whom}\lebnote{Literally “from them”} you are in fear \textit{before them}.\lebnote{Literally “from their presence”}
\verse And, moreover, Adonai your God will send the hornets\lebnote{Hebrew “hornet”} among them until both the survivors and the fugitives \textit{are destroyed}\lebnote{Literally “the destroying of”} \textit{before you}.\lebnote{Literally “from your faces”}
\verse You must not be in dread from the presence of them, because Adonai your God, who is in your midst, is a great and awesome God.
\verse And Adonai your God will clear away these nations \textit{from before you}\lebnote{Literally “from your faces”} little by little; you will not be able to finish them off quickly, \textit{lest}\lebnote{Literally “so that not”} the \textit{wild animals}\lebnote{Literally “animals of the field”} \textit{multiply}\lebnote{Literally “will become numerous”} \textit{against you}.\lebnote{Literally “on you”}
\verse But Adonai your God will \textit{give them to you},\lebnote{Literally “give them to your face”} and he will throw them into great panic \textit{until they are destroyed}.\lebnote{Literally “until their destroying”}
\verse And he will give their kings into your hand, and you shall blot out their names\lebnote{Hebrew “name”} from under the heaven; anyone will not be able to stand \textit{against you}\lebnote{Literally “in your faces”} \textit{until you destroy them}.\lebnote{Literally “until your destroying them”}
\verse You shall burn the images of their gods with fire; you shall not covet the silver or gold that is on them, and so you take it for yourself, so that you are not ensnared by it, for it is a detestable thing to Adonai your God.
\verse And you must not bring a detestable thing into your house, or you will become a thing devoted to destruction like it; you must utterly detest it, and you must utterly abhor it, for it is an object devoted to destruction.
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 8
\verse “All of the commandments\lebnote{Or “every commandment”} that I am commanding you today you must diligently observe, so that you may live and multiply, and you may go and take possession of the land that Adonai swore to your ancestors.\lebnote{Or “your fathers”}
\verse And you shall remember all of the way that Adonai your God led you during these forty years in the desert in order to humble you and to test you to know what is in your heart, whether you would diligently keep his commandments or not.
\verse And he humbled you and let you go hungry, and then he fed you with that which you did not know nor did your ancestors\lebnote{Or “fathers”} know, in order to make you know that not by bread alone but by all that goes out\lebnote{Or “by all of the going out of”} of the mouth of Adonai humankind shall live.
\verse Your clothing did not wear out \textit{on you},\lebnote{Literally “from on you”} and your feet did not swell during these forty years.
\verse And you should know with your heart that as a man disciplines his son, so Adonai your God is disciplining you.\lebnote{Or “has disciplined you”}
\verse So you must keep the commandments of Adonai your God by walking in his ways and by fearing him.
\verse For Adonai your God is bringing you to a good land with streams of water, springs and underground water, welling up in the valleys\lebnote{Hebrew “valley”} and in the hills,\lebnote{Hebrew “hill”}
\verse to a land of wheat and barley and vines\lebnote{Hebrew “vine”} and fig trees\lebnote{Hebrew “fig tree”} and pomegranate trees,\lebnote{Hebrew “pomegranate tree”} a land of olive trees,\lebnote{Hebrew “olive tree”} olive oil and honey;
\verse to a land where you may eat food in it \textit{without scarcity};\lebnote{Literally “not in scarcity”} you will not find anything lacking in it, a land where its stones are iron and from its mountains you can mine copper.
\verse And you will eat, and \textit{you will be satisfied},\lebnote{Literally “you will eat your fill”} and you will bless Adonai your God because of the good land that he has given to you.
\verse “Take care for yourself so that you not forget Adonai your God by not keeping his commandments and his regulations and his statutes that I am commanding you \textit{today},\lebnote{Literally “the day”}
\verse lest when you have eaten and you are satisfied and you have built good houses and you live in them,
\verse and your herds and your flocks have multiplied, and \textit{you have accumulated silver and gold},\lebnote{Literally “and silver and gold has multiplied for you”} and all that \textit{you have}\lebnote{Literally “is for you”} has multiplied,
\verse then your heart \textit{becomes proud}\lebnote{Literally “raises up”} and you forget Adonai your God, \textit{the one who brought you out}\lebnote{Literally “the one bringing you out”} from the land of Egypt, from the house of slavery,
\verse the one leading you in the great and terrible desert infested with \textit{dangerous}\lebnote{Literally “burning”} snakes\lebnote{Hebrew “snake”} and scorpions\lebnote{Hebrew “scorpion”} and parched ground, where there is\lebnote{Or “was”} no water, and the one bringing out water for you from flint rock,
\verse the one \textit{feeding you}\lebnote{Literally “causing you to eat”} manna in the desert, food that your ancestors\lebnote{Or “fathers”} did not know, in order to humble you and in order to test you so that he could do good to you \textit{in the future}.\lebnote{Literally “in the end/later”}
\verse And you may think\lebnote{Or “say”} in your heart, ‘My strength and the might of my hand \textit{acquired this wealth for}\lebnote{Literally “gave this wealth to”} me.’
\verse But you must remember Adonai your God, for he is the one giving you strength to acquire wealth in order to confirm his covenant that he swore to your ancestors\lebnote{Or “fathers”} \textit{as it is today}.\lebnote{Literally “as the day the this”}
\verse And it will happen that if you indeed forget Adonai your God and you go after other gods and you serve them and you bow down before them, I warn you today that you will surely perish.
\verse As with the nations that Adonai is destroying \textit{before you},\lebnote{Literally “from your faces”} so you shall perish because you would not obey the voice of Adonai your God.
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 9
\verse “Hear, Israel, you are about to cross the Jordan today to go to dispossess nations larger and more numerous than you, great cities fortified \textit{with high walls},\lebnote{Literally “in heaven/the sky”}
\verse a great and tall people, the sons of the Anakites, whom you know and of whom you have heard it said, ‘Who could stand before the sons of Anak?’
\verse You should know \textit{today}\lebnote{Literally “the day”} that Adonai your God is the one crossing \textit{ahead of you}\lebnote{Literally “to your faces”} as a devouring fire; he will destroy them, and he will subdue them before you; so you will dispossess them, and you will destroy them quickly, \textit{just as}\lebnote{Literally “as that”} Adonai \textit{promised}\lebnote{Literally “spoke”} you.
\verse “You shall not say \textit{to yourself}\lebnote{Literally “in your heart”} \textit{when Adonai your God is driving them out}\lebnote{Literally “at the driving out of Adonai, your God, them”} \textit{before you},\lebnote{Literally “from before your faces”} \textit{saying},\lebnote{Literally “to say”} ‘Because of my righteousness Adonai brought me to take possession of this land’; but because of the wickedness of these nations Adonai is driving them out \textit{before you}.\lebnote{Literally “from your faces”}
\verse It is not because of your righteousness and because of the uprightness of your heart that you are coming to take possession of their land, but because of the wickedness of these nations Adonai your God is driving them \textit{before you},\lebnote{Literally “from your faces”} and in order to confirm the \textit{promise}\lebnote{Literally “word”} that Adonai swore to your ancestors,\lebnote{Or “fathers”} to Abraham, to Isaac, and to Jacob.
\verse “So you should understand that it is not because of your righteousness that Adonai your God is giving you this good land to take possession of it, because \textit{you are a stubborn people}. \lebnote{Literally “a people stiff of neck you”}
\verse Remember, \textit{do not forget},\lebnote{Literally “not you shall forget”} that you provoked Adonai your God in the desert, and from the day that you went out from the land of Egypt until \textit{you came to this place}\lebnote{Literally “your coming up to this place the this” = until arrived at this location} you were rebelling against Adonai.
\verse “And remember at Horeb you provoked Adonai, and Adonai became angry enough to destroy you.
\verse \textit{When I went up the mountain}\lebnote{Literally “In/At my going up the mountain”} to receive \textit{the stone tablets},\lebnote{Literally “the tablets of stones”} the tablets of the covenant that Adonai \textit{made}\lebnote{Literally “cut”} with you, and remained on the mountain forty days\lebnote{Hebrew “day”} and forty nights,\lebnote{Hebrew “night”} I did not eat food and I did not drink water.
\verse And Adonai gave me the two tablets of stone written with the finger of God, and on them was writing according to all the words that Adonai spoke with you at the mountain, from the midst of the fire on the day of the assembly.
\verse \textit{And then}\lebnote{Literally “And it happened”} at the end of forty days\lebnote{Hebrew “day”} and forty nights,\lebnote{Hebrew “night”} Adonai gave me the two tablets of stone, the tablets of the covenant.
\verse And Adonai said to me, ‘Come now, go down quickly from this mountain because your people behave corruptly whom you brought out from Egypt, for they turned quickly from the way that I commanded them to follow; they have made for themselves a cast image.’
\verse And Adonai spoke to me, \textit{saying},\lebnote{Literally “to say”} ‘I have seen this people, and look! \textit{They are a stubborn people}.\lebnote{Literally “It is a people stiff of neck”}
\verse Leave me alone, and let me destroy them, and let me blot out their name from under heaven, and let me make you into a nation mightier and more numerous than they!’
\verse “And I turned, and I went down the mountain, as the mountain was burning with fire, and the two tablets of the covenant were in my two hands.
\verse And I looked,\lebnote{Or “saw”} and indeed you had sinned against Adonai your God; you had made for yourselves an image of a calf of cast metal; \lebnote{Or “molten calf”} you had turned quickly from the way that Adonai had commanded for you.
\verse And I took hold of the two tablets, and I threw them out \textit{of}\lebnote{Literally “from”} my two hands and smashed them before your eyes.
\verse And then I lay prostrate \textit{before}\lebnote{Literally “to the face of”} Adonai, as earlier, forty days\lebnote{Hebrew “day”} and forty nights;\lebnote{Hebrew “night”} I did not eat food and I did not drink water because of all your sins that you committed, by doing evil in the eyes of Adonai and so provoking him.
\verse For \textit{I was in dread}\lebnote{Literally “I dreaded”} from being in the presence of the anger and the wrath with which Adonai was angry with you so as to destroy you, but Adonai listened to me also \textit{at that time}.\lebnote{Literally “the occurrence the that”}
\verse And with Aaron Adonai was \textit{angry enough}\lebnote{Literally “very angry”} to destroy him, and I prayed also for Aaron at that time.
\verse And your sinful thing that you had made, the molten calf, I took and I burned it with fire, and I crushed it, grinding it thoroughly until it was crushed to dust, and I threw its dust into \textit{the stream that flowed down the mountain}.\lebnote{Literally “the stream the one going down from the mountain”}
\verse “And also at Taberah and at Massah and at Kibroth Hattaavah you provoked Adonai to anger.
\verse And when Adonai sent you out from Kadesh Barnea, \textit{saying},\lebnote{Literally “to say”} ‘Go up and take possession of the land that I have given you,’ you rebelled against the command of Adonai your God, and you did not believe him, and you did not listen to his voice.
\verse You have been rebellious toward Adonai \textit{from the day I have known you}.\lebnote{Literally “from the day of my knowing you”}
\verse “And I lay prostrate before Adonai through forty days,\lebnote{Hebrew “day”} and through forty nights\lebnote{Hebrew “night”} I prostrated myself, because Adonai intended to kill you.
\verse And I prayed to Adonai, and I said, ‘Lord Adonai, you must not destroy your people and your inheritance whom you redeemed in your greatness,\lebnote{Or “your great act”} whom you brought out from Egypt with a strong hand.
\verse Remember your servants, Abraham, Isaac, and Jacob; you must not \textit{pay attention to}\lebnote{Literally “turn toward”} the stubbornness of this people, to their wickedness and to their sin,
\verse lest the people of the land from which you brought us out from there say, “Because Adonai was not able to bring them to the land that he \textit{promised}\lebnote{Literally “spoken”} to them and because of his hatred toward them, he has brought them out to kill them in the desert.”
\verse For they are your people and your inheritance whom you brought with your great power and with your outstretched arm.’
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 10
\verse “At that time Adonai said to me, ‘Carve for yourself two tablets of stone just as the former ones, and come up the mountain to me, and you shall make for yourself an ark of wood.
\verse And I will write on the tablets the words that were on the former tablets, which you smashed, and you must put them in the ark.’
\verse And so I made an ark \textit{of acacia wood},\lebnote{Literally “of wood of acacia trees”} and I carved two tablets of stone like the former ones, and I went up the mountain with the two tablets in my hand.
\verse And he wrote upon the tablets \textit{according to the first writing},\lebnote{Literally “as the writing the first”} the ten words that Adonai spoke to you on the mountain from the midst of the fire on the day of the assembly, and Adonai gave them to me.
\verse And I turned, and I came down from the mountain, and I put the tablets in the ark that I had made, and they are there, just as Adonai commanded me.
\verse “And the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} journeyed from the wells of Bene-Yaqan to Moserah; there Aaron died and was buried, and Eleazar, his son, served as a priest in place of him.
\verse From there they journeyed to Gudgodah, and from Gudgodah to Jotbathah, a land flowing with streams of water.
\verse At that time Adonai set apart the tribe of Levi to carry the ark of the covenant of Adonai to stand \textit{before}\lebnote{Literally “the faces of”} Adonai, to serve him and to bless the people in his name until this day.
\verse Therefore \textit{there was not}\lebnote{Literally “it/he was not”} for Levi an allotment or an inheritance along with his brothers; rather Adonai is his inheritance just as Adonai your God \textit{promised}\lebnote{Literally “spoke”} to him.
\verse And I stayed on the mountain just as during the former forty days\lebnote{Hebrew “day”} and forty nights,\lebnote{Hebrew “night”} and Adonai listened to me also on that occasion;\lebnote{Or “occurrence” or “that time”} Adonai was not willing to destroy you.
\verse And Adonai said to me, ‘\textit{Come, continue}\lebnote{Literally “Arise, go”} your journey \textit{before the people}, \lebnote{Literally “in the faces of the people”} so that you may go and take possession of the land that I swore to their ancestors\lebnote{Or “fathers”} to give to them.’
\verse And now, Israel, what is Adonai your God asking\lebnote{Or “requiring”} from you, except\lebnote{Or “but only”} to revere Adonai your God, to go\lebnote{Or “to walk”} in all his ways and to love him and to serve Adonai your God with all your heart and with all your soul,\lebnote{Or “inner self”}
\verse to keep the commandments of Adonai and his statutes that I am commanding you today \textit{for your own good}.\lebnote{Literally “for good to you”}
\verse Look! For to Adonai your God \textit{belong}\lebnote{Literally “is”} heaven and the \textit{highest heavens},\lebnote{Literally “heavens of the heavens”} the earth and all that is in it.
\verse \textit{Yet}\lebnote{Literally “Only”} to your ancestors\lebnote{Or “fathers”} Adonai was very attached, so as to love them, and so he chose their offspring after them, namely you, from all the peoples, as it is \textit{today}.\lebnote{Literally “the day the this”}
\verse So you shall circumcise the foreskin of your heart, and \textit{do not be stubborn}.\lebnote{Literally “your neck you shall not make stiff any longer”}
\verse For Adonai your God, he is God of the gods and Lord of the lords, the great and mighty God, the awesome one who \textit{is not partial},\lebnote{Literally “does not lift up faces”} and he does not take bribes.
\verse And he executes justice for the orphan and widow, and he is one who loves the alien, to give to them food and clothing.
\verse And you shall love the alien, for you were aliens in the land of Egypt.
\verse Adonai your God, you shall revere him, you shall serve him, and to him you shall cling, and by his name you shall swear.
\verse He is your praise, and he is your God, who has done with you these great and awesome things that your eyes have seen.
\verse With only seventy persons your ancestors\lebnote{Or “fathers”} went down to Egypt, but now Adonai your God has made you as the stars of the heaven \textit{with respect to multitude}.\lebnote{Literally “as far as number”}
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 11
\verse “And you shall love Adonai your God, and you shall keep his obligations and his statutes and his regulations and his commandments \textit{always}.\lebnote{Literally “all of the days”}
\verse And you shall realize \textit{today}\lebnote{Literally “the day”} that it is not with your children who have not known and who have not seen the discipline of Adonai your God—his greatness, his strong hand, and his outstretched arm,
\verse and his signs and his deeds that he did in the midst of Egypt to Pharaoh, the king of Egypt, and to all of his land,
\verse and what he did to the army of Egypt and to their horses and to their chariots, and how he made the water of the \textit{Red Sea}\lebnote{Literally “sea of reed”} flow over them, \textit{when they pursued after them},\lebnote{Literally “in/at their pursuing after them”} and so Adonai has destroyed them, \textit{as is the case today},\lebnote{Literally “until the day the this”}
\verse and what he did to you in the desert until you came up to this place,
\verse and what he did to Dathan and to Abiram, the sons of Eliab, the son of Reuben, how the earth opened its mouth wide and swallowed them, their households and their tents, and all of the living creatures\lebnote{Hebrew “creature”} that were in their possession and that were \textit{following along with them}\lebnote{Literally “in their feet” = “living things traveling along with them”} in the midst of all of Israel.
\verse The fact of the matter is, \textit{your own eyes have seen}\lebnote{Literally “your eyes that are seeing”} all of the great deeds\lebnote{Hebrew “deed”} of Adonai that he did.
\verse “And you must keep all of the commandments\lebnote{Hebrew “commandment”} that I am commanding you \textit{today},\lebnote{Literally “the day”} so that you may have strength and you may go and you may take possession of that land \textit{to which you are crossing}\lebnote{Literally “you are crossing there”} to take possession of it,
\verse so that \textit{you may live long}\lebnote{Literally “you may make long days”} on the land that Adonai swore to your ancestors,\lebnote{Or “fathers”} to give it to them and to their offspring, a land flowing with milk and honey.
\verse For the fact is that the land \textit{that you are about to go into}\lebnote{Literally “you are about to go there”} to take possession of it is not like the land of Egypt, from which you have \textit{come out of},\lebnote{Literally “come out from there”} where you sow your seed and you give water by your foot, \textit{as in a vegetable garden}.\lebnote{Literally “as the garden of the vegetables”}
\verse But the land that you are about to cross into to take possession of it is a land of hills and valleys, and by the rain of the heaven it drinks water,
\verse a land that Adonai your God is caring for it; continually the eyes of Adonai your God are on it, from the beginning of the year up to the end of the year.
\verse “And it will happen that if you listen carefully to my commandments that I am commanding you \textit{today},\lebnote{Literally “the day”} to love Adonai your God, and to serve him with all of your heart and with all of your soul,\lebnote{Or “inner self”}
\verse then ‘I will send the rain for your land in its season, early rain and later rain, and you will gather your grain and your wine and your olive oil.
\verse And I will give vegetation in your field for your livestock, and \textit{you will eat your fill}.’\lebnote{Literally “you will eat and you will be satisfied”}
\verse \textit{Take care}\lebnote{Literally “Watch for yourself”} so that your heart is not easily deceived, and you turn away, and you serve other gods, and you bow down to them.
\verse And then \textit{the anger of Adonai will be kindled against you},\lebnote{Literally “the nose of Adonai become hot against you”} and he will shut up the heavens,\lebnote{Or “sky”} and there shall not be rain, and so the ground will not give its produce, and you will perish quickly from the good land that Adonai is giving to you.
\verse “And you shall put these, my words, on your heart and on your inner self,\lebnote{Or “soul”} and you shall bind them as a sign on your hand and let them be as an emblem between your eyes.
\verse And you shall teach them to your children\lebnote{Or “sons”} by talking about them when you sit in your house and when you travel on the road and when you lie down and when you get up.
\verse And you shall write them on the doorframes of your house and on your gates,
\verse so that they may multiply your days and the days of your children on the land that Adonai swore to your ancestors\lebnote{Or “fathers”} to give it to them \textit{as long as heaven endures over the earth}.\lebnote{Literally “like the days of heaven above the earth”}
\verse Yes, if you diligently keep all\lebnote{Or “every one of”} this commandment that I am commanding you \textit{to observe it}, \lebnote{Literally “to do it”} by loving Adonai your God \textit{by walking in}\lebnote{Literally “to walk in”} all his ways and by holding fast to him,
\verse then Adonai will drive out all of these nations \textit{before you},\lebnote{Literally “to the face of you”} and you will dispossess nations larger and more numerous than you.
\verse Every place on which the sole of your foot treads, it shall be yours; your boundary shall be from the desert and Lebanon from the river, the river Euphrates, on up to the western sea.\lebnote{That is, the Mediterranean Sea}
\verse No one can take a stand \textit{against you};\lebnote{Literally “to your faces”} your dread and your fear Adonai your God will put on the \textit{surface}\lebnote{Literally “faces”} of all the land \textit{where you tread},\lebnote{Literally “which you tread in it”} just as he \textit{promised}\lebnote{Literally “spoke”} to you.
\verse “See, I am setting \textit{before you}\lebnote{Literally “to the face of you”} \textit{today}\lebnote{Literally “the day”} a blessing and a curse:
\verse the blessing, if you listen to the commandments of Adonai your God that I am commanding you \textit{today},\lebnote{Literally “the day”}
\verse and the curse, if you do not listen to the commandments of Adonai your God, but rather you turn from the way that I am commanding you \textit{today}\lebnote{Literally “the day”} to go after other gods that you have not known.
\verse “And it will happen that when Adonai your God has brought you to \textit{the land that you are going to},\lebnote{Literally “the land that you are going there”} to take possession of it, then\lebnote{Hebrew “and”} you shall pronounce the blessing on Mount Gerizim and the curse on Mount Ebal.
\verse (Are they not beyond the Jordan, \textit{toward the west},\lebnote{Literally “after the way of the descent of the sun”} in the land of the Canaanites living in the Jordan Valley,\lebnote{Hebrew “in the Arabah”} opposite Gilgal beside the terebinth\lebnote{Or “oaks”} of Moreh?)
\verse For you are now about to cross the Jordan to go to take possession of the land that Adonai, your God, is giving to you, and you will take possession of it and live in it,
\verse and you must diligently \textit{observe}\lebnote{Literally “observe to do”} all the rules and the regulations that I am setting \textit{before you}\lebnote{Literally “before your face”} \textit{today}.”\lebnote{Literally “the day”}
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 12
\verseWithHeading{Detailed Stipulations: Purity and Unity} “These are the rules and the regulations \textit{that you must diligently observe}\lebnote{Literally “that you must observe to do”} in the land that Adonai, the God of your ancestors,\lebnote{Or “fathers”} has given to you to take possession of it, during all the days that you are living on the land.
\verse You must completely demolish all of the places there where they served their gods, that is, the nations whom you are about to dispossess, on the high mountains, and on the hills and under each leafy green tree.\lebnote{Or “spreading tree”}
\verse And you shall break down their altars, and you shall smash their stone pillars, and their Asherah poles you must burn with fire, and the images of their gods you shall hew down, and you shall blot out their names from that place.
\verse You shall not worship Adonai your God like this.
\verse \textit{But only}\lebnote{Literally “But if”} to the place that Adonai your God will choose from all of your tribes to place his name there as his dwelling shall you seek, and there you shall go.
\verse And you shall bring there your burnt offerings and your sacrifices and your tithes and \textit{your donations}\lebnote{Literally “the contributions of your hand”} and your votive gifts and your freewill offerings and the firstling of your herd and your flock.
\verse And you shall eat there \textit{before}\lebnote{Literally “before the faces of”} Adonai your God, and you shall rejoice \textit{in all your endeavors},\lebnote{Literally “in all of the sending of your hand”} you and your family in which Adonai your God has blessed you.
\verse “You must not do just as we are doing here \textit{today},\lebnote{Literally “the day”} \textit{each according to all that is right in his eyes}.\lebnote{Literally “each man all of the right in his eyes”}
\verse For you have not come up to now to the resting place and to the inheritance that Adonai your God is giving to you.
\verse But you will cross the Jordan, and you will settle in the land that Adonai your God is giving you as an inheritance, and he will give rest to you from all your enemies from all around, and you will live securely,
\verse \textit{and then}\lebnote{Literally “and it will happen”} at the place that Adonai your God will choose, to let his name dwell there, there you shall bring all the things I am commanding you, your burnt offerings and your sacrifices, your tithes and \textit{your donations},\lebnote{Literally “the contribution of your hand”} and all of the choice things, your votive gifts that you vow to Adonai.
\verse And you shall rejoice \textit{before}\lebnote{Literally “the face of”} Adonai your God, you and your sons and your daughters and your slaves and your slave women and the Levite who is in your \textit{towns},\lebnote{Literally “gates”} because there is not for him a plot of ground and an inheritance with you.
\verse “Take care for yourself so that you do not offer your burnt offerings at just any place that you happen to see,
\verse \textit{but only}\lebnote{Literally “but only if”} at the place that Adonai will choose among one of your tribes; there you shall offer your burnt offerings, and there you shall do all the things that I am commanding you.
\verse “But \textit{whenever you desire}\lebnote{Literally “in all the wanting of your soul/inner self”} you may slaughter, and you may eat meat according to the blessing of Adonai your God that he has given to you in all of your \textit{towns};\lebnote{Literally “gates”} the unclean and the clean may eat it just as they would the gazelle and as the deer.
\verse Only the blood you must not eat, but on the ground you must pour it like water.
\verse You are not allowed to eat in your \textit{towns}\lebnote{Literally “gates”} the tithe of your grain and your wine and your olive oil and the firstborn of your herd and your flock and all of your votive gifts that you vowed and your freewill offering and \textit{your donations}.\lebnote{Literally “the contribution of your hand”}
\verse But only \textit{before}\lebnote{Literally “before the face of”} Adonai your God you shall eat it, at the place that Adonai your God will choose,\lebnote{Hebrew “choose it”} you and your son and your daughter and your slave and your slave woman and the Levite who is in your \textit{towns},\lebnote{Literally “gate”} and you must rejoice \textit{before}\lebnote{Literally “before the face of”} your God \textit{in all your undertakings}.\lebnote{Literally “in all of the sending of your hand”}
\verse \textit{Take care}\lebnote{Literally “Watch for yourself”} so that you do not neglect the Levite all of your days on your land.
\verse “When Adonai your God enlarges your territory just as he has \textit{promised}\lebnote{Literally “spoken”} to you, and you say ‘I want to eat some meat,’ \textit{because you want it},\lebnote{Literally “because your soul/inner self desires it”} \textit{whenever you desire}\lebnote{Literally “at/in all the wanting of your soul/inner self”} you may eat meat.
\verse If the place that\lebnote{Or “where”} Adonai your God will choose to put his name there is too far from you, and you slaughter any of your herd and any of your flock that Adonai has given to you just as I have commanded you, then you may eat whenever you desire in your \textit{towns}.\lebnote{Literally “gates”}
\verse Surely just as the gazelle and the deer is eaten, so both the unclean and the clean together may eat it.
\verse Only, be sure not to eat the blood, because the blood is the life, and you shall not eat the life with the meat.
\verse You shall not eat it, but on the ground you shall pour it out like water.
\verse You shall not eat it, so that \textit{it will go well}\lebnote{Literally “it/he is good”} for you and your children after you, because then you will be doing what is\lebnote{Hebrew “the”} right in the eyes of Adonai.
\verse Only your holy objects \textit{that are yours}\lebnote{Literally “that are for you”} and your votive gifts you must carry and you must bring to the place that Adonai will choose.
\verse And you shall offer your burnt offerings, the flesh and the blood on the altar of Adonai your God, and the blood of your sacrifices you shall pour out on the altar of Adonai your God, but the meat you may eat.
\verse \textit{Be careful to obey}\lebnote{Literally “Watch carefully and listen”} all these things that I am commanding you, so that \textit{it will go well}\lebnote{Literally “he/it is good”} for you and for your children after you \textit{forever},\lebnote{Literally “until ever”} because then you will be doing what is\lebnote{Hebrew “the”} good and right\lebnote{Hebrew “the right”} in the eyes of Adonai your God.
\verse “When Adonai your God has cut off the nations whom you are \textit{about to go to},\lebnote{Literally “about to go to there”} to dispossess them \textit{before you},\lebnote{Literally “from the face of you”} and you have dispossessed them, and you live in their land,
\verse \textit{take care}\lebnote{Literally “Watch for yourself”} so that you are not ensnared \textit{into imitating them}\lebnote{Literally “after them”} after their being destroyed from \textit{before you},\lebnote{Literally “from the face of you”} and so that you not enquire concerning their gods, \textit{saying},\lebnote{Literally “to say”} ‘How did these nations serve their gods, and thus \textit{I myself}\lebnote{Emphatic use of the pronoun} want to do also.’
\verse You must not do so toward Adonai your God, because of every detestable thing they have done for their gods Adonai hates, for even their sons and their daughters they would burn in the fire to their gods.
\verse \lebnote{Deuteronomy 12:32–13:18 in the English Bible is 13:1–19 in the Hebrew Bible} All of the things\lebnote{Hebrew “thing”} that I am commanding you, \textit{you must diligently observe};\lebnote{Literally “you must observe to do it”} you shall not add to it, and you shall not take away from it.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 13
\verse “If a prophet stands up in your midst or a dreamer of dreams\lebnote{Hebrew “dream”} and he gives to you a sign or wonder,
\verse and the sign or\lebnote{Hebrew “and”} the wonder comes about that he spoke\lebnote{Or “promised/declared”} to you, \textit{saying},\lebnote{Literally “to say”} ‘Let us go after other gods (those whom you have not known), and let us serve them,’
\verse you must not listen to the words of that prophet or to that dreamer, for Adonai your God \textit{is testing you to know whether you love}\lebnote{Literally “a testing of Adonai your God concerning you in order to know whether there is you loving”} Adonai your God with all of your heart and with all of your inner self.\lebnote{Or “soul”}
\verse You shall go after Adonai your God, and him you shall revere, and his commandment you shall keep, and to his voice you shall listen, and him you shall serve, and to him you shall hold fast.
\verse But that prophet or the dreamer of that dream shall be executed, for he spoke falsely about Adonai your God, the one bringing you out from the land of Egypt and the one redeeming you from the house of slavery, in order to seduce you from the way that Adonai your God commanded you to go in it; so in this way you shall purge the evil from your midst.
\verse “If your brother, the son of your mother or your son or your daughter \textit{or your wife whom you embrace}\lebnote{Literally “or the wife of your embrace”} or \textit{your intimate friend}\lebnote{Literally “your friend who is like your soul/inner self”} in secrecy \textit{says},\lebnote{Literally “to say”} ‘Let us go and let us serve other gods!’ gods that you and your ancestors\lebnote{Or “fathers”} have not known,
\verse from among the gods of the people who are around you, \textit{those near you or those far from you},\lebnote{Literally “the near ones to you or the distant ones from you”} from one end of the earth and up to the other end of the earth,
\verse you must not give in to him, and you shall not listen to him, and your eye shall not take pity on him, and you shall not have compassion, and you shall not cover up\lebnote{Or “conceal” him} for him.
\verse But you shall certainly kill him; your hand shall be first against him to kill him and next the hand of all of the people.
\verse And you shall stone him with stones and let him die, for he tried to seduce you from Adonai your God, the one bringing you from the land of Egypt, from the house of slavery.
\verse And all of Israel shall hear, and they shall fear, and \textit{they shall not continue to act}\lebnote{Literally “they shall not add to do”} according to this evil thing in your midst.
\verse “If you hear in one of your towns which Adonai your God is giving to you to live in, someone \textit{saying that}\lebnote{Literally “to say”}
\verse \textit{worthless men}\lebnote{Literally “sons of Belial”} have gone out from your midst and have seduced the inhabitants of their town, \textit{saying},\lebnote{Literally “to say”} ‘Let us go and serve other gods!’ whom you have not known,
\verse then you shall inquire and examine and interrogate thoroughly, and, look! It is true; the thing has actually been done, this detestable thing in your midst,
\verse then you shall certainly strike down the inhabitants of that town with the \textit{edge}\lebnote{Literally “mouth”} of the sword; you shall destroy it and everything in it, its domestic animals with the \textit{edge}\lebnote{Literally “mouth”} of the sword.
\verse And then you shall gather all of its booty into the middle of its public square, and you shall burn the town and all of its war-booty totally for Adonai your God, and it shall be \textit{a pile of rubble forever};\lebnote{Literally “a tell/ruin forever”} it shall not be built again.
\verse And let not something cling to your hand from the things devoted to destruction, so that Adonai may turn back \textit{from his burning anger},\lebnote{Literally “from the burning of his nose”} and he may show compassion to you and he may continue to show compassion and so multiply you \textit{just as he swore}\lebnote{Literally “as that he swore”} to your ancestors,\lebnote{Or “fathers”}
\verse if you listen to the voice of Adonai your God, to keep all of his commandments that I am commanding to you \textit{today}\lebnote{Literally “the day”} so as to do the right thing in the eyes of Adonai your God.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 14
\verse “You are children\lebnote{Or “sons”} of Adonai your God; therefore you must not gash\lebnote{Or “cut”} yourself, and \textit{you must not make your forehead bald}\lebnote{Literally “you shall not make/place baldness between your eyes”} for the dead.
\verse For you are a people holy to Adonai your God, and you Adonai has chosen to be a treasured possession from among all of the peoples that are on the surface\lebnote{Or “face”} of the earth.
\verse You shall not eat any detestable thing.
\verse These are the animals you may eat: ox, \textit{sheep, goats},\lebnote{Literally “small livestock of the sheep and a small livestock of the goats”}
\verse deer, gazelle, roebuck, wild goat, ibex, antelope, and mountain sheep.
\verse And any animal having a split hoof\lebnote{Or “divides the hoof”} and so \textit{a dividing of the hoof into two parts}\lebnote{Literally “a dividing cleft creating two hoofs”} \textit{and that chews the cud}\lebnote{Literally “and that brings up the cud”} among the animals\lebnote{Hebrew “animal”}—that animal you may eat.
\verse Only these you may not eat from \textit{those chewing the cud}\lebnote{Literally “bringing up the cud”} and from \textit{those having a division of the hoof}:\lebnote{Literally “those having a division of the hoof divided”} the camel and the hare and the coney, because \textit{they chew the cud},\lebnote{Literally “they bringing up of the cud”} but they do not divide the hoof; they are therefore unclean for you.
\verse And also the pig \textit{because it has a division of the hoof}\lebnote{Literally “because a division of the hoof”} \textit{but does not chew the cud};\lebnote{Literally “but not a chewing cud”} it is unclean for you; from their meat you shall not eat, and you shall not touch their \textit{carcasses}.\lebnote{Literally “dead body”}
\verse “This is what you shall eat from all that is in the water: \textit{everything}\lebnote{Literally “all of that”} \textit{that has fins and scales}\lebnote{Literally “for it is fins and scales,” showing possession of these features} you may eat.
\verse But \textit{anything that does not have}\lebnote{Literally “all of that there is not for it,” showing lack of possession of these features} fins and scales, you may not eat, for it is unclean for you.
\verse “All of the birds that are clean you may eat.
\verse Now these are the ones you shall not eat \textit{any of them}:\lebnote{Literally “from them”} the eagle and the vulture and the short-toed eagle,\lebnote{This list of birds is difficult to translate since the terms are not definitely known: e.g., some translations render the last bird as a “buzzard” (NASV); other translations give different names for all three: griffon vulture, black vulture, bearded vulture (NEB)}
\verse and the red kite\lebnote{Various options are available: large bird, kite, red kite, glede, buzzard} and the black kite or \textit{any kind of falcon},\lebnote{Literally “or the falcon according to its kind”; other options for falcon: bird, falcon, kite (others as carrion-bird), vulture, crow or raven, buzzard}
\verse and any kind of crow\lebnote{Or others translate as “raven”} according to its kind,
\verse and the \textit{ostrich}\lebnote{Literally “daughter of the ostrich”; others “desert owl”} and the short-eared owl and the seagull\lebnote{Or “long-eared owl”} and the hawk according to its kind,
\verse the little owl and the great owl and the barn owl,\lebnote{Or “white owl”}
\verse and the desert owl\lebnote{Or “large bird” or “horned-owl”} and the carrion vulture\lebnote{Or “large bird”} and the cormorant,\lebnote{Or “large bird,” or “fisher-owl”}
\verse and the stork and the heron according to its kind and the hoopoe and the bat.
\verse And also all of \textit{the winged insects};\lebnote{Literally “the swarmers of the flyers” or “all the swarms of things that fly”} they are unclean for you; you shall not eat them.
\verse You may eat any clean bird.
\verse “You shall not eat any carcass;\lebnote{Or “corpse”} you may give it to the alien who is in your \textit{towns},\lebnote{Literally “gates”} and he may eat it, or you may sell it to a foreigner, for you are a holy people for Adonai your God; you may not boil a kid in its mother’s milk.
\verse “Certainly you must give a tithe of all the yield of your seed, \textit{which comes forth from your field year after year}.\lebnote{Literally “the going forth of the field year by year”}
\verse And you shall eat \textit{before}\lebnote{Literally “before the faces of”} Adonai your God in the place that he will choose to make to dwell his name there the tithe of your grain, your wine and your olive oil and the firstling of your herd and your flock, so that you may learn to revere Adonai your God \textit{always}.\lebnote{Literally “all of the days”}
\verse But if \textit{the distance is too great for you},\lebnote{Literally “it is great from you the journey that”} so that you are not able to transport\lebnote{Or “carry”} it, because the place that Adonai your God will choose to set his name there, it is too far from you, when Adonai your God will bless you,
\verse then in that case \textit{you may exchange for money},\lebnote{Literally “you may give it in for the money/silver”} and you shall take\lebnote{Or “bind”} the money to your hand and go to the place that Adonai your God will choose.
\verse You may spend the money for anything \textit{that you desire},\lebnote{Literally “that your soul/inner self desires”} for oxen or for sheep or for wine or for strong drink or for anything \textit{that you desire},\lebnote{Literally “that your soul/inner self desires”} and you shall eat it there \textit{before}\lebnote{Literally “to the face of”} Adonai your God, and you shall rejoice, you and your household.
\verse And as to the Levite who is in your \textit{towns},\lebnote{Literally “gates”} you shall not neglect him, because there is not a plot of ground for him and an inheritance along with you.
\verse “At the end of three years you shall bring out all of the tithe of your yield for that year, and you shall store it in your \textit{towns}.\lebnote{Literally “gates”}
\verse And so the Levite may come, because there is no plot of ground for him or an inheritance with you, and the alien also may come and the orphan and the widow that are in your \textit{towns},\lebnote{Literally “gates”} and \textit{they may eat their fill},\lebnote{Literally “they may eat and they may be satisfied”} so that Adonai your God may bless you in all of the work of your hand that you undertake.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 15
\verse “At the end of seven years you shall grant a remission of debt.
\verse And this is the manner of the remission of debt: every \textit{creditor}\lebnote{Literally “owner of the loan of his hand”} shall remit his claim that he holds against his neighbor, and he shall not exact payment from his brother because there\lebnote{Hebrew “it”} a remission of debt has been proclaimed unto\lebnote{Hebrew “for”} Adonai.
\verse With respect to the foreigner you may exact payment, but \textit{you must remit}\lebnote{Literally “your hand shall remit”} what shall be owed to you with respect to your brother.
\verse Nevertheless, there\lebnote{Hebrew “it”} shall not be among you a poor person, because Adonai will certainly bless you in the land that Adonai your God is giving to you as an inheritance, to take possession of it.
\verse If only you listen well to the voice of Adonai your God \textit{by observing diligently}\lebnote{Literally “to observe so as to do”} all of these commandments\lebnote{Hebrew “commandment”} that I am commanding you \textit{today}.\lebnote{Literally “the day”}
\verse When Adonai your God has blessed you, just as he \textit{promised}\lebnote{Literally “spoke”} to you, then you will lend to many nations, but you will not borrow from them, and you will rule over many nations, but they will not rule over you.
\verse If there is a poor person among you from among one of your brothers in one of your \textit{towns}\lebnote{Literally “gates”} that Adonai your God is giving to you, you shall not harden your heart, and you shall not shut your hand toward \textit{your brother who is poor}.\lebnote{Literally “from among your brothers, the poor one”}
\verse But you shall certainly open your hand for him, and \textit{you shall willingly lend}\lebnote{Literally “lending you shall lend”} to him enough to meet his need, \textit{whatever it is}.\lebnote{Literally “whatever is lacking for him”}
\verse \textit{Take care}\lebnote{Literally “Watch for yourself”} so that there\lebnote{Hebrew “it”} will not be \textit{a thought of wickedness}\lebnote{Literally “a thing in your heart wickedness”} in your heart, \textit{saying},\lebnote{Literally “to say”} ‘The seventh year, the year of the remission of debt is near,’ \textit{and you view your needy neighbor with hostility},\lebnote{Literally “is bad your eye against your brother who is needy”} and so you do not give to him, and he might cry out against you to Adonai, and \textit{you would incur guilt against yourself}.\lebnote{Literally “it will be against you a sin”}
\verse By all means you must give to him, and \textit{you must not be discontented}\lebnote{Literally “and not shall be bad/evil your heart at/when”} at your giving to him, because on account of this very thing, Adonai your God will bless you in all your work and \textit{in all that you undertake}.\lebnote{Literally “in all of the sending/putting forth of your hand”}
\verse For the poor\lebnote{Or “the needy person”} will not cease to be \textit{among you}\lebnote{Literally “from the midst of “} in the land; therefore I am commanding you, \textit{saying},\lebnote{Literally “to say”} ‘You shall willingly open your hand to your brother, to your needy and to your poor that are in your land.’
\verse If your relative\lebnote{Or “brother”} who is a Hebrew man or a Hebrew woman is sold to you, and he or she has served you six years, then in the seventh year you shall send that person out \textit{free}.\lebnote{Literally “free from with you”}
\verse And when you send him out free from you, you shall not send him away empty-handed.
\verse You shall generously supply him from among your flocks and from your threshing floor and from your press; according to that with which Adonai your God has blessed you, you shall give to him.
\verse And remember that you were a slave in the land of Egypt, and Adonai your God redeemed you; therefore I am commanding you thus \textit{today}.\lebnote{Literally “the day”}
\verse And then if it will happen that he says to you, ‘\textit{I do not want to go out}\lebnote{Literally “I will not go out”} from you,’ because he loves you and your family, because it is good for him to be with you;
\verse then you shall take an awl, and you shall thrust it through his earlobe and into the door, and he shall be to you \textit{a slave forever};\lebnote{Literally “a slave of eternity”} and you shall also do likewise for your slave woman.
\verse It shall not be hard in your eyes \textit{when you send him forth free},\lebnote{Literally “in/at you to send him forth free from being with you”} because for six years he has served you worth twice the wage of a hired worker; and Adonai your God will bless you \textit{in whatever you will do}.\lebnote{Literally “in all of that you will do”}
\verse “Every firstling male that is born of your herd and of your flock you shall consecrate to Adonai your God; you shall not do work with the firstling of your ox, and you shall not shear the firstling of your flock.
\verse Rather \textit{before Adonai}\lebnote{Literally “in the face of Adonai”} your God you shall eat it year by year at the place Adonai will choose, you and your household.
\verse But if there is a physical defect in it, such as lameness or blindness, any serious defect, you shall not sacrifice it to Adonai your God.
\verse In your \textit{towns}\lebnote{Literally “gates”} you shall eat it, the unclean and the clean together may eat it, just as they eat the gazelle and as they eat the deer.
\verse But you shall not eat its blood; you shall pour it on the ground like water.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 16
\verse “Observe the month of Abib, and you shall keep the Passover to Adonai your God, for in the month of Abib Adonai your God brought you out from Egypt by night.
\verse And you shall offer the Passover sacrifice to Adonai your God from among your flock and herd at the place that Adonai will choose, to let his name dwell there.
\verse You shall not eat \textit{with it}\lebnote{Literally “in addition to” or “upon it”} anything leavened; seven days you shall eat \textit{with it}\lebnote{Literally “in addition to” or “upon it”} unleavened bread of affliction, because in haste you went out from the land of Egypt, so that you will remember the day of your going out from the land of Egypt all the days of your life.
\verse And leaven shall not be seen with\lebnote{Or “for”} you in any of your territory\lebnote{Or “all of” your territory} for seven days, and none of the meat that you will slaughter on the evening on the first day shall remain overnight until morning.
\verse You are not allowed to offer the Passover sacrifice in one of your \textit{towns}\lebnote{Literally “gates”} that Adonai your God is giving to you,
\verse but only at the place that Adonai your God will choose, to let his name dwell there; you shall offer the Passover sacrifice \textit{in the evening at sunset},\lebnote{Literally “in the evening as the sun goes/sets”} at the designated time\lebnote{The Hebrew word here indicates the specific time that God had chosen to bring Israel out of Egypt} of your going out from Egypt.
\verse And you shall cook, and you shall eat it at the place that Adonai your God will choose; and you may turn in the morning and go to your tents.
\verse Six days you shall eat unleavened bread, and on the seventh day there shall be an assembly for Adonai your God; you shall not do work.
\verse “You shall count off seven weeks for you; \textit{from the time you begin to harvest the standing grain}\lebnote{Literally “from the beginning of the sickle against the standing grain”} you shall begin to count seven weeks.
\verse And then you shall celebrate the Feast of Weeks for Adonai your God with the measure of the freewill offering of your hand that you shall give just as Adonai your God has blessed you.
\verse And you shall rejoice before Adonai your God, you and your son and your daughter and your slave and your slave woman and the Levite that is in your \textit{towns}\lebnote{Literally “gates”} and the alien and the orphan and the widow who are in your midst in the place that Adonai your God will choose to let his name dwell there.
\verse And you shall remember that you were a slave in Egypt, and so \textit{you shall diligently observe}\lebnote{Literally “you shall observe and do”} these rules.
\verse “You shall celebrate the Feast of Booths for yourselves\lebnote{Hebrew “for you”} seven days \textit{at the gathering in of the produce}\lebnote{Literally “at your gathering of the produce”} from your threshing floor and from your press;
\verse and you shall rejoice at your feast, you and your son and your daughter and your slave and your slave woman and the Levite and the orphan and the widow that are in your \textit{towns}.\lebnote{Literally “gates”}
\verse Seven days you shall celebrate your feast to Adonai your God at the place Adonai will choose, for Adonai your God shall bless you in all of your produce\lebnote{Or “increase”} and in all of the work of your hand, and you shall surely be rejoicing.\lebnote{Or “joyful”}
\verse Three times in the year all of your males shall appear \textit{before }\lebnote{Literally “with the face of”} Adonai your God at the place that he will choose, at the Feast of Unleavened Bread and at the Feast of Weeks and at the Feast of Booths, and they shall not appear \textit{before Adonai}\lebnote{Literally “with the face of Adonai”} empty-handed.
\verse Each person \textit{shall give as he is able},\lebnote{Literally “according to the gift of his hand”} that is, according to the blessing of Adonai your God that he has given to you.
\verse “You shall appoint judges and officials for you in all your \textit{towns}\lebnote{Literally “gates”} that Adonai your God is giving to you throughout your tribes, and you shall render for the people \textit{righteous judgments}.\lebnote{Literally “a judgment based on righteousness”}
\verse You shall not subvert\lebnote{Or “distort/pervert”} justice; you shall not \textit{show partiality};\lebnote{Literally “recognize faces”} and you shall not take a bribe, for the bribe makes blind the eyes of the wise and misrepresents the words of the righteous.
\verse \textit{Justice, only justice}\lebnote{Literally “justice justice”} you shall pursue, so that you may live, and you shall take possession of the land that Adonai your God is giving to you.
\verse You shall not plant for yourselves\lebnote{Hebrew “for/to you” but with collective meaning} \textit{an Asherah pole}\lebnote{Literally “an Asherah of any wood/tree”} beside the altar of Adonai your God that you make for yourselves.\lebnote{Hebrew “for/to you” but with collective meaning}
\verse And you shall not set up for yourselves a stone pillar, a thing that Adonai your God hates.
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 17
\verse “You shall not sacrifice to Adonai your God an ox or sheep \textit{that has a physical defect}\lebnote{Literally “that is on it a physical defect”} \textit{of anything seriously wrong},\lebnote{Literally “of any thing bad/evil”} for that is a detestable thing to Adonai your God.
\verse If there is found in one of your \textit{towns}\lebnote{Literally “gates”} that Adonai your God is giving to you a man or a woman that does evil in the eyes of Adonai your God to transgress his covenant
\verse and by going and serving other gods and so he bows down to them and to the sun or to the moon or to any of the host of heaven \textit{which I have forbidden},\lebnote{Literally “which not I have commanded”}
\verse and it is reported to you or you hear about it and you enquire about it thoroughly and, indeed,\lebnote{Or “look!”} the trustworthiness of the deed\lebnote{Hebrew “the thing”} has been established, it \textit{has occurred},\lebnote{Literally “has been done”} this detestable thing, in Israel,
\verse then you shall bring out that man or that woman who has\lebnote{Hebrew “have”} done this evil thing to your gates; that is, the man or the woman, and you shall stone them with stones \textit{to death}.\lebnote{Literally “and so they die”}
\verse \textit{On the evidence of}\lebnote{Literally “By the mouth of”} two or three witnesses \textit{the person shall be put to death}.\lebnote{Literally “he/she shall be put to death, the dead person”} The person\lebnote{Hebrew “he/she”} shall not be put to death by the mouth of one witness.
\verse The hand of the witnesses shall be first against the person\lebnote{Hebrew “him”} to kill the person,\lebnote{Hebrew “him”} and afterward the hands\lebnote{Hebrew “hand”} of all the people, and so you shall purge the evil from your midst.
\verse “\textit{If a matter is too difficult for you},\lebnote{Literally “If is difficult/wonderful from you a thing for judgment”} for example disputes between blood and blood,\lebnote{Or “between one homicide and another”} between legal claim and legal claim\lebnote{Or “one kind of lawsuit and another”} and between assault and assault\lebnote{Or “one kind of abuse and another”} and between matters of discernment in your \textit{towns},\lebnote{Literally “gates”} then you shall get up and you shall go to the place that Adonai your God will choose;
\verse then you shall go to the priests and the Levites and to the judge who will be in office in those days, and you shall enquire, and they shall announce to you \textit{the verdict}.\lebnote{Literally “the word of the decision”}
\verse “And \textit{you shall carry out exactly the decision}\lebnote{Literally “and you shall do according to the mouth of the word”} that they announced to you from that place that Adonai will choose, and \textit{you shall diligently observe}\lebnote{Literally “and you shall observe and do”} according to all that they instruct you.
\verse And so according to \textit{the instruction of the law}\lebnote{Literally “mouth of the law”} that they teach you and according to the decisions that they say to you, you shall do; you shall not turn from the word that they tell you to the right or to the left.
\verse And the man who treats with contempt\lebnote{Or “acts presumptuously”} so as not to listen to the priest who is standing to minister on behalf of Adonai your God or to the judge, that man shall die; so you shall purge the evil from Israel.
\verse And all the people will hear and will be afraid, and they will not behave presumptuously again.
\verse “When you have come to that land that Adonai your God is giving to you and you have taken possession of it and you have settled in it, and you say, ‘I will set over me a king like all the nations that are around me,’
\verse indeed, you may set a king over you whom Adonai your God will choose, from the midst of your countrymen\lebnote{Or “brothers”} you must set a king over you; you are not allowed to appoint over you a man, a foreigner, who is not your countryman.\lebnote{Or “brother”}
\verse Except, he may \textit{not make numerous}\lebnote{Literally “not multiply”} for himself horses, and he may not allow the people to to go to Egypt \textit{in order to increase horses},\lebnote{Literally “in order to make numerous horse”} for Adonai has said to you that \textit{you may never return}.\lebnote{Literally “not you may do again to return”}
\verse And he must not \textit{acquire many}\lebnote{Or “make numerous”} wives for himself, so that his heart would turn aside; and \textit{he must not accumulate silver and gold for himself excessively}.\lebnote{Literally “gold and silver not he must make numerous for him very”}
\verse “\textit{And then}\lebnote{Literally “And it shall happen”} \textit{when he is sitting}\lebnote{Literally “as/when his sitting”} on the throne of his kingdom, then he shall write for himself a copy of this law on a scroll \textit{before}\lebnote{Literally “to the face of”} the Levitical priests.
\verse And it shall be with him, and he shall read it\lebnote{Hebrew “in it”} all the days of his life, so that he may learn to revere Adonai your God by \textit{diligently observing}\lebnote{Literally “by keeping ... to do them”} all the word of this law and these rules,
\verse so as not to exalt his heart above his countrymen\lebnote{Or “brothers”} and not to turn aside from the commandment to the right or to the left, so that \textit{he may reign long over his kingdom},\lebnote{Literally “he may make long his days over his kingdom”} he and his children in the midst of Israel.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 18
\verse “And there shall not be for the Levitical priests, the whole tribe of Levi, a plot of ground and an inheritance with Israel, rather they may eat an offering made by fire as their inheritance,\lebnote{The meaning of the Hebrew text here is uncertain; possibly it reads: “rather an offering by fire then they will eat as their inheritance/patrimony”} for Adonai.
\verse And there shall not be for them\lebnote{Hebrew “him”} an inheritance of land in the midst of his brothers; rather Adonai is his inheritance, just as he \textit{promised}\lebnote{Literally “spoke”} to them.\lebnote{Hebrew “him”}
\verse Now this shall be the share of the priest from the people, from \textit{those who sacrifice the sacrifice},\lebnote{Literally “the sacrificers of the sacrifice”} whether it is an ox, sheep, or goat, and they shall give the priest the shoulder and the jawbones and the stomach.
\verse The firstfruits\lebnote{Hebrew “firstfruit”} of your grain, your wine, and your olive oil and the firstfruits\lebnote{Hebrew “firstfruit”} of the fleece of your sheep you shall give to him.
\verse For Adonai your God has chosen him from among all your tribes to stand to minister in the name of Adonai, he and his sons \textit{forever}.\lebnote{Literally “all of the days”}
\verse And if a\lebnote{Hebrew “the”} Levite comes from one of your \textit{towns}\lebnote{Literally “gates”} from \textit{anywhere in Israel}\lebnote{Literally “all of Israel”} where he is residing, \textit{he may come whenever he desires},\lebnote{Literally “he may come in all the desire of his soul/inner self”} to the place that Adonai will choose,
\verse and he may minister in the name of Adonai his God, just like all his brothers, \textit{the Levites who stand there}\lebnote{Literally “the Levites the ones standing there”} \textit{before}\lebnote{Literally “to the face of”} Adonai.
\verse They shall eat \textit{equal portions},\lebnote{Literally “a portion like a portion”} apart from what he may receive from the sale of his patrimony.\lebnote{Hebrew meaning of these phrases/words is not certain; also could translate as “except what they receive from the sale of their fathers’ estates” (NASB); or “what he may inherit from his father’s family”}
\verse “When you come to the land that Adonai your God is giving to you, you must not learn to do like the detestable practices of those nations.
\verse There shall not be found among you one who makes his son or his daughter go through the fire, or \textit{one who practices divination},\lebnote{Literally “a diviner of divination”} or an interpreter of signs,\lebnote{Or “soothsayer”} or an augur,\lebnote{Or “an interpreter of omens”} or sorcerer,
\verse or one who casts magic spells, or one who consults a spirit of the dead,\lebnote{Or “medium”} or spiritist, or one who inquires of the dead.\lebnote{Or “necromancer”}
\verse For everyone doing these things is detestable to Adonai, and because of these detestable things Adonai your God is driving them out from \textit{before you}.\lebnote{Literally “the face of you”}
\verse You must be blameless before Adonai your God.
\verse For these nations that you are about to dispossess listen to interpreters of signs\lebnote{Or “practice witchcraft”} and to diviners, but Adonai your God has not allowed you to do the same.
\verse “Adonai your God will raise up for you a prophet like me from your midst, from your countrymen,\lebnote{Or “your brothers”} and to him you shall listen.
\verse This is \textit{according to all that you asked}\lebnote{Literally “just as all of that you asked”} from Adonai your God at Horeb, on the day of the assembly, \textit{saying},\lebnote{Literally “to say”} ‘\textit{I do not want again to hear}\lebnote{Literally “not I want to do again to hear”} the voice of Adonai my God, and I do not want to see again this great fire, so that I may not die!’
\verse And Adonai said to me, ‘They are right in what they have said.
\verse I will raise up a prophet for them \textit{from among their countrymen}\lebnote{Literally “from the midst of their brothers”} like you, and I will place\lebnote{Or “give”} my words into his mouth, and he shall speak to them \textit{everything that I command him}.\lebnote{Literally “all of what I command him”}
\verse \textit{And then}\lebnote{Literally “And it will happen”} the man that will not listen to my words that he shall speak in my name, I will hold accountable.
\verse However, the prophet that behaves presumptuously by speaking a word in my name that I have not commanded him to speak, and who speaks in the name of other gods, then that prophet shall die.’
\verse And if you say \textit{to yourself},\lebnote{Literally “in your heart”} ‘How can we know the word that Adonai has not spoken to him?’
\verse Whenever what the prophet spoke in the name of Adonai, the thing does not take place and does not come about, that is the thing that Adonai has not spoken to him.\lebnote{Or “that Adonai has not spoken it”} Presumptuously the prophet spoke it; you shall not fear that\lebnote{Hebrew “the”} prophet.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 19
\verse “When Adonai your God has \textit{exterminated}\lebnote{Literally “cut off”} the nations concerning whom Adonai your God is giving to you their land, and you have dispossessed them, and you have settled in their towns and in their houses,
\verse you shall set apart three cities for you in the midst of your land that Adonai your God is giving to you to take possession of it.
\verse You shall prepare the roads\lebnote{Hebrew “road”} for yourselves, and you shall divide the regions of your land into thirds that Adonai your God gives you as a possession, so that \textit{it will be available for any manslayer to flee there}.\lebnote{Literally “it shall be to flee there all/anyone killing”}
\verse “Now this is the case of the manslayer who may flee there and live there who has killed his neighbor \textit{unintentionally},\lebnote{Literally “with no knowledge”} and he did not hate him \textit{previously}.\lebnote{Literally “from yesterday and the day before”}
\verse \textit{For example},\lebnote{Literally “And as”} when somebody goes with his neighbor into the forest to cut wood, and the iron head slips from the handle of the tool and strikes his neighbor and he dies, then he may flee to one of these cities, and so he may live.
\verse He does this lest the avenger of blood might pursue after the killer, because \textit{he is hot with anger}\lebnote{Literally “is hot his heart”} and he overtakes him, because it is a long distance to the city of refuge, and so \textit{he kills him},\lebnote{Literally “he strikes him down as to his life/soul”} but \textit{he did not deserve a death sentence},\lebnote{Literally “for him there was not a judgment of death”} because he was not hating him \textit{before}.\lebnote{Literally “from yesterday and the day before”}
\verse Therefore I am commanding you, \textit{saying},\lebnote{Literally “to say”} ‘You shall set apart three cities.’
\verse Then if Adonai your God enlarges your territory just as he swore to your ancestors\lebnote{Or “fathers”} and gives\lebnote{Hebrew “will give”} to you all the land that he \textit{promised}\lebnote{Literally “spoke”} to give to your ancestors,\lebnote{Or “fathers”}
\verse then \textit{if you diligently observe this entire commandment}\lebnote{Literally “if you observe all of the commandment the this to do it”} that I am commanding you \textit{today}\lebnote{Literally “the day”} by loving Adonai your God and by going\lebnote{Or “walking”} in his ways \textit{at all times},\lebnote{Literally “all of the days”} then you shall add three more cities for yourselves to these three.
\verse Do this so that innocent blood will not be shed\lebnote{Hebrew “is not shed”} in the midst of your land that Adonai your God is giving to you as an inheritance and thereby bloodguilt would be on you.\lebnote{Hebrew “shall be on you,” but conditional sense of imperfect is clear}
\verse But if someone hates\lebnote{Hebrew “is hating”} his neighbor and lies in wait for him and rises up\lebnote{Or “gets up”} against him \textit{and murders him},\lebnote{Literally “and strikes him mortally with regard to his life and he dies”} and the murderer flees to one of these cities,
\verse then the elders of his city shall send and take him from there, and they shall give him into the hand of the avenger of blood, and he shall be put to death.
\verse Your eye shall not take pity on him, and you shall purge the guilt of innocent blood from Israel, \textit{so that good will be directed toward you}.\lebnote{Literally “and good shall be for you” or “it shall be good for you”}
\verse “You shall not move the boundary marker of your neighbor that \textit{former generations}\lebnote{Literally “the first settlers/ancestors”} set up on your property in the land that Adonai your God is giving to you to take possession of it.
\verse \textit{The testimony of a single witness may not be used to convict}\lebnote{Literally “Not shall get up only a single witness against a man”} with respect to any crime and for any wrongdoing in any offense that a person\lebnote{Hebrew “he”} committed; on the \textit{evidence}\lebnote{Literally “mouth”} of two witnesses or on the \textit{evidence}\lebnote{Literally “mouth”} of three witnesses \textit{a charge shall be sustained}.\lebnote{Literally “shall be established a case/charge”}
\verse If \textit{a malicious witness}\lebnote{Literally “a witness of violence”} gets up \textit{to accuse}\lebnote{Literally “against”} anyone to testify against him falsely,
\verse then the two men \textit{to whom the legal dispute pertains}\lebnote{Literally “who for them are the legal dispute”} shall stand \textit{before}\lebnote{Literally “to the face of”} Adonai, \textit{before}\lebnote{Literally “to the face of”} the priests and the judges who are in office in those days.
\verse Then judges shall make a thorough inquiry, and if it turns out\lebnote{Or “in fact”} that the witness is a false witness and he testified falsely against his brother,
\verse then you shall do to him as he meant\lebnote{Or “planned”} to do to his brother, and so you shall purge the evil from your midst.
\verse \textit{And the rest}\lebnote{Literally “and those remaining”} shall hear and shall fear, and \textit{they shall not continue to do such a thing again}\lebnote{Literally “and they shall not do again to do again”} as this evil thing in your midst.
\verse \textit{You must show no pity}:\lebnote{Literally “And not take pity your eye”} life for life, eye for eye, tooth for tooth, hand for hand, foot for foot.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 20
\verse “If you go out to war against your enemies and you see a horse and a chariot, \textit{an army}\lebnote{Literally “a people”} larger that you, you shall not be afraid because of them; for Adonai your God is with you, the one who brought you from the land of Egypt.
\verse \textit{And then}\lebnote{Literally “And it will happen”} when you approach the battle, then the priest shall come near and speak to the troops.
\verse And he shall say to them, ‘Hear, Israel, you are near \textit{today}\lebnote{Literally “the day”} to the battle against your enemies; \textit{do not lose heart};\lebnote{Literally “do not be faint/tender-hearted”} you shall not be afraid, and you shall not panic, and you shall not be terrified \textit{because of them},\lebnote{Literally “from their faces”}
\verse for Adonai your God is going with you to fight for you against your enemies to help you.’
\verse And the officials shall speak to the troops, \textit{saying},\lebnote{Literally “to say”} ‘Who is the man who has built a new house and has not dedicated it? Let him go and return to this house, so that he does not die in battle and \textit{another man}\lebnote{Literally “a man other”} dedicates it.
\verse And who is the man that has planted a vineyard and has not enjoyed it? Let him go and let him return to his house, so that he does not die in battle and \textit{another man}\lebnote{Literally “a man other”} enjoys it.
\verse And who is the man who got engaged to a woman and has not married her? Let him go and let him return to his house, so that he does not die in battle and \textit{another man}\lebnote{Literally “a man other”} marries her.’
\verse And the officials shall continue to speak to the troops, and they shall say, ‘\textit{What man}\lebnote{Literally “Who is the man”} is afraid \textit{and disheartened}?\lebnote{Literally “and soft/weak of the heart”} Let him go, and let him return to his house, and let him not cause the heart of his brothers to melt\lebnote{The verb has a causal meaning here} like his.’
\verse And \textit{when the officials have finished speaking}\lebnote{Literally “it will happen as to finish the officials to speak”} to the army troops, then they shall appoint commanders of divisions at the head of the troops.
\verse “When you approach a city to fight against it, \textit{you must offer it peace}.\lebnote{Literally “you should call it for peace”}
\verse \textit{And then}\lebnote{Literally “And it will happen”} if \textit{they accept your terms of peace}\lebnote{Literally “if peace they reply to you”} and \textit{they surrender to you},\lebnote{Literally “they open to you”} \textit{and then}\lebnote{Literally “and it will happen”} all the people \textit{inhabiting it}\lebnote{Literally “being found in it”} shall be forced labor for you, and they shall serve you.
\verse But if they \textit{do not accept your terms of peace}\lebnote{Literally “not they make peace with you”} and they want to make war with you, then you shall lay siege against it.\lebnote{That is, the city}
\verse And Adonai your God will give it into your hand, and you shall kill all its males with the \textit{edge}\lebnote{Literally “mouth”} of the sword.
\verse Only the women and the little children and the domestic animals\lebnote{Hebrew “animal”} and all that shall be in the city, all of its spoil you may loot for yourselves, and you may enjoy the spoil of your enemies that Adonai you God has given to you.
\verse Thus you shall do to all the far\lebnote{Or “distant”} cities from you, which are not from the cities of these nations located \textit{nearby}.\lebnote{Literally “here”}
\verse But from the cities of these peoples that Adonai your God is giving to you as an inheritance, you shall not let anything live that breathes.\lebnote{Or “is alive”}
\verse Rather, you shall utterly destroy them, the Hittites\lebnote{Hebrew “Hittite”} and the Amorites,\lebnote{Hebrew “Amorite”} the Canaanites\lebnote{Hebrew “Canaanite”} and the Perizzites,\lebnote{Hebrew “Perizzite”} the Hivites,\lebnote{Hebrew “Hivite”} and the Jebusites,\lebnote{Hebrew “Jebusite”} just as Adonai your God has commanded you,
\verse so that they may not teach you to do like all their detestable things that they do for their gods and thereby you sin against Adonai your God.
\verse “If\lebnote{Or “when”} you besiege a town for many days to make war against it in order to seize it, you shall not destroy its trees by wielding an ax against them,\lebnote{Hebrew “it”} for you may eat from them,\lebnote{Hebrew “it”} and so you must not cut them\lebnote{Hebrew “it”} down. Are the trees of the field humans that they should come in siege \textit{against you}?\lebnote{Literally “from your face”}
\verse Only\lebnote{Or “But”} the trees\lebnote{Hebrew “tree”} that you know \textit{are not fruit trees}\lebnote{Literally “not a tree of food”} you may destroy and you may cut down, and you may build siege works against that city that is making war with you \textit{until it falls}.”\lebnote{Literally “it to fall”}
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 21
\verse “If someone slain is found in the land that Adonai your God is giving to you to take possession of it and is lying in the field, and it is not known who \textit{killed him},\lebnote{Literally “struck/smote him”}
\verse then your elders and your judges shall go out and shall measure the distance to the cities that are around the slain one.
\verse \textit{And then}\lebnote{Literally “And it will happen”} the nearest city to the slain one, the elders of that city shall take a heifer of the herd that has not been worked with in the field, that has not pulled a yoke,
\verse and the elders of that city shall bring the heifer down to a \textit{wadi that flows with water all year}\lebnote{Literally “an ever-flowing wadi”} and that has not been plowed and has not been sown; then \textit{there they shall break the neck of the heifer in the wadi}.\lebnote{Literally “they shall break there the neck with respect to the heifer in the ever-flowing wadi”; the Hebrew verb carries the meaning “to break the neck of”}
\verse Then the priests, the descendants\lebnote{Or “sons”} of Levi, shall come near, for Adonai your God has chosen them to bless in the name of Adonai, and every legal dispute and every case of assault will be \textit{subject to their ruling}.\lebnote{Literally “on their mouth”}
\verse And all of the elders of that city nearest to the slain person shall wash their hands over the heifer with the broken neck in the wadi.\lebnote{A valley that is dry most of the year, but contains a stream during the rainy season}
\verse And they shall declare, and they shall say, ‘Our hands did not shed this blood, and our eyes did not see what was done.
\verse Forgive your people, Israel, whom you redeemed, Adonai, and do not \textit{allow}\lebnote{Literally “place/put”} the guilt of innocent blood in the midst of your people Israel, and let them be forgiven with regard to blood.’
\verse And so you shall purge the innocent blood from your midst, because you must do the right thing in the eyes of Adonai.
\verse “When you go out for battle against your enemies, and Adonai your God gives them into your hand, and you lead the captives\lebnote{Hebrew “his captive” but singular pronoun refers to the many captives taken with plural sense} away,
\verse and you see among the captives\lebnote{Hebrew “captive”} a woman beautiful in appearance, and you become attached to her and you want to take her as a wife,
\verse then you shall bring her into your household, and she shall shave her head, and she shall trim her nails.
\verse And she shall remove the clothing of her captivity from her, and she shall remain in your house, and she shall mourn her father and her mother \textit{a full month},\lebnote{Literally “a month of days”} and after this \textit{you may have sex with her},\lebnote{Literally “you may go into her”} and you may marry her, and she may \textit{become your wife}.\lebnote{Literally “become for you as wife”}
\verse And then if you do not take delight in her, then you shall let her go \textit{to do whatever she wants},\lebnote{Literally “according to her desire/soul”} but you shall not treat her as a slave, since you have dishonored\lebnote{Or “humbled”} her.
\verse “If a man has two wives, and the one is loved and the other one is disliked and the one loved and the one that is disliked have borne for him sons, if it happens that the firstborn son \textit{belongs to the one that is disliked},\lebnote{Literally “is to the wife who is hated”}
\verse nevertheless \textit{it will be the case that}\lebnote{Literally “it will happen”} on the day of bestowing his inheritance upon his sons, he will not be allowed to treat as the firstborn son the son of the beloved wife \textit{in preference to}\lebnote{Literally “over the faces of”} the son of the disliked wife, who is the firstborn son.
\verse But he shall acknowledge the firstborn son of the disliked wife \textit{by giving}\lebnote{Literally “to give”} him a double portion of \textit{all that he has},\lebnote{Literally “all that is found for him”} for he is the firstfruit of his vigor;\lebnote{Or “the beginning of his strength”} to him is the legal claim of the birthright.\lebnote{Or “the just claim of the firstborn”}
\verse “\textit{If a man has a stubborn and rebellious son}\lebnote{Literally “If shall be for a man, a son stubborn and rebellious”} who \textit{does not listen to}\lebnote{Literally “and there is no listening/obedience”} the voice of his father and to the voice of his mother, and they discipline him, and he does not obey\lebnote{Or “listen to”} them,
\verse then his father and his mother shall take hold of him, and they shall bring him out to the elders of his city and to the gate of his \textit{town},\lebnote{Literally “place”}
\verse and they shall say to the elders of his city, ‘This our son is stubborn and rebellious; \textit{he does not obey us},\lebnote{Literally “there is no listening to our voice”} and he is a glutton and a drunkard.’
\verse Then all the men of his city shall stone him with stones and let him die; and so you shall purge the evil from your midst, and all of Israel will hear, and they will fear.
\verse “And \textit{if a man commits a sin punishable by death},\lebnote{Literally “when shall be against a man a sin of judgment of death”} and so he is put to death and you hang him on a tree,
\verse his dead body shall not hang on the tree, but certainly you shall bury him on that day, for cursed by God is one that is being hung; so you shall not defile your land\lebnote{Or “ground”} that Adonai your God is giving to you as an inheritance.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 22
\verse “You shall not watch the ox of your neighbor or his sheep or goat straying and ignore them; certainly you shall return them to your neighbor.
\verse And if your countryman\lebnote{Or “brother”} is not near you or you do not know \textit{who he is},\lebnote{Literally “him”} then you shall bring it \textit{to your household},\lebnote{Literally “to the midst of your house”} and it shall be\lebnote{Or “remain”} with you \textit{until your countryman seeks after it},\lebnote{Literally “the seeking of your countryman after it”} and you shall return it to him.
\verse And thus also you shall do regarding\lebnote{Or “to”} his donkey, and thus you shall do concerning\lebnote{Or “to”} his garment, and so you shall do with respect to\lebnote{Or “to”} all of the lost property of your countryman\lebnote{Or “brother”} that is lost from him and you find it; you are not allowed to withhold help.
\verse “You shall not see the donkey of your neighbor or his ox fallen on the road and you ignore them; certainly you must help them get up along with him.
\verse “The apparel of a man shall not be put on\lebnote{Literally “on”} a woman, and a man shall not wear the clothing of a woman, because everyone who does these things is detestable to Adonai your God.
\verse “If a bird’s nest is found \textit{before you}\lebnote{Literally “before your face”} on the road in any tree or on the ground, and there are chicks or eggs, and the mother is lying down on the chicks or the eggs, you shall not take the mother along with the young;
\verse you shall certainly let the mother go, but you may take the young for yourselves;\lebnote{Hebrew “for you”} do this \textit{so that it may go well}\lebnote{Literally “he/it is good”} for you and \textit{you may live long in the land}.\lebnote{Literally “and you may make long/extend days”}
\verse “When you build a new house then\lebnote{Hebrew “and”} you shall make a parapet wall\lebnote{Or “fence/rail”} for your roof, so that you will not bring bloodguilt on your house \textit{if anyone should fall from it}.\lebnote{Literally “if should fall the falling from it”}
\verse “You shall not sow your vineyard with differing kinds of seed, so that you shall not forfeit \textit{the whole harvest},\lebnote{Literally “the entire fullness”} both the seed that you sowed and the yield of the vineyard.
\verse “You shall not plow with an ox and with a donkey yoked together.
\verse “You shall not wear woven material made of wool and linen mixed together.
\verse “You shall make tassels for yourselves\lebnote{Hebrew “for you”} on the four corners of your clothing with which you cover yourself.
\verse “If a man takes a woman and \textit{he has sex with her},\lebnote{Literally “and he goes unto her”} but he then \textit{dislikes her},\lebnote{Literally “hates her”}
\verse and \textit{he accuses her falsely},\lebnote{Literally “he puts to her deeds of words”} and \textit{he defames her},\lebnote{Literally “he brings forth against her a bad name”} and he says ‘This woman I took and I lay with her and \textit{I discovered that she was not a virgin},’\lebnote{Literally “I found not with her virginity”}
\verse then in defense the father of the young woman shall take, along with her mother, and together they must bring out the evidence of the virginity of the young woman to display it to the elders of the city \textit{at the city gate}.\lebnote{Literally “toward the city gate”}
\verse And then the father of the young woman shall say to the elders, ‘I gave my daughter to this man as wife, but he now \textit{dislikes}\lebnote{Literally “hates”} her,
\verse and now look \textit{he has accused her falsely},\lebnote{Literally “he put deeds of words”} saying, “I did not find \textit{your daughter a virgin},”\lebnote{Literally “to your daughter virginity”} but here is evidence of the virginity of my daughter’; and they shall spread the cloth out \textit{before}\lebnote{Literally “to the face of”} the elders of the city.
\verse Then the elders of that city shall take the man, and they shall discipline him.
\verse Then they shall fine him a hundred shekels of silver, and they shall give them to the father of the young woman, for \textit{he defamed an Israelite young woman},\lebnote{Literally “for he brought out a bad name against the virgin of Israel”} and \textit{she shall become his wife};\lebnote{Literally “and for him she shall become as wife”} he will not be allowed \textit{to divorce her}\lebnote{Literally “to send her out”} all his days.
\verse “But if \textit{this charge}\lebnote{Literally “the thing”} was true, \textit{and the signs of virginity were not found}\lebnote{Literally “and they were not found evidence of virginity”} for the young woman,
\verse and then they shall bring out the young woman to the doorway of the house of her father, and the men of her city shall stone her with stones, and she shall die, because she did a disgraceful thing in Israel \textit{by playing the harlot}\lebnote{Literally “to play the harlot/to prostitute herself”} in the house of her father, and so you shall purge the evil from your midst.
\verse “If a man is found lying \textit{with a married woman},\lebnote{Literally “with a woman, a young woman of a husband”} then they shall both die; \textit{both of them},\lebnote{Literally “also/even the two of them”} the man who lay with the woman and the woman also, so you shall purge the evil from Israel.
\verse “If it happens that a young woman, a virgin, is engaged to a man, and a man finds her in the town and lies with her,
\verse then you shall bring out \textit{both of them}\lebnote{Literally “the two of them”} to the gate of that city, and you shall stone them with stones so that they shall die, the young woman because she did not cry out in the town, and the man because\lebnote{Literally “because of the fact that”} he violated his neighbor’s wife; and so you shall purge the evil from your midst.
\verse “But if the man finds the young engaged woman in the field and the man overpowers her and \textit{he has sex with her},\lebnote{Literally “he lies down with her”} then the man only\lebnote{Or “alone”} must die who lay\lebnote{Or “slept”} with her.
\verse But you shall not do anything\lebnote{Hebrew “a thing”} to the young woman, for there is not reckoned against\lebnote{Hebrew “to”} the young woman \textit{a sin deserving death};\lebnote{Literally “a sin of death”} \textit{it is similar to when}\lebnote{Literally “for as that”} a man rises up against his neighbor and murders him, \textit{a fellow human being},\lebnote{Literally “a soul/individual person”} just so is this \textit{case},\lebnote{Literally “thing/matter”}
\verse for he found her in the field, the engaged young woman cried out, but there was no\lebnote{Hebrew “there was not”} rescuer \textit{to help her}.\lebnote{Literally “for her”}
\verse “If a man finds a young woman, a virgin who is not engaged, and he seizes her and \textit{he has sex with her}\lebnote{Literally “he lies with her”} and they are caught,
\verse then \textit{the man who lay with her}\lebnote{Literally “the man the one lying with her”} shall give to the father of the young woman fifty shekels of silver, and she shall become \textit{his wife}\lebnote{Literally “for/as a wife”} \textit{because}\lebnote{Literally “in place of”} he violated her, and he is not allowed to divorce her \textit{during his lifetime}.\lebnote{Literally “all of his days”}
\verse \lebnote{Deuteronomy 22:30–23:25 in the English Bible is 23:1–26 in the Hebrew Bible} A man may not take the wife of his father, and so \textit{he may not dishonor his father}.\lebnote{Literally “not he may reveal the skirt of the garment of his father”}
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 23
\verse “No man \textit{with crushed testicles}\lebnote{Literally “bruised of crushing”} or whose \textit{male organ is cut off}\lebnote{Literally “cut off of male organ”} may come into the assembly of Adonai.
\verse An illegitimate child may not come into the assembly of Adonai; even to the tenth generation none \textit{of his descendants}\lebnote{Literally “to him”} may come into the assembly of Adonai.
\verse An Ammonite or a Moabite may not come into the assembly of Adonai; even to the tenth generation none \textit{of his descendants}\lebnote{Literally “to him”} may come into the assembly of Adonai \textit{forever},\lebnote{Literally “until eternity”}
\verse \textit{because}\lebnote{Literally “because of the event when”} they did not come to meet you with food and with water \textit{when you came out of Egypt},\lebnote{Literally “at your going out of Egypt”} and also \textit{because}\lebnote{Literally “because of the event when”} they\lebnote{Hebrew “he,” but with plural meaning} hired Balaam, son of Beor, from Pethor, in Aram Naharaim\lebnote{Or “Mesopotamia” = “between the rivers”} to act against you to curse you.
\verse But Adonai your God was not willing to listen to Balaam, and Adonai your God turned the curse into a blessing for you, because Adonai your God loved you.
\verse You shall not promote their welfare or their prosperity all your days \textit{forever}.\lebnote{Literally “until eternity”}
\verse “You shall not abhor an Edomite, because he is your brother; you shall not abhor an Egyptian because you were an alien in his land.
\verse The children\lebnote{Or “sons”} that are born to them in the third generation may come \textit{representing them}\lebnote{Literally “for them” or possibly “become members of”} in the assembly of Adonai.
\verse “If\lebnote{Or “when”} you go out to encamp against your enemies, then you shall guard against doing anything evil.
\verse “If there is among you a man that is not clean because of a seminal emission \textit{during the night},\lebnote{Literally “of the night”} he shall go outside the camp; he shall not come within the camp.
\verse \textit{And then}\lebnote{Literally “And it will happen/come about that”} toward the \textit{coming}\lebnote{Literally “turning”} of the evening, he shall bathe with\lebnote{Hebrew “in” but this is instrumental use of this preposition} water, and at the going down\lebnote{The verb can mean “coming/going” and here it means “going”} of the sun, he may come to the midst of the camp.
\verse “And there shall be for you a designated place outside the camp; \textit{and you shall go there to relieve yourself},\lebnote{Literally “and you shall go out there, outside to relieve yourself”}
\verse and a digging tool\lebnote{Or “spade”} shall be included in addition to your other utensils\lebnote{Hebrew “utensil”} for yourself; \textit{and then}\lebnote{Literally “and it will happen”} \textit{when you relieve yourself}\lebnote{Literally “at your sitting”} outside the camp you shall dig with it, and then you shall turn, and you shall cover your excrement.
\verse For Adonai your God is walking about in the midst of your camp to deliver you and \textit{to hand your enemies over to you before you},\lebnote{Literally “to give your enemies to the face of you”} and so let your camp be holy, so that he shall not see in it \textit{anything indecent},\lebnote{Literally “nakedness of a thing”} and he shall turn away \textit{from going with you}.\lebnote{Literally “from behind/after you”}
\verse “And you shall not hand over a slave to his master who has escaped and fled to you from his master.
\verse He shall reside with you in your midst in the place that he chooses in one of \textit{your towns wherever he pleases};\lebnote{Literally “in your gates good for him”} you shall not oppress him.
\verse “No woman \textit{of Israel}\lebnote{Literally “from the daughters of Israel”} shall be a temple prostitute, and no man \textit{of Israel}\lebnote{Literally “from the sons of Israel”} shall be a male shrine prostitute.
\verse You may not bring the \textit{hire}\lebnote{Literally “gift”} of a prostitute or \textit{the earnings of a male prostitute}\lebnote{Literally “the earnings of a dog”} into the house of Adonai your God, for any vow offerings, because \textit{both}\lebnote{Literally “both, the two of them”} are a detestable thing to Adonai your God.
\verse “\textit{You shall not charge your brother interest on money},\lebnote{Literally “You shall not lend on interest to your brother, interest of money”} interest on\lebnote{Hebrew “of”} food, or interest on\lebnote{Hebrew “of”} anything that one could lend\lebnote{Or “can lend”} on interest.
\verse You may lend on interest to the foreigner, but to your countryman\lebnote{Or “brother”} you may not lend on interest, so that Adonai your God may bless you \textit{in all your undertakings}\lebnote{Literally “in all the sendings out of your hand”} in \textit{the land where you are going},\lebnote{Literally “the land that you are going there”} \textit{in order to take possession of it}.\lebnote{Literally “to take possession”}
\verse “\textit{When you make a vow}\lebnote{Literally “When/if you vow a vow”} to Adonai your God, you shall not postpone \textit{fulfillment of it},\lebnote{Literally “to fulfill it”} for certainly Adonai your God shall require it from you and if postponed \textit{you will incur guilt}.\lebnote{Literally “it shall become against you as sin”}
\verse And \textit{if you refrain from vowing},\lebnote{Literally “if you refrain to make a vow”} \textit{you shall not incur guilt}.\lebnote{Literally “it shall not become against you as sin”}
\verse The utterance of your lips \textit{you must perform diligently}\lebnote{Literally “you must observe and you shall do”} just as you have vowed freely to Adonai your God whatever it was that you promised\lebnote{Or “spoke”} with your mouth.
\verse “When\lebnote{Or “If”} you come into the vineyard of your neighbor, then you may eat grapes \textit{as you please}\lebnote{Literally “according to your desire”} and \textit{until you are full},\lebnote{Literally “to your satiation”} but you shall not put any into your container.
\verse “When\lebnote{Or “If”} you come into the standing grain of your neighbor, then you may pluck ears with your hand, but you may not \textit{swing}\lebnote{Literally “wave”} a sickle among the standing grain of your neighbor.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 24
\verse “When\lebnote{Or “If”} a man takes a wife and he marries her \textit{and then}\lebnote{Literally “and it will happen”} \textit{she does not please him},\lebnote{Literally “if not she finds favor in his eyes”} because he found \textit{something objectionable}\lebnote{Literally “shameful/repulsive thing”} and writes her a letter of divorce and puts it in her hand and sends her away from his house,
\verse and she goes from his house, and she goes out and becomes a wife \textit{for another man},\lebnote{Literally “for a man other”}
\verse and then the second man dislikes her and he writes her a letter of divorce and places it into her hand and sends her from his house, or if the second man dies who took her \textit{to himself}\lebnote{Literally “to for him”} as a wife,
\verse her first husband who sent her away is not allowed \textit{to take her again}\lebnote{Literally “to return to take her”} to become a wife to him after she has \textit{been defiled},\lebnote{Literally “become unclean”} for that is a detestable thing \textit{before}\lebnote{Literally “to the face of”} Adonai, and so you shall not mislead into sin the land that Adonai your God is giving to you as an inheritance.
\verse “When\lebnote{Or “If”} a man takes a new wife he shall not go out with the army, and \textit{he shall not be obligated with anything};\lebnote{Literally “he shall not come across come over upon him to anything”} he shall be free from obligation, \textit{to stay at home}\lebnote{Literally “for his house”} for one year, and he shall bring joy to his wife that he took.
\verse “A person\lebnote{Hebrew “he”} shall not take\lebnote{Or “require”} a pair of millstones or an upper millstone, for \textit{he is taking necessities of life as a pledge}.\lebnote{Literally “for a life he is taking as a pledge”}
\verse “If a man is \textit{caught}\lebnote{Literally “found”} kidnapping somebody from among his countrymen,\lebnote{Or “brothers”} the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and he treats him as a slave or he sells him, then that kidnapper shall die, and so you shall purge the evil \textit{from among you}.\lebnote{Literally “from your midst”}
\verse Be watchful\lebnote{Or “Be careful”} \textit{with respect to}\lebnote{Literally “against”} an outbreak of any infectious skin disease, by being very careful and by acting\lebnote{Or “doing”} according to all that the priests and the Levites have instructed you, just as I have commanded them, \textit{so you shall diligently observe}.\lebnote{Literally “so you shall observe to do”}
\verse So remember what Adonai your God did to Miriam on the journey \textit{when you went out from Egypt}.\lebnote{Literally “in you to go out from Egypt”}
\verse “When you make a loan to your neighbor, a loan of any kind, you shall not go into his house \textit{to take his pledge}.\lebnote{Literally “to pledge with respect to his pledge”}
\verse You shall wait outside, and the man to whom you are lending, he shall bring the pledge outside to you.
\verse And if he is a needy man, you shall not sleep in his pledge.\lebnote{“His pledge” refers to “a garment given as pledge”}
\verse You shall certainly return the pledge to him \textit{as the sun sets},\lebnote{Literally “as/at the moment of the going out of the sun”} so that he may sleep in his cloak and may bless you, and it shall be considered righteousness \textit{on your behalf}\lebnote{Literally “to you”} \textit{before}\lebnote{Literally “to the face of”} Adonai your God.
\verse “You shall not exploit a hired worker, who is needy and poor, from among your fellow men or from among your aliens\lebnote{Hebrew “alien”} who are in your land and in your \textit{towns}.\lebnote{Literally “gates”}
\verse On his day you shall give his wage, and the sun shall not go \textit{down},\lebnote{Literally “over him”} because he is poor and \textit{his life depends on it};\lebnote{Literally “and to him it is a lifting up with respect to his life/soul”} do this so that he does not cry out against you to Adonai, \textit{and you incur guilt}.\lebnote{Literally “and it becomes against you as sin}
\verse “Fathers shall not be put to death because of their children, and children shall not be put to death because of their fathers; each one shall be put to death for his own sin.
\verse You shall not subvert the rights of an alien or an orphan, and you shall not take as pledge the garment of a widow.
\verse And you shall remember that you were a slave in Egypt and that Adonai your God redeemed you from there; therefore I am commanding you to do this commandment.
\verse “When\lebnote{Or “If”} you reap your harvest in your field and you forget a sheaf in the field, you shall not return to get it, for it shall be for the alien, for the orphan, and for the widow, so that Adonai your God may bless you in all the work of your hands.
\verse When you beat off the fruit of your olive trees you shall not search through the branches afterward, for it\lebnote{That is, what is left} shall be for the alien, for the orphan, and for the widow.
\verse When you harvest grapes, you shall not glean your vineyards \textit{again};\lebnote{Literally “behind/after you”} it\lebnote{That is, what is left} shall be for the alien, for the orphan, and for the widow.
\verse And you shall remember that you were a slave in the land of Egypt, therefore I am commanding you to do this thing.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 25
\verse “When a legal dispute \textit{takes place}\lebnote{Literally “shall be”} between men and they come near to the court, and the judges judge with respect to them, then they shall declare the righteous to be in the right and they shall condemn the wicked,
\verse then it will happen if the guilty one \textit{deserves beating},\lebnote{Literally “is a son of beating”} then the judge shall make him lie, and he shall beat him \textit{before him},\lebnote{Literally “to the face of him”} \textit{according to}\lebnote{Literally “as”} \textit{the prescribed number of lashes proportionate to the offense}.\lebnote{Literally “as is sufficient/necessary, with respect to number, for his wickedness/wicked behavior”}
\verse He may beat him with forty lashes, and he shall not do more than these, so that he will not beat more in addition to these many blows,\lebnote{Hebrew “blow”} and your countryman\lebnote{Or “brother”} would be\lebnote{Hebrew “is”} degraded before your eyes.
\verse “You shall not muzzle an ox \textit{when he is threshing}.\lebnote{Literally “at his threshing”}
\verse “When brothers dwell together and one of them dies and has no son, the wife of the deceased shall not become the wife of a \textit{man of another family};\lebnote{Others: “strange man,” “man outside the family” (NEB)} her brother-in-law \textit{shall have sex with her},\lebnote{Literally “shall go to her”} and he shall take her \textit{to himself}\lebnote{Literally “to/for him”} as a wife, and he shall perform his duty as a brother-in-law with respect to her.
\verse And then the firstborn that she bears \textit{shall represent his dead brother},\lebnote{Literally “he shall stand upon the name of his brother the deceased”} so that his name is not blotted out from Israel.
\verse But if the man does not want to take his sister-in-law, then his sister-in-law shall go up to the gate, to the elders, and she shall say, ‘My brother-in-law refused \textit{to perpetuate his brother’s name}\lebnote{Literally “to cause a name to stand for his brother”} in Israel, for he is not willing \textit{to marry me}.’\lebnote{Literally “to consummate the marriage with the widow of a his brother”}
\verse Then the elders of his town shall summon him and speak to him, and if he persists and says, ‘\textit{I do not desire to}\lebnote{Literally “I do not delight in”} marry her’
\verse then his sister-in-law shall go near him before the eyes of the elders, and she shall pull off his sandal from his foot, and she shall spit in his face, and she shall \textit{declare}\lebnote{Literally “respond/answer”} and she shall say, ‘This is how it is done to the man who does not build the house of his brother.’
\verse And his \textit{family}\lebnote{Literally “name”} shall be called in Israel, ‘The house where the sandal was pulled off.’\lebnote{Or “the house of the man whose sandal was pulled off”}
\verse “If a man and his brother fight each other and the wife of the one man comes near to rescue her husband from the hand of his attacker and she stretches out her hand and she seizes his genitals,
\verse then you shall cut off her hand; your eye shall not take pity.
\verse “There shall not be \textit{for your use}\lebnote{Literally “to you”} in your bag \textit{two kinds of stone weights, a large one and a small one}.\lebnote{Literally “stone and stone large and small”}
\verse There shall not be in your house \textit{for your use}\lebnote{Literally “for you”} \textit{two kinds of measures}.\lebnote{Literally “ephah and ephah large and small”}
\verse Rather a full and honest weight shall be \textit{for your use};\lebnote{Literally “for you”} there shall be for you a full and honest \textit{measure},\lebnote{Literally “an ephah”} so that your days on the land that Adonai your God is giving to you may be long.
\verse For detestable to\lebnote{Or “for”} Adonai your God is everyone who is doing such things,\lebnote{Or “are all whoare doing”} everyone who is acting\lebnote{Or “all whoare acting”} dishonestly.
\verse “Remember what Amalek did to you on the journey when\lebnote{Or “at/in”} you went out from Egypt,
\verse that he met you on the journey and attacked you, all those lagging behind you and when you were weary and worn out, and he did not fear God.
\verse \textit{And when}\lebnote{Literally “And it will happen when”} Adonai your God gives rest to you from all your enemies from around about you in the land that Adonai your God is giving to you as an inheritance to take possession of it, you shall blot out the remembrance of Amalek from under the heavens; you shall not forget!”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 26
\verse “\textit{And then}\lebnote{Literally “And it shall happen”} when you come to the land that Adonai your God is giving to you as an inheritance, and you take possession of it and you settle in it,
\verse then you shall take from the firstfruit of all the fruit of the ground that you harvest from your land that Adonai your God is giving to you, and you shall put it in a basket, and you shall go to the place that Adonai your God will choose to make his name to dwell there.
\verse And you shall go to the priest who is in office in those days, and you shall say, ‘I declare \textit{today}\lebnote{Literally “the day”} to Adonai your God that I have come into the land that Adonai swore to our ancestors\lebnote{Or “fathers”} to give to us.’
\verse Then the priest takes the basket from your hand and places it \textit{before}\lebnote{Literally “to the face of”} the altar of Adonai your God.
\verse And \textit{you shall declare}\lebnote{Literally “answer”} and you shall say \textit{before}\lebnote{Literally “to the face of”} your God, ‘My ancestor\lebnote{Or “father”} was a wandering Aramean, and he went down to Egypt, and there he dwelt as an alien \textit{few in number},\lebnote{Literally “in people only a few”} and there he became a great nation, mighty and numerous.
\verse And the Egyptians treated us badly, and they oppressed us and imposed on us hard labor.
\verse And we cried to Adonai, the God of our ancestors,\lebnote{Or “fathers”} and Adonai heard our voice and saw our affliction and our toil and our oppression.
\verse And Adonai brought us out from Egypt with a strong hand and an outstretched arm and with great terror and with signs and with wonders.
\verse And he brought us to this place and gave to us this land, a land flowing with milk and honey.
\verse And now, look, I am bringing\lebnote{Hebrew “I bring”} the firstfruit of the fruit of the ground that you gave to me, Adonai,’ and you shall place it \textit{before}\lebnote{Literally “to the face of”} Adonai your God, and you shall bow down \textit{before}\lebnote{Literally “to the face of”} Adonai your God.
\verse And you shall celebrate with all of the bounty that Adonai your God gave to you and to your family, you and the Levite and the alien who is in your midst.
\verse “When you are finished \textit{giving a tithe},\lebnote{Literally “to give a tenth”} all of the tithe of your produce in the third year, the year of the tithe, then you shall give to the Levite, to the alien, to the orphan, and to the widow, so that they may eat in your towns and they may be satisfied.
\verse And you shall say \textit{before}\lebnote{Literally “to the face of”} Adonai your God, ‘I have removed the sacred portion from the\lebnote{Or “my” house since the definite article can show possession here} house and, moreover, I have given it to the Levite and to the alien and to the orphan and to the widow according to all your commandment that you commanded me; I have not transgressed any of your commandments, and I have not forgotten any of them.
\verse I have not eaten during my time of mourning, and I have not removed anything from it while being unclean, and I have not offered \textit{any of it}\lebnote{Literally “from it”} to someone who has died. I have listened to the voice of Adonai my God; I have done all that you commanded me to do.
\verse Look down from the dwelling place of your holiness, from heaven, and bless your people Israel, and the land that you have given to us, as you swore to our ancestors,\lebnote{Or “our fathers”} a land flowing with milk and honey.’
\verse “This day Adonai your God is commanding you to do these rules and regulations, \textit{and you must observe them diligently}\lebnote{Literally “and you shall observe and you shall do them”} with all your heart and with all your soul.\lebnote{Or “inner self”}
\verse Adonai you have declared \textit{today}\lebnote{Literally “the day”} to be for you as your God, and to go\lebnote{Or “to walk”} in his ways and to observe his rules and his commandments and his regulations and to listen to his voice.
\verse And Adonai has declared you \textit{today}\lebnote{Literally “the day”} to be for him as a people, a treasured possession, as he \textit{promised}\lebnote{Literally “spoke”} to you, and that you are to observe all his commandments,
\verse \textit{and that he then will set you}\lebnote{Literally “and to set you”} high above all the nations that he has made for his praise and \textit{for fame}\lebnote{Literally “for a name”} and for honor and \textit{for you to be a holy people}\lebnote{Literally “for you to be/your being a people holy”} to Adonai your God, as he \textit{promised}.”\lebnote{Literally “spoke”}
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 27
\verseWithHeading{Ceremonies} Then Moses and the elders of Israel charged the people, \textit{saying},\lebnote{Literally “to say”} “Keep all of the commandment that I am commanding you \textit{today}.\lebnote{Literally “the day”}
\verse And then on the day that you cross the Jordan to the land that Adonai your God is giving to you, then you shall set up \textit{for yourselves}\lebnote{Literally “for you”} large stones, and you shall paint\lebnote{Or “coat/cover them”} them with lime,
\verse and you shall write on them all the words of this law at your crossing, so that you may come into the land that Adonai your God is giving to you, a land flowing with milk and honey, as Adonai, the God of your ancestors,\lebnote{Or “fathers”} \textit{promised}\lebnote{Literally “spoke”} to you.
\verse And \textit{when you cross the Jordan},\lebnote{Literally “at/in your crossing the Jordan”} you shall set up these stones that I am commanding you about \textit{today}\lebnote{Literally “the day”} on Mount Ebal, and you shall paint\lebnote{Or “coat/cover”} them with lime.
\verse And you shall build an altar there for Adonai your God, an altar of stone, but\lebnote{Or “and”} \textit{you shall not use an iron tool to shape the stones}.\lebnote{Literally “and you shall not wave to and fro over them an iron tool”}
\verse You must build the altar of your God with unhewn stones, and you shall sacrifice on it burnt offerings to Adonai your God.
\verse And you shall sacrifice fellowship offerings, and you shall eat them there, and you shall rejoice \textit{before}\lebnote{Literally “to the face of”} Adonai your God.
\verse You shall write on the stone all of the words of this law very clearly.”
\verse Then Moses and the priests, the Levites, spoke to all Israel, saying,\lebnote{Literally “to say”} “Be silent and hear, Israel, for this day you have become \textit{a people}\lebnote{Literally “for a people”} for Adonai your God.
\verse And listen to the voice of Adonai your God, and observe his commandments and his rules that I am commanding you \textit{today}.”\lebnote{Literally “the day”}
\verseWithHeading{Blessings and Curses} And Moses charged the people on that day, \textit{saying},\lebnote{Literally “to say”}
\verse “These tribes shall stand on Mount Gerizim to bless the people \textit{when you cross}\lebnote{Literally “at/in crossing your”} the Jordan: Simeon and Levi and Judah and Issachar and Joseph and Benjamin.
\verse And these shall stand on Mount Ebal for delivering the curse: Reuben, Gad and Asher and Zebulun, Dan and Naphtali.
\verse And \textit{the Levites shall declare},\lebnote{Literally “they shall answer the Levites”} and they shall say to each man of Israel with a loud voice,
\verse ‘Cursed be the man that makes a divine image or a cast image, which is a detestable thing for Adonai, the work of the hand of a skilled craftsman, and then sets it in a hiding place.’\lebnote{Or “a secret place”} And \textit{all the people shall respond},\lebnote{Literally “and they shall answer all of the people”} ‘Amen.’
\verse ‘Cursed be the one who dishonors his father or his mother.’ And all of the people shall say, ‘Amen.’
\verse ‘Cursed be the one who moves the boundary marker of his neighbor.’ And all the people shall say, ‘Amen.’
\verse ‘Cursed be the one who misleads a blind person on the road.’ And all the people shall say, ‘Amen.’
\verse ‘Cursed be the one who deprives the alien, the orphan, and the widow of justice.’ And all the people shall say, ‘Amen.’
\verse ‘Cursed be the one who lies with the wife of his father, because \textit{he has dishonored his father’s bed}.’\lebnote{Literally “because he uncovered the hem of his father”} And all the people shall say, ‘Amen.’
\verse ‘Cursed be the one who lies with any kind of animal.’ And all the people shall say, ‘Amen.’
\verse ‘Cursed be the one who lies with his sister, the daughter of his father or the daughter of his mother.’ And all the people shall say, ‘Amen.’
\verse ‘Cursed be the one who lies with his mother-in-law.’ And all the people shall say, ‘Amen.’
\verse ‘Cursed be the one who strikes down his neighbor in secret.’ And all the people shall say, ‘Amen.’
\verse ‘Cursed be the one who takes a bribe \textit{to murder an innocent person}.’\lebnote{Literally “to strike down an individual person in blood”} And all the people shall say, ‘Amen.’
\verse ‘Cursed be the one \textit{who does not keep}\lebnote{Literally “who not keeps”} the words of this law, to observe them.’ And all the people shall say, ‘Amen.’ ”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 28
\verse “And it will happen that if you indeed listen to the voice of Adonai your God, \textit{to diligently observe}\lebnote{Literally “to observe to do”} all his commandments that I am commanding you \textit{today},\lebnote{Literally “the day”} then Adonai your God will set you above all the nations of the earth.
\verse And all of these blessings shall come upon you, and \textit{they shall have an effect on you}\lebnote{Literally “they shall overtake you”} if you listen to the voice of Adonai your God:
\verse “You will be blessed in the city, and you will be blessed in the field.
\verse “Blessed will be the fruit of your womb and the fruit of your ground and the fruit of your livestock, \textit{the calf of your cattle and the lambs of your flock}.\lebnote{Literally “what is dropped of your cattle and the offspring of your flocks”}
\verse “Blessed will be your basket and your kneading trough.
\verse “Blessed will you be \textit{when you come in and blessed will you be when you go out}.\lebnote{Literally “in/at your coming and you will be blessed in/at your going out”}
\verse “Adonai will cause your enemies \textit{who rise up against you to be defeated before you};\lebnote{Literally “the ones rising against you being defeated to the face of you”} on one road\lebnote{Or “way”} they shall come out against you, but on seven roads\lebnote{Or “ways”} they shall flee \textit{before you}.\lebnote{Literally “to the face of you”}
\verse Adonai will command \textit{concerning you}\lebnote{Literally “with you”} the blessing to be in your barns and \textit{in all your endeavors};\lebnote{Literally “in all of the sending forth of your hand”} and he will bless you in the land that Adonai your God is giving to you.
\verse Adonai will establish you for \textit{himself}\lebnote{Literally “to/for him”} as a holy people as he has sworn to you, if you keep the commandments of Adonai your God and you walk in his ways.
\verse And all of the peoples of the earth shall see that \textit{by the name of Adonai you are called},\lebnote{Literally “that the name of Adonai is called/assigned upon you”} and \textit{they shall fear you}.\lebnote{Literally “they shall be afraid/fearful from you”}
\verse And \textit{Adonai will make you successful and prosperous},\lebnote{Literally “Adonai will cause for you goodness”} in the fruit of your womb and on the land that Adonai swore to your ancestors\lebnote{Or “fathers”} to give to you.
\verse Adonai shall open for you his \textit{rich}\lebnote{Literally “good”} storehouse, even the heavens, to give the rain for\lebnote{Hebrew “of”} your land in its time and to bless all of the work of your hand, and you will lend to many nations; you will not borrow from them.
\verse And Adonai shall make you\lebnote{Literally “place/set you up”} as head and not the tail, and you shall be only at the top of the nations, and you shall not be at the bottom, if you listen to the commandments of Adonai your God that I am commanding you \textit{today}\lebnote{Literally “the day”} \textit{and diligently observe them}.\lebnote{Literally “to observe them and to do them”}
\verse And you shall not turn aside from \textit{any of}\lebnote{Literally “all of”} the words that I am commanding you \textit{today}\lebnote{Literally “the day”} to the right or left by going after other gods to serve them.
\verse “\textit{And then}\lebnote{Literally “and it will happen”} if you do not listen to the voice of Adonai your God \textit{by diligently observing}\lebnote{Literally “to observe and to do”} all of his commandments and his statutes that I am commanding you \textit{today},\lebnote{Literally “the day”} then all of these curses shall come upon you, and they shall overtake you:
\verse “You shall\lebnote{All of these curses have a future orientation that points to Israel in the land of Canaan} be cursed in the city, and you shall be cursed in the field.
\verse “Your basket shall be cursed and your kneading trough.
\verse “The fruit of your womb shall be cursed and the fruit of you ground, the calves of your cattle and the lambs of your flock.
\verse “You shall be cursed \textit{when you come in},\lebnote{Literally “at/in your coming”} and you shall be cursed \textit{when you go out}.\lebnote{Literally “at/in your going out”}
\verse “Adonai will send upon you\lebnote{Others translate as a wish: “may Adonai send upon you …”} the curse,\lebnote{These words are translated variously, e.g., “starvation,” “thirst,” “rebuke/dysentery” (NEB) or “curses,” “confusion,” “frustration” (NLT)} the panic,\lebnote{Or “confusion”} and the threat\lebnote{Or “rebuke”} \textit{in everything that you undertake},\lebnote{Literally “in all the sending out of your hand that you do”} \textit{until you are destroyed and until you perish quickly}\lebnote{Literally “until your to be destroyed and until your to perish quickly”} \textit{because of}\lebnote{Literally “from the face of the evil of your deeds”} the evil of your deeds in that\lebnote{Literally “which”} you have forsaken me.
\verse Adonai will cause the plague to cling to you \textit{until it consumes you}\lebnote{Literally “it to consume to”} from the land that you are going to,\lebnote{Hebrew “to there”} to take possession of it.
\verse Adonai will afflict you with the wasting diseases and with the fever and with the inflammation and with the scorching heat and with the sword\lebnote{Others translate “drought” (NLT, NEB)} and with the blight and with the mildew, and they shall pursue you \textit{until you perish}.\lebnote{Literally “until your/you to perish”}
\verse And your heavens that are over your heads shall be like bronze, and the earth that is under you shall be like iron.
\verse Adonai will change the rain of your land to fine dust and to sand; from the heaven it shall come down upon you \textit{until you are destroyed}.\lebnote{Literally “until your being destroyed”}
\verse “Adonai shall cause you to be defeated \textit{before}\lebnote{Literally “to the face of”} your enemies; on one road you shall go \textit{against}\lebnote{Literally “to/toward”} them,\lebnote{Hebrew “him”} but you will flee on seven roads \textit{before}\lebnote{Literally “to the face of”} them,\lebnote{Hebrew “him”} and you shall become a thing of horror to all of the kingdoms of the earth.
\verse And your dead bodies shall be as food for all of the birds of the heaven and to the animals of the earth, and there shall not be \textit{anyone to frighten them away}.\lebnote{Literally “one causing fright to them”}
\verse “Adonai shall afflict you with the boils of Egypt and with tumors and with the scurvy and with \textit{the skin rash that cannot be healed}.\lebnote{Literally “the skin rash that not it is able it to be healed”}
\verse Adonai shall afflict you with madness and with blindness and with confusion of heart.
\verse And you shall be groping at noon just as the blind person gropes in the dark, and you shall not succeed in finding your way, and you shall only be abused and robbed \textit{all the time},\lebnote{Literally “all the days”} and there will not be \textit{anyone who will rescue you}. \lebnote{Literally “one who delivers/rescues you”}
\verse You shall become engaged to a woman, but another man shall sleep\lebnote{Or “ravish/violate”} with her; you shall build a house, but you shall not live in it; a vineyard you shall plant, but you shall not enjoy it.
\verse Your ox shall be slaughtered before your eyes, and you shall not eat it; your donkey shall be stolen right \textit{before you},\lebnote{Literally “to the face of you”} and it shall not be returned to you; your sheep and your goats shall be given to your enemies, and \textit{there shall not be anyone who rescues you}.\lebnote{Literally “there shall not be for you one who rescues”}
\verse Your sons and your daughters shall be given to other people, and \textit{you will be looking on}\lebnote{Literally “your eyes will see/look”} \textit{longingly}\lebnote{Literally “wearing out”} for\lebnote{Hebrew “to”} them all day, \textit{but you will be powerless to do anything}.\lebnote{Literally “but there will not be for power of your hand”}
\verse A people that you do not know shall consume the harvest of your land and all your labor, and you will be only oppressed and crushed \textit{for the rest of your lives}.\lebnote{Literally “all of the days”}
\verse You shall become mad \textit{because of what your eyes shall see}.\lebnote{Literally “from/by the sight of your eyes”}
\verse Adonai shall strike you with grievous boils on the knees and on the upper thighs from which \textit{you will not be able to be healed},\lebnote{Literally “which not you are able to heal”} from the sole of your foot and up to your crown.
\verse Adonai will bring you and your king whom you set up over you to a nation that you or your ancestors\lebnote{Or “fathers”} have not known, and there you will serve other gods of wood and stone.
\verse And you will become a horror and a proverb and ridicule among all the peoples where Adonai drives you there.
\verse “You shall carry out much seed to the field, but you shall gather little produce, for the locust shall devour it.
\verse You shall plant vineyards and you shall dress\lebnote{Or “cultivate”} them, but you shall not drink wine and you shall not gather grapes, for the worm shall eat it.\lebnote{That is, the produce}
\verse There shall be olive trees for you in all of your territory, but you shall not anoint yourself, for your olives\lebnote{Hebrew “olive”} shall drop off.
\verse You shall bear sons and daughters, but they shall not be \textit{yours},\lebnote{Literally “for you”} for they shall go into captivity.
\verse The cricket shall take possession of all your trees and the fruit of your ground.
\verse The alien that is in your midst shall ascend over you, higher and higher, but you shall go down lower and lower.
\verse He shall lend to you, but you shall not lend to him; he shall be the head, but you shall be the tail.
\verse “And all of these curses shall come over you, and they shall pursue you, and they shall overtake you \textit{until you are destroyed},\lebnote{Literally “until you to be destroyed”} because \textit{you did not listen}\lebnote{Literally “not you listened”} to the voice of Adonai your God, \textit{by observing}\lebnote{Literally “to observe”} his commandments and his statutes that he commanded you.
\verse And they shall be among you as a sign and as a wonder and among your offspring \textit{forever}.\lebnote{Literally “until eternity,” but not in a timeless, philosophic sense}
\verse “\textit{Because}\lebnote{Literally “Under”} of the fact that you did not serve Adonai your God with joy and with gladness of heart for the abundance of everything,
\verse then you shall serve your enemies, whom Adonai will send against you under conditions of famine, in thirst, in nakedness, and in a lack of everything; and he shall place\lebnote{Literally “give”} a yoke of iron on your neck \textit{until he has destroyed you}.\lebnote{Literally “until his destroying of you”}
\verse Adonai will raise up against you a nation from far off, from the end of the earth, attacking as the eagle swoops down, \textit{a nation whose language you will not understand},\lebnote{Literally “a nation who not you will understand its language”}
\verse \textit{a grim-faced nation}\lebnote{Literally “a nation fierce/determined of face”} \textit{who does not show respect}\lebnote{Literally “who not lifts up faces”} to the old and the young and does not show pity.
\verse And it\lebnote{That is, the invading nation} shall consume the fruit of your livestock and the fruit of your ground \textit{until you are destroyed},\lebnote{Literally “until your destroying”} and who will not leave for you any grain, wine, and olive oil, \textit{calves of your herds},\lebnote{Literally “what is dropped of your herds”} and \textit{lambs of}\lebnote{Literally “the offspring of small animals”} your flock \textit{until it has destroyed you}.\lebnote{Literally “until his to destroy you”}
\verse And it shall besiege you in all your towns \textit{until your high and fortified walls fall},\lebnote{Literally “until the coming down of your walls, high and fortified”} which you are trusting in\lebnote{Hebrew “in them,” which is redundant} \textit{throughout your land};\lebnote{Literally “in all of your land”} and it shall besiege you in all of your towns in all of your land that Adonai your God has given to you.
\verse And you shall eat the fruit of your womb, the flesh of your sons and your daughters, whom Adonai your God gave to you, \textit{during the siege and during the distress}\lebnote{Literally “in siege and in distress”} your enemy inflicts upon you.
\verse The most refined and the very sensitive\lebnote{Or “kindly disposed”} man among you \textit{shall be mean with his brother}\lebnote{Literally “shall be bad/evil his eye against his brother”} and \textit{against his beloved wife}\lebnote{Literally “against the wife of his lap”} and against the rest\lebnote{Or “remainder”} of his children that he has left over,
\verse \textit{by refraining from giving}\lebnote{Literally “from giving”} to even one of them any of the meat of his children that he eats, because there is not anything that is left over for him \textit{during the siege and distress}\lebnote{Literally “in the siege and distress”} that your enemy inflicts upon you.
\verse The most refined and the most delicate woman among you, who shall\lebnote{Or “would”} not venture to put the sole of her foot on the ground from being so delicate and from such gentleness, \textit{shall be mean to her beloved husband}\lebnote{Literally “shall be bad her eye against the husband of her lap”} and against her son and against her daughter,
\verse and even concerning her afterbirth \textit{that goes out}\lebnote{Literally “the going out”} from between her feet and also concerning her children that she bears, because she eats them for lack of anything in secret \textit{during the siege and during the distress}\lebnote{Literally “in the siege and in the distress”} that your enemy inflicts upon her in your \textit{towns}.\lebnote{Literally “gates”}
\verse “If \textit{you do not diligently observe}\lebnote{Literally “you observe to do”} all the words of this law written in this scroll by revering this glorious and awesome name, Adonai your God,
\verse then Adonai shall overwhelm you with your plagues and the plagues of your offspring, severe plagues and lasting illnesses, grievous and enduring.
\verse And he shall bring back upon you all the diseases of Egypt concerning which \textit{you were in dread}\lebnote{Literally “you dreaded”} \textit{because of them}.\lebnote{Literally “from their presence”}
\verse Also any illness and any plague\lebnote{Or “each illness and each plague”} that is not written in the scroll of this law, he shall bring them, Adonai, upon you until you are destroyed.
\verse And you shall remain \textit{only a few people}\lebnote{Literally “with people of few”} in place of the fact you were formerly as the stars of heaven as far as number is concerned, because you did not listen to the voice of Adonai your God.
\verse \textit{And then}\lebnote{Literally “And it will happen”} as Adonai delighted over you \textit{to make you prosperous}\lebnote{Literally “to do good with you”} to make you numerous, so Adonai shall delight over you to exterminate you\lebnote{Or “to cause you to perish”} and to destroy you, and so you shall be plucked from the land that you are going there to take possession of it.
\verse And Adonai shall scatter you among all the nations from one end of the earth up to the other end of the earth, and there you shall serve other gods that you have not known nor\lebnote{Hebrew “and”} your ancestors,\lebnote{Or “your fathers”} gods of wood and stone.
\verse And among these nations you shall not find rest, and there shall not be a resting place for the sole of your foot, and Adonai shall give you there an anxious heart and a weakening of eyes\lebnote{Or “a failure of eyes”} and a languishing of your inner self.\lebnote{Or “soul”}
\verse And \textit{your life shall hang in doubt before you},\lebnote{Literally “they shall be your life hanging for you in front”} and you shall be startled\lebnote{Or “in fear”} night and day, and you shall not be confident of your life.
\verse In the morning you shall say, ‘\textit{If only it was evening}!’\lebnote{Literally “Who shall give evening?”} and in the evening you shall say ‘\textit{If only it was morning}!’\lebnote{Literally “Who shall give morning?”} because of the dread of your heart that you shall feel, and because of the sight of your eyes that you shall see.
\verse And Adonai shall bring you back to Egypt in ships by the route that I \textit{promised}\lebnote{Literally “spoke”} to you that ‘\textit{You shall not see it again}!’\lebnote{Literally “you shall not do again further to see it/her”} And you shall sell yourself there to your enemies as slaves and as female slaves, but there will not be a buyer.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 29
\verseWithHeading{Covenant Renewal, Oaths, Restoration, Charges to the Nation} \lebnote{Deuteronomy 29:1–29 in the English Bible is 28.69–29:28 in the Hebrew Bible} These are the words of the covenant that Adonai commanded Moses to make with the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} in the land of Moab \textit{besides}\lebnote{Literally “from to alone”} the covenant that he made with them at Horeb.
\verse And Moses summoned all of Israel and said to them, “You saw all that Adonai did before your eyes in the land of Egypt and to Pharaoh and to all his servants and to all his land;
\verse that is, the great trials that your eyes saw, and those great signs and wonders.
\verse But Adonai has not given to you a heart to understand, or eyes to see, or ears to hear, even \textit{to this day}.\lebnote{Literally “until the day the this”}
\verse And I have led you forty years\lebnote{Hebrew “year”} in the desert; your clothes have not worn out \textit{on you},\lebnote{Literally “from on you”} and your sandal has not worn out \textit{on your foot}.\lebnote{Literally “from on your foot”}
\verse You have not eaten bread, and you have not drunk wine and strong drink, so that you may know that I am Adonai your God.
\verse And when you came to this place then Sihon the king of Heshbon, and Og the king of Bashan, came out to meet you for battle, and we defeated them.
\verse And we took their land and gave it as an inheritance to the Reubenites\lebnote{Hebrew “Reubenite”} and to the Gadites\lebnote{Hebrew “Gadite”} and to the half-tribe of Manasseh.
\verse And \textit{you must diligently observe the words of this covenant},\lebnote{Literally “you must keep the words of the covenant the this and you must do them”} so that you may succeed in all that you do.
\verse “You are standing \textit{today},\lebnote{Literally “the day”} all of you, \textit{before}\lebnote{Literally “to the face of”} Adonai your God, your leaders, your tribes, your elders, and your officials, all the men of Israel,
\verse your little children, your women and your aliens\lebnote{Hebrew “alien”} who are in the midst of your camp, from the choppers\lebnote{Hebrew “chopper”} of your wood to the drawers\lebnote{Hebrew “drawer”} of your water,
\verse \textit{in order for you to enter into the covenant of Adonai your God},\lebnote{Literally “for you to go over into the covenant of Adonai your God”} and into\lebnote{Or “with”} his oath that Adonai your God is \textit{making with you}\lebnote{Literally “cutting with you”} \textit{today},\lebnote{Literally “the day”}
\verse in order to establish you \textit{today}\lebnote{Literally “the day”} \textit{to himself}\lebnote{Literally “for him”} as a people and so that he may be for you as God, just as he \textit{promised}\lebnote{Literally “spoke”} to you and \textit{just as}\lebnote{Literally “according to that which”} he swore to your ancestors,\lebnote{Or “your fathers”} to Abraham, to Isaac, and to Jacob.
\verse “Now I am not \textit{making this covenant}\lebnote{Literally “cutting this covenant”} and this oath \textit{with you alone}.\lebnote{Literally “with you to alone you”}
\verse But with \textit{whoever is standing here}\lebnote{Literally “with who he is here with us standing”} with us \textit{today}\lebnote{Literally “the day”} \textit{before}\lebnote{Literally “to the face of”} Adonai our God, and with \textit{whoever is not standing here}\lebnote{Literally “with who he is not standing here with us”} with us \textit{today}.\lebnote{Literally “the day”}
\verse For you know how we lived in the land of Egypt and how we traveled through the midst of the nations that you traveled\lebnote{Or “journeyed”} through.
\verse And you have seen their detestable things and their idols of wood and stone, silver, and gold that were among them,
\verse so that \textit{there is not}\lebnote{Literally “lest there be/develop”} among you a man or a woman or a clan or a tribe \textit{whose heart}\lebnote{Literally “his heart”} turns \textit{today}\lebnote{Literally “the day”} from being with Adonai our God to go to serve the gods of these nations, so that there is not among you a root sprouting poison and wormwood.
\verse And then when he hears the words of this oath, then\lebnote{Or “and”} \textit{he will assure himself}\lebnote{Literally “he will bless himself”; HALOT 160 suggests “to consider oneself fortunate”} in his heart, \textit{saying},\lebnote{Literally “to say”} ‘\textit{Safety shall be mine even though I go in the stubbornness of my heart},’\lebnote{Literally “Peace shall happen/be for me, although/even if in the stubbornness of my heart I go”} thereby destroying the well-watered land along with the parched.\lebnote{Some translators prefer to include the last clause as a part of the words of the wicked man (NASV vs. NEB)}
\verse Adonai will not be willing to forgive him, for by then the anger of Adonai will smoke, and his passion against that man and all the curses written in this scroll will descend on him, and Adonai will blot out his name from under heaven.
\verse And Adonai will single him out for calamity out of all the tribes of Israel, according to all the curses of the covenant written in the scroll of this law.
\verse “And the next generation, that is, your children who will rise up after you, and the foreigner who will come from a distant land, when\lebnote{Hebrew “and”} they will see the plagues of that land and its diseases that Adonai has inflicted upon it, will say,
\verse ‘All its land is brimstone and salt left by fire, \textit{none of its land will be sown},\lebnote{Literally “all of its land will not be sown”} and it will not make plants sprout out and it will not grow any vegetation; it is as the destruction of Sodom and Gomorrah, Adman and Zeboiim, which Adonai overturned in his anger and in his wrath.’
\verse And all the nations will say, ‘\textit{Why}\lebnote{Literally “On what basis”} has Adonai done \textit{such a thing}\lebnote{Literally “so”} to this land? What caused the fierceness of this great anger?’
\verse And they will say,\lebnote{Or “answer/respond”} ‘It is because they abandoned the covenant of Adonai, the God of their ancestors,\lebnote{Or “fathers”} which he \textit{made}\lebnote{Literally “cut”} with them \textit{when he brought them out}\lebnote{Literally “at/in his to bring them out from the land of Egypt”} from the land of Egypt.
\verse And they went and served other gods and bowed down to them, gods whom they did not know them and he\lebnote{That is, Adonai} had not allotted to them.
\verse So \textit{the anger of Adonai was kindled}\lebnote{Literally “became hot the nose of Adonai”} against that land to bring upon it all the curses written in this scroll,
\verse and Adonai uprooted them from their land in anger and in wrath and in great fury, and he cast them into another land, \textit{just as it is today}.’\lebnote{Literally “as the day the this”}
\verse “The hidden things \textit{belong to Adonai}\lebnote{Literally “are for Adonai”} our God, but the revealed things \textit{belong to us}\lebnote{Literally “are for us”} to know and to our children \textit{forever},\lebnote{Literally “until eternity”} in order to do all the words of this law.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 30
\verse “And then when all of these things come upon you, the blessing and the curse that I have set \textit{before you}\lebnote{Literally “to the face of you”} \textit{and you call them to mind}\lebnote{Literally “and you bring them back to your heart”} among the nations there where Adonai your God has scattered you,
\verse and you return to Adonai and you listen to his voice according to all that I am commanding you \textit{today},\lebnote{Literally “the day”} both you and your children,\lebnote{Or “sons”} with all your heart and with all your inner self,\lebnote{Or “soul”}
\verse and Adonai your God will restore your fortunes, and he will have compassion upon you, and \textit{he will again gather you together}\lebnote{Literally “he will do again and he will gather you together”} from all the peoples where Adonai your God scattered you there.
\verse “Even if \textit{you are outcasts}\lebnote{Literally “he/it shall be outcasting your”} at the end of the heavens, even from there Adonai your God shall gather you, and from there \textit{he shall bring you back}.\lebnote{Literally “he shall take/fetch you”}
\verse And Adonai your God will bring you to the land that your ancestors\lebnote{Or “fathers”} had taken possession of,\lebnote{Hebrew “of it”} and he will make you successful, and he will make you more numerous than your ancestors.\lebnote{Or “fathers”}
\verse “And Adonai your God will circumcise your heart and the heart of your offspring to love Adonai your God with all your heart and with all your inner self\lebnote{Or “soul”} \textit{so that you may live}.\lebnote{Literally “for the sake of your life”}
\verse And Adonai your God will put all these curses on your enemies and \textit{on those who hate you},\lebnote{Literally “the haters of you”} on \textit{those who harassed you}.\lebnote{Literally “who pursued after/persecuted you”}
\verse And \textit{you will again listen}\lebnote{Literally “you will return and you will listen”} to the voice of Adonai, and you will do all his commandments that I am commanding you \textit{today}.\lebnote{Literally “the day”}
\verse And Adonai your God will make you prosperous \textit{in all your undertakings},\lebnote{Literally “in all of the work of your hand”} and in the fruit of your livestock and in the fruit of your ground \textit{abundantly},\lebnote{Literally “to/for good”} for Adonai \textit{will again rejoice}\lebnote{Literally “he will return ... to rejoice”} over you, just as he rejoiced over your ancestors.\lebnote{Or “fathers”}
\verse He will do this if you listen to\lebnote{Or “if you obey/hearken to”} the voice of Adonai your God by keeping his commandment and his statutes written in the scroll of this law and if you return to Adonai your God with all your heart and with all your inner self.\lebnote{Or “soul”}
\verse “For this commandment that I am commanding you \textit{today}\lebnote{Literally “the day”} is \textit{not too wonderful for you},\lebnote{Literally “not is wonderful it from you”} and it is not too far from you.
\verse It is not in the heavens \textit{so that you might say},\lebnote{Literally “to say”} ‘Who will go up for us to the heavens and get it for us and cause us to hear it, so that we may do it?’
\verse And it is \textit{not beyond the sea},\lebnote{Literally “is not from beyond the sea”} \textit{so that you might say},\lebnote{Literally “to say”} ‘Who will cross for us to the other side of the sea and take it for us and cause us to hear it, so that we may do it?’
\verse But the word is very near you, even in your mouth and in your heart, \textit{so that you may do it}.\lebnote{Literally “to do it”}
\verse “See, I am setting \textit{before you}\lebnote{Literally “to the face of you”} \textit{today}\lebnote{Literally “the day”} life and prosperity and death and disaster;
\verse what I am commanding you \textit{today}\lebnote{Literally “the day”} is to love Adonai your God by going\lebnote{Or “walking”} in his ways and by keeping his commandments and his statutes and his regulations, and then you will live, and you will become numerous, and Adonai your God will bless you in the \textit{land where you are going}.\lebnote{Literally “the land where you are going there”}
\verse However, if your heart turns aside and you do not listen and you are lured away and you bow down to other gods and you serve them,
\verse I declare to you \textit{today}\lebnote{Literally “the day”} that you will certainly perish; \textit{you will not extend your time}\lebnote{Literally “you will not make long days”} on the land that you are crossing the Jordan to go there to take possession of it.
\verse I invoke as a witness against you \textit{today}\lebnote{Literally “the day”} the heaven and the earth: life and death I have set \textit{before you},\lebnote{Literally “to the face of you”} blessing and curse. So choose life, so that you may live, you and your offspring,\lebnote{Or “seed”}
\verse by loving Adonai your God by listening to his voice and by clinging to him, for he is your life and the length of your days in order for you to live on the land that Adonai swore to your ancestors,\lebnote{Or “fathers”} to Abraham, to Isaac, and to Jacob, to give to them.”
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 31
\verseWithHeading{Succession, Deposition, and Recitation of Text} And Moses went and spoke these words to all Israel.
\verse And he said to them, “I am a hundred and twenty years old \textit{today};\lebnote{Literally “the day”} I am not able to go out and to come in any longer, and Adonai said to me, ‘You may not cross this Jordan.’
\verse Adonai your God is about to cross \textit{before you};\lebnote{Literally “to the face of you”} he will destroy these nations \textit{before you},\lebnote{Literally “to the face of you”} and you shall dispossess them. Joshua will be crossing \textit{before you},\lebnote{Literally “to the face of you”} just as Adonai \textit{promised}.\lebnote{Literally “spoke”}
\verse And Adonai will do to them just as he did to Sihon and to Og, kings of the Amorites, and to their land, which he destroyed with them.
\verse And Adonai \textit{will deliver them to you before you},\lebnote{Literally “will give them Adonai to the face of you”} and you shall do to them according to every commandment that I have commanded you.
\verse Be strong and be courageous; you should not be afraid, and you \textit{should not be in dread from their presence},\lebnote{Literally “not you should dread from their face”} for Adonai your God is the one going with you; he will not leave you alone and he will not forsake you.”
\verse Then Moses summoned Joshua, and he said to him \textit{in the presence of all Israel},\lebnote{Literally “before the eyes of all Israel”} “Be strong and be courageous, for you will go with this people into the land that Adonai swore to their ancestors\lebnote{Or “their fathers”} to give to them, and you will give it to them as an inheritance.
\verse Adonai is the one going \textit{before you};\lebnote{Literally “to the face of you”} he will be with you, and he will not leave you alone, and he will not forsake you; you shall not be afraid, and you shall not be discouraged.”
\verse So Moses wrote this law, and he gave it to the priests, the descendants\lebnote{Or “sons”} of Levi, the ones carrying the ark of the covenant of Adonai, and to all the elders of Israel.
\verse Then Moses commanded them, \textit{saying},\lebnote{Literally “to say”} “At the end of seven years, in the time of the year for canceling debts during the Feast of Booths,
\verse \textit{when all Israel comes to appear before}\lebnote{Literally “at/in the coming of all of Israel to see”} Adonai their God at the place that he will choose, you shall read this law before all Israel \textit{in their hearing}.\lebnote{Literally “in their ears”}
\verse Assemble the people, the men and the women and the little children and your aliens that are in your \textit{towns},\lebnote{Literally “gates”} so that they may hear and so that they may learn and they may revere Adonai your God, and \textit{they shall diligently observe}\lebnote{Literally “they shall keep to do”} all the words of this law.
\verse And then their children,\lebnote{Hebrew “child”} who have not known, they too may hear, and they may learn to revere Adonai their God all the days \textit{that you live}\lebnote{Literally “that you are alive”} on the land that you are crossing the Jordan \textit{to get there}\lebnote{Literally “there”} to take possession of it.”
\verse Then Adonai said to Moses, “Look, \textit{you are about to die};\lebnote{Literally “are near your days to die”} call Joshua and present yourselves in the tent of assembly,\lebnote{Or “meeting”} so that I may instruct him.” So Moses and Joshua went and presented themselves in\lebnote{Or “at”} the tent of assembly.\lebnote{Or “meeting”}
\verse And Adonai appeared in the tent in a column of cloud, and the column of the cloud stood at the entrance of the tent.
\verse And Adonai said to Moses, “Look, you are about to lie down with your ancestors,\lebnote{Or “fathers”} and this people will arise and they\lebnote{Hebrew “he”} will play the prostitute after \textit{the foreign gods}\lebnote{Literally “the gods of the foreigner/foreign person of the land”} of the land \textit{to which they are going}.\lebnote{Literally “which he is going to”}
\verse And \textit{my anger shall flare up against them}\lebnote{Literally “shall become my nose against it/him”} on that day, and I will forsake them, and I will hide my face from them, and they\lebnote{Hebrew “he”} shall become as prey, and disasters and troubles shall find\lebnote{Or “encounter”} them,\lebnote{Hebrew “him”} and they\lebnote{Hebrew “he”} shall say on that day, ‘Is it not because our\lebnote{Hebrew “my” but represents all the people} God is not in our\lebnote{Hebrew “my”} midst that these disasters have found\lebnote{Or “encountered”} us?’\lebnote{Hebrew “me”}
\verse But I will certainly hide my face on that day, because of all of the evil that they\lebnote{Hebrew “he”} did because they\lebnote{Hebrew “he”} turned to other gods.
\verse “And so then write this song for yourselves and teach it to the \textit{Israelites};\lebnote{Literally “sons/children of Israel”} put it in their mouth, so that this song may be for me as a witness against the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse For when I bring them\lebnote{Hebrew “him,” referring to the people} into the land that I swore to their\lebnote{Hebrew “his”} ancestors,\lebnote{Or “fathers”} a land flowing with milk and honey, \textit{they will eat their fill},\lebnote{Literally “he will eat and he will be satisfied”} and they\lebnote{Hebrew “he”} will become fat, and then they\lebnote{Hebrew “he”} will turn to other gods, and they will serve them, and they will spurn me, and they\lebnote{Hebrew “he”} will break my covenant.
\verse \textit{And then}\lebnote{Literally “And it shall happen”} many disasters and troubles will come upon them,\lebnote{Hebrew “him”} and this song will give evidence before them\lebnote{Hebrew “him”} as a witness, because it will not be forgotten from out of the mouth of their descendants,\lebnote{Hebrew “his offspring/descendant”} for I know their\lebnote{Hebrew “his”} inclination that they\lebnote{Hebrew “he”} are having \textit{today}\lebnote{Literally “the day”} before I have brought them\lebnote{Hebrew “him”} into the land that I swore.”
\verse And Moses wrote this song on that day and taught it to the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse Then he\lebnote{That is, Adonai} told Joshua the son of Nun, and said to him, “Be strong and be courageous, for you shall bring the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} into the land that I swore to them, and I will be with you.”
\verse \textit{And then when Moses finished writing}\lebnote{Literally “And it happened as/when to finish Moses to write”} the words of this law on the scroll \textit{until they were complete},\lebnote{Literally “until their to be complete/”}
\verse then Moses commanded the Levites carrying the ark of the covenant of Adonai, \textit{saying},\lebnote{Literally “to say”}
\verse “Take the scroll of this law and put it at the side of the ark of the covenant of Adonai your God, and it will be there as a witness against you.
\verse For I know your rebellion and your stiff neck even now while I am still alive with you \textit{today},\lebnote{Literally “the day”} rebelling against Adonai, and \textit{how much more}\lebnote{Literally “and also for/indeed”} after my death.
\verse Assemble to me all the elders of your tribes and your officials, so that I may speak in their ears these words, and that I may call as witness against them heaven and earth.
\verse For I know that after my death you will certainly act corruptly, and you will turn aside from the way that I have commanded you, and the disaster in the future days will befall you, because you will do evil\lebnote{Hebrew “the evil”; the definite article indicates the general concept of evil} in the eyes of Adonai, \textit{provoking him with the work of your hands}.”\lebnote{Literally “to provoke him with/by the work of your hands”}
\verse So Moses spoke into the ears of the assembly of Israel the words of this song \textit{until they were complete}.\lebnote{Literally “until their to be complete”}
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 32
\verseWithHeading{The Song of Moses} “Give ear, O heavens, and I will speak, 
and let the earth hear the words of my mouth.
\verse May my teaching trickle like the dew, 
my words like rain showers on tender grass, 
and like spring showers on new growth.
\verse For I will proclaim the name of Adonai; 
ascribe greatness to our God!
\verse The Rock, his work is perfect, 
for all his ways are just; 
he is a faithful God, and \textit{without injustice};\lebnote{Literally “there is not injustice”} 
righteous and upright is he.
\verse They have behaved corruptly toward\lebnote{Or “with”} him; 
they are not his children; this is their flaw, 
a generation crooked and perverse.
\verse Like this do you treat Adonai, 
foolish and \textit{unwise}\lebnote{Literally “not wise”} people? 
Has he not, your father, created you? 
He made you, and he established you.
\verse Remember the old days, \textit{the years long past};\lebnote{Literally “the years of from generation to generation”} 
ask your father, and he will inform you, 
your elders and they will tell you.\lebnote{Hebrew “to you”}
\verse \textit{When the Most High apportioned}\lebnote{Literally “In/at the apportioning of the Most High”} the nations, 
at his dividing up of the sons of humankind, 
he fixed the boundaries of the peoples, 
according to the number of the children of Israel.\lebnote{LXX reads “the number of the angels of God”; Dead Sea Scrolls reads “the number of the sons of God”}
\verse For Adonai’s portion was his people, 
Jacob the share of his inheritance.
\verse He found him in a desert land, 
and in a howling, desert wasteland; 
he \textit{encircled him},\lebnote{Literally “moved around him”} he cared for him, 
he protected him like the \textit{apple}\lebnote{Literally “pupil”} of his eye.
\verse As an eagle stirs up its nest, 
hovers over its young, 
spreads out its wings, takes them,\lebnote{Hebrew “it” but used poetically and with plural sense in context} 
carries them\lebnote{Hebrew “it” but used poetically and with plural sense in context} on its pinions,\lebnote{Hebrew “pinion”}
\verse so Adonai alone guided him,\lebnote{That is, Jacob, standing for Israel} 
and \textit{there was no foreign god accompanying him}.\lebnote{Literally “and there was not with him a god foreign/strange”}
\verse And he set him on the high places of the land, 
and he fed him the crops\lebnote{Hebrew “crop”} of the field, 
and he nursed him with honey from crags,\lebnote{Hebrew “crag”} 
and with oil from flinty rock,
\verse With curds\lebnote{Hebrew “curd”} from the herd, 
and with milk from the flock, 
with the fat of young rams, 
and rams, the offspring of Bashan, 
and with goats along with the finest kernels of wheat, 
and from the blood of grapes\lebnote{Hebrew “grape”} you drank fermented wine.\lebnote{Or “partially fermented wine”; others translate simply as wine (NASB, NEB); HALOT 330, “still fermenting wine”}
\verse And Jeshurun grew fat, and he kicked; 
you grew fat, you bloated, and you became obstinate; 
and he abandoned God, his maker, 
and he scoffed at the rock of his salvation.
\verse They made him jealous with strange gods; 
with detestable things they provoked him.
\verse They sacrificed to the demons, not God, 
to gods whom\lebnote{Hebrew “them”} they had not known, 
new gods who\lebnote{Hebrew “they,” understood in the verb form} came from recent times; 
\textit{their ancestors had not known them}.\lebnote{Literally “their fathers not knew them”}
\verse The rock who\lebnote{Hebrew “he,” understood in verb form} bore you, you neglected, 
and you forgot God, the one giving you birth.
\verse Then Adonai saw, and he spurned them, 
because of the provocation of his sons and his daughters.
\verse So he said, ‘I will hide my face from them; 
I will see what will be their end, 
for they are a generation of perversity, 
\textit{children in whom there is no faithfulness}.\lebnote{Literally “sons not faithfulness is in them”}
\verse They annoyed me with what is not a god; 
they provoked me with their idols. 
So I will make them jealous with those not a people, 
with a foolish nation I will provoke them.
\verse For a fire was kindled\lebnote{Or “ignited”} by my anger, 
and it burned \textit{up to the depths of Sheol},\lebnote{Literally “up to Sheol depths”} 
and it devoured the earth and its produce, 
and it set afire the foundation of the mountains.
\verse I will heap disasters upon them; 
my arrows I will spend on them.
\verse They will become weakened by famine, 
and consumed by plague and bitter pestilence; 
and the teeth of wild animals I will send against them, 
with the poison of the creeping things in the dust;
\verse From outside her boundaries the sword will make her childless, 
and from inside, terror; 
both for the young man and also the young woman, 
the infant along with the gray-headed man.
\verse I thought, “I will wipe them out; 
\textit{I will make people forget they ever existed}.”\lebnote{Literally “I will blot out from among human being their remembrance”}
\verse If I had not feared a provocation of the enemy, 
lest their foes might misunderstand, \lebnote{Or “so that they not make a false construal” of what has happened} 
lest they should say, \lebnote{Or “so that they might not”} “Our hand is 
\textit{triumphant},\lebnote{Literally “raised/held high”} and Adonai did not do all this.” ’
\verse For \textit{they are a nation void of sense},\lebnote{Literally “they a nation perishing of counsel”} 
and there is not any understanding in them.
\verse If only they were wise, they would understand this; 
they would discern \textit{for themselves their end}.\lebnote{Literally “they would discern for end their”}
\verse How could one chase a thousand 
and two could cause a myriad to flee, 
if their Rock had not sold them, 
and Adonai had not given them up?
\verse For the fact of the matter is, 
their rock is not like our Rock, 
and our enemies recognize\lebnote{The meaning of this expression is uncertain, but “discerns” or “judges” seem good choices; see HALOT 932, which allows for “judges” or “in the estimation of our enemies”} this.
\verse For their vine is from the vine of Sodom, 
and from the terraces of Gomorrah; 
their grapes are grapes of poison; 
\textit{their clusters are bitter}.\lebnote{Literally “clusters of bitter are for them”}
\verse Their wine is the poison of snakes, 
and the deadly poison of horned vipers.
\verse ‘Is not this stored up with me, 
sealed in my treasuries?’\lebnote{Or “storehouses}
\verse \textit{Vengeance belongs to me}\lebnote{Literally “To me is vengeance”} and also recompense, 
\textit{for at the time their foot slips},\lebnote{Literally “at the time when it shall slip foot their”} 
because the day of their disaster is near, 
\textit{and fate comes quickly for them}.’\lebnote{Literally “and comes quickly/hurries events to come to them”}
\verse For Adonai will judge on behalf of his people, 
and concerning his servants; 
he will change his mind when he sees that their power has disappeared, 
and there is no one left, confined\lebnote{Or “bond/bound”} or free.\lebnote{Or “freed”}
\verse And he will say, ‘Where are their gods, 
their rock in whom\lebnote{Hebrew “in him”} they took refuge?
\verse Who ate the fat of their sacrifices 
and drank the wine of their libations? 
Let them rise up, and let them help you; 
\textit{Let them be to you a refuge}.\lebnote{Literally “let there be/him/them unto you as a shelter”}
\verse See, now, that I, even I am he, 
and there is not a god besides me; 
I put to death and I give life; 
I wound and I heal; 
there is not one who delivers from my hand!
\verse For indeed I lift up my hand to heaven, 
And I promise \textit{as I live forever},\lebnote{Literally “live I to eternity”}
\verse When I sharpen\lebnote{Or “have sharpened”} \textit{my flashing sword},\lebnote{Literally “the flashing of my sword”} 
and my hand takes hold\lebnote{Or “seizes”} of it in judgment, 
\textit{I will take reprisals against my foes},\lebnote{Literally “I will let return vengeance to foes my”} 
and my haters I will repay.
\verse I will make my arrows drunk with blood, 
and my sword shall devour flesh with the blood of the slain, 
and captives\lebnote{Hebrew “captive”} from the heads\lebnote{Hebrew “head”} of the leaders of the enemy.’
\verse Call for songs of joy, O nations, concerning his people,\lebnote{Dead Sea Scrolls reads: “Rejoice, O heavenly ones, with him! Bow down, all you gods, before him!”} 
for the blood of his servants he will avenge, 
\textit{and he will take reprisals against his foes},\lebnote{Literally “vengeance he will let return to his foes”} 
and he will make atonement for his land, his people.”
\verse And Moses came, and he spoke\lebnote{Or “recited”} all the words of this song in the ears of the people; that is, he and Joshua the son of Nun.
\verse \textit{And when Moses finished speaking}\lebnote{Literally “And he finished Moses to speak”} all these words to all Israel,
\verse then he said to them, “\textit{Take to heart all the words}\lebnote{Literally “Put heart your to all the words”} that I am admonishing against you \textit{today}\lebnote{Literally “the day”} concerning which you should instruct them with respect to your children\lebnote{Or “sons”} \textit{so that they will observe diligently}\lebnote{Literally “to keep to do”} all the words of this law,
\verse for it is not a trifling matter among you, but it is your life, and through this word \textit{you will live long in the land}\lebnote{Literally “you will make long days on the land”} that you are about to cross the Jordan to get there to take possession of it.”
\verseWithHeading{Instructions Concerning Moses’ Death} And Adonai said to Moses on exactly this day, \textit{saying},\lebnote{Literally “to say”}
\verse “Go up to this mountain of the Abarim range, Mount Nebo, which is \textit{opposite Jericho},\lebnote{Literally “on the face of Jericho”} and see the land of Canaan that I am giving to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} as a possession.
\verse You shall die on that mountain that you are about to go up there, and you will be gathered to your people, just as your brother Aaron died on \textit{Mount Hor}\lebnote{Literally “on Hor the mountain”} and he was gathered to his people,
\verse because of the fact that you broke faith with me in the midst of Israel at the waters of Meribah Kadesh, in the desert of Zin, because \textit{you did not treat me as holy}\lebnote{Literally “that not you treated me as holy”} in the midst of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse Yes, from afar you may view the land, but there you shall not enter there, that is, into the land that I am giving to the \textit{Israelites}.”\lebnote{Literally “sons/children of Israel”}
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 33
\verseWithHeading{Blessings of Moses on Israel} Now\lebnote{Or “And”} this is the blessing with which Moses, the man of God, blessed the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} \textit{before}\lebnote{Literally “to the face of”} his death.
\verse Then\lebnote{Or “And”} he said,
\verse “Adonai came from Sinai, 
and he dawned upon them from Seir; 
he shone forth from Mount Paran, 
and he came with myriads of holy ones, 
at his right hand a fiery law for them.\lebnote{This pointing of the MT is very difficult; here HALOT 93 suggests “fire a law for them” as a translation of the compound Hebrew word made up of “fire” + “law,” which I render as “a fiery law”}
\verse Moreover,\lebnote{Or “Indeed”} \textit{he loves his people},\lebnote{Literally “a lover of peoples”} 
all the holy ones were in your hand, 
and they bowed down to\lebnote{Or “at”} your feet, 
\textit{each one accepted directions from you}.\lebnote{Literally “he takes up from your words”}
\verse A law Moses instructed for us, 
as a possession for the assembly of Jacob.
\verse And then a king arose in Jeshurun, 
at the gathering of the leaders of the people, 
\textit{united were the tribes of Israel}.\lebnote{Literally “together the tribes of Israel”}
\verse “May Reuben live, and may he not die, 
\textit{and let his number not be few}.”\lebnote{Literally “let he be his people of number”}
\verse \textit{And he said this of Judah},\lebnote{Literally “and this concerning/to Judah, and he said”} 
“Hear, O Adonai, the voice of Judah, 
and bring him to his people; 
his own hands strive\lebnote{Or “is/are great” on his behalf} for him, 
and may you be a help \textit{against}\lebnote{Literally “from”} his foes.”
\verse And of Levi he said, 
“Your Thummim and your Urim 
are for \textit{your faithful one},\lebnote{Literally “man, your faithful one”} 
\textit{whom you tested at Massah};\lebnote{Literally “whom you tested him at Massah”} 
you contended with him 
at the waters of Meribah.
\verse The one saying of his father and of his mother, 
‘I have not regarded them,’ 
and his brothers he did not acknowledge, 
and his children\lebnote{Or “sons”} he did not know, 
but rather they observed your word, 
and your covenant they kept.
\verse They taught\lebnote{Or “teach/will teach”} your regulations to Jacob, 
and your law to Israel; 
they placed incense smoke \textit{before you},\lebnote{Literally “in your nose”} 
and whole burnt offerings on your altar.
\verse Bless, O Adonai, his substance, 
and with the work of his hands you must be pleased; 
smite the loins of those who attack him, 
and those hating him, \textit{so that they cannot arise}.”\lebnote{Literally “so that not they stand up”}
\verse Of Benjamin he said, 
“The beloved of Adonai dwells \textit{securely},\lebnote{Literally “in confidence”} 
the Most High\lebnote{This is the most likely reading in context (Most High God). Literally the Hebrew text reads “upon him,” so the NASV translates the first line as: dwell in security “by him”} shields all around him, 
all the day, 
and between his shoulders he dwells.”
\verse And of Joseph he said, 
“Blessed by Adonai is his land, 
with\lebnote{Hebrew “from”} the choice things of heaven, 
with\lebnote{Hebrew “from”} dew, and with\lebnote{Hebrew “from”} the deep lying down beneath,
\verse and with\lebnote{Hebrew “from”} the choice things of the fruits\lebnote{Or “produce of”} of the the sun, 
and with the choice things of the yield of the \textit{seasons},\lebnote{Literally “months”}
\verse and with\lebnote{Hebrew “from”} the finest things of the ancient mountains, 
and with\lebnote{Hebrew “from”} the choice things of the \textit{eternal hills},\lebnote{Literally “the hills of eternity never ending ages”}
\verse and with\lebnote{Hebrew “from”} the choice things of the earth and its fullness, 
and the favor of the one \textit{who dwelt}\lebnote{Literally “dwells” but context puts the event into the past; literally “dweller of the bush”} in the bush.\lebnote{Some scholars suggest that this expression must be read as: on Sinai} 
Let them come to the head of Joseph, 
and to the crown of the prince among his brothers.
\verse As the firstborn of his ox, majesty \textit{belongs to him},\lebnote{Literally “is for him”} 
and his horns are the horns of a wild ox; 
with them he drives people together,\lebnote{Or “all at once” (NASB)} 
and they are the myriads of Ephraim, 
and they are the thousands of Manasseh.”
\verse And of Zebulun he said, 
“Rejoice, Zebulun, in your going out, 
and rejoice, Issachar, in your tents;
\verse They summon people to the mountains;\lebnote{Hebrew “mountain”} 
there they sacrifice the sacrifices of righteousness, 
because the affluence of the seas they suck out, 
and \textit{the most hidden treasures of the sand}.”\lebnote{Literally “the covered of the hidden of the sand”}
\verse And of Gad he said, 
“Blessed be \textit{the one who enlarges Gad};\lebnote{Literally “the enlarger of Gad”} 
like a lion he dwells, 
and he tears an arm as well as a scalp.
\verse And he selected\lebnote{Or “provided”} the best part for himself, \lebnote{Hebrew “for him”} 
for there the portion of a ruler is included, 
and he came with the heads of the people;\lebnote{Or possibly, “the heads of the people came/assembled” (compare NEB)} 
he did\lebnote{Or “worked”} the righteousness of Adonai, 
and his regulations for Israel.”
\verse And of Dan he said, 
“Dan is a cub of a lion; 
he leaps from Bashan.”\lebnote{Hebrew “the Bashan,” referring to a well-known geographical and agricultural area}
\verse And of Naphtali, he said, 
“Oh, Naphtali, sated of favor, 
and full of the blessing of Adonai; 
take possession of the lake, 
and the land to the south.”
\verse And of Asher he said, 
“Blessed \textit{more than sons}\lebnote{Literally “blessed from/of sons is” features the comparative use the Hebrew preposition \textit{min}, “than”} is Asher; 
\textit{may he be the favorite}\lebnote{Literally “may he be favored”} of his brothers, 
\textit{dipping his feet in the oil}.\lebnote{Literally “and dipping in the oil his feet”}
\verse Your bars are\lebnote{Or “will be”} iron and bronze, 
and as your days, so is\lebnote{Or “so will be/may be”} your strength.”
\verse “There is no one like God, O, Jeshurun, 
who rides through the heavens to your help, 
and with his majesty through the skies.
\verse The God of \textit{ancient time}\lebnote{Literally “before/earlier times”} is a hiding place, 
and underneath are the arms of eternity,\lebnote{Or “everlasting ages” or “everlasting arms”} 
and he drove out \textit{from before}\lebnote{Literally “from your face”} you your enemy, 
and he said, ‘Destroy them!’
\verse So Israel dwells alone and carefree, 
the spring of Jacob in a land of grain and wine; 
his heavens even drip dew.
\end{biblechapter}

\begin{biblechapter} % Deuteronomy 34
\verseWithHeading{The Death of Moses and the Commissioning of Joshua} Then Moses went up from the desert plateau\lebnote{Or “plains/lowlands”} of Moab to Mount Nebo, to the top of Pisgah, \textit{which is opposite}\lebnote{Literally “which is on the face of”} Jericho, and Adonai showed him all of the land, Gilead all the way up to Dan,
\verse and all of Naphtali and the land of Ephraim and Manasseh and all of the land of Judah, up to the western sea,\lebnote{That is, the Mediterranean Sea}
\verse and the Negev and the plain of the valley of Jericho, the city of palms, on up to Zoar.
\verse And Adonai said to him, “This is the land that I swore to Abraham to Isaac and to Jacob, \textit{saying},\lebnote{Literally “to say”} ‘To your offspring I will give it.’ I have let you see it with your eyes, but you shall not cross \textit{into it}.”\lebnote{Literally “to there”}
\verse Then Moses, the servant of Adonai, died there in the land of Moab \textit{according to the command of Adonai}.\lebnote{Literally “at the mouth of Adonai”}
\verse And he\lebnote{That is, Adonai} buried him in the valley in the land of Moab opposite Beth Peor. But until this day no one knows his burial site.
\verse \textit{Now Moses was a hundred and twenty years old}\lebnote{Literally “and Moses was a son of a hundred and twenty year”} \textit{when he died};\lebnote{Literally “at/in his dying”} \textit{his sight was not impaired and his vigor had not abated}.\lebnote{Literally “not faded his eyes and not had fled away his vitality”}
\verse And the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} wept concerning Moses thirty days; finally the days of weeping and mourning for Moses were completed.
\verse Now\lebnote{Or “And”} Joshua the son of Nun was full of the spirit of wisdom because Moses had placed his hands on him, and the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} listened to him, and they did as Adonai had commanded Moses.
\verse And not again\lebnote{Or “since then”} has a prophet arisen in Israel like Moses, whom Adonai knew\lebnote{Hebrew “knew him”} face to face,
\verse as far as all the signs and the wonders Adonai sent him to do in the land of Egypt, against\lebnote{Hebrew “to”} Pharaoh and all of his servants and against\lebnote{Hebrew “to”} all of his land,
\verse and as far as all of \textit{the mighty deeds}\lebnote{Literally “the hand the strong”} and as far as \textit{the great awesome wonders}\lebnote{Literally “the terrifying the great actions”} Moses did before the eyes of all Israel.
\end{biblechapter}

