\biblebook{Ezekiel}

\begin{biblechapter} % Ezekiel 1
\verseWithHeading{Date of Vision of Living Creatures and Glory of the Throne of Adonai}{And it was in the thirtieth year, in the fourth month, on the fifth day of the month, and I was in the midst of\lebnote{Or “among”}the exiles\lebnote{Hebrew “exile”}by the \textit{Kebar River}.\lebnote{Literally “river of Kebar”}The heavens were opened, and I saw visions of God.}%
\verse{On the fifth day of the month—it was the fifth year of the exile of the king Jehoiachin—}%
\verse{the word of Adonai \textit{came}\lebnote{Literally “was”}clearly\lebnote{Or “certainly”}to Ezekiel the son of Buzi, the priest, in the land of the Chaldeans at the \textit{Kebar River},\lebnote{Literally “river of Kebar”}and the hand of Adonai was on him there.}%
\verse{And I looked,\lebnote{Or “saw”}and look! A storm wind was coming from the north, a great cloud, and fire flashing back and forth, and brightness around and \textit{within it},\lebnote{Literally “for it”}and from its midst\lebnote{Or “middle”}it was like the outward appearance of amber stone from the midst\lebnote{Or “middle”}of the fire.}%
\verse{And from its midst\lebnote{Or “middle”}was the likeness of four living creatures, and this was their appearance: \textit{a human form},\lebnote{Literally “this their appearance the likeness of a human to them”}}%
\verse{and \textit{each had four faces},\lebnote{Literally “and four faces to each”}and \textit{each of them had four wings}.\lebnote{Literally “and four wings to each for them”}}%
\verse{And their legs were straight legs, and the sole of their feet was like the sole of the foot of a calf, and they were sparkling like the outward appearance of polished bronze.}%
\verse{And under their wings were \textit{human hands}\lebnote{Literally “hand his a man”}on their four sides. And their faces and their wings for the four of them were as follows:}%
\verse{\textit{their wings were touching one another};\lebnote{Literally “their wings were touching each her sister”}\textit{each of them went straight forward},\lebnote{Literally “they did not turn at their going”}\textit{without turning right or left}.\lebnote{Literally “each toward the side of his face they went”}}%
\verse{The likeness of their faces was the face of a human in front, and the face of a lion on the right \textit{of each of them},\lebnote{Literally “of the four of them”}and the face of an ox on the left \textit{of each of them},\lebnote{Literally “of the four of them”}and the face of an eagle \textit{for each of them}.\lebnote{Literally “of the four of them”}}%
\verse{So were their faces; their wings were spread out \textit{upward};\lebnote{Literally “from to above”}each had two touching one another and two covering their bodies.}%
\verse{And each \textit{went straight forward};\lebnote{Literally “toward the side of his face they went”}\textit{wherever the spirit went}\lebnote{Literally “towards which it was there the spirit”}they went, and \textit{they did not turn as they went}.\lebnote{Literally “they did not turn at their going”}}%
\verse{As for\lebnote{Or “And”}the likeness of the living creatures, their appearance was like burning coals of fire, like the appearance of torches. It\lebnote{That is, the fire}was moving to and fro between the living creatures, and \textit{the fire was very bright}\lebnote{Literally “brightness was for the fire”}and lightning was going out from the fire.}%
\verse{And the living creatures \textit{were speeding to and fro}\lebnote{Literally “sped forth and returned”}like the appearance of lightning.}%
\verse{And I saw the living creatures, and look! A wheel was on the earth\lebnote{Or “ground”}beside \textit{each of the living creatures that had four faces}.\lebnote{Literally “the living creatures for the four of his/its faces.” The LXX has “for the four of them.”}}%
\verse{The appearance of the wheels and their construction was like the appearance of beryl,\lebnote{Or “chrysolite,” or “topaz”; this stone is difficult to identify exactly}and \textit{they all looked alike},\lebnote{Literally “likeness was one for the four of them”}and their appearance and their construction was \textit{like a wheel within a wheel}.\lebnote{Literally “as that it was a wheel in the midst of a wheel”}}%
\verse{When they moved, they went toward their four sides; they did not \textit{veer at all as they went}.\lebnote{Literally “turn away when they went”}}%
\verse{And their rims \textit{were high and awesome},\lebnote{Literally “height was for them and awesomeness was for them”}and \textit{all four of}\lebnote{Literally “of the four of them”}their rims were full of eyes all around.}%
\verse{And at the going of the living creatures, the wheels next to them also went, and \textit{when the creatures were lifted from the ground}\lebnote{Literally “and at the lifting up of the creatures from the earth”}the wheels also rose.}%
\verse{\textit{Wherever}\lebnote{Literally “Toward which”}the spirit went they would go there, and the wheels rose, for the spirit of the living creatures\lebnote{Hebrew “creature”}was in the wheels.}%
\verse{At their\lebnote{That is, the living creatures}going, they\lebnote{That is, the wheels}go, and at their standing, they stood, and at their being lifted up from on the earth,\lebnote{Or “ground”}the wheels rose,\lebnote{Or “were lifted up”}for the spirit of the living creatures\lebnote{Hebrew “creature”}was in the wheels.}%
\verse{Now\lebnote{Or “And”}the likeness above the heads of the living creatures was an expanse like the outward appearance of \textit{awesome}\lebnote{Literally “terrible”}\textit{ice}\lebnote{Literally “frozen water”}spread out above their heads \textit{upward}.\lebnote{Literally “from to upward”}}%
\verse{And under the expanse their wings were stretched out straight \textit{one toward the other};\lebnote{Literally “each towards its sister”}each had two wings covering them, and each had two wings covering their bodies.}%
\verse{And I heard the sound of their wings like the sound of many waters, like the voice of Shaddai,\lebnote{Often translated “Almighty”}and \textit{when they moved}\lebnote{Literally “at their going”}there was a sound of tumult like the sound of an army; \textit{when they stood still}\lebnote{Literally “at their standing”}they lowered their wings.}%
\verse{And there was\lebnote{Or “it was”}a sound from above the expanse that was above their heads, \textit{and when they stood}\lebnote{Literally “at their standing”}they lowered their wings.}%
\verse{And from above the expanse that was above their heads there was the likeness of a throne, \textit{looking like a sapphire},\lebnote{Literally “the appearance of a stone of sapphire”}and above the likeness of the throne was a likeness similar to\lebnote{Or “like”}the appearance of a human on it, but \textit{above it}.\lebnote{Literally “from to above”}}%
\verse{And I saw something like the outward appearance of amber, something like the appearance of fire, with a covering around it, from the likeness of his loins and \textit{upward}.\lebnote{Literally “to above”}And from the likeness of his loins and \textit{downward}\lebnote{Literally “to down”}I saw something like the appearance of fire, and \textit{it was radiant all around}.\lebnote{Literally “brightness was for it around”}}%
\verse{Like the appearance\lebnote{Or “likeness”}of a bow that is in the cloud on \textit{a rainy day},\lebnote{Literally “the day of the rain”}such was the radiance\lebnote{Or “brightness”}around it; thus was the appearance of the likeness of the glory of Adonai. And I saw, and I fell on my face, and I heard a voice speaking.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 2
\verseWithHeading{The Call of Ezekiel to Speak God’s Words}{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}stand on your feet, so that I can speak with you.”}%
\verse{And the Spirit\lebnote{Or “a spirit”}came into me as he was speaking to me, and it set me on my feet, and I heard the one speaking to me.}%
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}I am sending you to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}to nations\lebnote{Or “a people”}who are rebelling, who rebelled against me, they and their ancestors,\lebnote{Hebrew “fathers”}they transgressed against me \textit{until this very day}.\lebnote{Literally “until the day the this”}}%
\verse{And the children\lebnote{Or “sons”}are \textit{impudent}\lebnote{Literally “hard of face”}and \textit{stubborn},\lebnote{Literally “strong of heart”}and so I am sending you to them, and you must say to them, ‘Thus says the Lord Adonai!’}%
\verse{And they, whether they listen or whether they fail to listen, for they are a rebellious house, they will know that a prophet was in their midst.}%
\verse{And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}you must not be afraid of them, and you must not be afraid of their words, because\lebnote{Or “although”}briers and thorns are with you, and you are sitting among scorpions. You must not be afraid of their words, and you must not be dismayed because of their \textit{looks},\lebnote{Literally “faces”}for they are a rebellious house.}%
\verse{And you must speak my words to them whether they listen or whether they fail to listen, for they are rebellious.}%
\verse{And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}hear what I am speaking to you: you must not be rebellious like the house of rebellion. Open your mouth and eat what I am giving to you.”}%
\verse{And I looked,\lebnote{Or “saw”}and look! There was a hand stretched out to me, and look! In it was \textit{a scroll with writing}.\lebnote{Literally “a scroll of a book”}}%
\verse{And he rolled it out\lebnote{Or “he spread it out”}\textit{before me},\lebnote{Literally “to the face of me”}and it was written on the front and back, and there were written on it laments and mourning and wailing.\lebnote{Or “woe”}}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 3
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}what you find, eat! Eat this scroll, and go, speak to the house of Israel.”}%
\verse{And I opened my mouth, and he gave me this scroll to eat,}%
\verse{and he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}you must give your stomach this to eat, and you must fill your belly with this scroll that I am giving to you.” And I ate, and it became \textit{like sweet honey}\lebnote{Literally “like honey for sweetness”}in my mouth.}%
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}come! Go to the house of Israel, and you must speak to them with my words.}%
\verse{For you are sent to the house of Israel, not to a people \textit{of obscure speech}\lebnote{Literally “obscure of lip”}and \textit{of a difficult language},\lebnote{Literally “heavy/thick of tongue”}}%
\verse{and not to many nations \textit{of obscure speech}\lebnote{Literally “obscure of tongue”}and \textit{a difficult language}\lebnote{Literally “heavy/thick of tongue”}\textit{whose words}\lebnote{Literally “their words”}you do not understand, for if I had sent you to them they would have listened to you.}%
\verse{And the house of Israel, they are not willing to listen to you, for they are not willing to listen to me, for all of the house of Israel is hard of forehead, and they are hard of heart.}%
\verse{But look, I have made\lebnote{Or “set”}your face hard \textit{against}\lebnote{Literally “to correspond to”}their faces and your forehead hard \textit{against}\lebnote{Literally “to correspond to”}their forehead.}%
\verse{Like a diamond \textit{harder than flint}\lebnote{Literally “hard more than flint”}I have made your forehead; you must not fear them, and you must not be dismayed \textit{on account of them},\lebnote{Literally “from the face of them”}for they \textit{are a rebellious house}.\lebnote{Literally “are a house of rebellion”}}%
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}all of my words that I shall speak to you, receive into\lebnote{Hebrew “in”}your heart and hear with your ears.}%
\verse{And come, go to the exiles,\lebnote{Hebrew “exile”}to the children\lebnote{Or “sons”}of your people, and you must speak to them, and say to them, ‘Thus says the Lord Adonai!’ whether they listen or whether they fail to listen.”}%
\verse{And the Spirit lifted me up, and I heard behind me the sound of a great earthquake when the glory of Adonai rose from its place.}%
\verse{And it was the sound of the wings of the living creatures touching lightly\lebnote{Or “brushing against”}\textit{one against the other},\lebnote{Literally “each against each its sister}and the sound of the wheels beside them and the sound of a great earthquake.}%
\verse{And the Spirit lifted me and took me, and I went in bitterness in the heat of my spirit, and the hand of Adonai was strong on me.}%
\verse{And I went to the exiles\lebnote{Hebrew “exile”}at Tel Abib, who were dwelling near the \textit{Kebar River},\lebnote{Literally “river of Kebar”}and I sat where they were dwelling. I sat there seven days in the midst of them, stunned.}%
\verse{And then it happened\lebnote{Or “was”}at the end of seven days, \textit{the word of Adonai came}\lebnote{Literally “and it was the word of Adonai to me to say saying”}to me, saying,}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}I have appointed you as a watchman for the house of Israel. When you hear a word from my mouth, then you must warn them from me.}%
\verse{\textit{When I say}\lebnote{Literally “at/in my saying”}to the wicked, ‘Surely you will die,’ and you do not warn him and you do not speak to warn the wicked from his wicked way \textit{so that he may live},\lebnote{Literally “so that you let him live”}that wicked person will die because of his guilt, and from your hand I will seek his blood.}%
\verse{And you, if you do warn the wicked and he does not turn from his wickedness and from his wicked way, on account of his guilt he will die, and you yourself will have saved your life.\lebnote{Or “soul,” or “self”}}%
\verse{And \textit{when the righteous turns from his righteousness}\lebnote{Literally “and at the to turn of the righteous from his righteousness”}and does injustice, and I place\lebnote{Or “give”}a stumbling block \textit{before him},\lebnote{Literally “to the face of him”}he will die, for you did not warn him. Because of his sin he will die, and his righteousness\lebnote{Or “righteous deeds”}that he did will not be remembered, and his blood I shall seek from your hand.}%
\verse{And if you warn him, the righteous, not to sin, and the righteous does not sin, surely he will live, for he heeded a warning. And you will have saved your life.”\lebnote{Or “soul,” or “self”}}%
\verse{And the hand of Adonai was on me there, and he said to me, “Rise up, go out to the valley, and there I will speak with you.”}%
\verse{And I rose up, and I went to the valley, and, look, there the glory of Adonai was standing, like the glory that I saw near the \textit{Kebar River},\lebnote{Literally “river of Kebar”}and I fell on my face.}%
\verse{And the Spirit\lebnote{Or “a spirit”}came into me, and it made me stand on my feet, and he spoke with me and said to me, “Come, shut yourself \textit{inside your house},\lebnote{Literally “in the midst of your house”}}%
\verse{and you, son of man,\lebnote{Or “mortal,” or “son of humankind”}look, they will place\lebnote{Or “give”}cords on you, and they will tie you up with them. Then you will not go out into the midst of them.}%
\verse{And I will make your tongue cling to the roof of your mouth, and you will be silenced, and you will not be a reproving man for them, for they \textit{are a rebellious house}.\lebnote{Literally “a house of rebellion”}}%
\verse{And when I speak with you, I will open your mouth, and you must say to them, ‘Thus says the Lord Adonai: “The one hearing him, let him hear, and the one failing to hear, let him fail,” ’ for they \textit{are a rebellious house}.\lebnote{Literally “a house of rebellion”}}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 4
\verseWithHeading{Siege of Jerusalem Illustrated on a Brick}{“Now,\lebnote{Or “And”}son of man,\lebnote{Or “mortal,” or “son of humankind”}take for yourself a brick, and you must put it \textit{before you},\lebnote{Literally “to the face of you”}and you must portray on it a city, Jerusalem.}%
\verse{And you must build against it siege works, and you must build against it a bulwark, and you must heap against it a siege ramp, and you must set up against it camps and put against it a battering ram all around.}%
\verse{And take for yourself a plate of iron, and you must place it as a wall of iron between you and the city, and you must set your face against it, and it must be \textit{under siege},\lebnote{Literally “in the siege”}and you must lay the siege against it; it is a sign for the house of Israel.}%
\verse{And you, lie down on your left side, and you must put\lebnote{Or “you will put”}the guilt of the house of Israel on it. You will carry their guilt the number of days that you will lie on it.}%
\verse{And I will give to you the years of their guilt according to the number of the days, three hundred and ninety days, and you must bear the guilt\lebnote{Or “punishment”}of the house of Israel.}%
\verse{When you have completed these, then you must lie a second time on your right side; and you must bear the guilt\lebnote{Or “punishment”}of the house of Judah forty days, a day for each year, a day for each year I give it to you.}%
\verse{And toward the siege of Jerusalem you must set your face and your bared arm; then you must prophesy against it.}%
\verse{Now\lebnote{Or “and”}look! I will put on you cords, and you may not turn yourself from one side to your other side until you complete the days of your siege.}%
\verseWithHeading{Ezekiel’s Strange Defiled Meal}{“And you, take for yourself\lebnote{Hebrew “you”}wheat and barley and beans and lentils and millet\lebnote{Or “sorghum”}and spelt, and you must put them in one vessel, and you must make them for yourself\lebnote{Hebrew “you”}into a food during the number of days that you are lying on your side; three hundred and ninety days you shall eat it.}%
\verse{And your food that you will eat\lebnote{Hebrew “will eat it”}will be according to weight; twenty shekels for each day \textit{at fixed times}\lebnote{Literally “from time to time”}you shall eat it.}%
\verse{And \textit{an amount of water}\lebnote{Literally “water by amount”}you shall drink, a sixth of a hin; \textit{at fixed times}\lebnote{Literally “from time to time”}you shall drink it.}%
\verse{And as a bread-cake of barley you shall eat it, and \textit{with human excrement}\lebnote{Literally “with human dung of the excrement of the human”}you shall bake it before their eyes.”}%
\verse{And Adonai said, “Thus shall the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}eat their unclean food among the nations \textit{where I will scatter them}.”\lebnote{Literally “which I will scatter them there”}}%
\verse{And I said, “Ah, Lord Adonai! Look! I have not been defiling myself, and a dead body and mangled carcass I have not eaten from my childhood until now, and \textit{unclean meat}\lebnote{Literally “flesh of unclean meat”}has not come into my mouth!”}%
\verse{And he said to me, “See I will give you \textit{cattle manure}\lebnote{Literally “the manure of cattle”}in the place of the feces of a human, and you may prepare your food on it.”}%
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}look, I am going to break the \textit{supply}\lebnote{Literally “staff”}of bread in Jerusalem, and they will eat bread by weight, \textit{anxiously},\lebnote{Literally “and with worry”}and \textit{rationed water},\lebnote{Literally “and water by amount”}and they will drink with horror,}%
\verse{so that they will lack food and water, and they will be appalled \textit{with one another},\lebnote{Literally “a man and his brother”}and they will waste away because of their guilt.\lebnote{Or “punishment”}}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 5
\verse{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}take for yourself\lebnote{Hebrew “you”}a sword, sharp as \textit{a barber’s razor}.\lebnote{Literally “the razor of the barbers”}Take it for yourself, and you must cause it to pass over your head and over your beard, and you must take for yourself a set of scales for weighing, and you must divide them.\lebnote{That is, the hairs}}%
\verse{A third you must burn with fire in the midst\lebnote{Or “middle”}of the city at \textit{the completion}\lebnote{Literally “the completing”}of the days of the siege, and you must take a third, and you must strike it with the sword around it, and a third you must scatter to the wind, and I will draw a sword behind them.}%
\verse{And you must take from these a few in number, and you must tuck them in your hem.}%
\verse{And from them again you shall take some, and you must throw them in the middle of the fire, and you must burn them with fire; from it a fire will go out to all of the house of Israel.}%
\verse{Thus says the Lord Adonai: This is Jerusalem in the midst of the nations where I have put her, and countries are around her.}%
\verse{But she has rebelled against my regulations to the point of wickedness more than the nations, and my statutes more than the countries that are around her; for they rejected my regulations, and as for my statutes, they did not walk in them.}%
\verse{Therefore, thus says the Lord Adonai: Because of your commotion more than the nations that are around you—you did not walk in my statutes, and you did not do my regulations, and according to the regulations of the nations that are around you, you did not do.}%
\verse{Therefore thus says the Lord Adonai: Look! I, \textit{even I},\lebnote{Literally “surely I”}am against you, and I will execute judgment in the midst of you before the eyes of the nations,}%
\verse{and I will do with you that which I have not done, and which I will not do again, because of all of your detestable things.}%
\verse{Therefore\lebnote{Literally “to thus”}parents\lebnote{Or “fathers”}will eat children\lebnote{Or “sons”}in the midst of you, and children\lebnote{Or “sons”}will eat their parents,\lebnote{Or “fathers”}and I will execute judgment in you, and I will scatter your entire remnant to every wind.}%
\verse{Therefore \textit{as surely as I live},\lebnote{Literally “alive I am”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, \textit{Surely}\lebnote{Literally “If not”}because you have defiled my sanctuary with all of your vile idols and with all of your detestable things, \textit{now indeed I}\lebnote{Literally “and also I”}will reduce, and my eye will not take pity, and surely I will show no compassion.}%
\verse{A third of you will die because of the plague, and because of the famine they will perish in the midst of you, and a third will fall through the sword around you, and a third I will scatter to every direction of the wind, and I will draw the sword behind them.}%
\verse{And my anger will come to an end, and I will place my rage on them, and I will relent, and they will know that I, Adonai, have spoken in my passion \textit{when I fully vent my rage against them}.\lebnote{Literally “at my finishing my anger against them”}}%
\verse{And I will make you into a desolate place and into a disgrace among the nations that surround you before the eyes of \textit{every one who passes by}.\lebnote{Literally “every one of passing by”}}%
\verse{And it will be\lebnote{Or “and you will be”}an object of taunting\lebnote{Hebrew “taunt”}and an object of mockery, a warning and a horror to the nations that are around you whenever I execute judgments against you in anger and in wrath and in furious punishments! I, Adonai, have spoken!}%
\verse{When I send my arrows of deadly famine against them, which will be as destruction that I will send in order to destroy you, I will increase famine against you, and I will break \textit{the supply of food}\lebnote{Literally “staff of bread”}for you.}%
\verse{And I will send against you famine and fierce animals,\lebnote{Hebrew “animal”}and they will make you childless; and plague and blood will pass through you, and I will bring the sword upon you. I, Adonai, I have spoken!”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 6
\verseWithHeading{Idolatrous Worship and Idolatrous Objects Denounced}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, saying,}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}set your face to the mountains of Israel and prophesy against them,}%
\verse{and you must say, ‘Mountains of Israel, hear the word of the Lord Adonai, thus says the Lord Adonai to the mountains and to the hills, to the ravines and to the valleys: “Look, I am bringing upon you the sword, and I will destroy your high places,}%
\verse{and your altars will be desolate, and your incense altars will be broken, and I will throw down your slain ones \textit{before}\lebnote{Literally “to the face of”}your idols,}%
\verse{and I will place\lebnote{Or “give”}the corpses of the children of Israel \textit{before}\lebnote{Literally “to the face of”}their idols, and I will scatter your bones around your altars.}%
\verse{In all of your dwellings, the cities will be desolate and the high places will be ruined, so that your altars will be desolate and will suffer punishment. Your idols will be broken and will come to an end, and your incense altars will be cut down, and your works will be destroyed,}%
\verse{and the slain one will fall in the midst of you, and then you will know that I am Adonai.}%
\verse{But\lebnote{Or “And”}I will spare some, \textit{so there will be}\lebnote{Literally “at their being”}for you fugitives from the sword among the nations \textit{when you are scattered}\lebnote{Literally “at you to be scattered”}in the countries.}%
\verse{And your fugitives will remember me among the nations to which they were taken captive, that I was shattered by \textit{their adulterous heart}\lebnote{Literally “heart their which was prostituting itself”}which departed from me, and by \textit{their adulterous eyes}\lebnote{Literally “eyes their which were prostituting”}which went after their idols, and they will feel loathing \textit{for themselves},\lebnote{Literally “for their face”}for the evil that they did, for all of their detestable things.}%
\verse{And they will know that I am Adonai; not \textit{in vain}\lebnote{Literally “to nothing”}I spoke to bring to them this evil.” ’}%
\verse{Thus says the Lord Adonai: ‘Clap your hand and stamp with your foot, and say, “Alas, for all of the detestable things of the evil of the house of Israel, \textit{because of which}\lebnote{Literally “which”}they will fall with the sword, with the famine, and with the plague.}%
\verse{The one who is far away will die by the plague, and the one who is near will fall by the sword, and the one who is being left behind and being spared will die by the famine, and I will complete my rage on\lebnote{Or “toward”}them.}%
\verse{And you will know that I am Adonai \textit{when their slain ones are in the midst of their idols}\lebnote{Literally “in the being of their slain ones in the midst of their idols”}around their altars at every high hill, on the tops of all the mountains and under every green tree and under every leafy oak—the place at which they gave a pleasing scent for all of their idols.}%
\verse{And I will stretch out my hand against them, and I will make the land a desolation and a wasteland from the desert to Riblah in all of their dwellings, and they will know that I am Adonai.” ’ ”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 7
\verseWithHeading{Punishment for Abominations Throughout the Land}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, saying,}%
\verse{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}thus says the Lord Adonai to the land of Israel: ‘The end comes, the end on the four corners of the land.}%
\verse{Now the end is on you, and I will send my anger on you, and I will judge you according to your ways, and I will bring on you all your detestable things.}%
\verse{And my eye will not take pity on you, and I will not show compassion for your ways; on you I will bring your detestable things; they will be in the midst of you, and you will know that I am Adonai.’}%
\verse{Thus says the Lord Adonai: ‘Look! Disaster after disaster is coming!}%
\verse{The end comes, comes the end! It has awakened against you! Look! It comes!}%
\verse{Doom is coming against you, the dweller\lebnote{Or “inhabitant”}of the land; the time comes, the day is near, panic and not joy is on the mountains.}%
\verse{\textit{Soon}\lebnote{Literally “now from near”}I will pour out my rage on you, and I will fully vent my anger on you, and I will judge you according to your ways, and I will bring on you all of your detestable things.}%
\verse{And my eye will not take pity, and I will not show compassion. According to your ways I will \textit{deal with you},\lebnote{Literally “on you I will bring”}and your detestable things will be in the midst of you, and you will know that I am Adonai who strikes.\lebnote{Or “smites”}}%
\verse{Look! The day is coming; doom goes out, the staff blossoms, pride sprouts.}%
\verse{Violence has grown to become a staff of wickedness; \textit{none}\lebnote{Literally “not any”}from them will remain, and \textit{none}\lebnote{Literally “not any”}from their abundance nor from their wealth;\lebnote{Or “noise,” or “multitude”; the same word appears in each of the next three verses}and prominence will not be among them.}%
\verse{The time has come, the day has arrived; let not the buyer rejoice, and let the seller not mourn, for anger is on all their multitude.}%
\verse{For the seller will not return to the merchandise \textit{while they are still alive},\lebnote{Literally “while still is in the life, their life”}for the vision is about all of its multitude; it will not change, and a man because of his guilt will not be able to hold onto his life.}%
\verse{They shall blow on the trumpet and prepare \textit{everything},\lebnote{Literally “the all”}but there is no one going to the battle, for my anger is on all of their multitude.}%
\verse{The sword \textit{is outside},\lebnote{Literally “is in the outside place”}and the plague and the famine \textit{are inside};\lebnote{Literally “are from the house}the one who is in the field will die by the sword, and the one who is in the city, famine and plague will devour him.}%
\verse{And if their survivors will escape, they will be on the mountains, like the doves of the valley, all of them groaning, each because of his guilt.}%
\verse{All of the hands will hang limp, and all of \textit{the knees will be wet with urine}.\lebnote{Literally “the knees will flow with water”}}%
\verse{And they will wear sackcloth, and horror will cover them, and on all of the faces will be shame, and baldness on all of their heads.}%
\verse{Their silver they will discard on the streets, and filth will be their gold; their silver and their gold will not be able to rescue them on the day of the wrath of Adonai. They will not satisfy their \textit{hunger}\lebnote{Literally “selves,” or “desire”}and their stomachs they will not fill, for their guilt will be their\lebnote{Hebrew “the”}stumbling block.}%
\verseWithHeading{Judgment for the Profanation of the Temple and Sanctuary}{“ ‘And the beauty of its\lebnote{Or “their”}ornament \textit{he made into a prideful thing}\lebnote{Literally “to a prideful thing he made it”}and made with it their detestable things and their vile idols; therefore I will make it into an impure thing for them.}%
\verse{And I will give it into the hand of strangers as plunder and to the wicked people of the earth as spoil, and they will defile it.}%
\verse{And I will turn my face from them, and they will defile my treasured place, and violent ones will enter and defile it.}%
\verse{Make a chain for the land; it is full of \textit{bloody crimes}\lebnote{Literally “the judgment of blood”}and the city is full of violence.}%
\verse{And I will bring the worst of the nations, and they will take possession of their houses, and I will put to an end the pride of the mighty ones, and their sanctuaries will be defiled.}%
\verse{Anguish comes, and they will seek peace, and there will be none.}%
\verse{Calamity upon calamity will come, and rumor will be upon rumor. And they will seek a vision from a prophet, but instruction will perish from the priest and counsel from the elders.}%
\verse{The king will mourn, and the prince will be dressed with despair, and the hands of the people of the land will tremble. According to their way I will do to\lebnote{Or “deal with”}them, and according to their judgments I will judge them, and they will know that I am Adonai.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 8
\verseWithHeading{Abominations in the Temple and in Jerusalem}{\textit{And then}\lebnote{Literally “and it happened”}in the sixth year, in the sixth month, on the fifth day of the month, I was sitting in my house, and the elders of Judah were sitting \textit{before me}.\lebnote{Literally “to the face of me”}And the hand of the Lord Adonai fell on me there,}%
\verse{and I saw,\lebnote{Or “looked”}and look! A figure like the appearance of a man;\lebnote{Hebrew “fire”; but reading the BHS note 2b, “a man.”}from the appearance of his waist and below was fire, and from his waist and \textit{upward}\lebnote{Literally “to above”}was like the appearance of brightness,\lebnote{Or “shining metal”}like the \textit{outward appearance of}\lebnote{Literally “eye”}amber.}%
\verse{And he sent out the form of a hand, and he took me by a lock of hair of my head, and the Spirit lifted me between earth and heaven, and it brought me to Jerusalem in visions of God to the doorway of the inner gate \textit{that faced north},\lebnote{Literally “the one facing north”}at which there was the seat of the image of jealousy, which was making jealous.}%
\verse{And look! There was the glory of the God of Israel like the vision that I saw in the valley.}%
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}lift up now\lebnote{Particle of entreaty, “please/ please now”}your eyes \textit{toward the north}.”\lebnote{Literally “to the way of the north”}And I lifted up my eyes \textit{toward the north},\lebnote{Literally “to the way of the north”}and, look, there was to the north of the gate of the altar this image of jealousy at the entrance.}%
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}“\textit{Do you see}\lebnote{Literally “Are you seeing?”}what they are doing—great detestable things that the house of Israel is committing\lebnote{Or “doing”}here so as to drive me from\lebnote{Hebrew “from upon”}my sanctuary, and yet \textit{you will see again}\lebnote{Literally “you will return you will see”}greater\lebnote{Or “worse”}detestable things.”}%
\verse{And he brought me to the doorway of the courtyard, and I saw, and look! A hole in the wall.}%
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}dig now\lebnote{Or “please”}through the wall.” And I dug through the wall, and look! There was a doorway.}%
\verse{And he said to me, “Come and see the detestable things, the evil that they are doing\lebnote{Or “making”}here.}%
\verse{And I came, and I saw, and look, all kinds of creatures\lebnote{Hebrew “creature”}and detestable beasts;\lebnote{Hebrew “beast”}and all of the idols of the house of Israel were carved on the wall \textit{all around}.\lebnote{Literally “around around”}}%
\verse{And seventy men from the elders of the house of Israel, and Jaazaniah, the son of Shaphan, was standing in the midst of them, and they were standing \textit{before them}.\lebnote{Literally “to the face of them”}Each one had his censer in his hand, and the fragrance of the cloud of the incense was going up.}%
\verse{And he said to me, “Have you seen, son of man,\lebnote{Or “mortal,” or “son of humankind”}what the elders of the house of Israel are doing in the dark, each in the inner rooms of his idol, for they are saying, ‘Adonai is not seeing us; Adonai has abandoned the land.’ ”}%
\verse{And he said to me, “Still \textit{you will see again}\lebnote{Literally “you will return, you will see”}greater detestable things that they are doing.”}%
\verse{And he brought me to the doorway of the gate of the house of Adonai that is toward the north, and look! There were the women sitting weeping for Tammuz.}%
\verse{And he said to me, “Have you seen, son of man?\lebnote{Or “mortal,” or “son of humankind”}Still \textit{you will see again}\lebnote{Literally “you will return, you will see”}greater detestable things than these.”}%
\verse{And he brought me to the inner courtyard of the house of Adonai, and look, at the doorway of the temple of Adonai, between the portico and the altar, there were about twenty-five men with their backs to the temple of Adonai and their faces toward the east, and they were bowing down toward the east before the sun.}%
\verse{And he said to me, “Have you seen, son of man?\lebnote{Or “mortal,” or “son of humankind”}\textit{Was it too small a thing}\lebnote{Literally “was it too little”}for the house of Judah \textit{to do}\lebnote{Literally “from doing”}the detestable things that they did here? For they filled up the land with violence, and \textit{they provoked me to anger again},\lebnote{Literally “they returned to provoke me to anger”}and look! They are putting the branch to their nose.}%
\verse{And so I will act in rage, and my eye will not take pity, and I will not have compassion, and they will cry in my ear with a loud voice, and I will not hear them.”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 9
\verseWithHeading{Adonai’s Avenging Messenger Destroys the Wicked in the City}{And he called in my ears with a loud voice, \textit{saying},\lebnote{Literally “to say”}“They have come near,\lebnote{Or “Come near” (imperative)}the punishers of the city, and each with his weapon of destruction in his hand.}%
\verse{And look! Six men coming from the way of \textit{the upper gate}\lebnote{Literally “the gate of the upper”}that faced \textit{northward},\lebnote{Literally “faced north + directive”}and each with his weapon for\lebnote{Or “of”}shattering in his hand; and one man was in the midst of them, dressed in linen, and the\lebnote{Or “his”}writing case of the scribe was at his side. And they came and stood beside the bronze altar.}%
\verse{And the glory of the God of Israel lifted itself up from upon the cherub that he was on\lebnote{Hebrew “which he was on it”}and went to the threshold of the house,\lebnote{That is, the temple}and he called to the man who was clothed in linen \textit{with a scribal writing case at his side}.\lebnote{Literally “whom a writing case of a scribe was at his side”}}%
\verse{And Adonai said to him, “Go through in the midst of the city, in the midst of Jerusalem, and you must place a mark on the foreheads of the men \textit{who are groaning and lamenting}\lebnote{Literally “the men the groaning and the lamenting”}about all of the detestable things \textit{that are being done}\lebnote{Literally “the being done”}in the midst of her.”}%
\verse{And to the others he said in my hearing, “Go through the city behind him and kill! Your eyes shall not take pity, and you shall not have compassion.}%
\verse{You must kill \textit{totally}\lebnote{Literally “to destruction”}old man, young man and young woman, and little children and women, but \textit{concerning}\lebnote{Literally “upon”}every man \textit{with the mark on him}\lebnote{Literally “which is on him the mark”}you must not approach; and from my sanctuary you must begin.” And they began with the old men who \textit{were before}\lebnote{Literally “were to the face of”}the house.\lebnote{That is, the temple}}%
\verse{And he said to them, “Defile the house\lebnote{Or “temple”}and fill the courtyards with the dead; go out! And they went out, and they killed in the city.}%
\verse{\textit{And then}\lebnote{Literally “and it was”}\textit{as they were striking},\lebnote{Literally “in their striking”}and I was left behind, I fell on my face, and I cried out, and I said, “Ah, Lord Adonai! “Will you be destroying all of the remnant of Israel \textit{while you pour out your rage}\lebnote{Literally “when pouring out your rage”}on Jerusalem?”}%
\verse{And he said to me, “The guilt of the house of Israel and Judah is \textit{exceedingly}\lebnote{Literally “with very very”}great, and the land is filled with bloodguilt, and the city is full of injustice. For they say Adonai abandoned the land, and \textit{Adonai does not see}.\lebnote{Literally “not Adonai is seeing}}%
\verse{And I, my eye will not take pity, and I will not have compassion; their way I will bring on their head.”}%
\verse{And look! The man clothed in linen and \textit{with a writing case at his side}\lebnote{Literally “which/whom the writing-case was at his side”}was bringing back a word, \textit{saying},\lebnote{Literally “to say”}“I have done all that you commanded me.”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 10
\verseWithHeading{God’s Presence Leaves the Temple}{And I looked, and look! On the expanse that was above the head of the cherubim something like a stone of sapphire,\lebnote{Or “lapis lazuli”}and like the appearance of the shape of a throne it appeared above them.}%
\verse{And he spoke\lebnote{Or “said”}to the man clothed in linen and said, “Go \textit{in among the wheel area under the cherubim}\lebnote{Literally “to between to the wheel to under the cherub”}and fill the hollow of your hands with coals of fire from among the cherubim, and toss them on the city.” And he went right before my eyes.}%
\verse{Now\lebnote{Or “And”}the cherubim were standing on the south of the temple \textit{when the man went},\lebnote{Literally “at his coming the man”}and the cloud filled the inner courtyard.}%
\verse{And the glory of Adonai rose up from on the cherub toward\lebnote{Or “onto”}the threshold of the house,\lebnote{Or “temple”}and the house\lebnote{Or “temple”}was filled with the cloud, and the courtyard was filled with the brightness of the glory of Adonai.}%
\verse{And the sound of the wings of the cherubim was heard \textit{as far as}\lebnote{Literally “up to”}the outer courtyard; it was like the voice of God Shaddai\lebnote{Often translated “Almighty”}\textit{when he speaks}.\lebnote{Literally “in his speaking”}}%
\verse{And then at his command\lebnote{Or “when he commanded”}to the man clothed in linen, \textit{saying},\lebnote{Literally “to say”}“Take fire from among the wheel area from among the cherubim,” he went and stood beside the wheel.}%
\verse{Then the cherub stretched out his hand from among\lebnote{Or “between”}the cherubim toward the fire that was among\lebnote{Or “between”}the cherubim, and he lifted up and gave\lebnote{Or “placed”}it into the hollow of the hand of the man clothed with linen, and he took it, and he went out.}%
\verse{And there appeared for\lebnote{Literally “to”}the cherubim the form of \textit{a human hand}\lebnote{Literally “a hand of a man”}under their wings.}%
\verse{And I saw, and look, four wheels beside the cherubim, \textit{one wheel beside each cherub},\lebnote{Literally “wheel one beside the cherub one, and wheel one beside the cherub one”}and the appearance of the wheels was like the outward appearance of \textit{turquoise stone}.\lebnote{Literally “stone of turquoise”}}%
\verse{Their appearance\lebnote{That is, of the wheels}\textit{was the same for each of the four of them},\lebnote{Literally “likeness one was for the four of them”}as if the wheel was in the midst of a wheel.}%
\verse{\textit{When they went}\lebnote{Literally “In their going”}to the four of their directions that they went, they did not change direction \textit{when they went},\lebnote{Literally “At their going”}for the place to which the head\lebnote{Or “first” or “lead wheel”}turned, they went behind him; they did not change direction at their going.}%
\verse{And \textit{their whole body},\lebnote{Literally “the whole of their body”}and their rims, and their spokes, and their wings, and the wheels were full of eyes all around—\textit{the wheels for the four of them}.\lebnote{Literally “for the four of them their wheels”}}%
\verse{Concerning the wheels, he was calling them “the wheelwork” \textit{in my hearing}.\lebnote{Literally “in my ears”}}%
\verse{And \textit{each one had four faces};\lebnote{Literally “four faces were to each”}the face of the one was the face of a cherub, and the face of the second was the face of a human, and the face of the third was the face of a lion, and the fourth was the face of an eagle.}%
\verse{And the cherubim rose; that is, the living creatures\lebnote{Hebrew “creature”}that I saw at the \textit{Kebar River}.\lebnote{Literally “river of Kebar”}}%
\verse{And \textit{when the cherubim went},\lebnote{Literally “at the going of the cherubim”}the wheels beside them went; and \textit{when the cherubim lifted their wings to rise up from the earth},\lebnote{Literally “and at the lifting of the cherubim wings their to rise from on the earth”}the wheels also did not turn aside from beside them.}%
\verse{\textit{When they stood},\lebnote{Literally “at their standing”}they stood, and at their rising, they rose with them, for the spirit of the living creatures was in them.}%
\verse{And the glory of Adonai went out from\lebnote{Hebrew “from on”}the threshold of the temple, and it stood above the cherubim.}%
\verse{And the cherubim lifted up their wings, and they rose from the earth before my eyes. At their going out, the wheels were\lebnote{Hebrew “to”}beside them. And he stood at the doorway of the eastern gate of the temple\lebnote{Or “house”}of Adonai, and the glory of the God of Israel \textit{was over them}.\lebnote{Literally “on them from to above”}}%
\verse{This was\lebnote{Or “these were”}the living creature that I saw under the God of Israel at the \textit{Kebar River},\lebnote{Literally “river of Kebar”}and I knew that they were cherubim.}%
\verse{\textit{Each one had four faces},\lebnote{Literally “four faces were for one/each”}and there were four wings for each, and the likeness of \textit{human hands}\lebnote{Literally “hands of a man”}was under their wings.}%
\verse{And the likeness of their faces, they were the faces that I saw at the \textit{Kebar River};\lebnote{Litearlly “river of Kebar”}thus was their appearance, and they each went \textit{straight ahead}.\lebnote{Literally “to the opposite of face his”}}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 11
\verseWithHeading{Judgment On Evil Rulers}{And the Spirit lifted me up, and it brought me to the eastern gate, the one facing east, of the temple\lebnote{Or “house”}of Adonai. And look, there were twenty-five men in the doorway of the gate, and I saw Jaazaniah the son of Azzur in the midst of them, and Pelatiah the son of Benaiah, the commanders of the people.}%
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}these are the men \textit{who devise mischief},\lebnote{Literally “the ones devising mischief”}and who are offering bad counsel in this city,}%
\verse{who are saying, “The building of houses is not near; the city is the pot, and we are the flesh.}%
\verse{Therefore prophesy against them! Prophesy, son of man!”\lebnote{Or “mortal,” or “son of humankind”}}%
\verse{And the Spirit of Adonai fell on me, and he said to me, “Say, ‘thus says Adonai: “This is what you think, house of Israel, and I myself know them,\lebnote{Hebrew “it”}the thoughts of your spirit.}%
\verse{You made your slain ones numerous in this city, and you have filled its streets with slain ones.” ’\lebnote{Hebrew “slain one”}}%
\verse{Therefore thus says the Lord Adonai: “Your slain ones whom you put in its midst, they are the flesh, and it is the pot, and I\lebnote{Hebrew “he”}will bring you out from its midst.}%
\verse{The sword you have feared, and the sword I will bring against you!” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“And I will bring you out from its midst, and I will give you into the hand of strangers, and I will execute judgments against you.}%
\verse{By the sword you will fall at the border of Israel; I will judge you, and you will know that I am Adonai.}%
\verse{It will not be as a pot to you, and so you would be\lebnote{Or “will be”}in the midst of it as flesh, for at the border of Israel I will judge you.}%
\verse{And you will know that I am Adonai, whose\lebnote{Hebrew “which”}rules you did not follow, and whose\lebnote{Hebrew “my”}regulations you did not do, but according to the regulations of the nations that are around you, you acted.” ’\lebnote{Or “did”}}%
\verse{And it happened that \textit{as I was prophesying},\lebnote{Literally “at my prophesying”}Pelatiahu the son of Benaiahu died! And I fell on my face, and I cried with a loud voice, and I said, “Ah, Lord Adonai, you are making a complete destruction of the remnant of Israel!”}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Or “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}your brothers, your brothers, the men of your redemption, and all of the house of Israel, all of it, \textit{who said concerning the inhabitants of Jerusalem},\lebnote{Literally “which they said concerning them the inhabitants of Jerusalem”}‘They are far from Adonai, therefore to us this land was given as a possession.’}%
\verse{Therefore say, ‘Thus says the Lord Adonai: “Though I have removed them far away among the nations and though I have scattered them among the countries, yet I was a sanctuary to them for a little while in the countries \textit{to which they have gone}.’ ”\lebnote{Literally “which they came there”}}%
\verse{Therefore say, ‘Thus says the Lord Adonai: “And I will assemble you from the peoples, and I will gather you from the countries to which you were scattered among them, and I will give the land of Israel to you.}%
\verse{And when they come there, then they will remove all of its vile idols and all of its detestable things from it.}%
\verse{And I will give to them one heart, and a new spirit I will give\lebnote{Or “put”}in their inner parts. And I will remove their\lebnote{Hebrew “the”}heart of stone from their body, and I will give to them a heart of flesh,}%
\verse{so that they may walk in my statutes, and they will keep my regulations, and they will do them, and they will be to me a people, and I myself will be to them as God.}%
\verse{But to\lebnote{Or “and to”}the heart of their abominations and the detestable things their heart is going. I will bring their way on their head,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.’ ”}%
\verse{Then\lebnote{Or “And”}the cherubim lifted up their wings, and their\lebnote{Or “the”}wheels were beside them, and the glory of the God of Israel \textit{was over them}.\lebnote{Literally “from to above”}}%
\verse{And the glory of Adonai went up from the midst of the city, and it stood still on the mountain that is to the east of the city.}%
\verse{And the Spirit lifted me up, and it brought me to Chaldea, to the exiles,\lebnote{Hebrew “exile”}in the vision by the spirit of God; and the vision that I had seen \textit{left me}.\lebnote{Literally “went up from being upon me”}}%
\verse{And I spoke to the exiles\lebnote{Hebrew “exile”}all of the words of Adonai that he had shown me.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 12
\verseWithHeading{The Announcement of the Imminent Coming of the Exile}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}you are dwelling in the midst of the house of rebellion \textit{who has eyes to see and they do not see};\lebnote{Literally “which eyes are for them to see and not they see”}\textit{they have ears to hear},\lebnote{Literally “ears are for them to hear”}and they do not hear, for they are a house of rebellion.}%
\verse{And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}prepare for yourself\lebnote{Hebrew “you”}the baggage of an exile, and go into exile by day before their eyes. And you must go into\lebnote{Hebrew “in”}exile from your place to another place before their eyes; perhaps they will see that they are a house of rebellion.}%
\verse{And you must bring out your baggage like the baggage of an exile by day before their eyes, and you must go out in the evening before their eyes like \textit{those who go into exile}.\lebnote{Literally “like the goers out of an exile”}}%
\verse{Before their eyes dig through for yourself,\lebnote{Hebrew “you”}through the wall, and you must bring the baggage out through it.}%
\verse{Before their eyes, on your\lebnote{Literally “the”}shoulder, you must lift up the baggage in the dusk, and your face you must cover, so that you may not see the land, for I make you as a sign to the house of Israel.”}%
\verse{And I did \textit{just as}\lebnote{Literally “so as that”}I was commanded; my baggage, like the baggage of an exile, I brought out by day, and in the evening I dug through for myself\lebnote{Hebrew “me”}into\lebnote{Hebrew “in”}the wall with my hand in the dusk; I brought the baggage on my shoulder; I carried it before their eyes.}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me in the morning, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}did not they, the house of Israel, the house of rebellion, say to you, ‘What are you doing?’}%
\verse{Say to them, ‘Thus says the Lord Adonai: “This oracle is about the prince in Jerusalem and \textit{the entire}\lebnote{Literally “all of”}house of Israel \textit{who are among them}.” ’\lebnote{Literally “which they are in the midst of them”}}%
\verse{Say, ‘I am your sign, and \textit{just as}\lebnote{Literally “as that”}I did, so will it be done to them in the exile; into\lebnote{Hebrew “in”}captivity they will go.’}%
\verse{And the prince who is in the midst of them, on his shoulder he will carry the baggage in the dusk, and\lebnote{Or “in the dusk he will go out”}he will go out; \textit{the wall will be dug through}\lebnote{Literally “through the wall they will dig”}to bring him through it; and he will cover his face so that he will\lebnote{Or “may”}not see the land with his\lebnote{Hebrew “the”}eye.}%
\verse{And I will spread out my net on him, and he will be captured in my hunting snare, and I will bring him to Babylon, the land of the Chaldeans, but he will not see it, and there he will die.}%
\verse{And all who are around him, his help\lebnote{Or “support”}and all of his troops I will scatter \textit{in every direction},\lebnote{Literally “to every direction of the wind”}and I will draw\lebnote{Or “unsheathe”}the sword behind them.}%
\verse{And they will know that I am Adonai \textit{when I scatter them}\lebnote{Literally “at my scattering them”}among the nations and I scatter them in the countries.}%
\verse{But\lebnote{Or “And”}I will spare from them \textit{a few men}\lebnote{Literally “men of number”}from the sword, from the famine and from the plague, so that they may tell of all their abominations among the nations to which they will go, and they will know that I am Adonai.”}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}you must eat your food with trembling, and your water with shuddering, and with anxiety you must drink.}%
\verse{And you must say to the people of the land, ‘Thus says the Lord Adonai to the inhabitants of Jerusalem about the land of Israel: “They will eat their food with anxiety and their water they will drink with horror, because their land will be desolate from\lebnote{Or “of”}its fullness because of all the violence of \textit{those who are dwelling in it}.\lebnote{Literally “the ones dwelling in it”}}%
\verse{And \textit{the inhabited cites will be desolate},\lebnote{Literally “the being inhabited cities they will be desolate”}and the land will be a desolation, and you will know that I am Adonai.” ’ ”}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}what is this proverb \textit{you people have}\lebnote{Literally “to you”}about the land of Israel, \textit{saying},\lebnote{Literally “to you”}‘\textit{The days are prolonged},\lebnote{Literally “they are long the days”}and every vision has come to nothing.’}%
\verse{Therefore say to them, ‘Thus says the Lord Adonai: “I will put an end to this proverb, and they will not quote it as a proverb again in Israel.” ’ But\lebnote{Hebrew “But if”}say to them, ‘The days are near, and also the word of every vision.’}%
\verse{For there will not be any longer \textit{any false vision}\lebnote{Literally “every vision of falseness”}or flattering divination in the midst of the house of Israel.}%
\verse{For I, Adonai, I will speak what I will speak as a word, and it will be done. It will not prolong itself any longer, for in your days, house of rebellion, I will speak a word and I will fulfill it!” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}look! The house of Israel is saying, ‘The vision that he is seeing is for many days from now, and for distant times he is prophesying.’}%
\verse{Therefore say to them, ‘Thus says the Lord Adonai: “None of my words will be prolonged any longer that I speak as a word, and it will be fulfilled!” ’ ” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 13
\verseWithHeading{Condemnation of False Prophets}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Or “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}prophesy to\lebnote{Or “against”}the prophets of Israel \textit{who are prophesying},\lebnote{Literally “the ones prophesying”}and you must say to those who are prophets out of their own \textit{imagination},\lebnote{Literally “heart”}‘Hear the word of Adonai!’}%
\verse{Thus says the Lord Adonai: ‘Alas, for the foolish prophets who are going after their own spirit, \textit{and they did not see anything}!\lebnote{Literally “and not they saw”}}%
\verse{Your prophets have been like foxes\lebnote{Or “jackals”}among ruins.}%
\verse{You did not go up into\lebnote{Hebrew “in”}the breaches and repair a wall for the house of Israel to stand firm in the battle on the day of Adonai.}%
\verse{They saw falseness and \textit{a lying divination},\lebnote{Literally “a divination of a lie”}the ones saying, ‘Declaration of Adonai!’ and Adonai did not send them, and they wait \textit{for the confirmation of their word}.\lebnote{Literally “to confirm a word”}}%
\verse{Have you not seen \textit{a false vision}\lebnote{Literally “vision of falseness”}and spoken a \textit{lying divination},\lebnote{Literally “divination of a lie”}and you said, ‘Declaration of Adonai!’ but \textit{I myself did not speak}.\lebnote{Literally “I not I spoke”}}%
\verse{Therefore thus says the Lord Adonai: ‘Because of your speaking falseness and because you envisioned a lie, therefore, look! I am against you!” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“And it will be my hand against the prophets who are seeing falseness and who are practicing lying divination. They will not be in the council of my people, and they will not be written down in the record book of the house of Israel, and into\lebnote{Hebrew “in”}the land of Israel they will not come, and then you will know that I am the Lord Adonai.}%
\verse{Because, yes, because they led my people astray, \textit{saying}\lebnote{Literally “to say”}‘Peace!’ And there is not peace. And when anyone\lebnote{Hebrew “he”}builds a flimsy wall, look, they coat it with whitewash.}%
\verse{Say to those covering it with whitewash that it will fall; there will be \textit{a torrent of rain},\lebnote{Literally “rain flooding”}and I will give stones of hail; they will fall! \lebnote{Or “and I will give great hailstones. They will fall”}And a \textit{windstorm}\lebnote{Literally “wind of a storm”; or “with respect/concerning you stones of hail, they will fall”}will burst forth!}%
\verse{And look! When the wall falls, will it not be said to you, ‘Where is the whitewash with which you covered it?’}%
\verse{Therefore thus says the Lord Adonai: ‘And so I will let burst forth a \textit{windstorm}\lebnote{Literally “a wind of storm”}in my rage, and \textit{there will be a torrent of rain}\lebnote{Literally “rain flooding there will be”}in my anger, and hailstones in my rage \textit{for complete destruction}.\lebnote{Literally “to complete destruction”}}%
\verse{And I will break down the wall that you covered with whitewash, and I will knock it to the ground, and its foundation will be revealed, and it will fall, and you will come to an end in the midst of it, and you will know that I am Adonai!}%
\verse{And I will fully vent my rage against the wall and against those covering it with whitewash, and I will say to you, “\textit{The wall is no more},\lebnote{Literally “not is the wall”}and \textit{the people covering it are no more},\lebnote{Literally “not are the people whitewashing him”}}%
\verse{that is, the prophets of Israel, the ones prophesying concerning Jerusalem and the ones seeing visions of peace, and there is not peace!’ ” ” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}place your face toward the daughters of your people who prophesy from their \textit{imagination},\lebnote{Literally “heart”}and prophesy against them!}%
\verse{And you must say, ‘Thus says the Lord Adonai: “Woe to \textit{those who sew}\lebnote{Literally “the sewers of”}magic charm bands on all the wrists\lebnote{Or “joints”}of the hands of my people and \textit{those who make the veils}\lebnote{Literally “the makers of the veils”}that are on the head of people of every height, to ensnare people’s lives!\lebnote{Or “souls”}Will you ensnare the lives\lebnote{Or “souls”}of my people \textit{and keep yourselves alive}?\lebnote{Literally “and lives/souls for you, you keep alive”}}%
\verse{And you defiled me among my people for a handful of barley and for morsels of bread to kill persons who should not die and to keep alive persons who should not live by means of your lies to my people who are listening to your lies.” ’\lebnote{Hebrew “lie”}}%
\verse{Therefore thus says the Lord Adonai: ‘Look! I am against your magic charm bands \textit{with which you are ensnaring persons}\lebnote{Literally “which you are ensnaring with them the souls/individuals”}as birds; I will tear them from your arms, and I will release the persons that you are ensnaring, treating persons as birds.}%
\verse{And I will tear off your veils, and I will deliver my people from your hand, and they will not any longer be in your hand as prey, and you will know that I am Adonai.}%
\verse{Because you disheartened the heart of the righteous by deception, and I have not\lebnote{Or “did not”}caused him pain, and strengthened the hands of the wicked so that he did not turn\lebnote{Or “not to turn”}from his wicked way to save his life.}%
\verse{Therefore falseness\lebnote{Or “vanity” or “vain/false visions”}you will not see,\lebnote{Or “observe”}and divination you will not practice\lebnote{Or “you will indeed not practice divination”}any longer, and I will rescue my people from your hand, and you will know that I am Adonai!’ ”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 14
\verseWithHeading{Condemnation of Israel’s Idolatrous Elders}{And men from the elders of Israel came to me, and they sat \textit{before me}.\lebnote{Literally “to the face of me”}}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}these men took up their idols into their heart and \textit{they placed before themselves}\lebnote{Literally “they gave before their faces”}a stumbling block of their guilt.\lebnote{Or “iniquity”}\textit{Should I really let myself be consulted by them}?\lebnote{Literally “Surely I will let me be consulted by them?”}}%
\verse{Therefore speak with them, and you must say to them, ‘Thus says the Lord Adonai: “Every person from the house of Israel who brings up his idols into his heart and \textit{places before himself}\lebnote{Literally “they gave before their faces”}a stumbling block of his guilt\lebnote{Or “iniquity”}and yet he comes to the prophet, I Adonai, I will respond to him through this act\lebnote{Hebrew “it”}with respect to the multitude\lebnote{According to the \textit{Kethib}; the \textit{Qere} reads “I will respond to him, he who comes with the multitude of his idols”}of his idols,}%
\verse{so that I may take hold of the house of Israel by their heart, those who are estranged from me, all of them, through their idols.’}%
\verse{Therefore say to the house of Israel, ‘Thus says the Lord Adonai: “Return,\lebnote{Or “Repent”}and turn away from your idols and from all of your detestable things! Turn away your faces!” ’}%
\verse{For \textit{each person}\lebnote{Literally “a man, a man” or “each one, each one”}from the house of Israel, and from the alien who dwells as an alien in Israel, who separates himself \textit{from following me}\lebnote{Literally “from behind me”}and brings up his idols into his heart and \textit{places before himself}\lebnote{Literally “he puts beside his face”}a stumbling block of his guilt\lebnote{Or “iniquity”}, and yet he comes to the prophet to consult him concerning me, I Adonai will answer\lebnote{Or “am answering”}him myself.}%
\verse{And I will set my face against that person, and I will make him to\lebnote{Or “as”}be a sign and make him into\lebnote{Hebrew “to”}the subject of proverbs; and I will cut him off from the midst of my people, and you will know I am Adonai.}%
\verse{And the prophet, if he is deceived and he speaks a word, I Adonai, I have deceived that prophet, and I will stretch out my hand against him, and I will destroy him from the midst of my people Israel.}%
\verse{And they will bear their guilt, like\lebnote{Or “as”}the guilt of the inquirer so\lebnote{Or “like”}the guilt of the prophet will be,}%
\verse{so that the house of Israel will not go astray again \textit{from me},\lebnote{Literally “from behind me”}and they will not make themselves unclean again with all of their transgressions, and they will be for me a people, and I will be for them as God,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verseWithHeading{Jerusalem will not be Spared}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}when a land sins against me by \textit{acting very unfaithfully},\lebnote{Literally “acting unfaithfully doing infidelity”}then I will stretch out my hand against it, and I will break for it the \textit{supply of food},\lebnote{Literally “staff of bread”}and I will send against it famine, and I will cut it off, both human and animal.}%
\verse{And if even the three of these men were in the midst of it—Noah, Daniel, and Job—they, through their righteousness, would save only \textit{themselves}!”\lebnote{Literally “their souls”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“If a fierce animal I should let cross through the land, and it should make it childless and it will be a desolation, so that there will not be one crossing over the land \textit{due to}\lebnote{Literally “from the face of”}the presence of the animal,}%
\verse{even if these three men were in the midst of it, \textit{as surely as I live},”\lebnote{Literally “live I”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, “surely they will not save sons and daughters; \textit{they themselves alone},\lebnote{Literally “they to alone them”}they will be saved, but the land will be a desolation.}%
\verse{Or, if I bring a sword over that land, and I say, ‘Sword, let it cross over into\lebnote{Hebrew “in”}the land!’ And I will cut off from it both human and animal.}%
\verse{And if these three men were in the midst of it, \textit{as surely as I live},”\lebnote{Literally “live I”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, “they will not save sons and daughters, but \textit{they alone}\lebnote{Literally “they to alone them”}will be saved.}%
\verse{And if I were to send a plague to that land, and I pour out my rage on it with blood to cut it off, both human and animal,}%
\verse{and if Noah, Daniel, and Job were in the midst of her, \textit{as surely as I live},”\lebnote{Literally “live I”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, “surely not a son, surely not a daughter will they save by their righteousness; they would save \textit{themselves}.”\lebnote{Literally “their souls”}}%
\verse{For thus says the Lord Adonai: “How much more when I send my four punishments—the evil sword, and famine, and a fierce animal, and a plague—to Jerusalem to cut it off, both human and animal!}%
\verse{But\lebnote{Or “And”}look! A remnant will be left over in it, sons and daughters \textit{who will be brought out}.\lebnote{Literally “the ones being brought out”}Look! They are coming out to you, and you will see their way, and with their deeds you will be consoled with respect to\lebnote{Hebrew “for”}the evil that I brought over Jerusalem, all of it that I brought over\lebnote{Or “upon”}it.}%
\verse{And they will console you when you see their way and their deeds, and you will know that not for nothing I did all that I did in it,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 15
\verseWithHeading{Jerusalem a Useless Vine}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man, how will the wood of the vine be \textit{better than}\lebnote{Literally “more than”}\textit{any of the wood of}\lebnote{Literally “all of the wood of”}the branch which is among the trees of the forest?\lebnote{Or “wood”}}%
\verse{Can wood be taken from it to make anything, or can they take a tent peg from it to hang on it any object?\lebnote{Or “vessel”}}%
\verse{Look! It is given to the fire as fuel. The fire will consume two of its ends, and its middle will be charred. Is it useful for anything?}%
\verse{Look! \textit{When it is in perfect condition},\lebnote{Literally “In being it perfect”}it will not be used for anything, \textit{how much less}\lebnote{Literally “indeed”}\textit{when the fire has consumed it}\lebnote{Literally “for if fire has consumed it”}and it is charred; and then can it be used again for anything?}%
\verse{Therefore thus says the Lord Adonai: \textit{Just like}\lebnote{Literally “As that”}the wood of the vine among the trees of the forest\lebnote{Or “wood”}which I gave to the fire for fuel, so I have given the inhabitants of Jerusalem.}%
\verse{And I will set my face against them. From the fire they may\textit{escape},\lebnote{Literally “go out”}but the fire will yet consume them, and they will know that I am Adonai \textit{when I set my face against them}.\lebnote{Literally “in to set me my face against them”}}%
\verse{And I will make the land a desolation because \textit{they acted very unfaithfully}!”\lebnote{Literally “they displayed infidelity”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 16
\verseWithHeading{God’s Covenant Grace to Unfaithful and Rebellious Israel}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}make known to Jerusalem its detestable things,}%
\verse{and you must say, ‘Thus says the Lord Adonai to Jerusalem: Your origin and your birth were from the land of the Canaanites,\lebnote{Hebrew “Canaanite”}your father was an Amorite, and your mother was a Hittite.}%
\verse{And as for your birth, \textit{on the day you were born}\lebnote{Literally “on the day of being born you”}your umbilical cord was not cut, and \textit{you were not thoroughly washed clean with water},\lebnote{Literally “with water not you were cleansed for cleansing”}and \textit{you were not thoroughly rubbed with salt},\lebnote{Literally “being rubbed with salt you were not rubbed with salt”}and \textit{you were not carefully wrapped in strips of cloth}.\lebnote{Literally “being wrapped in strips of cloth not you were wrapped in strips of cloth”}}%
\verse{\textit{No eye took pity on you}\lebnote{Literally “not took pity on you an eye”}to do to you one of these things to show compassion for you, and you were thrown \textit{into the open field}\lebnote{Literally “to the face of the field”}in their despising of you \textit{on the day you were born}.\lebnote{Literally “on the day of being born you”}}%
\verse{“ ‘And I passed by you, and I saw you kicking about in your blood, and I said to you in your blood, “Live!}%
\verse{Grow up;\lebnote{Or “A myriad I have made you”}I will make you like a plant of the field.” And you grew up, and you became tall and \textit{reached full womanhood}.\lebnote{Literally “ornament of ornaments”}Your breasts were formed, and your hair had grown, but you were naked and bare.}%
\verse{“ ‘And I passed by you, and I saw you, and look, \textit{your time of lovemaking had come},\lebnote{Literally “your time was the time of lovemaking”}and so I spread out my hem over you, and I covered your nakedness, and I swore to you, and I entered into\lebnote{Hebrew “in”}a covenant with you,’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, ‘and \textit{you became mine}.\lebnote{Literally “and you were to me”}}%
\verse{And I washed you with water, and I rinsed off your blood from on you, and I anointed you with oil.}%
\verse{And I clothed you with beautiful finished cloth, and I put sandals on you of fine leather, and I bound you in fine linen, and I covered you with costly fabric.}%
\verse{And I adorned you with ornaments,\lebnote{Hebrew “ornament”}and I put a bracelet on your arms and a necklace on your neck.}%
\verse{And I put an ornamental ring on your nose and earrings on your ears and a \textit{beautiful crown}\lebnote{Literally “a crown of splendor”}on your head.}%
\verse{And you adorned yourself with gold and silver, and your clothing was fine linen and costly fabric and beautiful finished cloth; you ate finely milled flour and honey and olive oil, and you became exceedingly beautiful; \textit{you were fit to be a queen}.\lebnote{Literally “you could succeed to queenship”}}%
\verse{And a name went out for you among the nations because of your beauty, for it was perfect because of my majesty that I bestowed on you,’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{‘And you trusted in your beauty, and you prostituted on account of your name, and you poured out your fornication on every one passing by, saying, “Let it be his.”}%
\verse{And you took from your clothes and you made for yourself colorful\lebnote{Or “gaudy”}shrines, and you prostituted on them; \textit{this had not happened before},\lebnote{Literally “Not were happening”}and it \textit{will not continue to happen}.\lebnote{Literally “and not it will be”; Hebrew text unclear}}%
\verse{And you took \textit{your beautiful jewelry}\lebnote{Literally “the jewelry of splendor”}made from my gold and from my silver that I had given to you, and you made for yourself \textit{male images},\lebnote{Literally “images of a male”}and you prostituted with them.}%
\verse{And you took the clothes of your beautiful finished cloth, and you covered them and my oil and my incense you set \textit{before them}.\lebnote{Literally “to the face of them”}}%
\verse{And my bread that I gave to you, finely milled flour and oil and honey with which I fed you, then you set it \textit{before them}\lebnote{Literally “to the face of them”}as a \textit{pleasing odor},\lebnote{Literally “a scent of an incense”}and so it was,’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{‘And you took your sons and your daughters whom you had borne for me, and you sacrificed them to them to be eaten,\lebnote{Or “consumed”}\textit{as if your whorings were not enough}.\lebnote{Literally “it was too little from your whorings”}}%
\verse{And you slaughtered my children, and you gave them \textit{to be sacrificed to them}.\lebnote{Literally “to cause them to cross over to them”}}%
\verse{And in all of your detestable things and your fornication you did not remember the days of your childhood \textit{when you were naked and bare},\lebnote{Literally “at your being naked and bare”}when you were kicking about in your blood.}%
\verse{“ ‘And then after all of your evil! Woe, woe, to you!’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{‘And then you built for yourself a mound, and you made for yourself a high place in every public square.}%
\verse{At the head of every street you built your high place and you disgraced your beauty and you spread your feet for every passerby, and you increased your whoring.}%
\verse{And you prostituted with the \textit{Egyptians},\lebnote{Literally “sons/children of Egypt”}your neighbors \textit{who were sexually aroused},\lebnote{Literally “those being great of flesh”}and you increased your fornication to provoke me.}%
\verse{And look! I stretched out my hand against you, and I reduced your portion, and I gave you into\lebnote{Hebrew “in”}the desire of your haters, the daughters of the Philistines, who were ashamed because of your lewd \textit{conduct}.\lebnote{Literally “your way of life”}}%
\verse{And you prostituted with the \textit{Assyrians}\lebnote{Literally “sons/children of Assyria”}on account of \textit{your insatiable lust},\lebnote{Literally “of no your satisfaction”}and you prostituted with them, and still you were not satisfied.}%
\verse{And you increased your fornication to the land of traders,\lebnote{Hebrew “trader”}to Chaldea, and even with this you were not satisfied.}%
\verse{How hot with fever was your heart!’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai. ‘\textit{When you did}\lebnote{Literally “at/in you to do”}all of these actions \textit{of a headstrong female prostitute},\lebnote{Literally “deeds of a woman prostituting domineering”}}%
\verse{\textit{when you built}\lebnote{Literally “at/in building your”}your mound at the head of every street and your high place you made on every street, yet\lebnote{Hebrew “and”}you were\lebnote{According to the reading tradition (\textit{Qere})}not like a prostitute, as you were scorning your harlot’s wages.}%
\verse{O, adulterous woman! In the place of her husband she received strangers.}%
\verse{To all prostitutes they give a fee, but you, you gave your gifts to all of your lovers, and you bribed them to come to you from all around in your whorings!}%
\verse{And it was the opposite with you from the other women in your whorings; \textit{no one solicited you as a prostitute}.\lebnote{Literally “and after you not he solicited a prostitute”}\textit{You gave}\lebnote{Literally “And at your giving”}a harlot’s wages, and a harlot’s wages were not given to you, and so it was just the opposite.}%
\verse{“ ‘Therefore, prostitute, hear the word of Adonai.}%
\verse{Thus says the Lord Adonai: Because of the pouring out of your shame and because your nakedness was revealed in your whoring with your lovers, and on account of all of the idols of your detestable things, and according to the blood of your children whom you gave to them,}%
\verse{therefore, look! I am gathering all of your lovers \textit{whom you were pleased about},\lebnote{Literally “whom you were pleased about them”}and all of whom you loved in addition to all of whom you hated, and I will gather them against you from \textit{everywhere},\lebnote{Literally “all around”}and I will uncover your nakedness to them, and they will see all of your nakedness.}%
\verse{And I will judge you with the judgment of women committing adultery and shedding blood, and I will give you the blood of rage and jealousy.}%
\verse{And I will give you into their hand, and they will demolish your mound, and they will break down your high places, and they will strip you of your clothes, and they will take \textit{your beautiful jewelry},\lebnote{Literally “the objects of your splendor”}and they will leave you naked and bare.}%
\verse{And they will bring up against you a crowd, and they will stone you with stones,\lebnote{Hebrew “stone”}and they will cut you to pieces with their swords.}%
\verse{And they will burn your houses with fire, and they will execute against you judgments \textit{before}\lebnote{Literally “to the eyes of”}many women, and I will stop you from prostitution, and also a harlot’s wages you will not give\lebnote{Or “pay”}again.}%
\verse{And I will satisfy my rage on you, and my jealousy will turn away from you, and I will be calm, and I will not be angry any longer.}%
\verse{\textit{Because}\lebnote{Literally “because that”}you did not remember\lebnote{According to the reading tradition (\textit{Qere}) with BHS}the days of your childhood and you made me angry with all of these. And in turn I, look, I will return your way upon your head,’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, ‘and did\lebnote{According to the reading tradition (\textit{Qere}); cf. BHS}you not do wickedness\lebnote{Literally “the wickedness”}in addition to all of your detestable things?}%
\verse{Look! Everyone quoting a proverb\lebnote{Literally “proverbing a proverb”}against you, he will quote, \textit{saying},\lebnote{Literally “to say”}‘Like her mother is her daughter!’}%
\verse{You are a daughter of your mother who abhorred her husband and her children, and you are a sister of your sisters who abhorred their husbands and their children. Your mother was a Hittite, and your father was an Amorite.}%
\verse{And your elder sister, she is Samaria, and her daughters, who are dwelling on your north, and your younger sister is the one dwelling on your south; she is Sodom and her daughters.}%
\verse{And you have not only gone in their ways, but\lebnote{Or “and”}you also did according to their detestable things. \textit{In such a short time}\lebnote{Literally “as few little and”}you behaved more corruptly than they in all of your ways.}%
\verse{\textit{As surely as I live},’\lebnote{Literally “live I”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, ‘surely your sister Sodom and her daughters did not do as\lebnote{Hebrew “as that”}you and your daughters did.}%
\verse{Look! This was the iniquity of Sodom, your sister: Pride, abundance of food, and \textit{prosperous ease}\lebnote{Literally “and ease of granting relief”}was to her and to her daughters, and \textit{she did not sustain the needy and the poor}.\lebnote{Literally “And the hand of the needy and the poor not she did take hold of”}}%
\verse{And they were proud, and they did a detestable thing \textit{before me},\lebnote{Literally “to the face of me”}and I removed them \textit{because}\lebnote{Literally “as that”}I saw it.}%
\verse{And Samaria did not sin \textit{according to even half of your sins},\lebnote{Literally “about/as/like the half of sins your not did she sin”}and you caused your detestable things to increase more than they, and you made your sister righteous in comparison with all of your detestable things that you did.\lebnote{According to the reading tradition (\textit{Qere}); cf. BHS}}%
\verse{Also,\lebnote{Or “indeed”}you bear your disgrace, by which you furnished justification to your sisters\lebnote{Hebrew “sister”}through your sins by which you acted more abominably than they; they were more righteous than you, and also,\lebnote{Or “indeed”}you be ashamed and bear your disgrace through your making your sister righteous.}%
\verse{And I will restore their fortune, the fortunes of Sodom and her daughters, and the fortunes of Samaria and her daughters, and even the fortunes of your captivity in the midst of them,}%
\verse{in order that you may bear your disgrace and you may be put to shame because of all that you did at your consoling them.}%
\verse{And as for your sisters, Sodom and her daughters, they will return to their former state, and Samaria and her daughters, they will return to their former state, and you and your daughters will return to your former state.}%
\verse{And was not Sodom, your sister, a byword in your mouth in the day of your pride}%
\verse{\textit{before}\lebnote{Literally “at not yet”}your evil was uncovered? It was like the time of the scorn of the daughters of Aram and all those around her, and of the daughters of the Philistines, those all around who are despising you.}%
\verse{Your wickedness and your detestable things, you, you must bear them,’ \textit{declares}\lebnote{Literally “declaration of”}Adonai.}%
\verse{‘For thus says the Lord Adonai: And I will do it\lebnote{According to the reading tradition (\textit{Qere})}to you \textit{according to what you have done},\lebnote{Literally “as that you have done”}who despised the oath to break covenant.}%
\verse{And I, I will remember my covenant with you in the days of your youth, and I will establish with you \textit{an everlasting covenant}.\lebnote{Literally “a covenant of eternity”}}%
\verse{And you will remember your ways, and you will be ashamed when you take your sisters, \textit{both the older and the younger},\lebnote{Literally “the older from you to the younger from you”}and I give them to you as daughters, but not on account of your covenant.}%
\verse{And I, I will establish my covenant with you, and you will know that I am Adonai,}%
\verse{in order that you will remember, and you will be ashamed, and \textit{you will not open your mouth again}\lebnote{Literally “and not it will be again an opening of mouth”}\textit{because of}\lebnote{Literally “from the face of”}your disgrace when I forgive you for all that you have done!’ ” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 17
\verseWithHeading{Parable of Two Eagles and a Vine}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}tell a riddle and quote a proverb to the house of Israel,}%
\verse{and you must say, ‘Thus says the Lord Adonai: “\textit{The great eagle with great wings and long feathers}\lebnote{Literally “the eagle the great great of the wings long of the feather”}and \textit{full of variegated plumage}\lebnote{Literally “full of the plumage which is for him the variegated”}came to Lebanon and he took the treetop of the cedar.}%
\verse{He plucked the top of its new plant shoot, and he brought it to the land of Canaan and put it in a city of merchants,}%
\verse{and he took from the seed of the land and placed it in \textit{fertile soil}\lebnote{Literally “a field of seed”}on\lebnote{Or “by”}many waters; like a willow he planted it.}%
\verse{And it sprouted, and \textit{it became a vine spreading out},\lebnote{Literally “he was to vine spreading out”}low of height, turning its branches to him, and its roots were under it, and \textit{it became a vine},\lebnote{Literally “and it was to a vine”}and it made\lebnote{Or “sprouted”}branches, and it sent out foliage.}%
\verse{“ ‘ “And there was another great eagle, great of wings and with abundant plumage, and look! This vine stretched out its roots toward him and extended its branches to him to water it from the garden bed \textit{where it was planted}.\lebnote{Literally “from the garden bed of its planting”}}%
\verse{It was planted in\lebnote{Or “to”}a good field by\lebnote{Or “to”}many waters\lebnote{Hebrew “water”}to produce branches\lebnote{Hebrew “branch”}and to bear fruit to become\lebnote{Hebrew “be”}\textit{a beautiful vine}.” ’\lebnote{Literally “to a vine of splendor”}}%
\verse{Say, ‘Thus says the Lord Adonai: “Will it prosper? Will he not tear out its roots, and will he not make its fruit scaly, and it will wither, and all of the freshness of its vegetation will dry up? And to lift it from its roots \textit{will not require great strength or many people}.\lebnote{Literally “not with strong arm and with many people”}}%
\verse{And look! Though it is planted, will it prosper? \textit{When the east wind strikes it},\lebnote{Literally “at striking her/it the wind of the east”}will it not dry up completely? On the garden bed of its vegetation it will dry up!” ’ ”}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},}%
\verse{“Say now to the rebellious house of Israel, ‘Do you not know what these are?’\lebnote{Or “mean”}Say, ‘Look! The king of Babylon will come to Jerusalem, and he will take its king and its officials, and he will bring\lebnote{Or “take”}them to himself,\lebnote{Hebrew “him”}to Babylon.}%
\verse{And he took from the seed of the kingship, and he made with him a covenant, and \textit{he brought him under oath},\lebnote{Literally “and he brought him into oath”}and he took the rulers of the land,}%
\verse{so that they would be a humble kingdom and not lift itself up to keep his covenant \textit{in order for it to stand}.\lebnote{Literally “to standing her”; or “to let her stand”}}%
\verse{But he rebelled against him by sending his messengers to Egypt to give to him horses and a large army. Will he succeed? Will he escape doing these things, and can he break the covenant and escape?}%
\verse{\textit{As I live},’\lebnote{Literally “live I”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, ‘\textit{surely}\lebnote{Literally “if not”}in the place of the king \textit{who made}\lebnote{Literally “who he was making”}him king, who despised his oath and who broke his covenant with him—in the midst of Babylon he will die.}%
\verse{And not with a great army and with a great crowd will Pharaoh work\lebnote{Or “help/aid”}with him in the war, at the pouring out of a siege ramp and the building of siege works to destroy many lives.}%
\verse{And he despised the oath to break covenant. And, look, he gave his hand in pledge, and yet he did all of these things. He will not escape.’}%
\verse{Therefore thus says the Lord Adonai: ‘\textit{As I live},\lebnote{Literally “live I”}\textit{surely}\lebnote{Literally “if not”}my oath that he despised and my covenant that he broke I will return upon his head.}%
\verse{And I will spread my net over him, and he will be caught in my hunting net, and I will bring him to Babylon, and I will enter into judgment with him there about his infidelity that he displayed against me.}%
\verse{And all of his choice troops, among all of his troops, they will fall by the sword, and those who are remaining,\lebnote{Or “left over”}they will be scattered to all the directions of the wind, and you will know that I, Adonai, I have spoken.’}%
\verse{Thus says the Lord Adonai: ‘And I will take, even I, from the treetop of the high cedar, and I will plant it, from the head of its new plant shoot I will pluck a tender one, and I will plant it, even I, on a high and lofty mountain.}%
\verse{On the height of the mountain of Israel I will plant it, and it will carry\lebnote{Or “bear”}branches,\lebnote{Hebrew “branch”}and it will bear fruit, and \textit{it will become a noble cedar},\lebnote{Literally “he will be to a cedar noble”}and all of the birds of all wings\lebnote{Or “kinds”}will dwell under it in the shade of its branches.}%
\verse{And all of the trees of the field will know that I, Adonai, I will bring low a high tree, and I will exalt a low, fresh tree, and I will make a dry tree flourish. I, Adonai, I have spoken, and I will do it.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 18
\verseWithHeading{God’s Just Treatment of Individuals, Both Righteous and Wicked}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“\textit{What do you mean by}\lebnote{Literally “What is to you”}quoting this proverb about the land of Israel, \textit{saying},\lebnote{Literally “to say”}‘The fathers, they ate unripe fruit, and the teeth of the child became blunt.’\lebnote{Or “dull,” set on edge”}}%
\verse{\textit{As I live},\lebnote{Literally “life I”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, it will surely not any longer be \textit{appropriate for you}\lebnote{Literally “to you”}to quote this proverb in Israel!}%
\verse{Look! \textit{All lives are mine}.\lebnote{Literally “all of the lives are for me they”}\textit{The lives of father and son alike are mine}.\lebnote{Literally “Like the life of the father and like the life of the son is to me look”}The person\lebnote{Or “life,” or “soul”}sinning will die.}%
\verse{And if a man is righteous and does justice and righteousness,}%
\verse{and on the mountains he does not eat and he does not lift up his eyes to the idols of the house of Israel, and the wife of his neighbor he does not defile and he does not approach a woman of\lebnote{Or “in”}menstruation,}%
\verse{and he oppresses \textit{no one}\lebnote{Literally “a person not”}and he returns a pledge for his loan \textit{and he commits no robbery}\lebnote{Literally “robbed things not he seized”}and he gives\lebnote{Or “shares”}his bread to the hungry and he covers a naked person with a garment,}%
\verse{and \textit{he does not charge interest}\lebnote{Literally “with the interest not he gives”}and he takes no usury, and he holds back\lebnote{Or “brings back”}his hand from injustice and he executes a judgment of fairness between \textit{persons},\lebnote{Literally “man and man”}}%
\verse{and in my statutes he goes about and my regulations he keeps,\lebnote{Or “watches”}\textit{performing faithfully}\lebnote{Literally “to do with with faithfulness”}—then he is righteous, and certainly he will live,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“And now he has a son, a violent one, \textit{who sheds blood}\lebnote{Literally “shedding of blood”}\textit{and does any of these things}\lebnote{Literally “and he does like from one from these”}}%
\verse{(though\lebnote{Hebrew “And”}he\lebnote{That is, the father}did not do all of these things), for the son also\lebnote{Or “even”}eats on the mountains and he defiles the wife of his neighbor.}%
\verse{He oppresses the needy and the poor, and \textit{he commits robbery},\lebnote{Literally “robbed things he seized”}and he does not return a pledge for a loan, and he lifts his eyes to the idols so he does a detestable thing.}%
\verse{\textit{He charges interest}\lebnote{Literally “with the interest he gives”}and takes usury. Then, shall he live? He shall not live, for he did all of these detestable things. Surely he will die! His blood will be on him.}%
\verse{“And look! He has a son, and he sees all of the sin of his father that he did, and he sees it, but he does not do it.}%
\verse{On the mountains he does not eat, and he does not lift up his eyes to the idols of the house of Israel, and the wife of his neighbor he does not defile.}%
\verse{And he oppresses no one; he requires no pledge for a loan, and \textit{he does not commit robbery};\lebnote{Literally “robbed things not does he seize”}he gives his bread to the hungry, and he covers the naked person with a garment.}%
\verse{He brings back\lebnote{Or “holds back”}his hand from iniquity;\lebnote{Cf. verse 8}he does not take interest and usury; he does my regulations; he goes\lebnote{Or “walks”}in my statutes. He will not die because of the guilt of his father; he will surely live!}%
\verse{Because his father \textit{oppressed severely};\lebnote{Literally “he oppressed oppression”}\textit{he stole from his brother};\lebnote{Literally “he seized robbed things of a brother”}that which is not good he did in the midst of his people, and look! He will die through\lebnote{Or “in”}his guilt.}%
\verse{“Yet\lebnote{Or “And”}you say, ‘Why does the son not bear the guilt of the father?’ And since the son does justice and righteousness and he keeps all of my statutes and does them, he shall surely live!}%
\verse{The person,\lebnote{Or “soul,” or “life”}the one sinning, will die. A son shall not bear the guilt of the father, and a father shall not bear the guilt of the son. The righteousness of the righteous shall be on him; the wickedness of the wicked\lebnote{According to the reading tradition (\textit{Qere})}shall be on him.}%
\verse{But if the wicked returns from all of his sins\lebnote{According to the reading tradition (\textit{Qere})}that he has done and he keeps all of my statutes and he does justice and righteousness, he shall surely live; he shall not die!}%
\verse{All of his transgressions that he committed will not be remembered against\lebnote{Or “for”}him. Through\lebnote{Or “By means of”}his righteousness that he has done he shall live.}%
\verse{Have I delight by any means\lebnote{Or “in any way”}in the death of the wicked, \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, and not at his turning from his way, \textit{so that}\lebnote{Literally “and”}he lives?}%
\verse{And \textit{when the righteous turns}\lebnote{Literally “at turning of the righteous”}from his righteousness, so that\lebnote{Literally “and”}he does injustice, and does all of the detestable things that the wicked do, then\lebnote{Or “and”}will he live because of all of his righteousness\lebnote{According to the reading tradition (\textit{Qere})}that he did? Those things will not be remembered because of his infidelity that he displayed, and because of his \textit{sin that he committed}.\lebnote{Literally “sin that he sinned”}Through\lebnote{Or “in/by”}them he shall die.}%
\verse{“Yet\lebnote{Or “And”}you say, ‘The way of the Lord is not fair!’ Listen, now, house of Israel, is my way not fair? Is it not your ways that are not fair?}%
\verse{When the righteous turns from his righteousness, and he does injustice, then\lebnote{Or “and”}he will die because of them; because of his injustice that he did he will die!}%
\verse{And when the wicked turns from his wickedness that he did and he does justice and righteousness, \textit{he will preserve his life}.\lebnote{Literally “his soul/life he will keep alive”}}%
\verse{And if he sees and he returns from all of his transgressions that he did, surely he will live; he will not die!}%
\verse{And yet they, the house of Israel, say, ‘The way of the Lord is not fair!’ Are not my ways fair, house of Israel? \textit{Are not your ways unfair?}\lebnote{Literally “not ways your not are right”}}%
\verse{“Therefore I will judge you, house of Israel, each one\lebnote{Hebrew “man”}according to his ways,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai. “Repent\lebnote{Or “Turn/Return”}and turn around from all of your transgression, and it will not be as a stumbling block of iniquity to you.}%
\verse{Throw \textit{away from yourselves}\lebnote{Literally “from on you”}all of your transgressions that you committed,\lebnote{Hebrew “committed with them”}and make \textit{for yourselves}\lebnote{Literally “to you”}a new heart and new spirit, and so why will you die, house of Israel?}%
\verse{For I have no pleasure in the death of the dying,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai. “And so repent\lebnote{Or “turn/return”}and live!”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 19
\verseWithHeading{Lament for the Leaders of Israel}{“And you, raise a lament about\lebnote{Or “over”}the leaders of Israel,}%
\verse{and you must say,}%
\verse{‘What a lioness was your mother 
among the lions. 
She lay down in the midst of young lions, 
and she reared her lion cubs.}%
\verse{And she raised up one from her cubs; 
he became a fierce lion, 
and he learned to tear prey; 
he ate\lebnote{Or “devoured”}humans.\lebnote{Hebrew “human”}}%
\verse{And nations heard about him; 
in their pit he was caught, 
and they brought him with hooks 
to the land of Egypt.}%
\verse{And she saw\lebnote{Or “realized”}that she was waiting in vain; 
her hope was destroyed, 
and she took one from her cubs, 
and she made him a fierce lion.}%
\verse{And he walked about in the midst of lions; 
he became a fierce lion, 
and he learned to tear prey; 
he ate\lebnote{Or “devoured”}humans.\lebnote{Hebrew “human”}}%
\verse{And he knew their widows,\lebnote{Or “ravished/raped”; or “he ravaged their strongholds”}
and he devastated their cities, 
and the land was appalled, and \textit{everyone in it}\lebnote{Literally “its fullness”}
at the sound of his roar.}%
\verse{And nations set out against him 
\textit{from the surrounding provinces},\lebnote{Literally “all around from the provinces”}
and they spread their net over him, 
and he was caught in their pit.}%
\verse{And they put him in a collar with hooks, 
and they brought him to the king of Babylon; 
they brought him into\lebnote{Hebrew “in”}a prison, 
so that his voice would not be heard \textit{any longer}\lebnote{Literally “longer”}
on the mountains of Israel.}%
\verse{Your mother was like the vine in your vineyard;\lebnote{Or “in the vineyard”}
planted \textit{fruitfully}\lebnote{Literally “fruitful”}beside water, 
and it was full of branches 
from many waters.\lebnote{Or “from waters abundant/many”}}%
\verse{And \textit{she produced branches of strength}\lebnote{Literally “they became to her branches/rods of strength”}
to\lebnote{Or “for”}scepters of rulers; 
its height became tall 
between\lebnote{Or “among”}thick foliage, 
and it was seen\lebnote{Or “visible”}because of its tallness 
among the abundance of its branches.}%
\verse{But it was uprooted in rage; 
it was thrown to the earth, 
and the east wind dried up its fruit; 
they were stripped off, 
and its strong branch dried up; 
fire consumed it.}%
\verse{And now it is planted in the desert, 
in a \textit{dry and thirsty land}.\lebnote{Literally “a land of dryness and thirst”} This is a lament, and \textit{it will be used as a lament}.\lebnote{Literally “she is for a lament”}}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 20
\verseWithHeading{God’s Dealings with Israel for His Name’s Sake}{And then in the seventh year, in the fifth month, on the tenth day of the month, men from the elders of Israel came to consult Adonai, and they sat \textit{before me}.\lebnote{Literally “to the face of”}}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}speak with the elders of Israel, and you must say to them, ‘Thus says the Lord Adonai: “Are you coming to consult me? \textit{As I live},\lebnote{Literally “live I”}I will surely not allow myself\lebnote{Or “me”}to be consulted by you!” ’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{Will you judge them? Will you judge them, son of man?\lebnote{Or “mortal,” or “son of humankind”}Make known to them the detestable things of their ancestors.\lebnote{Or “fathers”}}%
\verse{And you must say to them, ‘Thus says the Lord Adonai: “On the day of my choosing Israel \textit{I swore}\lebnote{Literally “and I raised my hand”}to the offspring of the house of Jacob, and I made myself known to them in the land of Egypt, and \textit{I swore}\lebnote{Literally “and I raised my hand”}to them, \textit{saying},\lebnote{Literally “to say”}‘I am Adonai your God.’}%
\verse{On that day \textit{I swore}\lebnote{Literally “I raised my hand”}to them to bring them out from the land of Egypt to the land that I had searched out for them, flowing with milk and honey—it is the most beautiful of all of the lands.}%
\verse{Then I said to them, ‘Let each one throw away the detestable things of his eyes, and you must not make yourselves unclean with the idols of Egypt! I am Adonai your God.’}%
\verse{But they rebelled against me, and they were not willing to listen to me; each one \textit{did not throw away}\lebnote{Literally “not they threw away”}the detestable things of their eyes; and \textit{they did not abandon}\lebnote{Literally “not they abandoned”}the idols of Egypt, and I decided to pour out my rage on them, to fully vent my anger against them in the midst of the land of Egypt.}%
\verse{“But\lebnote{Or “And”}I acted for the sake of my name \textit{to keep it from being profaned}\lebnote{Literally “not to be profaned before the eyes of”}before the eyes of the nations \textit{among whom they lived},\lebnote{Literally “which they were in their midst”}where I made known to them before their eyes, to bring them out from the land of Egypt.}%
\verse{And I brought them out from the land of Egypt, and I brought them to the desert,}%
\verse{and I gave my statutes to them, and my regulations I made known to them, which, if a person does them, then he will live by them.}%
\verse{And also my Sabbaths I gave to them to be a sign between me and between them \textit{so they would know}\lebnote{Literally “know”}that I, Adonai, am the one sanctifying them.}%
\verse{“But\lebnote{Or “And”}in the desert the house of Israel rebelled against me; \textit{they did not walk in my statutes},\lebnote{Literally “in my statutes not they went”}and they rejected my regulations, which, if a person does them, he will live by them, and they greatly profaned my Sabbaths, and I decided to pour out my rage on them in the desert to destroy them,}%
\verse{and I acted for the sake of my name, that it not be profaned before the eyes of the nations \textit{before whom I brought them out}.\lebnote{Literally “which I brought out them before their eyes”}}%
\verse{And also \textit{I myself swore}\lebnote{Literally “I, I raised my hand”}to them in the desert not to bring them into the land that I had given to them, flowing with milk and honey—it is the most beautiful of all of the lands—}%
\verse{because they despised my judgments, and \textit{they did not walk in my statutes},\lebnote{Literally “and my statutes not they went in them”}and my Sabbaths they profaned, for their heart was going after their idols.}%
\verse{But my eye took pity on them by not destroying them, and \textit{I did not completely destroy them}\lebnote{Literally “and not I make them complete destruction”}in the desert.}%
\verse{“And I said to their children in the desert, ‘You must not go in the statutes of your parents;\lebnote{Or “fathers”}you must not keep their regulations, and you must not make yourself unclean with their idols.}%
\verse{I, Adonai, am your God, so go in my statutes and keep my regulations and do them.}%
\verse{And treat my Sabbaths as holy, and they will be a sign between me and between you \textit{that you may know}\lebnote{Literally “know”}that I, Adonai, am your God.’}%
\verse{But the children rebelled against me; \textit{they did not walk in my statutes},\lebnote{Literally “in statutes my not they went”}and \textit{they did not observe my regulations},\lebnote{Literally “my regulations not they kept to do them”}which if a person does them, then he will live by them. My Sabbaths they desecrated, and I decided to pour out my rage on them, to finish my anger against them in the desert.}%
\verse{But I withheld my hand, and I acted for the sake of my name not to be profaned before the eyes of the nations before whom I had brought them out before their eyes.}%
\verse{What is more, \textit{I swore}\lebnote{Literally “I will raise my hand”}to them in the desert to scatter them among the nations and to disperse them in the lands,}%
\verse{because they did not do my regulations, and my statutes they despised, and my Sabbaths they profaned, and their eyes were after the idols of their ancestors.\lebnote{Or “fathers”}}%
\verse{And in turn I gave to them rules that were not good and regulations \textit{by which they will not live}.\lebnote{Literally “regulations not they will live by them”}}%
\verse{And I defiled them through their gifts in sacrificing all of the first offspring of the womb, in order that\lebnote{Or “so that”}I will cause them to be stunned, so that they will know that I am Adonai.}%
\verse{“Therefore speak to the house of Israel, son of man,\lebnote{Or “mortal,” or “son of humankind”}and you must say to them, ‘Thus says the Lord Adonai: “Again in this your ancestors blasphemed me \textit{at\lebnote{Or “by”}their display of infidelity toward me}.” ’\lebnote{Literally “at/in acting unfaithfully they against me in/by unfaithfulness”}}%
\verse{And I brought them to the land that \textit{I swore}\lebnote{Literally “I raised my hand”}to give\lebnote{Hebrew “give it”}to them, and they saw every high hill and every leafy tree, and they offered their sacrifices, and they presented there the provocation of their offering, and they gave there their fragrant incense offering, and they poured out their libations there.}%
\verse{And I said to them, ‘What is the high place\lebnote{Hebrew Bamah, which became its name}to which you are going?’ And \textit{it is called}\lebnote{Literally “it is called name its”}Bamah until this day.}%
\verse{Therefore thus say to the house of Israel, ‘Thus says the Lord Adonai: “In the way of your ancestors\lebnote{Or “fathers”}will you defile yourself, and after their vile idols will you prostitute yourselves?}%
\verse{And \textit{when you lift up your gifts, sacrificing your children}\lebnote{Literally “at to lift up your gifts at causing to pass over your children”}through the fire, you are defiling yourself through all of your idols until \textit{today},\lebnote{Literally “the day”}and will I let myself\lebnote{Hebrew “me”}be consulted by you, house of Israel?” ’ \textit{As I live},”\lebnote{Literally “Live I”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, “I will not let myself\lebnote{Hebrew “me”}be consulted by you!}%
\verse{And \textit{what you are planning},\lebnote{Literally “what is going up on your spirit”}surely it will not be—that you are saying, ‘Let us be like the nations, like the clans\lebnote{Or “tribes”}of the lands, \textit{serving wood and stone}!’\lebnote{Literally “to serve wood and stone”}}%
\verse{“\textit{As I live},”\lebnote{Literally “live I”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, “\textit{surely}\lebnote{Literally “if not”}with a strong hand and with an outstretched arm and with \textit{rage pouring forth}\lebnote{Literally “rage poured out”}I will reign as king over you!}%
\verse{And I will bring you from the peoples, and I will gather you from the countries to which you were scattered with a strong hand and with an outstretched arm and with rage poured out.}%
\verse{Then I will bring you to the desert of the peoples, and I will execute justice on you there face to face.}%
\verse{\textit{Just as}\lebnote{Literally “Like what”}I executed justice on your ancestors\lebnote{Or “fathers”}in the desert of the land of Egypt, likewise I will execute justice on you!” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“And I will make you pass under the rod, and I will bring you into the bond of the covenant.}%
\verse{And I will purge the rebels from among you and the ones transgressing against me; I will bring them out from the land \textit{where they are living as aliens},\lebnote{Literally “from the land of living as an alien there”}but into the land of Israel they will not come, and then you will know that I am Adonai.}%
\verse{And you, house of Israel, thus says the Lord Adonai: “Let each one go serve his idols now and after, if you are not listening to me, but \textit{my holy name}\lebnote{Literally “the name of my holiness”}you will not profane any longer with your gifts and with your idols.}%
\verse{“For on \textit{my holy mountain},\lebnote{Literally “the mountain of my holiness”}on the mountain of the height of Israel,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, “there all of the house of Israel will serve me, all of them,\lebnote{Hebrew “it/him”}in the land. I will take pleasure in them, and there I will accept your contributions and the best of your portions with all of your holy objects.}%
\verse{I will accept you as \textit{a fragrant incense offering}\lebnote{Literally “a fragrance of an incense offering”}when I bring you out from the peoples and I gather you from the lands to which you were scattered, and I will show myself holy among you before the eyes of the nations.}%
\verse{And you will know that I am Adonai \textit{when I bring you} to the land of Israel, to the land that \textit{I swore}\lebnote{Literally “I raised my hand”}to give to your ancestors.}%
\verse{And you will remember there your ways, and all of your deeds \textit{by which you were made unclean},\lebnote{Literally “which you were made unclean by them”}and you will feel a loathing \textit{for yourself}\lebnote{Literally “for your face”}for all of your evils that you have done.}%
\verse{And you will know that I am Adonai \textit{when I deal with you}\lebnote{Literally “at/in doing my with you”}for the sake of my name and \textit{not according to your evil ways}\lebnote{Literally “not like your ways the evil”}or \textit{according to}\lebnote{Literally “like”}your corrupted deeds, house of Israel,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{\lebnote{Ezekiel 20:45–21:32 in the English Bible is 21:1–37 in the Hebrew Bible}And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}set your face toward the way of the south,\lebnote{Or “Teman”}and preach to the south, and prophesy against the forest of the territory of the Negev.}%
\verse{And you must say to the forest of the Negev, ‘Hear the word of Adonai, thus says the Lord Adonai: “Look! I am kindling against you a fire, and it will devour in you every fresh tree and every dry tree; the blaze of the flame will not be quenched, and all the surfaces from the south to the north will be scorched by it.}%
\verse{And all creatures will see that I, Adonai, I kindled it—it will not be quenched!” ’ ”}%
\verse{Then I said, “Ah, Lord Adonai, they are saying about me, ‘Is he not posing a parable?’ ”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 21
\verseWithHeading{A Vision of the Avenging Sword of Adonai}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}set your face toward Jerusalem, and preach to the sanctuaries, and prophesy to the land of Israel.}%
\verse{And you must say to the land of Israel, ‘Thus says Adonai: “Look! I am against you, and I will draw out my sword from its sheath, and I will cut off from you the righteous and the wicked.}%
\verse{Because\lebnote{Hebrew “Because that”}I will cut off from you both righteous and wicked, therefore my sword will go out from its sheath to\lebnote{Or “against”}all creatures\lebnote{Hebrew “creature”}from south to north.}%
\verse{And they will know, all creatures, that I, Adonai, I will bring out my sword from its sheath; it will not return again!” ’}%
\verse{And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}groan \textit{with shaking hips},\lebnote{Literally “with destruction of loins”}and you must groan with bitterness before their eyes.}%
\verse{\textit{And then}\lebnote{Literally “and it will be”}when they say to you, ‘On account of what are you groaning?’ then you must say, ‘On account of the report, for it is coming, and every heart will be weak and all hands will hang limp and every spirit will be disheartened, and all knees will go like water.’ Look! It is coming, and it will happen!” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}prophesy, and you must say, ‘Thus says the Lord,’ say, ‘A sword, a sword is sharpened and is also polished.}%
\verse{It is sharpened to slaughter a slaughter, \textit{polished to flash like lightning}!\lebnote{Literally “so that there be for her lightning”}Or will we rejoice?\lebnote{Or “be pleased”}A rod, my son, is despising every tree.\lebnote{Or “stick” of correction; Hebrew is uncertain in meaning here}}%
\verse{And he gives it to be polished, \textit{to be seized by the hand}.\lebnote{Literally “to take hold of in the palm of the hand”}It is sharpened—a sword—and it is polished to give it into the hand of the killer.}%
\verse{Cry and wail, son of man,\lebnote{Or “mortal,” or “son of humankind”}for it is against my people; it is against all of the princes of Israel. They are thrown to the sword with my people; therefore \textit{strike your thigh}.\lebnote{Literally “strike on the thigh”}}%
\verse{For \textit{examine}!\lebnote{Literally “it was tested”}And what if also the rod will not be despising?’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}prophesy \textit{and clap your hands}.\lebnote{Literally “strike hand to hand”}And the sword will \textit{strike twice};\lebnote{Literally “be doubled”}let it happen a third time. It is a sword of the dead, the sword of the great dead that is surrounding them,}%
\verse{so that a heart melts, and the fallen \textit{multiply}\lebnote{Literally “be made many”}at all of their gates. I gave a sword for slaughter, and alas! It is made for flashing, it is grasped\lebnote{Or “polished”}for slaughtering.}%
\verse{Gather together, strike to the right; cause to go to the left, where your \textit{edge}\lebnote{Literally “face”}is directed.}%
\verse{And also \textit{I myself}\lebnote{Literally “I, I”}will \textit{clap my hands},\lebnote{Literally “strike my hand to my hand”}and I will satisfy my rage! I, Adonai, I have spoken.”}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}mark out for yourself\lebnote{Hebrew “you”}two roads for the coming of the sword of the king of Babylon; they must \textit{both}\lebnote{Literally “the two of them”}go out from the same land. And hew out a \textit{signpost};\lebnote{Literally “hand”}hew it at the head of the road of the city.}%
\verse{You must mark a road for the coming of the sword to Rabbah of the \textit{Ammonites}\lebnote{Literally “sons of Ammon”}and to Judah, in Jerusalem the fortified.}%
\verse{For the king of Babylon stands at the fork of the road at the head of the two roads to practice divination. \textit{He shakes the arrows},\lebnote{Literally “he casts the lot with the arrows” (cf. NASB, NRSV)}he inquires with the teraphim, he \textit{examines}\lebnote{Literally “sees with”}the liver.}%
\verse{In his right hand is the divination for Jerusalem, to put up battering rams, to open a mouth for slaughter,\lebnote{Or “open the mouth of Sheol”; or “give the command to slaughter” (cf. NRSV, NASB)}\textit{to raise the battle cry},\lebnote{Literally “to raise a voice in shouting”}to put up battering rams against gates, to build a siege ramp, to build siege works.\lebnote{Hebrew “siege work”}}%
\verse{And it will be to them like practicing divination falsely\lebnote{Or “in vain”}in their eyes; \textit{they have sworn oaths for themselves}.\lebnote{Literally “swearers of oaths will be for them”}But\lebnote{Or “And”}he will bring their guilt to remembrance so as to seize them.}%
\verse{“Therefore thus says the Lord Adonai: ‘Because you have brought to remembrance your guilt by the uncovering of your transgressions, so that your sins in all of your deeds appear—because of your being remembered, you will be captured in the hand.’}%
\verse{And as for you, profane one, wicked prince of Israel, whose day has come with the time of the punishment of the end,}%
\verse{thus says the Lord Adonai: ‘Remove the turban and lift off the crown; things are no longer the same. Exalt the low and bring low the high.}%
\verse{A ruin, a ruin, a ruin I will make it! Also\lebnote{Or “indeed”}this has not ever happened;\lebnote{Or “will not happen until” (cf. NASB and NRSV)}it will remain until the coming of the one to whom the judgment belongs and I have given it to him.’}%
\verse{And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}prophesy, and you must say, ‘Thus says the Lord Adonai to the \textit{Ammonites}\lebnote{Literally “sons of Ammon”}and concerning their disgrace,’ and you must say, ‘A sword, a sword is drawn for slaughtering; it is polished for holding, \textit{for flashing like lightning},\lebnote{Literally “for the sake of lightening”}}%
\verse{when seeing a false vision for you in vain, when practicing divination for you falsely, to give\lebnote{Or “place”}you on the neck of the profane ones of the wicked, \textit{whose day has come},\lebnote{Literally “who has come their day”}\textit{the time of final punishment}.\lebnote{Literally “in the time of the punishment of the end”}}%
\verse{Return it to its sheath in the place where you were created. In the land of your origin I will judge you!}%
\verse{And I will pour out my anger on you; I will blow on you with the fire of my wrath, and I will give you into the hand of brutal men, skilled craftsmen of destruction.}%
\verse{You will be as fuel for the fire; your blood will be in the midst of the earth. You will not be remembered, for I, Adonai, I have spoken.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 22
\verseWithHeading{Highlighting the Sins and the Judgments of Israel}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}will you judge? Will you judge the \textit{bloody city}?\lebnote{Literally “city of blood”}Then you must make known to her all of her detestable things!}%
\verse{And you must say, ‘Thus says the Lord Adonai: “A city pouring out blood in the midst of her; \textit{its time has come},\lebnote{Literally “to come time her its”}and it made idols for itself, becoming unclean.}%
\verse{By your blood that you poured out you have become guilty, and by your idols that you made you have become unclean, and you have brought your days near, \textit{and your appointed years have come}.\lebnote{Literally “and it came to years your”}Therefore I have made you a disgrace for the nations and a laughingstock to all of the countries.}%
\verse{The people near and the people far from you will make fun of you, \textit{the unclean and the terrified}.\lebnote{Literally “the unclean of the name the great of the panic”}}%
\verse{Look! The princes of Israel, each one according to his strength,\lebnote{Or “influence”}they are in you \textit{for the shedding of blood}.\lebnote{Literally “to shedding of blood”}}%
\verse{They have treated father and mother with contempt in you; \textit{they violated}\lebnote{Literally “they did with extortion”}the alien with\lebnote{Or “by”}extortion; in the midst of you they mistreated the orphan and widow.}%
\verse{You despised my holy objects; my Sabbaths you profaned.}%
\verse{Slanderous men are \textit{among you}\lebnote{Literally “in you”}to shed blood, and they, \textit{among you},\lebnote{Literally “in you”}eat upon the mountains; they do wickedness in the midst of you.}%
\verse{They uncover the nakedness of a father among\lebnote{Hebrew “in”}you; they violate a woman unclean of menstruation among\lebnote{Hebrew “in”}you.}%
\verse{And a man does a detestable thing with the wife of his neighbor, and a man defiles his daughter-in-law in wickedness, and a man sexually violates among\lebnote{Hebrew “in”}you his sister, the daughter of his father.}%
\verse{They take a bribe among\lebnote{Hebrew “in”}you in order to shed blood; and you take usury, and you make gain from your neighbors by extortion, and so you have forgotten me, \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{And look! I strike my hand for your ill-gotten gain that you have made and at your blood that was in the midst of you.}%
\verse{Can your heart endure,\lebnote{Or “hold up”}or can your hands be strong at the days in which I am dealing with you? I, Adonai, I have spoken, and I will act!}%
\verse{I will scatter you among the nations, and I will disperse you through the countries, and I will purge your uncleanness from you.}%
\verse{And I will be profaned by you before the eyes of the nations, and you will know that I am Adonai.” ’ ”}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}the house of Israel has become as silver dross to me; all of them are as bronze and tin and iron and lead in the midst of a furnace, even as silver dross, silver dross they became!}%
\verse{Therefore thus says the Lord Adonai: ‘Because all of you have become as silver dross, therefore look! I am gathering you to the midst of Jerusalem,}%
\verse{like the gathering of silver and bronze and iron and lead and tin to the middle of a furnace to blow fire on it for melting. Thus I will gather in my anger and in my rage, and I will \textit{deposit}\lebnote{Literally “I will put/place”}you, and I will melt you.}%
\verse{And I will gather you, and I will blow on you with the fire of my wrath, and so I will melt you in the midst of her.}%
\verse{Thus you will be melted like the melting of silver in the midst of a furnace, and you will know that I, Adonai, I have poured out my rage on you.’ ”}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}say to her, ‘You are a land not cleansed; it is not rained upon in the day of indignation.’}%
\verse{The conspiracy of its prophets in the midst of her is like a roaring lion that is tearing prey. They devour people, and they take wealth and treasure; they make its widows numerous in the midst of her.}%
\verse{Its priests treat my law violently, and they profane my holy objects; they do not distinguish between a holy object and what is unholy, or between the clean and the unclean. They do not teach the difference, and they hide their eyes from my Sabbaths, and so I am profaned in the midst of them.}%
\verse{Its officials are like wolves tearing prey in its midst, to pour out blood, to destroy people, to make dishonest gain.}%
\verse{And for them its prophets plaster whitewash; they are seeing falseness and are practicing divination for them by lying, saying, ‘Thus says the Lord Adonai,’ and Adonai has not spoken.}%
\verse{They severely oppress the people of the land, and \textit{they committed robbery},\lebnote{Literally “and they seized a robbed things”}and they mistreated the needy and the poor, and they oppressed the alien \textit{without}\lebnote{Literally “with not”}justice.}%
\verse{And so I sought for them somebody, one repairing the wall and standing in the breach \textit{before me}\lebnote{Literally “to the face of me”}on behalf of the land not to destroy it, but I did not find anyone,}%
\verse{and so I poured out my indignation on them. With the fire of my wrath I destroyed them; I returned their way upon their head,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 23
\verseWithHeading{Oholah and Oholibah as Symbols of God’s Corrupt People}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}there were two women, the daughters of one mother,}%
\verse{and they prostituted themselves in Egypt in their childhood; they were prostituting themselves there, and their breasts were fondled, and there they caressed the bosoms of their virginity.}%
\verse{Now as for their names, the older was Oholah, and Oholibah was her sister. And \textit{they became mine},\lebnote{Literally “they were for me”}and they bore sons and daughters, and their names\lebnote{Hebrew “name”}are Samaria for Oholah, and Jerusalem for Oholibah.}%
\verse{And Oholah prostituted herself \textit{while she was still mine},\lebnote{Literally “under me”}and she lusted for her lovers, for Assyria who was nearby,}%
\verse{clothed in blue cloth, governors and prefects, \textit{handsome young men}\lebnote{Literally “young men of handsomeness/beauty”}and all of them horsemen, \textit{experts on horseback}.\lebnote{Literally “riders of horsemen”}}%
\verse{And she bestowed her fornication on them, on the choice ones of the \textit{Assyrians},\lebnote{Literally “sons of Assyria”}all of them, and with every one after which she lusted; with all of their idols she defiled herself.}%
\verse{And her whorings from the time of Egypt she did not abandon, for they slept with her in her childhood, and they caressed the bosoms of her virginity, and they poured out their fornication on her.}%
\verse{Therefore I gave her into the hand of her lovers, into the hand of the \textit{Assyrians}\lebnote{Literally “sons of Assyria”}\textit{after whom she lusted}.\lebnote{Literally “which she lusted for them”}}%
\verse{They uncovered her nakedness; they took her sons and her daughters, and they killed her with the sword. And she became a name for the women, and they executed judgments against her.}%
\verse{“And though Oholibah her sister saw, yet in her lust \textit{she behaved more corruptly than her},\lebnote{Literally “she behaved corruptly in lust from her”}and her whoring was more than the prostitution of her sister.}%
\verse{She lusted after the \textit{Assyrians},\lebnote{Literally “sons of Assyria”}governors and prefects, warriors\lebnote{Hebrew uncertain}clothed in\lebnote{Or “to”}perfection,\lebnote{Or “in full armor” or “wonderfully dressed”}\textit{expert horsemen},\lebnote{Literally “horsemen, riders of horses”}all of them \textit{handsome young men}.\lebnote{Literally “young men of handsomeness”}}%
\verse{And I saw that she was defiled; \textit{they had both taken the same path}!\lebnote{Literally “way one was for the two of them”}}%
\verse{And she increased her whorings, and she saw men carved\lebnote{Or “portrayed”}on the wall, images of Chaldeans\lebnote{According to the reading tradition (\textit{Qere})}carved in red,}%
\verse{belted with a belt at\lebnote{Or “around”}their waist with turbans on their heads, all of them giving the appearance of adjutants,\lebnote{Or “officers”}the image\lebnote{Or “likeness”}of the \textit{Babylonians};\lebnote{Literally “sons of Babylon”}Chaldea was the land of their birth.}%
\verse{And she lusted for them \textit{when her eyes saw them},\lebnote{Literally “at the sight of her eyes”}and she sent messengers to them at Chaldea.}%
\verse{And so the \textit{Babylonians}\lebnote{Literally “sons of Babylon”}came to her for the bed of lovemaking; and they defiled her with their fornication, and she was defiled by them, and she turned from them.}%
\verse{And she revealed her whorings, and she revealed her nakedness, and so I turned from her \textit{just as}\lebnote{Literally “as that”}I turned from her sister.}%
\verse{Yet\lebnote{Or “And”}she increased her whorings, \textit{recalling}\lebnote{Or “remembering”}the days of her childhood when she was prostituted\lebnote{Or “prostituted herself”}in the land of Egypt.}%
\verse{And she lusted after her male lovers whose genitalia were the genitalia of male donkeys and their seminal emission was the seminal emission of horses.}%
\verse{And you longed after the obscene conduct of your youth when your bosom was caressed by Egypt, fondling your young breasts.}%
\verse{“Therefore, Oholibah, thus says the Lord Adonai: ‘Look! I am stirring up your lovers against you concerning whom you turned away, and I will bring them against you from all around:}%
\verse{the \textit{Babylonians}\lebnote{Literally “sons of Babylon”}and all of the Chaldeans, Pekod and Shoa and Koa, all of the \textit{Assyrians}\lebnote{Literally “sons of Assyria”}along with them, \textit{handsome young men},\lebnote{Literally “young men of beauty/handsomeness”}governors and prefects, all of them adjutants\lebnote{Or “officers”}\textit{and excellent horsemen}.\lebnote{Literally “ones called riders of horses”}}%
\verse{And they will come against you with an army chariot and wagon and with a crowd\lebnote{Or “horde” or “host of”}of peoples; they will set themselves against you from all around with large shield and small shield and helmet. And I will give \textit{before them}\lebnote{Literally “to the face of them”}judgment, and they will judge you with their judgments.}%
\verse{And I will direct my zeal against you, and they will deal with you in anger; your nose and your ears they will remove, and \textit{those who are left},\lebnote{Literally “and remainder your”}they will fall by the sword, and they will take your sons and your daughters, and \textit{your remnant}\lebnote{Literally “and remainder your”}will be consumed by fire.}%
\verse{And they will strip you of your clothes, and they will take \textit{your splendid jewelry}.\lebnote{Literally “the jewelry of splendor your”; or “your beautiful jewelry”}}%
\verse{And I will put an end to your obscene conduct coming from you and your fornication from the land of Egypt, and you will not lift your eyes to them; and you will not remember Egypt again.’}%
\verse{For thus says the Lord Adonai: ‘Look! I am giving you into the hand of those you hated, into the hand of those from whom\lebnote{Literally “them”}you turned away.}%
\verse{And they will deal with you in hatred, and they will take all of your acquisitions, and they will leave you naked and in bareness; and the nakedness of your fornication and your obscene conduct and your whorings will be exposed.}%
\verse{These things are accomplished\lebnote{Or “done”}against\lebnote{Hebrew “in”}you since you prostituted yourself after the nations, and on account of that, you defiled yourself by their idols.}%
\verse{You went in the same way of your sister, and I will give her cup into your hand.” ’}%
\verse{Thus says the Lord Adonai: “You will drink the deep and wide\lebnote{Or place “drink deep and wide the cup”}cup of your sister; you will be as laughter and as scorn; \textit{the cup holds so much}!\lebnote{Literally “a large among to hold”}}%
\verse{You will be filled with drunkenness and sorrow, for a cup of horror and desolation is the cup of your sister, Samaria.}%
\verse{And you will drink it, and you will drain it, and its potsherds you will gnaw, and you will tear out your breasts, for \textit{I myself}\lebnote{Literally “I, I”}spoke,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{Therefore thus says the Lord Adonai: “Because you have forgotten me, and you threw me behind your back, now in turn\lebnote{Or “also”}you bear your obscene conduct and your whorings.”}%
\verse{And Adonai said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}will you judge Oholah and Oholibah and declare their abominable deeds to\lebnote{Or “for”}them?}%
\verse{For they committed adultery, and blood is on their hands, and they committed adultery with their idols, and even their children that they had borne for me—they sacrificed them as food!}%
\verse{Also they did this to me: they defiled my sanctuary on that day, and they profaned my Sabbaths,}%
\verse{and when they slaughtered their children for their idols, they came to my sanctuary on that day to profane it. And look, \textit{this is what they did in my house}!\lebnote{Literally “thus they did in the midst of house my”}}%
\verse{\textit{What is worse},\lebnote{Literally “and also for/indeed”}they sent for men \textit{who come}\lebnote{Literally “coming”}from a distant place, to whom a messenger was sent to them, and look! They came! Men for whom you bathed and painted your eyes, and you adorned yourself with an ornament.}%
\verse{And you sat on a magnificent couch and a table prepared \textit{before her},\lebnote{Literally “to the face of her”}and my incense and my olive oil you put on her.\lebnote{Or “it”}}%
\verse{And a sound of a carefree crowd was with it,\lebnote{Or “her”}and in addition to these men, \textit{a crowd of drunken men\lebnote{A problematic verse; reading according to the reading tradition (\textit{Qere})}was brought in}\lebnote{Literally “from an abundance of men being brought in drunken”}from the desert,\lebnote{Or “wilderness”}and they put bracelets on their arms and a crown of splendor on their heads.}%
\verse{And I said to\lebnote{Or “about”}the one worn out with adulteries, ‘Now they will prostitute her concerning her fornication, even her.’\lebnote{Or “that she did”}}%
\verse{And so they went to her like going to a female\lebnote{Or “a woman”}prostitute; and thus they went to Oholah and to Oholibah, the women of obscene conduct.}%
\verse{But righteous men, they will judge them with the judgment of committing adultery, and with the judgment of shedding blood; for they were committing adultery, and blood was on their hands.”}%
\verse{For thus says the Lord Adonai: “Bring up against them an assembly, and make them as a thing of horror and as plunder.}%
\verse{And an assembly must stone them with stones,\lebnote{Hebrew “stone”}and they must cut them down. With their swords they shall kill their sons and their daughters, and with fire they shall burn their houses.}%
\verse{And \textit{I will cause obscene conduct to cease from the land},\lebnote{Or “I will put an end to obscene conduct from the land”}and all of the women will be warned, and they will not do according to your wickedness.}%
\verse{And they will repay your obscene conduct upon you, and the guilt of your idols you will bear, and you will know that I am the Lord Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 24
\verseWithHeading{The Boiling Pot and the Death of Ezekiel’s Wife}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me in the ninth year, in the tenth month, on the tenth day of the month, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}write for yourself the name of the day, \textit{this very day}.\lebnote{Literally “exactly the day the this”}The king of Babylon laid siege to Jerusalem on exactly this day!}%
\verse{And \textit{deliver a proverb}\lebnote{Literally “quote a proverb”; “proverb a proverb”}to \textit{the rebellious house},\lebnote{Literally “the house of rebellion”}and you must say to them, ‘Thus says the Lord Adonai: “Place the pot! Place it and also pour water into it.}%
\verse{Gather its pieces to it, every good piece, thigh and shoulder, fill it with choice bones;}%
\verse{take the choicest of the flock, and also pile the bones\lebnote{Or possibly “logs”; the difference is one letter in Hebrew}under it; \textit{boil it vigorously};\lebnote{Literally “boil its boiling”}indeed,\lebnote{Or “also”}its bones boiled in the midst of it.” ’ ”}%
\verse{Therefore thus says the Lord Adonai: “Woe to the city of bloodguilt! A pot that has its rust in it; \textit{its rust did not go out from it}.\lebnote{Literally “rust her not went forth from her”}Bring it out \textit{piece by piece};\lebnote{Literally “according to its pieces”}\textit{one is as good as another}.\lebnote{Literally “no lot has fallen on it”}}%
\verse{For her blood was in the midst of her; \textit{she put it on a bare rock};\lebnote{Literally “on barrenness of a rock she put it”}she did not pour it on the ground to cover it with dust.\lebnote{Or “earth” or “dirt”}}%
\verse{To stir up rage, to avenge myself through vengeance, I placed its blood on the barrenness of a rock, \textit{so that it may not be covered}.”\lebnote{Literally “so as not to be covered”}}%
\verse{Therefore thus says the Lord Adonai: “Woe to the city of bloodguilt! I, even I, will make the pile of wood great!}%
\verse{Pile up the logs; kindle the fire; finish cooking the meat, and mix in the spices, and let the bones be burned.\lebnote{Or “charred”}}%
\verse{And \textit{make it stand}\lebnote{Literally “cause it to stand”}empty upon its burning coals so that it may become hot, and its copper may become molten and be melted in the midst of it,\lebnote{Literally “her”}so that its uncleanness and its rust be consumed.}%
\verse{Through effort she has caused weariness; \textit{its thick rust}\lebnote{Literally “the abundance of its rust”}went not out of it. Its rust! Into the fire with its rust!}%
\verse{In your uncleanness is obscene conduct, because I cleansed you, but you were not clean from your uncleanness; you will not be clean again until \textit{I fully vent my rage}\lebnote{Literally “until causing rest I to my rage”}on you.}%
\verse{I, Adonai, I have spoken. It is coming, and I did it; I will not disregard, and I will not take pity, and I will not reconsider. According to your ways and according to your deeds they will judge you!”\lebnote{Or “I will judge”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Literally “to say”}look! I am taking from you what is pleasing to your eyes with a plague,\lebnote{Or “blow”}but you shall not mourn, and you shall not weep, and your tears shall not run down.}%
\verse{Sigh in silence for the dead; you shall not make a mourning ceremony. Bind your turban on you, and you must put your sandals on your feet. You must not cover your upper lip,\lebnote{Or “your mustache” (NASB)}and the bread of \textit{mourners}\lebnote{Literally “men”}you shall not eat.”}%
\verse{And I spoke to the people in the morning, and my wife died in the evening, and I did in the morning \textit{just as}\lebnote{Literally “as that”}I was commanded.}%
\verse{And the people said to me, “Will you not make known to us what these things that you are doing mean for us?”}%
\verse{And I said to them that the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Say to the house of Israel, ‘Thus says the Lord Adonai: “Look! I will be profaning my sanctuary, the pride of your strength and the delight of your eyes and the object of your own\lebnote{Or “souls’,” or “selves’ ”}affection; and your sons and your daughters whom you left behind will fall by the sword,}%
\verse{and you shall do \textit{just as}\lebnote{Literally “as that”}I did: You shall not cover your upper lip,\lebnote{Or “mustache”}and you shall not eat the bread of \textit{mourners}.\lebnote{Literally “men”}}%
\verse{And your turban must be on your heads, and your sandals must be on your feet. You shall not mourn, and you shall not weep, but you shall waste away because of your iniquities, and you shall groan \textit{to one another}.\lebnote{Literally “a man to his brother”}}%
\verse{And Ezekiel will be to you as a sign; \textit{everything that he did}\lebnote{Literally “like all that he did”}you shall do; and when it comes you will know that I am the Lord Adonai.}%
\verse{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}will it not be that on the day \textit{when I take from them}\lebnote{Literally “on the day of to take me from them”}their stronghold, the joy of their glory, the delight of their eyes, and the longing of their inner self,\lebnote{Or “soul”}their sons and daughters—}%
\verse{on that day a survivor will come to you \textit{with the news}.\lebnote{Literally “with a message for ears”}}%
\verse{On that day your mouth shall be opened \textit{at the arrival of the fugitive},\lebnote{Literally “with the fugitive”; or “to the one who has escaped” (NRSV; cf. NASB)}and you shall speak, and you shall not be silent any longer, and you shall be to them as a sign, and they will know that I am Adonai.”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 25
\verseWithHeading{Judgment upon Various Gentile Nations}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}set your face toward the \textit{Ammonites}\lebnote{Literally “sons of Ammon”}and prophesy against them,}%
\verse{and you must say to the \textit{Ammonites},\lebnote{Literally “sons of Ammon”}‘Hear the word of the Lord Adonai! Thus says the Lord Adonai: “Because of your saying, ‘Ah!’ to my sanctuary when it was profaned, and to the land of Israel when it was\lebnote{Or “became”}desolate, and to the house of Judah when they went into the exile,}%
\verse{therefore look! I am giving you to the people\lebnote{Or “children”}of the East as a possession, and they will set up their encampments in you, and they will make in you their dwelling places; \textit{they themselves}\lebnote{Literally “they, they”}will eat your fruit and \textit{they themselves}\lebnote{Literally “they, they”}will drink your milk.}%
\verse{And I will make Rabbah as a pasture of camels and the \textit{Ammonites}\lebnote{Literally “sons of Ammon”}as a haunt of flocks, and they will know that I am Adonai.” ’ ”}%
\verse{For thus says the Lord Adonai: “Because of clapping your hand and stamping with your foot, and because you rejoiced in\lebnote{Or “to”}yourself with all of your malice over the land of Israel,}%
\verse{therefore look! I stretched out my hand against you, and I will give you as plunder\lebnote{According to the reading tradition (\textit{Qere})}to the nations, and I will cut you off from the peoples, and I will destroy you from the countries, and I will wipe you out, and you will know that I am Adonai.”}%
\verse{Thus says the Lord Adonai: “Because of Moab and Seir saying, ‘The house of Judah is like all of the nations,’}%
\verse{therefore, look! I am opening the side of\lebnote{Or “the flank”}Moab from the cities on its frontier, the glory of the land: Beth Jeshimoth, Baal Meon, and Kiriathaim.}%
\verse{I will give it to the people\lebnote{Or “children”}of the East in addition to the \textit{Ammonites}\lebnote{Literally “sons of Ammon”}as a possession, so that the \textit{Ammonites}\lebnote{Literally “sons of Ammon”}will not be remembered among the nations.}%
\verse{And on Moab I will execute punishments, and they will know that I am Adonai.”}%
\verse{Thus says the Lord Adonai: “Because of the doings of Edom in avenging himself with vengeance for\lebnote{Or “toward”}the house of Judah, and so they became very guilty and\lebnote{Or “since”}they avenged themselves on them,”}%
\verse{therefore thus says the Lord Adonai: “so\lebnote{Literally “and”}I will stretch out my hand against Edom and I will cut off from it both human and animal, and I will make it a ruin from Teman and Dedan; they will fall by the sword.}%
\verse{And I will exact my vengeance on Edom by the hand of my people Israel, and they will do in Edom according to my anger and according to my rage, and they will know my vengeance,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{Thus says the Lord Adonai: “Because of the \textit{acting}\lebnote{Literally “doing”}of the Philistines in vengeance, so that they avenged themselves relentlessly with malice in themselves for destruction \textit{with everlasting hostility},\lebnote{Literally “with hostility of ongoing ages/eternity”}}%
\verse{therefore thus says the Lord Adonai: ‘Look! I am stretching out my hand against the Philistines, and I will cut off the Kerethites, and I will destroy the remainder of \textit{the seacoast}.\lebnote{Literally “the coast of the sea”}}%
\verse{And I will execute on them great vengeance with punishments of rage, and they will know that I am Adonai when I exact my vengeance on them!”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 26
\verseWithHeading{Ezekiel Prophesies Against Tyre}{And it was in the \textit{eleventh}\lebnote{Literally “one ten”}year, on the first day of the month, the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}because\lebnote{Hebrew “because that”}Tyre said concerning Jerusalem, ‘Ah! The gates of the peoples are broken; it has swung open to me; I shall be filled, for it lies in ruins!’}%
\verse{Therefore thus says the Lord Adonai: ‘Look! I am against you, Tyre, and I will bring up against you many nations like the stirring up of the sea \textit{stirring up its waves}.\lebnote{Literally “with respect to its waves”}}%
\verse{And I will destroy the walls of Tyre, and they will demolish its towers, and I will scrape away its earthen dirt from it, and I will make it into \textit{a bare rock}.\lebnote{Literally “to barrenness of rock”}}%
\verse{It will become a place for spreading out dragnets in the midst of the sea; for I have spoken,’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai. ‘And it will become as plunder for the nations,}%
\verse{and its daughters who are in the field with the sword, they will be killed; and they will know that I am Adonai.’ ”}%
\verse{For thus says the Lord Adonai: “Look! I am bringing to Tyre Nebuchadnezzar, the king of Babylon, from the north, the king of kings, with horse and with chariot and with horsemen and his assembly and many people.}%
\verse{Your daughters he will kill in the field with the sword, and he will place\lebnote{Hebrew “give”}against you siege works, and he will build against you a siege ramp, and he will raise against you a shield,}%
\verse{and the thrust of his battering ram he will direct against your walls, and your towers he will break down with his weapons.}%
\verse{From the abundance of his horses he will cover you with their fine dust; at the sound of horseman and wheel and chariot your walls will shake, at his coming into your gates like the entrance of a city that is being broken through.}%
\verse{With the hooves of his horses he will trample all of your streets. He will kill your people with the sword, and \textit{your strong stone pillars}\lebnote{Literally “the stone pillars of strength your”}will tumble down to the earth.}%
\verse{And they will plunder your wealth, and they will loot your merchandise, and they will break down your walls, and the houses of your delight they will break down, and your stones and your timbers and your earthen dirt they will cast into the midst of the water.}%
\verse{And, I will put an end to the noise of your songs, and the sound of your lyres will not be heard any longer.}%
\verse{And I will make you into \textit{a bare rock},\lebnote{Literally “to barrenness of rock”}a place for the spreading out of dragnets. You will not be built again, for I, Adonai, I have spoken,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{Thus says the Lord Adonai to Tyre, “Will not the coastlands shake from the sound of your downfall, at the groaning of the wounded, \textit{at people being killed}\lebnote{Literally “at being killed a killing”}in the midst of you?}%
\verse{And all the princes of the sea will go down from their thrones, and they will remove their robes, and their \textit{beautiful garments}\lebnote{Literally “the garments of beautiful finished cloth”}of finished cloth they will take off. With terror they will be clothed, and on the ground they will sit, and they will tremble \textit{continually},\lebnote{Literally “for moments”}and they will be appalled over you.}%
\verse{And they will raise a lament over you, and they will say to you,}%
\verse{‘How you have been lost who was inhabited from\lebnote{Or “on”}the seas; 
the city that was praised, that was strong. 
It is located on the sea, 
and its inhabitants imposed their terror \textit{on all of its inhabitants}.\lebnote{Literally “to all of its inhabitants”}}%
\verse{For thus says the Lord Adonai: “When I make you a desolate city, like the cities that are not inhabited, when bringing up over you the deep, the \textit{great surging waters}\lebnote{Literally “the waters the many”}will cover you.}%
\verse{And I will bring you down with those who are going down to the grave,\lebnote{Or “pit”}\textit{an ancient people},\lebnote{Literally “a people of eternity/endless ages”}and I will cause you to dwell in the world\lebnote{Or “land”}of the depths, in the ruins from of old with those who are going down to the grave,\lebnote{Or “pit”}so that you will not be inhabited and \textit{have a place}\lebnote{Hebrew “and I will give beauty in the land of the living”}in the land of the living.}%
\verse{Sudden terrors I will bring on you, and you shall no longer exist; and you will be sought, and you will not be found again \textit{forever},”\lebnote{Literally “to eternity/endless ages”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 27
\verseWithHeading{A Lament for Doomed Tyre}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}raise a lament against Tyre.}%
\verse{And you must say to Tyre, the one who sits at the entrance of the sea as the merchant of the peoples to the many coastlands, ‘Thus says the Lord Adonai:}%
\verse{Tyre, \textit{you yourself said}\lebnote{Literally “you you said”}
“I am perfect \textit{in beauty}!”\lebnote{Literally “of beauty”}}%
\verse{In the heart of the seas are your boundaries; 
your builders perfected your beauty.}%
\verse{They built with pine trees from Senir 
all of your boards for you; 
they took cedars from Lebanon 
to make a sailing mast \textit{for you}.\lebnote{Literally “on you”}}%
\verse{They made your oars 
with oaks from Bashan; 
your deck they made with inlaid ivory, 
with cypress trees from the coastlands of Cyprus.\lebnote{According to the reading tradition (\textit{Qere})}}%
\verse{Your sail was fine linen with colorful weaving from Egypt 
\textit{to serve as a banner for you};\lebnote{Literally “to be for you as a banner”}
blue and purple cloth from the coastlands of Cyprus\lebnote{Or “Elishah”}
was your awning.}%
\verse{The inhabitants of Sidon and Arvan 
were your rowers; 
your \textit{skilled men},\lebnote{Literally “wise men”}O Tyre, were \textit{from your own people},\lebnote{Literally “in you”}
and they were your seamen.}%
\verse{The elders of Gebal and its \textit{skilled men}\lebnote{Literally “wise men”}were \textit{among you}\lebnote{Literally “in you”}
as the repairers of the seam of your boat; 
all of the ships of the sea and their mariners were \textit{among you}\lebnote{Literally “in you”}
to barter your wares.}%
\verse{Persia and Lud and Put 
were \textit{among your soldiers};\lebnote{Literally “in your army as the men of war your”}
small shield and helmet hung \textit{among you}\lebnote{Literally “in you”}
and gave to you your adornment.}%
\verse{“ ‘Tarshish was your trader; from the abundance of all of their wealth, with silver, iron, tin, and lead they exchanged for your merchandise.}%
\verse{Javan, Tubal, and Meshech were your traders; \textit{in exchange for people}\lebnote{Or “for a human person/soul”}and an object of bronze they gave you your wares.}%
\verse{From Beth Togarmah they exchanged horses and war horses and mules for your wares.}%
\verse{The people\lebnote{Or “children”}of Dedan were trading with you, many coastlands composed the region of your \textit{influence};\lebnote{Literally “hand”}they brought back horns of ivory and ebony as your payment.}%
\verse{Edom\lebnote{Reading Edom for Aram}was trading with you because of the abundance of your products, trading with malachite, purple wool yarn and colorful weaving, and fine white fabric and black corals and rubies; all these they exchange for your merchandise.}%
\verse{Judah and the land of Israel were trading with you with wheat from Minnith and millet and honey and olive oil and balm; all these they gave for your wares.}%
\verse{Damascus was trading with you \textit{because of}\lebnote{Literally “for”}the abundance of your products, because of the abundance of all of your wealth, trading with the wine of Helbon and white wool.}%
\verse{Vedan and Javan from Uzal, they exchanged wrought iron, cinnamon,\lebnote{Or “cassia”}and reed spice for your merchandise; all this was for your wares.}%
\verse{Dedan was trading with you, with garments of woven material for riding.}%
\verse{Arabia and all of the leaders of Kedar were \textit{your customers};\lebnote{Literally “traders of your hand”}with young rams and adult rams and goats they were trading with them with you.}%
\verse{The traders of Sheba and Raamah were trading with you, with the finest of every spice and with every precious stone and gold; they exchanged all these for your merchandise.}%
\verse{Haran and Canneh and Eden, the traders of Sheba, Assyria, and Kilmad were trading with you.}%
\verse{They were trading with you in finery, in mantles of blue cloth and colorful embroidered work and with rugs of variegated cloth in twisted cords and knotted tightly; \textit{these were among your merchandise}.\lebnote{Literally “in your trading/marketplace” (cf. NET)}}%
\verse{The ships of Tarshish were carrying for you your wares, and you were filled, and you became very heavy in the heart of the seas.}%
\verse{“ ‘Into many waters 
the rowers brought you; 
but \textit{the east wind}\lebnote{Literally “the wind of the east”}wrecked you 
in the heart of the seas.}%
\verse{Your wealth and your merchandise, your wares, 
your mariners, and your seamen, 
your shipwrights\lebnote{Literally “the repairers/strengtheners of your seams”}and the barterers of your wares 
and all of \textit{your soldiers}\lebnote{Literally “the men of your war”}who are in you, 
along\lebnote{Hebrew “and”}with all of your crew who are in the midst of you, 
they will fall in the heart of the seas on the day of your downfall.}%
\verse{At the sound of the shout\lebnote{Or “cry”}of your seamen, 
the pasturelands will shake.}%
\verse{And they will go down from their ships, 
all of those holding an oar, 
mariners, all of the seamen of the sea 
will stand on the land,}%
\verse{and they will lament\lebnote{Hebrew “lamented”}over you with their voice, 
and they will cry out\lebnote{Hebrew “cried out”}bitterly, 
and they will throw\lebnote{Hebrew “threw”}dust on their heads, 
and they will roll\lebnote{Hebrew “rolled”}in the dust.}%
\verse{\textit{They will shave themselves bald for your sake},\lebnote{Literally “And they made bald for you with baldness”}
and they will dress\lebnote{Hebrew “dressed”}themselves in sackcloth, 
and they will weep\lebnote{Hebrew “wept”}over you with a bitterness of soul\lebnote{Or “of their inner self”}
and with bitter wailing.}%
\verse{And they will raise\lebnote{Hebrew “raised”}over you with their wailing a lament, 
and they will chant\lebnote{Hebrew “chanted”}a lament over you: 
‘Who is like Tyre, 
like this destruction in the midst of the sea?’}%
\verse{When your merchandise went out from the seas, 
you satisfied many peoples with the abundance of your wealth, 
and by your wares 
you made rich the kings of the world.}%
\verse{Now you are broken by seas 
in the depths of waters; 
your wares and all of your crew 
in the midst of you have sunk.}%
\verse{All of the inhabitants of the coastlands 
are appalled over you, 
and their kings \textit{shudder}\lebnote{Literally “they have bristling hair”}in horror; 
\textit{they distort their faces}!\lebnote{Literally “they are distorted as to faces”}}%
\verse{The traders among the peoples hiss over you; 
you have become a horror, 
and \textit{you shall be no more forever}!’ ”\lebnote{Literally “there is not you until eternity/endless ages” (cf. NRSV)}}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 28
\verseWithHeading{A Prophecy Aimed at the King of Tyre}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}say to the leader of Tyre, ‘Thus says the Lord Adonai:}%
\verse{“Because your heart was haughty, 
and you said, ‘I am a god; 
I sit in the seat of the gods, 
I sit in the heart of the seas!’ 
But\lebnote{Or “And”}you are a human, not a god, 
and you gave your heart to be like the heart of a god.}%
\verse{Look, are you wiser than Daniel, 
so that \textit{no secret}\lebnote{Literally “all of hidden/secret”}is hidden from you?}%
\verse{By your wisdom and by your understanding 
you have gained for yourself wealth, 
and you have amassed gold and silver 
in your treasuries.}%
\verse{Therefore thus says the Lord Adonai:}%
\verse{“Because of your regarding your \textit{mind}\lebnote{Literally “heart”}
like the \textit{mind}\lebnote{Literally “heart”}of a god,}%
\verse{therefore look! I am bringing strangers over you, 
the most ruthless of the peoples, 
and they will draw their swords against the beauty\lebnote{Or “splendor”}of your wisdom, 
and they will defile your splendor.}%
\verse{They will thrust you down to the pit, 
and you will die \textit{a violent death}\lebnote{Literally “a death of a slain”}
in the heart of the seas.}%
\verse{\textit{Will you indeed still say}\lebnote{Literally “Will you say, will you say”}“I am a god!” 
before the face of your killers?\lebnote{Or “slayers”}
And in fact you are a human\lebnote{Or “man”}and not a god 
in the hand of \textit{those who pierce you}.\lebnote{Literally “piercers/slayers of you”} for \textit{I myself}\lebnote{Hebrew “I, I”}have spoken!” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}raise a lament over the king of Tyre, and you must say to him, ‘thus says the Lord Adonai:}%
\verse{“You were a perfect model of\lebnote{Or “for”}an example, 
full of wisdom and perfect of beauty.}%
\verse{You were in Eden, the garden of God, 
and every precious stone was your adornment: 
carnelian, topaz and moonstone,\lebnote{Meaning of Hebrew uncertain}
turquoise, onyx and jasper, 
sapphire,\lebnote{Or “lapis lazuli”}malachite and emerald. 
And gold was the craftsmanship of your settings 
and your mountings in you; 
on the day when you were created they were prepared.}%
\verse{You were an anointed guardian cherub, 
and I placed you on \textit{God’s holy mountain};\lebnote{Literally “on the mountain of the holiness of God”}
you walked in the midst of stones of fire.}%
\verse{You were blameless in your ways 
from the day \textit{when you were created},\lebnote{Literally “being created you”}
until wickedness was found in you.}%
\verse{In the abundance of your trading, 
they filled the midst of you with violence, and you sinned; 
and I cast you as a profane thing from the mountain of God, 
and I expelled you, the guardian\lebnote{Or “guarding”}cherub, 
from the midst of the stones of fire.}%
\verse{Your heart was proud because of your beauty; 
you ruined your wisdom because of your splendor. 
I threw you on the ground \textit{before}\lebnote{Literally “in the face of”}kings; 
I have exposed you \textit{for viewing}.\lebnote{Literally “to look at you”}}%
\verse{From the abundance of your iniquities 
in the dishonesty of your trading, 
I profaned your sanctuaries 
and I brought fire from your midst; 
it consumed you, and I have turned you to ashes on the earth 
\textit{before the eyes of everyone who sees you}.\lebnote{Literally “before the eyes of all seeing you”}}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}set your face toward Sidon and prophesy against it,}%
\verse{and you must say, ‘Thus says the Lord Adonai:}%
\verse{“Look! I am against you, Sidon, 
and I display my glory in the midst of you, 
and they will know that I am Adonai 
\textit{when I execute my judgments},\lebnote{Literally “in/at doing in her judgments”}
and I will show myself holy within it.}%
\verse{And \textit{there will not be any longer}\lebnote{Literally “and not it will be again”}a painful thorn and a sharp thornbush for\lebnote{Or “against”}the house of Israel from \textit{anywhere around them}\lebnote{Literally “from all of all around them”}from those \textit{who are despising them},\lebnote{Literally “the ones despising them”}and they will know that I am the Lord Adonai.” ’ ”}%
\verse{Thus says the Lord Adonai: “When I gather the house of Israel from the peoples to which they were scattered about in them, and I show myself holy in\lebnote{Or “among”}them before the eyes of the nations, then they will live on their soil, which I gave to my servant Jacob.}%
\verse{And they will live on it in safety, and they will build houses, and they will plant vineyards, and they will live in safety when I execute my judgments on all \textit{those who despise them from all around them},\lebnote{Literally “the ones despising them”}and then they will know that I am Adonai their God.”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 29
\verseWithHeading{Adonai’s Judgment upon Egypt}{In the tenth year, in the tenth month, on the \textit{twelfth}\lebnote{Literally “two ten”}day of the month, the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}set your face against Pharaoh, the king of Egypt, and prophesy against him and against Egypt, all of it.}%
\verse{Speak, and you must say, ‘thus says the Lord Adonai:}%
\verse{“Look! I am against you, 
Pharaoh, king of Egypt, 
the great sea monster, 
the one lying down 
in the midst of his Nile streams, 
who says to me, “It is my Nile, and I made it for myself.”\lebnote{Hebrew “me”}}%
\verse{And so I will put hooks in your jawbones, 
and I will make the fish of your Nile streams stick to your scales, 
and I will bring you up from the midst of your Nile streams, 
and all of the fish of your Nile streams which cling to your scales.}%
\verse{And I will fling you to the desert, 
you and all the fish of your Nile streams. 
On the surface of the field you will fall; 
you will not be gathered, 
and you will not be assembled. 
To the animals\lebnote{Hebrew “animal”}of the field 
and to the birds\lebnote{Hebrew “bird”}of the heavens 
I will give you as food.}%
\verse{And all of the inhabitants of Egypt will know that I am Adonai, 
because of their\lebnote{LXX, Syriac, and Vulgate: “you”}being a staff of reed for the house of Israel.}%
\verse{Therefore thus says the Lord Adonai: “Look! I am bringing on you a sword, and I will cut off from you human and animal.}%
\verse{And the land of Egypt will become\lebnote{Hebrew “be”}a desolation and ruins, and they will know that I am Adonai because he\lebnote{That is, Pharaoh}said, ‘The Nile is \textit{mine}!\lebnote{Literally “to me”}And I, I made it!’}%
\verse{Therefore look! I am against you and against your Nile streams, and I will make the land of Egypt into ruins, a pile of rubble, a desolation from Migdol to Syene and up to the boundary of Cush.}%
\verse{A foot of a human will not pass over it, and a foot of an animal will not pass over it, and so it will not be inhabited for forty years.\lebnote{Hebrew “year”}}%
\verse{And I will make the land of Egypt a desolation in the midst of desolated countries, and its cities to be in the midst of ruined cities. They will be a desolation for forty years, and I will scatter Egypt among the nations, and I will disperse them among the countries.”}%
\verse{For thus says the Lord Adonai: “At the end of forty years I will gather Egypt from the peoples among whom they were scattered.}%
\verse{And I will restore the fortunes of Egypt, and I will bring them back to the land of Pathros, to the land of their origin, and they will be a lowly kingdom there.}%
\verse{Of all the kingdoms it will be the most lowly, and it will not exalt itself again over the nations, and I will make them small so as not to rule over the nations.}%
\verse{And it will not be again for the house of Israel an object of trust, bringing to remembrance their guilt \textit{when they turned back to them},\lebnote{Literally “at their turning”}and they will know that I am the Lord Adonai.”}%
\verse{\textit{And then}\lebnote{Literally “and it happened”}in \textit{the twenty-seventh year},\lebnote{Literally “twenty and seven year”}in the first month, on the first day of the month, the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}Nebuchadnezzar the king of Babylon, he made his army labor hard against Tyre. Every head was rubbed bare, and every shoulder was rubbed raw, but\lebnote{Or “and”}a wage was not paid for him and for his army from Tyre for the labor that he did against it.”}%
\verse{Therefore thus says the Lord Adonai: “Look! I am giving to Nebuchadnezzar the king of Babylon the land of Egypt, and he will carry its wealth away, and he will plunder its plunder, and he will loot its loot, and it will be a wage for his army.}%
\verse{As his wages that he worked for, I will give to him the land of Egypt, because they worked for me,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“On that day I will cause \textit{power}\lebnote{Literally “a horn”}to grow up for the house of Israel, and for you I will give an opening of your mouth in the midst of them, and they will know that I am Adonai.”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 30
\verseWithHeading{A Lament over the Nation and Land of Egypt}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}prophesy, and you must say, ‘thus says the Lord Adonai:}%
\verse{“Wail, alas! For the day!}%
\verse{For a day is near; 
indeed, a day is near for Adonai. 
A day of cloud, 
a time of the nations it will be.}%
\verse{Cush and Put and Lud and all of Arabia and Kub and the people\lebnote{Or “children”}of the land of the treaty\lebnote{Or “covenant”}with them—by the sword they will fall.”}%
\verse{Thus says Adonai:}%
\verse{“And the supporters of Egypt will fall, 
and the majesty of its strength will go down 
from Migdol to Syene; 
by the sword they will fall in it,” 
\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“And they will be desolate in the midst of desolate countries, 
and its cities will be in the midst of ruined cities.}%
\verse{On that day messengers will go down from \textit{before me}\lebnote{Literally “to the face of me”}in the ships to terrify unsuspecting Cush, and anguish will be in them on the day of Egypt; for look, it is coming!”}%
\verse{Thus says the Lord Adonai:}%
\verse{“And I will put an end to the crowds\lebnote{Or “hordes”}of Egypt 
by the hand of Nebuchadnezzar the king of Babylon.}%
\verse{He and his people with him, the most ruthless of nations, 
will be brought to destroy the land. 
And they will draw their swords against Egypt, 
and they will fill the land with the slain, I, Adonai, I have spoken.”}%
\verse{Thus says the Lord Adonai:}%
\verse{“And I will destroy the idols, 
and I will put an end to the worthless idols from Memphis. 
And there will \textit{no longer be}\lebnote{Literally “not be any longer”}a prince from the land of Egypt, 
and I will put fear in the land of Egypt.}%
\verse{And I will lay waste Pathros, 
and I will set fire in Zoan, 
and I will execute judgments in No.\lebnote{That is, Thebes}}%
\verse{I will pour out my rage over Sin, 
the stronghold of Egypt, 
and I will cut off the crowd\lebnote{Or “horde”}of No.\lebnote{That is, Thebes}}%
\verse{And I will set fire in Egypt, 
certainly Sin will writhe, 
and No\lebnote{That is, Thebes}will be breached, 
and Memphis will face enemies \textit{daily}.\lebnote{Literally “of by day”}}%
\verse{The young men of On and Pi Beseth will fall by the sword, 
and they will go into captivity.}%
\verse{And at Tahpanhes the day will become dark 
when I break the yoke of Egypt, 
and in it the majesty of its strength 
will come to an end. 
A cloud will cover it, 
and its daughters will go into captivity.}%
\verse{\textit{And then}\lebnote{Literally “and it was”}\textit{in the eleventh}\lebnote{Literally “on one ten”}year, in the first month, on the seventh day of the month, the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}the arm of Pharaoh, the king of Egypt, I have broken. And look, it has not been bound up for giving of a remedy, or for the placing of a splint to bind it up to make it strong to take hold of the sword.”}%
\verse{Therefore thus says the Lord Adonai: “Look! I am against Pharaoh, the king of Egypt, and I will break his strong arm and the broken arm, and I will let the sword fall from his hand.}%
\verse{And I will scatter Egypt among the nations, and I will disperse them among the countries.}%
\verse{And I will strengthen the arm of the king of Babylon, and I will give my sword into his hand, and I will break the arms of Pharaoh, and he will groan with the groaning of the mortally wounded \textit{before him}.\lebnote{Literally “the face of him”}}%
\verse{And I will strengthen the arm of the king of Babylon, and the arms of Pharaoh will fall. And they will know that I am Adonai when I give my sword into the hand of the king of Babylon, and he will stretch it out to the land of Egypt.}%
\verse{And I will scatter Egypt among the nations, and I will disperse them into the countries, and they will know that I am Adonai.”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 31
\verseWithHeading{Pharaoh Is Warned Through Assyria’s Destruction}{\textit{And then}\lebnote{Literally “and it happened/it was”}\textit{in the eleventh}\lebnote{Literally “on one ten”}year, in the third month, on the first day of the month, the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}say to Pharaoh, the king of Egypt, and to his crowd,\lebnote{Or “hordes”}}%
\verse{‘To whom are you like in your greatness?}%
\verse{Look! Assyria was a cedar in Lebanon, 
with beautiful branches\lebnote{Hebrew “branch”}and a forest giving shade, 
and \textit{very high},\lebnote{Literally “high of height”}
and its treetop was between the clouds.}%
\verse{Waters made it great, 
the deep made it grow high; 
its rivers were going all around \textit{its planting area},\lebnote{Literally “base of its tree”; NRSV, “in the place where it was planted”}
and its channels it sent out 
to all of the trees of the field.}%
\verse{Therefore it became tall, 
with its height more than all of the trees of the field, 
and its branches became numerous, 
and its branches became long 
\textit{from its sending its shoots from abundant water}.\lebnote{Literally “from water much at sending shoots its”}}%
\verse{In its branches all the birds\lebnote{Hebrew “bird”}of the heaven\lebnote{Or “sky”}made their nest, 
and under its branches all the animals\lebnote{Hebrew “animal”}of the field gave birth, 
and in its shadow all the many nations lived.}%
\verse{And it was beautiful in its greatness, 
in the length of its branches, 
for its root was toward much water.}%
\verse{Cedars in the garden of God 
could not be equal to it; 
fir trees\lebnote{Or possibly, juniper}could not resemble its branches, 
and plane trees were not even like its branches; 
any tree even in the garden of God 
could not resemble it in its beauty.}%
\verse{Therefore thus says the Lord Adonai: “\textit{Because}\lebnote{Literally “because that”}it was tall in height and it set its treetop between\lebnote{Hebrew “to between”}thick clouds, and \textit{he took pride in his tallness},\lebnote{Literally “and his heart he raised in his tallness”}}%
\verse{then\lebnote{Or “and”}I gave it into the hand of the leader of nations; he \textit{dealt thoroughly}\lebnote{Literally “doing he did”}with it according to its wickedness. I drove it out.}%
\verse{And strangers cut it off, the most ruthless of nations, and they abandoned it. On the mountains and in all of the valleys its branches fell, and its branches were broken in all the river channels of the land, and all the peoples of the world went out from its shadow, and they abandoned it.}%
\verse{On its fallen trunk all the birds\lebnote{Hebrew “bird”}of the heaven\lebnote{Or “sky”}now dwell, and all the animals of the field were on its branches.}%
\verse{This occurred so that all of the trees with abundant water will not become tall, and they will not set their treetop between\lebnote{Or “among”}their thick foliage,\lebnote{Or “clouds”}and so that\textit{all of the trees that are abundantly watered}\lebnote{Literally “all of the drinkers of water”}\textit{will not stand up to them}\lebnote{Literally “will not stand to them”}in their tallness, for all of them, they have been given over to death, to the world below in the midst of \textit{mortals},\lebnote{Literally “sons of men,” or “children of humankind”}to the people going down to the grave.”\lebnote{Or “pit”}}%
\verse{Thus says the Lord Adonai: “On the day of its going down to Sheol, I caused mourning; I covered over it with the deep, and I withheld its rivers, and many waters were restrained, and I brought gloom over it; Lebanon and all of the trees of the field, they had fainted because of it.}%
\verse{From the sound of its downfall I caused nations to shake when I made it go down to Sheol, with the people going down to the grave,\lebnote{Or “the pit”}and so in the world below all of the trees of Eden, the choice and the best of Lebanon, \textit{all the well-watered trees}\lebnote{Literally “all ofthe drinkers of water”}were comforted!}%
\verse{They also went down with it to Sheol to \textit{those who died by the sword},\lebnote{Literally “to the slain of the sword”}and its army who had lived in its shadow in the midst of nations.}%
\verse{\textit{To whom could you be compared, whether in glory or in majesty}\lebnote{Literally “To who you are like as in glory and in majesty”}among the trees of Eden? And yet you will be brought down with the trees of Eden to the world below; in the midst of the uncircumcised you will lie with \textit{those who died by the sword}.\lebnote{Literally “the slain of the sword”}That is Pharaoh and \textit{his entire}\lebnote{Literally “all his hordes/crowd”}crowd!”\lebnote{Or “hordes”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 32
\verseWithHeading{A Lament for Pharaoh, King of Egypt}{\textit{And then}\lebnote{Literally “and it was/happened”}in \textit{the twelfth}\lebnote{Literally “two of ten”}year, in the \textit{twelfth}\lebnote{Literally “in two ten month”}month, on the first day of the month, the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}raise a lament over Pharaoh, the king of Egypt, and you must say to him, ‘With a fierce, strong lion among nations you compared yourself, 
and you are like the sea monster in the seas, 
and you thrash about in your rivers, 
and you make water turbid with your feet, 
and you make your rivers muddy.}%
\verse{Thus says the Lord Adonai:}%
\verse{Now I will spread my net over you 
in the assembly of many peoples, 
and I will bring you up in my dragnet.}%
\verse{And I will throw you on the ground; 
on the surface of the open field I will hurl you, 
and I will cause every bird of the heaven to dwell on you, 
and I will satisfy the animals\lebnote{Hebrew “animal”}of all of the world from you.}%
\verse{And I will put your flesh on the mountains, 
and I will fill the valleys with your carcass.}%
\verse{And I will water the land with your discharge from your blood 
on the mountains, 
and valleys will be filled from\lebnote{Or “with”}you.}%
\verse{And I will cover you at\lebnote{Or “by”}extinguishing your heavens, 
and I will make dark their stars, 
and I will cover the sun with the cloud, 
and the moon will not give its light. \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“And I will disturb the hearts\lebnote{Hebrew “heart”}of many peoples 
at my bringing about your captivity among the nations, 
to countries that you do not know.}%
\verse{And I will cause many peoples to be awestruck over you, 
and their kings will shudder over you in horror, 
at\lebnote{Or “when”}my brandishing my sword before their faces, 
and they will tremble \textit{continually},\lebnote{Literally “for moments/seconds”}
each person for his life, 
on the day of your downfall.”}%
\verse{For thus says the Lord Adonai:}%
\verse{“The sword of the king of Babylon, 
it will come to you.}%
\verse{By the swords of warriors I will cause to fall your hordes 
by the most ruthless of the nations, all of them, 
and they will devastate the majesty of Egypt, 
and all of its hordes will be destroyed.}%
\verse{And I will destroy all her domestic livestock 
\textit{beside}\lebnote{Literally “from on”}many waters. 
The feet of humans\lebnote{Hebrew “human”}will not make them turbid again, 
and the hooves of domestic livestock will not make them turbid. \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“Then I make the land of Egypt a desolation, 
so that the land will be stripped from its fullness 
when I strike all of those dwelling in it, 
and they will know that I am Adonai.}%
\verse{\textit{This is}\lebnote{Literally “she is”}a lament, and they will chant it as a lament; the daughters of the nations will chant it as a lament over Egypt, and over all of its hordes they will chant it as a lament,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{\textit{And then}\lebnote{Literally “and it happened/was”}in the \textit{twelfth}\lebnote{Literally “two of ten”}year, \textit{on the fifteenth}\lebnote{Literally “in the five ten of”}day of the month, the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}mourn over the hordes of Egypt, and make her go down, and with her the daughters of mighty nations, \textit{down to the deep underworld}\lebnote{Literally “to the world/earth deeps/depth”}with the people going down to the grave.\lebnote{Or “pit” or “netherworld”}}%
\verse{\textit{You are more lovely than whom}?\lebnote{Literally “from whom are you lovely”}Go down and be laid in rest with the uncircumcised.}%
\verse{In the midst of the people slain by the sword it\lebnote{Or “she”; that is, Egypt}is given they will fall to a sword; they carried her\lebnote{Or “it”}off and all of her\lebnote{Or “its”}hordes.}%
\verse{The chiefs of the warriors will speak to him from the middle of Sheol; with his helpers they have gone down; the uncircumcised lie still, \textit{those slain by the sword}.\lebnote{Literally “the slain of the sword”}}%
\verse{Assyria is there and all of its assembly; all around it are its graves, all of them killed, those fallen by the sword,}%
\verse{who\lebnote{That is, Assyria}will be given its graves in the remote areas of the pit, and its assembly will be all around its grave, all of them killed, fallen by the sword, those who spread terror in the land of the living.}%
\verse{Elam is there and all of its hordes, all around its grave, all of them dead who fell by the sword, those who went down uncircumcised to the depths of the underworld,\lebnote{Literally “land/earth/world”}those who spread their terror to the land of the living, and now they bear their disgrace with the people going down to the grave.\lebnote{Or “pit” or “sheol”}}%
\verse{In the midst of the slain they made a bed for her with all of her hordes all around its graves, all of them uncircumcised, slain by the sword, for their terror was spread in the land of the living, and they now bear their disgrace with the people going down to the grave;\lebnote{Or “pit”}in the midst of the slain it was placed.}%
\verse{Meshech and Tubal are there and all of its hordes, all around him its graves, all of them uncircumcised and killed by the sword, for they gave their terror in the land of the living.}%
\verse{And they do not lie with warriors fallen from long ago, who went down to Sheol with their weapons of war,\lebnote{Literally “with the weapons of war their”}and they placed their sword under their heads, and their shields\lebnote{Reading “their shields” for “their guilt”}were on their bones, for the terror of the warriors was in the land of the living.}%
\verse{And you\lebnote{Evidently refers to Egypt}too in the midst of uncircumcised people will be broken, and you will lie with those slain by the sword.}%
\verse{There is Edom, its kings and all of its leaders who are laid along with their might with those killed by the sword; they will lie with the uncircumcised and with the people going down to the grave.\lebnote{Or “pit”}}%
\verse{There are also the princes of the north, all of them, all of the Sidonians who have gone down, being ashamed to lie with the slain because\lebnote{Or “in spite of” (cf. NJPS, NIV)}of their terror that they caused from their might, and they lie uncircumcised with those killed by the sword, and they bear their disgrace along with the people going down to the grave.\lebnote{Or “pit”}}%
\verse{Pharaoh will see them, and he will be comforted over all of his hordes killed by the sword, Pharaoh and all of his army,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“For\lebnote{Or “although”}he spread\lebnote{Hebrew “I gave”}my terror in the land of the living, and so he will be laid down in the midst of the uncircumcised with those killed by the sword: Pharaoh and all of his hordes,”\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 33
\verseWithHeading{Ezekiel Appointed as a Watchman and His Responsibility}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}speak to \textit{your people},\lebnote{Literally “the children of your people”}and you must say to them, ‘A land, if I bring over it a sword and the people of the land take a man, one from their number, and they appoint him for them as a watchman,}%
\verse{and he sees the sword coming against the land, and he blows on the horn and he warns the people,}%
\verse{and \textit{anyone who listens}\lebnote{Literally “the listening”}hears the sound of the horn and \textit{he does not take warning}\lebnote{Literally “not he takes warning”}and the sword comes and it takes him, his blood will be on his own head.}%
\verse{For he heard the sound of the horn and he did not take warning; his blood will be on him. But if he took warning, he saved his life.}%
\verse{And as for the watchman, if he sees the sword coming and \textit{he does not blow}\lebnote{Literally “and not he blows on”}the horn, and the people are not warned, and the sword comes and it takes \textit{their lives},\lebnote{Literally “it takes from them a life”}he will be taken through his guilt, but his blood from the hand of the watchman I will seek.’}%
\verse{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}I have made you a watchman for the house of Israel; if you hear a word from my mouth, then you must warn them from me.}%
\verse{\textit{When I say}\lebnote{Literally “At my saying”}to the wicked, ‘Wicked one, you will certainly die,’ and you \textit{did not speak}\lebnote{Literally “and not you spoke”}to warn the wicked from his way, he, the wicked, will die by his guilt, but his blood I will seek from your hand.}%
\verse{But, you, if you warn the wicked from his way, to turn from it, and \textit{he does not turn}\lebnote{Literally “and not he turns”}from his way, he will die by his guilt, and you will have saved your life.}%
\verse{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}say to the house of Israel, \textit{saying},\lebnote{Literally “to say”}‘So you said, \textit{saying},\lebnote{Literally “to say”}“Indeed, our transgressions and our sins are on us, and because of them we are rotting, and how can we live?” ’}%
\verse{Say to them, ‘\textit{As I live},’\lebnote{Literally “live I”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, ‘\textit{Surely I have no delight}\lebnote{Literally “if”}in the death of the wicked, \textit{except}\lebnote{Literally “for it”}in the wicked returning from his way, and he lives. Turn back! Turn back from your ways, O evil ones, for why should you die, house of Israel?’}%
\verse{“So\lebnote{Or “And”}you, son of man,\lebnote{Or “mortal,” or “son of humankind”}say to \textit{your people}\lebnote{Literally “the children of your people”}that the righteousness of the righteous shall not save him in the day of his transgression, and the wickedness of the wicked \textit{will not cause him to stumble}\lebnote{Literally “not he will stumbled by it”}on the day of his returning from his wickedness; and the righteous will not be able to live by it on the day when he returns to his sin.}%
\verse{When I say to the righteous, ‘Certainly he will live,’ and he trusted in\lebnote{Or “on”}his righteousness, and he turns and he does injustice, all of his righteousness\lebnote{According to the reading tradition (\textit{Qere})}will not be remembered, and because of his injustice that he did, because of it he will die.}%
\verse{And when I say to the wicked, ‘Certainly you will die,’ but\lebnote{Or “and”}he returns from his sin and he does justice and righteousness—}%
\verse{for example, the wicked returns a pledge for a loan, he restores \textit{stolen property},\lebnote{Literally “robbed/stolen items”}he goes in the statutes of life so as not to do injustice—certainly he will live; he will not die.}%
\verse{All of his \textit{sins that he committed},\lebnote{Literally “sins that he sinned”}they will not be remembered against\lebnote{Or “for”}him, and he did justice and righteousness; certainly he will live.}%
\verse{“Yet\lebnote{Or “And”}\textit{your people}\lebnote{Literally “the children of your people”}say, ‘The way of the Lord is not fair!’ yet their way is not fair.}%
\verse{When the righteous turns from his righteousness and does injustice, then he will die \textit{because of it}.\lebnote{Literally “by/in them”}}%
\verse{And when the wicked turns from his wickedness, and he does justice and righteousness, because of it\lebnote{Hebrew “them”}\textit{he will certainly live}!\lebnote{Literally “he, he will live”}}%
\verse{Yet you said, ‘The way of the Lord is not fair!’ I will judge you, house of Israel, each person according to his ways.”}%
\verse{\textit{And then}\lebnote{Literally “and it happened”}it was in \textit{the twelfth}\lebnote{Literally “two of ten”}year, in the tenth month, on the fifth day of the month of our exile, a survivor from Jerusalem came to me, \textit{saying},\lebnote{Literally “to say”}“The city was destroyed!”}%
\verse{And the hand of Adonai was on me on the evening \textit{before}\lebnote{Literally “to the face of”}the coming of the survivor, and he opened my mouth \textit{before the survivor came to me in the morning},\lebnote{Literally “until coming to me in the morning”}and my mouth was opened, and I was \textit{no longer}\lebnote{Literally “and not I was dumb still”}dumb.\lebnote{Or “silent” or “speechless”}}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}the inhabitants of these ruins on the soil of Israel are saying, ‘Abraham was one man, and he took possession of the land, and we are many; the land has been given for\lebnote{Or “to”}us as a possession.’}%
\verse{Therefore say to them, ‘Thus says the Lord Adonai: “You eat blood in your meat, and you raise your eyes to your idols and you pour out blood, and yet you want to take possession of the land?\lebnote{Or “you would take possession of”}}%
\verse{You rely on your sword, you do a detestable thing, and \textit{each man defiles}\lebnote{Literally “a man the wife of neighbor you defile”}the wife of his neighbor, and yet you want to take possession of the land?” ’\lebnote{Or “you would take possession of the land”}}%
\verse{“Thus you must say to them, ‘Thus says the Lord Adonai: “\textit{As I live},\lebnote{Literally “I live”}\textit{surely}\lebnote{Literally “if not”}whoever is in the ruins, by the sword they will fall, and whoever is on the surface of the open field, I will give him to the animals\lebnote{Hebrew “animal”}to eat him, and whoever is in the stronghold and in the cave, they will die by the plague.}%
\verse{And I will make the land a desolation and a wasteland, and the pride of its strength will come to an end, and the mountains of Israel will be desolate \textit{with no one traveling through them}.” ’\lebnote{Literally “from there is not someone going over”}}%
\verse{And they will know that I am Adonai when I make my land a desolation and a wasteland, because of all of their detestable things that they have done.}%
\verse{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}\textit{your people},\lebnote{Literally “the children of your people”}the ones talking together concerning you beside the walls and in the doorways of the houses, each one with his brother \textit{saying},\lebnote{Literally “to say”}‘Please come and hear what is the word \textit{that is going out from Adonai}.’\lebnote{Literally “the going out from with Adonai”}}%
\verse{And they come to you \textit{as people do},\lebnote{Literally “like the coming of a people”}and they sit \textit{before you},\lebnote{Literally “to the face of you”}and my people hear your words, and they do not do them, for they are showing\lebnote{Or “expressing”}passion in their mouth, but their heart is going after ill-gotten gain.}%
\verse{Now, look! You are to them like \textit{a sensual song},\lebnote{Literally “like a song of lust”}beautiful of voice and \textit{played well on an instrument},\lebnote{Literally “doing good to play a music instrument”}and they hear your words, \textit{but they are not doing them}.\lebnote{Literally “and/but doing there is not they them”}}%
\verse{So\lebnote{Or “And”}\textit{when it comes},\lebnote{Literally “in/at her coming”}look! It is coming! And then they will know that a prophet was in the midst of them.”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 34
\verseWithHeading{Prophecy Concerning Israel’s Shepherds and Sheep}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}prophesy against the shepherds of Israel, prophesy, and you must say to them, to the shepherds, ‘Thus says the Lord Adonai: “Woe to the shepherds of Israel who were feeding themselves! Must not the shepherds feed the flock?}%
\verse{The fat you eat, and you clothe yourself with the wool; the well-nourished animals you slaughter, but you do not feed the flock.}%
\verse{The weak you have not strengthened, and the sick you have not healed, and \textit{with respect to}\lebnote{Literally “for”}the hurt\lebnote{Or “wounded”}you have not bound them up, and you have not brought back the scattered, and you have not sought the lost, but rather you ruled over them with force\lebnote{Or “violence”}and with ruthlessness.}%
\verse{And they were scattered \textit{without}\lebnote{Literally “from not”}a shepherd, and they were\lebnote{Or “became”}as food for all the animals\lebnote{Hebrew “animal”}of the field when they were scattered.}%
\verse{My flock went astray upon all of the mountains and on every high hill, and so upon all the surface of the world\lebnote{Or “earth”}my flock were scattered, and \textit{there was no one seeking them},\lebnote{Literally “there was not someone seeking them”}and \textit{there was no one searching for them}.”\lebnote{Literally “there was not someone searching for them”}}%
\verse{Therefore, hear, O shepherds, the word of Adonai:}%
\verse{“\textit{As I live},”\lebnote{Literally “live I”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, “\textit{Surely}\lebnote{Literally “if not”}because my flock have become as plunder, and my flock became as food to\lebnote{Or “for”}all the animals\lebnote{Hebrew “animal”}of the field, since there was not a shepherd, since my shepherds have not sought my flock, but the shepherds fed themselves, and they fed not my flock,’ ”}%
\verse{therefore, O shepherds, hear the word of Adonai,}%
\verse{‘Thus says the Lord Adonai: “Look! I am against the shepherds, and I will seek my flock from their hand, and I will put an end to them from shepherding flocks, and the shepherds will \textit{no longer}\lebnote{Literally “not any longer”}feed themselves, and I will deliver my flocks from their mouth, so that they will not be as food for them.”}%
\verse{“ ‘For thus says the Lord Adonai: “Look! I, even I, will seek my flock, and I will look after them,}%
\verse{just like the caring of a shepherd for his herd on\lebnote{Or “in”}the day when he is in the midst of his scattered flock. Thus I will look after my flock, and I will deliver them from all the places \textit{to which they were scattered}\lebnote{Literally “which they were scattered there”}on the day \textit{of storm and stress}.\lebnote{Literally “on the day of cloud and deep gloom”}}%
\verse{And I will bring them out from the peoples, and I will gather them from the countries, and I will bring them to their soil, and I will feed them on the mountains of Israel, in the valleys, and in all of the settlements of the land.}%
\verse{I will feed them in good pasture, and their pasture will be on the mountains of the heights of Israel; there they will lie down in good pasture, and on lush pasture they will feed on the mountains of Israel.}%
\verse{\textit{I myself}\lebnote{Literally “I, I”}will feed my flock and \textit{I myself}\lebnote{Literally “I, I”}will allow them to lie down,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“I will seek the lost, and I will bring back the scattered, and I will bind up the one hurt, and I will strengthen the sick; and the fat and the strong I will destroy. I will feed her with justice.”}%
\verse{“ ‘And you, my flock, thus says the Lord Adonai: “Look! I am judging between \textit{one sheep and another},\lebnote{Literally “small sheep and small sheep”}between the rams and between the he-goats.}%
\verse{\textit{Is it not enough for you}\lebnote{Literally “is it too little from you”}that you feed on the good pasture? And still you must trample the remainder of your pasture with your feet, and \textit{clear water}\lebnote{Literally “and settling of water”}you drink, and the \textit{leftover water}\lebnote{Literally “water being left”}you must make muddy with your feet.}%
\verse{And my flock must graze the pasture treaded upon by\lebnote{Hebrew “of”}your feet, and must drink the mud puddle stirred up by\lebnote{Hebrew “of”}your feet.”}%
\verse{“ ‘Therefore thus says the Lord Adonai to them: “Look! I, even I will judge between fat sheep and between lean sheep,}%
\verse{because with your flank and with your shoulder you shoved, and with your horns you pushed all of the sick animals \textit{until}\lebnote{Literally “until that”}you scattered it\lebnote{Or “them”}to the outside.}%
\verse{And so I will save my flock, and \textit{they will no longer be}\lebnote{Literally “not they will be longer for plunder”}for plunder, and I will judge between \textit{one sheep and another}.\lebnote{Literally “sheep and sheep”}}%
\verse{And I will set up over them one shepherd, and he will feed them; that is, my servant David. He will feed them, and he will be for them as\lebnote{Or “like”}a shepherd.}%
\verse{And I, Adonai, I will be for them \textit{as God},\lebnote{Literally “to a God”}and my servant David will be a leader in the midst of them. I, Adonai, I have spoken.}%
\verse{And I will make them a covenant of peace, and I will put an end to wild animals\lebnote{Hebrew “animal”}from the land, and they will dwell in the desert \textit{safely},\lebnote{Literally “with confidence/trust”}and they will sleep in the forest.}%
\verse{And I will make them and the area all around my hill a blessing, and I will let the rain go down \textit{at its appointed time};\lebnote{Literally “in time his”}they will be rains of blessing.}%
\verse{And the tree\lebnote{Or “trees”}of the field will give its fruit, and the land will give its produce, and they will be on their land \textit{safely},\lebnote{Literally “with/in conficence”}and they will know that I am Adonai when I break their yoke, and I will deliver them from the hand of the ones enslaving them.}%
\verse{And they will not be any longer plunder to the nations, and the animals of the land will not eat them, and they will dwell \textit{in safety},\lebnote{Literally “with confidence”}and there will not be anyone frightening them.}%
\verse{And I will raise for them a garden plot\lebnote{Or “planting”}of renown, and they will \textit{no longer}\lebnote{Literally “not any longer”}be victims of famine in the land, and they will not bear any more the insult of the nations.}%
\verse{And they will know that I, Adonai their God, am with them and they are my people, the house of Israel,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai,}%
\verse{“and you are my flock, the flock of my pasture. You are \textit{my people};\lebnote{Literally “humanity, Mankind, human, mankind”}I am your God,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 35
\verseWithHeading{Prophecy Directed at Edom}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}set your face against the mountain of Seir and prophesy against it,}%
\verse{and you must say to it, ‘Thus says the Lord Adonai: “Look! I am against you, mountain of Seir, and I will stretch out my hand against you, and I will make you a desolation and a wasteland.}%
\verse{Your cities I will make ruins, and you will be a desolation, and then you will know that I am Adonai,}%
\verse{because there has been to you \textit{an ancient hostility},\lebnote{Literally “enmity of eternity”}and you handed over the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}to the power of the sword at the time of their disaster, at the time of \textit{their final punishment}.\lebnote{Literally “the time of punishment of end”}}%
\verse{Therefore \textit{as I live},”\lebnote{Literally “live I”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, “Certainly to blood guilt I will prepare you and blood\lebnote{Or “blood vengeance”}will pursue; \textit{since}\lebnote{Literally “if not”}\textit{you did not hate blood},\lebnote{Literally “not blood you hated”}it will pursue you.}%
\verse{And I will make the mountain of Seir into a desolation and ruin, and I will cut off from it both he who is crossing over and he who is returning.}%
\verse{And I will fill its mountains with its slain; your hills and your valleys and all of your watercourses, the slain by the sword will fall in them.}%
\verse{\textit{An everlasting desolation}\lebnote{Literally “of eternity/endless ages”}I will make you, and your cities will not return, and you will know that I am Adonai.}%
\verse{I do this because of your saying,\lebnote{Or “thinking”}‘The two nations and the two lands, they will be \textit{mine},\lebnote{Literally “to me”}and we will take possession of it,’ and yet Adonai was there.}%
\verse{Therefore, \textit{as I live},”\lebnote{Literally “live I”}\textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, “so I will deal with you according to your anger and according to your jealousy that you did,\lebnote{Or “showed” or “acted on”}because of your hatred against them, and I will make myself known among them when I judge you.}%
\verse{And you will know that I, Adonai, I have heard all of your contemptible words that you said\lebnote{Or “spoke”}against the mountains of Israel, \textit{saying},\lebnote{Literally “to say”}‘They will be desolate; they are given to us as food.’}%
\verse{And you magnified yourself against me with your \textit{insolent speech},\lebnote{Literally “with your mouth”}and you spoke voluminously against me with your words—I heard!}%
\verse{Thus the Lord Adonai says, ‘\textit{As the whole world rejoices},\lebnote{Literally “as/like the rejoicing of all of the world”}I will make you a desolation.}%
\verse{\textit{Like your rejoicing over the inheritance of the house of Israel}\lebnote{Literally “as/like rejoicing your for the inheritance of the house of Israel”}\textit{because}\lebnote{Literally “because that”}it was desolate, I will do to you; you will be a desolation, mountain of Seir and all of Edom, indeed all of it; and they will know that I am Adonai.’ ” ’}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 36
\verseWithHeading{Prophecy to the Mountains of Israel and of Israel’s Restoration}{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}prophesy to the mountains of Israel; and so you must say, ‘Mountains of Israel, hear the word of Adonai;}%
\verse{thus says the Lord Adonai: “Because the enemy said concerning you, ‘Ah, and\lebnote{Or “now”}the ancient high places will be to us as a possession!’ ” ’}%
\verse{Therefore prophesy, and you must say, ‘Thus says the Lord Adonai: “\textit{Because indeed}\lebnote{Literally “with because”}\textit{when you were desolate},\lebnote{Literally “you being desolate”}then they crushed you from all around, so that you became a possession to the remainder\lebnote{Or “the rest of”}of the nations, and you became \textit{a byword and object of slander for the people}.”\lebnote{Literally “on lips of the tongue and slander of the people”}}%
\verse{Therefore, mountains of Israel, hear the word of the Lord Adonai; thus says the Lord Adonai to the mountains and to the hills, to the river channels and to the valleys and to the ruins that are desolate and to the cities that are abandoned and that have become as plunder and as scorn for the remainder\lebnote{Or “the rest of”}of the nations that are from all around.}%
\verse{Therefore thus says the Lord Adonai: “\textit{Certainly}\lebnote{Literally “if not”}in the fire of my passion I spoke against the rest of the nations and against all\lebnote{Literally “against Edom all of it”}Edom, who made my land as a possession, because of\lebnote{Or “in order to plunder its pastureland”}its pastureland for plunder, for themselves in all of the joy of their whole heart and in their inner disdain.”\lebnote{Or “malice/disrespect”}}%
\verse{Therefore prophesy against the land of Israel, and you must say to the mountains and to the hills, to the river channels and to the valleys, thus says the Lord Adonai: “Look! I spoke in my passion and in my anger the insult of the peoples you have endured.”}%
\verse{Therefore thus says the Lord Adonai: “\textit{I swear}\lebnote{Literally “I, raise my hand”}\textit{certainly}\lebnote{Literally “if not”}that the nations \textit{that are all around you},\lebnote{Literally “that are to you from all around”}they indeed will endure their disgrace.}%
\verse{“ ‘ “But you, mountains of Israel, your branch you will shoot, and your fruit you will carry\lebnote{Or “bear”}for my people Israel, for they are soon to come.}%
\verse{For look! I am for you, and I will turn you, and you will be tilled, and you will be planted.}%
\verse{And I will cause \textit{your population}\lebnote{Literally “humans/humankind”}to increase for you, all of the house of Israel, all of it, and the towns will be inhabited, and the ruins will be built.}%
\verse{And I will cause \textit{your population}\lebnote{Literally “humans/humankind”}to increase for you, and domestic animals, even they will be numerous, and they will be fruitful, and I will cause you to be inhabited like your \textit{ancient times},\lebnote{Literally “past days/times/ancient times”}and I will do good things more than in your former times, and you will know that I am Adonai.}%
\verse{And I will send over you my people Israel, and they will take possession of you, and you will be for them as an inheritance, and \textit{you will no longer bereave them of their children}.”\lebnote{Literally “not you will again longer to bring bereavement for them”}}%
\verse{“ ‘Thus says the Lord Adonai, because they are saying to you, “You are an eater of humans,\lebnote{Or “peoples”}and you are \textit{one who makes childless}\lebnote{Literally “maker of childless”}your people.”\lebnote{That is, Israel and Judah; Cf. Ezek 37:15–22}}%
\verse{Therefore humans\lebnote{Or “peoples”}you will not eat any longer, and your nation you will not make childless\lebnote{According to the reading tradition (\textit{Qere})}any longer,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“And I will not let you hear any longer the insult of the nations, and the disgrace of the peoples you will not bear any longer; you yourself will not cause your nation to stumble,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.’ ”}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}the house of Israel, they were dwelling on their soil, and they defiled it with their way and with their deeds; like the uncleanness of menstruation was their way \textit{before me}.\lebnote{Literally “to the face of me”}}%
\verse{And I poured out my rage on them for the blood that they poured on the land and for their idols with which they defiled it.}%
\verse{And I scattered them among the nations, and I dispersed them in the countries according to their way, and according to their deeds I judged them.}%
\verse{And they came\lebnote{Or “went”}into the nations \textit{to which they went},\lebnote{Literally “who/which they went there”}and they profaned \textit{my holy name}\lebnote{Literally “the name of my holiness”}when they said to them, ‘These are the people of Adonai, and from his land they went out.’}%
\verse{And I was concerned for \textit{my holy name},\lebnote{Literally “the name of my holiness”}which, the house of Israel defiled among the nations \textit{to which they went}.\lebnote{Literally “which they went there”}}%
\verse{“Therefore thus say to the house of Israel, ‘Thus says the Lord Adonai: “\textit{Not for your sake} \lebnote{Literally “not for the sake of/on behalf of you”}am I about to act, house of Israel, \textit{but}\lebnote{Literally “for it”}\textit{for my holy name},\lebnote{Literally “for the name of holiness my”}which you defiled among the nations \textit{to which you went}.\lebnote{Literally “which you went there”}}%
\verse{And I will consecrate my great name, which was profaned among the nations and which you have profaned in the midst of them, and the nations will know that I am Adonai!” ’ a declaration of the Lord Adonai, \textit{when I show myself holy}\lebnote{Literally “at/in showing/to show self holy my”}\textit{before}\lebnote{Literally “to their eyes”}their eyes.}%
\verse{“ ‘And I will take you from the nations, and I will gather you from all of the lands, and I will bring you to your land.}%
\verse{And I will sprinkle on you pure water, and you will be clean from all of your uncleanness, and I will cleanse you from all of your idols.}%
\verse{And I will give a new heart to you, and a new spirit I will give\lebnote{Or “put/place”}into your inner parts, and I will remove the\lebnote{Or “your”}heart of stone from your flesh, and I will give to you a heart of flesh.}%
\verse{And I will give\lebnote{Or “place/put”}my spirit into your inner parts, and I will make it so that you will go\lebnote{Or “walk”}in my rules, and my regulations you will remember, and you will do them.}%
\verse{And you will dwell in the land that I gave to your ancestors,\lebnote{Or “fathers”}and you will be to me as a people, and I will be to you as\lebnote{Hebrew “to”}God.}%
\verse{And I will save you from all of your uncleanness, and I will call to the grain, and I will cause it to increase, and I will not bring famine upon you.}%
\verse{And I will cause the fruit of the tree and the crop of the field to increase, so that you will not suffer again the disgrace of famine among the nations.}%
\verse{And you will remember your evil ways and your deeds that were not good, and you will loathe \textit{yourself}\lebnote{Literally “for face your”}over\lebnote{Or “because of”}your iniquities and over your detestable things.}%
\verse{But not for your sake am I acting,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai. “Let it be known to you, be ashamed, and be put to shame because of your ways, house of Israel.}%
\verse{“Thus says the Lord Adonai: ‘On the day when I cleanse you from all of your iniquities, I will cause the cities to be inhabited, and the ruins will be rebuilt.}%
\verse{And the land that was desolate will be cultivated in \textit{the very place that}\lebnote{Literally “in the place that/which/where”}it was desolate before the eyes of all of the persons crossing over.}%
\verse{And they will say, “\textit{This desolate land}\lebnote{Literally “the land this the being desolate”}has become like the garden of Eden, and the wasted and desolate and destroyed cities, now being refortified, are inhabited.”}%
\verse{And the nations who are left all around, you will know that I, Adonai, I built \textit{that which was destroyed};\lebnote{Literally “the destroyed”}I planted the desolate land; I, Adonai, I have spoken, and I will act.’}%
\verse{“Thus says the Lord Adonai: ‘Again this time I will let \textit{myself be inquired of}\lebnote{Literally “I will be/let me be searched”}by the house of Israel, to do something for them; I will cause them to increase \textit{their population like a flock}.\lebnote{Literally “them as the flock humankind”}}%
\verse{Like the flock of the sanctuary, like the flock of Jerusalem at its festival, so the desolate cities will be filled with flocks\lebnote{Hebrew “flock”}of people; and they will know that I am Adonai.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 37
\verseWithHeading{A Valley of Dried Bones; A Renewed Nation of Israel}{The hand of Adonai was upon me, and he brought me by the Spirit of Adonai, and he \textit{let me rest}\lebnote{Or “placed me”; literally “set me”}in the midst of the valley, and it was full of bones.}%
\verse{And he led me \textit{all around}\lebnote{Literally “all around, all around”}over them, and look, very many on the surface of the valley, and look, very dry.}%
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}can these bones live?” And I said, “Lord Adonai, you know.”}%
\verse{And he said to me, “Prophesy to these bones, and you must say to them, to the dry bones, ‘Hear the word of Adonai!}%
\verse{Thus says the Lord Adonai to these bones: “Look! I am bringing into you breath, and you will live!}%
\verse{And I will lay on you sinews, and I will let flesh come upon you, and I will cover you over with skin, and I will put breath into you, and you will live, and you will know that I am Adonai.” ’ ”}%
\verse{And I prophesied \textit{just as}\lebnote{Literally “as that”}I had been commanded, and there was a sound at my prophesying, and look! A rattling, and they came together—the bones! Bone to its bone!}%
\verse{And I looked, and indeed, sinews were on them, and flesh went up,\lebnote{Or “appeared”}and skin covered over\lebnote{Or “on”}them \textit{upward},\lebnote{Literally “from to above”}but breath was not in them.}%
\verse{And he said to me, “Prophesy to the breath; prophesy, son of man,\lebnote{Or “mortal,” or “son of humankind”}and you must say to the breath, ‘Thus says the Lord Adonai \textit{from the four winds},\lebnote{Literally “from four winds”}“Come, O spirit and breath, on these dead ones, so that they may live!” ’ ”}%
\verse{And I prophesied \textit{as}\lebnote{Literally “like what”}he commanded me, and the breath came into them, and they became alive, and they stood on their feet, a very, very large group.}%
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}these bones are all of the house of Israel; look! they are saying, ‘Our bones are dried up, and our hope is destroyed; \textit{we are cut off as far as we are concerned}.’\lebnote{Literally “we are cut in two for us”}}%
\verse{Therefore prophesy, and you must say to them, ‘Thus says the Lord Adonai: “Look! I am opening your graves, and I will bring you up from your graves, my people, and I will bring you to the land of Israel!}%
\verse{And you will know that I am Adonai when I open your graves when I bring you up from your graves, my people!}%
\verse{And I will put my breath into you so that you may live, and I will cause you to rest on your soil, and you will know that I, Adonai, I have spoken, and I will act!” ’ \textit{declares}\lebnote{Literally “declaration of”}Adonai.”}%
\verse{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}take for yourself \textit{a piece of wood},\lebnote{Literally “a piece of wood one”}and write on it, ‘For Judah and for the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}his associates,’ and take another piece of wood, and write on it, ‘For Joseph, the piece of wood of\lebnote{Or “for”}Ephraim and all of the house of Israel his associates.’}%
\verse{And join them one to the other with respect to you as one piece of wood, so that they may become one in your hand.}%
\verse{\textit{When}\lebnote{Literally “as that”}\textit{your people}\lebnote{Literally “the children of your people”}say to you, \textit{saying},\lebnote{Literally “to say”}‘Will you not inform us as to what these actions mean for you?’}%
\verse{Then speak to them, ‘Thus says the Lord Adonai: “Look! I am taking the piece of wood for Joseph that is in the hand of Ephraim and the tribes of Israel its associates, and I will put them on it, the piece of wood of Judah, and I will make them into one piece of wood, so that they be one in my hand.’ ”}%
\verse{And the pieces of wood on which you wrote will be in your hand before their eyes.}%
\verse{And speak to them, ‘Thus says the Lord Adonai: “Look! I am taking the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}from among \textit{the nations to which they went},\lebnote{Literally “which they went there”}and I will gather them from \textit{everywhere},\lebnote{Literally “all around”}and I will bring them to their own soil.}%
\verse{And I will make them into one nation in the land on the mountains of Israel, and \textit{they will all have one king as their king},\lebnote{Literally “and king one he will be for all of them to king”}and they\lebnote{According to the reading tradition (\textit{Qere})}will not again be two nations and will not again divide into two kingdoms again.}%
\verse{And they will not defile themselves again with their idols and with vile idols and with all of their transgressions, and I will save them from all of their apostasy \textit{by which they sinned},\lebnote{Or “which they sinned by them”}and I will cleanse them and they will be for me as a people and I, I will be for them as God.}%
\verse{“ ‘ “And my servant David will be king over them, and one shepherd will be for all of them, and in my regulations they will go, and my statutes they will observe, and they will do them.}%
\verse{And they will dwell on the land that I gave to my servant, to Jacob, \textit{in which your ancestors dwelled},\lebnote{Literally “which they dwelt in it your fathers”}and they will dwell on it, they and their children and the children of their children \textit{forever},\lebnote{Literally “until eternity” or “until unlimited time”}and my servant David will be a leader for them \textit{forever}.\lebnote{Literally “to eternity” or “to endless ages”}}%
\verse{And I will make with them a covenant of peace; \textit{an everlasting covenant}\lebnote{Literally “a covenant of eternity”}it will be with them. And I will establish them, and I will cause them to increase, and I will put my sanctuary in the midst of them \textit{forever}.\lebnote{Literally “to eternity” or “to unlimited ages”}}%
\verse{And my tabernacle will be with them, and I will be for them as God, and they will be to me as a nation.}%
\verse{And the nations will know that I, Adonai, am consecrating Israel when my sanctuary is in the midst of them \textit{forever}.” ’ ”\lebnote{Literally “to eternity” or “to unlimited ages”}}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 38
\verseWithHeading{God’s Message to Gog}{And the word of Adonai \textit{came}\lebnote{Literally “was”}to me, \textit{saying},\lebnote{Literally “to say”}}%
\verse{“Son of man,\lebnote{Or “mortal,” or “son of humankind”}set your face toward Gog of the land of Magog, the head\lebnote{Or “chief”}leader of Meshech and Tubal, and prophesy against him.}%
\verse{And you must say, ‘Thus says the Lord Adonai: “Look! I am against you, Gog, the head\lebnote{Or “chief”}leader of Meshach and Tubal,}%
\verse{and I will turn you around, and I will place\lebnote{Or “give”}hooks in your cheeks, and I will bring you out and all of your horses and horsemen \textit{fully armed},\lebnote{Literally “clothed of fullness”}all of them, a great crowd, holding a shield,\lebnote{Or “buckler”}and small shield, and holding\lebnote{Or “wielding”}swords, all of them.}%
\verse{Persia, Cush, and Put are with them, all of them, with a small shield and helmet.}%
\verse{Gomer and all of its troops, Beth Togarmah, the remote areas of the north, and with all of its troops and many peoples with them.}%
\verse{Be ready, and prepare \textit{yourselves},\lebnote{Literally “for you”}you and all of your assembly,\lebnote{Or “battalions” or “hoards”}the assembling\lebnote{Or “assemblies”}around you, and you must be\lebnote{Or “serve”}for them as a guard.}%
\verse{After many days you will be mustered; \textit{in the last years}\lebnote{Or “final years”; literally “in the last of the years”}you will come to a restored land from the sword, gathered from many peoples on the mountains of Israel which were as \textit{permanent ruins}\lebnote{Literally “ruins continually/permanently”}but from peoples it was brought out, and they will dwell \textit{in safety},\lebnote{Literally “with confidence/assurance/trust”}all of them.}%
\verse{And you will advance like a storm; you will come, and you will be like a cloud \textit{covering}\lebnote{Literally “to cover”}the land, and all of your troops and many nations along with you.”}%
\verse{“ ‘Thus says the Lord Adonai: “\textit{And then}\lebnote{Literally “and it will be”}on that day, things\lebnote{Or “thoughts”}will come up on\lebnote{Or “in”}your mind, and you will devise evil plans.}%
\verse{And you will say, ‘I will go up against a land of open country; I will come\lebnote{Or “go”}to the people being at rest \textit{in safety},\lebnote{Literally “with confidence”}all of them dwelling without a wall\lebnote{Or “walls”}and crossbars\lebnote{Hebrew “crossbar”}and \textit{without doors},\lebnote{Literally “and doors there are not for them”}}%
\verse{to\lebnote{Or “for”}loot loot,\lebnote{Or “war-booty”; this word is chosen because of the alliteration in Hebrew}and to plunder\lebnote{Or “seize”; this word is chosen because of the alliteration in Hebrew}plunder, \textit{to assail}\lebnote{Literally “to bring back your hand against”; NJPS, “to turn your hand against”}inhabited ruins and a people gathered together from various peoples and who are acquiring livestock and goods and dwelling at the center of the world.}%
\verse{Sheba and Dedan and traders of Tarshish and all its strong lions,\lebnote{Or “all of its violent leaders”}they will ask you, ‘To\lebnote{Or “for”}loot loot\lebnote{Or “war-booty” this word is chosen because of the alliteration in Hebrew}are you coming? To plunder\lebnote{Or “seize”; this word is chosen because of the alliteration in Hebrew}plunder have you summoned your assembly, to take away silver and gold, to take livestock and goods, to loot great amounts\lebnote{Hebrew “amount”}of loot?” ’}%
\verse{“Therefore prophesy, son of man,\lebnote{Or “mortal,” or “son of humankind”}and you must say to Gog, ‘Thus says the Lord Adonai: “Will you not realize\lebnote{LXX reads, “stir up/rouse yourself”}on that day when my people Israel are dwelling \textit{in safety},\lebnote{Literally “with confidence/trust”}}%
\verse{and so you will come from your place, from the remote areas of the north, you and many people with you, \textit{horsemen}\lebnote{Literally “riders of horsemen”}all of them, a great crowd and a vast army,}%
\verse{and you will advance against my people Israel like a cloud \textit{covering}\lebnote{Literally “to cover”}the land; it will be\lebnote{Or “occur”}in the last days, and I will bring you against my land, so that the nations know me, \textit{when I show myself holy through you} before their eyes, O Gog!” ’}%
\verse{“Thus says the Lord Adonai, ‘Are you he of whom I spoke in former days by the hand of my servants the prophets of Israel who were prophesying in those days for years that I would bring you against them?}%
\verse{And so then in that day, on the day of the coming of God against the land of Israel,’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, ‘my rage will come up in my anger.}%
\verse{And in my passion, in the fire of my wrath, I spoke\lebnote{Or “I asserted”}that \textit{certainly}\lebnote{Literally “if not”}on that day a great earthquake will be on the land of Israel.}%
\verse{And the fish of the sea and the birds\lebnote{Hebrew “bird”}of the heaven\lebnote{Or “sky”}and the animals\lebnote{Hebrew “animal”}of the field and all of the creeping things that creep on the earth and all of the humans\lebnote{Or “every human being”}who are on the surface of the earth will shake \textit{at my presence};\lebnote{Literally “from the face of me”}and the mountains will be demolished, and the steep mountain sides will fall, and every wall on\lebnote{Or “to”}the earth will fall.}%
\verse{And I will call against him\lebnote{Or “summon against him” (cf. NJPS)}in all of my mountains a sword,’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, ‘And the sword of each person will be against his brother.}%
\verse{And I will execute justice with\lebnote{Or “with respect to him”}him with a plague and with blood and \textit{torrents of rain},\lebnote{Literally “rain flooding/pouring down”}and hailstones; fire and sulfur I will cause to fall on him and on his troops and on many peoples who are with him.}%
\verse{And so I will exalt myself, and I will show myself holy, and I will make myself known before the eyes of many nations, and they will know that I am Adonai.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 39
\verseWithHeading{God’s Destruction of Gog and His Hordes}{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}prophesy against Gog, and you must say, ‘Thus says the Lord Adonai: “Look! I am against you Gog, the head\lebnote{Or “chief”}leader of Meshech and Tubal,}%
\verse{and I will turn you around, and I will drag you along, and I will bring you up from the remote areas of the north, and I will bring you against the mountains of Israel,}%
\verse{and I will strike your bow from \textit{your left hand},\lebnote{Literally “the hand of left your”}and your arrows from \textit{your right hand}\lebnote{Literally “the hand of right your”}I will cause to fall.}%
\verse{On the mountains of Israel you will fall, you and all of your troops, and the peoples who are with you, to birds\lebnote{Hebrew “bird”}of prey, birds of every wing,\lebnote{Or “kind”}and animals of the field I will give you as food.}%
\verse{On the surface of the field you will fall, for \textit{I myself}\lebnote{Literally “I, I”}have spoken,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“And I will send fire against Magog and among the people inhabiting the coastlands \textit{in safety},\lebnote{Literally “with confidence/trust”}and they will know that I am Adonai.}%
\verse{And \textit{my holy name}\lebnote{Literally “the name of my holiness”}I will make known in the midst of my people Israel, and I will not let the name of my holiness be profaned anymore, and the nations will know that I am Adonai, the holy one in Israel.}%
\verse{Look! it is coming, and it will happen,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai. “It is the day \textit{about which I have spoken}.\lebnote{Literally “which I have spoken of”}}%
\verse{“ʻ“And the inhabitants of the towns\lebnote{Or “cities”}of Israel will go out, and they will set ablaze, and they will kindle a fire with a weapon\lebnote{Or “weapons”}and a small shield and a shield,\lebnote{Or “shields and bucklers”; cf. NJPS}with a bow and with arrows, and with \textit{hand clubs}\lebnote{Literally “with club of hand”}and with spears,\lebnote{Hebrew “spear”}and so they will light a fire with them, \textit{a fire lasting seven years}.\lebnote{Literally “a fire of seven years”}}%
\verse{And they will not carry trees from the field, and they will not chop wood from the forests, for with the weapons\lebnote{Hebrew “weapon”}they will light a fire; and they will plunder those plundering them, and they will plunder those plundering them,” ’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“ ‘ “\textit{And then}\lebnote{Literally “and it will happen”}on that day I will give to Gog \textit{a grave there}\lebnote{Literally “a place there of a grave”}in Israel, The Valley of the Travelers, east of the sea,\lebnote{That is, the Dead Sea}and it \textit{will block}\lebnote{Literally “blocking”}the travelers and Gog and all of his hordes\lebnote{Hebrew “horde”}they will bury there, and they will call it the Valley of \textit{Hamon-Gog}.\lebnote{Literally “Gog’s Multitude”}}%
\verse{And the house of Israel will bury them for seven months to cleanse the land.}%
\verse{All of the people of the land will bury them, and it will be an honor for them on the day when I appear,” ’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{ʻ“And they will set apart \textit{men to continually}\lebnote{Literally “men of continuity/men who regularly”}go through the land burying the travelers,\lebnote{In a negative sense = “invaders”}the ones left over on the surface of the land to cleanse it; \textit{for the whole of}\lebnote{Literally “from the end of”}seven months they will explore it.}%
\verse{And the ones going through, they will go through within the land, and if anyone sees the bones of a person, then he will build beside it a sign to remain until the buriers have buried him in the Valley of Hamon-Gog.\lebnote{Or “Gog’s multitude”}}%
\verse{And furthermore the name of the city there is Hamonah,\lebnote{That is, “Multitude”}and thus they will cleanse the land.” ’}%
\verse{“And you, son of man,\lebnote{Or “mortal,” or “son of humankind”}thus says the Lord Adonai: ‘Say to \textit{all kinds of birds and animals of the field},\lebnote{Literally “to bird of every wind and to every animal of the field”}“Be gathered, come be gathered \textit{from everywhere}\lebnote{Literally “all around”}to my sacrifice that I am sacrificing for you, a great sacrifice on the mountains of Israel, and you will eat flesh, and you will drink blood.}%
\verse{Flesh of warriors you will eat, and the blood of leaders of the land you will drink; rams, young rams and goats, bulls, and the fattened animals of Bashan—all of them!}%
\verse{And you will eat fat \textit{until you are satiated},\lebnote{Literally “satiation/gluttony”}and you will drink blood \textit{until you are drunk}\lebnote{Literally “to drunkenness”}from my sacrifice that I sacrifice for you,}%
\verse{and you will be satisfied at my table with horse and horsemen and warriors and every man of war,” ’ \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“And I will display my glory among the nations, and all of the nations will see my judgment that I have executed and my hand that I have laid upon them.}%
\verse{“And the house of Israel will know that I am Adonai their God from that day and beyond,}%
\verse{and the nations will know that because of their guilt the house of Israel went into exile, \textit{because}\lebnote{Literally “because that”}they acted unfaithfully against me, and I hid my face from them, and I gave them into the hand of their foes, and they fell by the sword, all of them.}%
\verse{According to their uncleanness and according to their transgression I dealt with them, and I hid my face from them.}%
\verse{Therefore thus says the Lord Adonai: “Now I will restore the fortunes of Jacob, and I will have compassion on all of the house of Israel, and I will be jealous\lebnote{Or “zealous”}for \textit{my holy name}.\lebnote{Literally “for the name of holiness my”}}%
\verse{And they will forget their disgrace and all of their infidelity that they displayed against me when they dwelt on their soil \textit{in safety}\lebnote{Literally “with confidence”}\textit{with nobody making them afraid}.\lebnote{Literally “and not was somebody making afraid”}}%
\verse{When I restore them from the nations and I gather them from the countries of their enemies, then I will show myself holy through them before the eyes of many nations.}%
\verse{And they will know that I am Adonai their God, because of when I deported them into the nations, and I reassembled them to their soil, and \textit{I will not let any of them remain there any longer}.\lebnote{Literally “not I will leave over/let remain again/still from them there”}}%
\verse{And I will not hide my face again from them when I pour out my Spirit over\lebnote{Or “upon”}the house of Israel,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 40
\verseWithHeading{A Vision of the New Temple}{In the twenty-fifth year of our exile, at the beginning of the year, on the tenth day of the month, in the fourteenth year after the city was destroyed, in this day exactly, the hand of Adonai was on me, and he brought me there}%
\verse{in visions from God. He brought me to the land of Israel and put me on a very high mountain, and on it was something like a structure of a city to the south.}%
\verse{And he brought me there, and look, there was a man whose\lebnote{Hebrew “his”}appearance was like the appearance of bronze, and a cord of linen was in his hand and a reed for measurement; he was standing in the gate.}%
\verse{And the man spoke to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}look with your eyes and listen with your ears and apply\lebnote{Or “set/put”}your heart to all that I am showing you, for you were brought here in order to show you this; tell all that you are seeing to the house of Israel.”}%
\verseWithHeading{The Eastern Gate}{And there was a wall on the outside of the temple \textit{all the way around it},\lebnote{Literally “all around, all around”}and in the hand of the man the reed\lebnote{Or “rod”}for measurement was six long cubits, \textit{according to}\lebnote{Literally “with/in”}the cubit\lebnote{That is, about 18 inches}and a handbreadth,\lebnote{A handbreadth wasabout 3 inches; the reed/rod was 10.5 feet long}and he measured the width of the outer wall as one reed,\lebnote{That is, 10.5 feet}and the height as one reed.\lebnote{That is, about 10.5 feet}}%
\verse{And then he went toward\lebnote{Or “to”}the gate whose face was \textit{to the east}.\lebnote{Literally “to the way toward the east”}And he went up by\lebnote{Or “with”}its steps, and he measured the threshold of the gate, one reed\lebnote{That is, about 10.5 feet}wide.}%
\verse{And the alcove\lebnote{Or “niches” in the temple tower}was one reed\lebnote{That is, about 10.5 feet}long and one reed\lebnote{That is, about 10.5 feet}wide,\lebnote{That is, 10.5 x 10.5 feet}and between the alcoves was five cubits,\lebnote{That is, about 8.75 feet}and the threshold of the gate along the side of the portico of the gate \textit{on the inside}\lebnote{Literally “from the house”}was one reed.\lebnote{That is, about 10.5 feet}}%
\verse{And then he measured the portico of the gate \textit{on the inside}\lebnote{Literally “from the house”}as one reed.\lebnote{That is, about 10.5 feet.}}%
\verse{And he measured the portico of the gate as eight cubits,\lebnote{That is, about 14 feet}and its pilaster was two cubits\lebnote{That is, 3.5 feet}thick, and he measured the portico of the gate \textit{on the inside}.\lebnote{Literally “from the house”}}%
\verse{And the alcoves\lebnote{Hebrew “alcove”}of the gate toward the way\lebnote{Or “direction”}\textit{eastward}\lebnote{Literally “of the east”}were \textit{three on each side};\lebnote{Literally “was three from here and three from here”}\textit{the same measurements applied to all three of them};\lebnote{Literally “measurement one was to the three of them”}and \textit{the same measurement applied to the pilaster}\lebnote{Literally “measurement one was to the pilaster”}\textit{on each side}.\lebnote{Literally “from here and from there”}}%
\verse{And he measured the width of the doorway of the gate as ten cubits\lebnote{That is, 17.5 feet}and the length of the gateway\lebnote{Hebrew “gate”}was thirteen cubits.\lebnote{That is, 22.75 feet; or the inside width of the gateway was 22.75 feet}}%
\verse{And a wall\lebnote{Or “guard wall”; others prefer “curb” (NLT)}\textit{was before}\lebnote{Literally “was to the face of”}the alcoves; \textit{one cubit\lebnote{That is, about 18 inches}on either side}\lebnote{Literally “cubit one and cubit one”; or 21 inches}was the wall from here.\lebnote{Or “from the front”}And the alcove \textit{was six cubits\lebnote{That is, 10.5 feet}on each side}.\lebnote{Literally “was six cubits from here and six cubits from here”}}%
\verse{And he measured the gate from the top slab of the alcove to its opposite top slab as twenty-five cubits in width, \textit{from one entrance to the other one opposite it}.\lebnote{Literally “doorway opposite doorway”}}%
\verse{And he made the pilasters sixty cubits,\lebnote{That is, 105 feet (problematic in this context)}and to the pilaster of the courtyard \textit{all the way around the gate}.\lebnote{Literally “the gate all around, all around”}}%
\verse{\textit{And from the front of}\lebnote{Literally “and at the face of”}the gate at the entrance to \textit{the front of}\lebnote{Literally “the face of”}the portico of the inner gate was fifty cubits.\lebnote{That is, 87.5 feet}}%
\verse{And there were narrow windows for the alcoves and for their pilasters to the inside of the gate \textit{all the way around}\lebnote{Literally “all around, all around”}it. And likewise with respect to the porticos and windows were \textit{all the way around}\lebnote{Literally “all around, all around”}to the inside, and on a pilaster there were images of palm trees.\lebnote{Hebrew “tree”}}%
\verseWithHeading{The Outer Courtyard}{And he brought me to the outer courtyard, and there were chambers and a pavement made for the courtyard all around it, and thirty chambers were facing to the pavement.}%
\verse{And the pavement \textit{flanked}\lebnote{Literally “was to the side of”}the side of the gates \textit{and all along the gates};\lebnote{Literally “to along side the length of the gates”}this describes the lower pavement.}%
\verse{And he measured the width from \textit{the front of}\lebnote{Literally “to the face of”}the lower gate to \textit{the front of}\lebnote{Literally “to the face of”}the outside of the inner courtyard as a hundred cubits\lebnote{That is, 175 feet}to the east and to the north.}%
\verseWithHeading{Measurement of the North Gate}{And as for the gate that had its face \textit{toward the north}\lebnote{Literally “to the way of the north”}of the outer courtyard, he measured its length and its width.}%
\verse{And its alcoves were \textit{three on each side},\lebnote{Literally “three from here and three from here”}and its pilasters and its porticos \textit{had the same measurement}\lebnote{Literally “it was like the measurement of”}as the first gate: its length was fifty cubits,\lebnote{That is, 87.5 feet}and its width was twenty-five \textit{cubits}.\lebnote{Literally “in/by the cubit”; 43.5 feet}}%
\verse{And its windows and its porticos and its palm tree images were like the measurement of the gate that was \textit{facing toward the east};\lebnote{Literally “its face to the way of the east”}and \textit{it had}\lebnote{Literally “and with steps”}seven steps that go up it, and there were porticos \textit{before them}.\lebnote{Literally “to the face of them”}}%
\verse{And a gate led to the inner courtyard, opposite the gate to the north and to the east, and he measured from gate to gate as a hundred cubits.\lebnote{That is, 175 feet}}%
\verseWithHeading{The Gate to the South}{And then he took me \textit{toward the south},\lebnote{Literally “to the way of the south”}and look, a gate \textit{toward the south},\lebnote{Literally “to the way of the south”}and he measured its pilasters and its porticos; they had measurements just \textit{like the others}.\lebnote{Literally “like the measurements the these”}}%
\verse{And there were for it windows and for its porticos \textit{all the way around it}\lebnote{Literally “all around, all around”}like these windows; fifty cubits\lebnote{That is, 87.5 feet}long and twenty-five cubits\lebnote{That is, 43.75 feet}wide.}%
\verse{And seven steps were going up to it and its porticos \textit{before them}.\lebnote{Literally “to the face of them”}And \textit{it had palm tree images}\lebnote{Literally “and palm tree images were for it”}\textit{all along}\lebnote{Literally “from here and one from here to”}its pilasters.}%
\verse{And there was to the way of the south\lebnote{That is, “to the south”}a gate for the inner courtyard, and he measured from gate to the gate on the way of the south, a hundred cubits.\lebnote{That is, 175 feet}}%
\verseWithHeading{Gates to the Inner Courtyard}{And he brought me to the inner courtyard through\lebnote{Or “at”}the gate of the south, and he measured the south gate, and it had measurements like \textit{the others}.\lebnote{Literally “the measurements the these”}}%
\verse{And its alcoves and its pilasters and its porticos were just like these measurements. And it had windows for it and for its porticos\lebnote{Hebrew “portico”}\textit{all the way around},\lebnote{Literally “all around, all around”}fifty cubits\lebnote{Hebrew “cubit”}along\lebnote{Literally “long”}its length, and twenty-five cubits\lebnote{That is, 43.75 feet}wide.}%
\verse{And there were porticos \textit{all the way around}\lebnote{Literally “all around, all around”}its length, twenty-five cubits;\lebnote{That is, 43.75 feet}and its width was five cubits.\lebnote{That is, 8.75 feet, which differs from 40:9; the LXX and some Greek versions lack this verse altogether}}%
\verse{And its porticos were to the outer courtyard, and palm tree images were on its pilasters, and eight steps were for its stairs.}%
\verse{And he brought me to the inner courtyard by way of the east gate, and he measured the gate as just like \textit{the other measurements}.\lebnote{Literally “the measurements the these”}}%
\verse{And its alcoves and its pilasters and its porticos were also like these measurements, and windows were for it and for its porticos all around;\lebnote{Literally “all around, all around”}its length was fifty cubits,\lebnote{That is, 87.5 feet}and it was twenty-five cubits\lebnote{That is, 43.75 feet}wide.}%
\verse{And its porticos were toward the outer courtyard, and palm tree images were on its pilasters \textit{one each side},\lebnote{Literally “from here and from here”}and \textit{eight steps served as it stairs}.\lebnote{Literally “and eight steps its stairs”}}%
\verse{And he brought me to the gate of the north, and he measured it; and he measured \textit{the same measurements as these others},\lebnote{Literally “like the measurements the these”}}%
\verse{its alcoves, its pilasters and porticos, \textit{and its windows}\lebnote{Literally “its windows were for it”}\textit{all the way around},\lebnote{Literally “all around, all around”}and a length of fifty cubits\lebnote{That is, 87.5 feet}and twenty-five cubits wide.\lebnote{That is, 43.75 feet}}%
\verse{And its pilasters \textit{faced the outer courtyard},\lebnote{Literally “were to the courtyard the outer”}and \textit{it had palm tree images on its pilasters}\lebnote{Literally “and palm tree images were for its pilaster”}\textit{on each side};\lebnote{Literally “from here and from here”}and eight steps served as its stairs.}%
\verseWithHeading{Rooms for Sacrificial Preparations}{And a chamber with its doorway was in\lebnote{Or “set among”}the pilasters at the gates,\lebnote{Or “gate”}and there they rinsed off\lebnote{Or “washed”}the burnt offering.}%
\verse{And in the portico of the gate \textit{were two tables on each side}\lebnote{Literally “were two tables from here and two tables from here”}to slaughter the burnt offering on them and the sin offering and the guilt offering.}%
\verse{And on the outer side as one goes up to the doorway of the gate to the north were two tables, and on the other side, which is toward the portico of the gate, were two tables.}%
\verse{\textit{On each side of the gate were four tables},\lebnote{Literally “four tables from here and four tables from here”}eight tables in all, and on them they slaughtered sacrifices.}%
\verse{And four tables were for the burnt offering, and made of dressed stones, one cubit and a half\lebnote{That is, 31.5 inches}long and one cubit and a half wide and one cubit\lebnote{That is, 21 inches}high \textit{was their measurements},\lebnote{Literally “wasto them}and they placed the objects\lebnote{Or “instruments/utensils”}with which they slaughtered the burnt offering on them and the sacrifice.}%
\verse{And there were double-pronged hooks, one handbreadth in width, put in place in the house\lebnote{Or “inside”}all around, and on the tables was the flesh of the offering.}%
\verseWithHeading{Rooms for Priests and Other Worship Leaders}{And on the outside of the inner gate, there were chambers for\lebnote{Hebrew “of”}singers in the inner courtyard, which was to the side of the north gate, and their faces\lebnote{That is, of the chambers}\textit{were to the south}\lebnote{Literally “were to the way of the south”}with respect to one, and on the side of the east gate one facing to the north.}%
\verse{And he said to me, “This chamber \textit{with its face toward the south}\lebnote{Literally “which its face is to the way of the south”}is for the priests who are taking care of the responsibility of the temple.}%
\verse{And the chamber \textit{with its face to the north}\lebnote{Literally “with its face is to the way of the north”}is for the priests who are taking care of the responsibility of the altar. They are the descendants\lebnote{Or “sons”}of Zadok, the ones who approach from among the descendants\lebnote{Or “sons”}of Levi to Adonai to serve him.”}%
\verseWithHeading{Inner Courtyard and Temple Measurements}{And he measured the courtyard as to its length, a hundred cubits,\lebnote{That is, 175 feet}and a hundred cubits\lebnote{That is, 175 feet}wide, squared, and the altar \textit{is in front of}\lebnote{Literally “before face of”}the temple.}%
\verse{And he brought me to the portico of the temple, and he measured the pilaster of the portico, \textit{five cubits on each side},\lebnote{That is, 8.75 feet; literally “five cubits from here and five cubits from here”}and the width of the gate \textit{was three cubits on each side}.\lebnote{Literally “was three cubits from here and three cubits from here”}}%
\verse{The length of the portico was twenty cubits\lebnote{That is, 35 feet}and its width eleven cubits,\lebnote{That is, 19.25 feet; LXX reads 12 cubits or 21 feet}and with ten steps they went up to it, and the \textit{pilasters had pillars},\lebnote{Literally “and pillars were to the pilasters”}\textit{one on each side}.\lebnote{Literally “one from here and one from here”}}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 41
\verseWithHeading{Sanctuary and Temple}{And he brought me to the temple sanctuary,\lebnote{Or “main room” (of the temple of Jerusalem); “great hall” (JPS); “outer sanctuary” (NIV); word can refer to surrounding courts and halls; HALOT 245.}and he measured the pilasters, \textit{six cubits\lebnote{That is, 10.5 feet}wide on each side};\lebnote{Literally “six cubits wide from here and six cubits wide from here”}this was the width of the tent.\lebnote{Or “of the jambs”}}%
\verse{And the width of the doorway was ten cubits,\lebnote{That is, 17.5 feet}and the sidewall of the doorway \textit{was five cubits on each side},\lebnote{Literally “was five cubits from here and five cubits from here”; five cubits = 8.75 feet}and he measured its\lebnote{That is, the outer sanctuary}length as forty cubits,\lebnote{That is, 70 feet}and its width was twenty cubits.\lebnote{That is, 35 feet}}%
\verse{And he went into the inner room, and he measured the pilaster of the doorway as two cubits\lebnote{That is, 3.5 feet}and the doorway as six cubits\lebnote{That is, 10.5 feet}and the width of the doorway seven cubits.\lebnote{That is, 12.25 feet}}%
\verse{And he measured its length as twenty cubits\lebnote{That is, 35 feet}and its width as twenty cubits to the front of the temple, and he said to me, “This is \textit{the most holy place}.\lebnote{Literally “the holy of the holies”}}%
\verse{And he measured the wall of the temple as six cubits,\lebnote{That is, 10.5 feet}and the width of the side room as four cubits\lebnote{That is, 7 feet}\textit{all along the outside wall for the temple all around the wall}.\lebnote{Literally “all around, all around for the temple all around”}}%
\verse{And the side rooms were \textit{side by side}\lebnote{Literally “were side room to side room”}in three stories and a total of thirty \textit{rooms},\lebnote{Literally “times”}and there were offsets\lebnote{Meaning of word is uncertain}in the wall, which was for the temple for the side rooms, \textit{all the way around}\lebnote{Literally “all around, all around”}to be supports, and so they were not supports extending into the wall of the temple.}%
\verse{And each level widened, and it went around\lebnote{Or “was surrounded”}\textit{upward}\lebnote{Literally “to above, to above”}to the side rooms for the structure that surrounds\lebnote{Or “goes around”}the temple \textit{upward}\lebnote{Literally “to above, to above”}\textit{all the way around},\lebnote{Literally “all around, all around”}therefore the width increased to the temple \textit{upward},\lebnote{Literally “to above”}and thus the lower level goes up to the upper level by means of the middle story.}%
\verse{And I saw for the temple a platform \textit{all the way around}\lebnote{Literally “all around, all around”}the foundations of the side rooms; it was the length of a full reed, six cubits long.\lebnote{That is, 10.5 feet}}%
\verse{The width of the outside wall, which was for the side room to the outside, was five cubits,\lebnote{That is, 8.75 feet}and a space that was set between the side rooms, which was for the temple}%
\verse{and between the chambers was the width of twenty cubits\lebnote{That is, 35 feet}all around the temple, \textit{all the way around}\lebnote{Literally “all around, all around”}it.}%
\verse{And the doorway of the side room \textit{faced}\lebnote{Literally “was to”}the open area; one doorway \textit{faced}\lebnote{Literally “was to”}\textit{the north},\lebnote{Literally “the way of the north”}and one doorway was to the south, and the width of the place of the open area was five cubits\lebnote{That is, 8.75 feet}all around.\lebnote{Literally “all around, all around”}}%
\verse{And the building that \textit{faced}\lebnote{Literally “was to the front of”}the courtyard \textit{was toward the west},\lebnote{Literally “wasto the side of the way of the sea/west”}and its width was seventy cubits,\lebnote{That is, 122.5 feet}and the wall of the building was five cubits\lebnote{That is, 8.25 feet}wide \textit{all the way around},\lebnote{Literally “all around, all around”}and its length was ninety cubits.\lebnote{That is, 157.5 feet}}%
\verse{And he measured the temple, and its length was a hundred cubits,\lebnote{That is, 175 feet}and the courtyard and the building and its walls, their length was a hundred cubits.\lebnote{That is, 175 feet}}%
\verse{And the width of the front of the temple and the courtyard to the east was a hundred cubits.\lebnote{That is, 175 feet}}%
\verse{And he measured the length of the building \textit{facing the courtyard}\lebnote{Literally “to the front of the courtyard”}\textit{at the rear}\lebnote{Literally “which was on behind her/it”}and its galleries, \textit{a hundred cubits\lebnote{That is, 175 feet}on each side},\lebnote{Literally “from here and from here”}and also the temple, the court, and the porticos of the courtyard,}%
\verse{the thresholds and the framed windows and the galleries around the three of them. Before the threshold was a covering of wood all around,\lebnote{Literally “all around, all around”}and from the ground up to the windows, and all around the windows, they were covered.}%
\verse{\textit{Above}\lebnote{Literally “On from on”}the doorway and up to the inner temple\lebnote{Or “sanctuary”}and on the outside, and on all of the wall \textit{all the way around}\lebnote{Literally “all around, all around”}in the inner and in the outer areas were \textit{patterns},\lebnote{Literally “measurements”}}%
\verse{and it was made of cherubim and palm tree images; a palm tree image between cherub and cherub, and \textit{the cherub had two faces}.\lebnote{Literally “two faces were for the cherub”}}%
\verse{And the face of a human \textit{was toward}\lebnote{Literally “was to”}the palm tree image \textit{on the one side},\lebnote{Literally “from here”}and the face of a fierce strong lion \textit{faced}\lebnote{Literally “was to”}the palm tree image \textit{on the other side};\lebnote{Literally “from here”}\textit{this work was executed}\lebnote{Literally “being made”}for the entire temple \textit{all the way around}.\lebnote{Literally “all around, all around”}}%
\verse{From the ground up to \textit{above}\lebnote{Literally “up to form on”}the doorway,\lebnote{Or “entrance”}the cherubim and the palm tree images were made,\lebnote{Or “being made”}and also the outer wall of the temple.}%
\verse{As far as the temple is concerned its doorframe was squared, and \textit{before}\lebnote{Literally “in the front of”}the sanctuary was the appearance as it were the appearance of}%
\verse{\textit{a wooden altar}\lebnote{Literally “the altar made of wood”}that was three cubits\lebnote{That is, 5.25 feet}high, and its length was two cubits,\lebnote{That is, 3.5 feet}and its corners\lebnote{Hebrew “corner”}for it and its length and its walls were of wood. And he spoke to me, “This is the table that is \textit{before}\lebnote{Literally “to the face of”}Adonai.”}%
\verse{And the two doors were for the temple and for the sanctuary.}%
\verse{And two leaves\lebnote{That is, two swinging leaves for each door}of a door were for each of the doors, two hinged leaves of a door: two were for the first door, and two leaves of a door were for the other door.}%
\verse{And cherubim were made\lebnote{Or “carved”}on them, that is, on the doors of the temple and palm tree images like \textit{the ones prepared for the walls};\lebnote{Literally “like which they were made for the walls”}and an overhang of wood was on the surface of the porticos on the outside.}%
\verse{And narrow windows and palm tree images \textit{were on either side},\lebnote{Literally “were from here and from here”}and on the side walls of the portico, and the side rooms of the temple and their\lebnote{Literally “the”}overhang.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 42
\verseWithHeading{Description of Rooms for the Priests}{And he brought me out to the outer courtyard \textit{to the north},\lebnote{Literally “to the way of the north”}and he brought me to the chamber which was opposite the courtyard and which is opposite the building to the north.}%
\verse{As to\lebnote{Or “along the face of”}the face of the length of the building with the doorway to the north, it was a hundred cubits,\lebnote{That is, 175 feet}and its\lebnote{Or “the”}width was fifty cubits.\lebnote{That is, 87.5 feet}}%
\verse{Opposite the twenty cubits \textit{of the inner courtyard},\lebnote{Literally “which were to the courtyard the inner”}and opposite the pavement that was to the outer courtyard was a gallery \textit{facing}\lebnote{Literally “toward in the face/front of”}a gallery in the three stories.}%
\verse{And \textit{in front of}\lebnote{Literally “to the face of”}the chambers was a passageway ten cubits\lebnote{That is, 17.5 feet}in width toward the inside, a walkway of one cubit,\lebnote{One cubit of 18–21 inches is problematic; LXX and Syriac read 100 cubits (175 feet) running the length of the building}and their doorways were to the north.}%
\verse{And the upper chambers narrowed, for the galleries took away space from them more than they took from the lower levels and more than they took from the middle level in the building.}%
\verse{For they\lebnote{Or “there”}were three stories and \textit{they had no pillars}\lebnote{Literally “there were not for them pillars”}like\lebnote{Or “as”}the pillars of\lebnote{Or “for”}the courtyards; therefore they were smaller than the lower stories and than the middle stories from the ground up.}%
\verse{And there was a wall that was to the outside \textit{alongside}\lebnote{Literally “to the outside to corresponding”}the chambers on the walkway to\lebnote{Or “of”}the outer courtyard\lebnote{That is, “to the north”}\textit{in front of}\lebnote{Literally “to in face of”}the chambers; its length was fifty\lebnote{That is, 87.5 feet}cubits.\lebnote{Hebrew “cubit”}}%
\verse{For\lebnote{Or “while”}the length of the chambers which were to the outer courtyard was fifty\lebnote{Hebrew “cubit”}cubits,\lebnote{That is, 87.5 feet}and look!\lebnote{Or “yet!”}The chambers on the front of the temple were a hundred\lebnote{Hebrew “cubit”}cubits.\lebnote{That is, 175 feet}}%
\verse{And from under these chambers was the entrance\lebnote{Reading the Ketiv}from the east for them \textit{when one enters}\lebnote{Literally “at/in his/its coming”}from the outer courtyard.}%
\verse{All along the width of the wall of the courtyard \textit{eastward}\lebnote{Literally “on the way of the east”}in front of the courtyard to the front of the building were chambers.}%
\verse{And a walkway was \textit{before them}\lebnote{Literally “was to the face of them”}like the appearance of the chambers which were on \textit{the north},\lebnote{Literally “were on the way of the north”}just like them in their length, and so was their width and all their exits, and they were built like their arrangements and like their doorways,}%
\verse{and like the doorways of the chambers which were on the way of the south was a doorway at the head of the way\lebnote{Dropping one “way”}\textit{before} the stone wall, projecting\lebnote{Or “protecting wall”}on\lebnote{Or “toward”}the way of the east at their coming.\lebnote{That is, when people entered the complex}}%
\verse{And he said to me, “The chambers of the north and the chambers of the south which are \textit{before}\lebnote{Literally “to the front of”}the courtyard, they \textit{are the holy chambers}\lebnote{Literally “are the chambers of the holiness”}in which the priests, who are near to Adonai, will eat the \textit{most holy objects}.\lebnote{Literally “holy objects of the holy objects”}There they shall put the \textit{most holy objects},\lebnote{Literally “holy objects of the holy objects”}and the grain offering and the sin offering and the guilt offering, for the place is holy.}%
\verse{\textit{When the priests enter},\lebnote{Literally “at coming their”}then they shall not go out from the sanctuary to the outer courtyard; and there they must put\lebnote{Or “deposit/leave/remove”}their garments in which they serve because they are holy. They must put on other garments and then they may approach the area that is for the people.}%
\verse{And he completed the measurements of the inner temple, and he brought me to the walkway of the gate \textit{that faces toward the east}\lebnote{Literally “which its face is to the way of the east”}and he measured it \textit{all the way around}.\lebnote{Literally “all around, all around”}}%
\verse{He measured the east side with the reed\lebnote{Or “rod”}for\lebnote{Literally “of measuring”}measuring, five hundred cubits,\lebnote{Hebrew has “reeds” here (= 5,250 feet), which is not correct; 500 cubits = 875 feet.}with respect to reeds with the reed for measurement, he measured it all around.}%
\verse{He measured the north side as five hundred cubits, with respect to reeds with the reed for measurement all around.}%
\verse{Then he measured the south side as five hundred cubits, with respect to reeds with the reed for measurement.}%
\verse{He went around the west side\lebnote{Or “side toward/to the sea”}and he measured five hundred cubits, with respect to reeds with the reed for measurement.}%
\verse{\textit{Toward the four sides}\lebnote{Literally “toward the four winds”}he measured it; there was a wall for it \textit{all the way around}.\lebnote{Literally “all around, all around”}Its length was five hundred cubits\lebnote{That is, 875 feet}and its width was five hundred cubits, in order to make a separation between what is holy and what is common.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 43
\verseWithHeading{Return of the Lord’s Glory and Description of Sacrifices and Worship}{And he brought me to the gate which was \textit{facing east}.\lebnote{Literally “which was facing to the way of the east”}}%
\verse{And, look! The glory of the God of Israel, it came from the way of the east, and its sound was like the sound of many waters\lebnote{Hebrew “water”}and the land radiated due to his glory!}%
\verse{And\lebnote{Or “and as”; as is not needed in translation}the appearance of the vision which I saw was as\lebnote{Or “like”}the vision which I saw at his\lebnote{Or “my coming”}coming to destroy the city, and these visions were also as\lebnote{Or “like”}the vision which I saw by the \textit{Kebar River},\lebnote{Literally “river of Kebar”}and I fell on my face.}%
\verse{And then the glory of Adonai came to the temple \textit{by the way of the gate facing east}.\lebnote{Literally “by the way of the gate which its face was to the way of the east”}}%
\verse{And the Spirit lifted me and it brought me to the inner courtyard and, look! The glory of Adonai filled the temple!}%
\verse{And I heard someone speaking to me from the house,\lebnote{Or “from inside”}and a man was standing beside me.}%
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}this is the place of my throne and the place for the soles of my feet \textit{where I will dwell}\lebnote{Literally “which I will dwell there”}in the midst of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}to eternity, and they, the house of Israel, they and their kings, will not again defile \textit{my holy name}\lebnote{Literally “the name of my holiness”}with their fornication and with their offerings for the dead\lebnote{Or “corpses”}of their kings on their high places.}%
\verse{\textit{When they placed their threshold}\lebnote{Literally “in/at their placing their threshold”}with\lebnote{Or “beside/next to”}my threshold, their doorframe beside my doorframe and the\lebnote{Or “their”}wall was between me and between them, then they defiled \textit{my holy name},\lebnote{Literally “the name of my holiness”}with their detestable things that they did, and so I consumed them in my anger.}%
\verse{Now let them send their fornication far away and the offerings for the dead\lebnote{Or “corpses”}of their kings away from me, and I will dwell in the midst of them \textit{forever}.\lebnote{Literally “to eternity”}}%
\verse{You, son of man,\lebnote{Or “mortal,” or “son of humankind”}describe to the house of Israel the temple and let them be ashamed of their iniquities and let them measure the pattern.}%
\verse{And if they are ashamed of all that they did, then the plan of the temple and its arrangement and its exits and its entrances and all of its plans\lebnote{Or “its entire pattern”}and all of its statutes. And all of its plans\lebnote{Or “purposes”}and all of its laws\lebnote{According to the reading tradition (\textit{Qere}); cf. BHS}make known to them and write them before their eyes, so that they may remember all of its plans and all of its statutes, so that they do them!}%
\verse{This is the law\lebnote{Literally “torah” or “instruction”}of the temple: On the top of the mountain, all of its territory, \textit{all the way around it},\lebnote{Literally “all around, all around”}\textit{will be most holy}.\lebnote{Literally “will be a holiness of holinesses”}Look, this is the law of the temple.}%
\verse{And these are the measurements of the altar in the cubits\lebnote{Or “in long cubits”}(a cubit is a cubit and a handbreadth):\lebnote{That is, about 18 inches (regular cubit) + 3 inches = 21 inches (long cubit)}now its gutter\lebnote{Or “trench” (NJPS)}\textit{is a cubit\lebnote{That is, 21 inches}in depth by a cubit\lebnote{That is, 21 inches}in width}\lebnote{Literally “the cubit and cubit its width”}and its rim along its edge all around is one span,\lebnote{That is, 9 inches}and this is also the height of the altar.}%
\verse{And from the gutter\lebnote{Or “trench”}at the ground up to the lower ledge is two cubits,\lebnote{That is, 3 feet 6 inches}and its width is one cubit,\lebnote{That is, 21 inches}and from the small ledge up to the large ledge is four cubits,\lebnote{That is, 7 feet}and its width is one cubit.}%
\verse{And the altar hearth was four cubits,\lebnote{That is, 7 feet}and from the altar hearth and \textit{upwards}\lebnote{Literally “and to above”}were the four horns of the altar.}%
\verse{And the altar hearth was \textit{twelve}\lebnote{Literally “two ten”}cubits in length and \textit{twelve}\lebnote{Literally “two ten”}cubits in width\lebnote{That is, 21 feet square}; it was \textit{squared on its four sides}.\lebnote{Literally “squared to the four of its sides”}}%
\verse{And the ledge was \textit{fourteen}\lebnote{Literally “four ten”}cubits in length with \textit{fourteen}\lebnote{Literally “four ten”}cubits its width\lebnote{That is, 24.5 feet on all four sides, a square}to all four of its sides, and a rim was all around it of one-half cubit.\lebnote{That is, 10.5 inches}And the gutter\lebnote{Or “trench”}for it was a cubit\lebnote{That is, 21 inches}all around, and its steps were facing east.”}%
\verse{And he said to me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}thus says the Lord Adonai: ‘These are the statutes of the altar on the day \textit{when it is made},\lebnote{Literally “to be made it”}to sacrifice a burnt offering on it and to sprinkle blood on it.}%
\verse{And you must give to the Levitical priests who are from the offspring of Zadok, the ones coming near me,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, “to serve me, a bull, \textit{a calf}\lebnote{Literally “son of cattle,” or “son of the herd”}as a sin offering.}%
\verse{You must take some of its blood and put it on \textit{the four horns of the altar},\lebnote{Literally “on its four horns”}and the four corners of the ledge, and on the rim all around; thus you must purify it and make atonement for it.}%
\verse{And you must take the bull, the sin offering, and you must burn it in the designated place of the temple outside of the holy place.}%
\verse{And on the second day you must offer a he-goat without defect as a sin offering, and they must purify the altar \textit{like}\lebnote{Literally “as that”}they purified with the bull.}%
\verse{\textit{When you are finished}\lebnote{Literally “at your finishing”}from purifying, you must offer a bull, \textit{a calf},\lebnote{Literally “son of cattle,” or “son of the herd”}without defect, and a ram from the flock without defect.}%
\verse{And you must bring them near \textit{before}\lebnote{Literally “to the face of”}Adonai, and the priests must throw salt on them, and they must offer them as a burnt offering to Adonai.}%
\verse{For seven days you must provide a goat of sin offering for the day and a bull, \textit{a calf},\lebnote{Literally “son of cattle,” or “son of the herd”}and a ram from the flock without defect; you must provide them.}%
\verse{Seven days they must purify the altar and they must cleanse it and so \textit{they will consecrate it}.\lebnote{Literally “they must fill its hand”}}%
\verse{And they will finish the days, \textit{and then}\lebnote{Literally “and it will happen”}on the eighth day and beyond that the priests will offer on the altar your burnt offerings and your fellowship offerings, and then I will be pleased with you,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 44
\verseWithHeading{The Prince, the Levites, and the Priests Described}{And he brought me back by the way of the outer gate of the sanctuary that is facing east, and it was shut.}%
\verse{And Adonai said to me, “This gate will be shut. It shall not be opened, and \textit{no one}\lebnote{Literally “and a man not”}will go through it, for Adonai, the God of Israel, has entered it, and it will be shut.}%
\verse{The prince, he may sit in it to eat food \textit{before}\lebnote{Literally “to the face of”}Adonai; he will come from the way\lebnote{Or “direction”}of the portico of the gate and by means of its way he will also go out.”}%
\verse{Then he brought me by the way of the gate of the north to the front of the temple,\lebnote{Or “house”}and I looked,\lebnote{Or “saw”}and look! The glory of Adonai filled the temple\lebnote{Or “house”}of Adonai, and I fell on my face.}%
\verse{And Adonai said, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}set your heart\lebnote{Or “pay close attention” or “listen closely”}and look with your eyes and with your ears hear \textit{everything}\lebnote{Literally “all of which/what”}I am saying to you concerning all the statutes of the temple\lebnote{Or “house”}of Adonai and concerning all its laws,\lebnote{According to the reading tradition (\textit{Qere}); cf. BHS}and you must listen with your heart concerning the entrance of the temple\lebnote{Or “house”}with all of the exits of the sanctuary.}%
\verse{And you must say to the rebellious, to the house of Israel, ‘Thus says the Lord Adonai: “Enough for you, house of Israel, of\lebnote{Or “from”}all of your detestable things!}%
\verse{At\lebnote{Or “in”}your bringing \textit{foreigners}\lebnote{Literally “children/people of a foreign land”}who are uncircumcised of heart and uncircumcised of flesh to be in my sanctuary to profane it, my temple,\lebnote{Or “house”}\textit{as you offered my food, fat, and blood},\lebnote{Literally “in/at your offering my food and fat and blood”}so\lebnote{Or “and”}you broke my covenant by all of your detestable things.}%
\verse{And you did not observe the responsibility of my sanctuary, but\lebnote{Or “and”}you appointed them\lebnote{That is, foreigners}as the keepers of my responsibility in my sanctuary for you.”}%
\verse{Thus says the Lord Adonai: “Every \textit{foreigner}\lebnote{Literally “child of a foreign land”}uncircumcised of heart and uncircumcised of flesh shall not come into my sanctuary—not any of the \textit{foreigners}\lebnote{Literally “child of a foreign land”}who are in the midst of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}}%
\verse{\textit{But}\lebnote{Literally “For if”}the Levites who removed themselves from me \textit{at the going astray of Israel},\lebnote{Or “when Israel went astray”}who went astray from me and went after their idols, as a result they will bear their guilt.}%
\verse{Then they will be\lebnote{Or “may be”}in my sanctuary, those serving in my sanctuary, as sentries\lebnote{Or “guards”}at the gates of the temple\lebnote{Or “house”}and as those serving the temple;\lebnote{Or “house”}they will slaughter the burnt offering and the sacrifice for the people, and they will stand \textit{before them}\lebnote{Literally “to the face of them”}to serve them,}%
\verse{\textit{because}\lebnote{Literally “because that”}they used to serve them \textit{before}\lebnote{Literally “to the face of”}their idols and they were for the house of Israel like a stumbling block of iniquity. Therefore \textit{I swore}\lebnote{Literally “I raised my hand”}concerning them,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai, “that they will bear their iniquity.}%
\verse{And they shall not approach me to serve as a priest for me and to come near to all of\lebnote{Or “any of”}my holy objects, \textit{to the most holy objects}, \lebnote{Literally “to the holy objects of the holy objects”}and they will bear their disgrace and the results of their detestable things they have done.}%
\verse{And I will appoint them as the keepers of the responsibility of\lebnote{Or “toward”}the temple, for all of its work\lebnote{Or “tasks”}and \textit{everything} \lebnote{Literally “all”}which will be done in it.}%
\verse{But the Levitical priests, the descendants\lebnote{Or “sons”}of Zadok, who cared for the responsibility of my sanctuary \textit{when the Israelites went astray}\lebnote{Literally “at going astray the children of Israel”}from me, they will approach me to serve me, and they will stand \textit{before me}\lebnote{Literally “to the face of me”}to offer to me fat and blood,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“They shall come to my sanctuary, and they shall approach my table to serve me, and they will observe my requirement.}%
\verse{\textit{And then}\lebnote{Literally “and it will happen”}\textit{when they come}\lebnote{Literally “at their coming”}to the gates of the inner courtyards, they shall put on their\lebnote{Hebrew “the”}inner linen garments and not wear wool garments \textit{when they serve}\lebnote{Literally “at their serving”}in the gates of the inner courtyard and \textit{inside}\lebnote{Literally “and toward”}the temple.}%
\verse{Linen turbans shall be on their head and linen undergarments shall be on their waists; they shall not gird themselves with material causing perspiration.}%
\verse{When they go out to the outer courtyard to the people, they must take off their garment in which they were serving, and they must place them \textit{in the holy chambers},\lebnote{Literally “in the chambers of the holiness”}and they must put on other garments, so that they will not make the people holy with their garments.}%
\verse{And their head they shall not shave, or long hair they shall not let grow; short they shall clip their heads.}%
\verse{And no priest shall drink wine when they come into the inner courtyard.}%
\verse{And a widow or divorced woman they shall not take for themselves as wives, \textit{but only}\lebnote{Literally “but if”}a virgin from the offspring of the house of Israel, or the widow who is a widow for a priest they may take.}%
\verse{And they will teach my people the difference between what is holy and what is unholy, and the difference between unclean and clean they must show them.}%
\verse{And at a legal dispute they themselves shall stand to judge; employing\lebnote{Or “with”}my judgments they shall judge it, and with respect to my laws and my statutes. All my festivals they shall keep and my Sabbaths they shall consecrate.}%
\verse{And \textit{near a dead person}\lebnote{Literally “and to a dead person of a man”}he\lebnote{That is, the priest}shall not come to be defiled, \textit{but only}\lebnote{Literally “but if”}for a father and for a mother and for a son and for a daughter, for a brother and for a sister \textit{who was not married}\lebnote{Literally “who not she was for a man”}may they defile themselves.}%
\verse{And, after his\lebnote{That is, the priest}cleansing, they shall count for him seven days,}%
\verse{and then on the day of his coming into the sanctuary to the inner courtyard to serve in the sanctuary, he shall offer his sin offering,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{And thus it will be to them as regards inheritance, that I am their inheritance, and so you shall not give to them property in Israel; I am their property.}%
\verse{The grain offering and the sin offering and the guilt offering, they themselves may eat them, and also all the consecrated possessions\lebnote{Hebrew “possession”}in Israel will be theirs.}%
\verse{And also what is first of all of the firstfruits of everything, and of all of the contribution of everything from all of your contributions, to the priest \textit{it belongs},\lebnote{Literally “it shall be”}and what is first of your dough you shall give to the priest, so that a blessing may rest on your house.}%
\verse{Any dead body or mangled carcass from the birds\lebnote{Hebrew “bird”}or from the animals,\lebnote{Hebrew “animal”}the priests shall not eat.”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 45
\verseWithHeading{Instructions about Divisions of the Renewed Land}{And when you allocate the land as an inheritance, you shall provide a contribution for Adonai as a holy portion from the land, its length being twenty-five thousand cubits\lebnote{Or 8.33 miles}and its width ten thousand cubits;\lebnote{Hebrew 3.5 miles; LXX reads 6.67 miles in width, or 20,000 cubits (cf. 45:3, 5; 48:9)}it is holy in all its territory, all around.}%
\verse{And there shall be from this area five hundred cubits\lebnote{That is, “875 feet by 875 feet”}by five hundred cubits, squared all around, for the sanctuary; and fifty cubits\lebnote{Hebrew “cubit”}﻿\lebnote{That is, “87.5 feet”}of open space shall be for it all around it.}%
\verse{And from this measured area you shall measure a length of twenty-five\lebnote{According to the reading tradition (\textit{Qere})}thousand cubits\lebnote{That is, 8.33 miles; cf. 45:5}and a width of ten thousand cubits,\lebnote{That is, 3.5 miles; cf. 45:5}and in it will be the sanctuary, \textit{the most holy place}.\lebnote{Literally “the holiness of holinesses”}}%
\verse{It\lebnote{Or “this area”}is a holy portion from the land; it will be for the priests, the servants of the sanctuary \textit{who approach}\lebnote{Literally “the ones who approach”}to serve Adonai, and it will be for them a place for houses and a holy place for the sanctuary.}%
\verse{And an area twenty-five thousand cubits in length and ten thousand cubits in width will be for the Levites, the servants of the temple. It will serve as property for them \textit{as cities to dwell in}.\lebnote{Literally “twenty chambers”}}%
\verse{And as property of the city, you must set apart alongside the contribution of the sanctuary a portion five thousand cubits\lebnote{That is, 1.67 miles}in width and twenty-five thousand cubits\lebnote{That is, 8.33 miles}in length, and it shall be for the entire house of Israel.}%
\verse{And a portion will be for\lebnote{Hebrew “to”}the prince \textit{on both sides}\lebnote{Literally “from this and from this”}of \textit{the holy district},\lebnote{Literally “the contribution of the holiness”}and both sides of the property of the city, and \textit{alongside}\lebnote{Literally “to the front of”}the property of the city \textit{on the west},\lebnote{Literally “from/on the side of the sea”}and alongside the property \textit{on the east}.\lebnote{Literally “from/on the side of east”}And also its length\lebnote{Or “portion”}\textit{corresponds to one of the tribal portions to the west}\lebnote{Literally “corresponding to one of the portions from the boundary of the sea”}running \textit{to the eastern border}.\lebnote{Literally “to the boundary of the east”}}%
\verse{This shall be to him \textit{with respect to the land}\lebnote{Literally “to the land”}as property in Israel; and so my princes shall not again oppress my people, but they shall give the land to the house of Israel according to their tribes.”}%
\verse{Thus says the Lord Adonai, “Enough of this for you, the princes of Israel; put away violence and destruction and do justice and righteousness; revoke your acts of dispossession from upon\lebnote{Or “against”}my people!” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{There shall be for you \textit{an honest set of scales}\lebnote{Literally “a set of scales of justice/honesty”}and \textit{an honest ephah}\lebnote{Literally “an ephah of honesty”}and \textit{an honest bath}.\lebnote{Literally “a bath of honesty”; a bath is a liquid measure}}%
\verse{The ephah and the bath shall be one unit of measurement;\lebnote{Or “shall be the same size”}the tenth part of the homer is the bath, and the tenth of the homer is the ephah; so the homer shall be its\lebnote{Or “the”}unit of measurement.}%
\verse{And the shekel shall weigh twenty gerahs, twenty shekels\lebnote{Hebrew “shekel”}and five and twenty shekels\lebnote{Hebrew “shekel”}and ten and five shekels,\lebnote{Hebrew “shekel”}that shall make the mina for you.\lebnote{Meaning of Hebrew uncertain}}%
\verse{This is the contribution offering which you shall present: a sixth of the ephah from a\lebnote{Or “each”; literally “the”}homer of wheat, and a\lebnote{Literally “the”}sixth of the ephah from a\lebnote{Literally “the”}homer of barley.}%
\verse{And the quota of the olive oil, the bath of the olive oil, is the tenth part of a bath from a kor, which is ten baths, or a homer—for ten baths are equal to a homer.}%
\verse{And one\lebnote{Or “a”}sheep from the flock from among two hundred from the pastures of Israel will be taken as a grain offering and as a burnt offering and as a fellowship offering to make atonement for them,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{“All of the people of the land shall join in\lebnote{Or “participate/ take part in”}to this contribution with\lebnote{Or possibly “for/to”}the prince in Israel.}%
\verse{“But on the prince shall be the responsibility for the burnt offerings, and the grain offering, and the libation at the feasts, and at the New Moon festivals, and at\lebnote{Or “on”}the Sabbaths at all of the assemblies of the house of Israel; he shall provide the sin offering, and the grain offering, and the burnt offering, and the fellowship offering to make atonement for the house of Israel.}%
\verse{Thus says the Lord Adonai: ‘On the first month on the first day of the month, you shall take a bull, \textit{a calf}\lebnote{Literally “son of cattle,” or “son of the herd”}without defect, and you shall purify the sanctuary.}%
\verse{And the priest shall take from the blood of the sin offering, and he shall put it on the doorframe of the temple\lebnote{Or “house”}and on the four corners of the ledge of the altar and on the doorframe of the gate of the inner courtyard.}%
\verse{And so you shall do on the seventh day in the month for anyone doing wrong inadvertently or due to ignorance, and so you must make atonement for the temple.\lebnote{Or “house”}}%
\verse{In the first month, on the \textit{fourteenth}\lebnote{Literally “four ten”}day of the month, you shall have the Passover, a feast lasting for \textit{seven days},\lebnote{Literally “a week of days”}when you shall eat unleavened breads.}%
\verse{And the prince shall provide on that day for himself\lebnote{Hebrew “him”}and for all of the people of the land a bull as a sin offering.}%
\verse{And during the seven days of the religious feast he shall provide as a burnt offering to Adonai seven bulls and seven rams without defect for each of the seven days, and as a sin offering a he-goat for each day.}%
\verse{And as a grain offering, an ephah for\lebnote{Hebrew “to”}the bull and an ephah for\lebnote{Hebrew “to”}the ram he must provide, and a hin of olive oil for \textit{each}\lebnote{Literally “the”}ephah.}%
\verse{In the seventh month, on the \textit{fifteenth}\lebnote{Literally “five ten”}day of the month, during the feast, he shall provide just as he has for these seven days, as he did for the sin offering, as he did for the burnt offering, and as he did for the grain offering, and as he did for the olive oil.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 46
\verse{Thus says the Lord Adonai: “The gate of the inner courtyard facing east must be shut on\lebnote{Or “during”}the six of the days for work, but on the day of the Sabbath it must be opened and on the day of the new moon it must be opened.}%
\verse{And the prince shall come by way of the portico of the gate from the outside, and he shall stand at the doorframe of the gate and the priests shall prepare his burnt offering and his fellowship offerings, and he shall bow down at the threshold of the gate. Then he shall go out, but the gate shall not be closed until the evening.}%
\verse{And the people of the land will bow down at the doorway of that gate on the Sabbaths and on the new moons \textit{before}\lebnote{Literally “to the face of”}Adonai.}%
\verse{And the burnt offering which the prince will offer to Adonai on the Sabbath day must be six male lambs without defect and a ram without defect.}%
\verse{And the grain offering he will give shall be an ephah \textit{for each ram},\lebnote{Literally “to the ram”}and for the male lambs the grain offering shall be \textit{as much as he wants to give}\lebnote{Literally “the gift of his hand”}and a hin of olive oil \textit{for each ephah}.\lebnote{Literally “to the ephah”}}%
\verse{And on the day of the new moon he will offer a bull, \textit{a calf}\lebnote{Literally “son of cattle,” or “son of the herd”}without defect and six male lambs and a ram; they must be without defect.}%
\verse{And an ephah for each bull and an ephah for the ram he must provide as a grain offering, and also for the male lambs \textit{as much as he can afford}\lebnote{Literally “as that it stretches out his hand”}and a hin of olive oil for each ephah.}%
\verse{And when the prince comes, on\lebnote{Or “by”}the way of the portico of the gate he shall come and by this\lebnote{Or “his”}way he shall go out.}%
\verse{And when the people of the land come \textit{before}\lebnote{Literally “to the face of”}Adonai at the festivals, the one coming by way of the gate of the north to bow down\lebnote{Or “to worship”}must go out by the way of the gate of the south, and the one coming by way of the gate of the south must go out by the way of the gate of the north, so he must not return by the way of the gate by which he came,\lebnote{Hebrew “came by it”}but before it he\lebnote{According to the reading tradition (\textit{Qere})}shall go out opposite it.}%
\verse{And the prince will be in the midst of them; he will come in when they come in, and when they go out he will go out.}%
\verse{And at the festivals and at the appointed times, the grain offering will be an ephah with a bull and an ephah with the ram and with the male lambs, \textit{as much as he wants to give},\lebnote{Literally “the gift of his hand”}and a hin of olive oil \textit{for each ephah}.\lebnote{Literally “to the ephah”}}%
\verse{And when the prince makes a freewill offering, a burnt offering, or a fellowship offering as a freewill offering to Adonai, the gate facing east will be open for him. And he must offer his burnt offering and his fellowship offering \textit{just as}\lebnote{Literally “as that”}he does on the day of the Sabbath, and then he must go out and the gate will shut \textit{after he goes out}.\lebnote{Literally “after his going out”}}%
\verse{And a male lamb, \textit{a yearling}\lebnote{Literally “a son of his year”}without defect, he must provide as a burnt offering \textit{daily}\lebnote{Literally “for the day”}for Adonai; \textit{every morning}\lebnote{Literally “in the morning in the morning”}he must provide it.}%
\verse{And a grain offering he must provide in addition to it \textit{every morning},\lebnote{Literally “in the morning in the morning”}a\lebnote{Hebrew “the”}sixth of an\lebnote{Hebrew “the”}ephah and a third of a hin of olive oil to moisten the finely milled flour as a grain offering to Adonai \textit{as a perpetual statute}.\lebnote{Literally “regulations of everlasting continually”}}%
\verse{So they must provide the male lamb and the grain offering and the olive oil \textit{every morning}\lebnote{Literally “in the morning in the morning”}\textit{as a regular burnt offering}.”\lebnote{Literally “as burnt offering of regularity/continuity”}}%
\verse{Thus says the Lord Adonai: “If the prince gives a gift to any one from among his sons, it is his inheritance and it will belong\lebnote{Hebrew “be”}to his sons; it is their possession as an inheritance.}%
\verse{And if he give a gift from his inheritance to one from among his servants, then it will belong\lebnote{Hebrew “be”}to him until the year of release,\lebnote{That is, the Year of Jubilee}and then it will return to the prince. His inheritance with respect to his sons will be only for them.}%
\verse{But the prince shall not take from the inheritance of the people to rob\lebnote{Or “oppress”}them, separating them from their property; from his own property he shall bestow an inheritance on his sons, so that my people will not be driven away, any one, from his property.”}%
\verse{Then he brought me through the entrance which was on the side of the gate \textit{to the holy chambers}\lebnote{Literally “to the chamber of the holiness”}to the priests; the chambers were facing northwards and, look, there was a place at the extreme end westwards.}%
\verse{And he said to me, “This is the place where the priests will boil the guilt offering and the sin offering, and where they shall bake the grain offering in order not to bring them out to the outer courtyard to make the people holy.}%
\verse{And he brought me to the outer courtyard and led me past the four corners of the courtyard; and look, \textit{there was a courtyard in each corner of the courtyard}.\lebnote{Literally “a courtyard in the corner of the courtyard, a courtyard in the corner of the courtyard”}}%
\verse{In the four corners of the courtyard were small\lebnote{Hebrew uncertain}courtyards forty cubits in length and thirty cubits in width, each with \textit{the same measurements}\lebnote{Literally “measurement one”}for the four of them with corners.\lebnote{Or “being cornered”}}%
\verse{And a row was all around inside\lebnote{Hebrew “in”}them, and cooking-places were made under the rows of stones\lebnote{Possibly some kind of stone ledge}all around.}%
\verse{And he said to me, “These are the houses for\lebnote{Hebrew “of”}cooking, where the ones serving the temple shall cook the sacrifices of the people.”}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 47
\verseWithHeading{River of the Temple, Division of the Land and Its Boundaries}{And he brought me back to the doorway of the temple and, look! There was water coming out from under the threshold of the temple eastward, because the face of the temple was eastwards; and the water was coming down from under the south side of the temple, from south of the altar.}%
\verse{And he brought me out by the way of the north gate and he led me around by the \textit{way leading to the outside of the outer gate}\lebnote{Literally “way of the outside to the gate of the outside”}by way of the gate facing east, and look, there was water trickling from the south side.}%
\verse{As the man went eastward a measuring line was in his hand, and he measured a thousand \textit{cubits}.\lebnote{Literally “in the cubit”; about 1,750 feet}And he led me through on into the water; it was \textit{ankle deep}.\lebnote{Literally “water of ankles”}}%
\verse{And he measured a thousand cubits,\lebnote{About 1,750 feet}and he brought me over into the waters and it was \textit{knee deep}.\lebnote{Literally “water of knees”}And he measured a thousand cubits,\lebnote{About 1,750 feet}and he brought me over into the waters and it was \textit{waist deep}.\lebnote{Literally “water of waist”}}%
\verse{And he measured a thousand cubits,\lebnote{About 1,750 feet}and it became a stream that I was not able to cross, because the water rose, \textit{waters a person could swim in},\lebnote{Literally “water of swimming”}and became a stream that could not be crossed.}%
\verse{And he asked me, “Son of man,\lebnote{Or “mortal,” or “son of humankind”}did you see this?” And he made me go, and he brought me back along the bank of the stream.}%
\verse{\textit{When I returned},\lebnote{Literally “At my returning”}then look! On the bank of the river were many trees \textit{on both sides}.\lebnote{Literally “from this and from this”}}%
\verse{And he said to me, “These waters are going out to the eastern region, and they go down to the \textit{Jordan Valley},\lebnote{Literally “Arabah”}and they come to the sea\lebnote{Clearly refers to the Dead Sea}and flow into the sea where \textit{they issue out},\lebnote{Literally “being brought out/are issued out”}and the waters in the sea will be healed.\lebnote{Or “made fresh”}}%
\verse{And then every living creature with which it teems,\lebnote{Or “swarms”}\textit{to every place the waters come}\lebnote{Literally “to all of which they come there the water”}it will live; and very many fish will live because these waters came there, and so the water will be healed\lebnote{Or “fresh”}and everything will be alive \textit{everywhere that}\lebnote{Literally “all that”}\textit{the stream will come}.\lebnote{Literally “that it will come there the stream”}}%
\verse{And then fishers\lebnote{Or “fishermen”}from En Gedi and on up to En Eglaim will stand beside it, a sea-girt drying yard for nets; their fish for the dragnets will be \textit{of every kind},\lebnote{Literally “to the”}like the fish of the great sea,\lebnote{That is, the Mediterranean}and very many.}%
\verse{But its marshes and its swamps will not be cured, for they were given for salt.}%
\verse{And along the stream will go up on its banks \textit{from both sides}\lebnote{Literally “from this and from this”}every tree \textit{producing food};\lebnote{Literally “of food”}its leaf will not wither and it will not cease producing its fruit. \textit{Every month}\lebnote{Literally “for its months”}it will bear early fruit, for its waters are going out from the sanctuary, and its fruit will be as food, and its leaf for healing.”}%
\verse{Thus says the Lord Adonai: “This is the boundary by which you shall distribute the land for \textit{the twelve}\lebnote{Literally “two ten”}tribes of Israel: Two shares shall be for Joseph.}%
\verse{And you must take possession of it, each one a share like his brother, of this land which \textit{I swore}\lebnote{Literally “I raised my hand”}to give it to your fathers, and so this land shall fall to you as an inheritance.}%
\verse{And\lebnote{Or “Bow”}this is the boundary of the land: \textit{On the north side},\lebnote{Literally “to the side of the north”}from the great sea by the way of Hethlon \textit{until you come to Zedad},\lebnote{Literally “to the arriving at Zedad”}}%
\verse{Hamath\lebnote{Or “Lebo-Hamath to Zeded” (cf. NRSV, NJPS)}to Berothath to Sibraim, which is\lebnote{Or “are”}between the boundary of Damascus and the boundary of Hamath, on to Hazer Hatticon, which is on the boundary of Hauran.}%
\verse{And so the boundary will be from the sea\lebnote{That is, the Mediterranean}to Hazar Enan at the boundary of Damascus northwards and the boundary of Hamath to the north, and this is the boundary on the side of the north.}%
\verse{\textit{And the eastern boundary}\lebnote{Literally “and the side of the east”}will be between Hauran and Damascus, and from between Gilead and the land of Israel along the Jordan River from\lebnote{Or “to”}the boundary on the eastern sea to Tamar. And\lebnote{Or “Now”}\textit{this is the border on the east}.\lebnote{Literally “and this is the side of the east”}}%
\verse{\textit{And on the south side the border}\lebnote{Literally “and the side of the south to south”}will run from Tamar up to the waters of Meribot Kadesh and along the wadi\lebnote{That is, the Wadi of Egypt}to the Great Sea.\lebnote{That is, the Mediterranean}And this is \textit{the boundary to the south}.\lebnote{Literally “and this is the side of south to the south”}}%
\verse{\textit{And on the west side}\lebnote{Literally “and the side of the west”}the Great Sea\lebnote{That is, the Mediterranean}shall be the boundary up to opposite Lebo-Hamath. \textit{This is the western boundary}.\lebnote{Literally “this is the side of the west”}}%
\verse{You shall divide this land for yourselves according to the tribes of Israel.}%
\verse{And then you shall allocate it as an inheritance for yourselves and for the aliens dwelling as aliens in the midst of you who have children in the midst of you, and so they shall be to you just like full citizens among the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}With you they shall be allotted with an inheritance in the midst of the tribes of Israel.}%
\verse{And then in the tribe in which an alien who is with you dwells, there you shall give his inheritance,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Ezekiel 48
\verseWithHeading{The Allotment of the Land and the Gates of the City}{And\lebnote{Or “Now”}these are the names of the tribes: At the end of the north, \textit{along the way to Hethlon}\lebnote{Literally “to the side of the way of Hethlon”}from Lebo-hamath to Hazar Enan on the boundary of Damascus \textit{beside Hamath northward},\lebnote{Literally “northward to the side of Hamat,”}there will be one portion for Dan from the eastern border to the \textit{west}.\lebnote{Literally “sea”}}%
\verse{And next to the territory of Dan, from \textit{the eastern border}\lebnote{Literally “the side of the east}up to \textit{the western border},\lebnote{Literally “the side of the sea/west”}one portion for Asher.}%
\verse{And next to the territory of Asher, from \textit{the eastern border}\lebnote{Literally “the side of the east”}and up to \textit{the western border},\lebnote{Literally “the side of the west/sea”}one portion for Naphtali.}%
\verse{And next to the territory of Naphtali, from \textit{the eastern border}\lebnote{Literally “the side of the east”}up to \textit{the western border},\lebnote{Literally “the side of the sea/west”}one portion for Manasseh.}%
\verse{And next to the territory of Manasseh, from \textit{the eastern border}\lebnote{Literally “the side of the east”}up to \textit{the western border},\lebnote{Literally “the side of the sea/west”}one portion for Ephraim.}%
\verse{And next to the territory of Ephraim, from \textit{the eastern border}\lebnote{Literally “the side of the east”}and up to \textit{the western border},\lebnote{Literally “the side of the sea/west}one portion for Reuben.}%
\verse{And next to the territory of Reuben, from \textit{the eastern border}\lebnote{Literally “the side of the east}up to \textit{the western border},\lebnote{Literally “the side of the sea/west}one portion for Judah.}%
\verse{And next to the territory of Judah, from \textit{the eastern border}\lebnote{Literally “the side of the east}up to \textit{the western border},\lebnote{Literally “the side of the sea/west}shall be located the contribution that you set apart, twenty-five thousand cubits\lebnote{That is, 8.33 miles}in width, and its length just like one of the portions from \textit{the eastern border}\lebnote{Literally “the side of the east}up to \textit{the western border},\lebnote{Literally “the side of the sea/west}and the sanctuary shall be in the midst of it.}%
\verse{The contribution that you shall set apart for Adonai, its length shall be twenty-five thousand cubits,\lebnote{That is, 8.33 miles}and its width ten thousand cubits.\lebnote{That is, 3.5 miles; cf. 48:13 and 45:1}}%
\verse{And to these shall be \textit{the holy district}:\lebnote{Literally “the contribution of the holiness”}to the priests northwards twenty-five thousand cubits,\lebnote{That is, 8.33 miles}and westwards its width ten thousand cubits,\lebnote{That is, 3.5 miles}and eastwards its width ten thousand cubits,\lebnote{That is, 3.5 miles}and southwards its length twenty-five thousand cubits,\lebnote{That is, 8.33 miles}and the sanctuary of Adonai shall be in the midst of it.}%
\verse{To the priests, the ones being consecrated from the descendants\lebnote{Or “sons”}of Zadok, who took care of my service and who \textit{did not go astray}\lebnote{Literally “who not they went astray”}\textit{when the Israelites went astray},\lebnote{Literally “at going astray the children of Israel”}\textit{just as}\lebnote{Literally “as that”}the Levites went astray.}%
\verse{And it will be for them a special gift from the contribution of the land \textit{as a most holy object}\lebnote{Literally “as a holiness of holinesses”}next to the territory of the Levites.}%
\verse{And the Levites \textit{alongside}\lebnote{Literally “to corresponding”}the territory of the priests shall have twenty-five thousand cubits\lebnote{That is, 8.33 miles}in length and in width ten thousand cubits,\lebnote{That is, 3.5 miles}its \textit{full length},\lebnote{Literally “all/every of its length ….”}twenty-five thousand cubits\lebnote{That is, 8.33 miles}and its full width, ten thousand cubits.\lebnote{That is, 3.5 miles}}%
\verse{And they shall not sell any part from it, and shall not exchange, and shall not transfer the best of the land, for it is holy to Adonai.}%
\verse{And \textit{the remaining part}\lebnote{Literally “the being a part left over”}of five thousand cubits\lebnote{That is, 1.75 miles}in the width \textit{by}\lebnote{Literally “in front of”}twenty-five thousand cubits,\lebnote{That is, 8.33 miles}is unholy;\lebnote{Or “common”}it is for the city as dwelling and as pastureland; and the city will be located \textit{in the middle of it}.\lebnote{Literally “will be the city in its middle”}}%
\verse{And these shall be its measurements: on its side to the north, four thousand five hundred cubits;\lebnote{That is, 1.5 miles}and on its side to the south, four thousand five hundred cubits;\lebnote{That is, 1.5 miles}and on \textit{the eastern side},\lebnote{Literally “from the side of the east”}four thousand five hundred cubits;\lebnote{That is, 1.5 miles}and on \textit{the western side},\lebnote{Literally “and the side of the sea/west + directive”}four thousand five hundred cubits.\lebnote{That is, 1.5 miles}}%
\verse{And a pastureland shall be for the city northwards, two hundred and fifty cubits,\lebnote{That is, 150 yards}southwards two hundred and fifty cubits,\lebnote{That is, 150 yards}and eastwards two hundred and fifty cubits,\lebnote{That is, 150 yards}and westwards two hundred and fifty cubits.\lebnote{That is, 150 yards}}%
\verse{And the rest in its length \textit{alongside}\lebnote{Literally “to corresponding”}\textit{the holy district},\lebnote{Literally “the contribution of the holiness”}ten thousand cubits\lebnote{That is, 3.5 miles}eastwards and ten thousand\lebnote{That is, 3.5 miles}westwards; and it shall be \textit{alongside}\lebnote{Literally “to corresponding”}\textit{the holy district}\lebnote{Literally “the contribution of the holiness”}and its yield\lebnote{According to the reading tradition (\textit{Qere}); cf. BHS}shall be as food for the workers of the city.}%
\verse{And the workers of the city from all the tribes of Israel shall cultivate it.}%
\verse{All of the contribution, twenty-five thousand by twenty five thousand cubits\lebnote{That is, 8.33 square miles}square, you shall set apart as \textit{the holy district}\lebnote{Literally “the contribution of the holiness”}along with the property of the city.}%
\verse{And the rest shall be to\lebnote{Or “for”}the prince, \textit{on both sides}\lebnote{Literally “from this and from this”}\textit{of the holy district}\lebnote{Literally “of the contribution of the holiness”}and of the property of the city \textit{extending from}\lebnote{Literally “to in front of”}twenty-five thousand cubits of the contribution up to the boundary of the east, and westwards \textit{extending}\lebnote{Literally “to in front of”}twenty-five thousand cubits\lebnote{That is, 8.33 miles}to the boundary of the sea \textit{alongside}\lebnote{Literally “to corresponding”}the tribal portions. They shall be for the prince, and it shall be \textit{the holy district},\lebnote{Literally “the contribution of the holiness”}and the sanctuary of the temple\lebnote{Or “house”}will be in its middle.}%
\verse{And also the property of the Levites and the property of the city will be in the midst of the property which is for the prince; between the territory of Judah and between the territory of Benjamin the land will be for the prince.}%
\verse{And for the remainder of the tribes: From the side on the east up to the side on the west, Benjamin one portion;}%
\verse{and next to the territory of Benjamin, from the side on the east up to the side of the sea, Simeon one portion;}%
\verse{and next to the territory of Simeon, from the side of the east up to the side of the sea, Issachar one portion;}%
\verse{and next to the territory of Issachar, from the side of the east up to the side of the sea, Zebulun one portion;}%
\verse{and next to the territory of Zebulun, from the side of the east up to the side of the sea, Gad one portion;}%
\verse{and next to the territory of Gad, to the side of the south, \textit{southward};\lebnote{Literally “south”}and the territory will be from Tamar to the waters of Meribath Kadesh, toward the wadi,\lebnote{That is, the Wadi of Egypt}on to the Great Sea.\lebnote{That is, the Mediterranean}}%
\verse{This is the land that you shall allot as an inheritance to the tribes of Israel, and these are their portions,” \textit{declares}\lebnote{Literally “declaration of”}the Lord Adonai.}%
\verse{And these shall be the exits of the city: From\lebnote{Or “on” or “along”}\textit{the north side},\lebnote{Literally “the side of the north”}four thousand and five hundred cubits \lebnote{That is, 1.5 miles}by measurement.}%
\verse{And the gates of the city shall be according to the names of the tribes of Israel: three gates to the north, the gate of Reuben, one; the gate of Judah, one; the gate of Levi, one.}%
\verse{And on the east side, four thousand five hundred cubits,\lebnote{That is, 1.5 miles}and three gates, and so the gate of Joseph, one; the gate of Benjamin, one; the gate of Dan, one.}%
\verse{And on the south side, four thousand five hundred cubits\lebnote{That is, 1.5 miles}by measurement, and so three gates, the gate of Simeon, one; the gate of Issachar, one; the gate of Zebulun, one.}%
\verse{And on the west\lebnote{Or “sea”}side, four thousand five hundred cubits,\lebnote{That is, 1.5 miles}and their gates three, the gate of Gad, one; the gate of Asher, one; the gate of Naphtali, one.}%
\verse{All around the city is eighteen thousand cubits,\lebnote{That is, 6 miles}and the name of the city from that day is “Adonai Is There”!}%
\end{biblechapter}

