\biblebook{Ezra}

\begin{biblechapter} % Ezra 1
\verseWithHeading{The Decree of Cyrus} In the first year of Cyrus king of Persia, in order to accomplish the word of Adonai by the mouth of Jeremiah, Adonai stirred the spirit of Cyrus king of Persia and he sent a message to all of his kingdom and also put the message in writing:
\verse “Thus says Cyrus king of Persia: Adonai, the God of the heavens, has given me all the kingdoms of the earth. And he himself has appointed me to build a house for him in Jerusalem, which is in Judah.
\verse Whoever among you who is from all of his people, may his God be with him and may he go up to Jerusalem, which is in Judah, and may he build the house of Adonai, the God of Israel. He is the God who is in Jerusalem.
\verse And let every survivor, from wherever he \textit{resides} be assisted by the men of that place with silver and gold, with possessions and domestic animals, and with the freewill offering for the house of God which is in Jerusalem.”
\verse So the heads of the \textit{families} for Judah and Benjamin, and the priests and the Levites—to all whose spirit God had stirred—prepared to go up and build the house of Adonai which is in Jerusalem.
\verse And all of their neighbors \textit{helped them} with objects of silver, gold, possessions, domestic animals, and with valuable gifts—besides all of the freewill offering.
\verse And Cyrus the king brought out the objects of the house of Adonai that Nebuchadnezzar had brought from Jerusalem and placed in the house of his gods.
\verse Cyrus the king of Persia let them go out by the hand of Mithredath the treasurer, and he counted them out to Sheshbazzar the prince of Judah.
\verse Now these were the inventories: thirty gold metal dishes, one thousand silver metal dishes, twenty-nine vessels,
\verse thirty bowls of gold, four hundred and ten matching silver metal bowls, and one thousand other objects.
\verse All of the objects of gold and silver metal were five thousand four hundred. All this Sheshbazzar brought up along with the exiles from Babylonia to Jerusalem.
\end{biblechapter}

\begin{biblechapter} % Ezra 2
\verseWithHeading{The List of Returned Exiles} Now these were the people of the province who came up from the captivity of the exile whom Nebuchadnezzar the king of Babylon carried captive to Babylonia. They returned to Jerusalem and Judah, each to his own city.
\verse They came with Zerubbabel, Jeshua, Nehemiah, Seraiah, Reelaiah, Mordecai, Bilshan, Mispar, Bigvai, Rehum, and Baanah.
\verse The number of the men of the people of Israel:
\verse the descendants of Parosh, two thousand one hundred and seventy-two;
\verse the descendants of Shephatiah, three hundred and seventy-two;
\verse the descendants of Arah, seven hundred and seventy-five;
\verse the descendants of Pahath-moab, particularly the descendants of Jeshua and Joab, two thousand eight hundred and twelve;
\verse the descendants of Elam, one thousand two hundred and fifty-four;
\verse the descendants of Zattu, nine hundred and forty-five;
\verse the descendants of Zaccai, seven hundred and sixty;
\verse the descendants of Bani, six hundred and forty-two;
\verse the descendants of Bebai, six hundred and twenty-three;
\verse the descendants of Azgad, one thousand two hundred and twenty-two;
\verse the descendants of Adonikam, six hundred and sixty-six;
\verse the descendants of Bigvai, two thousand fifty-six;
\verse the descendants of Adin, four hundred and fifty-four;
\verse the descendants of Ater, particularly of Hezekiah, ninety-eight;
\verse the descendants of Bezai, three hundred and twenty-three;
\verse the descendants of Jorah, one hundred and twelve;
\verse the descendants of Hashum, two hundred and twenty-three;
\verse the descendants of Gibbar, ninety-five;
\verse the people of Bethlehem, one hundred and twenty-three;
\verse the men of Netophah, fifty-six;
\verse the men of Anathoth, one hundred and twenty-eight;
\verse the people of Azmaveth, forty-two;
\verse the people of Kiriath Αrim, Kephirah, and Beeroth, seven hundred and forty-three;
\verse the people of Ramah and Geba, six hundred and twenty-one;
\verse the men of Micmash, one hundred and twenty-two;
\verse the men of Bethel and Ai, two hundred and twenty-three;
\verse the people of Nebo, fifty-two;
\verse the people of Magbish, one hundred and fifty-six;
\verse the people of the other Elam, one thousand two hundred and fifty-four;
\verse the people of Harim, three hundred and twenty;
\verse the people of Lod, Hadid, and Ono, seven hundred and twenty-five;
\verse the people of Jericho, three hundred and forty-five;
\verse The priests: the descendants of Jedaiah, of the house of Jeshua, nine hundred and seventy-three;
\verse the descendants of Immer, one thousand fifty-two;
\verse the descendants of Pashhur, one thousand two hundred and forty-seven;
\verse the descendants of Harim, one thousand seventeen.
\verse The Levites: the descendants of Jeshua and Kadmiel, of the descendants of Hodaviah, seventy-four.
\verse The singers: the descendants of Asaph, one hundred and twenty-eight.
\verse The descendants of the gatekeepers: the descendants of Shallum, the descendants of Ater, the descendants of Talmon, the descendants of Akkub, the descendants of Hatita, and the descendants of Shobai; in total one hundred and thirty-nine.
\verse The temple servants: the descendants of Ziha, the descendants of Hasupha, the descendants of Tabbaoth,
\verse the descendants of Keros, the descendants of Siaha, the descendants of Padon,
\verse the descendants of Lebanah, the descendants of Hagabah, the descendants of Akkub,
\verse the descendants of Hagab, the descendants of Shamlai, the descendants of Hanan,
\verse the descendants of Giddel, the descendants of Gahar, the descendants of Reaiah,
\verse the descendants of Rezin, the descendants of Nekoda, the descendants of Gazzam,
\verse the descendants of Uzza, the descendants of Paseah, the descendants of Besai,
\verse the descendants of Asnah, the descendants of Meunim, the descendants of Nephisim,
\verse the descendants of Bakbuk, the descendants of Hakupha, the descendants of Harhur,
\verse the descendants of Bazluth, the descendants of Mehida, the descendants of Harsha,
\verse the descendants of Barkos, the descendants of Sisera, the descendants of Temah,
\verse the descendants of Neziah, and the descendants of Hatipha.
\verse The descendants of Solomon’s servants: the descendants of Sotai, the descendants of Hassophereth, the descendants of Peruda,
\verse the descendants of Jaalah, the descendants of Darkon, the descendants of Giddel,
\verse the descendants of Shephatiah, the descendants of Hattil, the descendants of Pochereth-hazzebaim, and the descendants of Ami.
\verse All the temple servants and the descendants of Solomon’s servants were three hundred and ninety-two.
\verse Now these were the ones who came up from Tel-melah, Tel-harsha, Cherub, Addan, and Immer, but they were not able to prove their fathers’ household and their descent, if they were from Israel:
\verse the descendants of Delaiah, the descendants of Tobiah, and the descendants of Nekoda, six hundred and fifty-two.
\verse And from the descendants of the priests: the descendants of Habaiah, the descendants of Hakkoz, and the descendants of Barzillai (who took a wife from the daughters of Barzillai the Gileadite, and was called by their name).
\verse These sought their record in the genealogy records, but they were not found, and \textit{were excluded from the priesthood as unclean}.
\verse The governor said to them that they could not eat from the holy food of the sanctuary until there was present a priest for the Urim and Thummim.
\verse The whole assembly together was forty-two thousand three hundred and sixty,
\verse apart from their male and female servants, of whom were seven thousand three hundred and thirty-seven; and they had two hundred male and female singers.
\verse Their horses numbered seven hundred and thirty-six, their mules numbered two hundred and forty-five,
\verse their camels numbered four hundred and thirty-five, their donkeys numbered six thousand seven hundred and twenty.
\verseWithHeading{Gifts Given for Temple Work} When some of the heads of \textit{families} came to the house of Adonai that is in Jerusalem, they gave freewill offerings for the house of God to erect it on its place.
\verse According to their ability they gave to the treasury room for the work sixty-one thousand darics of gold, five thousand minas of silver, and one hundred priestly tunics.
\verse The priests, the Levites, and some of the people lived in Jerusalem, and the singers, the gatekeepers, and the temple servants lived in their cities, and all Israel in their cities.
\end{biblechapter}

\begin{biblechapter} % Ezra 3
\verseWithHeading{Worship in Jerusalem Restored} Now the seventh month drew near and the \textit{Israelites} were in the cities, so the people gathered as \textit{one} in Jerusalem.
\verse And Jeshua son of Jehozadak and his brothers the priests stood up, and Zerubbabel the son of Shealtiel and his brothers built the altar of the God of Israel, in order to offer burnt offerings on it, as it is written in the law of Moses, the man of God.
\verse They set up the altar on its foundations, because \textit{they were in terror} because of the peoples of the lands. And they offered burnt offerings on it for Adonai, burnt offerings for the morning and the evening.
\verse And they kept the feast of booths, as it is written, and offered burnt offerings day by day in number according to the ordinance \textit{as described for each day}.
\verse After this, they presented the daily sacrifice of burnt offerings, the offerings for the New Moon Festival, and for all of the appointed times consecrated for Adonai and for all who gave a freewill offering to Adonai.
\verse From the first day of the seventh month they began to offer burnt offerings to Adonai. But the temple of Adonai was not yet founded.
\verse So they gave money to the stone craftsmen and skilled craftsmen, and food, drink, and olive oil to the Sidonians and Tyrians, to bring cedar trees from Lebanon to the sea of Joppa, according to the authorization over them of Cyrus king of Persia.
\verseWithHeading{Foundation of the Temple is Laid} In the second year after their coming to the house of God at Jerusalem, in the second month, Zerubbabel son of Shealtiel and Jeshua son of Jehozadak began their work, and the remainder of their brothers the priests and the Levites and all who came from the captivity to Jerusalem. They appointed the Levites \textit{from twenty years and older} to direct the work of the house of Adonai.
\verse And Jeshua with his sons and brothers, and Kadmiel and his sons—the sons of Judah—together directed the workers in the house of God, along with the sons of Henadad and their sons and brothers the Levites.
\verse And the builders laid the foundation of the temple of Adonai, and the priests in their apparel with the trumpets, and the Levites, the sons of Asaph, with the cymbals, were positioned to praise Adonai, \textit{as described by} King David of Israel.
\verse And they \textit{sang responsively}, with praising and thanksgiving to Adonai saying, “For he is good, for his loyal love is everlasting for Israel.”
\verse And all of the people responded with a great shout of joyful acclaim in praise to Adonai because the house of Adonai was laid.
\verse But many of the elderly priests, Levites, and heads of the families who had seen the first temple, wept with a loud voice when they saw the foundation of this house, but many \textit{shouted aloud with joy}.
\end{biblechapter}

\begin{biblechapter} % Ezra 4
\verseWithHeading{Resistance to Rebuilding the House of God} Now the enemies of Judah and Benjamin heard that the \textit{returned exiles} were building a temple for Adonai the God of Israel.
\verse And they approached Zerubbabel and the heads of the \textit{families}, and they said to them, “Let us build with you. Like you, we seek your God and have been sacrificing to him from the days of Esarhaddon the king of Assyria who brought us up here.
\verse But Zerubbabel, Jeshua, and the rest of the heads of the \textit{families} of Israel said to them, “It is not for you but for us to build a house for our God. For we ourselves alone will build it for Adonai the God of Israel, just as Cyrus the king of Persia has commanded us.”
\verse Then the people of the land discouraged \textit{the people of Judah} and made them afraid to build
\verse and bribed officials against them to frustrate their plan for all the days of Cyrus king of Persia until the reign of Darius king of Persia.
\verseWithHeading{Resistance to Rebuilding Jerusalem} In the reign of Ahasuerus, at the beginning of his reign, they wrote an accusation against the inhabitants of Judah and Jerusalem.
\verse And in the days of Artaxerxes, Bishlam, Mithredath, Tabeel, and the rest of their colleagues wrote to Artaxerxes king of Persia. The letter was written in Aramaic and translated from Aramaic.
\verse Rehum the royal officer and Shimshai the scribe wrote a letter against Jerusalem to King Artaxerxes as follows
\verse (then Rehum the royal officer, Shimshai the scribe, and the rest of their associates, the judges, the envoys, the officials, the Persians, the Erechs, the Babylonians, the Susians (that is the Elamites)
\verse and the rest of the nations whom the great and noble Osnappar deported and settled in the cities of Samaria and the rest of the province Beyond the River) and now
\verse this is the copy of the letter which they sent to him:
\verse “To King Artaxerxes from your servants, the men of the province Beyond the River. And now,
\verse be it known to the king that the Jews who have come up from near you to us have gone to Jerusalem. They are building the rebellious and wicked city. They are finishing the wall and repairing the foundation.
\verse Now be it known to the king that if this city is built and the walls are finished, they will not pay tribute and toll, and the royal revenue will be reduced.
\verse Now since we eat the salt of the palace and the dishonor of the king is not proper for us to see, we send and make this known to the king,
\verse so that it may be investigated in the book of records of your ancestors. You will find in the book of records and learn that this city is a rebellious city, hurtful to kings and provinces, and they have rebelled in its midst from ancient times. Because of that this city was destroyed.
\verse The king sent a reply: “To Rehum the royal officer, Shimshai the scribe, and the rest of their associates who live in Samaria and the rest of the province Beyond the River, greetings. And now
\verse the letter that you sent to us has been translated and read before me.
\verse And \textit{I issued} a decree, and they searched and found that this city from ancient days revolted against kings, and rebellion and sedition has been made in it.
\verse Mighty kings have ruled over Jerusalem, governing all the province Beyond the River, to whom tribute, duty, and tax has been given.
\verse So now, issue forth a decree that these men stop and this city not be built, until a decree is issued from me.
\verse And be careful not to be negligent on this matter. Why should damage grow to hurt kings?”
\verse Then when a copy of the letter of King Artaxerxes was read before Rehum, Shimshai the scribe, and their associates they returned in a hurry to Jerusalem against the Jews and they stopped them by force and power.
\verse Then the work on the house of God in Jerusalem stopped, and was discontinued until the second year of the reign of Darius king of Persia.
\end{biblechapter}

\begin{biblechapter} % Ezra 5
\verseWithHeading{Temple Restoration Resumed} Now the prophets, Haggai and Zechariah the son of Iddo, prophesied to the Jews who were in \textit{Judah} and in Jerusalem, in the name of the God of Israel who was over them.
\verse Then Zerubbabel the son of Shealtiel and Jeshua the son of Jozadak set out and began to build the house of God that is in Jerusalem. And with them the prophets of God were helping them.
\verse At the same time Tattenai, governor of the province Beyond the River, and Shethar-bozenai, and their associates came to them and spoke to them thus: “Who issued you all a decree to build this house and to finish this structure?”
\verse Then we asked them this: “What are the names of the men who are building this building?”
\verse And the eye of their God was on the elders of the Jews, and they did not stop them until the report came to Darius, and then answer was received.
\verse The copy of the letter that Tattenai, the governor of the province Beyond the River, and Shethar-bozenai and his associates the envoys who were in the province Beyond the River sent to Darius the king.
\verse They sent to him the report \textit{in which was written as follows}: “To Darius the king, all peace.
\verse May it be known to the king that we went to the province of Judah, the house of the great God. It is being built with stone blocks and timber is being put in the walls. This work is being done with diligence and is making progress in their hands.
\verse Then we asked those elders and said this to them, ‘Who issued forth to you a decree to build this house and to finish this structure?’
\verse We also asked them their names to make them known to you, that we might write down the \textit{name of their leaders}.
\verse And this is the answer they returned to us: ‘We are servants of the God of heaven and earth, and we are building the house that was built formerly many years ago, which a great king of Israel had built and finished.
\verse But because our ancestors angered the God of heaven, he gave them into the hand of the Chaldean King Nebuchadnezzar of Babylon. He destroyed this house and carried away the people to Babylonia.
\verse But in the first year of King Cyrus of Babylon’s reign he issued forth a decree to build this house of God.
\verse Also, the gold and silver vessels of the house of God, which Nebuchadnezzar took away from the temple in Jerusalem and brought to the temple in Babylonia, King Cyrus removed them from the temple in Babylonia and they were given to Sheshbazzar, whom he appointed governor.
\verse He said to him, ‘Take these vessels. Go put them in the temple in Jerusalem and let the house of God be built on its site.’
\verse Then this Sheshbazzar came and laid the foundation of the house of God in Jerusalem. And from that time until now it is being built, but not yet finished.’
\verse And now, if it seems good for the king, let it be investigated in the house of the treasury of the king in Babylonia to see if a decree was issued forth from King Cyrus to build this house of God in Jerusalem. And let the king send to us his desire on this matter.”
\end{biblechapter}

\begin{biblechapter} % Ezra 6
\verseWithHeading{Darius’ Search for the Decree} Then King Darius issued forth a decree, and they searched the house of the treasury of scrolls being stored in Babylonia.
\verse But it was in Ecbatana in the province of Media, in the citadel, that a certain scroll had written on it, “A record.
\verse In the first year of King Cyrus, he issued forth a decree concerning the house of God in Jerusalem. Let the house be built, the place where sacrifices are offered and let its foundations be raised. Its height shall be sixty cubits and its width sixty cubits,
\verse with three layers of great stones and a layer of timber. Let the new expenses be paid from the house of the king.
\verse Also, let the gold and silver vessels of the house of God, which Nebuchadnezzar took from the temple in Jerusalem and brought to Babylonia, be returned and brought to the temple in Jerusalem to its place. Put them in the house of God.”
\verseWithHeading{The Decree of Darius} “Now then, Tattenai governor of the province Beyond the River, Shethar-bozenai, and your associates, the envoys who are in the province Beyond the River—keep far away from there.
\verse Leave this work of the house of God alone. Let the governor of the Jews and the elders of the Jews build this house of God on its site.
\verse And I issue forth a decree for what you should do for these elders of the Jews to build this house of God. The full expense will be paid to these men from the riches of the king from the taxes of the province Beyond the River, without delay.
\verse Whatever may be needed—\textit{young bulls}, young rams, sheep for burnt offerings to the God of heaven, wheat, salt, wine, and oil for the priests in Jerusalem—let it be given to them day by day with no negligence,
\verse that they may offer incense offerings to the God of heaven, and pray for the life of the king and his children.
\verse Furthermore, I issue forth a decree that if any person violates this decree, let a beam be pulled out from his house and let him be impaled on it. And let his house be made a pile of rubble on account of this.
\verse May the God who has set his name there overthrow any king or people who sets his hand to alter or to destroy this house of God in Jerusalem. I, Darius, issue forth a decree. Let it be done with diligence.”
\verseWithHeading{Temple is Dedicated and Completed} Then Tattenai the governor of the province Beyond the River, Shethar-bozenai, and their associates consequently did with diligence what Darius the king ordered.
\verse So the elders of the Jews were building and prospering, through the prophecy of Haggai the prophet and Zechariah son of Iddo. They finished building by the command of the God of Israel and by the decree of Cyrus, Darius, and King Artaxerxes of Persia.
\verse This house was completed on the third day of the month of Adar, which was in the sixth year of the reign of King Darius.
\verse And the \textit{Israelites}, the priests, the Levites, and the rest of the \textit{returned exiles} celebrated the dedication of this house of God with joy.
\verse And they offered during the dedication of this house of God one hundred young bulls, two hundred young rams, four hundred lambs, and twelve male goats as a sin offering for Israel according to the number of the tribes of Israel.
\verse Then they set the priests in their divisions and the Levites in their sections for the work of God in Jerusalem, as it is written in the book of Moses.
\verseWithHeading{The Celebration of Passover} On the fourteenth day of the first month the returned exiles observed the Passover feast.
\verse For the priests and Levites together had consecrated themselves; all of them were clean. And they slaughtered the Passover sacrifice for all of the returned exiles, for their brothers the priests, and for themselves.
\verse The \textit{Israelites} who returned from the exile and all those who separated themselves from the uncleanness of the nations of the earth to seek Adonai the God of Israel, ate.
\verse With joy they celebrated the festival of unleavened bread for seven days, because Adonai had made them joyful and had turned the heart of the king of Assyria toward them in order to help them with the work of their hands on the house of the God of Israel.
\end{biblechapter}

\begin{biblechapter} % Ezra 7
\verseWithHeading{Ezra’s Arrival and Work} After these things during the reign of King Artaxerxes of Persia, Ezra the son of Seraiah, son of Azariah, son of Hilkiah,
\verse son of Shallum, son of Zadok, son of Ahitub,
\verse son of Amariah, son of Azariah, son of Meraioth,
\verse , son of Zerahiah, son of Uzzi, son of Bukki,
\verse son of Abishua, son of Phinehas, son of Eleazar, son of Aaron the chief priest—
\verse this Ezra went up from Babylonia. He was a scribe skilled in the law of Moses, which Adonai the God of Israel gave. The king granted him all his requests, for the hand of Adonai was upon him.
\verse Some of the \textit{Israelites}, some priests and Levites, the singers, the gatekeepers, and the temple servants went up to Jerusalem in the seventh year of King Artaxerxes.
\verse He came to Jerusalem in the fifth month, which was in the seventh year of the king.
\verse For on the first day of the first month he began the journey from Babylonia; and on the first day of the fifth month he came to Jerusalem, for the good hand of his God was on him.
\verse For Ezra had set his heart to seek the law of Adonai, to do it, and to teach the regulations and judgments in Israel.
\verseWithHeading{Artaxerxes’ Letter to Ezra} This is the copy of the letter that King Artaxerxes gave to Ezra the priest, the scribe of the scroll of the words of the commands of Adonai and his statutes for Israel:
\verse “Artaxerxes, the king of kings, to Ezra the priest, the scribe of the law of the God of the heavens. Peace. And now
\verse I issue forth a decree that any of the people of Israel, their priests, or their Levites in my kingdom who are willing to go to Jerusalem may go with you.
\verse For you are sent from the king and his seven counselors to enquire about Judah and Jerusalem concerning the law of your God, which is in your hand.
\verse Also, bring the silver and gold that the king and his advisors have freely offered to the God of Israel whose dwelling is in Jerusalem,
\verse and all of the silver and gold that you will find in the whole province of Babylonia, with the freewill offerings of the people and the priests giving willingly for the house of their God in Jerusalem.
\verse With this money, then, with diligence you must buy young bulls, young rams, lambs, and their offerings and libations. You must offer them on the altar that is in the house of your God in Jerusalem.
\verse You may do whatever seems best to you and your brothers to do with the remainder of the silver and gold according to the desire of your God.
\verse And the vessels that were given to you for the service of the house of your God you shall deliver before the God of Jerusalem.
\verse And the remainder of the needs for the house of your God that falls to you to provide, you may provide from the house of the king’s treasury.”
\verse “I, even I, King Artaxerxes, issue forth a decree to all the treasurers who are in the province Beyond the River. Whatever the priest Ezra, the scribe of the law of the God of heaven, may ask of you, let it be done with diligence,
\verse up to one hundred talents of silver, one hundred measures of wheat, one hundred baths of wine, one hundred baths of oil, and \textit{unlimited salt}.
\verse All that is commanded by the God of heaven, let it be done with diligence for the house of the God of heaven, otherwise wrath will come on the kingdom of the king and his sons.
\verse You also should be aware that it is not permitted to place tax, tribute, or duty upon all of the priests, the Levites, the singers, the doorkeepers, the temple servants, and other servants of the house of God.
\verse “You, Ezra, according to the wisdom of your God that you possess, appoint magistrates and judges who can judge all of the people in the province Beyond the River who know the laws of your God. And you will teach those who do not know.
\verse All who do not obey the law of your God and the law of the king, let judgment be executed on him with diligence, whether for death or for banishment or for confiscation of goods and for imprisonment.”
\verse Blessed be Adonai the God of our ancestors, who put this in the heart of the king to glorify the house of Adonai in Jerusalem
\verse and who extended to me loyal love before the king and his counselors, and before all of the mighty officials of the king. I took courage, for the hand of Adonai my God was upon me, and I gathered leaders from Israel to go up with me.
\end{biblechapter}

\begin{biblechapter} % Ezra 8
\verseWithHeading{Family Heads Who Returned With Ezra} These were the heads of their \textit{families} and the register of those returnees from Babylonia who came up with me in the reign of King Artaxerxes:
\verse From the descendants of Phinehas: Gershom. From the descendants of Ithamar: Daniel. From the descendants of David: Hattush.
\verse From the descendants of Shecaniah, from the descendants of Parosh: Zechariah and with him one hundred and fifty registered males.
\verse From the descendants of Pahath-Moab: Eliehoenai son of Zerahiah and with him two hundred males.
\verse From the descendants of Zattu: Shecaniah son of Jahaziel, and with him three hundred males.
\verse From the descendants of Adin: Ebed son of Jehonathan, and with him fifty males.
\verse From the descendants of Elam: Jeshaiah son of Athaliah, and with him seventy males.
\verse From the descendants of Shephatiah: Zebadiah son of Michael, and with him eighty males.
\verse From the descendants of Joab: Obadiah son of Jehiel, and with him two hundred and eighteen males.
\verse From the descendants of Bani: Shelomith son of Josiphiah, and with him one hundred and sixty males.
\verse From the descendants of Bebai: Zechariah son of Bebai, and with him twenty-eight males.
\verse From the descendants of Azgad: Jehohanan son of Haqqatan, and with him one hundred and ten males.
\verse From the descendants of Adonikam, those who came last, these were their names: Eliphelet, Jeiel, Shemaiah, and with them sixty males.
\verse From the descendants of Bigvai: Uthai son of Zabud, and with him seventy males.
\verseWithHeading{Servants for the Temple are Gathered} I gathered them by the river that goes to Ahava and we camped there for three days. And I looked at the people and the priests, I found no one there from the sons of Levi.
\verse And I sent for Eliezer, Ariel, Shemaiah, Elnathan, Jarib, Elnathan, Nathan, Zechariah, and Meshullam, who were family heads, and for Jehoiarib and Elnathan who were wise.
\verse I sent them to Iddo, the head of the place called Casiphia, and I placed in their mouths what words to say to Iddo and his brothers the temple servants in Casiphia, to send to us ministers for the house of our God.
\verse With the good hand of our God before us, they brought us a man of understanding from the sons of Mahli, son of Levi, son of Israel—Sherebiah and his sons and brothers, eighteen in total.
\verse Also Hashabiah and with him Jeshaiah, the sons of Merari, with his brothers and their sons, twenty in total.
\verse And from the temple servants, whom David and his officials had set up to serve the Levites, two hundred and twenty in total. All of them were registered by name.
\verseWithHeading{Ezra Proclaims a Fast} I proclaimed a fast there at the river Ahava to humble ourselves before our God in order to seek from him a safe journey for us, our children, and our possessions.
\verse For I was ashamed to ask the king for troops and horses to protect us from enemies on the way because we said to the king, “the hand of our God is favorable to all who seek him, but his strength and anger are against all who forsake them.”
\verse So we fasted and sought our God for this and he responded to our prayer.
\verseWithHeading{The Priests’ Gift for the Temple} I set apart twelve of the official priests: Sherebiah, Hashabiah, and ten of their brothers with them.
\verse Then I weighed out to them silver and gold and the vessels, the offering of the house of our God that the king, his counselors, his commanders, and all of Israel who was present offered.
\verse I weighed out into their hand six hundred and fifty talents of silver, one hundred vessels of silver, one hundred talents of gold,
\verse twenty gold bowls worth one thousand darics, and two vessels of good polished bronze as precious as gold.
\verse And I said to them, “You are holy to Adonai and the vessels are holy. The silver and gold are a freewill offering to Adonai, God of your ancestors.
\verse Guard and look after them until you weigh them before the chief priests, the Levites, and the heads of the \textit{families} of Israel in Jerusalem, in the chamber of the house of Adonai.”
\verse So the priests and Levites took over the weighed silver, gold, and the vessels to bring to Jerusalem for the house of our God.
\verseWithHeading{Return to Jerusalem} On the twelfth day of the first month we set out from the river of Ahava to go to Jerusalem. The hand of our God was on us and he delivered us from the hand of the enemy and from ambushing on the way.
\verse We came to Jerusalem and remained there three days.
\verse On the fourth day, the silver, gold, and vessels were weighed in the house of our God into the hand of priest Meremoth, the son of Uriah, and with him Eleazar son of Phinehas. With them were the Levites, Jehozabad son of Jeshua and Noadiah son of Binnui.
\verse The amount and weight of all the weighed items was recorded at that time.
\verse The exiles that returned from captivity offered burnt offerings to the God of Israel; twelve bulls for all Israel, ninety-six rams, seventy-seven lambs, and twelve goats as a sin offering. All was a burnt offering to Adonai.
\verse They delivered the customs of the king to the king’s satraps and to the governor of the province Beyond the River. And they supported the people of the house of God.
\end{biblechapter}

\begin{biblechapter} % Ezra 9
\verseWithHeading{Disdain Over Mixed Marriages} After finishing these things the officials approached me saying, “The people of Israel, the priests, and the Levites have not separated from the people of the lands with their detestable things, and from the Canaanites, Hittites, Perizzites, Jebusites, Ammonites, Moabites, Egyptians, and Amorites.
\verse For they have taken from their daughters for themselves and their sons. So the holy seed has mixed itself with the peoples of the lands, and the hand of the officials and prefects were foremost in this sin.
\verse When I heard this I tore my garment and my robe, and I pulled the hair out from my head and beard, and I sat appalled.
\verse Then all who trembled at the words of the God of Israel because of the sin of the returned exiles were gathered around me and I sat appalled until the evening offering.
\verseWithHeading{Ezra Prays} At the evening offering I got up from my mourning posture and, with my garment and robe torn, I fell down on my knees and I spread out my palms to Adonai my God
\verse and said, “My God, I am ashamed and embarrassed to lift my face to you, my God. For our sins have risen above our heads and our guilt has grown up to the heavens.
\verse From the days of our ancestors until this day we have been in great guilt, and because of our sins we ourselves, our kings, and our priests have been handed into the hand of the kings of the lands, to the sword, to captivity, to plunder, and to utter shame, as it is this day.
\verse But now for a brief moment mercy has been shown by Adonai our God, who left behind for us a remnant, and given us security in his holy place—for our God to brighten our eyes and to give us brief relief in our bondage.
\verse For we are slaves and in our bondage our God did not forsake us, and he has extended to us loyal love in the presence of the kings of Persia, to give to us deliverance and to set up the house of our God, to repair its ruins, and to give us a wall in Judea and Jerusalem.
\verse “And now our God what shall we say after this? For we have forsaken your commandments,
\verse which you commanded by the hand of your servants the prophets saying, ‘The land that you are entering to possess is a land of impurity with the impurity of the peoples of the lands, with their detestable things which they have filled from end to end with their uncleanness.
\verse Therefore, do not give your daughters to their sons, and do not take their daughters for your sons. Do not seek their peace and prosperity so that forever you may be strong and eat the good of the land and may give it as an inheritance to your sons.’
\verse After all that has come upon us for our evil practices and for our great guilt—you, our God, have held back less than our guilt deserved and you have given us a remainder such as this.
\verse Shall we again break your commandments and intermarry with the peoples who practice these detestable things? Would you not be angry with us until you destroy us with no remnant or remainder?
\verse Adonai, God of Israel, you are righteous, for we have been left this day as a remnant. Here we are before you in our guilt, for none can stand before you because of this.”
\end{biblechapter}

\begin{biblechapter} % Ezra 10
\verseWithHeading{The People Respond} While Ezra was praying and confessing, weeping and throwing himself down before the house of God, a very great assembly of men, women, and children from Israel gathered to him. And the people wept bitterly.
\verse Shecaniah son of Jehiel, of the sons of Elam, answered and said to Ezra, “We ourselves have broken faith with our God and have married foreign women from the peoples of the land, and even now there is hope for Israel in spite of this.
\verse So now let us make a covenant with our God to send away all of these women and their offspring, according to the advice of my lord and those who tremble at the commandment of our God. Let it be done according to the law.
\verse Arise, for it is your task and we are with you. Be strong and do it.”
\verse Then Ezra stood up and made the chief priests, Levites, and all of Israel swear to do according to what was said. So they swore this oath.
\verseWithHeading{Foreign Wives and Children are Sent Away} Ezra rose from before the house of God and went to the chamber of Jehohanan son of Eliashib. He did not eat food nor drink water because he was mourning over the sin of the exiles.
\verse And they sent a proclamation through Judah and Jerusalem to all the \textit{returned exiles} to assemble in Jerusalem.
\verse Anyone who did not come within three days, by decision of the officials and elders, all of his possessions would be devoted to God, and he himself would be excluded from the assembly of exiles.
\verse Then all the men of Judah and Benjamin assembled in Jerusalem within the three days. It was the ninth month on the twentieth day of the month. All the people sat in the public square of the house of God trembling because of this matter and from the rains.
\verse Then Ezra the priest stood up and said to them, “You have broke faith and married foreign women, increasing the guilt of Israel.
\verse Now make a confession to Adonai the God of your ancestors and do his will. Separate yourselves from the peoples of the land and from the foreign women.”
\verse Then all the assembly answered with a great voice and said, “It is so. We must do according to your words.
\verse But the people are many, and it is the time of rain; we are unable to stand outside. The task is not for one day or two, for we greatly rebelled in this matter.
\verse Please let our leaders stand for the whole assembly, and let all that are in our cities who have married foreign women come at set times, and with them the elders of and judges of each city, until the fierce wrath of our God because of this matter is averted from us.”
\verse Only Jehonathan son of Asahel and Jahzeiah son of Tikvah stood against this, and the Levites Meshullam and Shabbethai supported them.
\verse The \textit{returned exiles} did so. Ezra the priest selected men, the heads of the \textit{families} according to the house of their fathers, all of them by name. They sat down to examine the matter on the first day of the tenth month.
\verse They finished investigating all the men who married foreign women by the first day of the first month.
\verse There was found from the sons of the priests those who had married foreign women, from the sons of Jeshua son of Jehozadak and his brothers: Maaseiah, Eliezer, Jarib, and Gedaliah.
\verse \textit{They pledged themselves} to put away their wives, and their guilt offering was a ram of the flock for their guilt.
\verse From the sons of Immer: Hanani and Zebadiah.
\verse From the sons of Harim: Maaseiah, Elijah, Shemaiah, Jehiel, and Uzziah.
\verse From the sons of Pashhur: Elioenai, Maaseiah, Ishmael, Nethanel, Jozabad, and Eleasah.
\verse From the Levites: Jozabad, Shimei, Kelaiah (that is, Kelita), Pethahiah, Judah, and Eliezer.
\verse From the singers: Eliashib. From the gatekeepers: Shallum, Telem, and Uri.
\verse And from Israel, the sons of Parosh: Ramiah, Izziah, Malkijah, Mijamin, Eleazar, Malkijah, and Benaiah.
\verse From the sons of Elam: Mattaniah, Zechariah, Jehiel, Abdi, Jeremoth, and Elijah.
\verse From the sons of Zattu: Elioenai, Eliashib, Mattaniah, Jeremoth, Zabad, and Aziza.
\verse From the sons of Bebai: Jehohanan, Hananiah, Zabbai, and Athlai.
\verse From the sons of Bani: Meshullam, Malluch, Adaiah, Jashub, Sheal, and Jeremoth.
\verse From the sons of Pahath-Moab: Adna, Kelal, Benaiah, Maaseiah, Mattaniah, Bezalel, Binnui, and Manasseh.
\verse From the sons of Harim: Eliezer, Isshiah, Malkijah, Shemaiah, Simeon,
\verse Benjamin, Malluch, and Shemariah.
\verse From the sons of Hashum: Mattenai, Mattattah, Zabad, Eliphelet, Jeremai, Manasseh, and Shimei.
\verse From the sons of Bani: Maadai, Amram, Uel,
\verse Benaiah, Bedeiah, Keluhi,
\verse Vaniah, Meremoth, Eliashib,
\verse Mattaniah, Mattenai, Jaasu,
\verse Bani, Binnui, Shimei,
\verse Shelemiah, Nathan, Adaiah,
\verse Macnadebai, Shashai, Sharai,
\verse Azarel, Shelemiah, Shemariah,
\verse Shallum, Amariah, and Joseph.
\verse From the sons of Nebo: Jeiel, Mattithiah, Zabad, Zebina, Iddo, Joel, and Benaiah.
\verse All of these had married foreign wives, and some from among the wives bore children.
\end{biblechapter}

