\biblebook{Hosea}

\begin{biblechapter} % Hosea 1
\verseWithHeading{God Commands Hosea to Start a Family}{The word of Adonai that came to Hosea son of Beeri, in the days of Uzziah, Jotham, Ahaz, and Hezekiah, the kings of Judah, and in the days of Jeroboam the son of Joash, king of Israel.}%
\verse{\textit{At the beginning when Adonai spoke}\lebnote{Literally “At the beginning of the talking of Adonai”}through Hosea, 
Adonai said to Hosea, 
“Go, take for yourself a wife and children of whoredom, 
because the land commits great whoredom 
\textit{forsaking Adonai}.”\lebnote{Literally “from behind Adonai”}}%
\verse{So he went and took Gomer daughter of Diblaim, 
and she conceived and bore him a son.}%
\verse{And Adonai said to him, 
“\textit{Name him}\lnPDB{}Jezreel;\lnPDC{}
because in a little while \textit{I will punish}\lebnote{Literally “I will visit sin upon”}the house of Jehu\lebnote{Or “the dynasty of Jehu”}
for the blood\lebnote{Or “bloodshed”}of Jezreel,\lnPDC{}
and I will put an end to the kingdom of the house of Israel.\lebnote{Or “the dynasty of Israel”}}%
\verse{\textit{On that day}\lebnote{Literally “And it will happen on that day”}
I will break the bow of Israel 
in the Valley of Jezreel.”}%
\verse{And she conceived again and bore a daughter, and he\lnAHZ{}said to him,}%
\verse{“\textit{Name her}\lebnote{Literally “Call her name”}Lo-ruhamah,\lnPDF{}
for I will no longer have pity 
on the house of Israel 
or \textit{forgive them}.\lebnote{Literally “raise them up”}}%
\verse{But I will have pity on the house of Judah 
and I will save them by Adonai their God, 
and will not deliver them 
by bow, sword, war, 
horses, or horsemen.}%
\verse{And when she had weaned Lo-ruhamah,\lnPDF{}she conceived and bore a son.}%
\verse{And he\lnAHZ{}said, “\textit{Name him}\lnPDB{}Lo-ammi,\lebnote{Lo-ammi means “not my people”}
for you are not my people 
and I am not your God.\lebnote{Hebrew “not yours”}}%
\verse{\lebnote{Hosea 1:10–2:23 in the English Bible is 2:1–25 in the Hebrew Bible}The number of the children of Israel 
will be like the sand of the sea 
that cannot be measured or counted; 
and in the place where it is said to them, 
“You are not my people,” 
it will be said to them, 
“Children of the living God.”}%
\end{biblechapter}

\begin{biblechapter} % Hosea 2
\verseWithHeading{The Unfaithful Wife}{Say to your brother,\lnPDK{}“Ammi,”\lebnote{Ammi means “My people”}
and to your sister,\lebnote{Hebrew “sisters”}“Ruhamah.”\lebnote{Ruhamah means “Pitied”}}%
\verse{Plead with your mother, plead— 
because she is not my wife, 
and I am not her husband. 
Let her put away her whoring \textit{from before her},\lebnote{Literally “from her face”}
and her adultery from between her breasts.}%
\verse{Lest I strip her naked 
and expose her like the day when she was born; 
I will make her like the desert 
and turn her into parched land; 
I will kill her with thirst.}%
\verse{And I will not have pity on her children, 
because they are children of whoredom.}%
\verse{Because their mother was unfaithful;\lebnote{Or “their mother played the whore”}
she who conceived them has acted shamefully, 
for she said: 
“I will go after my lovers, 
the ones who give me my bread, my water, 
my wool, and my flax, 
my oil, and my drink.”}%
\verse{Therefore, Look! 
I am going to hedge her path\lebnote{Hebrew “your path”}\textit{with thorns},\lebnote{Literally “with the thorns”}
and I will build a stone wall, a stone wall against her, 
and she will not find her paths.}%
\verse{Then she will pursue \textit{her lovers},\lebnote{Literally “the ones who love her”}
but she will not overtake them; 
she will seek them and not find them; 
and she will say, “I will go 
and return to my first husband 
because it was better for me then than now.”}%
\verse{But she did not know 
that it was I who gave her 
grain, new wine, and oil, 
and who gave her silver in abundance, 
and gold which they made into a Baal.}%
\verse{Therefore I will take again 
my grain in its time, 
and my wine in its season; 
and I will take away my wool and my flax, 
which were to cover her nakedness.}%
\verse{And now I will uncover her nakedness 
before the eyes of her lovers, 
and no one will rescue her from my hand.}%
\verse{And I will put an end to all her mirth, 
her festivals,\lnPDL{}her new moons,\lebnote{Hebrew “new moon”}and her Sabbaths,\lebnote{Hebrew “Sabbath”}
and all her appointed festivals;\lnPDL{}}%
\verse{and I will lay waste to her vines\lnDYD{}and her fig trees,\lnERM{}
of which she said, 
“They are my payment for prostitution, 
which my lovers gave to me.” 
I will make them a forest, 
and the wild animals\lnASW{}of the field will devour them.}%
\verse{\textit{I will punish her}\lebnote{Literally “To visit sins upon her;” see 1:4}for the days of the Baals, 
to whom she burns incense, 
and she decked herself with her ornamental ring and jewelry, 
and she went after her lovers, and forgot me 
—a declaration of Adonai.}%
\verse{Therefore, Look! I am going to allure her 
and bring her into the desert, 
and \textit{I will speak tenderly to her}.\lebnote{Literally “I will speak to her heart”}}%
\verse{From there I will give her 
her vineyards, 
and the Valley of Achor as a doorway of hope. 
And there she will respond, 
as in the days of her youth, 
just as in the day of her coming out of 
the land of Egypt.}%
\verse{\textit{And on that day}—\lebnote{Literally, “And it will happen on that day”}
a declaration of Adonai— 
you will call me, “My husband;”\lebnote{Hebrew “Ishi”}
you will no longer call me, “My Baal.”\lebnote{Hebrew “Baali”}}%
\verse{I will remove the names 
of the Baals from her mouth, 
and they will no longer be mentioned by their name.}%
\verse{I will make a covenant for them on that day, 
with the animals\lnASW{}of the field, 
with the birds\lnJPI{}of the heaven, and 
with the creeping things\lebnote{Hebrew “creeping thing”}of the ground; 
the bow, the sword, and the war I will abolish from the land, and I will let them lie down in safety.}%
\verse{And I will take you as my wife forever; 
I will take you as a wife for myself in righteousness and in justice, 
in steadfast love and in mercy.}%
\verse{\textit{I will take you as my wife}\lebnote{Literally “I will take you as a wife to me”}in faithfulness, 
and you will know Adonai.}%
\verse{On that day I will answer, \textit{declares Adonai},\lebnote{Literally “a declaration of Adonai}
I will answer the heavens, 
and they will answer the earth.}%
\verse{And the earth will answer 
the grain, the wine, and the oil, 
and they will answer Jezreel.\lnPDC{}}%
\verse{I will sow her for myself in the land; 
I will have pity on Lo-ruhama;\lebnote{Lo-ruhama means “Not pitied”}
I will say to Lo-ammi,\lebnote{Lo-ammi means “Not my people”}
“You are my people,” 
and he himself will say, “you are my God.”}%
\end{biblechapter}

\begin{biblechapter} % Hosea 3
\verseWithHeading{Another Symbolic Marriage}{And Adonai said to me again, 
“Go, love a woman 
\textit{who has a lover}\lebnote{Literally “who is beloved of a friend”}and is committing adultery, 
just like the love of Adonai for the children of Israel, 
but they are turning to other gods 
and love raisin cakes.”}%
\verse{So I bought her for fifteen shekels of silver 
and a homer of barely and a measure of wine.\lebnote{Hebrew “a measure of barley”; the LXX has “a nebel of wine,” although some modern translations, e.g. NASB, have “a homer and a half of barley”}}%
\verse{And I said to her, 
“\textit{You must remain as mine}\lebnote{Literally “You must sit at home for me”}for many days, 
and you will not play the whore; 
\textit{you will not belong}\lebnote{Literally “you will not be”}to a man, 
and I will belong to you.”}%
\verse{The children of Israel will remain for many days 
without a king and prince, 
without sacrifice and stone pillar, ephod and teraphim.}%
\verse{Afterward the children of Israel will return 
and seek Adonai their God and David their king. 
They will come in fear to Adonai 
and to his goodness at the end of days.}%
\end{biblechapter}

\begin{biblechapter} % Hosea 4
\verseWithHeading{Adonai Has an Indictment}{Hear the word of Adonai, O children of Israel, 
for Adonai has an indictment against the inhabitants of the land. 
There is no faithfulness or loyalty, 
there is no knowledge of God in the land.}%
\verse{Swearing and lying, 
murdering and stealing, 
and adultery break out,\lebnote{The LXX adds “and adultery breaks out in the land”}
and bloodshed follows bloodshed.}%
\verse{Therefore the land mourns, 
and all those living in it languish 
with the animals of the field, 
and the birds of the heaven, 
and even the fish of the sea 
are being swept away.}%
\verse{Yet, let no one contend, 
and let no one accuse, 
for with you, O priest, is my contention.\lebnote{The Hebrew is unclear; the NASB translates “For your people are like those who contend with the priest”}}%
\verse{You will stumble by day, 
and the priest also will stumble 
with you at night. 
And I will destroy your mother;}%
\verse{My people are destroyed 
for lack of knowledge; 
because you have rejected knowledge, 
I reject you from acting as a priest for me.\lebnote{Or “to me”}
And since you have forgotten the law of your God, 
I will also forget your children.\lnJWK{}}%
\verse{When they became many, their sin increased against me; 
they changed\lebnote{Hebrew uncertain; MT “I will exchange”}their glory into shame.}%
\verse{\textit{They feed on the sin of my people},\lebnote{Literally “They eat the sins of my people”}
\textit{they are greedy}\lebnote{Literally “they lift up their throat”}for their iniquity.}%
\verse{And it will be like people, like priest; 
I will punish them\lnDUT{}for their\lnDMD{}ways, 
and I will requite them\lnDUT{}for their\lnDMD{}deeds.}%
\verse{They will eat and not be satisfied; 
they will play the whore and not multiply, 
because they have forsaken Adonai \textit{to devote themselves to}%
\verse{whoredom}.\lebnote{Literally “to watch over whoredom”}
Wine and new wine \textit{take away the understanding}.\lebnote{Literally “take away heart”}}%
\verse{My people, 
consult \textit{their wooden idols},\lebnote{Literally “their wood”}
and their divining rod gives them oracles; 
because a spirit of whoredom led them astray, 
and they played the whore from under their God.\lebnote{That is, “they played the whore forsaking their God”; See Ezek 23:5}}%
\verse{On the tops of the mountains they sacrifice, 
and on the hills they make offerings, 
under oak, poplar, and terebinth, 
because their\lnDTS{}shade is good. 
Therefore your daughters play the whore, 
and all your daughters-in-law\lnPDZ{}commit adultery.}%
\verse{\textit{I will not punish your daughters}\lebnote{Literally “I will not visit sins upon your daughters”}when they play the whore, 
or your daughters-in-law\lnPDZ{}when they commit adultery; 
because they go aside with whores, 
and they sacrifice with temple prostitutes, 
a people that does not understand comes to ruin.}%
\verse{Even though you, O Israel, are playing the whore, 
do not let Judah become guilty; 
do not enter Gilgal, 
or go up to Beth-aven; 
and do not swear, 
“\textit{As Adonai lives}!”\lnGNS{}}%
\verse{Indeed, like a stubborn heifer, 
Israel is stubborn; 
Now Adonai will feed them, 
like a lamb in broad pasture.\lebnote{Or “Can Adonai feed them, like a lamb in broad pasture?”}}%
\verse{Ephraim is joined to idols; 
let him alone.}%
\verse{When their drinking has ended, 
they surely indulge in sexual orgies. 
They love lewdness more than their\lnPEC{}glory.}%
\verse{A wind has wrapped them\lnPEC{}in its wings, 
and they will be ashamed because of their altars.}%
\end{biblechapter}

\begin{biblechapter} % Hosea 5
\verseWithHeading{The Priests Are a Snare}{Hear this, O priests! 
Give heed, O house of Israel! 
Listen, O house of the king! 
Because the judgment applies to you; 
because you have been a snare for Mizpah, 
and a net spread out on Tabor;}%
\verse{they dug a deep pit\lnNRZ{}in Shittim, 
but I am a punishment for all of them.}%
\verse{I myself know Ephraim, 
and Israel is not hidden from me; 
because now you have played the whore, O Ephraim— 
Israel is defiled.}%
\verse{Their deeds do not permit them 
to return them to God. 
Because a spirit of whoredom is in their midst, 
they do not know Adonai.}%
\verse{The pride of Israel testifies \textit{against him},\lebnote{Literally “against his face;” see 7:10}
and Israel and Ephraim stumble in their guilt, 
and Judah stumbles with them.}%
\verse{With their flocks\lnLHA{}and herds\lnLIR{}they will go 
to seek Adonai, but they will not find him; 
he has withdrawn from them.}%
\verse{They have dealt faithlessly with Adonai 
because they have borne illegitimate children. 
Now the new moon will devour them with their fields.}%
\verse{Blow the \textit{horn}\lnPEH{}in Gibeah, 
the trumpet in Ramah. 
Sound the alarm in Beth-aven; 
look behind you, Benjamin.}%
\verse{Ephraim will be a desolation 
in the day of punishment; 
among the tribes of Israel 
I will reveal what is true.}%
\verse{The princes of Judah have become 
like those who remove a landmark; 
on them I will pour out 
my wrath like water.}%
\verse{Ephraim is oppressed, 
crushed in judgment, 
because he was determined 
to go after filth.\lebnote{Hebrew uncertain; another possible translation is “that he follows human commands”}}%
\verse{But I am like a maggot to Ephraim 
and like rottenness to the house of Judah.}%
\verse{And when Ephraim saw his illness, 
and Judah his wound, 
Ephraim went to Assyria, 
he sent to the great king.\lebnote{Some translate literally “King Jareb” (see the NASB, NKJV)}
But he was unable to cure you 
and heal your wound.}%
\verse{Because I will be like a lion to Ephraim 
and like a fierce strong lion to the house of Judah. 
I myself will tear and I will go; 
I will carry off, and there is no one who delivers.}%
\verse{\textit{I will return again}\lebnote{Literally “I will go, I will return”}to my place 
until they acknowledge their guilt 
and seek my face; 
in their distress they will search\lebnote{Or “they will beg favor of”}me.}%
\end{biblechapter}

\begin{biblechapter} % Hosea 6
\verseWithHeading{A Call to Adonai}{Come, let us return to Adonai; 
because it is he who has torn, and he will heal us; 
he has struck us down and will bind us up.}%
\verse{He will revive us after two days; 
on the third day he will raise us up, 
that we may live in his presence.\lebnote{Or “before his face”}}%
\verse{Let us know, let us press on to know Adonai; 
his rising is sure like the dawn. 
He will come like the showers to us, 
like the spring rain that waters the earth.}%
\verse{What will I do with you, O Ephraim? 
What will I do with you, O Judah? 
Your love is like a morning cloud, 
like the dew that goes away early in the morning.}%
\verse{Therefore, I have hewn them by the prophets; 
I have killed them by the words of my mouth, 
and my judgment\lebnote{Hebrew “your judgments”}goes forth like the light.}%
\verse{Because I desire steadfast love and not sacrifice, 
and knowledge of God rather than burnt offerings.}%
\verse{But like Adam, they transgressed the covenant; 
there they dealt faithlessly with me.}%
\verse{Gilead is a city of evil, 
a cunning city because of blood.\lebnote{The Hebrew is difficult; NRSV translates “tracked with blood”}}%
\verse{Like bandits lying in wait, 
so is a band of priests; 
they murder on the road to Shechem; 
indeed, they commit a monstrous crime.}%
\verse{In the house of Israel I have seen something horrible; 
Ephraim’s unfaithfulness is there. 
Israel is defiled.}%
\verseWithHeading{Ephraim’s Guilt}{For you also, O Judah, a harvest is appointed, 
when I restore the fortunes of my people,}%
\end{biblechapter}

\begin{biblechapter} % Hosea 7
\verse{when I would heal Israel, 
and the corruption of Ephraim is revealed, 
and the wicked deeds of Samaria; 
because they deal in falsehood, 
the thief breaks in,\lebnote{Or “comes in”}
and the bandit raids the outside.}%
\verse{\textit{But they did not consider}\lebnote{Literally “But they did not say to their heart”}
that I remember all their wickedness. 
Now their deeds surround them; 
they are before my face.}%
\verse{By their wickedness they make the king glad, 
and the officials by their treacheries.}%
\verse{All of them commit adultery, 
like a burning oven 
whose baker has stopped from stirring the fire, 
and from kneading the dough until it is leavened.}%
\verse{On the day of our king, the princes 
became sick with the heat of wine;\lebnote{This phrase refers to the effects of wine on the body}
he stretched out his hand with mockers.}%
\verse{Because they are kindled\lebnote{Hebrew “they drew near;” other ancient translations read “they are kindled”}like an oven, 
their heart burns within them; 
all night their anger \textit{smolders},\lebnote{Literally “sleeps”}
in the morning it blazes like a flaming fire.}%
\verse{All of them are hot as an oven, 
and they devour\lebnote{Or “they eat”}their rulers. 
All their kings have fallen; 
there is none who calls to me amongst them.}%
\verse{Ephraim mixes himself 
with the nations; 
Ephraim is a bread cake 
not turned over.}%
\verse{Foreigners devour\lebnote{Or “Foreigners eat”}his strength, 
and he does not know it; 
\textit{mold}\lebnote{Literally “grey hair,” but in this case the word refers to the “hairs of mold” on food}is also sprinkled upon him, 
and he does not know it.}%
\verse{The pride of Israel \textit{testifies against him}\lebnote{Literally “answers against his face”}— 
they do not return to Adonai their God; 
they do not seek him for all of this.\lebnote{Or “in all of this”}}%
\verse{Ephraim was like a dove, 
silly, \textit{without sense};\lnLNK{}
they call to Egypt, 
they go to Assyria.}%
\verse{Just as they go, I will cast my net 
over them; 
I will bring them down 
like the birds\lnJPI{}of the heavens;\lnAAH{}
I will discipline them 
according to a report to their assembly.\lnNRZ{}}%
\verse{Woe to them, because they have strayed from me! 
Destruction to them, because they have rebelled against me! 
I myself would redeem them, 
but they speak lies against me.}%
\verse{They do not cry out to me from their heart, 
but they wail on their beds; 
because of grain and new wine they lacerate\lebnote{Or “gash”}themselves; 
they depart from me.}%
\verse{And I myself trained 
and strengthened their arms; 
but they plan evil against me.}%
\verse{They turn, not to the Most High,\lebnote{Hebrew uncertain; the NRSV translates “to that which does not profit”; the NASB translates “but not upward”}
like a slack bow; 
their officials will fall by the sword 
because of the anger of their tongue. 
This is their scorn\lebnote{The NRSV translates “So much for their babbling”}
in the land of Egypt.}%
\end{biblechapter}

\begin{biblechapter} % Hosea 8
\verseWithHeading{Israel Rejects the Good}{Put to your lips the trumpet 
like a vulture over the house of Adonai, 
because they have \textit{broken my covenant}\lebnote{Literally “transgressed my covenant”}
and rebelled against my law.}%
\verse{They cry out to me, 
“My God! We, Israel, know you!”}%
\verse{Israel has spurned the good; 
the enemy will pursue him.}%
\verse{They appointed kings, but not through me; 
they made officials, but without my knowledge. 
With their silver and gold 
they made idols for themselves 
for their\lnDMD{}own \textit{destruction}.\lnEYI{}}%
\verse{Your calf is rejected,\lebnote{Hebrew “he rejects” or “has rejected you”}O Samaria; 
my anger burns against them. 
How long\lebnote{Hebrew “Until when”}will they be incapable of innocence?}%
\verse{Because it is from Israel, 
an artisan made it, 
it is not a god; 
for the calf of Samaria 
will be broken to pieces.}%
\verse{Because they sow the wind, 
they will reap the whirlwind. 
The standing grain does not have heads;\lnFIR{}
it will not yield flour. 
And if it would yield, 
strangers would devour it.}%
\verse{Israel is swallowed up; 
now they are among the nations, 
like an object that no one desires.}%
\verse{For they have gone up to Assyria, 
a wild donkey alone to itself; 
Ephraim has sold itself for lovers.}%
\verse{Even though they have sold themselves to the nations, 
now I will gather them. 
They will soon writhe 
from the burden of kings\lebnote{Hebrew “king”}and princes.}%
\verse{When Ephraim multiplied altars 
to expiate sins, 
they became to him altars to sin on.}%
\verse{I write for him myriads of my instruction;\lebnote{Or “my law”}
they are regarded as a strange thing.}%
\verse{They offer sacrifices of my choice\lebnote{Hebrew uncertain; or “loved ones”}
and they eat flesh; 
Adonai does not accept them. 
Now he will remember their iniquity 
and punish their sins; 
they will return to Egypt.}%
\verse{Israel has forgotten his maker and built palaces, 
and Judah has multiplied fortified cities; 
but I will send fire on his cities 
and it will devour her strongholds.}%
\end{biblechapter}

\begin{biblechapter} % Hosea 9
\verseWithHeading{Ephraim Faces Punishment}{Do not rejoice, O Israel, 
do not exult like the nations! 
For you played the whore, departing from your God; 
you loved a harlot’s wage 
on all the threshing floors of grain.}%
\verse{Threshing floor and wine vat will not feed them, 
and new wine will fail her.\lnEAG{}}%
\verse{They will not remain 
in the land of Adonai. 
But Ephraim will return to Egypt, 
and in Assyria they will eat unclean food.}%
\verse{They will not pour drink offerings of wine to Adonai, 
and their sacrifices will not please him. 
They are like mourners’ bread for them; 
all those who eat it will be defiled. 
For their bread will be \textit{for their hunger};\lnJPG{}
it will not come to the house of Adonai.}%
\verse{What will you do on the day of the appointed time, 
and on the day of the festival of Adonai?}%
\verse{For look! If they flee from the destruction, 
Egypt will gather them; 
Memphis\lebnote{Hebrew “Noph”}will bury them. 
Nettles will possess 
their precious things of silver; 
thorns will be in their tents.}%
\verse{The days of punishment have come; 
the days of retribution have come; 
Israel knows! 
The prophet is a fool, 
the man of the spirit is acting like a madman. 
Because of the greatness of your sin, 
your hostility is great.}%
\verse{The prophet keeps watch over Ephraim for my God; 
the snare of a fowler is on all his ways, 
and hostility in the house of his God.}%
\verse{\textit{They deeply corrupted themselves}\lebnote{Literally “They made deep, they corrupted themselves”}
as in the days of Gibeah; 
he will remember their sin, 
he will punish their sins.}%
\verse{Like the grapes in the wilderness, 
I found Israel. 
Like early ripened fruit on the fig tree in the first season,\lebnote{Or “in the beginning”}
I saw your ancestors.\lnAHC{}
They themselves came to Baal Peor, 
and they consecrated themselves to shame. 
And they became detestable things, 
like \textit{the thing they love}.\lebnote{Literally “like loving things”}}%
\verse{Ephraim’s glory is like a bird; 
it will fly away— 
no birth, no pregnancy, and no conception.}%
\verse{Even though they bring up their children, 
I will bereave them before maturity.\lebnote{Or “until no one is left”}
Woe to them indeed, 
when I depart from them!}%
\verse{Ephraim, as I see it, 
is like a palm in a meadow; 
but Ephraim must bring out 
his children to the slayer.}%
\verse{Give them, O Adonai— 
what will you give them? 
Give them a miscarrying womb 
and dry breasts.}%
\verse{Every evil of theirs began at Gilgal, 
so I began to hate them there; 
because of the evil of their deeds 
I will drive them out from my house. 
I will love them no more; 
all their officials are rebels.}%
\verse{Ephraim is stricken, 
their root is dried up, 
they shall not bear fruit. 
Even if they give birth, 
I will kill the cherished offspring of their womb.}%
\verse{My God will reject them 
because they did not listen to him, 
and they will be wanderers among the nations.}%
\end{biblechapter}

\begin{biblechapter} % Hosea 10
\verseWithHeading{Israel Is Punished}{Israel is a luxuriant vine;\lebnote{Or “ravaged vine”}
he yields fruit for himself. 
The more his fruit increased, 
the more he made numerous altars. 
The more his land prospered, 
the more he\lnFOB{}improved\lebnote{Or “made beautiful”}his stone pillars.}%
\verse{Their heart is false; 
now they must bear their guilt. 
He himself\lebnote{That is, “Adonai himself”}will break down their altars; 
he will destroy their stone pillars.}%
\verse{For now they will say, 
“We have no king; 
indeed, we did not fear Adonai, 
and what can a king do for us?”}%
\verse{They utter words of vain oaths 
when \textit{making covenants},\lebnote{Literally “cutting covenant”}
and judgment blossoms like a poisonous plant 
on the furrows of the field.}%
\verse{The inhabitants\lnGBL{}of Samaria tremble 
for the calf\lebnote{Hebrew “calves”}of Beth-aven. 
Indeed, his people will mourn for it, 
and his idolatrous priests will wail\lebnote{Or “rejoice”}over it— 
over its glory because it has departed from it.}%
\verse{It will also be brought to Assyria, 
as tribute to the great king.\lebnote{See 5:13; some translate literally “King Jareb”}
Ephraim will obtain disgrace 
and Israel will be ashamed from his advice.\lebnote{Or “counsel”}}%
\verse{Samaria will be destroyed; 
her king is like a chip on the surface of the water.}%
\verse{The high places of Aven, 
the sin of Israel, will be destroyed. 
Thorn and thistle will grow 
on their altars. 
They will say to the mountains, “Cover us,” 
and to the hills, “Fall on us.”}%
\verse{From the days of Gibeah you have sinned, O Israel; 
there \textit{they have remained}.\lebnote{Literally “they stood”}
Will not war in Gibeah overtake them 
against the children of evil?}%
\verse{In my desire\lebnote{The LXX translates “I will come”}I will punish them; 
nations will be gathered against them 
when they are punished for \textit{their double iniquities}.\lebnote{So \textit{Qere}; Masoretic Hebrew text (\textit{Kethib}) “their two eyes”}}%
\verse{Ephraim was a trained heifer, 
that loved to thresh grain, 
and I myself \textit{spared}\lebnote{Literally “passed over”}
the fairness of her neck; 
I will make Ephraim break the ground, 
Judah will plow, 
Jacob must till for himself.}%
\verse{Sow for yourselves righteousness; 
\textit{reap loyal love}.\lebnote{Literally “reap according to loyal love”}
Break up for yourself fallow ground; 
it is time to seek Adonai 
so he will come and rain 
righteousness upon you.}%
\verse{You have plowed wickedness, 
you have reaped injustice, 
you have eaten the fruit of lies,\lnJES{}
because you have trusted in your strength, 
in the multitude of your warriors.}%
\verse{The tumult of war will rise up against your people, 
and all your fortresses will be destroyed, 
as Shalman destroyed Beth-arbel; 
on the day of war 
mothers were dashed to pieces with their children.\lebnote{Or “upon their children”}}%
\verse{So it will be done to you, O Bethel, 
because of the evil of your wickedness; 
at dawn, the king of Israel 
will be utterly destroyed.}%
\end{biblechapter}

\begin{biblechapter} % Hosea 11
\verseWithHeading{God Will Have Compassion}{When Israel was a child, I loved him, 
and out of Egypt I called my son.}%
\verse{When I called\lebnote{Hebrew “When they called”; the LXX and modern translations read “When I called”}them, they went from my face.\lebnote{Hebrew “their face”}
They sacrificed to the Baals, and they sacrificed to idols.}%
\verse{And I myself taught Ephraim to walk; 
I took them\lnNRZ{}in my\lnDMD{}arms, 
but they did not know that I healed them.}%
\verse{I drew them \textit{with human ties},\lebnote{Literally “ropes of a human”}
with the bands of love. 
I was to them like one who lifted up\lebnote{Or “like one who imposed”}a yoke on their jaws,\lebnote{Or “infants to their cheeks”}
though I bent down to them\lnDUT{}and let them\lnDUT{}eat.}%
\verse{He will return\lebnote{Or “They will not return”}to the land of Egypt, 
and Assyria will be his king,\lebnote{That is “Israel’s king”}
because they refused to return.}%
\verse{The sword rages in his cities; 
it consumes his false prophets 
and devours because of their plans.}%
\verse{My people are bent on backsliding from me. 
To the Most High they call, 
he does not raise them at all.}%
\verse{How can I give you up, O Ephraim? 
How can I hand you over, O Israel? 
How can I make you like Admah? 
How can I treat you like Zeboiim? 
My heart is disturbed\lebnote{Or “My heart is overturned”}within me; 
\textit{all}\lnJEQ{}my compassions are aroused.}%
\verse{\textit{I will not execute}\lebnote{Literally “I will not do”}\textit{my fierce anger},\lebnote{Literally “the anger of my nose”}
I will not again destroy Ephraim; 
because I am God 
and not a mortal,\lebnote{Or “human”}
the Holy One in your midst; 
and I will not come in wrath.}%
\verse{They will go after Adonai; 
he roars like a lion. 
When he roars, 
his children will come trembling 
from the sea.\lebnote{That is, “the West”}}%
\verse{They will tremble like birds\lnJPI{}from Egypt, 
and like doves\lebnote{Hebrew “dove”}from the land of Assyria; 
and I will let them return\lebnote{Hebrew “I will cause them to live”}to their homes— 
a declaration of Adonai.}%
\verseWithHeading{Israel’s History}{\lebnote{Hosea 11:12–12:14 in the English Bible is 12:1–15 in the Hebrew Bible}Ephraim has surrounded me with lies,\lnJES{}
and the house of Israel with deceit; 
and Judah is still wandering with God 
and is faithful to the Holy One.}%
\end{biblechapter}

\begin{biblechapter} % Hosea 12
\verse{Ephraim herds the wind\lebnote{Or “Ephraim befriends the wind”}
and is pursuing the east wind all day long; 
he multiplies deception and violence 
and he\lnFOB{}\textit{makes a treaty with Assyria},\lebnote{Literally “cuts a covenant with Assyria”}
and oil is brought to Egypt.}%
\verse{Adonai has a quarrel\lebnote{This word in Hebrew conveys a legal setting}with Judah 
and will punish Jacob according to his ways 
and repay him according to his deeds.}%
\verse{In the womb he deceived his brother, 
and in his manhood he struggled with God.}%
\verse{He struggled with the angel and prevailed; 
he pleaded for his mercy. 
He met him at Bethel, 
and there he spoke with him.\lnEMB{}}%
\verse{Adonai the God of hosts, 
Adonai is his renowned name!}%
\verse{But you, you must return to your God; 
keep love and justice, 
and wait continually for your God.}%
\verse{The trader, in his hand are scales of deceit;\lebnote{Or “false balances”}
he loves to oppress.}%
\verse{And Ephraim said, “Surely, I am rich, 
I gained\lebnote{Or “I found”}wealth for myself; 
in all my toil they have not found guilt in me 
that is sin.}%
\verse{But I am Adonai your God 
since\lnBIL{}the land of Egypt; 
I will make you live in tents again, 
like the days of the appointed festival.}%
\verse{I spoke to the prophets; 
I myself multiplied revelations\lebnote{Hebrew “revelation”}
and through the hand of the prophets I will destroy.}%
\verse{If in Gilead there is evil, 
surely they will come to nothing. 
In Gilgal they sacrifice bulls, 
also their altars will be like stone heaps 
on furrows of the field.}%
\verse{Jacob fled to the open field of Aram,\lebnote{Or “the land of Aram”}
and Israel served for a wife, 
and for a wife he watched over sheep.}%
\verse{And by a prophet Adonai brought 
Israel up from Egypt, 
and by a prophet 
he\lebnote{Israel}was watched over.}%
\verse{Ephraim has caused bitter provocation, 
and his Lord will \textit{hold him responsible for his crimes}\lebnote{Literally “will bring down on him his blood”}
and pay back to him his insults.}%
\end{biblechapter}

\begin{biblechapter} % Hosea 13
\verseWithHeading{Adonai’s Judgment on Idolatrous Israel}{When Ephraim spoke,\lebnote{Hebrew “speaks”}there was terror; 
he was exalted in Israel, 
but he incurred guilt through Baal and died.}%
\verse{And now they sin again, 
and they make for themselves a molten idol, 
idols from their silver metal according to their understanding, 
all of them the work of skilled craftsmen. 
To these they say, “Sacrifice!” 
People are kissing bull calves.}%
\verse{Therefore, they will be like the morning cloud, 
and like the dew of early morning going away, 
like chaff swirling from the threshing floor, 
or like smoke from a window.}%
\verse{I am Adonai your God 
since\lnBIL{}the land of Egypt; 
you know no god except me, 
and no one saves besides me.}%
\verse{I fed you\lebnote{So LXX; Hebrew “I knew you”}in the desert, 
in the land of drought.}%
\verse{When I fed them,\lebnote{Hebrew “According to their pasture”}they were satisfied; 
they were satisfied and their heart was lifted up; 
therefore they forgot me.}%
\verse{And I will be like a lion to them; 
I lie in wait beside the way, like a leopard.}%
\verse{I will attack them like a bear robbed of her offspring, 
and I will tear open the covering of their heart; 
there I will devour them like a lion, 
like an animal of the field would mutilate them.}%
\verse{I will destroy you,\lebnote{Hebrew “He destroys you”}O Israel; 
who will help you?\lebnote{Hebrew “for in me, in your help”}}%
\verse{Where now is your king that he may save you? 
Where in all your cities are your judges, 
of whom you said, “Give to 
me a king and rulers?”}%
\verse{I gave you a king in my anger, 
and I took him in my wrath.}%
\verse{The sin of Ephraim is wrapped up; 
his sin is concealed.}%
\verse{The labor pains of childbirth come for him; 
he is an unwise son 
because at the proper time, he does not present himself 
\textit{at the mouth of the womb}.\lebnote{Literally “the place where the sons burst forth”}}%
\verse{Should I redeem them from \textit{the power}\lebnote{Literally “the hand”}of Sheol? \lnPFG{}
Should I deliver them from death? 
Where are your plagues, O Death? 
Where is your destruction, O Sheol?\lnPFG{}
Compassion is hidden from my eyes.}%
\verse{Although he may flourish among reeds,\lnPDK{}
the east wind will come, a wind of Adonai 
rising from the desert; 
his fountain will dry up, 
his spring will be parched. 
It will plunder his treasury,\lebnote{Or “its treasury”}
every object of desire.}%
\verse{\lebnote{Hosea 13:16–14:9 in the English Bible is 14:1–10 in the Hebrew Bible}Samaria will be guilty, 
because she has rebelled against her God; 
they will fall by the sword, 
their children will be dashed to pieces, 
and their pregnant women will be ripped open.}%
\end{biblechapter}

\begin{biblechapter} % Hosea 14
\verseWithHeading{Return to Adonai}{Return to Adonai your God, O Israel, 
for you have stumbled because of your sin.}%
\verse{Take words with you, 
and return to Adonai. 
Say to him, 
“Take away all guilt; 
accept good, and we will offer 
the fruit\lnNRZ{}of our lips.}%
\verse{Assyria will not save us; 
we will not ride on horses,\lnATY{}
and we will say no more, “Our God,” 
to the work of our hands 
because in you the fatherless child finds mercy.}%
\verse{I will heal their disloyalty;\lebnote{Or “backsliding”}
I will love them freely 
because my anger has turned back 
from them.\lnDUT{}}%
\verse{I will be like the dew to Israel; 
he will blossom like the lily plant, 
and he will strike his roots like the trees of Lebanon.}%
\verse{His new plant shoots will spread out; 
his splendor will be like the olive tree, 
and his scent like the trees of Lebanon.}%
\verse{\textit{They will again dwell}\lebnote{Literally “They shall return, they shall dwell”}in my\lnDMD{}shadow; 
they will grow grain 
and they will blossom like the plant vine; 
his fame will be like the wine of Lebanon.}%
\verse{O Ephraim, \textit{what have I to do}\lebnote{Literally “what for me”}with idols? 
I myself have answered and looked after you.\lnDUT{}
I am like a luxuriant cypress; 
your fruit\lebnote{Or “your faithfulness”}comes from me.}%
\verse{Who is wise that he can understand these things? 
Who is discerning that he knows them? 
The ways of Adonai are right, 
and the righteous walk in them; 
but transgressors stumble in them.}%
\end{biblechapter}

