\biblebook{Isaiah}

\begin{biblechapter} % Isaiah 1
\verseWithHeading{Superscription} The vision of Isaiah son of Amoz, which he saw concerning Judah and Jerusalem in the days of Uzziah, Jotham, Ahaz, and Hezekiah, kings of Judah.
\verseWithHeading{Rebellious Judah} Hear, heavens, 
and listen, earth, 
for Adonai has spoken: 
“I reared children 
and I brought them up, 
but they rebelled against me.
\verse An ox knows its owner 
and a donkey the manger of its master. 
Israel does not know; 
my people do not understand.
\verse Ah, sinful nation, 
a people heavy with iniquity, 
offspring of evildoers, 
children who deal corruptly. 
They have forsaken Adonai; 
they have despised the holy one of Israel. 
They are estranged and gone backward.
\verse Why do you want to be beaten again? 
You continue in rebellion. 
The whole of the head is sick, 
and the whole of the heart is faint.
\verse From the sole of the foot and up to the head 
there is no health in it; 
bruise and sore and bleeding wound have not been cleansed, 
and they have not been bound up 
and not softened with the oil.
\verse Your country is desolate, 
your cities are burned with fire; 
As for your land, aliens are devouring it in your presence, 
and it is desolate, like devastation by foreigners.
\verse And the daughter of Zion is left like a booth in a vineyard, 
like a shelter in a cucumber field, 
like a city that is besieged.\lebnote{Or “preserved”}
\verse If Adonai of hosts had not left us survivors,\lebnote{Hebrew “survivor”} 
we would have been as few as Sodom, 
we would have become like Gomorrah.
\verse Hear the word of Adonai, rulers of Sodom! 
Listen to the teaching of our God, people of Gomorrah!
\verse What is the abundance of your sacrifices to me? says Adonai. 
I have had enough of burnt offerings of rams 
and the fat of fattened animals 
and I do not delight in the blood of bulls 
and ram-lambs and goats.
\verse When you come to appear before me, 
who asked for this from your hand: 
you trampling my courts?
\verse You must not \textit{continue}\lebnote{Literally “increase” or “add to”} to bring offerings\lebnote{Hebrew “offering”} of futility, 
incense—it is an abomination to me; 
new moon and Sabbath, the calling of a convocation— 
I cannot endure iniquity with solemn assembly.
\verse Your new moons and your appointed festivals my soul hates; 
they have become to me like a burden, 
I am not able to bear them.
\verse And when you stretch out your hands, 
I will hide my eyes from you; 
even though you make many prayers,\lebnote{Hebrew “prayer”} 
I will not be listening. 
Your hands are full of blood.
\verse Wash! Make yourselves clean! 
Remove the evil of your doings from before my eyes! 
Cease to do evil!
\verse Learn to do good! 
Seek justice! Rescue the oppressed! 
Defend the orphan! Plead for the widow!
\verse “Come now, and let us argue,” says Adonai. 
“Even though your sins are like scarlet, they will be white like snow; 
even though they are red like crimson, they shall become like wool.
\verse If you are willing and you are obedient, 
you shall eat the good of the land.
\verse But if you refuse and you rebel, you shall be devoured by the sword. For the mouth of Adonai has spoken.”
\verseWithHeading{Purifying Jerusalem} How has a faithful city become like a whore? 
Full of justice, righteousness lodged in her, but now murderers.
\verse Your silver has become as dross; 
Your wine is diluted with waters.
\verse Your princes are rebels 
and companions of thieves. 
Every one loves a bribe 
and runs after gifts. 
They do not defend the orphan 
and the legal dispute of the widow does not come before them.
\verse Therefore, the declaration of the Lord Adonai of hosts, the Mighty One of Israel: 
“Ah, I will be relieved of my enemies, 
and I will avenge myself on my foes.
\verse And I will turn my hand against you; 
I will purify your dross like lye, 
and I will remove all of your tin.
\verse And I will restore your judges, as at the first, 
and your counselors, as at the beginning. 
After this \textit{you will be called}\lebnote{Literally “it will be called for you”} the city of righteousness, 
faithful city.
\verse Zion will be redeemed by justice, 
and those of her who repent, by righteousness.
\verse But the destruction of rebels and sinners shall be together, 
and those who forsake Adonai will perish.
\verse For you\lebnote{Literally “they,” but a few manuscripts read “you,” which fits the context better} will be ashamed of the oaks in which you delighted, 
and you will be disgraced because of the gardens that you have chosen.
\verse For you shall be like an oak withering its leaves, 
and like a garden where there is no water for her.
\verse And the strong man shall become like tinder, 
and his work like a spark. 
And both of them shall burn together, 
and there is not one to quench them.”
\end{biblechapter}

\begin{biblechapter} % Isaiah 2
\verseWithHeading{The Mountain of Adonai} The word that Isaiah son of Amoz saw concerning Judah and Jerusalem:
\verse And it shall happen in the future of the days 
the mountain of the house\lebnote{Or “temple”} of Adonai shall be established; 
it will be among the highest\lebnote{Literally “head”} of the mountains, 
and it shall be raised from the hills. 
All of the nations shall travel to him;
\verse many peoples shall come. 
And they shall say, 
“Come, let us go up to the mountain of Adonai, 
to the house\lebnote{Or “temple”} of the God of Jacob, 
and may he teach us part of his ways, 
and let us walk in his paths.” 
For instruction shall go out from Zion, 
and the word of Adonai from Jerusalem.
\verse He shall judge between the nations 
and he shall arbitrate for many peoples. 
They shall beat their swords into ploughshares 
and their spears into pruning hooks. 
A nation shall not lift up a sword against a nation, 
and they shall not learn war again.
\verseWithHeading{The Day of Adonai} House of Jacob, come and let us walk in the light of Adonai.
\verse For you have forsaken your people, house of Jacob, 
because they are full\lebnote{Possibly “of diviners” was part of the original text here} from the east, 
and of soothsayers like the Philistines, 
and \textit{they make alliances}\lebnote{Literally “they clap hands”} with the offspring of foreigners.
\verse And its land is filled with silver and gold, 
and there is no end to its treasures; 
and its land is filled with horses, 
and there is no end to its chariots.
\verse Its land is filled with idols; 
they bow down to the work of their\lebnote{Hebrew “his”} hands, 
to what they made with their\lebnote{Hebrew “his”} fingers.
\verse So humanity is humbled; 
everyone is humbled, 
and you must not forgive them.
\verse Enter into the rock 
and hide yourself in the dust 
from the presence of the terror of Adonai 
and from the glory of his majesty.
\verse The \textit{haughty eyes}\lebnote{Literally “eyes of the haughtiness”} of humanity will\lebnote{The Hebrew is singular} be brought low, 
and the pride of everyone will be humbled, 
and Adonai alone will be exalted on that day.
\verse For there is a day for Adonai of hosts 
against all of the proud and the lofty 
and against all that is lifted up and humble,\lebnote{Or “it will be humbled”}
\verse and against all the lofty and lifted up cedars of Lebanon, 
and against all the large trees of Bashan,
\verse and against all the high mountains, 
and against all the lofty hills,
\verse and against every kind of high tower, 
and against every kind of fortified wall,
\verse and against all the ships of Tarshish, 
and against all the ships of desire.
\verse And the haughtiness of the people shall be humbled, 
and the pride of everyone shall be brought low, 
and Adonai alone will be exalted on that day.
\verse And the idols shall pass away entirely,
\verse and they will enter into the caves of the rocks 
and into the holes of the ground 
from the presence of the terror of Adonai 
and from the glory of his majesty 
\textit{when he rises}\lebnote{Literally “in his rising”} to terrify the earth.
\verse On that day humanity will throw away its idols of silver 
and its idols of gold, 
which they made for it to worship, 
to the rodents\lebnote{Or “moles”} and to the bats—
\verse to enter into the crevices of the rocks 
and into the clefts of the crags 
from the presence of the terror of Adonai 
and from the glory of his majesty, 
\textit{when he rises}\lebnote{Literally “in his rising”} to terrify the earth.
\verse Turn away from humanity, 
who has breath in its nostrils,\lebnote{Hebrew “nostril”} 
for by\lebnote{Or “in”} what is it esteemed?
\end{biblechapter}

\begin{biblechapter} % Isaiah 3
\verseWithHeading{Leaders of Judah and Jerusalem} For look, the Lord Adonai of hosts is removing \textit{every source of support}\lebnote{Literally “supplies and supplies”} 
from Jerusalem and from Judah: 
all of the supplies of bread 
and all of the supplies of water,
\verse mighty warrior and man of war, 
judge and prophet, 
and diviner and elder,
\verse captain of fifty and the honorable men of rank, 
and counselor and skillful magicians and skillful enchanter.
\verse And I will make boys their princes, 
and children shall rule over them.
\verse And the people will be oppressed \textit{by each other}\lebnote{Literally “man by man”} 
and a man by his neighbor. 
The boy will act arrogantly toward the elder, 
and the dishonorable toward the honorable.
\verse Indeed, a man will seize his brother 
in the house of his father: 
“\textit{You have a cloak};\lebnote{Literally “a cloak for you”} 
you shall be a leader for us, 
and this heap of ruins shall be under your hand!”
\verse He will lift up his voice on that day, saying, 
“I will not be a healer; 
in my house there is no bread 
and there is no cloak. 
You shall not make me the leader of the people!”
\verse For Jerusalem has stumbled, 
and Judah has fallen 
because their speech and their deeds are against Adonai, 
defying the eyes of his glory.
\verse The look on their faces testifies against them 
and they declare their sin like Sodom; 
they do not hide it. 
Woe to their soul! 
For they have dealt out evil to themselves.
\verse Tell the innocent\lebnote{Singular} that it is good 
for they shall eat the fruit of their deeds.
\verse Woe to the wicked!\lebnote{Singular} It is bad! 
For what is done by his hands will be done to him.
\verse My people—children are their oppressors, 
and women rule over them. 
My people, your leaders are misleading you, 
and they confuse the course of your paths.
\verse Adonai takes his stand to conduct a legal case 
and takes his stand to judge the peoples.
\verse Adonai enters into judgment with the elders of his people and its princes. 
“And you! You have devoured the vineyard; 
the spoil of the poor is in your houses!
\verse \textit{Why}\lebnote{Literally “What to you”} do you crush my people 
and grind the face of the poor?” \textit{declares}\lebnote{Literally “declaration of”} the Lord Adonai of hosts.
\verseWithHeading{The Pride of Jerusalem’s Women} And Adonai said: “Because\lebnote{There are two Hebrew words for “because” here} the daughters of Zion are haughty, 
and they walk with outstretched neck, 
and they give flirting glances with their eyes, 
\textit{mincing along as they go},\lebnote{Literally “they go walking and mincing along”} 
and \textit{with their feet they rattle their bangles},\lebnote{Literally “they tinkle with their feet”}
\verse the Lord will make the heads\lebnote{Hebrew “head”} of the daughters of Zion scabby, 
and Adonai will lay their foreheads bare.”
\verse In that day the Lord will take away the finery of the anklets 
and the headbands and the crescent necklaces,
\verse the pendants and the bracelets and the veils,
\verse the headdresses and the armlets and the sashes, 
and the \textit{perfume boxes}\lebnote{Literally “houses of the breath”} and the amulets,
\verse the signet rings and the nose rings,
\verse the festal robes and the mantles, 
and the cloaks and the handbags,
\verse and the mirrors and the linen garments, 
and the turbans and the wraps.
\verse And this shall happen: There will be 
a stench instead of perfume, 
and a rope instead of a sash, 
and baldness instead of a well-set hairdo, 
and a clothing wrap of sackcloth instead of a rich robe, 
branding instead of beauty.
\verse Your men shall fall by the sword, 
and your warriors in battle.
\verse And her\lebnote{That is, Jerusalem’s} gates shall lament and mourn, 
and she shall be banished; 
she shall sit upon the ground.
\end{biblechapter}

\begin{biblechapter} % Isaiah 4
\verse And seven women shall grasp at one man on that day, saying, 
“We will eat our own bread, 
and we will wear our own clothing; 
only \textit{let us be called by your name}!\lebnote{Literally “let your name be called upon us”} 
Take away our disgrace!”
\verseWithHeading{The Glory of the Branch of Adonai} On that day the branch of Adonai shall become beautiful and glorious, 
and the fruit of the land shall become the pride and glory of the survivors\lebnote{Hebrew “survivor”} of Israel.
\verse And this shall happen: He who is left in Zion 
and he who remains in Jerusalem will be called holy, 
everyone written for life in Jerusalem,
\verse when the Lord has washed away the filth of the daughters of Zion 
and cleansed the blood of Jerusalem from her midst 
by a spirit of judgment 
and by a spirit of burning.
\verse Then Adonai will create over all of the site\lebnote{Or “place”} of \textit{Mount Zion}\lebnote{Literally “the mountain of Zion”} 
and over her assembly a cloud by day 
and smoke and the brightness of flaming fire by night. 
Indeed, over all the glory there will be a canopy,
\verse and it will be a shelter for shade from the heat by day, 
and a refuge and a hiding place from rainstorm and from rain.
\end{biblechapter}

\begin{biblechapter} % Isaiah 5
\verseWithHeading{The Song of the Vineyard} Let me sing for my beloved 
a song of my love concerning his vineyard: 
\textit{My beloved had a vineyard}\lebnote{Literally “A vineyard was for my beloved”} on \textit{a fertile hill}.\lebnote{Literally “a horn of a son of olive oil.” The Hebrew for horn, \textit{qeren}, sounds like the Hebrew for vineyard, \textit{kerem|i|\lebnote{:
\verse And he dug it and cleared it of stones, 
and he planted it with choice vines,\lebnote{Hebrew “vine”} 
and he built a watchtower in the middle of it, 
and he even hewed out a wine vat in it, 
and he waited for it to yield grapes— 
but it yielded wild grapes.
\verse And now, inhabitants of Jerusalem 
and men\lebnote{Hebrew “man”} of Judah, judge between me and my vineyard.
\verse What more was there to do for my vineyard that I have not done in it? 
Why did I hope for it to yield grapes, and it yielded wild grapes?
\verse And now let me tell you what I myself am about to do to my vineyard.
\verse I will remove its hedge, and it shall become a devastation. 
I will break down its wall, and it shall become a trampling.
\verse And I will make it a wasteland; 
it shall not be pruned and hoed, 
and it shall be overgrown with briers\lebnote{Hebrew “brier”} and thornbushes.\lebnote{Hebrew “thornbush”} 
And concerning the clouds, I will command \textit{them not to send}\lebnote{Literally “from sending”} rain down upon it.
\verseWithHeading{Woes on the Wicked} Ah! Those who \textit{join}\lebnote{Literally “touch”} house with house, 
they join field together with field 
until \textit{there is no place}\lebnote{Literally “an end of place”} 
and you are caused to dwell alone in the midst of the land.
\verse Adonai of hosts said in my ears:
\verse \textit{Surely}\lebnote{Literally “If not”} many houses shall become a desolation, 
large and beautiful ones without inhabitant.
\verse For ten acres of vineyard shall yield one bath,\lebnote{A bath is a liquid measure} 
and the seed of a homer will yield an ephah.\lebnote{An ephah is a dry measure equal to one-tenth of a homer}
\verse Ah! Those who rise early in the morning, 
they pursue strong drink. 
Those who linger in the evening, 
wine inflames them.
\verse And there will be lyre and harp, 
tambourine and flute, 
and wine at their feasts, 
but they do not look at the deeds\lebnote{Hebrew “deed”} of Adonai, 
and they do not see the work of his hands.
\verse Therefore my people will go into exile without knowledge, 
and their\lebnote{Hebrew “its”} nobles\lebnote{Hebrew “noble”} will be men of hunger, 
and their\lebnote{Hebrew “its”} multitude is parched with thirst.
\verse Therefore Sheol has enlarged its throat, 
and it has opened wide its mouth without limit, 
and her\lebnote{That is, Jerusalem’s} nobles\lebnote{Hebrew “noble”} will go down, and her multitude, 
her tumult and those who revel in her.
\verse And humankind is bowed down, 
and man is brought low, 
and the eyes of the haughty are humiliated.
\verse But Adonai of hosts is exalted by justice, 
and the holy God shows himself holy by righteousness.
\verse And then the lambs will graze as in their pasture, 
and \textit{fatlings, kids\lebnote{Following the Septuagint, which reads the Hebrew \textit{grym} (resident aliens) as \textit{gdym} (young goats/sheep)} will eat among the sites of ruins.}\lebnote{Literally “and ruins, fatlings, resident aliens, will eat”}
\verse Ah! Those who drag iniquity along with the cords of falsehood 
and sin as with rope of the cart,
\verse those who say, 
“Let him make haste; 
let him hurry his work 
so that we may see it 
and let it draw near 
and let the plan of the holy one of Israel come 
so that we may know it!”
\verse Ah! Those who call evil good and good evil, 
those who put darkness for light and light for darkness, 
those who put bitter for sweet and sweet for bitter!
\verse Ah! Those who are wise in their own eyes 
and have understanding \textit{in their view}!\lebnote{Literally “before their faces”}
\verse Ah! Heroes at drinking wine, 
and men of capability at mixing strong drink!
\verse Those who acquit the guilty because of a bribe 
and remove the justice of the innocent from him.
\verse Therefore as the tongue of fire devours the stubble, 
and dry grass sinks down in the flame, 
so their root will become like the stench, 
and their blossom will go up like the dust. 
For they have rejected the instruction of Adonai of hosts, 
and they have treated the word of the holy one of Israel with contempt.
\verseWithHeading{Adonai’s Outstretched Hand} In all of this his anger has not turned back, 
and still his hand is stretched out.
\verse And he will raise a signal for a nation\lebnote{The Hebrew is plural, but the following verses refer to the nation as singular} from afar, 
and he will whistle for it from the end of the earth. 
And look! It comes quickly, swiftly!
\verse None is weary, 
and none among him stumbles; 
none slumbers and none sleeps. 
And no loincloth on his waist is opened, 
and no thong of his sandals is drawn away.
\verse Whose arrows are sharp, 
and all of his bows are bent. 
The hoofs of his horses are reckoned like flint, 
and his wheels like the storm wind.
\verse His roaring is like the lion, 
and he roars like young lions. 
And he growls and seizes his prey, 
and he carries it off, 
and not one can rescue it.
\end{biblechapter}

\begin{biblechapter} % Isaiah 6
\verseWithHeading{Isaiah’s Commission} In the year of the death of Uzziah the king, I saw the Lord sitting on a high and raised throne, and the hem of his robe was filling the temple.
\verse Seraphs were standing above him. \textit{Each had six wings}:\lebnote{Literally “six wings six wings for one”} with two he covered his face, and with two he covered his feet, and with two he flew.
\verse And the one called to the other and said, “Holy, holy, holy is Adonai of hosts! 
The \textit{whole earth is full of his glory}.”\lebnote{Literally “fullness of all of the earth glory him”}
\verse And the pivots of the thresholds shook from the sound of those who called, and the house\lebnote{Or “temple”} was filled with smoke.
\verse And I said, “Woe to me! For I am destroyed!\lebnote{Or “silenced”} For I am a man \textit{of unclean lips},\lebnote{Literally “unclean of lips”} and I am living among\lebnote{With an emphatic sense: “in the very heart and midst of”} a people \textit{of unclean lips},\lebnote{Literally “unclean of lips”} for my eyes have seen the king, Adonai of hosts!”
\verse Then one of the seraphs flew to me, and in his hand was a hot coal he had taken from the altar with tongs.
\verse And he touched my mouth, and he said, “Look! This has touched your lips 
and has removed your guilt, 
and your sin is annulled.”
\verse Then I heard the voice of the Lord saying, “Whom shall I send? 
And who will go for us?” And I said, “I am here! 
Send me!”
\verse And he said, “Go and say to this people,
\verse ‘Keep on listening and do not comprehend! 
And keep on looking and do not understand!’
\verse Then I said, “Until when, Lord?” And he said,
\verse “Until the cities lie wasted without inhabitant, 
and houses without people, 
and the land is ruined and a waste,
\verse and Adonai sends the people far away, 
and the abandonment is great in the midst of the land.
\end{biblechapter}

\begin{biblechapter} % Isaiah 7
\verseWithHeading{The Sign to Ahaz} This happened in the days of Ahaz, son of Jotham, son of Uzziah, king of Judah. Rezin, king of Aram, and Pekah, son of Remaliah, king of Israel, went\lebnote{The Hebrew is singular} up to Jerusalem for warfare against it,\lebnote{Literally “her”} but he was not able to fight against it.\lebnote{Literally “her”}
\verse When it was reported to the house of David, saying “Aram stands by Ephraim,” his heart and the heart of his people shook like the shaking of the trees of the forest because of the wind.
\verse Then Adonai said to Isaiah, “Go out to meet Ahaz, you and Shear-Jashub your son, at the end of the conduit of the upper pool on the highway of the washer’s field.
\verse And you must say to him, ‘Take heed and be quiet! You must not fear, and your heart must not be faint because of these two stumps of smoldering firebrands, \textit{because of the fierce anger of}\lebnote{Literally “because of the fierceness of the anger of”} Rezin and Aram and the son of Remaliah.
\verse Because Aram has plotted evil against you with Ephraim and the son of Remaliah, saying,
\verse “Let us go up against Judah and let us tear her apart, \textit{and let us lay it open and so bring it unto ourselves},\lebnote{Literally “and let us break through her to us”} and let us install the son of Tabeel as king in her midst.”
\verse Thus says the Lord Adonai, “It shall not stand, and it shall not come to pass.
\verse For the head of Aram is Damascus, and the head of Damascus is Rezin, and in sixty-five years from now Ephraim will be too shattered to be a people.
\verse And the head of Ephraim is Samaria, and the head of Samaria is the son of Remaliah. If you do not believe then you will not endure.” ’ ”
\verse And Adonai continued to speak to Ahaz, saying,
\verse “Ask for a sign for yourself from Adonai; make it deep as Sheol or make it high as above.”
\verse But Ahaz said, “I will not ask, and I will not put Adonai to the test.”
\verse Then he said, “Hear, house of David! Is it too little for you to make men weary, that you should also make my God weary?
\verse Therefore the Lord himself will give you\lebnote{The Hebrew is plural} a sign. Look! the virgin\lebnote{Or “young woman”} is with child and she is about to give birth to a son, and she shall call his name ‘God with us.’
\verse He shall eat curds and honey until he knows to reject the evil and to choose the good.
\verse For before the boy knows to reject the evil and to choose the good, \textit{the land whose two kings you dread will be abandoned.}\lebnote{Literally “the land which you dread will be abandoned because of the face of her two kings”}
\verseWithHeading{That Day} “Adonai will bring on you and on your people and on the house of your ancestor\lebnote{Or “father”} days that have not come since the day Ephraim departed from Judah: the king of Assyria.”
\verse And this shall happen: On that day, Adonai will whistle for the fly that is at the end of the stream of Egypt and the bee that is in the land of Assyria.
\verse And all of them will come and settle in the rivers of the cliffs and in the clefts of the rocks and on all of the thornbushes and watering places.
\verse On that day, the Lord will shave the head and the hair of the feet with a razor of the one hired from beyond the river—with the king of Assyria—and it will even take off the beard.
\verse And this shall happen: on that day, a young man will keep a young cow of the herd and two sheep alive.
\verse And this shall happen: because of the abundance of milk production, he will eat curds, for every one that is left in the midst of the land will eat curds and honey.
\verse And this shall happen on that day: Every place where there are a thousand vines\lebnote{Hebrew “vine”} for a thousand silver pieces will become briers,\lebnote{Hebrew “brier”} and it will be thornbushes.\lebnote{Hebrew “thornbush”}
\verse One will go there with arrows and bow, for all of the land will be briers\lebnote{Hebrew “brier”} and thornbushes.\lebnote{Hebrew “thornbush”}
\verse And as for all of the hills that they hoed with the hoe, you will not go there, for fear of briers\lebnote{Hebrew “brier”} and thornbushes.\lebnote{Hebrew “thornbush”} And it will become like pastureland for cattle and overtrodden land for sheep.
\end{biblechapter}

\begin{biblechapter} % Isaiah 8
\verseWithHeading{Signs of the Assyrian Invasion} Then Adonai said to me, “Take yourself a large tablet and write on it with a common stylus pen: Maher-Halal-Hash-Baz.
\verse And I will require reliable witnesses as a witness for me: Uriah the priest and Zechariah son of Jeberekiah.”
\verse And I approached the prophetess, and she conceived, and she gave birth to a son. And Adonai said to me, “Call his name Maher-Halal-Hash-Baz.
\verse For before the boy knows to call ‘my father’ and ‘my mother,’ one will carry away the wealth of Damascus and the spoil of Samaria in the presence of the king of Assyria.”
\verseWithHeading{Shiloah Waters and Euphrates Flood} And Adonai continued to speak to me again, saying,
\verse “Because this people has refused the gently flowing waters of Shiloah and rejoices over Rezin and the son of Remaliah,
\verse therefore look! The Lord is bringing up the waters of the great and mighty river against them, the king of Assyria and all his glory.
\verse And he will rise above all his channels, 
and he will flow over all his banks.
\verse And he will sweep into Judah; 
he will overflow and he will flood up to the neck. 
He will reach, and \textit{he will spread his wings out over your entire land},\lebnote{Literally “the outspreading of his wings will be the fullness of the breadth of your land”} God with us.”
\verse Be broken, you peoples, and be dismayed. 
And listen, all distant parts of the earth; 
gird yourselves and be dismayed; 
gird yourselves and be dismayed!
\verseWithHeading{Wait for Adonai} For Adonai said this to me \textit{while his hand weighed heavily on me},\lebnote{Literally “with the strength of the hand”} 
and he \textit{warned me not to walk}\lebnote{Literally “instructed me from walking”} in the way of this people, saying,
\verse “You must not call conspiracy everything that this people calls conspiracy, 
and you must not \textit{share its fear},\lebnote{Literally “fear its fear”} and you must not be in dread.
\verse You shall regard Adonai of hosts as holy, 
and he is your\lebnote{The Hebrew is plural} fear, and he is your dread.
\verse And he will become like a sanctuary and a stumbling-stone, 
and like a stumbling-rock for the two houses of Israel, 
like a trap and a snare for the inhabitants of Jerusalem.
\verse And many shall stumble among them, 
and they shall fall and they shall be broken, 
and they shall be ensnared and they shall be caught.”
\verse Bind up the testimony; 
seal the teaching among my disciples.
\verse And I will wait for Adonai, 
who hides his face from the house of Jacob, 
and I will await him.
\verse Look! I and the children whom Adonai has given to me are like signs and portents in Israel from Adonai of hosts, the one who dwells on the mountain of Zion.
\verse Now if they tell you, “Consult the ghosts and the spirits, those who chirp and those who mutter. Should not a people consult its gods, the dead on behalf of the living,
\verse for teaching and for testimony?” \textit{surely they who speak like this have no dawn}.\lebnote{Literally “if not, they say like this word that there is no dawn for him,” which could also mean “if not, let them say a word like this: that there is no dawn for him”}
\verse And it\lebnote{That is, the nation} will pass through it\lebnote{That is, the land} distressed and hungry, and this shall happen: when it is hungry, it will be enraged, and it will curse its king and its gods,\lebnote{Or “God”} and it will face upwar
\verse or look to the earth. But look! Distress and darkness, the gloom of affliction! And it will be thrust into darkness!
\end{biblechapter}

\begin{biblechapter} % Isaiah 9
\verseWithHeading{New Light: The Birth of a King} \lebnote{Isaiah 9:1–21 in the English Bible is 8:23–9:20 in the Hebrew Bible} \textit{But there will be no gloom for those who were in distress}.\lebnote{Literally “Indeed there is no gloom for the one to whom there was anxiety for her”} In former times he\lebnote{That is, Adonai} treated the land of Zebulun and Naphtali with contempt, but in the future he will honor the way of the sea beyond the Jordan, Galilee of the nations.
\verse The people who walked in darkness have seen a great light; 
light has shined on those who lived in a land of darkness.
\verse You have made the nation numerous; 
you have not\lebnote{The written text (\textit{Kethib}) is “not,” but the reading tradition (\textit{Qere}) is “for it”} made the joy great. 
They rejoice in your presence as with joy at the harvest, 
as they rejoice when they divide plunder.
\verse For you have shattered the yoke of its burden 
and the stick of its shoulder, 
the rod of its oppressor, on\lebnote{Hebrew “like”; the Hebrew letters for “like” and “on” look alike} the day of Midian.
\verse For every boot \textit{that marches and shakes the earth}\lebnote{Literally “marching with shaking,” which might mean “marching is with shaking”} 
and garment rolled in blood 
\textit{will}\lebnote{Literally “and it will”} be for burning—fire fuel.
\verse For a child has been born for us; 
a son has been given to us. 
And the dominion will be on his shoulder, 
and his name is called Wonderful Counselor, Mighty God, 
Everlasting Father, Prince of Peace.
\verse \textit{His dominion will grow continually, 
and to peace there will be no end}\lebnote{Literally “There is no end to the increase of the dominion and to peace”} 
on\lebnote{Or “for”} the throne of David and over\lebnote{Or “for”} his kingdom, 
to establish it\lebnote{That is, the kingdom} and sustain it 
with justice and righteousness 
now and forever. 
The zeal of Adonai of hosts will do this.
\verseWithHeading{Adonai’s Anger against Arrogance} The Lord has sent out a word against Jacob, 
and it fell on Israel.
\verse And all of the people knew it, 
Ephraim and the inhabitants of Samaria 
in pride and arrogance of heart, saying,
\verse “The bricks have fallen, but we will build with dressed stone. 
The sycamore-fig trees were felled, but we will replace them with cedars.”
\verse So Adonai strengthened \textit{the adversaries of Rezin}\lebnote{Or “his adversaries” if a copyist added “of Rezin” in error} against him, 
and he provoked his enemies—
\verse Aram from the east 
and Philistines\lebnote{Hebrew “Philistine”} from the west— 
and they devoured Israel with the whole mouth. 
He has not turned away his anger in all of this, 
and his hand is still stretched out.
\verse And the people did not turn to the one who struck it,\lebnote{That is, the people} 
and they did not seek Adonai of hosts.
\verse So Adonai cut off head and tail from Israel, 
palm branch and reed in one day.
\verse Elders\lebnote{Hebrew “Elder”} and \textit{the respectable}\lebnote{Literally “one lifted up of face”} are the head, 
and prophets\lebnote{Hebrew “prophet”} who teach lies\lebnote{Hebrew “lie”} are the tail.
\verse And the leaders of this people were misleading them, 
and those who were led were confused.
\verse Therefore the Lord did not rejoice over its young men, 
and he did not have compassion on its orphans and widows, 
for everyone was godless and an evildoer, 
and every mouth was speaking folly. 
In all of this his anger did not turn away, 
and still his hand is stretched out.
\verse For wickedness burned like fire; 
it consumed brier and thorn. 
And it kindled the thickets of the forest, 
and they swirled upward in a column of smoke.
\verse The land was burned through the wrath of Adonai of hosts, 
and the people became like fire fuel. 
\textit{People had no compassion toward each other}.\lebnote{Literally “A man does not have compassion toward his brother”}
\verse They devoured on the right but still were hungry 
and devoured on the left but they were not satisfied. 
Each one devoured the flesh of his arm,
\verse Manasseh devoured Ephraim, and Ephraim Manasseh; 
together they were against Judah. 
In all of this his anger has not turned away, 
and still his hand is stretched out.
\end{biblechapter}

\begin{biblechapter} % Isaiah 10
\verseWithHeading{Woes on the Wicked} Ah! Those who decree decrees of evil, 
and writers who have written harm,
\verse to guide the needy away from legal claims,\lebnote{Hebrew “claim”} 
and to rob the justice from the poor of my people, 
to make widows their spoil; 
and they plunder orphans.
\verse And what will you do at the day of punishment, 
and at calamity? It comes from afar! 
To whom will you flee for help, 
and where will you leave your wealth,
\verse \textit{save that they bow down under the prisoners 
and fall under the slain}?\lebnote{Literally “except he will bow down under a prisoner and under slain they will fall”; “under” could also mean “in the place of”} 
In all of this his anger has not turned away, 
and still his hand is stretched out.
\verseWithHeading{Judgment on Assyria’s Arrogance} Ah! Assyria, the rod of my anger, 
and a staff is in their hand: my wrath!
\verse I send him against a godless nation, 
and I command him against the people of my wrath, 
to capture spoil 
and to carry off plunder, 
and to make them\lebnote{Hebrew “him”} a trampling place, 
like the clay of the streets.
\verse But he does not think this,\lebnote{Or “so”} 
and his heart does not plan this. 
For it is in his heart to destroy 
and to cut off not a few nations.
\verse For he says, “Are not my commanders altogether kings?
\verse Is not Calno like Carchemish? 
Is not Hamath like Arpad? 
Is not Samaria like Damascus?
\verse As my hand has reached to the kingdoms of the idols\lebnote{Hebrew “idol”} 
—and their images were \textit{greater than those of}\lebnote{Literally “from”} Jerusalem and Samaria—
\verse shall I not do to Jerusalem and its idols 
what I have done to Samaria and her idols?”
\verse And this shall happen: when the Lord has finished all his work against \textit{Mount Zion}\lebnote{Literally “the mountain of Zion”} and Jerusalem, “I will punish the \textit{arrogance}\lebnote{Literally “fruit of the greatness of the heart”} of the king of Assyria and \textit{his haughtiness}.”\lebnote{Literally “the pride of the height of his eyes”}
\verse For he says,
\verse “I have done it by the strength of my hand 
and by my wisdom, for I have understanding, 
and I have removed the boundaries of peoples, 
and I have plundered their stores, 
and like a bull I have brought down \textit{the inhabitants}.\lebnote{Or, “those sitting,” that is, rulers sitting on thrones}
\verse And my hand has found, like a nest, the wealth of the peoples, 
and like the gathering of forsaken eggs, I myself have gathered all the earth. 
And there was no fluttering wing or open mouth or chirp.”
\verse Does the ax boast against the one who cuts with it, 
or the saw magnify itself against the one who moves it to and fro? 
\textit{As if a rod should move the one who lifts it}!\lebnote{Literally “As a rod waving even the one who lifts it up”} 
\textit{As if a staff should lift up that which is not wood}!\lebnote{Literally “As a staff lifting up not wood”}
\verse Therefore the Lord, Adonai of hosts, will send leanness among his sturdy warriors, 
and a burning like the burning of fire will burn under his glory.
\verse And the light of Israel will become like a fire, 
and his holy one like a flame, 
and it will burn and devour his thorns\lebnote{Hebrew “thorn”} and briers\lebnote{Hebrew “brier”} in one day.
\verse And he will destroy the glory of his forest and orchard \textit{completely},\lebnote{Literally “from soul and to body”} 
and it will be like the wasting away of one who is sick.
\verseWithHeading{The Return of the Remnant} And this shall happen: on that day, the remnant of Israel and the survivors\lebnote{Hebrew “survivor”} of the house of Jacob will not continue to lean on the one who struck it 
but will lean on Adonai, the holy one of Israel, in truth.
\verse A remnant will return— 
the remnant of Jacob—to the mighty God.
\verse For though your people Israel was like the sand of the sea, 
only a remnant of it will return. 
Annihilation is determined, 
overflowing with righteousness.
\verse For the Lord Adonai of hosts is about to make a complete destruction 
and a determined end in the midst of all the earth.
\verse Therefore thus says the Lord Adonai of hosts:
\verse “My people who live in Zion, 
you must not be afraid of Assyria. 
It beats you with the rod, 
and it lifts up its staff against you \textit{as the Egyptians did}.\lebnote{Literally “in the way of Egypt”}
\verse My indignation will come to an end \textit{in just a very little while},\lebnote{Literally “for still a little a trifle”} 
and my anger will be directed to their destruction.”
\verse And this shall happen: on that day,
\verse he will remove his burden from your shoulder 
and his yoke from your neck, 
\textit{and a yoke will be destroyed because of fat}.\lebnote{The meaning of this phrase is uncertain, leading to the conjecture that it belongs with the next verse and by a different word division could mean “he has gone up from Jeshimon”; alternatively, “fat” could be a metaphor for prosperity}
\verse He has come to Aiath, 
he has passed through Migron; 
at Micmash he deposited his baggage.
\verse They crossed over the pass; 
Geba is a place of overnight lodging for us. 
Ramah trembles; 
Gibeah of Saul has fled.
\verse Daughter of Gallim, cry out with your voice; 
Laishah, listen! 
\textit{Anathoth is poor}.\lebnote{Or with different vocalization, “Answer her, Anathoth!” which fits the parallelism better}
\verse Madmenah flees! 
The inhabitants of Gebim bring themselves into safety!
\verse \textit{This day}\lebnote{Literally “Yet today”} \textit{taking a stand}\lebnote{Literally “to stand”} at Nob, 
he will shake his fist at the mountain of the daughter\lebnote{Following the reading tradition (\textit{Qere}); the consonantal text has “house”} of Zion, 
at the hill of Jerusalem.
\verse Look! The Lord Adonai of hosts is about to lop off the branches\lebnote{Hebrew “branch”} with great power, 
and \textit{the towering trees}\lebnote{Literally “the haughty of the height”} will be felled, 
and the \textit{tall trees}\lebnote{Literally “height”} will be brought low.
\end{biblechapter}

\begin{biblechapter} % Isaiah 11
\verseWithHeading{The Branch’s Righteous Reign} And a shoot will come out from the stump of Jesse, 
and a branch from its roots will bear fruit.
\verse And the spirit of Adonai shall rest on him— 
a spirit of wisdom and understanding, 
a spirit of counsel and might, 
a spirit of knowledge and the fear of Adonai.
\verse And his \textit{breath}\lebnote{Hebrew “smelling,” perhaps metaphorically as “delight” as in Amos 5:14; perhaps this line was accidentally copied twice from the preceding line} is in the fear of Adonai. 
And he shall judge not by his eyesight, 
and he shall rebuke not by \textit{what he hears with}\lebnote{Literally “the rumors of”} his ears.
\verse But he shall judge the poor with righteousness, 
and he shall decide for the needy of the earth with rectitude. 
And he shall strike the earth with the rod of his mouth, 
and he shall kill the wicked person with the breath of his lips.
\verse And righteousness shall be the belt around his waist, 
and faithfulness the belt around his loins.
\verse And a wolf shall \textit{stay}\lebnote{Literally “dwell as an alien”} with a lamb, 
and a leopard shall lie down with a kid, 
and a calf and a lion and a fatling together 
\textit{as a small boy leads}\lebnote{Literally “and a small boy leading”} them.
\verse And a cow and a bear shall graze; 
their young shall lie down together. 
And a lion shall eat straw like the cattle.
\verse And \textit{an infant}\lebnote{Literally “one nursing”} shall play over a serpent’s hole, 
and \textit{a toddler}\lebnote{Literally “one who is weaned”} shall put his hand on a viper’s hole.
\verse They will not injure and they will not destroy on all of my \textit{holy mountain},\lebnote{Literally “mountain of holiness”} 
for the earth will be full of the knowledge of Adonai, 
as the waters cover the sea.
\verse And this shall happen on that day: the nations shall inquire of the root of Jesse, 
which shall be standing as a signal to the peoples, 
and his resting place shall be glorious.
\verseWithHeading{The Regathered Remnant} And this shall happen on that day:
\verse The Lord will again extend his hand a second time 
to acquire the remnant of his people that is left, 
from Assyria, Egypt, Pathros, Cush, Elam, Shinar, Hamath, and the coastlands of the sea,
\verse and he will raise a signal for the nations. 
And he will gather the outcasts of Israel, 
and he will gather the scattered ones of Judah together from the four \textit{corners}\lebnote{Literally “wings”} of the earth.
\verse And the jealousy of Ephraim shall depart, 
and the enemies of Judah shall be cut off. 
Ephraim shall not be jealous of Judah, 
and Judah shall not be an enemy of Ephraim.
\verse But they shall \textit{swoop}\lebnote{Literally “fly”} upon the Philistine shoulder,\lebnote{That is, the hill country west of Judah} \textit{westward}.\lebnote{Literally “sea”} 
Together they shall plunder the sons of the east. 
Edom and Moab \textit{will be under their command},\lebnote{Literally “the stretching of their hand”} 
and the sons of Ammon will be their subjugated people.
\verse And Adonai will divide\lebnote{Or “utterly destroy”} the tongue\lebnote{Probably the “gulf”} of the sea of Egypt,\lebnote{That is, the Red Sea} 
and he will wave his hand over the river\lebnote{That is, the Euphrates River} with his scorching wind; 
and he will strike it into seven streams, 
and he will \textit{make it passable by foot}.\lebnote{Literally “cause to walk with the sandal”}
\end{biblechapter}

\begin{biblechapter} % Isaiah 12
\verseWithHeading{A Song of Thanksgiving} And you will say on that day, 
“I will give you thanks, Adonai, 
for though you were angry with me, 
your anger turned away, 
and you comforted me.
\verse Look! God is my salvation; 
I will trust, and I will not be afraid, 
for my strength and might is Yah, Adonai; 
and he has become salvation for me.”
\verse And you will draw water from the wells of salvation in joy.
\verse And you will say on that day,
\verse “Give thanks to Adonai; 
call on his name. 
Make his deeds known among the peoples; 
bring to remembrance that his name is exalted.
\verse Sing praises to Adonai, for he has done a glorious thing; 
this is known in all the earth.
\end{biblechapter}

\begin{biblechapter} % Isaiah 13
\verseWithHeading{An Oracle against Babylon} The oracle of Babylon that Isaiah son of Amoz saw:
\verse Raise a signal on a bare hill, 
lift up your voice to them; 
wave the hand and may they enter the gateways of the noblemen.
\verse I myself I have commanded my consecrated ones, 
I have also summoned my mighty warriors concerning my anger, 
the \textit{ones who exalt over}\lebnote{Literally “exultant of”} my majesty.
\verse A sound, a noise is on the mountains, 
the likeness of many people! 
A sound of the roar of the kingdoms, 
of nations gathering! 
Adonai of hosts is mustering an army for battle.
\verse They are coming from a distant land, 
from the end of the heavens, 
Adonai and the weapons of his indignation, 
to destroy \textit{the whole earth}.\lebnote{Literally “all of the land/earth”}
\verse Wail, for the day of Adonai is near; 
it will come like destruction from Shaddai!\lebnote{Often translated “the Almighty”}
\verse Therefore all hands will grow slack, 
and every human heart will melt,
\verse and they will be dismayed. 
Pangs and labor pains will seize them; 
they will tremble like a woman giving birth. 
\textit{They will stare at one another},\lebnote{Literally “A man will stare at his neighbor”} 
\textit{their faces flushing}.\lebnote{Literally “faces of flames”}
\verse Look! The day of Adonai is coming, 
cruel and wrath and the burning of anger, 
to make the earth a desolation, 
and he will destroy its sinners from it.
\verse For the stars of the heavens and their constellations will not flash forth their light; 
the sun will keep back when it comes out, 
and the moon will not cause its light to shine.
\verse And I will punish the world for its evil 
and the wicked for their iniquity. 
And I will put an end to the pride of the arrogant, 
and I will bring the haughtiness of tyrants low.
\verse I will make humanity more rare than gold 
and humankind more than the gold of Ophir.
\verse Therefore I will make the heavens tremble, 
and the earth will quake from its place 
because of the wrath of Adonai of hosts, 
and in the day his anger burns.
\verse And this shall happen:
\verse like a hunted gazelle or sheep \textit{with none to gather them},\lebnote{Literally “and without one who gathers”} 
they will each turn to his own people, 
and they will each flee to his own land.
\verse Everyone who is found will be pierced through, 
and everyone who is carried away will fall by the sword.
\verse And their children will be dashed into pieces before their eyes; 
their houses will be plundered, and their wives will be raped.\lebnote{The reading tradition (\textit{Qere}) softens this to “slept with”}
\verse Look! I am stirring the Medes up against them, 
who do not value silver 
and do not delight in gold.
\verse And Babylon, the glory of kingdoms, the splendor of the Chaldeans’ pride, will be like \textit{when God overthrew}\lebnote{Literally “the overthrow by God of”} Sodom and Gomorrah.
\verse It will not be inhabited forever, 
and it will not be dwelled in \textit{forever};\lebnote{Literally “until generation and generation”} 
and no Arab will pitch a tent there, 
and shepherds will not allow their flocks to lie down there.
\verse But wild animals will lie down there, 
and their houses will be full of howling creatures, 
and the daughters of ostriches\lebnote{Hebrew “ostrich”} will live there, 
and goats will dance there.
\verse And hyenas will answer in its palaces, 
and jackals in the pleasure palaces; 
and its time \textit{is coming soon},\lebnote{Literally “close to come”} 
and its days will not be prolonged.
\end{biblechapter}

\begin{biblechapter} % Isaiah 14
\verseWithHeading{The Restoration of Israel} But Adonai will have compassion on Jacob, 
and he will again choose Israel 
and set them on their land, 
and the immigrant will join himself to them, 
and they will attach themselves to the house of Jacob.
\verse And the nations will take them 
and bring them to their place, 
and the house of Israel will take possession of them\lebnote{That is, the nations} in the land of Adonai 
as slaves and female slaves. And this will happen: they will take their captors captive 
and rule over their oppressors.
\verseWithHeading{The Downfall of the King of Babylon} And it shall happen on the day Adonai gives you rest from your pain and turmoil and hard labor which \textit{you had to perform},\lebnote{Literally “was worked by you”}
\verse that you will take this taunt against the king of Babylon, 
and you will say: 
“How the oppressor has ceased! 
his insolence\lebnote{The meaning of the word is uncertain; others translate it as “fury,” “hostility,” or even “golden city”} has ceased.
\verse Adonai has broken the staff of the wicked, 
the scepter of rulers,
\verse that struck the peoples in wrath, a blow without ceasing, 
that ruled the nations in anger 
\textit{with unrestrained persecution}.\lebnote{Literally “persecution without withholding”}
\verse All of the earth rests and is quiet; 
they break forth into singing.
\verse Even the cypresses rejoice over you, 
the cedars of Lebanon: 
‘Since you were laid down, 
no wood cutter comes up against us.’
\verse Sheol below is getting excited over you, 
to meet \textit{you when you come};\lebnote{Literally “your entrance”} 
it arouses the dead spirits for you, 
all of the leaders of the earth. 
It raises all of the kings of the nations from their thrones.
\verse All of them will respond and say to you, 
‘You yourself also were made weak like us! 
You have become the same as us!’
\verse Your pride is brought down to Sheol, 
and the sound of your harps; 
maggots\lebnote{Hebrew “maggot”} are spread out beneath you like a bed, 
and your covering is worms.\lebnote{Hebrew “worm”}
\verse How you have fallen from heaven, morning star, son of dawn! 
You are cut down to the ground, conqueror of nations!
\verse And you yourself said in your heart,
\verse ‘I will ascend to heaven; 
I will raise up my throne above the stars of God; 
and I will sit on the mountain of assembly 
on the summit of Zaphon;\lebnote{Or “of the north”}
\verse I will ascend to the high places of the clouds,\lebnote{Hebrew “cloud”} 
I will make myself like the Most High.’
\verse But you are brought down to Sheol, 
to the depths of the pit.
\verse Those who see you will stare at you, 
they will look closely at you: 
‘Is this the man who made the earth tremble, 
who caused kingdoms to shake,
\verse who made the world like the desert 
and destroyed its cities, 
who would not let his prisoners go home?’
\verse All the kings of the nations, all of them, lie in glory, 
each one in his house.
\verse But as for you, you are thrown away from your grave, 
like an abhorrent shoot, 
clothed with the slain, 
those pierced by the sword, 
those who go down to the stones of the pit, 
like a corpse that is trodden down. The descendants\lebnote{Hebrew “descendant”} of evildoers will not be mentioned for eternity!
\verse Prepare a place of slaughter for his sons 
because of the sin of their ancestors.\lebnote{Or “fathers”} 
Let them not rise and take possession of the earth 
or fill up the face of the world with cities.”
\verse “And I will rise up against them,” \textit{declares}\lebnote{Literally “declaration of”} Adonai of hosts, “and I will cut off name and a remnant from Babylon, 
and offspring and posterity,” \textit{declares}\lebnote{Literally “declaration of”} Adonai.
\verse “And I will make her a possession of the hedgehog, 
and pools of water, 
and I will sweep her away with the broom of destruction,” \textit{declares}\lebnote{Literally “declaration of”} Adonai of hosts.
\verseWithHeading{Oracle of Judgment on Assyria} Adonai of hosts has sworn, saying,
\verse “\textit{Surely}\lebnote{The oath formula begins literally “If not”} just as I have intended, so it shall be. 
And just as I have planned, it shall stand:
\verse to break Assyria in my land, 
and I will trample him down on my mountains; 
and he shall remove his yoke from them, 
and he shall remove his burden from his\lebnote{That is, their} shoulders.”\lebnote{Hebrew “shoulder”}
\verse This is the plan that is planned concerning all of the earth; 
and this is the hand that is stretched out over all of the nations.
\verseWithHeading{Oracle of Judgment on Philistia} In the year of the death of king Ahaz there was this oracle:
\verse You must not rejoice, all you Philistines, 
that the rod that struck you is broken, 
for a viper will come forth from the root of the snake, 
and its fruit will be a flying serpent.
\verse And the firstborn of the poor will graze, 
and the needy will lie down in security; 
but I will cause your root to die in famine, 
and it will kill your remnant.
\verse Wail, gate! Cry, city! 
Melt,\lebnote{With fear or despair} Philistia, all of you! 
For smoke is coming from the north, 
and there is no straggler in his ranks.
\verse And what will one answer the messengers of the nation? That Adonai has founded Zion, 
and the needy of his people will take refuge in it.
\end{biblechapter}

\begin{biblechapter} % Isaiah 15
\verseWithHeading{Oracle of Judgment on Moab} An oracle of Moab:
\verse Because Ar is devastated in a night, Moab is destroyed; 
because Kir of Moab is devastated in a night, it is destroyed.
\verse It has gone up to the house,\lebnote{Or “temple”} 
and Dibon to the high places 
for weeping over Nebo, 
and Moab wails over Medeba. 
\textit{Every head is bald},\lebnote{Literally “On all its heads baldness”} 
every beard is shaved.
\verse They gird themselves with sackcloth in its streets; 
on its roofs and public squares everyone wails, 
going down in weeping.
\verse And Heshbon and Elealeh cry\lebnote{The Hebrew is singular} out, 
their voice is heard as far as Jahaz; 
therefore the armed ones of Moab cry out; 
its soul quivers\lebnote{The Hebrew words for “cry out” and “quiver” are similar} for him.
\verse My heart cries out for Moab; 
its fugitives flee up to Zoar, 
to Eglath-shelishiyah. 
For at the ascent of Luhith it goes up it with weeping; 
for on the road of Horonaim they raise up a cry of destruction.
\verse For the waters of Nimrim are wastelands; 
for the grass has withered, 
the vegetation has vanished, 
there is no greenness.
\verse Therefore they carry the abundance it has made 
and their store of goods over the river of the poplars.
\verse For a cry for help has encircled the territory of Moab, 
her wailing is heard as far as Eglaim, 
and her wailing as far as Beer Elim.
\end{biblechapter}

\begin{biblechapter} % Isaiah 16
\verse Send a ram\lebnote{Possibly “rams” if a scribe accidentally omitted the Hebrew plural ending because the next word begins with that same letter} to the ruler of the land, 
from Sela across the desert to the mountain of \textit{daughter Zion}.\lebnote{Literally “the daughter of Zion”}
\verse And this shall be:
\verse like a bird fleeing from a thrust away nest 
shall be the daughters of Moab at the fords of Arnon.
\verse “Bring counsel, 
make a decision; 
make your shade like the night 
in the middle of noonday. 
Hide the outcasts; 
you must not betray the fugitive.
\verse Let my outcasts of Moab dwell as aliens among you; 
be a hiding place for them from the presence of the destroyer.” 
When the oppressor is no more, destruction has stopped, 
the \textit{one who tramples has}\lebnote{Literally “one who tramples have,” with mismatched singular and plural} disappeared from the land,
\verse then a throne shall be established in steadfast love, 
and one shall sit on it in faithfulness, in the tent of David, 
judging and seeking justice 
and zealous for righteousness.
\verse We have heard of the pride of Moab—exceedingly proud— 
of his arrogance, pride, and insolence; \textit{his boasting is not true}.\lebnote{Literally “not so his boasting”}
\verse Therefore Moab wails; 
all of it wails for Moab, 
for the raisin cakes of Kir-hareseth you moan, \textit{utterly devastated}.\lebnote{Literally “surely destroyed”}
\verse For Heshbon withers the fields, the vine of Sibmah; 
rulers of nations have broken down her tendrils, 
they reached up to Jazer, 
they wandered to the desert; 
her shoots spread abroad, 
they crossed over the sea.
\verse Therefore I weep with the weeping of Jazer for the vine of Sibmah. 
I drench you with my tears,\lebnote{Hebrew “tear”} Heshbon and Elealeh, 
for a jubilant shout has fallen over your summer fruit and harvest.
\verse And joy and gladness are\lebnote{The Hebrew is singular} taken away from the fruitful land, 
and in the vineyards no one exults, 
no one shouts for joy; 
no treader treads wine in the presses; 
I have put to an end to the jubilant shout.
\verse And this shall happen: when Moab appears, when it is weary upon the high place 
and it comes to its sanctuary to pray, it will not prevail.
\verse This was the word that Adonai spoke to Moab \textit{in the past}.\lebnote{Literally “from then”}
\verse But now Adonai speaks, saying, “In three years, like the years of a hired worker, the glory of Moab will become contemptible, with all of the great multitude, and the remnant will be a few, small, not strong.
\end{biblechapter}

\begin{biblechapter} % Isaiah 17
\verseWithHeading{Oracle of Judgment on Damascus} An oracle of Damascus:
\verse “Look! Damascus will cease being a city 
and will become a heap of ruins.
\verse The cities of Aroer will be deserted;\lebnote{These words in Hebrew (and “flocks” in the next line) all begin with the same letter, Ayin} 
they will be for the flocks, 
and they will lie down and \textit{no one will frighten}\lebnote{Literally “there is not one who frightens”} them. \textit{declares}\lebnote{Literally “declaration of”} Adonai of hosts.
\verse “And this shall happen:
\verse On that day, the glory of Jacob will be brought low, 
and the fat of his flesh will become lean.
\verse And gleanings will be left over in it, as \textit{when an olive tree is beaten},\lebnote{Literally “beating of an olive tree”} two or three ripe olive berries in the top of a branch, 
four or five on its fruitful branches,” \textit{declares}\lebnote{Literally “declaration of”} Adonai, the God of Israel.
\verse On that day, mankind will look to its maker, 
and its eyes will look to the holy one of Israel;
\verse it will not look to the altars, 
the work of its hands, 
and it will not see what its fingers made 
and the poles of Asherah worship and the incense altars.
\verse On that day, \textit{its fortified cities}\lebnote{Literally “the cities of his fortress”} will be like the \textit{abandonment of the wooded place and the summit},\lebnote{Perhaps this difficult phrase originally read “abandonment of the wooded heights of the Amorites”} which they deserted because of the children of Israel; and there will be desolation.
\verse For you have forgotten the God of your salvation, 
and you have not remembered the rock of your refuge; 
therefore you plant plants of pleasantness, 
and you \textit{plant}\lebnote{Literally “plant it”} a vine of a foreigner.
\verse On your planting day you make them grow, 
and in the morning of your sowing you bring them into bloom, 
yet the harvest will flee\lebnote{Reading the same consonants as a verb, \textit{nad}, rather than the noun \textit{ned,} which would mean “a heap ofthe harvest”} in a day of sickness and incurable pain.
\verseWithHeading{The Roar of the Peoples} Ah! The noise of many peoples, they make a noise like the noise of the seas! 
And the roar of nations, they roar like the roar of mighty waters!
\verse The nations roar like the roar of many waters, 
but he will rebuke him, and he will flee far away. 
And they are chased like chaff of the mountains before the wind 
and like tumbleweed before the storm.
\verse At the time of evening, and look, terror! 
Before morning he is no more. 
This is the fate of those who plunder us 
and the lot of those who plunder us.
\end{biblechapter}

\begin{biblechapter} % Isaiah 18
\verseWithHeading{Oracle of Judgment on Cush} Ah! land of the whirring of wings, 
which is beyond the rivers of Cush,
\verse that sends messengers by the sea 
and in vessels of papyrus on the surface of the waters! 
Go, swift messengers, to a \textit{tall}\lebnote{Literally “stretched out”} and \textit{smooth}\lebnote{Literally “peeled,” “wiped clean,” or “hairless”} nation, 
to a people feared \textit{near and far},\lebnote{Literally “from this and beyond”} 
a \textit{mighty, mighty}\lebnote{Perhaps this is a gibberish phrase in Hebrew, \textit{qaw-qaw|i|\lebnote{: and trampling\lebnote{Meaning uncertain} nation, 
whose land rivers divide.
\verse All you inhabitants of the world 
and dwellers of the earth, 
\textit{when a signal is raised on the}\lebnote{Literally “as a raising of a signal”} mountains, you must look, 
and \textit{when a trumpet is blown},\lebnote{Literally “as a blowing of a trumpet”} you must listen!
\verse For Adonai said this to me:
\verse “I will be quiet, 
and I will look from my dwelling place 
like \textit{clear heat because of light},\lebnote{The meaning of this phrase is uncertain; perhaps the “light” is sunlight} 
like a cloud of dew in the heat of harvest.”
\verse For before the harvest, \textit{when the blossom is complete}\lebnote{Literally “as a completion of a blossom”} 
and a blossom becomes ripening fruit, 
and one cuts off the shoots with pruning hooks, 
and one removes, tears away the tendrils.
\verse At that time, a gift\lebnote{Perhaps tribute is meant} will be brought to Adonai of hosts from a \textit{tall}\lebnote{Literally “stretched out”} and \textit{smooth}\lebnote{Literally “peeled,” “wiped clean,” or “hairless”} people, 
and from a people feared near and far, 
a mighty, mighty and trampling nation, 
whose land the rivers divide, 
to the place of the name of Adonai of hosts, 
the mountain of Zion.
\end{biblechapter}

\begin{biblechapter} % Isaiah 19
\verseWithHeading{Oracle of Judgment on Egypt} An oracle of Egypt:
\verse Look! Adonai is riding on a swift cloud 
and is coming to Egypt. 
And the idols of Egypt will tremble in front of him, 
and the heart of Egypt melts in his inner parts.
\verse “And I will stir up \textit{Egyptians}\lebnote{Literally “Egypt”} against \textit{Egyptians},\lebnote{Literally “Egypt”} 
and each one will fight against his brother 
and each one against his neighbor, 
city against city, kingdom against kingdom.
\verse And the spirit of \textit{the Egyptians}\lebnote{Literally “Egypt”} will be disturbed in his midst, 
and I will \textit{confuse}\lebnote{Literally “engulf”} his plans,\lebnote{Hebrew “plan”} 
and they will consult the idols and the spirits of the dead, 
and the ghosts and the spiritists.
\verse And I will hand over \textit{the Egyptians}\lebnote{Literally “Egypt”} into the hand of a hard master, 
and a powerful king will rule over them,” 
\textit{declares}\lebnote{Literally “declaration of”} the Lord Adonai of hosts.
\verse And the waters will be dried up from the sea, 
and the river will be parched and dry.
\verse And the rivers will become foul-smelling; 
the branches of the Nile of Egypt will become little and dry up; 
reed and rush will wither.
\verse Bare places by the Nile will be dried up, 
by the edge of the Nile and all the sown land of the Nile; 
it will be driven about, and \textit{it will be no more}.\lebnote{Literally “there is none of it”}
\verse And the fishers will mourn, 
and all of those who cast fishhooks\lebnote{Hebrew “fishhook”} in the Nile will lament, 
and those who spread out fishing nets\lebnote{Hebrew “net”} on the \textit{surface}\lebnote{Literally “face”} of the water will languish.
\verse And the workers of combed flax will be ashamed, 
and those who weave white linen.
\verse And her weavers will be crushed; 
all the \textit{hired workers}\lebnote{Literally “workers of wage”} will be grieved of heart.
\verse Surely the princes of Zoan are foolish; 
the wise of the counselors of Pharaoh give senseless counsel. 
How can you say to Pharaoh, 
“I myself am a son of sages, 
a \textit{descendant}\lebnote{Literally “son”} of ancient kings”?
\verse Where are your sages then? 
Now, let them tell you, 
and let them know what Adonai of hosts has planned against Egypt.
\verse The princes of Zoan have become foolish; 
the princes of Memphis are deceived; 
the leaders of her tribes have led Egypt astray.
\verse Adonai has mixed a spirit of confusion into her midst, 
and they have caused Egypt to stagger in all of its doings, 
as \textit{when a drunkard staggers}\lebnote{Literally “the staggering of a drunkard”} in his vomit.
\verseWithHeading{The Future Blessing of Egypt, Assyria, and Egypt} On that day, Egypt will be like women, and will tremble and be afraid before the waving hand of Adonai of hosts that he is waving against it.
\verse And the land of Judah will become a terror to Egypt, everyone to whom one mentions it will be afraid in himself because of the plan of Adonai of hosts that he is planning against him.
\verse On that day, there will be five cities in the land of Egypt that speak the \textit{language}\lebnote{Literally “lip”} of Canaan and swear an oath to Adonai of hosts. One will be called “City of the Sun.”
\verse On that day, there will be an altar for Adonai in in the middle of the land of Egypt, and a stone pillar for Adonai at her border.
\verse And it will be a sign and a witness to Adonai of hosts in the land of Egypt; when they cry out to Adonai because of oppressors, he will send them a savior and a \textit{defender},\lebnote{Literally “contender”} and he will deliver them.
\verse And Adonai will make himself known to Egypt, and Egypt will know Adonai on that day, and they will worship with sacrifice and offering, and they will make vows\lebnote{Hebrew “vow”} to Adonai, and they will fulfill them.
\verse And Adonai will strike Egypt, striking and healing; and they will return to Adonai, and he will respond to their prayer, and he will heal them.
\verse On that day, there will be a highway from Egypt to Assyria, and Assyria will come into Egypt, and Egypt into Assyria, and Egypt will worship together with Assyria.
\verse On that day, Israel will be the third with Egypt and Assyria, a blessing in the midst of the earth,
\verse whom Adonai of hosts blessed, saying, “May Egypt my people be blessed, and Assyria, the work of my hands, and my inheritance, Israel.”
\end{biblechapter}

\begin{biblechapter} % Isaiah 20
\verseWithHeading{A Sign regarding Egypt and Cush} In the year \textit{the commander-in-chief came}\lebnote{Literally “of the coming of the commander in chief”} to Ashdod, when Sargon the king of Assyria sent him, and he fought against Ashdod and he took it,
\verse at that time, Adonai had spoken by the hand of Isaiah son of Amoz, saying, “Go and loosen the sackcloth from your loins, 
and take off your sandals\lebnote{Hebrew “sandal”} from your feet,” and he had done so, walking naked and barefoot.
\verse Then\lebnote{Or “And”} Adonai said, “Just as my servant Isaiah has walked naked and barefoot three years as a sign and a portent against Egypt and Cush,
\verse so shall the king of Assyria lead the captives\lebnote{Hebrew “captive”} of Egypt and the exiles of Cush, young and old, naked and barefoot, \textit{with bared buttocks},\lebnote{Literally “and bare of buttocks”} the shame of Egypt.
\verse And they shall be dismayed, 
and they shall be ashamed 
because of Cush, their hope, 
and because of Egypt, their pride.
\verse And the inhabitant\lebnote{Hebrew “inhabitant”} of the coastland will say this on that day: ‘Look! This is our hope to whom we fled for help, to be delivered \textit{from}\lebnote{Literally “from the face of”} the king of Assyria, and how shall we escape?’ ”
\end{biblechapter}

\begin{biblechapter} % Isaiah 21
\verseWithHeading{The Fall of Babylon} The oracle of the wilderness of the sea: As storm winds passing over in the Negev, \textit{it comes}\lebnote{Literally “to come”} from the desert, 
from a frightful land.
\verse A hard revelation is told to me; the treacherous deals treacherously, 
and the destroyer destroys. 
Go up, Elam; 
lay siege, Media! I put an end to all of her sighing.
\verse Therefore my loins are filled with anguish; 
pangs have seized me, like the pangs of a woman giving birth. 
I am \textit{too bent to hear},\lebnote{Literally “bent from hearing,” which could also mean “bent because of what I heard”} 
I am \textit{too dismayed to see}.\lebnote{Literally “dismayed from seeing,” which could also mean “dismayed because of what I saw”}
\verse My \textit{mind}\lebnote{Literally “heart”} staggers; fear terrifies me; 
the twilight \textit{I desired}\lebnote{Literally “of my desire”} \textit{brought}\lebnote{Literally “put”} me fear.
\verse Set out the table in order! 
Spread out the rugs!\lebnote{Hebrew “rug”} 
Eat! drink! 
Rise up, commanders; 
smear\lebnote{That is, prepare with oil} the shield!
\verse For the Lord said this to me:
\verse “Go, set a watchman in position. 
He must announce what he sees.
\verse Then\lebnote{Or “And”} the \textit{watchman}\lebnote{Literally “lion”} called,
\verse “Lord, I am standing on the watchtower continually by day, 
and I am standing at my post \textit{throughout}\lebnote{Literally “all of”} the night. Then\lebnote{Or “and”} he responded and said,
\verse “It has fallen! Babylon has fallen! 
And all the images of her gods are\lebnote{Hebrew “is”} smashed on the ground!” I will announce to you what I have heard from Adonai of hosts, the God of Israel.
\verseWithHeading{Oracle regarding Dumah} The oracle of Dumah: One is calling to me from Seir, “Watchman, what \textit{of}\lebnote{Literally “from”} the night? 
Watchman, what \textit{of}\lebnote{Literally “from”} the night?”
\verse The watchman says, “Morning comes, 
and also the night. 
If you will inquire, inquire; 
\textit{come back again}.”\lebnote{Literally “return come”}
\verseWithHeading{Oracle against Arabia} An oracle \textit{concerning}\lebnote{Literally “in”} Arabia:
\verse You will spend the night in the thicket in a desert-plateau,\lebnote{The Hebrew for “in a desert-plateau” is the same as for “concerning Arabia” in the preceding line} 
caravans of Dedanites.
\verse \textit{When you happen to meet}\lebnote{Literally “Toward encountering”} the thirsty, bring water. 
Inhabitants of the land of Tema came to meet the fugitive with his bread.
\verse For the Lord said this to me: “\textit{In one more year},\lebnote{Literally “In yet a year”} like the years of a hired worker,
\verse all the glory of Kedar will come to an end. For Adonai, the God of Israel, has spoken.
\end{biblechapter}

\begin{biblechapter} % Isaiah 22
\verseWithHeading{Oracle against Jerusalem} The oracle of the valley of vision: \textit{What business do you have going}\lebnote{Literally “What to you then that you have gone”} up, all of you, to the roofs,
\verse \textit{noisy},\lebnote{Literally “full of noises”} tumultuous city, 
exultant town? 
Your slain are not slain by the sword, 
nor are they dead from battle.
\verse All of your rulers have fled together without a bow; 
all of \textit{you who were found}\lebnote{Literally “your found ones”} were captured. 
They were captured together; 
they had fled far away.
\verse Therefore I said,
\verse “Look away from me, 
let me \textit{weep bitterly};\lebnote{Literally “be bitter in weeping”} 
you must not insist on comforting me 
for the destruction of the daughter of my people.”
\verse For the Lord Adonai of hosts has a day of tumult and trampling and confusion\lebnote{These three Hebrew words are alliterative: \textit{m}\textit{e}\textit{hûm}\textit{â}\textit{ûm}\textit{e}\textit{ḇûs}\textit{â}\textit{ûm}\textit{e}\textit{ḇû}\textit{ḵâ|i|\lebnote{: in the valley of vision, 
a tearing down of walls\lebnote{Hebrew “wall”} and a cry for help to the mountains.\lebnote{Hebrew “mountain”}
\verse And this happened:
\verse the choicest of your valleys were full of chariots,\lebnote{Hebrew “chariot”} 
and the cavalry confidently stood at the gate.
\verse And he uncovered the covering of Judah. 
And you looked, on that day, to the weapons\lebnote{Hebrew “weapon”} of the House of the Forest,
\verse and you saw that the breaches in the walls of the city of David were many, 
and you gathered the waters of the lower pool.
\verse And you counted the houses of Jerusalem, 
and you broke down the houses to fortify the wall.
\verse And the Lord, Adonai of hosts, called on that day
\verse for weeping and mourning, 
and for baldness and girding \textit{with}\lebnote{Literally “of”} sackcloth.
\verse And it was revealed in my ears by Adonai of hosts: “Surely this sin will not be atoned for you until you die!” 
says the Lord, Adonai of hosts.
\verseWithHeading{Oracle regarding Shebna} The Lord, Adonai of hosts, says this:
\verse “Go! Go to this steward, to Shebna, 
who is over the house:
\verse ‘What \textit{business do you have}\lebnote{Literally “to you”} here, 
and who \textit{do you have}\lebnote{Literally “to you”} here, 
that you have cut a grave cutting here for yourself, 
carving his grave on the height, 
a dwelling place for him in the rock?
\verse Look! Adonai is about to \textit{really hurl}\lebnote{Literally “hurl a hurling”} you, man! 
And he is about to grasp you firmly;
\verse he will wind a winding tightly around you like a ball, to a \textit{wide land}.\lebnote{Literally “land broad of sides”} 
There you shall die, 
and there the chariots of your splendor will be, 
disgrace to your master’s house!
\verse And this shall happen:
\verse On that day I will call to my servant, Eliakim son of Hilkiah,
\verse and I will clothe him with your tunic, 
and I will bind your sash firmly about him, 
and I will put your authority into his hand, 
and he shall be like a father to the inhabitants\lebnote{Hebrew “inhabitant”} of Jerusalem 
and to the house of Judah.
\verse And I will put the key of the house of David on his shoulder, 
and he shall open and no one will be able to shut; 
and he shall shut and no one will be able to open.
\verse And I will drive him in like a peg into a secure place, 
and he will become like a throne of glory to the house of his father.
\verse On that day, \textit{declares}\lebnote{Literally “declaration of”} Adonai of hosts, the peg that was driven will move away into a secure place, and it will be cut down and fall, and the load that was on her will be cut off. For Adonai has spoken.’ ”
\end{biblechapter}

\begin{biblechapter} % Isaiah 23
\verseWithHeading{Oracle against Tyre} The oracle of Tyre:
\verse Wail, ships of Tarshish, for \textit{the house is destroyed so that no one can enter};\lebnote{Literally “it is destroyed from a house from entering”} 
it is announced to them from the land of Cyprus.
\verse Be still, inhabitants of the coast, 
merchant of Sidon, who travels over the sea, they filled you.
\verse And on the great waters is the seed of Shihor, 
the harvest of the Nile is its produce, 
and she was the merchandise of the nations.
\verse Be ashamed, Sidon, for the sea, 
the fortress of the sea said, saying, 
“I was not in labor, 
and I have not given birth, 
and I have not reared young men, 
brought\lebnote{Or “I have brought”} up young women.”
\verse When the report comes to Egypt, 
they will be in anguish at the report \textit{about}\lebnote{Literally “of”} Tyre.
\verse Cross over to Tarshish! 
Wail, inhabitants of the coast!
\verse \textit{Is this your exultant one},\lebnote{Literally “this to you, exultant”} her origin from the days of long ago? 
Her feet brought her to dwell afar as an alien.
\verse Who has planned this against Tyre, the bestower of crowns,\lebnote{Hebrew “crown”} 
whose merchants were princes, 
her traders the honored ones of the earth?
\verse Adonai of hosts has planned it: 
to defile the pride of all glory, 
to humble all the honored ones of the earth.
\verse Cross over your own land like the Nile, daughter of Tarshish; 
there is no longer a harbor.\lebnote{Hebrew “waistband” is emended to “harbor” by transposing two consonants}
\verse And he said,
\verse “You will not \textit{continue}\lebnote{Literally “do again longer”} to exult, crushed one, \textit{virgin}\lebnote{Literally “virgin of the”} daughter of Sidon. 
Arise, cross over to Cyprus! There will be no rest for you even there.”
\verse Look at the land of the Chaldeans! 
\textit{This people no longer exists. 
Assyria destined}\lebnote{Literally “this people was not Assyria destined,” which could be translated “It was this people! It was not Assyria. They destined”} it for wild animals. 
They erected its siege towers; 
they demolished its citadel fortresses. 
It made her like a ruin.
\verse And this will happen on that day: And Tyre will be forgotten seventy years,\lebnote{Hebrew “year”} like the days of one king. At the end of seventy years,\lebnote{Hebrew “year”} it will be for Tyre like the song of the prostitute:
\verse “Take a harp, 
go around the city, forgotten prostitute! 
Do it well, playing a stringed instrument! 
Make numerous songs,\lebnote{Hebrew “song”} that you may be remembered.”
\verse And this shall happen: at the end of seventy years,\lebnote{Hebrew “year”} Adonai will visit Tyre, and she will return to her harlot’s wages, 
and she will commit fornication 
with all the kingdoms of the world on the face of the earth.
\verse And this shall happen: her merchandise and her harlot’s wages will be set apart for Adonai; it will not be stored up, 
and it will not be hoarded, 
but her merchandise will be for those who live before the \textit{presence}\lebnote{Literally “face”} of Adonai, 
for eating to satiation and for fine clothing.
\end{biblechapter}

\begin{biblechapter} % Isaiah 24
\verseWithHeading{The Judgment on the Earth} Look! Adonai is about to lay the earth waste 
and is about to devastate it, 
and he will twist her surface, 
and he will scatter her inhabitants.
\verse And it shall be as with the people, so with the priest; 
as with the slave, so with his master; 
as with the female slave, so with her mistress; 
as with the buyer, so with the seller; 
as with the lender, so with the borrower; 
as with the creditor, so with the one to whom he lends.
\verse The earth shall be utterly laid waste, 
and it shall be utterly plundered, for Adonai has spoken this word.
\verse The earth dries up, it withers; 
the world languishes, it withers. 
The elevated of the people of the earth languish,
\verse and the earth is defiled beneath its inhabitants. 
For they have transgressed laws; 
they have passed by statutes;\lebnote{Hebrew “statute”} 
they have broken the everlasting covenant.
\verse Therefore a curse devours the earth, 
and the inhabitants in it suffer for their guilt. 
Therefore the inhabitants of the earth burn, 
and \textit{few men are left}.\lebnote{Literally “he is left a man of smallness”}
\verse The new wine dries up; 
the vine languishes. 
All the merry of heart sigh;
\verse the joy of the tambourine has stopped. 
The noise of the jubilant has ceased; 
the joy of the lyre has stopped.
\verse They do not drink wine with song; 
strong drink is bitter to those who drink it.
\verse The city of emptiness is broken; 
every house is shut \textit{so that no one can enter};\lebnote{Literally “from entering”}
\verse there is an outcry over the wine in the streets. 
All joy turns into \textit{darkness};\lebnote{Literally “evening”} 
the joy of the earth disappears.
\verse Desolation is left in the city; 
the gate is crushed into a state of ruin.
\verse For it shall be like this in the midst of the earth, 
among the nations, 
like the beating of an olive tree, 
like gleanings when a grape harvest is at an end.
\verse They lift up their voices;\lebnote{Hebrew “voice”} they sing for joy; 
they shout out from the west over the majesty of Adonai.
\verse Therefore glorify Adonai in the east, 
the name of Adonai the God of Israel in the coastlands of the sea.
\verse We hear songs from the \textit{edge}\lebnote{Literally “wing”} of the earth: 
“Glory to the righteous one!” 
But I say, “Ruin to me! 
Ruin to me! Woe to me! 
The treacherous ones deal treacherously, 
and the treacherous ones deal treacherously with treachery!”
\verse Terror and the pit and the snare are upon you, inhabitants\lebnote{Hebrew “inhabitant”} of the earth!
\verse And this shall happen:
\verse The one who flees from the sound of the terror shall fall into the pit, 
and the one who goes up from \textit{inside the pit}\lebnote{Literally “the middle of the pit”} shall be caught in the snare, 
for the windows from heaven are opened, 
and the foundations of the earth tremble.
\verse The earth is utterly broken; 
the earth is torn asunder; 
the earth is shaken violently.
\verse And this shall happen on that day:
\verse Adonai will punish the host of heaven in heaven, 
and the kings of the earth on the earth.
\verse And they will be gathered in a gathering, like a prisoner \textit{in}\lebnote{Literally “on”} a pit. 
And they will be shut \textit{in}\lebnote{Literally “on”} a prison and be punished \textit{after}\lebnote{Literally “from”} many days.
\end{biblechapter}

\begin{biblechapter} % Isaiah 25
\verseWithHeading{Praise for Salvation} Adonai, you are my God; I will exalt you. 
I will praise your name, 
for you have done wonderful things,\lebnote{Hebrew “thing”} 
plans\lebnote{Hebrew “plan”} \textit{of old},\lebnote{Literally “from far away “} in faithfulness, trustworthiness.
\verse For you have \textit{made}\lebnote{Literally “placed”} the city a heap, 
the fortified city a ruin, 
the palace of foreigners \textit{is no longer}\lebnote{Literally “from”} a city; 
it will \textit{never}\lebnote{Literally “to eternity not”} be rebuilt.
\verse Therefore strong peoples will glorify you; 
a city of ruthless nations, they will fear you.
\verse For you have been a refuge to the poor, 
a refuge to the needy in his distress, 
a shelter from the rainstorm, 
a shade from the heat. 
For the breath of the ruthless was like a rainstorm against a wall,\lebnote{The same consonants with different vowels can be translated “winter rainstorm”}
\verse the noise of foreigners like heat in a dry land. 
You subdued the heat with the shade of a cloud; 
the song of the ruthless was silenced.
\verse And on this mountain Adonai of hosts will make for all peoples a \textit{rich feast},\lebnote{Literally “feast of fat”} 
a feast of aged wines, fat filled with marrow, filtered aged wine.
\verse And on this mountain he will \textit{destroy}\lebnote{Literally “engulf”} the face of the shroud, 
the shroud over all peoples, 
and the woven covering over all nations.
\verse He will \textit{destroy}\lebnote{Literally “engulf”} death forever, 
and the Lord Adonai will wipe off the tears from all faces, 
and he will remove the disgrace of his people from all the earth, for Adonai has spoken.
\verse And one will say, on that day,
\verse “Look! This is our God! We have waited for him and he saved us! 
This is Adonai; we waited for him! 
Let us be glad, 
and let us rejoice in his salvation.”
\verse For the hand of Adonai will rest on this mountain, 
and Moab shall be trampled down under him 
as \textit{a heap of straw is trampled down}\lebnote{Literally “the trampling down of a heap of straw”} in \textit{waters of}\lebnote{These words are in the written Hebrew text, but not in the reading tradition (\textit{Qere})} a dung heap.
\verse And it\lebnote{Moab} will spread out its hands in the midst of it, 
just as the swimmer spreads out to swim, 
and its pride will be brought low with the movement\lebnote{Meaning uncertain} of its hands.
\end{biblechapter}

\begin{biblechapter} % Isaiah 26
\verseWithHeading{Judah’s Song of Praise} On that day, this song will be sung in the land of Judah: “\textit{We have a strong city};\lebnote{Literally “a city of strength for us”} 
he sets up victory like walls and ramparts!\lebnote{Hebrew “rampart”}
\verse Open the gates \textit{so that}\lebnote{Or “and”} the righteous nation who keeps faithfulness may enter!
\verse You will protect a firm inclination in peace, 
in peace because he trusts in you.
\verse Trust in Adonai forever, 
for in Yah, Adonai you have an everlasting rock.
\verse For he has thrown down the inhabitants of the height, 
he lays low the lofty city. 
He lays it low to the ground; 
he throws her to the dust.
\verse The foot tramples it, 
the feet of the poor, the steps of the needy.”
\verseWithHeading{Adonai’s People Vindicated} The way of the righteous is a straight path; 
\textit{you clear the level path of the righteous}.\lebnote{Literally “the level path of the righteous you make level”; the meaning of this phrase is uncertain}
\verse Surely we wait for you in the path of your judgments, Adonai, 
for your name and renown are the desire of the soul.
\verse I desire you with all my soul in the night; 
also I seek you with my spirit within me, 
for when your judgments are \textit{upon}\lebnote{Literally “to”} the earth, 
the inhabitants of the world learn righteousness.
\verse Though the wicked person is shown compassion, he does not learn righteousness; he acts unjustly in the land of uprightness, 
and he does not see the majesty of Adonai.
\verse Adonai, though your hand reaches high, they do not see it.
\verse Let them see, and let them be ashamed of the zeal of people. 
Indeed, let the fire of your enemies consume them.
\verse Adonai, you will establish peace for us, 
for you have done even all of our works for us.
\verse Adonai, our God, lords besides you ruled over us, 
\textit{but we praise your name alone}.\lebnote{Literally “alone in you we praise name you”}
\verse The dead do not live; 
dead spirits do not rise 
because you have punished and destroyed them, 
and you have destroyed all memory of them.
\verse You have added to the nation, Adonai. 
You have added to the nation; you are honored. 
You have extended all the ends of the land.
\verse Adonai, in distress they have visited you; 
they poured out an \textit{incantation};\lebnote{Literally “whispering”} 
your discipline was on them. So we were because of your presence, Adonai.
\verse We became pregnant, we writhed; 
we gave birth to wind. 
We cannot bring about deliverance on the earth, 
and no inhabitants of the world are born.
\verse Your dead shall live; \textit{their corpses}\lebnote{Literally “my corpse;” some manuscripts propose an emendation to the suffix to change it to third person masculine plural (“their corpses”), which makes better sense} shall rise. 
Wake up and sing for joy, dwellers of the dust, 
for your dew is \textit{celestial dew},\lebnote{Literally “dew of light”} 
and the earth will give birth to dead spirits.\lebnote{Or “you make the land of Rephaim fall”}
\verse Go, my people, enter into your chambers 
and shut your doors\lebnote{The reading tradition (\textit{Qere}) is singular} behind you; 
hide for \textit{a very little}\lebnote{Literally “a little a little”} while, 
until the wrath has passed over.
\verse For look! Adonai is about to come out from his place 
to punish the iniquity of the inhabitants\lebnote{Hebrew “inhabitant”} of the earth against him, 
and the earth will disclose her blood 
and will no longer cover her slain.
\end{biblechapter}

\begin{biblechapter} % Isaiah 27
\verseWithHeading{Israel Rescued} On that day, Adonai will punish with his cruel, great and strong sword Leviathan, the fleeing serpent, and Leviathan, the twisting serpent, and he will kill the sea monster that is in the sea.
\verse On that day:
\verse “A vineyard of beauty! Sing in praise of it!
\verse I, Adonai, am her keeper; 
I water it \textit{again and again}.\lebnote{Literally “by moments”} 
Lest one afflict harm on it, 
I guard it night and day;
\verse \textit{I have no wrath}.\lebnote{Literally “There is no wrath for me”} 
Whatever gives me thorns and briers, 
I will step forth against in battle. 
I will set it on fire altogether.
\verse Or let it grasp at my protection; 
let it make peace with me; 
peace let it make with me.”
\verse \textit{Days are coming},\lebnote{Literally “The coming ones”} let Jacob take root; 
Israel will blossom and send out shoots, 
and they will fill the face of the world with fruit.
\verse Does he strike him \textit{as he struck down those who struck him}?\lebnote{Literally “like the striking of one striking him”} 
Or is he killed \textit{as those who killed him were killed}?\lebnote{Literally “like the killing of his killed ones”}
\verse By expelling\lebnote{Meaning uncertain, derived from the following parallel expression} her, by her sending away, you argue with her. 
He removed them with his strong wind, 
in the day of the east wind.
\verse Therefore by this he will make atonement for the guilt of Jacob, 
and this will be all of the fruit of the removal of his sin: 
\textit{when he makes}\lebnote{Literally “in his setting”} all the stones of the altar like crushed stones of chalk, 
no poles of Asherah worship or incense altars will stand.
\verse For the fortified city is solitary, 
a settlement deserted and forsaken, like the wilderness; 
the calf grazes there, 
lies down there and destroys its branches. For it is not a people of understanding; therefore his maker will not have compassion on him, 
and his creator will not show him favor.
\verse And this shall happen: on that day, Adonai will thresh from the floodwaters of the Euphrates to the wadi\lebnote{A seasonal stream that is often dry} of Egypt, and as for you, you will be gathered one by one, sons of Israel.
\verse And this shall happen: on that day, a great trumpet will be blown, and those who were lost in the land of Assyria will come, 
and those who were scattered in the land of Egypt, and they will bow down to Adonai on the \textit{holy mountain}\lebnote{Literally “mountain of holiness”} at Jerusalem.
\end{biblechapter}

\begin{biblechapter} % Isaiah 28
\verseWithHeading{Judgment against the Leaders of Ephraim} Ah! The garland of the pride of the drunkards of Ephraim 
and the withering flower of the glory of his beauty, 
which is at the head of \textit{the rich valley},\lebnote{Literally “a valley of fat”} 
\textit{ones overcome}\lebnote{Literally “ones struck”} with wine!
\verse Look! The Lord has a mighty and strong one, 
like a rainstorm of hail, a wind storm of destruction, 
like a rainstorm of mighty overflowing waters, 
he will put them to the earth with his hand.
\verse The garland of the pride of the drunkards of Ephraim 
will be trampled\lebnote{The Hebrew is plural} by feet,
\verse and the withering flower of the glory of its beauty 
which is at the head of \textit{the rich valley}\lebnote{Literally “a valley of fat”} 
will be like its early fig before summer, 
\textit{which the one who sees it swallows}\lebnote{Literally “Which, when the one who sees it sees it, he swallows it”} while it is still in his hand.
\verse In that day, Adonai of hosts will become a garland of glory 
and a diadem of beauty to the remnant of his people,
\verse and a spirit of justice to the one who sits over judgment, 
and strength to those who turn back the battle at the gate.
\verse And these also stagger because of wine 
and stagger because of strong drink; 
priest and prophet stagger because of strong drink; 
they are confused\lebnote{Or “swallowed”} because of wine. 
They stagger because of strong drink; 
they err in vision. 
They stagger in the rendering of a decision,
\verse for all the tables are full of disgusting vomit, \textit{with no place left}.\lebnote{Literally “without a place”}
\verse To whom will he teach knowledge, 
and to whom will he explain the message? 
Those who are weaned from milk, 
those taken from the breast?
\verse For it is blah-blah upon blah-blah, 
blah-blah upon blah-blah, 
gah-gah upon gah-gah, 
gah-gah upon gah-gah,\lebnote{In this context, the Hebrew expressions \textit{tsaw-tsaw} and \textit{qaw-qaw} are likely meant to sound like baby talk, but they could mean “command upon command” and “rule upon rule”} 
a little here, a little there.
\verse For he will speak with \textit{stammering}\lebnote{Literally “stammer of lip”} and another tongue to this people,
\verse to whom he has said, “This is rest; give rest to the weary; 
and this is repose”; yet\lebnote{Or “and”} they were not willing to hear.
\verse And to them the word of Adonai will be blah-blah upon blah-blah 
blah-blah upon blah-blah 
gah-gah upon gah-gah 
gah-gah upon gah-gah,\lebnote{In this context, the Hebrew expressions \textit{tsaw-tsaw} and \textit{qaw-qaw} are likely meant to sound like baby talk, but they could mean “command upon command” and “rule upon rule”} 
a little here, a little there, 
so that they may go and stumble backward 
and be broken and ensnared and captured.
\verseWithHeading{The Cornerstone of Zion} Therefore hear the word of Adonai, men of mockery, 
rulers of these people in Jerusalem:
\verse Because you have said, “We have \textit{made}\lebnote{Literally “cut”} a covenant with death, 
and we have made an agreement with Sheol. The overwhelming flood, when it passes through, will not come to us, for we have made lies\lebnote{Hebrew “lie”} our refuge, 
and we have hidden ourselves in falsehood.”
\verse Therefore the Lord Adonai says this: “Look! I am laying a stone in Zion, 
a \textit{tested}\lebnote{Literally “testing”} stone, 
a precious cornerstone, 
a founded foundation: ‘The one who trusts will not \textit{panic}.’\lebnote{Literally “hurry”}
\verse And I will make justice the measuring line, 
and righteousness the plumb line; 
and hail will sweep away the refuge of lies,\lebnote{Hebrew “lie”} 
and waters will flood over the hiding place.
\verse And your covenant with death will be annulled, 
and your agreement with Sheol will not stand; 
you will become a trampling place for the overwhelming flood when it passes through.
\verse \textit{As often as it passes through},\lebnote{Literally “From enough his passing through”} it will take you, 
for \textit{morning by morning}\lebnote{Literally “in the morning in the morning”} it will pass through, 
in the day and in the night, and understanding the message will be only terror.”
\verse For the bed is too short \textit{to stretch out on},\lebnote{Literally “from stretching oneself”} 
and the covering is too narrow when wrapping oneself.
\verse For Adonai will rise up as at Mount Perazim; 
he will rave as in the valley at Gibeon 
to do his deed—his deed is strange— 
and to work his work—his work is alien!
\verse And now you must not scoff, 
or your bonds will be strong, for I have heard from the Lord, Adonai of hosts: complete destruction 
decreed upon all the land.
\verseWithHeading{A Parable of Adonai’s Work} Listen, and hear my voice! 
Listen attentively, and hear my word!
\verse Is it all day that the plowman plows, 
opens to sow seed, harrows his ground?
\verse When he has leveled its \textit{surface},\lebnote{Literally “face”} does he not scatter dill, 
and sow cumin seed, 
and \textit{plant}\lebnote{Literally “place”} wheat in planted rows,\lebnote{Hebrew “row”} 
and barley in an appointed place, 
and spelt grain as its border?
\verse And his God instructs him about the prescription;\lebnote{That is, “the proper way”} 
he teaches him.
\verse For dill is not threshed with a threshing sledge, 
nor is a wheel of a utility cart \textit{rolled}\lebnote{Literally “turned”} over cumin, 
but dill is beaten out with a stick, 
and cumin with a rod.
\verse Grain is crushed fine, 
but certainly one does not thresh it forever; 
and one drives the wheel of his cart, 
but his horses do not crush it.
\verse This also comes forth from Adonai of hosts. He is wonderful in advice; 
he makes great wisdom.
\end{biblechapter}

\begin{biblechapter} % Isaiah 29
\verseWithHeading{Woe to Jerusalem} Ah! Ariel, Ariel, the city where David encamped!
\verse Add year to year, 
let festivals recur.
\verse Yet\lebnote{Or “And”} I will inflict Ariel, 
and there shall be mourning and lamentation, 
and it shall be to me like an altar hearth.\lebnote{Hebrew “Ariel,” which might mean “altar hearth”}
\verse And I will encamp \textit{in a circle}\lebnote{The Hebrew text literally reads “like the ball”; the LXX supports an emendation to “David”} against you, 
and I will lay siege to you with towers\lebnote{Hebrew “tower”} 
and I will raise up siegeworks against you.
\verse Then\lebnote{Or “And”} you shall be low; 
you shall speak from the earth, 
and your words\lebnote{Hebrew “word”} will be low, from dust. 
And your voice will be from the earth, like a ghost, 
and your word will whisper from the dust. And it will happen in an instant, suddenly.
\verse You will be punished by Adonai of hosts
\verse with thunder and earthquake and great sound, 
storm wind and tempest and the flame of a devouring fire.
\verse And the multitude of all the nations who fight against Ariel, 
all those who fight against her and her stronghold, 
and those who inflict her shall be like a dream, a vision of the night.
\verse And it shall be as when the hungry person dreams—look, he is eating! 
And he wakes up and his inner self is empty. 
Or\lebnote{Or “And”} as when the thirsty person dreams—look, he is drinking! 
And he wakes up and look, he is faint, 
and his inner self is longing for water. 
So shall be the multitude of all the nations who fight against Mount Zion.
\verse Be astonished and be amazed! 
Blind yourselves and be blinded! 
They are drunk but\lebnote{Or “and”} not from wine; 
they stagger but\lebnote{Or “and”} not from strong drink.
\verse And the vision of all this has become for you like the words of a sealed document. When they give it to one who knows the\lebnote{The reading tradition (\textit{Qere}) has “a” instead of “the”} document, saying, “\textit{Read}\lebnote{Literally “Call”} this now!” He says, “I am not able, for it is sealed.”
\verse And if the document is given to one who does not \textit{know how to read},\lebnote{Literally “know a document”} saying, “\textit{Read}\lebnote{Literally “Call”} this now!” he says, “I do not \textit{know how to read}.”\lebnote{Literally “know a document”}
\verse And the Lord said, “Because this people draw near with its mouth, 
and with its lips it honors me, 
and its heart is far from me, 
and their fear of me is a commandment of men that has been taught,
\verse therefore look, I am again doing something spectacular
\verse and a spectacle with this spectacular people. 
And the wisdom of its wise men shall perish, 
and the discernment of its discerning ones shall keep itself hidden.”
\verse Ah! Those who make a plan deep, to hide it from Adonai, 
and their deeds are\lebnote{The Hebrew is singular} in a dark place. 
And they say, “Who sees us? 
And who knows us?”
\verseWithHeading{Blessing after Punishment} \textit{In a very little while}\lebnote{Literally “Not still a trifling of time”} shall not Lebanon change into a fruitful land, 
and the fruitful land be regarded as a forest?
\verse And on that day, the deaf shall hear the words of a scroll, 
and the eyes of the blind shall see out of gloom and darkness.
\verse And the meek \textit{have joy after joy}\lebnote{Literally “shall add joy”} in Adonai, 
and the needy of the people shall rejoice in the holy one of Israel.
\verse For the tyrant shall be no more, 
and the scoffer shall come to an end. 
And all those lying in wait for evil shall be cut off;
\verse those who mislead a person into sin with a word 
and set a trap for the \textit{arbitrator}\lebnote{Literally “one who argues”} in the gate 
and guide away the righteous by emptiness.
\verse Therefore Adonai, who redeemed Abraham, says this to the house of Jacob:
\verse “Jacob will no longer be ashamed, 
and his face will no longer grow pale.
\verse For when he sees his children, 
the work of my hands, in his midst, 
they will treat my name as holy, 
and they will treat the holy one of Jacob as holy, 
and they will stand in awe of the God of Israel.
\end{biblechapter}

\begin{biblechapter} % Isaiah 30
\verseWithHeading{Warning against Alliance with Egypt} “Oh rebellious children!” \textit{declares}\lebnote{Literally “declaration of”} Adonai,
\verse “to make a plan, but\lebnote{Or “and”} not from me, 
and pour out a libation, but\lebnote{Or “and”} not from my Spirit, 
\textit{so as to add}\lebnote{Literally “for the sake of adding”} sin to sin.
\verse Who go to go down to Egypt, 
but\lebnote{Or “and”} they do not ask of my mouth, 
to take refuge in the protection of Pharaoh 
and to take refuge in the shadow of Egypt.
\verse And the protection of Pharaoh shall be shame to you, 
and the refuge in the shadow of Egypt, humiliation.
\verse Everyone will start to stink because of a people that cannot profit them, not for help and not for profiting, 
but for shame and also for disgrace.”
\verseWithHeading{Oracle regarding the Negev} An oracle of the animals of the Negev:
\verse Through a land of trouble and distress, 
of lioness and lion, 
\textit{among}\lebnote{Literally “from”} them are snake and flying serpent; 
they carry their wealth on the \textit{backs}\lebnote{Literally “shoulder”} of male donkeys 
and their treasures on the humps\lebnote{Hebrew “hump”} of camels, 
to a people that cannot profit them.
\verse For\lebnote{Or “And”} \textit{Egyptians}\lebnote{Literally “Egypt”} help with vanity and emptiness, 
therefore I have called this one “Rahab, they are sitting.”
\verse Now go, write it on a tablet with them, 
and inscribe it on a scroll, 
\textit{that}\lebnote{Or “and”} it may be \textit{for the time to come},\lebnote{Literally “for last day”} 
forever, forever.
\verse For it is a people of rebellion, deceitful children, 
children who are not willing to hear the instruction of Adonai,
\verse who say to those who do see, “You must not see!” 
and to the seers, “You must not see truth for us; 
speak smooth things to us, 
see illusions,
\verse Therefore the holy one of Israel says this:
\verse “Because you are rejecting this word 
and you trust in oppression and \textit{cunning}\lebnote{Literally “going wrong”} 
and you rely on it,
\verse therefore this iniquity shall come for you like a breach about to fall, 
bulging out on a high wall \textit{that breaks}\lebnote{Literally “whose breaking comes”} suddenly, in an instant.
\verse For the Lord Adonai, the holy one of Israel, said this: “In returning and rest you shall be saved; 
your strength shall be in quietness and in trust.”
\verse But\lebnote{Or “And”} you were not willing,
\verse “No! For we will flee on horses!”\lebnote{Hebrew “horse”} Therefore you shall flee! 
And, “We will ride on swift horses!”\lebnote{Hebrew “horse”} Therefore your pursuers shall be swift!
\verseWithHeading{Adonai Will Show Mercy} Therefore Adonai waits to be gracious to you, 
and therefore he will rise up to show you mercy, 
for Adonai is a God of justice; 
blessed are all those who wait for him.
\verse For people will live in Zion; 
in Jerusalem, you will surely not weep. 
Surely he will be gracious to you; 
when he hears the sound of your cry, he will answer you.
\verse And the Lord will give you the bread of distress 
and the water of oppression, 
but\lebnote{Or “and”} your teachers will not hide themselves\lebnote{Hebrew “himself”} any longer. 
And your eyes \textit{shall see}\lebnote{Literally “be seeing”} your teachers.
\verse And your ears shall hear a word from behind you, saying, 
“this is the way; walk in it,” 
when you go to your right 
and when you go to your left.
\verse And you will defile the plating of your silver idols 
and the covering of your gold image. 
You will scatter them like contaminated things;\lebnote{Hebrew “contaminated thing”} 
you will say to it, “Filth!”\lebnote{Or “Get out!”}
\verse And he will give rain for your seed with which you sow the ground, 
and grain, the produce of the ground, and it will be rich and \textit{fertile}.\lebnote{Literally “fat”} 
On that day, your cattle will graze in broad pastures;\lebnote{Hebrew “pasture”}
\verse and the oxen and the donkeys that till\lebnote{Or “tilling”} the ground will eat fodder, sorrel 
that has been winnowed with shovel and pitchfork.
\verse And there will be streams on every high mountain and elevated hill, 
watercourses of water, on a day of great slaughter, when towers fall.
\verse And the light of the full moon will be like the light of the sun, 
and the light of the sun will be sevenfold, like the light of seven days, 
on the day when Adonai binds up the breakage of his people, 
and he heals the wound of his blow.
\verseWithHeading{Judgment against Assyria} Look! The name of Adonai comes from afar, 
burning with his anger and heaviness of cloud. 
His lips are full of indignation, 
and his tongue is like a devouring fire.
\verse And his breath is like an overflowing river; 
it reaches up to the neck 
to shake the nations with the sieve of worthlessness; 
and a bridle that leads astray is on the jawbones of the peoples.
\verse \textit{You shall have a song}\lebnote{Literally “There shall be a song for you”} as in the night when a holy festival is kept, 
and a gladness of heart like one who goes with the flute, 
to go to the mountain of Adonai, to the rock of Israel.
\verse And Adonai will cause the majesty of his voice to be heard, 
and he will cause the descent of his arm to be seen, 
in furious anger and a flame of devouring fire, 
with a cloudburst and a rainstorm and stones\lebnote{Hebrew “stone”} of hail.
\verse Indeed, Assyria will be shattered by the voice of Adonai; 
he strikes with the rod.
\verse And every stroke of the staff of foundation\lebnote{Some translations emend this to “discipline”} that Adonai lays will be on it with timbrels and lyres, 
and he will fight against it\lebnote{The reading tradition (\textit{Qere}) has “them”} with battles of brandishing.
\verse For Topheth has been prepared from yesterday; 
indeed, it is made ready for the king. 
He makes its pile of wood deep and wide; 
he makes fire and wood \textit{abundant}.\lebnote{Literally “numerous”} 
The breath of Adonai burns in it like a stream of sulfur.
\end{biblechapter}

\begin{biblechapter} % Isaiah 31
\verseWithHeading{The Egyptians are No Help} Ah! Those who go down to Egypt for help! They rely on horses
\verse and trust in chariots because they are many, 
and in horsemen because they are very numerous, 
and they do not look to the holy one of Israel, 
and they do not consult Adonai.
\verse And indeed, he is wise, and he brings disaster, 
and he does not remove his words, 
and he will rise against the house of evildoers 
and against the help of workers of iniquity.
\verse For Adonai said this to me: “As which a lion growls 
and a young lion over its prey when \textit{a full group}\lebnote{Literally “fullness”} of shepherds is called against him, it is not terrified by their voice, 
and to their noise it does not respond, 
so Adonai of hosts will come down to fight upon Mount Zion 
and upon its hill.
\verse Like birds flying overhead, so Adonai of hosts will protect Jerusalem; he will protect and deliver it; 
he will pass over and rescue it.
\verse Turn back to the one against whom the sons of Israel have made deep rebellion.
\verse For on that day, each one will reject his idols of silver 
and his idols of gold which your hands have made in sin for you.
\verse And Assyria shall fall by a sword not of a man, 
and a sword not of a human shall devour him; 
and he shall flee from the sword, 
and his young men shall be put to forced labor.
\verse And his rock will pass over because of terror, 
and his officers will be terrified because of the flag,” 
\textit{declares}\lebnote{Literally “declaration of”} Adonai, 
who has a fire in Zion 
and has a furnace in Jerusalem.
\end{biblechapter}

\begin{biblechapter} % Isaiah 32
\verseWithHeading{The Kingdom of Righteousness} See, a king will rule according to righteousness, 
and princes will rule according to justice.
\verse And each one will be like a hiding place from the wind 
and a covering from the rainstorm, 
like streams of water in a dry region, 
like the shade of a large rock in a weary land.
\verse And the eyes of those who see will not gaze,\lebnote{The same consonants with different vowels would be translated “be blinded,” which fits the context better} 
and the ears of those who hear will listen.
\verse And the \textit{minds}\lebnote{Literally “heart”} of the rash will understand \textit{knowledge},\lebnote{Literally “to know”} 
and the tongues\lebnote{Hebrew “tongue”} of stammerers will hasten to speak clearly.
\verse A fool will no longer be called noble, 
and a scoundrel will not be said to be eminent.
\verse For a fool speaks folly, 
and his \textit{mind}\lebnote{Literally “heart”} does iniquity: 
to \textit{behave wickedly},\lebnote{Literally “do ungodliness”} 
and to speak error concerning Adonai, 
to leave the throat of the hungry empty, 
and he deprives the thirsty of drink.
\verse And a scoundrel, his weapons are evil; 
he plans evil devices 
to ruin the poor with words of deception 
even\lebnote{Or “and”} when the speech of the needy is right.
\verse But\lebnote{Or “And”} the nobleman plans noble things, 
and he stands upon noble things.
\verseWithHeading{Against the Carefree Women} Women who are at ease, rise up; hear my voice! 
Carefree daughters, listen to my word!
\verse \textit{In a year}\lebnote{Literally “In days upon a year”} you will tremble, carefree ones,
\verse for the vintage will come to an end; 
the harvest will not come.
\verse Tremble, you who are at ease; 
tremble, carefree ones; 
strip, and strip yourself, 
and gird yourself on your loins,\lebnote{Although the carefree and at ease addressees in this verse are feminine plural (“you women”), the commands are masculine singular in form}
\verse mourning over breasts, 
over fields of delight, 
over the fruitful vine,
\verse over the soil of my people. 
It goes up in thorns\lebnote{Hebrew “thorn”} and briers,\lebnote{Hebrew “brier”} 
indeed over all of the houses of joy in the jubilant city.
\verse Until a spirit is poured out on us from on high,
\verse and the wilderness becomes a fruitful field, 
and the fruitful field is reckoned as the forest.
\verse Then\lebnote{Or “And”} justice will dwell in the wilderness, 
and righteousness will live in the fruitful field.
\verse And the work of righteousness will be peace, 
and the work of righteousness, quietness and security forever.
\verse And my people will dwell in a settlement of peace 
and in a dwelling place of security 
and in undisturbed resting places.
\verse And it hails when \textit{the forest comes down},\lebnote{Literally “the coming down of the forest”} 
and the city will become low in humiliation.
\end{biblechapter}

\begin{biblechapter} % Isaiah 33
\verseWithHeading{Adonai’s Judgment and Help} Ah, destroyer, and yourself not destroyed! 
And treacherous one, and no one has dealt treacherously with him!\lebnote{Many manuscripts suggest “you” rather than “him”} 
When you \textit{cease}\lebnote{Literally “finishing”} destroying, you will be destroyed. 
When you stop dealing treacherously, one will deal treacherously with you.
\verse Adonai, be gracious to us, we wait for you. 
Be our\lebnote{Hebrew “their”} arm in the mornings, 
indeed our salvation in the time of trouble.
\verse At the sound of tumult, peoples fled; 
because of your exaltation, nations scattered.
\verse And your spoil is gathered, 
as the gathering of the locust, 
as a swarm of locusts storming on it.
\verse Adonai is exalted, for he dwells on high; 
he filled Zion with justice and righteousness,
\verse and he will be the security of your times, 
an abundance of salvation, wisdom, and knowledge. The fear of Adonai is his treasure.
\verse Look! Their heroes cry out in the street; 
the messengers of peace weep bitterly.
\verse Highways are deserted; 
the traveler on the road ceases. 
One breaks a treaty, 
he rejects the cities,\lebnote{A Dead Sea Scroll reads “witnesses”} 
he does not hold man in high regard.
\verse The land mourns; it languishes. 
Lebanon feels abashed; it withers. 
Sharon is like the desert, 
and Bashan and Carmel \textit{are losing their leaves}.\lebnote{Literally “shaking off”}
\verse “Now I will arise,” says Adonai. 
“Now I will lift myself up proudly; 
now I will raise myself.
\verse You conceive dry grass, you bring forth stubble; 
your breath is a fire; it will consume you.
\verse And the peoples will be burning to lime— 
they are burned like thorns that have been cut down in the fire.
\verse You who are far away, hear what I have done; 
and you who are near, know my might!”
\verse Sinners are afraid in Zion; 
trembling has seized the godless: 
“Who of us can \textit{live}\lebnote{Literally “dwell as an alien”} with devouring fire? 
Who of us can \textit{live}\lebnote{Literally “dwell as an alien”} with everlasting consuming hearths?”
\verse He who walks in righteousness 
and speaks uprightness, 
who rejects the gain of extortion, 
who \textit{refuses}\lebnote{Literally “shakes his hand from the taking of”} a bribe, 
who stops up his ears\lebnote{Hebrew “ear”} from hearing \textit{bloodshed}\lebnote{Literally “blood”} 
and shuts his eyes from seeing evil.
\verse That one will live on the heights; 
the fortresses of rocks will be his refuge. 
His food will be given; 
his waters\lebnote{Hebrew “water”} will endure.
\verse Your eyes will see the king in his beauty; 
they will see a \textit{distant land}.\lebnote{Literally “land of distance”}
\verse Your \textit{mind}\lebnote{Literally “heart”} will meditate on the terror: 
“Where is the one who counted? 
Where is the one who weighed out? 
Where is the one who counted the towers?”
\verse You will not see the insolent people, 
the people \textit{whose language is too obscure to understand},\lebnote{Literally “obscure of lip than to hear”} 
whose stammering of tongue \textit{cannot be understood}.\lebnote{Literally “there is no understanding”}
\verse Look on Zion, the city of our appointed festivals! 
Your eyes will see Jerusalem, 
an undisturbed settlement, 
a tent \textit{that is not moved}.\lebnote{Literally “not he is loaded”} 
No one will ever pull out its tent pegs, 
and none of its ropes will be torn in two.
\verse Rather, there Adonai will be mighty for us, 
a place of rivers and \textit{broad streams},\lebnote{Literally “streams broad of hands”} 
a galley ship with\lebnote{Or “of”} oars\lebnote{Hebrew “oar”} cannot go in it, 
and a mighty ship cannot pass through it.
\verse For Adonai is our judge; Adonai is our lawgiver. 
Adonai is our king; he is the one who will save us.
\verse Your riggings hang slack; 
they do not hold the base of their mast firm, 
they do not spread out the sail. 
Then the prey of spoil in abundance will be divided; 
the lame will take plunder.
\verse And no inhabitant will say, “I am sick”; 
the people who live in it, their iniquity will be taken away.
\end{biblechapter}

\begin{biblechapter} % Isaiah 34
\verseWithHeading{Judgment on the Nations} Come near, nations, to hear; 
and peoples, listen attentively! 
Let the earth hear, and that which fills it; 
the world and all its offspring.
\verse For the anger of Adonai is against all the nations, 
and his wrath is against all their armies; 
he has put them under a ban, 
he has given them up for slaughter.
\verse And their slain shall be cast out; 
as for\lebnote{Or “and”} their corpses, their stench shall go up. 
And the mountains shall melt with\lebnote{Or “from”} their blood,
\verse and all the host of heaven shall rot. 
And the skies shall roll up like a scroll, 
and all their host shall wither 
like the withering of a leaf from a vine, 
or\lebnote{Or “and”} like the withering from a fig tree.
\verse When my sword is drenched in the heavens, look! It will descend upon Edom, 
and upon the people of my ban, for judgment.
\verse \textit{Adonai has a sword};\lebnote{Literally “A sword for Adonai”}
\verse it is full of blood. 
It is covered with\lebnote{Or “from”} fat, 
with\lebnote{Or “from”} the blood of lambs and goats, 
with\lebnote{Or “from”} the fat of the kidneys of rams, 
for \textit{Adonai has a sacrifice}\lebnote{Literally “a sacrifice for Adonai”} in Bozrah 
and a great slaughter in the land of Edom.
\verse And wild oxen shall go down with them, 
and steers with strong bulls. 
And their land shall be drenched with\lebnote{Or “from”} blood, 
and their soil shall be fattened with\lebnote{Or “from”} fat.
\verse For \textit{Adonai has a day of vengeance},\lebnote{Literally “a day of vengeance for Adonai”} 
a year of retribution for the strife of Zion.
\verse And its streams shall be changed to pitch and its soil to sulfur, 
and its land shall become like burning pitch.
\verse Night and day it shall not be quenched; 
its smoke shall go up forever. 
From generation to generation it shall be in ruins; 
forever and ever there will be no one who passes through her.
\verse But\lebnote{Or “And”} the large bird and the hedgehog shall take possession of it, 
and the owl and the raven shall live in it. 
And he shall stretch the measuring line of confusion out over it, 
and the plumb line of emptiness.
\verse Its nobles—but no kingdom is there—shall call, 
and all its princes shall be nothing.
\verse And thorns shall go up her citadel fortress, 
weeds\lebnote{Hebrew “weed”} and thistle plants\lebnote{Hebrew “thistle plant”} in her fortresses; 
and it shall be the settlement of jackals, 
green grass for the daughters of an ostrich.
\verse And desert creatures shall meet with hyenas, 
and a goat-demon shall call to his neighbor; 
surely there Lilith\lebnote{Hebrew transliteration; possibly a proper name for a Mesopotamian night-demon} shall repose, 
and she shall find a resting place for herself.
\verse Seek from the book of Adonai and \textit{read};\lebnote{Literally “call”}
\verse none of these shall be missing; 
none shall miss her mate. 
For my\lebnote{The Dead Sea Scroll has “his”} mouth \textit{is the one that}\lebnote{Literally “he”} has commanded, 
and his spirit \textit{is the one that}\lebnote{Literally “he”} has gathered them.
\end{biblechapter}

\begin{biblechapter} % Isaiah 35
\verseWithHeading{The Ransomed Return to Zion} Wilderness and dry land shall be glad,\lebnote{The Hebrew has an extra “them,” probably from an accidental of the first letter of the next word} 
and desert shall rejoice and blossom like the crocus.
\verse It shall blossom abundantly, 
and it shall rejoice indeed with rejoicing and exulting. 
The glory of Lebanon shall be given to it, 
the majesty of Carmel and Sharon. 
\textit{They are the ones who}\lebnote{Literally “they”} shall see the glory of Adonai, 
the majesty of our God.
\verse Strengthen the weak hands 
and make the staggering knees firm.
\verse Say to those who are hasty of heart,
\verse “Be strong; you must not fear! 
Look! your God will come with vengeance, 
with \textit{divine retribution}.\lebnote{Literally “the retribution of gods”} 
He \textit{is the one who}\lebnote{Literally “he”} will come and save you.”
\verse Then the eyes of the blind shall be opened, 
and the ears of the deaf shall be opened.
\verse Then the lame shall leap like the deer, 
and the tongue of the dumb shall sing for joy, 
for waters shall break forth in the wilderness 
and streams in the desert.
\verse And the parched ground shall become a pool, 
and the thirsty ground springs of water. 
Her resting place is in a settlement of jackals; 
the grass shall become like reeds\lebnote{Hebrew “reed”} and rushes.\lebnote{Hebrew “rush”}
\verse And a highway shall be there, and a way, 
and it shall be called the way of holiness. 
The unclean shall not travel through it, 
but\lebnote{Or “and”} it is for them, he who walks on the way; 
and fools shall not wander about.
\verse No lion shall be there, 
and no ferocious wild beast shall go up it. 
It shall not be found there, 
but\lebnote{Or “and”} the redeemed shall walk there.
\end{biblechapter}

\begin{biblechapter} % Isaiah 36
\verseWithHeading{Sennacherib Threatens Jerusalem} And this happened: In \textit{the fourteenth year}\lebnote{Literally “four ten year”} of King Hezekiah, Sennacherib king of Assyria went up against all the fortified cities of Judah, and he captured them.
\verse And the king of Assyria sent Rabshakeh\lebnote{Rabshekah is the title of a high Assyrian official} from Lachish to Jerusalem, to King Hezekiah, with a large army, and he stood by the conduit of the upper pool on the highway of the field of the washer.
\verse And Eliakim son of Hilkiah, who \textit{was in charge of the palace},\lebnote{Literally “was over the house”} came out to him, and Shebna the secretary, and Joah son of Asaph, the reminder.
\verse And Rabshakeh said to them, “Now say to Hezekiah, ‘Thus says the great king, the king of Assyria: “What is this confidence in which you trust?
\verse I said, ‘Only a word of lips! \textit{War has power and a plan}!’\lebnote{The Hebrew here is awkward; literally “Plan and power for war”} Now, in whom do you trust, that you have rebelled against me?
\verse Look, you trust in the staff of this broken reed, on Egypt, which if a man leans on it, goes into his hand and bores through it! Such is Pharaoh, king of Egypt, to all those who trust in him.
\verse And if you say to me, ‘We trust in Adonai our God,’ was it not he whose high places and altars Hezekiah removed? And he said to Judah and to Jerusalem, ‘You shall bow down in the \textit{presence}\lebnote{Literally “face”} of this altar.’ ”
\verse And now please make a wager with my master the king of Assyria, and I will give you two thousand horses, that is, if you are able \textit{put}\lebnote{Literally “give”} riders for yourself on them!
\verse But how can you drive back \textit{one governor among the least of my master’s servants},\lebnote{Literally “the face of the governor of the one of the insignificant servants of my master”} when\lebnote{Or “and”} you trust in Egypt for chariots\lebnote{Hebrew “chariot”} and horsemen?
\verse And now was it without Adonai that I have come up against this land to destroy it? Adonai said to me, “Go up against this land and destroy it!” ’ ”
\verse And Eliakim, Shebna, and Joah said to Rabshakeh, “Please speak to your servants in Aramaic, for we can understand\lebnote{Or “hear”} it, and you must not speak to us in Judean in the \textit{hearing}\lebnote{Literally “ear”} of the people who are on the wall.”
\verse But\lebnote{Or “And”} Rabshakeh said, “Has my master sent me to speak these words to your masters and you? Was it not for the people who sit on the wall, to eat their dung and drink their urine\lebnote{So Masoretic Hebrew text (\textit{Kethib}); the reading tradition (\textit{Qere}) has “feet-water”} with you?”
\verse Then\lebnote{Or “And”} Rabshakeh stood and called in a great voice in Judean and said, “Hear the words of the great king, the king of Assyria.
\verse Thus says the king: ‘Do not let Hezekiah deceive you, for he will not be able to deliver you!
\verse And do not let Hezekiah make you rely on Adonai, saying, “Surely Adonai will deliver us; this city will not be given into the hand of the king of Assyria!”
\verse You must not listen to Hezekiah, for thus says the king of Assyria: “Make a blessing\lebnote{That is, a gesture of surrender} with me, and come out to me, and each one will eat from his vine and from his fig tree and drink water from\lebnote{Or “of”} his cistern,
\verse until \textit{I come}\lebnote{Literally “my coming”} and take you to a land like your land, a land of grain and new wine, a land of bread and vineyards,
\verse lest Hezekiah mislead you, saying, ‘Adonai will save us!’ Did the gods of the nations each save his land from the hand of the king of Assyria?
\verse Where are the gods of Hamath and Arpad? Where are the gods of Sepharvaim? Indeed, have they delivered Samaria from my hand?
\verse Who are there among all the gods of these countries who have saved their land from my hand, that Adonai should save Jerusalem from my hand?” ’ ”
\verse But\lebnote{Or “And”} they were silent and did not answer him a word, for the command of the king was, “You must not answer him.”
\verse Then\lebnote{Or “And”} Eliakim son of Hilkiah, who was over the \textit{palace},\lebnote{Literally “house”} Shebna the secretary, and Joah son of Asaph, the reminder, came to Hezekiah with torn garments and told him the words of Rabshakeh.
\end{biblechapter}

\begin{biblechapter} % Isaiah 37
\verseWithHeading{Hezekiah Consults Isaiah} And this happened: When King Hezekiah heard, he tore his garments, covered himself with sackcloth, and entered the temple\lebnote{Or “house”} of Adonai.
\verse And he sent Eliakim, who was \textit{in charge of}\lebnote{Literally “over”} the palace,\lebnote{Or “house”} and Shebna the secretary, and the elders of the priests \textit{covered}\lebnote{Literally “covering themselves”} with sackcloth to Isaiah son of Amoz, the prophet.
\verse And they said to him, “Thus says Hezekiah: ‘This day is a day of distress, rebuke, and disgrace, for children have come to the cervical opening, and there is no strength to give birth.
\verse Maybe Adonai your God heard the words of Rabshakeh whom the king of Assyria, his master, has sent to taunt the living God, and he will rebuke the words that Adonai your God hears. And you must lift up a prayer for the benefit of the remnant that is found.’ ”
\verse When\lebnote{Or “And”} the servants of King Hezekiah came to Isaiah,
\verse Isaiah said to them, “You must say this to your master: ‘Thus says Adonai: “You must not be afraid because of the words that you have heard, with which the servants of the king of Assyria have blasphemed me.
\verse Look! I am about to \textit{put}\lebnote{Literally “give”} a spirit in him \textit{so that}\lebnote{Or “and”} he shall hear a rumor and he shall return to his land, and I will cause him to fall by the sword in his land.” ’ ”
\verse And Rabshakeh returned and found the king of Assyria fighting against Libnah, for he had heard that he had left from Lachish.
\verse Now\lebnote{Or “And”} he\lebnote{That is, the king} heard concerning Tirhakah the king of Cush, saying, “He has set out to fight against\lebnote{Or “with”} you.”
\verse When\lebnote{Or “And”} he heard it, he sent messengers to Hezekiah, saying,
\verse “You shall say this to Hezekiah, king of Judah: ‘Do not let your God in whom you trust in him deceive you by saying, “Jerusalem will not be given into the hand of the king of Assyria.”
\verse Look! you have heard what the kings of Assyria have done to all lands to destroy them, and you—shall you be delivered?
\verse Did the gods of the nations that my \textit{predecessors}\lebnote{Literally “fathers”} destroyed deliver them—Gozan, Haran, Rezeph, and the sons of Eden who were in Telassar?
\verseWithHeading{Hezekiah’s Prayer} And Hezekiah took the letter from the hand of the messengers, and he read\lebnote{Or “called”} it. Then\lebnote{Or “And”} he went up to the temple\lebnote{Or “house”} of Adonai, and Hezekiah spread it out before the \textit{presence}\lebnote{Literally “face”} of Adonai.
\verse And Hezekiah prayed to Adonai, saying,
\verse “Adonai of hosts, God of Israel \textit{who is enthroned on}\lebnote{Literally “sitting”} the cherubim, you are the one, God by yourself, of all the kingdoms of the earth; you are the one who made the heavens and the earth.
\verse Adonai, extend your ear and hear! Adonai, open your eyes and see, and hear all the words of Sennacherib that he has sent to taunt the living God!
\verse Truly, Adonai, the kings of Assyria have laid waste all the lands\lebnote{The parallel text in 2 Kings has “nations”} and their land,
\verse to \textit{set}\lebnote{Literally “give”} their gods in the fire, for they were not gods, but the work of human hands, wood and stone, and they destroyed them.
\verse So\lebnote{Or “And”} now, Adonai, our God, save us from his hand so that\lebnote{Or “and”} all the kingdoms of the earth may know that you are Adonai, you alone!”
\verseWithHeading{God’s Answer} And Isaiah son of Amoz sent to Hezekiah, saying, “Thus says Adonai the God of Israel: ‘Because you have prayed to me \textit{concerning}\lebnote{Literally “to”} Sennacherib, king of Assyria,
\verse this is the word that Adonai has spoken concerning him:
\verse She shows contempt for you; she derides you, virgin daughter of Zion; 
she shakes her head behind you, daughter of Jerusalem. To the holy one of Israel!
\verse By the hand of your servants you have taunted the Lord, 
and you have said, “With my many chariots,\lebnote{Hebrew “chariot”} 
I myself have gone up the height of the mountains, 
to the remote areas of Lebanon. 
And I cut off \textit{its tall cedars},\lebnote{Literally “the height of its cedars”} 
the choicest of its junipers. 
And I came to the height of its limit, 
the forest of its orchard.\lebnote{Or “Carmel”}
\verse I myself dug and drank waters, 
and I caused all the streams of Egypt to dry up by the sole of my feet.”
\verse Have you not heard from \textit{a long time ago}?\lebnote{Literally “distant”} I have made it 
from days of primeval time, and I formed it. 
Now I have brought it about, 
and it is for fortified cities to collapse into heaps of destroyed stones.
\verse And their inhabitants are \textit{weak};\lebnote{Literally “short of hand”} 
they are dismayed, and they are ashamed; 
they have become like plants\lebnote{Hebrew “plant”} of the field, 
and like greens of grass, 
like grass on\lebnote{Or “of”} the roofs 
and a cultivated field \textit{before}\lebnote{Literally “before the face of”} the standing grain.
\verse And I know your sitting down and your going out, 
and your coming in, and your raging \textit{against}\lebnote{Literally “to”} me.
\verse Because you were enraged \textit{against}\lebnote{Literally “to”} me, 
and your noise\lebnote{Or “complacency”} has come up to\lebnote{Literally “in”} my ears, 
I will put my hook in your nose 
and my bridle on your lips, 
and I will turn you back 
on the way by which you came.
\verse And this shall be the sign for you: the eating of volunteer plants\lebnote{Hebrew “plant”} this\lebnote{Literally “the”} year, and in the second year self-seeded plants,\lebnote{Hebrew “plant”} and in the third year sow, reap, plant vineyards, and eat their fruit.
\verse And the remnant of the house of Judah that remain shall \textit{grow}\lebnote{Literally “add”} roots\lebnote{Hebrew “root”} downwards and make fruit upwards.
\verse For a remnant shall go out from Jerusalem 
and survivors\lebnote{Hebrew “survivor”} from mountain Zion. The zeal of Adonai of hosts will do this.’
\verse Therefore thus says Adonai concerning\lebnote{Or “to”} the king of Assyria:
\verse ‘He shall not come to this city, 
and he shall not shoot an arrow there, 
and he shall not meet it with a shield, 
and he shall not heap a siege ramp up against her. \textit{declares}\lebnote{Literally “declaration of”} Adonai.
\verse ‘And I will defend this city, to save it for my sake and for the sake of David, my servant.’ ”
\verseWithHeading{Sennacherib’s Defeat} And the angel of Adonai set out and struck one hundred and eighty-five thousand in the camp of the Assyrians. When\lebnote{Or “And”} they rose in the morning, look! All of them were dead corpses.
\verse Then\lebnote{Or “And”} Sennacherib king of Assyria left, went, and returned and lived at Nineveh.
\verse And this happened: he was bowing in worship in the house of Nisroch his god, and Adrammelech and Sharezer, his sons, struck him with the sword. And they themselves escaped to the land of Ararat, and Esar-haddon his son reigned as king in his place.
\end{biblechapter}

\begin{biblechapter} % Isaiah 38
\verseWithHeading{Hezekiah’s Illness} In those days, Hezekiah became sick to death, and Isaiah son of Amoz, the prophet, came to him and said to him, “Thus says Adonai: ‘Order your house, for you are about to die, and you shall not recover.’ ”
\verse Then\lebnote{Or “And”} Hezekiah turned his face to the wall and prayed to Adonai,
\verse and he said, “O Adonai, please remember how\lebnote{Or “that”} I have walked before your \textit{presence}\lebnote{Literally “face”} in faithfulness with a whole heart, and I have done the good in your eyes!” And Hezekiah wept with great weeping.
\verse Then\lebnote{Or “And”} the word of Adonai came\lebnote{Or “was”} to Isaiah, saying,
\verse “Go and say to Hezekiah, ‘Thus says Adonai, the God of David your ancestor:\lebnote{Or “father”} “I have heard your prayer; I have seen your tears. \textit{Look, I am going to}\lebnote{Literally, “Behold me; he will add,” but in this context it makes better sense to change the “he” to “I”; most translations follow this emendation} add fifteen years to your days.
\verse And I will deliver you and this city from the hand of the king of Assyria, and I will defend this city.” ’
\verse And this is the sign to you from Adonai, that Adonai will do this thing that he has spoken:
\verse Look! I will cause the shadow of the steps, which it had gone down on the steps of Ahaz with the sun, to turn backwards ten steps.” And the sun turned back ten steps on the steps which it had gone down.
\verse A writing of Hezekiah, king of Judah, when he was sick and had recovered from his sickness:
\verse I was the one who said, “I must go in the quiet of my days; 
I am summoned through the gates of Sheol for the rest of my years.”
\verse I said, “I shall not see Yah! Yah in the land of the living! 
I shall no more look at humankind among the inhabitants of the world.
\verse My dwelling place is pulled up and removed from me like the tent of my shepherd; 
I have rolled up my life like a weaver. 
He cuts me off from the thrum; 
from day to night you bring me to an end.
\verse I lie down\lebnote{Or “cry out”} until morning; 
like a lion, so he breaks all my bones; 
from day to night you bring me to an end.
\verse Like a horse or a crane, so I chirp; 
I moan like a dove. 
My eyes are weak toward the height. 
Lord, I have oppression; lend me support!
\verse What can I say? For\lebnote{Or “And”} he has spoken to me, 
and he himself has done it. 
I will walk slowly all my years because of the bitterness of my soul.
\verse Lord, they live by them, and the life of my spirit belongs to all among them. 
And restore me to health and keep me alive!
\verse Look! Bitterness was bitter to me for peace. 
And you were the one who loved\lebnote{Or possibly “kept back,” which sounds similar in Hebrew} my life from the pit of destruction, 
for you have cast all my sins behind your back.
\verse For Sheol cannot praise you; death cannot praise you. 
Those who go down to the pit cannot hope for your faithfulness.
\verse The living, the living one praises you like me today; 
a father will make your faithfulness known to children.
\verse Adonai, save me, and we will play my music on stringed instruments all the days of our lives at the temple\lebnote{Or “house”} of Adonai.”
\verse And Isaiah said, “Let them \textit{take}\lebnote{Literally “lift up”} a lump of figs, and let them rub it on the boil \textit{so that}\lebnote{Or “and”} he may recover.”
\verse And Hezekiah said, “What is the sign that I shall go up to the temple\lebnote{Or “house”} of Adonai?”
\end{biblechapter}

\begin{biblechapter} % Isaiah 39
\verseWithHeading{The Delegation from Babylon} At that time, Merodach-Baladan, son of Baladan, king of Babylon, sent letters and a present to Hezekiah, for\lebnote{Or “and”} he heard that he had been sick and recovered.
\verse And Hezekiah rejoiced over them and showed them his house of aromatic gum, the silver, gold, spices, good oil, all the house of his weapons, and all that was found in his storehouses. There was nothing that Hezekiah had not shown them in his house or in all his dominion.
\verse And Isaiah the prophet came to King Hezekiah and said to him, “What did these men say, and from where did they come to you?” And Hezekiah \textit{answered},\lebnote{Literally “said”} “They came to me from a distant country, from Babylon.”
\verse And he\lebnote{That is, Isaiah} said, “What have they seen in your house?” And Hezekiah \textit{answered},\lebnote{Literally “said”} “They have seen all that is in my house. There is nothing that I have not shown them in my storehouses.”
\verse And Isaiah said to Hezekiah, “Hear the word of Adonai of hosts:
\verse ‘Look! days are coming, and all that is in your house and that which your ancestors\lebnote{Or “fathers”} have stored up to this day shall be carried off to Babylon. Nothing shall be left,’ says Adonai.
\verse ‘And some of your sons who go out from you, whom you fathered, shall be taken, and they shall be eunuchs in the palace of the king of Babylon.’ ”
\verse And Hezekiah said to Isaiah, “The word of Adonai that you have spoken is good,” for\lebnote{Or “and”} he \textit{thought},\lebnote{Literally “said”} “Surely there will be peace and security in my days.”
\end{biblechapter}

\begin{biblechapter} % Isaiah 40
\verseWithHeading{Comfort for God’s People} “Comfort; comfort my people,” says your God.
\verse “Speak to the heart of Jerusalem, and call to her,
\verse that her compulsory labor is fulfilled, that her sin is paid for, 
that she has \textit{received}\lebnote{Literally “taken”} from the hand of Adonai double for all her sins.”
\verse A voice is calling in the wilderness, “Clear the way of Adonai! 
Make a highway smooth in the desert for our God!
\verse Every valley shall be lifted up, 
and every mountain and hill shall become low, 
And the rough ground shall be like a plain, 
and the rugged ground like a valley-plain. for the mouth of Adonai has spoken.”
\verse A voice is saying, “Call!” 
And he said, “What shall I call?” 
All \textit{humankind}\lebnote{Literally “flesh”} are grass, 
and all his loyalty is like the flowers of the field.
\verse Grass withers; the flower withers 
when the breath of Adonai blows on it. 
Surely the people are grass.
\verse Grass withers; the flower withers, 
but\lebnote{Or “and”} the word of our God will stand forever.
\verse \textit{Get yourself}\lebnote{Literally “Go for you”} up to a high mountain, Zion, bringer of good news! 
Lift up your voice with strength, Jerusalem, bringer of good news! 
Lift it up; you must not fear! 
Say to the cities of Judah, “Here is your God!”
\verse Look! The Lord Adonai comes with \textit{strength},\lebnote{Literally “strong”} 
and his arm rules for him. 
Look! His reward is with him, 
and his recompense \textit{in his presence}.\lebnote{Literally “before his face”}
\verse He will feed his flock like a shepherd; 
he will gather the lambs in his arm, 
and he will carry them in his bosom; 
he will lead those who nurse.
\verse Who has measured the waters in the hollow of his hand 
and marked off the heavens with a span, 
comprehended the dust of the earth in a third of a measure 
and weighed out the mountains in the scales,\lebnote{Hebrew “scale”} and the hills in a balance?
\verse Who has measured up the spirit of Adonai 
or informed him as \textit{his counselor}\lebnote{Literally “the man of his counsel”}?
\verse With whom has he consulted, that\lebnote{Or “and”} he enlightened him\lebnote{Or “made him understand”} 
and taught him the path of justice, 
and taught him knowledge, 
and made the way of understanding known to him?
\verse Look! The nations are like a drop from a bucket, 
and they are counted like dust of the balances! 
Look! He weighs the islands like a thin covering.
\verse And Lebanon is not enough to light a fire, 
and its animals\lebnote{Hebrew “animal”} not enough for a burnt offering.
\verse All the nations are like nothing before him; 
they are counted by him as\lebnote{Or “from”} nothing and emptiness.
\verse And to whom will you liken God? 
And to what likeness will you compare him?
\verse A craftsman pours out the idol, 
and a \textit{goldsmith}\lebnote{Literally “refiner”} overlays it with gold, 
and he smelts chains of silver.
\verse The one who is too impoverished for a gift chooses wood that will not rot; 
he seeks a skillful artisan for himself to set up an image that will not be knocked over.
\verse Have you not known? 
Have you not heard? 
Has it not been told to you from the \textit{beginning}?\lebnote{Literally “head”} 
Have you not understood from the foundation of the earth?
\verse He is the one who sits above the circle of the earth, 
and its inhabitants are like grasshoppers; 
the one who stretches out the heavens like a veil 
and spreads them out like a tent to live in,
\verse the one who brings\lebnote{Or “gives”} princes to nothing; 
he makes rulers of the earth like nothing.
\verse Indeed, hardly are they planted; indeed, hardly are they sown; 
indeed, hardly has their shoot taken root in the earth 
when\lebnote{Or “also”} he blows on them and they wither, 
and the tempest carries them like stubble.
\verse “And to whom you will compare me, and am I equal?” says the holy one.
\verse Lift your eyes up on high, and see! Who created these? 
The one who brings out their host by number. 
He calls all them by name. 
Because he is great of power 
and mighty of power, no man is missing.
\verse \textit{Why}\lebnote{Literally “To what”} do you say, Jacob, 
and you speak, Israel, 
“My way is hidden from Adonai, 
and my judgment is passed over by my God?”
\verse Have you not known, 
or have you not heard? 
Adonai is the God of eternity, 
the creator of the ends of the earth! 
He is not faint, and he does not grow weary! 
There is no searching his understanding.
\verse He gives power to the weary, 
and he increases power for \textit{the powerless}.\lebnote{Literally “not power”}
\verse Even\lebnote{Or “And”} young people will be faint and grow weary, 
and the young will stumble, exhausted.
\verse But\lebnote{Or “And”} those who wait for Adonai shall renew their strength. 
They shall go up with wings\lebnote{Hebrew “wing”} like eagles; 
they shall run and not grow weary; 
they shall walk and not be faint.
\end{biblechapter}

\begin{biblechapter} % Isaiah 41
\verseWithHeading{God Helps Israel} Listen to me in silence, coastlands, 
and let nations renew their strength. 
Let them approach, then let them speak; 
let us draw near together for judgment.
\verse Who has roused salvation from the east, 
summoned\lebnote{Or “called”} him to his foot, 
gives nations \textit{in his presence},\lebnote{Literally “before his face”} 
and subjugates kings? 
He makes\lebnote{Or “gives”} them like the dust with his sword, 
like scattered stubble with his bow.
\verse He pursues them and passes on in peace; 
he does not enter the path with his feet.
\verse Who has accomplished and done this, 
calling the generations from the \textit{beginning}?\lebnote{Literally “head”} 
I, Adonai, am first; 
and I am the one with the last.
\verse The coastlands have seen and are afraid; 
the ends of the earth tremble. 
They have drawn near, 
and they have come.
\verse Each one helps\lebnote{Hebrew “help”} his neighbor; 
he says to his brother, “Take courage!”
\verse And the artisan encourages the \textit{goldsmith},\lebnote{Literally “refiner”} 
the one who makes smooth with the hammer encourages the one who strikes the anvil, 
saying of the soldering, “It is good!” 
And they strengthen it with nails so it cannot be knocked over.
\verse But\lebnote{Or “And”} you, Israel, my servant, Jacob, whom I have chosen, 
you, the \textit{offspring}\lebnote{Literally “seed”} of Abraham my \textit{friend},\lebnote{Literally “beloved”}
\verse you whom I grasped from the ends of the earth 
and called from its remotest parts 
and told, “You are my servant; 
I have chosen you and I have not rejected you.”
\verse You must not fear, for I am with you; 
you must not be afraid, for I am your God. 
I will strengthen you, indeed I will help you, 
indeed I will take hold of you with the right hand of my salvation.
\verse Look! All those who are angry with you shall be ashamed and humiliated; 
\textit{your opponents}\lebnote{Literally “the men of your strife”} shall be like nothing and shall become lost.
\verse You shall seek them, 
but\lebnote{Or “and”} you shall not find them; 
\textit{your opponents}\lebnote{Literally “the men of your strife”} shall be like nothing, 
and the men of your war like nothing.
\verse For I, Adonai your God, am grasping your right hand; 
\textit{it is I who say}\lebnote{Literally “the sayer”} to you, 
“You must not fear; 
I myself, I will help you.
\verse You must not fear, O worm of Jacob; 
people of Israel, 
I myself, I will help you,” \textit{declares}\lebnote{Literally “declaration of”} Adonai, 
“and your redeemer is the holy one of Israel.
\verse Look! I will make you into a new sharp threshing sledge, 
\textit{with}\lebnote{Literally “owner of”} sharp edges. 
You shall thresh and crush the mountains, 
and you shall make the hills like chaff.
\verse You shall winnow them and the wind shall carry them, 
and the tempest shall scatter them. 
And you yourself shall rejoice in Adonai; 
you shall boast in the holy one of Israel.
\verse The poor and the needy are seeking water and there is none; 
their tongue is dried up with thirst. 
I, Adonai, will answer them; 
I, the God of Israel, will not forsake them.
\verse I will open rivers on the barren heights 
and fountains in the midst of the valleys. 
I will make the wilderness like a pool of water 
and the land of dryness like springs of water.
\verse I will \textit{put}\lebnote{Literally “give”} the cedar, acacia, myrtle, and olive oil tree in the wilderness; 
I will set the cypress, elm, and box tree together in the desert
\verse so that they may see and know, 
and take to heart and understand together 
that the hand of Adonai has done this, 
and the holy one of Israel has created it.”
\verse “Present your legal case,” says Adonai. 
“Bring your evidence,” says the king of Jacob.
\verse Let them bring them, 
and let them tell us what will happen. 
Tell us what the former things are 
\textit{so that}\lebnote{Or “And”} we may take them to our heart 
and know their outcome.\lebnote{Or “end”} 
Declare to us the things to come;
\verse tell us the things coming \textit{hereafter},\lebnote{Literally “later”} 
that\lebnote{Or “and”} we may know that you are gods. 
Indeed, do good or\lebnote{Or “and”} do evil, 
that\lebnote{Or “and”} we may be afraid and see\lebnote{Following the reading tradition (\textit{Qere}); the written text (\textit{Kethib}) has “fear”} together.
\verse Look! you are nothing, 
and your work is something worthless; 
whoever chooses you is an abomination.
\verse I stirred up one from the north, 
and he has come from the rising of the sun. He shall call on my name,
\verse and he shall come on officials as on mortar, 
and as the \textit{potter}\lebnote{Literally “creator”} treads clay.
\verse First to Zion, look! Look at them! And I give a herald\lebnote{Or “bringer”} of good tidings to Jerusalem.
\verse But\lebnote{Or “And”} I look and there is no man, 
and I look among these and there is no counselor, 
that\lebnote{Or “and”} I might ask them and they might answer a word.
\verse Look! All of them are deception; 
their works are nothing; 
their images are wind and emptiness.
\end{biblechapter}

\begin{biblechapter} % Isaiah 42
\verseWithHeading{The Mission of Adonai’s Servant} Look! here is my servant; I hold him, 
my chosen one, in whom my soul delights. 
I have \textit{put}\lebnote{Literally “given”} my spirit on him; 
he will bring justice forth to the nations.
\verse He will not cry out and lift up 
and make his voice heard in the street.
\verse He will not break a broken reed, 
and he not will extinguish a dim wick. 
He will bring justice forth in\lebnote{Or “to”} faithfulness.
\verse He will not grow faint, 
and he will not be broken 
until he has established justice in the earth. 
And the coastlands wait for his teaching.
\verse Thus says the God, Adonai, 
who created the heavens 
and stretched them out, 
who spread out the earth and its offspring, 
who gives breath to the people upon it 
and spirit to those who walk in it.
\verse “I am Adonai; I have called you in righteousness, 
and I have grasped your hand and watched over you; 
and I have given you as a covenant of the people, 
as a light of the nations,
\verse to open the blind eyes, 
to bring the prisoner out from the dungeon, 
those who sit in darkness from the house of imprisonment.
\verse I am Adonai; that is my name,
\verse and I do not give my glory to another, 
nor\lebnote{Or “and”} my praise to the idols.
\verseWithHeading{A Song of Praise to Adonai} Sing a new song to Adonai; 
praise him from the end of the earth, 
you\lebnote{Or “those”} who go down to the sea and that which fills it, 
the coastlands and their inhabitants.
\verse Let the desert and its towns lift up their voice, 
the villages that Kedar inhabits. 
Let the inhabitants of Sela sing for joy; 
let them shout loudly from the \textit{top}\lebnote{Literally “head”} of the mountains.
\verse Let them give\lebnote{Or “establish”} glory to Adonai 
and declare his praise in the coastlands.
\verse Adonai goes forth like a mighty warrior; 
he stirs up zeal like a man of war. 
He raises the war cry, indeed he raises the battle shout; 
he prevails against his foes.
\verse I have been silent for a long time; 
I have kept silent; 
I have restrained myself like one giving birth; 
I will moan, pant, and gasp together.
\verse I will cause mountains and hills to dry up, 
and I will cause all their herbage to wither; 
and I will make rivers like islands, 
and I will cause pools to dry up.
\verse And I will lead the blind by a road they do not know; 
I will cause them to tread on paths they have not known. 
I will make darkness \textit{in their presence}\lebnote{Literally “before their face”} into light 
and rough places into level ground. 
These are the things I will do, 
and I will not forsake them.
\verse They shall turn back; 
they shall be \textit{greatly ashamed},\lebnote{Literally “ashamed with shame”} 
those who trust in an image, 
who say to a cast image, “You are our gods.”
\verseWithHeading{Blind and Deaf Israel} Deaf people, listen! 
And blind people, look to see!
\verse Who is blind but my servant 
or\lebnote{Or “and”} deaf like my messenger whom I sent? 
Who is blind like the one who is repaid 
or\lebnote{Or “and”} blind like the servant of Adonai?
\verse You see many things, but\lebnote{Or “and”} you do not observe. 
His ears are open, but\lebnote{Or “and”} he does not hear.
\verse Adonai was willing for the sake of his righteousness; 
he showed his teaching to be great and proved it to be glorious.
\verse But\lebnote{Or “And”} this is a people plundered and looted; all of them are trapped in holes, 
and they are kept hidden in houses of imprisonment. 
They have become like plunder, and there is no one who saves; 
like booty, and there is no one who says, “Restore!”
\verse Who among you will heed this, 
will listen attentively and listen, for \textit{the time to come}?\lebnote{Literally “later”}
\verse Who gave Jacob to a plunderer\lebnote{The reading tradition (\textit{Qere}) has “as plunder”} 
and Israel to those who plunder? Was it not Adonai, against whom\lebnote{Or “this”} we have sinned?
\verse And they were not willing to walk in his ways, 
and they would not \textit{obey}\lebnote{Literally “hear”} his law.
\end{biblechapter}

\begin{biblechapter} % Isaiah 43
\verseWithHeading{The Restorer of Israel} But now thus says Adonai,
\verse he who created you, Jacob, 
and he who formed you, Israel: 
“You must not fear, for I have redeemed you. 
I have called you by your name; \textit{you are mine}.\lebnote{Literally “to me you”}
\verse When you pass through the waters, I will be with you, 
and through the rivers, they shall not flow over you. 
When you walk through fire, you shall not be burned, 
and the flame shall not scorch\lebnote{Or “burn”} you.
\verse For I am Adonai, your God, 
the holy one of Israel, your savior. 
I give you Egypt as ransom, 
Cush and Seba in place of you.
\verse Because you are precious in my eyes, 
you are honored, and I myself love you, 
and I give people in place of you, 
and nations in place of your life.
\verse You must not fear, 
for I am with you. 
I will bring your \textit{offspring}\lebnote{Literally “seed”} from the east, 
and I will gather you from the west.
\verse I will say to the north, ‘Give!’ 
and to the south, ‘You must not withhold!’ 
Bring my sons from far away, 
and my daughters from the end of the earth—
\verse everyone who is called by my name, 
and whom I created for my glory, 
whom I formed, 
indeed whom I made.”
\verse Bring out the people blind yet\lebnote{Or “and”} with eyes, 
and deaf, though\lebnote{Or “and”} they have ears.
\verse Let all the nations gather together, 
and let the peoples assemble. 
Who among them has declared this, 
and \textit{declared}\lebnote{Literally “caused to hear”} the former things to us? 
Let them bring\lebnote{Or “give”} their witnesses, that\lebnote{Or “and”} they may be in the right, 
and let them hear and say, “It is true!”
\verse “You are my witnesses,” \textit{declares}\lebnote{Literally “declaration of”} Adonai, 
“and my servant whom I have chosen 
so that you may know and believe \textit{in}\lebnote{Literally “to”} me 
and understand that I am he. 
No god was formed before \textit{me}\lebnote{Literally “my face”}, 
and none shall be after me.
\verse I myself am Adonai, 
and there is no savior besides me!
\verse I myself declared and saved, 
and I \textit{proclaimed}.\lebnote{Literally “caused to hear”} And there was no \textit{strange god}\lebnote{Literally “stranger”} among you. 
And you are my witnesses,” \textit{declares}\lebnote{Literally “declaration of”} Adonai, 
“and I am God.
\verseWithHeading{Adonai Rescues His People} Thus says Adonai, your redeemer, 
the holy one of Israel: 
“For your sake I will send to Babylon, 
and I will cause all of them to fall down as fugitives, 
and \textit{the Chaldeans},\lebnote{Literally “Chaldea”} their rejoicing on the ships.
\verse I am Adonai, your holy one, 
the creator of Israel, your king.”
\verse Thus says Adonai, who \textit{makes}\lebnote{Literally “gives”} a way in the sea 
and a path in the mighty waters,
\verse who brings out chariot and horse, 
army and mighty one. 
Together they lie down; 
they cannot rise. 
They are extinguished, 
quenched like a wick.
\verse “You must not remember the former things, 
and you not must consider the former things.
\verse Look! I am about to do a new thing! Now it sprouts! 
Do you not perceive\lebnote{Or “know”} it? 
Indeed, I will \textit{make}\lebnote{Literally “put”} a way in the wilderness, 
rivers in the desert.
\verse The animals of the field will honor me, 
jackals and daughters of the ostrich, 
for I give water in the wilderness, 
rivers in the desert, 
to give a drink to my chosen people,
\verse this people whom I formed for myself, 
so they might \textit{make known}\lebnote{Literally “count out”} my praise.
\verse But\lebnote{Or “And”} you did not call me, Jacob; 
for you have become weary of me, Israel.
\verse You have not brought me your sheep for a burnt offering 
nor\lebnote{Or “and”} honored me with your sacrifice. 
I have not made you serve with offerings,\lebnote{Hebrew “offering”} 
nor have I made you weary with frankincense.
\verse You have not bought me spice reed with money 
or\lebnote{Or “and”} satisfied me with the fat of sacrifices. 
But you have burdened me with your sins; 
you have made me weary with your iniquities.
\verse I, I am the one who blots out your transgressions for my sake, 
and I will not remember your sins.
\verse Take me to court; let us enter into judgment together. 
You, \textit{make an account}\lebnote{Literally “count up”} so that you may be in the right.
\verse Your first ancestor\lebnote{Or “father”} sinned, 
and your \textit{representatives}\lebnote{Literally “scoffers” or “spokesmen”} transgressed against me.
\verse And I profaned the princes of the sanctuary, 
and I gave Jacob to destruction, and Israel to reviling.
\end{biblechapter}

\begin{biblechapter} % Isaiah 44
\verseWithHeading{Adonai Blesses Chosen Israel} “But now hear, Jacob my servant, 
and Israel, whom I have chosen.
\verse Thus says Adonai, who made you, 
and who formed you in\lebnote{Or “from”} the womb and will help you: 
you must not fear, my servant Jacob, 
and Jeshurun whom I have chosen.
\verse For I will pour out water on a thirsty land 
and streams on dry ground. 
I will pour my spirit out on your descendants\lebnote{Hebrew “descendant”} 
and my blessing on your offspring.
\verse And they shall sprout \textit{among}\lebnote{Literally “in between”} the grass 
like willows by a watercourse of water.
\verse This one will say, ‘I belong to Adonai!’ 
And that one will be called by the name of Jacob, 
and another will write on his hand ‘Adonai’s’ 
and \textit{take the name}\lebnote{Literally “he will be titled by the name”} of Israel.”
\verseWithHeading{Adonai the Only God} Thus says Adonai, the king of Israel, 
and its redeemer, Adonai of hosts: 
“I am the first, and I am the last, 
and there is no god besides me.
\verse And who is like me? Let him proclaim\lebnote{Or “call”} it! 
And let him declare it and set it in order for me 
\textit{since I established an eternal people}\lebnote{Literally “from my placing a people of eternity”} and things that are to come, 
and let them tell them the things that are coming.
\verse You must not tremble,\lebnote{Or “be startled”} 
and you must not be paralyzed with fear. 
Have I not made you hear from of old\lebnote{Or “then”} 
and declared it, and you are my witnesses? 
Is there a god besides me? 
And there is no rock! I know none!”
\verseWithHeading{Idolatry Is Ridiculous} All those who form an idol are nothing, 
and their delightful things do not profit. 
And their witnesses do not see or know, 
so they will be ashamed.
\verse Who would form a god 
and cast an image of which he cannot profit?
\verse Look! all his companions shall be ashamed, 
and the artisans are human! 
Let all of them assemble; 
let them stand up. 
They shall tremble;\lebnote{Or “be startled”} 
they shall be ashamed together.
\verse The \textit{ironsmith}\lebnote{Literally “craftsman of iron”} works in the coals with his tool 
and forms it with hammers. 
And he makes it with \textit{his strong arm};\lebnote{Literally “the arm of his strength”} 
indeed, he becomes hungry, and \textit{he lacks}\lebnote{Literally “there is no”} strength; 
he does not drink water, and he is faint.
\verse The \textit{woodworker}\lebnote{Literally “craftsman of wood”} stretches out a line; 
he makes an outline of it with a marker. 
He makes it with a knife 
and makes an outline of it with a compass. 
He makes it like the image of a man, 
like the beauty of a human, to dwell in a temple.\lebnote{Or “house”}
\verse Cutting down cedars for himself, 
he \textit{chooses}\lebnote{Literally “takes”} a holm tree and an oak, 
and he lets it grow strong for him among the trees of the forest. 
He plants a cedar, and the rain makes it grow.
\verse And it \textit{becomes fuel for a human},\lebnote{Literally “will become for a man as burning”} 
and he takes some of it and grows warm; 
also, he kindles a fire and bakes bread. 
Also, he makes a god and bows in worship; 
he makes himself an image and bows down to it!
\verse He burns half of it in the fire; 
he eats meat over half of it; 
he roasts a roast and is satisfied. 
Also he grows warm and says, “Ah! 
I am warm! I see the fire!”
\verse And he makes the remainder of it into a god! 
He bows down to his idol, 
and he bows in worship and prays to him, 
and he says, “Save me, for you are my god!”
\verse They do not know, 
and they do not understand, 
for their eyes are besmeared \textit{so that they cannot see},\lebnote{Literally “from seeing”} 
their \textit{minds}\lebnote{Literally “hearts”} \textit{so that they have no insight}.\lebnote{Literally “from having insight”}
\verse And \textit{no one takes it to heart},\lebnote{Literally “he does not bring back to his heart”} and there is no knowledge 
and no understanding to say, “I burned half of it in the fire 
and also I baked bread on its coals; 
I roasted meat, and I have eaten. 
And I shall make the rest of it into an abomination! 
I shall bow down to a block of wood!”
\verse He feeds on ashes; a \textit{deceived mind}\lebnote{Literally “heart deceived”} misleads him. 
And he cannot save \textit{himself},\lebnote{Literally “his soul”} and he cannot say, 
“Is this not an illusion in my right hand?”
\verseWithHeading{Adonai Remembers Israel} “Remember these things, Jacob, 
and Israel, for you are my servant: 
I formed you; you are my servant; 
Israel, you will not be forgotten by me!
\verse I have wiped your transgressions out like a cloud 
and your sins like mist. 
Return to me, 
for I have redeemed you!”
\verse Sing for joy, heavens, for Adonai has done it! 
Shout, depths of the earth! 
Break forth, mountains, in rejoicing, 
forest and every tree in it, 
for Adonai has redeemed Jacob, 
and he will show his glory in Israel!
\verseWithHeading{Jerusalem Restored through Cyrus} Thus says Adonai, your redeemer, 
and he who formed you in\lebnote{Or “from”} the womb: 
“I am Adonai, who made everything, 
who stretched out the heavens alone, 
who spread out the earth—who was with me?—
\verse who frustrates\lebnote{Or “breaks”} the signs of oracle priests 
and makes a fool of diviners, 
who drives the wise men back 
and makes their knowledge foolish,
\verse who keeps\lebnote{Or “erects”} the word of his servant 
and carries out the plan of his messengers, 
who says of Jerusalem, ‘It shall be inhabited,’ 
and of the cities of Judah, ‘They shall be rebuilt, 
and I will restore its ruins’;
\verse who says to the deep, ‘Dry up! 
And I will cause your rivers to dry up’;
\verse who says of Cyrus, ‘My shepherd,’ 
and he shall carry out all my wishes\lebnote{Hebrew “wish”}; 
and saying of Jerusalem, ‘It shall be rebuilt,’ 
and the temple, ‘It shall be founded.’ ”
\end{biblechapter}

\begin{biblechapter} % Isaiah 45
\verse Thus says Adonai to his anointed one, 
to Cyrus, whose right hand I have grasped 
to subjugate nations before him, 
and I uncover\lebnote{Or “let loose”} the loins of kings 
to open doors before him, 
and the gates shall not be shut:
\verse “I myself will go before you, 
and I will level the \textit{mountains}.\lebnote{Literally “swells”} 
I will break the doors of bronze 
and cut throughthe bars of iron.
\verse And I will give you the treasures of darkness 
and treasures of secret places 
so that you may know that I am Adonai, 
the one who calls you by your name, the God of Israel,
\verse for the sake of my servant Jacob, 
and Israel my chosen one. 
And I call you by your name; 
I give you a name of honor, though\lebnote{Or “and”} you do not know me.
\verse I am Adonai, and there is none besides me; 
besides me there is no god. 
I gird you though\lebnote{Or “and”} you do not know me,
\verse so that they may know from the rising of the sun 
and from the west that there is none besides me; 
I am Adonai and there is none besides me.
\verse I form light and I create darkness; 
I make peace and I create evil; 
I am Adonai; I do all these things.
\verse Trickle, O heavens, from above, 
and let clouds trickle with righteousness; 
let the earth open so that\lebnote{Or “and”} salvation may be fruitful, 
and let it cause righteousness to sprout along with it.\lebnote{Or “together”} 
I myself, Adonai, have created it.
\verse Woe to the one who strives with his \textit{maker},\lebnote{Literally “one who forms”} 
a potsherd among\lebnote{Or “with”} potsherds of earth! 
Does the clay say to the one who fashions it, 
‘What are you making?’ 
and ‘Your work has no hands’?
\verse Woe to the one who says to a father, ‘What you are begetting?’ 
or\lebnote{Or “and”} to a woman, ‘With what are you in labor?’ ”
\verse Thus says Adonai, the holy one of Israel, 
and its \textit{maker}:\lebnote{Literally “one who forms”} 
“Ask me of the things to come \textit{about}\lebnote{Literally “over”} my children, 
and you command me about the work of my hands.
\verse I myself made the earth, 
and I created humankind upon it. 
I, my hands, stretched out the heavens, 
and I commanded all their host.
\verse I myself have stirred him up in righteousness, 
and I will make all his paths smooth. 
He himself shall build my city, 
and he shall set my exiles free, not for price or a gift,” says Adonai of hosts.
\verseWithHeading{Adonai the Only Savior} Thus says Adonai: 
“The acquisition of Egypt 
and the merchandise of Cush 
and the Sabeans, tall men, 
shall pass over to you; 
they shall be yours, 
and they shall walk behind you. 
They shall pass over in chains, 
and they shall bow down to you; 
they will pray to you: 
‘Surely God is with you, and there is no other. 
Besides him there is no God.’ ”
\verse Surely you are a God who keeps yourself hidden, 
God of Israel, the savior.
\verse All of them are ashamed and indeed humiliated; 
the craftsmen of idols go together in insult.
\verse Israel is saved by Adonai with everlasting salvation; 
you shall not be ashamed, 
and you shall not be humiliated \textit{to all eternity}.\lebnote{Literally “for eternity, forever”}
\verse For thus says Adonai, \textit{who created}\lebnote{Literally “creating”} the heavens, 
he is God, \textit{who formed}\lebnote{Literally “forming”} the earth 
and \textit{who made}\lebnote{Literally “making”} it. 
He himself established it; 
he did not create it as emptiness— 
he formed it for inhabiting. “I am Adonai and there is none besides me.
\verse I have spoken not in secrecy, in a place, a land, of darkness, 
I have not said to the descendants of Jacob, ‘Seek me \textit{in vain}!’\lebnote{Literally “in emptiness,” as in verse 18} 
I, Adonai, am speaking righteousness, 
declaring uprightness.
\verse Assemble and come; 
draw near together, survivors of the nations! 
They do not know, those who carry \textit{their wooden idols}\lebnote{Literally “the wood of their idol”} 
and pray to a god who cannot save.
\verse Declare and present your case, 
also let them consult together! 
Who \textit{made this known}\lebnote{Literally “caused to hear this”} from \textit{former times},\lebnote{Literally “before”} 
declared it from \textit{of old}?\lebnote{Literally “then”} 
Was it not I, Adonai? 
And there is no other god besides me, 
a righteous God besides me, 
and no savior besides me.
\verse Turn to me and be saved, all the ends of the earth, 
for I am God and there is none besides me.
\verse I have sworn by myself; 
a word that\lebnote{Or “and”} shall not return has gone forth from my mouth in righteousness: 
‘Every knee shall kneel down to me; 
every tongue shall swear.’
\verse ‘Only in Adonai,’ one shall say to me, ‘are righteousness and strength.’ 
He shall come to him, and all those who were angry with him shall be ashamed.
\verse In Adonai all the offspring of Israel shall be in the right, 
and they shall boast.”
\end{biblechapter}

\begin{biblechapter} % Isaiah 46
\verseWithHeading{Babylon’s Idols} Bel bows down; 
Nebo is stooping. 
Their idols are \textit{on}\lebnote{Literally “to”} animals\lebnote{Hebrew “animal”} and \textit{on}\lebnote{Literally “to”} cattle; 
your \textit{cargo}\lebnote{Literally “things that are carried”} is carried as a burden \textit{on}\lebnote{Literally “to”} weary animals.
\verse They stoop; 
they bow down together. 
They are not able to save the burden, 
but\lebnote{Or “and”} \textit{they themselves go}\lebnote{Literally “their inner self goes”} in captivity.
\verse “Listen to me, house of Jacob, 
and all the remnant of the house of Israel 
who have been carried from the belly, 
who have been carried from the womb:
\verse Even\lebnote{Or “And”} to your old age I am he; 
even\lebnote{Or “And”} to your advanced age I myself will support you. 
I myself have made you, 
and I myself will carry you, 
and I myself will support you, 
and I will save you.
\verse To whom will you liken me, 
and count as equal, 
and compare with me, 
as though\lebnote{Or “and”} we were alike?
\verse Those who lavish gold from the purse 
and weigh out silver in the balance scales; 
hire \textit{a goldsmith}\lebnote{Literally “one who smelts”} 
and he makes him a god; 
they bow down, 
indeed they bow in worship.
\verse They carry it on their shoulder; 
they support it and they set it in its place, 
and it stands in position. 
It cannot be removed from its place; 
even when he cries out to it, it does not answer. 
It does not save him from his trouble.
\verse Remember this and pluck up courage! 
Call to \textit{mind},\lebnote{Literally “heart”} you transgressors!
\verse Remember the former things from \textit{a long time ago},\lebnote{Literally “forever”} 
for I am God and there is none besides me, 
God and there is none like me,
\verse who from the beginning declares the end, 
and from before, things that have not been done, 
who says, ‘My plan shall stand,’ 
and, ‘I will accomplish\lebnote{Or “do”} all my wishes,’\lebnote{Hebrew “wish”}
\verse who calls a bird of prey from the east, 
the man of his\lebnote{The reading tradition (\textit{Qere}) has “my”} plan from a country from afar. 
Indeed I have spoken; indeed I will bring it to being. 
I have formed it; indeed I will do it.
\verse Listen to me, strong of heart, 
far from righteousness!
\verse I bring my righteousness near; it is not far. 
And my salvation will not delay; 
and I will put\lebnote{Or “give”} salvation in Zion, 
for Israel my glory.”
\end{biblechapter}

\begin{biblechapter} % Isaiah 47
\verseWithHeading{Babylon’s Fall} Come down and sit on the dust, virgin daughter of Babylon! 
Sit on the ground without a throne, daughter of Chaldea! 
For \textit{they shall no longer call}\lebnote{Literally “you shall not add they shall call to”} you tender and delicate.
\verse Take the pair of mill stones and grind flour! 
Uncover your veil, 
strip off your skirt, 
uncover your thigh, 
pass through the rivers!
\verse Your nakedness shall be exposed; 
indeed, your shame shall become visible. 
I will take vengeance 
and I will not \textit{spare}\lebnote{Literally “plead with”} a person.
\verse Our redeemer, Adonai of hosts is his name, the holy one of Israel.
\verse Sit silently and go into the darkness, daughter of Chaldea, 
for \textit{they shall no longer call}\lebnote{Literally “you shall not add they shall call to”} you mistress of kingdoms.
\verse I was angry with my people; 
I profaned my inheritance, 
and I gave them into your hand. 
You did not \textit{give}\lebnote{Literally “set up”} them mercy; 
on the aged you made your yoke very heavy.
\verse And you said, “I shall be an eternal mistress forever!” 
You did not set these things upon your heart; 
you did not remember its end.
\verse Therefore now hear this, luxuriant one who sits\lebnote{Or “sitting”} in security, 
who says\lebnote{Or “saying”} in her heart, “I am, and besides me there is no one. 
I shall not sit as a widow, 
and I shall not know the loss of children.”
\verse And these two shall come to you in a moment, in one day: 
the loss of children and widowhood shall come on you \textit{completely},\lebnote{Literally “as their fullness”} 
in spite of your many sorceries, 
in spite of the power of your great enchantments.
\verse And you felt secure in your wickedness; 
you said, “\textit{No one}\lebnote{Literally “none”} sees\lebnote{Or “is seeing”} me.” 
Your wisdom and your knowledge led you astray, 
and you said in your heart, “I am, and besides me there is no one.”
\verse And evil shall come upon you, you will not know; 
it will be on the lookout for her. 
And disaster shall fall upon you; you will not be able to \textit{avert}\lebnote{Literally “make amends”} it. 
And ruin shall come on you suddenly; you do not know.
\verse Stand, now, in your enchantments, 
and in your many sorceries with which you have labored from your youth. 
Perhaps you may be able to benefit; 
perhaps you may \textit{scare away}.\lebnote{Literally “terrify”}
\verse You struggle with your many consultations;\lebnote{Or “advice”} 
let them stand, now, and save you— 
those who see the stars, divide the \textit{celestial sphere},\lebnote{Literally “heavens,” that is, for astrology} 
who inform by new moons— 
from those things that are coming upon you.
\verse Look! They are like stubble; 
the fire burns them completely. 
They cannot deliver \textit{themselves}\lebnote{Literally “their lives”} from the \textit{power}\lebnote{Literally “hand”} of the flame; 
there is no coal \textit{for warming oneself},\lebnote{Literally “to grow warm”} 
no fire before which to sit.
\verse So are to you those with whom you have labored, 
your traders from your youth. 
They wander, each to his side; 
there is no one who can save you.
\end{biblechapter}

\begin{biblechapter} % Isaiah 48
\verseWithHeading{Adonai Refines Stubborn Israel} Hear this, house of Jacob, 
who are called by the name of Israel 
and came out from the waters of Judah, 
who swear by the name of Adonai, 
and \textit{invoke}\lebnote{Literally “cause to remember”} the God of Israel, 
but not in truth 
and not in righteousness.
\verse For they call themselves \textit{after the holy city},\lebnote{Literally “from the city of the holiness”} 
and they lean on the God of Israel— 
Adonai of hosts is his name.
\verse “I declared the former things from \textit{of old},\lebnote{Literally “then”} 
and they went out from my mouth. 
And I \textit{announced}\lebnote{Literally “caused to hear”} them suddenly; 
I acted,\lebnote{Or “did”} and they came to pass,
\verse because \textit{I know}\lebnote{Literally “knowing me”} that you are obstinate, 
and your neck an iron sinew, 
and your forehead bronze.
\verse And I declared them to you from of old; 
I \textit{ announced}\lebnote{Literally “caused to hear”} them to you before they came to pass 
so that you would not say, ‘My idol did them, 
and my image and my cast image commanded them.’
\verse You have heard; see it all. 
And will you not declare it? 
I \textit{announce}\lebnote{Literally “cause to hear”} new things to you from this time 
and hidden things that\lebnote{Or “and”} you have not known.
\verse Now they are created, and not from \textit{of old},\lebnote{Literally “then”} 
and before today, and you have not heard them 
so that you could not say, “Look! I knew them.”
\verse Neither have you heard, nor have you known, 
nor from \textit{of old}\lebnote{Literally “then”} has your ear been opened. 
For I knew you would deal treacherously, very treacherously, 
and you are called a rebel from the womb.
\verse For the sake of my name I \textit{refrain from}\lebnote{Literally “make long”} my anger, 
and for my praise I restrain it for you 
so as not to cut you off.
\verse Look! I have refined you, but\lebnote{Or “and”} not like\lebnote{Or “with”} silver; 
I have chosen you in the furnace of misery.
\verse For my own sake, for my own sake I do it; 
for why should it\lebnote{That is, my name} be defiled? 
And I will not give my glory to another.”
\verseWithHeading{Adonai Frees Israel} “Listen to me, Jacob, 
and Israel, \textit{whom I called}:\lebnote{Literally “who was called by me”} 
I am he. I am the first; 
also I am the last.
\verse Indeed, my hand founded the earth, 
and my right hand spread out the heavens; 
when I summon\lebnote{Or “calling”} them, they stand in position together.
\verse Assemble, all of you, and hear! 
Who among them declared these things? 
Adonai loves him; 
he shall perform\lebnote{Or “do”} his wish against Babylon 
and his arm against \textit{the Chaldeans}.\lebnote{Literally “Chaldea”}
\verse I, I myself, I have spoken! 
Indeed, I have called him. 
I have brought him, 
and he will be successful in his way.
\verse Draw near to me; hear this! I have not spoken in secrecy from the \textit{beginning};\lebnote{Literally “head”} 
from the time \textit{it came to be},\lebnote{Literally “of becoming it”} there I have been; 
And now the Lord Adonai has sent me and his Spirit.”
\verse Thus says Adonai, your redeemer, the holy one of Israel: “I am Adonai your God, \textit{who teaches}\lebnote{Literally “teaching”} you to profit, 
leads you in the way you should go.
\verse O that you had listened attentively to my commandments!
\verse Then\lebnote{Or “and”} your prosperity would have been like a river, 
and your righteousness like the waves of the sea.
\verse And your offspring would have been like the sand, 
and the descendants of your \textit{body}\lebnote{Literally “intestines”} like its grains.\lebnote{Hebrew “grain”} 
It would not be cut off, 
and its name would not be destroyed from my presence.”
\verse Go out from Babylon! 
Flee from Chaldea! 
Proclaim\lebnote{Or “announce”} it with a shout\lebnote{Or “sound”} of rejoicing; proclaim\lebnote{Or “cause to hear”} this! 
\textit{Send it forth}\lebnote{Literally “Cause it to go out”} to the end of the earth; 
say, “Adonai has redeemed his servant Jacob!”
\verse “There is no peace,” says Adonai, “for the wicked.”
\end{biblechapter}

\begin{biblechapter} % Isaiah 49
\verseWithHeading{Adonai’s Servant Brings Salvation} Listen to me, coastlands, 
and listen attentively, peoples from far away! 
Adonai called me from the womb; 
from the \textit{body}\lebnote{Literally “intestines”} of my mother he made my name known.
\verse And he \textit{made}\lebnote{Literally “put”} my mouth like a sharp sword; 
he hid me in the shadow of his hand, 
and he \textit{made}\lebnote{Literally “put”} me like an sharpened arrow; 
he hid me in his quiver.
\verse And he said to me, “You are my servant, 
Israel, in whom I will show my glory.”
\verse But\lebnote{Or “And”} I myself said, “I have labored in\lebnote{Or “for”} vain; 
I have used up my strength for nothing and vanity! 
Nevertheless, my justice is with Adonai, 
and my reward is with my God.”
\verse And now Adonai says, 
who formed me from the womb as a servant for him, 
to bring Jacob back to him, 
and that Israel might not\lebnote{The reading tradition (\textit{Qere}) has “to him”} be gathered, 
for\lebnote{Or “and”} I am honored in the eyes of Adonai, 
and my God has become my strength.
\verse And he says, “It is trivial \textit{for you to be}\lebnote{Literally “from being you”} a servant for me, to raise up the tribes of Jacob 
and to bring back the preserved of Israel. 
I will give you as a light to the nations, 
to be my salvation to the end of the earth.”
\verse Thus says Adonai, the redeemer of Israel, his holy one, to the one who despises\lebnote{A Dead Sea Scroll reads “is despised with respect to”} life, 
to the one who abhors the nation, 
to the slave of rulers: 
“Kings shall see and stand up; 
princes, and they shall bow down, 
for the sake of Adonai, who is faithful, 
the holy one of Israel, and he has chosen you.”
\verseWithHeading{Adonai Restores the Afflicted} Thus says Adonai:
\verse “I have answered you in a time of favor, 
and helped you on a day of salvation, 
and watched over you, 
and given you as a covenant of the people, 
to raise up the land, 
to give the desolate hereditary property as an inheritance,
\verse saying to the \textit{prisoners},\lebnote{Literally “captured ones”} “Come out!” 
to those who are in darkness, “Show yourselves!” 
they shall feed \textit{along}\lebnote{Literally “on”} the ways, 
and their pasturage shall be on all the barren heights.
\verse They shall not be hungry or thirsty, 
and heat and sun shall not strike them, 
for he who takes pity on them will lead them, 
and he will guide them to springs of water.
\verse And I will \textit{make}\lebnote{Literally “put”} all my mountains like a road, 
and my highways shall lead up.\lebnote{Or “be high”}
\verse Look! These shall come from afar, 
And look! These from the north and from the west 
and these from the land of Sinim.”
\verseWithHeading{Adonai Remembers Zion} But\lebnote{Or “And”} Zion said, “Adonai has forsaken me, 
and the Lord has forgotten me!”
\verse Can a woman forget her suckling, 
refrain from having compassion on the child of her womb? 
Indeed, these may forget, 
but\lebnote{Or “and”} I, I will not forget you!
\verse Look, I have inscribed you on the palms of my hands; 
your walls are continually before me.
\verse Your children hasten; 
your destroyers and those who laid you waste depart\lebnote{Or “move away”} from you.
\verse Lift your eyes up all around and see; 
all of them gather; they come to you. 
\textit{As surely as I live},\lebnote{Literally “Life I”} 
\textit{declares}\lebnote{Literally “declaration of”} Adonai, 
surely you shall put on all of them like an ornament, 
and you shall bind them on like a bride.
\verse Surely your sites of ruins and desolate places and land of ruins,\lebnote{Hebrew “ruin”} surely now you will be \textit{too cramped for your}\lebnote{Literally “cramped by”} inhabitants,\lebnote{Hebrew “inhabitant”} 
and those who engulfed you will be far away.
\verse Yet the children \textit{born when you were bereaved}\lebnote{Literally “of your bereavement”} will say in your \textit{hearing},\lebnote{Literally “ears”} “The place is too cramped for me; 
\textit{make room}\lebnote{Literally “approach”} for me so that I can dwell.”
\verse Then\lebnote{Or “And”} you will say in your heart, “Who has borne me these?” 
And, “I was bereaved and barren, 
exiled and thrust away; 
so\lebnote{Or “and”} who raised these? 
Look at me! I was left alone; 
\textit{where have these come from}?”\lebnote{Literally “these where they”}
\verse Thus says the Lord Adonai:
\verse “Look! I will lift my hand up to the nations, 
and I will raise my signal to the peoples, 
and they shall bring your sons in their bosom, 
and your daughters shall be carried on their shoulders.\lebnote{Hebrew “shoulder”}
\verse And kings shall be your \textit{guardians},\lebnote{Literally “those who nurture”} 
and their queens your nurses. 
They shall bow down, \textit{faces}\lebnote{Literally “noses”} to the ground, to you, 
and they will lick up the dust of your feet. 
Then\lebnote{Or “And”} you will know that I am Adonai; 
those who await me shall not be ashamed.
\verse But thus says Adonai:
\verse “Indeed a captive of the mighty shall be taken, 
and the war-booty of the tyrant shall be rescued, 
for\lebnote{Or “and”} I myself will dispute with your opponent, 
and I myself will save your children.
\end{biblechapter}

\begin{biblechapter} % Isaiah 50
\verse Thus says Adonai:
\verse “Where is this divorce document of your mother’s divorce, with which I dismissed her? 
or to whom of my creditors did I sell you? 
Look! you were sold because of your sin, 
and your mother was dismissed because of your transgressions.
\verse Why was there no man when I came, 
no one who answered when I called? 
\textit{Do I lack the strength to save}?\lebnote{Literally “short is short my hand from redemption”} 
Or is there no power in me to deliver? 
Look! by my rebuke I dry up the sea; 
I \textit{make}\lebnote{Literally “put”} the rivers a desert; 
their fish stink because there is no water, 
and they\lebnote{Hebrew “she”} die because of thirst.
\verseWithHeading{The Servant’s Vindication} The Lord Adonai has given me the tongue of a pupil, 
to know how to help the weary with a word. 
He awakens \textit{morning by morning},\lebnote{Literally “in the morning in the morning”} 
awakens an ear for me to listen as do the pupils.
\verse The Lord Adonai has opened an ear for me, 
and I, I was not rebellious. 
I did not turn backwards;
\verse I gave my back to those who struck me, 
and my cheeks to those who pulled out my \textit{beard};\lebnote{Literally “face”} 
I did not hide my face from insults and spittle.
\verse And the Lord Adonai helps me,
\verse therefore I have not been put to shame; 
therefore I have set my face like flint. 
And I know that I shall not be ashamed;
\verse he who obtains rights for me is near. 
Who will contend with me? Let us stand together. 
Who is the master of my judgment? Let him approach me.
\verse Look! The Lord Adonai helps me. 
Who is the one who will declare me guilty? 
Look! All of them will be worn out like a garment; 
the moth will eat them.
\verse Who among you is in fear of Adonai, 
obeys\lebnote{Or “listens”} the voice of his servant? 
Who walks in darkness 
and has no light, 
trusts in the name of Adonai 
and depends on his God?
\end{biblechapter}

\begin{biblechapter} % Isaiah 51
\verseWithHeading{Adonai Comforts Zion} “Listen to me, you who pursue righteousness, 
who seek Adonai. 
Look to the rock from which you were hewn, 
and to the excavation of the pit from which you were quarried.
\verse Look to Abraham your father, 
and to Sarah; she brought you forth. 
For I called him \textit{alone},\lebnote{Literally “one”} 
but\lebnote{Or “and”} I blessed him and made him numerous.”
\verse For Adonai will comfort Zion; 
he will comfort all its sites of ruins. 
And he will \textit{make}\lebnote{Literally “put”} its wilderness like Eden, 
and its desert like the garden of Adonai. 
Joy and gladness will be found in it, 
thanksgiving and the \textit{sound}\lebnote{Literally “voice”} of song.
\verse “Listen attentively to me, my people, 
and my nation, listen to me! 
For a teaching will go out from me, 
and I will cause my justice to rest for a light to the peoples.
\verse My righteousness is near; my salvation has gone out, 
and my arms will judge the peoples. 
The coastlands wait for me, 
and for my arm they wait.
\verse Lift up your eyes to the heavens 
and look to the earth beneath, 
for the heavens will be torn to pieces like smoke, 
and the earth will be worn out like a garment, 
and those who inhabit her will die like gnats. 
But\lebnote{Or “And”} my salvation will be forever, 
and my righteousness will not be broken to pieces.
\verse Listen to me, you who know righteousness, 
people who have my teaching in their heart; 
you must not fear the reproach of men, 
or be terrified because of their abuse.
\verse For a moth will eat them like garments; 
a moth will devour\lebnote{Or “eat”} them like wool, 
but\lebnote{Or “and”} my righteousness will be forever, 
and my salvation for \textit{generation after generation}.\lebnote{Literally “generation of generations”}”
\verse Awake! Awake; put on strength, O arm of Adonai! 
Awake as in days of long ago, 
the generations of a long time back! 
Are you not the one who cut Rahab in pieces, 
the one who pierced the sea-dragon?
\verse Are you not the one who dried up the sea, 
the waters of the great deep, 
the one who \textit{made}\lebnote{Literally “placed”} the depths of the sea a way 
for those who are redeemed to cross over?
\verse So\lebnote{Or “And”} the redeemed ones of Adonai shall return, 
and they shall come to Zion with singing, 
and everlasting joy shall be on their heads.\lebnote{Hebrew “head”} 
Joy and gladness shall \textit{appear};\lebnote{Literally “reach”} 
sorrow and sighing shall flee away!
\verse “I, I am he who comforts you; 
who are you that\lebnote{Or “and”} you are afraid of man? He dies! 
And of the son of humankind? He is \textit{sacrificed}\lebnote{Literally “given”} as grass!
\verse And you have forgotten Adonai, your maker, 
who stretched out the heavens, 
and founded the earth. 
And you tremble continually, all day, 
because of the wrath of the oppressor 
when he \textit{takes aim}\lebnote{Literally “sets up”} to destroy. 
But\lebnote{Or “And”} where is the wrath of the oppressor?
\verse The fettered one shall make haste to be freed. 
And he shall not die in\lebnote{Or “to”} the pit, 
and he shall not lack his bread.
\verse For\lebnote{Or “And”} I am Adonai, your God, 
who stirs up\lebnote{Or “stirring up”} the sea, \textit{so that}\lebnote{Or “and”} its waves roar; 
Adonai of hosts is his name.
\verse And I have put my words in your mouth, 
and I have covered you in the shadow of my hand, 
to plant the heavens 
and to found the earth, 
saying to Zion, ‘You are my people.’ ”
\verse Rouse yourself! Rouse yourself! 
Stand up, Jerusalem, who have drunk from the hand of Adonai the cup of his wrath; 
you have drunk the goblet, the cup of staggering; 
you have drained it out.
\verse There is no one who guides her among\lebnote{Or “from”} all the children she has borne, 
and there is no one who grasps her by the hand among\lebnote{Or “from”} all the children she raised.
\verse Two things here have happened to you—who will \textit{show sympathy}\lebnote{Literally “wander”} for you?— 
devastation and destruction, famine and sword—who will comfort you?
\verse Your children have fainted; 
they lie at the head of all the streets, like an antelope in\lebnote{Or “of”} a snare, 
those who are full of the wrath of Adonai, 
the rebuke of your God.
\verse Therefore hear now this afflicted one 
and drunken one but\lebnote{Or “and”} not from wine.
\verse Thus says your Lord, Adonai, 
and your God pleads the cause of his people: 
“Look! I have taken from your hand the cup of staggering. 
You shall not \textit{continue}\lebnote{Literally “do again”} to drink the goblet, the cup of my wrath, any longer.
\verse And I will put it in the hand of your tormenters, 
who have said to \textit{you},\lebnote{Literally “your inner self”} ‘Bow down that\lebnote{Or “and”} we may \textit{pass}\lebnote{Or “stride”} over you!’ 
And you have \textit{made}\lebnote{Literally “placed”} your back like the ground, 
and like the street for those who \textit{pass}\lebnote{Or “stride”} over you.”
\end{biblechapter}

\begin{biblechapter} % Isaiah 52
\verseWithHeading{Adonai Redeems Zion} Awake! Awake; put on your strength, Zion! 
Put on the garments of your beauty, Jerusalem, \textit{holy city}!\lebnote{Literally “city of the holiness”} 
For the uncircumcised and the unclean shall not \textit{continue to}\lebnote{Literally “do again he shall”} enter you any longer.
\verse Shake yourself free from the dust! Rise up; sit, Jerusalem! 
Free yourselves from the bonds of your neck, captive daughter of Zion!
\verse For thus says Adonai: “You were sold for nothing, 
and you shall be redeemed without money.”
\verse For thus says the Lord Adonai:
\verse “My people went down to Egypt in the beginning, to dwell as aliens there, and Assyria oppressed him \textit{without cause}.\lebnote{Literally “in nothing”}
\verse And now what do I have here?”\textit{declares}\lebnote{Literally “declaration of”} Adonai, “for my people is taken without cause. Its rulers howl,” \textit{declares}\lebnote{Literally “declaration of”} Adonai—“and my name is reviled continually, all day.
\verse How delightful on the mountains are the feet of him who brings good news, 
who announces peace, who brings good news, who announces salvation, 
who says to Zion, “Your God reigns as a king.”
\verse The voices\lebnote{Hebrew “voice”} of your watchmen! They lift up their voices;\lebnote{Hebrew “voice”} 
together they sing for joy; 
for they \textit{clearly}\lebnote{Literally “eye in eye “} see Adonai’s return to Zion.
\verse Break forth, 
sing for joy together, ruins of Jerusalem, 
for Adonai has comforted his people; 
he has redeemed Jerusalem.
\verse Adonai has bared \textit{his holy arm}\lebnote{Literally “the arm of his holiness”} to the eyes of all the nations, 
and all the ends\lebnote{Hebrew “end”} of the earth shall see the salvation of our God.
\verse Depart, depart, go out from there! You must not touch any unclean thing. 
Go out from the midst of it, keep clean, you who carry the vessels of Adonai.
\verse For you shall not go out in haste, 
and you shall not go in flight, 
for Adonai is going before you, 
and your rear guard is the God of Israel.
\verseWithHeading{The Servant’s Suffering and Exaltation} Look, my servant shall achieve success; 
he shall be exalted, 
and he shall be lifted up, 
and he shall be very high.
\verse Just as many were appalled at you— 
such was his appearance beyond human disfigurement, 
and his form beyond the sons of mankind—
\verse so he shall sprinkle many nations; 
because of him, kings shall shut their mouths.\lebnote{Hebrew “mouth”} 
For they shall see what has not been told them, 
and they shall consider with full attention what they have not heard.
\end{biblechapter}

\begin{biblechapter} % Isaiah 53
\verse Who has believed our message, 
and to whom has the arm of Adonai been revealed?
\verse For\lebnote{Or “And”} he went\lebnote{Or “grew”} up like a shoot before him, 
and like a root from dry ground. 
He had no form and no majesty that\lebnote{Or “and”} we should see him, 
and no appearance that\lebnote{Or “and”} we should take pleasure in him.
\verse He was despised and rejected by men, 
a man of suffering, and acquainted with\lebnote{Or “knowledgeable of”} sickness, 
and like \textit{one from whom others hide their faces},\lebnote{Literally “hiding of face from him”} he was despised, 
and we did not hold him in high regard.
\verse However, he was the one who lifted up our sicknesses, 
and he carried our pain, 
yet\lebnote{Or “and”} we ourselves assumed him stricken, 
struck down by God and afflicted.
\verse But\lebnote{Or “And”} he was pierced\lebnote{Or “wounded”} because of our transgressions, 
crushed because of our iniquities; 
the chastisement for\lebnote{Or “of”} our peace\lebnote{Or “healing”} was upon him, 
and by his wounds\lebnote{Hebrew “wound”} \textit{we were healed}.\lebnote{Literally “it was healed for us”}
\verse All of us have wandered about like sheep; 
we each have turned to his own way; 
and Adonai let fall on him 
the iniquity of us all.
\verse He was oppressed and afflicted, 
yet\lebnote{Or “and”} he did not open his mouth; 
he was brought like a lamb to the slaughter, 
and like a sheep is dumb before its shearers, 
so\lebnote{Or “and”} he did not open his mouth.
\verse He was taken by restraint of justice, 
and who concerned himself with his generation? 
For he was cut off from the land of the living; 
he received a blow because of the transgression of my people.
\verse He made\lebnote{Or “gave”} his grave with the wicked, 
and with the rich in his death, 
although he had done no violence, 
and there was no deceit in his mouth.
\verse Yet\lebnote{Or “And”} Adonai was pleased to crush him; 
he \textit{made him sick}.\lebnote{Literally “he made sick”} 
If she\lebnote{Or “you”} places\lebnote{Or “makes”} his life a guilt offering,\lebnote{Or “compensation”} 
he will see offspring. 
He will prolong days, 
and the will of Adonai will succeed in his hand.
\verse From the trouble of his life\lebnote{Or “soul”} he will see;\lebnote{Dead Sea Scrolls add “light”} 
he will be satisfied. 
In his knowledge, the righteous one, my servant, shall declare many righteous,\lebnote{Or “right”} 
and he is the one who will bear their iniquities.
\verse Therefore, I will divide to him a portion among the many,\lebnote{Or “great”} 
and with the strong ones he will divide bounty, 
\textit{because}\lebnote{Literally “Instead of that”} he poured his life out to death 
and was counted with the transgressors; 
and\lebnote{Or “yet”} he was the one who bore the sin of many 
and will intercede for the transgressors.
\end{biblechapter}

\begin{biblechapter} % Isaiah 54
\verseWithHeading{The Fertile Wife of Adonai} “Sing for joy, barren woman; who has not borne! 
Burst forth into rejoicing and rejoice, she who has not been in labor! 
For the children of the desolate woman are more than the children of the married woman,” says Adonai.
\verse “Enlarge the site of your tent, 
and let them stretch out the tent curtains of your dwelling place. 
You must not spare; make your tent cords long 
and strengthen your pegs,
\verse for you will spread out to the right and to the left. 
And your descendants\lebnote{Hebrew “descendant”} will be heir to the nations, 
and they will inhabit desolate towns.
\verse You must not fear, for you will not be ashamed, 
and you must not be confounded, for you will not feel abashed, 
for you will forget the shame of your youth, 
and you will no longer remember the disgrace of your widowhood.
\verse For your husband\lebnote{The Hebrew is plural} is your maker,\lebnote{The Hebrew is plural} his name is Adonai of hosts; 
and your redeemer is the holy one of Israel, he is called the God of all of the earth.
\verse For Adonai has called you like a wife forsaken and hurt of spirit, 
like\lebnote{Or “and”} the wife of childhood when she is rejected, says your God.
\verse I abandoned you \textit{for a short}\lebnote{Literally “in a small”} moment, 
but\lebnote{Or “and”} I will gather you with great compassion.
\verse I hid my face from you for a moment, in the flowing of anger, 
but\lebnote{Or “and”} I will have compassion on you with everlasting faithfulness” says your redeemer, Adonai.
\verse “For this is like the waters of Noah to me, 
when I swore \textit{that the waters of Noah would never again pass}\lebnote{Literally “from the passing over of the waters of Noah again”} over the earth, 
so I swore \textit{that I would not be}\lebnote{Literally “from being “} angry at you and rebuke you.
\verse For the mountains may depart, 
and the hills may sway, 
but\lebnote{Or “and”} my faithfulness shall not depart from you, 
and my covenant of peace shall not sway,” 
says Adonai, who has compassion on you,
\verse “O afflicted one, driven away, who is not consoled. 
Look! I am about to set your stones in hard mortar, 
and I will lay your foundation with sapphires.
\verse And I will \textit{make}\lebnote{Literally “place”} your battlements of ruby, 
and your gates of stones of beryl, 
and all your wall of precious stones.
\verse And all your children shall be pupils of Adonai, 
and the peace of your children shall be great.
\verse In righteousness you shall be established.
\verse Be far from oppression, for you shall not fear, 
and from terror, for it shall not come near you.
\verse If indeed one attacks, it is not from me; 
whoever attacks you shall fall because of you.
\verse Look! I myself have created the craftsman who blows\lebnote{Or “blowing”} the fire of coals, 
and who produces\lebnote{Or “producing”} a weapon for his work; 
also\lebnote{Or “and”} I myself have created the destroyer to destroy. \textit{declares}\lebnote{Literally “declaration of”} Adonai.
\end{biblechapter}

\begin{biblechapter} % Isaiah 55
\verseWithHeading{Invitation to True Rewards} “Ho! Everyone thirsty, come to the waters! 
And whoever has no money, come, buy and eat, 
and come, buy without money, 
wine and milk without price!
\verse Why do you weigh out money for what is not food, 
and your labor for \textit{what cannot satisfy}?\lebnote{Literally “what is not for satiation”} 
Listen carefully to me, and eat what is good, 
and let your soul take pleasure in \textit{rich}\lebnote{Literally “fatty”} food.
\verse Extend your ear, and come to me! 
Listen so that\lebnote{Or “and”} your soul may live, 
and I will \textit{make}\lebnote{Literally “cut”} an everlasting covenant with\lebnote{or “for”} you, 
the enduring proofs of \textit{the mercies shown to}\lebnote{Literally “mercy of”} David.
\verse Look! I made\lebnote{Or “gave”} him a witness to the peoples, 
a leader and a commander for the peoples.
\verse Look! You shall call a nation that you do not know, 
and a nation that does not know you shall run to you, 
because of Adonai your God, 
and the holy one of Israel, for he has glorified you.”
\verse Seek Adonai \textit{while he lets himself be found};\lebnote{Literally “in his letting himself be found”} 
call him \textit{while he is}\lebnote{Literally “in his being”} near.
\verse Let the wicked forsake his way, 
and the man of sin his thoughts. 
And let him return to Adonai, that\lebnote{Or “and”} he may take pity on him, 
and to our God, for he will \textit{forgive manifold}.\lebnote{Literally “make numerous to”}
\verse “For my thoughts are not your thoughts, 
and your ways are not my ways,” \textit{declares}\lebnote{Literally “declaration of”} Adonai.
\verse “For as the heavens are \textit{higher}\lebnote{Literally “high”} than the earth, 
so my ways are \textit{higher}\lebnote{Literally “high”} than your ways, 
and my thoughts than your thoughts.
\verse For just as the rain and the snow come down from heaven, 
and they do not return there 
except they have watered the earth thoroughly 
and cause it to bring forth and sprout, 
and give seed to the sower 
and bread to the eater,
\verse so shall be my word that goes out from my mouth. 
It shall not return to me without success, 
but shall accomplish\lebnote{Or “do”} what I desire 
and be successful in the thing for which I sent it.
\verse For you shall go out in joy, 
and you shall be led in peace. 
The mountains and the hills shall break forth before you, rejoicing 
and all the trees of the field shall clap hands.\lebnote{Hebrew “hand”}
\verse Instead of the thorn bush, the juniper shall go up; 
instead of the brier, the myrtle shall go up, 
and it shall serve as a memorial to Adonai, 
for an everlasting sign that shall not be cut off.”
\end{biblechapter}

\begin{biblechapter} % Isaiah 56
\verseWithHeading{Salvation for All Who Obey} Thus says Adonai:
\verse “Observe justice 
and do righteousness, 
for my salvation is close to coming, 
and my justice to being revealed.
\verse Happy is the man who does this, 
and the son of humankind who keeps hold of it, 
who keeps the Sabbath \textit{so as not to profane}\lebnote{Literally “from profaning”} it, 
and who keeps his hand from doing any\lebnote{Or “all”} evil.”
\verse And do not let the \textit{foreigner}\lebnote{Literally “son of the foreign country”} who joins himself to Adonai say, “Surely Adonai will separate me from his people.” 
And do not let the eunuch say, “Look! I am a dry tree!”
\verse For thus says Adonai, “To the eunuchs who keep my Sabbaths, 
and choose that in which I delight, 
and who keep hold of my covenant.
\verse And I will give them a monument and a name in my house and within my walls, 
better than sons and daughters; 
I will give him an everlasting name 
that shall not be cut off.
\verse And the \textit{foreigners}\lebnote{Literally “sons of the foreign country”} who join themselves to Adonai to serve him 
and to love the name of Adonai, to become his servants, 
every one who keeps the Sabbath, \textit{so as not to profane}\lebnote{Literally “from profaning”} it, 
and those who keep hold of my covenant,
\verse \textit{declares}\lebnote{Literally “declaration of”} the Lord Adonai, who gathers\lebnote{Or “gathering”} the scattered ones of Israel (still I will gather to\lebnote{Or “upon”} him, to his gathered ones).
\verseWithHeading{Wicked Leaders} All wild animals in the field, come, 
to devour,\lebnote{Or “eat”} all wild animals in the forest!
\verse His watchmen are blind, 
none of them know. 
They are all dumb dogs; 
they are unable to bark, 
panting, lying down, 
loving to slumber.
\verse And the dogs \textit{have a greedy appetite};\lebnote{Literally “are strong of soul”} 
they \textit{are never satisfied}.\lebnote{Literally “do not know satiation”} 
And they are the shepherds! 
They do not have\lebnote{Or “know”} understanding. 
They all turn to their own way, 
each one for his own gain, \textit{every one of them}.\lebnote{Literally “from his end “}
\verse “Come, let me take wine, 
and let us carouse with intoxicating drink, 
and tomorrow will be like today, 
\textit{very great indeed}.”\lebnote{Literally “excessively”}
\end{biblechapter}

\begin{biblechapter} % Isaiah 57
\verseWithHeading{Idolatry is Adultery} The righteous one perishes, and there is no one who takes\lebnote{Or “sets”} it to heart. 
And men of faithfulness are gathered, while there is no one who understands, 
for the righteous is gathered from the presence of wickedness.
\verse he enters into peace; 
they will rest on their beds, 
walking straight ahead of him.
\verse “But\lebnote{Or “And”} you, come near here, you children of a soothsayer, 
offspring of an adulterer and she who commits fornication.
\verse At whom do you make fun? 
At whom do you \textit{open}\lebnote{Literally “widen”} your mouth and \textit{stick out}\lebnote{Literally “make long”} your tongue? 
Are you not children of transgression, 
offspring of deception,
\verse who burn with lust among the oaks, 
under every leafy tree, 
who slaughter children in the valleys, 
under the clefts of the rocks?
\verse Your portion is among the smooth stones of the valley; 
they, they are your lot; 
indeed, to them you have poured out a drink offering, 
you have brought a food offering. Shall I relent concerning these things?
\verse You have set your bed upon a high and lofty mountain; 
indeed, you went up there to slaughter sacrifice.
\verse And you have set your symbol behind the door and the doorpost; 
for you depart\lebnote{The reading tradition (\textit{Qere}) has “uncover,” which makes no sense in context, but with different vowels it would be “depart”} from me, and you go up; 
you make your bed wide, and you \textit{make a deal with}\lebnote{Literally “cut off for yourself from”} them, 
you have loved their bed; you have seen their \textit{genitals}.\lebnote{Literally “hand”}
\verse And you climbed down to the king with oil, 
and you made your perfumes numerous, 
and you sent your envoys \textit{far away},\lebnote{Literally “to from far”} 
and you \textit{sent down deep}\lebnote{Literally “humiliated”} to Sheol.
\verse You grow weary by the greatness of your way, 
but you did not say, ‘Despairing!’ 
You found the \textit{renewal}\lebnote{Literally “life”} of your \textit{strength},\lebnote{Literally “hand;” compare the “hand” in verse 8} 
therefore you do not grow weak.
\verse And of whom were you afraid and feared, that you deceived 
and did not remember me? 
Did you not place it on your heart? 
Have I not been silent, even from long ago, 
and so you do not fear me?
\verse I myself will declare your righteousness and your works, 
but\lebnote{Or “and”} they will not benefit you.
\verse When \textit{you cry},\lebnote{Literally “your crying”} let your collection deliver you, 
and the wind will carry all of them away; 
a breath will take them away. 
But\lebnote{Or “And”} he who takes refuge in me shall take possession of the land, 
and he shall inherit my \textit{holy mountain}.”\lebnote{Literally “mountain of holiness “}
\verseWithHeading{Peace for the Contrite} And one shall say, “Build up, build up! Clear the way! 
Remove the obstacles\lebnote{Hebrew “obstacle”} from the way of my people!”
\verse For thus says the high and lofty one who resides forever, and whose name is holy:
\verse “I reside in a high and holy place, 
and with the \textit{contrite}\lebnote{Literally “crushed”} and humble of spirit, 
to revive the spirit of the humble, 
and to revive the heart of the \textit{contrite}.\lebnote{Literally “crushed”}
\verse For I will not attack forever, 
and I will not be angry forever, 
for the spirit will grow faint before me, 
and the breaths that I myself I have made.
\verse I was angry because of his sin of gain, and I struck him; 
I hid and I was angry, but\lebnote{Or “and”} he walked apostate, in the ways\lebnote{Hebrew “way”} of his heart.
\verse I have seen his ways, but\lebnote{Or “and”} I will heal him; 
and I will lead him and give him and his mourners comfort as a recompense,
\verse creating fruit of lips. 
Peace, peace to the far and near,” says Adonai, “and I will heal him.
\verse But\lebnote{Or “And”} the wicked are like the churning sea, 
that is not able to keep quiet, 
and its waters toss up mire and mud.
\end{biblechapter}

\begin{biblechapter} % Isaiah 58
\verseWithHeading{True Fasts and Sabbaths} “Call with the throat; you must not keep back! 
lift up your voice like a trumpet, 
and declare\lebnote{Or “announce”} to my people their rebellion, 
and to the house of Jacob their sins.
\verse Yet\lebnote{Or “And”} they seek me day by day, 
and they desire the knowledge of my ways 
like a nation that \textit{practiced}\lebnote{Literally “did”} righteousness, 
and had not forsaken the judgment of its God; 
they ask me for \textit{righteous judgments},\lebnote{Literally “judgments of righteousness”} 
they desire the closeness of God.
\verse ‘Why do we fast, and you do not see it? 
We humiliate our soul, and you do not notice it?’ 
Look! You find delight on the day of your fast, 
and you oppress all your workers!
\verse Look! You fast to quarrel and strife, 
and to strike with a \textit{wicked fist}.\lebnote{Literally “fist of wickedness”} 
You shall not fast as you do \textit{today},\lebnote{Literally “the day”} 
to \textit{make your voice heard}\lebnote{Literally “cause to hear your voice”} on the height.
\verse Is the fast I choose like this, 
a day for humankind to humiliate \textit{himself}\lebnote{Literally “his soul”}? 
To bow his head like a reed, 
and \textit{make}\lebnote{Literally “he makes”} his bed on sackcloth and ashes; 
you call this a fast 
and a day of pleasure to Adonai?
\verse Is this not the fast I choose: to release the bonds of injustice, 
to untie the ropes of the yoke, 
and to let the oppressed go free, 
and \textit{tear}\lebnote{Literally “you must tear”} every yoke to pieces?
\verse Is it not to break your bread for the hungry? 
You must bring \textit{home}\lebnote{Literally “a house”} the poor, the homeless. 
When you see the naked, you must cover him, 
and you must not hide yourself from your \textit{relatives}.\lebnote{Literally “flesh”}
\verse Then your light shall break forth like the dawn, 
and your healing shall grow quickly. 
And your salvation shall go before you; 
the glory of Adonai will be your rear guard.
\verse Then you shall call, and Adonai himself will answer. 
You shall cry for help, and he will say, ‘Here I am!’ 
If you remove from among you the yoke, 
the \textit{finger-pointing}\lebnote{Literally “stretching out of the finger”} and \textit{evil speech},\lebnote{Literally “the speaking of evil”}
\verse if\lebnote{Or “and”} you offer your soul to the hungry, 
and you satisfy the appetite of the afflicted, 
then\lebnote{Or “and”} your light shall rise in the darkness, 
and your darkness will be like noon.
\verse And Adonai will lead you continually, and satisfy your soul in a barren land, 
and he will strengthen your bones, 
and you shall be like a well-watered garden, 
and like a spring of water whose water does not fail.
\verse And they shall rebuild ancient ruins from you; 
you shall erect the foundations of \textit{many generations,}\lebnote{Literally “generation and generation”} 
and \textit{you shall be called}\lebnote{Literally “shall be called to you”} the bricklayer of the breach, 
the restorer of paths to live in.
\verse If you hold your foot back from the Sabbath, 
from doing your affairs on \textit{my holy day},\lebnote{Literally “the day of my holiness”} 
if\lebnote{Or “and”} you call the Sabbath a pleasure, 
the holy day of Adonai honorable, 
if\lebnote{Literally “and”} you honor him \textit{more than}\lebnote{Literally “from”} doing your ways, 
than finding your affairs\lebnote{Hebrew “affair”} and speaking a word,
\verse then you shall take your pleasure in\lebnote{Or “on”} Adonai, 
and I will make you ride upon the heights of the earth, 
and I will feed you the heritage of Jacob your ancestor,\lebnote{Or “father”} for the mouth of Adonai has spoken.”
\end{biblechapter}

\begin{biblechapter} % Isaiah 59
\verseWithHeading{Rampant Transgression} Look! The hand of Adonai is not too short \textit{to save},\lebnote{Literally “from saving”} 
and his ear is not too dull \textit{to hear}.\lebnote{Literally “from hearing”}
\verse Rather, your iniquities have been \textit{barriers}\lebnote{Literally “dividers”} between you and your God, 
and your sins have hidden his face from you, from hearing.
\verse For your hands are defiled with blood, 
and your fingers with iniquity. 
Your lips have spoken lies, 
your tongue speaks wickedness.
\verse There is nobody \textit{who pleads with}\lebnote{Literally “calling in”} justice, 
and there is nobody who judges with honesty. 
They rely on nothing 
and speak vanity. 
They conceive trouble 
and beget iniquity;
\verse they hatch viper eggs, 
and they weave a spider web. 
One who eats their eggs dies, 
and that which is pressed is hatched as a serpent.
\verse Their webs cannot become clothing, 
and they cannot cover themselves with their works. 
Their works are works of iniquity, 
and deeds\lebnote{Hebrew “deed”} of violence are in their hands.
\verse Their feet run to evil, 
and they hasten to shed innocent blood. 
Their thoughts are thoughts of iniquity; 
devastation and destruction are in their highways.
\verse They do not know the way of peace, 
and there is no justice in their firm paths. 
They have made their paths crooked for themselves; 
everyone \textit{who walks}\lebnote{Literally “walking”} in it knows no peace.
\verse Therefore justice is far from us, 
and righteousness does not reach us. 
We wait for light, but\lebnote{Or “and”} look! there is darkness; 
for brightness, but we walk in darkness.
\verse We grope like the blind along a wall, 
and we grope \textit{as without}\lebnote{Literally “like there are no”} eyes. 
We stumble at noon as in the twilight; 
among the strong we are like the dead.
\verse We all groan like bears, 
and we coo mutteringly like doves. 
We wait for justice, but\lebnote{Or “and”} there is none; 
for salvation, but it is far from us.
\verse For our transgressions are numerous before you, 
and our sins \textit{testify}\lebnote{Or “answers”} against us. 
Indeed, our transgressions are with us, 
and we know our iniquities:
\verse transgressing and denying Adonai, 
and turning \textit{away from following}\lebnote{Literally “back from behind”} our God; 
speaking oppression and falsehood, 
conceiving and uttering words of deception from the heart.
\verse And justice is pushed back, 
and righteousness stands afar; 
for truth stumbles in the public square, 
and straightforwardness is unable to enter,
\verse and truth is missing, 
and he who turns aside from evil is plundered. 
And Adonai saw, 
and it was displeasing in his eyes that there was no justice
\verse And he saw that there was no man, 
and he was appalled that there was no one who intercedes, 
so\lebnote{Or “and”} his arm came to assist him, 
and his righteousness \textit{was what}\lebnote{Literally “it”} sustained him.
\verse And he put on righteousness like a breastplate, 
and a helmet of salvation on his head, 
and he put on garments of vengeance for clothing, 
and he wrapped himself in zeal as in a robe.
\verse According to deeds, so he will repay; 
wrath to his enemies, requital to those who are his \textit{enemies}.\lebnote{Literally “hostile”} 
He will repay requital to the coastlands.
\verse So\lebnote{Or “And”} they shall fear the name of Adonai from the west, 
and his glory from the sunrise, 
for he will come like a narrow stream; 
the wind of Adonai drives it on.
\verse “And a redeemer will come to Zion, 
to those in Jacob who turn away from transgression,” \textit{declares}\lebnote{Literally “declaration of”} Adonai.
\verse “And as for me, this is my covenant with them, says Adonai: my spirit that is upon you, 
and my words that I have placed in your mouth 
shall not depart from your mouth, 
or\lebnote{Or “and”} from the mouths\lebnote{Hebrew “mouth”} of your \textit{children},\lebnote{Literally “offspring”} 
or\lebnote{Or “and”} from the mouths\lebnote{Hebrew “mouth”} of \textit{your children’s children},”\lebnote{Literally “the offspring of your offspring”} 
says Adonai, “from now on and forever.”
\end{biblechapter}

\begin{biblechapter} % Isaiah 60
\verseWithHeading{Zion Glorified} “Arise, shine! For your light has come, 
and the glory of Adonai has risen on you.
\verse For look! darkness shall cover the earth, 
and thick darkness the peoples, 
but\lebnote{Or “and”} Adonai will rise on you, 
and his glory will appear over you.
\verse And nations shall come to your light, 
and kings to the bright light of your sunrise.
\verse Lift up your eyes all around and see! 
All of them gather; they come to you. 
Your sons shall come from afar, 
and your daughters shall be looked after on the hip.
\verse Then you shall see 
and you shall be radiant; 
and your heart shall tremble 
and open itself wide, 
because the abundance of the sea shall fall upon you; 
the wealth of the nations shall come to you.
\verse A multitude of camels shall cover you, 
the young male camels of Midian and Ephah. 
All those from Sheba shall come; 
they shall bring gold and frankincense, 
and they shall proclaim the praise of Adonai.
\verse All the flocks of Kedar shall be gathered to you; 
the rams of Nebaioth shall serve you. 
They shall present a sacrifice for favor on my altar, 
and I will glorify \textit{my honorable house}.\lebnote{Literally “house of my honor”}
\verse Who are these? They fly like a cloud, 
and like doves to their coops.
\verse Because the coastlands wait for me, 
and the ships of Tarshish are first 
to bring your children from afar, 
their silver and gold with them, 
for the name of Adonai your God, 
and for the holy one of Israel, because he has glorified you.
\verse And \textit{foreigners}\lebnote{Literally “sons of a foreign country”} shall build your walls, 
and their kings shall serve you, 
for in my anger I struck you, 
but\lebnote{Or “and”} in my favor I have taken pity on you.
\verse And your gates shall continually be open, 
day and night they shall not be shut, 
to bring you the wealth of nations, 
and their kings shall be led.
\verse For the nation and the kingdom that will not serve you shall perish, 
and the nations shall be utterly devastated.
\verse The glory of Lebanon shall come to you; 
the cypress, the plane, and the pine together, 
to glorify the place of my sanctuary, 
and I will do honor to the place of my feet.
\verse And the children of those who oppressed you shall come to you bending low, 
and all those who treated you disrespectfully shall bow down at the soles of your feet. 
And they shall call you the city of Adonai, 
Zion of the holy one of Israel.
\verse Instead of you being forsaken 
and hated \textit{with no one passing through,}\lebnote{Or “and there is not one who passes through”} 
I will \textit{make}\lebnote{Literally “place”} you an everlasting \textit{object of pride},\lebnote{Literally “pride”} 
a joy of \textit{coming generations}.\lebnote{Literally “generation and generation”}
\verse And you shall suck the milk of nations, 
and suck the breast of kings, 
and you shall know that I am Adonai your Savior, 
and your Redeemer, the Strong One of Jacob.
\verse Instead of bronze I will bring gold, 
and instead of iron I will bring silver, 
and instead of wood, bronze, 
and instead of stones, iron. 
And I will \textit{appoint}\lebnote{Literally “put”} peace as your overseer, 
and righteousness as \textit{your ruling body}.\lebnote{Literally “they who force you to work”}
\verse Violence shall no longer be heard in your land; 
devastation or\lebnote{Or “and”} destruction on your borders. 
And you shall call your walls Salvation, 
and your gates, Praise.
\verse The sun shall no longer be your light by day, 
and for bright light the moon shall not give you light, 
but\lebnote{Or “and”} Adonai will be your everlasting light, 
and your God your glory.
\verse Your sun shall no longer go down, 
and your moon shall not \textit{wane},\lebnote{Literally “be taken away”} 
for Adonai himself will be your everlasting light, 
and your days of mourning shall come to an end.
\verse And all your people shall be righteous; 
they shall take possession of the land forever, 
the shoot of his\lebnote{The reading tradition (\textit{Qere}) has “my”} planting, 
the work of my hands, to show my glory.
\verse The small one shall become a tribe, 
and the smallest one a mighty nation. I am Adonai; I will hasten it in its time.”
\end{biblechapter}

\begin{biblechapter} % Isaiah 61
\verseWithHeading{The Year of Adonai’s Favor} The Spirit of the Lord Adonai is upon me, 
because Adonai has anointed me, 
he has sent me to bring good news to the oppressed, 
to bind up \textit{the brokenhearted},\lebnote{Literally “those broken of heart”} 
to \textit{proclaim}\lebnote{Literally “call”} release to the captives 
and liberation to those who are bound,
\verse to \textit{proclaim}\lebnote{Literally “call”} the year of Adonai’s favor, 
and our God’s day of vengeance, 
to comfort all those in mourning,
\verse to \textit{give}\lebnote{Literally “place”} for those in mourning in Zion, 
to give them a head wrap instead of ashes, 
the oil of joy instead of mourning, 
a garment of praise instead of a faint spirit. 
And \textit{they will be called}\lebnote{Literally “it will be called to them”} oaks of righteousness, 
the planting of Adonai, to show his glory.
\verse And they shall build the ancient ruins, 
they shall erect the former deserted places. 
And they shall restore the \textit{devastated cities},\lebnote{Literally “cities of devastation”} 
the deserted places of \textit{many generations}.\lebnote{Literally “generation and generation”}
\verse And strangers shall stand and feed your flocks,\lebnote{Hebrew “flock”} 
and \textit{foreigners}\lebnote{Literally “sons of a foreign country”} shall be your farmers and vinedressers.
\verse But\lebnote{Or “And”} you shall be called the priests of Adonai, 
\textit{you will be called}\lebnote{Literally “it shall be said of you”} servers of our God. 
You shall eat the wealth of the nations, 
and you shall boast in their riches.
\verse Instead of your shame, a double portion, 
and instead of insult, they will rejoice over their portion. 
Therefore they will take a double portion in their land; 
\textit{they shall possess everlasting joy}.\lebnote{Literally “possession of joy everlasting shall be for them”}
\verse For I, Adonai, love justice, 
hate robbery and\lebnote{Or “in”} injustice, 
and I will \textit{faithfully}\lebnote{Literally “in the faithfulness”} give their reward, 
and I will \textit{make}\lebnote{Literally “cut”} an everlasting covenant \textit{with}\lebnote{Literally “to”} them.
\verse And their descendants\lebnote{Hebrew “descendant”} will be known among the nations, 
and their offspring in the midst of the peoples. 
All those who see them shall recognize them, 
that they are descendants\lebnote{Hebrew “descendant”} whom Adonai has blessed.
\verse I will rejoice greatly in Adonai; 
my being shall shout in exultation in my God. 
For he has clothed me with garments of salvation, 
he has covered me with the robe of righteousness, 
as a bridegroom adorns himself with a head wrap like a priest, 
and as a bride adorns herself with her jewelry.
\verse For as the earth \textit{produces}\lebnote{Literally “brings out”} its sprout, 
and as a garden makes its plants sprout, 
so the Lord Adonai will make righteousness sprout, 
and praise before all the nations.
\end{biblechapter}

\begin{biblechapter} % Isaiah 62
\verseWithHeading{Zion’s New Identity} For the sake of Zion I will not be silent, 
and for the sake of Jerusalem I will not maintain a quiet attitude, 
until her righteousness goes out like the bright light, 
and her salvation burns like a torch.
\verse And the nations shall see your righteousness, 
and all the kings your glory, 
and \textit{you will be called}\lebnote{Literally “it shall be called to you”} a new name 
that the mouth of Adonai will designate.
\verse And you shall be a crown of beauty in the hand of Adonai, 
and a headband of royalty in the hand of your God.
\verse It shall no longer be said of you, “Forsaken,” 
and it shall no longer be said of your land, “Desolation!” 
but \textit{you will}\lebnote{Literally “to you it shall”} be called “My Delight Is In Her,” 
and your land, “Married,” 
for Adonai delights in you, 
and your land shall be married.
\verse For as a young man marries a virgin, 
so shall your sons marry you, 
and as is the joy of the bridegroom over the bride, 
so shall your God rejoice over you.
\verse I have appointed watchmen upon your walls, Jerusalem; 
all day and all night they shall never be silent. 
You who profess Adonai have no rest.
\verse And you must not give him rest until he establishes, 
and until he \textit{makes}\lebnote{Literally “places”} Jerusalem \textit{an object of praise}\lebnote{Literally “a praise”} in the earth.
\verse Adonai has sworn by his right hand, 
and by \textit{his mighty arm},\lebnote{Literally “the arm of his might”} 
“Surely I will never again give your grain as food to \textit{your enemies},\lebnote{Literally “those who are hostile to you”} 
and surely \textit{foreigners}\lebnote{Literally “sons of a foreign country”} shall not drink your new wine at which you have labored.”
\verse But those who gather it shall eat it, and they shall praise Adonai, 
and those who gather it shall drink it in \textit{my holy courts}.\lebnote{Literally “the courts of my holiness”}
\verse Pass through, pass through the gates! 
Make the way clear \textit{for}\lebnote{Literally “of”} the people! 
Pile up, pile up the highway; clear it of stones!\lebnote{Hebrew “stone”} 
Lift up an ensign over the peoples!
\verse Look! Adonai has proclaimed to the end of the earth:
\verse Say to daughter Zion, “Look, your salvation is coming! 
Look, his reward is with him, 
and his reward before him.
\end{biblechapter}

\begin{biblechapter} % Isaiah 63
\verseWithHeading{Adonai’s Day of Vengeance} Who is this, coming from Edom, 
from Bozrah in bright red garments? 
Who is this honored in his garment, 
lying down in his great strength? 
“It is I, speaking in justice, 
\textit{mighty}\lebnote{Literally “great”} to save!”
\verse Why are your garments\lebnote{Hebrew “garment”} red, 
and your garments like he who treads in the winepress?
\verse “I have trodden the winepress alone, 
and there was no man from the peoples with me. 
And I trod them in my anger, 
and I trampled them in my wrath, 
and spattered their juice on my garments, 
and stained all my clothing,
\verse for the day of vengeance was in my heart, 
and the year of my blood-vengeance has come!
\verse And I looked, but\lebnote{Or “and”} there was no helper, 
and I was appalled, but\lebnote{Or “and”} there was no one who sustains, 
so\lebnote{Or “and”} my arm came to assist me, 
and my wrath \textit{was what}\lebnote{Literally “it”} sustained me.
\verse And I trampled peoples in my anger, 
and I made them drunk in my wrath, 
and I brought their juice down to the earth.”
\verseWithHeading{Remembering Adonai’s Mercy} I will mention the loyal love of Adonai, the praises of Adonai, according to all that Adonai has done for us, 
and the greatness of goodness to the house of Israel that he has done to them according to his mercy and the abundance of his loyal love.
\verse And he said, “Surely my people are children; they will not break faith.” And he became a Savior to them.
\verse In all their distress, \textit{there was no distress},\lebnote{The reading tradition (\textit{Qere}) has “he had distress”} 
and the messenger of his \textit{presence}\lebnote{Literally “face”} saved them, 
in his love and compassion he himself redeemed them, 
and he lifted them up, 
and he supported them all the days of old.
\verse But\lebnote{Or “And”} they \textit{were the ones who}\lebnote{Literally “they”} rebelled, 
and they grieved \textit{his Holy Spirit},\lebnote{Literally “the spirit of his holiness”} 
so\lebnote{Or “and”} he \textit{became an enemy to them};\lebnote{Literally “was changed to them to one who is hostile”} 
he himself fought against them.
\verse Then\lebnote{Or “And”} his people remembered the days of old, of Moses. 
Where is the one who led up them from the sea with the shepherds of his flock? 
Where is the one who puts \textit{his Holy Spirit}\lebnote{Literally “the spirit of his holiness”} inside him,
\verse \textit{who made}\lebnote{Literally “bringing”} his magnificent arm move \textit{at}\lebnote{Literally “to”} the right hand of Moses, 
\textit{who divided}\lebnote{Literally “dividing”} the waters \textit{before them},\lebnote{Literally “from their face”} 
to make an everlasting name for himself,
\verse \textit{who led}\lebnote{Literally “leading”} them through the depths? 
They did not stumble like a horse in the desert;
\verse like cattle in the valley that goes down, the Spirit of Adonai \textit{gave him rest},\lebnote{Literally “caused to rest”} 
so you lead your people to make a magnificent name for yourself.
\verseWithHeading{Prayer for Mercy} Look from heaven, 
and see from the lofty residence of your holiness and glory. 
Where are your zeal and strength? 
\textit{Your compassion}\lebnote{Literally “The agitation of your intestines”} and mercy to me hold themselves back.
\verse For you are our father, although Abraham does not know us, 
and Israel does not acknowledge us. 
You, Adonai are our father, 
Our Redeemer from of old is your name.
\verse Why do you make us wander, Adonai? 
You harden our heart from your ways \textit{so that we do not fear}\lebnote{Literally “from the fear of”} you. 
Turn back for the sake of your servants, 
the tribes of your inheritance.
\verse \textit{Your holy people}\lebnote{Literally “The people of your holiness”} took possession for a little while; 
our adversaries have trampled down your sanctuary.
\verse We have been since antiquity; you did not rule them; 
\textit{they were not called by your name}.\lebnote{Literally “your name was not called over them”}
\end{biblechapter}

\begin{biblechapter} % Isaiah 64
\verse \lebnote{Isaiah 64:1 in the English Bible begins here. In the Hebrew Bible, 63:19 continues} Would that you would tear the heavens and come down; 
the mountains would quake before you,
\verse \lebnote{Isaiah 64:2–12 in the English Bible is 64:1–11 in the Hebrew Bible} as fire kindles brushwood, 
the fire causes water to boil, 
to make your name known to your adversaries, 
that the nations might tremble from your \textit{presence}.\lebnote{Literally “face”}
\verse When you did terrible deeds which we did not \textit{expect},\lebnote{Literally “await”} you came down; 
the mountains quaked because of your \textit{presence}.\lebnote{Literally “face”}
\verse And since ancient times they have not heard, 
have not listened, 
no eye has seen a God except you; 
he \textit{acts}\lebnote{Literally “makes”} for the one who waits for him.
\verse You meet with the one who rejoices, 
one who does righteousness. 
In your ways they remember you. 
Look! You were angry and we sinned against them in ancient times and we were saved.
\verse And we all have become like the unclean, 
and all our deeds of justice like a menstrual cloth, 
And we all wither like a leaf, 
and our iniquities take us away like the wind.
\verse And there is no one who calls on your name, 
who pulls himself up to keep hold of you, 
for you have hidden your face from us,\lebnote{Or “him”} 
and melted us into the hand of our iniquity.
\verse Yet now Adonai, you are our father; 
we are the clay and you are our \textit{potter},\lebnote{Literally “one who fashions”} 
and we all are the work of your hand.
\verse You must not be exceedingly angry, Adonai, 
and you must not remember iniquity forever! Look! Behold, now! We all are your people!
\verse \textit{Your holy cities}\lebnote{Literally “The cities of your holiness”} have become a wilderness; 
Zion has become a wilderness, Jerusalem a desolation.
\verse \textit{Our holy and beautiful temple},\lebnote{Literally “the house of our holiness and our beauty”} where our ancestors\lebnote{Or “fathers”} praised you has \textit{been burned}\lebnote{Literally “been for burning”} by fire, 
and all our precious objects have become ruins.\lebnote{Hebrew “ruin”}
\verse Will you control yourself because of these, Adonai? 
Will you be silent? 
And will you oppress us exceedingly?
\end{biblechapter}

\begin{biblechapter} % Isaiah 65
\verseWithHeading{Judgment and Salvation} “I let myself be sought by those who did not ask; 
I let myself be found by those who did not seek me. 
I said, ‘Here I am; here I am!’ to a nation that did not call on my name;
\verse I spread out my hands all day to a stubborn people, 
those who walk after their thoughts in the way that is not good,
\verse the people \textit{who provoke}\lebnote{Literally “provoking”} me to anger continually to my face, 
slaughtering for sacrifices in the garden, 
and making smoke offerings on bricks,
\verse \textit{who sit}\lebnote{Literally “sitting”} in graves 
and spend the night in \textit{secret places},\lebnote{Literally “watches”} 
\textit{who eat}\lebnote{Literally “eating”} the flesh of swine 
with\lebnote{Or “and”} a fragment\lebnote{The reading tradition (\textit{Qere}) has “broth”} of impurity in their vessels,
\verse \textit{who say},\lebnote{Literally “saying”} “\textit{Keep to yourself}!\lebnote{Literally “Come near to you”} 
You must not come near me, for I am too holy for you!” 
These are a smoke in my \textit{nostrils},\lebnote{Literally “nose”} 
a fire burning all day.
\verse Look! It is written before me: I will not be silent, but I will repay; 
and I will repay in\lebnote{Or “on”} the fold of their garment
\verse your iniquities and the iniquities of your ancestors\lebnote{Or “fathers”} together, says Adonai, 
because they made smoke offerings on the mountains 
and they taunted me on the hills, 
I will measure their punishment from the beginning into the fold of their garment.”
\verse Thus says Adonai: “Just as the new wine is found in the cluster, and they say ‘You must not destroy it, for there is a blessing in it,’ so I will do for the sake of my servants \textit{by}\lebnote{Literally “to”} not destroying \textit{everyone}.\lebnote{Literally “all”}
\verse And I will bring descendant s\lebnote{Hebrew “descendant”} out from Jacob, 
and a \textit{people}\lebnote{Literally “man”} from Judah to take possession of my mountain, 
and my chosen ones shall inherit it, 
and my servants shall settle there.
\verse And Sharon shall become a pasture for\lebnote{Literally “of”} flocks,\lebnote{Hebrew “flock”} 
and the valley of Achor a resting place for\lebnote{Or “of”} herds\lebnote{Hebrew “herd”} 
for my people who have sought me.
\verse But\lebnote{Or “and”} you \textit{who forsake}\lebnote{Literally “forsaking”} Adonai, 
forgetting \textit{my holy mountain},\lebnote{Literally “the mountain of my holiness”} 
who set a table for Fortune, 
and who \textit{pour out}\lebnote{Literally “fill”} a jug of mixed wine for Destiny,
\verse and I will remit you to the sword, 
and all of you shall bow down to the slaughter, 
because I called, but\lebnote{Or “and”} you did not answer; 
I spoke, but\lebnote{Or “and”} you did not listen, 
but\lebnote{Or “and”} you did the evil in my eyes, 
and you chose that in which I do not delight.”
\verse Therefore thus says the Lord Adonai:
\verse “Look! My servants shall eat but\lebnote{Or “and”} you, you shall be hungry. 
Look! My servants shall drink but\lebnote{Or “and”} you, you shall be thirsty. 
Look! My servants shall rejoice but\lebnote{Or “and”} you, you shall be ashamed.
\verse Look! My servants shall shout \textit{for joy},\lebnote{Literally “from goodness of heart”} but\lebnote{Or “and”} you, you shall cry out \textit{for pain}\lebnote{Literally “from pain of heart”} and howl \textit{for sadness}.\lebnote{Literally “from a breaking of the spirit”}
\verse And you shall leave your name to my chosen ones as a curse, 
and the Lord Adonai will kill you, and he will \textit{give}\lebnote{Literally “call”} his servants another name.
\verseWithHeading{New Creation} For look! I am about to create new heavens and a new earth, 
and the former things shall not be remembered, 
and they shall not \textit{come to mind}.\lebnote{Literally “go up to heart”}
\verse But rejoice and shout in exultation forever and ever over what I am about to create! 
For look! I am about to create Jerusalem as a source of rejoicing, 
and her people as a source of joy.
\verse And I will shout in exultation over Jerusalem, 
and I will rejoice over my people, 
and the sound of weeping shall no longer be heard in it, 
or\lebnote{Or “and”} the sound of a cry for help.
\verse There will no longer be a nursing infant \textit{who lives only a few}\lebnote{Literally “of”} days, 
or\lebnote{Or “and”} an old man who does not fill his days, 
for the boy will die \textit{a hundred years old},\lebnote{Literally “a son of a hundred year”} 
and the one who \textit{fails to reach}\lebnote{Literally “misses”} \textit{a hundred years}\lebnote{Literally “a son of a hundred year”} will be \textit{considered}\lebnote{Literally “treated”} accursed.
\verse And they shall build houses and inhabit them, 
and they shall plant vineyards and eat their fruit.
\verse They shall not build and another inhabit; 
they shall not plant and another eat. 
For the days of my people shall be like the days of a tree, 
and my chosen ones shall enjoy the work of their hands.
\verse They shall not labor for nothing, 
and they shall not give birth to horror, 
for they \textit{shall be offspring blessed by}\lebnote{Literally “the offspring of the blessing of”} Adonai, 
and their descendants with them.
\verse And this will happen: before they call, I myself will answer; 
while still they are speaking, I myself will hear.
\verse The wolf and the lamb shall feed like one, 
and the lion shall eat straw like the ox, 
but\lebnote{Or “and”} dust shall be the serpent’s food. 
They shall do no evil, 
and they shall not destroy on all \textit{my holy mountain},”\lebnote{Literally “the mountain of my holiness”} says Adonai.
\end{biblechapter}

\begin{biblechapter} % Isaiah 66
\verseWithHeading{Proper Worship} Thus says Adonai:
\verse “Heaven is my throne, 
and the earth is the footstool for my feet. 
Where is this house that you would build for me? 
And where is this resting place for me? \textit{declares}\lebnote{Literally “declaration of”} Adonai, “but\lebnote{Or “and”} I look to this one:
\verse to the humble and the contrite of spirit 
and the one frightened at my word.
\verse The one who slaughters a bull strikes a man; 
the one who slaughters a lamb for sacrifice breaks the neck of a dog. 
The one who offers an offering, the blood of swine; 
the one who offers frankincense blesses an idol. 
Indeed, they themselves have chosen their ways, 
and their soul delights in their abhorrence.
\verse Hear the word of Adonai, you who are frightened at his word:
\verse Your brothers who hate you, 
who exclude you for my name’s sake have said, 
“Let Adonai be honored so that\lebnote{Or “and”} we may see your joy!” 
But\lebnote{Or “And”} they themselves shall be ashamed.
\verseWithHeading{Zion’s Delivery} Before she was in labor she gave birth; 
before labor pains\lebnote{Hebrew “pain”} came to her, she gave birth to a son.
\verse Who has heard anything like this? 
Who has seen anything like these things? 
Can a land be born in one day? 
Or can a nation be born in a moment? Yet when she was in labor, Zion indeed gave birth to her children.
\verse Shall I myself I break open and not \textit{deliver}?”\lebnote{Literally “cause to bring forth”} says Adonai, 
“or I who \textit{delivers}\lebnote{Literally “causes to bring forth”} lock up the womb?” says your God.
\verse “Rejoice with Jerusalem 
and shout in exultation with her, all those who love her! 
Rejoice with her in joy, all those who mourn over her,
\verse so that you may suck and be satisfied from her \textit{consoling breast},\lebnote{Literally “breast of consolation”} 
so that you may drink deeply and refresh yourselves from her \textit{heavy breast}.”\lebnote{Literally “breast of heaviness “}
\verse For thus says Adonai:
\verse “Look! I am about to spread prosperity out to her like a river, 
and the wealth of the nations like an overflowing stream, 
and you shall suck and be carried on the hip; 
and you shall you shall be played with on the knees.
\verseWithHeading{Adonai’s Final Judgment} And you shall see and your heart shall rejoice, 
and your bones shall \textit{flourish}\lebnote{Literally “sprout”} like the grass, 
and the hand of Adonai shall make itself known \textit{to}\lebnote{Literally “with”} his servants, 
and he shall curse his enemies.
\verse For look! Adonai will come in fire, 
and his chariots like the storm wind, 
to \textit{give back}\lebnote{Literally “cause to turn around”} his anger in wrath, 
and his rebuke in flames of fire.
\verse For Adonai enters into judgment on all flesh with fire and his sword, 
and \textit{those slain by}\lebnote{Literally “the slain of”} Adonai shall be many.
\verse Those who sanctify themselves 
and those who cleanse themselves \textit{to go into}\lebnote{Literally “toward”} the gardens after the one in the middle, 
eating the flesh of swine 
and detestable things\lebnote{Hebrew “thing”} and rodents\lebnote{Hebrew “rodent”} together shall come to an end!” \textit{declares}\lebnote{Literally “declaration of”} Adonai.
\verse “And I—their works and thoughts!—am about to come to gather all nations and tongues, 
and they shall come and see my glory.
\verse And I will set a sign among them, 
and I will send survivors from them to the nations: 
Tarshish, Pul, and Lud, who draw the bow; 
Tubal and Javan, the faraway coastlands 
that have not heard of my fame, 
and have not seen my glory.
\verse And they shall declare my glory among the nations,
\verse and bring all your countrymen\lebnote{Or “brothers”} from all the nations as an offering to Adonai on horses and chariots\lebnote{Hebrew “chariot”} and in litters and on mules and camels, to\lebnote{Or “on”} \textit{my holy mountain},\lebnote{Literally “the mountain of holiness me”} Jerusalem,” says Adonai, “just as the sons of Israel bring an offering in a clean vessel to the house of Adonai.
\verse And indeed, I will take some of them as priests and the Levites,” says Adonai.
\verse And this shall happen: From \textit{new moon to new moon}\lebnote{Literally “enough of new moon in his new moon”} and from \textit{Sabbath to Sabbath}\lebnote{Literally “enough of Sabbath in his Sabbath”} all flesh shall come to bow in worship before me,” says Adonai.
\verse “And they shall go out and look at the corpses of the people\lebnote{Or “men”} \textit{who have rebelled}\lebnote{Literally “rebelling”} against me, for their worm shall not die, 
and their fire shall not be quenched, 
and they shall be an abhorrence to all flesh.”
\end{biblechapter}

