\biblebook{Jeremiah}

\begin{biblechapter} % Jeremiah 1
\verseWithHeading{Adonai Appoints Jeremiah as a Prophet to the Nations} The words of Jeremiah, the son of Hilkiah, among the priests who were in Anathoth, in the land of Benjamin,
\verse to whom the word of Adonai came in the days of Josiah the son of Amon, the king of Judah, in the thirteenth year of his reign.
\verse It came also in the days of Jehoiakim, the son of Josiah, the king of Judah, until the end of the eleventh year of Zedekiah, the son of Josiah, the king of Judah, until the exile of Jerusalem in the fifth month.
\verse And the word of Adonai came to me, \textit{saying},\lebnote{Literally “to say”}
\verse “Before I formed you in the womb I knew you, 
and before you came out from the womb I consecrated you; 
I appointed you as a prophet to the nations.”
\verse Then I said, “Ah, Lord Adonai! Look, I do not know how to speak, for I am a youth.”\lebnote{Or “boy”}
\verse But Adonai said to me, “You must not say, ‘I am a youth,’\lebnote{Or “boy”} for to \textit{whomever}\lebnote{Literally “all of whom”} I send you, you will go, and \textit{whatever}\lebnote{Literally “all of that”} I command you, you will speak.
\verse \textit{Do not be afraid of them},\lebnote{Literally “you must not be afraid from the faces of them”} for I am with you to deliver you,” \textit{declares}\lebnote{Literally “declaration of”} Adonai.
\verse Then Adonai stretched out his hand and he touched my mouth, and Adonai said to me,
\verse “Look, I have put my words in your mouth.
\verseWithHeading{Adonai Confirms Jeremiah’s Call} And the word of Adonai came to me, \textit{saying},\lebnote{Literally “to say”} “What are you seeing, Jeremiah?” And I said, “I am seeing a branch of an almond tree.”
\verse Then Adonai said to me, “You did well to see, for I am watching over my word to perform it.”
\verse And the word of Adonai came to me a second time, \textit{saying},\lebnote{Literally “to say”} “What are you seeing?” And I said, “I am seeing a \textit{boiling}\lebnote{Literally “blown upon”} pot, and its face is from the face of the north.”
\verse Then Adonai said to me, “From the north disaster will be unleashed\lebnote{Or “opened”} on all the inhabitants of the land.
\verse For, look, I am calling to all the clans\lebnote{Or “families”} of the kingdoms of the north,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “and they will come and set each one his throne at the entrance of the gates of Jerusalem, and against all its surrounding walls, and against all the cities of Judah.
\verse And I will pass\lebnote{Or “pronounce”} my judgments on them because of all their wickedness in that they have forsaken me, and they made sacrifices to other gods, and they bowed in worship to the works of their hands.
\verse But you, you must gird your loins, and stand, and speak to them all that I command you. \textit{Do not be afraid of them},\lebnote{Literally “You must not be filled with terror from the faces of them”} or I will shatter you \textit{before}\lebnote{Literally “to the faces of”} them.
\verse Now look, I have made you \textit{today}\lebnote{Literally “the day”} as a fortified city and as an iron pillar and a bronze wall\lebnote{Hebrew “walls”} against all the land, against the kings of Judah, against its princes, against its priests, and against the people of the land.
\verse And they will fight against you but they will not prevail against you, for I am with you,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “to deliver you.”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 2
\verseWithHeading{Adonai Remembers Israel’s Faithfulness} And the word of Adonai came to me, \textit{saying},\lebnote{Literally “to say”}
\verse “Go and proclaim in the ears of Jerusalem, \textit{saying},\lebnote{Literally “to say”} ‘Thus says Adonai,
\verse “I remember concerning you the loyal love of your childhood,\lebnote{Or “youth”} 
the love of your betrothal-time, 
your going\lebnote{Or “walking”} after me in the desert,\lebnote{Or “wilderness”} 
in the land not sown.
\verseWithHeading{The Unfaithfulness of Israel’s Ancestors} Hear the word of Adonai, O house of Jacob, and all the clans\lebnote{Or “families”} of the house of Israel.
\verse Thus says Adonai:
\verse “What injustice did your ancestors\lebnote{Or “fathers”} find in me 
that they were far from me, 
and they went after the vanity, 
and they became vain?
\verse And they did not say, 
‘Where is Adonai, 
the one who brought us up from the land of Egypt, 
the one who led us in the desert,\lebnote{Or “wilderness”} 
in a land of desert plains and gorges,\lebnote{Hebrew “gorge”} 
in a land of dryness and deep darkness, 
in a land \textit{that no one passes through},\lebnote{Literally “no man passes through in it”} 
and no human lives there?’
\verse And I brought you to the land of the orchard, 
to eat its fruit, and its good things.\lebnote{Hebrew “good thing”} 
But you entered and defiled my land, 
and my inheritance you made a detestable thing.\lebnote{Or “abomination”}
\verseWithHeading{The Unfaithfulness of Contemporary Israel} \textit{Therefore}\lebnote{Literally “To thus”} I again bring a lawsuit against you,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“and with \textit{your children}\lebnote{Literally “sons of your sons”} I bring a lawsuit.
\verse For go over to the coasts of Cyprus and look, 
and send to Kedar and consider very closely, 
and see if there has been a thing like this:
\verse Has a nation exchanged gods? And they are not gods! 
But my people have exchanged their glory for that which does not profit.
\verse Be appalled, O heavens, at this, and shudder; 
be utterly desolate,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “For my people have committed two evils: 
they have forsaken me, the source\lebnote{Or “fountain”} of living water, 
to hew out for themselves cisterns, cracked cisterns that can hold no water.
\verse Is Israel a slave? Or \textit{a slave born in a house}?\lebnote{Literally “born of a house”} 
Why has he become plunder?
\verse The young lions have roared against him, 
they have \textit{raised}\lebnote{Literally “given”} their voices.\lebnote{Hebrew “voice”} 
And they have made his land as horror; 
his cities are destroyed, \textit{without}\lebnote{Literally “from not”} an inhabitant.
\verse Moreover, people of Memphis and Tahpanhes 
will \textit{shave}\lebnote{Literally “pasture”} your skull.\lebnote{Or “crown of your head”}
\verse Did you not do this to yourself, 
by forsaking Adonai, your God, 
at the time of your leading in the way?
\verse And now \textit{why do you go}\lebnote{Literally “what for you”} the way of Egypt 
to drink the waters of the Shihor? 
And \textit{why do you go}\lebnote{Literally “what for you”} the way of Assyria 
to drink the waters of the \textit{Euphrates}?\lebnote{Literally “river”}
\verse Your wickedness will chastise you, 
and your apostasies will reprove you. 
Therefore know and see that it is evil and bitter, 
your forsaking of Adonai, your God, 
and fear of me is not in you,” 
\textit{declares}\lebnote{Literally “a declaration of”} the Lord, Adonai of hosts.
\verse “For from long ago you have broken your yoke, 
you tore to pieces your bonds. 
And you said, ‘I will not serve!’ 
But on every high hill and under every leafy tree 
you were lying down as a prostitute.
\verse Yet I planted you as a choice vine, 
all of it a seed of trustworthiness. 
How then have you altered before me 
into something that is degenerate, the foreign vine?
\verse For if you wash with natron, 
and you \textit{use much soap},\lebnote{Literally “make numerous for yourself alkaline salt”} 
your guilt is sticking as a stain \textit{before}\lebnote{Literally “to the face of”} me,” 
\textit{declares}\lebnote{Literally “a declaration of”} the Lord Adonai.
\verse “How can you say, ‘I have not defiled myself, 
I have not gone after the Baals?’ 
Look at your way in the valley, 
know what you have done. 
You are a young she-camel, 
interweaving her ways.
\verse A wild ass accustomed to the desert,\lebnote{Or “wilderness”} 
in the desire of her soul. 
She gasps for wind in her rutting time. 
Who can quell\lebnote{Or “restrain”} her lust? 
All those who seek her will not grow weary, 
in her month they will find her.
\verse Restrain your foot from being barefoot, 
and your throat from thirst. 
But you said, ‘It is hopeless. 
No! For I have loved strangers, 
and after them I will go.’\lebnote{Or “walk”}
\verse As the shame of a thief when he is caught in the act, 
so the house of Israel will be ashamed. 
They, their kings, their officials, 
and their priests, and their prophets.
\verse Those who say to the tree, ‘You are my father,’ 
and to the stone, ‘You gave birth to me.’ 
For they have turned their \textit{backs}\lebnote{Literally “neck”} to me, and not their faces. 
But in the time of their trouble they say, ‘Arise and save us.’
\verse But where are your gods that you made for yourself? 
Let them arise if they can save you in the time of your trouble. 
For according to the number of your towns 
are your gods, O Judah.
\verse Why do you complain to me? 
All of you have \textit{rebelled against me},”\lebnote{Literally “broken with me”} \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “For \textit{in vain}\lebnote{Literally “the vanity”} I have struck your children. 
They accepted no discipline. 
Your sword devoured your prophets, 
like a ravening\lebnote{Or “destroying”} lion.
\verse O generation, you see the word of Adonai. 
Have I been a desert\lebnote{Or “wilderness”} to Israel? Or a land of great darkness? 
Why do my people say, ‘We roam about freely, 
we will not come again to you’?
\verse Can a young woman forget her ornaments?\lebnote{Or “jewelry”} Or a bride her sashes? 
Yet my people have forgotten me days without number.
\verse How well you do on your way to seek love! 
\textit{Therefore}\lebnote{Literally “To thus”} even to the wicked woman you can teach your ways.
\verse Also on your skirts is found 
the blood of the life of the innocent, the poor. 
You did not find them in the act of breaking into a house. 
But in spite of all these things,
\verse yet you say, ‘But I am innocent, 
surely his anger has turned away from me.’ 
Look, I am entering into judgment with you 
because of your saying, ‘I have not sinned.’
\verse How you go about so much changing your way! 
Also by Egypt you will be put to shame, 
just as you were put to shame by Assyria.
\verse Also from there you will come out 
with your hands on your head, 
for Adonai has rejected \textit{those in whom you trust},\lebnote{Literally “in your objects of trust”} 
and you will not prosper through them.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 3
\verse A \textit{saying}:\lebnote{Literally “to say”} ‘Look, if a man divorces his wife, 
and she goes from him \textit{and she becomes another man’s wife},\lebnote{Literally “and she becomes to a man another”} 
will he return to her again?’ 
Will not that land be greatly defiled? 
And you have prostituted yourself with many lovers, 
would you now return to me?” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “Lift up your eyes at the barren heights, 
and see where you have not been ravished. 
Beside the roads you sat for them like an Arab in the desert,\lebnote{Or “wilderness”} 
and you have defiled the land 
with your fornication 
and with your wickedness.
\verse Therefore rain showers have been withheld, 
and the spring rain has not come. 
Yet \textit{you have}\lebnote{Literally “it is for you”} the forehead of a woman prostitute, 
you refuse to be ashamed.
\verse Have you not just now called to me, 
‘My father, you are the close friend of my youth?
\verse Will he be angry \textit{forever}?\lebnote{Literally “to eternity”} 
Will he maintain it \textit{always}’?\lebnote{Literally “to duration”} 
Look, you have spoken, 
but you have done all the evil that you could.”\lebnote{Literally “you have done the evil and you were able”}
\verseWithHeading{Adonai Gives Israel a Letter of Divorce} Then Adonai said to me in the days of Josiah, the king, “Have you seen what apostate\lebnote{Or “unfaithful”} Israel has done? She has gone on every high hill and under every leafy tree and she has prostituted herself there.
\verse And I thought, ‘After her doing all these things to me she will return,’ but she did not return. And her treacherous sister Judah saw it.\lebnote{Here the direct object is supplied from context in the English translation}
\verse And I saw that \textit{for this very reason, that}\lebnote{Literally “because of all”} on account of apostate\lebnote{Or “unfaithful”} Israel committing adultery I divorced her and gave the letter of divorce to her. Yet her treacherous sister Judah was not afraid and she went and prostituted herself also.
\verse And it was because of the frivolity of her fornication that she defiled the land and committed adultery with the stone and with the tree.
\verse Yet even in all this her treacherous sister Judah did not return to me with all her heart, \textit{but only}\lebnote{Literally “but if”} in pretense,”\lebnote{Or “deception”} \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verseWithHeading{A Call for Repentance} Then Adonai said to me, “Apostate\lebnote{Or “unfaithful”} Israel has proved herself more upright\lebnote{Or “righteous”} than treacherous Judah.
\verse Go, and proclaim these words toward the north, and say,
\verse ‘Return, apostate\lebnote{Or “unfaithful”} Israel,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai. 
‘I will not cause my \textit{anger}\lebnote{Literally “faces”} to fall on you. 
For I am loyal,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai. 
‘I will not be angry \textit{forever}.\lebnote{Literally “to eternity”}
\verse “Return, apostate\lebnote{Or “unfaithful”} children,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai. “For \textit{I am your master},\lebnote{Literally “I rule over among you”} and I will take you one from a city and two from a clan,\lebnote{Or “family”} and I will bring you to Zion.
\verse Then I will give you shepherds after my own heart and they will feed you knowledge and insight.
\verse And it will be when you have multiplied and become fruitful in the land in those days,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “they will no longer say, ‘The ark of the covenant of Adonai.’ And \textit{it will not come to mind},\lebnote{Literally “it shall not go up on heart”} nor will they remember it, nor will they miss it,\lebnote{Here the direct object is supplied from context in the English translation} nor will it be made again.
\verse At that time they will call Jerusalem ‘The Throne of Adonai,’ and all the nations will be gathered to it, to the name of Adonai, to Jerusalem, and they will no longer go after the stubbornness of their evil heart.
\verse In those days the house of Judah will walk along with the house of Israel and they will come together from the land of the north to the land that I gave as an inheritance to your ancestors.\lebnote{Or “fathers”}
\verse Then I thought, ‘How I would set you among the children, 
and I would give you a land of desire, 
an inheritance of the glory of the hosts of nations.’ 
And I thought, ‘You would call me, “My father,” 
and you would not turn back from \textit{behind}\lebnote{Or “following”} me.’
\verse However, as a wife departs treacherously from her lover, 
so you have dealt treacherously with me, 
O house of Israel,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “A voice is heard on the barren heights, 
the weeping of the pleas for mercy of the children of Israel, 
because they have perverted their way, 
they have forgotten Adonai their God.
\verse Return, O apostate\lebnote{Or “unfaithful”} children, 
I will heal your backsliding. 
‘Look, we come to you, 
for you are Adonai our God.
\verse Surely, an illusion comes from the hills, 
the turmoil on the mountains. 
Surely, in Adonai our God 
is the salvation of Israel.
\verse But the shameful thing has devoured 
the labor of our ancestors\lebnote{Or “fathers”} from our youth, 
their flocks, and their cattle, 
their sons and their daughters.
\verse Let us lie down in our shame, 
and let our disgrace cover us. 
For against Adonai our God we have sinned, 
we and our ancestors,\lebnote{Or “fathers”} 
from our youth and until this day. 
and we have not obeyed the voice of Adonai our God.’
\end{biblechapter}

\begin{biblechapter} % Jeremiah 4
\verse If you return, O Israel,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“you return to me. 
And if you remove your abominations from my \textit{presence},\lebnote{Literally “faces”} 
and you do not wander,
\verse and you swear, ‘\textit{As Adonai lives},’\lebnote{Literally “life of Adonai”} 
in truth, in justice, and in righteousness, 
then nations will be blessed by him, 
and in him they will boast.”
\verse For thus says Adonai 
to the men\lebnote{Hebrew “man”} of Judah and to Jerusalem, 
“Break up for yourselves prepared virgin soil, 
and you must not sow among thornbushes.
\verse Circumcise yourselves to Adonai, 
and remove the foreskins of your hearts,\lebnote{Hebrew “heart”} 
men\lebnote{Hebrew “man”} of Judah and inhabitants of Jerusalem, 
lest my wrath goes forth like the fire and burns, 
and there is no one who extinguishes it,\lebnote{Here the direct object is supplied from context in the English translation} 
because of the evil of your deeds.”
\verseWithHeading{An Invasion from the North} Declare in Judah and proclaim in Jerusalem and say, 
“Blow a horn through the land, 
\textit{call with a loud voice}\lebnote{Literally “call fill”} and say, 
‘Be gathered and let us go into the fortified cities.’
\verse Lift up a signal toward Zion, 
bring yourselves to safety; you must not remain standing, 
for I am bringing evil from the north, 
and a great destruction.
\verse A lion has gone up from his thicket, 
and a destroyer of nations has set out. 
He has gone out from his place to make your land as a horror, 
your cities will go to ruin \textit{without}\lebnote{Literally “from not”} inhabitant.
\verse Because of this gird yourselves with sackcloth,\lebnote{Hebrew “sackcloths”} lament and wail, 
for \textit{the burning anger}\lebnote{Literally “the blaze of the nose”} of Adonai has not turned back from us.
\verse And then on that day,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“the heart of the king and the heart of the officials will be destroyed, 
and the priests will be appalled, 
and the prophets will be astonished.”
\verse Then I said, “Ah, Lord Adonai, surely you have utterly deceived this people and Jerusalem, saying, ‘It will be peace for you,’ while the sword reaches up to the throat.”
\verse At that time it will be said to this people and to Jerusalem, 
“\textit{A hot wind from the barren heights}\lebnote{Literally “a wind of shimmering of barren heights”} in the desert,\lebnote{Or “wilderness”} 
\textit{in the direction of}\lebnote{Literally “the way of”} the daughter of my people, 
not to winnow and not to cleanse,
\verse \textit{a wind too strong for}\lebnote{Literally “a wind strong more than”} these will come for me, 
now also I speak judgments against them.”
\verse Look, he comes up like the clouds, 
and his chariots are like the whirlwind, 
his horses are swifter than eagles, 
woe to us, for we are devastated.
\verse Cleanse your heart from wickedness, O Jerusalem, 
so that you may be saved. 
\textit{How long}\lebnote{Literally “until when”} will your thoughts of mischief 
dwell in your inner part?
\verse For a voice is declaring from Dan, 
and is proclaiming disaster from the mountain of Ephraim.
\verse “Report to the nations, ‘Here they are!’ 
Proclaim against Jerusalem, 
‘Besiegers are coming \textit{from a distant land},\lebnote{Literally “from a land the distance place”} 
and they raise their voice against the cities of Judah.’
\verse Like watchers of a field they were against her from all around, 
because she has rebelled against me,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “You obtained this for yourself 
because of your way and your deeds. 
This is your doom, for it is bitter, 
yes, it has reached up to your heart.”
\verse My bowels, my bowels! I writhe! 
The walls of my heart! 
My heart is restless within me, I cannot keep silent, 
for I hear in my inner self the sound of a horn, the alarm of war.
\verse Destruction on destruction is proclaimed, 
for all of the land is devastated. 
Suddenly my tents are devastated, 
my tent curtains in a moment.
\verse How long must I see the banner, 
and hear the sound of a horn?
\verse “For my people are foolish, 
they have not known me. 
They are foolish children, 
and they do not have insight. 
They are skillful at doing evil, 
and they do not know how to do good.”
\verse I looked at the earth, and behold, it was wasteland and emptiness, 
and to the heavens, and they were without their light.
\verse I looked at the mountains, and behold, they were quaking, 
and all of the hills were jolted to and fro.
\verse I looked and behold, there was no person, 
and all of the birds\lebnote{Hebrew “bird”} of the sky had fled.
\verse I looked and behold, the fruitful land was a desert,\lebnote{Or “wilderness”} 
and all of its cities were ruined before Adonai, 
before the face of \textit{his burning anger}.\lebnote{Literally “the blaze of his nose”}
\verse For thus says Adonai, “All of the land will be a desolation, 
yet I will not make a complete desolation.
\verse Because of this the earth will mourn, 
and the heavens above will become dark. 
Because I have spoken, I have planned, 
and I have not relented, and I will not turn back from it.”
\verse From the sound of a horseman and \textit{an archer}\lebnote{Literally “one who shoots arrows of a bow”} every town flees, 
they enter in the thickets and among the rocks they climb. 
Every town is forsaken, 
and there is no person who lives in them.
\verse And you, O devastated one, what do you do, 
that you put on crimson, 
that you adorn yourself with ornaments of gold, 
that you make your eyes look bigger with the eye make-up? 
In vain you beautify yourself. 
Your lovers reject you; 
they seek your life.
\verse For I heard a voice like an ill woman, 
anxiety like a woman who bears her first child, 
the voice of the daughter of Zion. 
She is gasping for breath, 
she is spreading out her hands: 
“Woe is me, for I am becoming tired\lebnote{Or “fainting”} before killers.”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 5
\verse “Roam about through the streets of Jerusalem, 
and look please, and take note, 
and search at its public squares, 
if you can find a person who does justice, 
who seeks honesty, 
so that I may forgive it.
\verse And though they say, ‘\textit{As Adonai lives},’\lebnote{Literally “Life of Adonai”} 
\textit{therefore}\lebnote{Literally “to thus”} they swear falsely.”
\verse O Adonai, do not your eyes look for honesty? 
You have struck them, but they felt no pain, 
you have destroyed them, they refused to take discipline. 
They have hardened their faces more than rock, 
they have refused to turn back.
\verse Then I said, “They are just simple people, 
they are foolish, 
for they know not the way of Adonai, 
the law\lebnote{Or “ordinance”} of their God.
\verse Let me go to the great, 
and let me speak with them, 
for they know the way of Adonai, 
the law\lebnote{Or “ordinance”} of their God.” 
However, they together have broken the yoke, 
they have torn to pieces the bonds.
\verse Therefore a lion from the forest will kill them, 
A wolf from the desert plains\lebnote{Or “Arabah”} will devastate them. 
A leopard is watching their cities, 
everyone who goes out from there will be torn, 
because their transgressions are many, 
their apostasies are numerous.
\verse “\textit{How}\lebnote{Literally “Why to this”} can I forgive you? 
Your children have forsaken me, 
and they have sworn by those who are not gods. 
But I fed them to the full, 
and they committed adultery, 
and flocked to the house of a prostitute.
\verse They were well-fed lusty horses, 
they neighed each to the wife of his neighbor.
\verse Because of these things shall I not punish?” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“and on a nation who is like this, shall I not take revenge?
\verse Go up through her vineyards and destroy, 
but you must not make complete destruction. 
Remove her branches, 
for they are not to Adonai.
\verse For the house of Israel and the house of Judah 
have dealt utterly treacherously with me,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “They have denied Adonai, and have said, 
‘Not he, \textit{for}\lebnote{Literally “and”} evil will not come on us, 
we will see neither sword nor famine.
\verse And the prophets are like wind, 
and the word is not in them.’ 
Thus it will be done to them.”
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai, the God of hosts, 
“Because you have spoken this word, 
look, I am making my words in your mouth like a fire, 
and this people wood, and it will devour them.
\verse Look, I am about to bring on you a nation from afar, 
O house of Israel,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai. 
“It is an enduring nation, it is \textit{an ancient nation}.\lebnote{Literally “a nation from a long time ago”} 
A nation whose language you do not know, 
and you cannot understand what they speak.
\verse Their quiver is like an open tomb, 
all of them are warriors.
\verse And they will eat your harvest and your food, 
they will eat your sons and your daughters, 
they will eat your flock and your herd, 
they will eat your vine and your fig tree, 
they will smash with the sword \textit{your fortified cities},\lebnote{Literally “the cities of your fortification”} 
in which you trust.
\verse “But even in those days,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “I will not make you a complete destruction.
\verse \textit{And then},\lebnote{Literally “And it will happen”} when you people say, ‘Why has Adonai our God done all of these things to us?’ Then you will say to them, ‘As you have forsaken me and have served gods of a foreign land in your land, so will you serve strangers in \textit{a land that does not belong to you}.’\lebnote{Literally “a land not to you”}
\verse “Declare this in the house of Jacob, 
and proclaim it in Judah, saying,
\verse ‘Hear this please, O foolish and \textit{senseless}\lebnote{Literally “without heart”} people, 
\textit{who have eyes}\lebnote{Literally “eyes for them”} but do not see, 
\textit{who have ears}\lebnote{Literally “ears for them”} but do not hear.’
\verse Do you not revere me?” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“Do you not tremble \textit{before me}?\lebnote{Literally “from the face of me”} 
For I have placed the sand as a boundary for the sea, 
\textit{as an everlasting limit},\lebnote{Literally “a limit of eternity”} and it cannot pass over it, 
and they rise and fall loudly, but they cannot prevail, 
and its waves roar, but they cannot pass over it.
\verse But for this people is a stubborn and rebellious heart, 
they have turned aside and have gone away.
\verse And they do not say in their hearts, 
‘Let us fear please Adonai our God, 
the one who gives the autumn rain\lebnote{Hebrew “early rain”} and the spring rain\lebnote{Hebrew “late rain”} in its season, 
\textit{the set times of the harvest}\lebnote{Literally “weeks due harvest”} he keeps for us.’
\verse Your iniquities have disturbed these, 
and your sins have kept away the good from you.
\verse For wicked people are found among my people, 
they lie in wait like the hiding of fowlers, 
they set up a trap, 
they catch humans.
\verse Like a cage full of birds,\lebnote{Hebrew “bird”} 
so their houses are full of fraud. 
\textit{Therefore}\lebnote{Literally “Unto thus”} they have become great, 
and they have become rich.
\verse They have grown fat, 
they have grown sleek, 
also, \textit{their evil deeds have no limit}.\lebnote{Literally “they cross over evil deeds”} 
They do not judge with justice, 
the legal cause of the orphan, 
or allow it to succeed, 
and the legal case of the poor, 
they do not defend.
\verse Because of these things shall I not punish?” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“and on a nation who is like this, shall I not take revenge?
\verse A horrific event and something horrible has happened in the land.
\verse The prophets prophesy \textit{falsely},\lebnote{Literally “in the lie”} 
and the priests rule \textit{by their own authority},\lebnote{Literally “by their hands”} 
and my people love it\lebnote{Here the direct object is supplied from context in the English translation} so much. 
But what will you do \textit{when the end comes}?\lebnote{Literally “to end her”}
\end{biblechapter}

\begin{biblechapter} % Jeremiah 6
\verse Flee for safety, O children of Benjamin, 
from the midst of Jerusalem. 
And in Tekoa blow a horn, 
and on Beth-haccherem lift up a signal, 
for evil looks down from the north, 
and a great destruction.
\verse The lovely and the delicate, 
the daughter of Jerusalem, I will destroy.
\verse Shepherds and their flocks will come against her, 
they will pitch their tents against her all around, 
they will pasture, each his \textit{portion}.\lebnote{Literally “hand”}
\verse \textit{Sanctify}\lebnote{Or “Consecrate”} war against her. 
Arise, and let us attack at noon. 
Woe to us, for the day turns, 
for the shadows of evening are lengthened.
\verse Arise, and let us attack by night, 
and let us destroy her citadel fortresses.”
\verse For thus says Adonai of hosts, 
“Cut down trees and heap up a siege ramp against Jerusalem. 
This is the city that must be punished, 
its oppression is in its midst.
\verse As a water well keeps its water cool, 
so she keeps cool her wickedness. 
Violence and destruction are heard within her, 
sickness and wounds\lebnote{Hebrew “wound”} are continually \textit{before me}.\lebnote{Literally “at the face of me”}
\verse Be warned, O Jerusalem, 
lest I turn away from you in disgust, 
lest I make you a desolation, 
a land that is not inhabited.”
\verse Thus says Adonai of hosts: 
“They will thoroughly glean the remnant of Israel as a vine, 
turn back your hand over the branches like a grape-gatherer.
\verse To whom shall I speak and admonish, 
that they may hear? 
Look, their ears are \textit{closed},\lebnote{Literally “uncircumcised”} 
and they are not able to listen attentively. 
Look, the word of Adonai is to them an object of scorn, 
they do not delight in it.
\verse But I am full with the wrath of Adonai, 
I struggle to hold it\lebnote{Here the direct object is supplied from context in the English translation} in. 
Pour it out on the children\lebnote{Hebrew “child”} in the street, 
and on the assemblies\lebnote{Hebrew “assembly”} of young men at the same time. 
For even husband with wife will get trapped, 
the old with him who is full of days.
\verse And their houses will be turned over to others, 
their fields and their wives together. 
For I will stretch out my hand 
against the inhabitants of the land,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “For from the smallest of them to the greatest of them, 
\textit{everyone}\lebnote{Literally “all of him”} makes profit for unlawful gain. 
And from the prophet to the priest, 
\textit{everyone}\lebnote{Literally “all of him”} practices deceit.
\verse And they have treated the wound of my people lightly, 
\textit{saying},\lebnote{Literally “to say”} ‘Peace, peace,’ but there is no peace.
\verse Have they acted ashamed, for they have committed a detestable thing? 
Not at all, nor were they ashamed; they did not know to feel humiliated. 
\textit{Therefore}\lebnote{Literally “To thus”} they will fall among those who fall at the time I punish them, 
they will stumble,” says Adonai.
\verse Thus says Adonai: “Stand at the roads and look, 
and ask for the ancient paths, where the way of the good is, 
and walk in it, and find rest for your inner selves.\lebnote{Hebrew “inner self”} 
But they said, ‘We will not walk in it.’\lebnote{Here the direct object is supplied from context in the English translation}
\verse And I raised up watchmen over you: 
‘Listen attentively to the sound of a horn.’ 
But they said, ‘We will not listen attentively.’
\verse \textit{Therefore}\lebnote{Literally “To thus”} hear, O nations, and know, O community, 
\textit{what will happen to them}.\lebnote{Literally “what among them”}
\verse Hear, O earth, 
look, I am about to bring disaster to this people, 
the fruit of their plans, 
because they have not listened attentively to my words, 
and my teaching, they have rejected it also.
\verse What is this to me? 
Frankincense comes from Sheba, 
and the precious spice reed from a distant land. 
Your burnt offerings are not \textit{acceptable},\lebnote{Literally “for acceptance”} 
your sacrifices are not pleasing to me.”
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai: 
“Look, I am about to put before this people stumbling blocks, 
and they will stumble against them, 
fathers and children together, 
neighbor and his friend will perish.”
\verse Thus says Adonai: 
“Look, a people is coming from the land of the north, 
a great nation is woken up from the farthest part of the earth.
\verse The bow and the short sword they grasp are cruel, 
and they show no mercy. 
Their sound roars like the sea, and on horses they ride, 
drawn up like a man for battle against you, O daughter of Zion.”
\verse We have heard news of it, 
our hands grow slack, 
anxiety has grasped us, 
pain as a woman who gives birth.
\verse You must not go out into the field, 
and you must not walk on the road, 
for \textit{the enemy has a sword},\lebnote{Literally “a sword for enemy”} 
terror is all around.
\verse O daughter of my people, 
gird yourself with sackcloth, 
and roll about in the ashes in mourning, 
a mourning ceremony as for an only child. 
Make for yourself wailing of bitterness, 
for suddenly the destroyer will come upon us.
\verse “I have made you an assayer among my people, a fortification, 
so that you may know and put to the test their ways.\lebnote{Hebrew “way”}
\verse All of them are stubborn rebels, 
going about talking slander. 
They are bronze and iron, 
they are all behaving corruptly.
\verse The bellows\lebnote{Hebrew “bellow”} are burned up, 
lead is consumed by fire. 
\textit{In vain}\lebnote{Literally “To the vanity”} \textit{he keeps on refining},\lebnote{Literally “he refines refining”} 
but the wicked are not torn apart.
\verse \textit{They are called}\lebnote{Literally “They call to them”} rejected silver, 
For Adonai has rejected them.”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 7
\verseWithHeading{Jeremiah’s Proclamation at the Gate of the House of Adonai} The word that came to Jeremiah from Adonai, \textit{saying},\lebnote{Literally “to say”}
\verse “Stand in the gate of the house of Adonai and proclaim there this word, and say, ‘Hear the word of Adonai, all of Judah, those who enter through these gates to bow in worship to Adonai.
\verse Thus says Adonai of hosts, the God of Israel, “Make your ways and your deeds good and let me dwell with you in this place.
\verse Do not trust in \textit{the deceitful words},\lebnote{Literally “the words of the deception”} \textit{saying},\lebnote{Literally “to say”} ‘The temple of Adonai, the temple of Adonai, they are the temple of Adonai.’
\verse For if you truly make your ways and your deeds good, if you truly do justice between a man and his neighbor,
\verse you do not oppress the alien, the orphan, and the widow, you do not shed innocent blood in this place, and you do not go after other gods \textit{to your harm},\lebnote{Literally “to harm to you”}
\verse then I will let you dwell in this place, in the land that I gave to your ancestors\lebnote{Or “fathers”} \textit{forever and ever}.\lebnote{Literally “from eternity and until eternity”}
\verse Look, you are relying for yourselves on \textit{deceitful words}\lebnote{Literally “the words of the deception”} without benefiting.
\verse Will you steal? Will you murder? And will you commit adultery? And will you swear \textit{falsely}?\lebnote{Literally “to the lie”} And will you make a smoke offering to the Baal? And will you go after other gods whom you have not known?
\verse And then you come and you stand \textit{before}\lebnote{Literally “to the face of”} me in this house, which is called \textit{by my name},\lebnote{Literally “my name over it”} and you say, ‘we are safe \textit{to go on doing}\lebnote{Literally “for the sake of the doing of”} all of these detestable things.’
\verse Has this house, which is called \textit{by my name},\lebnote{Literally “my name over it”} become a cave of robbers in your eyes? Look, I, even I, have seen it,” \lebnote{Here the direct object is supplied from context in the English translation} \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “For go now to my place that was in Shiloh, where I caused my name to dwell in the beginning, and see what I did to it \textit{because of}\lebnote{Literally “from face”} the wickedness of my people Israel.
\verse And now, because of your doing all these things,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “and \textit{I have spoken to you over and over again},\lebnote{Literally “I spoke to you doing early and speaking”} and you have not listened, and I called you, and you have not answered,
\verse therefore I will do to the house that is called \textit{by my name},\lebnote{Literally “my name over it”} in which you are trusting, and to the place that I gave to you and to your ancestors\lebnote{Or “fathers”} \textit{what}\lebnote{Literally “as that”} I did to Shiloh.
\verse And I will cast you \textit{out of my sight}\lebnote{Literally “from at my face”} just as I cast out all of your blood relatives,\lebnote{Or “brothers”} all of the offspring of Ephraim.” ’
\verse “And you, you must not pray for this people, and you must not lift up for them a cry of entreaty or a prayer, and you must not plead with me, for I will not hear you.
\verse Do you not see what they are doing in the towns of Judah and in the streets of Jerusalem?
\verse The children are gathering wood, and the fathers are kindling the fire, and the women are kneading dough to make sacrificial cakes for the queen of heaven, and they pour out libations to other gods for the sake of provoking me to anger.
\verse Are they provoking me to anger? ” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “Is it not themselves they hurt, for the sake of the shame of their faces?”
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says the Lord Adonai, “Look, my anger and my wrath will be poured out on this place, on the humankind, and on the animal, and on the tree of the field, and on the fruit of the ground, and it will not be extinguished.”
\verse Thus says Adonai of hosts, the God of Israel, “Add your burnt offerings to your sacrifices, and eat the flesh.
\verse For I did not speak with your ancestors,\lebnote{Or “fathers”} nor did I command them in the day of bringing them out from the land of Egypt, concerning the matter of burnt offering and sacrifice.
\verse But I only commanded them this word, \textit{saying},\lebnote{Literally “to say”} ‘Obey my voice, and I will be to you God, and you will be to me people, and you must walk in all of the way that I command you, so that it goes well with you.’
\verse Yet they did not obey, and they did not incline their ear, but they walked in their own plans, in the stubbornness of their evil heart, and \textit{they became worse instead of better}.\lebnote{Literally “they were to backward and not to face”}
\verse From the day that your ancestors\lebnote{Or “fathers”} came out from the land of Egypt until this day I have sent to you all my servants the prophets, day after day, \textit{sending again and again}.\lebnote{Literally “doing early and sending”}
\verse Yet they have not listened to me and they have not inclined their ear, but they have hardened their neck, they did more evil than their ancestors.\lebnote{Or “fathers”}
\verse So you shall speak to them all these words, but they will not listen to you, and you shall call to them, but they will not answer you.
\verseWithHeading{The Valley of the Slaughter} “And you shall say to them, ‘This is the nation that has not obeyed the voice of Adonai their God, and they have not accepted discipline. The truth has perished, and it is cut off from their mouths.\lebnote{Hebrew “mouth”}
\verse Shear your hair and throw it\lebnote{Here the direct object is supplied from context in the English translation} away, and lift up a lament on the barren heights, for Adonai has rejected and forsaken the generation of his wrath.’
\verse For the people \lebnote{Or “children”} of Judah have done evil in my eyes,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “They have set their abominations in the house that is called \textit{by my name},\lebnote{Literally “my name over it”} to defile it.
\verse And they built the high place of Topheth, which is in the Valley of Ben Hinnom, to burn their sons and their daughters in the fire, which I have not commanded, and it did not come into my mind.\lebnote{Or “heart”}
\verse Therefore look, days are about to come,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “and it will no longer be called Topheth or the Valley of Ben Hinnom, but the Valley of the Slaughter, and they will bury in Topheth \textit{until there is no more place}.\lebnote{Literally “from there is no place”}
\verse And the dead bodies\lebnote{Hebrew “body”} of this people will be as food for the birds\lebnote{Hebrew “bird”} of the heavens,\lebnote{Or “sky”} and for the animals\lebnote{Hebrew “animal”} of the earth, without anyone to scare them\lebnote{Here the direct object is supplied from context in the English translation} away.
\verse And I will cause to disappear from the towns of Judah and from the streets of Jerusalem the sound of jubilation, and the sound of joy, the voice of the bridegroom, and the voice of the bride, for the land will become a site of ruins.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 8
\verse “At that time,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “they shall bring out the bones of the kings of Judah, and the bones of its officials, and the bones of the priests, and the bones of the prophets, and the bones of the inhabitants of Judah from their graves.
\verse And they shall spread them out before the sun, and before the moon, and before all the host of heaven, which they have loved, and which they have served, and which they have gone after, and which they have inquired about, and to which they have bowed in worship. They shall not be gathered, and they shall not be buried; they will be like dung on the surface of the ground.
\verse And death will be preferred to life by all the remaining of this evil clan\lebnote{Or “family”} in all the remaining places where I have driven them away,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verseWithHeading{Adonai Questions His People} “And you shall say to them, ‘Thus asks\lebnote{Or “says”} Adonai:
\verse “Do men fall and not stand up? 
If one turns away does he not return?
\verse Why has this people, Jerusalem, turned away in enduring apostasy? 
They have kept hold of deceit, they have refused to return.\lebnote{Or “repent”}
\verse I have listened attentively, 
and I have listened. 
They do not speak honestly, 
there is no man who regrets his wickedness, saying, ‘What have I done?’ 
All of them are turning to their ways of running, 
like a horse plunging into battle.
\verse Even the stork in heaven knows its appointed time, 
and the turtledove, and the swallow, and the song bird, 
they observe the time of their coming. 
But my people do not know the ordinance\lebnote{Or “law”} of Adonai.
\verse How can you say, ‘We are wise and the law of Adonai is with us’? 
Look, surely \textit{the lying stylus of the scribes}\lebnote{Literally “stylus of lie of scribes”} has made it a lie.
\verse The wise will be put to shame, 
they will be dismayed, 
and they will get taken. 
Look, they have rejected the word of Adonai, 
and what is wisdom to them?
\verse Therefore I will give their wives to others, 
their fields to conquerors, 
because from the smallest to the greatest, 
all of them make profit for unlawful gain, 
from prophet to priest, 
all of them practice deceit.
\verse And they have treated the wound of the daughter of my people lightly, 
\textit{saying},\lebnote{Literally “to say”} ‘Peace, peace,’ but there is no peace.
\verse Were they ashamed, for they had committed a detestable thing? 
What’s worse,\lebnote{Or “also”} they were not ashamed at all, 
and they did not know to feel humiliated. 
\textit{Therefore}\lebnote{Literally “To thus”} they will fall among those who fall; 
at the time of their punishment they will stumble,” says Adonai.
\verseWithHeading{The Nation Longs for Adonai’s Help} Why are we sitting? 
Gather, and let us go into the fortified cities, 
and let us perish there, 
for Adonai our God has caused us to perish, 
and has provided drink for us, water of poison, 
because we have sinned against Adonai.
\verse We hope for peace, 
but there is no good, 
for a time of healing, 
but there is terror.
\verse “From Dan is heard the snorting of their horses, 
from the sound of the neighing of their stallions all the land quakes, 
for they come and devour the land, 
and that which fills the city, 
and those who live in it.
\verse For look, I am letting loose among you snakes, 
adders for which there is no incantation, 
and they will bite you,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “My comfort is gone from me, 
sorrow is on me, 
my heart is sick.
\verse Look, the sound of the cry for help of the daughter of my people 
from a distant land, 
‘Is Adonai not in Zion? 
Is her king not in her?’ 
Why have they provoked me to anger with their images, 
with idols of a foreign land?”
\verse The harvest has passed, 
the summer has come to an end, 
and we have not received help.\lebnote{Or “been saved”}
\verseWithHeading{Adonai’s Lament} “Because of the wound of the daughter of my people I am broken down, 
I mourn, horror has seized me.
\verse Is there no balm in Gilead? 
Is there no healer there? 
Why, then, has the healing of the daughter of my people not been restored?
\end{biblechapter}

\begin{biblechapter} % Jeremiah 9
\verse \lebnote{Jeremiah 9:1–26 in the English Bible is 8:23–9:25 in the Hebrew Bible} \textit{Oh that my head were waters},\lebnote{Literally “Who he gives my head waters”} 
and my eyes a fountain of tears, 
so that I might weep day and night, 
for the slain of the daughter of my people.
\verse \textit{Oh that I had}\lebnote{Literally “Who he gives me”} in the desert a place of overnight lodging for travelers, 
that I may leave my people and go from them, 
for all of them are adulterers, 
a band of traitors.
\verse And they bend their tongue like their bow, 
for falsehood and not truth is superior in the land, 
for they go forth\lebnote{Or “proceed”} from evil to evil, 
and they do not know me,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verseWithHeading{A Nation Characterized by Deceit} “Let everyone be on your guard against his neighbor, 
and you must not trust in any brother, 
for everyone surely betrays, 
and every neighbor goes about with slander.
\verse And everyone deceives his neighbor, 
and they do not speak the truth, 
they have taught their tongues\lebnote{Hebrew “tongue”} to speak lies,\lebnote{Hebrew “lie”} 
they are tired from going astray.
\verse Your dwelling is in the midst of deceit, 
upon\lebnote{Or “through”} deceit they refuse to know me,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse Therefore, thus says Adonai of hosts:
\verse “Look, I am about to refine them, 
and I will test them, 
for what else can I do, 
because of the \textit{presence}\lebnote{Literally “faces”} of the daughter of my people?
\verse Their tongue is a murderous arrow, 
it speaks deceit. 
With his mouth he speaks peace with his neighbor, 
but in his inner parts he sets up his ambush.
\verse Because of these things shall I not punish them?” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“On a nation that is like this shall I not take revenge?
\verse For the mountains I lift up weeping and wailing, 
and for the pastures of the desert a lament, 
because they are laid waste so that no man passes through. 
And the sounds\lebnote{Hebrew “sound”} of cattle are not heard, 
from the birds\lebnote{Hebrew “bird”} of heaven to the animals\lebnote{Hebrew “animal”} 
they have fled, they are gone.
\verse And I will make Jerusalem as heaps of ruins, a lair of jackals, 
and the towns of Judah I will make a desolation, \textit{without inhabitants}.\lebnote{Literally “from no inhabitant”}
\verseWithHeading{The Nation is Judged by Adonai} And Adonai said, “Because of their forsaking my law that I set \textit{before}\lebnote{Literally “to the face of”} them, and they have not obeyed my voice, and have not walked in it,
\verse but they went after the stubbornness of their heart, and after the Baals, which their ancestors\lebnote{Or “fathers”} taught them.”
\verse Therefore thus says Adonai of hosts, the God of Israel, “Look, I am feeding this people wormwood, and providing drink for them, water of poison,
\verse and I will scatter them among the nations that they have not known, they and their ancestors,\lebnote{Or “fathers”} and I will send the sword after them until I bring them to an end.”
\verse Thus says Adonai of hosts:
\verse “Consider closely, and call for the wailing women, so that they come, 
and for the skillful women, so that they come.
\verse And let them hasten, 
and let them lift up wailing over us, 
so that our eyes may melt with tears, 
and our eyelids may flow with water.
\verse For a sound of wailing is heard from Zion, 
‘How we are devastated! 
We are very ashamed because we have left the land, 
because they have overthrown our dwelling places.’
\verse For hear, O women, the word of Adonai, 
and let your ear receive the word of his mouth, 
and teach your daughters a lamentation, 
and each woman her neighbor a lament.
\verse Speak, ‘thus \textit{declares}\lebnote{Literally “a declaration of”} Adonai: “The dead body of the human will fall 
like dung upon the surface of the field, 
and like cut grain behind the reaper, 
and there is no one who gathers.” ’ ”
\verseWithHeading{The Things in Which Adonai Delights} Thus says Adonai,
\verse “The wise man must not boast in his wisdom, 
and the warrior must not boast in his might, 
the wealthy man must not boast in his wealth.
\verseWithHeading{The Way of the Nations} “Look, days are coming,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “and I will call into account all those who are circumcised in the foreskin:
\verse Egypt, Judah, and Edom, and the \textit{Ammonites},\lebnote{Literally “sons/children of Ammon”} and Moab, and all those who are trimmed to the side, those who live in the desert,\lebnote{Or “wilderness”} for all the nations are uncircumcised, and all the house of Israel are uncircumcised of heart.”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 10
\verse Hear the word that Adonai speaks to you, O house of Israel.
\verse Thus says Adonai:
\verse “You must not learn the way of the nations, 
and you must not be dismayed by the signs of the heavens, 
for the nations are dismayed by them.
\verse For the statutes of the peoples are vanity, 
for it is a tree cut down from the forest, 
the work of the hands of a craftsman with the tool.
\verse He decorates it with silver and gold, 
they strengthen it with nails and hammers, 
so that it does not stagger.
\verse They are like scarecrows\lebnote{Hebrew “scarecrow”} in a cucumber field, 
they cannot speak. 
Indeed, they must be carried, 
for they cannot march. 
You must not be afraid of them, 
for they cannot do evil; 
furthermore, to do good is not in them.”
\verse There is none like you, O Adonai, 
you are great and your name is great in might.
\verse Who would not revere you, O king of the nations? 
For you it is fitting. 
For among all the wise men of the nations, 
and in all their kingdoms,\lebnote{Hebrew “kingdom”} 
there is none like you.
\verse \textit{At the same time}\lebnote{Literally “And in one”} they are stupid and foolish, 
in the instruction of idols, it is wood.
\verse Silver beaten from Tarshish is brought, 
and gold from Uphaz, 
the work of the craftsman and the hands of the goldsmith. 
Blue and purple are their clothing, 
all of them are the work of skillful people.
\verse “\textit{Thus}\lebnote{Literally “Like this”} you shall say to them,
\verse ‘Gods who did not make the heavens and the earth 
will perish from the earth and from under these heavens.
\verse He is the maker of the earth by his power, 
who created the world by his wisdom, 
and by his understanding he stretched out heaven.
\verse \textit{When he utters his voice}\lebnote{Literally “At the voice of his giving”} there is a noise of water in the heavens, 
and he causes the mist to rise from the ends\lebnote{Hebrew “end”} of the earth. 
He makes lightning for the rain, 
and he causes the wind to go out from his storehouses.
\verse \textit{Everyone}\lebnote{Literally “All humankind”} is stupid, \textit{without knowledge},\lebnote{Literally “from knowledge”} 
every goldsmith is ashamed by his divine image, 
for his cast image is an illusion, 
and there is no breath in them.
\verse They are vanity, a work of mockery, 
at the time of their punishment, they will perish.
\verse The portion of Jacob is not like these, 
for he is the creator of everything, 
and Israel is the tribe of his inheritance, 
Adonai of hosts is his name.
\verse For thus says Adonai, “Look, I am about to sling out the inhabitants of the land at this time, 
and I will bring distress to them, so that they may feel it.”
\verseWithHeading{The Nation Accepts Its Judgment} Woe to me, because of my wound. 
My wound is incurable. 
But I said, “Surely this is my sickness, 
and I must bear it.”
\verse My tent is devastated, 
and all my tent cords are torn. 
My children have gone out from me, 
and they are not. 
There is no one who pitches my tent again, 
or one who puts up my tent curtains.
\verse For the shepherds have become stupid, 
they do not seek Adonai. 
\textit{Therefore}\lebnote{Literally “Unto thus”} they do not have insight, 
and all of their flock are scattered.
\verse Listen, news:\lebnote{Or “a report”} 
Look, it is coming, 
a great roar from the land of the north, 
to make the cities of Judah a desolation, 
a lair of jackals.
\verse I know, O Adonai, that to the human is not his own way, 
nor to a person is the walking and the directing of his own step.
\verse Chastise\lebnote{Or “Discipline”} me, O Adonai, but in moderation, 
not in your anger, lest you eradicate me.
\verse Pour out your wrath on the nations that do not know you, 
and on the peoples that do not call on your name, 
for they have devoured Jacob, 
they have devoured and consumed him, 
and they have caused his settlement to be desolate.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 11
\verseWithHeading{The Broken Covenant} The word that came to Jeremiah from Adonai, \textit{saying},\lebnote{Literally “to say”}
\verse “Hear the words of this covenant, and speak to the \textit{people}\lebnote{Literally “man”} of Judah and to the inhabitants of Jerusalem,
\verse and say to them, ‘Thus says Adonai, the God of Israel: “The man is cursed who does not obey the words of this covenant,
\verse which I commanded your ancestors\lebnote{Or “fathers”} in the day of my bringing them from the land of Egypt, from the furnace of iron, \textit{saying},\lebnote{Literally “to say”} ‘Listen to my voice and do according to all that I command you, \textit{so will you be my people, and I will be your God}.’\lebnote{Literally “and you shall be to me as people and I will be to you as God”}
\verse in order to keep the oath that I swore to your ancestors,\lebnote{Or “fathers”} to give them a land flowing with milk and honey, as it is this day.” ’ ” Then I answered and said, “\textit{Let it be so},\lebnote{Literally “Amen”} O Adonai.”
\verse And Adonai said to me, “Proclaim all these words in the cities of Judah and in the streets of Jerusalem, \textit{saying},\lebnote{Literally “to say”} ‘Hear the words of this covenant and do them,
\verse for I solemnly admonished your ancestors\lebnote{Or “fathers”} in the day of my leading them out from the land of Egypt and up to this day, \textit{admonishing persistently},\lebnote{Literally “do early and admonishing”} \textit{saying},\lebnote{Literally “to say”} “Obey my voice.”
\verse But they did not obey and they did not incline their ear. Instead each one walked in the stubbornness of his\lebnote{Hebrew “their”} evil heart. Therefore, I brought on them all the words of this covenant which I commanded them\lebnote{Here the direct object is supplied from context in the English translation} to do, and they did not do it.’ ”\lebnote{Here the direct object is supplied from context in the English translation}
\verse And Adonai said to me, “A conspiracy is found among the \textit{people}\lebnote{Literally “man”} of Judah and among the inhabitants of Jerusalem.
\verse They have turned back to the iniquities of their former ancestors,\lebnote{Or “fathers”} who refused to obey my words, and they have gone after other gods, to serve them. The house of Israel and the house of Judah have broken my covenant that I \textit{made}\lebnote{Literally “cut”} with their ancestors.”\lebnote{Or “fathers”}
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai, “I am about to bring on them disaster from which they will not be able to escape. Though they cry out to me, yet I will not listen to them.
\verse Then the cities of Judah and the inhabitants of Jerusalem will go and cry out to the gods to whom they are making smoke offerings, but they will certainly not save them in the time of their trouble.
\verse For the number of your towns are your gods, O Judah, and the number of the streets of Jerusalem are the altars you have set up to the shameful things, to make smoke offerings to Baal.
\verse Therefore you must not pray for this people, and you must not lift up for them a cry of entreaty, or a prayer, for I will not listen in the time of their calling to me, in the time of their trouble.
\verse \textit{What rights has my beloved in my house}\lebnote{Literally “What for my beloved in my house”} 
when she has done many wicked things?\lebnote{Hebrew “wicked thing”} 
Can the flesh of holiness pass over you? 
For \textit{you engage in your wickedness},\lebnote{Literally “your wickedness”} then you exult.
\verse Adonai called your name, ‘A leafy olive tree, \textit{beautiful with fine fruit}.’\lebnote{Literally “beautiful of fruit of appearance”} 
With the sound of a great storm he will set fire to it, 
and its branches will be good for nothing.
\verse And Adonai of hosts, the one who planted you, 
has spoken evil against you, 
because of the evil of the house of Israel and the house of Judah, 
that they have done to themselves, 
to provoke me to anger by making smoke offerings to Baal.”
\verseWithHeading{A Warning for the People of Anathoth} Moreover, Adonai let it be made known to me, and I knew. 
Then you showed me their deeds.
\verse And I was like a gentle ram-lamb that is brought to slaughter, 
and I did not know that they planned plans against me, saying, 
“Let us destroy the tree with its fruit, 
and let us cut him off from the land of the living, 
so that his name will no longer be remembered.”
\verse But O Adonai of hosts, who judges in righteousness, 
who tests the \textit{inmost being},\lebnote{Literally “kidney”} and the \textit{mind},\lebnote{Literally “heart”} 
let me see your retribution upon them, 
for to you I have revealed my legal case.
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai concerning the people\lebnote{Or “men”} of Anathoth who seek your life, \textit{saying},\lebnote{Literally “to say”} “You shall not prophesy in the name of Adonai \textit{or}\lebnote{Literally “and not”} you will die by our hand.”
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai of hosts: “Look, I am about to punish them. The young men will die by the sword. Their sons and their daughters will die by the famine.
\verse And a remnant will not be left for them, for I will bring disaster to the people\lebnote{Or “men”} of Anathoth, the year of their punishment.”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 12
\verseWithHeading{Jeremiah’s Complaint} You will be in the right, O Adonai, when I complain to you. 
Even so, let me speak my claims with you. 
Why does the way of the wicked succeed? 
All those who deal treacherously with treachery are at ease.
\verse Not only do you plant them, they take root. 
They grow, but also they produce fruit. 
You are near in their mouths, 
but far from their \textit{inmost beings}.\lebnote{Literally “kidneys”}
\verse But you, O Adonai, you know me, 
you see me, and you test that my heart is with you. 
Tear them apart like sheep for the slaughtering, 
and set them apart for the day of slaughter.
\verse \textit{How long}\lebnote{Literally “Until when”} will the land mourn, and the vegetation of every field dry up, 
because of the wickedness of those who live in it? 
The animals and the birds\lebnote{Hebrew “bird”} are swept away, 
because they have said, “He does not see our future.”
\verseWithHeading{The Lord’s Reply} “If you run with foot soldiers and they have made you weary, 
then how will you compete with horses? 
If you have fallen in \textit{a peaceful land},\lebnote{Literally “a land of peace”} 
then how will you do in the thickets of the Jordan?
\verse For even your relatives, and the house of your father, 
even they have dealt treacherously with you, 
even they call \textit{loudly}\lebnote{Literally “full”} after you. 
You must not trust in them, though they speak kindly to you.
\verse I have forsaken my house, I have abandoned my inheritance. 
I have given the beloved one of my \textit{heart}\lebnote{Literally “soul”} into the hand of her enemies.
\verse My inheritance has become to me like a lion in the forest. 
She has lifted up against me her voice; \textit{therefore}\lebnote{Literally “unto thus”} I hate her.
\verse The lair of the hyena is my inheritance to me, 
the birds of prey\lebnote{Hebrew “bird of prey”} are all around her. 
Go, gather all of the animals\lebnote{Hebrew “animal”} of the field, 
bring them \textit{to come and devour}.\lebnote{Literally “to food”}
\verse Many shepherds have destroyed my vineyard, 
they have trampled down my plot of land. 
They have made \textit{my pleasant field}\lebnote{Literally “the plot of land of my desire”} 
like the desert\lebnote{Or “wilderness”} of desolation.
\verse They have made it a desolation. 
Desolate, it mourns to me. 
All the land is made desolate, 
but there is no person taking it to heart.
\verse Upon all the barren heights in the desert\lebnote{Or “wilderness”} destroyers have come, 
for the sword of Adonai devours, 
from one end of the land to the other end of the land, 
there is no peace for all flesh.
\verse They have sown wheat and they have reaped thorns. 
They are exhausted, they do not profit. 
And they will be ashamed of your produce, 
because of \textit{the burning anger}\lebnote{Literally “the blaze of the nose”} of Adonai.”
\verse Thus says Adonai concerning all of the evil neighbors who touch the inheritance that I gave as an inheritance to my people Israel: “Look, I am about to drive them out from their land, and the house of Judah I will drive out from among them.
\verse \textit{And then}\lebnote{Literally “And it will happen”} after my driving them out, I will once again have compassion on them, and I will bring them back, each one to his inheritance, and each one to his land.
\verse \textit{And then},\lebnote{Literally “And it will happen”} if they diligently learn the ways of my people, to swear by my name, ‘\textit{as Adonai lives},’\lebnote{Literally “the life of Adonai”} even as they taught my people to swear by Baal, then they will live in the midst of my people.
\verse But if they do not listen, then \textit{I will completely uproot and destroy that nation}.”\lebnote{Literally “I will drive out the nation that driving out and destroying”} \textit{declares}\lebnote{Literally “declaration of”} Adonai.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 13
\verseWithHeading{The Linen Loincloth} Thus Adonai said to me: “Go and buy for yourself a loincloth of linen and put it on your loins, but you must not \textit{place}\lebnote{Literally “bring”} it in water.”
\verse So I bought the loincloth according to the word of Adonai and I put it\lebnote{Here the direct object is supplied from context in the English translation} on my loins.
\verse Now the word of Adonai came to me a second time, \textit{saying},\lebnote{Literally “to say”}
\verse “Take the loincloth that you bought, that is on your loins, and stand up, go to the Euphrates and hide it there in the cleft of the rock.”
\verse So I went and hid it by the Euphrates just as Adonai had commanded me.
\verse \textit{And then},\lebnote{Literally “And it happened”} \textit{after a long time},\lebnote{Literally “from the end of many days”} then Adonai said to me, “Stand up, go to the Euphrates and take from there the loincloth that I commanded you to hide there.”
\verse Then I went to the Euphrates and I dug, and I took the loincloth from the place where I had hidden it, and look, the loincloth was ruined; it was not good for anything.
\verse \textit{And then}\lebnote{Literally “And it happened”} the word of Adonai came to me, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai: ‘So I will ruin the pride of Judah and the great pride of Jerusalem.
\verse This evil people, who refuse to listen to my words, who go\lebnote{Or “walk”} in the stubbornness of their hearts,\lebnote{Hebrew “heart”} and have gone after other gods, to serve them, and to bow in worship to them, let them be like this loincloth which is not good for anything.
\verse For as the loincloth clings to the loins of a person, so I caused to cling to me all the house of Israel and all the house of Judah,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘to be for me a people, and a name, and a praise, and a glory, but they would not listen.’
\verse “Therefore, you shall say to them this word, ‘Thus says Adonai, the God of Israel: “Every jar must be filled with wine.” ’ And they will say to you, ‘Do we not certainly know that every jar should be filled with wine?’
\verse Then you shall say to them, ‘Thus says Adonai: “Look, I am about to fill all the inhabitants of this land, and the kings who sit on David’s throne, and the priests, and the prophets, and all the inhabitants of Jerusalem with drunkenness.
\verse And I will smash them, each one against his brother, and the fathers and the sons together,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai. “I will not have compassion, and I will not take pity, and I will not show compassion from destroying them.” ’ ”
\verse Listen, and pay attention, 
you must not be haughty, for Adonai has spoken.
\verse Give glory to Adonai your God \textit{before}\lebnote{Literally “at not yet”} he brings darkness, 
and \textit{before}\lebnote{Literally “at not yet”} your feet stumble on the mountains at twilight. 
And you will hope for light, but he makes it as gloom, 
he makes it\lebnote{Here the direct object is supplied from context in the English translation} as thick darkness.
\verse But if you will not listen to it, 
my inner self\lebnote{Or “soul”} will weep in secret places, 
because of \textit{the presence of}\lebnote{Literally “the face of”} your pride. 
And my eyes\lebnote{Hebrew “eye”} will weep bitterly, 
and they will melt in tears, 
because the flock of Adonai has been taken captive.
\verseWithHeading{A Message for the King and the Queen Mother} “Say to the king and to the queen mother, ‘\textit{Take a lower seat},\lebnote{Literally “Bring low sit”} 
for \textit{your beautiful crown}\lebnote{Literally “the crown of your beauty”} has come down from your head.’\lebnote{Or “at the head of you”}
\verse The towns of the Negev are shut up, 
and there is no one who opens them.\lebnote{Here the direct object is supplied from context in the English translation} 
All of Judah is deported; 
it is deported in completeness.
\verse Lift up your eyes 
and see those coming from the north. 
Where is the flock that was given to you, 
\textit{your magnificent flock}?\lebnote{Literally “the flock of your splendor”}
\verse What will you say when he sets over you— 
and you have taught them—allies as head over you? 
Will not labor pains take hold of you, 
like a woman giving birth?
\verse And if you say in your heart, 
‘Why have these things happened to me?’ 
Because of the greatness of your iniquity your skirts are uncovered; 
your heels have endured violence.
\verse Can a Cushite change his skin, 
or a leopard his spots? 
Then you also will be able to do good, 
who are accustomed to doing evil.
\verse And I will scatter them 
like stubble scattered by the wind of the desert.\lebnote{Or “wilderness”}
\verse This is your lot, 
the portion of your measure from me,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“because you have forgotten me, 
and you have trusted in the lie.
\verse So in turn I will bare your skirts over your face, 
and your shame will become visible.
\verse Your adulteries, and your neighings, 
the shameful behavior of your fornications 
on the hills in the countryside, 
I have seen your abominations. 
Woe to you, O Jerusalem! 
\textit{How long will you continue to be unclean}?\lebnote{Literally “Not you are clean after how long still”}”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 14
\verseWithHeading{The Severe Drought} That which came as the word of Adonai to Jeremiah concerning the matter\lebnote{Hebrew “matters”} of the severe drought.
\verse “Judah mourns, and her gates languish, 
they are in mourning on the ground, 
and the cry of lament of Jerusalem goes up.
\verse And their nobles send their servants for water. 
They come to the pits, they find no water. 
They return with their vessels \textit{empty}.\lebnote{Literally “without success”} 
They are ashamed, and they are confounded, 
and they cover their heads.\lebnote{Hebrew “head”}
\verse Because of the ground, which is cracked 
because there was no rain on the ground. 
The farmers are ashamed, 
they cover their heads.\lebnote{Hebrew “head”}
\verse For even the doe in the field gives birth and forsakes her young,\lebnote{Here the direct object is supplied from context in the English translation} 
because there is no green growth.
\verse And wild asses stand on the barren heights, 
they gasp for breath like the jackals. 
Their eyes fail 
because there is no vegetation.”
\verse Although our iniquities testify against us, O Adonai, 
act for the sake of your name. 
Yes, our apostasies are many; 
we have sinned against you.
\verse You are the hope of Israel, 
its savior in time of distress. 
Why should you be like an alien in the land, 
or like a traveler who spreads out his tent \lebnote{Here the direct object is supplied from context in the English translation} to spend the night?
\verse Why should you be like a confused man, 
like a warrior who is not able to help? 
Yet you are in our midst, O Adonai, 
and your name is called upon us. 
You must not leave us!
\verse Thus says Adonai concerning this people: “They have loved so much to wander, 
they have not spared their feet. 
Therefore Adonai is not favorable to them, 
now he will remember their iniquity and punish their sins.”
\verseWithHeading{Lying Prophets} So Adonai said to me, “You must not pray for this people, for their happiness.\lebnote{Or “good”}
\verse Although they fast, I am not listening to their cry of entreaty. And although they present burnt offering and grain offering, I am not favorable to them. But by the sword, and by the famine, and by the plague, I am about to destroy them.”
\verse Then I said, “Ah, Lord Adonai, look, the prophets are saying to them, ‘You will not see the sword, and famine will not be yours, but reliable peace I will give to you in this place.’ ”
\verse And Adonai said to me, “The prophets are prophesying lies\lebnote{Hebrew “lie”} in my name. I have not sent them, and I have not commanded them, and I have not spoken to them. They are prophesying to you a lying vision, and a divination, worthless, and the deceitfulness of their \textit{minds}.”\lebnote{Literally “heart”}
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai concerning the prophets who prophecy in my name though I have not sent them, yet they are saying, “Sword and famine will not be in this land.” “By the sword and by the famine those prophets will perish.
\verse And the people to whom they are prophesying will be cast out into the streets of Jerusalem \textit{because of}\lebnote{Literally “from the face of”} the famine and the sword. And there will be no one who buries them, their wives, or their sons, or their daughters. And I will pour out on them their wickedness.
\verseWithHeading{The Nation Hopes in Adonai} “And you shall say to them this word,
\verse ‘Let my eyes melt with tears night and day, 
and let them not cease, 
for with a great wound the virgin daughter of my people is broken, 
with a very incurable wound.
\verse If I go out into the field, 
then look, those slain by the sword. 
And if I enter the city, 
then look, the sickness of the famine. 
For both prophet and priest wander around through the land, 
\textit{they do not know what is going on}.’ ”\lebnote{Literally “and not they know”}
\verse Have you completely rejected Judah? 
Or has your soul\lebnote{Or “inner self”} loathed Zion? 
Why have you struck us 
so that there is no healing for us? 
We hope for peace, but there is no good, 
and for a time of healing, but look, there is terror.
\verse We know, O Adonai, our wickedness, 
the iniquity of our ancestors,\lebnote{Or “fathers”} 
for we have sinned against you.
\verse You must not spurn us\lebnote{Here the direct object is supplied from context in the English translation} for the sake of your name, 
you must not dishonor \textit{your glorious throne}.\lebnote{Literally “the throne of your glory”} 
Remember us!\lebnote{Here the direct object is supplied from context in the English translation} 
You must not break your covenant with us.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 15
\verseWithHeading{Four Types of Calamity} Then Adonai said to me, “Even if Moses and Samuel stood \textit{before me},\lebnote{Literally “to the face of me”} my heart would not be to this people. Send them away from my \textit{sight},\lebnote{Literally “face”} and let them go out.
\verse \textit{And then}\lebnote{Literally “And it will happen”} if they say to you, ‘Where shall we go?’ Then you shall say to them, ‘Thus says Adonai, “Those who are destined for the plague, to the plague, 
and those who are destined for the sword, to the sword, 
and those who are destined for the famine, to the famine, 
and those who are destined for the captivity, to the captivity.” ’
\verse “And I will summon\lebnote{Or “appoint”} over them four types of calamity,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “the sword to kill, and the dogs to drag away, and the birds\lebnote{Hebrew “bird”} of the air and the wild animals\lebnote{Hebrew “wild animal”} of the earth to devour and to destroy.
\verse And I will make them a terror to all of the kingdoms of the earth because of Manasseh, the son of Hezekiah, the king of Judah, because of what he did in Jerusalem.
\verse “For who will have compassion on you, O Jerusalem, 
or who will show sympathy to you, 
or who will turn aside \textit{to ask about your welfare}?\lebnote{Literally “to ask to the welfare of you”}
\verse You have forsaken me,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai. 
“You go backward, so I have stretched my hand against you, 
and I have destroyed you. 
I am tired of becoming remorseful.
\verse And I have winnowed them with a winnowing fork in the gates of the land. 
I am bereaved. 
I have destroyed my people. 
They did not turn back from their ways.
\verse Their widows\lebnote{Hebrew “widow”} will be more numerous before me than the sand of the seas. 
I have brought to them, against the mothers\lebnote{Hebrew “mother”} of young men,\lebnote{Hebrew “man”} 
a destroyer at noon. 
I will suddenly drop on her anguish and horror.\lebnote{Hebrew “horrors”}
\verse She who gave birth to seven has withered away. 
She gasps her breath. 
Her sun \textit{set}\lebnote{Literally “came”} while still day. 
She has been put to shame and disgraced. 
And the remainder of them I will give to the sword 
\textit{before}\lebnote{Literally “to the face of”} their enemies,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verseWithHeading{Jeremiah’s Lament and Adonai’s Response} Woe to me, O my mother, that you gave birth to me, 
a man of contention and quarreling \textit{to the whole land}.\lebnote{Literally “to all of the land”} 
I have not lent out and \textit{I have not borrowed}.\lebnote{Literally “not they have lent out for me”} 
All of them are cursing me.
\verse Adonai said,
\verse “\textit{Surely}\lebnote{Literally “If not”} I will set you free \textit{to be good}\lebnote{Literally “for good”} for you, 
\textit{surely}\lebnote{Literally “if not”} I will cause the enemy to intercede with you 
in a time of trouble, 
and in a time of distress.
\verse Can anyone break iron, 
iron from the north, or bronze?
\verse I will give your wealth and your treasures 
as plunder without price, 
even because of all your sins, 
and throughout all your territories.
\verse And I will make you serve your enemies 
in a land that you do not know, 
for a fire will kindle in my anger. 
Upon you it will be kindled.”
\verse You who know, O Adonai, remember me, and attend to me, 
and take revenge for me against my persecutors \textit{in your forbearance}.\lebnote{Literally “in the length of your nose”} 
You must take me away. 
Know that I am carrying disgrace because of you.
\verse Your words were found and I ate them, 
and your words became to me as jubilation, 
even as the joy of my heart, 
for your name is called on me, O Adonai of hosts.
\verse I sat not in the assembly of jokers, 
nor did I exult. 
Because of the \textit{presence}\lebnote{Literally “face”} of your hand I sat alone, 
for you filled me with indignation.
\verse Therefore thus says Adonai,
\verse “If you turn back, then I will take you back. 
You will stand \textit{before me}.\lebnote{Literally “to the face of me”} 
And if you utter what is precious and not what is worthless, 
you will be as my mouth. 
They will turn back to you, 
and you will not turn back to them.
\verse And I will make you to this people 
a wall of fortified bronze, 
and they will fight against you, 
but they will not prevail over you, 
for I am with you, to save you, 
and to deliver you,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 16
\verseWithHeading{Jeremiah is Commanded Not to Marry} And the word of Adonai came to me, \textit{saying},\lebnote{Literally “to say”}
\verse “You shall not take for yourself a wife, there shall not be for you sons and daughters in this place.”
\verse For thus says Adonai concerning the sons and concerning the daughters born in this place, and concerning their mothers who gave birth to them, and concerning their fathers who fathered them in this land:
\verse “They will die by diseases, they will not be mourned, and they will not be buried. They will be as dung on the face of the earth. And they will perish by the sword and they will perish by the famine. And their dead bodies\lebnote{Hebrew “dead body”} will become food for the birds\lebnote{Hebrew “bird”} of the air, and for the wild animals\lebnote{Hebrew “wild animal”} of the earth.”
\verse For thus says Adonai, “You must not enter the house of the funeral meal, and you must not go to lament, and you must not show sympathy for them, for I have taken away my peace from this people,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “the loyal love and the compassion.
\verse Both the great and small will die in this land. They will not be buried, and they will not mourn for them, and no one will cut himself with a blade, and no one will shave himself for them.
\verse And they will not break bread for the mourning ceremony for them, to comfort him because of the dead. And they will not give drink to them, the cup of desolation, for their fathers,\lebnote{Hebrew “father”} or for their mothers.\lebnote{Hebrew “mother”}
\verse And you shall not enter the house of the banquet, to sit with them, to eat and drink.”
\verse For thus says Adonai of hosts, the God of Israel, “Look, I am about to cause to disappear from this place, before your eyes and in your days, the voice of jubilation, and the voice of joy, and the voice of the bridegroom, and the voice of the bride.
\verse \textit{And then}\lebnote{Literally “And it will happen”} when you tell this people all these words and they ask you, ‘\textit{Why}\lebnote{Literally “On what”} has Adonai pronounced against us all this great evil, and what is our iniquity, and what is our sin that we have sinned against Adonai our God?’
\verse Then you shall say to them, ‘\textit{Because}\lebnote{Literally “On that”} your ancestors\lebnote{Or “fathers”} have forsaken me,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘and they have gone after other gods, and they have served them, and they have bowed down in worship to them, and they have forsaken me, and they have not kept my laws.
\verse And \textit{you have behaved badly}\lebnote{Literally “you have done evil to do”} more than your ancestors,\lebnote{Or “fathers”} and look, you are going each one after the stubbornness of his evil heart, without listening to me.
\verse So I will throw you from this land to the land that you have not known, you or your ancestors,\lebnote{Or “fathers”} and there you will serve other gods day and night, for I will not show you favor.
\verseWithHeading{The Nation’s Return to the Land} \textit{Therefore}\lebnote{Literally “To thus”} look, days are coming,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘and it will no longer be said, “\textit{As Adonai lives},\lebnote{Literally “The life of Adonai”} who led up the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} from the land of Egypt,”
\verse \textit{but only}\lebnote{Literally “but if”} “\textit{As Adonai lives},\lebnote{Literally “The life of Adonai”} who led up the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} from the land of the north, and from all the lands where he had driven them,” for I will bring them back to their ground that I gave to their ancestors.\lebnote{Or “fathers”}
\verse Look, I am sending for many fishermen,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘and they will fish them out, and \textit{afterward}\lebnote{Literally “after thus”} I will send for many hunters, and they will hunt them from every mountain, and from every hill, and from the clefts of the cliffs.
\verse For my eyes are on all their ways, they are not hidden \textit{from before me},\lebnote{Literally “from to the face of me”} and their iniquity is not concealed from before my eyes.
\verse And \textit{I will first doubly repay}\lebnote{Literally “I will reward first second”} their iniquity and their sin because of their profaning my land with the dead bodies\lebnote{Hebrew “dead body”} of their abhorrences, and with their abominations they have filled up my inheritance.’ ”
\verse O Adonai, my strength, and my stronghold, 
and my refuge in the day of distress, 
to you the nations will come from the ends of the earth, 
and they will say, “Our ancestors\lebnote{Or “fathers”} have inherited only lies,\lebnote{Hebrew “lie”} 
vanity, and there is no profit in them.
\verse Can a human make for himself gods? 
Yet they are not gods!”
\verse “\textit{Therefore}\lebnote{Literally “To thus”} look, I am about to let them know, 
this time I am going to let them know 
my \textit{power}\lebnote{Literally “hand”} and my might, 
and they will know that my name is Adonai.”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 17
\verseWithHeading{The Sin of Judah} “The sin of Judah is written with an iron stylus, 
with a point of diamond it is engraved 
on the tablet of their heart, 
and on the horns of their altars.
\verse As the remembering of their children, 
so is the remembering of their altars 
and their poles of Asherah worship 
beside the leafy trees\lebnote{Hebrew “tree”} on the high hills.
\verse O my mountain in the open country, 
I will give your wealth, all your treasures, for spoil, 
your high places, because of your sin, 
throughout all your territories.
\verse And \textit{you will let go your hand from your inheritance}\lebnote{Literally “you shall loose and in you from your inheritance”} 
that I gave to you, 
and I will make you serve your enemies 
in the land that you do not know, 
for you kindled a fire in my anger, 
until eternity it will be kindled.”
\verseWithHeading{The One Who Trusts in Humankind Contrasted With the One Who Trusts in Adonai} Thus says Adonai,
\verse “Cursed is the person who trusts in humankind 
and makes flesh his \textit{strength},\lebnote{Literally “arm”} 
and turns aside his heart from Adonai.
\verse And he will be like a juniper in the wilderness,\lebnote{Or “Arabah”} 
and he will not see when good comes, 
and he will dwell in the parched places in the desert,\lebnote{Or “wilderness”} 
in a land of salt flats,\lebnote{Hebrew “salt flat”} where no one lives.
\verse Blessed be the person who trusts in Adonai, 
and Adonai is his trust.
\verse For he will be like a tree planted by water, 
and to the stream it sends its roots, 
and it will not fear when heat comes, 
and its leaves\lebnote{Hebrew “leaf”} will be luxuriant, 
and in the year of drought it will not be anxious, 
and it will not cease from the bearing of fruit.
\verse The heart is deceitful more than anything else, 
and it is disastrous. Who can understand it?
\verse I, Adonai, examine the \textit{mind},\lebnote{Literally “heart”} 
I test the \textit{heart},\lebnote{Literally “kidneys”} 
and give to each one according to his way, 
according to the fruit of his deeds.
\verseWithHeading{The Fountain of Living Water} A lofty throne of glory from the beginning 
is the place of our sanctuary.
\verse O Adonai, the hope of Israel, 
all those who forsake you will be put to shame, 
and those who turn aside from you in the earth will be recorded, 
for they have forsaken the fountain of living water, Adonai.
\verse Heal me, O Adonai, and I will be healed, 
save me, and I will be saved, 
for you are my praise.
\verse Look, they are saying to me, 
“Where is the word of Adonai? 
Please let it come.”
\verse But I, I have not run away from being a shepherd \textit{who follows you},\lebnote{Literally “after you”} 
and I have not desired the disastrous day. 
You, you know the pronouncement of my lips, 
it was before your face.
\verse You must not become to me a terror, 
you are my refuge in the day of disaster.
\verse Let my persecutors be put to shame, 
\textit{but as for me},\lebnote{Literally “I”} do not let me be put to shame. 
Let them be dismayed, 
\textit{but as for me},\lebnote{Literally “I”} do not let me be dismayed. 
Bring on them the day of disaster, 
and with double destruction, destroy them.
\verseWithHeading{The Sabbath Day} Thus said Adonai to me, “Go, and you must stand in the \textit{People’s Gate}\lebnote{Literally “gate of the sons/children of the people”} through which the kings of Judah enter, and through which they go out, and in all the gates of Jerusalem.
\verse And you must say to them, ‘Hear the word of Adonai, kings of Judah, and all Judah, and all the inhabitants of Jerusalem who enter through these gates.
\verse Thus says Adonai, “Be on your guard for the sake of yourselves, that you must not carry a burden on the day of the Sabbath, and you must not bring it\lebnote{Here the direct object is supplied from context in the English translation} through the gates of Jerusalem.
\verse And you must not carry a burden from your houses on the day of the Sabbath, and you must not do any work. But you must declare holy the day of Sabbath, just as I commanded your ancestors.\lebnote{Or “fathers”}
\verse Yet they did not listen, and they did not incline their ear, and they hardened their neck so as to not hear, nor to receive discipline.
\verse “And it will happen, if you carefully listen to me,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “to bring no burden through the gates of this city on the day of the Sabbath, but to declare holy the day of Sabbath, not to do on it any work,
\verse then kings and officials who sit on the throne of David will enter through the gates of this city riding in chariots\lebnote{Hebrew “chariot”} and on horses, they, and their officials, the people of Judah, and the inhabitants of Jerusalem, and this city will be inhabited \textit{forever}.\lebnote{Literally “to eternity”}
\verse And they will come from the towns of Judah, and from all around Jerusalem, and from the land of Benjamin, and from the Shephelah, and from the hill country, and from the Negev, bringing burnt offerings,\lebnote{Hebrew “burnt offering”} and sacrifices,\lebnote{Hebrew “sacrifice”} and grain offerings,\lebnote{Hebrew “grain offering”} and frankincense, and bringing thank offerings\lebnote{Hebrew “thank offering”} to the house of Adonai.
\verse But if you do not listen to me, to declare holy the day of the Sabbath, and to not carry a burden and enter through the gates of Jerusalem on the day of the Sabbath, then I will kindle a fire in its gates, and it will devour the palaces of Jerusalem, and it will not be quenched.” ’ ”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 18
\verseWithHeading{The Potter and the Clay} The word that came to Jeremiah from Adonai, \textit{saying},\lebnote{Literally “to say”}
\verse “Stand up and go down to the house of the potter, and there I will let you hear my words.”
\verse So I went down to the house of the potter and look, there he was doing work at the potter’s wheels.
\verse And the vessel that he was making with the clay was spoiled in the hand of the potter, so \textit{he made again}\lebnote{Literally “he turned back and he made it”} another vessel, as it was right in the eyes of the potter to make.
\verse And the word of Adonai came to me, \textit{saying},\lebnote{Literally “to say”}
\verse “Like this potter, am I not able to do to you, O house of Israel?” \textit{declares}\lebnote{Literally “a declaration of”} Adonai. “Look, like the clay in the hand of the potter, so are you in my hand, O house of Israel.
\verse One moment I speak concerning a nation or concerning a kingdom, to uproot, and to tear down, and to destroy it.\lebnote{Here the direct object is supplied from context in the English translation}
\verse But if that nation turns back from its evil that I have threatened against it, then I will relent concerning the disaster that I planned to do to it.
\verse And the next moment I speak concerning a nation or concerning a kingdom, to build, and to plant it.\lebnote{Here the direct object is supplied from context in the English translation}
\verse But if it does evil in my sight, to not listen to my voice, then I will relent concerning the good that I said I would do to it.
\verse “So now then, say, please, to the \textit{people}\lebnote{Literally “man”} of Judah and to the inhabitants of Jerusalem, \textit{saying},\lebnote{Literally “to say”} ‘Thus says Adonai, “Look, I am preparing evil against you, and I am planning a plan against you. Please turn back, each one from his evil way, and walk rightly in your ways and your deeds.” ’
\verse But they will say, ‘It is hopeless, for we will go after our own plans, and each one of us will act according to the stubbornness of his evil heart.’
\verse “Therefore thus says Adonai,
\verse ‘Please ask among the nations, “Who has heard \textit{the like of this}?”\lebnote{Literally “like these”} 
The virgin of Israel has done something very horrible.
\verse Does the snow of Lebanon leave from the crags\lebnote{Hebrew “crag”} of Sirion? 
Or are the cold waters flowing from distant mountains dried up?
\verse But my people have forgotten me. 
They make smoke offerings to the idols,\lebnote{Hebrew “idol”} 
and they caused them to stumble in their ways, the ancient ways, 
to go into \textit{bypaths},\lebnote{Literally “paths of way”} not one that is built up.
\verse To make their land a horror, 
an object of whistling for eternity. 
All who pass by it will be appalled, 
and he will shake his head.
\verseWithHeading{Jeremiah’s Imprecatory Prayer} Then they said, “Come and let us plan plans against Jeremiah, for instruction will not be lost from the priest, nor advice from the wise man, nor the word from the prophet. Come and \textit{let us bring charges against him},\lebnote{Literally “let us strike him with the tongue”} and let us not listen attentively to any of his words.”
\verse Listen attentively to me, O Adonai, 
and listen to the voice of my opponents.
\verse Should good be repaid in place of evil? 
Yet they have dug a pit for me. 
Remember my standing \textit{before you}\lebnote{Literally “to the face of you”} to speak good for them, 
to avert your wrath from them.
\verse \textit{Therefore}\lebnote{Literally “To thus”} give their children to the famine, 
and hand them over to the \textit{power}\lebnote{Literally “hand”} of the sword, 
and let their wives be bereaved and widows, 
and let their men be killed by death, 
their young men struck dead 
by the sword in the battle.
\verse Let a cry for help be heard from their houses 
when you suddenly bring upon them the raiding band, 
for they have dug a pit to catch me, 
and they have fixed secretly a trap for my feet.
\verse But you, O Adonai, you know 
\textit{all their plans of assassination against me}.\lebnote{Literally “all of their plan against me to the death”} 
You must not make atonement for their iniquity, 
and you must not cause their sin to be blotted out from \textit{before you}.\lebnote{Literally “to the face of you”} 
But let them be overthrown \textit{before you}.\lebnote{Literally “to the face of you”} 
Deal with them in the time of your anger.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 19
\verseWithHeading{The Valley of the Slaughter} Thus said Adonai, “Go and buy a potter’s earthenware jar, and take some of the elders of the people, and some of the leaders of the priests,
\verse and go out to the Valley of Ben Hinnom, which is at the entrance of the Gate of the Potsherd, and proclaim there the words that I speak to you.
\verse And you shall say, ‘Hear the word of Adonai, O kings of Judah and inhabitants of Jerusalem. Thus says Adonai of hosts, the God of Israel: “Look, I am about to bring disaster upon this place so that \textit{everyone who hears it},\lebnote{Literally “all hears it”} his ears will ring.
\verse Because they have forsaken me, and they have defaced this place, and they have made smoke offerings in it to other gods whom they have not known, they, nor their ancestors,\lebnote{Or “fathers”} nor the kings of Judah, and they have filled up this place with the blood of the innocent,
\verse and they have built the high places of Baal, to burn their children in the fire, burnt offerings to Baal, which I commanded not, and I ordered not, and it did not come to my \textit{mind}.\lebnote{Literally “heart”}
\verse “\textit{Therefore}\lebnote{Literally “To thus”} look, days are about to come,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “when this place will no longer be called Topheth or the Valley of Ben Hinnom, \textit{but}\lebnote{Literally “surely if”} the Valley of the Slaughter.
\verse And I will lay waste the plans\lebnote{Hebrew “plan”} of Judah and Jerusalem in this place, and I will bring them to ruin by the sword \textit{before}\lebnote{Literally “to the face of”} their enemies, and by the hand of those who seek their life, and I will give their dead bodies\lebnote{Hebrew “dead body”} as food to the birds\lebnote{Hebrew “bird”} of heaven and to the animals\lebnote{Hebrew “animal”} of the earth.
\verse And I will make this city a horror, and an object of hissing, everyone who passes by it will be appalled, and will hiss because of all its wounds.
\verse And I will cause them to eat the flesh of their sons, and the flesh of their daughters, and each one will eat the flesh of his neighbor in the siege and in the distress which their enemies and those who seek their life inflict on them.” ’
\verse “Then you shall break the jar before the eyes of the men who go with you.
\verse And you shall say to them, ‘Thus says Adonai of hosts: “So I will break this people and this city as one breaks the vessel of the potter, so that it is not able to be repaired again. And in Topheth they will bury \textit{until there is no room to bury}.\lebnote{Literally “from there is no place to bury”}
\verse Thus will I do to this place,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “and to its inhabitants, to make this city like Topheth.
\verse And the houses of Jerusalem and the houses of the kings of Judah will be unclean like the place of Topheth, all the houses where they made smoke offerings upon their roofs to all the host of heaven, and where they poured out libations to other gods.” ’ ”
\verse Then Jeremiah came from Topheth, where Adonai had sent him to prophesy, and he stood in the courtyard of the house of Adonai and said to all the people,
\verse “Thus says Adonai of hosts, the God of Israel, ‘Look, I am about to bring to this city and upon all its towns all the disaster that I have pronounced against it, because they have hardened their neck to not hear my words.’ ”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 20
\verseWithHeading{A Message for Pashhur} Now Pashhur, the son of Immer, the priest who was officer in charge in the temple of Adonai, heard Jeremiah prophesying these things.
\verse Then Pashhur struck Jeremiah the prophet and put him in the stocks that were in the upper Gate of Benjamin, which was by the temple of Adonai.
\verse \textit{And then}\lebnote{Literally “And it was”} on the next day, when Pashhur brought Jeremiah out from the stocks, Jeremiah said to him, “Adonai will not call your name Pashhur, \textit{but}\lebnote{Literally “but if”} Terror From All Around.
\verse For thus says Adonai, ‘Look, I am making you a terror to yourself and to all your friends, and they will fall by the sword of their enemies, and your eyes will be seeing it.\lebnote{Here the direct object is supplied from context in the English translation} And all Judah I will give into the hand of the king of Babylon, and he will deport them to Babylon, and he will strike them with the sword.
\verse And I will give all the wealth of this city, and all its acquisitions, and all its precious items, and all of the treasures of the kings of Judah I will give into the hand of their enemies, and they will plunder them, and they will seize them, and they will bring them to Babylon.
\verse And you, Pashhur, and all those who live in your house will go into captivity, and you will go to Babylon, and there you will die, and there you will be buried, you, and all your friends to whom you have prophesied \textit{falsely}.\lebnote{Literally “in the lie”}’ ”
\verseWithHeading{Jeremiah’s Lament} You have persuaded me, O Adonai, and I was persuaded. 
You have overpowered me, and you have prevailed. 
I have become a laughingstock all day long. 
\textit{Everyone}\lebnote{Literally “All of him”} is mocking me.
\verse For \textit{as often as}\lebnote{Literally “from enough”} I speak, 
I must cry out violence and destruction. 
I must shout, for the word of Adonai has become for me 
a disgrace and derision all day long.
\verse But if I say, “I will not mention him 
and I will no longer speak in his name,” 
then it becomes in my heart like a fire burning, 
locked up in my bones, 
and I struggle to contain it,\lebnote{Here the direct object is supplied from context in the English translation} 
and I am not able.
\verse For I hear the rumor of many, 
“Terror is from all around, denounce him,\lebnote{Here the direct object is supplied from context in the English translation} 
yes, let us denounce him!” 
\textit{All my close friends}\lebnote{Literally “All the human being of my peace”} are watchers of my stumbling, 
saying, “perhaps he can be persuaded, 
and we can prevail over him, 
and we can take our revenge on him.”
\verse But Adonai is with me like a powerful warrior. 
\textit{Therefore}\lebnote{Literally “Unto thus”} my persecutors will stumble and will not prevail. 
They will be very ashamed, for they will not achieve success. 
Their \textit{everlasting insult}\lebnote{Literally “insult of eternity”} will not be forgotten.
\verse Yet, O Adonai of hosts, who tests the righteous, 
who sees the \textit{heart}\lebnote{Literally “kidneys”} and the \textit{mind},\lebnote{Literally “heart”} 
let me see your retribution upon them, 
for to you I have revealed my legal case.
\verse Sing to Adonai, praise Adonai, 
for he has delivered the life of the poor from the hand of evildoers.
\verse Cursed be the day on which I was born, 
let not the day on which my mother gave birth to me be blessed.
\verse Cursed be the man who brought the news to my father, 
\textit{saying},\lebnote{Literally “to say”} “A child is born to you, a son!” 
and he made him very glad.
\verse And let that man be like the cities 
that Adonai demolished without regret, 
and let him hear a cry for help in the morning, 
and an alarm at the time of noon.
\verse Because he did not kill me in the womb, 
so that my mother would have been for me my grave, 
and her womb would be pregnant forever.
\verse \textit{Why}\lebnote{Literally “To what this”} did I come out from the womb, 
to see toil, and sorrow, and to \textit{end my days in shame}?\lebnote{Literally “my days will come to an end in shame”}
\end{biblechapter}

\begin{biblechapter} % Jeremiah 21
\verseWithHeading{A Message for Zedekiah} The word that came to Jeremiah from Adonai when King Zedekiah sent to him Pashhur, the son of Malchiah, and Zephaniah, the son of Maaseiah, the priest, \textit{saying},\lebnote{Literally “to say”}
\verse “Please inquire of Adonai on behalf of us, for Nebuchadnezzar the king of Babylon is fighting against us. Perhaps Adonai will do with us according to all his miraculous acts, so that he may go up from against us.”
\verse Then Jeremiah said to them, “Thus you shall say to Zedekiah:
\verse ‘Thus says Adonai, the God of Israel: “Look, I am about to turn back the weapons of war that are in your hands,\lebnote{Hebrew “hand”} with which you are fighting against them, the king of Babylon and the Chaldeans who are laying siege to you from outside the wall, and I will gather them into the center of this city.
\verse And I myself will fight against you with outstretched hand, and with strong arm, and in \textit{anger},\lebnote{Literally “nose”} and in fury, and in great wrath.
\verse And I will strike the inhabitants of this city, both humankind and animal; by a great plague they will die.
\verse And \textit{afterward},”\lebnote{Literally “after thus”} \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “I will give Zedekiah, the king of Judah, and his servants, and the people, and those who remain in this city from the plague, from the sword, and from the famine, into the hand of Nebuchadnezzar, the king of Babylon, and into the hand of their enemies, and into the hand of those who seek their lives, and he will strike them with the \textit{edge}\lebnote{Literally “mouth”} of the sword. He will not take pity on them, and he will not have compassion, nor will he show compassion.” ’
\verseWithHeading{The Way of Life and the Way of Death} “And to this people you shall say, ‘Thus says Adonai, “Look, I am setting \textit{before you}\lebnote{Literally “to the face of you”} the way of life and the way of death.
\verse He who stays in this city will die by the sword, or by the famine, or by the plague. And he who goes out and goes over to the Chaldeans who are laying siege to you will live, and his life will be to him as booty.
\verse For I have set my face against this city for evil and not for good,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai. It will be given into the hand of the king of Babylon, and he will burn it with fire.” ’
\verseWithHeading{A Series of Messages for the House of the King of Judah} “And to the house of the king of Judah, ‘Hear the word of Adonai,
\verse O house of David, thus says Adonai,
\verse “Execute judgment in the morning, justice, 
and deliver the one who has been seized 
from the hand of the oppressor, 
lest my wrath goes forth like the fire and it burns, 
and there is no one who quenches it\lebnote{Here the direct object is supplied from context in the English translation} 
because of the evil of your deeds.
\verse Look, I am against you, 
O inhabitants\lebnote{Hebrew “inhabitant”} of the valley, 
O rock of the plain,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai; 
“you who say, ‘Who can descend against us, 
or who can enter into our hiding place?’
\end{biblechapter}

\begin{biblechapter} % Jeremiah 22
\verse Thus says Adonai: “Go down to the house of the king of Judah and you must speak this word there.
\verse And you must say, ‘Hear the word of Adonai, O king of Judah, who sits on the throne of David, you, and your servants, and your people who enter through these gates.
\verse Thus says Adonai, “Act with justice and righteousness, and deliver the one who has been seized from the hand of the oppressor. And you must not oppress the immigrant, the orphan, and the widow. And you must not shed innocent blood in this place.
\verse And if you indeed obey this word, then kings who sit for David on his throne will enter through the gates of this house, riding in chariots\lebnote{Hebrew “chariot”} and on horses, he, and his servants, and his people.
\verse But if you do not obey these words, I swear by myself,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “that this house will become a site of ruins.” ’ ”
\verse For thus says Adonai concerning the house of the king of Judah,
\verse “You are like Gilead to me, 
like the head of the Lebanon. 
\textit{Surely}\lebnote{Literally “If not”} I will make you a desert,\lebnote{Or “wilderness”} 
like cities that are not inhabited.
\verse And many nations will pass by this city and they will say each one to his neighbor, ‘\textit{Why}\lebnote{Literally “Over what”} has Adonai done in this manner to this great city?’
\verse Then they will answer, ‘Because they abandoned the covenant of Adonai their God, and bowed down in worship to other gods, and served them.’ ”
\verse You must not weep for the dead person, 
and you must not show sympathy for him. 
Weep bitterly for the one who goes away, 
for he will not return, or see the land of his birth again.
\verse For thus says Adonai concerning Shallum, the son of Josiah, the king of Judah, who reigned as king in place of Josiah his father, who went out from this place: “He will not return here again.
\verse But in the place where they have deported him, there he will die, and he will not see this land again.
\verse Woe to the one who builds his house without righteousness, 
and his upper rooms without justice. 
His fellow countryman, he works for nothing, 
and he does not give to him his wages.\lebnote{Hebrew “wage”}
\verse Who says ‘I will build for myself 
a spacious house with large upper rooms,’ 
and he cuts windows for it, 
and it is\lebnote{Here the subject is supplied from context in the English translation} paneled with cedar, 
and he paints it\lebnote{Here the direct object is supplied from context in the English translation} with vermilion.
\verse Do you reign as king because you are competing in cedar? 
Did not your father eat and drink, 
and he did justice and righteousness, 
then it was well with him?
\verse He judged the legal cause of the needy and the poor, 
and then it was well. 
\textit{Is that not what it means to know me}?\lebnote{Literally “Not that the knowledge me”}” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “But there is nothing in your eyes and your heart 
\textit{except}\lebnote{Literally “but if”} your unlawful gain, 
and on shedding the blood of the innocent, 
and on committing oppression and extortion.”
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai concerning Jehoiakim, the son of Josiah, the king of Judah:
\verse “They will not lament for him, 
‘Alas, my brother,’ or ‘Alas, sister.’ 
They will not lament for him, 
‘Alas, lord,’ or ‘alas, his majesty.’
\verse He will be buried with the burial of a donkey. 
He will be\lebnote{Here the subject is supplied from context in the English translation} dragged away and thrown \textit{outside}\lebnote{Literally “from beyond”} the gates of Jerusalem.
\verse Go up to Lebanon and cry out, 
and in Bashan lift up your voice, 
and cry out from Abarim, 
for all your lovers are shattered.
\verse I spoke to you \textit{in the times you were secure},\lebnote{Literally “in your security”} 
and you said, ‘I will not listen!’ 
This was your way from your youth, 
for you have not obeyed my voice.
\verse All your shepherds will shepherd the wind, 
and your lovers will go into captivity. 
Yes, then you will be ashamed and humiliated 
because of all your wickedness.
\verse “As I live,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “surely if Coniah, the son of Jehoiakim, the king of Judah, were the seal on my right hand, surely from there I would wrench you off.
\verse And I would give you into the hand of those who seek your life, and into the hand of those from whom you are frightened by their presence, and into the hand of Nebuchadnezzar, the king of Babylon, and into the hand of the Chaldeans.
\verse And I will throw you and your mother who gave birth to you to another country where you were not born, and there you will die.
\verse As for the land to which they are \textit{longing}\lebnote{Literally “lifting their inner self”} to return, they will not return.
\verse Is this man Coniah a despised, shattered vessel? 
Or a vessel in which there is no delight? 
Why are he and his offspring thrown far, 
and cast away to the land that they do not know?”
\verse O land, land, land, hear the word of Adonai.
\verse Thus says Adonai: “Record this man as childless, 
a man who will not succeed in his days, 
for no man from his offspring will succeed him,\lebnote{Here the direct object is supplied from context in the English translation} 
sitting on the throne of David and ruling again in Judah.”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 23
\verseWithHeading{Woe to the Evil Shepherds} “Woe to the shepherds who destroy and scatter the flock of my pasture,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai, the God of Israel concerning the shepherds who shepherd my people, “You yourselves have scattered my flock, and you have driven them away, and you do not attend to them. Look, I will punish you for the evil of your deeds,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “Then I myself will gather together the remnant of my flock from all the lands where I have driven them, and I will bring them back to their grazing place, and they will be fruitful, and they will become numerous.
\verse And I will raise up over them shepherds, and they will shepherd them, and they will no longer fear, and they will not be dismayed, and they will not be missing,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verseWithHeading{The Righteous Branch} “Look, days are coming,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“when I will raise up for David a righteous branch, 
and he will reign as king, and he will achieve success, 
and he will do justice and righteousness in the land.
\verse In his days Judah will be saved, 
and Israel will dwell \textit{in safety},\lebnote{Literally “with confidence”} 
and this is his name by which he will be called: 
‘Adonai is our righteousness.’
\verse “\textit{Therefore}\lebnote{Literally “To thus”} look, days are coming,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “when they will no longer say, ‘\textit{As Adonai lives},\lebnote{Literally “the life of Adonai”} who led up the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} from the land of Egypt,’
\verse but ‘\textit{As Adonai lives},\lebnote{Literally “the life of Adonai”} who led up, and who brought the offspring of the house of Israel from the land of the north and from all the lands where he had driven them.’ Then they will live in their land.”
\verseWithHeading{The Unfaithful Prophets} Concerning the prophets:
\verse My heart is broken in my midst. 
All my bones tremble. 
I have become like a drunken man, 
even like a man over whom wine has passed, 
because of Adonai, 
and because of \textit{his holy words}.\lebnote{Literally “words of his holiness”}
\verse For the land is full of adulterers; 
for the land mourns because of a curse. 
The pastures of the desert\lebnote{Or “wilderness”} are dry, 
and their evil has been their way of running, 
and their power is not right.
\verse “For both prophet as well as priest are godless, 
even in my temple I have found their wickedness,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “\textit{Therefore}\lebnote{Literally “To thus”} their way will be to them like the slippery places, 
they will be pushed in the darkness, 
and they will fall into it, 
for I will bring disaster on them in the year of their punishment,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “Now in the prophets of Samaria I saw a disgusting thing. 
They prophesied by Baal and they caused my people Israel to err.
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai concerning the prophets, “Look, I am going to let them eat wormwood 
and I will give them water of poison to drink, 
for from the prophets of Jerusalem 
went out ungodliness to all the land.”
\verse Thus says Adonai of hosts,
\verse “You must not listen to the words of the prophets who prophesy to you. 
They are deluding you with visions of their \textit{mind},\lebnote{Literally “heart”} 
They do not speak from the mouth of Adonai.
\verse They are\lebnote{Here the subject and verb are supplied from context in the English translation} continually saying to those who disregard the word of Adonai, 
‘Peace it will be to you,’ 
and to \textit{each one}\lebnote{Literally “all”} who walks in the stubbornness of his heart they say, 
‘Calamity will not come upon you.’
\verse For who has stood in the council of Adonai, 
that he has seen and heard his word? 
Who has listened attentively to his word 
and heard it?\lebnote{Here the direct object is supplied from context in the English translation}
\verse Look, the storm of Adonai has gone forth in wrath, 
even a whirling tempest. 
It will whirl upon the head of the wicked.
\verse The \textit{anger}\lebnote{Literally “nose”} of Adonai will not turn back 
until his doing and until his keeping the plans of his \textit{mind}.\lebnote{Literally “heart”} 
\textit{In latter days}\lebnote{Literally “In the last of the days”} you will look closely at it with understanding.
\verse I have not sent the prophets, yet they ran. 
I have not spoken to them, yet they prophesied.
\verse But if they had stood in my council, 
then they would have proclaimed my words to my people, 
and they would have caused them to turn from their evil way, 
and from the evil of their deeds.
\verse Am I a God from near,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“and not a God from far?
\verse “I have heard what the prophets who prophesy lies\lebnote{Hebrew “lie”} have said in my name, \textit{saying},\lebnote{Literally “to say”} ‘I have dreamed! I have dreamed!’
\verse \textit{How long}\lebnote{Literally “Until when”} will this be in the hearts\lebnote{Hebrew “heart”} of the prophets who prophesy lies,\lebnote{Hebrew “the lie”} even the prophets of the deceitfulness of their hearts,\lebnote{Hebrew “heart”}
\verse who plan to make my people forget my name by their dreams that they tell each one to his neighbor, just as their ancestors\lebnote{Or “fathers”} forgot my name through Baal?
\verse The prophet who has with him a dream, let him tell the dream. But the prophet\lebnote{Here the subject is supplied from context in the English translation} who has my word with him, let him speak my word faithfully. \textit{What is straw compared to wheat}?”\lebnote{Literally “What to the straw in comparison with the wheat”} \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “Is not my word like fire?” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “and like a blacksmith’s hammer that breaks a rock into pieces?
\verse “\textit{Therefore}\lebnote{Literally “To thus”} look, I am against the prophets,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “who steal my words each one from his neighbor.
\verse Look, I am against the prophets,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “who take their tongues\lebnote{Hebrew “tongue”} and declare as prophets, ‘Adonai \textit{declares}.’\lebnote{Literally “a declaration of”}
\verse Look, I am against those who prophesy dreams of lies,”\lebnote{Hebrew “lie”} \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “and tell them, and they caused my people to err through their lies, and in their recklessness, when I myself have not sent them nor commanded them, so they profit not this people at all,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “Now when this people, or a prophet, or a priest, ask you, \textit{saying},\lebnote{Literally “to say”} ‘What is the burden\lebnote{Or “oracle”} of Adonai?’ then you shall say to them, ‘\textit{You are the burden},\lebnote{Literally “What burden”} and I will forsake you,’ ” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “And the prophet, and the priest, and the people who say, ‘The burden\lebnote{Or “oracle”} of Adonai,’ I will punish that man and his house.
\verse Thus you shall say, each one to his neighbor and each one to his brother: ‘What has Adonai answered?’ or, ‘What has Adonai spoken?’
\verse But you shall not mention the burden\lebnote{Or “oracle”} of Adonai, for the burden\lebnote{Or “oracle”} is to each one his word, and so you twist the words of the living God, Adonai of hosts, our God.
\verse Thus you shall say to that prophet, ‘What has Adonai answered you?’ or, ‘What has Adonai spoken?’
\verse But if you say, ‘The burden\lebnote{Or “oracle”} of Adonai,’ \textit{therefore}\lebnote{Literally “to thus”} thus says Adonai, ‘because of your saying these words,\lebnote{Hebrew “word”} “The burden\lebnote{Or “oracle”} of Adonai,” then I will send to you, \textit{saying},\lebnote{Literally “to say”} “You shall not say, ‘The burden\lebnote{Or “oracle”} of Adonai.’ ” ’
\verse \textit{Therefore}\lebnote{Literally “To thus”} look, here I am, and I will surely lift you up and forsake you and the city that I gave to you and to your ancestors\lebnote{Or “fathers”} from \textit{my presence}\lebnote{Literally “my face”},
\verse and I will bring upon you \textit{an everlasting disgrace}\lebnote{Literally “a disgrace of eternity”} and \textit{an everlasting shame}\lebnote{Literally “a shame of eternity”} that will not be forgotten.”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 24
\verseWithHeading{Two Baskets of Figs} Adonai showed me, and look, there were two baskets of figs placed \textit{before}\lebnote{Literally “to the face of”} the temple of Adonai—after Nebuchadnezzar the king of Babylon had deported Jeconiah the son of Jehoiakim, the king of Judah, with the officials of Judah, and the craftsmen, and the smiths,\lebnote{Hebrew “smith”} from Jerusalem and had brought them to Babylon.
\verse The one basket had very good figs, like \textit{early figs},\lebnote{Literally “the figs of the early ripened fruit”} and the other basket had very bad figs that could not be eaten because of their bad quality.
\verse And Adonai asked me, “What are you seeing, Jeremiah?” And I said, “Figs—the good figs, very good, and the bad figs, very bad, that cannot be eaten because of their bad quality.”
\verse Then the word of Adonai came to me, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai, the God of Israel, ‘Like these good figs, so I will regard as good the exiles\lebnote{Hebrew “exile”} of Judah whom I have sent away from this place to the land of the Chaldeans.
\verse For I will set my eyes\lebnote{Hebrew “eye”} on them for good, and I will bring them back to this land. And I will build them and not annihilate them, and I will plant them and not uproot them.
\verse And I will give to them a heart to know me, that I am Adonai, and \textit{they will be my people},\lebnote{Literally “they shall be for me as a people”} and \textit{I will be their God},\lebnote{Literally “I will be for them as God”} for they will return to me with the whole of their heart.
\verse But like the bad figs that cannot be eaten because of their bad quality—for thus says Adonai—so I will treat Zedekiah the king of Judah, and his officials, and the remnant of Jerusalem who remain in this land, and those who live in the land of Egypt.
\verse And I will make them as a terror, an evil to all the kingdoms of the earth, as a disgrace and a proverb, as a taunt and a curse, in all the places where I will drive them.
\verse And I will send among them the sword, the famine, and the plague, until they perish from the land that I gave to them and their ancestors.’ ”\lebnote{Or “fathers”}
\end{biblechapter}

\begin{biblechapter} % Jeremiah 25
\verseWithHeading{Seventy Years of Servitude} The word that came to Jeremiah concerning all the people of Judah in the fourth year of Jehoiakim the son of Josiah, the king of Judah, that was the first year of Nebuchadnezzar the king of Babylon,
\verse which Jeremiah the prophet spoke to all the people of Judah and to all the inhabitants of Jerusalem, \textit{saying},\lebnote{Literally “to say”}
\verse “From the thirteenth year of Josiah, the son of Amon, the king of Judah, even up to this day, these twenty-three years\lebnote{Hebrew “year”} the word of Adonai has come to me, and \textit{I have spoken to you over and over again},\lebnote{Literally “I have spoken to you doing early and speaking”} but you have not listened.
\verse And though Adonai has sent all his servants the prophets to you \textit{over and over again}\lebnote{Literally “do early and sending”} you have not listened, and you have not inclined your ear to hear,
\verse \textit{saying},\lebnote{Literally “to say”} ‘Turn back please, each one from his evil way, and from the evil of your deeds, and live on the land that Adonai has given to you and to your ancestors\lebnote{Or “fathers”} from a long time back and until forever,
\verse and you must not go after other gods to serve them, and to bow in worship to them, and you must not provoke me to anger with the work of your hands, then I will not do something bad to you.’
\verse ‘Yet you have not listened to me,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘so that you\lebnote{Hebrew “they”} have provoked me to anger with the work of your hands for your own harm.’
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai of hosts, ‘Because you have not obeyed my words,
\verse look, I am going to send and take all the clans\lebnote{Or “families”} of the north,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘and I will send\lebnote{Here the subject and verb are supplied from context in the English translation} to Nebuchadnezzar the king of Babylon, my servant, and I will bring them against this land, and against its inhabitants, and against all these nations all around, and I will destroy them, and I will make them a horror, and an object of hissing, and \textit{everlasting ruins}.\lebnote{Literally “sites of ruins of eternity”}
\verse And I will exterminate from them the sound of jubilation, and the sound of joy, the voice of the bridegroom, and the voice of the bride, the sound of the millstones, and the light of the lamp.
\verse And all this land will become a site of ruins, a desolation, and these nations will serve the king of Babylon seventy years.\lebnote{Hebrew “year”}
\verseWithHeading{The Cup of Wrath} \textit{And then}\lebnote{Literally “And it will happen”} when the seventy years\lebnote{Hebrew “year”} are fulfilled, I will punish the king of Babylon and that nation,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘for their iniquity, and the land of the Chaldeans, and I will make it \textit{an everlasting waste}.\lebnote{Literally “a waste of eternity”}
\verse And I will bring upon that land all my words that I have spoken against it, everything that is written in this scroll which Jeremiah prophesied against all the nations.
\verse For many nations and great kings will let them work, even them, and I will repay them according to their deeds, and according to the work of their hands.’ ”
\verse For thus said Adonai, the God of Israel, to me, “Take this cup of the wine of wrath from my hand, and you must give it to all the nations to whom I am sending you to drink.
\verse And they will drink, and they will stagger, and they will act like madmen because of the \textit{presence}\lebnote{Literally “face”} of the sword that I am sending among them.”
\verse So I took the cup from the hand of Adonai and I gave it\lebnote{Here the direct object is supplied from context in the English translation} to all the nations to whom Adonai sent me to drink:
\verse Jerusalem and the towns of Judah, and its kings, and its officials, to make them a site of ruins, a horror, an object of hissing, and a curse, as it is this day;
\verse Pharaoh, the king of Egypt, and his servants, and his officials, and all his people;
\verse and all the other\lebnote{Or “foreign”} people; and all the kings of the land of Uz; and all the kings of the land of the Philistines—\lebnote{Hebrew “Philistine”}Ashkelon, and Gaza, and Ekron, and the remnant of Ashdod;
\verse Edom, and Moab, and the \textit{Ammonites};\lebnote{Literally “sons of Ammon”}
\verse and all the kings of Tyre, and all the kings of Sidon, and the kings of the coastland that is across the sea;
\verse and Dedan, and Tema, and Buz, and all those who are trimmed to the side;
\verse and all the kings of Arabia, and all the kings of the other\lebnote{Or “foreign”} people who dwell in the desert;\lebnote{Or “wilderness”}
\verse and all the kings of Zimri, and all the kings of Elam, and all the kings of Media;
\verse and all the kings of the north, near and far, each one to his brother, and all the kingdoms of the world that are on the face of the earth; and the king of Sheshach, he will drink after them.
\verse “And you shall say to them, ‘Thus says Adonai of hosts, the God of Israel, “Drink and become drunk, and vomit, and fall, and you must not rise because of the \textit{presence}\lebnote{Literally “face”} of the sword that I am sending among you.” ’
\verse And it will happen, if they refuse to take the cup from your hand to drink, then you shall say to them, ‘Thus says Adonai of hosts, “You must certainly drink!
\verse For look, on the city that is called \textit{by}\lebnote{Literally “on it”} my name I am beginning to inflict harm, and you indeed, will you go unpunished? You will not go unpunished, for I am summoning a sword against all the inhabitants of the earth,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai of hosts.’
\verse And you yourself shall prophesy against them all these words, and you shall say to them:
\verse ‘Adonai will roar from on high, 
and from \textit{his holy dwelling place}\lebnote{Literally “the dwelling place of his holiness”} he will give his voice. 
He will roar mightily against his settlement, 
a jubilant shout like those who tread grapes, 
he will answer against all the inhabitants of the earth.
\verse Thus says Adonai of hosts,
\verse “Look, disaster is going out from nation to nation, 
and a great tempest is stirred up from the farthest parts of the earth.
\verse And the slain ones of Adonai will be on that day 
from one end of the earth to the other end of the earth, 
and they will not be mourned, 
and they will not be gathered, 
and they will not be buried. 
They will become like dung on the \textit{surface}\lebnote{Literally “face”} of the ground.
\verse Wail, you shepherds, and cry out. 
And roll about in mourning in the dust, O leaders\lebnote{Or “nobles”} of the flock. 
For your days of slaughter and your dispersions have arrived, 
and you will fall like a vessel of desire.
\verse And flight will perish from the shepherds, 
and escape from the leaders\lebnote{Or “nobles”} of the flock.
\verse A sound, the cry of distress of the shepherds 
and the wailing of the leaders\lebnote{Or “nobles”} of the flock, 
for Adonai is despoiling their pasture.
\verse And the \textit{peaceful settlements}\lebnote{Literally “settlements of the peace”} are devastated 
because of \textit{the burning anger of Adonai}.\lebnote{Literally “the blaze of the nose of Adonai”}
\end{biblechapter}

\begin{biblechapter} % Jeremiah 26
\verseWithHeading{Jeremiah Escapes Death} At the beginning of the reign of Jehoiakim the son of Josiah, the king of Judah, this word came from Adonai, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai: ‘Stand in the courtyard of the house of Adonai, and you must speak to all the cities of Judah that come to bow in worship in the house of Adonai all the words that I command you to speak to them. You must not omit a word.
\verse Perhaps they will listen and turn back each from his evil way, and I will relent of the disaster that I am planning to do to them because of their evil deeds.
\verse And you shall say to them, “Thus says Adonai: ‘If you will not listen to me, to walk in my law that I have set\lebnote{Or “given”} \textit{before you},\lebnote{Literally “to the face of you”}
\verse to listen to the words of my servants the prophets whom I have sent to you, \textit{over and over again},\lebnote{Literally “and doing early and sending”} though you have not listened,
\verse then I will make this house like Shiloh, and this city I will make a curse for all the nations of the earth.’ ” ’ ”
\verse And the priests and the prophets and all the people heard Jeremiah speaking these words in the temple of Adonai.
\verse \textit{And then}\lebnote{Literally “And it happened”} as Jeremiah finished speaking all that Adonai had commanded him to speak to all the people, then the priests and the prophets and all the people laid hold of him, \textit{saying},\lebnote{Literally “to say”} “You will die!
\verse Why have you prophesied in the name of Adonai, \textit{saying},\lebnote{Literally “to say”} ‘This house will be like Shiloh, and this city will be in ruins, \textit{without}\lebnote{Literally “from there is not”} inhabitant’?” And all the people gathered around Jeremiah in the temple of Adonai.
\verse When the officials of Judah heard these things, they came up from the house of the king to the temple of Adonai, and they sat in the entrance of the New Gate of Adonai’s temple.
\verse Then the priests and the prophets said to the officials and to all the people, \textit{saying},\lebnote{Literally “to say”} “\textit{This man deserves the death sentence},\lebnote{Literally “A sentence of death is to this man”} because he has prophesied against this city as that which you have heard with your ears.”
\verse Then Jeremiah said to all the officials and to all the people, \textit{saying},\lebnote{Literally “to say”} “Adonai sent me to prophesy against this house and against this city all the words that you have heard.
\verse Now \textit{therefore}\lebnote{Literally “so then”} amend your ways and your deeds, and obey the voice of Adonai your God, and Adonai will relent of the disaster that he has spoken over you.
\verse But as for me, look, I am in your hand, do to me what is good and right in your eyes.
\verse Only you must certainly know that if you put me to death, you will bring on yourselves innocent blood, and on this city and on its inhabitants, for \textit{truly}\lebnote{Literally “in truth”} Adonai sent me to you to speak all these words in your ears.”
\verse Then the officials and all the people said to the priests and to the prophets, “\textit{This man does not deserve}\lebnote{Literally “Not to this man”} a sentence of death, for in the name of Adonai our God, he has spoken to us.”
\verse Then men from the elders of the land arose and said to all the assembly of the people, \textit{saying},\lebnote{Literally “to say”}
\verse “Micah the Morashtite was prophesying in the days of Hezekiah, the king of Judah, and he said to all the people of Judah, \textit{saying}:\lebnote{Literally “to say”} ‘Thus says Adonai of hosts, 
“Zion will be plowed, 
and Jerusalem will become a heap of ruins, 
and the mountain of the \textit{temple}\lebnote{Literally “house”} as high places of wood.” ’
\verse Did Hezekiah, the king of Judah, and all Judah actually put him to death? Was he not in fear of Adonai? And he entreated the face of Adonai, and Adonai relented of the disaster that he had spoken against them. But we are about to do great disaster to ourselves.”
\verse Indeed, there also was a man prophesying in the name of Adonai, Uriah, the son of Shemaiah, from Kiriath-Jearim, and he prophesied against this city and against this land like all the words of Jeremiah.
\verse And when King Jehoiakim, and all his warriors, and all the officials heard his words, then the king sought to put him to death. But Uriah heard, and he was afraid, and he fled and went to Egypt.
\verse Then King Jehoiakim sent men to Egypt. Elnathan, the son of Achbor, and men with him went to Egypt.
\verse And they brought out Uriah from Egypt and they brought him to King Jehoiakim, and he struck him down with the sword, and he threw his dead body into the burial sites of the sons of the people.
\verse However, the hand of Ahikam the son of Shaphan was with Jeremiah, so that he was not given into the hand of the people to put him to death.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 27
\verseWithHeading{Jeremiah Uses Fetters and Yokes to Illustrate His Message of Submission} In the beginning of the reign of Jehoiakim, the son of Josiah, the king of Judah, this word came to Jeremiah from Adonai, \textit{saying}\lebnote{Literally “to say”}—
\verse thus said Adonai to me—“Make for yourself fetters and yokes and put them on your neck,
\verse and send them to the king of Edom, and to the king of Moab, and to the king of the \textit{Ammonites},\lebnote{Literally “sons of Ammon”} and to the king of Tyre, and to the king of Sidon in the hand of the envoys who have come to Jerusalem, to Zedekiah, the king of Judah.
\verse And you must command them for their masters, \textit{saying},\lebnote{Literally “to say”} ‘Thus says Adonai of hosts, the God of Israel, this you shall say to your masters:
\verse “I have made the earth with humankind and animals\lebnote{Hebrew “animal”} that are on the face of the earth by my great power and by my outstretched arm, and I give it to whomever is right in my eyes.
\verse And now I myself have given all these lands into the hand of Nebuchadnezzar the king of Babylon, my servant, and also the animals\lebnote{Hebrew “animal”} of the field I have given to him to serve him.
\verse And all the nations will serve him, and his son, and \textit{his grandson},\lebnote{Literally “the son of his son”} until the coming of the time of \textit{his own}\lebnote{Literally “also his”} land. Then many nations and great kings will let him work.
\verse “But it will be that the nation or kingdom that will not serve him, Nebuchadnezzar the king of Babylon, and that will not put his neck under the yoke of the king of Babylon, I will punish that nation with the sword, and with the famine, and with the plague,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “until I have destroyed it with my hand.
\verse And you, you must not listen to your prophets, and to your diviners, and to your dreamers, and to your interpreters of signs, and to your sorcerers who are \textit{saying},\lebnote{Literally “to say”} ‘You will not serve the king of Babylon.’
\verse For they are prophesying to you a lie, so that you will be removed from your land, and I will drive you away, and you will perish.
\verse But the nation that will bring its neck under the yoke of the king of Babylon and will serve him, yet will I leave it on its land,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “and they will till it, and they will live in it.” ’ ”
\verse And I spoke words like these to Zedekiah, the king of Judah, \textit{saying},\lebnote{Literally “to say”} “Bring your necks under the yoke of the king of Babylon, and serve him and his people, and live.
\verse Why should you die—you and your people—by the sword, by the famine, and by the plague, as Adonai has spoken concerning the nation that will not serve the king of Babylon?
\verse And you must not listen to the words of the prophets who are speaking to you, \textit{saying},\lebnote{Literally “to say”} ‘You must not serve the king of Babylon,’ for they are prophesying a lie to you.
\verse For I have not sent them,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “but they are prophesying in my name \textit{falsely},\lebnote{Literally “to the lie”} so that I will drive them away. And you will perish—you and the prophets who are prophesying to you.”
\verseWithHeading{The Vessels of the Temple} Then I spoke to the priests and to all this people, \textit{saying},\lebnote{Literally “to say”} “Thus says Adonai, ‘You must not listen to the words of your prophets who are prophesying to you, \textit{saying},\lebnote{Literally “to say”} “Look, the vessels of the house of Adonai are about to be quickly brought back from Babylon”, for they are prophesying a lie to you.
\verse And you must not listen to them; serve the king of Babylon and live. Why should this city become a site of ruins?
\verse But if they are prophets, and if there is with them the word of Adonai, let them please plead with Adonai of hosts, that the vessels that are left over in the \textit{temple}\lebnote{Literally “house”} of Adonai, and the house of the king of Judah, and in Jerusalem, must not go to Babylon.’
\verse For thus says Adonai of hosts concerning the pillars, and concerning the sea, and concerning the stands, and concerning the rest of the vessels that are left in this city,
\verse which Nebuchadnezzar the king of Babylon, did not take when he deported Jeconiah, the son of Jehoiakim, the king of Judah, from Jerusalem to Babylon, \textit{along with}\lebnote{Literally “and”} all the nobles of Judah and Jerusalem.
\verse For thus says Adonai of hosts, the God of Israel, concerning the vessels that are left in the house of Adonai, and in the house of the king of Judah, and in Jerusalem:
\verse ‘They will be brought to Babylon, and there they will stay until the day of my attending to them,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai. ‘Then I will bring them up and restore them to this place.’ ”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 28
\verseWithHeading{Hananiah, the False Prophet} And it was in that same year, at the beginning of the reign of Zedekiah, the king of Judah, in the fourth year, in the fifth month, that Hananiah, the son of Azzur, the prophet who was from Gibeon, said to me in the house of Adonai before the eyes of the priests and all the people, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai of hosts, the God of Israel, \textit{saying},\lebnote{Literally “to say”} ‘I have broken the yoke of the king of Babylon.
\verse \textit{Within two years}\lebnote{Literally “In still two years days”} I will bring back to this place all the vessels of the house of Adonai which Nebuchadnezzar the king of Babylon took away from this place and brought to Babylon.
\verse And Jeconiah the son of Jehoiakim, the king of Judah, and all the exiles\lebnote{Hebrew “exile”} from Judah who went to Babylon, I will bring back to this place,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘For I will break the yoke of the king of Babylon.’ ”
\verse Then Jeremiah the prophet said to Hananiah the prophet before the eyes of the priests and before the eyes of all the people who were standing in the house of Adonai,
\verse and Jeremiah the prophet said, “Amen! May Adonai do so; may Adonai fulfill your words that you have prophesied, to bring back the vessels of the house of Adonai and all the exiles\lebnote{Hebrew “exile”} from Babylon to this place.
\verse Nevertheless listen please to this word that I am speaking in your \textit{hearing}\lebnote{Literally “ears”} and in the \textit{hearing}\lebnote{Literally “ears”} of all the people,
\verse The prophets who were \textit{before}\lebnote{Literally “to the face of”} me and \textit{before}\lebnote{Literally “to the face of”} you from ancient times prophesied against many countries and against great kingdoms of war, and of disaster, and of plague.
\verse The prophet who prophesies peace, at the coming of the word of the prophet, will become known as the prophet that Adonai has \textit{truly}\lebnote{Literally “in truth”} sent.”
\verse Then Hananiah the prophet took the yoke from the neck of Jeremiah the prophet and broke it.
\verse Then Hananiah said before the eyes of all the people, \textit{saying},\lebnote{Literally “to say”} “Thus says Adonai, ‘This is how I will break the yoke of Nebuchadnezzar the king of Babylon, \textit{within two years},\lebnote{Literally “in still two years days”} from the neck of all the nations.’ ” And Jeremiah the prophet went on his way.
\verse And the word of Adonai came to Jeremiah after Hananiah the prophet broke the yoke from the neck of Jeremiah the prophet, \textit{saying},\lebnote{Literally “to say”}
\verse “Go, and you must say to Hananiah, \textit{saying},\lebnote{Literally “to say”} ‘Thus says Adonai, “You have broken yoke bars of wood, but you have made in place of them yoke bars of iron.”
\verse For thus says Adonai of hosts, the God of Israel, “I have put a yoke of iron on the neck of all these nations, to serve Nebuchadnezzar, the king of Babylon, and they will serve him, and I have even given the animals\lebnote{Hebrew “animal”} of the open field to him.” ’ ”
\verse Then Jeremiah the prophet said to Hananiah the prophet, “Please listen, Hananiah, Adonai has not sent you, and you have made this people trust in a lie.
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai, ‘Look, I am going to send you from the face of the earth. This year you will die because you have spoken rebellion against Adonai.’ ”
\verse And Hananiah the prophet died in that same year in the seventh month.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 29
\verseWithHeading{Jeremiah’s Letter to the Exiles in Babylon} And these are the words of the letter that Jeremiah the prophet sent from Jerusalem to the remainder of the exiles,\lebnote{Hebrew “exile”} and to the priests, and to the prophets, and to all the people whom Nebuchadnezzar had deported from Jerusalem to Babylon,
\verse after the going out of Jeconiah the king, and the queen mother, and the court officials, the leaders of Judah and Jerusalem, and the artisans,\lebnote{Hebrew “artisan”} and the smiths\lebnote{Hebrew “smith”} from Jerusalem,
\verse by the hand of Elasah, the son of Shaphan, and Gemariah, the son of Hilkiah, whom Zedekiah, the king of Judah, sent to Nebuchadnezzar, the king of Babylon, to Babylon, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai of hosts, the God of Israel, to all the exiles\lebnote{Hebrew “exile”} whom I have deported from Jerusalem to Babylon,
\verse ‘Build houses and live in them,\lebnote{Here the direct object is supplied from context in the English translation} and plant gardens and eat their fruit.
\verse Take wives and father sons and daughters, and take for your sons wives, and give your daughters to men that they may bear sons and daughters, and multiply there, and you must not be few.
\verse And seek the prosperity of the city where I have deported you, and pray on behalf of it to Adonai, for in its prosperity you will have prosperity.’
\verse For thus says Adonai of hosts, the God of Israel, ‘Do not let your prophets who are in your midst, and your diviners, deceive you, and you must not listen to your dreams that you are causing them to dream.
\verse For they are prophesying \textit{falsely}\lebnote{Literally “in a lie”} to you in my name; I have not sent them,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse For thus says Adonai, ‘\textit{As soon as the time has passed},\lebnote{Literally “When to the mouth of being full”} seventy years\lebnote{Hebrew “year”} for Babylon, I will attend to you, and I will fulfill my good word to you, to bring you back to this place.
\verse For I know the plans that I am planning concerning you,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘plans for prosperity and not for harm, to give to you a future and a hope.
\verse Then when you call me, and you come and pray to me, then I will listen to you.
\verse When you search for me, then you will find me, if you seek me with all your heart.
\verse And I will let myself be found by you,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘and I will restore your fortunes,\lebnote{Hebrew “fortune”} and I will gather you from all the nations and from all the places to which I have driven you,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘and I will bring you back to the place from which I deported you.’
\verse Because you have said, ‘Adonai has raised up prophets for us in Babylon’—
\verse for thus says Adonai concerning the king who sits on the throne of David and concerning all the people who live in this city, your fellow kinsmen who did not go with you into the exile—
\verse thus says Adonai of hosts, ‘Look, I am going to send among them the sword, the famine, and the plague, and I will make them like rotten figs that cannot be eaten because of their bad quality.
\verse And I will pursue them with the sword, with the famine, and with the plague, and I will make them a terror to all the kingdoms of the earth, a curse, and a horror, and an object of hissing, and a disgrace among all the nations to which I have driven them,
\verse because they did not listen to my words,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘when I sent to them my servants the prophets, \textit{sending over and over again},\lebnote{Literally “doing early and sending”} and they would not listen,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse And you, hear the word of Adonai, all you exiles\lebnote{Hebrew “exile”} whom I sent away from Jerusalem to Babylon.
\verse Thus says Adonai of hosts, the God of Israel, concerning Ahab, the son of Kolaiah, and concerning Zedekiah, the son of Maaseiah, who are prophesying to you in my name a lie, ‘Look, I am going to give them into the hand of Nebuchadnezzar the king of Babylon, and he will strike them before your eyes.
\verse And a curse will be taken up because of them by all the exiles\lebnote{Hebrew “exile”} of Judah who are in Babylon, \textit{saying},\lebnote{Literally “to say”} “May Adonai make you like Zedekiah and like Ahab, whom the king of Babylon roasted in the fire,”
\verse because they have done a disgraceful thing in Israel, and they have committed adultery with the wives of their neighbors, and they have spoken words\lebnote{Hebrew “word”} in my name, lies\lebnote{Hebrew “lie”} that I have not commanded them, and I am he who knows, and I am a witness,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai.”
\verseWithHeading{Shemaiah’s Letter to Zephaniah the Priest} And to Shemaiah the Nehelamite you shall say, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai of hosts, the God of Israel, \textit{saying},\lebnote{Literally “to say”} ‘Because you yourself sent a letter\lebnote{Hebrew “letters”} in your name to all the people who are in Jerusalem, and to Zephaniah, the son of Maaseiah, the priest, and to all the priests, \textit{saying},\lebnote{Literally “to say”}
\verse “Adonai made you priest instead of Jehoiada the priest, so that there are overseers in the house of Adonai over any mad man who exhibits the behavior of a prophet, and you must put him into the stocks and into the neck iron,
\verse and so then why have you not rebuked Jeremiah the Anathothite who exhibits the behavior of a prophet for you?
\verse \textit{Because}\lebnote{Literally “Seeing that unto thus”} he has sent to us in Babylon, \textit{saying},\lebnote{Literally “to say”} ‘It will be a long time, build houses and live in them,\lebnote{Here the direct object is supplied from context in the English translation} and plant gardens and eat their fruit.’ ” ’ ”
\verse And Zephaniah the priest read this letter in the \textit{hearing}\lebnote{Literally “ears”} of Jeremiah the prophet.
\verseWithHeading{Jeremiah’s Response to Shemaiah’s Letter} And the word of Adonai came to Jeremiah, \textit{saying},\lebnote{Literally “to say”}
\verse “Send to all the exiles,\lebnote{Hebrew “exile”} \textit{saying},\lebnote{Literally “to say”} ‘Thus says Adonai concerning Shemaiah the Nehelamite: “Because Shemaiah prophesied to you, though I have not sent him, and he has made you trust in a lie,”
\verse \textit{therefore}\lebnote{Literally “to thus”} thus says Adonai, “Look, I am going to punish Shemaiah the Nehelamite and his offspring. There will not be for him a man who lives in the midst of this people, and he will not see the good that I am going to do to my people,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “For he has spoken rebellion against Adonai.” ’ ”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 30
\verseWithHeading{Israel and Judah Will Be Delivered from Their Distress} The word that came to Jeremiah from Adonai, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai, the God of Israel, \textit{saying},\lebnote{Literally “to say”} ‘Write for yourself all the words that I have spoken to you in a scroll.
\verse For look, days are coming,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘when I will restore the fortunes\lebnote{Hebrew “fortune”} of my people Israel and Judah,’ says Adonai, ‘and I will bring them back to the land that I gave to their ancestors\lebnote{Or “fathers”} and they will take possession of it.’ ”
\verse Now these are the words that Adonai spoke concerning Israel and concerning Judah.
\verse “For thus says Adonai:
\verse ‘We have heard a sound of trembling, 
terror, and there is no peace.
\verse Ask please and see whether a male can bear a child. 
Why do I see every strong man 
with his hands on his loins like a woman giving birth? 
And why are all their faces changed to paleness?
\verse Alas! For that day is great, 
\textit{there is none like it}.\lebnote{Literally “from none like it”} 
And it is a time of distress for Jacob, 
yet from it he will be delivered.
\verse And it will happen on that day,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai of hosts, 
‘I will break his yoke from your neck 
and your bonds I will tear to pieces. 
And strangers will no longer let him work as a slave.
\verse But they will serve Adonai their God, 
and David their king, whom I will raise up for them.
\verse But you must not fear, my servant Jacob,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
‘and you must not be dismayed, Israel, 
for look, I am going to save you from far, 
and your offspring from the land of their captivity. 
And Jacob will return, 
and he will be at rest, 
and he will be at ease, 
and there will be no one who makes him\lebnote{Here the direct object is supplied from context in the English translation} afraid.
\verse For thus says Adonai,
\verse ‘Your injury is incurable, 
your wound is overcome by sickness.
\verse There is no one who pleads your cause, 
for your boil there is no healing for you.
\verse All your lovers have forgotten you, 
they do not care for you. 
I have struck you with the blow of an enemy, 
the punishment of the cruel, 
because of the greatness of your guilt, 
your sins are numerous.
\verse Why do you cry because of your wound? 
Your pain is incurable. 
Because of the greatness of your guilt—your sins are vast— 
I have done these things to you.
\verse \textit{Therefore}\lebnote{Literally “To thus”} all those who devour you will be devoured, 
and all your foes, all of them will go into captivity, 
and those who plunder you will be as plunder, 
and all those who plunder you I will make as plunder.
\verse Thus says Adonai:
\verse ‘Look, I will restore the fortunes\lebnote{Hebrew “fortune”} of the tents of Jacob, 
and I will have compassion on his dwellings, 
and the city will be rebuilt upon its mound, 
and the citadel fortress will stand on its rightful site.
\verse And thanksgiving and the sound of merrymakers 
will come out from them, 
and I will make them numerous, 
and they will not be few. 
And I will make them honored, 
and they will not be lowly.
\verse And their children will be as of old, 
and their community will be established \textit{before me}.\lebnote{Literally “to the face of me”} 
And I will punish all their oppressors.
\verse And their noble will be from them, 
and their ruler will come out from their midst. 
And I will bring him near and he will approach me. 
For who is he that would pledge his heart to approach me?’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse ‘And you will be to me a people, 
and I will be to you God.’ ”
\verse Look, the storm of Adonai! 
Wrath has gone forth, 
a whirling tempest. 
It will whirl upon the head of the wicked.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 31
\verse “At that time,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “I will be the God of all the clans\lebnote{Or “families”} of Israel, 
and they will be to me a people.”
\verse Thus says Adonai:
\verse “The people of the survivors of the sword found grace in the desert;\lebnote{Or “wilderness”} 
Israel going to find their rest.”
\verse From afar Adonai appeared to me, saying, 
“I have loved you with an everlasting love. 
\textit{Therefore}\lebnote{Literally “Unto thus”} I have drawn you with loyal love.
\verse I will again build you, 
and you will be built, O virgin Israel. 
You will again adorn yourself with your tambourines, 
and you will go forth in the dancing in a ring of the merrymakers.
\verse You will again plant vineyards on the mountains of Samaria. 
The planters will plant, and they will enjoy it.\lebnote{Here the direct object is supplied from context in the English translation}
\verse For thus says Adonai,
\verse “Sing for joy for Jacob with gladness, 
and shout out for the head of the nations. 
Proclaim, praise, and say, 
‘Save, O Adonai, your people, the remnant of Israel.’
\verse Look, I am going to bring them from the land of the north, 
and I will gather them from the remotest part of the earth. 
Among them the blind, and the lame, 
those who are pregnant, and those who give birth, 
together, a great assembly, 
they will return here.
\verse With weeping they will come, 
and with pleas for mercy I will bring them; 
I will let them walk by streams of water in a straight path. 
They will not stumble in it, 
for I have become to Israel a father, 
and Ephraim, he is my firstborn.”
\verse Hear the word of Adonai, O nations, 
and declare in the coastlands from afar, and say, 
“The scatterer of Israel will gather him, 
and he will keep him as a shepherd his flock.
\verse For Adonai has ransomed Jacob, 
and he has redeemed him from the hand of the person \textit{stronger}\lebnote{Literally “strong”} than he.
\verse And they will come, 
and they will sing for joy on the height of Zion, 
and they will be radiant over the goodness of Adonai, 
over the grain, and over the wine, and over the olive oil, 
and over the young ones of the flock, and over the cattle. 
And their life will become like a well-watered garden, 
and they will never languish again.”
\verse “Then the young woman will rejoice in the dancing in a ring, 
and the young men and the old men together. 
And I will turn their mourning to jubilation, 
and I will comfort them, 
and I will gladden them from their sorrow.
\verse And I will saturate the appetite of the priests with fatness, 
and my people will be satisfied with my goodness,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse Thus says Adonai:
\verse “Restrain your voice from weeping, 
and your eyes from tears, 
for there is a reward for your work,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“and they will return from the land of the enemy.
\verse And there is hope for your future,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“and your children will return to their territory.
\verse Indeed I heard Ephraim pitying themselves, 
‘You disciplined me, and I was disciplined, 
like a calf that is not trained. 
Bring me back and let me return, 
for you are Adonai my God.
\verse For after my turning back I repented, 
and after coming to understand I struck my thigh. 
I was ashamed and also humiliated, 
because I bore the disgrace of my youth.’
\verse Is Ephraim my dear son, 
or the child of my delight? 
For \textit{as often as}\lebnote{Literally “from enough”} I have earnestly spoken against him, 
I still remember him. 
\textit{Therefore}\lebnote{Literally “Unto thus”} my bowels are turbulent for him, 
surely I will have compassion on him,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “Set up for yourself road markers, 
make for yourself signposts, 
set your \textit{mind}\lebnote{Literally “heart”} to the main road, 
the road by which you went. 
Return, O virgin of Israel, 
return to these your cities.
\verse Thus says Adonai of hosts, the God of Israel, “They will again say these words\lebnote{Hebrew “word”} in the land of Judah and in its towns at my restoring their fortunes,\lebnote{Hebrew “fortune”}
\verse ‘Adonai bless you, O settlement of righteousness, 
O hill of holiness.’
\verse And Judah and all of its towns together will live in it, 
farmers and those who travel with the flocks.\lebnote{Hebrew “flock”}
\verse For I will saturate the thirst of the weary, 
and every person who languishes I will replenish.”
\verseWithHeading{The New Covenant} “Look, the days are coming,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “and I will sow the house of Israel and the house of Judah with the seed of humankind, and with the seed of animals.\lebnote{Hebrew “animal”}
\verse \textit{And then}\lebnote{Literally “And it will happen”} as I have watched over them to pull up, and to tear down, and to annihilate, and to destroy, and to do evil, so I will watch over them to build and to plant,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “In those days they will say no longer, ‘\textit{Parents}\lebnote{Literally “Fathers”} have eaten unripe fruit, and the teeth of the children \textit{are set on edge}\lebnote{Literally “have become blunt”}.’
\verse \textit{But}\lebnote{Literally “But if”} each will die because of his iniquity, \textit{everyone}\lebnote{Literally “all of the humankind”} who eats the unripe fruit, their teeth will \textit{be set on edge}.\lebnote{Literally “become blunt”}
\verse Look, the days are coming,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “and I will \textit{make}\lebnote{Literally “cut”} a new covenant with the house of Israel and with the house of Judah,
\verse not like the covenant that I \textit{made}\lebnote{Literally “cut”} with their ancestors\lebnote{Or “fathers”} on the day of my grasping them\lebnote{Here the direct object is supplied from context in the English translation} by their hand, bringing them out from the land of Egypt, my covenant that they themselves broke, though I myself was a master over them,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “But this is the covenant that I will \textit{make}\lebnote{Literally “cut”} with the house of Israel after those days,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai: “I will put my law in their inward parts and on their hearts\lebnote{Hebrew “heart”} I will write it, and I will be to them God, and they themselves will be to me people.
\verse And they will no longer teach each one his neighbor, or each one his brother, \textit{saying},\lebnote{Literally “to say”} ‘Know Adonai,’ for all of them will know me, from their \textit{smallest}\lebnote{Literally “small”} and up to their \textit{greatest},”\lebnote{Literally “great”} \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “for I will forgive their iniquity and their sin I will no longer remember.”
\verse Thus says Adonai, who gives the sun for light by day, the regulations of the moon and the stars for light by night, who stirs up the sea and its waves roar—Adonai of hosts is his name.
\verse “If these rules would cease from \textit{before me},”\lebnote{Literally “to the face of me”} \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “also the offspring of Israel would cease from being a nation \textit{before me}\lebnote{Literally “to the face of me”} \textit{forever}.”\lebnote{Literally “all the days”}
\verse Thus says Adonai, “If the heavens above can be measured, and the foundations of the earth below can be explored, also I will reject all the offspring of Israel because of all that they have done,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “Look, days are coming,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “and the city will be rebuilt for Adonai, from the Tower of Hananel to the Corner Gate.
\verse And \textit{the measuring line}\lebnote{Literally “the string of the measurement”} will still go out immediately in front of it to the hill of Gareb, and it will turn to Goah.
\verse And the whole of the valley of the corpses, and the ashes, and all the cultivated fields up to the wadi of Kidron, up to the corner of the Gate of the Horses toward the east will be holy to Adonai. It will not be uprooted, and it will not be overthrown again \textit{forever}.”\lebnote{Literally “to eternity”}
\end{biblechapter}

\begin{biblechapter} % Jeremiah 32
\verseWithHeading{Jeremiah Buys a Field} The word that came to Jeremiah from Adonai in the tenth year of Zedekiah the king of Judah, that was the eighteenth year of Nebuchadnezzar.
\verse Now at that time the army of the king of Babylon was laying siege to Jerusalem and Jeremiah the prophet was confined in the courtyard of the guard that was in the palace of the king of Judah,
\verse where Zedekiah, the king of Judah, had confined him, \textit{saying},\lebnote{Literally “to say”} “Why are you prophesying, \textit{saying},\lebnote{Literally “to say”} ‘Thus says Adonai, “Look, I am going to give this city into the hand of the king of Babylon, and he will capture it,
\verse and Zedekiah, the king of Judah, will not escape from the hand of the Chaldeans, but surely he will be given into the hand of the king of Babylon, and \textit{he will speak face to face with him}\lebnote{Literally “his mouth shall speak with his mouth”} and \textit{he will see him eye to eye},\lebnote{Literally “his eyes shall see his eyes”}
\verse and to Babylon he will bring Zedekiah, and there he will be until my attending to him,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai. “If you fight against the Chaldeans, you will not be successful” ’?”
\verse And Jeremiah said, “The word of Adonai came to me, \textit{saying},\lebnote{Literally “to say”}
\verse ‘Look, Hanamel, the son of Shallum, your uncle, is going to come to you, \textit{saying},\lebnote{Literally “to say”} “Buy for yourself my field that is at Anathoth, for \textit{you have}\lebnote{Literally “to you is”} the right of redemption to buy it.” ’\lebnote{Here the direct object is supplied from context in the English translation}
\verse Then Hanamel, the son of my uncle, came to me, to the courtyard of the guard \textit{according to}\lebnote{Literally “as”} the word of Adonai, and he said to me, ‘Please buy my field that is at Anathoth, that is in the land of Benjamin, for to you is the claim of possession, and to you the redemption; buy it\lebnote{Here the direct object is supplied from context in the English translation} for yourself.’ Then I knew that this was the word of Adonai.
\verse And I bought the field from Hanamel, the son of my uncle, that was at Anathoth. And I weighed out to him the money, seventeen silver shekels.
\verse And I signed on the letter\lebnote{Or “scroll”} and sealed it,\lebnote{Here the direct object is supplied from context in the English translation} and I called witnesses as witness, and I weighed out the money on a set of scales.
\verse Then I took the deed\lebnote{Or “scroll”} of the purchase, the sealed copy containing the commandments\lebnote{Hebrew “commandment”} and the rules, together with the one that was open.
\verse And I gave the deed of the purchase to Baruch the son of Neriah, the son of Mahseiah, \textit{in the presence of}\lebnote{Literally “before the eyes of”} Hanamel, the son of my uncle, and \textit{in the presence of}\lebnote{Literally “before the eyes of”} the witnesses who signed the deed of the purchase, \textit{in the presence of}\lebnote{Literally “before the eyes of”} the Judeans\lebnote{Hebrew “Judean”} who were sitting in the courtyard of the guard.
\verse And I commanded Baruch \textit{in their presence},\lebnote{Literally “before their eyes”} \textit{saying},\lebnote{Literally “to say”}
\verse ‘Thus says Adonai of hosts, the God of Israel, “Take these deeds,\lebnote{Or “scrolls”} this deed of the purchase, the sealed one, and this opened deed, and you must put them in an \textit{earthenware jar}\lebnote{Literally “jar of earthenware”} so that they may be kept preserved many days.”
\verse For thus says Adonai of hosts, the God of Israel: “Houses and fields and vineyards will again be bought in this land.” ’
\verse And I prayed to Adonai—after giving my deed of the purchase to Baruch the son of Neriah—\textit{saying},\lebnote{Literally “to say”}
\verse ‘Ah Lord Adonai! Look, you made the heavens and the earth by your great power and by your outstretched arm. \textit{Nothing}\lebnote{Literally “Not all a thing”} is too difficult for you,
\verse showing loyal love to the thousands, and repaying the guilt of \textit{parents}\lebnote{Literally “fathers”} into the laps\lebnote{Hebrew “lap”} of their children after them. Powerful, great, mighty, his name is Adonai of hosts,
\verse great in counsel and great in deed, whose eyes are opened to all the ways of the children of humankind, to give to each one according to his ways and according to the fruit of his deeds,
\verse who accomplished signs and wonders in the land of Egypt, up to this day, and in Israel, and among humankind, and you have made for yourself a name, as it is this day.
\verse And you brought out your people Israel from the land of Egypt with signs, and with wonders, and with a strong hand, and with an arm stretched out, and with great terror.
\verse And you gave to them this land, which you swore to their ancestors\lebnote{Or “fathers”} to give to them, a land flowing with milk and honey.
\verse And they entered and took possession of it, but they did not listen to your voice, and they did not follow your law. They did nothing of all that you commanded to them to do, and you caused to happen to them all this disaster.
\verse Look, the siege ramps have come up to the city to capture it, and the city has been given into the hand of the Chaldeans who are fighting against it, because of the sword, the famine, and the plague, and what you spoke happened, and look, you are seeing it.\lebnote{Here the direct object is supplied from context in the English translation}
\verse Yet you have said to me, Lord Adonai, “Buy for yourself the field with the money, and call witnesses as witness,” though the city has been given into the hand of the Chaldeans.’ ”
\verse And the word of Adonai came to Jeremiah, \textit{saying},\lebnote{Literally “to say”}
\verse “Look, I am Adonai, the God of all flesh; is \textit{anything}\lebnote{Literally “all thing”} too difficult for me?”
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai: “Look, I am going to give this city into the hand of the Chaldeans, and into the hand of Nebuchadnezzar, the king of Babylon, and he will capture it.
\verse And the Chaldeans, who are fighting against this city, will come and set this city on fire, and they will burn it, and the houses where they have made smoke offerings on their roofs to Baal, and where they have devoted libations to other gods, in order to provoke me to anger.
\verse For the people\lebnote{Or “sons/children”} of Israel and the people\lebnote{Or “sons/children”} of Judah were doing only evil in my eyes from their youth, for the people\lebnote{Or “sons/children”} of Israel were only provoking me to anger by the work of their hands,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “For this city has been for me a cause of my anger and of my wrath from the day that they built it even until this day, so I will remove it from my \textit{sight},\lebnote{Literally “face”}
\verse because of all the evil of the people\lebnote{Or “sons/children”} of Israel and the people\lebnote{Or “sons/children”} of Judah that they did to provoke me to anger—they, their kings, their officials, their priests, and their prophets, and the men\lebnote{Hebrew “man”} of Judah, and the inhabitants of Jerusalem.
\verse And they have turned to me their backs\lebnote{Hebrew “back”} and not their faces, though \textit{I have taught them over and over again},\lebnote{Literally “I have taught them doing early and teaching”} but they were not listening to accept discipline.
\verse And they set up their abominations in the house that is called by my name, to defile it.
\verse And they built the high places of Baal that are in the Valley of Ben Hinnom to present as offerings their sons and their daughters to Molech, which I had not commanded them, and it had not come to my \textit{mind}\lebnote{Literally “heart”} to do this detestable thing in order to cause Judah to sin.”
\verse “So now \textit{therefore},\lebnote{Literally “to thus”} thus says Adonai, the God of Israel, concerning this city, of which you are saying, ‘It will be given into the hand of the king of Babylon by the sword, and by the famine, and by the plague’:
\verse Look, I am going to gather them from all the lands to which I driven them in my anger, and in my fury, and in great wrath, and I will bring them back to this place, and I will cause them to dwell \textit{in safety}.\lebnote{Literally “with confidence”}
\verse And they will be for me a people, and I will be for them God.
\verse And I will give to them one heart and one way, to revere me \textit{forever},\lebnote{Literally “all the days”} for good to them, and to their children after them.
\verse And I will \textit{make}\lebnote{Literally “cut”} with them \textit{an everlasting covenant}\lebnote{Literally “a covenant of eternity”} that I will not turn away from them, my doing good to them, and my reverence I will put in their hearts\lebnote{Hebrew “heart”} so that they will not turn aside from me.
\verse And I will rejoice over them to do good to them, and I will plant them in this land in faithfulness with all my heart, and with all my inner self.”\lebnote{Or “soul”}
\verse “For thus says Adonai, ‘Just as I have brought to this people all this great disaster, so I will bring upon them all the good that I promise to them.
\verse And the fields\lebnote{Hebrew “field”} will be bought in this land of which you are saying, “It is a desolation, \textit{without humankind or animals}.\lebnote{Literally “from there is not humankind and animal”} It has been given into the hand of the Chaldeans.”
\verse They will buy fields with money, and they will sign the deeds,\lebnote{Hebrew “scroll”} and they will seal them,\lebnote{Here the direct object is supplied from context in the English translation} and they will call witnesses as witness in the land of Benjamin, and in the surroundings of Jerusalem, and in the cities of Judah, and in the cities of the hill country, and in the cities of the Shephelah, and in the cities of the Negev, for I will restore their fortunes,’\lebnote{Hebrew “fortune”} \textit{declares}\lebnote{Literally “a declaration of”} Adonai.”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 33
\verseWithHeading{The Fortunes of Judah and Israel Will Be Restored} And the word of Adonai came to Jeremiah a second time while he was still held back in the courtyard of the guard, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai who made the earth,\lebnote{Here the direct object is supplied from context in the English translation} Adonai who formed it to establish it, Adonai is his name:
\verse ‘Call to me, and I will answer you, and I will tell you great things and inaccessible things that you have not known.’
\verse For thus says Adonai, the God of Israel, concerning the houses of this city and concerning the houses of the kings of Judah, that were torn down to make a defense against the siege ramps and against the sword:
\verse ‘They are coming to fight against the Chaldeans, and to fill them with the corpses of the people whom I will strike in my anger and in my wrath, for whom I have hidden my face from this city because of all their wickedness.
\verse Look, I am going to bring healing and health to it, and I will heal them, and I will reveal to them abundance of peace and reliability.
\verse And I will restore the fortunes\lebnote{Hebrew “fortune”} of Judah and the fortunes\lebnote{Hebrew “fortune”} of Israel, and I will rebuild them as in the beginning.
\verse And I will cleanse them from all their guilt that they sinned against me, and I will forgive all their iniquities that they sinned against me, and that they rebelled against me.
\verse And it will be to me a name of jubilation, a praise, and a glory before all the nations of the earth, who will hear all the good that I will do for them, and they will fear,\lebnote{Or “tremble”} and they will tremble because of all the good, and because of all the prosperity that I will provide for it.’
\verse “Thus says Adonai: ‘Again will be heard in this place, of which you are saying, “It is a waste, \textit{without}\lebnote{Literally “from there is not”} people and \textit{without}\lebnote{Literally “from there is not”} animals,”\lebnote{Hebrew “animal”} in the towns of Judah and in the streets of Jerusalem that are desolate, \textit{without}\lebnote{Literally “from there is not”} people and \textit{without}\lebnote{Literally “from there is not”} inhabitants\lebnote{Hebrew “inhabitant”} and \textit{without}\lebnote{Literally “from there is not”} animals,\lebnote{Hebrew “animal”}
\verse the voice of jubilation, and the voice of joy, the voice of the bridegroom, and the voice of the bride, the voices\lebnote{Hebrew “voice”} of those who say, “Praise Adonai of hosts, for Adonai is good, for his loyal love is \textit{forever},”\lebnote{Literally “to eternity”} the voices of those who bring thank offerings\lebnote{Hebrew “thank offering”} to the house of Adonai, for I will restore the fortunes\lebnote{Hebrew “fortune”} of the land as in the beginning,’ says Adonai.
\verse Thus says Adonai of hosts: ‘Again there will be in this waste place, \textit{without}\lebnote{Literally “from there is not”} people and animals,\lebnote{Hebrew “animal”} and in all its towns, pasture for shepherds allowing their flock to lie down.
\verse In the towns of the hill country, in the towns of the Shephelah, and in the towns of the Negev, and in the land of Benjamin, and in the surroundings of Jerusalem, and in the towns of Judah, flocks will again pass under the hands of the counter,’ says Adonai.
\verse ‘Look, days are coming,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘and I will fulfill the good promise\lebnote{Or “word”} that I promised to the house of Israel and to the house of Judah.
\verse In those days and in that time I will make a branch of righteousness sprout for David, and he will execute justice and righteousness in the land.
\verse In those days Judah will be saved, and Jerusalem will dwell \textit{safely},\lebnote{Literally “with confidence”} and this is what they shall call it: “Adonai is our righteousness.” ’
\verse For thus says Adonai: ‘A man who sits on the throne of the house of Israel will not be cut off for David.
\verse And for the priests, a man from the Levites who offers burnt offerings\lebnote{Hebrew “burnt offering”} and who burns grain offerings\lebnote{Hebrew “grain offering”} and who brings sacrifices\lebnote{Hebrew “sacrifice”} will not be cut off \textit{before me}\lebnote{Literally “to the face of me”} \textit{forever}.’ ”\lebnote{Literally “all the days”}
\verse And the word of Adonai came to Jeremiah, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai: ‘If you could break my covenant with the day, and my covenant with the night, so that day and night would not come at their time,
\verse then my covenant could be broken with David my servant, from \textit{having}\lebnote{Literally “being for him”} a son who could rule on his throne, and with the Levites, the priests, my ministers.
\verse As the host of heaven cannot be counted, and the sand of the sea cannot be measured, so I will make numerous the offspring of David my servant and the Levites who minister to me.’ ”
\verse And the word of Adonai came to Jeremiah, \textit{saying},\lebnote{Literally “to say”}
\verse “Have you not seen how these people speak, \textit{saying},\lebnote{Literally “to say”} ‘The two clans\lebnote{Or “families”} whom Adonai chose, he has also rejected them’? Thus they spurn my people from being a nation \textit{before them}\lebnote{Literally “to the face of them”} any longer.
\verse Thus says Adonai: ‘If my covenant with day and with night, the regulations of heaven and earth, I had not established,
\verse then the offspring of Jacob and David my servant I would reject, from choosing rulers from his offspring over the offspring of Abraham, Isaac, and Jacob; for I will restore their fortunes\lebnote{Hebrew “fortune”} and I will have compassion on them.’ ”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 34
\verseWithHeading{A Message for Zedekiah} The word that came to Jeremiah from Adonai, when Nebuchadnezzar the king of Babylon, and all his army, with all the kingdoms of the earth under the dominion of his hand, and all the peoples were fighting against Jerusalem and against all its cities, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai, the God of Israel: ‘Go and say to Zedekiah the king of Judah, now you must say to him, “Thus says Adonai: ‘Look, I am going to give this city into the hand of the king of Babylon, and he will burn it with fire.
\verse And you will not escape from his hand, but surely you will be captured, and into his hand you will be given, and \textit{you will see the king of Babylon eye to eye},\lebnote{Literally “your eyes shall see with the eyes of the king of Babylon”} and \textit{you will speak face to face with him},\lebnote{Literally “his mouth shall speak with your mouth”} and to Babylon you will go.’
\verse However, hear the word of Adonai, O Zedekiah, the king of Judah. Thus says Adonai concerning you: ‘You will not die by the sword.
\verse In peace you will die, and as there was burning for your ancestors,\lebnote{Or “fathers”} the former kings who were \textit{before you},\lebnote{Literally “to the face of you”} so they will burn for you, and they will lament for you, “Alas, lord!” For the word I have spoken,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai.” ’ ”
\verse Then Jeremiah the prophet spoke to Zedekiah the king of Judah all these words in Jerusalem
\verse when the army of the king of Babylon was fighting against Jerusalem and against all the cities of Judah that were left over—Lachish and Azekah, for these remained among the cities of Judah, the cities of fortification.
\verseWithHeading{Zedekiah’s Covenant Concerning Slaves} The word that came to Jeremiah from Adonai after king Zedekiah \textit{made}\lebnote{Literally “cut”} a covenant with all the people who were in Jerusalem to proclaim release to them,
\verse to let go each one his male slave and each one his female slave, the Hebrew and the free Hebrew, so that no one among the Judeans\lebnote{Hebrew “Judean”} should enslave his fellow countryman.
\verse And all the officials and all the people obeyed, who had entered into the covenant to let go each one his male slave and each one his female slave, not enslaving them again, and they obeyed and they let them\lebnote{Here the direct object is supplied from context in the English translation} go.
\verse But \textit{afterward}\lebnote{Literally “after thus”} they turned back and they brought back the male slaves and the female slaves whom they had let go free, and they subdued them as male slaves and female slaves.
\verse And the word of Adonai came to Jeremiah from Adonai, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai, the God of Israel: ‘I \textit{made}\lebnote{Literally “cut”} a covenant with your ancestors\lebnote{Or “fathers”} on the day of my bringing them out from the land of Egypt, from the house of slaves, \textit{saying},\lebnote{Literally “to say”}
\verse “At the end of seven years you must let go each one his fellow countryman, the Hebrew who has been sold to you and who has served you six years, and you must let him go free from you.” But your ancestors\lebnote{Or “fathers”} did not listen to me, and they did not incline their ears.\lebnote{Hebrew “ear”}
\verse And you turned back \textit{recently}\lebnote{Literally “the day”} and you did right in my eyes, to proclaim release each one to his neighbor, and you \textit{made}\lebnote{Literally “cut”} a covenant \textit{before me}\lebnote{Literally “to the face of me”} in the house that is called \textit{by}\lebnote{Literally “on it”} my name.
\verse But you turned back and you profaned my name when you brought back each one his male slave and each one his female slave, whom you had let go free according to their desire, and you subdued them to be to you as male slaves and as female slaves.’
\verse “\textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai, ‘You have not listened to me to proclaim release each one to his fellow countryman and each one to his neighbor. Look, I am going to proclaim to you a release,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘to the sword, to the plague, and to the famine, and I will make you a terror to all the kingdoms of the earth.
\verse And I will make the men who transgressed my covenant, who have not kept the words of the covenant that they \textit{made}\lebnote{Literally “cut”} \textit{before me},\lebnote{Literally “to the face of me”} like the calf which they cut in two and they passed between its parts—
\verse the officials of Judah, and the officials of Jerusalem, the eunuchs, and the priests, and all the people of the land who passed between the parts of the calf—
\verse and I will give them into the hand of their enemies, and into the hand of the seekers of their lives, and their dead bodies\lebnote{Hebrew “dead body”} will become as food for the birds\lebnote{Hebrew “bird”} of the heavens\lebnote{Or “sky”} and for the animals\lebnote{Hebrew “animal”} of the earth.
\verse And Zedekiah the king of Judah and his officials I will give into the hand of their enemies, and into the hand of the seekers of their lives, and into the hand of the army of the king of Babylon, who retreated from you.
\verse Look, I am going to command,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘and I will bring them back to this city, and they will fight against it, and they will capture it, and they will burn it with fire, and the towns of Judah I will make a desolation \textit{without}\lebnote{Literally “from there is not”} an inhabitant.’ ”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 35
\verseWithHeading{The Faithfulness of the Rechabites} The word that came to Jeremiah from Adonai in the days of Jehoiakim, the son of Josiah, the king of Judah, \textit{saying},\lebnote{Literally “to say”}
\verse “Go to the house of the Rechabites, and speak with them, and bring them to the house of Adonai into one of the chambers, and give them wine to drink.”
\verse So I took Jaazaniah, the son of Jeremiah, the son of Habazziniah, and his brothers, and all his sons, and the whole of the house of the Rechabites,
\verse and I brought them to the house of Adonai, into the chamber of the sons of Hanan, the son of Igdaliah, the man of God, which was beside the chamber of the officials, which was above the chamber of Maaseiah, the son of Shallum, the keeper of the threshold.
\verse Then I \textit{set}\lebnote{Literally “gave”} \textit{before}\lebnote{Literally “to the face of”} the house of the Rechabites bowls full of wine and cups, and I said to them, “Drink wine.”
\verse But they answered, “We will not drink wine, for Jonadab, the son of Rechab, our ancestor,\lebnote{Or “father”} commanded us, \textit{saying},\lebnote{Literally “to say”} ‘You shall not drink wine, you or your children, \textit{forever}.\lebnote{Literally “until eternity”}
\verse And you shall not build a house, and you shall not sow seed, and you shall not plant a vineyard, and it shall not be for you, but in tents you shall live all your days, so that you may live many days on the surface of the land where you are dwelling as aliens.’
\verse And we obeyed the voice of Jonadab, the son of Rechab, our ancestor,\lebnote{Or “father”} in all that he commanded us, not to drink wine all our days, ourselves, our wives, our sons, or our daughters,
\verse and not to build houses for our living, and there is not vineyard, or field, or seed for us.
\verse But we have lived in tents, and we have obeyed, and we have done all that Jehonadab, our ancestor,\lebnote{Or “father”} commanded us.
\verse \textit{But then}\lebnote{Literally “And it was”} at the coming up against the land of Nebuchadnezzar the king of Babylon, then we said, ‘Come and let us go to Jerusalem \textit{before}\lebnote{“from the face of”} the army of the Chaldeans, and \textit{before}\lebnote{“from the face of”} the army of the Arameans.’\lebnote{Hebrew “Aramean”} That is why we are living in Jerusalem.”
\verse And the word of Adonai came to Jeremiah, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai of hosts, the God of Israel: ‘Go and say to the people of Judah and to the inhabitants of Jerusalem, “Can you not learn a lesson to listen to my words?” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “The words of Jonadab, the son of Rechab, that he commanded his descendants\lebnote{Or “children”} to not drink, have been carried out, and they have not drunk until this day, for they have obeyed the command of their ancestor.\lebnote{Or “father”} But \textit{I have spoken to you over and over again},\lebnote{Literally “I have spoken to you doing early and speaking”} and you have not listened to me.
\verse And I have sent to you all my servants the prophets, \textit{I have sent them over and over again},\lebnote{Literally “doing early and sending”} \textit{saying},\lebnote{Literally “to say”} ‘Turn back please each one from his evil way, and amend your deeds, and you must not go after other gods to serve them, \textit{so that you may live}\lebnote{Literally “and live”} on the land that I gave to you and to your ancestors.’\lebnote{Or “fathers”} But you did not incline your ear and you did not listen to me.
\verse For the descendants\lebnote{Or “children”} of Jonadab, the son of Rechab, have carried out the command of their ancestor\lebnote{Or “father”} that he commanded them, but this people did not listen to me.”
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai, the God of hosts, the God of Israel, “Look, I am going to bring to Judah and to all the inhabitants of Jerusalem every disaster that I have spoken against them, because I have spoken to them and they have not listened, and I have called to them and they have not answered.” ’ ”
\verse But to the house of the Rechabites Jeremiah said, “Thus says Adonai of hosts, the God of Israel, ‘Because you have listened to the command of Jonadab your ancestor,\lebnote{Or “father”} and you have kept all his commands, and you have done all that he commanded you,’
\verse \textit{therefore}\lebnote{Literally “to thus”} thus says Adonai of hosts, the God of Israel: ‘A man will not be cut off for Jonadab, the son of Rechab, to stand \textit{before me}\lebnote{Literally “to the face of me”} \textit{always}.’ ”\lebnote{Literally “all the days”}
\end{biblechapter}

\begin{biblechapter} % Jeremiah 36
\verseWithHeading{Jehoiakim Burns the Scroll of Jeremiah’s Prophecies} \textit{And then}\lebnote{Literally “And it was”} in the fourth year of Jehoiakim the son of Josiah, the king of Judah, this word came to Jeremiah from Adonai, \textit{saying},\lebnote{Literally “to say”}
\verse “Take for yourself \textit{a scroll}\lebnote{Literally “a scroll of a scroll”} and you must write on it all the words that I have spoken to you against Israel, and against Judah, and against all the nations, from the day that I spoke to you, from the days of Josiah, and until this day.
\verse Perhaps when the house of Judah hears all the disasters\lebnote{Hebrew “disaster”} that I am planning to do to them, then they may turn back each one from his evil way, and I will forgive their guilt and their sin.”
\verse Then Jeremiah called Baruch the son of Neriah, and Baruch wrote from the mouth of Jeremiah all the words of Adonai that he had spoken to him on \textit{a scroll}.\lebnote{Literally “a scroll of a scroll”}
\verse And Jeremiah instructed Baruch, \textit{saying},\lebnote{Literally “to say”} “I am held back, I am not able to enter the temple of Adonai.
\verse So you must go and you shall read aloud from the scroll that you have written from my mouth the words of Adonai in the hearing of the people in the temple of Adonai on a day of fast, and also you shall read aloud in the hearing of all those of Judah who came from their towns.
\verse Perhaps their plea will fall \textit{before}\lebnote{Literally “to the face of”} Adonai and each one will turn away from his evil way, for great is the anger and wrath that Adonai pronounced against this people.”
\verse And Baruch the son of Neriah did all that Jeremiah the prophet instructed him, to read aloud from the scroll the words of Adonai in the temple of Adonai.
\verse \textit{And then}\lebnote{Literally “And it was”} in the fifth year of Jehoiakim, the son of Josiah, the king of Judah, in the ninth month, all the people in Jerusalem and all the people who came from the towns of Judah to Jerusalem proclaimed a fast \textit{before}\lebnote{Literally “to the face of”} Adonai.
\verse Then Baruch read aloud from the scroll the words of Jeremiah in the temple of Adonai, in the chamber of Gemariah, the son of Shaphan, the secretary, in the upper courtyard at the entrance of the New Gate of the temple of Adonai in the hearing of all the people.
\verse When Micaiah, the son of Gemariah, the son of Shaphan, heard all the words of Adonai from the scroll,
\verse he went down to the house of the king, to the chamber of the secretary, and look, all the officials were sitting there: Elishama the secretary, and Delaiah, the son of Shemaiah, and Elnathan, the son of Achbor, and Gemariah, the son of Shaphan, and Zedekiah, the son of Hananiah, and all the other officials.
\verse And Micaiah told them all the words that he had heard at the reading aloud of Baruch from the scroll in the hearing of the people.
\verse Then all the officials sent Jehudi, the son of Nethaniah, the son of Shelemiah, the son of Cushi, to Baruch, \textit{saying},\lebnote{Literally “to say”} “The scroll that you read aloud from in the hearing of the people, take it in your hand and come.” And Baruch the son of Neriah took the scroll in his hand and he came to them.
\verse And they said to him, “Sit please and read it aloud in our hearing.” So Baruch read aloud in their hearing.
\verse \textit{And then},\lebnote{Literally “And it was”} the moment of their hearing all the words, \textit{they turned to one another in alarm}\lebnote{Literally “they approached in trepidation each one to his neighbor”} and they said to Baruch, “We must certainly report all these words to the king!”
\verse Then they asked Baruch, \textit{saying},\lebnote{Literally “to say”} “Tell us please, how did you write all these words, from his mouth?”
\verse And Baruch said to them, “From his mouth. He dictated to me all these words and I was writing on the scroll with the ink.”
\verse Then the officials said to Baruch, “Go, hide yourself, you and Jeremiah, and let not a man know where you are.”
\verse And they went to the king, to the courtyard, and they \textit{put}\lebnote{Literally “gave”} the scroll for safe-keeping in the chamber of Elishama the secretary, and they reported all the words in the hearing of the king.
\verse Then the king sent Jehudi to take the scroll, and he took it from the chamber of Elishama the secretary, and Jehudi read it aloud in the hearing of the king, and in the hearing of all the officials who stood next to the king.
\verse Now the king was sitting in the quarters of the winter in the ninth month, and a fire-pot was burning \textit{before}\lebnote{Literally “to the face of”} him.
\verse \textit{And then},\lebnote{Literally “and it was”} as Jehudi read three or four columns, he would cut it up in pieces with the knife of the scribe, and he would throw it\lebnote{Here the direct object is supplied from context in the English translation} into the fire that was in the fire-pot until the whole of the scroll was consumed in the fire that was in the fire-pot.
\verse And the king and any of his servants who heard all these words were not startled, and they did not tear their garments.
\verse And even when Elnathan, and Delaiah, and Gemariah urged the king not to burn the scroll, he would not listen to them.
\verse And the king commanded Jerahmeel, the son of the king, and Seraiah, the son of Azriel, and Shelemiah, the son of Abdeel, to arrest Baruch the secretary and Jeremiah the prophet, but Adonai hid them.
\verse And the word of Adonai came to Jeremiah after the king burned the scroll and the words that Baruch wrote from the mouth of Jeremiah, \textit{saying},\lebnote{Literally “to say”}
\verse “\textit{Take again}\lebnote{Literally “Turn back take”} for yourself another scroll and write on it all the former words that were in the first scroll which Jehoiakim, the king of Judah, has burned.
\verse And concerning Jehoiakim, the king of Judah, you shall say, ‘Thus says Adonai, “You have burned this scroll, \textit{saying},\lebnote{Literally “to say”} ‘Why have you written in it, \textit{saying},\lebnote{Literally “to say”} “The king of Babylon will certainly come and he will destroy this land, and he will cause to disappear from it humankind and animals”?’ ”\lebnote{Hebrew “animal”}
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai concerning Jehoiakim, the king of Judah, “There will not be for him one who sits on the throne of David. And his dead body will be thrown out to the heat in the day and to the frost in the night.
\verse And I will punish him, and his offspring, and his servants for their guilt, and I will bring on them, and on the inhabitants of Jerusalem, and on the people of Judah all the disaster with which I have threatened them, but they would not listen.” ’ ”
\verse Then Jeremiah took another scroll and gave it to Baruch the son of Neriah, the secretary, and he wrote on it from the mouth of Jeremiah all the words of the scroll that Jehoiakim, the king of Judah, had burned in the fire, and furthermore was added to them many words like these.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 37
\verseWithHeading{The Babylonians Withdraw to Meet a Threat from Egypt} And king Zedekiah, the son of Josiah, whom Nebuchadnezzar, the king of Babylon, had made king, reigned as king in the land of Judah in place of Coniah, the son of Jehoiakim.
\verse But he and his servants and the people of the land did not listen to the words of Adonai that he spoke by the hand of Jeremiah the prophet.
\verse And king Zedekiah sent Jehucal the son of Shelemiah, and Zephaniah the son of Maaseiah, the priest, to Jeremiah the prophet, \textit{saying},\lebnote{Literally “to say”} “Please pray for us to Adonai our God.”
\verse Now Jeremiah was coming and going out in the midst of the people and they had not put him in the house of imprisonment.
\verse And the army of Pharaoh had come out from Egypt, and the Chaldeans, who were laying siege to Jerusalem, heard their report and they withdrew from Jerusalem.
\verse And the word of Adonai came to Jeremiah the prophet, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai, the God of Israel, ‘This is what you shall say to the king of Judah, who sent you to me to inquire of me, “Look, the army of Pharaoh, which set out to help you, is going to return to his land Egypt.
\verse And the Chaldeans will return, and they will fight against this city, and they will capture it, and they will burn it with fire.” ’
\verse Thus says Adonai: ‘You must not deceive yourselves, \textit{saying},\lebnote{Literally “to say”} “Surely the Chaldeans will go from us,” for they will not go.
\verse For even if you struck the whole army of the Chaldeans who are fighting against you, and only men pierced through remained among them, each one in his tent, they would rise up and they would burn this city with fire.’ ”
\verseWithHeading{Jeremiah is Arrested and Put in Prison} \textit{And then},\lebnote{Literally “And it will happen”} at the withdrawing of the army of the Chaldeans from Jerusalem \textit{before}\lebnote{Literally “from the face of”} the army of Pharaoh,
\verse Jeremiah set out from Jerusalem to go to the land of Benjamin to receive a portion from there in the midst of the people.
\verse \textit{And when}\lebnote{Literally “And it was”} he was at the Gate of Benjamin, there was \textit{a sentry on duty}\lebnote{Literally “an owner of supervision”} whose name was Irijah, the son of Shelemiah, the son of Hananiah, and he seized Jeremiah the prophet, \textit{saying},\lebnote{Literally “to say”} “You are deserting to the Chaldeans!”
\verse And Jeremiah said, “That is a lie! I am not deserting to the Chaldeans.” But he would not listen to him. So Irijah seized Jeremiah and brought him to the officials.
\verse And the officials were angry at Jeremiah and they struck him. And they put him in \textit{prison},\lebnote{Literally “the house of the fetters”} in the house of Jonathan the secretary, for they had converted it to the \textit{prison}\lebnote{Literally “house of the imprisonment”}.
\verse So Jeremiah came to the \textit{dungeon}\lebnote{Literally “house of the cistern”}, even to the vaulted cells, and Jeremiah stayed there many days.
\verse Then King Zedekiah sent and fetched him. And the king questioned him in secrecy in his house, and he said, “Is there a word from Adonai?” And Jeremiah said, “There is.” And he said, “You will be given into the hand of the king of Babylon.”
\verse Then Jeremiah said to King Zedekiah, “What did I do wrong to you, or to your servants, or to this people, that you have put me in \textit{prison}?\lebnote{Literally “the house of the imprisonment”}
\verse And where are your prophets who prophesied to you, \textit{saying},\lebnote{Literally “to say”} ‘The king of Babylon will not come against you and against this land’?
\verse Now then, my lord the king, please let my plea fall \textit{before you}.\lebnote{Literally “to the face of you”} You must not send me back to the house of Jonathan the secretary, so that I will not die there.”
\verse So King Zedekiah commanded, and they handed Jeremiah over in the courtyard of the guard, and they gave to him a round loaf of bread from the street of the bakers \textit{every day}\lebnote{Literally “for the day”} until the finishing of all the bread from the city. And Jeremiah stayed in the courtyard of the guard.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 38
\verseWithHeading{Jeremiah is Thrown into the Pit of Malchiah} Now Shephatiah the son of Mattan, and Gedaliah the son of Pashhur, and Jehucal the son of Shelemiah, and Pashhur the son of Malchiah heard the words that Jeremiah was saying to all the people, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai, ‘The one who stays in this city will die by the sword, by the famine, and by the plague. But the one who goes out to the Chaldeans will live. And his life will be for him as booty, and he will live.’
\verse Thus says Adonai, ‘Surely this city will be given into the hand of the army of the king of Babylon, and he will capture it.’ ”
\verse Then the officials said to the king, “Please, this man must be killed, \textit{because}\lebnote{Literally “unto thus”} he is making slack the hands of \textit{the soldiers}\lebnote{Literally “the men of the war”} who are left in this city, and the hands of all the people, by speaking to them words like these, for this man is not seeking for welfare\lebnote{Or “peace”} to this people, \textit{but only}\lebnote{Literally “but if”} for harm.”
\verse And Zedekiah the king said, “Look, he is in your hand, for the king is not able to do a thing against you.”
\verse So they took Jeremiah and threw him into the pit\lebnote{Or “cistern”} of Malchiah, the son of the king, which was in the courtyard of the guard. And they let Jeremiah down by ropes. Now in the pit\lebnote{Or “cistern”} there was no water,\lebnote{Hebrew “waters”} \textit{but only}\lebnote{Literally “but if”} mud, and Jeremiah sank in the mud.
\verse When Ebed-melech the Cushite, \textit{a eunuch}\lebnote{Literally “a man of a eunuch”} who was in the house of the king, heard that they had put Jeremiah into the pit\lebnote{Or “cistern”}—now the king was sitting at the Gate of Benjamin—
\verse Ebed-melech went out from the house of the king and spoke to the king, \textit{saying},\lebnote{Literally “to say”}
\verse “My lord the king, these men have done evil in all that they have done to Jeremiah the prophet, in that they have thrown him into the pit,\lebnote{Or “cistern”} and he will die \textit{there}\lebnote{Literally “in it”} \textit{because of}\lebnote{Literally “from the face of”} starvation, for there is no longer any bread in the city.”
\verse Then the king commanded Ebed-melech the Cushite, \textit{saying},\lebnote{Literally “to say”} “Take \textit{with you}\lebnote{Literally “in your hand”} these thirty men and pull Jeremiah the prophet up from the pit\lebnote{Or “cistern”} \textit{before}\lebnote{Literally “at not yet”} he dies.”
\verse So Ebed-melech took the men \textit{with him}\lebnote{Literally “in his hand”} and went to the palace of the king, to a place beneath the storehouse, and he took from there \textit{rags}\lebnote{Literally “the rags of rags”} and \textit{worn-out clothes}.\lebnote{Literally “rags of ragged clothes”} And he let them down into the pit\lebnote{Or “cistern”} by ropes to Jeremiah.
\verse Then Ebed-melech the Cushite said to Jeremiah, “Please put \textit{the rags}\lebnote{Literally “rags of the rags”} and \textit{worn-out clothes}\lebnote{Literally “ragged clothes”} under the joints of your arms under the ropes,” and Jeremiah did so.
\verse And they pulled Jeremiah by the ropes and brought him up from the pit.\lebnote{Or “cistern”} And Jeremiah stayed in the courtyard of the guard.
\verse And Zedekiah the king sent and made someone bring Jeremiah the prophet to him, to the third entrance that was at the \textit{temple}\lebnote{Literally “house”} of Adonai. And the king said to Jeremiah, “I am asking you \textit{something},\lebnote{Literally “a thing”} you must not hide \textit{anything}\lebnote{Literally “a thing”} from me.”
\verse And Jeremiah said to Zedekiah, “If I tell you, will you not surely kill me? \textit{Besides},\lebnote{Literally “And”} if I advise you, you will not listen to me.
\verse So Zedekiah the king swore to Jeremiah in secret, \textit{saying},\lebnote{Literally “to say”} “\textit{As Adonai lives},\lebnote{Literally “Life of Adonai”} who has made for us this life, I will not kill you and I will not give you into the hand of these men who are seeking your life.”
\verse Then Jeremiah said to Zedekiah, “Thus says Adonai, the God of hosts, the God of Israel, ‘If only you will go surrender to the officials of the king of Babylon, then \textit{you}\lebnote{Literally “your life” or “your soul”} will live, and this city will not be burned with fire, and you will live, you and your house.
\verse But if you do not go surrender to the officials of the king of Babylon, then this city will be given into the hand of the Chaldeans, and they will burn it with fire, and you will not escape from their hand.’ ”
\verse And Zedekiah the king said to Jeremiah, “I am afraid of the Judeans who have deserted to the Chaldeans, lest they give me into their hand and they abuse me.”
\verse And Jeremiah said, “They will not give you over. Please obey the voice of Adonai with regard to what I am saying to you, and it will go well with you, and \textit{you}\lebnote{Literally “your life” or “your soul”} will live.
\verse But if you are refusing to surrender, this is the thing that Adonai has shown me.
\verse Now look, all the women who remain in the house of the king of Judah are being led out to the officials of the king of Babylon. And look, they are saying, ‘\textit{Your trusted friends}\lebnote{Literally “The men of your peace”} have misled you, 
and they have prevailed against you. 
Your feet are stuck in the mud, 
so they turned backward.’
\verse And all your wives and your children will be led out to the Chaldeans, and you will not escape from their hand, but by the hand of the king of Babylon you will be seized, and this city will burn with fire.”
\verse And Zedekiah said to Jeremiah, “No man must know about these words so that you will not die.
\verse And if the officials hear that I have spoken with you, and they come to you and say to you, ‘Please tell us what you said to the king. You must not conceal it from us so that we will not kill you. And what did the king say to you?’
\verse Then you shall say to them, ‘I was presenting my plea \textit{before}\lebnote{Literally “to the face of”} the king, to not cause me to return to the house of Jonathan to die there.’ ”
\verse And all the officials came to Jeremiah and questioned him, and he informed them like all these words that the king commanded. So they fell silent, for the \textit{conversation}\lebnote{Literally “word”} was not heard.
\verse And Jeremiah stayed in the courtyard of the guard until the day that Jerusalem was captured. And it happened that Jerusalem was captured.\lebnote{Jeremiah 38:28b in the Hebrew Bible is 39:1a in the English Bible}
\end{biblechapter}

\begin{biblechapter} % Jeremiah 39
\verseWithHeading{The Fall of Jerusalem} In the ninth year of Zedekiah, the king of Judah, in the tenth month, Nebuchadnezzar the king of Babylon and all his army came against Jerusalem and laid siege to it.
\verse In the eleventh year of Zedekiah, in the fourth month, on the ninth day of the month, the city was taken by assault.
\verse And all the officials of the king of Babylon came and sat in the Middle Gate: Nergal-sharezer, Samgar-nebo, Sarsechim the chief officer, Nergal-sharezer the high official, with all the rest of the officials of the king of Babylon.
\verse \textit{And then}\lebnote{Literally “And it was”} \textit{when}\lebnote{Literally “as that”} Zedekiah the king of Judah saw them, and all \textit{the soldiers}\lebnote{Literally “the men of the war”} with him, they fled and went out at night from the city by the way of the garden of the king through the gate between the walls. And they went out toward the Jordan Valley.
\verse But the army of the Chaldeans pursued after them and overtook Zedekiah in the plains of Jericho. And they took him and brought him up to Nebuchadnezzar the king of Babylon at Riblah in the land of Hamath. And he pronounced \textit{sentence}\lebnote{Literally “judgments”} on him.
\verse And the king of Babylon slaughtered the sons of Zedekiah at Riblah before his eyes. The king of Babylon also slaughtered all the nobles of Judah.
\verse Then he blinded the eyes of Zedekiah and tied him up with bronze fetters to bring him to Babylon.
\verse And the Chaldeans burned the palace\lebnote{Or “house”} of the king and the houses\lebnote{Hebrew “house”} of the people with fire and broke down the walls of Jerusalem.
\verse Then the rest of the people who were left in the city, and those deserting who had deserted to him, and the rest of the people who remained, Nebuzaradan, the captain of the guard,\lebnote{Hebrew “guards”} deported to Babylon.
\verse And some of the poor people, \textit{who had nothing},\lebnote{Literally “who there was not for them something”} Nebuzaradan, the captain of the guard,\lebnote{Hebrew “guards”} left in the land of Judah. And he gave them vineyards and fields on that day.
\verse And Nebuchadnezzar the king of Babylon commanded concerning Jeremiah \textit{through}\lebnote{Literally “in the hand of”} Nebuzaradan, the captain of the guard,\lebnote{Hebrew “guards”} \textit{saying},\lebnote{Literally “to say”}
\verse “Take him and set your eyes on him. And you must not do something bad to him, \textit{but only}\lebnote{Literally “but if”} that which he speaks to you, so do with him.”
\verse So Nebuzaradan, the captain of the guard,\lebnote{Hebrew “guards”} sent word, \textit{along with}\lebnote{Literally “and”} Nebushazban the chief officer, and Nergal-sharezer the high official, and all the chief officers of the king of Babylon.
\verse And they sent and took Jeremiah from the courtyard of the guard and gave him to Gedaliah the son of Ahikam, the son of Shaphan, to take him to the house, so he stayed in the midst of the people.
\verseWithHeading{Ebed-melech is Rewarded for His Faith} And the word of Adonai came to Jeremiah at his confinement in the courtyard of the guard, \textit{saying},\lebnote{Literally “to say”}
\verse “Go and say to Ebed-melech the Cushite, \textit{saying},\lebnote{Literally “to say”} ‘Thus says Adonai of hosts, the God of Israel: “Look, I am about to bring my words to pass against this city for evil and not for good. And they will be \textit{before you}\lebnote{Literally “to the face of you”} on that day.
\verse But I will rescue you on that day,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “and you will not be given into the hand of the men \textit{of whom you are frightened}.\lebnote{Literally “who you are frightened from the face of them”}
\verse For surely I will save you, and you will not fall by the sword. But your life will be for you as booty because you have trusted in me,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.’ ”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 40
\verseWithHeading{Jeremiah Decides to Remain in the Land} The word that came to Jeremiah from Adonai after Nebuzaradan, the captain of the guard,\lebnote{Hebrew “guards”} had let him go from Ramah, \textit{where he had been taken}\lebnote{Literally “at his taking him} bound in chains in the midst of all the exiles\lebnote{Hebrew “exile”} of Jerusalem and Judah who were being deported to Babylon.
\verse And the captain of the guard\lebnote{Hebrew “guards”} took Jeremiah and said to him, “Adonai your God threatened this disaster to this place,
\verse and now he has brought it about,\lebnote{Here the direct object is supplied from context in the English translation} and Adonai has done \textit{just as}\lebnote{Literally “as that”} he threatened. Because you sinned against Adonai and did not listen to his voice, so this thing has happened to you.
\verse And so then look, I have released you \textit{today}\lebnote{Literally “the day”} from the chains that were on your hands.\lebnote{Hebrew “hand”} If it is good in your eyes to come with me to Babylon, then come, and \textit{I will take care of you}.\lebnote{Literally “I will place my eye on you”} But if it is bad in your eyes to come with me to Babylon, then refrain. Look, the whole land is \textit{before you}.\lebnote{Literally “to the face of you”} To wherever it is good and right in your eyes to go, then go there.”
\verse While he still had not turned back, Nebuzaradan added, “Return to Gedaliah the son of Ahikam, the son of Shaphan, whom the king of Babylon has appointed in an official position over the towns of Judah, and stay with him in the midst of the people. Or to wherever it is right in your eyes to go, then go there.”\lebnote{Here the direct object is supplied from context in the English translation} Then the captain of the guard\lebnote{Hebrew “guards”} gave him an allowance of provisions and a present, and let him go.
\verse So Jeremiah went to Gedaliah the son of Ahikam, at Mizpah and stayed with him in the midst of the people who were left in the land.
\verseWithHeading{The Assassination of Gedaliah} When all the commanders of the armies who were in the open country and their men heard that the king of Babylon had appointed Gedaliah the son of Ahikam in an official position in the land, and that \textit{he had put him in charge of}\lebnote{Literally “he had appointed to him”} men, and women, and little children, and of the poor of the land, of all those who had not been deported to Babylon,
\verse then they went to Gedaliah at Mizpah—Ishmael the son of Nethaniah, and Johanan, and Jehonathan the son of Kareah, and Seraiah the son of Tanhumeth, and the sons of Ephai the Netophathite, and Jezaniah the son of the Maacathite, they and their men.
\verse And Gedaliah the son of Ahikam, the son of Shaphan, swore to them and to their men, \textit{saying},\lebnote{Literally “to say”} “You must not be afraid of serving the Chaldeans. Stay in the land and serve the king of Babylon, and it will go well with you.
\verse \textit{As for me},\lebnote{Literally “And I”} look, I am staying at Mizpah to represent you \textit{before}\lebnote{Literally “to the face of”} the Chaldeans who come to us. But you, gather wine and summer fruit and oil, and put them in your vessels, and live in your towns that you have seized.”
\verse And also all the Judeans who were in Moab, and among the \textit{Ammonites},\lebnote{Literally “sons of Ammon”} and in Edom, and who were in all the lands, when they heard that the king of Babylon had given a remnant to Judah and that he had appointed Gedaliah the son of Ahikam, the son of Shaphan, in an official position over them,
\verse then all the Judeans returned from all the places to which they were scattered. And they came to the land of Judah, to Gedaliah at Mizpah, and they gathered wine and summer fruit that yielded \textit{in great abundance}.\lebnote{Literally “very numerous”}
\verse And Johanan the son of Kareah and all the commanders of the armies who were in the open country came to Gedaliah at Mizpah
\verse and said to him, “\textit{Are you at all aware}\lebnote{Literally “know you know”} that Baalis, the king of the \textit{Ammonites},\lebnote{Literally “sons of Ammon”} has sent Ishmael the son of Nethaniah, to kill you?” But Gedaliah the son of Ahikam would not believe them.
\verse Then Johanan the son of Kareah said to Gedaliah in secrecy at Mizpah, \textit{saying},\lebnote{Literally “to say”} “Please let me go and kill Ishmael the son of Nethaniah, and \textit{nobody}\lebnote{Literally “not a man”} will know. Why should he kill you, so that all of Judah who are gathered to you will be scattered, and the remnant of Judah will perish?”
\verse But Gedaliah the son of Ahikam said to Johanan the son of Kareah, “You must not do this thing, for you are telling a lie about Ishmael.”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 41
\verse \textit{And then}\lebnote{Literally “And it was”} in the seventh month Ishmael the son of Nethaniah, the son of Elishama, from the offspring of the kingship, and one of the chief officers of the king, came to Gedaliah the son of Ahikam at Mizpah, \textit{along with}\lebnote{Literally “and”} ten men. And they ate bread together there at Mizpah.
\verse And Ishmael the son of Nethaniah got up, \textit{along with}\lebnote{Literally “and”} the ten men who were with him, and they struck Gedaliah the son of Ahikam, the son of Shaphan, with the sword and killed him whom the king of Babylon had appointed in an official position over the land.
\verse Then Ishmael killed all the Judeans who were with Gedaliah at Mizpah, \textit{along with}\lebnote{Literally “and”} the Chaldeans who were found there, \textit{the soldiers}.\lebnote{Literally “the men of the war”}
\verseWithHeading{The Massacre of the Mourners} \textit{And then}\lebnote{Literally “And it was”} on the second day of the killing of Gedaliah—and \textit{no one}\lebnote{Literally “not a man”} knew—
\verse then men came from Shechem, from Shiloh, and from Samaria, eighty men with shaven beards\lebnote{Hebrew “beard”} and torn garments, who had cut themselves with blades, \textit{having}\lebnote{Literally “and”} grain offerings\lebnote{Hebrew “grain offering”} and frankincense in their hands\lebnote{Hebrew “hand”} to bring to the \textit{temple}\lebnote{Literally “house”} of Adonai.
\verse And Ishmael the son of Nethaniah came out to meet them from Mizpah, \textit{weeping as he came}.\lebnote{Literally “coming he came and weeping”} \textit{And then}\lebnote{Literally “And it was”} as he was meeting them, he said to them, “Come to Gedaliah, the son of Ahikam.”
\verse \textit{And then},\lebnote{Literally “And it was”} the moment of their coming to the middle of the city, then Ishmael the son of Nethaniah, \textit{along with}\lebnote{Literally “and”} the men who were with him, slaughtered\lebnote{The Hebrew verb is singular} them and threw them to the middle of the pit.\lebnote{Or “cistern”}
\verse But ten men were found among them, and they said to Ishmael, “You must not kill us, for \textit{we have}\lebnote{Literally “there is to us”} hidden treasures in the field, wheat, and barley, and oil, and honey. So he refrained and he did not kill them in the midst of their fellow countrymen.
\verse Now the pit\lebnote{Or “cistern”} into which Ishmael threw all the corpses of the men whom he had killed \textit{along with}\lebnote{Literally “by the hand of”} Gedaliah was the same one that King Asa had made because of Baasha the king of Israel, who was against him. Ishmael the son of Nethaniah filled it with the slain ones.
\verse Then Ishmael took captive all the rest of the people who were in Mizpah, the daughters of the king and all the people who were left at Mizpah, over whom Nebuzaradan, the captain of the guard,\lebnote{Hebrew “guards”} had appointed Gedaliah the son of Ahikam. And Ishmael took them captive and set out to cross over to the \textit{Ammonites}.\lebnote{Literally “sons of Ammon”}
\verse When Johanan the son of Kareah and all the commanders of the armies who were with him, heard\lebnote{The Hebrew verb is singular} all the evil that Ishmael the son of Nethaniah had done,
\verse then they took all the men and went to fight against Ishmael the son of Nethaniah. And they met him at the great pool that is in Gibeon.
\verse \textit{And then},\lebnote{Literally “And it was”} the moment that all the people who were with Ishmael saw Johanan the son of Kareah and all the commanders of the armies who were with him, they were glad.
\verse So all the people whom Ishmael had taken captive from Mizpah turned around and returned and went to Johanan the son of Kareah.
\verse But Ishmael the son of Nethaniah escaped with eight men \textit{from}\lebnote{Literally “from the face of”} Johanan, and they went to the \textit{Ammonites}.\lebnote{Literally “sons of Ammon”}
\verse Then Johanan the son of Kareah and all the commanders of the armies who were with him took from Mizpah all the rest of the people whom he had recovered from Ishmael the son of Nethaniah, after he had killed Gedaliah the son of Ahikam, strong men, \textit{soldiers},\lebnote{Literally “the men of the war”} and women, and little children, and eunuchs whom he brought back from Gibeon.
\verse And they set out and stayed at the lodging place of Chimham that is near Bethlehem, \textit{intending}\lebnote{Literally “going”} to go to Egypt
\verse because of the Chaldeans. For they were afraid \textit{of them},\lebnote{Literally “from the face of them”} because Ishmael the son of Nethaniah had killed Gedaliah the son of Ahikam, whom the king of Babylon had appointed in an official position over the land.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 42
\verseWithHeading{The Remnant Asks Jeremiah to Pray to Adonai for Direction} Then all the commanders of the armies, and Johanan the son of Kareah, and Jezaniah the son of Hoshaiah, and all the people from the smallest to the greatest, approached
\verse and said to Jeremiah the prophet, “May our plea please fall \textit{before you},\lebnote{Literally “to the face of you”} and pray for us to Adonai your God, for all this remnant, for we are left but a few of many, as your eyes are seeing us.
\verse And let Adonai your God inform us the way in which we should go and the thing that we should do.”
\verse And Jeremiah the prophet said to them, “I have heard. Look, I am going to pray to Adonai your God according to your words, \textit{and then}\lebnote{Literally “and it will happen”} all the words\lebnote{Hebrew “word”} that Adonai answers you I will tell you; I will not withhold from you a word.”
\verse Then they said to Jeremiah, “May Adonai be against us as \textit{a true and faithful witness}\lebnote{Literally “a witness of faithfulness and trustworthy”} if we do not do according to all the words\lebnote{Hebrew “word”} that Adonai sends you for us.
\verse Whether good or bad, we will listen to the voice of Adonai our God, to whom we are sending you, so that it may go well with us when we listen to the voice of Adonai our God.”
\verseWithHeading{The Remnant Rejects the Word of Adonai} \textit{And then}\lebnote{Literally “And it was”} at the end of ten days the word of Adonai came to Jeremiah.
\verse And he summoned Johanan the son of Kareah, and all the commanders of the armies who were with him, and all the people from the smallest to the greatest,
\verse and said to them, “Thus says Adonai, the God of Israel, to whom you sent me to cause your plea to fall \textit{before him}:\lebnote{Literally “to the face of him”}
\verse ‘If only you will stay in this land, then I will build you and I will not tear you down, and I will plant you and I will not pluck you up, for I relent of the disaster that I have brought to you.
\verse You must not be afraid \textit{of}\lebnote{Literally “from the face of”} the king of Babylon whom you are \textit{afraid of}.\lebnote{Literally “in fear from the presence of him”} You must not be afraid of him,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘for I am with you to save you and to deliver you from his hand.
\verse And I will show you compassion, and he will have compassion on you and will restore you to your soil.
\verse But if you are saying, “We will not stay in this land,” so as to not listen to the voice of Adonai your God,
\verse saying, “No, for we will go to the land of Egypt where we will not see war, and we will not hear the sound of a horn, and we will not be hungry for bread, and there we will stay,”
\verse then \textit{therefore}\lebnote{Literally “to thus”} hear the word of Adonai, O remnant of Judah. Thus says Adonai of hosts, the God of Israel: “\textit{If you are determined}\lebnote{Literally “If you indeed set your face”} to go to Egypt, and you go to dwell as aliens there,
\verse then it will be that the sword that you are in fear of will overtake you there in the land of Egypt, and the famine that you are anxious of will pursue after you there into Egypt, and there you will die.
\verse So all the people who \textit{are determined}\lebnote{Literally “have set their faces”} to go to Egypt to dwell as aliens there will die by the sword, by the famine, and by the plague, and there will not be for them an escapee or a survivor from the disaster that I am bringing upon them.”
\verse For thus says Adonai of hosts, the God of Israel: “As my anger and my wrath were poured out on the inhabitants of Jerusalem, so my wrath will pour out on you at your going to Egypt, and you will become a curse, and a horror, and a curse formula, and a disgrace, and you will no longer see this place.” ’
\verse Adonai has spoken to you, O remnant of Judah, ‘You must not go to Egypt.’ You must certainly know that I have warned you \textit{today}\lebnote{Literally “the day”}
\verse that \textit{you have made a fatal mistake},\lebnote{Literally “you have caused them to err for your lives”} for you sent me to Adonai your God, \textit{saying},\lebnote{Literally “to say”} ‘Pray for us to Adonai our God, and all that Adonai our God says, so tell us and we will do it.’\lebnote{Here the direct object is supplied from context in the English translation}
\verse So I have told you \textit{today},\lebnote{Literally “the day”} and you have not listened to the voice of Adonai your God, or of \textit{anything}\lebnote{Literally “all”} that he sent me for you.
\verse Now then, certainly you must know that by the sword, by the famine, and by the plague you will die in the place where you desire to go, to dwell as aliens there.”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 43
\verse \textit{And then},\lebnote{Literally “And it was”} when Jeremiah had finished speaking to all the people all the words of Adonai their God, for which Adonai their God had sent him to them with all these words,
\verse Azariah the son of Hoshaiah, and Johanan the son of Kareah, and all the insolent men spoke,\lebnote{The Hebrew verb is singular} saying to Jeremiah, “You are telling a lie! Adonai our God did not send you, \textit{saying},\lebnote{Literally “to say”} ‘You must not go to Egypt to dwell as aliens there.’
\verse But Baruch the son of Neriah is inciting you against us in order to give us into the hand of the Chaldeans, to kill us and to deport us to Babylon.”
\verse So Johanan the son of Kareah and all the commanders of the armies, and all the people did not listen\lebnote{The Hebrew verb is singular} to the voice of Adonai to stay in the land of Judah.
\verse But Johanan the son of Kareah and all the commanders of the armies took\lebnote{The Hebrew verb is singular} all the remnant of Judah who had returned from all the nations to which they had been scattered to dwell as aliens in the land of Judah—
\verse the men, and the women, and the little children, and the daughters of the king, and \textit{everyone}\lebnote{Literally “all of the person”} whom Nebuzaradan the captain of the \textit{guard}\lebnote{Hebrew “guards”} had left with Gedaliah the son of Ahikam, the son of Shaphan, and Jeremiah the prophet, and Baruch the son of Neriah—
\verse and they came into the land of Egypt, for they did not listen to the voice of Adonai. And they came up to Tahpanhes.
\verseWithHeading{The Throne of Nebuchadnezzar Will be Set Up at Tahpanhes} And the word of Adonai came to Jeremiah in Tahpanhes, \textit{saying},
\verse “Take in your hands\lebnote{Hebrew “hand”} large stones and bury them in the mortar in the clay floor that is at the entrance of the palace\lebnote{Or “house”} of Pharaoh in Tahpanhes before the eyes of the people\lebnote{Or “men”} of the Judeans,
\verse and say to them, ‘Thus says Adonai of hosts, the God of Israel: “Look, I am going to send and take Nebuchadnezzar the king of Babylon, my servant, and I will set his throne above these stones that I have hidden. And he will spread out his canopy over them.
\verse And he will come and ravage the land of Egypt, and deliver those who are appointed for the plague, to the plague, and those who are appointed for the captivity, to the captivity, and those who are appointed for the sword, to the sword.
\verse Then he will kindle a fire in the \textit{temples}\lebnote{Literally “houses”} of the gods of Egypt, and he will burn them and take them captive. And he will wrap up the land of Egypt just as the shepherd wraps up his cloak. And he will go forth from there in peace.
\verse And he will break the stone pillars of Heliopolis, which is in the land of Egypt. And he will burn the \textit{temples}\lebnote{Literally “houses”} of the gods of Egypt with fire.” ’ ”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 44
\verseWithHeading{A Message of Rebuke for the Rebellious Remnant} The word that came to Jeremiah concerning all the Judeans who were living in the land of Egypt, who were living at Migdol, and Tahpanhes, and Memphis, and in the land of Pathros, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai of hosts, the God of Israel: ‘You yourselves have seen all the disaster that I have brought on Jerusalem, and on all the towns of Judah. And look, they are a site of ruins this day, and there is not in them an inhabitant,
\verse \textit{because of}\lebnote{Literally “from the face of”} their wickedness that they committed to provoke me to anger by going to make smoke offerings, serving other gods whom they had not known, neither they, you, or your ancestors.\lebnote{Or “fathers”}
\verse And I sent to you all my servants the prophets, \textit{sending over and over again},\lebnote{Literally “doing early and sending”} \textit{saying},\lebnote{Literally “to say”} “Please, you must not do this detestable thing that I hate.”
\verse But they did not listen and they did not incline their ears\lebnote{Hebrew “ear”} to turn back from their wickedness, to not make smoke offerings to other gods.
\verse So my wrath and my anger were poured out\lebnote{The Hebrew verb is singular} and burned\lebnote{The Hebrew verb is singular} in the towns of Judah and in the streets of Jerusalem, and they became as a site of ruins, as a desolation, as they are this day.’
\verse So then, thus says Adonai, the God of hosts, the God of Israel, ‘Why are you doing great harm to \textit{yourselves},\lebnote{Literally “your lives”} to cut off for yourselves man and woman, child and suckling, from the midst of Judah, to not leave over for yourselves a remnant,
\verse provoking me to anger with the works of your hands, to make smoke offerings to other gods in the land of Egypt where you have come to dwell as aliens, so as to cut off yourselves, and so that you are becoming as a curse, and as a disgrace among all the nations of the earth?
\verse Have you forgotten the wicked things of your ancestors,\lebnote{Or “fathers”} and the wicked things of the kings of Judah, and the wicked things of their wives, and your wicked things, and the wicked things of your wives, which they committed in the land of Judah and in the streets of Jerusalem?
\verse They have not shown contrition up to this day, and they have not shown reverence, and they have not walked in my law and in my statutes that I set \textit{before you}\lebnote{Literally “to the face of you”} and \textit{before}\lebnote{Literally “to the face of”} your ancestors.’\lebnote{Or “fathers”}
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai of hosts, the God of Israel: ‘Look, \textit{I am determined to bring disaster on you},\lebnote{Literally “I am setting my face against you for disaster”} even to exterminate all Judah.
\verse And I will take away the remnant of Judah \textit{who are determined}\lebnote{Literally “who set their faces”} to come to the land of Egypt to dwell as aliens there, and everyone will perish. They will fall in the land of Egypt by the sword; they will perish by the famine, from the smallest to the greatest. By the sword and by the famine they will die and become as a curse, as a horror, and as a curse, and as a disgrace.
\verse So I will punish those who live in the land of Egypt just as I have punished Jerusalem, with the sword, with the famine, and with the plague.
\verse And there will not be a survivor or an escapee of the remnant of Judah, those who have come to dwell as an alien there in the land of Egypt, to return to the land of Judah where they are \textit{longing}\lebnote{Literally “lifting up their soul”} to return to dwell there, for they will not turn back, \textit{but only}\lebnote{Literally “but if”} some survivors.’ ”
\verse Then all the men who knew that their wives were making smoke offerings to other gods, and all the women who stood by, a great assembly, and all the people who lived in the land of Egypt in Pathros, answered Jeremiah, \textit{saying},\lebnote{Literally “to say”}
\verse “Concerning the word that you have spoken to us in the name of Adonai, we are not going to listen to you!
\verse But certainly we will do \textit{everything}\lebnote{Literally “all the word”} that went out from our mouths,\lebnote{Hebrew “mouth”} to make smoke offerings to the queen of heaven and to pour out to her libations just as we did, we, and our ancestors,\lebnote{Or “fathers”} our kings, and our officials, in the towns of Judah and in the streets of Jerusalem, for then \textit{we had enough to eat},\lebnote{Literally “we satisfied ourselves with food”} and we were well-off, and we did not see disaster.
\verse But \textit{since}\lebnote{Literally “from”} we stopped making smoke offerings to the queen of heaven and pouring out libations to her, we have lacked everything, and by the sword and by the famine we have perished.
\verse And indeed, when we were making smoke offerings to the queen of heaven and \textit{pouring out}\lebnote{Literally “to pour out”} to her libations, was it not \textit{with the consent of}\lebnote{Literally “from apart from”} our husbands that we made for her sacrificial cakes \textit{marked with her image},\lebnote{Literally “to the copying of her”} and we poured out to her libations?”
\verse Then Jeremiah said to all the people, to the men and to the women, and to all the people who answered him a word, \textit{saying},\lebnote{Literally “to say”}
\verse “The offering that you made, smoke offerings in the towns of Judah and in the streets of Jerusalem, you, and your ancestors,\lebnote{Or “fathers”} your kings, and your officials, and the people of the land, did not Adonai remember them, and did it not come to his \textit{mind}?\lebnote{Literally “heart”}
\verse And Adonai was no longer able to bear\lebnote{Or “endure”} it, \textit{because of}\lebnote{Literally “from the face of”} the evil of your deeds, \textit{because of}\lebnote{Literally “from the face of”} the detestable things that you committed. Thus your land became as a site of ruins, and as a horror, and as a curse, \textit{without}\lebnote{Literally “from there is not”} inhabitants, as it is this day.
\verse \textit{Because of}\lebnote{Literally “From the face of”} the fact that you made smoke offerings, and that you sinned against Adonai, and you did not listen to the voice of Adonai, and you have not walked in his law, and in his statute, and in his legal provisions, \textit{therefore}\lebnote{Literally “unto thus”} this disaster has happened to you, as it is this day.”
\verse Then Jeremiah said to all the people and to all the women, “Hear the word of Adonai, all Judah who are in the land of Egypt.
\verse Thus says Adonai of hosts, the God of Israel, \textit{saying},\lebnote{Literally “to say”} ‘You and your wives have promised with your mouths\lebnote{Hebrew “mouth”} and with your hands\lebnote{Hebrew “hand”} you have fulfilled, \textit{saying},\lebnote{Literally “to say”} “We certainly will carry out our vows that we vowed, to make smoke offerings to the queen of the heavens,\lebnote{Hebrew “heaven”} and to pour out to her libations.” You must certainly keep your vows and you must certainly carry out your vows.’
\verse \textit{Therefore}\lebnote{Literally “To thus”} hear the word of Adonai, all Judah, those who live in the land of Egypt: ‘Look, I hereby swear by my great name,’ says Adonai, ‘surely my name will no longer be called in the mouth of any man of Judah, saying, “\textit{as the Lord Adonai lives},”\lebnote{Literally “the life of Lord Adonai”} in all the land of Egypt.
\verse Look, I am going to watch over them for harm and not for good, and all people\lebnote{Or “men”} of Judah who are in the land of Egypt will perish by the sword and by the famine until their perishing.
\verse And the survivors of the sword will return from the land of Egypt to the land of Judah \textit{few in number}.\lebnote{Literally “males of number”} And all the remnant of Judah who have come to the land of Egypt to dwell as aliens there will know whose word will endure, from me or from them.
\verse And this will be to you the sign,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai, ‘that I am going to punish you in this place, so that you may know that surely my words will endure against you for evil.’
\verse Thus says Adonai: ‘Look, I am going to give Pharaoh Hophra, the king of Egypt, into the hand of his enemies, and into the hand of those who seek his life, just as I gave Zedekiah the king of Judah into the hand of Nebuchadnezzar the king of Babylon, his enemy and the one who sought his life.’ ”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 45
\verseWithHeading{A Message for Baruch} The word that Jeremiah the prophet spoke to Baruch, the son of Neriah, at his writing these words on a scroll from the mouth of Jeremiah, in the fourth year of Jehoiakim, the son of Josiah, the king of Judah, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai, the God of Israel, to you, O Baruch:
\verse ‘You said, “Woe please to me, for Adonai has added sorrow to my pain. I am weary in my groaning and I do not find rest.” ’
\verse Thus you will say to him: ‘Thus says Adonai, “Look, what I have built I am about to tear down, and what I have planted I am about to pluck up, it is all the land.
\verse And you, do you seek great things for yourself? Do not seek great things,\lebnote{Here the direct object is supplied from context in the English translation} for look, I am about to bring disaster upon all flesh,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “but to you I will give your life as booty in all the places where you may go.” ’ ”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 46
\verseWithHeading{Messages Concerning Egypt} The word of Adonai that came to Jeremiah the prophet concerning the nations.
\verse Concerning Egypt: Concerning the army of Pharaoh Neco, the king of Egypt, which was by the Euphrates River at Carchemish, which Nebuchadnezzar the king of Babylon defeated in the fourth year of Jehoiakim the son of Josiah, the king of Judah.
\verse “Prepare small shield and shield, 
and approach for the battle!
\verse Harness the horses and mount the steeds! 
And take your stand with helmets! 
Polish the spears! 
Put on the body armor!
\verse Why do I see them terrified? 
They are turning back 
and their warriors are beaten down. 
And they have fled to a place of refuge 
and they do not turn back. 
Terror is from all around,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse The swift cannot flee, 
and the warrior cannot escape. 
They have stumbled and they have fallen, 
by the bank of the Euphrates River.
\verse Who is this that rises like the Nile, 
like the rivers whose waters surge?
\verse Egypt rises like the Nile, 
even like the rivers whose waters surge. 
And he says, “Let me rise, 
let me cover the earth, 
let me destroy cities, 
and the inhabitants in them.”
\verse Go up, O horses, 
and drive madly, O chariots,\lebnote{Hebrew “chariot”} 
and let the warriors go forth; 
Cush and Put, who wield the small shield, 
and Lud, who wield and bend the bow.
\verse For that day is to the Lord Adonai of hosts 
a day of retribution, 
to take revenge on his foes. 
And the sword will devour and be satisfied, 
and it will drink its fill of their blood, 
for a sacrifice is for the Lord Adonai of hosts 
in the land of the north by the Euphrates River.
\verse Go up to Gilead and take balm, 
O virgin of the daughter of Egypt. 
\textit{In vain}\lebnote{Literally “To the vanity”} you make use of many medicines; 
there is no healing for you.
\verse The nations have heard your shame, 
and your cry of lament fills the earth. 
For warrior has stumbled against warrior, 
together they have fallen, the two of them.
\verse The word that Adonai spoke to Jeremiah the prophet about the coming of Nebuchadnezzar the king of Babylon to attack the land of Egypt:
\verse “Declare in Egypt, and proclaim in Migdol, 
and proclaim in Memphis and in Tahpanhes, 
say, ‘Take your stand and be prepared, 
for the sword will devour those all around you.’
\verse Why are your bulls defeated? 
He did not stand because Adonai thrust him down.
\verse He multiplied the ones\lebnote{Hebrew “one”} stumbling. 
Furthermore, each one fell to his neighbor, 
and said, ‘Stand up, and let us return to our people 
and to the land of our birth, 
\textit{because of}\lebnote{Literally “from”} the \textit{presence}\lebnote{Literally “face”} of the sword of the oppressor.’
\verse They will call there, 
‘Pharaoh, the king of Egypt is but a roar, 
\textit{he missed the opportunity}.’\lebnote{Literally “he has allowed to pass over the appointed time”}
\verse \textit{As I live},”\lebnote{Literally “Life I”} \textit{declares}\lebnote{Literally “a declaration of”} the king, 
Adonai of hosts is his name, 
“one will come like Tabor among the mountains 
and like Carmel by the sea.
\verse Pack your baggage for exile, 
O inhabitants\lebnote{Hebrew “inhabitant”} of the daughter of Egypt, 
for Memphis will become as a horror, 
and it will be destroyed, \textit{without}\lebnote{Literally “from there is not”} inhabitant.
\verse Egypt is a beautiful heifer, 
but a gadfly from the north will come against her.
\verse Even her mercenaries in her midst are like calves of fattening, 
for they also have turned away, 
they have fled together. 
They stand not, 
for the day of their calamity has come, 
the time of their punishment.
\verse Her sound is like a snake that glides away, 
for they march in force. 
And with axes they come to her 
like those who chop trees.
\verse They will cut down her forest,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“for \textit{it is impenetrable}.\lebnote{Literally “it cannot be searched out”} 
Yes, they are more numerous than locusts,\lebnote{Hebrew “locust”} 
and \textit{they are without number}.\lebnote{Literally “there is not to them a number”}
\verse The daughter of Egypt will be ashamed, 
she will be given into the hand of the people of the north.”
\verse Adonai of hosts, the God of Israel, said, “Look, I am going to punish Amon of Thebes, and Pharaoh, and Egypt, and her gods, and her kings, even Pharaoh and those who trust in him.
\verse And I will give them into the hand of the seekers of their lives,\lebnote{Hebrew “life”} and into the hand of Nebuchadnezzar the king of Babylon, and into the hand of his troops. But \textit{afterward}\lebnote{Literally “after thus”} it will be inhabited, as in days of old,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “But as for you, my servant Jacob, you must not be afraid, 
and Israel, you must not be dismayed, 
for look, I am going to save you from far, 
and your offspring from the land of their captivity. 
And Jacob will return, 
and he will be quiet, 
and he will be at ease, 
and without one who makes him afraid.
\verse You must not be afraid, my servant Jacob,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“for I am with you. 
For I will make a complete destruction in all the nations where I scattered you, 
but you I will not make a complete destruction, 
though I will chastise you to the measure, 
and I will certainly not leave you unpunished.”
\end{biblechapter}

\begin{biblechapter} % Jeremiah 47
\verseWithHeading{A Message Concerning the Philistines} The word of Adonai that came to Jeremiah the prophet concerning the Philistines \textit{before}\lebnote{Literally “at not yet”} Pharaoh attacked Gaza.
\verse Thus says Adonai:
\verse “Look, waters are rising from the north, 
and they will become as a raging torrent, 
and they will overflow the land, 
and that which fills it, 
the city and those who live in it. 
And the people will cry out, 
and every inhabitant of the land will wail,
\verse from the sound of the stamping of the hoofs of his stallions, 
from the clattering of his chariots,\lebnote{Hebrew “chariot”} 
the noise of their wheels. 
Fathers do not turn to their children 
because of the slackness of their hands.
\verse Because of the day that is coming 
to destroy all the Philistines, 
to cut off from Tyre and from Sidon 
every survivor who could help, 
for Adonai is destroying the Philistines, 
the remnant of the coastland of Caphtor.
\verse Baldness has come upon Gaza, 
Ashkelon is destroyed. 
O remnant of their valley, 
\textit{how long}\lebnote{Literally “until when”} will you cut yourself with a blade?
\verse Ah, sword of Adonai, 
\textit{how long}\lebnote{Literally “until when”} will you not be quiet? 
Gather to your sheath, 
keep quiet and stand still.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 48
\verseWithHeading{A Message Concerning Moab} Concerning Moab, thus says Adonai of hosts, the God of Israel:
\verse “Woe to Nebo, 
for it is devastated. 
It is ashamed, 
Kiriathaim is captured. 
Misgab is ashamed, 
and it is filled with terror.
\verse The glory of Moab is no longer. 
In Heshbon they planned evil against her. 
Come and let us cut her off from being a nation. 
Also, Madmen, you will be silent. 
After you will go the sword.
\verse The sound of a cry of distress arises from Horonaim, 
devastation and great destruction.
\verse Moab is broken, 
her little ones \textit{make heard}\lebnote{Literally “cause to hear”} a cry for help.
\verse For the ascent of Luhith, 
in weeping they go weeping, 
For on the slope of Horonaim they have heard 
the distresses of the cry of distress of the destruction.
\verse Flee! Save your life! 
\textit{For}\lebnote{Literally “And”} you must not be like the juniper tree in the desert.
\verse For because of your trusting in your works and in your treasures, 
even you yourselves will be taken, 
and Chemosh will go out into the exile, 
his priests and his officials together.
\verse And the destroyer will come to every town, 
and a town will not escape. 
And the valley will be destroyed, 
and the plain will be destroyed, as Adonai has spoken.
\verse Give salt for Moab, 
for certainly she will go to ruin, 
and her towns will become as a desolation, 
\textit{without}\lebnote{Literally “from there is not”} an inhabitant in them.
\verse The one who does the work of Adonai with slackness is cursed, 
and the one who keeps away blood from his sword is cursed.
\verse Moab has been at ease from his youth, 
and he has been quiet on his dregs, 
and he has not been poured out from vessel to vessel, 
and he has not gone into exile. 
\textit{Therefore}\lebnote{Literally “Unto thus”} his taste has remained unchanged, 
and his fragrance has not changed.
\verse \textit{Therefore}\lebnote{Literally “To thus”} look, days are coming,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“and I will send to him \textit{decanters},\lebnote{Literally “those who tilt”} 
and they will decant him, 
and his vessels they will empty out, 
and their jars they will smash.
\verse Then Moab will be ashamed of Chemosh, 
just as the house of Israel were ashamed of Bethel, their trust.
\verse How can you say, ‘we are warriors, 
and men of might for the battle.’
\verse Moab is destroyed, 
and his towns have gone up, 
and the choice of his young men have gone down to the slaughtering,” 
\textit{declares}\lebnote{Literally “a declaration of”} the King, Adonai of hosts is his name.
\verse “The disaster of Moab is near to come, 
and his disaster \textit{approaches very quickly}.\lebnote{Literally “hastens greatly”}
\verse Mourn for him, all his surroundings, 
and all those who knew his name. 
Say, ‘How the staff of might is broken, 
the staff of glory.’
\verse Come down from glory and sit on the parched ground, 
O inhabitant, daughter of Dibon, 
for the destroyer of Moab has come up against you, 
he has destroyed your fortifications.
\verse Stand by the road and watch, 
O inhabitant of Aroer. 
Ask the man who is fleeing and the woman who is escaping, 
say, ‘What has happened?’
\verse Judgment has also come to the land of the plain, to Holon, and to Jahaz, and on Mephaath,
\verse and on Dibon, and on Nebo, and on Beth-diblathaim,
\verse and on Kiriathaim, and on Beth-gamul, and on Beth-meon,
\verse and on Kerioth, and on Bozrah, and on all the towns of the land of Moab, the far and the near.
\verse The horn of Moab is cut off and his arm is broken,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “Make him drunk, 
because he magnified himself against Adonai, 
and let Moab wallow in his vomit. 
And he also will become a laughingstock.
\verse Now was not Israel the laughingstock to you? 
Or was he found among thieves? 
For \textit{as often as}\lebnote{Literally “from enough”} you spoke over him 
you shook your head.
\verse Leave the towns and dwell on the rock, 
O inhabitants of Moab, 
and be like the dove, 
it nests on the sides of the mouth of a ravine.
\verse We have heard of the pride of Moab 
—he is very proud— 
his height, and his pride, and his arrogance, 
and the haughtiness of his heart.
\verse I know his insolence,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“and his boastings are \textit{false};\lebnote{Literally “not right”} 
they act \textit{falsely}.\lebnote{Literally “not right”}
\verse \textit{Therefore}\lebnote{Literally “Unto thus”} over Moab I wail, 
and for all of Moab I cry out, 
for the people\lebnote{Or “men”} of Kir-heres I moan.
\verse More than the weeping of Jazer 
I weep for you, O vine Sibmah. 
Your branches cross over the sea, 
up to the sea of Jazer they reach. 
Upon your summer fruit and upon your vintage 
the destroyer has fallen.
\verse And joy and rejoicing have been taken away from the fruitful land, 
even from the land of Moab. 
And the wine from the wine presses 
I have caused to disappear. 
He will not tread with exuberant shouts.\lebnote{Hebrew “exuberant shout”} 
Exuberant shouts\lebnote{Hebrew “Exuberant shout”} will not be exuberant shouts.\lebnote{Hebrew “exuberant shout”}
\verse From the cry for help of Heshbon, up to Elealeh, 
up to Jahaz, they give their voice, 
from Zoar up to Horonaim, up to Eglath-shelishiyah, 
for even the waters of Nimrim have become as a desolation.
\verse And I will cause to disappear for Moab,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“the one who presents a sacrifice at a high place, 
and the one who burns a smoke offering to his gods.
\verse \textit{Therefore}\lebnote{Literally “Unto thus”} my heart moans for Moab like the flute.\lebnote{Hebrew “flutes”} 
It moans for the people\lebnote{Or “men”} of Kir-heres like the flute.\lebnote{Hebrew “flutes”} 
\textit{Therefore}\lebnote{Literally “Unto thus”} the wealth they gained has perished.
\verse For on every head is baldness, 
and every beard is trimmed. 
On all the hands, slashes, 
and on the loins, sackcloth.
\verse On all the roofs of Moab, and in its public squares, 
there is wailing everywhere. 
For I have broken Moab like a vessel, 
there is no delight in it,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “How it is filled with terror! How they wail! 
How Moab has turned his back! He is disgraced! 
So Moab has become as a derision, 
and as a terror for all his surrounding peoples.”
\verse For thus says Adonai,
\verse “Look, like an eagle he will swoop down, 
and he will spread out his wings against Moab.
\verse Kerioth will be taken, 
and the strongholds will be conquered, 
and the hearts\lebnote{Hebrew “heart”} of the warriors of Moab will be on that day 
like the heart of a woman who is in labor.
\verse And Moab will be destroyed from being a people, 
because against Adonai he magnified himself.
\verse Dread, and pit, and a trapping net are against you, 
O inhabitant of Moab,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “He who flees \textit{before}\lebnote{Literally “from the face of”} the dread 
will fall into the pit, 
and the one who climbs up from the pit 
will be caught in the trapping net, 
for I will bring on Moab 
the year of their punishment,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “In the shadow of Heshbon 
fugitives stand without strength, 
for a fire has gone out from Heshbon, 
and a flame from Sihon. 
And it has eaten the forehead of Moab, 
and the skulls\lebnote{Hebrew “skull”} of the people\lebnote{Or “sons”} of tumult.
\verse Woe to you, Moab! 
The people of Chemosh have perished. 
For your sons have been taken into captivity, 
and your daughters into captivity. \textit{Thus far}\lebnote{Literally “Up to here”} is the judgment of Moab.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 49
\verseWithHeading{A Message Concerning Ammon} Concerning the \textit{Ammonites}:\lebnote{Literally “sons of Ammon”} Thus says Adonai,
\verse “Are there no sons for Israel? 
Or is there no heir for him? 
Why has Milcom taken possession of Gad, 
and his people dwelled in its towns?
\verse \textit{Therefore}\lebnote{Literally “To thus”} look, days are coming,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“and I will sound against Rabbah, the \textit{Ammonites},\lebnote{Literally “sons of Ammon”} 
the alarm of the war, 
and it will become as a mound of desolation, 
and its daughters will burn in the fire. 
Then Israel will dispossess his dispossessors,” says Adonai.
\verse “Wail, Heshbon, for Ai is devastated. 
Cry out, O daughters of Rabbah, put on sackcloth. 
Lament and run back and forth among the walls. 
For Milcom will go into exile, 
his priests and his officials together.
\verse Why do you boast in your valleys? 
Your valleys\lebnote{Hebrew “valley”} are ebbing. 
O unfaithful daughter, who trusted in her treasures, 
who said, ‘Who will come against me?’
\verse Look, I am going to bring dread on you,” 
\textit{declares}\lebnote{Literally “a declaration of”} Lord Adonai of hosts, 
“from all your surrounding neighbors. 
And you will be scattered, 
each one \textit{before it},\lebnote{Literally “to the face of it”} 
and there is no one who gathers the fugitives.\lebnote{Hebrew “fugitive”}
\verseWithHeading{A Message Concerning Edom} Concerning Edom, thus says Adonai of hosts:
\verse “Is there no longer wisdom in Teman? 
Has counsel become lost from those who understand? 
Is their wisdom rotting?
\verse Flee! Turn back! \textit{Get down low}!\lebnote{Literally “make deep to dwell”} 
O inhabitants of Dedan. 
For I will bring upon him the disaster of Esau 
at the time when I will punish him.
\verse If grape-gatherers come to you, 
would they not leave gleanings? 
If thieves in the night, 
they destroy only \textit{enough for them}.\lebnote{Literally “their enough”}
\verse But I have stripped Esau bare. 
I have uncovered his hiding places, 
and he is not able to hide himself. 
His offspring are devastated, \textit{along with}\lebnote{Literally “and”} his brothers, 
and his neighbors, and \textit{he is no more}.\lebnote{Literally “and not he”}
\verse For thus says Adonai, “Look, those \textit{who are not condemned}\lebnote{Literally “who there is not their justice”} to drink the cup must certainly drink it.\lebnote{Here the direct object is supplied from context in the English translation} And are you the one who will go entirely unpunished? You will not go unpunished, for you must certainly drink it.\lebnote{Here the direct object is supplied from context in the English translation}
\verse For I have sworn by myself,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, “that Bozrah will become as a horror, as a disgrace, as a waste, and as a curse. And all her towns will be \textit{everlasting sites of ruins}.”\lebnote{Literally “sites of ruins of eternity”}
\verse I have heard a message from Adonai, 
and a messenger has been sent among the nations. 
“Gather and come against her, 
and rise for the battle.”
\verse “For look, I will make you small among the nations, 
despised by the humankind.
\verse Your terror has deceived you, 
the pride of your heart, 
O you who dwell in the clefts of the rock, 
you who seize the height of the hill. 
Though you make your nest high like the eagle, 
from there I will bring you down,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “And Edom will become as a horror, 
\textit{everyone who passes by it}\lebnote{Literally “all who passes by it”} will be appalled, 
and will hiss because of all its blows.
\verse As the overthrow of Sodom and Gomorrah 
and their neighbors,” says Adonai, 
“\textit{no one}\lebnote{Literally “not a man”} will live there, 
and the son of humankind will not dwell as an alien in it.
\verse Look, like a lion he comes up from the thickets of the Jordan 
to \textit{a pastureland by a constantly flowing stream}.\lebnote{Literally “a pasture of everflowing”} 
So \textit{I will chase him away hurriedly}\lebnote{Literally “I will do in an instant I will chase him away”} from it. 
And whoever is chosen I will appoint over it. 
For who is like me, and who can summon me? 
And who is this, the shepherd who can stand \textit{before me}?”\lebnote{Literally “to the face of me”}
\verse \textit{Therefore}\lebnote{Literally “To thus”} hear the plan of Adonai that he has planned against Edom, 
and his plans that he has planned against the inhabitants of Teman. 
\textit{Surely}\lebnote{Literally “If not”} they will drag them away, 
the little ones of the flock. 
\textit{Surely}\lebnote{Literally “If not”} he will cause to be desolated over them 
their grazing place.
\verse The earth will quake from the sound of their falling. 
A cry of distress! Their voice will be heard at \textit{the Red Sea}.\lebnote{Hebrew “Sea of Reeds”}
\verse Look, like an eagle he will go up and he will swoop down, 
and he will spread out his wings against Bozrah. 
And the heart of the warriors of Edom will be in that day 
like the heart of a woman who is in labor.
\verseWithHeading{A Message Concerning Damascus} Concerning Damascus:
\verse “Hamath and Arpad are ashamed, 
for they have heard bad news; 
they melt. 
There is concern in the sea. 
It is not able to keep quiet.
\verse Damascus has grown slack. 
She has turned away to flee and panic. 
Anxiety has seized her, 
and labor pains have grasped her, 
like a woman in labor.
\verse How the city of praise has not been abandoned! 
The city of my joy!
\verse \textit{Therefore}\lebnote{Literally “To thus”} her young men will fall in her public squares, 
and all \textit{the soldiers}\lebnote{Literally “the men of the battle”} will perish in that day,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai of hosts.
\verseWithHeading{A Message Concerning Kedar and the Kingdoms of Hazor} Concerning Kedar and concerning the kingdoms of Hazor that Nebuchadnezzar the king of Babylon defeated. Thus says Adonai:
\verse “Rise up, go up against Kedar 
and destroy the people\lebnote{Or “sons”} of the east.
\verse They will take their tents, and their flocks, 
their tent curtains, and all their equipment, 
and their camels they will carry away for themselves, 
and they will call to them, ‘Terror is from all around.’
\verse Flee! Wander far away! \textit{Get down low}!\lebnote{Literally “Make deep to dwell”} 
O inhabitants of Hazor,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai. 
“For Nebuchadnezzar, the king of Babylon, has planned a plan against you. 
And he has schemed\lebnote{Or “planned”} a scheme\lebnote{Or “plan”} against you.
\verse Rise up, go up against a nation which is at ease, 
living \textit{securely},”\lebnote{Literally “with confidence”} \textit{declares}\lebnote{Literally “a declaration of”} Adonai. 
“There are no gates and there are no bars\lebnote{Hebrew “bar”} for them, 
they dwell alone.
\verse And their camels will become as plunder, 
and the multitude of their livestock as spoil. 
And I will scatter to every wind those who are trimmed to the side, 
and from \textit{every side}\lebnote{Literally “all his sides”} I will bring their disaster,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verseWithHeading{A Message Concerning Elam} The word of Adonai that came to Jeremiah the prophet concerning Elam at the beginning of the reign of Zedekiah the king of Judah, \textit{saying},\lebnote{Literally “to say”}
\verse “Thus says Adonai of hosts:
\verse ‘Look, I am going to break the bow of Elam, 
the first and best of their might.
\verse And I will bring to Elam four winds, 
from the four corners of heaven, 
and I will scatter them to all these winds, 
and there will not be a nation 
where the scattered people of Elam will not go.
\verse So I will shatter Elam \textit{before}\lebnote{Literally “to the face of”} their enemies, 
and \textit{before}\lebnote{Literally “to the face of”} the seekers of their life. 
And I will bring disaster upon them, 
\textit{my burning anger},\lebnote{Literally “the blaze of my nose”}’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai. 
‘And I will send after them the sword 
until I have destroyed them.
\verse And I will set my throne in Elam 
and I will destroy from there king and officials,’ \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 50
\verseWithHeading{A Message Concerning Babylon} The word that Adonai spoke concerning Babylon, concerning the land of the Chaldeans, by the hand of Jeremiah the prophet:
\verse “Declare among the nations and proclaim, 
and lift up a banner. 
Proclaim, you must not conceal it. 
Say, ‘Babylon is captured, 
Bel is ashamed, 
Merodach is filled with terror, 
her idols are ashamed, 
her idols are filled with terror.
\verse For a nation will come up against her from the north. 
It will make her land as a horror, 
and an inhabitant will not be in her. 
From humans\lebnote{Hebrew “human”} to animals\lebnote{Hebrew “animal”} they all will wander off. 
They will go away.
\verse In those days and in that time,’ ” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“the people\lebnote{Or “sons/children”} of Israel will come, 
they and the people\lebnote{Or “sons/children”} of Judah together. 
\textit{Weeping as they go},\lebnote{Literally “They shall go and weeping”} they will go, 
and Adonai their God they will seek.
\verse They will ask the way to Zion, 
turning their faces there. 
They will come and join themselves to Adonai 
by \textit{an everlasting covenant}\lebnote{Literally “a covenant of eternity”} that will not be forgotten.
\verse My people have become lost sheep, 
their shepherds have caused them to go astray. 
They led them away to the mountains. 
From mountain to hill they have gone, 
they have forgotten their resting place.
\verse All those who found them have devoured them. 
And their foes have said, ‘We are not guilty, 
\textit{because}\lebnote{Literally “under that”} they have sinned against Adonai, the \textit{true pasture},\lebnote{Literally “settlement of righteousness”} 
even Adonai, the hope of their ancestors.’\lebnote{Or “fathers”}
\verse Flee from the midst of Babylon, 
and from the land of the Chaldeans. 
Go out and be like he-goats 
\textit{before}\lebnote{Literally “to the face of”} the flock.
\verse For look, I am going to stir up, 
and I am going to bring against Babylon 
a contingent of great nations 
from the land of the north. 
And they will draw up a battle formation against her, 
from there she will be captured. 
Their arrows are like a warrior who has achieved success, 
he does not return without success.
\verse And Chaldea will be as booty, 
all those who plunder her will be satisfied,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “Though you rejoice, though you exult, 
O plunderers of my inheritance, 
though you frisk about like a heifer on the grass, 
and you neigh like the powerful stallions,
\verse your mother will be very ashamed. 
She who gave birth to you will be humiliated.\lebnote{Or “ashamed”} 
Look, she will be the least of the nations, 
a desert,\lebnote{Or “wilderness”} dry land, and wilderness.\lebnote{Or “Arabah”}
\verse Because of the anger of Adonai she will not be inhabited, 
but she will be a wasteland, all of her. 
Everyone who passes by Babylon will be appalled 
and will hiss because of all her wounds.
\verse Draw up a battle formation at Babylon all around, 
all those who bend the bow, shoot at her. 
You must not spare arrows,\lebnote{Hebrew “arrow”} 
for she has sinned against Adonai.
\verse Raise a war cry against her on all sides. 
\textit{She has surrendered}.\lebnote{Literally “She has given her hand”} 
Her towers have fallen, 
her walls are ruined. 
For this is the vengeance of Adonai. 
Take revenge on her. 
As she has done to others, 
so do to her.
\verse Cut off the sower from Babylon, 
and the one who uses the sickle in the time of harvest. 
\textit{Because of}\lebnote{Literally “From the face of”} the sword of the oppressor 
each one will turn to their people, 
and each one will flee to their land.
\verse Israel is a sheep scattered, 
lions drove them away. 
The first who devoured it was the king of Assyria, 
and now at the end Nebuchadnezzar, the king of Babylon, gnawed its bones.”
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai of hosts, the God of Israel,
\verse “Look, I am going to punish the king of Babylon and his land 
just as I punished the king of Assyria.
\verse And I will restore Israel to its pasture, 
and it will feed on Carmel, and in Bashan, 
and on the hills\lebnote{Hebrew “hill”} of Ephraim, and in Gilead 
its hunger will be satisfied.
\verse In those days and at that time,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“the guilt of Israel will be sought, but there is none, 
and the sins of Judah, but they will not be found, 
for I will forgive those I left behind.
\verse To the land of Merathaim, 
go up against her, 
and against the inhabitants of Pekod, 
massacre and destroy them,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“and do according to all that I have commanded you.
\verse The sound of battle is in the land, 
and great destruction.
\verse How the blacksmith’s hammer of all the earth is cut down and broken! 
How Babylon has become as a horror among the nations!
\verse I laid a bird snare for you, and you were caught, O Babylon. 
But you yourself did not know. 
You were discovered and seized, 
because with Adonai you measured yourself.
\verse Adonai has opened his armory, 
and he has brought out the weapons of his wrath, 
for it is a word for my Lord Adonai of hosts 
in the land of the Chaldeans.
\verse Come against her from the end, 
open her granaries. 
Pile her up like heaps and destroy her. 
Let there be no remnant for her.
\verse Massacre all her bulls, 
let them go down to the slaughtering. 
Woe to them, for their day has come, 
the time of their punishment.
\verse There is the sound of fugitives and survivors 
from the land of Babylon, 
to declare in Zion the vengeance of Adonai our God, 
the vengeance for his temple.
\verse Summon archers against Babylon, 
all those who bend the bow. 
Encamp all around her, 
there must not be for her an escape. 
Take revenge on her according to her deeds. 
According to all that she has done, so do to her. 
For against Adonai she has behaved insolently, 
against the Holy One of Israel.
\verse \textit{Therefore}\lebnote{Literally “To thus”} her young men will fall in her public squares, 
and all \textit{her soldiers}\lebnote{Literally “her the men of war”} will perish on that day,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “Look, I am against you, O arrogant one,” 
\textit{declares}\lebnote{Literally “a declaration of”} Lord Adonai of hosts, 
“for your day has come, 
the time when I will punish you.
\verse Thus says Adonai of hosts:
\verse “The people\lebnote{Or “sons/children”} of Israel are oppressed, 
and the people\lebnote{Or “sons/children”} of Judah likewise, 
\textit{for}\lebnote{Literally “and”} all their captors have seized them, 
they refuse to let them go free.
\verse Their redeemer is strong, 
his name is Adonai of hosts. 
Surely he will plead their case, 
in order to make peace for the earth, 
but to cause unrest to the inhabitants of Babylon.
\verse A sword against the Chaldeans,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“and against the inhabitants of Babylon, 
and against her officials, 
and against her wise men.
\verse A sword against the oracle priests, 
and they will become foolish. 
A sword against her warriors, 
and they will be filled with terror.
\verse A sword against his horses, 
and against his chariots,\lebnote{Hebrew “chariot”} 
and against all the foreign troops who are in the midst of her, 
so that they may become as women. 
A sword against her treasures, 
so that they may be plundered.
\verse A drought will come against her waters, 
and they will dry up, 
for it is a land of images, 
and because of the frightful objects, they act like madmen.
\verse \textit{Therefore}\lebnote{Literally “To thus”} desert creatures will live there\lebnote{Here the direct object is supplied from context in the English translation} with jackals, 
and daughters of ostriches\lebnote{Hebrew “ostrich”} will inhabit her. 
And she will not be inhabited again \textit{forever},\lebnote{Literally “to eternity”} 
and she will not be dwelt in \textit{for all generations}.\lebnote{Literally “until generation and generation”}
\verse As when God overthrew Sodom and Gomorrah 
and their neighbors,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“no one will live there, 
and the son of humankind will not dwell as an alien in her.
\verse Look, a people is about to come 
from the north, 
and a great nation and many kings are woken up 
from the remotest part of the earth.
\verse They keep hold of bow and short sword. 
They are cruel and they have no mercy. 
Their sound is like the sea, it roars, 
and upon horses they ride. 
Drawn up for battle as a man for the battle, 
against you, O daughter of Babylon.
\verse The king of Babylon heard their report 
and his hands grew slack. 
Anxiety has seized him, 
fear and pain like the woman who gives birth.
\verse Look, like a lion he comes up from the thickets\lebnote{Hebrew “thicket”} of the Jordan 
against \textit{a pastureland by a constantly flowing stream},\lebnote{Literally “a pasture of ever-flowing”} 
so \textit{I will chase them away quickly}\lebnote{Literally “I will do in an instant I will cause them to run”} from her, 
and I will appoint whoever is chosen over her. 
For who is like me, and who can summon me, 
and who is this shepherd who can stand \textit{before me}?\lebnote{Literally “to the face of me”}
\verse \textit{Therefore}\lebnote{Literally “To thus”} hear the plan of Adonai 
that he has planned against Babylon, 
and his plans 
that he has planned against the land of the Chaldeans. 
\textit{Surely}\lebnote{Literally “If not”} they will drag them away, 
the little ones of the flock. 
\textit{Surely}\lebnote{Literally “If not”} he will cause their grazing place 
to be desolate over them.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 51
\verse Thus says Adonai:
\verse “Look, I am going to stir up against Babylon 
and against the inhabitants of Leb-qamai a destructive wind.
\verse And I will send winnowers to Babylon, 
and they will winnow her, 
and they will lay waste her land, 
for they will be against her from all sides on the day of disaster.
\verse Let not him who bends the bow shoot his bow, 
and let him not rise high in his body armor. 
And you must not spare her young men; 
destroy her whole army.
\verse And they will fall slain in the land of the Chaldeans, 
and pierced through in her streets.
\verse For neither Israel nor Judah is a widower from their God, 
from Adonai of hosts, 
though their land is full of guilt 
against the Holy One of Israel.”
\verse Flee from the midst of Babylon and save each one his life. 
You must not perish because of her guilt. 
For this is the time of vengeance for Adonai, 
he will repay her what is deserved.
\verse Babylon was a cup of gold in the hand of Adonai, 
making drunk all the earth. 
The nations drank of her wine. 
\textit{Therefore}\lebnote{Literally “Unto thus”} the nations acted like madmen.
\verse Suddenly Babylon has fallen and she is shattered. 
Wail over her! 
Take balm for her wounds, 
perhaps she may be healed.
\verse We tried to heal Babylon, and she was not healed. 
Forsake her and let us go each one to his country, 
for her judgment has reached to the heavens, 
and it has been lifted up to the skies.
\verse Adonai has brought forth our vindication. 
Come and let us make known in Zion the work of Adonai our God.
\verse Sharpen the arrows, fill the quivers! 
Adonai has stirred up the spirit of the kings of the Medes. 
Because his plan concerning Babylon is to destroy it, 
for that is the vengeance of Adonai, 
the vengeance for his temple.
\verse Against the walls of Babylon raise a banner; 
\textit{post a strong watch},\lebnote{Literally “make strong the guard”} post watchmen, prepare the ambushes, 
for Adonai has both planned as well as performed 
what he has spoken concerning the inhabitants of Babylon.
\verse O you who live by mighty waters, rich in treasures, 
your end has come, \textit{the measure of your life}.\lebnote{Literally “the cubit of your severing”}
\verse Adonai of hosts has sworn by himself, 
“Surely I will fill you with troops like the creeping locusts,\lebnote{Hebrew “locust”} 
and they will sing over you a war cry.”
\verse He is the one who made the earth by his power, 
the one who created the world by his wisdom, 
and by his understanding he stretched out the heavens.
\verse At the giving of his voice 
there is a roar of waters in the heavens, 
and he causes the patches of mist 
to go up from the end of the earth. 
He makes lightning for the rain, 
and he brings out the wind from his storehouses.
\verse All humankind turns out to be stupid, 
without knowledge. 
Every goldsmith is put to shame 
by the divine image. 
For his cast image is a lie, 
and there is no breath in them.
\verse They are worthless, 
a work of mockery. 
At the time of their punishment, 
they will perish.
\verse The portion of Jacob is not like these, 
for he is the creator of \textit{everything},\lebnote{Literally “the all”} 
and the tribe of his inheritance. 
Adonai of hosts is his name.
\verse He says, “You are a war club for me, 
a weapon of battle, 
and I smash nations with you, 
and I destroy kingdoms with you.
\verse And I smash the horse with you, 
and its rider, 
and I smash the chariot with you, 
and its rider.
\verse And I smash man and woman with you, 
and I smash the old man and the boy with you, 
and I smash the young man and the young woman with you.
\verse And I smash the shepherd and his flock with you, 
and I smash the farmer and his team with you, 
and I smash the governors and the officials with you.
\verse And I will repay Babylon, 
and all the inhabitants of Chaldea, 
all their wickedness that they have done in Zion 
before your eyes,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “Look, I am against you, 
O mountain of the destruction,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“the one that destroys the whole earth. 
And I will stretch out my hand against you, 
and I will roll you down from the cliffs, 
and I will make you as a mountain burned away.
\verse They will not take from you 
a stone for a corner, 
and a stone for a foundation, 
for you will be \textit{an everlasting desolation},”\lebnote{Literally “a desolation of everlasting”} \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse Raise a banner in the land; 
blow a horn among the nations; 
prepare for holy war against her; 
summon the nations against her, 
the kingdoms of Ararat, Minni, and Ashkenaz; 
summon against her an official; 
bring up horses\lebnote{Hebrew “horse”} like bristling creeping locusts.\lebnote{Hebrew “locust”}
\verse Prepare the nations for holy war against her, 
the kings of the Medes, 
with their governors and all their officials, 
and every land of their dominion.
\verse And the land quakes and writhes, 
for the plans of Adonai against Babylon stand, 
to make the land of Babylon as a horror, 
\textit{without}\lebnote{Literally “from there is not”} inhabitant.
\verse The warriors of Babylon have ceased to fight, 
they remain in the strongholds, 
their power has dried up, 
they have become as women. 
Her dwelling places are set on fire, 
her bars are broken.
\verse one runner runs to meet another runner, 
and one messenger to meet another messenger, 
to tell the king of Babylon that his city has been captured, 
\textit{from end to end}.\lebnote{Literally “from end”}
\verse For thus says Adonai of hosts, the God of Israel,
\verse “The daughter of Babylon is like a threshing floor 
at the time it is trodden down, 
\textit{just a little while}\lebnote{Literally “still a little”} 
and the time of the harvest will come for her.”
\verse Nebuchadnezzar the king of Babylon 
has devoured me and sucked me dry; 
he has made me an empty vessel; 
he has swallowed me like the sea monster; 
he has filled his belly with my delicacies; 
he has rinsed me.
\verse \textit{Therefore}\lebnote{Literally “To thus”} thus says Adonai,
\verse “Look, I am going to contend your case, 
and I will avenge your vengeance, 
and I will cause her sea to dry up, 
and I will cause her fountain to dry up.
\verse And Babylon will become as a heap of stones, 
a lair of jackals, 
an object of horror, and an object of hissing, 
\textit{without}\lebnote{Literally “from there is not”} inhabitants.
\verse They will roar together like the young lions, 
they will growl like the cubs of lions.
\verse At their becoming hot I will set their banquets, 
and I will make them drunk, 
so that they will fall into a swoon, 
and they will sleep \textit{an everlasting sleep},\lebnote{Literally “a sleep of eternity”} 
and they will not wake up,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse “I will bring them down like young rams to slaughter, 
like rams and goats.
\verse How Sheshach is captured, 
and the glory of all the earth is seized! 
How Babylon has become as an object of horror 
among the nations!
\verse The sea has risen over Babylon, 
she has been covered by the roar of its waves.
\verse Her cities have become as an object of horror, 
a dry land and a wilderness,\lebnote{Or “Arabah”} 
a land in which no person lives, 
nor does a son of humankind pass through it.
\verse And I will punish Bel in Babylon, 
and I will wrench out from his mouth \textit{what he has swallowed}.\lebnote{Literally “what has been swallowed by him”} 
And the nations will not stream towards him any longer 
\textit{what’s worse},\lebnote{Literally “also”} the wall of Babylon has fallen.
\verse Come out from her midst, my people, 
and save each one his life 
from \textit{the burning anger of}\lebnote{Literally “the blazing of the nose of”} Adonai.
\verse Now \textit{so that you are not fainthearted},\lebnote{Literally “so that your heart is not timid”} 
and you are afraid at the rumors\lebnote{Hebrew “rumor”} that are heard in the land— 
and in the year the rumor comes, 
and in the year after it the rumor, 
and violence is in the land, 
with ruler against ruler—
\verse \textit{therefore}\lebnote{Literally “to thus”} look, days are coming, 
and I will punish the images of Babylon, 
and all her land will be put to shame, 
and all her slain ones will fall in the midst of her.
\verse Then the heaven and the earth and all that is in them 
will shout for joy over Babylon, 
for from the north 
the destroyers will come to it,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse Babylon must fall not only because of the slain ones of Israel, 
but also because of Babylon the slain ones of all the earth have fallen.
\verse Survivors of the sword, go! 
You must not linger! 
Remember Adonai from far away, 
and let Jerusalem come to your \textit{mind}.\lebnote{Literally “heart”}
\verse We are ashamed, 
for we have heard taunts,\lebnote{Hebrew “taunt”} 
disgrace covers our faces, 
for strangers have come to the holy places of the house of Adonai.
\verse “\textit{Therefore}\lebnote{Literally “To thus”} look, days are coming,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“and I will punish her images, 
and the fatally wounded will sigh through all her land.
\verse Though Babylon goes up to the heavens, 
and though she fortifies the high place of her strength, 
from me destroyers will come to her,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai.
\verse The sound of a cry for help from Babylon, 
and a great collapse from the land of the Chaldeans.
\verse For Adonai is devastating Babylon, 
and he obliterates her loud noise\lebnote{Or “voice”} from her. 
Their waves roar like mighty waters, 
the roar of their voices\lebnote{Hebrew “voice”} resounds.
\verse For a destroyer is coming against her, against Babylon, 
and her warriors are captured. 
Their bows are broken; 
for Adonai is a God of recompense, 
certainly he will repay.
\verse Thus says Adonai of hosts: “The broad walls of Babylon will be utterly demolished, 
and her high gates will burn with fire, 
and the peoples will labor \textit{for nothing},\lebnote{Literally “for enough of empty”} 
and the nations \textit{for fire},\lebnote{Literally “for enough of fire”} 
and they will grow weary.”
\verseWithHeading{The Scroll Concerning Babylon is Thrown into the Euphrates} The word that Jeremiah the prophet commanded Seraiah, the son of Neriah, the son of Mahseiah, at his going with Zedekiah, the king of Judah, to Babylon in the fourth year of his reign. Now Seraiah was the \textit{quartermaster}.\lebnote{Literally “official of the resting place”}
\verse And Jeremiah wrote all the disasters\lebnote{Hebrew “disaster”} that would come on Babylon in one scroll, all these words that are written concerning Babylon.
\verse And Jeremiah said to Seraiah, “At your coming to Babylon, then you must see that you read aloud all these words.
\verse And you must say, ‘Adonai, you yourself spoke against this place, to destroy it, so that there will not be in it anything living, from humankind to animals,\lebnote{Hebrew “animal”} for it will be \textit{an everlasting desolation}.’\lebnote{Literally “a desolation of eternity”}
\verse \textit{And then}\lebnote{Literally “And it will happen”} when you finish reading aloud this scroll, you must tie a stone on it, and you must throw it into the middle of the Euphrates.
\verse And you must say, ‘Babylon will sink, and she will not rise, \textit{because of}\lebnote{Literally “from”} the face of the disasters\lebnote{Hebrew “disaster”} that I am bringing on her, and they will grow weary.’ ” \textit{Thus far}\lebnote{Literally “Up to here”} the words of Jeremiah.
\end{biblechapter}

\begin{biblechapter} % Jeremiah 52
\verseWithHeading{The Fall of Jerusalem} Zedekiah was \textit{twenty-one years old}\lebnote{Literally “a son of twenty-one year”} at his beginning to reign, and he reigned eleven years\lebnote{Hebrew “year”} in Jerusalem. And the name of his mother was Hamutal, the daughter of Jeremiah of Libnah.
\verse And he did evil in the eyes of Adonai like all that Jehoiakim had done.
\verse For because of the \textit{anger}\lebnote{Literally “nose”} of Adonai this happened in Jerusalem and Judah until his casting them from his \textit{presence}.\lebnote{Literally “face”} And Zedekiah rebelled against the king of Babylon.
\verse \textit{And then}\lebnote{Literally “And it was”} in the ninth year of his reign, in the tenth month, on the tenth day of the month, Nebuchadnezzar the king of Babylon came against Jerusalem, he and all his army. And they laid siege to it, and built siege works\lebnote{Literally “siege work”} against it all around.
\verse So the city came \textit{under siege}\lebnote{Literally “into the siege”} until the eleventh year of King Zedekiah.
\verse In the fourth month, on the ninth day of the month, the famine in the city became severe and there was no food for the people of the land.
\verse Then the city was breached, and all \textit{the soldiers}\lebnote{Literally “the men of the battle”} fled and went out from the city by night by the way of the gate between the two walls that are at the garden of the king, though the Chaldeans were all around the city. And they went in the direction of the Jordan Valley.\lebnote{Or “Arabah”}
\verse But the army of the Chaldeans pursued after the king and they overtook Zedekiah in the plains of Jericho, and all his army was scattered from him.
\verse Then they captured the king and brought him up to the king of Babylon at Riblah in the land of Hamath, and \textit{he passed sentence on him}.\lebnote{Literally “he spoke to him judgments”}
\verse And the king of Babylon slaughtered the sons of Zedekiah before his eyes, and he also slaughtered all the officials of Judah at Riblah.
\verse Then he made blind the eyes of Zedekiah, and they tied him up with bronze fetters, and the king of Babylon brought him to Babylon. And he put him in \textit{prison}\lebnote{Literally “the house of the watch”} until the day of his death.
\verse Now in the fifth month, on the tenth day of the month, which was the nineteenth year of King Nebuchadnezzar the king of Babylon, Nebuzaradan the captain of the guard,\lebnote{Hebrew “guards”} who stood \textit{before}\lebnote{Literally “to the face of”} the king of Babylon, entered into Jerusalem.
\verse And he burned the \textit{temple}\lebnote{Literally “house”} of Adonai, and the palace\lebnote{Or “house”} of the king, and all the houses of Jerusalem, even every great house he burned with fire.
\verse And all the army of the Chaldeans who were with the captain of the guard\lebnote{Hebrew “guards”} broke down all the walls of Jerusalem all around.
\verse And Nebuzaradan the captain of the guard\lebnote{Hebrew “guards”} deported some of the poor of the people, and the rest of the people who were left in the city, and the deserters who deserted to the king of Babylon, along with the rest of the craftsmen.
\verse But Nebuzaradan the captain of the guard\lebnote{Hebrew “guards”} left some of the poor of the land to serve as vinedressers and farmers.
\verse And the Chaldeans broke the pillars of bronze that were in the \textit{temple}\lebnote{Literally “house”} of Adonai, and the kettle stands and the sea of bronze that were in the \textit{temple}\lebnote{Literally “house”} of Adonai, and they carried all their bronze to Babylon.
\verse And they took with them the pots, and the shovels, and the snuffers, and the sprinkling bowls, and the pans, and all the vessels of bronze which were used in temple service.
\verse And the captain of the guard\lebnote{Hebrew “guards”} took the bowls, and the firepans, and the sprinkling bowls, and the pots, and the lampstands, and the pans, and the libation bowls, \textit{those made of solid gold}\lebnote{Literally “which gold gold”} and \textit{those made of solid silver}.\lebnote{Literally “which silver silver”}
\verse The two pillars, the one sea, and the twelve bronze oxen that were under the kettle stands which King Solomon had made for the \textit{temple}\lebnote{Literally “house”} of Adonai—there was not a weight for the bronze of all these vessels!
\verse Now the pillars, the height of one pillar was eighteen cubits,\lebnote{Hebrew “cubit”} and a thread of twelve cubits\lebnote{Hebrew “cubit”} surrounded it, and its thickness was four fingers, hollowed out.
\verse And a capital upon it was bronze and the height of the one capital was five cubits, and latticework and pomegranates were on the capital on all sides, all of bronze. And like these was the second pillar with pomegranates.
\verse And there were ninety-six pomegranates \textit{on the sides};\lebnote{Literally “breath”} all the pomegranates on the latticework on all sides were a hundred.
\verse Then the captain of the guard\lebnote{Hebrew “guards”} took Seraiah \textit{the chief priest},\lebnote{Literally “the priest of the head”} and Zephaniah \textit{the second priest},\lebnote{Literally “the priest of the second”} and three keepers of the threshold.
\verse And from the city he took one high official who was chief officer over \textit{the soldiers},\lebnote{Literally “the men of the battle”} and seven men of \textit{the king’s advisors}\lebnote{Literally “those who see the face of the king”} who were found in the city, and the secretary of the commander of the army who levied for military service the people of the land, and sixty men\lebnote{Hebrew “man”} of the people of the land who were found in the midst of the city.
\verse Then Nebuzaradan the captain of the guard\lebnote{Hebrew “guards”} took them and brought them to the king of Babylon at Riblah.
\verse And the king of Babylon struck them down and killed them at Riblah in the land of Hamath. So Judah left from its land.
\verse This is the number of the people whom Nebuchadnezzar deported: in the seventh year, three thousand twenty-three Judeans;
\verse in the eighteenth year of Nebuchadnezzar, eight hundred and thirty-two persons\lebnote{Hebrew “person”} from Jerusalem;
\verse in the twenty-third year of Nebuchadnezzar, Nebuzaradan, the captain of the guard,\lebnote{Hebrew “guards”} deported seven hundred and forty-five Judean persons;\lebnote{Hebrew “person”} there were four thousand six hundred persons\lebnote{Hebrew “person”} in all.
\verseWithHeading{An Allowance for Jehoiachin} \textit{And then}\lebnote{Literally “and it was”} in the thirty-seventh year of the exile of Jehoiachin, the king of Judah, in the twelfth month, on the twenty-fifth day of the month, Evil-merodach, the king of Babylon, in the first year of his reign, lifted up the head of Jehoiachin, the king of Judah, and brought him out from \textit{prison}.\lebnote{Literally “the house of the imprisonment”}
\verse Then he spoke with him kindly and gave his seat above the seats\lebnote{Hebrew “seat”} of the kings who were with him in Babylon.
\verse So he changed the garments of his imprisonment and he ate food \textit{before him}\lebnote{Literally “to the face of him”} continually all the days of his life.
\verse And his allowance, a continual allowance was given to him by the king of Babylon \textit{on a daily basis}\lebnote{Literally “a matter of a day in its day”} all the days of his life up to the day of his death.
\end{biblechapter}

