\biblebook{Job}

\begin{biblechapter} % Job 1
\verseWithHeading{Job’s Character and Greatness}{There was a man in the land of Uz whose\lebnote{Hebrew “his”}name was Job. That\lebnote{Hebrew “And that”}man was blameless and upright and God-fearing and turning away from evil.}%
\verse{And seven sons and three daughters were born to him.}%
\verse{Then\lebnote{Hebrew “And”}his livestock came to be seven thousand sheep and goats and three thousand camels and five hundred pairs\lebnote{Hebrew “pair”}of oxen and five hundred female donkeys, and he had very many slaves, and that man was greater than all the people of the east.}%
\verse{And his sons used to go and hold a feast \textit{at each other’s house}\lebnote{Literally “at the house of a man,” or “at the house of each”}on his day, and they would send, and they would invite their three sisters to eat and to drink with them.}%
\verse{\textit{Then when}\lebnote{Literally “And it would happen that,” or “And it would happen when”}the days of the feast had run their course, \textit{Job would send},\lebnote{Literally “and Job would send”}and he would sanctify them. Thus\lebnote{Hebrew “And”}he would arise early in the morning and offer burnt offerings according to the number of all of them, because Job thought,\lebnote{Or “said”}“Perhaps my children have sinned and \textit{cursed}\lebnote{Literally “blessed”— negative meaning by context}God in their heart.” This is what Job used to do \textit{all the time}.\lebnote{Or “always”; literally “all the days”}}%
\verse{And it happened \textit{one day}\lebnote{Literally “the day”}\textit{that}\lebnote{Literally “and”}the sons of God\lebnote{Or “the heavenly beings,” or “the angels”}came to present themselves before Adonai, and Satan\lebnote{Hebrew “the accuser,” or “the adversary”}also came into their midst.}%
\verse{So\lebnote{Hebrew “And”}Adonai said to Satan,\lebnote{Hebrew “the accuser,” or “the adversary”}“From where have you come?” Then\lebnote{Hebrew “And”}Satan\lebnote{Hebrew “the accuser,” or “the adversary”}answered Adonai and said, “From roaming on the earth and from walking about in it.”}%
\verse{So\lebnote{Hebrew “And”}Adonai said to Satan,\lebnote{Hebrew “the accuser,” or “the adversary”}“Have you \textit{considered}\lebnote{Literally “set your heart”}my servant Job? Indeed,\lebnote{Or “For,” or “Because”}there is no one like him on the earth—a blameless man and upright and God-fearing and turning away from evil.”}%
\verse{Then\lebnote{Hebrew “And”}Satan\lebnote{Hebrew “the accuser,” or “the adversary”}answered Adonai and said, “Does Job fear God for nothing?}%
\verse{Have you\lebnote{Emphatic personal pronoun could be translated “you yourself”}not put a fence around him and his household and around \textit{all that belongs to him}\lebnote{Literally “all that is his”}\textit{on every side}?\lebnote{Literally “from around”}You have blessed the work of his hands, and his livestock\lebnote{Or “possessions”}has increased in the land.}%
\verse{But,\lebnote{Hebrew “And”}on the other hand, stretch out\lebnote{Or “please”}your hand and touch \textit{all that belongs to him}\lebnote{Literally “all that is his”}and see \textit{whether}\lebnote{Literally “if not”}he will \textit{curse}\lebnote{Literally “bless”— negative meaning by context}you to your face.”}%
\verse{So\lebnote{Hebrew “And”}Adonai said to Satan,\lebnote{Hebrew “the accuser,” or “the adversary”}“Look, \textit{all that belongs to him}\lebnote{Literally “all that is his”}is in your \textit{power}.\lebnote{Literally “hand”}Only do not stretch out your hand \textit{against}\lebnote{Literally “to,” or “unto”}him.” So\lebnote{Hebrew “And”}Satan\lebnote{Hebrew “the accuser,” or “the adversary”}went out from Adonai’s \textit{presence}.\lebnote{Literally “faces of”}}%
\verseWithHeading{The Adversary’s Attack on Job’s Possessions}{And then there was one day when\lebnote{Hebrew “and”}his sons and his daughters were eating and drinking wine in their firstborn brother’s house.}%
\verse{And a messenger came to Job and said, “The oxen were plowing, and the female donkeys were feeding \textit{beside them}.\lebnote{Literally “on their hands”}}%
\verse{Then\lebnote{Hebrew “And”}the Sabeans \textit{attacked},\lebnote{Literally “fell upon them”}and they took them, and they slew the servants \textit{by the edge of the sword}.\lebnote{Literally “with the mouth of the sword”}But\lebnote{Hebrew “And”}I escaped, even I alone, to tell you.”}%
\verse{While this one was still speaking, \textit{another}\lebnote{Literally “this”}came and said, “The fire of God fell from the heavens, and it blazed up against the sheep and goats and against the servants, and it consumed them. But I escaped, even I alone, to tell you.”}%
\verse{While this one was still speaking, \textit{another}\lebnote{Literally “this”}came and said, “The Chaldeans formed three divisions, and they made a raid on the camels, and they carried them away, and they struck your servants \textit{by the edge of the sword},\lebnote{Literally “with the mouth of the sword”}but I escaped, even I alone, to tell you.”}%
\verse{At the time\lebnote{Hebrew “Until”}this one was speaking, \textit{another}\lebnote{Literally “this”}came and said, “Your sons and your daughters were eating and drinking wine in their firstborn brother’s house.}%
\verse{And behold,\lebnote{Or “Look”}a great wind came from across the desert, and it struck the four corners of the house \textit{so that}\lebnote{Literally “and”}it fell upon the young people, and they died. But I escaped, even I alone, to tell you.”}%
\verse{Then\lebnote{Hebrew “And”}Job arose and tore his outer garment\lebnote{Or “robe”}and shaved his head; \textit{then}\lebnote{Hebrew “and”}he fell upon the ground and he worshiped.}%
\verse{Then\lebnote{Hebrew “And”}he said, “Naked I came out from my mother’s womb, and naked I will return there. Adonai gives, and Adonai takes. Let Adonai’s name be blessed.”}%
\verse{In all this, Job did not sin and did not \textit{charge God with wrongdoing}.\lebnote{Literally “give unseemliness to God”}}%
\end{biblechapter}

\begin{biblechapter} % Job 2
\verseWithHeading{The Adversary’s Attack on Job’s Person}{\textit{And then}\lebnote{Literally “And it came to be,” or “and it happened”}one day the sons of God\lebnote{Or “the heavenly beings,” or “the angels”}came to present themselves before Adonai, and Satan\lebnote{Hebrew “the accuser,” or “the adversary”}also came into their midst to present himself before Adonai.}%
\verse{So\lebnote{Hebrew “And”}Adonai \textit{asked}\lebnote{Literally “said to”}Satan,\lebnote{Hebrew “the accuser,” or “the adversary”}“From where have you come?” And Satan\lebnote{Hebrew “the accuser,” or “the adversary”}answered Adonai and said, “From roaming on the earth and from walking about in it.”}%
\verse{So\lebnote{Hebrew “And”}Adonai \textit{asked}\lebnote{Literally “said to”}Satan,\lebnote{Hebrew “the accuser,” or “the adversary”}“Have you \textit{considered}\lebnote{Literally “set your heart”}my servant Job? Indeed,\lebnote{Or “For” or “Because”}there is no one like him on the earth—a blameless man and upright and God-fearing and turning away from evil. And still he persists in his blamelessness \textit{even though}\lebnote{Hebrew “and”}you incited me against him to destroy him for nothing.”}%
\verse{Then\lebnote{Hebrew “And”}Satan\lebnote{Hebrew “the accuser,” or “the adversary”}answered him and said, “Skin for skin! All that \textit{that}\lebnote{Literally “the”}man has he will give for his life.}%
\verse{But stretch out\lebnote{Or “please”}your hand and touch his bones and his flesh, and see \textit{whether}\lebnote{Literally “if not”}he will \textit{curse}\lebnote{Literally “bless”—negative meaning by context}you to your face.”\lebnote{Hebrew “your faces”}}%
\verse{So\lebnote{Literally “And it happened the/one day”}Adonai said to Satan,\lebnote{Hebrew “the accuser,” or “the adversary”}“\textit{All right},\lebnote{Literally “Look”}he is in your \textit{power}.\lebnote{Literally “hand”}Only spare his life.”}%
\verseWithHeading{Job’s Blameless Behavior}{So\lebnote{Hebrew “And”}Satan\lebnote{Hebrew “the accuser,” or “the adversary”}went out from \textit{Adonai’s presence},\lebnote{Literally “the faces of Adonai”}and he inflicted Job with loathsome skin sores from the sole of his foot up to the crown of his head.}%
\verse{\textit{So}\lebnote{Hebrew “And”}he took for himself a potsherd with which to scrape himself, and he sat in the midst of the ashes.}%
\verse{Then\lebnote{Hebrew “And”}his wife said to him, “Are you still persisting in your blamelessness? \textit{Curse}\lebnote{Literally “Bless”—negative meaning by context}God and die.”}%
\verse{So\lebnote{Hebrew “And”}he said to her, “You speak like one of the foolish women speaks. Indeed, should we receive the good from God, but\lebnote{Or “and”}not receive the evil?” In all this, Job did not sin with his lips.}%
\verse{Thus\lebnote{Hebrew “And”}Job’s three friends heard of this calamity that had come upon him. So\lebnote{Hebrew “And”}each set out from his own place: Eliphaz the Temanite and Bildad the Shuhite and Zophar the Naamathite. And they met\lebnote{Or “they agreed”}together to come to console him and to comfort him.}%
\verse{Thus\lebnote{Hebrew “And”}they lifted up their eyes\lebnote{Or “they gazed”}from afar, but\lebnote{Hebrew “and”}they did not recognize him, so\lebnote{Hebrew “and”}they raised their voice, and they wept, and each man tore his outer garment\lebnote{Or “robe”}and threw dust on their heads \textit{toward the sky}.\lebnote{Or “in the air”; literally “to the heavens”}}%
\verse{Then\lebnote{Hebrew “And”}they sat with him on the ground for seven days and seven nights, but\lebnote{Hebrew “and”}no one spoke\lebnote{Or “there was no speaking”}a word to him because they saw that \textit{his}\lebnote{Literally “the”}suffering was very great.}%
\end{biblechapter}

\begin{biblechapter} % Job 3
\verseWithHeading{Job Regrets His Birth}{\textit{Afterward}\lebnote{Literally “After thus”}Job opened his mouth and cursed\lebnote{A different term than that employed in 1:5, 11; 2:5, 9}his day.}%
\verse{Thus\lebnote{Hebrew “And”}Job \textit{spoke up}\lebnote{Literally “answered”}and said,}%
\verse{“Let the day perish on which I was born, 
and the night that said, ‘A man-child is conceived.’}%
\verse{Let that day become\lebnote{Or “be”}darkness; 
may God not seek it from above, 
nor may daylight shine on it.}%
\verse{Let darkness and deep shadow claim it; 
let clouds\lebnote{Literally “cloud”}settle on it; 
let them\lebnote{Masculine plural referring to all three entities mentioned in this verse}terrify it with the blackness\lebnote{Literally “blacknesses”}of day.}%
\verse{Let darkness seize that night;\lebnote{Literally “That night, let darkness seize it”}
let it not rejoice among the days of the year; 
let it not enter among the number of the months.}%
\verse{Look, let that night become\lebnote{Or “be”}barren; 
let a joyful song not enter it.}%
\verse{Let those who curse the day curse it, 
those who are skilled at rousing Leviathan.}%
\verse{Let the stars of its dawn be dark; 
let it hope for light but\lebnote{Hebrew “and”}there be none, 
and let it not see the eyelids of dawn}%
\verse{because it did not shut the doors of my mother’s womb, 
nor\lebnote{Hebrew “And”}did it hide trouble from my eyes.}%
\verseWithHeading{Job Wishes He Had Died}{“Why did I not die at\lebnote{Literally “from”}birth? 
Why did I not come forth from the womb and expire?}%
\verse{Why did the knees receive me 
and the breasts, that I could suck?}%
\verse{For now I would lie down, and I would be at peace; 
I would be asleep; then \textit{I would be at rest}\lebnote{Literally “it would be at rest for me”}}%
\verse{with kings and counselors of the earth, 
who \textit{rebuild} \lebnote{Literally “build”}ruins for themselves,}%
\verse{or with high officials \textit{who have gold},\lebnote{Literally “gold is for them”}
who fill up their houses with silver.}%
\verse{Or why was I not hidden like a miscarriage, 
like infants who did not see the light?}%
\verse{There the wicked cease from troubling, 
and there the \textit{weary}\lebnote{Literally “weary of strength”}are at rest;}%
\verse{the prisoners are at ease together; 
they do not hear the oppressor’s voice.}%
\verse{The small and the great are there, 
and the slave is free from his masters.\lebnote{Or “master”}}%
\verseWithHeading{Job Wishes He Might Die}{“Why does he\lebnote{Most likely God}give light to one in misery 
and life to those bitter of soul,}%
\verse{who wait for death, but\lebnote{Hebrew “and”}\textit{it does not come},\lebnote{Literally “it is not”}
and search\lebnote{Or “dig”}for it more than for treasures,}%
\verse{who rejoice \textit{exceedingly},\lebnote{Literally “unto rejoicing”}
and they are glad when they find the grave?}%
\verse{Why does he\lebnote{Most likely God}give light to a man whose way is hidden, 
and God has fenced him in all around?}%
\verse{For\lebnote{Or perhaps emphatic, “Indeed”}my sighing comes \textit{before}\lebnote{Or “in place of” (NET); literally “to the faces of”}my bread,\lebnote{Or “food”}
and my groanings gush forth like water}%
\verse{because the dread that I \textit{feel}\lebnote{Literally “dread”}has come upon me, 
and what I feared befalls me.}%
\verse{I am not at ease, and I am not at peace, 
and I do not have rest, thus\lebnote{Hebrew “and”}turmoil has come.”}%
\end{biblechapter}

\begin{biblechapter} % Job 4
\verseWithHeading{Eliphaz’s First Response to Job}{Then\lebnote{Hebrew “And”}Eliphaz the Temanite answered and said,}%
\verse{“If someone would test a word with you, would you be offended? 
But\lebnote{Hebrew “And”}who can refrain from speaking?}%
\verse{Look, you have instructed many, 
and you have strengthened weak hands.}%
\verse{Your words have raised up the one who stumbles, 
and you have strengthened knees giving way.}%
\verse{But now it has come to you, and you are worn out; 
it touches you, and you are horrified.}%
\verse{Is not your fear in God your confidence? 
Is not your hope even\lebnote{Hebrew “and”}the integrity of your ways?}%
\verse{“\textit{Think}\lebnote{Literally “Remember”}now, who has perished who is innocent? 
Or\lebnote{Hebrew “And”}where are the upright destroyed?}%
\verse{Just as I have seen, plowers of mischief 
and sowers of trouble will reap it.}%
\verse{By\lebnote{Or “From”}the breath of God they perish, 
and by\lebnote{Or “from”}the blast of his anger they come to an end.}%
\verse{The roar of the lion and the voice of a lion in its prime, 
and the teeth of the young lions are broken.}%
\verse{The lion is perishing without\lebnote{Or “from lack of”}prey, 
and the lion’s whelps are scattered.}%
\verse{“And a word came stealing to me, 
and my ear received the whisper from it.}%
\verse{Amid troubling thoughts from night visions, 
at the falling of deep sleep on men,}%
\verse{dread met me, and trembling, 
and it made many of my bones shake.}%
\verse{And a spirit glided before my face; 
the hair of my flesh\lebnote{Or “body”}bristled.}%
\verse{It stood still, but\lebnote{Hebrew “and”}I could not recognize its appearance; 
a form was \textit{before}\lebnote{Literally “to before”}my eyes; 
there was a hush, and I heard a voice:}%
\verse{‘Can a human being be more righteous than God, 
or can a man be more pure than his Maker?}%
\verse{Look, he does not trust in his servants 
and he charges his angels with error.}%
\verse{How much more dwellers in clay houses, 
whose foundation is in the dust? 
They are crushed like a moth.}%
\verse{\textit{Between morning and evening}\lebnote{Literally “from morning to the evening”}they are destroyed; 
without anyone regarding it they perish forever.}%
\verse{Is not their tent cord pulled up within them? 
They die, but\lebnote{Hebrew “and”}not in\lebnote{Or “with”}wisdom.’}%
\end{biblechapter}

\begin{biblechapter} % Job 5
\verseWithHeading{Eliphaz’s Response Continues}{“Call now, is there anyone answering you? 
And to which of the holy ones will you turn?}%
\verse{For vexation will slay the fool, 
and jealousy will kill the simple.}%
\verse{I have seen a fool taking root, 
but\lebnote{Hebrew “And”}suddenly I cursed his dwelling.}%
\verse{His children are far from deliverance, 
and they are crushed in the gate, 
and there is no deliverer—}%
\verse{whose harvest the hungry eats, 
and he takes it \textit{from behind}\lebnote{Literally “to from”}the thorns; 
and the thirsty pants after their wealth.}%
\verse{Indeed,\lebnote{Or “For”}mischief does not come from the dust, 
and trouble does not sprout from the earth.}%
\verse{But a human being is born to trouble, 
and \textit{they soar aloft}\lebnote{Literally “they make high to fly”}like \textit{sparks}.\lebnote{Literally “children of flame”}}%
\verse{“But I myself\lebnote{Emphatic personal pronoun}will seek God, 
and to God I would commit my cause.}%
\verse{He is doing great and \textit{unsearchable}\lebnote{Literally “there is no searching”}things, 
marvelous things \textit{without number}\lebnote{Literally “until there is no number”}—}%
\verse{the one who is giving rain on the \textit{surface of}\lebnote{Literally “faces of”}the earth 
and is sending water on the \textit{surface of}\lebnote{Literally “faces of”}the fields,}%
\verse{to set the lowly on high, 
and those mourning are lifted to safety.}%
\verse{He is frustrating the devices of the crafty, 
and their hands do not achieve success.}%
\verse{He is capturing the wise in their craftiness, 
and the schemes of the wily are rushed.}%
\verse{In the daytime they meet with darkness, 
and they grope at noon\lebnote{Or “the noon”}as in the night.}%
\verse{And\lebnote{Or “Thus,” or “So”}he saves from the sword of\lebnote{Literally “from”}their mouth, 
\textit{even}\lebnote{Literally “and”; the entire verse could be translated “So he saves the poor from the sword of their mouth and from the hand of the strong”}the poor from the hand of the strong.}%
\verse{So\lebnote{Or “And”}there is hope\lebnote{Or “hope exists”}for the powerless, 
and wickedness shuts its mouth.}%
\verse{“Look, happy is the human being whom God reproves; 
and you must not despise the discipline of Shaddai,}%
\verse{for he himself\lebnote{Emphatic personal pronoun}wounds, but\lebnote{Hebrew “and”}he binds up; 
he strikes, but\lebnote{Hebrew “and”}his hands heal.}%
\verse{\textit{From}\lebnote{Literally “In,” or “By”}six troubles he will deliver you, 
and in seven evil shall not touch you.}%
\verse{In famine he will redeem you from death, 
and in war from the \textit{power of}\lebnote{Literally “hand of”}the sword.}%
\verse{From the scourge of the tongue you shall be hidden, 
and you shall not be afraid of\lebnote{Hebrew “from”}destruction when it comes.}%
\verse{At destruction and famine\lebnote{Hebrew “at famine”}you shall laugh, 
and you shall not fear the wild animals of the earth.}%
\verse{For your covenant will be with the stones of the field, 
and the wild animals\lebnote{Collective singular}of the field will be at peace with you.}%
\verse{And you shall know that your tent is safe, 
and you will inspect your fold, and you shall not be missing anything.}%
\verse{And you shall know that your offspring are many, 
and your descendants like the vegetation of the earth.}%
\verse{You shall come in maturity to the grave, 
as the raising up of a stack of sheaves in its season.}%
\verse{“Look, we have searched this out—it is true; 
hear it and know it \textit{yourself}.”\lebnote{Literally “you for yourself”}}%
\end{biblechapter}

\begin{biblechapter} % Job 6
\verseWithHeading{Job’s Second Speech: A Response to Eliphaz}{ Then\lebnote{Hebrew “And”}Job answered and said,}%
\verse{“If only my vexation could be well weighed, 
and my calamity could be lifted up together with it in the balances,}%
\verse{for then it would be heavier than the sand of the seas; 
therefore my words have been rash,}%
\verse{for the arrows of Shaddai are in me; 
my spirit drinks their poison; 
the terrors of God are arrayed against me.}%
\verse{Does the wild ass bray over grass, 
or the ox bellow over its fodder?}%
\verse{Can tasteless food be eaten without\lebnote{Hebrew “from without”}salt, 
or is there taste in the white of a marshmallow plant?}%
\verse{\textit{I refused}\lebnote{Literally “My soul/throat refuses”}to touch them; 
they are like \textit{food that will make me ill}.\lebnote{Literally “the illness of my bread/food”}}%
\verse{“\textit{O that}\lebnote{Literally “Who would give”}my request may come, 
and that God may grant my hope,}%
\verse{that\lebnote{Hebrew “and”}God would decide that\lebnote{Hebrew “and”}he would crush me, 
that he would let loose his hand and \textit{kill me}.\lebnote{Literally “he would cut me off”}}%
\verse{But\lebnote{Hebrew “And”}it will still be my consolation, 
and I would recoil in \textit{unrelenting}\lebnote{Literally “he/it does not have compassion,” or “he/it does not have pity,” or “he/it does not spare”}pain, 
for I have not denied the words of the Holy One.}%
\verse{What is my strength, that I should wait? 
And what is my end, that \textit{I should hold out}?\lebnote{Literally “I should make my self long,” or “I should lengthen my self”}}%
\verse{Or is my strength like the strength of stones? 
Or is my flesh bronze?}%
\verse{\textit{Indeed},\lebnote{An interrogative marker plus “if”}my help is not in me, 
and any success is driven from me.}%
\verse{“Loyal love\lebnote{Or “Kindness”}should come for the afflicted from his friend, 
even if\lebnote{Hebrew “and”}he forsakes the fear of Shaddai.}%
\verse{My companions are treacherous like a torrent-bed; 
like a streambed of wadis\lebnote{A seasonal stream that is often dry}they flow away,}%
\verse{which are growing dark because of ice upon them, 
it will pile up snow.}%
\verse{In time they dry up, they disappear; 
when it is hot, they vanish from their place.}%
\verse{The paths of their way wind around; 
they go up into the wasteland, and they perish.}%
\verse{The caravans of Tema looked; 
the traveling merchants of Sheba hope for them.}%
\verse{They are disappointed, because they trusted; 
they came \textit{here}\lebnote{Literally “up to it”}and they are confounded.}%
\verse{“For now you\lebnote{Plural throughout the rest of this chapter}have become \textit{such};\lebnote{Literally “to it,” or “for it”}
you see terrors, and you fear.}%
\verse{Is it because I have said, ‘Give to me,’ 
or,\lebnote{Hebrew “and”}‘Offer a bribe for me from your wealth’?}%
\verse{or,\lebnote{Hebrew “and”}‘Save me from the foe’s hand,’ 
or,\lebnote{Hebrew “And”}‘Ransom me from the tyrants’ hand’?}%
\verse{Teach me, and I myself\lebnote{Emphatic personal pronoun}will be silent; 
and make me understand how I have gone astray.}%
\verse{How painful are \textit{upright words}!\lebnote{Literally “words of uprightness”}
But\lebnote{Hebrew “And”}what does \textit{your reproof}\lebnote{Literally “reproving from you”}reprove?}%
\verse{Do you intend to reprove my words\lebnote{Or “Do you intend to reprove with words”}
and consider the words of a desperate man as wind?}%
\verse{Even over the orphan you would cast the lot, 
and you would bargain over your friend.}%
\verse{“\textit{Therefore}\lebnote{Literally “And now,” or “And so then”}be prepared, turn to me, 
and I surely will not lie to your face.\lebnote{Hebrew “faces”}}%
\verse{Please turn, let no injustice happen; 
indeed,\lebnote{Hebrew “and”}turn, \textit{my righteousness is still intact}.\lebnote{Literally “still my righteousness is in it”}}%
\verse{Is there injustice on my tongue? 
Or can my palate not discern calamity?\lebnote{Or “calamities”}}%
\end{biblechapter}

\begin{biblechapter} % Job 7
\verseWithHeading{Job’s Second Speech: A Response to Eliphaz}{“Does not \textit{a human being\lebnote{Or a collective singular, “human beings”}have hard service}\lebnote{Literally “hard service for a human being”}on earth? 
And are not his\lebnote{Or a collective singular, “their”}days like the days of a laborer?}%
\verse{Like a slave he longs for the shadow, 
and like a laborer he waits for his wages.}%
\verse{So \textit{I had to inherit}\lebnote{Literally “I am allotted to me”}months of worthlessness, 
and nights of misery are apportioned to me.}%
\verse{When I lie down, I say,\lebnote{Hebrew “and I say”}‘When shall I rise?’ 
But\lebnote{Hebrew “And”}the night is long, 
and I have my fill of tossing until dawn.}%
\verse{My body is clothed with maggots and clods of dust; 
my skin hardens, then\lebnote{Hebrew “and”}it gives way again.}%
\verse{“My days are swifter than a weaver’s shuttle, 
and they come to an end \textit{without hope}.\lebnote{Literally “with nothing hope”}}%
\verse{Remember that my life is a breath; 
my eye will not return to see good.}%
\verse{The eye of the one seeing me will not see me; 
your eyes are upon me, but\lebnote{Hebrew “and”}\textit{I will be gone}.\lebnote{Literally “there is not me,” or “I am not”}}%
\verse{A cloud vanishes, and it goes away, 
so he who goes down to Sheol will not come up.}%
\verse{He does not return again to his house, 
and his place does not recognize him again.}%
\verse{“Even\lebnote{Or “I on my part”}I will not restrain my mouth; 
I will speak in my spirit’s anguish; 
I will complain in my inner self’s\lebnote{Or “soul’s”}bitterness.}%
\verse{Am I the sea, or a sea monster, 
that you set a guard over me?}%
\verse{When I say, ‘My bed will comfort me, 
and my couch\lebnote{Or “bed”}will ease my complaint,’}%
\verse{then\lebnote{Hebrew “and”}you terrify me with dreams,\lebnote{Or “the dreams”}
and with visions you terrify\lebnote{Or “dishearten,” or “frighten”}me.}%
\verse{So\lebnote{Or “And”}my inner self\lebnote{Or “soul”}will choose\lebnote{Or “I will choose”}strangling— 
death more than my \textit{existence}.\lebnote{Literally “bones”}}%
\verse{I loathe my life; I would not live forever; 
depart from me, for my days are a breath.}%
\verse{“What is a human being that you make him great 
and that \textit{you fix your mind on him},\lebnote{Literally “you set on him your heart”}}%
\verse{so that\lebnote{Hebrew “and”}you visit him \textit{every morning},\lebnote{Literally “for mornings,” or “at mornings”}
you test him \textit{every moment}?\lebnote{Literally “for moments,” or “at moments”}}%
\verse{\textit{How long}\lebnote{Literally “Like what”}will you not turn away from me? 
Or not leave me alone until I swallow\lebnote{Or “my swallowing”}my spit?}%
\verse{If I have sinned, what have I done to you, watcher of humanity? 
Why have you made me as a target for yourself, 
so that\lebnote{Hebrew “and”}I have become a burden to myself?\lebnote{Some translations translate as “you” (e.g., ESV, NRSV, NIV, NET)}}%
\verse{And why do you not pardon my transgression 
and take away my guilt? 
For now I shall lie in the dust, 
and you will seek me, but\lebnote{Hebrew “and”}\textit{I will be no more}.”\lebnote{Literally “there is not me,” or “I am not”}}%
\end{biblechapter}

\begin{biblechapter} % Job 8
\verseWithHeading{Bildad’s First Response to Job}{Then\lebnote{Hebrew “And”}Bildad the Shuhite answered and said,\lebnote{Hebrew “he said”}}%
\verse{“\textit{How long}\lebnote{Literally “Until when”}will you say these things, 
and the words of your mouth be a mighty wind?}%
\verse{Does God pervert justice, 
\textit{or}\lebnote{Literally “and if”}Shaddai pervert righteousness?}%
\verse{If your children sinned against him, 
then\lebnote{Hebrew “and”}he sent them into the \textit{power}\lebnote{Literally “hand”}of their transgression.}%
\verse{If you yourself\lebnote{Emphatic personal pronoun}would seek God, 
then\lebnote{Hebrew “and”}you would plead to Shaddai for grace.}%
\verse{If you are pure and upright, 
indeed,\lebnote{Or “even,” or “yes”}now he will rouse himself for you, 
and he will restore your righteous abode.}%
\verse{Though\lebnote{Hebrew “And”}your beginning was small, 
your end\lebnote{Hebrew “and your end”}will be very great.}%
\verse{“Indeed,\lebnote{Or “For”}please inquire of former generations, 
and consider \textit{what their ancestors have found},\lebnote{Literally “as to the finding of their fathers”}}%
\verse{for we are of yesterday, and we do not know, 
for our days on earth are a shadow.}%
\verse{Will they themselves\lebnote{Emphatic personal pronoun}not teach you and \textit{tell you}\lebnote{Literally “they will say to you”}
and \textit{utter}\lebnote{Literally “they will bring out”}words from their heart?}%
\verse{Can papyrus grow tall where there is not a marsh? 
Will reeds flourish without water?\lebnote{Or “where there is no water”}}%
\verse{While it is in its flower and is not plucked, 
yet\lebnote{Hebrew “and”}it withers \textit{before}\lebnote{Literally “to the faces of”}all grass.}%
\verse{So are the paths of all who forget God; 
and the hope of the godless will perish,}%
\verse{whose confidence is cut off 
and \textit{whose trust}\lebnote{Literally “his trust”}is a spider’s house.}%
\verse{He will lean himself against his house, but\lebnote{Hebrew “and”}it will not stand; 
he will take hold of it, but\lebnote{Hebrew “and”}it will not endure.}%
\verse{He is \textit{thriving}\lebnote{Literally “soaked,” or “well-watered”}\textit{before}\lebnote{Literally “to the faces of”}the sun, 
and his plant shoots spread over his garden.}%
\verse{His roots twine over a stone heap; 
he sees a house of stone.}%
\verse{If he destroys him from his place, 
then\lebnote{Hebrew “and”}it deceives him, saying, ‘I have never seen you.’}%
\verse{Look, this is the joy of his way, 
and from dust others will spring up.}%
\verse{“Look, God will not reject the blameless, 
and he will not uphold the hand of evildoers.}%
\verse{Yet he will fill your mouth with laughter 
and your lips with a shout of joy.}%
\verse{Those who hate you will be clothed with shame, 
and the tent of the wicked \textit{will be no more}.”\lebnote{Literally “there is not it,” or “it is not”}}%
\end{biblechapter}

\begin{biblechapter} % Job 9
\verseWithHeading{Job’s Third Speech: A Response to Bildad}{Then\lebnote{Hebrew “And”}Job answered and said,}%
\verse{“Truly I know that it is so, 
but\lebnote{Hebrew “and”}how can a human being be just before God?}%
\verse{If he wants to contend with him, 
he cannot answer him one time in a thousand.}%
\verse{He is wise in\lebnote{Hebrew “of”}heart and mighty in\lebnote{Hebrew “of”}strength; 
who has resisted him and succeeded?\lebnote{Literally “he succeeded”}}%
\verse{“He is the one who moves mountains, and they do not know how, 
who overturns them in his anger.}%
\verse{He is the one who shakes the earth from its place, 
and its pillars tremble.}%
\verse{He is the one who commands the sun, and it does not rise, 
\textit{and he seals up the stars}.\lebnote{Literally “and behind the stars he seals up”}}%
\verse{He is the one who alone stretches out the heavens 
and who tramples on the waves of the sea.}%
\verse{He is the one who made the Bear and Orion, 
the Pleiades and the constellations of the south.}%
\verse{He is the one who does great things \textit{beyond understanding}\lebnote{Literally “until there is not understanding”}
and marvelous things \textit{beyond number}.\lebnote{Literally “until there is not number”}}%
\verse{“\textit{If}\lebnote{Literally “Look”}he passes by me, \textit{I would not see him};\lebnote{Literally “and I would not see”}
and if he should move on, \textit{I would not recognize him}.\lebnote{Literally “and I would not recognize him”}}%
\verse{\textit{If}\lebnote{Literally “Look”}he would snatch away, who could turn him? 
Who could say to him, ‘What are you doing?’}%
\verse{God will not turn back his anger; 
beneath him the helpers of Rahab bow.}%
\verse{\textit{How much less}\lebnote{Literally “Also for”}can I myself\lebnote{Emphatic personal pronoun}answer him? 
How can I choose my words with him,}%
\verse{whom I cannot answer, even though I am righteous? 
From\lebnote{Hebrew “To”}my judge I must implore grace.}%
\verse{If I summon him, and he should answer me, 
I do not believe that he will listen to my voice—}%
\verse{who crushes me with a tempest 
and multiplies\lebnote{Hebrew “he multiplies”}my wounds without cause.}%
\verse{He will not allow me to \textit{catch}\lebnote{Literally “return,” or “regain”}my breath; 
rather, he will fill me with bitterness.}%
\verse{If \textit{it is a matter of}\lebnote{Literally “for”; see NASB, NIV, NET}strength, look, he is mighty. 
But\lebnote{Hebrew “And”}if \textit{it is a matter of}\lebnote{Literally “for”; NASB, NIV, NET, NRSV}justice, who can summon me?\lebnote{Or “arraign me”}}%
\verse{Even though I am righteous, my mouth will condemn me; 
even though I am blameless, yet\lebnote{Hebrew “and”}it would pronounce me guilty.}%
\verse{“I am blameless; I do not care about myself;\lebnote{Perhaps the meaning is “but it makes no difference to me” (NLT)}
I loathe my life.}%
\verse{It is all one; therefore I say, 
‘He destroys both the blameless and the wicked.’}%
\verse{When the whip\lebnote{Or “scourge”}kills suddenly, 
he mocks at the despair of the innocent.}%
\verse{The earth is given into the hand of the wicked; 
he covers the face of its judge— 
if it is not he, then who is it?}%
\verse{“And my days are swifter than a runner; 
they flee away; they do not see good.}%
\verse{They go by like papyrus skiffs, 
like an eagle swoops down on its prey.}%
\verse{Though\lebnote{Or “If”}\textit{I say},\lebnote{Literally “my saying”}‘I will forget my complaint; 
I will change my expression, and I will rejoice,’}%
\verse{I become afraid of all my sufferings; 
I know that you do not consider me innocent.}%
\verse{If I shall be declared guilty, 
why then should I labor in vain?}%
\verse{If I wash myself with soap,\lebnote{Or “with snow,” or “in snow”}
and I cleanse my hands with lye,}%
\verse{then you plunge me into the slime pit, 
and my clothes abhor me.}%
\verse{“For\lebnote{Or the emphatic “Indeed”}he is not a mortal like me that I can answer him, 
that \textit{we can come to trial together}.\lebnote{Literally “we come together in the justice,” or “we come together in the judgment”}}%
\verse{There is no arbiter between us 
that he might lay his hand on both of us.}%
\verse{May he remove his rod from me, 
and let his dread not terrify me;}%
\verse{then I would speak and not fear him,\lebnote{Hebrew “and I will not fear him”}
for \textit{in myself I am not fearful}.\lebnote{Literally “not so I with me”}}%
\end{biblechapter}

\begin{biblechapter} % Job 10
\verseWithHeading{Job Continues His Response to Bildad}{“My inner self\lebnote{Or “soul”}loathes my life; 
\textit{I want to give vent to my complaint};\lebnote{Literally “I want to let go upon me my complaint”}
I want to speak out of the bitterness of my inner self.\lebnote{Or “soul”}}%
\verse{I will say to God, ‘You should not condemn me; 
let me know why you contend against me.}%
\verse{Is it good for you that you oppress, 
that you despise the labor of your hands, 
and you smile over the schemes of the wicked?}%
\verse{\textit{Do you have eyes of flesh}?\lebnote{Literally “Are eyes of flesh for you”}
Or \textit{do you see as human beings see}?\lebnote{Literally “as seeing of human beings you see”}}%
\verse{Are your days as the days of human beings, 
or your years as the days of man,}%
\verse{that you seek out my iniquity, 
and you search for my sin}%
\verse{because of your knowledge that I am not guilty, 
and there is no escaping from your hand?}%
\verse{Your hands fashioned me and made me\lebnote{Hebrew “they made me”}\textit{altogether},\lebnote{Literally “together all around”}
then\lebnote{Or “and”}you destroyed me.}%
\verse{Please\lebnote{Or an untranslatable particle expressing urgency}remember that you made me like clay, 
but\lebnote{Hebrew “and”}you turn me into dust again?}%
\verse{Did you not pour me out like milk 
and \textit{curdle me}\lebnote{Literally “you did curdle me”}like cheese?}%
\verse{You clothed me with skin and flesh, 
and you knit me together with bones and sinews.}%
\verse{You have granted me life and loyal love, 
and your care has preserved my spirit.}%
\verse{“ ‘Yet\lebnote{Hebrew “And”}you hid these things in your heart; 
I know that this \textit{was your purpose}.\lebnote{Literally “was with you”}}%
\verse{If I had sinned, then\lebnote{Hebrew “and”}you would be watching me, 
and you would not acquit me of my guilt.}%
\verse{If I am guilty, woe to me! 
But\lebnote{Hebrew “And”}if I am righteous, I dare not lift my head; 
I am filled with shame, and just look at my misery!}%
\verse{And if my head grows bold, you would hunt me as the lion in its prime; 
\textit{and you repeat your exploits against me}.\lebnote{Literally “and you do again you display marvelous powers against me”}}%
\verse{You renew your witnesses against me, 
and you increase your vexation against me; 
\textit{relief forces}\lebnote{Literally “substitute and forces”}are against me.}%
\verse{So\lebnote{Hebrew “And”}why did you bring me forth from the womb? 
I should have passed away, \textit{and no eye should have seen me}.\lebnote{Literally “and an eye should not have seen me”}}%
\verse{I should have been as though I had not been; 
I should have been brought from the womb to the grave.}%
\verse{Are not my days few? Let him leave me alone; 
let him turn from me, and let me rejoice a little.}%
\verse{Before I go—and I will not return— 
to the land of darkness and deep shadow,}%
\verse{to the land of darkness, 
like the darkness of a deep shadow and \textit{chaos},\lebnote{Literally “and not order”}
so that\lebnote{Or “and”}it shines forth like darkness.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Job 11
\verseWithHeading{Zophar’s First Response to Job}{Then\lebnote{Hebrew “And”}Zophar the Naamathite answered and said,}%
\verse{“Should an abundance of words go unanswered, 
or \textit{a man full of talk}\lebnote{Literally “a man of lips”}be vindicated?\lebnote{Hebrew “should he be vindicated”}}%
\verse{Should your loose talk put people to silence? 
\textit{And when you mock, shall no one put you to shame}?\lebnote{Literally “And you will mock there is not putting to shame”}}%
\verse{For you say, ‘My teaching is pure, 
and I am clean in your sight.’}%
\verse{But,\lebnote{Hebrew “And but”}\textit{O that}\lebnote{Literally “who shall give”}God might speak, 
and that he would open his lips to you,}%
\verse{and that he would tell you the secrets of wisdom, 
for \textit{insight has many sides}.\lebnote{Literally “double to sound wisdom”}
And know that God \textit{on your behalf}\lebnote{Literally “for you”}has forgotten \textit{some of}\lebnote{Literally “from”}your guilt.}%
\verse{“Can you find out the essence of God, 
or can you find out \textit{the ultimate limits}\lebnote{Literally “up to the limit”}of Shaddai?}%
\verse{\textit{It is higher than the heaven};\lebnote{Literally “heights of heaven”}what can you do? 
It is deeper than Sheol; what can you know?}%
\verse{Its measure is longer than the earth 
and broader than the sea.}%
\verse{“If he passes through and imprisons someone\lebnote{Hebrew “and he imprisons”}
\textit{and summons the assembly},\lebnote{Literally “and he summons”}then\lebnote{Hebrew “and”}who can hinder him?}%
\verse{For he knows \textit{those who are worthless};\lebnote{Literally “men of worthlessness”}
\textit{when he sees}\lebnote{Literally “And he will see”}iniquity, \textit{he will not consider it}.\lebnote{Literally “and he will not consider”}}%
\verse{But\lebnote{Hebrew “And”}\textit{an empty-headed person}\lebnote{Literally “a man hollowed out”}will get understanding 
when\lebnote{Hebrew “and”}a wild donkey’s colt is born as a human being.}%
\verse{“If you yourself\lebnote{Emphatic personal pronoun}direct your heart 
and stretch out your hands to him—}%
\verse{if iniquity\lebnote{See v. 11, where the same word is used}is in your hand, put it far away, 
and you must not let wickedness reside in your tents—}%
\verse{surely then you will lift up your face without\lebnote{Hebrew “from”}blemish, 
and you will be firmly established and will not fear.\lebnote{Hebrew “and not will you fear”}}%
\verse{For you yourself\lebnote{Emphatic personal pronoun}will forget your misery; 
you will remember it as water that has flowed past.}%
\verse{“\textit{And your life will be brighter than noon};\lebnote{Literally “And from noon lifespan will arise”}
its darkness will be like the morning.}%
\verse{And you will have confidence because there is hope; 
and you will be well protected—\textit{you will sleep in safety}.\lebnote{Literally “you will lie down securely”}}%
\verse{And you will lie down, and \textit{no one will make you afraid};\lebnote{Literally “there is not making afraid”}
\textit{and many will entreat your favor}.\lebnote{Literally “and they will appease/implore your faces many”}}%
\verse{But\lebnote{Hebrew “And”}the eyes of the wicked will fail; 
and refuge will be lost to them, 
and their hope is \textit{to breathe their last breath}.”\lebnote{Literally “the expiring of soul”}}%
\end{biblechapter}

\begin{biblechapter} % Job 12
\verseWithHeading{Job’s Fourth Speech}{ Then\lebnote{Hebrew “And”}Job answered and said,}%
\verse{“Truly indeed you\lebnote{Plural}are the people, 
and wisdom will die with you.\lebnote{Plural}}%
\verse{\textit{I also have insight}\lebnote{Literally “Also for me heart”}like you;\lebnote{Plural}
I am not more inferior than you.\lebnote{Plural}
And \textit{who does not know things like these}?\lebnote{Literally “and who there is not like these”}}%
\verse{I am a laughingstock to my friends:\lebnote{Hebrew “to his friends”}
‘He calls on God, and he answers him.’ 
A righteous, blameless man is a laughingstock.}%
\verse{\textit{Those at ease have contempt}\lebnote{Literally “Contempt is according to the thought of the complacent”}for the thought of \textit{disaster},\lebnote{Literally “a torch”}
but it is ready for those unstable of foot.}%
\verse{The tents of the destroyers are at peace, 
and there is security for those who provoke God, 
for those whom God brings into his hand.\lebnote{Or “power”}}%
\verse{“But\lebnote{Hebrew “And”}ask\lebnote{Or “ask now,” or “please ask”}the animals, and they will teach you, 
and the birds of the heaven, and they will tell you;}%
\verse{or ask the earth, and it will teach you, 
and the fishes of the sea will declare to you.}%
\verse{Who among all of these does not know 
that Adonai’s hand has done this?}%
\verse{In whose hand is the life of all living things 
and \textit{the breath of every human being}?\lebnote{Literally “the breath of all the flesh of man”}}%
\verse{Does not the ear test words 
and the palate taste food for itself?}%
\verse{Wisdom is with the aged, 
and understanding is in length of days.}%
\verse{“With him are wisdom and powerful deeds, 
and to him belong counsel and understanding.}%
\verse{If he tears down, then\lebnote{Hebrew “and”}it will not be rebuilt; 
if he shuts a man in, then\lebnote{Hebrew “and”}he cannot be freed.}%
\verse{Look, if he withholds the water,\lebnote{Or “waters”}then\lebnote{Hebrew “and”}they dry up; 
and if he sends them out, then\lebnote{Hebrew “and”}they overwhelm the land.}%
\verse{“Strength and sound wisdom are with him; 
the deceived and the deceiver are his.\lebnote{Or “for him”}}%
\verse{He leads counselors away stripped, 
and he makes fools of judges.}%
\verse{He loosens the fetters of kings, 
and he binds a loincloth on their loins.}%
\verse{He leads priests away stripped, 
and he overthrows \textit{the members of ancient families}.\lebnote{Literally “the constant”}}%
\verse{He deprives the trustworthy of speech, 
and he takes away the discretion of elders.}%
\verse{He pours contempt on noblemen, 
and he loosens the girdle of the mighty.}%
\verse{“He uncovers mysteries out of darkness, 
and he brings deep shadow to the light.}%
\verse{He makes the nations great, then\lebnote{Hebrew “and”}he destroys them; 
he expands the nations, then\lebnote{Hebrew “and”}he guides them.}%
\verse{He strips away the insight of the heads of the earth’s people, 
and he makes them wander in \textit{a pathless wasteland}.\lebnote{Literally “in a wasteland not a way”}}%
\verse{They grope in the dark \textit{without}\lebnote{Literally “and not”}light, 
and he makes them stagger like a\lebnote{Hebrew “the”}drunkard.}%
\end{biblechapter}

\begin{biblechapter} % Job 13
\verseWithHeading{Job’s Fourth Speech Continues}{“Look, my eye has seen everything; 
my ear has heard and has understood it.}%
\verse{\textit{What you\lebnote{Plural}know},\lebnote{Literally “Like your knowledge”}I myself\lebnote{Emphatic personal pronoun}also know— 
I am not more inferior than you.\lebnote{Plural}}%
\verse{But I would speak to Shaddai, 
and I desire to argue with God.}%
\verse{“But\lebnote{Or “And but”}you\lebnote{Plural}\textit{whitewash with lies};\lebnote{Literally “coverers of lies”}
all of you\lebnote{Plural}are \textit{worthless healers}.\lebnote{Literally “healers of worthlessness”}}%
\verse{\textit{O that}\lebnote{Literally “Who shall give”}you\lebnote{Plural}would keep completely silent, 
and that it would become wisdom for you.\lebnote{Plural}}%
\verse{Please hear my argument, 
and listen attentively to the pleadings of my lips.}%
\verse{“Will you\lebnote{Plural}speak falsely for God? 
And will you\lebnote{Plural}speak deceitfully for him?}%
\verse{\textit{Will you\lebnote{Plural}show partiality for him}?\lebnote{Literally “his faces will you raise”}
Or do you\lebnote{Plural}want to plead God’s case?}%
\verse{Will it be well, if he examines you?\lebnote{Plural}
Or can you\lebnote{Plural}deceive him like deceiving a human being?}%
\verse{“Surely he will rebuke you\lebnote{Plural}
if \textit{you\lebnote{Plural}show partiality}\lebnote{Literally “faces you lift up”}in secret.}%
\verse{Will not his majesty terrify you,\lebnote{Plural}
and his dread fall upon you?\lebnote{Plural}}%
\verse{Your\lebnote{Plural}maxims are proverbs of ashes; 
your\lebnote{Plural}defenses are defenses of clay.}%
\verse{“\textit{Let me have silence},\lebnote{Literally “Be silent away from me”}and I myself\lebnote{Emphatic personal pronoun}will speak, 
and let come over me whatever may.}%
\verse{Why should I take my flesh in my teeth 
and\lebnote{Hebrew “now”}put\lebnote{Or “should I put,” or “will I put”}my life in my hand?}%
\verse{Look,\lebnote{Or “Though”}though he kill me, I will hope in\lebnote{Or “for”}him; 
however, I will defend my ways \textit{before him}.\lebnote{Literally “to his faces”}}%
\verse{Moreover, this is salvation to me, 
that the godless would not come \textit{before him}.\lebnote{Literally “to his faces”}}%
\verse{“Listen carefully to my words, 
and let my exposition be in your ears.}%
\verse{Please look,\lebnote{Or “Look here”}I have prepared my case; 
I know that I myself\lebnote{Emphatic personal pronoun}will be vindicated.}%
\verse{Who is he who will contend with me? 
For \textit{then} I would be silent, and I would pass away.}%
\verseWithHeading{Job Argues His Case with God}{“Only you\lebnote{Singular; the antecedent is God}must not do these two things to me; 
then I will not hide from your face:}%
\verse{withdraw your\lebnote{Singular}hand from me, 
and let not your\lebnote{Singular}dread terrify me.}%
\verse{Then\lebnote{Hebrew “And”}call, and I myself\lebnote{Emphatic personal pronoun}will answer; 
or let me speak, then\lebnote{Hebrew “and”}reply to me.}%
\verse{“\textit{How many}\lebnote{Literally “Like what”}are my iniquities and sins? 
Make known to me my transgression and my sin.}%
\verse{Why do you\lebnote{Singular}hide your\lebnote{Singular}face 
and count me as your\lebnote{Singular}enemy?}%
\verse{Will you\lebnote{Singular}terrify a blown leaf? 
And will you\lebnote{Singular}pursue dry stubble?}%
\verse{“Indeed, you\lebnote{Singular}write bitter things against me, 
and you\lebnote{Singular}make me reap the iniquities of my childhood.}%
\verse{And you\lebnote{Singular}put my feet in the block, 
and you\lebnote{Singular}watch all my paths; 
you carve a mark on the soles of my feet.\lebnote{Or “You mark/note my footprints”}}%
\verse{And he himself \lebnote{Emphatic personal pronoun}wastes away like something rotten, 
like a garment that the moth has eaten.}%
\end{biblechapter}

\begin{biblechapter} % Job 14
\verseWithHeading{Job Continues to Argue His Case with God}{“A human being born of a woman 
is short of days and full of troubles.\lebnote{Literally “trouble”}}%
\verse{Like a flower he comes up, and he withers away; 
and he flees like a\lebnote{Or “the”}shadow, but\lebnote{Hebrew “and”}he does not last.}%
\verse{Even on such a one you fix your eyes, 
and you bring me into judgment with you.}%
\verse{“Who can bring a clean thing from an unclean thing? 
No one!\lebnote{Or “Not one”}}%
\verse{If his days are determined, the number of his months is with you; 
you have appointed his boundaries, and he cannot cross them.}%
\verse{Look away from him, and let him desist 
until he enjoys his days like a laborer.}%
\verse{“Indeed, there is hope for a tree: 
if it is cut down, then\lebnote{Hebrew “and”}it will sprout again, 
and its new shoots will not cease;}%
\verse{though its root grows old in the earth, 
and its stump dies in the dust,}%
\verse{at\lebnote{Hebrew “from”}the scent of water it will bud, 
and it will put forth branches like a young plant.}%
\verse{“But\lebnote{Hebrew “And”}a man dies, and he dwindles away; 
thus\lebnote{Hebrew “And”}a human being passes away, and where is he?\lebnote{Or “where is he then?”}}%
\verse{As water disappears from a lake, 
and a river withers away and dries up,\lebnote{Or “it dries up”}}%
\verse{\textit{so}\lebnote{Literally “and”}a man lies down, and he does not arise. 
\textit{Until the heavens are no more},\lebnote{Literally “Until not heavens”}they will not awaken, 
and they will not be roused out of their sleep.}%
\verse{“\textit{O that}\lebnote{Literally “Who shall give”}you would conceal me in Sheol, 
that you would hide me \textit{until your wrath is past},\lebnote{Literally “until the return of your nose”}
that you would appoint a set time for me and remember me.}%
\verse{If a man dies, will he live again? 
All the days of my compulsory service I will wait, 
until the coming of my relief.}%
\verse{You would call, and I myself\lebnote{Emphatic personal pronoun}would answer you; 
you would long for the work of your hand.}%
\verse{For then\lebnote{Hebrew “now”}you would count my steps, 
but you would not keep watch over my sin.}%
\verse{My transgression would be sealed in a bag, 
and you would cover over my guilt.}%
\verse{“But\lebnote{Or “And but”}a mountain falls; it crumbles away, 
and a rock moves away from its place.}%
\verse{Water wears away stones; 
its torrents wash away the soil of the earth; 
so\lebnote{Hebrew “and”}you destroy the hope of human beings.}%
\verse{You overpower him forever, and he passes away; 
you change \textit{his countenance},\lebnote{Literally “his faces”}\textit{then}\lebnote{Hebrew “and”}you send him away.}%
\verse{His children may come to honor, but\lebnote{Hebrew “and”}he does not know it; 
or\lebnote{Hebrew “and”}they may become lowly, but\lebnote{Hebrew “and”}he does not realize it.\lebnote{Or “them”}}%
\verse{He feels only the pain of his own body,\lebnote{Hebrew “his body on him”}
and his inner self\lebnote{Or “soul”}mourns for him.”\lebnote{Or “for himself”}}%
\end{biblechapter}

\begin{biblechapter} % Job 15
\verseWithHeading{Eliphaz’s Second Response to Job}{Then\lebnote{Hebrew “And”}Eliphaz the Temanite answered and said,}%
\verse{“Should the wise answer with windy knowledge, 
and should he fill his stomach with the east wind?}%
\verse{Should he argue in talk that is not profitable 
or in words with which he cannot do good?}%
\verse{“\textit{What is worse},\lebnote{Literally “Also,” or “Even”}you yourself\lebnote{Emphatic personal pronoun}are doing away with fear, 
and you are lessening meditation \textit{before}\lebnote{Literally “to the faces of”}God.}%
\verse{For your iniquity teaches your mouth, 
and you choose the tongue of the crafty.}%
\verse{Your mouth condemns you, and not I; 
and your lips testify against you.}%
\verse{“Were you born the firstborn of the human race? 
And were you brought forth \textit{before}\lebnote{Literally “to the faces of”}the hills?}%
\verse{Have you listened in God’s confidential discussion? 
And do you limit wisdom to yourself?}%
\verse{What do you know that we do not know? 
What do you understand that is not clear to us?}%
\verse{Both the gray-haired and the old are among us— 
\textit{those older than your father}.\lebnote{Literally “more aged than your father in days”}}%
\verse{“Are the consolations of God too small for you, 
a word spoken gently with you?}%
\verse{Why does your heart carry you away? 
And why do your eyes flash,}%
\verse{that you turn your spirit against God, 
and you let such words go out of your mouth?}%
\verse{“What is a human being, that he can be clean, 
or\lebnote{Or “and”}that one born of a woman can be righteous?}%
\verse{Look, he does not trust his holy ones, 
and the heavens are not clean in his eyes.}%
\verse{\textit{How much less}\lebnote{Literally “Also for,” or “Indeed for,” or “Indeed that”}he who is abominable and corrupt,\lebnote{Or “he who is corrupt”}
a man drinking wickedness like water.}%
\verse{“I will show you, listen to me; 
and what I have seen, I will tell\lebnote{Literally “and I will tell”}—}%
\verse{what wise men have told, 
and they have not hidden that which is from their ancestors,}%
\verse{\textit{to whom alone}\lebnote{Literally “to them to alone them”}the land was given, 
and no stranger passed through their midst.}%
\verse{“All of the wicked one’s days he is writhing, 
even\lebnote{Hebrew “and”}through the number of years that are laid up for the tyrant.}%
\verse{Sounds of terror are in his ears; 
in prosperity the destroyer will come against him.}%
\verse{\textit{He cannot trust that he will return}\lebnote{Literally “He cannot trust to return,” or “He is not certain to return,” or “He cannot be sure to return” (compare NJPS)}from darkness, 
and he himself\lebnote{Emphatic personal pronoun}is destined for the sword.}%
\verse{“He is wandering for bread, saying, ‘Where is it?’ 
He knows that a day of darkness is ready \textit{at hand}.\lebnote{Hebrew “at his hand”; the meaning seems to be, “his ruin is certain”}}%
\verse{Anguish and distress terrify him; 
they\lebnote{Hebrew “it”}overpower him like a king ready for the battle.}%
\verse{Because he stretched out his hand against God, 
and he was arrogant to Shaddai;}%
\verse{he \textit{stubbornly}\lebnote{Literally “with neck”}runs against him 
\textit{with his thick-bossed shield}.\lebnote{Literally “with the thickness of the boss of his shield”}}%
\verse{“Because he has covered his face with his fat 
and has gathered fat upon his loins,}%
\verse{\textit{he will dwell}\lebnote{Hebrew “and he will dwell”}in desolate cities, 
in houses that they should not inhabit, 
which are destined for rubble.}%
\verse{He will not become rich, and his wealth will not endure, 
and their possessions will not stretch across the earth.\lebnote{See HCSB, ESV; or “will not reach to the earth”; perhaps “his possessions will not go to the underworld”}}%
\verse{“He will not escape from darkness; 
a flame will dry up his new shoot, 
and by the wind of his mouth he shall be removed.}%
\verse{Let him not trust in emptiness—he will be deceiving himself— 
for worthlessness will be his recompense.}%
\verse{It will be paid in full \textit{before his time},\lebnote{Literally “in not his day,” or “without his day”}
and his branch will not flourish.}%
\verse{“He will shake off his unripe fruit like the vine, 
and he will cast off his blossom like the olive tree;}%
\verse{for the company of the godless is barren, 
and fire consumes the tents of those who accept bribes.}%
\verse{They conceive trouble and bring forth mischief, 
and their womb prepares deceit.”}%
\end{biblechapter}

\begin{biblechapter} % Job 16
\verseWithHeading{Job’s Fifth Speech}{Then\lebnote{Hebrew “And”}Job answered and said,}%
\verse{“I have heard many things like these; 
all of you are \textit{miserable comforters}.\lebnote{Literally “comforters of trouble”}}%
\verse{Is there a limit to windy words? 
What provokes you that you answer?}%
\verse{I myself\lebnote{Emphatic personal pronoun;}also could talk as you, 
if \textit{you were in my place};\lebnote{Literally “there is your soul in place of my soul”}
I could join against you with words, 
and I could shake at you with my head.}%
\verse{I could\lebnote{Or “would”}encourage you with my mouth, 
and the solace of my lips would\lebnote{Or “should”}ease the pain.}%
\verse{If I speak, my pain is not relieved; 
and if I cease, how much will leave me?}%
\verse{“Surely now he has worn me out; 
you\lebnote{Singular}have devastated all my company.}%
\verse{Thus\lebnote{Hebrew “And”}you shriveled me up;\lebnote{Or “you have seized me”}
it became a witness. 
And my leanness has risen up against me; 
it testifies to my face.}%
\verse{His wrath has torn, and he has been hostile toward me; 
he gnashed at me with his teeth. 
My foe sharpens his eyes against me.}%
\verse{They gaped at me with their mouth; 
they struck my cheeks with disgrace; 
they have massed themselves together against me.}%
\verse{God delivers me to an evil one, 
and he casts me into the hands of the wicked.}%
\verse{“I was at ease, then\lebnote{Hebrew “and”}he broke me in two, 
and he seized me by my neck; 
then\lebnote{Hebrew “and”}he shattered me 
and set me up as a target for him.}%
\verse{His archers surround me; 
he slashes open my kidneys, and he does not have compassion; 
he pours out my gall on the ground.}%
\verse{He breached me \textit{breach upon breach};\lebnote{Literally “breach upon the faces of breach”}
he rushes at me like a warrior.}%
\verse{“I have sewed sackcloth on my skin, 
and I have inserted \textit{my pride}\lebnote{Literally “my horn”}in the dust.}%
\verse{My face is red because of weeping, 
and deep shadows are on my eyelids,}%
\verse{\textit{although}\lebnote{Literally “upon,” or “because”}violence is not on my hands, 
and my prayer is pure.}%
\verse{“O earth, you should not cover my blood, 
and let there be no place\lebnote{Or “do not let it become a place”}for my cry for help.}%
\verse{So now look, my witness is in the heavens, 
and he who vouches for me is in the heights.}%
\verse{My friends scorn me; 
my eye pours out tears to God,}%
\verse{and it argues\lebnote{Or “but may someone argue”}for a mortal with God, 
and as \textit{a human}\lebnote{Literally “a son of man”}for his friend.}%
\verse{Indeed, after \textit{a few years}\lebnote{Literally “years of number”}have come, 
then\lebnote{Hebrew “and”}I will go the way from which I will not return.}%
\end{biblechapter}

\begin{biblechapter} % Job 17
\verseWithHeading{Job’s Fifth Speech, Continued}{“My spirit is pulled down; my days are extinguished; 
the graveyard is for me.}%
\verse{\textit{Surely}\lebnote{Literally “If not”}mockery is with me, 
and my eye \textit{rests}\lebnote{Literally “spends the night”}on their provocation.}%
\verse{Please lay down a pledge for me with yourself; 
who is he who will give security for my hand?\lebnote{Or “who will give security for me?”}}%
\verse{Indeed,\lebnote{Or “For”}you have closed their mind from understanding; 
therefore, you will not let them triumph.}%
\verse{He denounces friends for reward, 
so\lebnote{Hebrew “and”}his children’s eyes will fail.}%
\verse{“And he has made me a proverb for\lebnote{Hebrew “of”}the peoples, 
\textit{and I am one before whom people spit}.\lebnote{Literally “and spitting to the faces I am”}}%
\verse{And my eye has grown dim from grief, 
and the limbs of my body are all\lebnote{Hebrew “all of them”}like a shadow.}%
\verse{The upright are appalled at this, 
and the innocent excites himself over the godless.}%
\verse{But\lebnote{Hebrew “And”}the righteous holds on to his way, 
and \textit{he who has clean hands}\lebnote{Literally “the clean of hands”}increases in strength.}%
\verse{But\lebnote{Literally “And but”}\textit{all of you must return}\lebnote{Literally “all of them you must return”}—\textit{please come}!\lebnote{Literally “and please come,” or “come on!”}
But\lebnote{Hebrew “And”}I shall not find a wise person among you.}%
\verse{“My days are past; my plans are broken down— 
even the desires of my heart.}%
\verse{They make night into day, 
saying, ‘Light is \textit{near to darkness}.’\lebnote{Literally “near from the faces of darkness”}}%
\verse{If I hope for Sheol as my house, 
if I spread my couch in the darkness,}%
\verse{if I call to the pit, ‘You are my father,’ 
to the maggot, ‘You are my mother or\lebnote{Or “and”}my sister,’}%
\verse{where\lebnote{Hebrew “and where”}then is my hope? 
And \textit{who will see my hope}?\lebnote{Literally “my hope who will see it”}}%
\verse{Will they go down to the bars of Sheol? 
Or shall we descend together into the dust?”}%
\end{biblechapter}

\begin{biblechapter} % Job 18
\verseWithHeading{Bildad’s Second Speech}{Then\lebnote{Hebrew “And”}Bildad the Shuhite answered and said,}%
\verse{“\textit{How long}\lebnote{Literally “Until when”}\textit{will you\lebnote{Plural}hunt for words}?\lebnote{Literally “will you set a snare for words”}
Consider,\lebnote{Or “you should consider”; plural second person}and afterward we shall speak.}%
\verse{Why are we considered as animals? 
Why are we taken as\lebnote{Or “considered”}stupid in your\lebnote{Plural}eyes?}%
\verse{“You who are tearing \textit{yourself}\lebnote{Literally “his soul,” or “his individual self”}in your\lebnote{Hebrew “his”}anger, 
will the earth be forsaken because of you?\lebnote{Singular}
Or\lebnote{Hebrew “And”}will the rock be removed from its place?}%
\verse{Furthermore,\lebnote{Or “Also”}the light of the wicked is put out, 
and the flame of his fire will not shine.}%
\verse{The light becomes dark in his tent, 
and his lamp above him is put out.}%
\verse{“\textit{His strong steps}\lebnote{Literally “The steps of his power”}are shortened, 
and his own schemes\lebnote{Collective singular}throw him down,}%
\verse{for he is thrust into a net by his feet, 
and he walks into a pitfall.}%
\verse{A trap seizes him by the heel; 
a snare takes hold of him.}%
\verse{His rope is hidden in the ground, 
and his trap on the path.}%
\verse{“Sudden terrors terrify him all around, 
and they chase him at his heels.}%
\verse{His wealth will become hunger, 
and disaster is ready for his stumbling.}%
\verse{It consumes parts of his skin; 
the firstborn of death consumes his limbs.}%
\verse{He is torn from his tent \textit{in which he trusted},\lebnote{Literally “his trust”}
and it brought him to the king of terrors.}%
\verse{“\textit{Nothing}\lebnote{Literally “From not”}remains for him in his tent; 
sulfur is scattered upon his dwelling place.}%
\verse{His roots dry up \textit{below},\lebnote{Literally “from under”}
and its branches\lebnote{Singular collective}wither away above.\lebnote{Hebrew “from above”}}%
\verse{His remembrance perishes from the earth, 
and there is not a name for him on the street.}%
\verse{“They thrust him from light into darkness, 
and they drive him out from the world.}%
\verse{There is no offspring for him nor\lebnote{Hebrew “and not”}a descendant among his people, 
and there is not a survivor in his abode.}%
\verse{Those of the west are appalled over \textit{his fate},\lebnote{Literally “his day”}
and those of the east \textit{are seized with horror}.\lebnote{Literally “they seize horror”}}%
\verse{Surely these are the dwellings of the godless, 
and this is the dwelling place of him who knows not God.”}%
\end{biblechapter}

\begin{biblechapter} % Job 19
\verseWithHeading{Job’s Sixth Speech: A Response to Bildad}{ Then\lebnote{Hebrew “And”}Job answered and said,}%
\verse{“\textit{How long}\lebnote{Literally “Until when”}will you\lebnote{Plural}torment me\lebnote{Hebrew “my soul,” or “my self”}
and crush me with words?}%
\verse{These ten times you\lebnote{Plural}have disgraced me; 
you\lebnote{Plural}are not ashamed that you\lebnote{Plural}have attacked\lebnote{Or “astonished”}me.}%
\verse{And what is more, if I have truly erred, 
my error remains with me.}%
\verse{If indeed you\lebnote{Plural}must magnify yourselves against me, 
and you\lebnote{Plural}must let my disgrace argue against me,}%
\verse{know then that God has wronged me 
and has surrounded me with his net.}%
\verse{“Look, I cry out, ‘Violence!’ but\lebnote{Or “and”}I am not answered; 
I cry out, but\lebnote{Hebrew “and”}there is no justice.}%
\verse{He has walled up my way so that\lebnote{Hebrew “and”}I cannot pass; 
and he has set darkness upon my paths.}%
\verse{He has taken my glory from me, 
and he has removed the crown of\lebnote{Or “from”}my head.}%
\verse{He has broken me down all around, and I am gone. 
And he has uprooted my hope like a tree,}%
\verse{and he has kindled his wrath against me, 
and \textit{he has counted me as one of his foes}.\lebnote{Literally “he has counted me for him as his foes”}}%
\verse{His troops have come together 
and have thrown up their rampart\lebnote{Or “siege works”; compare NRSV}against me 
and have encamped around my tent.}%
\verse{“He has removed my kinsfolk from me, 
and my acquaintances have only\lebnote{Or “completely”}turned aside from me.}%
\verse{My relatives have failed, 
and my close friends have forgotten me.}%
\verse{The sojourners in\lebnote{Or “of”}my house and my slave women count me as a stranger; 
I have become a foreigner in their eyes.}%
\verse{I call to my servant, but\lebnote{Hebrew “and”}he does not answer; 
I must \textit{personally}\lebnote{Literally “with my mouth”}plead with him.}%
\verse{My breath is repulsive to my wife, 
and I am loathsome to \textit{my own family}.\lebnote{Literally “to the children of my womb”}}%
\verse{Little boys also despise me; 
when I rise, then\lebnote{Hebrew “and”}they talk against me.}%
\verse{All \textit{my intimate friends}\lebnote{Literally “the men of the circle of my confidants”}abhor me, 
and these whom I have loved have turned against me.}%
\verse{My bones\lebnote{Collective singular}cling to my skin and to my flesh, 
and I have escaped by the skin of my teeth.}%
\verse{“Pity me, pity me, you my friends, 
for God’s hand has touched me.}%
\verse{Why do you\lebnote{Plural}pursue me like God? 
And are not satisfied with my flesh?}%
\verse{“\textit{O that}\lebnote{Literally “Who then shall give and”}my words could be written down! 
\textit{O that they could be inscribed in a scroll}!\lebnote{Literally “Who shall give in the scroll and they could be inscribed”}}%
\verse{That with a pen of iron and with lead 
they might be engraved on a rock forever!}%
\verse{But\lebnote{Hebrew “And”}I myself\lebnote{Emphatic personal pronoun}know that my redeemer is alive, 
and at the last he will stand up upon \textit{the earth}.\lebnote{Literally “dust”}}%
\verse{And after my skin has been thus destroyed, 
but\lebnote{Hebrew “and”}from\lebnote{Or “without”}my flesh I will see God,}%
\verse{whom \textit{I will see for myself},\lebnote{Literally “I myself will see for myself”; emphatic personal pronoun as subject to the singular verb}
and whom my eyes will see and not a stranger.\lebnote{Or “another”}
\textit{My heart faints within me}.\lebnote{Literally “My kidneys fail in my lap”; see NRSV}}%
\verse{“If you\lebnote{Plural}say, ‘How will we persecute him?’ 
And ‘The root of the trouble is found’ in me,\lebnote{Or “ ‘The root of the trouble is found’ in him”}}%
\verse{be afraid for yourselves \textit{because of the sword},\lebnote{Literally “from the faces of the sword”}
for wrath brings punishment by\lebnote{Literally “of”}the sword, 
so that you may know that there is judgment.”}%
\end{biblechapter}

\begin{biblechapter} % Job 20
\verseWithHeading{Zophar’s Second Speech}{ Then\lebnote{Hebrew “And”}Zophar the Naamathite answered and said,}%
\verse{“Therefore my disquieting thoughts bring me back 
\textit{for the sake of}\lebnote{Literally “and in because of”}my inward excitement.\lebnote{Or “my inward haste”}}%
\verse{I hear discipline that insults me, 
and a spirit beyond my understanding answers me.}%
\verse{“Did you\lebnote{Singular}know this from of old, 
since the setting of the human being on earth,}%
\verse{that the rejoicing of the wicked is \textit{short},\lebnote{Literally “from near”}
and the joy of the godless \textit{lasts only a moment}?\lebnote{Literally “is until a moment”}}%
\verse{Even though his stature mounts up to the heaven, 
and his head reaches to the clouds,}%
\verse{he will perish forever like his dung; 
those who have seen him will say, ‘Where is he?’}%
\verse{He will fly away like a dream, and they will not find him, 
and he will be chased away like a vision of the night.}%
\verse{The eye that saw him \textit{will not see him again},\lebnote{Literally “and it will not do again”}
and his place will no longer behold him.}%
\verse{His children will seek favors from the poor, 
and his hands will return his wealth.}%
\verse{His bones were full of his vigor, 
but\lebnote{Hebrew “and”}it will lie down with him on\lebnote{Or “in”}the dust.}%
\verse{“Though wickedness tastes sweet in his mouth, 
and he hides it under his tongue,}%
\verse{though he spares it and does not let it go 
and holds it back in the midst of his palate,}%
\verse{in his bowels his food is turned, 
the venom of horned vipers is \textit{within him}.\lebnote{Literally “in his entrails”}}%
\verse{He swallows riches, but\lebnote{Hebrew “and”}he vomits them\lebnote{Hebrew “it”}up; 
God drives them\lebnote{Hebrew “it”}out from his stomach.}%
\verse{He will suck the poison of horned vipers; 
the viper’s tongue will kill him.}%
\verse{\textit{He will not enjoy the streams},\lebnote{Literally “He will not look on streams”}
the torrents of honey and curds.}%
\verse{Returning the products of his toil, he will not swallow;\lebnote{Hebrew “and he will not swallow”}
according to the profit of his trade, \textit{he will not enjoy},\lebnote{Literally “and he will not taste”}}%
\verse{for he has oppressed; he has abandoned the poor; 
he has seized a house but\lebnote{Or “and”}did not build it.}%
\verse{Because he has not known satisfaction in his stomach, 
\textit{he lets nothing escape that he desires}.\lebnote{Literally “in his desire he does not let be saved”}}%
\verse{\textit{There is nothing left after he has eaten};\lebnote{Literally “There is not a remnant at his eating”}
therefore his prosperity will not endure.}%
\verse{In the fullness of his excess \textit{he will be in distress};\lebnote{Literally “it will be in distress for him”}
all of misery’s power will come upon him.}%
\verse{\textit{When his stomach fills up},\lebnote{Literally “It will be to fill up his stomach”}God will send \textit{his burning anger}\lebnote{Literally “the blaze of his nose/anger”}upon him, 
and he will let it rain down upon him\lebnote{Or “them”}as his food.}%
\verse{“He will flee from an iron weapon, 
but an arrow of bronze will pierce him.}%
\verse{He draws it forth, and it comes out from his body, 
and the glittering point comes from his gall-bladder; 
terrors come upon him.}%
\verse{\textit{Total darkness}\lebnote{Literally “All darkness”}is hidden for his treasures;\lebnote{Or “His treasures are hidden by total darkness”}
\textit{an unfanned fire}\lebnote{Literally “a fire not fanned”}will devour him; 
the remnant\lebnote{Of his treasures}\textit{will be consumed}\lebnote{Literally “will be evil,” or “will be bad”}in his tent.}%
\verse{The heavens will reveal his guilt, 
and the earth will rise up against him.}%
\verse{The products of his house will be carried away\lebnote{Or “the produce of his house will go away into exile”}
like gushing waters\lebnote{Or “by gushing waters”}on the day of his wrath.}%
\verse{This is a wicked human being’s portion from God 
and the inheritance of his decree from God.”\lebnote{Or “and the inheritance decreed for him by God” (compare ESV, NJPS)}}%
\end{biblechapter}

\begin{biblechapter} % Job 21
\verseWithHeading{Job’s Seventh Speech: A Response to Zophar}{Then\lebnote{Hebrew “And”}Job answered and said,}%
\verse{“Listen carefully to my words, 
and let this be your\lebnote{Plural}consolation.}%
\verse{Bear with me, and I myself\lebnote{Emphatic personal pronoun}will speak; 
then\lebnote{Hebrew “and”}after my speaking you can mock.}%
\verse{As for me, is my complaint for human beings? 
And if so, why cannot I be impatient?}%
\verse{Turn to me and be appalled, 
and place your hand on your mouth.}%
\verse{And when I think of it, I am horrified,\lebnote{Hebrew “and I am horrified”}
and shuddering seizes my flesh.}%
\verse{“Why do the wicked live, 
grow old, even grow mighty in power?}%
\verse{With them their offspring are established \textit{before them},\lebnote{Literally “to their faces”}
and their descendants before their eyes.}%
\verse{Their houses are safe without fear, 
and the rod of God is not upon them.}%
\verse{His bull breeds and does not fail; 
his cow calves and does not miscarry.}%
\verse{They send out their little ones like the flock, 
and their children dance around.}%
\verse{They \textit{sing}\lebnote{Literally “they lift up their voices”}to the tambourine and lyre, 
and they rejoice to the sound of the long flute.}%
\verse{They spend their days in prosperity, 
and in peace they go down to Sheol.}%
\verse{And they say to God, ‘Turn away from us, 
for\lebnote{Hebrew “and”}we do not desire to know your ways.}%
\verse{Who is Shaddai that we should serve him, 
or\lebnote{Hebrew “and”}what would we benefit when we plead with him?’}%
\verse{Look, their prosperity is not in their hands; 
the schemes of the wicked are repugnant\lebnote{Or “the plan of the wicked is repugnant”}to me.}%
\verse{“How often is the lamp of the wicked put out, 
and their disaster comes upon them? 
He\lebnote{The antecedent is “God,” or possibly “the Almighty”}distributes pains in his anger.}%
\verse{How often are they like straw \textit{before}\lebnote{Literally “to the faces of”}the wind, 
and like chaff that the storm carries away?}%
\verse{‘God stores up his iniquity for his children’?\lebnote{Or “You say, ‘God stores up his iniquity for his children.’ ”}
Then let him repay it to him that\lebnote{Hebrew “and”}he may know.}%
\verse{Let his eyes see his decay, 
and let him drink from the wrath of Shaddai,}%
\verse{for what \textit{does he care for}\lebnote{Literally “his joy in”}his house after him, 
when the number of his months is\lebnote{Hebrew “are”}cut off?}%
\verse{Can anyone teach knowledge to God, 
since\lebnote{Hebrew “and”}he himself\lebnote{Emphatic personal pronoun}judges high ones?}%
\verse{This one dies \textit{in full prosperity},\lebnote{Literally “in the vigor of his prosperity”}
\textit{completely}\lebnote{Literally “all of him”}at ease and secure.}%
\verse{His vats are full with milk 
and the marrow of his bones is moist.}%
\verse{Yet\lebnote{Hebrew “And”}another dies with a bitter inner self\lebnote{Or “soul”}
and has not tasted prosperity.}%
\verse{They lie down together in the dust, 
and maggots cover them.}%
\verse{“Look, I know your\lebnote{Plural}thoughts 
and the schemes you devise against me.}%
\verse{For you say, ‘Where is the nobleman’s house, 
and where are \textit{the tents in which the wicked dwell}?’\lebnote{Literally “tent of dwelling places of wicked”}}%
\verse{Do you not ask those traveling the roads, 
and do you not take notice of their accounts?}%
\verse{Indeed, the wicked is spared from the day of disaster; 
he is delivered from the day of wrath.}%
\verse{Who denounces his way to his face? 
And who repays him for what he has done?}%
\verse{When\lebnote{Hebrew “And”}he is brought to the grave, 
then\lebnote{Hebrew “and”}someone stands guard over the tomb.}%
\verse{The clods of the valley are sweet to him; 
everyone will follow after him, 
and \textit{before}\lebnote{Literally “to the faces of”}him \textit{they are innumerable}.\lebnote{Literally “there is not number”}}%
\verse{So\lebnote{Hebrew “And”}how will you comfort me with emptiness, 
when\lebnote{Hebrew “and”}fraud is left in your answers?”}%
\end{biblechapter}

\begin{biblechapter} % Job 22
\verseWithHeading{Eliphaz’s Third Speech}{Then\lebnote{Hebrew “And”}Eliphaz the Temanite answered and said,}%
\verse{“Can a man be of use to God? 
Yes, can the wise be of use to him?}%
\verse{Is it a pleasure to Shaddai if you are righteous, 
or\lebnote{Hebrew “and”}a gain if you make your ways blameless?}%
\verse{Does he reprove you because of your reverence 
or enter into judgment with you?}%
\verse{Is not your wickedness great, 
and there is no end to your iniquities?}%
\verse{“Indeed, you have required a pledge from your family for nothing, 
and you have stripped off the clothes of the naked.}%
\verse{You have given no water for the weary to drink, 
and you withheld food from the hungry.}%
\verse{And the land \textit{belongs to the man of power},\lebnote{Literally “man of power for him”}
and \textit{the favored}\lebnote{Literally “lifted up of face”}lives in it.}%
\verse{You have sent widows away empty-handed, 
and the arms of orphans were crushed.}%
\verse{Therefore trapping nets are all around you, 
and panic suddenly terrifies you,}%
\verse{or it is so dark you cannot see, 
and a flood of water covers you.}%
\verse{“Is not God in the height of the heavens? 
But\lebnote{Hebrew “And”}see how lofty are \textit{the highest stars}.\lebnote{Literally “top of stars,” or “head of stars”}}%
\verse{And you ask, ‘What does God know? 
Can he judge through deep gloom?}%
\verse{Thick clouds are a covering for him, so that he does not see; 
and he walks about on the dome of heaven.’}%
\verse{Will you keep to the way of old 
that the people of mischief have trod,}%
\verse{who were snatched away \textit{before their time},\lebnote{Literally “and not time”}
whose foundation was washed away by a current?}%
\verse{Those who said to God, ‘Turn away from us,’ 
and ‘What can Shaddai do to us?’}%
\verse{Yet\lebnote{Hebrew “And”}he himself\lebnote{Emphatic personal pronoun}filled up their houses with good things, 
but\lebnote{Hebrew “and”}the schemes of the wicked are repugnant to me.}%
\verse{The righteous see, and they rejoice, 
but\lebnote{Hebrew “and”}the innocent one mocks at them.}%
\verse{\textit{Surely}\lebnote{Literally “If not”}our foe perishes, 
and fire has consumed their remainder.\lebnote{Or “their wealth”}}%
\verse{“Please reconcile yourself with him, 
and be at peace; \textit{in this way},\lebnote{Literally “by them”}good will come to you.}%
\verse{Please receive instruction from his mouth, 
and place his words in your heart.}%
\verse{If you return to Shaddai, you will be restored; 
if you remove wickedness from your tent,}%
\verse{and if you put gold ore in the dust, 
and the gold of Ophir in the rock of wadis,}%
\verse{then\lebnote{Hebrew “And”}Shaddai will be your gold ore 
and \textit{your precious silver}.\lebnote{Literally “silver of the best for you,” or “silver of the best belonging to you”}}%
\verse{Indeed, then you will delight yourself in Shaddai, 
and \textit{you will expect that God will be good to you}.\lebnote{Literally “you will lift up to God your face”}}%
\verse{You will pray to him, and he will hear you, 
and you will pay your vows.}%
\verse{And you will decide on a matter, and it will be established for you, 
and light will shine on your ways.}%
\verse{When they are humiliated, \textit{you say},\lebnote{Literally “and you say,” or “then you say”}‘It is pride,’ 
for\lebnote{Hebrew “and”}he saves \textit{the humble}.\lebnote{Literally “downward of eyes”}}%
\verse{He will deliver the guilty, 
and he will escape because of the cleanness of your hands.”}%
\end{biblechapter}

\begin{biblechapter} % Job 23
\verseWithHeading{Job’s Eighth Speech: A Response to Eliphaz}{Then\lebnote{Hebrew “And”}Job answered and said,}%
\verse{“Even today my complaint is bitter; 
my hand is heavy in addition to my groaning.}%
\verse{\textit{O that}\lebnote{Literally “Who shall give”}I knew and that I might find him; 
O that I might come to his dwelling.}%
\verse{I would lay my case \textit{before him},\lebnote{Literally “to his face”}
and I would fill my mouth with arguments.}%
\verse{I want to know the words with which he would answer me, 
and I want to understand what he would say to me.}%
\verse{Would he contend with me in the greatness of his strength? 
No, but he himself\lebnote{Emphatic personal pronoun}would give heed to me.}%
\verse{There an upright person could reason with him, 
and I would be acquitted forever by my judge.}%
\verse{“When I go forward, \textit{he is not there},\lebnote{Literally “and there is not he,” or “and he is not”}
or\lebnote{Hebrew “and”}backward, I cannot see him.\lebnote{Hebrew “and I cannot see him”}}%
\verse{\textit{When he works}\lebnote{Literally “At his working”}on the left, \textit{I cannot perceive}\lebnote{Literally “and I cannot see”}him; 
he turns to the right, but\lebnote{Hebrew “and”}I cannot see him.}%
\verse{But he knows \textit{the way that I take};\lebnote{Literally “way with me”}
he tests me—I shall come out like gold.}%
\verse{My foot has held on to his steps; 
I have kept his way, and I have not turned aside.}%
\verse{From the commandment of his lips, indeed\lebnote{Hebrew “and”}I have not departed; 
I have treasured the words of his mouth more than my daily food.}%
\verse{“But\lebnote{Hebrew “And”}he is \textit{alone},\lebnote{Literally “as one”}and who can dissuade him? 
And \textit{whatever he desires},\lebnote{Literally “his inner self desires,” or “his soul desires”}\textit{indeed},\lebnote{Hebrew “and”}he does it.}%
\verse{For he will carry out \textit{what he appoints for me},\lebnote{Literally “my portion,” or “my allotment”}
and many things like these are with him.}%
\verse{Therefore, I am horrified because of his presence; 
when I consider, I tremble\lebnote{Hebrew “and I tremble,” or “then I tremble”}because of him.}%
\verse{Indeed,\lebnote{Hebrew “And”}God has made my heart faint, 
and Shaddai has terrified me.}%
\verse{Indeed, I have not been silenced \textit{because of darkness},\lebnote{Literally “from the face of darkness”}
and because of me he\lebnote{Or “it”}conceals thick darkness.}%
\end{biblechapter}

\begin{biblechapter} % Job 24
\verseWithHeading{Job’s Eighth Speech, Continued}{“Why are not times kept by Shaddai, 
and why do not those who know him see his days?}%
\verse{They\lebnote{Or “Some” (NASU, ESV) or “Men” (NIV, NET)}remove border stones; 
they seize flocks, and they pasture them.}%
\verse{They drive away the donkey of orphans; 
they take the widow’s ox as a pledge.}%
\verse{They thrust the poor off the road; 
the needy of the earth hide themselves together.\lebnote{Or “altogether”}}%
\verse{“Look, like wild donkeys in the desert 
they\lebnote{That is, the poor}go out to their labor as searchers for the prey; 
the wilderness is\lebnote{Or “provides”}\textit{their}\lebnote{Hebrew “for their”}food for the young.}%
\verse{They reap their\lebnote{Hebrew “his”}fodder in the field, 
and they glean in the vineyard of the wicked.}%
\verse{They spend the night naked, \textit{without}\lebnote{Literally “from not”}clothing, 
and they have no garment in the cold.}%
\verse{They are wet from the rainstorm of the mountains,\lebnote{Or “from the mountain rainstorm”}
and they cling to the rock \textit{without}\lebnote{Literally “from not”}refuge.}%
\verse{“They\lebnote{That is, those who oppress the poor}snatch the orphan from the breast, 
and they take a pledge against the needy.\lebnote{Or “And they take the child of the needy as a pledge”}}%
\verse{They\lebnote{That is, the poor}go about naked, without clothing, 
and hungry, they carry the sheaves.\lebnote{Or “and they carry the sheaves though hungry”}}%
\verse{Between their terraces they press out oil; 
they tread the presses, but\lebnote{Hebrew “and”}they are thirsty.}%
\verse{From the city people groan, 
and the throat\lebnote{Or “soul,” or “inner self”}of the wounded cries for help; 
yet\lebnote{Hebrew “and”}God does not \textit{regard it as unseemly}.\lebnote{Literally “he puts offensiveness”}}%
\verse{“Those are among the ones rebelling against the light; 
they do not recognize his ways, 
and they do not stay in his paths.}%
\verse{\textit{At dusk}\lebnote{Literally “At the light”}the murderer rises; 
he kills the needy and the poor, 
and in the night he is like a\lebnote{Or “the”}thief.}%
\verse{And the eye of the adulterer waits for dusk, 
saying, \textit{‘No eye will see me,’}\lebnote{Literally “ ‘An eye will not see me’ ”}
and he places a covering on his face.}%
\verse{He digs through houses in the darkness; 
by day they shut themselves in— 
they do not know the light}%
\verse{because morning likewise is to them deep darkness; 
indeed, they know about the terrors of deep darkness.}%
\verse{“He himself\lebnote{Emphatic personal pronoun}is swift on the water’s surface; 
their portion is cursed in the land. 
\textit{No one turns toward the path of their vineyards}.\lebnote{Literally “He does not turn toward the path of the vineyard”}}%
\verse{Drought and heat snatch away the snow waters, 
like Sheol snatches away those\lebnote{Or “they/them”}who have sinned.}%
\verse{The womb forgets him. 
The maggot feasts on him until he is no longer remembered, 
and wickedness is broken like a\lebnote{Or “the”}tree.}%
\verse{He feeds on the barren, who does not have a child, 
and does no good to the widow.}%
\verse{Yet\lebnote{Hebrew “And”}he\lebnote{The probable antecedent is God}carries off the tyrants by his strength; 
if he rises up, then\lebnote{Hebrew “and”}he\lebnote{The antecedent of the singular here probably changes to the wicked person}cannot be certain of life.\lebnote{Hebrew “the life”}}%
\verse{He\lebnote{The probable antecedent is God}gives security to him, and he is\lebnote{Or “they are”}supported, 
but\lebnote{Hebrew “and”}his eyes are upon their ways.}%
\verse{They are exalted a little while, then\lebnote{Literally “and”}\textit{they are gone}.\lebnote{Literally “he is not”}
And they are brought low; they are cut off like all others, 
and like the heads of grain they wither away.}%
\verse{“And if it is not so, then who can prove me a liar 
and reduce my word to nothing?”}%
\end{biblechapter}

\begin{biblechapter} % Job 25
\verseWithHeading{Bildad’s Third Speech}{Then\lebnote{Hebrew “And”}Bildad the Shuhite answered and \textit{said},\lebnote{Hebrew “he said”}}%
\verse{“Domination and dread are with him 
who makes peace in his high heavens.}%
\verse{Is there a number to his troops? 
And upon whom does his light not rise?}%
\verse{“Indeed,\lebnote{Hebrew “And”}how can a human being be righteous before God? 
And how will he who is born of a woman be pure?}%
\verse{Look, even the moon \textit{is not bright},\lebnote{Literally “and it is not bright”}
and the stars are not pure in his sight.}%
\verse{\textit{How much less}\lebnote{Literally “Also”}for a human being who is a maggot, 
and \textit{a human}\lebnote{Literally “a son of man,” or “a son of Adam”}who is a worm?”}%
\end{biblechapter}

\begin{biblechapter} % Job 26
\verseWithHeading{Job’s Ninth Speech: A Response to Bildad}{Then\lebnote{Hebrew “And”}Job answered and said,}%
\verse{“How you\lebnote{Singular}have helped \textit{one who has no power}!\lebnote{Literally “for not power”}
How you\lebnote{Singular}have assisted the arm \textit{that has no strength}!\lebnote{Literally “not strength”}}%
\verse{How you\lebnote{Singular}have advised \textit{one who has no wisdom}!\lebnote{Literally “for not wisdom”}
And what sound wisdom you\lebnote{Singular}have made known \textit{in abundance}!\lebnote{Literally “as far as the multitude,” or “for multitude”}}%
\verse{\textit{With whose help}\lebnote{Literally “With whom”}have you\lebnote{Singular}uttered words, 
and whose breath has come forth from you?\lebnote{Singular}}%
\verse{“The spirits of the dead tremble 
below the waters and their inhabitants.}%
\verse{Sheol is naked before him, 
and there is no covering for Abaddon.}%
\verse{He stretches out the north over emptiness; 
he hangs the earth \textit{over nothing}.\lebnote{Literally “over not what”}}%
\verse{He ties up the water in its clouds, 
and the cloud is not torn open beneath it.\lebnote{Hebrew “them”; antecedent for “them” is “water” (literally “waters”)}}%
\verse{He covers the face of the full moon;\lebnote{Text reads “throne”; “full moon” is based on a change of the vowels}
he spreads his cloud over it.}%
\verse{\textit{He has described a circle}\lebnote{NRSV; literally “a limit he has circled”}on the face of the water 
\textit{between light and darkness}.\lebnote{Literally “at the end of light with darkness”}}%
\verse{“The pillars of heaven tremble, 
and they are astounded at his rebuke.}%
\verse{By his power he stilled the sea, 
and by his understanding he struck down Rahab.}%
\verse{By his breath \textit{the heavens were made clear};\lebnote{Literally “heavens clearness”}
his hand pierced the fleeing snake.}%
\verse{Look, these are the outer fringes of his ways, 
and \textit{how faint is the word}\lebnote{Literally “how whisper of word”}that we hear of him! 
But\lebnote{Hebrew “And”}who can understand the thunder of his power?”}%
\end{biblechapter}

\begin{biblechapter} % Job 27
\verseWithHeading{Job Continues His Final Speech}{Then\lebnote{Hebrew “And”}Job again took up his discourse and said,}%
\verse{“\textit{As God lives},\lebnote{Literally “The life of God”}he has removed my justice, 
and Shaddai has made my inner self\lebnote{Or “soul”}bitter.}%
\verse{For \textit{as long as}\lebnote{Literally “all of still”}my breath is in me, 
and the spirit of God is in my nose,}%
\verse{my lips surely will not speak falseness, 
and my tongue surely will not utter deceit.}%
\verse{Far be it from me that \textit{I would say that you\lebnote{Plural}are right};\lebnote{Literally “I would declare you as in the right”}
until I pass away, I will not put away from me my blamelessness.}%
\verse{I hold fast to my righteousness, and I will not let it go; 
my heart will not blame any of my days.}%
\verse{“Let my enemy be like the wicked 
and my opponent like the unrighteous,}%
\verse{for what is the hope of the godless when he cuts them off, 
when God takes away his life?}%
\verse{Will God hear his cry of distress 
when distress comes upon him?}%
\verse{Or, in Shaddai will he find delight? 
Will he call upon God at all times?}%
\verse{“I will teach you\lebnote{Plural}about\lebnote{Or “concerning”}God’s hand; 
I will not conceal that which is with Shaddai.}%
\verse{Look, you all have seen, 
and \textit{why in the world}\lebnote{Literally “why this”}\textit{have you become altogether vain}?\lebnote{Literally “vain you have become vain”}}%
\verse{“This is the portion of the wicked human being with God, 
and they receive from Shaddai the inheritance of the ruthless.}%
\verse{If their children multiply, it is for the sword, 
and his offspring \textit{do not have enough to eat}.\lebnote{Literally “they are not satisfied with bread”}}%
\verse{Their\lebnote{Hebrew “His”}survivors are buried through\lebnote{Or “because of”}the plague, 
and their\lebnote{Hebrew “his”}widows do not weep.}%
\verse{If he heaps up silver like the dust 
and fashions clothing like the clay,}%
\verse{he makes it ready, and the righteous will wear it, 
and the innocent will divide the silver.}%
\verse{“He builds his house like the moth, 
and like a booth that a watchman has made.}%
\verse{He goes to bed with wealth, but\lebnote{Hebrew “and”}\textit{he will do so no more};\lebnote{Literally “he will not do again”}
he opens his eyes, and \textit{it is gone}.\lebnote{Literally “it is not”}}%
\verse{Terrors overtake him like the water; 
a storm wind carries him off in the night.}%
\verse{The east wind lifts him up, and he is gone, 
and it sweeps him away from his place.}%
\verse{And it hurls at him, and it has no compassion; 
he will quickly flee from its power.}%
\verse{It claps its hands over him, 
and it hisses at him from its place.}%
\end{biblechapter}

\begin{biblechapter} % Job 28
\verseWithHeading{Job’s Discourse on Wisdom}{“Indeed, there is a mine for silver 
and a place \textit{for gold to be refined}.\lebnote{Literally “for the gold they refine”}}%
\verse{Iron is taken from dust, 
and he pours out copper from ore.}%
\verse{He puts an end to darkness, 
and he searches out the farthest limits 
for the ore in gloom and deep shadow.}%
\verse{He breaks open a mine shaft \textit{away from where people dwell};\lebnote{Literally “from with dwelling”}
those who are forgotten \textit{by travelers},\lebnote{Literally “by feet”}
they dangle, they sway far away from human beings.}%
\verse{As for the earth, from it comes food,\lebnote{Or “bread”}
but\lebnote{Hebrew “and”}underneath it, it is turned up as by fire.}%
\verse{Its stones are the place of sapphire, 
and the earthen dirt \textit{has}\lebnote{Literally “for it”}gold.}%
\verse{“It is a path a bird of prey does not know 
and the black kite’s eye has not seen.}%
\verse{\textit{Proud wild animals}\lebnote{Literally “The sons of pride”}have not trodden it; 
the lion in its prime has not prowled over it.}%
\verse{He puts his hand on the hard rock; 
he overturns mountains by the roots.}%
\verse{He cuts out tunnels in the rocks, 
and his eye sees every treasure.}%
\verse{He dams up rivers from their sources, 
and he brings secret things to the light.}%
\verse{“But\lebnote{Hebrew “And”}from where will wisdom be found? 
And \textit{where in the world}\lebnote{Literally “where this”}is the place of understanding?}%
\verse{A human being does not know its proper value, 
and it is not found in the land of the living.}%
\verse{The deep says, ‘It is not in me,’ 
and the sea says, ‘It is not with me.’}%
\verse{“Refined gold cannot be gotten in its place, 
and silver cannot be weighed out as its price.}%
\verse{It cannot be bought for the gold of Ophir, 
for precious onyx or\lebnote{Hebrew “and”}sapphire.}%
\verse{Gold and glass cannot be compared with it, 
and its substitution cannot be an ornament of refined gold.}%
\verse{Black corals and crystal will not be mentioned, 
and wisdom’s price is more than red corals.}%
\verse{The topaz of Cush cannot be compared with it; 
it cannot be bought for pure gold.}%
\verse{Indeed,\lebnote{Hebrew “And”}from where does wisdom come? 
And \textit{where in the world}\lebnote{Literally “where this”}is the place of understanding?}%
\verse{It is hidden from the eyes of all living, 
and it is concealed from the birds of the heaven.}%
\verse{Abaddon and Death say, 
‘We heard its rumor with our ears.’}%
\verse{“God understands its way, 
and he knows its place,}%
\verse{for he himself\lebnote{Emphatic personal pronoun}looks to the end of the earth; 
he sees under all the heaven.}%
\verse{\textit{When he gave}\lebnote{Literally “To give”}weight to the wind 
and he apportioned the waters by measure,}%
\verse{\textit{when he made}\lebnote{Literally “at his making”}a rule for the rain 
and a way for the thunder’s lightning bolt,}%
\verse{then he saw it and talked about it; 
he established it, and moreover, he explored it.}%
\verse{And to the human beings he said, 
‘Look, the fear of the Lord is wisdom, 
and to depart from evil is understanding.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Job 29
\verseWithHeading{Job’s Final Defense}{Then\lebnote{Hebrew “And”}Job again took up his discourse and said,}%
\verse{“\textit{O that I were}\lebnote{Literally “Who shall give me”}as in the months before, 
as in the days when God watched over me,}%
\verse{when his shining lamp was over my head— 
by his light I walked through darkness—}%
\verse{as when I was in the days of my prime, 
when God’s confiding was over my \textit{house},\lebnote{Literally “tent”}}%
\verse{when Shaddaiwas still with me, 
my children were all around me,}%
\verse{when my paths were washed in sour milk, 
and the rock poured out streams of oil for me.}%
\verse{“At my going out the gate to the city, 
I secured my seat in the square.}%
\verse{Young men saw me and stepped aside, 
and the aged rose up and stood.}%
\verse{Officials refrained from talking, 
and they laid their hand on their mouth.}%
\verse{The voices\lebnote{Hebrew “voice”; collective singular by context, governing a plural verb}of nobles were hushed, 
and their tongue stuck to their palate.}%
\verse{“When the ear heard and commended me, 
and the eye saw and testified in support of me}%
\verse{because I saved the needy who cried for help, 
and I saved\lebnote{Supplying the elided verb from the preceding parallel line}the orphan for whom there was no helper.}%
\verse{The blessing of the wretched came upon me, 
and I caused the widow’s heart to sing for joy.}%
\verse{I put on righteousness, and it clothed me; 
my justice was like a robe and a headband.}%
\verse{“I was eyes to the blind, 
and I was feet to the lame.}%
\verse{I was a father to the poor, 
and I investigated \textit{the stranger’s}\lebnote{Literally “the one I knew not”}legal dispute.}%
\verse{And I broke the evil one’s jaw bones, 
and I made his prey drop from his teeth.}%
\verse{“And I thought, ‘I will pass away in my nest, 
and like the phoenix I shall multiply my days.}%
\verse{My roots were open to water, 
and dew spent the night on my branches;}%
\verse{My glory was new with me, 
and \textit{I was revitalized regularly}.’\lebnote{See NJPS; literally “my bow in my hand was new”}}%
\verse{“They listened to me and waited, 
and they kept silent for my counsel.}%
\verse{After my word, they did not speak again, 
and my word dropped down like dew upon them.}%
\verse{And they waited for me as for the rain, 
and they opened their mouth wide as for the spring rain.}%
\verse{I smiled for them when they had no confidence in anything, 
and they did not extinguish the light of my face.}%
\verse{I chose their way, and I sat as head, 
and I dwelled like a king among the troops, 
like one who comforts mourners.}%
\end{biblechapter}

\begin{biblechapter} % Job 30
\verseWithHeading{Job’s Final Defense Continued}{“But now those younger than I, as far as days, laugh at me, 
whose fathers I rejected for setting with the dogs of my sheep and goats.}%
\verse{Moreover, \textit{what use to me is the strength of their hands}?\lebnote{Literally “the strength of their hands what is for me”}
With them, vigor is destroyed.}%
\verse{Through want and through barren hunger 
they are gnawing in the dry region in the darkness of desolation and waste.}%
\verse{They are picking salt herbs, the leaves of bushes, 
and the roots of broom trees to warm themselves.}%
\verse{They were driven out from fellow people; 
they shout at them as at a\lebnote{Hebrew “the”}thief,}%
\verse{\textit{so that they dwell}\lebnote{Literally “to dwell”}
in holes of the ground and in the rocks.}%
\verse{They bray among the bushes; 
they are gathered under the nettles.}%
\verse{\textit{A senseless crowd},\lebnote{Literally “sons of good-for-nothing”}yes, \textit{a disreputable brood},\lebnote{Literally “sons of no name”}
they were cast out from the land.}%
\verse{“But now I am their mocking song, 
and I have become a byword for them.}%
\verse{They abhor me; they keep aloof from me, 
and they do not withhold spit from my face}%
\verse{because he has loosened his bowstring and humbled me, 
and they have thrown off restraint \textit{in my presence}.\lebnote{Literally “from my face”}}%
\verse{On the right hand the brood rises\lebnote{Hebrew “rise”}up; 
\textit{they put me to flight},\lebnote{Literally “my feet they send away”}
and they build up their \textit{siege ramps}\lebnote{Literally “the ways/paths of their disaster”}against me.}%
\verse{They destroy my path; 
they promote my destruction; 
they have no helper.}%
\verse{As through a wide breach they come; 
amid a crash they rush on.}%
\verse{Terrors are turned upon me; 
my honor is pursued as by the wind, 
and my hope of deliverance passed by like a cloud.}%
\verse{“And now my life is poured out onto me; 
days of misery have taken hold of me.}%
\verse{\textit{At night I am in great pain};\lebnote{Literally “night my bones he pierces upon me,” or “night pierces my bones upon me”}
my pains do not take a rest.}%
\verse{He seizes my clothing with \textit{great power};\lebnote{Literally “much of strength”}
he grasps me by my tunic’s collar.}%
\verse{He has cast me into the dirt, 
and I have become like dust\lebnote{Hebrew “the dust”}and ashes.}%
\verse{I cry to you for help, but\lebnote{Hebrew “and”}you do not answer me; 
I stand, and you merely look at me.}%
\verse{You have turned cruel to me; 
you persecute me with your hand’s might.}%
\verse{You lift me up to the wind—you make me ride it, 
and you toss me about in the storm.}%
\verse{Indeed, I know that you will bring me to death 
and to the house of assembly for all the living.}%
\verse{“Surely someone must not send a hand against the needy 
when, in his misfortune, there is a cry of help for them.}%
\verse{Have I not wept for \textit{the unfortunate},\lebnote{Literally “hard of day”}
and grieved myself over the poor?}%
\verse{Indeed, I hoped for good, but evil came, 
and I waited for light, but darkness came.}%
\verse{My bowels are in turmoil, and they are not still; 
days of misery come to confront me.}%
\verse{I go about in mourning garb, but not in sunlight; 
I stand up in the assembly, and I cry for help.}%
\verse{I am a companion for the jackals 
and a companion for \textit{ostriches}.\lebnote{Literally “daughters of ostrich”}}%
\verse{My skin turns black on me, 
and my bones burn with heat.}%
\verse{So\lebnote{Hebrew “And”}my lyre came to be used for mourning, 
and my flute for the voice of those who weep.}%
\end{biblechapter}

\begin{biblechapter} % Job 31
\verseWithHeading{Job’s Final Defense Continued}{“I made a covenant with my eyes, 
so\lebnote{Hebrew “and”}how could I look closely upon a virgin?}%
\verse{And what is the portion of God from above 
or\lebnote{Hebrew “and”}the heritage of Shaddai from on high?}%
\verse{Is not disaster for the evil one 
and ruin for the workers of mischief?}%
\verse{Does he not see my ways 
and count all my steps?}%
\verse{“If I have walked with falseness, 
and my foot has hastened to deceit,}%
\verse{let him weigh me in the balance of justice, 
and let God know my blamelessness.}%
\verse{If my steps have turned aside from the way, 
and my heart has walked after my eyes, 
and my hand has clung to a spot,}%
\verse{let me sow, and let another eat, 
and let my crops be rooted out.}%
\verse{“If my heart has been enticed by a woman, 
and at my neighbor’s doorway I have lain in wait,}%
\verse{let my wife grind for another, 
and let other men kneel over her,}%
\verse{for that is a shameful act, 
and that is \textit{a criminal offense}.\lebnote{Literally “an iniquity of judges”}}%
\verse{Indeed, that is a fire that will consume up to Abaddon, 
and it would uproot all my crop.}%
\verse{“If I have rejected my male or\lebnote{Hebrew “and”}female slave’s case 
when their complaint was against me,}%
\verse{then\lebnote{Hebrew “and”}what shall I do when God rises up? 
And when he enquires, how shall I answer him?}%
\verse{Did not he who made me in the womb make them? 
And did not one fashion us in the womb?}%
\verse{“If I have withheld the desire of the poor from them, 
or\lebnote{Hebrew “and”}I have caused the widow’s eyes to fail,}%
\verse{or I have eaten my morsel \textit{alone},\lebnote{Literally “to alone me”}
and the orphan has not eaten from it}%
\verse{(for from my childhood he\lebnote{Probably referring to “the orphan”}\textit{grew up with}\lebnote{Literally “reared”}me like a father, 
and from my mother’s womb I guided her\lebnote{Probably referring to “the widow”}),}%
\verse{if I have seen the one who perishes because of no clothing 
or\lebnote{Hebrew “and”}that there is no covering for the poor,}%
\verse{if his loins have not blessed me, 
or\lebnote{Hebrew “and”}by means of my sheep’s fleece he has warmed himself,}%
\verse{if I have raised my hand against an orphan 
because I saw my supporters at the gate,}%
\verse{then let my shoulder blade fall from my shoulder, 
and let my arm be broken from its socket.}%
\verse{Indeed, the disaster from God was a dread for me, 
and \textit{I was powerless}\lebnote{Literally “I was not able”}because of his majesty.}%
\verse{“If I have made gold my trust, 
or\lebnote{Hebrew “and”}I have called fine gold my security,}%
\verse{if I have rejoiced because my wealth was great 
or\lebnote{Hebrew “and”}because my hand found a fortune,}%
\verse{if I looked at the sun when it shone 
or\lebnote{Hebrew “and”}at the moon moving in splendor,}%
\verse{and \textit{I was secretly enticed},\lebnote{Literally “it was gullible in the secrecy of my heart”}
and \textit{my hand threw them a kiss},\lebnote{Literally “my hand kissed for my mouth”}}%
\verse{this also is an iniquity to be judged, 
for I have deceived God above.}%
\verse{“If I have rejoiced at the ruin of the one who hated me 
or\lebnote{Hebrew “and”}have exulted when evil overtook him—}%
\verse{no,\lebnote{Hebrew “and”}I have not allowed my mouth to sin, 
to ask his life with a curse.}%
\verse{\textit{Have the people of my tent not said},\lebnote{Literally “If the people of my tent have not said”}
‘\textit{O that}\lebnote{Literally “who shall give”}someone\lebnote{Or “he”}had not been satisfied with his meat’?}%
\verse{An alien has not lodged in the street; 
I have opened my door to the traveler.}%
\verse{\textit{Have I concealed}\lebnote{Literally “If I have concealed”}my transgressions as other human beings 
to hide my iniquity in my bosom}%
\verse{because I dreaded the great multitude, 
and the contempt of clans terrified me, 
so that I kept quiet, I did not go out of the doorway?}%
\verse{\textit{O that}\lebnote{Literally “Who shall give”}\textit{I had}\lebnote{Literally “for me”}someone hearing me! 
Here is my signature; 
let Shaddai answer me! 
\textit{As for}\lebnote{Hebrew “And,” or “But”}the written communication that \textit{my adversary}\lebnote{Literally “the man of my indictment”}has written,}%
\verse{I would \textit{surely}\lebnote{Literally “if not”}carry it on my shoulder; 
I would bind it on me like a crown.}%
\verse{I would give him an account of my steps; 
I would approach him like a noble.}%
\verse{If my land has cried out against me, 
and its furrows have wept together,}%
\verse{if I have eaten its yield without payment, 
or\lebnote{Hebrew “and”}I have caused the breath of its owners to die,}%
\verse{let thorns grow in place of wheat 
and noxious weeds in place of barley.” 
The words of Job are ended.}%
\end{biblechapter}

\begin{biblechapter} % Job 32
\verseWithHeading{Elihu Rebukes Job and His Three Friends}{Then these three men ceased from answering Job because he was righteous in his own eyes.}%
\verse{So Elihu the son of Barakel the Buzite, from the clan of Ram, \textit{became angry}.\lebnote{Literally “the nose of Elihu became hot”}\textit{He became angry}\lebnote{Literally “His nose became hot”}at Job because he justified himself rather than God,}%
\verse{and \textit{he became angry}\lebnote{Literally “his nose became hot”}at his three friends because they had not found an answer, and they had declared Job guilty.}%
\verse{Thus Elihu had waited for Job \textit{to speak}\lebnote{Literally “in words”}because \textit{they were older than he}.\lebnote{Literally “they were older than he as far as days”}}%
\verse{When Elihu saw that there was no answer in the mouth of the three men, \textit{he became angry}.\lebnote{Literally “his nose became hot”}}%
\verse{Then Elihu the son of Barakel the Buzite \textit{spoke up}\lebnote{Literally “answered”; see NET}and said,}%
\verse{“\textit{I am young},\lebnote{Literally “Young I am as far as days”}but\lebnote{Or “and”}you\lebnote{Plural}are old; 
therefore I feared and became afraid of explaining my knowledge to you.\lebnote{Plural}}%
\verse{I thought,\lebnote{Or “said”}‘Let days\lebnote{Compare the idiom of “as far as days” in vv. 4 and 6}speak, 
and let many years teach wisdom.’}%
\verse{“Truly, it is a spirit in a human being, 
and the breath of Shaddai teaches them.}%
\verse{It is not \textit{the aged}\lebnote{Literally “many in days”}who are wise, 
or\lebnote{Hebrew “and”}it is not the elders who understand justice.}%
\verse{Therefore I say, ‘Listen to me; 
let me also declare my knowledge myself.’}%
\verse{“Look, I waited for your\lebnote{Plural}words; 
I listened for your\lebnote{Plural}insight, 
until you searched out \textit{what to say}.\lebnote{Literally “words,” or “sentences”}}%
\verse{And I directed my attention to you;\lebnote{Plural}
and, look, there is no one for Job who refuted, 
answering his words among you.}%
\verse{So do not say, ‘We have found wisdom; 
let God refute him, not a man.’}%
\verse{But\lebnote{Hebrew “And”}he did not direct his words\lebnote{Or “sentences,” or “what he said”}to me, 
and I will not answer him with your\lebnote{Plural}words.}%
\verse{“They are dismayed; they no longer answer; 
\textit{they have nothing to say}.\lebnote{Literally “words fail from them,” or “sentences fail from them,” or “what to say fails from them”}}%
\verse{And I have waited because they do not speak, 
because they stand there and no longer answer.}%
\verse{I myself will answer my share also; 
I myself will declare my knowledge also,}%
\verse{for I am full of words;\lebnote{Or “sentences,” or “what to say”}
\textit{the spirit within me}\lebnote{Literally “the spirit of my stomach”}urges me.}%
\verse{“Look, my internal organs are like unopened wine, 
like new wineskins it\lebnote{Antecedent is “wine”}is ready to burst open.}%
\verse{Let me speak \textit{that I may find relief};\lebnote{Literally “and let it find relief for me”}
let me open my lips, and let me answer.}%
\verse{\textit{Please let me not show partiality to anyone},\lebnote{Literally “Not please I will lift up face of a man”}
and let me not show flattery to human beings,}%
\end{biblechapter}

\begin{biblechapter} % Job 33
\verseWithHeading{Elihu Rebukes Job}{“But now, please hear my speeches,\lebnote{Or “my sentences,” or “what I have to say”}Job, 
and hear\lebnote{Or “give ear to”}all my words.}%
\verse{Please look, I open my mouth; 
my tongue in my mouth speaks.}%
\verse{My words declare my heart’s uprightness, 
and my lips sincerely speak \textit{what my lips know}.\lebnote{Literally “the knowledge of my lips”}}%
\verse{The Spirit of God has made me, 
and the breath of Shaddai gives life to me.}%
\verse{“If you are able, answer me. 
Present your argument \textit{before me};\lebnote{Literally “to the face of me”}take your stand.}%
\verse{Look, before God \textit{I am like you};\lebnote{Literally “I am like your mouth”}
I myself was also formed from clay.}%
\verse{Look, dread of me should not terrify you, 
and my hand will not be heavy upon you.}%
\verse{“Surely you have spoken in my ears, 
and I have heard the sound of your words:\lebnote{Or “sentences,” or “what you say”}}%
\verse{‘I am clean, without transgression; 
I am pure, and there is no guilt in me.}%
\verse{Look, he finds fault against me; 
he reckons me as his enemy;}%
\verse{he puts my feet in the block; 
he watches all my paths.’}%
\verse{“Look, in this you are not right—I will answer you: 
Indeed, God is greater than a human being.}%
\verse{Why do you contend against him, 
that he will not answer all a person’s words?}%
\verse{Indeed, God speaks in one way, 
even\lebnote{Hebrew “and”}in two,\lebnote{Or “time and again” (NJPS, HCSB)}yet someone does not perceive it.}%
\verse{“In a dream, a vision of the night, 
when a deep sleep falls on men slumbering on their bed,}%
\verse{then he opens the ear of men, 
and \textit{he frightens them with a warning}\lebnote{Literally “with their warning he frightens”}}%
\verse{to turn human beings aside from their deeds, 
and he keeps man from pride.}%
\verse{He spares his life from the pit 
and his life from passing over the river of death.}%
\verse{“And he is reproved with pain on his bed, 
even\lebnote{Hebrew “and”}with the strife of his bones continually,}%
\verse{\textit{so that}\lebnote{Hebrew “and”}his life loathes bread, 
and his inner self\lebnote{Or “soul”}loathes \textit{appetizing food}.\lebnote{Literally “food of desire”}}%
\verse{His flesh is wasted away from sight, 
and his bones, which are invisible, are bared.}%
\verse{And \textit{he}\lebnote{Literally “his inner self/soul”}draws near to the pit 
and his life to the killers.}%
\verse{“If there is a messenger beside him, a mediator, one of a thousand,\lebnote{That is, “unique,” or “one of a kind”}
to declare to a human being his uprightness}%
\verse{\textit{so that}\lebnote{Hebrew “and”}he is gracious to him, and he says, 
‘Deliver him from descending into the pit; 
I have found a ransom.’}%
\verse{His flesh is renewed with his youth; 
he returns to the days of his youthful strength.}%
\verse{He prays to God, then he accepts him, 
and he sees his face with a shout of joy, 
and he repays to the human being his righteousness.}%
\verse{“He will sing to men, and he will say, 
‘I have sinned and have perverted what is right, 
and it was not paid back to me.}%
\verse{He redeemed my life from going down into the pit, 
so\lebnote{Hebrew “and”}\textit{I will enjoy the light}.’\lebnote{Literally “my life in the light will see”}}%
\verse{Look, God does all these things 
twice, three times\lebnote{Or “time and again,” or “oftentimes” (NASB, NASU), or “several times”}with a person}%
\verse{to bring his life back from the pit 
\textit{so that he may enjoy the light of life}.\lebnote{Literally “to be lighted with the light of the life”}}%
\verse{“Listen attentively, Job; listen to me; 
be silent, and I will speak.}%
\verse{If \textit{you have anything to say},\lebnote{Literally “If there are words,” or “If there are sentences”}\textit{answer me};\lebnote{Literally “return to me”}
speak, for I desire \textit{to justify you}.\lebnote{Literally “your righteousness”}}%
\verse{If not, you listen to me; 
be silent, and I will teach you wisdom.”}%
\end{biblechapter}

\begin{biblechapter} % Job 34
\verseWithHeading{Elihu Asserts God’s Justice}{Thus\lebnote{Hebrew “And”}Elihu \textit{spoke up}\lebnote{Literally “answered”; see NET}and said,}%
\verse{“Hear my words, you wise men, 
and listen to me, you who know,}%
\verse{for the ear tests words, 
and the palate tastes \textit{food}.\lebnote{Literally “to eat”}}%
\verse{Let us choose justice for ourselves; 
let us determine among ourselves what is good—}%
\verse{for Job has said, ‘I am righteous, 
but God has taken away my justice.}%
\verse{Should I lie concerning my justice? 
Though I am without transgression, my wound caused by an arrow is incurable.’}%
\verse{What man is like Job, 
who drinks scorn like water?}%
\verse{And he goes on the road in company with instigators of mischief, 
and walks with men of wickedness.}%
\verse{Indeed, he says, ‘A man does not profit 
when he takes delight in God.’}%
\verse{“Therefore, listen to me, \textit{people who have sense}:\lebnote{Literally “men of heart”}
far be it from God \textit{that he should commit wickedness}\lebnote{Literally “from wickedness”}
and that Shaddai \textit{should do wrong}.\lebnote{Literally “from wrong”}}%
\verse{Indeed, he will repay according to a human being’s deeds, 
and according to a man’s way he will let it happen to him.}%
\verse{What’s more, God truly does not act wickedly, 
and Shaddai does not pervert justice.}%
\verse{Who gave him charge over the earth? 
Or\lebnote{Hebrew “And”}who laid on him the whole world?}%
\verse{If he should set his heart to it, 
and he should gather his spirit and his breath to himself,}%
\verse{all flesh would perish together, 
and humankind would return to dust.}%
\verse{“But\lebnote{Hebrew “And”}if you have understanding, hear this; 
listen to \textit{what I say}.\lebnote{Literally “the voice/sound of my words”}}%
\verse{Shall he who hates justice really govern? 
Or will you\lebnote{Singular}declare the Righteous One, the Mighty, guilty?—}%
\verse{the one who says to a king, 
‘You worthless man,’ to noblemen, ‘You wicked man,’}%
\verse{who \textit{shows no partiality to officials}\lebnote{Literally “he does not lift the face of officials”}
and who \textit{does not take note of the noble more than of the insignificant},\lebnote{Literally “he does not regard the eminent to the face of the poor/insignificant”}
for all of them are the work of his hands?}%
\verse{In a moment they die, 
and in the middle of the night the people are shaken, and they pass away, 
and they take away the mighty \textit{effortlessly}.\lebnote{Literally “not by a hand”}}%
\verse{“Indeed, his eyes are upon the ways of man, 
and he sees all his steps.}%
\verse{There is no darkness, and there is no deep shadow 
where instigators of mischief might hide themselves.}%
\verse{Indeed, he has not yet appointed a time for man 
to go to God in the judgment.}%
\verse{He shatters the mighty without investigation, 
and he sets others in their place.}%
\verse{Therefore he knows their works, 
and he overturns them in the night 
\textit{so that}\lebnote{Hebrew “and”}they are crushed.}%
\verse{He strikes them \textit{where the wicked stood},\lebnote{Literally “among the wicked”}
\textit{where there are onlookers},\lebnote{Literally “in the place of seers”}}%
\verse{because they have turned aside from him 
and have not understood any of his ways,}%
\verse{so that they cause the cry of distress from the helpless to come to him, 
and he hears the cry of distress from the needy,}%
\verse{but he remains quiet, and who can condemn? 
And he hides his face, and who can behold him? 
Yet\lebnote{Hebrew “And”}he is over a nation and over a person alike,}%
\verse{so that godless human beings should not reign, 
nor those who ensnare the people.}%
\verse{“Indeed, does anyone say to God, ‘I have endured chastisement; 
I will not act corruptly again;}%
\verse{teach me yourself \textit{what I cannot see};\lebnote{Literally “apart from what I can see”}
if I have done wrong, I will not repeat it’?}%
\verse{\textit{According to your judgment},\lebnote{Literally “from with you”}must he repay it because you rejected this? 
Indeed, you must choose, and not I, so\lebnote{Hebrew “And”}declare what you know.}%
\verse{\textit{Those who have sense}\lebnote{Literally “Men of heart”}say to me, 
and the wise man hearing me says:}%
\verse{‘Job speaks without knowledge, 
and his words are without insight.}%
\verse{Would that Job were tested up to the end 
because his answers are like those from men of mischief,}%
\verse{for he adds transgression to his sin; 
he claps hands among us, 
and he multiplies his words against God.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Job 35
\verseWithHeading{Elihu Condemns Job}{Moreover,\lebnote{Hebrew “And”}Elihu \textit{spoke up}\lebnote{Literally “answered”; see NET}and said,}%
\verse{“Do you think this is justice when you say, 
‘\textit{I am right}\lebnote{Literally “My righteousness”}before God’?}%
\verse{If you ask what it profits you: 
‘How do I benefit \textit{by refraining from my sin}?’\lebnote{Literally “from my sin”}}%
\verse{I myself\lebnote{Emphatic personal pronoun}\textit{will give answer to you}\lebnote{Literally “will return you words”}
and to your friends with you.}%
\verse{“Look at the heavens, and see; 
and observe the clouds that are higher than you.}%
\verse{If you have sinned, what do you accomplish against him? 
And if your transgressions are multiplied, what do you do to him?}%
\verse{If you are righteous, what do you give to him, 
or what does he receive from your hand?}%
\verse{Your wickedness affects a person like yourself, 
and your righteousness affects humans.}%
\verse{“Because of the multitude of oppressions, they cry out; 
they cry for help because of the arm of the mighty.}%
\verse{But\lebnote{Hebrew “And”}no one says, ‘Where is God my Maker, 
who gives songs in the night,}%
\verse{who teaches us more than the animals of the earth 
and makes us wiser than the birds of the heavens?’}%
\verse{There they cry out, but\lebnote{Hebrew “and”}he does not answer 
\textit{because of the pride of evildoers}.\lebnote{Literally “from the face of the pride of evildoers”}}%
\verse{“Surely God does not hear an empty plea, 
and Shaddai does not regard it.}%
\verse{\textit{How much less}\lebnote{Literally “Also for”}when you say that you do not see him, 
that the case is \textit{before him},\lebnote{Literally “to the face of him”}and you are waiting for him!}%
\verse{But\lebnote{Hebrew “And”}now, because his anger does not punish, 
and he does not acknowledge the transgression at all,}%
\verse{\textit{therefore}\lebnote{Hebrew “and,” or “so”}Job opens his mouth in empty talk— 
without knowledge he multiplies words.”}%
\end{biblechapter}

\begin{biblechapter} % Job 36
\verseWithHeading{Elihu Extols God’s Greatness}{Then\lebnote{Hebrew “And”}Elihu continued and said,}%
\verse{“Bear with me a little longer, and \textit{I will explain},\lebnote{Literally “I will show you”}
for \textit{I still have something to say on God’s behalf}.\lebnote{Literally “yet on behalf of God words/sentences”}}%
\verse{I will bring my knowledge from far away, 
and I will ascribe righteousness to my maker,}%
\verse{for truly my words\lebnote{Or “what I have to say”}are not falsehood; 
one who has perfect knowledge is with you.\lebnote{In other words, “Somebody who knows what he is talking about is addressing you”}}%
\verse{Look, God is mighty but\lebnote{Hebrew “and”}does not despise anyone; 
he is mighty and \textit{good in understanding}.\lebnote{Literally “strength of heart”}}%
\verse{He does not keep the wicked alive 
but\lebnote{Hebrew “and”}grants justice to the afflicted.}%
\verse{He does not withdraw his eyes from the righteous, 
but\lebnote{Hebrew “and”}he sets them forever with kings on the throne, and they are exalted.}%
\verse{“And if they are tied up with fetters, 
if they are caught in the cords of misery,}%
\verse{then\lebnote{Hebrew “and”}he declares their work to them, 
and their transgressions, that they are behaving arrogantly.}%
\verse{So\lebnote{Hebrew “And”}he opens their ear to the discipline, 
and he commands that they return from mischief.}%
\verse{If they listen and serve him, 
they complete their days in prosperity and their years in pleasantness.}%
\verse{But\lebnote{Hebrew “And”}if they do not listen, they perish by the sword, 
and they pass away without knowledge.}%
\verse{And the godless of heart cherish anger; 
they do not cry for help when he binds them.}%
\verse{\textit{They die in their youth},\lebnote{Literally “It dies in the youth of their self”}
and their life ends among the male shrine prostitutes.}%
\verse{He delivers the afflicted by his misery, 
and he opens their ears by the adversity.}%
\verse{And what’s more, he allured you \textit{out of distress}\lebnote{Literally “from the mouth of distress”}
into a broad place, where in place of it was no constraint; 
and what was set on your table is full of fatness.}%
\verse{But\lebnote{Hebrew “And”}you are obsessed with the judgment of the wicked; 
judgment and justice take hold of you.}%
\verse{Yes,\lebnote{Or “Indeed”}it is wrath, so that it will not incite you into mockery; 
and do not let the ransom’s greatness turn you aside.}%
\verse{Will your cry for help sustain you without distress, 
or\lebnote{Hebrew “and”}all the efforts of your strength?}%
\verse{You must not long for the night, 
to cut off people \textit{in their place}.\lebnote{Literally “in place of them”}}%
\verse{Take care, you must not turn to mischief, 
for because of this you have been tried by misery.}%
\verse{“Look, God is exalted in his power; 
who is a teacher like him?}%
\verse{Who has prescribed his way for him? 
Or\lebnote{Hebrew “And”}who says, ‘You have done wrong’?}%
\verse{Remember that you should extol his work, 
of which people have sung.}%
\verse{All human beings have looked on it; 
everyone watches from afar.}%
\verse{“Look, God is exalted, and we do not know him; 
the number of his years is unsearchable.}%
\verse{Indeed, he draws up the drops of water; 
he distills the rain into its mist,}%
\verse{which the clouds pour down; 
they shower abundantly on human beings.}%
\verse{“Moreover, \textit{can anyone understand}\lebnote{Literally “if he can understand”}the spreading of clouds, 
the thundering of his dwelling place?}%
\verse{Look, he scatters his lightning around him, 
and he covers the roots of the sea.}%
\verse{Indeed, he judges people by them; 
he gives food \textit{in abundance}.\lebnote{Literally “as far as providing in abundance”}}%
\verse{He covers his hands with lightning, 
and he commands it the place at which to strike.}%
\verse{His thundering tells about him; 
the livestock also tell \textit{concerning what rises}.\lebnote{Literally “about going up”}}%
\end{biblechapter}

\begin{biblechapter} % Job 37
\verseWithHeading{Elihu Extols God’s Majesty}{“About this also my heart trembles, 
and it leaps from its place.}%
\verse{Listen carefully to his voice’s thunder 
and the rumbling that goes out from his mouth.}%
\verse{He lets it loose under all the heavens, 
and his lightning to the earth’s corners.}%
\verse{After it, his voice roars; 
it thunders \textit{with his majestic voice},\lebnote{Literally “with the voice of his majesty”}
and he does not restrain it when his voice is heard.}%
\verse{“God thunders with his voice in marvelous ways; 
he does great things, and we cannot comprehend.}%
\verse{For to the snow he says, ‘Fall on the earth’; 
and \textit{the shower of rain, his heavy shower of rain}\lebnote{Literally “the shower of rain and the show of rains of his strength”}—}%
\verse{\textit{he stops all human beings from working}\lebnote{Literally “on the hand of all man he seals”}
\textit{so that everyone whom he has made may know it}.\lebnote{Literally “to know all men of his work”}}%
\verse{Then\lebnote{Hebrew “And”}the animal goes into its den, 
and it remains in its den.}%
\verse{“The storm wind comes from its chamber 
and cold from the north wind.}%
\verse{By God’s breath, ice is given, 
and the broad waters are frozen.}%
\verse{Also, he loads down thick clouds with moisture; 
his lightning scatters the clouds.}%
\verse{And \textit{they}\lebnote{Collective “clouds” in the first line takes the singular “it” here with the plural meaning}turn around by his guidance 
to accomplish all that he has commanded them 
on the face of \textit{the habitable world}.\lebnote{Literally “world of earth”}}%
\verse{Whether as \textit{correction}\lebnote{Literally “a rod”}or for his land, 
or as loyal love, he lets it happen.}%
\verse{“Hear this, Job; 
stand still and consider carefully God’s wondrous works.}%
\verse{Do you know \textit{how God commands them}\lebnote{Literally “at the putting of God upon them”}
and how he causes his cloud’s lightning to shine?}%
\verse{Do you know about the hovering of the clouds, 
\textit{the marvelous works of the one with perfect knowledge}?\lebnote{Literally “the marvelous works of perfect knowledge”}}%
\verse{You whose garments are hot, 
when the earth is being still because of the south wind,}%
\verse{with him can you spread out the skies, 
hard as a molten mirror?}%
\verse{“Teach us what we should say to him; 
we cannot draw up our case because of the presence of darkness.}%
\verse{Should he be told that I want to speak? 
Or did a man say that he would be communicated with?}%
\verse{So then, they do not look at the light when it is bright in the skies, 
when\lebnote{Hebrew “and”}the wind has passed and has cleansed them.}%
\verse{From the north comes gold— 
awesome majesty is around God.}%
\verse{As for Shaddai, we cannot attain him; 
he is exalted in power, 
and he does not oppress justice and abundant righteousness.}%
\verse{Therefore people revered him; 
he does not regard any\lebnote{Or “all”}\textit{who think that they are wise}.”\lebnote{Literally “the wise of heart”}}%
\end{biblechapter}

\begin{biblechapter} % Job 38
\verseWithHeading{Adonai Challenges Job}{Then\lebnote{Hebrew “And”}Adonai answered Job from the storm, and he said,}%
\verse{“Who is this darkening counsel 
by words without knowledge?}%
\verse{\textit{Prepare yourself for a difficult task like a man},\lebnote{Literally “Gird up like a man your loins”}
and I will question you, and you shall declare to me.}%
\verseWithHeading{Adonai Interrogates Job}{“Where were you at my laying the foundation of the earth? 
Tell me, if you possess understanding.}%
\verse{Who determined its measurement? Yes, you do know. 
Or who stretched the measuring line upon it?}%
\verse{On what were its bases sunk? 
Or who laid its cornerstone,}%
\verse{when the morning stars were singing together 
and all the sons of God\lebnote{Or “heavenly beings”}shouted for joy?}%
\verse{“Or who shut the sea in with doors 
at its bursting, when it went out of the womb,}%
\verse{at my making the clouds its garment 
and thick darkness its swaddling band,}%
\verse{and I prescribed my rule for it, 
and I set bars and doors,}%
\verse{and I said, ‘You shall come up to here, but\lebnote{Hebrew “and”}you shall not go further, 
and here it will set a boundary \textit{for your proud surging waves}’?\lebnote{Literally “for the pride of your surging waves”}}%
\verse{“Have you \textit{ever in your life}\lebnote{Literally “from your days”}commanded the morning? 
Have you made the dawn know its place,}%
\verse{to take hold of the earth’s skirts 
so that\lebnote{Hebrew “and”}the wicked might be shaken off from it?}%
\verse{It is changed like clay under a seal, 
and they appear like a garment.}%
\verse{And their light is withheld from the wicked, 
and their uplifted arm is broken.}%
\verse{“Have you entered into the sea’s sources? 
Or have you walked around in the recesses of the deep?}%
\verse{Have the gates of death been revealed to you? 
Or have you seen the gates of deep shadow?}%
\verse{Have you considered closely the earth’s vast expanse? 
Declare it, if you know all of it.}%
\verse{“Where then\lebnote{Or “Where exactly”}is the way where the light dwells? 
And where then\lebnote{Or “where exactly”}is its place,}%
\verse{that you may take it to its territory, 
and that you might discern the paths to its home?}%
\verse{You know, for you were born then, 
and the number of your days is great.}%
\verse{Have you entered into the storehouses of the snow, 
or\lebnote{Hebrew “and”}have you seen the storehouses of the hail,}%
\verse{which I have reserved for the time of trouble, 
for the day of battle and war?}%
\verse{Where then\lebnote{Or “Where exactly”}is the way where the light is distributed, 
where he scatters the east wind upon the earth?}%
\verse{“Who has cut open a channel for the torrents 
and a way for the \textit{thunder bolts},\lebnote{Literally “bolt of thunders”}}%
\verse{to bring rain on a land \textit{where no one lives},\lebnote{Literally “not a man”}
a desert \textit{where no humans live},\lebnote{Literally “not humans in it”}}%
\verse{to satisfy desert and wasteland, 
and to cause the ground to put forth the rising of grass?}%
\verse{Is there a father for the rain, 
or who fathered the drops of dew?}%
\verse{From whose womb did the ice come forth, 
and who fathered the frost of heaven?}%
\verse{Like stone the waters become hard, 
and the faces of the deep freeze.}%
\verse{“Can you bind the chains of the Pleiades, 
or can you loosen the cords of Orion?}%
\verse{Can you lead forth the southern constellations at their appointed time, 
or\lebnote{Hebrew “and”}can you lead the Bear with its children?}%
\verse{Do you know heaven’s statutes, 
or can you establish their rule on the earth?}%
\verse{Can you lift up your voice to the clouds 
so that\lebnote{Hebrew “and”}a flood of water may cover you?}%
\verse{Can you send forth lightnings, that they may go? 
And will they say to you, ‘Here we are’?}%
\verse{Who has put wisdom in the ibis,\lebnote{Or “inner parts”}
or who has given understanding to the rooster?\lebnote{Or “mind”}}%
\verse{Who can number the clouds with wisdom? 
Or\lebnote{Hebrew “And”}who can tilt heaven’s jars,}%
\verse{at the flowing of the dust into a cast 
and the clods cling together?}%
\verse{“Can you hunt prey for the lion? 
And can you satisfy the hunger of strong lions}%
\verse{when they are crouched in the dens, 
when they lie in the thicket in an ambush?}%
\verse{Who prepares for the crow its prey, 
when its young ones cry to God for help, 
and they wander around \textit{for lack of food}?\lebnote{Literally “for no food”}}%
\end{biblechapter}

\begin{biblechapter} % Job 39
\verse{“Do you know the time when the goats of the rocks give birth? 
Do you observe the doe deer’s giving birth?}%
\verse{Can you number the months they fulfill, 
and do you know the time of its giving birth?}%
\verse{When they crouch, they bring forth their young ones; 
they get rid of their labor pains.\lebnote{Or “deliver their fetuses”}}%
\verse{Their young ones grow strong; they grow up in the open; 
they go forth and do not return to them.}%
\verse{“Who has sent forth the wild ass free? 
And who has released the wild donkey’s bonds,}%
\verse{to which I have given the wilderness as its house 
and the salt flat as its dwelling place?}%
\verse{It scorns the city’s turmoil; 
it does not hear the driver’s shouts.}%
\verse{It explores the mountains as its pasture 
and searches after every kind of green plant.}%
\verse{“Is the wild ox willing to serve you, 
or will he spend the night at your feeding trough?}%
\verse{Can you tie the wild ox with its rope to a furrow, 
or will it harrow the valleys after you?}%
\verse{Can you trust it because its strength is great, 
or will you hand your labor over to it?}%
\verse{Can you rely on it that it will return your grain 
and that it will gather it to your threshing floor?}%
\verse{“The wings\lebnote{Hebrew “wing”}of the female ostrich flap\lebnote{Or “flaps”}— 
are they\lebnote{Or “if,” or “or”}the pinions of the stork or\lebnote{Hebrew “and”}the falcon?}%
\verse{Indeed, it leaves its eggs to the earth, 
and it lets them be warmed on the ground,}%
\verse{and it forgets that a foot might crush an egg,\lebnote{Hebrew “it”; or a collective singular (“them”) referring to “eggs” in v. 14}
and \textit{a wild animal}\lebnote{Literally “an animal of the field”}might trample it.\lebnote{Or a collective singular (“them”) referring to “eggs” in v. 14}}%
\verse{It deals cruelly with its young ones, as if they were not its own, 
as if without fear that its labor were in vain,}%
\verse{because God made it forget wisdom, 
and he did not give it a share in understanding.}%
\verse{When it spreads its wings \textit{aloft},\lebnote{Literally “in the height”}
it laughs at the horse and its rider.}%
\verse{“Do you give power to the horse? 
Do you clothe its neck with a mane?}%
\verse{Do you make it leap like the locust? 
The majesty of its snorting is terrifying.}%
\verse{They paw in the valley, and it exults with strength; 
it goes out to meet the battle.}%
\verse{It laughs at danger and is not dismayed, 
and it does not turn back from \textit{before}\lebnote{Literally “from faces of”}the sword.}%
\verse{Upon it the quiver rattles 
along with the flash of the spear and the short sword.}%
\verse{With roar and rage \textit{it races over the ground},\lebnote{Or “it paws the ground”; literally “it swallows the earth/ground”}
and it cannot stand still at the sound of the horn.}%
\verse{\textit{Whenever}\lebnote{Literally “At enough”}a horn sounds, it says, ‘Aha!’ 
And it smells the battle from a distance— 
the thunder of the commanders and the war cry.}%
\verse{“Does the hawk soar by your wisdom? 
Does it spread its wings to the south?}%
\verse{Or does the eagle fly high at your command 
and construct its nest high?}%
\verse{It lives on the rock and spends the night 
\textit{on the rock point and the mountain stronghold}.\lebnote{Literally “on the tooth of the rock and the stronghold”}}%
\verse{From there it spies out the prey; 
its eyes look from far away.}%
\verse{And its young ones lick blood greedily, 
and where the dead carcasses are, there they are.”}%
\end{biblechapter}

\begin{biblechapter} % Job 40
\verse{Then\lebnote{Hebrew “And”}Adonai answered Job and said,}%
\verse{“Shall a faultfinder contend with Shaddai? 
Anyone who argues with God must answer it.”}%
\verseWithHeading{Job Responds to Adonai}{So Job answered Adonai and said,}%
\verse{“Look, I am insignificant. 
What shall I answer you? 
I lay my hand on my mouth.}%
\verse{Once I have spoken, and I will not answer; 
even\lebnote{Hebrew “And”}twice, but\lebnote{Or “and”}I will not proceed.”}%
\verseWithHeading{Adonai Challenges Job Again}{Then\lebnote{Or “And”}Adonai answered Job from the storm, and he said,}%
\verse{“\textit{Prepare yourself for a difficult task like a man},\lebnote{Literally “Gird up like a man your loins”}
and I will question you, and you shall declare to me.}%
\verseWithHeading{Adonai Interrogates Job Again}{“Indeed, would you annul my justice? 
Would you condemn me, so that you might be righteous?}%
\verse{Or do you have an arm like God, 
and can you thunder with a voice like his?}%
\verse{Adorn\lebnote{The particle sometimes translated “please” is present, but an almighty God or sovereign king does not say “please” when commanding his subjects}yourself with pride and dignity, 
and clothe yourself with splendor.}%
\verse{Pour out the overflowing of your anger, 
and look at all the proud, and humble them.}%
\verse{Look at all the proud, humble them, 
and tread down the wicked \textit{where they stand}.\lebnote{Literally “in place of them”}}%
\verse{Hide them in the dust together; 
bind their faces in the grave.}%
\verse{And I will also praise you, 
that your own right hand can save you.}%
\verse{“Look, Behemoth,\lebnote{Some interpret as a hippopotamus}which I have made just as I made you; 
it eats grass like the ox.}%
\verse{Look, its strength is in its loins 
and its power in the muscles of its stomach.}%
\verse{It keeps its tail straight like a cedar; 
the sinews of its thighs are tightly wound.}%
\verse{Its bones are tubes of copper, 
its limbs like rods of iron.}%
\verse{“It is the first of God’s actions; 
the one who made him furnishes it with his sword.\lebnote{See NET; nearly all other English translations: “Only his Maker can draw the sword against him” (NJPS; compare NLT), or “only its Maker can approach it with the sword” (NRSV, NIV, ESV, HCSB, NASU)}}%
\verse{Yes, the mountains yield produce for it, 
and all \textit{wild animals}\lebnote{Literally “all the animals of the field”}play there.}%
\verse{Under the lotus tree it lies, 
in the hiding place of the reeds and in the marsh.}%
\verse{The lotus trees cover it with their shade; 
the wadi’s\lebnote{A seasonal stream that is often dry}poplar trees surround it.}%
\verse{Look, if the river is turbulent, it is not frightened; 
it is confident even though the Jordan rushes against its mouth.}%
\verse{Can anyone take it by its eyes? 
Can he pierce its nose with a snare?}%
\end{biblechapter}

\begin{biblechapter} % Job 41
\verse{\lebnote{Job 41:1–34 in the English Bible is 40:25–41:26 in the Hebrew Bible}“Can you draw out Leviathan with a fishhook? 
Or\lebnote{Hebrew “And”}\textit{can you tie down its mouth with a cord}?\lebnote{Literally “with a cord can you press down its tongue”}}%
\verse{Can you put a rope in its nose? 
Or\lebnote{Hebrew “And”}can you pierce its jawbone with a hook?}%
\verse{Will it make numerous pleas for mercy to you? 
Or will it speak gentle words to you?}%
\verse{Will it make a covenant with you? 
Will you take it as a slave forever?}%
\verse{Will you play with it as with birds 
and put it on a leash for your girls?}%
\verse{Will guildsmen bargain over it? 
Will they divide it between tradesmen?}%
\verse{Can you fill its kin with harpoons 
or\lebnote{Hebrew “and”}its head with fish spears?}%
\verse{Lay your hands on it; 
think about the battle—you will not do it again!}%
\verse{“Look, \textit{the hope of capturing it}\lebnote{Literally “its hope”}is false. 
Will one be hurled down even at its sight?}%
\verse{Is it not fierce when somebody stirs it? 
Who then is he who would stand \textit{before it}?\lebnote{Or “Who is he thenwho could stand before me?”; literally “to the face of me”}}%
\verse{Who has come to confront me, that\lebnote{Hebrew “and”}I should repay him? 
Under all the heavens, \textit{it belongs to me}.\lebnote{Literally “to me”}}%
\verse{“I will not keep quiet concerning its limbs 
or\lebnote{Hebrew “and”}concerning the extent of its might and the gracefulness of its frame.}%
\verse{Who can strip off \textit{its outer covering}?\lebnote{Literally “the surface of its garment”}
Who can penetrate its double harness?}%
\verse{Who can open the doors of its face? 
Its teeth all around are fearsome.}%
\verse{Its back\lebnote{LXX; or “pride”}has scales of shields; 
it is shut up closely as with a seal.}%
\verse{They are close \textit{to one another}—\lebnote{Literally “one to one”}
even\lebnote{Hebrew “and”}the air cannot come between them.}%
\verse{They are joined \textit{one to another};\lebnote{Literally “a man to his brother”}
they cling together and cannot be separated.}%
\verse{“Its snorting flashes forth light, 
and its eyes are \textit{red like dawn}.\lebnote{Literally “like eyelids of dawn”}}%
\verse{Torches go from its mouth; 
sparks of fire shoot out.}%
\verse{Smoke comes from its nostrils 
as from a kettle boiling and burning bulrushes.}%
\verse{Its breath kindles charcoal, 
and a flame comes from its mouth.}%
\verse{“Strength abides in its neck, 
and dismay\lebnote{Or “strength”}dances \textit{before it}.\lebnote{Literally “to the face of him/it”}}%
\verse{Its flesh’s folds of skin cling together; 
it is cast on it—it will not be moved.}%
\verse{Its heart is cast as stone; 
yes,\lebnote{Hebrew “and”}it is cast as the lower millstone.}%
\verse{\textit{When it raises itself},\lebnote{Literally “From his/its rising up”}the mighty ones are terrified; 
they retreat because of its thrashing.}%
\verse{Reaching it with the sword does not avail, 
nor with the spear, the dart, or\lebnote{Hebrew “and”}the javelin.}%
\verse{It regards iron as straw, 
bronze as rotten wood.}%
\verse{\textit{An arrow}\lebnote{Literally “A son of a bow”}will not make it flee; 
sling stones are turned to stubble for it.}%
\verse{Clubs are regarded as stubble, 
and it laughs at the short sword’s rattle.}%
\verse{“Its underparts are shards of a potsherd; 
it moves over mud like a threshing sledge.}%
\verse{It makes the deep boil like a cooking pot; 
it makes the sea like a pot of ointment.}%
\verse{Behind it, \textit{it leaves a glistening wake};\lebnote{Literally “it lights up the path”}
one would think that the deep has gray hair.}%
\verse{“\textit{On the ground it has no equal}\lebnote{Literally “There is not on the ground its likeness”}— 
a\lebnote{Hebrew “the”}creature without fear.}%
\verse{It observes all the lofty; 
it is king \textit{over all that are proud}.”\lebnote{Literally “over all sons of pride”}}%
\end{biblechapter}

\begin{biblechapter} % Job 42
\verseWithHeading{Job’s Repentance and Restoration}{Then\lebnote{Hebrew “And”}Job answered Adonai and said,}%
\verse{“I know that you can do all things, 
and any scheme from you will not be thwarted.}%
\verse{‘Who is this darkening counsel without knowledge?’ 
Therefore I uttered, but\lebnote{Hebrew “and”}I did not understand; 
\textit{things too wonderful for me},\lebnote{Literally “wonderful things from me”}but I did not know.}%
\verse{‘Hear and I will speak; 
I will question you, then\lebnote{Hebrew “and”}inform me.’}%
\verse{By the ear’s hearing I heard of you, 
but\lebnote{Hebrew “and”}now my eye has seen you.}%
\verse{Therefore I despise myself 
and repent in dust and ashes.”}%
\verse{\textit{And then}\lebnote{Literally “And it happened”}after Adonai spoke these words to Job, Adonai said to Eliphaz the Temanite, “\textit{My wrath has been kindled}\lebnote{Literally “My nose became hot”}against you\lebnote{Singular}and against the two of your friends, for you\lebnote{Plural}have not spoken to me what is right as my servant Job has.}%
\verse{\textit{So then},\lebnote{Or “And now,” or “Therefore”}take for yourselves seven bulls and seven rams, and go to my servant Job and offer a burnt offering for yourselves. And my servant Job will pray for you,\lebnote{Plural}for \textit{I will certainly accept his prayer},\lebnote{Literally “if his face I will lift up,” or “indeed his face I will lift up” (“if” used as an oath particle)}so that it will not be done with you\lebnote{Plural}according to your folly, for you\lebnote{Plural}have not spoken to me what is right as my servant Job has.”}%
\verse{So\lebnote{Hebrew “And”}Eliphaz the Temanite and Bildad the Shuhite and Zophar the Naamathite went and did just as Adonai had told them, and \textit{Adonai accepted Job’s prayer}.\lebnote{Literally “Adonai lifted up the face of Job”}}%
\verse{Then\lebnote{Hebrew “And”}Adonai returned Job’s fortune when he prayed to him on behalf of his friends. Thus\lebnote{Hebrew “And”}Adonai increased \textit{all that Job had twice as much as before}.\lebnote{Literally “all which for Job to twice”}}%
\verse{So\lebnote{Hebrew “And”}all his brothers and all his sisters and all those who had known him \textit{before}\lebnote{Literally “to faces”}came to him, and they ate bread\lebnote{Or “food”}with him in his house and showed sympathy to him and comforted him for all the disaster\lebnote{Or “evil”}that Adonai had brought upon him. Then\lebnote{Hebrew “And”}each one gave to him one piece of money, and each one gave to him one ornamental ring of gold.}%
\verse{So\lebnote{Hebrew “And”}Adonai blessed Job’s latter days more than his beginning. \textit{Thus he had}\lebnote{Literally “And it came to be for him,” or “And it was for him”}fourteen thousand sheep and goats and six thousand camels and a thousand pair of oxen and a thousand female donkeys.}%
\verse{\textit{And he had}\lebnote{Literally “And it came to be for him,” or “And it was for him”}seven sons and three daughters.}%
\verse{And he called the name of the first Jemimah and the name of the second Keziah and the name of the third Qeren-Happuk.}%
\verse{And beautiful women were not found in all the land like Job’s daughters, and their father gave to them an inheritance in the midst of their brothers.}%
\verse{And Job lived after this one hundred and forty years, and he saw his sons and his \textit{grandsons}\lebnote{Literally “the sons of his sons”}for four generations.}%
\verse{Then\lebnote{Hebrew “And”}Job died old and full of days.}%
\end{biblechapter}

