\biblebook{Joel}

\begin{biblechapter} % Joel 1
\verse The word of Adonai that came to Joel son of Pethuel.
\verseWithHeading{A Lament Over the Land} Hear this, O elders, 
and give ear, all the inhabitants of the land. 
Has this happened in your days? 
Or the days of your ancestors?
\verse Tell it to your children, 
and your children to their children, 
and their children to the following generation.
\verse What the cutting locust left, 
the swarming locust has eaten. 
What the swarming locust left, 
the hopping locust has eaten. 
And what the hopping locust left, 
the destroying locust has eaten.
\verse Wake up, drunkards, and weep! 
Wail, all drinkers of wine, over the new wine, 
for it is cut off from your mouth.
\verse Because a nation has \textit{invaded}\lebnote{Literally “come up against”} my land, 
strong \textit{and beyond counting}.\lebnote{Literally “without number”} 
Its teeth are the teeth of a lion, 
and its fangs are those of a lioness.
\verse It has made my vine a desolation, 
and my fig tree a completely splintered stump. 
It has stripped them bare and thrown them down; 
their branches have turned white.
\verse Lament like a virgin girded in sackcloth 
for the husband of her youth.
\verse The offering and libation are withheld 
from the house\lebnote{Or “temple”} of Adonai. 
The priests mourn, 
the ministers of Adonai.
\verse The field is destroyed; 
the earth mourns 
because the grain is destroyed, 
the new wine dries up, 
the olive oil languishes.
\verse Be ashamed, farmers; 
Wail, vinedressers, 
over the wheat and over the barley, 
because the harvest of the field is ruined.
\verse The vine withers 
and the fig tree droops. 
The pomegranate tree, and also the palm tree, the apple tree 
—all the trees of the field—are dried up. 
Indeed, joy is dried up 
among the sons of men.
\verse Gird yourselves and lament, O priests! 
Wail, O ministers of the altar! 
Come spend the night in sackcloth, 
O ministers of my God, 
because offering and libation 
are withheld from the house of your God.
\verse Solemnize a fast! 
Call an assembly! 
Gather the elders, 
all of the inhabitants of the land 
in the house of Adonai your God, 
and cry out to Adonai.
\verse Ah! For the day! 
For the day of Adonai is near. 
It will come like destruction from Shaddai.\lebnote{Often translated “the Almighty”}
\verse Is not food cut off 
before our eyes, 
from the house of our God, 
joy and gladness?
\verse The seeds shrivel under their clods; 
the storehouses are desolate. 
The grain storage places are destroyed 
because grain has dried out.
\verse How the beasts\lebnote{Hebrew “beast”} groan; 
the herds of cattle wander around 
because there is no pasture for them; 
the flocks of sheep are in distress.
\verse To you, O Adonai, I cry out, 
because fire has devoured 
the pastures of the desert, 
and flames\lebnote{Hebrew “flame”} burned 
all the trees of the field.
\verse Also, the beasts of the field 
long for you, 
because the courses of water 
are dried up, 
and fire has devoured 
the pastures of the desert.
\end{biblechapter}

\begin{biblechapter} % Joel 2
\verseWithHeading{The Day of Adonai is Near} Blow the trumpet\lebnote{Hebrew “shofar”} in Zion, 
and sound the alarm on \textit{my holy mountain}!\lebnote{Literally “mountain of my holiness”} 
Let all the inhabitants of the land tremble, 
for the day of Adonai is coming 
—it is indeed near.
\verse A day of darkness and gloom, 
a day of cloud and thick darkness, 
like the dawn spreads on the mountains, 
a great and strong army! 
There has been nothing like it from old, 
and after it nothing will be again \textit{for generations to come}.\lebnote{Literally “until years of generations”}
\verse \textit{Before them}\lebnote{Literally “before the presence of it”} a fire devours, 
and behind them a flame burns. 
Like the garden of Eden is the land \textit{before them},\lebnote{Literally “before the presence of it”} 
and after them it is like a desolate desert, 
and nothing can escape them.
\verse Like the appearance of horses is their appearance, 
and like horsemen they run;
\verse like the sound of chariots on the tops of the mountains, they leap about; 
like the sound of a flame of fire 
devouring stubble; 
like a strong army arranged 
in rows for battle.
\verse \textit{From before them}\lebnote{Or “From in front of it,” or “From the presence of it”} nations writhe, 
all faces \textit{turn pale}.\lebnote{Literally “gather paleness”}
\verse They run like mighty warriors,\lebnote{That is, the “great and strong army” (see 2:2)} 
they scale the wall like men of war; 
each goes on its own way, 
and they do not swerve from their paths.
\verse They do not jostle \textit{one another};\lebnote{Literally “each man his brother”} 
each goes on its own trail; 
and through the falling weapons,\lebnote{Hebrew “weapon”} 
they are not halted.
\verse In the city they rush forth; 
on the walls\lebnote{Hebrew “wall”} they run. 
Into the houses they climb up; 
through the windows 
they enter like a thief.
\verse \textit{Before them}\lebnote{Literally “Before the presence of it”} the earth quakes; 
the heavens tremble; 
the sun and the moon grow dark, 
and the stars \textit{have withheld}\lebnote{Literally “gathered”} their splendor
\verse And Adonai \textit{utters}\lebnote{Literally “gives”} his voice before his army, 
because his encampment is very large; 
strong is the one who carries out his decree, 
for great is the day of Adonai 
and exceedingly fearful. 
Who can endure it?
\verseWithHeading{A Call to Repentance} “And even now,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“return to me with all your heart, 
with fasting, and weeping, and wailing.”
\verse Rend your hearts\lebnote{Hebrew “heart”} and not your garments, 
and return to Adonai your God, 
because he is gracious and compassionate, 
\textit{slow to anger}\lebnote{Or “patient”; literally “long of anger”} and great in loyal love, 
and relenting from harm.
\verse Who knows whether he will turn and relent, 
and leave a blessing behind him, 
an offering and a libation, 
for Adonai your God?
\verse Blow the trumpet\lebnote{Hebrew “Shofar”} in Zion, 
sanctify a fast, call an assembly;
\verse gather the people, consecrate the assembly; 
assemble the elders, gather the children, 
even those who are breast-feeding; 
let the bridegroom come out from his private room, 
and the bride from her canopy.
\verse Between the colonnade and the altar, 
let the priests, the ministers of Adonai, weep. 
And let them say, “Take pity, Adonai, on your people. 
Do not make your inheritance a reproach, 
a byword among the nations. 
Why should they say among the nations, 
‘Where is their God?’ ”
\verseWithHeading{The Response and Promise of Adonai} Then Adonai became jealous for his land 
and took pity on his people.
\verse And Adonai answered and said to his people, 
“Look at me, I am sending to you 
grain, new wine, and olive oil, 
and you will be satisfied by it. 
I will not give you over any more 
as a disgrace among the nations.
\verse \textit{The northerners}\lebnote{Or “The northern army”} 
I will remove from you; 
I will drive them to a desert 
and desolate land, 
its\lebnote{That is, the northern army} front to the eastern sea, 
and its rear into the western sea; 
its stench and odor will rise up 
because he\lebnote{That is, Adonai (see v. 21)} has done great things.
\verse Do not fear, O land, 
rejoice and be glad, 
because Adonai has done great things.
\verse Do not fear, wild animals of the field, 
because the pastures of the desert have put forth new green shoots, 
because the tree has produced its fruit, 
the fig tree and the vine 
have yielded their produce.
\verse Be glad, O children of Zion, 
be glad and rejoice in Adonai your God, 
because he has given for you 
the autumn rains for your righteousness,\lebnote{Or “vindication”} 
and he has \textit{poured down}\lebnote{Literally “brought down”} for you rainwater, 
the autumn and spring rains, as before.
\verse The threshing floors will be full with grain, 
and the vats will overflow 
with new wine and olive oil.
\verse I will repay you the years 
that the locust has eaten, 
the hopper, the destroyer, and the cutter, 
my mighty troops that I sent against you.
\verse And you will eat abundantly and be satisfied, 
and praise the name of Adonai your God, 
who has dealt with you \textit{wondrously}.\lebnote{Literally “to do a wonder”} 
My people \textit{will never be ashamed}.\lebnote{Literally “will forever not be ashamed”}
\verse And you will know that I am in the midst of Israel, 
and I am Adonai, 
your God, and there is no other. 
My people \textit{will never be ashamed again}.\lebnote{Literally “will forever no longer be ashamed”}
\verseWithHeading{Adonai’s Spirit Poured Out} \lebnote{Joel 2:28–32 in the English Bible is 3:1–5 in the Hebrew Bible} And it will happen afterward thus: 
I will pour out my Spirit on all flesh, 
and your sons and your daughters will prophesy, 
and your elders will dream dreams; 
your young men shall see visions.
\verse And also on the male slaves and on the female slaves, 
I will pour out my Spirit in those days.
\verse And I will \textit{set}\lebnote{Literally “give”} wonders in the heavens, and on earth, blood and fire and columns of smoke.
\verse The sun will be changed to darkness, and the moon to blood, before the coming of the great and awesome day of Adonai.
\verse And it will happen—everyone who calls on the name of Adonai will be rescued, because on Mount Zion and in Jerusalem there will be \textit{those who escape},\lebnote{Literally, “an escape”} as Adonai said, and among the survivors whom Adonai is calling.
\end{biblechapter}

\begin{biblechapter} % Joel 3
\verseWithHeading{Adonai’s Judgment on the Nations} \lebnote{Joel 3:1–21 in the English Bible is 4:1–21 in the Hebrew Bible} For look! In those days, and in that time, when I will return\lebnote{Or “restore”} the fortunes of Judah and Jerusalem,
\verse I will gather all the nations, and I will bring them down to the valley of Jehoshaphat, and I will argue \textit{a case against them}\lebnote{Literally “a matter with them”} there concerning my people and my inheritance Israel whom they have scattered among the nations, and my land that they have divided.
\verse For my people they cast lots,\lebnote{Hebrew “lot”} and \textit{they traded}\lebnote{Literally “gave”} the male child for the prostitute, and the female child they sold for wine and they drank it.
\verse What are you to me, O Tyre and O Sidon, and all of the regions of Philistia? Are you repaying to me what is deserved? If you are recompensing me, I will return swiftly and quickly what you deserve on your head!
\verse For you have taken my silver and my gold, and my beautiful treasures you have carried into your temples.
\verse And the sons of Judah and Jerusalem you sold to the sons of the Greeks, in order to remove them from their border.
\verse Look! I am rousing them from the place where you have sold them, and I will return what you deserve on your head!
\verse I will sell your sons and your daughters into the hand of the children of Judah, and they will sell them to the Sabeans and to a nation far away, for Adonai has spoken.
\verseWithHeading{Adonai’s Judgment in the Valley of Jehoshaphat} Proclaim this among the nations: 
\textit{Prepare for war}!\lebnote{Literally “Sanctify a battle”} 
Stir up the mighty warriors; 
let them approach and come up; 
let all the men of war approach.
\verse Beat your cutting tools of iron into swords 
and your pruning hooks into spears; 
let the weakling say, ‘I am a mighty warrior!’
\verse Hurry and come, 
all the nations, from all around, 
and gather yourselves there. 
Bring down your mighty warriors, O Adonai!
\verse Let the nations be roused and let them come up 
to the valley of Jehoshaphat, 
for there I will sit to judge 
all the nations from all around.
\verse Send forth the sickle, 
for the harvest is ripe! 
Go tread, 
for the winepress is full! 
The vats overflow, 
because their evil is great!
\verse Commotion, commotion 
in the valley of decision! 
For the day of Adonai is near 
in the valley of decision!
\verse The sun and the moon grow dark, 
and the stars \textit{have withheld}\lebnote{Literally “gathered”} their splendor.
\verse And Adonai roars from Zion; 
from Jerusalem he utters his voice, 
and the heavens and the earth shake. 
But Adonai is a refuge for his people, 
and a protection for the children of Israel.
\verse And you will know that I, Adonai your God, 
am dwelling in Zion, my holy mountain. 
And Jerusalem will be a place of holiness, 
and strangers will pass through it no longer.
\verse And it will happen on that day; 
the mountains will drip new wine, 
and the hills will flow with milk, 
and all the channels of Judah will flow with water. 
A spring from the house of Adonai will come forth, 
and it will water the valley of Acacia Trees.\lebnote{Or “the wadi Shittim”}
\verse Egypt will become a desolation, 
and Edom will become a \textit{desolate desert},\lebnote{Literally “desert of desolation”} 
because of the violence they did against the children of Judah, 
in whose land they have shed innocent blood.
\verse But Judah will be inhabited forever, 
and Jerusalem \textit{for all generations}.\lebnote{Or “for each generation”}
\verse \textit{I will cleanse their bloodguilt}\lebnote{Literally “I will not leave unpunished their blood”} that \textit{I did not cleanse},\lebnote{Or “I did not consider innocent”} 
for Adonai is dwelling in Zion.
\end{biblechapter}

