\biblebook{Leviticus}

\begin{biblechapter} % Leviticus 1
\verseWithHeading{Laws for Burnt Offerings}{Then\lebnote{Or “And”}Adonai called to Moses and spoke to him from the tent of assembly, saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}and say to them, ‘When a person\lebnote{Or “man”—the singular noun is generic, thus the “you” and “your” in the remainder of the verse are all plural}from you presents an offering to Adonai, you shall present your offering from domestic animals, from the cattle\lebnote{Or “herd”}or from the flock.\lebnote{The Hebrew term refers collectively to both sheep and goats (small livestock animals)}}%
\verse{If his offering is a burnt offering from the cattle,\lebnote{Or “herd”}then he must present\lebnote{Hebrew “present it”}an unblemished male; he must present it at the door of the tent of assembly for his acceptance \textit{before}\lebnote{Literally “to the face of”}Adonai.}%
\verse{“ ‘He\lebnote{Or “And he”}must lay his hand on the head of the burnt offering and it will be accepted\lebnote{Same root word as “acceptance” in v. 3}for him in order to make atonement for him.}%
\verse{He\lebnote{Or “And he”}must slaughter \textit{the young bull}\lebnote{Literally “the son of the herd” or “the son of the cattle”}\textit{before}\lebnote{Literally “to the face of”}Adonai, then\lebnote{Or “and”}Aaron’s sons, the priests, will present the blood and sprinkle the blood all around the altar that is at the door of the tent of assembly.}%
\verse{Then\lebnote{Or “And”}he must remove the skin of the burnt offering and cut it into its pieces.}%
\verse{The sons of\lebnote{Or “And the sons of”}Aaron the priest will put fire on the altar and arrange the wood on the fire.}%
\verse{Then\lebnote{Or “And”}Aaron’s sons, the priests, will arrange the pieces of meat,\lebnote{Verse 6 in the context uses the same word with reference to the meat}the head, and the suet on the wood that is on the fire that is on the altar.}%
\verse{Then\lebnote{Or “And”}he must wash its inner parts\lebnote{Or “entrails”}and its lower leg bones with water, and the priest will turn into smoke the whole animal on the altar as a burnt offering by fire, as an appeasing fragrance for Adonai.}%
\verse{“ ‘But if his offering is from the flock,\lebnote{The Hebrew term refers collectively to both sheep and goats (small livestock animals)}from the sheep or from the goats for a burnt offering, he must present\lebnote{Hebrew “present it”}an unblemished\lebnote{Same word as in v. 3}male.}%
\verse{He\lebnote{Or “And he”}must slaughter it on the north side of the altar \textit{before}\lebnote{Literally “to the face of”}Adonai; then\lebnote{Or “and”}Aaron’s sons, the priests, will sprinkle its blood all around the altar.}%
\verse{Then\lebnote{Or “And”}he must cut it into pieces along with its head and its suet; and the priest will arrange them on the wood that is on the fire that is on the altar.}%
\verse{Then\lebnote{Or “And”}he must wash the inner parts\lebnote{Or “entrails”}and the lower leg bones with water and the priest shall present the whole animal and will turn it\lebnote{Understood direct object}into smoke on the altar; it is a burnt offering by fire as an appeasing fragrance for Adonai.}%
\verse{“ ‘But if his offering for Adonai is a burnt offering from the birds, then he must present his offering from the turtledoves or from \textit{the young doves}.\lebnote{Literally “the sons of the dove”}}%
\verse{The priest\lebnote{Or “And the priest”}will present it at the altar and must wring off its head and turn it into smoke on the altar, and its blood will be drained out on the wall\lebnote{Or “side”}of the altar.}%
\verse{He\lebnote{Or “And he”}must remove its crop with its plumage and throw it to the east beside the altar on the place of the fatty ashes.}%
\verse{Then\lebnote{Or “And”}he must tear it apart by its wings but must not sever it; then\lebnote{Or “and”}the priest will turn it into smoke on the altar. It is a burnt offering by fire as an appeasing fragrance for Adonai.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 2
\verseWithHeading{Laws for Grain Offerings}{“ ‘When a person\lebnote{Or “a soul”}brings a grain offering to Adonai, his offering must be finely milled flour, and he must pour out oil on it and place frankincense on it.}%
\verse{And he shall bring it to the sons of Aaron, the priests, and he\lebnote{That is, the priest—see v. 9}shall take \textit{his handful from its finely milled flour}\lebnote{Literally “from there his handful from its finely milled flour”}and from its oil in addition to all its frankincense. The priest\lebnote{Or “And the priest”}shall turn its token portion into smoke on the altar as an offering made by fire, as an appeasing fragrance for Adonai.}%
\verse{The remainder\lebnote{Or “And the remainder”}of the grain offering \textit{belongs to}\lebnote{Literally “for”}Aaron and to his sons—\textit{it is a most holy thing}\lebnote{Literally “a holiness of holinesses”}from the offerings made by fire for\lebnote{Hebrew “of”}Adonai.}%
\verse{“ ‘But if you bring a grain offering of something oven-baked, it must be of finely milled flour as ring-shaped unleavened bread mixed with oil or wafers of unleavened bread smeared with oil.}%
\verse{If your offering is a grain offering baked on a\lebnote{Hebrew “the”}flat baking pan, it must be finely milled flour, unleavened bread mixed with oil;}%
\verse{break it into pieces and pour out oil on it; it is a grain offering.}%
\verse{“ ‘If\lebnote{Or “And if”}your offering is a grain offering prepared in a cooking pan, it must be with finely milled flour in oil.}%
\verse{And you shall bring the grain offering that is made from these things to Adonai, and the offerer\lebnote{Changing from 2ms in the first verb to 3ms in the second verb apparently moves from general to specific; NET takes the second verb as an imperative (“Present it”), and NJPS translates it as an indefinite 3ms, making it passive (“it shall be brought”)}shall bring it to the priest, and he shall bring it to the altar.}%
\verse{And the priest shall take away from the grain offering its token portion, and he shall turn it into smoke on the altar as an offering made by fire, as an appeasing fragrance for Adonai.}%
\verse{And the remainder of the grain offering \textit{belongs to}\lebnote{Literally “for”}Aaron and to his sons—\textit{it is a most holy thing}\lebnote{Literally “a holiness of holinesses”}from the offerings made by fire for\lebnote{Hebrew “of”}Adonai.}%
\verse{“ ‘Every grain offering you\lebnote{The first time a plural 2m verb has been employed since 1:2}bring to Adonai must not be made of yeasted food, because you must not turn into smoke any yeast or any honey from\lebnote{Hebrew “from it”}an offering made by fire for Adonai.}%
\verse{As an offering of the choicest portion, you\lebnote{Another occurrence of the 2mp}may bring them to Adonai, but they must not be offered on the altar as an appeasing fragrance.}%
\verse{Also all\lebnote{Or “And all”}of your grain offerings you must season with salt; you\lebnote{Or “and you”}must not omit the salt of your God’s covenant from your offering.}%
\verse{“ ‘And if you bring to Adonai a grain offering of firstfruits, you must bring an ear of new grain roasted by fire, coarsely crushed ripe grain, as the grain offering of your firstfruits.}%
\verse{And you shall put oil on it and place frankincense on it; it is a grain offering.}%
\verse{The\lebnote{Or “And the”}priest shall turn into smoke its token portion from its coarsely crushed grain together with all of its frankincense—it is an offering made by fire for Adonai.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 3
\verseWithHeading{Laws for Fellowship Offerings}{“ ‘Now if\lebnote{Or “And if”}his offering is a sacrifice of fellowship offering, if he brings it from the cattle,\lebnote{Or “the herd”}whether male or female, he must bring it without defect before Adonai.}%
\verse{He\lebnote{Or “And he”}must lay his hand on the head of his offering and slaughter it at the entrance of the tent of assembly, and Aaron’s sons shall sprinkle the blood on the altar all around.}%
\verse{He shall present\lebnote{Or “And he shall present”}from the sacrifice of the fellowship offering an offering made by fire for Adonai consisting of\lebnote{The Hebrew accusative implies this transition}the fat covering the inner parts\lebnote{Or “entrails”}and all the fat that is on the inner parts,\lebnote{Or “entrails”}}%
\verse{the two kidneys,\lebnote{Or “and the two kidneys”}and the fat that is on them, which is on the loins, and he must remove the lobe on the liver in addition to the kidneys.}%
\verse{Aaron’s sons shall turn it to smoke on the altar in addition to the burnt offering that is on the wood, which is on the fire; it is an offering made by fire as an appeasing fragrance for Adonai.}%
\verse{“ ‘But if his offering for a sacrifice of fellowship offering for Adonai is from the flock,\lebnote{The Hebrew term refers collectively to both sheep and goats (small livestock animals)}he must bring a male or a female without defect.}%
\verse{If he brings a sheep as his offering, then\lebnote{Or “and”}he shall present it before Adonai,}%
\verse{and he shall lay his hand on the head of his offering, and he shall slaughter it before the tent of assembly, and Aaron’s sons shall sprinkle its blood on the altar all around.}%
\verse{He shall present\lebnote{Or “And he shall present”}from the sacrifice of the fellowship offering an offering made by fire for Adonai: he must remove its fat, the entire fat tail near the tailbone, and the fat covering the inner parts\lebnote{Or “entrails”}and all the fat that is on the inner parts,\lebnote{Or “entrails”}}%
\verse{the two kidneys,\lebnote{Or “and the two kidneys”}and the fat that is on them, which is on the loins, and he must remove the lobe on the liver in addition to the kidneys.}%
\verse{The priest\lebnote{Or “And the priest”}shall turn it into smoke on the altar as a food offering made by fire for Adonai.}%
\verse{“ ‘And\lebnote{Or “But”}if his offering is a goat, then\lebnote{Or “and”}he shall bring it before Adonai,}%
\verse{and he shall lay his hand on the head of his offering, and he shall slaughter it before the tent of assembly, and Aaron’s sons shall sprinkle its blood on the altar all around.}%
\verse{He shall present\lebnote{Or “And he shall present”}his offering from it as an offering made by fire for Adonai: the fat covering the inner parts\lebnote{Or “entrails”}and all the fat that is on the inner parts,\lebnote{Or “entrails”}}%
\verse{the two kidneys,\lebnote{Or “and the two kidneys”}and the fat that is on them, which is on the loins, and he must remove the lobe on the liver in addition to the kidneys.}%
\verse{The priest\lebnote{Or “And the priest”}shall turn them into smoke on the altar as a food offering;\lebnote{Compare v. 11}all the fat is an offering made by fire as an appeasing fragrance for Adonai.}%
\verse{This is\lebnote{Understood by context}a lasting statute for your generations in all your dwellings: you must not eat any fat or any blood.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 4
\verseWithHeading{Laws for Sin Offerings}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}saying, ‘If a person\lebnote{Or “a soul”}sins by an unintentional wrong from any of Adonai’s commands that should not be \textit{violated},\lebnote{Literally “done”}and he \textit{violates}\lebnote{Literally “does”}\textit{any of them}\lebnote{Literally “from one from them”; see NET}—}%
\verse{if the anointed priest sins, \textit{bringing guilt on the people},\lebnote{Literally “to the guilt of the people”}then,\lebnote{Or “and”}concerning the sin that he has \textit{committed},\lebnote{Literally “sinned”}he shall bring \textit{a young bull}\lebnote{Literally “a bull, a son of cattle”}without defect for Adonai as a sin offering.}%
\verse{He shall bring\lebnote{Or “And he shall bring”}the bull to the tent of assembly’s entrance before Adonai, place\lebnote{Or “And place”}his hand on the bull’s head, and slaughter the bull before Adonai.}%
\verse{The anointed priest\lebnote{Or “And the anointed priest”}shall take \textit{some of}\lebnote{Literally “from”}the bull’s blood and shall bring it to the tent of assembly,}%
\verse{and the priest shall dip his finger in the blood and shall spatter \textit{some of}\lebnote{Literally “from”}the blood seven times before Adonai in front of the sanctuary’s curtain.}%
\verse{The priest\lebnote{Or “And the priest”}shall put \textit{some of}\lebnote{Literally “from”}the blood on the horns of the altar of fragrant incense before Adonai, which is in the tent of assembly, and all the rest\lebnote{Implied by the context}of the bull’s blood he must pour out on the base of the altar of the burnt offering, which is at the entrance of the tent of assembly.}%
\verse{“ ‘He must remove\lebnote{Or “And he must remove”}all the fat from the bull of the sin offering: the fat that covers the inner parts\lebnote{Or “entrails”}and all of the fat that is on the inner parts,\lebnote{Or “entrails”}}%
\verse{the two\lebnote{Or “and the two”}kidneys and the fat that is on them, and the liver’s lobe that he must remove in addition to the kidneys—}%
\verse{just as it is removed from the ox of the fellowship offerings’ sacrifice—and the priest shall turn them into smoke on the altar of the burnt offering.}%
\verse{But\lebnote{Or “And”}the bull’s skin and its meat, in addition to its head, \textit{its lower leg bones},\lebnote{Literally “and in addition to its lower leg bones”}its\lebnote{Or “and its”}inner parts,\lebnote{Or “entrails”}and its offal—}%
\verse{he shall carry\lebnote{Or “and he shall carry out”}all of the bull \textit{outside the camp}\lebnote{Literally “to from an outside place of the camp”}to a ceremonially clean place, to the fatty ashes’ dump, and he shall burn it on wood in the fire; it must be burned up on the fatty ashes’ dump.}%
\verse{“ ‘If\lebnote{Or “And if”}Israel’s whole assembly did wrong unintentionally and the matter\lebnote{Or “thing”}was concealed from the assembly’s eyes, and they acted\lebnote{Or “did”}against \textit{any of}\lebnote{Literally “one from all of”}Adonai’s commands that should not be \textit{violated},\lebnote{Literally “done”}so that\lebnote{Or “and”}they incur guilt,}%
\verse{when\lebnote{Or “and”}the sin that they have \textit{committed}\lebnote{Literally “sinned”}against that command\lebnote{The 3fs pronominal suffix may refer to “one of Adonai’s commands”—“one” is also fs}becomes known, the assembly\lebnote{Or “and the assembly”}shall present \textit{a young bull}\lebnote{Literally “a bull, a son of cattle”}as a sin offering, and they shall bring it before the tent of assembly.}%
\verse{And the elders of the community shall place their hands on the bull’s head before Adonai, and he\lebnote{Or “someone”; see NET—the 3ms refers to either one of the elders or the priest}shall slaughter the bull before Adonai.}%
\verse{Then the anointed priest shall bring \textit{some of}\lebnote{Literally “from”}the bull’s blood to the tent of assembly,}%
\verse{and the priest shall dip his finger in\lebnote{Or “from”}the blood and shall spatter it seven times before Adonai in front of the curtain.}%
\verse{He must put\lebnote{Or “And he must put”}\textit{some of}\lebnote{Literally “from”}the blood on the altar’s horns before Adonai \textit{in the tent of assembly},\lebnote{Literally “which is in the tent of assembly”—the reference is to the altar}and all the rest\lebnote{Indicated by the context}of the blood he must pour out on the base of the altar of the burnt offering, which is at the tent of assembly’s entrance.}%
\verse{He must remove\lebnote{Or “And he must remove”}all its fat from it, and he shall turn it into smoke on the altar.}%
\verse{He shall do\lebnote{Or “And he shall do”}to the bull \textit{just as}\lebnote{Literally “as that”}he did to the sin offering’s bull, so he must do to it. The priest\lebnote{Or “And the priest”}shall make atonement for them, and \textit{they will be forgiven}.\lebnote{Literally “it shall be forgiven to them”}}%
\verse{He shall bring\lebnote{Or “And he shall bring out”}the rest of\lebnote{Indicated by the context}the bull \textit{outside the camp},\lebnote{Literally “to from an outside place of the camp”}and he shall burn it \textit{just as}\lebnote{Literally “as that”}he burned the first bull; it is the sin offering for the assembly.}%
\verse{“ ‘When a leader sins and commits\lebnote{Or “does”}an unintentional wrong against \textit{any of}\lebnote{Literally “one from all of”}the commands of Adonai his God that should not be \textit{violated},\lebnote{Literally “done”}so that\lebnote{Or “and”}he incurs guilt,}%
\verse{or his sin he has \textit{committed}\lebnote{Literally “sinned”}is made known to him, he shall bring\lebnote{Or “and he shall bring” or “then he shall bring”}\textit{a male goat}\lebnote{Literally “a he-goat of goats a male”}without defect as his offering.}%
\verse{He shall place\lebnote{Or “And he shall place”}his hand on the he-goat’s head and slaughter it in the place where he slaughtered the burnt offering before Adonai; it is a sin offering.}%
\verse{The priest\lebnote{Or “And the priest”}shall take \textit{some of}\lebnote{Literally “from”}the sin offering’s blood with his finger, and he shall put it on the horns of the altar of the burnt offering, and he must pour out the rest of\lebnote{Indicated by context}its blood on the base of the altar of the burnt offering.}%
\verse{He\lebnote{Or “And he”; the antecedent is the priest (cp. vv. 10, 31)}must turn all of its fat into smoke on the altar like the fat of the fellowship offerings’ sacrifice, and the priest shall make atonement for him \textit{because of}\lebnote{Literally “from”}his sin, and \textit{he will be forgiven}.\lebnote{Literally “it shall be forgiven to him”}}%
\verse{“ ‘If\lebnote{Or “And if”}\textit{anyone}\lebnote{Literally “a soul one”}of the people of the land sins by an unintentional wrong by \textit{violating}\lebnote{Literally “doing”}one of Adonai’s commands that should not be \textit{violated},\lebnote{Literally “done”}so that\lebnote{Or “and”}he incurs guilt,}%
\verse{or his sin he has \textit{committed}\lebnote{Literally “sinned”}is made known to him, he shall bring\lebnote{Or “and he shall bring” or “then he shall bring”}as his offering \textit{a female goat without defect}\lebnote{Literally “a she-goat of goats without defect a female”}as his offering for his sin that he \textit{committed}.\lebnote{Literally “sinned”}}%
\verse{He shall place\lebnote{Or “And he shall place”}his hand on the sin offering’s head and slaughter the sin offering in the place of the burnt offering.}%
\verse{The priest\lebnote{Or “And the priest”}shall take \textit{some of}\lebnote{Literally “from”}its blood with his finger, and he shall put it on the horns of the altar of the burnt offering, and he must pour out all the rest of\lebnote{Indicated by context}its blood on the altar’s base.}%
\verse{He\lebnote{Or “And he”}must remove all of its fat \textit{just as}\lebnote{Literally “as that”}the fat was removed from\lebnote{Or “from on” or “from upon”}the fellowship offerings’ sacrifice, and the priest shall turn it into smoke on the altar as an appeasing fragrance for Adonai. The priest\lebnote{Or “And the priest”}shall make atonement for him, and \textit{he will be forgiven}.\lebnote{Literally “it shall be forgiven to him”}}%
\verse{“ ‘But\lebnote{Or “And”}if he brings a lamb as his offering for a sin offering, he must bring a female without defect.}%
\verse{He shall place\lebnote{Or “And he shall place”}his hand on the sin offering’s head, and he shall slaughter it as a sin offering in the place where he slaughtered the burnt offering.}%
\verse{The priest\lebnote{Or “And the priest”}shall take \textit{some of}\lebnote{Literally “from”}the sin offering’s blood with his finger, and he shall put it on the horns of the altar of the burnt offering, and he must pour out all the rest\lebnote{Indicated by context}of its blood on the altar’s base.}%
\verse{He must remove\lebnote{Or “And he must remove”}all of its fat \textit{just as}\lebnote{Literally “as that”}the lamb’s fat from the fellowship offerings’ sacrifice was removed, and the priest shall turn them into smoke on the altar upon Adonai’s offerings made by fire; and the priest shall make atonement for him because of\lebnote{Or “for”}his sin that he \textit{committed},\lebnote{Literally “sinned”}and \textit{he will be forgiven}.’ ”\lebnote{Literally “it will be forgiven to him”}}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 5
\verseWithHeading{More Laws Regarding Sin Offerings}{“ ‘When a person\lebnote{Or “a soul”}sins in that\lebnote{Or “and”}he hears the utterance of a curse and he is a witness or he sees or he knows, if he does not make it known, then\lebnote{Or “and”}he shall bear his guilt.}%
\verse{Or if a person\lebnote{Or “a soul”}touches anything unclean, \textit{whether}\lebnote{Literally “or”}an unclean wild\lebnote{Implied by the following phrase specifying domestic animals}animal’s dead body or an unclean domestic animal’s dead body or an unclean swarmer’s dead body, but\lebnote{Or “and”}\textit{he is unaware of it},\lebnote{Literally “it is concealed from him”}he is unclean and he is guilty.}%
\verse{Or when he touches human uncleanness, \textit{namely}\lebnote{Literally “for” or “to” (see HALOT 510)}any uncleanness of his by which he might become unclean, but\lebnote{Or “and”}\textit{he is unaware of it},\lebnote{Literally “it is concealed from him”}and he himself finds out, then\lebnote{Or “and”}he will be guilty.}%
\verse{Or when a person\lebnote{Or “a soul”}swears, speaking thoughtlessly with his lips, to do evil or to do good \textit{with regard to}\lebnote{Literally “for” or “to” (see HALOT 510)}anything that \textit{the person}\lebnote{Literally “the man”}in a sworn oath speaks thoughtlessly, but\lebnote{Or “and”}\textit{he is unaware of it},\lebnote{Literally “it is concealed from him”}he will be guilty \textit{in any of}\lebnote{Literally “for one of”}these.}%
\verse{When he becomes guilty \textit{in any of}\lebnote{Literally “for one of”}these, he shall confess\lebnote{Or “and he shall confess”}what he has sinned \textit{regarding}\lebnote{Literally “upon” or “against”}it,}%
\verse{and he shall bring his guilt offering to\lebnote{Or “for”}Adonai for his sin that he has \textit{committed}:\lebnote{Literally “sinned”}a female from the flock,\lebnote{The Hebrew term refers collectively to both sheep and goats (small livestock animals)}a ewe-lamb or \textit{a she-goat},\lebnote{Literally “a she-goat of goats”}as a sin offering, and the priest shall make atonement for him for\lebnote{Hebrew “from”}his sin.}%
\verse{“ ‘If\lebnote{Or “And if”}\textit{he cannot afford a sheep},\lebnote{Literally “his hand does not touch enough of small livestock”}he shall bring as his guilt offering for what he sinned two turtledoves or two \textit{young doves}\lebnote{Literally “sons of dove” or “children of dove”}for Adonai, one for a sin offering and one for a burnt offering.}%
\verse{He shall bring\lebnote{Or “And he shall bring”}them to the priest, and he shall present that which is for the sin offering first, and the priest\lebnote{Required by the previous action (see NET)}shall wring its head off \textit{at the back of its neck},\lebnote{Literally “from opposite its neck”}but\lebnote{Or “and”}he must not sever it,}%
\verse{and he shall spatter \textit{some of}\lebnote{Literally “from”}the sin offering’s blood on the altar’s side, and the leftover blood must be drained out on the altar’s base; it is a sin offering.}%
\verse{The second\lebnote{Or “And the second”}bird he must prepare as a burnt offering according to the regulation, and the priest shall make atonement for him for his sin that he has \textit{committed},\lebnote{Literally “sinned”}and he shall be forgiven.}%
\verse{“ ‘But\lebnote{Or “And”}if \textit{he cannot afford}\lebnote{Literally “his hand cannot produce for”}two turtledoves or two \textit{young doves},\lebnote{Literally “sons of dove” or “children of dove”}then,\lebnote{Or “and”}because he has sinned, he shall bring as his offering a tenth of an\lebnote{Hebrew “the”}ephah of finely milled flour as a sin offering. He must not put\lebnote{Or “And he must not put”}oil on it, nor should he put frankincense on it, because it is a sin offering.}%
\verse{He shall bring\lebnote{Or “And he shall bring”}it to the priest, and the priest \textit{shall take a handful of it}\lebnote{Literally “shall scoop up from it the fullness of his handful”}for its token portion, and he shall turn it to smoke on the altar \textit{in addition to}\lebnote{Literally “on” or “upon”}the offerings made by fire to Adonai; it is a sin offering.}%
\verse{Thus\lebnote{Or “And”}the priest shall make atonement for him because of the sin that he has \textit{committed}\lebnote{Literally “sinned”}\textit{in any of these},\lebnote{Literally “from one from these”}and he shall be forgiven. It shall be\lebnote{Or “And it shall be”}for the priest, like the grain offering.’ ”}%
\verseWithHeading{Laws for Guilt Offerings}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“When a person\lebnote{Or “a soul”}\textit{displays infidelity}\lebnote{Literally “acts unfaithfully infidelity” or “is unfaithful in unfaithfulness”}and he sins in an unintentional wrong \textit{in any of}\lebnote{Literally “from”}Adonai’s holy things, then\lebnote{Or “and”}he shall bring his guilt offering to\lebnote{Or “for”}Adonai: a ram without defect from the flock\lebnote{The Hebrew term refers collectively to both sheep and goats (small livestock animals)}as a guilt offering by your valuation in silver shekels\lebnote{Or “convertible into silver shekels” (NET, NRSV, TNK), “of the proper value in silver” (NIV; cp. ESV “valued in silver shekels,” CSB), or “or you may buy one of equal value with silver” (NLT)}according to the sanctuary shekel.}%
\verse{And he shall make restitution for what he sinned \textit{because of}\lebnote{Literally “from” (see HALOT 598)}a holy thing and shall add one-fifth of \textit{its value}\lebnote{Literally “it”}onto it and shall give it to the priest. The priest\lebnote{Or “And the priest”}shall make atonement for him with the ram of the guilt offering, and he will be forgiven.}%
\verse{“If\lebnote{Or “And if”}a person\lebnote{Or “a soul”}when he sins \textit{violates}\lebnote{Literally “and she/it does”}one from all of Adonai’s commands that should not \textit{be violated},\lebnote{Literally “they are to be done”}but\lebnote{Or “and”}he did not know, then\lebnote{Or “and”}he is guilty and he shall bear his guilt.}%
\verse{He shall bring\lebnote{Or “And he shall bring”}to the priest a ram without defect from the flock\lebnote{The Hebrew term refers collectively to both sheep and goats (small livestock animals)}as a guilt offering by your valuation, and the priest shall make atonement for him because of his unintentional wrong (although\lebnote{Or “and”}he himself did not know), and he will be forgiven.}%
\verse{It is a guilt offering; he certainly was guilty before Adonai.”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 6
\verseWithHeading{Additional Laws for Burnt Offerings}{\lebnote{Leviticus 6:1–30 in the English Bible is 5:20–6:23 in the Hebrew Bible} Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“When a person\lebnote{Or “a soul”}\textit{displays infidelity}\lebnote{Literally “acts unfaithfully infidelity” or “is unfaithful in unfaithfulness”}against Adonai and he deceives his fellow citizen regarding\lebnote{Literally “in”}something entrusted or \textit{a pledge}\lebnote{Literally “a pledge of a hand”}or stealing or he exploits his fellow citizen,}%
\verse{or he finds lost property and lies about it and swears \textit{falsely}\lebnote{Literally “in accordance with deception”}\textit{in regard to}\lebnote{Literally “on” or “upon”}any one of these things by which \textit{a person}\lebnote{Literally “the man”}might commit sin,}%
\verse{\textit{and when}\lebnote{Literally “and it will be when”}he sins and is guilty, then\lebnote{Or “and”}he shall bring back \textit{the things he had stolen}\lebnote{Literally “the robbed things that he had stolen”}or \textit{what he had extorted}\lebnote{Literally “the extortion that he had extorted”}or \textit{something with which he had been entrusted}\lebnote{Literally “something entrusted that had been entrusted to him”}or the lost property that he had found,}%
\verse{or \textit{regarding}\lebnote{Literally “from”}anything about which he has sworn \textit{falsely},\lebnote{Literally “in accordance with deception”}then\lebnote{Or “and”}he shall repay it \textit{according to}\lebnote{Literally “in”}its value and shall add one-fifth of its value to it—he must give it \textit{to whom it belongs}\lebnote{Literally “to whom it is to him”}on the day of his guilt offering.}%
\verse{And he must bring as his guilt offering to Adonai a ram without defect from the flock\lebnote{The Hebrew term refers collectively to both sheep and goats (small livestock animals)}by your valuation\lebnote{See 5:15 and 18}as a guilt offering to the priest,}%
\verse{and the priest shall make atonement for him \textit{before}\lebnote{Literally “to the faces of”}Adonai, and he shall be forgiven \textit{anything}\lebnote{Literally “one”}from all that he might do \textit{by which he might incur guilt}.”\lebnote{Literally “which he might do for incurring guilt by it”}}%
\verse{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Command Aaron and his sons, saying, ‘This is the regulation of the burnt offering: \textit{The burnt offering must remain on the hearth}\lebnote{Literally “It shall be the burnt offering on a hearth”}on the altar all night until the morning, and the altar’s fire must be kept burning on it.}%
\verse{And the priest shall put on his linen robe, and he must put his linen undergarments on his body, and he shall take away the fatty ashes of the burnt offering that the fire has consumed on the altar, and he shall place them\lebnote{Hebrew “him/it”—plural required by the English “fatty ashes”}beside the altar.}%
\verse{And he shall take off his garments and put on other garments, and he shall bring out the fatty ashes \textit{outside the camp}\lebnote{Literally “to from an outside place of the camp”}to a ceremonially clean place,}%
\verse{but\lebnote{Or “and”}the fire on the altar must be kept burning on it; it must not be quenched. And the priest must burn wood \textit{every morning}\lebnote{Literally “in the morning in the morning”}on it,\lebnote{Antecedent for this 3fs suffix is “fire” (“altar” is ms)}and he shall arrange the burnt offering on it,\lebnote{Antecedent for this 3fs suffix is “fire” (“altar” is ms)}and he shall turn into smoke the fat portions of the fellowship offerings on it.\lebnote{Antecedent for this 3fs suffix is “fire” (“altar” is ms)}}%
\verse{A perpetual fire must be kept burning on the altar; it must not be quenched.}%
\verseWithHeading{Additional Laws for Grain Offerings}{“ ‘And this is the regulation of the grain offering. Aaron’s sons shall present it \textit{before}\lebnote{Literally “to the faces of”}Adonai \textit{in front of}\lebnote{Literally “to the faces of”}the altar,}%
\verse{and he\lebnote{That is, the priest; understood by context and 3ms verb}in his fist shall take away from it \textit{some of}\lebnote{Literally “from”}the grain offering’s finely milled flour, and \textit{some of}\lebnote{Literally “from”}its oil and all of the frankincense that is on the grain offering, and he shall turn into smoke its token portion on the altar as an appeasing fragrance to Adonai.}%
\verse{And Aaron and his sons must eat the remainder of it; they must eat it as unleavened bread in a holy place—in the tent of assembly’s courtyard they must eat it.}%
\verse{It must not be baked with yeast. I have given it as their share from my offerings made by fire. It is \textit{a most holy thing},\lebnote{Literally “a holiness of holinesses”}like the sin offering and like the guilt offering.}%
\verse{Every male among Aaron’s sons may eat it as a lasting rule among your generations from the offerings made by fire \textit{belonging to}\lebnote{Literally “of”}Adonai. Anything that\lebnote{Or “Everyone who”}touches them will become holy.’ ”}%
\verse{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“This is the offering of Aaron and his sons that they shall present to Adonai on the day of his being anointed: a tenth of an\lebnote{Hebrew “the”}ephah of finely milled flour as a perpetual grain offering, half of it in the morning and half of it in the evening.}%
\verse{It must be made\lebnote{Or “prepared”}in\lebnote{Or “with”}oil on a flat baking pan; you\lebnote{Singular masculine}must bring it well-mixed; you must present pieces of a grain offering’s baked goods\lebnote{Or “broken bits” (JPS, NET, NIV)}as an appeasing fragrance to Adonai.}%
\verse{And the anointed priest taking his place from among his sons must do it. As a lasting rule, it must be turned into smoke totally for Adonai.}%
\verse{And every grain offering of a priest must be a whole burnt offering; it must not be eaten.”}%
\verseWithHeading{Additional Laws for Sin Offerings}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to Aaron and his sons, saying, ‘This is the regulation of the sin offering: In the place where the sin offering is slaughtered, the sin offering must be slaughtered \textit{before}\lebnote{Literally “to the faces of”}Adonai; it is \textit{a most holy thing}.\lebnote{Literally “a holiness of holinesses”}}%
\verse{The priest \textit{who offers the sin offering}\lebnote{Literally “one who offers it a sin offering”}must eat it in a holy place—in the tent of assembly’s courtyard.}%
\verse{Anything that\lebnote{Or “Everyone who”}touches its flesh will become holy, and when \textit{some of}\lebnote{Literally “from”}its blood spatters on a\lebnote{Hebrew “the”}garment, what was spattered on it you\lebnote{Singular masculine}shall wash in a holy place.}%
\verse{And a clay vessel in which it was boiled must be broken, but\lebnote{Or “and”}if it was boiled in a bronze vessel, then\lebnote{Or “and”}it shall be thoroughly scoured and rinsed with water.}%
\verse{Any male among the priests may eat it; it is \textit{a most holy thing}.\lebnote{Literally “a holiness of holinesses”}}%
\verse{But\lebnote{Or “And”}any sin offering from which \textit{some of}\lebnote{Literally “from”}its blood is brought to the tent of assembly to make atonement in the sanctuary must not be eaten; it must be burned in the fire.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 7
\verseWithHeading{Additional Laws for Guilt Offerings}{“ ‘And this is the regulation of the guilt offering; it is \textit{a most holy thing}.\lebnote{Literally “a holiness of holinesses”}}%
\verse{In the place where they slaughter the burnt offering,\lebnote{Or, taking the verb as an indefinite (thus passive) imperfect 3mp, “the sin offering is slaughtered”}they must slaughter the guilt offering,\lebnote{Or, taking the verb as an indefinite (thus passive) imperfect 3mp, “the guilt offering must be slaughtered”}and he\lebnote{That is, the priest; understood by context and 3ms verb}must sprinkle its blood upon the altar all around.}%
\verse{And he must present all of its fat:\lebnote{Hebrew “all of its fat from it”}the fat tail and the fat that covers the inner parts,\lebnote{Or “entrails”}}%
\verse{and the two kidneys, and the fat that is on them, which is on the loins, and he must remove the lobe on the liver in addition to the kidneys.}%
\verse{And the priest shall turn it into smoke it on the altar as a food offering made by fire for Adonai; it is a guilt offering.}%
\verse{Every male among the priests may eat it; it must be eaten in a holy place; it is \textit{a most holy thing}.\lebnote{Literally “a holiness of holinesses”}}%
\verse{The instruction is \textit{the same for the guilt offering as for the sin offering};\lebnote{Literally “as the sin offering as the guilt offering one for them”}\textit{it belongs to}\lebnote{Literally “for him it shall be” or “it will become his”}the priest, who makes atonement with it.}%
\verseWithHeading{Portions for the Priests}{“ ‘And\lebnote{Or “As for”}the priest who presents \textit{a person’s}\lebnote{Literally “of a man”}burnt offering, to that\lebnote{Hebrew “the”}priest \textit{belongs}\lebnote{Literally “to/for him it shall be”}the skin of the burnt offering that he presented.}%
\verse{And every grain offering that is baked in the oven and all that is prepared in a\lebnote{Hebrew “the”}cooking pan or\lebnote{Or “and”}on a flat baking pan \textit{belongs to}\lebnote{Literally “for him it shall be” or “it will become his”}the priest who presented it.}%
\verse{And every grain offering, whether mixed with oil or dry, shall be for all of Aaron’s sons \textit{equally}.\lebnote{Literally “each as his brother”}}%
\verseWithHeading{Additional Laws for Fellowship Offerings}{“ ‘And this is the regulation of the fellowship offerings that he must present to Adonai:}%
\verse{If he presents it for thanksgiving, in addition to the thanksgiving sacrifice he shall present ring-shaped unleavened bread mixed with oil and unleavened bread wafers smeared with oil and well-mixed ring-shaped bread cakes of finely milled flour mixed with oil.}%
\verse{In addition to ring-shaped cakes of bread with yeast, he must present his grain\lebnote{Implied by v. 12}offering together with\lebnote{Or “in addition to”}his sacrifice of thanksgiving peace offerings.}%
\verse{And he shall present one of each kind of grain\lebnote{Implied by v. 12}offering as a contribution for Adonai; \textit{it belongs to}\lebnote{Literally “for him it shall be” or “it will become his”}the priest who sprinkles the fellowship offerings’ blood.}%
\verse{And the meat of the sacrifice of his thanksgiving fellowship offerings must be eaten on the day of his offering; he must not leave it until morning.}%
\verse{“ ‘But\lebnote{Or “And”}if his sacrifice is for a vow or as a freewill offering, it must be eaten on the day of his presenting his sacrifice, and on the next day the remainder\lebnote{Or “and the remainder”}from it may be eaten,}%
\verse{but\lebnote{Or “and”}the remainder from the sacrifice’s meat must be burned up in the fire on the third day.}%
\verse{And if indeed some of the meat of his fellowship offerings’ sacrifice is eaten on the third day, it will not be accepted; it will not be considered of benefit for the one who presented it—it shall be unclean meat, and the person\lebnote{Or “soul”}who eats it shall bear his guilt.}%
\verse{And the meat that touches anything unclean must not be eaten; it must be burned with fire, and as for the clean\lebnote{Understood by context}meat, anyone who is clean may eat the meat.}%
\verse{And the person\lebnote{Or “the soul”}who eats meat from the fellowship offerings’ sacrifice, which is for Adonai, and whose uncleanness is upon him—that person\lebnote{Or “the soul”}shall be cut off from his people.}%
\verse{And when a person\lebnote{Or “a soul”}touches anything unclean, whether human uncleanness or an unclean animal or any unclean detestable thing, and he eats from the meat of the fellowship offerings’ sacrifice, which is for Adonai, then\lebnote{Or “and”}that person\lebnote{Or “the soul”}shall be cut off from his people.’ ”}%
\verseWithHeading{Instructions for the People}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}saying, ‘You\lebnote{Plural}must not eat any fat of ox, or\lebnote{Or “and”}sheep, or\lebnote{Or “and”}goat;}%
\verse{and a dead body’s fat or\lebnote{Or “and”}mangled carcass’s fat may be used for any \textit{purpose},\lebnote{Literally “work”}but\lebnote{Or “and”}you\lebnote{Plural}certainly must not eat it.}%
\verse{When anyone eats fat from the domestic animal from which he presented an offering made by fire for Adonai, then\lebnote{Or “and”}that person\lebnote{Or “the soul”}who ate shall be cut off from his people.}%
\verse{And in any of your\lebnote{Plural}dwellings, you must not eat any blood \textit{belonging to}\lebnote{Literally “to/of”}birds\lebnote{Hebrew “the bird”; generic article with a collective noun}or\lebnote{Or “and”}domestic animals.\lebnote{Hebrew “the domestic animal”; generic article with a collective noun}}%
\verse{Any person\lebnote{Or “any soul” or “all soul(s)”}who eats any blood, that person\lebnote{Or “and that soul”}shall be cut off from his people.’ ”}%
\verseWithHeading{Portions of Fellowship Offerings for Priests}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}saying, ‘The one who presents his fellowship offerings’ sacrifice for Adonai shall bring his offering to Adonai from his fellowship offerings’ sacrifice.}%
\verse{His own hands must bring Adonai’s offerings made by fire. He must bring the fat in addition to the breast section to wave the breast section as a wave offering before Adonai,}%
\verse{and the priest shall turn the fat into smoke on the altar, and the breast section shall be for Aaron and his\lebnote{Hebrew “for his”}sons.}%
\verse{And the right upper thigh you\lebnote{Plural}must give as a contribution for the priest from your\lebnote{Plural}fellowship offerings’ sacrifice.}%
\verse{As for the one from Aaron’s sons who presents the blood of the fellowship offerings and the fat, the right upper thigh \textit{shall belong to him}\lebnote{Literally “for him it shall be”}as his share,}%
\verse{because I have taken the wave offering’s breast section and the contribution offering’s upper thigh from the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}out of their fellowship offerings’ sacrifices, and I have given them to Aaron the priest and his\lebnote{Hebrew “to his”}sons from the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}as a lasting rule.’ ”}%
\verse{This is Aaron’s allotted portion and his sons’ allotted portion from Adonai’s offerings made by fire \textit{when}\lebnote{Literally “in a day”}he brought them forward to serve as priests for Adonai.}%
\verse{This is what Adonai commanded to give them from the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}on the day of his anointing them; it is a lasting statute for their generations.}%
\verseWithHeading{Concluding Summary Concerning the Offerings}{This is the regulation for the burnt offering, for the grain offering and for the sin offering and for the guilt offering and for the consecration offering and for the fellowship offerings’ sacrifice,}%
\verse{which Adonai commanded Moses on \textit{Mount Sinai}\lebnote{Literally “the mountain of Sinai”}on the day of his commanding the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}to present their offerings to Adonai in the desert of Sinai.}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 8
\verseWithHeading{Installing the Priests}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Take Aaron and his sons with him, and the garments and the anointing oil and the bull of the sin offering and the two rams and the basket of the unleavened bread,}%
\verse{and summon all of the community to the entrance to\lebnote{Or “of”}the tent of assembly.”}%
\verse{So\lebnote{Or “And”}Moses did just as Adonai commanded him, and the community gathered by the entrance to\lebnote{Or “of”}the tent of assembly.}%
\verse{Then\lebnote{Or “And”}Moses said to the community, “This is the word that Adonai has commanded to be done.”\lebnote{Or “to do” or “me to do”}}%
\verse{So\lebnote{Or “And”}Moses brought Aaron and his sons near, and he washed them with water.}%
\verse{Then\lebnote{Or “And”}he put the tunic on him and tied\lebnote{Or “he tied”}the sash around him; then\lebnote{Or “and”}he clothed him with the robe and put\lebnote{Or “he put”}the ephod on him; then\lebnote{Or “and”}he tied the ephod’s waistband around him and fastened\lebnote{Or “he fastened”}the ephod\lebnote{Understood by context}to him with it.}%
\verse{Then\lebnote{Or “And”}he placed the breastpiece on him and put\lebnote{Or “he put”}the Urim and the Thummim into the breastpiece;}%
\verse{and he placed the turban on his head, and on \textit{the front of}\lebnote{Literally “to the front of his faces”}the turban he placed the gold rosette, the holy diadem, just as Adonai had commanded Moses.}%
\verse{Then\lebnote{Or “And”}Moses took the anointing oil and anointed\lebnote{Or “he anointed”}the tabernacle and all that was in it, and he consecrated them.}%
\verse{And he spattered \textit{part of}\lebnote{Literally “from”}it on the altar seven times—thus\lebnote{Or “and”}he anointed the altar and all of its utensils, and the basin and its stand, to consecrate them.}%
\verse{Then\lebnote{Or “And”}he poured out \textit{part of}\lebnote{Literally “from”}the anointing oil on Aaron’s head—thus\lebnote{Or “and”}he anointed him in order to consecrate him.}%
\verse{Then\lebnote{Or “And”}Moses brought Aaron’s sons near and clothed\lebnote{Or “he clothed”}them with tunics and tied\lebnote{Or “he tied”}a sash around \textit{each one},\lebnote{Literally “them”}and he bound headbands on\lebnote{Or “to”}them, just as Adonai had commanded Moses.}%
\verseWithHeading{The Offerings for Consecration}{Then\lebnote{Or “And”}he brought forth the bull of the sin offering, and Aaron and his sons placed their hands on the head of the bull of the sin offering,}%
\verse{and he slaughtered it, and Moses took the blood and put\lebnote{Or “he put”}it with his finger on the altar’s horns all around and purified the altar; then\lebnote{Or “and”}he poured the blood out on the altar’s base—thus\lebnote{Or “and”}he consecrated it in order to make atonement for it.\lebnote{Or “to make atonement upon it” (see NET, Tanakh)}}%
\verse{Then\lebnote{Or “And”}he took all the fat that was on the inner parts\lebnote{Or “entrails”}and the lobe on the liver and the two kidneys and their fat, and Moses turned them into smoke on the altar,}%
\verse{but\lebnote{Or “and”}he burned the bull and its skin and its meat and its offal in the fire \textit{outside the camp},\lebnote{Literally “from an outside place of the camp”}just as Adonai had commanded Moses.}%
\verse{Then\lebnote{Or “And”}he brought the ram of the burnt offering near, and Aaron and his sons placed their hands on the ram’s head,}%
\verse{and he slaughtered it. Then\lebnote{Or “And”}Moses sprinkled the blood on the altar all around.}%
\verse{Then\lebnote{Or “And”}he cut the ram into pieces, and Moses turned into smoke the head and the pieces and the suet,}%
\verse{but\lebnote{Or “and”}he washed the inner parts\lebnote{Or “entrails”}and the lower leg bones with water, and Moses turned into smoke all of the ram on the altar; it was a burnt offering as an appeasing fragrance, an offering made by fire for\lebnote{Or “to”}Adonai, just as Adonai had commanded Moses.}%
\verse{Then\lebnote{Or “And”}he brought the second ram near, the ram of the consecration, and Aaron and his sons placed their hands on the ram’s head,}%
\verse{and he slaughtered it. Then\lebnote{Or “And”}Moses took \textit{some of}\lebnote{Literally “from”}its blood and put\lebnote{Or “he put”}it on Aaron’s right ear lobe and on his right hand’s thumb and on his right foot’s big toe.}%
\verse{Then\lebnote{Or “And”}he brought Aaron’s sons near, and Moses put \textit{some of}\lebnote{Literally “from”}the blood on their right ear lobe and on their right hand’s thumb and on their right foot’s big toe, and Moses sprinkled the blood on the altar all around.}%
\verse{Then\lebnote{Or “And”}he took the fat and the fat tail and all of the fat that was on the inner parts\lebnote{Or “entrails”}and the lobe of the liver and the two kidneys and their fat and the right upper thigh;}%
\verse{and from the basket of the unleavened bread that was before Adonai he took one ring-shaped unleavened bread and one ring-shaped bread with\lebnote{Or “of”}oil and one wafer, and he placed them on the \textit{fat parts}\lebnote{Literally “fats”}and on the right upper thigh.}%
\verse{Then\lebnote{Or “And”}he put \textit{all of these}\lebnote{Literally “the all”}on Aaron’s palms and on his sons’ palms, and he waved them as a wave offering \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{Then\lebnote{Or “And”}Moses took them from upon their palms, and he turned them into smoke upon the burnt offering on the altar; they were a consecration offering as an appeasing fragrance—it was an offering made by fire for\lebnote{Or “to”}Adonai.}%
\verse{Then\lebnote{Or “And”}Moses took the breast section, and he waved it as a wave offering \textit{before}\lebnote{Literally “to the faces of”}Adonai from the ram of the consecration offering; it was Moses’ share, just as Adonai had commanded Moses.}%
\verseWithHeading{Anointing the Priests and Their Garments}{Then\lebnote{Or “And”}Moses took \textit{some of}\lebnote{Literally “from”}the anointing oil and \textit{some of}\lebnote{Literally “from”}the blood that was on the altar, and he spattered them on Aaron, on his garments, and on Aaron’s sons and on his sons’ garments with him—thus\lebnote{Or “and”}he consecrated Aaron, his garments, and his sons and his sons’ garments with him.}%
\verse{Then\lebnote{Or “And”}Moses said to Aaron and to his sons, “Boil the meat in the entrance to\lebnote{Or “of”}the tent of assembly, and there you must eat it and the bread that is in the basket of the consecration offering, just as I have commanded, saying, ‘Aaron and his sons must eat it,’}%
\verse{but\lebnote{Or “and”}the remainder of the meat and the bread you\lebnote{Plural}must burn in the fire.}%
\verse{And you must not go out from the entrance to\lebnote{Or “of”}the tent of assembly for seven days, until the day of fulfilling the days of your\lebnote{Plural}consecration, because \textit{it will take seven days to ordain you}.\lebnote{Literally “seven of days it will fill your (plural) hand”}}%
\verse{Just as was done\lebnote{Or “he did”}on this day, Adonai commanded to be done\lebnote{Or “to do”}in order to make atonement for you.\lebnote{Plural}}%
\verse{And you\lebnote{Plural}must stay at the entrance to\lebnote{Or “of”}the tent of assembly day and night for seven days, and you\lebnote{Plural}shall keep the obligation from\lebnote{Or “of”}Adonai, so you\lebnote{Plural}might not die, for thus I have been commanded.”}%
\verse{So\lebnote{Or “And”}Aaron and his sons did all the things that Adonai had commanded \textit{through}\lebnote{Literally “by the hand of”}Moses.}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 9
\verseWithHeading{Worship at the Tent of Assembly}{Then\lebnote{Or “And it happened”}on the eighth day Moses summoned Aaron and his sons and Israel’s elders,}%
\verse{and he said to Aaron, “Take for yourself \textit{a bull calf}\lebnote{Literally “a bull-calf a son of cattle” or “a bull-calf a son of the herd”}as a sin offering and a ram as a burnt offering, without defect,\lebnote{An adjective in masculine plural to modify both animals; see NET}and present them \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{Then\lebnote{Or “And”}you\lebnote{Singular}must speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}saying, ‘Take \textit{a he-goat}\lebnote{Literally “a he-goat of goats”}as a sin offering and a bull calf and a male sheep, \textit{yearlings}\lebnote{Literally “sons of a year”}without defect,\lebnote{An adjective in masculine plural to modify both animals; see NET}as a burnt offering,}%
\verse{and an ox and a ram as fellowship offerings to sacrifice \textit{before}\lebnote{Literally “to the faces of”}Adonai, and a grain offering mixed with oil, because today Adonai will appear to you.”\lebnote{Plural}}%
\verse{So\lebnote{Or “And”}they took what Moses had commanded to the \textit{front of}\lebnote{Literally “to the faces of”}the tent of assembly, and the whole community presented themselves, and they stood \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{Then\lebnote{Or “And”}Moses said, “This is the word that Adonai commanded you\lebnote{Plural}to do \textit{so that}\lebnote{Or “and”}the glory of Adonai might appear to you.”\lebnote{Plural}}%
\verse{Then\lebnote{Or “And”}Moses said to Aaron, “Approach\lebnote{Or “Draw near unto”}the altar and \textit{sacrifice}\lebnote{Literally “do” or “make”}your sin offering and your burnt offering, and make atonement for yourself and for the people. And \textit{sacrifice}\lebnote{Literally “do” or “make”}the people’s offering and make atonement for them, just as Adonai has commanded.”}%
\verse{Then\lebnote{Or “And”}Aaron approached\lebnote{Or “drew near unto”}the altar, and he slaughtered the bull calf of the sin offering, which was for himself.}%
\verse{Then\lebnote{Or “And”}Aaron’s sons presented the blood to him, and he dipped his finger in the blood, and he put it on the altar’s horns, and he poured out the blood on the altar’s base.}%
\verse{And the fat and the kidneys and the lobe \textit{from the sin offering’s liver}\lebnote{Literally “from the liver from the sin offering”}he turned into smoke on the altar, just as Adonai had commanded Moses,}%
\verse{but\lebnote{Or “and”}the meat and the skin he burned \textit{with fire}\lebnote{Literally “in the fire”}\textit{outside the camp}.\lebnote{Literally “from an outside place to/of the camp”}}%
\verse{Then\lebnote{Or “And”}he slaughtered the burnt offering, and Aaron’s sons brought the blood to him, and he sprinkled it on the altar all around;}%
\verse{and they brought the burnt offering to him by its pieces, as well as\lebnote{Or “and”}the head, and he turned them into smoke on the altar;}%
\verse{and he washed the inner parts\lebnote{Or “entrails”}and the lower leg bones, then\lebnote{Or “and”}he turned them into smoke upon the burnt offering on the altar.}%
\verse{Then\lebnote{Or “And”}he presented the people’s offering, and he took the goat of the sin offering, which was for the people, and he slaughtered it and offered\lebnote{Or “he offered”}it as a sin offering like the first one.}%
\verse{Then\lebnote{Or “And”}he presented the burnt offering, and he \textit{sacrificed}\lebnote{Literally “did” or “made”}it according to the regulation.}%
\verse{Then\lebnote{Or “And”}he presented the grain offering, and he filled his palm \textit{with some of}\lebnote{Literally “from”}it, and he turned it into smoke on the altar besides the morning’s burnt offering.}%
\verse{Then\lebnote{Or “And”}he slaughtered the ox and the ram, the fellowship offerings that are for the people, and Aaron’s sons brought the blood to him, and he sprinkled it on the altar all around.}%
\verse{And as for the fat portions from the ox and from the ram (the fat tail and the layer of fat and the kidneys and the lobe of the liver),}%
\verse{they placed the fat portions on the breast sections, and he turned the fat portions into smoke on the altar.}%
\verse{Then\lebnote{Or “And”}Aaron waved the breast sections and the right upper thigh as a wave offering \textit{before}\lebnote{Literally “to the faces of”}Adonai, just as Moses had commanded.}%
\verse{Then\lebnote{Or “And”}Aaron lifted his hand toward\lebnote{Or “to”}the people, and he blessed them, and he came down \textit{after}\lebnote{Literally “from”}\textit{sacrificing}\lebnote{Literally “doing” or “making”}the sin offering and the burnt offering and the fellowship offerings.}%
\verse{Then\lebnote{Or “And”}Moses and Aaron entered the tent of assembly. When\lebnote{Or “And”}they came out, they blessed\lebnote{Or “and they blessed”}the people, and Adonai’s glory appeared to all the people.}%
\verse{Then\lebnote{Or “And”}a fire went out \textit{from before}\lebnote{Literally “from to the faces of”}Adonai, and it consumed the burnt offering and the fat portions on the altar. And all the people saw it, so\lebnote{Or “and”}they shouted for joy, and they fell on their faces.}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 10
\verseWithHeading{The Deaths of Nadab and Abihu}{And Aaron’s sons Nadab and Abihu each took his censer, and they put fire in them and placed incense on it;\lebnote{That is, the fire}then\lebnote{Or “and”}they presented \textit{before}\lebnote{Literally “to the faces of”}Adonai illegitimate fire, which he had not commanded them.}%
\verse{So\lebnote{Or “And”}fire went out \textit{from before}\lebnote{Literally “from to the faces of”}Adonai, and it consumed them so that\lebnote{Or “and”}they died \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{Therefore\lebnote{Or “And”}Moses said to Aaron, “This is what Adonai spoke, saying, ‘Among those who are close to me I will show myself holy, and \textit{in the presence of}\lebnote{Literally “upon the faces of”}all the people I will display my glory.’ ”\lebnote{Or “I will be glorified”}So\lebnote{Or “And”}Aaron was silent.}%
\verse{Then\lebnote{Or “And”}Moses summoned Mishael and Elzaphan the sons of Uzziel, Aaron’s uncle, and he said to them, “Come forward.\lebnote{Or “Come near” or “Approach”}Carry your brothers from \textit{the front of}\lebnote{Literally “the faces of”}the sanctuary to \textit{outside the camp}.”\lebnote{Literally “from an outside place of the camp”}}%
\verse{So\lebnote{Or “And”}they came forward,\lebnote{Or “came near” or “approached”}and they carried them \textit{outside the camp}\lebnote{Literally “to from an outside place of the camp”}in their tunics, just as Moses had ordered.}%
\verse{Then\lebnote{Or “And”}Moses said to Aaron and to his sons Eleazar and Ithamar, “You must not let your hair hang loosely, and you must not tear your garments, so that\lebnote{Or “and”}you will not die and he\lebnote{That is, God}will be angry with all the community. But\lebnote{Or “And”}your brothers, all the house of Israel, may weep because of \textit{the burning that Adonai caused},\lebnote{Literally “the burning that Adonai burned”}}%
\verse{but\lebnote{Or “and”}you must not go out from the entrance to\lebnote{Or “of”}the tent of assembly lest you die, because Adonai’s anointing oil is on you.” So\lebnote{Or “And”}they did according to Moses’ word.}%
\verseWithHeading{Lasting Statutes}{Then\lebnote{Or “And”}Adonai spoke to Aaron, saying,}%
\verse{“You and your sons with you may not drink wine or\lebnote{Or “and”}strong drink when you come to the tent of assembly, so that\lebnote{Or “and”}you will not die—it is a lasting statute for your\lebnote{Plural}generations—}%
\verse{and to distinguish between the holy and the unholy,\lebnote{Or “the common”}as well as\lebnote{Or “and”}between the unclean and the clean,}%
\verse{and to teach the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}all the rules that Adonai has spoken to them \textit{through}\lebnote{Literally “by the hand of”}Moses.”}%
\verse{Then\lebnote{Or “And”}Moses spoke to Aaron and to his sons Eleazar and Ithamar, “As for the remaining parts,\lebnote{Implied by plural form of noun and the immediate context}take the remainder of the grain offering from Adonai’s offerings made by fire and eat it, the unleavened bread, beside the altar, because it is \textit{a most holy thing}.\lebnote{Literally “a holiness of holinesses”}}%
\verse{And you shall eat it in a holy place, because it is your allotted portion and the allotted portion of your sons from Adonai’s offerings made by fire, for so I have been commanded.}%
\verse{And the wave offering’s breast section and the upper thigh of the contribution offering you\lebnote{Plural}must eat in a clean place, you and your sons and your daughters with you, because they are given as your allotted portion and your sons’ allotted portion from the sacrifices of the \textit{Israelites}’\lebnote{Literally “sons/children of Israel”}fellowship offerings.}%
\verse{They must bring the thigh of the contribution offering and the breast section of the wave offering in addition to the offerings made by fire, consisting of the fat portions, to wave as a wave offering \textit{before}\lebnote{Literally “to the faces of”}Adonai; and it will be for you and for your sons with you as a lasting rule, just as Adonai had commanded.”}%
\verseWithHeading{The Problem of the Uneaten Sin Offering}{Then\lebnote{Or “And”}Moses sought all over for the goat of the sin offering and behold, it was burned up. So\lebnote{Or “And”}he was angry with Aaron’s remaining sons Eleazar and Ithamar, saying,}%
\verse{“Why did you not eat the sin offering on the sanctuary’s site, because it is \textit{a most holy thing}?\lebnote{Literally “a holiness of holinesses”}And he gave it to you to remove the community’s guilt, to make atonement for them \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{Look, its blood was not brought \textit{inside the sanctuary}.\lebnote{Or “to the sanctuary inside” or “to the sanctuary within”}Certainly you should have eaten it in the sanctuary, as I commanded.”}%
\verse{So\lebnote{Or “And”}Aaron said to Moses, “Look, today they presented their sin offering and their burnt offering \textit{before}\lebnote{Literally “to the faces of”}Adonai, and things such as these have happened to me, and if I were to eat a sin offering today, would it have been good in Adonai’s eyes?”}%
\verse{When\lebnote{Or “And”}Moses heard, it\lebnote{Or “and it”}was good in his eyes.}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 11
\verseWithHeading{Clean and Unclean Animals}{Then\lebnote{Or “And”}Adonai spoke to Moses and to Aaron, saying to them,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}saying, ‘These are the animals that you may eat from all the animals that are on the land:}%
\verse{Any among the animals that has a divided hoof and has a split cleft in\lebnote{Hebrew “of”}the hoof, such\lebnote{Hebrew “her/it”}you may eat.}%
\verse{However,\lebnote{Or “Only”}these\lebnote{By context; Hebrew “this”}you may not eat from those that chew the cud and from those that have a\lebnote{Hebrew “the”}divided hoof: the camel, because it is a chewer of cud but it does not have a hoof that is divided—it is unclean for you;}%
\verse{and the coney, because it is a chewer of cud but it does not have a hoof that is divided—it is unclean for you;}%
\verse{and the hare, because it is a chewer of cud but it does not have a hoof that is divided—it is unclean for you;}%
\verse{and the pig, because it has a divided hoof and has a split cleft in\lebnote{Hebrew “of”}the hoof but it does not chew cud—it is unclean for you.}%
\verse{You must not eat from their meat, and you must not touch their dead body—they are unclean for you.}%
\verse{“ ‘These\lebnote{By context; Hebrew “This”}you may eat from all that are in the water: any in the water that has a fin and scales, whether in the seas or\lebnote{Or “and”}in the streams—such\lebnote{Hebrew “them”}you may eat.}%
\verse{But any that does not have a fin and scales, whether in the seas or\lebnote{Or “and”}in the streams, \textit{among}\lebnote{Literally “from”}all the water’s swarmers among all the living creatures that are in the water—they are a detestable thing to you.}%
\verse{And they shall be detestable to you; you must not eat from their meat, and you must detest their dead body.}%
\verse{Any that does not have a fin and scales in the water—it is a detestable thing to you.}%
\verse{“ ‘And these you must detest from the birds; they must not be eaten—they are detestable: the eagle and the vulture and the short-toed eagle,}%
\verse{and the red kite and the black kite according to its kind,}%
\verse{every crow according to its kind,}%
\verse{and \textit{the ostrich}\lebnote{Literally “the daughter of the ostrich”}and the short-eared owl and the seagull and the hawk according to its kind,}%
\verse{and the little owl and the cormorant and the great owl,}%
\verse{and the barn owl and the desert owl and the carrion vulture,}%
\verse{and the stork, the heron according to its kind and the hoopoe and the bat.}%
\verse{“ ‘Any \textit{winged insect}\lebnote{Literally “swarmer of the wing”}that walks on all fours is detestable to you.}%
\verse{Only this may you eat from any of \textit{the winged insects}\lebnote{Literally “the swarmer of the wing”}that walk on all fours—that which has jointed legs above its feet for leaping upon the land.}%
\verse{\textit{From these}\lebnote{Literally “These from them”}you may eat the locust according to its kind and the bald locust according to its kind and the cricket according to its kind and the grasshopper according to its kind.}%
\verse{But\lebnote{Or “And”}any other\lebnote{Implied by context}\textit{winged insect}\lebnote{Literally “swarmer of wing”}that has four legs is detestable to you.}%
\verse{And by these you shall become unclean—anyone who touches their dead body shall become unclean until the evening,}%
\verse{and anyone who carries their dead body must wash his garments, and he shall be unclean until the evening.}%
\verse{“ ‘With regard to any animal that has a divided hoof but does not split the hoof, or\lebnote{Or “and”}does not have a cud for chewing\lebnote{So HALOT 830 s.v. 4}—they are unclean for you; anyone who touches them shall become unclean.}%
\verse{And anything that walks upon its paws among any of the animals\lebnote{Collective singular = plural by context}that walks on all fours—they are unclean for you; anyone who touches their dead body shall become unclean until the evening,}%
\verse{and the one who carries their dead body must wash his garments, and he shall be unclean until the evening—they are unclean for you.}%
\verse{“ ‘And these\lebnote{By context; Hebrew “this”}are the unclean for you among the swarmers\lebnote{Collective singular = plural}that swarm on the land: the weasel and the mouse and the thorn-tailed lizard according to its kind,}%
\verse{and the gecko and the land crocodile and the lizard and the sand lizard and the chameleon.}%
\verse{These are the unclean for you among all the swarmers; anyone who touches them at their death shall become unclean until the evening.}%
\verse{And anything on which \textit{one of them}\lebnote{Literally “from them”}falls at their death shall become unclean: any object of wood or garment or skin or sackcloth—any object that has performed work—must be placed in water, and it shall be unclean until the evening, and then it shall be clean.}%
\verse{And any clay vessel\lebnote{Context indicates a vessel as distinguished from a tool or utensil}into which it falls shall become unclean, and you must break it.}%
\verse{Any of the food that could be eaten on which water from such a vessel comes shall become unclean, and any liquid that could be drunk in any such vessel shall become unclean.}%
\verse{And anything on which \textit{one of their dead bodies}\lebnote{Literally “from their dead body”}falls shall become unclean: an oven or\lebnote{Or “and”}a stove must be broken\lebnote{Or “smashed” (NASB, HCSB, NET, NJPS) or “broken in pieces” (ASV, ESV, NRSV)}—they are unclean and shall be unclean for you.}%
\verse{Surely\lebnote{So HALOT 45}a spring or\lebnote{Or “and”}a cistern collecting water shall be clean, but that which touches their dead body shall become unclean.}%
\verse{And when \textit{one of their dead bodies}\lebnote{Literally “from their dead body”}falls on any \textit{seed for sowing},\lebnote{Literally “seed plant that is to be sown”}it is clean.}%
\verse{But\lebnote{Or “And”}when water is put on the seed and \textit{one of their dead bodies}\lebnote{Literally “from their dead body”}falls on it, it is unclean for you.}%
\verse{“ ‘And when \textit{one of the animals}\lebnote{Literally “from the animal”}dies that is for you to eat, the one who touches its dead body shall become unclean until the evening.}%
\verse{And the one who eats \textit{some of}\lebnote{Literally “from”}its dead body must wash his garments, and he shall be unclean until the evening; and the one who carries its dead body must wash his garments, and he shall be unclean until the evening.}%
\verse{“ ‘And any swarmer that swarms on the land is detestable; it must not be eaten.}%
\verse{You must not eat\lebnote{Hebrew “eat them”}anything that moves upon its belly or\lebnote{Or “and”}that walks on all fours, even any with numerous feet belonging to any swarmer that swarms on the land, because they are detestable.}%
\verse{You must not defile yourselves with any swarmer that swarms, and you must not make yourselves unclean by them and so be made unclean by them,}%
\verse{because I am Adonai your God, and you must keep yourselves sanctified, so that\lebnote{Or “and”}you shall be holy, because I am holy. And you must not make yourselves unclean with any swarmer that moves along on the land,}%
\verse{because I am Adonai, who brought you up from the land of Egypt to be for you as God. Thus\lebnote{Or “And”}you shall be holy, because I am holy.}%
\verse{“ ‘This is the regulation of the animals\lebnote{Collective singulars in this verse are plural by context}and the birds and all living creatures that move along in the water and \textit{concerning}\lebnote{Literally “of”}all the creatures that swarm on the land,}%
\verse{to distinguish between the unclean and the clean and between the animal that is to be eaten and the animal that must not be eaten.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 12
\verseWithHeading{Purification After Childbirth}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}saying, ‘When a woman becomes pregnant and she gives birth to a male,\lebnote{Or “son”}then\lebnote{Or “and”}she shall be unclean seven days—as in the time of her menstrual bleeding, she shall become unclean.}%
\verse{And on the eighth day his foreskin’s flesh shall be circumcised.}%
\verse{And for thirty-three days she shall stay in the blood of her cleansing; she must not touch any holy object, and she may not come to the sanctuary until the fulfilling of the days of her cleansing.}%
\verse{But\lebnote{Or “And”}if she gives birth to a female,\lebnote{Or “daughter”}then\lebnote{Or “and”}she shall be unclean for two weeks as in her menstruation, and for sixty-six days she shall stay \textit{through}\lebnote{Or “at” or “on”}the blood of her cleansing.}%
\verse{And at the fulfilling of the days of her cleansing, whether for a son or for a daughter, she must bring to the priest at the tent of assembly’s entrance a \textit{yearling}\lebnote{Literally “a son of his year”}male lamb as a burnt offering and \textit{young dove}\lebnote{Literally “son of a dove”}or a turtledove as a sin offering.}%
\verse{And the priest shall present it \textit{before}\lebnote{Literally “to the faces of”}Adonai, and he shall make atonement for her, so that\lebnote{Or “and”}she shall be clean from the flow of her blood. This is the regulation of childbearing\lebnote{Literally “the childbearing”}for the male\lebnote{Or “son”}or for the female.\lebnote{Or “daughter”}}%
\verse{And if \textit{she cannot afford}\lebnote{Literally “her hand does not find enough”}a sheep,\lebnote{Or “small livestock”}then\lebnote{Or “and”}she shall take two turtledoves or two \textit{young doves}\lebnote{Literally “sons of a dove”}—one as a burnt offering and one as a sin offering—and the priest shall make atonement for her, so that\lebnote{Or “and”}she shall be clean.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 13
\verseWithHeading{Regulations About Defiling Skin Diseases}{Then\lebnote{Or “And”}Adonai spoke to Moses and to Aaron, saying,}%
\verse{“When a \textit{person}\lebnote{Literally “man”}has on his body’s skin a swelling or an epidermal eruption or a spot and it becomes\lebnote{Perfect of הָיָה followed by לְ; see HALOT 244 s.v. 7.c}an infectious skin disease on his body’s skin, then\lebnote{Or “and”}he shall be brought to Aaron the priest or to one of his sons the priests.}%
\verse{And the priest shall examine the infection on his body’s skin, and if the hair in the infection turns white and the appearance of the infection is deeper than his body’s skin, it is an infectious skin disease, and the priest shall examine it, and he shall declare him unclean.}%
\verse{But\lebnote{Or “And”}if a spot is white on his body’s skin and its appearance is not deeper than the skin and its hair does not turn white, then\lebnote{Or “and”}the priest shall confine the afflicted person for seven days.}%
\verse{And the priest shall examine it on the seventh day, and \textit{if},\lebnote{Literally “look” or “behold”}in his eyes, the infection has stayed unchanged, the infection has not spread on the skin, then\lebnote{Or “and”}the priest shall confine him for seven days a second time.}%
\verse{And the priest shall examine him on the seventh day for a second time, and \textit{if}\lebnote{Literally “look” or “behold”}the infection has faded and the infection has not spread on the skin, then\lebnote{Or “and”}the priest shall declare him clean—it is an epidermal eruption; and he shall wash his garments, and so he shall be clean.}%
\verse{But\lebnote{Or “And”}if the epidermal eruption spreads further on the skin after showing himself to the priest for his cleansing, then\lebnote{Or “and”}he shall appear a second time to the priest.}%
\verse{And the priest shall examine it,\lebnote{The direct object is supplied from context in the English translation}and \textit{if}\lebnote{Literally “look” or “behold”}the epidermal eruption has spread on the skin, then\lebnote{Or “and”}the priest shall declare him unclean—it is an infectious skin disease.}%
\verse{“When an infectious skin disease is on a person and he is brought to the priest,}%
\verse{the priest shall examine it,\lebnote{The direct object is supplied from context in the English translation}and \textit{if}\lebnote{Literally “look” or “behold”}a white swelling is on the skin and it turns the hair white and \textit{raw flesh}\lebnote{Literally “living of living flesh”}is in the swelling,}%
\verse{it is a chronic infectious skin disease on his body’s skin, and the priest shall declare him unclean; he shall not confine him, because he is unclean.}%
\verse{And if the infectious skin disease breaks out all over on the skin and the infectious skin disease covers all of the afflicted person’s skin from his head to his feed, \textit{so far as the priest can see},\lebnote{Literally “for all the sight of the eyes of the priest”}}%
\verse{then\lebnote{Or “and”}the priest shall examine it,\lebnote{The direct object is supplied from context in the English translation}and \textit{if}\lebnote{Literally “look” or “behold”}the infectious skin disease covers his whole body, then\lebnote{Or “and”}he shall pronounce the afflicted person clean—all of it has turned white; he is clean.}%
\verse{But\lebnote{Or “And”}\textit{whenever}\lebnote{Literally “on a day”}\textit{raw flesh}\lebnote{Literally “living flesh”}appears on him, he shall become unclean.}%
\verse{And the priest shall examine the \textit{raw flesh},\lebnote{Literally “living flesh”}and he shall pronounce him unclean—the \textit{raw flesh}\lebnote{Literally “living flesh”}is unclean; it is an infectious skin disease.}%
\verse{Or, when the \textit{raw flesh}\lebnote{Literally “living flesh”}returns and it has changed to white, then\lebnote{Or “and”}he shall come to the priest,}%
\verse{and the priest shall examine him, and \textit{if}\lebnote{Literally “look” or “behold”}the infection has changed to white, then\lebnote{Or “and”}the priest shall pronounce the afflicted person clean—he is clean.}%
\verse{“And when someone’s body \textit{has}\lebnote{Literally “becomes on it”}a skin sore on his skin and it is healed}%
\verse{and a white swelling or a \textit{pinkish}\lebnote{Literally “white red”}spot appears in the skin sore’s place, then\lebnote{Or “and”}he shall show himself to the priest.}%
\verse{And the priest shall examine it,\lebnote{The direct object is supplied from context in the English translation}and \textit{if}\lebnote{Literally “look” or “behold”}its appearance is deeper than the skin and its hair has changed to white, then\lebnote{Or “and”}the priest shall declare him unclean—it is an infectious skin disease; it has broken out in the skin sore.}%
\verse{And if the priest examines it and \textit{if}\lebnote{Literally “look” or “behold”}there is no white hair in it and it is not deeper than the skin and it is faded, then\lebnote{Or “and”}the priest shall confine him for seven days.}%
\verse{But\lebnote{Or “And”}if it has spread further on the skin, then\lebnote{Or “and”}the priest shall declare him unclean—it is an infection.}%
\verse{But\lebnote{Or “And”}if the spot has stayed unchanged, it has not spread, it is the skin sore’s scar, so\lebnote{Or “and”}the priest shall declare him clean.}%
\verse{“Or when a body \textit{has}\lebnote{Literally “it becomes”}a burn-spot \textit{from}\lebnote{Literally “of”}fire on its skin and the \textit{raw flesh}\lebnote{Literally “living”}of the burn-spot is \textit{pinkish}\lebnote{Literally “white red”}or white,}%
\verse{then\lebnote{Or “and”}the priest shall examine it, and \textit{if}\lebnote{Literally “look” or “behold”}the hair turns white in the spot and its appearance is deeper than the skin, it is an infectious skin disease—it has broken out in the burn-spot; so\lebnote{Or “and”}the priest shall declare him unclean—it is an infectious skin disease.}%
\verse{But\lebnote{Or “And”}if the priest examines it and \textit{if}\lebnote{Literally “look” or “behold”}there is not white hair in the spot and it is not deeper than the skin and it is faded, then\lebnote{Or “and”}the priest shall confine him for seven days.}%
\verse{And the priest shall examine him on the seventh day; if it has spread further on the skin, then\lebnote{Or “and”}the priest shall declare him unclean—it is an infectious skin disease.}%
\verse{But\lebnote{Or “And”}if it the spot has stayed unchanged in its place, it has not spread on the skin and it is faded, then it is the burn-spot’s swelling, so\lebnote{Or “and”}the priest shall declare him clean, because it is the burn-spot’s scar.}%
\verse{“And when a man or a woman \textit{has}\lebnote{Literally “becomes on him”}an infection on the head or in the beard,}%
\verse{then\lebnote{Or “and”}the priest shall examine the infection, and \textit{if}\lebnote{Literally “look” or “behold”}its appearance is deeper than the skin and in it is thin bright red hair, then\lebnote{Or “and”}the priest shall declare it unclean—it is a diseased area of skin; it is an infectious skin disease of the head or the beard.}%
\verse{But\lebnote{Or “And”}if the priest examines the diseased area of the skin’s infection and \textit{if}\lebnote{Literally “look” or “behold”}its appearance is not deeper than the skin and there is no black hair in it, then\lebnote{Or “and”}the priest shall confine the afflicted person with the diseased area of skin for seven days.}%
\verse{And the priest shall examine the infection on the seventh day, and \textit{if}\lebnote{Literally “look” or “behold”}the diseased area of skin has not spread and it does not have bright red hair in it and the diseased area of the skin’s appearance is not deeper than the skin,}%
\verse{then\lebnote{Or “and”}he shall shave himself, but\lebnote{Or “and”}he shall not shave the diseased area of skin, and the priest shall confine the person with\lebnote{The direct object is supplied from context in the English translation}the diseased area of skin a second time for seven days.}%
\verse{And the priest shall examine the diseased area of skin on the seventh day, and \textit{if}\lebnote{Literally “look” or “behold”}the diseased area has not spread on the skin and its appearance is not deeper than the skin, then\lebnote{Or “and”}the priest shall pronounce him clean, and he shall wash his garments, and he shall be clean.}%
\verse{But\lebnote{Or “And”}if the diseased area of skin has not spread further on the skin after his cleansing,}%
\verse{then\lebnote{Or “and”}the priest shall examine him, and \textit{if}\lebnote{Literally “look” or “behold”}the diseased area of skin has spread on the skin, the priest shall not inspect for bright\lebnote{Hebrew “the bright”}red hair—he is unclean.}%
\verse{But\lebnote{Or “And”}if, in his eyes, the diseased area of skin has stayed unchanged and black hair has grown in it, the diseased area of skin is healed—he is clean, and the priest shall pronounce him clean.}%
\verse{“And when a man or a woman \textit{has}\lebnote{Literally “becomes”}spots on their body’s skin, white spots,}%
\verse{then\lebnote{Or “and”}the priest shall examine them,\lebnote{The direct object is supplied from context in the English translation}and \textit{if}\lebnote{Literally “look” or “behold”}the spots on their body’s skin are a faded white, it is a skin rash; it has broken out on the skin—it is clean.}%
\verse{“And if a man becomes bald, his head is bald, he is clean.}%
\verse{And if he becomes bald \textit{from his forehead},\lebnote{Literally “from the forehead of his face”}his head is bald, he is clean.}%
\verse{But\lebnote{Or “And”}if a \textit{pinkish}\lebnote{Literally “white red”}infection occurs\lebnote{Or “becomes” or “happens”}on the bald spot or on the bald forehead, it is an infectious skin disease that sprouts on his bald spot or on his bald forehead.}%
\verse{So\lebnote{Or “And”}the priest shall examine him, and \textit{if}\lebnote{Literally “look” or “behold”}the infection’s swelling is \textit{pinkish}\lebnote{Literally “white red”}on his bald spot or on his bald forehead, like the appearance of an infectious skin disease on\lebnote{Literally “of”}the body,}%
\verse{he is a man afflicted with a skin disease—he is unclean; the priest certainly shall declare him unclean—his infection is on his head.}%
\verse{“As for\lebnote{Or “And”}the person who is afflicted with a skin disease, his garments must be torn and his \textit{hair}\lebnote{Literally “head”}must be allowed to hang loosely, and he must cover his upper lip, and he must call out, ‘Unclean! Unclean!’}%
\verse{For all the days during which the infection is on him, he shall be unclean; he must live alone; his dwelling must be \textit{outside the camp}.”\lebnote{Literally “from an outside place of the camp”}}%
\verseWithHeading{Regulations About Contaminated Fabrics}{“And when the garment \textit{has}\lebnote{Literally “becomes”}an infectious skin disease\lebnote{Perhaps better translated “mold” rather than “skin disease”}on it, on a wool garment\lebnote{Literally “a garment of wool”}or on a linen garment,\lebnote{Literally “a garment of linen”}}%
\verse{or on woven material or on a linen fabric, or\lebnote{Or “and”}on wool or on leather or on any work of leather,}%
\verse{and if the infection is yellowish green or reddish on the garment or on the leather or on the woven material or on the fabric or on any leather object,\lebnote{Literally “an object of leather”}it is an infectious skin disease\lebnote{Perhaps better translated “mold” rather than “skin disease”}and it shall be shown to the priest.}%
\verse{And the priest shall examine the infection, and he shall confine the infected article for seven days.}%
\verse{And he shall examine the infection on the seventh day; if the infection has spread on the garment or on the woven material or on the fabric or on the leather, for any work for which the leather is used, the infection is a destructive skin disease\lebnote{Perhaps better translated “mold” rather than “skin disease”}—it is unclean.}%
\verse{And he shall burn the garment or the woven material or the fabric, \textit{whether wool or linen},\lebnote{Literally “in/on the wool or in/on the linen”}or any leather object that \textit{has}\lebnote{Literally “becomes on/in it”}the infection, because it is an infectious skin disease,\lebnote{Perhaps better translated “mold” rather than “skin disease”}which is destructive—it must be burned in the fire.}%
\verse{“But\lebnote{Or “And”}if the priest examines it\lebnote{The direct object is supplied from context in the English translation}and \textit{if}\lebnote{Literally “look” or “behold”}the infection has not spread on the garment or on the woven material or on the fabric or on any leather object,}%
\verse{then\lebnote{Or “and”}the priest shall command, and \textit{someone}\lebnote{Literally “he”}shall wash that on which the infection is, and he shall confine it a second time for seven days.}%
\verse{And the priest shall examine it\lebnote{The direct object is supplied from context in the English translation}after the infection has been washed off, and \textit{if}\lebnote{Literally “look” or “behold”}the infection has not changed its outward appearance and the infection has not spread, it is unclean; he must burn it in the fire; it is a fungus on its back or on its front.}%
\verse{But\lebnote{Or “And”}if the priest examines it\lebnote{The direct object is supplied from context in the English translation}and \textit{if}\lebnote{Literally “look” or “behold”}the infection is faded after it has been washed off, then\lebnote{Or “and”}he shall tear it from the garment or from the leather or from the woven material or from the fabric.}%
\verse{And if it appears again on the garment or on the woven material or on the fabric or on any leather object, it is spreading; you\lebnote{Singular}must burn in the fire that which has the infection in it.}%
\verse{And the garment or the woven material or the fabric or any leather object that he might wash and the infection is removed from them then\lebnote{Or “and”}shall be washed a second time, and it shall be clean.”}%
\verse{This is the regulation of the infectious skin disease\lebnote{Perhaps better translated “mold” rather than “skin disease”}in\lebnote{Hebrew “of”}the wool garment or the linen or the woven material or the fabric or any leather object to declare it clean or to declare it unclean.}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 14
\verseWithHeading{Instructions for Cleansing Infectious Skin Diseases}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“This is the regulation of the person afflicted with a skin disease \textit{at the time of}\lebnote{Or “on the day of”}his cleansing. And he shall be brought to the priest,}%
\verse{and the priest shall go \textit{outside the camp},\lebnote{Literally “to from an outside place of the camp”}and the priest shall examine him,\lebnote{The direct object is supplied from context in the English translation}and \textit{if}\lebnote{Literally “look” or “behold”}the skin disease’s infection is healed on\lebnote{Hebrew “from”}the afflicted person,}%
\verse{then\lebnote{Or “and”}the priest shall command, and he shall take two living, clean birds and \textit{cedar wood}\lebnote{Literally “wood of cedar”}and \textit{a crimson thread}\lebnote{Literally “crimson thread of the worm”}and hyssop for the one who presents himself for cleansing.}%
\verse{Then\lebnote{Or “And”}the priest shall command someone\lebnote{The direct object is supplied from context in the English translation}to slaughter one bird over fresh water in a clay vessel.}%
\verse{He must take the living bird and the \textit{cedar wood}\lebnote{Literally “wood of cedar”}and \textit{the crimson thread}\lebnote{Literally “the crimson thread of the worm”}and the hyssop, and he shall dip them and the living bird in the bird’s blood slaughtered over the fresh water.}%
\verse{And he shall spatter the blood\lebnote{The direct object is supplied from context in the English translation}seven times on the one who presents himself for cleansing from the infectious skin disease, and he shall declare him clean, and he shall send the living bird \textit{into the open field}.\lebnote{Literally “toward the faces of the field”}}%
\verse{Then\lebnote{Or “And”}the one who presents himself for cleansing shall wash his garments, and he shall shave off all his hair, and he shall wash himself in the water; thus\lebnote{Or “and”}he shall be clean, and afterward he shall enter the camp, but\lebnote{Or “and”}he shall stay \textit{outside his tent}\lebnote{Literally “from an outside place of his tent”}for seven days.}%
\verse{\textit{And then}\lebnote{Literally “And it shall be”}on the seventh day he must shave off all his hair—he must shave his head and his beard and \textit{his eyebrows}\lebnote{Literally “the rims of his eyes”}and all the rest\lebnote{Implied by context}of his hair—and he shall wash his garments, and he shall wash his body in the water; thus\lebnote{Or “and”}he shall be clean.}%
\verse{“And on the eighth day he must take two male lambs without defect and one ewe-lamb \textit{in its first year}\lebnote{Literally “a daughter of her year”}without defect and three-tenths of an ephah\lebnote{Implied by context}of finely milled flour mixed with oil as a grain offering and one log of oil.}%
\verse{And the priest who cleanses him\lebnote{The direct object is supplied from context in the English translation}shall present the man who presents himself for cleansing and \textit{these things}\lebnote{Literally “with them”}\textit{before}\lebnote{Literally “to the faces of”}the tent of assembly’s entrance.}%
\verse{Then\lebnote{Or “And”}the priest shall take the one male lamb, and he shall present it as a guilt offering, and the log of oil, and he shall wave them as a wave offering \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{And he shall slaughter the male lamb in the place where he slaughters the sin offering and the burnt offering in the sanctuary’s space,\lebnote{Or “place” or “area”}because as the sin offering belongs to the priest, so also the guilt offering—\textit{it is a most holy thing}.\lebnote{Literally “a holiness of holinesses is it”}}%
\verse{And the priest shall take \textit{some of}\lebnote{Literally “from”}the guilt offering’s blood, and the priest shall put it\lebnote{The direct object is supplied from context in the English translation}on the right ear’s lobe of the one who presents himself for cleansing and on his right hand’s thumb and on his right foot’s big toe.}%
\verse{And the priest shall take \textit{some of}\lebnote{Literally “from”}the log of oil, and he shall pour it\lebnote{The direct object is supplied from context in the English translation}on his\lebnote{Hebrew “the priest’s”}left palm;}%
\verse{and the priest shall dip his right finger in the oil that is on his left palm, and he shall spatter \textit{some of}\lebnote{Literally “from”}the oil with his finger seven times \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{Then\lebnote{Or “And”}the priest shall put \textit{some of}\lebnote{Literally “from”}the remaining oil, which is on his palm, on the right ear’s lobe of the one to be cleansed and on his right hand’s thumb and on his right foot’s big toe, on top of the guilt offering’s blood.\lebnote{See v. 14}}%
\verse{And the remaining oil that is on the priest’s palm he shall put on the head of the one who presents himself for cleansing, and the priest shall make atonement for him \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{Thus\lebnote{Or “And”}the priest shall \textit{sacrifice}\lebnote{Literally “do” or “make”}the sin offering, and he shall make atonement for the one who presents himself for cleansing from his uncleanness, and afterward he shall slaughter the burnt offering.}%
\verse{Then\lebnote{Or “And”}the priest shall offer the burnt offering and the grain offering on the altar, and the priest shall make atonement for him, and so he shall be clean.}%
\verse{“But\lebnote{Or “And”}if he is poor and \textit{he cannot afford}\lebnote{Literally “his hand is not producing”}it,\lebnote{The direct object is supplied from context in the English translation}then\lebnote{Or “and”}he shall take one male lamb for a guilt offering as a wave offering to make atonement for himself and one-tenth of an ephah\lebnote{Implied by context}of finely milled flour mixed with oil for a grain offering, and a log of oil,}%
\verse{and two turtledoves or two \textit{young doves}\lebnote{Literally “sons of dove”}that \textit{he can afford},\lebnote{Literally “his hand can produce”}and one shall be a sin offering and the \textit{other}\lebnote{Literally “one”}a burnt offering.}%
\verse{And he shall bring them to the priest at the tent of assembly’s entrance \textit{before}\lebnote{Literally “to the faces of”}Adonai on the eighth day for his cleansing.}%
\verse{And the priest shall take the male lamb for the guilt offering and the log of oil, and the priest shall wave them as a wave offering \textit{before}\lebnote{Literally “to the faces of”}Adonai;}%
\verse{and he shall slaughter the male lamb of the guilt offering, and the priest shall take \textit{some of}\lebnote{Literally “from”}the guilt offering’s blood, and he shall put it\lebnote{The direct object is supplied from context in the English translation}on the right ear’s lobe of the one who presents himself for cleansing and on his right hand’s thumb and on his right foot’s big toe.}%
\verse{Then\lebnote{Or “And”}the priest shall pour out \textit{some of}\lebnote{Literally “from”}the oil on his own\lebnote{Hebrew “the priest’s”}left palm,}%
\verse{and with his right finger the priest shall spatter \textit{some of}\lebnote{Literally “from”}the oil that is on his left palm seven times \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{Then\lebnote{Or “And”}the priest shall put \textit{some of}\lebnote{Literally “from”}the oil that is on his palm on the right ear’s lobe of the one who presents himself for cleansing and on his right hand’s thumb and on his right foot’s big toe on the place of the guilt offering’s blood.}%
\verse{And the remaining oil that is on the priest’s palm he shall put on the head of the one who presents himself for cleansing to make atonement for him \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{And he shall \textit{sacrifice}\lebnote{Literally “do” or “make”}one of the turtledoves or the \textit{young doves}\lebnote{Literally “sons of dove”}that \textit{he can afford},\lebnote{Literally “his hand can produce”}}%
\verse{even\lebnote{Implied by context}what \textit{he can afford},\lebnote{Literally “his hand can produce”}the one as a sin offering and the \textit{other}\lebnote{Literally “one”}as a burnt offering in addition to the grain offering, and the priest shall make atonement for the one who presents himself for cleansing \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{This is the regulation of the one on whom is an infectious skin disease who \textit{cannot afford}\lebnote{Literally “his hand cannot produce”}the cost\lebnote{The direct object is supplied from context in the English translation}for his cleansing.”}%
\verseWithHeading{Instructions for Cleansing Houses}{Then\lebnote{Or “And”}Adonai spoke to Moses and to Aaron, saying,}%
\verse{“When you come into the land of Canaan, which I am about to give to you as your possession, and I put \textit{mildew}\lebnote{Literally “an infection of skin disease”}in a house in the land of your possession,}%
\verse{then\lebnote{Or “and”}the one who \textit{owns the house}\lebnote{Literally “the house is for him”}shall come and tell the priest, saying, ‘It appears to me that an infection is in the house.’}%
\verse{And the priest shall issue a command, and they shall clear out the house before the priest comes to examine the infection, so that\lebnote{Or “and”}all that is in the house might not become unclean; and \textit{afterward}\lebnote{Literally “after thus”}the priest shall go to examine the house.}%
\verse{And he shall examine the infection, and \textit{if}\lebnote{Literally “look” or “behold”}the infection on the house’s wall has yellowish-green or reddish spots and its appearance is deeper than the surface of\lebnote{Implied by the context}the wall,}%
\verse{then\lebnote{Or “and”}the priest shall go out from the house to the house’s entrance, and he shall confine\lebnote{Or “close up”}the house for seven days.}%
\verse{And the priest shall return on the seventh day, and he shall examine the infection,\lebnote{The direct object is supplied from context in the English translation}and \textit{if}\lebnote{Literally “look” or “behold”}the infection has spread on the house’s wall,}%
\verse{the priest shall issue a command and they shall remove the stones on which is the infection, and they shall throw them \textit{outside the city}\lebnote{Literally “to from an outside place of the city”}on an unclean place.}%
\verse{Then\lebnote{Or “And”}they\lebnote{Hebrew “he”}shall scrape off the plaster\lebnote{The direct object is supplied from context in the English translation}from all around the house, and they shall pour out the plaster,\lebnote{See HALOT 862 s.v. 3.b}which they scraped off, \textit{outside the city}\lebnote{Literally “to from an outside place of the city”}on an unclean place.}%
\verse{And they shall take other stones, and they shall \textit{put}\lebnote{Literally “bring”}them in place of those stones, and they\lebnote{Hebrew “he”}shall take other plaster,\lebnote{See HALOT 862 s.v. 3.b}and they\lebnote{Hebrew “he”}shall replaster\lebnote{Implied by the context; Hebrew “plaster”}the house.}%
\verse{“But\lebnote{Or “And”}if the infection should return and it breaks out in the house after they\lebnote{Hebrew “he”}have removed the stones and after scraping off the plaster\lebnote{The direct object is supplied from context in the English translation}of the house and after it has been replastered,\lebnote{Implied by the context; or “plaster”}}%
\verse{then\lebnote{Or “and”}the priest shall come, and he shall examine the infection,\lebnote{The direct object is supplied from context in the English translation}and \textit{if}\lebnote{Literally “look” or “behold”}the infection has spread in the house, it is a destructive \textit{mildew}\lebnote{Literally “infectious disease”}in the house—it is unclean.}%
\verse{So\lebnote{Or “And”}he shall break down the house, its stones and its wood and all of the house’s plaster,\lebnote{See HALOT 862 s.v. 3.b}and he shall bring it all\lebnote{The direct object is supplied from context in the English translation}\textit{outside the city}\lebnote{Literally “to from an outside place of the city”}to an unclean place.}%
\verse{And the person who enters into the house during all the days that he\lebnote{That is, the priest}confined\lebnote{Or “closed up”}it shall become unclean until the evening.}%
\verse{And the person who sleeps in the house must wash his garments, and the person who eats in the house shall wash his garments.}%
\verse{“And if the priest comes again and examines the house\lebnote{The direct object is supplied from context in the English translation}and \textit{if}\lebnote{Literally “look” or “behold”}the infection has not spread in the house after being replastered,\lebnote{Implied by the context; or “plaster”}then\lebnote{Or “and”}the priest shall pronounce the house clean, because the infection is healed.}%
\verse{And he shall take two birds and \textit{cedar wood}\lebnote{Literally “wood of cedar”}and a \textit{crimson thread}\lebnote{Literally “crimson thread of the worm”}and hyssop to cleanse the house;}%
\verse{and he shall slaughter the first bird over fresh water on a clay vessel.}%
\verse{Then\lebnote{Or “And”}he shall take the \textit{cedar wood}\lebnote{Literally “wood of cedar”}and the hyssop and \textit{the crimson thread}\lebnote{Literally “the crimson thread of the worm”}and the living bird, and he shall dip them in the slaughtered bird’s blood and in the fresh water, and he shall spatter them\lebnote{The direct object is supplied from context in the English translation}on the house seven times.}%
\verse{Thus\lebnote{Or “And”}he shall purify the house with the bird’s blood and with the fresh water and with the living bird and with the \textit{cedar wood}\lebnote{Literally “wood of cedar”}and with the hyssop and with \textit{the crimson thread}.\lebnote{Literally “the crimson thread of the worm”}}%
\verse{And he shall send the living bird \textit{outside the city}\lebnote{Literally “to from an outside place of the city”}\textit{into the open field};\lebnote{Literally “to the faces of the field”}and so he shall make atonement for the house, and it shall be clean.}%
\verse{“This is the instruction for any infectious skin disease and for the diseased area of skin,}%
\verse{and for a \textit{mildew}\lebnote{Literally “infectious skin disease”}of the garment and for the house,}%
\verse{and for the swelling and for the epidermal eruption and for the spot,}%
\verse{to teach \textit{when something is unclean and when something is clean}.\lebnote{Literally “on the day of the unclean and on the day of the clean”}This is the regulation of the infectious skin disease.”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 15
\verseWithHeading{Instructions About Bodily Discharges}{Then\lebnote{Or “And”}Adonai spoke to Moses and to Aaron, saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}and you shall say to them, ‘\textit{Any man}\lebnote{Literally “a man a man”}when a fluid discharge occurs\lebnote{Or “becomes” or “shall be”}from his body, his fluid discharge is unclean.}%
\verse{And this becomes his uncleanness in his fluid discharge: whether his body secretes his fluid discharge or his body blocks his fluid discharge, it is his uncleanness.}%
\verse{Any bed upon which the person who discharges lies down becomes unclean, and any object upon which he sits becomes unclean.}%
\verse{And anyone who touches his bed must wash his garments and shall wash\lebnote{Or “bathe”}himself\lebnote{The direct object is supplied from context in the English translation}with water, and he shall be unclean until evening.}%
\verse{And the person who sits on the object upon which the person who discharges has sat must wash his garments, and he shall wash\lebnote{Or “bathe”}himself\lebnote{The direct object is supplied from context in the English translation}with water, and he shall be unclean until the evening.}%
\verse{And the person who touches the body of the person who discharges must wash his garments, and he shall wash\lebnote{Or “bathe”}himself\lebnote{The direct object is supplied from context in the English translation}with water, and he shall be unclean until the evening.}%
\verse{And if the person who discharges spits on one who is\lebnote{The generic article designating a class of persons or things}clean, then\lebnote{Or “and”}that one\lebnote{Hebrew “he”}shall wash is garments, and he shall wash himself\lebnote{The direct object is supplied from context in the English translation}with water, and he shall be unclean until the evening.}%
\verse{And any saddle upon which the person who discharges rides becomes unclean.}%
\verse{And any person who touches anything that happened to be under him becomes unclean until the evening, and the person who carries them must wash his garments, and he shall wash himself\lebnote{The direct object is supplied from context in the English translation}with water, and he shall be unclean until the evening.}%
\verse{And anyone whom the person who discharges might touch \textit{without}\lebnote{Literally “and not”}rinsing off his hands with water shall wash his garments, and he shall wash himself\lebnote{The direct object is supplied from context in the English translation}with water, and he shall be unclean until the evening.}%
\verse{But\lebnote{Or “And”}a clay vessel that the person who discharges touches must be broken, and any \textit{wood object}\lebnote{Literally “object of wood”}must be rinsed with water.}%
\verse{“ ‘And when the person who discharges becomes clean from his body fluid discharge, he shall count\lebnote{Literally “and he shall count”}for himself seven days for his cleansing; then\lebnote{Or “and”}he shall wash his garments, and he shall wash his body with \textit{fresh}\lebnote{Or “running”; literally “living”}water, and he shall be clean.}%
\verse{Then\lebnote{Or “And”}on the eighth day he shall take for himself two turtledoves or two \textit{young doves},\lebnote{Literally “sons of dove”}and he shall come \textit{before}\lebnote{Literally “to the faces of”}Adonai at the tent of assembly’s entrance, and he shall give them to the priest.}%
\verse{And the priest shall \textit{sacrifice}\lebnote{Literally “do” or “make”}one as a sin offering and \textit{the other}\lebnote{Literally “the one”}as a burnt offering, and so the priest shall make atonement for him \textit{before}\lebnote{Literally “to the faces of”}Adonai from his body fluid discharge.}%
\verse{“ ‘And if an emission of semen goes out from anyone, then\lebnote{Or “and”}he shall wash all of his body with water, and he shall be unclean until the evening.}%
\verse{And any garment and anything leather on which is an emission of semen shall be washed with water, and it shall be unclean until the evening.}%
\verse{If\lebnote{Or “And”}there is a woman with whom a man lies down and there is an emission of semen, then\lebnote{Or “and”}they shall wash themselves\lebnote{The direct object is supplied from context in the English translation}with water, and they shall be unclean until the evening.}%
\verse{“ ‘And when a woman \textit{is menstruating},\lebnote{Literally “is discharging blood”}her body fluid discharge occurs in\lebnote{Or “from”}her body; for seven days she shall be in her menstruation, and any person who touches her shall become unclean until the evening.}%
\verse{And anything upon which she lies down during her menstruation shall become unclean, and anything upon which she sits shall become unclean.}%
\verse{And any person who touches her bed must wash his garments, and he shall wash himself\lebnote{The direct object is supplied from context in the English translation}with water, and he shall be unclean until the evening.}%
\verse{And any person who touches any object on which she sat must wash his garments, and he shall wash himself\lebnote{The direct object is supplied from context in the English translation}with water, and he shall be unclean until the evening.}%
\verse{And if it is on the bed or on the object on which she sits, at his touching it he becomes unclean until the evening.}%
\verse{And if a man indeed lies with her and her menstruation occurs on him, then\lebnote{Or “and”}he shall be unclean for seven days, and any bed on which he lies down becomes unclean.}%
\verse{“ ‘And when a woman discharges a body fluid consisting of her blood for many days, but not at the time of her menstruation, or when she discharges in addition to\lebnote{Or “beyond”}her menstruation, all the days of her unclean body fluid discharge she shall become unclean as in the days of her menstruation.}%
\verse{Any bed on which she lies during all the days of her body fluid discharge shall become for her as her bed of menstruation, and any object on which she sits becomes unclean as her menstruation’s uncleanness.}%
\verse{And any person who touches them becomes unclean, and he shall wash his garments, and he shall wash himself\lebnote{The direct object is supplied from context in the English translation}with water, and he shall be unclean until the evening.}%
\verse{“ ‘And if she is clean from her body fluid discharge, then\lebnote{Or “and”}she shall count for herself seven days, and afterward she becomes clean.}%
\verse{And on the eighth day she shall take for herself two turtledoves or two \textit{young doves},\lebnote{Literally “sons of dove”}and she shall bring them to the priest at the tent of assembly’s entrance.}%
\verse{And the priest shall \textit{sacrifice}\lebnote{Literally “do” or “make”}the one as a sin offering and the \textit{other}\lebnote{Literally “one”}as a burnt offering, and so the priest shall make atonement for her \textit{before}\lebnote{Literally “the the faces of”}Adonai from her unclean body fluid discharge.’}%
\verse{“And you shall keep the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}separate from their uncleanness so that they might not die because of\lebnote{Or “by” or “in”}their uncleanness by their making my tabernacle, which is in their midst, unclean.}%
\verse{“This is the regulation of the one with the body fluid discharge and the one from whom an emission of semen goes out so that he becomes unclean by it}%
\verse{and \textit{concerning}\lebnote{Literally “of”}the menstruating woman in her bleeding\lebnote{See similar idiom at 12:2}and the person who discharges his body fluid, for the male and for the female and for a man who lies with an unclean woman.”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 16
\verseWithHeading{The Day of Atonement}{Then\lebnote{Or “And”}Adonai spoke to Moses after the death of Aaron’s two sons, when they had come near \textit{before}\lebnote{Literally “before faces of”}Adonai and they died.}%
\verse{And Adonai said to Moses, “Tell\lebnote{Or “Speak to”}your brother Aaron that he should not enter at any time into the sanctuary \textit{behind}\lebnote{Literally “from the interior/inside to”}the curtain \textit{in front of}\lebnote{Literally “to the faces of”}the atonement cover that is on the ark, so that he might not die, because I appear in the cloud over the atonement cover.}%
\verse{“Aaron must enter the sanctuary with this: \textit{a young bull}\lebnote{Literally “a bull a son of cattle”}as a sin offering and a ram as a burnt offering.}%
\verse{He must put on \textit{a holy linen tunic},\lebnote{Literally “a tunic of linen of holiness”}and linen undergarments must be on his body, and he must fasten himself\lebnote{The direct object is supplied from context in the English translation}with a linen sash, and he must wrap a linen turban around his head\lebnote{Supplied from context in the English translation}—they are holy garments, and he shall wash his body with water, then\lebnote{Or “and”}he shall put them on.}%
\verse{And he must take from the \textit{Israelites}’\lebnote{Literally “sons/children of Israel”}community two he-goats as a sin offering and one ram as a burnt offering.}%
\verse{“And Aaron shall present the sin offering’s bull, which is for himself, and so he shall make atonement for himself and for his family.}%
\verse{And he shall take the two goats, and he shall present them \textit{before}\lebnote{Literally “to the faces of”}Adonai at the tent of assembly’s entrance.}%
\verse{Then\lebnote{Or “And”}Aaron shall cast lots for the two goats: one lot for Adonai and one for Azazel.}%
\verse{And Aaron shall present the goat on which the lot for Adonai fell, and he shall \textit{sacrifice}\lebnote{Literally “do” or “make”}it as a sin offering.}%
\verse{But\lebnote{Or “And”}he must present alive \textit{before}\lebnote{Literally “to the faces of”}Adonai the goat on which the lot for Azazel fell to make atonement for himself, to send it away into the desert to Azazel.}%
\verse{“And Aaron shall present the sin offering’s bull, which is for himself, and so he shall make atonement for himself and for his family; then\lebnote{Or “and”}he shall slaughter the sin offering’s bull, which is for himself.}%
\verse{And he shall take a\lebnote{Hebrew “the”}censer full of \textit{burning charcoal}\lebnote{Literally “burning charcoal of fire”}from upon the altar from \textit{before}\lebnote{Literally “to the faces of”}Adonai and \textit{two handfuls}\lebnote{Literally “the fullness of the hollow of his hands”}of incense of powdered fragrant perfumes, and he shall bring it \textit{from behind}\lebnote{Literally “from the interior/inside of”}the curtain,}%
\verse{and he shall put the incense on the fire \textit{before}\lebnote{Literally “to the faces of”}Adonai so that the cloud of incense might cover the atonement cover, which is on the \textit{covenant text},\lebnote{Literally “testimony”}so that he might not die.}%
\verse{And he shall take \textit{some of}\lebnote{Literally “from”}the bull’s blood, and he shall spatter it\lebnote{The direct object is supplied from context in the English translation}with his finger on the atonement cover’s surface on the eastern side, and \textit{before}\lebnote{Literally “to the faces of”}the atonement cover he shall spatter \textit{some of}\lebnote{Literally “from”}the blood with his finger seven times.}%
\verse{“And he shall slaughter the sin offering’s goat, which is for the people, and he shall bring its blood \textit{from behind}\lebnote{Literally “from the interior/inside of”}the curtain, and he shall do with its blood as that which he did with\lebnote{Hebrew “to”}the bull’s blood, and he shall spatter it on the atonement cover and \textit{before}\lebnote{Literally “to the faces of”}the atonement cover.}%
\verse{Thus\lebnote{Or “And”}he shall make atonement for the sanctuary from the \textit{Israelites}’\lebnote{Literally “sons/children of Israel”}impurities and from their transgressions for all their sins; and so he must do for the tent of assembly, which dwells with them in the midst of their impurities.}%
\verse{And \textit{no person}\lebnote{Literally “any man not”}shall be in the tent of assembly when he enters to make atonement in the sanctuary until he comes out, and so he shall make atonement for himself and for his family and for all of Israel’s assembly.}%
\verse{“Then\lebnote{Or “And”}he shall go out to the altar that is \textit{before}\lebnote{Literally “to the faces of”}Adonai, and he shall make atonement for it; and he shall take \textit{some of}\lebnote{Literally “from”}the bull’s blood and \textit{some of}\lebnote{Literally “from”}the goat’s blood, and he shall put it\lebnote{The direct object is supplied from context in the English translation}all around on the altar’s horns.}%
\verse{And he shall spatter \textit{some of}\lebnote{Literally “from”}the blood on it seven times with his finger, and he shall cleanse it and consecrate\lebnote{Literally “he shall consecrate”}it from the \textit{Israelites}’\lebnote{Literally “of the sons/children of Israel”}impurities.}%
\verse{“And he shall finish making atonement for the sanctuary and the tent of assembly and the altar; then\lebnote{Or “and”}he shall present the living goat.}%
\verse{And Aaron shall place his two hands on the living goat’s head, and he shall confess over it all the \textit{Israelites}’\lebnote{Literally “of the sons/of the children of Israel”}iniquities and all their transgressions for all their sins, and he shall put them on the goat’s head, and he shall send it\lebnote{The direct object is supplied from context in the English translation}away into the desert \textit{with}\lebnote{Literally “by the hand of”}a man standing ready.\lebnote{Or “at hand”}}%
\verse{Thus\lebnote{Or “And”}the goat shall bear on it to \textit{a barren region}\lebnote{Literally “a land of separation” or “a land of infertility”}all their guilt, and he shall send the goat away into the desert.}%
\verse{“And Aaron shall enter the tent of assembly, and he shall take off the linen garments that he put on at his coming to the sanctuary, and he shall leave them there.}%
\verse{And he shall wash his body with water in a holy place, and he shall put on his garments and go out and \textit{sacrifice}\lebnote{Literally “he shall do” or “he shall make”}his burnt offering and the people’s burnt offering, and so he shall make atonement for himself and for people.}%
\verse{And he must turn into smoke the sin offering’s fat on the altar.}%
\verse{“And the person who sends out the goat for Azazel shall wash his garments, and he shall wash his body with water, and \textit{afterward}\lebnote{Literally “after thus”}he shall come to the camp.}%
\verse{And the sin offering’s bull and the sin offering’s goat, whose blood was brought to make atonement in the sanctuary, shall be brought \textit{outside the camp},\lebnote{Literally “to from an outside place of the camp”}and they shall burn their hide and their flesh and their offal in the fire.}%
\verse{And the person who burns them shall wash his garments, and he shall wash his body with water, and \textit{afterward}\lebnote{Literally “after thus”}he must come to the camp.}%
\verse{“And this shall be \textit{a lasting statute}\lebnote{Literally “a statute of eternity” or “a statute of long duration”}for you: in the seventh month, on the tenth of the month, you must deny yourselves and you must not do any work, whether the native or\lebnote{Or “and”}the alien who is dwelling in your midst,}%
\verse{because on this day he shall make atonement for you to cleanse you; you must be clean from all your sins \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{It is \textit{a Sabbath of complete rest}\lebnote{Literally “a Sabbath of ‘Sabbathation.’ ” “Sabbathation” is not a real word, but it is devised as an attempt to convey the sounds of the related nouns in the Hebrew phrase}for you, and you shall deny yourselves—it is \textit{a lasting statute}.\lebnote{Literally “a statute of eternity” or “a statute of long duration”}}%
\verse{And the priest who is anointed and who is \textit{ordained}\lebnote{Literally “filled his hand”}to serve as a priest in place of his father shall make atonement; thus\lebnote{Or “and”}he shall put on the linen garments, the holy garments,}%
\verse{and he shall make atonement for the sanctuary’s holy place, and he shall make atonement for the tent of assembly and the altar, and he shall make atonement for the priests and for all of the assembly’s people.}%
\verse{And this shall be \textit{a lasting statute}\lebnote{Literally “a statute of eternity” or “a statute of long duration”}for you to make atonement for the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}one time in a year from all their sins.”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 17
\verseWithHeading{The Place of Sacrifice}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to Aaron and to his sons and to all the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}and you shall say to them, ‘This is the word that Adonai has commanded, saying,}%
\verse{“\textit{Any man}\lebnote{Literally “A man a man”}from the house of Israel who slaughters an ox or a sheep or a goat in the camp or who slaughters it \textit{outside the camp}\lebnote{Literally “from an outside place of the camp”}}%
\verse{and he does not bring it to the tent of assembly’s entrance to present an offering to Adonai \textit{before}\lebnote{Literally “to the faces of”}Adonai’s tabernacle, then\lebnote{Or “and”}that man shall be accounted bloodguilty—he has poured out blood, and that man shall be cut off from the midst of his people.}%
\verse{This is so that the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}may bring their sacrifices that they are sacrificing \textit{in the open field}\lebnote{Literally “on the faces of the field”}and bring\lebnote{Or “they shall bring”}them for Adonai to the tent of assembly’s entrance to the priest, and they shall sacrifice fellowship offerings for Adonai with them.}%
\verse{And the priest shall sprinkle the blood on Adonai’s altar at the tent of assembly’s entrance, and he shall \textit{burn}\lebnote{Literally “turn into smoke”}the fat as an appeasing fragrance for Adonai.}%
\verse{And \textit{they may no longer sacrifice}\lebnote{Literally “not they may sacrifice again”}their sacrifices to the goat-idols after which they were prostituting. This is \textit{a lasting statute}\lebnote{Literally “a statute of eternity” or “a statute of long duration”}for them \textit{throughout}\lebnote{Literally “to”}their generations.” ’}%
\verseWithHeading{Instructions Against Eating Blood}{“And you shall say to them, ‘if there is \textit{anyone}\lebnote{Literally “A man a man”}from the house of Israel or\lebnote{Or “and”}from the alien who dwells in their midst who offers a burnt offering or a sacrifice}%
\verse{and he does not bring it to the tent of assembly’s entrance to \textit{sacrifice}\lebnote{Literally “do” or “make”}it for Adonai, then\lebnote{Or “and”}that man shall be cut off from his people.}%
\verse{And if there is \textit{anyone}\lebnote{Literally “a man a man”}from the house of Israel or\lebnote{Or “and”}from the alien who is dwelling in their midst who eats any blood, then\lebnote{Or “and”}I will set my face against the person who eats the blood, and I will cut him off from among his people.}%
\verse{Indeed\lebnote{Or “Because”}the flesh’s life is in the blood, and I have given it to you on the altar to make atonement for your lives, because it is the blood with the life that makes atonement.}%
\verse{\textit{Therefore}\lebnote{Literally “Unto thus”}I said to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}‘\textit{None of you}\lebnote{Literally “all of individual self from you not”}may eat blood, nor\lebnote{Or “and not”}may the alien who is dwelling in your midst eat blood.’}%
\verse{“And if there is \textit{anyone}\lebnote{Literally “a man a man”}from the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}or\lebnote{Or “and”}from the alien who is dwelling in their midst who hunts a wild game animal or a bird that may be eaten, then\lebnote{Or “and”}he shall pour out its blood, and he shall cover it with the soil.}%
\verse{Indeed,\lebnote{Or “Because”}the life of all flesh, its blood, is in its life, so\lebnote{Or “and”}I said to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}‘You may not eat the blood of any flesh, because the life of all flesh is its blood; anyone who eats it must be cut off.’}%
\verse{“And if there is any person who eats a dead body or\lebnote{Or “and”}a mangled carcass, whether among the native or\lebnote{Or “and”}among the alien, then\lebnote{Or “and”}he shall wash his garments, and he shall wash himself\lebnote{The direct object is supplied from context in the English translation}with water, and he shall be unclean until the evening, and he shall be clean.}%
\verse{But\lebnote{Or “And”}if he does not wash his garments\lebnote{The direct object is supplied from context in the English translation}and he does not wash his body, then\lebnote{Or “and”}he shall bear his guilt.”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 18
\verseWithHeading{Unlawful Sexual Relations}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}and say to them, ‘I am Adonai your God.}%
\verse{You must not \textit{carry out}\lebnote{Literally “do”}the practices of the land of Egypt, in which you lived, and you must not \textit{carry out}\lebnote{Literally “do”}the practices of the land of Canaan, to which I am bringing you; and you must not follow their statutes.}%
\verse{You must \textit{carry out}\lebnote{Literally “do”}my regulations, and you must observe\lebnote{Or “keep”}my statutes by following them; I am Adonai your God.}%
\verse{And you shall observe\lebnote{Or “keep”}my statutes and my regulations \textit{by which the person doing them shall live};\lebnote{Literally “which the man does them and he shall live in them”}I am Adonai.}%
\verse{“ ‘\textit{None}\lebnote{Literally “a man a man … not”}of you shall approach anyone who is \textit{his close relative}\lebnote{Literally “the flesh of his body”}to expose nakedness; I am Adonai.}%
\verse{You must not expose your father’s nakedness or\lebnote{Or “and”}your mother’s nakedness—she is your mother; you must not expose her nakedness.}%
\verse{You must not expose the nakedness of your father’s wife—it is your father’s nakedness.}%
\verse{As for your sister’s nakedness, whether your father’s daughter or your mother’s daughter, whether \textit{born at home}\lebnote{Literally “a relative of house”}or \textit{born abroad},\lebnote{Literally “a relative of an outside place”}you must not expose their nakedness.}%
\verse{As for the nakedness of your son’s daughter or your daughter’s daughter, you must not expose their nakedness, because they are your nakedness.}%
\verse{As for the nakedness of the daughter of your father’s wife, she is your sister, a relative of your father; you must not expose her nakedness.}%
\verse{You must not expose the nakedness of your father’s sister; she is \textit{your father’s close relative}.\lebnote{Literally “the flesh of your father”}}%
\verse{You must not expose the nakedness of your mother’s sister, because she is \textit{your mother’s close relative}.\lebnote{Literally “the flesh of your mother”}}%
\verse{You must not expose the nakedness of your father’s brother; you must not \textit{have sex with}\lebnote{Literally “approach”}his wife—she is your aunt.}%
\verse{You must not expose your daughter-in-law’s nakedness; she is your son’s wife; you must not expose her nakedness.}%
\verse{You must not expose the nakedness of your brother’s wife; she is your brother’s nakedness.}%
\verse{You must not expose the nakedness of a woman and her daughter, or her son’s daughter, or her daughter’s daughter; you must not take her as wife\lebnote{The verb לקח is used to speak of marrying (“taking a wife”)}to expose her nakedness; they are \textit{close relatives}\lebnote{Literally “flesh”; see vv. 6, 12, 13}—that is wickedness.}%
\verse{And you must not take as wife\lebnote{The verb לקח is used to speak of marrying (“taking a wife”)}a woman with her sister, to be a rival-wife, to expose her nakedness before her \textit{during}\lebnote{Literally “in”}her life.}%
\verse{“ ‘And you must not \textit{have sex with}\lebnote{Literally “approach”}a woman to expose her nakedness \textit{during}\lebnote{Literally “in”}her menstrual uncleanness.}%
\verse{And \textit{you must not have sex}\lebnote{Literally “you shall not give your lying down for semen”}with your fellow citizen’s\lebnote{Or “neighbor’s”}wife, becoming unclean with her.}%
\verse{“ ‘And you shall not give \textit{any of}\lebnote{Literally “from”}your offspring\lebnote{Or “descendants”}in order to sacrifice them to Molech, nor\lebnote{Or “and not”}shall you profane the name of your God; I am Adonai.}%
\verse{And you shall not lie with a male as lying \textit{with}\lebnote{Literally “of”}a woman; that is a detestable thing.}%
\verse{And you shall not \textit{have sexual relations}\lebnote{Literally “give your lying down”}with any animal, becoming unclean with it; and a woman shall not stand \textit{before}\lebnote{Literally “to the faces of”}an animal to copulate with it—that is a perversion.}%
\verse{“ ‘You must not make yourself unclean in any of these things, because the nations whom I am driving out from your presence were made unclean by all of these.}%
\verse{So\lebnote{Or “And”}the land became unclean, and \textit{I have brought the punishment of}\lebnote{Literally “I have visited”}its guilt upon it, and the land has vomited out its inhabitants.}%
\verse{But\lebnote{Or “And”}you (neither the native nor\lebnote{Or “and”}the alien who is dwelling in your midst) shall keep my statutes and my regulations, and you shall not \textit{practice}\lebnote{Literally “do”}any of these detestable things}%
\verse{(because the people of the land, who were \textit{before you},\lebnote{Literally “to the faces of you”}did all these detestable things, so the land became unclean),}%
\verse{so that the land will not vomit you out when you make it unclean just as it vomited out the nation that was \textit{before you}.\lebnote{Literally “to the faces of you”}}%
\verse{Indeed, anyone who does any of these detestable things, even\lebnote{Or “and”}those persons who do so shall be cut off from the midst of their people.}%
\verse{Thus\lebnote{Or “And”}you shall keep my obligation to not do any of the statutes \textit{regarding}\lebnote{Literally “of”}the detestable things that they did \textit{before you},\lebnote{Literally “to the faces of you”}so that you will not make yourselves unclean by them; I am Adonai your God.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 19
\verseWithHeading{Adonai Is Holy}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to all the community of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}and say to them, ‘You\lebnote{Plural}must be holy, because I, Adonai your\lebnote{Plural}God, am holy.}%
\verse{Each of you must revere your mother\lebnote{Hebrew “his mother”}and your father,\lebnote{Hebrew “his father”}and you\lebnote{Plural}must keep my Sabbaths; I am Adonai your\lebnote{Plural}God.}%
\verse{You\lebnote{Plural}must not turn to idols, and you\lebnote{Plural}must not make for yourselves gods of cast metal; I am Adonai your\lebnote{Plural}God.}%
\verse{“ ‘And when you\lebnote{Plural}sacrifice a sacrifice of fellowship offerings to Adonai, you\lebnote{Plural}must sacrifice it for your\lebnote{Plural}acceptance.}%
\verse{It must be eaten on the day of your\lebnote{Plural}sacrifice and the next day; but\lebnote{Or “and”}the remainder must be burned up in the fire by the third day.}%
\verse{And if it is indeed eaten on the third day, it is unclean meat; it shall not be regarded as accepted.}%
\verse{And the one who eats it shall bear his guilt, because he has profaned Adonai’s holiness, and that person shall be cut off from his people.}%
\verseWithHeading{Love Your Neighbor as Yourself}{“ ‘And at your\lebnote{Plural}reaping the harvest of your\lebnote{Plural}land you\lebnote{Singular}must not finish reaping the edge of your\lebnote{Singular}field, and you\lebnote{Singular}must not glean the remnants of your\lebnote{Singular}harvest.}%
\verse{And you\lebnote{Singular}must not glean your\lebnote{Singular}vineyard, and you\lebnote{Singular}must not gather your\lebnote{Singular}vineyard’s fallen grapes; you\lebnote{Singular}must leave them behind for the needy and for the alien; I am Adonai your\lebnote{Plural}God.}%
\verse{“ ‘You\lebnote{Plural}shall not steal, and you\lebnote{Plural}shall not deceive, and you\lebnote{Plural}shall not lie \textit{to one another};\lebnote{Literally “a man to his fellow citizen”}}%
\verse{and you\lebnote{Plural}shall not swear \textit{falsely}\lebnote{Literally “to the deception”}in my name, and so one of you\lebnote{Singular}profane\lebnote{Or “you shall profane”}the name of your\lebnote{Singular}God; I am Adonai.}%
\verse{“ ‘You\lebnote{Singular}shall not exploit your\lebnote{Singular}neighbor, and you\lebnote{Singular}shall not rob him; a hired worker’s wage you\lebnote{Singular}shall not \textit{withhold}\lebnote{Literally “leave with you”}overnight until morning.}%
\verse{You\lebnote{Singular}shall not curse the deaf, and you\lebnote{Singular}shall not put a stumbling block \textit{before}\lebnote{Literally “to the faces of”}a blind person, but\lebnote{Or “and”}you\lebnote{Singular}shall revere your\lebnote{Singular}God; I am Adonai.}%
\verse{“ ‘You\lebnote{Plural}shall not do injustice in judgment; \textit{you\lebnote{Singular}shall not show partiality to the powerless};\lebnote{Literally “you shall not lift up the faces of the poor/powerless”}you\lebnote{Singular}shall not give preference \textit{to the powerful};\lebnote{Literally “faces of the great”}you\lebnote{Singular}shall judge your\lebnote{Singular}fellow citizen with justice.}%
\verse{You\lebnote{Singular}shall not go about with slander among your\lebnote{Singular}people; \textit{you\lebnote{Singular}shall not endanger your\lebnote{Singular}neighbor’s life};\lebnote{Literally “you shall not stand on the blood of your neighbor”}I am Adonai.}%
\verse{“ ‘You\lebnote{Singular}shall not hate your\lebnote{Singular}brother in your\lebnote{Singular}heart; you\lebnote{Singular}shall surely rebuke your\lebnote{Singular}fellow citizen, so that you\lebnote{Singular}do not incur sin \textit{along with}\lebnote{Or “in addition to him”; literally “upon him”}him.}%
\verse{You\lebnote{Singular}shall not seek vengeance, and you\lebnote{Singular}shall not harbor a grudge against \textit{your\lebnote{Singular}fellow citizens};\lebnote{Literally “the sons of your people”}and you\lebnote{Singular}shall love your\lebnote{Singular}neighbor like yourself; \lebnote{Singular}I am Adonai.}%
\verseWithHeading{You Shall Keep My Statutes}{“ ‘You\lebnote{Plural}must keep my statutes: as for your\lebnote{Singular}domestic animals, you\lebnote{Singular}shall not cause two differing kinds to breed; as for your\lebnote{Singular}field, you\lebnote{Singular}shall not sow two differing kinds of seed; and, a garment of two differing kinds of woven material should not be worn on you.\lebnote{Singular}}%
\verse{“ ‘And when a man lies with a woman and there is an emission of semen and she is a female slave promised to a man, but\lebnote{Or “and”}she indeed has not been ransomed or freedom has not be given to her, there shall be an obligation to compensate; they shall not be put to death, because she has not been freed.}%
\verse{And he shall bring his guilt offering to Adonai at the tent of assembly’s entrance: a ram for a guilt offering.}%
\verse{And the priest shall make atonement for him \textit{before}\lebnote{Literally “to the faces of”}Adonai with the ram of the guilt offering for his sin that he \textit{committed},\lebnote{Literally “sinned”}and so his sin that he \textit{committed}\lebnote{Literally “sinned”}shall be forgiven him.}%
\verse{“ ‘And when you\lebnote{Plural}have come into the land and you\lebnote{Plural}plant any tree for\lebnote{Hebrew “of”}food, \textit{you\lebnote{Plural}shall regard its fruit as unharvestable};\lebnote{Literally “and you shall regard its foreskin as uncircumcised its fruit”}for three years it shall be forbidden for you; \lebnote{Plural}it shall not be eaten.}%
\verse{But\lebnote{Or “And”}in the fourth year all its fruit shall be holy, offerings of praise for Adonai.}%
\verse{And in the fifth year you\lebnote{Plural}shall eat its fruit to increase its yield for you; \lebnote{Plural}I am Adonai your\lebnote{Plural}God.}%
\verse{“ ‘You\lebnote{Plural}must not eat anything with the blood; you\lebnote{Plural}shall not practice divination, nor shall you\lebnote{Plural}interpret signs.}%
\verse{You\lebnote{Plural}shall not round off the corner hair of your\lebnote{Plural}head, and you\lebnote{Singular}shall not trim the corner of your\lebnote{Singular}beard.}%
\verse{And you\lebnote{Plural}shall not make a slash in your\lebnote{Plural}body for a dead person, nor shall you\lebnote{Plural}make on yourselves a tattoo’s mark; I am Adonai.}%
\verse{“ ‘You\lebnote{Singular}shall not profane your\lebnote{Singular}daughter by making her a prostitute, \textit{lest the land be prostituted and the land fill up with depravity}.\lebnote{Literally “and the land does not prostitute and the land fills up depravity” or “so that the land does not prostitute and the land fills up depravity”}}%
\verse{You\lebnote{Plural}shall keep my Sabbaths, and you\lebnote{Plural}shall revere my sanctuary; I am Adonai.}%
\verse{“ ‘You\lebnote{Plural}shall not turn to the mediums and to the soothsayers; you\lebnote{Plural}shall not seek them to become unclean with them; I am Adonai your\lebnote{Plural}God.}%
\verse{“ ‘\textit{Before}\lebnote{Literally “from the faces of”}old age you\lebnote{Singular}shall get up, and you\lebnote{Singular}shall show respect for an old person; and you\lebnote{Singular}shall revere your God; I am Adonai.}%
\verse{“ ‘And when an alien dwells with you\lebnote{Singular}in your\lebnote{Plural}land, you\lebnote{Plural}shall not oppress him.}%
\verse{The alien who is dwelling with you\lebnote{Plural}shall be like a native among you,\lebnote{Plural}and you\lebnote{Singular}shall love him like yourself, \lebnote{Singular}because you\lebnote{Plural}were aliens in the land of Egypt; I am Adonai your\lebnote{Plural}God.}%
\verse{“ ‘You\lebnote{Plural}shall not commit injustice in regulation, in measurement, in weight, or\lebnote{Or “and”}volume.}%
\verse{You\lebnote{Plural}must have honest balances, honest weights, an honest ephah, and an honest hin; I am Adonai your\lebnote{Plural}God who brought you\lebnote{Plural}out from the land of Egypt.}%
\verse{“ ‘Thus you\lebnote{Plural}shall keep all my statutes and all my regulations, and you\lebnote{Plural}shall do them; I am Adonai.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 20
\verseWithHeading{Molech Worship and Spiritism}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“And to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}you shall say, ‘If there is \textit{anyone}\lebnote{Literally “a man a man”}from the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}or\lebnote{Or “and”}from the alien who is dwelling in Israel, who gives \textit{any of}\lebnote{Literally “from”}his offspring to Molech, he must surely be put to death; the people of the land must stone him with stones.\lebnote{Hebrew “the stone”}}%
\verse{And I myself will set my face against that man, and I will cut him off from the midst of his people, because he has given \textit{some of}\lebnote{Literally “from”}his offspring to Molech, so that \textit{he makes my sanctuary unclean}\lebnote{Literally “to make unclean my sanctuary”}and profanes\lebnote{Hebrew “to profane”}\textit{my holy name}.\lebnote{Literally “the name of my holiness”}}%
\verse{And if the people of the land ever shut their eyes from\lebnote{Or “disregard”}that man at his giving \textit{some of}\lebnote{Literally “from”}his offspring to Molech, not putting him to death,}%
\verse{then\lebnote{Or “and”}I myself will set my face against that man and against his clan, and I will cut him off and all those from the midst of their people who prostitute after Molech.}%
\verse{As for\lebnote{Or “And”}the person who turns to the mediums and the soothsayers to prostitute after them, I will set\lebnote{Hebrew “and I will set”}my face against that person, and I will cut him off from the midst of his people.}%
\verse{“ ‘And you shall consecrate yourselves, and you shall be holy, because I am Adonai your God.}%
\verse{And you shall keep my statutes, and you shall do them; I am Adonai who consecrates you.}%
\verseWithHeading{Family and Sexual Offenses}{“ ‘If there is \textit{anyone}\lebnote{Literally “a man a man”}who curses his father or his mother, he shall surely be put to death; he has cursed his father and his mother—his blood is upon him.}%
\verse{“ ‘As for\lebnote{Or “And”}a man who commits adultery with a man’s wife, who commits adultery with his neighbor’s wife, both the man who commits adultery and the woman who commits adultery shall surely be put to death.}%
\verse{As for\lebnote{Or “And”}a man who lies with his father’s wife, he has exposed his father’s nakedness; both of them shall be put to death—their blood is on them.}%
\verse{As for\lebnote{Or “And”}a man who lies with his daughter-in-law, both of them shall be put to death; they have committed a perversion—their blood is on them.}%
\verse{“ ‘As for\lebnote{Or “And”}the man who lies with a male as lying with a woman, they have committed a detestable thing; they shall surely be put to death—their blood is on them.}%
\verse{“ ‘As for\lebnote{Or “And”}a man who marries a woman and her mother, that is depravity; they shall burn him and them, so that it shall not become depravity in the midst of you all.}%
\verse{“ ‘As for\lebnote{Or “And”}a man who \textit{has sexual relations}\lebnote{Literally “he gives his lying down”}with an animal, he shall surely be put to death, and you must kill the animal.}%
\verse{As for\lebnote{Or “And”}a woman who approaches any animal to copulate with it, you shall kill\lebnote{Or “and you shall kill”}the woman and the animal; they shall surely be put to death—their blood is on them.}%
\verse{“ ‘As for\lebnote{Or “And”}a man who takes his sister, his father’s daughter or his mother’s daughter, and he sees her nakedness and she herself sees his nakedness, it is a disgrace, and they shall be cut off \textit{before}\lebnote{Literally “in”}the eyes of \textit{their people};\lebnote{Literally “the sons/children of their people”}he has exposed his sister’s nakedness—he must bear his guilt.}%
\verse{As for\lebnote{Or “And”}a man who lies with a menstruating woman, he exposes\lebnote{Or “and he exposes”}her nakedness—her source he exposes and she herself reveals her blood’s source—both of them shall be cut off from the midst of their people.}%
\verse{And you shall not expose the nakedness of your mother’s sister, and you shall not expose your father’s sister, because such a person\lebnote{Hebrew “he”}has dishonored his close relative—they must bear their guilt.}%
\verse{As for\lebnote{Or “And”}a man who lies with his aunt, he has exposed his uncle’s nakedness—they shall bear their sin; they shall die childless.}%
\verse{As for\lebnote{Or “And”}a man who marries his brother’s wife, it is an abomination; he has exposed his brother’s nakedness—they shall be childless.}%
\verseWithHeading{You Shall Be Holy}{“ ‘And you shall keep all my statutes and all my regulations, and you shall do them, so that\lebnote{Or “and”}the land, to which I am bringing you to inhabit it, shall not vomit you out.}%
\verse{And you shall not follow the statutes of the nation that I am driving out \textit{from before you},\lebnote{Literally “from your faces”}because they did all these things, and I detested them.}%
\verse{So\lebnote{Or “And”}I said to you, “You yourselves shall take possession of their land, and I myself shall give it to you to possess it—a land flowing with milk and honey”; I am Adonai your God, who \textit{has set you apart}\lebnote{Literally “I have set you apart”}from the nations.}%
\verse{And you shall distinguish between the clean and the unclean animal and between the unclean and the clean bird; and you shall not defile yourselves with the animal or\lebnote{Or “and”}with the bird or\lebnote{Or “and”}with anything that moves along the ground that I have set apart for you \textit{as unclean}.\lebnote{Literally “to make unclean”}}%
\verse{And you shall be holy for me, because I, Adonai, am holy, and I have singled you out from the nations to be mine.}%
\verse{“ ‘And a man or a woman, if a spirit of the dead or a spirit of divination is in them, they shall surely be put to death; they shall stone them with stones\lebnote{Hebrew “the stone”}—their blood is on them.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 21
\verseWithHeading{Regulations Concerning Priests}{Then\lebnote{Or “And”}Adonai said to Moses, “Speak to the priests, Aaron’s sons, and say to them, ‘One must not make himself unclean for a dead person among his own people,}%
\verse{except for his direct relative closest to him: his mother and his father, and his son and his daughter, and his brother,}%
\verse{and for his sister, a virgin, who is closest to him, \textit{who has not had a husband}\lebnote{Literally “who is not for a man/husband”}—for her he may defile himself.}%
\verse{He must not make himself unclean as \textit{a kinsman by marriage},\lebnote{Literally “a husband among his people”}defiling himself.}%
\verse{“ ‘And they shall not shave bald patches on their head, and they shall not shave off the corner of their beard, and they shall not make a cut in their body.}%
\verse{They shall be holy to their God, and they shall not profane the name of their God, because they are bringing near the offerings made by fire to\lebnote{Hebrew “of”}Adonai—their God’s food—and they shall be holy.}%
\verse{“ ‘They shall not \textit{marry}\lebnote{Literally “take”}a woman who is a prostitute and defiled, nor shall they \textit{marry}\lebnote{Literally “take”}a woman divorced from her husband, because each priest\lebnote{Hebrew “he”; singular antecedent specified from the context}is holy for his God.}%
\verse{And you shall consecrate him, because he is bringing near your God’s food; he shall be holy to you, since I, Adonai, who consecrates you, am holy.}%
\verse{“ ‘As for\lebnote{Or “And”}the daughter of any priest, if she is defiled by prostituting, she is disgracing her father—she shall be burned in the fire.}%
\verse{“ ‘As for\lebnote{Or “And”}the priest who is higher than his brothers, on whose head the oil of anointment is poured and \textit{he was ordained}\lebnote{Literally “he has filled up his hand”}to wear the garments, he shall not dishevel his head, and he shall not tear his garments.}%
\verse{And he shall not go near any dead person, nor shall he make himself unclean for his father or\lebnote{Or “and”}for his mother.}%
\verse{And he shall not go out from the sanctuary, and he shall not profane his God’s sanctuary, because the dedication of his God’s oil of anointment is on him; I am Adonai.}%
\verse{“ ‘And he himself must take a wife in her virginity.}%
\verse{A widow or\lebnote{Or “and”}a divorced woman or\lebnote{Or “and”}a defiled woman, a prostitute—these he must not take; he shall take only\lebnote{Or “instead”}a virgin from his people as wife.}%
\verse{And he shall not profane his offspring among his people, because I am Adonai, who consecrates him.’ ”}%
\verse{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to Aaron, saying, ‘A man from your offspring throughout their generations, in whom is a physical defect, shall not come near to present your God’s food.}%
\verse{Indeed,\lebnote{Or “For”}any man in whom is a physical defect shall not come near: a blind man or lame or disfigured or deformed,}%
\verse{or a man in whom is a broken foot or a broken hand,}%
\verse{or a hunchback or a dwarf, or a spot in his eye or a skin disorder or a skin eruption or a crushed testicle.}%
\verse{Any man from Aaron the priest’s offspring in whom is a physical defect shall not come near to present offerings made by fire to\lebnote{Hebrew “of”}Adonai; a physical defect is in him; he shall not come near to present his God’s food.}%
\verse{He may eat his God’s food, from \textit{the most holy things}\lebnote{Literally “the holy things of the holy things”}and from the holy things.}%
\verse{But he must not enter the curtain, and he must not come near to the altar, because a physical defect is in him, and he must not profane my sanctuary, because I am Adonai, who consecrates them.’ ”}%
\verse{Thus\lebnote{Or “And”}Moses spoke to Aaron and to his sons and to all the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 22
\verseWithHeading{Priests and Their Food}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Tell\lebnote{Or “Speak to”}Aaron and his sons\lebnote{Literally “to Aaron and to his sons”}that\lebnote{Or “and”}they must deal respectfully with the \textit{Israelites}’\lebnote{Literally “sons/children of Israel”}votive offerings, and they must not profane my holy name, which they are consecrating to me; I am Adonai.}%
\verse{“Say to them, ‘Throughout your generations, any man from any of your offspring who comes near the votive offerings that the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}consecrate to Adonai with\lebnote{Or “and”}his uncleanness on him, that person shall be cut off\lebnote{Or “and that person shall be cut off”}from \textit{before me};\lebnote{Literally “to the faces of me”}I am Adonai.}%
\verse{“ ‘\textit{Anyone}\lebnote{Literally “A man a man”}from Aaron’s offspring, if\lebnote{Or “and”}he is afflicted with a skin disease or a fluid discharge, shall not eat in the sanctuary \textit{until}\lebnote{Literally “until which”}he is clean; and the one who touches any unclean person or a man from whom an emission of semen goes out,}%
\verse{or a man who touches any swarmer that is unclean for him or who touches a person who is unclean for him \textit{due to}\lebnote{Literally “for”}whatever his uncleanness,}%
\verse{a person who touches such a thing\lebnote{Hebrew “him” or “it”}shall be unclean until the evening, and he shall not eat from the votive offerings, except\lebnote{Or “but if” or “but rather”}when he washes\lebnote{Or “bathes”}his body with water}%
\verse{and \textit{the sun sets},\lebnote{Literally “the sun goes” or “the sun enters”}and he shall be clean; then\lebnote{Or “and”}afterward he may eat from the votive offerings, because it is his food.}%
\verse{He shall not eat a naturally dead body or\lebnote{Or “and”}a mangled carcass, \textit{so that he becomes unclean}\lebnote{Literally “to become unclean”}by it; I am Adonai.}%
\verse{“ ‘And they shall keep my obligation, and they shall not incur guilt because of it, so that\lebnote{Or “and”}they die through it, because they have profaned it; I am Adonai who consecrates them.}%
\verse{“ ‘\textit{No stranger shall eat}\lebnote{Literally “And any stranger shall not eat”}the votive offering; nor shall a temporary resident with\lebnote{Hebrew “of”}a priest or\lebnote{Or “and”}a hired worker eat the votive offering.}%
\verse{But\lebnote{Or “And”}a priest, if with his money he buys a person as \textit{his possession},\lebnote{Literally “property of”}that one may eat it, and the descendants of his house themselves may eat his food.}%
\verse{And a priest’s daughter, when \textit{she marries a layman},\lebnote{Literally “she becomes for a strange man”}she herself may not eat \textit{the votive offering}.\lebnote{Literally “the offering/lifting of the votive offering”}}%
\verse{But\lebnote{Or “And”}a priest’s daughter, when she becomes a widow or\lebnote{Or “and”}divorced or there is no offspring for her, and she returns to her father’s house as in her childhood, she may eat from her father’s food, but\lebnote{Or “and”}\textit{no layman may eat it}.\lebnote{Literally “any stranger shall not eat it”}}%
\verse{And if a man eats the votive offering unintentionally, then\lebnote{Or “and”}he shall add to it a fifth of it, and he shall give the votive offering to the priest.}%
\verse{And they shall not profane the \textit{Israelites}’\lebnote{Literally “sons/children of Israel”}votive offerings that they present to Adonai,}%
\verse{and so cause them, by their eating their votive offerings, to bear guilt requiring a guilt offering, because I am Adonai, who consecrates them.’ ”}%
\verseWithHeading{Acceptable Offerings}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to Aaron and to his sons and to all the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}and say to them, ‘Anyone from the house of Israel or\lebnote{Or “and”}from the alien in Israel who presents his offering for any of their vows or\lebnote{Or “and”}for any of their freewill offerings that they present to Adonai as a burnt offering,}%
\verse{it must be without defect \textit{to be acceptable for you}:\lebnote{Literally “for your acceptance”}a male among the cattle, among the sheep, or\lebnote{Or “and”}among the goats.}%
\verse{You shall not present any animal in which is a physical defect, because it shall not be \textit{acceptable}\lebnote{Literally “for acceptance”}for you.}%
\verse{And if anyone brings a sacrifice of fellowship offerings for Adonai to fulfill a vow or as a freewill offering from\lebnote{Or “among”}the cattle or from\lebnote{Or “among”}the flock,\lebnote{The Hebrew term refers collectively to both sheep and goats (small livestock animals)}it must be without defect \textit{to be acceptable};\lebnote{Literally “for acceptance”}there must not be any physical defect in it.}%
\verse{The blind or the injured or the maimed or the seeping or one with a skin disorder or one with a skin eruption—these you shall not present to Adonai, nor shall you give from them an offering made by fire on the altar for Adonai.}%
\verse{As for\lebnote{Or “And”}an ox or sheep that is deformed or\lebnote{Or “and”}that is stunted, you may present it as a freewill offering, but\lebnote{Or “and”}for a vow it will not be accepted.}%
\verse{And you shall not present anything for Adonai with bruised or\lebnote{Or “and”}shattered or\lebnote{Or “and”}torn or\lebnote{Or “and”}cut-off testicles, and you shall not \textit{sacrifice}\lebnote{Literally “do” or “make”}such in your land.}%
\verse{And you shall not present your God’s food from any of these by the hand of \textit{a foreigner},\lebnote{Literally “a son of a foreign land”}because their deformity is in them; a physical defect is in them; they shall not be accepted for you.’ ”}%
\verse{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“When an ox or a sheep or a goat is born, then\lebnote{Or “and”}it shall be under its mother for seven days, and from the eighth day and beyond it is acceptable as an offering made by fire for Adonai.}%
\verse{And you shall not slaughter an ox or a sheep and \textit{its young}\lebnote{Literally “his son” or “his child”}on \textit{the same day}.\lebnote{Literally “on day one” or “in one day”}}%
\verse{And when you sacrifice a sacrifice of thanksgiving to Adonai, you must sacrifice it to be acceptable for you.}%
\verse{It must be eaten on that day; you must not leave over anything from it until morning; I am Adonai.}%
\verse{“Thus\lebnote{Or “And”}you shall keep my commands, and you shall do them; I am Adonai.}%
\verse{“And you shall not profane my holy name, so that I may be consecrated in the midst of the \textit{Israelites};\lebnote{Literally “sons/children of Israel”}I am Adonai, who consecrates you,}%
\verse{the one who brought you out from the land of Egypt to be as God for you; I am Adonai.”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 23
\verseWithHeading{Adonai’s Feasts}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}and say to them, ‘The festivals of Adonai that you shall proclaim are holy assemblies; these are my appointed times.}%
\verseWithHeading{Adonai’s Sabbath}{“ ‘For six days work is to be done, and on the seventh day shall be \textit{a Sabbath of complete rest},\lebnote{Literally “a Sabbath of ‘Sabbathation.’ ” “Sabbathation” is not a real word, but it is devised as an attempt to convey the sounds of the related nouns in the Hebrew phrase}a holy assembly; you shall not do any work; it shall be a Sabbath for Adonai in all your dwellings.}%
\verseWithHeading{The Passover}{“ ‘These are Adonai’s appointed times, holy assemblies, which you shall proclaim at\lebnote{Or “on”}their appointed time.}%
\verse{In the first month, on the fourteenth of the month at the evening is Adonai’s Passover.}%
\verse{And on the fifteenth day of this month is Adonai’s Feast of Unleavened Bread; for seven days you shall eat unleavened bread.}%
\verse{On the first day there shall be a holy assembly for you; you shall not do \textit{any regular work}.\lebnote{Literally “all work of labor”}}%
\verse{And you shall present an offering for Adonai made by fire for seven days; on the seventh day there shall be a holy assembly; you shall not do \textit{any regular work}.’ ”\lebnote{Literally “all work of labor”}}%
\verseWithHeading{The Feast of Firstfruits}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}and say to them, ‘When you come to the land that I am about to give to you and you reap its harvest, then\lebnote{Or “and”}you shall bring a sheaf of the firstfruit of your harvest to the priest.}%
\verse{And he shall wave the sheaf \textit{before}\lebnote{Literally “to the faces of”}Adonai for your acceptance; the priest shall wave it \textit{on the day after}\lebnote{Literally “from the next day of”}the Sabbath.}%
\verse{And on the day of your waving the sheaf you shall \textit{offer}\lebnote{Or “sacrifice”; literally “do” or “make”}a \textit{yearling}\lebnote{Literally “a son of its year”}male lamb without defect as a burnt offering to Adonai.}%
\verse{And its grain offering shall be two-tenths of an ephah\lebnote{Supplied by context}of finely milled flour mixed with oil, an offering made by fire for Adonai, an appeasing fragrance; and its libation shall be a fourth of a hin\lebnote{Hebrew “the hin”}of wine.}%
\verse{And you shall not eat bread or\lebnote{Or “and”}roasted grain or\lebnote{Or “and”}ripe grain until \textit{this very same day},\lebnote{Literally “the exactly of this day”}until you present your God’s offering. This must be \textit{a lasting statute}\lebnote{Literally “a statute of eternity” or “a statute of long duration”}for your generations in all your dwellings.}%
\verseWithHeading{The Feast of Weeks}{“ ‘And you shall count for yourselves \textit{from the day after}\lebnote{Literally “from the next day of”}the Sabbath, from the day of your bringing the wave offering’s sheaf—there shall be seven full weeks.}%
\verse{Until \textit{the day after}\lebnote{Literally “from the next day of”}the seventh Sabbath you shall count fifty days; then\lebnote{Or “and”}you shall present a new grain offering for Adonai.}%
\verse{You shall bring from your dwellings for a wave offering two loaves of bread \textit{made with}\lebnote{Literally “they shall be”}two-tenths of an ephah\lebnote{Supplied by context}of finely milled flour; they must be baked with leaven—the firstfruits \textit{belonging to}\lebnote{Literally “to”}Adonai.}%
\verse{And, in addition to the bread, you shall present seven \textit{yearling}\lebnote{Literally “a son of its year”}male lambs without defects and one \textit{young bull}\lebnote{Literally “a bull a son of cattle”}and two rams—they shall be a burnt offering for Adonai with\lebnote{Or “and”}their grain offering and their libations, an offering made by fire, an appeasing fragrance for Adonai.}%
\verse{And you shall \textit{offer}\lebnote{Or “sacrifice”; literally “do” or “make”}one he-goat as a sin offering and two \textit{yearling}\lebnote{Literally “a son of its year”}male lambs as a sacrifice of fellowship offerings.}%
\verse{And the priest shall wave them with the bread of the firstfruits as a wave offering \textit{before}\lebnote{Literally “to the faces of”}Adonai; in addition to the two male lambs, they shall be holy for Adonai for the priest.}%
\verse{And \textit{you shall make a proclamation}\lebnote{Literally “you shall proclaim”}on \textit{this very same day};\lebnote{Literally “the exactly of this day”}it shall be a holy assembly for you; you shall not do \textit{any regular work};\lebnote{Literally “all work of labor”}this is a \textit{lasting statute}\lebnote{Literally “a statute of eternity” or “a statute of long duration”}in all your dwellings throughout\lebnote{Or “for”}your generations.}%
\verse{And when you\lebnote{Plural}reap the harvest of your\lebnote{Plural}land, you\lebnote{Singular}must not finish the edge of your\lebnote{Singular}field at your reaping, and you\lebnote{Singular}must not glean the remnants of your\lebnote{Singular}harvest—you\lebnote{Singular}shall leave them behind for the needy and for the alien; I am Adonai your\lebnote{Plural}God.’ ”}%
\verseWithHeading{The Feast of Trumpets}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}saying, ‘In the seventh month, on the first day of the month, \textit{you must have}\lebnote{Literally “it must be for yourselves”}a rest period, a remembrance of the trumpet blast, a holy assembly.}%
\verse{You must not do \textit{any regular work},\lebnote{Literally “all work of labor”}and you shall present an offering made by fire to Adonai.’ ”}%
\verseWithHeading{The Day of Atonement}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Surely the Day of Atonement is on the tenth day of the seventh month; it shall be a holy assembly for you, and you shall deny yourselves, and you shall present an offering made by fire to Adonai.}%
\verse{And you must not do \textit{any regular work}\lebnote{Literally “all work of labor”}on \textit{this very same day},\lebnote{Literally “the exactly of this day”}because it is the Day of Atonement to make atonement for you \textit{before}\lebnote{Literally “to the faces of”}Adonai your God.}%
\verse{If there is any person who does not deny himself on \textit{this very same day},\lebnote{Literally “the exactly of this day”}then\lebnote{Or “and”}he shall be cut off from his people.}%
\verse{As for\lebnote{Or “And”}any person who does any work on \textit{this very same day},\lebnote{Literally “the exactly of this day”}I will exterminate that person from the midst of his people.}%
\verse{You must not do any work; it is a \textit{lasting statute}\lebnote{Literally “a statute of eternity” or “a statute of long duration”}throughout your generations in all your dwellings.}%
\verse{It is \textit{a Sabbath of complete rest}\lebnote{Literally “a Sabbath of ‘Sabbathation.’ ” “Sabbathation” is not a real word, but it is devised as an attempt to convey the sounds of the related nouns in the Hebrew phrase}for you, and you shall deny yourselves on the ninth day of the month in the evening—from evening to evening you must observe your extraordinary\lebnote{Supplied from the immediate context within this verse}Sabbath.”}%
\verseWithHeading{The Feast of Booths}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}saying, ‘On the fifteenth day of the seventh month, this shall be the Feast of Booths for seven days for Adonai.}%
\verse{On the first day there shall be a holy assembly; you must not do any \textit{regular work}.\lebnote{Literally “work of labor”}}%
\verse{For seven days you must present an offering made by fire to Adonai. On the eighth day it shall be a holy assembly for you, and you shall present an offering made by fire to Adonai; it is a celebration; you must not do \textit{any regular work}.\lebnote{Literally “all work of labor”}}%
\verseWithHeading{Summary}{“ ‘These are Adonai’s festivals, which you must proclaim, holy assemblies to present an offering made by fire to Adonai—burnt offering and grain offering, sacrifice and libations, \textit{each on its proper day}\lebnote{Literally “a thing of a day on its day”}—}%
\verse{\textit{besides}\lebnote{Literally “from to alone”}Adonai’s Sabbaths and \textit{besides}\lebnote{Literally “from to alone”}your gifts and \textit{besides}\lebnote{Literally “from to alone”}your vows and \textit{besides}\lebnote{Literally “from to alone”}all your freewill offerings that you give to Adonai.}%
\verse{“ ‘Surely on the fifteenth day of the seventh month, at your gathering the land’s produce, you shall hold Adonai’s festival for seven days; on the first day there shall be a rest period and on the eighth day a rest period.}%
\verse{And on the first day you shall take for yourselves the first fruit of majestic trees, branches of palm trees and branches\lebnote{Collective singular}of a leafy tree and of a brook’s poplar trees, and you shall rejoice \textit{before}\lebnote{Literally “to the faces of”}Adonai your God for seven days.}%
\verse{And you must hold it as a festival for Adonai for seven days in the year; it shall be a \textit{lasting statute}\lebnote{Literally “a statute of eternity” or “a statute of long duration”}throughout your generations; in the seventh month you must hold it.}%
\verse{You must live in the booths for seven days; all the natives in Israel must live in the booths,}%
\verse{so that your generations shall know that I made the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}live in booths when I brought them from the land of Egypt; I am Adonai your God.’ ”}%
\verse{Thus\lebnote{Or “And”}Moses announced to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}Adonai’s appointed times.}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 24
\verseWithHeading{The Sanctuary’s Lamp and Bread}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Command the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}that\lebnote{Or “and”}they should bring pure olive oil from beaten olives for the light to \textit{present}\lebnote{Literally “to cause raising up”; or “set up” (HALOT 830 s.v. 4.d)}a lamp continually.}%
\verse{Aaron shall arrange for it \textit{outside}\lebnote{Literally “from the outside of to”}the curtain of the testimony in the tent of assembly from evening until morning \textit{before}\lebnote{Literally “to the faces of”}Adonai continually; it shall be a \textit{lasting statute}\lebnote{Literally “a statute of eternity” or “a statute of long duration”}throughout your generations.}%
\verse{On the pure golden\lebnote{Supplied by context and the description of the lampstand in Exod 25:31}lampstand he shall arrange for the lamps \textit{before}\lebnote{Literally “to the faces of”}Adonai continually.}%
\verse{“And you shall take finely milled flour, and you shall bake with it twelve ring-shaped bread cakes: each one shall be two-tenths of an ephah.}%
\verse{And you shall place them in two rows, six to the row, on the pure gold\lebnote{Supplied by context and the description of the table in Exod 25:24}table \textit{before}\lebnote{Literally “to the faces of”}Adonai.}%
\verse{And you shall put pure frankincense on each\lebnote{Hebrew “the”}row so that\lebnote{Or “and”}it shall be for the bread as a memorial offering, an offering made by fire for Adonai.}%
\verse{\textit{On every Sabbath}\lebnote{Literally “On the day of the Sabbath on the day of the Sabbath”}he shall arrange it in rows \textit{before}\lebnote{Literally “to the faces of”}Adonai continually; they are from the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}as an everlasting\lebnote{Or “eternal” or “enduring” or “perpetual”}covenant.}%
\verse{And it shall be for Aaron and for his sons, and they shall eat it in a holy place, because it is \textit{a most holy thing}\lebnote{Literally “a holy thing of holy things”}for him from Adonai’s offerings made by fire—a \textit{lasting rule}.”\lebnote{Literally “rule of eternity” or “rule of long duration”}}%
\verseWithHeading{Punishment for Blasphemy}{And an Israelite woman’s son, and he was an Egyptian man’s son, went out in the midst of the \textit{Israelites};\lebnote{Literally “sons/children of Israel”}and the Israelite woman’s son and an\lebnote{Hebrew “the”}Israelite man fought in the camp.}%
\verse{Then\lebnote{Or “And”}the Israelite woman’s son blasphemed the name, and he cursed, so\lebnote{Or “and”}they brought him to Moses—and the name of his mother was Shelomith the daughter of Dibri, of the tribe of Dan.}%
\verse{Then\lebnote{Or “And”}they put him in custody \textit{so that}\lebnote{Literally “to”}the matter might be made clear to them in accordance with the mouth of Adonai.}%
\verse{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Bring the curser \textit{outside the camp},\lebnote{Literally “to from an outside place of the camp”}and all the hearers shall place their hands on his head, and the whole community shall stone him.}%
\verse{And you shall speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}saying, ‘\textit{Any man}\lebnote{Literally “A man a man”}when he curses his God shall bear\lebnote{Or “and he shall bear”}his sin.}%
\verse{And he who blasphemes Adonai’s name certainly shall be put to death; the whole community certainly shall stone him. As the alien, so the native shall be put to death at blaspheming his name.}%
\verse{“ ‘And when a man kills any human being, he certainly shall be put to death.}%
\verse{And \textit{he who kills}\lebnote{Literally “he who kills the soul of” or “he who kills the life of”}a domestic animal must repay for it life in place of life.}%
\verse{And when a man \textit{causes}\lebnote{Literally “gives”}a physical defect in his fellow citizen \textit{according to}\lebnote{Literally “as that”}what he has done, so it shall be done to him:}%
\verse{fracture in place of fracture, eye in place of eye, tooth in place of tooth—\textit{according to}\lebnote{Literally “as that”}the physical defect he \textit{causes}\lebnote{Literally “gives”}to the person, likewise it shall be \textit{caused}\lebnote{Literally “given”}to him.}%
\verse{And a killer of a domestic animal must repay for it, and a killer of a human shall be put to death.}%
\verse{\textit{You must have}\lebnote{Literally “it shall be for you”}one norm; as for the alien, so\lebnote{Hebrew “as”}it must be for the native, because I am Adonai your God.’ ”}%
\verse{Thus\lebnote{Or “And”}Moses spoke to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}and they brought the curser \textit{outside the camp},\lebnote{Literally “to from an outside place of the camp”}and they stoned him with stones, and the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}did just as Adonai had commanded Moses.}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 25
\verseWithHeading{The Sabbath Year}{Then\lebnote{Or “And”}Adonai spoke to Moses on \textit{Mount Sinai},\lebnote{Literally “the mountain of Sinai”}saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}and say to them, ‘When you\lebnote{Plural}come into the land that I am about to give to you,\lebnote{Plural}then\lebnote{Or “and”}the land shall observe a Sabbath for Adonai.}%
\verse{Six years you\lebnote{Singular throughout this verse}shall sow your field, and six years you shall prune your vineyard, and you shall gather its yield.}%
\verse{But\lebnote{Or “And”}in the seventh year it shall be \textit{a Sabbath of complete rest}\lebnote{Literally “a Sabbath of ‘Sabbathation.’ ” “Sabbathation” is not a real word, but it is devised as an attempt to convey the sounds of the related nouns in the Hebrew phrase}for the land—a Sabbath for Adonai; you\lebnote{Singular throughout this verse}must not sow your field, and you must not prune your vineyard.}%
\verse{You\lebnote{Singular throughout this verse}must not reap your harvest’s aftergrowth, and you must not harvest the grapes of your unpruned vines—it shall be \textit{a year of complete rest}\lebnote{Literally “a year of a rest period”}for the land.}%
\verse{And a Sabbath of the land shall be for food for you:\lebnote{Plural}for you\lebnote{Singular here and through the rest of this verse}and for your slave and for your slave woman and for your hired worker and for your temporary residents\lebnote{Collective singular; Hebrew “temporary resident”}who are dwelling as aliens with you;}%
\verse{and all its yield shall be for your\lebnote{Singular throughout this verse}domestic animal and for the wild animal, which are in your land to eat.}%
\verseWithHeading{The Year of Jubilee}{“ ‘And you\lebnote{Singular throughout this verse}shall count for yourself seven Sabbaths of years, seven times seven years, and they shall be for you \textit{time periods of}\lebnote{Literally “days of”}years: \textit{forty-nine}\lebnote{Literally “nine and forty”}years.}%
\verse{And you\lebnote{Singular}shall cause \textit{a loud horn blast}\lebnote{Literally “a ram’s horn of a blast”}to be heard on the seventh month on the tenth of the month; on the Day of Atonement you\lebnote{Plural}shall cause a ram’s horn to be heard in all your\lebnote{Plural}land.}%
\verse{And you\lebnote{Plural throughout this verse}shall consecrate the fiftieth year, and you shall proclaim a release in the land for all its inhabitants. It is a Jubilee; it shall be for you, and you shall return. You must return—everyone to his property and everyone to his clan.\lebnote{Or “each of you must return to his property and to his clan”}}%
\verse{\textit{You\lebnote{Plural throughout this verse}shall have the fiftieth year as a Jubilee};\lebnote{Literally “a Jubilee it the year of the fiftieth year it shall be for you”}you must not reap its aftergrowth, and you must not harvest its unpruned vines.}%
\verse{Because it is a Jubilee, it shall be holy to you. You must eat its\lebnote{That is, the field’s}produce from the field.}%
\verse{“ ‘In this Year of Jubilee \textit{each of you\lebnote{Plural}shall return}\lebnote{Literally “you shall return a man” or “you shall return everyone”}to his property.}%
\verse{And when you\lebnote{Plural}sell something to your\lebnote{Singular}fellow citizen or you buy from your\lebnote{Singular}neighbor’s hand, you\lebnote{Plural}must not oppress \textit{one another}.\lebnote{Literally “a man his brother”}}%
\verse{You\lebnote{Singular throughout this verse}must buy from your fellow citizen according to the number of years after the Jubilee; he must sell to you according to the number of years of yield.}%
\verse{You\lebnote{Singular throughout this verse}must increase its price \textit{according to a greater number of years},\lebnote{Literally “to the mouth of many of the years”}but\lebnote{Or “and”}you must decrease its price \textit{according to a lesser number of years},\lebnote{Literally “to the mouth of few of the years”}because he is selling its yields to you.}%
\verse{And you\lebnote{Plural}must not oppress \textit{one another},\lebnote{Literally “a man his citizen”}but\lebnote{Or “and”}you\lebnote{Singular}shall revere your God, because I am Adonai, your\lebnote{Plural}God.}%
\verse{“ ‘And you\lebnote{Plural throughout this verse}shall do my statutes, and you must keep my regulations, and you shall do them, so that\lebnote{Or “and”}you shall live \textit{securely}\lebnote{Literally “with confidence”}on the land.}%
\verse{And the land shall give its fruit, and you\lebnote{Plural throughout this verse}shall eat \textit{your fill},\lebnote{Literally “to contentment”}and you shall live \textit{securely}\lebnote{Literally “with confidence”}on it.}%
\verse{And if you\lebnote{Plural}should say, “What shall we eat in the seventh year, \textit{if}\lebnote{Literally “look”}we do not sow and we do not gather its yield?”}%
\verse{then\lebnote{Or “and”}I will command my blessing for you\lebnote{Plural}in the sixth year, so that\lebnote{Or “and”}it will make\lebnote{Or “produce”}the yield for three years.}%
\verse{And you\lebnote{Plural throughout this verse}will sow in the eighth year, and you shall eat from the old yield;\lebnote{The verse divides here}until the ninth year, until the coming of its yield, you shall eat the old yield.}%
\verse{“ ‘But\lebnote{Or “And”}the land must not be sold in\lebnote{Or “for”}perpetuity, because the land is mine, because you\lebnote{Plural}are aliens and temporary residents with me.}%
\verse{And in all your\lebnote{Plural}property’s land you\lebnote{Plural}must provide redemption for the land.}%
\verse{“ ‘When your\lebnote{Singular}brother becomes poor and he sells part of his property, then\lebnote{Or “and”}\textit{his nearest redeemer}\lebnote{Literally “his redeemer the nearest to him”}shall come, and he shall redeem the thing sold by his brother.}%
\verse{But\lebnote{Or “And”}if\lebnote{Or “when”}a man \textit{does not have}\lebnote{Literally “it is not for him”}a redeemer, then\lebnote{Or “and”}\textit{he prospers}\lebnote{Literally “his hand produces”}and he finds enough for his redemption,}%
\verse{then\lebnote{Or “and”}he shall calculate the years of its selling, and he shall refund the balance to the man to whom he sold it, and he shall return to his property.}%
\verse{But\lebnote{Or “And”}if his hand does not find enough to refund to him, then\lebnote{Or “and”}\textit{what he has sold}\lebnote{Literally “his thing sold”}shall be in the buyer’s hand until the Year of Jubilee; and it shall go out of the buyer’s hand\lebnote{Meaning derived from context; an alternative translation could be “shall not be released” (cf. NKJV, NRSV, ESV, NJPS) or “shall not revert” (NASB, NET)}in the Jubilee, and he shall return to his property.}%
\verse{“ ‘And if a man sells \textit{a residential house in a walled city},\lebnote{Literally “a house of a dwelling of a city wall”}then\lebnote{Or “and”}it shall be his redemption until completing \textit{a year after his selling};\lebnote{Literally “a year of his selling”}its redemption \textit{shall last}\lebnote{Literally “shall be”}\textit{a year}.\lebnote{Literally “days”}}%
\verse{But\lebnote{Or “And”}if it is not redeemed \textit{before a full year has passed},\lebnote{Literally “until the fulfilling of for him/it an entire year”}then\lebnote{Or “and”}the house that is \textit{in the walled city}\lebnote{Literally “in the city that for it a wall”}shall belong to the buyer in\lebnote{Hebrew “for”}perpetuity throughout his generations; it shall not go out of the buyer’s hand\lebnote{Meaning derived from context; an alternative translation could be “shall not be released” (cf. NKJV, NRSV, ESV, NJPS) or “shall not revert” (NASB, NET)}in the Jubilee.}%
\verse{However,\lebnote{Or “And” or “But”}village houses that have no surrounding wall shall be considered \textit{open country};\lebnote{Literally “the field of the land”}there is redemption for it, and in the Jubilee it shall go out of the buyer’s hand.\lebnote{Meaning derived from context; an alternative translation could be “shall not be released” (cf. NKJV, NRSV, ESV, NJPS) or “shall not revert” (NASB, NET)}}%
\verse{“ ‘As for\lebnote{Or “And”}the cities of the Levites, that is, the houses in\lebnote{Hebrew “of”}their property’s cities, it shall be \textit{a lasting redemption}\lebnote{Or “a permanent redemption”; literally “redemption of eternity” or “redemption of long duration”}for the Levites.}%
\verse{And whatever anyone redeems from the Levites then\lebnote{Or “and”}must go out of the buyer’s hand\lebnote{Meaning derived from context; an alternative translation could be “shall not be released” (cf. NKJV, NRSV, ESV, NJPS) or “shall not revert” (NASB, NET)}in the Jubilee, including a house’s selling \textit{in his city’s property},\lebnote{Literally “the city of his property”}because the houses in\lebnote{Hebrew “of”}the cities of the Levites are their property in the midst of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}}%
\verse{But\lebnote{Or “And”}a\lebnote{Hebrew “the”}field of their cities’ pastureland must not be sold, because \textit{it is their property for all time}.\lebnote{Literally “a property of eternity it is for them” or “a property of long duration it is for them”}}%
\verse{“ ‘And if your\lebnote{Singular throughout this verse}countryman\lebnote{Or “brother”}becomes poor and \textit{if he becomes dependent on you},\lebnote{Literally “and his hand is shaky with you”}then\lebnote{Or “and”}you shall support him like an alien and like a temporary resident, and he shall live with you.}%
\verse{You\lebnote{Singular throughout this verse}must not take interest or\lebnote{Or “and”}usury from him, but\lebnote{Or “and”}you shall revere your God, and your countryman\lebnote{Or “brother”}shall live with you.}%
\verse{You\lebnote{Singular throughout this verse}must not give your money to him with interest or\lebnote{Or “and”}give your food for\lebnote{Or “in” or “with”}profit.}%
\verse{I am Adonai your\lebnote{Plural throughout this verse}God, who brought you out from the land of Egypt to give \textit{you}\lebnote{Literally “to you”}the land of Canaan, to be as God for you.}%
\verse{“ ‘And if your\lebnote{Singular throughout this verse}countryman\lebnote{Or “brother”}who is with you becomes poor, and he is sold to you, \textit{you shall not treat him as a slave}.\lebnote{Literally “you shall not let him work the work of a slave” or “you shall not enslave him the work of a slave”}}%
\verse{He shall be with you like a hired worker, like a temporary resident; he shall work with you until the Year of Jubilee.}%
\verse{And he and his sons with him shall go out from you, and he shall return to his clan, and to the property of his ancestors\lebnote{Or “fathers”}he shall return.}%
\verse{Because they are my servants whom I brought out from the land of Egypt, they shall not be sold \textit{as a slave}.\lebnote{Literally “a selling of a slave”}}%
\verse{You\lebnote{Singular throughout this verse}shall not rule over him with ruthlessness, but\lebnote{Or “and”}you shall revere your God.}%
\verse{“ ‘As for\lebnote{Or “And”}your\lebnote{Singular}slave and your\lebnote{Singular}slave woman who are\lebnote{Or “may be” (permission; cf. NASU, ESV, NRSV) or “can be” (ability)}yours,\lebnote{Singular}from the nations that are all around you,\lebnote{Plural}from them you\lebnote{Plural}may buy a slave or\lebnote{Or “and”}a slave woman.}%
\verse{And you\lebnote{Plural throughout this verse}may buy also from the children\lebnote{Or “sons”}of the temporary residents who are dwelling with you as aliens and from their clan who are with you, who have children in your land; indeed,\lebnote{Or “and”}they may be as property for you.}%
\verse{And you\lebnote{Plural}may pass them on as an inheritance to your\lebnote{Plural}sons\lebnote{Or “children”}after you\lebnote{Plural}to take possession of as property \textit{for all time}\lebnote{Literally “to eternity” or “forever” or “for a long duration”}—you\lebnote{Plural}may let them work. But\lebnote{Or “And”}as for your\lebnote{Plural}countrymen,\lebnote{Or “brothers”}the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}you\lebnote{Singular}shall not rule with ruthlessness over \textit{one another}.\lebnote{Literally “a man over his brother”}}%
\verse{“ ‘And if the alien or\lebnote{Or “and”}the temporary resident who are with you\lebnote{Singular throughout this verse}\textit{prosper},\lebnote{Literally “the hand produces”}but\lebnote{Or “and”}your countryman\lebnote{Or “brother”}who is with him becomes poor and he is sold to an alien, a temporary resident who is with you, or to a descendant of an alien’s clan,}%
\verse{after he is sold redemption shall be for him; one of his brothers may redeem him,}%
\verse{or his uncle or his uncle’s son may redeem him, or \textit{one of}\lebnote{Literally “from”}\textit{his close relatives}\lebnote{Literally “the direct relative of his flesh”}from his clan may redeem him; or if \textit{he prospers},\lebnote{Literally “his hand produces”}he may redeem himself.}%
\verse{And he shall calculate with his buyer from the year of \textit{his selling himself}\lebnote{Literally “his being sold for him”}until the Jubilee; and the value of his selling shall be according to the number of years—it shall be with him like\lebnote{Or “as”}a hired worker’s days.}%
\verse{If there are still many years, \textit{in keeping with them}\lebnote{Literally “to the mouth of them”}he shall restore his redemption \textit{in proportion to his purchase price}.\lebnote{Literally “from the price of his acquisition”}}%
\verse{And if there are a few years left until the Year of Jubilee, then\lebnote{Or “and”}he shall calculate for himself; he shall restore his redemption \textit{according to the number of his years}.\lebnote{Literally “according to the mouth of his years”}}%
\verse{He shall be with him \textit{as a yearly hired worker};\lebnote{Literally “as a hired worker of a year in a year”}he shall not rule over him with ruthlessness \textit{in your\lebnote{Singular}sight}.\lebnote{Literally “to your eyes” or “for your eyes”}}%
\verse{And if he is not redeemed by\lebnote{Or “in”}any of these ways,\lebnote{Bracketed words provided from the sense of the context}then\lebnote{Or “and”}he and his sons with him shall go out in the Year of Jubilee.}%
\verse{Indeed,\lebnote{Emphatic use of כִּי, since there is no indication of direct causation to only what precedes in vv. 47–54; or “Because”}the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}are servants for me; they are my servants whom I brought out from the land of Egypt. I am Adonai your\lebnote{Plural}God.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 26
\verseWithHeading{Blessings for Obedience}{“ ‘You\lebnote{Plural throughout the entire chapter}shall not make for yourselves idols and divine images, and you shall not raise up stone pillars for yourselves, and you shall not put a sculptured stone in your land in order to\lebnote{Or “to”}worship before it, because I am Adonai your God.}%
\verse{“ ‘You shall keep my Sabbaths, and you shall revere my sanctuary; I am Adonai.}%
\verse{“ ‘If you walk in my statutes and you keep my commands and you do them,}%
\verse{then\lebnote{Or “and”}I will give you rains in their time, and the land shall give its produce, and the trees of the field shall give their fruit.}%
\verse{And for you the threshing season shall overtake the grape harvest, and the grape harvest shall overtake the sowing, and you shall eat your food \textit{to your fill}\lebnote{Literally “to contentment”}and you shall live \textit{securely}\lebnote{Literally “with confidence”}in your land.}%
\verse{And I will give peace in the land, and you shall lie down, and there shall not be \textit{anybody who makes you afraid},\lebnote{Literally “who makes afraid” or “making afraid”}and I will remove harmful animals from the land, and \textit{no sword shall pass through your land}.\lebnote{Literally “a sword shall not pass through in your land”}}%
\verse{And you shall pursue your enemies, and they shall fall by the sword \textit{before you}.\lebnote{Literally “to your faces”}}%
\verse{And five of\lebnote{Hebrew “from”}you shall pursue a hundred, and a hundred of\lebnote{Hebrew “from”}you shall pursue a myriad;\lebnote{Or “ten thousand”}and your enemies shall fall by the sword \textit{before you}.\lebnote{Literally “to your faces”}}%
\verse{And I will turn to you, and I will make you fruitful, and I will make you numerous; and I will keep my covenant with you.}%
\verse{And you shall eat \textit{old grain},\lebnote{Literally “old what is stale”}and \textit{you shall clear away the old before the new}.\lebnote{Literally “old from the faces of new you shall bring out”}}%
\verse{And I will put my dwelling place in your midst, and my inner self\lebnote{Or “soul”}shall not abhor you.}%
\verse{And I will walk about in your midst, and I shall be your God,\lebnote{Or “as a God for you”}and you shall be my people.\lebnote{Or “as a people for me”}}%
\verse{I am Adonai, your God who brought you out from the land of Egypt, from being their slaves; and I broke the bars of your yoke, and I caused you to walk erectly.}%
\verseWithHeading{Punishment for Disobedience}{“ ‘But\lebnote{Or “And”}if you do not listen to me and you do not carry out\lebnote{Or “you do not do” or “you do not observe” or “you do not perform”}all these commands,}%
\verse{and if you reject my statutes and if your inner self\lebnote{Or “soul”}abhors my regulations, to not carry out\lebnote{Or “do” or “observe” or “perform”}all my commands by your breaking my covenant,}%
\verse{I \textit{in turn}\lebnote{Literally “also” or “indeed”}will do this to you: then\lebnote{Or “and”}I will summon onto you horror, the wasting disease, and the fever that wastes eyes and that drains away life; and you shall sow your seed \textit{in vain},\lebnote{Literally “for the emptiness”}and your enemies shall eat it.}%
\verse{And I will set my face against you, and you shall be defeated \textit{before}\lebnote{Literally “to the faces of”}your enemies; and your haters shall rule over you, and you shall flee away, but\lebnote{Or “and”}there shall not be \textit{anybody who is pursuing}\lebnote{Literally “who pursues” or “pursuing”}you.}%
\verse{“ ‘And if in spite of these things you do not listen to me, then\lebnote{Or “and”}I will continue to discipline you seven times for your sins.}%
\verse{And I will break the pride of your strength; and I will make your heaven like iron\lebnote{Or “iron ore”}and your land like copper.}%
\verse{And your strength shall be consumed \textit{in vain};\lebnote{Literally “to the emptiness”}and your land shall not give its produce, and the land’s trees shall not give their fruit.}%
\verse{“ ‘And if you go against me in hostility and you are not willing to listen to me, then\lebnote{Or “and”}I will add a plague onto you seven times according to your sins.}%
\verse{And I will send \textit{wild animals}\lebnote{Literally “the animals of the field”}out among you, and they shall make you childless, and they shall cut down your domestic animals, and they shall make you fewer; and your roads shall be desolate.}%
\verse{“ ‘And if you do not accept correction from\lebnote{Hebrew “for” or “to”}me through these things, but\lebnote{Or “and”}you go against me in hostility,}%
\verse{then\lebnote{Or “and”}I myself\lebnote{Emphatic personal pronoun}will also go against you in hostility, and I myself\lebnote{Emphatic personal pronoun}also will strike you seven times for your sins.}%
\verse{And I will bring upon you a sword that seeks vengeance for the covenant, and you shall be gathered to your cities; and I will send a plague in your midst, and you shall be given into the hand of an enemy.}%
\verse{At my breaking the \textit{supply}\lebnote{Literally “staff”}of bread\lebnote{Or “food”}for you, then\lebnote{Or “and”}ten women shall bake your bread in one oven, and they shall return your bread by weight; and you shall eat it, and you shall not be satisfied.}%
\verse{“ ‘And if through this you do not listen to me and you go against me in hostility,}%
\verse{then\lebnote{Or “and”}I will go against you in hostile anger, and also\lebnote{Or “surely”}I myself\lebnote{Emphatic personal pronoun}will discipline you seven times for your sins.}%
\verse{And you shall eat the flesh of your sons; and the flesh of your daughters you shall eat.}%
\verse{And I will destroy your high places, and I will cut down your incense altars, and I will place your corpses on your idols’ corpses; and my inner self\lebnote{Or “soul”}shall abhor you.}%
\verse{And I will lay your cities in ruins, and I will lay waste your sanctuaries; and I shall not smell your sacrifices’\lebnote{Implied by the use of the same phraseology in regard to the sacrifices in the early chapters of the book}appeasing fragrance.}%
\verse{And I myself\lebnote{Emphatic personal pronoun}will lay waste the land, and your enemies who are living in it shall be appalled over it.}%
\verse{And I will scatter you among the nations, and I will draw a sword behind you; and your land shall be a desolation, and your cities shall be a ruin.}%
\verse{Then the land shall enjoy its Sabbaths all the days of its lying desolate, and you shall be in the land of your enemies; then the land shall rest, and it shall enjoy its Sabbaths.}%
\verse{All the days of its lying desolate it shall rest for the time\lebnote{Implied by context}that it had not rested during your Sabbaths while you were living on it.}%
\verse{As for\lebnote{Or “And”}the ones who remain among you, I will bring\lebnote{Or “and I will bring”}fearfulness in their hearts in the land of their enemies; and a sound of a windblown leaf shall pursue them, and they shall flee like flight \textit{before}\lebnote{Literally “of”}a sword, and they shall fall, but\lebnote{Or “and”}there shall not be a pursuer.}%
\verse{And they shall stumble over \textit{one another}\lebnote{Literally “a man on his brother”}as \textit{from before}\lebnote{Literally “from the faces of”}a sword, but\lebnote{Or “and”}there shall not be a pursuer; and \textit{you shall have no resistance}\lebnote{Literally “it shall not be for you resistance”}\textit{before}\lebnote{Literally “to the faces of”}your enemies.}%
\verse{And you shall perish among the nations, and the land of your enemies shall eat you.}%
\verse{And because of their guilt, the ones among you who remain shall decay in the land of their enemies; and also because of the iniquities of their ancestors,\lebnote{Or “fathers”}they shall decay with them.}%
\verse{“ ‘But\lebnote{Or “And”}when they confess their guilt and the guilt of their ancestors\lebnote{Or “fathers”}in their infidelity that they displayed against me, and moreover that they went against me in hostility—}%
\verse{I myself\lebnote{Emphatic personal pronoun}also\lebnote{Or “surely”}went against them in hostility, and I brought them into the land of their enemies—or if then their uncircumcised heart is humbled and then they pay for their guilt,}%
\verse{I will remember\lebnote{Or “and I will remember”}my covenant with Jacob; and I will remember also my covenant with Isaac and also my covenant with Abraham, and I will remember the land.}%
\verse{And the land shall be deserted by them, and it shall enjoy its Sabbaths in its being desolate from them, and they themselves\lebnote{Emphatic personal pronoun}must pay for their guilt, \textit{simply because}\lebnote{Literally “because and in because”}they rejected my regulations, and their inner self\lebnote{Or “soul”}abhorred my statutes.}%
\verse{And \textit{in spite of}\lebnote{Literally “also even” or “moreover also” or “moreover even”}this, \textit{when they are}\lebnote{Literally “in their being”}in the land of their enemies I will not reject them, and I will not abhor them to destroy them, to break my covenant with them, because I am Adonai their God.}%
\verse{And I will remember the first covenant for them\lebnote{Or “on behalf of them” or “on their behalf”}—whom I brought out from the land of Egypt \textit{in the sight of}\lebnote{Literally “to the eyes of” or “for the eyes of”}the nations \textit{to be their God}.\lebnote{Literally “to be for them for God” or “to be for them as a God”}I am Adonai.’ ”}%
\verse{These are the rules and the regulations and the laws that Adonai gave between himself and the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}on \textit{Mount Sinai}\lebnote{Literally “the mountain of Sinai”}\textit{through}\lebnote{Literally “in/by the hand of”}Moses.}%
\end{biblechapter}

\begin{biblechapter} % Leviticus 27
\verseWithHeading{Instructions About Vows}{Then\lebnote{Or “And”}Adonai spoke to Moses, saying,}%
\verse{“Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”}and say\lebnote{Or “and you shall say”}to them, ‘When a man makes a vow according to your\lebnote{Singular when modifying “proper value” throughout the entire chapter}proper value of persons to Adonai,}%
\verse{if\lebnote{Or “and”}your proper value is for a male\lebnote{Hebrew “the male”}from \textit{twenty years of age}\lebnote{Literally “a son of twenty years”}up to\lebnote{Or “and up to”}\textit{sixty years of age},\lebnote{Literally “a son of sixty years”}then\lebnote{Or “and”}your proper value shall be fifty shekels of money according to the sanctuary’s shekel.}%
\verse{But\lebnote{Or “And”}if it is for a female, then\lebnote{Or “and”}your proper value shall be thirty shekels.}%
\verse{And if from \textit{five years of age}\lebnote{Literally “a son of five years”}up to\lebnote{Or “and up to”}\textit{twenty years of age},\lebnote{Literally “a son of twenty years”}then\lebnote{Or “and”}your proper value shall be twenty shekels for the male and ten shekels for the female.}%
\verse{And if from \textit{a month of age}\lebnote{Literally “a son of a month”}up to\lebnote{Or “and up to”}\textit{five years of age},\lebnote{Literally “a son of five years”}then\lebnote{Or “and”}your proper value shall be five shekels of money for the male, and your proper value for the female shall be three shekels of money.}%
\verse{And if from \textit{sixty years of age}\lebnote{Literally “a son of sixty years”}and above: if a male, then\lebnote{Or “and”}your proper value shall be fifteen shekels; and for the female, ten shekels.}%
\verse{But\lebnote{Or “And”}if he is poorer than your proper value, then\lebnote{Or “and”}he shall present himself \textit{before}\lebnote{Literally “to the faces of”}the priest, and the priest shall set a value on him; the priest shall value him \textit{according to}\lebnote{Literally “on a mouth of”}what the person who made a vow \textit{can afford}.\lebnote{Literally “his hand produces”}}%
\verse{“ ‘And if it is a domestic animal from which they present an offering for Adonai, all that he gives from it for Adonai shall be a holy object.}%
\verse{He shall not replace it, nor shall he exchange it, either good with bad or bad with good; and if he indeed exchanges a domestic animal with a domestic animal, then\lebnote{Or “and”}\textit{it and its substitution shall be a holy object}.\lebnote{Literally “it shall be and its substitution shall be holy”}}%
\verse{But\lebnote{Or “And”}if it is any unclean animal from which they may not present an offering for Adonai, then\lebnote{Or “and”}he shall present the animal \textit{before}\lebnote{Literally “to the faces of”}the priest.}%
\verse{And the priest shall set a value on it, \textit{either good or bad};\lebnote{Literally “between good and between bad”}as the priest sets your proper value, so it shall be.}%
\verse{And if he indeed wants to redeem it, then\lebnote{Or “and”}he shall add a fifth of it onto your proper value.}%
\verse{“ ‘And if a man consecrates his house as a holy object for Adonai, then\lebnote{Or “and”}the priest shall set a value on it, \textit{either good or bad};\lebnote{Literally “between good and between bad”}just as the priest sets a value on it, so it shall remain.}%
\verse{But\lebnote{Or “And”}if the one who consecrates it wants to redeem his house, then\lebnote{Or “and”}he shall add a fifth of your proper value’s money onto it, and it shall be his.}%
\verse{“ ‘And if a man consecrates \textit{some of}\lebnote{Literally “from”}his property’s fields\lebnote{Collective singular; Hebrew “field”}for Adonai, then\lebnote{Or “and”}your proper value shall be \textit{in accordance with its seed requirements}:\lebnote{Literally “to the number of its seed”}a homer of barley seed for fifty shekels of money.}%
\verse{If he consecrates his field from the Year of Jubilee, it shall stand as your proper value.}%
\verse{But\lebnote{Or “And”}if he consecrates his field after the Jubilee, then\lebnote{Or “and”}the priest shall calculate the money for him \textit{according to the number of years}\lebnote{Literally “on the mouth of the years”}that are left over until the Year of Jubilee; and it shall be deducted from your proper value.}%
\verse{And if he indeed redeems the field that is consecrated, then\lebnote{Or “and”}he shall add a fifth of your proper value’s money onto it, and it shall stand for\lebnote{Or “remain for” or “belong to”}him.}%
\verse{And if he does not redeem the field and if he sells the field to another man, it may not be redeemed again,}%
\verse{and the field shall be a holy object for Adonai when it goes out\lebnote{Or “is released” or “reverts”}in the Jubilee, like a devoted\lebnote{Or “permanently set apart”; a different Hebrew word than previously translated “consecrated” in this chapter}field; \textit{it shall be the priest’s property}.\lebnote{Literally “to/for the priest it shall be his property”}}%
\verse{“ ‘And if he consecrates for Adonai his acquired\lebnote{Or “purchased”}field that is not the field of his inherited possession,}%
\verse{then\lebnote{Or “and”}the priest shall calculate for him the \textit{amount}\lebnote{Literally “number”}of your proper value until the year of the Jubilee, and he shall give your proper value on that day as a holy object for Adonai.}%
\verse{In the Year of the Jubilee the field shall return to the one who bought it from him, to the one whose property the land is.}%
\verse{And every proper value of yours shall be in the sanctuary’s shekel—the shekel shall be twenty gerahs.}%
\verse{“ ‘However, a man shall not consecrate a firstborn among livestock, which belongs as firstborn to Adonai; whether an ox \textit{or}\lebnote{Literally “whether” or “if”}small livestock, it is for Adonai.}%
\verse{And if it is among the unclean animals, then\lebnote{Or “and”}he shall ransom it according to your proper value, and he shall add a fifth of its value onto it; and if it is not redeemed, then\lebnote{Or “and”}it shall be sold according to your proper value.}%
\verse{However, anything devoted\lebnote{Or “permanently set apart”; a different Hebrew word than previously translated “consecrated” in this chapter}that a man has devoted to Adonai \textit{from all that he has},\lebnote{Literally “from all that is for him”}from human or\lebnote{Or “and”}animal, or\lebnote{Or “and”}from the field of his property, may not be sold, and it may not be redeemed; anything devoted is \textit{a most holy thing}\lebnote{Literally “a holy thing of holy things”}for Adonai.}%
\verse{Anyone devoted who is devoted from \textit{human beings}\lebnote{Literally “the human”}cannot be ransomed—he shall surely be put to death.}%
\verse{“ ‘And any tithe of the land from the land’s seed or from the fruit of the trees is for Adonai; it is a holy object for Adonai.}%
\verse{And if a man indeed redeems from his tithe, he shall add a fifth of its value onto it.}%
\verse{As for\lebnote{Or “And”}every tithe of cattle or\lebnote{Or “and”}of the flock,\lebnote{The Hebrew term refers collectively to both sheep and goats (small livestock animals)}all which crosses under the rod, the tenth shall be a holy object for Adonai.}%
\verse{He shall not inspect between the good and the bad, and he shall not exchange it; but\lebnote{Or “and”}if he indeed exchanges it, then\lebnote{Or “and”}\textit{it and its substitution shall be}\lebnote{Literally “it shall be and its substitution shall be”}a holy object—it shall not be redeemed.’ ”}%
\verse{These are the commands that Adonai commanded Moses for the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}on \textit{Mount Sinai}.\lebnote{Literally “the mountain of Sinai”}}%
\end{biblechapter}

