\biblebook{Malachi}

\begin{biblechapter} % Malachi 1
\verse{\textit{An oracle}.\lebnote{“A burden”} The word of Adonai to Israel \textit{through}\lnAXK{} Malachi.\lebnote{The name “Malachi” means “my messenger” in Hebrew}}%
\verseWithHeading{An Oracle Against Edom}{“I have loved you,” says Adonai, but you say, “How have you loved us?” “Is Esau not Jacob’s brother?” \textit{declares}\lnDJI{} Adonai. “I have loved Jacob,}%
\verse{but Esau I have hated. I have made his\lebnote{That is, “Esau’s”} mountain ranges a desolation, and given his inheritance to the jackals of the desert.”}%
\verse{If Edom says, “We are shattered, but we will return and rebuild the ruins,” Adonai of hosts says this: “They may build, but I will tear down; and they will be called a territory of wickedness, and the people with whom Adonai is angry forever.”}%
\verse{Your eyes will see this, and you will say, “Adonai is great beyond the borders\lebnote{Hebrew “boundary”} of Israel.”}%
\verseWithHeading{An Oracle Against the Temple Priesthood}{“A son honors his father, and a slave his master;\lnIEG{} but if I am a father, where is my honor, and if I am a master,\lnIEG{} where is my reverence?” says Adonai of hosts to you, O priests, \textit{who despise}\lebnote{“despisers of”} my name. “But you say, ‘How have we despised your name?’}%
\verse{You are presenting defiled food on my altar! But you ask, ‘How have we defiled you?’ By saying that the table of Adonai is despised!}%
\verse{When you offer a blind animal for sacrifice, is that not wrong?\lnMOA{} And when you offer the lame and the one who is ill, is that not wrong?\lnMOA{} Present it, please, to your governor! Will he be pleased with you? \textit{Will he show you favor}?”\lebnote{“Will he lift your face?”} says Adonai of hosts.}%
\verse{So then, \textit{implore the favor of God}\lebnote{“appease, please, the face of God”} so that he will be gracious to us. “\textit{This is what you have done}.\lebnote{“From your hand was this”} \textit{Will he show favor to any of you}?”\lebnote{“will he lift up from you a face”} says Adonai of hosts.}%
\verse{“Who also among you will shut the temple doors so that you will not kindle fire in vain on my altar? I take no pleasure in you,” says Adonai of hosts, “and \textit{I will not accept}\lebnote{“I will not take pleasure in”} an offering from your hand.}%
\verse{From the rising of the sun to its setting, my name is great among the nations, and in every place incense is being presented to my name, and a pure offering. For my name is great among the nations,” says Adonai of hosts.}%
\verse{“But you are profaning it by saying the table of the Lord is defiled, and its fruit—its food\lebnote{The Hebrew here is difficult; “its food” may function to explain “its fruit”}—is despised!}%
\verse{And you say, ‘Look! This is a weariness,’ and you sniff with disdain at it,” says Adonai of hosts. “And you bring the stolen, the lame, and the one that is sick—this you bring as the offering! Should I \textit{accept it}\lebnote{“take pleasure in it”} from your hand?” says Adonai.}%
\verse{“Cursed is the one who cheats, who has in his flock a male and vows it, but instead sacrifices a blemished one to the Lord! For I am a great king,” says Adonai of hosts, “and my name is awesome among the nations.”}%
\end{biblechapter}

\begin{biblechapter} % Malachi 2
\verse{“And so then, O priests, this command is for you:}%
\verse{If you will not listen, and if you will not \textit{take it to heart}\lebnote{“place it on your heart”} to give glory to my name,” says Adonai of hosts, “then I will send the curse on you, and I will curse your blessings; moreover I have already cursed them because \textit{you are not taking it to heart}.\lebnote{“you are not placing it on your heart”}}%
\verse{Look! I am going to rebuke \textit{your offspring},\lebnote{“for you the offspring”} and I will scatter offal on your faces, the offal of your religious feasts, and you will be carried to it.\lebnote{The Hebrew is difficult. The subject is indefinite, and “it” likely refers to a refuse or waste pile}}%
\verse{“Then you will know that I have sent to you this command, that my covenant with Levi \textit{continues},”\lebnote{“be”} says Adonai of hosts.}%
\verse{“My covenant with him was one of life and peace, and I gave them to him. This required reverence, and he revered\lebnote{Or “feared”} me and stood in awe \textit{before}\lnIQE{} my name.}%
\verse{The instruction of truth\lebnote{Or “law of truth”} was in his mouth, and wickedness was not found on his lips. In peace and in uprightness\lebnote{Or “rectitude,” or “fairness”} he walked with me, and he brought back many from sin.\lnKUX{}}%
\verse{For the lips of the priest should guard knowledge, and they should seek instruction\lebnote{Or “law”} from his mouth, for he is the messenger of Adonai of hosts.}%
\verse{But you, you have turned from the way; you have caused many to stumble by your instruction;\lebnote{Or “the law”} you have ruined the covenant of Levi,” says Adonai of hosts.}%
\verse{“And in turn I have made you despised and humbled before all the people, inasmuch as you are not keeping\lebnote{Or “observing”} my way, but are \textit{favoring your own instruction}.”\lebnote{“lifting faces in the instruction/law”}}%
\verseWithHeading{Adonai Hates Faithlessness}{\textit{Do we not all have}\lebnote{“Is for all of us not”} one father? Has not one God created us? Why then are we unfaithful \textit{to one another},\lebnote{“each man to his brother”} profaning the covenant of our ancestors?\lnAGP{}}%
\verse{Judah has been faithless, and a detestable thing has been done in Israel and in Jerusalem. Judah has profaned the sanctuary of Adonai which he loves, and has married the daughter of a foreign\lebnote{Or “pagan”} god.}%
\verse{May Adonai cut off from the tents of Jacob the man who does this, the one who is awake and who answers, and the one who brings an offering to Adonai of hosts!}%
\verse{And this you do \textit{as well}:\lebnote{“a second time”} you cover the altar of Adonai with tears, with weeping and with groaning, because he\lnAHK{} \textit{no longer regards the offering}\lebnote{“there is no longer turning to the offering”} or accepts it with favor from your hand.}%
\verse{But you ask, “\textit{For what reason}?”\lebnote{“Because of what?”} Because Adonai stands as a witness between you and the wife of your youth, against whom you have been unfaithful, even though she is your marriage partner and \textit{your wife by covenant}.\lebnote{“the wife of your covenant”}}%
\verse{\textit{Did not one God make them?}\lebnote{“Not one did he make”} But a remnant of the spirit is his.\lebnote{The meaning of this sentence is difficult; literally “And a remnant of the spirit to him”} And what does the one God desire? An offspring of God.\lebnote{Or “a godly offspring”} \textit{You must be attentive to}\lnMOG{} your spirit, and you must not be unfaithful to the wife of your youth.}%
\verse{“For I hate divorce,”\lebnote{Hebrew “sending away”} says Adonai, the God of Israel, “and he who covers his clothing with violence,” says Adonai of hosts. “\textit{You must be attentive to}\lnMOG{} your spirit and you must not be unfaithful.”}%
\verse{You have wearied Adonai with your words, but you ask, “With what have we wearied him?” You have done so \textit{when you say},\lebnote{“at your saying”} “Everyone who does evil is good in the \textit{sight}\lebnote{“eyes”} of Adonai, and in them he delights,” or “Where is the God of justice?”}%
\end{biblechapter}

\begin{biblechapter} % Malachi 3
\verseWithHeading{Adonai’s Messenger is Coming}{“Look! I am going to send my messenger,\lebnote{Or “my angel”} and he will prepare\lebnote{Or “clear”} the way \textit{before me}.\lebnote{“before my face”} And the Lord whom you are seeking will come suddenly to his temple, and the messenger of the covenant, in whom you are taking pleasure—look!—he is about to come,” says Adonai of hosts.}%
\verse{And who can endure the day of his coming? And who is the one who can stand \textit{when he appears}?\lebnote{“at his appearing”} For he is like a refiner’s fire, like launderers’ alkali.\lebnote{Or “lye,” or “soap”}}%
\verse{He will sit as a refiner and purifier of silver; he will purify the children of Levi, and he will refine them like gold and like silver, and they will present to Adonai offerings\lnFAR{} in righteousness.}%
\verse{And the offering of Judah and Jerusalem will be pleasing to Adonai, like in the days of old and like in former years.}%
\verse{“Then I will approach you for judgment, and I will be a swift witness against the sorcerers and against the adulterers, and against those who swear \textit{falsely},\lebnote{“as far as the falseness”} and against the oppressors of the hired worker with his wages, the widow and the orphan, and \textit{the abusers of}\lebnote{“those pushing aside”} the alien,\lebnote{Or “sojourner”} and yet do not fear me,” says Adonai of hosts.}%
\verseWithHeading{An Overflowing Blessing}{“For I, Adonai, have not changed,\lebnote{Or “I do not change”} and you, O children of Jacob, have not perished.\lebnote{Or “are not destroyed”}}%
\verse{From the days of your ancestors\lnAGP{} you have turned aside from my rules, and have not kept\lebnote{Or “observed”} them! Return to me and I will return to you,” says Adonai of hosts, “but you say, ‘How shall we return?’}%
\verse{Will a human dare to rob God? Yet you are robbing me! And you say, ‘How have we robbed you?’ In the tithes and the contributions!}%
\verse{You are being cursed with a curse, for you are robbing me, the whole nation of you!}%
\verse{Bring the whole tithe to \textit{the storehouse},\lebnote{Hebrew “the house of the treasure”} so that there will be food in my house, and test me please in this,” says Adonai of hosts, “if I will not open for you the windows of heaven, and pour forth for you \textit{an overflowing blessing}.\lebnote{“a blessing up to not enough”}}%
\verse{I will rebuke the devourer\lebnote{NRSV, JPS interpret this word as “locust”} for you; it will not destroy the fruit of your soil; your vine in the field will not be unfruitful,” says Adonai of hosts.}%
\verse{“And all the nations will call you blessed, because you will be \textit{a land in which one takes joy},”\lebnote{“a land of desire”} says Adonai of hosts.}%
\verse{“Your words have been harsh\lebnote{Hebrew “strong”} against me,” says Adonai. “But you say, ‘How have we spoken together against you?’}%
\verse{You have said, ‘It is useless\lebnote{Or “vain”} to serve God! What is the gain if we keep\lebnote{Or “observe”} his requirements, and if we walk as mourners \textit{before}\lnIQE{} Adonai of hosts?}%
\verse{And now we are calling the arrogant blessed! Not only do those who do wickedness prosper; they also test God and they escape!’ ”}%
\verseWithHeading{A Book of Remembrance is Written}{Then those who revered\lnMOL{} Adonai spoke \textit{with one another}.\lebnote{“each with his friend/companion”} And Adonai listened attentively\lebnote{Or “gave heed”} and heard, and a scroll of remembrance was written \textit{before him}\lebnote{“to the face of him”} of those who revere\lnMOL{} Adonai and ponder his name.}%
\verse{“They will be mine,” says Adonai of hosts, “on the day that I am acting, my treasured possession. I will have compassion on them as a man has compassion on his son who serves him.}%
\verse{You will return and see the difference between the righteous and the wicked, between the one who serves God and the one who does not serve him.}%
\end{biblechapter}

\begin{biblechapter} % Malachi 4
\verse{\lebnote{Malachi 4:1–6 in the English Bible is 3:19–24 in the Hebrew Bible} “For look! The day is about to come, burning like an oven, and all the arrogant and every \textit{evildoer}\lebnote{“doer of wickedness”} will be stubble. The coming day will consume\lebnote{Or “devour”} them,” says Adonai of hosts. “It will not leave behind for them root or branch.}%
\verse{But for you who revere\lnMOL{} my name, the sun of righteousness will rise, with healing in its wings, and you will go out and leap like \textit{fattened calves}.\lebnote{“bull-calves of the animal stall” (on the idiom see HALOT 631 s.v.)}}%
\verse{You will trample down the wicked, for they will be ashes under the soles of your feet on the day that I am going to act,” says Adonai of hosts.}%
\verse{“Remember the instruction\lebnote{Or “law”; Hebrew \textit{torah}} of my servant Moses, which I commanded him at Horeb to all Israel, the rules and regulations.}%
\verse{Look! I am going to send to you Elijah the prophet before the coming of the great and awesome day of Adonai!}%
\verse{And he will bring back\lebnote{Or “And he will restore”} the hearts of the fathers to the sons, and the hearts of the sons to their fathers, so that I will not come and strike the land with a ban.”}%
\end{biblechapter}

