\biblebook{Micah}

\begin{biblechapter} % Micah 1
\verseWithHeading{The Impending Doom of Samaria and Jerusalem} The word of Adonai that came to Micah the Morashtite in the days of Jotham, Ahaz, and Hezekiah, the kings of Judah, that he saw concerning Samaria and Jerusalem:
\verse Hear, all you peoples; 
give heed, O earth and its fullness. 
And let the Lord Adonai be against you as a witness, 
the Lord from \textit{his holy temple}.\lebnote{Literally “the temple of his holiness”}
\verse For behold, Adonai is coming out from his place, 
and he will come down and tread upon the high places of the earth.
\verse The mountains will melt under him 
and the valleys will burst open, 
like wax before the fire, 
like water rushing down a slope.
\verse All this is for the rebellion of Jacob, 
and for the sins of the house of Israel. 
What is the rebellion of Jacob? 
Is it not Samaria? 
And what are the high places of Judah? 
Are they not Jerusalem?
\verse So I will make Samaria as a heap of rubble in the field, 
a place for planting a vineyard. 
And I will pour down her stones into the valley 
and uncover her foundations.
\verse Then all her idols will be broken in pieces, 
and all her prostitution wages will be burned in the fire, 
and all her idols I will make a desolation, 
For from the wage of a prostitute she gathered them, 
and to the wage of a prostitute they will return.
\verse On account of this I will lament and wail. 
I will go about barefoot and naked. 
I will make a lamentation like the jackals, 
and a mourning ceremony like the \textit{ostriches}.\lebnote{Literally “daughters of an ostrich”}
\verse For her wounds are incurable, 
because it has come to Judah. 
It has reached to the gate of my people, to Jerusalem.
\verse Do not tell it in Gath; 
you must not weep at all. 
In Beth-le-Aphrah 
roll yourself in dust.
\verse Pass on, you inhabitants of Shaphir, 
in naked shame. 
The inhabitants of Zaanan do not come forth; 
the wailing of Beth-ha-Ezel will take its protection from you.
\verse For the inhabitants of Maroth 
writhed for good, 
because disaster has come down from Adonai 
to the gate of Jerusalem.
\verse Harness the chariot to the team of horses, 
O inhabitants of Lachish; 
it is the beginning of sin 
for the daughter of Zion, 
for the transgressions of Israel 
were found in you.
\verse Therefore you will give parting gifts 
to Moresheth-Gath; 
the houses of Achzib will be a deception 
to the kings of Israel.
\verse I will again bring the conqueror upon you, 
O inhabitants of Mareshah; 
the glory of Israel 
will come to Adullam.
\verse Make yourselves bald 
and cut off your hair 
for the children of your pleasure. 
Expand your baldness as the eagle, 
for they will go into exile away from you.
\end{biblechapter}

\begin{biblechapter} % Micah 2
\verseWithHeading{Woe to the Oppressors of God’s People} Woe to those who plan wickedness 
and evil deeds upon their beds! 
In the light of the morning they did it, 
because they have power in their hands.
\verse They covet fields and seize them, 
and houses, and they take them away. 
They oppress a man and his house; 
a man and his inheritance.
\verse Therefore, thus says Adonai: Look! I am planning disaster against this family from which you will not be able to remove your necks. You will not walk proudly, for it is a time of disaster.
\verse In that day they will raise a proverb against you, 
and will wail a bitter wailing, saying, 
“We are utterly ruined; 
he exchanges the portion of my people. 
How he removes it from me; 
he apportions our field to an apostate.”
\verse Therefore you will have no one casting a line by lot 
in the assembly of Adonai.
\verse “Do not preach!” they are preaching. 
They should not preach to these; 
disgrace will not overtake us.
\verse Should this be said, O house of Jacob? 
Is the patience\lebnote{Or “spirit”} of Adonai shortened? 
Are these his deeds? 
Are my words not pleasing 
to him who walks upright?
\verse But \textit{recently}\lebnote{Literally “yesterday”} my people have risen up as an enemy; 
from before the cloak you strip off the robe 
from those passing by in confidence, 
returning from war.
\verse You have driven out the women of my people 
from the houses of their\lebnote{Hebrew “her”} pleasure. 
From their\lebnote{Hebrew “her”} children you have taken away 
my glory forever.
\verse Arise and go, 
for this is no resting place, 
on account of uncleanness that destroys 
with painful destruction.
\verse If a man walks about in a spirit of deception and lies, 
saying “I will preach to you of wine and strong drink,” 
then he would be a preacher for this people!
\verse I will surely gather all of you, O Jacob; 
I will surely gather together the remainder of Israel. 
I will set them like sheep of Bozrah. 
Like a herd in the midst of their pasture 
they will be in tumult from people.
\verse The one who breaks out before them goes up; 
they break through and pass the gate, 
going out through it. 
Their king passes before them, 
Adonai at their head.
\end{biblechapter}

\begin{biblechapter} % Micah 3
\verseWithHeading{Israel’s Leaders Denounced} And I said, 
“Listen, O heads of Jacob 
and leaders of the house of Israel! 
Is it not for you to know justice?
\verse Those of you who are haters of good 
and lovers of evil, 
who tear their skin from them 
and their flesh from their bones,
\verse and who eat the flesh of my people, 
and strip their skin from them, 
and break their bones, 
and chop them like meat in the pot, 
and like flesh in the midst of a cooking pot.”
\verse Then they will cry to Adonai, 
but he will not answer them. 
He will hide his face from them at that time, 
when they have made their deeds evil.
\verse Thus says Adonai concerning the prophets 
who are leading my people astray, 
who are biting with their teeth 
and proclaim, “Peace,” 
but whoever puts nothing into their mouths 
they declare war against.
\verse Therefore it will be as night to you, without vision, 
and darkness to you, without divination. 
And the sun will set on the prophets, 
and the day will grow dark over them.
\verse Then the seers will be disgraced, 
and those practicing divination will be put to shame. 
All of them will cover their lips, 
for there is no reply from God.
\verse But I, I am filled with power, 
with the Spirit of Adonai, 
and with justice and might, 
to declare to Jacob his rebellion, 
and to Israel his sin.
\verse Hear this, O rulers of the house of Jacob 
and leaders of the house of Israel, 
those detesting justice 
and perverting all that is right,
\verse he who builds Zion with blood 
and Jerusalem with wickedness.
\verse Its rulers judge for a bribe; 
its priests teach for a price; 
its prophets practice divination for money. 
But they lean on Adonai, saying, 
“Is not Adonai in our midst? 
Disaster will not come upon us.”
\verse Therefore on account of you 
Zion will be plowed as a field, 
and Jerusalem will be a heap of rubble, 
and \textit{the temple mount}\lebnote{Literally “the mountain of the house”} as a high place in a forest.
\end{biblechapter}

\begin{biblechapter} % Micah 4
\verseWithHeading{The Mountain of Adonai} And it will be that at the end of those days, 
the mountain of Adonai 
will be established as the highest of the mountains, 
and it will be lifted up above the hills, 
and people will stream to it.
\verse And many nations will come and say, 
“Come! Let us go up to the mountain of Adonai, 
and to the \textit{temple}\lebnote{Literally “house”} of the God of Jacob, 
that he may teach us his ways, 
and that we may walk in his paths,” 
for the law will go out from Zion, 
and the word of Adonai from Jerusalem.
\verse And he will judge between many peoples 
and will arbitrate for strong nations far away; 
and they will beat their swords into plowshares 
and their spears into pruning hooks; 
nation will not lift up a sword against a nation, 
and they will no longer learn war.
\verse But they will sit, each under his vine 
and under his fig tree, 
and no one will make them afraid, 
for the mouth of Adonai of hosts has spoken.
\verse For all the nations walk, 
each in the name of its god, 
but we will walk in the name of Adonai our God, 
forever and ever.
\verseWithHeading{Adonai Will Deliver Zion} “In that day,” \textit{declares Adonai,}\lebnote{Literally “a declaration of Adonai”} 
“I will assemble the one who limps, 
and I will gather the one who has been scattered, 
and those whom I have mistreated.
\verse And I will make the one who limps a remnant, 
and the one driven far away a strong nation, 
and Adonai will reign over them on Mount Zion 
from now to forever.
\verse And you, O Migdal-Eder,\lebnote{Or “tower of the flock”} 
hill of the daughter of Zion, 
to you it will come, 
and the former dominion will come, 
the reign of the daughter of Jerusalem.
\verse So then, why do you shout a loud shout? 
Is there no king in you? 
Has your counselor perished 
that pangs like a woman in labor have seized you?
\verse Writhe and groan, O daughter of Zion, 
like a woman in labor. 
For now you will go forth from the city, 
and you will camp in the field; 
you will go to Babylon. 
There you will be rescued; 
there Adonai will redeem you 
from the hand of your enemies.
\verse And now, many nations are gathered against you 
who are saying, ‘Let her be defiled, 
and let our eyes gaze upon Zion.’
\verse But they do not know the thoughts of Adonai, 
and they do not understand his plan, 
that he has gathered them as sheaves 
to his threshing floor.
\verse Arise and thresh, O daughter of Zion, 
for your horn I will make as iron 
and your hooves as bronze. 
And you will break many peoples in pieces, 
and their gain you will devote to destruction to Adonai, 
and their wealth to the Lord of the whole earth.”
\end{biblechapter}

\begin{biblechapter} % Micah 5
\verseWithHeading{A Ruler from Bethlehem} \lebnote{Micah 5:1–15 in the English Bible is 4:14–5:14 in the Hebrew Bible} Now muster troops, O daughter of troops; 
a siege he puts against us. 
They strike the ruler of Israel 
with a rod on the cheek.
\verse But you, O Bethlehem Ephrathah, 
too small to be among the clans of Judah, 
from you one will go out for me, 
to be ruler in Israel; 
and his origins are from of old, 
from ancient days.
\verse Therefore he will give them up 
until the time of she who is with child has given birth. 
And the rest of his brothers will return 
to the children of Israel.
\verse And he will stand and shepherd his flock 
in the strength of Adonai, 
in the majesty of the name of Adonai his God. 
And they will live, 
for now he will be great unto the ends of the earth.
\verse And this one will be peace. 
As for the Assyrian,\lebnote{Hebrew “Assyria”} when he comes into our land 
and when he treads on our fortresses, 
then we will raise up against him seven shepherds 
and eight leaders of men.
\verse And they will shepherd the land of Assyria with the sword, 
and the land of Nimrod at its entrances. 
And he will rescue us from the Assyrian\lebnote{Hebrew “Assyria”} 
when he comes into our land, 
and when they tread upon our border.
\verseWithHeading{A Remnant Delivered} And the remnant of Jacob will be 
in the midst of many nations, 
like dew from Adonai, 
like showers upon the grass 
which does not wait for a man, 
nor delays for the children of humankind.
\verse And the remnant of Jacob will be among the nations, 
in the midst of many peoples, 
like a lion among the wild animals of the forest, 
like a young lion among the herds of sheep 
which, when it passes through and treads down 
and tears in pieces, there is none who can deliver.
\verse Your hand will be lifted high over your enemies, 
and all your foes will be cut off.
\verse “And it will happen that in that day,” \textit{declares}\lebnote{Literally “a declaration of”} Adonai, 
“then I will cut off your horses from among you, 
and I will destroy your chariots.
\verse And I will cut off the cities of your land, 
and I will demolish all your fortifications.
\verse And I will cut off sorceries from your hand, 
and you will not have soothsayers.
\verse And I will cut off your idols 
and your stone pillars from among you, 
and you will no longer bow down 
to the work of your hands.
\verse And I will uproot your Asherahs from among you, 
and I will destroy your cities.
\verse And in anger and in wrath I will execute vengeance 
on the nations who did not obey.”
\end{biblechapter}

\begin{biblechapter} % Micah 6
\verseWithHeading{Adonai Brings a Charge against Israel} Hear now what Adonai says: 
“Arise! Plead your case with the mountains, 
and let the hills hear your voice.”
\verse Hear, O mountains, the indictment of Adonai, 
and you eternal foundations of the earth, 
for Adonai has an indictment against his people, 
and against Israel he contends.
\verse “O my people, what have I done to you, 
and how have I wearied you? Answer me!
\verse For I brought you up from the land of Egypt, 
and from the house of slavery I redeemed you. 
And I sent Moses, Aaron, and Miriam before you.
\verse O my people, remember what Balak the king of Moab devised, 
and what Balaam the son of Beor answered him, 
and what happened from Shittim to Gilgal, 
so that you might know the righteous acts of Adonai.”
\verseWithHeading{What Does Adonai Require?} With what shall I approach Adonai, 
and bow down to God on high? 
Shall I approach him with burnt offerings, 
with bull calves a year old?
\verse Will Adonai be pleased with thousands of rams, 
with myriads of rivers of oil? 
Shall I give my firstborn for my transgression, 
the fruit of my body for the sin of my soul?
\verse He has told you, O mortal, what is good, 
and what does Adonai ask from you 
but to do justice, and to love kindness, 
and to walk humbly with your God?
\verseWithHeading{The Destruction of the Wicked} The voice of Adonai calls to the city 
(it is sound judgment to fear your name): 
“Hear, O staff!\lebnote{Or “tribe”} Now who has appointed it?
\verse Is there any longer a man in the house of the wicked 
treasures of wickedness, 
and the ephah of scarcity which is accursed?\lebnote{The meaning of the Hebrew text is uncertain}
\verse Shall I regard as pure the one with scales of wickedness, 
and with a bag of deceitful weights?
\verse Because her rich are full of violence, 
and her inhabitants speak lies, 
and their tongue is deceitful in their mouth.
\verse And I also \textit{have made you sick by striking you down},\lebnote{Literally “I have made sick to strike you down”} 
making you desolate because of your sins.
\verse You yourself will eat but not be satisfied; 
your hunger will be in your midst, 
and you will put away, but you will not save, 
and what you save I will hand over to the sword.
\verse You yourself will sow, but you will not reap; 
you will tread olives, but you will not anoint yourself with oil; 
you will tread grapes, but you will not drink wine.
\verse For you have observed the regulations of Omri 
and all the works of the house of Ahab. 
And you have walked in their counsels, 
so that I am making you a desolation 
and your inhabitants an object of scorn. 
So you will bear the scorn of my people.
\end{biblechapter}

\begin{biblechapter} % Micah 7
\verseWithHeading{Wait Upon Adonai} Woe is me! For I have become 
like the gatherings of summer, 
like the gleanings of the grape harvest, 
when there is no cluster of grapes to eat 
or early ripened fruit that my soul desires.
\verse The faithful person has perished from the land, 
and there is none who is upright among humankind. 
All of them lie in wait; 
each hunts his brother with a net.
\verse Their hands are upon evil, to do it well; 
the official and the judge ask for the bribe, 
and the great man utters the evil desire of his soul; 
and they weave it together.
\verse The best of them is like a brier; 
the most upright worse than a thorn hedge. 
The day of your watchman, your punishment, has come; 
now their confusion will come.
\verse Do not put faith in a friend; 
put no trust in a close friend. 
Guard the doorways of your mouth 
from the one who lies in your lap.
\verse For a son treats a father with contempt; 
a daughter rises up against her mother, 
a daughter-in-law against her mother-in-law; 
the enemies of a man are the men of his own house.
\verse But as for me, I will look to Adonai; 
I will wait for the God of my salvation. 
My God will hear me.
\verse You should not rejoice over me, O my enemy! 
When I fall I will stand up; 
when I sit in darkness 
Adonai will be a light for me.
\verse I will bear the rage of Adonai, 
for I have sinned against him, 
until he pleads my cause 
and executes my justice. 
He will bring me out to the light; 
I will see his righteousness.
\verse Then my enemy will see, 
and shame will cover her who said to me, 
“Where is Adonai your God?” 
My eyes will look upon her; 
now she will become a trampling place, 
like mud in the streets.
\verse A day for building your walls; 
on that day he will extend your boundary.
\verse On that day he will come to you 
from Assyria and the cities of Egypt, 
and from Egypt to the River, 
and from sea to sea 
and mountain to mountain.
\verse But the earth will be a desolation 
because of its inhabitants, 
for the fruit of their deeds.
\verse Shepherd your people with your staff, 
the flock of your inheritance, 
those dwelling alone in a forest 
in the midst of Carmel. 
Let them feed in Bashan and Gilead, 
as in the days of old.
\verse As in the days when you came out 
from the land of Egypt 
I will show him wondrous things.
\verse The nations will see and be ashamed 
because of all their might. 
They will lay the hand on the mouth; 
their ears will be deaf.
\verse They will lick the dust like the serpent, 
like the crawling things of the earth. 
They will come trembling from their strongholds 
to Adonai our God. 
Let them fear and be afraid of you.
\verse Who is a God like you, forgiving sin 
and passing over rebellion for the remnant of his inheritance? 
He does not retain his anger forever, 
for he delights in loyal love.
\verse \textit{He will again have compassion}\lebnote{Literally “He will return, he will have compassion”} on us; 
he will trample our iniquities. 
And you will hurl all their sins 
in the depths of the sea.
\verse You will show faithfulness to Jacob, 
and loyal love to Abraham, 
as you have sworn to our ancestors\lebnote{Or “fathers”} 
from the days of old.
\end{biblechapter}

