\biblebook{Nahum}

\begin{biblechapter} % Nahum 1
\verseWithHeading{Superscription} The oracle concerning Nineveh— 
the scroll of the vision of Nahum the Elkoshite:
\verseWithHeading{Half-Acrostic: Adonai Takes Vengeance against His Enemies} Adonai is a jealous God and avenging; 
Adonai is avenging and \textit{full of wrath}. 
Adonai takes vengeance against his enemies; 
he rages against his adversaries.
\verse Adonai is slow to anger but great in power; 
he will certainly not allow the guilty to go unpunished. 
\textit{He marches} in storm wind and in gale; 
storm clouds are the dust of his feet.
\verse He rebukes the sea and he dries up the rivers; 
he makes all the rivers run dry. 
Bashan and Carmel wither; 
the blossom of Lebanon languishes.
\verse Mountains quake before him; 
the hills shake apart. 
The earth heaves before him— 
the world and all her inhabitants.
\verse His indignation—who can stand before it? 
Who can endure \textit{his fierce anger}? 
His wrath is poured out like fire; 
rocks are shattered before him.
\verse Adonai is good—a refuge in the day of distress; 
he knows those who take refuge in him.
\verse But with a rushing torrent he will bring Nineveh to an end; 
he will chase his enemies into darkness.
\verseWithHeading{Adonai Will Destroy Nineveh} What do you plot against Adonai? 
\textit{He will completely destroy it}; 
trouble will not rise up a second time!
\verse For like entangled thorns, 
and like their drink which is drunk, 
they will be consumed like fully dry chaff.
\verse From you goes out one who plots evil against Adonai, 
one who plans wickedness.
\verseWithHeading{Adonai Will Deliver Judah and Destroy Nineveh} Adonai says this:
\verse “Even though they are powerful and likewise many, 
even so they will be cut off and pass away. 
Though I have afflicted you, 
I will afflict you no longer.
\verse Adonai has commanded concerning you:
\verse “Your name will no longer be sown. 
I will cut off \textit{the idols and images} from the \textit{temple} of your gods; 
\textit{I will send you to the grave} because you are worthless.”
\end{biblechapter}

\begin{biblechapter} % Nahum 2
\verseWithHeading{The Divine Warrior Will Attack Nineveh} \textit{One who shatters} has come up against you! 
Guard the fortification! 
Watch the road! 
Gird your loins! 
\textit{Muster} all your strength!
\verse For Adonai will restore the majesty of Jacob 
like the majesty of Israel; 
for ravagers have ravaged them 
and ruined their branches.
\verseWithHeading{The Enemy Army Conquers the City of Nineveh} The shields of his warriors are dyed red; 
The powerful men are dressed in scarlet. 
The metal of the chariots shines like fire \textit{on the day of battle}, 
and their spears quiver.
\verse The chariots race madly through the streets; 
they rush back and forth in the public squares. 
Their appearance like lightning bolts, 
they dart about like flashes of lightning.
\verse He calls his officers; 
they stumble as they march; 
they rush to her wall; 
they set the covering in place.
\verse The gates of the river are opened; 
the palace trembles.
\verse \textit{Her goddess is taken out and taken into exile}; 
her maidservants moan like doves; 
they beat on their breasts.
\verse Nineveh is like a pool of water without its water. 
\textit{As they flee}, she cries, “Stop! Stop!” 
But there is no one who turns back.
\verse Plunder the silver! Plunder the gold! 
There is no end to the spoils, 
an abundance of \textit{everything one could want}!
\verse Emptiness and plundering and devastation! 
Their hearts faint and their knees tremble, 
\textit{All their loins shake} and all their faces \textit{turn} pale.
\verseWithHeading{The Divine Warrior Will Hunt down the Mighty Lions} Where now is the den of the lions 
and the cave of the fierce lions? 
There the lioness, the cub, and the lion once \textit{prowled}, 
\textit{and no one disturbed them}.
\verse The lion tore apart enough prey for his cubs, 
he strangled prey for his lioness; 
he filled \textit{his lair} with prey 
and his den with mangled carcass.
\verse “Look! I am against you!” \textit{declares} Adonai of hosts. 
“I will burn her chariots \textit{with fire}; 
the sword will devour fierce lions. 
I will cut off your prey from the earth; 
the voice of your messengers will no longer be heard.”
\end{biblechapter}

\begin{biblechapter} % Nahum 3
\verseWithHeading{The Doom of the Wicked City is Certain} Woe to the city that has shed \textit{much blood}! 
\textit{She is a deceiver}, 
She is filled with plunder, 
\textit{She has hoarded her spoils of war}.
\verse The crack of the whip! 
The rumbling of the chariot wheel! 
The galloping of the horse! 
The racing of the chariot!
\verse Chariots charge! 
Swords flash! 
Spears glitter! 
\textit{Many corpses are piled high}! 
There is no end to the slain! 
They stumble over their dead!
\verseWithHeading{Harlot Nineveh Will Get What She Deserves} “Because of the many idolatries of the prostitute, 
the beautiful charm of a mistress of witchcraft, 
she who enslaves nations by her harlotries, 
and peoples by her sorceries,
\verse “Look! I am against you!” \textit{declares}Adonai. 
“I will strip up your skirts over your face; 
I will let nations look at your nakedness and your shame.
\verse “I will throw filth upon you, 
I will treat you with contempt, 
I will make you a spectacle.
\verse “And it will be that everyone who sees you will flee from you, 
And \textit{they} will say, ‘Nineveh is destroyed!’ 
Who will mourn for her? 
From whence shall I seek comforters for you?”
\verseWithHeading{Nineveh Will Suffer the Same Fate She Inflicted on Thebes} Are you better than Thebes? 
She who sits at the Nile, surrounded by her waters, 
her rampart was the sea and water was her wall.
\verse Cush was her strength, Egypt without end; 
Put and Libya were among \textit{your allies}.
\verse Yet she went into captivity as an exile; 
her children were dashed to pieces at the head of the streets; 
they cast lots for all of her nobles; 
all of her dignitaries were bound with chains.
\verse You also will behave like a drunkard; 
You also will hide yourself; 
You also will seek refuge from the enemy.
\verseWithHeading{Nineveh’s Defenses Will Fail} All of your fortifications are like fig trees with ripe firstfruits— 
if they are shaken, they will fall into the mouth of the eater.
\verse Look! Your troops are like women in your midst. 
The gates of your land are wide open to your enemies; 
fire will consume the bars of your gates.
\verse Draw water for a siege! 
Strengthen your fortifications! 
Go to the mud pit! 
Trample the clay! 
Grasp the brick mold!
\verse There fire will consume you; 
the sword will cut you off. 
It will consume you like the locust.
\verseWithHeading{Nineveh’s Defenders Will Flee} Multiply yourself like the locust! 
Multiply yourself like the grasshopper!
\verse You have increased your merchants more than the stars of heaven; 
like the locust they will shed their skin and fly away.
\verseWithHeading{Nineveh Will Come to an End} Your shepherds are sleeping, O king of Assyria! 
Your nobles slumber. 
Your people are scattered on the mountains; 
no one can gather them.
\verse There is no healing for your wound; 
your injury is fatal. 
All who hear the report of you will clap their hands for joy concerning you, 
\textit{For who has not suffered at the hands of your endless cruelty}?
\end{biblechapter}

