\biblebook{Numbers}

\begin{biblechapter} % Numbers 1
\verseWithHeading{God Commands Moses to Take a Census} Adonai spoke to Moses in the desert of Sinai, in the tent of assembly, on the first of the month, in the second year \textit{after they came out}\lebnote{Literally “of their coming out”} of the land of Egypt, saying,
\verse “\textit{Take a census of}\lebnote{Literally “Lift up the head of”} the entire community of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} according to their clans and \textit{their families},\lebnote{Literally “the house of their fathers”} according to the number of names, every male individually
\verse from \textit{twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone in Israel who is able to go to war. You and Aaron must muster them\lebnote{Or “count them,” or “summon them,” or “enroll them”} for their wars.
\verse A man from each tribe will be with you, each man the head of \textit{his family}.\lebnote{Literally “the house of his father”}
\verse And these are the names of the men who will \textit{assist you}:\lebnote{Literally “stand with you”} from Reuben, Elizur son of Shedeur;
\verse from Simeon, Shelumiel son of Zurishaddai;
\verse from Judah, Nahshon son of Amminadab;
\verse from Issachar, Nethanel son of Zuar;
\verse from Zebulun, Eliab son of Helon.
\verse From the descendants of Joseph: from Ephraim, Elishama son of Ammihud; from Manasseh, Gamaliel son of Pedahzur.
\verse From Benjamin, Abidan son of Gideoni;
\verse from Dan, Ahiezer son of Ammishaddai;
\verse from Asher, Pagiel son of Ocran;
\verse from Gad, Eliasaph son of Deuel;
\verse and from Naphtali, Ahira son of Enan.”
\verse These are the ones summoned from the community, the leaders of their ancestors’\lebnote{Or “fathers’ ”} tribes; they are the heads of Israel’s clans.
\verse So Moses and Aaron took these men who had been designated by name,
\verse and they summoned the entire community on the first day of the second month. And they registered themselves among their clans according to \textit{their families},\lebnote{Literally “the house of their fathers”} according to the number of names from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above individually,
\verse just as Adonai commanded Moses. And he counted them in the desert of Sinai.
\verse The descendants of Reuben, the firstborn of Israel, their genealogies according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} according to the number of names, every male individually from \textit{twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone who is able to go to war:
\verse those who were counted from the tribe of Reuben were forty-six thousand five hundred.
\verse From the descendants of Simeon, their genealogies according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} those who were counted according to the number of their names, every individual male from \textit{twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone who is able to go to war:
\verse those who were counted from the tribe of Simeon were fifty-nine thousand three hundred.
\verse From the descendants of Gad, their genealogies according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} according to the number of names, from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone who is able to go to war:
\verse those who were counted from the tribe of Gad were forty-five thousand six hundred and fifty.
\verse From the descendants of Judah, their genealogies according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} according to the number of names, from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone who is able to go to war:
\verse those who were counted from the tribe of Judah were seventy-four thousand six hundred.
\verse From the descendants of Issachar, their genealogies according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} according to the number of names, from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone who is able to go to war:
\verse those who were counted from the tribe of Issachar were fifty-four thousand four hundred.
\verse From the descendants of Zebulun, their genealogies according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} according to the number of names, from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone who is able to go to war:
\verse those who were counted from the tribe of Zebulun were fifty-seven thousand four hundred.
\verse From the descendants of Joseph: from the descendants of Ephraim, their genealogies according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} according to the number of names, from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone who is able to go to war:
\verse those who were counted from the tribe of Ephraim were forty thousand five hundred.
\verse From the descendants of Manasseh, their genealogies according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} according to the number of names, from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone who is able to go to war:
\verse those who were counted from the tribe of Manasseh were thirty-two thousand two hundred.
\verse From the descendants of Benjamin, their genealogies according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} according to the number of names, from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone who is able to go to war:
\verse those who were counted from the tribe of Benjamin were thirty-five thousand four hundred.
\verse From the descendants of Dan, their genealogies according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} according to the number of names, from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone who is able to go to war:
\verse those who were counted from the tribe of Dan were sixty-two thousand seven hundred.
\verse From the descendants of Asher, their genealogies according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} according to the number of names, from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone who is able to go to war:
\verse those who were counted from the tribe of Asher were forty-one thousand five hundred.
\verse From the descendants of Naphtali, their genealogies according to their clans, according to \textit{their families},\lebnote{Literally “their fathers”} according to the number of names, from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone who is able to go to war:
\verse those who were counted from the tribe of Naphtali were fifty-three thousand four hundred.
\verse These are the ones counted whom Moses and Aaron mustered,\lebnote{Or “counted,” or “summoned,” or “enrolled”} with the twelve leaders of Israel, each one from \textit{his family}.\lebnote{Literally “the house of his fathers”}
\verse So all those who were counted from the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} according to \textit{their families},\lebnote{Literally “the house of their fathers”} from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, everyone in Israel who is able to go to war.
\verse All of the ones counted were six hundred and three thousand, five hundred and fifty.
\verse The Levites from their ancestors’\lebnote{Or “fathers’ ”} tribe were not mustered\lebnote{Or “counted,” or “summoned,” or “enrolled”} in their midst.
\verse And Adonai spoke to Moses, saying,
\verse “You will not muster\lebnote{Or “count,” or “summon,” or “enroll”} the tribe of Levi, and you will not \textit{take a census of}\lebnote{Literally “Lift up the head of”} them in the midst of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse You will \textit{appoint}\lebnote{Or “count,” or “summon,” or “enroll”} them over the tabernacle of the testimony,\lebnote{Other modern translations read “tabernacle of the covenant”} over all its vessels, and over all that belongs to it. They will carry the tabernacle and all its vessels, and they will care for it; and they will camp around the tabernacle.
\verse And when the tabernacle is set out, the Levites will \textit{take it down},\lebnote{Literally “lower it”} and when encamping the tabernacle the Levites will set it up; the stranger\lebnote{Or “outsider”} that approaches it will be put to death.
\verse The \textit{Israelites}\lebnote{Literally “sons/children of Israel”} will encamp, each in their own camp, and each by their\lebnote{Hebrew “his” or “its”} own banner according to their divisions.
\verse But the Levites will encamp around the tabernacle of the testimony,\lebnote{Other modern translations read “tabernacle of the covenant”} and there will not be wrath on the community of the \textit{Israelites};\lebnote{Literally “sons/children of Israel”} and the Levites will keep the requirements of the tabernacle of the testimony.”\lebnote{Other modern translations read “tabernacle of the covenant”}
\verse And the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} did thus; they did everything that Adonai commanded Moses.
\end{biblechapter}

\begin{biblechapter} % Numbers 2
\verseWithHeading{The Arrangement of the Camps} Adonai spoke to Moses and Aaron, saying,
\verse “The \textit{Israelites}\lebnote{Literally “sons/children of Israel”} will encamp each with his standard, with a banner according to \textit{their families};\lebnote{Literally “the house of their fathers”} they will encamp around the tent of assembly.
\verse The ones who encamp on the eastern side, toward the sunrise, will be of the standard of the camp of Judah according to their divisions; and the leader of the descendants of Judah will be Nahshon son of Amminadab,
\verse and his division and \textit{the ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} are seventy-four thousand six hundred.
\verse And the ones who encamp next to him will be the tribe of Issachar. And the leader of the descendants of Issachar will be Nethanel son of Zuar,
\verse and his division are fifty-four thousand four hundred.
\verse For the tribe of Zebulun: the leader of the descendants of Zebulun will be Eliab son of Helon,
\verse and his division and the \textit{ones counted}\lebnote{Literally “the ones counted of him,” or “the ones mustered of him”} are fifty-seven thousand four hundred.
\verse All those counted from the camp of Judah are one hundred and eighty-six thousand four hundred. They will set out first according to their divisions.
\verse “The standard of the camp of Reuben will be to the south according to their divisions. The leader of the descendants will be Elizur son of Shedeur.
\verse And his division and the \textit{ones counted}\lebnote{Literally “the ones counted of him,” or “the ones mustered of him”} are forty-six thousand five hundred.
\verse Those encamped next to him will be the tribe of Simeon. The leader of the descendants of Simeon will be Shelumiel son of Zurishaddai.
\verse And his division and the \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} are fifty-nine thousand three hundred.
\verse For the tribe of Gad: the leader of the descendants of Gad will be Eliasaph son of Reuel.
\verse And his division and the \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} are forty-five thousand six hundred and fifty.
\verse All \textit{those counted}\lebnote{Or “those mustered”} from the camp of Reuben are one hundred and fifty-one thousand four hundred and fifty. They will set out second according to their divisions.
\verse “The tent of assembly the camp of the Levites will set out in the midst of the camps; they will set out just as they encamped, \textit{each according to their standards}.\lebnote{Literally “each man on his hand according to their standards”}
\verse “The standard of the camp of Ephraim according to their divisions will be to the west. The leader of the descendants of Ephraim will be Elishama son of Ammihud.
\verse And his division and the \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} are forty thousand five hundred.
\verse The tribe of Manasseh will be next to him. The leader of the descendants of the tribe of Manasseh will be Camaliel son of Pedahzur.
\verse And his division and the \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} are thirty-two thousand two hundred.
\verse For the tribe of Benjamin: the leader of the descendants of Benjamin will be Abidan son of Gideoni.
\verse And his division and the \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} are thirty-five thousand four hundred.
\verse All \textit{those counted}\lebnote{Or “those mustered”} from the camp of Ephraim are one hundred and eighty thousand one hundred. They will set out third according to their divisions.
\verse “The standard of the camp of Dan according to their divisions will be to the west. The leader of the descendants of Dan will be Ahiezer son of Ammishaddai.
\verse And his division and the \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} are sixty-two thousand seven hundred.
\verse Those encamped next to him will be the tribe of Asher. The leader of the descendants of Asher will be Pagiel son of Ocran.
\verse And his division and the \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} are forty-one thousand five hundred.
\verse For the tribe of Naphtali: the leader of the descendants of Naphtali will be Ahira son of Enan.
\verse And his division and the \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} are fifty-three thousand four hundred.
\verse All the \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} from the camp of Dan are one hundred and fifty-seven thousand six hundred. They will set out \textit{last}\lebnote{Literally “from behind”} according to their divisions.”
\verse These were the ones counted of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} according to \textit{their families};\lebnote{Literally “the house of their fathers”} all those counted from the camps according to their divisions were six hundred and three thousand five hundred.
\verse The Levites were not counted in the midst of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} just as Adonai commanded Moses.
\verse And the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} did everything that Adonai commanded Moses. They encamped according to their standards, and they\lebnote{Hebrew “each one,” or “each man”} set out each one according to their clans\lebnote{Hebrew “his clans”} among \textit{their families}.\lebnote{Literally “the house of his fathers”}
\end{biblechapter}

\begin{biblechapter} % Numbers 3
\verseWithHeading{Aaron’s Sons} These are the genealogies of Aaron and Moses \textit{at the time}\lebnote{Literally “on a day”} when Adonai spoke to Moses on Mount Sinai.
\verse These are the names of the descendants of Aaron: Nadab the firstborn, Abihu, Eleazar, and Ithamar.
\verse These are the names of the descendants of Aaron, the priests, the anointed ones whom \textit{he consecrated as priests}.\lebnote{Literally “he filled their hands to serve as a priest”}
\verse Nadab and Abihu died \textit{before Adonai}\lebnote{Literally “before the face of Adonai”} when they presented a strange fire \textit{before Adonai}\lebnote{Literally “before the face of Adonai”} in the desert of Sinai, and \textit{they had no children}.\lebnote{Literally “sons were not for them”} Eleazar and Ithamar served as priest during the presence of Aaron their father.
\verse Adonai spoke to Moses, saying,
\verse “Bring near the tribe of Levi, and \textit{set the tribe}\lebnote{Literally “cause it to stand”} \textit{before Aaron}\lebnote{Literally “in the presence of Aaron”} the priest, and they will minister to him.
\verse They shall observe his duties and the duties of the entire community before the tent of assembly, to do the work of the tabernacle.
\verse And they will keep all the vessels of the tent of assembly and the responsibilities of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} to do the work\lebnote{Or “service”} of the tabernacle.
\verse You will give the Levites to Aaron and to his descendants; they are surely assigned to him from among the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse But you will count Aaron and his descendants; they will keep their priesthood, and the stranger\lebnote{Or “outsider”} who approaches will be put to death.”
\verse Adonai spoke to Moses saying,
\verse “I myself receive the Levites from the midst of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} in the place of all the firstborn of the offspring of the womb from the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”} The Levites will be mine
\verse because all the firstborn are mine; on the day of my killing all the firstborn in the land of Egypt, I consecrated for myself all the firstborn in Israel, \textit{both humankind and animal};\lebnote{Literally “from humankind to animal”} they will be mine. I am Adonai.”
\verse Adonai spoke to Moses in the desert of Sinai, saying,
\verse “Muster\lebnote{Or “count,” or “summon,” or “enroll”} the descendants of Levi according to \textit{their families},\lebnote{Literally “the house of their fathers”} according to their clans. You will count every male from \textit{one month}\lebnote{Literally “the son of a month”} and above.”
\verse So Moses mustered\lebnote{Or “counted,” or “summoned,” or “enrolled”} them according to the \textit{command of Adonai},\lebnote{Literally “mouth of Adonai”} just as he commanded.
\verse These were the sons of Levi according to their names: Gershon, Kohath, and Merari.
\verse And these are the names of the sons of Gershon according to their clans: Libni and Shimei.
\verse And the sons of Kohath according to their clans: Amram, Izhar, Hebron, and Uzziel.
\verse The sons of Merari according to their tribes: Mahli and Mushi. These are the clans of the Levites according to \textit{their families}.\lebnote{Literally “the house of their fathers”}
\verse \textit{To Gershon belonged}\lebnote{Literally “To Gershon was”} the clan of the Libnites and the clan of the Shimeites; these are the clans of the Gershonites.
\verse The \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} according to the number of every male from \textit{one month}\lebnote{Literally “the son of a month and above”} and above were seven thousand five hundred.
\verse The clans of the Gershonites will camp behind the tabernacle to the west,
\verse and the leader of \textit{the family}\lebnote{Literally “the house of his father”} of the Gershonites is Eliasaph son of Lael.
\verse And the responsibility of the descendants of Gershon in the tent of assembly is the tabernacle, and the tent covering it and the curtain of the doorway of the tent of the assembly,
\verse and the hangings of the courtyard and the curtain of the doorway of the courtyard that is around the tabernacle and the altar, and its ten cords, all of its use.
\verse \textit{To Kohath belonged}\lebnote{Literally “For Kohath was”} the clan of Amramites,\lebnote{Hebrew “Amramite”} the clan of the Izharites,\lebnote{Hebrew “Izharite”} the clan of the Hebronites,\lebnote{Hebrew “Hebronite”} and the clan of the Uzzielites;\lebnote{Hebrew “Uzzielite”} these were the clans of the Kohathites.\lebnote{Hebrew “Kohathite”}
\verse According to the number of every male from \textit{one month}\lebnote{Literally “the son of a month”} and above there were eight thousand six hundred keeping the responsibility of the sanctuary.
\verse The clan of the descendants of Kohath will encamp on the side of the tabernacle to the south.
\verse The leader of \textit{his family}\lebnote{Literally “the house of his father”} according to the clans of the Kohathites\lebnote{Hebrew “Kohathite”} is Elizaphan the son of Uzziel.
\verse Their responsibility was the ark, the table, the lampstand, the altar, and the vessels of the sanctuary, with which they ministered, and the curtain, and all of its use.
\verse The \textit{chief of the leaders}\lebnote{Literally “the leader of leaders”} of the Levites\lebnote{Hebrew “Levite”} was Eleazar son of Aaron the priest who had oversight of those keeping the responsibility of the sanctuary.
\verse \textit{To Merari belonged}\lebnote{Literally “for Merari was”} the clan of Mahlites\lebnote{Hebrew “Mahlite”} and the clan of the Mushites:\lebnote{Hebrew “Mushite”} these are the clans of Merari.
\verse The \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} according to the number of every male from \textit{one month}\lebnote{Literally “the son of a month”} and above were six thousand two hundred.
\verse The leader of \textit{the family}\lebnote{Literally “the house of his father”} according to the clans of Merari is Zuriel son of Abihail; they will encamp of the side of the tabernacle to the north.
\verse The responsibility of the sons of Merari was the supervision of the frames of the tabernacle, its bars, pillars, bases, and all its vessels and all its service,
\verse and the pillars around the courtyard, and their bases, pegs, and cords.
\verse Those encamped before the tabernacle to the east—before the tent of assembly to the east—were Moses and Aaron and his sons; they will keep the responsibility of the sanctuary \textit{for the Israelites};\lebnote{Literally “for the responsibility of the children of Israel”} and the stranger\lebnote{Or “outsider”} who approaches will be put to death.
\verse All those counted from the Levites whom Moses and Aaron mustered\lebnote{Or “counted,” or “summoned,” or “enrolled”} according to \textit{the word of Adonai},\lebnote{Literally “the mouth of Adonai”} according to their clans, every male from \textit{one month}\lebnote{Literally “the son of a month”} and above were twenty-two thousand.
\verse And Adonai said to Moses, “Muster every firstborn male from the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} from \textit{one month}\lebnote{Literally “the son of a month”} and above and \textit{count}\lebnote{Literally “and lift up the number of”} their names.
\verse And you will receive the Levites for me—I am Adonai—in the place of all the firstborn among the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and the animals\lebnote{Hebrew “animal”} of the Levites in the place of all the firstborn among the animals\lebnote{Hebrew “animal”} among the \textit{Israelites}.”\lebnote{Literally “sons/children of Israel”}
\verse So Moses mustered\lebnote{Or “counted” or “summoned” or “enrolled”} all the firstborn among the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} just as Adonai commanded him.
\verse And all the firstborn males\lebnote{Hebrew “male”} among the number of names from \textit{one month}\lebnote{Literally “the son of a month”} and above, the \textit{ones counted},\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} were twenty-two thousand two hundred and seventy-three.
\verse Adonai spoke to Moses, saying,
\verse “Receive the Levites in the place of all the firstborn among the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and the animals of the Levites in the place of their animals; the Levites will be mine. I am Adonai.
\verse And the ransom of the two hundred and seventy-three of the firstborn of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} who are excessive over the Levites,
\verse you will receive five shekels a person, in the sanctuary shekel; you will collect twenty gerahs\lebnote{Hebrew “gerah”} per shekel.
\verse You will give the money to Aaron, and to his sons the ransom of the ones who are excessive among them.”
\verse And Moses received the money of the redemption from the ones who were excessive from those redeemed of the Levites.
\verse From the firstborn of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} he took the money, one thousand three hundred and sixty-five shekels, in the sanctuary shekel.
\verse And Moses gave the money of the ransom to Aaron and to his sons according to the \textit{word}\lebnote{Literally “mouth”} of Adonai just as Adonai commanded Moses.
\end{biblechapter}

\begin{biblechapter} % Numbers 4
\verseWithHeading{The Census of the Kohathites} Adonai spoke to Moses and Aaron, saying,
\verse “\textit{Take a census}\lebnote{Literally “and lift up the number of”} of the descendants\lebnote{Or “sons”} of Kohath from the midst of the descendants\lebnote{Or “sons”} of Levi, according to their clans and \textit{their families},\lebnote{Literally “the house of their fathers”}
\verse from \textit{thirty years old}\lebnote{Literally “a son of thirty years”} and above, up to \textit{fifty years old},\lebnote{Literally “a son of fifty years”} everyone who comes to the service to do the work in the tent of assembly.
\verse This is the work of the descendants\lebnote{Or “sons”} of Kohath in the tent of assembly, concerning the holiness of the sanctuary:
\verse When setting out the camp, Aaron and his sons will go and lower the curtain of the covering and cover with it the ark of the testimony.\lebnote{Or “the ark of the covenant”}
\verse They will put on it a covering of \textit{fine leather},\lebnote{Literally “the hide of a sea cow”} and they will spread a cloth of perfect blue over it, and they will place its poles.
\verse And over the table of the presence they will spread out a blue cloth and put on it the plates, dishes, and libation bowls, and the pitchers of the libation; and the bread of continuity will be on it.
\verse They will spread over it a scarlet cloth, and they will cover it with a covering of \textit{fine leather},\lebnote{Literally “the hide of a sea cow”} and they will place its poles.
\verse They will take a blue cloth and cover the lampstand for the light source, its lamps, a pair of its tongs, its small pans, and all the vessels of its oil with which they attend to it.
\verse They will put it and all its vessels inside a covering of \textit{fine leather}\lebnote{Literally “the hide of a sea cow”} and put it on the carrying frame.
\verse Over the altar of gold they will spread a blue cloth, and they will cover it with a covering of \textit{fine leather}\lebnote{Literally “the hide of a sea cow”} and place its poles.
\verse They will take all the vessels of the cultic service with which they serve in the sanctuary and put them on a blue cloth, and they will cover them with a covering of \textit{fine leather};\lebnote{Literally “the hide of a sea cow”} and they will put them on the carrying frame.
\verse They will remove the fat-soaked ashes from the altar and spread a purple cloth over it;
\verse they will put on it all the vessels with which they serve, the fire pans, forks, shovels, and bowls—all the vessels of the altar. They will spread on it a covering of \textit{fine leather};\lebnote{Literally “the hide of a sea cow”} and they will place its poles.
\verse And when Aaron and his sons have finished covering the sanctuary and all the vessels of the sanctuary when the camp sets out, the descendants\lebnote{Or “sons”} of Kohath will come after to carry these, but they must not touch the sanctuary, or they will die. These are the load of the descendants\lebnote{Or “sons”} of Kohath in the tent of assembly.
\verse “Eleazar son of Aaron the priest is to supervise the oil of the light source, the incense, \textit{the regular grain offering},\lebnote{Literally “the grain offering of continuity”} the oil of anointment, the supervision of all the tabernacle and all that is in it, in the sanctuary and in its vessels.”
\verse Adonai spoke to Moses and Aaron, saying,
\verse “You must not cut off the tribe of the clan of the Kohathites\lebnote{Hebrew “Kohathite”} from the midst of the Levites.
\verse Do this to them and they will live and not die when they come near the most holy things. Aaron and his sons will go and appoint them, each one to his task and burden.
\verse But they must not go and look \textit{for a moment}\lebnote{Literally “as devouring”} at the holy objects.”
\verse Adonai spoke to Moses and Aaron, saying,
\verse “\textit{Take a census}\lebnote{Literally “And lift up the number of”} of the descendants\lebnote{Or “sons”} of Gershon also, according to \textit{their families}\lebnote{Literally “the house of their fathers”} and clans,
\verse from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above until \textit{fifty years old};\lebnote{Literally “a son of fifty years”} you will muster\lebnote{Or “count,” or “summon,” or “enroll”} them, all who come to help to do the work of the tent of assembly.
\verse This is the work of the clans of the Gershonites:\lebnote{Hebrew “Gershonite”} to serve and to carry.
\verse They will carry the curtains of the tabernacle and the tent of assembly and its covering and the covering of \textit{fine leather},\lebnote{Literally “the hide of a sea cow”} which \textit{is on top of it},\lebnote{Literally “is above upon it”} and the curtain of the doorway of the tent of assembly,
\verse and the curtains of the courtyard, and the covering for the doorway of the gate of the courtyard, which is all around on the tabernacle and altar, and their cords and all the vessels of their work; and all that is done to them they will do.
\verse And all the work of the descendants\lebnote{Or “sons”} of the Gershonites\lebnote{Hebrew “Gershonite”} will be at \textit{the command}\lebnote{Literally “the mouth”} of Aaron and his sons, for all they are to carry and for all their work, and you will appoint to them responsibility for all they are to carry.
\verse This is the work of the clan of the descendants\lebnote{Or “sons”} of the Gershonites\lebnote{Hebrew “Gershonite”} in the tent of assembly, and their responsibility lies \textit{under the direction}\lebnote{Literally “in the hand”} of Ithamar son of Aaron the priest.
\verse “For the descendants\lebnote{Or “sons”} of Merari according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} you will muster\lebnote{Or “count,” or “summon,” or “enroll”} them;
\verse from \textit{those thirty years old}\lebnote{Literally “a son of thirty years”} and above until \textit{fifty years old};\lebnote{Literally “a son of fifty years”} you will muster\lebnote{Or “count,” or “summon,” or “enroll”} them, all who come to do the work of the tent of assembly.
\verse And this is the responsibility of \textit{those who are to carry},\lebnote{Literally “them carrying”} all their work in the tent of assembly: the frames of the tabernacle and its bars, pillars, and bases,
\verse and the pillars of the courtyard all around, and their bases, pegs, and cords, with all their vessels and for all their work. You will appoint by name the vessels that they are responsible to carry.
\verse This is the work of the clan of the descendants\lebnote{Or “sons”} of Merari, for all their work in the tent of assembly \textit{under the direction}\lebnote{Literally “in the hand”} of Ithamar son of Aaron the priest.”
\verse And Moses and Aaron mustered the leaders of the community according to the house of \textit{their families},\lebnote{Literally “the house of their fathers”}
\verse from \textit{those thirty years old}\lebnote{Literally “a son of thirty years”} and above until \textit{fifty years old};\lebnote{Literally “a son of fifty years”} everyone who comes to the service to work in the tent of assembly,
\verse the \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} were two thousand seven hundred and fifty.
\verse These were those counted of the clans of the Kohathites,\lebnote{Hebrew “Kohathite”} everyone who served in the tent of assembly whom Moses and Aaron mustered\lebnote{Or “counted,” or “summoned,” or “enrolled”} according to \textit{the command}\lebnote{Literally “the mouth”} of Adonai by the hand of Moses.\lebnote{Or “through Moses”}
\verse And the descendants\lebnote{Or “sons”} of Gershon counted according to their clans and according to \textit{their families},\lebnote{Literally “the house of their fathers”}
\verse from \textit{those thirty years old}\lebnote{Literally “a son of thirty years”} and above until \textit{fifty years old},\lebnote{Literally “a son of fifty years”} everyone who comes to the service to work in the tent of assembly;
\verse the \textit{ones counted},\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”} were two thousand six hundred and thirty.
\verse These were those counted of the clans of the descendants\lebnote{Or “sons”} of Gershon, everyone who serves in the tent of assembly whom Moses and Aaron mustered\lebnote{Or “counted,” or “summoned,” or “enrolled”} according to \textit{the command}\lebnote{Literally “the mouth”} of Adonai.
\verse Those counted of the clans of the descendants\lebnote{Or “sons”} of Merari according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”}
\verse from \textit{those thirty years old}\lebnote{Literally “thirty years and above”} and above until \textit{fifty years old},\lebnote{Literally “a son of fifty years”} everyone who comes to the service to work in the tent of assembly,
\verse the \textit{ones counted},\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} according to their clans, were three thousand two hundred.
\verse These were those counted of the clans of the descendants\lebnote{Or “sons”} of Merari, whom Moses and Aaron mustered\lebnote{Or “counted,” or “summoned,” or “enrolled”} according to \textit{the command}\lebnote{Literally “the mouth”} of Adonai by the hand of Moses.\lebnote{Or “through Moses”}
\verse All those counted of the Levites whom Moses and Aaron and all the leaders of Israel mustered\lebnote{Or “counted,” or “summoned,” or “enrolled”} according to their clans, according to \textit{their families},\lebnote{Literally “the house of their fathers”}
\verse from \textit{those thirty years old}\lebnote{Literally “a son of thirty years”} and above until \textit{fifty years old},\lebnote{Literally “a son of fifty years”} everyone who comes to the service to do the work of the service carrying in the tent of assembly,
\verse the \textit{ones counted}\lebnote{Literally “the ones counted of them,” or “the ones mustered of them”} were eighty thousand five hundred and eighty.
\verse According to \textit{the command}\lebnote{Literally “the mouth”} of Adonai by the hand of Moses\lebnote{Or “through Moses”} \textit{they were mustered},\lebnote{Hebrew “he mustered”} each man according to his service and according to their\lebnote{Hebrew “his”} service and according to their\lebnote{Hebrew “his”} burden; and so they were counted by him just as Adonai commanded Moses.
\end{biblechapter}

\begin{biblechapter} % Numbers 5
\verseWithHeading{Rules Concerning Those Unclean} Adonai spoke to Moses, saying,
\verse “Command the \textit{Israelites}:\lebnote{Literally “sons/children of Israel”} they must send everyone from the camp who is afflicted with a rash,\lebnote{The precise meaning is uncertain; many modern translations suggest “leprosy”} everyone with a fluid discharge, and everyone unclean through contact with a corpse.
\verse You will send away \textit{both male and female};\lebnote{Literally “from male until female”} you will send them \textit{outside the camp}.\lebnote{Literally “to an outside place of the camp”} They must not make unclean their camps where I am dwelling in their midst.”
\verse So the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} did so. They sent them away \textit{outside the camp};\lebnote{Literally “to an outside place of the camp”} just as Adonai spoke to Moses, so did the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verseWithHeading{Rules of Restitution} Adonai spoke to Moses, saying,
\verse “Speak to the \textit{Israelites}:\lebnote{Literally “sons/children of Israel”} ‘When a man or woman \textit{commits}\lebnote{Literally “does”} any of the sins of humankind by acting unfaithfully, it is a sin against Adonai, and that person will be guilty;
\verse they will confess their sin that they did and will make restitution for their\lebnote{Hebrew “his”} guilt by adding a fifth \textit{to it}\lebnote{Literally “on top of it”} and giving it \textit{to whomever was wronged}.\lebnote{Literally “to whomever he was guilty”}
\verse But if the man does not have a redeemer to make restitution to him for the reparation, the reparation is to be given to Adonai for the priest, in addition to the ram of atonement by which atonement is made for him.
\verse And every contribution of all the holy objects of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} that they bring to the priest for him will be his.
\verse The holy objects of a man will be for him;\lebnote{That is, the priest} whatever he gives to the priest will be for him.’ ”
\verseWithHeading{Rules Concerning an Unfaithful Wife} Adonai spoke to Moses, saying,
\verse “Speak to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} and say to them, ‘If any man’s wife goes astray and acts unfaithfully to him,
\verse and a man sleeps with her and ejaculates and it is hidden from the eyes of her husband and she is concealed, although she is defiled, and there is no witness against her and she was not caught,
\verse if a spirit of jealousy comes over him, and he is jealous of his wife and she is defiled; or if a spirit of jealousy comes over him and he is jealous of his wife but she is not defiled,
\verse he will bring his wife to the priest. And he will bring her offering for her, one-tenth of an ephah of flour. He will not pour oil on it, and he will not put frankincense on it because it is a grain offering of jealousy, a grain offering of remembering,\lebnote{Or “memorial”} a reminding of guilt.
\verse “ ‘Then the priest will bring her near and present her \textit{before}\lebnote{Literally “before the face of”} Adonai;
\verse the priest will take holy waters in a clay vessel, and from the dust that is on the floor of the tabernacle, and the priest will put it into the waters.
\verse And the priest will present the woman \textit{before}\lebnote{Literally “before the face of”} Adonai, and he will uncover the head of the woman; he will then put in her hands the grain offering of the remembering—which is the grain offering of jealousy—and in the hand of the priest will be the waters of bitterness that brings a curse.
\verse Then the priest will make her swear an oath, and he will say to the woman, “If a man has not slept with you, and if you have not had an impurity affair under your husband, go unpunished from the waters of bitterness that brings this curse.
\verse But if you have had an affair under your husband, and if you are defiled and a man other than your husband had intercourse with you,”
\verse the priest will make the woman swear an oath of the sworn oath of the curse, the priest will say to the woman, “May Adonai give you a curse and a sworn oath in the midst of your people with Adonai making\lebnote{Hebrew “giving”} your hip fall away\lebnote{Or “waste away”} and your stomach swollen;
\verse and these waters that bring a curse will go into your intestines to cause your womb to swell and to make your hip fall away.”\lebnote{Or “waste away”} And the women will say, “Amen. Amen.”
\verse “ ‘And the priests will write these curses on the scroll, and he will wipe them off into the waters of the bitterness.
\verse He will make the woman drink the waters of the bitterness that brings\lebnote{Hebrew “bring”} a curse, and the waters of bitterness that bring a curse will go into her.
\verse The priest will take the grain offering of jealousy from the hand of the woman, and he will wave the grain offering \textit{before Adonai},\lebnote{Literally “before the face of Adonai”} and he will present it to the altar;
\verse the priest will grasp her memorial offering from the grain offering, and he will turn it into smoke on the altar, and afterward he will make the woman drink the waters.
\verse When he has made her drink the waters, it will come about, if she has defiled herself and acted unfaithfully to her husband and the waters of bitterness that bring a curse go into her and her stomach swells and her hip falls away,\lebnote{Or “wastes away”} the woman will be as a curse in the midst of her people.
\verse And if the woman is not defiled, and she is pure, she will go unpunished and be able to conceive children.
\verse “ ‘This is the regulation of jealousy, when a woman has an affair under her husband and she is defiled,
\verse or when a spirit of jealousy comes over a man and he is jealous of his wife, he will present the woman \textit{before Adonai},\lebnote{Literally “before the face of”}and the priest will do to her all of this law.
\verse The man will go unpunished from guilt, and the woman, she will bear her guilt.’ ”
\end{biblechapter}

\begin{biblechapter} % Numbers 6
\verseWithHeading{Rules Concerning Nazirites} Adonai spoke to Moses, saying,
\verse “Speak to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} and say to them, ‘When a man or a woman takes a special vow, a vow of a \textit{Nazirite},\lebnote{Literally “one separated”} to keep separate for Adonai,
\verse he will abstain from wine and fermented drink; he will not drink wine vinegar or vinegar of fermented drink; he will not drink the fruit juice of grapes or eat fresh or dry grapes.
\verse All the days of his separation\lebnote{That is, “the Nazirite”} you will not eat from anything that is made from the grapevine, from sour grapes to the skin of grapes.
\verse “ ‘All the days of the vow of his separation a razor will not pass over his head. Until fulfilling the days that he separated himself to Adonai he will be holy and grow long the locks of the hair of his head.
\verse “ ‘All the days of keeping himself separated for Adonai he will not go to a person who is dead;
\verse for even his father, mother, brother, or sister he will not make himself unclean \textit{by their death},\lebnote{Literally “with their dying”} because the separation to his God is on his head.
\verse He will be holy for Adonai all the days of his separation.\lebnote{That is, “the Nazirite”}
\verse “ ‘If someone dies suddenly and makes the head of his separation\lebnote{That is, “the Nazirite”} unclean, he will shave off the hair of his head on the day of his cleansing; on the seventh day he will shave it off.
\verse On the eighth day he will bring two turtledoves or two \textit{young pigeons}\lebnote{Literally “the sons of doves”} to the priest by the doorway of the tent of assembly,
\verse and the priest will offer one for a sin offering and one for a burnt offering, and he will make atonement for him because he sinned concerning the corpse. He will consecrate his head on that day.
\verse He will rededicate to Adonai the days of his separation\lebnote{That is, “the Nazirite”} and bring a ram-lamb \textit{in its first year}\lebnote{Literally “a son of his year”} as a guilt offering. The former days of his vow will fall away because his separation was defiled.
\verse “ ‘This is the regulation of the Nazirite for the day of the fulfilling of the days of his separation: one will bring him to the doorway of the tent of the assembly.
\verse He will present his offering to Adonai, one ram-lamb \textit{in its first year}\lebnote{Literally “a son of his year”} without defect as a burnt offering, and one ewe-lamb \textit{in its first year}\lebnote{Literally “a daughter of his year”} without defect as a sin offering, and one ram without defect as a fellowship offering;
\verse and a basket of unleavened bread, finely milled flour of ring-shaped bread cakes mixed with oil, and wafers of unleavened bread smeared with oil, and their grain offering and their libations.
\verse The priest will present \textit{before Adonai}\lebnote{Or “before the face of Adonai”} and offer his sin offering, his burnt offering;
\verse he will offer a ram as a sacrifice of a fellowship offering to Adonai, in addition to the basket of the unleavened bread; the priest will offer his grain offering and his libation.
\verse The Nazirite will shave off the hair of \textit{his consecrated head}\lebnote{Literally “the head of his separation”} at the doorway of the tent of assembly, and he will take the hair of \textit{his consecrated head},\lebnote{Literally “the head of his separation”} and he will put it on the fire that is beneath the sacrifice of the fellowship offering.
\verse The priest will take the shoulder from the ram that is boiled, and one ring-shaped bread cake of unleavened bread from the basket, and one wafer of unleavened bread, and he will put them on the hands of the Nazirite after he has shaved \textit{his consecrated head}.\lebnote{Literally “his separation”}
\verse The priest will wave them as a wave offering \textit{before the presence of}\lebnote{Literally “before the face of”} Adonai; they\lebnote{Hebrew “it”} are a holy object to the priest, in addition to the breast section of the wave offering, and in addition to the upper thigh of the contribution; and afterward the Nazirite may drink wine.
\verse “ ‘This is the regulation of the Nazirite who has made a vow of his offering to Adonai according to his separation, \textit{in addition to what he can afford}.\lebnote{Literally “except from that which overtakes his hand”} In accordance to the word of his vow that he vowed, he will do, concerning the instruction of his separation.’ ”
\verseWithHeading{The Priestly Blessing} Adonai spoke to Moses, saying,
\verse “Speak to Aaron and his sons, saying, ‘You will bless the \textit{Israelites}:\lebnote{Literally “sons/children of Israel”} You will say to them:
\verse Adonai will bless you 
and keep you;
\verse Adonai will make shine his face on you 
and be gracious to you;
\verse Adonai will lift up his face upon you, 
and he will give you peace.’
\verse And they will put my name on the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and I will bless them.”
\end{biblechapter}

\begin{biblechapter} % Numbers 7
\verseWithHeading{The Leaders Make Offerings} On the day Moses finished setting up the tabernacle and appointed and consecrated it and all its vessels, and the altar and its vessels, and he appointed them and consecrated them,
\verse the leaders of Israel, the heads of \textit{the families},\lebnote{Literally “the house of their fathers”} presented an offering; they were the leaders of the tribes and were the ones in charge of the counting.
\verse They brought their offering before the presence of\lebnote{Or “before the face of”} Adonai, six covered utility carts and twelve cattle, a utility cart for two of the leaders, and a bull for each; and they presented them \textit{before}\lebnote{Literally “before the face of”} the tabernacle.
\verse Adonai said to Moses, saying,
\verse “\textit{Take them},\lebnote{Literally “Take from them”} and they will be used to do the work of the tent of the assembly; and you will give them to the Levites, each \textit{according to his required service}.”\lebnote{Literally “according to the mouth of his work”}
\verse So Moses took the utility carts and the cattle, and he gave them to the Levites.
\verse Two utility carts and four cattle he gave to the descendants\lebnote{Or “sons”} of Gershon \textit{according to their required service};\lebnote{Literally “according to the mouth of their work”}
\verse and four utility carts and eight cattle he gave to the descendants\lebnote{Or “sons”} of Merari \textit{according to their required service},\lebnote{Literally “according to the mouth of their work”} \textit{under the authority of}\lebnote{Literally “in the hand of”} Ithamar son of Aaron the priest.
\verse But to the descendants\lebnote{Or “sons”} of Kohath he did not give anything because the work of the sanctuary they carried upon them on their shoulders.\lebnote{Hebrew “the shoulder”}
\verse The leaders presented offerings for the dedication of the altar on the day of its anointing, and the leaders presented their offerings \textit{before}\lebnote{Literally “in the presence of”} the altar.
\verse Adonai said to Moses, “\textit{One leader for each day}\lebnote{Literally “One leader for the day one leader for the day”} will present their offering for the dedication of the altar.”
\verse And it happened, the one who presented his offering on the first day was Nahshon son of Amminadab from the tribe of Judah.
\verse His offering was one plate of silver—its weight was one hundred and thirty shekels—and one silver bowl weighing seventy shekels according to the sanctuary shekel, \textit{both of them}\lebnote{Literally “the two of them”} filled with finely milled flour mixed with oil as a grain offering;
\verse one golden dish weighing ten shekels filled with incense;
\verse \textit{one young bull},\lebnote{Literally “one bull, a son of a cattle”} one ram, one male lamb \textit{in its first year}\lebnote{Literally “the son of its year”} as a burnt offering;
\verse one he-goat as a sin offering;
\verse and as a sacrifice of the fellowship offering, two cattle, five rams, five he-goats, and five male lambs \textit{a year old}.\lebnote{Literally “sons of a year”} This was the offering of Nahshon son of Amminadab.
\verse On the second day Nethanel son of Zuar, leader of Issachar, presented an offering.
\verse He presented as his offering one silver plate—its weight one hundred and thirty shekels—and one silver bowl for drinking wine weighing seventy shekels\lebnote{Hebrew “shekel”} according to the sanctuary shekel, \textit{both of them}\lebnote{Literally “the two of them”} filled with finely milled flour mixed with oil as a grain offering.
\verse One dish weighing ten shekels filled with incense;
\verse one \textit{young }\lebnote{Literally “a son of cattle”} bull, one ram, a male lamb \textit{in its first year}\lebnote{Literally “the son of its year”} as a burnt offering;
\verse one he-goat as a sin offering;
\verse and for the sacrifice of the fellowship offering, two cattle, five rams, five he-goats, and five male lambs \textit{in their first year}.\lebnote{Literally “sons of a year”} This was the offering of Nethanel son of Zuar.
\verse On the third day Eliab son of Helon, leader of the descendants\lebnote{Or “sons”} of Zebulun:
\verse his offering was one silver plate—its weight one hundred and thirty shekels—and one silver bowl for drinking wine weighing seventy shekels\lebnote{Hebrew “shekel”} according to the sanctuary shekel, \textit{both of them}\lebnote{Literally “the two of them”} filled with finely milled flour mixed with oil as a grain offering;
\verse \textit{one golden dish weighing ten shekels}\lebnote{Literally “one dish of ten gold”} filled with incense;
\verse one \textit{young}\lebnote{Literally “a son of cattle”} bull, one ram, a male lamb \textit{in its first year}\lebnote{Literally “the son of its year”} as a burnt offering;
\verse one he-goat as a sin offering;
\verse and for the sacrifice of the fellowship offering, two cattle, five rams, five he-goats, and five male lambs \textit{in their first year}.\lebnote{Literally “sons of a year”} This was the offering of Eliab son of Helon.
\verse On the fourth day Elizur son of Shedeur, leader of the descendants\lebnote{Or “sons”} of Reuben:
\verse his offering was one silver plate—its weight one hundred and thirty shekels—and one silver bowl for drinking weighing seventy shekels\lebnote{Hebrew “shekel”} according to the sanctuary shekel, \textit{both of them}\lebnote{Literally “the two of them”} filled with finely milled flour mixed with oil as a grain offering;
\verse \textit{one golden dish weighing ten shekels}\lebnote{Literally “one dish of ten gold”} filled with incense;
\verse one \textit{young}\lebnote{Literally “a son of cattle”} bull, one ram, a male lamb \textit{in its first year}\lebnote{Literally “the son of its year”} as a burnt offering;
\verse one he-goat as a sin offering;
\verse and for the sacrifice of the fellowship offering, two cattle, five rams, five he-goats, and five male lambs \textit{in their first year}.\lebnote{Literally “sons of a year”} This was the offering of Elizur son of Shedeur.
\verse On the fifth day Shelumiel son of Zurishaddai, leader of the descendants\lebnote{Or “sons”} of Simeon:
\verse his offering was one silver plate—its weight one hundred and thirty shekels—and one silver bowl for drinking weighing seventy shekels\lebnote{Hebrew “shekel”} according to the sanctuary shekel, \textit{both of them}\lebnote{Literally “the two of them”} filled with finely milled flour mixed with oil as a grain offering;
\verse \textit{one golden dish weighing ten shekels}\lebnote{Literally “one dish of ten gold”} filled with incense;
\verse one \textit{young}\lebnote{Literally “a son of cattle”} bull, one ram, a male lamb \textit{in its first year}\lebnote{Literally “the son of its year”} as a burnt offering;
\verse one he-goat as a sin offering;
\verse and for the sacrifice of the fellowship offering, two cattle, five rams, five he-goats, and five male lambs \textit{in their first year}.\lebnote{Literally “sons of a year”} This was the offering of Shelumiel son of Zurishaddai.
\verse On the sixth day Eliasaph son of Deuel, leader of the descendants\lebnote{Or “sons”} of Gad:
\verse his offering was one silver plate—its weight one hundred and thirty shekels—and one silver bowl for drinking weighing seventy shekels\lebnote{Hebrew “shekel”} according to the sanctuary shekel, \textit{both of them}\lebnote{Literally “the two of them”} filled with finely milled flour mixed with oil as a grain offering;
\verse \textit{one golden dish weighing ten shekels}\lebnote{Literally “one dish of ten gold”} filled with incense;
\verse one \textit{young}\lebnote{Literally “a son of cattle”} bull, one ram, a male lamb \textit{in its first year}\lebnote{Literally “the son of its year”} as a burnt offering;
\verse one he-goat as a sin offering;
\verse and for the sacrifice of the fellowship offering, two cattle, five rams, five he-goats, and five male lambs \textit{in their first year}.\lebnote{Literally “sons of a year”} This was the offering of Eliasaph son of Deuel.
\verse On the seventh day Elishama son of Ammihud, leader of the descendants\lebnote{Or “sons”} of Ephraim:
\verse his offering was one silver plate—its weight one hundred and thirty shekels—and one silver bowl for drinking weighing seventy shekels\lebnote{Hebrew “shekel”} according to the sanctuary shekel, \textit{both of them}\lebnote{Literally “the two of them”} filled with finely milled flour mixed with oil as a grain offering;
\verse \textit{one golden dish weighing ten shekels}\lebnote{Literally “one dish of ten gold”} filled with incense;
\verse one \textit{young}\lebnote{Literally “a son of cattle”} bull, one ram, a male lamb \textit{in its first year}\lebnote{Literally “the son of its year”} as a burnt offering;
\verse one he-goat as a sin offering;
\verse and for the sacrifice of the fellowship offering, two cattle, five rams, five he-goats, and five male lambs \textit{in their first year}.\lebnote{Literally “sons of a year”} This was the offering of Elishama son of Ammihud.
\verse On the eighth day Gamaliel son of Pedahzur, leader of the descendants\lebnote{Or “sons”} of Manasseh:
\verse his offering was one silver plate—its weight one hundred and thirty shekels—and one silver bowl for drinking weighing seventy shekels according to the sanctuary shekel, \textit{both of them}\lebnote{Literally “the two of them”} filled with finely milled flour mixed with oil as a grain offering;
\verse \textit{one golden dish weighing ten shekels}\lebnote{Literally “one dish of ten gold”} filled with incense;
\verse one \textit{young}\lebnote{Literally “a son of cattle”} bull, one ram, a male lamb \textit{in its first year}\lebnote{Literally “the son of its year”} as a burnt offering;
\verse one he-goat as a sin offering;
\verse and for the sacrifice of the fellowship offering, two cattle, five rams, five he-goats, and five male lambs \textit{in their first year}.\lebnote{Literally “sons of a year”} This was the offering of Gamaliel son of Pedahzur.
\verse On the ninth day Abidan son of Gideoni, leader of the descendants\lebnote{Or “sons”} of Benjamin:
\verse his offering was one silver plate—its weight one hundred and thirty shekels—and one silver bowl for drinking weighing seventy shekels\lebnote{Hebrew “shekel”} according to the sanctuary shekel, \textit{both of them}\lebnote{Literally “the two of them”} filled with finely milled flour mixed with oil as a grain offering;
\verse \textit{one golden dish weighing ten shekels}\lebnote{Literally “one dish of ten gold”} filled with incense;
\verse one \textit{young}\lebnote{Literally “a son of cattle”} bull, one ram, a male lamb \textit{in its first year}\lebnote{Literally “the son of its year”} as a burnt offering;
\verse one he-goat as a sin offering;
\verse and for the sacrifice of the fellowship offering, two cattle, five rams, five he-goats, and five male lambs \textit{in their first year}.\lebnote{Literally “sons of a year”} This was the offering of Abidan son of Gideoni.
\verse On the tenth day Ahiezer son of Ammishaddai, leader of the descendants\lebnote{Or “sons”} of Dan:
\verse his offering was one silver plate—its weight one hundred and thirty shekels—and one silver bowl for drinking weighing seventy shekels\lebnote{Hebrew “shekel”} according to the sanctuary shekel, \textit{both of them}\lebnote{Literally “the two of them”} filled with finely milled flour mixed with oil as a grain offering;
\verse \textit{one golden dish weighing ten shekels}\lebnote{Literally “one dish of ten gold”} filled with incense;
\verse one \textit{young}\lebnote{Literally “a son of cattle”} bull, one ram, a male lamb \textit{in its first year}\lebnote{Literally “the son of its year”} as a burnt offering;
\verse one he-goat as a sin offering;
\verse and for the sacrifice of the fellowship offering, two cattle, five rams, five he-goats, and five male lambs \textit{in their first year}.\lebnote{Literally “sons of a year”} This was the offering of Ahiezer son of Ammishaddai.
\verse On the eleventh day Pagiel son of Ocran, leader of the descendants\lebnote{Or “sons”} of Asher:
\verse his offering was one silver plate—its weight one hundred and thirty shekels—and one silver bowl for drinking weighing seventy shekels\lebnote{Hebrew “shekel”} according to the sanctuary shekel, \textit{both of them}\lebnote{Literally “the two of them”} filled with finely milled flour mixed with oil as a grain offering;
\verse \textit{one golden dish weighing ten shekels}\lebnote{Literally “one dish of ten gold”} filled with incense;
\verse one \textit{young}\lebnote{Literally “a son of cattle”} bull, one ram, a male lamb \textit{in its first year}\lebnote{Literally “the son of its year”} as a burnt offering;
\verse one he-goat as a sin offering;
\verse and for the sacrifice of the fellowship offering, two cattle, five rams, five he-goats, and five male lambs \textit{in their first year}.\lebnote{Literally “sons of a year”} This was the offering of Pagiel son of Ocran.
\verse On the twelfth day Ahira son of Enan, leader of the descendants\lebnote{Or “sons”} of Naphtali:
\verse his offering was one silver plate—its weight one hundred and thirty shekels—and one silver bowl for drinking weighing seventy shekels\lebnote{Hebrew “shekel”} according to the sanctuary shekel, \textit{both of them}\lebnote{Literally “the two of them”} filled with finely milled flour mixed with oil as a grain offering;
\verse \textit{one golden dish weighing ten shekels}\lebnote{Literally “one dish of ten gold”} filled with incense;
\verse one \textit{young}\lebnote{Literally “a son of cattle”} bull, one ram, a male lamb \textit{in its first year}\lebnote{Literally “the son of its year”} as a burnt offering;
\verse one he-goat as a sin offering;
\verse and for the sacrifice of the fellowship offering, two cattle, five rams, five he-goats, and five male lambs \textit{in their first year}.\lebnote{Literally “sons of a year”} This was the offering of Ahira son of Enan.
\verse This was the dedication of the altar on the day of anointing it, from the leaders of Israel: twelve silver plates, twelve silver bowls for drinking wine, twelve golden dishes;
\verse each plate of silver weighed one hundred and thirty shekels, and each bowl for drinking seventy, all the silver of the vessels two thousand four hundred shekels, according to the sanctuary shekel;
\verse the twelve golden dishes filled with incense, each dish weighing ten shekels according to the sanctuary shekel, all the gold of the dishes one hundred and twenty;
\verse all the cattle for the burnt offering twelve bulls, twelve rams, twelve male lambs \textit{in their first year},\lebnote{Literally “sons of a year”} and their grain offering; and twelve he-goats as a sin offering;
\verse and all the cattle of the sacrifice of the fellowship offering twenty-four bulls, sixty rams, sixty he-goats, sixty male lambs \textit{in their first year}.\lebnote{Literally “sons of a year”} These were the dedication of the altar after its anointing.
\verse And when Moses came to the tent of assembly to speak with him,\lebnote{That is, Adonai} he would hear the voice speaking to him from the atonement cover,\lebnote{Some modern translations have “mercy seat” (see, for example, the NRSV, NASB)} which is on the ark of the testimony, from between the two cherubim, and he would speak to him.
\end{biblechapter}

\begin{biblechapter} % Numbers 8
\verseWithHeading{The Seven Lamps} Adonai spoke to Moses, saying,
\verse “Speak to Aaron, and say to him: ‘When you are setting up the lamps, the seven lamps will give light in front of the face of the lampstand.’ ”
\verse And Aaron did so; he set up the lampstand in front of the face of its lamps, just as Adonai commanded Moses.
\verse And this is \textit{how the lampstand was made},\lebnote{Literally “the work of the lampstand”} a hammered-work of gold; from its base up to its blossom,\lebnote{Or “flower”} it was hammered-work according to the pattern that Adonai showed Moses; so he made the lampstand.
\verseWithHeading{Moses Consecrates the Levites} Adonai spoke to Moses, saying,
\verse “Take the Levites from the midst of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} and purify them.
\verse So you will do to them, to purify them: sprinkle on them waters of purification, and \textit{they will shave their whole body}\lebnote{Literally “they will send a razor on all their body”} and wash their garments.
\verse And they will take a \textit{young bull}\lebnote{Literally “a bull, a son of cattle”} and its grain offering of finely milled flour mixed with oil, and you will take a second \textit{young bull}\lebnote{Literally “a bull, a son of cattle”} as a sin offering.
\verse You will bring the Levites \textit{before}\lebnote{Literally “in the presence of”} the tent of assembly, and you will summon the entire community of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse And you will bring the Levites \textit{before Adonai},\lebnote{Literally “in the presence of Adonai”} and the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} will lay their hands on the Levites,
\verse and \textit{Aaron will offer}\lebnote{Literally “Aaron will wave”} the Levites as a wave offering \textit{before Adonai}\lebnote{Literally “in the presence of Adonai”} from the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and they will do the work of Adonai.
\verse And the Levites will lay their hands on the head of the one bull and offer it as a sin offering and the other one as a burnt offering to Adonai, to make atonement for the Levites.
\verse And you will present the Levites \textit{before}\lebnote{Literally “in the presence of”} Aaron and \textit{before}\lebnote{Literally “in the presence of”} his sons, and he \textit{will offer}\lebnote{Literally “Aaron will wave”} them as a wave offering to Adonai.
\verse “And you will separate the Levites from the midst of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and the Levites will be for me.
\verse And after this the Levites will come to serve at the tent of assembly, and you will purify them, and you \textit{will offer}\lebnote{Literally “Aaron will wave”} them as a wave offering.
\verse For they are given to me exclusively from the midst of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”} I have taken them for myself in place of the firstborn of every womb, every firstborn from the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse For every firstborn among the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} is mine, both humankind and animal. On the day I destroyed every firstborn in the land of Egypt I consecrated them to me,
\verse and I have taken the Levites in the place of every firstborn among the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse And I have given the Levites; they are given to Aaron and his sons from the midst of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} to do the work of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} in the tent of the assembly and to make atonement for the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} so a plague will not be among the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} when the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} come near the sanctuary.”
\verse And Moses and Aaron and the entire community of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} did to the Levites; everything that Adonai commanded Moses concerning the Levites, the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} did to them.
\verse And the Levites purified themselves, and they washed their garments, and \textit{Aaron offered them}\lebnote{Literally “Aaron waved them”} as a wave offering \textit{before Adonai};\lebnote{Literally “in the presence of Adonai”} and Aaron made atonement for them to purify them.
\verse After this the Levites came to do their work in the tent of assembly before Aaron and his sons. Just as Adonai commanded Moses concerning the Levities, so they did to them.
\verse Adonai spoke to Moses, saying,
\verse “This is what is for the Levites: \textit{those twenty-five years old}\lebnote{Literally “from a son of twenty-five years”} and above will\lebnote{Hebrew “and he will”} come to help with the service in the work of the tent of assembly;
\verse and \textit{those fifty years old}\lebnote{Literally “from a son of fifty years”} will\lebnote{Hebrew “he will”} return from the service of the work and will serve no longer.
\verse They\lebnote{Hebrew “he”} can attend\lebnote{Or “assist”} their brothers in the tent of assembly to keep their responsibilities, but they\lebnote{Hebrew “he”} will not do work. This is what you will do concerning the Levities and their responsibilities.”
\end{biblechapter}

\begin{biblechapter} % Numbers 9
\verseWithHeading{The Passover} Adonai spoke to Moses in the desert of Sinai, in the second year after they came out from the land of Egypt, in the first month, saying,
\verse “Let the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} observe the Passover at its appointed time.
\verse On the fourteenth day of this month \textit{at twilight}\lebnote{Literally “between the two evenings”} you will perform it at its appointed time according to all its decrees; and according to all its stipulations you will observe it.”
\verse So Moses spoke to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} to observe the Passover.
\verse And they observed the Passover on the fourteenth day of the month \textit{at twilight}\lebnote{Literally “between the two evenings”} in the desert of Sinai. According to all that Adonai commanded Moses, thus the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} did.
\verse And it happened, men who were unclean \textit{by a dead person}\lebnote{Literally “by a life of a person”} were not able to perform the Passover on that day. And they came \textit{before}\lebnote{Literally “in the presence of”} Moses and Aaron on that day.
\verse And those men said to him, “Although we are unclean \textit{by a dead person},\lebnote{Literally “by a life of a person”} why are we hindered from presenting the offering of Adonai at its appointed time in the midst of the \textit{Israelites}?”\lebnote{Literally “sons/children of Israel”}
\verse Moses said to them, “Stay. I will hear what Adonai commands to you.”
\verse And Adonai spoke to Moses, saying,
\verse “Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} saying, ‘Each man that is unclean \textit{by a dead person}\lebnote{Literally “by a life of a person”} or is on a far journey, you or your \textit{descendants},\lebnote{Literally “generations”} he will observe the Passover of Adonai.
\verse On the second month on the fourteenth day \textit{at twilight}\lebnote{Literally “between the two evenings”} they will observe it; they will eat it with unleavened bread and bitter plants.
\verse They will leave none of it until morning, and they will not break a bone in it; they will observe it according to every decree of the Passover.
\verse But the man who is clean and not on a journey, and he fails to observe the Passover, that person will be cut off from the people because he did not present the offering of Adonai on its appointed time. That man will bear his guilt.
\verse If an alien dwells with you he will observe the Passover of Adonai according to the decree of the Passover and according to its stipulation; thus you will have one decree for you, for the alien and for the native of the land.’ ”
\verseWithHeading{The Cloud and the Fire} And on a day setting up the tabernacle, the cloud covered the tent of the tabernacle, the tent of the testimony; in the evening it was on the tabernacle as an appearance of fire until morning.
\verse So it was\lebnote{Hebrew “it will be”} continually; the cloud would cover it and the appearance of fire by night.
\verse Whenever the cloud lifted up from on the tent, after that the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} would set out, and in the place where the cloud dwelled, there the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} camped.
\verse On the \textit{command of Adonai}\lebnote{Literally “mouth of Adonai”} the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} would set out, and on the \textit{command of Adonai}\lebnote{Literally “mouth of Adonai”} they encamped; all the days that the cloud dwelled on the tabernacle they encamped.
\verse And when the cloud prolonged on the tabernacle many days the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} kept the kept requirement of Adonai and did not set out.
\verse When the cloud \textit{remained}\lebnote{Literally “it was there”} a number of days on the tabernacle, on the \textit{command of Adonai}\lebnote{Literally “mouth of Adonai”} they encamped; and on the \textit{command of Adonai}\lebnote{Literally “mouth of Adonai”} they set out.
\verse When the cloud \textit{remained}\lebnote{Literally “it was there”} from evening until morning, and the cloud lifted up in the morning, they would set out, or if it remained in the daytime and at night, when the cloud lifted up they set out.
\verse When it was two days, a month, \textit{or a year}\lebnote{Literally “or days”} that the cloud prolonged to dwell on the tabernacle, the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} encamped, and they did not set out; when it lifted up they set out.
\verse On the \textit{command of Adonai}\lebnote{Literally “mouth of Adonai”} they encamped, and on the \textit{command of Adonai}\lebnote{Literally “mouth of Adonai”} they set out. They kept the requirement of Adonai, on the \textit{command of Adonai}\lebnote{Literally “mouth of Adonai”} in the hand of Moses.\lebnote{Or “through Moses”}
\end{biblechapter}

\begin{biblechapter} % Numbers 10
\verseWithHeading{The Silver Trumpets} Adonai spoke to Moses, saying,
\verse “Make yourself two silver trumpets; make them of hammered-work. \textit{You will use them}\lebnote{Literally “They will be for you”} for calling the community and for breaking the camp.
\verse You will blow them, and all the community will assemble to the doorway of the tent of assembly.
\verse But if they blow only one, the leaders, the heads of the thousands of Israel, will assemble to you.
\verse When you will blow a blast, the camps that are camping on the east will set out;
\verse when you blow a second blast, the camps that are camping on the south will set out; they will blow a blast for their journeys.
\verse But when summoning the assembly, you will blow, but you will not signal with a loud noise.
\verse The sons of Aaron, the priests, will blow on the trumpets; this will be an eternal decree for your generations.
\verse If you go to war in your land against the enemy who attacks you, you will signal with a loud noise on the trumpets. You will be remembered \textit{before}\lebnote{Literally “in the presence of”} Adonai your God, and you will be rescued from your enemies.
\verse “And on the day of your joy and in your appointed times, at the beginning of your months, you will blow on the trumpets in addition to your burnt offerings and in addition to the sacrifices of your fellowship offerings. And they will be as a memorial for you \textit{before}\lebnote{Literally “in the presence of”} your God; I am Adonai your God.”
\verseWithHeading{The Israelites Depart from Sinai} And it happened, in the second year, in the second month, on the twentieth of the month the cloud was lifted from upon the tabernacle of the testimony.\lebnote{Some modern translations (e.g., the NRSV) have “tabernacle of the covenant”}
\verse And the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} set out for their journey\lebnote{Hebrew “journeys”} from the desert of Sinai, and the cloud dwelled in the desert of Paran.
\verse They set out \textit{for the first time}\lebnote{Literally “in the beginning”} on the command of Adonai in the hand of Moses.\lebnote{Or “through Moses”}
\verse The standard of the camp of the descendants\lebnote{Or “sons”} of Judah set out for the first time according to their divisions, with Nahshon son of Amminadab over its division.
\verse And Nathanel son of Zuar was over the division of the descendants\lebnote{Or “sons”} of Issachar;
\verse Eliab son of Helon was over the division of the tribe of the descendants\lebnote{Or “sons”} of Zebulun.
\verse The tabernacle was taken down, and the sons of Gershon and the sons of Merari, the bearers of the tabernacle, set out.
\verse And the standard of the camp of Reuben according to their divisions; Elizur son of Shedeur was over their division.
\verse Shelumiel son of Zurishaddai was over the division of the sons of the tribe of Simeon.
\verse Eliasaph son of Deuel was over the division of the tribe of the descendants\lebnote{Or “sons”} of Gad.
\verse The Kohathites, the bearers of the sanctuary, set out, and they set up the tabernacle before they arrived.
\verse And the stand of the camp of the descendants\lebnote{Or “sons”} of Ephraim set out according to their divisions; Elishama son of Ammihud was over its division.
\verse Gamaliel son of Pedahzur was over the division of the tribe of the descendants\lebnote{Or “sons”} of Manasseh.
\verse Abidan son of Gideoni was over the division of the tribe of the descendants\lebnote{Or “sons”} of Benjamin.
\verse Then the standard of the camp of the descendants\lebnote{Or “sons”} of Dan, who formed a rear guard for all the camps, set out according to their divisions; Ahiezer son of Ammishaddai was over its division.
\verse Pagiel son of Ocran was over the division of the tribe of the descendants\lebnote{Or “sons”} of Asher.
\verse Ahira son of Enan was over the division of the tribe of the descendants\lebnote{Or “sons”} of Naphtali.
\verse These were the departures of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} according to their divisions; and so they set out.
\verse Moses said to Hobab son of Reuel the Midianite, the father-in-law of Moses, “We are setting out to the place that Adonai said, ‘I will give it to you’; go with us, and we will \textit{treat you well}\lebnote{Literally “do good to you”} because \textit{Adonai promised}\lebnote{Literally “Adonai spoke”} good concerning Israel.”
\verse But he said to him, “I will not go. I will only go to my land and to my family.”
\verse He\lebnote{That is, Moses} said, “Please, do not abandon us because you know our encampment in the desert, and \textit{you should be our guide}.\lebnote{Literally “you should be our eyes”}
\verse Moreover, if you go with us, the good that Adonai will do to us we will do to you.”
\verse And so they set out from the mountain of Adonai a journey of three days, with the ark of the covenant of Adonai setting out ahead of them\lebnote{Or “before them”} three days’ journey to search out a resting place for them;
\verse and the cloud of Adonai was over them by day when they set out from the camp.
\verse And whenever the ark was setting out Moses would say,
\verse “Rise up, Adonai! 
May your enemies be scattered; 
may the ones that hate you flee from your presence.”
\end{biblechapter}

\begin{biblechapter} % Numbers 11
\verseWithHeading{The Israelites Complain} And it happened, the people were like those who \textit{complain of hardship}\lebnote{Literally “complain of bad”} \textit{in the hearing}\lebnote{Literally “in the ears”} of Adonai, and Adonai \textit{became angry},\lebnote{Literally “his nose became hot”} and the fire of Adonai burned among them, and it consumed the edge of the camp.
\verse Then the people cried out to Moses, and Moses prayed to Adonai, and the fire died down.
\verse And he called the name of that place Taberah\lebnote{This word is difficult, but some modern translations suggest the word in Hebrew means “burning” (see NRSV, NASB)} because the fire of Adonai burned among them.
\verse The riff-raff that were in their midst \textit{had a strong desire};\lebnote{Literally “desired a desire”} and the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} turned back and also wept, and they said, “Who will feed us meat?
\verse We remember the fish that we ate in Egypt for nothing, the cucumber, melon, leek, the onions, and the garlic.
\verse But now \textit{our strength is dried up};\lebnote{Literally “our life is dry”} there is nothing whatsoever except \textit{for the manna before us}.”\lebnote{Literally “for the manna of our eyes”}
\verse Now the manna was like coriander seed, and its outward appearance was like that of bdellium-gum.
\verse The people went about and gathered it, and they ground it with mills or crushed it with mortar. Then they boiled it in a pot and made it into bread-cakes; and it tasted like olive oil cakes.
\verse When the dew came down on the camp at night, the manna came down with it.
\verse Moses heard the people weeping according to their\lebnote{Hebrew “its”} clans, each at the doorway of their tents. Then \textit{Adonai became very angry},\lebnote{Literally “the nose of Adonai became very hot”} and in the eyes of Moses it was bad.
\verse And Moses said to Adonai, “Why have you brought trouble to your servant? Why have I not found favor in your eyes, that the burdens of all these people have been placed on me?
\verse Did I conceive all these people? If I have fathered them,\lebnote{Hebrew “him/it”} that you could say to me, ‘Carry them\lebnote{Hebrew “him/it”} in your lap, just as a foster-father carries the suckling on the land that you swore an oath to their ancestors?’\lebnote{Or “fathers”}
\verse From where do I have meat to give all these people? They weep before me, saying, ‘Give us meat and let us eat!’
\verse I am not able to carry all these people along alone; they are too heavy for me.
\verse If this is how you are going to treat me, please kill me immediately if I find favor in your eyes, and do not let me see my misery.”
\verse And Adonai said to Moses, “Gather for me seventy men from the elders of Israel whom you know are elders of the people and their\lebnote{Hebrew “his/its”} officials; take them to the tent of assembly, and they will stand there with you.
\verse I will come down and speak with you there; I will take away from the spirit that is on you, and I will place it on them; and they will bear the burdens of the people with you; you will not bear it alone.
\verse And you will say to the people, ‘Sanctify yourselves tomorrow, for you will eat meat because you have wept in the ears of Adonai, saying, “Who will feed us good meat? It was good for us in Egypt.” Adonai will give to you meat, and you will eat.
\verse You will eat, not one day, or two days, or five days, or ten days, or twenty days,
\verse \textit{but for a whole month},\lebnote{Literally “until a period of one month”} until it comes out from your nose and becomes as nausea to you; because you have rejected Adonai, who is in your midst, and you wept before \textit{his presence},\lebnote{Literally “before his face”} saying, “Why did we ever leave Egypt?” ’ ”
\verse But Moses said, “There are six hundred thousand on foot, among whom I am in the midst, and you yourself said, ‘I will give meat to them, and they will eat for a whole month.’
\verse Should flocks and cattle be slaughtered for them? Should all the fish of the sea be gathered together for them, to be enough for them?”
\verse And Adonai said to Moses, “\textit{Is Adonai’s power limited}?\lebnote{Literally “Is Adonai’s hand short?”} Now you will see if my word will happen or not.”
\verse So Moses went out, and he spoke the words of Adonai to the people, and he gathered together seventy men from the elders of the people, and he \textit{made them stand}\lebnote{Literally “caused them to stand”} all around the tent.
\verse Then Adonai went down in the cloud and spoke to him, and he took away the spirit that was on him, and he \textit{put it}\lebnote{Or “gave it”} on the seventy elders. And as soon as the spirit was resting on them they prophesied, but they did not do it again.
\verse But two men were left in the camp; the name of one was Eldad, and the name of the second was Medad, and the spirit rested on them; they were among those who were written down, but they did not go out to the tent, so they prophesied in the camp.
\verse So a boy\lebnote{Hebrew “the boy”} ran and told Moses and said, “Eldad and Medad are prophesying in the camp.”
\verse And Joshua son of Nun, the assistant of Moses from time of his youth, answered, “Moses, my lord, stop them.”
\verse But Moses said to him, “Are you jealous for my sake? Would that he\lebnote{That is, Adonai} give all Adonai’s people prophets, that Adonai put his spirit on them!”
\verse Then Moses and the elders of Israel were gathered to the camp.
\verseWithHeading{The Quail} Then a wind set out from Adonai, and it drove quails from the west, and he spread them out on the camp about a day’s journey on one side and about a day’s journey on the other, all around the camp, about two cubits on the surface of the land.
\verse And so the people \textit{worked}\lebnote{Literally “arose”} all day and all night and all the next day, and they gathered the quail (the least of the ones collecting gathered ten homers).\lebnote{HALOT 330, “a dry measure”}
\verse While the meat was still between their teeth, before it was consumed, Adonai was angry with the people, and Adonai struck a very great plague among the people.
\verse And he called the name of that place Kibroth Hattaavah\lebnote{Hebrew “the graves of greediness”} because they buried the people that \textit{were greedy}.\lebnote{Literally “craved”}
\verse From Kibroth Hattaavah\lebnote{Hebrew “the graves of greediness”} the people set out to Hazeroth; and they stayed\lebnote{Hebrew “they were”} in Hazeroth.
\end{biblechapter}

\begin{biblechapter} % Numbers 12
\verseWithHeading{Aaron and Miriam Murmur Against Moses} And Miriam and Aaron spoke against Moses because of the Cushite woman whom he took (because he took a Cushite wife);
\verse and they said, “Has Adonai spoken only through Moses? Has not Adonai also spoken through us?” And Adonai heard it.
\verse Now the man, Moses, was more humble than any other person on the face of the earth,
\verse and Adonai said suddenly to Moses, Aaron, and Miriam, “Go out, you three, to the tent of assembly.” So the three of them when out.
\verse And Adonai went down in a column of cloud and stood at the doorway of the tent, and he called Aaron and Miriam, and the two of them went,
\verse and he said,
\verse “Please hear my words: 
If there is a prophet among you, I, Adonai, 
will make myself known to him in a vision. 
I will speak to him in a dream.\lebnote{Hebrew “the dream”}
\verse Not so with my servant Moses; 
in all my house he is faithful. Why were you not afraid to speak against my servant, against Moses?”
\verse \textit{And Adonai became very angry}\lebnote{Literally “And the nose of Adonai became hot”} with them, and he went away.
\verse And the cloud departed from on the tent, and behold, Miriam was infected with \textit{a skin disease}\lebnote{The precise meaning is uncertain; many modern translations suggest “leprosy”} white like snow; when Aaron turned toward Miriam, behold, she was afflicted with a skin disease.
\verse So Aaron said to Moses, “Please, my lord, please do not put on us this sin in which we were foolish and in which we have sinned.
\verse Please do not let her be like the dead, whose flesh is half consumed when coming out from the womb of its mother.”
\verse And Moses cried to Adonai, saying, “God, \textit{please heal her}!”\lebnote{Literally “Please heal please her”}
\verse But Adonai said to Moses, “If her father had surely spit in her face, would she not bear her shame for seven days? Let her be confined for seven days to an outside place of the camp, and afterward she may be gathered.”
\verse So Miriam was confined to the outside place of the camp seven days, and the people did not set out until Miriam was gathered.
\verse And afterward the people set out from Hazeroth, and they encamped in the desert of Paran.
\end{biblechapter}

\begin{biblechapter} % Numbers 13
\verseWithHeading{Spies Sent to Spy Out the Land of Canaan} And Adonai spoke to Moses, saying,
\verse “Send for yourself men, and let them explore the land of Canaan, which I am about to give to the \textit{Israelites};\lebnote{Literally “sons/children of Israel”} \textit{from each tribe of his father send one man},\lebnote{Literally “one man one man from the tribe of his father”} everyone a leader among them.”
\verse So Moses sent them from the desert of Paran on the command of Adonai; all of the men were \textit{leaders}\lebnote{Literally “heads”} of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse And these are their names: from the tribe of Reuben, Shammua son of Zaccur;
\verse from the tribe of Simeon, Shaphat son of Hori;
\verse from the tribe of Judah, Caleb son of Jephunneh;
\verse from the tribe of Issachar, Igal son of Joseph;
\verse from the tribe of Ephraim, Hoshea son of Nun;
\verse from the tribe of Benjamin, Palti son of Raphu;
\verse from the tribe of Zebulun, Gaddiel son of Sodi;
\verse from the tribe of Joseph, from the tribe of Manasseh, Gaddi son of Susi;
\verse from the tribe of Dan, Ammiel son of Gemalli;
\verse from the tribe of Asher, Sethur son of Michael;
\verse from the tribe of Naphtali, Nahbi son of Vophsi;
\verse from the tribe of Gad, Geuel son of Maki.
\verse These are the names of the men whom Moses sent to explore the land. And Moses called Hoshea son of Nun Joshua.
\verse Moses sent them to explore the land of Canaan, and he said to them, “Go up like this to the Negev,\lebnote{An arid region south of the Judean hills} and go up into the hill country,
\verse and you will see what the land is like and if the people who inhabit it are strong or weak, or whether they are few or many,
\verse and whether the land that they are inhabiting is good or bad, and whether the cities they are inhabiting are camps or fortifications,
\verse and whether the land is fertile or lean, and whether there are trees on it or not. You will show yourself courageous, and you will take some of the fruit of the land.” It was the time of first ripe grapes.
\verse So they went up and explored the land from the desert of Zin until Rehob, at Lebo Hamath.\lebnote{Or “near Hamath”}
\verse They went up through the Negev\lebnote{An arid region south of the Judean hills} and came to Hebron, where\lebnote{Hebrew “and there”} Ahiman, Sheshai, and Talmai the descendants of the Anakites were. (Hebron was built seven years before Zoan in Egypt.)
\verse And they came up to the valley\lebnote{Or “wadi”} of Eshcol, and they cut off a vine branch and one cluster of grapes from there; they carried it on a pole between two men, with pomegranates and figs.
\verse That place he called the valley\lebnote{Or “wadi”} of Eshcol on account of the cluster of grapes that the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} cut off from there.
\verseWithHeading{The Spies Return} They returned from exploring the land at the end of forty days.\lebnote{Hebrew “day”}
\verse And they came\lebnote{Hebrew “they went and came”} to Moses and Aaron and to the entire community of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} in the desert of Paran at Kadesh; they brought back word to them and to all the community, and they showed them the fruit of the land.
\verse And they told him,\lebnote{Hebrew “they told him and said”} “We came to the land that you sent us, and it is flowing of milk and honey; this is its fruit.
\verse Yet the people who are inhabiting it are strong and the cities are fortified and very large; moreover, we saw the descendants of the Anakites there.
\verse The Amalekites are living in the land of the Negev;\lebnote{An arid region south of the Judean hills} the Hittites, Jebusites, and the Amorites are living in the hill country; and the Canaanites are living at the sea and on the banks of the Jordan.”
\verse And Caleb silenced the people before Moses and said, “Surely, let us go up and let us take possession of it because surely we will be able to prevail over it.”
\verse And the men who went up with him said, “We are not able to go up to the people because they are stronger than us.”
\verse And they presented the report of the land that they explored to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} saying, “The land that we went through to explore is a land that eats its inhabitants, and all the people whom we saw in its midst are \textit{men of great size}.\lebnote{Literally “men of measurements”}
\verse There we saw the Nephilim (the descendants\lebnote{Or “sons”} of Anak came from the Nephilim), and we were like grasshoppers in our own sight, and so we were in their eyes.”
\end{biblechapter}

\begin{biblechapter} % Numbers 14
\verseWithHeading{The People Complain} Then all the community \textit{lifted up their voices},\lebnote{Hebrew “they lifted up and gave their voice”} and the people wept during that night.
\verse And all the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} grumbled against Moses and Aaron, and all the community said to them, “If only we had died in the land of Egypt or in this desert!
\verse Why did Adonai bring us into this land to fall by the sword? Our wives and our little children will become plunder; would it not be better for us to return to Egypt?”
\verse \textit{They said to each other},\lebnote{Literally “A man said to his brother”} “Let us appoint a leader, and we will return to Egypt.”
\verse Then Moses and Aaron fell on their faces \textit{before}\lebnote{Literally “in the presence of”} the assembly of the community of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse Joshua son of Nun and Caleb son of Jephunneh, from the explorers of the land, tore their garments.
\verse And they said to all the community of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} “The land that we went through to explore is an \textit{exceptionally good land}.\lebnote{Literally “the land is very very good”}
\verse If Adonai delights in us, then he will bring us into this land, and he will give it to us, a land that is flowing with milk and honey.
\verse Only do not rebel against Adonai, and you will not fear the people of the land, because they will be our food. \textit{Their protection}\lebnote{Literally “Their shadow”} has been turned from them; Adonai is with us. You should not fear them.”
\verse And all the community said to stone them with stones, but the glory of Adonai appeared in the tent of assembly among the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse And Adonai said to Moses, “How long until this people will despise me, and how long until they will not believe in me, and in all the signs that I have done in their\lebnote{Hebrew “his”} midst?
\verse I will strike them\lebnote{Hebrew “him”} with disease, and I will dispossess them;\lebnote{Hebrew “him”} I will make you into a greater and stronger nation than them.”\lebnote{Hebrew “him”}
\verse And Moses said to Adonai, “Then the Egyptians will hear that you brought up this people from their\lebnote{Hebrew “his”} midst in your power,
\verse and they will \textit{tell it}\lebnote{Literally “say”} to the inhabitants of this land. They heard that you, Adonai, are in the midst of this people, that you are seen eye to eye, and your cloud is standing over them, and in a column of cloud you go before them by day and in a column of fire at night.
\verse But if you destroy this people \textit{all at once},\lebnote{Literally “as one man”} the nations that will have heard your message will say,
\verse ‘Adonai was unable to bring this people in the land that he swore by an oath, and he slaughtered them in the desert.’
\verse But now, please, let the power of my Lord be great, just has you spoke,
\verse ‘Adonai is \textit{slow to anger}\lebnote{Literally “slow of noses”} 
and great of loyal love, 
\textit{forgiving}\lebnote{Literally “lifting up”} sin and rebellion; 
but surely he leaves nothing unpunished, 
visiting the sin of the fathers on the sons 
to the third and fourth generations.’
\verse Please forgive the sin of this people according to the greatness of your loyal love, just as you \textit{forgave}\lebnote{Literally “lifted up”} this people, from Egypt until now.”
\verse Adonai said, “I have forgiven them according to your word;
\verse but as I am alive, the glory of Adonai will fill all the earth.
\verse But because all the men who have seen my glory and my signs that I did in Egypt and in the desert yet tested me these ten times and did not listen to my voice,
\verse they will not see the land that I swore by oath to their ancestors,\lebnote{Or “fathers”} and all those who despised me will not see it.
\verse But my servant Caleb, because another spirit was with him, he remained true after me, and I will bring him into the land that \textit{he entered},\lebnote{Or “he went to”} and his offspring will take possession of it.
\verse And the Amalekites and the Canaanites live in the valleys; tomorrow turn and set out for the desert by way of the \textit{Red Sea}.”\lebnote{Literally “sea of reed”}
\verse And Adonai spoke to Moses and Aaron, saying,
\verse “How long will I bear this evil community who are grumbling against me? I have heard the grumbling of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} which \textit{they are making}\lebnote{Literally “they are grumbling”} against me.
\verse Say to them, ‘Surely as I live,’ \textit{declares}\lebnote{Literally “declaration of”} Adonai, ‘just as you spoke \textit{in my hearing},\lebnote{Literally “in my ears”} so I will do to you;
\verse in this desert your corpses will fall, and all your counted ones, according to all your number, from \textit{twenty years old}\lebnote{Literally “a son of twenty years”} and above who grumbled against me.
\verse You yourselves will not come into the land that \textit{I swore by oath}\lebnote{Literally “I lifted up my hand”} to make you to dwell in it, but Caleb son of Jephunneh and Joshua son of Nun.
\verse But your little children, whom you said would be plunder, I will bring them, and they will know the land that you rejected.
\verse But for you, all your corpses will fall in this desert.
\verse And your children will be shepherds in the desert forty years,\lebnote{Hebrew “year”} and you will bear your unfaithfulness until \textit{all your corpses have fallen}\lebnote{Literally “until to complete your corpses”} in the desert.
\verse According to the number of the days\lebnote{Hebrew “day”} that you explored the land, forty days,\lebnote{Hebrew “day”} \textit{a day for each year},\lebnote{Literally “a day for a year a day for a year”} you will bear your sins forty years,\lebnote{Hebrew “year”} and you will know my opposition.’
\verse I, Adonai, have spoken; I will surely do this to all this evil community who has banded together against me. In this desert they will come to an end, and there they will die.”
\verse As for the men whom Moses sent to explore the land, who returned and made the community grumble against him by spreading a report over the land,
\verse the men who spread the evil report of the land died by the plague \textit{before Adonai}.\lebnote{Literally “in the presence of Adonai”}
\verse But Joshua son of Nun and Caleb son of Jephunneh lived from among the men who went to explore the land.
\verse And Moses spoke words to all the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and the people mourned greatly.
\verse They rose early in the morning and went to the top of the mount, saying, “Here we are. We will go up to the place that Adonai said, because we have sinned.”
\verse But Moses said, “Why are you going against the command of Adonai? It will not succeed.
\verse You should not go up because Adonai is not in your midst; do not let yourselves be defeated in the presence of your enemies,
\verse because the Amalekites\lebnote{Hebrew “Amalekite”} and the Canaanites\lebnote{Hebrew “Canaanite”} are there \textit{before you},\lebnote{Literally “in your presence”} and you will fall by the sword; because you have turned back from Adonai, and Adonai will not be with you.”
\verse But they dared to go to the top of the mountain, and the ark of the covenant of Adonai and Moses did not depart from the midst of the camp.
\verse So the Amalekites\lebnote{Hebrew “Amalekite”} and the Canaanites\lebnote{Hebrew “Canaanite”} who were living on the mountain descended, and they beat them down, up to Hormah.
\end{biblechapter}

\begin{biblechapter} % Numbers 15
\verseWithHeading{Various Sacrifices and Offerings} Adonai spoke to Moses, saying,
\verse “Speak to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} and say to them, ‘When you come into the land of your dwellings that I am about to give to you,
\verse you will make an offering by fire for Adonai from the cattle or from the flock, a burnt offering or a sacrifice to fulfill a vow, or as a freewill offering or at your feasts, to make a fragrance of appeasement for Adonai.
\verse And the one who presents an offering\lebnote{Hebrew “his offering”} for Adonai, he will present a grain offering of finely milled flour; a tenth will be mixed with a fourth of the liquid measure of oil;
\verse and you will add a fourth of wine for the libation upon the burnt offering, or to the sacrifice for each ram-lamb.
\verse Or for the ram you will make a grain offering of two-tenths of finely milled flour mixed into a third of a liquid measure of oil.
\verse You will present a third of the liquid measure of wine for the libation, a fragrance of appeasement for Adonai.
\verse When you prepare \textit{a bull}\lebnote{Literally “a son of cattle”} as a burnt offering or a sacrifice to fulfill a vow or a fellowship offering for Adonai,
\verse you will present with \textit{the bull}\lebnote{Literally “the son of the cattle”} a grain offering of three-tenths of finely milled flour mixed with half a liquid measure of oil,
\verse and you will present half a liquid measure of wine as a libation, as an offering made by fire, a fragrance of appeasement for Adonai.
\verse “ ‘This is how it should be done for each bull, or for the each ram, or for the small four-footed mammal, or ram-lambs, or goats.
\verse According to the number that you prepare, so should you do to each according to their number.
\verse Every native must do these things to present an offering made by fire, a fragrance of appeasement for Adonai.
\verse If an alien dwells among you, or whoever is in your midst throughout your generations,\lebnote{Hebrew “for your generations”} and prepares an offering made by fire, a fragrance of appeasement for Adonai, he should do as you do.
\verse For the assembly, there will be one decree for you and for the alien who dwells among you; it is an eternal decree for all your generations. \textit{You as well as the alien}\lebnote{Literally “like you like the alien”} will be \textit{before Adonai}.\lebnote{Literally “in the presence of Adonai”}
\verse There will be one law and one stipulation for you and for the alien dwelling among you.’ ”
\verse Adonai spoke to Moses, saying,
\verse “Speak to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} and say to them, ‘When you come into the land to which I am about to bring you,
\verse whenever you eat from the food of the land, you will lift up a contribution to Adonai.
\verse You must lift up a contribution of the first batch of your ring-shaped dough bread; you must lift it up as a contribution of the threshing floor.
\verse You will give to Adonai a contribution from the first of your dough throughout your generations.\lebnote{Hebrew “for your generations”}
\verse “ ‘But if you go astray and you do not follow\lebnote{Or “do”} all these commandments that Adonai commanded to Moses,
\verse all that Adonai commanded you by the hand of Moses\lebnote{Or “through Moses”} from the day that Adonai commanded and beyond, throughout your generations,\lebnote{Hebrew “for your generations”}
\verse and if it was done unintentionally \textit{without the knowledge}\lebnote{Literally “from the eyes”} of the community, then the entire community must prepare one \textit{young bull}\lebnote{Literally “a bull a son of cattle”} as a burnt offering, as a fragrance of appeasement for Adonai, and its grain offering and its libation, according to the stipulation, and one male goat as a sin offering.
\verse The priest will make atonement for all of the community of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and \textit{they will be forgiven}\lebnote{Literally “it will be forgiven to them”} because it was unintentional; they will bring their offering, an offering made by fire for Adonai, their sin offering \textit{before Adonai}\lebnote{Literally “in the presence of Adonai”} for their unintentional sin.
\verse All of the community of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} will be forgiven, as well as the alien that dwells in their midst, because the whole community was involved in the unintentional wrong.
\verse “ ‘If one person sins unintentionally, that person will present a female goat \textit{in its first year}\lebnote{Literally “a daughter of a year”} as a sin offering.
\verse And the priest will make atonement for the person who \textit{sinned unintentionally}\lebnote{Literally “sinned unintentionally when sinning an unintentional wrong”} \textit{before Adonai},\lebnote{Literally “in the presence of Adonai”} to make atonement for him, and he will be forgiven.
\verse For the native among the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} and the alien that dwells in their midst, there will be one law for anyone who commits an unintentional wrong.
\verse But the one \textit{who acts presumptuously}\lebnote{Literally “who acts with a high hand”} from among the native or alien blasphemes against Adonai, and that person must be cut off from the midst of the people.
\verse Because he despised the word of Adonai and broke his command, that person will be surely cut off and bear the guilt.’ ”
\verseWithHeading{Violation of the Sabbath} When the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} were in the desert, they found a man who was gathering wood on the day of the Sabbath.
\verse The ones who found him gathering wood brought him to Moses, Aaron, and to all the community.
\verse And they put him under watch because it was not made clear what should be done to him.
\verse And Adonai said to Moses, “Surely the man must be put to death by stoning him; all the community must stone him with stones from outside the camp.”
\verse So the entire community brought him out to a place outside the camp, and \textit{they stoned him to death}\lebnote{Literally “they stoned him with stones and he died”} just as Adonai commanded Moses.
\verseWithHeading{Garment Fringes} Adonai spoke to Moses, saying,
\verse “Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and tell them to make for themselves tassels\lebnote{Hebrew “tassel”} on the hems of their garments throughout their generations\lebnote{Hebrew “for their generations”} and to put a blue cord on the tassel of the hem.
\verse You will have a tassel \textit{for you to look at}\lebnote{Literally “and you will look at it”} and remember all the commands of Adonai and do them, and not follow \textit{after the unfaithfulness of your own heart and eyes},\lebnote{Literally “after your heart and after your eyes, which you are unfaithful after them”}
\verse so that you will remember and do all my commandments, and you will be holy for your God.
\verse I am Adonai your God, who brought you out of the land of Egypt, to be your God; I am Adonai your God.”
\end{biblechapter}

\begin{biblechapter} % Numbers 16
\verseWithHeading{Korah, Dathan, and Abiram Rebel} Now Korah son of Izhar, son of Kohath, son of Levi, and Dathan and Abiram sons of Eliab, and On son of Peleth, the descendants\lebnote{Or “sons”} of Reuben,
\verse took two hundred and fifty men from the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} leaders of the community summoned from the assembly, \textit{renowned men},\lebnote{Literally “men of name”} and \textit{they confronted}\lebnote{Literally “they rose up before”} Moses.
\verse They were assembled in front of Moses and Aaron, and they said to them, “\textit{You take too much upon yourselves}!\lebnote{Literally “It is much for you”} All of the community is holy, every one of them, and Adonai is in their midst, so why do you raise yourselves over the assembly of Adonai?”
\verse When Moses heard this, he fell on his face.
\verse And he said to Korah and to his entire company, saying, “Tomorrow morning Adonai will make known who is his and who is holy, and he will bring him near to him, whomever he chooses he will bring near to him.
\verse Do this: take for yourselves censers, Korah and all of your company;\lebnote{Hebrew “his company”}
\verse tomorrow put fire in them and place incense on them \textit{before}\lebnote{Literally “in the presence of”} Adonai; the man whom Adonai chooses will be the holy one. You take too much upon yourselves, sons of Levi!”
\verse And Moses said to Korah, “Please listen, sons of Levi!
\verse Is it too little for you that the God of Israel set you apart from the community of Israel \textit{to allow you to approach him}\lebnote{Literally “to bring you near to him”} to do the work of the tabernacle of Adonai, to stand \textit{before}\lebnote{Literally “in the presence of”} the community to serve them?
\verse \textit{He has allowed you to approach him},\lebnote{Literally “He has brought you near”} you with all your brothers, the descendants\lebnote{Or “sons”} of Levi, but yet you also seek the priesthood.
\verse Therefore you and your company that has banded together against Adonai. What is Aaron that you grumble against him?”
\verse Moses sent to call for Dathan and Abiram son of Eliab, but they said, “We will not come!\lebnote{Hebrew “We will not come up!”}
\verse Is it too little that you have brought us from a land that flows with milk and honey to kill us in the desert, and that you also appoint yourself as a ruler over us?
\verse Surely, you have not brought us to a land that flows with milk and honey, and you have not given us the inheritance of fields and a vineyard. Will you gouge out the eyes of these men? We will not come!”\lebnote{Hebrew “We will not come up!”}
\verse Then Moses became angry, and he said to Adonai, “Do not notice their grain offering. I have not offered one donkey from them, and I have not mistreated one of them.”
\verse And Moses said to Korah, “You and your entire company will be \textit{before}\lebnote{Literally “in the presence of”} Adonai tomorrow, you and they and Aaron.
\verse Each one take his censer, and put incense on it\lebnote{Hebrew “on them”} and you will present it \textit{before}\lebnote{Literally “in the presence of”} Adonai, and each of you bring his censer, two hundred and fifty censers, you and Aaron, each his censer.”
\verse So each of them took his censer, and they put fire on them, and they placed incense on them; they stood at the doorway of the tent of the assembly of Moses and Aaron.
\verse And Korah summoned them, the entire community, by the doorway of the tent of assembly, and the glory of Adonai appeared to all the community.
\verse And Adonai spoke to Moses and Aaron, saying,
\verse “Separate yourselves from the midst of this community, that I can destroy them in a moment.”
\verse And they fell on their faces, and they said, “God, God of the spirits of all flesh, will one man sin and you become angry toward the entire community?”
\verse Adonai spoke to Moses, saying,
\verse “Speak to the community, saying, ‘Move away from the dwelling of Korah, Dathan, and Abiram.’ ”
\verse So Moses stood up and went to Dathan and Abiram; the elders of Israel followed after him.
\verse He said to the community, saying, “Please turn away from the tents of these wicked men, and do not touch anything \textit{that belongs to them},\lebnote{Literally “that is to them”} or you will be destroyed with all their sins.”
\verse And so they moved away from around the dwellings\lebnote{Hebrew “dwelling”} of Korah, Dathan, and Abiram; and Dathan and Abiram came out standing at the doorway of their tents, with their wives, sons, and little children.
\verse And Moses said, “In this you will know that Adonai has sent me to do all these works; it is not from my heart.
\verse \textit{If they die a natural death}\lebnote{Literally “If they die like the death of every human”} or \textit{if a natural fate is visited upon them},\lebnote{Literally “If the fate of every human is visited upon them”} Adonai has not sent me.
\verse But if Adonai creates something new, and the ground opens its mouth and swallows them up and \textit{all that belongs to them},\lebnote{Literally “All that is for them”} and they go down alive to Sheol, and you will know that these men have despised Adonai.”
\verse And it happened, as soon as he finished \textit{speaking}\lebnote{Literally “to speak”} all these words, the ground that was under them split open.
\verse The land opened its mouth and swallowed them up with their houses and every person \textit{that belonged to Korah}\lebnote{Literally “that was to Korah”} and all the property.
\verse They went down alive to Sheol, they and all that belonged to them, and the land covered over them, and they perished from the midst of the assembly.
\verse All Israel who were around them fled at their cry, because they said, “Lest the land swallow us up!”
\verse And fire went out from Adonai, and it consumed the two hundred and fifty men presenting the incense.
\verse \lebnote{Numbers 16:36–17:13 in the English Bible is 17:1–28 in the Hebrew Bible} And Adonai spoke to Moses, saying,
\verse “Say to Eleazar son of Aaron the priest, ‘\textit{Take out}\lebnote{Literally “raise up”} the censers from among the place of burning because they are sacred, and scatter the fire outside.
\verse The censers of these \textit{who have sinned}\lebnote{Literally “sinners”} at the cost of their lives, let them be made into gilded leafing plating for the altar; because they presented them \textit{before Adonai},\lebnote{Literally “in the presence of Adonai”} they are holy; and they will be a sign for the \textit{Israelites}.’ ”\lebnote{Literally “sons/children of Israel”}
\verse Eleazar the priest took the bronze censers that the ones who were burned presented, and they hammered them out thinly as plating for the altar;
\verse it was a memorial for the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} so that \textit{no strange man}\lebnote{NASB translates “no layman”} who is not from \textit{the offspring}\lebnote{Literally “the seed”} of Aaron should approach the presence of Adonai to burn a smoke offering;\lebnote{That is, an incense offering} he will not be like Korah and his company, just as Adonai had spoken to him by the hand of Moses.\lebnote{Or “through Moses”}
\verseWithHeading{The Israelites Grumble} The next day all the community of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} grumbled against Moses and Aaron, saying, “You have killed the people of Adonai!”
\verse Then, when the community had gathered against Moses and Aaron, they turned to the tent of assembly, and behold, the cloud covered it, and the glory of Adonai appeared.
\verse And Moses and Aaron came to the front of the tent of assembly,
\verse and Adonai spoke to Moses, saying,
\verse “Get away from the midst of this community, and I will finish them in an instant,”\lebnote{Or “in a moment”} but they fell on their faces.
\verse And Moses and Aaron said, “Take the censer, and put fire on it from the altar. Place incense on it, and bring it quickly to the community, and make atonement for them, because wrath went out from the presence of Adonai, and a plague has begun.”
\verse And so Aaron took it just as Moses had spoken, and he ran into the midst of the assembly, for behold, the plague had begun among the people; so he gave the incense and made atonement for the people.
\verse He stood between the dead and between the living, and the plague was stopped.
\verse Those who died by the plague were fourteen thousand seven hundred, besides those who died on account of Korah.
\verse Then Aaron returned to Moses at the doorway of the tent of assembly, and the plague was stopped.
\end{biblechapter}

\begin{biblechapter} % Numbers 17
\verseWithHeading{Aaron’s Staff Is Chosen} Adonai spoke to Moses, saying,
\verse “Speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and take from among them twelve staffs, \textit{a staff from each family}\lebnote{Literally “a staff from a house of a father”} from among all their leaders according to their families’ households. Write the name of each man on his staff,
\verse and the name of Aaron on the staff of Levi, because one staff is for the head of each of \textit{their families}.\lebnote{Literally “the house of their fathers”}
\verse You must then put them in the tent of assembly \textit{before}\lebnote{Literally “in the presence of”} the testimony\lebnote{Or “the statute”} where I meet with you.
\verse And it will happen, the man whom I will choose, his staff will blossom, and so I will rid from upon myself the grumblings of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} who are grumbling against you.”
\verse Moses spoke to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and all their leaders gave him a staff for each leader, one from \textit{each of their families},\lebnote{Literally “each from the house of their fathers”} twelve staffs, and the staff of Aaron was in the midst of their tribes.
\verse And Moses put the staffs before Adonai in the tent of testimony.
\verse Then the next day, Moses went into the tent of the testimony, and behold the staff of Aaron for the house of Levi blossomed and put forth a flower and produced blossoms, and it produced almonds.
\verse Then Moses brought out to all the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} all the staffs before the presence of Adonai, and they saw, and each man took his staff.
\verse And Adonai said to Moses, “Bring back the staff of Aaron before the testimony\lebnote{Or “the statute”} as a guard and sign for the children of rebellion, and let them finish their grumblings before me and not die.”
\verse So Moses did; just as Adonai commanded him, so he did.
\verse And the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} said to Moses, saying, “Look! We will die! We will be destroyed! All of us will perish!
\verse Anyone who approaches the tabernacle of Adonai will die. \textit{Will we all die}?”\lebnote{Literally “Will we all die to perish?”}
\end{biblechapter}

\begin{biblechapter} % Numbers 18
\verseWithHeading{The Duties of the Priests and Levities} Adonai said to Aaron, “You, your sons, and your family with you will bear the guilt of the sanctuary, and you and your sons with you will bear the guilt of your priesthood.
\verse Moreover, bring your brothers with you, the tribe of Levi the tribe of your father, that they may be joined to you and minister to you, you and your sons with you before the tent of testimony.
\verse They will keep your responsibility and the responsibility of all the tent, only they may not come near the vessels of the sanctuary and the altar, so both you and they will not die.
\verse They will be joined to you, and they will keep the responsibility of the tent of assembly for the entire service of the tent; a stranger may not come near you.
\verse You will keep the responsibility of the sanctuary and the responsibility of the altar, and there will no longer be wrath on the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse Look, I myself have chosen your brothers the Levites from the midst of the children. They are a gift to you given from Adonai to perform the work of the tent of assembly.
\verse But you with your sons will keep your priesthood to perform your priestly duties for everything at the altar\lebnote{Or “for all the things of the altar”} and for \textit{the area behind the curtain}.\lebnote{Literally “the house of the curtain”} I give you the priesthood as a gift, but the stranger who approaches will be put to death.”
\verseWithHeading{Portions for the Priests} Adonai spoke to Aaron, “Behold, I myself have given to you the responsibility of my contributions for all the holy objects of the \textit{Israelites};\lebnote{Literally “sons/children of Israel”} I have given them as a portion to you and your sons as an eternal decree.
\verse This will be for you from the sanctuary of the holy things from the fire; all of their offerings, from every grain offering, from every sin offering, and from every guilt offering which they will bring to me is \textit{a most holy thing}\lebnote{Literally “a holy object of holiness”} for you and your sons.
\verse You will eat it in the most holy place;\lebnote{Alternatively “as a holy object of holiness”} every male will eat it. It will be a holy object to you.
\verse This is also for you: the contribution of their gift of the wave offerings of the children Israel. I have given them to you and your sons and your daughters with you as an eternal decree; whoever is clean in your house may eat it.
\verse All the finest olive oil and all the finest new wine and their best grain that they have given to Adonai, I have given them to you.
\verse The firstfruits of all that is in their land that they present to Adonai will be for you; whoever is clean in your house may eat it.
\verse All consecrated possessions\lebnote{Hebrew “possession”} in Israel will be for you.
\verse All the first offspring of a womb of any creature that they offer to Adonai, whether human or animal, will be yours; you will surely redeem the firstborn of the human and the unclean firstborn of the animal.
\verse As to their price of redemption, from \textit{a one-month-old}\lebnote{Literally “a son of a month”} you will redeem them according to your proper value, five shekels of silver according to the shekel of the sanctuary, which is twenty gerah.
\verse Only the firstborn of an ox or the firstborn of a sheep or the firstborn of a goat you will not redeem; they are holy. Their blood you will sprinkle over the altar, and their fat you will turn into smoke as an offering made by fire, a fragrance of appeasement for Adonai.
\verse But their flesh will be for you like the breast section of the wave offering, and it will be for you like the right upper thigh.
\verse All the contributions of holiness that the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} offer to Adonai I have given to you and your sons and your daughters with you as an eternal decree; it is an eternal covenant of salt \textit{before}\lebnote{Literally “in the presence of”} Adonai to you and your offspring with you.”
\verse Then Adonai said to Aaron, “You will not receive an inheritance in their land, and there will not be a plot of ground for you in the midst of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse “Behold, I have given to the descendants\lebnote{Or “sons”} of Levi every tithe in Israel as an inheritance in return for their service, which they are doing, the work of the tent of assembly.
\verse The \textit{Israelites}\lebnote{Literally “sons/children of Israel”} will not come near again to the tent of assembly, or \textit{they will bear sin}\lebnote{The NRSV translates “they will incur guilt”} and die.
\verse The Levites\lebnote{Hebrew “Levite”} will perform the service of the tent of assembly, and they will bear their guilt, an eternal decree for all your generations. But they will not receive an inheritance in the midst of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”}
\verse because the tithes\lebnote{Hebrew “tithe”} of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} that are \textit{offered}\lebnote{Hebrew “raised up”} to Adonai as a contribution, I have given to the Levites as an inheritance; therefore I said to them, ‘They will not receive an inheritance in the midst of the \textit{Israelites}.’ ”\lebnote{Literally “sons/children of Israel”}
\verse Adonai spoke to Moses, saying,
\verse “You will speak to the Levites and say to them, ‘When you receive the tithe from the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} that I have given to you from them for your inheritance, \textit{you will present}\lebnote{Literally “you will raise up”} a contribution from it to Adonai, a tithe from a tithe.
\verse Your contribution will be credited to you like the grain from the threshing floor and like the produce from the press.
\verse So \textit{you will present}\lebnote{Literally “you will raise up”} your own contribution to Adonai from all your tithes that you receive from the \textit{Israelites};\lebnote{Literally “sons/children of Israel”} from it you will give the contribution of Adonai to Aaron the priest.
\verse From all your gifts \textit{you will present}\lebnote{Literally “you will raise up”} every contribution of Adonai, from all its fat, the part that is sacred.’
\verse You will say to them, ‘When \textit{you are presenting}\lebnote{Literally “you are raising up”} its fat, the rest will be credited to the Levites like a yield of the threshing floor and like a yield from the press.
\verse You may eat it anywhere, you and your household, because it is a wage in return for your service in the tent of assembly.
\verse You will not bear any sin because \textit{you have presented}\lebnote{Literally “you have raised up”} its fat; you will not defile the holy objects of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} or you will die.’ ”
\end{biblechapter}

\begin{biblechapter} % Numbers 19
\verseWithHeading{Ashes of the Red Heifer} And Adonai spoke to Moses and Aaron, saying,
\verse “This is the decree of the law that Adonai has commanded, saying, ‘Speak to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} and let them take to you a red heifer without a physical defect, on which a yoke \textit{has not been placed}.\lebnote{Literally “has not gone up”}
\verse And you will give it to Eleazar the priest, and it will be brought\lebnote{Or “he will bring it out”} out to a place outside the camp, and it will be slaughtered\lebnote{Or “he will slaughter it”} in his presence.
\verse Then Eleazar the priest will take some of its blood on his finger and spatter it toward the mouth of the tent of assembly seven times.
\verse The heifer will be burned\lebnote{Hebrew “The heifer will burn”} in his sight; its skin, its meat, and its blood, in addition to its offal, will burn.
\verse The priest will take cedar wood, hyssop, and crimson thread, and he will throw them in the midst of \textit{the burning heifer}.\lebnote{Literally “the burning of the heifer”}
\verse The priest will wash his garments and his body in the water, and afterward he will come to the camp; the priest will be unclean until the evening.
\verse The one who burns it will wash his garments and his body in water; he will be unclean until the evening.
\verse A clean man will gather the ashes of the heifer, and he will put them in a clean \textit{place outside the camp};\lebnote{Literally “an outside place of the camp”} it will be for the community of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} as a requirement for waters of impurity; it is a purification offering.
\verse The one who gathers the ashes of the heifer will wash his garments; he will be unclean until evening. It will be an eternal decree for the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} and for one who dwells as an alien in their midst.
\verse “ ‘The one who touches a corpse of \textit{any person}\lebnote{Literally “any human person”} will be unclean for seven days.
\verse He will purify himself on the third day, and on the seventh day he will be clean. If he does not purify himself on the third day, he will not be clean on the seventh day.
\verse Anyone who touches a corpse, the person of a human being who died, and does not purify himself, defiles the tabernacle of Adonai, and that person will be cut off from Israel because the waters of impurity were not sprinkled on him. He will still be unclean, and uncleanness is on him.
\verse “ ‘This is the law of a person who dies in a tent: everyone who comes into the tent and all who are in the tent will be unclean seven days.
\verse Every container that is opened that does not have a lid cord\lebnote{That is, “that does not have a lid tied shut”} on it is unclean.
\verse Anyone \textit{in the open field}\lebnote{Literally “upon the face of a field”} who touches \textit{one who has been slain},\lebnote{Literally “the dead of sword”} or a corpse, or a bone of a person, or a burial site, he will be unclean for seven days.
\verse For the unclean person they will take\lebnote{Hebrew “he will take”} from the powder of the \textit{burnt purification offering},\lebnote{Or “burning of the sin offering”} and they will put\lebnote{Hebrew “he will put”} running water into a container.
\verse A clean person will take hyssop and dip it into the water and sprinkle it on the tent and on all the objects and persons who were there, and on one who touched the bone, or the one slain, or the dead, or the burial site.
\verse The clean person will spatter the unclean on the third day and on the seventh day; and on the seventh day he will purify him, and he will wash his garments; he will bathe in the waters, and in the evening he will be clean.
\verse But the man who is unclean and does not purify himself, that person will be cut off from the midst of the assembly because he defiled the sanctuary of Adonai; the water of impurity was not sprinkled on him; he is unclean.
\verse “ ‘It will be an eternal decree for them. The one who spatters the waters of impurity will wash his garments, and the one who touches the waters of impurity will be unclean until the evening.
\verse Anything that the unclean person touches will be unclean, and the person who touches it will be unclean until the evening.’ ”
\end{biblechapter}

\begin{biblechapter} % Numbers 20
\verseWithHeading{Miriam Dies} Then the entire community of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} came to the desert of Zin on the first month, and the people stayed in Kadesh; Miriam died and was buried there.
\verse There was no water for the community, and they were gathered before Moses and Aaron.
\verse And the people quarreled with Moses and spoke, saying, “If only we died when our brothers were dying \textit{before}\lebnote{Literally “in the presence of”} Adonai!
\verse Why have you brought the assembly of Adonai, us and our livestock, into this desert to die here?
\verse Why have you brought us from Egypt to bring us to this bad place? It is not a place of seed or figs\lebnote{Hebrew “fig”} or vines\lebnote{Hebrew “vine”} or pomegranate trees,\lebnote{Hebrew “tree”} and there is not water to drink.”
\verse And Moses and Aaron went from the presence of the assembly to the doorway of the tent of assembly. They fell on their faces, and the glory of Adonai appeared to them.
\verse Adonai spoke to Moses, saying,
\verse “Take the staff and summon the community, you and Aaron your brother, and speak to the rock before their eyes, and it will give water. Bring out for them water from the rock, and let the community and their livestock drink.”
\verse So Moses took the staff from \textit{before}\lebnote{Literally “in the presence of”} Adonai just as he command him,
\verse and Moses and Aaron summoned the assembly to the presence of the rock, and he said to them, “Please listen, you rebels; can we bring out water for you from this rock?”
\verse Then Moses lifted up his hand and struck the rock with his staff twice. And abundant water went out, and the community and their livestock drank.
\verse But Adonai said to Moses and Aaron, “Because you have not trusted in me, to regard me as holy \textit{in the sight of}\lebnote{Or “before the eyes of”} the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} you will not bring this assembly into the land that I have given to them.”
\verse Those were the waters of Meribah, where the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} quarreled with Adonai, and he showed himself holy among them.
\verse From Kadesh Moses sent messengers to the king of Edom: “Thus your brother Israel has said, ‘You know all the hardship that has found us;
\verse our ancestors\lebnote{Or “fathers”} went down to Egypt, and we lived in Egypt \textit{a long time},\lebnote{Literally “many days”} and the Egyptians mistreated us and our ancestors.\lebnote{Or “fathers”}
\verse Then we cried to Adonai, and he heard our voice; he sent an angel and brought us out from Egypt. And look, we are in Kadesh, a city on the edge of your territory.
\verse Please let us go through your land. We will not go through a field or vineyard, and we will not drink water from a well. We will go along the road of the king; we will not turn aside right or left until we have gone through your territory.’ ”
\verse Then Edom said to him, “You will not pass through us\lebnote{Hebrew “me”} lest \textit{we will go out}\lebnote{Hebrew “I will go out”} to meet you with the sword.”
\verse The \textit{Israelites}\lebnote{Literally “sons/children of Israel”} said to him, “We will go up on the main road, and if we\lebnote{Hebrew “I”} and our livestock\lebnote{Hebrew “my livestock”} drink your water, we will pay for it.\lebnote{Hebrew “I will give their worth”} It is only a small matter; let us pass through on our feet.”\lebnote{Hebrew “I will go through on my feet”}
\verse But he said, “You will not go through.” And Edom went out to meet \textit{them}\lebnote{Hebrew “him”} with a large army and a strong hand.
\verse So Edom refused to give Israel passage through his territory, and Israel turned aside from him.
\verseWithHeading{Aaron Dies} And they set out from Kadesh. The \textit{Israelites},\lebnote{Literally “sons/children of Israel”} the whole community, came to Mount Hor.
\verse Adonai said to Moses and to Aaron on Mount Hor, on the boundary of the land of Edom, saying,
\verse “Let Aaron be gathered to his people; he will not come into the land that I have given to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} because you rebelled against \textit{my word}\lebnote{Literally “my mouth”} at the waters of Meribah.
\verse Take Aaron and Eleazar his son, and take them up Mount Hor.
\verse Strip off Aaron’s garments, and put them on Eleazar his son; Aaron will be gathered to his people, and he will die there.”
\verse So Moses did just as Adonai commanded, and they went up to Mount Hor before the eyes of all the community.
\verse And Moses stripped off Aaron’s garments and put them on Eleazar his son. Aaron died there on the top of the mountain; and Moses and Eleazar went down from the mountain.
\verse All the community saw that Aaron died; so all the house of Israel wept for Aaron thirty days.\lebnote{Hebrew “day”}
\end{biblechapter}

\begin{biblechapter} % Numbers 21
\verseWithHeading{Arad Captured} The Canaanite king of Arad, who was dwelling in the Negev,\lebnote{An arid region south of the Judean hills} heard that Israel came along the way of Atharim; he fought against Israel and took some of them captive.
\verse Israel made a vow to Adonai, and they said, “If you will surely give this people into our\lebnote{Hebrew “my”} hand, then we\lebnote{Hebrew “I”} \textit{will destroy}\lebnote{Literally “devote to God”} their cities.”
\verse Adonai heard the voice of Israel; he gave to them the Canaanites, and \textit{they destroyed them}\lebnote{Literally “they devoted to God”} and their cities. They called the name of the place Hormah.
\verse They set out from Mount Hor by the way of the \textit{Red Sea}\lebnote{Literally “sea of reed”} to go around the land of Edom; but \textit{the people became impatient}\lebnote{Literally “the life of the people became short”} along the way.
\verse The people spoke against God and against Moses, “Why have you brought us from Egypt to die in the desert? There is no food and no water, and our hearts detest this miserable food.”
\verseWithHeading{The Bronze Serpent} And Adonai sent among the people poisonous snakes; they bit the people, and many people from Israel died.
\verse The people came to Moses and said, “We have sinned because we have spoken against Adonai and against you. Pray to Adonai and let him remove the snakes\lebnote{Hebrew “snake”} from among us.” So Moses prayed for the people.
\verse And Adonai said to Moses, “Make for yourself a snake and place it on a pole. When\lebnote{Hebrew “And it will happen”} anyone is bitten and looks at it, that person will live.”
\verse So Moses made a snake of bronze, and he placed it on the pole; whenever\lebnote{Hebrew “And it will happen”} a snake bit someone, and that person looked at the snake of bronze, he lived.
\verse The \textit{Israelites}\lebnote{Literally “sons/children of Israel”} set out and encamped at Oboth.
\verse They set out from Oboth and encamped at Iye Abarim in the desert, which was in front of Moab \textit{toward the sunrise}.\lebnote{Literally “from the east of the sun”}
\verse From there they set out and encamped at the valley of Zered.
\verse From there they set out and encamped beyond Arnon, which is in the desert that goes out from the boundary of the Amorites,\lebnote{Hebrew “Amorite”} because Arnon is the boundary of Moab, between Moab and the Amorites.\lebnote{Hebrew “Amorite”}
\verse Therefore thus it is said in the scroll of the Wars of Adonai,
\verse “Waheb in Suphah, 
and the wadis of Arnon,
\verse From there they went to Beer, which is the water well where Adonai spoke to Moses, “Gather the people, that I may give them water.”
\verse Then Israel sang this song, “Arise, well water! Sing to it!
\verse Well water that the princes dug, that the leaders of the people dug, with a staff and with their rods.” And from the desert they continued to Mattanah,
\verse and from Mattanah to Nahaliel, and from Nahaliel to Bamoth;
\verse and from Bamoth to the valley that is in the territory of Moab, by the top of Pisgah, which overlooks the surface of the wasteland.
\verseWithHeading{Sihon and Og Defeated} Israel sent messengers to Sihon, the king of the Amorites,\lebnote{Hebrew “Amorite”} saying,
\verse “Let us go through your land; we will not turn aside into a field or vineyard; we will not drink well water along the way of the king until we have gone through your territory.”
\verse But Sihon did not allow Israel to go through his territory. Sihon gathered all his people and went out to meet Israel; he came to the desert, to Jahaz, and he fought against Israel.
\verse But Israel struck him with the edge of the sword, and they took possession of his land from Arnon to Jabbok, until the \textit{Ammonites},\lebnote{Literally “sons of Ammon”} because the boundary of the \textit{Ammonites}\lebnote{Literally “sons of Ammon”} was strong.
\verse Israel took all these cities, and Israel inhabited all the cities of the Amorites,\lebnote{Hebrew “Amorite”} in Heshbon, and in all its environs.\lebnote{Hebrew “her daughters;” other modern versions translate “its villages”}
\verse Because Heshbon was the city of Sihon king of the Amorites,\lebnote{Hebrew “Amorite”} who had fought against the former king of Moab and taken all his land from his hand until Arnon.
\verse Thus the ones who quote proverbs say,
\verse “Come to Heshbon! Let it be built! 
And let the city of Sihon be established.
\verse Because fire went out from Heshbon, 
a flame from the city of Sihon; 
it consumed Ar of Moab, 
the lords of the\lebnote{Or “the dominant”} high places of Arnon.
\verse Woe to you, Moab! 
You have perished, people of Chemosh. 
He has given his sons as fugitives, 
and his daughters into captivity, 
to the king of the Amorites,\lebnote{Hebrew “Amorite”} Sihon.
\verse Thus Israel lived in the land of the Amorites.\lebnote{Hebrew “Amorite”}
\verse Moses sent to explore Jaazer; they captured \textit{its environs}\lebnote{Hebrew “her daughters;” other modern versions translate “its villages”} and dispossessed the Amorites\lebnote{Hebrew “Amorite”} who were there.
\verse Then they turned and went up by the way of the Bashan, and Og king of the Bashan and all his people went out to meet them for battle at Edrei.
\verse And Adonai said to Moses, “Do not fear him because I will give him and all his people and all his land into your hand. You will do to him just as you did to Sihon king of the Amorites,\lebnote{Hebrew “Amorite”} who was living in Heshbon.”
\verse And so they destroyed him and his sons, and all his people until they had not spared a survivor; and they took possession of his land.
\end{biblechapter}

\begin{biblechapter} % Numbers 22
\verseWithHeading{Balak and Balaam} The \textit{Israelites}\lebnote{Literally “sons/children of Israel”} set out, and they encamped on the desert-plateau of Moab, across from Jericho beyond the Jordan.
\verse Balak son of Zippor saw all that Israel did to the Amorites,\lebnote{Hebrew “Amorite”}
\verse and Moab was very terrified in the presence of the people because \textit{they}\lebnote{Hebrew “he” or “it”} were numerous; and Moab dreaded the presence of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse And Moab said to the elders of Midian, “Now the crowd will lick up all around us, like a bull devours the grass of the field.” And Balak son of Zippor was king of Moab at that time.
\verse He sent messengers to Balaam son of Beor at Pethor, which is by the river,\lebnote{That is, the Euphrates} in the land of the children of his people, to summon him, saying, “Look! A people went out from Egypt. Look! They cover \textit{the surface of the land};\lebnote{Literally “the eye of the land”} they are about to dwell opposite me.
\verse Now, please go, curse this people for me because they\lebnote{Hebrew “he”} are stronger than me; perhaps I will be able to strike them\lebnote{Hebrew “he”} and drive them\lebnote{Hebrew “he”} out from the land because I know whoever you bless is blessed, and whoever you cursed is cursed.”
\verse So the elders of Moab and the elders of Midian went with a fee for divination in their hand; they came to Balaam and spoke the words of Balak to him.
\verse He said to them, “Spend the night here, and I will return, and I will return word to you, just as Adonai speaks to me.” So the princes of Moab stayed with Balaam.
\verse And God came to Balaam and said, “Who are these men with you?”
\verse And Balaam said to God, “Balak son of Zippor, king of Moab, sent word to me,
\verse ‘Look! A people went out from Egypt. Look! They cover \textit{the surface of the land}.\lebnote{Literally “the eye of the land”} Now, go, curse them\lebnote{Hebrew “him”} for me. Perhaps I will be able to attack them\lebnote{Hebrew “him”} and drive them\lebnote{Hebrew “him”} out.”
\verse God said to Balaam, “You will not go with them; you will not curse the people, because they\lebnote{Hebrew “he”} are blessed.”
\verse Balaam got up in the morning, and he said to the princes of Balak, “Go to your land, because Adonai refused to allow me to go with you.”
\verse The princes of Moab got up and went to Balak, and they said, “Balaam refused to come with us.”
\verse Balak again sent many princes, who were more honored \textit{than the former}.\lebnote{Literally “than these”}
\verse They came to Balaam and said to him, “Thus says Balak son of Zippor, ‘Please, let nothing keep you from coming to me
\verse because I will surely honor you greatly, and all that you say to me I will do. Please, come; curse this people for me.’ ”
\verse Balaam answered and said to the servants of Balak, “Even though Balak gives to me his house full of silver and gold, I am not able to go beyond \textit{the command of Adonai}\lebnote{Literally “the mouth of Adonai”} my God to do a little or a lot.
\verse And now please, \textit{you also stay here}\lebnote{Literally “please stay in this”} the night, and \textit{let me find out}\lebnote{Literally “let me know”} again what Adonai will say with me.”
\verse And God came to Balaam at night, and he said to him, “If the men have come to call you, get up and go with them; but only the word that I will speak to you, you will do.”
\verse So Balaam got up in the morning and saddled his donkey, and he went with the princes of Moab.
\verseWithHeading{Balaam and the Angel} But \textit{God became angry}\lebnote{Literally “God’s nose became hot”} because he was going, and the angel of Adonai stood in the road as an adversary to him; he was riding on his donkey, and two servants were with him.
\verse The donkey saw the angel of Adonai standing in the road with his sword drawn in his hand, and the donkey turned aside from the road and went into the field. And Balaam struck the donkey to turn her back to the road.
\verse The angel of Adonai stood in the narrow path of the vineyards, with \textit{a wall on either side}.\lebnote{Literally “a wall from this and a wall from this”}
\verse When the donkey saw the angel of Adonai, she pressed herself into the wall, and she pressed the foot of Balaam into the wall, so he struck her again.
\verse Then the angel of Adonai went further ahead and stood in a narrow place where there was not a way to turn aside to the right or left.
\verse When the donkey saw the angel of Adonai, she lay down under Balaam, so \textit{Balaam became angry},\lebnote{Literally “Balaam’s nose became hot”} and he struck the donkey with his staff.
\verse Adonai opened the mouth of the donkey, and she said to Balaam, “What did I do to you that you struck me these three times?”
\verse Balaam said to the donkey, “Because you made a mockery of me! If only I had a sword in my hand, I would kill you right now!”
\verse The donkey said to Balaam, “Am I not your donkey on which you have ridden all your life until this day? Have I been in the habit of doing this to you?” He said, “No.”
\verse Then Adonai exposed the eyes of Balaam, and he saw the angel of Adonai standing in the road with his sword drawn in his hand, and he bowed down and worshiped to his face.
\verse The angel of Adonai said to him, “Why have you struck this donkey three times? Look, I have come out as an adversary because your conduct is perverse before me.
\verse The donkey saw me and turned aside from me these three times. If she had not turned aside from my face, then I would have killed you and kept her alive.”
\verse Balaam said to the angel of Adonai, “I have sinned because I did not know that you were standing to meet me in the road. Now, \textit{if it is displeasing to you},\lebnote{Literally “if it is evil in your eyes”} I will turn back.”
\verse The angel of Adonai said to Balaam, “Go with the men, but speak only the word that I will speak to you.” So Balaam went with the princes of Balak.
\verse When Balak heard that Balaam was coming, he went out to meet him by the city of Moab, which was on the boundary of Aaron at the end of the territory.
\verse And Balak said to Balaam, “Did I not urgently send to meet with you? Why did you not come to me? Am I really not able to honor you?”
\verse Balaam said to Balak, “Look, I came to you now. Am I really able to speak anything at all? I speak the word that God puts in my mouth.”
\verse Balaam went with Balak, and they came to Kiriath-Huzoth.
\verse And Balak sacrificed cattle and sheep, and he sent them to Balaam and to the princes who were with him.
\verse And it happened, in the morning Balak took Balaam and took him up to Bamoth-Baal, and he saw from there the end of the nation.
\end{biblechapter}

\begin{biblechapter} % Numbers 23
\verseWithHeading{Balaam’s Oracles} Balaam said to Balak, “Build for me this: seven altars. And prepare for me this: seven bulls and seven rams.”
\verse And Balak did just as Balaam spoke, and Balak offered Balaam a bull and a ram on the altar.
\verse And Balaam said to Balak, “Station yourself at your burnt offering, and I will go; perhaps Adonai will come to meet me, and whatever he shows me I will tell to you.” So he went to a barren height.
\verse And God met with Balaam, and he said to him, “I have arranged seven altars, and I have offered a bull and a ram on the altar.”
\verse Adonai put a word in the mouth of Balaam and said, “Return to Balak, and you must speak thus.”
\verse So he returned to him, and behold, he was standing beside his burnt offering, he and all the leaders of Moab.
\verse And he lifted up his oracle and said,
\verse “From Aram Balak lead me, 
from the mountains of the east the king of Moab, 
‘Go for me, curse Jacob, 
and go, denounce Israel.’
\verse How can I curse whom God has not cursed, 
and how can I denounce whom Adonai has not denounced?
\verse Because from the top of the rocks I see him, 
from hilltops I watch him. 
Behold, a people who dwell alone, 
they do not consider themselves among the nations.
\verse And Balak said to Balaam, “What have you done to me? I took you to curse my enemies, and look, you have surely blessed them!”
\verse He answered and said, “\textit{Should I not speak}\lebnote{Literally “Should I not observe to speak”} what Adonai puts in my mouth?”
\verse Then Balak said, “Please walk with me to another place where you will see them, but you will only see part of them and will not see all of them; and curse them for me from there.”
\verse So he took him to the field of Zophim to the top of Pisgah, and he built seven altars, and he offered a bull and a ram on each altar.
\verse Balaam\lebnote{Hebrew “He”} said to Balak, “Station yourself here at the burnt offering while I myself meet with Adonai there.”
\verse Then Adonai met with Balaam, and he put a word in his mouth, and he said, “Return to Balak, and you must speak thus.”
\verse He came to him, and behold, he was standing at his burnt offering, and the princes of Moab with him. And Balak said to him, “What has Adonai spoken?”
\verse Then \textit{he uttered}\lebnote{Literally “he lifted up”} his oracle, and said,
\verse “Stand up, Balak, and hear; 
listen to me, son of Zippor!
\verse God is not a man, that he should lie, 
nor a son of humankind, 
that he should change his mind. 
Has he said, and will he not do it? 
And has he spoken, and will he not fulfill it?
\verse Behold, I have received a command to bless; 
when he has blessed, I cannot cause it to return.
\verse He has no regard for evil in Jacob, 
and he does not see trouble in Israel; 
Adonai his God is with him, 
and a shout\lebnote{Or “a blast”} of a king is among them.\lebnote{Hebrew “him”}
\verse God, who brings them out from Egypt, 
is like the strength\lebnote{Or “like the horns”} of a wild ox for them.\lebnote{Hebrew “him”}
\verse Because there is no sorcery against Jacob, 
and there is no divination against Israel. 
Now\lebnote{Or “At the right time”} it will be said to Jacob and Israel, 
what God has done!
\verse Then Balak said to Balaam, “Do not curse them\lebnote{Hebrew “him”} at all, nor bless them\lebnote{Hebrew “him”} at all!”
\verse But Balaam answered and said to Balak, “Did I not speak to you, saying, ‘Whatever Adonai speaks I will do’?”
\verse Then Balak said to Balaam, “Please, come, I will take you to another place; perhaps \textit{it will be acceptable to}\lebnote{Literally “it will be right in the eyes of”} God, and you will curse for me from there.”
\verse So Balak took Balaam to the top of Peor, which looks down on the face of the Jeshimon.\lebnote{Or “the wasteland”}
\verse And Balaam said to Balak, “Build for me these seven altars, and prepare for me these seven bulls and seven rams.”
\verse Balak did just as Balaam said, and he offered a bull and a ram on each altar.
\end{biblechapter}

\begin{biblechapter} % Numbers 24
\verseWithHeading{Balaam Continues to Utter Oracles} And Balaam saw that \textit{it pleased}\lebnote{Literally “it was good in the eyes of Adonai”} Adonai to bless Israel, and he did not go \textit{as other times}\lebnote{Literally “as time on time”} \textit{to seek out}\lebnote{Literally “to meet”} sorcery; instead, he set his face toward the desert.
\verse Balaam lift up his eyes, and he saw Israel dwelling according to its tribes, and the spirit of God was upon it.\lebnote{That is, Israel}
\verse He \textit{uttered}\lebnote{Literally “lifted up”} his oracle and said,
\verse “The declaration of Balaam son Beor, 
the declaration of the man whose eyes are closed,
\verse the declaration of the hearer of God’s words,\lebnote{Or “God’s sayings”} 
who sees the revelation of Shaddai,\lebnote{Often translated “the Almighty”} 
falling down but whose eyes are uncovered.
\verse How good are your tents, O Jacob, 
your dwellings, O Israel!
\verse They are spread out like valleys, 
like gardens on a river, 
like aloes planted by Adonai, 
like cedars at the waters.
\verse He will pour water from his buckets, 
and his offspring will be like many waters; 
his king will be higher than Agag, 
and his kingdom will be exalted.
\verse God, who brings him out from Egypt, 
is like the strength\lebnote{Or “the horns”} of a wild ox for him. 
He will devour the nations who are his enemies; 
he will break their bones; 
he will pierce them with his arrows.
\verse Then \textit{Balak became angry with}\lebnote{Literally “the nose of Balak became hot against”} Balaam, and he clapped his hands and said to Balaam, “I called you to curse my enemies, but look, you have surely blessed them these three times.
\verse \textit{Flee}\lebnote{Literally “Flee for yourself”} to your place now. I said I would richly honor you, but look, Adonai has withheld honor from you.”
\verse Balaam said to Balak, “Did I not speak to your messengers whom you sent to me, saying,
\verse ‘If Balak gave to me the fullness of his house full of silver and gold, I am not able to go beyond \textit{the command of Adonai}\lebnote{Literally “the mouth of Adonai”} to do good or evil, from my heart; what Adonai speaks, I will speak’?\lebnote{Hebrew “I will speak it”}
\verse And now, look, I am about to go to my people; I will advise you what this people will do to your people \textit{in the following days}.”\lebnote{Literally “in the last of the days”}
\verse And he \textit{uttered}\lebnote{Literally “lifted up”} his oracle and said,
\verse “The declaration of Balaam son of Beor, 
and the declaration of the man whose eye is closed,
\verse the declaration of the hearer of God’s words,\lebnote{Or “God’s sayings”} 
and the knower of the knowledge of the Most High, 
who sees the vision of Shaddai,\lebnote{Often translated “the Almighty”} 
who is falling, and his eyes are revealed.
\verse I see him, but not now; 
I behold him, but not near; 
a star will go out from Jacob, 
and a scepter will rise from Israel; 
it will crush the foreheads of Moab 
and destroy all the children of Seth.
\verse Edom will be a captive; 
Seir, its enemies, will be a captive, 
and Israel will be acting \textit{courageously}.\lebnote{Literally “with physical strength”}
\verse Someone\lebnote{Hebrew “He”} from Jacob will rule 
and will destroy a remnant\lebnote{Or “survivor”} from the city.”
\verse And he looked at the Kenites,\lebnote{Hebrew “Kenite”} \textit{uttered}\lebnote{Literally “lifted up”} his oracle, and said,
\verse “Steady is your dwelling place; 
in the rock is your nest.
\verse Again he \textit{uttered}\lebnote{Literally “lifted up”} his oracle and said,
\verse “Woe, who will live when God establishes this?\lebnote{Hebrew “it”}
\verse Then Balaam got up and went and returned to his place, and Balak also went on his way.
\end{biblechapter}

\begin{biblechapter} % Numbers 25
\verseWithHeading{The Plague of Israel} When Israel dwelled in Shittim, the people began to prostitute\lebnote{Or “have sexual relations”} themselves with the daughters of Moab.
\verse And they invited the people to the sacrifices of their gods, and the people ate and worshiped their gods.
\verse So Israel was joined together to Baal Peor, and \textit{Adonai became angry}\lebnote{Literally “the nose of Adonai became hot”} with Israel.
\verse Adonai said to Moses, “Take all \textit{the leaders}\lebnote{Literally “the heads”} of the people and kill them before the sun, so the fierce anger of Adonai will turn from Israel.”
\verse So Moses said to the judges of Israel, “Each of you kill his men who are joined together with Baal Peor.”
\verse And behold, a man from the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} came and brought to his brothers a Midianite woman before the eyes of Moses and before the eyes of all of the community of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and they were weeping at the doorway of the tent of assembly.
\verse When Phinehas son of Eleazar son of Aaron the priest saw, he got up from the midst of the community and took a spear in his hand.
\verse He went after the man of Israel into the woman’s section of the tent, and he drove the two of them, the man of Israel and the woman, into her belly. And the plague among the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} stopped.
\verse The ones who died in the plague were twenty-four thousand.
\verse Adonai spoke to Moses, saying,
\verse “Phinehas son of Eleazar, son of Aaron the priest, turned away my anger from among the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} when he was jealous with my jealousy in their midst, and I did not destroy the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} with my jealousy.
\verse Therefore say, ‘Behold, I am giving to him my covenant of peace,
\verse and it will be for him and his \textit{offspring}\lebnote{Literally “seed”} after him a covenant of an eternal priesthood because he was jealous for his God and made atonement for the \textit{Israelites}.’ ”\lebnote{Literally “sons/children of Israel”}
\verse The name of the man of Israel who was struck with the Midianite woman was Zimri son of Salu, a leader of \textit{the family}\lebnote{Or “the father’s house” or “the ancestor’s house”} of the Simeonites.\lebnote{Hebrew “Simeonite”}
\verse The name of the Midianite woman who was struck was Cozbi daughter of Zur, \textit{a leader}\lebnote{Literally “a head”} of a tribe of \textit{the family}\lebnote{Or “the father’s house” or “the ancestor’s house”} in Midian.
\verse Adonai spoke to Moses, saying,
\verse “Attack the Midianites and strike them
\verse because they were attacking you with their deception, with which they have deceived you on the matter of Peor and on the matter of Cozbi the daughter of the leader of Midian, their sister who was struck on the day of the plague because of the matter of Peor.”
\end{biblechapter}

\begin{biblechapter} % Numbers 26
\verseWithHeading{A New Census} \lebnote{Numbers 26:1a in the English Bible is 25:19b in the Hebrew Bible} And it happened after the plague, \lebnote{In the Hebrew Bible, Numbers 26:1 begins here}Adonai said to Moses and to Eleazar son of Aaron the priest, saying,
\verse “\textit{Take a census}\lebnote{Literally “Lift up the heads”} of the community of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, according to \textit{their families},\lebnote{Literally “the house of their fathers”} all who are able to go out to war in Israel.”
\verse So Moses and Eleazar the priest spoke with them on the desert-plateau of Moab by the Jordan across from Jericho, saying,
\verse “Take a census of the community from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, just as Adonai commanded Moses.” The Israelites who went out from the land of Egypt were:
\verse Reuben, the firstborn of Israel, the descendants\lebnote{Or “sons”} of Reuben: of Hanoch, the clan of the Hanochites; of Pallu, the clan of the Palluites;\lebnote{Hebrew “Palluite”}
\verse of Hezron, the clan of the Hezronites;\lebnote{Hebrew “Hezronite”} of Carmi, the clan of the Carmites.\lebnote{Hebrew “Carmite”}
\verse These are the clans of the Reubenites,\lebnote{Hebrew “Reubenite”} and the ones counted of them were forty-three thousand seven hundred and thirty.
\verse The children of Pallu: Eliab.
\verse The children of Eliab: Nemuel, Dathan, and Abiram. These are the same Dathan and Abiram who were appointed of the community, who rebelled against Moses and Aaron in the company of Korah, when they rebelled against Adonai,
\verse and the land opened its mouth and swallowed them with Korah, when that company died, when the fire consumed two hundred and fifty men, and they were a sign.\lebnote{That is, a warning sign}
\verse The children\lebnote{Or “sons”} of Korah, however, did not die.
\verse The descendants\lebnote{Or “sons”} of Simeon, according to their clans: of Nemuel, the clan of the Nemuelites;\lebnote{Hebrew “Nemuelite”} of Jamin, the clan of the Jaminites;\lebnote{Hebrew “Jaminite”} of Jakin, the clan of the Jakinites;\lebnote{Hebrew “Jakinite”}
\verse of Zerah, the clan of the Zerahites;\lebnote{Hebrew “Zerahite”} of Shaul, the clan of the Shaulites.\lebnote{Hebrew “Shaulite”}
\verse These were the clans of the Simeonites,\lebnote{Hebrew “Simeonite”} twenty-two thousand two hundred.
\verse The descendants\lebnote{Or “sons”} of Gad according to their clans: of Zephon, the clan of the Zephonites;\lebnote{Hebrew “Zephonite”} of Haggi, the clan of the Haggites;\lebnote{Hebrew “Haggite”} of Shuni, the clan of the Shunites;\lebnote{Hebrew “Shunite”}
\verse of Ozni, the clan of the Oznites;\lebnote{Hebrew “Oznite”} of Eri, the clan of the Erites;\lebnote{Hebrew “Erite”}
\verse of Arod, the clan of the Arodites;\lebnote{Hebrew “Arodite”} of Areli, the clan of the Arelites;\lebnote{Hebrew “Arelite”}
\verse These were the clans of the descendants\lebnote{Or “sons”} of Gad according to the ones counted of them, forty thousand five hundred.
\verse The sons of Judah: Er and Onan; but Er and Onan died in the land of Canaan.
\verse The descendants\lebnote{Or “sons”} of Judah according to their clans were: of Shelah, the clan of the Shelanites;\lebnote{Hebrew “Shelanite”} of Perez, the clan of the Perezites;\lebnote{Hebrew “Perezite”} of Zerah, the clan of the Zerahites.\lebnote{Hebrew “Zerahite”}
\verse The children of Perez were: of Hezron, the clan of the Hezronites;\lebnote{Hebrew “Hezronite”} of Hamul, the clan of the Hamulites.\lebnote{Hebrew “Hamulite”}
\verse These were the clans of Judah according to the ones counted of them, seventy-six thousand five hundred.
\verse The descendants\lebnote{Or “sons”} of Issachar according to their clans: of Tola, the clan of the Tolaites;\lebnote{Hebrew “Tolaite”} of Puvah, the clan of the Punites;\lebnote{Hebrew “Punite”}
\verse of Jashub, the clans of the Jashubites;\lebnote{Hebrew “Jashubite”} of Shimron, the clan of the Shimronites.\lebnote{Hebrew “Shimronite”}
\verse These were the clans of Issachar according to the ones counted of them, sixty-four thousand three hundred.
\verse The descendants\lebnote{Or “sons”} of Zebulun according to their clans: of Sered, the clan of the Seredites;\lebnote{Hebrew “Seredite”} of Elon, the clan of the Elonites;\lebnote{Hebrew “Elonite”} of Jahleel, the clan of the Jahleelites.\lebnote{Hebrew “Jahleelite”}
\verse These were the clans of the Zebulunites\lebnote{Hebrew “Zebulunite”} according to the ones counted of them, sixty thousand five hundred.
\verse The descendants\lebnote{Or “sons”} of Joseph according to their clans: Manasseh and Ephraim.
\verse The descendants\lebnote{Or “sons”} of Manasseh: of Makir, the clan of the Makirites.\lebnote{Hebrew “Makirite”} And Makir fathered Gilead; of Gilead, the clan of the Gileadites.\lebnote{Hebrew “Gileadite”}
\verse These were the descendants\lebnote{Or “sons”} of Gilead: of Iezer, the clan of the Iezerites;\lebnote{Hebrew “Iezerite”} of Helek, the clan of the Helekites;\lebnote{Hebrew “Helekite”}
\verse and of Asriel, the clan of the Asrielites;\lebnote{Hebrew “Asrielite”} and of Shechem, the clan of the Shechemites;\lebnote{Hebrew “Shechemite”}
\verse and of Shemida, the clan of the Shemidaites;\lebnote{Hebrew “Shemidaite”} and of Hepher, the clan of the Hepherites.\lebnote{Hebrew “Hepherite”}
\verse Zelophehad son of Hepher did not have sons, but only daughters; and the names\lebnote{Hebrew “name”} of the daughters of Zelophehad were Mahlah, Noah, Hoglah, Milcah, and Tirzah.
\verse These were the clans of Manasseh, and the ones counted of them were fifty-two thousand seven hundred.
\verse These were the descendants\lebnote{Or “sons”} of Ephraim according to their clans: of Shuthelah, the clan of the Shuthelahites;\lebnote{Hebrew “Shuthelahite”} of Beker, the clan of the Bekerites;\lebnote{Hebrew “Bekerite”} of Tahan, the clan of the Tahanites.\lebnote{Hebrew “Tahanite”}
\verse And these were the descendants\lebnote{Or “sons”} of Shuthelah: of Eran, the family of the Eranites.\lebnote{Hebrew “Eranite”}
\verse These were the clans of the descendants\lebnote{Or “sons”} of Ephraim according to the ones counted of them, thirty-two thousand five hundred. These were the descendants\lebnote{Or “sons”} of Joseph according to their clans.
\verse The descendants\lebnote{Or “sons”} of Benjamin according to their clans: of Bela, the clan of the Belaites;\lebnote{Hebrew “Belaite”} of Ashbel, the clan of the Ashbelites;\lebnote{Hebrew “Ashbelite”} of Ahiram, the clan of the Ahiramites;\lebnote{Hebrew “Ahiramite”}
\verse of Shephupham, the clan of the Shuphamites;\lebnote{Hebrew “Shuphamite”} of Hupham, the clan of the Huphamites.\lebnote{Hebrew “Huphamite”}
\verse The sons of Bela were Ard and Naaman: of Ard, the clan of the Ardites;\lebnote{Hebrew “Ardite”} of Naaman, the clan of the Naamites.\lebnote{Hebrew “Naamite”}
\verse These were the descendants\lebnote{Or “sons”} of Benjamin according to their clans. And the ones counted of them were forty-five thousand six hundred.
\verse These were the descendants\lebnote{Or “sons”} of Dan according to their clans: of Shuham, the clan of the Shuhamites.\lebnote{Hebrew “Shuhamite”} These were the clans of Dan according to their clans.
\verse All the clans of the Shuhamites,\lebnote{Hebrew “Shumhamite”} according to the ones counted of them, were sixty-four thousand four hundred.
\verse The descendants\lebnote{Or “sons”} of Asher according to their clans: of Imnah, the clan of the Imnahites;\lebnote{Hebrew “Imnahite”} of Ishvi, the clan of the Ishvites;\lebnote{Hebrew “Ishvite”} of Beriah, the clan of the Beriahites.\lebnote{Hebrew “Beriahite”}
\verse The descendants\lebnote{Or “sons”} of Beriah: of Heber, the clan of the Heberites;\lebnote{Hebrew “Heberite”} of Malkiel, the clan of the Malkielites.\lebnote{Hebrew “Malkielite”}
\verse The name of the daughter of Asher was Serah.
\verse These were the clans of the descendants\lebnote{Or “sons”} of Asher according to the ones counted of them, fifty-three thousand four hundred.
\verse The descendants\lebnote{Or “sons”} of Naphtali according to their clans: of Jahzeel, the clan of the Jahzeelites;\lebnote{Hebrew “Jahzeelite”} of Guni, the clan of the Gunites;\lebnote{Hebrew “Gunite”}
\verse of Jezer, the clan of the Jezerites;\lebnote{Hebrew “Jezerite”} of Shillem, the clan of the Shillemites.\lebnote{Hebrew “Shillemite”}
\verse These were the clans of Naphtali according to their clans, the ones counted of them, forty-five thousand four hundred.
\verse These were the ones counted of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} six hundred and one thousand seven hundred and thirty.
\verse Then Adonai spoke to Moses, saying,
\verse “For these the land must be divided as an inheritance according to the number of names.
\verse For the larger group you must increase their inheritance, and for the smaller group you must make smaller their inheritance; each must be given \textit{their}\lebnote{Hebrew “his”} inheritance according to the number of the ones counted of them.
\verse Surely the land will be divided by lot. They will inherit according to the names of the tribes of their ancestors.\lebnote{Or “fathers”}
\verse \textit{Their}\lebnote{Hebrew “His”} inheritance must be divided according to the lot between the larger and smaller groups.”
\verse These are the ones counted of the Levites according to their clans: of Gershon, the clan of the Gershonites;\lebnote{Hebrew “Gershonite”} of Kohath, the clan of the Kohathites;\lebnote{Hebrew “Kohathite”} of Merari, the clan of the Merarites.\lebnote{Hebrew “Merarite”}
\verse These are the clans of Levi: the clan of the Libnites,\lebnote{Hebrew “Libnite”} the clan of the Hebronites,\lebnote{Hebrew “Hebronite”} the clan of the Mahlites,\lebnote{Hebrew “Mahlite”} the clan of the Mushites,\lebnote{Hebrew “Mushite”} the clan of the Korahites.\lebnote{Hebrew “Korahite”} Kohath fathered Amram.
\verse The name of the wife of Amram was Jochebed, the daughter of Levi, whose mother bore her for Levi in Egypt; she bore to Amram: Aaron and Moses and their sister Miriam.
\verse To Aaron were born Nadab and Abihu, Eleazar and Ithamar.
\verse But Nadab and Abihu died when they presented strange fire \textit{before}\lebnote{Literally “in the presence of”} Adonai.
\verse The ones counted were twenty-three thousand, every male from \textit{a month old}\lebnote{Literally “a son of one month”} and above, because they were not counted in the midst of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} since no inheritance was given to them in the midst of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse These were the ones counted by Moses and Eleazar the priest, who counted the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} on the desert-plateau of Moab on the Jordan across Jericho.
\verse And among these there was not a man of those counted by Moses and Aaron the priest, who counted the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} in the desert of Sinai.
\verse For Adonai said to them, “They will surely die in the desert.” And not a man was left over from them, except Caleb son of Jephunneh and Joshua son of Nun.
\end{biblechapter}

\begin{biblechapter} % Numbers 27
\verseWithHeading{Laws of Inheritance} Then the daughters of Zelophehad, the son of Hepher, the son of Gilead, the son of Makir, the son of Manasseh, of the clan of Manasseh the son of Joseph, came near; and these were the names of his daughters: Mahlah, Noah, Hoglah, and Tirzah.
\verse They stood before Moses and before Eleazar the priest and before the leaders of the entire community at the doorway of the tent of assembly,\lebnote{Or “meeting”} saying,
\verse “Our father died in the desert; he was not in the midst of the company of those who banded together against Adonai in the company of Korah, but he died in his own sin, and he had no sons.
\verse Why should the name of our father disappear from the midst of his clan because he does not have a son? Give us property in the midst of the brothers of our father.”
\verse So Moses brought their case before Adonai.
\verse And Adonai said to Moses, saying,
\verse “\textit{The statements of the daughters of Zelophehad are right}.\lebnote{Literally “Such as the daughters of Zelophehad are speaking”} You must surely give them the property of an inheritance in the midst of their father’s brothers, and you must transfer the inheritance of their father to them.
\verse And you must speak to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} saying, ‘If a man dies and has no son, you must transfer his inheritance to his daughter.
\verse And if he has no daughter, you must give his inheritance to his brothers.
\verse If he has no brothers, then you must give his inheritance to his father’s brothers.
\verse If his father has no brothers, then you must give his inheritance to his nearest relative from his own clan, and he will take possession of it. It will be as a decree of stipulation for the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} just as Adonai commanded Moses.’ ”
\verse Adonai said to Moses, “Go up to this mountain of Abarim, and see the land that I have given to the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse When you see it, you will be gathered to your people, just as Aaron your brother was gathered,
\verse because you rebelled against my word in the desert of Zin when the community quarreled regarding my holiness at the waters.” (These are the waters of Meribah-Kadesh in the desert of Zin.)
\verseWithHeading{Joshua Succeeds Moses} Adonai spoke to Moses, saying,
\verse “Let Adonai, the God of the spirits of all flesh, appoint a man over the community
\verse who will go out before them and will come in before them, and who will lead them out and bring them in, so the community of Adonai will not be like a flock that does not have a shepherd.”
\verse Then Adonai said to Moses, “Take Joshua son of Nun, a man in whom is the spirit, and place your hand on him.
\verse Have him stand before Eleazar the priest and before the entire community, and \textit{commission him}\lebnote{Literally “command him”} \textit{in their sight}.\lebnote{Literally “before their eyes”}
\verse You will give to him from your authority so that the entire community of Israel will \textit{obey him}.\lebnote{Literally “hear him”}
\verse He will stand before Eleazar the priest, who will ask for him by the decision\lebnote{Or “judgment”} of the Urim before Adonai. On \textit{his command}\lebnote{Literally “his word”} they will go out, and at \textit{his command}\lebnote{Literally “his word”} they will come in, both he and all of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} with him, the entire community.”
\verse Moses did just as Adonai commanded him, and he took Joshua and set him before Eleazar the priest and before the entire community.
\verse And he placed his hands on him and \textit{commissioned him}\lebnote{Literally “commanded him”} just as Adonai spoke by the hand of Moses.\lebnote{Or “through Moses”}
\end{biblechapter}

\begin{biblechapter} % Numbers 28
\verseWithHeading{Daily Sacrifice} Adonai spoke to Moses, saying,
\verse “Command the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} and say to them, ‘\textit{You will be careful to present}\lebnote{Literally “you will observe to present”} my offering, my food of my offerings made by fire, of a fragrance of appeasement to me, at its appointed time.’
\verse You will say to them, ‘This is the offering made by fire that you will offer to Adonai: two male lambs without defect \textit{in their first year}\lebnote{Literally “sons of a year”} as a continual burnt offering each day.
\verse You will offer one male lamb in the morning, and the second male lamb you will offer \textit{at twilight},\lebnote{Literally “between the two evenings”}
\verse and a tenth of an ephah of finely milled flour as a grain offering, mixed with a fourth of a measure of beaten oil.
\verse It is a continual burnt offering that was \textit{ordained}\lebnote{Literally “done”} on Mount Sinai as a fragrance of appeasement, an offering made by fire for Adonai.
\verse The libation with it will be a fourth of a liquid measure for each male lamb; in the sanctuary you will pour out the libation of fermented drink for Adonai.
\verse And the second male lamb you will offer \textit{at twilight};\lebnote{Literally “between the two evenings”} as the grain offering of the morning and as its libation you will offer it, an offering made by fire, a fragrance of appeasement for Adonai.
\verse “ ‘On the day of the Sabbath, two male lambs without defect \textit{in their first year},\lebnote{Literally “sons of a year”} and two-tenths of finely milled flour mixed with oil for a grain offering and its libation.
\verse This is the burnt offering every Sabbath in addition to the continual burnt offering and its libation.
\verse “ ‘And at the beginning of each of your months, you will present a burnt offering for Adonai: two bulls and one ram, seven male lambs without defect \textit{in their first year};\lebnote{Literally “sons of a year”}
\verse and three-tenths of finely milled flour mixed with oil for a grain offering, for each bull; and two-tenths of finely milled flour mixed with oil for a grain offering for the one ram;
\verse and a tenth of finely milled flour mixed with oil as a grain offering for each male lamb, for a burnt offering of a fragrance of appeasement, an offering of fire for Adonai.
\verse Their libations will be half a liquid measure of wine for the bull and a third of a liquid measure of wine for the ram and a fourth of a liquid measure of wine for the male lamb; this is the burnt offering for every month for the months of the year.
\verse And one male goat as a sin offering for Adonai; it will be offered in addition to the continual burnt offering and its libation.
\verse “ ‘On the fourteenth day of the first month is the Passover for Adonai.
\verse On the fifteenth day of this month is a religious feast, unleavened bread must be eaten for seven days.
\verse On the first day there will be a holy convocation; \textit{you will not do any regular work}.\lebnote{Literally “you will not do work of labor”}
\verse You will present an offering by fire, a burnt offering for Adonai: two bulls and one ram and seven male lambs \textit{in their first year};\lebnote{Literally “sons of a year”} they will be for you without defect.
\verse For their grain offering, you will offer finely milled flour mixed with oil: three-tenths for the bull and two-tenths for the ram.
\verse You will offer a tenth for each of the seven male lambs;
\verse and a goat for one sin offering to make atonement for you.
\verse You will offer these besides the burnt offering of the morning, which is for the continual burnt offering.
\verse Like this you will offer daily, for seven days, the food of the offering made by fire, a fragrance of appeasement for Adonai; it will be offered in addition to the continual burnt offering and its libation.
\verse On the seventh day you will have a holy convocation; \textit{you will not do any regular work}.\lebnote{Literally “you will not do work of labor”}
\verseWithHeading{Offerings for the Festival of Weeks} “ ‘And on the day of firstfruits, when you are presenting a new offering for Adonai during your Festival of Weeks, you will have a holy convocation; \textit{you will not do any regular work}.\lebnote{Literally “you will not do work of labor”}
\verse You will present a burnt offering for a fragrance of appeasement for Adonai: two bulls, one ram, seven male lambs \textit{in their first year};\lebnote{Literally “sons of a year”}
\verse and their grain offering will be finely milled flour mixed with oil: three-tenths for each bull, two-tenths for one ram,
\verse a tenth for each of the male lambs;
\verse and one male goat to make atonement for you.
\verse In addition to the continual burnt offering and its grain offering, you will offer them without defect with their libation.
\end{biblechapter}

\begin{biblechapter} % Numbers 29
\verseWithHeading{Offers for the Seventh Month} “ ‘On the seventh month, on the first day of the month, you will have a holy convocation; \textit{you will not do any regular work}.\lebnote{Literally “you will not do work of labor”} It will be a day for you of blowing trumpets.
\verse You will offer a burnt offering as a fragrance of appeasement for Adonai: one bull, one ram, and seven male lambs \textit{in their first year};\lebnote{Literally “sons of a year”} they will be without defect.
\verse Their grain offering will be finely milled flour mixed with oil: three-tenths for the bull, two-tenths for the ram;
\verse and one-tenth for each of the seven male lambs;
\verse with one male goat for a sin offering, to make atonement for you,
\verse in addition to the burnt offering of the new moon and its grain offering, the continual burnt offering and its grain offering, and their libations, according to their stipulations, as a fragrance of appeasement by fire for Adonai.
\verseWithHeading{Offerings for the Day of Atonement} “ ‘And on the tenth of this seventh month you will have a holy convocation, and \textit{you will afflict yourselves};\lebnote{Literally “you will afflict your lives” or “you will afflict your souls”} you will not do any work.
\verse You will present a burnt offering for Adonai, a fragrance of appeasement: one bull, one ram, seven male lambs \textit{in their first year};\lebnote{Literally “sons of a year”} they will be without defect.
\verse And their grain offering will be of finely milled flour mixed with oil: three-tenths for the bull, two-tenths for the one ram,
\verse one-tenth for each of the seven male lambs;
\verse one male goat for a sin offering, in addition to the sin offering of atonement and the continual burnt offering and its grain offering, and their libations.
\verse “ ‘Then on the fifteenth day of the seventh month you will have a holy convocation; \textit{you will not do any regular work},\lebnote{Literally “you will not do work of labor”} and you will hold a religious feast for Adonai for seven days.
\verse You will present a burnt offering, an offering made by fire as a fragrance of appeasement for Adonai: thirteen bulls, two rams, fourteen male lambs \textit{in their first year};\lebnote{Literally “sons of a year”} they will be without defect.
\verse And their grain offering will be of finely milled flour mixed with oil: three-tenths for the bull, two-tenths for the one ram,
\verse one-tenth for each of the seven male lambs;
\verse and one male goat for a sin offering, in addition to the continual burnt offering, its grain offering, and its libation.
\verse “ ‘On the second day: twelve bulls, two rams, fourteen male lambs \textit{in their first year};\lebnote{Literally “sons of a year”} they will be without defect;
\verse and their grain offering and their libations for the bulls, for the rams, and for the male lambs, by their number according to the stipulation;
\verse and one male goat for a sin offering, in addition to the continual burnt offering and its grain offering, and their libations.
\verse “ ‘On the third day: eleven bulls, two rams, fourteen male lambs without defect \textit{in their first year};\lebnote{Literally “sons of a year”}
\verse and their grain offering and their libations for the bulls, for the rams, and for the male lambs, by their number according to the stipulation;
\verse and one male goat for a sin offering, in addition to the continual burnt offering, its grain offering, and its libation.
\verse “ ‘On the fourth day: ten bulls, two rams, fourteen male lambs without defect \textit{in their first year};\lebnote{Literally “sons of a year”}
\verse and their grain offering and their libations for the bulls, for the rams, and for the male lambs by their number according to the stipulation;
\verse and one male goat for a sin offering, in addition to the continual burnt offering, its grain offering, and its libation.
\verse “ ‘On the fifth day: nine bulls, two rams, fourteen male lambs without defect \textit{in their first year};\lebnote{Literally “sons of a year”}
\verse and their grain offering and their libations for the bulls, for the rams, and for the male lambs by their number according to the stipulation;
\verse and one male goat for a sin offering, in addition to the continual burnt offering, its grain offering, and its libation.
\verse “ ‘On the sixth day: eight bulls, two rams, fourteen male lambs without defect \textit{in their first year};\lebnote{Literally “sons of a year”} and their grain offering and their libations for the bulls, for the rams, and for the male lambs by their number according to the stipulation;
\verse and their grain offering and their libations for the bulls, for the rams, and for the male lambs by their number according to the stipulation;
\verse and one male goat for a sin offering, in addition to the continual burnt offering, its grain offering, and its libation.
\verse “ ‘On the seventh day: seven bulls, two rams, fourteen male lambs without defect \textit{in their first year};\lebnote{Literally “sons of a year”}
\verse and their grain offering and their libations for the bulls, for the rams, and for the male lambs by their number according to the stipulation;
\verse and one male goat for a sin offering, in addition to the continual burnt offering, its grain offering, and its libation.
\verse “ ‘On the eighth day you will have an assembly; \textit{you will not do any regular work}.\lebnote{Literally “you will not do work of labor”}
\verse You will present a burnt offering, an offering made by fire as a fragrance of appeasement for Adonai: one bull, one ram, seven male lambs without defect \textit{in their first year};\lebnote{Literally “sons of a year”}
\verse and their grain offering and their libations for the bulls, for the rams, and for the male lambs by their number according to the stipulation;
\verse and one male goat for a sin offering, in addition to the continual burnt offering, its grain offering, and its libation.
\verse “ ‘You will present these to Adonai at your appointed time, in addition to your vows and your freewill offerings, for your burnt offerings and for you grain offerings and for your libations and for your fellowship offerings.’ ”
\verse \lebnote{Numbers 29:40–30:16 in the English Bible is 30:1–17 in the English Bible} So Moses said to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} in accordance with all that Adonai commanded Moses.
\end{biblechapter}

\begin{biblechapter} % Numbers 30
\verseWithHeading{Laws for Vows} Then Moses spoke to \textit{the leaders}\lebnote{Literally “the heads”} of the tribes concerning the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} saying, “This is the word that Adonai commanded:
\verse if a man makes a vow for Adonai or swears an oath with a binding pledge on himself, he must not render his word invalid; he must do all that went out from his mouth.
\verse “If a woman makes a vow to Adonai, and she binds a pledge on herself in her father’s house in your childhood,
\verse but if her father hears her vow or her pledge that she bound on herself and says nothing to her, then all her vows will stand, and every pledge that she binds on her life will stand.
\verse If her father forbids her on the day he hears of it, all her vows or her pledges that she bound on herself will not stand, and Adonai will forgive her because her father has forgiven her.
\verse “If \textit{she has a husband}\lebnote{Literally “she is to a man”} while bound by her vows or a rash promise of her lips,
\verse and her husband hears of it and is silent on the day he hears it, her vows will stand, and her pledge that she bound upon herself will stand.
\verse But if on the day her husband hears of it, he forbids her, then he will nullify her vow that she is under, and the rash promise of her lips that she bound on herself; and Adonai will forgive her.
\verse “But the vow of a widow or a woman who is divorced, all that she binds on herself will stand on her.
\verse But if she made a vow in her husband’s house, or bound herself on a pledge with a sworn oath,
\verse and her husband heard it but was silent to her, and he did not forbid her, all her vows will stand and every pledge that she bound on herself will stand.
\verse But if her husband nullified them on the day he hears them, all her vows going out of her lips concerning her vows or the pledge on herself will not stand; her husband has nullified them, and Adonai will forgive her.
\verse “Any vow and any sworn oath of a pledge to inflict on herself, her husband can confirm it or her husband can nullify it.
\verse But if her husband is completely silent from day to day, then he confirms all her vows or all her pledges that are on her; he confirms them because he was silent to her on the day he heard them.
\verse But if he indeed nullifies them after he hears them, then he will bear her guilt.”
\verse These are the decrees that Adonai commanded Moses, as between a husband and his wife, and between a father and his daughter, while her childhood is in her father’s house.
\end{biblechapter}

\begin{biblechapter} % Numbers 31
\verseWithHeading{War Against the Midianites} Adonai spoke to Moses, saying,
\verse “Seek vengeance for the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} on the Midianites; afterward you will be gathered to your people.”
\verse Moses spoke to the people, saying, “Arm yourself from among your men for the battle, so that they will \textit{go}\lebnote{Literally “be”} against Midian to mete out the vengeance of Adonai on Midian.
\verse A thousand from each tribe of every tribe of Israel you will send to battle.”
\verse So theywere assigned from the thousands of Israel, a thousand from each tribe, twelve thousand equipped for battle.
\verse Moses sent them, a thousand from each tribe, to the battle, and Phinehas son of Eleazar the priest to the battle with them, and the vessels of the sanctuary and the trumpets of the blast were in his hand.
\verse And they fought against Midian just as Adonai commanded Moses, and they killed every male.
\verse They killed the kings of Midian in addition to the ones they had slain: Evi and Rekem and Zur and Hur and Reba, the five kings of Midian; they also killed Balaam son of Beor by the sword.
\verse The \textit{Israelites}\lebnote{Literally “sons/children of Israel”} took captive the women of Midian and their children, and they plundered all their domestic animals and all their livestock and all their wealth.
\verse They burned all their cities where they dwelled and all their camps with fire.
\verse They took all the plunder and all the war-booty with the humans\lebnote{Hebrew “human”} and domestic animals.\lebnote{Hebrew “animal”}
\verse They brought the captives, the war-booty, and the plunder to Moses, and to Eleazar the priest, and to the community of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} to the camp to the desert-plateau of Moab, which was on the Jordan across Jericho.
\verse And Moses and Eleazar the priest and all the leaders of the community went out to meet them outside the camp.
\verse But Moses was angry toward the leaders of the troops, the commanders of the thousands and the commanders of the hundreds, who came from the battle of the war.
\verse And Moses said to them, “You have kept alive every female?
\verse Behold, these women \textit{caused}\lebnote{Literally “were to”} the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} by the word of Balaam, to be in apostasy against Adonai in the matter of Peor, so that the plague was among the community of Adonai.
\verse Now kill every male among the little children, and kill every woman who has \textit{had sexual intercourse with a man}.\lebnote{Literally “who has known the bed of a male”}
\verse But all the females who have not \textit{had sexual intercourse with a man},\lebnote{Literally “who has known the bed of a male”} keep alive for yourselves.
\verse And you, camp outside the camp seven days; all who killed a person and all who touched the slain purify yourselves on the third day and on the seventh day, you and your captives.
\verse You will purify yourselves and every garment and every object of hide and all the work of goats’ hair, and every object of wood.”
\verse Then Eleazar the priest said to the men of the battle who came from the war, “This is the decree of the law that Adonai commanded Moses.
\verse Only the gold and the silver, the bronze, the iron, the tin, and the lead—
\verse everything that will go through the fire—you will pass through the fire, and it will be clean, and only in waters of impurity will it be purified. Whatever does not go into the fire you will pass through the waters.
\verse And you will wash your garments on the seventh day and be clean, and afterward you will come into the camp.”
\verseWithHeading{Division of the War-Booty} Adonai said to Moses, saying,
\verse “You and Eleazar the priest and the leaders of the \textit{families}\lebnote{Literally “fathers”} of the community, \textit{take count}\lebnote{Literally “lift up the head”} of the war-booty that was captured, both humans\lebnote{Hebrew “human”} and the domestic animals;\lebnote{Hebrew “animal”}
\verse divide the war-booty between those who engaged in the war, who went out to the battle, and all the community.
\verse Exact a tribute for Adonai from the men of the war, those who went out to the battle, one from five hundred persons, and from the cattle and from the male donkeys and from the flock;
\verse take from their half and give it to Eleazar the priest as a contribution to Adonai.
\verse From half of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} take one share drawn by lot from the fifty from the humans,\lebnote{Hebrew “human”} from the cattle, from the male donkeys, from the flock, from all the domestic animals,\lebnote{Hebrew “animal”} and give them to the Levities who keep the responsibilities of the tabernacle of Adonai.”
\verse Moses and Eleazar the priest did just as Adonai commanded Moses.
\verse Thus the war-booty that remained of the plunder that the people of the battle plundered was six hundred and seventy-five thousand flocks of sheep,
\verse seventy-two thousand cattle,
\verse sixty-one thousand male donkeys,
\verse and the life of humankind, from the women who did not \textit{have sexual intercourse with a man},\lebnote{Literally “who has known the bed of a male”} all the persons\lebnote{Hebrew “person”} were thirty-two thousand.
\verse The half of the share that was going out to the battle: the number of the flock of sheep was thee hundred and thirty-seven thousand five hundred;
\verse the tribute to Adonai from the flock was six hundred and seventy-five;
\verse and the cattle were thirty-six thousand; and the tribute to Adonai was seventy-two.
\verse Of the male donkeys there were thirty thousand five hundred, and the tribute to Adonai was sixty-one;
\verse the humans\lebnote{Hebrew “human”} were sixteen thousand, and the tribute to Adonai was thirty-two persons.
\verse And Moses gave away the tribute of the contribution of Adonai to Eleazar the priest, just as Adonai commanded Moses.
\verse From the half of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} which Moses divided from the men who were fighting,
\verse the half that belonged to the community was three hundred and thirty-seven thousand five hundred from the flock,
\verse and thirty-six thousand cattle,
\verse and thirty thousand five hundred male donkeys,
\verse and sixteen thousand humans.\lebnote{Hebrew “human”}
\verse From the half that belonged to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} Moses took one share drawn by lot out of every fifty humans\lebnote{Hebrew “human”} and domestic animals, and he gave them to the Levites, who keep the responsibility of the tabernacle of Adonai, just as Adonai commanded Moses.
\verse Then the leaders of the thousands of the army, the commanders of the thousands and the commanders of the hundreds, approached Moses,
\verse and they said to Moses, “Your servants have \textit{taken count}\lebnote{Literally “lifted up the head”} of the men of war who were \textit{in our charge},\lebnote{Literally “in our hand”} and no man is missing from us.
\verse So we brought the offering of Adonai, what each man found, objects of gold, bangles,\lebnote{Hebrew “bangle”} bracelets,\lebnote{Hebrew “bracelet”} rings,\lebnote{Hebrew “ring”} earrings,\lebnote{Hebrew “earring”} and female ornaments,\lebnote{Hebrew “ornament”} to make atonement for ourselves \textit{before}\lebnote{Literally “in the presence of”} Adonai.”
\verse Moses and Eleazar the priest took the gold from them, all objects of work.
\verse All the gold of the contribution that they raised up to Adonai, from the commanders of the thousands and the commanders of the hundreds, was sixteen thousand seven hundred and fifty shekels.
\verse The men of battle plundered each for himself.
\verse So Moses and Eleazar the priest took the gold from the commanders of the thousands and hundreds, and they brought it to the tent of the assembly as a memorial for the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} before Adonai.
\end{biblechapter}

\begin{biblechapter} % Numbers 32
\verseWithHeading{Gad and Reuben Inherit Gilead} The descendants\lebnote{Or “sons”} of Reuben and the descendants\lebnote{Or “sons”} of Gad had a very large number of livestock. And they saw the land of Jazer and the land of Gilead, and behold it was a place for livestock.
\verse The descendants\lebnote{Or “sons”} of Gad and the descendants\lebnote{Or “sons”} of Reuben came, and they said to Moses and to Eleazar the priest and to the leaders of the community, saying,
\verse “Ataroth, Dibon, Jazer, Nimrah, Heshbon, Elealeh, Sebam, Nebo, and Beon,
\verse the land that Adonai struck before the community of Israel, is a land of livestock, and your servants have livestock.”
\verse They said, “If we have found favor in \textit{your sight},\lebnote{Literally “your eyes”} let this land be given to your servants as property; do not lead us across the Jordan.”
\verse But Moses said to the descendants\lebnote{Or “sons”} of Gad and to the descendants\lebnote{Or “sons”} of Reuben, “Will your brothers go to war while you yourselves live here?
\verse Why are you discouraging the hearts of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} from crossing into the land that Adonai gave to them?
\verse This is what your fathers did when I sent them from Kadesh Barnea to see the land.
\verse When they went up to the valley\lebnote{Or “the wadi”} of Eshcol and saw the land, they discouraged the heart of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} so that they did not come to the land that Adonai gave to them.
\verse So \textit{Adonai’s anger burned}\lebnote{Literally “Adonai’s nose became hot”} on that day, and he swore an oath, saying,
\verse ‘The men who went up from Egypt, from \textit{those twenty years old}\lebnote{Literally “a son of twenty years”} and above, will not see the land that I swore with an oath to Abraham, Isaac, and Jacob because they have not wholly followed me,
\verse except Caleb son of Jephunneh the Kenizzite and Joshua son of Nun, because they followed Adonai wholly.’
\verse And \textit{Adonai became angry},\lebnote{Literally “Adonai’s nose became hot”} and he made them wander in the desert forty years until the entire generation who did evil in the \textit{sight of Adonai}\lebnote{Literally “eyes of Adonai”} \textit{had died}.\lebnote{Literally “had completed”}
\verse Behold, you stand in the place of your fathers, a brood of sinful men, to increase still more \textit{Adonai’s fierce anger}\lebnote{Literally “the fierce anger of Adonai’s nose”} against Israel.
\verse If you turn from following him, he will again abandon them\lebnote{Hebrew “him”} in the wilderness, and you would have destroyed all these people.”
\verse They came near to him and said, “We will build sheep pens here for the flock of our livestock and cities for our little children;
\verse but we ourselves will become armed and ready before the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} until we have brought them to their place, and our little children will live in the fortified cities because of the inhabitants of the land.
\verse We will not return to our houses until the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} each obtain their\lebnote{Hebrew “his”} inheritance for themselves.
\verse For we will not take possession with them from across the Jordan and beyond because our inheritance has come to us from across the Jordan to the east.”
\verse So Moses said to them, “If you do this thing, if you arm yourselves \textit{before}\lebnote{Literally “in the presence of”} Adonai for the war,
\verse and everyone of you armed cross the Jordan \textit{before}\lebnote{Literally “in the presence of”} Adonai until he has driven out his enemies from before him,
\verse and the land is subdued \textit{before}\lebnote{Literally “in the presence of”} Adonai, then afterward you will return and be free of obligation from Adonai and from Israel, and this land will be your property \textit{before}\lebnote{Literally “in the presence of”} Adonai.
\verse But if you do not do so, behold, you have sinned against Adonai, and know that your sin will find you.
\verse Build for yourselves cities for your little children and sheep pens for your flocks; what has gone out from your mouth you will do.”
\verse So the descendants\lebnote{Or “sons”} of Gad and the descendants\lebnote{Or “sons”} of Reuben said to Moses, saying, “Your servants will do just as my lord commands.
\verse Our little children, our wives, our livestock, and all of our animals\lebnote{Hebrew “animal”} \textit{will remain}\lebnote{Literally “will be there”} in the cities of Gilead,
\verse but your servants, everyone who is armed for battle, will cross over \textit{before}\lebnote{Literally “in the presence of”} Adonai to the war, just as my lord says.”
\verse So Moses commanded them, Eleazar the priest, Joshua son of Nun, and the heads of the \textit{families}\lebnote{Literally “fathers”} of the tribes of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse Moses said to them, “If the descendants\lebnote{Or “sons”} of Gad and the descendants\lebnote{Or “sons”} of Reuben, everyone who is armed for the war, cross over the Jordan \textit{before}\lebnote{Literally “in the presence of”} Adonai, and the land is subdued before you, you will give them the land of Gilead as property.
\verse But if they will not cross over with you armed, they will acquire land in your midst in Canaan.”
\verse The descendants\lebnote{Or “sons”} of Gad and the descendants of Reuben answered and said, “What Adonai has commanded your servants, we will do.
\verse We ourselves will cross over armed \textit{before}\lebnote{Literally “in the presence of”} Adonai to the land of Canaan, and the property of our inheritance will remain with us beyond the Jordan.”
\verse So Moses gave to them, to the descendants\lebnote{Or “sons”} of Gad and the descendants\lebnote{Or “sons”} of Reuben, and to half of the tribe of Joseph’s son Manasseh, the kingdom of Sihon the king of Amorites and the kingdom of Og the king of the Bashan, the land with its cities and their territories, the cities of the surrounding land.
\verse The descendants\lebnote{Or “sons”} of Gad rebuilt Dibon, Ataroth, and Aroer,
\verse and Atroth Shophan, Jazer, and Jogbehah,
\verse and Beth Nimrah and Beth Haran, the cities of Mibzar, and the sheep pens for flocks.
\verse The descendants\lebnote{Or “sons”} of Reuben rebuilt Heshbon, Elealeh, and Kiriathaim,
\verse and Nebo, Baal Meon (their names\lebnote{Hebrew “name”} were changed), and Sibmah, and \textit{they renamed}\lebnote{Literally “they called to names the names”} the cities that they rebuilt.
\verse The descendants\lebnote{Or “sons”} of Makir son of Manasseh went to Gilead, and they captured it and drove out the Amorites\lebnote{Hebrew “Amorite”} who were in it.
\verse So Moses gave Gilead to Makir son of Manasseh, and he lived in it.
\verse And Jair son of Manasseh went and captured their unwalled villages, and he called them Havvoth Jair.
\verse Nobah went and captured Kenath and its villages, and he called it Nobah after his own name.
\end{biblechapter}

\begin{biblechapter} % Numbers 33
\verseWithHeading{The Travels Are Recounted} These were the journeys of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} who went out from the land of Egypt according to their divisions, by the hand of Moses and Aaron.
\verse Moses wrote down their movements according to their journeys on the command of Adonai, and these are their journeys according to their movements.
\verse They set out from Rameses on the first month, on the fifteenth day of the first month; on the next day after the Passover the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} went out \textit{boldly}\lebnote{Literally “with a hand that was raised”} \textit{in the sight}\lebnote{Literally “for the eyes”} of all the Egyptians
\verse while the Egyptians were burying all the firstborn among them whom Adonai struck. Adonai also executed punishments among their gods.
\verse Then the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} set out from Rameses, and they camped in Succoth.
\verse They journeyed from Succoth and camped in Etham, which is on the edge of the desert.
\verse Then they set out from Etham and returned to Pi-Hahiroth, which faces Baal Zephon, and they camped before Migdol.
\verse They set out from Pi-Hahiroth and went through the midst of the sea into the desert; and they went a journey of three days into the desert of Etham and camped at Marah.
\verse They set out from Marah and came to Elim, and in Elim there were twelve springs of water and seventy palm trees, and they camped there.
\verse They set out from Elim, and they camped at the \textit{Red Sea}.\lebnote{Literally “sea of reed”}
\verse They set out from the \textit{Red Sea}\lebnote{Literally “sea of reed”} and camped at the desert of Sin.
\verse They set out from the desert of Sin and camped at Dophkah.
\verse They set out from Dophkah and camped at Alush.
\verse They set out from Alush and encamped at Rephidim; and it was there that the people had no water to drink.
\verse They set out from Rephidim and camped in the desert of Sinai.
\verse The set out from the desert of Sinai and camped at Kibroth Hattaavah.
\verse They set out from Kibroth Hattaavah and camped at Hazeroth.
\verse They set out from Hazeroth and camped at Rithmah.
\verse They set out from Rithmah and camped at Rimmon Perez.
\verse They set out from Rimmon Perez and camped at Libnah.
\verse They set out from Libnah and camped at Rissah.
\verse They set out from Rissah and camped at Kehelathah.
\verse They set out from Kehelathah and camped at Mount Shapher.
\verse They set out from Mount Shapher and camped at Haradah.
\verse They set out from Haradah and camped at Makheloth.
\verse They set out from Makheloth and camped at Tahath.
\verse They set out from Tahath and camped at Terah.
\verse They set out from Terah and camped at Mithcah.
\verse They set out from Mithcah and camped at Hashmonah.
\verse They set out from Hashmonah and camped at Moserah.
\verse They set out from Moserah and camped at Bene-Jaakan.
\verse They set out from Bene-Jaakan and camped at Hor Haggidgad.
\verse They set out from Hor Haggidgad and camped at Jotbathah.
\verse They set out from Jotbathah and camped at Abronah.
\verse They set out from Abronah and camped at Ezion Geber.
\verse They set out from Ezion Geber and camped in the desert of Zin, that is, Kadesh.
\verse They set out from Kadesh and camped at Mount Hor, at the edge of the land of Edom.
\verse Aaron the priest went up to Mount Hor at the \textit{command}\lebnote{Literally “mouth”} of Adonai, and he died there in the fortieth year after the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} had gone out from the land of Egypt, in the fifth month on the first day of the month.
\verse Aaron was one hundred and twenty-three years old when he died on Mount Hor.
\verse Now the Canaanite, the king of Arad, who was living in the Negev\lebnote{An arid region south of the Judean hills} in the land of Canaan, heard of the coming of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse Then they set out from Mount Hor and camped at Zalmonah.
\verse They set out from Zalmonah and camped at Punon.
\verse They set out from Punon and camped at Oboth.
\verse They set out from Oboth and camped at Iye Abarim, the boundary of Moab.
\verse They set out from Iyim and camped at Dibon Gad.
\verse They set out from Dibon Gad and camped at Almon-Diblatayim.
\verse They set out from Almon-Diblatayim and camped in the mountains of Abarim, before Nebo.
\verse They set out from the mountains of Abarim and camped on the desert-plateau of Moab by the Jordan across Jericho.
\verse They camped by the Jordan, from Beth-Jeshimoth up to Abel Shittim, on the desert-plateau of Moab.
\verse Then Adonai spoke to Moses on the desert-plateau of Moab by the Jordan across Jericho, saying,
\verse “Speak to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} and say to them, ‘When you cross the Jordan into the land of Canaan,
\verse you will drive out the inhabitants of the land from your presence, and you will destroy all their idols and all the images of their molten idols, and you will demolish all their high places;
\verse you will dispossess the land and live in it because I have given the land to you to possess it.
\verse You will distribute the land by lot according to your clans; to the larger group you will give a larger inheritance, and to the smaller group you will give less inheritance. However the lot falls for him, there the lot will be. You will distribute it according to the tribes of your ancestors.\lebnote{Or “fathers”}
\verse But if you do not drive out the inhabitants of the land from your presence, then it will happen that whomever you let remain of them will be like irritants in your eyes and like thorns in your sides; they will be your enemies in the land in which you live.
\verse And just as I planned to do to them, I will do to you.’ ”
\end{biblechapter}

\begin{biblechapter} % Numbers 34
\verseWithHeading{The Land of Canaan Is Divided} Then Adonai spoke to Moses, saying,
\verse “Command the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} and say to them, ‘When you come into the land of Canaan, this is the land that was allotted to you as an inheritance, the land of Canaan according to its boundaries.
\verse Your southern edge will be from the desert of Zin toward the side of Edom, and your southern border will be from the end of the Salt Sea\lebnote{That is, the Dead Sea} to the east;
\verse your boundary will turn from the south to the ascent of Akrabbim\lebnote{Or “Scorpions”} and will pass over to Zin, and its limits will be from the south of Kadesh Barnea; it will continue to Hazar Addar and pass over to Azmon.
\verse The boundary will turn from Azmon to the valley of Egypt, and its limits will be to the sea.
\verse “ ‘Your western boundary will be the Great Sea;\lebnote{That is, the Mediterranean} this will be your western boundary.
\verse Your northern border will be from the Great Sea;\lebnote{That is, the Mediterranean} you will make a boundary from the Great Sea\lebnote{That is, the Mediterranean} to Mount Hor.
\verse From Mount Hor you will make a boundary to reach Hamath; the limits of the territory will be at Zedad.
\verse The boundary will go out to Ziphron, and its limits will be at Hazar Enan. This will be your boundary to the north.
\verse “ ‘You will mark out your eastern boundary from Hazar Enan to Shepham;
\verse the boundary will go down from Shepham to Riblah from the east side of Ain, and the boundary will go down and border on the eastern side of the Sea of Kinnereth.
\verse The boundary will go down to the Jordan, and its limits will be at the Salt Sea.\lebnote{That is, the Dead Sea} This will be your land according to its boundaries all around.’ ”
\verse So Moses commanded the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} saying, “This is the land that you will obtain as an inheritance for yourself by lot, which Adonai commanded to give to the nine and a half tribes.
\verse For the tribe of the children of the Reubenites,\lebnote{Hebrew “Reubenite”} the children of the Gadites,\lebnote{Hebrew “Gadite”} and the half-tribe of Manasseh took their inheritance according to the house of their families.\lebnote{Or “their fathers”}
\verse The two and a half tribes have taken their inheritance from beyond the Jordan across Jericho, east toward the sunrise.”
\verse Adonai spoke to Moses, saying,
\verse “These are the names of the men who divide up the land for your inheritance: Eleazar the priest and Joshua son of Nun.
\verse You will take one leader from every tribe to divide up the land for inheritance.
\verse These are the names of the men: of the tribe of Judah, Caleb son of Jephunneh;
\verse of the tribe of the descendants of Simeon, Samuel son of Ammihud;
\verse of the tribe of Benjamin, Elidad son of Chislon;
\verse of the tribe of the descendants\lebnote{Or “sons”} of Dan, the leader Bukki son of Jogli;
\verse of the descendants\lebnote{Or “sons”} of Joseph, the tribe of the descendants\lebnote{Or “sons”} of Manasseh, the leader Hanniel son of Ephod.
\verse Of the tribe of the descendants\lebnote{Or “sons”} of Ephraim, the leader Kemuel son of Shiphtan;
\verse of the tribe of the descendants\lebnote{Or “sons”} of Zebulum, the leader Elizaphan son of Parnach;
\verse of the tribe of the descendants\lebnote{Or “sons”} of Issachar, the leader Paltiel son of Azzan;
\verse of the tribe of the descendants\lebnote{Or “sons”} of Asher, the leader Ahihud son of Shelomi;
\verse of the tribe of the descendants\lebnote{Or “sons”} of Naphtali, the leader Pedahel son of Ammihud.”
\verse These are those whom Adonai commanded to allot to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} the land of Canaan.
\end{biblechapter}

\begin{biblechapter} % Numbers 35
\verseWithHeading{Cities for the Levites} Adonai spoke to Moses on the desert plains of Moab beyond the Jordan across Jericho, saying,
\verse “Command the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} that they give to the Levites from the inheritance of their property cities to live in; and you will give to the Levites pastureland all around the cities.\lebnote{Hebrew “them”}
\verse The cities will be theirs to live in, and their pasturelands will be for their domestic animals, for their possessions, and their animals.\lebnote{Hebrew “animal”}
\verse “The pasturelands of the cities that you will give to the Levites will extend from the wall of the city to a distance of a thousand cubits all around.
\verse You will measure outside the city the eastern edge two thousand cubits, for the southern edge two thousand cubits, for the western edge two thousand cubits, and for the northern edge two thousand cubits, with the city in the middle; this will be for them the pasturelands of the cities.
\verse “All the cities that you will give the Levites will be six cities of refuge, to which the killer can flee; in addition to them you will give forty-two cities.
\verse All the cities that you will give to the Levites will be forty-eight cities, them with their pasturelands.
\verse And the cities that you will give from the property of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} you will take more from the larger group and less from the smaller group; each of them will give according to the portion of their inheritance according to the portion that he inherits.”
\verseWithHeading{Cities of Refuge} Adonai spoke to Moses, saying,
\verse “Speak to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} and say to them, ‘When you cross the Jordan into the land of Canaan,
\verse you will select for yourselves cities for your cities of refuge, that a killer who has killed a person unintentionally can flee there.
\verse The cities will be to you a refuge from a redeemer, so that the killer will not die until he stands before the community for judgment.
\verse The cities that you are to give will be your six cities of refuge.
\verse You will give three cities across the Jordan and three cities in the land of Canaan; they will be cities of refuge.
\verse To the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} to the alien, and to the temporary resident in their midst there will be these six cities as a refuge to which anyone who unintentionally kills a person may flee.
\verse “ ‘But if he hit him with an object of iron, so that he dies, the killer must surely be put to death.
\verse And if he hit him with a stone in the hand, by which he will die, and he does die, he is a killer; the killer must surely be put to death.
\verse Or if he hit him with a wooden object, by which he will die, and he does die, he is a killer; the killer must surely be put to death.
\verse The blood avenger himself will put the killer to death; he must put him to death when meeting him.
\verse If he shoves him in hatred, or he throws something at him with intention, and he dies,
\verse or if he hits him in hostility with his hand, and he dies, the one that struck him will put to death the killer when meeting him.
\verse “ ‘Or if in an instant he shoved him, not in hostility, or threw something at him without intention,
\verse or with any stone, without seeing it dropped on him so that he dies, while he was not seeking his injury,
\verse then the community will judge between the striker and between the blood avenger according to these ordinances.
\verse The community will deliver the killer from the hand of the blood avenger, and the community will restore him to the city of his refuge to which he fled; and he will live there in it until the death of the high priest who was anointed with holy oil.
\verse But if the killer surely goes out of the territory of the city of his refuge to which he fled,
\verse and the blood avenger finds him outside the territory of the city of his refuge, and the blood avenger kills the killer, \textit{he will not be guilty of blood}\lebnote{Literally “there will not be blood for him”}
\verse because he must live in the city of his refuge until the death of the high priest. But after the death of the high priest the killer will return to the land of his property.
\verse These things will be as a decree of justice for you for your generations in all your dwellings.
\verse “ ‘If anyone kills a person, the killer will be put to death \textit{according to the testimony}\lebnote{Literally “according to the mouth”} of witnesses, but someone cannot die on testimony of one person.
\verse Also, you will not take a ransom payment for the life of a killer who is guilty of death; indeed, he must surely be put to death.
\verse You will not take a ransom payment for the one that flees to the city of his refuge, so that he may return to live in the land before the death of the priest.
\verse So you will not pollute the land in which you are; because blood pollutes the land, and no atonement can be made for the land for the blood that is poured out on it except with the blood of the one who poured it out.
\verse You will not defile the land on which you are living because I am living in the midst of it; I am Adonai; I am living in the midst of the \textit{Israelites}.’ ”\lebnote{Literally “sons/children of Israel”}
\end{biblechapter}

\begin{biblechapter} % Numbers 36
\verseWithHeading{Inheritance Through Marriage} \textit{The leaders}\lebnote{Literally “The heads”} of the families\lebnote{Or “fathers”} of the clans of descendants\lebnote{Or “sons”} of Gilead the son of Makir, the son of Manasseh, of the clans of the descendants\lebnote{Or “sons”} of Joseph came near and spoke \textit{before}\lebnote{Literally “before the face of”} Moses and \textit{before}\lebnote{Literally “before the face of”} \textit{the leaders}\lebnote{Literally “the heads”} of the families\lebnote{Hebrew “of the fathers”} of the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse And they said, “Adonai commanded my lord to give the land by lot as an inheritance to the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} and my lord was commanded by Adonai to give the inheritance of Zelophehad our brother to his daughters.
\verse But if they become wives to one of the sons from another tribe of the \textit{Israelites},\lebnote{Literally “sons/children of Israel”} their inheritance will disappear from the inheritance of our ancestors,\lebnote{Or “fathers”} and it will be added to the inheritance of the tribe to which they belong; the lot of our inheritance would disappear.
\verse When the Jubilee of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} \textit{comes},\lebnote{Literally “will be”} it will be added to the inheritance of the tribe to which they belong; and their inheritance will disappear from the tribe of our father.”
\verse Then Moses commanded the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} by the command of Adonai, saying, “The tribe of the descendants\lebnote{Or “sons”} of Joseph is right regarding what they are speaking.
\verse This is the word that Adonai commanded the daughters of Zelophehad, saying, ‘\textit{Let them marry}\lebnote{Literally “Let them be as wives”} \textit{whomever they like};\lebnote{Literally “they will be as the good in their eyes”} only \textit{they must marry}\lebnote{Literally “they must be as wives”} from within the clan of the tribe of their father.
\verse Thus an inheritance of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} will not go around from tribe to tribe. Rather, the inheritance of each tribe of his father will remain with the \textit{Israelites}.\lebnote{Literally “sons/children of Israel”}
\verse Every daughter who possesses an inheritance from the tribes of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} will \textit{marry}\lebnote{Literally “be a wife to”} one of the clan of the tribe of her father, so that the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} will possess the inheritance of his ancestors.\lebnote{Or “fathers”}
\verse Therefore an inheritance will not go around from one tribe to another tribe because the tribes of the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} will each hold to their own inheritance.’ ”
\verse Just as Adonai commanded to Moses, so the daughters of Zelophehad did:
\verse Mahlah, Tirzah, Hoglah, Milcah, and Noah, the daughters of Zelophehad, \textit{married}\lebnote{Literally “were as wives to”} the sons of their uncles.
\verse \textit{They married}\lebnote{Literally “They were as wives to”} those from the sons of Manasseh son of Joseph, and their inheritance \textit{remained}\lebnote{Literally “was”} among the tribe of the clan of their ancestors.\lebnote{Or “fathers”}
\verse These were the commands and the stipulations that Adonai commanded by the hand of Moses\lebnote{Or “through Moses”} to the \textit{Israelites}\lebnote{Literally “sons/children of Israel”} on the desert-plateaus of Moab by the Jordan across Jericho.
\end{biblechapter}

