\biblebook{Obadiah}

\begin{biblechapter} % Obadiah 1
\verseWithHeading{God’s Judgment on Edom} The vision of Obadiah. Thus says my Lord Adonai concerning Edom:
\verseWithHeading{Edom’s Approaching Destruction} We have heard a report from Adonai, and a messenger has been sent among the nations: “Rise up and let us rise against it for battle.”
\verse “Look, I will make you insignificant among the nations. You \textit{will be} utterly despised!
\verse The pride of your heart has deceived you, you who live in the clefts of a rock, the heights of its dwelling, \textit{you who} say in your heart: ‘Who can bring me down \textit{to the} ground?’
\verse \textit{Even} if you soar like the eagle, \textit{even} if your nest is set among \textit{the} stars, from there I will bring you down!” \textit{declares} Adonai:
\verse “If thieves came to you, if plunderers of \textit{the} night—How you have been destroyed!—would they not steal \textit{what they wanted}? If grape gatherers came, would they not leave gleanings?
\verse How Esau has been pillaged; his treasures have been ransacked!
\verse All of your \textit{allies} have driven you up to the boundary; your \textit{confederates} have deceived you and have prevailed against you. \textit{Those who eat} your bread have set an ambush for you, there is no understanding of it.
\verse On that day,” \textit{declares} Adonai, “will I not destroy the wise men from Edom, and understanding from the mountain of Esau?
\verseWithHeading{Edom’s Treachery against Judah} “Because of the violence done to your brother Jacob, shame will cover you and you will be cut off forever.
\verse On \textit{the day you stood nearby}, \textit{on the day strangers took} his wealth, and foreigners entered his gates and cast lots over Jerusalem, you were also like one of them.
\verse But you should not have gloated over your brother’s day, on the day of his misfortune, and you should not have rejoiced over the \textit{people} of Judah on the day of their perishing, and you should not have opened your mouth wide on the day of distress.
\verse You should not have entered the gate of my people on the day of their disaster. You also should not have gloated over his misery on the day of his disaster, and you should not have stretched out \textit{your hands} on the day of his disaster.
\verse And you should not have stood at the crossroad\textit{s} to cut off his fugitives and you should not have handed over his survivors on the day of distress.
\verseWithHeading{Future Judgment of the Nations} “For the day of Adonai \textit{is} near against all the nations! Just as you have done, it will be done to you. Your deeds will return on your \textit{own} head.
\verse For just as you have drunk on \textit{my holy mountain}, all the nations will drink continually. They will drink and they will slurp, and they will be as \textit{if} they had never been.
\verse But on Mount Zion there will be an escape, and it will be holy, and the house of Jacob will take possession \textit{of} their dispossessors.
\verse And the house of Jacob will be a fire and the house of Joseph a flame, and the house of Esau stubble; and they will set them on fire and will consume them. And there will not be a survivor for the house of Esau,” for Adonai has spoken.
\verseWithHeading{Future Blessing for Israel} Those of the Negev will take possession of the mountain of Esau, and those of the Shephelah \textit{will possess the land of} the Philistines, and they shall take possession of the territory of Ephraim and the territory of Samaria, and Benjamin \textit{will take possession of} Gilead.
\verse And the exiles of this army of the people of Israel \textit{will possess} Canaan up to Zarephath, and the exiles of Jerusalem who \textit{are} in Sepharad will take possession of the cities of the Negev.
\verse And those who have been saved will go up on Mount Zion to rule the mountain of Esau. And the kingdom \textit{will belong to} Adonai.
\end{biblechapter}

