\biblebook{Proverbs}

\begin{biblechapter} % Proverbs 1
\verseWithHeading{Prologue} Proverbs of Solomon, son of David, king of Israel:
\verse To know wisdom and instruction, 
to understand sayings of understanding,
\verse to gain \textit{insightful instruction}, 
righteousness and justice and equity,
\verse to give shrewdness to the simple, 
knowledge and purpose to the young,
\verse may the wise hear and increase learning, 
and the one who understands gain direction,
\verse to understand a proverb and an expression, 
words of wisdom and their riddles.
\verse Fear of Adonai \textit{is the} beginning of knowledge; 
wisdom and instruction, fools despise.
\verseWithHeading{Lecture Against Gang Behavior} My child, may you keep your father’s instruction, 
and do not reject your mother’s teachings,
\verse for they \textit{are} a garland of favor for your head, 
and pendants for your neck.
\verse My child, if sinners entice you, do not consent.
\verse If they say, 
“Come with us! 
We shall lie in wait for blood; 
we shall ambush the innocent without cause.
\verse Like Sheol, we will swallow them alive 
and whole, \textit{like those who descend to the pit}.
\verse We shall find all precious wealth, 
we shall fill our houses \textit{with} booty,
\verse you shall throw your lot in our midst, 
there will be one purse for all of us.”
\verse My child, do not walk in their way. 
Keep your foot from their paths,
\verse for their feet run to evil, 
and they hurry to shed blood,
\verse for “in vain is the net scattered, 
in the sight of any \textit{winged bird}.”
\verse They lie in wait for their \textit{own} blood. 
They ambush their \textit{own} lives.
\verse Thus are the ways of all who are greedy \textit{for} gain— 
it will take the life of its possessors.
\verseWithHeading{The Call of Lady Wisdom} Wisdom calls out in the streets, 
in the squares she raises her voice.
\verse On a \textit{busy corner} she cries out, 
at the entrances of the gates in the city, she speaks her sayings:
\verse “\textit{How long}, O simple ones, will you love simplicity? 
And \textit{how long} will scoffers delight in their scoffing, 
and fools hate knowledge?
\verse May you turn to my argument! 
Behold, I shall pour out my spirit upon you; 
I will make my words known to you.
\verse Because I called out and you refused me, 
I stretched out my hand, yet there is none who heeds.
\verse You have ignored all my counsel, 
and my reproof you are not willing \textit{to accept}.
\verse I will also laugh at your calamity; 
I will mock when panic comes \textit{upon} you.
\verse When your panic comes like a storm, 
and your calamity arrives like a whirlwind, 
when distress and anguish come upon you,
\verse then they will call me, but I will not answer; 
they will seek me diligently but not find me.
\verse Since they hated knowledge, 
and did not choose the fear of Adonai,
\verse they were not willing \textit{to accept} my counsel, 
they despised all my reproof.
\verse They shall eat from the fruit of their way, 
and they shall be sated from their own schemes,
\verse for the waywardness of the simple ones will kill them, 
and the complacency of fools will destroy them.
\verse Whoever listens to me will dwell in security 
and rest securely from dread and disaster.”
\end{biblechapter}

\begin{biblechapter} % Proverbs 2
\verseWithHeading{The Benefits of Wisdom} My child, if you will receive my sayings, 
and hide my commands with you,
\verse \textit{in order} to incline your ear toward wisdom, 
\textit{then} you shall apply your heart to understanding.
\verse For if you cry out for understanding, 
\textit{if} you lift your voice for insight,
\verse if you seek her like silver 
and search her out \textit{like treasure},
\verse then you will understand the fear of Adonai, 
and the knowledge of God you will find.
\verse For Adonai will give wisdom; 
from his mouth \textit{come} knowledge and understanding.
\verse For the upright, he stores sound judgment, a shield for those who walk uprightly,
\verse \textit{in order} to guard paths of justice 
and keep the way of his faithful \textit{ones}.
\verse Then you will understand righteousness and justice 
and uprightness—every good course—
\verse for wisdom will enter your heart, 
and knowledge will be pleasing to your soul.
\verse Discretion will watch over you; 
understanding will protect you,
\verse \textit{in order} to deliver you from the way of evil, 
from a man who speaks devious things—
\verse those who forsake the paths of uprightness 
to walk in ways of darkness,
\verse those who are happy to do evil, 
\textit{for} they delight in \textit{the} deviousness of evil,
\verse \textit{who are crooked in their ways}, 
and devious in their paths;
\verse \textit{in order} to deliver you from a strange woman, 
from a foreign woman \textit{who} flatters \textit{with} her sayings,
\verse she who forsakes the partner of her youth 
and has forgotten the covenant of her God,
\verse for her house sinks to death, 
and to the dead are her paths.
\verse \textit{Of} all who go to her, none shall return, 
nor do they reach paths of life.
\verse So that you will walk on the road of those who are good, 
and the paths of those who are righteous you shall keep.
\verse For those who are upright will dwell \textit{in the} land, 
and those who are blameless will remain in it.
\verse And those who are wicked will be cut off from the land, 
and those who are treacherous will be uprooted from it.
\end{biblechapter}

\begin{biblechapter} % Proverbs 3
\verseWithHeading{Wisdom and Piety} My child, do not forget my instruction, 
and may your heart guard my commands.
\verse For length of days, years of life, 
and peace they shall add to you.
\verse May loyal love and truth not forsake you; 
bind them around your neck, 
write them upon your heart.
\verse And you shall find favor and good sense 
in the eyes of God and humankind.
\verse \textit{Trust Adonai} with all your heart; 
do not lean toward your own understanding.
\verse In all your ways acknowledge him, 
and he will straighten your paths.
\verse Do not be wise in your own eyes; 
fear Adonai and retreat from evil.
\verse There shall be healing for your \textit{flesh}, 
and refreshment for your body.
\verse Honor Adonai from your substance, 
and from the firstfruits of all that will come to you,
\verse and your barns shall be full of plenty, 
and your vats shall burst \textit{with} new wine.
\verseWithHeading{Happy Is the One Who Finds Wisdom} Do not despise the discipline of Adonai, my child. 
Do not be weary of his reproof
\verse because whomever Adonai will love, he will rebuke, 
as a father delights in his son.
\verse Happy \textit{is} \textit{the one} who finds wisdom, 
and one who obtains understanding.
\verse For her income is better than the income of silver, 
and her revenue than \textit{that of} gold.
\verse She is more precious than rubies, 
and all your desires shall not compare with her.
\verse Length of days \textit{is} in her right hand; 
in her left hand \textit{are} riches and honor.
\verse Her ways \textit{are} ways of pleasantness, 
and all her paths \textit{are} peace.
\verse She is a tree of life for those who seize her; 
those who take hold of her are considered happy.
\verseWithHeading{Role of Wisdom in Creation and Society} Adonai in wisdom founded the earth; 
he established the heavens in understanding.
\verse With his knowledge, depths broke open, 
and clouds dropped dew.
\verse My child, may they not escape from your \textit{sight}; 
may you keep sound wisdom and prudence.
\verse They shall be life for your soul 
and adornment for your neck.
\verse Then you will walk in the confidence of your ways, 
and your foot will not stumble.
\verse If you sit down, you will not panic, 
and \textit{if} you lie down, then your sleep shall be sweet.
\verse Do not be afraid of sudden panic, 
or the storm of wickedness that will come.
\verse Adonai will be \textit{your confidence} 
and guard your foot from capture.
\verse Do not withhold good from its owner 
when it is in the power of your hand to do.
\verse Do not say to your neighbor, 
“Go and return and tomorrow I will give \textit{it},” 
\textit{when it is with you}.
\verse Do not plot harm against your neighbor 
\textit{who} dwells in confidence beside you.
\verse Do not quarrel with anyone without cause, 
when he did not do you harm.
\verse Do not envy a man of violence, 
and do not choose any of his ways,
\verse for he who is perverse \textit{is} an abomination of Adonai, 
but those who are upright \textit{are} his confidence.
\verse The curse of Adonai \textit{is} on the house of the wicked, 
and the abode of the righteous ones he blesses.
\verse With those who scorn, he is scornful, 
but to those who are humble, he gives favor.
\verse They will inherit the honor of the wise, 
but stubborn fools, disgrace.
\end{biblechapter}

\begin{biblechapter} % Proverbs 4
\verseWithHeading{The Father’s Wisdom} Children, listen \textit{to the} instruction of a father, 
and be attentive \textit{in order} to know insight.
\verse For I have given you good instruction; 
do not forsake my teaching.
\verse When I was a son to my father, 
tender and alone before my mother,
\verse he taught me and said to me, 
“May your heart hold fast \textit{to} my words; 
guard my commandments and live.
\verse Get wisdom and insight; 
do not forget and do not turn from the sayings of my mouth.
\verse Do not forsake her—then she will guard you; 
love her and she will keep you.
\verse The beginning of wisdom: Get wisdom! 
With all \textit{that is in} your possession, gain insight.
\verse Cherish her and she will exalt you; 
she will honor you if you embrace her.
\verse She will give a garland for your head; 
she shall bestow a crown of glory upon you.”
\verseWithHeading{The Right Path} Listen, my child, take my sayings, 
and \textit{the} years of your life shall be great.
\verse In the way of wisdom I have instructed you; 
I have led you in the path of uprightness.
\verse \textit{When you walk}, your step will not be hampered, 
and if you run, you will not stumble.
\verse Seize the instruction! Do not let go! 
Guard her, for she is your life.
\verse In the path of the wicked do not go; 
do not walk in the way of those who do evil.
\verse Avoid it; do not transgress it; 
turn away from it and pass by.
\verse For they will not sleep if they have not done wrong, 
and they are robbed of their sleep if they do not cause stumbling.
\verse For they ate the bread of wickedness, 
and they drank the wine of violence.
\verse But the path of the righteous ones \textit{is} like the light of dawn, 
leading and shining until the day \textit{is full}.
\verse The way of the wicked ones \textit{is} like deep darkness; 
they do not know what they stumble over.
\verseWithHeading{Staying the Course} My child, be attentive to my words; 
to my sayings incline your ear.
\verse May they not escape from your \textit{sight}; 
keep them in \textit{the} midst of your heart.
\verse For they are life to those who find them 
and healing \textit{to the entire body}.
\verse With all vigilance, keep your heart, 
for from it \textit{comes} the source of life.
\verse Remove from yourself \textit{deceitful speech}, 
and abolish \textit{devious talk} from yourself.
\verse May your eyes look forward 
and your gaze be straight before you.
\verse May the path of your foot be balanced 
and all your ways be sure.
\verse Do not swerve right or left; 
remove your foot from evil.
\end{biblechapter}

\begin{biblechapter} % Proverbs 5
\verse My child, be attentive to my wisdom, 
and to my understanding incline your ear;
\verse \textit{in order} to keep prudence, 
and knowledge will guard your lips.
\verse For the lips of the strange woman will drip honey, 
and smoother than oil is her mouth.
\verse But her end is bitter as the wormwood, 
sharp as a two-edged sword.
\verse Her feet go down \textit{to} death; 
her steps take hold \textit{of} Sheol.
\verse She does not observe the path of life; 
her ways wander, \textit{and} she does not know \textit{it}.
\verseWithHeading{Do Not Commit Adultery Against Wisdom} Now, O children, listen to me; 
do not depart from the sayings of my mouth.
\verse Keep your paths far from her, 
and do not go near to the door of her house,
\verse lest you give your honor to the others, 
and your years to the merciless,
\verse lest strangers take \textit{their} fill of your strength, 
and your labors \textit{go} to the house of a foreigner,
\verse and you groan at your end, 
when your flesh and body are consumed,
\verse and say “How I hated discipline, 
and I despised reproof!”
\verse and “I did not listen to the voice of my teachers, 
and I did not incline my ear to my instructors!
\verse I was almost at utter ruin 
in the midst of the assembly and congregation.”
\verse Drink water from your \textit{own} cistern 
and flowing waters from inside your own well.
\verse Shall your springs be scattered outward? 
In the streets, \textit{shall there be} streams of water?
\verse May they be yours alone, 
and not for strangers \textit{who are} with you.
\verse May your fountain be blessed, 
and rejoice in the wife of your youth.
\verse \textit{She is} a deer of love and a doe of grace; 
may her breasts satisfy you \textit{always}; 
by her love may you be intoxicated continually.
\verse Why should you be intoxicated, my child, by a strange woman, 
and embrace the bosom of a foreigner?
\verse For before the eyes of Adonai \textit{are} \textit{human ways}, 
and all his paths he examines.
\verse His iniquities shall ensnare him, the evildoer, 
and in the vanity of his sin he shall be caught.
\verse He shall die \textit{for lack of} discipline, 
and in the greatness of his folly he shall be lost.
\end{biblechapter}

\begin{biblechapter} % Proverbs 6
\verseWithHeading{Against Pledges} My child, if you have pledged to your neighbor, 
\textit{if} you have bound \textit{yourself} to the stranger,
\verse \textit{if} you are snared by the sayings of your mouth, 
\textit{if} you are caught by the sayings of your mouth,
\verse do this, then, my child, and save yourself, 
for you have come into \textit{the palm of your neighbor’s hand}: 
Go, humble yourself, plead with your neighbor.
\verse Do not give sleep to your eyes, 
or slumber to your eyelids.
\verse Save yourself like a gazelle from a hand, 
or like a bird from the hand of a fowler.
\verseWithHeading{Against Sloth} Go to the ant, lazy! 
Consider its ways and be wise.
\verse It has no chief, 
officer, or ruler.
\verse In the summer, it prepares its food; 
in the harvest, it gathers its sustenance.
\verse How long will you lie down, lazy? 
When will you rise up from your sleep?
\verse A little sleep, a little slumber, 
a little folding of the hands for rest—
\verse like a \textit{robber} shall your poverty come, 
and what you lack like an armed man.
\verseWithHeading{Against Worthlessness} A worthless man, an evil man, 
goes around with \textit{deceitful speech}.
\verse Winking in his eye, shuffling in his foot, 
pointing in his fingers,
\verse perversion in his heart, he devises evil; 
at all times he will send out discord.
\verse Upon \textit{such a man}, suddenly shall his calamity come; 
in a moment he will be damaged and there is no healing.
\verseWithHeading{What Adonai Hates} There are six \textit{things} Adonai hates, 
and seven \textit{things are} abominations of his soul:
\verse haughty eyes, a lying tongue, 
and hands that shed innocent blood,
\verse a devising heart, plans of deception, 
feet that hurry to run to evil,
\verse a false witness who breathes lies 
and sends out discord between brothers.
\verseWithHeading{Commandment and Instruction as Guardians} My child, keep \textit{the} commandment of your father, 
and do not disregard \textit{the} instruction of your mother.
\verse Bind them on your heart continually; 
tie them upon your neck.
\verse \textit{When you walk}, she will lead you, 
\textit{When you lie down}, she will watch over you, 
and \textit{when} you awake, she will converse \textit{with} you.
\verse For \textit{like} a lamp \textit{is} a commandment, and instruction \textit{is} light, 
and the way of life \textit{is the} reproof of discipline,
\verse \textit{in order} to preserve you from an evil woman, 
from the smoothness of \textit{the} tongue of \textit{an adulteress}.
\verse Do not desire her beauty in your heart; 
may she not capture you with her eyelashes.
\verse For \textit{the} price of a woman, a prostitute, \textit{is the} price of a loaf of bread, 
but the \textit{woman belonging to a man} hunts precious life.
\verseWithHeading{Warning Against Relations with a Married Woman} Can a man carry fire in his lap, 
and his clothes not burn?
\verse If a man walks upon the hot coals, 
will his feet not be burned?
\verse Thus, he who goes to the wife of his neighbor, 
any who touches her shall not go unpunished.
\verse \textit{People} do not despise a thief when he steals 
to fill himself when he is hungry.
\verse But \textit{if} he is found, he will pay sevenfold, 
every possession of his house he shall give.
\verse He who commits adultery \textit{with} a woman lacks \textit{sense}, 
he destroys himself who does it.
\verse A wound and dishonor he will find, 
and his disgrace will not be wiped out.
\verse For jealousy \textit{is the} fury of a husband, 
and he will not show restraint on the day of revenge.
\verse He will not accept \textit{any compensation}, 
and he will not be willing, though the bribe is large.
\end{biblechapter}

\begin{biblechapter} % Proverbs 7
\verseWithHeading{Warning Against the Strange Woman} My child, guard my sayings; 
store my commandments with you.
\verse Keep my commands and live, 
and my teaching like \textit{the} \textit{apple of your eye}.
\verse Bind them on your fingers; 
write them on the tablet of your heart.
\verse Say to wisdom, “you are my sister,” 
and you shall call insight, “\textit{intimate friend}.”
\verse \textit{In order} to guard yourself from \textit{an adulteress}, 
from the foreigner who \textit{makes her words smooth}.
\verse For at the window of my house, 
through my lattice, I looked down.
\verse And I saw among the simple, 
I observed among the youth, 
a young man lacking \textit{sense},
\verse passing on the street \textit{at} her corner, 
and he takes the road \textit{to} her house,
\verse at twilight, at \textit{the} day’s evening, 
in \textit{the} midst of night and the darkness.
\verse Then behold! A woman \textit{comes} to meet him 
\textit{with the} garment of a prostitute and \textit{a secret heart}.
\verse She is loud and stubborn; 
her feet do not stay at her house.
\verse Now in the street, now in the square, 
\textit{at} every corner she lies in wait.
\verse She took hold of  him and kissed him. 
Her face was impudent, and she said to him,
\verse “Sacrifices of peace offerings \textit{are} upon me; 
\textit{today} I completed my vows.
\verse So I have come out to meet you, 
to seek your face, and I have found you.
\verse \textit{With} coverings I have adorned my couch, 
spreads of the linen of Egypt;
\verse I have perfumed my bed \textit{with} myrrh, 
aloes, and cinnamon.
\verse Come, let us take \textit{our} fill of love making, 
until the morning let us delight in love.
\verse For there is no man in his home; 
he has gone on a \textit{long journey}.
\verse The bag of money he took in his hand, 
for \textit{on} the day of the full moon he will come home.”
\verse She persuades him with the greatness of her teachings; 
with her smooth lips she compels him.
\verse He goes after her suddenly; 
like an ox to the slaughter he goes, 
and like a stag to the instruction of a fool,
\verse until an arrow pierces his \textit{entrails}, 
like a bird rushing into a snare, 
but he does not know that \textit{it will cost him his life}.
\verseWithHeading{Reiteration of the Warning Against a Strange Woman} And now, my children, listen to me, 
and be attentive to the sayings of my mouth.
\verse May your heart not turn aside to her ways; 
do not stray into her path.
\verse For many slain she has laid low, 
and countless \textit{are} all of her killings.
\verse The ways of Sheol \textit{are} her house, 
descending to chambers of death.
\end{biblechapter}

\begin{biblechapter} % Proverbs 8
\verseWithHeading{Wisdom Calls} Does not wisdom call, 
and understanding raise its voice?
\verse \textit{Atop the heights} beside the road, 
at the crossroads she stands.
\verse Beside gates, before towns, 
\textit{at the} entrance of doors, she cries out:
\verse “To you, O \textit{people}, I call, 
and my cry is to the children of humankind.
\verse Learn prudence, O simple ones; 
fools, learn \textit{intelligence}.
\verse Listen! For noble things I will speak, 
and upright things from the opening of my lips.
\verse My mouth will utter truth, 
and wickedness \textit{is} an abomination to my lips.
\verse All sayings of my mouth \textit{are} in righteousness; 
none of them are twisted and crooked.
\verse All of them are straight to him who understands, 
and upright to those who find knowledge.
\verse Take my teaching and not silver; 
may you choose knowledge rather than choice gold.
\verse For wisdom is better than jewels, 
and all desires shall not compare with her.
\verse I, wisdom, live with prudence, 
and I find knowledge and discretion.
\verse The fear of Adonai \textit{is} hatred of evil, 
pride, and arrogance and an evil way. 
And I hate a mouth of perversity.
\verse \textit{Advice and sound judgment} are mine; 
I am understanding, \textit{strength is mine}.
\verse By me kings reign, 
and rulers decree righteousness.
\verse By me rulers rule, 
and nobles—all judges of righteousness.
\verse I love \textit{those who love me}, 
and those who seek me diligently shall find me.
\verse Fortune and glory \textit{are} with me, 
enduring wealth and righteousness.
\verse My fruit is better than gold, even refined gold, 
and my yield than choice silver.
\verse In the way of righteousness I walk, 
in the midst of paths of justice,
\verse \textit{in order} to endow those who love me \textit{with} wealth, 
and I will fill their treasuries.
\verseWithHeading{Wisdom at Creation} “Adonai possessed me, \textit{the} first of his ways, 
before his acts \textit{of old}.
\verse From eternity, I was set up from \textit{the} first, 
from the beginning of the earth.
\verse When there were no depths, I was brought forth, 
when there were no springs of \textit{abounding} water.
\verse Before mountains had been shaped, 
before hills, I was brought forth.
\verse When he had not yet made earth and fields, 
or the first dust of the world,
\verse \textit{when he established} the heavens, there I \textit{was}, 
\textit{when he drew} a circle upon the face of the deep,
\verse \textit{when he made} skies from above, 
when he founded fountains of the deep,
\verse \textit{when he assigned} his limits to the sea, 
that waters shall not transgress his \textit{command}, 
\textit{when he marked} the foundations of the earth,
\verse I was \textit{beside} him, a master workman, 
and I was delighting day by day, 
rejoicing before him \textit{always},
\verse rejoicing in the world of his earth, 
and my delight \textit{was} with the children of humankind.
\verseWithHeading{Benefits of Following Wisdom} “And now, children, listen to me; 
happy \textit{are} those who will keep my ways.
\verse Hear teaching and be wise; 
do not neglect \textit{it}.
\verse Happy \textit{is the} person who listens to me, 
\textit{in order} to keep watch at my doors day by day, 
\textit{in order} to guard the frames of my entrances.
\verse For he who finds me \textit{is} he who finds life, 
and he obtains favor from Adonai.
\verse But he who misses me injures himself. 
All those who hate me love death.”
\end{biblechapter}

\begin{biblechapter} % Proverbs 9
\verseWithHeading{Wisdom’s Banquet} Wisdom has built her house; 
she has hewn her seven pillars.
\verse She has slaughtered her slaughtering, mixed her wine, 
\textit{and} also set her table.
\verse She has sent her servant girls, she calls 
upon the wings of the high places of town,
\verse “Whoever \textit{is} simple, let him turn here.” 
As for the one who lacks \textit{sense}, she says to him,
\verse “Come, eat with my bread; 
drink with the wine I have mixed.
\verse Lay aside simplicity and live; 
walk in the way of understanding.”
\verse He who corrects a scoffer gains abuse for himself, 
and he who rebukes the wicked \textit{gets hurt}.
\verse Do not rebuke a scoffer, lest he hate you; 
rebuke the wise and he will love you.
\verse Give to a wise one and he will become more wise; 
\textit{teach} a righteous one and he will increase learning.
\verseWithHeading{Foolishness’ Banquet} The start of wisdom is fear of Adonai, 
and knowledge of the Holy One, insight.
\verse For by me your days shall increase, 
and years of life shall multiply for you.
\verse If you are wise, you are wise for yourself, 
and \textit{if} you scoff, alone you shall bear \textit{it}.
\verse A woman of foolishness is loud, 
simple, and does not know \textit{it}.
\verse She sits at the door of her house, 
upon a throne \textit{at the} high places of town,
\verse \textit{in order} to call to those who pass by the road, 
those who go straight \textit{on} their way:
\verse “Whoever is simple, may he turn here!” 
As for he who lacks \textit{sense}, she says to him,
\verse “Stolen waters are sweet, 
and bread of secrecy is pleasant.”
\verse But he does not know that the dead \textit{are} there, 
in the depths of Sheol \textit{are} her guests.
\end{biblechapter}

\begin{biblechapter} % Proverbs 10
\verseWithHeading{Proverbs of Solomon} The proverbs of Solomon: 
A wise child makes a father glad, 
but a foolish child grieves his mother.
\verse Treasures of wickedness do not profit, 
but righteousness delivers from death.
\verse Adonai will not cause a righteous person to go hungry, 
but the craving of the wicked he will thwart.
\verse A slack hand causes poverty, 
but the hand of the diligent enriches.
\verse He who gathers in the summer \textit{is} a child who is prudent; 
he who sleeps at the harvest \textit{is} a child who brings shame.
\verse Blessings \textit{belong} to the head of the righteous, 
but the mouth of the wicked conceals violence.
\verse The memory of righteousness \textit{is} like a blessing, 
but the name of the wicked will rot.
\verse A heart of wisdom will heed commandments, 
but a babbling fool will come to ruin.
\verse Whoever walks in integrity will walk securely, 
but whoever follows perversity, his ways will be made known.
\verse The winking of an eye causes trouble, 
and the foolishness of lips comes to ruin.
\verse A fountain of life \textit{is} a mouth of righteousness, 
and a mouth of wickedness conceals violence.
\verse Hatred stirs up strife, 
but love covers over all offenses.
\verse On the lips of one who has understanding, wisdom is found, 
but a rod \textit{is} for the back of one who lacks \textit{sense}.
\verse Those who are wise lay up knowledge, 
but \textit{to} the mouth of the fool, ruin draws near.
\verse The wealth of the rich \textit{is} the city of his strength; 
the ruin of the poor \textit{is} their poverty.
\verse The wage of the righteous \textit{leads} to life; 
the gain of the wicked to sin.
\verse \textit{On} the path to life is he who guards instruction, 
but he who rejects rebuke goes astray.
\verse He who conceals hatred \textit{has} lips of deceit, 
and he who utters slander—he is a fool.
\verse In many words, transgression is not lacking, 
but he who restrains his lips is prudent.
\verse Choice silver is a tongue of righteousness, 
a heart of wickedness is \textit{of little worth}.
\verse Lips of righteousness feed many, 
but fools \textit{die for lack of sense}.
\verse The blessing of Adonai makes \textit{one} rich, 
and he does not increase sorrow with it.
\verse \textit{It is} like a sport for a fool to do wrong, 
wisdom for a person of understanding.
\verse The dread of the wicked will come upon him, 
but the desire of the righteous will be granted.
\verse With the passing of the tempest, there is no wickedness, 
but \textit{the} righteous have a foundation forever.
\verse Like vinegar to the tooth and like smoke to the eyes, 
thus \textit{is} the lazy to one who employs him.
\verse The fear of Adonai adds life, 
but the years of the wicked are shortened.
\verse The hope of the righteous \textit{is} gladness, 
but the expectation of the wicked comes to nothing.
\verse A stronghold for the upright is the way of Adonai, 
but ruin \textit{belongs} to \textit{evildoers}.
\verse The righteous one is forever; he will not be removed. 
But the wicked will not remain in the land.
\verse The mouth of the righteous brings forth wisdom, 
but a tongue of perversity will be cut off.
\verse The lips of the righteous know \textit{the} acceptable, 
but the mouth of the wicked, the perverse.
\end{biblechapter}

\begin{biblechapter} % Proverbs 11
\verse Balances of deceit \textit{are} an abomination of Adonai, 
but \textit{an accurate weight} delights him.
\verse Pride comes, then disgrace comes, 
but wisdom \textit{is} with the humble.
\verse The integrity of the upright guides them, 
but the crookedness of the treacherous destroys them.
\verse Wealth does not profit on the day of wrath, 
but righteousness will deliver from death.
\verse The righteousness of the blameless will keep his ways straight, 
but the wicked will fall by his wickedness.
\verse The righteousness of the upright will save them, 
but by a scheme the treacherous will be taken captive.
\verse With the death of a wicked person, hope will die, 
and the expectation of the godless perishes.
\verse The righteous is delivered from trouble, 
but the wicked enters into it.
\verse With a mouth, the godless shall destroy his neighbor, 
but by knowledge the righteous are delivered.
\verse When good \textit{is with} the righteous, \textit{the} city rejoices, 
and with the perishing of the wicked, jubilation.
\verse By the blessing of the upright, a city will be exalted, 
but by the mouth of the wicked, it will be overthrown.
\verse He who lacks \textit{sense} belittles his neighbor, 
but a person of intelligence will remain silent.
\verse A gossip walks \textit{about} telling a secret, 
but the trustworthy in spirit keeps \textit{the} matter.
\verse Where there is no guidance, a nation shall fall, 
but \textit{there is} safety in an abundance of counsel.
\verse He will suffer trouble when he loans to a stranger, 
but he who refuses a pledge is safe.
\verse A woman of grace receives honor, 
but the ruthless gets wealth.
\verse A person of kindness rewards himself, 
but a cruel \textit{person} harms his own flesh.
\verse The wicked earns \textit{deceptive gain}, 
but he who sows righteousness, a \textit{true reward}.
\verse He who is steadfast in righteousness \textit{is} to life 
as he who pursues evil \textit{is} to death.
\verse An abomination of Adonai \textit{are} the \textit{crooked of heart}, 
but his delight \textit{are} \textit{those with blameless ways}.
\verse \textit{Rest assured}, the wicked will not go unpunished, 
but the offspring of the righteous will escape.
\verse A ring of gold in \textit{the} snout of a pig 
\textit{is} a woman \textit{who is} beautiful but without discretion.
\verse The desire of the righteous \textit{is} only good, 
but the expectation of the wicked, wrath.
\verse There is one who gives yet \textit{grows richer}, 
but he who withholds \textit{what is right} only \textit{finds} need.
\verse A person of blessing will be enriched, 
and he who gives water also will be refreshed.
\verse He who withholds grain, the people curse him, 
but a blessing \textit{is} for the head of him who sells.
\verse He who diligently seeks good seeks favor, 
but he who inquires of evil, it will come to him.
\verse He who trusts in his wealth \textit{is} he who will fall, 
but like a green leaf the righteous will flourish.
\verse He who brings trouble \textit{to} his household, he will inherit wind, 
and a fool will serve the wise of heart.
\verse The fruit of righteousness is a tree of life, 
and he who captures souls \textit{is} wise.
\verse If the righteous on earth will be repaid, 
\textit{how much more} the wicked and sinner.
\end{biblechapter}

\begin{biblechapter} % Proverbs 12
\verse He who loves discipline loves knowledge, 
but he who hates rebuke \textit{is} stupid.
\verse The good obtains favor from Adonai, 
but anyone who schemes, he condemns.
\verse A person will not be established by wickedness, 
but the root of the righteous will not be moved.
\verse A woman of strength \textit{is} the crown of her master, 
but like rot in his bones is she who brings shame.
\verse The thoughts of the righteous are just; 
the advice of the wicked is treacherous.
\verse The words of the wicked \textit{are} an ambush of blood, 
but the mouth of the upright delivers them.
\verse The wicked are overthrown and \textit{are no more}, 
but the house of the righteous shall stand.
\verse For his mouth of good sense, a man will be recommended, 
but he who is of perverse \textit{mind} will be despised.
\verse It is better to be lowly and a servant \textit{to someone} 
than self-glorifying and lacking food.
\verse The righteous knows the life of his animal, 
but the compassion of the wicked is cruel.
\verse He who works his land will have plenty of food, 
but he who follows worthless things lacks \textit{sense}.
\verse The wicked covets the proceeds of evil, 
but the root of the righteous \textit{bears fruit}.
\verse By the transgression of lips, evil is ensnared, 
but the righteous escapes from trouble.
\verse From the fruit of the mouth of a man, he is filled with good, 
and the reward of \textit{a man’s labor} will return to him.
\verse The way of a fool is upright in his \textit{own} eyes, 
but he who listens to advice \textit{is} wise.
\verse As for a fool, \textit{on that very day} he makes his anger known, 
but he who ignores an insult \textit{is} prudent.
\verse He who will speak truth will reveal righteousness, 
but the witness of falsehood, deceit.
\verse There is one who speaks rashly, like the thrust of a sword, 
but the tongue of the wise brings healing.
\verse A lip of truth endures forever, 
but a tongue of deception lasts only a moment.
\verse Deceit \textit{is} in the heart of \textit{those who plan} evil, 
but to \textit{those who plan} peace, \textit{there is} joy.
\verse \textit{No evil will happen} to the righteous, 
but the wicked are filled \textit{with} trouble.
\verse An abomination of Adonai are lips of deceit, 
but they who act faithfully \textit{are} his delight.
\verse A clever person conceals knowledge, 
but the heart of a fool announces folly.
\verse The hand of the diligent ones will rule, 
but the lazy will belong to forced labor.
\verse Anxiety in the heart of a man will weigh him down, 
but a good word will cheer him.
\verse A righteous person will seek out his neighbor, 
but the way of the wicked will lead them astray.
\verse The lazy will not roast his game, 
but diligence is the precious wealth of a man.
\verse On the road of righteousness is life, 
and \textit{on} the way of the path, \textit{may there be} no death.
\end{biblechapter}

\begin{biblechapter} % Proverbs 13
\verse A wise child \textit{hears} the discipline of a father, 
but a scoffer does not listen to a rebuke.
\verse From the fruit of the mouth of a man, he shall eat what is good, 
but the desire of the treacherous, wrongdoing.
\verse He who keeps his mouth guards his life; 
he who opens his lips, ruin \textit{belongs} to him.
\verse The soul of the lazy craves, but there is nothing, 
but the person of diligence is richly supplied.
\verse The righteous hates a word of falsehood; 
the wicked will bring shame and \textit{disgrace}.
\verse Righteousness will guard the upright of way, 
but wickedness will overthrow sin.
\verse There is one who acts rich but \textit{has nothing}; 
another who pretends to be poor but \textit{has} wealth.
\verse The ransom of the life of a man \textit{is} his wealth, 
but the poor does not receive a threat.
\verse The light of the righteous will rejoice, 
but the lamp of the wicked will die out.
\verse Only by insolence is strife set up, 
and wisdom \textit{is} with those who take advice.
\verse Wealth \textit{gained} from haste will dwindle, 
but he who gathers \textit{little by little} will increase it.
\verse Hope that is deferred makes the heart sick, 
but a desire fulfilled \textit{is} a tree of life.
\verse He who despises a word will bring destruction on himself, 
but he who respects a commandment will be rewarded.
\verse The teaching of the wise \textit{is} a fountain of life, 
\textit{in order} to avoid the snares of death.
\verse Good sense grants favor, 
but the way of the faithless is coarse.
\verse Anyone who is clever will act with intelligence, 
but the fool will display folly.
\verse A messenger of wickedness will fall into trouble, 
but an envoy of the faithful \textit{brings} healing.
\verse Poverty and disgrace \textit{belong to} him who ignores instruction, 
but he who guards reproof will be honored.
\verse A desire \textit{fulfilled} will be sweet to the soul, 
but an abomination of fools is turning from evil.
\verse Walk with the wise and be wise, 
but as for the companion of fools, he will suffer harm.
\verse Misfortune will pursue sinners, 
but the righteous \textit{will be rewarded with prosperity}.
\verse He who is good will leave an inheritance to \textit{his grandchildren}, 
and stored up for the righteous \textit{is} the wealth of a sinner.
\verse \textit{There is} much food \textit{in} the field of the poor, 
but \textit{it is swept away} by \textit{injustice}.
\verse He who withholds his rod hates his child, 
but he who loves him \textit{gives him} discipline.
\verse The righteous eats to satisfy his life, 
but the belly of the wicked will lack.
\end{biblechapter}

\begin{biblechapter} % Proverbs 14
\verse The \textit{wisest of women} builds her house, 
but the foolish tears it down with her hands.
\verse He who walks in uprightness fears Adonai, 
but he who is devious \textit{in} his ways displeases him.
\verse In the mouth of a fool is the rod of pride, 
but the lips of the wise preserve them.
\verse \textit{When there are no} oxen the manger is empty, 
but an abundance of crops \textit{comes} by the strength of an ox.
\verse A faithful witness does not lie, 
but he who breathes out falsehood is a witness of deceit.
\verse A scoffer seeks wisdom, but there is none, 
but knowledge \textit{comes easily to him who understands}.
\verse Leave \textit{the presence of a foolish man}, 
\textit{for} you will not \textit{come to} know \textit{words of knowledge}.
\verse The wisdom of the clever is understanding his ways, 
but the folly of fools is deceit.
\verse Fools mock \textit{the} guilt offering, 
but among the upright, it is favorable.
\verse The heart knows the bitterness of its soul, 
but in its joy, it will not share itself with a stranger.
\verse The house of the wicked will be destroyed, 
but the tent of the upright will flourish.
\verse There is a way \textit{that seems} upright \textit{to} a man, 
but its end \textit{is} the way of death.
\verse Even in laughter, a heart may be sad, 
and the end of joy \textit{may be} grief.
\verse From his ways, the perverse of heart will be satisfied, 
and \textit{from his own}, \textit{so shall} a good man.
\verse The simple will believe every word, 
but the clever will consider his step.
\verse The wise is cautious and turns from evil, 
but the fool throws off restraint and is confident.
\verse He who is short of \textit{temper} will act foolishly, 
and the man who schemes will be hated.
\verse The simple are adorned \textit{with} folly, 
but the clever are crowned with knowledge.
\verse The evil bow down before the good, 
and the wicked at the gates of the righteous.
\verse The poor is disliked even by his neighbor, 
but the lovers of the rich are many.
\verse He who despises his neighbor is a sinner, 
but he who has mercy on the poor blesses him.
\verse Have they not erred, those who plan evil? 
But loyalty and faithfulness \textit{belong to} those who plan good.
\verse In all toil, there is profit, 
but the \textit{talk} of lips \textit{leads} only to poverty.
\verse The crown of the wise \textit{is} their wealth; 
the folly of fools is folly.
\verse He who saves lives \textit{is} a witness of truth, 
but he who utters lies \textit{is} a betrayer.
\verse In the fear of Adonai, there is confidence of strength, 
and for his children, there will be refuge.
\verse The fear of Adonai \textit{is} a fountain of life, 
\textit{in order} to turn from the snares of death.
\verse In the multitude of people is the glory of the king, 
but without a population, a prince \textit{is} ruined.
\verse \textit{He who is} slow to anger \textit{has} great understanding, 
but the hasty of spirit exalts folly.
\verse A heart of tranquility \textit{is} life \textit{to the} flesh, 
but causes bones of passion to rot.
\verse He who oppresses the poor insults him who made him, 
but he who has mercy on the poor honors him.
\verse By his evildoing, the wicked will be overthrown, 
and the righteous will find refuge in his death.
\verse In the heart of him who has understanding, wisdom rests, 
but \textit{even} in the midst of fools it becomes known.
\verse Righteousness will exalt a nation, 
but sin \textit{is} a reproach to a people.
\verse The favor of a king \textit{is} for the servant who deals wisely, 
but his wrath will be \textit{on} him who acts shamefully.
\end{biblechapter}

\begin{biblechapter} % Proverbs 15
\verse A soft answer will turn away wrath, 
but a word of trouble will stir \textit{anger}.
\verse The tongue of the wise will dispense knowledge, 
but the mouth of fools will pour out folly.
\verse In every place, the eyes of Adonai 
keep watch over the evil and the good.
\verse Gentleness of tongue is a tree of life, 
but perverseness in it \textit{causes} a break in spirit.
\verse A fool will despise the instruction of his father, 
but he who guards reproof is prudent.
\verse \textit{In} the house of the righteous \textit{there is} much treasure, 
but the income of the wicked brings trouble.
\verse The lips of the wise will spread knowledge, 
but the heart of fools, not so.
\verse The sacrifice of the wicked is an abomination of Adonai, 
but the prayer of the upright \textit{is} his delight.
\verse An abomination of Adonai is the way of the wicked, 
but he who pursues righteousness he will love.
\verse Severe discipline \textit{belongs} to him who forsakes the way; 
he who hates a rebuke will die.
\verse Sheol and Abaddon \textit{are} before Adonai, 
\textit{how much more} the hearts of the children of men!
\verse A scoffer does not like \textit{his rebuke}; 
to the wise he will not go.
\verse A heart of gladness will make good \textit{countenance}, 
but in sorrow of heart a spirit is broken.
\verse The heart of him who understands will seek knowledge, 
but the faces of fools, they will feed on folly.
\verse All the days of the poor \textit{are} hard, 
but goodness of heart \textit{is} a continuous feast.
\verse Better \textit{is} little with the fear of Adonai 
than great treasure and trouble with it.
\verse Better \textit{is} a dinner of vegetables when love \textit{is} there 
than a fattened ox and hatred with it.
\verse A man who is hot-tempered will stir up strife, 
but he who is slow \textit{to} \textit{anger}, he will calm contention.
\verse The way of the lazy is like a hedge of thorns, 
but the path of the upright \textit{is} a highway.
\verse A child of wisdom will make a father glad, 
but a \textit{foolish person}, he despises his mother.
\verse Folly is a joy to him who lacks \textit{sense}, 
and a person of understanding \textit{will walk upright}.
\verse Plans go wrong when there is no counsel, 
but with many advisors it will succeed.
\verse Joy \textit{belongs} to a man with answers in his mouth, 
and a word in its time, how good \textit{it is}!
\verse The path of life \textit{leads} upward for him who has insight, 
in order to turn away from Sheol below.
\verse The house of the proud, Adonai will tear \textit{it} down, 
but he will maintain the \textit{property line} of the widow.
\verse Plans of evil \textit{are} an abomination of Adonai, 
but gracious words are pure.
\verse He who makes trouble for his house \textit{is} he who is greedy for unjust gain, 
but he who hates bribes will live.
\verse A heart of righteousness will ponder the answer, 
but a mouth of wickedness will pour out deceit.
\verse Adonai is far from the wicked, 
but the prayers of the righteous he will hear.
\verse From the light of the eyes, the heart will rejoice, 
and good news will enliven the bones.
\verse The ear of him who listens \textit{to} admonitions of life, 
in the midst of the wise it will lodge.
\verse He who ignores instruction despises himself, 
but he who hears admonition gains heart.
\verse Fear of Adonai \textit{is} the instruction of the wise, 
and before honor \textit{comes} humility.
\end{biblechapter}

\begin{biblechapter} % Proverbs 16
\verse To mortals \textit{belong} the plans of the heart, 
but from Adonai \textit{comes} the answer of the tongue.
\verse All the ways of a man \textit{are} pure in his own eyes, 
but Adonai weighs the spirit.
\verse Commit your work to Adonai, 
and your plans will be established.
\verse All Adonai has made \textit{is} for his purpose, 
and even the wicked for the day of trouble.
\verse An abomination of Adonai \textit{are} all who are arrogant of heart; 
\textit{rest assured}, he will not go unpunished.
\verse By loyalty and faithfulness, iniquity will be covered over, 
and by fear of Adonai \textit{one turns} from evil.
\verse When the ways of a man are pleasing \textit{to} Adonai, 
even his enemies he will cause to make peace with him.
\verse Better is little with righteousness 
than great income with no justice.
\verse The \textit{mind} of a person will plan his ways, 
and Adonai will direct his steps.
\verse A decision is upon the lips of a king; 
in judgment his mouth will not sin.
\verse A balance and scales of justice \textit{belong} to Adonai; 
all the weights of the bag \textit{are} his work.
\verse An abomination to kings \textit{is} doing evil, 
for by righteousness the throne will be established.
\verse The delight of kings \textit{are} the lips of righteousness, 
and he who speaks what is upright he will love.
\verse The wrath of a king \textit{is} a messenger of death, 
but one who is wise will appease it.
\verse In the light of the face of the king \textit{there is} life, 
and his favor \textit{is} like a cloud of spring rain.
\verse Getting wisdom: \textit{how much} better than gold! 
And getting understanding: it is chosen \textit{over} silver.
\verse The highway of the upright, it turns from evil, 
he who guards himself keeps his way.
\verse Before destruction \textit{comes} pride, 
and before a fall, a haughty spirit.
\verse Better a lowly spirit with the poor 
than dividing the spoil with the proud.
\verse He who is attentive to a matter will find goodness, 
and he who trusts in Adonai, his own happiness.
\verse The wise of heart is called perceptive, 
but he who is pleasant of lips will increase persuasiveness.
\verse A fountain of life is wisdom for its owner, 
but the instruction of fools is folly.
\verse The heart of the wise will make his mouth judicious, 
and upon his lips, it will add persuasiveness.
\verse Pleasant sayings \textit{are} a honeycomb, 
sweetness to the soul and healing to the bones.
\verse There is a way \textit{that seems} upright \textit{to} a man, 
but its end \textit{is} the way of death.
\verse The life of a worker works for him, 
for his \textit{hunger} urges him.
\verse A man of wickedness concocts evil, 
and his lips \textit{are} like a scorching fire.
\verse A person of perversity will spread dissent, 
and he who whispers separates a close friend.
\verse A person of violence will entice his neighbor 
and cause him to walk on a way \textit{that is} not good.
\verse He who winks his eyes \textit{does so in order} to plan perverse things; 
he who purses his lips will bring evil to pass.
\verse A crown of glory is gray hair; 
by a righteous life it is gained.
\verse He who is slow to anger \textit{is} better than him who is mighty, 
and he who controls his spirit than him who captures a city.
\verse The lot will be cast into the lap, 
but all of its decisions are from Adonai.
\end{biblechapter}

\begin{biblechapter} % Proverbs 17
\verse Better a dry morsel and quiet with it 
than a house filled \textit{with} feasts of strife.
\verse A slave who deals wisely will rule over a child who acts shamefully, 
and in the midst of brothers he will share his inheritance.
\verse A crucible \textit{is} for the silver, and a furnace \textit{is} for the gold, 
but Adonai will test hearts.
\verse He who does evil listens to lips of wickedness, 
\textit{and} the liar gives heed to the tongue of mischief.
\verse He who mocks the poor insults him who made him; 
he who rejoices at calamity will not go unpunished.
\verse The crown of the elderly \textit{are} \textit{grandchildren}, 
and the glory of children \textit{is} their fathers.
\verse \textit{Fine speech} \textit{is} not becoming a fool, 
\textit{still less} \textit{is} \textit{false speech} for a ruler.
\verse The bribe \textit{is} a stone of magic in the eyes of its owner; 
\textit{everywhere} he will turn, he will prosper.
\verse He who forgives an affront fosters love, 
but he who waits on a matter will alienate a friend.
\verse A rebuke strikes him who understands 
deeper than one hundred blows to a fool.
\verse An evil person will seek only rebellion, 
and a cruel messenger will be sent against him.
\verse \textit{May} a man meet a she-bear robbed of offspring 
and not a fool in his folly.
\verse \textit{For} he who returns evil for good, 
evil will not depart from his house.
\verse \textit{Like} the release of water \textit{is} the beginning of strife; 
before it breaks out, stop the quarrel.
\verse He who justifies the wicked and he who condemns the righteous, 
the two of them \textit{are} both abominations of Adonai.
\verse Why \textit{is} this? A price in the hand of a fool, 
\textit{in order} to buy wisdom where there is no \textit{sense}.
\verse The friend loves at all times, 
but a brother is born for adversity.
\verse A person who lacks \textit{sense} \textit{pledges}; 
he becomes security before his neighbor.
\verse He who loves transgression loves strife; 
he who builds his high thresholds seeks destruction.
\verse He who is crooked of heart will not find goodness, 
and he who is perverse, by his tongue he will fall into calamity.
\verse He who begets a fool, \textit{there is} trouble for him; 
the father of a fool will not rejoice.
\verse A cheerful heart is good medicine, 
but a downcast spirit will dry out bones.
\verse The wicked will accept a bribe from the lap, 
\textit{in order} to pervert the ways of justice.
\verse He who understands sets \textit{his} face \textit{toward} wisdom, 
but the eyes of a fool, to the end of the earth.
\verse A grief to his father \textit{is} the child of a fool, 
and bitterness to her who bore him.
\verse Also, imposing a fine on the righteous is not good, 
\textit{nor} to flog nobles for uprightness.
\verse He who spares his sayings knows knowledge, 
and a cool spirit is a man of understanding.
\verse Even a fool who keeps silent \textit{shall be considered wise}; 
he who closes his lips \textit{is} intelligent.
\end{biblechapter}

\begin{biblechapter} % Proverbs 18
\verse He who is selfish seeks a craving; 
against all sound judgment he shows contempt.
\verse A fool will not take pleasure in understanding, 
but in expressing his heart.
\verse With the coming of wickedness comes contempt also, 
and with dishonor, disgrace.
\verse Deep waters \textit{are} words of the mouth of a man; 
a gushing stream \textit{is} a fountain of wisdom.
\verse Being partial \textit{to} faces of evil \textit{is} not good, 
\textit{nor} to subvert the righteous at the judgment.
\verse The lips of a fool will bring strife, 
and his mouth calls out for a flogging.
\verse The mouth of a fool \textit{is} ruin to him, 
and his lips \textit{are} a snare to his soul.
\verse The words of a whisper \textit{are} like delicious morsels, 
and they themselves go down \textit{to} inner parts of the body.
\verse Even he who is slack in his work, 
he is brother to a master of destruction.
\verse A tower of strength is the name of Adonai; 
into him the righteous will run and be safe.
\verse The wealth of the rich \textit{is} \textit{his strong city}, 
and like a wall, it is high in his imagination.
\verse In the presence of destruction, the heart of a man will be haughty, 
but in the presence of honor, humble.
\verse He who returns a word before he will hear, 
folly itself \textit{belongs} to him as well as shame.
\verse The spirit of a man will endure his sickness, 
but a broken spirit, who may bear it?
\verse An intelligent \textit{mind} will acquire knowledge, 
and the ear of the wise will seek knowledge.
\verse The gift of a person will open doors for him, 
and before the great, it gives him access.
\verse The first in his dispute is \textit{deemed} righteous, 
\textit{but} his neighbor will come and examine him.
\verse The lot will put an end to disputes, 
and between powerful contenders it will decide.
\verse A brother who is offended \textit{is worse} than a city of strength, 
and quarrels \textit{are} like the bars of a fortification.
\verse From the fruit of a man’s mouth, his stomach will be satisfied, 
as for the yield of his lips, it will satisfy.
\verse Death and life \textit{are} in the \textit{power} of the tongue, 
and those who love her will eat of her fruit.
\verse He who finds a wife finds good, 
and he will obtain favor from Adonai.
\verse The poor may speak entreaties, 
but the rich will answer roughly.
\verse A man of \textit{many} friends will come to ruin, 
but there is a friend who sticks closer than a brother.
\end{biblechapter}

\begin{biblechapter} % Proverbs 19
\verse Better a poor person walking in integrity 
than one who is perverse \textit{in his speech} and is a fool.
\verse Also, a life without knowledge is not good, 
and he who moves quickly with \textit{his} feet \textit{misses the mark}.
\verse As for the folly of humankind, its way leads to ruin, 
and against Adonai his heart will rage.
\verse Wealth adds many friends, 
but the poor will be left by his friends.
\verse A witness of falsehood will not go unpunished, 
and he who breathes lies will not escape.
\verse Many will seek favor before the generous, 
and everyone \textit{is} the friend of a man of gifts.
\verse All the brothers of the poor, \textit{if} they hate him, 
how much more will his friends keep away from him. 
He pursues \textit{them with} words, \textit{and} they are gone.
\verse He who acquires wisdom loves himself; 
he who guards understanding \textit{loves} to find good.
\verse A false witness will not go unpunished, 
and he who breathes lies will perish.
\verse For a fool living in luxury is not fitting, 
\textit{any} more \textit{than it is} for a slave to rule over princes.
\verse The understanding of a person makes him slow to his \textit{anger}, 
and his glory overlooks offense.
\verse The rage of a king growls like a lion, 
but his favor is like dew on the grass.
\verse A foolish child is a ruin to his father, 
and the quarreling of a woman is a continuous dripping.
\verse A house and wealth \textit{are} an inheritance \textit{from} fathers, 
but from Adonai \textit{comes} a woman who is prudent.
\verse Laziness will bring on a deep sleep, 
and a person of idleness will suffer hunger.
\verse He who guards commandments guards his life; 
he who is careless of his ways will be killed.
\verse He who lends to Adonai \textit{is} he who is kind to the poor, 
and his benefits he will repay to him.
\verse Discipline your child, for there is hope, 
but on his destruction do not set your desire.
\verse A hot-tempered person pays a penalty; 
if you rescue \textit{him}, you will do \textit{it} yet again.
\verse Listen \textit{to} advice and accept instruction 
so that you will gain wisdom for your \textit{future}.
\verse Many plans \textit{are} in the heart of a man, 
but the purpose of Adonai will be established.
\verse The craving of a man \textit{is} his steadfast loyalty, 
and \textit{it is} better \textit{to be} poor than a \textit{liar}.
\verse Fear of Adonai \textit{leads} to life; 
he who is filled \textit{with it} will rest—he will not suffer harm.
\verse A lazy person buries his hand in the dish, 
and even to his mouth he will not bring it back.
\verse The scoffer you shall strike, and the simple, may they learn prudence, 
and reprove the intelligent \textit{and} he will gain knowledge.
\verse He who does violence to a father, he who chases away a mother, 
\textit{is} a child who causes shame and brings reproach.
\verse Cease to listen to instruction, my child, 
\textit{and you will stray} from sayings of knowledge.
\verse A worthless witness will mock justice, 
and the mouth of the wicked will devour iniquity.
\verse Judgments are prepared for the scoffers, 
and flogging for the back of fools.
\end{biblechapter}

\begin{biblechapter} % Proverbs 20
\verse Wine \textit{is} a mocker, strong drink a brawler, 
and any who go astray by it are not wise.
\verse Roaring like the lion \textit{is the} dreaded anger of a king; 
he who provokes him forfeits his life.
\verse \textit{It is} honorable for the man to refrain from strife, 
but every fool will be quick to quarrel.
\verse The lazy person will not plow in season; 
he will expect at the harvest, but there \textit{will be} nothing.
\verse Deep waters \textit{are like} purpose in the heart of a man, 
and a man of understanding will draw it out.
\verse Many a person will proclaim his loyalty \textit{for} himself, 
but a man \textit{who is} trustworthy, who can find?
\verse He who walks in his integrity \textit{is} righteous; 
happy are his children who follow him.
\verse A king who sits on \textit{the} throne of judgment 
winnows all evil with his eyes.
\verse Who will say “I have made my heart clean; 
I am pure from my sin”?
\verse Stone and stone, measure and measure, 
\textit{both of them} \textit{are} an abomination of Adonai.
\verse Even by his acts, a \textit{young man} will make himself known, 
whether his acts are pure and upright.
\verse The ear that hears and the eye that sees, 
Adonai has made \textit{them both}.
\verse Do not love sleep, lest you become poor; 
open your eyes \textit{and} have plenty bread.
\verse “Bad, bad,” the buyer will say, 
but \textit{when} one goes to him, then he will boast.
\verse There is gold and many costly stones, 
but precious jewels are lips of knowledge.
\verse Take his garment, for he has given security \textit{to} a stranger, 
and on behalf of a foreigner—take it as pledge.
\verse Bread gained by deceit is sweet for the man, 
but afterward, his mouth will be filled \textit{with} gravel.
\verse A plan will be established by advice, 
and with guidance make war.
\verse He reveals a secret, he who walks \textit{about with} gossip, 
and do not associate with a babbler’s lips.
\verse He who curses his father and his mother, 
his lamp will be extinguished \textit{in the midst of darkness}.
\verse An inheritance acquired \textit{hastily} \textit{at the beginning} 
will not be blessed at its end.
\verse Do not say “I will repay evil”; 
wait for Adonai and he will deliver you.
\verse An abomination of Adonai \textit{is} \textit{a stone and a weight}, 
and scales of falsehood are not good.
\verse \textit{Away} from Adonai are the steps of a strong man, 
and how will humankind understand his ways?
\verse \textit{It is} a snare to humankind to say rashly “\textit{It is} holy,” 
and after vows, to scrutinize.
\verse A wise king winnows the wicked, 
and he will drive a wheel over them.
\verse The lamp of Adonai \textit{is} the spirit of humankind, 
he who searches every \textit{innermost part}.
\verse Loyalty and faithfulness will preserve a king, 
and he is upheld with the righteousness of his throne.
\verse The glory of young men \textit{is} their strength, 
but the beauty of the aged \textit{is} gray hair.
\verse The blows of a wound will cleanse evil, 
as will beatings of the \textit{innermost part}.
\end{biblechapter}

\begin{biblechapter} % Proverbs 21
\verse Streams of water \textit{are} the heart of a king in the hand of Adonai; 
\textit{wherever} he will desire, he will turn.
\verse Every way of a man is upright in his \textit{own} eyes, 
but Adonai weighs hearts.
\verse Doing righteousness and justice is \textit{more} acceptable 
to Adonai than sacrifice.
\verse Haughtiness of \textit{the} eyes and pride of heart, 
\textit{the} lamp of the wicked \textit{are} sin.
\verse The plans of the diligent only \textit{lead} to abundance, 
but all who are hasty, only to want.
\verse He who makes treasure by a lying tongue 
\textit{is a} fleeting vapor \textit{and} seeker of death.
\verse The violence of the wicked will sweep them away, 
for they refuse to do justice.
\verse Crooked is the way of a man and a foreigner, 
but the pure \textit{is} upright \textit{in} his conduct.
\verse Better to dwell on the corner of a roof 
than \textit{to} share a house with a woman of contention.
\verse The soul of the wicked desires evil; 
his neighbor will not find mercy in his eyes.
\verse With the punishment of a scoffer, the simple will become wise, 
and with the instruction of the wise, he will obtain knowledge.
\verse The righteous observes the house of the wicked; 
he throws the wicked to ruin.
\verse He who closes his ear from the cry of the poor, 
he also will cry out and not be heard.
\verse A gift in secret will avert \textit{anger}, 
and a \textit{concealed bribe}, strong wrath.
\verse \textit{It is} a joy to the righteous to do justice, 
but dismay to those who do evil.
\verse Whoever wanders from the way of understanding, 
in the assembly of the dead he will rest.
\verse A man of want \textit{is} he who loves pleasure; 
he who loves wine and oil will not become rich.
\verse A ransom for the righteous is the wicked, 
and \textit{the} faithless instead of the upright.
\verse Better to live in a land of wilderness 
than \textit{with} a wife of quarrels and provocation.
\verse Precious treasure and oil \textit{are} in the house of the wise, 
but the foolish person will devour them.
\verse He who pursues righteousness and kindness 
will find life, righteousness, and honor.
\verse To a city of warriors, the wise ascends, 
and he will bring down the stronghold—its object of trust.
\verse He who guards his mouth and his tongue, 
he guards his life from danger.
\verse The proud, haughty one: “scoffer” \textit{is} his name; 
he acts with arrogance of pride.
\verse The craving of a lazy person will kill him, 
for his hands refuse to work.
\verse \textit{All day} he craves a craving, 
but the righteous will give and not hold back.
\verse The sacrifice of the wicked \textit{is} an abomination; 
how much more when he brings it in divisiveness!
\verse A false witness will perish, 
but a man who listens will testify with success.
\verse A wicked man is strong in \textit{his countenance}, 
but the upright will appoint his paths.
\verse There is no wisdom, nor understanding, 
nor counsel to oppose Adonai.
\verse A horse is prepared for the day of battle, 
but to Adonai \textit{belongs} the victory.
\end{biblechapter}

\begin{biblechapter} % Proverbs 22
\verse A \textit{reputation} \textit{is} to be chosen \textit{rather than} great riches; 
\textit{favor is better than silver and gold}.
\verse Rich and poor have \textit{much} in common; 
Adonai is the maker of all of them.
\verse The clever sees danger and hides, 
but the simple go on and suffer.
\verse The reward of humility \textit{is} the fear of Adonai— 
wealth and honor and life.
\verse Thorns \textit{and} snares \textit{are} in the way of the perverse; 
he who guards himself will keep away from them.
\verse Train the child \textit{concerning his way}; 
even when he is old, he will not stray from it.
\verse The rich will rule over the poor, 
and the borrower is a slave of \textit{the lender}.
\verse He who sows injustice will reap calamity, 
and the rod of his anger will fail.
\verse He \textit{who} \textit{is generous} will be blessed, 
for he gives to the poor from his \textit{own} bread.
\verse Drive out a scoffer and strife will go out; 
quarrel and abuse will cease.
\verse He who loves purity of heart 
\textit{and has}\textit{gracious speech}, his friend \textit{is the} king.
\verse The eyes of Adonai keep watch over knowledge, 
but he will overthrow the words of the faithless.
\verse A lazy person says “A lion in the street! 
In the middle of the highway, I shall be killed!”
\verse A deep pit is the mouth of an \textit{adulteress}, 
\textit{he with whom} Adonai is angry will fall there.
\verse Folly is bound up in the heart of a boy; 
\textit{the} rod of discipline will drive it from him.
\verse He who oppresses the poor \textit{in order} to enrich himself, 
\textit{or} gives to the rich, \textit{will come to poverty}.
\verseWithHeading{Words of the Wise} Incline your ear and hear the words of the wise; 
you shall apply your heart to my teaching.
\verse For \textit{it is} pleasant if you guard them \textit{within you}; 
together they will be ready upon your lips.
\verse \textit{In order} for your trust to be in Adonai, 
I have made \textit{them} known to you \textit{today}—even you.
\verse Have I not written for you thirty \textit{sayings} 
with admonitions and knowledge?
\verse \textit{In order} to show you what is right—sayings of truth— 
\textit{in order} to return a true saying to him who sent you.
\verse Do not rob the poor because he is poor, 
and do not crush the afflicted at the gate;
\verse For Adonai will plead their case 
and despoil those who despoil them of life.
\verse Do not befriend an owner of \textit{anger}, 
and with a man of wrath you shall not associate;
\verse lest you learn his way 
and become entangled in a snare to yourself.
\verse Do not be with those who \textit{give a pledge} 
\textit{by becoming} surety.
\verse If there is nothing for you to pay, 
why will he take your bed from under you?
\verse Do not remove an ancient boundary marker 
which your ancestors made.
\verse A man who is skillful in his work, you shall see: 
before kings, he will serve; 
he will not serve before the commoners.
\end{biblechapter}

\begin{biblechapter} % Proverbs 23
\verse When you sit to eat with a ruler, 
you shall surely observe what \textit{is} before you,
\verse and you shall put a knife to your throat 
if you have a \textit{big appetite}.
\verse Do not desire his delicacies, 
for it is food of deception.
\verse Do not tire \textit{in order} to become rich; 
out of your understanding, \textit{may you} desist.
\verse Your eyes will \textit{alight} on it, but there is nothing \textit{to} it, 
for suddenly it will make for itself wings 
like an eagle and it will be exhausted \textit{in} the heavens.
\verse Do not eat the bread of \textit{the stingy}, 
and do not desire his delicacies.
\verse For, like hair in his \textit{throat}, so it is. 
“Eat and drink!” he will say to you, 
but his heart will not be with you.
\verse Your morsel you have eaten, you will vomit it up, 
and you will waste your pleasant words.
\verse In the ears of a fool do not speak, 
for he will despise the wisdom of your words.
\verse Do not remove an ancient boundary marker, 
and on the fields of orphans do not encroach;
\verse For their redeemer is strong, 
he himself will plead their cause against you.
\verse Apply your heart to instruction, 
and your ear to sayings of knowledge.
\verse Do not withhold discipline from a child, 
if you will beat him with the rod, he will not die.
\verse As for you, with the rod you shall beat him, 
and his life you will save from Sheol.
\verse My child, if your heart is wise, 
my heart will be glad—even me!
\verse And my \textit{insides} will rejoice 
when your lips speak what is upright.
\verse May your heart not envy the sinners, 
but \textit{live} in fear of Adonai \textit{all day}.
\verse \textit{Surely} there is a future, 
and your hope will not be cut off.
\verse You, my child, hear and be wise, 
and direct your heart on the road.
\verse Do not be among drinkers of wine, 
among gluttonous eaters of \textit{their meat}.
\verse For the drunkard and gluttonous, they will become poor, 
and \textit{with} rags, drowsiness will clothe them.
\verse Listen to your father—he who \textit{gave you life}, 
and do not despise your mother when she is old.
\verse Buy truth and do not sell \textit{it}, 
wisdom and instruction and understanding.
\verse The father of the righteous will surely rejoice; 
he who bears a wise person will be happy with him.
\verse May your father and your mother be glad, 
and may she who bore you rejoice.
\verse My child, may you give your heart to me, 
and may your eyes delight in my ways.
\verse For a deep pit is a prostitute, 
and a narrow well is \textit{an adulteress}.
\verse She \textit{is} also like a robber lying in wait, 
and the faithless among mankind she increases.
\verse To whom \textit{is} woe? To whom \textit{is} sorrow? To whom \textit{are} quarrels? To whom \textit{is} complaint? 
To whom \textit{are} wounds without cause? To whom \textit{is} redness of \textit{the} eyes?
\verse To those who linger over wine, 
to those who come to try mixed wine.
\verse Do not look at wine when it is red, 
when it \textit{sparkles} on the cup, 
going down smoothly.
\verse \textit{In the end}, it will bite like a serpent, 
and it will sting like an adder.
\verse Your eyes will see strange things, 
and your heart will speak perverse things.
\verse And you will be like him who lies down in the heart of the sea, 
and like him who lies down on top of a mast.
\verse “They struck me; I was not hurt. They beat me; I did not know \textit{it}. 
When I will awake, I will continue; I will seek it again.”
\end{biblechapter}

\begin{biblechapter} % Proverbs 24
\verse Do not envy men of evil, 
and do not desire to be with them.
\verse For their minds will devise violence, 
and their lips will speak mischief.
\verse By wisdom a house is built, 
and by understanding it is established.
\verse And by knowledge, rooms are filled 
\textit{with} all riches, precious and pleasant.
\verse The warrior of wisdom \textit{is} in strength, 
and a man of knowledge is strong \textit{in} power.
\verse For with wise guidance you shall make war for yourself, 
and victory \textit{is} in an abundance of counsel.
\verse Wisdom \textit{is too} high for fools; 
at the gate he will not open his mouth.
\verse He who plans to do evil for himself, 
they will call \textit{him} “master of mischief.”
\verse Devising folly \textit{is} a sin, 
and an abomination to humankind \textit{is} a scoffer.
\verse \textit{If} you faint on the day of adversity, 
little \textit{is} your strength.
\verse Rescue those who are led away to the death 
and those who stagger to the slaughter. If you hold back,
\verse if you say, “Look, we do not know this,” 
does not he who weighs hearts perceive \textit{it}? 
And he who keeps your soul, 
he knows and will repay humankind according to his deeds.
\verse My child, eat honey, for \textit{it is} good, 
and the dripping of the honeycomb \textit{is} sweet to your taste.
\verse Thus know wisdom for \textit{the sake of} your soul, 
if you find \textit{it}, then there is a future, 
and your hope will not be cut off.
\verse Do not lie in wait \textit{like} an outlaw against the home of the righteous; 
do not do violence \textit{to} his dwelling place.
\verse For seven \textit{times} the righteous will fall, but he will rise, 
but the wicked will be overthrown by calamity.
\verse While your enemies are falling, do not rejoice; 
when he trips himself, may your heart not be glad
\verse lest Adonai see and \textit{it} be evil in his eyes, 
and turn his anger away from him.
\verse Do not fret because of the evildoers; 
do not envy the wicked.
\verse For there will not be a future for the evil; 
the lamp of the wicked will die out.
\verse Fear Adonai, my son, and the king; 
with those who change, do not associate.
\verse For suddenly their disaster will come, 
and the ruin of \textit{both of them}, who knows?
\verse These \textit{sayings} are also for the wise: 
\textit{Partiality} in judgment is not good.
\verse Whoever says to the guilty, “You are righteous,” the people will curse him; 
the nations will abhor him.
\verse But they who rebuke will have delight, 
and upon them blessings of goodness will come.
\verse He will kiss \textit{the} lips, 
he who gives an honest answer.
\verse Prepare your work in the street and get it ready for yourself in the field; 
afterward, then you shall build your house.
\verse Do not be a witness without cause against your neighbor 
nor deceive with your lips.
\verse Do not say, “Just as he has done to me, so shall I do to him; 
I will pay back the man according to his deed.”
\verse I passed by the field of a lazy person, 
and over the vineyard of a person lacking \textit{sense};
\verse and behold, it was overgrown—all of it was covered \textit{with} thorns, its surface with nettles, 
and \textit{its stone wall} was broken down.
\verse Then I myself saw and my heart considered; 
I looked, and I took hold of instruction:
\verse A little sleep, a little slumber, 
a little folding of the hands for rest,
\verse and your poverty will come running, 
and your lack like an armed warrior.
\end{biblechapter}

\begin{biblechapter} % Proverbs 25
\verseWithHeading{More Proverbs of Solomon} These are also proverbs of Solomon which officials of Hezekiah king of Judah copied:
\verse The glory of God conceals things, 
but the glory of kings searches out things.
\verse As heaven is to height and the earth is to depth, 
so \textit{is} the heart of kings—there is no searching.
\verse Remove the dross from silver, 
and it will become a vessel for the smith.
\verse Remove the wicked before a king, 
and his throne will be established in righteousness.
\verse Do not promote yourself before the king, 
and in the place of the great ones do not stand.
\verse For \textit{it is} better \textit{that} he say to you, “Ascend here,” 
than he humble you before a noble. 
What your eyes have seen,
\verse do not hastily bring out to court, 
for what will you do at its end, 
when your neighbor puts you to shame?
\verse Argue your argument with your neighbor \textit{himself}, 
the secret of another do not disclose,
\verse lest he who hears shame you 
and your ill repute will not end.
\verse Apples of gold in a setting of silver 
is a matter spoken at its proper time.
\verse A ring of gold and an ornament of fine gold 
\textit{is} a rebuke of the wise to the ear of a listener.
\verse Like the cold of snow on a day of harvest 
\textit{is} a faithful messenger to those who send him, 
and the soul of his master is refreshed.
\verse \textit{Like} clouds and wind when there is no rain, 
\textit{so too is} a man who boasts in a gift of deception.
\verse \textit{With patience} a ruler may be persuaded, 
and a soft tongue will break a bone.
\verse \textit{If} you find honey, eat what is sufficient for you, 
lest you have your fill of it and vomit it out.
\verse Make your foot scarce in the house of your neighbor, 
lest he become weary of you and hate you.
\verse \textit{Like} a club and sword and a sharp arrow 
\textit{is} a man who bears false witness against his neighbor.
\verse A bad tooth and a lame foot 
\textit{is} the trust of a faithless person \textit{in a time} of trouble.
\verse \textit{Like} one who removes a garment on a cold day, \textit{or like} vinegar on natron, 
\textit{is} he who sings songs to a heavy heart.
\verse If your enemy \textit{is} hungry, feed him bread, 
and if thirsty, let him drink water.
\verse For coals of fire you will heap upon his head, 
and Adonai will reward you.
\verse The wind of the north produces rain, 
and a backbiting tongue, angry faces.
\verse Better to live upon the corner of a roof 
than \textit{with} a woman of contention and \textit{in a} shared house.
\verse \textit{Like} cold water upon a weary soul, 
\textit{so too is} good news from a distant place.
\verse \textit{Like} a muddied spring or a polluted fountain 
is the righteous who gives way before the wicked.
\verse To eat much honey is not good, 
nor \textit{is} seeking one’s honor honorable.
\verse A breached city \textit{where} there is no wall 
\textit{is like} a man who \textit{has no} self-control for his spirit.
\end{biblechapter}

\begin{biblechapter} % Proverbs 26
\verse Like snow in the summer and like rain at the harvest, 
so honor is not fitting for a fool.
\verse Like the sparrow \textit{is} to fluttering and like the swallow \textit{is} to flying, 
so an undeserved curse does not go forth.
\verse A whip for the horse, a bridle for the donkey, 
and a rod for the back of fools.
\verse Do not answer a fool according to his folly 
lest you become like him—even you.
\verse Answer a fool according to his folly, 
or else he will be wise in his own eyes.
\verse \textit{Like} cutting off feet \textit{or} drinking violence, 
so \textit{is} he who sends messages in the hand of a fool.
\verse \textit{Like} legs that hang limp from a lame person, 
so \textit{is} a proverb in the mouth of fools.
\verse Like binding a stone in a sling, 
so \textit{is} giving honor to a fool.
\verse \textit{Like} a thorn that goes up in the hand of a drunkard, 
so \textit{is} a proverb in the mouth of fools.
\verse \textit{Like} an archer who wounds everyone, 
so \textit{is} he who hires a fool or he who hires passersby.
\verse Like a dog returning to his vomit 
\textit{is} a fool reverting to his folly.
\verse \textit{Do} you see a man wise in his own eyes? 
\textit{There is more} hope for a fool than for him.
\verse A lazy person says “A lion \textit{is} in the road! 
A lion among the streets!”
\verse The door turns on its hinge, 
and a lazy person on his bed.
\verse A lazy person buries his hands in the dish; 
he is \textit{too} tired to return it to his mouth.
\verse A lazy person is wiser in his eyes 
than seven who answer discreetly.
\verse Grabbing onto the ears of a dog 
passing by \textit{is} one who meddles in a quarrel \textit{that is} not his own.
\verse Like a maniac who shoots 
firebrands, arrows, and death,
\verse so \textit{is} a man who deceives his neighbor, 
but says “Am I not joking?”
\verse For lack of wood, a fire goes out, 
and where there is no whisperer, quarreling will cease.
\verse \textit{As} charcoal \textit{is} to hot embers and wood \textit{is} to fire, 
so a man of quarrels \textit{is} to kindling strife.
\verse The words of a whisperer \textit{are} like delicious morsels, 
and they go down to the inner parts of the body.
\verse \textit{Like} \textit{impure silver} which overlays an earthen vessel, 
\textit{so are} smooth lips and an evil heart.
\verse On his lips, an enemy will pretend, 
but \textit{inside} he will harbor deceit.
\verse When he makes his voice gracious, do not believe him, 
for seven abominations \textit{are} in his heart.
\verse Though hatred is covered with guile, 
its evil will be exposed in the assembly.
\verse He who digs a pit, in it he will fall, 
and he who rolls a stone, on him it will come back.
\verse A tongue of deceit hates its victim, 
and a flattering mouth makes ruin.
\end{biblechapter}

\begin{biblechapter} % Proverbs 27
\verse Do not boast \textit{about tomorrow}, 
for you do not know what the day will bring.
\verse May another praise you and not your own mouth, 
a stranger and not your own lips.
\verse Heavy \textit{is} a stone and weighty \textit{is} sand, 
but \textit{the} provocation of a fool \textit{is} heavier than both of them.
\verse Cruel \textit{is} wrath and overwhelming \textit{is} anger, 
but who will stand before jealousy?
\verse Better a rebuke that is open 
than a love that is hidden.
\verse The wounds of a friend mean well, 
but the kisses of an enemy are profane.
\verse An \textit{appetite} that is sated spurns honey, 
but \textit{to} an \textit{appetite} that is ravenous, all bitterness \textit{is} sweet.
\verse Like a bird that strays from its nest, 
so is a man who strays from his place.
\verse Perfume and incense will gladden a heart, 
and the pleasantness of one’s friend \textit{is} \textit{personal advice}.
\verse As for your friend and a friend of your father, do not forsake \textit{them}, 
and the house of your brother, do not enter on the day of your calamity. 
Better is a close neighbor than a distant brother.
\verse Be wise, my child, and make my heart glad, 
and I will answer him who reproaches me \textit{with} a word.
\verse \textit{When} the clever sees danger, he hides; 
the simple go on \textit{and} suffer.
\verse Take his garment, for he gives surety \textit{to} a stranger, 
and to \textit{an adulteress}—\textit{so} take his pledge.
\verse He who blesses his neighbor with a loud voice 
early in the morning, 
a curse will be reckoned to him.
\verse Dripping constantly on a day of heavy rain 
and a woman of contention are alike.
\verse \textit{In} restraining her, he restrains wind, 
and his right hand will grasp oil.
\verse \textit{As} iron sharpens iron, 
so \textit{one man sharpens another}.
\verse He who tends a fig tree will eat its fruit, 
and he who guards his master will be honored.
\verse As the waters \textit{reflect} \textit{face to face}, 
so the heart of a person \textit{reflects} the person.
\verse Sheol and Abaddon will not be satisfied, 
and the eyes of a person will not be satisfied \textit{either}.
\verse A crucible \textit{is} for the silver, and a furnace for the gold, 
but a man \textit{is tested by} the mouth of him who praises him.
\verse If you crush a fool in the mortar with the pestle \textit{along with} the crushed grain, 
it will not drive folly from upon him.
\verse You will surely know the \textit{condition} of your flock; 
your heart attends to the herds.
\verse For riches \textit{are} not forever, 
nor a crown for generation \textit{after} generation.
\verse When the grass is gone, then green growth will appear, 
and the herbs of the mountains will be gathered.
\verse Lambs \textit{will be} your clothing, 
and goats the price of the field.
\verse And \textit{there will be} enough goats’ milk for your food, 
for \textit{the} food of your household and \textit{the} \textit{nourishment} of your maidservants.
\end{biblechapter}

\begin{biblechapter} % Proverbs 28
\verse The wicked flee, but no one pursues, 
but the righteous is bold like a lion.
\verse By the rebellion of a land, her rulers increase, 
but by a person of intelligence who knows justice, it will last.
\verse A man \textit{who is} poor and oppresses the impoverished 
\textit{is} a beating rain \textit{that leaves} no food.
\verse Those who forsake instruction will praise the wicked, 
but they who guard instruction will struggle against them.
\verse Men of evil do not understand justice, 
but seekers of Adonai understand completely.
\verse Better \textit{to be} poor and walking in one’s integrity 
than to be crooked of ways when one \textit{is} rich.
\verse He who keeps instruction \textit{is} a child of understanding, 
but the companion of gluttons will shame his father.
\verse He who augments his wealth with interest and with usury 
gathers it for him who is kind to the poor.
\verse He who turns his ear from listening to instruction, 
even his prayer \textit{is} an abomination.
\verse He who misleads the upright onto the way of evil, 
into his pits he will fall. 
But as for the blameless, they will inherit good.
\verse A man of wealth \textit{is} wise in his own eyes, 
but the intelligent poor sees through him.
\verse When the righteous triumphs, great \textit{is} the glory, 
but with \textit{the} rising of the wicked, a person will be hidden.
\verse He who conceals his transgression will not prosper, 
but he who confesses and forsakes will obtain mercy.
\verse Happy is the person who fears continuously, 
but he who is stubborn of heart, will fall into calamity.
\verse \textit{Like} a roaring lion and a charging bear 
\textit{is} a wicked ruler over a poor people.
\verse A ruler who lacks understanding is a cruel oppressor, 
\textit{but} those who hate unjust gain will have long days.
\verse A person who is burdened with \textit{the} blood of \textit{another}, 
until death he will flee; 
do not take hold of him.
\verse He who walks in integrity will be safe, 
but he who takes crooked paths will fall in one.
\verse He who tills his ground will have plenty bread, 
but he who follows fantasies will have plenty \textit{of} poverty.
\verse A man of faithfulness \textit{has} abundant blessings, 
but he who hurries to become rich will not go unpunished.
\verse \textit{Showing partiality} \textit{is} not good, 
and over a morsel of bread, a strong man will do wrong.
\verse He who hurries for wealth is \textit{a man with an evil eye}, 
but he does not know that poverty will come upon him.
\verse He who rebukes a person will afterward find \textit{more} favor 
than he who flatters \textit{with the} tongue.
\verse He who robs his father and his mother and says, “There is no crime,” 
is partner to a man who corrupts.
\verse The \textit{greedy person} will stir up strife, 
but he who trusts in Adonai will be enriched.
\verse He who trusts in his own heart is a fool, 
but he who walks in wisdom will be saved.
\verse \textit{For} he who gives to the poor, there is no lacking, 
but \textit{for} he who turns his eyes, \textit{there are} many curses.
\verse With \textit{the} rising of the wicked, humankind will be hidden, 
and with their perishing, \textit{the} righteous will multiply.
\end{biblechapter}

\begin{biblechapter} % Proverbs 29
\verse A man who is reproved, being stubborn of neck— 
suddenly he will be broken, and there will not be healing.
\verse When the righteous are numerous, the people will rejoice, 
but when the wicked are ruling, people will groan.
\verse A man who loves wisdom will make his parents glad, 
but the friend of prostitutes will squander his wealth.
\verse By justice a king gives stability \textit{to} a land, 
but a man of bribes will ruin it.
\verse A strong man who flatters his neighbor 
is spreading a net for his feet.
\verse In transgression, an evil man \textit{is} a snare, 
but \textit{the} righteous will sing and rejoice.
\verse \textit{The} righteous knows \textit{the} case of the poor, 
but \textit{the} wicked does not understand knowledge.
\verse Men of scoffing set a city aflame, 
but the wise turn away wrath.
\verse \textit{If} a wise man goes to court with a foolish man, 
then there is rankling and ridicule, but there is no relief.
\verse Men of blood hate \textit{the} blameless, 
and they seek the life of the upright.
\verse A fool gives all his breath, 
but \textit{the} wise holds back in quiet.
\verse A ruler listening to a word of falsehood, 
all his officials are wicked.
\verse The poor and a man of oppression have \textit{this} in common: 
Adonai gives \textit{light to the eyes of them both}.
\verse A king who judges with truthfulness \textit{to} the poor, 
his throne will be established forever.
\verse As for a rod and reproof, they will give wisdom, 
but a neglected child is disgraced \textit{by} his mother.
\verse With the increase of the wicked, transgression will increase, 
but the righteous will look on his downfall.
\verse Discipline your children, and they will give you rest, 
and they will give delight to your soul.
\verse When there is no prophecy, the people cast off restraint, 
but as for he who guards instruction, happiness \textit{is} his.
\verse By words, a servant is not disciplined, 
for he will understand, but there is no giving heed.
\verse You see a man who is hasty in his words: 
\textit{there is more} hope for a fool than him.
\verse He who pampers his servant from childhood, 
\textit{arrogance will be his end}.
\verse A man of anger will stir strife, 
and the owner of anger, much transgression.
\verse The pride of a person will bring him humiliation, 
and the lowly of spirit will obtain honor.
\verse Being a partner with a thief is hating one’s life; 
a curse he will hear, but not disclose.
\verse The fear of a person will lay a snare, 
but he who trusts in Adonai will be secure.
\verse Many are those who seek the \textit{favor} of a ruler, 
but from Adonai \textit{one obtains justice}.
\verse A man of injustice is an abomination to the righteous, 
but the \textit{upright} is an abomination to the wicked.
\end{biblechapter}

\begin{biblechapter} % Proverbs 30
\verseWithHeading{The Oracle of Agur} The words of Agur, son of Yakeh, the oracle. 
Thus says the man to Ithiel, to Ithiel, and Ucal:
\verse Certainly I \textit{am more} stupid than a man, 
and the understanding of humankind \textit{is} not for me.
\verse And I have not learned wisdom, 
nor will I know knowledge of the Holy One.
\verse Who has ascended \textit{to} heaven and come down? 
Who has gathered \textit{the} wind in the hollow of his hand? 
Who has wrapped water in a garment? 
Who has established all the ends of the earth? 
What \textit{is} his name and what is the name of his child? 
For \textit{surely} you know.
\verse Every word of God \textit{is} flawless; 
he \textit{is} a shield for him who takes refuge in him.
\verse Do not add to his words 
lest he rebuke you and you be found a liar.
\verse Two things I ask from you; 
do not deny me before I die:
\verse Keep falsehood and a \textit{lying word} far from me; 
do not give me poverty or wealth; 
provide me \textit{with} food \textit{only for} my need.
\verse Or else I will be satisfied and will deny \textit{him} 
and say “Who \textit{is} Adonai?” 
Or I will be poor and will steal 
and profane the name of my God.
\verse Do not slander a servant to his master, 
or else he will curse you and you will be guilty.
\verse \textit{There is} a generation \textit{that} will curse its father, 
and its mother it will not bless.
\verse \textit{There is} a generation \textit{that is} pure in its own eyes, 
but from its filthiness it will not be cleansed.
\verse \textit{There is} a generation—how lofty \textit{are} their eyes! 
And their eyelids they will lift.
\verse \textit{There is} a generation whose teeth \textit{are} swords, 
and its jawbones, knives, 
\textit{in order} to devour the poor from the earth 
and the needy from humankind.
\verse For the leech, \textit{there are} two daughters; 
“Give, give!” \textit{they cry}. 
As for three of these, they are not satisfied; 
as for four, they do not say enough.
\verse Sheol and barrenness of womb, 
the land is not satisfied \textit{with} water, 
and fire does not say “enough!”
\verse The eye \textit{that} mocks a father 
and scorns the obedience of a mother— 
the ravens of the valley will peck it out, 
and the offspring of vultures will eat it.
\verse Three of these are \textit{too} wonderful for me, 
and four, I do not understand them:
\verse the way of the eagle in the sky, 
the way of a snake on a rock, 
the way of a ship in the heart of the sea, 
and the way of a man with a young woman.
\verse This is the way of a woman committing adultery: 
she eats and wipes her mouth, 
and says “I have not done wrong.”
\verse Under three \textit{things} the earth trembles, 
and under four, it is not able to bear up:
\verse under a slave when he becomes king, 
and a fool when he is satisfied \textit{with} food;
\verse under an unloved woman when she gets married, 
and a maid when she succeeds her mistress.
\verse There \textit{are} four small things on the earth, 
and they \textit{are} \textit{exceedingly wise}:
\verse The ants \textit{are} a people \textit{who are} not strong, 
yet they prepare their food in the summer;
\verse the badgers \textit{are} a people \textit{who are} not mighty, 
yet they set their house on the rock;
\verse there is no king for the locust, 
yet it marches in rank;
\verse a lizard you \textit{can} seize with hands, 
yet it is in palaces of kings.
\verse There \textit{are} three \textit{things} that are magnificent of stride, 
and four that are magnificent \textit{when} moving:
\verse a mighty lion among the beasts, 
but he will not turn back from \textit{any face};
\verse a strutting rooster or he-goat, 
and a king \textit{whose} army \textit{is} with him.
\verse If you have been foolish by exalting yourself, 
and if you have devised evil, \textit{put your} hand to \textit{your} mouth.
\verse For pressing milk produces curd, 
and pressing the nose produces blood, 
so pressing \textit{anger} produces strife.
\end{biblechapter}

\begin{biblechapter} % Proverbs 31
\verseWithHeading{The Oracle of King Lemuel} The words of Lemuel, \textit{the} king—an oracle 
that his mother taught him:
\verse What, my son? And what, my son in my womb? 
And what, son of my vows?
\verse Do not give your strength to the women, 
and your ways to destroy kings.
\verse \textit{It is} not for the kings, O Lemuel; 
drinking wine \textit{is} not for the kings, 
nor \textit{is} strong drink for rulers.
\verse Or else he will drink and forget what has been decreed, 
and he will pervert the rights of all the \textit{afflicted}.
\verse Give strong drink to him who is perishing, 
and wine to those \textit{in bitter distress}.
\verse He will drink and forget his poverty, 
and his misery he will not remember \textit{any} more.
\verse Open your mouth for the mute, 
for the rights of all the \textit{needy}.
\verse Open your mouth, judge righteousness, 
and defend \textit{the} poor and needy.
\verseWithHeading{An Excellent Woman} A woman of excellence, who will find? 
For her worth \textit{is} far more than precious jewels.
\verse The heart of her husband trusts in her, 
and gain he will not lack.
\verse She does him good, but not harm 
all the days of her life.
\verse She seeks wool and flax, 
and she works with the diligence of \textit{her hands}.
\verse She is like the ships of a merchant; 
from far off she brings her food;
\verse And she arises while \textit{it is} still night, 
and gives food to her household, 
and tasks to her servant girls.
\verse She considers a field and buys it, 
from the fruit of \textit{her hand} she plants a vineyard.
\verse She girds her waist in strength, 
and makes her arms strong.
\verse She perceives that her merchandise \textit{is} good; 
her lamp does not go out in the night.
\verse Her hands she puts onto the distaff, 
and her palms hold a spindle.
\verse Her palm she opens to the poor, 
and her hand she reaches \textit{out} to the needy.
\verse She does not fear for her house \textit{when} it snows, 
for \textit{her entire household} is clothed \textit{in} crimson.
\verse She makes for herself coverings; 
her clothing \textit{is} fine linen and purple.
\verse Her husband is known at the gates, 
in his seat among the elders of the land.
\verse Linen garments she makes and sells, 
and she supplies sashes for the merchants.
\verse Strength and dignity are her clothing, 
and she laughs \textit{at the future}.
\verse She opens her mouth with wisdom, 
and instruction of kindness \textit{is} upon her tongue.
\verse She looks after the ways of her household, 
and the bread of idleness she will not eat.
\verse Her children rise and consider her happy, 
her husband \textit{also}, and he praises her;
\verse “Many daughters have done excellence, 
but \textit{you surpass} all of them.”
\verse Charm \textit{is} deceit and beauty \textit{is} vain; 
\textit{but} a woman who fears Adonai shall be praised.
\verse Give to her from the fruit of her hand, 
and may they praise her works in the city gates.
\end{biblechapter}

