\biblebook{Psalms}

\begin{biblechapter} % Psalms 1
\verseWithHeading{The Ways of the Righteous and the Wicked}{Blessed is the man who does not walk in the advice of the wicked; 
nor does he stand in the way of sinners; 
nor does he sit in the assembly of mockers.}%
\verse{Instead, in the law of Adonai is his delight, 
and on his law he meditates day and night.}%
\verse{And so, he is like a tree planted by streams of water 
that gives its fruit in its season; 
its leaf also does not wither. 
Therefore all that he does prospers.}%
\verse{Not so the wicked. 
Instead, they are like the chaff that the wind scatters.}%
\verse{Therefore the wicked will not stand in the judgment, 
nor sinners in the congregation of the righteous;}%
\verse{for Adonai knows the way of the righteous,\lebnote{Plural}
but the way of the wicked will perish.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 2
\verseWithHeading{The Messiah’s Reign}{Why are nations in tumult, 
and countries plotting\lebnote{Or “muttering”}in vain?\lebnote{Or “an empty thing”}}%
\verse{The kings of the earth establish themselves, 
and the rulers conspire\lebnote{Or “make themselves firm”}together 
against Adonai and his anointed:}%
\verse{“Let us tear off their bonds, 
and cast their cords from us!”}%
\verse{He who sits enthroned in the heavens laughs. 
The Lord derides them.}%
\verse{Then he speaks to them in his wrath, 
and in his fury he terrifies them:}%
\verse{“But as for me, I have set my king 
on Zion, my holy mountain.”}%
\verse{I will tell the decree; 
Adonai said to me: 
“You are my son; 
today I have begotten you.}%
\verse{Ask from me and I will make the nations your heritage, 
and your possession the ends of the earth.}%
\verse{You will break them with an iron rod. 
Like a potter’s vessel you will shatter them.”}%
\verse{So then, O kings, be wise. 
Be warned, O rulers of the earth.}%
\verse{Serve Adonai with fear, 
and rejoice with trembling.}%
\verse{Kiss the Son 
lest he be angry and you perish on the way, 
for his anger burns quickly. 
Blessed are all who take refuge in him.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 3
\verseWithHeading{A psalm of David at his fleeing from the presence of Absalom, his son.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Adonai, how many are my enemies; 
many are rising against me.}%
\verse{Many are saying about my soul, 
“There is no deliverance for him from\lebnote{Or “in”}God. \textit{Selah}\lebnote{Meaning uncertain; probably a musical or liturgical direction}}%
\verse{But\lebnote{Or “And”}you, Adonai, are a shield around me, 
my glory and the one who lifts up my head.}%
\verse{With my voice I call to Adonai 
and he answers me from \textit{his holy hill}.\lebnote{Literally “the hill of his holiness”}\textit{Selah}}%
\verse{I lay down and slept; 
I woke up because Adonai sustains me.}%
\verse{I am not afraid of the ten thousands of people 
who all around have set themselves against me.}%
\verse{Rise up, O Adonai; deliver me, O my God; 
for you strike all my enemies on the cheek. 
The teeth of the wicked you break.}%
\verse{To Adonai belongs deliverance; 
may your blessing be over your people. \textit{Selah}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 4
\verseWithHeading{For the music director; with stringed instruments. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{When I call, answer me, O God of my righteousness.\lebnote{Or “vindication”}
In trouble \textit{deliver me}.\lebnote{Literally “make wide for me”}
Be gracious to me and hear my prayer.}%
\verse{O sons of man, how long will my honor\lebnote{Or “glory”}be a disgrace? 
How long will you love vanity? 
How long will you seek lies?\lebnote{Or “falsehood”}\textit{Selah}}%
\verse{But know that Adonai has set apart\lebnote{Or “treated specially”}for himself the faithful one.\lebnote{Or “pious, godly one”}
Adonai hears when I call to him.}%
\verse{Be disturbed but do not sin. 
Commune in your heart\lebnote{That is, “think it over”}on your bed and be silent. \textit{Selah}}%
\verse{Offer correct sacrifices,\lebnote{Or “right sacrifices”}
and \textit{trust in}\lebnote{Literally “trust to”}Adonai.}%
\verse{Many are saying, “Who will show us something good?” 
Lift up over\lebnote{Or “upon”}us the light of your face, O Adonai.}%
\verse{You have put gladness in my heart 
more than \textit{when}\lebnote{Literally “the time of”}their grain and their wine abound.}%
\verse{In peace I will lie down and sleep \textit{at once},\lebnote{Literally “together”}
for you alone, O Adonai, make me dwell safely.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 5
\verseWithHeading{For the music director; with the flutes. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Hear my words, O Adonai. 
Give heed to my sighing.\lebnote{Or “murmuring”}}%
\verse{Listen to the sound of my pleading,\lebnote{Or “cry for help”}my king and my God, 
for to you I pray.}%
\verse{O Adonai, in the morning you will hear my voice. 
In the morning I will set forth\lebnote{Or “present”}my case to you and I will watch.\lebnote{Or “look expectantly”}}%
\verse{For you are not a God who desires\lebnote{Or “takes delight in”}wickedness. 
Evil cannot dwell with you.}%
\verse{The boastful do not stand before\lebnote{Or “in front of”}your eyes. 
You hate all evildoers.}%
\verse{You destroy speakers of lies.\lebnote{Hebrew “lie”}
\textit{A man of bloodshed}\lebnote{Literally “a man of bloods”}and deceit Adonai abhors.}%
\verse{\textit{But as for me},\lebnote{Literally “And I”}through the abundance of your steadfast love\lebnote{Or “covenant love”}
I will enter your house.\lebnote{Or “temple”}
I will bow down\lebnote{Or “I will worship”}toward \textit{your holy temple}\lebnote{Literally “the temple of your holiness”}in awe of you.}%
\verse{O Adonai, lead me in your righteousness because of my enemies; 
make straight before me your way.}%
\verse{For there is not anything reliable in his mouth; 
their inner part is destruction.\lebnote{Hebrew “destructions”}
Their throat is an open grave; 
with their tongue they speak deceit.\lebnote{Or “flattery; smooth words”}}%
\verse{Treat them as guilty, O God; 
let them fall because of their plans. 
Because of the abundance of their transgressions cast them out, 
for they have rebelled against you.}%
\verse{But let all who take shelter in you rejoice. 
Let them ever\lebnote{Or “perpetually”}sing for joy, 
\textit{because}\lebnote{Literally “and”}you spread protection over them; 
And let those\lebnote{Or “So that those”}who love your name exult in you.}%
\verse{For you bless the righteous. 
O Adonai, like a\lebnote{Hebrew “the”}shield you surround him with good favor.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 6
\verseWithHeading{For the music director; with stringed instruments, on the Sheminith.\lebnote{Hebrew “upon the eighth”; meaning uncertain}
A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{O Adonai, do not rebuke me in your anger, 
and do not discipline me in your wrath.}%
\verse{Be gracious to me, O Adonai, because I am feeble. 
Heal me, O Adonai, for my bones are terrified.}%
\verse{My soul is also very terrified. 
But you, O Adonai, how long?}%
\verse{Turn, O Adonai; deliver my life.\lebnote{Or “my soul”; same word as in v. 3}
Save me for the sake of your steadfast love.\lebnote{Or “loyal love”}}%
\verse{For there is no remembrance of you in death. 
In Sheol, who will give thanks to you?}%
\verse{I am weary with my groaning; 
\textit{I flood}\lebnote{Literally “I make my bed float”}my bed every night. 
With my tears \textit{I drench}\lebnote{Literally “I cause to melt”}my couch.}%
\verse{My eye wastes away because of vexation; 
it grows old because of all my oppressors.}%
\verse{Depart from me, all workers of evil, 
for Adonai has heard the sound of my weeping.}%
\verse{Adonai has heard my plea; 
Adonai has accepted my prayer.}%
\verse{All my enemies shall be ashamed and shall be very terrified. 
They shall turn back; they shall suddenly be ashamed.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 7
\verseWithHeading{A Shiggaion\lebnote{Meaning uncertain; perhaps a lamentation}of David which he sang to Adonai 
on account of Cush, a Benjaminite.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{O Adonai, my God, in you I have taken refuge. 
Save me from all who pursue me, and deliver me.}%
\verse{Or else he will tear me apart like a lion, 
ripping to pieces, with none to deliver.}%
\verse{O Adonai, my God, if I have done this, 
if there is wrong in my hands,}%
\verse{if I have repaid \textit{my ally}\lebnote{Literally “one being at peace with me”}with harm, 
or if I have plundered my enemy without cause,}%
\verse{let the enemy of my soul pursue, 
and overtake and trample my life into the ground, 
and lay my honor in the dust. \textit{Selah}}%
\verse{Rise up, O Adonai, in your anger, 
and lift yourself up against the wrath\lebnote{Or “outbursts”}of my oppressors, 
and awake for me, 
since you have commanded judgment.}%
\verse{And let the assembly of peoples surround you, 
and over it return on high.}%
\verse{Adonai judges the nations. 
O Adonai, vindicate me 
according to my innocence\lebnote{Or “righteousness”}and according to my integrity within me.}%
\verse{Let the evil of the wicked come to an end, 
but establish the righteous, 
and test the hearts and innermost being, O righteous God.}%
\verse{My shield is with God 
who saves the upright of heart.}%
\verse{God is a righteous judge, 
and a God who has indignation every day.}%
\verse{If he\lebnote{That is, a person}does not repent, he\lebnote{That is, God}will sharpen his sword. 
He has bent his bow and he has \textit{strung}\lebnote{Literally “prepared”}it.}%
\verse{And he has prepared for him weapons of death. 
He has made his arrows fiery shafts.}%
\verse{See, he\lebnote{That is, the wicked}travails with\lebnote{Or “conceives”}evil. 
He is pregnant with trouble, 
and he gives birth to deception.}%
\verse{He makes a pit and digs it out, 
then falls in the trap he has made.}%
\verse{His trouble comes back on his head, 
and his violence comes down on his skull.}%
\verse{I will thank Adonai according to his righteousness, 
and I will sing praise to the name of Adonai, the Most High.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 8
\verseWithHeading{For the music director, on the Gittith.\lebnote{Meaning uncertain}A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Adonai, our Lord, how majestic is your name in all the earth, 
who put your splendor above the heavens.}%
\verse{From the mouth of children and infants you have founded\lebnote{Or “ordained”}strength 
on account of your enemies, 
to silence the enemy and the avenger.}%
\verse{When I look at your heavens, the work of your fingers, 
the moon and the stars which you set in place—}%
\verse{what is a human being that you think of him? 
and a child of humankind that you care for him?}%
\verse{And you made him a little lower than heavenly beings,\lebnote{Or “than God”}
and with glory and with majesty you crowned him.}%
\verse{You make him over the works of your hands; 
all things you have placed under his feet:}%
\verse{sheep and cattle, all of them, 
and also the wild animals of the field,}%
\verse{the birds of the sky and the fish of the sea, 
everything that passes along the paths of seas.}%
\verse{Adonai, our Lord, 
how majestic is your name in all of the earth!}%
\end{biblechapter}

\begin{biblechapter} % Psalms 9
\verseWithHeading{For the music director, according to Muth-labben.\lebnote{Meaning uncertain}A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{I will thank Adonai with all my heart. 
I will tell of all your wonderful deeds.}%
\verse{I will be glad and rejoice in you. 
I will sing the praise of your name, O Most High.}%
\verse{When my enemies turn back, 
they will stumble and perish because of your presence.}%
\verse{For you have maintained my just cause; 
you have sat on the throne judging correctly.}%
\verse{You have rebuked the nations; 
you have destroyed the wicked. 
Their name you have blotted out 
forever and ever.}%
\verse{The enemies are destroyed in ruins forever, 
and you have uprooted their cities; 
their very memory has perished.}%
\verse{But Adonai sits enthroned forever. 
He has established his throne for judgment.}%
\verse{And he will judge the world with righteousness. 
He will judge the peoples with equity.}%
\verse{And Adonai will be a stronghold for the oppressed, 
a stronghold in times of distress.}%
\verse{And those who know your name will trust in you, 
for you do not forsake those who seek you, O Adonai.}%
\verse{Sing praises to Adonai who sits enthroned\lebnote{Or “who dwells”}in Zion. 
Tell his deeds among the peoples,}%
\verse{for he who \textit{avenges bloodshed}\lebnote{Literally “seeks bloods”}remembers them. 
He does not forget the distressed cry of the afflicted.}%
\verse{Be gracious to me, O Adonai. 
See my suffering from those who hate me, 
you who lift me up from the gates of death,}%
\verse{so that I may tell of all your praises. 
In the gates of the daughter of Zion 
let me rejoice in your salvation.}%
\verse{The nations have fallen in the pit that they made; 
their foot is caught in the net that they hid.}%
\verse{Adonai has made himself known; 
he has executed judgment. 
The wicked is snared by the work of his hands. \textit{Higgaion}\lebnote{Hebrew musical term}\textit{Selah}}%
\verse{The wicked shall turn back to Sheol, 
all the nations forgetting God,}%
\verse{for the needy shall not always be forgotten; 
the hope of the poor shall never perish.}%
\verse{Rise up, O Adonai, let not humans\lebnote{Hebrew “human”}prevail; 
let the nations be judged before you.}%
\verse{O Adonai, put them in fear. 
Let the nations know they are merely human. \textit{Selah}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 10
\verseWithHeading{A Prayer for God to Throw down the Wicked}{Why, O Adonai, do you stand far off? 
Why do you hide\lebnote{Or “shut your eyes”}during times of distress?}%
\verse{In arrogance the wicked persecutes\lebnote{Or “pursues”}the poor. 
Let them be caught in the schemes that they devised,}%
\verse{for the wicked boasts about the desire of his \textit{heart},\lebnote{Literally “soul”}
and the one greedy for gain curses and treats Adonai with contempt.}%
\verse{\textit{With bald-faced pride}\lebnote{Literally “According to the height of his nose”}the wicked will not seek God. 
There is no God in any of his thoughts.}%
\verse{His ways\lebnote{According to the reading tradition (\textit{Qere})}endure at all times.\lebnote{Hebrew “time”}
Your judgments are aloof from him. 
As for all his enemies, he scoffs at them.}%
\verse{He says in his heart, “I shall not be moved 
\textit{throughout all generations},\lebnote{Literally “for a generation and a generation”}during which I will have no trouble.”}%
\verse{His mouth is filled with cursing, 
with deceits and oppression; 
under his tongue are trouble and evil.}%
\verse{He sits in ambush in villages; 
in the hiding places he kills the innocent. 
His eyes \textit{lurk}\lebnote{Literally “lie hid”}for the helpless.}%
\verse{He lies in ambush \textit{secretly,}\lebnote{Literally “in the secret places”}like a lion in a thicket. 
He lies in ambush to seize the poor; 
he seizes the poor by catching him in his net.}%
\verse{He is crushed;\lebnote{Or “and he the wicked stoops”}he is bowed down;\lebnote{Or “he the wicked crouches”}
so\lebnote{Or “and”}the helpless host\lebnote{According to the reading tradition (\textit{Qere})}falls by his \textit{might}.\lebnote{Literally “mighty ones”}}%
\verse{He says in his heart, “God has forgotten. 
He has hidden his face. 
He never sees.”}%
\verse{Rise up, O Adonai; 
O God, lift up your hand. 
Do not forget the afflicted.}%
\verse{Why does the wicked treat God with contempt? 
He says in his heart, “You will not call me to account.”}%
\verse{But you have seen; indeed you have noted trouble and grief 
to take it into your hand. 
The helpless abandons himself upon you; 
you have been the helper for the orphan.}%
\verse{Break the arm of the wicked, 
and as for the evil man— 
seek out his wickedness until you find none.}%
\verse{Adonai is king forever and ever; 
the nations have perished from his land.}%
\verse{The longing of the afflicted you have heard, O Adonai. 
You will make their heart secure. You will \textit{listen attentively}\lebnote{Literally “cause your ear to hear”}}%
\verse{to render judgment for the fatherless and the oppressed 
so that a mere mortal from the earth will no longer cause terror.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 11
\verseWithHeading{For the music director. Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{In Adonai I have taken refuge; 
how can you say to my soul, 
“Flee to your mountain like a bird”?}%
\verse{For look: The wicked \textit{string the bow}.\lebnote{Literally “tread the bow,” that is, placing the foot on the bow to bend and string it}
They have fitted their arrow to the string, 
to shoot from the darkness 
at the upright of heart.}%
\verse{When the foundations are destroyed 
what can the righteous do?\lebnote{Or “what has the righteous done”}}%
\verse{Adonai is in his holy temple; 
Adonai is in the heavens on his throne. 
His eyelids see; 
they test the children of humankind.}%
\verse{Adonai tests the righteous, 
but the wicked and the lover of violence 
his soul hates.}%
\verse{He will rain coals\lebnote{According to Symmachus (compare also \textit{Targum Psalms}) a form slightly different from the MT’s “snares”}on the wicked. 
Burning sulfur and scorching wind 
will be the portion of their cup.}%
\verse{For Adonai is righteous; 
he loves righteous deeds. 
The upright shall see his face.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 12
\verseWithHeading{For the music director; on the Sheminith.\lebnote{Meaning uncertain}A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Save, O Adonai, for the pious have ceased to be; 
for the faithful have vanished 
from among the children of humankind.}%
\verse{They speak falseness \textit{to each other}.\lebnote{Literally “each with his companion”}
With flattering lips, 
with \textit{a double heart}\lebnote{Literally “a heart and a heart”}they speak.}%
\verse{May Adonai cut off all flattering lips, 
the tongue speaking great boasts—}%
\verse{those who say, “With our tongue we will prevail. 
Our lips are on our side. 
Who is master over us?”}%
\verse{“Because of the oppression of the afflicted, 
because of the groaning of the poor, 
now I will rise up,” Adonai says. 
“I shall put them in the safety for which they\lebnote{Hebrew “he”}long.”}%
\verse{The words of Adonai are pure words 
like silver refined in the crucible on the ground, 
refined seven times.}%
\verse{You, O Adonai, will protect them. 
You will preserve him\lebnote{Or “us”}
from this generation always.}%
\verse{The wicked prowl about 
when vileness is exalted among the children of humankind.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 13
\verseWithHeading{For the music director. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{How long, O Adonai? Will you forget me forever? 
How long will you hide your face from me?}%
\verse{How long must I take counsel\lebnote{Hebrew pl.; with a slight modification of the Hebrew this reads “set pains”}in my soul, 
and sorrow in my heart all the day? 
How long will my enemy be exalted over me?}%
\verse{Consider\lebnote{Literally “Look”}and answer me, O Adonai my God. 
Give light to my eyes 
lest I sleep the sleep of death,}%
\verse{and lest my enemy should say, “I have overcome him,” 
lest my enemies rejoice because I am shaken.}%
\verse{But as for me, I have trusted in your steadfast love.\lebnote{Or, “loyal love”}
My heart will rejoice in your deliverance.}%
\verse{\lebnote{In the Hebrew Bible, the previous verse continues}I will sing to Adonai 
because he has dealt bountifully with me.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 14
\verseWithHeading{For the music director. Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{The fool says in his heart, “There is no God.” 
They are corrupt. 
They do abominable deeds. 
There is none who does good.}%
\verse{Adonai looks down from heaven upon the children of humankind 
to see whether there is one who has insight, 
one who cares about God.}%
\verse{All have gone astray; 
they are altogether corrupt. 
There is not one who does good; 
there is not even one.}%
\verse{All who do evil—do they not know, 
they who eat my people as though they were eating bread? 
They do not call on Adonai.}%
\verse{There \textit{they are very fearful}\lebnote{Literally “they feared a fear”}
because God is with the generation of the righteous.}%
\verse{You would put to shame the plan of the poor, 
because\lebnote{Or “but”}Adonai is his refuge.}%
\verse{\textit{Oh that from Zion}\lebnote{Literally “Who will give that from Zion”}would come salvation for Israel! 
When Adonai returns the fortunes of his people, 
Jacob will rejoice; Israel will be happy.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 15
\verseWithHeading{A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{O Adonai, who may reside in your tent? 
Who may dwell on your holy mountain?}%
\verse{He who walks blamelessly and does what is right. 
He who speaks honestly in his heart.}%
\verse{He does not slander with his tongue. 
He does not harm his friend, 
nor \textit{bring up}\lebnote{Literally “lift up”}a disgrace against his neighbor.}%
\verse{In his eyes a rejected one is contemptible, 
but he honors those who respect Adonai. 
He takes an oath to his own injury and does not retract it.}%
\verse{He does not lend his money at interest, 
and does not take a bribe against the innocent. 
He who does these things 
will never be shaken.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 16
\verseWithHeading{A miktam\lebnote{Meaning uncertain}of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Protect me, O God, 
for I take refuge in you.}%
\verse{Oh my soul you have said\lebnote{Reading a feminine singular verb; some Hebrew manuscripts and translations read “I said”}to Adonai, “You are my Lord. 
I have no good apart from you.”}%
\verse{With regard to the saints who are in the land, 
they are the noble ones in whom is all my desire.}%
\verse{Those who hurry after\lebnote{Or “they who acquire”}another god increase their sorrows. 
I will not pour out their drink offerings of blood, 
nor take up their names on my lips.}%
\verse{Adonai is the portion which is my share and my cup. 
You hold my lot.}%
\verse{The measuring lines have fallen for me in pleasant places. 
Yes, my inheritance is delightful for me.}%
\verse{I will bless Adonai who advises me; 
yes, at night my \textit{innermost being}\lebnote{Literally “kidneys”}instructs me.}%
\verse{I have set Adonai before me always. 
Because he is at my right hand I will not be shaken.}%
\verse{Therefore my heart is glad and my glory rejoices. 
Yes, my body will dwell in safety,}%
\verse{for you will not abandon my soul to Sheol; 
you will not give your faithful one to see the grave.}%
\verse{You will make known to me the path of life. 
In your presence is fullness of joy. 
At your right hand are pleasures forever.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 17
\verseWithHeading{A prayer of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{O Adonai, hear a just cause. 
Hear my cry; heed my prayer 
I make without deceitful lips.}%
\verse{Let my vindication come forth from you; 
let your eyes see fairness.}%
\verse{You have tried my heart; 
you have examined me by night; 
you have tested me; you found nothing. 
I have decided that my mouth will not transgress.}%
\verse{As for the works of humankind, 
by the word of your lips, 
I have kept from the ways of the violent.}%
\verse{I have held my steps in your path 
My feet will not slip.}%
\verse{As for me, I have called on you 
because you will answer me, O God. 
Incline your ear to me. 
Hear my words.\lebnote{Hebrew “word”}}%
\verse{Show wondrously your acts of loyal love, 
O Savior of those who take refuge 
at your right hand\lebnote{Or “saving by your right hand those”}
from those who rise up against them.}%
\verse{Keep me as \textit{the apple of your eye}.\lebnote{Literally “the little man of the daughter of your eye”}
Hide me in the shadow of your wings}%
\verse{from the \textit{presence}\lebnote{Literally “face”}of the wicked who destroy me, 
those enemies against my life, 
they that surround me.}%
\verse{They have shut off their calloused\lebnote{Or “fat”}heart; 
with their mouth they speak arrogantly.}%
\verse{Now they surround us at our every step. 
They \textit{intend}\lebnote{Literally “set their eyes”}
\textit{to pin}\lebnote{Literally “to stretch”}me to the ground.}%
\verse{\textit{He is like}\lebnote{Literally “His image”}a lion; he longs to tear apart, 
and like a strong lion crouching in hiding places.}%
\verse{Rise up, O Adonai, confront \textit{him}.\lebnote{Literally “his face”}
Make him bow down. 
Rescue with your sword my life from the wicked,}%
\verse{from men by your hand, O Adonai, from men of this world. 
Their share is in \textit{this life},\lebnote{Literally “the life”}and you fill their stomach with your treasure. 
They are satisfied with children. 
They bequeath their excess to their children.}%
\verse{By contrast, I in righteousness shall see your face. 
Upon awakening I will be satisfied seeing your form.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 18
\verseWithHeading{For the music director. Of David, the servant of Adonai, 
who spoke to Adonai the words of this song 
\textit{when}\lebnote{Literally “on the day”}Adonai delivered him from the hand of all his enemies, 
and from the hand of Saul.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{And he said: 
“I love you, O Adonai, my strength.”}%
\verse{Adonai is my rock and my fortress and my deliverer, 
my God, my rock in whom I have taken refuge, 
my shield and the horn of my deliverance, my stronghold.}%
\verse{I called upon Adonai, who is worthy to be praised, 
and I was saved from my enemies.}%
\verse{The ropes of death encircled me, 
and streams of ruin overwhelmed me.}%
\verse{The ropes of Sheol surrounded me; 
the snares of death confronted me.}%
\verse{In my trouble I called on Adonai, 
and to my God I cried for help. 
He heard my voice from his temple, 
and my cry for help 
came up before his presence into his ears.}%
\verse{Then the earth shook and quaked, 
and the foundations of the mountains trembled, 
and they staggered because he was angry.}%
\verse{Smoke went forth from his nose, 
and fire from his mouth consumed. 
Burning coals blazed from him.}%
\verse{So he bowed the heavens and came down 
with a thick cloud under his feet.}%
\verse{And he mounted a cherub and flew, 
and he swooped down on wings of wind.}%
\verse{He made darkness his hiding place; 
all about him his covering 
was a darkness of waters, \textit{thick clouds}.\lebnote{Literally “clouds of clouds”}}%
\verse{From the brightness before him 
his clouds passed over 
with hail and coals of fire.}%
\verse{And Adonai thundered from the heavens, 
and the Most High uttered his voice 
with hail and coals of fire.}%
\verse{And he shot his arrows and scattered them,\lebnote{That is, the enemy}
and many lightning bolts and routed them.}%
\verse{Then the channels of the sea became visible, 
and the foundations of the world were uncovered 
by your rebuke, O Adonai, 
by the \textit{blast}\lebnote{Or “breath”}of the wind of your nose.}%
\verse{He reached\lebnote{Or “sent”}from on high; he seized me. 
He drew me from many waters.}%
\verse{He delivered me from my strong enemy, 
and from those who hated me 
because they were stronger than I.}%
\verse{They confronted me on the day of disaster, 
but Adonai was my support.}%
\verse{So he brought me out to a spacious place. 
He delivered me because he delighted in me.}%
\verse{Adonai has rewarded me according to my righteousness; 
According to the cleanness of my hands 
he has repaid me.}%
\verse{Because I have kept the ways of Adonai, 
and have not acted wickedly against my God,}%
\verse{because all his ordinances are before me, 
and his statutes I have not removed from me,}%
\verse{and I was blameless before him, 
and I kept myself from my guilt.}%
\verse{Therefore Adonai has repaid me according to my righteousness, 
according to the cleanness of my hands in his sight.}%
\verse{To the loyal you show yourself loyal. 
To the blameless you show yourself blameless.}%
\verse{To the pure you show yourself pure. 
but to the wicked you show yourself shrewd.}%
\verse{For you deliver humble people, 
but haughty eyes you humiliate.}%
\verse{For you light my lamp; 
Adonai, my God, lights up my darkness.}%
\verse{For with you I can charge a troop, 
and with my God I can scale a wall.}%
\verse{This\lebnote{Hebrew “The”}God—his way is blameless. 
The word of Adonai is flawless.\lebnote{Or “refined”}
He is a shield for all those who take refuge in him.}%
\verse{For who is God apart from Adonai, 
and who is a rock except our God?}%
\verse{The God who girded me with strength, 
and made safe\lebnote{Or “blameless”}my way,}%
\verse{he made my feet like the deer’s, 
and on my high places he makes me stand.}%
\verse{He trains my hands for war, 
so that my arms can bend a bow of bronze.}%
\verse{You also gave me the shield of your salvation, 
and your right hand supported me, 
and your humility made me great.}%
\verse{You enlarged my step under me, 
and my ankles did not slip.}%
\verse{I chased my enemies and overtook them; 
I did not turn back until they were finished off.}%
\verse{I struck them and they were not able to stand up; 
they fell under my feet.}%
\verse{And you girded me with strength for the battle; 
those rising against me you made to kneel under me.}%
\verse{And you \textit{made my enemies turn their backs},\lebnote{Literally “turned to me my enemies’ neck”}
then I destroyed those who hated me.}%
\verse{They cried out but there was no rescuer, 
even to Adonai, but he did not answer them.}%
\verse{And I ground them like dust \textit{in the wind},\lebnote{Literally “on the face of the wind”}
I emptied them\lebnote{Many Hebrew manuscripts, the LXX, Syriac, and 2 Sam 22:43 read “I crushed them”}out like mud of the streets.}%
\verse{You rescued me from strife with people; 
you set me as head (over) nations. 
People I did not know served me;}%
\verse{as soon as \textit{they heard},\lebnote{Literally “hearing of ear”}they obeyed me. 
The children of a foreign land feigned obedience before me.}%
\verse{The children of a foreign land \textit{lost heart},\lebnote{Literally “sunk down”}
and they came trembling from their strongholds.}%
\verse{Adonai lives, and blessed be my rock, 
and may the God of my salvation be exalted—}%
\verse{the God who executed vengeance for me, 
and subdued peoples under me,}%
\verse{who delivered me from my enemies. 
Indeed, you exalted me higher than those who rose against me. 
You delivered me from the man of violence.}%
\verse{Therefore I will give thanks to you among the nations, O Adonai, 
and I will sing the praise of your name:}%
\verse{the one who makes great victories for his king, 
and who shows loyal love for his anointed, 
to David and to his descendants forever.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 19
\verseWithHeading{For the music director. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{The heavens are telling the glory of God, 
and the firmament proclaims the work of his hands.}%
\verse{\textit{Every day}\lebnote{Literally “day to day”}they pour forth speech, 
and \textit{every night}\lebnote{Literally “night to night”}they tell knowledge.}%
\verse{There is no speech and there are no words; 
their sound is inaudible.}%
\verse{Yet in all the world their line\lebnote{LXX “line”}goes out, 
and their words to the end of the world. 
In them\lebnote{That is, the heavens}he has pitched a tent for the sun,}%
\verse{and it is like a bridegroom 
who comes out of his bridal chamber. 
It is glad like a strong man 
to run its course.}%
\verse{Its rising is from one\lebnote{Hebrew “the”}end of the heavens 
and its circuit to the other end\lebnote{Hebrew “ends”}of them; 
and nothing is hidden from its heat.}%
\verse{The law of Adonai is perfect, reviving life.\lebnote{Or “soul”}
The testimony of Adonai is firm, making wise the simple.}%
\verse{The precepts of Adonai are right, making the heart rejoice. 
The command of Adonai is pure, enlightening the eyes.}%
\verse{The fear of Adonai is pure, enduring forever. 
The ordinances of Adonai are true; they are righteous altogether,}%
\verse{more desirable than gold, even much fine gold; 
and sweeter than honey, even honey in the comb.}%
\verse{Moreover, your servant is warned\lebnote{Or “illumined”}by them; 
in keeping them is great reward.}%
\verse{Who can perceive his errors? 
Acquit me from hidden faults.}%
\verse{Also, keep back your servant from arrogant sins;\lebnote{Usage elsewhere might suggest “arrogant people”}
let them not rule over me. 
Then I shall be blameless, 
and I shall be innocent of great transgression.}%
\verse{Let the words of my mouth and the meditation of my heart 
be acceptable in your presence, 
O Adonai, my rock and my redeemer.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 20
\verseWithHeading{For the music director. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{May Adonai answer you in the day of trouble. 
May the name of Jacob’s God protect\lebnote{The sense is that of making something inaccessibly high, like a fortress}you.}%
\verse{May he send you help\lebnote{Hebrew “your help”}from the sanctuary, 
and from Zion may he sustain you.}%
\verse{May he remember all your offerings, 
and your burnt offering may he \textit{regard favorably}.\lebnote{Literally “find to be fat”}}%
\verse{May he give to you \textit{your heart’s desire},\lebnote{Literally “according to your heart”}
and your every plan may he fulfill. \textit{Selah}}%
\verse{May we shout for joy over your victory, 
and in the name of our God may we set up banners. 
May Adonai fulfill all your requests.}%
\verse{Now I know that Adonai will help his anointed.\lebnote{Hebrew “messiah”}
He will answer him from his holy heaven 
with the victorious power of his right hand.}%
\verse{\textit{Some boast in chariots and others in horses},\lebnote{Literally “These in the chariots and those in the horses”}
but we boast in the name of Adonai, our God.}%
\verse{They will collapse and fall, 
and we will rise and stand firm.}%
\verse{Rescue,\lebnote{Or “Grant victory”}O Adonai. 
Let the king answer us when we call.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 21
\verseWithHeading{For the music director. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{O Adonai, the king will rejoice in your strength, 
and how greatly he will delight in your help.\lebnote{Or “victory”}}%
\verse{You have given him the desire of his heart, 
and have not withheld the request of his lips. \textit{Selah}}%
\verse{For you meet him with blessings of good things. 
You set on his head a crown of fine gold.}%
\verse{He asked life from you; you gave it to him— 
length of days forever and ever.}%
\verse{His honor is great because of your help.\lebnote{Or “victory”}
Splendor and majesty you have bestowed upon him.}%
\verse{For you set on him blessings forever. 
You make him glad with the joy of your presence.}%
\verse{Because the king trusts Adonai, 
and through the steadfast love\lebnote{Or “loyal love”}of the Most High 
he shall not be moved.}%
\verse{Your hand will find all your enemies; 
your right hand will find those who hate you.}%
\verse{You will make them like your fiery furnace 
at the time of your appearance. 
Adonai will swallow them in his wrath 
and fire will consume them.}%
\verse{You will destroy their \textit{offspring}\lebnote{Literally “fruit”}from the earth, 
and their \textit{descendants}\lebnote{Literally “seed”}from among the children of humankind.}%
\verse{Though they have \textit{plotted}\lebnote{Literally “spread out”}evil against you, 
though they have planned a scheme, they will not prevail.}%
\verse{For you will \textit{turn them to flight};\lebnote{Literally “make them a shoulder”}
you will aim arrows on your bowstrings at their faces.}%
\verse{Be exalted, O Adonai, in your strength, 
and we sing and praise your power.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 22
\verseWithHeading{For the music director; according to The Doe of the Dawn. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{My God, my God why have you forsaken me? 
Why are you far from helping me, far from the words of my groaning?}%
\verse{O my God, I call by day and you do not answer, 
and by night but I have no rest.\lebnote{Or “I am not silent”}}%
\verse{Yet you are holy, 
enthroned on the praises of Israel.}%
\verse{Our ancestors\lebnote{Or “fathers”}trusted you; 
they trusted and you delivered them.}%
\verse{They cried to you and were saved; 
they trusted you and were not ashamed.}%
\verse{But I am a worm and not a man, 
scorned by humankind and despised by people.}%
\verse{All who see me mock me. 
They open wide their lips; 
they shake the head, saying:}%
\verse{“\textit{He trusts Adonai}.\lebnote{Literally “Commit to Adonai”}Let him rescue him. 
Let him deliver him because he delights in him.”}%
\verse{Yet you took me from the belly; 
you made me trust while on my mother’s breasts.}%
\verse{On you I was cast from the womb. 
From my mother’s belly you have been my God.}%
\verse{Do not be far from me 
because trouble is near; 
because there is no helper.}%
\verse{Many bulls have encircled me; 
mighty bulls of Bashan have surrounded me.}%
\verse{They open their mouth against me 
like a lion tearing and roaring.}%
\verse{I am poured out like water, 
and all my bones are out of joint. 
My heart is like wax; 
it is melted \textit{within me}.\lebnote{Literally “in the midst of my intestines”}}%
\verse{My strength is dry like a potsherd, 
and my tongue is sticking to my jaws; 
and you have placed me in the dust of death.}%
\verse{Because dogs have surrounded me; 
a gang of evildoers has encircled me. 
Like the lion\lebnote{Or “they gouge out” or “they bind.” Not the same word for “lion” found in vv. 14(13), 22(21). The Hebrew text presents such problems that any interpretation must be lightly held.}they are at my hands and my feet.}%
\verse{I can count all my bones; 
they gaze, they look at me.}%
\verse{They divide my garments among them, 
and for my clothing they cast lots.\lebnote{Hebrew “lot”}}%
\verse{But you, O Adonai, do not remain distant. 
O my help, hasten to help me.}%
\verse{Rescue my life from the sword, 
my only life from the power of the dogs.\lebnote{Hebrew “dog”}}%
\verse{Save me from the mouth of the lion, 
and from the horns of the wild oxen answer me.}%
\verse{I will tell your name to my brothers; 
inside the assembly I will praise you.}%
\verse{You who revere Adonai, praise him! 
Glorify him, all you seed of Jacob, 
and be in awe of him, all you seed of Israel,}%
\verse{because he has not despised nor abhorred the affliction of the afflicted, 
and has not hid his face from him; 
but he listened to him when he cried for help.}%
\verse{From you is my praise. 
In the great assembly, 
I will pay my vows before those who revere him.}%
\verse{The afflicted will eat and will be satisfied. 
Those who seek him will praise Adonai. 
May your\lebnote{Plural}heart live forever.}%
\verse{All the ends of the earth 
will remember and turn to Adonai. 
All the families of the nations 
will worship before you.}%
\verse{Because the kingship belongs to Adonai, 
and he rules over the nations.}%
\verse{All the \textit{healthy ones}\lebnote{Or “prosperous”; Literally “fat ones”}of the earth will eat and worship. 
Before him all of those descending into the dust will kneel, 
even he who cannot keep his soul alive.}%
\verse{Descendants\lebnote{Hebrew “Descendant”}will serve him. 
Regarding the Lord, it will be told to the next generation.}%
\verse{They will come and tell his saving\lebnote{Or “righteousness”}deeds 
to a people yet to be born, that he has done it.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 23
\verseWithHeading{A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Adonai is my shepherd; 
I will not lack for anything.}%
\verse{In grassy pastures he makes me lie down; 
by quiet waters he leads me.}%
\verse{He restores my life.\lebnote{Or “my soul”}
He leads me in \textit{correct paths}\lebnote{Literally “paths of righteousness”}
for the sake of his name.}%
\verse{Even when I walk in a dark valley, I fear no evil 
because you are with me. 
Your rod and your staff, they comfort me.}%
\verse{You prepare before me a table 
in the presence of my oppressors. 
You anoint my head with oil; 
my cup is overflowing.}%
\verse{Surely goodness and loyal love will pursue me 
all the days of my life, 
and I will stay\lebnote{Hebrew “return”}in the house\lebnote{Or “temple”}of Adonai 
\textit{for a very long time}.\lebnote{Literally “for length of days”}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 24
\verseWithHeading{A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{The earth is Adonai’s, with its fullness, 
the world and those who live in it,}%
\verse{because he has founded it on the seas, 
and has established it on the rivers.}%
\verse{Who may ascend the mountain of Adonai? 
And who may stand in his holy place?}%
\verse{He who is innocent of hands and pure of heart, 
who does not lift up his soul to falseness, 
and does not swear deceitfully.}%
\verse{He will receive blessing from Adonai, 
and justice\lebnote{Or “vindication”}from the God of his salvation.}%
\verse{Such is the sort of those who seek\lebnote{According to the reading tradition (\textit{Qere})}him, 
those who seek your face, even Jacob. \textit{Selah}}%
\verse{Lift up your heads, O gates, 
and rise up, O ancient doorways, 
that the king of glory\lebnote{Literally “the king of the glory”}may enter.}%
\verse{Who is the king of glory?\lebnote{Literally “the king of the glory”}
Adonai, strong and mighty; 
Adonai, mighty in war!}%
\verse{Lift up your heads, O gates, 
and lift up, O ancient doorways, 
that the king of glory\lebnote{Literally “the king of the glory”}may enter.}%
\verse{Who is the king of glory?\lebnote{Literally “the king of the glory”}
Adonai of hosts, 
He is the king of glory!\lebnote{Literally “the king of the glory”}\textit{Selah}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 25
\verseWithHeading{Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{To you, O Adonai, I lift up my soul.}%
\verse{O my God, I trust you; let me not be put to shame. 
Do not let my enemies exult over me.}%
\verse{Indeed, none who wait for you should be put to shame. 
Those who betray without cause should be put to shame.}%
\verse{Make me know your ways, O Adonai. 
Teach me your paths.}%
\verse{Cause me to walk in your truth and teach me, 
because you are the God of my salvation. 
I await you \textit{all day long}.\lebnote{Literally “all of the day”}}%
\verse{Remember your compassion, O Adonai, 
and your acts of loyal love, 
because they are from of old.}%
\verse{Do not remember 
the sins of my youth or my transgressions. 
According to your loyal love remember me if you will, 
for the sake of your goodness, O Adonai.}%
\verse{Good and right is Adonai; 
therefore he instructs sinners in the way.}%
\verse{He causes the humble to walk in justice, 
and teaches the humble his way.}%
\verse{All the paths of Adonai are loyal love and faithfulness 
for those who keep his covenant and statutes.\lebnote{Or “covenant statutes”}}%
\verse{Also, for the sake of your name, O Adonai, 
forgive my sin, because it is great.}%
\verse{Who is the man fearing Adonai? 
He will instruct him in the way he should choose.}%
\verse{His soul will lodge in prosperity, 
and his offspring will possess the land.}%
\verse{Intimate fellowship with Adonai is for those who fear him, 
and he makes known his covenant to them.}%
\verse{My eyes are continually toward Adonai, 
because he will take my feet from the net.}%
\verse{Turn to me and have mercy on me 
because I am lonely and afflicted.}%
\verse{Remove the troubles of my heart; 
bring me out from my distresses.}%
\verse{Consider my affliction and trouble, 
and forgive\lebnote{Or “take away”}all my sins.}%
\verse{Consider that my enemies are many, 
and they hate me with violent hatred.}%
\verse{Protect my life and deliver me. 
Let me not be put to shame, because I take shelter in you.}%
\verse{Let integrity and uprightness preserve me, 
because I wait for you.}%
\verse{O God, redeem Israel 
from all its troubles.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 26
\verseWithHeading{Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Judge me, O Adonai, because I have walked in my integrity, 
and I have trusted Adonai and not wavered.}%
\verse{Prove me, O Adonai, and test me. 
Try my \textit{innermost being}\lebnote{Literally “kidneys,” the locus of intense thoughts and emotions in biblical Hebrew}and my \textit{mind}.\lebnote{Literally “heart”}}%
\verse{Because your loyal love is before my eyes, 
and I walk about in your faithfulness.}%
\verse{I do not sit with deceitful people, 
nor will I go about with hypocrites.}%
\verse{I hate the crowd of evildoers, 
and with the wicked I will not sit.}%
\verse{I will wash my hands in innocence, 
and I will walk about your altar, O Adonai,}%
\verse{to declare with a voice of thanksgiving, 
and tell of all your wondrous deeds.}%
\verse{O Adonai, I love the dwelling of your house,\lebnote{Or “temple”}
and \textit{the place where your glory abides}.\lebnote{Literally “the place of the dwelling place of your glory”}}%
\verse{Do not destroy me\lebnote{Or “gather my soul for judgment”}with the sinners, 
nor my life with men of bloodshed,}%
\verse{in whose hands is an evil plan, 
and whose right hand is full of bribes.\lebnote{Hebrew “bribe”}}%
\verse{But as for me, I will walk in my integrity. 
Redeem me and be gracious to me.}%
\verse{My foot stands on level ground. 
In assemblies I will bless Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 27
\verseWithHeading{Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Adonai is my light and my salvation; 
whom shall I fear? 
Adonai is the refuge of my life; 
of whom shall I be afraid?}%
\verse{When evildoers drew near against me to eat my flesh— 
my adversaries and my enemies who drew near to me— 
they themselves stumbled and fell.}%
\verse{Though an army encamp against me, 
my heart will not fear. 
Though war arise against me, 
even in this I will remain confident.}%
\verse{One thing I have asked from Adonai; 
it I will seek: 
that I may dwell in the house\lebnote{Or “temple”}of Adonai all the days of my life, 
to behold the beauty of Adonai, 
and to consider\lebnote{Or “inquire in”}his temple.}%
\verse{Because he will hide me in his shelter in the day of trouble. 
He will conceal me in the hiding place of his tent. 
He will set me high upon a rock.}%
\verse{And now my head will be high over my enemies round about me. 
And I will offer in his tent sacrifices with shouts of joy. 
I will sing, yes, I will sing praises to Adonai.}%
\verse{Hear, O Adonai, my voice when I call, 
and be gracious to me and answer me.}%
\verse{On your behalf my heart says, “Seek my face.” 
Your face, O Adonai, I do seek.}%
\verse{Do not hide your face from me; 
do not turn your servant away in anger. 
You have been my help; do not abandon nor forsake me, 
O God of my salvation.}%
\verse{If my father or my mother forsake me, 
then Adonai will receive me.}%
\verse{Teach me, O Adonai, your way, 
and lead me on a level path because of my enemies.}%
\verse{Do not give me over to the desire of my enemies, 
because false witnesses have arisen against me, 
and each breathing out violence.\lebnote{Or “and each a witness of violence”}}%
\verse{Surely\lebnote{Hebrew text is uncertain}I believe that I will see the goodness of Adonai 
in the land of the living.}%
\verse{Wait for Adonai. 
Be strong and let your heart show strength, 
and wait for Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 28
\verseWithHeading{A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{To you, O Adonai, I call. 
O my rock, do not be deaf to me. 
Or else, if you are silent to me, 
then I will become like those descending to the pit.}%
\verse{Hear the voice of my supplications 
when I cry to you for help, 
when I lift up my hands 
toward your holy inner sanctuary.}%
\verse{Do not drag me away with the wicked 
or with the workers of evil, 
who speak of peace with their neighbors, 
while evil is in their heart.}%
\verse{Give to them according to their work, 
even according to the evil of their deeds. 
Give to them according to the work of their hands; 
repay them their rightful due.}%
\verse{Because they do not regard the works of Adonai, 
nor the work of his hands, 
he will destroy them 
and not rebuild them.}%
\verse{Blessed is Adonai, 
because he has heard the voice of my supplications.}%
\verse{Adonai is my strength and my shield. 
My heart trusts him and I am helped. 
So my heart rejoices, 
and with my song I will give thanks to him.}%
\verse{Adonai is their strength, 
and he is the refuge for the salvation of his anointed one.}%
\verse{Save your people 
and bless your heritage. 
Shepherd them also and carry them always.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 29
\verseWithHeading{A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Ascribe to Adonai, O sons of God,\lebnote{Or “heavenly beings”}
ascribe to Adonai glory and strength.}%
\verse{Ascribe to Adonai the glory \textit{due}\lebnote{Literally “of”}his name. 
Worship Adonai in holy array.}%
\verse{The voice of Adonai is over the waters; 
the God of glory thunders. 
Adonai is over many waters.}%
\verse{The voice of Adonai is powerful. 
The voice of Adonai is majestic.}%
\verse{The voice of Adonai breaks the cedars. 
Adonai shatters even the cedars of Lebanon.}%
\verse{And he makes them skip like a calf; 
Lebanon and Sirion like a young wild ox.}%
\verse{The voice of Adonai flashes\lebnote{Or “divides”}flames of fire.}%
\verse{The voice of Adonai shakes the wilderness. 
Adonai shakes the wilderness of Kadesh.}%
\verse{The voice of Adonai causes deer\lebnote{A slightly different Hebrew word here yields, “makes the oaks tremble”}to give birth 
and strips the forests bare. 
And at his temple all in it say, “Glory!”}%
\verse{Adonai sits enthroned at the flood, 
and Adonai sits as king forever.}%
\verse{May Adonai give strength to his people. 
May Adonai bless his people with peace.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 30
\verseWithHeading{A psalm; a song for the dedication of the house.\lebnote{Or “temple”}Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{I will exalt you, O Adonai, 
because you have drawn me up, 
and have not made my enemies rejoice over me.}%
\verse{O Adonai, my God, 
I cried for help to you, 
and you healed me.}%
\verse{O Adonai, you have brought my life up from Sheol. 
You preserved me alive 
from among those going down \lebnote{According to the \textit{Kethib}; the reading tradition (\textit{Qere}) reads “so I would not go down”}to the pit.}%
\verse{Sing praises to Adonai, you his faithful ones, 
and give thanks \textit{to his holy fame}.\lebnote{Literally “to the fame of his holiness”}}%
\verse{For there is a moment in his anger; 
there is a lifetime in his favor. 
Weeping lodges for the evening, 
but in the morning comes rejoicing.}%
\verse{But as for me, I had said in my prosperity, 
“I shall not be moved ever.”}%
\verse{O Adonai, by your favor 
you caused my strong mountain to stand. 
You hid your face. I was bewildered.}%
\verse{To you, O Adonai, I called, 
and to the Lord I pleaded for grace saying,}%
\verse{“What gain is there in my \textit{death},\lebnote{Literally “blood”}
in my going down into the pit? 
Will the dust praise you? 
Will it tell of your faithfulness?}%
\verse{“O Adonai, hear and be gracious to me. 
O Adonai, be my helper.”}%
\verse{You have turned my wailing into my dancing. 
You have removed my sackcloth 
and clothed me with joy}%
\verse{so that \textit{I}\lebnote{Literally “my glory”}may sing praises to you 
and not be quiet. 
O Adonai, my God, 
I will give thanks to you forever.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 31
\verseWithHeading{For the music director. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{In you, O Adonai, I have taken refuge. 
Let me not be put to shame ever. 
Deliver me by your righteousness.}%
\verse{Incline your ear to me. 
Quickly deliver me. 
Become my rock of refuge, 
\textit{a fortified keep}\lebnote{Literally “a house of strongholds”}to save me.}%
\verse{For you are my rock and my fortress. 
So, for the sake of your name, 
lead me and guide me.}%
\verse{Bring me out of the net that they have secretly set for me, 
for you are my refuge.}%
\verse{Into your hand I commit my spirit. 
You have redeemed me, O Adonai, faithful God.\lebnote{Or “God of truth”}}%
\verse{I hate those devoted to useless idols, 
but I trust Adonai.}%
\verse{I will exult and rejoice in your loyal love. 
Because you have seen my misery, 
you know the distresses of my life.\lebnote{Or “soul”}}%
\verse{And you have not delivered me 
into the hand of the enemy. 
You have set my feet in a broad place.}%
\verse{Be gracious to me, O Adonai, 
because I have distress. 
My eye wastes away because of vexation, 
along with my soul and my \textit{body}.\lebnote{Literally “belly”}}%
\verse{For my life is at an end with sorrow, 
and my years with sighing. 
My strength stumbles because of my iniquity, 
and my bones waste away.}%
\verse{Because of all my adversaries I have become a disgrace, 
especially to my neighbors, 
and a dread to my acquaintances. 
Those who see me in the street flee from me.}%
\verse{I have become forgotten like one dead, out of \textit{mind}.\lebnote{Literally “heart”}
I am like a destroyed vessel.}%
\verse{For I hear the rumor of many, 
“Terror on every side!” 
When conspiring together against me, 
they have plotted to take my life.}%
\verse{But as for me, I trust you, O Adonai. 
I say, “You are my God.”}%
\verse{My times\lebnote{Or “destiny”}are in your hand. 
Deliver me from the hand of my enemies 
and from those who pursue me.}%
\verse{Shine your face upon your servant. 
Save me by your loyal love.}%
\verse{O Adonai, let me not be put to shame, for I call on you. 
Let the wicked be put to shame. 
Let them go silently\lebnote{Or “let them with wailing enter”}to Sheol.}%
\verse{Let lying lips be dumb, 
that speak against the righteous\lebnote{Or “the innocent one”}unrestrained 
with arrogance and contempt.}%
\verse{How abundant is your goodness 
that you have stored up for those who fear you, 
that you perform for those who take refuge in you 
before the children of humankind.}%
\verse{You will hide them in the protection of your presence 
from the plots of man. 
You will hide them in a shelter from the strife of tongues.}%
\verse{Blessed is Adonai, 
because he has worked marvelously his loyal love to me 
in a besieged city.}%
\verse{As for me, I said in my alarm, 
“I am cut off from before your eyes.” 
However you heard the voice of my supplications 
when I cried to you for help.}%
\verse{Love Adonai, all you his faithful ones. 
Adonai preserves the faithful 
but repays abundantly the one who acts arrogantly.}%
\verse{Be strong and let your\lebnote{Plural}heart show strength, 
all you who wait for Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 32
\verseWithHeading{Of David. A maskil.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Happy is he whose transgression is taken away, 
whose sin is covered.}%
\verse{Happy is a person to whom Adonai does not impute iniquity 
and in whose spirit there is not deceit.}%
\verse{When I kept silent, my bones were worn out 
due to my groaning all the day.}%
\verse{For day and night your hand was heavy upon me. 
My vigor was changed into the dry heat of summer. \textit{Selah}}%
\verse{I made known my sin to you, and my iniquity I did not cover. 
I said, “I will confess concerning my transgressions to Adonai,” 
and you took away the guilt of my sin. \textit{Selah}}%
\verse{Therefore let all the faithful pray to you 
at the time for finding you. 
Surely at the flood of many waters they will not reach him.}%
\verse{You are my hiding place; 
from trouble you preserve me. 
With cries of deliverance you surround me. \textit{Selah}}%
\verse{I\lebnote{Adonai is speaking in vv. 8–9}will instruct you and teach you 
in the way that you should go. 
I will advise you with my eye upon you.}%
\verse{Do not be like a horse or like a mule, without understanding; 
that needs his tackle—bridle and rein—for restraint 
or he would not come near you.}%
\verse{Many are the pains of the wicked, 
but for the one who trusts Adonai 
loyal love surrounds him.}%
\verse{Be glad in Adonai and rejoice, you righteous, 
and shout for joy, all you upright of heart.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 33
\verseWithHeading{Praise to Adonai for His Character and Creation}{Exult in Adonai, O you righteous, 
for praise is fitting for the upright.}%
\verse{Give thanks to Adonai with the lyre; 
with a harp of ten strings play to him.}%
\verse{Sing to him a new song; 
play skillfully with jubilant shout.}%
\verse{For the word of Adonai is upright, 
and all his work is done in faithfulness.}%
\verse{He loves righteousness and justice. 
The earth is full of the loyal love of Adonai.}%
\verse{By the word of Adonai the heavens were made, 
and by the breath of his mouth all their host.}%
\verse{He gathered the waters of the sea like a heap. 
He put the deeps in storehouses.}%
\verse{Let all the earth fear Adonai. 
Let all the inhabitants of the world stand in awe of him.}%
\verse{For \textit{he himself spoke}\lebnote{Literally “he, he spoke”}and it came to pass. 
\textit{He himself commanded}\lebnote{Literally “He, he commanded”}and it stood firm.}%
\verse{Adonai brings to nothing\lebnote{Or “breaks”}the plan of nations; 
he frustrates\lebnote{Or “nullifies”}the intents of peoples.}%
\verse{The plan of Adonai stands firm forever, 
the intents of his heart \textit{from one generation to the next}.\lebnote{Literally “to a generation and a generation”}}%
\verse{Blessed is the nation whose God is Adonai, 
the people he has chosen for his inheritance.}%
\verse{From heaven Adonai looks; 
he sees all the children of humankind.}%
\verse{From his dwelling place\lebnote{Or “the place where he sits enthroned”}he gazes 
on all the inhabitants of the earth,}%
\verse{he who fashions altogether their heart, 
he who understands all their works.}%
\verse{The king is not saved by the greatness of an army. 
The warrior is not delivered by the greatness of strength.}%
\verse{The horse is a false hope for victory, 
nor can it save by the greatness of its power.}%
\verse{Behold, the eye of Adonai is on those who fear him, 
on those who hope\lebnote{Or “wait expectantly”}for his loyal love}%
\verse{to deliver their soul from death 
and to keep them alive in famine.}%
\verse{Our soul waits for Adonai; 
he is our help and our shield.}%
\verse{For in him our heart rejoices 
because in his holy name we trust.}%
\verse{Let your loyal love, O Adonai, be upon us, 
\textit{even as}\lebnote{Literally “as which”}we hope in you.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 34
\verseWithHeading{Of David, when he \textit{pretended to be crazy}\lebnote{Literally “he changed his discernment”}in front of Abimelech\lebnote{Literally “my father is king” whereas 1 Sam 21:10(9) has “Achish.” “Abimelech” is perhaps an honorific title, since its literal meaning suggests a legitimate right to rule}
so that he drove him out, and so he departed.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{I will bless Adonai at all times; 
his praise shall be in my mouth continually.}%
\verse{My soul makes its boast in Adonai; 
let the humble hear and be glad.}%
\verse{Magnify Adonai with me, 
and let us exalt his name together.}%
\verse{I sought Adonai and he answered me, 
and from all my terrors he delivered me.}%
\verse{They looked to him and were radiant, 
and their faces shall not be ashamed.\lebnote{Or “and let not their faces be ashamed”}}%
\verse{This poor man called and Adonai heard, 
and saved him from all his troubles.}%
\verse{The angel of Adonai encamps 
around those who fear him, 
and he rescues them.}%
\verse{Taste and see that Adonai is good; 
blessed is the man who takes refuge in him.}%
\verse{Fear Adonai, you his saints,\lebnote{Or “holy ones”}
for there is no lack for those who fear him.}%
\verse{The young lions are in want and suffer hunger, 
but those who seek Adonai will not lack for any good thing.}%
\verse{Come, you children, listen to me; 
I will teach you the fear of Adonai.}%
\verse{Who is the man who desires life, 
who loves many days in order to see good?}%
\verse{Keep your tongue from evil 
and your lips from speaking deceit.}%
\verse{Turn from evil and do good. 
Seek peace and pursue it.}%
\verse{The eyes of Adonai are toward the righteous, 
and his ears toward their cry for help.}%
\verse{The face of Adonai is against evildoers, 
to cut off the remembrance of them from the earth.}%
\verse{They\lebnote{That is, the righteous}cry out and Adonai hears 
and delivers them from all of their troubles.}%
\verse{Adonai is near to those who are heartbroken 
and saves those who are crushed in spirit.}%
\verse{Many are the distresses of the righteous, 
but Adonai delivers him out of them all.}%
\verse{He protects all his bones; 
not one of them is broken.}%
\verse{Evil will slay the wicked, 
and those who hate the righteous will incur guilt.}%
\verse{Adonai redeems the life of his servants, 
and none who take refuge in him will incur guilt.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 35
\verseWithHeading{Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Contend, O Adonai, with my contenders; 
fight those who fight me.}%
\verse{Grasp buckler and shield 
and rise to my aid.}%
\verse{And draw the spear and javelin to meet those who pursue me. 
Say to my soul, “I am your salvation.”}%
\verse{Let those who seek my life be shamed and humiliated. 
Let those who plot calamity against me be repulsed and ashamed.}%
\verse{Let them be like chaff before the wind, 
with the angel of Adonai \textit{driving them}.\lebnote{Literally “pushing”}}%
\verse{Let their way be dark and slippery, 
with the angel of Adonai pursuing them.}%
\verse{For without cause they secretly hide the pit with their net for me; 
without cause they dug it for my life.}%
\verse{Let \textit{unforeseen ruin}\lebnote{Literally “ruin he does not know”}come on him, 
and his net that he hid, let it catch him. 
Let him fall into it in ruin.}%
\verse{Then my soul will rejoice in Adonai; 
it will rejoice in his salvation.}%
\verse{All of my bones shall say, “O Adonai, who is like you, 
who delivers the poor from one stronger than he 
and the poor and needy from the one who robs him?”}%
\verse{Violent witnesses rise up; 
they ask me concerning what I do not know.}%
\verse{They repay me evil in place of good. 
It is bereavement to my soul.}%
\verse{But as for me, when they were sick, my clothing was sackcloth. 
I weakened my soul with fasting, 
and my prayer \textit{returned to me unanswered}.\lebnote{Literally “upon my lap it will return”}}%
\verse{I \textit{behaved}\lebnote{Literally “walked about”}as though he were a friend or as a brother to me. 
As one lamenting a mother, I was bowed down in mourning.}%
\verse{But at my stumbling they rejoiced and gathered together; 
smiters whom I did not know gathered against me. 
They tore and did not cease.}%
\verse{Among the ungodly of the mockers at feasts,\lebnote{With a slight change the Hebrew might read, “The ungodly boldly mocked”}
they gnashed at me with their teeth.}%
\verse{My Lord, how long will you watch? 
Restore my life from their ravages,\lebnote{Or “roaring”}
my only life from the young lions.}%
\verse{I will give thanks to you in the great assembly; 
among the mighty people I will praise you.}%
\verse{Let not those who are wrongfully my enemies\lebnote{Or “those who oppose me with falsehood”}rejoice over me. 
Nor let those who hate me without cause wink the eye.}%
\verse{For they do not speak peace, 
but against the quiet ones of the land 
they plan deceitful words.}%
\verse{They also made wide their mouths\lebnote{Hebrew “mouth”}against me. 
They said, “Aha! Aha! 
Our eyes have seen it.”}%
\verse{You have seen, O Adonai. Do not be deaf. 
O Lord, do not be far from me.}%
\verse{Wake up and rouse yourself for my right, 
for my cause, O my God and my Lord.}%
\verse{Vindicate me according to your righteousness, 
O Adonai my God, 
and do not let them rejoice over me.}%
\verse{Do not let them say in their hearts,\lebnote{Hebrew “heart”}“Aha, our desire.” 
Let them not say, “We have swallowed him up.”}%
\verse{Let them be shamed and abashed altogether, 
who rejoice at my misfortune. 
Let them put on shame and insult, 
who magnify themselves against me.}%
\verse{Let them shout for joy and be glad, 
who delight in my vindication; 
and let them say continually, “Adonai is great, 
who delights in the welfare of his servant.”}%
\verse{Then my tongue will proclaim your righteousness, 
and your praise all day.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 36
\verseWithHeading{For the music director. Of David, the servant of Adonai.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{An oracle: the wicked has rebellion 
in the midst of his\lebnote{Reading with a few Hebrew manuscripts and some versions; MT reads “my”}heart. 
There is no fear of God before his eyes.}%
\verse{For he flatters himself in his eyes, 
hating to detect his iniquity.\lebnote{Or “when finding his iniquity and when hating”}}%
\verse{The words of his mouth are wickedness and deceit. 
He has ceased to have insight and to do good.}%
\verse{He plans sin on his bed. 
He puts himself on a way that is not good. 
He does not reject evil.}%
\verse{O Adonai, your loyal love extends into the heavens, 
your faithfulness unto the clouds.}%
\verse{Your righteousness is \textit{like the mighty mountains},\lebnote{Literally “like the mountains of God”}
your judgments like the great deep. 
You save man and beast, O Adonai.}%
\verse{How precious is your loyal love, O God, 
and the children of humankind 
take refuge in the shadow of your wings.}%
\verse{They are refreshed with the \textit{fullness}\lebnote{Literally “fatness”}of your house,\lebnote{Or “temple”}
and you give them drink from the river of your delights.}%
\verse{For with you is the fountain of life; 
in your light we see light.}%
\verse{Prolong your loyal love to those who know you, 
and your righteousness to the upright of heart.}%
\verse{Do not let a foot of pride come against me, 
nor let a wicked hand make me to wander homeless.}%
\verse{There doers of evil have fallen; 
they are thrust down and not able to rise.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 37
\verseWithHeading{Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Do not fret because of evildoers; 
do not be envious of doers of wickedness.}%
\verse{For like the grass they will dry up quickly, 
and like green vegetation they will wither.}%
\verse{Trust Adonai and do good. 
Abide in the land and feed on faithfulness.\lebnote{That is, God’s faithfulness}}%
\verse{Take pleasure in Adonai as well, 
and he will give to you the requests of your heart.}%
\verse{\textit{Commit to}\lebnote{Literally “Roll upon”}Adonai your way; 
Trust also on him and he will act.}%
\verse{Then he will bring forth your righteousness like the light, 
and your justice like the noonday.}%
\verse{Be quiet before Adonai and wait for him. 
Do not fret about one who succeeds in his way, 
about a man making plots.}%
\verse{Refrain from anger and forsake wrath. 
Do not fret; it only brings evil.}%
\verse{For evildoers will be cut off, 
but those who wait for Adonai— 
they will possess the land.}%
\verse{And\lebnote{Or “But”}yet a little while and the wicked will not be, 
and you will look carefully upon his place, but he will not be.}%
\verse{But the afflicted will possess the land, 
and they will take pleasure in an abundance of peace.}%
\verse{The wicked plans evil against the righteous, 
and gnashes at him with his teeth.}%
\verse{The Lord laughs at him, 
for he sees that his day is coming.}%
\verse{The wicked have drawn the sword and have bent their bow 
to throw down the poor and the needy, 
to kill those upright in their way.}%
\verse{Their sword will enter into their own heart, 
and their bows shall be broken.}%
\verse{The little belonging to the righteous is better 
than the wealth of many wicked.}%
\verse{For the arms of the wicked shall be broken, 
but Adonai supports the righteous.}%
\verse{Adonai knows the days of the blameless, 
and their inheritance will be forever.}%
\verse{They will not be put to shame in distressing times,\lebnote{Hebrew “time”}
and in the days of famine they will be satisfied.}%
\verse{But the wicked will perish, 
and the enemies of Adonai will be like the splendor of the pastures. 
They vanish; in smoke they vanish.}%
\verse{The wicked borrows and does not repay, 
but the righteous is gracious and gives.}%
\verse{For those blessed by him shall possess the land, 
but those cursed by him shall be cut off.}%
\verse{The steps of a man are established by Adonai, 
and he takes pleasure in his way.}%
\verse{Though he fall, he will not be cast headlong, 
for Adonai supports him with his hand.}%
\verse{I was a youth, but I am now old; 
yet I have not seen the righteous forsaken 
or his children \textit{begging for}\lebnote{Literally “searching”}bread.}%
\verse{All the day he is gracious and lends, 
and his children are a blessing.}%
\verse{Turn aside from evil and do good 
and so abide forever.}%
\verse{For Adonai loves justice 
and will not forsake his faithful ones. 
They are protected forever. 
But the children of the wicked will be cut off.}%
\verse{The righteous will possess the land 
and abide in it forever.}%
\verse{The mouth of the righteous utters wisdom, 
and his tongue speaks justice.}%
\verse{The law of his God is in his heart. 
His steps will not slip.}%
\verse{The wicked watches for the righteous 
and seeks to kill him.}%
\verse{Adonai will not abandon him into his hand, 
and he will not let him be condemned when he is judged.}%
\verse{Wait for Adonai and keep his way, 
and he will exalt you to possess the land. 
When the wicked are cut off, you will see it.}%
\verse{I have seen the wicked acting violently 
and spreading himself out 
like a luxuriant native tree.}%
\verse{Then he passed on\lebnote{Dead Sea Scrolls and the versions read “I passed by”}and behold, he was not. 
And I sought him, but he could not be found.}%
\verse{Observe the blameless and look at the upright, 
for there is a future for a man of peace.}%
\verse{But transgressors shall be destroyed altogether. 
The future of the wicked shall be cut off.}%
\verse{And\lebnote{Some manuscripts omit “and,” preserving the acrostic}the salvation of the righteous is from Adonai, 
their refuge in the time of trouble.}%
\verse{And Adonai helps them and he rescues them. 
He rescues them from the wicked and saves them, 
because they take refuge in him.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 38
\verseWithHeading{A psalm of David. To bring to remembrance.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{O Adonai, do not rebuke me in your anger 
or chastise me in your wrath.}%
\verse{For your arrows have sunk into me, 
and your hand has pressed down on me.}%
\verse{There is no soundness in my flesh because of your indignation. 
There is no health in my bones because of my sin.}%
\verse{For my iniquities have passed over my head; 
like a heavy burden they are too heavy for me.}%
\verse{My wounds start to stink; they rot 
because of my foolishness.}%
\verse{I am bowed down; I am bent over greatly. 
All the day I go about mourning.}%
\verse{For my loins are full of burning, 
and there is no soundness in my flesh.}%
\verse{I am faint and crushed greatly; 
I groan because of the roaring of my heart.}%
\verse{O Lord, all my longing is before you, 
and my sighing is not hidden from you.}%
\verse{My heart throbs violently, my strength leaves me; 
and the light of my eyes, that also is not with me.}%
\verse{My friends and companions stand aloof from my affliction, 
and my relatives stand afar off.}%
\verse{Those who seek my life lay snares as well, 
and those intent on my harm speak threats. 
They also plot deceit all day.}%
\verse{But as for me, like the deaf I cannot hear, 
and I am like the mute who cannot open his mouth.}%
\verse{And so I am like a man who hears not, 
and in whose mouth there are no retorts.}%
\verse{Rather for you I wait, O Adonai. 
You will answer, O Lord my God.}%
\verse{For I said, “Help, lest they rejoice over me, 
lest they boast against me when my foot slips.”}%
\verse{For I am ready to stumble, 
and my pain is before me continually.}%
\verse{For my iniquity I confess; 
I am anxious because of my sin.}%
\verse{And my enemies without cause\lebnote{Or “vigorous”}are numerous, 
and those who hate me wrongfully\lebnote{Or “those who hate me with falsehood”}are many.}%
\verse{And those who repay evil in return for good 
accuse me in return for my pursuing good.}%
\verse{Do not forsake me, O Adonai. 
O my God, do not be far from me.}%
\verse{Hurry to help me, 
O Lord, my salvation.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 39
\verseWithHeading{For the music director. For Jeduthun.\lebnote{One of David’s musicians (1 Chr 16:41)}A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{I said, “I will guard my ways 
\textit{that I may not sin}\lebnote{Literally “from sinning”}with my tongue. 
I will keep a muzzle over my mouth 
as long as the wicked are before me.”}%
\verse{I was mute with silence. I was silent even from saying good things, 
and my pain was stirred up.}%
\verse{My heart grew hot inside me; 
in my sighing a fire burned. 
Then I spoke with my tongue,}%
\verse{“Let me know, O Adonai, my end, 
and what is the measure of my days. 
Let me know how transient I am.”}%
\verse{Look, you have made my days mere handbreadths, 
and my lifespan as nothing next to you. 
Surely every person standing firm is complete vanity. \textit{Selah}}%
\verse{Surely a man walks about as a mere \textit{shadow};\lebnote{Literally “image”}
surely in vain they bustle about. 
He heaps up possessions but does not know who will gather them in.}%
\verse{And now, O Lord, for what do I wait? 
My hope is for you.}%
\verse{From all my transgressions deliver me; 
do not make me the taunt of the fool.}%
\verse{I am mute. I do not open my mouth, 
for you, yourself, have done it.}%
\verse{Remove from me your affliction. 
By the opposition of your hand I perish.}%
\verse{When with rebukes you chastise a man for sin, 
you\lebnote{Hebrew “and/then you”}\textit{consume}\lebnote{Literally “cause to melt”}like a moth his delightful things. 
Surely everyone is a mere vapor.\lebnote{The word translated “vanity” in vv. 6, 7}\textit{Selah}}%
\verse{Hear my prayer, O Adonai, and listen to my cry for help; 
do not be deaf to my tears. 
For I am an alien\lebnote{Or “stranger”}with you, 
a sojourner like all my ancestors.\lebnote{Or “fathers”}}%
\verse{Look away from me that I may be cheerful, 
before I depart and I am no more.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 40
\verseWithHeading{For the music director. Of David. A psalm.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{I waited patiently for Adonai, 
And he inclined to me 
and heard my cry for help.}%
\verse{And so he brought me up from the roaring pit,\lebnote{Some read “pit of destruction”}
from the miry clay. 
And he put my feet upon a rock; 
he made my steps steady.}%
\verse{Then he put a new song in my mouth, 
a praise to our God. 
Many will see and fear, 
and will trust Adonai.}%
\verse{Blessed is the man who makes Adonai his trust 
and does not turn to the proud and to those who fall away to a lie.\lebnote{Or “false god”}}%
\verse{Many things, O Adonai my God, you have done— 
your wonderful deeds and your thoughts toward us. 
There is none to compare with you. 
If I tried to proclaim and tell of them, 
they would be too numerous to count.}%
\verse{A sacrifice and offering you do not desire. 
\textit{My ears you have opened}.\lebnote{Literally “Ears you have dug for me.”}
Burnt offering and sin offering you have not demanded.}%
\verse{Then I said, “Look, I come. 
In the scroll of the book 
it is written concerning me:}%
\verse{‘I delight to do your will, O my God, 
and your law is \textit{deep within me}.’ ”\lebnote{Literally “in the midst of my intestines”}}%
\verse{I have brought good tidings of righteousness in the great congregation. 
Look, I have not shut my lips. 
O Adonai, \textit{you surely know that}.\lebnote{Literally “you, you know”}}%
\verse{Your righteousness I have not hidden in the midst of my heart. 
I have spoken of your faithfulness and your salvation. 
I have not concealed your loyal love or your faithfulness 
from the great congregation.}%
\verse{As for you, O Adonai, do not withhold your mercies from me. 
Let your loyal love and your faithfulness 
continually preserve me.}%
\verse{For evils without number have encompassed me. 
My iniquities have overtaken me, so that I am not able to see. 
They are more than the hairs of my head, 
and my heart fails me.}%
\verse{Be pleased, O Adonai, to deliver me. 
O Adonai, hurry to help me.}%
\verse{Let them be shamed and abashed altogether 
who seek to take away my life.\lebnote{Or “soul”}
Let them be repulsed and humiliated 
who desire my harm.}%
\verse{Let them be appalled because of their shame, 
those who say to me, “Aha! Aha!”}%
\verse{Let them rejoice and be glad in you, 
all those who seek you. 
Let them say continually, “Adonai is great!” 
—those who love your salvation.}%
\verse{But I am poor and needy. 
Let my Lord consider me. 
You are my help and my deliverer. 
O my God, do not delay.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 41
\verseWithHeading{For the music director. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Blessed is the one who has regard for the poor; 
in the day of disaster, Adonai delivers him.}%
\verse{Adonai protects him and keeps him alive; 
he is blessed in the land, 
and you do not give him 
into the will\lebnote{Hebrew “soul”}of his enemies.}%
\verse{Adonai sustains him on his sick bed. 
In his illness, you \textit{restore to health}.\lebnote{Literally “his every bed you change”}}%
\verse{As for me, I said, “O Adonai, be gracious to me. 
Heal me,\lebnote{Or “my soul”}for I have sinned against you.”}%
\verse{My enemies speak evil about me, 
“When will he die and his name perish?”}%
\verse{And when one comes to see me, he speaks falsely; 
his heart gathers disaster for itself. 
He goes out to the street; he speaks.}%
\verse{All who hate me speak together against me. 
Against me they assume the \textit{worst}\lebnote{Literally “evil”}for me:}%
\verse{“A ruinous thing is poured out on him, 
and now that he lies down, he will not rise up again.”}%
\verse{Even \textit{my close friend},\lebnote{Literally “the man of my peace”}whom I trusted, 
who ate my bread, 
has lifted his heel against me.}%
\verse{But you, O Adonai, be gracious to me and raise me up 
that I may repay them.}%
\verse{By this I know that you delight in me: 
because my enemy has not shouted in triumph over me.}%
\verse{As for me, you have upheld me in my integrity, 
and you have set me in your presence forever.}%
\verse{Blessed be Adonai, the God of Israel, 
from everlasting to everlasting. 
Amen and Amen.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 42
\verseWithHeading{For the music director. A maskil of the sons of Korah.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{As a deer longs for streams of water, 
so my soul longs for you, O God.}%
\verse{My soul thirsts for God, for the living God. 
When shall I come and appear before God?\lebnote{Some Hebrew manuscripts and versions read “and see the face of God?”}}%
\verse{My tears have been my food day and night, 
while they say to me all day long, 
“Where is your God?”}%
\verse{These I remember and I pour out my soul within me: 
that I would go with the multitude; 
I led them in procession to the house\lebnote{Or “temple”}of God, 
with a voice of rejoicing and thanksgiving, 
a crowd celebrating a festival.}%
\verse{Why are you \textit{in despair},\lebnote{Literally “bowed down”}O my soul, 
and disturbed within me? 
Hope in God, because I will again praise him, 
for the salvation of his presence.}%
\verse{O my God, within me my soul is \textit{in despair};\lebnote{Literally “bowed down”}
therefore I remember you from the land of Jordan 
and the heights of Hermon, from the mountain of Mizar.}%
\verse{Deep is calling to deep 
at\lebnote{Or “to”}the thunder of your waterfalls. 
All your breakers and your waves 
have passed over me.}%
\verse{By day Adonai commands his loyal love, 
and in the night his song is with me, 
a prayer to the God of my life.}%
\verse{I say to God, my rock, 
“Why have you forgotten me? 
Why must I walk about mourning 
because of the oppression of the enemy?”}%
\verse{As with a shattering in my bones 
my oppressors taunt me, 
while they say to me \textit{all day},\lebnote{Literally “all the day”}
“Where is your God?”}%
\verse{Why are you \textit{in despair},\lebnote{Literally “bowed down”}O my soul? 
And why are you disturbed within me? 
Hope in God, because I shall again praise him, 
\textit{my salvation}\lebnote{Literally “the salvation of my face”}and my God.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 43
\verseWithHeading{A Prayer for Rescue}{Judge me, O God, and plead my case 
against\lebnote{Or “because of”}an unfaithful\lebnote{Or “disloyal”}nation. 
From a man of deceit and wickedness rescue me,}%
\verse{because you are the God of my refuge. 
Why have you rejected me? 
Why must I go about mourning 
because of the oppression of the enemy?}%
\verse{Send your light and your truth; 
they shall lead\lebnote{Or “let them lead”}me. 
They shall bring me to \textit{your holy mountain}\lebnote{Literally “the mountain of your holiness”}
and to your dwelling places.}%
\verse{Then\lebnote{Or “And”}I will go to the altar of God, 
to God, \textit{my surpassing joy},\lebnote{Literally “the joy of my gladness”}
and I will praise you with lyre, 
O God, my God.}%
\verse{Why are you \textit{in despair},\lebnote{Literally “bowed down”}O my soul? 
And why are you disturbed within me? 
Hope in God, because I will again praise him, 
\textit{my salvation}\lebnote{Hebrew “the salvation of my face”}and my God.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 44
\verseWithHeading{For the music director. Of the sons of Korah. A maskil.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{O God, we have heard with our ears; 
our ancestors\lebnote{Or “fathers”}have told us 
of work you worked in their days, 
in days of old.}%
\verse{You with your hand drove out the nations, 
but them\lebnote{That is, Israel}you planted. 
You harmed the peoples, 
but them\lebnote{That is, Israel}you let spread out.}%
\verse{For not with their sword 
did they possess the land, 
and their arm did not give them victory. 
Rather it was your right hand and your arm 
and the light of your presence, 
because you delighted in them.}%
\verse{You are my king, O God. 
Command\lebnote{Or “who commands,” see LXX}victories for Jacob.}%
\verse{By you we push down our enemies; 
by your name we tread down those who rise up against us.}%
\verse{For I do not trust my bow, 
and my sword cannot give me victory.}%
\verse{Rather you have saved us from our enemies,\lebnote{Or “made us victorious over our enemies”}
and have humiliated those who hate us.}%
\verse{In God we boast all the day, 
and we will give thanks to your name forever. \textit{Selah}}%
\verse{Surely you have rejected and disgraced us, 
and have not gone out with our armies.}%
\verse{You have caused us to pull back from the enemy, 
and so those who hate us have plundered for themselves.}%
\verse{You have given us as sheep for food, 
and among the nations you have scattered us.}%
\verse{You have sold your people \textit{cheaply},\lebnote{Literally “for what is not wealth”}
and did not profit by their price.}%
\verse{You have made us a taunt to our neighbors, 
a derision and a scorn to those around us.}%
\verse{You have made us a byword among the nations, 
a shaking of the head among the peoples.}%
\verse{\textit{All day long}\lebnote{Literally “All of the day”}my disgrace is before me, 
and the shame of my face covers me,}%
\verse{because of the voice of the taunter and the reviler, 
because of the enemy and the avenger.}%
\verse{All this has befallen us, though we have not forgotten you, 
and we have not been false to your covenant.}%
\verse{Our heart has not turned back, 
and our steps have not turned aside from your way.}%
\verse{But you have crushed us in a place of jackals, 
and have covered us with deep shadow.}%
\verse{If we had forgotten the name of our God, 
or had spread out our hands in prayer to a foreign god,}%
\verse{would not God discover this, 
for he knows the secrets of the heart?}%
\verse{Rather, on account of you we are killed \textit{all day long};\lebnote{Literally “all of the day”}
we are accounted as sheep for slaughter.}%
\verse{Wake up! Why do you sleep, O Lord? 
Awake! Do not reject forever.}%
\verse{Why do you hide your face? 
Have you forgotten our misery and our oppression?}%
\verse{For our soul is bowed down to the dust. 
Our \textit{body}\lebnote{Literally “belly”}clings to the ground.}%
\verse{Rise up! Be a help for us, 
and redeem us for the sake of your loyal love.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 45
\verseWithHeading{For the music director; according to The Lilies. Of the sons of Korah. 
A maskil. A song of love.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{My heart is moved with a good word; 
I recite my \textit{compositions}\lebnote{Literally “works,” that is, the song he has composed}to the king. 
My tongue is the pen of a skilled scribe.}%
\verse{You are the most handsome of the sons of humankind; 
grace is poured out on your lips; 
therefore God has blessed you forever.}%
\verse{Gird your sword on your thigh, O mighty one, 
in your splendor and your majesty.}%
\verse{And in your majesty \textit{ride victoriously},\lebnote{Literally “Succeed! Ride!”}
because of truth and humility and righteousness. 
And let your right hand teach you awesome deeds.}%
\verse{Your arrows are sharp; 
peoples fall under you 
in the midst of the king’s enemies.}%
\verse{Your throne, O God, is forever and ever. 
A scepter of uprightness is 
the scepter of your kingdom.}%
\verse{You love righteousness and hate wickedness. 
Therefore God, your God, has anointed you 
from among your companions with festive oil.}%
\verse{All your robes are scented with myrrh and aloes and cassia. 
From palaces of ivory stringed instruments gladden you.}%
\verse{Kings’ daughters are among your noble ladies. 
The queen stands at your right hand in gold of Ophir.}%
\verse{Hear, O daughter, and \textit{discern}\lebnote{Literally “see”}and incline your ear, 
and forget your people and your father’s house.}%
\verse{Let the king desire your beauty.\lebnote{Or “Should the king desire … then bow down …”}
Because he is your lord, therefore\lebnote{Hebrew “and”}bow down to him.}%
\verse{Even the daughter of Tyre will come with a gift. 
The rich from among people will seek your \textit{favor}.\lebnote{Literally “face”}}%
\verse{The king’s daughter is all glorious within;\lebnote{That is, within her bridal chamber}
her garment is of gold embroidered cloth.}%
\verse{She is brought to the king in colorful garments. 
The young women behind her, her attendants, 
are being brought to you.}%
\verse{They are led with joy and gladness. 
They enter the palace of the king.}%
\verse{In place of your\lebnote{The pronouns and verbal forms are masculine in this verse}fathers will be your sons. 
You will make them princes in all the land.\lebnote{Or “earth”}}%
\verse{I will cause your name to be remembered in all generations; 
therefore peoples will praise you forever and ever.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 46
\verseWithHeading{For the music director. Of the sons of Korah. 
According to Alamoth. A song.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{God is our refuge and strength, 
a very sufficient help in troubles.}%
\verse{Therefore we will not fear though the earth change, 
and though the mountains totter into the \textit{midst}\lebnote{Literally “heart”}of the sea,}%
\verse{though its waters roar and foam, 
though mountains shake with its surging water. \textit{Selah}}%
\verse{There is a river whose streams gladden the city of God, 
the holiest of the dwellings of the Most High.}%
\verse{God is in the midst of her; 
she will not be made to totter. 
God will help her \textit{at daybreak}.\lebnote{Literally “at the turning of the morning”}}%
\verse{Nations roar, kingdoms shake; 
he utters his voice, the earth melts.}%
\verse{Adonai of hosts is with us; 
our high stronghold is the God of Jacob. \textit{Selah}}%
\verse{Come, see the works of Adonai, 
who has placed desolations on the earth.}%
\verse{He makes wars cease to the ends of the earth; 
he breaks the bow and cuts off the spear. 
The wagons of war he burns with fire.}%
\verse{Be still, and know that I am God. 
I will be exalted among the nations; 
I will be exalted in the earth.}%
\verse{Adonai of Hosts is with us; 
the God of Jacob is our high stronghold.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 47
\verseWithHeading{For the music director. Of the sons of Korah. A psalm.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{All you peoples, clap your hands.\lebnote{Hebrew “hand”}
Shout to God with a voice of rejoicing.}%
\verse{For Adonai Most High is awesome,\lebnote{Or “to be feared”}
a great king over all the earth.}%
\verse{He subdues peoples under us 
and nations under our feet.}%
\verse{He chooses for us our inheritance, 
the pride of Jacob whom he loves.}%
\verse{God has gone up with a shout, 
Adonai with the sound of a trumpet.}%
\verse{Sing praises to God, sing praises. 
Sing praises to our king, sing praises.}%
\verse{For God is king of all the earth. 
Sing praises with understanding.\lebnote{Or “sing a maskil”}}%
\verse{God reigns over nations; 
God sits on his holy throne.}%
\verse{The princes of the peoples are gathered together 
with the people of Abraham’s God. 
For the shields of the earth belong to God. 
He is very exalted.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 48
\verseWithHeading{A song. A psalm of the sons of Korah.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Adonai is great and very worthy of praise 
in the city of our God, in his \textit{holy mountain}.\lebnote{Literally “mountain of holiness”}}%
\verse{Beautiful in elevation, the joy of the whole earth, 
is \textit{Mount Zion},\lebnote{Literally “the mountain of Zion”}\textit{in the far north},\lebnote{Literally “far of north”; the word translated “north” could refer to Mount Zaphon}
the city of the great king.}%
\verse{God is in her citadels; 
he is known as a high stronghold.}%
\verse{For see, the kings assembled;\lebnote{That is, preparing for battle}
they advanced together.}%
\verse{\textit{They themselves saw}\lebnote{Literally “They, they saw”}it, so they were astonished. 
They were terrified; they ran off.}%
\verse{Trembling seized them there— 
pain as of a woman in labor.}%
\verse{With an east wind 
you shatter\lebnote{A few Hebrew manuscripts read “like an east wind that shatters”}the ships of Tarshish.\lebnote{That is, ships that can sail to Tarshish, i.e. well built, able to traverse great distances}}%
\verse{As we have heard, so we have seen 
in the city of Adonai of hosts, in the city of our God. 
God will establish her forever. \textit{Selah}}%
\verse{We have pondered your loyal love, O God, 
in the midst of your temple.}%
\verse{As is your name, O God, so is your praise 
to the ends of the earth. 
Your right hand is full of righteousness.}%
\verse{Let \textit{Mount Zion}\lebnote{Literally “the mountain of Zion”}rejoice; 
let the daughters of Judah rejoice 
because of your judgments.}%
\verse{Walk about Zion and circle it; 
count her towers.}%
\verse{\textit{Consider well}\lebnote{Literally “Set your heart to”}her ramparts. 
Go through her citadels 
so that you can tell the next generation}%
\verse{that this is God, our God forever and ever. 
\textit{He himself}\lebnote{Literally “he, he will …”}will guide us until death.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 49
\verseWithHeading{For the music director. Of the sons of Korah. A psalm.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Hear this, all you peoples; 
give ear, all you inhabitants of the world,}%
\verse{both low and high, 
rich and poor together.}%
\verse{My mouth will speak wisdom, 
and the meditation of my heart will be understanding.}%
\verse{I will incline my ear to a proverb; 
I will \textit{propound}\lebnote{Literally “open”}my riddle on a lyre.}%
\verse{Why should I fear in \textit{times of calamity},\lebnote{Literally “the days of evil”}
when iniquity surrounds me at my heels,}%
\verse{those who trust their wealth 
and boast about the abundance of their riches?}%
\verse{Surely a man cannot redeem a brother. 
He cannot give to God his ransom}%
\verse{(since\lebnote{Hebrew “and”}the redemption price for their life is costly 
and it always fails),}%
\verse{so that he may stay alive forever 
and not see the pit.}%
\verse{For he sees that the wise die, 
together with the fool and brute they perish, 
and leave their wealth to the next generation.}%
\verse{Within them they think their houses are forever, 
their dwelling places from generation to generation. 
They \textit{name}\lebnote{Literally “call”}their lands by their own names.}%
\verse{But man cannot continue in his pomp. 
He is like the beasts that perish.}%
\verse{This is the journey\lebnote{Hebrew “way”}of those who have foolish confidence, 
and those after them who accept their sayings. \textit{Selah}}%
\verse{Like sheep they are destined to Sheol; 
death will shepherd them. 
But the upright will rule over them in the morning, 
and their forms will be for Sheol to consume, 
far from his lofty abode.}%
\verse{Surely God will ransom my life 
from the \textit{power}\lebnote{Literally “hand”}of Sheol, 
because he will receive me. \textit{Selah}}%
\verse{Do not fear when a man becomes rich, 
when the wealth\lebnote{Or “honor”}of his house increases,}%
\verse{because when he dies he will not take away any of it. 
His wealth will not follow down after him.}%
\verse{Though he congratulated his soul while he was living 
—and people will praise you when you do well for yourself—}%
\verse{it\lebnote{That is, his soul}will go to the generation of his fathers. 
Never will they see light.}%
\verse{Humankind in its pomp, but does not understand, 
is like the beasts that perish.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 50
\verseWithHeading{A psalm of Asaph.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{The Supreme God, God, Adonai, has spoken 
and summoned the earth, 
from the rising of the sun 
to its setting.}%
\verse{From Zion, the perfection of beauty, 
God shines forth.}%
\verse{Our God comes and he is not silent. 
Before him fire devours, 
and around him it is very tempestuous.}%
\verse{He summons the heavens above 
and the earth that he might judge his people:}%
\verse{“Gather to me my loyal ones, 
those who have made a covenant with me by sacrifice.”}%
\verse{And the heavens declare his righteousness, 
because God himself is judge. \textit{Selah}}%
\verse{“Hear, O my people, and I will speak, 
O Israel, and I will testify against you. 
I am God, your God.}%
\verse{It is not concerning a lack of your sacrifices that I rebuke you, 
and your burnt offerings are before me continually.}%
\verse{I will not take from your house a bull 
or from your stalls a he-goat,}%
\verse{because every animal of the forest is mine, 
the cattle on a thousand hills.}%
\verse{I know every bird of the mountains, 
and every moving creature in the field is mine.}%
\verse{If I were hungry I would not tell you, 
because the world and its fullness are mine.}%
\verse{Do I eat the flesh of bulls 
or drink the blood of goats?}%
\verse{Offer to God a thank offering 
and pay your vows to the Most High.}%
\verse{And call me in the day of trouble; 
I will deliver you, and you will glorify me.”}%
\verse{But to the wicked God says, 
“What right have you to recite my statutes 
and \textit{mention my covenant with your mouth},\lebnote{Literally “take my covenant on your mouth”}}%
\verse{while you yourself hate discipline, 
and cast my words behind you?}%
\verse{When you see a thief, then you are pleased with him, 
and your association is with adulterers.}%
\verse{You give your mouth free rein for evil, 
and you harness your tongue to deceit.}%
\verse{You sit and speak against your brother; 
you slander your mother’s son.}%
\verse{These things you have done, and I have been silent; 
You imagined that I was just like you. 
I will rebuke you and present an argument before your eyes.}%
\verse{Now consider this, you who forget God, 
lest I tear you apart, and there will be none to deliver.}%
\verse{He who sacrifices a thank offering honors me, 
and he who orders his way; 
I will show him the salvation of God.”}%
\end{biblechapter}

\begin{biblechapter} % Psalms 51
\verseWithHeading{For the music director. A psalm of David. 
When Nathan the prophet came to him, after he had gone in to Bathsheba.\lebnote{The Hebrew Bible counts the superscription as the first two verses of the psalm; the English verse number is reduced by two}}{Be gracious to me, O God, according to your loyal love. 
According to your abundant mercies, 
blot out my transgressions.}%
\verse{Wash me thoroughly from my iniquity, 
and from my sin cleanse me.}%
\verse{For \textit{I myself know}\lebnote{Literally “I, I know”}my transgressions,\lebnote{Or “rebellions”}
and my sin is ever before me.}%
\verse{Against you, only you, I have sinned 
and have done this evil\lebnote{Hebrew “the evil”}in your eyes, 
so that you are correct when you speak, 
you are blameless when you judge.}%
\verse{Behold, in iniquity I was born, 
and in sin my mother conceived me.}%
\verse{Behold, you delight in truth in the inward parts, 
and in the hidden parts you make me to know wisdom.}%
\verse{Purify me with hyssop, and I shall be clean. 
Wash me, and I shall be whiter than snow.}%
\verse{Make me hear joy and gladness; 
let the bones you have crushed rejoice.}%
\verse{Hide your face from my sins, 
and all my iniquities blot out.}%
\verse{Create a clean heart for me, O God, 
and renew a steadfast spirit \textit{within me}.\lebnote{Literally “in my inner parts”}}%
\verse{Do not cast me away from your presence, 
and do not take your Holy Spirit from me.}%
\verse{Restore to me the joy of your salvation, 
and with a willing spirit sustain me.}%
\verse{Then I will teach transgressors your ways, 
and sinners will turn back to you.}%
\verse{Deliver me from the guilt of bloodshed, O God, 
the God of my salvation; 
then my tongue will sing aloud of your righteousness.}%
\verse{O Lord, open my lips, 
and my mouth will proclaim your praise.}%
\verse{For you do not delight in sacrifice or I would give it. 
With a burnt offering you are not pleased.}%
\verse{The sacrifices of God are a broken spirit; 
A broken and contrite heart, 
O God, you will not despise.}%
\verse{Do good in your favor toward Zion. 
Build the walls of Jerusalem.}%
\verse{Then you will delight in righteous sacrifices, 
burnt offering and whole burnt offering. 
Then bulls will be offered on your altar.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 52
\verseWithHeading{For the music director. A maskil of David. 
When Doeg the Edomite came and informed Saul. 
And he said to him, “David has come to the house of Ahimelech.”\lebnote{The Hebrew Bible counts the superscription as the first two verses of the psalm; the English verse number is reduced by two}}{Why do you boast about evil, O mighty man? 
The loyal love of God endures \textit{continually}.\lebnote{Literally “all the day”}}%
\verse{Your tongue plans destruction, 
like a sharp razor, working deceit.}%
\verse{You love evil more than good, 
a lie more than speaking \textit{what is right}.\lebnote{Literally “righteousness”}\textit{Selah}}%
\verse{You love all devouring words, 
O deceitful tongue,}%
\verse{but God will pull you down forever. 
He will snatch you and tear you away from your tent, 
and he will uproot you from the land of the living. \textit{Selah}}%
\verse{And the righteous will see and fear, 
and will laugh at him, saying,}%
\verse{“Look, the man who would not make God his refuge, 
but he trusted in the greatness of his wealth; 
he took refuge in his destructiveness.”}%
\verse{But I am like an olive tree flourishing 
in the house\lebnote{Or “temple”}of God. 
I trust the loyal love of God forever and ever.}%
\verse{I will give thanks to you forever, because of what you have done, 
and I will wait on your name, because it is good, 
in the presence of your faithful ones.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 53
\verseWithHeading{For the music director, according to Mahalath. 
A maskil of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{The fool says in his heart, “There is no God.” 
They are corrupt and they have done abominable iniquity. 
There is none who does good.}%
\verse{God looks down from heaven upon the children of humankind 
to see whether there is one who has insight, 
one who seeks God.}%
\verse{\textit{All of them}\lebnote{Literally “The whole of it,” i.e. all of humankind}have turned back. 
They are altogether corrupt. 
There is none who does good; 
there is not even one.}%
\verse{Do not evildoers know, 
they who eat my people as though they were eating bread? 
They do not call on God.}%
\verse{There \textit{they are very fearful}\lebnote{Literally “they feared a fear”}
where no fear had been, 
because God has scattered the bones of him who encamps against you. 
You have put them to shame, because God has rejected them.}%
\verse{\textit{Oh, that from Zion}\lebnote{Literally “Who will give that from Zion”}would come salvation for Israel! 
When God returns the fortunes\lebnote{Or “captivity”}of his people, 
let Jacob rejoice, let Israel be glad.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 54
\verseWithHeading{For the music director, with stringed instruments. 
A maskil of David, 
when the Ziphites went and said to Saul, 
“Is not David hiding himself among us?”\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{O God, by your name save me, 
and by your power vindicate me.}%
\verse{O God, hear my prayer; 
heed the words of my mouth.}%
\verse{For foreigners have risen against me, 
and ruthless men seek my life. 
They have not set God before them. \textit{Selah}}%
\verse{See, God is my helper; 
The Lord is with those who sustain my life.\lebnote{Or “The Lord is the sustainer of my life.”}}%
\verse{He will repay\lebnote{According to the reading tradition (\textit{Qere})}my enemies for their\lebnote{Hebrew “the”}evil; 
in your faithfulness destroy them.}%
\verse{I will freely sacrifice to you; 
I will give thanks to your name, 
O Adonai, because it is good.}%
\verse{Because he has delivered me from all trouble, 
and my eye has looked with satisfaction on my enemies.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 55
\verseWithHeading{For the music director, with stringed instruments. 
A maskil of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Give ear, O God, to my prayer, 
and do not hide yourself from my plea.}%
\verse{Attend to me and answer me. 
I am restless in my lamenting and I groan,}%
\verse{because of the voice of an enemy, 
because of the oppression of the wicked, 
because they bring down evil on me, 
and in anger they hold a grudge against me.}%
\verse{My heart trembles within me, 
and deathly terrors fall on me.}%
\verse{Fear and trembling come on me, 
and horror overwhelms me.}%
\verse{So I say, “\textit{Oh, that}\lebnote{Literally “Who will give that”}I had wings like a dove. 
I would fly away and be at rest.}%
\verse{Look, I would flee far away. 
I would dwell in the wilderness. \textit{Selah}}%
\verse{I would hurry to my refuge 
from the raging wind and storm.”}%
\verse{Confuse, O Lord; divide their speech, 
because I see violence and strife in the city.}%
\verse{Day and night they go around it on its walls, 
and iniquity and trouble are in its midst.}%
\verse{Destruction is within it, 
and oppression and deceit 
do not depart from its public square.}%
\verse{For it is not an enemy that taunts me, 
or I could bear it. 
It is not one who hates me that magnifies himself over me, 
or I could hide myself from him.}%
\verse{But it is you, a man my equal, 
my friend and confidant.}%
\verse{We who would take sweet counsel together; 
in the house\lebnote{Or “temple”}of God 
we would walk with the throng.}%
\verse{Let death deceive them. May they descend to Sheol alive, 
because evil is in their home and \textit{heart}.\lebnote{Literally “inward part”}}%
\verse{As for me, I will call to God, 
and Adonai will save me.}%
\verse{\textit{Morning, noon and night}\lebnote{Literally “evening and morning and noon”}I will lament and groan loudly, 
and he will hear my voice.}%
\verse{He safely redeems my life from the battle against me,\lebnote{Or “my battle”}
because those standing against me are among many.}%
\verse{God will hear and answer them,\lebnote{Or “and humble them”}
he who is enthroned from of old, \textit{Selah} 
Because they do not change, 
and they do not fear God.}%
\verse{He has put forth his hands against \textit{his friends};\lebnote{Literally “those at peace with him”}
he has defiled his covenant.}%
\verse{The \textit{buttery words}\lebnote{Literally “butter”}of his mouth were smooth, 
but there was battle in his heart. 
His words were smoother than oil, 
but they were drawn swords.}%
\verse{Cast your burden on Adonai, 
and he will sustain you. 
He will never allow the righteous to be moved.\lebnote{Or “shaken”}}%
\verse{But you, O God, you will bring them down to the pit of corruption. 
The men of bloodshed and deceit will not live half their days, 
but I will trust you.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 56
\verseWithHeading{For the music director, according to The Silent Dove of Distant Lands. 
Of David. A miktam. 
When the Philistines seized him in Gath.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Be gracious to me, O God, because humankind has trampled me; 
fighting all the day he oppresses me.}%
\verse{My enemies\lebnote{Or “lurking foes”}trample all day, 
because many are attacking me proudly.}%
\verse{\textit{When}\lebnote{Literally “The day”}I fear, I trust you.}%
\verse{God, whose word I praise, 
God I trust; I do not fear. 
What can mere flesh do to me?}%
\verse{All day they twist my words; 
all their thoughts are against me for evil.}%
\verse{They attack,\lebnote{Or “they stir up strife”}they hide, they watch my \textit{steps},\lebnote{Literally “heels”}
as they lie in wait for my life.}%
\verse{Because of iniquity will they escape? 
In anger cast down the peoples, O God.}%
\verse{You have kept count of my wonderings. 
Put my tears in your bottle; 
are they not in your book?}%
\verse{Then my enemies will turn back \textit{when}\lebnote{Literally “in the day”}I call.\lebnote{i.e., call on God}
This I know because\lebnote{Or “that”}God is for me.}%
\verse{God, whose word I praise, 
Adonai, whose word I praise,}%
\verse{God I trust; I do not fear. 
What can mere humankind do to me?}%
\verse{My vows to you, O God, are binding upon me. 
I will pay thank offerings to you,}%
\verse{because you have delivered my soul from death. 
Have you not kept my feet from stumbling, 
that I may walk before God 
in the light of the living?}%
\end{biblechapter}

\begin{biblechapter} % Psalms 57
\verseWithHeading{For the music director, according to Do Not Destroy. 
Of David. A miktam. 
When he fled from Saul into the cave.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Be gracious to me, O God, be gracious to me, 
because in you my soul takes refuge. 
In the shadow of your wings I will take refuge 
until destruction passes by.}%
\verse{I will call to God Most High, 
to God who accomplishes things concerning me.}%
\verse{He will send from heaven and save me; 
he will reproach the one who tramples me. \textit{Selah} 
God will send his loyal love and his faithfulness.}%
\verse{My soul is among lions. 
I lie down among those who devour— 
the children\lebnote{Or “sons”}of humankind whose teeth are spears and arrows 
and whose tongues are sharp swords.}%
\verse{Be exalted above the heavens, O God. 
Let your glory be above all the earth.}%
\verse{They have set a net for my steps; 
my soul is bowed down. 
They have dug a pit before me; 
they have fallen into the midst of it. \textit{Selah}}%
\verse{My heart is steadfast, O God; 
My heart is steadfast. 
I will sing and give praise.}%
\verse{Awake, my glory; 
Awake, harp and lyre. 
I will awake the dawn.\lebnote{Or “at dawn”}}%
\verse{I will give you thanks among the peoples, O Lord; 
I will give you praise among the nations.}%
\verse{Because your loyal love is high to the heavens, 
and your faithfulness to the clouds.}%
\verse{Be exalted above the heavens, O God. 
Let your glory be above all the earth.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 58
\verseWithHeading{For the music director, according to Do not Destroy. 
Of David. A miktam.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Do you really speak what is right when silent?\lebnote{A slight emendation of the Hebrew text yields “what is right, O gods,” or “O mighty ones”}
Do you judge fairly the children of humankind?\lebnote{Or “Do you judge fairly, O sons of humankind?”}}%
\verse{No, in your heart you plan injustices; 
in the land\lebnote{Or “on the earth”}you weigh out the violence of your hands.}%
\verse{The wicked are estranged from the womb. 
They go astray from the belly, speaking lies.}%
\verse{Their venom is like snake venom; 
They are like a deaf viper that closes its ear}%
\verse{so that it does not hear the voice of charmers 
or the skilled caster of spells.}%
\verse{O God, break their teeth in their mouth. 
Break off the fangs of the young lions, O Adonai.}%
\verse{Let them run away like water that runs off. 
When he bends the bow, let his arrows be as though they were cut off.\lebnote{Or “as though they were withered stalks”}}%
\verse{Let them be like a snail that melts away as it goes; 
like the stillborn of woman that do not see the sun.}%
\verse{Before your pots can feel the heat of a thornbush, 
whether green or \textit{dry},\lebnote{Literally “burning”}he will sweep it away.\lebnote{The Hebrew of this verse is difficult to interpret}}%
\verse{The righteous will rejoice when he sees the vengeance; 
he will wash his feet in the blood of the wicked.}%
\verse{And people\lebnote{Hebrew “a person”}will say, “Surely there is a \textit{reward}\lebnote{Literally “fruit”}for the righteous. 
Surely there is a God who judges in the land.”\lebnote{Or “on the earth”}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 59
\verseWithHeading{For the music director, according to Do Not Destroy. 
Of David. A miktam. 
When Saul dispatched men and they watched the house to kill him.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Deliver me from my enemies, O my God. 
Protect\lebnote{The sense is that of making something inaccessibly high, like a fortress}me from those who rise up against me.}%
\verse{Deliver me from workers of iniquity, 
and from men of bloodshed save me.}%
\verse{For look, they lie in wait for my life.\lebnote{Hebrew “soul”}
The mighty attack\lebnote{Or “stir up strife”}against me, 
not because of my transgression or my sin, O Adonai.}%
\verse{Without guilt on my part they run and ready themselves. 
Awake to meet me and see.}%
\verse{And you, Adonai, God of hosts, are the God of Israel. 
Rouse yourself to punish all the nations. 
Do not be gracious to any who treacherously plot evil. \textit{Selah}}%
\verse{They return at evening; they howl like dogs\lebnote{Hebrew “dog”}
while they prowl the city.}%
\verse{Look, they pour out talk with their mouth. 
Swords are on their lips, 
for they say, “Who hears?”}%
\verse{But you, O Adonai, will laugh at them; 
you will mock all the nations.}%
\verse{O my strength,\lebnote{Reading with a few Hebrew manuscripts and the LXX, Targum and v. 17}I will watch for you, 
because you, O God, are my (high) stronghold.}%
\verse{My\lebnote{According to the reading tradition (\textit{Qere}), some Hebrew manuscripts and the LXX}God of loyal love will meet me; 
God will cause me to look in triumph on my enemies.}%
\verse{Do not kill them, lest my people forget. 
Make them to wander by your power, 
and bring them down, O Lord, our shield.}%
\verse{By the sin of their mouth and the words of their lips, 
even in their pride, let them be trapped, 
and for the curses\lebnote{Hebrew “curse”}and lies\lebnote{Hebrew “lie”}they proclaim.}%
\verse{Destroy in anger; destroy so they are no more, 
so that they may know that God is ruling in Jacob 
to the ends of the earth. \textit{Selah}}%
\verse{They return at evening; they howl like dogs\lebnote{Hebrew “dog”}
while they prowl the city.}%
\verse{As for them, they wander for food. 
If they are not satisfied, then they continue all night.\lebnote{Or “then they growl”}}%
\verse{But as for me, I will sing of your strength, 
and I will hail your loyal love in the morning, 
because you have been my high stronghold 
and a refuge in my \textit{time}\lebnote{Literally “day”}of trouble.}%
\verse{O my strength, to you I will give praise, 
because God is my high stronghold, 
my God of loyal love.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 60
\verseWithHeading{For the music director, according to Shushan Eduth.\lebnote{Perhaps “The Lily of Testimony”}
A miktam of David. To teach. 
When he fought Mesopotamia and Aram Zobah, and Joab returned and struck Edom in the Valley of Salt, twelve thousand persons.\lebnote{The Hebrew Bible counts the superscription as the first two verses of the psalm; the English verse number is reduced by two}}{O God, you have rejected us. You have broken us. 
You have been angry. Restore us!}%
\verse{You have made the land quake. You have split it open. 
Heal its fissures, because it totters.}%
\verse{You have shown your people hard things; 
You have given us wine that staggers.}%
\verse{You have rallied those who fear you round a banner 
\textit{out of bowshot},\lebnote{Literally “from the presence of the bow”}\textit{Selah}}%
\verse{so that your beloved ones may be rescued. 
Save by your right hand and answer us.}%
\verse{God has spoken in his holiness, 
“I will rejoice; 
I will divide up Shechem, 
and portion out the valley of Succoth.}%
\verse{Gilead is mine, and Manasseh is mine, 
and Ephraim is the \textit{helmet for}\lebnote{Literally “protection of”}my head; 
Judah is my scepter.\lebnote{Or “commander’s rod”}}%
\verse{Moab is my washing pot; 
over Edom, I will cast my sandal. 
On account of me, O Philistia, raise a shout.”}%
\verse{Who will bring me to the fortified city? 
Who will lead me to Edom?}%
\verse{Have not you yourself rejected us, O God, 
and not gone out with our armies, O God?}%
\verse{Give us help against the adversary, 
for the help of humankind is futile.}%
\verse{Through God we will do \textit{valiantly},\lebnote{Literally “might”}
and it is he who will tread down our enemies.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 61
\verseWithHeading{For the music director, upon stringed instruments.\lebnote{Reading with many Hebrew manuscripts and several translations}Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Hear, O God, my cry; 
attend to my prayer.}%
\verse{From the end of the earth\lebnote{Or “the far reaches of the land”}I call to you 
when my heart grows faint. 
You will lead me to a rock that is higher than I.}%
\verse{For you have been my refuge, 
a strong tower from the enemy.}%
\verse{Let me reside in your tent forever. 
Let me take refuge under the covering of your wings. \textit{Selah}}%
\verse{For you, O God, you have heard my vows. 
You have given me the heritage of those who fear your name.}%
\verse{You will add days to the days of the king, 
his years like those of \textit{many generations}.\lebnote{Literally “a generation and a generation”}}%
\verse{May he live\lebnote{Or “be enthroned”}forever in the presence of God. 
Appoint loyal love and faithfulness to preserve him.}%
\verse{So I will sing the praise of your name forever, 
that I may fulfill\lebnote{Or “pay”}my vows \textit{daily}.\lebnote{Literally “day, day”}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 62
\verseWithHeading{For the music director, on Jeduthun.\lebnote{One of David’s musicians (1 Chr 16:41)}
A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Only for God my soul waits in silence. 
From him is my salvation.}%
\verse{Only he is my rock and my salvation, 
my high stronghold; I shall not be greatly shaken.}%
\verse{How long will you attack a man? 
All of you will be shattered\lebnote{Or “All of you shatter him”}
like a leaning wall, a tottering fence.}%
\verse{They only\lebnote{This word appears first in the Hebrew verse.}plan to thrust him down from his high position; 
they are pleased with deception. 
With their mouths\lebnote{Hebrew “his mouth”}they bless, 
but \textit{inwardly}\lebnote{Literally “in the midst of them”}they curse. \textit{Selah}}%
\verse{Only for God wait in silence, O my soul, 
because my hope is from him.}%
\verse{Only he is my rock and my salvation, 
my high stronghold; I shall not be shaken.}%
\verse{On God rests my salvation and my glory; 
God\lebnote{Or “In God”}is my strong rock, my refuge.}%
\verse{Trust him at all times, O people; 
pour out before him your heart. 
God is a refuge for us. \textit{Selah}}%
\verse{Only a vapor\lebnote{Or “breath”}are men of low degree, 
a deception are men of high degree. 
\textit{Weighed}\lebnote{Literally “To go up”}in a balance, 
together they are lighter than a vapor.}%
\verse{Do not trust in extortion, 
and do not put vain confidence in robbery. 
If wealth increases, 
do not set your heart on it.}%
\verse{Once God has spoken; 
twice I have heard this: 
that strength belongs to God.}%
\verse{And to you belongs loyal love, O Lord, 
because you will render to each according to his work.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 63
\verseWithHeading{A psalm of David, when he was in the wilderness of Judah.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{O God, you are my God; 
I will seek you diligently. 
My soul thirsts for you; my flesh longs for you 
as in a dry and weary land without water.}%
\verse{Thus I have seen you in the sanctuary, 
beholding your strength and glory.}%
\verse{Because your loyal love is better than life, 
my lips will praise you.}%
\verse{So I will bless you while I live. 
I will lift up my \textit{hands}\lebnote{Hebrew “palms”}in your name.}%
\verse{My soul will be satisfied as with the best and richest food, 
and with joyful lips my mouth will praise.}%
\verse{When I remember you on my bed, 
I meditate on you in the night watches.}%
\verse{Because you have been my help, 
therefore\lebnote{Hebrew “and”}in the shadow of your wings I will sing for joy.}%
\verse{My soul clings to\lebnote{Or “follows hard after”}you; 
your right hand upholds me.}%
\verse{But as for them who seek my life for ruin, 
they will go into the depths of the earth.}%
\verse{They who would \textit{deliver him to the sword}\lebnote{Literally “pour him out to the hands of the sword”}
will become a portion for jackals.}%
\verse{However, the king will rejoice in God. 
Everyone who swears an oath by him will boast, 
because the mouth of the liars will be stopped.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 64
\verseWithHeading{For the music director. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Hear my voice, O God, in my lamenting. 
Preserve my life from the fear of the enemy.}%
\verse{Hide me from the secret plots\lebnote{Hebrew “plot”}of evildoers, 
from the unrest of those who do iniquity,}%
\verse{who sharpen their tongue like a sword. 
They bend their bow to shoot their arrows—bitter words,}%
\verse{to shoot from\lebnote{Hebrew “in”}their\lebnote{Hebrew “the”}hiding places at the blameless. 
Suddenly they shoot at him and do not fear.}%
\verse{They encourage themselves in the evil matter. 
They talk of secretly laying snares. 
They say, “Who will see them?”}%
\verse{\textit{They carefully plot}\lebnote{Literally “They search out”}injustices saying, 
“We have perfected \textit{the perfect plot}!”\lebnote{Literally “a plotted plot”}
Both the \textit{inner thought}\lebnote{Literally “inward part”}and heart of a man are deep.}%
\verse{But God will shoot them; 
with an arrow they will suddenly be wounded.}%
\verse{\lebnote{The Hebrew of this verse is very difficult}And they who would cause him\lebnote{That is, the psalmist}to stumble, 
their own tongue will be against them. 
All who gaze on them will shake.\lebnote{Perhaps a shaking of the head; a similar Hebrew form could be translated “will flee”}}%
\verse{And then\lebnote{Hebrew “and”}all men\lebnote{Hebrew “man”}will fear, 
and will declare the working of God, 
and will understand his doing.}%
\verse{The righteous will rejoice in Adonai and take refuge in him, 
and all the upright in heart will boast.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 65
\verseWithHeading{For the music director. A psalm of David. A song.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Praise awaits you, O God, in Zion, 
and to you the vow shall be fulfilled.}%
\verse{O you who hear prayer, 
to you all flesh will come.}%
\verse{Iniquities\lebnote{Hebrew “matters of iniquities”}prevail over me. 
As for our transgressions, you will \textit{forgive}\lebnote{Literally “provide atonement for”}them.}%
\verse{Blessed is one whom you choose and bring near, 
that he may abide in your courts. 
We will be satisfied with the goodness of your house, 
your holy temple.}%
\verse{By awesome deeds in righteousness you will answer us, 
O God of our salvation, 
you who are the confidence of all the ends of the earth 
and of the far-off seas,}%
\verse{who established the mountains by his strength, 
the one who is girded with might,}%
\verse{who stills the roar of the seas, 
the roar of their waves, 
and the commotion of the peoples,}%
\verse{so that\lebnote{Hebrew “and”}the inhabitants of the farthest reaches are in awe of your signs. 
You make the dawn and sunset sing for joy.}%
\verse{You care for\lebnote{Or “visit”}the land\lebnote{Or “earth”}and water it; 
you greatly enrich it. 
The stream of God is filled with waters. 
You provide their grain, 
for so you have established it.}%
\verse{You drench its furrows, 
penetrating its ridges. 
With rains you soften it; 
its growth you bless.}%
\verse{You crown the year with your \textit{bounty},\lebnote{Literally “goodness”}
and your wagon paths drip with \textit{richness}.\lebnote{Literally “fatness”}}%
\verse{They drop on the pastures of the wilderness, 
and the hills gird themselves with joy.}%
\verse{The pasturelands put on flocks, 
and the valleys clothe themselves with grain. 
They shout in triumph; 
they even sing.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 66
\verseWithHeading{For the music director. A song. A psalm.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Shout joyfully to God, all the earth.}%
\verse{Sing the glory of his name. 
\textit{Set forth his glorious praise}.\lebnote{Literally “set for glory—his praise”}}%
\verse{Say to God, “How awesome are your works! 
Because of the greatness of your strength, 
your enemies will cringe before you.}%
\verse{All the earth will bow in worship to you, 
and sing praise to you. 
They will sing the praise of your name.” \textit{Selah}}%
\verse{Come and \textit{consider}\lebnote{Literally “see”}the works of God; 
he is awesome in his dealings\lebnote{Hebrew “dealing”}with the children of humankind.}%
\verse{He turned the sea into dry ground; 
they passed through the river on foot. 
There we rejoiced in him.}%
\verse{He rules by his might forever; 
his eyes keep watch on the nations. 
Do not let the rebellious exalt themselves.\lebnote{According to the reading tradition (\textit{Qere})}\textit{Selah}}%
\verse{Bless our God, O peoples, 
and cause the sound of his praise to be heard,}%
\verse{the one who has kept our soul among the living, 
and has not allowed our foot to slip.}%
\verse{For you have tested us, O God; 
you have tried us as silver is tried.}%
\verse{You brought us into the net; 
you placed a \textit{heavy burden}\lebnote{Literally “oppression”}on our backs.}%
\verse{You let men\lebnote{Hebrew “man”}ride over our heads. 
We went through fire and through water, 
but you have brought us out to the place of abundance.\lebnote{A slightly different Hebrew word yields “a broad place”}}%
\verse{I will come into your house\lebnote{Or “temple”}with burnt offerings. 
I will pay to you my vows}%
\verse{that my lips uttered, 
and my mouth spoke in my distress.}%
\verse{Burnt offerings of fat animals I will offer to you, 
with the smoke of rams. 
I will do\lebnote{That is, “offer”}cattle with rams. \textit{Selah}}%
\verse{Come and hear, all you God-fearers, and I will tell 
what he has done for me.\lebnote{Hebrew “my soul”}}%
\verse{I called to him with my mouth, 
and he was extolled with my tongue.\lebnote{A slightly different Hebrew word yields “his exaltation was under my tongue”}}%
\verse{If I had considered evil in my heart, 
the Lord would not have listened.}%
\verse{However, God has listened; 
he has attended to the sound of my prayer.}%
\verse{Blessed be God, 
who has not turned aside my prayer, 
or his loyal love from me.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 67
\verseWithHeading{For the music director, with stringed instruments. A psalm. A song.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{May God be gracious to us and bless us. 
May he cause his face to shine toward us, \textit{Selah}}%
\verse{that your way may be known on the earth, 
your salvation among all nations.}%
\verse{Let the peoples praise you, O God; 
let all of the peoples praise you.}%
\verse{Let the nations be glad and sing for joy, 
because you judge the peoples with equity 
and guide the nations on the earth. \textit{Selah}}%
\verse{Let the peoples praise you, O God; 
let all the peoples praise you.}%
\verse{The earth has yielded its produce. 
God, our God, will bless us.}%
\verse{God will bless us, 
and all the ends of the earth will fear him.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 68
\verseWithHeading{For the music director. Of David. A psalm. A song.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{God will arise; his enemies will be scattered, 
and those who hate him will flee from before him.}%
\verse{As smoke is driven off, you will drive them off. 
As wax melts before fire, 
the wicked will perish before God.}%
\verse{But the righteous will be glad; they will exalt before God, 
and will rejoice with joy.}%
\verse{Sing to God, sing praises to his name. 
Lift up a song to the rider on the clouds—his name is Yah—\lebnote{A shortened form of “Adonai”}
and rejoice before him.}%
\verse{A father to orphans and a judge for widows 
is God in his holy habitation.}%
\verse{God settles the lonely in a home; 
he brings prisoners out into prosperity. 
But the rebellious abide in a barren land.}%
\verse{O God, when you went out before your people, 
when you marched through the desert, \textit{Selah}}%
\verse{the earth shook; the heavens also dripped rain 
at the presence of God who was at Sinai, 
at the presence of God, the God of Israel.}%
\verse{You caused abundant rain to fall, O God. 
When your inheritance was weary you revived it.}%
\verse{Your community dwelled in it. 
You provided in your goodness for the poor, O God.}%
\verse{The Lord gives the word. 
Great is the company of women who announce:}%
\verse{“The kings of armies flee, they flee, 
and she who remains at home divides the spoil.}%
\verse{Though you men lie down between the sheepfolds, 
you will be like the wings of a dove covered in silver 
and its pinions in yellow gold.”}%
\verse{When Shaddai scattered kings on it, 
on Zalmon it snowed.\lebnote{Or “you caused it to snow on”}}%
\verse{A mountain of God\lebnote{Or “A mighty mountain”}is the mountain of Bashan; 
a mountain of many peaks is the mountain of Bashan.}%
\verse{Why do you look with hostility, O many-peaked mountains? 
This\lebnote{Hebrew “The”}mountain God desires for his dwelling. 
Yes, Adonai will abide in it forever.}%
\verse{The chariots of God 
are twice ten thousand, with thousands doubled. 
The Lord is among them at Sinai, distinctive in victory.\lebnote{Or “as at Sinai, in the sanctuary”}}%
\verse{You have ascended on high; you have led away captives. 
You have received gifts from among humankind, 
and even from the rebellious, so that Yah\lebnote{A shortened form of “Adonai”}God may dwell there.}%
\verse{Blessed be the Lord. Daily he loads us with benefits,\lebnote{Or “daily carries our burden”}
the God of our salvation.\lebnote{Or “victory”}\textit{Selah}}%
\verse{Our God is a God of deliverances, 
and to the Adonai the Lord belong escapes from death.}%
\verse{But God will shatter the head of his enemies, 
the hairy scalp of the one who \textit{continues on}\lebnote{Literally “is walking about”}in his guilt.}%
\verse{The Lord said, “I will bring them back from Bashan; 
I will bring them back from the depths of the sea,}%
\verse{so that your foot may shatter them bloody 
for the tongues of your dogs to have their share from the enemies.”}%
\verse{They have seen your processions, O God, 
the processions of my God, my king, distinctive in victory.\lebnote{Or “into the sanctuary”}}%
\verse{Singers went up front, those playing stringed instruments last, 
between them young women playing tambourines.}%
\verse{Bless God in the assemblies, 
Adonai from the fountain of Israel.}%
\verse{There is little Benjamin ruling them, 
with the princes of Judah in their throng, 
the princes of Zebulun and the princes of Naphtali.}%
\verse{Your God has commanded your strength. 
Show yourself strong, O God, by what you perform\lebnote{Or “you who have worked”}for us.}%
\verse{Because your temple is above Jerusalem, 
kings will bring tribute to you.}%
\verse{Rebuke the beasts in\lebnote{Hebrew “of”}the reeds, 
the herd of bulls with the calves of the peoples, 
trampling the pieces of silver.\lebnote{Or “until each humbles himself by offering pieces of silver”}
Scatter the nations who delight in battles.}%
\verse{Ambassadors\lebnote{Reading with the LXX; meaning uncertain}will come from Egypt; 
Cush will quickly stretch out her hands to God.}%
\verse{O kingdoms of the earth, sing to God; 
sing praise to the Lord, \textit{Selah}}%
\verse{to the one who rides in the highest heavens of old. 
See, he gives forth his voice, a mighty voice.}%
\verse{Ascribe strength to God. 
His majesty is over Israel, 
and his strength is in the clouds.}%
\verse{Awesome are you, O God, from your sanctuary.\lebnote{Hebrew “sanctuaries”}
The God of Israel, it is he who gives strength and might to the people. 
Blessed be God.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 69
\verseWithHeading{For the music director, according to The Lilies. Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Save me, O God, 
because waters have come up to my neck.\lebnote{Hebrew “soul” or “life”}}%
\verse{I sink in the mud of the deep, and there is no foothold. 
I have come to watery depths, 
and the torrent floods over me.}%
\verse{I am weary with my calling out; my throat is parched. 
My eyes are exhausted in my waiting for my God.}%
\verse{More numerous than the hairs of my head 
are those hating me without a cause. 
Those who are destroying me—my enemies wrongfully\lebnote{Or, “those who oppose me with falsehood”}— are mighty. 
What I did not steal, I then must restore.}%
\verse{O God, \textit{you yourself know}\lebnote{Literally “you, you know”}my foolishness, 
and my guilty deeds are not hidden from you.}%
\verse{Let those who wait for you not be put to shame because of me, 
O Lord Adonai of hosts. 
Let those who seek you not be disgraced because of me, 
O God of Israel.}%
\verse{Because on account of you I have borne reproach; 
disgrace has covered my face.}%
\verse{I have become a stranger to my brothers 
and a foreigner to my mother’s sons,}%
\verse{because the zeal for your house\lebnote{Or “temple”}has consumed me, 
and the reproaches of those reproaching you have fallen on me.}%
\verse{When I wept in the fasting of my soul, 
it became reproaches for me.}%
\verse{When I made sackcloth my clothing, 
I became for them a byword.}%
\verse{Those sitting at the gate talk about me 
as also\lebnote{Hebrew “and”}the songs of the drunkards.}%
\verse{But as for me, my prayer is to you, O Adonai, for a favorable time, 
O God, according to the abundance of your loyal love. 
Answer me with the faithfulness of your salvation.}%
\verse{Deliver me from the mud and do not let me sink. 
Let me be delivered from those who hate me 
and from the watery depths.}%
\verse{Do not let the torrent of waters flood over me, 
or the deep swallow me, 
or the pit close its mouth over me.}%
\verse{Answer me, O Adonai, because your loyal love is good; 
according to your abundant mercies, turn to me,}%
\verse{and do not hide your face from your servant. 
Because I am in trouble, answer me quickly.}%
\verse{Draw near to my soul; redeem it. 
Because of my enemies, ransom me.}%
\verse{You know my reproach, my shame and my disgrace. 
\textit{Fully known}\lebnote{Literally “right in front of”}to you are all my adversaries.}%
\verse{Reproach has broken my heart and I am sick. 
And I hoped for sympathy, but there was none, 
and for comforters, but I found none.}%
\verse{They also gave me gall for food,\lebnote{Or “they put poison in my food”}
and for my thirst they gave me vinegar to drink.}%
\verse{Let their table before them be a trap, 
and their times of peace a snare.}%
\verse{Let their eyes be dark so they cannot see, 
and make their loins continually tremble.}%
\verse{Pour out your indignation on them, 
and let your burning anger overtake them.}%
\verse{Let their camp be desolate. 
Let none dwell in their tents,}%
\verse{because they persecute those whom you, yourself, have struck, 
and they tell of the pain of those you have wounded.}%
\verse{\textit{Add guilt on top of their guilt},\lebnote{Literally “Give guilt on their guilt”}
and do not let them \textit{be acquitted}.\lebnote{Literally “come into your righteousness”}}%
\verse{Let them be blotted out of the book of the living, 
and let them not be recorded with the righteous.}%
\verse{But as for me, though I am afflicted and pained, 
your salvation will protect\lebnote{The sense is that of making something inaccessibly high, like a fortress}me, O God.}%
\verse{I will praise the name of God in song, 
and magnify him with thanksgiving.}%
\verse{For Adonai it will be better than an ox or bull, 
horned and hoofed.\lebnote{That is, mature and ritually clean}}%
\verse{The afflicted will see and rejoice. 
O God seekers, let your heart revive,\lebnote{Or “live”}}%
\verse{because Adonai hears the needy 
and does not despise his own who are prisoners.}%
\verse{Let heavens and earth praise him, 
the seas and all that moves in them,}%
\verse{because God will save Zion 
and build the cities of Judah, 
that they may dwell there and possess it.}%
\verse{And the offspring of his servants will inherit it, 
and those who love his name will abide in it.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 70
\verseWithHeading{For the music director. Of David. To bring to remembrance.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{O God, make haste to deliver me. 
O Adonai, make haste to help me.}%
\verse{Let them be ashamed and abashed 
who seek my life. 
Let them be repulsed and humiliated 
who desire my harm.}%
\verse{let them turn back because of their shame, 
those who say, “Aha! Aha!”}%
\verse{Let them rejoice and be glad in you, 
all those who seek you. 
Let them say continually, “God is great!” 
those who love your salvation.}%
\verse{But I am poor and needy; 
O God hasten to me. 
You are my help and my deliverer. 
O Adonai, do not delay.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 71
\verseWithHeading{A Prayer to God the Rock of Refuge}{In you, O Adonai, I have taken refuge; 
let me never be put to shame.}%
\verse{In your righteousness deliver me and rescue me; 
incline your ear to me and save me.}%
\verse{Be for me a rock of refuge to resort to always; 
you have ordained\lebnote{Or “commanded”}to save me, 
because you are my rock and my fortress.}%
\verse{My God, rescue me from the hand of the wicked, 
from the \textit{grasp}\lebnote{Literally “palm”}of the evildoer and the oppressor,}%
\verse{because you are my hope, 
O Lord Adonai, my confidence from my youth.}%
\verse{Upon you I have leaned \textit{from birth}.\lebnote{Literally “from the belly”}
It was you who \textit{took}\lebnote{Literally “cut,” that is, the umbilical cord}me from the womb of my mother. 
My praise is of you continually.}%
\verse{I have become a wonder to many, 
but you are my strong refuge.}%
\verse{My mouth is filled with your praise, 
with your glory all the day.}%
\verse{Do not cast me away in the time of old age; 
when my strength fails do not abandon me.}%
\verse{For my enemies talk concerning me, 
and those who watch for my life conspire together,}%
\verse{saying, “His God has abandoned him. 
Pursue and seize him, because there is no deliverer.”}%
\verse{O God, do not be far from me. 
My God, hurry to help me.}%
\verse{Let them be ashamed; let them perish 
who are \textit{my adversaries}.\lebnote{Literally “adversaries of my soul”}
Let them wrap themselves with scorn and disgrace 
who seek my harm.}%
\verse{But as for me, I will hope continually 
and increase your praise.}%
\verse{My mouth will tell of your righteousness, 
your salvation \textit{all day long},\lebnote{Literally “all the day”}
though I do not know the full sum of them.}%
\verse{I will come in to tell the mighty deeds of Lord Adonai. 
I will \textit{make known}\lebnote{Literally “bring to remembrance”}your righteousness, yours only.}%
\verse{O God, you have taught me from my youth, 
and up to now I have proclaimed your wonderful deeds.}%
\verse{And even when I am old and gray, 
O God, do not abandon me 
until I proclaim your \textit{strength}\lebnote{Literally “arm”}to this generation, 
your power to every one that comes after.}%
\verse{And your righteousness, O God, is to the height of heaven. 
You who have done great things, 
O God, who is like you?}%
\verse{You who have caused me\lebnote{According to the reading tradition (\textit{Qere})}to see many troubles and evils, 
you will again revive me.\lebnote{According to the reading tradition (\textit{Qere})}
And from the depths of the earth 
you will again bring me up.}%
\verse{You will increase my greatness, 
and you will comfort me all around.\lebnote{A slightly different vocalization yields, “… comfort me again”; compare the LXX.}}%
\verse{On my part, I will praise you with a stringed instrument, 
and your faithfulness, O my God. 
I will sing praises to you with a lyre, 
O Holy One of Israel.}%
\verse{My lips will sing for joy when I sing praises to you, 
and my soul, which you have redeemed.}%
\verse{My tongue also 
will speak of your righteousness all the day, 
because they have been put to shame, because they have been humiliated 
who seek my harm.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 72
\verseWithHeading{Of Solomon.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{O God, give your judgments to the king, 
and your righteousness to the king’s son.}%
\verse{May he judge your people with righteousness, 
and your poor with justice.}%
\verse{Let the mountains yield prosperity\lebnote{Or “peace”}for the people, 
and the hills in righteousness.}%
\verse{May he provide justice for the poor of the people, 
save the children of the needy, 
and crush the oppressor.}%
\verse{May he live long\lebnote{This presupposes the reading, suggested by the LXX, that might indicate that two of the Hebrew root letters have been inverted. The Hebrew reads “They will/Let them fear”}\textit{while the sun endures}\lebnote{Literally “with the sun”}
\textit{as long as the moon}\lebnote{Literally “in the presence of the moon”}\textit{for all generations}.\lebnote{Literally “generation of generations”}}%
\verse{May he descend like rain on mown grass, 
like showers watering the earth.}%
\verse{May what is right blossom in his days 
and an abundance of peace, until the moon is no more.}%
\verse{And may he rule from sea up to sea, 
and from the River\lebnote{That is, the Euphrates}to the edges of the land.}%
\verse{Let the desert dwellers bow down before him, 
and his enemies lick the dust.}%
\verse{Let the kings of Tarshish and the islands bring tribute. 
Let the kings of Sheba and Seba present gifts,}%
\verse{and let all kings bow down to him. 
Let all nations serve him.}%
\verse{Indeed he will deliver the needy who is crying for help, 
and the afflicted who has no helper.}%
\verse{He will take pity on the helpless and needy, 
and the lives of the needy he will save.}%
\verse{From oppression and from violence 
he will redeem their lives, 
and their blood will be precious in his eyes.}%
\verse{So may he live, and may gold from Sheba be given to him, 
and may prayers be offered for him continually. 
May blessings be invoked for him \textit{all day long}.\lebnote{Literally “all the day”}}%
\verse{May there be an abundance of grain in the land 
even on the top of the mountains. 
May his crop\lebnote{Or “May his descendants”}sway like the trees of Lebanon, 
and may those from the city blossom like the grass of the earth.}%
\verse{May his name endure forever. 
May his name increase \textit{as long as the sun shines},\lebnote{Literally “in the presence of the sun”}
and let them be blessed in him. 
Let all nations call him blessed.}%
\verse{Blessed be Adonai, the God of Israel, 
who alone does wonderful things.}%
\verse{And blessed be his glorious name forever, 
and may the whole earth be filled with his glory. 
Amen and Amen.}%
\verse{The prayers of David, the son of Jesse, are completed.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 73
\verseWithHeading{A song of Asaph.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Surely God is good to Israel, 
to those pure of heart.}%
\verse{But as for me, my feet had almost stumbled. 
My steps had nearly slipped,}%
\verse{because I envied the boastful 
when I saw the well-being\lebnote{Or “peace” or “prosperity”}of the wicked.}%
\verse{For there are no pains up to their death, 
and their bodies are \textit{healthy}.\lebnote{Literally “fat”}}%
\verse{They do not have \textit{ordinary trouble},\lebnote{Literally “the trouble of man”}
and they are not plagued as other people.\lebnote{Hebrew “humankind”}}%
\verse{Therefore pride is their necklace; 
an outfit of violence covers them.}%
\verse{Their eye bulges from fat.\lebnote{A slightly different Hebrew reading might be reflected in the LXX’s “their iniquity comes out of fatness”}
Imaginings overflow their heart.}%
\verse{They mock and speak maliciously of oppression; 
they speak as though from on high.}%
\verse{They set their mouth against the heavens, 
and their tongue roams the earth.}%
\verse{Therefore his people turn there,\lebnote{That is, toward the wicked}
and abundant waters are slurped up by them.}%
\verse{And they say, “How does God know?” 
and, “Does the Most High have knowledge?”}%
\verse{See, these are the wicked, 
and they increase wealth, ever carefree.}%
\verse{Surely in vain I have kept my heart pure, 
and washed my hands in innocence.}%
\verse{And I have been plagued all day 
and rebuked \textit{every morning}.\lebnote{Literally “to the mornings”}}%
\verse{If I had said, “I will speak thus,” 
behold, I would have acted treacherously 
against your children’s generation.}%
\verse{When I thought about how to understand this, 
it was troubling in my eyes}%
\verse{until I went into the sanctuary of God. 
Then I understood their fate.}%
\verse{Surely you set them on slippery places. 
You cause them to fall onto ruin.\lebnote{Or “fall for deceptions”}}%
\verse{How they become a desolation in a moment! 
They come to a complete end by terrors.}%
\verse{Like a dream upon awakening, 
when you wake up, O Lord, 
you will despise their fleeting form.\lebnote{Or “image”}}%
\verse{When my heart was embittered 
and I felt stabbed in my kidneys,}%
\verse{then I was brutish and ignorant. 
With you I was like the beasts.}%
\verse{But I am continually with you; 
you have hold of my right hand.}%
\verse{You will guide me with your advice, 
and afterward you will take me into honor.\lebnote{Or “to glory”}}%
\verse{Whom do I have in the heavens except you? 
And with you I have no other desire on earth.}%
\verse{My flesh and heart failed, 
but God is the \textit{strength}\lebnote{Literally “rock”}of my heart and my reward forever.}%
\verse{For indeed, those distancing themselves from you will be ruined. 
You destroy each who abandons you for harlotry.}%
\verse{But as for me, the approach to God is for my good. 
I have set the Lord Adonai as my refuge, 
in order to tell all your works.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 74
\verseWithHeading{A maskil of Asaph.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Why, O God, have you rejected us forever? 
Why does your anger smoke 
against the sheep of your pasture?}%
\verse{Remember your congregation that you bought long ago, 
when you redeemed the tribe of your inheritance. 
Remember \textit{Mount Zion}\lebnote{Literally “the mountain of Zion”}where you have dwelt.}%
\verse{Lift your steps to the perpetual ruins, 
to all that the enemy has ruined in the sanctuary.}%
\verse{Your enemies have roared in the midst of your meeting place; 
they have set up their signs for signs.}%
\verse{They are\lebnote{Hebrew “He is”}known to be like those who \textit{wield}\lebnote{Literally “lift up”}
axes in a thicket of trees.}%
\verse{And now\lebnote{According to the reading tradition (\textit{Qere})}its\lebnote{That is, the temple’s}carved works altogether 
they have smashed with axe and hammer.}%
\verse{They have set fire to your sanctuary. 
They have defiled to the ground, 
the dwelling place for your name.}%
\verse{They have said in their heart, 
“We will completely\lebnote{Hebrew “altogether”}oppress them.” 
They burned all the meeting places of God in the land.}%
\verse{We do not see our signs, 
and there is no longer a prophet. 
No one with us knows how long.}%
\verse{How long, O God, will the adversary taunt? 
Will the enemy treat your name with contempt forever?}%
\verse{Why do you draw back your hand, even your right hand? 
Take it from your bosom;\lebnote{According to the reading tradition (\textit{Qere})}destroy them!}%
\verse{But God has been my king from long ago, 
working salvation in the midst of the earth.}%
\verse{You split open the sea by your strength; 
You broke the heads of the sea monsters in the waters.}%
\verse{You crushed the heads of Leviathan; 
you gave him as food to the desert dwelling creatures.}%
\verse{You split open spring and wadi.\lebnote{A seasonal stream that is often dry}
You dried up ever-flowing rivers.}%
\verse{Yours is the day, yours is the night also. 
You established light and the sun.}%
\verse{You defined\lebnote{Or “set”}all the boundaries of the earth; 
Summer and winter—you formed them.}%
\verse{O Adonai, remember this: the enemy taunts, 
and foolish people treat your name with contempt.}%
\verse{Do not give to beasts the life of your dove; 
do not ever forget the life of your afflicted ones.}%
\verse{Have regard for the covenant, 
because the dark places of the land are full of 
the haunts of violence.}%
\verse{Do not let the oppressed turn back humiliated; 
let the poor and needy praise your name.}%
\verse{Rise up, O God, plead your cause; 
remember the reproaching of you by the foolish \textit{all day long}.\lebnote{Literally “all the day”}}%
\verse{Do not forget the sound of your adversaries, 
the roar of those rising up against you ascending continually.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 75
\verseWithHeading{For the music director, according to Do Not Destroy. 
A psalm of Asaph. A song.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{We give thanks to you, O God; 
we give thanks, and your name is near. 
Your wonderful deeds are told.\lebnote{Hebrew “They tell your wonderful deeds”}}%
\verse{“I will indeed set an appointed time; 
I will judge fairly.\lebnote{Hebrew “I, I will judge” might be emphatic}}%
\verse{The earth and all its inhabitants are shaking;\lebnote{Or “melting away”}
I steady\lebnote{Hebrew “I, I steady” might be emphatic}its columns. \textit{Selah}}%
\verse{I say to the boastful, ‘Do not boast!’ 
and to the wicked, ‘Do not lift up your horn!}%
\verse{Do not lift up high your\lebnote{Plural}horn. 
Do not speak with \textit{arrogant pride}.’ ”\lebnote{Hebrew “an unrestrained neck”}}%
\verse{For it is not from the east or the west 
and not from the south\lebnote{Hebrew “wilderness”}that lifting up comes,}%
\verse{rather God is the judge; 
one he brings low, and another he lifts up.}%
\verse{For there is a cup in the hand of Adonai 
with\lebnote{Hebrew “and”}wine that foams, fully mixed, 
and he pours out from this. 
Surely all the wicked of the land 
\textit{will quaff it down to its dregs}. \lebnote{Literally “they will drain, they will drink its dregs”}}%
\verse{But as for me, I will proclaim forever; 
I will sing praise to the God of Jacob.}%
\verse{“And all the horns of the wicked I will cut off. 
The horns of the righteous will be lifted up.”}%
\end{biblechapter}

\begin{biblechapter} % Psalms 76
\verseWithHeading{For the music director, with stringed instruments. 
A psalm of Asaph. A song.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{God is known in Judah. 
His name is great in Israel.}%
\verse{His den\lebnote{Or “shelter”; but the image of the lion appears in view}has been in Salem, 
his lair\lebnote{Or “refuge”}in Zion.}%
\verse{There he broke the flames of the bow, 
the shield and sword and battle. \textit{Selah}}%
\verse{Radiant\lebnote{A possible change in the Hebrew text yields, “You are feared, majestic …” a form found in vv. 8(7), 13(12). This is possibly reflected in the reading of Theodotian}you are, majestic, 
from\lebnote{Or “more than”}the mountains of prey.}%
\verse{The stouthearted are plundered; 
they sleep their sleep, 
and all the able men cannot \textit{use}\lebnote{Literally “find”}their hands.}%
\verse{At your rebuke, O God of Jacob, 
both rider\lebnote{Hebrew “chariot,” that is, charioteer}and horse slumber.}%
\verse{\textit{You, indeed, are to be feared},\lebnote{Literally “You, feared you are”}
and who can stand before you 
when you are angry? \lebnote{A slight change in the vocalization of the MT yields, “from the strength of your anger”}}%
\verse{From heaven you proclaimed judgment. 
The earth feared and was still}%
\verse{when God rose for justice, 
to save all the afflicted of the land. \textit{Selah}}%
\verse{For the anger of humankind will praise you. 
You will put on the remnant of anger.\lebnote{Hebrew “angers”}}%
\verse{Make vows and fulfill them to Adonai, your God; 
let all surrounding him 
bring tribute to the one to be feared.\lebnote{Hebrew “to the fear”}}%
\verse{He cuts off the spirit of leaders. 
He is to be feared by the kings of the earth.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 77
\verseWithHeading{For the music director, on Jeduthun.\lebnote{One of David’s musicians (1 Chr 16:41)}
Of Asaph. A psalm.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{I cry out with my voice to God; 
with my voice to God, that he may hear me.}%
\verse{In the day I have trouble, I seek\lebnote{Or “make supplication to”}the Lord. 
At night my hand stretches out \textit{continually};\lebnote{Literally “it does not grow weary”}
my soul refuses to be comforted.}%
\verse{I remember God and I groan loudly; 
I meditate and my spirit grows faint.}%
\verse{You hold open my eyelids. 
I am troubled and cannot speak.}%
\verse{I think about the days from long ago, 
the years of ancient times.}%
\verse{I remember my song in the night. 
With my heart I meditate, 
and my spirit searches to understand.}%
\verse{Will the Lord reject us forever, 
and will he never be pleased with us again?}%
\verse{Has his loyal love ceased forever? 
Is his promise\lebnote{Hebrew “word”}ended throughout generations?}%
\verse{Has God forgotten to have compassion? 
Or has he closed off his mercies in anger? \textit{Selah}}%
\verse{So I said, “This pierces me—\lebnote{Or “this is my sickness”}
the right hand of the Most High has changed.”}%
\verse{I will remember the deeds of Yah.\lebnote{A shortened form of “Adonai”}
Surely I will remember your wonders\lebnote{Hebrew “wonder”}from long ago.}%
\verse{I will also muse on all your work, 
and meditate on your deeds.}%
\verse{O God, your way is distinctive.\lebnote{Or “holy” or “set apart”}
Who is a great god like our God?}%
\verse{You are the God who works wonders;\lebnote{Hebrew “wonder”}
you have made known your might among the peoples.}%
\verse{With your arm you redeemed your people, 
the children of Jacob and Joseph. \textit{Selah}}%
\verse{Waters saw you, O God; 
waters saw you and they trembled. 
Surely the deeps shook.}%
\verse{The clouds poured out water. 
The skies \textit{thundered}.\lebnote{Literally “gave voice”}
Your arrows also \textit{flew about}.\lebnote{Literally “walked back and forth”}}%
\verse{The sound of your thunder was in the whirlwind;\lebnote{Or “in your chariot wheel”}
lightnings lit the world; 
the earth shook and quaked.}%
\verse{Your way was through the sea, 
and your path\lebnote{According to the reading tradition (\textit{Qere})}through many waters. 
Yet your footprints were not discerned.\lebnote{Hebrew “were not known”}}%
\verse{You led your people like a flock 
by the hand of Moses and Aaron.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 78
\verseWithHeading{A maskil of Asaph.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Listen, O my people, to my teaching. 
Incline your ears to the words of my mouth.}%
\verse{I will offer\lebnote{Hebrew “open”}a parable with my mouth. 
I will pour out riddles from long ago,}%
\verse{that we have heard and known, 
and our ancestors\lebnote{Or “fathers”}have told us.}%
\verse{We will not hide them from their children,\lebnote{Or “descendants”}
telling the next generation the praises of Adonai, 
and his power and his wonders that he has done.}%
\verse{For he established a testimony in Jacob, 
and appointed a law\lebnote{Hebrew \textit{torah}}in Israel, 
which he commanded our ancestors\lebnote{Or “fathers”}
to teach to their children,}%
\verse{so that the next generation might know— 
children yet to be born— 
that they might rise up and tell their children,}%
\verse{that they might set their confidence in God, 
and not forget the deeds of God, 
but keep his commandments,}%
\verse{and not be like their ancestors,\lebnote{Or “fathers”}
a stubborn and rebellious generation, 
a generation that did not make ready its heart, 
whose spirit was not faithful to God.}%
\verse{The sons of Ephraim, \textit{armed with archers},\lebnote{Literally “armed with shooters of the bow”}
turned back on the day of battle.}%
\verse{They did not keep the covenant of God 
and refused to go in his law.\lebnote{Hebrew \textit{torah}}}%
\verse{They also forgot his deeds, 
and his wonders that he had shown them.}%
\verse{In front of their ancestors\lebnote{Or “fathers”}he did a wonder, 
in the land of Egypt, in the region of Zoan.}%
\verse{He split the sea and caused them to go over, 
and he caused waters to stand like a heap.}%
\verse{And he led them with the cloud by day, 
and all night with a fiery light.}%
\verse{He caused rocks to split in the wilderness 
and provided drink abundantly as from the depths.}%
\verse{And he brought streams out of the rock 
and caused water to flow down like rivers.}%
\verse{But they sinned still further against him 
by rebelling against the Most High in the desert.}%
\verse{And they tested God in their heart 
by asking food \textit{for their craving}.\lebnote{Literally “for their soul”}}%
\verse{And they spoke against God. 
They said, “Is God able 
to prepare a table in the wilderness?}%
\verse{Yes, he struck the rock and water flowed 
and streams gushed out, 
but can he also give food 
or provide meat for his people?”}%
\verse{Therefore Adonai heard 
and he was very angry, 
and a fire was kindled against Jacob, 
and his anger also rose up against Israel,}%
\verse{because they did not believe God, 
and they did not trust his salvation.}%
\verse{Nevertheless, he commanded the skies above 
and opened the doors of heaven,}%
\verse{and rained down on them manna to eat 
and gave them the grain of heaven.}%
\verse{Humankind ate the bread of \textit{angels}.\lebnote{Literally “mighty ones”}
He sent them food enough to be satisfied.}%
\verse{He caused the east wind to blow in the heavens 
and drove along the south wind by his strength.}%
\verse{Then he rained meat on them like dust, 
even winged birds\lebnote{Hebrew “bird”}like the sand of the seas.}%
\verse{He caused them to fall in the midst of his camp, 
all around his dwellings.}%
\verse{So they ate and were well filled, 
and he brought about what they craved.}%
\verse{They had not yet turned aside from their craving, 
while their food was still in their mouth,}%
\verse{the anger of God rose against them, 
and he killed some of the stoutest of them, 
even the young men of Israel he caused to bow down in death.}%
\verse{In spite of all this they sinned further 
and did not believe his wonders.}%
\verse{And he consumed their days with futility \lebnote{Or “he ended their days like a breath”}
their years with terror.\lebnote{Or “and their years suddenly”}}%
\verse{When he killed some of them, then they sought him, 
and repented and earnestly sought God.}%
\verse{And they remembered that God was their rock, 
and God Most High their redeemer.}%
\verse{But they enticed him with their mouth 
and lied to him with their tongue.}%
\verse{For their heart was not steadfast with him, 
nor were they faithful to his covenant.}%
\verse{But he was compassionate; he pardoned\lebnote{Or “covered,” or “atoned for”}their guilt 
and did not destroy them. 
And many times he turned back his anger 
and did not stir up all his wrath,}%
\verse{for he remembered that they were flesh, 
a passing wind that does not return.}%
\verse{How often they rebelled against him in the wilderness 
and vexed him in the wasteland!}%
\verse{So they again tested God 
and distressed\lebnote{Or perhaps, “pressed to the limit”}the Holy One of Israel.}%
\verse{And they did not remember his \textit{power}\lebnote{Literally “hand”}
\textit{when}\lebnote{Literally “the day that”}he redeemed them from the enemy,}%
\verse{how he performed\lebnote{Hebrew “set”}his signs in Egypt 
and his wonders in the region of Zoan,}%
\verse{when he turned their rivers\lebnote{Normally the word for the Nile, with the plural here suggesting perhaps it and its canals}to blood 
so they could not drink from their streams.}%
\verse{He sent among them flies that devoured them 
and frogs that destroyed them.}%
\verse{And he gave their crop to the locusts 
and their labor to the locust.\lebnote{Two different words translated “locusts”}}%
\verse{He destroyed their vines with hail 
and their sycamore trees with sleet.\lebnote{The word for “sleet” is only used here, and the translation is a guess based on context}}%
\verse{He also handed their cattle over to the hail 
and their livestock to the lightning bolts.}%
\verse{He sent against them his fierce anger, 
rage and indignation and trouble, 
a band of \textit{destroying}\lebnote{Literally “evil”}angels.}%
\verse{He cleared a path for his anger. 
He did not spare them\lebnote{Hebrew “their soul”}from death 
but handed their life over to the plague.}%
\verse{And he struck down all the firstborn in Egypt, 
the first of their virility in the tents of Ham.}%
\verse{Then he led out his people like sheep 
and guided them like a herd in the wilderness.}%
\verse{And he led them safely and they were not afraid, 
but the sea covered their enemies.}%
\verse{So he brought them to his holy territory,\lebnote{Hebrew “border”}
this mountain his right hand acquired.\lebnote{Or “had created”}}%
\verse{And he drove out nations before them 
and allocated them for an inheritance by boundary line, 
and settled the tribes of Israel in their tents.}%
\verse{But they tested and rebelled against God Most High 
and did not keep his statutes.}%
\verse{And they turned and were treacherous like their ancestors.\lebnote{Or “fathers”}
They twisted like a crooked\lebnote{Or “deceitful”}bow.}%
\verse{For they provoked him to anger with their high places, 
and made him jealous with their images.}%
\verse{God heard and he was very angry 
and rejected Israel utterly.}%
\verse{So he abandoned the dwelling place at Shiloh, 
the tent he had placed among humankind.}%
\verse{And he gave his strength into captivity 
and his splendor into the hand of the enemy.}%
\verse{He also handed his people over to the sword, 
and he was very angry with his inheritance.}%
\verse{Fire devoured his young men, 
and his young women\lebnote{Or “virgins”}were not praised.}%
\verse{His priests fell by the sword, 
and his widows did not weep.}%
\verse{Then the Lord awoke like one who had been asleep, 
awoke like a warrior who had been drunk with wine.\lebnote{Or “like a warrior shouting because of wine”}}%
\verse{And he beat back his enemies; 
he gave them over to perpetual scorn.}%
\verse{And he rejected the tent of Joseph, 
and did not chose the tribe of Ephraim,}%
\verse{but chose the tribe of Judah, 
\textit{Mount Zion}\lebnote{Literally “the mountain of Zion”}that he loved.}%
\verse{And he built his sanctuary like the heights, 
like the earth that he established forever.}%
\verse{And he chose David his servant 
and took him from the sheepfolds.}%
\verse{He brought him from following nursing ewes 
to shepherd Jacob, his people, 
and Israel, his inheritance.}%
\verse{And he shepherded them according to the integrity of his heart, 
and led them by the skillfulness of his hands.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 79
\verseWithHeading{A psalm of Asaph.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{O God, the nations have entered your inheritance; 
they have defiled your holy temple; 
they have reduced Jerusalem to ruins.}%
\verse{They have given the bodies of your servants 
as food for the birds of the heavens, 
the flesh of your faithful to the beasts of the earth.}%
\verse{They have poured out their blood like water 
all around Jerusalem, 
and there was none to bury them.}%
\verse{We have become a taunt to our neighbors, 
a derision and a scorn to those around us.}%
\verse{How long, O Adonai? Will you be angry forever? 
Will your jealousy burn like fire?}%
\verse{Pour out your anger on the nations 
that do not know you, 
and on the kingdoms 
that do not call on your name,}%
\verse{because they\lebnote{Hebrew “he”}have devoured Jacob 
and have laid waste his habitation.}%
\verse{Do not remember against us former iniquities; 
let your mercies meet us quickly 
because we are brought very low.}%
\verse{Help us, O God of our salvation, 
for the glory of your name; 
and deliver us and forgive\lebnote{Hebrew “make atonement for”}our sins 
for the sake of your name.}%
\verse{Why should the nations say, 
“Where is their God?” 
Let it\lebnote{That is, God’s name}be known among the nations before our eyes, 
by the avenging of the blood of your servants 
that was poured out.}%
\verse{Let the groaning of the prisoner come before you. 
According to the greatness of your \textit{power},\lebnote{Literally “arm”}
spare\lebnote{Or “leave a remnant of”}the children appointed to death.}%
\verse{And return to our neighbors sevenfold \textit{upon them}\lebnote{Hebrew “into their bosom”}
their taunts with which they taunted you, O Lord.}%
\verse{Then we, your people and the flock of your pasture, 
we will give thanks to you forever. 
\textit{Generation after generation}\lebnote{Literally “to a generation and a generation”}
we will tell of your praise.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 80
\verseWithHeading{For the music director, according to The Lilies. 
A testimony. Of Asaph. A psalm.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Give ear, O shepherd of Israel, 
who leads Joseph like a flock. 
Shine forth, you who sits enthroned above the cherubim.}%
\verse{Before Ephraim, Benjamin, and Manasseh, 
stir up your power 
and come for our salvation.}%
\verse{O God, restore us, 
and cause your face to shine that we may be saved.}%
\verse{O Adonai of hosts, 
\textit{how long will you be angry}\lebnote{Literally “until when your anger smoke”}
against the prayer of your people?}%
\verse{You have fed them the bread of tears; 
you have given them tears to drink in full measure.\lebnote{The Hebrew “threefold” or “a third of a measure” is difficult}}%
\verse{You have made us an object of strife to our neighbors, 
and our enemies mock among themselves.}%
\verse{O God of hosts, restore us 
and cause your face to shine that we may be saved.}%
\verse{You uprooted a vine from Egypt; 
you drove out the nations and planted it.}%
\verse{You prepared a place before it, 
and \textit{it took deep root}\lebnote{Literally “its root rooted”}and filled the land.}%
\verse{The mountains were covered with its shade, 
and the mighty cedars with its boughs.}%
\verse{It spread its branches to the sea 
and its shoots to the river.}%
\verse{Why have you broken down its walls, 
so that all who pass on the road pluck fruit from it?}%
\verse{Swine from the forests devour\lebnote{Hebrew “devours”}it 
and creatures of the field feed on it.}%
\verse{Please return, O God of hosts. 
Observe from heaven and see, 
and pay attention to this vine,}%
\verse{eventhe stalk that your right hand planted, 
and concerning the shoot\lebnote{Hebrew “son”}you strengthened for yourself.}%
\verse{It is burned with fire, cut down. 
They perish at the rebuke of your face.}%
\verse{Let your hand be on the man of your right hand, 
on the son of humankind whom you made strong for yourself.}%
\verse{Then we will not turn back from you. 
Restore us to life, and we will proclaim your name.}%
\verse{O Adonai of hosts, restore us; 
cause your face to shine that we may be saved.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 81
\verseWithHeading{For the music director; on the Gittith. Of Asaph.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Shout out to God our strength; 
shout joyfully to the God of Jacob.}%
\verse{Lift up a song and strike\lebnote{Hebrew “give”}the tambourine, 
the pleasant lyre, together with the harp.}%
\verse{Blow the horn at new moon, 
at full moon, for our feast day,}%
\verse{because it is a statute for Israel, 
an ordinance of the God of Jacob.}%
\verse{He made it a statute\lebnote{Hebrew “testimony” or “witness”}in Joseph 
when he went out against the land of Egypt, 
where I heard a language I did not know.\lebnote{The meaning of this line is uncertain. The above is based on perhaps a similar thought in Psa 114:1}}%
\verse{“I removed his shoulder from a burden. 
His hands were freed from the basket.}%
\verse{In this\lebnote{Hebrew “the”}trouble you called, and I rescued you. 
Within the secret place of thunder I answered you; 
I tested you at the waters of Meribah. \textit{Selah}}%
\verse{Hear, O my people, and I will admonish you; 
O Israel, if you would but listen to me.}%
\verse{There shall be no strange god among you, 
and you shall not bow down to a foreign god.}%
\verse{I am Adonai your God, 
who brought you up from the land of Egypt. 
Open wide your mouth and I will fill it.}%
\verse{But my people did not listen to my voice, 
and Israel did not yield to me.}%
\verse{So I let them\lebnote{Hebrew “him”}go in the stubbornness of their heart; 
they walked in their counsels.}%
\verse{Oh that my people would listen to me; 
that Israel would walk in my ways.}%
\verse{I would subdue their enemies quickly, 
and turn my hand against their adversaries.}%
\verse{Those who hate Adonai would cringe before him, 
and their fate\lebnote{Hebrew “time”}would be forever.}%
\verse{But he would feed him\lebnote{That is, Israel}from the choicest wheat,\lebnote{Hebrew “the fat of wheat”}
and I would satisfy you with honey from a rock.”}%
\end{biblechapter}

\begin{biblechapter} % Psalms 82
\verseWithHeading{A psalm of Asaph.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{God stands in the divine assembly;\lebnote{Or “assembly of \textit{El} ”}
he administers judgment in the midst of the gods.\lebnote{Hebrew \textit{elohim}}}%
\verse{“How long will you judge unjustly 
and \textit{show favoritism to the wicked}?\lebnote{Literally “lift up the faces of the wicked”}\textit{Selah}}%
\verse{Judge on behalf of the helpless and the orphan; 
provide justice to the afflicted and the poor.}%
\verse{Rescue the helpless and the needy; 
deliver them from the hand of the wicked.”}%
\verse{They do not know or consider.\lebnote{Or “understand”}
They go about in the darkness, 
so that all the foundations of the earth are shaken.}%
\verse{I have said,\lebnote{Hebrew “I, I have said”}“You are gods, 
and sons of the Most High, all of you.}%
\verse{However, you will die like men,\lebnote{Singular}
and you will fall like one of the princes.”}%
\verse{Rise up, O God, judge the earth, 
because you shall inherit\lebnote{Or “receive an inheritance”}all the nations.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 83
\verseWithHeading{A song. A psalm of Asaph.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{O God, do not rest silently. 
Do not keep silent or be still, O God.}%
\verse{For look, your enemies roar, 
and those who hate you have lifted their head.}%
\verse{They devise cunning schemes\lebnote{Hebrew “scheme”}against your people, 
and consult together against your protected ones.\lebnote{Or “your treasured ones”}}%
\verse{They say, “Come, let us annihilate them from being a nation, 
so that the name of Israel will be remembered no more.”}%
\verse{For they have consulted together \textit{with a unified purpose}.\lebnote{Literally “with a heart together”}
They have made a covenant against you:}%
\verse{the tents of Edom and the Ishmaelites, 
Moab and the Hagrites,}%
\verse{Gebal and Ammon and Amalek, 
Philistia, with the inhabitants of Tyre.}%
\verse{Assyria also has joined with them. 
They \textit{provide help}\lebnote{Literally “they have become an arm for”}to the children of Lot. \textit{Selah}}%
\verse{Do to them as you did with Midian, 
as with Sisera, as with Jabin at the wadi of Kishon.}%
\verse{They were destroyed at En-dor; 
they became dung for the ground.}%
\verse{Make their leaders like Oreb and Zeeb, 
and all their chiefs like Zebah and Zalmunna,}%
\verse{who said, “Let us take as our possession 
the pastures of God.”}%
\verse{O my God, make them like the tumbleweed, 
like the chaff before wind.}%
\verse{As fire burns a forest, 
and as a flame sets afire mountains,}%
\verse{so pursue them with your tempest 
and terrify them with your storm.}%
\verse{Fill their faces with shame, 
that they may seek your name, O Adonai.}%
\verse{Let them be ashamed and terrified forever, 
and let them be humiliated and perish}%
\verse{that they may know that you, 
whose name is Adonai, you alone, 
are the Most High over the whole earth.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 84
\verseWithHeading{For the music director; on the Gittith. 
Of the sons of Korah. A psalm.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{How lovely are your dwelling places, 
O Adonai of hosts!}%
\verse{My soul longs and even fails 
for the courtyards of Adonai. 
My heart and flesh sing for joy 
to the living God.}%
\verse{Even a bird finds a home, and a swallow a nest for herself, 
where she may lay her young, 
near your altars, O Adonai of hosts, 
my king and my God.}%
\verse{Blessed are those who dwell in your house;\lebnote{Or “temple”}
they can ever praise you. \textit{Selah}}%
\verse{Blessed is the man whose strength is in you; 
in their heart are the highways to Zion.}%
\verse{Passing through the Valley of Baca,\lebnote{Or “Valley of Weeping”}
they make it a spring. 
The early rain covers it with blessings\lebnote{A change in vocalization of the Hebrew yields “it with pools”}as well.}%
\verse{They go from strength to strength, 
until each appears before God in Zion.\lebnote{Or “the God of gods will be seen in Zion”}}%
\verse{O Adonai, God of hosts, hear my prayer; 
give ear, O God of Jacob. \textit{Selah}}%
\verse{Look at our shield, O God, 
and have regard for the face of your anointed one.}%
\verse{Because better is a day in your courtyards 
than a thousand elsewhere. 
I would rather be at the threshold of the house\lebnote{Or “temple”}of my God 
than to dwell in the tents of wickedness.}%
\verse{Because Adonai is a sun and a shield; 
Adonai gives grace and honor. 
He does not withhold good from those who walk blamelessly.}%
\verse{O Adonai of hosts, 
blessed is the man who trusts you.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 85
\verseWithHeading{For the music director. Of the sons of Korah. A psalm.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{O Adonai, you favored your land. 
You restored the fortunes\lebnote{Or “captivity”}of Jacob.}%
\verse{You took away the guilt\lebnote{Or “You forgave the iniquity”}of your people; 
you covered all their sin. \textit{Selah}}%
\verse{You withdrew all your wrath; 
you turned from your burning anger.}%
\verse{Restore us, O God of our salvation, 
and annul your vexation with us.}%
\verse{Will you be angry against us forever? 
Will you prolong your anger generation after generation?}%
\verse{Will you not again revive us, 
that your people might rejoice in you?}%
\verse{Show us, O Adonai, your loyal love, 
and grant us your salvation.}%
\verse{I will hear what God, Adonai, will speak, 
because he will speak peace 
to his people, even his faithful ones,\lebnote{Hebrew “and to his faithful”}
but let them not return to folly.}%
\verse{Surely his salvation is near for those who fear him, 
that glory may abide in our land.}%
\verse{Loyal love and faithfulness\lebnote{Or “truth”}will meet one another; 
righteousness and peace will kiss.}%
\verse{Faithfulness\lebnote{Or “Truth”}will sprout from the ground, 
and righteousness will look down from heaven.}%
\verse{Yes, Adonai will give what is good, 
and our land will give its produce.}%
\verse{Righteousness will go before him, 
and it will make his steps a pathway.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 86
\verseWithHeading{A prayer of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Incline, O Adonai, your ear and answer me, 
because I am poor and needy.}%
\verse{Watch over my life because I am faithful. 
You are my God; save your servant. 
I am the one who trusts you.}%
\verse{Be gracious to me, O Lord, 
because I call to you \textit{all day long}.\lebnote{Literally “all the day”}}%
\verse{Make glad the soul of your servant, 
because \textit{I desire you},\lebnote{Literally “I lift up my soul to you”}O Lord.}%
\verse{For you, O Lord, are good and forgiving, 
and abundant in loyal love\lebnote{Hebrew “and great with respect to loyal love”}for all who call to you.}%
\verse{Heed, O Adonai, my prayer, 
and attend to the voice of my supplications.}%
\verse{In the day of my trouble I call to you, 
because you answer me.}%
\verse{There is none like you among the gods, O Lord, 
and there are no works like yours.}%
\verse{All the nations that you have made will come 
and bow down\lebnote{Or “worship”}before you, O Lord, 
and glorify your name.}%
\verse{For you are great and doing wondrous things; 
you alone are God.}%
\verse{Teach me, O Adonai, your way, 
that I may walk in your truth. 
Unite my heart\lebnote{That is, give singleness of heart}to fear your name.}%
\verse{I will give you thanks, O Lord my God, with all my heart, 
and glorify your name forever,}%
\verse{because your loyal love is great toward me, 
and you will have delivered my life from Sheol\lebnote{The nether world, the place of departed spirits.}below.}%
\verse{O God, arrogant men have risen up against me, 
even a gang of ruthless men seek my life, 
but they do not set you before them.\lebnote{That is, they do not factor God into their actions}}%
\verse{But you, O Lord, are a compassionate and gracious God, 
\textit{slow to anger}\lebnote{Literally “long of nostrils”}and abundant in loyal love and faithfulness.}%
\verse{Turn to me and be gracious to me. 
Give your strength to your servant, 
and grant victory to the son\lebnote{Or “Save the son”}of your maidservant.}%
\verse{Do a sign that benefits me, 
that those who hate me may see and be put to shame, 
because you, O Adonai, have helped me and comforted me.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 87
\verseWithHeading{Of the sons of Korah. A psalm. A song.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{His foundation is on the holy mountains.}%
\verse{Adonai loves the gates of Zion 
more than all the dwellings of Jacob.}%
\verse{Glorious things are spoken about\lebnote{Or “in”}you, 
O city of God. \textit{Selah}}%
\verse{I will record\lebnote{Or “mention”}those who know me 
in Rahab\lebnote{That is, Egypt; compare Isa 30:7}and Babylon, 
behold in Philistia and Tyre with Cush, 
“This one was born there.”}%
\verse{But with respect to Zion it shall be said, 
“\textit{Each one}\lebnote{Literally “A man and a man”}was born in her,” 
for the Most High himself will establish her.}%
\verse{Adonai will record in writing the peoples, 
“This one was born there.” \textit{Selah}}%
\verse{And while dancing, singers will sing, 
“All my springs are in you.”}%
\end{biblechapter}

\begin{biblechapter} % Psalms 88
\verseWithHeading{A song. A psalm of the sons of Korah. 
For the music director; according to Mahalath Leannoth. 
A maskil of Heman the Ezrahite.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{O Adonai, God of my salvation, 
I cry out by day and through the night before you.}%
\verse{Let my prayer come before you; 
incline your ear to my cry.}%
\verse{For my soul is full with troubles, 
and my life approaches Sheol.}%
\verse{I am reckoned with those descending to the pit. 
I am like a man\lebnote{The Hebrew suggests a strong, vigorous man}without strength,\lebnote{Or “without help”}}%
\verse{set free\lebnote{Or “forsaken”}among the dead, 
like the slain lying in the grave, 
whom you no longer remember, 
even those cut off from your hand.}%
\verse{You have set me in the pit below, 
in dark places, in the depths.}%
\verse{Your wrath presses upon me, 
and you afflict me with all your waves. \textit{Selah}}%
\verse{You have removed my acquaintances far from me. 
You have made me detestable\lebnote{Hebrew “an abomination”}to them. 
I am confined and cannot go out.}%
\verse{My eye languishes from misery. 
I call on you, O Adonai, every day; 
I spread out my hands\lebnote{Hebrew “palms”}to you.}%
\verse{Do you work wonders\lebnote{Hebrew “wonder”}for the dead? 
Or do the departed spirits\lebnote{Hebrew \textit{rephaim}, sometimes translated “shades”; the weakened state of those who have died}rise up to praise you? \textit{Selah}}%
\verse{Is your loyal love told in the grave, 
or your faithfulness in the underworld?\lebnote{Hebrew \textit{abaddon}, the place of destruction}}%
\verse{Are your wonders\lebnote{Hebrew “wonder”}known in the darkness, 
or your righteousness\lebnote{Or “deed of justice”}in the land of forgetfulness?}%
\verse{But as for me, I cry for help to you, O Adonai, 
and in the morning my prayer comes before you.}%
\verse{Why do you reject my soul, O Adonai? 
Why do you hide your face from me?}%
\verse{I am afflicted and perishing from my youth. 
I bear your terrors. I am distraught.\lebnote{Hebrew meaning uncertain}}%
\verse{Your burning anger has passed over me; 
your sudden fears have destroyed me.}%
\verse{They surround me like water all the day; 
they circle about me altogether.}%
\verse{You have removed loved one and friend far from me, 
my acquaintances far from my darkness.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 89
\verseWithHeading{A maskil of Ethan the Ezrahite.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{I will sing forever of Adonai’s acts of loyal love. 
\textit{From generation to generation}\lebnote{Literally “To a generation and a generation”}
I will make known your faithfulness with my mouth.}%
\verse{For I say, “Forever your loyal love is built up. 
The heavens you have established with your faithfulness in them.”}%
\verse{“I made a covenant with my chosen one; 
I swore an oath to David my servant:}%
\verse{‘I will establish your descendants forever, 
and I will build up your throne \textit{from generation to generation}.’ ”\lebnote{Literally “To a generation and a generation”}\textit{Selah}}%
\verse{And so the heavens will praise your wonderful deed, O Adonai, 
even your faithfulness, in the assembly of the holy ones.}%
\verse{For who in the sky\lebnote{Hebrew singular collective “clouds”}is equal to Adonai? 
Who is like Adonai among the sons of God,}%
\verse{a God feared greatly in the council of the holy ones, 
and awesome above all surrounding him?}%
\verse{O Adonai of hosts, 
who is mighty like you, O Yah, 
with\lebnote{Hebrew “and”}your faithfulness surrounding you?}%
\verse{You are ruling the surging of the sea. 
When its waves rise, \textit{you yourself still}\lebnote{Literally “you, you calm”}them.}%
\verse{\textit{You yourself crushed}\lebnote{Literally “You, you crushed”}Rahab like one who is slain; 
you scattered your enemies with your strong arm.}%
\verse{The heavens are yours, the earth yours as well, 
the world with its fullness, because you founded them.}%
\verse{North and south, you created them; 
Tabor and Hermon shout joyfully over your name.}%
\verse{You have a mighty arm; 
Your hand is strong; your right hand reaches high.}%
\verse{Righteousness and justice are the foundation of your throne. 
Loyal love and faithfulness\lebnote{Or “truth”}come before your face.\lebnote{That is, enjoy continual access}}%
\verse{Blessed are the people who know the joyful shout; 
O Adonai, they walk about in the light of your countenance.}%
\verse{In your name they rejoice all the day, 
and in your righteousness they rise up.}%
\verse{For you are the glory of their strength, 
and by your favor our horn rises up,\lebnote{According to the reading tradition (\textit{Qere}). The \textit{Kethib} reads “you exalt our horn”}}%
\verse{because our shield belongs to Adonai,\lebnote{Or “Adonai is our shield.”}
and our king to the Holy One of Israel.\lebnote{Or “The Holy One of Israel is our king.”}}%
\verse{Formerly you spoke in a vision 
to your faithful ones, and said, 
“I have bestowed help on a mighty man; 
I have exalted one chosen from the people.}%
\verse{I have found David, my servant. 
With my holy oil I have anointed him,}%
\verse{with whom my hand will be steadfast. 
Surely my arm will strengthen him.}%
\verse{The enemy will not deceive him, 
and no \textit{evil man}\lebnote{Literally “a son of wickedness”}will afflict him.}%
\verse{But I will crush his adversaries before him, 
and I will strike those who hate him.}%
\verse{And so my faithfulness 
and my loyal love will be with him, 
and in my name his horn will rise up.}%
\verse{And I will set his hand on the sea 
and his right hand on the rivers.}%
\verse{He will call out to me, ‘You are my Father, 
my God, and the rock of my salvation.’}%
\verse{I will also make him the firstborn, 
the highest of the kings of the earth.}%
\verse{Forever I will keep my loyal love for him, 
and my covenant will be reliable for him.}%
\verse{I will also establish his offspring forever, 
and his throne as the days of the heavens.}%
\verse{If his sons forsake my law 
and do not walk in my judgments,}%
\verse{if they defile\lebnote{Hebrew “profane; treat as common”}my statutes 
and do not keep my commandments,}%
\verse{then I will punish their transgression with a rod, 
and their guilt with wounds.}%
\verse{But I will not break off my loyal love from him, 
and I will not be false against my faithfulness.}%
\verse{I will not defile\lebnote{Hebrew “profane; treat as common”}my covenant, 
or alter what proceeded from my lips.}%
\verse{Once\lebnote{Hebrew “One”}I have sworn by my holiness, 
‘I will surely not lie to David,}%
\verse{His offspring will endure\lebnote{Hebrew “be”}forever, 
and his throne as the sun before me.}%
\verse{Like the moon it will be steadfast forever, 
and like an enduring witness in the sky.’ ” \textit{Selah}}%
\verse{But you have spurned and rejected; 
you are very angry with your anointed one.}%
\verse{You have repudiated the covenant with your servant; 
you have defiled\lebnote{Hebrew “profane; treat as common”}his crown in\lebnote{Hebrew “to”}the dirt.}%
\verse{You have broken down all his walls; 
you have made his fortifications a ruin.}%
\verse{All passing along the road plunder him; 
he has become a taunt to his neighbors.}%
\verse{You have raised the right hand\lebnote{That is, given them victory}of his adversaries. 
You have made all his enemies rejoice.}%
\verse{You have also turned back the edge of his sword 
and have not helped him stand his ground in the battle.}%
\verse{You have made his splendor to cease, 
and cast his throne to the ground.}%
\verse{You have cut short the days of his youth; 
you have covered him with shame. \textit{Selah}}%
\verse{How long, O Adonai? Will you hide yourself forever? 
Will your wrath burn like fire forever?}%
\verse{Remember what my lifespan is. 
Remember for what vanity 
you have created all the children of humankind.}%
\verse{What man can live on and not see death? 
Can he deliver his life from the \textit{power}\lebnote{Literally “hand””}of Sheol? \textit{Selah}}%
\verse{Where are your former acts of loyal love, O Lord, 
that you swore to David by your faithfulness?}%
\verse{Remember, O Lord the taunting of your servants, 
how I bear in my bosom the taunts of all the many peoples,}%
\verse{by which your enemies taunt, O Adonai, 
by which they taunt the steps of your anointed one.}%
\verse{Blessed be Adonai forever. 
Amen and Amen.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 90
\verseWithHeading{A prayer of Moses, the man of God.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{O Lord, you have been our help\lebnote{Or “dwelling place”}\textit{in all generations}.\lebnote{Literally “in a generation and a generation”}}%
\verse{Before the mountains were born 
and you brought forth the earth and the world, 
even from everlasting to everlasting, you are God.}%
\verse{You return man to the dust, 
\textit{saying},\lebnote{Literally “and you say”}“Return, O sons of man.”}%
\verse{For a thousand years in your eyes 
are like yesterday when it passes, 
or like a watch in the night.}%
\verse{You sweep them away like a flood. 
They fall asleep.\lebnote{The interpretation of 5a is difficult}
In the morning they are like grass that sprouts anew.}%
\verse{In the morning it blossoms and sprouts anew; 
by evening it withers and dries up.}%
\verse{For we are brought to an end by your anger, 
and we hasten off\lebnote{Or “we are terrified”}by your wrath.}%
\verse{You have put our iniquities before you, 
our hidden sins into the light of your countenance.}%
\verse{For all of our days dwindle away in your rage; 
we complete our years like a sigh.}%
\verse{As for the days of our years, within them are seventy years 
or if by strength eighty years, and their pride\lebnote{Or “span”}is trouble and disaster, 
for it passes quickly and we fly away.}%
\verse{Who knows the strength of your anger, 
and your rage consistent with\lebnote{Hebrew “according to”}the fear due you?}%
\verse{So teach us to number our days 
that we may gain a heart of wisdom.}%
\verse{Return,\lebnote{Or “Turn back your wrath”}O Adonai. How long? 
And have compassion on\lebnote{Or “change your mind concerning”}your servants.}%
\verse{Satisfy us in the morning with your loyal love, 
that we may sing for joy and be glad all our days.}%
\verse{Make us glad for as many days as you have afflicted us, 
for as many years as we have seen calamity.}%
\verse{Let your work be visible to your servants, 
and your majesty to their children.}%
\verse{And let the beauty of the Lord our God be upon us, 
and establish for us the work of our hands, 
yes, the work of our hands, establish it.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 91
\verseWithHeading{God’s Protection in Times of Crisis}{One who lives in the secret place\lebnote{Or “shelter”}of the Most High 
will lodge in the shadow of Shaddai.\lebnote{Often translated “the Almighty,” this is perhaps better understood as “God of the Mountain”}}%
\verse{I will say to Adonai, “You are my refuge and my fortress, 
my God in whom I trust.”}%
\verse{For he will deliver\lebnote{Hebrew “he, he will deliver” is perhaps emphatic, “He will certainly deliver”}you from the snare of the fowler, 
from the plague of destruction.}%
\verse{With his feathers he will cover you, 
and under his wings you can take refuge. 
His faithfulness will be a shield and a buckler.\lebnote{That is, a small, round shield, or perhaps a surrounding bulwark}}%
\verse{You need not fear the terror of the night, 
or the arrow that flies by day,}%
\verse{or the plague that spreads in the darkness, 
or the destruction that devastates at noon.}%
\verse{A thousand may fall at your side, 
and ten thousand at your right hand, 
but it will not come near you.}%
\verse{You will only look with your eyes, 
and see the punishment of the wicked.}%
\verse{For you, O Adonai, are my refuge. 
You have made the Most High your dwelling place.\lebnote{This is the most straightforward translation of the Hebrew, though the “you” of each line would refer to a different person. A more unusual rendering would be “Because Adonai, my refuge, even the Most High, you have made your dwelling place …”}}%
\verse{No harm will befall you, 
and no plague will come near your tent.}%
\verse{For he will command his angels concerning you, 
to watch over you in all your ways.}%
\verse{In their hands they will bear you up, 
lest you strike your foot against a stone.}%
\verse{You will tread on lion and viper; 
you will trample young lion and serpent.}%
\verse{Because he loves me, therefore I will deliver him; 
I will protect\lebnote{The sense is that of making something inaccessibly high, like a fortress}him because he knows my name.}%
\verse{He will call upon me and I will answer him; 
I will be with him in trouble; 
I will rescue and honor him.}%
\verse{With \textit{long life}\lebnote{Literally “a length of days”}I will satisfy him, 
and show him my salvation.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 92
\verseWithHeading{A psalm. A song. For the Sabbath day.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{It is good to give thanks to Adonai, 
and to sing praise concerning your name, O Most High;}%
\verse{to declare in the morning your loyal love 
and your faithfulness in the night,}%
\verse{on the ten string, and on the harp, 
with a melody on the lyre.}%
\verse{For you, O Adonai, have made me glad by your work; 
by the deeds of your hands I sing for joy.}%
\verse{How great are your deeds, O Adonai; 
how very deep are your thoughts.}%
\verse{The brutish man does not know, 
and the fool cannot understand this.}%
\verse{When the wicked flourish like grass 
and all the workers of evil blossom, 
it is so they can be destroyed forever.}%
\verse{But you are on high forever, O Adonai.}%
\verse{For behold, your enemies, O Adonai, 
for behold, your enemies will perish. 
All the workers of evil will be scattered.}%
\verse{But you have exalted my horn like that of a wild ox; 
you have anointed me\lebnote{The Hebrew’s “that I have moistened with fresh oil” seems unlikely. The above translation presupposes the loss of a letter in the Hebrew textual tradition (\textit{ballotani} for the MT’s \textit{balloti})}with fresh oil.}%
\verse{And so my eye looks on my enemies.\lebnote{This word for “enemies” suggests those watching for any advantage}
My ears hear those evildoers who rise up against me.}%
\verse{The righteous\lebnote{Hebrew singular, but used as a collective. The verbs of this verse are also singular}will flourish like the date palm. 
They will grow like a cedar in Lebanon.}%
\verse{Planted in the house\lebnote{Or “temple”}of Adonai, 
they will flourish in the courts of our God.}%
\verse{They will still prosper in old age. 
They will be fat and luxuriant,\lebnote{This is the same word used of the oil in v. 11 and translated “fresh”}}%
\verse{to declare that Adonai is upright. 
He is my rock, and there is no injustice\lebnote{According to the reading tradition (\textit{Qere})}in him.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 93
\verseWithHeading{Adonai Is King Over All the Earth}{Adonai is king; he clothes himself with majesty. 
Adonai clothes himself; he girds himself with might. 
Yes, the world is established so that it will not be moved.}%
\verse{Your throne is established from of old; 
you are from everlasting.}%
\verse{The rivers have lifted up, O Adonai; 
the rivers have lifted up their \textit{rumbling};\lebnote{Literally “voice”}
the rivers have lifted up their pounding.}%
\verse{Mightier than the rumblings of many waters, 
mightier than the mighty breakers of the sea, 
Adonai on high is mighty.}%
\verse{Your testimonies are fully reliable. 
Holiness is fitting for your house,\lebnote{Or “temple”}
O Adonai, \textit{forever}.\lebnote{Literally “for a length of days”}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 94
\verseWithHeading{A Prayer for Retribution against Oppressors}{O Adonai, God of vengeance, 
God of vengeance, shine forth.}%
\verse{Rise up, O Judge of the earth. 
Repay upon the proud what is their rightful due.}%
\verse{How long will the wicked, O Adonai, 
how long will the wicked exult?}%
\verse{\textit{They gush words}\lebnote{Literally “They gush, they speak,” which is probably a hendiadys}unrestrained. 
All the evildoers boast.}%
\verse{They crush your people, O Adonai; 
they oppress your inheritance.}%
\verse{They kill widow and stranger, 
and they murder orphans,}%
\verse{while\lebnote{Hebrew “and”}they say, “Yah\lebnote{A shortened form of “Adonai”}does not see,” 
and “The God of Jacob does not pay attention.”}%
\verse{You pay attention, O brutes among the people. 
And you fools, when you will show insight?}%
\verse{Will the one who planted the ear not hear? 
Will the one who formed the eye not see?}%
\verse{Will the one who instructs nations not rebuke, 
the one who teaches humankind knowledge?}%
\verse{Adonai knows the thoughts of humankind, 
that \textit{they are to no purpose}.\lebnote{Literally “they are a breath” or “a vanity”}}%
\verse{Blessed is the man, O Yah, whom you instruct 
and teach from your law,}%
\verse{to give him rest from days of trouble, 
until a pit is dug for the wicked.}%
\verse{For Adonai will not abandon his people, 
nor forsake his inheritance.}%
\verse{For judgment will return to righteousness,\lebnote{That is, the court system will operate correctly}
and all the upright in heart will follow after it.}%
\verse{Who rose up for me against the wicked? 
Who stood up for me against the workers of iniquity?}%
\verse{If Adonai had not been my help, 
my soul would soon have dwelt in silence.}%
\verse{When I thought,\lebnote{Hebrew “If/When I said,” that is, said to myself}“My foot is slipping,” 
your loyal love, O Adonai, supported me.}%
\verse{When my troubled thoughts were many \textit{within me},\lebnote{Literally “in my inward parts”}
your consolations cheered my soul.}%
\verse{Can there be allied with you a throne of destruction, 
one that forms trouble based on statute?}%
\verse{They band together against the life of the righteous, 
and declare the blameless guilty of blood.\lebnote{Or “condemn innocent blood”}}%
\verse{But Adonai has become my high stronghold, 
and my God has become my rock of refuge.}%
\verse{And he will repay on them their iniquity, 
and by\lebnote{Or “in”}their evil he will destroy them. 
Adonai our God will destroy them.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 95
\verseWithHeading{A Call to Worship and Obey}{Come, let us sing for joy to Adonai; 
let us shout joyfully to the rock of our salvation.}%
\verse{Let us come before his presence with thanksgiving; 
with songs let us shout joyfully to him.}%
\verse{For Adonai is the\lebnote{The Hebrew does not have the definite article used twice in the translation of this verse}great God, 
and the great king over all gods,}%
\verse{in whose hand are the unexplored places\lebnote{Or “the deep places”}of the earth, 
and the heights of the mountains are his,}%
\verse{to whom belongs the sea that he made, \lebnote{Hebrew “and he made it”}
and the dry land that his hands formed.}%
\verse{Come in, let us worship and bow down; 
let us kneel before Adonai, our maker.}%
\verse{For he is our God, 
and we are the people of his pasture 
and the sheep of his hand. 
\textit{Today}\lebnote{Literally “The day”}if you will hear his voice:\lebnote{Or “If only you would hear his voice today”}}%
\verse{“Do not harden your heart as at Meribah, 
as in the day of Massah in the wilderness,}%
\verse{when your ancestors\lebnote{Or “fathers”}tried me. 
They put me to the test, 
even though they had seen my work.}%
\verse{For forty years I loathed that generation, 
and said, ‘They are a people whose heart wanders. 
And my ways they do not know.’}%
\verse{Therefore I swore in my anger, 
‘They shall surely not enter into my rest.’ ”}%
\end{biblechapter}

\begin{biblechapter} % Psalms 96
\verseWithHeading{Adonai the King Comes in Judgment}{Sing to\lebnote{Or “concerning”}Adonai a new song; 
sing to\lebnote{Or “concerning”}Adonai, all the earth.}%
\verse{Sing to\lebnote{Or “concerning”}Adonai; bless his name. 
Announce\lebnote{Or “bring good news concerning”}his salvation from day to day.}%
\verse{Tell his glory among the nations, 
his marvelous works among all the peoples.}%
\verse{For Adonai is great and very worthy of praise; 
he is to be feared above all gods.}%
\verse{For all the gods of the peoples are idols,\lebnote{This word for idols suggests something worthless or insignificant}
but Adonai made the heavens.}%
\verse{Splendor and majesty are before him; 
strength and beauty are in his sanctuary.}%
\verse{Ascribe to Adonai, you families of the peoples, 
ascribe to Adonai glory and strength.}%
\verse{Ascribe to Adonai the glory due his name; 
bring an offering and come into his courts.}%
\verse{Worship Adonai in holy array; 
tremble before him, all the earth.}%
\verse{Say among the nations, “Adonai is king! 
Yes, the world is established so that it will not be moved. 
He will judge the peoples fairly.”}%
\verse{Let the heavens be glad and the earth rejoice. 
Let the sea with\lebnote{Hebrew “and”}its fullness roar.}%
\verse{Let the field with\lebnote{Hebrew “and”}all that is in it exult. 
Then all the trees of the forests will sing for joy}%
\verse{before Adonai, for he is coming; 
for he is coming to judge the earth. 
He will judge the world with righteousness, 
and the peoples with his faithfulness.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 97
\verseWithHeading{Adonai’s Glorious Reign}{Adonai is king! Let the earth rejoice; 
let many coastlands be glad.}%
\verse{Cloud and thick darkness are surrounding him; 
righteousness and justice are the foundation of his throne.}%
\verse{Fire goes before him, 
and devours his enemies round about.}%
\verse{His lightnings light the world; 
the earth sees and trembles.}%
\verse{The mountains melt like wax at the presence of Adonai, 
at the presence of the Lord of all the earth.}%
\verse{The heavens declare his righteousness, 
and all the peoples see his glory.}%
\verse{Let all who serve an image be ashamed, 
those who boast about idols.\lebnote{This word for idols suggests something worthless or insignificant}
Worship him, all you gods.}%
\verse{Zion hears and is glad, 
and the daughters of Judah rejoice 
because of your judgments, O Adonai.}%
\verse{For you, O Adonai, are most high over all the earth. 
You are highly\lebnote{Hebrew “very”}exalted above all gods.}%
\verse{You who love Adonai, hate evil. 
He protects the lives of his faithful; 
he delivers them from the hand of the wicked.}%
\verse{Light is sown for the righteous, 
and joy for the upright in heart.}%
\verse{Be glad in Adonai, you righteous, 
and give thanks \textit{to his holy name}.\lebnote{Literally “to his holy remembrance” or “to his memorable holiness”}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 98
\verseWithHeading{A psalm.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Sing to\lebnote{Or “sing concerning”}Adonai a new song, for he has done wonders. 
His right hand and his holy arm have secured his victory.}%
\verse{Adonai has made known his salvation;\lebnote{Or “victory”}
to the eyes of the nations he has revealed his righteousness.}%
\verse{He has remembered his loyal love and his faithfulness 
to the house of Israel. 
All the ends of the earth have seen 
the salvation\lebnote{Or “victory”}of our God.}%
\verse{Shout joyfully to Adonai, all the earth. 
Be cheerful\lebnote{Or “Burst forth”}and sing for joy and sing praises.}%
\verse{Sing praises to Adonai with lyre, 
with lyre and melodious sound.}%
\verse{With trumpets and sound of horn, 
shout joyfully before the king, Adonai.}%
\verse{Let the sea with\lebnote{Hebrew “and”}its fullness roar, 
the world and those who live in it.}%
\verse{Let the rivers clap their hands. 
Let the hills sing joyfully together}%
\verse{before Adonai, for he is coming 
to judge the earth. 
He will judge the world with righteousness 
and the peoples with equity.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 99
\verseWithHeading{Adonai Is a Holy King}{Adonai is king; let the peoples tremble. 
He sits enthroned\lebnote{Or “dwells”}between the cherubim. Let the earth shake.}%
\verse{Adonai is great in Zion, 
and he is exalted over all the peoples.}%
\verse{Let them praise your great and fearful name. 
He is holy.}%
\verse{And the strength of the king loves justice.\lebnote{The interpretation of this line is uncertain}
You have established equity; 
you have executed justice and righteousness in Jacob.}%
\verse{Exalt Adonai our God, 
and worship at his footstool. 
He is holy.}%
\verse{Moses and Aaron were among his priests; 
Samuel also was among those who called on his name. 
They called to Adonai, and he answered them.}%
\verse{He spoke to them in the pillar of cloud. 
They kept his testimonies and the statute he gave to them.}%
\verse{O Adonai our God, you answered them. 
You were a forgiving God to them, 
but an avenger of their wrong deeds.}%
\verse{Exalt Adonai our God, 
and worship at his holy mountain, 
for Adonai our God is holy.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 100
\verseWithHeading{A psalm of thanksgiving.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Shout in triumph to Adonai, all the earth.}%
\verse{Serve\lebnote{Or “Worship”}Adonai with joy; 
come into his presence with exultation.}%
\verse{Know that Adonai, he is God; 
he made us and we are his.\lebnote{According to the reading tradition (\textit{Qere}); the \textit{Kethib} reads “and not we ourselves”}
We are his people and the sheep of his pasture.}%
\verse{Enter his gates with thanksgiving, 
his courts with praise. 
Give thanks to him; bless his name.}%
\verse{For Adonai is good; his loyal love is forever, 
and his faithfulness is \textit{from generation to generation}.\lebnote{Literally “unto a generation and a generation”}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 101
\verseWithHeading{Of David. A psalm.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{I will sing of loyal love and justice; 
I will sing praises to you, O Adonai.}%
\verse{I will give attention to the way of integrity. 
When will you come to me? 
I will walk in the integrity of my heart 
in the midst of my house.}%
\verse{I will not set before my eyes 
any wicked thing. 
I hate the work of those who fall away;\lebnote{The term “those who fall away” is a rare noun. The verb from the same root is found in Psa 40:5(4)}
it will not cling to me.}%
\verse{A perverse heart will depart from me; 
I will not know evil.}%
\verse{One who slanders his neighbor in secret, 
him I will destroy. 
One haughty of eyes and \textit{arrogant}\lebnote{Literally “wide”}of heart, 
him I will not endure.}%
\verse{My eyes will be on the faithful of the land, 
that they may live with me. 
He who walks in the way of integrity, 
he shall minister to me.}%
\verse{There will not live in the midst of my house 
a worker of deceit. 
One who speaks lies 
will not remain\lebnote{Hebrew “be established”}before my eyes.}%
\verse{\textit{Each morning},\lebnote{Literally “to the mornings”}
I will destroy all the wicked of the land, 
cutting off from the city of Adonai all evildoers.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 102
\verseWithHeading{A prayer of one afflicted, when he grows faint 
and pours out his lament before Adonai.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{O Adonai, hear my prayer, 
and let my cry for help come to you.}%
\verse{Do not hide your face from me 
in the day of my trouble. 
Incline your ear to me. 
In the day I call, answer me quickly,}%
\verse{for my days vanish in smoke, 
and my bones are charred like a hearth.}%
\verse{My heart is struck like grass and withers. 
Indeed, I forget to eat\lebnote{A rare homonym might yield “I am too wilted to eat”}my bread.}%
\verse{Because of the sound of my groaning 
my bones\lebnote{Hebrew “bone”}cling to my skin.\lebnote{Hebrew “flesh”}}%
\verse{I am like an owl\lebnote{The identity of the birds in this verse is uncertain}of the wilderness; 
I am like a little owl of the ruins.}%
\verse{I lie awake and I am 
like a lone bird on a roof.}%
\verse{All the day my enemies reproach me; 
those who mock me swear oaths against me.}%
\verse{Indeed, I eat ashes like bread 
and mix my drink with tears\lebnote{Hebrew “weeping”}}%
\verse{because of your indignation and anger, 
for you have picked me up and thrown me away.}%
\verse{My days are like a lengthened shadow, 
and I wither like grass.}%
\verse{But you, O Adonai, abide\lebnote{Or “sit enthroned”}forever, 
and your remembrance\lebnote{Or “name”}\textit{from generation to generation}.\lebnote{Literally “to a generation and a generation”}}%
\verse{You rise up and take pity on Zion, 
because it is time to favor it, 
for the appointed time has come.}%
\verse{Your servants take pleasure in her stones, 
and show favor to its dust.}%
\verse{Then the nations will fear the name of Adonai, 
and all the kings of the earth your glory.}%
\verse{For Adonai will rebuild Zion; 
he will appear in his glory.}%
\verse{He will turn his attention to the prayer of the destitute 
and will not despise their prayer.}%
\verse{Let this be written for the next generation, 
so that a people yet to be created may praise Yah,\lebnote{A shortened form of “Adonai”}}%
\verse{that he looked down from his holy height. 
Adonai looked from heaven over the earth}%
\verse{to hear the groaning of the prisoner, 
to liberate \textit{those destined to die},\lebnote{Hebrew “the sons of death”}}%
\verse{so that they\lebnote{That is, the generation that reads what is written}may make known in Zion the name of Adonai, 
and his praise in Jerusalem,}%
\verse{when the peoples assemble, 
together with\lebnote{Hebrew “and”}kingdoms, to serve Adonai.}%
\verse{He has broken my strength along the way; 
he has cut short my days.}%
\verse{I say, “My God, do not carry me off 
from my life in the middle of my days.” 
Your years continue throughout all generations.}%
\verse{Long ago you laid the foundation of the earth, 
and the heavens are the work of your hands.}%
\verse{They will perish, but you will endure. 
And like a garment they will all wear out, 
you will replace them like clothing, and they will be set aside.}%
\verse{But you are the same, 
and your years do not end.}%
\verse{The children of your servants will continue, 
and their \textit{descendants}\lebnote{Literally “seed”}will be established before you.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 103
\verseWithHeading{Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Bless Adonai, O my soul, 
and all within me, bless his holy name.}%
\verse{Bless Adonai, O my soul, 
and do not forget all his benefits:}%
\verse{who forgives all your iniquity, 
who heals all your diseases,}%
\verse{who redeems your life from the pit, 
who crowns you with loyal love and mercies,}%
\verse{who satisfies your life with good 
so that your youth is renewed like the eagle’s.}%
\verse{Adonai does deeds of justice 
and judgments for all who are oppressed,}%
\verse{who made known his ways to Moses, 
his deeds to the \textit{people}\lebnote{Literally “sons”}of Israel.}%
\verse{Adonai is compassionate and gracious, 
\textit{slow to anger}\lebnote{Literally “long of nostrils”}and abundant in loyal love.}%
\verse{He does not dispute continually, 
nor keep his anger forever.}%
\verse{He has not dealt with us according to our sins, 
nor repaid us according to our iniquities.}%
\verse{For as the heavens are high above the earth, 
so his loyal love prevails over those who fear him.}%
\verse{As far as east is from west, 
so he has removed far from us the guilt of our transgressions.}%
\verse{As a father pities\lebnote{Or “shows compassion to”}his children, 
so Adonai pities\lebnote{Or “shows compassion to”}those who fear him.}%
\verse{For he knows our frame.\lebnote{That is, how we are formed}
He remembers that we are dust.}%
\verse{As for man, his days are like the grass. 
As the flower of the field, so he blossoms.}%
\verse{When the wind passes over it, it is no more, 
and its place knows it no longer.}%
\verse{But the loyal love of Adonai 
is from everlasting to everlasting on those who fear him, 
and his righteousness to their children’s children,}%
\verse{to those who keep his covenant 
and remember to do his precepts.}%
\verse{Adonai has established his throne in the heavens, 
and his kingdom rules over all.}%
\verse{Bless Adonai, you his angels, 
you mighty heroes who do his word 
by \textit{obeying}\lebnote{Literally “hearing”}the sound of his word.}%
\verse{Bless Adonai, all you his hosts, 
you his attendants who do his will.}%
\verse{Bless Adonai, all his works,\lebnote{That is, created things}
in all the places of his dominion. 
Bless Adonai, O my soul.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 104
\verseWithHeading{Praise to Adonai for His Creation and Providence}{Bless Adonai, O my soul. 
O Adonai my God, you are very great. 
You clothe yourself with splendor and majesty,}%
\verse{you who cover yourself with light as with a garment, 
who stretch out the heavens like a tent curtain,}%
\verse{the one who sets beams in the waters for his upper chambers, 
who makes clouds his chariot, 
who rides\lebnote{Hebrew “goes”}on the wings of the wind,}%
\verse{who makes his messengers the winds, 
his attendants a flame of fire.}%
\verse{He established the earth on her foundations, 
so that it will not be moved forever and ever.}%
\verse{You covered it with the deep as with a garment. 
The waters stood above the mountains.}%
\verse{At your rebuke they fled; 
at the sound of your thunder they ran off.}%
\verse{They ascended the mountains and drained though the valleys 
to the place that you established\lebnote{Or “appointed”}for them.}%
\verse{You set a boundary that they may not cross over, 
so that they would not return to cover the earth.}%
\verse{You are the one who sends forth springs into the valleys; 
they flow between the mountains.}%
\verse{They give drink for every beast of the field. 
The wild donkeys \textit{quench}\lebnote{Literally “break”}their thirst.}%
\verse{Along them the birds of the heavens abide. 
From among the branches they \textit{sing}.\lebnote{Literally “give voice”}}%
\verse{You are the one who \textit{waters}\lebnote{Literally “give drink”}the mountains 
from his upper chambers. 
The earth is full\lebnote{Or “is sated”}with the fruit of your labors:}%
\verse{who causes grass to grow for the cattle 
and herbs for the service of humankind, 
to bring forth food from the earth,}%
\verse{and wine that makes glad the heart of man, 
so that their faces shine from oil, 
and bread that strengthens the heart of man.}%
\verse{The trees of Adonai drink their fill, 
the cedars of Lebanon that he planted,}%
\verse{where birds make their nest. 
The stork has its home in the fir trees.}%
\verse{The high mountains are for the wild goats; 
the cliffs are a refuge for the rock badgers.}%
\verse{He made the moon for appointed times; 
the sun knows its time for setting.}%
\verse{You make darkness, and it is night, 
when\lebnote{Hebrew “in it”}all the animals of the forest creep about.}%
\verse{The young lions are roaring for the prey 
and seeking their food from God.}%
\verse{When the sun rises, they gather together 
and lie down in their dens.}%
\verse{Humankind goes out to its work, 
and to its labor until evening.}%
\verse{How many are your works, O Adonai; 
all of them you have done in wisdom. 
The earth is full of your creatures.}%
\verse{This is the great and \textit{wide}\lebnote{Literally “broad of hands”}sea, 
in which are moving animals without number, 
living things small and great.}%
\verse{There the ships sail. 
Leviathan is there that you formed to play with.\lebnote{For this construction meaning “to play with,” see Job 40:29 (Eng: 41:5)}}%
\verse{They all wait for you 
to give them their food \textit{at the proper time}.\lebnote{Literally “in its season”}}%
\verse{You give to them; they gather it. 
You open your hand, they are filled with what is good.}%
\verse{You hide your face, they are terrified. 
You take away their breath, they die 
and return to their dust.}%
\verse{You send forth your Spirit,\lebnote{Or “breath”}they are created, 
and you renew the face of the ground.}%
\verse{May the glory of Adonai endure forever. 
May Adonai be glad in his works,}%
\verse{the one who looks at the earth and it quakes, 
and touches the mountains and they smoke.}%
\verse{I will sing to Adonai throughout my life; 
I will sing praise to my God while \textit{I remain alive}.\lebnote{Literally “while I am still”}}%
\verse{May my meditation be pleasing to him. 
I will be glad in Adonai.}%
\verse{Let sinners perish completely from the earth, 
and the wicked \textit{not remain alive}.\lebnote{Literally “they are not still”}
Bless Adonai, O my soul. 
Praise Yah.\lebnote{Hebrew \textit{hallelujah}}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 105
\verseWithHeading{Praise to Adonai for His Work on Behalf of Israel}{Give thanks to Adonai; proclaim his name; 
make known his deeds among the peoples.}%
\verse{Sing to him; sing praises concerning him; 
tell of all his wonderful works.}%
\verse{Boast about his holy name. 
Let the heart of those who seek Adonai rejoice.}%
\verse{Make supplication to Adonai and his might; 
seek his face continually.}%
\verse{Remember his wonders that he has done, 
his signs and the judgments of his mouth,}%
\verse{O offspring of Abraham his servant, 
you descendants\lebnote{Or “children,” or “sons”}of Jacob, his chosen ones.}%
\verse{He is Adonai our God; 
his judgments are in all the earth.}%
\verse{He remembers his covenant forever, 
the word that he commanded 
for a thousand generations,}%
\verse{that he made with Abraham, 
and by his oath swore to Isaac.}%
\verse{He then confirmed it to Jacob as a statute, 
to Israel as an everlasting covenant,}%
\verse{saying, “To you I will give the land of Canaan 
as the allotment that is your\lebnote{Hebrew plural}inheritance.”}%
\verse{When they were \textit{few in number}\lebnote{Literally “men of number,” that is, so few you could count them}— 
a trifle—and were sojourners in it,}%
\verse{and they wandered about \textit{among the nations},\lebnote{Literally “from people to people”}
from one kingdom to another people,}%
\verse{he allowed no one to oppress them, 
and he rebuked kings on account of them,}%
\verse{“Do not touch my anointed ones, 
nor harm my prophets.”}%
\verse{Then he called for a famine upon the land; 
he broke every \textit{supply}\lebnote{Literally “staff”}of bread.}%
\verse{He sent a man on ahead of them; 
Joseph was sold as a slave.}%
\verse{They clamped\lebnote{Hebrew “oppressed, hurt”}his feet in fetters.\lebnote{Hebrew “fetter”}
His neck\lebnote{Hebrew \textit{nephesh} can designate the life, soul, appetite, desire, or throat/neck.}went into irons.\lebnote{Hebrew “iron”}}%
\verse{Until the time his word came about, 
the word of Adonai tested\lebnote{Or “refined”}him.}%
\verse{The king sent and he freed him; 
the ruler of the peoples sent and let him loose.}%
\verse{He made him lord of his house 
and ruler of all his possessions,}%
\verse{to obligate\lebnote{Hebrew “to bind, imprison” The LXX’s “to instruct” presupposes \textit{ysr} instead of \textit{'sr}, and might be correct in light of the parallelism that is fairly tight in this psalm}his officials \textit{as he saw fit}\lebnote{Literally “in his desire soul”}
and teach his elders wisdom.}%
\verse{Then Israel came to Egypt, 
and Jacob was a sojourner in the land of Ham.}%
\verse{And he\lebnote{That is, God}made his people very fruitful. 
He also made them\lebnote{Hebrew “him”}stronger than their\lebnote{Hebrew “his”}enemies.}%
\verse{He turned their heart to hate his people, 
to deal deceitfully against his servants.}%
\verse{He sent his servant Moses, 
together with Aaron whom he had chosen.}%
\verse{They set before them the words concerning his signs 
and portents in the land of Ham.}%
\verse{He sent darkness and made it dark, 
and they did not rebel against his words.}%
\verse{He turned their waters into blood 
and it killed their fish.}%
\verse{Their land swarmed with frogs, 
even in the chambers of their kings.}%
\verse{He spoke, and there came flies 
and gnats throughout all their territory.\lebnote{Hebrew “borders”}}%
\verse{He gave hail for their rains 
and the fire of lightning in their land.}%
\verse{And he struck their vines and fig trees, 
and shattered the trees of their territory.}%
\verse{He spoke and there came locusts 
and young locusts without number.}%
\verse{And they ate up all the vegetation in their land, 
and they ate up the fruit of their ground.}%
\verse{And he struck down all the firstborn in their land, 
the first of all their virility.}%
\verse{Then he brought them out with silver and gold, 
and there was none among his tribes who stumbled.}%
\verse{Egypt was glad when they departed, 
because the fear of them\lebnote{That is, Israel}had fallen upon them.}%
\verse{He spread out a cloud for a covering, 
and fire to give light at night.}%
\verse{They\lebnote{Hebrew “He”}asked, and he brought quail, 
and satisfied them with the bread of heaven.}%
\verse{He opened the rock and waters flowed; 
they coursed\lebnote{Or “surged”}through the dry places like a river.}%
\verse{For he remembered his holy \textit{promise};\lebnote{Literally “word”}
he remembered Abraham his servant.}%
\verse{And so he brought out his people with joy, 
his chosen ones with rejoicing.}%
\verse{And he gave to them the lands of the nations, 
and they inherited the labor of the peoples,}%
\verse{so that they might observe his statutes, 
and keep his laws. 
Praise Yah.\lebnote{Hebrew \textit{hallelujah}}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 106
\verseWithHeading{Praise to Adonai for His Faithfulness in Israel’s History}{Praise Yah.\lebnote{Hebrew \textit{hallelujah}}Give thanks to Adonai, for he is good, 
for his loyal love is forever.}%
\verse{Who can utter the mighty deeds of Adonai, 
or proclaim all his praise?}%
\verse{Blessed are those who observe justice, 
he who does righteousness at all times.}%
\verse{Remember me, O Adonai, when you show favor to your people. 
Look after me when you deliver,}%
\verse{that I may see the good done your chosen ones, 
to be glad in the joy of your nation, 
to glory together with your inheritance.}%
\verse{We have sinned along with our ancestors.\lebnote{Or “fathers”}
We have committed iniquity; we have incurred guilt.}%
\verse{Our ancestors\lebnote{Or “fathers”}in Egypt did not understand your wonderful works. 
They did not remember your many acts of loyal love, 
and so they rebelled by the sea at \textit{the Red Sea}.\lebnote{Literally “sea of reed”}}%
\verse{Yet he saved them for the sake of his name, 
to make known his might.}%
\verse{So he rebuked \textit{the Red Sea},\lebnote{Literally “sea of reed”}and it dried up, 
and he led them through the deep as through a desert.}%
\verse{Thus he saved them from the hand of the hater 
and redeemed them from the hand of the enemy.}%
\verse{But waters covered their adversaries; 
not one of them survived.}%
\verse{Then they believed his words; 
they sang his praise.}%
\verse{They quickly forgot his works; 
they did not wait for his counsel.}%
\verse{And they \textit{craved intensely}\lebnote{Literally “craved a craving”}in the wilderness, 
and tested God in the desert.}%
\verse{So he gave to them their request, 
but he sent leanness into their souls.\lebnote{Hebrew “soul” or “inner self”}}%
\verse{And they were jealous of Moses in the camp, 
and of Aaron, the holy one of Adonai.}%
\verse{The earth opened and swallowed up Dathan, 
and it covered over the gang of Abiram.}%
\verse{Also fire burned in their assembly; 
the flame devoured the wicked.}%
\verse{They made a calf at Horeb 
and bowed down to a cast image.}%
\verse{And so they exchanged their glory 
for an image of an ox that eats grass.}%
\verse{They forgot God their Savior, 
who had done great things in Egypt,}%
\verse{wonders in the land of Ham, 
awesome deeds by \textit{the Red Sea}.\lebnote{Literally “sea of reed”}}%
\verse{So he said he would exterminate them, 
had not Moses, his chosen one, 
stood in the breach before him, 
to reverse his wrath from destroying them.}%
\verse{Then they refused the desirable land; 
they did not believe his word,}%
\verse{but grumbled in their tents. 
They did not obey the voice of Adonai.}%
\verse{So \textit{he made an oath}\lebnote{Literally “he lifted his hand,” an act accompanying an oath}against them, 
to make them drop in the wilderness,}%
\verse{and to disperse\lebnote{Or “let drop”}their \textit{descendants}\lebnote{Literally “seed”}among the nations 
and to scatter them among the lands.}%
\verse{They also attached themselves to Baal of Peor, 
and they ate sacrifices offered to the dead.\lebnote{Hebrew “sacrifices of dead”}}%
\verse{Thus they provoked to anger by their deeds, 
and a plague broke out among them.}%
\verse{Then Phinehas stood up and intervened, 
and so the plague was stopped,}%
\verse{and it was reckoned to him as righteousness 
\textit{throughout all generations}.\lebnote{Literally “for a generation and a generation”}}%
\verse{They also angered God at the waters of Meribah, 
and it went badly for Moses on account of them,}%
\verse{because they rebelled against his Spirit,\lebnote{A slightly different verbal form yields “they embittered his spirit”}
and he spoke thoughtlessly with his lips.}%
\verse{They did not exterminate the peoples, 
as Adonai had commanded them, \lebnote{Hebrew “said to them”}}%
\verse{but they mingled with the nations 
and learned their works,}%
\verse{and served their idols, 
which became\lebnote{Hebrew “and they became”}a snare to them.}%
\verse{They even sacrificed their sons and daughters 
to the demons,}%
\verse{and they poured out innocent blood, 
the blood of their sons and daughters, 
whom they sacrificed to the idols of Canaan, 
and so the land was defiled with the blood.\lebnote{Hebrew “bloods,” that is, bloodshed}}%
\verse{And they became unclean by their works, 
and were unfaithful in their deeds.}%
\verse{So Adonai’s anger burned against his people, 
and he abhorred his inheritance.}%
\verse{Then he gave them into the hand of the nations, 
and those who hated them ruled over them.}%
\verse{And their enemies oppressed them, 
and they were subdued under their hand.}%
\verse{Many times he delivered them, 
but they rebelled in their counsel, 
and were brought low by their iniquity.}%
\verse{Yet he looked upon their distress 
when he heard their cry.}%
\verse{And he remembered his covenant with them, 
and relented based on the abundance of his loyal love.}%
\verse{And he \textit{let them find compassion}\lebnote{Literally “gave them to have mercies”}
before all their captors.}%
\verse{Save us, O Adonai our God, 
and gather us from the nations, 
so that we may give thanks to your holy name 
and boast in your praise.}%
\verse{Blessed is Adonai, the God of Israel, 
from everlasting and to everlasting. 
And let all the people say, “Amen!” 
Praise Yah.\lebnote{Hebrew \textit{hallelujah}}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 107
\verseWithHeading{Thanksgiving to Adonai for His of Deliverance}{Give thanks to Adonai, for he is good, 
for his loyal love is forever.}%
\verse{Let the redeemed of Adonai declare\lebnote{Hebrew “say”}it, 
those whom he has redeemed from the hand of the enemy}%
\verse{and gathered from the lands, 
from east and from west, from north and from south.\lebnote{Hebrew “sea” (\textit{yam}), which usually means “west” in a listing of directions. The word “south” is expected (Hebrew \textit{yamin}). Either the MT was corrupted early (along with Isa 49:12; the LXX reads “sea” in both) or it looks to the bodies of waters to the south (see also the Targum)}}%
\verse{They wandered in the wilderness, in a desert. 
They could find no way\lebnote{The MT reads “way” with the previous line.}to a city to inhabit.}%
\verse{Hungry and thirsty, 
their soul grew faint within them.}%
\verse{Then they cried out to Adonai in their trouble. 
He delivered them from their distresses}%
\verse{and led them by a straight way 
to get to a city to inhabit.}%
\verse{Let them give thanks to Adonai for his loyal love, 
and his wonderful deeds for the children of humankind,}%
\verse{for he satisfies the longing soul,\lebnote{Thirst is perhaps in view}
and the hungry soul he fills with good.}%
\verse{Those who sat in darkness and gloom, 
prisoners of misery and iron—}%
\verse{because they rebelled against the words of God 
and spurned the counsel of the Most High,}%
\verse{he therefore humbled their heart with trouble. 
They stumbled and there was no helper.}%
\verse{Then they called to Adonai for help in their trouble; 
he saved them from their distresses.}%
\verse{He brought them out of darkness and gloom, 
and tore off their bonds.}%
\verse{Let them give thanks to Adonai for his loyal love 
and his wonderful deeds for the children of humankind,}%
\verse{for he shatters the doors of bronze, 
and cuts through the bars of iron.}%
\verse{Fools, because of their rebellious way 
and their iniquities, were afflicted.}%
\verse{Their soul abhorred all food, 
and they approached the gates of death.}%
\verse{Then they called to Adonai for help in their trouble. 
He saved them from their distresses.}%
\verse{He sent his word and healed them, 
and he delivered them from their pits.\lebnote{This word is only found elsewhere in Lam 4:20 with the sense of “traps”}}%
\verse{Let them give thanks to Adonai for his loyal love, 
and his wonderful deeds for the children of humankind,}%
\verse{and let them offer sacrifices of thanksgiving, 
and tell of his works with rejoicing.}%
\verse{Those who went down to the sea into ships, 
doing business on the \textit{high seas},\lebnote{Literally “many waters”}}%
\verse{they saw the works of Adonai, 
and his wonderful deeds in the deep.}%
\verse{For he spoke and raised up a stormy wind, 
and it whipped up its waves.}%
\verse{They rose to the heavens; they plunged to the depths. 
Their soul melted in their calamity.}%
\verse{They reeled and staggered like a drunkard, 
and \textit{they were at their wits’ end}.\lebnote{Literally “all their wisdom was swallowed up”}}%
\verse{Then they cried out to Adonai in their trouble, 
and he brought them out of their distresses.}%
\verse{He made the storm be still 
and their waves became calm.}%
\verse{Then they were glad because they grew silent, 
so he guided them to their desired harbor.}%
\verse{Let them give thanks to Adonai for his loyal love 
and his wonderful deeds for the children of humankind,}%
\verse{and let them exalt him in the congregation of the people, 
and praise him in the assembly of the elders.}%
\verse{He turns rivers into a wilderness 
and springs of water into thirsty ground,}%
\verse{a fruitful land into a salty place, 
because of the evil of its inhabitants.}%
\verse{He turns a wilderness into a pool of water 
and a dry land into springs of water.}%
\verse{And he settles the hungry there, 
so that they may establish a city to inhabit,}%
\verse{and sow fields and plant vineyards, 
\textit{that they may yield fruit at harvest}.\lebnote{Hebrew “and make a fruit of produce”}}%
\verse{And he blesses them and they multiply greatly, 
and he does not let their cattle become few.}%
\verse{When they become few and they are bent down 
from the oppression of calamity and grief,}%
\verse{he pours contempt on princes 
and causes them to wander in a trackless waste.}%
\verse{But he protects\lebnote{The sense is that of making something inaccessibly high, like a fortress}the needy from misery, 
and he makes their families like a flock.}%
\verse{The upright see it and are glad, 
and all wickedness shuts its mouth.}%
\verse{Whoever is wise, then let him observe these things, 
and let them consider Adonai’s acts of loyal love.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 108
\verseWithHeading{A song. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{My heart is steadfast, O God. 
I will sing and give praise, even with my glory.}%
\verse{Awake, harp and lyre. 
I will awake the dawn.\lebnote{Or “I will awake at dawn”}}%
\verse{I will give you thanks among the peoples, O Adonai; 
I will give you praise among the nations,}%
\verse{because your loyal love is great above the heavens, 
and your faithfulness reaches to the clouds.}%
\verse{Be exalted above the heavens, O God, 
and let your glory be above all the earth,}%
\verse{so that your beloved ones may be rescued. 
Save by your right hand and answer me.}%
\verse{God has spoken in his holiness, 
“I will exult, I will divide up Shechem, 
and portion out the valley of Succoth.}%
\verse{Gilead is mine. Manasseh is mine. 
And Ephraim is the \textit{helmet for}\lebnote{Literally “protection of”}my head. 
Judah is my scepter;\lebnote{Or “commander’s rod”}}%
\verse{Moab is my washing pot. 
Over Edom I will cast my sandal; 
Over Philistia I will shout in triumph.”}%
\verse{Who will bring me to the fortified city? 
Who will lead me to Edom?}%
\verse{Have you not rejected us, O God, 
and not gone out with our armies, O God?}%
\verse{Give us help against the enemy, 
for the help of humankind is futile.}%
\verse{Through God we will do valiantly,\lebnote{Hebrew “do might”}
and it is he who will tread down our enemies.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 109
\verseWithHeading{For the music director. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{O God of my praise, do not keep silent,}%
\verse{for \textit{wicked and deceitful mouths}\lebnote{Literally “a wicked mouth and a deceitful mouth”}
have opened against me. 
They speak to me with a lying tongue.}%
\verse{They also surround me with words of hate, 
and fight me without cause.}%
\verse{In return for my love they accuse me, 
though I am in prayer.\lebnote{Or perhaps “I have a plea of innocence”}}%
\verse{So they inflicted evil against me in return for good 
and hatred in return for my love.}%
\verse{Appoint over him a wicked man, 
and let an accuser stand at his right hand.}%
\verse{When he is judged, let him come out guilty, 
and let his prayer\lebnote{Or perhaps “plea of innocence”}become as sin.}%
\verse{Let his days be few; 
let another take his office.}%
\verse{Let his children be orphans, 
and his wife a widow,}%
\verse{and let his children wander aimlessly and beg, 
and let them plead from their ruins.}%
\verse{Let the creditor seize all that is his, 
and let strangers plunder his property.}%
\verse{Let there be none who extend to him loyal love, 
nor any who pities his orphans.}%
\verse{Let his descendants\lebnote{Or “posterity”; Hebrew is singular}be cut off. 
Let their name be blotted out in the next generation.}%
\verse{Let the iniquity of his ancestors\lebnote{Or “fathers”}be remembered before Adonai, 
and let the sin of his mother not be blotted out.}%
\verse{Let them be before Adonai continually, 
that he may cut off their\lebnote{That is, the family}memory from the earth,}%
\verse{because he did not remember to show loyal love, 
but he pursued anyone,\lebnote{Hebrew “a man”}poor or needy 
or brokenhearted, to slay them.}%
\verse{Because he loved cursing, let it come upon him. 
Because he did not delight in blessing, 
let it be far from him.}%
\verse{Because he wore a curse as his robe, 
let it enter \textit{his body}\lebnote{Literally “his inward part”}like water, 
and into his bones like oil.}%
\verse{May it be for him like a garment in which he wraps, 
and a belt \textit{he continually wears}.\lebnote{Literally “continually he girds it”}}%
\verse{Let this be the punishment for my accusers from Adonai, 
even those who speak evil against my life.}%
\verse{But you, O Adonai my Lord, 
deal with me for your name’s sake. 
Because your loyal love is good, deliver me,}%
\verse{for I am poor and needy, 
and my heart is wounded within me.}%
\verse{Like a lengthening shadow I am passing away; 
I am shaken off like a locust.}%
\verse{My knees \textit{buckle}\lebnote{Literally “stumble”}from fasting, 
and my body grows lean without fat.}%
\verse{And so I am a disgrace to them; 
when they see me, they shake their heads.\lebnote{Hebrew “head”}}%
\verse{Help me, O Adonai my God; 
save me according to your loyal love,}%
\verse{that they may know that this is your hand, 
that you, O Adonai, you have done it.}%
\verse{Let them curse, but you bless. 
When they arise, let them be put to shame, 
that your servant may be glad.}%
\verse{Let my accusers put on disgrace, 
and let them cover themselves with their shame as with a robe.}%
\verse{I will give thanks to Adonai exceedingly with my mouth, 
and in the midst of many I will praise him,}%
\verse{for he stands at the right hand of the needy, 
to save him from those judging his life.\lebnote{Or “soul”}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 110
\verseWithHeading{Of David. A psalm.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{A declaration of Adonai to my lord, 
“Sit at my right hand 
until I make your enemies your footstool.”}%
\verse{Adonai will send out your mighty scepter from Zion; 
rule in the midst of your enemies.}%
\verse{Your people \textit{will volunteer}\lebnote{Literally “will be freewill offerings”; see the related verbal form in Judg 5:2}in the day of your power.\lebnote{Or “day of your army,” or “day of battle”}
In holy splendor, from the womb of the dawn, 
you will have the dew of your youth.\lebnote{The meaning of this verse is obscure}}%
\verse{Adonai has sworn and he will not change his mind, 
“You are a priest forever 
according to the manner of Melchizedek.”}%
\verse{O Lord, at your right hand 
he will shatter kings in the day of his anger.}%
\verse{He will judge among the nations; 
he will fill them with corpses. 
He will shatter the \textit{rulers}\lebnote{Literally “head”}of many countries.}%
\verse{He will drink from the stream by the road; 
therefore he will lift up his head.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 111
\verseWithHeading{Praise to God for His Work and Commands}{Praise Yah!\lebnote{Hebrew \textit{hallelujah}}
I will give thanks to Adonai with all my heart, 
in the assembly of the upright and the congregation.}%
\verse{The works of Adonai are great, 
studied\lebnote{Or “pondered”}by all who delight in them.}%
\verse{Full of splendor and majesty is his work, 
and his righteousness endures forever.}%
\verse{He has made his wonders \textit{to be remembered};\lebnote{Literally “a remembrance”}
Adonai is gracious and compassionate.}%
\verse{He gives food to those who fear him; 
he remembers his covenant forever.}%
\verse{He has declared the power of his works for his people, 
by giving to them the inheritance of the nations.}%
\verse{The works of his hands are faithfulness and justice; 
all his precepts are reliable.}%
\verse{They are sustained forever and ever, 
done with faithfulness and uprightness.}%
\verse{He has sent redemption to his people; 
he has commanded his covenant forever; 
holy and fearful is his name.}%
\verse{The fear of Adonai is the beginning of wisdom; 
all who do them\lebnote{That is, his precepts (v. 7), or the stipulations of his covenant (v. 9)}have a good understanding. 
His praise endures forever.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 112
\verseWithHeading{The Path of the Righteous and the Path of the Wicked}{Praise Yah!\lebnote{Hebrew \textit{hallelujah}}
Blessed is the man who fears Adonai; 
he takes great delight in his commands.}%
\verse{His \textit{descendants}\lebnote{Literally “seed”}will be mighty in the land; 
the generation of the upright will be blessed.}%
\verse{Wealth and riches are in his house, 
and his righteousness endures forever.}%
\verse{His light rises in the darkness for the upright; 
he is gracious and compassionate and righteous.}%
\verse{It goes well for a man who is gracious and lends, 
who conducts his \textit{business properly}.\lebnote{Literally “matters in justice”}}%
\verse{For he will not be shaken forever; 
the righteous \textit{will be remembered forever}.\lebnote{Literally “will be for an eternal remembrance”}}%
\verse{He is not afraid of bad news; 
his heart is steadfast, secure\lebnote{Or “full of trust”}in Adonai.}%
\verse{His heart is sustained; he will not be afraid, 
until he looks in triumph on his enemies.}%
\verse{\textit{He gives lavishly}\lebnote{Literally “He distributes freely, he gives”}to the needy; 
his righteousness endures forever. 
His horn rises high in honor.}%
\verse{The wicked\lebnote{Singular}sees it and is vexed. 
He gnashes his teeth and \textit{becomes weak}.\lebnote{Literally “and melts”}
The desire of the wicked\lebnote{Plural}perishes.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 113
\verseWithHeading{God’s Majesty and Care for the Needy}{Praise Yah!\lebnote{Hebrew \textit{hallelujah}}
Praise, O servants of Adonai, 
praise the name of Adonai.}%
\verse{Let the name of Adonai be blessed, 
from now until forever.}%
\verse{From the rising of the sun to its setting, 
let the name of Adonai be blessed.}%
\verse{Adonai is high above all nations; 
his glory is above the heavens.}%
\verse{Who is like Adonai our God, 
who is enthroned on high,\lebnote{Or “who dwells on high”}}%
\verse{who \textit{condescends to look at}\lebnote{Hebrew “makes low to see”}
what is in the heavens and in the earth?}%
\verse{He raises the helpless from the dust; 
he lifts the needy from the ash heap,}%
\verse{to seat them with princes, 
with the princes of his people.}%
\verse{He causes the barren woman of the house to dwell 
as the happy mother of children. 
Praise Yah!\lebnote{Hebrew \textit{hallelujah}}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 114
\verseWithHeading{Praise to God for His Works During the Exodus}{When Israel went out from Egypt, 
the house of Jacob from a people of a strange language,}%
\verse{Judah became his sanctuary, 
Israel his dominion.}%
\verse{The sea looked and fled; 
the Jordan turned back.}%
\verse{The mountains skipped like rams, 
the hills like \textit{lambs}.\lebnote{Literally “sons of the flock”}}%
\verse{What’s with you, O sea, that you fled? 
O Jordan, that you turned back?}%
\verse{O mountains, that you skipped like rams? 
O hills, like lambs?}%
\verse{At the presence of the Lord writhe, O earth,\lebnote{Or “O land”}
at the presence of the God of Jacob,}%
\verse{who turned the rock into a pool of water, 
flinty stone into a spring of water.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 115
\verseWithHeading{Dead Idols and the Living God}{Not to us, O Adonai, not to us, 
but to your name give glory, 
because of your loyal love, 
because of your faithfulness.}%
\verse{Why should the nations say, 
“Where, \textit{pray tell},\lebnote{Literally “please”}is their God?”}%
\verse{But our God is in the heavens; 
all that he desires, he does.}%
\verse{Their idols are silver and gold, 
the work of human hands.}%
\verse{They have mouths, but cannot speak; 
they have eyes, but cannot see;}%
\verse{they have ears, but cannot hear; 
they have noses, but cannot smell;}%
\verse{they have their hands, but they cannot feel, 
their feet, but they cannot walk; 
they cannot utter a sound in their throats.}%
\verse{Those who make them become like them, 
as does everyone who trusts in them.}%
\verse{O Israel, trust Adonai; 
he is their help and their shield.}%
\verse{O house of Aaron, trust in Adonai; 
he is their help and their shield.}%
\verse{You who fear Adonai, trust in Adonai; 
he is their help and their shield.}%
\verse{Adonai remembers us; he will bless us. 
He will bless the house of Israel; 
he will bless the house of Aaron.}%
\verse{He will bless those who fear Adonai, 
the small with the great.}%
\verse{May Adonai give you increase, 
you and your children.}%
\verse{May you be blessed by Adonai, 
who made heaven and earth.}%
\verse{The heavens are Adonai’s heavens, 
but the earth he has given to the children of humankind.}%
\verse{The dead do not praise Yah,\lebnote{A shortened form of “Adonai”}
nor any descending into silence.}%
\verse{It is we, we will bless Yah\lebnote{A shortened form of “Adonai”}
from now until forever. 
Praise Yah!\lebnote{Hebrew \textit{hallelujah}}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 116
\verseWithHeading{Thanksgiving for God’s Deliverance}{I love him, because Adonai has heard 
\textit{the voice of my supplications}.\lebnote{Literally “my voice, my supplications”}}%
\verse{Because he has inclined his ear to me, 
I will call \textit{all my days}.\lebnote{Literally “in my days”}}%
\verse{The ropes of death encircled me, 
and the distresses of Sheol found me. 
I found trouble and grief.}%
\verse{Then I called on the name of Adonai: 
“O Adonai, please save my life!”}%
\verse{Gracious is Adonai and righteous, 
and our God is merciful.}%
\verse{Adonai watches over the simple. 
I was brought low, but he gave me victory.}%
\verse{Return, O my soul, to your repose, 
for Adonai has dealt bountifully with you.}%
\verse{For you have rescued me\lebnote{Hebrew “my soul.” If “soul” is accepted, then “death” means “the place of death,” that is, the netherworld.}from death, 
my eyes\lebnote{Hebrew “eye”}from tears, my feet from stumbling.}%
\verse{I will walk before Adonai 
in the land\lebnote{Hebrew “lands”}of the living.}%
\verse{I believed when I spoke, 
“I am afflicted greatly.”}%
\verse{I said in my haste, 
“\textit{Everyone}\lebnote{Literally “All of the humankind”}is a liar.”}%
\verse{What shall I give back to Adonai 
for all his benefits to me?}%
\verse{I will lift up the cup of salvation 
and proclaim the name of Adonai.}%
\verse{I will pay my vows made to Adonai 
in the presence of all his people.}%
\verse{Costly \textit{in Adonai’s view}\lebnote{Literally “in the eyes of Adonai”}is 
the death of his faithful ones.}%
\verse{Ah, Adonai, I am indeed your servant; 
I am your servant, the child of your maidservant. 
You have loosed my bonds.}%
\verse{I will offer to you a sacrifice of thanksgiving, 
and proclaim the name of Adonai.}%
\verse{I will pay my vows made to Adonai 
in the presence of all his people,}%
\verse{in the courts of the house\lebnote{Or “temple”}of Adonai, 
in the midst of you, O Jerusalem. 
Praise Yah!\lebnote{Hebrew \textit{hallelujah}}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 117
\verseWithHeading{Let All Peoples Praise Adonai}{Praise Adonai, all you nations; 
extol him, all you peoples.}%
\verse{For his loyal love is mighty on our behalf, 
and the faithfulness of Adonai is forever. 
Praise Yah!\lebnote{Hebrew \textit{hallelujah}}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 118
\verseWithHeading{Praise to God for His Loyal Love}{Give thanks to Adonai for he is good, 
for his loyal love is forever.}%
\verse{Let Israel say, 
“His loyal love is forever.”}%
\verse{Let the house of Aaron say, 
“His loyal love is forever.”}%
\verse{Let those who fear Adonai say, 
“His loyal love is forever.”}%
\verse{Out of my distress I called to Yah.\lebnote{A shortened form of “Adonai”}
Yah answered me, setting me in a broad place.}%
\verse{Adonai is for me; I do not fear. 
What can mere humans\lebnote{Or “humankind”}do to me?}%
\verse{Adonai is for me as my helper, 
and so I will look in triumph on those who hate me.}%
\verse{It is better to take refuge in Adonai 
than to trust in humans.\lebnote{Or “humankind”}}%
\verse{It is better to take refuge in Adonai 
than to trust princes.}%
\verse{All nations surrounded me. 
In the name of Adonai I opposed them indeed.}%
\verse{They surrounded me; yes, they surrounded me. 
In the name of Adonai I opposed them indeed.}%
\verse{They surrounded me like bees; 
they were extinguished like a fire of thorns. 
In the name of Adonai I opposed them indeed.}%
\verse{You pushed me hard\lebnote{Hebrew intensive “pushing, you pushed me”}to make me fall, 
but Adonai helped me.}%
\verse{Yah\lebnote{A shortened form of “Adonai”}is my strength and my song,\lebnote{Or “my might”}
and he has become my salvation.\lebnote{Or “victory”}}%
\verse{The sound of rejoicing and salvation\lebnote{Or “victory”}
is in the tents of the righteous; 
the right hand of Adonai has done valiantly.}%
\verse{The right hand of Adonai has exalted; 
the right hand of Adonai has done valiantly.}%
\verse{I will not die but live, 
and tell of the works of Yah.\lebnote{A shortened form of “Adonai”}}%
\verse{Yah has disciplined me severely, 
but he did not consign\lebnote{Hebrew “give”}me to death.}%
\verse{Open to me the gates of righteousness, 
that I may enter through them 
and give thanks to Yah.\lebnote{A shortened form of “Adonai”}}%
\verse{This is the gate of Adonai, 
through which the righteous will enter.}%
\verse{I will give thanks to you for you have answered me, 
and you have become my salvation.\lebnote{Or “victory”}}%
\verse{The stone the builders rejected 
has become the chief cornerstone.}%
\verse{This is from Adonai; 
it is wonderful in our eyes.}%
\verse{This is the day Adonai has worked;\lebnote{Or “made”}
let us rejoice and be glad in him.\lebnote{Or “in it”}}%
\verse{O Adonai, please save; 
O Adonai, please grant success.}%
\verse{Blessed is he who comes in the name of Adonai. 
We bless you\lebnote{Hebrew plural}from the house\lebnote{Or “temple”}of Adonai.}%
\verse{Adonai is God, and he has given us light.\lebnote{Or “made his light to shine upon to us”}
Bind the festal sacrifice\lebnote{Or “festal procession”}with cords\lebnote{Or “branches”}
to the horns\lebnote{Or “and lead up to the horns”}of the altar.}%
\verse{You are my God, and I will give thanks to you. 
You are my God; I will exalt you.}%
\verse{Give thanks to Adonai for he is good, 
for his loyal love is forever.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 119
\verseWithHeading{Aleph\lebnote{Psalm 119 is an acrostic psalm, with twenty-two stanzas corresponding to the letters of the Hebrew alphabet}}{Blessed are those whose way is blameless, 
who walk in the law of Adonai.}%
\verse{Blessed are those who keep his testimonies; 
they seek him with a whole heart.}%
\verse{They also do no wrong; 
they walk in his ways.}%
\verse{You have commanded\lebnote{Perhaps emphatic; Hebrew “you, you have commanded”}your precepts, 
that we should keep them diligently.}%
\verse{Oh that my ways were steadfast, 
to keep your statutes!}%
\verse{Then I would not be ashamed 
when I look to all your commands.}%
\verse{I will give you thanks with an upright heart, 
when I learn your righteous laws.}%
\verse{I will heed your statutes; 
do not utterly forsake me.}%
\verseWithHeading{Beth}{How can a young man keep his way pure? 
By taking heed according to your word.}%
\verse{With my whole heart I have sought you; 
do not let me stray from your commands.}%
\verse{In my heart I have hidden your word,\lebnote{Or “promise”}
so that I may not sin against you.}%
\verse{Blessed are you, O Adonai. 
Teach me your statutes.}%
\verse{With my lips I have told of 
all the ordinances of your mouth.}%
\verse{In the way of your testimonies I have rejoiced 
as I would over all riches.}%
\verse{I will meditate on your precepts 
and look at\lebnote{That is, “examine”}your ways.}%
\verse{I will take delight in your statutes; 
I will not forget your word.}%
\verseWithHeading{Gimel}{Deal bountifully with your servant, that I may live 
and heed your word.}%
\verse{Uncover my eyes, that I may look at\lebnote{That is, “examine”}
wonderful things from your law.}%
\verse{I am a sojourner on the earth;\lebnote{That is, “here for a short time”}
do not hide your commands from me.}%
\verse{My soul is crushed with longing 
for your ordinances at all times.}%
\verse{You rebuke the arrogant, the accursed, 
who stray from your commands.}%
\verse{\textit{Remove from}\lebnote{Literally “Roll off of”}me scorn and contempt, 
for I have kept your testimonies.}%
\verse{Even though princes sit and speak against me, 
your servant meditates on your statutes.}%
\verse{Your testimonies are my delight, 
\textit{my counselors}.\lebnote{Literally “the men of my counsel”}}%
\verseWithHeading{Daleth}{My soul clings to the dust; 
revive me according to your word.}%
\verse{I told of my ways, and you answered me; 
teach me your statutes.}%
\verse{Make me understand the way of your precepts, 
that I may meditate on your wonderful things.}%
\verse{My soul weeps\lebnote{Or “is sleepless”}because of grief; 
strengthen me according to your word.}%
\verse{Remove from me the deceptive way, 
and graciously give me your law.}%
\verse{I have chosen the faithful way; 
I have set your ordinances before me.}%
\verse{I cling to your testimonies; 
O Adonai, do not let me be put to shame.}%
\verse{I will run the way of your commands, 
for you will enlarge my heart.\lebnote{That is, “give greater understanding”}}%
\verseWithHeading{Heʼ}{Teach me, O Adonai, the way of your statutes, 
and I will keep it to the end.}%
\verse{Give me understanding, that I may keep your law 
and heed it with my whole heart.}%
\verse{Cause me to walk in the path of your commands, 
for I delight in it.}%
\verse{Incline my heart to your testimonies, 
and not to dishonest gain.}%
\verse{Turn away my eyes from looking at what is worthless; 
revive me in your ways.}%
\verse{Fulfill your word\lebnote{Or “your promise”}to your servant, 
which is to bring about a reverence for you.}%
\verse{Turn away my disgrace that I dread, 
for your ordinances are good.}%
\verse{Look, I have longed for your precepts; 
revive me in your righteousness.}%
\verseWithHeading{Waw}{Let your acts of loyal love also come to me, O Adonai, 
your salvation according to your word.\lebnote{Or “your promise”}}%
\verse{Then I will have an answer for the one who taunts me, 
for I trust your word.}%
\verse{And do not snatch the word of truth utterly from my mouth, 
for I hope in your ordinances.}%
\verse{So I will heed your law continually, 
forever and ever.}%
\verse{And I will go about \textit{freely},\lebnote{Literally “in the wide place”}
for I have sought your precepts.}%
\verse{And I will speak of your testimonies before kings, 
and will not be ashamed.}%
\verse{And I will take delight in your commands, 
which I love.}%
\verse{And I will lift up my hands\lebnote{Hebrew “palms”}to your commands, which I love, 
and I will meditate on your statutes.}%
\verseWithHeading{Zayin}{Remember your word to your servant, 
upon which you have caused me to hope.}%
\verse{This is my comfort in my misery: 
that your word\lebnote{Or “your promise”}preserves my life.}%
\verse{The arrogant utterly deride me; 
I have not turned aside from your law.}%
\verse{I remember your ordinances of old, O Adonai, 
and I take comfort.}%
\verse{Rage seizes me because of the wicked, 
those who forsake your law.}%
\verse{Your statutes have been my songs 
in the house of my sojourning.}%
\verse{I remember your name in the night, O Adonai, 
and I heed your law.}%
\verse{This has been mine, 
that I have kept your precepts.}%
\verseWithHeading{Heth}{Adonai is my portion; 
I intend to heed your words.}%
\verse{I \textit{seek your favor}\lebnote{Literally “entreat your face”}with my whole heart; 
be gracious to me according to your word.\lebnote{Or “your promise”}}%
\verse{I think about my ways, 
and turn my feet to your testimonies.}%
\verse{I hurry and do not delay 
to heed your commands.}%
\verse{The cords of the wicked surround me, 
but I do not forget your law.}%
\verse{In the middle of the night I rise to give you thanks, 
because of your righteous ordinances.}%
\verse{I am a companion of all who fear you 
and heed your precepts.}%
\verse{The earth, O Adonai, is full of your loyal love. 
Teach me your statutes.}%
\verseWithHeading{Teth}{You have dealt well with your servant, 
O Adonai, according to your word.}%
\verse{Teach me good discernment and knowledge, 
for I believe your commands.}%
\verse{Before I was afflicted, I was going astray, 
but now I heed your word.}%
\verse{You are good and do good; 
teach me your statutes.}%
\verse{The arrogant smear me with lies;\lebnote{Hebrew “lie”}
I keep your precepts with my whole heart.}%
\verse{Their heart is insensitive like fat;\lebnote{See Isa 6:10 for a similar idea, but a different word for “fat”}
As for me, I take delight in your law.}%
\verse{It is good for me that I was afflicted, 
so that I might learn your statutes.}%
\verse{The law of your mouth is better to me 
than thousands of gold and silver coins.}%
\verseWithHeading{Yod}{Your hands have made me and established me; 
give me understanding that I may learn your commands.}%
\verse{Those who fear you will see me and be glad, 
because I hope in your word.}%
\verse{I know, O Adonai, that your judgments are right, 
and that in faithfulness you afflicted me.}%
\verse{Please let your loyal love comfort me, 
according to your word\lebnote{Or “your promise”}to your servant.}%
\verse{Let your mercies come to me, that I may live, 
for your law is my delight.}%
\verse{Let the arrogant be put to shame, 
because they have wronged me with lies.\lebnote{Hebrew “lie”}
As for me, I will meditate on your precepts.}%
\verse{Let those who fear you turn to me, 
even those who know\lebnote{According to the reading tradition (\textit{Qere})}your testimonies.}%
\verse{May my heart be blameless in your statutes, 
so that I may not be ashamed.}%
\verseWithHeading{Kaph}{My soul languishes for your salvation; 
I hope in your word.}%
\verse{My eyes long for\lebnote{Or “long to see”}your word,\lebnote{Or “your promise”}
saying, “When will you comfort me?”}%
\verse{For I have become like a wineskin in the smoke, 
yet I have not forgotten your statutes.}%
\verse{How many are the days of your servant? 
When will you execute justice against those who persecute me?}%
\verse{The arrogant have dug pits\lebnote{That is, traps}for me, 
which is not according to\lebnote{Or “who are not in accord with”}your law.}%
\verse{All your commands are faithful. 
They persecute me deceitfully; help me!}%
\verse{They have almost destroyed me on the earth, 
but I have not forsaken your precepts.}%
\verse{According to your loyal love preserve me alive, 
so that I may heed the testimony from your mouth.}%
\verseWithHeading{Lamed}{Forever, O Adonai, 
your word \textit{is settled}\lebnote{Literally “stands firm”}in heaven.}%
\verse{Your faithfulness endures \textit{throughout all generations}.\lebnote{Literally “for a generation and a generation”}
You have established the earth and it stands.}%
\verse{By your ordinances they stand today, 
for all are your servants.}%
\verse{Unless your law had been my delight, 
then I would have perished in my misery.}%
\verse{I will not ever forget your precepts, 
for by them you revive me.\lebnote{Or “give me life”}}%
\verse{I am yours. Save me, 
for I have sought your precepts.}%
\verse{For me the wicked lie in wait to destroy me. 
I give attention to your testimonies.}%
\verse{I have seen a limit to every perfection;\lebnote{The meaning of this line is uncertain.}
your command is very broad.}%
\verseWithHeading{Mem}{How I love your law! 
The whole day it is my meditation.}%
\verse{Your commands make me wiser than my enemies, 
because they\lebnote{Hebrew “it is”}are ever with me.}%
\verse{I have more insight than all my teachers, 
for your testimonies are my meditation.}%
\verse{I understand more than the elders, 
for I keep your precepts.}%
\verse{I have restrained my feet from every evil way, 
so that I may heed your word.}%
\verse{I have not turned aside from your ordinances, 
for you have taught me.}%
\verse{How smooth\lebnote{Or “sweet”}are your words\lebnote{Hebrew “word”}to my palate, 
more than honey in my mouth.}%
\verse{From your precepts I get understanding, 
therefore I hate every false way.}%
\verseWithHeading{Nun}{Your word is a lamp to my feet 
and a light to my path.}%
\verse{I have sworn an oath and confirmed it, 
to heed \textit{your righteous ordinances}.\lebnote{Literally “the ordinances of your righteousness”}}%
\verse{I am very much afflicted; 
O Adonai, revive me according to your word.}%
\verse{Please accept the freewill offerings of my mouth, O Adonai, 
and teach me your ordinances.}%
\verse{My life is \textit{in danger}\lebnote{Literally “in my hand”}continually, 
yet I do not forget your law.}%
\verse{The wicked have laid a snare for me, 
yet I do not wander from your precepts.}%
\verse{I have taken as my own your testimonies forever, 
for they are the joy of my heart.}%
\verse{I have inclined my heart to do your statutes 
forever, to the end.}%
\verseWithHeading{Samek}{I hate \textit{the double-minded},\lebnote{Literally “the divided”}
but I love your law.}%
\verse{You are my hiding place and my shield; 
I hope in your word.}%
\verse{Turn aside from me, you evildoers, 
for I will keep the commands of my God.}%
\verse{Sustain me according to your word,\lebnote{Or “promise”}that I may live, 
and do not let me be ashamed of my hope.}%
\verse{Uphold me, that I may be delivered, 
and have regard for your statutes continually.}%
\verse{You reject all who stray from your statutes, 
for their deceit is a breach of faith.}%
\verse{You remove all the wicked of the earth like dross, 
therefore I love your testimonies.}%
\verse{My flesh trembles for fear of you, 
and I am afraid of your judgments.}%
\verseWithHeading{ʽAyin}{I have done justice and righteousness;\lebnote{Or “done what is proper and right”}
do not leave me to my oppressors.}%
\verse{Pledge good for your servant; 
do not let the arrogant oppress me.}%
\verse{My eyes long for\lebnote{Or “long to see”}your salvation, 
and your righteous word.\lebnote{Or “promise”}}%
\verse{Deal with your servant according to your loyal love, 
and teach me your statutes.}%
\verse{I am your servant; give me understanding, 
that I may know your statutes.}%
\verse{It is time for Adonai to act,\lebnote{Or “time to act, O Adonai”}
for they have broken your law.}%
\verse{Therefore I love your commands 
more than gold, even fine gold.}%
\verse{Therefore all your precepts of everything I regard as right; 
I hate every false way.}%
\verseWithHeading{Pe}{Your testimonies are miracles, 
therefore my soul keeps them.}%
\verse{The unfolding of your words gives light, 
giving understanding to the simple.}%
\verse{I open my mouth and pant, 
because I long for your commands.}%
\verse{Turn to me and be gracious to me, 
as is proper for those who love your name.}%
\verse{Direct my steps in your word,\lebnote{Or “through your promise”}
and do not let any iniquity gain power over me.}%
\verse{Redeem me from the oppression of humankind, 
that I may heed your precepts.}%
\verse{Shine your face on your servant, 
and teach me your statutes.}%
\verse{Streams of water flow down from my eyes, 
because people do not heed your law.}%
\verseWithHeading{Tsade}{You are righteous, O Adonai, 
and your ordinances are upright.}%
\verse{You have ordained your testimonies in righteousness, 
and in exceeding faithfulness.}%
\verse{My zeal consumes me, 
because my enemies have forgotten your words.}%
\verse{Your word is very pure, 
and your servant loves it.}%
\verse{I am small and despised, 
yet I do not forget your precepts.}%
\verse{Your righteousness is an everlasting righteousness, 
and your law is truth.}%
\verse{Trouble and anguish have found me; 
your commands delight me.}%
\verse{The righteousness of your testimonies is forever; 
give me understanding that I may live.}%
\verseWithHeading{Qoph}{I call with my whole heart; answer me, O Adonai. 
I will keep your statutes.}%
\verse{I call to you. Save me, 
and I will heed your testimonies.}%
\verse{I rise early in the dawn and cry for help; 
I hope in your word.\lebnote{According to the reading tradition (\textit{Qere})}}%
\verse{My eyes anticipate the night watches, 
that I may meditate on your word.}%
\verse{Hear my voice according to your loyal love; 
O Adonai, preserve my life according to your justice.}%
\verse{Those who pursue a wicked purpose draw near; 
they are far from your law.}%
\verse{You are near, O Adonai, 
and all your commands are truth.}%
\verse{Concerning your testimonies, I knew long ago 
that you have established them forever.}%
\verseWithHeading{Resh}{See my misery and rescue me, 
for I do not forget your law.}%
\verse{Plead my cause and redeem me; 
preserve my life according to your word.\lebnote{Or “promise”}}%
\verse{Salvation is far from the wicked, 
for they do not seek your statutes.}%
\verse{Your mercies are great, O Adonai; 
preserve my life according to your ordinances.}%
\verse{My persecutors and enemies are many, 
yet I do not turn aside from your testimonies.}%
\verse{I see the treacherous and I feel disgust, 
because they do not heed your word.}%
\verse{Consider\lebnote{Hebrew “See”}that I love your precepts; 
O Adonai, according to your loyal love preserve my life.}%
\verse{The \textit{whole}\lebnote{Literally “head”}of your word is truth, 
and your every righteous judgment endures forever.}%
\verseWithHeading{Sin/Shin}{Princes persecute me without cause, 
but my heart trembles because of your word.\lebnote{According to the reading tradition (\textit{Qere})}}%
\verse{I am rejoicing over your word, 
like one who finds great spoil.}%
\verse{I hate and abhor falsehood; 
I love your law.}%
\verse{I praise you seven times in the day 
for your righteous ordinances.}%
\verse{Great peace is for those who love your law, 
and they do not have a cause for stumbling.}%
\verse{I hope for your salvation, O Adonai, 
and do your commands.}%
\verse{My soul heeds your testimonies, 
and I love them exceedingly.}%
\verse{I heed your precepts and testimonies, 
because all my ways are before you.}%
\verseWithHeading{Taw}{Let my cry come before you, O Adonai; 
give me understanding according to your word.}%
\verse{Let my plea come before you; 
Deliver me according to your word.\lebnote{Or “promise”}}%
\verse{Let my lips pour out praise, 
because you teach me your statutes.}%
\verse{Let my tongue sing of your word,\lebnote{Or “promise”}
because all your commands are right.}%
\verse{Let your hand be my help,\lebnote{Hebrew “to help me”}
because I have chosen your precepts.}%
\verse{I long for your salvation, O Adonai, 
and your law is my delight.}%
\verse{Let my soul\lebnote{Or “my life”}live that it may praise you, 
and let your ordinances help me.}%
\verse{I have wandered like a lost sheep; seek your servant, 
because I do not forget your commands.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 120
\verseWithHeading{A song of ascents.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{In my distress I called to Adonai, 
and he answered me.}%
\verse{“Deliver my life, O Adonai, from lying lips, 
from a deceitful tongue.”}%
\verse{What shall be given to you,\lebnote{That is, by God against the false accuser}
and what more shall be done to you, 
deceitful tongue?}%
\verse{The sharpened arrows of a warrior, 
with burning charcoals from broom trees.}%
\verse{Woe to me, that I sojourn in Meshech, 
that I dwell among the tents of Kedar.}%
\verse{Too long my soul has had its dwelling 
near one who hates peace.}%
\verse{I am for peace, but when I speak, 
they are for war.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 121
\verseWithHeading{A song for the ascents.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{I lift up my eyes to the mountains; 
whence will my help come?}%
\verse{My help is from Adonai, 
maker of heaven and earth.}%
\verse{He will not allow your foot to be moved; 
he who protects you will not slumber.}%
\verse{Look, he will not slumber and he will not sleep— 
he who protects Israel.}%
\verse{Adonai is your protector; 
Adonai is your shade at your right hand.}%
\verse{The sun will not strike you by day, 
nor the moon by night.}%
\verse{Adonai will protect you from all evil;\lebnote{Or “all calamity”}
he will protect your life.}%
\verse{Adonai will protect your going out and your coming in 
from now until forever.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 122
\verseWithHeading{A song of ascents. Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{I rejoiced in those who said to me, 
“Let us go to the house\lebnote{Or “temple”}of Adonai.”}%
\verse{Our feet are standing 
within your gates, O Jerusalem—}%
\verse{Jerusalem that is built 
as a city that is joined together,}%
\verse{where the tribes go up, 
the tribes of Yah\lebnote{A shortened form of “Adonai”}as a testimony for Israel, 
to give thanks to the name of Adonai.}%
\verse{For there the thrones sit\lebnote{That is, rulers or judges on thrones}for judgment, 
thrones of David’s house.}%
\verse{Pray\lebnote{Hebrew “Ask”}for the peace of Jerusalem: 
“May those who love you be at ease.}%
\verse{May peace be within your walls, 
security within your palaces.”}%
\verse{For the sake of my brothers and my friends, 
I will say, “Peace be within you.”}%
\verse{For the sake of the house\lebnote{Or “temple”}of Adonai our God, 
I will seek your good.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 123
\verseWithHeading{A song of ascents.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{I lift up my eyes to you, 
the one enthroned\lebnote{Or “the one who dwells”}in the heavens.}%
\verse{Behold, as the eyes of servants 
look to the hand of their master,\lebnote{Hebrew “masters”}
as the eyes of a maid to the hand of her mistress, 
so our eyes look to Adonai our God, 
until he is gracious to us.}%
\verse{Be gracious to us, O Adonai, be gracious to us, 
for long enough we have had our fill of contempt.}%
\verse{For long enough our soul has had its fill of 
the derision of the self-confident, 
the contempt of the arrogant.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 124
\verseWithHeading{A song of ascents. Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{“If it had not been Adonai who was on our side,” 
do let Israel say,}%
\verse{“If it had not been Adonai who was on our side, 
when men rose up against us,}%
\verse{then they would have swallowed us alive, 
when their anger was kindled against us.}%
\verse{Then the waters would have flooded over us, 
the torrent would have passed over our soul.}%
\verse{Then over our soul would have passed 
the raging waters.”}%
\verse{Blessed be Adonai, 
who has not made\lebnote{Hebrew “given”}us prey for their teeth.}%
\verse{Our soul has escaped like a bird 
from the snare of fowlers. 
The snare is broken, and we have escaped.}%
\verse{Our help is in the name of Adonai, 
maker of heaven and earth.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 125
\verseWithHeading{A song of ascents.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Those who trust in Adonai 
are like Mount Zion, which cannot be moved, 
but abides forever.}%
\verse{As mountains are round about Jerusalem, 
so Adonai is round about his people, 
from now until forever.}%
\verse{For the scepter of wickedness shall not rest 
on the \textit{land}\lebnote{Literally “lot,” that is, land allotted}of the righteous, 
lest the righteous stretch out 
their hands to do wickedness.\lebnote{Hebrew “in wickedness”}}%
\verse{Do good, O Adonai, to the good, 
and to those upright in their hearts.}%
\verse{But regarding those who turn aside to their crooked ways, 
Adonai will lead them away with the evildoers. 
Peace be upon Israel.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 126
\verseWithHeading{A song of ascents.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{When Adonai restored the fortunes\lebnote{Or “captivity”}of Zion,\lebnote{Or “those returned of belonging to Zion”}
we were like dreamers.}%
\verse{Then our mouth was filled with laughter, 
and our tongue with rejoicing. 
Then they said among the nations, 
“Adonai has done great things for these people.”}%
\verse{Adonai has done great things for us; 
we are glad.}%
\verse{Restore, O Adonai, our fortunes\lebnote{Or “captivity”}
like the streams in the Negeb.\lebnote{Or “South”}}%
\verse{Those who sow with tears 
shall reap with rejoicing.}%
\verse{He who diligently goes out with weeping, 
carrying the seed bag, 
shall certainly come in with rejoicing, 
carrying his sheaves.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 127
\verseWithHeading{A song of ascents. Of Solomon.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Unless Adonai builds a house, 
its builders labor at it in vain. 
Unless Adonai guards a city, 
a guard watches in vain.}%
\verse{It is in vain for you who rise early and sit late, 
eating the bread of anxious toil, 
when thus he\lebnote{That is, Adonai}provides\lebnote{Hebrew “gives”}for his beloved in his sleep.\lebnote{Or “so he gives sleep to his beloved”}}%
\verse{Look, children\lebnote{Hebrew “sons”}are the heritage\lebnote{That is, gift}of Adonai; 
the fruit of the womb is a reward.}%
\verse{Like arrows in the hand of a warrior, 
so are the children\lebnote{Hebrew “sons”}of one’s youth.}%
\verse{Blessed is the man who fills his quiver with them.\lebnote{Or “whose quiver he Adonai fills with them”}
They shall not be put to shame 
when they speak with enemies at the gate.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 128
\verseWithHeading{A song of ascents.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Blessed is everyone who fears Adonai, 
who walks in his ways.}%
\verse{You will indeed eat of the labor of your hands; 
you will be happy and it will be well with you.}%
\verse{Your wife will be like a fruitful vine 
\textit{within}\lebnote{Literally “in the recesses of”}your house. 
Your children\lebnote{Hebrew “sons”}will be like olive shoots 
about your table.}%
\verse{Look, for thus shall a man be blessed 
\textit{who fears Adonai}.\lebnote{Literally “a fearer of Adonai”}}%
\verse{May Adonai bless you from Zion, 
that you may see the good of Jerusalem 
all the days of your life,}%
\verse{and that you may see your children’s children. 
May peace be upon Israel.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 129
\verseWithHeading{A song of ascents.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{“Too often\lebnote{Or “Greatly”}they have attacked me from my youth.” 
Let Israel say,}%
\verse{“Too often\lebnote{Or “Greatly”}they have attacked me from my youth, 
yet they have not prevailed against me.}%
\verse{On my back plowmen have plowed. 
They have made their furrows\lebnote{According to the reading tradition (\textit{Qere})}long.”}%
\verse{Adonai is righteous. 
He has cut the ropes\lebnote{That is, the yoke and tackle on a beast of burden}of the wicked.}%
\verse{Let all be put to shame and repulsed 
who hate Zion.}%
\verse{Let them be like grass on the housetops, 
that withers before it grows up,}%
\verse{with which a reaper cannot fill his hand, 
nor the binder of sheaves his \textit{arms},\lebnote{Literally “bosom”}}%
\verse{so that passersby do not say, 
“The blessing of Adonai be upon you. 
We bless you in the name of Adonai.”}%
\end{biblechapter}

\begin{biblechapter} % Psalms 130
\verseWithHeading{A song of ascents.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Out of the depths I call to you, O Adonai.}%
\verse{“O Lord, hear my voice. 
Let your ears be attentive 
to the voice of my supplications.}%
\verse{If you, O Yah,\lebnote{A shortened form of “Adonai”}should \textit{keep track of}\lebnote{Literally “keep”}iniquities, 
O Lord, who could stand?}%
\verse{But with you is forgiveness, 
so that you may be feared.”}%
\verse{I await Adonai; my soul awaits, 
and I wait\lebnote{Or “hope,” here and v. 7, though the Hebrew uses a different verbal stem in v. 7}for his word.}%
\verse{My soul waits for the Lord 
more than watchmen for the morning. 
Yes, more than watchmen for the morning.}%
\verse{O Israel, wait\lebnote{Or “hope”}for Adonai. 
For with Adonai there is loyal love, 
and with him there is abundant redemption.}%
\verse{And he will redeem Israel 
from all its iniquities.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 131
\verseWithHeading{A song of ascents. Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{My heart is not haughty nor my eyes \textit{arrogant}.\lebnote{Literally “raised”}
And I do not \textit{concern myself}\lebnote{Literally “walk about”}
with things too great and difficult for me.}%
\verse{Rather I have soothed and quieted my soul, 
like a weaned child with its mother, 
like the weaned child is my soul with me.}%
\verse{O Israel, hope in Adonai 
from now until forever.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 132
\verseWithHeading{A song of ascents.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{O Adonai, remember with regard to David 
all his affliction.\lebnote{Hebrew “his being afflicted”}}%
\verse{How he swore to Adonai, 
he vowed to the Mighty One of Jacob:}%
\verse{“I will not enter into the tent of my house, 
I will not go up to the couch of my bed,}%
\verse{I will give no sleep to my eyes, 
no slumber to my eyelids,}%
\verse{until I find a place for Adonai, 
a dwelling place for the Mighty One of Jacob.”}%
\verse{Look, we heard of it in Ephrathah; 
we found her in the fields of Jaar.\lebnote{Or “the fields of the woods”}}%
\verse{Let us go to his dwelling places;\lebnote{That is, the temple}
Let us worship at his footstool.}%
\verse{Arise, O Adonai, to your resting place, 
you and your mighty ark.}%
\verse{Let your priests clothe themselves with righteousness, 
and let your faithful sing for joy.}%
\verse{For the sake of your servant David, 
do not turn away the face of your anointed one.}%
\verse{Adonai swore in truth to David; 
he will not turn back from it. 
“One from the fruit of your \textit{body}\lebnote{Literally “belly”}
I will set on your throne.}%
\verse{If your sons will heed my covenant 
and my testimonies that I will teach them, 
their sons also forever 
will sit on your throne.”}%
\verse{For Adonai has chosen Zion; 
he has desired it for his habitation.}%
\verse{“This is my resting place forever; 
here I will dwell, for I have desired it.}%
\verse{I will bless its provisions abundantly; 
I will satisfy its poor with bread.}%
\verse{I will also clothe its priests with salvation, 
and its faithful will shout exuberantly for joy.}%
\verse{There I will cause a horn to grow for David; 
I will set a lamp for my anointed one.}%
\verse{I will clothe his enemies with shame, 
but on him his crown will flourish.”\lebnote{Or “shine”}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 133
\verseWithHeading{A song of ascents. Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Look, how good and how pleasant it is 
when brothers dwell together in unity.}%
\verse{It is like the fragrant oil upon the head, 
running down upon the beard, the beard of Aaron, 
that runs down upon the edges of his robes.}%
\verse{It is like the dew of Hermon that runs down 
upon the mountains of Zion, 
because there Adonai commanded the blessing— 
life forever.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 134
\verseWithHeading{A song of ascents.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Look, bless Adonai, all you the servants of Adonai, 
who \textit{serve}\lebnote{Literally “stand”}in the house\lebnote{Or “temple”}of Adonai by night.}%
\verse{Lift up your hands in the sanctuary 
and bless Adonai.}%
\verse{May Adonai bless you from Zion, 
he who made heaven and earth.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 135
\verseWithHeading{Praise to God for His Power and Redemption}{Praise Yah!\lebnote{Hebrew \textit{hallelujah}}Praise the name of Adonai. 
Praise him, O servants of Adonai,}%
\verse{who stand in the house\lebnote{Or “temple”}of Adonai, 
in the courts of the house\lebnote{Or “temple”}of our God.}%
\verse{Praise Yah,\lebnote{A shortened form of “Adonai”}for Adonai is good. 
Sing praises concerning his name, for it is pleasant.}%
\verse{For Yah has chosen Jacob for himself, 
Israel as his special possession.}%
\verse{For I know that Adonai is great, 
and our Lord is greater than all gods.}%
\verse{All that Adonai desires, he does, 
in heaven and on earth, 
in the seas and all the depths.}%
\verse{He causes clouds to rise from the edge of the earth; 
he makes lightnings for the rain; 
he brings out the wind from his storehouses.}%
\verse{Who struck the firstborn of Egypt, 
both of humans\lebnote{Hebrew “human”}and animals.\lebnote{Hebrew “animal”}}%
\verse{He sent signs and wonders into your midst, O Egypt, 
against Pharaoh and all his servants.}%
\verse{Who struck many nations 
and killed mighty kings—}%
\verse{Sihon the king of the Amorites, 
and Og the king of Bashan, 
and all the kingdoms of Canaan—}%
\verse{and he gave their land as an inheritance, 
an inheritance for his people Israel.}%
\verse{O Adonai, your name endures forever, 
your \textit{renown}\lebnote{Literally “remembrance”}\textit{throughout all generations},\lebnote{Literally “for a generation and a generation”}}%
\verse{for Adonai will plead the cause of his people 
and will have compassion on his servants.}%
\verse{The idols of the nations are silver and gold, 
the work of the hands of humankind.}%
\verse{They have mouths, but cannot speak; 
they have eyes, but cannot see;}%
\verse{they have ears, but cannot hear; 
there is not even breath in their mouths.}%
\verse{Those who make them become like them, 
as does everyone who trusts in them.}%
\verse{O house of Israel, bless Adonai. 
O house of Aaron, bless Adonai.}%
\verse{O house of Levi, bless Adonai. 
You who fear Adonai, bless Adonai.}%
\verse{Blessed be Adonai from Zion, 
who dwells in Jerusalem. Praise Yah!\lebnote{Hebrew \textit{hallelujah}}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 136
\verseWithHeading{Praise to God for His Creation and Deliverance}{Give thanks to Adonai, for he is good, 
for his loyal love endures forever.}%
\verse{Give thanks to the God of gods, 
for his loyal love endures forever.}%
\verse{Give thanks to the Lord of lords, 
for his loyal love endures forever.}%
\verse{To him who alone does great wonders, 
for his loyal love endures forever.}%
\verse{To him who made the heavens with skill,\lebnote{Or “understanding”}
for his loyal love endures forever.}%
\verse{To him who spread out the earth above the waters, 
for his loyal love endures forever.}%
\verse{To him who made the great lights, 
for his loyal love endures forever.}%
\verse{The sun to rule the day, 
for his loyal love endures forever.}%
\verse{The moon and stars to rule the night, 
for his loyal love endures forever.}%
\verse{To him who struck Egypt through their firstborn, 
for his loyal love endures forever.}%
\verse{And he brought Israel out from among them, 
for his loyal love endures forever.}%
\verse{With a strong hand and an outstretched arm, 
for his loyal love endures forever.}%
\verse{To him who divided the \textit{Red Sea}\lebnote{Literally “sea of reed”}in two, 
for his loyal love endures forever.}%
\verse{And he let Israel cross over through the midst of it, 
for his loyal love endures forever.}%
\verse{But he tossed Pharaoh and his army into the \textit{Red Sea},\lebnote{Literally “sea of reed”}
for his loyal love endures forever.}%
\verse{To him who led his people through the wilderness, 
for his loyal love endures forever.}%
\verse{To him who struck great kings, 
for his loyal love endures forever.}%
\verse{And he killed mighty kings, 
for his loyal love endures forever.}%
\verse{Sihon the king of the Amorites, 
for his loyal love endures forever.}%
\verse{And Og the king of Bashan, 
for his loyal love endures forever.}%
\verse{And he gave their land as an inheritance, 
for his loyal love endures forever.}%
\verse{An inheritance to Israel his servant, 
for his loyal love endures forever.}%
\verse{Who remembered us in our low estate, 
for his loyal love endures forever.}%
\verse{And he rescued us from our enemies, 
for his loyal love endures forever.}%
\verse{The one who gives food to all flesh, 
for his loyal love endures forever.}%
\verse{Give thanks to the God of heaven, 
for his loyal love endures forever.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 137
\verseWithHeading{Lament During the Babylonian Exile}{By the rivers of Babylon, 
there we sat, yes, we wept, 
when we remembered Zion.}%
\verse{On the willows\lebnote{Species of tree uncertain}in her midst, 
we hung up our lyres.}%
\verse{For there our captors asked of us 
words of a song, 
and our tormentors\lebnote{Only occurring once in the MT, this interpretation is uncertain. The LXX’s “carried us captive” is a guess. If related to the Hebrew root \textit{yll} it might have the idea of “mockers/yammerers”}asked of us jubilation, 
“Sing for us from a song of Zion.”}%
\verse{How could we sing the song of Adonai 
in \textit{a foreign land}?\lebnote{Literally “a land of a foreigner”}}%
\verse{If I forget you, O Jerusalem, 
let my right hand forget.\lebnote{That is, how to play a musical instrument}}%
\verse{Let my tongue cling to the roof of my mouth, 
if I do not remember you, 
if do not I exalt Jerusalem 
above my highest joy.}%
\verse{Remember, O Adonai, against the sons of Edom 
the day of Jerusalem, 
the ones who said, “Lay it bare! Lay it bare 
to its foundation!”}%
\verse{O daughter of Babylon, about to be devastated, 
happy shall be he who pays back to you 
\textit{what you paid out to us}.\lebnote{Literally “with what you treated us”}}%
\verse{Happy shall be he who seizes 
and smashes your children 
against the rock.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 138
\verseWithHeading{Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{I give you thanks with my whole heart; 
before the gods I sing your praise.}%
\verse{I bow down toward your holy temple 
and give thanks to your name, 
because of your loyal love and faithfulness, 
for you have magnified your word\lebnote{Or “promise”}according to all your name.\lebnote{Or “beyond all your name”}}%
\verse{\textit{When}\lebnote{Literally “In the day”}I called and you answered me, 
you emboldened me in my soul with strength.}%
\verse{All the kings of the earth 
will praise you, O Adonai, 
when they have heard the words\lebnote{Or “heard about the words”}of your mouth,}%
\verse{and they will sing of the ways of Adonai, 
because Adonai’s glory is great.}%
\verse{Though Adonai is high, yet he sees the lowly, 
but the proud he perceives\lebnote{Hebrew “knows”}at a distance.}%
\verse{Though I walk in the midst of trouble, 
you preserve me alive. 
You stretch out your hand against the anger of my enemies, 
and you save me with your right hand.}%
\verse{Adonai accomplishes things for my benefit. 
O Adonai, your loyal love endures forever; 
do not abandon the works of your hands.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 139
\verseWithHeading{For the music director. Of David. A psalm.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{O Adonai, you have searched me, and you know me.}%
\verse{You know my sitting down and my rising up. 
You understand my thought from afar.}%
\verse{You search out\lebnote{Hebrew “winnow”; or perhaps from another root, “measure”}my wandering and my lying down, 
and are acquainted with all my ways.}%
\verse{For there is not a word yet on my tongue, 
but behold, O Adonai, you know it completely.}%
\verse{You barricade me behind and in front, 
and set your hand upon me.}%
\verse{Such knowledge is too wonderful for me. 
It is set high; I cannot prevail against it.}%
\verse{Where I can go from your Spirit, 
or where can I flee from your presence?}%
\verse{If I ascend to heaven, there you are, 
and if I make my bed in Sheol, look! There you are.}%
\verse{If I lift up the wings of the dawn, 
and I alight on the far side of the sea,}%
\verse{even there your hand would lead me, 
and your right hand would hold me fast.}%
\verse{And if I should say, “Surely darkness will cover me, 
and the light around me will be as night,”}%
\verse{even the darkness is not too dark for you,\lebnote{Or “does not obscure from you”}
and the night shines as the day— 
the darkness and the light are alike for you.}%
\verse{Indeed you created my \textit{inward parts};\lebnote{Literally “kidneys”}
you wove me in my mother’s womb.}%
\verse{I praise you, because I am fearfully 
and wonderfully made.\lebnote{Or “I am made wonderful fearfully”}
Wonderful are your works, 
and my soul knows it well.}%
\verse{My \textit{frame}\lebnote{Literally “bone”}was not hidden from you, 
when I was created secretly,\lebnote{Or “in the secret place”}
and intricately woven 
in the depths of the earth.}%
\verse{Your eyes saw my embryo,\lebnote{Hebrew “unformed substance”}
and in your book they all were written— 
days fashioned for me when there was not one of them.}%
\verse{And to me, how precious\lebnote{Or “difficult”}are your thoughts, O God; 
how vast is their sum.}%
\verse{If I should count them, 
they would outnumber the sand. 
I awaken, and I am still with you.}%
\verse{If only you would kill the wicked, O God— 
so get away from me, you \textit{bloodthirsty men}—\lebnote{Literally “men of bloods”}}%
\verse{who speak against you deceitfully. 
Your enemies take your name in vain.}%
\verse{Do I not hate those who hate you, O Adonai? 
And do I not loathe those who rise up against you?}%
\verse{I hate them with a complete hatred; 
they have become my enemies.}%
\verse{Search me, O God, and know my heart; 
test me and know my anxious thoughts.}%
\verse{And see if there is in me \textit{the worship of false gods},\lebnote{Literally “a way of an idol”}
and lead me in the way everlasting.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 140
\verseWithHeading{For the music director. A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{Rescue me, O Adonai, from evil men.\lebnote{The several uses of the singular “man” in this psalm are probably collective}
Preserve me from violent men,}%
\verse{who plan evil things in their heart. 
They stir up wars continually.}%
\verse{They sharpen their tongue as sharp as a snake’s; 
the venom of a viper is under their lips. \textit{Selah}}%
\verse{Protect me, O Adonai, from the hands of the wicked. 
Preserve me from violent men, 
who have planned to \textit{make me stumble}.\lebnote{Literally “push my steps”}}%
\verse{The proud have hidden a trap for me, and cords. 
They have spread out a net along the \textit{side}\lebnote{Literally “hand”}of the path. 
They have set snares for me. \textit{Selah}}%
\verse{I say to Adonai, “You are my God.” 
Listen, O Adonai, to the voice of my supplications.}%
\verse{O Adonai, my Lord, the strength of my salvation, 
you have covered my head in the day of battle.}%
\verse{Do not grant, O Adonai, the desires of the wicked. 
Do not allow them to attain their\lebnote{Hebrew “his”}plan, lest they be exalted. \textit{Selah}}%
\verse{With respect to the head of those who surround me, 
may the harm of their lips cover them.}%
\verse{Let burning coals fall on them; 
let them be dropped into the fire, 
into bottomless pits from which they cannot rise.}%
\verse{Do not let a \textit{slanderer}\lebnote{Literally “man of tongue”}be established in the land; 
as for the man of violence, let evil hunt him quickly.}%
\verse{I know\lebnote{According to the reading tradition (\textit{Qere})}that Adonai will maintain the cause of the afflicted, 
and justice for the needy.}%
\verse{Surely the righteous will give thanks to your name; 
the upright will dwell in your presence.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 141
\verseWithHeading{A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{I call on you, O Adonai; hasten to me. 
Listen to my voice when I call to you.}%
\verse{Let my prayer be set before you as incense, 
the lifting up of my palms as the evening offering.}%
\verse{Set a guard, O Adonai, over my mouth; 
keep watch over the door of my lips.}%
\verse{Do not incline my heart to any evil thing, 
to practice wicked deeds 
with men who do iniquity; 
and do not let me eat of their delicacies.}%
\verse{Let a righteous one strike me in kindness, 
and let him chasten me. 
It is oil for my head;\lebnote{Or “It is choice oil”}let not my head refuse. 
For still my prayer is against their evil deeds.}%
\verse{When their judges are thrown down the sides of a cliff,\lebnote{Or “are dropped into the hands of the Rock,” that is, Adonai. The interpretation of vv. 6–7 is uncertain.}
then they will understand that my words were pleasant.}%
\verse{As when one plows and breaks up the earth, 
so our bones are scattered at the mouth of Sheol.}%
\verse{But\lebnote{Or “Because”}my eyes are toward you, O Adonai, my Lord; 
I have taken refuge in you. Do not lay bare my soul.\lebnote{Or “Do not raze my life”}}%
\verse{Protect me from the \textit{grasp}\lebnote{Literally “hand”}of the trap they have laid for me, 
and from the snares of evildoers.}%
\verse{Let the wicked fall into their\lebnote{Hebrew “his”}nets, 
while I escape altogether.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 142
\verseWithHeading{A maskil of David. 
When he was in the cave. A prayer.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm; the English verse number is reduced by one}}{I cry out with my voice to Adonai; 
I implore favor with my voice to Adonai.}%
\verse{I pour out my complaint before him; 
I declare my trouble before him.}%
\verse{When my spirit faints within me, 
you know my way. 
On the path where I walk, 
they have hidden a trap for me;}%
\verse{look to my right and see. 
There is no one looking out for me; 
there is no escape for me;\lebnote{Hebrew “a refuge has perished from me”}
no one cares for my soul.\lebnote{Or “life”}}%
\verse{I cry out to you, O Adonai. 
I say, “You are my refuge, 
my portion in the land of the living.”}%
\verse{Attend to my cry, 
for I am brought very low. 
Deliver me from my pursuers, 
for they are stronger than I.}%
\verse{Bring me\lebnote{Hebrew “my soul”}out of prison, 
that I may give thanks to your name. 
The righteous will encircle me, 
because you will deal bountifully with me.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 143
\verseWithHeading{A psalm of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{O Adonai, hear my prayer; 
listen to my supplications. 
In your faithfulness answer me, 
and in your righteousness.}%
\verse{And do not enter into judgment with your servant, 
because no one alive is righteous before you.}%
\verse{For the enemy has pursued my soul; 
he has crushed my life to the ground. 
He has made me dwell in dark places 
like those long dead.}%
\verse{And so my spirit grows faint within me; 
my heart \textit{within}\lebnote{Literally “in the midst of”}me is desolate.}%
\verse{I remember the days of long ago; 
I meditate on all your doings.\lebnote{Hebrew “doing”}
I muse on the labor of your hands.}%
\verse{I stretch out my hands to you; 
my soul longs for you like a dry land. \textit{Selah}}%
\verse{Quickly answer me, O Adonai; 
my spirit fails. 
Do not hide your face from me, 
or I will become 
like those descending to the pit.}%
\verse{Cause me to hear your loyal love in the morning, 
for I trust you. 
Cause me to know the way that I should go, 
for I lift up my soul to you.}%
\verse{Deliver me from my enemies, O Adonai. 
I take refuge in you.\lebnote{This reading for the difficult Hebrew “I cover to you” is suggested by the LXX}}%
\verse{Teach me to do your will, for you are my God; 
your Spirit is good. 
Lead me onto level ground.}%
\verse{For your name’s sake, O Adonai, preserve my life; 
in your righteousness bring me\lebnote{Hebrew “my soul”}out of trouble.}%
\verse{And in your loyal love destroy my enemies, 
and exterminate all the adversaries of my soul, 
for I am your servant.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 144
\verseWithHeading{Of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{Blessed be Adonai, my rock, 
the one who trains my hands for battle, 
my fingers for war—}%
\verse{my loyal love and my fortress, 
my stronghold and my deliverer, 
my shield, and one in whom I take refuge, 
the one who subdues peoples\lebnote{Reading with many Hebrew manuscripts and versions; MT reads “my people”}under me.}%
\verse{O Adonai, what is humankind that you take knowledge of him, 
or the son of man that you take thought of him?}%
\verse{Humankind is like a breath, 
his days like a passing shadow.}%
\verse{O Adonai, bow the heavens and come down; 
touch the mountains so that they smoke.}%
\verse{Flash forth lightning and scatter them; 
dispatch your arrows and rout them.}%
\verse{Stretch out your hands from on high; 
Rescue me and deliver me from many waters, 
from the hand of foreigners,}%
\verse{whose mouth speaks falsely, 
and their right hand is a false right hand.}%
\verse{O God, I will sing a new song to you. 
With a lyre of ten strings I will sing praise to you,}%
\verse{who gives victory to kings, 
who rescues David his servant 
from the evil sword.}%
\verse{Rescue me and deliver me 
from the hand of foreigners, 
whose mouth speaks falsely, 
and whose right hand is a false right hand,}%
\verse{that our sons may be like plants, 
full grown in their youth, 
our daughters like corner pillars, 
carved in the style of a palace,}%
\verse{that our granaries may be full, 
providing produce of all kinds,\lebnote{Hebrew “from kind to kind”}
that our sheep may produce by the thousands, 
by the tens of thousands in our open fields,}%
\verse{that our cattle may be pregnant; 
that there be no breach in our walls, 
and no going out in exile, 
and no outcry in our plazas.}%
\verse{Blessed are the people who have it thus. 
Blessed are the people whose God is Adonai.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 145
\verseWithHeading{A praise of David.\lebnote{The Hebrew Bible counts the superscription as the first verse of the psalm}}{I will exalt you, my God, O king, 
and I will bless\lebnote{That is, praise; here and v. 2}your name forever and ever.}%
\verse{Every day I will bless you 
and praise your name forever and ever.}%
\verse{Great is Adonai, and very worthy of praise, 
and his greatness is unsearchable.}%
\verse{One generation will laud your works to another, 
and will declare your mighty deeds.}%
\verse{On the splendor of the glory of your majesty, 
and on your wonderful deeds, I will meditate.}%
\verse{And they will speak of the power of your awesome deeds,\lebnote{That is, deeds that evoke a fearful reverence}
and I will tell of your greatness.\lebnote{According to the reading tradition (\textit{Qere})}}%
\verse{They will utter \lebnote{Hebrew “pour forth”}the renown of your abundant goodness, 
and they will proclaim with joy your righteousness.}%
\verse{Adonai is gracious and compassionate, 
\textit{slow to anger}\lebnote{Literally “long of nostrils”}and great in loyal love.}%
\verse{Adonai is good to all, 
and his mercies are over all his works.}%
\verse{All your works will praise you, O Adonai, 
and your faithful ones will bless\lebnote{That is, praise}you.}%
\verse{They will speak of the glory of your kingdom, 
and talk of your power,}%
\verse{to make known to the children of humankind his mighty deeds, 
and the glory of the majesty of his kingdom.}%
\verse{Your kingdom is an everlasting kingdom, 
and your dominion endures \textit{throughout all generations}.\lebnote{Literally “for a generation and a generation”}}%
\verse{Adonai upholds all who are falling, 
and raises up all who are bowed down.}%
\verse{The eyes of all look hopefully to you, 
and you give them their food in due time,}%
\verse{opening your hand, 
and satisfying the desire of every living creature.}%
\verse{Adonai is righteous in all his ways, 
and loyal in all his works.}%
\verse{Adonai is near to all who call on him, 
to all who call on him in truth.}%
\verse{He fulfills the desire of those who fear him; 
he also hears their cry for help, and saves them.}%
\verse{Adonai protects all those who love him, 
but all the wicked he will exterminate.}%
\verse{My mouth will speak the praise of Adonai, 
and all flesh will bless his holy name 
forever and ever.}%
\end{biblechapter}

\begin{biblechapter} % Psalms 146
\verseWithHeading{Praise to Adonai for His Help}{Praise Yah.\lebnote{Hebrew \textit{hallelujah}}
Praise Adonai, O my soul.}%
\verse{I will praise Adonai while I live; 
I will sing praises to my God while I am still alive.}%
\verse{Do not place trust in princes, 
in a son of humankind with whom there is no deliverance.}%
\verse{His breath departs; he returns to his plot;\lebnote{A Hebrew wordplay: “humankind,” \textit{adam} (v. 3), returns to his ground/plot, \textit{adamah}}
on that day his plans perish.}%
\verse{Blessed is the one whose help is the God of Jacob, 
whose hope is on Adonai as his God,}%
\verse{who made heaven and earth, 
the sea and all that is in them, 
the one who keeps faith\lebnote{Or “remains faithful”}forever,}%
\verse{who executes justice for the oppressed, 
who gives food for the hungry. 
Adonai sets prisoners free;}%
\verse{Adonai opens the eyes of the blind; 
Adonai raises up those bowed down; 
Adonai loves the righteous;}%
\verse{Adonai protects\lebnote{Or “watches over”}the strangers. 
He helps up the orphan and the widow, 
but the way of the wicked he \textit{thwarts}.\lebnote{Literally “makes crooked”}}%
\verse{Adonai will reign forever, 
Your God, O Zion, \textit{throughout all generations}.\lebnote{Literally “for a generation and a generation”}
Praise Yah.\lebnote{Hebrew \textit{hallelujah}}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 147
\verseWithHeading{Praise to Adonai for His Providence}{Praise Yah.\lebnote{Hebrew \textit{hallelujah}; here and v. 20}
For it is good to sing praises to our God; 
for it is pleasant; praise is fitting.}%
\verse{Adonai is building Jerusalem; 
he gathers the scattered ones of Israel.}%
\verse{He is the one who heals the brokenhearted, 
and binds up their wounds.}%
\verse{He counts the number of the stars; 
he gives names to all of them.}%
\verse{Great is our Lord, and abundant in power; 
his understanding is unlimited.}%
\verse{Adonai helps the afflicted up; 
he brings the wicked down to the ground.}%
\verse{Sing to Adonai with thanksgiving; 
sing praises to our God with lyre,}%
\verse{who covers the heavens with clouds, 
who provides rain for the earth, 
who causes grass to grow on the mountains.}%
\verse{He gives to the animal its food, 
and to the young ravens that cry.}%
\verse{He does not delight in the strength of the horse; 
he takes no pleasure in the legs of the man.}%
\verse{Adonai takes pleasure in those who fear him, 
the ones who hope\lebnote{That is, “wait expectantly”}for his loyal love.}%
\verse{Laud Adonai, O Jerusalem; 
praise your God, O Zion,}%
\verse{for he strengthens the bars of your gates. 
He blesses your children within you;}%
\verse{he makes your border peaceful; 
he satisfies you with the \textit{finest of wheat}.\lebnote{Literally “the fat of the wheat”}}%
\verse{He sends out his command\lebnote{Or “utterance”}to the earth; 
his word runs swiftly.}%
\verse{He gives snow like wool; 
he scatters frost like ashes;}%
\verse{he throws his hail like crumbs. 
Who can stand before his cold?\lebnote{With slight alteration of the Hebrew, the line might read “before his cold water stands”}}%
\verse{He sends out his word and melts them; 
he blows his breath, the water flows.}%
\verse{He declares his word to Jacob, 
his statutes and his ordinances to Israel.}%
\verse{He has not done so for any nation, 
and they do not know his ordinances. 
Praise Yah.\lebnote{Hebrew \textit{hallelujah}}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 148
\verseWithHeading{Let All Creation Praise Adonai}{Praise Yah.\lebnote{Hebrew \textit{hallelujah}}
Praise Adonai from the heavens; 
praise him in the heights.}%
\verse{Praise him, all his angels; 
praise him, all his hosts.\lebnote{According to the reading tradition (\textit{Qere})}}%
\verse{Praise him, sun and moon; 
praise him, all stars of light.}%
\verse{Praise him, \textit{highest heavens},\lebnote{Literally “heavens of the heavens”}
and waters above the heavens.}%
\verse{Let them praise the name of Adonai, 
because he commanded and they were created.}%
\verse{And he put them in place forever and ever, 
by a decree he gave that will not pass away.}%
\verse{Praise Adonai from the earth— 
great sea creatures and all deeps,}%
\verse{fire and hail, snow and cloud,\lebnote{The meaning of “cloud” is uncertain}
stormy wind doing his bidding,\lebnote{Hebrew “his word”}}%
\verse{the mountains and all hills, 
fruit trees and all cedars,}%
\verse{beasts and all cattle,\lebnote{That is, wild and domesticated}
creeping things and \textit{flying birds},\lebnote{Literally “bird of wing”}}%
\verse{kings of the earth and all peoples, 
princes and all judges of the earth,}%
\verse{young men and young women as well, 
the old together with the young—}%
\verse{let them praise the name of Adonai, 
because his name alone is is exalted. 
His splendor is above earth and heavens.}%
\verse{And he has raised high a horn for his people, 
praise for all his faithful— 
for the children of Israel, 
a people close to him. 
Praise Yah.\lebnote{Hebew \textit{hallelujah}}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 149
\verseWithHeading{Praise to God for His Future Judgment}{Praise Yah.\lebnote{Hebrew \textit{hallelujah}}
Sing to Adonai a new song, 
his praise in the assembly of the faithful.}%
\verse{Let Israel be glad in its maker; 
let the children of Zion rejoice in their king.}%
\verse{Let them praise his name with dancing; 
let them sing praises to him with tambourine and lyre.}%
\verse{For Adonai takes pleasure in his people; 
he glorifies the afflicted with salvation.}%
\verse{Let the faithful exult in glory; 
let them sing for joy on their beds.}%
\verse{Let the extolling of God be in their throat, 
and a double-edged sword in their hand,}%
\verse{to execute vengeance on the nations 
and punishment on the peoples,\lebnote{Reading against the MT and with the LXX}}%
\verse{to bind their kings with chains 
and their \textit{nobles}\lebnote{Literally “honorable ones”}with fetters of iron,}%
\verse{to execute on them the judgment that is decreed.\lebnote{Hebrew “written”}
This will be honor for all his faithful ones; 
Praise Yah.\lebnote{Hebrew \textit{hallelujah}}}%
\end{biblechapter}

\begin{biblechapter} % Psalms 150
\verseWithHeading{Let Everything Praise Adonai}{Praise Yah.\lebnote{Hebrew \textit{hallelujah}}
Praise God in his sanctuary; 
praise him in his mighty firmament.}%
\verse{Praise him for his mighty deeds; 
praise him according to the abundance of his greatness.}%
\verse{Praise him with blast of horn; 
praise him with harp and lyre.}%
\verse{Praise him with tambourine and dancing; 
praise him with strings and flute.}%
\verse{Praise him with sounding cymbals; 
praise him with clashing cymbals.}%
\verse{Every breathing thing, 
let it praise Yah.\lebnote{A shortened form of “Adonai”}
Praise Yah.\lebnote{Hebrew \textit{hallelujah}}}%
\end{biblechapter}

