\biblebook{Zephaniah}

\begin{biblechapter} % Zephaniah 1
\verseWithHeading{Judah’s Impending Judgment} The word of Adonai that came to Zephaniah the son of Cushi, the son of Gedaliah, the son of Amariah, the son of Hezekiah, in the days of Josiah son of Amon, king of Judah.
\verse “I will surely destroy everything 
from the face of the earth”—a declaration of Adonai.
\verse “I will destroy humanity and beast; 
I will destroy the birds of the sky 
and the fish of the sea, 
and the stumbling blocks with the wicked. 
And I will cut off humankind 
from the face of the earth”—a declaration of Adonai.
\verse “And I will stretch out my hand against Judah 
and against all the inhabitants of Jerusalem. 
And I will cut off from this place 
the remnant of Baal, 
and the name of idolatrous priests with the priests,
\verse and those who bow down on the rooftops 
to the host of heaven, 
and those who bow down, swearing to Adonai 
but also swearing by Milkom,
\verse and those who turned back from \textit{following} Adonai, 
and who did not seek Adonai, 
and did not inquire of him.”
\verse Be silent before the Lord Adonai, 
for the day of Adonai draws near, 
for Adonai has provided a sacrifice 
and has consecrated his guests.
\verse And it shall be that on the day of the sacrifice of Adonai, 
I will punish the officials 
and the sons of the king 
and those who dress in foreign clothing.
\verse And on that day I will punish 
all who leap over the threshold, 
who fill up the house of their master 
with violence and deceit.
\verse “And there shall be on that day”—a declaration of Adonai— 
“a \textit{loud outcry} from the Fish Gate, 
and a wailing from the Second District, 
and a loud crashing from the hills.
\verse The inhabitants of the Mortar shall wail, 
for all the traders have perished; 
all who trade with silver have been cut off.
\verse And it shall be that at that time 
I will search Jerusalem with lamps, 
and I will punish the men 
\textit{whose senses are dulled from drinking}, 
who say in their hearts, 
‘Adonai will not do good, nor will he do evil.’
\verse Then their wealth shall be as plunder, 
and their homes as desolation. 
And they shall build their houses and not inhabit them, 
and they shall plant vineyards and not drink their wine.”
\verse The great day of Adonai draws near; 
it is near and coming very swiftly. 
The sound of the day of Adonai is bitterness; 
there, a warrior cries out!
\verse That day will be a day of wrath, 
a day of trouble and distress, 
a day of disaster and ruination, 
a day of darkness and deep gloom, 
a day of clouds and thick clouds,
\verse a day of trumpet and trumpet blast 
against the fortified cities 
and against the high corner towers.
\verse And I will bring distress to all humankind, 
and they shall walk about like the blind, 
for they sinned against Adonai; 
their blood shall be poured out like dust 
and their entrails like dung.
\verse Moreover, their silver and their gold 
will not be able to save them 
on the day of the wrath of Adonai. 
And in the fire of his zeal, 
the whole land shall be consumed, 
for a terrifying end he shall make 
for all the inhabitants of the land.
\end{biblechapter}

\begin{biblechapter} % Zephaniah 2
\verseWithHeading{Judgment of Judah’s Enemies} Gather yourselves together! Now gather together, 
O nation having no shame!
\verse Before the birth of the decree, 
the day flies away like chaff; 
before the fierce anger of Adonai overtakes you; 
before the day of the anger of Adonai overtakes you.
\verse Seek Adonai, all you afflicted of the land 
who have fulfilled his law; 
seek righteousness; 
seek humility. 
Perhaps you will be concealed 
on the day of the anger of Adonai.
\verse For Gaza will be abandoned, 
and Ashkelon as a desolation. 
As for Ashdod, at noon they will drive her away, 
and Ekron shall be uprooted.
\verse Woe to you, inhabitants of the region of the sea, 
people of the Kerethites! 
The word of Adonai is against you, 
O Canaan, land of the Philistines. 
I will destroy you \textit{until there is no inhabitant}.
\verse And you, O region of the sea, shall become 
pastures and meadows for shepherds 
and sheep pens for the flocks.
\verse And it shall become a region 
for the remnant of the house of Judah; 
upon them they shall graze, 
and in the houses of Ashkelon 
they shall lie down in the evening. 
For Adonai their God shall be mindful of them, 
and he will restore their fortunes.
\verse I have heard the reproaches of Moab 
and the scorning of the \textit{Ammonites}, 
with which they have taunted my nation 
and made boasts against their territory.
\verse “Therefore, as I live”—a declaration of Adonai of hosts, 
the God of Israel— 
“Moab shall be as Sodom, 
and the \textit{Ammonites} like Gomorrah, 
a place of weeds and salt pits 
and a desolation forever. 
The remnant of my people shall plunder them, 
and the remainder of my nation shall possess them.”
\verse This shall be for them in place of their pride 
because they mocked and boasted 
against the people of Adonai of hosts.
\verse Adonai will be awesome against them, 
for he will destroy all the gods of the earth 
and all the lands of the nations; 
each in its place will bow down to him.
\verse You also, O Cushites, 
they shall be killed by my sword.
\verse And he will stretch out his hand against the north, 
and he will destroy Assyria 
and will make Nineveh a desolation, 
a dry wasteland like the desert.
\verse Herds shall lie down in her midst, 
every wild animal of a nation; 
even the desert owl and the screech owl 
shall lodge on her capitals. 
A voice shall hoot in the window; 
rubble on the threshold; 
for the cedar is laid bare.
\verse This is the city of rejoicing 
that lived securely; 
the one saying in her heart, 
“\textit{I am, and there is none besides me}.” 
How she has become a desolation, 
a lair for wild animals! 
Everyone who passes by her hisses and shakes his fist.
\end{biblechapter}

\begin{biblechapter} % Zephaniah 3
\verseWithHeading{Judgment of Jerusalem and the Nations} Woe to you, O rebellious and defiled one! 
The oppressing city!
\verse She listens to no voice; 
she does not accept correction. 
In Adonai she does not trust; 
to her God she does not draw near.
\verse Her officials in her midst are roaring lions; 
her judges are evening wolves; 
they leave nothing until the morning.
\verse Her prophets are arrogant men of treachery; 
her priests have defiled that which is holy; 
they do violence to the law.
\verse Adonai is righteous in her midst; 
he does not do wickedness. 
\textit{Morning by morning} he renders his judgment; 
at dawn he does not fail, 
but an evil one knows no shame.
\verse “I have cut off nations; 
their strongholds are deserted. 
I have laid waste their streets, 
so that none pass through. 
Their cities have been destroyed 
and are without a man, without an inhabitant.
\verse I have said, ‘Surely she will fear me; 
she will accept discipline. 
Then her dwelling place will not be cut down, 
nor all that I have brought upon her.’ 
Surely they rise early; 
they make all their deeds corrupt.
\verse “Therefore, wait for me”—a declaration of Adonai— 
“for the day of my rising as a witness. 
For my decision is to gather nations, 
to assemble kingdoms, 
to pour out my wrath upon them, all my burning anger. 
For by the fire of my anger all the land shall be consumed.
\verse Because then I will \textit{make the speech of the nations pure}; 
that all of them might call on the name of Adonai, 
to serve him \textit{in unison}.
\verse From beyond the rivers of Cush 
my worshipers and the daughter of my scattered ones 
shall bring my offering.
\verse On that day you shall not be humiliated 
on account of all your deeds 
by which you have rebelled against me. 
For then I shall remove from your midst 
those exulting in your pride, 
and you shall no longer be haughty 
in \textit{my holy mountain}.
\verse And I will leave in your midst 
a people afflicted and poor, 
and they shall take refuge 
in the name of Adonai.
\verse The remnant of Israel 
shall not do wickedness; 
they shall not speak deception, 
nor shall a deceitful tongue be found in their mouth. 
For they shall graze and lie down, 
and none shall frighten them.”
\verse Shout for joy, O daughter of Zion! 
Cry aloud, O Israel! 
Rejoice and be jubilant with all your heart, 
O daughter of Jerusalem!
\verse Adonai has annulled your judgments; 
he has turned away your enemies. 
The king of Israel, Adonai, is in your midst; 
you shall no longer fear misfortune.
\verse On that day it shall be said to Jerusalem, 
“Fear not, O Zion; 
your hands shall not hang limp.
\verse Adonai your God is in your midst; 
a mighty warrior who saves. 
He shall rejoice over you with joy; 
he renews you in his love; 
he will exult over you with singing.
\verse I will gather those of you grieving on account of the feast; 
they were raising against her a reproach.
\verse Behold, at that time I will deal with all your oppressors; 
I will save the lame and gather the outcast. 
I will change them from shame 
to glory and \textit{renown} throughout the whole world.
\verse At that time I will bring you in; 
at the time of my gathering you together. 
For I will make you \textit{renowned} and praised 
among all the nations of the earth 
when I restore your fortunes before your eyes,” 
says Adonai.
\end{biblechapter}

