\biblebook{Philemon}
\begin{biblechapter}% Philemon 1
\verseWithHeading{Greeting}{Paul, a prisoner for Christ Jesus, and Timothy our brother, To Philemon our beloved fellow worker}%
\verse{and Apphia our sister and Archippus our fellow soldier, and the church in your house:}%
\verse{Grace to you and peace from God our Father and the Lord Jesus Christ.}%
\verseWithHeading{Philemon’s Love and Faith}{I thank my God always when I remember you in my prayers,}%
\verse{because I hear of your love and of the faith that you have toward the Lord Jesus and for all the saints,}%
\verse{and I pray that the sharing of your faith may become effective for the full knowledge of every good thing that is in us for the sake of Christ.}%
\verse{For I have derived much joy and comfort from your love, my brother, because the hearts of the saints have been refreshed through you.}%
\verseWithHeading{Paul’s Plea for Onesimus}{Accordingly, though I am bold enough in Christ to command you to do what is required,}%
\verse{yet for love's sake I prefer to appeal to you— I, Paul, an old man and now a prisoner also for Christ Jesus—}%
\verse{I appeal to you for my child, Onesimus, whose father I became in my imprisonment.}%
\verse{(Formerly he was useless to you, but now he is indeed useful to you and to me.)}%
\verse{I am sending him back to you, sending my very heart.}%
\verse{I would have been glad to keep him with me, in order that he might serve me on your behalf during my imprisonment for the gospel,}%
\verse{but I preferred to do nothing without your consent in order that your goodness might not be by compulsion but of your own accord.}%
\verse{For this perhaps is why he was parted from you for a while, that you might have him back forever,}%
\verse{no longer as a bondservant but more than a bondservant, as a beloved brother— especially to me, but how much more to you, both in the flesh and in the Lord.}%
\verse{So if you consider me your partner, receive him as you would receive me.}%
\verse{If he has wronged you at all, or owes you anything, charge that to my account.}%
\verse{I, Paul, write this with my own hand: I will repay it— to say nothing of your owing me even your own self.}%
\verse{Yes, brother, I want some benefit from you in the Lord. Refresh my heart in Christ.}%
\verse{Confident of your obedience, I write to you, knowing that you will do even more than I say.}%
\verse{At the same time, prepare a guest room for me, for I am hoping that through your prayers I will be graciously given to you.}%
\verseWithHeading{Final Greetings}{Epaphras, my fellow prisoner in Christ Jesus, sends greetings to you,}%
\verse{and so do Mark, Aristarchus, Demas, and Luke, my fellow workers.}%
\verse{The grace of the Lord Jesus Christ be with your spirit.}%
\end{biblechapter}%
