\biblebook{Mark}
\begin{biblechapter}% Mark 1
\verseWithHeading{John the Baptist Begins His Ministry}{The beginning of the gospel of Jesus Christ.\lebnote{Some manuscripts add “the Son of God”}}%
\verse{Just as it is written in the prophet Isaiah, “Behold, I am sending my messenger before your face, who will prepare your way,}%
\verse{the voice of one shouting in the wilderness, ‘Prepare the way of the Lord, make straight his paths!’”\lebnote{from Isa 40:3}}%
\verse{John was there baptizing in the wilderness, proclaiming\lebnote{Some manuscripts have “and proclaiming”} a baptism of repentance for the forgiveness of sins.}%
\verse{And all the Judean region and all the inhabitants of Jerusalem went out to him and were being baptized by him in the Jordan River, confessing their sins.}%
\verse{And John was dressed in camel’s hair and a belt made of leather around his waist, and he ate locusts and wild honey.}%
\verse{And he was preaching, saying, “One who is more powerful than I is coming after me, of whom I am not worthy to bend down and\lebnote{“\textit{and}” supplied because previous participle “bend down” translated as an infinitive} untie the strap of his sandals.}%
\verse{I baptized you with water, but he will baptize you with the Holy Spirit.”}%
\verseWithHeading{The Baptism of Jesus}{And it happened that in those days Jesus came from Nazareth in Galilee and was baptized in the Jordan by John.}%
\verse{And immediately as he\lebnote{“\textit{as}” is supplied as a component of the participle (“was coming up”) which is understood as temporal} was coming up out of the water, he saw the heavens being split apart and the Spirit descending like a dove on him.}%
\verse{And a voice came from heaven, “You are my beloved Son; with you I am well pleased.”}%
\verseWithHeading{The Temptation of Jesus}{And immediately the Spirit drove him out into the wilderness.}%
\verse{And he was in the wilderness forty days being tempted by Satan. And he was with the wild animals, and the angels were ministering to him.}%
\verseWithHeading{Public Ministry in Galilee}{And after\lebnote{Some manuscripts have “Now after”} John had been taken into custody,\lebnote{“had been handed over”} Jesus went into Galilee proclaiming the gospel of God}%
\verse{and saying, \JesusWords{“The time is fulfilled and the kingdom of God has come near. Repent and believe in the gospel!”}}%
\verseWithHeading{Jesus Calls His First Disciples}{And as he\lebnote{“\textit{as}” is supplied as a component of the participle (“was passing by”) which is understood as temporal} was passing by along the Sea of Galilee, he saw Simon and Andrew, Simon’s brother, casting a net\lnCYZ{} into the sea (for they were fishermen).}%
\verse{And Jesus said to them, \JesusWords{“Follow\lebnote{“come after”} me and I will make you become fishers of people.”}}%
\verse{And immediately they left their nets and\lnCZA{} followed him.}%
\verse{And going on a little farther, he saw James the son of Zebedee and his brother John, and they were in the boat mending the nets.}%
\verse{And immediately he called them, and they left their father Zebedee in the boat with the hired men and\lnCZA{} went away after him.}%
\verse{And they went into Capernaum and immediately on the Sabbath he began to teach in the synagogue.\lebnote{Some manuscripts have “he went into the synagogue and began to teach”}}%
\verseWithHeading{A Man with an Unclean Spirit Healed}{And they were amazed at his teaching, because he was teaching them like one who had authority, and not like the scribes.}%
\verse{And so then there was a man in their synagogue with an unclean spirit, and he cried out,}%
\verse{saying, “Leave us alone,\lebnote{“what to us and to you”} Jesus the Nazarene! Have you come to destroy us? I know who you are — the Holy One of God!”}%
\verse{And Jesus rebuked him, saying, \JesusWords{“Be silent, and come out of him!”}}%
\verse{And after\lebnote{“\textit{after}” is supplied as a component of the participle (“convulsing”) which is understood as temporal} convulsing him and crying out with a loud voice, the unclean spirit came out of him.}%
\verse{And they were all amazed, so that they began to discuss with one another, saying, “What is this? A new teaching with authority! He even commands the unclean spirits and they obey him.”}%
\verse{And the report about him then went out everywhere in the whole surrounding region of Galilee.}%
\verseWithHeading{Many at Capernaum Are Healed}{And so then he departed from the synagogue and\lebnote{“\textit{and}” supplied because previous participle “departed” translated as a finite verb} came into the house of Simon and Andrew with James and John.}%
\verse{Now Simon’s mother-in-law was lying down, suffering with a fever, and at once they told him about her.}%
\verse{And he came and\lebnote{“\textit{and}” supplied because previous participle “came” translated as a finite verb} raised her up by\lebnote{“\textit{by}” is supplied as a component of the participle (“taking hold of”) which is understood as means} taking hold of her\lebnote{“the”: the Greek article is used here as a possessive pronoun} hand, and the fever left her, and she began to serve them.}%
\verse{Now when it\lebnote{“\textit{when}” is supplied as a component of the temporal genitive absolute participle (“was”)} was evening, when the sun had set, they began bringing\lebnote{Imperfect tense as ingressive (“began bringing”)} to him all those who were sick\lnCZB{} and those who were demon-possessed.}%
\verse{And the whole town was gathered together at the door.}%
\verse{And he healed many who were sick\lnCZB{} with various diseases and expelled many demons. And he did not permit the demons to speak, because they knew him.}%
\verseWithHeading{Preaching Throughout Galilee}{And getting up early in the morning while it was very dark, he departed and went to a deserted place, and there he was praying.}%
\verse{And Simon and those who were with him searched diligently for him.}%
\verse{And they found him and said to him, “Everyone is looking for you!”}%
\verse{And he said to them, \JesusWords{“Let us go elsewhere, into the neighboring rural towns, so that I can preach there also, because I have come out for this very reason.”}}%
\verse{And he went into all Galilee preaching in their synagogues and expelling demons.}%
\verseWithHeading{A Leper Cleansed}{And a leper came to him, entreating him and kneeling down, saying\lebnote{Some manuscripts have “and saying”} to him, “If you are willing, you are able to make me clean.”}%
\verse{And becoming angry,\lebnote{Some manuscripts have “having compassion”} he stretched out his hand and\lebnote{“\textit{and}” supplied because previous participle “stretched out” translated as a finite verb} touched him\lnCYZ{}, and said to him, \JesusWords{“I am willing; be made clean.”}}%
\verse{And immediately the leprosy went away from him and he was made clean.}%
\verse{And warning him sternly, he sent him away at once.}%
\verse{And he said to him, \JesusWords{“See to it that you say nothing to anyone, but go, show yourself to the priest and bring for your cleansing the things which Moses commanded, for a testimony to them.}}%
\verse{But he went out and\lebnote{“\textit{and}” supplied because previous participle “went out” translated as a finite verb} began to proclaim it\lnCYZ{} freely and to spread abroad the account, so that he was no longer able to enter publicly into a town. But he was staying outside in deserted places, and they were coming to him from all directions.}%
\end{biblechapter}%
\begin{biblechapter}% Mark 2
\verseWithHeading{A Paralytic Healed}{And when he\lebnote{“\textit{when}” is supplied as a component of the participle (“entered”) which is understood as temporal} entered again into Capernaum after some days, it became known that he was at home.}%
\verse{And many had gathered, so that there was no longer room, not even at the door, and he was speaking the word to them.}%
\verse{And they came bringing to him a paralytic, carried by four of them.}%
\verse{And when\lebnote{“\textit{when}” is supplied as a component of the participle (“able”) which is understood as temporal} they were not able to bring him\lnCZC{} to him because of the crowd, they removed the roof where he was. And after\lebnote{“\textit{after}” is supplied as a component of the participle (“digging through”) which is understood as temporal} digging through, they lowered the stretcher on which the paralytic was lying.}%
\verse{And when\lnCZD{} Jesus saw their faith, he said to the paralytic, \JesusWords{“Child, your sins are forgiven.”}}%
\verse{Now some of the scribes were sitting there and reasoning in their hearts,}%
\verse{“Why does this man speak like this? He is blaspheming! Who is able to forgive sins except God alone?”}%
\verse{And immediately Jesus, perceiving in his spirit that they were reasoning like this within themselves, said to them, \JesusWords{“Why are you considering these things in your hearts?}}%
\verse{\JesusWords{Which is easier to say to the paralytic, ‘Your sins are forgiven,’ or to say ‘Get up and pick up your stretcher and walk’?}}%
\verse{\JesusWords{But so that you may know that the Son of Man has authority on earth to forgive sins,”} — he said to the paralytic —}%
\verse{\JesusWords{“I say to you, get up, pick up your stretcher, and go to your home.”}}%
\verse{And he got up and immediately picked up his\lebnote{“the”: the Greek article is used here as a possessive pronoun} stretcher and\lebnote{“\textit{and}” supplied because previous participle “picked up” translated as a finite verb} went out in front of them all, so that they were all amazed and glorified God, saying, “We have never seen anything\lnCZC{} like this!”}%
\verseWithHeading{Levi Called to Follow Jesus}{And he went out again beside the sea,\lebnote{That is, the Sea of Galilee} and all the crowd was coming to him, and he began to teach\lebnote{Imperfect tense as ingressive (“began to teach”)} them.}%
\verse{And as he\lebnote{“\textit{as}” is supplied as a component of the participle (“was passing by”) which is understood as temporal} was passing by, he saw Levi the son of Alphaeus sitting at the tax booth, and he said to him, \JesusWords{“Follow me!”} And he stood up and\lebnote{“\textit{and}” supplied because previous participle “stood up” translated as a finite verb} followed him.}%
\verse{And it happened that he was dining\lebnote{“was reclining for a meal”} in his house, and many tax collectors and sinners were dining with\lebnote{“were reclining at table with”} Jesus and his disciples, for there were many and they were following him.}%
\verse{And the scribes of the Pharisees, when they\lnCZD{} saw that he was eating with sinners and tax collectors, began to say\lnCZE{} to his disciples, “Why does he eat with tax collectors and sinners?”}%
\verse{And when\lebnote{“\textit{when}” is supplied as a component of the participle (“heard”) which is understood as temporal} Jesus heard it\lnCZC{}, he said to them, \JesusWords{“Those who are healthy do not have need of a physician, but those who are sick.\lebnote{“having badly”} I have not come to call the righteous, but sinners.”}}%
\verseWithHeading{On Fasting}{And John’s disciples and the Pharisees were fasting, and they came and said to him, “Why\lebnote{“for what” \textit{reason}} do the disciples of John and the disciples of the Pharisees fast, but your disciples do not fast?”}%
\verse{And Jesus said to them, \JesusWords{“The bridegroom’s attendants\lebnote{“the sons of the bridal chamber”} are not able to fast while the bridegroom is with them, are they?\lebnote{The negative construction in Greek anticipates a negative answer here, indicated in the translation by the phrase “\textit{are they}”} As long a time as they have the bridegroom with them, they are not able to fast.}}%
\verse{\JesusWords{But days will come when the bridegroom is taken away from them, and then they will fast in that day.}}%
\verse{\JesusWords{No one sews a patch of unshrunken cloth on an old garment. Otherwise\lnCZF{} the patch pulls away from it — the new from the old — and the tear becomes worse.}}%
\verse{\JesusWords{And no one puts new wine into old wineskins. Otherwise\lnCZF{} the wine will burst the wineskins and the wine is destroyed and the wineskins too. But new wine is put into new wineskins.”}}%
\verseWithHeading{Plucking Grain on the Sabbath}{And it happened that he was going through the grain fields on the Sabbath, and his disciples began to make their way while\lebnote{“\textit{while}” is supplied as a component of the participle (“picking”) which is understood as temporal} plucking off the heads of grain.}%
\verse{And the Pharisees began to say\lnCZE{} to him, “Behold, why are they doing what is not permitted on the Sabbath?”}%
\verse{And he said to them, \JesusWords{“Have you never read what David did when he had need and he and those who were with him were hungry —}}%
\verse{\JesusWords{how he entered into the house of God in the time of Abiathar the high priest and ate the bread of the presentation, which it is not permitted to eat (except the priests) and also gave it\lnCZC{} to those who were with him?”}}%
\verse{And he said to them, \JesusWords{“The Sabbath was established for people, and not people for the Sabbath.}}%
\verse{\JesusWords{So then, the Son of Man is lord even of the Sabbath.”}}%
\end{biblechapter}%
\begin{biblechapter}% Mark 3
\verseWithHeading{A Man with a Withered Hand Healed}{And he entered into the synagogue again, and a man who had a withered hand was there.}%
\verse{And they were watching him closely to see if he would heal him on the Sabbath, in order that they could accuse him.}%
\verse{And he said to the man who had the withered hand, \JesusWords{“Come into the middle.”}}%
\verse{And he said to them, \JesusWords{“Is it permitted on the Sabbath to do good or to do evil, to save life or to kill?”} But they were silent.}%
\verse{And looking around at them with anger, grieved at the hardness of their hearts, he said to the man, \JesusWords{“Stretch out your\lebnote{“the”: the Greek article is used here as a possessive pronoun} hand.”} And he stretched it\lebnote{supplied from English context} out, and his hand was restored.}%
\verse{And the Pharisees went out immediately with the Herodians and began to conspire\lebnote{“began to give counsel”; the imperfect tense has been translated as ingressive (“began to”)} against him with regard to how they could destroy him.}%
\verseWithHeading{Jesus Heals Crowds by the Sea}{And Jesus went away with his disciples to the sea,\lebnote{That is, the Sea of Galilee} and a great crowd from Galilee followed him.\lnCZG{} And from Judea}%
\verse{and from Jerusalem and from Idumea and the other side of the Jordan and around Tyre and Sidon a great crowd came to him because they\lebnote{“\textit{because}” is supplied as a component of the participle (“heard”) which is understood as causal} heard all that he was doing.}%
\verse{And he told his disciples that a small boat should stand ready for him because of the crowd, so that they would not press upon him.}%
\verse{For he had healed many, so that all those who were suffering from diseases\lebnote{“had suffering”} pressed about him in order that they could touch him.}%
\verse{And the unclean spirits, whenever they saw him, were falling down before him and crying out, saying, “You are the Son of God!”}%
\verse{And he warned them strictly that they should not make him known.}%
\verseWithHeading{The Selection of the Twelve Apostles}{And he went up on the mountain and summoned those whom he wanted, and they came to him.}%
\verse{And he appointed twelve,\lebnote{Some manuscripts add “whom he also named apostles”} so that they would be with him and so that he could send them out to preach}%
\verse{and to have authority to expel demons.}%
\verse{And he appointed the twelve.\lebnote{Most Greek manuscripts omit the phrase “and he appointed the twelve”} And to Simon he gave the name Peter,}%
\verse{and James the son of Zebedee and John the brother of James (and he gave to them the name Boanerges, that is, “Sons of Thunder”),}%
\verse{and Andrew, and Philip, and Bartholomew, and Matthew, and Thomas, and James the son of Alphaeus, and Thaddaeus, and Simon the Zealot,\lebnote{“the Cananean,” but according to BDAG 507 s.v., this term has no relation at all to the geographical terms for Cana or Canaan, but is derived from the Aramaic term for “enthusiast, zealot” (see Luke 6:15; Acts 1:13)}}%
\verse{and Judas Iscariot, who also betrayed him.}%
\verseWithHeading{A House Divided Cannot Stand}{And he went home, and the crowd gathered again, so that they were not even able to eat a meal.}%
\verse{And when\lebnote{“\textit{when}” is supplied as a component of the participle (“heard”) which is understood as temporal} his family\lebnote{\textit{those} “close to him”} heard this,\lnCZG{} they went out to restrain him, for they were saying, “He has lost his mind!”}%
\verse{And the scribes who had come down from Jerusalem were saying, “He is possessed by Beelzebul!” and “By the ruler of the demons he expels the demons!”}%
\verse{And he called them to himself and\lebnote{“\textit{and}” supplied because previous participle “called” translated as a finite verb} was speaking to them in parables, \JesusWords{“How can Satan expel Satan?}}%
\verse{\JesusWords{And if a kingdom is divided against itself, that kingdom is not able to stand.}}%
\verse{\JesusWords{And if a house is divided against itself, that house will not be able to stand.}}%
\verse{\JesusWords{And if Satan has risen up against himself and is divided, he is not able to stand, but is at an end!}}%
\verse{\JesusWords{But no one is able to enter into the house of a strong man and\lebnote{“\textit{and}” supplied because previous participle “enter” translated as a finite verb} plunder his property unless he first ties up the strong man, and then he can thoroughly plunder his house.}}%
\verse{\JesusWords{“Truly I say to you that all the sins and the blasphemies will be forgiven the sons of men, however much they blaspheme.}}%
\verse{\JesusWords{But whoever blasphemes against the Holy Spirit does not have forgiveness forever,\lebnote{“for the age”} but is guilty of an eternal sin” —}}%
\verse{because they were saying, “He has an unclean spirit.”}%
\verseWithHeading{Jesus’ Mother and Brothers}{And his mother and his brothers arrived, and standing outside, they sent word to him to summon him.}%
\verse{And a crowd was sitting around him, and they told him, “Behold, your mother and your brothers\lebnote{Some manuscripts add “and your sisters”} are outside looking for you.”}%
\verse{And he answered them and\lebnote{“\textit{and}” supplied because previous participle “answered” translated as a finite verb} said, \JesusWords{“Who is my mother or\lebnote{Some manuscripts have “and”} my brothers?”}}%
\verse{And looking around at those who were sitting around him in a circle, he said, \JesusWords{“Behold, my mother and my brothers!}}%
\verse{\JesusWords{For whoever does the will of God, this person is my brother and sister and mother.”}}%
\end{biblechapter}%
\begin{biblechapter}% Mark 4
\verseWithHeading{The Parable of the Sower}{And again he began to teach beside the sea,\lebnote{That is, the Sea of Galilee} and a very large crowd was gathered to him, so that he got into a boat and\lebnote{“and” supplied because previous participle “got” translated as a finite verb} sat on the sea, and the whole crowd was at the sea on the land.}%
\verse{And he began to teach\lebnote{Imperfect tense as ingressive (“began to teach”)} them many things in parables, and was saying to them in his teaching,}%
\verse{\JesusWords{“Listen! Behold, the sower went out to sow.}}%
\verse{\JesusWords{And it happened that while he was sowing, some seed\lebnote{“\textit{some of} which”} fell on the side of the path, and the birds came and devoured it.}}%
\verse{\JesusWords{And other seed fell on the rocky ground where it did not have much soil, and it sprang up at once, because it did not have any depth of soil.}}%
\verse{\JesusWords{And when the sun rose it was scorched, and because it did not have enough root, it withered.}}%
\verse{\JesusWords{And other seed fell among the thorn plants, and the thorn plants came up and choked it, and it did not produce grain.}\lnCZH{}}%
\verse{\JesusWords{And other seed fell on the good soil, and produced grain,\lnCZH{} coming up and increasing, and it bore a crop\lnCZI{} — one thirty and one sixty and one a hundred times as much.\lnCZJ{}}}%
\verse{And he said, \JesusWords{“Whoever has ears to hear, let him hear!”}}%
\verseWithHeading{The Reason for the Parables}{And when he was alone, those around him together with the twelve began asking\lebnote{Imperfect tense as ingressive (“began asking”)} him about the parables.}%
\verse{And he said to them, \JesusWords{“To you has been granted the secret of the kingdom of God, but to those who are outside everything is in parables,}}%
\verse{\JesusWords{so that ‘they may look closely\lebnote{“seeing they may see”} and not perceive, and they may listen carefully\lebnote{“hearing they may hear”} and not understand, lest they turn and it be forgiven them.’”\lebnote{from Isa 6:9–10} }}%
\verseWithHeading{The Parable of the Sower Interpreted}{And he said to them, \JesusWords{“Do you not understand this parable? And how will you understand all the parables?}}%
\verse{\JesusWords{The sower sows the word.}}%
\verse{\JesusWords{And these are the ones beside the path where the word is sown, and whenever they hear it,\lnCZK{} immediately Satan comes and takes away the word that was sown in them.}}%
\verse{\JesusWords{And these are like\lebnote{Some manuscripts omit “like”} the ones sown on the rocky ground, who whenever they hear the word immediately receive it with joy.}}%
\verse{\JesusWords{And they have no root in themselves, but are temporary. Then when\lebnote{“\textit{when}” is supplied as a component of the temporal genitive absolute participle (“comes”)} affliction or persecution comes because of the word, immediately they fall away.}}%
\verse{\JesusWords{And others are the ones sown among the thorn plants — these are the ones who hear the word,}}%
\verse{\JesusWords{and the cares of the world and the deceitfulness of wealth and the desires for other things come in and\lebnote{“\textit{and}” supplied because previous participle “come in” translated as a finite verb} choke the word and it becomes unproductive.}}%
\verse{\JesusWords{And those are the ones sown on the good soil, who hear the word and receive it\lnCZI{} and bear fruit — one thirty and one sixty and one a hundred times as much.”\lnCZJ{}}}%
\verseWithHeading{The Parable of the Lamp}{And he said to them, \JesusWords{“Surely a lamp is not brought so that it may be put under a bushel basket or under a bed, is it?\lebnote{The negative construction in Greek anticipates a negative answer here, indicated in the translation by the phrase “\textit{is it}”} Is it not\lebnote{The negative construction in Greek anticipates a positive answer here} so that it may be put on a lampstand?}}%
\verse{\JesusWords{For nothing is secret except so that it may be revealed, nor has become hidden except so that it will come to light.}}%
\verse{\JesusWords{If anyone has ears to hear, let him hear!”}}%
\verse{And he said to them, \JesusWords{“Take care what you hear! With the measure by which you measure out, it will be measured out to you, and will be added to you.}}%
\verse{\JesusWords{For whoever has, more will be given to him, and whoever does not have, even what he has will be taken away from him.”}}%
\verseWithHeading{The Parable of the Seed that Grows by Itself}{And he said, \JesusWords{“The kingdom of God is like this: like a man scatters seed on the ground.}}%
\verse{\JesusWords{And he sleeps and gets up, night and day, and the seed sprouts and grows — he does not know how.\lebnote{“\textit{in such a way} as he does not know”}}}%
\verse{\JesusWords{By itself the soil produces a crop: first the grass, then the head of grain, then the full grain in the head.}}%
\verse{\JesusWords{But when the crop permits, he sends in the sickle right away, because the harvest has come.”}}%
\verseWithHeading{The Parable of the Mustard Seed}{And he said, \JesusWords{“With what can we compare the kingdom of God, or by what parable can we present it?}}%
\verse{\JesusWords{It is like a mustard seed that when sown on the ground, although it\lebnote{“\textit{although}” is supplied as a component of the participle (“is”) which is understood as concessive} is the smallest of all the seeds that are on the ground,}}%
\verse{\JesusWords{but when it is sown it grows up and becomes the largest of all the garden herbs, and sends out large branches so that the birds of the sky are able to nest in its shade.”}}%
\verse{And with many parables such as these he was speaking the word to them, as they were able to hear it.\lnCZK{}}%
\verse{And he did not speak to them without a parable, but in private he explained everything to his own disciples.}%
\verseWithHeading{Calming of a Storm}{And on that day, when it\lebnote{“\textit{when}” is supplied as a component of the temporal genitive absolute participle (“was”)} was evening, he said to them, \JesusWords{“Let us cross over to the other side.”}}%
\verse{And leaving the crowd, they took him along, as he was, in the boat. And other boats were with him.}%
\verse{And a great storm of wind developed, and the waves were breaking into the boat, so that the boat was already being filled with water.\lebnote{*The words “\textit{with water}” are not in the Greek text but are implied}}%
\verse{And he was in the stern sleeping on the cushion, and they woke him up and said to him, “Teacher, is it not a concern to you that we are perishing?”}%
\verse{And he woke up and\lebnote{“\textit{and}” supplied because previous participle “woke up” translated as a finite verb} rebuked the wind, and said to the sea, \JesusWords{“Be quiet! Be silent!”} And the wind abated and there was a great calm.}%
\verse{And he said to them, \JesusWords{“Why are you fearful? Do you not yet have faith?”}}%
\verse{And they were terribly frightened\lebnote{“they feared a great fear”} and began to say\lebnote{Imperfect tense as ingressive (“began to say”)} to one another, “Who then is this, that even the wind and the sea obey him?”}%
\end{biblechapter}%
\begin{biblechapter}% Mark 5
\verseWithHeading{A Demon-possessed Gerasene Healed}{And they came to the other side of the sea,\lebnote{That is, the Sea of Galilee} to the region of the Gerasenes.\lebnote{Most later manuscripts read “Gadarenes,” while other manuscripts read “Gergesenes” here}}%
\verse{And as\lebnote{“\textit{as}” is supplied as a component of the temporal genitive absolute participle (“was getting out”)} he was getting out of the boat, immediately from the tombs a man with an unclean spirit went to meet him,}%
\verse{who lived\lebnote{“had his dwelling”} among the tombs. And no one was able to bind him any longer, not even with a chain,}%
\verse{because he had often been bound with shackles and chains, and the chains had been torn apart by him, and the shackles had been shattered. And no one was strong enough to subdue him.}%
\verse{And during every night and day among the tombs and on the mountains he was crying out and cutting himself with stones.}%
\verse{And when he\lnCZL{} saw Jesus from a distance, he ran and knelt down before him.}%
\verse{And crying out with a loud voice he said, “What have I to do with you\lebnote{“what to me and to you”}, Jesus, Son of the Most High God? I implore you by God, do not torment me!”}%
\verse{(For he was saying to him, \JesusWords{“Come out of the man, unclean spirit!”})}%
\verse{And he was asking him \JesusWords{“What is your name?”}\lebnote{“what name to you”} And he said to him, “My name is Legion, because we are many.”}%
\verse{And he was imploring him many times that he would not send them out of the region.}%
\verse{Now a large herd of pigs was there at the hill feeding,}%
\verse{and they implored him, saying, “Send us to the pigs so that we may enter into them.”}%
\verse{And he permitted them. And the unclean spirits came out and\lebnote{“\textit{and}” supplied because previous participle “came out” translated as a finite verb} entered into the pigs, and the herd — about two thousand — rushed headlong down the steep slope into the sea and were drowned in the sea.}%
\verse{And their herdsmen fled and reported it\lnCZM{} in the town and in the countryside, and they came to see what it was that had happened.}%
\verse{And they came to Jesus and saw the demon-possessed man sitting there clothed and in his right mind — the one who had had the legion — and they were afraid.}%
\verse{And those who had seen it\lnCZM{} described to them what had happened to the demon-possessed man, and about the pigs.}%
\verse{And they began to urge him to depart from their region.}%
\verse{And as\lebnote{“\textit{as}” is supplied as a component of the temporal genitive absolute participle (“was getting”)} he was getting into the boat, the man who had been demon-possessed began to implore\lebnote{Imperfect tense as ingressive (“began to implore”)} him that he could go with him.}%
\verse{And he did not permit him, but said to him, \JesusWords{“Go to your home to your people and tell them all that the Lord has done for you, and that he has had mercy on you.”}}%
\verse{And he went away and began to proclaim in the Decapolis all that Jesus had done for him, and they were all astonished.}%
\verseWithHeading{A Woman Healed and a Daughter Raised}{And after\lebnote{“\textit{after}” is supplied as a component of the temporal genitive absolute participle (“had crossed over”)} Jesus had crossed over again in the boat to the other side, a large crowd gathered to him, and he was beside the sea.}%
\verse{And one of the rulers of the synagogue came — Jairus by name — and when he\lnCZL{} saw him, he fell down at his feet.}%
\verse{And he was imploring him many times, saying, “My little daughter is at the point of death!\lebnote{“has finally”} Come, lay your\lnCZN{} hands on her, so that she will get well and will live.”}%
\verse{And he went with him, and a large crowd was following him and pressing around him.}%
\verse{And there was a woman who was suffering from hemorrhages\lebnote{“with a flow of blood”} twelve years.}%
\verse{And she had endured many things under many physicians, and had spent all that she had\lebnote{“all with her”} and had received no help at all, but instead became worse.\lebnote{“came back for the worse”}}%
\verse{When she\lebnote{“\textit{when}” is supplied as a component of the participle (“heard”) which is understood as temporal} heard about Jesus, she came up in the crowd behind him and\lebnote{“\textit{and}” supplied because previous participle “came up” translated as a finite verb} touched his cloak,}%
\verse{for she was saying, “If I touch just his clothing, I will be healed!”}%
\verse{And immediately her hemorrhage stopped\lebnote{“flow of blood was dried up”} and she realized in her\lnCZN{} body that she was healed of her\lnCZN{} suffering.}%
\verse{And immediately Jesus, perceiving in himself that power had gone out from himself, turned around in the crowd and\lebnote{“\textit{and}” supplied because previous participle “turned around” translated as a finite verb} said, \JesusWords{“Who touched my clothing?”}}%
\verse{And his disciples said to him, “You see the crowd pressing upon you, and you say ‘Who touched me?’”}%
\verse{And he was looking around to see the one who had done this.}%
\verse{So the woman, frightened and trembling, knowing what had happened to her, came and fell down before him and told him the whole truth.}%
\verse{But he said to her, \JesusWords{“Daughter, your faith has healed you. Go in peace and be well from your suffering.”}}%
\verse{While\lebnote{“\textit{while}” is supplied as a component of the temporal genitive absolute participle (“speaking”)} he was still speaking, they came from the synagogue ruler’s house\lebnote{“\textit{house}” is implied since the synagogue ruler himself is directly addressed (and therefore present) in the remainder of this verse} saying, “Your daughter has died. Why trouble the Teacher further?”}%
\verse{But Jesus, ignoring\lebnote{Or “overhearing”} what was said,\lebnote{“the report that was spoken”} told the ruler of the synagogue, \JesusWords{“Do not be afraid — only believe!”}}%
\verse{And he did not allow anyone to follow along with him except Peter and James and John, the brother of James.}%
\verse{And they came to the house of the ruler of the synagogue and saw a commotion, and people weeping and wailing loudly.}%
\verse{And when he\lebnote{“\textit{when}” is supplied as a component of the participle (“entered”) which is understood as temporal} entered, he said to them, \JesusWords{“Why are you agitated and weeping? The child is not dead, but is sleeping.”}}%
\verse{And they began laughing\lebnote{Imperfect tense as ingressive (“began laughing”)} at him. But he sent them all out and\lebnote{“\textit{and}” supplied because previous participle “sent … out” translated as a finite verb} took along the father and mother of the child, and those who were with him, and went in to where the child was.}%
\verse{And taking hold of the child’s hand, he said to her, \JesusWords{“Talitha koum!”} (which is translated, \JesusWords{“Little girl, I say to you, get up!”}),}%
\verse{and immediately the girl stood up and began walking around\lebnote{Imperfect tense as ingressive (“began walking around”)} (for she was twelve years old). And immediately they were utterly and completely astonished.\lebnote{“immediately they were astonished with great astonishment”}}%
\verse{And he commanded them strictly that no one should learn of this, and said to give her something\lnCZM{} to eat.}%
\end{biblechapter}%
\begin{biblechapter}% Mark 6
\verseWithHeading{Rejected at Nazareth}{And he went out from there and came to his hometown, and his disciples followed him.}%
\verse{And when\lnCZO{} the Sabbath came, he began to teach in the synagogue, and many who heard him\lnCZP{} were amazed, saying, “Where did this man get these things? And what is this wisdom that has been granted to this man, and the miracles such as these performed through his hands?}%
\verse{Is not this the carpenter, the son of Mary and brother of James and Joses and Judas and Simon? And are not his sisters here with us?” And they were offended by him.}%
\verse{And Jesus said to them, \JesusWords{“A prophet is not without honor except in his hometown, and among his relatives, and in his own household.”}}%
\verse{And he was not able to do any miracle in that place except to lay his\lnCZQ{} hands on a few sick people and\lebnote{“\textit{and}” supplied because previous participle “lay” translated as a finite verb} heal them.\lnCZR{}}%
\verse{And he was astonished because of their unbelief.\innerVerseHeading{The Twelve Commissioned and Sent Out}And he was going around among the villages teaching.}%
\verse{And he summoned the twelve and began to send them out two by two, and gave them authority over the unclean spirits.}%
\verse{And he commanded them that they \JesusWords{take along nothing for the journey except only a staff — no bread, no traveler’s bag, no money in their\lnCZQ{} belts —}}%
\verse{\JesusWords{but to put on sandals and not to wear two tunics.}}%
\verse{And he said to them, \JesusWords{“Whenever you enter into a house, stay there until you depart from there.}}%
\verse{\JesusWords{And whatever place does not welcome you or listen to you, as you\lebnote{“\textit{as}” is supplied as a component of the participle (“go out”) which is understood as temporal} go out from there, shake off the dust that is on your feet for a testimony against them.”}}%
\verse{And they went out and\lnCZS{} proclaimed that people\lebnote{“\textit{people}” is supplied as the subject because a third-person pronoun (“they”) would be ambiguous} should repent.}%
\verse{And they were expelling many demons and anointing many sick people with olive oil and healing them.\lnCZR{}}%
\verseWithHeading{Herod Kills John the Baptist}{And King Herod heard it,\lnCZR{} because his name had become known. And they were saying, “John, the one who baptizes, has been raised from the dead, and because of this these miraculous powers are at work in him.”}%
\verse{But others were saying, “He is Elijah,” and others were saying, “He is a prophet like one of the prophets.”}%
\verse{But when\lnCZT{} Herod heard it,\lnCZR{} he said, “John whom I beheaded — this one has been raised!”}%
\verse{For Herod himself had sent and\lebnote{“\textit{and}” supplied because previous participle “had sent” translated as a finite verb} arrested John and bound him in prison because of Herodias, the wife of Philip his brother, because he had married her.}%
\verse{For John had been saying to Herod, “It is not permitted for you to have your brother’s wife.”}%
\verse{So Herodias held a grudge against him and was wanting to kill him, and was not able to do so.}%
\verse{For Herod was afraid of John, because he\lebnote{“\textit{because}” is supplied as a component of the participle (“knew”) which is understood as causal} knew him to be a righteous and holy man and protected him. And when he\lebnote{“\textit{when}” is supplied as a component of the participles (“came in” and “danced”) which are understood as temporal} listened to him, he was greatly perplexed, and yet he listened to him gladly.}%
\verse{And a suitable day came when Herod, on his birthday, gave a banquet for his courtiers and military tribunes and the most prominent men of Galilee.}%
\verse{And when\lebnote{“\textit{when}” is supplied as a component of the temporal genitive absolute participle (“came in”)} the daughter of Herodias herself\lebnote{In place of “the daughter of Herodias herself” some manuscripts have “his daughter Herodias”} came in and danced and pleased\lebnote{Some manuscripts have “danced, she pleased”} Herod and his dinner guests,\lebnote{“those reclining at table with \textit{him}”} the king said to the girl, “Ask me for whatever you want, and I will give it\lnCZP{} to you.”}%
\verse{And he swore to her,\lebnote{Some manuscripts add “at length”} “Anything whatever you ask me for I will give you, up to half my kingdom!”}%
\verse{And she went out and\lnCZS{} said to her mother, “What should I ask for?” And she said, “The head of John the baptizer.”}%
\verse{And she came in immediately with haste to the king and\lebnote{“\textit{and}” supplied because previous participle “came in” translated as a finite verb} asked, saying, “I want you to give me the head of John the Baptist on a platter immediately.”}%
\verse{And although he\lebnote{“\textit{although}” is supplied as a component of the participle (“was”) which is understood as concessive} was deeply grieved, the king, because of his\lnCZQ{} oaths and dinner guests,\lebnote{“those who were reclining at table”} did not want to refuse her.}%
\verse{And immediately the king sent an executioner and\lebnote{“\textit{and}” supplied because previous participle “sent” translated as a finite verb} ordered him\lnCZP{} to bring his head. And he went and\lebnote{“\textit{and}” supplied because previous participle “went” translated as a finite verb} beheaded him in the prison.}%
\verse{And he brought his head on a platter and gave it to the girl, and the girl gave it to her mother.}%
\verse{And when\lnCZT{} his disciples heard this,\lnCZR{} they came and took away his corpse and placed it in a tomb.}%
\verseWithHeading{The Feeding of Five Thousand}{And the apostles regathered to Jesus and reported to him everything that they had done and that they had taught.}%
\verse{And he said to them, \JesusWords{“You yourselves come privately to an isolated place and rest for a short time.”} For those who were coming and going were many, and they did not even have time to eat.}%
\verse{And they went away in the boat to an isolated place by themselves.}%
\verse{And many people saw them leaving and recognized them,\lnCZR{} and ran there together by land from all the towns, and arrived ahead of them.}%
\verse{And getting out of the boat\lebnote{The words “\textit{of the boat}” are not in the Greek text but are implied by this verb, which refers to disembarking from a boat or ship} he saw the large crowd and had compassion on them, because they were like sheep without\lebnote{“not having”} a shepherd, and he began to teach them many things.}%
\verse{And the hour had already become late when\lebnote{“\textit{when}” is supplied as a component of the participle (“came up”) which is understood as temporal} his disciples came up to him, saying, “The place is desolate and the hour is already late.}%
\verse{Send them away so that they can go into the surrounding farms and villages and\lnCZU{} purchase something to eat for themselves.”}%
\verse{But he answered and\lebnote{“\textit{and}” supplied because previous participle “answered” translated as a finite verb} said to them, \JesusWords{“You give them something\lnCZP{} to eat.”} And they said to him, \JesusWords{“Should we go and\lnCZU{} purchase bread for two hundred denarii and give it\lnCZP{} to them to eat?”}}%
\verse{And he said to them, \JesusWords{“How many loaves do you have? Go look!”} And when they\lebnote{“\textit{when}” is supplied as a component of the participle (“found out”) which is understood as temporal} found out, they said, “Five, and two fish.”}%
\verse{And he ordered them all to recline in groups on the green grass.}%
\verse{And they reclined in groups, by hundreds and by fifties.}%
\verse{And taking the five loaves and the two fish and\lebnote{“\textit{and}” is supplied in the translation because of English style} looking up to heaven, he gave thanks and broke the loaves and gave them\lnCZP{} to his disciples so that they could set them\lnCZP{} before them. And he distributed the two fish to them all.}%
\verse{And they all ate and were satisfied.}%
\verse{And they picked up the broken pieces, twelve baskets full, and of the fish.}%
\verse{And those who ate the loaves were five thousand men.}%
\verseWithHeading{Jesus Walks on the Water}{And immediately he made his disciples get into the boat and go on ahead to the other side, to Bethsaida, while he himself dismissed the crowd.}%
\verse{And after he\lebnote{“\textit{after}” is supplied as a component of the participle (“went away”) which is understood as temporal} had said farewell to them, he went away to the mountain to pray.}%
\verse{And when\lnCZO{} evening came, the boat was in the middle of the sea and he was alone on the land.}%
\verse{And he saw them being beaten in their rowing\lebnote{Or “being held up in their progress”} because the wind was against them. Around the fourth watch of the night he came to them, walking on the sea, and he was wanting to pass by them.}%
\verse{But when\lebnote{“\textit{when}” is supplied as a component of the participle (“saw”) which is understood as temporal} they saw him walking on the sea, they thought that it was a ghost, and they cried out.}%
\verse{For they all saw him and were terrified. But immediately he spoke with them and said to them, \JesusWords{“Have courage, I am he! Do not be afraid!”}}%
\verse{And he went up with them into the boat, and the wind abated. And they were extraordinarily\lebnote{“exceedingly extremely”} astounded within themselves,}%
\verse{because they did not understand concerning the loaves, but their hearts were hardened.}%
\verseWithHeading{Many Healed at Gennesaret}{And after they\lebnote{“\textit{after}” is supplied as a component of the participle (“had crossed over”) which is understood as temporal} had crossed over, they came to land at Gennesaret and anchored there.}%
\verse{And as they were getting out of the boat, people\lebnote{“\textit{people}” is supplied as the subject of the verb because the third person pronoun “they” could be confused with the disciples getting out of the boat earlier in this verse} recognized him immediately.}%
\verse{They ran about through that whole region and began to carry around those who were sick\lebnote{“who were having badly”} on stretchers, wherever they heard that he was.}%
\verse{And wherever he would go, into villages or into towns or to farms, they would put those who were sick in the marketplaces and would implore him that if they could touch even the edge of his cloak. And all those who touched it were healed.}%
\end{biblechapter}%
\begin{biblechapter}% Mark 7
\verseWithHeading{Human Traditions and God’s Commandments}{And the Pharisees and some of the scribes who had come from Jerusalem gathered to him.}%
\verse{And they saw that some of his disciples were eating their\lnCZV{} bread with unclean — that is, unwashed — hands.}%
\verse{(For the Pharisees and all the Jews do not eat unless they wash their hands ritually,\lebnote{“with the fist”; although the exact meaning of the phrase is uncertain, there is general agreement it has to do with the ritual or ceremonial nature of the handwashing} thus\lebnote{“\textit{thus}” is supplied as a component of the participle (“holding fast to”) which is understood as result} holding fast to the traditions of the elders.}%
\verse{And when they come\lebnote{The phrase “\textit{when they come}” is not in the Greek text but is implied} from the marketplace, they do not eat unless they wash. And there are many other traditions\lebnote{The word “\textit{traditions}” is not in the Greek text but is implied} which they have received and\lebnote{“\textit{and}” supplied because previous participle “have received” translated as a finite verb} hold fast to — for example,\lebnote{The phrase “\textit{for example}” is not in the Greek text but is supplied as a clarification in the English translation} the washing of cups and pitchers and bronze kettles and dining couches.)\lebnote{Several important manuscripts omit “and dining couches”}}%
\verse{And the Pharisees and the scribes asked him, “Why do your disciples not live according to the tradition of the elders, but eat their\lnCZV{} bread with unclean hands?”}%
\verse{So he said to them, \JesusWords{“Isaiah prophesied correctly about you hypocrites, as it is written, ‘This people honors me with their\lnCZV{} lips, but their heart is far, far away from me.}}%
\verse{\JesusWords{And they worship me in vain, teaching as doctrines the commandments of men.’}\lebnote{from Isa 29:13}}%
\verse{\JesusWords{Abandoning the commandment of God, you hold fast to the tradition of men.”}}%
\verse{And he said to them, \JesusWords{“You splendidly ignore the commandment of God so that you can keep\lebnote{Some manuscripts have “you can maintain”} your tradition.}}%
\verse{\JesusWords{For Moses said, ‘Honor your father and your mother,’\lebnote{from Exod 20:12; Deut 5:16} and, ‘The one who speaks evil of father or mother must certainly die\lebnote{“let him die the death”}.’\lebnote{from Exod 21:17; Lev 20:9}}}%
\verse{\JesusWords{But you say, ‘If a man says to his\lnCZV{} father or to his\lnCZV{} mother, “Whatever benefit you would have received\lebnote{“you would have been benefited”} from me is corban”\lebnote{A Hebrew term referring to something consecrated as a gift to God and thus not available for ordinary use} (that is, a gift to God),}}%
\verse{\JesusWords{you no longer permit him to do anything for his\lnCZV{} father or his\lnCZV{} mother,}}%
\verse{\JesusWords{thus\lebnote{“\textit{thus}” is supplied as a component of the participle (“making void”) which is understood as result} making void the word of God by your tradition that you have handed down, and you do many similar things such as this.”}}%
\verseWithHeading{Defilement from Within}{And summoning the crowd again, he said to them, \JesusWords{“Listen to me, all of you, and understand:}}%
\verse{\JesusWords{There is nothing outside of a person that is able to defile him by\lebnote{“\textit{by}” is supplied as a component of the participle (“going”) which is understood as means} going into him. But the things that go out of a person are the things that defile a person.”}\lebnote{Most later manuscripts add v. 16, “If anyone has ears to hear, let him hear.”}}%
\verse{\JesusWords{}}%
\verse{And when he entered into the house away from the crowd, his disciples asked him about the parable.}%
\verse{And he said to them, \JesusWords{“So are you also without understanding? Do you not understand that everything that is outside that goes into a person is not able to defile him?}}%
\verse{\JesusWords{For it does not enter into his heart but into his\lnCZV{} stomach, and goes out into the latrine” — thus\lebnote{“\textit{thus}” is supplied as a component of the participle (“declaring”) which is understood as result} purifying all the food.}}%
\verse{And he said, \JesusWords{“What comes out of a person, that defiles a person.}}%
\verse{\JesusWords{For from within, from the heart of people, come evil plans, sexual immoralities, thefts, murders,}}%
\verse{\JesusWords{adulteries, acts of greed, malicious deeds, deceit, licentiousness, envy,\lebnote{“the evil eye”} abusive speech, pride, foolishness.}}%
\verse{\JesusWords{All these evil things come from within and defile a person.”}}%
\verseWithHeading{A Syrophoenician Woman’s Great Faith}{And from there he set out and\lebnote{“\textit{and}” supplied because previous participle “set out” translated as a finite verb} went to the region of Tyre. And when he\lebnote{“\textit{when}” is supplied as a component of the participle (“entered”) which is understood as temporal} entered into a house, he wanted no one to know, and yet he was not able to escape notice.}%
\verse{But immediately a woman whose young daughter was possessed by an unclean spirit, when she\lebnote{“\textit{when}” is supplied as a component of the participle (“heard”) which is understood as temporal} heard about him, came and\lebnote{“\textit{and}” supplied because previous participle “came” translated as a finite verb} fell down at his feet,}%
\verse{Now the woman was a Greek — a Syrophoenician by nationality — and she was asking him that he would expel the demon from her daughter.}%
\verse{And he said to her, \JesusWords{“Let the children be satisfied first, for it is not right to take the children’s bread and throw it\lnCZW{} to the dogs!”}}%
\verse{But she answered and said to him, “Lord, even the dogs under the table eat the children’s crumbs.”}%
\verse{And he said to her, \JesusWords{“Because of this statement, go! The demon has gone out of your daughter.”}}%
\verse{And when she\lebnote{“\textit{when}” is supplied as a component of the participle (“went”) which is understood as temporal} went to her home, she found the child lying on the bed and the demon gone.}%
\verseWithHeading{A Man Deaf and Unable to Speak Healed}{And again he went away from the region of Tyre and\lebnote{“\textit{and}” supplied because previous participle “went away” translated as a finite verb} came through Sidon to the Sea of Galilee, within the region of the Decapolis.}%
\verse{And they brought to him a man who was deaf and had difficulty speaking, and they were imploring him that he would place his\lnCZV{} hand on him.}%
\verse{And he took him away from the crowd by himself and\lebnote{“\textit{and}” supplied because previous participle “took … away” translated as a finite verb} put his fingers into his ears, and after\lebnote{“\textit{after}” is supplied as a component of the participle (“spitting”) which is understood as temporal} spitting, he touched his tongue.}%
\verse{And looking up to heaven, he sighed and said to him, \JesusWords{“Ephphatha!”} (that is, \JesusWords{“Be opened!”}).}%
\verse{And\lebnote{Some manuscripts have “And immediately”} his ears were opened and his difficulty in speaking was removed\lebnote{“the bond of his tongue was loosened”} and he began to speak normally.}%
\verse{And he ordered them that they should say nothing, but as much as he ordered them not to, they proclaimed it\lnCZW{} even more instead.}%
\verse{And they were amazed beyond all measure, saying, “He has done all things well! He even makes the deaf hear and the mute speak!”}%
\end{biblechapter}%
\begin{biblechapter}% Mark 8
\verseWithHeading{The Feeding of Four Thousand}{In those days there was\lebnote{the participle (“was”) is translated as a finite verb because of English style} again a large crowd, and they did not have\lebnote{the participle (“have”) is translated as a finite verb because of English style} anything they could eat. Summoning the disciples, he said to them,}%
\verse{\JesusWords{“I have compassion on the crowd, because they have remained with me three days already and do not have anything to eat.}}%
\verse{\JesusWords{And if I send them away hungry to their homes they will give out on the way, and some of them have come from far away.”}}%
\verse{And his disciples answered him, “Where is anyone able to feed these people with bread here in the desert?”}%
\verse{And he asked them, \JesusWords{“How many loaves do you have?”} So they said, “Seven.”}%
\verse{And he commanded the crowd to recline for a meal on the ground, and taking the seven loaves, after he\lebnote{“\textit{after}” is supplied as a component of the participle (“had given thanks”) which is understood as temporal} had given thanks he broke them\lnCZX{} and began giving\lebnote{Imperfect tense as ingressive (“began giving”)} them\lnCZX{} to his disciples so that they could set them\lnCZX{} before them.\lnCZY{} And they set them\lnCZX{} before the crowd.}%
\verse{And they had a few small fish, and after\lebnote{“\textit{after}” is supplied as a component of the participle (“giving thanks”) which is understood as temporal} giving thanks for them, he said to set these before them also.}%
\verse{And they ate and were satisfied, and they picked up the broken pieces that were left, seven baskets full.}%
\verse{Now there were about four thousand. And he sent them away.}%
\verse{And immediately he got into the boat with his disciples and\lebnote{“\textit{and}” supplied because previous participle “got” translated as a finite verb} went to the district of Dalmanutha.}%
\verseWithHeading{Pharisees Demand a Sign}{And the Pharisees came and began to argue with him, demanding from him a sign from heaven in order to\lebnote{“\textit{in order to}” is supplied as a component of the participle (“test”) which is understood as purpose} test him.}%
\verse{And sighing deeply in his spirit, he said, \JesusWords{“Why does this generation demand a sign? Truly I say to you, no sign will be given to this generation!”}}%
\verse{And he left them, got into the boat\lebnote{the words “\textit{the boat}” are not in the Greek text but must be supplied} again, and\lebnote{“\textit{and}” supplied because previous participles “left” and “got into” translated as finite verbs} went to the other side.}%
\verseWithHeading{Beware the Leaven of the Pharisees and Herod}{And they had forgotten to take bread, and except for one loaf, they did not have any\lnCZX{} with them in the boat.}%
\verse{And he ordered them, saying, \JesusWords{“Watch out! Beware of the leaven of the Pharisees and the leaven of Herod!”}}%
\verse{And they began to discuss with one another that they had no bread.}%
\verse{And knowing this,\lnCZY{} he said to them, \JesusWords{“Why are you discussing that you have no bread? Do you not yet perceive or understand? Have your hearts been hardened?}}%
\verse{\JesusWords{Although you\lnCZZ{} have eyes, do you not see? And although you\lnCZZ{} have ears, do you not hear? And do you not remember?}}%
\verse{\JesusWords{When I broke the five loaves for the five thousand how many baskets full of broken pieces did you pick up?”} They said to him, “Twelve.”}%
\verse{\JesusWords{“When I also\lebnote{Some manuscripts omit “also”} broke\lebnote{The words “\textit{I broke}” are not in the Greek text, but are understood based on the previous verse} the seven loaves\lebnote{The word “\textit{loaves}” is not in the Greek text, but is understood based on the previous verse} for the four thousand, how many baskets full of broken pieces did you pick up?”} And they said to him, “Seven.”}%
\verse{And he said to them, \JesusWords{“Do you not yet understand?”}}%
\verseWithHeading{A Blind Man Healed at Bethsaida}{And they came to Bethsaida. And they brought to him a blind man and implored him that he would touch him.}%
\verse{And he took hold of the blind man’s hand and\lebnote{“\textit{and}” supplied because previous participle “took hold of” translated as a finite verb} led him outside the village, and after\lebnote{“\textit{after}” is supplied as a component of the participle (“spitting”) which is understood as temporal} spitting in his eyes, he placed his hands on him and\lebnote{“\textit{and}” supplied because previous participle “placed” translated as a finite verb} asked him, \JesusWords{“Do you see anything?”}}%
\verse{And looking up he said, “I see people, for I see them\lnCZX{} like trees walking around.”}%
\verse{Then he placed his\lebnote{“the”: the Greek article is used here as a possessive pronoun} hands on his eyes again, and he opened his eyes and was cured, and could see everything clearly.}%
\verse{And he sent him to his home, saying, \JesusWords{“Do not even go into the village.”}}%
\verseWithHeading{Peter’s Confession at Caesarea Philippi}{And Jesus and his disciples went out to the villages of Caesarea Philippi, and on the way he asked his disciples, saying to them, \JesusWords{“Who do people say that I am?”}}%
\verse{And they told him, saying, “John the Baptist, and others Elijah, and others that you are one of the prophets.”}%
\verse{And he asked them, \JesusWords{“But who do you say that I am?”} Peter answered and\lebnote{“\textit{and}” supplied because previous participle “answered” translated as a finite verb} said to him, “You are the Christ!”}%
\verse{And he warned them that they should tell no one about him.}%
\verseWithHeading{Jesus Predicts His Death and Resurrection}{And he began to teach them that it was necessary for the Son of Man to suffer many things and to be rejected by the elders and the chief priests and the scribes, and to be killed, and after three days to rise.}%
\verse{And he was speaking openly about the subject, and Peter took him aside and\lebnote{“\textit{and}” supplied because previous participle “took … aside” translated as a finite verb} began to rebuke him.}%
\verse{But turning around and seeing his disciples, he rebuked Peter and said, \JesusWords{“Get behind me, Satan, because you are not setting your mind on the things of God, but the things of people!”}}%
\verseWithHeading{Taking Up One’s Cross to Follow Jesus}{And summoning the crowd together with his disciples, he said to them, \JesusWords{“If anyone wants to come\lebnote{Some manuscripts have “to follow”} after me, let him deny himself and take up his cross and follow me.}}%
\verse{\JesusWords{For whoever wants to save his life will lose it, but whoever loses his life on account of me and of the gospel will save it.}}%
\verse{\JesusWords{For what does it benefit a person to gain the whole world and forfeit his life?}}%
\verse{\JesusWords{For what can a person give in exchange for his life?}}%
\verse{\JesusWords{For whoever is ashamed of me and my words in this adulterous and sinful generation, the Son of Man will also be ashamed of him when he comes in the glory of his Father with the holy angels.”}}%
\end{biblechapter}%
\begin{biblechapter}% Mark 9
\verse{And he said to them, \JesusWords{“Truly I say to you, that there are some of those standing here who will never experience death until they see the kingdom of God having come with power.”}}%
\verseWithHeading{The Transfiguration}{And after six days, Jesus took along Peter and James and John, and led them to a high mountain by themselves alone. And he was transfigured before them,}%
\verse{and his clothing became radiant — extremely white, like no cloth refiner on earth can make so white.\lebnote{“make white like this”}}%
\verse{And Elijah appeared to them together with Moses, and they were talking with Jesus.}%
\verse{And Peter answered and\lnDAA{} said to Jesus, “Rabbi, it is good that we are here! And let us make three shelters, one for you and one for Moses and one for Elijah.”}%
\verse{(For he did not know what he should answer, because they were terrified.)}%
\verse{And a cloud came, overshadowing them, and a voice came from the cloud, “This is my beloved Son. Listen to him!”}%
\verse{And suddenly, looking around, they no longer saw anyone with them but Jesus alone.}%
\verse{And as\lebnote{“\textit{as}” is supplied as a component of the temporal genitive absolute participle (“were coming down”)} they were coming down from the mountain, he ordered them that they should tell no one the things that they had seen, except when the Son of Man had risen from the dead.}%
\verse{And they kept the matter to themselves, discussing what this rising from the dead meant.\lebnote{“is”}}%
\verse{And they asked him, saying, “Why do\lebnote{“\textit{what is it} that”} the scribes say that Elijah must come first?”}%
\verse{And he said to them, \JesusWords{“Elijah indeed does come first and\lebnote{“\textit{and}” supplied because previous participle “does come” translated as a finite verb} restores all things. And how is it written concerning the Son of Man that he should suffer many things and be treated with contempt?}}%
\verse{\JesusWords{But I tell you that indeed Elijah has come, and they did to him whatever they wanted, just as it is written about him.”}}%
\verseWithHeading{A Demon-possessed Boy Healed}{And when they\lebnote{“\textit{when}” is supplied as a component of the participle (“came”) which is understood as temporal} came to the disciples, they saw a large crowd around them and scribes arguing with them.}%
\verse{And immediately the whole crowd, when they\lnDAB{} saw him, were amazed, and ran up to him\lnDAC{} and\lebnote{“\textit{and}” supplied because previous participle “ran up to” translated as a finite verb} greeted him.}%
\verse{And he asked them, \JesusWords{“What are you arguing about with them?”}}%
\verse{And one individual from the crowd answered him, “Teacher, I brought to you my son who has a spirit that makes him mute.\lebnote{“a mute spirit”}}%
\verse{And whenever it seizes him, it throws him down and he foams at the mouth and grinds his\lebnote{“the”: the Greek article is used here as a possessive pronoun} teeth and becomes paralyzed. And I told your disciples that they should expel it, and they were not able to do so.\lebnote{*The words “\textit{to do so}” are not in the Greek text but are implied}}%
\verse{And he answered them and\lnDAA{} said, \JesusWords{“O unbelieving generation! How long\lnDAD{} will I be with you? How long\lnDAD{} must I put up with you? Bring him to me!”}}%
\verse{And they brought him to him. And when he\lnDAB{} saw him, the spirit immediately convulsed him, and falling on the ground, he began to roll around, foaming at the mouth.}%
\verse{And he asked his father \JesusWords{how long it was since this had been happening to him.} And he said, “From childhood.}%
\verse{And often it has thrown him both into fire and into water, in order that it could destroy him. But if you are able to do anything, have compassion on us and\lebnote{“\textit{and}” is supplied and the following participle (“have compassion”) has been translated as a finite verb and placed before the verb “help” in keeping with English style} help us!”}%
\verse{But Jesus said to him, \JesusWords{“If you are able! All things are possible for the one who believes!”}}%
\verse{Immediately the father of the child cried out and\lebnote{“\textit{and}” supplied because previous participle “cried out” translated as a finite verb} said, “I believe! Help my unbelief!”}%
\verse{Now when\lnDAB{} Jesus saw that a crowd was running together, he rebuked the unclean spirit, saying to it, \JesusWords{“Mute and deaf spirit, I command you, come out of him, and enter into him no more!”}}%
\verse{And it came out, screaming and convulsing him\lnDAC{} greatly, and he became as if he were dead, so that most of them said, “He has died!”}%
\verse{But Jesus took hold of his hand and\lebnote{“\textit{and}” supplied because previous participle “took hold of” translated as a finite verb} raised him up, and he stood up.}%
\verse{And after\lebnote{“\textit{after}” is supplied as a component of the temporal genitive absolute participle (“had entered”)} he had entered into the house, his disciples asked him privately, “Why were we not able to expel it?”}%
\verse{And he said to them, \JesusWords{“This kind can come out by nothing except by prayer.”}\lebnote{Some manuscripts add “and fasting”}}%
\verseWithHeading{Jesus Predicts His Death and Resurrection a Second Time}{And from there they went out and\lebnote{“\textit{and}” supplied because previous participle “went out” translated as a finite verb} passed through Galilee. And he did not want anyone to know,}%
\verse{for he was teaching his disciples and was telling them, \JesusWords{“The Son of Man is being betrayed into the hands of men, and they will kill him. And when he\lebnote{“\textit{when}” is supplied as a component of the participle (“is killed”) which is understood as temporal} is killed, after three days he will rise.”}}%
\verse{But they did not understand the statement, and they were afraid to ask him.}%
\verseWithHeading{The Question About Who Is Greatest}{And they came to Capernaum. And after he\lebnote{“\textit{after}” is supplied as a component of the participle (“was”) which is understood as temporal} was in the house, he asked them, \JesusWords{“What were you discussing on the way?”}}%
\verse{But they were silent, because they had argued with one another on the way about who was greatest.}%
\verse{And he sat down and\lebnote{“\textit{and}” supplied because previous participle “sat down” translated as a finite verb} called the twelve and said to them, \JesusWords{“If anyone wants to be first, he will be last of all and servant of all.”}}%
\verse{And he took a young child and\lebnote{“\textit{and}” supplied because previous participle “took” translated as a finite verb} had him stand among them.\lebnote{“in the midst of them”} And taking him in his arms, he said to them,}%
\verse{\JesusWords{“Whoever welcomes one of the young children such as these in my name welcomes me, and whoever welcomes me does not welcome me, but the one who sent me.”}}%
\verseWithHeading{Whoever Is Not Against Us Is for Us}{John said to him, “Teacher, we saw someone expelling demons in your name, and we tried to prevent him because he was not following us.”}%
\verse{But Jesus said, \JesusWords{“Do not prevent him, because there is no one who does a miracle in my name and will be able soon afterward to speak evil of me.}}%
\verse{\JesusWords{For whoever is not against us is for us.}}%
\verse{\JesusWords{For whoever gives you a cup of water to drink in my name because you are Christ’s, truly I say to you that he will never lose his reward.}}%
\verseWithHeading{Temptations to Sin}{\JesusWords{“And whoever causes one of these little ones who believe in me to sin, it is better for him if instead a large millstone\lebnote{“a millstone of a donkey”} is placed around his neck and he is thrown into the sea.}}%
\verse{\JesusWords{And if your hand causes you to sin, cut it off! It is better for you to enter into life crippled than, having two hands, to go into hell — into the unquenchable fire!\lebnote{Most later Greek manuscripts add v. 44 after v. 43, “where their worm does not die and the fire is not extinguished” (identical to v. 48)}}}%
\verse{\JesusWords{}}%
\verse{\JesusWords{And if your foot causes you to sin, cut it off! It is better for you to enter into life lame than, having two feet, to be thrown into hell!}\lebnote{Most later Greek manuscripts add v. 46 after v. 45, “where their worm does not die and the fire is not extinguished” (identical to v. 48)}}%
\verse{\JesusWords{}}%
\verse{\JesusWords{And if your eye causes you to sin, tear it out! It is better for you to enter into the kingdom of God with one eye than, having two eyes, to be thrown into hell,}}%
\verse{\JesusWords{‘where their worm does not die and the fire is not extinguished.’}\lebnote{from Isa 66:24}}%
\verse{\JesusWords{For everyone will be salted with fire.}}%
\verse{\JesusWords{Salt is good, but if the salt becomes deprived of its salt content, by what can you make it salty? Have salt among yourselves, and be at peace with one another.”}}%
\end{biblechapter}%
\begin{biblechapter}% Mark 10
\verseWithHeading{On Divorce}{And from there he set out and\lebnote{“\textit{and}” supplied because previous participle “set out” translated as a finite verb} came to the region of Judea and the other side of the Jordan, and again crowds came together to him. And again, as he was accustomed to do,\lebnote{*The words “\textit{to do}” are not in the Greek text but are implied} he began to teach\lebnote{Imperfect tense as ingressive (“began to teach”)} them.}%
\verse{And they asked\lebnote{Some manuscripts have “And Pharisees came up and asked”} him if it was permitted for a man to divorce his\lebnote{The pronoun “\textit{his}” is not in the Greek text but is implied} wife, in order to\lebnote{“\textit{in order to}” is supplied as a component of the participle (“test”) which is understood as purpose} test him.}%
\verse{And he answered and\lnDAE{} said to them, \JesusWords{“What did Moses command you?”}}%
\verse{So they said, “Moses permitted a man\lnDAF{} to write a certificate of divorce and to send her\lnDAF{} away.”}%
\verse{But Jesus said to them, \JesusWords{“He wrote this commandment for you because of your hardness of heart.}}%
\verse{\JesusWords{But from the beginning of creation ‘he made them male and female.}\lebnote{from Gen 1:27; 5:2}}%
\verse{\JesusWords{Because of this a man will leave his father and mother and will be joined to his wife,}\lebnote{The earliest and most important manuscripts do not contain the phrase “and be joined to his wife”}}%
\verse{\JesusWords{and the two will become one flesh,’\lebnote{from Gen 2:24} so that they are no longer two but one flesh.}}%
\verse{\JesusWords{Therefore what God has joined together, man must not separate.”}}%
\verse{And in the house again the disciples began to ask\lebnote{Imperfect tense as ingressive (“began to ask”)} him about this.}%
\verse{And he said to them, \JesusWords{“Whoever divorces his wife and marries another commits adultery against her.}}%
\verse{\JesusWords{And if she divorces her husband and\lebnote{“\textit{and}” supplied because previous participle “divorces” translated as a finite verb} marries another, she commits adultery.”}}%
\verseWithHeading{Little Children Brought to Jesus}{And they were bringing young children to him so that he could touch them, but the disciples rebuked them.}%
\verse{But when\lebnote{“\textit{when}” is supplied as a component of the participle (“saw”) which is understood as temporal} Jesus saw it,\lnDAG{} he was indignant, and said to them, \JesusWords{“Let the young children come to me. Do not forbid them, for to such belongs\lebnote{“for of such is”} the kingdom of God.}}%
\verse{\JesusWords{Truly I say to you, whoever does not welcome the kingdom of God like a young child will never enter into it.”}}%
\verse{And after\lebnote{“\textit{after}” is supplied as a component of the participle (“taking … into his arms”) which is understood as temporal} taking them\lnDAF{} into his arms, he blessed them, placing his\lebnote{“the”: the Greek article is used here as a possessive pronoun} hands on them.}%
\verseWithHeading{A Rich Young Man}{And as\lnDAH{} he was setting out on his way, one individual ran up and knelt down before him and\lebnote{“\textit{and}” supplied because two previous participles “ran up” and “knelt down before” translated as finite verbs} asked him, “Good Teacher, what must I do so that I will inherit eternal life?”}%
\verse{So Jesus said to him, \JesusWords{“Why do you call me good? No one is good except God alone.}}%
\verse{\JesusWords{You know the commandments: ‘Do not murder, do not commit adultery, do not steal, do not give false testimony, do not defraud, honor your father and mother.’”\lebnote{from Exod 20:12–16; Deut 5:16–20, except for “do not defraud” which is an allusion to Deut 24:14}}}%
\verse{And he said to him, “Teacher, all these I have observed from my youth.”}%
\verse{And Jesus, looking at him, loved him, and said to him, \JesusWords{“You lack one thing: Go, sell all that you have, and give the proceeds\lnDAF{} to the poor — and you will have treasure in heaven — and come, follow me.”}}%
\verse{But he looked gloomy at the statement and\lebnote{“\textit{and}” supplied because previous participle “looked gloomy” translated as a finite verb} went away sorrowful, because he had\lebnote{“because he was having”} many possessions.}%
\verse{And Jesus looked around and\lebnote{“\textit{and}” supplied because previous participle “looked around” translated as a finite verb} said to his disciples, \JesusWords{“How difficult it is for\lebnote{“with difficulty”} those who possess wealth to enter into the kingdom of God!”}}%
\verse{And the disciples were astounded at his words. But Jesus answered and\lnDAE{} said to them again, \JesusWords{“Children, how difficult it is to enter into the kingdom of God!}}%
\verse{\JesusWords{It is easier for a camel to go through the eye of a needle than for a rich person to enter into the kingdom of God.”}}%
\verse{And they were very astounded, saying to one another, “And who can be saved?”}%
\verse{Jesus looked at them and\lebnote{“\textit{and}” supplied because previous participle “looked at” translated as a finite verb} said, \JesusWords{“With human beings it is impossible, but not with God. For all things are possible with God.”}}%
\verse{Peter began to say to him, “Behold, we have left everything and followed you.”}%
\verse{Jesus said, \JesusWords{“Truly I say to you, there is no one who has left house or brothers or sisters or mother or father or children or fields on account of me and on account of the gospel}}%
\verse{\JesusWords{who will not\lebnote{“unless \textit{he will} not”} receive a hundred times as much now in this time — houses and brothers and sisters and mothers and children and fields, together with persecutions — and in the age to come, eternal life.}}%
\verse{\JesusWords{But many who are first will be last, and the last first.”}}%
\verseWithHeading{Jesus Predicts His Death and Resurrection a Third Time}{Now they were on the road going up to Jerusalem, and Jesus was going on ahead of them. And they were astounded, but those who were following him\lnDAF{} were afraid. And taking aside the twelve again, he began to tell them the things that were about to happen to him:}%
\verse{\JesusWords{“Behold, we are going up to Jerusalem, and the Son of Man will be handed over to the chief priests and the scribes, and they will condemn him to death and will hand him over to the Gentiles.}}%
\verse{\JesusWords{And they will mock him and spit on him and flog him and kill him,\lnDAG{} and after three days he will rise.”}}%
\verseWithHeading{A Request by James and John}{And James and John, the sons of Zebedee, came up to him and\lebnote{“\textit{and}” supplied because participle “said” translated as a finite verb in keeping with English style} said to him, “Teacher, we want you to do for us whatever we ask you.”}%
\verse{And he said to them, \JesusWords{“What do you want that I do\lebnote{Some manuscripts have “do you want me to do”} for you?”}}%
\verse{So they said to him, “Grant to us that we may sit one at your right hand and one at your left in your glory.”}%
\verse{But Jesus said to them, \JesusWords{“You do not know what you are asking! Are you able to drink the cup that I drink, or to be baptized with the baptism that I am baptized with?”}}%
\verse{And they said to him, “We are able.” So Jesus said to them, \JesusWords{“You will drink the cup that I drink, and you will be baptized with the baptism that I am baptized with,}}%
\verse{\JesusWords{but to sit at my right hand or at my left is not mine to grant, but is for those for whom it has been prepared.”}}%
\verse{And when they\lnDAI{} heard this,\lnDAG{} the ten began to be indignant about James and John.}%
\verse{And Jesus called them to himself and\lebnote{“\textit{and}” supplied because previous participle “called … to himself” translated as a finite verb} said to them, \JesusWords{“You know that those who are considered to rule over the Gentiles lord it over them, and their people in high positions exercise authority over them.}}%
\verse{\JesusWords{But it is not like this among you! But whoever wants to become great among you must be your servant,}}%
\verse{\JesusWords{and whoever wants to be most prominent among you must be the slave of all.}}%
\verse{\JesusWords{For even the Son of Man did not come to be served, but to serve, and to give his life as a ransom for many.”}}%
\verseWithHeading{A Blind Man Healed at Jericho}{And they came to Jericho. And as\lnDAH{} he was setting out from Jericho along with his disciples and a large crowd, a blind beggar, Bartimaeus the son of Timaeus, was sitting beside the road.}%
\verse{And when he\lnDAI{} heard that it was Jesus the Nazarene, he began to cry out and say, “Jesus, Son of David, have mercy on me!”}%
\verse{And many people warned him that he should be quiet. But he was crying out even more loudly,\lebnote{“by much more”} “Son of David, have mercy on me!”}%
\verse{And Jesus stopped and\lebnote{“\textit{and}” supplied because previous participle “stopped” translated as a finite verb} said, \JesusWords{“Call him.”} And they called the blind man and\lebnote{“\textit{and}” supplied because previous participle “called” translated as a finite verb} said to him, “Have courage! Get up! He is calling you.”}%
\verse{And he threw off his cloak, jumped up, and\lebnote{“\textit{and}” supplied because two previous participles “threw off” and “jumped up” translated as finite verbs} came to Jesus.}%
\verse{And Jesus answered him and\lnDAE{} said, \JesusWords{“What do you want me to do\lebnote{“that I do”} for you?”} And the blind man said to him, “Rabboni,\lebnote{The Aramaic term is an elevated form of Rabbi} that I may regain my sight.”}%
\verse{And Jesus said to him, \JesusWords{“Go, your faith has healed you.”} And immediately he regained his sight and began to follow\lebnote{Imperfect tense as ingressive (“began to follow”)} him on the road.}%
\end{biblechapter}%
\begin{biblechapter}% Mark 11
\verseWithHeading{The Triumphal Entry}{And when they came near to Jerusalem, to Bethphage and Bethany at the Mount of Olives, he sent two of his disciples}%
\verse{and said to them, \JesusWords{“Go into the village before you, and right away as you\lebnote{“\textit{as}” is supplied as a component of the participle (“enter”) which is understood as temporal} enter into it you will find a colt tied, on which no one has ever sat. Untie it and bring it.}\lnDAJ{}}%
\verse{\JesusWords{And if anyone says to you, ‘Why are you doing this?’ say\lebnote{Some manuscripts omit “that” after “say” here; though understood to be present in the underlying Greek text, it introduces direct discourse here and is left untranslated, functioning much like English quotation marks} ‘The Lord has need of it, and will send it here again at once.’”}}%
\verse{And they went away and found a colt tied at a door outside in the street, and they untied it.}%
\verse{And some of those who were standing there said to them, “What are you doing, untying the colt?”}%
\verse{So they told them, just as Jesus had said, and they allowed them to take it.\lebnote{The phrase “\textit{to take it}” is not in the Greek text, but is implied}}%
\verse{And they brought the colt to Jesus and threw their cloaks over it, and he sat on it.}%
\verse{And many people spread their cloaks on the road, and others spread\lebnote{the verb “\textit{spread}” is an understood repetition of the verb earlier in this verse} leafy branches they\lebnote{the participle “had cut” has been translated as a finite verb; it agrees in number, gender, and case with “others,” so “\textit{they}’ has been supplied to indicate this} had cut from the fields.}%
\verse{And those who went ahead and those who were following were shouting, “Hosanna! Blessed is the one who comes in the name of the Lord!\lebnote{from Ps 118:25–26}}%
\verse{Blessed is the coming kingdom of our father David! Hosanna in the highest heaven!”\lebnote{*Here “heaven” is understood}}%
\verse{And he went into Jerusalem to the temple, and after\lebnote{“\textit{after}” is supplied as a component of the participle (“looking around”) which is understood as temporal} looking around at everything, because\lebnote{“\textit{because}” is supplied as a component of the participle (“was”) which is understood as causal} the hour was already late, he went out to Bethany with the twelve.}%
\verseWithHeading{A Barren Fig Tree Cursed}{And on the next day as\lebnote{“\textit{as}” is supplied as a component of the temporal genitive absolute participle (“were departing”)} they were departing from Bethany, he was hungry.}%
\verse{And when he\lebnote{“\textit{when}” is supplied as a component of the participle (“saw”) which is understood as temporal} saw from a distance a fig tree that had leaves, he went to see if perhaps he would find anything on it. And when he\lebnote{“\textit{when}” is supplied as a component of the participle (“came up”) which is understood as temporal} came up to it he found nothing except leaves, because it was not the season for figs.}%
\verse{And he responded and\lebnote{“\textit{and}” supplied because previous participle “responded” translated as a finite verb} said to it, \JesusWords{“Let no one eat fruit from you any more forever!”}\lebnote{“for the age”} And his disciples heard it.\lnDAJ{}}%
\verseWithHeading{The Cleansing of the Temple}{And they came to Jerusalem. And he entered into the temple courts\lebnote{“\textit{courts}” is supplied to distinguish this area from the interior of the temple building itself} and\lebnote{“\textit{and}” supplied because previous participle “entered” translated as a finite verb} began to drive out those who were selling and those who were buying in the temple courts,\lnDAK{} and overturned the tables of the money changers and the chairs of those who were selling doves.}%
\verse{And he did not permit anyone to carry objects\lebnote{Or “merchandise”} through the temple courts.\lnDAK{}}%
\verse{And he began to teach\lebnote{Imperfect tense as ingressive (“began to teach”)} and was saying to them, \JesusWords{“Is it not written, ‘My house will be called a house of prayer for all the nations,’\lebnote{from Isa 56:7} but you have made it a cave of robbers!”}}%
\verse{And the chief priests and the scribes heard it,\lnDAJ{} and began considering\lebnote{Imperfect tense as ingressive (“began considering”)} how they could destroy him. For they were afraid of him because the whole crowd was astounded by his teaching.}%
\verse{And when evening came they went out of the city.}%
\verseWithHeading{The Barren Fig Tree Withered}{And as they\lebnote{“\textit{as}” is supplied as a component of the participle (“passed by”) which is understood as temporal} passed by early in the morning, they saw the fig tree withered from the roots.}%
\verse{And Peter remembered and\lebnote{“\textit{and}” supplied because previous participle “remembered” translated as a finite verb} said to him, “Rabbi, look! The fig tree that you cursed has withered!”}%
\verse{And Jesus answered and\lebnote{“\textit{and}” supplied because previous participle “answered” translated as a finite verb} said to them, \JesusWords{“Have faith in God!}}%
\verse{\JesusWords{Truly I say to you that whoever says to this mountain, ‘Be lifted up and thrown into the sea!’ and does not doubt in his heart, but believes that what he says will happen, it will be done for him.}}%
\verse{\JesusWords{For this reason I say to you, whatever you pray and ask for, believe that you have received it,\lnDAJ{} and it will be done for you.}}%
\verse{\JesusWords{And whenever you stand praying, if you have anything against anyone, forgive him,\lnDAJ{} so that your Father who is in heaven will also forgive you your sins.”}\lebnote{Most later Greek manuscripts add v. 26 after v. 25, “But if you do not forgive, neither will your Father in heaven forgive your sins”}}%
\verseWithHeading{Jesus’ Authority Challenged}{}%
\verse{And they came again to Jerusalem. And as\lebnote{“\textit{as}” is supplied as a component of the temporal genitive absolute participle (“was walking”)} he was walking in the temple courts,\lnDAK{} the chief priests and the scribes and the elders came up to him}%
\verse{and said to him, “By what authority are you doing these things, or who gave you this authority that you do these things?”}%
\verse{So Jesus said to them, \JesusWords{“I will ask you one question. Answer me and I will tell you by what authority I am doing these things.}}%
\verse{\JesusWords{The baptism of John — was it from heaven or from men? Answer me!”}}%
\verse{And they began to discuss\lebnote{Imperfect tense as ingressive (“began to discuss”)} this\lebnote{supplied from English context} with one another, saying, “What should we say?\lebnote{Some manuscripts omit “What should we say?”} If we say ‘From heaven,’ he will say, ‘Why then did you not believe him?’}%
\verse{But if we say, ‘From men’” — they were afraid of the crowd, because they all looked upon John as truly a prophet.\lebnote{“that he was truly a prophet”}}%
\verse{And they replied to Jesus saying, “We do not know.” And Jesus said to them, \JesusWords{“Neither will I tell you by what authority I am doing these things.”}}%
\end{biblechapter}%
\begin{biblechapter}% Mark 12
\verseWithHeading{The Parable of the Tenant Farmers in the Vineyard}{And he began to speak to them in parables: \JesusWords{“A man planted a vineyard, and put a fence around it, and dug a trough for the winepress, and built a watchtower, and leased it to tenant farmers, and went on a journey.}}%
\verse{\JesusWords{And he sent a slave to the tenant farmers at the proper time, so that he could collect some of the fruit of the vineyard from the tenant farmers.}}%
\verse{\JesusWords{And they seized him and\lnDAL{} beat him\lnDAM{} and sent him\lnDAM{} away empty-handed.}}%
\verse{\JesusWords{And again he sent to them another slave, and that one they struck on the head and dishonored.}}%
\verse{\JesusWords{And he sent another, and that one they killed. And he sent\lebnote{The words “\textit{he sent}” are not in the Greek text, but are an implied repetition from earlier in the verse} many others, some of whom they beat and some of whom they killed.}}%
\verse{\JesusWords{He had one more, a beloved son. Last of all he sent him to them, saying, ‘They will respect my son.’}}%
\verse{\JesusWords{But those tenant farmers said to one another, ‘This is the heir. Come, let us kill him and the inheritance will be ours!’}}%
\verse{\JesusWords{And they seized and\lnDAL{} killed him and threw him out of the vineyard.}}%
\verse{\JesusWords{What\lebnote{Some manuscripts have “What then”} will the owner of the vineyard do? He will come and destroy the tenant farmers and give the vineyard to others.}}%
\verse{\JesusWords{Have you not read this scripture: ‘The stone which the builders rejected, this has become the cornerstone.}\lebnote{“the head of the corner”}}%
\verse{\JesusWords{This came about from the Lord, and it is marvelous in our eyes’?”}\lebnote{from Ps 118:22–23}}%
\verse{And they were seeking to arrest him, and they were afraid of the crowd, because they knew that he had told the parable with reference to them. And they left him and\lebnote{“\textit{and}” supplied because previous participle “left” translated as a finite verb} went away.}%
\verseWithHeading{Paying Taxes to Caesar}{And they sent some of the Pharisees and the Herodians to him so that they could catch him unawares in a statement.}%
\verse{And when they\lebnote{“\textit{when}” is supplied as a component of the participle (“came”) which is understood as temporal} came, they said to him, “Teacher, we know that you are truthful and you do not care what anyone thinks,\lebnote{“it is not a care to you concerning anyone”} because you do not regard the opinion of people\lebnote{“because you do not look at the face of men”} but teach the way of God in truth. Is it permitted to pay taxes to Caesar or not? Should we pay or should we not pay?”}%
\verse{But because he\lebnote{“\textit{because}” is supplied as a component of the participle (“knew”) which is understood as causal} knew their hypocrisy, he said to them, \JesusWords{“Why are you testing me? Bring me a denarius so that I can look at it!”\lnDAN{}}}%
\verse{So they brought one.\lnDAN{} And he said to them, \JesusWords{“Whose image and inscription is this?”} And they said to him, “Caesar’s.”}%
\verse{And Jesus said to them, \JesusWords{“Give to Caesar the things of Caesar, and to God the things of God!”} And they were utterly amazed at him.}%
\verseWithHeading{A Question About Marriage and the Resurrection}{And Sadducees — who say there is no resurrection — came up to him and began to ask\lebnote{Imperfect tense as ingressive (“began to ask”)} him, saying,}%
\verse{“Teacher, Moses wrote for us that if someone’s brother dies and he leaves behind a wife and does not leave a child, that his brother should take the wife and father\lebnote{“raise up”} descendants for his brother.}%
\verse{There were seven brothers, and the first took a wife. And when he\lebnote{“\textit{when}” is supplied as a component of the participle (“died”) which is understood as temporal} died, he did not leave descendants.}%
\verse{And the second took her, and he died without leaving descendants. And the third likewise.}%
\verse{And the seven did not leave descendants. Last of all the woman also died.}%
\verse{In the resurrection, when they rise, whose\lebnote{“who of them”} wife will she be? For the seven had her as wife.}%
\verse{Jesus said to them, \JesusWords{“Are you not deceived because of this, because you\lebnote{“\textit{because}” is supplied as a component of the participle (“know”) which is understood as causal} do not know the scriptures or the power of God?}}%
\verse{\JesusWords{For when they rise from the dead, they neither marry nor are given in marriage, but are like angels in heaven.}}%
\verse{\JesusWords{Now concerning the dead, that they are raised, have you not read in the book of Moses in the passage about the bush\lebnote{“at the bush”} how God spoke to him, saying, ‘I am the God of Abraham and the God of Isaac and the God of Jacob’?}\lebnote{from Exod 3:6}}%
\verse{\JesusWords{He is not God of the dead, but of the living. You are very much mistaken!”}}%
\verseWithHeading{The Greatest Commandment}{And one of the scribes came up and\lebnote{“\textit{and}” supplied because previous participle “came up” translated as a finite verb} heard them debating. When he\lnDAO{} saw that he answered them well, he asked him, “Which commandment is the most important of all?”}%
\verse{Jesus answered, \JesusWords{“The most important is, ‘Listen, Israel! The Lord our God, the Lord is one.}}%
\verse{\JesusWords{And you shall love the Lord your God from your whole heart and from your whole soul and from your whole mind and from your whole strength.’}\lebnote{from Deut 6:4–5; Josh 22:5}}%
\verse{\JesusWords{The second is this: ‘You shall love your neighbor as yourself.’\lebnote{from Lev 19:18} There is no other commandment greater than these.”}}%
\verse{And the scribe said to him, “That is true, Teacher. You have said correctly\lebnote{“in accordance with truth”} that he is one and there is no other except him.}%
\verse{And to love him from your\lnDAP{} whole heart and from your\lnDAP{} whole understanding and from your\lnDAP{} whole strength, and to love your\lnDAP{} neighbor as yourself, is much more than all whole burnt offerings and sacrifices.”}%
\verse{And Jesus, when he\lnDAO{} saw that he had answered thoughtfully, said to him, \JesusWords{“You are not far from the kingdom of God.”} And no one dared to put a question to him any longer.}%
\verseWithHeading{David’s Son and Lord}{And continuing, Jesus said while\lebnote{“\textit{while}” is supplied as a component of the participle (“teaching”) which is understood as temporal} teaching in the temple courts,\lebnote{*Here “\textit{courts}” is supplied to distinguish this area from the interior of the temple building itself} \JesusWords{“How can the scribes say that the Christ is David’s son?}}%
\verse{\JesusWords{David himself said by the Holy Spirit, ‘The Lord said to my Lord, “Sit at my right hand, until I put your enemies under your feet.”’\lebnote{from Ps 110:1} }}%
\verse{David himself calls him ‘Lord,’ and how is he his son?” And the large crowd was listening to him gladly.}%
\verseWithHeading{Warning to Beware of the Scribes}{And in his teaching he said, \JesusWords{“Beware of the scribes, who like walking around in long robes and greetings in the marketplaces}}%
\verse{\JesusWords{and the best seats in the synagogues and the places of honor at banquets,}}%
\verse{\JesusWords{who devour the houses of widows and pray lengthy prayers for the sake of appearance. These will receive more severe condemnation!”}}%
\verseWithHeading{A Poor Widow’s Offering}{And he sat down opposite the contribution box and\lebnote{“\textit{and}” supplied because previous participle “sat down” translated as a finite verb} was observing how the crowd was putting coins into the contribution box. And many rich people were putting in many coins.\lebnote{Although often translated “large sums,” the plural here suggests large numbers of individual coins, which would make an impressive noise}}%
\verse{And one poor widow came and\lebnote{“\textit{and}” supplied because previous participle “came” translated as a finite verb} put in two small copper coins\lebnote{This coin was the \textit{lepton}, worth 1/128 of a denarius} (that is, a penny).\lebnote{This coin was the \textit{quadrans}, the smallest Roman coin, worth 2 \textit{lepta}}}%
\verse{And summoning his disciples, he said to them, \JesusWords{“Truly I say to you that this poor widow put in more than all those who put offerings\lnDAM{} into the contribution box.}}%
\verse{\JesusWords{For they all contributed\lebnote{“put in”} out of their abundance, but she out of her poverty put in everything she had, her whole means of subsistence.”}}%
\end{biblechapter}%
\begin{biblechapter}% Mark 13
\verseWithHeading{The Destruction of the Temple Predicted}{And as\lebnote{“\textit{as}” is supplied as a component of the temporal genitive absolute participle (“was going out”)} he was going out of the temple courts,\lebnote{*Here “\textit{courts}” is supplied to distinguish this area from the interior of the temple building itself} one of his disciples said to him, “Teacher, look! What great stones and what wonderful buildings!”}%
\verse{And Jesus said to him, \JesusWords{“Do you see these great buildings? Not one stone will be left here on another stone that will not be thrown down!”}}%
\verseWithHeading{Signs of the End of the Age}{And as\lebnote{“\textit{as}” is supplied as a component of the temporal genitive absolute participle (“was sitting”)} he was sitting on the Mount of Olives opposite the temple, Peter and James and John and Andrew asked him privately,}%
\verse{“Tell us, when will these things happen, and what will be the sign when all these things are about to be accomplished?”}%
\verse{So Jesus began to say to them, \JesusWords{“Watch out that no one deceives you!}}%
\verse{\JesusWords{Many will come in my name, saying, ‘I am he,’ and they will deceive many.}}%
\verse{\JesusWords{And when you hear about wars and rumors of wars, do not be alarmed. This must happen, but the end is not yet.}}%
\verse{\JesusWords{For nation will rise up against nation and kingdom against kingdom. There will be earthquakes in various places. There will be famines. These things are the beginning of birth pains.}}%
\verseWithHeading{Persecution of Disciples Predicted}{\JesusWords{“But you, watch out for yourselves! They will hand you over to councils and you will be beaten in the synagogues and will have to stand before governors and kings because of me, for a witness to them.}}%
\verse{\JesusWords{And the gospel must first be proclaimed to all the nations.}\lebnote{Or “Gentiles”; the same Greek word can be translated “nations” or “Gentiles” depending on the context}}%
\verse{\JesusWords{And when they arrest you and\lebnote{“\textit{and}” supplied because participle “hand you over” translated as a finite verb in keeping with English style} hand you over, do not be anxious beforehand what you should say, but whatever is given to you at that hour, say this. For you are not the ones who are speaking, but the Holy Spirit.}}%
\verse{\JesusWords{And brother will hand over brother to death, and a father his\lebnote{The word “\textit{his}” is not in the Greek text but is implied} child, and children will rise up against parents and have them put to death.}}%
\verse{\JesusWords{And you will be hated by all because of my name. But the one who endures to the end — this one will be saved.}}%
\verseWithHeading{The Abomination of Desolation}{\JesusWords{“But when you see the abomination of desolation standing where it should not be” (let the one who reads understand), “then those in Judea must flee to the mountains!}}%
\verse{\JesusWords{The one\lebnote{Some manuscripts have “And the one”} who is on his\lnDAQ{} housetop must not come down or go inside to take anything out of his house,}}%
\verse{\JesusWords{and the one who is in the field must not turn back to pick up his cloak.}}%
\verse{\JesusWords{And woe to those who are pregnant\lebnote{“who have in the womb”} and to those who are nursing their babies\lebnote{The words “\textit{their babies}” are not in the Greek text but are supplied as a necessary clarification} in those days!}}%
\verse{\JesusWords{But pray that it will not happen in winter.}}%
\verse{\JesusWords{For in those days there will be tribulation of such a kind as has not happened from the beginning of the creation that God created until now, and never will happen.}}%
\verse{\JesusWords{And if the Lord had not shortened the days, no human being would be saved.\lebnote{“every flesh would not be saved”} But for the sake of the elect, whom he chose, he has shortened the days.}}%
\verse{\JesusWords{“And at that time if anyone should say to you, “Behold, here is the Christ,’ ‘Behold, there he is,’ do not believe him!\lebnote{*supplied from English context}}}%
\verse{\JesusWords{For false messiahs and false prophets will appear, and will produce signs and wonders in order to mislead, if possible, the elect.}}%
\verse{\JesusWords{But you, watch out! I have told you everything ahead of time!}}%
\verseWithHeading{The Arrival of the Son of Man}{\JesusWords{“But in those days, after that tribulation, ‘the sun will be darkened and the moon will not give its light,}}%
\verse{\JesusWords{and the stars will be falling from heaven, and the powers in the heavens will be shaken.’}\lebnote{from Isa 13:10; 34:4}}%
\verse{\JesusWords{And then they will see the Son of Man arriving in the clouds with great power and glory.}}%
\verse{\JesusWords{And then he will send out the angels, and will gather the\lebnote{Some manuscripts have “his”} elect together from the four winds, from the end of the earth to the end of heaven.}}%
\verseWithHeading{The Parable of the Fig Tree}{\JesusWords{“Now learn the parable from the fig tree: Whenever its branch has already become tender and puts forth its\lnDAQ{} leaves, you know that summer is near.}}%
\verse{\JesusWords{So also you, when you see these things happening, know\lebnote{Or “you know”} that he is near, at the door.}}%
\verse{\JesusWords{Truly I say to you that this generation will never pass away until all these things take place!}}%
\verse{\JesusWords{Heaven and earth will pass away, but my words will never pass away.}}%
\verseWithHeading{The Unknown Day and Hour}{\JesusWords{“But concerning that day or hour no one knows — not even the angels in heaven nor the Son — except the Father.}}%
\verse{\JesusWords{Watch out! Be alert, because you do not know when the time is!}}%
\verse{\JesusWords{It is like a man away on a journey, who left his house and gave his slaves authority — to each one his work — and to the doorkeeper he gave orders that he should be on the alert.}}%
\verse{\JesusWords{Therefore be on the alert, for you do not know when the master of the house is coming — whether in the evening, or at midnight, or when the rooster crows, or early in the morning —}}%
\verse{\JesusWords{lest he arrive suddenly and\lebnote{“\textit{and}” supplied because previous participle “arrive” translated as a finite verb} find you sleeping.}}%
\verse{\JesusWords{And what I say to you, I say to everyone: Be on the alert!”}}%
\end{biblechapter}%
\begin{biblechapter}% Mark 14
\verseWithHeading{The Chief Priests and Scribes Plot to Kill Jesus}{Now after two days it was the Passover and the feast of Unleavened Bread, and the chief priests and the scribes were seeking how, after\lebnote{“\textit{after}” is supplied as a component of the participle (“arresting”) which is understood as temporal} arresting him by stealth, they could kill him.\lnDAR{}}%
\verse{For they said, “Not at the feast, lest there be an uproar by the people.”}%
\verseWithHeading{Jesus’ Anointing at Bethany}{And while\lnDAS{} he was at Bethany in the house of Simon the leper, as\lebnote{“\textit{as}” is supplied as a component of the temporal genitive absolute participle (“was reclining for a meal”)} he was reclining for a meal, a woman came holding an alabaster flask of very costly perfumed oil of genuine nard. After\lebnote{“\textit{after}” is supplied as a component of the participle (“breaking”) which is understood as temporal} breaking the alabaster flask, she poured it\lnDAT{} out on his head.}%
\verse{But some were expressing indignation to one another:\lebnote{Or perhaps “within themselves”} “Why has there been this waste of perfumed oil?}%
\verse{For this perfumed oil could have been sold for more than three hundred denarii and given to the poor!” And they began to scold\lebnote{Imperfect tense as ingressive (“began to scold”)} her.}%
\verse{But Jesus said, \JesusWords{“Leave her alone. Why do you cause trouble for her? She has done a good deed to me.}}%
\verse{\JesusWords{For the poor you always have with you, and you can do good for them whenever you want, but you do not always have me.}}%
\verse{\JesusWords{She has done what she could; she has anointed my body beforehand\lebnote{“she has anticipated to anoint my body”} for burial.}}%
\verse{\JesusWords{And truly I say to you, wherever the gospel is proclaimed in the whole world, what she has done will also be told in memory of her.}}%
\verseWithHeading{Judas Arranges to Betray Jesus}{And Judas Iscariot, who was one of the twelve, went to the chief priests in order to betray him to them.}%
\verse{And when\lebnote{“\textit{when}” is supplied as a component of the participle (“heard”) which is understood as temporal} they heard this,\lnDAR{} they were delighted, and promised to give him money. And he began seeking\lebnote{Imperfect tense as ingressive (“began seeking”)} how he could betray him conveniently.}%
\verseWithHeading{Jesus’ Final Passover with the Disciples}{And on the first day of the feast of Unleavened Bread, when they sacrificed the Passover lamb, his disciples said to him, “Where do you want us to go and\lebnote{“and” supplied because previous participle “go” translated as an English infinitive} prepare, so that you can eat the Passover?”}%
\verse{And he sent two of his disciples and said to them, \JesusWords{“Go into the city and a man carrying a jar of water will meet you. Follow him,}}%
\verse{\JesusWords{and wherever he enters, say to the master of the house, ‘The Teacher says, “Where is my guest room where I may eat the Passover with my disciples?”’}}%
\verse{\JesusWords{And he will show you a large upstairs room furnished\lebnote{Or perhaps “paved” or “panelled”} and\lnDAU{} ready, and prepare for us there.”}}%
\verse{And the disciples went out and came into the city and found everything\lnDAT{} just as he had told them, and they prepared the Passover.}%
\verse{And when it\lebnote{“\textit{when}” is supplied as a component of the temporal genitive absolute participle (“was”)} was evening, he arrived with the twelve.}%
\verse{And while\lebnote{“\textit{while}” is supplied as a component of the temporal genitive absolute participle (“were reclining at table”)} they were reclining at table and eating, Jesus said, \JesusWords{“Truly I say to you, that one of you who is eating with me will betray me.”}}%
\verse{They began to be distressed and to say to him one by one, “Surely not I?”\lebnote{The negative construction in Greek anticipates a negative answer here}}%
\verse{But he said to them, \JesusWords{“It is one of the twelve — the one who is dipping bread\lnDAT{} into the bowl with me.}}%
\verse{\JesusWords{For the Son of Man is going just as it is written about him, but woe to that man by whom the Son of Man is betrayed! It would be better for him if that man had not been born.”}}%
\verseWithHeading{The Lord’s Supper}{And while\lebnote{“\textit{while}” is supplied as a component of the temporal genitive absolute participle (“were eating”)} they were eating, he took bread and,\lebnote{*Here “\textit{and}” is supplied in the translation because of English style} after\lebnote{“\textit{after}” is supplied as a component of the participle (“giving thanks”) which is understood as temporal} giving thanks, he broke it\lnDAT{} and gave it\lnDAT{} to them and said, \JesusWords{“Take it,\lnDAR{} this is my body.”}}%
\verse{And after\lebnote{“\textit{after}” is supplied as a component of the participle (“taking”) which is understood as temporal} taking the cup and\lnDAU{} giving thanks, he gave it\lnDAT{} to them, and they all drank from it.}%
\verse{And he said to them, \JesusWords{“This is my blood of the covenant which is poured out for many.}}%
\verse{\JesusWords{Truly I say to you that I will never drink of the fruit of the vine any longer until that day when I drink it new in the kingdom of God.”}}%
\verse{And after they\lebnote{“\textit{after}” is supplied as a component of the participle (“had sung the hymn”) which is understood as temporal} had sung the hymn, they went out to the Mount of Olives.}%
\verseWithHeading{Jesus Predicts Peter’s Denial}{And Jesus said to them, \JesusWords{“You will all fall away, because it is written, ‘I will strike the shepherd and the sheep will be scattered.’}\lebnote{from Zech 13:7}}%
\verse{\JesusWords{But after I am raised, I will go ahead of you into Galilee.”}}%
\verse{But Peter said to him, “Even if they all fall away, certainly I will not!”}%
\verse{And Jesus said to him, \JesusWords{“Truly I say to you that today — this night — before the rooster crows twice, you will deny me three times!”}}%
\verse{But he kept saying emphatically, “If it is necessary for me to die with you, I will never deny you!” And they all were saying the same thing also.}%
\verseWithHeading{The Prayer in Gethsemane}{And they came to a place named\lebnote{“the name of which”} Gethsemane, and he said to his disciples, \JesusWords{“Sit here while I pray.”}}%
\verse{And he took along Peter and James and John with him, and he began to be distressed and troubled.}%
\verse{And he said to them, \JesusWords{“My soul is deeply grieved, to the point of death. Remain here and stay awake.”}}%
\verse{And going forward a little he fell to the ground and began to pray\lebnote{Imperfect tense as ingressive (“began to pray”)} that, if it were possible, the hour would pass from him.}%
\verse{And he said, \JesusWords{“Abba,\lebnote{The word “Abba” means “father” in Aramaic} Father, all things are possible for you! Take away this cup from me! Yet not what I will, but what you will.”}\lebnote{*Here the verb “\textit{will}” is an understood repetition of the verb earlier in this verse}}%
\verse{And he came and found them sleeping, and he said to Peter, \JesusWords{“Simon, are you sleeping? Were you not able to stay awake one hour?}}%
\verse{\JesusWords{Stay awake and pray that you will not enter into temptation. The spirit is willing, but the flesh is weak!”}}%
\verse{And again he went away and\lebnote{“\textit{and}” supplied because previous participle “went away” translated as a finite verb} prayed, saying the same thing.}%
\verse{And again he came and\lebnote{“\textit{and}” supplied because previous participle “came” translated as a finite verb} found them sleeping, for they could not keep their eyes open,\lebnote{“for their eyes were weighed down”} and they did not know what to reply to him.}%
\verse{And he came the third time and said to them, \JesusWords{“Are you still sleeping and resting? It is enough! The hour has come. Behold, the Son of Man is being betrayed into the hands of sinners.}}%
\verse{\JesusWords{Get up, let us go! Behold, the one who is betraying me is approaching!”}}%
\verseWithHeading{The Betrayal and Arrest of Jesus}{And immediately, while\lebnote{“\textit{while}” is supplied as a component of the temporal genitive absolute participle (“was … speaking”)} he was still speaking, Judas — one of the twelve — arrived, and with him a crowd with swords and clubs, from the chief priests and the scribes and the elders.}%
\verse{Now the one who was betraying him had given them a sign, saying, “The one whom I kiss — he is the one.\lebnote{*Here the predicate nominative (“\textit{the one}”) is implied} Arrest him and lead him\lnDAT{} away under guard!”}%
\verse{And when he\lebnote{“\textit{when}” is supplied as a component of the participle (“arrived”) which is understood as temporal} arrived, he came up to him immediately and\lebnote{“\textit{and}” supplied because previous participle “came up” translated as a finite verb} said, “Rabbi,” and kissed him.}%
\verse{So they laid hands on him and arrested him.}%
\verse{But a certain one of the bystanders, drawing his\lnDAV{} sword, struck the slave of the high priest and cut off his ear.}%
\verse{And Jesus answered and\lebnote{“\textit{and}” supplied because previous participle “answered” translated as a finite verb} said to them, \JesusWords{“Have you come out with swords and clubs, as against a robber, to arrest me?}}%
\verse{\JesusWords{Every day I was with you in the temple courts\lebnote{“\textit{courts}” is supplied to distinguish this area from the interior of the temple building itself} teaching, and you did not arrest me! But this has happened\lebnote{The phrase “\textit{this has happened}” is not in the Greek text, but is understood and must be supplied in the translation because of English style; cf. the parallel in Matt 26:56} in order that the scriptures would be fulfilled.}}%
\verse{And they all abandoned him and\lebnote{“\textit{and}” supplied because previous participle “abandoned” translated as a finite verb} fled.}%
\verse{And a certain young man was following him, clothed only in a linen cloth on his naked body. And they attempted to seize\lebnote{the present tense is translated as a conative present (“attempted to”)} him,}%
\verse{but he left behind the linen cloth and\lebnote{“\textit{and}” supplied because previous participle “left behind” translated as a finite verb} fled naked.}%
\verseWithHeading{Jesus Before the Sanhedrin}{And they led Jesus away to the high priest, and all the chief priests and the elders and the scribes came together.}%
\verse{And Peter followed him from a distance, right inside, into the courtyard of the high priest. And he was sitting with the officers and warming himself by the fire.}%
\verse{Now the chief priests and the whole Sanhedrin were looking for testimony against Jesus in order to put him to death, and they did not find it.\lnDAR{}}%
\verse{For many gave false testimony against him, and their\lnDAV{} testimony was not consistent.}%
\verse{And some stood up and\lnDAW{} began to give false testimony\lebnote{Imperfect tense as ingressive (“began to give false testimony”)} against him, saying,}%
\verse{“We heard him saying, \JesusWords{‘I will destroy this temple made by hands, and within three days I will build another not made by hands.”}}%
\verse{And their testimony was not even consistent about this.}%
\verse{And the high priest stood up in the midst of them and\lnDAW{} asked Jesus, saying, “Do you not reply anything? What are these people testifying against you?”}%
\verse{But he was silent and did not reply anything. Again the high priest asked him and said to him, “Are you the Christ, the Son of the Blessed One?”}%
\verse{And Jesus said, \JesusWords{“I am, and you will see the Son of Man sitting at the right hand of the Power\lebnote{An indirect way of referring to God} and coming with the clouds of heaven.”}}%
\verse{And the high priest tore his clothes and\lebnote{“\textit{and}” supplied because previous participle “tore” translated as a finite verb} said, “What further need do we have of witnesses?}%
\verse{You have heard the blasphemy! What do you think?”\lebnote{“does it seem to you”} And they all condemned him as deserving death.\lebnote{“to be deserving of death”}}%
\verse{And some began to spit on him and to cover his face and to strike him with their fists, and to say to him “Prophesy!” And the officers received him with slaps in the face.\lebnote{Or “with blows” (either meaning is possible here)}}%
\verseWithHeading{Peter Denies Jesus Three Times}{And while\lnDAS{} Peter was below in the courtyard, one of the female slaves of the high priest came up}%
\verse{And when\lnDAX{} she saw Peter warming himself, she looked intently at him and\lebnote{“\textit{and}” supplied because previous participle “looked intently at” translated as a finite verb} said, “You also were with the Nazarene, Jesus.”}%
\verse{But he denied it,\lnDAR{} saying, “I neither know nor understand what you mean!” And he went out into the gateway, and a rooster crowed.\lebnote{Several important and early manuscripts lack the words “and a rooster crowed”}}%
\verse{And the female slave, when she\lnDAX{} saw him, began to say again to the bystanders, “This man is one of them!”}%
\verse{But he denied it\lnDAT{} again. And after a little while, again the bystanders began to say\lebnote{Imperfect tense as ingressive (“began to say”)} to Peter, “You really are one of them, because you also are a Galilean, and your accent shows it!”\lebnote{“is like”}\lebnote{Some manuscripts omit “and your accent shows it”}}%
\verse{And he began to curse and to swear with an oath, “I do not know this man whom you are talking about!”}%
\verse{And immediately a rooster crowed for the second time. And Peter remembered the statement, how Jesus had said to him, \JesusWords{“Before the rooster crows twice, you will deny me three times,”} and throwing himself down, he began to weep.\lebnote{Imperfect tense as ingressive (“began to weep”)}}%
\end{biblechapter}%
\begin{biblechapter}% Mark 15
\verseWithHeading{Jesus Taken to Pilate}{And as soon as morning came, after\lebnote{“\textit{after}” is supplied as a component of the participle (“formulating”) which is understood as temporal} formulating a plan, the chief priests, with the elders and scribes and the whole Sanhedrin, tied up Jesus, led him\lnDAY{} away, and handed him\lnDAY{} over to Pilate.}%
\verse{And Pilate asked him, “Are you the king of the Jews?” And he answered him and\lnDAZ{} said, \JesusWords{“You say so.”}}%
\verse{And the chief priests began to accuse\lebnote{Imperfect tense as ingressive (“began to accuse”)} him of many things.}%
\verse{So Pilate asked him again, saying, “Do you not answer anything? See how many charges\lebnote{The word “charges” is not in the Greek text but is implied} they are bringing against you!”}%
\verse{But Jesus did not answer anything further, so that Pilate was astonished.}%
\verseWithHeading{Pilate Releases Barabbas}{Now at each feast he customarily released\lebnote{The imperfect tense has been translated as customary here (“customarily released”)} for them one prisoner whom they requested.}%
\verse{And the one named Barabbas\lebnote{“Barabbas” means “son of the father” in Aramaic} was imprisoned with the rebels who had committed murder in the rebellion.}%
\verse{And the crowd came up and\lebnote{“\textit{and}” supplied because previous participle “came up” translated as a finite verb} began to ask him to do as he customarily did\lebnote{The imperfect tense has been translated as customary here (“customarily did”)} for them.}%
\verse{So Pilate answered them, saying, “Do you want me to release for you the king of the Jews?”}%
\verse{(For he realized that the chief priests had handed him over because of envy.)}%
\verse{But the chief priests incited the crowd so that he would release for them Barabbas\lnDAZ{} instead.}%
\verse{So Pilate answered and said to them again, “Then what do you want me to do with the one whom you call the king of the Jews?”}%
\verse{And they shouted again, “Crucify him!”}%
\verse{And Pilate said to them, “Why? What evil has he done?” But they shouted even louder, “Crucify him!”}%
\verse{So Pilate, because he\lebnote{“\textit{because}” is supplied as a component of the participle (“wanted”) which is understood as causal} wanted to satisfy\lebnote{“to make sufficient”} the crowd, released for them Barabbas. And after\lebnote{“\textit{after}” is supplied as a component of the participle (“flogged”) which is understood as temporal} he had Jesus flogged, he handed him\lnDAY{} over so that he could be crucified.}%
\verseWithHeading{Jesus Is Mocked}{So the soldiers led him away into the palace (that is, the governor’s residence) and called together the whole cohort.}%
\verse{And they put a purple cloak on him, and after\lebnote{“\textit{after}” is supplied as a component of the participle (“weaving”) which is understood as temporal} weaving a crown of thorns they placed it\lnDAY{} on him.}%
\verse{And they began to greet him, “Hail, king of the Jews!”}%
\verse{And they repeatedly struck\lebnote{The imperfect tense has been translated as iterative here (“repeatedly struck”)} him on the head with a reed, and were spitting on him, and they knelt down\lebnote{“bending the knees”} and\lebnote{“\textit{and}” supplied because previous participle “knelt down” translated as a finite verb} did obeisance to him.}%
\verse{And when they had mocked him, they stripped him of the purple cloak and put his own clothes\lebnote{Some manuscripts have “his clothes” in place of “his own clothes”} on him, and they led him out so that they could crucify him.}%
\verseWithHeading{Jesus Is Crucified}{And they forced a certain man who was passing by, Simon of Cyrene (the father of Alexander and Rufus), who was coming from the country, to carry his cross.}%
\verse{And they brought him to the place Golgotha (which is translated “Place of a Skull”).}%
\verse{And they attempted to give\lebnote{the imperfect tense is translated as a conative imperfect (“attempted to give”)} him wine mixed with myrrh, but he did not take it.}%
\verse{And they crucified him and divided his clothes among themselves\lebnote{“among themselves” reflects the middle voice of the verb “divided”} by\lebnote{“\textit{by}” is supplied as a component of the participle (“casting”) which is understood as means} casting lots for them to see who should take what.}%
\verse{Now it was the third hour when they crucified him.}%
\verse{And the inscription of the charge against him was written, “The king of the Jews.”}%
\verse{And with him they crucified two robbers, one on his right and one on his left.\lebnote{Most later Greek manuscripts add v. 28 (a quotation from Isa 53:12) after v. 27, “And the scripture was fulfilled that says, ‘And he was counted with the lawless ones’”}}%
\verse{}%
\verse{And those who passed by reviled him, shaking their heads and saying, “Aha! The one who would destroy the temple and rebuild it\lnDAY{} in three days,}%
\verse{save yourself by\lebnote{“\textit{by}” is supplied as a component of the participle (“coming down”) which is understood as means} coming down from the cross!”}%
\verse{In the same way also the chief priests, along with the scribes, were mocking him\lnDAY{} to one another, saying, “He saved others; he is not able to save himself!}%
\verse{Let the Christ, the king of Israel, come down now from the cross, so that we may see and believe! Even those who were crucified with him were reviling him.}%
\verseWithHeading{Jesus Dies on the Cross}{And when\lebnote{“\textit{when}” is supplied as a component of the temporal genitive absolute participle (“came”)} the sixth hour came, darkness came over the whole land until the ninth hour.}%
\verse{And at the ninth hour Jesus cried out with a loud voice, \JesusWords{“Eloi, Eloi, lema sabachthani?”} (which is translated, \JesusWords{“My God, my God, why\lebnote{“for what \textit{reason}”} have you forsaken me?”})\lebnote{from Ps 22:1}}%
\verse{And some of the bystanders, when they\lebnote{“\textit{when}” is supplied as a component of the participle (“heard”) which is understood as temporal} heard it,\lebnote{*supplied from English context} said, “Behold, he is summoning Elijah!”}%
\verse{And someone ran and filled a sponge with sour wine, put it\lnDAY{} on a reed, and\lebnote{“\textit{and}” supplied because two previous participles “ran” and “filled” translated as finite verbs} gave it\lnDAY{} to him to drink, saying, “Leave him\lnDAY{} alone! Let us see if Elijah is coming to take him down.”}%
\verse{But Jesus uttered a loud cry and\lebnote{“\textit{and}” supplied because previous participle “uttered” translated as a finite verb} expired.}%
\verse{And the curtain of the temple was torn in two from top to bottom.}%
\verse{And when\lebnote{“\textit{when}” is supplied as a component of the participle (“saw”) which is understood as temporal} the centurion who was standing opposite him saw that he expired like this, he said, “Truly this man was God’s Son!”}%
\verse{And there were also women observing from a distance, among whom were Mary Magdalene, and Mary the mother of James the younger\lebnote{Or perhaps “the short,” referring to stature} and Joses,\lebnote{This name appears in Matt 27:56 as “Joseph”} and Salome,}%
\verse{who used to follow\lebnote{The imperfect tense has been translated as customary here (“used to follow”)} him and serve him when he was in Galilee, and many other women who went up with him to Jerusalem.}%
\verseWithHeading{Jesus Is Buried}{And when it\lebnote{“\textit{when}” is supplied as a component of the temporal genitive absolute participle (“was”)} was already evening, since it was the day of preparation (that is, the day before the Sabbath),}%
\verse{Joseph of Arimathea, a prominent member of the council who was also himself looking forward to\lebnote{Or “waiting for”} the kingdom of God, came acting courageously and\lebnote{“\textit{and}” supplied because previous participle “came” translated as a finite verb} went in to Pilate and asked for the body of Jesus.}%
\verse{And Pilate was surprised that he was already dead, and summoning the centurion, asked him whether he had died already.}%
\verse{And when he\lebnote{“\textit{when}” is supplied as a component of the participle (“learned of”) which is understood as temporal} learned of it\lnDAY{} from the centurion, he granted the corpse to Joseph.}%
\verse{And after\lebnote{“\textit{after}” is supplied as a component of the participle (“purchasing”) which is understood as temporal} purchasing a linen cloth and\lebnote{“\textit{and}” supplied because participle “taking...down” translated as a finite verb in keeping with English style} taking him down, he wrapped him\lnDAY{} in the linen cloth and placed him in a tomb that had been cut from the rock. And he rolled a stone over the entrance of the tomb.}%
\verse{Now Mary Magdalene and Mary the mother of Joses saw where he was placed.}%
\end{biblechapter}%
\begin{biblechapter}% Mark 16
\verseWithHeading{Jesus Is Raised}{And when\lebnote{“\textit{when}” is supplied as a component of the temporal genitive absolute participle (“was over”)} the Sabbath was over, Mary Magdalene, and Mary the mother of James, and Salome purchased fragrant spices so that they could go and\lnDBA{} anoint him.}%
\verse{And very early in the morning on the first day of the week they came to the tomb after\lebnote{“\textit{after}” is supplied as a component of the temporal genitive absolute participle (“had risen”)} the sun had risen.}%
\verse{And they were saying to one another, “Who will roll away the stone for us from the entrance of the tomb?”}%
\verse{And when they\lebnote{“\textit{when}” is supplied as a component of the participle (“looked up”) which is understood as temporal} looked up, they saw that the stone had been rolled away (for it was very large).}%
\verse{And as they\lebnote{“\textit{as}” is supplied as a component of the participle (“going”) which is understood as temporal} were going into the tomb, they saw a young man dressed in a white robe sitting on the right side, and they were alarmed.}%
\verse{But he said to them, “Do not be alarmed. You are looking for Jesus the Nazarene who was crucified. He has been raised, he is not here! See the place where they laid him!}%
\verse{But go, tell his disciples and Peter that he is going ahead of you to Galilee. You will see him there, just as he told you.”}%
\verse{And they went out and\lnDBB{} fled from the tomb, because trembling and amazement had seized them. And they said nothing to anyone, because they were afraid.\lebnote{The Gospel of Mark ends at this point in some manuscripts, including two of the most important ones, while other manuscripts supply a shorter ending (sometimes included as part of v. 8), others supply the traditional longer ending (vv. 9–20), and still other manuscripts supply both the shorter ending and vv. 9–20; due to significant questions about the authenticity of these alternative endings, many scholars regard 16:8 as the last verse of the Gospel of Mark}\innerVerseHeading{The Shorter Ending of Mark}So they promptly reported all the things they had been commanded to those around Peter. And after these things, Jesus himself also sent out through them from the east even as far as the west the holy and imperishable proclamation of eternal salvation. Amen.}%
\verseWithHeading{The Longer Ending of Mark}{Now early on the first day of the week, after he\lebnote{“\textit{after}” is supplied as a component of the participle (“rose”) which is understood as temporal} rose, he appeared first to Mary Magdalene, from whom he had expelled seven demons.}%
\verse{She went out and\lnDBB{} announced it\lnDBC{} to those who were with him while they\lebnote{“\textit{while}” is supplied as a component of the participle (“were”) which is understood as temporal} were mourning and weeping.}%
\verse{And those, when they\lebnote{“\textit{when}” is supplied as a component of the participle (“heard”) which is understood as temporal} heard that he was alive and had been seen by her, refused to believe it.\lebnote{*supplied from English context}}%
\verse{And after these things, he appeared in a different form to two of them as they\lebnote{“\textit{as}” is supplied as a component of the participle (“were walking”) which is understood as temporal} were walking, while they\lebnote{“\textit{while}” is supplied as a component of the participle (“were going out”) which is understood as temporal} were going out into the countryside.}%
\verse{And these went and\lebnote{“\textit{and}” supplied because previous participle “went” translated as a finite verb} reported it\lnDBC{} to the others, and they did not believe them.}%
\verse{And later, while\lebnote{“\textit{while}” is supplied as a component of the participle (“were reclining at table”) which is understood as temporal} they were reclining at table, he appeared to the eleven. And he reprimanded their unbelief and hardness of heart, because they did not believe those who had seen him after he\lebnote{“\textit{after}” is supplied as a component of the participle (“had been raised”) which is understood as temporal} had been raised.}%
\verse{And he said to them, \JesusWords{“Go\lebnote{As a participle of attendant circumstance this participle carries imperatival force picked up from the main verb (“preach”)} into all the world and\lnDBA{} preach the gospel to all creation.}}%
\verse{\JesusWords{The one who believes and is baptized will be saved, but the one who refuses to believe will be condemned.}}%
\verse{\JesusWords{And these signs will accompany those who believe: in my name they will expel demons, they will speak in new tongues,}}%
\verse{\JesusWords{they will pick up\lebnote{Some manuscripts have “and they will pick up”} snakes.\lebnote{Some manuscripts add “with their hands”} And if they drink any deadly poison it will never hurt them; they will lay hands on the sick and they will get\lebnote{“they will have”} well.”}}%
\verse{Then the Lord Jesus, after he had spoken to them, was taken up into heaven and sat down at the right hand of God.}%
\verse{And they went out and\lnDBB{} proclaimed everywhere, while\lebnote{“\textit{while}” is supplied as a component of the temporal genitive absolute participle (“was working together with”)} the Lord was working together with them\lnDBC{} and confirming the message through the accompanying signs.}%
\end{biblechapter}%
