\biblebook{2 Samuel}
\begin{biblechapter}% 2 Samuel 1
\verseWithHeading{The Report of Saul’s Death by the Amalekite}{ויהי Now it came to pass אחרי after מות שׁאול of Saul, ודוד when David שׁב was returned מהכות from the slaughter את העמלק of the Amalekites, וישׁב had abode דוד and David בצקלג in Ziklag; ימים days שׁנים׃ two}%
\verse{ויהי It came even to pass ביום day, השׁלישׁי on the third והנה that, behold, אישׁ a man בא came מן out of המחנה the camp מעם from שׁאול Saul ובגדיו with his clothes קרעים rent, ואדמה and earth על upon ראשׁו his head: ויהי and it was, בבאו when he came אל to דוד David, ויפל that he fell ארצה to the earth, וישׁתחו׃ and did obeisance.}%
\verse{ויאמר said לו דוד And David אי מזהבוא comest ויאמר thou? And he said אליו unto ממחנה ישׂראל of Israel נמלטתי׃ am I escaped.}%
\verse{ויאמר said אליו unto דוד And David מה him, How היה went הדבר the matter? הגד tell נא I pray thee, לי ויאמר me. And he answered, אשׁר That נס are fled העם the people מן from המלחמה the battle, וגם also הרבה and many נפל are fallen מן of העם the people וימתו and dead; וגם also. שׁאול and Saul ויהונתן and Jonathan בנו his son מתו׃ are dead}%
\verse{ויאמר said דוד And David אל unto הנער the young man המגיד that told לו איך him, How ידעת knowest כי thou that מת be dead? שׁאול Saul ויהונתן and Jonathan בנו׃ his son}%
\verse{ויאמר him said, הנער And the young man המגיד that told לו נקרא by chance נקריתי As I happened בהר upon mount הגלבע Gilboa, והנה behold, שׁאול Saul נשׁען leaned על upon חניתו his spear; והנה and, lo, הרכב the chariots ובעלי and horsemen הפרשׁים and horsemen הדבקהו׃ followed hard}%
\verse{ויפן And when he looked אחריו behind ויראני him, he saw ויקרא me, and called אלי unto ואמר me. And I answered, הנני׃}%
\verse{ויאמר And he said לי מי unto me, Who אתה thou? ויאמר And I answered אליו him, עמלקי an Amalekite. אנכי׃ I}%
\verse{ויאמר He said אלי unto עמד me again, Stand, נא I pray thee, עלי upon ומתתני me, and slay כי me: for אחזני is come השׁבץ anguish כי upon me, because כל whole עוד yet נפשׁי׃ my life}%
\verse{ואעמד So I stood עליו upon ואמתתהו him, and slew כי him, because ידעתי I was sure כי that לא he could not יחיה live אחרי after that נפלו he was fallen: ואקח and I took הנזר the crown אשׁר that על upon ראשׁו his head, ואצעדה and the bracelet אשׁר that על on זרעו his arm, ואביאם and have brought אל unto אדני my lord. הנה׃ them hither}%
\verse{ויחזק took hold דוד Then David בבגדו on his clothes, ויקרעם and rent וגם them; and likewise כל all האנשׁים the men אשׁר that אתו׃ with}%
\verse{ויספדו And they mourned, ויבכו and wept, ויצמו and fasted עד until הערב even, על for שׁאול Saul, ועל and for יהונתן Jonathan בנו his son, ועל and for עם the people יהוה ועל and for בית the house ישׂראל of Israel; כי because נפלו they were fallen בחרב׃ by the sword.}%
\verse{ויאמר said דוד And David אל unto הנער the young man המגיד that told לו אי מזהתה thou? ויאמר And he answered, בן the son אישׁ of a stranger, גר of a stranger, עמלקי an Amalekite. אנכי׃ I}%
\verse{ויאמר said אליו unto דוד And David איך him, How לא wast thou not יראת afraid לשׁלח to stretch forth ידך thine hand לשׁחת to destroy את משׁיח anointed? יהוה׃ the LORD's}%
\verse{ויקרא called דוד And David לאחד one מהנערים of the young men, ויאמר and said, גשׁ Go near, פגע fall בו ויכהו upon him. And he smote וימת׃ him that he died.}%
\verse{ויאמר said אליו unto דוד And David דמיך him, Thy blood על upon ראשׁך thy head; כי for פיך thy mouth ענה hath testified בך לאמר against thee, saying, אנכי I מתתי have slain את משׁיח anointed. יהוה׃ the LORD's}%
\verseWithHeading{David Laments Jonathan with the “Song of the Bow”}{ויקנן lamented דוד And David את with הקינה lamentation הזאת this על over שׁאול Saul ועל and over יהונתן Jonathan בנו׃ his son:}%
\verse{ויאמר (Also he bade ללמד them teach בני the children יהודה of Judah קשׁת the bow: הנה behold, כתובה written על in ספר the book הישׁר׃ of Jasher.)}%
\verse{הצבי The beauty ישׂראל of Israel על upon במותיך thy high places: חלל is slain איך how נפלו fallen! גבורים׃ are the mighty}%
\verse{אל not תגידו Tell בגת in Gath, אל not תבשׂרו publish בחוצת in the streets אשׁקלון of Askelon; פן lest תשׂמחנה rejoice, בנות the daughters פלשׁתים of the Philistines פן lest תעלזנה triumph. בנות the daughters הערלים׃ of the uncircumcised}%
\verse{הרי Ye mountains בגלבע of Gilboa, אל no טל dew, ואל neither מטר rain, עליכם upon ושׂדי you, nor fields תרומת of offerings: כי for שׁם there נגעל is vilely cast away, מגן the shield גבורים of the mighty מגן the shield שׁאול of Saul, בלי not משׁיח בשׁמן׃ with oil.}%
\verse{מדם חללים of the slain, מחלב from the fat גבורים of the mighty, קשׁת the bow יהונתן of Jonathan לא not נשׂוג turned אחור back, וחרב and the sword שׁאול of Saul לא not תשׁוב returned ריקם׃ empty.}%
\verse{שׁאול Saul ויהונתן and Jonathan הנאהבים lovely והנעימם and pleasant בחייהם in their lives, ובמותם and in their death לא they were not נפרדו divided: מנשׁרים than eagles, קלו they were swifter מאריות than lions. גברו׃ they were stronger}%
\verse{בנות Ye daughters ישׂראל of Israel, אל over שׁאול Saul, בכינה weep המלבשׁכם who clothed שׁני you in scarlet, עם with עדנים delights, המעלה who put on עדי ornaments זהב of gold על upon לבושׁכן׃ your apparel.}%
\verse{איך How נפלו fallen גברים are the mighty בתוך in the midst המלחמה of the battle! יהונתן O Jonathan, על in במותיך thine high places. חלל׃ slain}%
\verse{צר I am distressed לי עליך for אחי thee, my brother יהונתן Jonathan: נעמת pleasant לי מאד very נפלאתה to me was wonderful, אהבתך hast thou been unto me: thy love לי מאהבת passing the love נשׁים׃ of women.}%
\verse{איך How נפלו fallen, גבורים are the mighty ויאבדו perished! כלי and the weapons מלחמה׃ of war}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 2
\verseWithHeading{David Moves to Hebron}{ויהי And it came to pass אחרי after כן this, וישׁאל inquired דוד that David ביהוה of the LORD, לאמר saying, האעלה Shall I go up באחת into any ערי of the cities יהודה of Judah? ויאמר said יהוה And the LORD אליו unto עלה him, Go up. ויאמר said, דוד And David אנה Whither אעלה shall I go up? ויאמר And he said, חברנה׃ Unto Hebron.}%
\verse{ויעל went up שׁם thither, דוד So David וגם also, שׁתי and his two נשׁיו wives אחינעם Ahinoam היזרעלית the Jezreelitess, ואביגיל and Abigail אשׁת wife נבל Nabal's הכרמלי׃ the Carmelite.}%
\verse{ואנשׁיו And his men אשׁר that עמו with העלה bring up, דוד him did David אישׁ every man וביתו with his household: וישׁבו and they dwelt בערי in the cities חברון׃ of Hebron.}%
\verseWithHeading{David Anointed King over Judah at Hebron}{ויבאו came, אנשׁי And the men יהודה of Judah וימשׁחו they anointed שׁם and there את דוד David למלך king על over בית the house יהודה of Judah. ויגדו And they told לדוד David, לאמר saying, אנשׁי the men יבישׁ גלעדשׁר that קברו buried את שׁאול׃ Saul.}%
\verse{וישׁלח sent דוד And David מלאכים messengers אל unto אנשׁי the men יבישׁ גלעדיאמר and said אליהם unto ברכים them, Blessed אתם ye ליהוה of the LORD, אשׁר that עשׂיתם ye have showed החסד kindness הזה this עם unto אדניכם your lord, עם unto שׁאול Saul, ותקברו and have buried אתו׃}%
\verse{ועתה And now יעשׂ show יהוה the LORD עמכם unto חסד kindness ואמת and truth וגם also אנכי you: and I אעשׂה will requite אתכם הטובה kindness, הזאת you this אשׁר because עשׂיתם ye have done הדבר thing. הזה׃ this}%
\verse{ועתה Therefore now תחזקנה be strengthened, ידיכם let your hands והיו and be לבני ye valiant: חיל ye valiant: כי for מת is dead, אדניכם your master שׁאול Saul וגם and also אתי משׁחו have anointed בית the house יהודה of Judah למלך me king עליהם׃ over}%
\verseWithHeading{Ish-Bosheth over Israel}{ואבנר But Abner בן the son נר of Ner, שׂר captain צבא host, אשׁר לשׁאול of Saul's לקח took את אישׁ בשׁת Ish-bosheth בן the son שׁאול of Saul, ויעברהו and brought him over מחנים׃ to Mahanaim;}%
\verse{וימלכהו And made him king אל over הגלעד Gilead, ואל and over האשׁורי ואל and over יזרעאל Jezreel, ועל and over אפרים Ephraim, ועל and over בנימן Benjamin, ועל and over ישׂראל Israel. כלה׃ all}%
\verse{בן son ארבעים forty שׁנה years אישׁ בשׁת Ish-bosheth בן old שׁאול Saul's במלכו when he began to reign על over ישׂראל Israel, ושׁתים two שׁנים years. מלך and reigned אך But בית the house יהודה of Judah היו followed אחרי followed דוד׃ David.}%
\verse{ויהי was מספר הימיםשׁר that היה was דוד David מלך king בחברון in Hebron על over בית the house יהודה of Judah שׁבע seven שׁנים years ושׁשׁה and six חדשׁים׃ months.}%
\verseWithHeading{War between Judah and Israel}{ויצא went out אבנר And Abner בן the son נר of Ner, ועבדי and the servants אישׁ בשׁת of Ish-bosheth בן the son שׁאול of Saul, ממחנים גבעונה׃ to Gibeon.}%
\verse{ויואב And Joab בן the son צרויה of Zeruiah, ועבדי and the servants דוד of David, יצאו went out, ויפגשׁום and met על by ברכת the pool גבעון of Gibeon: יחדו together וישׁבו and they sat down, אלה the one על of הברכה the pool, מזה on the one side ואלה and the other על of הברכה the pool. מזה׃ on the other side}%
\verse{ויאמר said אבנר And Abner אל to יואב Joab, יקומו arise, נא now הנערים Let the young men וישׂחקו and play לפנינו before ויאמר said, יואב us. And Joab יקמו׃ Let them arise.}%
\verse{ויקמו Then there arose ויעברו and went over במספר by number שׁנים twelve עשׂר twelve לבנימן of Benjamin, ולאישׁ בשׁת which to Ish-bosheth בן the son שׁאול of Saul, ושׁנים and twelve עשׂר and twelve מעבדי of the servants דוד׃ of David.}%
\verse{ויחזקו And they caught אישׁ every one בראשׁ by the head, רעהו his fellow וחרבו and his sword בצד side; רעהו in his fellow's ויפלו so they fell down יחדו together: ויקרא was called למקום place ההוא wherefore that חלקת הצרים Helkath-hazzurim, אשׁר which בגבעון׃ in Gibeon.}%
\verse{ותהי And there was המלחמה battle קשׁה sore עד a very מאד a very ביום day; ההוא that וינגף was beaten, אבנר and Abner ואנשׁי and the men ישׂראל of Israel, לפני before עבדי the servants דוד׃ of David.}%
\verse{ויהיו were שׁם And there שׁלשׁה three בני sons צרויה of Zeruiah יואב there, Joab, ואבישׁי and Abishai, ועשׂהאל and Asahel: ועשׂהאל and Asahel קל light ברגליו of foot כאחד as a הצבים roe. אשׁר wild בשׂדה׃ wild}%
\verse{וירדף pursued עשׂהאל And Asahel אחרי after אבנר Abner; ולא not נטה he turned ללכת and in going על to הימין the right ועל hand nor to השׂמאול the left מאחרי from following אבנר׃ Abner.}%
\verse{ויפן looked אבנר Then Abner אחריו behind ויאמר him, and said, האתה thou זה עשׂהאל Asahel? ויאמר And he answered, אנכי׃ I}%
\verse{ויאמר said לו אבנר And Abner נטה to him, Turn thee aside לך על to ימינך thy right או hand or על to שׂמאלך thy left, ואחז and lay thee hold לך אחד on one מהנערים of the young men, וקח and take לך את חלצתו thee his armor. ולא not אבה would עשׂהאל But Asahel לסור turn aside מאחריו׃ from following}%
\verse{ויסף again עוד אבנר And Abner לאמר said אל to עשׂהאל Asahel, סור Turn thee aside לך מאחרי from following למה me: wherefore אככה should I smite ארצה thee to the ground? ואיך how אשׂא then should I hold up פני my face אל to יואב Joab אחיך׃ thy brother?}%
\verse{וימאן Howbeit he refused לסור to turn aside: ויכהו smote אבנר wherefore Abner באחרי with the hinder end of החנית the spear אל him under החמשׁ the fifth ותצא came out החנית that the spear מאחריו behind ויפל him; and he fell down שׁם there, וימת and died תחתו in the same place: ויהי and it came to pass, כל as many הבא as came אל to המקום the place אשׁר where נפל fell down שׁם עשׂהאל Asahel וימת and died ויעמדו׃ stood still.}%
\verse{וירדפו pursued יואב Joab ואבישׁי also and Abishai אחרי after אבנר Abner: והשׁמשׁ and the sun באה went down והמה when they באו were come עד to גבעת the hill אמה of Ammah, אשׁר that על before פני before גיח Giah דרך by the way מדבר of the wilderness גבעון׃ of Gibeon.}%
\verse{ויתקבצו gathered themselves together בני And the children בנימן of Benjamin אחרי after אבנר Abner, ויהיו and became לאגדה troop, אחת one ויעמדו and stood על on ראשׁ the top גבעה hill. אחת׃ of a}%
\verse{ויקרא called אבנר Then Abner אל to יואב Joab, ויאמר and said, הלנצח forever? תאכל devour חרב Shall the sword הלוא thou not ידעתה knowest כי that מרה bitterness תהיה it will be באחרונה in the latter end? ועד how long מתי לא shall it be then, ere תאמר thou bid לעם the people לשׁוב return מאחרי from following אחיהם׃ their brethren?}%
\verse{ויאמר said, יואב And Joab חי liveth, האלהים God כי unless לולא unless דברת thou hadst spoken, כי surely אז then מהבקר in the morning נעלה had gone up העם the people אישׁ every one מאחרי from following אחיו׃ his brother.}%
\verse{ויתקע blew יואב So Joab בשׁופר a trumpet, ויעמדו stood כל and all העם the people ולא no ירדפו still, and pursued עוד more, אחרי after ישׂראל Israel ולא neither יספו they any more. עוד להלחם׃ fought}%
\verse{ואבנר And Abner ואנשׁיו הלכו walked בערבה through the plain, כל all הלילה night ההוא that ויעברו and passed over את הירדן Jordan, וילכו and went through כל all הבתרון Bithron, ויבאו and they came מחנים׃ to Mahanaim.}%
\verse{ויואב And Joab שׁב returned מאחרי from following אבנר Abner: ויקבץ אתל העםיפקדו there lacked מעבדי דודשׁעה nineteen עשׂר nineteen אישׁ men ועשׂהאל׃ and Asahel.}%
\verse{ועבדי But the servants דוד of David הכו had smitten מבנימן ובאנשׁי men, אבנר and of Abner's שׁלשׁ three מאות hundred ושׁשׁים and threescore אישׁ men מתו׃ died.}%
\verse{וישׂאו And they took up את עשׂהאל Asahel, ויקברהו and buried בקבר him in the sepulcher אביו of his father, אשׁר which בית לחם Bethlehem. וילכו went כל all הלילה night, יואב And Joab ואנשׁיו and his men ויאר at break of day. להם בחברון׃ and they came to Hebron}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 3
\verseWithHeading{The House of David Grows Stronger}{ותהי Now there was המלחמה war ארכה long בין between בית the house שׁאול of Saul ובין בית and the house דוד of David: ודוד but David הלך waxed stronger and stronger, וחזק waxed stronger and stronger, ובית and the house שׁאול of Saul הלכים waxed ודלים׃ weaker and weaker.}%
\verse{וילדו born לדוד And unto David בנים were sons בחברון in Hebron: ויהי was בכורו and his firstborn אמנון Amnon, לאחינעם of Ahinoam היזרעאלת׃ the Jezreelitess;}%
\verse{ומשׁנהו And his second, כלאב Chileab, לאביגל of Abigail אשׁת the wife נבל of Nabal הכרמלי the Carmelite; והשׁלשׁי and the third, אבשׁלום Absalom בן the son מעכה of Maacah בת the daughter תלמי of Talmai מלך king גשׁור׃ of Geshur;}%
\verse{והרביעי And the fourth, אדניה Adonijah בן the son חגית of Haggith; והחמישׁי and the fifth, שׁפטיה Shephatiah בן the son אביטל׃ of Abital;}%
\verse{והשׁשׁי And the sixth, יתרעם Ithream, לעגלה by Eglah אשׁת wife. דוד David's אלה These ילדו were born לדוד to David בחברון׃ in Hebron.}%
\verse{ויהי And it came to pass, בהיות while there was המלחמה war בין between בית the house שׁאול of Saul ובין בית and the house דוד of David, ואבנר that Abner היה מתחזק made himself strong בבית for the house שׁאול׃ of Saul.}%
\verse{ולשׁאול And Saul פלגשׁ had a concubine, ושׁמה whose name רצפה Rizpah, בת the daughter איה of Aiah: ויאמר and said אל to אבנר Abner, מדוע Wherefore באתה hast thou gone אל in unto פילגשׁ concubine? אבי׃ my father's}%
\verse{ויחר wroth לאבנר Then was Abner מאד very על for דברי the words אישׁ בשׁת of Ish-bosheth, ויאמר and said, הראשׁ head, כלב a dog's אנכי I אשׁר which ליהודה against Judah היום this day אעשׂה do show חסד kindness עם unto בית the house שׁאול of Saul אביך thy father, אל to אחיו his brethren, ואל and to מרעהו his friends, ולא and have not המציתך delivered ביד thee into the hand דוד of David, ותפקד that thou chargest עלי that thou chargest עון with a fault האשׁה concerning this woman? היום׃ me today}%
\verse{כה So יעשׂה do אלהים God לאבנר to Abner, וכה also, יסיף and more לו כי except, כאשׁר as נשׁבע hath sworn יהוה the LORD לדוד to David, כי even כן so אעשׂה׃ I do}%
\verse{להעביר To translate הממלכה the kingdom מבית from the house שׁאול of Saul, ולהקים and to set up את כסא the throne דוד of David על over ישׂראל Israel ועל and over יהודה Judah, מדן ועד even to באר שׁבע׃ Beer-sheba.}%
\verse{ולא not יכל And he could עוד again, להשׁיב answer את אבנר Abner דבר a word מיראתו because he feared אתו׃}%
\verseWithHeading{Abner Pledges Support for David}{וישׁלח sent אבנר And Abner מלאכים messengers אל to דוד David תחתו on his behalf, לאמר saying, למי Whose ארץ the land? לאמר saying כרתה Make בריתך thy league אתי with והנה me, and, behold, ידי my hand עמך with להסב thee, to bring about אליך unto את כל all ישׂראל׃ Israel}%
\verse{ויאמר And he said, טוב Well; אני I אכרת will make אתך with ברית a league אך thee: but דבר thing אחד one אנכי I שׁאל require מאתך of לאמר thee, that is, לא Thou shalt not תראה see את פני my face, כי except אם except לפני thou first הביאך bring את מיכל Michal בת daughter, שׁאול Saul's בבאך when thou comest לראות to see את פני׃ my face.}%
\verse{וישׁלח sent דוד And David מלאכים messengers אל to אישׁ בשׁת Ish-bosheth בן son, שׁאול Saul's לאמר saying, תנה Deliver את אשׁתי my wife את מיכל Michal, אשׁר which ארשׂתי I espoused לי במאה to me for a hundred ערלות foreskins פלשׁתים׃ of the Philistines.}%
\verse{וישׁלח sent, אישׁ בשׁת And Ish-bosheth ויקחה and took מעם her from אישׁ husband, מעם from פלטיאל Phaltiel בן the son לושׁ׃ of Laish.}%
\verse{וילך went אתה with אישׁה And her husband הלוך her along ובכה weeping אחריה behind עד her to בחרים Bahurim. ויאמר Then said אליו unto אבנר Abner לך him, Go, שׁוב return. וישׁב׃ And he returned.}%
\verse{ודבר communication אבנר And Abner היה had עם with זקני the elders ישׂראל of Israel, לאמר saying, גם in times תמול in times גם past שׁלשׁם past הייתם מבקשׁיםת דוד for David למלך king עליכם׃ over}%
\verse{ועתה Now עשׂו then do כי for יהוה the LORD אמר hath spoken אל of דוד David, לאמר saying, ביד By the hand דוד David עבדי of my servant הושׁיע I will save את עמי my people ישׂראל Israel מיד out of the hand פלשׁתים of the Philistines, ומיד and out of the hand כל of all איביהם׃ their enemies.}%
\verse{וידבר spoke גם also אבנר And Abner באזני in the ears בנימין of Benjamin: וילך went גם also אבנר and Abner לדבר to speak באזני in the ears דוד of David בחברון in Hebron את כל all אשׁר that טוב seemed good בעיני seemed good ישׂראל to Israel, ובעיני and that seemed good כל to the whole בית house בנימן׃ of Benjamin.}%
\verse{ויבא came אבנר So Abner אל to דוד David חברון to Hebron, ואתו with עשׂרים and twenty אנשׁים men ויעשׂ made דוד him. And David לאבנר Abner ולאנשׁים and the men אשׁר that אתו with משׁתה׃ him a feast.}%
\verse{ויאמר said אבנר And Abner אל unto דוד David, אקומה I will arise ואלכה and go, ואקבצה and will gather אל unto אדני my lord המלך the king, את כל all ישׂראל Israel ויכרתו that they may make אתך with ברית a league ומלכת thee, and that thou mayest reign בכל over all אשׁר that תאוה desireth. נפשׁך thine heart וישׁלח sent דוד And David את אבנר Abner וילך away; and he went בשׁלום׃ in peace.}%
\verse{והנה And, behold, עבדי the servants דוד of David ויואב and Joab בא came מהגדוד from a troop, ושׁלל spoil רב in a great עמם with הביאו and brought ואבנר them: but Abner איננו not עם with דוד David בחברון in Hebron; כי for שׁלחו he had sent him away, וילך and he was gone בשׁלום׃ in peace.}%
\verse{ויואב When Joab וכל and all הצבא the host אשׁר that אתו with באו him were come, ויגדו they told ליואב Joab, לאמר saying, בא came אבנר Abner בן the son נר of Ner אל to המלך the king, וישׁלחהו and he hath sent him away, וילך and he is gone בשׁלום׃ in peace.}%
\verseWithHeading{Joab Assassinates Abner}{ויבא came יואב Then Joab אל to המלך the king, ויאמר and said, מה What עשׂיתה hast thou done? הנה behold, בא came אבנר Abner אליך unto למה thee; why זה שׁלחתו it thou hast sent him away, וילך and he is quite gone? הלוך׃}%
\verse{ידעת Thou knowest את אבנר Abner בן the son נר of Ner, כי that לפתתך to deceive בא he came ולדעת thee, and to know את מוצאך thy going out ואת מבואךלדעת and to know את כל all אשׁר that אתה thou עשׂה׃ doest.}%
\verse{ויצא was come out יואב And when Joab מעם from דוד David, וישׁלח he sent מלאכים messengers אחרי after אבנר Abner, וישׁבו אתובור from the well הסרה of Sirah: ודוד but David לא not. ידע׃ knew}%
\verse{וישׁב was returned אבנר And when Abner חברון to Hebron, ויטהו took him aside יואב Joab אל in תוך in השׁער the gate לדבר to speak אתו with בשׁלי him quietly, ויכהו and smote שׁם him there החמשׁ under the fifth וימת that he died, בדם for the blood עשׂה אל of Asahel אחיו׃ his brother.}%
\verse{וישׁמע heard דוד when David מאחרי כןיאמר he said, נקי guiltless אנכי I וממלכתי and my kingdom מעם before יהוה the LORD עד עולם forever מדמי from the blood אבנר of Abner בן the son נר׃ of Ner:}%
\verse{יחלו Let it rest על on ראשׁ the head יואב of Joab, ואל and on כל all בית house; אביו his father's ואל and let there not יכרת fail מבית from the house יואב of Joab זב one that hath an issue, ומצרע or that is a leper, ומחזיק or that leaneth בפלך on a staff, ונפל or that falleth בחרב on the sword, וחסר or that lacketh לחם׃ bread.}%
\verse{ויואב So Joab ואבישׁי and Abishai אחיו his brother הרגו slew לאבנר Abner, על because אשׁר because המית he had slain את עשׂהאל Asahel אחיהם their brother בגבעון at Gibeon במלחמה׃ in the battle.}%
\verse{ויאמר said דוד And David אל to יואב Joab, ואל and to כל all העם the people אשׁר that אתו with קרעו him, Rend בגדיכם your clothes, וחגרו and gird שׂקים you with sackcloth, וספדו and mourn לפני before אבנר Abner. והמלך And king דוד David הלך אחרי followed המטה׃ the bier.}%
\verse{ויקברו And they buried את אבנר Abner בחברון in Hebron: וישׂא lifted up המלך and the king את קולו his voice, ויבך and wept אל at קבר the grave אבנר of Abner; ויבכו wept. כל and all העם׃ the people}%
\verse{ויקנן lamented המלך And the king אל over אבנר Abner, ויאמר and said, הכמות dieth? נבל as a fool ימות Died אבנר׃ Abner}%
\verse{ידך Thy hands לא not אסרות bound, ורגליך thy feet לא nor לנחשׁתים into fetters: הגשׁו put כנפול falleth לפני before בני as a man עולה wicked נפלת fellest ויספו again כל thou. And all העם the people לבכות wept עליו׃ over}%
\verse{ויבא came כל And when all העם the people להברות to eat את דוד to cause David לחם meat בעוד while it was yet היום day, וישׁבע swore, דוד David לאמר saying, כה So יעשׂה do לי אלהים God וכה also, יסיף to me, and more כי if אם if לפני till בוא be down. השׁמשׁ the sun אטעם I taste לחם bread, או or כל aught else, מאומה׃ aught else,}%
\verse{וכל And all העם the people הכירו took notice וייטב and it pleased them: בעיניהם , and it pleased them: ככל as whatsoever אשׁר as whatsoever עשׂה did המלך the king בעיני pleased כל all העם the people. טוב׃}%
\verse{וידעו understood כל For all העם the people וכל and all ישׂראל Israel ביום day ההוא that כי that לא not היתה it was מהמלך of the king להמית to slay את אבנר Abner בן the son נר׃ of Ner.}%
\verse{ויאמר said המלך And the king אל unto עבדיו his servants, הלוא ye not תדעו Know כי that שׂר there is a prince וגדול and a great man נפל fallen היום day הזה this בישׂראל׃ in Israel?}%
\verse{ואנכי And I היום this day רך weak, ומשׁוח though anointed מלך king; והאנשׁים men האלה and these בני the sons צרויה of Zeruiah קשׁים too hard ממני for ישׁלם shall reward יהוה me: the LORD לעשׂה the doer הרעה of evil כרעתו׃ according to his wickedness.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 4
\verseWithHeading{Ish-Bosheth is Assassinated}{וישׁמע heard בן son שׁאול And when Saul's כי that מת was dead אבנר Abner בחברון in Hebron, וירפו were feeble, ידיו his hands וכל and all ישׂראל the Israelites נבהלו׃ were troubled.}%
\verse{ושׁני two אנשׁים men שׂרי captains גדודים of bands: היו had בן son שׁאול And Saul's שׁם the name האחד of the one בענה Baanah, ושׁם and the name השׁני of the other רכב Rechab, בני the sons רמון of Rimmon הבארתי a Beerothite, מבני of the children בנימן of Benjamin: כי (for גם also בארות Beeroth תחשׁב was reckoned על to בנימן׃ Benjamin:}%
\verse{ויברחו fled הבארתים And the Beerothites גתימה to Gittaim, ויהיו and were שׁם there גרים sojourners עד until היום day.) הזה׃ this}%
\verse{וליהונתן And Jonathan, בן son, שׁאול Saul's בן had a son נכה lame רגלים of feet. בן old חמשׁ five שׁנים years היה He was בבא came שׁמעת when the tidings שׁאול of Saul ויהונתן and Jonathan מיזרעאל ותשׂאהו took him up, אמנתו and his nurse ותנס and fled: ויהי and it came to pass, בחפזה as she made haste לנוס to flee, ויפל that he fell, ויפסח and became lame. ושׁמו And his name מפיבשׁת׃ Mephibosheth.}%
\verse{וילכו went, בני And the sons רמון of Rimmon הבארתי the Beerothite, רכב Rechab ובענה ויבאו and came כחם about the heat היום of the day אל to בית the house אישׁ בשׁת of Ish-bosheth, והוא who שׁכב lay את משׁכב on a bed הצהרים׃ at noon.}%
\verse{והנה באו And they came עד into תוך the midst הבית of the house, לקחי they would have fetched חטים wheat; ויכהו and they smote אל him under החמשׁ the fifth ורכב and Rechab ובענה and Baanah אחיו his brother נמלטו׃ escaped.}%
\verse{ויבאו For when they came הבית into the house, והוא he שׁכב lay על on מטתו his bed בחדר in his bedchamber, משׁכבו in his bedchamber, ויכהו and they smote וימתהו him, and slew ויסירו אתאשׁו his head, ויקחו him, and took את ראשׁוילכו and got them away דרך through הערבה the plain כל all הלילה׃ night.}%
\verse{ויבאו And they brought את ראשׁ the head אישׁ בשׁת of Ish-bosheth אל unto דוד David חברון to Hebron, ויאמרו and said אל to המלך the king, הנה Behold ראשׁ the head אישׁ בשׁת of Ish-bosheth בן the son שׁאול of Saul איבך thine enemy, אשׁר which בקשׁ sought את נפשׁך thy life; ויתן hath avenged יהוה and the LORD לאדני my lord המלך the king נקמות hath avenged היום day הזה this משׁאול ומזרעו׃ and of his seed.}%
\verse{ויען answered דוד And David את רכב Rechab ואת בענה and Baanah אחיו his brother, בני the sons רמון of Rimmon הבארתי the Beerothite, ויאמר and said להם חי liveth, יהוה unto them, the LORD אשׁר who פדה hath redeemed את נפשׁי my soul מכל out of all צרה׃ adversity,}%
\verse{כי When המגיד one told לי לאמר me, saying, הנה Behold, מת is dead, שׁאול Saul והוא היה to have כמבשׂר brought good tidings, בעיניו thinking ואחזה I took hold בו ואהרגהו of him, and slew בצקלג him in Ziklag, אשׁר who לתתי that I would have given לו בשׂרה׃ him a reward for his tidings:}%
\verse{אף How much more, כי when אנשׁים men רשׁעים wicked הרגו have slain את אישׁ person צדיק a righteous בביתו in his own house על upon משׁכבו his bed? ועתה therefore now הלוא shall I not אבקשׁ require את דמו his blood מידכם of your hand, ובערתי אתכםן of your hand, הארץ׃ the earth?}%
\verse{ויצו commanded דוד And David את הנערים his young men, ויהרגום and they slew ויקצצו them, and cut off את ידיהם their hands ואת רגליהם and their feet, ויתלו and hanged up על over הברכה the pool בחברון in Hebron. ואת ראשׁ the head אישׁ בשׁת of Ish-bosheth, לקחו But they took ויקברו and buried בקבר in the sepulcher אבנר of Abner בחברון׃ in Hebron.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 5
\verseWithHeading{David Anointed as King over All of Israel}{ויבאו Then came כל all שׁבטי the tribes ישׂראל of Israel אל to דוד David חברונה unto Hebron, ויאמרו and spoke, לאמר saying, הננו עצמך thy bone ובשׂרך and thy flesh. אנחנו׃ we}%
\verse{גם Also אתמול in time past, גם שׁלשׁום in time past, בהיות was שׁאול when Saul מלך king עלינו over אתה us, thou הייתה wast מוציא he that leddest out והמבי and broughtest את ישׂראל in Israel: ויאמר said יהוה and the LORD לך אתה to thee Thou תרעה shalt feed את עמי my people את ישׂראל Israel, ואתה and thou תהיה shalt be לנגיד a captain על over ישׂראל׃ Israel.}%
\verse{ויבאו came כל So all זקני the elders ישׂראל of Israel אל to המלך the king חברונה to Hebron; ויכרת made להם המלך and king דוד David ברית a league בחברון with them in Hebron לפני before יהוה the LORD: וימשׁחו and they anointed את דוד David למלך king על over ישׂראל׃ Israel.}%
\verse{בן old שׁלשׁים thirty שׁנה years דוד David במלכו when he began to reign, ארבעים forty שׁנה years. מלך׃ he reigned}%
\verse{בחברון In Hebron מלך he reigned על over יהודה Judah שׁבע seven שׁנים years ושׁשׁה and six חדשׁים months: ובירושׁלם and in Jerusalem מלך he reigned שׁלשׁים thirty ושׁלשׁ and three שׁנה years על over כל all ישׂראל Israel ויהודה׃ and Judah.}%
\verseWithHeading{The Capture and Growth of Jerusalem}{וילך went המלך And the king ואנשׁיו and his men ירושׁלם to Jerusalem אל unto היבסי the Jebusites, יושׁב the inhabitants הארץ of the land: ויאמר which spoke לדוד unto David, לאמר saying, לא thou shalt not תבוא come in הנה hither: כי אםסירך thou take away העורים the blind והפסחים לאמר thinking, לא cannot יבוא come in דוד David הנה׃ hither.}%
\verse{וילכד took דוד Nevertheless David את מצדת the stronghold ציון of Zion: היא the same עיר the city דוד׃ of David.}%
\verse{ויאמר said דוד And David ביום day, ההוא on that כל Whosoever מכה and smiteth יבסי the Jebusites, ויגע getteth up בצנור to the gutter, ואת הפסחים and the lame ואת העורים and the blind, שׂנאו hated נפשׁ soul, דוד of David's על כןאמרו they said, עור The blind ופסח and the lame לא shall not יבוא come אל into הבית׃ the house.}%
\verse{וישׁב dwelt דוד So David במצדה in the fort, ויקרא and called לה עיר it the city דוד of David. ויבן built דוד And David סביב round about מן from המלוא Millo וביתה׃ and inward.}%
\verse{וילך went on, דוד And David הלוך went on, וגדול and grew great, ויהוה and the LORD אלהי God צבאות of hosts עמו׃ with}%
\verse{וישׁלח sent חירם And Hiram מלך king צר of Tyre מלאכים messengers אל to דוד David, ועצי trees, ארזים and cedar וחרשׁי and carpenters, עץ and carpenters, וחרשׁי and masons: אבן קיריבנו and they built בית a house. לדוד׃ David}%
\verse{וידע perceived דוד And David כי that הכינו had established יהוה the LORD למלך him king על over ישׂראל Israel, וכי and that נשׂא he had exalted ממלכתו his kingdom בעבור sake. עמו for his people ישׂראל׃ Israel's}%
\verse{ויקח took דוד And David עוד more פלגשׁים concubines ונשׁים and wives מירושׁלם אחרי after באו he was come מחברון ויולדו born עוד and there were yet לדוד to David. בנים sons ובנות׃ and daughters}%
\verse{ואלה And these שׁמות the names הילדים of those that were born לו בירושׁלם unto him in Jerusalem; שׁמוע Shammua, ושׁובב and Shobab, ונתן and Nathan, ושׁלמה׃ and Solomon,}%
\verse{ויבחר Ibhar ואלישׁוע also, and Elishua, ונפג and Nepheg, ויפיע׃ and Japhia,}%
\verse{ואלישׁמע And Elishama, ואלידע and Eliada, ואליפלט׃ and Eliphalet.}%
\verseWithHeading{War with the Philistines}{וישׁמעו heard פלשׁתים But when the Philistines כי that משׁחו they had anointed את דוד David למלך king על over ישׂראל Israel, ויעלו came up כל all פלשׁתים the Philistines לבקשׁ to seek את דוד David; וישׁמע heard דוד and David וירד and went down אל to המצודה׃ the hold.}%
\verse{ופלשׁתים The Philistines באו also came וינטשׁו and spread themselves בעמק in the valley רפאים׃ of Rephaim.}%
\verse{וישׁאל inquired דוד And David ביהוה of the LORD, לאמר saying, האעלה Shall I go up אל to פלשׁתים the Philistines? התתנם wilt thou deliver בידי them into mine hand? ויאמר said יהוה And the LORD אל unto דוד David, עלה Go up: כי for נתן אתןת הפלשׁתים the Philistines בידך׃ into thine hand.}%
\verse{ויבא came דוד And David בבעל פרצים to Baal-perazim, ויכם smote שׁם them there, דוד and David ויאמר and said, פרץ hath broken forth upon יהוה The LORD את איבי mine enemies לפני before כפרץ me, as the breach מים of waters. על כןרא he called שׁם the name המקום place ההוא of that בעל פרצים׃ Baal-perazim.}%
\verse{ויעזבו they left שׁם And there את עצביהם their images, וישׂאם burned דוד and David ואנשׁיו׃}%
\verse{ויספו again, עוד yet פלשׁתים And the Philistines לעלות came up וינטשׁו and spread themselves בעמק in the valley רפאים׃ of Rephaim.}%
\verse{וישׁאל inquired דוד And when David ביהוה of the LORD, ויאמר he said, לא Thou shalt not תעלה go up; הסב fetch a compass אל fetch a compass אחריהם behind ובאת them, and come להם ממול upon them over against בכאים׃ the mulberry trees.}%
\verse{ויהי And let it be, בשׁמעך when thou hearest את קול the sound צעדה of a going בראשׁי in the tops הבכאים of the mulberry trees, אז that then תחרץ thou shalt bestir כי thyself: for אז then יצא go out יהוה shall the LORD לפניך before להכות thee, to smite במחנה the host פלשׁתים׃ of the Philistines.}%
\verse{ויעשׂ did דוד And David כן so, כאשׁר as צוהו had commanded יהוה the LORD ויך him; and smote את פלשׁתים the Philistines מגבע עדאך until thou come גזר׃ to Gazer.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 6
\verseWithHeading{David Brings the Ark of Adonai to Jerusalem}{ויסף עוד Again, דוד David את כל all בחור בישׂראל of Israel, שׁלשׁים thirty אלף׃ thousand.}%
\verse{ויקם arose, וילך and went דוד And David וכל with all העם the people אשׁר that אתו with מבעלי יהודה להעלות to bring up משׁם from thence את ארון the ark האלהים of God, אשׁר whose נקרא is called שׁם name שׁם by the name יהוה of the LORD צבאות of hosts ישׁב that dwelleth הכרבים the cherubims. עליו׃ that dwelleth}%
\verse{וירכבו And they set את ארון the ark האלהים of God אל upon עגלה cart, חדשׁה a new וישׂאהו and brought מבית it out of the house אבינדב of Abinadab אשׁר that בגבעה ועזא and Uzzah ואחיו and Ahio, בני the sons אבינדב of Abinadab, נהגים drove את העגלה cart. חדשׁה׃ the new}%
\verse{וישׂאהו And they brought מבית it out of the house אבינדב of Abinadab אשׁר which בגבעה עם accompanying ארון the ark האלהים of God: ואחיו and Ahio הלך went לפני before הארון׃ the ark.}%
\verse{ודוד And David וכל and all בית the house ישׂראל of Israel משׂחקים played לפני before יהוה the LORD בכל on all עצי wood, ברושׁים manner of fir ובכנרות even on harps, ובנבלים and on psalteries, ובתפים and on timbrels, ובמנענעים and on cornets, ובצלצלים׃ and on cymbals.}%
\verse{ויבאו And when they came עד to גרן threshingfloor, נכון Nachon's וישׁלח put forth עזא Uzzah אל to ארון the ark האלהים of God, ויאחז and took hold בו כי of it; for שׁמטו shook הבקר׃ the oxen}%
\verse{ויחר was kindled אף And the anger יהוה of the LORD בעזה against Uzzah; ויכהו smote שׁם him there האלהים and God על for השׁל error; וימת he died שׁם and there עם by ארון the ark האלהים׃ of God.}%
\verse{ויחר was displeased, לדוד And David על because אשׁר because פרץ had made יהוה the LORD פרץ a breach בעזה upon Uzzah: ויקרא and he called the name למקום place ההוא of the פרץ עזה Perez-uzzah עד to היום day. הזה׃ this}%
\verse{וירא was afraid דוד And David את יהוה of the LORD ביום day, ההוא that ויאמר and said, איך How יבוא come אלי to ארון shall the ark יהוה׃ of the LORD}%
\verse{ולא not אבה would דוד So David להסיר remove אליו unto את ארון the ark יהוה of the LORD על him into עיר the city דוד of David: ויטהו carried it aside דוד but David בית into the house עבד אדום of Obed-edom הגתי׃ the Gittite.}%
\verse{וישׁב continued ארון And the ark יהוה of the LORD בית in the house עבד אדם of Obed-edom הגתי the Gittite שׁלשׁה three חדשׁים months: ויברך blessed יהוה and the LORD את עבד אדם Obed-edom, ואת כל and all ביתו׃ his household.}%
\verse{ויגד And it was told למלך king דוד David, לאמר saying, ברך hath blessed יהוה The LORD את בית the house עבד אדם of Obed-edom, ואת כל and all אשׁר that לו בעבור unto him, because ארון of the ark האלהים of God. וילך went דוד So David ויעל and brought up את ארון the ark האלהים of God מבית from the house עבד אדם of Obed-edom עיר into the city דוד of David בשׂמחה׃ with gladness.}%
\verse{ויהי And it was כי that when צעדו had gone נשׂאי they that bore ארון the ark יהוה of the LORD שׁשׁה six צעדים paces, ויזבח he sacrificed שׁור oxen ומריא׃ and fatlings.}%
\verse{ודוד And David מכרכר danced בכל with all עז might; לפני before יהוה the LORD ודוד and David חגור girded אפוד ephod. בד׃ with a linen}%
\verse{ודוד So David וכל and all בית the house ישׂראל of Israel מעלים brought up את ארון the ark יהוה of the LORD בתרועה with shouting, ובקול and with the sound שׁופר׃ of the trumpet.}%
\verse{והיה And as ארון the ark יהוה of the LORD בא came עיר into the city דוד of David, ומיכל Michal בת daughter שׁאול Saul's נשׁקפה looked בעד through החלון a window, ותרא and saw את המלך king דוד David מפזז leaping ומכרכר and dancing לפני before יהוה the LORD; ותבז and she despised לו בלבה׃ him in her heart.}%
\verse{ויבאו And they brought in את ארון the ark יהוה of the LORD, ויצגו and set אתו במקומו it in his place, בתוך in the midst האהל of the tabernacle אשׁר that נטה had pitched לו דוד David ויעל offered דוד for it: and David עלות burnt offerings לפני before יהוה the LORD. ושׁלמים׃ and peace offerings}%
\verse{ויכל had made an end דוד And as soon as David מהעלות of offering העולה burnt offerings והשׁלמים and peace offerings, ויברך he blessed את העם the people בשׁם in the name יהוה of the LORD צבאות׃ of hosts.}%
\verse{ויחלק And he dealt לכל among all העם the people, לכל among the whole המון multitude ישׂראל of Israel, למאישׁ as men, ועד as well to אשׁה the women לאישׁ to every one חלת cake לחם of bread, אחת a ואשׁפר good piece אחד and a ואשׁישׁה flagon אחת and a וילך departed כל So all העם the people אישׁ every one לביתו׃ to his house.}%
\verse{וישׁב returned דוד Then David לברך to bless את ביתו his household. ותצא came out מיכל And Michal בת the daughter שׁאול of Saul לקראת to meet דוד David, ותאמר and said, מה How נכבד glorious היום today, מלך was the king ישׂראל of Israel אשׁר who נגלה uncovered himself היום today לעיני in the eyes אמהות of the handmaids עבדיו of his servants, כהגלות נגלותחד as one הרקים׃ of the vain fellows}%
\verse{ויאמר said דוד And David אל unto מיכל Michal, לפני before יהוה the LORD, אשׁר which בחר chose בי מאביך me before thy father, ומכל and before all ביתו his house, לצות to appoint אתי נגיד me ruler על over עם the people יהוה of the LORD, על over ישׂראל Israel: ושׂחקתי therefore will I play לפני before יהוה׃ the LORD.}%
\verse{ונקלתי be more vile עוד And I will yet מזאת than thus, והייתי and will be שׁפל base בעיני in mine own sight: ועם and of האמהות the maidservants אשׁר which אמרת thou hast spoken עמם אכבדה׃ of, of them shall I be had in honor.}%
\verse{ולמיכל Therefore Michal בת the daughter שׁאול of Saul לא no היה had לה ילד child עד unto יום the day מותה׃ of her death.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 7
\verseWithHeading{Adonai Makes a Covenant with David}{ויהי And it came to pass, כי when ישׁב sat המלך the king בביתו in his house, ויהוה and the LORD הניח had given him rest לו מסביב round about מכל from all איביו׃ his enemies;}%
\verse{ויאמר said המלך That the king אל unto נתן Nathan הנביא the prophet, ראה See נא now, אנכי I יושׁב dwell בבית in a house ארזים of cedar, וארון but the ark האלהים of God ישׁב dwelleth בתוך within היריעה׃ curtains.}%
\verse{ויאמר said נתן And Nathan אל to המלך the king, כל all אשׁר that בלבבך in thine heart; לך Go, עשׂה do כי for יהוה the LORD עמך׃ with}%
\verse{ויהי And it came to pass בלילה night, ההוא that ויהי came דבר that the word יהוה of the LORD אל unto נתן Nathan, לאמר׃ saying,}%
\verse{לך Go ואמרת and tell אל and tell עבדי my servant אל דוד David, כה Thus אמר saith יהוה the LORD, האתה Shalt thou תבנה build לי בית me a house לשׁבתי׃ for me to dwell}%
\verse{כי Whereas לא I have not ישׁבתי dwelt בבית in house למיום since the time העלתי that I brought up את בני the children ישׂראל of Israel ממצרים ועד even to היום day, הזה this ואהיה but have מתהלך walked באהל in a tent ובמשׁכן׃ and in a tabernacle.}%
\verse{בכל In all אשׁר wherein התהלכתי I have walked בכל with all בני the children ישׂראל of Israel הדבר I a word דברתי spoke את with אחד any שׁבטי of the tribes ישׂראל of Israel, אשׁר whom צויתי I commanded לרעות to feed את עמי my people את ישׂראל Israel, לאמר saying, למה Why לא ye not בניתם build לי בית me a house ארזים׃ of cedar?}%
\verse{ועתה Now כה therefore תאמר so shalt thou say לעבדי unto my servant לדוד David, כה Thus אמר saith יהוה the LORD צבאות of hosts, אני I לקחתיך took מן thee from הנוה the sheepcote, מאחר from following הצאן the sheep, להיות to be נגיד ruler על over עמי my people, על over ישׂראל׃ Israel:}%
\verse{ואהיה And I was עמך with בכל thee whithersoever אשׁר thee whithersoever הלכת thou wentest, ואכרתה and have cut off את כל all איביך thine enemies מפניך out of thy sight, ועשׂתי and have made לך שׁם name, גדול thee a great כשׁם like unto the name הגדלים of the great אשׁר that בארץ׃ in the earth.}%
\verse{ושׂמתי Moreover I will appoint מקום a place לעמי for my people לישׂראל Israel, ונטעתיו and will plant ושׁכן them, that they may dwell תחתיו in a place ולא no ירגז of their own, and move עוד more; ולא neither יסיפו them any more, בני shall the children עולה of wickedness לענותו afflict כאשׁר as בראשׁונה׃ formerly,}%
\verse{ולמן And as since היום the time אשׁר that צויתי I commanded שׁפטים judges על over עמי my people ישׂראל Israel, והניחתי and have caused thee to rest לך מכל from all איביך thine enemies. והגיד telleth לך יהוה Also the LORD כי thee that בית thee a house. יעשׂה he will make לך יהוה׃}%
\verse{כי And when ימלאו be fulfilled, ימיך thy days ושׁכבת and thou shalt sleep את with אבתיך thy fathers, והקימתי I will set up את זרעך thy seed אחריך after אשׁר thee, which יצא shall proceed ממעיך out of thy bowels, והכינתי and I will establish את ממלכתו׃ his kingdom.}%
\verse{הוא He יבנה shall build בית a house לשׁמי for my name, וכננתי and I will establish את כסא the throne ממלכתו of his kingdom עד forever. עולם׃ forever.}%
\verse{אני I אהיה will be לו לאב his father, והוא and he יהיה shall be לי לבן my son. אשׁר If בהעותו he commit iniquity, והכחתיו I will chasten בשׁבט him with the rod אנשׁים of men, ובנגעי and with the stripes בני of the children אדם׃ of men:}%
\verse{וחסדי But my mercy לא shall not יסור depart away ממנו from כאשׁר him, as הסרתי I took מעם from שׁאול Saul, אשׁר whom הסרתי I put away מלפניך׃ before}%
\verse{ונאמן shall be established ביתך And thine house וממלכתך and thy kingdom עד forever עולם forever לפניך before כסאך thee: thy throne יהיה shall be נכון established עד forever. עולם׃ forever.}%
\verseWithHeading{David Responds to Adonai’s Covenant}{ככל According to all הדברים words, האלה these וככל and according to all החזיון vision, הזה this כן so דבר speak נתן did Nathan אל unto דוד׃ David.}%
\verse{ויבא Then went המלך king דוד David וישׁב in, and sat לפני before יהוה ויאמר and he said, מי Who אנכי I, אדני O Lord יהוה the LORD, ומי and what ביתי my house, כי that הביאתני thou hast brought עד me hitherto? הלם׃ me hitherto?}%
\verse{ותקטן a small עוד was yet זאת And this בעיניך thing in thy sight, אדני O Lord יהוה ותדבר but thou hast spoken גם also אל of בית house עבדך thy servant's למרחוק for a great while to come. וזאת And this תורת the manner האדם of man, אדני O Lord יהוה׃}%
\verse{ומה And what יוסיף say דוד can David עוד more לדבר say אליך unto ואתה thee? for thou, ידעת knowest את עבדך thy servant. אדני Lord יהוה׃ GOD,}%
\verse{בעבור דברךכלבך and according to thine own heart, עשׂית hast thou done את כל all הגדולה great things, הזאת these להודיע know את עבדך׃ to make thy servant}%
\verse{על כןדלת thou art great, אדני God: יהוה כי for אין none כמוך like thee, ואין neither אלהים God זולתך beside בכל thee, according to all אשׁר that שׁמענו we have heard באזנינו׃ with our ears.}%
\verse{ומי And what כעמך like thy people, כישׂראל like Israel, גוי nation אחד one בארץ in the earth אשׁר whom הלכו went אלהים God לפדות to redeem לו לעם for a people ולשׂום to himself, and to make לו שׁם him a name, ולעשׂות and to do לכם הגדולה for you great ונראות things and terrible, לארצך for thy land, מפני before עמך thy people, אשׁר which פדית thou redeemedst לך ממצרים גוים the nations ואלהיו׃ and their gods?}%
\verse{ותכונן For thou hast confirmed לך את עמך to thyself thy people ישׂראל Israel לך לעם a people עד unto thee forever: עולם unto thee forever: ואתה and thou, יהוה LORD, היית art become להם לאלהים׃ their God.}%
\verse{ועתה And now, יהוה O LORD אלהים God, הדבר the word אשׁר that דברת thou hast spoken על concerning עבדך thy servant, ועל and concerning ביתו his house, הקם establish עד forever, עולם , forever, ועשׂה and do כאשׁר as דברת׃ thou hast said.}%
\verse{ויגדל be magnified שׁמך And let thy name עד forever, עולם forever, לאמר saying, יהוה The LORD צבאות of hosts אלהים the God על over ישׂראל Israel: ובית and let the house עבדך of thy servant דוד David יהיה be נכון established לפניך׃ before}%
\verse{כי For אתה thou, יהוה O LORD צבאות of hosts, אלהי God ישׂראל of Israel, גליתה hast revealed את אזן hast revealed עבדך to thy servant, לאמר saying, בית thee a house: אבנה I will build לך על therefore כן therefore מצא found עבדך hath thy servant את לבו in his heart להתפלל to pray אליך unto את התפלה prayer הזאת׃ this}%
\verse{ועתה And now, אדני O Lord יהוה GOD, אתה thou הוא that האלהים God, ודבריך and thy words יהיו be אמת true, ותדבר and thou hast promised אל unto עבדך thy servant: את הטובה goodness הזאת׃ this}%
\verse{ועתה Therefore now הואל let it please וברך thee to bless את בית the house עבדך of thy servant, להיות that it may continue לעולם forever לפניך before כי thee: for אתה thou, אדני O Lord יהוה GOD, דברת hast spoken ומברכתך : and with thy blessing יברך be blessed בית let the house עבדך of thy servant לעולם׃ forever.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 8
\verseWithHeading{David’s Military Victories Continue}{ויהי it came to pass, אחרי And after כן this ויך smote דוד that David את פלשׁתים the Philistines, ויכניעם and subdued ויקח took דוד them: and David את מתג האמה Metheg-ammah מיד out of the hand פלשׁתים׃ of the Philistines.}%
\verse{ויך And he smote את מואב Moab, וימדדם and measured בחבל them with a line, השׁכב אותםרצה to the ground; וימדד measured שׁני even with two חבלים lines להמית he to put to death, ומלא and with one full החבל line להחיות to keep alive. ותהי became מואב And the Moabites לדוד David's לעבדים servants, נשׂאי brought מנחה׃ gifts.}%
\verse{ויך smote דוד David את הדדעזר also Hadadezer, בן the son רחב of Rehob, מלך king צובה of Zobah, בלכתו as he went להשׁיב to recover ידו his border בנהר־׃ at the river}%
\verse{וילכד took דוד And David ממנו from אלף him a thousand ושׁבע and seven מאות hundred פרשׁים horsemen, ועשׂרים and twenty אלף thousand אישׁ footmen: רגלי footmen: ויעקר hamstrung דוד and David את כל all הרכב the chariot ויותר but reserved ממנו of מאה them a hundred רכב׃ chariots.}%
\verse{ותבא came ארם And when the Syrians דמשׂק of Damascus לעזר to succor להדדעזר Hadadezer מלך king צובה of Zobah, ויך slew דוד David בארם of the Syrians עשׂרים and twenty ושׁנים two אלף thousand אישׁ׃ men.}%
\verse{וישׂם put דוד Then David נצבים garrisons בארם in Syria דמשׂק of Damascus: ותהי became ארם and the Syrians לדוד to David, לעבדים servants נושׂאי brought מנחה gifts. וישׁע preserved יהוה And the LORD את דוד David בכל whithersoever אשׁר whithersoever הלך׃ he went.}%
\verse{ויקח took דוד And David את שׁלטי the shields הזהב of gold אשׁר that היו were אל on עבדי the servants הדדעזר of Hadadezer, ויביאם and brought ירושׁלם׃ them to Jerusalem.}%
\verse{ומבטח ומברתירי cities הדדעזר of Hadadezer, לקח took המלך king דוד David נחשׁת brass. הרבה much מאד׃ exceeding}%
\verse{וישׁמע heard תעי When Toi מלך king חמת of Hamath כי that הכה had smitten דוד David את כל all חיל הדדעזר׃ of Hadadezer,}%
\verse{וישׁלח sent תעי Then Toi את יורם Joram בנו his son אל unto המלך king דוד David, לשׁאל to salute לו לשׁלום to salute ולברכו him, and to bless על him, because אשׁר him, because נלחם he had fought בהדדעזר against Hadadezer, ויכהו and smitten כי him: for אישׁ מלחמות wars תעי with Toi. היה had הדדעזר Hadadezer ובידו with היו And brought כלי him vessels כסף of silver, וכלי and vessels זהב of gold, וכלי and vessels נחשׁת׃ of brass:}%
\verse{גם Which also אתם הקדישׁ did dedicate המלך king דוד David ליהוה unto the LORD, עם with הכסף the silver והזהב and gold אשׁר that הקדישׁ he had dedicated מכל of all הגוים nations אשׁר which כבשׁ׃ he subdued;}%
\verse{מארם וממואבמבני and of the children עמון of Ammon, ומפלשׁתים ומעמלקמשׁלל and of the spoil הדדעזר of Hadadezer, בן son רחב of Rehob, מלך king צובה׃ of Zobah.}%
\verse{ויעשׂ got דוד And David שׁם a name בשׁבו when he returned מהכותו from smiting את ארם of the Syrians בגיא in the valley מלח of salt, שׁמונה eighteen עשׂר eighteen אלף׃ thousand}%
\verse{וישׂם And he put באדום in Edom; נצבים garrisons בכל throughout all אדום Edom שׂם put נצבים he garrisons, ויהי became כל and all אדום they of Edom עבדים servants. לדוד David's ויושׁע preserved יהוה And the LORD את דוד David בכל whithersoever אשׁר whithersoever הלך׃ he went.}%
\verse{וימלך reigned דוד And David על over כל all ישׂראל Israel; ויהי דוד and David עשׂה executed משׁפט judgment וצדקה and justice לכל unto all עמו׃ his people.}%
\verse{ויואב And Joab בן the son צרויה of Zeruiah על over הצבא the host; ויהושׁפט and Jehoshaphat בן the son אחילוד of Ahilud מזכיר׃ recorder;}%
\verse{וצדוק And Zadok בן the son אחיטוב of Ahitub, ואחימלך and Ahimelech בן the son אביתר of Abiathar, כהנים the priests; ושׂריה and Seraiah סופר׃ the scribe;}%
\verse{ובניהו And Benaiah בן the son יהוידע of Jehoiada והכרתי both the Cherethites והפלתי and the Pelethites; ובני sons דוד and David's כהנים chief rulers. היו׃ were}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 9
\verseWithHeading{David Cares for Mephibosheth}{ויאמר said, דוד And David הכי ישׁ Is there עוד yet אשׁר any נותר that is left לבית of the house שׁאול of Saul, ואעשׂה עמוסד him kindness בעבור sake? יהונתן׃}%
\verse{ולבית And of the house שׁאול of Saul עבד a servant ושׁמו whose name ציבא Ziba. ויקראו And when they had called לו אל him unto דוד David, ויאמר said המלך the king אליו unto האתה him, thou ציבא Ziba? ויאמר And he said, עבדך׃ Thy servant}%
\verse{ויאמר said, המלך And the king האפס there not עוד yet אישׁ any לבית of the house שׁאול of Saul, ואעשׂה that I may show עמו unto חסד the kindness אלהים of God ויאמר said ציבא him? And Ziba אל unto המלך the king, עוד hath yet בן a son, ליהונתן Jonathan נכה lame רגלים׃ on feet.}%
\verse{ויאמר said לו המלך And the king איפה unto him, Where הוא he? ויאמר said ציבא And Ziba אל unto המלך the king, הנה Behold, הוא he בית in the house מכיר of Machir, בן the son עמיאל of Ammiel, בלו דבר׃ in Lo-debar.}%
\verse{וישׁלח sent, המלך Then king דוד David ויקחהו and fetched מבית him out of the house מכיר of Machir, בן the son עמיאל of Ammiel, מלו דבר׃}%
\verse{ויבא was come מפיבשׁת Now when Mephibosheth, בן the son יהונתן of Jonathan, בן the son שׁאול of Saul, אל unto דוד David, ויפל he fell על on פניו his face, וישׁתחו and did reverence. ויאמר said, דוד And David מפיבשׁת Mephibosheth. ויאמר And he answered, הנה Behold עבדך׃ thy servant!}%
\verse{ויאמר said לו דוד And David אל not: תירא unto him, Fear כי for עשׂה אעשׂהמך חסד thee kindness בעבור sake, יהונתן for Jonathan אביך thy father's והשׁבתי and will restore לך את כל thee all שׂדה the land שׁאול of Saul אביך thy father; ואתה and thou תאכל shalt eat לחם bread על at שׁלחני my table תמיד׃ continually.}%
\verse{וישׁתחו And he bowed himself, ויאמר and said, מה What עבדך thy servant, כי that פנית thou shouldest look אל upon הכלב dog המת a dead אשׁר such כמוני׃ as I}%
\verse{ויקרא called המלך Then the king אל to ציבא Ziba, נער servant, שׁאול Saul's ויאמר and said אליו unto כל all אשׁר that היה pertained לשׁאול to Saul ולכל and to all ביתו his house. נתתי him, I have given לבן son אדניך׃ unto thy master's}%
\verse{ועבדת shall till לו את האדמה the land אתה Thou ובניך therefore, and thy sons, ועבדיך and thy servants, והבאת for him, and thou shalt bring והיה may have לבן son אדניך in that thy master's לחם food ואכלו to eat: ומפיבשׁת but Mephibosheth בן son אדניך thy master's יאכל shall eat תמיד always לחם bread על at שׁלחני my table. ולציבא Now Ziba חמשׁה had fifteen עשׂר had fifteen בנים sons ועשׂרים and twenty עבדים׃ servants.}%
\verse{ויאמר Then said ציבא Ziba אל unto המלך the king, ככל According to all אשׁר that יצוה hath commanded אדני my lord המלך the king את עבדו his servant, כן so יעשׂה do. עבדך shall thy servant ומפיבשׁת As for Mephibosheth, אכל he shall eat על at שׁלחני my table, כאחד as one מבני המלך׃}%
\verse{ולמפיבשׁת And Mephibosheth בן son, קטן had a young ושׁמו whose name מיכא Micha. וכל And all מושׁב that dwelt בית in the house ציבא of Ziba עבדים servants למפיבשׁת׃ unto Mephibosheth.}%
\verse{ומפיבשׁת So Mephibosheth ישׁב dwelt בירושׁלם in Jerusalem: כי for על at שׁלחן table; המלך the king's תמיד continually הוא he אכל did eat והוא פסח and was lame שׁתי on both רגליו׃ his feet.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 10
\verseWithHeading{The Ammonites Refuse David’s Loyal Love}{ויהי And it came to pass אחרי after כן this, וימת died, מלך that the king בני of the children עמון of Ammon וימלך reigned חנון and Hanun בנו his son תחתיו׃ in his stead.}%
\verse{ויאמר Then said דוד David, אעשׂה I will show חסד kindness עם unto חנון Hanun בן the son נחשׁ of Nahash, כאשׁר as עשׂה showed אביו his father עמדי חסד kindness וישׁלח sent דוד me. And David לנחמו to comfort ביד him by the hand עבדיו of his servants אל for אביו his father. ויבאו came עבדי servants דוד And David's ארץ into the land בני of the children עמון׃ of Ammon.}%
\verse{ויאמרו said שׂרי And the princes בני of the children עמון of Ammon אל unto חנון Hanun אדניהם their lord, המכבד doth honor דוד thou that David את אביך thy father, בעיניך Thinkest כי that שׁלח he hath sent לך מנחמים comforters הלוא unto thee? hath not בעבור thee, to חקור search את העיר the city, ולרגלה and to spy it out, ולהפכה and to overthrow שׁלח sent דוד David את עבדיו his servants אליך׃ unto}%
\verse{ויקח took חנון Wherefore Hanun את עבדי servants, דוד David's ויגלח and shaved off את חצי the one half זקנם of their beards, ויכרת and cut off את מדויהם their garments בחצי in the middle, עד to שׁתותיהם their buttocks, וישׁלחם׃ and sent them away.}%
\verse{ויגדו When they told לדוד unto David, וישׁלח he sent לקראתם to meet כי them, because היו were האנשׁים נכלמים ashamed: מאד greatly ויאמר said, המלך and the king שׁבו Tarry בירחו at Jericho עד until יצמח be grown, זקנכם your beards ושׁבתם׃ and return.}%
\verseWithHeading{Israel Fights Ammon and Aram}{ויראו saw בני And when the children עמון of Ammon כי that נבאשׁו they stank בדוד before David, וישׁלחו sent בני the children עמון of Ammon וישׂכרו and hired את ארם the Syrians בית רחוב of Beth-rehob, ואת ארם and the Syrians צובא of Zoba, עשׂרים twenty אלף thousand רגלי footmen, ואת מלך and of king מעכה Maacah אלף a thousand אישׁ men, ואישׁ men. טוב שׁנים twelve עשׂר twelve אלף אישׁ׃ thousand}%
\verse{וישׁמע heard דוד And when David וישׁלח of he sent את יואב Joab, ואת כל and all הצבא the host הגברים׃ of the mighty men.}%
\verse{ויצאו came out, בני And the children עמון of Ammon ויערכו in array מלחמה and put the battle פתח at the entering השׁער in of the gate: וארם and the Syrians צובא of Zoba, ורחוב and of Rehob, ואישׁ טובמעכה and Maacah, לבדם by themselves בשׂדה׃ in the field.}%
\verse{וירא saw יואב When Joab כי that היתה was אליו against פני the front המלחמה of the battle מפנים him before ומאחור and behind, ויבחר he chose מכל of all בחורי בישׂראל of Israel, ויערך and put in array לקראת against ארם׃ the Syrians:}%
\verse{ואת יתר And the rest העם of the people נתן he delivered ביד into the hand אבשׁי of Abishai אחיו his brother, ויערך that he might put in array לקראת against בני the children עמון׃ of Ammon.}%
\verse{ויאמר And he said, אם If תחזק be too strong ארם the Syrians ממני for והיתה me, then thou shalt לי לישׁועה and help ואם me: but if בני the children עמון of Ammon יחזקו be too strong ממך for והלכתי thee, then I will come להושׁיע׃}%
\verse{חזק Be of good courage, ונתחזק and let us play the men בעד for עמנו our people, ובעד and for ערי the cities אלהינו of our God: ויהוה and the LORD יעשׂה do הטוב him good. בעיניו׃ that which seemeth}%
\verse{ויגשׁ drew nigh, יואב And Joab והעם and the people אשׁר that עמו with למלחמה him, unto the battle בארם against the Syrians: וינסו and they fled מפניו׃ before}%
\verse{ובני And when the children עמון of Ammon ראו saw כי that נס were fled, ארם the Syrians וינסו then fled מפני they also before אבישׁי Abishai, ויבאו and entered העיר into the city. וישׁב returned יואב So Joab מעל from בני the children עמון of Ammon, ויבא and came ירושׁלם׃ to Jerusalem.}%
\verseWithHeading{The Arameans Regroup for Attack}{וירא saw ארם And when the Syrians כי that נגף they were smitten לפני before ישׂראל Israel, ויאספו they gathered themselves יחד׃ together.}%
\verse{וישׁלח sent, הדדעזר ויצא and brought out את ארם the Syrians אשׁר that מעבר beyond הנהר the river: ויבאו and they came חילם to Helam; ושׁובך and Shobach שׂר the captain צבא of the host הדדעזר לפניהם׃ before}%
\verse{ויגד And when it was told לדוד David, ויאסף אתל ישׂראליעבר and passed over את הירדן Jordan, ויבא and came חלאמה to Helam. ויערכו set themselves in array ארם And the Syrians לקראת against דוד David, וילחמו and fought עמו׃ with}%
\verse{וינס fled ארם And the Syrians מפני before ישׂראל Israel; ויהרג slew דוד and David מארם שׁבע seven מאות hundred רכב chariots וארבעים and forty אלף thousand פרשׁים horsemen, ואת שׁובך Shobach שׂר the captain צבאו of their host, הכה and smote וימת who died שׁם׃ there.}%
\verse{ויראו saw כל And when all המלכים the kings עבדי servants הדדעזר כי that נגפו they were smitten לפני before ישׂראל Israel, וישׁלמו they made peace את with ישׂראל Israel, ויעבדום and served ויראו feared ארם them. So the Syrians להושׁיע to help עוד any more. את בני the children עמון׃ of Ammon}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 11
\verseWithHeading{David Commits Adultery with Bathsheba}{ויהי And it came to pass, לתשׁובת was expired, השׁנה after the year לעת at the time צאת go forth המלאכים when kings וישׁלח sent דוד that David את יואב Joab, ואת עבדיו and his servants עמו with ואת כל him, and all ישׂראל Israel; וישׁחתו and they destroyed את בני the children עמון of Ammon, ויצרו and besieged על and besieged רבה Rabbah. ודוד But David יושׁב tarried בירושׁלם׃ still at Jerusalem.}%
\verse{ויהי And it came to pass לעת in an eveningtide, הערב in an eveningtide, ויקם arose דוד that David מעל from off משׁכבו his bed, ויתהלך and walked על upon גג the roof בית house: המלך of the king's וירא he saw אשׁה a woman רחצת washing מעל and from הגג the roof והאשׁה herself; and the woman טובת beautiful מראה to look מאד׃ very}%
\verse{וישׁלח sent דוד And David וידרשׁ and inquired לאשׁה after the woman. ויאמר And said, הלוא not זאת this בת־שׁבע Bath-sheba, בת the daughter אליעם of Eliam, אשׁת the wife אוריה of Uriah החתי׃ the Hittite?}%
\verse{וישׁלח sent דוד And David מלאכים messengers, ויקחה and took ותבוא her; and she came אליו in unto וישׁכב him, and he lay עמה with her; והיא for she מתקדשׁת was purified מטמאתה from her uncleanness: ותשׁב and she returned אל unto ביתה׃ her house.}%
\verse{ותהר conceived, האשׁה And the woman ותשׁלח and sent ותגד and told לדוד David, ותאמר and said, הרה with child. אנכי׃ I}%
\verse{וישׁלח sent דוד And David אל to יואב Joab, שׁלח sent אלי to את אוריה me Uriah החתי the Hittite. וישׁלח יואב And Joab את אוריה Uriah אל דוד׃ David.}%
\verse{ויבא was come אוריה And when Uriah אליו unto וישׁאל demanded דוד him, David לשׁלום did, יואב how Joab ולשׁלום did, העם and how the people ולשׁלום prospered. המלחמה׃ and how the war}%
\verse{ויאמר said דוד And David לאוריה to Uriah, רד Go down לביתך to thy house, ורחץ and wash רגליך thy feet. ויצא departed אוריה And Uriah מבית המלך from the king. ותצא and there followed אחריו and there followed משׂאת him a mess המלך׃}%
\verse{וישׁכב slept אוריה But Uriah פתח at the door בית house המלך of the king's את with כל all עבדי the servants אדניו of his lord, ולא not ירד and went אל down to ביתו׃ his house.}%
\verse{ויגדו And when they had told לדוד David, לאמר saying, לא went not down ירד went not down אוריה Uriah אל unto ביתו his house, ויאמר said דוד David אל unto אוריה Uriah, הלוא not מדרך from journey? אתה thou בא Camest מדוע why לא didst thou not ירדת go down אל unto ביתך׃ thine house?}%
\verse{ויאמר said אוריה And Uriah אל unto דוד David, הארון The ark, וישׂראל and Israel, ויהודה and Judah, ישׁבים abide בסכות in tents; ואדני and my lord יואב Joab, ועבדי and the servants אדני of my lord, על in פני the open השׂדה fields; חנים are encamped ואני shall I אבוא then go אל into ביתי mine house, לאכל to eat ולשׁתות and to drink, ולשׁכב and to lie עם with אשׁתי my wife? חיך thou livest, וחי liveth, נפשׁך and thy soul אם אעשׂה I will not do את הדבר thing. הזה׃ this}%
\verse{ויאמר said דוד And David אל to אוריה Uriah, שׁב Tarry בזה here גם also, היום today ומחר and tomorrow אשׁלחך I will let thee depart. וישׁב abode אוריה So Uriah בירושׁלם in Jerusalem ביום day, ההוא that וממחרת׃ and the morrow.}%
\verse{ויקרא had called לו דוד And when David ויאכל him, he did eat לפניו before וישׁת and drink וישׁכרהו him; and he made him drunk: ויצא he went out בערב and at even לשׁכב to lie במשׁכבו on his bed עם with עבדי the servants אדניו of his lord, ואל down to ביתו his house. לא not ירד׃ but went}%
\verse{ויהי And it came to pass בבקר in the morning, ויכתב wrote דוד that David ספר a letter אל to יואב Joab, וישׁלח and sent ביד by the hand אוריה׃ of Uriah.}%
\verse{ויכתב And he wrote בספר in the letter, לאמר saying, הבו Set את אוריה ye Uriah אל in מול the forefront פני the forefront המלחמה battle, החזקה of the hottest ושׁבתם and retire מאחריו ye from ונכה him, that he may be smitten, ומת׃ and die.}%
\verse{ויהי And it came to pass, בשׁמור observed יואב when Joab אל observed העיר the city, ויתן that he assigned את אוריה Uriah אל unto המקום a place אשׁר where ידע he knew כי that אנשׁי men חיל valiant שׁם׃ where}%
\verse{ויצאו went out, אנשׁי העיר of the city וילחמו and fought את with יואב Joab: ויפל and there fell מן of העם the people מעבדי of the servants דוד of David; וימת died גם also. אוריה and Uriah החתי׃ the Hittite}%
\verse{וישׁלח sent יואב Then Joab ויגד and told לדוד David את כל all דברי the things המלחמה׃ concerning the war;}%
\verse{ויצו And charged את המלאך the messenger, לאמר saying, ככלותך When thou hast made an end את כלברי the matters המלחמה of the war לדבר of telling אל unto המלך׃ the king,}%
\verse{והיה so be אם And if תעלה arise, חמת wrath המלך that the king's ואמר and he say לך מדוע unto thee, Wherefore נגשׁתם approached אל ye so nigh unto העיר the city להלחם when ye did fight? הלוא ye not ידעתם knew את אשׁר that ירו they would shoot מעל from החומה׃ the wall?}%
\verse{מי Who הכה smote את אבימלך Abimelech בן the son ירבשׁת of Jerubbesheth? הלוא did not אשׁה a woman השׁליכה cast עליו upon פלח a piece רכב of a millstone מעל him from החומה the wall, וימת that he died בתבץ in Thebez? למה why נגשׁתם went ye nigh אל went ye nigh החומה the wall? ואמרת then say גם also. עבדך thou, Thy servant אוריה Uriah החתי the Hittite מת׃ is dead}%
\verse{וילך went, המלאך So the messenger ויבא and came ויגד and showed לדוד David את כל all אשׁר that שׁלחו had sent יואב׃ Joab}%
\verse{ויאמר said המלאך And the messenger אל unto דוד David, כי Surely גברו prevailed עלינו against האנשׁים the men ויצאו us, and came out אלינו unto השׂדה us into the field, ונהיה and we were עליהם upon עד them even unto פתח the entering השׁער׃ of the gate.}%
\verse{ויראו המוראיםל upon עבדך thy servants; מעל from off החומה the wall וימותו be dead, מעבדי and thy servant המלך וגם also. עבדך אוריה Uriah החתי the Hittite מת׃ is dead}%
\verse{ויאמר said דוד Then David אל unto המלאך the messenger, כה Thus תאמר shalt thou say אל unto יואב Joab, אל Let not ירע בעיניך displease את הדבר thing הזה this כי thee, for כזה as well as another: וכזה one תאכל devoureth החרב the sword החזק more strong מלחמתך make thy battle אל against העיר the city, והרסה and overthrow וחזקהו׃ it: and encourage}%
\verse{ותשׁמע heard אשׁת And when the wife אוריה of Uriah כי that מת was dead, אוריה Uriah אישׁה her husband ותספד she mourned על for בעלה׃ her husband.}%
\verse{ויעבר was past, האבל And when the mourning וישׁלח sent דוד David ויאספה and fetched אל her to ביתו his house, ותהי and she became לו לאשׁה his wife, ותלד and bore לו בן him a son. וירע הדבר But the thing אשׁר that עשׂה had done דוד David בעיני displeased יהוה׃ the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 12
\verseWithHeading{Nathan Reproves David}{וישׁלח sent יהוה And the LORD את נתן Nathan אל unto דוד David. ויבא And he came אליו unto ויאמר him, and said לו שׁני two אנשׁים היו unto him, There were בעיר city; אחת in one אחד the one עשׁיר rich, ואחד and the other ראשׁ׃ poor.}%
\verse{לעשׁיר The rich היה had צאן flocks ובקר and herds: הרבה many מאד׃ exceeding}%
\verse{ולרשׁ But the poor אין had nothing, כל had nothing, כי save אם save כבשׂה ewe lamb, אחת one קטנה little אשׁר which קנה bought ויחיה and nourished up: ותגדל and it grew up עמו with ועם him, and with בניו his children; יחדו together מפתו of his own meat, תאכל it did eat ומכסו of his own cup, תשׁתה and drank ובחיקו in his bosom, תשׁכב and lay ותהי he had לו כבת׃ unto him as a daughter.}%
\verse{ויבא And there came הלך a traveler לאישׁ man, העשׁיר unto the rich ויחמל and he spared לקחת to take מצאנו of his own flock ומבקרו and of his own herd, לעשׂות to dress לארח for the wayfaring man הבא that was come לו ויקח unto him; but took את כבשׂת lamb, האישׁ man's הראשׁ the poor ויעשׂה and dressed לאישׁ it for the man הבא that was come אליו׃ to}%
\verse{ויחר kindled אף anger דוד And David's באישׁ against the man; מאד was greatly ויאמר and he said אל to נתן Nathan, חי liveth, יהוה the LORD כי shall surely die: בן מותאישׁ the man העשׂה that hath done זאת׃ this}%
\verse{ואת הכבשׂה the lamb ישׁלם And he shall restore ארבעתים fourfold, עקב because אשׁר because עשׂה he did את הדבר thing, הזה this ועל and because אשׁר and because לא he had no חמל׃ pity.}%
\verse{ויאמר said נתן And Nathan אל to דוד David, אתה Thou האישׁ the man. כה Thus אמר saith יהוה the LORD אלהי God ישׂראל of Israel, אנכי I משׁחתיך anointed למלך thee king על over ישׂראל Israel, ואנכי and I הצלתיך delivered מיד thee out of the hand שׁאול׃ of Saul;}%
\verse{ואתנה And I gave לך את בית house, אדניך thee thy master's ואת נשׁי wives אדניך and thy master's בחיקך into thy bosom, ואתנה and gave לך את בית thee the house ישׂראל of Israel ויהודה and of Judah; ואם and if מעט too little, ואספה I would moreover have given לך כהנה unto thee such וכהנה׃ and such}%
\verse{מדוע Wherefore בזית hast thou despised את דבר the commandment יהוה of the LORD, לעשׂות to do הרע evil בעינו in his sight? את אוריה Uriah החתי the Hittite הכית thou hast killed בחרב with the sword, ואת אשׁתו his wife לקחת and hast taken לך לאשׁה thy wife, ואתו הרגת and hast slain בחרב him with the sword בני of the children עמון׃ of Ammon.}%
\verse{ועתה Now לא shall never תסור depart חרב therefore the sword מביתך from thine house; עד עולםקב because כי because בזתני thou hast despised ותקח me, and hast taken את אשׁת the wife אוריה of Uriah החתי the Hittite להיות to be לך לאשׁה׃ thy wife.}%
\verse{כה Thus אמר saith יהוה the LORD, הנני מקים I will raise up עליך against רעה evil מביתך thee out of thine own house, ולקחתי and I will take את נשׁיך thy wives לעיניך before thine eyes, ונתתי and give לרעיך unto thy neighbor, ושׁכב and he shall lie עם with נשׁיך thy wives לעיני in the sight השׁמשׁ sun. הזאת׃ of this}%
\verse{כי For אתה thou עשׂית didst בסתר secretly: ואני but I אעשׂה will do את הדבר thing הזה this נגד before כל all ישׂראל Israel, ונגד and before השׁמשׁ׃ the sun.}%
\verseWithHeading{David Repents, But the Child Dies}{ויאמר said דוד And David אל unto נתן Nathan, חטאתי I have sinned ליהוה against the LORD. ויאמר said נתן And Nathan אל unto דוד David, גם also יהוה The LORD העביר hath put away חטאתך thy sin; לא thou shalt not תמות׃ die.}%
\verse{אפס Howbeit, כי because נאץ thou hast given great occasion נאצת to blaspheme, את איבי to the enemies יהוה of the LORD בדבר deed הזה by this גם also הבן the child הילוד born לך מות unto thee shall surely die. ימות׃ unto thee shall surely die.}%
\verse{וילך departed נתן And Nathan אל unto ביתו his house. ויגף struck יהוה And the LORD את הילד the child אשׁר that ילדה bore אשׁת wife אוריה Uriah's לדוד unto David, ויאנשׁ׃ and it was very sick.}%
\verse{ויבקשׁ therefore besought דוד David את האלהים God בעד for הנער the child; ויצם fasted, דוד and David צום fasted, ובא and went ולן night ושׁכב in, and lay ארצה׃ upon the earth.}%
\verse{ויקמו arose, זקני And the elders ביתו of his house עליו to להקימו him, to raise him up מן from הארץ the earth: ולא not, אבה but he would ולא neither ברא אתם with לחם׃ bread}%
\verse{ויהי And it came to pass ביום day, השׁביעי on the seventh וימת died. הילד that the child ויראו feared עבדי And the servants דוד of David להגיד to tell לו כי him that מת was dead: הילד the child כי for אמרו they said, הנה Behold, בהיות was הילד while the child חי yet alive, דברנו we spoke אליו unto ולא him, and he would not שׁמע hearken בקולנו unto our voice: ואיך how נאמר himself, if we tell אליו himself, if we tell מת is dead? הילד him that the child ועשׂה will he then vex רעה׃ will he then vex}%
\verse{וירא saw דוד But when David כי that עבדיו his servants מתלחשׁים whispered, ויבן perceived דוד David כי that מת was dead: הילד the child ויאמר said דוד therefore David אל unto עבדיו his servants, המת dead? הילד Is the child ויאמרו And they said, מת׃ He is dead.}%
\verse{ויקם arose דוד Then David מהארץ from the earth, וירחץ and washed, ויסך and anointed ויחלף and changed שׂמלתו his apparel, ויבא and came בית into the house יהוה of the LORD, וישׁתחו and worshiped: ויבא then he came אל to ביתו his own house; וישׁאל and when he required, וישׂימו they set לו לחם bread ויאכל׃ before him, and he did eat.}%
\verse{ויאמרו Then said עבדיו his servants אליו unto מה him, What הדבר thing הזה this אשׁר that עשׂיתה thou hast done? בעבור while הילד for the child, חי alive; צמת thou didst fast ותבך and weep וכאשׁר but when מת was dead, הילד the child קמת thou didst rise ותאכל and eat לחם׃ bread.}%
\verse{ויאמר And he said, בעוד was yet הילד While the child חי alive, צמתי I fasted ואבכה and wept: כי for אמרתי I said, מי Who יודע can tell יחנני will be gracious יהוה GOD וחי may live? הילד׃ to me, that the child}%
\verse{ועתה But now מת he is dead, למה wherefore זה wherefore אני should I צם fast? האוכל can להשׁיבו I bring him back עוד again? אני I הלך shall go אליו to והוא him, but he לא shall not ישׁוב return אלי׃ to}%
\verse{וינחם comforted דוד And David את בת־שׁבע Bath-sheba אשׁתו his wife, ויבא and went אליה in unto וישׁכב her, and lay עמה with ותלד her: and she bore בן a son, ויקרא and he called את שׁמו his name שׁלמה Solomon: ויהוה and the LORD אהבו׃ loved}%
\verse{וישׁלח And he sent ביד by the hand נתן of Nathan הנביא the prophet; ויקרא and he called את שׁמו his name ידידיה Jedidiah, בעבור because יהוה׃ of the LORD.}%
\verseWithHeading{Battle with the Ammonites}{וילחם fought יואב And Joab ברבת against Rabbah בני of the children עמון of Ammon, וילכד and took את עיר city. המלוכה׃ the royal}%
\verse{וישׁלח sent יואב And Joab מלאכים messengers אל to דוד David, ויאמר and said, נלחמתי I have fought ברבה against Rabbah, גם and לכדתי have taken את עיר the city המים׃ of waters.}%
\verse{ועתה Now אסף therefore gather את יתר the rest העם of the people וחנה together, and encamp על against העיר the city, ולכדה and take פן it: lest אלכד take אני I את העיר the city, ונקרא and it be called שׁמי my name. עליה׃ after}%
\verse{ויאסף gathered דוד And David את כל all העם the people וילך together, and went רבתה to Rabbah, וילחם and fought בה וילכדה׃ against it, and took}%
\verse{ויקח And he took את עטרת crown מלכם their king's מעל from off ראשׁו his head, ומשׁקלה the weight ככר whereof a talent זהב of gold ואבן stones: יקרה with the precious ותהי and it was על on ראשׁ head. דוד David's ושׁלל the spoil העיר of the city הוציא And he brought forth הרבה abundance. מאד׃ in great}%
\verse{ואת העם the people אשׁר that בה הוציא And he brought forth וישׂם therein, and put במגרה under saws, ובחרצי and under harrows הברזל of iron, ובמגזרת and under axes הברזל of iron, והעביר אותםמלכן the brickkiln: וכן and thus יעשׂה did לכל he unto all ערי the cities בני of the children עמון of Ammon. וישׁב returned דוד So David וכל and all העם the people ירושׁלם׃ unto Jerusalem.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 13
\verseWithHeading{Amnon Assaults His Sister Tamar}{ויהי And it came to pass אחרי after כן this, ולאבשׁלום that Absalom בן the son דוד of David אחות sister, יפה had a fair ושׁמה whose name תמר Tamar; ויאהבה loved אמנון and Amnon בן the son דוד׃ of David}%
\verse{ויצר לאמנון And Amnon להתחלות that he fell sick בעבור for תמר Tamar; אחתו his sister כי for בתולה a virgin; היא she ויפלא it hard בעיני thought אמנון and Amnon לעשׂות for him to do לה מאומה׃ any thing}%
\verse{ולאמנון But Amnon רע had a friend, ושׁמו whose name יונדב Jonadab, בן the son שׁמעה of Shimeah אחי brother: דוד David's ויונדב and Jonadab אישׁ man. חכם subtle מאד׃ a very}%
\verse{ויאמר And he said לו מדוע unto him, Why אתה thou, ככה lean דל lean בן son, המלך the king's בבקר from day בבקר to day? הלוא wilt thou not תגיד tell לי ויאמר said לו אמנון me? And Amnon את תמר Tamar, אחות sister. אבשׁלם Absalom's אחי my brother אני unto him, I אהב׃ love}%
\verse{ויאמר said לו יהונדב שׁכב unto him, Lay thee down על on משׁכבך thy bed, והתחל and make thyself sick: ובא cometh אביך and when thy father לראותך to see ואמרת thee, say אליו unto תבא come, נא him, I pray thee, תמר Tamar אחותי let my sister ותברני and give לחם me meat, ועשׂתה and dress לעיני in my sight, את הבריה the meat למען that אשׁר that אראה I may see ואכלתי and eat מידה׃ at her hand.}%
\verse{וישׁכב lay down, אמנון So Amnon ויתחל and made himself sick: ויבא was come המלך and when the king לראתו to see ויאמר said אמנון him, Amnon אל unto המלך the king, תבוא come, נא I pray thee, תמר let Tamar אחתי my sister ותלבב and make לעיני in my sight, שׁתי me a couple לבבות of cakes ואברה that I may eat מידה׃ at her hand.}%
\verse{וישׁלח sent דוד Then David אל to תמר Tamar, הביתה home לאמר saying, לכי Go נא now בית house, אמנון Amnon's אחיך to thy brother ועשׂי and dress לו הבריה׃ him meat.}%
\verse{ותלך went תמר So Tamar בית house; אמנון Amnon's אחיה to her brother והוא and he שׁכב was laid down. ותקח And she took את הבצק flour, ותלושׁ and kneaded ותלבב and made cakes לעיניו in his sight, ותבשׁל and did bake את הלבבות׃ the cakes.}%
\verse{ותקח And she took את המשׂרת a pan, ותצק and poured out לפניו before וימאן him; but he refused לאכול to eat. ויאמר said, אמנון And Amnon הוציאו Have out כל all אישׁ men מעלי from ויצאו me. And they went out כל every אישׁ man מעליו׃ from}%
\verse{ויאמר said אמנון And Amnon אל unto תמר Tamar, הביאי Bring הבריה the meat החדר into the chamber, ואברה that I may eat מידך of thine hand. ותקח took תמר And Tamar את הלבבות the cakes אשׁר which עשׂתה she had made, ותבא and brought לאמנון to Amnon אחיה her brother. החדרה׃ into the chamber}%
\verse{ותגשׁ And when she had brought אליו unto לאכל him to eat, ויחזק he took hold בה ויאמר of her, and said לה בואי unto her, Come שׁכבי lie עמי with אחותי׃ me, my sister.}%
\verse{ותאמר And she answered לו אל him, Nay, אחי my brother, אל do not תענני force כי me; for לא no יעשׂה ought to be done כן such thing בישׂראל in Israel: אל not תעשׂה do את הנבלה folly. הזאת׃ thou this}%
\verse{ואני And I, אנה whither אוליך to go? את חרפתי shall I cause my shame ואתה and as for thee, תהיה thou shalt be כאחד as one הנבלים of the fools בישׂראל in Israel. ועתה Now דבר speak נא therefore, I pray thee, אל unto המלך the king; כי for לא he will not ימנעני withhold ממך׃ me from}%
\verse{ולא not אבה Howbeit he would לשׁמע hearken בקולה unto her voice: ויחזק but, being stronger ממנה ויענה than she, forced וישׁכב her, and lay with אתה׃}%
\verse{וישׂנאה hated אמנון Then Amnon שׂנאה her exceedingly; גדולה her greater מאד כי so that גדולה השׂנאה the hatred אשׁר wherewith שׂנאה he hated מאהבה than the love אשׁר wherewith אהבה he had loved ויאמר said לה אמנון her. And Amnon קומי unto her, Arise, לכי׃ be gone.}%
\verse{ותאמר And she said לו אל unto him, no אודת cause: הרעה evil הגדולה greater הזאת this מאחרת than the other אשׁר that עשׂית thou didst עמי unto לשׁלחני in sending me away ולא not אבה me. But he would לשׁמע׃ hearken}%
\verse{ויקרא Then he called את נערו his servant משׁרתו that ministered ויאמר unto him, and said, שׁלחו Put נא now את זאת this מעלי החוצהנעל and bolt הדלת the door אחריה׃ after}%
\verse{ועליה upon כתנת And a garment פסים of divers colors כי her: for כן with such תלבשׁן appareled. בנות daughters המלך were the king's הבתולת virgins מעילים robes ויצא brought אותה משׁרתו Then his servant החוץ her out, ונעל and bolted הדלת the door אחריה׃ after}%
\verse{ותקח put תמר And Tamar אפר ashes על on ראשׁה her head, וכתנת her garment הפסים of divers colors אשׁר that עליה on קרעה and rent ותשׂם her, and laid ידה her hand על on ראשׁה her head, ותלך and went on הלוך and went on וזעקה׃ crying.}%
\verse{ויאמר said אליה unto אבשׁלום And Absalom אחיה her brother האמינון her, Hath Amnon אחיך thy brother היה been עמך with ועתה now אחותי thy peace, my sister: החרישׁי thee? but hold אחיך thy brother; הוא he אל not תשׁיתי אתבך לדבר thing. הזה this ותשׁב remained תמר So Tamar ושׁממה בית house. אבשׁלום Absalom's אחיה׃ in her brother}%
\verseWithHeading{Absalom Kills Amnon}{והמלך But when king דוד David שׁמע heard את כל of all הדברים things, האלה these ויחר wroth. לו מאד׃ he was very}%
\verse{ולא neither דבר spoke אבשׁלום And Absalom עם unto אמנון his brother Amnon למרע bad: ועד nor טוב good כי for שׂנא hated אבשׁלום Absalom את אמנון Amnon, על because דבר אשׁרנה he had forced את תמר Tamar. אחתו׃ his sister}%
\verse{ויהי And it came to pass לשׁנתים after two full years, ימים after two full years, ויהיו had גזזים sheepshearers לאבשׁלום that Absalom בבעל חצור in Baal-hazor, אשׁר which עם beside אפרים Ephraim: ויקרא invited אבשׁלום and Absalom לכל all בני sons. המלך׃ the king's}%
\verse{ויבא came אבשׁלום And Absalom אל to המלך the king, ויאמר and said, הנה Behold נא now, גזזים hath sheepshearers; לעבדך thy servant ילך go נא I beseech thee, המלך let the king, ועבדיו and his servants עם with עבדך׃ thy servant.}%
\verse{ויאמר said המלך And the king אל to אבשׁלום Absalom, אל Nay, בני my son, אל let us not נא now נלך go, כלנו all ולא lest נכבד we be chargeable עליך unto ויפרץ thee. And he pressed בו ולא not אבה him: howbeit he would ללכת go, ויברכהו׃ but blessed}%
\verse{ויאמר Then said אבשׁלום Absalom, ולא If not, ילך go נא I pray thee, אתנו with אמנון Amnon אחי let my brother ויאמר said לו המלך us. And the king למה unto him, Why ילך should he go עמך׃ with}%
\verse{ויפרץ pressed בו אבשׁלום But Absalom וישׁלח go אתו with את אמנון him, that he let Amnon ואת כל and all בני sons המלך׃ the king's}%
\verse{ויצו had commanded אבשׁלום Now Absalom את נעריו his servants, לאמר saying, ראו Mark נא ye now כטוב לב heart אמנון when Amnon's ביין with wine, ואמרתי and when I say אליכם unto הכו you, Smite את אמנון Amnon; והמתם then kill אתו אל not: תיראו him, fear הלוא have not כי אנכי I צויתי commanded אתכם חזקו you? be courageous, והיו and be לבני valiant. חיל׃ valiant.}%
\verse{ויעשׂו did נערי And the servants אבשׁלום of Absalom לאמנון unto Amnon כאשׁר as צוה had commanded. אבשׁלום Absalom ויקמו arose, כל Then all בני sons המלך the king's וירכבו got him up אישׁ and every man על upon פרדו his mule, וינסו׃ and fled.}%
\verse{ויהי And it came to pass, המה while they בדרך were in the way, והשׁמעה that tidings באה came אל to דוד David, לאמר saying, הכה hath slain אבשׁלום Absalom את כל all בני sons, המלך the king's ולא and there is not נותר them left. מהם אחד׃ one}%
\verse{ויקם arose, המלך Then the king ויקרע and tore את בגדיו his garments, וישׁכב and lay ארצה on the earth; וכל and all עבדיו his servants נצבים stood by קרעי rent. בגדים׃ with their clothes}%
\verse{ויען answered יונדב And Jonadab, בן the son שׁמעה of Shimeah אחי brother, דוד David's ויאמר and said, אל Let not יאמר suppose אדני my lord את כל all הנערים the young men בני sons; המלך the king's המיתו they have slain כי for אמנון Amnon לבדו only מת is dead: כי for על by פי the appointment אבשׁלום of Absalom היתה this hath been שׂומה determined מיום from the day ענתו that he forced את תמר Tamar. אחתו׃ his sister}%
\verse{ועתה Now אל therefore let not ישׂם take אדני my lord המלך the king אל to לבו his heart, דבר the thing לאמר to think כל that all בני sons המלך the king's מתו are dead: כי for אם for אמנון Amnon לבדו only מת׃ is dead.}%
\verseWithHeading{Absalom Flees}{ויברח fled. אבשׁלום But Absalom וישׂא lifted up הנער And the young man הצפה that kept the watch את עינו his eyes, וירא and looked, והנה and, behold, עם people רב much הלכים there came מדרך by the way אחריו behind מצד ההר׃}%
\verse{ויאמר said יונדב And Jonadab אל unto המלך the king, הנה Behold, בני sons המלך the king's באו come: כדבר said, עבדך as thy servant כן so היה׃ it is.}%
\verse{ויהי And it came to pass, ככלתו as soon as he had made an end לדבר of speaking, והנה that, behold, בני sons המלך the king's באו came, וישׂאו and lifted up קולם their voice ויבכו and wept: וגם also המלך and the king וכל and all עבדיו his servants בכו wept בכי גדול sore. מאד׃ very}%
\verse{ואבשׁלום But Absalom ברח fled, וילך and went אל to תלמי Talmai, בן the son עמיחור מלך king גשׁור of Geshur. ויתאבל And mourned על for בנו his son כל every הימים׃ day.}%
\verse{ואבשׁלום So Absalom ברח fled, וילך and went גשׁור to Geshur, ויהי and was שׁם there שׁלשׁ three שׁנים׃ years.}%
\verse{ותכל longed דוד David המלך And king לצאת to go forth אל unto אבשׁלום Absalom: כי for נחם he was comforted על concerning אמנון Amnon, כי seeing מת׃ he was dead.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 14
\verseWithHeading{Joab Plots to Reconcile David with Absalom}{וידע perceived יואב Now Joab בן the son צריה of Zeruiah כי that לב heart המלך the king's על toward אבשׁלום׃ Absalom.}%
\verse{וישׁלח sent יואב And Joab תקועה to Tekoah, ויקח and fetched משׁם thence אשׁה woman, חכמה a wise ויאמר and said אליה unto התאבלי feign thyself to be a mourner, נא her, I pray thee, ולבשׁי and put on נא now בגדי apparel, אבל mourning ואל not תסוכי and anoint שׁמן thyself with oil, והיית but be כאשׁה as a woman זה that ימים time רבים had a long מתאבלת mourned על for מת׃ the dead:}%
\verse{ובאת And come אל to המלך the king, ודברת and speak אליו on כדבר manner הזה this וישׂם put יואב him. So Joab את הדברים the words בפיה׃ in her mouth.}%
\verse{ותאמר spoke האשׁה And when the woman התקעית of Tekoah אל to המלך the king, ותפל she fell על on אפיה her face ארצה to the ground, ותשׁתחו and did obeisance, ותאמר and said, הושׁעה Help, המלך׃ O king.}%
\verse{ויאמר said לה המלך And the king מה unto her, What לך ותאמר aileth thee? And she answered, אבל indeed אשׁה woman, אלמנה a widow אני I וימת is dead. אישׁי׃ and mine husband}%
\verse{ולשׁפחתך And thy handmaid שׁני had two בנים sons, וינצו strove together שׁניהם and they two בשׂדה in the field, ואין and none מציל to part ביניהם to part ויכו smote האחד them, but the one את האחד the other, וימת and slew אתו׃}%
\verse{והנה And, behold, קמה is risen כל the whole המשׁפחה family על against שׁפחתך thine handmaid, ויאמרו and they said, תני Deliver את מכה him that smote אחיו his brother, ונמתהו that we may kill בנפשׁ him, for the life אחיו of his brother אשׁר whom הרג he slew; ונשׁמידה and we will destroy גם also: את היורשׁ the heir וכבו and so they shall quench את גחלתי my coal אשׁר which נשׁארה is left, לבלתי and shall not שׂום לאישׁיׁם name ושׁארית nor remainder על upon פני upon האדמה׃ the earth.}%
\verse{ויאמר said המלך And the king אל unto האשׁה the woman, לכי Go לביתך to thine house, ואני and I אצוה will give charge עליך׃ concerning}%
\verse{ותאמר said האשׁה And the woman התקועית of Tekoah אל unto המלך the king, עלי on אדני My lord, המלך O king, העון the iniquity ועל me, and on בית house: אבי my father's והמלך and the king וכסאו and his throne נקי׃ guiltless.}%
\verse{ויאמר said, המלך And the king המדבר Whosoever saith אליך unto והבאתו thee, bring אלי him to ולא me, and he shall not יסיף thee any עוד more. לגעת׃ touch}%
\verse{ותאמר Then said יזכר remember נא she, I pray thee, המלך let the king את יהוה the LORD אלהיך thy God, מהרבית any more, גאל suffer the revengers הדם of blood לשׁחת to destroy ולא that thou wouldest not ישׁמידו lest they destroy את בני my son. ויאמר And he said, חי liveth, יהוה the LORD אם there shall not יפל fall משׂערת one hair בנך of thy son ארצה׃ to the earth.}%
\verse{ותאמר said, האשׁה Then the woman תדבר speak נא I pray thee, שׁפחתך Let thine handmaid, אל unto אדני my lord המלך the king. דבר word ויאמר And he said, דברי׃ Say on.}%
\verse{ותאמר said, האשׁה And the woman ולמה Wherefore חשׁבתה then hast thou thought כזאת such על a thing against עם the people אלהים of God? ומדבר doth speak המלך for the king הדבר thing הזה this כאשׁם as one which is faulty, לבלתי doth not השׁיב fetch home again המלך in that the king את נדחו׃ his banished.}%
\verse{כי For מות we must needs die, נמות we must needs die, וכמים and as water הנגרים spilt ארצה on the ground, אשׁר which לא cannot יאספו be gathered up ולא again; neither ישׂא respect אלהים doth God נפשׁ person: וחשׁב yet doth he devise מחשׁבות means, לבלתי ידח that his banished ממנו from נדח׃ be not expelled}%
\verse{ועתה Now אשׁר therefore that באתי I am come לדבר to speak אל unto המלך the king, אדני my lord את הדבר thing הזה of this כי because יראני have made me afraid: העם the people ותאמר said, שׁפחתך and thy handmaid אדברה speak נא I will now אל unto המלך the king; אולי it may be יעשׂה will perform המלך that the king את דבר the request אמתו׃ of his handmaid.}%
\verse{כי For ישׁמע will hear, המלך the king להציל to deliver את אמתו his handmaid מכף out of the hand האישׁ of the man להשׁמיד destroy אתי ואתני me and my son יחד together מנחלת out of the inheritance אלהים׃ of God.}%
\verse{ותאמר said, שׁפחתך Then thine handmaid יהיה be נא shall now דבר The word אדני of my lord המלך the king למנוחה comfortable: כי for כמלאך as an angel האלהים of God, כן so אדני my lord המלך the king לשׁמע to discern הטוב good והרע and bad: ויהוה therefore the LORD אלהיך thy God יהי will be עמך׃ with}%
\verse{ויען answered המלך Then the king ויאמר and said אל unto האשׁה the woman, אל not נא me, I pray thee, תכחדי Hide ממני from דבר the thing אשׁר that אנכי I שׁאל shall ask אתך ותאמר said, האשׁה thee. And the woman ידבר speak. נא now אדני Let my lord המלך׃ the king}%
\verse{ויאמר said, המלך And the king היד the hand יואב of Joab אתך with בכל thee in all זאת this? ותען answered האשׁה And the woman ותאמר and said, חי liveth, נפשׁך thy soul אדני my lord המלך the king, אם none can turn אשׁ none can turn להמין to the right ולהשׂמיל hand or to the left מכל from aught אשׁר that דבר hath spoken: אדני my lord המלך the king כי for עבדך thy servant יואב Joab, הוא he צוני bade והוא me, and he שׂם put בפי in the mouth שׁפחתך of thine handmaid: את כל all הדברים words האלה׃ these}%
\verse{לבעבור To סבב fetch about את פני this form הדבר of speech עשׂה done עבדך hath thy servant יואב Joab את הדבר thing: הזה this ואדני and my lord חכם wise, כחכמת according to the wisdom מלאך of an angel האלהים of God, לדעת to know את כל all אשׁר that בארץ׃ in the earth.}%
\verse{ויאמר said המלך And the king אל unto יואב Joab, הנה Behold נא now, עשׂיתי I have done את הדבר thing: הזה this ולך go השׁב therefore, bring את הנער the young man את אבשׁלום׃ Absalom}%
\verse{ויפל fell יואב And Joab אל on פניו his face, ארצה to the ground וישׁתחו and bowed himself, ויברך and thanked את המלך the king: ויאמר said, יואב and Joab היום Today ידע knoweth עבדך thy servant כי that מצאתי I have found חן grace בעיניך in thy sight, אדני my lord, המלך O king, אשׁר in that עשׂה hath fulfilled המלך the king את דבר the request עבדו׃ of his servant.}%
\verseWithHeading{Absalom Returns to Jerusalem}{ויקם arose יואב So Joab וילך and went גשׁורה to Geshur, ויבא and brought את אבשׁלום Absalom ירושׁלם׃ to Jerusalem.}%
\verse{ויאמר said, המלך And the king יסב Let him turn אל to ביתו his own house, ופני my face. לא and let him not יראה see ויסב returned אבשׁלום So Absalom אל to ביתו his own house, ופני face. המלך the king's לא not ראה׃ and saw}%
\verse{וכאבשׁלום as Absalom לא none היה there was אישׁ none יפה for his beauty: בכל But in all ישׂראל Israel להלל praised מאד to be so much מכף from the sole רגלו of his foot ועד even to קדקדו the crown of his head לא no היה there was בו מום׃ blemish}%
\verse{ובגלחו And when he polled את ראשׁו his head, והיה (for it was מקץ ימיםימים אשׁר that יגלח he polled כי because כבד was heavy עליו on וגלחו him, therefore he polled ושׁקל it:) he weighed את שׂער the hair ראשׁו of his head מאתים at two hundred שׁקלים shekels באבן weight. המלך׃ after the king's}%
\verse{ויולדו there were born לאבשׁלום And unto Absalom שׁלושׁה three בנים sons, ובת daughter, אחת and one ושׁמה whose name תמר Tamar: היא she היתה was אשׁה a woman יפת of a fair מראה׃ countenance.}%
\verse{וישׁב dwelt אבשׁלום So Absalom בירושׁלם in Jerusalem, שׁנתים two full years ימים two full years ופני face. המלך the king's לא not ראה׃ and saw}%
\verse{וישׁלח sent אבשׁלום Therefore Absalom אל for יואב Joab, לשׁלח to have sent אתו אל him to המלך the king; ולא not אבה but he would לבוא come אליו to וישׁלח him: and when he sent עוד again שׁנית the second time, ולא not אבה he would לבוא׃ come.}%
\verse{ויאמר Therefore he said אל unto עבדיו his servants, ראו See, חלקת field יואב Joab's אל is near ידי mine, ולו שׁם there; שׂערים and he hath barley לכו go והוצתיה and set באשׁ it on fire. ויצתו set עבדי servants אבשׁלום And Absalom's את החלקה the field באשׁ׃ on fire.}%
\verse{ויקם arose, יואב Then Joab ויבא and came אל to אבשׁלום Absalom הביתה unto house, ויאמר and said אליו unto למה him, Wherefore הציתו set עבדך have thy servants את החלקה my field אשׁר לי באשׁ׃ on fire?}%
\verse{ויאמר answered אבשׁלום And Absalom אל answered יואב Joab, הנה Behold, שׁלחתי I sent אליך unto לאמר thee, saying, בא Come הנה hither, ואשׁלחה that I may send אתך אל thee to המלך the king, לאמר to say, למה Wherefore באתי come מגשׁור טוב good לי עד still: אני am I שׁם for me there ועתה now אראה therefore let me see פני face; המלך the king's ואם and if ישׁ there be בי עון iniquity והמתני׃ in me, let him kill}%
\verse{ויבא came יואב So Joab אל to המלך the king, ויגד and told לו ויקרא him: and when he had called אל for אבשׁלום Absalom, ויבא he came אל to המלך the king, וישׁתחו and bowed לו על himself on אפיו his face ארצה to the ground לפני before המלך the king: וישׁק kissed המלך and the king לאבשׁלום׃ Absalom.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 15
\verseWithHeading{Absalom Leads a Rebellion}{ויהי And it came to pass מאחרי after כן this, ויעשׂ prepared לו אבשׁלום that Absalom מרכבה him chariots וססים and horses, וחמשׁים and fifty אישׁ men רצים to run לפניו׃ before}%
\verse{והשׁכים rose up early, אבשׁלום And Absalom ועמד and stood על beside יד beside דרך the way השׁער of the gate: ויהי and it was כל that when any האישׁ man אשׁר that יהיה had לו ריב a controversy לבוא came אל to המלך the king למשׁפט for judgment, ויקרא called אבשׁלום then Absalom אליו unto ויאמר him, and said, אי what מזה עיר city אתה thou? ויאמר And he said, מאחד of one שׁבטי of the tribes ישׂראל of Israel. עבדך׃ Thy servant}%
\verse{ויאמר said אליו unto אבשׁלום And Absalom ראה him, See, דברך thy matters טובים good ונכחים and right; ושׁמע to hear אין but no לך מאת המלך׃ man of the king}%
\verse{ויאמר said אבשׁלום Absalom מי moreover, Oh that ישׂמני I were made שׁפט judge בארץ in the land, ועלי unto יבוא might come כל that every אישׁ man אשׁר which יהיה hath לו ריב any suit ומשׁפט or cause והצדקתיו׃ me, and I would do him justice!}%
\verse{והיה And it was בקרב came nigh אישׁ that when any man להשׁתחות to do him obeisance, לו ושׁלח he put forth את ידו his hand, והחזיק and took לו ונשׁק׃ him, and kissed}%
\verse{ויעשׂ did אבשׁלום Absalom כדבר manner הזה And on this לכל to all ישׂראל Israel אשׁר that יבאו came למשׁפט for judgment: אל to המלך the king ויגנב stole אבשׁלום so Absalom את לב the hearts אנשׁי of the men ישׂראל׃ of Israel.}%
\verse{ויהי And it came to pass מקץ after ארבעים forty שׁנה years, ויאמר said אבשׁלום that Absalom אל unto המלך the king, אלכה let me go נא I pray thee, ואשׁלם and pay את נדרי my vow, אשׁר which נדרתי I have vowed ליהוה unto the LORD, בחברון׃ in Hebron.}%
\verse{כי For נדר a vow נדר vowed עבדך thy servant בשׁבתי while I abode בגשׁור at Geshur בארם in Syria, לאמר saying, אם If ישׁיב shall bring me again ישׁיבני יהוה the LORD ירושׁלם indeed to Jerusalem, ועבדתי then I will serve את יהוה׃ the LORD.}%
\verse{ויאמר said לו המלך And the king לך unto him, Go בשׁלום in peace. ויקם So he arose, וילך and went חברונה׃ to Hebron.}%
\verse{וישׁלח sent אבשׁלום But Absalom מרגלים spies בכל throughout all שׁבטי the tribes ישׂראל of Israel, לאמר saying, כשׁמעכם As soon as ye hear את קול the sound השׁפר of the trumpet, ואמרתם then ye shall say, מלך reigneth אבשׁלום Absalom בחברון׃ in Hebron.}%
\verse{ואת And with אבשׁלום Absalom הלכו went מאתים two hundred אישׁ men מירושׁלם קראים called; והלכים and they went לתמם in their simplicity, ולא not ידעו and they knew כל any דבר׃ thing.}%
\verse{וישׁלח sent אבשׁלום And Absalom את אחיתפל for Ahithophel הגילני the Gilonite, יועץ counselor, דוד David's מעירו from his city, מגלה בזבחו while he offered את הזבחים sacrifices. ויהי was הקשׁר And the conspiracy אמץ strong; והעם for the people הולך continually ורב increased את with אבשׁלום׃ Absalom.}%
\verseWithHeading{David Flees from Jerusalem}{ויבא And there came המגיד a messenger אל to דוד David, לאמר saying, היה are לב The hearts אישׁ of the men ישׂראל of Israel אחרי after אבשׁלום׃ Absalom.}%
\verse{ויאמר said דוד And David לכל unto all עבדיו his servants אשׁר that אתו with בירושׁלם him at Jerusalem, קומו Arise, ונברחה כי for לא not תהיה we shall לנו פליטה escape מפני אבשׁלום Absalom: מהרו make speed ללכת to depart, פן lest ימהר us suddenly, והשׂגנו he overtake והדיח and bring עלינו upon את הרעה evil והכה us, and smite העיר the city לפי with the edge חרב׃ of the sword.}%
\verse{ויאמרו said עבדי servants המלך And the king's אל unto המלך the king, ככל whatsoever אשׁר whatsoever יבחר shall appoint. אדני my lord המלך the king הנה Behold, עבדיך׃ thy servants}%
\verse{ויצא went forth, המלך And the king וכל and all ביתו his household ברגליו after ויעזב left המלך him. And the king את עשׂר ten נשׁים women, פלגשׁים concubines, לשׁמר to keep הבית׃ the house.}%
\verse{ויצא went forth, המלך And the king וכל and all העם the people ברגליו after ויעמדו him, and tarried בית המרחק׃}%
\verse{וכל And all עבדיו his servants עברים passed on על beside ידו beside וכל him; and all הכרתי the Cherethites, וכל and all הפלתי the Pelethites, וכל and all הגתים the Gittites, שׁשׁ six מאות hundred אישׁ men אשׁר which באו came ברגלו after מגת עברים passed on על before פני before המלך׃ the king.}%
\verse{ויאמר Then said המלך the king אל to אתי Ittai הגתי the Gittite, למה Wherefore תלך goest גם also אתה thou אתנו with שׁוב us? return ושׁב and abide עם with המלך the king: כי for נכרי a stranger, אתה thou וגם and also גלה an exile. אתה למקומך׃ to thy place,}%
\verse{תמול yesterday, בואך Whereas thou camest והיום should I this day אנועך and down עמנו with ללכת make thee go up ואני us? seeing I הולך go על whither אשׁר whither אני I הולך may, שׁוב return והשׁב thou, and take back את אחיך thy brethren: עמך with חסד mercy ואמת׃ and truth}%
\verse{ויען answered אתי And Ittai את המלך the king, ויאמר and said, חי liveth, יהוה the LORD וחי liveth, אדני and my lord המלך the king כי surely אם surely במקום place אשׁר in what יהיה shall be, שׁם there אדני my lord המלך the king אם whether למות in death אם or לחיים life, כי even שׁם יהיה be. עבדך׃ also will thy servant}%
\verse{ויאמר said דוד And David אל to אתי Ittai, לך Go ועבר and pass over. ויעבר passed over, אתי And Ittai הגתי the Gittite וכל and all אנשׁיו his men, וכל and all הטף the little ones אשׁר that אתו׃ with}%
\verse{וכל And all הארץ the country בוכים wept קול voice, גדול with a loud וכל and all העם the people עברים passed over: והמלך the king עבר also himself passed over בנחל the brook קדרון Kidron, וכל and all העם the people עברים passed over, על toward פני toward דרך the way את המדבר׃ of the wilderness.}%
\verseWithHeading{The Priests Offer Sacrifices for David}{והנה And lo גם also, צדוק Zadok וכל and all הלוים the Levites אתו with נשׂאים him, bearing את ארון the ark ברית of the covenant האלהים of God: ויצקו and they set down את ארון the ark האלהים of God; ויעל went up, אביתר and Abiathar עד until תם had done כל all העם the people לעבור passing מן out of העיר׃ the city.}%
\verse{ויאמר said המלך And the king לצדוק unto Zadok, השׁב Carry back את ארון the ark האלהים of God העיר into the city: אם if אמצא I shall find חן favor בעיני in the eyes יהוה of the LORD, והשׁבני he will bring me again, והראני and show אתו ואתוהו׃ me it, and his habitation:}%
\verse{ואם But if כה he thus יאמר say, לא I have no חפצתי delight בך הנני יעשׂה I, let him do לי כאשׁר to me as טוב good בעיניו׃ seemeth}%
\verse{ויאמר said המלך The king אל also unto צדוק Zadok הכהן the priest, הרואה a seer? אתה thou שׁבה return העיר into the city בשׁלום in peace, ואחימעץ you, Ahimaaz בנך sons ויהונתן and Jonathan בן thy son, אביתר of Abiathar. שׁני and your two בניכם the son אתכם׃ with}%
\verse{ראו See, אנכי I מתמהמה will tarry בעברות המדבר of the wilderness, עד until בוא there come דבר word מעמכם from להגיד׃ you to certify}%
\verse{וישׁב carried צדוק Zadok ואביתר therefore and Abiathar את ארון the ark האלהים of God ירושׁלם again to Jerusalem: וישׁבו and they tarried שׁם׃ there.}%
\verseWithHeading{Hushai Offers to Serve King David}{ודוד And David עלה went up במעלה by the ascent הזיתים of Olivet, עלה as he went up, ובוכה and wept וראשׁ and had his head לו חפוי covered, והוא and he הלך went יחף barefoot: וכל and all העם the people אשׁר that אתו with חפו him covered אישׁ every man ראשׁו his head, ועלו and they went up, עלה as they went up. ובכה׃ weeping}%
\verse{ודוד David, הגיד And told לאמר saying, אחיתפל Ahithophel בקשׁרים among the conspirators עם with אבשׁלום Absalom. ויאמר said, דוד And David סכל into foolishness. נא I pray thee, את עצת turn the counsel אחיתפל of Ahithophel יהוה׃ O LORD,}%
\verse{ויהי And it came to pass, דוד that David בא was come עד to הראשׁ the top אשׁר where ישׁתחוה he worshiped שׁם לאלהים God, והנה behold, לקראתו came to meet חושׁי Hushai הארכי the Archite קרוע rent, כתנתו him with his coat ואדמה and earth על upon ראשׁו׃ his head:}%
\verse{ויאמר said, לו דוד Unto whom David אם If עברת thou passest on אתי with והית me, then thou shalt be עלי unto למשׂא׃ a burden}%
\verse{ואם But if העיר to the city, תשׁוב thou return ואמרת and say לאבשׁלום unto Absalom, עבדך thy servant, אני I המלך O king; אהיה will be עבד servant אביך thy father's ואני I מאז hitherto, ועתה now ואני so I עבדך also thy servant: והפרתה then mayest thou for me defeat לי את עצת the counsel אחיתפל׃ of Ahithophel.}%
\verse{והלוא And not עמך with שׁם there צדוק thee Zadok ואביתר and Abiathar הכהנים the priests? והיה therefore it shall be, כל what הדבר thing אשׁר soever תשׁמע thou shalt hear מבית המלךגיד thou shalt tell לצדוק to Zadok ולאביתר and Abiathar הכהנים׃ the priests.}%
\verse{הנה Behold, שׁם there עמם with שׁני them their two בניהם sons, אחימעץ Ahimaaz לצדוק Zadok's ויהונתן and Jonathan לאביתר Abiathar's ושׁלחתם them ye shall send בידם and by אלי unto כל me every דבר thing אשׁר that תשׁמעו׃ ye can hear.}%
\verse{ויבא came חושׁי So Hushai רעה friend דוד David's העיר into the city, ואבשׁלם and Absalom יבא came ירושׁלם׃ into Jerusalem.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 16
\verseWithHeading{Ziba Brings Provisions}{ודוד And when David עבר past מעט was a little מהראשׁ the top והנה behold, ציבא Ziba נער the servant מפי בשׁת of Mephibosheth לקראתו וצמד him, with a couple חמרים of asses חבשׁים saddled, ועליהם and upon מאתים them two hundred לחם of bread, ומאה and a hundred צמוקים bunches of raisins, ומאה and a hundred קיץ of summer fruits, ונבל and a bottle יין׃ of wine.}%
\verse{ויאמר said המלך And the king אל unto ציבא Ziba, מה What אלה meanest thou by these? לך ויאמר said, ציבא And Ziba החמורים The asses לבית household המלך for the king's לרכב to ride on; ולהלחם and the bread והקיץ and summer fruit לאכול to eat; הנערים for the young men והיין and the wine, לשׁתות may drink. היעף that such as be faint במדבר׃ in the wilderness}%
\verse{ויאמר said, המלך And the king ואיה And where בן son? אדניך thy master's ויאמר said ציבא And Ziba אל unto המלך the king, הנה Behold, יושׁב he abideth בירושׁלם at Jerusalem: כי for אמר he said, היום Today ישׁיבו restore לי בית shall the house ישׂראל of Israel את ממלכות me the kingdom אבי׃ of my father.}%
\verse{ויאמר Then said המלך the king לצבא to Ziba, הנה Behold, לך כל thine all אשׁר that למפי בשׁת unto Mephibosheth. ויאמר said, ציבא And Ziba השׁתחויתי I humbly beseech אמצא thee I may find חן grace בעיניך in thy sight, אדני my lord, המלך׃ O king.}%
\verseWithHeading{Shimei Curses David}{ובא came המלך And when king דוד David עד to בחורים Bahurim, והנה behold, משׁם thence אישׁ a man יוצא came out ממשׁפחת of the family בית of the house שׁאול of Saul, ושׁמו whose name שׁמעי Shimei, בן the son גרא of Gera: יצא he came forth, יצוא still as he came. ומקלל׃ and cursed}%
\verse{ויסקל And he cast באבנים stones את דוד at David, ואת כל and at all עבדי the servants המלך of king דוד David: וכל and all העם the people וכל and all הגברים the mighty men מימינו on his right hand ומשׂמאלו׃ and on his left.}%
\verse{וכה And thus אמר said שׁמעי Shimei בקללו when he cursed, צא Come out, צא come out, אישׁ man, הדמים thou bloody ואישׁ and thou man הבליעל׃ of Belial:}%
\verse{השׁיב hath returned עליך upon יהוה The LORD כל thee all דמי the blood בית of the house שׁאול of Saul, אשׁר in whose מלכת thou hast reigned; תחתו stead ויתן hath delivered יהוה and the LORD את המלוכה the kingdom ביד into the hand אבשׁלום of Absalom בנך thy son: והנך ברעתך thou in thy mischief, כי because אישׁ man. דמים a bloody אתה׃ thou}%
\verse{ויאמר Then said אבישׁי Abishai בן the son צרויה of Zeruiah אל unto המלך the king, למה Why יקלל curse הכלב dog המת dead הזה should this את אדני my lord המלך the king? אעברה let me go over, נא I pray thee, ואסירה and take off את ראשׁו׃ his head.}%
\verse{ויאמר said, המלך And the king מה What לי ולכם בני have I to do with you, ye sons צריה of Zeruiah? כי because יקלל let him curse, וכי יהוה the LORD אמר hath said לו קלל unto him, Curse את דוד David. ומי Who יאמר shall then say, מדוע Wherefore עשׂיתה hast thou done כן׃ so?}%
\verse{ויאמר said דוד And David אל to אבישׁי Abishai, ואל and to כל all עבדיו his servants, הנה Behold, בני my son, אשׁר which יצא came forth ממעי of my bowels, מבקשׁ seeketh את נפשׁי my life: ואף how much more כי how much more עתה now בן הימיני Benjamite הנחו לו ויקלל and let him curse; כי for אמר hath bidden לו יהוה׃ the LORD}%
\verse{אולי It may be יראה will look יהוה that the LORD בעוני והשׁיב will requite יהוה and that the LORD לי טובה me good תחת for קללתו his cursing היום day. הזה׃ this}%
\verse{וילך went דוד And as David ואנשׁיו and his men בדרך by the way, ושׁמעי Shimei הלך went along בצלע side ההר on the hill's לעמתו over against הלוך as he went, ויקלל him, and cursed ויסקל and threw באבנים stones לעמתו at ועפר him, and cast בעפר׃ dust.}%
\verse{ויבא him, came המלך And the king, וכל and all העם the people אשׁר that אתו with עיפים weary, וינפשׁ and refreshed themselves שׁם׃ there.}%
\verseWithHeading{Hushai Comes to Absalom}{ואבשׁלום And Absalom, וכל and all העם the people אישׁ the men ישׂראל of Israel, באו came ירושׁלם to Jerusalem, ואחיתפל and Ahithophel אתו׃ with}%
\verse{ויהי And it came to pass, כאשׁר when בא was come חושׁי Hushai הארכי the Archite, רעה friend, דוד David's אל unto אבשׁלום Absalom, ויאמר said חושׁי that Hushai אל unto אבשׁלום Absalom, יחי God save המלך the king, יחי God save המלך׃ the king.}%
\verse{ויאמר said אבשׁלום And Absalom אל to חושׁי Hushai, זה this חסדך thy kindness את with רעך to thy friend? למה why לא thou not הלכת wentest את רעך׃ thy friend?}%
\verse{ויאמר said חושׁי And Hushai אל unto אבשׁלם Absalom, לא Nay; כי but אשׁר whom בחר choose, יהוה the LORD, והעם people, הזה and this וכל and all אישׁ the men ישׂראל of Israel, לא אהיה his will I be, ואתו and with אשׁב׃ him will I abide.}%
\verse{והשׁנית And again, למי whom אני should I אעבד serve? הלוא not לפני in the presence בנו of his son? כאשׁר as עבדתי I have served לפני presence, אביך in thy father's כן so אהיה will I be לפניך׃ in thy presence.}%
\verse{ויאמר Then said אבשׁלום Absalom אל to אחיתפל Ahithophel, הבו Give לכם עצה counsel מה among you what נעשׂה׃ we shall do.}%
\verse{ויאמר said אחיתפל And Ahithophel אל unto אבשׁלם Absalom, בוא Go אל in unto פלגשׁי concubines, אביך thy father's אשׁר which הניח לשׁמור to keep הבית the house; ושׁמע shall hear כל and all ישׂראל Israel כי that נבאשׁת thou art abhorred את of אביך thy father: וחזקו thee be strong. ידי then shall the hands כל of all אשׁר that אתך׃ with}%
\verse{ויטו So they spread לאבשׁלום Absalom האהל a tent על upon הגג the top ויבא went אבשׁלום of the house; and Absalom אל in unto פלגשׁי concubines אביו his father's לעיני in the sight כל of all ישׂראל׃ Israel.}%
\verse{ועצת And the counsel אחיתפל of Ahithophel, אשׁר which יעץ he counseled בימים days, ההם in those כאשׁר as ישׁאל had inquired בדבר at the oracle האלהים of God: כן so כל all עצת the counsel אחיתפל of Ahithophel גם both לדוד with David גם and לאבשׁלם׃ with Absalom.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 17
\verseWithHeading{Hushai Frustrates the Counsel of Ahithophel}{ויאמר said אחיתפל Moreover Ahithophel אל unto אבשׁלם Absalom, אבחרה choose out נא Let me now שׁנים twelve עשׂר twelve אלף thousand אישׁ men, ואקומה and I will arise וארדפה and pursue אחרי after דוד David הלילה׃ this night:}%
\verse{ואבוא And I will come עליו upon והוא him while he יגע weary ורפה and weak ידים handed, והחרדתי אתונס him shall flee; כל and all העם the people אשׁר that אתו with והכיתי and I will smite את המלך the king לבדו׃ only:}%
\verse{ואשׁיבה And I will bring back כל all העם the people אליך unto כשׁוב returned: הכל as if all האישׁ thee: the man אשׁר whom אתה thou מבקשׁ seekest כל all העם the people יהיה shall be שׁלום׃ in peace.}%
\verse{ויישׁר pleased הדבר And the saying בעיני well, אבשׁלם Absalom ובעיני כל and all זקני the elders ישׂראל׃ of Israel.}%
\verse{ויאמר Then said אבשׁלום Absalom, קרא Call נא now גם also, לחושׁי Hushai הארכי the Archite ונשׁמעה and let us hear מה what בפיו saith. גם likewise הוא׃ he}%
\verse{ויבא was come חושׁי And when Hushai אל to אבשׁלום Absalom, ויאמר spoke אבשׁלום Absalom אליו unto לאמר him, saying, כדבר manner: הזה after this דבר hath spoken אחיתפל Ahithophel הנעשׂה shall we do את דברו his saying? אם if אין not; אתה thou. דבר׃ speak}%
\verse{ויאמר said חושׁי And Hushai אל unto אבשׁלום Absalom, לא not טובה good העצה The counsel אשׁר that יעץ hath given אחיתפל Ahithophel בפעם time. הזאת׃ at this}%
\verse{ויאמר For, said חושׁי Hushai, אתה thou ידעת knowest את אביך thy father ואת אנשׁיו and his men, כי that גברים mighty men, המה they ומרי chafed נפשׁ in their minds, המה and they כדב as a bear שׁכול robbed of her whelps בשׂדה in the field: ואביך and thy father אישׁ a man מלחמה of war, ולא and will not ילין lodge את with העם׃ the people.}%
\verse{הנה Behold, עתה now הוא he נחבא is hid באחת in some הפחתים pit, או or באחד in some המקומת place: והיה and it will come to pass, כנפל when some of them be overthrown בהם בתחלה at the first, ושׁמע that whosoever heareth השׁמע that whosoever heareth ואמר it will say, היתה There is מגפה a slaughter בעם among the people אשׁר that אחרי follow אבשׁלם׃ Absalom.}%
\verse{והוא And he גם also בן valiant, חיל valiant, אשׁר whose לבו heart כלב as the heart האריה of a lion, המס shall utterly melt: ימס shall utterly melt: כי for ידע knoweth כל all ישׂראל Israel כי that גבור a mighty man, אביך thy father ובני men. חיל him valiant אשׁר and which אתו׃ with}%
\verse{כי Therefore יעצתי I counsel האסף be generally יאסף gathered עליך unto כל that all ישׂראל Israel מדן ועד even to באר שׁבע Beer-sheba, כחול as the sand אשׁר that על by הים the sea לרב for multitude; ופניך in thine own person. הלכים and that thou go בקרב׃ to battle}%
\verse{ובאנו So shall we come אליו upon באחת him in some המקומת place אשׁר where נמצא he shall be found, שׁם where ונחנו and we will light עליו upon כאשׁר him as יפל falleth הטל the dew על on האדמה the ground: ולא him there shall not נותר be left בו ובכל and of him and of all האנשׁים אשׁר that אתו with גם so much אחד׃ as one.}%
\verse{ואם Moreover, if אל into עיר a city, יאסף he be gotten והשׂיאו bring כל then shall all ישׂראל Israel אל to העיר city, ההיא that חבלים ropes וסחבנו and we will draw אתו עד it into הנחל the river, עד until אשׁר until לא there be not נמצא found שׁם there. גם one small צרור׃ stone}%
\verse{ויאמר said, אבשׁלום And Absalom וכל and all אישׁ the men ישׂראל of Israel טובה better עצת The counsel חושׁי of Hushai הארכי the Archite מעצת than the counsel אחיתפל of Ahithophel. ויהוה For the LORD צוה had appointed להפר to defeat את עצת counsel אחיתפל of Ahithophel, הטובה the good לבעבור to the intent הביא might bring יהוה that the LORD אל upon אבשׁלום Absalom. את הרעה׃ evil}%
\verseWithHeading{Hushai Sends Word to David}{ויאמר Then said חושׁי Hushai אל unto צדוק Zadok ואל and to אביתר Abiathar הכהנים the priests, כזאת Thus וכזאת and thus יעץ counsel אחיתפל did Ahithophel את אבשׁלם Absalom ואת זקני and the elders ישׂראל of Israel; וכזאת and thus וכזאת and thus יעצתי counseled. אני׃ have I}%
\verse{ועתה Now שׁלחו therefore send מהרה quickly, והגידו and tell לדוד David, לאמר saying, אל not תלן Lodge הלילה this night בערבות in the plains המדבר of the wilderness, וגם but עבור speedily תעבור pass over; פן lest יבלע be swallowed up, למלך the king ולכל and all העם the people אשׁר that אתו׃ with}%
\verse{ויהונתן Now Jonathan ואחימעץ and Ahimaaz עמדים stayed בעין רגל by En-rogel; והלכה went השׁפחה and a wench והגידה and told להם והם them; and they ילכו went והגידו and told למלך king דוד David. כי for לא not יוכלו they might להראות be seen לבוא to come העירה׃ into the city:}%
\verse{וירא saw אתם נער Nevertheless a lad ויגד them, and told לאבשׁלם Absalom: וילכו but they went שׁניהם both מהרה of them away quickly, ויבאו and came אל to בית house אישׁ a man's בבחורים in Bahurim, ולו באר which had a well בחצרו in his court; וירדו they went down. שׁם׃ whither}%
\verse{ותקח took האשׁה And the woman ותפרשׂ and spread את המסך a covering על over פני mouth, הבאר the well's ותשׁטח and spread עליו thereon; הרפות ground corn ולא was not נודע known. דבר׃ and the thing}%
\verse{ויבאו came עבדי servants אבשׁלום And when Absalom's אל to האשׁה the woman הביתה to the house, ויאמרו they said, איה Where אחימעץ Ahimaaz ויהונתן and Jonathan? ותאמר said להם האשׁה And the woman עברו unto them, They be gone over מיכל the brook המים of water. ויבקשׁו And when they had sought ולא and could not מצאו find וישׁבו they returned ירושׁלם׃ to Jerusalem.}%
\verse{ויהי And it came to pass, אחרי after לכתם they were departed, ויעלו that they came up מהבאר out of the well, וילכו and went ויגדו and told למלך king דוד David, ויאמרו and said אל unto דוד David, קומו Arise, ועברו and pass quickly over מהרה and pass quickly over את המים the water: כי for ככה thus יעץ counseled עליכם against אחיתפל׃ hath Ahithophel}%
\verse{ויקם arose, דוד Then David וכל and all העם the people אשׁר that אתו with ויעברו him, and they passed over את הירדן Jordan: עד by אור light הבקר the morning עד אחד one לא not נעדר there lacked אשׁר of them that לא was not עבר gone over את הירדן׃ Jordan.}%
\verse{ואחיתפל And when Ahithophel ראה saw כי that לא was not נעשׂתה followed, עצתו his counsel ויחבשׁ he saddled את החמור ass, ויקם and arose, וילך and got אל him home to ביתו his house, אל to עירו his city, ויצו and put אל and put ביתו his household ויחנק in order, and hanged himself, וימת and died, ויקבר and was buried בקבר in the sepulcher אביו׃ of his father.}%
\verse{ודוד Then David בא came מחנימה to Mahanaim. ואבשׁלם And Absalom עבר passed over את הירדן Jordan, הוא he וכל and all אישׁ the men ישׂראל of Israel עמו׃ with}%
\verse{ואת עמשׂא Amasa שׂם made אבשׁלם And Absalom תחת instead יואב of Joab: על captain הצבא of the host ועמשׂא which Amasa בן son, אישׁ a man's ושׁמו whose name יתרא Ithra הישׂראלי an Israelite, אשׁר that בא went אל in to אביגל Abigail בת the daughter נחשׁ of Nahash, אחות sister צרויה to Zeruiah אם mother. יואב׃ Joab's}%
\verse{ויחן pitched ישׂראל So Israel ואבשׁלם and Absalom ארץ in the land הגלעד׃ of Gilead.}%
\verse{ויהי And it came to pass, כבוא was come דוד when David מחנימה to Mahanaim, ושׁבי that Shobi בן the son נחשׁ of Nahash מרבת בני of the children עמון of Ammon, ומכיר and Machir בן the son עמיאל of Ammiel מלא דבר וברזלי and Barzillai הגלעדי the Gileadite מרגלים׃}%
\verse{משׁכב beds, וספות and basins, וכלי vessels, יוצר and earthen וחטים and wheat, ושׂערים and barley, וקמח and flour, וקלי and parched ופול and beans, ועדשׁים and lentils, וקלי׃ and parched}%
\verse{ודבשׁ And honey, וחמאה and butter, וצאן and sheep, ושׁפות and cheese בקר of kine, הגישׁו לדוד for David, ולעם and for the people אשׁר that אתו with לאכול him, to eat: כי for אמרו they said, העם The people רעב hungry, ועיף and weary, וצמא and thirsty, במדבר׃ in the wilderness.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 18
\verseWithHeading{Absalom Dies in Battle}{ויפקד numbered דוד And David את העם the people אשׁר that אתו with וישׂם him, and set עליהם over שׂרי captains אלפים of thousands ושׂרי and captains מאות׃ of hundreds}%
\verse{וישׁלח sent forth דוד And David את העם of the people השׁלשׁית a third part ביד under the hand יואב of Joab, והשׁלשׁית and a third part ביד under the hand אבישׁי of Abishai בן the son צרויה of Zeruiah, אחי brother, יואב Joab's והשׁלשׁת and a third part ביד under the hand אתי of Ittai הגתי the Gittite. ויאמר said המלך And the king אל unto העם the people, יצא אצאם also. אני you myself עמכם׃ with}%
\verse{ויאמר answered, העם But the people לא Thou shalt not תצא go forth: כי for אם if נס we flee away, ננוס we flee away, לא they will not ישׂימו care אלינו for לב care ואם if ימתו of us die, חצינו half לא us; neither ישׂימו will they care אלינו for לב will they care כי us: but עתה now כמנו worth עשׂרה ten אלפים thousand ועתה of us: therefore now טוב better כי that תהיה לנו מעיר us out of the city. לעזיר׃ thou succor}%
\verse{ויאמר said אליהם unto המלך And the king אשׁר them, What ייטב you best בעיניכם seemeth אעשׂה I will do. ויעמד stood המלך And the king אל by יד side, השׁער the gate וכל and all העם the people יצאו למאות by hundreds ולאלפים׃ and by thousands.}%
\verse{ויצו commanded המלך And the king את יואב Joab ואת אבישׁי and Abishai ואת אתי and Ittai, לאמר saying, לאט gently לי לנער for my sake with the young man, לאבשׁלום with Absalom. וכל And all העם the people שׁמעו heard בצות charge המלך when the king את כל gave all השׂרים the captains על concerning דבר concerning אבשׁלום׃ Absalom.}%
\verse{ויצא went out העם So the people השׂדה into the field לקראת against ישׂראל Israel: ותהי was המלחמה and the battle ביער in the wood אפרים׃ of Ephraim;}%
\verse{וינגפו were slain שׁם Where עם the people ישׂראל of Israel לפני before עבדי the servants דוד of David, ותהי and there was שׁם there המגפה slaughter גדולה a great ביום day ההוא that עשׂרים of twenty אלף׃ thousand}%
\verse{ותהי was שׁם there המלחמה For the battle נפצית scattered על over פני the face כל of all הארץ the country: וירב more היער and the wood לאכל devoured בעם people מאשׁר אכלה devoured. החרב than the sword ביום day ההוא׃ that}%
\verse{ויקרא met אבשׁלום And Absalom לפני עבדי the servants דוד of David. ואבשׁלום And Absalom רכב rode על upon הפרד a mule, ויבא went הפרד and the mule תחת under שׂובך the thick boughs האלה oak, הגדולה of a great ויחזק caught hold ראשׁו and his head באלה of the oak, ויתן and he was taken up בין between השׁמים the heaven ובין הארץ and the earth; והפרד and the mule אשׁר that תחתיו under עבר׃ him went away.}%
\verse{וירא saw אישׁ man אחד And a certain ויגד and told ליואב Joab, ויאמר and said, הנה Behold, ראיתי I saw את אבשׁלם Absalom תלוי hanged באלה׃ in an oak.}%
\verse{ויאמר said יואב And Joab לאישׁ unto the man המגיד that told לו והנה him, And, behold, ראית thou sawest ומדוע and why לא didst thou not הכיתו smite שׁם him there ארצה to the ground? ועלי לתת and I would have given לך עשׂרה thee ten כסף of silver, וחגרה girdle. אחת׃ and a}%
\verse{ויאמר said האישׁ And the man אל unto יואב Joab, ולא Though אנכי I שׁקל should receive על in כפי mine hand, אלף a thousand כסף of silver לא would I not אשׁלח put forth ידי mine hand אל against בן son: המלך the king's כי for באזנינו in our hearing צוה charged המלך the king אתך ואתבישׁי thee and Abishai ואת אתי and Ittai, לאמר saying, שׁמרו מינער the young man באבשׁלום׃ Absalom.}%
\verse{או Otherwise עשׂיתי I should have wrought בנפשׁו against mine own life: שׁקר falsehood וכל דברא for there is no matter יכחד hid מן from המלך the king, ואתה and thou thyself תתיצב wouldest have set thyself מנגד׃ against}%
\verse{ויאמר Then said יואב Joab, לא I may not כן thus אחילה tarry לפניך with ויקח thee. And he took שׁלשׁה three שׁבטים darts בכפו in his hand, ויתקעם and thrust בלב them through the heart אבשׁלום of Absalom, עודנו while חי he yet alive בלב in the midst האלה׃ of the oak.}%
\verse{ויסבו compassed about עשׂרה And ten נערים young men נשׂאי that bore כלי armor יואב Joab's ויכו and smote את אבשׁלום Absalom, וימיתהו׃ and slew}%
\verse{ויתקע blew יואב And Joab בשׁפר the trumpet, וישׁב returned העם and the people מרדף from pursuing אחרי after ישׂראל Israel: כי for חשׂך held back יואב Joab את העם׃ the people.}%
\verse{ויקחו And they took את אבשׁלום Absalom, וישׁליכו and cast אתו ביער in the wood, אל him into הפחת pit הגדול a great ויצבו and laid עליו upon גל heap אבנים of stones גדול great מאד a very וכל him: and all ישׂראל Israel נסו fled אישׁ every one לאהלו׃ to his tent.}%
\verse{ואבשׁלם Now Absalom לקח had taken ויצב and reared up לו בחיו in his lifetime את מצבת for himself a pillar, אשׁר which בעמק dale: המלך in the king's כי for אמר he said, אין I have no לי בן son בעבור to הזכיר keep my name in remembrance: שׁמי keep my name in remembrance: ויקרא and he called למצבת the pillar על after שׁמו his own name: ויקרא and it is called לה יד place. אבשׁלם Absalom's עד unto היום day, הזה׃ this}%
\verse{ואחימעץ Ahimaaz בן the son צדוק of Zadok, אמר Then said ארוצה run, נא Let me now ואבשׂרה and bear את המלך the king כי tidings, how that שׁפטו hath avenged יהוה the LORD מיד איביו׃ him of his enemies.}%
\verse{ויאמר said לו יואב And Joab לא shalt not אישׁ bear tidings בשׂרה bear tidings אתה unto him, Thou היום day, הזה this ובשׂרת but thou shalt bear tidings ביום day: אחר another והיום day הזה but this לא thou shalt bear no tidings, תבשׂר thou shalt bear no tidings, כי because על because בן son המלך the king's מת׃ is dead.}%
\verse{ויאמר Then said יואב Joab לכושׁי to Cushi, לך Go הגד tell למלך the king אשׁר what ראיתה thou hast seen. וישׁתחו bowed himself כושׁי And Cushi ליואב unto Joab, וירץ׃ and ran.}%
\verse{ויסף again עוד yet אחימעץ Ahimaaz בן the son צדוק of Zadok ויאמר Then said אל to יואב Joab, ויהי מהרצה run נא pray גם thee, also אני let me, I אחרי after הכושׁי Cushi. ויאמר said, יואב And Joab למה זהתה wilt thou רץ run, בני my son, ולכה אין seeing that thou hast no בשׂורה tidings מצאת׃ ready?}%
\verse{ויהי מהרוץ let me run. ויאמר And he said לו רוץ unto him, Run. וירץ ran אחימעץ Then Ahimaaz דרך by the way הככר of the plain, ויעבר and overran את הכושׁי׃ Cushi.}%
\verse{ודוד And David יושׁב sat בין between שׁני the two השׁערים gates: וילך went up הצפה and the watchman אל to גג the roof השׁער over the gate אל unto החומה the wall, וישׂא and lifted up את עיניו his eyes, וירא and looked, והנה and behold אישׁ a man רץ running לבדו׃ alone.}%
\verse{ויקרא cried, הצפה And the watchman ויגד and told למלך the king. ויאמר said, המלך And the king אם If לבדו he alone, בשׂורה tidings בפיו in his mouth. וילך And he came הלוך apace, וקרב׃ and drew near.}%
\verse{וירא saw הצפה And the watchman אישׁ man אחר another רץ running: ויקרא called הצפה and the watchman אל unto השׁער the porter, ויאמר and said, הנה Behold אישׁ man רץ running לבדו alone. ויאמר said, המלך And the king גם also זה He מבשׂר׃ bringeth tidings.}%
\verse{ויאמר said, הצפה And the watchman אני Me ראה thinketh את מרוצת the running הראשׁון of the foremost כמרצת is like the running אחימעץ of Ahimaaz בן the son צדוק of Zadok. ויאמר said, המלך And the king אישׁ man, טוב a good זה He ואל בשׂורה tidings. טובה with good יבוא׃ and cometh}%
\verse{ויקרא called, אחימעץ And Ahimaaz ויאמר and said אל unto המלך the king, שׁלום All is well. וישׁתחו למלך before the king, לאפיו upon his face ארצה to the earth ויאמר and said, ברוך Blessed יהוה the LORD אלהיך thy God, אשׁר which סגר hath delivered up את האנשׁים the men אשׁר that נשׂאו lifted up את ידם their hand באדני against my lord המלך׃ the king.}%
\verse{ויאמר said, המלך And the king שׁלום safe? לנער Is the young man לאבשׁלום Absalom ויאמר answered, אחימעץ And Ahimaaz ראיתי I saw ההמון tumult, הגדול a great לשׁלח sent את עבד servant, המלך the king's יואב When Joab ואת עבדך and thy servant, ולא not ידעתי but I knew מה׃ what}%
\verse{ויאמר said המלך And the king סב Turn aside, התיצב stand כה here. ויסב And he turned aside, ויעמד׃ and stood still.}%
\verse{והנה And, behold, הכושׁי Cushi בא came; ויאמר said, הכושׁי and Cushi יתבשׂר Tidings, אדני my lord המלך the king: כי for שׁפטך hath avenged יהוה the LORD היום thee this day מיד of כל all הקמים them that rose up עליך׃ against}%
\verse{ויאמר said המלך And the king אל unto הכושׁי Cushi, השׁלום safe? לנער Is the young man לאבשׁלום Absalom ויאמר answered, הכושׁי And Cushi יהיו be כנער as young man איבי The enemies אדני of my lord המלך the king, וכל and all אשׁר that קמו rise עליך against לרעה׃ thee to do hurt,}%
\verse{וירגז was much moved, המלך And the king ויעל and went up על to עלית the chamber השׁער over the gate, ויבך and wept: וכה thus אמר he said, בלכתו and as he went, בני O my son אבשׁלום Absalom, בני my son, בני my son אבשׁלום Absalom! מי יתןותי had died אני I תחתיך for אבשׁלום thee, O Absalom, בני my son, בני׃ my son!}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 19
\verseWithHeading{King David Weeps for Absalom}{ויגד And it was told ליואב Joab, הנה Behold, המלך the king בכה weepeth ויתאבל and mourneth על for אבשׁלם׃ Absalom.}%
\verse{ותהי that התשׁעה And the victory ביום day ההוא that לאבל was into mourning לכל unto all העם the people: כי for שׁמע heard העם the people ביום day ההוא לאמר say נעצב was grieved המלך how the king על for בנו׃ his son.}%
\verse{ויתגנב them by stealth העם And the people ביום day ההוא that לבוא got העיר into the city, כאשׁר as יתגנב steal away העם people הנכלמים being ashamed בנוסם when they flee במלחמה׃ in battle.}%
\verse{והמלך But the king לאט covered את פניו his face, ויזעק cried המלך and the king קול voice, גדול with a loud בני O my son אבשׁלום Absalom, אבשׁלום O Absalom, בני my son, בני׃ my son!}%
\verse{ויבא came יואב And Joab אל to המלך the king, הבית into the house ויאמר and said, הבשׁת Thou hast shamed היום this day את פני the faces כל of all עבדיך thy servants, הממלטים have saved את נפשׁך thy life, היום which this day ואת נפשׁ and the lives בניך of thy sons ובנתיך and of thy daughters, ונפשׁ and the lives נשׁיך of thy wives, ונפשׁ and the lives פלגשׁיך׃ of thy concubines;}%
\verse{לאהבה In that thou lovest את שׂנאיך thine enemies, ולשׂנא and hatest את אהביך thy friends. כי For הגדת thou hast declared היום this day, כי that אין thou regardest neither לך שׂרים princes ועבדים nor servants: כי for ידעתי I perceive, היום this day כי that לא if אבשׁלום Absalom חי had lived, וכלנו and all היום this day, מתים we had died כי אז then ישׁר it had pleased thee well. בעיניך׃ it had pleased thee well.}%
\verse{ועתה Now קום therefore arise, צא go forth, ודבר and speak על comfortably לב comfortably עבדיך unto thy servants: כי for ביהוה by the LORD, נשׁבעתי I swear כי if אינך thou go not forth, יוצא thou go not forth, אם there will not ילין tarry אישׁ one אתך with הלילה thee this night: ורעה the evil לך זאת and that מכל unto thee than all הרעה אשׁר that באה befell עליך befell מנעריך thee from thy youth עד until עתה׃ now.}%
\verse{ויקם arose, המלך Then the king וישׁב and sat בשׁער in the gate. ולכל unto all העם the people, הגידו And they told לאמר saying, הנה Behold, המלך the king יושׁב doth sit בשׁער in the gate. ויבא came כל And all העם the people לפני before המלך the king: וישׂראל for Israel נס had fled אישׁ every man לאהליו׃ to his tent.}%
\verseWithHeading{David Returns from Exile}{ויהי were כל And all העם the people נדון at strife בכל throughout all שׁבטי the tribes ישׂראל of Israel, לאמר saying, המלך The king הצילנו saved מכף us out of the hand איבינו of our enemies, והוא and he מלטנו delivered מכף us out of the hand פלשׁתים of the Philistines; ועתה and now ברח he is fled מן us out of the hand הארץ the land מעל for אבשׁלום׃ Absalom.}%
\verse{ואבשׁלום And Absalom, אשׁר whom משׁחנו we anointed עלינו over מת us, is dead במלחמה in battle. ועתה Now למה therefore why אתם ye מחרשׁים speak להשׁיב not a word of bringing את המלך׃ the king}%
\verse{והמלך And king דוד David שׁלח sent אל to צדוק Zadok ואל and to אביתר Abiathar הכהנים the priests, לאמר saying, דברו Speak אל unto זקני the elders יהודה of Judah, לאמר saying, למה Why תהיו are אחרנים ye the last להשׁיב to bring את המלך the king אל back to ביתו his house? ודבר seeing the speech כל of all ישׂראל Israel בא is come אל to המלך the king, אל to ביתו׃ his house.}%
\verse{אחי my brethren, אתם Ye עצמי my bones ובשׂרי and my flesh: אתם ye ולמה wherefore תהיו then are אחרנים ye the last להשׁיב to bring back את המלך׃ the king?}%
\verse{ולעמשׂא ye to Amasa, תמרו And say הלוא not עצמי of my bone, ובשׂרי and of my flesh? אתה thou כה so יעשׂה do לי אלהים God וכה also, יוסיף to me, and more אם if לא not שׂר captain צבא of the host תהיה thou be לפני before כל me continually הימים me continually תחת in the room יואב׃ of Joab.}%
\verse{ויט And he bowed את לבב the heart כל of all אישׁ the men יהודה of Judah, כאישׁ man; אחד even as one וישׁלחו so that they sent אל unto המלך the king, שׁוב Return אתה thou, וכל and all עבדיך׃ thy servants.}%
\verse{וישׁב returned, המלך So the king ויבא and came עד to הירדן Jordan. ויהודה And Judah בא came הגלגלה to Gilgal, ללכת to go לקראת to meet המלך the king, להעביר אתמלך אתירדן׃ Jordan.}%
\verse{וימהר hasted שׁמעי And Shimei בן the son גרא of Gera, בן הימיני a Benjamite, אשׁר which מבחורים וירד and came down עם with אישׁ the men יהודה of Judah לקראת to meet המלך king דוד׃ David.}%
\verse{ואלף And a thousand אישׁ men עמו with מבנימן וציבא him, and Ziba נער the servant בית of the house שׁאול of Saul, וחמשׁת and his fifteen עשׂר and his fifteen בניו sons ועשׂרים and his twenty עבדיו servants אתו with וצלחו him; and they went over הירדן Jordan לפני before המלך׃ the king.}%
\verse{ועברה And there went over העברה a ferry boat לעביר to carry over את בית household, המלך the king's ולעשׂות and to do הטוב good. בעינו what he thought ושׁמעי And Shimei בן the son גרא of Gera נפל fell down לפני before המלך the king, בעברו as he was come over בירדן׃ Jordan;}%
\verse{ויאמר And said אל unto המלך the king, אל Let not יחשׁב impute לי אדני my lord עון iniquity ואל unto me, neither תזכר do thou remember את אשׁר that which העוה did perversely עבדך thy servant ביום the day אשׁר that יצא went out אדני my lord המלך the king מירושׁלם לשׂום should take המלך that the king אל it to לבו׃ his heart.}%
\verse{כי For ידע doth know עבדך thy servant כי that אני I חטאתי have sinned: והנה therefore, behold, באתי I am come היום this day ראשׁון the first לכל of all בית the house יוסף of Joseph לרדת to go down לקראת to meet אדני my lord המלך׃ the king.}%
\verse{ויען answered אבישׁי But Abishai בן the son צרויה of Zeruiah ויאמר and said, התחת for זאת this, לא Shall not יומת be put to death שׁמעי Shimei כי because קלל he cursed את משׁיח anointed? יהוה׃ the LORD's}%
\verse{ויאמר said, דוד And David מה What לי ולכם בני have I to do with you, ye sons צרויה of Zeruiah, כי that תהיו be לי היום ye should this day לשׂטן adversaries היום this day יומת be put to death אישׁ unto me? shall there any man בישׂראל in Israel? כי for הלוא do not ידעתי I know כי that היום this day אני I מלך king על over ישׂראל׃ Israel?}%
\verse{ויאמר said המלך Therefore the king אל unto שׁמעי Shimei, לא Thou shalt not תמות die. וישׁבע swore לו המלך׃ And the king}%
\verse{ומפבשׁת And Mephibosheth בן the son שׁאול of Saul ירד came down לקראת to meet המלך the king, ולא and had neither עשׂה dressed רגליו his feet, ולא nor עשׂה trimmed שׂפמו his beard, ואת בגדיו his clothes, לא nor כבס washed למן from היום the day לכת departed המלך the king עד until היום the day אשׁר בא he came בשׁלום׃ in peace.}%
\verse{ויהי And it came to pass, כי when בא he was come ירושׁלם to Jerusalem לקראת to meet המלך the king, ויאמר said לו המלך that the king למה unto him, Wherefore לא not הלכת wentest עמי thou with מפיבשׁת׃ me, Mephibosheth?}%
\verse{ויאמר And he answered, אדני My lord, המלך O king, עבדי my servant רמני deceived כי me: for אמר said, עבדך thy servant אחבשׁה I will saddle לי החמור me an ass, וארכב that I may ride עליה thereon, ואלך and go את to המלך the king; כי because פסח lame. עבדך׃ thy servant}%
\verse{וירגל And he hath slandered בעבדך thy servant אל unto אדני my lord המלך the king; ואדני but my lord המלך the king כמלאך as an angel האלהים of God: ועשׂה do הטוב therefore good בעיניך׃ in thine eyes.}%
\verse{כי For לא היה were כל all בית house אבי my father's כי but אם but אנשׁי men מות dead לאדני my lord המלך the king: ותשׁת yet didst thou set את עבדך thy servant באכלי among them that did eat שׁלחנך at thine own table. ומה What ישׁ therefore have לי עוד any more צדקה right ולזעק to cry עוד אל unto המלך׃ the king?}%
\verse{ויאמר said לו המלך And the king למה unto him, Why תדבר speakest עוד thou any more דבריך of thy matters? אמרתי I have said, אתה Thou וציבא and Ziba תחלקו divide את השׂדה׃ the land.}%
\verse{ויאמר said מפיבשׁת And Mephibosheth אל unto המלך the king, גם Yea, את הכל all, יקח let him take אחרי forasmuch אשׁר forasmuch בא is come again אדני as my lord המלך the king בשׁלום in peace אל unto ביתו׃ his own house.}%
\verse{וברזלי And Barzillai הגלעדי the Gileadite ירד came down מרגלים ויעבר and went over את with המלך the king, הירדן Jordan לשׁלחו to conduct him over את בירדן׃ Jordan.}%
\verse{וברזלי Now Barzillai זקן aged מאד was a very בן old: שׁמנים man, fourscore שׁנה years והוא and he כלכל of sustenance את המלך had provided the king בשׁיבתו while he lay במחנים at Mahanaim; כי for אישׁ man. גדול great הוא he מאד׃ a very}%
\verse{ויאמר said המלך And the king אל unto ברזלי Barzillai, אתה thou עבר Come אתי over with וכלכלתי me, and I will feed אתך עמדיירושׁלם׃ me in Jerusalem.}%
\verse{ויאמר said ברזלי And Barzillai אל unto המלך the king, כמה How ימי long שׁני חיי have I to live, כי that אעלה I should go up את with המלך the king ירושׁלם׃ unto Jerusalem?}%
\verse{בן old: שׁמנים fourscore שׁנה years אנכי I היום this day האדע can I discern בין between טוב good לרע and evil? אם can יטעם taste עבדך thy servant את אשׁר what אכל I eat ואת אשׁר or what אשׁתה I drink? אם can אשׁמע I hear עוד any more בקול the voice שׁרים of singing men ושׁרות and singing women? ולמה wherefore יהיה be עבדך then should thy servant עוד yet למשׂא a burden אל unto אדני my lord המלך׃ the king?}%
\verse{כמעט will go a little way over יעבר will go a little way over עבדך Thy servant את הירדן Jordan את with המלך the king: ולמה and why יגמלני recompense המלך should the king הגמולה a reward? הזאת׃ it me with such}%
\verse{ישׁב turn back again, נא I pray thee, עבדך Let thy servant, ואמת that I may die בעירי in mine own city, עם by קבר the grave אבי of my father ואמי and of my mother. והנה But behold עבדך thy servant כמהם Chimham; יעבר let him go over עם with אדני my lord המלך the king; ועשׂה and do לו את אשׁר to him what טוב good בעיניך׃ shall seem}%
\verse{ויאמר answered, המלך And the king אתי with יעבר shall go over כמהם Chimham ואני me, and I אעשׂה will do לו את הטוב good בעיניך to him that which shall seem וכל unto thee: and whatsoever אשׁר unto thee: and whatsoever תבחר thou shalt require עלי of אעשׂה׃ me, will I do}%
\verse{ויעבר went over כל And all העם the people את הירדן Jordan. והמלך And when the king עבר was come over, וישׁק kissed המלך the king לברזלי Barzillai, ויברכהו and blessed וישׁב him; and he returned למקמו׃ unto his own place.}%
\verse{ויעבר went on המלך Then the king הגלגלה to Gilgal, וכמהן and Chimham עבר went on עמו with וכל him: and all עם the people יהודה of Judah ויעברו conducted את המלך the king, וגם and also חצי half עם the people ישׂראל׃ of Israel.}%
\verse{והנה And, behold, כל all אישׁ the men ישׂראל of Israel באים came אל to המלך the king, ויאמרו and said אל unto המלך the king, מדוע Why גנבוך stolen thee away, אחינו have our brethren אישׁ the men יהודה of Judah ויעברו and have brought את המלך the king, ואת ביתו and his household, את הירדן him, over Jordan? וכל and all אנשׁי men דוד David's עמו׃ with}%
\verse{ויען answered כל And all אישׁ the men יהודה of Judah על answered אישׁ the men ישׂראל of Israel, כי Because קרוב near of kin המלך the king אלי to ולמה us: wherefore זה us: wherefore חרה then be ye angry לך על for הדבר matter? הזה this האכול have we eaten at all אכלנו have we eaten at all מן of המלך the king's אם or נשׂאת hath he given us any gift? נשׂא׃ hath he given us any gift?}%
\verse{ויען answered אישׁ And the men ישׂראל of Israel את אישׁ the men יהודה of Judah, ויאמר and said, עשׂר We have ten ידות parts לי במלך in the king, וגם have also בדוד more in David אני and we ממך than ומדוע ye: why הקלתני then did ye despise ולא should not היה be דברי us, that our advice ראשׁון first לי להשׁיב had in bringing back את מלכי our king? ויקשׁ were fiercer דבר And the words אישׁ of the men יהודה of Judah מדבר than the words אישׁ of the men ישׂראל׃ of Israel.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 20
\verseWithHeading{Sheba Leads a Revolt}{ושׁם And there נקרא happened אישׁ to be there a man בליעל of Belial, ושׁמו whose name שׁבע Sheba, בן the son בכרי of Bichri, אישׁ every man ימיני ויתקע and he blew בשׁפר a trumpet, ויאמר and said, אין We have no לנו חלק part בדוד in David, ולא neither נחלה have we inheritance לנו בבן in the son ישׁי of Jesse: אישׁ לאהליו to his tents, ישׂראל׃ O Israel.}%
\verse{ויעל went up כל So every אישׁ man ישׂראל of Israel מאחרי from after דוד David, אחרי followed שׁבע Sheba בן the son בכרי of Bichri: ואישׁ but the men יהודה of Judah דבקו cleaved במלכם unto their king, מן from after הירדן Jordan ועד even to ירושׁלם׃ Jerusalem.}%
\verse{ויבא came דוד And David אל to ביתו his house ירושׁלם at Jerusalem; ויקח took המלך and the king את עשׂר the ten נשׁים women פלגשׁים concubines, אשׁר whom הניח לשׁמר to keep הבית the house, ויתנם and put בית them in ward, משׁמרת them in ward, ויכלכלם and fed ואליהם in unto לא not בא them, but went ותהיינה them. So they were צררות shut up עד unto יום the day מתן אלמנות in widowhood. חיות׃ living}%
\verseWithHeading{Joab Assassinates Amasa}{ויאמר Then said המלך the king אל to עמשׂא Amasa, הזעק Assemble לי את אישׁ me the men יהודה of Judah שׁלשׁת within three ימים days, ואתה and be thou פה here עמד׃ present.}%
\verse{וילך went עמשׂא So Amasa להזעיק to assemble את יהודה Judah: וייחר מן than המועד the set time אשׁר which יעדו׃ he had appointed}%
\verse{ויאמר said דוד And David אל to אבישׁי Abishai, עתה Now ירע do us more harm לנו שׁבע shall Sheba בן the son בכרי of Bichri מן than אבשׁלום Absalom: אתה thou קח take את עבדי servants, אדניך thy lord's ורדף and pursue אחריו after פן him, lest מצא he get לו ערים cities, בצרות him fenced והציל and escape עיננו׃ and escape}%
\verse{ויצאו And there went out אחריו after אנשׁי men, יואב him Joab's והכרתי and the Cherethites, והפלתי and the Pelethites, וכל and all הגברים the mighty men: ויצאו and they went out מירושׁלם לרדף to pursue אחרי after שׁבע Sheba בן the son בכרי׃ of Bichri.}%
\verse{הם When they עם at האבן stone הגדולה the great אשׁר which בגבעון in Gibeon, ועמשׂא Amasa בא went לפניהם before ויואב them. And Joab's חגור was girded מדו garment לבשׁו that he had put on ועלו unto חגור חרב a sword מצמדת fastened על him, and upon מתניו his loins בתערה in the sheath והוא thereof; and as he יצא went forth ותפל׃ it fell out.}%
\verse{ויאמר said יואב And Joab לעמשׂא to Amasa, השׁלום in health, אתה thou אחי my brother? ותחז took יד hand ימין with the right יואב And Joab בזקן by the beard עמשׂא Amasa לנשׁק׃ to kiss}%
\verse{ועמשׂא But Amasa לא took no נשׁמר heed בחרב to the sword אשׁר that ביד hand: יואב in Joab's ויכהו so he smote בה אל him therewith in החמשׁ the fifth וישׁפך and shed out מעיו his bowels ארצה to the ground, ולא him not שׁנה and struck לו וימת again; and he died. ויואב So Joab ואבישׁי and Abishai אחיו his brother רדף pursued אחרי after שׁבע Sheba בן the son בכרי׃ of Bichri.}%
\verse{ואישׁ And one עמד עליו by מנערי יואב Joab, ויאמר him, and said, מי He אשׁר that חפץ favoreth ביואב Joab. ומי and he אשׁר that לדוד for David, אחרי after יואב׃}%
\verse{ועמשׂא And Amasa מתגלל wallowed בדם in blood בתוך in the midst המסלה of the highway. וירא saw האישׁ And when the man כי that עמד stood still, כל all העם the people ויסב he removed את עמשׂא Amasa מן out of המסלה the highway השׂדה into the field, וישׁלך and cast עליו upon בגד a cloth כאשׁר him, when ראה he saw כל that every הבא one that came עליו by ועמד׃ him stood still.}%
\verse{כאשׁר When הגה he was removed מן out of המסלה the highway, עבר went on כל all אישׁ the people אחרי after יואב Joab, לרדף to pursue אחרי after שׁבע Sheba בן the son בכרי׃ of Bichri.}%
\verseWithHeading{Wisdom from a Woman under Siege}{ויעבר And he went through בכל all שׁבטי the tribes ישׂראל of Israel אבלה unto Abel, ובית מעכה and to Beth-maachah, וכל and all הברים the Berites: ויקלהו and they were gathered together, ויבאו and went אף also אחריו׃ after}%
\verse{ויבאו And they came ויצרו and besieged עליו and besieged באבלה בית המעכהישׁפכו and they cast up סללה a bank אל against העיר the city, ותעמד and it stood בחל in the trench: וכל and all העם the people אשׁר that את with יואב Joab משׁחיתם battered להפיל to throw it down. החומה׃ the wall,}%
\verse{ותקרא Then cried אשׁה woman חכמה a wise מן out of העיר the city, שׁמעו Hear, שׁמעו hear; אמרו say, נא I pray you, אל unto יואב Joab, קרב Come near עד hither, הנה hither, ואדברה that I may speak אליך׃ with}%
\verse{ויקרב And when he was come near אליה unto ותאמר said, האשׁה her, the woman האתה thou יואב Joab? ויאמר And he answered, אני I ותאמר Then she said לו שׁמע unto him Hear דברי the words אמתך of thine handmaid. ויאמר And he answered, שׁמע do hear. אנכי׃ I}%
\verse{ותאמר Then she spoke, לאמר saying, דבר They were wont ידברו to speak בראשׁנה in old time, לאמר saying, שׁאל ישׁאלואבל at Abel: וכן and so התמו׃ they ended}%
\verse{אנכי I שׁלמי peaceable אמוני faithful ישׂראל in Israel: אתה thou מבקשׁ seekest להמית to destroy עיר a city ואם and a mother בישׂראל in Israel: למה why תבלע wilt thou swallow up נחלת the inheritance יהוה׃ of the LORD?}%
\verse{ויען answered יואב And Joab ויאמר and said, חלילה Far be it, חלילה far be it לי אם from me, that אבלע I should swallow up ואם or אשׁחית׃ destroy.}%
\verse{לא not כן so: הדבר The matter כי but אישׁ a man מהר of mount אפרים Ephraim, שׁבע Sheba בן the son בכרי of Bichri שׁמו by name, נשׂא hath lifted up ידו his hand במלך against the king, בדוד against David: תנו deliver אתו לבדו him only, ואלכה and I will depart מעל from העיר the city. ותאמר said האשׁה And the woman אל unto יואב Joab, הנה Behold, ראשׁו his head משׁלך shall be thrown אליך to בעד thee over החומה׃ the wall.}%
\verse{ותבוא went האשׁה Then the woman אל unto כל all העם the people בחכמתה in her wisdom. ויכרתו And they cut off את ראשׁ the head שׁבע of Sheba בן the son בכרי of Bichri, וישׁלכו and cast out אל to יואב Joab. ויתקע And he blew בשׁופר a trumpet, ויפצו and they retired מעל from העיר the city, אישׁ every man לאהליו to his tent. ויואב And Joab שׁב returned ירושׁלם to Jerusalem אל unto המלך׃ the king.}%
\verse{ויואב Now Joab אל over כל all הצבא the host ישׂראל of Israel: ובניה and Benaiah בן the son יהוידע of Jehoiada על over הכרי ועל and over הפלתי׃ the Pelethites:}%
\verse{ואדרם And Adoram על over המס the tribute: ויהושׁפט and Jehoshaphat בן the son אחילוד of Ahilud המזכיר׃ recorder:}%
\verse{ושׁיא And Sheva ספר scribe: וצדוק and Zadok ואביתר and Abiathar כהנים׃ the priests:}%
\verse{וגם also עירא And Ira היארי the Jairite היה was כהן a chief ruler לדוד׃ about David.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 21
\verseWithHeading{The Famine Brings Justice}{ויהי Then there was רעב a famine בימי in the days דוד of David שׁלשׁ three שׁנים years, שׁנה year אחרי after שׁנה year; ויבקשׁ inquired דוד and David את פני inquired יהוה of the LORD. ויאמר answered, יהוה And the LORD אל for שׁאול Saul, ואל and for בית house, הדמים bloody על because אשׁר because המית he slew את הגבענים׃ the Gibeonites.}%
\verse{ויקרא called המלך And the king לגבענים the Gibeonites, ויאמר and said אליהם unto והגבענים them; (now the Gibeonites לא not מבני of the children ישׂראל of Israel, המה כי but אם but מיתר of the remnant האמרי of the Amorites; ובני and the children ישׂראל of Israel נשׁבעו had sworn להם ויבקשׁ sought שׁאול unto them: and Saul להכתם to slay בקנאתו them in his zeal לבני to the children ישׂראל of Israel ויהודה׃ and Judah.)}%
\verse{ויאמר said דוד Wherefore David אל unto הגבענים the Gibeonites, מה What אעשׂה shall I do לכם ובמה for you? and wherewith אכפר shall I make the atonement, וברכו that ye may bless את נחלת the inheritance יהוה׃ of the LORD?}%
\verse{ויאמרו said לו הגבענים And the Gibeonites אין unto him, We will have no לי כסף silver וזהב nor gold עם of שׁאול Saul, ועם nor of ביתו his house; ואין neither לנו אישׁ any man להמית for us shalt thou kill בישׂראל in Israel. ויאמר And he said, מה What אתם ye אמרים shall say, אעשׂה׃ will I do}%
\verse{ויאמרו אלמלך the king, האישׁ The man אשׁר that כלנו consumed ואשׁר us, and that דמה devised לנו נשׁמדנו against us we should be destroyed מהתיצב from remaining בכל in any גבל of the coasts ישׂראל׃ of Israel,}%
\verse{ינתן be delivered לנו שׁבעה Let seven אנשׁים men מבניו of his sons והוקענום unto us, and we will hang ליהוה them up unto the LORD בגבעת in Gibeah שׁאול of Saul, בחיר did choose. יהוה the LORD ויאמר said, המלך And the king אני I אתן׃ will give}%
\verse{ויחמל spared המלך But the king על spared מפיבשׁת Mephibosheth, בן the son יהונתן of Jonathan בן the son שׁאול of Saul, על because שׁבעת oath יהוה of the LORD's אשׁר that בינתם between בין them, between דוד David ובין יהונתן and Jonathan בן the son שׁאול׃ of Saul.}%
\verse{ויקח took המלך But the king את שׁני the two בני sons רצפה of Rizpah בת the daughter איה of Aiah, אשׁר whom ילדה she bore לשׁאול unto Saul, את ארמני Armoni ואת מפבשׁת and Mephibosheth; ואת חמשׁת and the five בני sons מיכל of Michal בת the daughter שׁאול of Saul, אשׁר whom ילדה she brought up לעדריאל for Adriel בן the son ברזלי of Barzillai המחלתי׃ the Meholathite:}%
\verse{ויתנם And he delivered ביד them into the hands הגבענים of the Gibeonites, ויקיעם and they hanged בהר them in the hill לפני before יהוה the LORD: ויפלו and they fell שׁבעתים seven יחד together, והם המתו and were put to death בימי in the days קציר of harvest, בראשׁנים in the first תחלת in the beginning קציר harvest. שׂערים׃ of barley}%
\verse{ותקח took רצפה And Rizpah בת the daughter איה of Aiah את השׂק sackcloth, ותטהו and spread לה אל it for her upon הצור the rock, מתחלת from the beginning קציר of harvest עד until נתך dropped מים water עליהם upon מן from the beginning השׁמים heaven, ולא neither נתנה and suffered עוף the birds השׁמים of the air לנוח to rest עליהם on יומם them by day, ואת חית nor the beasts השׂדה of the field לילה׃ by night.}%
\verse{ויגד And it was told לדוד David את אשׁר what עשׂתה had done. רצפה Rizpah בת the daughter איה of Aiah, פלגשׁ the concubine שׁאול׃ of Saul,}%
\verse{וילך went דוד And David ויקח and took את עצמות the bones שׁאול of Saul ואת עצמות and the bones יהונתן of Jonathan בנו his son מאת בעלי the men יבישׁ גלעדשׁר which גנבו had stolen אתם מרחב them from the street בית שׁן of Beth-shan, אשׁר where תלום שׁם where הפלשׁתים the Philistines ביום them, when הכות had slain פלשׁתים the Philistines את שׁאול Saul בגלבע׃ in Gilboa:}%
\verse{ויעל And he brought up משׁם from thence את עצמות the bones שׁאול of Saul ואת עצמות and the bones יהונתן of Jonathan בנו his son; ויאספו and they gathered את עצמות the bones המוקעים׃ of them that were hanged.}%
\verse{ויקברו buried את עצמות And the bones שׁאול of Saul ויהונתן and Jonathan בנו his son בארץ they in the country בנימן of Benjamin בצלע in Zelah, בקבר in the sepulcher קישׁ of Kish אביו his father: ויעשׂו and they performed כל all אשׁר that צוה commanded. המלך the king ויעתר was entreated אלהים God לארץ for the land. אחרי כן׃}%
\verseWithHeading{Battles with the Philistines Recounted}{ותהי had עוד מלחמה yet war לפלשׁתים Moreover the Philistines את again with ישׂראל Israel; וירד went down, דוד and David ועבדיו and his servants עמו with וילחמו him, and fought against את פלשׁתים the Philistines: ויעף waxed faint. דוד׃ and David}%
\verse{וישׁבו בנב And Ishbibenob, אשׁר which בילידי of the sons הרפה ומשׁקל the weight קינו of whose spear שׁלשׁ three מאות hundred משׁקל in weight, נחשׁת of brass והוא he חגור being girded חדשׁה with a new ויאמר thought להכות to have slain את דוד׃ David.}%
\verse{ויעזר succored לו אבישׁי But Abishai בן the son צרויה of Zeruiah ויך him, and smote את הפלשׁתי the Philistine, וימיתהו and killed אז him. Then נשׁבעו swore אנשׁי the men דוד of David לו לאמר unto him, saying, לא no תצא Thou shalt go עוד more אתנו למלחמה out with us to battle, ולא not תכבה that thou quench את נר the light ישׂראל׃ of Israel.}%
\verse{ויהי And it came to pass אחרי after כן this, ותהי that there was עוד again המלחמה a battle בגוב at Gob: עם with פלשׁתים the Philistines אז then הכה slew סבכי Sibbechai החשׁתי the Hushathite את סף Saph, אשׁר which בילדי of the sons הרפה׃}%
\verse{ותהי And there was עוד again המלחמה a battle בגוב in Gob עם with פלשׁתים the Philistines, ויך slew אלחנן where Elhanan בן the son יערי ארגים of Jaare-oregim, בית הלחמי a Bethlehemite, את גלית Goliath הגתי the Gittite, ועץ the staff חניתו of whose spear כמנור beam. ארגים׃ like a weaver's}%
\verse{ותהי And there was עוד yet מלחמה a battle בגת in Gath, ויהי where was אישׁ a man מדין ואצבעת fingers, ידיו that had on every hand ואצבעות toes, רגליו and on every foot שׁשׁ six ושׁשׁ six עשׂרים and twenty וארבע four מספר in number; וגם also הוא and he ילד was born להרפה׃}%
\verse{ויחרף And when he defied את ישׂראל Israel, ויכהו slew יהונתן Jonathan בן the son שׁמעי אחי the brother דוד׃ of David}%
\verse{את ארבעת four אלה These ילדו were born להרפה בגת in Gath, ויפלו and fell ביד by the hand דוד of David, וביד and by the hand עבדיו׃ of his servants.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 22
\verseWithHeading{The Victory Song of David}{וידבר spoke דוד And David ליהוה unto the LORD את דברי the words השׁירה song הזאת of this ביום in the day הציל had delivered יהוה the LORD אתו מכף him out of the hand כל of all איביו his enemies, ומכף and out of the hand שׁאול׃ of Saul:}%
\verse{ויאמר And he said, יהוה The LORD סלעי my rock, ומצדתי and my fortress, ומפלטי׃ and my deliverer;}%
\verse{אלהי The God צורי of my rock; אחסה in him will I trust: בו מגני my shield, וקרן and the horn ישׁעי of my salvation, משׂגבי my high tower, ומנוסי and my refuge, משׁעי my savior; מחמס me from violence. תשׁעני׃ thou savest}%
\verse{מהלל worthy to be praised: אקרא I will call on יהוה the LORD, ומאיבי from mine enemies. אושׁע׃ so shall I be saved}%
\verse{כי When אפפני compassed משׁברי the waves מות of death נחלי me, the floods בליעל of ungodly יבעתני׃ men made me afraid;}%
\verse{חבלי The sorrows שׁאול of hell סבני compassed me about; קדמני prevented מקשׁי the snares מות׃ of death}%
\verse{בצר In my distress לי אקרא I called upon יהוה the LORD, ואל to אלהי my God: אקרא and cried וישׁמע and he did hear מהיכלו out of his temple, קולי my voice ושׁועתי and my cry באזניו׃ into his ears.}%
\verse{ותגעשׁ shook ותרעשׁ and trembled; הארץ Then the earth מוסדות the foundations השׁמים of heaven ירגזו moved ויתגעשׁו and shook, כי because חרה׃ he was wroth.}%
\verse{עלה There went up עשׁן a smoke באפו out of his nostrils, ואשׁ and fire מפיו out of his mouth תאכל devoured: גחלים coals בערו were kindled ממנו׃ out of his mouth}%
\verse{ויט He bowed שׁמים the heavens וירד also, and came down; וערפל and darkness תחת under רגליו׃ his feet.}%
\verse{וירכב And he rode על upon כרוב a cherub, ויעף and did fly: וירא and he was seen על upon כנפי the wings רוח׃ of the wind.}%
\verse{וישׁת And he made חשׁך darkness סביבתיו round about סכות pavilions חשׁרת him, dark מים waters, עבי thick clouds שׁחקים׃ of the skies.}%
\verse{מנגה נגדו before בערו kindled. גחלי him were coals אשׁ׃ of fire}%
\verse{ירעם thundered מן from שׁמים heaven, יהוה ועליון and the most High יתן uttered קולו׃ his voice.}%
\verse{וישׁלח And he sent out חצים arrows, ויפיצם and scattered ברק them; lightning, ויהמם׃ and discomfited}%
\verse{ויראו appeared, אפקי And the channels ים of the sea יגלו were discovered, מסדות the foundations תבל of the world בגערת at the rebuking יהוה of the LORD, מנשׁמת at the blast רוח of the breath אפו׃ of his nostrils.}%
\verse{ישׁלח He sent ממרום from above, יקחני he took ימשׁני me; he drew ממים רבים׃}%
\verse{יצילני He delivered מאיבי עזשׂנאי from them that hated כי me: for אמצו they were too strong ממני׃ me from my strong enemy,}%
\verse{יקדמני They prevented ביום me in the day אידי of my calamity: ויהי was יהוה but the LORD משׁען׃ my stay.}%
\verse{ויצא He brought me forth למרחב אתיחלצני he delivered כי me, because חפץ׃ he delighted}%
\verse{יגמלני rewarded יהוה The LORD כצדקתי me according to my righteousness: כבר according to the cleanness ידי of my hands ישׁיב׃ hath he recompensed}%
\verse{כי For שׁמרתי I have kept דרכי the ways יהוה of the LORD, ולא and have not רשׁעתי wickedly מאלהי׃}%
\verse{כי For כל all משׁפטו his judgments לנגדי before וחקתיו me: and his statutes, לא I did not אסור depart ממנה׃ from}%
\verse{ואהיה I was תמים also upright לו ואשׁתמרה before him and have kept myself מעוני׃ from mine iniquity.}%
\verse{וישׁב hath recompensed יהוה Therefore the LORD לי כצדקתי me according to my righteousness; כברי according to my cleanness לנגד in עיניו׃ his eye sight.}%
\verse{עם With חסיד the merciful תתחסד thou wilt show thyself merciful, עם with גבור man תמים the upright תתמם׃ thou wilt show thyself upright.}%
\verse{עם With נבר thou wilt show thyself pure; תתבר ועם and with עקשׁ the froward תתפל׃ thou wilt show thyself unsavory.}%
\verse{ואת עם people עני And the afflicted תושׁיע thou wilt save: ועיניך but thine eyes על upon רמים the haughty, תשׁפיל׃ thou mayest bring down.}%
\verse{כי For אתה thou נירי my lamp, יהוה O LORD: ויהוה and the LORD יגיה will lighten חשׁכי׃ my darkness.}%
\verse{כי For בכה ארוץ by thee I have run גדוד through a troop: באלהי by my God אדלג have I leaped שׁור׃ over a wall.}%
\verse{האל God, תמים perfect; דרכו his way אמרת the word יהוה of the LORD צרופה tried: מגן a buckler הוא he לכל to all החסים׃ them that trust}%
\verse{כי For מי who אל God, מבלעדי save יהוה the LORD? ומי and who צור a rock, מבלעדי save אלהינו׃ our God?}%
\verse{האל God מעוזי my strength חיל power: ויתר and he maketh תמים perfect. דרכו׃ my way}%
\verse{משׁוה He maketh רגליו my feet כאילות like hinds' ועל me upon במותי my high places. יעמדני׃ and setteth}%
\verse{מלמד He teacheth ידי my hands למלחמה to war; ונחת is broken קשׁת so that a bow נחושׁה of steel זרעתי׃ by mine arms.}%
\verse{ותתן Thou hast also given לי מגן me the shield ישׁעך of thy salvation: וענתך תרבני׃ hath made me great.}%
\verse{תרחיב Thou hast enlarged צעדי my steps תחתני under ולא did not מעדו slip. קרסלי׃ me; so that my feet}%
\verse{ארדפה I have pursued איבי mine enemies, ואשׁמידם and destroyed ולא them; and turned not again אשׁוב them; and turned not again עד until כלותם׃ I had consumed}%
\verse{ואכלם ואמחצם them, and wounded ולא them, that they could not יקומון arise: ויפלו yea, they are fallen תחת under רגלי׃ my feet.}%
\verse{ותזרני For thou hast girded חיל me with strength למלחמה to battle: תכריע against me hast thou subdued קמי them that rose up תחתני׃ under}%
\verse{ואיבי of mine enemies, תתה Thou hast also given לי ערף me the necks משׂנאי them that hate ואצמיתם׃ that I might destroy}%
\verse{ישׁעו They looked, ואין but none משׁיע to save; אל unto יהוה the LORD, ולא them not. ענם׃ but he answered}%
\verse{ואשׁחקם Then did I beat כעפר them as small as the dust ארץ of the earth, כטיט them as the mire חוצות of the street, אדקם I did stamp ארקעם׃ did spread them abroad.}%
\verse{ותפלטני Thou also hast delivered מריבי me from the strivings עמי of my people, תשׁמרני thou hast kept לראשׁ me head גוים of the heathen: עם a people לא not ידעתי I knew יעבדני׃ shall serve}%
\verse{בני נכרתכחשׁו shall submit themselves לי לשׁמוע unto me: as soon as they hear, אזן unto me: as soon as they hear, ישׁמעו׃ they shall be obedient}%
\verse{בני נכרבלו shall fade away, ויחגרו and they shall be afraid ממסגרותם׃ out of their close places.}%
\verse{חי liveth; יהוה The LORD וברוך and blessed צורי my rock; וירם and exalted אלהי be the God צור of the rock ישׁעי׃ of my salvation.}%
\verse{האל It God הנתן that avengeth נקמת that avengeth לי ומוריד עמים the people תחתני׃ under}%
\verse{ומוציאי And that bringeth me forth מאיבי from mine enemies: ומקמי above them that rose up תרוממני thou also hast lifted me up on high מאישׁ חמסיםצילני׃ against me: thou hast delivered}%
\verse{על כןודך I will give thanks יהוה unto thee, O LORD, בגוים among the heathen, ולשׁמך unto thy name. אזמר׃ and I will sing praises}%
\verse{מגדיל ישׁועות of salvation מלכו for his king: ועשׂה and showeth חסד mercy למשׁיחו to his anointed, לדוד unto David, ולזרעו and to his seed עד forevermore. עולם׃ forevermore.}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 23
\verseWithHeading{David Extols Adonai}{ואלה Now these דברי words דוד of David האחרנים the last נאם said, דוד בן the son ישׁי of Jesse ונאם הגבר and the man הקם raised up על on high, משׁיח the anointed אלהי of the God יעקב of Jacob, ונעים and the sweet זמרות psalmist ישׂראל׃ of Israel,}%
\verse{רוח The Spirit יהוה of the LORD דבר spoke בי ומלתו by me, and his word על in לשׁוני׃ my tongue.}%
\verse{אמר said, אלהי The God ישׂראל of Israel לי דבר spoke צור the Rock ישׂראל of Israel מושׁל to me, He that ruleth over באדם men צדיק just, מושׁל ruling יראת in the fear אלהים׃ of God.}%
\verse{וכאור And as the light בקר of the morning, יזרח riseth, שׁמשׁ the sun בקר a morning לא without עבות clouds; מנגה by clear shining ממטר after rain. דשׁא the tender grass מארץ׃ out of the earth}%
\verse{כי Although לא not כן so ביתי my house עם with אל God; כי yet ברית covenant, עולם with me an everlasting שׂם he hath made לי ערוכה ordered בכל in all ושׁמרה and sure: כי for כל all ישׁעי my salvation, וכל and all חפץ desire, כי although לא he make not יצמיח׃ to grow.}%
\verse{ובליעל But of Belial כקוץ of them as thorns מנד thrust away, כלהם all כי because לא they cannot ביד with hands: יקחו׃ be taken}%
\verse{ואישׁ But the man יגע shall touch בהם ימלא them must be fenced ברזל with iron ועץ and the staff חנית of a spear; ובאשׁ with fire שׂרוף and they shall be utterly burned ישׂרפו and they shall be utterly burned בשׁבת׃ in the place.}%
\verseWithHeading{David’s Faithful Soldiers}{אלה These שׁמות the names הגברים of the mighty men אשׁר whom לדוד David ישׁב that sat בשׁבת in the seat, תחכמני had: The Tachmonite ראשׁ chief השׁלשׁי among the captains; הוא the same עדינו Adino העצנו the Eznite: על against שׁמנה eight מאות hundred, חלל whom he slew בפעם time. אחד׃ at one}%
\verse{ואחרו And after אלעזר him Eleazar בן the son דדי of Dodo בן אחחישׁלשׁה of the three גברים mighty men עם with דוד David, בחרפם when they defied בפלשׁתים the Philistines נאספו gathered together שׁם were there למלחמה to battle, ויעלו were gone away: אישׁ and the men ישׂראל׃ of Israel}%
\verse{הוא He קם arose, ויך and smote בפלשׁתים the Philistines עד until כי until יגעה was weary, ידו his hand ותדבק cleaved ידו and his hand אל unto החרב the sword: ויעשׂ wrought יהוה and the LORD תשׁועה victory גדולה a great ביום day; ההוא that והעם and the people ישׁבו returned אחריו after אך him only לפשׁט׃ to spoil.}%
\verse{ואחריו And after שׁמא בן the son אגא of Agee הררי the Hararite. ויאספו were gathered together פלשׁתים And the Philistines לחיה into a troop, ותהי was שׁם where חלקת a piece השׂדה of ground מלאה עדשׁים of lentils: והעם and the people נס fled מפני from פלשׁתים׃ the Philistines.}%
\verse{ויתיצב But he stood בתוך in the midst החלקה of the ground, ויצילה and defended ויך it, and slew את פלשׁתים the Philistines: ויעשׂ wrought יהוה and the LORD תשׁועה victory. גדולה׃ a great}%
\verse{וירדו went down, שׁלשׁים of the thirty מהשׁלשׁים ראשׁ chief ויבאו and came אל to קציר the harvest time אל in דוד David אל unto מערת the cave עדלם of Adullam: וחית and the troop פלשׁתים of the Philistines חנה pitched בעמק in the valley רפאים׃ of Rephaim.}%
\verse{ודוד And David אז then במצודה in a hold, ומצב and the garrison פלשׁתים of the Philistines אז then בית לחם׃ Bethlehem.}%
\verse{ויתאוה longed, דוד And David ויאמר and said, מי Oh that one ישׁקני would give me drink מים of the water מבאר בית לחם of Bethlehem, אשׁר which בשׁער׃ by the gate!}%
\verse{ויבקעו broke through שׁלשׁת And the three הגברים mighty men במחנה the host פלשׁתים of the Philistines, וישׁאבו and drew מים water מבאר בית לחם of Bethlehem, אשׁר that בשׁער by the gate, וישׂאו and took ויבאו and brought אל to דוד David: ולא not אבה nevertheless he would לשׁתותם drink ויסך אתםיהוה׃ unto the LORD.}%
\verse{ויאמר And he said, חלילה Be it far לי יהוה from me, O LORD, מעשׂתי things did זאת this: הדם the blood האנשׁים ההלכים that went בנפשׁותם in jeopardy of their lives? ולא not אבה therefore he would לשׁתותם drink אלה it. These עשׂו שׁלשׁת these three הגברים׃ mighty men.}%
\verse{ואבישׁי And Abishai, אחי the brother יואב of Joab, בן the son צרויה of Zeruiah, הוא And he ראשׁ was chief השׁלשׁי among three. והוא עורר lifted up את חניתו his spear על against שׁלשׁ three מאות hundred, חלל slew ולו שׁם and had the name בשׁלשׁה׃ among three.}%
\verse{מן most השׁלשׁה of three? הכי Was he not נכבד honorable ויהי therefore he was להם לשׂר their captain: ועד unto השׁלשׁה the three. לא not בא׃ howbeit he attained}%
\verse{ובניהו And Benaiah בן the son יהוידע of Jehoiada, בן the son אישׁ man, חי רב who had done many acts, פעלים who had done many acts, מקבצאל הוא he הכה slew את שׁני two אראל lionlike men מואב of Moab: והוא he ירד went down והכה also and slew את האריה a lion בתוך in the midst הבאר of a pit ביום in time השׁלג׃ of snow:}%
\verse{והוא And he הכה slew את אישׁ man: מצרי and the Egyptian אשׁר מראה a goodly וביד in his hand; המצרי חנית had a spear וירד but he went down אליו to בשׁבט him with a staff, ויגזל and plucked את החנית the spear מיד המצרייהרגהו בחניתו׃ him with his own spear.}%
\verse{אלה These עשׂה did בניהו Benaiah בן the son יהוידע of Jehoiada, ולו שׁם and had the name בשׁלשׁה among three הגברים׃ mighty men.}%
\verse{מן than השׁלשׁים the thirty, נכבד He was more honorable ואל to השׁלשׁה the three. לא not בא but he attained וישׂמהו set דוד And David אל him over משׁמעתו׃ his guard.}%
\verseWithHeading{The Mighty Men of David}{עשׂה אל Asahel אחי the brother יואב of Joab בשׁלשׁים one of the thirty; אלחנן Elhanan בן the son דדו of Dodo בית לחם׃ of Bethlehem,}%
\verse{שׁמה Shammah החרדי the Harodite, אליקא Elika החרדי׃ the Harodite,}%
\verse{חלץ Helez הפלטי the Paltite, עירא Ira בן the son עקשׁ of Ikkesh התקועי׃ the Tekoite,}%
\verse{אביעזר Abiezer הענתתי the Anethothite, מבני Mebunnai החשׁתי׃ the Hushathite,}%
\verse{צלמון Zalmon האחחי the Ahohite, מהרי Maharai הנטפתי׃ the Netophathite,}%
\verse{חלב Heleb בן the son בענה of Baanah, הנטפתי a Netophathite, אתי Ittai בן the son ריבי of Ribai מגבעת בני of the children בנימן׃ of Benjamin,}%
\verse{בניהו Benaiah פרעתני the Pirathonite, הדי Hiddai מנחלי of the brooks געשׁ׃ of Gaash,}%
\verse{אבי עלבון Abi-albon הערבתי the Arbathite, עזמות Azmaveth הברחמי׃ the Barhumite,}%
\verse{אליחבא Eliahba השׁעלבני the Shaalbonite, בני of the sons ישׁן of Jashen, יהונתן׃ Jonathan,}%
\verse{שׁמה Shammah ההררי the Hararite, אחיאם Ahiam בן the son שׁרר of Sharar האררי׃ the Hararite,}%
\verse{אליפלט Eliphelet בן the son אחסבי of Ahasbai, בן the son המעכתי of the Maachathite, אליעם Eliam בן the son אחיתפל of Ahithophel הגלני׃ the Gilonite,}%
\verse{חצרו Hezrai הכרמלי the Carmelite, פערי Paarai הארבי׃ the Arbite,}%
\verse{יגאל Igal בן the son נתן of Nathan מצבה בני Bani הגדי׃ the Gadite,}%
\verse{צלק Zelek העמני the Ammonite, נחרי Naharai הבארתי the Beerothite, נשׂאי armorbearer כלי armorbearer יואב to Joab בן the son צריה׃ of Zeruiah,}%
\verse{עירא Ira היתרי an Ithrite, גרב Gareb היתרי׃ an Ithrite,}%
\verse{אוריה Uriah החתי the Hittite: כל in all. שׁלשׁים thirty ושׁבעה׃ and seven}%
\end{biblechapter}%
\begin{biblechapter}% 2 Samuel 24
\verseWithHeading{David and the Census of the People}{ויסף And again אף the anger יהוה of the LORD לחרות was kindled בישׂראל against Israel, ויסת and he moved את דוד David בהם לאמר against them to say, לך Go, מנה number את ישׂראל Israel ואת יהודה׃ and Judah.}%
\verse{ויאמר said המלך For the king אל to יואב Joab שׂר the captain החיל of the host, אשׁר which אתו with שׁוט him, Go נא now בכל through all שׁבטי the tribes ישׂראל of Israel, מדן ועד even to באר שׁבע Beer-sheba, ופקדו and number את העם ye the people, וידעתי that I may know את מספר the number העם׃ of the people.}%
\verse{ויאמר said יואב And Joab אל unto המלך the king, ויוסף add יהוה Now the LORD אלהיך thy God אל unto העם the people, כהם how many soever they be וכהם how many soever they be מאה a hundredfold, פעמים a hundredfold, ועיני and that the eyes אדני of my lord המלך the king ראות may see ואדני doth my lord המלך the king למה but why חפץ delight בדבר thing? הזה׃ in this}%
\verse{ויחזק prevailed דבר word המלך Notwithstanding the king's אל against יואב Joab, ועל and against שׂרי the captains החיל of the host. ויצא went out יואב And Joab ושׂרי and the captains החיל of the host לפני from the presence המלך of the king, לפקד to number את העם the people את ישׂראל׃ of Israel.}%
\verse{ויעברו And they passed over את הירדן Jordan, ויחנו and pitched בערוער in Aroer, ימין on the right side העיר of the city אשׁר that בתוך in the midst הנחל of the river הגד of Gad, ואל and toward יעזר׃ Jazer:}%
\verse{ויבאו Then they came הגלעדה to Gilead, ואל and to ארץ the land תחתים חדשׁי of Tahtim-hodshi; ויבאו and they came דנה יען to Dan-jaan, וסביב and about אל to צידון׃ Zidon,}%
\verse{ויבאו And came מבצר to the stronghold צר of Tyre, וכל and to all ערי the cities החוי of the Hivites, והכנעני and of the Canaanites: ויצאו and they went out אל נגב the south יהודה Judah, באר שׁבע׃ to Beer-sheba.}%
\verse{וישׁטו So when they had gone בכל through all הארץ the land, ויבאו they came מקצה at the end תשׁעה of nine חדשׁים months ועשׂרים and twenty יום days. ירושׁלם׃ to Jerusalem}%
\verse{ויתן gave up יואב And Joab את מספר the sum מפקד of the number העם of the people אל unto המלך the king: ותהי and there were ישׂראל in Israel שׁמנה eight מאות hundred אלף thousand אישׁ men חיל valiant שׁלף that drew חרב the sword; ואישׁ and the men יהודה of Judah חמשׁ five מאות hundred אלף thousand אישׁ׃ men.}%
\verse{ויך smote לב heart דוד And David's אתו אחרי him after כן that ספר he had numbered את העם the people. ויאמר said דוד And David אל unto יהוה the LORD, חטאתי I have sinned מאד greatly אשׁר in that עשׂיתי I have done: ועתה and now, יהוה O LORD, העבר take away נא I beseech thee, את עון the iniquity עבדך of thy servant; כי for נסכלתי foolishly. מאד׃ I have done very}%
\verse{ויקם was up דוד For when David בבקר in the morning, ודבר the word יהוה of the LORD היה came אל unto גד Gad, הנביא the prophet חזה seer, דוד David's לאמר׃ saying,}%
\verse{הלוך Go ודברת saith אל unto דוד David, כה Thus אמר and say יהוה the LORD, שׁלשׁ thee three אנכי I נוטל offer עליך offer בחר choose לך אחת thee one מהם ואעשׂה׃ them, that I may}%
\verse{ויבא came גד So Gad אל to דוד David, ויגד and told לו ויאמר him, and said לו התבוא come לך שׁבע unto him, Shall seven שׁנים years רעב of famine בארצך unto thee in thy land? אם or שׁלשׁה three חדשׁים months נסך wilt thou flee לפני before צריך thine enemies, והוא while they רדפך pursue ואם thee? or היות that there be שׁלשׁת three ימים days' דבר pestilence בארצך in thy land? עתה now דע advise, וראה and see מה what אשׁיב I shall return שׁלחי to him that sent דבר׃ answer}%
\verse{ויאמר said דוד And David אל unto גד Gad, צר strait: לי מאד I am in a great נפלה let us fall נא now ביד into the hand יהוה of the LORD; כי for רבים great: רחמו his mercies וביד into the hand אדם of man. אל and let me not אפלה׃ fall}%
\verse{ויתן sent יהוה So the LORD דבר a pestilence בישׂראל upon Israel מהבקר from the morning ועד even to עת the time מועד appointed: וימת and there died מן of העם the people מדן ועד even to באר שׁבע Beer-sheba שׁבעים seventy אלף thousand אישׁ׃ men.}%
\verse{וישׁלח stretched out ידו his hand המלאך And when the angel ירושׁלם upon Jerusalem לשׁחתה to destroy וינחם repented him יהוה it, the LORD אל of הרעה the evil, ויאמר and said למלאך to the angel המשׁחית that destroyed בעם the people, רב It is enough: עתה now הרף stay ידך thine hand. ומלאך And the angel יהוה of the LORD היה was עם by גרן the threshingplace האורנה of Araunah היבסי׃ the Jebusite.}%
\verse{ויאמר spoke דוד And David אל unto יהוה the LORD בראתו when he saw את המלאך the angel המכה that smote בעם the people, ויאמר and said, הנה Lo, אנכי I חטאתי have sinned, ואנכי and I העויתי have done wickedly: ואלה but these הצאן sheep, מה what עשׂו have they done? תהי be נא I pray thee, ידך let thine hand, בי ובבית house. אבי׃ against me, and against my father's}%
\verse{ויבא came גד And Gad אל to דוד David, ביום day ההוא that ויאמר and said לו עלה unto him, Go up, הקם rear ליהוה unto the LORD מזבח an altar בגרן in the threshingfloor ארניה of Araunah היבסי׃ the Jebusite.}%
\verse{ויעל went up דוד And David, כדבר according to the saying גד of Gad, כאשׁר as צוה commanded. יהוה׃ the LORD}%
\verse{וישׁקף looked, ארונה And Araunah וירא and saw את המלך the king ואת עבדיו and his servants עברים coming on עליו toward ויצא went out, ארונה him: and Araunah וישׁתחו and bowed himself למלך before the king אפיו on his face ארצה׃ upon the ground.}%
\verse{ויאמר said, ארונה And Araunah מדוע Wherefore בא come אדני is my lord המלך the king אל to עבדו his servant? ויאמר said, דוד And David לקנות To buy מעמך of את הגרן the threshingfloor לבנות thee, to build מזבח an altar ליהוה unto the LORD, ותעצר may be stayed המגפה that the plague מעל from העם׃ the people.}%
\verse{ויאמר said ארונה And Araunah אל unto דוד David, יקח take ויעל and offer up אדני Let my lord המלך the king הטוב what good בעינו unto him: ראה behold, הבקר oxen לעלה for burnt sacrifice, והמרגים and threshing instruments וכלי and instruments הבקר of the oxen לעצים׃ for wood.}%
\verse{הכל All נתן give ארונה these did Araunah, המלך a king, למלך the king. ויאמר said ארונה And Araunah אל unto המלך unto the king, יהוה The LORD אלהיך thy God ירצך׃ accept}%
\verse{ויאמר said המלך And the king אל unto ארונה Araunah, לא Nay; כי but קנו bought אקנה מאותךמחיר thee at a price: ולא neither אעלה will I offer ליהוה unto the LORD אלהי my God עלות burnt offerings חנם doth cost me nothing. ויקן דוד So David את הגרן the threshingfloor ואת הבקר and the oxen בכסף of silver. שׁקלים shekels חמשׁים׃ for fifty}%
\verse{ויבן built שׁם there דוד And David מזבח an altar ליהוה unto the LORD, ויעל and offered עלות burnt offerings ושׁלמים and peace offerings. ויעתר was entreated יהוה So the LORD לארץ for the land, ותעצר was stayed המגפה and the plague מעל from ישׂראל׃ Israel.}%
\end{biblechapter}%
