\biblebook{Ecclesiastes}
\begin{biblechapter}% Ecclesiastes 1
\verseWithHeading{Prologue}{דברי The words קהלת of the Preacher, בן the son דוד of David, מלך king בירושׁלם׃ in Jerusalem.}%
\verseWithHeading{Motto Introduced}{הבל Vanity הבלים of vanities, אמר saith קהלת the Preacher, הבל vanity הבלים of vanities; הכל all הבל׃ vanity.}%
\verseWithHeading{All Toil is Profitless and Repetitious}{מה What יתרון profit לאדם hath a man בכל of all עמלו his labor שׁיעמל which he taketh תחת under השׁמשׁ׃ the sun?}%
\verse{דור generation הלך passeth away, ודור and generation בא cometh: והארץ but the earth לעולם forever. עמדת׃ abideth}%
\verse{וזרח also ariseth, השׁמשׁ The sun ובא goeth down, השׁמשׁ and the sun ואל to מקומו his place שׁואף and hasteth זורח arose. הוא he שׁם׃ where}%
\verse{הולך goeth אל toward דרום the south, וסובב and turneth about אל unto צפון the north; סובב continually, סבב continually, הולך it whirleth about הרוח The wind ועל according סביבתיו to his circuits. שׁב returneth again הרוח׃ and the wind}%
\verse{כל All הנחלים the rivers הלכים run אל into הים the sea; והים yet the sea איננו not מלא full; אל unto מקום the place שׁהנחלים from whence the rivers הלכים come, שׁם thither הם they שׁבים return ללכת׃ again.}%
\verse{כל All הדברים things יגעים full of labor; לא cannot יוכל cannot אישׁ man לדבר utter לא is not תשׂבע satisfied עין the eye לראות with seeing, ולא nor תמלא filled אזן the ear משׁמע׃ with hearing.}%
\verse{מה The thing שׁהיה that hath been, הוא it שׁיהיה which shall be; ומה and that שׁנעשׂה which is done הוא that שׁיעשׂה which shall be done: ואין and no כל and no חדשׁ new תחת under השׁמשׁ׃ the sun.}%
\verse{ישׁ Is there דבר thing שׁיאמר whereof it may be said, ראה See, זה this חדשׁ new? הוא it כבר already היה hath been לעלמים of old time, אשׁר which היה was מלפננו׃ before}%
\verse{אין no זכרון remembrance לראשׁנים of former וגם neither לאחרנים after. שׁיהיו shall there be לא ; neither יהיה of that are to come להם זכרון remembrance עם with שׁיהיו that shall come לאחרנה׃}%
\verseWithHeading{Qohelet Introduces His Quest}{אני I קהלת the Preacher הייתי was מלך king על over ישׂראל Israel בירושׁלם׃ in Jerusalem.}%
\verse{ונתתי And I gave את לבי my heart לדרושׁ to seek ולתור and search out בחכמה by wisdom על concerning כל all אשׁר that נעשׂה are done תחת under השׁמים heaven: הוא this ענין travail רע sore נתן given אלהים hath God לבני to the sons האדם of man לענות׃ to be exercised}%
\verse{ראיתי I have seen את כל all המעשׂים the works שׁנעשׂו that are done תחת under השׁמשׁ the sun; והנה and, behold, הכל all הבל vanity ורעות and vexation רוח׃ of spirit.}%
\verse{מעות crooked לא cannot יוכל cannot לתקן be made straight: וחסרון and that which is wanting לא cannot יוכל cannot להמנות׃ be numbered.}%
\verse{דברתי communed אני I עם with לבי mine own heart, לאמר saying, אני I הנה Lo, הגדלתי am come to great estate, והוספתי and have gotten more חכמה wisdom על than כל all אשׁר that היה have been לפני before על me in ירושׁלם Jerusalem: ולבי yea, my heart ראה experience הרבה had great חכמה of wisdom ודעת׃ and knowledge.}%
\verse{ואתנה And I gave לבי my heart לדעת to know חכמה wisdom, ודעת and to know הוללות madness ושׂכלות and folly: ידעתי I perceived שׁגם also זה that this הוא רעיון vexation רוח׃ of spirit.}%
\verse{כי For ברב in much חכמה wisdom רב כעס grief: ויוסיף and he that increaseth דעת knowledge יוסיף increaseth מכאוב׃ sorrow.}%
\end{biblechapter}%
\begin{biblechapter}% Ecclesiastes 2
\verseWithHeading{Qohelet’s Investigation of Self-Indulgence}{אמרתי said אני I בלבי in mine heart, לכה Go to נא now, אנסכה I will prove בשׂמחה thee with mirth, וראה therefore enjoy בטוב pleasure: והנה and, behold, גם also הוא this הבל׃ vanity.}%
\verse{לשׂחוק of laughter, אמרתי I said מהולל mad: ולשׂמחה and of mirth, מה What זה it? עשׂה׃ doeth}%
\verse{תרתי I sought בלבי in mine heart למשׁוך to give ביין unto wine, את בשׂרי myself ולבי mine heart נהג yet acquainting בחכמה with wisdom; ולאחז and to lay hold בסכלות on folly, עד till אשׁר till אראה I might see אי what זה that טוב good לבני for the sons האדם of men, אשׁר which יעשׂו they should do תחת under השׁמים the heaven מספר all ימי the days חייהם׃ of their life.}%
\verseWithHeading{Qohelet’s Investigation of Personal Accomplishment}{הגדלתי I made me great מעשׂי works; בניתי I built לי בתים me houses; נטעתי I planted לי כרמים׃ me vineyards:}%
\verse{עשׂיתי I made לי גנות me gardens ופרדסים and orchards, ונטעתי and I planted בהם עץ trees כל in them of all פרי׃ fruits:}%
\verse{עשׂיתי I made לי ברכות me pools מים of water, להשׁקות to water מהם יער the wood צומח that bringeth forth עצים׃ trees:}%
\verse{קניתי I got עבדים servants ושׁפחות and maidens, ובני servants born בית in my house; היה and had לי גם also מקנה possessions בקר of great וצאן and small cattle הרבה great היה I had לי מכל above all שׁהיו that were לפני before בירושׁלם׃ in Jerusalem}%
\verse{כנסתי I gathered לי גם me also כסף silver וזהב and gold, וסגלת and the peculiar treasure מלכים of kings והמדינות and of the provinces: עשׂיתי I got לי שׁרים me men singers ושׁרות and women singers, ותענוגת and the delights בני of the sons האדם of men, שׁדה musical instruments, ושׁדות׃ musical instruments,}%
\verse{וגדלתי So I was great, והוספתי and increased more מכל than all שׁהיה that were לפני before בירושׁלם me in Jerusalem: אף also חכמתי my wisdom עמדה׃ remained}%
\verse{וכל from any אשׁר שׁאלו desired עיני mine eyes לא not אצלתי I kept מהם לא not מנעתי them, I withheld את לבי my heart מכל in all שׂמחה joy; כי for לבי my heart שׂמח מכל of all עמלי my labor: וזה and this היה was חלקי my portion מכל עמלי׃ my labor.}%
\verse{ופניתי looked אני Then I בכל on all מעשׂי the works שׁעשׂו to do: ידי ובעמל and on the labor שׁעמלתי לעשׂותהנה and, behold, הכל all הבל vanity ורעות and vexation רוח of spirit, ואין and no יתרון profit תחת under השׁמשׁ׃ the sun.}%
\verseWithHeading{The Living Must Abandon the Work of their Hands to Others at Death}{ופניתי turned אני And I לראות myself to behold חכמה wisdom, והוללות and madness, וסכלות and folly: כי for מה what האדם the man שׁיבוא that cometh אחרי after המלך the king? את אשׁר that which כבר hath been already עשׂוהו׃ done.}%
\verse{וראיתי saw אני Then I שׁישׁ that יתרון excelleth לחכמה wisdom מן folly, הסכלות folly, כיתרון excelleth האור as far as light מן darkness. החשׁך׃ darkness.}%
\verse{החכם The wise man's עיניו eyes בראשׁו in his head; והכסיל but the fool בחשׁך in darkness: הולך walketh וידעתי perceived גם also אני and I myself שׁמקרה אחדקרה happeneth את כלם׃ them all.}%
\verse{ואמרתי Then said אני I בלבי in my heart, כמקרה As it happeneth הכסיל to the fool, גם even אני was I יקרני it happeneth ולמה to me; and why חכמתי wise? אני אז then יותר more ודברתי Then I said בלבי in my heart, שׁגם also זה that this הבל׃ vanity.}%
\verse{כי For אין no זכרון remembrance לחכם of the wise עם more than הכסיל of the fool לעולם forever; בשׁכבר seeing that which now הימים in the days הבאים to come הכל shall all נשׁכח be forgotten. ואיך And how ימות dieth החכם the wise עם as הכסיל׃ the fool.}%
\verse{ושׂנאתי Therefore I hated את החיים life; כי because רע grievous עלי unto המעשׂה the work שׁנעשׂה that is wrought תחת under השׁמשׁ the sun כי me: for הכל all הבל vanity ורעות and vexation רוח׃ of spirit.}%
\verse{ושׂנאתי hated אני Yea, I את כל all עמלי my labor שׁאני עמל had taken תחת under השׁמשׁ the sun: שׁאניחנו לאדם it unto the man שׁיהיה that shall be אחרי׃ after}%
\verse{ומי And who יודע knoweth החכם a wise יהיה whether he shall be או or סכל a fool? וישׁלט yet shall he have rule בכל over all עמלי my labor שׁעמלתי ושׁחכמתיחת under השׁמשׁ the sun. גם also זה This הבל׃ vanity.}%
\verse{וסבותי went about אני Therefore I ליאשׁ to despair את לבי to cause my heart על of כל all העמל the labor שׁעמלתי תחת under השׁמשׁ׃ the sun.}%
\verse{כי For ישׁ there is אדם a man שׁעמלו whose labor בחכמה in wisdom, ובדעת and in knowledge, ובכשׁרון and in equity; ולאדם yet to a man שׁלא that hath not עמל labored בו יתננו therein shall he leave חלקו it his portion. גם also זה This הבל vanity ורעה evil. רבה׃ and a great}%
\verse{כי For מה what הוה hath לאדם man בכל of all עמלו his labor, וברעיון and of the vexation לבו of his heart, שׁהוא wherein he עמל hath labored תחת under השׁמשׁ׃ the sun?}%
\verse{כי For כל all ימיו his days מכאבים sorrows, וכעס grief; ענינו and his travail גם yea, בלילה in the night. לא taketh not שׁכב rest לבו his heart גם is also זה This הבל vanity. הוא׃}%
\verseWithHeading{It is Best to Simply Enjoy the Passing Pleasures of Life as Reward for Pleasing God}{אין nothing טוב better באדם for a man, שׁיאכל that he should eat ושׁתה and drink, והראה אתפשׁו טוב good בעמלו in his labor. גם also זה This ראיתי saw, אני I כי that מיד from the hand האלהים of God. היא׃ it}%
\verse{כי For מי who יאכל can eat, ומי or who יחושׁ else can hasten חוץ more ממני׃ than}%
\verse{כי For לאדם to a man שׁטוב that good לפניו in his sight נתן giveth חכמה wisdom, ודעת and knowledge, ושׂמחה and joy: ולחוטא but to the sinner נתן he giveth ענין travail, לאסוף to gather ולכנוס and to heap up, לתת that he may give לטוב to good לפני before האלהים God. גם also זה This הבל vanity ורעות and vexation רוח׃ of spirit.}%
\end{biblechapter}%
\begin{biblechapter}% Ecclesiastes 3
\verseWithHeading{God Has Ordained the Ebb and Flow of Human Activities}{לכל To every זמן a season, ועת and a time לכל to every חפץ purpose תחת under השׁמים׃ the heaven:}%
\verse{עת A time ללדת to be born, ועת and a time למות to die; עת a time לטעת to plant, ועת and a time לעקור to pluck up נטוע׃ planted;}%
\verse{עת A time להרוג to kill, ועת and a time לרפוא to heal; עת a time לפרוץ to break down, ועת and a time לבנות׃ to build up;}%
\verse{עת A time לבכות to weep, ועת and a time לשׂחוק to laugh; עת a time ספוד to mourn, ועת and a time רקוד׃ to dance;}%
\verse{עת A time להשׁליך to cast away אבנים stones, ועת and a time כנוס to gather stones together; אבנים to gather stones together; עת a time לחבוק to embrace, ועת and a time לרחק to refrain מחבק׃ from embracing;}%
\verse{עת A time לבקשׁ to get, ועת and a time לאבד to lose; עת a time לשׁמור to keep, ועת and a time להשׁליך׃ to cast away;}%
\verse{עת A time לקרוע to rend, ועת and a time לתפור to sew; עת a time לחשׁות to keep silence, ועת and a time לדבר׃ to speak;}%
\verse{עת A time לאהב to love, ועת and a time לשׂנא to hate; עת a time מלחמה of war, ועת and a time שׁלום׃ of peace.}%
\verse{מה What יתרון profit העושׂה hath he that worketh באשׁר in that wherein הוא he עמל׃ laboreth?}%
\verseWithHeading{No One Understands God’s Mysterious Plan}{ראיתי I have seen את הענין the travail, אשׁר which נתן hath given אלהים God לבני to the sons האדם of men לענות׃ to be exercised}%
\verse{את הכל every עשׂה He hath made יפה beautiful בעתו in his time: גם also את העלם the world נתן he hath set בלבם in their heart, מבלי no אשׁר so that לא ימצא can find out האדם man את המעשׂה the work אשׁר that עשׂה maketh האלהים God מראשׁ from the beginning ועד to סוף׃ the end.}%
\verse{ידעתי I know כי that אין no טוב good בם כי for אם in them, but לשׂמוח to rejoice, ולעשׂות and to do טוב good בחייו׃ in his life.}%
\verse{וגם And also כל of all האדם שׁיאכלשׁתה and drink, וראה and enjoy טוב the good בכל עמלו his labor, מתת the gift אלהים of God. היא׃ it}%
\verse{ידעתי I know כי that, כל whatsoever אשׁר whatsoever יעשׂה doeth, האלהים God הוא it יהיה shall be לעולם forever: עליו to אין nothing להוסיף can be put וממנו from אין it, nor לגרע any thing taken והאלהים it: and God עשׂה doeth שׁיראו , that should fear מלפניו׃ before}%
\verse{מה שׁהיה is to be כבר is now; הוא and that ואשׁר which להיות been; כבר hath already היה והאלהים and God יבקשׁ requireth את נרדף׃ that which is past.}%
\verseWithHeading{God’s Mysterious Plan Allows Injustice to Exist in the World}{ועוד And moreover ראיתי I saw תחת under השׁמשׁ the sun מקום the place המשׁפט of judgment, שׁמה there; הרשׁע wickedness ומקום and the place הצדק of righteousness, שׁמה there. הרשׁע׃ iniquity}%
\verse{אמרתי said אני I בלבי in mine heart, את הצדיק the righteous ואת הרשׁע and the wicked: ישׁפט shall judge האלהים God כי for עת a time לכל for every חפץ purpose ועל and for כל every המעשׂה work. שׁם׃ there}%
\verse{אמרתי said אני I בלבי in mine heart על concerning דברת the estate בני of the sons האדם of men, לברם might manifest האלהים that God ולראות might see שׁהם them, and that they בהמה themselves are beasts. המה להם׃}%
\verse{כי For מקרה that which befalleth בני the sons האדם of men ומקרה befalleth הבהמה beasts; ומקרה befalleth אחד even one thing להם כמות dieth, זה them: as the one כן so מות dieth זה the other; ורוח breath; אחד one לכל yea, they have all ומותר preeminence האדם so that a man מן above הבהמה a beast: אין hath no כי for הכל all הבל׃ vanity.}%
\verse{הכל All הולך go אל unto מקום place; אחד one הכל all היה are מן of העפר the dust, והכל and all שׁב turn to dust again. אל העפר׃}%
\verse{מי Who יודע knoweth רוח the spirit בני האדם of man העלה goeth היא that למעלה upward, ורוח and the spirit הבהמה of the beast הירדת goeth היא that למטה downward לארץ׃ to the earth?}%
\verse{וראיתי Wherefore I perceive כי that אין nothing טוב better, מאשׁר than that ישׂמח should rejoice האדם a man במעשׂיו in his own works; כי for הוא that חלקו his portion: כי for מי who יביאנו shall bring לראות him to see במה what shall be שׁיהיה אחריו׃ after}%
\end{biblechapter}%
\begin{biblechapter}% Ecclesiastes 4
\verseWithHeading{The Existence of Oppression in the World Makes Human Existence Miserable}{ושׁבתי returned, אני So I ואראה and considered את כל all העשׁקים the oppressions אשׁר that נעשׂים are done תחת under השׁמשׁ the sun: והנה and behold דמעת the tears העשׁקים of oppressed, ואין and they had no להם מנחם comforter; ומיד and on the side עשׁקיהם of their oppressors כח power; ואין but they had no להם מנחם׃ comforter.}%
\verse{ושׁבח praised אני Wherefore I את המתים the dead שׁכבר which are already מתו dead מן more than החיים the living אשׁר which המה which חיים alive. עדנה׃ are yet}%
\verse{וטוב Yea, better משׁניהם than both את אשׁר they, which עדן yet לא hath not היה been, אשׁר who לא hath not ראה seen את המעשׂה work הרע the evil אשׁר that נעשׂה is done תחת under השׁמשׁ׃ the sun.}%
\verseWithHeading{People Need Balance in Their Approach to Labor}{וראיתי considered אני Again, I את כל all עמל travail, ואת כל and every כשׁרון right המעשׂה work, כי that היא for this קנאת is envied אישׁ a man מרעהו of his neighbor. גם also זה This הבל vanity ורעות and vexation רוח׃ of spirit.}%
\verse{הכסיל The fool חבק אתדיו ואכל and eateth את בשׂרו׃ his own flesh.}%
\verse{טוב Better מלא full כף נחת quietness, ממלא חפנים than both the hands עמל travail ורעות and vexation רוח׃ of spirit.}%
\verseWithHeading{Wealth without Someone with Which to Enjoy It is Futile}{ושׁבתי returned, אני Then I ואראה and I saw הבל vanity תחת under השׁמשׁ׃ the sun.}%
\verse{ישׁ There is אחד one ואין and not שׁני a second; גם yea, בן child ואח nor brother: אין he hath neither לו ואין yet no קץ end לכל of all עמלו his labor; גם also עיניו is his eye לא neither תשׂבע satisfied עשׁר with riches; ולמי neither For whom אני do I עמל labor, ומחסר and bereave את נפשׁי my soul מטובה of good? גם זה This הבל vanity, וענין travail. רע a sore הוא׃ yea, it}%
\verseWithHeading{Friends and Family Can Help One Another in Life}{טובים better השׁנים Two מן than האחד one; אשׁר because ישׁ they have להם שׂכר reward טוב a good בעמלם׃ for their labor.}%
\verse{כי For אם if יפלו they fall, האחד the one יקים will lift up את חברו his fellow: ואילו but woe האחד to him alone שׁיפול when he falleth; ואין for not שׁני another להקימו׃ to help him up.}%
\verse{גם Again, אם if ישׁכבו lie together, שׁנים two וחם then they have heat: להם ולאחד can one איך but how יחם׃ be warm}%
\verse{ואם And if יתקפו prevail against האחד one השׁנים him, two יעמדו shall withstand נגדו shall withstand והחוט cord המשׁלשׁ him; and a threefold לא is not במהרה quickly ינתק׃ broken.}%
\verseWithHeading{One Must Be Willing to Listen to Counsel}{טוב Better ילד child מסכן a poor וחכם and a wise ממלך king, זקן than an old וכסיל and foolish אשׁר who לא no ידע will להזהר be admonished. עוד׃ more}%
\verse{כי For מבית הסוריםצא he cometh למלך to reign; כי whereas גם also במלכותו in his kingdom נולד born רשׁ׃ becometh poor.}%
\verse{ראיתי I considered את כל all החיים the living המהלכים which walk תחת under השׁמשׁ the sun, עם with הילד child השׁני the second אשׁר that יעמד shall stand up תחתיו׃ in his stead.}%
\verse{אין no קץ end לכל of all העם the people, לכל of all אשׁר that היה have been לפניהם before גם them: they also האחרונים that come after לא shall not ישׂמחו rejoice בו כי in him. Surely גם also זה this הבל vanity ורעיון and vexation רוח׃ of spirit.}%
\end{biblechapter}%
\begin{biblechapter}% Ecclesiastes 5
\verseWithHeading{Listen to God Rather Than Uttering Rash Vows}{שׁמר Keep רגליך thy foot כאשׁר when תלך thou goest אל to בית the house האלהים of God, וקרוב לשׁמע to hear, מתת than to give הכסילים of fools: זבח the sacrifice כי that אינם not יודעים for they consider לעשׂות they do רע׃ evil.}%
\verse{אל Be not תבהל rash על with פיך thy mouth, ולבך thine heart אל and let not ימהר be hasty להוציא to utter דבר thing לפני before האלהים God: כי for האלהים God בשׁמים in heaven, ואתה and thou על upon הארץ earth: על therefore כן therefore יהיו be דבריך let thy words מעטים׃ few.}%
\verse{כי For בא cometh החלום a dream ברב through the multitude ענין of business; וקול voice כסיל and a fool's ברב by multitude דברים׃ of words.}%
\verse{כאשׁר When תדר thou vowest נדר a vow לאלהים unto God, אל not תאחר defer לשׁלמו to pay כי it; for אין no חפץ pleasure בכסילים in fools: את אשׁר that which תדר thou hast vowed. שׁלם׃ pay}%
\verse{טוב Better אשׁר that לא thou shouldest not תדר vow, משׁתדור ולא and not תשׁלם׃ pay.}%
\verse{אל not תתן Suffer את פיך thy mouth לחטיא to sin; את בשׂרך to cause thy flesh ואל neither תאמר say לפני thou before המלאך the angel, כי that שׁגגה an error: היא it למה wherefore יקצף be angry האלהים should God על at קולך thy voice, וחבל and destroy את מעשׂה the work ידיך׃ of thine hands?}%
\verse{כי For ברב in the multitude חלמות of dreams והבלים also vanities: ודברים words הרבה and many כי but את האלהים thou God. ירא׃ fear}%
\verseWithHeading{Powerful Bureaucrats Exploit the Helpless Poor}{אם If עשׁק the oppression רשׁ of the poor, וגזל and violent perverting משׁפט of judgment וצדק and justice תראה thou seest במדינה in a province, אל not תתמה marvel על at החפץ the matter: כי for גבה higher מעל than גבה the highest שׁמר regardeth; וגבהים and higher עליהם׃ than}%
\verse{ויתרון Moreover the profit ארץ of the earth בכל is for all: היא מלך the king לשׂדה by the field. נעבד׃ is served}%
\verseWithHeading{There is Never Enough Money to Satisfy}{אהב He that loveth כסף silver לא shall not ישׂבע be satisfied כסף with silver; ומי nor he אהב that loveth בהמון abundance לא תבואה with increase: גם also זה this הבל׃ vanity.}%
\verse{ברבות increase, הטובה When goods רבו אוכליה that eat ומה them: and what כשׁרון good לבעליה to the owners כי thereof, saving אם thereof, saving ראית the beholding עיניו׃ with their eyes?}%
\verse{מתוקה sweet, שׁנת The sleep העבד of a laboring man אם whether מעט little ואם or הרבה much: יאכל he eat והשׂבע but the abundance לעשׁיר of the rich איננו will not מניח לו לישׁון׃ him to sleep.}%
\verseWithHeading{Hoarding Wealth Can Backfire}{ישׁ There is רעה evil חולה a sore ראיתי I have seen תחת under השׁמשׁ the sun, עשׁר riches שׁמור kept לבעליו for the owners לרעתו׃ thereof to their hurt.}%
\verse{ואבד perish העשׁר riches ההוא But those בענין travail: רע by evil והוליד and he begetteth בן a son, ואין and nothing בידו in his hand. מאומה׃ and nothing}%
\verse{כאשׁר As יצא he came forth מבטן womb, אמו of his mother's ערום naked ישׁוב shall he return ללכת to go כשׁבא as he came, ומאומה nothing לא nothing ישׂא and shall take בעמלו of his labor, שׁילך which he may carry way בידו׃ in his hand.}%
\verse{וגם also זה And this רעה evil, חולה a sore כל in all עמת points שׁבא as he came, כן so ילך shall he go: ומה and what יתרון profit לו שׁיעמל hath he that hath labored לרוח׃ for the wind?}%
\verse{גם also כל All ימיו his days בחשׁך in darkness, יאכל he eateth וכעס הרבה and much וחליו with his sickness. וקצף׃ and wrath}%
\verseWithHeading{If You Have Wealth, Enjoy It as God Enables}{הנה Behold אשׁר which ראיתי have seen: אני I טוב good אשׁר which יפה and comely לאכול to eat ולשׁתות and to drink, ולראות and to enjoy טובה the good בכל of all עמלו his labor שׁיעמל that he taketh תחת under השׁמשׁ the sun מספר all ימי the days חיו of his life, אשׁר נתן giveth לו האלהים God כי him: for הוא it חלקו׃ his portion.}%
\verse{גם also כל Every האדם man אשׁר to whom נתן hath given לו האלהים God עשׁר riches ונכסים and wealth, והשׁליטו and hath given him power לאכל to eat ממנו thereof, ולשׂאת and to take את חלקו his portion, ולשׂמח and to rejoice בעמלו in his labor; זה this מתת the gift אלהים of God. היא׃}%
\verse{כי For לא he shall not הרבה much יזכר remember את ימי the days חייו of his life; כי because האלהים God מענה בשׂמחת in the joy לבו׃ of his heart.}%
\end{biblechapter}%
\begin{biblechapter}% Ecclesiastes 6
\verseWithHeading{Those Who Have Wealth but Do Not Enjoy It Are Pitiful}{ישׁ There is רעה an evil אשׁר which ראיתי I have seen תחת under השׁמשׁ the sun, ורבה common היא and it על among האדם׃ men:}%
\verse{אישׁ A man אשׁר to whom יתן hath given לו האלהים God עשׁר riches, ונכסים wealth, וכבוד and honor, ואיננו nothing חסר so that he wanteth לנפשׁו for his soul מכל of all אשׁר that יתאוה he desireth, ולא giveth him not power ישׁליטנו giveth him not power האלהים yet God לאכל to eat ממנו of all כי but אישׁ a stranger נכרי a stranger יאכלנו eateth זה it: this הבל vanity, וחלי disease. רע an evil הוא׃ and it}%
\verse{אם If יוליד beget אישׁ a man מאה a hundred ושׁנים years, רבות many יחיה and live ורב many, שׁיהיו be ימי so that the days שׁניו of his years ונפשׁו and his soul לא be not תשׂבע filled מן with הטובה good, וגם and also קבורה burial; לא no היתה he have לו אמרתי I say, טוב better ממנו than הנפל׃ an untimely birth}%
\verse{כי For בהבל in with vanity, בא he cometh ובחשׁך in darkness, ילך and departeth ובחשׁך with darkness. שׁמו and his name יכסה׃ shall be covered}%
\verse{גם Moreover שׁמשׁ the sun, לא he hath not ראה seen ולא nor ידע known נחת hath more rest לזה this מזה׃ than the other.}%
\verse{ואלו Yea, חיה though he live אלף a thousand שׁנים years פעמים twice וטובה good: לא no ראה yet hath he seen הלא do not אל to מקום place? אחד one הכל all הולך׃ go}%
\verseWithHeading{One Must Learn to Be Content with What One Has}{כל All עמל the labor האדם of man לפיהו for his mouth, וגם and yet הנפשׁ the appetite לא is not תמלא׃ filled.}%
\verse{כי For מה what יותר more לחכם hath the wise מן than הכסיל the fool? מה what לעני hath the poor, יודע that knoweth להלך to walk נגד before החיים׃ the living?}%
\verse{טוב Better מראה the sight עינים of the eyes מהלך than the wandering נפשׁ of the desire: גם also זה this הבל vanity ורעות and vexation רוח׃ of spirit.}%
\verseWithHeading{It is Futile for Humans to Complain about God’s Irresistible Will}{מה That שׁהיה which hath been כבר already, נקרא is named שׁמו is named ונודע and it is known אשׁר that הוא it אדם man: ולא neither יוכל may לדין he contend עם with שׁהתקיף ממנו׃ than}%
\verse{כי Seeing ישׁ there be דברים things הרבה many מרבים that increase הבל vanity, מה what יתר the better? לאדם׃ man}%
\verseWithHeading{The Future is Inscrutable to Humans}{כי For מי who יודע knoweth מה what טוב good לאדם for man בחיים in life, מספר all ימי the days חיי life הבלו of his vain ויעשׂם which he spendeth כצל as a shadow? אשׁר for מי who יגיד can tell לאדם a man מה what יהיה shall be אחריו after תחת him under השׁמשׁ׃ the sun?}%
\end{biblechapter}%
\begin{biblechapter}% Ecclesiastes 7
\verseWithHeading{People Generally Do Not Know What is Best for Them}{טוב better שׁם A good name משׁמן ointment; טוב than precious ויום and the day המות of death מיום than the day הולדו׃ of one's birth.}%
\verse{טוב better ללכת to go אל to בית the house אבל of mourning, מלכת than to go אל to בית the house משׁתה of feasting: באשׁר for הוא that סוף the end כל of all האדם men; והחי and the living יתן will lay אל to לבו׃ his heart.}%
\verse{טוב better כעס Sorrow משׂחק than laughter: כי for ברע by the sadness פנים of the countenance ייטב is made better. לב׃ the heart}%
\verse{לב The heart חכמים of the wise בבית in the house אבל of mourning; ולב but the heart כסילים of fools בבית in the house שׂמחה׃ of mirth.}%
\verse{טוב better לשׁמע to hear גערת the rebuke חכם of the wise, מאישׁ than for a man שׁמע to hear שׁיר the song כסילים׃ of fools.}%
\verse{כי For כקול as the crackling הסירים of thorns תחת under הסיר a pot, כן so שׂחק the laughter הכסיל of the fool: וגם also זה this הבל׃ vanity.}%
\verseWithHeading{Wisdom—Although Vulnerable—is Beneficial}{כי Surely העשׁק oppression יהולל mad; חכם maketh a wise man ויאבד destroyeth את לב the heart. מתנה׃ and a gift}%
\verse{טוב Better אחרית the end דבר of a thing מראשׁיתו than the beginning טוב better ארך thereof: the patient רוח in spirit מגבה than the proud רוח׃ in spirit.}%
\verse{אל Be not תבהל hasty ברוחך in thy spirit לכעוס to be angry: כי for כעס anger בחיק in the bosom כסילים of fools. ינוח׃ resteth}%
\verse{אל not תאמר Say מה thou, What היה is שׁהימים הראשׁניםיו were טובים better מאלה than these? כי for לא thou dost not מחכמה wisely שׁאלת inquire על concerning זה׃ this.}%
\verse{טובה good חכמה Wisdom עם with נחלה an inheritance: ויתר and profit לראי to them that see השׁמשׁ׃ the sun.}%
\verse{כי For בצל a defense, החכמה wisdom בצל a defense: הכסף money ויתרון but the excellency דעת of knowledge החכמה wisdom תחיה giveth life בעליה׃ to them that have}%
\verseWithHeading{Humans Must Accept God’s Will and Make the Best of It}{ראה Consider את מעשׂה the work האלהים of God: כי for מי who יוכל can לתקן make straight, את אשׁר which עותו׃ he hath made crooked?}%
\verse{ביום In the day טובה of prosperity היה be בטוב joyful, וביום but in the day רעה of adversity ראה consider: גם also את זה the one לעמת over against זה the other, עשׂה hath set האלהים God על to דברת the end שׁלא ימצא should find האדם that man אחריו after מאומה׃}%
\verseWithHeading{The Law of Retribution Does Not Always Work—but It Does Sometimes}{את הכל All ראיתי have I seen בימי in the days הבלי of my vanity: ישׁ there is צדיק a just אבד that perisheth בצדקו in his righteousness, וישׁ and there is רשׁע a wicked מאריך that prolongeth ברעתו׃ in his wickedness.}%
\verse{אל not תהי Be צדיק righteous הרבה over much; ואל neither תתחכם make thyself over wise: יותר make thyself over wise: למה why תשׁומם׃ shouldest thou destroy thyself?}%
\verse{אל Be not תרשׁע wicked, הרבה over much ואל neither תהי be סכל thou foolish: למה why תמות shouldest thou die בלא before עתך׃ thy time?}%
\verse{טוב good אשׁר that תאחז thou shouldest take hold בזה of this; וגם yea, also מזה from this אל not תנח אתדך thine hand: כי for ירא he that feareth אלהים God יצא shall come forth את כלם׃ of them all.}%
\verseWithHeading{Wisdom is Valuable, but No One is Completely Righteous}{החכמה Wisdom תעז strengtheneth לחכם the wise מעשׂרה more than ten שׁליטים mighty אשׁר which היו are בעיר׃ in the city.}%
\verse{כי For אדם man אין not צדיק a just בארץ upon earth, אשׁר that יעשׂה doeth טוב good, ולא not. יחטא׃ and sinneth}%
\verse{גם Also לכל unto all הדברים words אשׁר that ידברו are spoken; אל no תתן take לבך heed אשׁר lest לא lest תשׁמע thou hear את עבדך thy servant מקללך׃ curse}%
\verse{כי For גם also פעמים oftentimes רבות oftentimes ידע knoweth לבך thine own heart אשׁר that גם likewise את thou thyself קללת hast cursed אחרים׃ others.}%
\verseWithHeading{Absolute Wisdom is Unattainable}{כל All זה this נסיתי have I proved בחכמה by wisdom: אמרתי I said, אחכמה I will be wise; והיא but it רחוקה far ממני׃ from}%
\verse{רחוק far off, מה שׁהיהעמק and exceeding deep, עמק and exceeding deep, מי who ימצאנו׃ can find it out?}%
\verse{סבותי applied אני I ולבי mine heart לדעת to know, ולתור and to search, ובקשׁ and to seek out חכמה wisdom, וחשׁבון and the reason ולדעת רשׁע the wickedness כסל of folly, והסכלות even of foolishness הוללות׃ madness:}%
\verse{ומוצא find אני And I מר more bitter ממות than death את האשׁה the woman, אשׁר whose היא whose מצודים snares וחרמים and nets, לבה heart אסורים bands: ידיה her hands טוב whoso pleaseth לפני whoso pleaseth האלהים God ימלט shall escape ממנה than death וחוטא her; but the sinner ילכד׃ shall be taken}%
\verse{ראה Behold, זה this מצאתי have I found, אמרה saith קהלת the preacher, אחת one לאחת by one, למצא to find out חשׁבון׃ the account:}%
\verse{אשׁר Which עוד yet בקשׁה seeketh, נפשׁי my soul ולא not: מצאתי but I find אדם man אחד one מאלף among a thousand מצאתי have I found; ואשׁה but a woman בכל among all אלה those לא have I not מצאתי׃ found.}%
\verse{לבד only ראה Lo, זה this מצאתי have I found, אשׁר that עשׂה hath made האלהים God את האדם man ישׁר upright; והמה but they בקשׁו have sought out חשׁבנות inventions. רבים׃ many}%
\end{biblechapter}%
\begin{biblechapter}% Ecclesiastes 8
\verseWithHeading{Wisdom is Valuable}{מי Who כהחכם as the wise ומי and who יודע knoweth פשׁר the interpretation דבר of a thing? חכמת wisdom אדם a man's תאיר to shine, פניו maketh his face ועז and the boldness פניו of his face ישׁנא׃ shall be changed.}%
\verse{אני I פי commandment, מלך the king's שׁמור to keep ועל and in regard דברת and in regard שׁבועת of the oath אלהים׃ of God.}%
\verse{אל Be not תבהל hasty מפניו out of his sight: תלך to go אל not תעמד stand בדבר thing; רע in an evil כי for כל whatsoever אשׁר whatsoever יחפץ pleaseth יעשׂה׃ he doeth}%
\verse{באשׁר Where דבר the word מלך of a king שׁלטון power: ומי and who יאמר may say לו מה unto him, What תעשׂה׃ doest}%
\verse{שׁומר Whoso keepeth מצוה the commandment לא no ידע shall feel דבר thing: רע evil ועת both time ומשׁפט and judgment. ידע discerneth לב heart חכם׃ and a wise man's}%
\verse{כי Because לכל to every חפץ purpose ישׁ there is עת time ומשׁפט and judgment, כי therefore רעת the misery האדם of man רבה great עליו׃ upon}%
\verseWithHeading{No One Knows the Future}{כי For איננו not ידע he knoweth מה that שׁיהיה which shall be: כי for כאשׁר him when יהיה it shall be? מי who יגיד׃ can tell}%
\verse{אין no אדם man שׁליט that hath power ברוח over the spirit לכלוא to retain את הרוח the spirit; ואין neither שׁלטון power ביום in the day המות of death: ואין and no משׁלחת discharge במלחמה in war; ולא neither ימלט deliver רשׁע shall wickedness את בעליו׃ those that are given}%
\verseWithHeading{The World Marred by Oppression and Injustice}{את כל All זה this ראיתי have I seen, ונתון and applied את לבי my heart לכל unto every מעשׂה work אשׁר that נעשׂה is done תחת under השׁמשׁ the sun: עת a time אשׁר wherein שׁלט ruleth over האדם one man באדם another לרע׃ to his own hurt.}%
\verse{ובכן And so ראיתי I saw רשׁעים the wicked קברים buried, ובאו who had come וממקום from the place קדושׁ of the holy, יהלכו and gone וישׁתכחו and they were forgotten בעיר in the city אשׁר where כן they had so עשׂו done: גם also זה this הבל׃ vanity.}%
\verseWithHeading{Although Evil is Not Punished Swiftly, God Does Eventually Punish Sinners}{אשׁר Because אין is not נעשׂה executed פתגם sentence against מעשׂה work הרעה an evil מהרה speedily, על therefore כן therefore מלא is fully set לב the heart בני of the sons האדם of men בהם לעשׂות in them to do רע׃ evil.}%
\verse{אשׁר Though חטא a sinner עשׂה do רע evil מאת a hundred times, ומאריך and his be prolonged, לו כי yet גם surely יודע know אני I אשׁר that יהיה it shall be טוב well ליראי fear האלהים God, אשׁר which ייראו מלפניו׃ before}%
\verse{וטוב well לא But it shall not יהיה be לרשׁע with the wicked, ולא neither יאריך shall he prolong ימים days, כצל as a shadow; אשׁר because איננו not ירא מלפני before אלהים׃ God.}%
\verse{ישׁ There is הבל a vanity אשׁר which נעשׂה is done על upon הארץ the earth; אשׁר that ישׁ there be צדיקים just אשׁר unto whom מגיע it happeneth אלהם , unto whom כמעשׂה according to the work הרשׁעים of the wicked; וישׁ again, there be רשׁעים wicked שׁמגיע whom it happeneth אלהם to כמעשׂה according to the work הצדיקים of the righteous: אמרתי I said שׁגם זה that this also הבל׃ vanity.}%
\verseWithHeading{Humans Should Enjoy the Life That God Gives to Them}{ושׁבחתי commended אני Then I את השׂמחה mirth, אשׁר because אין hath no טוב better לאדם a man תחת thing under השׁמשׁ the sun, כי than אם than לאכול to eat, ולשׁתות and to drink, ולשׂמוח and to be merry: והוא for that ילונו shall abide בעמלו with him of his labor ימי the days חייו of his life, אשׁר which נתן giveth לו האלהים God תחת him under השׁמשׁ׃ the sun.}%
\verseWithHeading{No One Can Discover the Rhyme and Reason for Things}{כאשׁר When נתתי I applied את לבי mine heart לדעת to know חכמה wisdom, ולראות and to see את הענין the business אשׁר that נעשׂה is done על upon הארץ the earth: כי (for גם also ביום day ובלילה nor night שׁנה sleep בעיניו with his eyes:) איננו neither ראה׃ seeth}%
\verse{וראיתי Then I beheld את כל all מעשׂה the work האלהים of God, כי that לא cannot יוכל cannot האדם a man למצוא find out את המעשׂה the work אשׁר that נעשׂה is done תחת under השׁמשׁ the sun: בשׁל because אשׁר though יעמל labor האדם a man לבקשׁ to seek out, ולא yet he shall not ימצא find וגם yea אם further; though יאמר think החכם a wise לדעת to know לא yet shall he not יוכל be able למצא׃ to find}%
\end{biblechapter}%
\begin{biblechapter}% Ecclesiastes 9
\verseWithHeading{The Same Fate—Death—Awaits Everyone}{כי For את כל all זה this נתתי I considered אל in לבי my heart ולבור even to declare את כל all זה this, אשׁר that הצדיקים the righteous, והחכמים and the wise, ועבדיהם and their works, ביד in the hand האלהים of God: גם either אהבה love גם or שׂנאה hatred אין no יודע knoweth האדם man הכל all לפניהם׃ before}%
\verse{הכל All כאשׁר alike לכל to all: מקרה event אחד one לצדיק to the righteous, ולרשׁע and to the wicked; לטוב to the good ולטהור and to the clean, ולטמא and to the unclean; ולזבח to him that sacrificeth, ולאשׁר and to him that sacrificeth איננו not: זבח and to him that sacrificeth כטוב as the good, כחטא so the sinner; הנשׁבע he that sweareth, כאשׁר as שׁבועה an oath. ירא׃ that feareth}%
\verse{זה This רע an evil בכל among all אשׁר that נעשׂה are done תחת under השׁמשׁ the sun, כי that מקרה event אחד one לכל unto all: וגם yea, also לב the heart בני of the sons האדם of men מלא is full רע of evil, והוללות and madness בלבבם in their heart בחייהם while they live, ואחריו and after אל that to המתים׃ the dead.}%
\verseWithHeading{Death Deprives Humans of Everything in Life}{כי For מי to him אשׁר that יבחר אל to כל all החיים the living ישׁ there is בטחון hope: כי for לכלב dog חי a living הוא טוב is better מן than האריה lion. המת׃ a dead}%
\verse{כי For החיים the living יודעים know שׁימתו that they shall die: והמתים but the dead אינם not יודעים know מאומה any thing, ואין neither עוד have they any more להם שׂכר a reward; כי for נשׁכח of them is forgotten. זכרם׃ the memory}%
\verse{גם Also אהבתם their love, גם and שׂנאתם their hatred, גם and קנאתם their envy, כבר is now אבדה perished; וחלק a portion אין neither להם עוד have they any more לעולם forever בכל in any אשׁר that נעשׂה is done תחת under השׁמשׁ׃ the sun.}%
\verseWithHeading{Enjoy Life While It Lasts}{לך Go אכל thy way, eat בשׂמחה with joy, לחמך thy bread ושׁתה and drink בלב heart; טוב with a merry יינך thy wine כי for כבר now רצה accepteth האלהים God את מעשׂיך׃ thy works.}%
\verse{בכל always עת always יהיו be בגדיך Let thy garments לבנים white; ושׁמן ointment. על ראשׁך and let thy head אל no יחסר׃ lack}%
\verse{ראה חיים of the life עם with אשׁה the wife אשׁר whom אהבת thou lovest כל all ימי the days חיי in life, הבלך of thy vanity, אשׁר which נתן he hath given לך תחת thee under השׁמשׁ the sun, כל all ימי the days הבלך of thy vanity: כי for הוא that חלקך thy portion בחיים ובעמלך and in thy labor אשׁר which אתה thou עמל takest תחת under השׁמשׁ׃ the sun.}%
\verse{כל אשׁר whither תמצא findeth ידך thy hand לעשׂות to do, בכחך with thy might; עשׂה do כי for אין no מעשׂה work, וחשׁבון nor device, ודעת nor knowledge, וחכמה nor wisdom, בשׁאול in the grave, אשׁר אתה thou הלך goest. שׁמה׃ whither}%
\verseWithHeading{The Injustice of Time and Chance}{שׁבתי I returned, וראה and saw תחת under השׁמשׁ the sun, כי that לא not לקלים to the swift, המרוץ the race ולא nor לגבורים to the strong, המלחמה the battle וגם yet לא neither לחכמים to the wise, לחם bread וגם yet לא nor לנבנים to men of understanding, עשׁר riches וגם yet לא nor לידעים to men of skill; חן favor כי but עת time ופגע and chance יקרה happeneth את כלם׃ to them all.}%
\verse{כי For גם also לא not ידע knoweth האדם man את עתו his time: כדגים as the fishes שׁנאחזים that are taken במצודה net, רעה in an evil וכצפרים and as the birds האחזות that are caught בפח in the snare; כהם יוקשׁים snared בני so the sons האדם of men לעת time, רעה in an evil כשׁתפול when it falleth עליהם upon פתאם׃ suddenly}%
\verseWithHeading{Wisdom—Although Vulnerable—is Superior to Power}{גם also זה This ראיתי have I seen חכמה wisdom תחת under השׁמשׁ the sun, וגדולה great היא and it אלי׃ unto}%
\verse{עיר city, קטנה a little ואנשׁים בה מעט and few ובא within it; and there came אליה against מלך king גדול a great וסבב it, and besieged אתה ובנה it, and built עליה against מצודים bulwarks גדלים׃ great}%
\verse{ומצא Now there was found בה אישׁ man, מסכן in it a poor חכם wise ומלט delivered הוא and he את העיר the city; בחכמתו by his wisdom ואדם man לא yet no זכר remembered את האישׁ man. המסכן poor ההוא׃ that same}%
\verse{ואמרתי Then said אני I, טובה better חכמה Wisdom מגבורה than strength: וחכמת wisdom המסכן nevertheless the poor man's בזויה despised, ודבריו and his words אינם are not נשׁמעים׃ heard.}%
\verseWithHeading{Wisdom—Although Vulnerable—is Superior to Folly}{דברי The words חכמים of wise בנחת in quiet נשׁמעים heard מזעקת more than the cry מושׁל of him that ruleth בכסילים׃ among fools.}%
\verse{טובה better חכמה Wisdom מכלי than weapons קרב of war: וחוטא sinner אחד but one יאבד destroyeth טובה good. הרבה׃ much}%
\end{biblechapter}%
\begin{biblechapter}% Ecclesiastes 10
\verse{זבובי flies מות Dead יבאישׁ a stinking savor: יביע to send forth שׁמן cause the ointment רוקח of the apothecary יקר him that is in reputation מחכמה for wisdom מכבוד honor. סכלות folly מעט׃ a little}%
\verse{לב heart חכם A wise man's לימינו at his right hand; ולב heart כסיל but a fool's לשׂמאלו׃ at his left.}%
\verse{וגם Yea also, בדרך by the way, כשׁהסכל when he that is a fool הלך walketh לבו his wisdom חסר faileth ואמר and he saith לכל to every one סכל a fool. הוא׃ he}%
\verse{אם If רוח the spirit המושׁל of the ruler תעלה rise up עליך against מקומך thy place; אל not תנח כי for מרפא yielding יניח חטאים offenses. גדולים׃ great}%
\verse{ישׁ There is רעה an evil ראיתי I have seen תחת under השׁמשׁ the sun, כשׁגגה as an error שׁיצא which proceedeth מלפני from השׁליט׃ the ruler:}%
\verse{נתן is set הסכל Folly במרומים dignity, רבים in great ועשׁירים and the rich בשׁפל in low place. ישׁבו׃ sit}%
\verse{ראיתי I have seen עבדים servants על upon סוסים horses, ושׂרים and princes הלכים walking כעבדים as servants על upon הארץ׃ the earth.}%
\verseWithHeading{Accidents Happen—Even to Professionals}{חפר He that diggeth גומץ a pit בו יפול shall fall ופרץ into it; and whoso breaketh גדר a hedge, ישׁכנו shall bite נחשׁ׃ a serpent}%
\verse{מסיע Whoso removeth אבנים stones יעצב shall be hurt בהם בוקע therewith; he that cleaveth עצים wood יסכן׃ shall be endangered}%
\verseWithHeading{Hard Work and Skill Alone Cannot Succeed—Wisdom is Necessary}{אם If קהה be blunt, הברזל the iron והוא and he לא do not פנים the edge, קלקל whet וחילים more strength: יגבר then must he put to ויתרון profitable הכשׁיר to direct. חכמה׃ but wisdom}%
\verse{אם Surely ישׁך will bite הנחשׁ the serpent בלוא without לחשׁ enchantment; ואין is no יתרון better. לבעל and a babbler הלשׁון׃ and a babbler}%
\verseWithHeading{The Consequences of Foolishness}{דברי The words פי mouth חכם of a wise man's חן gracious; ושׂפתות but the lips כסיל of a fool תבלענו׃ will swallow up}%
\verse{תחלת The beginning דברי of the words פיהו of his mouth סכלות foolishness: ואחרית and the end פיהו of his talk הוללות madness. רעה׃ mischievous}%
\verse{והסכל A fool ירבה also is full דברים of words: לא cannot ידע tell האדם a man מה what shall be; שׁיהיה shall be ואשׁר and what יהיה מאחריו after מי him, who יגיד׃ can tell}%
\verse{עמל The labor הכסילים of the foolish תיגענו wearieth אשׁר every one of them, because לא not ידע he knoweth ללכת how to go אל to עיר׃ the city.}%
\verse{אי Woe לך ארץ to thee, O land, שׁמלכך when thy king נער a child, ושׂריך and thy princes בבקר in the morning! יאכלו׃ eat}%
\verse{אשׁריך Blessed ארץ thou, O land, שׁמלכך when thy king בן the son חורים of nobles, ושׂריך and thy princes בעת in due season, יאכלו eat בגבורה for strength, ולא and not בשׁתי׃ for drunkenness!}%
\verse{בעצלתים By much slothfulness ימך decayeth; המקרה the building ובשׁפלות and through idleness ידים of the hands ידלף droppeth through. הבית׃ the house}%
\verse{לשׂחוק for laughter, עשׂים is made לחם A feast ויין and wine ישׂמח maketh merry: חיים maketh merry: והכסף but money יענה answereth את הכל׃ all}%
\verse{גם no במדעך not in thy thought; מלך the king, אל not תקלל Curse ובחדרי in thy bedchamber: משׁכבך in thy bedchamber: אל not תקלל and curse עשׁיר the rich כי for עוף a bird השׁמים of the air יוליך shall carry את הקול the voice, ובעל and that which hath הכנפים wings יגיד shall tell דבר׃ the matter.}%
\end{biblechapter}%
\begin{biblechapter}% Ecclesiastes 11
\verseWithHeading{Living in the Light of the Limits of Human Knowledge}{שׁלח Cast לחמך thy bread על upon פני upon המים the waters: כי for ברב it after many הימים days. תמצאנו׃ thou shalt find}%
\verse{תן Give חלק a portion לשׁבעה to seven, וגם and also לשׁמונה to eight; כי for לא not תדע thou knowest מה what יהיה shall be רעה evil על upon הארץ׃ the earth.}%
\verse{אם If ימלאו be full העבים the clouds גשׁם of rain, על upon הארץ the earth: יריקו they empty ואם and if יפול fall עץ the tree בדרום toward the south, ואם or בצפון toward the north, מקום in the place שׁיפול falleth, העץ where the tree שׁם there יהוא׃ it}%
\verse{שׁמר He that observeth רוח the wind לא shall not יזרע sow; וראה and he that regardeth בעבים the clouds לא shall not יקצור׃ reap.}%
\verse{כאשׁר As אינך not יודע thou knowest מה what דרך the way הרוח of the spirit, כעצמים how the bones בבטן in the womb המלאה of her that is with child: ככה even so לא not תדע thou knowest את מעשׂה the works האלהים of God אשׁר who יעשׂה maketh את הכל׃ all.}%
\verse{בבקר In the morning זרע sow את זרעך thy seed, ולערב and in the evening אל not תנח ידך thine hand: כי for אינך not יודע thou knowest אי whether זה whether יכשׁר shall prosper, הזה either this או or זה that, ואם or whether שׁניהם they both כאחד alike טובים׃ good.}%
\verseWithHeading{Enjoy Life to the Fullest within the Auspices of the Fear of God}{ומתוק sweet, האור Truly the light וטוב and a pleasant לעינים for the eyes לראות to behold את השׁמשׁ׃ the sun:}%
\verse{כי But אם if שׁנים years, הרבה many יחיה live האדם a man בכלם in them all; ישׂמח rejoice ויזכר yet let him remember את ימי the days החשׁך of darkness; כי for הרבה many. יהיו they shall be כל All שׁבא that cometh הבל׃ vanity.}%
\verse{שׂמח Rejoice, בחור O young man, בילדותיך in thy youth; ויטיבך cheer לבך and let thy heart בימי thee in the days בחורותך of thy youth, והלך and walk בדרכי in the ways לבך of thine heart, ובמראי and in the sight עיניך of thine eyes: ודע but know כי thou, that על for כל all אלה these יביאך will bring האלהים God במשׁפט׃ thee into judgment.}%
\verse{והסר Therefore remove כעס sorrow מלבך from thy heart, והעבר and put away רעה evil מבשׂרך from thy flesh: כי for הילדות childhood והשׁחרות and youth הבל׃ vanity.}%
\end{biblechapter}%
\begin{biblechapter}% Ecclesiastes 12
\verseWithHeading{Advice to the Young: Life is Short and Then You Die}{וזכר Remember את בוראיך now thy Creator בימי in the days בחורתיך of thy youth, עד while אשׁר while לא not, יבאו come ימי days הרעה the evil והגיעו draw nigh, שׁנים nor the years אשׁר when תאמר thou shalt say, אין I have no לי בהם חפץ׃ pleasure}%
\verse{עד אשׁרא be not תחשׁך darkened, השׁמשׁ the sun, והאור or the light, והירח or the moon, והכוכבים or the stars, ושׁבו return העבים nor the clouds אחר after הגשׁם׃ the rain:}%
\verse{ביום In the day שׁיזעו shall tremble, שׁמרי when the keepers הבית of the house והתעותו shall bow themselves, אנשׁי men החיל and the strong ובטלו cease הטחנות and the grinders כי because מעטו וחשׁכו be darkened, הראות and those that look בארבות׃ out of the windows}%
\verse{וסגרו shall be shut דלתים And the doors בשׁוק in the streets, בשׁפל קול when the sound הטחנה of the grinding ויקום and he shall rise up לקול at the voice הצפור of the bird, וישׁחו shall be brought low; כל and all בנות the daughters השׁיר׃ of music}%
\verse{גם Also מגבה of high, יראו they shall be afraid וחתחתים and fears בדרך in the way, וינאץ shall flourish, השׁקד and the almond tree ויסתבל shall be a burden, החגב and the grasshopper ותפר shall fail: האביונה and desire כי because הלך goeth האדם man אל to בית home, עולמו his long וסבבו go about בשׁוק the streets: הספדים׃ and the mourners}%
\verse{עד Or אשׁר ever לא ever ירחק be loosed, חבל cord הכסף the silver ותרץ be broken, גלת bowl הזהב or the golden ותשׁבר be broken כד or the pitcher על at המבוע the fountain, ונרץ broken הגלגל or the wheel אל at הבור׃ the cistern.}%
\verse{וישׁב return העפר Then shall the dust על to הארץ the earth כשׁהיה as it was: והרוח and the spirit תשׁוב shall return אל unto האלהים God אשׁר who נתנה׃ gave}%
\verseWithHeading{Motto Restated}{הבל Vanity הבלים of vanities, אמר saith הקוהלת the preacher; הכל all הבל׃ vanity.}%
\verseWithHeading{Epilogue}{ויתר And moreover, שׁהיה was קהלת because the preacher חכם wise, עוד he still למד taught דעת knowledge; את העם the people ואזן yea, he gave good heed, וחקר and sought out, תקן set in order משׁלים proverbs. הרבה׃ many}%
\verse{בקשׁ sought קהלת The preacher למצא to find out דברי words: חפץ acceptable וכתוב and written ישׁר upright, דברי words אמת׃ of truth.}%
\verse{דברי The words חכמים of the wise כדרבנות as goads, וכמשׂמרות and as nails נטועים fastened בעלי the masters אספות of assemblies, נתנו are given מרעה אחד׃}%
\verse{ויתר And further, מהמה by these, בני my son, הזהר be admonished: עשׂות of making ספרים books הרבה many אין no קץ end; ולהג study הרבה and much יגעת a weariness בשׂר׃ of the flesh.}%
\verse{סוף the conclusion דבר matter: הכל of the whole נשׁמע Let us hear את האלהים God, ירא Fear ואת מצותיו his commandments: שׁמור and keep כי for זה this כל the whole האדם׃ of man.}%
\verse{כי For את כל every מעשׂה work האלהים God יבא shall bring במשׁפט into judgment, על with כל every נעלם secret thing, אם whether טוב good, ואם or whether רע׃ evil.}%
\end{biblechapter}%
