\biblebook{Genesis}
\begin{biblechapter}% Genesis 1
\verseWithHeading{The Creation}{בראשׁית In the beginning ברא created אלהים God את השׁמים the heaven ואת הארץ׃ and the earth.}%
\verse{והארץ And the earth היתה was תהו without form, ובהו and void; וחשׁך and darkness על upon פני the face תהום of the deep. ורוח And the Spirit אלהים of God מרחפת moved על upon פני the face המים׃ of the waters.}%
\verse{ויאמר said, אלהים And God יהי Let there be אור light: ויהי and there was אור׃ light.}%
\verse{וירא saw אלהים And God את האור the light, כי that טוב good: ויבדל divided אלהים and God בין divided האור the light ובין from החשׁך׃ the darkness.}%
\verse{ויקרא called אלהים And God לאור the light יום Day, ולחשׁך and the darkness קרא he called לילה Night. ויהי were ערב And the evening ויהי בקר and the morning יום day. אחד׃ the first}%
\verse{ויאמר said, אלהים And God יהי Let there be רקיע a firmament בתוך in the midst המים of the waters, ויהי and let מבדיל it divide בין it divide מים the waters למים׃ from the waters.}%
\verse{ויעשׂ made אלהים And God את הרקיע the firmament, ויבדל and divided בין and divided המים the waters אשׁר which מתחת under לרקיע the firmament ובין from המים the waters אשׁר which מעל above לרקיע the firmament: ויהי and it was כן׃ so.}%
\verse{ויקרא called אלהים And God לרקיע the firmament שׁמים Heaven. ויהי were ערב And the evening ויהי בקר and the morning יום day. שׁני׃ the second}%
\verse{ויאמר said, אלהים And God יקוו be gathered together המים Let the waters מתחת under השׁמים the heaven אל unto מקום place, אחד one ותראה appear: היבשׁה and let the dry ויהי and it was כן׃ so.}%
\verse{ויקרא called אלהים And God ליבשׁה the dry ארץ Earth; ולמקוה and the gathering together המים of the waters קרא called ימים he Seas: וירא saw אלהים and God כי that טוב׃ good.}%
\verse{ויאמר said, אלהים And God תדשׁא bring forth הארץ Let the earth דשׁא grass, עשׂב the herb מזריע yielding זרע seed, עץ tree פרי the fruit עשׂה yielding פרי fruit למינו after his kind, אשׁר whose זרעו seed בו על in itself, upon הארץ the earth: ויהי and it was כן׃ so.}%
\verse{ותוצא brought forth הארץ And the earth דשׁא grass, עשׂב herb מזריע yielding זרע seed למינהו after his kind, ועץ and the tree עשׂה yielding פרי fruit, אשׁר whose זרעו seed בו למינהו in itself, after his kind: וירא saw אלהים and God כי that טוב׃ good.}%
\verse{ויהי were ערב And the evening ויהי בקר and the morning יום day. שׁלישׁי׃ the third}%
\verse{ויאמר said, אלהים And God יהי Let there be מארת lights ברקיע in the firmament השׁמים of the heaven להבדיל to divide בין to divide היום the day ובין from הלילה the night; והיו and let them be לאתת for signs, ולמועדים and for seasons, ולימים and for days, ושׁנים׃ and years:}%
\verse{והיו And let them be למאורת ברקיע in the firmament השׁמים of the heaven להאיר to give light על upon הארץ the earth: ויהי and it was כן׃ so.}%
\verse{ויעשׂ made אלהים And God את שׁני two המארת lights; הגדלים great את המאור light הגדל the greater לממשׁלת to rule היום the day, ואת המאור light הקטן and the lesser לממשׁלת to rule הלילה the night: ואת הכוכבים׃ the stars}%
\verse{ויתן set אתם אלהים And God ברקיע them in the firmament השׁמים of the heaven להאיר to give light על upon הארץ׃ the earth,}%
\verse{ולמשׁל And to rule ביום over the day ובלילה and over the night, ולהבדיל and to divide בין and to divide האור the light ובין from החשׁך the darkness: וירא saw אלהים and God כי that טוב׃ good.}%
\verse{ויהי were ערב And the evening ויהי בקר and the morning יום day. רביעי׃ the fourth}%
\verse{ויאמר said, אלהים And God ישׁרצו bring forth abundantly המים Let the waters שׁרץ the moving creature נפשׁ that hath life, חיה that hath life, ועוף and fowl יעופף may fly על above הארץ the earth על in פני the open רקיע firmament השׁמים׃ of heaven.}%
\verse{ויברא created אלהים And God את התנינם whales, הגדלים great ואת כל and every נפשׁ creature החיה living הרמשׂת that moveth, אשׁר which שׁרצו brought forth abundantly, המים the waters למינהם after their kind, ואת כל and every עוף fowl כנף winged למינהו after his kind: וירא saw אלהים and God כי that טוב׃ good.}%
\verse{ויברך blessed אתם אלהים And God לאמר them, saying, פרו Be fruitful, ורבו and multiply, ומלאו and fill את המים the waters בימים in the seas, והעוף and let fowl ירב multiply בארץ׃ in the earth.}%
\verse{ויהי were ערב And the evening ויהי בקר and the morning יום day. חמישׁי׃ the fifth}%
\verse{ויאמר said, אלהים And God תוצא bring forth הארץ Let the earth נפשׁ creature חיה the living למינה after his kind, בהמה cattle, ורמשׂ and creeping thing, וחיתו and beast ארץ of the earth למינה after his kind: ויהי and it was כן׃ so.}%
\verse{ויעשׂ made אלהים And God את חית the beast הארץ of the earth למינה after his kind, ואת הבהמה and cattle למינה after their kind, ואת כל and every thing רמשׂ that creepeth האדמה upon the earth למינהו after his kind: וירא saw אלהים and God כי that טוב׃ good.}%
\verse{ויאמר said, אלהים And God נעשׂה Let us make אדם man בצלמנו in our image, כדמותנו after our likeness: וירדו and let them have dominion בדגת over the fish הים of the sea, ובעוף and over the fowl השׁמים of the air, ובבהמה and over the cattle, ובכל and over all הארץ the earth, ובכל and over every הרמשׂ creeping thing הרמשׂ that creepeth על upon הארץ׃ the earth.}%
\verse{ויברא created אלהים So God את האדם man בצלמו in his image, בצלם in the image אלהים of God ברא created אתו זכר he him; male ונקבה and female ברא created אתם׃}%
\verse{ויברך blessed אתם אלהים And God ויאמר said להם אלהים them, and God פרו unto them, Be fruitful, ורבו and multiply, ומלאו and replenish את הארץ the earth, וכבשׁה and subdue ורדו it: and have dominion בדגת over the fish הים of the sea, ובעוף and over the fowl השׁמים of the air, ובכל and over every חיה living thing הרמשׂת that moveth על upon הארץ׃ the earth.}%
\verse{ויאמר said, אלהים And God הנה Behold, נתתי I have given לכם את כל you every עשׂב herb זרע bearing זרע seed, אשׁר which על upon פני the face כל of all הארץ the earth, ואת כל and every העץ tree, אשׁר in the which בו פרי the fruit עץ of a tree זרע yielding זרע seed; לכם יהיה to you it shall be לאכלה׃ for meat.}%
\verse{ולכל And to every חית beast הארץ of the earth, ולכל and to every עוף fowl השׁמים of the air, ולכל and to every thing רומשׂ that creepeth על upon הארץ the earth, אשׁר wherein בו נפשׁ life, חיה life, את כל every ירק green עשׂב herb לאכלה for meat: ויהי and it was כן׃ so.}%
\verse{וירא saw אלהים And God את כל every thing אשׁר that עשׂה he had made, והנה and, behold, טוב good. מאד very ויהי were ערב And the evening ויהי בקר and the morning יום day. השׁשׁי׃ the sixth}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 2
\verse{ויכלו were finished, השׁמים Thus the heavens והארץ and the earth וכל and all צבאם׃ the host}%
\verse{ויכל ended אלהים God ביום day השׁביעי And on the seventh מלאכתו his work אשׁר which עשׂה he had made; וישׁבת and he rested ביום day השׁביעי on the seventh מכל from all מלאכתו his work אשׁר which עשׂה׃ he had made.}%
\verse{ויברך blessed אלהים And God את יום day, השׁביעי the seventh ויקדשׁ and sanctified אתו כי it: because בו שׁבת that in it he had rested מכל from all מלאכתו his work אשׁר which ברא created אלהים God לעשׂות׃ and made.}%
\verseWithHeading{The Generations of Heaven and Earth}{אלה These תולדות the generations השׁמים of the heavens והארץ and of the earth בהבראם when they were created, ביום in the day עשׂות made יהוה that the LORD אלהים God ארץ the earth ושׁמים׃ and the heavens,}%
\verse{וכל And every שׂיח plant השׂדה of the field טרם before יהיה it was בארץ in the earth, וכל and every עשׂב herb השׂדה of the field טרם before יצמח it grew: כי for לא had not המטיר caused it to rain יהוה the LORD אלהים God על upon הארץ the earth, ואדם a man אין and not לעבד to till את האדמה׃ the ground.}%
\verse{ואד a mist יעלה But there went up מן from הארץ the earth, והשׁקה and watered את כל the whole פני face האדמה׃ of the ground.}%
\verse{וייצר formed יהוה And the LORD אלהים God את האדם man עפר the dust מן of האדמה the ground, ויפח and breathed באפיו into his nostrils נשׁמת the breath חיים of life; ויהי became האדם and man לנפשׁ soul. חיה׃ a living}%
\verse{ויטע planted יהוה And the LORD אלהים God גן a garden בעדן in Eden; מקדם eastward וישׂם he put שׁם and there את האדם the man אשׁר whom יצר׃ he had formed.}%
\verse{ויצמח to grow יהוה made the LORD אלהים God מן And out of האדמה the ground כל every עץ tree נחמד that is pleasant למראה to the sight, וטוב and good למאכל for food; ועץ the tree החיים of life בתוך also in the midst הגן of the garden, ועץ and the tree הדעת of knowledge טוב of good ורע׃ and evil.}%
\verse{ונהר And a river יצא went out מעדן להשׁקות to water את הגן the garden; ומשׁם and from thence יפרד it was parted, והיה and became לארבעה into four ראשׁים׃ heads.}%
\verse{שׁם The name האחד of the first פישׁון Pison: הוא that it הסבב which compasseth את כל the whole ארץ land החוילה of Havilah, אשׁר where שׁם where הזהב׃ gold;}%
\verse{וזהב And the gold הארץ land ההוא of that טוב good: שׁם there הבדלח bdellium ואבן stone. השׁהם׃ and the onyx}%
\verse{ושׁם And the name הנהר river השׁני of the second גיחון Gihon: הוא the same הסובב it that compasseth את כל the whole ארץ land כושׁ׃ of Ethiopia.}%
\verse{ושׁם And the name הנהר river השׁלישׁי of the third חדקל Hiddekel: הוא that it ההלך which goeth קדמת toward the east אשׁור of Assyria. והנהר river הרביעי And the fourth הוא פרת׃ Euphrates.}%
\verse{ויקח took יהוה And the LORD אלהים God את האדם the man, וינחהו בגן him into the garden עדן of Eden לעבדה to dress ולשׁמרה׃ it and to keep}%
\verse{ויצו commanded יהוה And the LORD אלהים God על commanded האדם the man, לאמר saying, מכל עץ tree הגן of the garden אכל thou mayest freely eat: תאכל׃ thou mayest freely eat:}%
\verse{ומעץ הדעת of the knowledge טוב of good ורע and evil, לא thou shalt not תאכל eat ממנו of כי it: for ביום in the day אכלך that thou eatest ממנו thereof מות thou shalt surely die. תמות׃ thou shalt surely die.}%
\verse{ויאמר said, יהוה And the LORD אלהים God לא not טוב good היות should be האדם that the man לבדו alone; אעשׂה I will make לו עזר him a help כנגדו׃ meet for him.}%
\verse{ויצר formed יהוה the LORD אלהים God מן And out of האדמה the ground כל every חית beast השׂדה of the field, ואת כל and every עוף fowl השׁמים of the air; ויבא and brought אל unto האדם לראות to see מה what יקרא he would call לו וכל them: and whatsoever אשׁר them: and whatsoever יקרא called לו האדם נפשׁ creature, חיה every living הוא that שׁמו׃ the name}%
\verse{ויקרא gave האדם שׁמות names לכל to all הבהמה cattle, ולעוף and to the fowl השׁמים of the air, ולכל and to every חית beast השׂדה of the field; ולאדם לא there was not מצא found עזר a help כנגדו׃ meet for him.}%
\verse{ויפל to fall יהוה And the LORD אלהים God תרדמה caused a deep sleep על upon האדם ויישׁן and he slept: ויקח and he took אחת one מצלעתיו of his ribs, ויסגר and closed up בשׂר the flesh תחתנה׃ instead}%
\verse{ויבן made יהוה the LORD אלהים God את הצלע And the rib, אשׁר which לקח had taken מן from האדם man, לאשׁה he a woman, ויבאה and brought אל her unto האדם׃ the man.}%
\verse{ויאמר said, האדם זאת This הפעם now עצם bone מעצמי of my bones, ובשׂר and flesh מבשׂרי of my flesh: לזאת she יקרא shall be called אשׁה Woman, כי because מאישׁ לקחה was taken זאת׃ she}%
\verse{על כןעזב leave אישׁ shall a man את אביו his father ואת אמו and his mother, ודבק and shall cleave באשׁתו unto his wife: והיו and they shall be לבשׂר flesh. אחד׃ one}%
\verse{ויהיו And they were שׁניהם both ערומים naked, האדם the man ואשׁתו and his wife, ולא and were not יתבשׁשׁו׃ ashamed.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 3
\verseWithHeading{The Fall}{והנחשׁ Now the serpent היה was ערום more subtle מכל than any חית beast השׂדה of the field אשׁר which עשׂה had made. יהוה the LORD אלהים God ויאמר And he said אל unto האשׁה the woman, אף כימר said, אלהים hath God לא Ye shall not תאכלו eat מכל of every עץ tree הגן׃ of the garden?}%
\verse{ותאמר said האשׁה And the woman אל unto הנחשׁ the serpent, מפרי of the fruit עץ of the trees הגן of the garden: נאכל׃ We may eat}%
\verse{ומפרי העץ of the tree אשׁר which בתוך in the midst הגן of the garden, אמר hath said, אלהים God לא Ye shall not תאכלו eat ממנו of ולא it, neither תגעו shall ye touch בו פן it, lest תמתון׃ ye die.}%
\verse{ויאמר said הנחשׁ And the serpent אל unto האשׁה the woman, לא Ye shall not מות surely die: תמתון׃ surely die:}%
\verse{כי For ידע doth know אלהים God כי that ביום in the day אכלכם ye eat ממנו thereof, ונפקחו shall be opened, עיניכם then your eyes והייתם and ye shall be כאלהים as gods, ידעי knowing טוב good ורע׃ and evil.}%
\verse{ותרא saw האשׁה And when the woman כי that טוב good העץ the tree למאכל for food, וכי and that תאוה pleasant הוא it לעינים to the eyes, ונחמד to be desired העץ and a tree להשׂכיל to make wise, ותקח she took מפריו of the fruit ותאכל thereof, and did eat, ותתן and gave גם also לאישׁה עמה with ויאכל׃ her; and he did eat.}%
\verse{ותפקחנה were opened, עיני And the eyes שׁניהם of them both וידעו and they knew כי that עירמם naked; הם they ויתפרו and they sewed עלה leaves תאנה fig ויעשׂו together, and made להם חגרת׃ themselves aprons.}%
\verse{וישׁמעו And they heard את קול the voice יהוה of the LORD אלהים God מתהלך walking בגן in the garden לרוח in the cool היום of the day: ויתחבא hid themselves האדם ואשׁתו and his wife מפני from the presence יהוה of the LORD אלהים God בתוך amongst עץ the trees הגן׃ of the garden.}%
\verse{ויקרא called יהוה And the LORD אלהים God אל unto האדם ויאמר and said לו איכה׃ unto him, Where}%
\verse{ויאמר And he said, את קלך thy voice שׁמעתי I heard בגן in the garden, ואירא and I was afraid, כי because עירם naked; אנכי I ואחבא׃ and I hid myself.}%
\verse{ויאמר And he said, מי Who הגיד told לך כי thee that עירם naked? אתה thou המן of העץ the tree, אשׁר whereof צויתיך I commanded לבלתי thee that thou shouldest not אכל Hast thou eaten ממנו whereof אכלת׃ eat?}%
\verse{ויאמר said, האדם And the man האשׁה The woman אשׁר whom נתתה thou gavest עמדי הוא me, she נתנה gave לי מן me of העץ the tree, ואכל׃ and I did eat.}%
\verse{ויאמר said יהוה And the LORD אלהים God לאשׁה unto the woman, מה What זאת this עשׂית thou hast done? ותאמר said, האשׁה And the woman הנחשׁ The serpent השׁיאני beguiled ואכל׃ me, and I did eat.}%
\verse{ויאמר said יהוה And the LORD אלהים God אל unto הנחשׁ the serpent, כי Because עשׂית thou hast done זאת this, ארור cursed אתה thou מכל above all הבהמה cattle, ומכל and above every חית beast השׂדה of the field; על upon גחנך thy belly תלך shalt thou go, ועפר and dust תאכל shalt thou eat כל all ימי the days חייך׃ of thy life:}%
\verse{ואיבה enmity אשׁית And I will put בינך between ובין and between האשׁה thee and the woman, ובין זרעך thy seed ובין זרעה and her seed; הוא it ישׁופך shall bruise ראשׁ thy head, ואתה and thou תשׁופנו shalt bruise עקב׃ his heel.}%
\verse{אל Unto האשׁה the woman אמר he said, הרבה I will greatly ארבה multiply עצבונך thy sorrow והרנך and thy conception; בעצב in sorrow תלדי thou shalt bring forth בנים children; ואל to אישׁך thy husband, תשׁוקתך and thy desire והוא and he ימשׁל׃ shall rule}%
\verse{ולאדם And unto Adam אמר he said, כי Because שׁמעת thou hast hearkened לקול unto the voice אשׁתך of thy wife, ותאכל and hast eaten מן of העץ the tree, אשׁר of which צויתיך I commanded לאמר thee, saying, לא Thou shalt not תאכל eat ממנו of ארורה it: cursed האדמה the ground בעבורך for thy sake; בעצבון in sorrow תאכלנה shalt thou eat כל it all ימי the days חייך׃ of thy life;}%
\verse{וקוץ Thorns ודרדר also and thistles תצמיח shall it bring forth לך ואכלת to thee; and thou shalt eat את עשׂב the herb השׂדה׃ of the field;}%
\verse{בזעת In the sweat אפיך of thy face תאכל shalt thou eat לחם bread, עד till שׁובך thou return אל unto האדמה the ground; כי for ממנה out of לקחת it wast thou taken: כי for עפר dust אתה thou ואל and unto עפר dust תשׁוב׃ shalt thou return.}%
\verse{ויקרא called האדם שׁם name אשׁתו his wife's חוה Eve; כי because הוא she היתה was אם the mother כל of all חי׃ living.}%
\verse{ויעשׂ make יהוה did the LORD אלהים God לאדם ולאשׁתו also and to his wife כתנות coats עור of skins, וילבשׁם׃ and clothed}%
\verse{ויאמר said, יהוה And the LORD אלהים God הן Behold, האדם the man היה is become כאחד as one ממנו of לדעת us, to know טוב good ורע and evil: ועתה and now, פן lest ישׁלח he put forth ידו his hand, ולקח and take גם also מעץ of the tree החיים of life, ואכל and eat, וחי and live לעלם׃ forever:}%
\verse{וישׁלחהו sent him forth יהוה Therefore the LORD אלהים God מגן from the garden עדן of Eden, לעבד to till את האדמה the ground אשׁר לקח he was taken. משׁם׃}%
\verse{ויגרשׁ So he drove out את האדם the man; וישׁכן and he placed מקדם at the east לגן of the garden עדן of Eden את הכרבים Cherubims, ואת להט and a flaming החרב sword המתהפכת which turned every way, לשׁמר to keep את דרך the way עץ of the tree החיים׃ of life.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 4
\verseWithHeading{Cain and Abel}{והאדם ידע knew את חוה Eve אשׁתו his wife; ותהר and she conceived, ותלד and bore את קין Cain, ותאמר and said, קניתי I have gotten אישׁ a man את from יהוה׃ the LORD.}%
\verse{ותסף And she again ללדת bore את אחיו his brother את הבל Abel. ויהי was הבל And Abel רעה a keeper צאן of sheep, וקין but Cain היה was עבד a tiller אדמה׃ of the ground.}%
\verse{ויהי it came to pass, מקץ ימים of time ויבא brought קין that Cain מפרי of the fruit האדמה of the ground מנחה an offering ליהוה׃ unto the LORD.}%
\verse{והבל And Abel, הביא brought גם also הוא he מבכרות of the firstlings צאנו of his flock ומחלבהן and of the fat וישׁע had respect יהוה thereof. And the LORD אל unto הבל Abel ואל and to מנחתו׃ his offering:}%
\verse{ואל But unto קין Cain ואל and to מנחתו his offering לא he had not respect. שׁעה he had not respect. ויחר wroth, לקין And Cain מאד was very ויפלו fell. פניו׃ and his countenance}%
\verse{ויאמר said יהוה And the LORD אל unto קין Cain, למה Why חרה art thou wroth? לך ולמה and why נפלו fallen? פניך׃ is thy countenance}%
\verse{הלוא shalt thou not אם If תיטיב thou doest well, שׂאת be accepted? ואם and if לא thou doest not תיטיב well, לפתח at the door. חטאת sin רבץ lieth ואליך And unto תשׁוקתו thee his desire, ואתה and thou תמשׁל׃ shalt rule}%
\verse{ויאמר talked קין And Cain אל with הבל Abel אחיו his brother: ויהי and it came to pass, בהיותם when they were בשׂדה in the field, ויקם rose up קין that Cain אל against הבל Abel אחיו his brother, ויהרגהו׃ and slew}%
\verse{ויאמר said יהוה And the LORD אל unto קין Cain, אי Where הבל Abel אחיך thy brother? ויאמר And he said, לא not: ידעתי I know השׁמר keeper? אחי my brother's אנכי׃ I}%
\verse{ויאמר And he said, מה What עשׂית hast thou done? קול the voice דמי blood אחיך of thy brother's צעקים crieth אלי unto מן me from האדמה׃ the ground.}%
\verse{ועתה And now ארור cursed אתה thou מן from האדמה the earth, אשׁר which פצתה hath opened את פיה her mouth לקחת to receive את דמי blood אחיך thy brother's מידך׃ from thy hand;}%
\verse{כי When תעבד thou tillest את האדמה the ground, לא it shall not תסף henceforth תת yield כחה unto thee her strength; לך נע a fugitive ונד and a vagabond תהיה shalt thou be בארץ׃ in the earth.}%
\verse{ויאמר said קין And Cain אל unto יהוה the LORD, גדול greater עוני My punishment מנשׂוא׃}%
\verse{הן Behold, גרשׁת אתייום this day מעל from פני the face האדמה of the earth; ומפניך and from thy face אסתר shall I be hid; והייתי and I shall be נע a fugitive ונד and a vagabond בארץ in the earth; והיה and it shall come to pass, כל every one מצאי that findeth יהרגני׃ me shall slay}%
\verse{ויאמר said לו יהוה And the LORD לכן unto him, Therefore כל whosoever הרג slayeth קין Cain, שׁבעתים on him sevenfold. יקם vengeance shall be taken וישׂם set יהוה And the LORD לקין upon Cain, אות a mark לבלתי lest הכות him should kill אתו כל any מצאו׃ finding}%
\verse{ויצא went out קין And Cain מלפני from the presence יהוה of the LORD, וישׁב and dwelt בארץ in the land נוד of Nod, קדמת on the east עדן׃ of Eden.}%
\verse{וידע knew קין And Cain את אשׁתו his wife; ותהר and she conceived, ותלד and bore את חנוך Enoch: ויהי and he built בנה and he built עיר a city, ויקרא and called שׁם the name העיר of the city, כשׁם after the name בנו of his son, חנוך׃ Enoch.}%
\verse{ויולד was born לחנוך And unto Enoch את עירד Irad: ועירד and Irad ילד begot את מחויאל Mehujael: ומחייאל and Mehujael ילד begot את מתושׁאל Methusael: ומתושׁאל and Methusael ילד begot את למך׃ Lamech.}%
\verse{ויקח took לו למך And Lamech שׁתי unto him two נשׁים wives: שׁם the name האחת of the one עדה Adah, ושׁם and the name השׁנית of the other צלה׃ Zillah.}%
\verse{ותלד bore עדה And Adah את יבל Jabal: הוא he היה was אבי the father ישׁב of such as dwell אהל in tents, ומקנה׃ and cattle.}%
\verse{ושׁם name אחיו And his brother's יובל Jubal: הוא he היה was אבי the father כל of all תפשׂ such as handle כנור the harp ועוגב׃ and organ.}%
\verse{וצלה And Zillah, גם also הוא she ילדה bore את תובל קין Tubal-cain, לטשׁ an instructor כל of every חרשׁ artificer נחשׁת in brass וברזל and iron: ואחות and the sister תובל קין of Tubal-cain נעמה׃ Naamah.}%
\verse{ויאמר said למך And Lamech לנשׁיו unto his wives, עדה Adah וצלה and Zillah, שׁמען Hear קולי my voice; נשׁי ye wives למך of Lamech, האזנה hearken אמרתי unto my speech: כי for אישׁ הרגתי I have slain לפצעי to my wounding, וילד and a young man לחברתי׃ to my hurt.}%
\verse{כי If שׁבעתים sevenfold, יקם shall be avenged קין Cain ולמך truly Lamech שׁבעים seventy ושׁבעה׃ and sevenfold.}%
\verse{וידע knew אדם And Adam עוד again; את אשׁתו his wife ותלד and she bore בן a son, ותקרא and called את שׁמו his name שׁת Seth: כי For שׁת hath appointed לי אלהים God, זרע seed אחר me another תחת instead of הבל Abel, כי הרגו slew. קין׃ whom Cain}%
\verse{ולשׁת And to Seth, גם also הוא to him ילד there was born בן a son; ויקרא and he called את שׁמו his name אנושׁ Enos: אז then הוחל began לקרא men to call בשׁם upon the name יהוה׃ of the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 5
\verseWithHeading{Adam’s Descendants to Noah}{זה This ספר the book תולדת of the generations אדם of Adam. ביום In the day ברא created אלהים that God אדם man, בדמות in the likeness אלהים of God עשׂה made אתו׃}%
\verse{זכר Male ונקבה and female בראם created ויברך he them; and blessed אתם ויקרא them, and called את שׁמם their name אדם ביום in the day הבראם׃ when they were created.}%
\verse{ויחי lived אדם And Adam שׁלשׁים and thirty ומאת a hundred שׁנה years, ויולד and begot בדמותו in his own likeness, כצלמו after his image; ויקרא and called את שׁמו his name שׁת׃ Seth:}%
\verse{ויהיו were ימי And the days אדם of Adam אחרי after הולידו he had begotten את שׁת Seth שׁמנה eight מאת hundred שׁנה years: ויולד and he begot בנים sons ובנות׃ and daughters:}%
\verse{ויהיו were כל And all ימי the days אדם Adam אשׁר that חי תשׁע nine מאות hundred שׁנה years: ושׁלשׁים and thirty שׁנה וימת׃ and he died.}%
\verse{ויחי lived שׁת And Seth חמשׁ and five שׁנים years, ומאת a hundred שׁנה ויולד and begot את אנושׁ׃ Enos:}%
\verse{ויחי lived שׁת And Seth אחרי after הולידו he begot את אנושׁ Enos שׁבע and seven שׁנים years, ושׁמנה eight מאות hundred שׁנה ויולד and begot בנים sons ובנות׃ and daughters:}%
\verse{ויהיו were כל And all ימי the days שׁת of Seth שׁתים and twelve עשׂרה and twelve שׁנה years: ותשׁע nine מאות hundred שׁנה וימת׃ and he died.}%
\verse{ויחי lived אנושׁ And Enos תשׁעים ninety שׁנה years, ויולד and begot את קינן׃ Cainan:}%
\verse{ויחי lived אנושׁ And Enos אחרי after הולידו he begot את קינן Cainan חמשׁ and fifteen עשׂרה and fifteen שׁנה years, ושׁמנה eight מאות hundred שׁנה ויולד and begot בנים sons ובנות׃ and daughters:}%
\verse{ויהיו were כל And all ימי the days אנושׁ of Enos חמשׁ and five שׁנים years: ותשׁע nine מאות hundred שׁנה וימת׃ and he died.}%
\verse{ויחי lived קינן And Cainan שׁבעים seventy שׁנה years, ויולד and begot את מהללאל׃ Mahalaleel:}%
\verse{ויחי lived קינן And Cainan אחרי after הולידו he begot את מהללאל Mahalaleel ארבעים and forty שׁנה years, ושׁמנה eight מאות hundred שׁנה ויולד and begot בנים sons ובנות׃ and daughters:}%
\verse{ויהיו were כל And all ימי the days קינן of Cainan עשׂר and ten שׁנים years: ותשׁע nine מאות hundred שׁנה וימת׃ and he died.}%
\verse{ויחי lived מהללאל And Mahalaleel חמשׁ and five שׁנים years, ושׁשׁים sixty שׁנה ויולד and begot את ירד׃ Jared:}%
\verse{ויחי lived מהללאל And Mahalaleel אחרי after הולידו he begot את ירד Jared שׁלשׁים and thirty שׁנה years, ושׁמנה eight מאות hundred שׁנה ויולד and begot בנים sons ובנות׃ and daughters:}%
\verse{ויהיו were כל And all ימי the days מהללאל of Mahalaleel חמשׁ and five ותשׁעים ninety שׁנה years: ושׁמנה eight מאות hundred שׁנה וימת׃ and he died.}%
\verse{ויחי lived ירד And Jared שׁתים and two ושׁשׁים sixty שׁנה years, ומאת a hundred שׁנה ויולד and he begot את חנוך׃ Enoch:}%
\verse{ויחי lived ירד And Jared אחרי after הולידו he begot את חנוך Enoch שׁמנה eight מאות hundred שׁנה years, ויולד and begot בנים sons ובנות׃ and daughters:}%
\verse{ויהיו were כל And all ימי the days ירד of Jared שׁתים and two ושׁשׁים sixty שׁנה years: ותשׁע nine מאות hundred שׁנה וימת׃ and he died.}%
\verse{ויחי lived חנוך And Enoch חמשׁ and five ושׁשׁים sixty שׁנה years, ויולד and begot את מתושׁלח׃ Methuselah:}%
\verse{ויתהלך walked חנוך And Enoch את with האלהים God אחרי after הולידו he begot את מתושׁלח Methuselah שׁלשׁ three מאות hundred שׁנה years, ויולד and begot בנים sons ובנות׃ and daughters:}%
\verse{ויהי were כל And all ימי the days חנוך of Enoch חמשׁ and five ושׁשׁים sixty שׁנה years: ושׁלשׁ three מאות hundred שׁנה׃}%
\verse{ויתהלך walked חנוך And Enoch את with האלהים God: ואיננו and he not; כי for לקח took אתו אלהים׃ God}%
\verse{ויחי lived מתושׁלח And Methuselah שׁבע and seven ושׁמנים eighty שׁנה years, ומאת a hundred שׁנה ויולד and begot את למך׃ Lamech:}%
\verse{ויחי lived מתושׁלח And Methuselah אחרי after הולידו he begot את למך Lamech שׁתים and two ושׁמונים eighty שׁנה years, ושׁבע seven מאות hundred שׁנה ויולד and begot בנים sons ובנות׃ and daughters:}%
\verse{ויהיו were כל And all ימי the days מתושׁלח of Methuselah תשׁע nine ושׁשׁים sixty שׁנה years: ותשׁע and nine מאות hundred שׁנה וימת׃ and he died.}%
\verse{ויחי lived למך And Lamech שׁתים and two ושׁמנים eighty שׁנה years, ומאת a hundred שׁנה ויולד and begot בן׃ a son:}%
\verse{ויקרא And he called את שׁמו his name נח Noah, לאמר saying, זה This ינחמנו shall comfort ממעשׂנו us concerning our work ומעצבון ידינו of our hands, מן us concerning our work האדמה the ground אשׁר which אררה hath cursed. יהוה׃ the LORD}%
\verse{ויחי lived למך And Lamech אחרי after הולידו he begot את נח Noah חמשׁ five ותשׁעים ninety שׁנה years, וחמשׁ and five מאת hundred שׁנה ויולד and begot בנים sons ובנות׃ and daughters:}%
\verse{ויהי were כל And all ימי the days למך of Lamech שׁבע seven ושׁבעים seventy שׁנה years: ושׁבע and seven מאות hundred שׁנה וימת׃ and he died.}%
\verse{ויהי was נח And Noah בן old: חמשׁ five מאות hundred שׁנה years ויולד begot נח and Noah את שׁם Shem, את חם Ham, ואת יפת׃ and Japheth.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 6
\verseWithHeading{Prelude to the Flood}{ויהי And it came to pass, כי when החל began האדם men לרב to multiply על on פני the face האדמה of the earth, ובנות and daughters ילדו׃ were born}%
\verse{ויראו saw בני That the sons האלהים of God את בנות the daughters האדם of men כי that טבת fair; הנה they ויקחו and they took להם נשׁים them wives מכל of all אשׁר which בחרו׃ they chose.}%
\verse{ויאמר said, יהוה And the LORD לא shall not ידון strive רוחי My spirit באדם with man, לעלם always בשׁגם also הוא for that he בשׂר flesh: והיו shall be ימיו yet his days מאה a hundred ועשׂרים and twenty שׁנה׃ years.}%
\verse{הנפלים giants היו There were בארץ in the earth בימים days; ההם in those וגם and also אחרי after כן that, אשׁר when יבאו came in בני the sons האלהים of God אל unto בנות the daughters האדם of men, וילדו and they bore להם המה to them, the same הגברים mighty men אשׁר which מעולם of old, אנשׁי men השׁם׃ of renown.}%
\verse{וירא saw יהוה And GOD כי that רבה great רעת the wickedness האדם of man בארץ in the earth, וכל and every יצר imagination מחשׁבת of the thoughts לבו of his heart רק only רע evil כל continually. היום׃ continually.}%
\verse{וינחם And it repented יהוה the LORD כי that עשׂה he had made את האדם man בארץ on the earth, ויתעצב and it grieved אל him at לבו׃ his heart.}%
\verse{ויאמר said, יהוה And the LORD אמחה I will destroy את האדם man אשׁר whom בראתי I have created מעל from פני the face האדמה of the earth; מאדם both man, עד and בהמה beast, עד and רמשׂ the creeping thing, ועד and עוף the fowls השׁמים of the air; כי for נחמתי it repenteth כי me that עשׂיתם׃ I have made}%
\verse{ונח But Noah מצא found חן grace בעיני in the eyes יהוה׃ of the LORD.}%
\verse{אלה These תולדת the generations נח of Noah: נח Noah אישׁ צדיק a just תמים perfect היה was בדרתיו in his generations, את with האלהים God. התהלך walked נח׃ Noah}%
\verse{ויולד begot נח And Noah שׁלשׁה three בנים sons, את שׁם Shem, את חם Ham, ואת יפת׃ and Japheth.}%
\verse{ותשׁחת also was corrupt הארץ The earth לפני before האלהים God, ותמלא was filled הארץ and the earth חמס׃ with violence.}%
\verse{וירא looked upon אלהים And God את הארץ the earth, והנה and, behold, נשׁחתה it was corrupt; כי for השׁחית had corrupted כל all בשׂר flesh את דרכו his way על upon הארץ׃ the earth.}%
\verse{ויאמר said אלהים And God לנח unto Noah, קץ The end כל of all בשׂר flesh בא is come לפני before כי me; for מלאה is filled הארץ the earth חמס with violence מפניהם through them; והנני משׁחיתם I will destroy את them with הארץ׃ the earth.}%
\verse{עשׂה Make לך תבת thee an ark עצי wood; גפר of gopher קנים rooms תעשׂה shalt thou make את התבה in the ark, וכפרת and shalt pitch אתה מבית it within ומחוץ and without בכפר׃ with pitch.}%
\verse{וזה And this אשׁר which תעשׂה thou shalt make אתה שׁלשׁ three מאות hundred אמה cubits, ארך it The length התבה of the ark חמשׁים of it fifty אמה cubits, רחבה the breadth ושׁלשׁים of it thirty אמה cubits. קומתה׃ and the height}%
\verse{צהר A window תעשׂה shalt thou make לתבה to the ark, ואל and in אמה a cubit תכלנה shalt thou finish מלמעלה it above; ופתח and the door התבה of the ark בצדה in the side תשׂים shalt thou set תחתים thereof; lower, שׁנים second, ושׁלשׁים and third תעשׂה׃ shalt thou make}%
\verse{ואני I, even I, הנני מביא do bring את המבול a flood מים of waters על upon הארץ the earth, לשׁחת to destroy כל all בשׂר flesh, אשׁר wherein בו רוח the breath חיים of life, מתחת from under השׁמים heaven; כל every thing אשׁר that בארץ in the earth יגוע׃ shall die.}%
\verse{והקמתי thee will I establish את בריתי my covenant; אתך But with ובאת and thou shalt come אל into התבה the ark, אתה thou, ובניך and thy sons, ואשׁתך and thy wife, ונשׁי wives בניך and thy sons' אתך׃ with}%
\verse{ומכל of all החי living thing מכל of every בשׂר flesh, שׁנים two מכל תביא shalt thou bring אל into התבה the ark, להחית to keep alive אתך with זכר male ונקבה and female. יהיו׃ thee; they shall be}%
\verse{מהעוף למינהו after their kind, ומן and of הבהמה cattle למינה after their kind, מכל of every רמשׂ creeping thing האדמה of the earth למינהו after his kind, שׁנים two מכל of every יבאו shall come אליך unto להחיות׃ thee, to keep alive.}%
\verse{ואתה thou קח And take לך מכל unto thee of all מאכל food אשׁר that יאכל is eaten, ואספת and thou shalt gather אליך to והיה thee; and it shall be לך ולהם לאכלה׃ for food}%
\verse{ויעשׂ Thus did נח Noah; ככל according to all אשׁר that צוה commanded אתו אלהים God כן him, so עשׂה׃ did}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 7
\verse{ויאמר said יהוה And the LORD לנח unto Noah, בא Come אתה thou וכל and all ביתך thy house אל into התבה the ark; כי for אתך ראיתי thee have I seen צדיק righteous לפני before בדור generation. הזה׃ me in this}%
\verse{מכל הבהמה beast הטהורה clean תקח thou shalt take לך שׁבעה to thee by sevens, שׁבעה to thee by sevens, אישׁ the male ואשׁתו and his female: ומן and of הבהמה beasts אשׁר that לא not טהרה clean הוא שׁנים by two, אישׁ the male ואשׁתו׃ and his female.}%
\verse{גם also מעוף השׁמים of the air שׁבעה by sevens, שׁבעה by sevens, זכר the male ונקבה and the female; לחיות alive זרע to keep seed על upon פני the face כל of all הארץ׃ the earth.}%
\verse{כי For לימים days, עוד yet שׁבעה seven אנכי and I ממטיר will cause it to rain על upon הארץ the earth ארבעים forty יום days וארבעים and forty לילה nights; ומחיתי will I destroy את כל and every היקום living substance אשׁר that עשׂיתי I have made מעל from off פני the face האדמה׃ of the earth.}%
\verse{ויעשׂ did נח And Noah ככל according unto all אשׁר that צוהו commanded יהוה׃ the LORD}%
\verseWithHeading{The Flood}{ונח And Noah בן old שׁשׁ six מאות hundred שׁנה years והמבול when the flood היה was מים of waters על upon הארץ׃ the earth.}%
\verse{ויבא went in, נח And Noah ובניו and his sons, ואשׁתו and his wife, ונשׁי wives בניו and his sons' אתו with אל him, into התבה the ark, מפני because מי of the waters המבול׃ of the flood.}%
\verse{מן Of הבהמה beasts, הטהורה clean ומן and of הבהמה beasts אשׁר that איננה not טהרה clean, ומן and of העוף fowls, וכל and of every thing אשׁר that רמשׂ creepeth על upon האדמה׃ the earth,}%
\verse{שׁנים two שׁנים and two באו There went in אל unto נח Noah אל into התבה the ark, זכר the male ונקבה and the female, כאשׁר as צוה had commanded אלהים God את נח׃ Noah.}%
\verse{ויהי And it came to pass לשׁבעת after seven הימים days, ומי that the waters המבול of the flood היו were על upon הארץ׃ the earth.}%
\verse{בשׁנת year שׁשׁ In the six מאות hundredth שׁנה לחיי life, נח of Noah's בחדשׁ month, השׁני in the second בשׁבעה the seventeenth עשׂר the seventeenth יום day לחדשׁ of the month, ביום day הזה נבקעו broken up, כל were all מעינות the fountains תהום deep רבה of the great וארבת and the windows השׁמים of heaven נפתחו׃ were opened.}%
\verse{ויהי was הגשׁם And the rain על upon הארץ the earth ארבעים forty יום days וארבעים and forty לילה׃ nights.}%
\verse{בעצם היום day הזה בא entered נח Noah, ושׁם and Shem, וחם and Ham, ויפת and Japheth, בני the sons נח of Noah, ואשׁת wife, נח and Noah's ושׁלשׁת and the three נשׁי wives בניו of his sons אתם with אל them, into התבה׃ the ark;}%
\verse{המה They, וכל and every החיה beast למינה after his kind, וכל and all הבהמה the cattle למינה after their kind, וכל and every הרמשׂ creeping thing הרמשׂ that creepeth על upon הארץ the earth למינהו after his kind, וכל and every העוף fowl למינהו after his kind, כל every צפור bird כל of every כנף׃ sort.}%
\verse{ויבאו And they went in אל unto נח Noah אל into התבה the ark, שׁנים two שׁנים and two מכל of all הבשׂר flesh, אשׁר wherein בו רוח the breath חיים׃ of life.}%
\verse{והבאים And they that went in, זכר male ונקבה and female מכל of all בשׂר flesh, באו went in כאשׁר as צוה had commanded אתו אלהים God ויסגר shut him in. יהוה him: and the LORD בעדו׃ shut him in.}%
\verse{ויהי was המבול And the flood ארבעים forty יום days על upon הארץ the earth; וירבו increased, המים and the waters וישׂאו and bore up את התבה the ark, ותרם and it was lifted up מעל above הארץ׃ the earth.}%
\verse{ויגברו prevailed, המים And the waters וירבו and were increased מאד greatly על upon הארץ the earth; ותלך went התבה and the ark על upon פני the face המים׃ of the waters.}%
\verse{והמים And the waters גברו prevailed מאד exceedingly מאד exceedingly על upon הארץ the earth; ויכסו were covered. כל and all ההרים hills, הגבהים the high אשׁר that תחת under כל the whole השׁמים׃ heaven,}%
\verse{חמשׁ עשׂרהמה cubits מלמעלה upward גברו prevail; המים did the waters ויכסו were covered. ההרים׃ and the mountains}%
\verse{ויגוע died כל And all בשׂר flesh הרמשׂ that moved על upon הארץ the earth, בעוף both of fowl, ובבהמה and of cattle, ובחיה and of beast, ובכל and of every השׁרץ creeping thing השׁרץ that creepeth על upon הארץ the earth, וכל and every האדם׃ man:}%
\verse{כל All אשׁר that נשׁמת the breath רוח the breath חיים of life, באפיו in whose nostrils מכל of all אשׁר בחרבה in the dry מתו׃ died.}%
\verse{וימח was destroyed את כל And every היקום living substance אשׁר which על was upon פני the face האדמה of the ground, מאדם both man, עד and בהמה cattle, עד and רמשׂ the creeping things, ועד and עוף the fowl השׁמים of the heaven; וימחו and they were destroyed מן both man, הארץ the earth: וישׁאר remained אך only נח and Noah ואשׁר and they that אתו with בתבה׃ him in the ark.}%
\verse{ויגברו prevailed המים And the waters על upon הארץ the earth חמשׁים and fifty ומאת a hundred יום׃ days.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 8
\verseWithHeading{The Flood Subsides}{ויזכר remembered אלהים And God את נח Noah, ואת כל and every החיה living thing, ואת כל and all הבהמה the cattle אשׁר that אתו with בתבה him in the ark: ויעבר to pass אלהים and God רוח made a wind על over הארץ the earth, וישׁכו assuaged; המים׃ and the waters}%
\verse{ויסכרו were stopped, מעינת The fountains תהום also of the deep וארבת and the windows השׁמים of heaven ויכלא was restrained; הגשׁם and the rain מן from השׁמים׃ heaven}%
\verse{וישׁבו returned המים And the waters מעל from off הארץ the earth הלוך continually: ושׁוב continually: ויחסרו were abated. המים the waters מקצה and after the end חמשׁים and fifty ומאת of the hundred יום׃ days}%
\verse{ותנח rested התבה And the ark בחדשׁ month, השׁביעי in the seventh בשׁבעה on the seventeenth עשׂר on the seventeenth יום day לחדשׁ of the month, על upon הרי the mountains אררט׃ of Ararat.}%
\verse{והמים And the waters היו continually הלוך continually וחסור decreased עד until החדשׁ month: העשׂירי the tenth בעשׂירי in the tenth באחד on the first לחדשׁ of the month, נראו seen. ראשׁי were the tops ההרים׃ of the mountains}%
\verse{ויהי And it came to pass מקץ at the end ארבעים of forty יום days, ויפתח opened נח that Noah את חלון the window התבה of the ark אשׁר which עשׂה׃ he had made:}%
\verse{וישׁלח And he sent forth את הערב a raven, ויצא which went forth יצוא to and fro, ושׁוב to and fro, עד until יבשׁת were dried up המים the waters מעל from off הארץ׃ the earth.}%
\verse{וישׁלח Also he sent forth את היונה a dove מאתו from לראות him, to see הקלו were abated המים if the waters מעל from off פני the face האדמה׃ of the ground;}%
\verse{ולא no מצאה found היונה But the dove מנוח rest לכף for the sole רגלה of her foot, ותשׁב and she returned אליו unto אל him into התבה the ark, כי for מים the waters על on פני the face כל of the whole הארץ earth: וישׁלח then he put forth ידו his hand, ויקחה and took ויבא אתהליו unto אל him into התבה׃ the ark.}%
\verse{ויחל And he stayed עוד yet שׁבעת seven ימים days; אחרים other ויסף and again שׁלח he sent forth את היונה the dove מן out of התבה׃ the ark;}%
\verse{ותבא came in אליו to היונה And the dove לעת him in ערב the evening; והנה and, lo, עלה leaf זית an olive טרף plucked off: בפיה in her mouth וידע knew נח so Noah כי that קלו were abated המים the waters מעל from off הארץ׃ the earth.}%
\verse{וייחל And he stayed עוד yet שׁבעת seven ימים days; אחרים other וישׁלח and sent forth את היונה the dove; ולא not יספה again שׁוב which returned אליו unto עוד׃ him any more.}%
\verse{ויהי And it came to pass באחת and first ושׁשׁ in the six מאות hundredth שׁנה year, בראשׁון in the first באחד the first לחדשׁ of the month, חרבו were dried up המים the waters מעל from off הארץ the earth: ויסר removed נח and Noah את מכסה the covering התבה of the ark, וירא and looked, והנה and, behold, חרבו פני the face האדמה׃ of the ground}%
\verse{ובחדשׁ month, השׁני And in the second בשׁבעה on the seven ועשׂרים and twentieth יום day לחדשׁ of the month, יבשׁה dried. הארץ׃ was the earth}%
\verse{וידבר spoke אלהים And God אל unto נח Noah, לאמר׃ saying,}%
\verse{צא Go forth מן of התבה the ark, אתה thou, ואשׁתך and thy wife, ובניך and thy sons, ונשׁי wives בניך and thy sons' אתך׃ with}%
\verse{כל thee every החיה living thing אשׁר that אתך with מכל thee, of all בשׂר flesh, בעוף of fowl, ובבהמה and of cattle, ובכל and of every הרמשׂ creeping thing הרמשׂ that creepeth על upon הארץ the earth; הוצא Bring forth אתך with ושׁרצו that they may breed abundantly בארץ in the earth, ופרו and be fruitful, ורבו and multiply על upon הארץ׃ the earth.}%
\verse{ויצא went forth, נח And Noah ובניו and his sons, ואשׁתו and his wife, ונשׁי wives בניו and his sons' אתו׃ with}%
\verse{כל Every החיה beast, כל every הרמשׂ creeping thing, וכל and every העוף fowl, כל whatsoever רומשׂ creepeth על upon הארץ the earth, למשׁפחתיהם after their kinds, יצאו went forth מן out of התבה׃ the ark.}%
\verse{ויבן built נח And Noah מזבח an altar ליהוה unto the LORD; ויקח and took מכל of every הבהמה beast, הטהרה clean ומכל and of every העוף fowl, הטהור clean ויעל and offered עלת burnt offerings במזבח׃ on the altar.}%
\verse{וירח smelled יהוה And the LORD את ריח savor; הניחח a sweet ויאמר said יהוה and the LORD אל in לבו his heart, לא I will not אסף again לקלל curse עוד any more את האדמה the ground בעבור האדם of man's כי for יצר the imagination לב heart האדם רע evil מנעריו from his youth; ולא neither אסף will I again עוד any more להכות smite את כל every thing חי living, כאשׁר as עשׂיתי׃ I have done.}%
\verse{עד While כל remaineth, ימי remaineth, הארץ the earth זרע seedtime וקציר and harvest, וקר and cold וחם and heat, וקיץ and summer וחרף and winter, ויום and day ולילה and night לא shall not ישׁבתו׃ cease.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 9
\verseWithHeading{God’s Covenant with Noah and Humankind}{ויברך blessed אלהים And God את נח Noah ואת בניו and his sons, ויאמר and said להם פרו unto them, Be fruitful, ורבו and multiply, ומלאו and replenish את הארץ׃ the earth.}%
\verse{ומוראכם And the fear וחתכם of you and the dread יהיה of you shall be על upon כל every חית beast הארץ of the earth, ועל and upon כל every עוף fowl השׁמים of the air, בכל upon all אשׁר that תרמשׂ moveth האדמה the earth, ובכל and upon all דגי the fishes הים of the sea; בידכם into your hand נתנו׃ are they delivered.}%
\verse{כל Every רמשׂ moving thing אשׁר that הוא חי liveth לכם יהיה shall be לאכלה meat כירק for you; even as the green עשׂב herb נתתי have I given לכם את כל׃ you all things.}%
\verse{אך But בשׂר flesh בנפשׁו with the life דמו thereof, the blood לא thereof, shall ye not תאכלו׃ eat.}%
\verse{ואך And surely את דמכם your blood לנפשׁתיכם of your lives אדרשׁ will I require; מיד at the hand כל of every חיה beast אדרשׁנו will I require ומיד it, and at the hand האדם of man; מיד at the hand אישׁ of every man's אחיו brother אדרשׁ will I require את נפשׁ the life האדם׃ of man.}%
\verse{שׁפך Whoso sheddeth דם blood, האדם man's באדם by man דמו shall his blood ישׁפך be shed: כי for בצלם in the image אלהים of God עשׂה made את האדם׃ he man.}%
\verse{ואתם And you, פרו be ye fruitful, ורבו and multiply; שׁרצו bring forth abundantly בארץ in the earth, ורבו׃ and multiply}%
\verse{ויאמר spoke אלהים And God אל unto נח Noah, ואל and to בניו his sons אתו with לאמר׃ him, saying,}%
\verse{ואני And I, הנני מקים I establish את בריתי my covenant אתכם with ואת you, and with זרעכם your seed אחריכם׃ after}%
\verse{ואת And with כל every נפשׁ creature החיה living אשׁר that אתכם with בעוף you, of the fowl, בבהמה of the cattle, ובכל and of every חית beast הארץ of the earth אתכם with מכל you; from all יצאי that go out התבה of the ark, לכל to every חית beast הארץ׃ of the earth.}%
\verse{והקמתי And I will establish את בריתי my covenant אתכם with ולא you; neither יכרת be cut off כל shall all בשׂר flesh עוד any more ממי by the waters המבול of a flood; ולא neither יהיה be עוד shall there any more מבול a flood לשׁחת to destroy הארץ׃ the earth.}%
\verse{ויאמר said, אלהים And God זאת This אות the token הברית of the covenant אשׁר which אני I נתן make ביני between וביניכם וביןל me and you and every נפשׁ creature חיה living אשׁר that אתכם with לדרת generations: עולם׃ you, for perpetual}%
\verse{את קשׁתי my bow נתתי I do set בענן in the cloud, והיתה and it shall be לאות for a token ברית of a covenant ביני between ובין הארץ׃ me and the earth.}%
\verse{והיה And it shall come to pass, בענני when I bring ענן a cloud על over הארץ the earth, ונראתה shall be seen הקשׁת that the bow בענן׃ in the cloud:}%
\verse{וזכרתי And I will remember את בריתי my covenant, אשׁר which ביני between וביניכם וביןל me and you and every נפשׁ creature חיה living בכל of all בשׂר flesh; ולא shall no יהיה become עוד more המים and the waters למבול a flood לשׁחת to destroy כל all בשׂר׃ flesh.}%
\verse{והיתה shall be הקשׁת And the bow בענן in the cloud; וראיתיה and I will look upon לזכר it, that I may remember ברית covenant עולם the everlasting בין between אלהים God ובין כל and every נפשׁ creature חיה living בכל of all בשׂר flesh אשׁר that על upon הארץ׃ the earth.}%
\verse{ויאמר said אלהים And God אל unto נח Noah, זאת This אות the token הברית of the covenant, אשׁר which הקמתי I have established ביני between ובין כל me and all בשׂר flesh אשׁר that על upon הארץ׃ the earth.}%
\verseWithHeading{Noah’s Descendants}{ויהיו were בני And the sons נח of Noah, היצאים that went forth מן of התבה the ark, שׁם Shem, וחם and Ham, ויפת and Japheth: וחם and Ham הוא אבי the father כנען׃ of Canaan.}%
\verse{שׁלשׁה the three אלה These בני sons נח of Noah: ומאלה and of them נפצה overspread. כל was the whole הארץ׃ earth}%
\verse{ויחל began נח And Noah אישׁ האדמה an husbandman, ויטע and he planted כרם׃ a vineyard:}%
\verse{וישׁת And he drank מן of היין the wine, וישׁכר and was drunken; ויתגל and he was uncovered בתוך within אהלה׃ his tent.}%
\verse{וירא saw חם And Ham, אבי the father כנען of Canaan, את ערות the nakedness אביו of his father, ויגד and told לשׁני his two אחיו brethren בחוץ׃ without.}%
\verse{ויקח took שׁם And Shem ויפת and Japheth את השׂמלה a garment, וישׂימו and laid על upon שׁכם their shoulders, שׁניהם both וילכו and went אחרנית backward, ויכסו and covered את ערות the nakedness אביהם of their father; ופניהם and their faces אחרנית backward, וערות nakedness. אביהם their father's לא not ראו׃ and they saw}%
\verse{וייקץ awoke נח And Noah מיינו from his wine, וידע and knew את אשׁר what עשׂה had done לו בנו son הקטן׃ his younger}%
\verse{ויאמר And he said, ארור Cursed כנען Canaan; עבד a servant עבדים of servants יהיה shall he be לאחיו׃ unto his brethren.}%
\verse{ויאמר And he said, ברוך Blessed יהוה the LORD אלהי God שׁם of Shem; ויהי shall be כנען and Canaan עבד׃ his servant.}%
\verse{יפת shall enlarge אלהים God ליפת Japheth, וישׁכן and he shall dwell באהלי in the tents שׁם of Shem; ויהי shall be כנען and Canaan עבד׃ his servant.}%
\verse{ויחי lived נח And Noah אחר after המבול the flood שׁלשׁ three מאות hundred שׁנה years. וחמשׁים and fifty שׁנה׃}%
\verse{ויהי were כל And all ימי the days נח of Noah תשׁע nine מאות hundred שׁנה years: וחמשׁים and fifty שׁנה וימת׃ and he died.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 10
\verseWithHeading{The Descendants of the Sons of Noah}{ואלה Now these תולדת the generations בני of the sons נח of Noah, שׁם Shem, חם Ham, ויפת and Japheth: ויולדו born להם בנים and unto them were sons אחר after המבול׃ the flood.}%
\verse{בני The sons יפת of Japheth; גמר Gomer, ומגוג and Magog, ומדי and Madai, ויון and Javan, ותבל and Tubal, ומשׁך and Meshech, ותירס׃ and Tiras.}%
\verse{ובני And the sons גמר of Gomer; אשׁכנז Ashkenaz, וריפת and Riphath, ותגרמה׃ and Togarmah.}%
\verse{ובני And the sons יון of Javan; אלישׁה Elishah, ותרשׁישׁ and Tarshish, כתים Kittim, ודדנים׃ and Dodanim.}%
\verse{מאלה נפרדו divided איי הגוים of the Gentiles בארצתם in their lands; אישׁ every one ללשׁנו after his tongue, למשׁפחתם after their families, בגויהם׃ in their nations.}%
\verse{ובני And the sons חם of Ham; כושׁ Cush, ומצרים and Mizraim, ופוט and Phut, וכנען׃ and Canaan.}%
\verse{ובני And the sons כושׁ of Cush; סבא Seba, וחוילה and Havilah, וסבתה and Sabtah, ורעמה and Raamah, וסבתכא and Sabtecha: ובני and the sons רעמה of Raamah; שׁבא Sheba, ודדן׃ and Dedan.}%
\verse{וכושׁ And Cush ילד begot את נמרד Nimrod: הוא he החל began להיות to be גבר a mighty one בארץ׃ in the earth.}%
\verse{הוא He היה was גבר a mighty ציד hunter לפני before יהוה the LORD: על wherefore כן wherefore יאמר it is said, כנמרד Even as Nimrod גבור the mighty ציד hunter לפני before יהוה׃ the LORD.}%
\verse{ותהי was ראשׁית And the beginning ממלכתו of his kingdom בבל Babel, וארך and Erech, ואכד and Accad, וכלנה and Calneh, בארץ in the land שׁנער׃ of Shinar.}%
\verse{מן Out of הארץ land ההוא that יצא went forth אשׁור Asshur, ויבן and built את נינוה Nineveh, ואת רחבת Rehoboth, עיר and the city ואת כלח׃ and Calah,}%
\verse{ואת רסן And Resen בין between נינוה Nineveh ובין כלח and Calah: הוא the same העיר city. הגדלה׃ a great}%
\verse{ומצרים And Mizraim ילד begot את לודים Ludim, ואת ענמים and Anamim, ואת להבים and Lehabim, ואת נפתחים׃ and Naphtuhim,}%
\verse{ואת פתרסים And Pathrusim, ואת כסלחים and Casluhim, אשׁר יצאו came משׁם פלשׁתים Philistim,) ואת כפתרים׃ and Caphtorim.}%
\verse{וכנען And Canaan ילד begot את צידן Sidon בכרו his firstborn, ואת חת׃ and Heth,}%
\verse{ואת היבוסי And the Jebusite, ואת האמרי and the Amorite, ואת הגרגשׁי׃ and the Girgasite,}%
\verse{ואת החוי And the Hivite, ואת הערקי and the Arkite, ואת הסיני׃ and the Sinite,}%
\verse{ואת הארודי And the Arvadite, ואת הצמרי and the Zemarite, ואת החמתי and the Hamathite: ואחר and afterward נפצו spread abroad. משׁפחות were the families הכנעני׃ of the Canaanites}%
\verse{ויהי was גבול And the border הכנעני of the Canaanites מצידן באכה as thou comest גררה to Gerar, עד unto עזה Gaza; באכה as thou goest, סדמה unto Sodom, ועמרה and Gomorrah, ואדמה and Admah, וצבים and Zeboim, עד even unto לשׁע׃ Lasha.}%
\verse{אלה These בני the sons חם of Ham, למשׁפחתם after their families, ללשׁנתם after their tongues, בארצתם in their countries, בגויהם׃ in their nations.}%
\verse{ולשׁם Unto Shem ילד were born. גם even הוא to him אבי also the father כל of all בני the children עבר of Eber, אחי the brother יפת of Japheth הגדול׃ the elder,}%
\verse{בני The children שׁם of Shem; עילם Elam, ואשׁור and Asshur, וארפכשׁד and Arphaxad, ולוד and Lud, וארם׃ and Aram.}%
\verse{ובני And the children ארם of Aram; עוץ Uz, וחול and Hul, וגתר and Gether, ומשׁ׃ and Mash.}%
\verse{וארפכשׁד And Arphaxad ילד begot את שׁלח Salah; ושׁלח and Salah ילד begot את עבר׃ Eber.}%
\verse{ולעבר And unto Eber ילד were born שׁני two בנים sons: שׁם the name האחד of one פלג Peleg; כי for בימיו in his days נפלגה divided; הארץ was the earth ושׁם name אחיו and his brother's יקטן׃ Joktan.}%
\verse{ויקטן And Joktan ילד begot את אלמודד Almodad, ואת שׁלף and Sheleph, ואת חצרמות and Hazarmaveth, ואת ירח׃ and Jerah,}%
\verse{ואת הדורם And Hadoram, ואת אוזל and Uzal, ואת דקלה׃ and Diklah,}%
\verse{ואת עובל And Obal, ואת אבימאל and Abimael, ואת שׁבא׃ and Sheba,}%
\verse{ואת אופר And Ophir, ואת חוילה and Havilah, ואת יובב and Jobab: כל all אלה these בני the sons יקטן׃ of Joktan.}%
\verse{ויהי was מושׁבם And their dwelling ממשׁא באכה as thou goest ספרה unto Sephar הר a mount הקדם׃ of the east.}%
\verse{אלה These בני the sons שׁם of Shem, למשׁפחתם after their families, ללשׁנתם after their tongues, בארצתם in their lands, לגויהם׃ after their nations.}%
\verse{אלה These משׁפחת the families בני of the sons נח of Noah, לתולדתם after their generations, בגויהם in their nations: ומאלה and by these נפרדו divided הגוים were the nations בארץ in the earth אחר after המבול׃ the flood.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 11
\verseWithHeading{The Tower of Babel}{ויהי was כל And the whole הארץ earth שׂפה language, אחת of one ודברים speech. אחדים׃ and of one}%
\verse{ויהי And it came to pass, בנסעם as they journeyed מקדם from the east, וימצאו that they found בקעה a plain בארץ in the land שׁנער of Shinar; וישׁבו and they dwelt שׁם׃ there.}%
\verse{ויאמרו And they said אישׁ one אל to רעהו another, הבה Go to, נלבנה let us make לבנים brick, ונשׂרפה and burn לשׂרפה them throughly. ותהי And they had להם הלבנה brick לאבן for stone, והחמר and slime היה had להם לחמר׃ they for mortar.}%
\verse{ויאמרו And they said, הבה Go to, נבנה let us build לנו עיר us a city ומגדל and a tower, וראשׁו whose top בשׁמים unto heaven; ונעשׂה and let us make לנו שׁם us a name, פן lest נפוץ we be scattered abroad על upon פני the face כל of the whole הארץ׃ earth.}%
\verse{וירד came down יהוה And the LORD לראת to see את העיר the city ואת המגדל and the tower, אשׁר which בנו built. בני the children האדם׃ of men}%
\verse{ויאמר said, יהוה And the LORD הן Behold, עם the people אחד one, ושׂפה language; אחת one לכלם and they have all וזה and this החלם they begin לעשׂות to do: ועתה and now לא nothing יבצר will be restrained מהם כל nothing אשׁר them, which יזמו they have imagined לעשׂות׃ to do.}%
\verse{הבה Go to, נרדה let us go down, ונבלה confound שׁם and there שׂפתם their language, אשׁר that לא they may not ישׁמעו understand אישׁ שׂפת speech. רעהו׃ another's}%
\verse{ויפץ יהוה So the LORD אתם משׁם from thence על upon פני the face כל of all הארץ the earth: ויחדלו and they left off לבנת to build העיר׃ the city.}%
\verse{על upon כן קרא of it called שׁמה is the name בבל Babel; כי because שׁם did there בלל confound יהוה the LORD שׂפת the language כל of all הארץ the earth: ומשׁם and from thence הפיצם scatter them abroad יהוה did the LORD על פני the face כל of all הארץ׃ the earth.}%
\verseWithHeading{The Descendants of Shem}{אלה These תולדת the generations שׁם of Shem: שׁם Shem בן old, מאת a hundred שׁנה years ויולד and begot את ארפכשׁד Arphaxad שׁנתים two years אחר after המבול׃ the flood:}%
\verse{ויחי lived שׁם And Shem אחרי after הולידו he begot את ארפכשׁד Arphaxad חמשׁ five מאות hundred שׁנה years, ויולד and begot בנים sons ובנות׃ and daughters.}%
\verse{וארפכשׁד And Arphaxad חי חמשׁ five ושׁלשׁים and thirty שׁנה years, ויולד and begot את שׁלח׃ Salah:}%
\verse{ויחי lived ארפכשׁד And Arphaxad אחרי after הולידו he begot את שׁלח Salah שׁלשׁ and three שׁנים years, וארבע four מאות hundred שׁנה ויולד and begot בנים sons ובנות׃ and daughters.}%
\verse{ושׁלח And Salah חי שׁלשׁים thirty שׁנה years, ויולד and begot את עבר׃ Eber:}%
\verse{ויחי lived שׁלח And Salah אחרי after הולידו he begot את עבר Eber שׁלשׁ and three שׁנים years, וארבע four מאות hundred שׁנה ויולד and begot בנים sons ובנות׃ and daughters.}%
\verse{ויחי lived עבר And Eber ארבע four ושׁלשׁים and thirty שׁנה years, ויולד and begot את פלג׃ Peleg:}%
\verse{ויחי lived עבר And Eber אחרי after הולידו he begot את פלג Peleg שׁלשׁים and thirty שׁנה years, וארבע four מאות hundred שׁנה ויולד and begot בנים sons ובנות׃ and daughters.}%
\verse{ויחי lived פלג And Peleg שׁלשׁים thirty שׁנה years, ויולד and begot את רעו׃ Reu:}%
\verse{ויחי lived פלג And Peleg אחרי after הולידו he begot את רעו Reu תשׁע and nine שׁנים years, ומאתים two hundred שׁנה ויולד and begot בנים sons ובנות׃ and daughters.}%
\verse{ויחי lived רעו And Reu שׁתים two ושׁלשׁים and thirty שׁנה years, ויולד and begot את שׂרוג׃ Serug:}%
\verse{ויחי lived רעו And Reu אחרי after הולידו he begot את שׂרוג Serug שׁבע and seven שׁנים years, ומאתים two hundred שׁנה ויולד and begot בנים sons ובנות׃ and daughters.}%
\verse{ויחי lived שׂרוג And Serug שׁלשׁים thirty שׁנה years, ויולד and begot את נחור׃ Nahor:}%
\verse{ויחי lived שׂרוג And Serug אחרי after הולידו he begot את נחור Nahor מאתים two hundred שׁנה years, ויולד and begot בנים sons ובנות׃ and daughters.}%
\verse{ויחי lived נחור And Nahor תשׁע nine ועשׂרים and twenty שׁנה years, ויולד and begot את תרח׃ Terah:}%
\verse{ויחי lived נחור And Nahor אחרי after הולידו he begot את תרח Terah תשׁע and nineteen עשׂרה and nineteen שׁנה years, ומאת a hundred שׁנה ויולד and begot בנים sons ובנות׃ and daughters.}%
\verse{ויחי lived תרח And Terah שׁבעים seventy שׁנה years, ויולד and begot את אברם Abram, את נחור Nahor, ואת הרן׃ and Haran.}%
\verseWithHeading{The Descendants of Terah}{ואלה Now these תולדת the generations תרח of Terah: תרח Terah הוליד begot את אברם Abram, את נחור Nahor, ואת הרן and Haran; והרן and Haran הוליד begot את לוט׃ Lot.}%
\verse{וימת died הרן And Haran על before פני before תרח Terah אביו his father בארץ in the land מולדתו of his nativity, באור in Ur כשׂדים׃ of the Chaldees.}%
\verse{ויקח took אברם And Abram ונחור and Nahor להם נשׁים them wives: שׁם the name אשׁת wife אברם of Abram's שׂרי Sarai; ושׁם and the name אשׁת wife, נחור of Nahor's מלכה Milcah, בת the daughter הרן of Haran, אבי the father מלכה of Milcah, ואבי and the father יסכה׃ of Iscah.}%
\verse{ותהי was שׂרי But Sarai עקרה barren; אין she no לה ולד׃ child.}%
\verse{ויקח took תרח And Terah את אברם Abram בנו his son, ואת לוט and Lot בן the son הרן of Haran בן his son's בנו son, ואת שׂרי and Sarai כלתו his daughter-in-law, אשׁת wife; אברם Abram's בנו his son ויצאו and they went forth אתם with מאור כשׂדים of the Chaldees, ללכת to go ארצה into the land כנען of Canaan; ויבאו and they came עד unto חרן Haran, וישׁבו and dwelt שׁם׃ there.}%
\verse{ויהיו were ימי And the days תרח of Terah חמשׁ and five שׁנים years: ומאתים two hundred שׁנה וימת died תרח and Terah בחרן׃ in Haran.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 12
\verseWithHeading{The Call of Abram}{ויאמר had said יהוה Now the LORD אל unto אברם Abram, לך Get thee out לך מארצך of thy country, וממולדתך and from thy kindred, ומבית אביךל unto הארץ a land אשׁר that אראך׃ I will show}%
\verse{ואעשׂך And I will make לגוי nation, גדול of thee a great ואברכך and I will bless ואגדלה thee, and make thy name great; שׁמך thee, and make thy name great; והיה and thou shalt be ברכה׃ a blessing:}%
\verse{ואברכה And I will bless מברכיך them that bless ומקללך him that curseth אאר thee, and curse ונברכו be blessed. בך כל thee: and in thee shall all משׁפחת families האדמה׃ of the earth}%
\verseWithHeading{Abram’s Journey}{וילך departed, אברם So Abram כאשׁר as דבר had spoken אליו unto יהוה the LORD וילך went אתו with לוט him; and Lot ואברם him: and Abram בן old חמשׁ and five שׁנים years ושׁבעים seventy שׁנה בצאתו when he departed מחרן׃}%
\verse{ויקח took אברם And Abram את שׂרי Sarai אשׁתו his wife, ואת לוט and Lot בן son, אחיו his brother's ואת כל and all רכושׁם their substance אשׁר that רכשׁו they had gathered, ואת הנפשׁ and the souls אשׁר that עשׂו they had gotten בחרן in Haran; ויצאו and they went forth ללכת to go ארצה into the land כנען of Canaan; ויבאו they came. ארצה and into the land כנען׃ of Canaan}%
\verse{ויעבר passed through אברם And Abram בארץ the land עד unto מקום the place שׁכם of Sichem, עד unto אלון the plain מורה of Moreh. והכנעני And the Canaanite אז then בארץ׃ in the land.}%
\verse{וירא appeared יהוה And the LORD אל unto אברם Abram, ויאמר and said, לזרעך Unto thy seed אתן will I give את הארץ land: הזאת this ויבן built שׁם and there מזבח he an altar ליהוה unto the LORD, הנראה who appeared אליו׃ unto}%
\verse{ויעתק And he removed משׁם from thence ההרה unto a mountain מקדם on the east לבית אל of Bethel, ויט and pitched אהלה his tent, בית אל Bethel מים on the west, והעי and Hai מקדם on the east: ויבן he built שׁם and there מזבח an altar ליהוה unto the LORD, ויקרא and called בשׁם upon the name יהוה׃ of the LORD.}%
\verse{ויסע journeyed, אברם And Abram הלוך going on ונסוע still הנגבה׃ toward the south.}%
\verse{ויהי And there was רעב a famine בארץ in the land: וירד went down אברם and Abram מצרימה into Egypt לגור to sojourn שׁם there; כי for כבד grievous הרעב the famine בארץ׃ in the land.}%
\verse{ויהי And it came to pass, כאשׁר when הקריב he was come near לבוא to enter מצרימה into Egypt, ויאמר that he said אל unto שׂרי Sarai אשׁתו his wife, הנה Behold נא now, ידעתי I know כי that אשׁה woman יפת a fair מראה to look upon: את׃ thou}%
\verse{והיה Therefore it shall come to pass, כי when יראו shall see אתך המצרים the Egyptians ואמרו thee, that they shall say, אשׁתו his wife: זאת This והרגו and they will kill אתי ואתךחיו׃ me, but they will save thee alive.}%
\verse{אמרי Say, נא I pray thee, אחתי my sister: את thou למען that ייטב it may be well לי בעבורך with me for thy sake; וחיתה shall live נפשׁי and my soul בגללך׃ because of}%
\verse{ויהי And it came to pass, כבוא was come אברם that, when Abram מצרימה into Egypt, ויראו beheld המצרים the Egyptians את האשׁה the woman כי that יפה fair. הוא she מאד׃ very}%
\verse{ויראו saw אתה שׂרי The princes פרעה also of Pharaoh ויהללו her, and commended אתה אל her before פרעה Pharaoh: ותקח was taken האשׁה and the woman בית house. פרעה׃ into Pharaoh's}%
\verse{ולאברם היטיבעבורה for her sake: ויהי and he had לו צאן sheep, ובקר and oxen, וחמרים and he asses, ועבדים and menservants, ושׁפחת and maidservants, ואתנת and she asses, וגמלים׃ and camels.}%
\verse{וינגע plagued יהוה And the LORD את פרעה Pharaoh נגעים plagues גדלים with great ואת ביתו and his house על because of דבר because of שׂרי Sarai אשׁת wife. אברם׃ Abram's}%
\verse{ויקרא called פרעה And Pharaoh לאברם Abram, ויאמר and said, מה What זאת this עשׂית thou hast done לי למה unto me? why לא didst thou not הגדת tell לי כי me that אשׁתך thy wife? הוא׃ she}%
\verse{למה Why אמרת saidst אחתי my sister? הוא thou, She ואקח so I might have taken אתה לי לאשׁה her to me to wife: ועתה now therefore הנה behold אשׁתך thy wife, קח take ולך׃ and go thy way.}%
\verse{ויצו commanded עליו concerning פרעה And Pharaoh אנשׁים men וישׁלחו אתואת אשׁתו and his wife, ואת כל and all אשׁר׃ that}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 13
\verseWithHeading{The Parting of Abram and Lot}{ויעל went up אברם And Abram ממצרים הוא he, ואשׁתו and his wife, וכל and all אשׁר that לו ולוט he had, and Lot עמו with הנגבה׃ him, into the south.}%
\verse{ואברם And Abram כבד מאד very במקנה in cattle, בכסף in silver, ובזהב׃ and in gold.}%
\verse{וילך And he went למסעיו on his journeys מנגב from the south ועד even to בית אל Bethel, עד unto המקום the place אשׁר where היה had been שׁם where אהלה his tent בתחלה at the beginning, בין between בית אל Bethel ובין העי׃ and Hai;}%
\verse{אל Unto מקום the place המזבח of the altar, אשׁר which עשׂה he had made שׁם there בראשׁנה at the first: ויקרא called שׁם and there אברם Abram בשׁם on the name יהוה׃ of the LORD.}%
\verse{וגם also, ללוט And Lot ההלך which went את with אברם Abram, היה had צאן flocks, ובקר and herds, ואהלים׃ and tents.}%
\verse{ולא was not נשׂא able to bear אתם הארץ And the land לשׁבת them, that they might dwell יחדו together: כי for היה was רכושׁם their substance רב great, ולא not יכלו so that they could לשׁבת dwell יחדו׃ together.}%
\verse{ויהי And there was ריב a strife בין between רעי מקנה cattle אברם of Abram's ובין רעיקנה cattle: לוט of Lot's והכנעני and the Canaanite והפרזי and the Perizzite אז then ישׁב dwelled בארץ׃ in the land.}%
\verse{ויאמר said אברם And Abram אל unto לוט Lot, אל no נא I pray thee, תהי Let there be מריבה strife, ביני between ובינך me and thee, and between ובין רעיבין רעיךי for אנשׁים אחים brethren. אנחנו׃ we}%
\verse{הלא not כל the whole הארץ land לפניך before הפרד thee? separate thyself, נא I pray thee, מעלי from אם me: if השׂמאל the left hand, ואימנה then I will go to the right; ואם or if הימין to the right hand, ואשׂמאילה׃ then I will go to the left.}%
\verse{וישׂא lifted up לוט And Lot את עיניו his eyes, וירא and beheld את כל all ככר the plain הירדן of Jordan, כי that כלה every where, משׁקה it well watered לפני before שׁחת destroyed יהוה the LORD את סדם Sodom ואת עמרה and Gomorrah, כגן as the garden יהוה of the LORD, כארץ like the land מצרים of Egypt, באכה as thou comest צער׃ unto Zoar.}%
\verse{ויבחר chose לו לוט Then Lot את כל him all ככר the plain הירדן of Jordan; ויסע journeyed לוט and Lot מקדם east: ויפרדו and they separated themselves אישׁ the one מעל from אחיו׃ the other.}%
\verse{אברם Abram ישׁב dwelled בארץ in the land כנען of Canaan, ולוט and Lot ישׁב dwelled בערי in the cities הככר of the plain, ויאהל and pitched tent עד toward סדם׃ Sodom.}%
\verse{ואנשׁי But the men סדם of Sodom רעים wicked וחטאים and sinners ליהוה before the LORD מאד׃ exceedingly.}%
\verse{ויהוה And the LORD אמר said אל unto אברם Abram, אחרי after that הפרד was separated לוט Lot מעמו from שׂא him, Lift up נא now עיניך thine eyes, וראה and look מן from המקום the place אשׁר where אתה thou שׁם where צפנה art northward, ונגבה and southward, וקדמה and eastward, וימה׃ and westward:}%
\verse{כי For את כל all הארץ the land אשׁר which אתה thou ראה seest, לך אתננה to thee will I give ולזרעך it, and to thy seed עד forever. עולם׃ forever.}%
\verse{ושׂמתי And I will make את זרעך thy seed כעפר as the dust הארץ of the earth: אשׁר so that אם if יוכל can אישׁ למנות number את עפר the dust הארץ of the earth, גם also זרעך shall thy seed ימנה׃ be numbered.}%
\verse{קום Arise, התהלך walk בארץ through the land לארכה in the length ולרחבה of it and in the breadth כי of it; for לך אתננה׃ I will give}%
\verse{ויאהל removed tent, אברם Then Abram ויבא and came וישׁב and dwelt באלני in the plain ממרא of Mamre, אשׁר which בחברון in Hebron, ויבן and built שׁם there מזבח an altar ליהוה׃ unto the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 14
\verseWithHeading{Abram Rescues Lot}{ויהי And it came to pass בימי in the days אמרפל of Amraphel מלך king שׁנער of Shinar, אריוך Arioch מלך king אלסר of Ellasar, כדרלעמר Chedorlaomer מלך king עילם of Elam, ותדעל and Tidal מלך king גוים׃ of nations;}%
\verse{עשׂו made מלחמה war את with ברע Bera מלך king סדם of Sodom, ואת and with ברשׁע Birsha מלך king עמרה of Gomorrah, שׁנאב Shinab מלך king אדמה of Admah, ושׁמאבר and Shemeber מלך king צביים of Zeboiim, ומלך and the king בלע of Bela, היא which צער׃ is Zoar.}%
\verse{כל All אלה these חברו were joined together אל in עמק the vale השׂדים of Siddim, הוא which ים sea. המלח׃ is the salt}%
\verse{שׁתים עשׂרה and in the thirteenth שׁנה years עבדו they served את כדרלעמר Chedorlaomer, ושׁלשׁ and in the thirteenth עשׂרה שׁנה year מרדו׃ they rebelled.}%
\verse{ובארבע עשׂרהׁנה year בא came כדרלעמר Chedorlaomer, והמלכים and the kings אשׁר that אתו with ויכו him, and smote את רפאים the Rephaims בעשׁתרת קרנים in Ashteroth Karnaim, ואת הזוזים and the Zuzims בהם in Ham, ואת האימים and the Emims בשׁוה קריתים׃ in Shaveh Kiriathaim,}%
\verse{ואת החרי And the Horites בהררם שׂעיר Seir, עד unto איל פארן El-paran, אשׁר which על by המדבר׃ the wilderness.}%
\verse{וישׁבו And they returned, ויבאו and came אל to עין משׁפט En-mishpat, הוא which קדשׁ Kadesh, ויכו and smote את כל all שׂדה the country העמלקי of the Amalekites, וגם and also את האמרי the Amorites, הישׁב that dwelt בחצצן תמר׃ in Hazezon-tamar.}%
\verse{ויצא And there went out מלך the king סדם of Sodom, ומלך and the king עמרה of Gomorrah, ומלך and the king אדמה of Admah, ומלך and the king צביים of Zeboiim, ומלך and the king בלע of Bela הוא (the same צער Zoar) ויערכו and they joined אתם with מלחמה battle בעמק them in the vale השׂדים׃ of Siddim;}%
\verse{את With כדרלעמר Chedorlaomer מלך the king עילם of Elam, ותדעל and with Tidal מלך king גוים of nations, ואמרפל and Amraphel מלך king שׁנער of Shinar, ואריוך and Arioch מלך king אלסר of Ellasar; ארבעה four מלכים kings את with החמשׁה׃ five.}%
\verse{ועמק And the vale השׂדים of Siddim בארת slime pits; בארת חמרינסו fled, מלך and the kings סדם of Sodom ועמרה and Gomorrah ויפלו and fell שׁמה there; והנשׁארים and they that remained הרה to the mountain. נסו׃ fled}%
\verse{ויקחו And they took את כל all רכשׁ the goods סדם of Sodom ועמרה and Gomorrah, ואת כל and all אכלם their victuals, וילכו׃ and went their way.}%
\verse{ויקחו And they took את לוט Lot, ואת רכשׁו and his goods, בן son, אחי brother's אברם Abram's וילכו and departed. והוא who ישׁב dwelt בסדם׃ in Sodom,}%
\verse{ויבא And there came הפליט one that had escaped, ויגד and told לאברם Abram העברי the Hebrew; והוא for he שׁכן dwelt באלני in the plain ממרא of Mamre האמרי the Amorite, אחי brother אשׁכל of Eshcol, ואחי and brother ענר of Aner: והם and these בעלי confederate ברית confederate אברם׃ with Abram.}%
\verse{וישׁמע heard אברם And when Abram כי that נשׁבה was taken captive, אחיו his brother וירק he armed את חניכיו his trained ילידי born ביתו in his own house, שׁמנה and eighteen, עשׂר and eighteen, ושׁלשׁ three מאות hundred וירדף and pursued עד unto דן׃ Dan.}%
\verse{ויחלק And he divided himself עליהם against לילה by night, הוא them, he ועבדיו and his servants, ויכם and smote וירדפם them, and pursued עד them unto חובה Hobah, אשׁר which משׂמאל on the left hand לדמשׂק׃ of Damascus.}%
\verse{וישׁב And he brought back את כל all הרכשׁ the goods, וגם and also את לוט Lot, אחיו his brother ורכשׁו and his goods, השׁיב brought again וגם also, את הנשׁים and the women ואת העם׃ and the people.}%
\verseWithHeading{Abram Meets Melchizedek}{ויצא went out מלך And the king סדם of Sodom לקראתו to meet אחרי him after שׁובו his return מהכות from the slaughter את כדר לעמר of Chedorlaomer, ואת המלכים and of the kings אשׁר that אתו with אל him, at עמק the valley שׁוה of Shaveh, הוא which עמק dale. המלך׃ the king's}%
\verse{ומלכי צדק And Melchizedek מלך king שׁלם of Salem הוציא brought forth לחם bread ויין and wine: והוא and he כהן the priest לאל God. עליון׃ of the most high}%
\verse{ויברכהו And he blessed ויאמר him, and said, ברוך Blessed אברם Abram לאל God, עליון of the most high קנה possessor שׁמים of heaven וארץ׃ and earth:}%
\verse{וברוך And blessed אל God, עליון be the most high אשׁר which מגן hath delivered צריך thine enemies בידך into thy hand. ויתן And he gave לו מעשׂר him tithes מכל׃ of all.}%
\verse{ויאמר said מלך And the king סדם of Sodom אל unto אברם Abram, תן Give לי הנפשׁ me the persons, והרכשׁ the goods קח׃ and take}%
\verse{ויאמר said אברם And Abram אל to מלך the king סדם of Sodom, הרמתי I have lifted up ידי mine hand אל unto יהוה the LORD, אל God, עליון the most high קנה the possessor שׁמים of heaven וארץ׃ and earth,}%
\verse{אם That I will not מחוט from a thread ועד even to שׂרוך a shoelatchet, נעל a shoelatchet, ואם and that I will not אקח take מכל any thing אשׁר that לך ולא thine, lest תאמר thou shouldest say, אני I העשׁרתי אתברם׃}%
\verse{בלעדי Save רק only אשׁר that which אכלו have eaten, הנערים the young men וחלק and the portion האנשׁים of the men אשׁר which הלכו went אתי with ענר me, Aner, אשׁכל Eshcol, וממרא and Mamre; הם let them יקחו take חלקם׃ their portion.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 15
\verseWithHeading{Adonai’s Covenant with Abram}{אחר After הדברים things האלה these היה came דבר the word יהוה of the LORD אל unto אברם Abram במחזה in a vision, לאמר saying, אל not, תירא Fear אברם Abram: אנכי I מגן thy shield, לך שׂכרך reward. הרבה great מאד׃ thy exceeding}%
\verse{ויאמר said, אברם And Abram אדני Lord יהוה GOD, מה what תתן wilt thou give לי ואנכי me, seeing I הולך go ערירי childless, ובן and the steward משׁק and the steward ביתי of my house הוא this דמשׂק of Damascus? אליעזר׃ Eliezer}%
\verse{ויאמר said, אברם And Abram הן Behold, לי לא no נתתה to me thou hast given זרע seed: והנה and, lo, בן one born ביתי in my house יורשׁ is mine heir. אתי׃}%
\verse{והנה And, behold, דבר the word יהוה of the LORD אליו unto לאמר him, saying, לא shall not יירשׁך be thine heir; זה This כי but אם but אשׁר he that יצא shall come forth ממעיך out of thine own bowels הוא יירשׁך׃ shall be thine heir.}%
\verse{ויוצא אתוחוצה abroad, ויאמר and said, הבט Look נא now השׁמימה toward heaven, וספר and tell הכוכבים the stars, אם if תוכל thou be able לספר to number אתם ויאמר them: and he said לו כה unto him, So יהיה be. זרעך׃ shall thy seed}%
\verse{והאמן And he believed ביהוה in the LORD; ויחשׁבה and he counted לו צדקה׃ it to him for righteousness.}%
\verse{ויאמר And he said אליו unto אני him, I יהוה the LORD אשׁר that הוצאתיך brought מאור כשׂדים of the Chaldees, לתת to give לך את הארץ land הזאת thee this לרשׁתה׃ to inherit}%
\verse{ויאמר And he said, אדני Lord יהוה במה whereby אדע shall I know כי that אירשׁנה׃ I shall inherit}%
\verse{ויאמר And he said אליו unto קחה him, Take לי עגלה me an heifer משׁלשׁת of three years old, ועז and a she goat משׁלשׁת of three years old, ואיל and a ram משׁלשׁ of three years old, ותר and a turtledove, וגוזל׃ and a young pigeon.}%
\verse{ויקח And he took לו את כל unto him all אלה these, ויבתר and divided אתם בתוך them in the midst, ויתן and laid אישׁ each בתרו piece לקראת one against רעהו another: ואת הצפר but the birds לא he not. בתר׃ divided}%
\verse{וירד came down העיט And when the fowls על upon הפגרים the carcasses, וישׁב אתםברם׃ Abram}%
\verse{ויהי השׁמשׁ And when the sun לבוא was going down, ותרדמה a deep sleep נפלה fell על upon אברם Abram; והנה and, lo, אימה a horror חשׁכה darkness גדלה of great נפלת fell עליו׃ upon}%
\verse{ויאמר And he said לאברם unto Abram, ידע תדעי that גר a stranger יהיה shall be זרעך thy seed בארץ in a land לא not להם ועבדום theirs, and shall serve וענו them; and they shall afflict אתם ארבע them four מאות hundred שׁנה׃ years;}%
\verse{וגם And also את הגוי that nation, אשׁר whom יעבדו they shall serve, דן judge: אנכי will I ואחרי and afterward כן and afterward יצאו shall they come out ברכשׁ substance. גדול׃ with great}%
\verse{ואתה And thou תבוא shalt go אל to אבתיך thy fathers בשׁלום in peace; תקבר thou shalt be buried בשׂיבה old age. טובה׃ in a good}%
\verse{ודור generation רביעי But in the fourth ישׁובו they shall come hither again: הנה they shall come hither again: כי for לא not שׁלם full. עון the iniquity האמרי of the Amorites עד yet הנה׃ yet}%
\verse{ויהי And it came to pass, השׁמשׁ that, when the sun באה went down, ועלטה dark, היה and it was והנה behold תנור furnace, עשׁן a smoking ולפיד lamp אשׁ and a burning אשׁר that עבר passed בין between הגזרים pieces. האלה׃ those}%
\verse{ביום day ההוא In the same כרת made יהוה the LORD את with אברם Abram, ברית a covenant לאמר saying, לזרעך Unto thy seed נתתי have I given את הארץ land, הזאת this מנהר from the river מצרים of Egypt עד unto הנהר river, הגדל the great נהר the river פרת׃ Euphrates:}%
\verse{את הקיני The Kenites, ואת הקנזי and the Kenizzites, ואת הקדמני׃ and the Kadmonites,}%
\verse{ואת החתי And the Hittites, ואת הפרזי and the Perizzites, ואת הרפאים׃ and the Rephaims,}%
\verse{ואת האמרי And the Amorites, ואת הכנעני and the Canaanites, ואת הגרגשׁי and the Girgashites, ואת היבוסי׃ and the Jebusites.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 16
\verseWithHeading{Sarai and Hagar}{ושׂרי Now Sarai אשׁת wife אברם Abram's לא bore him no children: ילדה bore him no children: לו ולה שׁפחה and she had an handmaid, מצרית an Egyptian, ושׁמה whose name הגר׃ Hagar.}%
\verse{ותאמר said שׂרי And Sarai אל unto אברם Abram, הנה Behold נא now, עצרני hath restrained יהוה the LORD מלדת me from bearing: בא go in נא I pray thee, אל unto שׁפחתי my maid; אולי it may be אבנה that I may obtain children ממנה me from bearing: וישׁמע hearkened אברם her. And Abram לקול to the voice שׂרי׃ of Sarai.}%
\verse{ותקח took שׂרי And Sarai אשׁת wife אברם Abram's את הגר Hagar המצרית the Egyptian, שׁפחתה her maid מקץ after עשׂר ten שׁנים years לשׁבת had dwelt אברם Abram בארץ in the land כנען of Canaan, ותתן and gave אתה לאברם Abram אישׁה לו לאשׁה׃ to be his wife.}%
\verse{ויבא And he went in אל unto הגר Hagar, ותהר and she conceived: ותרא and when she saw כי that הרתה she had conceived, ותקל was despised גברתה her mistress בעיניה׃ in her eyes.}%
\verse{ותאמר said שׂרי And Sarai אל unto אברם Abram, חמסי My wrong עליך upon אנכי thee: I נתתי have given שׁפחתי my maid בחיקך into thy bosom; ותרא and when she saw כי that הרתה she had conceived, ואקל I was despised בעיניה in her eyes: ישׁפט judge יהוה the LORD ביני between וביניך׃}%
\verse{ויאמר said אברם But Abram אל unto שׂרי Sarai, הנה Behold, שׁפחתך thy maid בידך in thy hand; עשׂי do לה הטוב to her as it pleaseth בעיניך to her as it pleaseth ותענה dealt hardly שׂרי thee. And when Sarai ותברח with her, she fled מפניה׃ from her face.}%
\verseWithHeading{Hagar and the Angel of Adonai}{וימצאה found מלאך And the angel יהוה of the LORD על her by עין a fountain המים of water במדבר in the wilderness, על by העין the fountain בדרך in the way שׁור׃ to Shur.}%
\verse{ויאמר And he said, הגר Hagar, שׁפחת maid, שׂרי Sarai's אי whence מזה באת camest ואנה thou? and whither תלכי wilt thou go? ותאמר And she said, מפני from the face שׂרי Sarai. גברתי of my mistress אנכי I ברחת׃ flee}%
\verse{ויאמר said לה מלאך And the angel יהוה of the LORD שׁובי unto her, Return אל to גברתך thy mistress, והתעני and submit thyself תחת under ידיה׃ her hands.}%
\verse{ויאמר said לה מלאך And the angel יהוה of the LORD הרבה ארבהת זרעךלא that it shall not יספר be numbered מרב׃ for multitude.}%
\verse{ויאמר said לה מלאך And the angel יהוה of the LORD הנך unto her, Behold, הרה thou with child, וילדת and shalt bear בן a son, וקראת and shalt call שׁמו his name ישׁמעאל Ishmael; כי because שׁמע hath heard יהוה the LORD אל hath heard עניך׃ thy affliction.}%
\verse{והוא And he יהיה will be פרא a wild אדם man; ידו his hand בכל against every man, ויד hand כל and every man's בו ועל in the presence פני in the presence כל of all אחיו his brethren. ישׁכן׃ against him; and he shall dwell}%
\verse{ותקרא And she called שׁם the name יהוה of the LORD הדבר that spoke אליה unto אתה her, Thou אל God ראי seest כי me: for אמרה she said, הגם Have I also הלם here ראיתי looked אחרי after ראי׃ him that seeth}%
\verse{על כןרא was called לבאר the well באר לחי ראי Beer-lahai-roi; הנה behold, בין between קדשׁ Kadesh ובין ברד׃ and Bered.}%
\verse{ותלד bore הגר And Hagar לאברם Abram בן a son: ויקרא called אברם and Abram שׁם name, בנו his son's אשׁר which ילדה bore, הגר Hagar ישׁמעאל׃ Ishmael.}%
\verse{ואברם And Abram בן old, שׁמנים fourscore שׁנה years ושׁשׁ and six שׁנים בלדת bore הגר when Hagar את ישׁמעאל Ishmael לאברם׃ to Abram.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 17
\verseWithHeading{Abram and Circumcision, the Sign of the Covenant}{ויהי was אברם And when Abram בן old תשׁעים ninety שׁנה years ותשׁע and nine, שׁנים וירא appeared יהוה the LORD אל to אברם Abram, ויאמר and said אליו unto אני him, I אל God; שׁדי the Almighty התהלך walk לפני before והיה me, and be תמים׃ thou perfect.}%
\verse{ואתנה And I will make בריתי my covenant ביני between ובינך וארבה me and thee, and will multiply אותך במאד thee exceedingly. מאד׃ thee exceedingly.}%
\verse{ויפל fell אברם And Abram על on פניו his face: וידבר talked אתו with אלהים and God לאמר׃ him, saying,}%
\verse{אני As for me, הנה behold, בריתי my covenant אתך with והיית thee, and thou shalt be לאב a father המון of many גוים׃ nations.}%
\verse{ולא Neither יקרא be called עוד any more את שׁמך shall thy name אברם Abram, והיה shall be שׁמך but thy name אברהם Abraham; כי for אב a father המון of many גוים nations נתתיך׃ have I made}%
\verse{והפרתי fruitful, אתך במאדאד ונתתיך and I will make לגוים nations ומלכים of thee, and kings ממך of יצאו׃ shall come out}%
\verse{והקמתי And I will establish את בריתי my covenant ביני between ובינך וביןרעך me and thee and thy seed אחריך after לדרתם thee in their generations לברית covenant, עולם for an everlasting להיות to be לך לאלהים a God ולזרעך unto thee, and to thy seed אחריך׃ after}%
\verse{ונתתי And I will give לך ולזרעך unto thee, and to thy seed אחריך after את ארץ thee, the land מגריך wherein thou art a stranger, את כל all ארץ the land כנען of Canaan, לאחזת possession; עולם for an everlasting והייתי and I will be להם לאלהים׃ their God.}%
\verse{ויאמר said אלהים And God אל unto אברהם Abraham, ואתה Thou את בריתי my covenant תשׁמר shalt keep אתה therefore, thou, וזרעך and thy seed אחריך after לדרתם׃ thee in their generations.}%
\verse{זאת This בריתי my covenant, אשׁר which תשׁמרו ye shall keep, ביני between וביניכם וביןרעך me and you and thy seed אחריך after המול among you shall be circumcised. לכם כל thee; Every זכר׃ man child}%
\verse{ונמלתם אתשׂר the flesh ערלתכם of your foreskin; והיה and it shall be לאות a token ברית of the covenant ביני between וביניכם׃}%
\verse{ובן old שׁמנת And he that is eight ימים days ימול shall be circumcised לכם כל among you, every זכר man child לדרתיכם in your generations, יליד he that is born בית in the house, ומקנת or bought כסף with money מכל of any בן stranger, נכר stranger, אשׁר which לא not מזרעך of thy seed. הוא׃}%
\verse{המול must needs be circumcised: ימול must needs be circumcised: יליד He that is born ביתך in thy house, ומקנת and he that is bought כספך with thy money, והיתה shall be בריתי and my covenant בבשׂרכם in your flesh לברית covenant. עולם׃ for an everlasting}%
\verse{וערל And the uncircumcised זכר man child אשׁר whose לא is not ימול circumcised, את בשׂר flesh ערלתו of his foreskin ונכרתה shall be cut off הנפשׁ soul ההוא that מעמיה from his people; את בריתי my covenant. הפר׃ he hath broken}%
\verse{ויאמר said אלהים And God אל unto אברהם Abraham, שׂרי As for Sarai אשׁתך thy wife, לא thou shalt not תקרא call את שׁמה her name שׂרי Sarai, כי but שׂרה Sarah שׁמה׃ her name}%
\verse{וברכתי And I will bless אתה וגם also נתתי her, and give ממנה of לך בן thee a son וברכתיה her: yea, I will bless והיתה her, and she shall be לגוים of nations; מלכי kings עמים of people ממנה of יהיו׃ shall be}%
\verse{ויפל fell אברהם Then Abraham על upon פניו his face, ויצחק and laughed, ויאמר and said בלבו in his heart, הלבן old? מאה unto him that is a hundred שׁנה years יולד Shall be born ואם שׂרה and shall Sarah, הבת old, תשׁעים that is ninety שׁנה years תלד׃ bear?}%
\verse{ויאמר said אברהם And Abraham אל unto האלהים God, לו O that ישׁמעאל Ishmael יחיה might live לפניך׃ before}%
\verse{ויאמר said, אלהים And God אבל indeed; שׂרה Sarah אשׁתך thy wife ילדת shall bear לך בן thee a son וקראת and thou shalt call את שׁמו his name יצחק Isaac: והקמתי and I will establish את בריתי my covenant אתו with לברית covenant, עולם him for an everlasting לזרעו with his seed אחריו׃ after}%
\verse{ולישׁמעאל And as for Ishmael, שׁמעתיך I have heard הנה thee: Behold, ברכתי I have blessed אתו והפריתיתו והרביתי and will multiply אתו במאד him exceedingly; מאד him exceedingly; שׁנים twelve עשׂר twelve נשׂיאם princes יוליד shall he beget, ונתתיו and I will make לגוי nation. גדול׃ him a great}%
\verse{ואת בריתי But my covenant אקים will I establish את with יצחק Isaac, אשׁר which תלד shall bear לך שׂרה Sarah למועד set time הזה unto thee at this בשׁנה year. האחרת׃ in the next}%
\verse{ויכל And he left off לדבר talking אתו with ויעל went up אלהים him, and God מעל from אברהם׃ Abraham.}%
\verse{ויקח took אברהם And Abraham את ישׁמעאל Ishmael בנו his son, ואת כל and all ילידי that were born ביתו in his house, ואת כל and all מקנת that were bought כספו with his money, כל every זכר male באנשׁי among the men בית house; אברהם of Abraham's וימל and circumcised את בשׂר the flesh ערלתם of their foreskin בעצם in the selfsame היום day, הזה in the selfsame כאשׁר as דבר had said אתו unto אלהים׃ God}%
\verse{ואברהם And Abraham בן old תשׁעים ninety ותשׁע and nine, שׁנה years בהמלו when he was circumcised בשׂר in the flesh ערלתו׃ of his foreskin.}%
\verse{וישׁמעאל And Ishmael בנו his son בן old, שׁלשׁ thirteen עשׂרה thirteen שׁנה years בהמלו when he was circumcised את בשׂר in the flesh ערלתו׃ of his foreskin.}%
\verse{בעצם היום day הזה נמול circumcised, אברהם was Abraham וישׁמעאל and Ishmael בנו׃ his son.}%
\verse{וכל And all אנשׁי the men ביתו of his house, יליד born בית in the house, ומקנת and bought כסף with money מאת בן the stranger, נכר the stranger, נמלו were circumcised אתו׃ of}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 18
\verseWithHeading{Adonai Appears to Abraham as a Man}{וירא appeared אליו unto יהוה And the LORD באלני him in the plains ממרא of Mamre: והוא and he ישׁב sat פתח door האהל in the tent כחם in the heat היום׃ of the day;}%
\verse{וישׂא And he lifted up עיניו his eyes וירא and looked, והנה and, lo, שׁלשׁה three אנשׁים men נצבים stood עליו by וירא him: and when he saw וירץ he ran לקראתם to meet מפתח האהלישׁתחו and bowed himself ארצה׃ toward the ground,}%
\verse{ויאמר And said, אדני אם if נא now מצאתי I have found חן favor בעיניך in thy sight, אל pass not away, נא I pray thee, תעבר pass not away, מעל from עבדך׃ thy servant:}%
\verse{יקח be fetched, נא I pray you, מעט Let a little מים water, ורחצו and wash רגליכם your feet, והשׁענו and rest yourselves תחת under העץ׃ the tree:}%
\verse{ואקחה And I will fetch פת a morsel לחם of bread, וסעדו and comfort לבכם ye your hearts; אחר after that תעברו ye shall pass on: כי for על therefore כן therefore עברתם are ye come על to עבדכם your servant. ויאמרו And they said, כן So תעשׂה do, כאשׁר as דברת׃ thou hast said.}%
\verse{וימהר hastened אברהם And Abraham האהלה into the tent אל unto שׂרה Sarah, ויאמר and said, מהרי Make ready quickly שׁלשׁ three סאים measures קמח meal, סלת of fine לושׁי knead ועשׂי and make עגות׃ cakes upon the hearth.}%
\verse{ואל unto הבקר the herd, רץ ran אברהם And Abraham ויקח and fetched בן a calf בקר a calf רך tender וטוב and good, ויתן and gave אל unto הנער a young man; וימהר and he hasted לעשׂות to dress אתו׃}%
\verse{ויקח And he took חמאה butter, וחלב and milk, ובן and the calf הבקר and the calf אשׁר which עשׂה he had dressed, ויתן and set לפניהם before והוא them; and he עמד stood עליהם by תחת them under העץ the tree, ויאכלו׃ and they did eat.}%
\verse{ויאמרו And they said אליו unto איה him, Where שׂרה Sarah אשׁתך thy wife? ויאמר And he said, הנה Behold, באהל׃ in the tent.}%
\verse{ויאמר And he said, שׁוב אשׁובליך unto כעת thee according to the time חיה of life; והנה and, lo, בן shall have a son. לשׂרה Sarah אשׁתך thy wife ושׂרה And Sarah שׁמעת heard פתח door, האהל in the tent והוא which אחריו׃ behind}%
\verse{ואברהם Now Abraham ושׂרה and Sarah זקנים old באים well stricken בימים in age; חדל it ceased להיות to be לשׂרה with Sarah ארח after the manner כנשׁים׃ of women.}%
\verse{ותצחק laughed שׂרה Therefore Sarah בקרבה within לאמר herself, saying, אחרי After בלתי I am waxed old היתה shall I have לי עדנה pleasure, ואדני my lord זקן׃ being old}%
\verse{ויאמר said יהוה And the LORD אל unto אברהם Abraham, למה זהחקה laugh, שׂרה did Sarah לאמר saying, האף אמנםלד bear a child, ואני which זקנתי׃ am old?}%
\verse{היפלא too hard מיהוה דבר Is any thing למועד At the time appointed אשׁוב I will return אליך unto כעת thee, according to the time חיה of life, ולשׂרה and Sarah בן׃ shall have a son.}%
\verse{ותכחשׁ denied, שׂרה Then Sarah לאמר saying, לא not; צחקתי I laughed כי for יראה she was afraid. ויאמר And he said, לא Nay; כי but צחקת׃ thou didst laugh.}%
\verse{ויקמו rose up משׁם from thence, האנשׁים וישׁקפו and looked על toward פני toward סדם Sodom: ואברהם and Abraham הלך went עמם with לשׁלחם׃ them to bring them on the way.}%
\verse{ויהוה And the LORD אמר said, המכסה hide אני Shall I מאברהם אשׁר that thing which אני I עשׂה׃ do;}%
\verse{ואברהם Seeing that Abraham היו shall surely become יהיה shall surely become לגוי nation, גדול a great ועצום and mighty ונברכו shall be blessed בו כל and all גויי the nations הארץ׃ of the earth}%
\verse{כי For ידעתיו I know למען him, that אשׁר him, that יצוה he will command את בניו his children ואת ביתו and his household אחריו after ושׁמרו him, and they shall keep דרך the way יהוה of the LORD, לעשׂות to do צדקה justice ומשׁפט and judgment; למען that הביא may bring יהוה the LORD על upon אברהם Abraham את אשׁר that which דבר he hath spoken עליו׃ of}%
\verse{ויאמר said, יהוה And the LORD זעקת the cry סדם of Sodom ועמרה and Gomorrah כי Because רבה וחטאתם their sin כי and because כבדה grievous; מאד׃ is very}%
\verse{ארדה I will go down נא now, ואראה and see הכצעקתה according to the cry הבאה of it, which is come אלי unto עשׂו whether they have done כלה altogether ואם me; and if לא not, אדעה׃ I will know.}%
\verseWithHeading{Abraham Intercedes for Sodom}{ויפנו turned their faces משׁם from thence, האנשׁים And the men וילכו and went סדמה toward Sodom: ואברהם but Abraham עודנו yet עמד stood לפני before יהוה׃ the LORD.}%
\verse{ויגשׁ drew near, אברהם And Abraham ויאמר and said, האף Wilt thou also תספה destroy צדיק the righteous עם with רשׁע׃ the wicked?}%
\verse{אולי Peradventure ישׁ there be חמשׁים fifty צדיקם righteous בתוך within העיר the city: האף wilt thou also תספה destroy ולא and not תשׂא spare למקום the place למען for חמשׁים the fifty הצדיקם righteous אשׁר that בקרבה׃ therein?}%
\verse{חללה That be far לך מעשׂת from thee to do כדבר manner, הזה after this להמית to slay צדיק the righteous עם with רשׁע the wicked: והיה should be כצדיק and that the righteous כרשׁע as the wicked, חללה that be far לך השׁפט the Judge כל of all הארץ the earth לא from thee: Shall not יעשׂה do משׁפט׃ right?}%
\verse{ויאמר said, יהוה And the LORD אם If אמצא I find בסדם in Sodom חמשׁים fifty צדיקם righteous בתוך within העיר the city, ונשׂאתי then I will spare לכל all המקום the place בעבורם׃ for their sakes.}%
\verse{ויען answered אברהם And Abraham ויאמר and said, הנה Behold נא now, הואלתי I have taken upon לדבר me to speak אל unto אדני the Lord, ואנכי which עפר dust ואפר׃ and ashes:}%
\verse{אולי Peradventure יחסרון there shall lack חמשׁים of the fifty הצדיקם righteous: חמשׁה five התשׁחית wilt thou destroy בחמשׁה for five? את כל all העיר the city ויאמר And he said, לא I will not אשׁחית destroy אם If אמצא I find שׁם there ארבעים forty וחמשׁה׃ and five,}%
\verse{ויסף again, עוד him yet לדבר And he spoke אליו unto ויאמר and said, אולי Peradventure ימצאון found שׁם there. ארבעים there shall be forty ויאמר And he said, לא I will not אעשׂה do בעבור הארבעים׃}%
\verse{ויאמר And he said אל let not נא Oh יחר be angry, לאדני the Lord ואדברה and I will speak: אולי Peradventure ימצאון be found שׁם there. שׁלשׁים there shall thirty ויאמר And he said, לא I will not אעשׂה do אם if אמצא I find שׁם there. שׁלשׁים׃ thirty}%
\verse{ויאמר And he said, הנה Behold נא now, הואלתי I have taken upon לדבר me to speak אל unto אדני the Lord: אולי Peradventure ימצאון found שׁם there. עשׂרים there shall be twenty ויאמר And he said, לא I will not אשׁחית destroy בעבור העשׂרים׃}%
\verse{ויאמר And he said, אל let not נא Oh יחר be angry, לאדני the Lord ואדברה and I will speak אך yet הפעם but this once: אולי Peradventure ימצאון shall be found שׁם there. עשׂרה ten ויאמר And he said, לא I will not אשׁחית destroy בעבור העשׂרה׃}%
\verse{וילך went his way, יהוה And the LORD כאשׁר as soon as כלה he had left לדבר communing אל with אברהם Abraham: ואברהם and Abraham שׁב returned למקמו׃ unto his place.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 19
\verseWithHeading{The Rescue of Lot from Sodom}{ויבאו And there came שׁני two המלאכים angels סדמה to Sodom בערב at even; ולוט and Lot ישׁב sat בשׁער in the gate סדם of Sodom: וירא seeing לוט and Lot ויקם rose up לקראתם to meet וישׁתחו them; and he bowed himself אפים with his face ארצה׃ toward the ground;}%
\verse{ויאמר And he said, הנה Behold נא now, אדני my lords, סורו turn in, נא I pray you, אל into בית house, עבדכם your servant's ולינו and tarry all night, ורחצו and wash רגליכם your feet, והשׁכמתם and ye shall rise up early, והלכתם and go לדרככם on your ways. ויאמרו And they said, לא Nay; כי but ברחוב we will abide in the street all night. נלין׃ we will abide in the street all night.}%
\verse{ויפצר And he pressed בם מאד upon them greatly; ויסרו and they turned in אליו unto ויבאו him, and entered אל into ביתו his house; ויעשׂ and he made להם משׁתה them a feast, ומצות unleavened bread, אפה and did bake ויאכלו׃ and they did eat.}%
\verse{טרם But before ישׁכבו they lay down, ואנשׁי העיר of the city, אנשׁי סדם of Sodom, נסבו compassed על compassed הבית the house מנער and young, ועד זקן round, both old כל all העם the people מקצה׃}%
\verse{ויקראו And they called אל unto לוט Lot, ויאמרו and said לו איה unto him, Where האנשׁים the men אשׁר which באו came in אליך to הלילה thee this night? הוציאם bring them out אלינו unto ונדעה us, that we may know אתם׃}%
\verse{ויצא went out אלהם unto לוט And Lot הפתחה at the door והדלת the door סגר them, and shut אחריו׃ after}%
\verse{ויאמר And said, אל do not so wickedly. נא I pray you, אחי brethren, תרעו׃ do not so wickedly.}%
\verse{הנה Behold נא now, לי שׁתי I have two בנות daughters אשׁר which לא have not ידעו known אישׁ man; אוציאה נא let me, I pray you, אתהן אליכם unto ועשׂו you, and do להן כטוב ye to them as good בעיניכם in your eyes: רק only לאנשׁים men האל unto these אל nothing; תעשׂו do דבר nothing; כי for על therefore כן therefore באו came בצל they under the shadow קרתי׃ of my roof.}%
\verse{ויאמרו And they said, גשׁ Stand הלאה back. ויאמרו And they said האחד This one בא came in לגור to sojourn, וישׁפט and he will needs be a judge: שׁפוט and he will needs be a judge: עתה now נרע will we deal worse לך מהם ויפצרו with them. And they pressed באישׁ בלוט Lot, מאד sore ויגשׁו and came near לשׁבר to break הדלת׃ the door.}%
\verse{וישׁלחו put forth האנשׁים But the men את ידם their hand, ויביאו and pulled את לוט Lot אליהם to הביתה into the house ואת הדלת the door. סגרו׃ them, and shut to}%
\verse{ואת האנשׁים the men אשׁר that פתח at the door הבית of the house הכו And they smote בסנורים with blindness, מקטן both small ועד גדול and great: וילאו so that they wearied themselves למצא to find הפתח׃ the door.}%
\verse{ויאמרו said האנשׁים And the men אל unto לוט Lot, עד besides מי any לך פה Hast thou here חתן a son-in-law, ובניך and thy sons, ובנתיך and thy daughters? וכל and whatsoever אשׁר and whatsoever לך בעיר thou hast in the city, הוצא bring מן out of המקום׃ this place:}%
\verse{כי For משׁחתים will destroy אנחנו we את המקום place, הזה this כי because גדלה of them is waxen great צעקתם the cry את פני before the face יהוה of the LORD; וישׁלחנו hath sent יהוה and the LORD לשׁחתה׃ us to destroy}%
\verse{ויצא went out, לוט And Lot וידבר and spoke אל unto חתניו his sons-in-law, לקחי which married בנתיו his daughters, ויאמר and said, קומו Up, צאו get you out מן of המקום place; הזה this כי for משׁחית will destroy יהוה the LORD את העיר this city. ויהי But he seemed כמצחק as one that mocked בעיני unto חתניו׃ his sons-in-law.}%
\verse{וכמו And when השׁחר the morning עלה arose, ויאיצו hastened המלאכים then the angels בלוט Lot, לאמר saying, קום Arise, קח take את אשׁתך thy wife, ואת שׁתי and thy two בנתיך daughters, הנמצאת which are here; פן lest תספה thou be consumed בעון in the iniquity העיר׃ of the city.}%
\verse{ויתמהמה And while he lingered, ויחזיקו laid hold האנשׁים the men בידו upon his hand, וביד and upon the hand אשׁתו of his wife, וביד and upon the hand שׁתי of his two בנתיו daughters; בחמלת being merciful יהוה the LORD עליו unto ויצאהו him: and they brought him forth, וינחהו מחוץ him without לעיר׃ the city.}%
\verse{ויהי And it came to pass, כהוציאם אתםחוצה abroad, ויאמר that he said, המלט Escape על for נפשׁך thy life; אל not תביט look אחריך behind ואל thee, neither תעמד stay בכל thou in all הככר the plain; ההרה to the mountain, המלט escape פן lest תספה׃ thou be consumed.}%
\verse{ויאמר said לוט And Lot אלהם unto אל not נא them, Oh, אדני׃ so, my Lord:}%
\verse{הנה Behold נא now, מצא hath found עבדך thy servant חן grace בעיניך in thy sight, ותגדל and thou hast magnified חסדך thy mercy, אשׁר which עשׂית thou hast showed עמדי להחיות me in saving את נפשׁי my life; ואנכי and I לא cannot אוכל cannot להמלט escape ההרה to the mountain, פן lest תדבקני take הרעה some evil ומתי׃ me, and I die:}%
\verse{הנה Behold נא now, העיר city הזאת this קרבה near לנוס to flee שׁמה unto, והוא and it מצער a little one: אמלטה let me escape נא Oh, שׁמה thither, הלא not מצער a little one?) הוא (it ותחי shall live. נפשׁי׃ and my soul}%
\verse{ויאמר And he said אליו unto הנה him, See, נשׂאתי פניךם also, לדבר thing הזה thee concerning this לבלתי that I will not הפכי overthrow את העיר this city, אשׁר for the which דברת׃ thou hast spoken.}%
\verse{מהר Haste המלט thee, escape שׁמה thither; כי for לא אוכלעשׂות do דבר any thing עד till באך thou be come שׁמה thither. על כןרא was called שׁם the name העיר of the city צוער׃ Zoar.}%
\verseWithHeading{The Destruction of Sodom}{השׁמשׁ The sun יצא was risen על upon הארץ the earth ולוט when Lot בא entered צערה׃ into Zoar.}%
\verse{ויהוה Then the LORD המטיר rained על upon סדם Sodom ועל and upon עמרה Gomorrah גפרית brimstone ואשׁ and fire מאת יהוה the LORD מן from השׁמים׃ heaven;}%
\verse{ויהפך And he overthrew את הערים cities, האל those ואת כל and all הככר the plain, ואת כל and all ישׁבי the inhabitants הערים of the cities, וצמח and that which grew האדמה׃ upon the ground.}%
\verse{ותבט looked אשׁתו But his wife מאחריו back from behind ותהי him, and she became נציב a pillar מלח׃ of salt.}%
\verse{וישׁכם got up early אברהם And Abraham בבקר in the morning אל to המקום the place אשׁר where עמד he stood שׁם where את פני before יהוה׃ the LORD:}%
\verse{וישׁקף And he looked על toward פני toward סדם Sodom ועמרה and Gomorrah, ועל and toward כל all פני and toward ארץ the land הככר of the plain, וירא and beheld, והנה and, lo, עלה went up קיטר the smoke הארץ of the country כקיטר as the smoke הכבשׁן׃ of a furnace.}%
\verse{ויהי And it came to pass, בשׁחת destroyed אלהים when God את ערי the cities הככר of the plain, ויזכר remembered אלהים that God את אברהם Abraham, וישׁלח and sent את לוט Lot מתוך out of the midst ההפכה of the overthrow, בהפך when he overthrew את הערים the cities אשׁר in the which ישׁב dwelt. בהן in the which לוט׃ Lot}%
\verseWithHeading{Lot and His Daughters}{ויעל went up לוט And Lot מצוער in Zoar: וישׁב and dwelt בהר in the mountain, ושׁתי and his two בנתיו daughters עמו with כי him; for ירא he feared לשׁבת to dwell בצוער וישׁב and he dwelt במערה in a cave, הוא he ושׁתי and his two בנתיו׃ daughters.}%
\verse{ותאמר said הבכירה And the firstborn אל unto הצעירה the younger, אבינו Our father זקן old, ואישׁ a man אין and not בארץ in the earth לבוא to come in עלינו unto כדרך us after the manner כל of all הארץ׃ the earth:}%
\verse{לכה Come, נשׁקה drink את אבינו let us make our father יין wine, ונשׁכבה and we will lie עמו with ונחיה him, that we may preserve מאבינו of our father. זרע׃ seed}%
\verse{ותשׁקין drink את אביהן And they made their father יין wine בלילה night: הוא that ותבא went in, הבכירה and the firstborn ותשׁכב and lay את with אביה her father; ולא not ידע and he perceived בשׁכבה when she lay down, ובקומה׃ nor when she arose.}%
\verse{ויהי And it came to pass ממחרת on the morrow, ותאמר said הבכירה that the firstborn אל unto הצעירה the younger, הן Behold, שׁכבתי I lay אמשׁ last night את with אבי my father: נשׁקנו let us make him drink יין wine גם also; הלילה this night ובאי and go thou in, שׁכבי lie עמו with ונחיה him, that we may preserve מאבינו of our father. זרע׃ seed}%
\verse{ותשׁקין drink גם also: בלילה night ההוא that את אביהן And they made their father יין wine ותקם arose, הצעירה and the younger ותשׁכב and lay עמו with ולא not ידע him; and he perceived בשׁכבה when she lay down, ובקמה׃ nor when she arose.}%
\verse{ותהרין with child שׁתי Thus were both בנות the daughters לוט of Lot מאביהן׃ by their father.}%
\verse{ותלד bore הבכירה And the firstborn בן a son, ותקרא and called שׁמו his name מואב Moab: הוא the same אבי the father מואב of the Moabites עד unto היום׃ this day.}%
\verse{והצעירה And the younger, גם also הוא she ילדה bore בן a son, ותקרא and called שׁמו his name בן עמי Ben-ammi: הוא the same אבי the father בני of the children עמון of Ammon עד unto היום׃ this day.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 20
\verseWithHeading{Abraham and Abimelech}{ויסע journeyed משׁם from thence אברהם And Abraham ארצה country, הנגב toward the south וישׁב and dwelled בין between קדשׁ Kadesh ובין שׁור and Shur, ויגר and sojourned בגרר׃ in Gerar.}%
\verse{ויאמר said אברהם And Abraham אל of שׂרה Sarah אשׁתו his wife, אחתי my sister: הוא She וישׁלח sent, אבימלך and Abimelech מלך king גרר of Gerar ויקח and took את שׂרה׃ Sarah.}%
\verse{ויבא came אלהים But God אל to אבימלך Abimelech בחלום in a dream הלילה by night, ויאמר and said לו הנך to him, Behold, מת thou a dead man, על for האשׁה the woman אשׁר which לקחת thou hast taken; והוא for she בעלת wife. בעל׃ a man's}%
\verse{ואבימלך But Abimelech לא had not קרב come near אליה come near ויאמר her: and he said, אדני Lord, הגוי nation? גם also צדיק a righteous תהרג׃ wilt thou slay}%
\verse{הלא not הוא he אמר Said לי אחתי my sister? הוא unto me, She והיא and she, גם even הוא she herself אמרה said, אחי my brother: הוא He בתם in the integrity לבבי of my heart ובנקין and innocency כפי of my hands עשׂיתי have I done זאת׃ this.}%
\verse{ויאמר said אליו unto האלהים And God בחלם him in a dream, גם Yea, אנכי I ידעתי know כי that בתם in the integrity לבבך of thy heart; עשׂית thou didst זאת this ואחשׂך withheld גם also אנכי for I אותך מחטו thee from sinning לי על against me: therefore כן against me: therefore לא I thee not נתתיך suffered לנגע to touch אליה׃ to touch}%
\verse{ועתה Now השׁב therefore restore אשׁת wife; האישׁ the man כי for נביא a prophet, הוא he ויתפלל and he shall pray בעדך for וחיה thee, and thou shalt live: ואם and if אינך not, משׁיב thou restore דע know כי thou that מות thou shalt surely die, תמות thou shalt surely die, אתה thou, וכל and all אשׁר׃ that}%
\verse{וישׁכם rose early אבימלך Therefore Abimelech בבקר in the morning, ויקרא and called לכל all עבדיו his servants, וידבר and told את כל all הדברים things האלה these באזניהם in their ears: וייראו were sore afraid. האנשׁים and the men מאד׃ were sore afraid.}%
\verse{ויקרא called אבימלך Then Abimelech לאברהם Abraham, ויאמר and said לו מה unto him, What עשׂית hast thou done לנו ומה unto us? and what חטאתי have I offended לך כי thee, that הבאת thou hast brought עלי on ועל me and on ממלכתי my kingdom חטאה sin? גדלה a great מעשׂים deeds אשׁר me that לא ought not יעשׂו thou hast done עשׂית to be done. עמדי׃}%
\verse{ויאמר said אבימלך And Abimelech אל unto אברהם Abraham, מה What ראית sawest כי thou, that עשׂית thou hast done את הדבר thing? הזה׃ this}%
\verse{ויאמר said, אברהם And Abraham כי Because אמרתי I thought, רק Surely אין not יראת the fear אלהים of God במקום place; הזה in this והרגוני and they will slay על me for דבר sake. אשׁתי׃ my wife's}%
\verse{וגם And yet אמנה indeed אחתי my sister; בת the daughter אבי of my father, הוא she אך but לא not בת the daughter אמי of my mother; ותהי and she became לי לאשׁה׃ my wife.}%
\verse{ויהי And it came to pass, כאשׁר when התעו caused me to wander אתי אלהים God מבית אביאמר that I said לה זה unto her, This חסדך thy kindness אשׁר which תעשׂי thou shalt show עמדי אל me; at כל every המקום place אשׁר whither נבוא we shall come, שׁמה whither אמרי say לי אחי my brother. הוא׃ of me, He}%
\verse{ויקח took אבימלך And Abimelech צאן sheep, ובקר and oxen, ועבדים and menservants, ושׁפחת and womenservants, ויתן and gave לאברהם unto Abraham, וישׁב and restored לו את שׂרה him Sarah אשׁתו׃ his wife.}%
\verse{ויאמר said, אבימלך And Abimelech הנה Behold, ארצי my land לפניך before בטוב where it pleaseth בעיניך where it pleaseth שׁב׃ thee: dwell}%
\verse{ולשׂרה And unto Sarah אמר he said, הנה Behold, נתתי I have given אלף a thousand כסף of silver: לאחיך thy brother הנה behold, הוא he לך כסות to thee a covering עינים of the eyes, לכל unto all אשׁר that אתך with ואת thee, and with כל all ונכחת׃ thus she was reproved.}%
\verse{ויתפלל prayed אברהם So Abraham אל unto האלהים God: וירפא healed אלהים and God את אבימלך Abimelech, ואת אשׁתו and his wife, ואמהתיו and his maidservants; וילדו׃ and they bore}%
\verse{כי For עצר had fast closed up עצר יהוה the LORD בעד כל all רחם the wombs לבית of the house אבימלך of Abimelech, על because דבר because שׂרה of Sarah אשׁת wife. אברהם׃ Abraham's}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 21
\verseWithHeading{The Birth of Isaac}{ויהוה And the LORD פקד visited את שׂרה Sarah כאשׁר as אמר he had said, ויעשׂ did יהוה and the LORD לשׂרה unto Sarah כאשׁר as דבר׃ he had spoken.}%
\verse{ותהר conceived, ותלד and bore שׂרה For Sarah לאברהם Abraham בן a son לזקניו in his old age, למועד at the set time אשׁר of which דבר had spoken אתו אלהים׃ God}%
\verse{ויקרא called אברהם And Abraham את שׁם the name בנו of his son הנולד that was born לו אשׁר unto him, whom ילדה bore לו שׂרה Sarah יצחק׃ to him, Isaac.}%
\verse{וימל circumcised אברהם And Abraham את יצחק Isaac בנו his son בן old, שׁמנת being eight ימים days כאשׁר as צוה had commanded אתו אלהים׃ God}%
\verse{ואברהם And Abraham בן old, מאת was a hundred שׁנה years בהולד was born לו את יצחק Isaac בנו׃ when his son}%
\verse{ותאמר said, שׂרה And Sarah צחק me to laugh, עשׂה hath made לי אלהים God כל all השׁמע that hear יצחק׃ will laugh}%
\verse{ותאמר And she said, מי Who מלל would have said לאברהם unto Abraham, היניקה should have given children suck? בנים should have given children suck? שׂרה that Sarah כי for ילדתי I have born בן a son לזקניו׃ in his old age.}%
\verseWithHeading{Hagar and Ishmael}{ויגדל grew, הילד And the child ויגמל and was weaned: ויעשׂ made אברהם and Abraham משׁתה feast גדול a great ביום the day הגמל was weaned. את יצחק׃ that Isaac}%
\verse{ותרא saw שׂרה And Sarah את בן the son הגר of Hagar המצרית the Egyptian, אשׁר which ילדה she had born לאברהם unto Abraham, מצחק׃ mocking.}%
\verse{ותאמר Wherefore she said לאברהם unto Abraham, גרשׁ Cast out האמה bondwoman הזאת this ואת בנה and her son: כי for לא shall not יירשׁ be heir בן the son האמה bondwoman הזאת of this עם with בני my son, עם with יצחק׃ Isaac.}%
\verse{וירע was very grievous הדבר And the thing מאד was very grievous בעיני sight אברהם in Abraham's על because אודת because בנו׃ of his son.}%
\verse{ויאמר said אלהים And God אל unto אברהם Abraham, אל Let it not ירע be grievous בעיניך in thy sight על because of הנער the lad, ועל and because of אמתך thy bondwoman; כל in all אשׁר that תאמר hath said אליך unto שׂרה Sarah שׁמע thee, hearken בקלה unto her voice; כי for ביצחק in Isaac יקרא be called. לך זרע׃ shall thy seed}%
\verse{וגם And also את בן of the son האמה of the bondwoman לגוי a nation, אשׂימנו will I make כי because זרעך thy seed. הוא׃ he}%
\verse{וישׁכם rose up early אברהם And Abraham בבקר in the morning, ויקח and took לחם bread, וחמת and a bottle מים of water, ויתן and gave אל unto הגר Hagar, שׂם putting על on שׁכמה her shoulder, ואת הילד and the child, וישׁלחה and sent her away: ותלך and she departed, ותתע and wandered במדבר in the wilderness באר שׁבע׃ of Beer-sheba.}%
\verse{ויכלו was spent המים And the water מן in החמת the bottle, ותשׁלך and she cast את הילד the child תחת under אחד one השׂיחם׃ of the shrubs.}%
\verse{ותלך And she went, ותשׁב and sat her down לה מנגד over against הרחק a good way off, כמטחוי as it were a bowshot: קשׁת as it were a bowshot: כי for אמרה she said, אל Let me not אראה see במות the death הילד of the child. ותשׁב And she sat מנגד over against ותשׂא and lifted up את קלה her voice, ותבך׃ and wept.}%
\verse{וישׁמע heard אלהים And God את קול the voice הנער of the lad; ויקרא called מלאך and the angel אלהים of God אל to הגר Hagar מן out of השׁמים heaven, ויאמר and said לה מה unto her, What לך הגר aileth thee, Hagar? אל not; תיראי fear כי for שׁמע hath heard אלהים God אל hath heard קול the voice הנער of the lad באשׁר where הוא he שׁם׃ where}%
\verse{קומי Arise, שׂאי lift up את הנער the lad, והחזיקי and hold את ידך thine hand; בו כי for לגוי nation. גדול him a great אשׂימנו׃ I will make}%
\verse{ויפקח opened אלהים And God את עיניה her eyes, ותרא and she saw באר a well מים of water; ותלך and she went, ותמלא and filled את החמת the bottle מים with water, ותשׁק drink. את הנער׃ and gave the lad}%
\verse{ויהי was אלהים And God את with הנער the lad; ויגדל and he grew, וישׁב and dwelt במדבר in the wilderness, ויהי and became רבה קשׁת׃}%
\verse{וישׁב And he dwelt במדבר in the wilderness פארן of Paran: ותקח took לו אמו and his mother אשׁה him a wife מארץ out of the land מצרים׃ of Egypt.}%
\verseWithHeading{The Covenant Between Abraham and Abimelech}{ויהי And it came to pass בעת time, ההוא at that ויאמר spoke אבימלך that Abimelech ופיכל and Phichol שׂר the chief captain צבאו of his host אל unto אברהם Abraham, לאמר saying, אלהים God עמך with בכל thee in all אשׁר that אתה thou עשׂה׃ doest:}%
\verse{ועתה Now השׁבעה therefore swear לי באלהים by God הנה unto me here אם תשׁקר that thou wilt not deal falsely לי ולניני with me, nor with my son, ולנכדי nor with my son's son: כחסד according to the kindness אשׁר that עשׂיתי I have done עמך unto תעשׂה thee, thou shalt do עמדי unto ועם me, and to הארץ the land אשׁר wherein גרתה׃ thou hast sojourned.}%
\verse{ויאמר said, אברהם And Abraham אנכי I אשׁבע׃ will swear.}%
\verse{והוכח reproved אברהם And Abraham את אבימלך Abimelech על because of אדות because of באר a well המים of water, אשׁר which גזלו had violently taken away. עבדי servants אבימלך׃ Abimelech's}%
\verse{ויאמר said, אבימלך And Abimelech לא not ידעתי I know מי who עשׂה hath done את הדבר thing: הזה this וגם yet אתה didst thou לא neither הגדת tell לי וגם אנכי I לא me, neither שׁמעתי heard בלתי but היום׃ today.}%
\verse{ויקח took אברהם And Abraham צאן sheep ובקר and oxen, ויתן and gave לאבימלך them unto Abimelech; ויכרתו of them made שׁניהם and both ברית׃ a covenant.}%
\verse{ויצב set אברהם And Abraham את שׁבע seven כבשׂת ewe lambs הצאן of the flock לבדהן׃ by themselves.}%
\verse{ויאמר said אבימלך And Abimelech אל unto אברהם Abraham, מה What הנה שׁבע seven כבשׂת ewe lambs האלה these אשׁר which הצבת thou hast set לבדנה׃ by themselves?}%
\verse{ויאמר And he said, כי For את שׁבע seven כבשׂת ewe lambs תקח shalt thou take מידי of my hand, בעבור that תהיה they may be לי לעדה a witness כי unto me, that חפרתי I have digged את הבאר well. הזאת׃}%
\verse{על כןרא he called למקום place ההוא that באר שׁבע Beer-sheba; כי because שׁם there נשׁבעו they swore שׁניהם׃ both}%
\verse{ויכרתו Thus they made ברית a covenant בבאר שׁבע at Beer-sheba: ויקם rose up, אבימלך then Abimelech ופיכל and Phichol שׂר the chief captain צבאו of his host, וישׁבו and they returned אל into ארץ the land פלשׁתים׃ of the Philistines.}%
\verse{ויטע And planted אשׁל a grove בבאר שׁבע in Beer-sheba, ויקרא and called שׁם there בשׁם on the name יהוה of the LORD, אל God. עולם׃ the everlasting}%
\verse{ויגר sojourned אברהם And Abraham בארץ land פלשׁתים in the Philistines' ימים days. רבים׃ many}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 22
\verseWithHeading{God Tests Abraham}{ויהי And it came to pass אחר after הדברים things, האלה these והאלהים that God נסה did tempt את אברהם Abraham, ויאמר and said אליו unto אברהם him, Abraham: ויאמר and he said, הנני׃ Behold,}%
\verse{ויאמר And he said, קח Take נא now את בנך thy son, את יחידך thine only אשׁר whom אהבת thou lovest, את יצחק Isaac, ולך and get לך אל thee into ארץ the land המריה of Moriah; והעלהו and offer שׁם him there לעלה for a burnt offering על upon אחד one ההרים of the mountains אשׁר which אמר I will tell אליך׃ thee of.}%
\verse{וישׁכם rose up early אברהם And Abraham בבקר in the morning, ויחבשׁ and saddled את חמרו his ass, ויקח and took את שׁני two נעריו of his young men אתו with ואת יצחק him, and Isaac בנו his son, ויבקע and cleaved עצי the wood עלה for the burnt offering, ויקם and rose up, וילך and went אל unto המקום the place אשׁר of which אמר had told לו האלהים׃ God}%
\verse{ביום day השׁלישׁי Then on the third וישׂא lifted up אברהם Abraham את עיניו his eyes, וירא and saw את המקום the place מרחק׃ afar off.}%
\verse{ויאמר said אברהם And Abraham אל unto נעריו his young men, שׁבו Abide לכם פה ye here עם with החמור the ass; ואני and I והנער and the lad נלכה will go עד yonder כה yonder ונשׁתחוה and worship, ונשׁובה and come again אליכם׃ to}%
\verse{ויקח took אברהם And Abraham את עצי the wood העלה of the burnt offering, וישׂם and laid על upon יצחק Isaac בנו his son; ויקח and he took בידו in his hand, את האשׁ the fire ואת המאכלת and a knife; וילכו and they went שׁניהם both יחדו׃ of them together.}%
\verse{ויאמר spoke יצחק And Isaac אל unto אברהם Abraham אביו his father, ויאמר and said, אבי My father: ויאמר and he said, הנני Here בני I, my son. ויאמר And he said, הנה Behold האשׁ the fire והעצים and the wood: ואיה but where השׂה the lamb לעלה׃ for a burnt offering?}%
\verse{ויאמר said, אברהם And Abraham אלהים God יראה will provide לו השׂה himself a lamb לעלה for a burnt offering: בני My son, וילכו so they went שׁניהם both יחדו׃ of them together.}%
\verse{ויבאו And they came אל to המקום the place אשׁר which אמר had told לו האלהים God ויבן built שׁם there, אברהם him of; and Abraham את המזבח an altar ויערך אתעצים the wood. ויעקד and bound את יצחק Isaac בנו his son, וישׂם and laid אתו על him on המזבח the altar ממעל upon לעצים׃}%
\verse{וישׁלח stretched forth אברהם And Abraham את ידו his hand, ויקח and took את המאכלת the knife לשׁחט to slay את בנו׃ his son.}%
\verse{ויקרא called אליו unto מלאך And the angel יהוה of the LORD מן him out of השׁמים heaven, ויאמר and said, אברהם Abraham, אברהם Abraham: ויאמר and he said, הנני׃ Here}%
\verse{ויאמר And he said, אל not תשׁלח Lay ידך thine hand אל upon הנער the lad, ואל neither תעשׂ do לו מאומה thou any thing כי unto him: for עתה now ידעתי I know כי that ירא fearest אלהים God, אתה thou ולא seeing thou hast not חשׂכת withheld את בנך thy son, את יחידך thine only ממני׃ from}%
\verse{וישׂא lifted up אברהם And Abraham את עיניו his eyes, וירא and looked, והנה and behold איל a ram אחר behind נאחז caught בסבך in a thicket בקרניו by his horns: וילך went אברהם and Abraham ויקח and took את האיל the ram, ויעלהו and offered him up לעלה for a burnt offering תחת in the stead of בנו׃ his son.}%
\verse{ויקרא called אברהם And Abraham שׁם the name המקום place ההוא of that יהוה of the LORD יראה it shall be seen. אשׁר as יאמר it is said היום this day, בהר In the mount יהוה יראה׃}%
\verse{ויקרא called מלאך And the angel יהוה of the LORD אל unto אברהם Abraham שׁנית the second time, מן out of השׁמים׃ heaven}%
\verse{ויאמר And said, בי נשׁבעתי By myself have I sworn, נאם saith יהוה the LORD, כי for יען because אשׁר because עשׂית thou hast done את הדבר thing, הזה this ולא and hast not חשׂכת withheld את בנך thy son, את יחידך׃ thine only}%
\verse{כי That ברך in blessing אברכך I will bless והרבה thee, and in multiplying ארבה I will multiply את זרעך thy seed ככוכבי as the stars השׁמים of the heaven, וכחול and as the sand אשׁר which על upon שׂפת shore; הים the sea וירשׁ shall possess זרעך and thy seed את שׁער the gate איביו׃ of his enemies;}%
\verse{והתברכו be blessed; בזרעך And in thy seed כל shall all גויי the nations הארץ of the earth עקב because אשׁר because שׁמעת thou hast obeyed בקלי׃ my voice.}%
\verse{וישׁב returned אברהם So Abraham אל unto נעריו his young men, ויקמו and they rose up וילכו and went יחדו together אל to באר שׁבע Beer-sheba; וישׁב dwelt אברהם and Abraham בבאר שׁבע׃ at Beer-sheba.}%
\verse{ויהי And it came to pass אחרי after הדברים things, האלה these ויגד that it was told לאברהם Abraham, לאמר saying, הנה Behold, ילדה born מלכה Milcah, גם hath also הוא she בנים children לנחור Nahor; אחיך׃ unto thy brother}%
\verse{את עוץ Huz בכרו his firstborn, ואת בוז and Buz אחיו his brother, ואת קמואל and Kemuel אבי the father ארם׃ of Aram,}%
\verse{ואת כשׂד And Chesed, ואת חזו and Hazo, ואת פלדשׁ and Pildash, ואת ידלף and Jidlaph, ואת בתואל׃ and Bethuel.}%
\verse{ובתואל And Bethuel ילד begot את רבקה Rebekah: שׁמנה eight אלה these ילדה did bear מלכה Milcah לנחור to Nahor, אחי brother. אברהם׃ Abraham's}%
\verse{ופילגשׁו And his concubine, ושׁמה whose name ראומה Reumah, ותלד bore גם also הוא she את טבח Tebah, ואת גחם and Gaham, ואת תחשׁ and Thahash, ואת מעכה׃ and Maachah.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 23
\verseWithHeading{Sarah’s Death and Burial}{ויהיו was חיי old: שׂרה And Sarah מאה a hundred שׁנה years ועשׂרים and twenty שׁנה the years ושׁבע and seven שׁנים שׁנייי of the life שׂרה׃ of Sarah.}%
\verse{ותמת died שׂרה And Sarah בקרית ארבע in Kirjath-arba; הוא the same חברון Hebron בארץ in the land כנען of Canaan: ויבא came אברהם and Abraham לספד to mourn לשׂרה for Sarah, ולבכתה׃ and to weep}%
\verse{ויקם stood up אברהם And Abraham מעל פניתו his dead, וידבר and spoke אל unto בני the sons חת of Heth, לאמר׃ saying,}%
\verse{גר a stranger ותושׁב and a sojourner אנכי I עמכם with תנו you: give לי אחזת me a possession קבר of a burial place עמכם with ואקברה you, that I may bury מתי my dead מלפני׃ out of my sight.}%
\verse{ויענו answered בני And the children חת of Heth את אברהם Abraham, לאמר׃ saying}%
\verse{שׁמענו Hear אדני us, my lord: נשׂיא prince אלהים a mighty אתה thou בתוכנו among במבחר us: in the choice קברינו of our sepulchers קבר bury את מתך thy dead; אישׁ ממנו of את קברו thee his sepulcher, לא none יכלה us shall withhold ממך from מקבר but that thou mayest bury מתך׃ thy dead.}%
\verse{ויקם stood up, אברהם And Abraham וישׁתחו and bowed himself לעם to the people הארץ of the land, לבני to the children חת׃ of Heth.}%
\verse{וידבר And he communed אתם with לאמר them, saying, אם If ישׁ it be את נפשׁכם your mind לקבר that I should bury את מתי my dead מלפני out of my sight; שׁמעוני hear ופגעו me, and entreat לי בעפרון for me to Ephron בן the son צחר׃ of Zohar,}%
\verse{ויתן That he may give לי את מערת me the cave המכפלה of Machpelah, אשׁר which לו אשׁר he hath, which בקצה in the end שׂדהו of his field; בכסף for as much money מלא as it is worth יתננה he shall give לי בתוככם amongst לאחזת it me for a possession קבר׃ of a burial place}%
\verse{ועפרון And Ephron ישׁב dwelt בתוך among בני the children חת of Heth: ויען answered עפרון and Ephron החתי the Hittite את אברהם Abraham באזני in the audience בני of the children חת of Heth, לכל of all באי that went in שׁער at the gate עירו of his city, לאמר׃ saying,}%
\verse{לא Nay, אדני my lord, שׁמעני hear השׂדה me: the field נתתי give לך והמערה I thee, and the cave אשׁר that בו לך נתתיה therein, I give לעיני it thee; in the presence בני of the sons עמי of my people נתתיה give לך קבר I it thee: bury מתך׃ thy dead.}%
\verse{וישׁתחו bowed down himself אברהם And Abraham לפני before עם the people הארץ׃ of the land.}%
\verse{וידבר And he spoke אל unto עפרון Ephron באזני in the audience עם of the people הארץ of the land, לאמר saying, אך But אם if אתה thou לו I pray thee, שׁמעני hear נתתי me: I will give כסף thee money השׂדה for the field; קח take ממני of ואקברה me, and I will bury את מתי my dead שׁמה׃ there.}%
\verse{ויען answered עפרון And Ephron את אברהם Abraham, לאמר׃ saying}%
\verse{אדני My lord, שׁמעני hearken ארץ unto me: the land ארבע four מאת hundred שׁקל shekels כסף of silver; ביני between ובינך מה what הוא that ואת מתך therefore thy dead. קבר׃ me and thee? bury}%
\verse{וישׁמע hearkened אברהם And Abraham אל unto עפרון Ephron; וישׁקל weighed אברהם and Abraham לעפרן to Ephron את הכסף the silver, אשׁר which דבר he had named באזני in the audience בני of the sons חת of Heth, ארבע four מאות hundred שׁקל shekels כסף of silver, עבר current לסחר׃ with the merchant.}%
\verse{ויקם were made sure שׂדה And the field עפרון of Ephron, אשׁר which במכפלה in Machpelah, אשׁר which לפני before ממרא Mamre, השׂדה the field, והמערה and the cave אשׁר which בו וכל therein, and all העץ the trees אשׁר that בשׂדה in the field, אשׁר that בכל in all גבלו the borders סביב׃ round about,}%
\verse{לאברהם Unto Abraham למקנה for a possession לעיני in the presence בני of the children חת of Heth, בכל before all באי that went in שׁער at the gate עירו׃ of his city.}%
\verse{ואחרי And after כן this, קבר buried אברהם Abraham את שׂרה Sarah אשׁתו his wife אל in מערת the cave שׂדה of the field המכפלה of Machpelah על before פני before ממרא Mamre: הוא the same חברון Hebron בארץ in the land כנען׃ of Canaan.}%
\verse{ויקם therein, were made sure השׂדה And the field, והמערה and the cave אשׁר that בו לאברהם unto Abraham לאחזת for a possession קבר of a burial place מאת בני the sons חת׃ of Heth.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 24
\verseWithHeading{Isaac and Rebekah}{ואברהם And Abraham זקן was old, בא well stricken בימים in age: ויהוה and the LORD ברך had blessed את אברהם Abraham בכל׃ in all things.}%
\verse{ויאמר said אברהם And Abraham אל unto עבדו servant זקן his eldest ביתו of his house, המשׁל that ruled בכל over all אשׁר that לו שׂים he had, Put, נא I pray thee, ידך thy hand תחת under ירכי׃ my thigh:}%
\verse{ואשׁביעך And I will make thee swear ביהוה by the LORD, אלהי the God השׁמים of heaven, ואלהי and the God הארץ of the earth, אשׁר that לא thou shalt not תקח take אשׁה a wife לבני unto my son מבנות of the daughters הכנעני of the Canaanites, אשׁר whom אנכי I יושׁב dwell: בקרבו׃ among}%
\verse{כי But אל unto ארצי my country, ואל and to מולדתי my kindred, תלך thou shalt go ולקחת and take אשׁה a wife לבני unto my son ליצחק׃ Isaac.}%
\verse{ויאמר said אליו unto העבד And the servant אולי him, Peradventure לא will not תאבה be willing האשׁה the woman ללכת to follow אחרי to follow אל me unto הארץ land: הזאת this ההשׁב אשׁיבת בנך thy son אל again unto הארץ the land אשׁר from whence יצאת thou camest? משׁם׃}%
\verse{ויאמר said אליו unto אברהם And Abraham השׁמר him, Beware לך פן thou that תשׁיב thou bring את בני not my son שׁמה׃ thither}%
\verse{יהוה The LORD אלהי God השׁמים of heaven, אשׁר which לקחני took מבית אבימארץ and from the land מולדתי of my kindred, ואשׁר and which דבר spoke לי ואשׁר unto me, and that נשׁבע swore לי לאמר unto me, saying, לזרעך Unto thy seed אתן will I give את הארץ land; הזאת this הוא he ישׁלח shall send מלאכו his angel לפניך before ולקחת thee, and thou shalt take אשׁה a wife לבני unto my son משׁם׃ from thence.}%
\verse{ואם And if לא will not תאבה be willing האשׁה the woman ללכת to follow אחריך to follow ונקית thee, then thou shalt be clear משׁבעתי זאתק only את בני my son לא not תשׁב bring שׁמה׃ thither}%
\verse{וישׂם put העבד And the servant את ידו his hand תחת under ירך the thigh אברהם of Abraham אדניו his master, וישׁבע and swore לו על to him concerning הדבר matter. הזה׃ that}%
\verse{ויקח took העבד And the servant עשׂרה ten גמלים camels מגמלי of the camels אדניו of his master, וילך and departed; וכל for all טוב the goods אדניו of his master בידו in his hand: ויקם and he arose, וילך and went אל to ארם נהרים Mesopotamia, אל unto עיר the city נחור׃ of Nahor.}%
\verse{ויברך to kneel down הגמלים And he made his camels מחוץ without לעיר the city אל by באר a well המים of water לעת at the time ערב of the evening, לעת the time צאת that women go out השׁאבת׃ to draw}%
\verse{ויאמר And he said, יהוה O LORD אלהי God אדני of my master אברהם Abraham, הקרה send me good speed נא I pray thee, לפני send me good speed היום this day, ועשׂה and show חסד kindness עם unto אדני my master אברהם׃ Abraham.}%
\verse{הנה Behold, אנכי I נצב stand על by עין the well המים of water; ובנות and the daughters אנשׁי of the men העיר of the city יצאת come out לשׁאב to draw מים׃ water:}%
\verse{והיה And let it come to pass, הנער that the damsel אשׁר to whom אמר and she shall say, אליה הטי Let down נא I pray thee, כדך thy pitcher, ואשׁתה that I may drink; ואמרה שׁתה Drink, וגם also: גמליך and I will give thy camels אשׁקה drink אתה הכחת she thou hast appointed לעבדך for thy servant ליצחק Isaac; ובה אדע and thereby shall I know כי that עשׂית thou hast showed חסד kindness עם unto אדני׃ my master.}%
\verse{ויהי And it came to pass, הוא he טרם before כלה had done לדבר speaking, והנה that, behold, רבקה Rebekah יצאת came out, אשׁר who ילדה was born לבתואל to Bethuel, בן son מלכה of Milcah, אשׁת the wife נחור of Nahor, אחי brother, אברהם Abraham's וכדה with her pitcher על upon שׁכמה׃ her shoulder.}%
\verse{והנער And the damsel טבת fair מראה to look upon, מאד very בתולה a virgin, ואישׁ had any man לא neither ידעה known ותרד her: and she went down העינה to the well, ותמלא and filled כדה her pitcher, ותעל׃ and came up.}%
\verse{וירץ ran העבד And the servant לקראתה to meet ויאמר her, and said, הגמיאיני drink נא Let me, I pray thee, מעט a little מים water מכדך׃ of thy pitcher.}%
\verse{ותאמר And she said, שׁתה Drink, אדני my lord: ותמהר and she hasted, ותרד and let down כדה her pitcher על upon ידה her hand, ותשׁקהו׃ and gave him drink.}%
\verse{ותכל And when she had done להשׁקתו giving him drink, ותאמר she said, גם also, לגמליך for thy camels אשׁאב I will draw עד until אם until כלו they have done לשׁתת׃ drinking.}%
\verse{ותמהר And she hasted, ותער and emptied כדה her pitcher אל into השׁקת the trough, ותרץ and ran עוד again אל unto הבאר the well לשׁאב to draw ותשׁאב and drew לכל for all גמליו׃ his camels.}%
\verse{והאישׁ And the man משׁתאה wondering לה מחרישׁ at her held his peace, לדעת to know ההצליח prosperous יהוה whether the LORD דרכו had made his journey אם or לא׃ not.}%
\verse{ויהי And it came to pass, כאשׁר as כלו had done הגמלים the camels לשׁתות drinking, ויקח took האישׁ that the man נזם earring זהב a golden בקע of half a shekel משׁקלו weight, ושׁני and two צמידים bracelets על for ידיה her hands עשׂרה of ten זהב of gold; משׁקלם׃ weight}%
\verse{ויאמר And said, בת daughter מי Whose את thou? הגידי tell נא me, I pray thee: לי הישׁ is there בית house אביך thy father's מקום room לנו ללין׃ for us to lodge in?}%
\verse{ותאמר And she said אליו unto בת the daughter בתואל of Bethuel אנכי him, I בן the son מלכה of Milcah, אשׁר which ילדה she bore לנחור׃ unto Nahor.}%
\verse{ותאמר She said אליו moreover unto גם him, We have both תבן straw גם and מספוא provender רב enough, עמנו גם and מקום room ללון׃ to lodge in.}%
\verse{ויקד bowed down his head, האישׁ And the man וישׁתחו and worshiped ליהוה׃ the LORD.}%
\verse{ויאמר And he said, ברוך Blessed יהוה the LORD אלהי God אדני of my master אברהם Abraham, אשׁר who לא hath not עזב left destitute חסדו of his mercy ואמתו and his truth: מעם אדני my master אנכי I בדרך in the way, נחני led יהוה the LORD בית me to the house אחי brethren. אדני׃ of my master's}%
\verse{ותרץ ran, הנער And the damsel ותגד and told לבית house אמה her mother's כדברים things. האלה׃ these}%
\verse{ולרבקה And Rebekah אח had a brother, ושׁמו and his name לבן Laban: וירץ ran לבן and Laban אל unto האישׁ the man, החוצה out אל unto העין׃ the well.}%
\verse{ויהי And it came to pass, כראת when he saw את הנזם the earring ואת הצמדים and bracelets על upon ידי hands, אחתו his sister's וכשׁמעו and when he heard את דברי the words רבקה of Rebekah אחתו his sister, לאמר saying, כה Thus דבר spoke אלי unto האישׁ the man ויבא me; that he came אל unto האישׁ the man; והנה and, behold, עמד he stood על by הגמלים the camels על at העין׃ the well.}%
\verse{ויאמר And he said, בוא Come in, ברוך thou blessed יהוה of the LORD; למה wherefore תעמד standest בחוץ thou without? ואנכי for I פניתי have prepared הבית the house, ומקום and room לגמלים׃ for the camels.}%
\verse{ויבא came האישׁ And the man הביתה into the house: ויפתח and he ungirded הגמלים his camels, ויתן and gave תבן straw ומספוא and provender לגמלים for the camels, ומים and water לרחץ to wash רגליו his feet, ורגלי feet האנשׁים אשׁר that אתו׃ with}%
\verse{ויישׂם לפניו before לאכל him to eat: ויאמר but he said, לא I will not אכל eat, עד until אם until דברתי I have told דברי mine errand. ויאמר And he said, דבר׃ Speak on.}%
\verse{ויאמר And he said, עבד servant. אברהם Abraham's אנכי׃ I}%
\verse{ויהוה And the LORD ברך hath blessed את אדני my master מאד greatly; ויגדל and he is become great: ויתן and he hath given לו צאן him flocks, ובקר and herds, וכסף and silver, וזהב and gold, ועבדם and menservants, ושׁפחת and maidservants, וגמלים and camels, וחמרים׃ and asses.}%
\verse{ותלד bore שׂרה And Sarah אשׁת wife אדני my master's בן a son לאדני to my master אחרי when זקנתה she was old: ויתן and unto him hath he given לו את כל all אשׁר׃ that}%
\verse{וישׁבעני made me swear, אדני And my master לאמר saying, לא Thou shalt not תקח take אשׁה a wife לבני to my son מבנות of the daughters הכנעני of the Canaanites, אשׁר in whose אנכי I ישׁב dwell: בארצו׃ land}%
\verse{אם לאל unto בית house, אבי my father's תלך thou shalt go ואל and to משׁפחתי my kindred, ולקחת and take אשׁה a wife לבני׃ unto my son.}%
\verse{ואמר And I said אל unto אדני my master, אלי Peradventure לא will not תלך follow האשׁה the woman אחרי׃ follow}%
\verse{ויאמר And he said אלי unto יהוה me, The LORD, אשׁר whom התהלכתי I walk, לפניו before ישׁלח will send מלאכו his angel אתך with והצליח thee, and prosper דרכך thy way; ולקחת and thou shalt take אשׁה a wife לבני for my son ממשׁפחתי of my kindred, ומבית אבי׃}%
\verse{אז Then תנקה shalt thou be clear מאלתי from my oath, כי when תבוא thou comest אל to משׁפחתי my kindred; ואם and if לא not יתנו they give לך והיית thee thou shalt be נקי clear מאלתי׃ from my oath.}%
\verse{ואבא And I came היום this day אל unto העין the well, ואמר and said, יהוה O LORD אלהי God אדני of my master אברהם Abraham, אם if ישׁך thou do נא now מצליח prosper דרכי my way אשׁר which אנכי I הלך go: עליה׃ which}%
\verse{הנה Behold, אנכי I נצב stand על by עין the well המים of water; והיה and it shall come to pass, העלמה that when the virgin היצאת cometh forth לשׁאב to draw ואמרתי and I say אליה to השׁקיני to drink; נא her, Give me, I pray thee, מעט a little מים water מכדך׃ of thy pitcher}%
\verse{ואמרה And she say אלי to גם me, Both אתה thou, שׁתה drink וגם and I will also לגמליך for thy camels: אשׁאב draw הוא the same האשׁה the woman אשׁר whom הכיח hath appointed out יהוה the LORD לבן son. אדני׃ for my master's}%
\verse{אני I טרם And before אכלה had done לדבר speaking אל in לבי mine heart, והנה behold, רבקה Rebekah יצאת came forth וכדה with her pitcher על on שׁכמה her shoulder; ותרד and she went down העינה unto the well, ותשׁאב and drew ואמר and I said אליה unto השׁקיני her, Let me drink, נא׃ I pray thee.}%
\verse{ותמהר And she made haste, ותורד and let down כדה her pitcher מעליה from ותאמר her and said, שׁתה Drink, וגם also: גמליך and I will give thy camels אשׁקה drink ואשׁת so I drank, וגם also. הגמלים and she made the camels השׁקתה׃}%
\verse{ואשׁאל And I asked אתה ואמר her, and said, בת daughter מי Whose את thou? ותאמר And she said, בת The daughter בתואל of Bethuel, בן son, נחור Nahor's אשׁר whom ילדה bore לו מלכה Milcah ואשׂם unto him: and I put הנזם the earring על upon אפה her face, והצמידים and the bracelets על upon ידיה׃ her hands.}%
\verse{ואקד And I bowed down my head, ואשׁתחוה and worshiped ליהוה the LORD, ואברך and blessed את יהוה the LORD אלהי God אדני of my master אברהם Abraham, אשׁר which הנחני had led בדרך way אמת me in the right לקחת to take את בת daughter אחי brother's אדני my master's לבנו׃ unto his son.}%
\verse{ועתה And now אם if ישׁכם ye will עשׂים deal חסד kindly ואמת and truly את with אדני my master, הגידו tell לי ואם me: and if לא not, הגידו tell לי ואפנה me; that I may turn על to ימין the right hand, או or על to שׂמאל׃ the left.}%
\verse{ויען answered לבן Then Laban ובתואל and Bethuel ויאמרו and said, מיהוה יצא proceedeth הדבר The thing לא we cannot נוכל we cannot דבר speak אליך unto רע thee bad או or טוב׃ good.}%
\verse{הנה Behold, רבקה Rebekah לפניך before קח thee, take ולך and go, ותהי אשׁה wife, לבן son's אדניך thy master's כאשׁר as דבר hath spoken. יהוה׃ the LORD}%
\verse{ויהי And it came to pass, כאשׁר that, when שׁמע heard עבד servant אברהם Abraham's את דבריהם their words, וישׁתחו he worshiped ארצה to the earth. ליהוה׃ the LORD,}%
\verse{ויוצא brought forth העבד And the servant כלי jewels כסף of silver, וכלי and jewels זהב of gold, ובגדים and raiment, ויתן and gave לרבקה to Rebekah: ומגדנת precious things. נתן he gave לאחיה also to her brother ולאמה׃ and to her mother}%
\verse{ויאכלו And they did eat וישׁתו and drink, הוא he והאנשׁים and the men אשׁר that עמו with וילינו him, and tarried all night; ויקומו and they rose up בבקר in the morning, ויאמר and he said, שׁלחני Send me away לאדני׃ unto my master.}%
\verse{ויאמר said, אחיה And her brother ואמה and her mother תשׁב abide הנער Let the damsel אתנו with ימים us days, או at the least עשׂור ten; אחר after that תלך׃ she shall go.}%
\verse{ויאמר And he said אלהם unto אל me not, תאחרו them, Hinder אתי ויהוה seeing the LORD הצליח hath prospered דרכי my way; שׁלחוני send me away ואלכה that I may go לאדני׃ to my master.}%
\verse{ויאמרו And they said, נקרא We will call לנער the damsel, ונשׁאלה and inquire את פיה׃ at her mouth.}%
\verse{ויקראו And they called לרבקה >font color="#000000">Rebekah, ויאמרו and said אליה unto התלכי her, Wilt thou go עם with האישׁ man? הזה this ותאמר And she said, אלך׃ I will go.}%
\verse{וישׁלחו And they sent away את רבקה Rebekah אחתם their sister, ואת מנקתה and her nurse, ואת עבד servant, אברהם and Abraham's ואת אנשׁיו׃ and his men.}%
\verse{ויברכו And they blessed את רבקה Rebekah, ויאמרו and said לה אחתנו our sister, את unto her, Thou היי be לאלפי thou of thousands רבבה of millions, ויירשׁ possess זרעך and let thy seed את שׁער the gate שׂנאיו׃ of those which hate}%
\verse{ותקם arose, רבקה And Rebekah ונערתיה and her damsels, ותרכבנה and they rode על upon הגמלים the camels, ותלכנה and followed אחרי and followed האישׁ the man: ויקח took העבד and the servant את רבקה Rebekah, וילך׃ and went his way.}%
\verse{ויצחק And Isaac בא came מבוא from the way באר לחי ראי of the well Lahai-roi; והוא for he יושׁב dwelt בארץ country. הנגב׃ in the south}%
\verse{ויצא went out יצחק And Isaac לשׂוח to meditate בשׂדה in the field לפנות at the eventide: ערב at the eventide: וישׂא and he lifted up עיניו his eyes, וירא and saw, והנה and, behold, גמלים the camels באים׃ coming.}%
\verse{ותשׂא lifted up רבקה And Rebekah את עיניה her eyes, ותרא and when she saw את יצחק Isaac, ותפל she lighted מעל off הגמל׃ the camel.}%
\verse{ותאמר For she said אל unto העבד the servant, מי What האישׁ man הלזה this ההלך that walketh בשׂדה in the field לקראתנו to meet ויאמר said, העבד us? And the servant הוא It אדני my master: ותקח therefore she took הצעיף a veil, ותתכס׃ and covered herself.}%
\verse{ויספר told העבד And the servant ליצחק Isaac את כל all הדברים things אשׁר that עשׂה׃ he had done.}%
\verse{ויבאה brought יצחק And Isaac האהלה tent, שׂרה Sarah's אמו her into his mother ויקח and took את רבקה Rebekah, ותהי and she became לו לאשׁה his wife; ויאהבה and he loved וינחם was comforted יצחק her: and Isaac אחרי after אמו׃ his mother's}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 25
\verseWithHeading{The Death and Descendants of Abraham}{ויסף Then again אברהם Abraham ויקח took אשׁה a wife, ושׁמה and her name קטורה׃ Keturah.}%
\verse{ותלד And she bore לו את זמרן him Zimran, ואת יקשׁן and Jokshan, ואת מדן and Medan, ואת מדין and Midian, ואת ישׁבק and Ishbak, ואת שׁוח׃ and Shuah.}%
\verse{ויקשׁן And Jokshan ילד begot את שׁבא Sheba, ואת דדן and Dedan. ובני And the sons דדן of Dedan היו were אשׁורם Asshurim, ולטושׁם and Letushim, ולאמים׃ and Leummim.}%
\verse{ובני And the sons מדין of Midian; עיפה Ephah, ועפר and Epher, וחנך and Hanoch, ואבידע and Abida, ואלדעה and Eldaah. כל All אלה these בני the children קטורה׃ of Keturah.}%
\verse{ויתן gave אברהם And Abraham את כל all אשׁר that לו ליצחק׃ he had unto Isaac.}%
\verse{ולבני But unto the sons הפילגשׁים of the concubines, אשׁר which לאברהם Abraham נתן gave אברהם had, Abraham מתנת gifts, וישׁלחם and sent them away מעל from יצחק Isaac בנו his son, בעודנו while he yet חי lived, קדמה eastward, אל unto ארץ country. קדם׃ the east}%
\verse{ואלה And these ימי the days שׁני of the years חיי life אברהם of Abraham's אשׁר which חי he lived, מאת a hundred שׁנה years. ושׁבעים שׁנהחמשׁ שׁנים׃}%
\verse{ויגוע gave up the ghost, וימת and died אברהם Then Abraham בשׂיבה old age, טובה in a good זקן an old man, ושׂבע and full ויאסף and was gathered אל to עמיו׃ his people.}%
\verse{ויקברו buried אתו יצחק Isaac וישׁמעאל and Ishmael בניו And his sons אל him in מערת the cave המכפלה of Machpelah, אל in שׂדה the field עפרן of Ephron בן the son צחר of Zohar החתי the Hittite, אשׁר which על before פני before ממרא׃ Mamre;}%
\verse{השׂדה The field אשׁר which קנה purchased אברהם Abraham מאת בני the sons חת of Heth: שׁמה there קבר buried, אברהם was Abraham ושׂרה and Sarah אשׁתו׃ his wife.}%
\verse{ויהי And it came to pass אחרי after מות the death אברהם of Abraham, ויברך blessed אלהים that God את יצחק Isaac; בנו his son וישׁב dwelt יצחק and Isaac עם by באר לחי ראי׃ the well Lahai-roi.}%
\verse{ואלה Now these תלדת the generations ישׁמעאל of Ishmael, בן son, אברהם Abraham's אשׁר whom ילדה bore הגר Hagar המצרית the Egyptian, שׁפחת handmaid, שׂרה Sarah's לאברהם׃ unto Abraham:}%
\verse{ואלה And these שׁמות the names בני of the sons ישׁמעאל of Ishmael, בשׁמתם by their names, לתולדתם according to their generations: בכר the firstborn ישׁמעאל of Ishmael, נבית Nebajoth; וקדר and Kedar, ואדבאל and Adbeel, ומבשׂם׃ and Mibsam,}%
\verse{ומשׁמע And Mishma, ודומה and Dumah, ומשׂא׃ and Massa,}%
\verse{חדד ותימא and Tema, יטור Jetur, נפישׁ Naphish, וקדמה׃ and Kedemah:}%
\verse{אלה These הם בני the sons ישׁמעאל of Ishmael, ואלה and these שׁמתם their names, בחצריהם by their towns, ובטירתם and by their castles; שׁנים twelve עשׂר twelve נשׂיאם princes לאמתם׃ according to their nations.}%
\verse{ואלה And these שׁני the years חיי of the life ישׁמעאל of Ishmael, מאת a hundred שׁנה years: ושׁלשׁים and thirty שׁנה ושׁבע and seven שׁנים ויגוע and he gave up the ghost וימת and died; ויאסף and was gathered אל unto עמיו׃ his people.}%
\verse{וישׁכנו And they dwelt מחוילה עד unto שׁור Shur, אשׁר that על before פני before מצרים Egypt, באכה as thou goest אשׁורה toward Assyria: על in פני the presence כל of all אחיו his brethren. נפל׃ he died}%
\verseWithHeading{Jacob and Esau}{ואלה And these תולדת the generations יצחק of Isaac, בן son: אברהם Abraham's אברהם Abraham הוליד begot את יצחק׃ Isaac:}%
\verse{ויהי was יצחק And Isaac בן old ארבעים forty שׁנה years בקחתו when he took את רבקה Rebekah בת the daughter בתואל of Bethuel הארמי the Syrian מפדן ארם אחות the sister לבן to Laban הארמי the Syrian. לו לאשׁה׃ to wife,}%
\verse{ויעתר entreated יצחק And Isaac ליהוה the LORD לנכח for אשׁתו his wife, כי because עקרה barren: הוא she ויעתר was entreated לו יהוה and the LORD ותהר conceived. רבקה of him, and Rebekah אשׁתו׃ his wife}%
\verse{ויתרצצו struggled together הבנים And the children בקרבה within ותאמר her; and she said, אם If כן so, למה why זה thus? אנכי I ותלך And she went לדרשׁ to inquire את יהוה׃ of the LORD.}%
\verse{ויאמר said יהוה And the LORD לה שׁני unto her, Two גיים nations בבטנך in thy womb, ושׁני and two לאמים manner of people ממעיך from thy bowels; יפרדו shall be separated ולאם and people מלאם than people; יאמץ shall be stronger ורב and the elder יעבד shall serve צעיר׃ the younger.}%
\verse{וימלאו were fulfilled, ימיה And when her days ללדת to be delivered והנה behold, תומם twins בבטנה׃ in her womb.}%
\verse{ויצא came out הראשׁון And the first אדמוני red, כלו all over כאדרת garment; שׂער like an hairy ויקראו and they called שׁמו his name עשׂו׃ Esau.}%
\verse{ואחרי כןצא came אחיו his brother וידו out, and his hand אחזת took hold בעקב heel; עשׂו on Esau's ויקרא was called שׁמו and his name יעקב Jacob: ויצחק and Isaac בן old שׁשׁים threescore שׁנה years בלדת when she bore אתם׃}%
\verse{ויגדלו grew: הנערים And the boys ויהי was עשׂו and Esau אישׁ a man ידע a cunning ציד hunter, אישׁ man, שׂדה of the field; ויעקב and Jacob אישׁ תם a plain ישׁב dwelling אהלים׃ in tents.}%
\verse{ויאהב loved יצחק And Isaac את עשׂו Esau, כי because ציד of venison: בפיו he did eat ורבקה but Rebekah אהבת loved את יעקב׃ Jacob.}%
\verse{ויזד sod יעקב And Jacob נזיד pottage: ויבא came עשׂו and Esau מן from השׂדה the field, והוא and he עיף׃ faint:}%
\verse{ויאמר said עשׂו And Esau אל to יעקב Jacob, הלעיטני Feed נא me, I pray thee, מן with האדם red האדם הזה that same כי for עיף faint: אנכי I על therefore כן therefore קרא called שׁמו was his name אדום׃ Edom.}%
\verse{ויאמר said, יעקב And Jacob מכרה Sell כיום me this day את בכרתך׃ thy birthright.}%
\verse{ויאמר said, עשׂו And Esau הנה Behold, אנכי I הולך at the point למות to die: ולמה and what profit זה shall this לי בכרה׃ birthright}%
\verse{ויאמר said, יעקב And Jacob השׁבעה Swear לי כיום to me this day; וישׁבע and he swore לו וימכר unto him: and he sold את בכרתו his birthright ליעקב׃ unto Jacob.}%
\verse{ויעקב Then Jacob נתן gave לעשׂו Esau לחם bread ונזיד and pottage עדשׁים of lentils; ויאכל and he did eat וישׁת and drink, ויקם and rose up, וילך and went his way: ויבז despised עשׂו thus Esau את הבכרה׃ birthright.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 26
\verseWithHeading{Isaac and Abimelech}{ויהי And there was רעב a famine בארץ in the land, מלבד beside הרעב famine הראשׁון the first אשׁר that היה was בימי in the days אברהם of Abraham. וילך went יצחק And Isaac אל unto אבימלך Abimelech מלך king פלשׁתים of the Philistines גררה׃ unto Gerar.}%
\verse{וירא appeared אליו unto יהוה And the LORD ויאמר him, and said, אל תרדצרימה into Egypt; שׁכן dwell בארץ in the land אשׁר which אמר I shall tell אליך׃ thee of:}%
\verse{גור Sojourn בארץ land, הזאת in this ואהיה and I will be עמך with ואברכך thee, and will bless כי thee; for לך ולזרעך unto thee, and unto thy seed, אתן I will give את כל all הארצת countries, האל these והקמתי and I will perform את השׁבעה the oath אשׁר which נשׁבעתי I swore לאברהם unto Abraham אביך׃ thy father;}%
\verse{והרביתי to multiply את זרעך And I will make thy seed ככוכבי as the stars השׁמים of heaven, ונתתי and will give לזרעך unto thy seed את כל all הארצת countries; האל these והתברכו be blessed; בזרעך and in thy seed כל shall all גויי the nations הארץ׃ of the earth}%
\verse{עקב Because אשׁר that שׁמע obeyed אברהם Abraham בקלי my voice, וישׁמר and kept משׁמרתי my charge, מצותי my commandments, חקותי my statutes, ותורתי׃ and my laws.}%
\verse{וישׁב dwelt יצחק And Isaac בגרר׃ in Gerar:}%
\verse{וישׁאלו asked אנשׁי And the men המקום of the place לאשׁתו of his wife; ויאמר and he said, אחתי my sister: הוא She כי for ירא he feared לאמר to say, אשׁתי my wife; פן lest, יהרגני should kill אנשׁי the men המקום of the place על me for רבקה Rebekah; כי because טובת fair מראה to look upon. הוא׃ she}%
\verse{ויהי And it came to pass, כי when ארכו a long לו שׁם he had been there הימים time, וישׁקף looked out אבימלך that Abimelech מלך king פלשׁתים of the Philistines בעד at החלון a window, וירא and saw, והנה and, behold, יצחק Isaac מצחק sporting את רבקה with Rebekah אשׁתו׃ his wife.}%
\verse{ויקרא called אבימלך And Abimelech ליצחק Isaac, ויאמר and said, אך of a surety הנה Behold, אשׁתך thy wife: הוא she ואיך and how אמרת saidst אחתי my sister? הוא thou, She ויאמר said אליו unto יצחק And Isaac כי him, Because אמרתי I said, פן Lest אמות I die עליה׃ for}%
\verse{ויאמר said, אבימלך And Abimelech מה What זאת this עשׂית thou hast done לנו כמעט might lightly שׁכב have lain אחד unto us? one העם of the people את with אשׁתך thy wife, והבאת and thou shouldest have brought עלינו upon אשׁם׃ guiltiness}%
\verse{ויצו charged אבימלך And Abimelech את כל all העם people, לאמר saying, הנגע He that toucheth באישׁ man הזה this ובאשׁתו or his wife מות shall surely be put to death. יומת׃ shall surely be put to death.}%
\verse{ויזרע sowed יצחק Then Isaac בארץ land, ההוא in that וימצא and received בשׁנה year ההוא in the same מאה a hundredfold: שׁערים a hundredfold: ויברכהו blessed יהוה׃ and the LORD}%
\verse{ויגדל waxed great, האישׁ And the man וילך and went הלוך forward, וגדל he became very great: עד until כי until גדל מאד׃ he became very great:}%
\verse{ויהי For he had לו מקנה possession צאן of flocks, ומקנה and possession בקר of herds, ועבדה of servants: רבה and great store ויקנאו envied אתו פלשׁתים׃ and the Philistines}%
\verse{וכל For all הבארת the wells אשׁר which חפרו had digged עבדי servants אביו his father's בימי in the days אברהם of Abraham אביו his father, סתמום had stopped פלשׁתים the Philistines וימלאום them, and filled עפר׃ them with earth.}%
\verse{ויאמר said אבימלך And Abimelech אל unto יצחק Isaac, לך Go מעמנו from כי us; for עצמת thou art much mightier ממנו from מאד׃ thou art much mightier}%
\verse{וילך departed משׁם thence, יצחק And Isaac ויחן and pitched his tent בנחל in the valley גרר of Gerar, וישׁב and dwelt שׁם׃ there.}%
\verse{וישׁב again יצחק And Isaac ויחפר digged את בארת the wells המים of water, אשׁר which חפרו they had digged בימי in the days אברהם of Abraham אביו his father; ויסתמום had stopped פלשׁתים for the Philistines אחרי them after מות the death אברהם of Abraham: ויקרא and he called להן שׁמות their names כשׁמת after the names אשׁר by which קרא had called להן אביו׃ his father}%
\verse{ויחפרו digged עבדי servants יצחק And Isaac's בנחל in the valley, וימצאו and found שׁם there באר a well מים water. חיים׃ of springing}%
\verse{ויריבו did strive רעי גרר of Gerar עם with רעי יצחק Isaac's לאמר saying, לנו המים The water ויקרא ours: and he called שׁם the name הבאר of the well עשׂק Esek; כי because התעשׂקו they strove עמו׃ with}%
\verse{ויחפרו And they digged באר well, אחרת another ויריבו and strove גם that also: עליה for ויקרא and he called שׁמה the name שׂטנה׃ of it Sitnah.}%
\verse{ויעתק And he removed משׁם from thence, ויחפר and digged באר well; אחרת another ולא not: רבו that they strove עליה and for ויקרא and he called שׁמה the name רחבות of it Rehoboth; ויאמר and he said, כי For עתה now הרחיב hath made room יהוה the LORD לנו ופרינו for us, and we shall be fruitful בארץ׃ in the land.}%
\verse{ויעל And he went up משׁם from thence באר שׁבע׃ to Beer-sheba.}%
\verse{וירא appeared אליו unto יהוה And the LORD בלילה night, ההוא him the same ויאמר and said, אנכי I אלהי the God אברהם of Abraham אביך thy father: אל not, תירא fear כי for אתך with אנכי I וברכתיך thee, and will bless והרביתי thee, and multiply את זרעך thy seed בעבור for אברהם Abraham's עבדי׃ my servant}%
\verse{ויבן And he built שׁם there, מזבח an altar ויקרא and called בשׁם upon the name יהוה of the LORD, ויט and pitched שׁם there: אהלו his tent ויכרו digged שׁם and there עבדי servants יצחק Isaac's באר׃ a well.}%
\verse{ואבימלך Then Abimelech הלך went אליו to מגרר ואחזת and Ahuzzath מרעהו one of his friends, ופיכל and Phichol שׂר the chief captain צבאו׃ of his army.}%
\verse{ויאמר said אלהם unto יצחק And Isaac מדוע them, Wherefore באתם come אלי ye to ואתם me, seeing ye שׂנאתם hate אתי ותשׁלחוני me, and have sent me away מאתכם׃}%
\verse{ויאמרו And they said, ראו ראינוי that היה was יהוה the LORD עמך with ונאמר thee: and we said, תהי Let there be נא now אלה an oath בינותינו between בינינו us, between ובינך ונכרתה us and thee, and let us make ברית a covenant עמך׃ with}%
\verse{אם That תעשׂה thou wilt do עמנו unto רעה us no hurt, כאשׁר as לא we have not נגענוך touched וכאשׁר thee, and as עשׂינו we have done עמך רק thee nothing but טוב good, ונשׁלחך and have sent thee away בשׁלום in peace: אתה thou עתה now ברוך the blessed יהוה׃ of the LORD.}%
\verse{ויעשׂ And he made להם משׁתה them a feast, ויאכלו and they did eat וישׁתו׃ and drink.}%
\verse{וישׁכימו And they rose up quickly בבקר in the morning, וישׁבעו and swore אישׁ one לאחיו to another: וישׁלחם sent them away, יצחק and Isaac וילכו and they departed מאתו בשׁלום׃ him in peace.}%
\verse{ויהי And it came to pass ביום day, ההוא the same ויבאו came, עבדי servants יצחק that Isaac's ויגדו and told לו על him concerning אדות him concerning הבאר the well אשׁר which חפרו they had digged, ויאמרו and said לו מצאנו unto him, We have found מים׃ water.}%
\verse{ויקרא And he called אתה שׁבעה it Shebah: על therefore כן therefore שׁם the name העיר of the city באר שׁבע Beer-sheba עד unto היום day. הזה׃ this}%
\verse{ויהי was עשׂו And Esau בן old ארבעים forty שׁנה years ויקח when he took אשׁה to wife את יהודית Judith בת the daughter בארי of Beeri החתי the Hittite, ואת בשׂמת and Bashemath בת the daughter אילן of Elon החתי׃ the Hittite:}%
\verse{ותהיין Which were מרת a grief רוח of mind ליצחק unto Isaac ולרבקה׃ and to Rebekah.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 27
\verseWithHeading{Jacob Steals Esau’s Blessing}{ויהי And it came to pass, כי that זקן was old, יצחק when Isaac ותכהין were dim, עיניו and his eyes מראת so that he could not see, ויקרא he called את עשׂו Esau בנו son, הגדל his eldest ויאמר and said אליו unto בני him, My son: ויאמר and he said אליו unto הנני׃ him, Behold,}%
\verse{ויאמר And he said, הנה Behold נא now, זקנתי I am old, לא not ידעתי I know יום the day מותי׃ of my death:}%
\verse{ועתה Now שׂא therefore take, נא I pray thee, כליך thy weapons, תליך thy quiver וקשׁתך and thy bow, וצא and go out השׂדה to the field, וצודה and take לי צידה׃}%
\verse{ועשׂה And make לי מטעמים me savory meat, כאשׁר such as אהבתי I love, והביאה and bring לי ואכלה to me, that I may eat; בעבור that תברכך may bless נפשׁי my soul בטרם thee before אמות׃ I die.}%
\verse{ורבקה And Rebekah שׁמעת heard בדבר spoke יצחק when Isaac אל to עשׂו Esau בנו his son. וילך went עשׂו And Esau השׂדה to the field לצוד to hunt ציד venison, להביא׃ to bring}%
\verse{ורבקה And Rebekah אמרה spoke אל unto יעקב Jacob בנה her son, לאמר saying, הנה Behold, שׁמעתי I heard את אביך thy father מדבר speak אל unto עשׂו Esau אחיך thy brother, לאמר׃ saying,}%
\verse{הביאה Bring לי ציד me venison, ועשׂה and make לי מטעמים me savory meat, ואכלה that I may eat, ואברככה and bless לפני thee before יהוה the LORD לפני before מותי׃ my death.}%
\verse{ועתה Now בני therefore, my son, שׁמע obey בקלי my voice לאשׁר according to that which אני I מצוה command אתך׃}%
\verse{לך Go נא now אל to הצאן the flock, וקח and fetch לי משׁם me from thence שׁני two גדיי kids עזים of the goats; טבים good ואעשׂה and I will make אתם מטעמים them savory meat לאביך for thy father, כאשׁר such as אהב׃ he loveth:}%
\verse{והבאת And thou shalt bring לאביך to thy father, ואכל that he may eat, בעבר and that אשׁר and that יברכך he may bless לפני thee before מותו׃ his death.}%
\verse{ויאמר said יעקב And Jacob אל to רבקה Rebekah אמו his mother, הן Behold, עשׂו Esau אחי my brother אישׁ man, שׂער a hairy ואנכי and I אישׁ man: חלק׃ a smooth}%
\verse{אולי peradventure ימשׁני will feel אבי My father והייתי me, and I shall seem בעיניו to him כמתעתע as a deceiver; והבאתי and I shall bring עלי upon קללה a curse ולא me, and not ברכה׃ a blessing.}%
\verse{ותאמר said לו אמו And his mother עלי unto him, Upon קללתך me thy curse, בני my son: אך only שׁמע obey בקלי my voice, ולך and go קח׃ fetch}%
\verse{וילך And he went, ויקח and fetched, ויבא and brought לאמו to his mother: ותעשׂ made אמו and his mother מטעמים savory meat, כאשׁר such as אהב loved. אביו׃ his father}%
\verse{ותקח took רבקה And Rebekah את בגדי raiment עשׂו Esau, בנה son הגדל of her eldest החמדת אשׁר which אתה with בבית her in the house, ותלבשׁ and put them upon את יעקב Jacob בנה son: הקטן׃ her younger}%
\verse{ואת ערת the skins גדיי of the kids העזים of the goats הלבישׁה And she put על upon ידיו his hands, ועל and upon חלקת the smooth צואריו׃ of his neck:}%
\verse{ותתן And she gave את המטעמים the savory meat ואת הלחם and the bread, אשׁר which עשׂתה she had prepared, ביד into the hand יעקב Jacob. בנה׃ of her son}%
\verse{ויבא And he came אל unto אביו his father, ויאמר and said, אבי My father: ויאמר and he said, הנני Here מי I; who אתה thou, בני׃ my son?}%
\verse{ויאמר said יעקב And Jacob אל unto אביו his father, אנכי I עשׂו Esau בכרך thy firstborn; עשׂיתי I have done כאשׁר according as דברת thou biddest אלי thou biddest קום me: arise, נא I pray thee, שׁבה sit ואכלה and eat מצידי of my venison, בעבור that תברכני may bless נפשׁך׃ thy soul}%
\verse{ויאמר said יצחק And Isaac אל unto בנו his son, מה How זה it מהרת so quickly, למצא that thou hast found בני my son? ויאמר And he said, כי Because הקרה brought יהוה the LORD אלהיך thy God לפני׃ to}%
\verse{ויאמר said יצחק And Isaac אל unto יעקב Jacob, גשׁה Come near, נא I pray thee, ואמשׁך that I may feel בני thee, my son, האתה whether thou זה my very בני son עשׂו Esau אם or לא׃ not.}%
\verse{ויגשׁ went near יעקב And Jacob אל unto יצחק Isaac אביו his father; וימשׁהו and he felt ויאמר him, and said, הקל The voice קול voice, יעקב Jacob's והידים but the hands ידי the hands עשׂו׃ of Esau.}%
\verse{ולא him not, הכירו And he discerned כי because היו were ידיו his hands כידי hands: עשׂו Esau's אחיו as his brother שׂערת hairy, ויברכהו׃ so he blessed}%
\verse{ויאמר And he said, אתה thou זה my very בני son עשׂו Esau? ויאמר And he said, אני׃ I}%
\verse{ויאמר And he said, הגשׁה Bring near לי ואכלה to me, and I will eat מציד venison, בני of my son's למען that תברכך may bless נפשׁי my soul ויגשׁ thee. And he brought near לו ויאכל to him, and he did eat: ויבא and he brought לו יין him wine, וישׁת׃ and he drank.}%
\verse{ויאמר said אליו unto יצחק Isaac אביו And his father גשׁה him, Come near נא now, ושׁקה and kiss לי בני׃ me, my son.}%
\verse{ויגשׁ And he came near, וישׁק and kissed לו וירח him: and he smelled את ריח the smell בגדיו of his raiment, ויברכהו and blessed ויאמר him, and said, ראה See, ריח the smell בני of my son כריח as the smell שׂדה of a field אשׁר which ברכו hath blessed: יהוה׃ the LORD}%
\verse{ויתן give לך האלהים Therefore God מטל thee of the dew השׁמים of heaven, ומשׁמני הארץ of the earth, ורב and plenty דגן of corn ותירשׁ׃ and wine:}%
\verse{יעבדוך serve עמים Let people וישׁתחו bow down לך לאמים thee, and nations הוה to thee: be גביר lord לאחיך over thy brethren, וישׁתחוו bow down לך בני sons אמך and let thy mother's ארריך to thee: cursed ארור every one that curseth ומברכיך thee, and blessed ברוך׃ he that blesseth}%
\verse{ויהי And it came to pass, כאשׁר as soon as כלה had made an end יצחק Isaac לברך of blessing את יעקב Jacob, ויהי was אך yet יצא scarce gone out יצא scarce gone out יעקב and Jacob מאת פני the presence יצחק of Isaac אביו his father, ועשׂו that Esau אחיו his brother בא came in מצידו׃ from his hunting.}%
\verse{ויעשׂ had made גם also הוא And he מטעמים savory meat, ויבא and brought לאביו it unto his father, ויאמר and said לאביו unto his father, יקם arise, אבי Let my father ויאכל and eat מציד venison, בנו of his son's בעבר that תברכני may bless נפשׁך׃ thy soul}%
\verse{ויאמר said לו יצחק And Isaac אביו his father מי unto him, Who אתה thou? ויאמר And he said, אני I בנך thy son, בכרך thy firstborn עשׂו׃ Esau.}%
\verse{ויחרד trembled יצחק And Isaac חרדה very exceedingly, גדלה עדאד ויאמר and said, מי Who? אפוא where הוא he הצד that hath taken ציד venison, ויבא and brought לי ואכל me, and I have eaten מכל of all בטרם before תבוא thou camest, ואברכהו and have blessed גם him? yea, ברוך blessed. יהיה׃ he shall be}%
\verse{כשׁמע heard עשׂו And when Esau את דברי the words אביו of his father, ויצעק he cried צעקה cry, גדלה with a great ומרה bitter עד and מאד exceeding ויאמר and said לאביו unto his father, ברכני Bless גם also, אני me, me אבי׃ O my father.}%
\verse{ויאמר And he said, בא came אחיך Thy brother במרמה with subtlety, ויקח and hath taken away ברכתך׃ thy blessing.}%
\verse{ויאמר And he said, הכי Is not he rightly קרא named שׁמו named יעקב Jacob? ויעקבני for he hath supplanted זה me these פעמים two times: את בכרתי my birthright; לקח he took away והנה and, behold, עתה now לקח he hath taken away ברכתי my blessing. ויאמר And he said, הלא Hast thou not אצלת reserved לי ברכה׃ a blessing}%
\verse{ויען answered יצחק And Isaac ויאמר and said לעשׂו unto Esau, הן Behold, גביר him thy lord, שׂמתיו I have made לך ואת כל and all אחיו his brethren נתתי have I given לו לעבדים to him for servants; ודגן and with corn ותירשׁ and wine סמכתיו have I sustained ולכה אפוא now מה him: and what אעשׂה shall I do בני׃ unto thee, my son?}%
\verse{ויאמר said עשׂו And Esau אל unto אביו his father, הברכה blessing, אחת Hast thou but one הוא לך אבי my father? ברכני bless גם also, אני me, me אבי O my father. וישׂא lifted up עשׂו And Esau קלו his voice, ויבך׃ and wept.}%
\verse{ויען answered יצחק And Isaac אביו his father ויאמר and said אליו unto הנה him, Behold, משׁמני הארץ of the earth, יהיה shall be מושׁבך thy dwelling ומטל and of the dew השׁמים of heaven מעל׃}%
\verse{ועל And by חרבך thy sword תחיה shalt thou live, ואת אחיך thy brother; תעבד and shalt serve והיה and it shall come to pass כאשׁר when תריד thou shalt have the dominion, ופרקת that thou shalt break עלו his yoke מעל from off צוארך׃ thy neck.}%
\verse{וישׂטם hated עשׂו And Esau את יעקב Jacob על because of הברכה the blessing אשׁר wherewith ברכו blessed אביו his father ויאמר said עשׂו him: and Esau בלבו in his heart, יקרבו are at hand; ימי The days אבל of mourning אבי for my father ואהרגה then will I slay את יעקב Jacob. אחי׃ my brother}%
\verse{ויגד were told לרבקה to Rebekah: את דברי And these words עשׂו of Esau בנה son הגדל her elder ותשׁלח and she sent ותקרא and called ליעקב Jacob בנה son, הקטן her younger ותאמר and said אליו unto הנה him, Behold, עשׂו Esau, אחיך thy brother מתנחם as touching thee, doth comfort himself, לך להרגך׃ to kill}%
\verse{ועתה Now בני therefore, my son, שׁמע obey בקלי my voice; וקום and arise, ברח flee לך אל thou to לבן Laban אחי my brother חרנה׃ to Haran;}%
\verse{וישׁבת And tarry עמו with ימים days, אחדים him a few עד until אשׁר until תשׁוב turn away; חמת fury אחיך׃ thy brother's}%
\verse{עד Until שׁוב turn away אף anger אחיך thy brother's ממך from ושׁכח thee, and he forget את אשׁר which עשׂית thou hast done לו ושׁלחתי to him: then I will send, ולקחתיך and fetch משׁם thee from thence: למה why אשׁכל should I be deprived גם also שׁניכם of you both יום day? אחד׃ in one}%
\verse{ותאמר said רבקה And Rebekah אל to יצחק Isaac, קצתי I am weary בחיי of my life מפני because בנות of the daughters חת of Heth: אם if לקח take יעקב Jacob אשׁה a wife מבנות of the daughters חת of Heth, כאלה such as these מבנות of the daughters הארץ of the land, למה what good לי חיים׃ shall my life}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 28
\verseWithHeading{Jacob Flees to Haran}{ויקרא called יצחק And Isaac אל called יעקב Jacob, ויברך and blessed אתו ויצוהו him, and charged ויאמר him, and said לו לא unto him, Thou shalt not תקח take אשׁה a wife מבנות of the daughters כנען׃ of Canaan.}%
\verse{קום Arise, לך go פדנה ארם to Padan-aram, ביתה to the house בתואל of Bethuel אבי father; אמך thy mother's וקח and take לך משׁם from thence אשׁה thee a wife מבנות of the daughters לבן of Laban אחי brother. אמך׃ thy mother's}%
\verse{ואל And God שׁדי Almighty יברך bless אתך ויפרך thee, and make thee fruitful, וירבך and multiply והיית thee, that thou mayest be לקהל a multitude עמים׃ of people;}%
\verse{ויתן And give לך את ברכת thee the blessing אברהם of Abraham, לך ולזרעך to thee, and to thy seed אתך with לרשׁתך thee; that thou mayest inherit את ארץ the land מגריך wherein thou art a stranger, אשׁר which נתן gave אלהים God לאברהם׃ unto Abraham.}%
\verse{וישׁלח sent away יצחק And Isaac את יעקב Jacob: וילך and he went פדנה ארם to Padan-aram אל unto לבן Laban, בן son בתואל of Bethuel הארמי the Syrian, אחי the brother רבקה of Rebekah, אם mother. יעקב Jacob's ועשׂו׃ and Esau's}%
\verse{וירא saw עשׂו When Esau כי that ברך had blessed יצחק Isaac את יעקב Jacob, ושׁלח אתודנה ארם to Padan-aram, לקחת to take לו משׁם from thence; אשׁה him a wife בברכו and that as he blessed אתו ויצו him he gave him a charge, עליו him he gave him a charge, לאמר saying, לא Thou shalt not תקח take אשׁה a wife מבנות of the daughters כנען׃ of Canaan;}%
\verse{וישׁמע obeyed יעקב And that Jacob אל obeyed אביו his father ואל אמו and his mother, וילך and was gone פדנה ארם׃ to Padan-aram;}%
\verse{וירא seeing עשׂו And Esau כי that רעות pleased not בנות the daughters כנען of Canaan בעיני pleased not יצחק Isaac אביו׃ his father;}%
\verse{וילך Then went עשׂו Esau אל unto ישׁמעאל Ishmael, ויקח and took את מחלת which he had Mahalath בת the daughter ישׁמעאל of Ishmael בן son, אברהם Abraham's אחות the sister נביות of Nebajoth, על unto נשׁיו the wives לו לאשׁה׃ to be his wife.}%
\verseWithHeading{Jacob’s Dream}{ויצא went out יעקב And Jacob מבאר שׁבע וילך and went חרנה׃ toward Haran.}%
\verse{ויפגע And he lighted במקום upon a certain place, וילן and tarried there all night, שׁם and tarried there all night, כי because בא was set; השׁמשׁ the sun ויקח and he took מאבני of the stones המקום of that place, וישׂם and put מראשׁתיו his pillows, וישׁכב and lay down במקום place ההוא׃ in that}%
\verse{ויחלם And he dreamed, והנה and behold סלם a ladder מצב set up ארצה on the earth, וראשׁו and the top מגיע of it reached השׁמימה to heaven: והנה and behold מלאכי the angels אלהים of God עלים ascending וירדים׃ and descending}%
\verse{והנה And, behold, יהוה the LORD נצב stood עליו above ויאמר it, and said, אני I יהוה the LORD אלהי God אברהם of Abraham אביך thy father, ואלהי and the God יצחק of Isaac: הארץ the land אשׁר whereon אתה thou שׁכב liest, עליה whereon לך אתננה to thee will I give ולזרעך׃ it, and to thy seed;}%
\verse{והיה shall be זרעך And thy seed כעפר as the dust הארץ of the earth, ופרצת and thou shalt spread abroad ימה to the west, וקדמה and to the east, וצפנה and to the north, ונגבה and to the south: ונברכו be blessed. בך כל shall all משׁפחת the families האדמה of the earth ובזרעך׃ and in thee and in thy seed}%
\verse{והנה And, behold, אנכי I עמך with ושׁמרתיך thee, and will keep בכל thee in all אשׁר whither תלך thou goest, והשׁבתיך and will bring thee again אל into האדמה land; הזאת this כי for לא I will not אעזבך leave עד thee, until אשׁר which אם עשׂיתי I have done את אשׁרברתי׃ I have spoken}%
\verse{וייקץ awaked יעקב And Jacob משׁנתו out of his sleep, ויאמר and he said, אכן Surely ישׁ is יהוה the LORD במקום place; הזה in this ואנכי and I לא not. ידעתי׃ knew}%
\verse{ויירא And he was afraid, ויאמר and said, מה How נורא dreadful המקום place! הזה this אין none זה this כי other but אם other but בית the house אלהים of God, וזה and this שׁער the gate השׁמים׃ of heaven.}%
\verse{וישׁכם rose up early יעקב And Jacob בבקר in the morning, ויקח and took את האבן the stone אשׁר that שׂם he had put מראשׁתיו his pillows, וישׂם אתהצבה a pillar, ויצק and poured שׁמן oil על upon ראשׁה׃ the top}%
\verse{ויקרא And he called את שׁם the name המקום place ההוא of that בית אל Bethel: ואולם but לוז Luz שׁם the name העיר of that city לראשׁנה׃ at the first.}%
\verse{וידר vowed יעקב And Jacob נדר a vow, לאמר saying, אם If יהיה will be אלהים God עמדי ושׁמרני me, and will keep בדרך way הזה me in this אשׁר that אנכי I הולך go, ונתן and will give לי לחם me bread לאכל to eat, ובגד and raiment ללבשׁ׃ to put on,}%
\verse{ושׁבתי So that I come again בשׁלום in peace; אל to בית house אבי my father's והיה be יהוה then shall the LORD לי לאלהים׃ my God:}%
\verse{והאבן stone, הזאת And this אשׁר which שׂמתי I have set מצבה a pillar, יהיה shall be בית house: אלהים God's וכל and of all אשׁר that תתן thou shalt give לי עשׂר אעשׂרנו׃}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 29
\verseWithHeading{Jacob Flees to Haran}{וישׂא went on his journey, יעקב Then Jacob רגליו went on his journey, וילך and came ארצה into the land בני of the people קדם׃ of the east.}%
\verse{וירא And he looked, והנה and behold באר a well בשׂדה in the field, והנה and, lo, שׁם there שׁלשׁה three עדרי flocks צאן of sheep רבצים lying עליה by כי it; for מן out of הבאר well ההוא that ישׁקו they watered העדרים the flocks: והאבן stone גדלה and a great על upon פי mouth. הבאר׃ the well's}%
\verse{ונאספו gathered: שׁמה And thither כל were all העדרים the flocks וגללו and they rolled את האבן the stone מעל from פי mouth, הבאר the well's והשׁקו and watered את הצאן the sheep, והשׁיבו and put את האבן the stone על again upon פי mouth הבאר the well's למקמה׃ in his place.}%
\verse{ויאמר said להם יעקב And Jacob אחי unto them, My brethren, מאין whence אתם ye? ויאמרו And they said, מחרן אנחנו׃}%
\verse{ויאמר And he said להם הידעתם unto them, Know את לבן ye Laban בן the son נחור of Nahor? ויאמרו And they said, ידענו׃ We know}%
\verse{ויאמר And he said להם השׁלום unto them, he well? לו ויאמרו And they said, שׁלום well: והנה and, behold, רחל Rachel בתו his daughter באה cometh עם with הצאן׃ the sheep.}%
\verse{ויאמר And he said, הן Lo, עוד yet היום day, גדול high לא neither עת time האסף should be gathered together: המקנה that the cattle השׁקו water הצאן ye the sheep, ולכו and go רעו׃ feed}%
\verse{ויאמרו And they said, לא נוכלד until אשׁר until יאספו be gathered together, כל all העדרים the flocks וגללו and they roll את האבן the stone מעל from פי mouth; הבאר the well's והשׁקינו then we water הצאן׃ the sheep.}%
\verse{עודנו And while he yet מדבר spoke עמם with ורחל them, Rachel באה came עם with הצאן sheep: אשׁר לאביה her father's כי for רעה kept הוא׃ she}%
\verse{ויהי And it came to pass, כאשׁר when ראה saw יעקב Jacob את רחל Rachel בת the daughter לבן of Laban אחי brother, אמו his mother's ואת צאן and the sheep לבן of Laban אחי brother, אמו his mother's ויגשׁ went near, יעקב that Jacob ויגל and rolled את האבן the stone מעל from פי mouth, הבאר the well's וישׁק and watered את צאן the flock לבן of Laban אחי brother. אמו׃ his mother's}%
\verse{וישׁק kissed יעקב And Jacob לרחל Rachel, וישׂא and lifted up את קלו his voice, ויבך׃ and wept.}%
\verse{ויגד told יעקב And Jacob לרחל Rachel כי that אחי brother, אביה her father's הוא he וכי and that בן son: רבקה Rebekah's הוא he ותרץ and she ran ותגד and told לאביה׃ her father.}%
\verse{ויהי And it came to pass, כשׁמע heard לבן when Laban את שׁמע the tidings יעקב of Jacob בן son, אחתו his sister's וירץ that he ran לקראתו to meet ויחבק him, and embraced לו וינשׁק him, and kissed לו ויביאהו him, and brought אל him to ביתו his house. ויספר And he told ללבן Laban את כל all הדברים things. האלה׃ these}%
\verse{ויאמר said לו לבן And Laban אך to him, Surely עצמי my bone ובשׂרי and my flesh. אתה thou וישׁב And he abode עמו with חדשׁ of a month. ימים׃ him the space}%
\verseWithHeading{Jacob’s Marriages}{ויאמר said לבן And Laban ליעקב unto Jacob, הכי Because אחי my brother, אתה thou ועבדתני shouldest thou therefore serve חנם me for naught? הגידה tell לי מה me, what משׂכרתך׃ thy wages}%
\verse{וללבן And Laban שׁתי had two בנות daughters: שׁם the name הגדלה of the elder לאה Leah, ושׁם and the name הקטנה of the younger רחל׃ Rachel.}%
\verse{ועיני eyed; לאה Leah רכות tender ורחל but Rachel היתה was יפת beautiful תאר beautiful ויפת and well מראה׃ favored.}%
\verse{ויאהב loved יעקב And Jacob את רחל Rachel; ויאמר and said, אעבדך I will serve שׁבע thee seven שׁנים years ברחל for Rachel בתך daughter. הקטנה׃ thy younger}%
\verse{ויאמר said, לבן And Laban טוב better תתי that I give אתה לך מתתיתה לאישׁחר her to another שׁבה abide עמדי׃}%
\verse{ויעבד served יעקב And Jacob ברחל for Rachel; שׁבע seven שׁנים years ויהיו and they seemed בעיניו and they seemed כימים days, אחדים unto him a few באהבתו אתה׃}%
\verse{ויאמר said יעקב And Jacob אל unto לבן Laban, הבה Give את אשׁתי my wife, כי for מלאו are fulfilled, ימי my days ואבואה that I may go in אליה׃ unto}%
\verse{ויאסף gathered together לבן And Laban את כל all אנשׁי המקום of the place, ויעשׂ and made משׁתה׃ a feast.}%
\verse{ויהי And it came to pass בערב in the evening, ויקח that he took את לאה Leah בתו his daughter, ויבא and brought אתה אליו her to ויבא him; and he went in אליה׃ unto}%
\verse{ויתן gave לבן And Laban לה את זלפה Zilpah שׁפחתו his maid ללאה Leah בתו unto his daughter שׁפחה׃ a handmaid.}%
\verse{ויהי And it came to pass, בבקר that in the morning, והנה behold, הוא it לאה Leah: ויאמר and he said אל to לבן Laban, מה What זאת this עשׂית thou hast done לי הלא unto me? did not ברחל thee for Rachel? עבדתי I serve עמך with ולמה wherefore רמיתני׃ then hast thou beguiled}%
\verse{ויאמר said, לבן And Laban לא It must not יעשׂה done כן be so במקומנו in our country, לתת to give הצעירה the younger לפני before הבכירה׃ the firstborn.}%
\verse{מלא Fulfill שׁבע week, זאת her ונתנה and we will give לך גם also את זאת thee this בעבדה for the service אשׁר which תעבד thou shalt serve עמדי עוד me yet שׁבע seven שׁנים years. אחרות׃ other}%
\verse{ויעשׂ did יעקב And Jacob כן so, וימלא and fulfilled שׁבע week: זאת her ויתן and he gave לו את רחל him Rachel בתו his daughter לו לאשׁה׃ to wife}%
\verse{ויתן gave לבן And Laban לרחל to Rachel בתו his daughter את בלהה Bilhah שׁפחתו his handmaid לה לשׁפחה׃ to be her maid.}%
\verse{ויבא And he went in גם also אל unto רחל Rachel, ויאהב and he loved גם also את רחל Rachel מלאה ויעבד and served עמו with עוד him yet שׁבע seven שׁנים years. אחרות׃ other}%
\verseWithHeading{Jacob’s Children}{וירא saw יהוה And when the LORD כי that שׂנואה hated, לאה Leah ויפתח he opened את רחמה her womb: ורחל but Rachel עקרה׃ barren.}%
\verse{ותהר conceived, לאה And Leah ותלד and bore בן a son, ותקרא and she called שׁמו his name ראובן Reuben: כי for אמרה she said, כי Surely ראה hath looked יהוה the LORD בעניי upon my affliction; כי therefore עתה now יאהבני will love אישׁי׃ my husband}%
\verse{ותהר And she conceived עוד again, ותלד and bore בן a son; ותאמר and said, כי Because שׁמע hath heard יהוה the LORD כי that שׂנואה hated, אנכי I ויתן he hath therefore given לי גם also: את זה me this ותקרא and she called שׁמו his name שׁמעון׃ Simeon.}%
\verse{ותהר And she conceived עוד again, ותלד and bore בן a son; ותאמר and said, עתה Now הפעם this time ילוה be joined אישׁי will my husband אלי unto כי me, because ילדתי I have born לו שׁלשׁה him three בנים sons: על therefore כן therefore קרא called שׁמו was his name לוי׃ Levi.}%
\verse{ותהר And she conceived עוד again, ותלד and bore בן a son: ותאמר and she said, הפעם Now אודה will I praise את יהוה the LORD: על therefore כן therefore קראה she called שׁמו his name יהודה Judah; ותעמד and left מלדת׃ bearing.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 30
\verseWithHeading{Jacob’s Children}{ותרא saw רחל And when Rachel כי that לא ילדהיעקב Jacob, ותקנא envied רחל Rachel באחתה her sister; ותאמר and said אל unto יעקב הבה Give לי בנים me children, ואם or אין else מתה die. אנכי׃ I}%
\verse{ויחר was kindled אף anger יעקב And Jacob's ברחל against Rachel: ויאמר and he said, התחת stead, אלהים in God's אנכי I אשׁר who מנע hath withheld ממך from פרי thee the fruit בטן׃ of the womb?}%
\verse{ותאמר And she said, הנה Behold אמתי my maid בלהה Bilhah, בא go in אליה unto ותלד her; and she shall bear על upon ברכי my knees, ואבנה have children גם may also אנכי that I ממנה׃ by}%
\verse{ותתן And she gave לו את בלהה him Bilhah שׁפחתה her handmaid לאשׁה to wife: ויבא went in אליה unto יעקב׃ and Jacob}%
\verse{ותהר conceived, בלהה And Bilhah ותלד and bore ליעקב Jacob בן׃ a son.}%
\verse{ותאמר said, רחל And Rachel דנני hath judged אלהים God וגם me, and hath also שׁמע heard בקלי my voice, ויתן and hath given לי בן me a son: על therefore כן therefore קראה called שׁמו she his name דן׃ Dan.}%
\verse{ותהר conceived עוד again, ותלד and bore בלהה And Bilhah שׁפחת maid רחל Rachel's בן son. שׁני a second ליעקב׃ Jacob}%
\verse{ותאמר said, רחל And Rachel נפתולי wrestlings אלהים With great נפתלתי have I wrestled עם with אחתי my sister, גם and יכלתי I have prevailed: ותקרא and she called שׁמו his name נפתלי׃ Naphtali.}%
\verse{ותרא saw לאה When Leah כי that עמדה she had left מלדת bearing, ותקח she took את זלפה Zilpah שׁפחתה her maid, ותתן and gave אתה ליעקב her Jacob לאשׁה׃ to wife.}%
\verse{ותלד bore זלפה And Zilpah שׁפחת maid לאה Leah's ליעקב Jacob בן׃ a son.}%
\verse{ותאמר said, לאה And Leah בגד ותקרא and she called את שׁמו his name גד׃ Gad.}%
\verse{ותלד bore זלפה And Zilpah שׁפחת maid לאה Leah's בן son. שׁני a second ליעקב׃ Jacob}%
\verse{ותאמר said, לאה And Leah באשׁרי Happy כי am I, for אשׁרוני will call me blessed: בנות the daughters ותקרא and she called את שׁמו his name אשׁר׃ Asher.}%
\verse{וילך went ראובן And Reuben בימי in the days קציר harvest, חטים of wheat וימצא and found דודאים mandrakes בשׂדה in the field, ויבא and brought אתם אל them unto לאה Leah. אמו his mother ותאמר said רחל Then Rachel אל to לאה Leah, תני Give נא me, I pray thee, לי מדודאי mandrakes. בנך׃ of thy son's}%
\verse{ותאמר And she said לה המעט unto her, a small matter קחתך that thou hast taken את אישׁילקחת and wouldest thou take away גם also? את דודאי mandrakes בני my son's ותאמר said, רחל And Rachel לכן Therefore ישׁכב he shall lie עמך with הלילה thee tonight תחת for דודאי mandrakes. בנך׃ thy son's}%
\verse{ויבא came יעקב And Jacob מן out of השׂדה the field בערב in the evening, ותצא went out לאה and Leah לקראתו to meet ותאמר him, and said, אלי unto תבוא Thou must come in כי me; for שׂכר שׂכרתיךדודאי mandrakes. בני thee with my son's וישׁכב And he lay עמה with בלילה night. הוא׃ her that}%
\verse{וישׁמע hearkened אלהים And God אל unto לאה Leah, ותהר and she conceived, ותלד and bore ליעקב Jacob בן son. חמישׁי׃ the fifth}%
\verse{ותאמר said, לאה And Leah נתן hath given אלהים God שׂכרי me my hire, אשׁר because נתתי I have given שׁפחתי my maiden לאישׁי to my husband: ותקרא and she called שׁמו his name ישׂשכר׃ Issachar.}%
\verse{ותהר conceived עוד again, לאה And Leah ותלד and bore בן son. שׁשׁי the sixth ליעקב׃ Jacob}%
\verse{ותאמר said, לאה And Leah זבדני hath endued אלהים God אתי זבד dowry; טוב me a good הפעם now יזבלני dwell with אישׁי will my husband כי me, because ילדתי I have born לו שׁשׁה him six בנים sons: ותקרא and she called את שׁמו his name זבלון׃ Zebulun.}%
\verse{ואחר And afterwards ילדה she bore בת a daughter, ותקרא and called את שׁמה her name דינה׃ Dinah.}%
\verse{ויזכר remembered אלהים And God את רחל Rachel, וישׁמע hearkened אליה to אלהים and God ויפתח her, and opened את רחמה׃ her womb.}%
\verse{ותהר And she conceived, ותלד and bore בן a son; ותאמר and said, אסף hath taken away אלהים God את חרפתי׃ my reproach:}%
\verse{ותקרא And she called את שׁמו his name יוסף Joseph; לאמר and said, יסף shall add יהוה The LORD לי בן son. אחר׃ to me another}%
\verseWithHeading{Jacob’s Prosperity}{ויהי And it came to pass, כאשׁר when ילדה had born רחל Rachel את יוסף Joseph, ויאמר said יעקב that Jacob אל unto לבן Laban, שׁלחני Send me away, ואלכה that I may go אל unto מקומי mine own place, ולארצי׃ and to my country.}%
\verse{תנה Give את נשׁי my wives ואת ילדי and my children, אשׁר for whom עבדתי I have served אתך בהן ואלכה thee, and let me go: כי for אתה thou ידעת knowest את עבדתי my service אשׁר which עבדתיך׃ I have done}%
\verse{ויאמר said אליו unto לבן And Laban אם if נא him, I pray thee, מצאתי I have found חן favor בעיניך in thine eyes, נחשׁתי I have learned by experience ויברכני hath blessed יהוה that the LORD בגללך׃ me for thy sake.}%
\verse{ויאמר And he said, נקבה שׂכרך me thy wages, עלי ואתנה׃ and I will give}%
\verse{ויאמר And he said אליו unto אתה him, Thou ידעת knowest את אשׁר how עבדתיך I have served ואת אשׁר thee, and how היה was מקנך thy cattle אתי׃ with}%
\verse{כי For מעט little אשׁר which היה thou hadst לך לפני before ויפרץ I and it is increased לרב unto a multitude; ויברך hath blessed יהוה and the LORD אתך לרגלי thee since my coming: ועתה and now מתי when אעשׂה provide גם also? אנכי shall I לביתי׃ for mine own house}%
\verse{ויאמר And he said, מה What אתן shall I give לך ויאמר said, יעקב thee? And Jacob לא Thou shalt not תתן give לי מאומה me any thing: אם if תעשׂה thou wilt do לי הדבר thing הזה this אשׁובה for me, I will again ארעה feed צאנך thy flock: אשׁמר׃ keep}%
\verse{אעבר I will pass through בכל all צאנך thy flock היום today, הסר removing משׁם from thence כל all שׂה cattle, נקד the speckled וטלוא and spotted וכל and all שׂה cattle חום the brown בכשׂבים among the sheep, וטלוא and the spotted ונקד and speckled בעזים among the goats: והיה and shall be שׂכרי׃ my hire.}%
\verse{וענתה answer בי צדקתי So shall my righteousness ביום for me in time מחר to come, כי when תבוא it shall come על for שׂכרי my hire לפניך before thy face: כל every one אשׁר that איננו not נקד speckled וטלוא and spotted בעזים among the goats, וחום and brown בכשׂבים among the sheep, גנוב shall be counted stolen הוא that אתי׃ with}%
\verse{ויאמר said, לבן And Laban הן Behold, לו I would יהי it might be כדברך׃ according to thy word.}%
\verse{ויסר And he removed ביום day ההוא that את התישׁים the he goats העקדים that were ringstreaked והטלאים and spotted, ואת כל and all העזים the she goats הנקדות that were speckled והטלאת and spotted, כל every one אשׁר that לבן had white בו וכל in it, and all חום the brown בכשׂבים among the sheep, ויתן and gave ביד into the hand בניו׃ of his sons.}%
\verse{וישׂם And he set דרך journey שׁלשׁת three ימים days' בינו between ובין יעקב himself and Jacob: ויעקב and Jacob רעה fed את צאן flocks. לבן of Laban's הנותרת׃ the rest}%
\verse{ויקח took לו יעקב And Jacob מקל him rods לבנה poplar, לח of green ולוז and of the hazel וערמון and chestnut tree; ויפצל and peeled בהן פצלות streaks לבנות white מחשׂף appear הלבן and made the white אשׁר which על in המקלות׃ the rods.}%
\verse{ויצג And he set את המקלות the rods אשׁר which פצל he had peeled ברהטים in the gutters בשׁקתות troughs המים in the watering אשׁר when תבאן came הצאן the flocks לשׁתות to drink, לנכח before הצאן the flocks ויחמנה that they should conceive בבאן when they came לשׁתות׃ to drink.}%
\verse{ויחמו conceived הצאן And the flocks אל before המקלות the rods, ותלדן and brought forth הצאן cattle עקדים ringstreaked, נקדים speckled, וטלאים׃ and spotted.}%
\verse{והכשׂבים the lambs, הפריד did separate יעקב And Jacob ויתן and set פני the faces הצאן of the flocks אל toward עקד the ringstreaked, וכל and all חום the brown בצאן in the flock לבן of Laban; וישׁת and he put לו עדרים his own flocks לבדו by themselves, ולא them not שׁתם and put על unto צאן cattle. לבן׃ Laban's}%
\verse{והיה And it came to pass, בכל whensoever יחם did conceive, הצאן cattle המקשׁרות the stronger ושׂם laid יעקב that Jacob את המקלות the rods לעיני before the eyes הצאן of the cattle ברהטים in the gutters, ליחמנה that they might conceive במקלות׃ among the rods.}%
\verse{ובהעטיף were feeble, הצאן But when the cattle לא not ישׂים he put והיה were העטפים in: so the feebler ללבן Laban's, והקשׁרים and the stronger ליעקב׃ Jacob's.}%
\verse{ויפרץ increased האישׁ And the man מאד exceedingly, מאד exceedingly, ויהי and had לו צאן cattle, רבות much ושׁפחות and maidservants, ועבדים and menservants, וגמלים and camels, וחמרים׃ and asses.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 31
\verseWithHeading{Jacob Flees from Laban}{וישׁמע And he heard את דברי the words בני sons, לבן of Laban's לאמר saying, לקח hath taken away יעקב Jacob את כל all אשׁר that לאבינו our father's; ומאשׁר and of which לאבינו our father's עשׂה hath he gotten את כל all הכבד glory. הזה׃ this}%
\verse{וירא beheld יעקב And Jacob את פני the countenance לבן of Laban, והנה and, behold, איננו it not עמו toward כתמול him as before. שׁלשׁום׃ him as before.}%
\verse{ויאמר said יהוה And the LORD אל unto יעקב Jacob, שׁוב Return אל unto ארץ the land אבותיך of thy fathers, ולמולדתך and to thy kindred; ואהיה and I will be עמך׃ with}%
\verse{וישׁלח sent יעקב And Jacob ויקרא and called לרחל Rachel וללאה and Leah השׂדה to the field אל unto צאנו׃ his flock,}%
\verse{ויאמר And said להן ראה see אנכי unto them, I את פני countenance, אביכן your father's כי that איננו it not אלי toward כתמל me as before; שׁלשׁם me as before; ואלהי but the God אבי of my father היה hath been עמדי׃}%
\verse{ואתנה And ye ידעתן know כי that בכל with all כחי my power עבדתי I have served את אביכן׃ your father.}%
\verse{ואביכן And your father התל hath deceived בי והחלף me, and changed את משׂכרתי my wages עשׂרת ten מנים times; ולא him not נתנו suffered אלהים but God להרע to hurt עמדי׃ to hurt}%
\verse{אם If כה thus, יאמר he said נקדים The speckled יהיה shall be שׂכרך thy wages; וילדו bore כל then all הצאן the cattle נקדים speckled: ואם and if כה thus, יאמר he said עקדים The ringstreaked יהיה shall be שׂכרך thy hire; וילדו then bore כל all הצאן the cattle עקדים׃ ringstreaked.}%
\verse{ויצל hath taken away אלהים Thus God את מקנה the cattle אביכם of your father, ויתן׃ and given}%
\verse{ויהי And it came to pass בעת at the time יחם conceived, הצאן that the cattle ואשׂא that I lifted up עיני mine eyes, וארא and saw בחלום in a dream, והנה and, behold, העתדים the rams העלים which leaped על upon הצאן the cattle עקדים ringstreaked, נקדים speckled, וברדים׃ and grizzled.}%
\verse{ויאמר spoke אלי unto מלאך And the angel האלהים of God בחלום me in a dream, יעקב Jacob: ואמר And I said, הנני׃ Here}%
\verse{ויאמר And he said, שׂא Lift up נא now עיניך thine eyes, וראה and see, כל all העתדים the rams העלים which leap על upon הצאן the cattle עקדים ringstreaked, נקדים speckled, וברדים and grizzled: כי for ראיתי I have seen את כל all אשׁר that לבן Laban עשׂה׃ doeth}%
\verse{אנכי I האל the God בית אל of Bethel, אשׁר where משׁחת thou anointedst שׁם where מצבה the pillar, אשׁר where נדרת thou vowedst לי שׁם where נדר a vow עתה unto me: now קום arise, צא get thee out מן from הארץ land, הזאת this ושׁוב and return אל unto ארץ the land מולדתך׃ of thy kindred.}%
\verse{ותען answered רחל And Rachel ולאה and Leah ותאמרנה and said לו העוד unto him, yet לנו חלק any portion ונחלה or inheritance בבית house? אבינו׃ for us in our father's}%
\verse{הלוא Are we not נכריות of him strangers? נחשׁבנו counted לו כי for מכרנו he hath sold ויאכל us, and hath quite devoured גם also אכול us, and hath quite devoured את כספנו׃ our money.}%
\verse{כי For כל all העשׁר the riches אשׁר which הציל hath taken אלהים God מאבינו from our father, לנו הוא that ולבנינו ours, and our children's: ועתה now כל then, whatsoever אשׁר then, whatsoever אמר hath said אלהים God אליך unto עשׂה׃ thee, do.}%
\verse{ויקם rose up, יעקב Then Jacob וישׂא and set את בניו his sons ואת נשׁיו and his wives על upon הגמלים׃ camels;}%
\verse{וינהג And he carried away את כל all מקנהו his cattle, ואת כל and all רכשׁו his goods אשׁר which רכשׁ he had gotten, מקנה the cattle קנינו of his getting, אשׁר which רכשׁ he had gotten בפדן ארם in Padan-aram, לבוא for to go אל to יצחק Isaac אביו his father ארצה in the land כנען׃ of Canaan.}%
\verse{ולבן And Laban הלך went לגזז to shear את צאנו his sheep: ותגנב had stolen רחל and Rachel את התרפים the images אשׁר that לאביה׃ her father's.}%
\verse{ויגנב stole away יעקב And Jacob את לב unawares לבן to Laban הארמי the Syrian, על in that בלי him not הגיד he told לו כי that ברח fled. הוא׃ he}%
\verse{ויברח fled הוא So he וכל with all אשׁר that לו ויקם he had; and he rose up, ויעבר and passed over את הנהר the river, וישׂם and set את פניו his face הר the mount הגלעד׃ Gilead.}%
\verse{ויגד And it was told ללבן Laban ביום day השׁלישׁי on the third כי that ברח was fled. יעקב׃ Jacob}%
\verse{ויקח And he took את אחיו his brethren עמו with וירדף him, and pursued אחריו after דרך journey; שׁבעת him seven ימים days' וידבק and they overtook אתו בהר him in the mount הגלעד׃ Gilead.}%
\verse{ויבא came אלהים And God אל to לבן Laban הארמי the Syrian בחלם in a dream הלילה by night, ויאמר and said לו השׁמר unto him, Take heed לך פן that תדבר thou speak עם not to יעקב Jacob מטוב either good עד or רע׃ bad.}%
\verse{וישׂג overtook לבן Then Laban את יעקב Jacob. ויעקב Now Jacob תקע had pitched את אהלו his tent בהר in the mount: ולבן and Laban תקע pitched את with אחיו his brethren בהר in the mount הגלעד׃ of Gilead.}%
\verse{ויאמר said לבן And Laban ליעקב to Jacob, מה What עשׂית hast thou done, ותגנב that thou hast stolen away את לבבי unawares ותנהג to me, and carried away את בנתי my daughters, כשׁביות as captives חרב׃ with the sword?}%
\verse{למה Wherefore נחבאת didst thou flee away לברח secretly, ותגנב and steal away אתי ולא from me; and didst not הגדת tell לי ואשׁלחך me, that I might have sent thee away בשׂמחה with mirth, ובשׁרים and with songs, בתף with tabret, ובכנור׃ and with harp?}%
\verse{ולא And hast not נטשׁתני suffered לנשׁק me to kiss לבני my sons ולבנתי and my daughters? עתה thou hast now הסכלת done foolishly עשׂו׃ in doing.}%
\verse{ישׁ It is לאל in the power ידי of my hand לעשׂות to do עמכם not to רע you hurt: ואלהי but the God אביכם of your father אמשׁ me last night, אמר spoke אלי unto לאמר saying, השׁמר Take thou heed לך מדבר that thou speak עם יעקב Jacob מטוב either good עד or רע׃ bad.}%
\verse{ועתה And now, הלך thou wouldest needs be gone, הלכת thou wouldest needs be gone, כי because נכסף thou sore longedst נכספתה thou sore longedst לבית house, אביך after thy father's למה wherefore גנבת hast thou stolen את אלהי׃ my gods?}%
\verse{ויען answered יעקב And Jacob ויאמר and said ללבן to Laban, כי Because יראתי I was afraid: כי for אמרתי I said, פן Peradventure תגזל thou wouldest take by force את בנותיך thy daughters מעמי׃ from}%
\verse{עם With אשׁר whomsoever תמצא thou findest את אלהיך thy gods, לא let him not יחיה live: נגד before אחינו our brethren הכר discern לך מה thou what עמדי וקח me, and take לך ולא not ידע knew יעקב to thee. For Jacob כי that רחל Rachel גנבתם׃ had stolen}%
\verse{ויבא went לבן And Laban באהל tent, יעקב into Jacob's ובאהל tent, לאה and into Leah's ובאהל tents; שׁתי and into the two האמהת maidservants' ולא not. מצא but he found ויצא Then went מאהל tent. לאה ויבא and entered באהל רחל׃ into Rachel's}%
\verse{ורחל Now Rachel לקחה had taken את התרפים the images, ותשׂמם and put בכר furniture, הגמל them in the camel's ותשׁב and sat עליהם upon וימשׁשׁ searched לבן them. And Laban את כל all האהל the tent, ולא not. מצא׃ but found}%
\verse{ותאמר And she said אל to אביה her father, אל Let it not יחר displease בעיני displease אדני my lord כי that לוא not אוכל לקום rise up מפניך before כי thee; for דרך the custom נשׁים of women לי ויחפשׂ upon me. And he searched, ולא מצא but found את התרפים׃ the images.}%
\verse{ויחר was wroth, ליעקב And Jacob וירב and chided בלבן with Laban: ויען answered יעקב and Jacob ויאמר and said ללבן to Laban, מה What פשׁעי my trespass? מה what חטאתי my sin, כי that דלקת thou hast so hotly pursued אחרי׃ after}%
\verse{כי Whereas משׁשׁת thou hast searched את כל all כלי my stuff, מה what מצאת hast thou found מכל of all כלי stuff? ביתך thy household שׂים set כה here נגד before אחי my brethren ואחיך and thy brethren, ויוכיחו that they may judge בין between שׁנינו׃ us both.}%
\verse{זה This עשׂרים twenty שׁנה years אנכי I עמך with רחליך thee; thy ewes ועזיך and thy she goats לא have not שׁכלו cast their young, ואילי and the rams צאנך of thy flock לא have I not אכלתי׃ eaten.}%
\verse{טרפה That which was torn לא not הבאתי I brought אליך unto אנכי thee; I אחטנה bore the loss מידי of it; of my hand תבקשׁנה didst thou require גנבתי it, stolen יום by day, וגנבתי or stolen לילה׃ by night.}%
\verse{הייתי I was; ביום in the day אכלני consumed חרב the drought וקרח me, and the frost בלילה by night; ותדד departed שׁנתי and my sleep מעיני׃ from mine eyes.}%
\verse{זה Thus לי עשׂרים have I been twenty שׁנה years בביתך in thy house; עבדתיך I served ארבע thee fourteen עשׂרה thee fourteen שׁנה years בשׁתי for thy two בנתיך daughters, ושׁשׁ and six שׁנים years בצאנך for thy cattle: ותחלף and thou hast changed את משׂכרתי my wages עשׂרת ten מנים׃ times.}%
\verse{לולי Except אלהי the God אבי of my father, אלהי the God אברהם of Abraham, ופחד and the fear יצחק of Isaac, היה had been לי כי with me, surely עתה now ריקם empty. שׁלחתני thou hadst sent me away את עניי mine affliction ואת יגיע and the labor כפי of my hands, ראה hath seen אלהים God ויוכח and rebuked אמשׁ׃ last night.}%
\verse{ויען answered לבן And Laban ויאמר and said אל unto יעקב Jacob, הבנות daughters בנתי my daughters, והבנים and children בני my children, והצאן and cattle צאני my cattle, וכל and all אשׁר that אתה thou ראה seest לי הוא ולבנתי my daughters, מה mine: and what אעשׂה can I do לאלה unto these היום this day או or לבניהן unto their children אשׁר which ילדו׃ they have born?}%
\verse{ועתה Now לכה therefore come נכרתה thou, let us make ברית a covenant, אני I ואתה and thou; והיה and let it be לעד for a witness ביני between ובינך׃}%
\verse{ויקח took יעקב And Jacob אבן a stone, וירימה and set it up מצבה׃ a pillar.}%
\verse{ויאמר said יעקב And Jacob לאחיו unto his brethren, לקטו Gather אבנים stones; ויקחו and they took אבנים stones, ויעשׂו and made גל an heap: ויאכלו and they did eat שׁם there על upon הגל׃ the heap.}%
\verse{ויקרא called לו לבן And Laban יגר שׂהדותא it Jegar-sahadutha: ויעקב but Jacob קרא called לו גלעד׃ it Galeed.}%
\verse{ויאמר said, לבן And Laban הגל heap הזה This עד a witness ביני between ובינך היום me and thee this day. על כןרא of it called שׁמו was the name גלעד׃ Galeed;}%
\verse{והמצפה And Mizpah; אשׁר for אמר he said, יצף watch יהוה The LORD ביני between ובינך כי me and thee, when נסתר we are absent אישׁ מרעהו׃ from another.}%
\verse{אם If תענה thou shalt afflict את בנתי my daughters, ואם or if תקח thou shalt take נשׁים wives על beside בנתי my daughters, אין no אישׁ man עמנו with ראה us; see, אלהים God עד witness ביני between ובינך׃}%
\verse{ויאמר said לבן And Laban ליעקב to Jacob, הנה Behold הגל heap, הזה this והנה and behold המצבה pillar, אשׁר which יריתי I have cast ביני between ובינך׃}%
\verse{עד witness, הגל heap הזה This ועדה witness, המצבה and pillar אם that אני I לא will not אעבר pass over אליך to את הגל heap הזה this ואם thee, and that אתה thou לא shalt not תעבר pass over אלי unto את הגל heap הזה this ואת המצבה pillar הזאת and this לרעה׃ me, for harm.}%
\verse{אלהי The God אברהם of Abraham, ואלהי and the God נחור of Nahor, ישׁפטו judge בינינו between אלהי the God אביהם of their father, וישׁבע swore יעקב us. And Jacob בפחד by the fear אביו of his father יצחק׃ Isaac.}%
\verse{ויזבח offered יעקב Then Jacob זבח sacrifice בהר upon the mount, ויקרא and called לאחיו his brethren לאכל to eat לחם bread: ויאכלו and they did eat לחם bread, וילינו and tarried all night בהר׃ in the mount.}%
\verse{וישׁכם rose up, לבן Laban בבקר And early in the morning וינשׁק and kissed לבניו his sons ולבנותיו and his daughters, ויברך and blessed אתהם וילך departed, וישׁב and returned לבן them: and Laban למקמו׃ unto his place.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 32
\verseWithHeading{Jacob Fears Esau}{ויעקב And Jacob הלך went לדרכו on his way, ויפגעו met בו מלאכי and the angels אלהים׃ of God}%
\verse{ויאמר them, he said, יעקב Jacob כאשׁר And when ראם saw מחנה host: אלהים God's זה This ויקרא and he called שׁם the name המקום place ההוא of that מחנים׃ Mahanaim.}%
\verse{וישׁלח sent יעקב And Jacob מלאכים messengers לפניו before אל him to עשׂו Esau אחיו his brother ארצה unto the land שׂעיר of Seir, שׂדה the country אדום׃ of Edom.}%
\verse{ויצו And he commanded אתם לאמר them, saying, כה Thus תאמרון shall ye speak לאדני unto my lord לעשׂו Esau; כה thus, אמר saith עבדך Thy servant יעקב Jacob עם with לבן Laban, גרתי I have sojourned ואחר and stayed there עד until עתה׃ now:}%
\verse{ויהי And I have לי שׁור oxen, וחמור and asses, צאן flocks, ועבד and menservants, ושׁפחה and womenservants: ואשׁלחה and I have sent להגיד to tell לאדני my lord, למצא that I may find חן grace בעיניך׃ in thy sight.}%
\verse{וישׁבו returned המלאכים And the messengers אל to יעקב Jacob, לאמר saying, באנו We came אל to אחיך thy brother אל עשׂו Esau, וגם and also הלך he cometh לקראתך to meet וארבע thee, and four מאות hundred אישׁ men עמו׃ with}%
\verse{ויירא was greatly afraid יעקב Then Jacob מאד was greatly afraid ויצר and distressed: לו ויחץ and he divided את העם the people אשׁר that אתו with ואת הצאן him, and the flocks, ואת הבקר and herds, והגמלים and the camels, לשׁני into two מחנות׃ bands;}%
\verse{ויאמר And said, אם If יבוא come עשׂו Esau אל to המחנה company, האחת the one והכהו and smite והיה המחנה it, then the other company הנשׁאר which is left לפליטה׃ shall escape.}%
\verse{ויאמר said, יעקב And Jacob אלהי O God אבי of my father אברהם Abraham, ואלהי and God אבי of my father יצחק Isaac, יהוה the LORD האמר which saidst אלי unto שׁוב me, Return לארצך unto thy country, ולמולדתך and to thy kindred, ואיטיבה and I will deal well עמך׃ with}%
\verse{קטנתי I am not worthy מכל of the least of all החסדים the mercies, ומכל and of all האמת the truth, אשׁר which עשׂית thou hast showed את עבדך unto thy servant; כי for במקלי with my staff עברתי I passed over את הירדן Jordan; הזה this ועתה and now הייתי I am become לשׁני two מחנות׃ bands.}%
\verse{הצילני Deliver נא me, I pray thee, מיד from the hand אחי of my brother, מיד from the hand עשׂו of Esau: כי for ירא אנכי I אתו פן him, lest יבוא he will come והכני and smite אם me, the mother על with בנים׃ the children.}%
\verse{ואתה And thou אמרת saidst, היטב איטיבמך ושׂמתי and make את זרעך thy seed כחול as the sand הים of the sea, אשׁר which לא cannot יספר be numbered מרב׃ for multitude.}%
\verse{וילן And he lodged שׁם there בלילה night; ההוא that same ויקח and took מן of הבא that which came בידו to his hand מנחה a present לעשׂו for Esau אחיו׃ his brother;}%
\verse{עזים she goats, מאתים Two hundred ותישׁים he goats, עשׂרים and twenty רחלים ewes, מאתים two hundred ואילים rams, עשׂרים׃ and twenty}%
\verse{גמלים camels מיניקות milch ובניהם with their colts, שׁלשׁים Thirty פרות kine, ארבעים forty ופרים bulls, עשׂרה and ten אתנת she asses, עשׂרים twenty ועירם foals. עשׂרה׃ and ten}%
\verse{ויתן And he delivered ביד into the hand עבדיו of his servants, עדר every drove עדר every drove לבדו by themselves; ויאמר and said אל unto עבדיו his servants, עברו Pass over לפני before ורוח a space תשׂימו me, and put בין between עדר drove ובין עדר׃ and drove.}%
\verse{ויצו And he commanded את הראשׁון the foremost, לאמר saying, כי When יפגשׁך meeteth עשׂו Esau אחי my brother ושׁאלך thee, and asketh לאמר thee, saying, למי Whose אתה thou? ואנה and whither תלך goest ולמי thou? and whose אלה these לפניך׃ before}%
\verse{ואמרת Then thou shalt say, לעבדך thy servant ליעקב Jacob's; מנחה a present הוא it שׁלוחה sent לאדני unto my lord לעשׂו Esau: והנה and, behold, גם also הוא he אחרינו׃ behind}%
\verse{ויצו commanded גם And so את השׁני he the second, גם and את השׁלישׁי the third, גם and את כל all ההלכים that followed אחרי that followed העדרים the droves, לאמר saying, כדבר manner הזה On this תדברון shall ye speak אל unto עשׂו Esau, במצאכם when ye find אתו׃}%
\verse{ואמרתם And say גם ye moreover, הנה Behold, עבדך thy servant יעקב Jacob אחרינו behind כי us. For אמר he said, אכפרה פניו before במנחה him with the present ההלכת that goeth לפני his face; ואחרי me, and afterward כן me, and afterward אראה I will see פניו he will accept אולי peradventure ישׂא he will accept פני׃}%
\verse{ותעבר So went המנחה the present על over before פניו over before והוא him: and himself לן lodged בלילה night ההוא that במחנה׃ in the company.}%
\verseWithHeading{Jacob Wrestles with God}{ויקם And he rose up בלילה night, הוא that ויקח and took את שׁתי his two נשׁיו wives, ואת שׁתי and his two שׁפחתיו womenservants, ואת אחד and his eleven עשׂר and his eleven ילדיו sons, ויעבר and passed over את מעבר the ford יבק׃ Jabbok.}%
\verse{ויקחם And he took ויעברם them, and sent them over את הנחל the brook, ויעבר and sent over את אשׁר׃ that}%
\verse{ויותר was left יעקב And Jacob לבדו alone; ויאבק and there wrestled אישׁ עמו with עד him until עלות the breaking השׁחר׃ of the day.}%
\verse{וירא And when he saw כי that לא not יכל he prevailed לו ויגע against him, he touched בכף the hollow ירכו of his thigh; ותקע was out of joint, כף and the hollow ירך thigh יעקב of Jacob's בהאבקו as he wrestled עמו׃ with}%
\verse{ויאמר And he said, שׁלחני Let me go, כי for עלה breaketh. השׁחר the day ויאמר And he said, לא I will not אשׁלחך let thee go, כי except אם except ברכתני׃ thou bless}%
\verse{ויאמר And he said אליו unto מה him, What שׁמך thy name? ויאמר And he said, יעקב׃ Jacob.}%
\verse{ויאמר And he said, לא no יעקב Jacob, יאמר shall be called עוד more שׁמך Thy name כי but אם but ישׂראל Israel: כי for שׂרית as a prince hast thou power עם with אלהים God ועם and with אנשׁים men, ותוכל׃ and hast prevailed.}%
\verse{וישׁאל asked יעקב And Jacob ויאמר and said, הגידה Tell נא I pray thee, שׁמך thy name. ויאמר And he said, למה Wherefore זה it תשׁאל thou dost ask לשׁמי after my name? ויברך And he blessed אתו שׁם׃ him there.}%
\verse{ויקרא called יעקב And Jacob שׁם the name המקום of the place פניאל Peniel: כי for ראיתי I have seen אלהים God פנים face אל to פנים face, ותנצל is preserved. נפשׁי׃ and my life}%
\verse{ויזרח rose לו השׁמשׁ the sun כאשׁר And as עבר he passed over את פנואל Penuel והוא upon him, and he צלע halted על upon ירכו׃ his thigh.}%
\verse{על upon כן לא not יאכלו eat בני the children ישׂראל of Israel את גיד the sinew הנשׁה which shrank, אשׁר which על כף the hollow הירך of the thigh, עד unto היום day: הזה this כי because נגע he touched בכף the hollow ירך thigh יעקב of Jacob's בגיד in the sinew הנשׁה׃ that shrank.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 33
\verseWithHeading{Jacob Meets Esau and Settles at Shechem}{וישׂא lifted up יעקב And Jacob עיניו his eyes, וירא and looked, והנה and, behold, עשׂו Esau בא came, ועמו and with ארבע him four מאות hundred אישׁ men. ויחץ And he divided את הילדים the children על unto לאה Leah, ועל and unto רחל Rachel, ועל and unto שׁתי the two השׁפחות׃ handmaids.}%
\verse{וישׂם And he put את השׁפחות the handmaids ואת ילדיהן and their children ראשׁנה foremost, ואת לאה and Leah וילדיה and her children אחרנים after, ואת רחל and Rachel ואת יוסף and Joseph אחרנים׃ hindmost.}%
\verse{והוא And he עבר passed over לפניהם before וישׁתחו them, and bowed himself ארצה to the ground שׁבע seven פעמים times, עד until גשׁתו he came near עד to אחיו׃ his brother.}%
\verse{וירץ ran עשׂו And Esau לקראתו to meet ויחבקהו him, and embraced ויפל him, and fell על on צוארו his neck, וישׁקהו and kissed ויבכו׃ him: and they wept.}%
\verse{וישׂא And he lifted up את עיניו his eyes, וירא and saw את הנשׁים the women ואת הילדים and the children; ויאמר and said, מי Who אלה those לך ויאמר with thee? And he said, הילדים The children אשׁר which חנן hath graciously given אלהים God את עבדך׃ thy servant.}%
\verse{ותגשׁן came near, השׁפחות Then the handmaidens הנה they וילדיהן and their children, ותשׁתחוין׃ and they bowed themselves.}%
\verse{ותגשׁ came near, גם also לאה And Leah וילדיה with her children וישׁתחוו and bowed themselves: ואחר and after נגשׁ יוסףרחל and Rachel, וישׁתחוו׃ and they bowed themselves.}%
\verse{ויאמר And he said, מי What לך כל thou by all המחנה drove הזה this אשׁר which פגשׁתי I met? ויאמר And he said, למצא to find חן grace בעיני in the sight אדני׃ of my lord.}%
\verse{ויאמר said, עשׂו And Esau ישׁ I have לי רב enough, אחי my brother; יהי keep לך אשׁר׃ that}%
\verse{ויאמר said, יעקב And Jacob אל Nay, נא I pray thee, אם if נא now מצאתי I have found חן grace בעיניך in thy sight, ולקחת then receive מנחתי my present מידי at my hand: כי for על therefore כן therefore ראיתי I have seen פניך thy face, כראת as though I had seen פני the face אלהים of God, ותרצני׃ and thou wast pleased}%
\verse{קח Take, נא I pray thee, את ברכתי my blessing אשׁר that הבאת is brought לך כי to thee; because חנני hath dealt graciously אלהים God וכי with me, and because ישׁ I have לי כל enough. ויפצר And he urged בו ויקח׃ him, and he took}%
\verse{ויאמר And he said, נסעה Let us take our journey, ונלכה and let us go, ואלכה and I will go לנגדך׃ before}%
\verse{ויאמר And he said אליו unto אדני him, My lord ידע knoweth כי that הילדים the children רכים tender, והצאן and the flocks והבקר and herds עלות with young עלי with ודפקום me: and if men should overdrive יום day, אחד them one ומתו will die. כל all הצאן׃ the flock}%
\verse{יעבר pass over נא I pray thee, אדני Let my lord, לפני before עבדו his servant: ואני and I אתנהלה will lead on לאטי softly, לרגל according as המלאכה the cattle אשׁר that לפני goeth before ולרגל be able to endure, הילדים me and the children עד until אשׁר until אבא I come אל unto אדני my lord שׂעירה׃ unto Seir.}%
\verse{ויאמר said, עשׂו And Esau אציגה leave נא Let me now עמך with מן thee of העם the folk אשׁר that אתי with ויאמר me. And he said, למה What זה needeth it? אמצא let me find חן grace בעיני in the sight אדני׃ of my lord.}%
\verse{וישׁב returned ביום day ההוא that עשׂו So Esau לדרכו on his way שׂעירה׃ unto Seir.}%
\verse{ויעקב And Jacob נסע journeyed סכתה to Succoth, ויבן and built לו בית him a house, ולמקנהו for his cattle: עשׂה and made סכת booths על therefore כן therefore קרא is called שׁם the name המקום of the place סכות׃ Succoth.}%
\verse{ויבא came יעקב And Jacob שׁלם to Shalem, עיר a city שׁכם of Shechem, אשׁר which בארץ in the land כנען of Canaan, בבאו when he came מפדן ארם ויחן and pitched his tent את פני before העיר׃ the city.}%
\verse{ויקן And he bought את חלקת a parcel השׂדה of a field, אשׁר where נטה he had spread שׁם where אהלו his tent, מיד at the hand בני of the children חמור of Hamor, אבי father, שׁכם Shechem's במאה for a hundred קשׂיטה׃ pieces of money.}%
\verse{ויצב And he erected שׁם there מזבח an altar, ויקרא and called לו אל אלהישׂראל׃}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 34
\verseWithHeading{The Rape of Dinah and the Massacre at Shechem}{ותצא went out דינה And Dinah בת the daughter לאה of Leah, אשׁר which ילדה she bore ליעקב unto Jacob, לראות to see בבנות the daughters הארץ׃ of the land.}%
\verse{וירא saw אתה שׁכם And when Shechem בן the son חמור of Hamor החוי the Hivite, נשׂיא prince הארץ of the country, ויקח her, he took אתה וישׁכב her, and lay with אתה ויענה׃ her, and defiled}%
\verse{ותדבק cleaved נפשׁו And his soul בדינה unto Dinah בת the daughter יעקב of Jacob, ויאהב and he loved את הנער the damsel, וידבר and spoke על kindly לב kindly הנער׃ unto the damsel.}%
\verse{ויאמר spoke שׁכם And Shechem אל unto חמור Hamor, אביו his father לאמר saying, קח Get לי את הילדה damsel הזאת me this לאשׁה׃ to wife.}%
\verse{ויעקב And Jacob שׁמע heard כי that טמא he had defiled את דינה Dinah בתו his daughter: ובניו now his sons היו were את with מקנהו his cattle בשׂדה in the field: והחרשׁ held his peace יעקב and Jacob עד until באם׃ they were come.}%
\verse{ויצא went out חמור And Hamor אבי the father שׁכם אל unto יעקב Jacob לדבר to commune אתו׃ with}%
\verse{ובני And the sons יעקב of Jacob באו came מן out of השׂדה the field כשׁמעם when they heard ויתעצבו were grieved, האנשׁים and the men ויחר and they were very wroth, להם מאד and they were very wroth, כי because נבלה folly עשׂה he had wrought בישׂראל in Israel לשׁכב in lying with את בת daughter; יעקב Jacob's וכן which thing לא ought not יעשׂה׃ to be done.}%
\verse{וידבר communed חמור And Hamor אתם with לאמר them, saying, שׁכם Shechem בני of my son חשׁקה longeth נפשׁו The soul בבתכם for your daughter: תנו give נא I pray you אתה לו לאשׁה׃ her him to wife.}%
\verse{והתחתנו And make ye marriages אתנו with בנתיכם your daughters תתנו us, give לנו ואת בנתינו our daughters תקחו׃ unto us, and take}%
\verse{ואתנו with תשׁבו And ye shall dwell והארץ us: and the land תהיה shall be לפניכם before שׁבו you; dwell וסחרוה and trade והאחזו׃ ye therein, and get you possessions}%
\verse{ויאמר said שׁכם And Shechem אל unto אביה her father ואל and unto אחיה her brethren, אמצא Let me find חן grace בעיניכם in your eyes, ואשׁר and what תאמרו ye shall say אלי unto אתן׃ me I will give.}%
\verse{הרבו עליאד me never so much מהר dowry ומתן and gift, ואתנה and I will give כאשׁר according as תאמרו ye shall say אלי unto ותנו me: but give לי את הנער me the damsel לאשׁה׃ to wife.}%
\verse{ויענו answered בני And the sons יעקב of Jacob את שׁכם Shechem ואת חמור and Hamor אביו his father במרמה deceitfully, וידברו and said, אשׁר because טמא he had defiled את דינה Dinah אחתם׃ their sister:}%
\verse{ויאמרו And they said אליהם unto לא נוכלעשׂות do הדבר thing, הזה this לתת to give את אחתנו our sister לאישׁ אשׁר that לו ערלה is uncircumcised; כי for חרפה a reproach הוא׃ that}%
\verse{אך But בזאת in this נאות will we consent לכם אם unto you: If תהיו ye will be כמנו להמל of you be circumcised; לכם כל as we that every זכר׃ male}%
\verse{ונתנו Then will we give את בנתינו our daughters לכם ואת בנתיכם your daughters נקח unto you, and we will take לנו וישׁבנו to us, and we will dwell אתכם with והיינו you, and we will become לעם people. אחד׃ one}%
\verse{ואם But if לא ye will not תשׁמעו hearken אלינו unto להמול us, to be circumcised; ולקחנו then will we take את בתנו our daughter, והלכנו׃ and we will be gone.}%
\verse{וייטבו pleased דבריהם And their words בעיני pleased חמור Hamor, ובעיני שׁכם and Shechem בן son. חמור׃ Hamor's}%
\verse{ולא not אחר deferred הנער And the young man לעשׂות to do הדבר the thing, כי because חפץ he had delight בבת daughter: יעקב in Jacob's והוא and he נכבד more honorable מכל than all בית the house אביו׃ of his father.}%
\verse{ויבא came חמור And Hamor ושׁכם and Shechem בנו his son אל unto שׁער the gate עירם of their city, וידברו and communed אל with אנשׁי עירם of their city, לאמר׃ saying,}%
\verse{האנשׁים men האלה These שׁלמים peaceable הם אתנו with וישׁבו us; therefore let them dwell בארץ in the land, ויסחרו and trade אתה והארץ therein; for the land, הנה behold, רחבת large enough ידים large enough לפניהם for את בנתם their daughters נקח them; let us take לנו לנשׁים to us for wives, ואת בנתינו them our daughters. נתן׃ and let us give}%
\verse{אך Only בזאת herein יאתו consent לנו האנשׁים will the men לשׁבת unto us for to dwell אתנו with להיות us, to be לעם people, אחד one בהמול among us be circumcised, לנו כל if every זכר male כאשׁר as הם they נמלים׃ circumcised.}%
\verse{מקנהם their cattle וקנינם and their substance וכל and every בהמתם beast הלוא not לנו הם אך of theirs ours? only נאותה let us consent להם וישׁבו unto them, and they will dwell אתנו׃ with}%
\verse{וישׁמעו hearkened אל And unto חמור Hamor ואל and unto שׁכם Shechem בנו his son כל all יצאי that went out שׁער of the gate עירו of his city; וימלו was circumcised, כל and every זכר male כל all יצאי that went out שׁער of the gate עירו׃ of his city.}%
\verse{ויהי And it came to pass ביום day, השׁלישׁי on the third בהיותם when they were כאבים sore, ויקחו took שׁני that two בני of the sons יעקב of Jacob, שׁמעון Simeon ולוי and Levi, אחי brethren, דינה Dinah's אישׁ each man חרבו his sword, ויבאו and came על upon העיר the city בטח boldly, ויהרגו and slew כל all זכר׃ the males.}%
\verse{ואת חמור Hamor ואת שׁכם and Shechem בנו his son הרגו And they slew לפי with the edge חרב of the sword, ויקחו and took את דינה Dinah מבית house, שׁכם out of Shechem's ויצאו׃ and went out.}%
\verse{בני The sons יעקב of Jacob באו came על upon החללים the slain, ויבזו and spoiled העיר the city, אשׁר because טמאו they had defiled אחותם׃ their sister.}%
\verse{את צאנם their sheep, ואת בקרם and their oxen, ואת חמריהם and their asses, ואת אשׁר and that which בעיר in the city, ואת אשׁר and that which בשׂדה in the field, לקחו׃ They took}%
\verse{ואת כל And all חילם their wealth, ואת כל and all טפם their little ones, ואת נשׁיהם and their wives שׁבו took they captive, ויבזו and spoiled ואת כל even all אשׁר that בבית׃ in the house.}%
\verse{ויאמר said יעקב And Jacob אל to שׁמעון Simeon ואל לוי and Levi, עכרתם Ye have troubled אתי להבאישׁני me to make me to stink בישׁב among the inhabitants הארץ of the land, בכנעני among the Canaanites ובפרזי and the Perizzites: ואני and I מתי few מספר in number, ונאספו they shall gather themselves together עלי against והכוני me, and slay ונשׁמדתי me; and I shall be destroyed, אני I וביתי׃ and my house.}%
\verse{ויאמרו And they said, הכזונה as with an harlot? יעשׂה Should he deal with את אחותנו׃ our sister}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 35
\verseWithHeading{Jacob Goes Back to Bethel}{ויאמר said אלהים And God אל unto יעקב Jacob, קום Arise, עלה go up בית אל to Bethel, ושׁב and dwell שׁם there: ועשׂה and make שׁם there מזבח an altar לאל unto God, הנראה that appeared אליך unto בברחך thee when thou fleddest מפני from the face עשׂו of Esau אחיך׃ thy brother.}%
\verse{ויאמר said יעקב Then Jacob אל unto ביתו his household, ואל and to כל all אשׁר that עמו with הסרו him, Put away את אלהי gods הנכר the strange אשׁר that בתככם among והטהרו you, and be clean, והחליפו and change שׂמלתיכם׃ your garments:}%
\verse{ונקומה And let us arise, ונעלה and go up בית אל to Bethel; ואעשׂה and I will make שׁם there מזבח an altar לאל unto God, הענה who answered אתי ביום me in the day צרתי of my distress, ויהי and was עמדי בדרך me in the way אשׁר which הלכתי׃ I went.}%
\verse{ויתנו And they gave אל unto יעקב Jacob את כל all אלהי gods הנכר the strange אשׁר which בידם in their hand, ואת הנזמים and earrings אשׁר which באזניהם in their ears; ויטמן hid אתם יעקב and Jacob תחת them under האלה the oak אשׁר which עם by שׁכם׃ Shechem.}%
\verse{ויסעו And they journeyed: ויהי was חתת and the terror אלהים of God על upon הערים the cities אשׁר that סביבותיהם round about ולא them, and they did not רדפו pursue אחרי after בני the sons יעקב׃ of Jacob.}%
\verse{ויבא came יעקב So Jacob לוזה to Luz, אשׁר which בארץ in the land כנען of Canaan, הוא that בית אל Bethel, הוא he וכל and all העם the people אשׁר that עמו׃ with}%
\verse{ויבן And he built שׁם there מזבח an altar, ויקרא and called למקום the place אל בית אלי because שׁם there נגלו appeared אליו unto האלהים God בברחו him, when he fled מפני from the face אחיו׃ of his brother.}%
\verse{ותמת died, דברה But Deborah מינקת nurse רבקה Rebekah's ותקבר and she was buried מתחת beneath לבית אל Bethel תחת under האלון an oak: ויקרא of it was called שׁמו and the name אלון בכות׃ Allon-bachuth.}%
\verse{וירא appeared אלהים And God אל unto יעקב Jacob עוד again, בבאו when he came מפדן ארם ויברך and blessed אתו׃}%
\verse{ויאמר said לו אלהים And God שׁמך unto him, Thy name יעקב Jacob: לא shall not יקרא be called שׁמך thy name עוד any more יעקב Jacob, כי but אם but ישׂראל Israel יהיה shall be שׁמך thy name: ויקרא and he called את שׁמו his name ישׂראל׃ Israel.}%
\verse{ויאמר said לו אלהים And God אני unto him, I אל God שׁדי Almighty: פרה be fruitful ורבה and multiply; גוי a nation וקהל and a company גוים of nations יהיה shall be ממך of ומלכים thee, and kings מחלציך of thy loins; יצאו׃ shall come out}%
\verse{ואת הארץ And the land אשׁר which נתתי I gave לאברהם Abraham וליצחק and Isaac, לך אתננה to thee I will give ולזרעך it, and to thy seed אחריך after אתן thee will I give את הארץ׃ the land.}%
\verse{ויעל went up מעליו from אלהים And God במקום him in the place אשׁר where דבר he talked אתו׃ with}%
\verse{ויצב set up יעקב And Jacob מצבה a pillar במקום in the place אשׁר where דבר he talked אתו with מצבת him, a pillar אבן of stone: ויסך and he poured עליה thereon, נסך a drink offering ויצק and he poured עליה thereon. שׁמן׃ oil}%
\verse{ויקרא called יעקב And Jacob את שׁם the name המקום of the place אשׁר where דבר spoke אתו with שׁם אלהים God בית אל׃ him, Bethel.}%
\verseWithHeading{The Death of Rachel}{ויסעו And they journeyed מבית אל ויהי and there was עוד but כברת a little way הארץ a little way לבוא to come אפרתה to Ephrath: ותלד travailed, רחל and Rachel ותקשׁ and she had hard בלדתה׃ labor.}%
\verse{ויהי And it came to pass, בהקשׁתה when she was in hard בלדתה labor, ותאמר said לה המילדת that the midwife אל not; תיראי unto her, Fear כי גם also. זה thou shalt have this לך בן׃ son}%
\verse{ויהי And it came to pass, בצאת was in departing, נפשׁה as her soul כי (for מתה she died) ותקרא that she called שׁמו his name בן אוני Ben-oni: ואביו but his father קרא called לו בנימין׃ him Benjamin.}%
\verse{ותמת died, רחל And Rachel ותקבר and was buried בדרך in the way אפרתה to Ephrath, הוא which בית לחם׃ Bethlehem.}%
\verse{ויצב set יעקב And Jacob מצבה a pillar על upon קברתה her grave: הוא that מצבת the pillar קברת grave רחל of Rachel's עד unto היום׃ this day.}%
\verse{ויסע journeyed, ישׂראל And Israel ויט and spread אהלה his tent מהלאה beyond למגדל עדר׃ of Edar.}%
\verse{ויהי And it came to pass, בשׁכן dwelt ישׂראל when Israel בארץ land, ההוא in that וילך went ראובן that Reuben וישׁכב and lay with את בלהה Bilhah פילגשׁ concubine: אביו his father's וישׁמע heard ישׂראל and Israel ויהיו were בני Now the sons יעקב of Jacob שׁנים twelve: עשׂר׃ twelve:}%
\verse{בני The sons לאה of Leah; בכור firstborn, יעקב Jacob's ראובן Reuben, ושׁמעון and Simeon, ולוי and Levi, ויהודה and Judah, וישׂשכר and Issachar, וזבלון׃ and Zebulun:}%
\verse{בני The sons רחל of Rachel; יוסף Joseph, ובנימן׃ and Benjamin:}%
\verse{ובני And the sons בלהה of Bilhah, שׁפחת handmaid; רחל Rachel's דן Dan, ונפתלי׃ and Naphtali:}%
\verse{ובני And the sons זלפה of Zilpah, שׁפחת handmaid; לאה Leah's גד Gad, ואשׁר and Asher: אלה these בני the sons יעקב of Jacob, אשׁר which ילד were born לו בפדן ארם׃ to him in Padan-aram.}%
\verseWithHeading{The Death of Isaac}{ויבא came יעקב And Jacob אל unto יצחק Isaac אביו his father ממרא unto Mamre, קרית הארבע unto the city of Arbah, הוא which חברון Hebron, אשׁר where גר sojourned. שׁם where אברהם Abraham ויצחק׃ and Isaac}%
\verse{ויהיו were ימי And the days יצחק of Isaac מאת a hundred שׁנה years. ושׁמנים and fourscore שׁנה׃}%
\verse{ויגוע gave up the ghost, יצחק And Isaac וימת and died, ויאסף and was gathered אל unto עמיו his people, זקן old ושׂבע and full ימים of days: ויקברו buried אתו עשׂו Esau ויעקב and Jacob בניו׃ and his sons}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 36
\verseWithHeading{The Descendants of Esau}{ואלה Now these תלדות the generations עשׂו of Esau, הוא who אדום׃ Edom.}%
\verse{עשׂו Esau לקח took את נשׁיו his wives מבנות of the daughters כנען of Canaan; את עדה Adah בת the daughter אילון of Elon החתי the Hittite, ואת אהליבמה and Aholibamah בת the daughter ענה of Anah בת the daughter צבעון of Zibeon החוי׃ the Hivite;}%
\verse{ואת בשׂמת And Bashemath בת daughter, ישׁמעאל Ishmael's אחות sister נביות׃ of Nebajoth.}%
\verse{ותלד bore עדה And Adah לעשׂו to Esau את אליפז Eliphaz; ובשׂמת and Bashemath ילדה bore את רעואל׃ Reuel;}%
\verse{ואהליבמה And Aholibamah ילדה bore את יעישׁ Jeush, ואת יעלם and Jaalam, ואת קרח and Korah: אלה these בני the sons עשׂו of Esau, אשׁר which ילדו were born לו בארץ unto him in the land כנען׃ of Canaan.}%
\verse{ויקח took עשׂו And Esau את נשׁיו his wives, ואת בניו and his sons, ואת בנתיו and his daughters, ואת כל and all נפשׁות the persons ביתו of his house, ואת מקנהו and his cattle, ואת כל and all בהמתו his beasts, ואת כל and all קנינו his substance, אשׁר which רכשׁ he had got בארץ in the land כנען of Canaan; וילך and went אל into ארץ the country מפני from the face יעקב Jacob. אחיו׃ of his brother}%
\verse{כי For היה were רכושׁם their riches רב more than משׁבת that they might dwell יחדו together; ולא not יכלה could ארץ and the land מגוריהם wherein they were strangers לשׂאת bear אתם מפני them because מקניהם׃ of their cattle.}%
\verse{וישׁב Thus dwelt עשׂו Esau בהר in mount שׂעיר Seir: עשׂו Esau הוא אדום׃ Edom.}%
\verse{ואלה And these תלדות the generations עשׂו of Esau אבי the father אדום of the Edomites בהר in mount שׂעיר׃ Seir:}%
\verse{אלה These שׁמות the names בני sons; עשׂו of Esau's אליפז Eliphaz בן the son עדה of Adah אשׁת the wife עשׂו of Esau, רעואל Reuel בן the son בשׂמת of Bashemath אשׁת the wife עשׂו׃ of Esau.}%
\verse{ויהיו were בני And the sons אליפז of Eliphaz תימן Teman, אומר Omar, צפו Zepho, וגעתם and Gatam, וקנז׃ and Kenaz.}%
\verse{ותמנע And Timna היתה was פילגשׁ concubine לאליפז to Eliphaz בן son; עשׂו Esau's ותלד and she bore לאליפז to Eliphaz את עמלק Amalek: אלה these בני the sons עדה of Adah אשׁת wife. עשׂו׃ Esau's}%
\verse{ואלה And these בני the sons רעואל of Reuel; נחת Nahath, וזרח and Zerah, שׁמה Shammah, ומזה and Mizzah: אלה these היו were בני the sons בשׂמת of Bashemath אשׁת wife. עשׂו׃ Esau's}%
\verse{ואלה And these היו were בני the sons אהליבמה of Aholibamah, בת the daughter ענה of Anah בת the daughter צבעון of Zibeon, אשׁת wife: עשׂו Esau's ותלד and she bore לעשׂו to Esau את יעישׁ Jeush, ואת יעלם and Jaalam, ואת קרח׃ and Korah.}%
\verse{אלה These אלופי dukes בני of the sons עשׂו of Esau: בני the sons אליפז of Eliphaz בכור the firstborn עשׂו of Esau; אלוף duke תימן Teman, אלוף duke אומר Omar, אלוף duke צפו Zepho, אלוף duke קנז׃ Kenaz,}%
\verse{אלוף Duke קרח Korah, אלוף duke געתם Gatam, אלוף duke עמלק Amalek: אלה these אלופי the dukes אליפז of Eliphaz בארץ in the land אדום of Edom; אלה these בני the sons עדה׃ of Adah.}%
\verse{ואלה And these בני the sons רעואל of Reuel בן son; עשׂו Esau's אלוף duke נחת Nahath, אלוף duke זרח Zerah, אלוף duke שׁמה Shammah, אלוף duke מזה Mizzah: אלה these אלופי the dukes רעואל of Reuel בארץ in the land אדום of Edom; אלה these בני the sons בשׂמת of Bashemath אשׁת wife. עשׂו׃ Esau's}%
\verse{ואלה And these בני the sons אהליבמה of Aholibamah אשׁת wife; עשׂו Esau's אלוף duke יעושׁ Jeush, אלוף duke יעלם Jaalam, אלוף duke קרח Korah: אלה these אלופי the dukes אהליבמה of Aholibamah בת the daughter ענה of Anah, אשׁת wife. עשׂו׃ Esau's}%
\verse{אלה These בני the sons עשׂו of Esau, ואלה and these אלופיהם their dukes. הוא who אדום׃ Edom,}%
\verse{אלה These בני the sons שׂעיר of Seir החרי the Horite, ישׁבי who inhabited הארץ the land; לוטן Lotan, ושׁובל and Shobal, וצבעון and Zibeon, וענה׃ and Anah,}%
\verse{ודשׁון And Dishon, ואצר and Ezer, ודישׁן and Dishan: אלה these אלופי the dukes החרי of the Horites, בני the children שׂעיר of Seir בארץ in the land אדום׃ of Edom.}%
\verse{ויהיו were בני And the children לוטן of Lotan חרי Hori והימם and Hemam; ואחות sister לוטן and Lotan's תמנע׃ Timna.}%
\verse{ואלה these; בני And the children שׁובל of Shobal עלון Alvan, ומנחת and Manahath, ועיבל and Ebal, שׁפו Shepho, ואונם׃ and Onam.}%
\verse{ואלה And these בני the children צבעון of Zibeon; ואיה both Ajah, וענה and Anah: הוא this ענה Anah אשׁר that מצא found את הימם the mules במדבר in the wilderness, ברעתו as he fed את החמרים the asses לצבעון of Zibeon אביו׃ his father.}%
\verse{ואלה these; בני And the children ענה of Anah דשׁן Dishon, ואהליבמה and Aholibamah בת the daughter ענה׃ of Anah.}%
\verse{ואלה And these בני the children דישׁן חמדן Hemdan, ואשׁבן and Esh-ban, ויתרן and Ithran, וכרן׃ and Cheran.}%
\verse{אלה these; בני The children אצר of Ezer בלהן Bilhan, וזעון and Zaavan, ועקן׃ and Akan.}%
\verse{אלה these; בני The children דישׁן of Dishan עוץ Uz, וארן׃ and Aran.}%
\verse{אלה These אלופי the dukes החרי of the Horites; אלוף duke לוטן Lotan, אלוף duke שׁובל Shobal, אלוף duke צבעון Zibeon, אלוף duke ענה׃ Anah,}%
\verse{אלוף Duke דשׁן Dishon, אלוף duke אצר Ezer, אלוף duke דישׁן Dishan: אלה these אלופי the dukes החרי of Hori, לאלפיהם among their dukes בארץ in the land שׂעיר׃ of Seir.}%
\verseWithHeading{The Kings of Edom}{ואלה And these המלכים the kings אשׁר that מלכו reigned בארץ in the land אדום of Edom, לפני before מלך there reigned מלך any king לבני over the children ישׂראל׃ of Israel.}%
\verse{וימלך reigned באדום in Edom: בלע And Bela בן the son בעור of Beor ושׁם and the name עירו of his city דנהבה׃ Dinhabah.}%
\verse{וימת died, בלע And Bela וימלך reigned תחתיו in his stead. יובב and Jobab בן the son זרח of Zerah מבצרה׃}%
\verse{וימת died, יובב And Jobab וימלך reigned תחתיו in his stead. חשׁם and Husham מארץ of the land התימני׃ of Temani}%
\verse{וימת died, חשׁם And Husham וימלך reigned תחתיו in his stead: הדד and Hadad בן the son בדד of Bedad, המכה who smote את מדין Midian בשׂדה in the field מואב of Moab, ושׁם and the name עירו of his city עוית׃ Avith.}%
\verse{וימת died, הדד And Hadad וימלך reigned תחתיו in his stead. שׂמלה and Samlah ממשׂרקה׃}%
\verse{וימת died, שׂמלה And Samlah וימלך reigned תחתיו in his stead. שׁאול and Saul מרחבות הנהר׃ the river}%
\verse{וימת died, שׁאול And Saul וימלך reigned תחתיו in his stead. בעל חנן and Baalhanan בן the son עכבור׃ of Achbor}%
\verse{וימת died, בעל חנן And Baalhanan בן the son עכבור of Achbor וימלך reigned תחתיו in his stead: הדר and Hadar ושׁם and the name עירו of his city פעו Pau; ושׁם name אשׁתו and his wife's מהיטבאל Mehetabel, בת the daughter מטרד of Matred, בת the daughter מי זהב׃ of Mezahab.}%
\verse{ואלה And these שׁמות the names אלופי of the dukes עשׂו of Esau, למשׁפחתם according to their families, למקמתם after their places, בשׁמתם by their names; אלוף duke תמנע Timnah, אלוף duke עלוה Alvah, אלוף duke יתת׃ Jetheth,}%
\verse{אלוף Duke אהליבמה Aholibamah, אלוף duke אלה Elah, אלוף duke פינן׃ Pinon,}%
\verse{אלוף Duke קנז Kenaz, אלוף duke תימן Teman, אלוף duke מבצר׃ Mibzar,}%
\verse{אלוף Duke מגדיאל Magdiel, אלוף duke עירם Iram: אלה these אלופי the dukes אדום of Edom, למשׁבתם according to their habitations בארץ in the land אחזתם of their possession: הוא he עשׂו Esau אבי the father אדום׃ of the Edomites.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 37
\verseWithHeading{The Dreams of Joseph}{וישׁב dwelt יעקב And Jacob בארץ in the land מגורי was a stranger, אביו wherein his father בארץ in the land כנען׃ of Canaan.}%
\verse{אלה These תלדות the generations יעקב of Jacob. יוסף Joseph, בן old, שׁבע seventeen עשׂרה seventeen שׁנה years היה was רעה feeding את with אחיו his brethren; בצאן the flock והוא נער and the lad את with בני the sons בלהה of Bilhah, ואת and with בני the sons זלפה of Zilpah, נשׁי wives: אביו his father's ויבא brought יוסף and Joseph את דבתם report. רעה their evil אל unto אביהם׃ his father}%
\verse{וישׂראל Now Israel אהב loved את יוסף Joseph מכל more than all בניו his children, כי because בן the son זקנים of his old age: הוא he לו ועשׂה and he made לו כתנת him a coat פסים׃ of colors.}%
\verse{ויראו saw אחיו And when his brethren כי that אתו אהב loved אביהם their father מכל him more than all אחיו his brethren, וישׂנאו they hated אתו ולא not יכלו him, and could דברו speak לשׁלם׃ peaceably}%
\verse{ויחלם dreamed יוסף And Joseph חלום a dream, ויגד and he told לאחיו his brethren: ויוספו the more. עוד him yet שׂנא and they hated אתו׃}%
\verse{ויאמר And he said אליהם unto שׁמעו them, Hear, נא I pray you, החלום I have dreamed: הזה this אשׁר which חלמתי׃ dream}%
\verse{והנה For, behold, אנחנו we מאלמים binding אלמים sheaves בתוך in השׂדה the field, והנה and, lo, קמה arose, אלמתי my sheaf וגם and also נצבה stood upright; והנה and, behold, תסבינה stood round about, אלמתיכם your sheaves ותשׁתחוין and made obeisance לאלמתי׃ to my sheaf.}%
\verse{ויאמרו said לו אחיו And his brethren המלך תמלךלינו over אם us? or משׁול shalt thou indeed have dominion תמשׁל shalt thou indeed have dominion בנו ויוספו the more עוד him yet שׂנא over us? And they hated אתו על for חלמתיו his dreams, ועל and for דבריו׃ his words.}%
\verse{ויחלם And he dreamed עוד yet חלום dream, אחר another ויספר and told אתו לאחיו it his brethren, ויאמר and said, הנה Behold, חלמתי I have dreamed חלום a dream עוד more; והנה and, behold, השׁמשׁ the sun והירח and the moon ואחד and the eleven עשׂר and the eleven כוכבים stars משׁתחוים׃ made obeisance}%
\verse{ויספר And he told אל to אביו his father, ואל and to אחיו his brethren: ויגער rebuked בו אביו and his father ויאמר him, and said לו מה unto him, What החלום dream הזה this אשׁר that חלמת thou hast dreamed? הבוא indeed come נבוא indeed come אני Shall I ואמך and thy mother ואחיך and thy brethren להשׁתחות to bow down ourselves לך ארצה׃ to thee to the earth?}%
\verse{ויקנאו envied בו אחיו And his brethren ואביו him; but his father שׁמר observed את הדבר׃ the saying.}%
\verseWithHeading{Joseph Sold Into Slavery by his Brothers}{וילכו went אחיו And his brethren לרעות to feed את צאן flock אביהם their father's בשׁכם׃ in Shechem.}%
\verse{ויאמר said ישׂראל And Israel אל unto יוסף Joseph, הלוא Do not אחיך thy brethren רעים feed בשׁכם in Shechem? לכה come, ואשׁלחך and I will send אליהם thee unto ויאמר them. And he said לו הנני׃ to him, Here}%
\verse{ויאמר And he said לו לך to him, Go, נא I pray thee, ראה see את שׁלום whether it be well אחיך with thy brethren, ואת שׁלום and well הצאן with the flocks; והשׁבני and bring דבר me word וישׁלחהו again. So he sent מעמק him out of the vale חברון of Hebron, ויבא and he came שׁכמה׃ to Shechem.}%
\verse{וימצאהו found אישׁ והנה him, and, behold, תעה wandering בשׂדה in the field: וישׁאלהו asked האישׁ לאמר him, saying, מה What תבקשׁ׃ seekest}%
\verse{ויאמר And he said, את אחי my brethren: אנכי I מבקשׁ seek הגידה tell נא me, I pray thee, לי איפה where הם they רעים׃ feed}%
\verse{ויאמר said, האישׁ And the man נסעו They are departed מזה hence; כי for שׁמעתי I heard אמרים them say, נלכה Let us go דתינה to Dothan. וילך went יוסף And Joseph אחר after אחיו his brethren, וימצאם and found בדתן׃ them in Dothan.}%
\verse{ויראו And when they saw אתו מרחק him afar off, ובטרם even before יקרב he came near אליהם unto ויתנכלו them, they conspired אתו להמיתו׃ against him to slay}%
\verse{ויאמרו And they said אישׁ one אל to אחיו another, הנה Behold, בעל dreamer החלמות dreamer הלזה this בא׃ cometh.}%
\verse{ועתה now לכו Come ונהרגהו therefore, and let us slay ונשׁלכהו him, and cast באחד him into some הברות pit, ואמרנו and we will say, חיה beast רעה Some evil אכלתהו hath devoured ונראה him: and we shall see מה what יהיו will become חלמתיו׃ of his dreams.}%
\verse{וישׁמע heard ראובן And Reuben ויצלהו and he delivered מידם him out of their hands; ויאמר and said, לא Let us not נכנו kill נפשׁ׃}%
\verse{ויאמר said אלהם unto ראובן And Reuben אל תשׁפכו them, Shed דם blood, השׁליכו cast אתו אל him into הבור pit הזה this אשׁר that במדבר in the wilderness, ויד hand אל no תשׁלחו and lay בו למען upon him; that הציל he might rid אתו מידם him out of their hands, להשׁיבו to deliver אל him to אביו׃ his father}%
\verse{ויהי And it came to pass, כאשׁר when בא was come יוסף Joseph אל unto אחיו his brethren, ויפשׁיטו that they stripped את יוסף Joseph את כתנתו out of his coat, את כתנת coat הפסים of colors אשׁר that עליו׃ on}%
\verse{ויקחהו And they took וישׁלכו him, and cast אתו הברה him into a pit: והבור and the pit רק empty, אין no בו מים׃ water}%
\verse{וישׁבו And they sat down לאכל to eat לחם bread: וישׂאו and they lifted up עיניהם their eyes ויראו and looked, והנה and, behold, ארחת a company ישׁמעאלים of Ishmaelites באה came מגלעד וגמליהם with their camels נשׂאים bearing נכאת spicery וצרי and balm ולט and myrrh, הולכים going להוריד to carry down מצרימה׃ to Egypt.}%
\verse{ויאמר said יהודה And Judah אל unto אחיו his brethren, מה What בצע profit כי if נהרג we slay את אחינו our brother, וכסינו and conceal את דמו׃ his blood?}%
\verse{לכו Come, ונמכרנו and let us sell לישׁמעאלים him to the Ishmaelites, וידנו our hand אל and let not תהי be בו כי upon him; for אחינו our brother בשׂרנו our flesh. הוא he וישׁמעו were content. אחיו׃ And his brethren}%
\verse{ויעברו Then there passed by אנשׁים מדיניםחרים merchantmen; וימשׁכו and they drew ויעלו and lifted up את יוסף Joseph מן out of הבור the pit, וימכרו and sold את יוסף Joseph לישׁמעאלים to the Ishmaelites בעשׂרים for twenty כסף of silver: ויביאו and they brought את יוסף Joseph מצרימה׃ into Egypt.}%
\verse{וישׁב returned ראובן And Reuben אל unto הבור the pit; והנה and, behold, אין not יוסף Joseph בבור in the pit; ויקרע and he rent את בגדיו׃ his clothes.}%
\verse{וישׁב And he returned אל unto אחיו his brethren, ויאמר and said, הילד The child איננו not; ואני and I, אנה whither אני shall I בא׃ go?}%
\verse{ויקחו And they took את כתנת coat, יוסף Joseph's וישׁחטו and killed שׂעיר a kid עזים of the goats, ויטבלו and dipped את הכתנת the coat בדם׃ in the blood;}%
\verse{וישׁלחו And they sent את כתנת the coat הפסים of colors, ויביאו and they brought אל to אביהם their father; ויאמרו and said, זאת This מצאנו have we found: הכר know נא now הכתנת coat בנך thy son's הוא whether it אם or לא׃ no.}%
\verse{ויכירה And he knew ויאמר it, and said, כתנת coat; בני my son's חיה beast רעה an evil אכלתהו hath devoured טרף is without doubt rent in pieces. טרף is without doubt rent in pieces. יוסף׃ him; Joseph}%
\verse{ויקרע rent יעקב And Jacob שׂמלתיו his clothes, וישׂם and put שׂק sackcloth במתניו upon his loins, ויתאבל and mourned על for בנו his son ימים days. רבים׃ many}%
\verse{ויקמו rose up כל And all בניו his sons וכל and all בנתיו his daughters לנחמו to comfort וימאן him; but he refused להתנחם to be comforted; ויאמר and he said, כי For ארד I will go down אל unto בני my son אבל mourning. שׁאלה into the grave ויבך wept אתו אביו׃ Thus his father}%
\verse{והמדנים מכרו sold אתו אל him into מצרים Egypt לפוטיפר unto Potiphar, סריס an officer פרעה of Pharaoh's, שׂר captain הטבחים׃ of the guard.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 38
\verseWithHeading{Judah and Tamar}{ויהי And it came to pass בעת time, ההוא at that וירד went down יהודה that Judah מאת אחיו his brethren, ויט and turned in עד to אישׁ a certain עדלמי Adullamite, ושׁמו whose name חירה׃ Hirah.}%
\verse{וירא saw שׁם there יהודה And Judah בת a daughter אישׁ of a certain כנעני Canaanite, ושׁמו whose name שׁוע Shuah; ויקחה and he took ויבא her, and went in אליה׃ unto}%
\verse{ותהר And she conceived, ותלד and bore בן a son; ויקרא and he called את שׁמו his name ער׃ Er.}%
\verse{ותהר And she conceived עוד again, ותלד and bore בן a son; ותקרא and she called את שׁמו his name אונן׃ Onan.}%
\verse{ותסף conceived, עוד And she yet again ותלד and bore בן a son; ותקרא and called את שׁמו his name שׁלה Shelah: והיה and he was בכזיב at Chezib, בלדתה when she bore אתו׃}%
\verse{ויקח took יהודה And Judah אשׁה a wife לער for Er בכורו his firstborn, ושׁמה whose name תמר׃ Tamar.}%
\verse{ויהי was ער And Er, בכור firstborn, יהודה Judah's רע wicked בעיני in the sight יהוה of the LORD; וימתהו slew יהוה׃ and the LORD}%
\verse{ויאמר said יהודה And Judah לאונן unto Onan, בא Go in אל unto אשׁת wife, אחיך thy brother's ויבם and marry אתה והקם her, and raise up זרע seed לאחיך׃ to thy brother.}%
\verse{וידע knew אונן And Onan כי that לא should not לו יהיה be הזרע the seed והיה his; and it came to pass, אם when בא he went in אל unto אשׁת wife, אחיו his brother's ושׁחת that he spilled ארצה on the ground, לבלתי lest נתן that he should give זרע seed לאחיו׃ to his brother.}%
\verse{וירע displeased בעיני displeased יהוה the LORD: אשׁר And the thing which עשׂה he did וימת wherefore he slew גם him also. אתו׃}%
\verse{ויאמר Then said יהודה Judah לתמר to Tamar כלתו his daughter-in-law, שׁבי Remain אלמנה a widow בית house, אביך at thy father's עד till יגדל be grown: שׁלה Shelah בני my son כי for אמר he said, פן Lest peradventure ימות die גם also, הוא he כאחיו as his brethren ותלך went תמר And Tamar ותשׁב and dwelt בית house. אביה׃ in her father's}%
\verse{וירבו הימיםתמת died; בת the daughter שׁוע of Shuah אשׁת wife יהודה Judah's וינחם was comforted, יהודה and Judah ויעל and went up על unto גזזי his sheepshearers צאנו his sheepshearers הוא he וחירה Hirah רעהו and his friend העדלמי the Adullamite. תמנתה׃ to Timnath,}%
\verse{ויגד And it was told לתמר Tamar, לאמר saying, הנה Behold חמיך thy father-in-law עלה goeth up תמנתה to Timnath לגז to shear צאנו׃ his sheep.}%
\verse{ותסר And she put בגדי garments אלמנותה her widow's מעליה off from ותכס her, and covered בצעיף her with a veil, ותתעלף and wrapped herself, ותשׁב and sat בפתח place, עינים אשׁר which על by דרך the way תמנתה to Timnath; כי for ראתה she saw כי that גדל was grown, שׁלה Shelah והוא and she לא was not נתנה given לו לאשׁה׃ unto him to wife.}%
\verse{ויראה saw יהודה When Judah ויחשׁבה her, he thought לזונה her a harlot; כי because כסתה she had covered פניה׃ her face.}%
\verse{ויט And he turned אליה unto אל her by הדרך the way, ויאמר and said, הבה Go to, נא I pray thee, אבוא let me come in אליך unto כי thee; (for לא not ידע he knew כי that כלתו his daughter-in-law.) הוא she ותאמר And she said, מה What תתן wilt thou give לי כי me, that תבוא thou mayest come in אלי׃ unto}%
\verse{ויאמר And he said, אנכי I אשׁלח will send גדי a kid עזים a kid מן from הצאן the flock. ותאמר And she said, אם תתן Wilt thou give ערבון a pledge, עד till שׁלחך׃ thou send}%
\verse{ויאמר And he said, מה What הערבון pledge אשׁר that אתן shall I give לך ותאמר thee? And she said, חתמך Thy signet, ופתילך and thy bracelets, ומטך and thy staff אשׁר בידך in thine hand. ויתן And he gave לה ויבא her, and came in אליה unto ותהר׃ her, and she conceived}%
\verse{ותקם And she arose, ותלך and went away, ותסר and laid by צעיפה her veil מעליה from ותלבשׁ her, and put on בגדי the garments אלמנותה׃ of her widowhood.}%
\verse{וישׁלח sent יהודה And Judah את גדי the kid העזים the kid ביד by the hand רעהו of his friend העדלמי the Adullamite, לקחת to receive הערבון pledge מיד hand: האשׁה from the woman's ולא her not. מצאה׃ but he found}%
\verse{וישׁאל Then he asked את אנשׁי >> ont> מקמה of that place, לאמר saying, איה Where הקדשׁה the harlot, הוא that בעינים openly על by הדרך the way side? ויאמרו And they said, לא no היתה There was בזה in this קדשׁה׃ harlot}%
\verse{וישׁב And he returned אל to יהודה Judah, ויאמר and said, לא I cannot מצאתיה find וגם her; and also אנשׁי the men המקום of the place אמרו said, לא no היתה there was בזה in this קדשׁה׃ harlot}%
\verse{ויאמר said, יהודה And Judah תקח Let her take לה פן to her, lest נהיה we be לבוז shamed: הנה behold, שׁלחתי I sent הגדי kid, הזה this ואתה and thou לא hast not מצאתה׃ found}%
\verse{ויהי And it came to pass כמשׁלשׁ about three חדשׁים months ויגד after, that it was told ליהודה Judah, לאמר saying, זנתה hath played the harlot; תמר Tamar כלתך thy daughter-in-law וגם and also, הנה behold, הרה she with child לזנונים by whoredom. ויאמר said, יהודה And Judah הוציאוה Bring her forth, ותשׂרף׃ and let her be burnt.}%
\verse{הוא When she מוצאת brought forth, והיא she שׁלחה sent אל to חמיה her father-in-law, לאמר saying, לאישׁ אשׁר whose אלה these לו אנכי I הרה with child: ותאמר and she said, הכר Discern, נא I pray thee, למי whose החתמת the signet, והפתילים and bracelets, והמטה and staff. האלה׃ these,}%
\verse{ויכר acknowledged יהודה And Judah ויאמר and said, צדקה She hath been more righteous ממני than כי I; because על that כן that לא her not נתתיה I gave לשׁלה to Shelah בני my son. ולא no יסף her again עוד more. לדעתה׃ And he knew}%
\verse{ויהי And it came to pass בעת in the time לדתה of her travail, והנה that, behold, תאומים twins בבטנה׃ in her womb.}%
\verse{ויהי And it came to pass, בלדתה when she travailed, ויתן that put out יד hand: ותקח took המילדת and the midwife ותקשׁר and bound על upon ידו his hand שׁני a scarlet thread, לאמר saying, זה This יצא came out ראשׁנה׃ first.}%
\verse{ויהי And it came to pass, כמשׁיב as he drew back ידו his hand, והנה that, behold, יצא came out: אחיו his brother ותאמר and she said, מה How פרצת hast thou broken forth? עליך upon פרץ breach ויקרא was called שׁמו thee: therefore his name פרץ׃ Pharez.}%
\verse{ואחר And afterward יצא came out אחיו his brother, אשׁר that על upon ידו his hand: השׁני had the scarlet thread ויקרא was called שׁמו and his name זרח׃ Zarah.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 39
\verseWithHeading{Joseph in Potiphar’s House}{ויוסף And Joseph הורד was brought down מצרימה to Egypt; ויקנהו bought פוטיפר and Potiphar, סריס an officer פרעה of Pharaoh, שׂר captain הטבחים of the guard, אישׁ מצרייד him of the hands הישׁמעאלים of the Ishmaelites, אשׁר which הורדהו had brought him down שׁמה׃ thither.}%
\verse{ויהי was יהוה And the LORD את with יוסף Joseph, ויהי and he was אישׁ מצליח a prosperous ויהי and he was בבית in the house אדניו of his master המצרי׃ the Egyptian.}%
\verse{וירא saw אדניו And his master כי that יהוה the LORD אתו with וכל made all אשׁר that הוא he עשׂה did יהוה him, and that the LORD מצליח to prosper בידו׃ in his hand.}%
\verse{וימצא found יוסף And Joseph חן grace בעיניו in his sight, וישׁרת and he served אתו ויפקדהו him: and he made him overseer על over ביתו his house, וכל and all ישׁ he had לו נתן he put בידו׃ into his hand.}%
\verse{ויהי And it came to pass מאז from the time הפקיד אתוביתו in his house, ועל and over כל all אשׁר that ישׁ he had, לו ויברך blessed יהוה that the LORD את בית house המצרי the Egyptian's בגלל יוסףיהי was ברכת and the blessing יהוה of the LORD בכל upon all אשׁר that ישׁ he had לו בבית in the house, ובשׂדה׃ and in the field.}%
\verse{ויעזב And he left כל all אשׁר that לו ביד hand; יוסף he had in Joseph's ולא not ידע and he knew אתו he had, מאומה aught כי save אם save הלחם the bread אשׁר which הוא he אוכל did eat. ויהי was יוסף And Joseph יפה goodly תאר goodly ויפה and well מראה׃ favored.}%
\verse{ויהי And it came to pass אחר after הדברים things, האלה these ותשׂא cast אשׁת wife אדניו that his master's את עיניה her eyes אל upon יוסף Joseph; ותאמר and she said, שׁכבה Lie עמי׃ with}%
\verse{וימאן But he refused, ויאמר and said אל unto אשׁת wife, אדניו his master's הן Behold, אדני my master לא not ידע knoweth אתי with מה what בבית me in the house, וכל all אשׁר that ישׁ he hath לו נתן and he hath committed בידי׃ to my hand;}%
\verse{איננו none גדול greater בבית house הזה in this ממני than ולא I; neither חשׂך hath he kept back ממני from מאומה any thing כי me but אם me but אותך באשׁר thee, because את thou אשׁתו his wife: ואיך how אעשׂה then can I do הרעה wickedness, הגדלה great הזאת this וחטאתי and sin לאלהים׃ against God?}%
\verse{ויהי And it came to pass, כדברה as she spoke אל to יוסף Joseph יום day יום by day, ולא not שׁמע that he hearkened אליה unto לשׁכב her, to lie אצלה by להיות her, to be עמה׃ with}%
\verse{ויהי And it came to pass כהיום time, הזה about this ויבא that went הביתה into the house לעשׂות to do מלאכתו his business; ואין and none אישׁ and none מאנשׁי of the men הבית of the house שׁם there בבית׃ within.}%
\verse{ותתפשׂהו And she caught בבגדו him by his garment, לאמר saying, שׁכבה Lie עמי with ויעזב me: and he left בגדו his garment בידה in her hand, וינס and fled, ויצא and got החוצה׃ him out.}%
\verse{ויהי And it came to pass, כראותה when she saw כי that עזב he had left בגדו his garment בידה in her hand, וינס and was fled החוצה׃ forth,}%
\verse{ותקרא That she called לאנשׁי unto the men ביתה of her house, ותאמר and spoke להם לאמר unto them, saying, ראו See, הביא he hath brought in לנו אישׁ עבריצחק unto us to mock בנו בא us; he came in אלי unto לשׁכב me to lie עמי with ואקרא me, and I cried בקול voice: גדול׃ with a loud}%
\verse{ויהי And it came to pass, כשׁמעו when he heard כי that הרימתי I lifted up קולי my voice ואקרא and cried, ויעזב that he left בגדו his garment אצלי with וינס me, and fled, ויצא and got החוצה׃ him out.}%
\verse{ותנח בגדו his garment אצלה by עד her, until בוא came אדניו his lord אל came ביתו׃ home.}%
\verse{ותדבר And she spoke אליו unto כדברים words, האלה him according to these לאמר saying, בא thou hast brought אלי unto העבד servant, העברי The Hebrew אשׁר which הבאת unto us, came in לנו לצחק׃ me to mock}%
\verse{ויהי And it came to pass, כהרימי as I lifted up קולי my voice ואקרא and cried, ויעזב that he left בגדו his garment אצלי with וינס me, and fled החוצה׃ out.}%
\verse{ויהי And it came to pass, כשׁמע heard אדניו when his master את דברי the words אשׁתו of his wife, אשׁר which דברה she spoke אליו unto לאמר him, saying, כדברים manner האלה After this עשׂה did לי עבדך thy servant ויחר was kindled. אפו׃ to me; that his wrath}%
\verse{ויקח took אדני master יוסף And Joseph's אתו ויתנהו him, and put אל him into בית the prison, הסהר the prison, מקום a place אשׁר where אסורי bound: המלך the king's אסורים ויהי and he was שׁם there בבית in the prison. הסהר׃ in the prison.}%
\verse{ויהי was יהוה But the LORD את with יוסף Joseph, ויט and showed אליו and showed חסד him mercy, ויתן and gave חנו him favor בעיני in the sight שׂר of the keeper בית of the prison. הסהר׃ of the prison.}%
\verse{ויתן committed שׂר And the keeper בית of the prison הסהר of the prison ביד hand יוסף to Joseph's את כל all האסירם the prisoners אשׁר that בבית in the prison; הסהר in the prison; ואת כל and whatsoever אשׁר and whatsoever עשׂים they did שׁם there, הוא he היה was עשׂה׃ the doer}%
\verse{אין not שׂר The keeper בית of the prison הסהר of the prison ראה looked את כל to any thing מאומה to any thing בידו under his hand; באשׁר because יהוה the LORD אתו was with ואשׁר him, and which הוא he עשׂה did, יהוה the LORD מצליח׃ made to prosper.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 40
\verseWithHeading{Joseph Interprets Dreams in Prison}{ויהי And it came to pass אחר after הדברים things, האלה these חטאו had offended משׁקה the butler מלך of the king מצרים of Egypt והאפה and baker לאדניהם their lord למלך the king מצרים׃ of Egypt.}%
\verse{ויקצף was wroth פרעה And Pharaoh על against שׁני two סריסיו his officers, על against שׂר the chief המשׁקים of the butlers, ועל and against שׂר the chief האופים׃ of the bakers.}%
\verse{ויתן And he put אתם במשׁמר them in ward בית in the house שׂר of the captain הטבחים of the guard, אל into בית the prison, הסהר the prison, מקום the place אשׁר where יוסף Joseph אסור bound. שׁם׃ where}%
\verse{ויפקד charged שׂר And the captain הטבחים of the guard את יוסף Joseph אתם with וישׁרת them, and he served אתם ויהיו them: and they continued ימים a season במשׁמר׃ in ward.}%
\verse{ויחלמו And they dreamed חלום a dream שׁניהם both אישׁ חלמו his dream בלילה night, אחד in one אישׁ כפתרון according to the interpretation חלמו of his dream, המשׁקה the butler והאפה and the baker אשׁר which למלך of the king מצרים of Egypt, אשׁר אסורים bound בבית in the prison. הסהר׃ in the prison.}%
\verse{ויבא came in אליהם unto יוסף And Joseph בבקר them in the morning, וירא and looked upon אתם והנםעפים׃ they sad.}%
\verse{וישׁאל And he asked את סריסי officers פרעה Pharaoh's אשׁר that אתו with במשׁמר him in the ward בית house, אדניו of his lord's לאמר saying, מדוע Wherefore פניכם look רעים ye sadly היום׃ today?}%
\verse{ויאמרו And they said אליו unto חלום a dream, חלמנו him, We have dreamed ופתר interpreter אין and no אתו ויאמר said אלהם unto יוסף of it. And Joseph הלוא them, not לאלהים to God? פתרנים interpretations ספרו tell נא׃ me I pray you.}%
\verse{ויספר told שׂר And the chief המשׁקים butler את חלמו his dream ליוסף to Joseph, ויאמר and said לו בחלומי to him, In my dream, והנה behold, גפן a vine לפני׃ before}%
\verse{ובגפן And in the vine שׁלשׁה three שׂריגם branches: והוא her כפרחת and it as though it budded, עלתה shot forth; נצה blossoms הבשׁילו thereof brought forth אשׁכלתיה and the clusters ענבים׃ ripe grapes:}%
\verse{וכוס cup פרעה And Pharaoh's בידי in my hand: ואקח and I took את הענבים the grapes, ואשׂחט and pressed אתם אל them into כוס cup, פרעה Pharaoh's ואתן and I gave את הכוס the cup על into כף hand. פרעה׃ Pharaoh's}%
\verse{ויאמר said לו יוסף And Joseph זה unto him, This פתרנו the interpretation שׁלשׁת of it: The three השׂרגים branches שׁלשׁת three ימים days: הם׃}%
\verse{בעוד Yet שׁלשׁת within three ימים days ישׂא lift up פרעה shall Pharaoh את ראשׁך thine head, והשׁיבך and restore על thee unto כנך thy place: ונתת and thou shalt deliver כוס cup פרעה Pharaoh's בידו into his hand, כמשׁפט manner הראשׁון after the former אשׁר when היית thou wast משׁקהו׃ his butler.}%
\verse{כי אםכרתני think on אתך כאשׁר me when ייטב it shall be well לך ועשׂית with thee, and show נא I pray thee, עמדי חסד kindness, והזכרתני me, and make mention אל of me unto פרעה Pharaoh, והוצאתני and bring me out מן of הבית house: הזה׃ this}%
\verse{כי For גנב גנבתיארץ out of the land העברים of the Hebrews: וגם also פה and here לא nothing עשׂיתי have I done מאומה nothing כי that שׂמו they should put אתי בבור׃ me into the dungeon.}%
\verse{וירא saw שׂר When the chief האפים baker כי that טוב was good, פתר the interpretation ויאמר he said אל unto יוסף Joseph, אף also אני I בחלומי in my dream, והנה and, behold, שׁלשׁה three סלי baskets חרי white על on ראשׁי׃ my head:}%
\verse{ובסל basket העליון And in the uppermost מכל of all manner מאכל of bakedmeats פרעה for Pharaoh; מעשׂה אפההעוף and the birds אכל did eat אתם מן of all manner הסל the basket מעל upon ראשׁי׃ my head.}%
\verse{ויען answered יוסף And Joseph ויאמר and said, זה This פתרנו the interpretation שׁלשׁת thereof: The three הסלים baskets שׁלשׁת three ימים days: הם׃}%
\verse{בעוד Yet שׁלשׁת within three ימים days ישׂא lift up פרעה shall Pharaoh את ראשׁך thy head מעליך from off ותלה thee, and shall hang אותך על thee on עץ a tree; ואכל shall eat העוף and the birds את בשׂרך thy flesh מעליך׃ from off}%
\verse{ויהי And it came to pass ביום day, השׁלישׁי the third יום birthday, הלדת birthday, את פרעה Pharaoh's ויעשׂ that he made משׁתה a feast לכל unto all עבדיו his servants: וישׂא and he lifted up את ראשׁ the head שׂר of the chief המשׁקים butler ואת ראשׁׂר and of the chief האפים baker בתוך among עבדיו׃ his servants.}%
\verse{וישׁב And he restored את שׂר the chief המשׁקים butler על unto משׁקהו his butlership ויתן again; and he gave הכוס the cup על into כף hand: פרעה׃ Pharaoh's}%
\verse{ואת שׂר the chief האפים baker: תלה But he hanged כאשׁר as פתר had interpreted להם יוסף׃ Joseph}%
\verse{ולא Yet did not זכר remember שׂר the chief המשׁקים butler את יוסף Joseph, וישׁכחהו׃ but forgot}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 41
\verseWithHeading{Joseph Interprets Pharaoh’s Dreams}{ויהי And it came to pass מקץ at the end שׁנתים of two full years, ימים of two full years, ופרעה that Pharaoh חלם dreamed: והנה and, behold, עמד he stood על by היאר׃ the river.}%
\verse{והנה And, behold, מן out of היאר the river עלת there came up שׁבע seven פרות kine יפות well מראה favored ובריאת and fatfleshed; בשׂר and fatfleshed; ותרעינה and they fed באחו׃ in a meadow.}%
\verse{והנה And, behold, שׁבע seven פרות kine אחרות other עלות came up אחריהן after מן them out of היאר the river, רעות ill מראה favored ודקות and leanfleshed; בשׂר and leanfleshed; ותעמדנה and stood אצל by הפרות the kine על upon שׂפת the brink היאר׃ of the river.}%
\verse{ותאכלנה did eat up הפרות kine רעות And the ill המראה favored ודקת and leanfleshed הבשׂר and leanfleshed את שׁבע the seven הפרות kine. יפת well המראה favored והבריאת and fat וייקץ awoke. פרעה׃ So Pharaoh}%
\verse{ויישׁן And he slept ויחלם and dreamed שׁנית the second time: והנה and, behold, שׁבע seven שׁבלים ears of corn עלות came up בקנה stalk, אחד upon one בריאות rank וטבות׃ and good.}%
\verse{והנה And, behold, שׁבע seven שׁבלים ears דקות thin ושׁדופת and blasted קדים with the east wind צמחות sprung up אחריהן׃ after}%
\verse{ותבלענה devoured השׁבלים ears הדקות And the seven thin את שׁבע the seven השׁבלים ears. הבריאות rank והמלאות and full וייקץ awoke, פרעה And Pharaoh והנה and, behold, חלום׃ a dream.}%
\verse{ויהי And it came to pass בבקר in the morning ותפעם was troubled; רוחו that his spirit וישׁלח and he sent ויקרא and called for את כל all חרטמי the magicians מצרים of Egypt, ואת כל and all חכמיה the wise men ויספר told פרעה thereof: and Pharaoh להם את חלמו them his dream; ואין but none פותר that could interpret אותם לפרעה׃ them unto Pharaoh.}%
\verse{וידבר Then spoke שׂר the chief המשׁקים butler את פרעה unto Pharaoh, לאמר saying, את חטאי my faults אני I מזכיר do remember היום׃ this day:}%
\verse{פרעה Pharaoh קצף was wroth על with עבדיו his servants, ויתן and put אתי במשׁמר me in ward בית house, שׂר in the captain הטבחים of the guard's אתי ואתׂר me and the chief האפים׃ baker:}%
\verse{ונחלמה And we dreamed חלום a dream בלילה night, אחד in one אני I והוא and he; אישׁ each man כפתרון according to the interpretation חלמו of his dream. חלמנו׃ we dreamed}%
\verse{ושׁם And there אתנו with נער us a young man, עברי a Hebrew, עבד servant לשׂר to the captain הטבחים of the guard; ונספר and we told לו ויפתר him, and he interpreted לנו את חלמתינו to us our dreams; אישׁ כחלמו according to his dream פתר׃ he did interpret.}%
\verse{ויהי And it came to pass, כאשׁר as פתר he interpreted לנו כן to us, so היה it was; אתי השׁיב me he restored על unto כני mine office, ואתו תלה׃ and him he hanged.}%
\verse{וישׁלח sent פרעה Then Pharaoh ויקרא and called את יוסף Joseph, ויריצהו and they brought him hastily מן out of הבור the dungeon: ויגלח and he shaved ויחלף and changed שׂמלתיו his raiment, ויבא and came in אל unto פרעה׃ Pharaoh.}%
\verse{ויאמר said פרעה And Pharaoh אל unto יוסף Joseph, חלום a dream, חלמתי I have dreamed ופתר that can interpret אין and none אתו ואני it: and I שׁמעתי have heard עליך of לאמר say תשׁמע thee, thou canst understand חלום a dream לפתר to interpret אתו׃}%
\verse{ויען answered יוסף And Joseph את פרעה Pharaoh, לאמר saying, בלעדי not in אלהים me: God יענה אתׁלום of peace. פרעה׃}%
\verse{וידבר said פרעה And Pharaoh אל unto יוסף Joseph, בחלמי In my dream, הנני עמד I stood על upon שׂפת the bank היאר׃ of the river:}%
\verse{והנה And, behold, מן out of היאר the river עלת there came up שׁבע seven פרות kine, בריאות fatfleshed בשׂר fatfleshed ויפת and well תאר favored; ותרעינה and they fed באחו׃ in a meadow:}%
\verse{והנה And, behold, שׁבע seven פרות kine אחרות other עלות came up אחריהן after דלות ורעות ill תאר favored מאד and very ורקות and leanfleshed, בשׂר and leanfleshed, לא as I never ראיתי saw כהנה such בכל in all ארץ the land מצרים of Egypt לרע׃ for badness:}%
\verse{ותאכלנה did eat up הפרות kine הרקות And the lean והרעות and the ill favored את שׁבע seven הפרות kine: הראשׁנות the first הבריאת׃ fat}%
\verse{ותבאנה they had eaten אל קרבנהלא it could not נודע be known כי that באו אלרבנה ומראיהן them; but they still ill favored, רע them; but they still ill favored, כאשׁר as בתחלה at the beginning. ואיקץ׃ So I awoke.}%
\verse{וארא And I saw בחלמי in my dream, והנה and, behold, שׁבע seven שׁבלים ears עלת came up בקנה stalk, אחד in one מלאת full וטבות׃ and good:}%
\verse{והנה And, behold, שׁבע seven שׁבלים ears, צנמות withered, דקות thin, שׁדפות blasted קדים with the east wind, צמחות sprung up אחריהם׃ after}%
\verse{ותבלען devoured השׁבלים ears הדקת And the thin את שׁבע the seven השׁבלים ears: הטבות good ואמר and I told אל unto החרטמים the magicians; ואין but none מגיד׃ that could declare}%
\verse{ויאמר said יוסף And Joseph אל unto פרעה Pharaoh, חלום The dream פרעה of Pharaoh אחד one: הוא אתשׁר what האלהים God עשׂה he about to do. הגיד hath showed לפרעה׃ Pharaoh}%
\verse{שׁבע The seven פרת kine הטבת good שׁבע seven שׁנים years; הנה ושׁבע and the seven השׁבלים ears הטבת good שׁבע seven שׁנים years: הנה חלום the dream אחד one. הוא׃}%
\verse{ושׁבע And the seven הפרות kine הרקות thin והרעת and ill favored העלת that came up אחריהן after שׁבע them seven שׁנים years; הנה ושׁבע and the seven השׁבלים ears הרקות empty שׁדפות blasted הקדים with the east wind יהיו shall be שׁבע seven שׁני years רעב׃ of famine.}%
\verse{הוא This הדבר the thing אשׁר which דברתי I have spoken אל unto פרעה Pharaoh: אשׁר What האלהים God עשׂה about to do הראה he showeth את פרעה׃ unto Pharaoh.}%
\verse{הנה Behold, שׁבע seven שׁנים years באות there come שׂבע plenty גדול of great בכל throughout all ארץ the land מצרים׃ of Egypt:}%
\verse{וקמו And there shall arise שׁבע them seven שׁני years רעב of famine; אחריהן after ונשׁכח shall be forgotten כל and all השׂבע the plenty בארץ in the land מצרים of Egypt; וכלה shall consume הרעב and the famine את הארץ׃ the land;}%
\verse{ולא shall not יודע be known השׂבע And the plenty בארץ in the land מפני by reason of הרעב famine ההוא that אחרי following; כן following; כי for כבד grievous. הוא it מאד׃ very}%
\verse{ועל And for השׁנות was doubled החלום that the dream אל unto פרעה Pharaoh פעמים twice; כי because נכון established הדבר the thing מעם by האלהים God, וממהר will shortly האלהים and God לעשׂתו׃ bring it to pass.}%
\verse{ועתה Now ירא look out פרעה therefore let Pharaoh אישׁ a man נבון discreet וחכם and wise, וישׁיתהו and set על him over ארץ the land מצרים׃ of Egypt.}%
\verse{יעשׂה do פרעה Let Pharaoh ויפקד and let him appoint פקדים officers על over הארץ the land, וחמשׁ and take up the fifth part את ארץ of the land מצרים of Egypt בשׁבע in the seven שׁני years. השׂבע׃ plenteous}%
\verse{ויקבצו And let them gather את כל all אכל the food השׁנים years הטבות good הבאת that come, האלה of those ויצברו and lay up בר corn תחת under יד the hand פרעה of Pharaoh, אכל food בערים in the cities. ושׁמרו׃ and let them keep}%
\verse{והיה shall be האכל And that food לפקדון for store לארץ to the land לשׁבע against the seven שׁני years הרעב of famine, אשׁר which תהיין shall be בארץ in the land מצרים of Egypt; ולא not תכרת perish הארץ that the land ברעב׃ through the famine.}%
\verseWithHeading{Joseph Rises to Power}{וייטב was good הדבר And the thing בעיני in the eyes פרעה of Pharaoh, ובעיני and in the eyes כל of all עבדיו׃ his servants.}%
\verse{ויאמר said פרעה And Pharaoh אל unto עבדיו his servants, הנמצא Can we find כזה as this אישׁ a man אשׁר in whom רוח the Spirit אלהים׃ of God}%
\verse{ויאמר said פרעה And Pharaoh אל unto יוסף Joseph, אחרי Forasmuch הודיע hath showed אלהים as God אותך אתל thee all זאת this, אין none נבון so discreet וחכם and wise כמוך׃ as thou}%
\verse{אתה Thou תהיה shalt be על over ביתי my house, ועל and according unto פיך thy word ישׁק be ruled: כל shall all עמי my people רק only הכסא in the throne אגדל will I be greater ממך׃ than}%
\verse{ויאמר said פרעה And Pharaoh אל unto יוסף Joseph, ראה See, נתתי I have set אתך על thee over כל all ארץ the land מצרים׃ of Egypt.}%
\verse{ויסר took off פרעה And Pharaoh את טבעתו his ring מעל from ידו his hand, ויתן and put אתה על it upon יד hand, יוסף Joseph's וילבשׁ and arrayed אתו בגדי him in vestures שׁשׁ of fine linen, וישׂם and put רבד chain הזהב a gold על about צוארו׃ his neck;}%
\verse{וירכב And he made him to ride אתו במרכבת chariot המשׁנה in the second אשׁר which לו ויקראו he had; and they cried לפניו before אברך him, Bow the knee: ונתון and he made אתו על him over כל all ארץ the land מצרים׃ of Egypt.}%
\verse{ויאמר said פרעה And Pharaoh אל unto יוסף Joseph, אני I פרעה Pharaoh, ובלעדיך and without לא thee shall no ירים lift up אישׁ man את ידו his hand ואת רגלו or foot בכל in all ארץ the land מצרים׃ of Egypt.}%
\verse{ויקרא called פרעה And Pharaoh שׁם name יוסף Joseph's צפנת פענח Zaphnath-paaneah; ויתן and he gave לו את אסנת Asenath בת the daughter פוטי פרע of Poti-pherah כהן priest אן of On. לאשׁה him to wife ויצא went out יוסף And Joseph על over ארץ the land מצרים׃ of Egypt.}%
\verse{ויוסף And Joseph בן old שׁלשׁים thirty שׁנה years בעמדו when he stood לפני before פרעה Pharaoh מלך king מצרים of Egypt. ויצא went out יוסף And Joseph מלפני from the presence פרעה of Pharaoh, ויעבר and went בכל throughout all ארץ the land מצרים׃ of Egypt.}%
\verse{ותעשׂ brought forth הארץ the earth בשׁבע And in the seven שׁני years השׂבע plenteous לקמצים׃ by handfuls.}%
\verse{ויקבץ And he gathered up את כל all אכל the food שׁבע of the seven שׁנים years, אשׁר which היו were בארץ in the land מצרים of Egypt, ויתן and laid up אכל the food בערים in the cities: אכל the food שׂדה of the field, העיר every city, אשׁר which סביבתיה round about נתן laid he up בתוכה׃ in the same.}%
\verse{ויצבר gathered יוסף And Joseph בר corn כחול as the sand הים of the sea, הרבה much, מאד very עד until כי for חדל he left לספר numbering; כי אין without מספר׃ number.}%
\verse{וליוסף And unto Joseph ילד were born שׁני two בנים sons בטרם before תבוא came, שׁנת the years הרעב of famine אשׁר which ילדה bore לו אסנת Asenath בת the daughter פוטי פרע of Poti-pherah כהן priest און׃ of On}%
\verse{ויקרא called יוסף And Joseph את שׁם the name הבכור of the firstborn מנשׁה Manasseh: כי For נשׁני hath made me forget אלהים God, את כל all עמלי my toil, ואת כל and all בית house. אבי׃ my father's}%
\verse{ואת שׁם And the name השׁני of the second קרא called אפרים he Ephraim: כי For הפרני hath caused me to be fruitful אלהים God בארץ in the land עניי׃ of my affliction.}%
\verse{ותכלינה were ended. שׁבע And the seven שׁני years השׂבע of plenteousness, אשׁר that היה was בארץ in the land מצרים׃ of Egypt,}%
\verse{ותחלינה began שׁבע And the seven שׁני years הרעב of dearth לבוא to come, כאשׁר according as אמר had said: יוסף Joseph ויהי was רעב and the dearth בכל in all הארצות lands; ובכל but in all ארץ the land מצרים of Egypt היה there was לחם׃ bread.}%
\verse{ותרעב was famished, כל And when all ארץ the land מצרים of Egypt ויצעק cried העם the people אל to פרעה Pharaoh ללחם for bread: ויאמר said פרעה and Pharaoh לכל unto all מצרים לכו Go אל unto יוסף Joseph; אשׁר what יאמר he saith לכם תעשׂו׃ to you, do.}%
\verse{והרעב And the famine היה was על over כל all פני the face הארץ of the earth: ויפתח opened יוסף And Joseph את כל all אשׁר the storehouses, בהם וישׁבר and sold למצרים ויחזק waxed sore הרעב and the famine בארץ in the land מצרים׃ unto the Egyptians;}%
\verse{וכל And all הארץ countries באו came מצרימה into Egypt לשׁבר for to buy אל to יוסף Joseph כי because חזק was sore הרעב that the famine בכל in all הארץ׃ lands.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 42
\verseWithHeading{Joseph’s Brothers Go to Egypt for Food}{וירא saw יעקב Now when Jacob כי that ישׁ there was שׁבר corn במצרים in Egypt, ויאמר said יעקב Jacob לבניו unto his sons, למה Why תתראו׃ do ye look one upon another?}%
\verse{ויאמר And he said, הנה Behold, שׁמעתי I have heard כי that ישׁ there is שׁבר corn במצרים in Egypt: רדו get you down שׁמה thither, ושׁברו and buy לנו משׁם for us from thence; ונחיה that we may live, ולא and not נמות׃ die.}%
\verse{וירדו went down אחי brethren יוסף And Joseph's עשׂרה ten לשׁבר to buy בר corn ממצרים׃}%
\verse{ואת בנימין But Benjamin, אחי brother, יוסף Joseph's לא not שׁלח sent יעקב Jacob את with אחיו his brethren; כי for אמר he said, פן Lest peradventure יקראנו befall אסון׃ mischief}%
\verse{ויבאו came בני And the sons ישׂראל of Israel לשׁבר to buy בתוך among הבאים those that came: כי for היה was הרעב the famine בארץ in the land כנען׃ of Canaan.}%
\verse{ויוסף And Joseph הוא he השׁליט the governor על over הארץ the land, הוא המשׁביר that sold לכל to all עם the people הארץ of the land: ויבאו came, אחי brethren יוסף and Joseph's וישׁתחוו and bowed down themselves לו אפים before him their faces ארצה׃ to the earth.}%
\verse{וירא saw יוסף And Joseph את אחיו his brethren, ויכרם and he knew ויתנכר them, but made himself strange אליהם unto וידבר them, and spoke אתם unto קשׁות roughly ויאמר them; and he said אלהם unto מאין באתם come ויאמרו ye? And they said, מארץ כנען of Canaan לשׁבר to buy אכל׃ food.}%
\verse{ויכר knew יוסף And Joseph את אחיו his brethren, והם but they לא not הכרהו׃ knew}%
\verse{ויזכר remembered יוסף And Joseph את החלמות the dreams אשׁר which חלם he dreamed להם ויאמר of them, and said אלהם unto מרגלים spies; אתם them, Ye לראות to see את ערות the nakedness הארץ of the land באתם׃ ye are come.}%
\verse{ויאמרו And they said אליו unto לא him, Nay, אדני my lord, ועבדיך are thy servants באו come. לשׁבר but to buy אכל׃ food}%
\verse{כלנו all בני sons; אישׁ man's אחד one נחנו We כנים true אנחנו we לא no היו are עבדיך thy servants מרגלים׃ spies.}%
\verse{ויאמר And he said אלהם unto לא them, Nay, כי but ערות the nakedness הארץ of the land באתם ye are come. לראות׃ to see}%
\verse{ויאמרו And they said, שׁנים twelve עשׂר twelve עבדיך Thy servants אחים brethren, אנחנו בני the sons אישׁ man אחד of one בארץ in the land כנען of Canaan; והנה and, behold, הקטן the youngest את with אבינו our father, היום this day והאחד and one איננו׃ not.}%
\verse{ויאמר said אלהם unto יוסף And Joseph הוא them, That אשׁר that דברתי I spoke אלכם unto לאמר you, saying, מרגלים spies: אתם׃ Ye}%
\verse{בזאת Hereby תבחנו ye shall be proved: חי By the life פרעה of Pharaoh אם except תצאו ye shall not go forth מזה hence, כי except אם בבוא come אחיכם brother הקטן your youngest הנה׃ hither.}%
\verse{שׁלחו Send מכם of אחד one ויקח you, and let him fetch את אחיכם your brother, ואתם and ye האסרו shall be kept in prison, ויבחנו may be proved, דבריכם that your words האמת whether truth אתכם in ואם you: or else לא you: or else חי by the life פרעה of Pharaoh כי surely מרגלים spies. אתם׃ ye}%
\verse{ויאסף אתםל into משׁמר ward שׁלשׁת three ימים׃ days.}%
\verse{ויאמר said אלהם unto יוסף And Joseph ביום day, השׁלישׁי them the third זאת This עשׂו do, וחיו and live; את האלהים God: אני I ירא׃}%
\verse{אם If כנים true אתם ye אחיכם of your brethren אחד let one יאסר be bound בבית in the house משׁמרכם of your prison: ואתם ye, לכו go הביאו carry שׁבר corn רעבון for the famine בתיכם׃ of your houses:}%
\verse{ואת אחיכם brother הקטן your youngest תביאו But bring אלי unto ויאמנו be verified, דבריכם me; so shall your words ולא and ye shall not תמותו die. ויעשׂו And they did כן׃ so.}%
\verse{ויאמרו And they said אישׁ one אל to אחיו another, אבל verily אשׁמים guilty אנחנו We על concerning אחינו our brother, אשׁר in that ראינו we saw צרת the anguish נפשׁו of his soul, בהתחננו when he besought אלינו when he besought ולא us, and we would not שׁמענו hear; על therefore כן therefore באה come אלינו upon הצרה distress הזאת׃ is this}%
\verse{ויען answered ראובן And Reuben אתם לאמר them, saying, הלוא I not אמרתי Spoke אליכם unto לאמר you, saying, אל Do not תחטאו sin בילד against the child; ולא and ye would not שׁמעתם hear? וגם also דמו his blood הנה therefore, behold, נדרשׁ׃ is required.}%
\verse{והם And they לא not ידעו knew כי that שׁמע understood יוסף Joseph כי for המליץ them by an interpreter. בינתם׃ he spoke unto}%
\verse{ויסב And he turned himself about מעליהם from ויבך them, and wept; וישׁב and returned אלהם to וידבר them again, and communed אלהם with ויקח them, and took מאתם אתׁמעון them Simeon, ויאסר and bound אתו לעיניהם׃ him before their eyes.}%
\verse{ויצו commanded יוסף Then Joseph וימלאו to fill את כליהם their sacks בר with corn, ולהשׁיב and to restore כספיהם money אישׁ every man's אל into שׂקו his sack, ולתת and to give להם צדה them provision לדרך for the way: ויעשׂ did להם כן׃ and thus}%
\verse{וישׂאו אתׁברם with the corn, על חמריהם their asses וילכו and departed משׁם׃ thence.}%
\verse{ויפתח of them opened האחד And as one את שׂקו his sack לתת to give מספוא provender לחמרו his ass במלון in the inn, וירא he espied את כספו his money; והנה for, behold, הוא it בפי mouth. אמתחתו׃ in his sack's}%
\verse{ויאמר And he said אל unto אחיו his brethren, הושׁב is restored; כספי My money וגם even הנה and, lo, באמתחתי in my sack: ויצא failed לבם and their heart ויחרדו and they were afraid, אישׁ one אל to אחיו another, לאמר saying מה What זאת this עשׂה hath done אלהים׃ God}%
\verse{ויבאו And they came אל unto יעקב Jacob אביהם their father ארצה unto the land כנען of Canaan, ויגידו and told לו את כל him all הקרת that befell אתם לאמר׃ unto them; saying,}%
\verse{דבר spoke האישׁ אדני the lord הארץ of the land, אתנו to קשׁות roughly ויתן us, and took אתנו כמרגלים us for spies את הארץ׃ of the country.}%
\verse{ונאמר And we said אליו unto כנים true אנחנו him, We לא no היינו we are מרגלים׃ spies:}%
\verse{שׁנים twelve עשׂר twelve אנחנו We אחים brethren, בני sons אבינו of our father; האחד one איננו not, והקטן and the youngest היום this day את with אבינו our father בארץ in the land כנען׃ of Canaan.}%
\verse{ויאמר said אלינו unto האישׁ And the man, אדני the lord הארץ of the country, בזאת us, Hereby אדע shall I know כי that כנים true אתם ye אחיכם of your brethren האחד one הניחו אתי with ואת רעבון the famine בתיכם of your households, קחו me, and take ולכו׃ and be gone:}%
\verse{והביאו And bring את אחיכם brother הקטן your youngest אלי unto ואדעה me: then shall I know כי that לא no מרגלים spies, אתם ye כי but כנים true אתם ye את אחיכם you your brother, אתן will I deliver לכם ואת הארץ in the land. תסחרו׃ and ye shall traffic}%
\verse{ויהי And it came to pass הם as they מריקים emptied שׂקיהם their sacks, והנה that, behold, אישׁ every man's צרור bundle כספו of money בשׂקו in his sack: ויראו saw את צררות the bundles כספיהם of money, המה and when they ואביהם and their father וייראו׃ they were afraid.}%
\verse{ויאמר said אלהם unto יעקב And Jacob אביהם their father אתי שׁכלתם them, Me have ye bereaved יוסף Joseph איננו not, ושׁמעון and Simeon איננו not, ואת בנימן Benjamin תקחו and ye will take עלי against היו these things are כלנה׃ all}%
\verse{ויאמר spoke ראובן And Reuben אל unto אביו his father, לאמר saying, את שׁני my two בני sons, תמית Slay אם if לא him not אביאנו I bring אליך to תנה thee: deliver אתו על him into ידי my hand, ואני and I אשׁיבנו will bring אליך׃ him to}%
\verse{ויאמר And he said, לא shall not ירד go down בני My son עמכם with כי you; for אחיו his brother מת is dead, והוא and he לבדו alone: נשׁאר is left וקראהו befall אסון if mischief בדרך him by the way אשׁר in the which תלכו ye go, בה והורדתם then shall ye bring down את שׂיבתי my gray hairs ביגון with sorrow שׁאולה׃ to the grave.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 43
\verseWithHeading{Joseph’s Brothers Return to Egypt}{והרעב And the famine כבד sore בארץ׃ in the land.}%
\verse{ויהי And it came to pass, כאשׁר when כלו they had eaten up לאכל they had eaten up את השׁבר the corn אשׁר which הביאו they had brought ממצרים ויאמר said אליהם unto אביהם their father שׁבו them, Go again, שׁברו buy לנו מעט us a little אכל׃ food.}%
\verse{ויאמר spoke אליו unto יהודה And Judah לאמר him, saying, העד did solemnly protest העד did solemnly protest בנו האישׁ The man לאמר unto us, saying, לא Ye shall not תראו see פני my face, בלתי except אחיכם your brother אתכם׃ with}%
\verse{אם If ישׁך thou wilt משׁלח send את אחינו our brother אתנו with נרדה us, we will go down ונשׁברה and buy לך אכל׃ thee food:}%
\verse{ואם But if אינך thou wilt not משׁלח send לא we will not נרד go down: כי for האישׁ the man אמר said אלינו unto לא us, Ye shall not תראו see פני my face, בלתי except אחיכם your brother אתכם׃ with}%
\verse{ויאמר said, ישׂראל And Israel למה Wherefore הרעתם dealt ye ill לי להגיד with me, to tell לאישׁ the man העוד whether ye had yet לכם אח׃ a brother?}%
\verse{ויאמרו And they said, שׁאול asked us straitly שׁאל asked us straitly האישׁ The man לנו ולמולדתנו of our state, and of our kindred, לאמר saying, העוד yet אביכם your father חי alive? הישׁ have לכם אח ye brother? ונגד and we told לו על him according to פי the tenor הדברים words: האלה of these הידוע could we certainly know נדע could we certainly know כי that יאמר he would say, הורידו אתחיכם׃}%
\verse{ויאמר said יהודה And Judah אל unto ישׂראל Israel אביו his father, שׁלחה Send הנער the lad אתי with ונקומה me, and we will arise ונלכה and go; ונחיה that we may live, ולא and not נמות die, גם both אנחנו we, גם and אתה thou, גם also טפנו׃ our little ones.}%
\verse{אנכי I אערבנו will be surety מידי for him; of my hand תבקשׁנו shalt thou require אם him: if לא him not הביאתיו I bring אליך unto והצגתיו thee, and set לפניך him before וחטאתי thee, then let me bear the blame לך כל forever: הימים׃ forever:}%
\verse{כי For לולא except התמהמהנו we had lingered, כי surely עתה now שׁבנו we had returned זה this פעמים׃ second time.}%
\verse{ויאמר said אלהם unto ישׂראל Israel אביהם And their father אם them, If כן so אפוא now, זאת this; עשׂו do קחו take מזמרת of the best fruits הארץ in the land בכליכם in your vessels, והורידו and carry down לאישׁ מנחה a present, מעט a little צרי balm, ומעט and a little דבשׁ honey, נכאת spices, ולט and myrrh, בטנים nuts, ושׁקדים׃ and almonds:}%
\verse{וכסף money משׁנה double קחו And take בידכם in your hand; ואת הכסף and the money המושׁב that was brought again בפי in the mouth אמתחתיכם of your sacks, תשׁיבו carry again בידכם in your hand; אולי peradventure משׁגה an oversight: הוא׃ it}%
\verse{ואת אחיכם also your brother, קחו Take וקומו and arise, שׁובו go again אל unto האישׁ׃ the man:}%
\verse{ואל And God שׁדי Almighty יתן give לכם רחמים you mercy לפני before האישׁ the man, ושׁלח that he may send away לכם את אחיכם brother, אחר your other ואת בנימין and Benjamin. ואני I כאשׁר If שׁכלתי be bereaved שׁכלתי׃ I am bereaved.}%
\verse{ויקחו took האנשׁים And the men את המנחה present, הזאת that ומשׁנה double כסף money לקחו and they took בידם in their hand, ואת בנימן and Benjamin; ויקמו and rose up, וירדו and went down מצרים to Egypt, ויעמדו and stood לפני before יוסף׃ Joseph.}%
\verse{וירא saw יוסף And when Joseph אתם with את בנימין Benjamin ויאמר them, he said לאשׁר to the ruler of על to the ruler of ביתו his house, הבא Bring את האנשׁיםביתה home, וטבח and slay, טבח and slay, והכן and make ready; כי for אתי with יאכלו shall dine האנשׁים בצהרים׃ me at noon.}%
\verse{ויעשׂ did האישׁ And the man כאשׁר as אמר bade; יוסף Joseph ויבא brought האישׁ and the man את האנשׁים the men ביתה house. יוסף׃ into Joseph's}%
\verse{וייראו were afraid, האנשׁים And the men כי because הובאו they were brought בית house; יוסף into Joseph's ויאמרו and they said, על against דבר הכסף the money השׁב that was returned באמתחתינו in our sacks בתחלה at the first time אנחנו are we מובאים brought in; להתגלל that he may seek occasion עלינו upon ולהתנפל us, and fall עלינו ולקחת us, and take אתנו לעבדים us for bondmen, ואת חמרינו׃ and our asses.}%
\verse{ויגשׁו And they came near אל to האישׁ the steward אשׁר עלית house, יוסף of Joseph's וידברו and they communed אליו with פתח him at the door הבית׃ of the house,}%
\verse{ויאמרו And said, בי O אדני sir, ירד we came indeed down ירדנו we came indeed down בתחלה at the first time לשׁבר to buy אכל׃ food:}%
\verse{ויהי And it came to pass, כי when באנו we came אל to המלון the inn, ונפתחה that we opened את אמתחתינו our sacks, והנה and, behold, כסף money אישׁ man's בפי in the mouth אמתחתו of his sack, כספנו our money במשׁקלו in full weight: ונשׁב אתוידנו׃ in our hand.}%
\verse{וכסף money אחר And other הורדנו have we brought down בידנו in our hands לשׁבר to buy אכל food: לא we cannot ידענו tell מי who שׂם put כספנו our money באמתחתינו׃ in our sacks.}%
\verse{ויאמר And he said, שׁלום Peace לכם אל not: תיראו to you, fear אלהיכם your God, ואלהי and the God אביכם of your father, נתן hath given לכם מטמון you treasure באמתחתיכם in your sacks: כספכם your money. בא I had אלי unto ויוצא אלהםת שׁמעון׃}%
\verse{ויבא brought האישׁ And the man את האנשׁים the men ביתה house, יוסף into Joseph's ויתן and gave מים water, וירחצו and they washed רגליהם their feet; ויתן and he gave מספוא provender. לחמריהם׃ their asses}%
\verse{ויכינו And they made ready את המנחה the present עד against בוא came יוסף Joseph בצהרים at noon: כי for שׁמעו they heard כי that שׁם there. יאכלו they should eat לחם׃ bread}%
\verse{ויבא came יוסף And when Joseph הביתה home, ויביאו they brought לו את המנחה him the present אשׁר which בידם in their hand הביתה into the house, וישׁתחוו and bowed themselves לו ארצה׃ to him to the earth.}%
\verse{וישׁאל And he asked להם לשׁלום them of welfare, ויאמר and said, השׁלום well, אביכם your father הזקן the old man אשׁר of whom אמרתם ye spoke? העודנו he yet חי׃ alive?}%
\verse{ויאמרו And they answered, שׁלום in good health, לעבדך Thy servant לאבינו our father עודנו he yet חי alive. ויקדו And they bowed down their heads, וישׁתחו׃ and made obeisance.}%
\verse{וישׂא And he lifted up עיניו his eyes, וירא and saw את בנימין Benjamin, אחיו his brother בן son, אמו his mother's ויאמר and said, הזה this אחיכם brother, הקטן your younger אשׁר of whom אמרתם ye spoke אלי unto ויאמר me? And he said, אלהים God יחנך be gracious בני׃ unto thee, my son.}%
\verse{וימהר made haste; יוסף And Joseph כי for נכמרו did yearn רחמיו his bowels אל upon אחיו his brother: ויבקשׁ and he sought לבכות to weep; ויבא and he entered into החדרה chamber, ויבך and wept שׁמה׃ there.}%
\verse{וירחץ And he washed פניו his face, ויצא and went out, ויתאפק and refrained himself, ויאמר and said, שׂימו Set on לחם׃ bread.}%
\verse{וישׂימו And they set on לו לבדו for him by himself, ולהם לבדם and for them by themselves, ולמצרים and for the Egyptians, האכלים which did eat אתו with לבדם him, by themselves: כי because לא not יוכלון might המצרים the Egyptians לאכל eat את העברים the Hebrews; לחם bread כי for תועבה an abomination הוא that למצרים׃}%
\verse{וישׁבו And they sat לפניו before הבכר him, the firstborn כבכרתו according to his birthright, והצעיר and the youngest כצערתו according to his youth: ויתמהו marveled האנשׁים and the men אישׁ one אל at רעהו׃ another.}%
\verse{וישׂא And he took משׂאת messes מאת פניו before אלהם unto ותרב was five times so much as משׂאת mess בנימן him: but Benjamin's ממשׂאת כלם any חמשׁ ידותישׁתו of theirs. And they drank, וישׁכרו and were merry עמו׃ with}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 44
\verseWithHeading{Joseph Tests His Brothers}{ויצו And he commanded את אשׁר the steward על the steward ביתו of his house, לאמר saying, מלא Fill את אמתחת sacks האנשׁים the men's אכל food, כאשׁר as much as יוכלון they can שׂאת carry, ושׂים and put כסף money אישׁ every man's בפי mouth. אמתחתו׃ in his sack's}%
\verse{ואת גביעי my cup, גביע cup, הכסף the silver תשׂים And put בפי mouth אמתחת in the sack's הקטן of the youngest, ואת כסף money. שׁברו and his corn ויעשׂ And he did כדבר according to the word יוסף Joseph אשׁר that דבר׃ had spoken.}%
\verse{הבקר As soon as the morning אור והאנשׁיםׁלחו were sent away, המה they וחמריהם׃ and their asses.}%
\verse{הם when they יצאו were gone out of את העיר the city, לא not הרחיקו far off, ויוסף Joseph אמר said לאשׁר unto his steward, על ביתוום Up, רדף follow אחרי after האנשׁים the men; והשׂגתם and when thou dost overtake ואמרת them, say אלהם unto למה them, Wherefore שׁלמתם have ye rewarded רעה evil תחת for טובה׃ good?}%
\verse{הלוא not זה this אשׁר in which ישׁתה drinketh, אדני my lord בו והוא and whereby indeed he divineth? נחשׁ ינחשׁו הרעתם ye have done evil אשׁר in so עשׂיתם׃ doing.}%
\verse{וישׂגם And he overtook וידבר them, and he spoke אלהם unto את הדברים words. האלה׃ them these same}%
\verse{ויאמרו And they said אליו unto למה him, Wherefore ידבר saith אדני my lord כדברים words? האלה these חלילה God forbid לעבדיך that thy servants מעשׂות should do כדבר thing: הזה׃ according to this}%
\verse{הן Behold, כסף the money, אשׁר which מצאנו we found בפי mouths, אמתחתינו in our sacks' השׁיבנו we brought again אליך unto מארץ thee out of the land כנען of Canaan: ואיך how נגנב then should we steal מבית אדניךסף silver או or זהב׃ gold?}%
\verse{אשׁר whomsoever ימצא it be found, אתו With מעבדיך of thy servants ומת both let him die, וגם also אנחנו and we נהיה will be לאדני my lord's לעבדים׃ bondmen.}%
\verse{ויאמר And he said, גם also עתה Now כדבריכם according unto your words: כן also הוא it אשׁר whom ימצא it is found אתו he with יהיה shall be לי עבד my servant; ואתם and ye תהיו shall be נקים׃ blameless.}%
\verse{וימהרו Then they speedily ויורדו took down אישׁ every man את אמתחתו his sack ארצה to the ground, ויפתחו and opened אישׁ every man אמתחתו׃ his sack.}%
\verse{ויחפשׂ And he searched, בגדול at the eldest, החל began ובקטן at the youngest: כלה and left וימצא was found הגביע and the cup באמתחת sack. בנימן׃ in Benjamin's}%
\verse{ויקרעו Then they rent שׂמלתם their clothes, ויעמס and laded אישׁ every man על חמרו his ass, וישׁבו and returned העירה׃ to the city.}%
\verse{ויבא came יהודה And Judah ואחיו and his brethren ביתה house; יוסף to Joseph's והוא for he עודנו yet שׁם there: ויפלו and they fell לפניו before ארצה׃ him on the ground.}%
\verse{ויאמר said להם יוסף And Joseph מה unto them, What המעשׂה deed הזה this אשׁר that עשׂיתם ye have done? הלוא ye not ידעתם know כי that נחשׁ can certainly divine? ינחשׁ can certainly divine? אישׁ such a man אשׁר כמני׃ as I}%
\verse{ויאמר said, יהודה And Judah מה What נאמר shall we say לאדני unto my lord? מה what נדבר shall we speak? ומה or how נצטדק shall we clear ourselves? האלהים God מצא hath found out את עון the iniquity עבדיך of thy servants: הננו עבדים servants, לאדני we my lord's גם both אנחנו we, גם and also אשׁר with whom נמצא is found. הגביע the cup בידו׃ with whom}%
\verse{ויאמר And he said, חלילה God forbid לי מעשׂות זאת so: האישׁ the man אשׁר in whose נמצא is found, הגביע the cup בידו hand הוא he יהיה shall be לי עבד my servant; ואתם and as for you, עלו get you up לשׁלום in peace אל unto אביכם׃ your father.}%
\verse{ויגשׁ came near אליו unto him, יהודה Then Judah ויאמר and said, בי Oh אדני my lord, ידבר speak נא I pray thee, עבדך let thy servant, דבר a word באזני ears, אדני in my lord's ואל and let not יחר burn אפך thine anger בעבדך against thy servant: כי for כמוך thou even as כפרעה׃ Pharaoh.}%
\verse{אדני My lord שׁאל asked את עבדיו his servants, לאמר saying, הישׁ Have לכם אב ye a father, או or אח׃ a brother?}%
\verse{ונאמר And we said אל unto אדני my lord, ישׁ We have לנו אב a father, זקן an old man, וילד and a child זקנים of his old age, קטן a little one; ואחיו and his brother מת is dead, ויותר is left הוא and he לבדו alone לאמו of his mother, ואביו and his father אהבו׃ loveth}%
\verse{ותאמר And thou saidst אל unto עבדיך thy servants, הורדהו Bring him down אלי unto ואשׂימה me, that I may set עיני mine eyes עליו׃ upon}%
\verse{ונאמר And we said אל unto אדני my lord, לא cannot יוכל cannot הנער The lad לעזב leave את אביו his father: ועזב for he should leave את אביו his father, ומת׃ would die.}%
\verse{ותאמר And thou saidst אל unto עבדיך thy servants, אם לא no ירד come down אחיכם brother הקטן your youngest אתכם with לא תספון more. לראות you, ye shall see פני׃ my face}%
\verse{ויהי And it came to pass כי when עלינו we came up אל unto עבדך thy servant אבי my father, ונגד we told לו את דברי him the words אדני׃ of my lord.}%
\verse{ויאמר said, אבינו And our father שׁבו Go again, שׁברו buy לנו מעט us a little אכל׃ food.}%
\verse{ונאמר And we said, לא not נוכל we may לרדת go down: אם if ישׁ be אחינו brother הקטן our youngest אתנו with וירדנו us, then will we go down: כי for לא נוכלראות see פני face, האישׁ the man's ואחינו brother הקטן our youngest איננו except אתנו׃ with}%
\verse{ויאמר said עבדך And thy servant אבי my father אלינו unto אתם us, Ye ידעתם know כי that שׁנים me two ילדה bore לי אשׁתי׃ my wife}%
\verse{ויצא went out האחד And the one מאתי ואמר me, and I said, אך Surely טרף he is torn in pieces; טרף he is torn in pieces; ולא him not ראיתיו and I saw עד since: הנה׃ since:}%
\verse{ולקחתם And if ye take גם also את זה this מעם from פני me, וקרהו befall אסון and mischief והורדתם him, ye shall bring down את שׂיבתי my gray hairs ברעה with sorrow שׁאלה׃ to the grave.}%
\verse{ועתה Now כבאי therefore when I come אל to עבדך thy servant אבי my father, והנער and the lad איננו not אתנו with ונפשׁו us; seeing that his life קשׁורה is bound up בנפשׁו׃ in the lad's life;}%
\verse{והיה It shall come to pass, כראותו when he seeth כי that אין not הנער the lad ומת that he will die: והורידו shall bring down עבדיך and thy servants את שׂיבת the gray hairs עבדך of thy servant אבינו our father ביגון with sorrow שׁאלה׃ to the grave.}%
\verse{כי For עבדך thy servant ערב became surety את הנער for the lad מעם unto אבי my father, לאמר saying, אם If לא him not אביאנו I bring אליך unto וחטאתי thee, then I shall bear the blame לאבי to my father כל forever. הימים׃ forever.}%
\verse{ועתה Now ישׁב abide נא therefore, I pray thee, עבדך let thy servant תחת instead of הנער the lad עבד a bondman לאדני to my lord; והנער and let the lad יעל go up עם with אחיו׃ his brethren.}%
\verse{כי For איך how אעלה shall I go up אל to אבי my father, והנער and the lad איננו not אתי with פן me? lest peradventure אראה I see ברע the evil אשׁר that ימצא shall come on את אבי׃ my father.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 45
\verseWithHeading{Joseph Reveals His Identity}{ולא not יכל could יוסף Then Joseph להתאפק refrain himself לכל before all הנצבים them that stood עליו by ויקרא him; and he cried, הוציאו to go out כל Cause every אישׁ man מעלי from ולא no עמד me. And there stood אישׁ man אתו with בהתודע made himself known יוסף him, while Joseph אל unto אחיו׃ his brethren.}%
\verse{ויתן אתלו בבכיישׁמעו heard. מצרים וישׁמעית and the house פרעה׃ of Pharaoh}%
\verse{ויאמר said יוסף And Joseph אל unto אחיו his brethren, אני I יוסף Joseph; העוד yet אבי doth my father חי live? ולא not יכלו could אחיו And his brethren לענות answer אתו כי him; for נבהלו they were troubled מפניו׃ at his presence.}%
\verse{ויאמר said יוסף And Joseph אל unto אחיו his brethren, גשׁו Come near נא me, I pray you. אלי to ויגשׁו And they came near. ויאמר And he said, אני I יוסף Joseph אחיכם your brother, אשׁר whom מכרתם ye sold אתי מצרימה׃ into Egypt.}%
\verse{ועתה Now אל therefore be not תעצבו grieved, ואל nor יחר angry בעיניכם with yourselves, כי that מכרתם ye sold אתי הנה me hither: כי for למחיה you to preserve life. שׁלחני did send אלהים God לפניכם׃ me before}%
\verse{כי For זה these שׁנתים two years הרעב the famine בקרב in הארץ the land: ועוד and yet חמשׁ five שׁנים years, אשׁר in the which אין neither חרישׁ earing וקציר׃ nor harvest.}%
\verse{וישׁלחני sent אלהים And God לפניכם me before לשׂום you to preserve לכם שׁארית you a posterity בארץ in the earth, ולהחיות and to save your lives לכם לפליטה deliverance. גדלה׃ by a great}%
\verse{ועתה So now לא not אתם you שׁלחתם sent אתי הנה me hither, כי but האלהים God: וישׂימני and he hath made לאב me a father לפרעה to Pharaoh, ולאדון and lord לכל of all ביתו his house, ומשׁל and a ruler בכל throughout all ארץ the land מצרים׃ of Egypt.}%
\verse{מהרו Haste ועלו ye, and go up אל to אבי my father, ואמרתם and say אליו unto כה him, Thus אמר saith בנך thy son יוסף Joseph, שׂמני hath made אלהים God לאדון me lord לכל of all מצרים Egypt: רדה come down אלי unto אל not: תעמד׃ me, tarry}%
\verse{וישׁבת And thou shalt dwell בארץ in the land גשׁן of Goshen, והיית and thou shalt be קרוב near אלי unto אתה me, thou, ובניך and thy children, ובני and thy children's בניך children, וצאנך and thy flocks, ובקרך and thy herds, וכל and all אשׁר׃ that}%
\verse{וכלכלתי will I nourish אתך שׁם And there כי thee; for עוד yet חמשׁ five שׁנים years רעב of famine; פן lest תורשׁ thou hast, come to poverty. אתה thou, וביתך and thy household, וכל and all אשׁר׃ that}%
\verse{והנה And, behold, עיניכם your eyes ראות see, ועיני and the eyes אחי of my brother בנימין Benjamin, כי that פי my mouth המדבר that speaketh אליכם׃ unto}%
\verse{והגדתם And ye shall tell לאבי my father את כל of all כבודי my glory במצרים in Egypt, ואת כל and of all אשׁר that ראיתם ye have seen; ומהרתם and ye shall haste והורדתם and bring down את אבי my father הנה׃ hither.}%
\verse{ויפל And he fell על upon צוארי neck, בנימן Benjamin's אחיו his brother ויבך and wept; ובנימן and Benjamin בכה wept על upon צואריו׃ his neck.}%
\verse{וינשׁק Moreover he kissed לכל all אחיו his brethren, ויבך and wept עלהם upon ואחרי them: and after כן that דברו talked אחיו his brethren אתו׃ with}%
\verse{והקל And the fame נשׁמע thereof was heard בית house, פרעה in Pharaoh's לאמר saying, באו are come: אחי brethren יוסף Joseph's וייטב בעינירעה ובעיניבדיו׃ and his servants.}%
\verse{ויאמר said פרעה And Pharaoh אל unto יוסף Joseph, אמר Say אל unto אחיך thy brethren, זאת This עשׂו do טענו ye; lade את בעירכם your beasts, ולכו and go, באו get ארצה you unto the land כנען׃ of Canaan;}%
\verse{וקחו And take את אביכם your father ואת בתיכם and your households, ובאו and come אלי unto ואתנה me: and I will give לכם את טוב you the good ארץ of the land מצרים of Egypt, ואכלו and ye shall eat את חלב the fat הארץ׃ of the land.}%
\verse{ואתה Now thou צויתה art commanded, זאת this עשׂו do קחו ye; take לכם מארץ out of the land מצרים of Egypt עגלות you wagons לטפכם for your little ones, ולנשׁיכם and for your wives, ונשׂאתם and bring את אביכם your father, ובאתם׃ and come.}%
\verse{ועינכם אל not תחס עלליכם your stuff; כי for טוב the good כל of all ארץ the land מצרים of Egypt לכם הוא׃}%
\verse{ויעשׂו did כן so: בני And the children ישׂראל of Israel ויתן gave להם יוסף and Joseph עגלות them wagons, על according to פי the commandment פרעה of Pharaoh, ויתן and gave להם צדה them provision לדרך׃ for the way.}%
\verse{לכלם To all נתן of them he gave לאישׁ חלפות changes שׂמלת of raiment; ולבנימן but to Benjamin נתן he gave שׁלשׁ three מאות hundred כסף of silver, וחמשׁ and five חלפת changes שׂמלת׃ of raiment.}%
\verse{ולאביו And to his father שׁלח he sent כזאת after this עשׂרה ten חמרים asses נשׂאים laden מטוב with the good things מצרים of Egypt, ועשׂר and ten אתנת she asses נשׂאת laden בר with corn ולחם and bread ומזון and meat לאביו for his father לדרך׃ by the way.}%
\verse{וישׁלח So he sent את אחיו his brethren וילכו away, and they departed: ויאמר and he said אלהם unto אל תרגזודרך׃ by the way.}%
\verse{ויעלו And they went up ממצרים ויבאו and came ארץ into the land כנען of Canaan אל unto יעקב Jacob אביהם׃ their father,}%
\verse{ויגדו And told לו לאמר him, saying, עוד yet יוסף Joseph חי alive, וכי for הוא and he משׁל governor בכל over all ארץ the land מצרים of Egypt. ויפג fainted, לבו And Jacob's heart כי לא them not. האמין׃ he believed}%
\verse{וידברו he had said אליו unto את כל him all דברי the words יוסף of Joseph, אשׁר which דבר אלהםירא them: and when he saw את העגלות the wagons אשׁר which שׁלח had sent יוסף Joseph לשׂאת to carry אתו ותחי revived: רוח him, the spirit יעקב of Jacob אביהם׃ their father}%
\verse{ויאמר said, ישׂראל And Israel רב enough; עוד yet יוסף Joseph בני my son חי alive: אלכה I will go ואראנו and see בטרם him before אמות׃ I die.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 46
\verseWithHeading{Jacob and His Offspring Go to Egypt}{ויסע took his journey ישׂראל And Israel וכל with all אשׁר that לו ויבא he had, and came בארה שׁבע to Beer-sheba, ויזבח and offered זבחים sacrifices לאלהי unto the God אביו of his father יצחק׃ Isaac.}%
\verse{ויאמר spoke אלהים And God לישׂראל unto Israel במראת in the visions הלילה of the night, ויאמר and said, יעקב Jacob, יעקב Jacob. ויאמר And he said, הנני׃ Here}%
\verse{ויאמר And he said, אנכי I האל God, אלהי the God אביך of thy father: אל not תירא fear מרדה to go down מצרימה into Egypt; כי for לגוי nation: גדול of thee a great אשׂימך make שׁם׃ I will there}%
\verse{אנכי I ארד will go down עמך with מצרימה thee into Egypt; ואנכי and I אעלך surely bring thee up גם will also עלה surely bring thee up ויוסף and Joseph ישׁית shall put ידו his hand על upon עיניך׃ thine eyes.}%
\verse{ויקם rose up יעקב And Jacob מבאר שׁבע וישׂאו carried בני and the sons ישׂראל of Israel את יעקב Jacob אביהם their father, ואת טפם and their little ones, ואת נשׁיהם and their wives, בעגלות in the wagons אשׁר which שׁלח had sent פרעה Pharaoh לשׂאת to carry אתו׃}%
\verse{ויקחו And they took את מקניהם their cattle, ואת רכושׁם and their goods, אשׁר which רכשׁו they had gotten בארץ in the land כנען of Canaan, ויבאו and came מצרימה into Egypt, יעקב Jacob, וכל and all זרעו his seed אתו׃ with}%
\verse{בניו His sons, ובני and his sons' בניו sons אתו with בנתיו him, his daughters, ובנות daughters, בניו and his sons' וכל and all זרעו his seed הביא brought אתו he with מצרימה׃ him into Egypt.}%
\verse{ואלה And these שׁמות the names בני of the children ישׂראל of Israel, הבאים which came מצרימה into Egypt, יעקב Jacob ובניו and his sons: בכר firstborn. יעקב Jacob's ראובן׃ Reuben,}%
\verse{ובני And the sons ראובן of Reuben; חנוך Hanoch, ופלוא and Phallu, וחצרן and Hezron, וכרמי׃ and Carmi.}%
\verse{ובני And the sons שׁמעון of Simeon; ימואל Jemuel, וימין and Jamin, ואהד and Ohad, ויכין and Jachin, וצחר and Zohar, ושׁאול and Shaul בן the son הכנענית׃ of a Canaanitish woman.}%
\verse{ובני And the sons לוי of Levi; גרשׁון Gershon, קהת Kohath, ומררי׃ and Merari.}%
\verse{ובני And the sons יהודה of Judah; ער Er, ואונן and Onan, ושׁלה and Shelah, ופרץ and Pharez, וזרח and Zerah: וימת died ער but Er ואונן and Onan בארץ in the land כנען of Canaan. ויהיו were בני And the sons פרץ of Pharez חצרן Hezron וחמול׃ and Hamul.}%
\verse{ובני And the sons ישׂשכר of Issachar; תולע Tola, ופוה and Phuvah, ויוב and Job, ושׁמרן׃ and Shimron.}%
\verse{ובני And the sons זבלון of Zebulun; סרד Sered, ואלון and Elon, ויחלאל׃ and Jahleel.}%
\verse{אלה These בני the sons לאה of Leah, אשׁר which ילדה she bore ליעקב unto Jacob בפדן ארם in Padan-aram, ואת with דינה Dinah: בתו his daughter כל all נפשׁ the souls בניו of his sons ובנותיו and his daughters שׁלשׁים thirty ושׁלשׁ׃ and three.}%
\verse{ובני And the sons גד of Gad; צפיון Ziphion, וחגי and Haggi, שׁוני Shuni, ואצבן and Ezbon, ערי Eri, וארודי and Arodi, ואראלי׃ and Areli.}%
\verse{ובני And the sons אשׁר of Asher; ימנה Jimnah, וישׁוה and Ishuah, וישׁוי and Isui, ובריעה and Beriah, ושׂרח and Serah אחתם their sister: ובני and the sons בריעה of Beriah; חבר Heber, ומלכיאל׃ and Malchiel.}%
\verse{אלה These בני the sons זלפה of Zilpah, אשׁר whom נתן gave לבן Laban ללאה to Leah בתו his daughter, ותלד she bore את אלה and these ליעקב unto Jacob, שׁשׁ sixteen עשׂרה sixteen נפשׁ׃ souls.}%
\verse{בני The sons רחל of Rachel אשׁת wife; יעקב Jacob's יוסף Joseph, ובנימן׃ and Benjamin.}%
\verse{ויולד were born ליוסף And unto Joseph בארץ in the land מצרים of Egypt אשׁר which ילדה bore לו אסנת Asenath בת the daughter פוטי פרע of Poti-pherah כהן priest אן of On את מנשׁה Manasseh ואת אפרים׃ and Ephraim,}%
\verse{ובני And the sons בנימן of Benjamin בלע Belah, ובכר and Becher, ואשׁבל and Ashbel, גרא Gera, ונעמן and Naaman, אחי Ehi, וראשׁ and Rosh, מפים Muppim, וחפים and Huppim, וארד׃ and Ard.}%
\verse{אלה These בני the sons רחל of Rachel, אשׁר which ילד were born ליעקב to Jacob: כל all נפשׁ the souls ארבעה fourteen. עשׂר׃ fourteen.}%
\verse{ובני And the sons דן of Dan; חשׁים׃ Hushim.}%
\verse{ובני And the sons נפתלי of Naphtali; יחצאל Jahzeel, וגוני and Guni, ויצר and Jezer, ושׁלם׃ and Shillem.}%
\verse{אלה These בני the sons בלהה of Bilhah, אשׁר which נתן gave לבן Laban לרחל unto Rachel בתו his daughter, ותלד and she bore את אלה these ליעקב unto Jacob: כל all נפשׁ the souls שׁבעה׃ seven.}%
\verse{כל All הנפשׁ the souls הבאה that came ליעקב with Jacob מצרימה into Egypt, יצאי which came out ירכו of his loins, מלבד besides נשׁי wives, בני sons' יעקב Jacob's כל all נפשׁ the souls שׁשׁים threescore ושׁשׁ׃ and six;}%
\verse{ובני And the sons יוסף of Joseph, אשׁר which ילד were born לו במצרים him in Egypt, נפשׁ souls: שׁנים two כל all הנפשׁ the souls לבית of the house יעקב of Jacob, הבאה which came מצרימה into Egypt, שׁבעים׃ threescore and ten.}%
\verse{ואת יהודה Judah שׁלח And he sent לפניו before אל him unto יוסף Joseph, להורת to direct לפניו his face גשׁנה unto Goshen; ויבאו and they came ארצה into the land גשׁן׃ of Goshen.}%
\verse{ויאסר made ready יוסף And Joseph מרכבתו his chariot, ויעל and went up לקראת to meet ישׂראל Israel אביו his father, גשׁנה to Goshen, וירא and presented himself אליו unto ויפל him; and he fell על on צואריו his neck, ויבך and wept על on צואריו his neck עוד׃ a good while.}%
\verse{ויאמר said ישׂראל And Israel אל unto יוסף Joseph, אמותה let me die, הפעם Now אחרי since ראותי I have seen את פניך thy face, כי because עודך thou yet חי׃ alive.}%
\verse{ויאמר said יוסף And Joseph אל unto אחיו his brethren, ואל and unto בית house, אביו his father's אעלה I will go up, ואגידה and show לפרעה Pharaoh, ואמרה and say אליו unto אחי him, My brethren, ובית house, אבי and my father's אשׁר which בארץ in the land כנען of Canaan, באו are come אלי׃ unto}%
\verse{והאנשׁים רעי shepherds, צאן shepherds, כי for אנשׁי מקנה to feed cattle; היו hath been וצאנם their flocks, ובקרם and their herds, וכל and all אשׁר that להם הביאו׃ and they have brought}%
\verse{והיה And it shall come to pass, כי when יקרא shall call לכם פרעה Pharaoh ואמר you, and shall say, מה What מעשׂיכם׃ your occupation?}%
\verse{ואמרתם That ye shall say, אנשׁי מקנה about cattle היו hath been עבדיך Thy servants' מנעורינו from our youth ועד even until עתה now, גם both אנחנו we, גם also אבתינו our fathers: בעבור that תשׁבו ye may dwell בארץ in the land גשׁן of Goshen; כי for תועבת an abomination מצרים כל every רעה shepherd צאן׃ shepherd}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 47
\verseWithHeading{Jacob Settles in Goshen}{ויבא came יוסף Then Joseph ויגד and told לפרעה Pharaoh, ויאמר and said, אבי My father ואחי and my brethren, וצאנם and their flocks, ובקרם and their herds, וכל and all אשׁר that להם באו they have, are come מארץ out of the land כנען of Canaan; והנם בארץ they in the land גשׁן׃ of Goshen.}%
\verse{ומקצה some אחיו of his brethren, לקח And he took חמשׁה five אנשׁים ויצגם and presented לפני them unto פרעה׃ Pharaoh.}%
\verse{ויאמר said פרעה And Pharaoh אל unto אחיו his brethren, מה What מעשׂיכם your occupation? ויאמרו And they said אל unto פרעה Pharaoh, רעה shepherds, צאן shepherds, עבדיך Thy servants גם both אנחנו we, גם also אבותינו׃ our fathers.}%
\verse{ויאמרו They said אל moreover unto פרעה Pharaoh, לגור For to sojourn בארץ in the land באנו are we come; כי for אין have no מרעה pasture לצאן for their flocks; אשׁר for לעבדיך thy servants כי for כבד sore הרעב the famine בארץ in the land כנען of Canaan: ועתה now ישׁבו dwell נא therefore, we pray thee, עבדיך let thy servants בארץ in the land גשׁן׃ of Goshen.}%
\verse{ויאמר spoke פרעה And Pharaoh אל unto יוסף Joseph, לאמר saying, אביך Thy father ואחיך and thy brethren באו are come אליך׃ unto}%
\verse{ארץ The land מצרים of Egypt לפניך before הוא במיטב thee; in the best הארץ of the land הושׁב to dwell; את אביך make thy father ואת אחיך and brethren ישׁבו let them dwell: בארץ in the land גשׁן of Goshen ואם and if ידעת thou knowest וישׁ בם אנשׁי men חיל of activity ושׂמתם among them, then make שׂרי them rulers מקנה cattle. על over my אשׁר׃ over my}%
\verse{ויבא brought in יוסף And Joseph את יעקב Jacob אביו his father, ויעמדהו and set לפני him before פרעה Pharaoh: ויברך blessed יעקב and Jacob את פרעה׃ Pharaoh.}%
\verse{ויאמר said פרעה And Pharaoh אל unto יעקב Jacob, כמה How ימי old שׁני חייך׃}%
\verse{ויאמר said יעקב And Jacob אל unto פרעה Pharaoh, ימי The days שׁני of the years מגורי of my pilgrimage שׁלשׁים and thirty ומאת a hundred שׁנה years: מעט few ורעים and evil היו been, ימי have the days שׁני of the years חיי of my life ולא and have not השׂיגו attained unto את ימי the days שׁני of the years חיי of the life אבתי of my fathers בימי in the days מגוריהם׃ of their pilgrimage.}%
\verse{ויברך blessed יעקב And Jacob את פרעה Pharaoh, ויצא and went out מלפני from before פרעה׃ Pharaoh.}%
\verse{ויושׁב placed יוסף And Joseph את אביו his father ואת אחיו and his brethren, ויתן להם אחזה them a possession בארץ in the land מצרים of Egypt, במיטב in the best הארץ of the land, בארץ in the land רעמסס of Rameses, כאשׁר as צוה had commanded. פרעה׃ Pharaoh}%
\verse{ויכלכל nourished יוסף And Joseph את אביו his father, ואת אחיו and his brethren, ואת כל and all בית household, אביו his father's לחם with bread, לפי according to הטף׃ families.}%
\verseWithHeading{The Famine in Egypt Continues}{ולחם bread אין And no בכל in all הארץ the land; כי for כבד sore, הרעב the famine מאד very ותלה fainted ארץ so that the land מצרים of Egypt וארץ and the land כנען of Canaan מפני by reason of הרעב׃ the famine.}%
\verse{וילקט gathered up יוסף And Joseph את כל all הכסף the money הנמצא that was found בארץ in the land מצרים of Egypt, ובארץ and in the land כנען of Canaan, בשׁבר for the corn אשׁר which הם they שׁברים bought: ויבא brought יוסף and Joseph את הכסף the money ביתה house. פרעה׃ into Pharaoh's}%
\verse{ויתם failed הכסף And when money מארץ in the land מצרים of Egypt, ומארץ and in the land כנען of Canaan, ויבאו came כל all מצרים אל unto יוסף Joseph, לאמר and said, הבה Give לנו לחם us bread: ולמה for why נמות should we die נגדך in thy presence? כי for אפס faileth. כסף׃ the money}%
\verse{ויאמר said, יוסף And Joseph הבו Give מקניכם your cattle; ואתנה and I will give לכם במקניכם you for your cattle, אם if אפס fail. כסף׃ money}%
\verse{ויביאו And they brought את מקניהם their cattle אל unto יוסף Joseph: ויתן gave להם יוסף and Joseph לחם them bread בסוסים for horses, ובמקנה and for the flocks, הצאן and for the flocks, ובמקנה and for the cattle הבקר of the herds, ובחמרים and for the asses: וינהלם and he fed בלחם them with bread בכל for all מקנהם their cattle בשׁנה year. ההוא׃ for that}%
\verse{ותתם was ended, השׁנה year ההוא When that ויבאו they came אליו unto בשׁנה year, השׁנית him the second ויאמרו and said לו לא unto him, We will not נכחד hide מאדני my lord, כי how that אם how that תם is spent; הכסף our money ומקנה our herds הבהמה of cattle; אל also hath אדני my lord לא there is not נשׁאר aught left לפני in the sight אדני of my lord, בלתי but אם but גויתנו our bodies, ואדמתנו׃ and our lands:}%
\verse{למה Wherefore נמות shall we die לעיניך before thine eyes, גם both אנחנו we גם and אדמתנו our land? קנה buy אתנו ואתדמתנו us and our land בלחם for bread, ונהיה will be אנחנו and we ואדמתנו and our land עבדים servants לפרעה unto Pharaoh: ותן and give זרע seed, ונחיה that we may live, ולא and not נמות die, והאדמה that the land לא be not desolate. תשׁם׃ be not desolate.}%
\verse{ויקן bought יוסף And Joseph את כל all אדמת the land מצרים of Egypt לפרעה for Pharaoh; כי for מכרו sold מצרים אישׁׂדהו his field, כי because חזק prevailed עלהם over הרעב the famine ותהי became הארץ them: so the land לפרעה׃ Pharaoh's.}%
\verse{ואת העם And as for the people, העביר he removed אתו לערים them to cities מקצה from end גבול of the borders מצרים of Egypt ועד even to קצהו׃ the end}%
\verse{רק Only אדמת the land הכהנים of the priests לא he not; קנה bought כי for חק had a portion לכהנים the priests מאת פרעה Pharaoh, ואכלו and did eat את חקם their portion אשׁר which נתן gave להם פרעה Pharaoh על them: wherefore כן them: wherefore לא not מכרו they sold את אדמתם׃ their lands.}%
\verse{ויאמר said יוסף Then Joseph אל unto העם the people, הן Behold, קניתי I have bought אתכם היום you this day ואת אדמתכם and your land לפרעה for Pharaoh: הא lo, לכם זרע seed וזרעתם for you, and ye shall sow את האדמה׃ the land.}%
\verse{והיה And it shall come to pass בתבואת in the increase, ונתתם that ye shall give חמישׁית the fifth לפרעה unto Pharaoh, וארבע and four הידת parts יהיה shall be לכם לזרע your own, for seed השׂדה of the field, ולאכלכם and for your food, ולאשׁר and for them בבתיכם of your households, ולאכל and for food לטפכם׃ for your little ones.}%
\verse{ויאמרו And they said, החיתנו Thou hast saved our lives: נמצא let us find חן grace בעיני in the sight אדני of my lord, והיינו and we will be עבדים servants. לפרעה׃ Pharaoh's}%
\verse{וישׂם made אתה יוסף And Joseph לחק it a law עד unto היום day, הזה this על over אדמת the land מצרים of Egypt לפרעה Pharaoh לחמשׁ should have the fifth רק except אדמת the land הכהנים of the priests לבדם only, לא not היתה became לפרעה׃ Pharaoh's.}%
\verse{וישׁב dwelt ישׂראל And Israel בארץ in the land מצרים of Egypt, בארץ in the country גשׁן of Goshen; ויאחזו and they had possessions בה ויפרו therein, and grew, וירבו and multiplied מאד׃ exceedingly.}%
\verse{ויחי lived יעקב And Jacob בארץ in the land מצרים of Egypt שׁבע seventeen עשׂרה seventeen שׁנה years: ויהי was ימי so the whole age יעקב of Jacob שׁני years. חייו שׁבע and seven שׁנים וארבעים forty ומאת a hundred שׁנה׃}%
\verse{ויקרבו drew nigh ימי And the time ישׂראל that Israel למות must die: ויקרא and he called לבנו his son ליוסף Joseph, ויאמר and said לו אם unto him, If נא now מצאתי I have found חן grace בעיניך in thy sight, שׂים put, נא I pray thee, ידך thy hand תחת under ירכי my thigh, ועשׂית and deal עמדי חסד kindly ואמת and truly אל me not, נא I pray thee, תקברני me; bury במצרים׃ in Egypt:}%
\verse{ושׁכבתי But I will lie עם with אבתי my fathers, ונשׂאתני and thou shalt carry ממצרים וקברתני and bury בקברתם me in their burial place. ויאמר And he said, אנכי I אעשׂה will do כדברך׃ as thou hast said.}%
\verse{ויאמר And he said, השׁבעה Swear לי וישׁבע unto me. And he swore לו וישׁתחו bowed himself ישׂראל unto him. And Israel על upon ראשׁ head. המטה׃ the bed's}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 48
\verseWithHeading{Jacob Blesses Ephraim and Manasseh}{ויהי And it came to pass אחרי after הדברים things, האלה these ויאמר that told ליוסף Joseph, הנה Behold, אביך thy father חלה sick: ויקח and he took את שׁני him his two בניו sons, עמו with את מנשׁה Manasseh ואת אפרים׃ and Ephraim.}%
\verse{ויגד And told ליעקב Jacob, ויאמר and said, הנה Behold, בנך thy son יוסף Joseph בא cometh אליך unto ויתחזק strengthened himself, ישׂראל thee: and Israel וישׁב and sat על upon המטה׃ the bed.}%
\verse{ויאמר said יעקב And Jacob אל unto יוסף Joseph, אל God שׁדי Almighty נראה appeared אלי unto בלוז me at Luz בארץ in the land כנען of Canaan, ויברך and blessed אתי׃}%
\verse{ויאמר And said אלי unto הנני מפרך I will make thee fruitful, והרביתך and multiply ונתתיך thee, and I will make לקהל of thee a multitude עמים of people; ונתתי and will give את הארץ land הזאת this לזרעך to thy seed אחריך after אחזת possession. עולם׃ thee an everlasting}%
\verse{ועתה And now שׁני thy two בניך sons, הנולדים which were born לך בארץ unto thee in the land מצרים of Egypt עד before באי I came אליך unto מצרימה thee into Egypt, לי הם אפרים Ephraim ומנשׁה and Manasseh, כראובן mine; as Reuben ושׁמעון and Simeon, יהיו׃ they shall be}%
\verse{ומולדתך And thy issue, אשׁר which הולדת thou begettest אחריהם after לך יהיו them, shall be על after שׁם the name אחיהם of their brethren יקראו thine, shall be called בנחלתם׃ in their inheritance.}%
\verse{ואני And as for me, בבאי when I came מפדן מתה died עלי by רחל Rachel בארץ me in the land כנען of Canaan בדרך in the way, בעוד when yet כברת but a little ארץ way לבא to come אפרתה unto Ephrath: ואקברה and I buried שׁם her there בדרך in the way אפרת of Ephrath; הוא the same בית לחם׃ Bethlehem.}%
\verse{וירא beheld ישׂראל And Israel את בני sons, יוסף Joseph's ויאמר and said, מי Who אלה׃ these?}%
\verse{ויאמר said יוסף And Joseph אל unto אביו his father, בני my sons, הם They אשׁר whom נתן hath given לי אלהים God בזה me in this ויאמר And he said, קחם Bring נא them, I pray thee, אלי unto ואברכם׃ me, and I will bless}%
\verse{ועיני Now the eyes ישׂראל of Israel כבדו were dim מזקן for age, לא not יוכל he could לראות see. ויגשׁ אתםליו unto וישׁק him; and he kissed להם ויחבק׃ them, and embraced}%
\verse{ויאמר said ישׂראל And Israel אל unto יוסף Joseph, ראה to see פניך thy face: לא I had not פללתי thought והנה and, lo, הראה hath showed אתי אלהים God גם me also את זרעך׃ thy seed.}%
\verse{ויוצא יוסף And Joseph אתם מעם from between ברכיו his knees, וישׁתחו and he bowed himself לאפיו with his face ארצה׃ to the earth.}%
\verse{ויקח took יוסף And Joseph את שׁניהם them both, את אפרים Ephraim בימינו in his right hand משׂמאל left hand, ישׂראל toward Israel's ואת מנשׁה and Manasseh בשׂמאלו in his left hand מימין right hand, ישׂראל toward Israel's ויגשׁ and brought near אליו׃ unto}%
\verse{וישׁלח stretched out ישׂראל And Israel את ימינו his right hand, וישׁת and laid על upon ראשׁ head, אפרים Ephraim's והוא who הצעיר the younger, ואת שׂמאלו and his left hand על upon ראשׁ head, מנשׁה Manasseh's שׂכל אתדיו כי for מנשׁה Manasseh הבכור׃ the firstborn.}%
\verse{ויברך And he blessed את יוסף Joseph, ויאמר and said, האלהים God, אשׁר whom התהלכו did walk, אבתי my fathers לפניו before אברהם Abraham ויצחק and Isaac האלהים the God הרעה which fed אתי מעודי me all my life long עד unto היום day, הזה׃ this}%
\verse{המלאך The Angel הגאל which redeemed אתי מכל me from all רע evil, יברך bless את הנערים the lads; ויקרא be named בהם שׁמי and let my name ושׁם on them, and the name אבתי of my fathers אברהם Abraham ויצחק and Isaac; וידגו and let them grow לרב into a multitude בקרב in the midst הארץ׃ of the earth.}%
\verse{וירא saw יוסף And when Joseph כי that ישׁית laid אביו his father יד hand ימינו his right על upon ראשׁ the head אפרים of Ephraim, וירע it displeased בעיניו it displeased ויתמך him: and he held up יד hand, אביו his father's להסיר to remove אתה מעל it from ראשׁ head אפרים Ephraim's על unto ראשׁ head. מנשׁה׃ Manasseh's}%
\verse{ויאמר said יוסף And Joseph אל unto אביו his father, לא Not כן so, אבי my father: כי for זה this הבכר the firstborn; שׂים put ימינך thy right hand על upon ראשׁו׃ his head.}%
\verse{וימאן refused, אביו And his father ויאמר and said, ידעתי I know בני my son, ידעתי I know גם also הוא he יהיה shall become לעם a people, וגם also הוא and he יגדל shall be great: ואולם but truly אחיו brother הקטן his younger יגדל shall be greater ממנו than וזרעו he, and his seed יהיה shall become מלא a multitude הגוים׃ of nations.}%
\verse{ויברכם And he blessed ביום day, ההוא them that לאמור saying, בך יברך bless, ישׂראל In thee shall Israel לאמר saying, ישׂמך make אלהים God כאפרים thee as Ephraim וכמנשׁה and as Manasseh: וישׂם and he set את אפרים Ephraim לפני before מנשׁה׃ Manasseh.}%
\verse{ויאמר said ישׂראל And Israel אל unto יוסף Joseph, הנה Behold, אנכי I מת die: והיה shall be אלהים but God עמכם with והשׁיב אתכםל unto ארץ the land אבתיכם׃ of your fathers.}%
\verse{ואני Moreover I נתתי have given לך שׁכם portion אחד to thee one על above אחיך thy brethren, אשׁר which לקחתי I took מיד out of the hand האמרי of the Amorite בחרבי with my sword ובקשׁתי׃ and with my bow.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 49
\verseWithHeading{Jacob Blesses His Twelve Sons}{ויקרא called יעקב And Jacob אל unto בניו his sons, ויאמר and said, האספו Gather yourselves together, ואגידה that I may tell לכם את אשׁר you which יקרא shall befall אתכם באחרית you in the last הימים׃ days.}%
\verse{הקבצו Gather yourselves together, ושׁמעו and hear, בני ye sons יעקב of Jacob; ושׁמעו and hearken אל unto ישׂראל Israel אביכם׃ your father.}%
\verse{ראובן Reuben, בכרי my firstborn, אתה thou כחי my might, וראשׁית and the beginning אוני of my strength, יתר the excellency שׂאת of dignity, ויתר and the excellency עז׃ of power:}%
\verse{פחז Unstable כמים as water, אל thou shalt not תותר excel; כי because עלית thou wentest up משׁכבי bed; אביך to thy father's אז then חללת defiledst יצועי to my couch. עלה׃ thou he went up}%
\verse{שׁמעון Simeon ולוי and Levi אחים brethren; כלי instruments חמס of cruelty מכרתיהם׃ their habitations.}%
\verse{בסדם thou into their secret; אל not תבא come נפשׁי O my soul, בקהלם unto their assembly, אל be not תחד thou united: כבדי mine honor, כי for באפם in their anger הרגו they slew אישׁ a man, וברצנם and in their self-will עקרו they digged down שׁור׃}%
\verse{ארור Cursed אפם their anger, כי for עז fierce; ועברתם and their wrath, כי for קשׁתה it was cruel: אחלקם I will divide ביעקב them in Jacob, ואפיצם and scatter בישׂראל׃ them in Israel.}%
\verse{יהודה Judah, אתה thou יודוך shall praise: אחיך whom thy brethren ידך thy hand בערף in the neck איביך of thine enemies; ישׁתחוו shall bow down לך בני children אביך׃ thy father's}%
\verse{גור whelp: אריה a lion's יהודה Judah מטרף from the prey, בני my son, עלית thou art gone up: כרע he stooped down, רבץ he couched כאריה as a lion, וכלביא and as an old lion; מי who יקימנו׃ shall rouse him up?}%
\verse{לא shall not יסור depart שׁבט The scepter מיהודה ומחקק nor a lawgiver מבין from between רגליו his feet, עד until כי until יבא come; שׁילה Shiloh ולו יקהת and unto him the gathering עמים׃ of the people}%
\verse{אסרי Binding לגפן unto the vine, עירה his foal ולשׂרקה unto the choice vine; בני colt אתנו and his ass's כבס he washed ביין in wine, לבשׁו his garments ובדם in the blood ענבים of grapes: סותה׃ and his clothes}%
\verse{חכלילי red עינים His eyes מיין with wine, ולבן white שׁנים and his teeth מחלב׃ with milk.}%
\verse{זבולן Zebulun לחוף at the haven ימים of the sea; ישׁכן shall dwell והוא and he לחוף for a haven אנית of ships; וירכתו and his border על unto צידן׃ Zidon.}%
\verse{ישׂשכר Issachar חמר ass גרם a strong רבץ couching down בין between המשׁפתים׃ two burdens:}%
\verse{וירא And he saw מנחה rest כי that טוב good, ואת הארץ and the land כי that נעמה pleasant; ויט and bowed שׁכמו his shoulder לסבל to bear, ויהי and became למס unto tribute. עבד׃ a servant}%
\verse{דן Dan ידין shall judge עמו his people, כאחד as one שׁבטי of the tribes ישׂראל׃ of Israel.}%
\verse{יהי shall be דן Dan נחשׁ a serpent עלי by דרך the way, שׁפיפן an adder עלי in ארח the path, הנשׁך that biteth עקבי heels, סוס the horse ויפל shall fall רכבו so that his rider אחור׃ backward.}%
\verse{לישׁועתך for thy salvation, קויתי I have waited יהוה׃ O LORD.}%
\verse{גד Gad, גדוד a troop יגודנו shall overcome והוא him: but he יגד shall overcome עקב׃ at the last.}%
\verse{מאשׁר שׁמנה fat, לחמו his bread והוא and he יתן shall yield מעדני dainties. מלך׃ royal}%
\verse{נפתלי Naphtali אילה a hind שׁלחה let loose: הנתן he giveth אמרי words. שׁפר׃ goodly}%
\verse{בן bough, פרת a fruitful יוסף Joseph בן bough פרת a fruitful עלי by עין a well; בנות branches צעדה run עלי over שׁור׃ the wall:}%
\verse{וימררהו have sorely grieved ורבו him, and shot וישׂטמהו and hated בעלי חצים׃}%
\verse{ותשׁב abode באיתן in strength, קשׁתו But his bow ויפזו were made strong זרעי and the arms ידיו of his hands מידי by the hands אביר of the mighty יעקב of Jacob; משׁם רעה the shepherd, אבן the stone ישׂראל׃ of Israel:)}%
\verse{מאל אביך of thy father, ויעזרך who shall help ואת שׁדי thee; and by the Almighty, ויברכך who shall bless ברכת thee with blessings שׁמים of heaven מעל ברכת blessings תהום of the deep רבצת that lieth תחת under, ברכת blessings שׁדים of the breasts, ורחם׃ and of the womb:}%
\verse{ברכת The blessings אביך of thy father גברו have prevailed על above ברכת the blessings הורי of my progenitors עד unto תאות the utmost bound גבעת hills: עולם of the everlasting תהיין they shall be לראשׁ on the head יוסף of Joseph, ולקדקד and on the crown of the head נזיר of him that was separate from אחיו׃ his brethren.}%
\verse{בנימין Benjamin זאב a wolf: יטרף shall ravin בבקר in the morning יאכל he shall devour עד the prey, ולערב and at night יחלק he shall divide שׁלל׃ the spoil.}%
\verseWithHeading{The Death and Burial of Jacob}{כל All אלה these שׁבטי tribes ישׂראל of Israel: שׁנים the twelve עשׂר the twelve וזאת and this אשׁר that דבר spoke להם אביהם their father ויברך unto them, and blessed אותם אישׁשׁר them; every one כברכתו according to his blessing ברך he blessed אתם׃}%
\verse{ויצו And he charged אותם ויאמר them, and said אלהם unto אני them, I נאסף am to be gathered אל unto עמי my people: קברו bury אתי אל me with אבתי my fathers אל in המערה the cave אשׁר that בשׂדה in the field עפרון of Ephron החתי׃ the Hittite,}%
\verse{במערה In the cave אשׁר that בשׂדה in the field המכפלה of Machpelah, אשׁר which על before פני before ממרא Mamre, בארץ in the land כנען of Canaan, אשׁר which קנה bought אברהם Abraham את השׂדה the field מאת עפרן Ephron החתי the Hittite לאחזת for a possession קבר׃ of a burial place.}%
\verse{שׁמה There קברו they buried את אברהם Abraham ואת שׂרה and Sarah אשׁתו his wife; שׁמה there קברו they buried את יצחק Isaac ואת רבקה and Rebekah אשׁתו his wife; ושׁמה and there קברתי I buried את לאה׃ Leah.}%
\verse{מקנה The purchase השׂדה of the field והמערה and of the cave אשׁר that בו מאת בני the children חת׃ of Heth.}%
\verse{ויכל had made an end יעקב And when Jacob לצות of commanding את בניו his sons, ויאסף he gathered up רגליו his feet אל into המטה the bed, ויגוע and yielded up the ghost, ויאסף and was gathered אל unto עמיו׃ his people.}%
\end{biblechapter}%
\begin{biblechapter}% Genesis 50
\verseWithHeading{Jacob’s Funeral and Joseph’s Remaining Time in Egypt}{ויפל fell יוסף And Joseph על upon פני face, אביו his father's ויבך and wept עליו upon וישׁק׃ him, and kissed}%
\verse{ויצו commanded יוסף And Joseph את עבדיו his servants את הרפאים the physicians לחנט to embalm את אביו his father: ויחנטו embalmed הרפאים and the physicians את ישׂראל׃ Israel.}%
\verse{וימלאו were fulfilled לו ארבעים And forty יום days כי for him; for כן so ימלאו are fulfilled ימי the days החנטים of those which are embalmed: ויבכו mourned אתו מצריםׁבעים for him threescore and ten יום׃ days.}%
\verse{ויעברו were past, ימי And when the days בכיתו of his mourning וידבר spoke יוסף Joseph אל unto בית the house פרעה of Pharaoh, לאמר saying, אם If נא now מצאתי I have found חן grace בעיניכם in your eyes, דברו speak, נא I pray you, באזני in the ears פרעה of Pharaoh, לאמר׃ saying,}%
\verse{אבי My father השׁביעני made me swear, לאמר saying, הנה Lo, אנכי I מת die: בקברי in my grave אשׁר which כריתי I have digged לי בארץ for me in the land כנען of Canaan, שׁמה there תקברני shalt thou bury ועתה me. Now אעלה therefore let me go up, נא I pray thee, ואקברה and bury את אבי my father, ואשׁובה׃ and I will come again.}%
\verse{ויאמר said, פרעה And Pharaoh עלה Go up, וקבר and bury את אביך thy father, כאשׁר according as השׁביעך׃ he made thee swear.}%
\verse{ויעל went up יוסף And Joseph לקבר to bury את אביו his father: ויעלו him went up אתו and with כל all עבדי the servants פרעה of Pharaoh, זקני the elders ביתו of his house, וכל and all זקני the elders ארץ of the land מצרים׃ of Egypt,}%
\verse{וכל And all בית the house יוסף of Joseph, ואחיו and his brethren, ובית house: אביו and his father's רק only טפם their little ones, וצאנם and their flocks, ובקרם and their herds, עזבו they left בארץ in the land גשׁן׃ of Goshen.}%
\verse{ויעל And there went up עמו with גם him both רכב chariots גם and פרשׁים horsemen: ויהי and it was המחנה company. כבד great מאד׃ a very}%
\verse{ויבאו And they came עד to גרן the threshingfloor האטד of Atad, אשׁר which בעבר beyond הירדן Jordan, ויספדו they mourned שׁם and there מספד lamentation: גדול with a great וכבד sore מאד and very ויעשׂ and he made לאביו for his father אבל a mourning שׁבעת seven ימים׃ days.}%
\verse{וירא saw יושׁב And when the inhabitants הארץ of the land, הכנעני the Canaanites, את האבל the mourning בגרן in the floor האטד of Atad, ויאמרו they said, אבל mourning כבד a grievous זה This למצרים על wherefore כן wherefore קרא of it was called שׁמה the name אבל מצרים Abel-mizraim, אשׁר which בעבר beyond הירדן׃ Jordan.}%
\verse{ויעשׂו did בניו And his sons לו כן unto him according as כאשׁר unto him according as צום׃ he commanded}%
\verse{וישׂאו carried אתו בניו For his sons ארצה him into the land כנען of Canaan, ויקברו and buried אתו במערת him in the cave שׂדה of the field המכפלה of Machpelah, אשׁר which קנה bought אברהם Abraham את השׂדה the field לאחזת for a possession קבר of a burial place מאת עפרן Ephron החתי the Hittite, על before פני before ממרא׃ Mamre.}%
\verse{וישׁב returned יוסף And Joseph מצרימה into Egypt, הוא he, ואחיו and his brethren, וכל and all העלים that went up אתו with לקבר him to bury את אביו his father, אחרי after קברו אתביו׃ his father.}%
\verse{ויראו saw אחי brethren יוסף And when Joseph's כי that מת was dead, אביהם their father ויאמרו they said, לו will peradventure ישׂטמנו hate יוסף Joseph והשׁב us, and will certainly requite ישׁיב us, and will certainly requite לנו את כל us all הרעה the evil אשׁר which גמלנו we did אתו׃}%
\verse{ויצוו And they sent a messenger אל unto יוסף Joseph, לאמר saying, אביך Thy father צוה did command לפני before מותו he died, לאמר׃ saying,}%
\verse{כה So תאמרו shall ye say ליוסף unto Joseph, אנא I pray thee שׂא Forgive, נא now, פשׁע the trespass אחיך of thy brethren, וחטאתם and their sin; כי for רעה unto thee evil: גמלוך they did ועתה and now, שׂא forgive נא we pray thee, לפשׁע the trespass עבדי of the servants אלהי of the God אביך of thy father. ויבך wept יוסף And Joseph בדברם when they spoke אליו׃ unto}%
\verse{וילכו went גם also אחיו And his brethren ויפלו and fell down לפניו before his face; ויאמרו and they said, הננו Behold, לך לעבדים׃ we thy servants.}%
\verse{ויאמר said אלהם unto יוסף And Joseph אל not: תיראו them, Fear כי for התחת in the place אלהים of God? אני׃ I}%
\verse{ואתם But as for you, חשׁבתם ye thought עלי against רעה evil אלהים me; God חשׁבה meant לטבה it unto good, למען to עשׂה bring to pass, כיום day, הזה as this להחית to save much people alive. עם רב׃}%
\verse{ועתה Now אל ye not: תיראו therefore fear אנכי I אכלכל will nourish אתכם ואתפכם you, and your little ones. וינחם And he comforted אותם וידבר them, and spoke על kindly לבם׃ kindly}%
\verseWithHeading{The Death of Joseph}{וישׁב dwelt יוסף And Joseph במצרים in Egypt, הוא he, ובית house: אביו and his father's ויחי lived יוסף and Joseph מאה a hundred ועשׂר and ten שׁנים׃ years.}%
\verse{וירא saw יוסף And Joseph לאפרים Ephraim's בני children שׁלשׁים of the third גם also בני the children מכיר of Machir בן the son מנשׁה of Manasseh ילדו were brought up על upon ברכי knees. יוסף׃ Joseph's}%
\verse{ויאמר said יוסף And Joseph אל unto אחיו his brethren, אנכי I מת die: ואלהים and God פקד will surely visit יפקד will surely visit אתכם והעלה you, and bring אתכם מן you out of הארץ land הזאת this אל unto הארץ the land אשׁר which נשׁבע he swore לאברהם to Abraham, ליצחק to Isaac, וליעקב׃ and to Jacob.}%
\verse{וישׁבע took an oath יוסף And Joseph את בני of the children ישׂראל of Israel, לאמר saying, פקד will surely visit יפקד will surely visit אלהים God אתכם והעלתם you, and ye shall carry up את עצמתי my bones מזה׃ from hence.}%
\verse{וימת died, יוסף So Joseph בן old: מאה a hundred ועשׂר and ten שׁנים years ויחנטו and they embalmed אתו ויישׂם him, and he was put בארון in a coffin במצרים׃ in Egypt.}%
\end{biblechapter}%
