\biblebook{Habakkuk}
\begin{biblechapter}% Habakkuk 1
\verseWithHeading{Habakkuk’s Complaint}{המשׂא The burden אשׁר which חזה did see. חבקוק Habakkuk הנביא׃ the prophet}%
\verse{עד how long אנה how long יהוה O LORD, שׁועתי shall I cry, ולא and thou wilt not תשׁמע hear! אזעק cry out אליך unto חמס thee violence, ולא and thou wilt not תושׁיע׃ save!}%
\verse{למה Why תראני dost thou show און me iniquity, ועמל grievance? תביט and cause to behold ושׁד for spoiling וחמס and violence לנגדי before ויהי me: and there are ריב strife ומדון and contention. ישׂא׃ raise up}%
\verse{על therefore כן therefore תפוג is slacked, תורה the law ולא doth never יצא go forth: לנצח doth never משׁפט and judgment כי for רשׁע the wicked מכתיר doth compass about את הצדיק the righteous; על כןצא proceedeth. משׁפט judgment מעקל׃ wrong}%
\verseWithHeading{God’s Answer to Habakkuk}{ראו Behold בגוים ye among the heathen, והביטו and regard, והתמהו and wonder marvelously: תמהו and wonder marvelously: כי for פעל a work פעל will work בימיכם in your days, לא ye will not תאמינו believe, כי though יספר׃ it be told}%
\verse{כי For, הנני מקים I raise up את הכשׂדים the Chaldeans, הגוי nation, המר bitter והנמהר and hasty ההולך which shall march למרחבי through the breadth ארץ of the land, לרשׁת to possess משׁכנות the dwelling places לא׃ not}%
\verse{אים terrible ונורא and dreadful: הוא They ממנו of משׁפטו their judgment ושׂאתו and their dignity יצא׃ shall proceed}%
\verse{וקלו also are swifter מנמרים than the leopards, סוסיו Their horses וחדו and are more fierce מזאבי wolves: ערב than the evening ופשׁו shall spread פרשׁיו and their horsemen ופרשׁיו themselves, and their horsemen מרחוק from far; יבאו shall come יעפו they shall fly כנשׁר as the eagle חשׁ hasteth לאכול׃ to eat.}%
\verse{כלה all לחמס for violence: יבוא They shall come מגמת shall sup up פניהם their faces קדימה the east wind, ויאסף and they shall gather כחול as the sand. שׁבי׃ the captivity}%
\verse{והוא And they במלכים at the kings, יתקלס shall scoff ורזנים and the princes משׂחק shall be a scorn לו הוא unto them: they לכל every מבצר stronghold; ישׂחק shall deride ויצבר for they shall heap עפר dust, וילכדה׃ and take}%
\verse{אז Then חלף change, רוח shall mind ויעבר and he shall pass over, ואשׁם and offend, זו this כחו his power לאלהו׃}%
\verseWithHeading{Habakkuk’s Second Complaint}{הלוא not אתה thou מקדם from everlasting, יהוה אלהי my God, קדשׁי mine Holy One? לא we shall not נמות die. יהוה למשׁפט them for judgment; שׂמתו thou hast ordained וצור and, O mighty God, להוכיח them for correction. יסדתו׃ thou hast established}%
\verse{טהור עינים eyes מראות than to behold רע evil, והביט look אל on עמל iniquity: לא not תוכל and canst למה wherefore תביט lookest בוגדים thou upon them that deal treacherously, תחרישׁ holdest thy tongue בבלע devoureth רשׁע when the wicked צדיק more righteous ממנו׃ than to behold}%
\verse{ותעשׂה And makest אדם men כדגי as the fishes הים of the sea, כרמשׂ as the creeping things, לא no משׁל׃ ruler}%
\verse{כלה all בחכה of them with the angle, העלה They take up יגרהו they catch בחרמו them in their net, ויאספהו and gather במכמרתו them in their drag: על therefore כן therefore ישׂמח they rejoice ויגיל׃ and are glad.}%
\verse{על כןזבח they sacrifice לחרמו unto their net, ויקטר and burn incense למכמרתו unto their drag; כי because בהמה by them שׁמן fat, חלקו their portion ומאכלו and their meat בראה׃ plenteous.}%
\verse{העל כןריק empty חרמו their net, ותמיד continually להרג to slay גוים the nations? לא and not יחמול׃ spare}%
\end{biblechapter}%
\begin{biblechapter}% Habakkuk 2
\verseWithHeading{The Righteous Will Live by Faith}{על upon משׁמרתי my watch, אעמדה I will stand ואתיצבה and set על me upon מצור the tower, ואצפה and will watch לראות to see מה what ידבר he will say בי ומה unto me, and what אשׁיב I shall answer על when תוכחתי׃ I am reproved.}%
\verse{ויענני answered יהוה And the LORD ויאמר me, and said, כתוב Write חזון the vision, ובאר and make plain על upon הלחות tables, למען that ירוץ he may run קורא׃ that readeth}%
\verse{כי For עוד yet חזון the vision למועד for an appointed time, ויפח it shall speak, לקץ but at the end ולא and not יכזב lie: אם though יתמהמה it tarry, חכה wait לו כי for it; because בא it will surely come, יבא it will surely come, לא it will not יאחר׃ tarry.}%
\verse{הנה Behold, עפלה is lifted up לא is not upright ישׁרה is not upright נפשׁו his soul בו וצדיק in him: but the just באמונתו by his faith. יחיה׃ shall live}%
\verse{ואף Yea כי also, because היין by wine, בוגד he transgresseth גבר man, יהיר a proud ולא neither ינוה keepeth at home, אשׁר who הרחיב enlargeth כשׁאול as hell, נפשׁו his desire והוא כמות and as death, ולא and cannot ישׂבע be satisfied, ויאסף but gathereth אליו unto כל him all הגוים nations, ויקבץ and heapeth אליו unto כל him all העמים׃ people:}%
\verse{הלוא Shall not אלה these כלם all עליו against משׁל a parable ישׂאו take up ומליצה him, and a taunting חידות proverb לו ויאמר against him, and say, הוי Woe המרבה to him that increaseth לא not לו עד מתימכביד and to him that ladeth עליו and to him that ladeth עבטיט׃ himself with thick clay!}%
\verse{הלוא Shall they not פתע suddenly יקומו rise up נשׁכיך that shall bite ויקצו מזעזעיך that shall vex והיית thee, and thou shalt be למשׁסות׃ for booties}%
\verse{כי Because אתה thou שׁלות hast spoiled גוים nations, רבים many ישׁלוך shall spoil כל all יתר the remnant עמים of the people מדמי blood, אדם thee; because of men's וחמס and the violence ארץ of the land, קריה of the city, וכל and of all ישׁבי׃ that dwell}%
\verse{הוי Woe בצע to him that coveteth בצע covetousness רע an evil לביתו to his house, לשׂום that he may set במרום on high, קנו his nest להנצל that he may be delivered מכף from the power רע׃ of evil!}%
\verse{יעצת Thou hast consulted בשׁת shame לביתך to thy house קצות by cutting off עמים people, רבים many וחוטא and hast sinned נפשׁך׃ thy soul.}%
\verse{כי For אבן the stone מקיר of the wall, תזעק shall cry out וכפיס and the beam מעץ out of the timber יעננה׃ shall answer}%
\verse{הוי Woe בנה to him that buildeth עיר a town בדמים with blood, וכונן and establisheth קריה a city בעולה׃ by iniquity!}%
\verse{הלוא not הנה Behold, מאת יהוה the LORD צבאות of hosts וייגעו shall labor עמים that the people בדי in the very אשׁ fire, ולאמים and the people בדי themselves for very ריק vanity? יעפו׃ shall weary}%
\verse{כי For תמלא shall be filled הארץ the earth לדעת אתבוד of the glory יהוה of the LORD, כמים as the waters יכסו cover על cover ים׃ the sea.}%
\verse{הוי Woe משׁקה unto him that giveth his neighbor drink, רעהו unto him that giveth his neighbor drink, מספח that puttest חמתך thy bottle ואף also, שׁכר to and makest drunken למען that הביט thou mayest look על on מעוריהם׃ their nakedness!}%
\verse{שׂבעת Thou art filled קלון with shame מכבוד for glory: שׁתה drink גם also, אתה thou והערל and let thy foreskin be uncovered: תסוב shall be turned עליך unto כוס the cup ימין right hand יהוה of the LORD's וקיקלון thee, and shameful spewing על on כבודך׃ thy glory.}%
\verse{כי For חמס the violence לבנון of Lebanon יכסך shall cover ושׁד thee, and the spoil בהמות of beasts, יחיתן made them afraid, מדמי blood, אדם because of men's וחמס and for the violence ארץ of the land, קריה of the city, וכל and of all ישׁבי׃ that dwell}%
\verse{מה What הועיל profiteth פסל the graven image כי that פסלו thereof hath graven יצרו of his work מסכה it; the molten image, ומורה and a teacher שׁקר of lies, כי that בטח trusteth יצר the maker יצרו the maker עליו therein, לעשׂות to make אלילים idols? אלמים׃ dumb}%
\verse{הוי Woe אמר unto him that saith לעץ to the wood, הקיצה Awake; עורי Arise, לאבן stone, דומם to the dumb הוא it יורה shall teach! הנה Behold, הוא it תפושׂ laid over זהב with gold וכסף and silver, וכל at all רוח breath אין and no בקרבו׃ in the midst}%
\verse{ויהוה But the LORD בהיכל temple: קדשׁו in his holy הס keep silence מפניו before כל let all הארץ׃ the earth}%
\end{biblechapter}%
\begin{biblechapter}% Habakkuk 3
\verseWithHeading{The Prayer of Habakkuk}{תפלה A prayer לחבקוק of Habakkuk הנביא the prophet על upon שׁגינות׃ Shigionoth.}%
\verse{יהוה O LORD, שׁמעתי I have heard שׁמעך thy speech, יראתי was afraid: יהוה O LORD, פעלך thy work בקרב in the midst שׁנים of the years, חייהו revive בקרב in the midst שׁנים of the years תודיע make known; ברגז in wrath רחם mercy. תזכור׃ remember}%
\verse{אלוה God מתימן יבוא came וקדושׁ and the Holy One מהר from mount פארן Paran. סלה Selah. כסה covered שׁמים the heavens, הודו His glory ותהלתו of his praise. מלאה was full הארץ׃ and the earth}%
\verse{ונגה And brightness כאור as the light; תהיה was קרנים he had horns מידו out of his hand: לו ושׁם and there חביון the hiding עזה׃ of his power.}%
\verse{לפניו Before ילך him went דבר the pestilence, ויצא went forth רשׁף and burning coals לרגליו׃ at his feet.}%
\verse{עמד He stood, וימדד ארץ the earth: ראה he beheld, ויתר and drove asunder גוים the nations; ויתפצצו were scattered, הררי mountains עד and the everlasting שׁחו did bow: גבעות hills עולם the perpetual הליכות his ways עולם׃ everlasting.}%
\verse{תחת in און affliction: ראיתי I saw אהלי the tents כושׁן of Cushan ירגזון did tremble. יריעות the curtains ארץ of the land מדין׃ of Midian}%
\verse{הבנהרים against the rivers? חרה displeased יהוה Was the LORD אם בנהרים against the rivers? אפך thine anger אם בים against the sea, עברתך thy wrath כי that תרכב thou didst ride על upon סוסיך thine horses מרכבתיך thy chariots ישׁועה׃ of salvation?}%
\verse{עריה was made quite naked, תעור was made quite naked, קשׁתך Thy bow שׁבעות to the oaths מטות of the tribes, אמר word. סלה Selah. נהרות with rivers. תבקע Thou didst cleave ארץ׃ the earth}%
\verse{ראוך saw יחילו thee, they trembled: הרים The mountains זרם the overflowing מים of the water עבר passed by: נתן uttered תהום the deep קולו his voice, רום on high. ידיהו his hands נשׂא׃ lifted up}%
\verse{שׁמשׁ The sun ירח moon עמד stood still זבלה in their habitation: לאור at the light חציך of thine arrows יהלכו they went, לנגה at the shining ברק of thy glittering חניתך׃ spear.}%
\verse{בזעם in indignation, תצעד Thou didst march through ארץ the land באף in anger. תדושׁ thou didst thresh גוים׃ the heathen}%
\verse{יצאת Thou wentest forth לישׁע for the salvation עמך of thy people, לישׁע for salvation את משׁיחך thine anointed; מחצת thou woundedst ראשׁ the head מבית out of the house רשׁע of the wicked, ערות by discovering יסוד the foundation עד unto צואר the neck. סלה׃ Selah.}%
\verse{נקבת Thou didst strike through במטיו with his staves ראשׁ the head פרזו of his villages: יסערו they came out as a whirlwind להפיצני to scatter עליצתם me: their rejoicing כמו as לאכל to devour עני the poor במסתר׃ secretly.}%
\verse{דרכת Thou didst walk בים through the sea סוסיך with thine horses, חמר the heap מים waters. רבים׃ of great}%
\verse{שׁמעתי When I heard, ותרגז trembled; בטני my belly לקול at the voice: צללו quivered שׂפתי my lips יבוא entered רקב rottenness בעצמי into my bones, ותחתי in ארגז and I trembled אשׁר myself, that אנוח I might rest ליום in the day צרה of trouble: לעלות when he cometh up לעם unto the people, יגודנו׃ he will invade}%
\verse{כי Although תאנה the fig tree לא shall not תפרח blossom, ואין neither יבול fruit בגפנים in the vines; כחשׁ shall fail, מעשׂה the labor זית of the olive ושׁדמות and the fields לא no עשׂה shall yield אכל meat; גזר shall be cut off ממכלה from the fold, צאן the flock ואין and no בקר herd ברפתים׃ in the stalls:}%
\verse{ואני Yet I ביהוה in the LORD, אעלוזה will rejoice אגילה I will joy באלהי in the God ישׁעי׃ of my salvation.}%
\verse{יהוה אדני God חילי my strength, וישׂם and he will make רגלי my feet כאילות like hinds' ועל upon במותי mine high places. ידרכני and he will make me to walk למנצח To the chief singer בנגינותי׃ on my stringed instruments.}%
\end{biblechapter}%
