\biblebook{Job}
\begin{biblechapter}% Job 1
\verseWithHeading{Job’s Character and Greatness}{אישׁ a man היה There was בארץ in the land עוץ of Uz, איוב Job; שׁמו whose name והיה was האישׁ man ההוא and that תם perfect וישׁר and upright, וירא and one that feared אלהים God, וסר and eschewed מרע׃ evil.}%
\verse{ויולדו And there were born לו שׁבעה unto him seven בנים sons ושׁלושׁ and three בנות׃ daughters.}%
\verse{ויהי also was מקנהו His substance שׁבעת seven אלפי thousand צאן sheep, ושׁלשׁת and three אלפי thousand גמלים camels, וחמשׁ and five מאות hundred צמד yoke בקר of oxen, וחמשׁ and five מאות hundred אתונות she asses, ועבדה household; רבה great מאד and a very ויהי was האישׁ man ההוא so that this גדול the greatest מכל of all בני the men קדם׃ of the east.}%
\verse{והלכו went בניו And his sons ועשׂו and feasted משׁתה and feasted בית houses, אישׁ every one יומו his day; ושׁלחו and sent וקראו and called לשׁלשׁת for their three אחיתיהם sisters לאכל to eat ולשׁתות and to drink עמהם׃ with}%
\verse{ויהי And it was כי so, when הקיפו were gone about, ימי the days המשׁתה of feasting וישׁלח sent איוב that Job ויקדשׁם and sanctified והשׁכים them, and rose up early בבקר in the morning, והעלה and offered עלות burnt offerings מספר to the number כלם of them all: כי for אמר said, איוב Job אולי It may be חטאו have sinned, בני that my sons וברכו and cursed אלהים God בלבבם in their hearts. ככה Thus יעשׂה did איוב Job כל continually. הימים׃ continually.}%
\verse{ויהי Now there was היום a day ויבאו came בני when the sons האלהים of God להתיצב to present themselves על before יהוה the LORD, ויבוא came גם also השׂטן and Satan בתוכם׃ among}%
\verse{ויאמר said יהוה אל unto השׂטן Satan, מאין תבא comest ויען answered השׂטן thou? Then Satan את יהוהיאמר and said, משׁוט בארץ in the earth, ומהתהלך׃ and from walking up and down}%
\verse{ויאמר said יהוה And the LORD אל unto השׂטן Satan, השׂמת לבךל עבדי my servant איוב Job, כי that אין none כמהו like him בארץ in the earth, אישׁ man, תם a perfect וישׁר and an upright ירא one that feareth אלהים God, וסר and escheweth מרע׃ evil?}%
\verse{ויען answered השׂטן Then Satan את יהוה the LORD, ויאמר and said, החנם for naught? ירא fear איוב Doth Job אלהים׃ God}%
\verse{הלא Hast not את thou שׂכת made a hedge בעדו about ובעד him, and about ביתו his house, ובעד and about כל all אשׁר that לו מסביב he hath on every side? מעשׂה the work ידיו of his hands, ברכת thou hast blessed ומקנהו and his substance פרץ is increased בארץ׃ in the land.}%
\verse{ואולם But שׁלח put forth נא now, ידך thine hand וגע and touch בכל all אשׁר that לו אם he hath, and לא על thee to פניך thy face. יברכך׃ he will curse}%
\verse{ויאמר said יהוה And the LORD אל unto השׂטן Satan, הנה Behold, כל all אשׁר that לו בידך he hath in thy power; רק only אליו upon אל himself put not forth תשׁלח himself put not forth ידך thine hand. ויצא went forth השׂטן So Satan מעם from פני the presence יהוה׃ of the LORD.}%
\verseWithHeading{The Adversary’s Attack on Job’s Possessions}{ויהי And there was היום a day ובניו when his sons ובנתיו and his daughters אכלים eating ושׁתים and drinking יין wine בבית house: אחיהם brother's הבכור׃ in their eldest}%
\verse{ומלאך a messenger בא And there came אל unto איוב Job, ויאמר and said, הבקר The oxen היו were חרשׁות plowing, והאתנות and the asses רעות feeding על beside ידיהם׃ beside}%
\verse{ותפל fell שׁבא And the Sabeans ותקחם and took them away; ואת הנערים the servants הכו yea, they have slain לפי with the edge חרב of the sword; ואמלטה am escaped רק only אני and I לבדי alone להגיד׃ to tell}%
\verse{עוד yet זה While he מדבר speaking, וזה also another, בא there came ויאמר and said, אשׁ The fire אלהים of God נפלה is fallen מן from השׁמים heaven, ותבער and hath burned up בצאן the sheep, ובנערים and the servants, ותאכלם and consumed ואמלטה am escaped רק only אני them; and I לבדי alone להגיד׃ to tell}%
\verse{עוד yet זה While he מדבר speaking, וזה also another, בא there came ויאמר and said, כשׂדים The Chaldeans שׂמו made שׁלשׁה out three ראשׁים bands, ויפשׁטו and fell על upon הגמלים the camels, ויקחום and have carried them away, ואת הנערים the servants הכו yea, and slain לפי with the edge חרב of the sword; ואמלטה am escaped רק only אני and I לבדי alone להגיד׃ to tell}%
\verse{עד זה While he מדבר speaking, וזה also another, בא there came ויאמר and said, בניך Thy sons ובנותיך and thy daughters אכלים eating ושׁתים and drinking יין wine בבית house: אחיהם brother's הבכור׃ in their eldest}%
\verse{והנה And, behold, רוח wind גדולה a great באה there came מעבר from המדבר the wilderness, ויגע and smote בארבע the four פנות corners הבית of the house, ויפל and it fell על upon הנערים the young men, וימותו and they are dead; ואמלטה am escaped רק only אני and I לבדי alone להגיד׃ to tell}%
\verse{ויקם arose, איוב Then Job ויקרע and rent את מעלו his mantle, ויגז and shaved את ראשׁו his head, ויפל and fell down ארצה upon the ground, וישׁתחו׃ and worshiped,}%
\verse{ויאמר And said, ערם Naked יצתי came I out מבטן אמיערם and naked אשׁוב shall I return שׁמה thither: יהוה the LORD נתן gave, ויהוה and the LORD לקח hath taken away; יהי be שׁם the name יהוה of the LORD. מברך׃ blessed}%
\verse{בכל In all זאת this לא not, חטא sinned איוב Job ולא nor נתן charged תפלה foolishly. לאלהים׃ God}%
\end{biblechapter}%
\begin{biblechapter}% Job 2
\verseWithHeading{The Adversary’s Attack on Job’s Person}{ויהי Again there was היום a day ויבאו came בני when the sons האלהים of God להתיצב to present themselves על before יהוה the LORD, ויבוא came גם also השׂטן and Satan בתכם among להתיצב them to present himself על before יהוה׃ the LORD.}%
\verse{ויאמר said יהוה And the LORD אל unto השׂטן Satan, אי מזהבא comest ויען answered השׂטן thou? And Satan את יהוה the LORD, ויאמר and said, משׁט בארץ in the earth, ומהתהלך׃ and from walking up and down}%
\verse{ויאמר said יהוה And the LORD אל unto השׂטן Satan, השׂמת לבךל עבדי my servant איוב Job, כי that אין none כמהו like him בארץ in the earth, אישׁ man, תם a perfect וישׁר and an upright ירא one that feareth אלהים God, וסר and escheweth מרע evil? ועדנו and still מחזיק he holdeth fast בתמתו his integrity, ותסיתני although thou movedst בו לבלעו me against him, to destroy חנם׃ him without cause.}%
\verse{ויען answered השׂטן And Satan את יהוה the LORD, ויאמר and said, עור בעד for עור וכל yea, all אשׁר that לאישׁ a man יתן hath will he give בעד נפשׁו׃ his life.}%
\verse{אולם But שׁלח put forth נא now, ידך thine hand וגע and touch אל and touch עצמו his bone ואל thee to בשׂרו and his flesh, אם and לא אלניך thy face. יברכך׃ he will curse}%
\verse{ויאמר said יהוה אל unto השׂטן Satan, הנו בידך he in thine hand; אך but את נפשׁו his life. שׁמר׃ save}%
\verseWithHeading{Job’s Blameless Behavior}{ויצא השׂטןאת פני the presence יהוה of the LORD, ויך and smote את איוב Job בשׁחין boils רע with sore מכף from the sole רגלו of his foot עד unto קדקדו׃ his crown.}%
\verse{ויקח And he took לו חרשׂ him a potsherd להתגרד to scrape himself בו והוא withal; and he ישׁב sat down בתוך among האפר׃ the ashes.}%
\verse{ותאמר Then said לו אשׁתו his wife עדך unto him, Dost thou still מחזיק retain בתמתך thine integrity? ברך curse אלהים God, ומת׃ and die.}%
\verse{ויאמר But he said אליה unto כדבר her, Thou speakest אחת as one הנבלות of the foolish women תדברי speaketh. גם What? את הטוב good נקבל shall we receive מאת האלהים of God, ואת הרע evil? לא and shall we not נקבל receive בכל In all זאת this לא did not חטא sin איוב Job בשׂפתיו׃ with his lips.}%
\verse{וישׁמעו heard שׁלשׁת three רעי friends איוב Now when Job's את כל of all הרעה evil הזאת this הבאה that was come עליו upon ויבאו him, they came אישׁ every one ממקמו from his own place; אליפז Eliphaz התימני the Temanite, ובלדד and Bildad השׁוחי the Shuhite, וצופר and Zophar הנעמתי the Naamathite: ויועדו for they had made an appointment יחדו together לבוא to come לנוד to mourn לו ולנחמו׃ with him and to comfort}%
\verse{וישׂאו And when they lifted up את עיניהם their eyes מרחוק afar off, ולא him not, הכירהו and knew וישׂאו they lifted up קולם their voice, ויבכו and wept; ויקרעו and they rent אישׁ every one מעלו his mantle, ויזרקו and sprinkled עפר dust על upon ראשׁיהם their heads השׁמימה׃ toward heaven.}%
\verse{וישׁבו So they sat down אתו with לארץ him upon the ground שׁבעת seven ימים days ושׁבעת and seven לילות nights, ואין and none דבר spoke אליו unto דבר a word כי him: for ראו they saw כי that גדל was very great. הכאב grief מאד׃ was very great.}%
\end{biblechapter}%
\begin{biblechapter}% Job 3
\verseWithHeading{Job Regrets His Birth}{אחרי After כן this פתח opened איוב Job את פיהו his mouth, ויקלל and cursed את יומו׃ his day.}%
\verse{ויען spoke, איוב And Job ויאמר׃ and said,}%
\verse{יאבד perish יום Let the day אולד wherein I was born, בו והלילה and the night אמר it was said, הרה גבר׃ There is a man child}%
\verse{היום day ההוא Let that יהי be חשׁך darkness; אל let not ידרשׁהו regard אלוה God ממעל it from above, ואל neither תופע shine עליו upon נהרה׃ let the light}%
\verse{יגאלהו stain חשׁך Let darkness וצלמות and the shadow of death תשׁכן dwell עליו upon עננה it; let a cloud יבעתהו terrify כמרירי it; let the blackness יום׃ of the day}%
\verse{הלילה night, ההוא As that יקחהו seize upon אפל let darkness אל it; let it not יחד be joined בימי unto the days שׁנה of the year, במספר into the number ירחים of the months. אל let it not יבא׃ come}%
\verse{הנה Lo, הלילה night ההוא let that יהי be גלמוד solitary, אל let no תבא come רננה׃ joyful voice}%
\verse{יקבהו Let them curse אררי it that curse יום the day, העתידים who are ready ערר to raise up לויתן׃ their mourning.}%
\verse{יחשׁכו thereof be dark; כוכבי Let the stars נשׁפו of the twilight יקו let it look לאור for light, ואין but none; ואל neither יראה let it see בעפעפי the dawning שׁחר׃ of the day:}%
\verse{כי Because לא it shut not up סגר it shut not up דלתי the doors בטני of my womb, ויסתר nor hid עמל sorrow מעיני׃ from mine eyes.}%
\verseWithHeading{Job Wishes He Had Died}{למה Why לא I not מרחם from the womb? אמות died מבטן of the belly? יצאתי when I came out ואגוע׃ did I give up the ghost}%
\verse{מדוע Why קדמוני prevent ברכים did the knees ומה me? or why שׁדים the breasts כי that אינק׃ I should suck?}%
\verse{כי For עתה now שׁכבתי should I have lain still ואשׁקוט and been quiet, ישׁנתי I should have slept: אז then ינוח׃ had I been at rest,}%
\verse{עם With מלכים kings ויעצי and counselors ארץ of the earth, הבנים which built חרבות׃ desolate places}%
\verse{או Or עם with שׂרים princes זהב that had gold, להם הממלאים who filled בתיהם their houses כסף׃ with silver:}%
\verse{או Or כנפל untimely birth טמון as a hidden לא I had not אהיה been; כעללים as infants לא never ראו saw אור׃ light.}%
\verse{שׁם There רשׁעים the wicked חדלו cease רגז troubling; ושׁם and there ינוחו be at rest. יגיעי the weary כח׃ the weary}%
\verse{יחד together; אסירים the prisoners שׁאננו rest לא not שׁמעו they hear קול the voice נגשׂ׃ of the oppressor.}%
\verse{קטן The small וגדול and great שׁם are there; הוא ועבד and the servant חפשׁי free מאדניו׃ from his master.}%
\verseWithHeading{Job Wishes He Might Die}{למה Wherefore יתן given לעמל to him that is in misery, אור is light וחיים and life למרי unto the bitter נפשׁ׃ soul;}%
\verse{המחכים Which long למות for death, ואיננו but it not; ויחפרהו and dig ממטמונים׃ for it more than for hid treasures;}%
\verse{השׂמחים אלי exceedingly, גיל exceedingly, ישׂישׂו are glad, כי when ימצאו they can find קבר׃ the grave?}%
\verse{לגבר to a man אשׁר whose דרכו way נסתרה is hid, ויסך hath hedged in? אלוה God בעדו׃ and whom}%
\verse{כי For לפני before לחמי I eat, אנחתי my sighing תבא cometh ויתכו are poured out כמים like the waters. שׁאגתי׃ and my roarings}%
\verse{כי For פחד פחדתייאתיני is come upon ואשׁר me, and that which יגרתי I was afraid of יבא׃ is come}%
\verse{לא I was not שׁלותי in safety, ולא neither שׁקטתי had I rest, ולא neither נחתי was I quiet; ויבא came. רגז׃ yet trouble}%
\end{biblechapter}%
\begin{biblechapter}% Job 4
\verseWithHeading{Eliphaz’s First Response to Job}{ויען answered אליפז Then Eliphaz התימני the Temanite ויאמר׃ and said,}%
\verse{הנסה דברליך with תלאה thee, wilt thou be grieved? ועצר withhold במלין himself from speaking? מי but who יוכל׃ can}%
\verse{הנה Behold, יסרת thou hast instructed רבים many, וידים hands. רפות the weak תחזק׃ and thou hast strengthened}%
\verse{כושׁל him that was falling, יקימון have upheld מליך Thy words וברכים knees. כרעות the feeble תאמץ׃ and thou hast strengthened}%
\verse{כי But עתה now תבוא it is come אליך upon ותלא thee, and thou faintest; תגע it toucheth עדיך it toucheth ותבהל׃ thee, and thou art troubled.}%
\verse{הלא not יראתך כסלתך thy confidence, תקותך thy hope, ותם and the uprightness דרכיך׃ of thy ways?}%
\verse{זכר Remember, נא I pray thee, מי who הוא נקי being innocent? אבד perished, ואיפה or where ישׁרים were the righteous נכחדו׃ cut off?}%
\verse{כאשׁר Even as ראיתי I have seen, חרשׁי they that plow און iniquity, וזרעי and sow עמל wickedness, יקצרהו׃ reap}%
\verse{מנשׁמת אלוה of God יאבדו they perish, ומרוח and by the breath אפו of his nostrils יכלו׃ are they consumed.}%
\verse{שׁאגת The roaring אריה of the lion, וקול and the voice שׁחל of the fierce lion, ושׁני and the teeth כפירים of the young lions, נתעו׃ are broken.}%
\verse{לישׁ The old lion אבד perisheth מבלי for lack טרף of prey, ובני whelps לביא and the stout lion's יתפרדו׃ are scattered abroad.}%
\verse{ואלי to דבר Now a thing יגנב was secretly brought ותקח received אזני me, and mine ear שׁמץ a little מנהו׃ thereof.}%
\verse{בשׂעפים In thoughts מחזינות from the visions לילה of the night, בנפל falleth תרדמה when deep sleep על on אנשׁים׃ men,}%
\verse{פחד Fear קראני came upon ורעדה me, and trembling, ורב which made all עצמותי my bones הפחיד׃ to shake.}%
\verse{ורוח Then a spirit על before פני my face; יחלף passed תסמר stood up: שׂערת the hair בשׂרי׃ of my flesh}%
\verse{יעמד It stood still, ולא but I could not אכיר discern מראהו the form תמונה thereof: an image לנגד before עיני mine eyes, דממה silence, וקול a voice, אשׁמע׃ and I heard}%
\verse{האנושׁ מאלוהצדק אםעשׂהו than his maker? יטהר be more pure גבר׃ shall a man}%
\verse{הן Behold, בעבדיו in his servants; לא he put no trust יאמין he put no trust ובמלאכיו and his angels ישׂים he charged תהלה׃ with folly:}%
\verse{אף How much less שׁכני them that dwell in בתי houses חמר of clay, אשׁר whose בעפר in the dust, יסודם foundation ידכאום are crushed לפני before עשׁ׃ the moth?}%
\verse{מבקר from morning לערב to evening: יכתו They are destroyed מבלי without any משׂים regarding לנצח forever יאבדו׃ they perish}%
\verse{הלא Doth not נסע in them go away? יתרם their excellency בם ימותו they die, ולא even without בחכמה׃ wisdom.}%
\end{biblechapter}%
\begin{biblechapter}% Job 5
\verseWithHeading{Eliphaz’s Response Continues}{קרא Call נא now, הישׁ if there be עונך any that will answer ואל thee; and to מי which מקדשׁים of the saints תפנה׃ wilt thou turn?}%
\verse{כי For לאויל the foolish יהרג killeth כעשׂ wrath ופתה the silly תמית slayeth קנאה׃ man, and envy}%
\verse{אני I ראיתי have seen אויל the foolish משׁרישׁ taking root: ואקוב I cursed נוהו his habitation. פתאם׃ but suddenly}%
\verse{ירחקו are far בניו His children מישׁע from safety, וידכאו and they are crushed בשׁער in the gate, ואין neither מציל׃ any to deliver}%
\verse{אשׁר Whose קצירו harvest רעב יאכל eateth up, ואל it even out of מצנים the thorns, יקחהו and taketh ושׁאף swalloweth up צמים and the robber חילם׃ their substance.}%
\verse{כי Although לא cometh not forth יצא cometh not forth מעפר of the dust, און affliction ומאדמה out of the ground; לא neither יצמח spring עמל׃ doth trouble}%
\verse{כי Yet אדם man לעמל unto trouble, יולד is born ובני as the sparks רשׁף as the sparks יגביהו upward. עוף׃ fly}%
\verse{אולם אני I אדרשׁ would seek אל unto אל God, ואל and unto אלהים God אשׂים would I commit דברתי׃ my cause:}%
\verse{עשׂה Which doeth גדלות great things ואין and unsearchable; חקר and unsearchable; נפלאות marvelous things עד without אין without מספר׃ number:}%
\verse{הנתן Who giveth מטר rain על upon פני upon ארץ the earth, ושׁלח and sendeth מים waters על upon פני upon חוצות׃ the fields:}%
\verse{לשׂום To set up שׁפלים those that be low; למרום on high וקדרים that those which mourn שׂגבו may be exalted ישׁע׃ to safety.}%
\verse{מפר He disappointeth מחשׁבות the devices ערומים of the crafty, ולא cannot תעשׂינה perform ידיהם so that their hands תושׁיה׃ enterprise.}%
\verse{לכד He taketh חכמים the wise בערמם in their own craftiness: ועצת and the counsel נפתלים of the froward נמהרה׃ is carried headlong.}%
\verse{יומם in the daytime, יפגשׁו They meet חשׁך with darkness וכלילה as in the night. ימשׁשׁו and grope בצהרים׃ in the noonday}%
\verse{וישׁע But he saveth מחרב from the sword, מפיהם from their mouth, ומיד and from the hand חזק of the mighty. אביון׃ the poor}%
\verse{ותהי hath לדל So the poor תקוה hope, ועלתה and iniquity קפצה stoppeth פיה׃ her mouth.}%
\verse{הנה Behold, אשׁרי happy אנושׁ the man יוכחנו correcteth: אלוה whom God ומוסר thou the chastening שׁדי of the Almighty: אל not תמאס׃ therefore despise}%
\verse{כי For הוא he יכאיב maketh sore, ויחבשׁ and bindeth up: ימחץ he woundeth, וידו and his hands תרפינה׃ make whole.}%
\verse{בשׁשׁ thee in six צרות troubles: יצילך He shall deliver ובשׁבע yea, in seven לא there shall no יגע touch בך רע׃ evil}%
\verse{ברעב In famine פדך he shall redeem ממות thee from death: ובמלחמה and in war מידי from the power חרב׃ of the sword.}%
\verse{בשׁוט from the scourge לשׁון of the tongue: תחבא Thou shalt be hid ולא neither תירא shalt thou be afraid משׁד of destruction כי when יבוא׃ it cometh.}%
\verse{לשׁד At destruction ולכפן and famine תשׂחק thou shalt laugh: ומחית of the beasts הארץ of the earth. אל neither תירא׃ shalt thou be afraid}%
\verse{כי For עם with אבני the stones השׂדה of the field: בריתך thou shalt be in league וחית and the beasts השׂדה of the field השׁלמה׃ shall be at peace}%
\verse{וידעת And thou shalt know כי שׁלום in peace; אהלך that thy tabernacle ופקדת and thou shalt visit נוך thy habitation, ולא and shalt not sin. תחטא׃ and shalt not sin.}%
\verse{וידעת Thou shalt know כי also that רב great, זרעך thy seed וצאצאיך and thine offspring כעשׂב as the grass הארץ׃ of the earth.}%
\verse{תבוא Thou shalt come בכלח in a full age, אלי to קבר grave כעלות cometh in גדישׁ like as a shock of corn בעתו׃ in his season.}%
\verse{הנה Lo זאת this, חקרנוה we have searched כן it, so היא it שׁמענה hear ואתה thou דע׃ it, and know}%
\end{biblechapter}%
\begin{biblechapter}% Job 6
\verseWithHeading{Job’s Second Speech: A Response to Eliphaz}{ויען answered איוב But Job ויאמר׃ and said,}%
\verse{לו Oh that שׁקול were throughly weighed, ישׁקל were throughly weighed, כעשׂי my grief והיתי במאזנים in the balances ישׂאו laid יחד׃ together!}%
\verse{כי For עתה now מחול than the sand ימים of the sea: יכבד it would be heavier על therefore כן therefore דברי my words לעו׃ are swallowed up.}%
\verse{כי For חצי the arrows שׁדי of the Almighty עמדי within אשׁר whereof חמתם me, the poison שׁתה drinketh up רוחי my spirit: בעותי the terrors אלוה of God יערכוני׃ do set themselves in array against}%
\verse{הינהק bray פרא Doth the wild ass עלי over דשׁא when he hath grass? אם or יגעה loweth שׁור the ox על בלילו׃ his fodder?}%
\verse{היאכל be eaten תפל Can that which is unsavory מבלי without מלח salt? אם or ישׁ is there טעם taste בריר in the white חלמות׃ of an egg?}%
\verse{מאנה refused לנגוע to touch נפשׁי my soul המה The things כדוי as my sorrowful לחמי׃ meat.}%
\verse{מי Oh that יתן I might תבוא have שׁאלתי my request; ותקותי the thing that I long for! יתן would grant אלוה׃ and that God}%
\verse{ויאל Even that it would please אלוה God וידכאני to destroy יתר me; that he would let loose ידו his hand, ויבצעני׃ and cut me off!}%
\verse{ותהי have עוד Then should I yet נחמתי comfort; ואסלדה yea, I would harden בחילה myself in sorrow: לא let him not יחמול spare; כי for לא I have not כחדתי concealed אמרי the words קדושׁ׃ of the Holy One.}%
\verse{מה What כחי my strength, כי that איחל I should hope? ומה and what קצי mine end, כי that אאריך I should prolong נפשׁי׃ my life?}%
\verse{אם כח my strength אבנים of stones? כחי the strength אם בשׂרי or my flesh נחושׁ׃ of brass?}%
\verse{האם אין not עזרתי my help בי ותשׁיה in me? and is wisdom נדחה driven ממני׃ quite from}%
\verse{למס To him that is afflicted מרעהו from his friend; חסד pity ויראת the fear שׁדי of the Almighty. יעזוב׃ but he forsaketh}%
\verse{אחי My brethren בגדו have dealt deceitfully כמו as נחל a brook, כאפיק as the stream נחלים of brooks יעברו׃ they pass away;}%
\verse{הקדרים Which are blackish מני by reason of קרח the ice, עלימו wherein יתעלם is hid: שׁלג׃ the snow}%
\verse{בעת What time יזרבו they wax warm, נצמתו they vanish: בחמו נדעכו they are consumed ממקומם׃ out of their place.}%
\verse{ילפתו are turned aside; ארחות The paths דרכם of their way יעלו they go בתהו to nothing, ויאבדו׃ and perish.}%
\verse{הביטו looked, ארחות The troops תמא of Tema הליכת the companies שׁבא of Sheba קוו׃ waited}%
\verse{בשׁו They were confounded כי because בטח they had hoped; באו they came עדיה thither, ויחפרו׃ and were ashamed.}%
\verse{כי For עתה now הייתם ye are לא nothing; תראו ye see חתת casting down, ותיראו׃ and are afraid.}%
\verse{הכי Did אמרתי I say, הבו Bring לי ומכחכם me of your substance? שׁחדו unto me? or, Give a reward בעדי׃ for}%
\verse{ומלטוני Or, Deliver מיד me from the hand צר ומידריצים of the mighty? תפדוני׃ or, Redeem}%
\verse{הורוני Teach ואני me, and I אחרישׁ will hold my tongue: ומה wherein שׁגיתי I have erred. הבינו׃ and cause me to understand}%
\verse{מה How נמרצו forcible אמרי words! ישׁר are right ומה but what יוכיח doth your arguing הוכח reprove? מכם׃ reprove?}%
\verse{הלהוכח to reprove מלים words, תחשׁבו Do ye imagine ולרוח as wind? אמרי and the speeches נאשׁ׃ of one that is desperate,}%
\verse{אף Yea, על ye overwhelm יתום the fatherless, תפילו ye overwhelm ותכרו and ye dig על for ריעכם׃ your friend.}%
\verse{ועתה Now הואילו therefore be content, פנו look בי ועל upon me; for evident unto פניכם upon me; for evident unto אם you if אכזב׃ I lie.}%
\verse{שׁבו Return, נא I pray you, אל let it not תהי be עולה iniquity; ושׁבי yea, return עוד again, צדקי׃ my righteousness}%
\verse{הישׁ Is there בלשׁוני in my tongue? עולה iniquity אם חכי my taste לא cannot יבין discern הוות׃ perverse things?}%
\end{biblechapter}%
\begin{biblechapter}% Job 7
\verseWithHeading{Job’s Second Speech: A Response to Eliphaz}{הלא not צבא an appointed time לאנושׁ to man על upon ארץ וכימי his days שׂכיר of a hireling? ימיו׃ also like the days}%
\verse{כעבד As a servant ישׁאף earnestly desireth צל the shadow, וכשׂכיר and as a hireling יקוה looketh for פעלו׃ his work:}%
\verse{כן So הנחלתי am I made to possess לי ירחי months שׁוא of vanity, ולילות nights עמל and wearisome מנו׃ are appointed}%
\verse{אם When שׁכבתי I lie down, ואמרתי I say, מתי When אקום shall I arise, ומדד be gone? ערב and the night ושׂבעתי and I am full נדדים of tossings to and fro עדי unto נשׁף׃ the dawning of the day.}%
\verse{לבשׁ is clothed בשׂרי My flesh רמה with worms וגישׁ and clods עפר of dust; עורי my skin רגע is broken, וימאס׃ and become loathsome.}%
\verse{ימי My days קלו are swifter מני than ארג a weaver's shuttle, ויכלו and are spent באפס without תקוה׃ hope.}%
\verse{זכר O remember כי that רוח wind: חיי my life לא shall no תשׁוב more עיני mine eye לראות see טוב׃ good.}%
\verse{לא me no תשׁורני me shall see עין The eye ראי of him that hath seen עיניך thine eyes בי ואינני׃ upon me, and I not.}%
\verse{כלה is consumed ענן the cloud וילך and vanisheth away: כן so יורד he that goeth down שׁאול to the grave לא no יעלה׃ shall come up}%
\verse{לא no ישׁוב He shall return עוד more לביתו to his house, ולא neither יכירנו know עוד him any more. מקמו׃ shall his place}%
\verse{גם Therefore אני I לא will not אחשׂך refrain פי my mouth; אדברה I will speak בצר in the anguish רוחי of my spirit; אשׂיחה I will complain במר in the bitterness נפשׁי׃ of my soul.}%
\verse{הים a sea, אני I אם or תנין a whale, כי that תשׂים thou settest עלי over משׁמר׃ a watch}%
\verse{כי When אמרתי I say, תנחמני shall comfort ערשׂי My bed ישׂא shall ease בשׂיחי משׁכבי׃ me, my couch}%
\verse{וחתתני Then thou scarest בחלמות me with dreams, ומחזינות me through visions: תבעתני׃ and terrifiest}%
\verse{ותבחר chooseth מחנק strangling, נפשׁי So that my soul מות death מעצמותי׃ rather than my life.}%
\verse{מאסתי I loathe לא I would not לעלם always: אחיה live חדל let me alone; ממני let me alone; כי for הבל vanity. ימי׃ my days}%
\verse{מה What אנושׁ man, כי that תגדלנו thou shouldest magnify וכי him? and that תשׁית thou shouldest set אליו upon לבך׃ thine heart}%
\verse{ותפקדנו And thou shouldest visit לבקרים him every morning, לרגעים him every moment? תבחננו׃ try}%
\verse{כמה How long לא wilt thou not תשׁעה depart ממני from לא me, nor תרפני let me alone עד till בלעי I swallow down רקי׃ my spittle?}%
\verse{חטאתי I have sinned; מה what אפעל shall I do לך נצר unto thee, O thou preserver האדם of men? למה why שׂמתני hast thou set למפגע me as a mark לך ואהיה עלי to למשׂא׃ a burden}%
\verse{ומה And why לא dost thou not תשׂא pardon פשׁעי my transgression, ותעביר and take away את עוני mine iniquity? כי for עתה now לעפר in the dust; אשׁכב shall I sleep ושׁחרתני and thou shalt seek me in the morning, ואינני׃ but I not}%
\end{biblechapter}%
\begin{biblechapter}% Job 8
\verseWithHeading{Bildad’s First Response to Job}{ויען Then answered בלדד Bildad השׁוחי the Shuhite, ויאמר׃ and said,}%
\verse{עד אןמלל wilt thou speak אלה these ורוח wind? כביר a strong אמרי and the words פיך׃ of thy mouth}%
\verse{האל Doth God יעות pervert משׁפט judgment? ואם or שׁדי doth the Almighty יעות pervert צדק׃ justice?}%
\verse{אם If בניך thy children חטאו have sinned לו וישׁלחם against him, and he have cast them away ביד for פשׁעם׃ their transgression;}%
\verse{אם If אתה thou תשׁחר אל to אל ואלׁדי the Almighty; תתחנן׃ and make thy supplication}%
\verse{אם If זך pure וישׁר and upright; אתה thou כי surely עתה now יעיר he would awake עליך for ושׁלם prosperous. נות thee, and make the habitation צדקך׃ of thy righteousness}%
\verse{והיה was ראשׁיתך Though thy beginning מצער small, ואחריתך yet thy latter end ישׂגה increase. מאד׃ should greatly}%
\verse{כי For שׁאל inquire, נא I pray thee, לדר age, רישׁון of the former וכונן and prepare לחקר thyself to the search אבותם׃ of their fathers:}%
\verse{כי (For תמול yesterday, אנחנו we ולא nothing, נדע and know כי because צל a shadow:) ימינו our days עלי upon ארץ׃ earth}%
\verse{הלא Shall not הם they יורוך teach יאמרו thee, tell לך ומלבם out of their heart? יוצאו thee, and utter מלים׃ words}%
\verse{היגאה grow up גמא Can the rush בלא without בצה mire? ישׂגה grow אחו can the flag בלי without מים׃ water?}%
\verse{עדנו Whilst it yet באבו in his greenness, לא not יקטף cut down, ולפני before כל any חציר herb. ייבשׁ׃ it withereth}%
\verse{כן So ארחות the paths כל of all שׁכחי that forget אל God; ותקות hope חנף and the hypocrite's תאבד׃ shall perish:}%
\verse{אשׁר Whose יקוט shall be cut off, כסלו hope ובית web. עכבישׁ a spider's מבטחו׃ and whose trust}%
\verse{ישׁען He shall lean על upon ביתו his house, ולא but it shall not יעמד stand: יחזיק he shall hold it fast, בו ולא but it shall not יקום׃ endure.}%
\verse{רטב green הוא He לפני before שׁמשׁ the sun, ועל in גנתו his garden. ינקתו and his branch תצא׃ shooteth forth}%
\verse{על about גל the heap, שׁרשׁיו His roots יסבכו are wrapped בית the place אבנים of stones. יחזה׃ seeth}%
\verse{אם If יבלענו he destroy ממקומו him from his place, וכחשׁ then shall deny בו לא him, I have not ראיתיך׃ seen}%
\verse{הן Behold, הוא this משׂושׂ the joy דרכו of his way, ומעפר and out of the earth אחר shall others יצמחו׃ grow.}%
\verse{הן Behold, אל God לא will not ימאס cast away תם a perfect ולא neither יחזיק will he help ביד will he help מרעים׃ the evildoers:}%
\verse{עד Till ימלה he fill שׂחוק with laughing, פיך thy mouth ושׂפתיך and thy lips תרועה׃ with rejoicing.}%
\verse{שׂנאיך They that hate ילבשׁו thee shall be clothed בשׁת with shame; ואהל and the dwelling place רשׁעים of the wicked איננו׃ shall come to naught.}%
\end{biblechapter}%
\begin{biblechapter}% Job 9
\verseWithHeading{Job’s Third Speech: A Response to Bildad}{ויען answered איוב Then Job ויאמר׃ and said,}%
\verse{אמנם of a truth: ידעתי I know כי so כן so ומה but how יצדק be just אנושׁ should man עם with אל׃ God?}%
\verse{אם If יחפץ he will לריב contend עמו with לא him, he cannot יעננו answer אחת him one מני of אלף׃ a thousand.}%
\verse{חכם wise לבב in heart, ואמיץ and mighty כח in strength: מי who הקשׁה hath hardened אליו against וישׁלם׃ him, and hath prospered?}%
\verse{המעתיק Which removeth הרים the mountains, ולא not: ידעו and they know אשׁר which הפכם overturneth באפו׃ them in his anger.}%
\verse{המרגיז Which shaketh ארץ the earth ממקומה out of her place, ועמודיה and the pillars יתפלצון׃ thereof tremble.}%
\verse{האמר Which commandeth לחרס the sun, ולא not; יזרח and it riseth ובעד כוכבים the stars. יחתם׃ and sealeth up}%
\verse{נטה spreadeth out שׁמים the heavens, לבדו Which alone ודורך and treadeth על upon במתי the waves ים׃ of the sea.}%
\verse{עשׂה Which maketh עשׁ Arcturus, כסיל Orion, וכימה and Pleiades, וחדרי and the chambers תמן׃ of the south.}%
\verse{עשׂה Which doeth גדלות great things עד past אין past חקר finding out; ונפלאות yea, and wonders עד without אין without מספר׃ number.}%
\verse{הן Lo, יעבר he goeth עלי by ולא not: אראה me, and I see ויחלף he passeth on ולא him not. אבין׃ also, but I perceive}%
\verse{הן Behold, יחתף he taketh away, מי who ישׁיבנו can hinder מי him? who יאמר will say אליו unto מה him, What תעשׂה׃ doest}%
\verse{אלוה God לא will not ישׁיב withdraw אפו his anger, תחתו under שׁחחו do stoop עזרי helpers רהב׃}%
\verse{אף כינכי shall I אעננו answer אבחרה him, choose out דברי my words עמו׃ with}%
\verse{אשׁר Whom, אם though צדקתי I were righteous, לא would I not אענה answer, למשׁפטי to my judge. אתחנן׃ I would make supplication}%
\verse{אם If קראתי I had called, ויענני and he had answered לא me; would I not אאמין believe כי that יאזין he had hearkened קולי׃ unto my voice.}%
\verse{אשׁר For בשׂערה me with a tempest, ישׁופני he breaketh והרבה and multiplieth פצעי my wounds חנם׃ without cause.}%
\verse{לא He will not יתנני suffer השׁב me to take רוחי my breath, כי but ישׂבעני filleth ממררים׃ me with bitterness.}%
\verse{אם If לכח of strength, אמיץ strong: הנה lo, ואם and if למשׁפט of judgment, מי who יועידני׃ shall set me a time}%
\verse{אם If אצדק I justify פי myself, mine own mouth ירשׁיעני shall condemn תם perfect, אני me: I ויעקשׁני׃ it shall also prove me perverse.}%
\verse{תם perfect, אני I לא would I not אדע know נפשׁי my soul: אמאס I would despise חיי׃ my life.}%
\verse{אחת one היא This על therefore כן , therefore אמרתי I said תם the perfect ורשׁע and the wicked. הוא He מכלה׃ destroyeth}%
\verse{אם If שׁוט the scourge ימית slay פתאם suddenly, למסת at the trial נקים of the innocent. ילעג׃ he will laugh}%
\verse{ארץ The earth נתנה is given ביד into the hand רשׁע of the wicked: פני the faces שׁפטיה of the judges יכסה he covereth אם thereof; if לא not, אפוא where, מי who הוא׃ he?}%
\verse{וימי Now my days קלו are swifter מני than רץ a post: ברחו they flee away, לא no ראו they see טובה׃ good.}%
\verse{חלפו They are passed away עם as אניות ships: אבה the swift כנשׁר as the eagle יטושׂ hasteth עלי to אכל׃ the prey.}%
\verse{אם If אמרי I say, אשׁכחה I will forget שׂיחי my complaint, אעזבה I will leave off פני my heaviness, ואבליגה׃ and comfort}%
\verse{יגרתי I am afraid כל of all עצבתי my sorrows, ידעתי I know כי that לא thou wilt not תנקני׃ hold me innocent.}%
\verse{אנכי I ארשׁע be wicked, למה why זה then הבל I in vain? איגע׃ labor}%
\verse{אם If התרחצתי I wash myself במו שׁלג water, והזכותי so clean; בבר never כפי׃ and make my hands}%
\verse{אז Yet בשׁחת me in the ditch, תטבלני shalt thou plunge ותעבוני shall abhor שׂלמותי׃ and mine own clothes}%
\verse{כי For לא not אישׁ a man, כמני as I אעננו I should answer נבוא him, we should come יחדו together במשׁפט׃ in judgment.}%
\verse{לא ישׁ is there בינינו between מוכיח any daysman ישׁת us, might lay ידו his hand על upon שׁנינו׃ us both.}%
\verse{יסר מעלי from שׁבטו ואמתו his fear אל me, and let not תבעתני׃ terrify}%
\verse{אדברה would I speak, ולא and not איראנו fear כי him; but לא not כן so אנכי עמדי׃ with}%
\end{biblechapter}%
\begin{biblechapter}% Job 10
\verseWithHeading{Job Continues His Response to Bildad}{נקטה is weary נפשׁי My soul בחיי of my life; אעזבה I will leave עלי upon שׂיחי my complaint אדברה myself; I will speak במר in the bitterness נפשׁי׃ of my soul.}%
\verse{אמר I will say אל unto אלוה God, אל Do not תרשׁיעני condemn הודיעני me; show על me wherefore מה me wherefore תריבני׃ thou contendest}%
\verse{הטוב good לך כי unto thee that תעשׁק thou shouldest oppress, כי that תמאס thou shouldest despise יגיע the work כפיך of thine hands, ועל upon עצת the counsel רשׁעים of the wicked? הופעת׃ and shine}%
\verse{העיני Hast thou eyes בשׂר of flesh? לך אם or כראות seest אנושׁ thou as man תראה׃ seeth?}%
\verse{הכימי thy days אנושׁ of man? ימיך as the days אם שׁנותיך thy years כימי days, גבר׃ as man's}%
\verse{כי That תבקשׁ thou inquirest לעוני after mine iniquity, ולחטאתי after my sin? תדרושׁ׃ and searchest}%
\verse{על דעתך Thou knowest כי that לא I am not ארשׁע wicked; ואין and none מידך out of thine hand. מציל׃ that can deliver}%
\verse{ידיך Thine hands עצבוני have made ויעשׂוני me and fashioned יחד me together סביב round about; ותבלעני׃ yet thou dost destroy}%
\verse{זכר Remember, נא I beseech thee, כי that כחמר me as the clay; עשׂיתני thou hast made ואל עפרשׁיבני׃ and wilt thou bring me into dust again?}%
\verse{הלא Hast thou not כחלב as milk, תתיכני poured me out וכגבנה me like cheese? תקפיאני׃ and curdled}%
\verse{עור me with skin ובשׂר and flesh, תלבישׁני Thou hast clothed ובעצמות me with bones וגידים and sinews. תסככני׃}%
\verse{חיים me life וחסד and favor, עשׂית עמדיפקדתך and thy visitation שׁמרה hath preserved רוחי׃ my spirit.}%
\verse{ואלה And these צפנת hast thou hid בלבבך in thine heart: ידעתי I know כי that זאת this עמך׃ with}%
\verse{אם If חטאתי I sin, ושׁמרתני then thou markest ומעוני me from mine iniquity. לא me, and thou wilt not תנקני׃ acquit}%
\verse{אם If רשׁעתי I be wicked, אללי woe לי וצדקתי unto me; and I be righteous, לא will I not אשׂא lift up ראשׁי my head. שׂבע קלון of confusion; וראה therefore see עניי׃ thou mine affliction;}%
\verse{ויגאה For it increaseth. כשׁחל me as a fierce lion: תצודני Thou huntest ותשׁב and again תתפלא׃ thou showest thyself marvelous}%
\verse{תחדשׁ Thou renewest עדיך thy witnesses נגדי against ותרב me, and increasest כעשׂך thine indignation עמדי חליפות me; changes וצבא and war עמי׃ upon}%
\verse{ולמה Wherefore מרחם out of the womb? הצאתני then hast thou brought me forth אגוע Oh that I had given up the ghost, ועין eye לא and no תראני׃ had seen}%
\verse{כאשׁר as though לא I had not הייתי I should have been אהיה been; מבטן from the womb לקבר to the grave. אובל׃ I should have been carried}%
\verse{הלא not מעט few? ימי my days יחדל cease ישׁית let me alone, ממני let me alone, ואבליגה that I may take comfort מעט׃ a little,}%
\verse{בטרם Before אלך I go ולא I shall not אשׁוב return, אל to ארץ the land חשׁך of darkness וצלמות׃ and the shadow of death;}%
\verse{ארץ A land עיפתה of darkness, כמו as אפל darkness צלמות of the shadow of death, ולא without סדרים any order, ותפע and the light כמו as אפל׃ darkness.}%
\end{biblechapter}%
\begin{biblechapter}% Job 11
\verseWithHeading{Zophar’s First Response to Job}{ויען Then answered צפר Zophar הנעמתי the Naamathite, ויאמר׃ and said,}%
\verse{הרב the multitude דברים of words לא Should not יענה be answered? ואם and should אישׁ a man שׂפתים full of talk יצדק׃ be justified?}%
\verse{בדיך Should thy lies מתים make men יחרישׁו hold their peace? ותלעג and when thou mockest, ואין shall no מכלם׃ man make thee ashamed?}%
\verse{ותאמר For thou hast said, זך pure, לקחי My doctrine ובר clean הייתי and I am בעיניך׃ in thine eyes.}%
\verse{ואולם But מי oh that יתן would אלוה God דבר speak, ויפתח and open שׂפתיו his lips עמך׃ against}%
\verse{ויגד And that he would show לך תעלמות thee the secrets חכמה of wisdom, כי that כפלים double לתושׁיה to that which is! ודע Know כי therefore that ישׁה exacteth לך אלוה God מעונך׃ of thee than thine iniquity}%
\verse{החקר Canst thou by searching אלוה God? תמצא find out אם עד unto תכלית perfection? שׁדי the Almighty תמצא׃ canst thou find out}%
\verse{גבהי as high שׁמים as heaven; מה what תפעל canst thou do? עמקה deeper משׁאול than hell; מה what תדע׃ canst thou know?}%
\verse{ארכה thereof longer מארץ than the earth, מדה The measure ורחבה and broader מני than the earth, ים׃ the sea.}%
\verse{אם If יחלף he cut off, ויסגיר and shut up, ויקהיל or gather together, ומי then who ישׁיבנו׃ can hinder}%
\verse{כי For הוא he ידע knoweth מתי men: שׁוא vain וירא he seeth און wickedness ולא also; will he not יתבונן׃ then consider}%
\verse{ואישׁ man נבוב For vain ילבב would be wise, ועיר colt. פרא a wild ass's אדם though man יולד׃ be born}%
\verse{אם If אתה thou הכינות prepare לבך thine heart, ופרשׂת and stretch out אליו toward כפך׃ thine hands}%
\verse{אם If און iniquity בידך in thine hand, הרחיקהו put it far away, ואל and let not תשׁכן dwell באהליך in thy tabernacles. עולה׃ wickedness}%
\verse{כי For אז then תשׂא shalt thou lift up פניך thy face ממום without spot; והיית yea, thou shalt be מצק steadfast, ולא and shalt not תירא׃ fear:}%
\verse{כי Because אתה thou עמל misery, תשׁכח shalt forget כמים as waters עברו pass away: תזכר׃ remember}%
\verse{ומצהרים than the noonday; יקום shall be clearer חלד And age תעפה thou shalt shine forth, כבקר as the morning. תהיה׃ thou shalt be}%
\verse{ובטחת And thou shalt be secure, כי because ישׁ there תקוה is hope; וחפרת yea, thou shalt dig לבטח in safety. תשׁכב׃ thou shalt take thy rest}%
\verse{ורבצת Also thou shalt lie down, ואין and none מחריד shall make afraid; וחלו shall make suit פניך unto רבים׃ yea, many}%
\verse{ועיני But the eyes רשׁעים of the wicked תכלינה shall fail, ומנוס and they shall not escape, אבד מנהםתקותם and their hope מפח the giving up נפשׁ׃ of the ghost.}%
\end{biblechapter}%
\begin{biblechapter}% Job 12
\verseWithHeading{Job’s Fourth Speech}{ויען answered איוב And Job ויאמר׃ and said,}%
\verse{אמנם No doubt כי but אתם ye עם the people, ועמכם with תמות shall die חכמה׃ and wisdom}%
\verse{גם But לי לבב I have understanding כמוכם as well as you; לא not נפל inferior אנכי I מכם to ואת you: yea, מי who אין knoweth not כמו such things as these? אלה׃ such things as these?}%
\verse{שׂחק one mocked לרעהו of his neighbor, אהיה I am קרא who calleth לאלוה upon God, ויענהו and he answereth שׂחוק laughed to scorn. צדיק him: the just תמים׃ upright}%
\verse{לפיד a lamp בוז despised לעשׁתות in the thought שׁאנן of him that is at ease. נכון He that is ready למועדי to slip רגל׃ with feet}%
\verse{ישׁליו prosper, אהלים The tabernacles לשׁדדים of robbers ובטחות are secure; למרגיזי and they that provoke אל God לאשׁר into whose הביא bringeth אלוה God בידו׃ hand}%
\verse{ואולם But שׁאל ask נא now בהמות the beasts, ותרך and they shall teach ועוף thee; and the fowls השׁמים of the air, ויגד׃ and they shall tell}%
\verse{או Or שׂיח speak לארץ to the earth, ותרך and it shall teach ויספרו shall declare לך דגי thee: and the fishes הים׃ of the sea}%
\verse{מי Who לא not ידע knoweth בכל in all אלה these כי that יד the hand יהוה עשׂתה hath wrought זאת׃ this?}%
\verse{אשׁר In whose בידו hand נפשׁ the soul כל of every חי living thing, ורוח and the breath כל of all בשׂר mankind. אישׁ׃ mankind.}%
\verse{הלא Doth not אזן the ear מלין words? תבחן try וחך and the mouth אכל his meat? יטעם׃ taste}%
\verse{בישׁישׁים With the ancient חכמה wisdom; וארך and in length ימים of days תבונה׃ understanding.}%
\verse{עמו With חכמה him wisdom וגבורה and strength, לו עצה he hath counsel ותבונה׃ and understanding.}%
\verse{הן Behold, יהרוס he breaketh down, ולא and it cannot יבנה be built again: יסגר he shutteth up על he shutteth up אישׁ a man, ולא and there can be no opening. יפתח׃ and there can be no opening.}%
\verse{הן Behold, יעצר he withholdeth במים the waters, ויבשׁו and they dry up: וישׁלחם also he sendeth them out, ויהפכו and they overturn ארץ׃ the earth.}%
\verse{עמו With עז him strength ותושׁיה and wisdom: לו שׁגג the deceived ומשׁגה׃ and the deceiver}%
\verse{מוליך He leadeth יועצים counselors שׁולל away spoiled, ושׁפטים and maketh the judges יהולל׃ fools.}%
\verse{מוסר the bond מלכים of kings, פתח He looseth ויאסר and girdeth אזור with a girdle. במתניהם׃ their loins}%
\verse{מוליך He leadeth כהנים princes שׁולל away spoiled, ואתנים the mighty. יסלף׃ and overthroweth}%
\verse{מסיר He removeth away שׂפה the speech לנאמנים of the trusty, וטעם the understanding זקנים of the aged. יקח׃ and taketh away}%
\verse{שׁופך He poureth בוז contempt על upon נדיבים princes, ומזיח the strength אפיקים of the mighty. רפה׃ and weakeneth}%
\verse{מגלה He discovereth עמקות deep things מני out of חשׁך darkness, ויצא and bringeth out לאור to light צלמות׃ the shadow of death.}%
\verse{משׂגיא He increaseth לגוים the nations, ויאבדם and destroyeth שׁטח them: he enlargeth לגוים the nations, וינחם׃ and straiteneth}%
\verse{מסיר He taketh away לב the heart ראשׁי of the chief עם of the people הארץ of the earth, ויתעם and causeth them to wander בתהו in a wilderness לא no דרך׃ way.}%
\verse{ימשׁשׁו They grope חשׁך in the dark ולא without אור light, ויתעם and he maketh them to stagger כשׁכור׃ like drunken}%
\end{biblechapter}%
\begin{biblechapter}% Job 13
\verseWithHeading{Job’s Fourth Speech Continues}{הן Lo, כל all ראתה hath seen עיני mine eye שׁמעה hath heard אזני mine ear ותבן׃ and understood}%
\verse{כדעתכם What ye know, ידעתי know גם also: אני do I לא not נפל inferior אנכי I מכם׃ unto}%
\verse{אולם Surely אני I אל to שׁדי the Almighty, אדבר would speak והוכח to reason אל with אל God. אחפץ׃ and I desire}%
\verse{ואולם But אתם ye טפלי forgers שׁקר of lies, רפאי physicians אלל of no value. כלכם׃ ye all}%
\verse{מי יתן ye would החרשׁ תחרישׁוןתהי and it should be לכם לחכמה׃ your wisdom.}%
\verse{שׁמעו Hear נא now תוכחתי my reasoning, ורבות to the pleadings שׂפתי of my lips. הקשׁיבו׃ and hearken}%
\verse{הלאל for God? תדברו Will ye speak עולה wickedly ולו תדברו and talk רמיה׃ deceitfully}%
\verse{הפניו his person? תשׂאון Will ye accept אם לאל for God? תריבון׃ will ye contend}%
\verse{הטוב Is it good כי that יחקר אתכםם or כהתל mocketh באנושׁ תהתלו׃ another, do ye mock}%
\verse{הוכח יוכיחתכם אם you, if בסתר ye do secretly פנים persons. תשׂאון׃ accept}%
\verse{הלא Shall not שׂאתו his excellency תבעת make you afraid? אתכם ופחדו and his dread יפל fall עליכם׃ upon}%
\verse{זכרניכם Your remembrances משׁלי אפר unto ashes, לגבי your bodies חמר of clay. גביכם׃ to bodies}%
\verse{החרישׁו Hold your peace, ממני let me alone, ואדברה may speak, אני that I ויעבר and let come עלי on מה׃ me what}%
\verse{על מהשׂא do I take בשׂרי my flesh בשׁני in my teeth, ונפשׁי my life אשׂים and put בכפי׃ in mine hand?}%
\verse{הן Though יקטלני he slay לא איחל me, yet will I trust אך in him: but דרכי mine own ways אל before פניו before אוכיח׃ I will maintain}%
\verse{גם also הוא He לי לישׁועה my salvation: כי for לא shall not לפניו before חנף a hypocrite יבוא׃ come}%
\verse{שׁמעו שׁמועלתי my speech, ואחותי and my declaration באזניכם׃ with your ears.}%
\verse{הנה Behold נא now, ערכתי I have ordered משׁפט cause; ידעתי I know כי that אני I אצדק׃ shall be justified.}%
\verse{מי Who הוא he יריב will plead עמדי with כי me? for עתה now, אחרישׁ if I hold my tongue, ואגוע׃ I shall give up the ghost.}%
\verseWithHeading{Job Argues His Case with God}{אך Only שׁתים two אל תעשׂ do עמדי unto אז me: then מפניך from לא will I not אסתר׃ hide myself}%
\verse{כפך thine hand מעלי far from הרחק Withdraw ואמתך thy dread אל me: and let not תבעתני׃ make me afraid.}%
\verse{וקרא Then call ואנכי thou, and I אענה will answer: או or אדבר let me speak, והשׁיבני׃ and answer}%
\verse{כמה How לי עונות many mine iniquities וחטאות and sins? פשׁעי my transgression וחטאתי and my sin. הדיעני׃ make me to know}%
\verse{למה Wherefore פניך thou thy face, תסתיר hidest ותחשׁבני and holdest לאויב׃ me for thine enemy?}%
\verse{העלה a leaf נדף driven to and fro? תערוץ Wilt thou break ואת קשׁ stubble? יבשׁ the dry תרדף׃ and wilt thou pursue}%
\verse{כי For תכתב thou writest עלי against מררות bitter things ותורישׁני me, and makest me to possess עונות the iniquities נעורי׃ of my youth.}%
\verse{ותשׂם Thou puttest בסד also in the stocks, רגלי my feet ותשׁמור and lookest narrowly כל unto all ארחותי my paths; על upon שׁרשׁי the heels רגלי of my feet. תתחקה׃ thou settest a print}%
\verse{והוא And he, כרקב as a rotten thing, יבלה consumeth, כבגד as a garment אכלו eaten. עשׁ׃ that is moth}%
\end{biblechapter}%
\begin{biblechapter}% Job 14
\verseWithHeading{Job Continues to Argue His Case with God}{אדם Man ילוד born אשׁה of a woman קצר of few ימים days, ושׂבע רגז׃ of trouble.}%
\verse{כציץ like a flower, יצא He cometh forth וימל and is cut down: ויברח he fleeth כצל also as a shadow, ולא not. יעמוד׃ and continueth}%
\verse{אף And על upon זה such a one, פקחת dost thou open עינך thine eyes ואתי תביא and bringest במשׁפט me into judgment עמך׃ with}%
\verse{מי Who יתן can bring טהור a clean מטמא out of an unclean? לא not אחד׃ one.}%
\verse{אם Seeing חרוצים determined, ימיו his days מספר the number חדשׁיו of his months אתך with חקו his bounds עשׂית thee, thou hast appointed ולא that he cannot יעבור׃ pass;}%
\verse{שׁעה Turn מעליו from ויחדל him, that he may rest, עד till ירצה he shall accomplish, כשׂכיר as a hireling, יומו׃ his day.}%
\verse{כי For ישׁ there is לעץ of a tree, תקוה hope אם if יכרת it be cut down, ועוד again, יחליף that it will sprout וינקתו and that the tender לא branch thereof will not תחדל׃ cease.}%
\verse{אם Though יזקין thereof wax old בארץ in the earth, שׁרשׁו the root ובעפר in the ground; ימות thereof die גזעו׃ and the stock}%
\verse{מריח מים of water יפרח it will bud, ועשׂה and bring forth קציר boughs כמו like נטע׃ a plant.}%
\verse{וגבר But man ימות dieth, ויחלשׁ and wasteth away: ויגוע giveth up the ghost, אדם yea, man ואיו׃ and where}%
\verse{אזלו fail מים the waters מני from ים the sea, ונהר and the flood יחרב decayeth ויבשׁ׃ and drieth up:}%
\verse{ואישׁ So man שׁכב lieth down, ולא not: יקום and riseth עד till בלתי no more, שׁמים the heavens לא they shall not יקיצו awake, ולא nor יערו be raised משׁנתם׃ out of their sleep.}%
\verse{מי O that יתן thou wouldest בשׁאול me in the grave, תצפנני hide תסתירני that thou wouldest keep me secret, עד until שׁוב be past, אפך thy wrath תשׁית that thou wouldest appoint לי חק me a set time, ותזכרני׃ and remember}%
\verse{אם If ימות die, גבר a man היחיה shall he live כל all ימי the days צבאי of my appointed time איחל will I wait, עד till בוא come. חליפתי׃ my change}%
\verse{תקרא Thou shalt call, ואנכי and I אענך will answer למעשׂה to the work ידיך of thine hands. תכסף׃ thee: thou wilt have a desire}%
\verse{כי For עתה now צעדי my steps: תספור thou numberest לא dost thou not תשׁמור watch על over חטאתי׃ my sin?}%
\verse{חתם sealed up בצרור in a bag, פשׁעי My transgression ותטפל and thou sewest up על and thou sewest up עוני׃ mine iniquity.}%
\verse{ואולם And surely הר the mountain נופל falling יבול cometh to naught, וצור and the rock יעתק is removed ממקמו׃ out of his place.}%
\verse{אבנים the stones: שׁחקו wear מים The waters תשׁטף thou washest away ספיחיה the things which grow עפר of the dust ארץ of the earth; ותקות the hope אנושׁ of man. האבדת׃ and thou destroyest}%
\verse{תתקפהו Thou prevailest לנצח forever ויהלך against him, and he passeth: משׁנה thou changest פניו his countenance, ותשׁלחהו׃ and sendest him away.}%
\verse{יכבדו come to honor, בניו His sons ולא not; ידע and he knoweth ויצערו and they are brought low, ולא not יבין׃ but he perceiveth}%
\verse{אך But בשׂרו his flesh עליו upon יכאב him shall have pain, ונפשׁו and his soul עליו within תאבל׃ him shall mourn.}%
\end{biblechapter}%
\begin{biblechapter}% Job 15
\verseWithHeading{Eliphaz’s Second Response to Job}{ויען Then answered אליפז Eliphaz התימני the Temanite, ויאמר׃ and said,}%
\verse{החכם Should a wise יענה man utter דעת knowledge, רוח vain וימלא and fill קדים with the east wind? בטנו׃ his belly}%
\verse{הוכח Should he reason בדבר talk? לא with unprofitable יסכון with unprofitable ומלים or with speeches לא wherewith he can do no יועיל׃ good?}%
\verse{אף Yea, אתה thou תפר castest off יראה fear, ותגרע and restrainest שׂיחה prayer לפני before אל׃ God.}%
\verse{כי For יאלף uttereth עונך thine iniquity, פיך thy mouth ותבחר and thou choosest לשׁון the tongue ערומים׃ of the crafty.}%
\verse{ירשׁיעך condemneth פיך Thine own mouth ולא thee, and not אני I: ושׂפתיך yea, thine own lips יענו׃ testify}%
\verse{הראישׁון thou the first אדם man תולד was born? ולפני before גבעות the hills? חוללת׃ or wast thou made}%
\verse{הבסוד the secret אלוה of God? תשׁמע Hast thou heard ותגרע and dost thou restrain אליך to חכמה׃ wisdom}%
\verse{מה What ידעת knowest ולא not? נדע thou, that we know תבין understandest ולא not עמנו in הוא׃ thou, which}%
\verse{גם With us both שׂב the grayheaded גם and ישׁישׁ very aged men, בנו כביר much elder מאביך than thy father. ימים׃ much elder}%
\verse{המעט small ממך with תנחמות the consolations אל of God ודבר thing לאט thee? is there any secret עמך׃ with}%
\verse{מה Why יקחך carry thee away? לבך doth thine heart ומה and what ירזמון wink at, עיניך׃ do thy eyes}%
\verse{כי That תשׁיב thou turnest אל against אל God, רוחך thy spirit והצאת go out מפיך of thy mouth? מלין׃ and lettest words}%
\verse{מה What אנושׁ man, כי that יזכה he should be clean? וכי that יצדק he should be righteous? ילוד and born אשׁה׃ of a woman,}%
\verse{הן Behold, בקדשׁו in his saints; לא he putteth no trust יאמין he putteth no trust ושׁמים yea, the heavens לא are not זכו clean בעיניו׃ in his sight.}%
\verse{אף כיתעב abominable ונאלח and filthy אישׁ man, שׁתה which drinketh כמים like water? עולה׃ iniquity}%
\verse{אחוך I will show שׁמע thee, hear לי וזה me; and that חזיתי I have seen ואספרה׃ I will declare;}%
\verse{אשׁר Which חכמים wise יגידו men have told ולא and have not כחדו hid מאבותם׃ from their fathers,}%
\verse{להם לבדם Unto whom alone נתנה was given, הארץ the earth ולא and no עבר passed זר stranger בתוכם׃ among}%
\verse{כל all ימי days, רשׁע The wicked man הוא מתחולל travaileth with pain ומספר and the number שׁנים of years נצפנו is hidden לעריץ׃ to the oppressor.}%
\verse{קול sound פחדים A dreadful באזניו in his ears: בשׁלום in prosperity שׁודד the destroyer יבואנו׃ shall come upon}%
\verse{לא not יאמין He believeth שׁוב that he shall return מני out of חשׁך darkness, וצפו is waited for הוא and he אלי of חרב׃ the sword.}%
\verse{נדד wandereth abroad הוא He ללחם for bread, איה ידע he knoweth כי that נכון is ready בידו at his hand. יום the day חשׁך׃ of darkness}%
\verse{יבעתהו shall make him afraid; צר Trouble ומצוקה and anguish תתקפהו they shall prevail against כמלך him, as a king עתיד ready לכידור׃ to the battle.}%
\verse{כי For נטה he stretcheth out אל against אל God, ידו his hand ואל against שׁדי the Almighty. יתגבר׃ and strengtheneth himself}%
\verse{ירוץ He runneth אליו upon בצואר him, on neck, בעבי upon the thick גבי bosses מגניו׃ of his bucklers:}%
\verse{כי Because כסה he covereth פניו his face בחלבו with his fatness, ויעשׂ and maketh פימה collops of fat עלי on כסל׃ flanks.}%
\verse{וישׁכון And he dwelleth in ערים cities, נכחדות desolate בתים in houses which לא no ישׁבו man inhabiteth, למו אשׁר which התעתדו are ready לגלים׃ to become heaps.}%
\verse{לא He shall not יעשׁר be rich, ולא neither יקום continue, חילו shall his substance ולא neither יטה shall he prolong לארץ thereof upon the earth. מנלם׃ the perfection}%
\verse{לא He shall not יסור depart מני out of חשׁך darkness; ינקתו his branches, תיבשׁ shall dry up שׁלהבת the flame ויסור shall he go away. ברוח and by the breath פיו׃ of his mouth}%
\verse{אל Let not יאמן trust בשׁו in vanity: נתעה him that is deceived כי for שׁוא vanity תהיה shall be תמורתו׃ his recompense.}%
\verse{בלא before יומו his time, תמלא It shall be accomplished וכפתו and his branch לא shall not רעננה׃ be green.}%
\verse{יחמס He shall shake off כגפן as the vine, בסרו his unripe grape וישׁלך and shall cast off כזית as the olive. נצתו׃ his flower}%
\verse{כי For עדת the congregation חנף of hypocrites גלמוד desolate, ואשׁ and fire אכלה shall consume אהלי the tabernacles שׁחד׃ of bribery.}%
\verse{הרה They conceive עמל mischief, וילד and bring forth און vanity, ובטנם and their belly תכין prepareth מרמה׃ deceit.}%
\end{biblechapter}%
\begin{biblechapter}% Job 16
\verseWithHeading{Job’s Fifth Speech}{ויען answered איוב Then Job ויאמר׃ and said,}%
\verse{שׁמעתי I have heard כאלה such things: רבות many מנחמי comforters עמל miserable כלכם׃ ye all.}%
\verse{הקץ have an end? לדברי words רוח Shall vain או or מה what ימריצך emboldeneth כי thee that תענה׃ thou answerest?}%
\verse{גם also אנכי I ככם אדברה could speak לו as ye if ישׁ were נפשׁכם your soul תחת נפשׁיחבירה I could heap up עליכם against במלים words ואניעה you, and shake עליכם at במו mine ראשׁי׃ head}%
\verse{אאמצכם I would strengthen במו you with פי my mouth, וניד and the moving שׂפתי of my lips יחשׂך׃ should assuage}%
\verse{אם Though אדברה I speak, לא is not יחשׂך assuaged: כאבי my grief ואחדלה and I forbear, מה what מני what יהלך׃ am I eased?}%
\verse{אך But עתה now הלאני he hath made me weary: השׁמות thou hast made desolate כל all עדתי׃ my company.}%
\verse{ותקמטני And thou hast filled me with wrath, לעד a witness היה is ויקם rising up בי כחשׁי and my leanness בפני to my face. יענה׃ in me beareth witness}%
\verse{אפו in his wrath, טרף He teareth וישׂטמני who hateth חרק me: he gnasheth עלי upon בשׁניו me with his teeth; צרי mine enemy ילטושׁ sharpeneth עיניו׃ his eyes}%
\verse{פערו They have gaped עלי upon בפיהם me with their mouth; בחרפה reproachfully; הכו they have smitten לחיי me upon the cheek יחד together עלי against יתמלאון׃ they have gathered themselves}%
\verse{יסגירני hath delivered אל God אל me to עויל the ungodly, ועל into ידי the hands רשׁעים of the wicked. ירטני׃ and turned me over}%
\verse{שׁלו at ease, הייתי I was ויפרפרני but he hath broken me asunder: ואחז he hath also taken בערפי by my neck, ויפצפצני and shaken me to pieces, ויקימני and set me up לו למטרה׃ for his mark.}%
\verse{יסבו compass me round about, עלי compass me round about, רביו His archers יפלח he cleaveth כליותי my reins ולא asunder, and doth not יחמול spare; ישׁפך he poureth out לארץ upon the ground. מררתי׃ my gall}%
\verse{יפרצני He breaketh פרץ me with breach על upon פני upon פרץ breach, ירץ he runneth עלי upon כגבור׃ me like a giant.}%
\verse{שׂק sackcloth תפרתי I have sewed עלי upon גלדי my skin, ועללתי and defiled בעפר in the dust. קרני׃ my horn}%
\verse{פני My face חמרמרה is foul מני with בכי weeping, ועל and on עפעפי my eyelids צלמות׃ the shadow of death;}%
\verse{על for לא Not חמס injustice בכפי in mine hands: ותפלתי also my prayer זכה׃ pure.}%
\verse{ארץ O earth, אל not תכסי cover דמי thou my blood, ואל no יהי have מקום place. לזעקתי׃ and let my cry}%
\verse{גם Also עתה now, הנה behold, בשׁמים in heaven, עדי my witness ושׂהדי and my record במרומים׃ on high.}%
\verse{מליצי scorn רעי My friends אל unto אלוה God. דלפה poureth out עיני׃ me: mine eye}%
\verse{ויוכח O that one might plead לגבר for a man עם with אלוה God, ובן as a man אדם as a man לרעהו׃ for his neighbor!}%
\verse{כי When שׁנות years מספר a few יאתיו are come, וארח the way לא I shall not אשׁוב return. אהלך׃ then I shall go}%
\end{biblechapter}%
\begin{biblechapter}% Job 17
\verseWithHeading{Job’s Fifth Speech, Continued}{רוחי My breath חבלה is corrupt, ימי my days נזעכו are extinct, קברים׃ the graves}%
\verse{אם לא not התלים mockers עמדי with ובהמרותם in their provocation? תלן continue עיני׃ me? and doth not mine eye}%
\verse{שׂימה Lay down נא now, ערבני put me in a surety עמך with מי thee; who הוא he לידי hands יתקע׃ will strike}%
\verse{כי For לבם their heart צפנת thou hast hid משׂכל from understanding: על therefore כן therefore לא shalt thou not תרמם׃ exalt}%
\verse{לחלק flattery יגיד He that speaketh רעים to friends, ועיני even the eyes בניו of his children תכלנה׃ shall fail.}%
\verse{והצגני He hath made למשׁל me also a byword עמים of the people; ותפת as a tabret. לפנים and formerly אהיה׃ I was}%
\verse{ותכה also is dim מכעשׂ by reason of sorrow, עיני Mine eye ויצרי my members כצל as a shadow. כלם׃ and all}%
\verse{ישׁמו shall be astonished ישׁרים Upright על at זאת this, ונקי and the innocent על against חנף the hypocrite. יתערר׃ shall stir up himself}%
\verse{ויאחז also shall hold צדיק The righteous דרכו on his way, וטהר ידים hands יסיף shall be stronger and stronger. אמץ׃ shall be stronger and stronger.}%
\verse{ואולם But כלם as for you all, תשׁבו do ye return, ובאו and come נא now: ולא for I cannot אמצא find בכם חכם׃ wise}%
\verse{ימי My days עברו are past, זמתי my purposes נתקו are broken off, מורשׁי the thoughts לבבי׃ of my heart.}%
\verse{לילה the night ליום into day: ישׂימו They change אור the light קרוב short מפני because חשׁך׃ of darkness.}%
\verse{אם If אקוה I wait, שׁאול the grave ביתי mine house: בחשׁך in the darkness. רפדתי I have made יצועי׃ my bed}%
\verse{לשׁחת to corruption, קראתי I have said אבי my father: אתה Thou אמי my mother, ואחתי and my sister. לרמה׃ to the worm,}%
\verse{ואיה And where אפו now תקותי my hope? ותקותי as for my hope, מי who ישׁורנה׃ shall see}%
\verse{בדי to the bars שׁאל of the pit, תרדנה They shall go down אם when יחד together על in עפר the dust. נחת׃ rest}%
\end{biblechapter}%
\begin{biblechapter}% Job 18
\verseWithHeading{Bildad’s Second Speech}{ויען Then answered בלדד Bildad השׁחי the Shuhite, ויאמר׃ and said,}%
\verse{עד אנהשׂימון ye make קנצי an end למלין of words? תבינו mark, ואחר and afterwards נדבר׃ we will speak.}%
\verse{מדוע Wherefore נחשׁבנו are we counted כבהמה as beasts, נטמינו reputed vile בעיניכם׃ in your sight?}%
\verse{טרף He teareth נפשׁו himself באפו in his anger: הלמענך for תעזב be forsaken ארץ shall the earth ויעתק be removed צור thee? and shall the rock ממקמו׃ out of his place?}%
\verse{גם Yea, אור the light רשׁעים of the wicked ידעך shall be put out, ולא shall not יגה shine. שׁביב and the spark אשׁו׃ of his fire}%
\verse{אור The light חשׁך shall be dark באהלו in his tabernacle, ונרו and his candle עליו with ידעך׃ shall be put out}%
\verse{יצרו shall be straitened, צעדי The steps אונו of his strength ותשׁליכהו shall cast him down. עצתו׃ and his own counsel}%
\verse{כי For שׁלח he is cast ברשׁת into a net ברגליו by his own feet, ועל upon שׂבכה a snare. יתהלך׃ and he walketh}%
\verse{יאחז shall take בעקב by the heel, פח The gin יחזק shall prevail עליו against צמים׃ the robber}%
\verse{טמון laid בארץ for him in the ground, חבלו The snare ומלכדתו and a trap עלי for him in נתיב׃ the way.}%
\verse{סביב on every side, בעתהו shall make him afraid בלהות Terrors והפיצהו and shall drive לרגליו׃ him to his feet.}%
\verse{יהי shall be רעב hunger-bitten, אנו His strength ואיד and destruction נכון ready לצלעו׃ at his side.}%
\verse{יאכל It shall devour בדי the strength עורו of his skin: יאכל shall devour בדיו his strength. בכור the firstborn מות׃ of death}%
\verse{ינתק shall be rooted out מאהלו of his tabernacle, מבטחו His confidence ותצעדהו and it shall bring למלך him to the king בלהות׃ of terrors.}%
\verse{תשׁכון It shall dwell באהלו in his tabernacle, מבלי because none לו יזרה shall be scattered על upon נוהו his habitation. גפרית׃ of his: brimstone}%
\verse{מתחת beneath, שׁרשׁיו His roots יבשׁו shall be dried up וממעל and above ימל be cut off. קצירו׃ shall his branch}%
\verse{זכרו His remembrance אבד shall perish מני from ארץ the earth, ולא and he shall have no שׁם name לו על in פני the street. חוץ׃ the street.}%
\verse{יהדפהו He shall be driven מאור from light אל into חשׁך darkness, ומתבל out of the world. ינדהו׃ and chased}%
\verse{לא He shall neither נין have son לו ולא nor נכד nephew בעמו among his people, ואין nor שׂריד any remaining במגוריו׃ in his dwellings.}%
\verse{על at יומו his day, נשׁמו shall be astonished אחרנים They that come after וקדמנים as they that went before אחזו שׂער׃ were frightened.}%
\verse{אך Surely אלה such משׁכנות the dwellings עול of the wicked, וזה and this מקום the place לא not ידע knoweth אל׃ God.}%
\end{biblechapter}%
\begin{biblechapter}% Job 19
\verseWithHeading{Job’s Sixth Speech: A Response to Bildad}{ויען answered איוב Then Job ויאמר׃ and said,}%
\verse{עד אנהוגיון will ye vex נפשׁי my soul, ותדכאונני and break me in pieces במלים׃ with words?}%
\verse{זה These עשׂר ten פעמים times תכלימוני have ye reproached לא me: ye are not תבשׁו ashamed תהכרו׃ ye make yourselves strange}%
\verse{ואף And אמנם be it indeed שׁגיתי I have erred, אתי with תלין remaineth משׁוגתי׃ mine error}%
\verse{אם If אמנם indeed עלי against תגדילו ye will magnify ותוכיחו me, and plead עלי against חרפתי׃ me my reproach:}%
\verse{דעו Know אפו now כי that אלוה God עותני hath overthrown ומצודו עלי me, and hath compassed הקיף׃ me, and hath compassed}%
\verse{הן Behold, אצעק I cry out חמס of wrong, ולא but I am not אענה heard: אשׁוע I cry aloud, ואין but no משׁפט׃ judgment.}%
\verse{ארחי my way גדר He hath fenced up ולא that I cannot אעבור pass, ועל in נתיבותי my paths. חשׁך darkness ישׂים׃ and he hath set}%
\verse{כבודי my glory, מעלי הפשׁיטיסר and taken עטרת the crown ראשׁי׃ my head.}%
\verse{יתצני He hath destroyed סביב me on every side, ואלך and I am gone: ויסע hath he removed כעץ like a tree. תקותי׃ and mine hope}%
\verse{ויחר He hath also kindled עלי against אפו his wrath ויחשׁבני me, and he counteth לו כצריו׃ me unto him as his enemies.}%
\verse{יחד together, יבאו come גדודיו His troops ויסלו and raise up עלי against דרכם their way ויחנו me, and encamp סביב round about לאהלי׃ my tabernacle.}%
\verse{אחי מעלי from הרחיק וידעי me, and mine acquaintance אך are verily זרו estranged ממני׃ from}%
\verse{חדלו have failed, קרובי My kinsfolk ומידעי and my familiar friends שׁכחוני׃ have forgotten}%
\verse{גרי They that dwell ביתי in mine house, ואמהתי and my maids, לזר me for a stranger: תחשׁבני count נכרי an alien הייתי I am בעיניהם׃ in their sight.}%
\verse{לעבדי my servant, קראתי I called ולא and he gave no יענה answer; במו him with פי my mouth. אתחנן׃ I entreated}%
\verse{רוחי My breath זרה is strange לאשׁתי to my wife, וחנתי though I entreated לבני for the children's בטני׃ of mine own body.}%
\verse{גם Yea, עוילים young children מאסו despised בי אקומה me; I arose, וידברו׃ and they spoke}%
\verse{תעבוני abhorred כל All מתי friends סודי my inward וזה me: and they whom אהבתי I loved נהפכו׃ are turned}%
\verse{בעורי to my skin ובבשׂרי and to my flesh, דבקה cleaveth עצמי My bone ואתמלטה and I am escaped בעור with the skin שׁני׃ of my teeth.}%
\verse{חנני Have pity upon חנני me, have pity upon אתם me, O ye רעי my friends; כי for יד the hand אלוה of God נגעה׃ hath touched}%
\verse{למה Why תרדפני do ye persecute כמו me as אל God, ומבשׂרי with my flesh? לא and are not תשׂבעו׃ satisfied}%
\verse{מי Oh that יתן were אפו now ויכתבון written! מלי my words מי oh that יתן they were בספר in a book! ויחקו׃ printed}%
\verse{בעט pen ברזל with an iron ועפרת and lead לעד forever! בצור in the rock יחצבון׃ That they were graven}%
\verse{ואני For I ידעתי know גאלי my redeemer חי liveth, ואחרון at the latter על upon עפר the earth: יקום׃ and he shall stand}%
\verse{ואחר And after עורי my skin נקפו destroy זאת this ומבשׂרי , yet in my flesh אחזה shall I see אלוה׃ God:}%
\verse{אשׁר Whom אני I אחזה shall see לי ועיני for myself, and mine eyes ראו shall behold, ולא and not זר another; כלו be consumed כליתי my reins בחקי׃ within}%
\verse{כי But תאמרו ye should say, מה Why נרדף persecute לו ושׁרשׁ we him, seeing the root דבר of the matter נמצא׃ is found}%
\verse{גורו Be ye afraid לכם מפני of חרב the sword: כי for חמה wrath עונות the punishments חרב of the sword, למען that תדעון ye may know שׁדין׃ a judgment.}%
\end{biblechapter}%
\begin{biblechapter}% Job 20
\verseWithHeading{Zophar’s Second Speech}{ויען Then answered צפר Zophar הנעמתי the Naamathite, ויאמר׃ and said,}%
\verse{לכן Therefore שׂעפי do my thoughts ישׁיבוני cause me to answer, ובעבור and for חושׁי׃ I make haste.}%
\verse{מוסר the check כלמתי of my reproach, אשׁמע I have heard ורוח and the spirit מבינתי of my understanding יענני׃ causeth me to answer.}%
\verse{הזאת thou this ידעת Knowest מני of עד old, מני since שׂים was placed אדם man עלי upon ארץ׃ earth,}%
\verse{כי That רננת the triumphing רשׁעים of the wicked מקרוב short, ושׂמחת and the joy חנף of the hypocrite עדי for רגע׃ a moment?}%
\verse{אם Though יעלה mount up לשׁמים to the heavens, שׂיאו his excellency וראשׁו and his head לעב unto the clouds; יגיע׃ reach}%
\verse{כגללו like his own dung: לנצח forever יאבד he shall perish ראיו they which have seen יאמרו him shall say, איו׃ Where}%
\verse{כחלום as a dream, יעוף He shall fly away ולא and shall not ימצאוהו be found: וידד yea, he shall be chased away כחזיון as a vision לילה׃ of the night.}%
\verse{עין The eye שׁזפתו also saw ולא him shall no תוסיף more; ולא neither עוד any more תשׁורנו behold מקומו׃ shall his place}%
\verse{בניו His children ירצו shall seek to please דלים the poor, וידיו and his hands תשׁבנה shall restore אונו׃ their goods.}%
\verse{עצמותיו His bones מלאו are full עלומו of his youth, ועמו with על him in עפר the dust. תשׁכב׃ which shall lie down}%
\verse{אם Though תמתיק be sweet בפיו in his mouth, רעה wickedness יכחידנה he hide תחת it under לשׁונו׃ his tongue;}%
\verse{יחמל עליהלא it not; יעזבנה it, and forsake וימנענה but keep it still בתוך within חכו׃ his mouth:}%
\verse{לחמו his meat במעיו in his bowels נהפך is turned, מרורת the gall פתנים of asps בקרבו׃ within}%
\verse{חיל riches, בלע He hath swallowed down ויקאנו and he shall vomit them up again: מבטנו his belly. יורשׁנו shall cast them out אל׃ God}%
\verse{ראשׁ the poison פתנים of asps: יינק He shall suck תהרגהו shall slay לשׁון tongue אפעה׃ the viper's}%
\verse{אל He shall not ירא see בפלגות the rivers, נהרי the floods, נחלי the brooks דבשׁ of honey וחמאה׃ and butter.}%
\verse{משׁיב shall he restore, יגע That which he labored for ולא and shall not יבלע swallow down: כחיל according to substance תמורתו the restitution ולא and he shall not יעלס׃ rejoice}%
\verse{כי Because רצץ he hath oppressed עזב hath forsaken דלים the poor; בית a house גזל he hath violently taken away ולא not; יבנהו׃ which he built}%
\verse{כי Surely לא he shall not ידע feel שׁלו quietness בבטנו in his belly, בחמודו of that which he desired. לא he shall not ימלט׃ save}%
\verse{אין There shall none שׂריד be left; לאכלו of his meat על therefore כן therefore לא shall no יחיל man look טובו׃ for his goods.}%
\verse{במלאות In the fullness שׂפקו of his sufficiency יצר he shall be in straits: לו כל every יד hand עמל of the wicked תבואנו׃ shall come}%
\verse{יהי he למלא is about to fill בטנו his belly, ישׁלח shall cast בו חרון the fury אפו of his wrath וימטר upon him, and shall rain עלימו upon בלחומו׃ him while he is eating.}%
\verse{יברח He shall flee מנשׁק weapon, ברזל from the iron תחלפהו shall strike him through. קשׁת the bow נחושׁה׃ of steel}%
\verse{שׁלף It is drawn, ויצא and cometh out מגוה of the body; וברק yea, the glittering sword ממררתו out of his gall: יהלך cometh עליו upon אמים׃ terrors}%
\verse{כל All חשׁך darkness טמון hid לצפוניו in his secret places: תאכלהו shall consume אשׁ a fire לא not נפח blown ירע שׂריד with him that is left באהלו׃ in his tabernacle.}%
\verse{יגלו shall reveal שׁמים The heaven עונו his iniquity; וארץ and the earth מתקוממה׃ shall rise up}%
\verse{יגל shall depart, יבול The increase ביתו of his house נגרות shall flow away ביום in the day אפו׃ of his wrath.}%
\verse{זה This חלק the portion אדם man רשׁע of a wicked מאלהים ונחלת and the heritage אמרו appointed מאל׃}%
\end{biblechapter}%
\begin{biblechapter}% Job 21
\verseWithHeading{Job’s Seventh Speech: A Response to Zophar}{ויען answered איוב But Job ויאמר׃ and said,}%
\verse{שׁמעו שׁמועלתי my speech, ותהי be זאת and let this תנחומתיכם׃ your consolations.}%
\verse{שׂאוני Suffer ואנכי me that I אדבר may speak; ואחר and after that דברי I have spoken, תלעיג׃ mock on.}%
\verse{האנכי As for me, לאדם to man? שׂיחי my complaint ואם and if מדוע why לא should not תקצר be troubled? רוחי׃ my spirit}%
\verse{פנו אליהשׁמו me, and be astonished, ושׂימו and lay יד hand על upon פה׃ mouth.}%
\verse{ואם Even when זכרתי I remember ונבהלתי I am afraid, ואחז taketh hold בשׂרי on my flesh. פלצות׃ and trembling}%
\verse{מדוע Wherefore רשׁעים do the wicked יחיו live, עתקו become old, גם yea, גברו are mighty חיל׃ in power?}%
\verse{זרעם Their seed נכון is established לפניהם in their sight עמם with וצאצאיהם them, and their offspring לעיניהם׃ before their eyes.}%
\verse{בתיהם Their houses שׁלום safe מפחד from fear, ולא neither שׁבט the rod אלוה of God עליהם׃ upon}%
\verse{שׁורו Their bull עבר engendereth, ולא not; יגעל and faileth תפלט calveth, פרתו their cow ולא and casteth not her calf. תשׁכל׃ and casteth not her calf.}%
\verse{ישׁלחו They send forth כצאן like a flock, עויליהם their little ones וילדיהם and their children ירקדון׃ dance.}%
\verse{ישׂאו They take כתף the timbrel וכנור and harp, וישׂמחו and rejoice לקול at the sound עוגב׃ of the organ.}%
\verse{יבלו בטוב in wealth, ימיהם their days וברגע and in a moment שׁאול to the grave. יחתו׃ go down}%
\verse{ויאמרו Therefore they say לאל unto God, סור Depart ממנו from ודעת the knowledge דרכיך of thy ways. לא not חפצנו׃ us; for we desire}%
\verse{מה What שׁדי the Almighty, כי that נעבדנו we should serve ומה him? and what נועיל profit כי should we have, if נפגע׃ we pray}%
\verse{הן Lo, לא not בידם in their hand: טובם their good עצת the counsel רשׁעים of the wicked רחקה is far מני׃ from}%
\verse{כמה How oft נר is the candle רשׁעים of the wicked ידעך put out! ויבא and cometh עלימו upon אידם their destruction חבלים sorrows יחלק them! distributeth באפו׃ in his anger.}%
\verse{יהיו They are כתבן as stubble לפני before רוח the wind, וכמץ and as chaff גנבתו carrieth away. סופה׃ that the storm}%
\verse{אלוה God יצפן layeth up לבניו for his children: אונו his iniquity ישׁלם he rewardeth אליו he rewardeth וידע׃ him, and he shall know}%
\verse{יראו shall see עינו His eyes כידו his destruction, ומחמת of the wrath שׁדי of the Almighty. ישׁתה׃ and he shall drink}%
\verse{כי For מה what חפצו pleasure בביתו he in his house אחריו after ומספר him, when the number חדשׁיו of his months חצצו׃ is cut off in the midst?}%
\verse{הלאל God ילמד Shall teach דעת knowledge? והוא seeing he רמים those that are high. ישׁפוט׃ judgeth}%
\verse{זה One ימות dieth בעצם strength, תמו in his full כלו being wholly שׁלאנן at ease ושׁליו׃ and quiet.}%
\verse{עטיניו His breasts מלאו are full חלב of milk, ומח with marrow. עצמותיו and his bones ישׁקה׃ are moistened}%
\verse{וזה And another ימות dieth בנפשׁ of his soul, מרה in the bitterness ולא and never אכל eateth בטובה׃ with pleasure.}%
\verse{יחד alike על in עפר the dust, ישׁכבו They shall lie down ורמה and the worms תכסה shall cover עליהם׃ shall cover}%
\verse{הן Behold, ידעתי I know מחשׁבותיכם your thoughts, ומזמות and the devices עלי against תחמסו׃ ye wrongfully imagine}%
\verse{כי For תאמרו ye say, איה Where בית the house נדיב of the prince? ואיה and where אהל places משׁכנות the dwelling רשׁעים׃ of the wicked?}%
\verse{הלא Have ye not שׁאלתם asked עוברי them that go דרך by the way? ואתתם their tokens, לא and do ye not תנכרו׃ know}%
\verse{כי That ליום to the day איד of destruction? יחשׂך is reserved רע the wicked ליום to the day עברות of wrath. יובלו׃ they shall be brought forth}%
\verse{מי Who יגיד shall declare על to פניו his face? דרכו his way והוא him he עשׂה hath done? מי and who ישׁלם׃ shall repay}%
\verse{והוא Yet shall he לקברות to the grave, יובל be brought ועל in גדישׁ the tomb. ישׁקוד׃ and shall remain}%
\verse{מתקו shall be sweet לו רגבי The clods נחל of the valley ואחריו after כל unto him, and every אדם man ימשׁוך shall draw ולפניו before אין him, as innumerable מספר׃ him, as innumerable}%
\verse{ואיך How תנחמוני then comfort הבל ye me in vain, ותשׁובתיכם seeing in your answers נשׁאר there remaineth מעל׃ falsehood?}%
\end{biblechapter}%
\begin{biblechapter}% Job 22
\verseWithHeading{Eliphaz’s Third Speech}{ויען answered אליפז Then Eliphaz התמני the Temanite ויאמר׃ and said,}%
\verse{הלאל unto God, יסכן be profitable גבר Can a man כי as he that יסכן may be profitable עלימו unto משׂכיל׃ is wise}%
\verse{החפץ any pleasure לשׁדי to the Almighty, כי that תצדק thou art righteous? ואם or בצע gain כי that תתם thou makest thy ways perfect? דרכיך׃ thou makest thy ways perfect?}%
\verse{המיראתך thee for fear יכיחך Will he reprove יבוא of thee? will he enter עמך with במשׁפט׃ thee into judgment?}%
\verse{הלא not רעתך thy wickedness רבה great? ואין infinite? קץ infinite? לעונתיך׃ and thine iniquities}%
\verse{כי For תחבל thou hast taken a pledge אחיך from thy brother חנם for naught, ובגדי of their clothing. ערומים the naked תפשׁיט׃ and stripped}%
\verse{לא Thou hast not מים given water עיף to the weary תשׁקה to drink, ומרעב תמנע and thou hast withheld לחם׃ bread}%
\verse{ואישׁ man, זרוע But the mighty לו הארץ he had the earth; ונשׂוא and the honorable man פנים and the honorable man ישׁב׃ dwelt}%
\verse{אלמנות שׁלחתיקם empty, וזרעות and the arms יתמים of the fatherless ידכא׃ have been broken.}%
\verse{על כןביבותיך round about פחים snares ויבהלך troubleth פחד fear פתאם׃ thee, and sudden}%
\verse{או Or חשׁך darkness, לא thou canst not תראה see; ושׁפעת and abundance מים of waters תכסך׃ cover}%
\verse{הלא not אלוה God גבה in the height שׁמים of heaven? וראה and behold ראשׁ the height כוכבים of the stars, כי how רמו׃ high they are!}%
\verse{ואמרת And thou sayest, מה How ידע know? אל doth God הבעד through ערפל the dark cloud? ישׁפוט׃ can he judge}%
\verse{עבים Thick clouds סתר a covering לו ולא not; יראה to him, that he seeth וחוג in the circuit שׁמים of heaven. יתהלך׃ and he walketh}%
\verse{הארח way עולם the old תשׁמר Hast thou marked אשׁר which דרכו have trodden? מתי men און׃ wicked}%
\verse{אשׁר Which קמטו were cut down ולא out of עת time, נהר with a flood: יוצק was overflown יסודם׃ whose foundation}%
\verse{האמרים Which said לאל unto God, סור Depart ממנו from ומה us: and what יפעל do שׁדי׃ can the Almighty}%
\verse{והוא Yet he מלא filled בתיהם their houses טוב with good ועצת but the counsel רשׁעים of the wicked רחקה is far מני׃ from}%
\verse{יראו see צדיקים The righteous וישׂמחו and are glad: ונקי and the innocent ילעג׃ laugh them to scorn.}%
\verse{אם Whereas לא is not נכחד cut down, קימנו our substance ויתרם but the remnant אכלה consumeth. אשׁ׃ of them the fire}%
\verse{הסכן Acquaint נא now עמו thyself with ושׁלם him, and be at peace: בהם תבואתך shall come טובה׃ thereby good}%
\verse{קח Receive, נא I pray thee, מפיו from his mouth, תורה the law ושׂים and lay up אמריו his words בלבבך׃ in thine heart.}%
\verse{אם If תשׁוב thou return עד to שׁדי the Almighty, תבנה thou shalt be built up, תרחיק far עולה thou shalt put away iniquity מאהלך׃ from thy tabernacles.}%
\verse{ושׁית Then shalt thou lay up על as עפר dust, בצר gold ובצור as the stones נחלים of the brooks. אופיר׃ and the of Ophir}%
\verse{והיה shall be שׁדי Yea, the Almighty בצריך thy defense, וכסף of silver. תועפות׃ and thou shalt have plenty}%
\verse{כי For אז then על in שׁדי the Almighty, תתענג shalt thou have thy delight ותשׂא and shalt lift up אל unto אלוה God. פניך׃ thy face}%
\verse{תעתיר Thou shalt make thy prayer אליו unto וישׁמעך him, and he shall hear ונדריך thy vows. תשׁלם׃ thee, and thou shalt pay}%
\verse{ותגזר Thou shalt also decree אומר a thing, ויקם and it shall be established לך ועל upon דרכיך thy ways. נגה shall shine אור׃ unto thee: and the light}%
\verse{כי When השׁפילו are cast down, ותאמר then thou shalt say, גוה lifting up; ושׁח the humble person. עינים the humble person. יושׁע׃ and he shall save}%
\verse{ימלט He shall deliver אי the island נקי of the innocent: ונמלט and it is delivered בבר by the pureness כפיך׃ of thine hands.}%
\end{biblechapter}%
\begin{biblechapter}% Job 23
\verseWithHeading{Job’s Eighth Speech: A Response to Eliphaz}{ויען answered איוב Then Job ויאמר׃ and said,}%
\verse{גם Even היום today מרי bitter: שׂחי my complaint ידי my stroke כבדה is heavier על than אנחתי׃ my groaning.}%
\verse{מי יתןדעתי I knew ואמצאהו where I might find אבוא him! I might come עד to תכונתו׃ his seat!}%
\verse{אערכה I would order לפניו before משׁפט cause ופי my mouth אמלא him, and fill תוכחות׃ with arguments.}%
\verse{אדעה I would know מלים the words יענני he would answer ואבינה me, and understand מה what יאמר׃ he would say}%
\verse{הברב כח power? יריב Will he plead עמדי against לא No; אך but הוא he ישׂם׃ would put}%
\verse{שׁם There ישׁר the righteous נוכח might dispute עמו with ואפלטה him; so should I be delivered לנצח forever משׁפטי׃ from my judge.}%
\verse{הן Behold, קדם forward, אהלך I go ואיננו but he not ואחור and backward, ולא but I cannot אבין׃ perceive}%
\verse{שׂמאול On the left hand, בעשׂתו where he doth work, ולא but I cannot אחז behold יעטף he hideth ימין himself on the right hand, ולא that I cannot אראה׃ see}%
\verse{כי But ידע he knoweth דרך the way עמדי that I take: בחנני he hath tried כזהב as gold. אצא׃ me, I shall come forth}%
\verse{באשׁרו his steps, אחזה hath held רגלי My foot דרכו his way שׁמרתי have I kept, ולא and not אט׃ declined.}%
\verse{מצות from the commandment שׂפתיו of his lips; ולא Neither אמישׁ have I gone back מחקי more than my necessary צפנתי I have esteemed אמרי the words פיו׃ of his mouth}%
\verse{והוא But he באחד in one ומי and who ישׁיבנו can turn ונפשׁו him? and his soul אותה desireth, ויעשׂ׃ even he doeth.}%
\verse{כי For ישׁלים he performeth חקי appointed וכהנה such רבות for me: and many עמו׃ with}%
\verse{על כןפניו at his presence: אבהל am I troubled אתבונן when I consider, ואפחד I am afraid ממנו׃ at his presence:}%
\verse{ואל For God הרך maketh my heart soft לבי maketh my heart soft ושׁדי and the Almighty הבהילני׃ troubleth}%
\verse{כי Because לא I was not נצמתי cut off מפני before חשׁך the darkness, ומפני from my face. כסה hath he covered אפל׃ the darkness}%
\end{biblechapter}%
\begin{biblechapter}% Job 24
\verseWithHeading{Job’s Eighth Speech, Continued}{מדוע Why, משׁדי לא are not נצפנו hidden עתים seeing times וידעו do they that know לא him not חזו see ימיו׃ his days?}%
\verse{גבלות the landmarks; ישׂיגו remove עדר flocks, גזלו they violently take away וירעו׃ and feed}%
\verse{חמור the ass יתומים of the fatherless, ינהגו They drive away יחבלו שׁורלמנה׃}%
\verse{יטו They turn אביונים the needy מדרך out of the way: יחד themselves together. חבאו hide עניי ארץ׃ of the earth}%
\verse{הן Behold, פראים wild asses במדבר in the desert, יצאו go they forth בפעלם to their work; משׁחרי rising quickly לטרף for a prey: ערבה the wilderness לו לחם food לנערים׃ for them for children.}%
\verse{בשׂדה in the field: בלילו his corn יקצירו They reap וכרם the vintage רשׁע of the wicked. ילקשׁו׃ and they gather}%
\verse{ערום They cause the naked ילינו to lodge מבלי without לבושׁ clothing, ואין that no כסות covering בקרה׃ in the cold.}%
\verse{מזרם with the showers הרים of the mountains, ירטבו They are wet ומבלי for want מחסה of a shelter. חבקו and embrace צור׃ the rock}%
\verse{יגזלו They pluck משׁד from the breast, יתום the fatherless ועל עני the poor. יחבלו׃ and take a pledge}%
\verse{ערום naked הלכו They cause to go בלי without לבושׁ clothing, ורעבים נשׂאו and they take away עמר׃ the sheaf}%
\verse{בין within שׁורתם their walls, יצהירו make oil יקבים winepresses, דרכו tread ויצמאו׃ and suffer thirst.}%
\verse{מעיר from out of the city, מתים Men ינאקו groan ונפשׁ and the soul חללים of the wounded תשׁוע crieth out: ואלוה yet God לא not ישׂים layeth תפלה׃ folly}%
\verse{המה They היו are במרדי of those that rebel אור against the light; לא not הכירו they know דרכיו the ways ולא thereof, nor ישׁבו abide בנתיבתיו׃ in the paths}%
\verse{לאור with the light יקום rising רוצח The murderer יקטל killeth עני the poor ואביון and needy, ובלילה and in the night יהי is כגנב׃ as a thief.}%
\verse{ועין The eye נאף also of the adulterer שׁמרה waiteth נשׁף for the twilight, לאמר saying, לא No תשׁורני shall see עין eye וסתר me: and disguiseth פנים face. ישׂים׃ me: and disguiseth}%
\verse{חתר they dig through בחשׁך In the dark בתים houses, יומם for themselves in the daytime: חתמו they had marked למו לא not ידעו they know אור׃ the light.}%
\verse{כי For יחדו to them even as בקר the morning למו צלמות the shadow of death: כי if יכיר know בלהות the terrors צלמות׃ of the shadow of death.}%
\verse{קל swift הוא He על as פני as מים the waters; תקלל is cursed חלקתם their portion בארץ in the earth: לא not יפנה he beholdeth דרך the way כרמים׃ of the vineyards.}%
\verse{ציה Drought גם and חם heat יגזלו consume מימי waters: שׁלג the snow שׁאול the grave חטאו׃ have sinned.}%
\verse{ישׁכחהו shall forget רחם The womb מתקו shall feed sweetly רמה him; the worm עוד more לא on him; he shall be no יזכר remembered; ותשׁבר shall be broken כעץ as a tree. עולה׃ and wickedness}%
\verse{רעה He evil entreateth עקרה the barren לא not: תלד beareth ואלמנה to the widow. לא and doeth not ייטיב׃ good}%
\verse{ומשׁך He draweth אבירים also the mighty בכחו with his power: יקום he riseth up, ולא and no יאמין is sure בחיין׃ of life.}%
\verse{יתן it be given לו לבטח him in safety, וישׁען whereon he resteth; ועיניהו yet his eyes על upon דרכיהם׃ their ways.}%
\verse{רומו They are exalted מעט for a little while, ואיננו but are gone והמכו and brought low; ככל as all יקפצון they are taken out of the way וכראשׁ as the tops שׁבלת of the ears of corn. ימלו׃ and cut off}%
\verse{ואם And if לא not אפו now, מי who יכזיבני will make me a liar, וישׂם and make לאל nothing מלתי׃ my speech}%
\end{biblechapter}%
\begin{biblechapter}% Job 25
\verseWithHeading{Bildad’s Third Speech}{ויען Then answered בלדד Bildad השׁחי the Shuhite, ויאמר׃ and said,}%
\verse{המשׁל Dominion ופחד and fear עמו with עשׂה him, he maketh שׁלום peace במרומיו׃ in his high places.}%
\verse{הישׁ Is there מספר any number לגדודיו of his armies? ועל and upon מי whom לא doth not יקום arise? אורהו׃ his light}%
\verse{ומה How יצדק be justified אנושׁ then can man עם with אל God? ומה or how יזכה can he be clean ילוד born אשׁה׃ of a woman?}%
\verse{הן Behold עד even ירח to the moon, ולא not; יאהיל and it shineth וכוכבים yea, the stars לא are not pure זכו are not pure בעיניו׃ in his sight.}%
\verse{אף כינושׁ man, רמה a worm? ובן and the son אדם of man, תולעה׃ a worm?}%
\end{biblechapter}%
\begin{biblechapter}% Job 26
\verseWithHeading{Job’s Ninth Speech: A Response to Bildad}{ויען answered איוב But Job ויאמר׃ and said,}%
\verse{מה How עזרת hast thou helped ללא without כח power? הושׁעת savest זרוע thou the arm לא no עז׃ strength?}%
\verse{מה How יעצת hast thou counseled ללא no חכמה wisdom? ותושׁיה the thing לרב and hast thou plentifully הודעת׃ declared}%
\verse{את To מי whom הגדת hast thou uttered מלין words? ונשׁמת spirit מי and whose יצאה came ממך׃ from}%
\verse{הרפאים Dead יחוללו are formed מתחת from under מים the waters, ושׁכניהם׃ and the inhabitants}%
\verse{ערום naked שׁאול Hell נגדו before ואין hath no כסות covering. לאבדון׃ him, and destruction}%
\verse{נטה He stretcheth out צפון the north על over תהו the empty place, תלה hangeth ארץ the earth על upon בלי nothing. מה׃}%
\verse{צרר He bindeth up מים the waters בעביו in his thick clouds; ולא is not rent נבקע is not rent ענן and the cloud תחתם׃ under}%
\verse{מאחז He holdeth back פני the face כסה of his throne, פרשׁז spreadeth עליו upon עננו׃ his cloud}%
\verse{חק with bounds, חג עלני the waters מים the waters עד until תכלית come to an end. אור the day עם חשׁך׃ and night}%
\verse{עמודי The pillars שׁמים of heaven ירופפו tremble ויתמהו and are astonished מגערתו׃ at his reproof.}%
\verse{בכחו with his power, רגע He divideth הים the sea ובתובנתו and by his understanding מחץ he smiteth through רהב׃ the proud.}%
\verse{ברוחו By his spirit שׁמים the heavens; שׁפרה he hath garnished חללה hath formed ידו his hand נחשׁ serpent. בריח׃ the crooked}%
\verse{הן Lo, אלה these קצות parts דרכו of his ways: ומה but how שׁמץ little דבר a portion נשׁמע is heard בו ורעם of him? but the thunder גבורתו of his power מי who יתבונן׃ can understand?}%
\end{biblechapter}%
\begin{biblechapter}% Job 27
\verseWithHeading{Job Continues His Final Speech}{ויסף Moreover איוב Job שׂאת continued משׁלו his parable, ויאמר׃ and said,}%
\verse{חי liveth, אל God הסיר hath taken away משׁפטי my judgment; ושׁדי and the Almighty, המר hath vexed נפשׁי׃ my soul;}%
\verse{כי כל All עוד the while נשׁמתי my breath בי ורוח in me, and the spirit אלוה of God באפי׃ in my nostrils;}%
\verse{אם shall not תדברנה speak שׂפתי My lips עולה wickedness, ולשׁוני my tongue אם nor יהגה utter רמיה׃ deceit.}%
\verse{חלילה God forbid לי אם that אצדיק I should justify אתכם עד you: till אגוע I die לא I will not אסיר remove תמתי mine integrity ממני׃ from}%
\verse{בצדקתי My righteousness החזקתי I hold fast, ולא and will not ארפה let it go: לא shall not יחרף reproach לבבי my heart מימי׃}%
\verse{יהי be כרשׁע as the wicked, איבי Let mine enemy ומתקוממי and he that riseth up against כעול׃ me as the unrighteous.}%
\verse{כי For מה what תקות the hope חנף of the hypocrite, כי though יבצע he hath gained, כי when ישׁל taketh away אלוה God נפשׁו׃ his soul?}%
\verse{הצעקתו his cry ישׁמע hear אל Will God כי when תבוא cometh עליו upon צרה׃ trouble}%
\verse{אם על in שׁדי the Almighty? יתענג Will he delight himself יקרא call upon אלוה God? בכל will he always עת׃ will he always}%
\verse{אורה I will teach אתכם ביד you by the hand אל of God: אשׁר which עם with שׁדי the Almighty לא will I not אכחד׃ conceal.}%
\verse{הן Behold, אתם ye yourselves כלכם all חזיתם have seen ולמה why זה then are ye thus הבל altogether תהבלו׃ vain?}%
\verse{זה This חלק the portion אדם man רשׁע of a wicked עם with אל God, ונחלת and the heritage עריצים of oppressors, משׁדי יקחו׃ they shall receive}%
\verse{אם If ירבו be multiplied, בניו his children למו for חרב the sword: וצאצאיו and his offspring לא shall not ישׂבעו be satisfied לחם׃ with bread.}%
\verse{שׂרידו Those that remain במות in death: יקברו of him shall be buried ואלמנתיו and his widows לא shall not תבכינה׃ weep.}%
\verse{אם Though יצבר he heap up כעפר as the dust, כסף silver וכחמר as the clay; יכין and prepare מלבושׁ׃ raiment}%
\verse{יכין He may prepare וצדיק but the just ילבשׁ shall put on, וכסף the silver. נקי and the innocent יחלק׃ shall divide}%
\verse{בנה He buildeth כעשׁ as a moth, ביתו his house וכסכה and as a booth עשׂה maketh. נצר׃ the keeper}%
\verse{עשׁיר The rich man ישׁכב shall lie down, ולא but he shall not יאסף be gathered: עיניו his eyes, פקח he openeth ואיננו׃ and he not.}%
\verse{תשׂיגהו take hold כמים on him as waters, בלהות Terrors לילה in the night. גנבתו stealeth him away סופה׃ a tempest}%
\verse{ישׂאהו carrieth him away, קדים The east wind וילך and he departeth: וישׂערהו and as a storm hurleth ממקמו׃ him out of his place.}%
\verse{וישׁלך For shall cast עליו upon ולא him, and not יחמל spare: מידו out of his hand. ברוח he would fain flee יברח׃ he would fain flee}%
\verse{ישׂפק shall clap עלימו at כפימו their hands וישׁרק him, and shall hiss עליו him, and shall hiss ממקמו׃ him out of his place.}%
\end{biblechapter}%
\begin{biblechapter}% Job 28
\verseWithHeading{Job’s Discourse on Wisdom}{כי Surely ישׁ there is לכסף for the silver, מוצא a vein ומקום and a place לזהב for gold יזקו׃ they fine}%
\verse{ברזל Iron מעפר out of the earth, יקח is taken ואבן the stone. יצוק molten נחושׁה׃ and brass}%
\verse{קץ an end שׂם He setteth לחשׁך to darkness, ולכל all תכלית perfection: הוא and searcheth out חוקר and searcheth out אבן the stones אפל of darkness, וצלמות׃ and the shadow of death.}%
\verse{פרץ breaketh out נחל The flood מעם from גר the inhabitant; הנשׁכחים forgotten מני from רגל the foot: דלו they are dried up, מאנושׁ from men. נעו׃ they are gone away}%
\verse{ארץ the earth, ממנה out of יצא it cometh לחם bread: ותחתיה and under נהפך it is turned up כמו as it were אשׁ׃ fire.}%
\verse{מקום of it the place ספיר of sapphires: אבניה The stones ועפרת and it hath dust זהב׃ of gold.}%
\verse{נתיב a path לא which no ידעו knoweth, עיט fowl ולא hath not שׁזפתו seen: עין eye איה׃ and which the vulture's}%
\verse{לא have not הדריכהו trodden בני whelps שׁחץ The lion's לא it, nor עדה passed עליו by שׁחל׃ the fierce lion}%
\verse{בחלמישׁ upon the rock; שׁלח He putteth forth ידו his hand הפך he overturneth משׁרשׁ by the roots. הרים׃ the mountains}%
\verse{בצורות among the rocks; יארים rivers בקע He cutteth out וכל every יקר precious thing. ראתה seeth עינו׃ and his eye}%
\verse{מבכי from overflowing; נהרות the floods חבשׁ He bindeth ותעלמה and hid יצא bringeth he forth אור׃ to light.}%
\verse{והחכמה shall wisdom מאין תמצא be found? ואי and where זה and where מקום the place בינה׃ of understanding?}%
\verse{לא not ידע knoweth אנושׁ Man ערכה the price ולא thereof; neither תמצא is it found בארץ in the land החיים׃ of the living.}%
\verse{תהום The depth אמר saith, לא not בי היא It וים in me: and the sea אמר saith, אין not עמדי׃ with}%
\verse{לא It cannot יתן be gotten סגור תחתיהלא neither ישׁקל be weighed כסף shall silver מחירה׃ the price}%
\verse{לא It cannot תסלה be valued בכתם with the gold אופיר of Ophir, בשׁהם onyx, יקר with the precious וספיר׃ or the sapphire.}%
\verse{לא cannot יערכנה equal זהב The gold וזכוכית and the crystal ותמורתה it: and the exchange כלי of it jewels פז׃ of fine gold.}%
\verse{ראמות of coral, וגבישׁ or of pearls: לא No יזכר mention shall be made ומשׁך for the price חכמה of wisdom מפנינים׃ above rubies.}%
\verse{לא shall not יערכנה equal פטדת The topaz כושׁ of Ethiopia בכתם gold. טהור with pure לא it, neither תסלה׃ shall it be valued}%
\verse{והחכמה wisdom? מאין תבוא then cometh ואי and where זה and where מקום the place בינה׃ of understanding?}%
\verse{ונעלמה Seeing it is hid מעיני from the eyes כל of all חי living, ומעוף from the fowls השׁמים of the air. נסתרה׃ and kept close}%
\verse{אבדון Destruction ומות and death אמרו say, באזנינו thereof with our ears. שׁמענו We have heard שׁמעה׃ the fame}%
\verse{אלהים God הבין understandeth דרכה the way והוא thereof, and he ידע knoweth את מקומה׃ the place}%
\verse{כי For הוא he לקצות to the ends הארץ of the earth, יביט looketh תחת under כל the whole השׁמים heaven; יראה׃ seeth}%
\verse{לעשׂות To make לרוח for the winds; משׁקל the weight ומים the waters תכן and he weigheth במדה׃ by measure.}%
\verse{בעשׂתו When he made למטר for the rain, חק a decree ודרך and a way לחזיז for the lightning קלות׃ of the thunder:}%
\verse{אז Then ראה did he see ויספרה it, and declare הכינה it; he prepared וגם it, yea, חקרה׃ and searched it out.}%
\verse{ויאמר he said, לאדם And unto man הן Behold, יראת the fear אדני of the Lord, היא that חכמה wisdom; וסור and to depart מרע from evil בינה׃ understanding.}%
\end{biblechapter}%
\begin{biblechapter}% Job 29
\verseWithHeading{Job’s Final Defense}{ויסף Moreover איוב Job שׂאת continued משׁלו his parable, ויאמר׃ and said,}%
\verse{מי Oh that יתנני I were כירחי as months קדם past, כימי as the days אלוה God ישׁמרני׃ preserved}%
\verse{בהלו shined נרו When his candle עלי upon ראשׁי my head, לאורו by his light אלך I walked חשׁך׃ darkness;}%
\verse{כאשׁר As הייתי I was בימי in the days חרפי of my youth, בסוד when the secret אלוה of God עלי upon אהלי׃ my tabernacle;}%
\verse{בעוד yet שׁדי When the Almighty עמדי with סביבותי about נערי׃ me, my children}%
\verse{ברחץ When I washed הליכי my steps בחמה with butter, וצור and the rock יצוק poured me out עמדי poured me out פלגי rivers שׁמן׃ of oil;}%
\verse{בצאתי When I went out שׁער to the gate עלי through קרת the city, ברחוב in the street! אכין I prepared מושׁבי׃ my seat}%
\verse{ראוני saw נערים The young men ונחבאו me, and hid themselves: וישׁישׁים and the aged קמו arose, עמדו׃ stood up.}%
\verse{שׂרים The princes עצרו refrained במלים talking, וכף hand ישׂימו and laid לפיהם׃ on their mouth.}%
\verse{קול their peace, נגידים The nobles נחבאו held ולשׁונם and their tongue לחכם to the roof of their mouth. דבקה׃ cleaved}%
\verse{כי When אזן the ear שׁמעה heard ותאשׁרני then it blessed ועין me; and when the eye ראתה saw ותעידני׃ it gave witness}%
\verse{כי Because אמלט I delivered עני the poor משׁוע that cried, ויתום and the fatherless, ולא and none עזר׃ to help}%
\verse{ברכת The blessing אבד of him that was ready to perish עלי upon תבא came ולב heart אלמנה me: and I caused the widow's ארנן׃ to sing for joy.}%
\verse{צדק righteousness, לבשׁתי I put on וילבשׁני and it clothed כמעיל as a robe וצניף and a diadem. משׁפטי׃ me: my judgment}%
\verse{עינים eyes הייתי I was לעור to the blind, ורגלים and feet לפסח to the lame. אני׃ I}%
\verse{אב a father אנכי I לאביונים to the poor: ורב and the cause לא not ידעתי I knew אחקרהו׃ I searched out.}%
\verse{ואשׁברה And I broke מתלעות the jaws עול of the wicked, ומשׁניו out of his teeth. אשׁליך and plucked טרף׃ the spoil}%
\verse{ואמר Then I said, עם in קני my nest, אגוע I shall die וכחול as the sand. ארבה and I shall multiply ימים׃ days}%
\verse{שׁרשׁי My root פתוח spread out אלי by מים the waters, וטל and the dew ילין lay all night בקצירי׃ upon my branch.}%
\verse{כבודי My glory חדשׁ fresh עמדי in וקשׁתי me, and my bow בידי in my hand. תחליף׃ was renewed}%
\verse{לי שׁמעו Unto me gave ear, ויחלו and waited, וידמו and kept silence למו at עצתי׃ my counsel.}%
\verse{אחרי After דברי my words לא they spoke not again ישׁנו they spoke not again ועלימו upon תטף dropped מלתי׃ and my speech}%
\verse{ויחלו And they waited כמטר for me as for the rain; לי ופיהם their mouth פערו and they opened למלקושׁ׃ wide for the latter rain.}%
\verse{אשׂחק I laughed אלהם on לא not; יאמינו them, they believed ואור and the light פני of my countenance לא they cast not down. יפילון׃ they cast not down.}%
\verse{אבחר I chose out דרכם their way, ואשׁב and sat ראשׁ chief, ואשׁכון and dwelt כמלך as a king בגדוד in the army, כאשׁר as אבלים the mourners. ינחם׃ one comforteth}%
\end{biblechapter}%
\begin{biblechapter}% Job 30
\verseWithHeading{Job’s Final Defense Continued}{ועתה But now שׂחקו עליעירים younger ממני than לימים younger אשׁר whose מאסתי I would have disdained אבותם fathers לשׁית to have set עם with כלבי the dogs צאני׃ of my flock.}%
\verse{גם Yea, כח the strength ידיהם of their hands למה whereto לי עלימו me, in whom אבד was perished? כלח׃ old age}%
\verse{בחסר For want ובכפן and famine גלמוד solitary; הערקים fleeing ציה into the wilderness אמשׁ in former time שׁואה desolate ומשׁאה׃ and waste.}%
\verse{הקטפים Who cut up מלוח mallows עלי by שׂיח the bushes, ושׁרשׁ roots רתמים and juniper לחמם׃ their meat.}%
\verse{מן from גו among יגרשׁו They were driven forth יריעו (they cried עלימו after כגנב׃ them as a thief;)}%
\verse{בערוץ in the clefts נחלים of the valleys, לשׁכן To dwell חרי caves עפר of the earth, וכפים׃ and the rocks.}%
\verse{בין Among שׂיחים the bushes ינהקו they brayed; תחת under חרול the nettles יספחו׃ they were gathered together.}%
\verse{בני children נבל of fools, גם yea, בני children בלי of base men: שׁם of base men: נכאו they were viler מן than הארץ׃ the earth.}%
\verse{ועתה And now נגינתם I their song, הייתי am ואהי yea, I am להם למלה׃ their byword.}%
\verse{תעבוני They abhor רחקו me, they flee far מני from ומפני in my face. לא not חשׂכו me, and spare רק׃ to spit}%
\verse{כי Because יתרו my cord, פתח he hath loosed ויענני and afflicted ורסן the bridle מפני before שׁלחו׃ me, they have also let loose}%
\verse{על Upon ימין right פרחח the youth; יקומו rise רגלי my feet, שׁלחו they push away ויסלו and they raise up עלי against ארחות me the ways אידם׃ of their destruction.}%
\verse{נתסו They mar נתיבתי my path, להותי my calamity, יעילו they set forward לא they have no עזר׃ helper.}%
\verse{כפרץ breaking in רחב as a wide יאתיו They came תחת in שׁאה the desolation התגלגלו׃ they rolled themselves}%
\verse{ההפך are turned עלי upon בלהות Terrors תרדף me: they pursue כרוח as the wind: נדבתי my soul וכעב as a cloud. עברה passeth away ישׁעתי׃ and my welfare}%
\verse{ועתה And now עלי upon תשׁתפך is poured out נפשׁי my soul יאחזוני have taken hold upon ימי me; the days עני׃ of affliction}%
\verse{לילה me in the night season: עצמי My bones נקר are pierced מעלי in וערקי and my sinews לא take no rest. ישׁכבון׃ take no rest.}%
\verse{ברב כח force יתחפשׂ changed: לבושׁי is my garment כפי as the collar כתנתי of my coat. יאזרני׃ it bindeth me about}%
\verse{הרני He hath cast לחמר me into the mire, ואתמשׁל and I am become כעפר like dust ואפר׃ and ashes.}%
\verse{אשׁוע I cry אליך unto ולא thee, and thou dost not תענני hear עמדתי me: I stand up, ותתבנן׃ and thou regardest}%
\verse{תהפך Thou art become לאכזר cruel לי בעצם to me: with thy strong ידך hand תשׂטמני׃ thou opposest thyself against}%
\verse{תשׂאני Thou liftest me up אל to רוח the wind; תרכיבני thou causest me to ride ותמגגני and dissolvest תשׁוה׃ my substance.}%
\verse{כי For ידעתי I know מות me death, תשׁיבני thou wilt bring ובית and the house מועד appointed לכל for all חי׃ living.}%
\verse{אך Howbeit לא he will not בעי to the grave, ישׁלח stretch out יד hand אם though בפידו in his destruction. להן שׁוע׃ they cry}%
\verse{אם לא Did not בכיתי I weep לקשׁה for him that was in trouble? יום for him that was in trouble? עגמה grieved נפשׁי was my soul לאביון׃ for the poor?}%
\verse{כי When טוב good, קויתי I looked for ויבא came רע then evil ואיחלה and when I waited לאור for light, ויבא there came אפל׃ darkness.}%
\verse{מעי My bowels רתחו boiled, ולא not: דמו and rested קדמני prevented ימי the days עני׃ of affliction}%
\verse{קדר mourning הלכתי I went בלא without חמה the sun: קמתי I stood up, בקהל in the congregation. אשׁוע׃ I cried}%
\verse{אח a brother הייתי I am לתנים to dragons, ורע and a companion לבנות to owls. יענה׃ to owls.}%
\verse{עורי My skin שׁחר is black מעלי upon ועצמי me, and my bones חרה are burned מני upon חרב׃ heat.}%
\verse{ויהי also is לאבל to mourning, כנרי My harp ועגבי and my organ לקול into the voice בכים׃ of them that weep.}%
\end{biblechapter}%
\begin{biblechapter}% Job 31
\verseWithHeading{Job’s Final Defense Continued}{ברית a covenant כרתי I made לעיני with mine eyes; ומה why אתבונן then should I think על upon בתולה׃ a maid?}%
\verse{ומה For what חלק portion אלוה of God ממעל from above? ונחלת and inheritance שׁדי of the Almighty ממרמים׃ from on high?}%
\verse{הלא not איד destruction לעול to the wicked? ונכר and a strange לפעלי to the workers און׃ of iniquity?}%
\verse{הלא Doth not הוא he יראה see דרכי my ways, וכל all צעדי my steps? יספור׃ and count}%
\verse{אם If הלכתי I have walked עם with שׁוא vanity, ותחשׁ hath hasted על to מרמה deceit; רגלי׃ or if my foot}%
\verse{ישׁקלני Let me be weighed במאזני balance, צדק in an even וידע may know אלוה that God תמתי׃ mine integrity.}%
\verse{אם If תטה hath turned אשׁרי my step מני out of הדרך the way, ואחר after עיני mine eyes, הלך walked לבי and mine heart ובכפי to mine hands; דבק hath cleaved מאום׃ and if any blot}%
\verse{אזרעה let me sow, ואחר and let another יאכל eat; וצאצאי yea, let my offspring ישׁרשׁו׃ be rooted out.}%
\verse{אם If נפתה have been deceived לבי mine heart על by אשׁה a woman, ועל at פתח door; רעי my neighbor's ארבתי׃ or I have laid wait}%
\verse{תטחן grind לאחר unto another, אשׁתי let my wife ועליה upon יכרעון bow down אחרין׃ and let others}%
\verse{כי For הוא this זמה a heinous crime; והיא yea, it עון an iniquity פלילים׃ the judges.}%
\verse{כי For אשׁ a fire היא it עד to אבדון destruction, תאכל consumeth ובכל all תבואתי mine increase. תשׁרשׁ׃ and would root out}%
\verse{אם If אמאס I did despise משׁפט the cause עבדי of my manservant ואמתי or of my maidservant, ברבם when they contended עמדי׃ with}%
\verse{ומה What אעשׂה then shall I do כי when יקום riseth up? אל God וכי and when יפקד he visiteth, מה what אשׁיבנו׃ shall I answer}%
\verse{הלא Did not בבטן me in the womb עשׂני he that made עשׂהו make ויכננו fashion ברחם us in the womb? אחד׃ him? and did not one}%
\verse{אם If אמנע I have withheld מחפץ from desire, דלים the poor ועיני or have caused the eyes אלמנה of the widow אכלה׃ to fail;}%
\verse{ואכל Or have eaten פתי my morsel לבדי myself alone, ולא hath not אכל eaten יתום and the fatherless ממנה׃ thereof;}%
\verse{כי (For מנעורי from my youth גדלני he was brought up כאב with me, as a father, ומבטן אמי her from my mother's אנחנה׃ and I have guided}%
\verse{אם If אראה I have seen אובד any perish מבלי for want לבושׁ of clothing, ואין without כסות covering; לאביון׃ or any poor}%
\verse{אם If לא have not ברכוני blessed חלצו his loins ומגז with the fleece כבשׂי of my sheep; יתחמם׃ me, and he were warmed}%
\verse{אם If הניפותי I have lifted up על against יתום the fatherless, ידי my hand כי when אראה I saw בשׁער in the gate: עזרתי׃ my help}%
\verse{כתפי let mine arm משׁכמה from my shoulder blade, תפול fall ואזרעי and mine arm מקנה from the bone. תשׁבר׃ be broken}%
\verse{כי For פחד a terror אלי to איד destruction אל God ומשׂאתו me, and by reason of his highness לא not אוכל׃ I could}%
\verse{אם If שׂמתי I have made זהב gold כסלי my hope, ולכתם to the fine gold, אמרתי or have said מבטחי׃ my confidence;}%
\verse{אם If אשׂמח I rejoiced כי because רב great, חילי my wealth וכי and because כביר much; מצאה had gotten ידי׃ mine hand}%
\verse{אם If אראה I beheld אור the sun כי when יהל it shined, וירח or the moon יקר brightness; הלך׃ walking}%
\verse{ויפת enticed, בסתר hath been secretly לבי And my heart ותשׁק hath kissed ידי my hand: לפי׃ or my mouth}%
\verse{גם also הוא This עון an iniquity פלילי the judge: כי for כחשׁתי I should have denied לאל the God ממעל׃ above.}%
\verse{אם If אשׂמח I rejoiced בפיד at the destruction משׂנאי of him that hated והתעררתי me, or lifted up myself כי when מצאו found רע׃ evil}%
\verse{ולא Neither נתתי have I suffered לחטא to sin חכי my mouth לשׁאל by wishing באלה a curse נפשׁו׃ to his soul.}%
\verse{אם If לא not, אמרו said מתי the men אהלי of my tabernacle מי Oh that יתן we had מבשׂרו לא we cannot נשׂבע׃ be satisfied.}%
\verse{בחוץ in the street: לא did not ילין lodge גר The stranger דלתי my doors לארח to the traveler. אפתח׃ I opened}%
\verse{אם If כסיתי I covered כאדם as Adam, פשׁעי my transgressions לטמון by hiding בחבי in my bosom: עוני׃ mine iniquity}%
\verse{כי אערוץ Did I fear המון multitude, רבה a great ובוז or did the contempt משׁפחות of families יחתני terrify ואדם me, that I kept silence, לא went not out אצא went not out פתח׃ of the door?}%
\verse{מי Oh that יתן one would לי שׁמע hear לי הן me! behold, תוי my desire שׁדי the Almighty יענני would answer וספר a book. כתב had written אישׁ me, and mine adversary ריבי׃ me, and mine adversary}%
\verse{אם Surely לא על it upon שׁכמי my shoulder, אשׂאנו I would take אענדנו bind עטרות׃ it a crown}%
\verse{מספר unto him the number צעדי of my steps; אגידנו I would declare כמו as נגיד a prince אקרבנו׃ would I go near unto}%
\verse{אם If עלי against אדמתי my land תזעק cry ויחד likewise תלמיה me, or that the furrows יבכיון׃ thereof complain;}%
\verse{אם If כחה the fruits אכלתי I have eaten בלי thereof without כסף money, ונפשׁ their life: בעליה or have caused the owners הפחתי׃ thereof to lose}%
\verse{תחת instead חטה of wheat, יצא grow חוח Let thistles ותחת instead שׂערה of barley. באשׁה and cockle תמו are ended. דברי The words איוב׃ of Job}%
\end{biblechapter}%
\begin{biblechapter}% Job 32
\verseWithHeading{Elihu Rebukes Job and His Three Friends}{וישׁבתו ceased שׁלשׁת three האנשׁים האלה So these מענות to answer את איוב Job, כי because הוא he צדיק righteous בעיניו׃ in his own eyes.}%
\verse{ויחר Then was kindled אף the wrath אליהוא of Elihu בן the son ברכאל of Barachel הבוזי the Buzite, ממשׁפחת of the kindred רם of Ram: באיוב against Job חרה kindled, אפו was his wrath על because צדקו he justified נפשׁו himself מאלהים׃}%
\verse{ובשׁלשׁת Also against his three רעיו friends חרה kindled, אפו was his wrath על because אשׁר because לא no מצאו they had found מענה answer, וירשׁיעו and had condemned את איוב׃ Job.}%
\verse{ואליהו Now Elihu חכה had waited את איוב till Job בדברים had spoken, כי because זקנים elder המה they ממנו than לימים׃ elder}%
\verse{וירא saw אליהוא When Elihu כי that אין no מענה answer בפי in the mouth שׁלשׁת of three האנשׁים men, ויחר was kindled. אפו׃ then his wrath}%
\verse{ויען answered אליהוא And Elihu בן the son ברכאל of Barachel הבוזי the Buzite ויאמר and said, צעיר young, אני I לימים very old; ואתם and ye ישׁישׁים very old; על wherefore כן wherefore זחלתי I was afraid, ואירא and durst מחות not show דעי you mine opinion. אתכם׃}%
\verse{אמרתי I said, ימים Days ידברו should speak, ורב and multitude שׁנים of years ידיעו should teach חכמה׃ wisdom.}%
\verse{אכן But רוח a spirit היא באנושׁ in man: ונשׁמת and the inspiration שׁדי of the Almighty תבינם׃ giveth them understanding.}%
\verse{לא men are not רבים Great יחכמו wise: וזקנים neither do the aged יבינו understand משׁפט׃ judgment.}%
\verse{לכן Therefore אמרתי I said, שׁמעה Hearken לי אחוה will show דעי mine opinion. אף also אני׃ to me; I}%
\verse{הן Behold, הוחלתי I waited לדבריכם for your words; אזין I gave ear עד to תבונתיכם your reasons, עד whilst תחקרון ye searched out מלין׃ what to say.}%
\verse{ועדיכם unto אתבונן Yea, I attended והנה you, and, behold, אין none לאיוב Job, מוכיח you that convinced עונה that answered אמריו his words: מכם׃ of}%
\verse{פן Lest תאמרו ye should say, מצאנו We have found out חכמה wisdom: אל God ידפנו thrusteth him down, לא not אישׁ׃ man.}%
\verse{ולא Now he hath not ערך directed אלי against מלין words ובאמריכם him with your speeches. לא me: neither אשׁיבנו׃ will I answer}%
\verse{חתו They were amazed, לא no ענו they answered עוד more: העתיקו they left off מהם מלים׃ speaking.}%
\verse{והוחלתי When I had waited, כי (for לא not, ידברו they spoke כי but עמדו stood still, לא no ענו answered עוד׃ more;)}%
\verse{אענה will answer אף also אני I חלקי my part, אחוה will show דעי mine opinion. אף also אני׃ I}%
\verse{כי For מלתי I am full מלים of matter, הציקתני me constraineth רוח the spirit בטני׃ within}%
\verse{הנה Behold, בטני my belly כיין as wine לא hath no יפתח vent; כאבות bottles. חדשׁים like new יבקע׃ it is ready to burst}%
\verse{אדברה I will speak, וירוח that I may be refreshed: לי אפתח I will open שׂפתי my lips ואענה׃ and answer.}%
\verse{אל Let me not, נא I pray you, אשׂא accept פני person, אישׁ any man's ואל unto אדם man. לא neither אכנה׃ let me give flattering titles}%
\verse{כי For לא not ידעתי I know אכנה to give flattering titles; כמעט would soon ישׂאני take me away. עשׂני׃ my maker}%
\end{biblechapter}%
\begin{biblechapter}% Job 33
\verseWithHeading{Elihu Rebukes Job}{ואולם Wherefore, שׁמע hear נא I pray thee, איוב Job, מלי my speeches, וכל to all דברי my words. האזינה׃ and hearken}%
\verse{הנה Behold, נא now פתחתי I have opened פי my mouth, דברה hath spoken לשׁוני my tongue בחכי׃ in my mouth.}%
\verse{ישׁר the uprightness לבי of my heart: אמרי My words ודעת knowledge שׂפתי and my lips ברור clearly. מללו׃ shall utter}%
\verse{רוח The Spirit אל of God עשׂתני hath made ונשׁמת me, and the breath שׁדי of the Almighty תחיני׃ hath given me life.}%
\verse{אם If תוכל thou canst השׁיבני answer ערכה me, set n order לפני before התיצבה׃ me, stand up.}%
\verse{הן Behold, אני I כפיך according to thy wish לאל in God's מחמר out of the clay. קרצתי am formed גם also אני׃ stead: I}%
\verse{הנה Behold, אמתי my terror לא shall not תבעתך make thee afraid, ואכפי shall my hand עליך upon לא neither יכבד׃ be heavy}%
\verse{אך Surely אמרת thou hast spoken באזני in mine hearing, וקול the voice מלין of words, אשׁמע׃ and I have heard}%
\verse{זך am clean אני I בלי without פשׁע transgression, חף innocent; אנכי I ולא neither עון׃ iniquity}%
\verse{הן Behold, תנואות occasions עלי against ימצא he findeth יחשׁבני me, he counteth לאויב׃ me for his enemy,}%
\verse{ישׂם He putteth בסד in the stocks, רגלי my feet ישׁמר he marketh כל all ארחתי׃ my paths.}%
\verse{הן Behold, זאת this לא thou art not צדקת just: אענך I will answer כי thee, that ירבה is greater אלוה God מאנושׁ׃ than man.}%
\verse{מדוע Why אליו against ריבות dost thou strive כי him? for כל of any דבריו of his matters. לא he giveth not יענה׃ account}%
\verse{כי For באחת once, ידבר speaketh אל God ובשׁתים yea twice, לא it not. ישׁורנה׃ perceiveth}%
\verse{בחלום In a dream, חזיון in a vision לילה of the night, בנפל falleth תרדמה when deep sleep על upon אנשׁים men, בתנומות in slumberings עלי upon משׁכב׃ the bed;}%
\verse{אז Then יגלה he openeth אזן the ears אנשׁים ובמסרם their instruction, יחתם׃ and sealeth}%
\verse{להסיר That he may withdraw אדם man מעשׂה purpose, וגוה pride מגבר from man. יכסה׃ and hide}%
\verse{יחשׂך He keepeth back נפשׁו his soul מני from שׁחת the pit, וחיתו and his life מעבר from perishing בשׁלח׃ by the sword.}%
\verse{והוכח He is chastened במכאוב also with pain על upon משׁכבו his bed, וריב עצמיו of his bones אתן׃ with strong}%
\verse{וזהמתו abhorreth חיתו So that his life לחם bread, ונפשׁו and his soul מאכל meat. תאוה׃ dainty}%
\verse{יכל is consumed away, בשׂרו His flesh מראי that it cannot be seen; ושׁפי עצמותיו and his bones לא were not ראו׃ seen}%
\verse{ותקרב draweth near לשׁחת unto the grave, נפשׁו Yea, his soul וחיתו and his life לממתים׃ to the destroyers.}%
\verse{אם If ישׁ there be עליו with מלאך a messenger מליץ him, an interpreter, אחד one מני among אלף a thousand, להגיד to show לאדם unto man ישׁרו׃ his uprightness:}%
\verse{ויחננו Then he is gracious ויאמר unto him, and saith, פדעהו Deliver מרדת him from going down שׁחת to the pit: מצאתי I have found כפר׃ a ransom.}%
\verse{רטפשׁ shall be fresher בשׂרו His flesh מנער ישׁוב he shall return לימי to the days עלומיו׃ of his youth:}%
\verse{יעתר He shall pray אל unto אלוה God, וירצהו and he will be favorable וירא unto him: and he shall see פניו his face בתרועה with joy: וישׁב for he will render לאנושׁ unto man צדקתו׃ his righteousness.}%
\verse{ישׁר He looketh על upon אנשׁים ויאמר and say, חטאתי I have sinned, וישׁר right, העויתי and perverted ולא me not; שׁוה׃ and it profited}%
\verse{פדה He will deliver נפשׁי his soul מעבר from going בשׁחת into the pit, וחיתי and his life באור the light. תראה׃ shall see}%
\verse{הן Lo, כל all אלה these יפעל worketh אל God פעמים oftentimes שׁלושׁ oftentimes עם with גבר׃ man,}%
\verse{להשׁיב To bring back נפשׁו his soul מני from שׁחת the pit, לאור to be enlightened באור with the light החיים׃ of the living.}%
\verse{הקשׁב Mark well, איוב O Job, שׁמע hearken לי החרשׁ unto me: hold thy peace, ואנכי and I אדבר׃ will speak.}%
\verse{אם If ישׁ thou hast מלין any thing to say, השׁיבני answer דבר me: speak, כי for חפצתי I desire צדקך׃ to justify}%
\verse{אם If אין not, אתה שׁמע hearken לי החרשׁ unto me: hold thy peace, ואאלפך and I shall teach חכמה׃ thee wisdom.}%
\end{biblechapter}%
\begin{biblechapter}% Job 34
\verseWithHeading{Elihu Asserts God’s Justice}{ויען answered אליהוא Furthermore Elihu ויאמר׃ and said,}%
\verse{שׁמעו Hear חכמים O ye wise מלי my words, וידעים unto me, ye that have knowledge. האזינו׃ and give ear}%
\verse{כי For אזן the ear מלין words, תבחן trieth וחך as the mouth יטעם tasteth לאכל׃ meat.}%
\verse{משׁפט to us judgment: נבחרה Let us choose לנו נדעה let us know בינינו among מה ourselves what טוב׃ good.}%
\verse{כי For אמר hath said, איוב Job צדקתי I am righteous: ואל and God הסיר hath taken away משׁפטי׃ my judgment.}%
\verse{על against משׁפטי my right? אכזב Should I lie אנושׁ incurable חצי my wound בלי without פשׁע׃ transgression.}%
\verse{מי What גבר man כאיוב like Job, ישׁתה drinketh up לעג scorning כמים׃ like water?}%
\verse{וארח Which goeth לחברה in company עם with פעלי the workers און of iniquity, וללכת and walketh עם with אנשׁי men. רשׁע׃ wicked}%
\verse{כי For אמר he hath said, לא nothing יסכן It profiteth גבר a man ברצתו that he should delight עם himself with אלהים׃ God.}%
\verse{לכן Therefore אנשׁי unto me, ye men לבב of understanding: שׁמעו hearken לי חללה far be it לאל from God, מרשׁע wickedness; ושׁדי and the Almighty, מעול׃ iniquity.}%
\verse{כי For פעל the work אדם of a man ישׁלם shall he render לו וכארח according to ways. אישׁ unto him, and cause every man ימצאנו׃ to find}%
\verse{אף Yea, אמנם surely אל God לא will not ירשׁיע do wickedly, ושׁדי will the Almighty לא neither יעות pervert משׁפט׃ judgment.}%
\verse{מי Who פקד hath given him a charge עליו hath given him a charge ארצה over the earth? ומי or who שׂם hath disposed תבל world? כלה׃ the whole}%
\verse{אם If ישׂים he set אליו upon לבו his heart רוחו himself his spirit ונשׁמתו and his breath; אליו unto יאסף׃ man, he gather}%
\verse{יגוע shall perish כל All בשׂר flesh יחד together, ואדם and man על unto עפר dust. ישׁוב׃ shall turn again}%
\verse{ואם If בינה now understanding, שׁמעה hear זאת this: האזינה hearken לקול to the voice מלי׃ of my words.}%
\verse{האף Shall even שׂונא he that hateth משׁפט right יחבושׁ govern? ואם צדיק just? כביר him that is most תרשׁיע׃ and wilt thou condemn}%
\verse{האמר to say למלך to a king, בליעל wicked? רשׁע ungodly? אל to נדיבים׃ princes,}%
\verse{אשׁר that לא not נשׂא accepteth פני the persons שׂרים of princes, ולא nor נכר regardeth שׁוע the rich לפני more than דל the poor? כי for מעשׂה the work ידיו of his hands. כלם׃ they all}%
\verse{רגע In a moment ימתו shall they die, וחצות at midnight, לילה at midnight, יגעשׁו shall be troubled עם and the people ויעברו and pass away: ויסירו shall be taken away אביר and the mighty לא without ביד׃ hand.}%
\verse{כי For עיניו his eyes על upon דרכי the ways אישׁ of man, וכל all צעדיו his goings. יראה׃ and he seeth}%
\verse{אין no חשׁך darkness, ואין nor צלמות shadow of death, להסתר may hide themselves. שׁם where פעלי the workers און׃ of iniquity}%
\verse{כי For לא he will not על upon אישׁ man ישׂים lay עוד more להלך that he should enter אל with אל God. במשׁפט׃ into judgment}%
\verse{ירע He shall break in pieces כבירים mighty men לא without חקר number, ויעמד and set אחרים others תחתם׃ in their stead.}%
\verse{לכן Therefore יכיר he knoweth מעבדיהם their works, והפך and he overturneth לילה in the night, וידכאו׃ so that they are destroyed.}%
\verse{תחת them as רשׁעים wicked men ספקם He striketh במקום in the open ראים׃ sight}%
\verse{אשׁר עלן סרו they turned back מאחריו from וכל any דרכיו of his ways: לא him, and would not השׂכילו׃ consider}%
\verse{להביא to come עליו unto צעקת So that they cause the cry דל of the poor וצעקת the cry עניים of the afflicted. ישׁמע׃ him, and he heareth}%
\verse{והוא When he ישׁקט giveth quietness, ומי who ירשׁע then can make trouble? ויסתר and when he hideth פנים face, ומי who ישׁורנו then can behold ועל him? whether against גוי a nation, ועל or against אדם a man יחד׃ only:}%
\verse{ממלך reign not, אדם חנףמקשׁי be ensnared. עם׃ lest the people}%
\verse{כי Surely it is meet אל unto אל God, האמר to be said נשׂאתי I have borne לא I will not אחבל׃ offend}%
\verse{בלעדי not אחזה I see אתה thou הרני teach אם me: if עול iniquity, פעלתי I have done לא no אסיף׃ I will do}%
\verse{המעמך ישׁלמנה he will recompense כי it, whether מאסת thou refuse, כי or whether אתה thou תבחר choose; ולא and not אני I: ומה what ידעת thou knowest. דבר׃ therefore speak}%
\verse{אנשׁי לבב of understanding יאמרו tell לי וגבר man חכם me, and let a wise שׁמע׃ hearken}%
\verse{איוב Job לא without בדעת knowledge, ידבר hath spoken ודבריו and his words לא without בהשׂכיל׃ wisdom.}%
\verse{אבי My desire יבחן may be tried איוב Job עד unto נצח the end על because תשׁבת of answers באנשׁי men. און׃ for wicked}%
\verse{כי For יסיף he addeth על unto חטאתו his sin, פשׁע rebellion בינינו among יספוק he clappeth וירב us, and multiplieth אמריו his words לאל׃ against God.}%
\end{biblechapter}%
\begin{biblechapter}% Job 35
\verseWithHeading{Elihu Condemns Job}{ויען spoke אליהו Elihu ויאמר׃ moreover, and said,}%
\verse{הזאת thou this חשׁבת Thinkest למשׁפט to be right, אמרת thou saidst, צדקי My righteousness מאל׃}%
\verse{כי For תאמר thou saidst, מה What יסכן advantage לך מה will it be unto thee? What אעיל profit מחטאתי׃}%
\verse{אני I אשׁיבך will answer מלין will answer ואת רעיך thee, and thy companions עמך׃ with}%
\verse{הבט Look שׁמים unto the heavens, וראה and see; ושׁור and behold שׁחקים the clouds גבהו are higher ממך׃ than}%
\verse{אם If חטאת thou sinnest, מה what תפעל doest בו ורבו be multiplied, פשׁעיך thou against him? or thy transgressions מה what תעשׂה׃ doest}%
\verse{אם If צדקת thou be righteous, מה what תתן givest לו או thou him? or מה what מידך he of thine hand? יקח׃ receiveth}%
\verse{לאישׁ a man כמוך as thou רשׁעך Thy wickedness ולבן the son אדם of man. צדקתך׃ and thy righteousness}%
\verse{מרב עשׁוקים of oppressions יזעיקו they make to cry: ישׁועו they cry out מזרוע by reason of the arm רבים׃ of the mighty.}%
\verse{ולא But none אמר saith, איה אלוה God עשׂי my maker, נתן who giveth זמרות songs בלילה׃ in the night;}%
\verse{מלפנו Who teacheth מבהמות us more than the beasts ארץ of the earth, ומעוף than the fowls השׁמים of heaven? יחכמנו׃ and maketh us wiser}%
\verse{שׁם There יצעקו they cry, ולא but none יענה giveth answer, מפני because גאון of the pride רעים׃ of evil men.}%
\verse{אך Surely שׁוא vanity, לא will not ישׁמע hear אל God ושׁדי will the Almighty לא neither ישׁורנה׃ regard}%
\verse{אף כיאמר thou sayest לא thou shalt not תשׁורנו see דין him, judgment לפניו before ותחולל׃ him; therefore trust}%
\verse{ועתה But now, כי because אין not פקד he hath visited אפו in his anger; ולא not ידע yet he knoweth בפשׁ extremity: מאד׃ in great}%
\verse{ואיוב Therefore doth Job הבל in vain; יפצה open פיהו his mouth בבלי without דעת knowledge. מלין words יכבר׃ he multiplieth}%
\end{biblechapter}%
\begin{biblechapter}% Job 36
\verseWithHeading{Elihu Extols God’s Greatness}{ויסף also proceeded, אליהוא Elihu ויאמר׃ and said,}%
\verse{כתר Suffer לי זעיר me a little, ואחוך and I will show כי thee that עוד yet לאלוה on God's מלים׃ to speak}%
\verse{אשׂא I will fetch דעי my knowledge למרחוק from afar, ולפעלי to my Maker. אתן and will ascribe צדק׃ righteousness}%
\verse{כי For אמנם truly לא not שׁקר false: מלי my words תמים he that is perfect דעות in knowledge עמך׃ with}%
\verse{הן Behold, אל God כביר mighty, ולא not ימאס and despiseth כביר mighty כח in strength לב׃ wisdom.}%
\verse{לא יחיהשׁע of the wicked: ומשׁפט right עניים to the poor. יתן׃ but giveth}%
\verse{לא not יגרע He withdraweth מצדיק from the righteous: עיניו his eyes ואת but with מלכים kings לכסא on the throne; וישׁיבם yea, he doth establish לנצח them forever, ויגבהו׃ and they are exalted.}%
\verse{ואם And if אסורים bound בזקים in fetters, ילכדון be holden בחבלי in cords עני׃ of affliction;}%
\verse{ויגד Then he showeth להם פעלם them their work, ופשׁעיהם and their transgressions כי that יתגברו׃ they have exceeded.}%
\verse{ויגל He openeth אזנם also their ear למוסר to discipline, ויאמר and commandeth כי that ישׁבון they return מאון׃ from iniquity.}%
\verse{אם If ישׁמעו they obey ויעבדו and serve יכלו they shall spend ימיהם their days בטוב in prosperity, ושׁניהם and their years בנעימים׃ in pleasures.}%
\verse{ואם But if לא not, ישׁמעו they obey בשׁלח by the sword, יעברו they shall perish ויגועו and they shall die בבלי without דעת׃ knowledge.}%
\verse{וחנפי But the hypocrites לב in heart ישׂימו heap up אף wrath: לא not ישׁועו they cry כי when אסרם׃ he bindeth}%
\verse{תמת die בנער in youth, נפשׁם They וחיתם and their life בקדשׁים׃ among the unclean.}%
\verse{יחלץ He delivereth עני the poor בעניו in his affliction, ויגל and openeth בלחץ in oppression. אזנם׃ their ears}%
\verse{ואף Even so הסיתך would he have removed מפי צרחב a broad place, לא no מוצק straitness; תחתיה where ונחת and that which should be set שׁלחנך on thy table מלא full דשׁן׃ of fatness.}%
\verse{ודין the judgment רשׁע of the wicked: מלאת But thou hast fulfilled דין judgment ומשׁפט and justice יתמכו׃ take hold}%
\verse{כי Because חמה wrath, פן lest יסיתך he take thee away בספק with stroke: ורב then a great כפר ransom אל cannot יטך׃ deliver}%
\verse{היערך Will he esteem שׁועך thy riches? לא not בצר gold, וכל nor all מאמצי the forces כח׃ of strength.}%
\verse{אל not תשׁאף Desire הלילה the night, לעלות are cut off עמים when people תחתם׃ in their place.}%
\verse{השׁמר Take heed, אל תפן regard not אל און iniquity: כי for על for זה this בחרת hast thou chosen מעני׃ rather than affliction.}%
\verse{הן Behold, אל God ישׂגיב exalteth בכחו by his power: מי who כמהו like him? מורה׃ teacheth}%
\verse{מי Who פקד hath enjoined עליו hath enjoined דרכו him his way? ומי or who אמר can say, פעלת Thou hast wrought עולה׃ iniquity?}%
\verse{זכר Remember כי that תשׂגיא thou magnify פעלו his work, אשׁר which שׁררו אנשׁים׃ men}%
\verse{כל Every אדם man חזו may see בו אנושׁ it; man יביט may behold מרחוק׃ afar off.}%
\verse{הן Behold, אל God שׂגיא great, ולא not נדע and we know מספר can the number שׁניו of his years ולא neither חקר׃ be searched out.}%
\verse{כי For יגרע he maketh small נטפי the drops מים of water: יזקו they pour down מטר rain לאדו׃ according to the vapor}%
\verse{אשׁר Which יזלו do drop שׁחקים the clouds ירעפו distill עלי upon אדם man רב׃ abundantly.}%
\verse{אף אםבין can understand מפרשׂי the spreadings עב of the clouds, תשׁאות the noise סכתו׃ of his tabernacle?}%
\verse{הן Behold, פרשׂ he spreadeth עליו upon אורו his light ושׁרשׁי the bottom הים of the sea. כסה׃ it, and covereth}%
\verse{כי For בם ידין by them judgeth עמים he the people; יתן he giveth אכל meat למכביר׃ in abundance.}%
\verse{על With כפים clouds כסה he covereth אור the light; ויצו and commandeth עליה it by במפגיע׃ that cometh between.}%
\verse{יגיד thereof showeth עליו concerning רעו The noise מקנה it, the cattle אף also על concerning עולה׃ the vapor.}%
\end{biblechapter}%
\begin{biblechapter}% Job 37
\verseWithHeading{Elihu Extols God’s Majesty}{אף also לזאת At this יחרד trembleth, לבי my heart ויתר and is moved ממקומו׃ out of his place.}%
\verse{שׁמעו שׁמוערגז the noise קלו of his voice, והגה and the sound מפיו of his mouth. יצא׃ goeth out}%
\verse{תחת it under כל the whole השׁמים heaven, ישׁרהו He directeth ואורו and his lightning על unto כנפות the ends הארץ׃ of the earth.}%
\verse{אחריו After ישׁאג roareth: קול it a voice ירעם he thundereth בקול with the voice גאונו of his excellency; ולא and he will not יעקבם stay כי them when ישׁמע is heard. קולו׃ his voice}%
\verse{ירעם thundereth אל God בקולו with his voice; נפלאות marvelously עשׂה doeth גדלות great things ולא he, which we cannot נדע׃ comprehend.}%
\verse{כי For לשׁלג to the snow, יאמר he saith הוא Be ארץ thou the earth; וגשׁם rain, מטר likewise to the small וגשׁם rain מטרות and to the great עזו׃ of his strength.}%
\verse{ביד the hand כל of every אדם man; יחתום He sealeth up לדעת may know כל that all אנשׁי men מעשׂהו׃ his work.}%
\verse{ותבא go חיה Then the beasts במו into ארב dens, ובמעונתיה in their places. תשׁכן׃ and remain}%
\verse{מן Out of החדר the south תבוא cometh סופה the whirlwind: וממזרים out of the north. קרה׃ and cold}%
\verse{מנשׁמת אל of God יתן is given: קרח frost ורחב and the breadth מים of the waters במוצק׃ is straitened.}%
\verse{אף Also ברי by watering יטריח he wearieth עב the thick cloud: יפיץ he scattereth ענן cloud: אורו׃ his bright}%
\verse{והוא And it מסבות round about מתהפך is turned בתחבולתו by his counsels: לפעלם כל whatsoever אשׁר whatsoever יצום he commandeth על them upon פני the face תבל of the world ארצה׃ in the earth.}%
\verse{אם whether לשׁבט for correction, אם or לארצו for his land, אם or לחסד for mercy. ימצאהו׃ He causeth it to come,}%
\verse{האזינה Hearken זאת unto this, איוב O Job: עמד stand still, והתבונן and consider נפלאות the wondrous אל׃ works of God.}%
\verse{התדע Dost thou know בשׂום disposed אלוה when God עליהם disposed והופיע to shine? אור them, and caused the light עננו׃ of his cloud}%
\verse{התדע עלפלשׂי the balancings עב of the clouds, מפלאות the wondrous works תמים of him which is perfect דעים׃ in knowledge?}%
\verse{אשׁר How בגדיך thy garments חמים warm, בהשׁקט when he quieteth ארץ the earth מדרום׃ by the south}%
\verse{תרקיע him spread out עמו Hast thou with לשׁחקים the sky, חזקים strong, כראי looking glass? מוצק׃ as a molten}%
\verse{הודיענו Teach מה us what נאמר we shall say לו לא unto him; we cannot נערך order מפני by reason חשׁך׃ of darkness.}%
\verse{היספר Shall it be told לו כי him that אדבר I speak? אם if אמר speak, אישׁ a man כי surely יבלע׃ he shall be swallowed up.}%
\verse{ועתה And now לא not ראו see אור light בהיר the bright הוא which בשׁחקים in the clouds: ורוח but the wind עברה passeth, ותטהרם׃ and cleanseth}%
\verse{מצפון out of the north: זהב Fair weather יאתה cometh על with אלוה God נורא terrible הוד׃ majesty.}%
\verse{שׁדי the Almighty, לא we cannot מצאנהו find him out: שׂגיא excellent כח in power, ומשׁפט and in judgment, ורב and in plenty צדקה of justice: לא he will not יענה׃ afflict.}%
\verse{לכן do therefore יראוהו fear אנשׁים לא not יראה him: he respecteth כל any חכמי wise לב׃ of heart.}%
\end{biblechapter}%
\begin{biblechapter}% Job 38
\verseWithHeading{Adonai Challenges Job}{ויען answered יהוה Then the LORD את איוב Job מנהסערה the whirlwind, ויאמר׃ and said,}%
\verse{מי Who זה this מחשׁיך that darkeneth עצה counsel במלין by words בלי without דעת׃ knowledge?}%
\verse{אזר Gird up נא now כגבר like a man; חלציך thy loins ואשׁאלך for I will demand והודיעני׃ of thee, and answer}%
\verseWithHeading{Adonai Interrogates Job}{איפה Where היית wast ביסדי thou when I laid the foundations ארץ of the earth? הגד declare, אם if ידעת thou hast understanding. בינה׃ thou hast understanding.}%
\verse{מי Who שׂם hath laid ממדיה the measures כי thereof, if תדע thou knowest? או or מי who נטה hath stretched עליה upon קו׃ the line}%
\verse{על מהדניה are the foundations הטבעו thereof fastened? או or מי who ירה laid אבן stone פנתה׃ the corner}%
\verse{ברן sang יחד together, כוכבי stars בקר When the morning ויריעו shouted for joy? כל and all בני the sons אלהים׃ of God}%
\verse{ויסך Or shut up בדלתים with doors, ים the sea בגיחו when it broke forth, מרחם of the womb? יצא׃ it had issued out}%
\verse{בשׂומי When I made ענן the cloud לבשׁו the garment וערפל thereof, and thick darkness חתלתו׃ a swaddling band}%
\verse{ואשׁבר And broke up עליו for חקי it my decreed ואשׂים and set בריח bars ודלתים׃ and doors,}%
\verse{ואמר And said, עד פה and here תבוא shalt thou come, ולא but no תסיף further: ופא ישׁית be stayed? בגאון shall thy proud גליך׃ waves}%
\verse{המימיך since thy days; צוית Hast thou commanded בקר the morning ידעתה to know שׁחר מקמו׃ his place;}%
\verse{לאחז That it might take hold בכנפות of the ends הארץ of the earth, וינערו might be shaken רשׁעים that the wicked ממנה׃ out of}%
\verse{תתהפך It is turned כחמר as clay חותם the seal; ויתיצבו and they stand כמו as לבושׁ׃ a garment.}%
\verse{וימנע is withheld, מרשׁעים אורם their light וזרוע arm רמה and the high תשׁבר׃ shall be broken.}%
\verse{הבאת Hast thou entered עד into נבכי the springs ים of the sea? ובחקר in the search תהום of the depth? התהלכת׃ or hast thou walked}%
\verse{הנגלו been opened לך שׁערי Have the gates מות of death ושׁערי the doors צלמות of the shadow of death? תראה׃ unto thee? or hast thou seen}%
\verse{התבננת עדחבי the breadth ארץ of the earth? הגד declare אם if ידעת thou knowest כלה׃ it all.}%
\verse{אי where זה where הדרך the way ישׁכן dwelleth? אור light וחשׁך and darkness, אי זהקמו׃ the place}%
\verse{כי That תקחנו thou shouldest take אל it to גבולו the bound וכי thereof, and that תבין thou shouldest know נתיבות the paths ביתו׃ the house}%
\verse{ידעת Knowest כי thou because אז thou wast then תולד born? ומספר or the number ימיך of thy days רבים׃ great?}%
\verse{הבאת Hast thou entered אל into אצרות the treasures שׁלג of the snow? ואצרות the treasures ברד of the hail, תראה׃ or hast thou seen}%
\verse{אשׁר Which חשׂכתי I have reserved לעת against the time צר of trouble, ליום against the day קרב of battle ומלחמה׃ and war?}%
\verse{אי זהדרך way יחלק parted, אור is the light יפץ scattereth קדים the east wind עלי upon ארץ׃ the earth?}%
\verse{מי Who פלג hath divided לשׁטף for the overflowing of waters, תעלה a watercourse ודרך or a way לחזיז for the lightning קלות׃ of thunder;}%
\verse{להמטיר To cause it to rain על on ארץ the earth, לא no אישׁ man מדבר the wilderness, לא wherein no אדם׃ man;}%
\verse{להשׂביע To satisfy שׁאה the desolate ומשׁאה and waste ולהצמיח to spring forth? מצא and to cause the bud דשׁא׃ of the tender herb}%
\verse{הישׁ Hath למטר the rain אב a father? או or מי who הוליד hath begotten אגלי the drops טל׃ of dew?}%
\verse{מבטן מי who יצא came הקרח the ice? וכפר and the hoary frost שׁמים of heaven, מי ילדו׃ hath engendered}%
\verse{כאבן as a stone, מים The waters יתחבאו are hid ופני and the face תהום of the deep יתלכדו׃ is frozen.}%
\verse{התקשׁר Canst thou bind מעדנות the sweet influences כימה of Pleiades, או or משׁכות the bands כסיל of Orion? תפתח׃ loose}%
\verse{התציא Canst thou bring forth מזרות Mazzaroth בעתו in his season? ועישׁ Arcturus על with בניה his sons? תנחם׃ or canst thou guide}%
\verse{הידעת Knowest חקות thou the ordinances שׁמים of heaven? אם תשׂים canst thou set משׁטרו the dominion בארץ׃ thereof in the earth?}%
\verse{התרים Canst thou lift up לעב to the clouds, קולך thy voice ושׁפעת that abundance מים of waters תכסך׃ may cover}%
\verse{התשׁלח Canst thou send ברקים lightnings, וילכו that they may go, ויאמרו and say לך הננו׃}%
\verse{מי Who שׁת hath put בטחות in the inward parts? חכמה wisdom או or מי who נתן hath given לשׂכוי to the heart? בינה׃ understanding}%
\verse{מי Who יספר can number שׁחקים the clouds בחכמה in wisdom? ונבלי the bottles שׁמים of heaven, מי or who ישׁכיב׃ can stay}%
\verse{בצקת groweth עפר When the dust למוצק into hardness, ורגבים and the clods ידבקו׃ cleave fast together?}%
\verse{התצוד Wilt thou hunt ללביא for the lion? טרף the prey וחית the appetite כפירים of the young lions, תמלא׃ or fill}%
\verse{כי When ישׁחו they couch במעונות in dens, ישׁבו abide בסכה in the covert למו to lie in wait? ארב׃ to lie in wait?}%
\verse{מי Who יכין provideth לערב for the raven צידו his food? כי when ילדו his young ones אל God, אל unto ישׁועו cry יתעו they wander לבלי for lack אכל׃ of meat.}%
\end{biblechapter}%
\begin{biblechapter}% Job 39
\verse{הידעת Knowest עת thou the time לדת bring forth? יעלי when the wild goats סלע of the rock חלל do calve? אילות when the hinds תשׁמר׃ canst thou mark}%
\verse{תספר Canst thou number ירחים the months תמלאנה they fulfill? וידעת or knowest עת thou the time לדתנה׃ when they bring forth?}%
\verse{תכרענה They bow ילדיהן their young ones, תפלחנה themselves, they bring forth חבליהם their sorrows. תשׁלחנה׃ they cast out}%
\verse{יחלמו are in good liking, בניהם Their young ones ירבו they grow up בבר with corn; יצאו they go forth, ולא not שׁבו׃ and return}%
\verse{מי Who שׁלח hath sent out פרא the wild ass חפשׁי free? ומסרות the bands ערוד of the wild ass? מי or who פתח׃ hath loosed}%
\verse{אשׁר Whose שׂמתי I have made ערבה the wilderness, ביתו house ומשׁכנותיו his dwellings. מלחה׃ and the barren land}%
\verse{ישׂחק He scorneth להמון the multitude קריה of the city, תשׁאות he the crying נוגשׂ of the driver. לא neither ישׁמע׃ regardeth}%
\verse{יתור The range הרים of the mountains מרעהו his pasture, ואחר after כל every ירוק green thing. ידרושׁ׃ and he searcheth}%
\verse{היאבה be willing רים Will the unicorn עבדך to serve אם thee, or ילין abide על by אבוסך׃ thy crib?}%
\verse{התקשׁר Canst thou bind רים the unicorn בתלם in the furrow? עבתו with his band אם or ישׂדד will he harrow עמקים the valleys אחריך׃ after}%
\verse{התבטח Wilt thou trust בו כי him, because רב great? כחו his strength ותעזב or wilt thou leave אליו to יגיעך׃ thy labor}%
\verse{התאמין Wilt thou believe בו כי him, that ישׁוב he will bring home זרעך thy seed, וגרנך thy barn? יאסף׃ and gather}%
\verse{כנף wings רננים unto the peacocks? נעלסה the goodly אם or אברה wings חסידה and feathers ונצה׃ unto the ostrich?}%
\verse{כי Which תעזב leaveth לארץ in the earth, בציה her eggs ועל them in עפר dust, תחמם׃ and warmeth}%
\verse{ותשׁכח And forgetteth כי that רגל the foot תזורה may crush וחית beast השׂדה them, or that the wild תדושׁה׃ may break}%
\verse{הקשׁיח She is hardened בניה against her young ones, ללא as though not לה לריק is in vain יגיעה hers: her labor בלי without פחד׃ fear;}%
\verse{כי Because השׁה hath deprived אלוה God חכמה her of wisdom, ולא neither חלק hath he imparted לה בבינה׃ to her understanding.}%
\verse{כעת What time במרום on high, תמריא she lifteth up herself תשׂחק she scorneth לסוס the horse ולרכבו׃ and his rider.}%
\verse{התתן Hast thou given לסוס the horse גבורה strength? התלבישׁ hast thou clothed צוארו his neck רעמה׃ with thunder?}%
\verse{התרעישׁנו Canst thou make him afraid כארבה as a grasshopper? הוד the glory נחרו of his nostrils אימה׃ terrible.}%
\verse{יחפרו He paweth בעמק in the valley, וישׂישׂ and rejoiceth בכח in strength: יצא he goeth on לקראת to meet נשׁק׃ the armed men.}%
\verse{ישׂחק He mocketh לפחד at fear, ולא and is not יחת frightened; ולא neither ישׁוב turneth he back מפני from חרב׃ the sword.}%
\verse{עליו against תרנה rattleth אשׁפה The quiver להב him, the glittering חנית spear וכידון׃ and the shield.}%
\verse{ברעשׁ with fierceness ורגז and rage: יגמא He swalloweth ארץ the ground ולא neither יאמין believeth כי he that קול the sound שׁופר׃ of the trumpet.}%
\verse{בדי among שׁפר the trumpets, יאמר He saith האח Ha, ha; ומרחוק afar off, יריח and he smelleth מלחמה the battle רעם the thunder שׂרים of the captains, ותרועה׃ and the shouting.}%
\verse{המבינתך by thy wisdom, יאבר fly נץ Doth the hawk יפרשׂ stretch כנפו her wings לתימן׃ toward the south?}%
\verse{אם על at פיך thy command, יגביה mount up נשׁר Doth the eagle וכי ירים and make her nest on high? קנו׃ and make her nest on high?}%
\verse{סלע on the rock, ישׁכן She dwelleth ויתלנן and abideth על upon שׁן the crag סלע of the rock, ומצודה׃ and the strong place.}%
\verse{משׁם חפר she seeketh אכל the prey, למרחוק afar off. עיניו her eyes יביטו׃ behold}%
\verse{ואפרחו Her young ones יעלעו also suck up דם blood: ובאשׁר and where חללים the slain שׁם there הוא׃ she.}%
\end{biblechapter}%
\begin{biblechapter}% Job 40
\verse{ויען answered יהוה Moreover the LORD את איוב Job, ויאמר׃ and said,}%
\verse{הרב Shall he that contendeth עם with שׁדי the Almighty יסור instruct מוכיח he that reproveth אלוה God, יעננה׃ let him answer}%
\verseWithHeading{Job Responds to Adonai}{ויען answered איוב Then Job את יהוה the LORD, ויאמר׃ and said,}%
\verse{הן Behold, קלתי I am vile; מה what אשׁיבך shall I answer ידי mine hand שׂמתי thee? I will lay למו upon פי׃ my mouth.}%
\verse{אחת Once דברתי have I spoken; ולא but I will not אענה answer: ושׁתים yea, twice; ולא אוסיף׃}%
\verseWithHeading{Adonai Challenges Job Again}{ויען Then answered יהוה the LORD את איוב unto Job מנסערה the whirlwind, ויאמר׃ and said,}%
\verse{אזר Gird up נא now כגבר like a man: חלציך thy loins אשׁאלך I will demand והודיעני׃ of thee, and declare}%
\verseWithHeading{Adonai Interrogates Job Again}{האף Wilt thou also תפר disannul משׁפטי my judgment? תרשׁיעני wilt thou condemn למען me, that תצדק׃ thou mayest be righteous?}%
\verse{ואם זרוע Hast thou an arm כאל like God? לך ובקול with a voice כמהו like him? תרעם׃ or canst thou thunder}%
\verse{עדה Deck נא thyself now גאון majesty וגבה and excellency; והוד thyself with glory והדר and beauty. תלבשׁ׃ and array}%
\verse{הפץ Cast abroad עברות the rage אפך of thy wrath: וראה and behold כל every one גאה proud, והשׁפילהו׃ and abase}%
\verse{ראה Look on כל every one גאה proud, הכניעהו bring him low; והדך and tread down רשׁעים the wicked תחתם׃ in their place.}%
\verse{טמנם Hide בעפר them in the dust יחד together; פניהם their faces חבשׁ bind בטמון׃ in secret.}%
\verse{וגם also אני Then will I אודך confess כי unto thee that תושׁע can save לך ימינך׃ thine own right hand}%
\verse{הנה Behold נא now בהמות behemoth, אשׁר which עשׂיתי I made עמך with חציר grass כבקר as an ox. יאכל׃ thee; he eateth}%
\verse{הנה Lo נא now, כחו his strength במתניו in his loins, ואנו and his force בשׁרירי in the navel בטנו׃ of his belly.}%
\verse{יחפץ He moveth זנבו his tail כמו like ארז a cedar: גידי the sinews פחדו of his stones ישׂרגו׃ are wrapped together.}%
\verse{עצמיו His bones אפיקי strong pieces נחושׁה of brass; גרמיו his bones כמטיל like bars ברזל׃ of iron.}%
\verse{הוא He ראשׁית the chief דרכי of the ways אל of God: העשׂו he that made יגשׁ to approach חרבו׃ him can make his sword}%
\verse{כי Surely בול food, הרים the mountains ישׂאו bring him forth לו וכל all חית the beasts השׂדה of the field ישׂחקו play. שׁם׃ where}%
\verse{תחת under צאלים the shady trees, ישׁכב He lieth בסתר in the covert קנה of the reed, ובצה׃ and fens.}%
\verse{יסכהו cover צאלים The shady trees צללו him their shadow; יסבוהו compass him about. ערבי the willows נחל׃ of the brook}%
\verse{הן Behold, יעשׁק he drinketh up נהר a river, לא not: יחפוז hasteth יבטח he trusteth כי that יגיח he can draw up ירדן Jordan אל into פיהו׃ his mouth.}%
\verse{בעיניו it with his eyes: יקחנו He taketh במוקשׁים through snares. ינקב pierceth אף׃ nose}%
\end{biblechapter}%
\begin{biblechapter}% Job 41
\verse{תמשׁך Canst thou draw out לויתן leviathan בחכה with a hook? ובחבל with a cord תשׁקיע thou lettest down? לשׁנו׃ or his tongue}%
\verse{התשׂים Canst thou put אגמון a hook באפו into his nose? ובחוח through with a thorn? תקוב or bore לחיו׃ his jaw}%
\verse{הירבה Will he make many אליך unto תחנונים supplications אם ידבר thee? will he speak אליך unto רכות׃ soft}%
\verse{היכרת Will he make ברית a covenant עמך with תקחנו thee? wilt thou take לעבד him for a servant עולם׃ forever?}%
\verse{התשׂחק Wilt thou play בו כצפור with him as a bird? ותקשׁרנו or wilt thou bind לנערותיך׃ him for thy maidens?}%
\verse{יכרו make a banquet עליו of חברים Shall the companions יחצוהו him? shall they part בין him among כנענים׃ the merchants?}%
\verse{התמלא Canst thou fill בשׂכות with barbed irons? עורו his skin ובצלצל spears? דגים with fish ראשׁו׃ or his head}%
\verse{שׂים Lay עליו upon כפך thine hand זכר him, remember מלחמה the battle, אל do no תוסף׃ more.}%
\verse{הן Behold, תחלתו the hope נכזבה of him is in vain: הגם even אל מראיו the sight יטל׃ shall not be cast down}%
\verse{לא None אכזר fierce כי that יעורנו dare stir him up: ומי who הוא לפני before יתיצב׃ then is able to stand}%
\verse{מי Who הקדימני hath prevented ואשׁלם me, that I should repay תחת under כל the whole השׁמים heaven לי הוא׃}%
\verse{לא I will not אחרישׁ conceal בדיו his parts, ודבר nor גבורות his power, וחין nor his comely ערכו׃ proportion.}%
\verse{מי Who גלה can discover פני the face לבושׁו of his garment? בכפל with his double רסנו bridle? מי who יבוא׃ can come}%
\verse{דלתי the doors פניו of his face? מי Who פתח can open סביבות round about. שׁניו his teeth אימה׃ terrible}%
\verse{גאוה pride, אפיקי מגניםגור shut up together חותם seal. צר׃ a close}%
\verse{אחד One באחד to another, יגשׁו is so near ורוח air לא that no יבוא can come ביניהם׃ between}%
\verse{אישׁ one באחיהו to another, ידבקו They are joined יתלכדו they stick together, ולא that they cannot יתפרדו׃ be sundered.}%
\verse{עטישׁתיו By his sneezes תהל doth shine, אור a light ועיניו and his eyes כעפעפי like the eyelids שׁחר׃ of the morning.}%
\verse{מפיו לפידים burning lamps, יהלכו go כידודי sparks אשׁ of fire יתמלטו׃ leap out.}%
\verse{מנחיריו יצא goeth עשׁן smoke, כדוד pot נפוח as of a seething ואגמן׃ or caldron.}%
\verse{נפשׁו His breath גחלים coals, תלהט kindleth ולהב and a flame מפיו of his mouth. יצא׃ goeth out}%
\verse{בצוארו In his neck ילין remaineth עז strength, ולפניו before תדוץ is turned into joy דאבה׃ and sorrow}%
\verse{מפלי The flakes בשׂרו of his flesh דבקו are joined together: יצוק they are firm עליו in בל themselves; they cannot ימוט׃ be moved.}%
\verse{לבו His heart יצוק is as firm כמו as אבן a stone; ויצוק yea, as hard כפלח as a piece תחתית׃ of the nether}%
\verse{משׂתו יגורו are afraid: אלים himself, the mighty משׁברים by reason of breakings יתחטאו׃ they purify themselves.}%
\verse{משׂיגהו of him that layeth חרב The sword בלי at him cannot תקום hold: חנית the spear, מסע the dart, ושׁריה׃ nor the habergeon.}%
\verse{יחשׁב He esteemeth לתבן as straw, ברזל iron לעץ wood. רקבון as rotten נחושׁה׃ brass}%
\verse{לא cannot יבריחנו make him flee: בן קשׁתקשׁ with him into stubble. נהפכו are turned לו אבני slingstones קלע׃ slingstones}%
\verse{כקשׁ as stubble: נחשׁבו are counted תותח Darts וישׂחק he laugheth לרעשׁ at the shaking כידון׃ of a spear.}%
\verse{תחתיו under חדודי Sharp חרשׂ stones ירפד him: he spreadeth חרוץ sharp pointed things עלי upon טיט׃ the mire.}%
\verse{ירתיח to boil כסיר like a pot: מצולה He maketh the deep ים the sea ישׂים he maketh כמרקחה׃ like a pot of ointment.}%
\verse{אחריו after יאיר to shine נתיב He maketh a path יחשׁב him; would think תהום the deep לשׂיבה׃ hoary.}%
\verse{אין there is not על Upon עפר earth משׁלו his like, העשׂו who is made לבלי without חת׃ fear.}%
\verse{את כל all גבה high יראה He beholdeth הוא he מלך a king על over כל all בני the children שׁחץ׃ of pride.}%
\end{biblechapter}%
\begin{biblechapter}% Job 42
\verseWithHeading{Job’s Repentance and Restoration}{ויען answered איוב Then Job את יהוה the LORD, ויאמר׃ and said,}%
\verse{ידעת I know כי that כל every תוכל thou canst do ולא and no יבצר can be withheld ממך from מזמה׃ thought}%
\verse{מי Who זה he מעלים that hideth עצה counsel בלי without דעת knowledge? לכן therefore הגדתי have I uttered ולא not; אבין that I understood נפלאות things too wonderful ממני for ולא not. אדע׃ me, which I knew}%
\verse{שׁמע Hear, נא I beseech thee, ואנכי and I אדבר will speak: אשׁאלך I will demand והודיעני׃ of thee, and declare}%
\verse{לשׁמע I have heard אזן of the ear: שׁמעתיך of thee by the hearing ועתה but now עיני mine eye ראתך׃ seeth}%
\verse{על in כן אמאס I abhor ונחמתי and repent על עפר dust ואפר׃ and ashes.}%
\verse{ויהי And it was אחר that after דבר had spoken יהוה the LORD את הדברים words האלה these אל unto איוב Job, ויאמר said יהוה the LORD אל to אליפז Eliphaz התימני the Temanite, חרה is kindled אפי My wrath בך ובשׁני against thee, and against thy two רעיך friends: כי for לא ye have not דברתם spoken אלי of נכונה me right, כעבדי as my servant איוב׃ Job}%
\verse{ועתה unto you now קחו Therefore take לכם שׁבעה seven פרים bullocks ושׁבעה and seven אילים rams, ולכו and go אל to עבדי my servant איוב Job, והעליתם and offer up עולה yourselves a burnt offering; בעדכם for ואיוב Job עבדי and my servant יתפלל shall pray עליכם for כי you: for אם you: for פניו him אשׂא will I accept: לבלתי lest עשׂות I deal עמכם with נבלה you folly, כי in that לא ye have not דברתם spoken אלי of נכונה me right, כעבדי like my servant איוב׃ Job.}%
\verse{וילכו went, אליפז So Eliphaz התימני the Temanite ובלדד and Bildad השׁוחי the Shuhite צפר Zophar הנעמתי the Naamathite ויעשׂו and did כאשׁר according as דבר commanded אליהם commanded יהוה the LORD וישׂא also accepted יהוה them: the LORD את פני also accepted איוב׃ Job.}%
\verse{ויהוה And the LORD שׁב turned את שׁבית the captivity איוב of Job, בהתפללו when he prayed בעד for רעהו his friends: ויסף gave יהוה also the LORD את כל as much as he had before. אשׁר as much as he had before. לאיוב Job למשׁנה׃ twice}%
\verse{ויבאו Then came אליו there unto כל him all אחיו his brethren, וכל and all אחיתיו his sisters, וכל and all ידעיו they that had been of his acquaintance לפנים before, ויאכלו and did eat עמו with לחם bread בביתו him in his house: וינדו and they bemoaned לו וינחמו him, and comforted אתו על him over כל all הרעה the evil אשׁר that הביא had brought יהוה the LORD עליו upon ויתנו also gave לו אישׁ him: every man קשׂיטה him a piece of money, אחת him a piece of money, ואישׁ and every one נזם earring זהב of gold. אחד׃ an}%
\verse{ויהוה So the LORD ברך blessed את אחרית the latter end איוב of Job מראשׁתו more than his beginning: ויהי for he had לו ארבעה fourteen עשׂר fourteen אלף thousand צאן sheep, ושׁשׁת and six אלפים thousand גמלים camels, ואלף and a thousand צמד yoke בקר of oxen, ואלף and a thousand אתונות׃ she asses.}%
\verse{ויהי He had לו שׁבענה also seven בנים sons ושׁלושׁ and three בנות׃ daughters.}%
\verse{ויקרא And he called שׁם the name האחת of the first, ימימה Jemima; ושׁם and the name השׁנית of the second, קציעה Kezia; ושׁם and the name השׁלישׁית of the third, קרן הפוך׃ Keren-happuch.}%
\verse{ולא were no נמצא found נשׁים women יפות fair כבנות as the daughters איוב of Job: בכל And in all הארץ the land ויתן gave להם אביהם and their father נחלה them inheritance בתוך among אחיהם׃ their brethren.}%
\verse{ויחי lived איוב Job אחרי After זאת this מאה a hundred וארבעים and forty שׁנה years, וירא and saw את בניו his sons, ואת בני andhis sons' בניו sons, ארבעה four דרות׃ generations.}%
\verse{וימת died, איוב So Job זקן old ושׂבע and full ימים׃ of days.}%
\end{biblechapter}%
