\biblebook{Judges}
\begin{biblechapter}% Judges 1
\verseWithHeading{Israel Continues Its Conquest}{ויהי it came to pass, אחרי Now after מות the death יהושׁע of Joshua וישׁאלו asked בני that the children ישׂראל of Israel ביהוה the LORD, לאמר saying, מי Who יעלה shall go up לנו אל for us against הכנעני the Canaanites בתחלה first, להלחם׃ to fight}%
\verse{ויאמר said, יהוה And the LORD יהודה Judah יעלה shall go up: הנה behold, נתתי I have delivered את הארץ the land בידו׃ into his hand.}%
\verse{ויאמר said יהודה And Judah לשׁמעון unto Simeon אחיו his brother, עלה Come up אתי with בגורלי me into my lot, ונלחמה that we may fight בכנעני against the Canaanites; והלכתי will go גם likewise אני and I אתך with בגורלך thee into thy lot. וילך went אתו with שׁמעון׃ So Simeon}%
\verse{ויעל went up; יהודה And Judah ויתן delivered יהוה and the LORD את הכנעני the Canaanites והפרזי and the Perizzites בידם into their hand: ויכום and they slew בבזק of them in Bezek עשׂרת ten אלפים thousand אישׁ׃ men.}%
\verse{וימצאו And they found את אדני בזק Adoni-bezek בבזק in Bezek: וילחמו and they fought בו ויכו against him, and they slew את הכנעני the Canaanites ואת הפרזי׃ and the Perizzites.}%
\verse{וינס fled; אדני בזק But Adoni-bezek וירדפו and they pursued אחריו after ויאחזו him, and caught אתו ויקצצו him, and cut off את בהנות his thumbs ידיו his thumbs ורגליו׃ and his great toes.}%
\verse{ויאמר said, אדני בזק And Adoni-bezek שׁבעים Threescore and ten מלכים kings, בהנות having their thumbs ידיהם having their thumbs ורגליהם and their great toes מקצצים cut off, היו מלקטים gathered תחת under שׁלחני my table: כאשׁר as עשׂיתי I have done, כן so שׁלם hath requited לי אלהים God ויביאהו me. And they brought ירושׁלם him to Jerusalem, וימת he died. שׁם׃ and there}%
\verse{וילחמו had fought בני Now the children יהודה of Judah בירושׁלם against Jerusalem, וילכדו and had taken אותה ויכוה it, and smitten לפי it with the edge חרב of the sword, ואת העיר the city שׁלחו and set באשׁ׃ on fire.}%
\verse{ואחר And afterward ירדו went down בני the children יהודה of Judah להלחם to fight בכנעני against the Canaanites, יושׁב that dwelt ההר in the mountain, והנגב and in the south, והשׁפלה׃ and in the valley.}%
\verse{וילך went יהודה And Judah אל הכנעני the Canaanites היושׁב that dwelt בחברון in Hebron: ושׁם (now the name חברון of Hebron לפנים before קרית ארבע Kirjath-arba:) ויכו and they slew את שׁשׁי Sheshai, ואת אחימן and Ahiman, ואת תלמי׃ and Talmai.}%
\verse{וילך he went משׁם אל against יושׁבי the inhabitants דביר of Debir: ושׁם and the name דביר of Debir לפנים before קרית ספר׃ Kirjath-sepher:}%
\verse{ויאמר said, כלב And Caleb אשׁר He that יכה smiteth את קרית ספר Kirjath-sepher, ולכדה and taketh ונתתי it, to him will I give לו את עכסהתי my daughter לאשׁה׃ to wife.}%
\verse{וילכדה took עתניאל And Othniel בן the son קנז of Kenaz, אחי brother, כלב Caleb's הקטן younger ממנו ויתן it: and he gave לו את עכסהתו his daughter לאשׁה׃ to wife.}%
\verse{ויהי And it came to pass, בבואה when she came ותסיתהו that she moved לשׁאול him to ask מאת אביה her father השׂדה a field: ותצנח and she lighted מעל from off החמור ass; ויאמר said לה כלב and Caleb מה׃ unto her, What}%
\verse{ותאמר And she said לו הבה unto him, Give לי ברכה me a blessing: כי for ארץ land; הנגב me a south נתתני thou hast given ונתתה give לי גלת me also springs מים of water. ויתן gave לה כלב And Caleb את גלת springs עלית her the upper ואת גלת springs. תחתית׃ and the nether}%
\verse{ובני And the children קיני of the Kenite, חתן father-in-law, משׁה Moses' עלו went up מעיר out of the city התמרים of palm trees את with בני the children יהודה of Judah מדבר into the wilderness יהודה of Judah, אשׁר which בנגב in the south ערד of Arad; וילך and they went וישׁב and dwelt את among העם׃ the people.}%
\verse{וילך went יהודה And Judah את with שׁמעון Simeon אחיו his brother, ויכו and they slew את הכנעני the Canaanites יושׁב that inhabited צפת Zephath, ויחרימו and utterly destroyed אותה ויקרא was called את שׁם it. And the name העיר of the city חרמה׃ Hormah.}%
\verse{וילכד took יהודה Also Judah את עזה Gaza ואת גבולה withthe coast ואת אשׁקלון thereof, and Askelon ואת גבולה with the coast ואת עקרון thereof, and Ekron ואת גבולה׃ with the coast}%
\verse{ויהי was יהוה And the LORD את with יהודה Judah; וירשׁ and he drove out את ההר the mountain; כי but לא could not להורישׁ drive out את ישׁבי the inhabitants העמק of the valley, כי because רכב they had chariots ברזל׃ of iron.}%
\verse{ויתנו And they gave לכלב unto Caleb, את חברון Hebron כאשׁר as דבר said: משׁה Moses ויורשׁ and he expelled משׁם thence את שׁלשׁה the three בני sons הענק׃ of Anak.}%
\verse{ואת היבוסי the Jebusites ישׁב that inhabited ירושׁלם Jerusalem; לא did not הורישׁו drive out בני And the children בנימן of Benjamin וישׁב dwell היבוסי but the Jebusites את with בני the children בנימן of Benjamin בירושׁלם in Jerusalem עד unto היום day. הזה׃ this}%
\verse{ויעלו went up בית And the house יוסף of Joseph, גם also הם they בית אל against Bethel: ויהוה and the LORD עמם׃ with}%
\verse{ויתירו sent to descry בית And the house יוסף of Joseph בבית אל Bethel. ושׁם (Now the name העיר of the city לפנים before לוז׃ Luz.)}%
\verse{ויראו saw השׁמרים And the spies אישׁ a man יוצא come forth מן out of העיר the city, ויאמרו and they said לו הראנו unto him, Show נא us, we pray thee, את מבוא the entrance העיר into the city, ועשׂינו and we will show עמך and we will show חסד׃ thee mercy.}%
\verse{ויראם And when he showed את מבוא them the entrance העיר into the city, ויכו they smote את העיר the city לפי with the edge חרב of the sword; ואת האישׁ the man ואת כל and all משׁפחתו his family. שׁלחו׃ but they let go}%
\verse{וילך went האישׁ And the man ארץ into the land החתים of the Hittites, ויבן and built עיר a city, ויקרא and called שׁמה the name לוז thereof Luz: הוא which שׁמה the name עד thereof unto היום day. הזה׃ this}%
\verse{ולא Neither הורישׁ drive out מנשׁה did Manasseh את בית שׁאן Beth-shean ואת בנותיה and her towns, ואת תענך nor Taanach ואת בנתיה and her towns, ואת ישׁב nor the inhabitants דור of Dor ואת בנותיה and her towns, ואת יושׁבי nor the inhabitants יבלעם of Ibleam ואת בנתיה and her towns, ואת יושׁבי nor the inhabitants מגדו of Megiddo ואת בנותיה and her towns: ויואל would הכנעני but the Canaanites לשׁבת dwell בארץ land. הזאת׃ in that}%
\verse{ויהי And it came to pass, כי when חזק was strong, ישׂראל Israel וישׂם that they put את הכנעני the Canaanites למס to tribute, והורישׁ utterly drive them out. לא and did not הורישׁו׃ utterly drive them out.}%
\verse{ואפרים did Ephraim לא Neither הורישׁ drive out את הכנעני the Canaanites היושׁב that dwelt בגזר in Gezer; וישׁב dwelt הכנעני but the Canaanites בקרבו among בגזר׃ in Gezer}%
\verse{זבולן did Zebulun לא Neither הורישׁ drive out את יושׁבי the inhabitants קטרון of Kitron, ואת יושׁבי nor the inhabitants נהלל of Nahalol; וישׁב dwelt הכנעני but the Canaanites בקרבו among ויהיו them, and became למס׃ tributaries.}%
\verse{אשׁר did Asher לא Neither הורישׁ drive out את ישׁבי the inhabitants עכו of Accho, ואת יושׁבי nor the inhabitants צידון of Zidon, ואת אחלב nor of Ahlab, ואת אכזיב nor of Achzib, ואת חלבה nor of Helbah, ואת אפיק nor of Aphik, ואת רחב׃ nor of Rehob:}%
\verse{וישׁב dwelt האשׁרי But the Asherites בקרב among הכנעני the Canaanites, ישׁבי the inhabitants הארץ of the land: כי for לא they did not הורישׁו׃ drive them out.}%
\verse{נפתלי did Naphtali לא Neither הורישׁ drive out את ישׁבי the inhabitants בית שׁמשׁ of Beth-shemesh, ואת ישׁבי nor the inhabitants בית ענת of Beth-anath; וישׁב but he dwelt בקרב among הכנעני the Canaanites, ישׁבי the inhabitants הארץ of the land: וישׁבי nevertheless the inhabitants בית שׁמשׁ of Beth-shemesh ובית ענת and of Beth-anath היו became להם למס׃ tributaries}%
\verse{וילחצו forced האמרי And the Amorites את בני the children דן of Dan ההרה into the mountain: כי for לא they would not נתנו suffer לרדת them to come down לעמק׃ to the valley:}%
\verse{ויואל would האמרי But the Amorites לשׁבת dwell בהר in mount חרס Heres באילון in Aijalon, ובשׁעלבים and in Shaalbim: ותכבד prevailed, יד yet the hand בית of the house יוסף of Joseph ויהיו so that they became למס׃ tributaries.}%
\verse{וגבול And the coast האמרי of the Amorites ממעלה עקרבים מהסלע from the rock, ומעלה׃ and upward.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 2
\verseWithHeading{Israel Disobeys Adonai}{ויעל came up מלאך And an angel יהוה of the LORD מן from הגלגל Gilgal אל to הבכים Bochim, ויאמר and said, אעלה אתכםמצרים ואביא and have brought אתכם אל you unto הארץ the land אשׁר which נשׁבעתי I swore לאבתיכם unto your fathers; ואמר and I said, לא אפר break בריתי my covenant אתכם with לעולם׃}%
\verse{ואתם And ye לא no תכרתו shall make ברית league ליושׁבי with the inhabitants הארץ land; הזאת of this מזבחותיהם their altars: תתצון ye shall throw down ולא but ye have not שׁמעתם obeyed בקלי my voice: מה why זאת this? עשׂיתם׃ have ye done}%
\verse{וגם Wherefore I also אמרתי said, לא I will not אגרשׁ אותםפניכם from before והיו you; but they shall be לכם לצדים in your sides, ואלהיהם and their gods יהיו shall be לכם למוקשׁ׃ a snare}%
\verse{ויהי And it came to pass, כדבר spoke מלאך when the angel יהוה of the LORD את הדברים words האלה these אל unto כל all בני the children ישׂראל of Israel, וישׂאו lifted up העם that the people את קולם their voice, ויבכו׃ and wept.}%
\verse{ויקראו And they called שׁם the name המקום place ההוא of that בכים Bochim: ויזבחו and they sacrificed שׁם there ליהוה׃ unto the LORD.}%
\verseWithHeading{Joshua Dies}{וישׁלח go, יהושׁע And when Joshua את העם had let the people וילכו went בני the children ישׂראל of Israel אישׁ every man לנחלתו unto his inheritance לרשׁת to possess את הארץ׃ the land.}%
\verse{ויעבדו served העם And the people את יהוה the LORD כל all ימי the days יהושׁע of Joshua, וכל and all ימי the days הזקנים of the elders אשׁר that האריכו outlived ימים אחריהושׁוע Joshua, אשׁר who ראו had seen את כל all מעשׂה works יהוה of the LORD, הגדול the great אשׁר that עשׂה he did לישׂראל׃ for Israel.}%
\verse{וימת died, יהושׁע And Joshua בן the son נון of Nun, עבד the servant יהוה of the LORD, בן old. מאה a hundred ועשׂר and ten שׁנים׃ years}%
\verse{ויקברו And they buried אותו בגבול him in the border נחלתו of his inheritance בתמנת חרס in Timnath-heres, בהר in the mount אפרים of Ephraim, מצפון on the north side להר of the hill געשׁ׃ Gaash.}%
\verse{וגם And also כל all הדור generation ההוא that נאספו were gathered אל unto אבותיו their fathers: ויקם and there arose דור generation אחר another אחריהם after אשׁר them, which לא not ידעו knew את יהוה the LORD, וגם nor yet את המעשׂה the works אשׁר which עשׂה he had done לישׂראל׃ for Israel.}%
\verseWithHeading{Israel Worships the Baals}{ויעשׂו did בני And the children ישׂראל of Israel את הרע evil בעיני in the sight יהוה of the LORD, ויעבדו and served את הבעלים׃ Baalim:}%
\verse{ויעזבו And they forsook את יהוה the LORD אלהי God אבותם of their fathers, המוציא אותםארץ of the land מצרים of Egypt, וילכו and followed אחרי and followed אלהים gods, אחרים other מאלהי of the gods העמים of the people אשׁר that סביבותיהם round about וישׁתחוו them, and bowed themselves להם ויכעסו unto them, and provoked את יהוה׃ the LORD}%
\verse{ויעזבו And they forsook את יהוה the LORD, ויעבדו and served לבעל Baal ולעשׁתרות׃ and Ashtaroth.}%
\verse{ויחר was hot אף And the anger יהוה of the LORD בישׂראל against Israel, ויתנם and he delivered ביד them into the hands שׁסים of spoilers וישׁסו that spoiled אותם וימכרם them, and he sold ביד them into the hands אויביהם of their enemies מסביב round about, ולא not יכלו so that they could עוד any longer לעמד stand לפני before אויביהם׃ their enemies.}%
\verse{בכל אשׁר as יצאו they went out, יד the hand יהוה of the LORD היתה was בם לרעה against them for evil, כאשׁר and as דבר had said, יהוה the LORD וכאשׁר נשׁבע had sworn יהוה the LORD להם ויצר distressed. להם מאד׃ unto them: and they were greatly}%
\verse{ויקם raised up יהוה Nevertheless the LORD שׁפטים judges, ויושׁיעום which delivered מיד them out of the hand שׁסיהם׃ of those that spoiled}%
\verse{וגם And yet אל unto שׁפטיהם their judges, לא they would not שׁמעו hearken כי but זנו they went a whoring אחרי after אלהים gods, אחרים other וישׁתחוו and bowed themselves להם סרו unto them: they turned מהר quickly מן out of הדרך the way אשׁר which הלכו walked אבותם their fathers לשׁמע in, obeying מצות the commandments יהוה of the LORD; לא not עשׂו they did כן׃ so.}%
\verse{וכי And when הקים raised them up יהוה the LORD להם שׁפטים judges, והיה was יהוה then the LORD עם with השׁפט the judge, והושׁיעם and delivered מיד them out of the hand איביהם of their enemies כל all ימי the days השׁופט of the judge: כי for ינחם it repented יהוה the LORD מנאקתם because of their groanings מפני by reason of them לחציהם that oppressed ודחקיהם׃ them and vexed}%
\verse{והיה And it came to pass, במות was dead, השׁופט when the judge ישׁבו they returned, והשׁחיתו and corrupted מאבותם more than their fathers, ללכת in following אחרי in following אלהים gods אחרים other לעבדם to serve ולהשׁתחות them, and to bow down להם לא not הפילו unto them; they ceased ממעלליהם from their own doings, ומדרכם הקשׁה׃}%
\verse{ויחר was hot אף And the anger יהוה of the LORD בישׂראל against Israel; ויאמר and he said, יען Because אשׁר that עברו hath transgressed הגוי people הזה this את בריתי my covenant אשׁר which צויתי I commanded את אבותם their fathers, ולא and have not שׁמעו hearkened לקולי׃ unto my voice;}%
\verse{גם also אני I לא will not אוסיף henceforth להורישׁ drive out אישׁ any מפניהם from before מן from before הגוים the nations אשׁר which עזב left יהושׁע Joshua וימת׃ when he died:}%
\verse{למען That נסות through them I may prove בם את ישׂראל Israel, השׁמרים will keep הם whether they את דרך the way יהוה of the LORD ללכת to walk בם כאשׁר therein, as שׁמרו did keep אבותם their fathers אם or לא׃ not.}%
\verse{וינח יהוה Therefore the LORD את הגוים nations, האלה those לבלתי without הורישׁם driving them out מהר hastily; ולא neither נתנם delivered ביד he them into the hand יהושׁע׃ of Joshua.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 3
\verseWithHeading{Some Nations Remain in the Land}{ואלה Now these הגוים the nations אשׁר which הניח יהוה the LORD לנסות to prove בם את ישׂראל Israel את כל by them, as many אשׁר by them, as many לא as had not ידעו known את כל all מלחמות the wars כנען׃ of Canaan;}%
\verse{רק Only למען that דעת דרות the generations בני of the children ישׂראל of Israel ללמדם to teach מלחמה them war, רק at the least אשׁר such as לפנים before לא nothing ידעום׃ might know,}%
\verse{חמשׁת five סרני lords פלשׁתים of the Philistines, וכל and all הכנעני the Canaanites, והצידני and the Sidonians, והחוי and the Hivites ישׁב that dwelt הר in mount הלבנון Lebanon, מהר from mount בעל חרמון Baal-hermon עד unto לבוא the entering in חמת׃ of Hamath.}%
\verse{ויהיו And they were לנסות to prove בם את ישׂראל Israel לדעת by them, to know הישׁמעו whether they would hearken את מצות unto the commandments יהוה of the LORD, אשׁר which צוה he commanded את אבותם their fathers ביד by the hand משׁה׃ of Moses.}%
\verse{ובני And the children ישׂראל of Israel ישׁבו dwelt בקרב among הכנעני the Canaanites, החתי Hittites, והאמרי and Amorites, והפרזי and Perizzites, והחוי and Hivites, והיבוסי׃ and Jebusites:}%
\verse{ויקחו And they took את בנותיהם their daughters להם לנשׁים to be their wives, ואת בנותיהם their daughters נתנו and gave לבניהם to their sons, ויעבדו and served את אלהיהם׃ their gods.}%
\verseWithHeading{Othniel}{ויעשׂו did בני And the children ישׂראל of Israel את הרע evil בעיני in the sight יהוה of the LORD, וישׁכחו and forgot את יהוה the LORD אלהיהם their God, ויעבדו and served את הבעלים Baalim ואת האשׁרות׃ and the groves.}%
\verse{ויחר was hot אף Therefore the anger יהוה of the LORD בישׂראל against Israel, וימכרם and he sold ביד them into the hand כושׁן רשׁעתים of Cushan-rishathaim מלך king ארם נהרים of Mesopotamia: ויעבדו served בני and the children ישׂראל of Israel את כושׁן רשׁעתים Cushan-rishathaim שׁמנה eight שׁנים׃ years.}%
\verse{ויזעקו cried בני And when the children ישׂראל of Israel אל unto יהוה the LORD, ויקם raised up יהוה the LORD מושׁיע a deliverer לבני to the children ישׂראל of Israel, ויושׁיעם who delivered את עתניאל them, Othniel בן the son קנז of Kenaz, אחי brother. כלב Caleb's הקטן younger ממנו׃}%
\verse{ותהי came עליו upon רוח And the Spirit יהוה of the LORD וישׁפט him, and he judged את ישׂראל Israel, ויצא and went out למלחמה to war: ויתן delivered יהוה and the LORD בידו into his hand; את כושׁן רשׁעתים Cushan-rishathaim מלך king ארם ותעז prevailed ידו and his hand על against כושׁן רשׁעתים׃ Cushan-rishathaim.}%
\verse{ותשׁקט had rest הארץ And the land ארבעים forty שׁנה years. וימת died. עתניאל And Othniel בן the son קנז׃ of Kenaz}%
\verseWithHeading{Ehud}{ויספו again בני And the children ישׂראל of Israel לעשׂות did הרע evil בעיני in the sight יהוה of the LORD: ויחזק strengthened יהוה and the LORD את עגלון Eglon מלך the king מואב of Moab על against ישׂראל Israel, על because כי because עשׂו they had done את הרע evil בעיני in the sight יהוה׃ of the LORD.}%
\verse{ויאסף And he gathered אליו unto את בני him the children עמון of Ammon ועמלק and Amalek, וילך and went ויך and smote את ישׂראל Israel, ויירשׁו and possessed את עיר the city התמרים׃ of palm trees.}%
\verse{ויעבדו served בני So the children ישׂראל of Israel את עגלון Eglon מלך the king מואב of Moab שׁמונה eighteen עשׂרה eighteen שׁנה׃ years.}%
\verse{ויזעקו cried בני But when the children ישׂראל of Israel אל unto יהוה the LORD, ויקם raised them up יהוה the LORD להם מושׁיע a deliverer, את אהודן the son גרא of Gera, בן הימיני אישׁ a man אטר יד and by him ימינו וישׁלחו sent בני the children ישׂראל of Israel בידו מנחה a present לעגלון unto Eglon מלך the king מואב׃ of Moab.}%
\verse{ויעשׂ made לו אהוד חרב him a dagger ולה שׁני which had two פיות edges, גמד of a cubit ארכה length; ויחגר and he did gird אותה מתחת it under למדיו his raiment על upon ירך thigh. ימינו׃ his right}%
\verse{ויקרב And he brought את המנחה the present לעגלון unto Eglon מלך king מואב of Moab: ועגלון and Eglon אישׁ man. בריא fat מאד׃ a very}%
\verse{ויהי כאשׁר And when כלה he had made an end להקריב to offer את המנחה the present, וישׁלח he sent away את העם the people נשׂאי that bore המנחה׃ the present.}%
\verse{והוא But he שׁב himself turned again מן from הפסילים the quarries אשׁר that את by הגלגל Gilgal, ויאמר and said, דבר errand סתר I have a secret לי אליך המלך unto thee, O king: ויאמר who said, הס Keep silence. ויצאו by him went out מעליו from כל And all העמדים that stood עליו׃}%
\verse{ואהוד בא came אליו unto והוא him; and he ישׁב was sitting בעלית parlor, המקרה in a summer אשׁר which לו לבדו he had for himself alone. ויאמר said, אהוד דבר I have a message אלהים from God לי אליך unto ויקם thee. And he arose מעל out of הכסא׃ seat.}%
\verse{וישׁלח put forth אהוד אתד hand, שׂמאלו his left ויקח and took את החרב the dagger מעל from ירך thigh, ימינו his right ויתקעה and thrust בבטנו׃ it into his belly:}%
\verse{ויבא went in גם also הנצב And the haft אחר after הלהב the blade; ויסגר closed החלב and the fat בעד upon הלהב the blade, כי so that לא he could not שׁלף draw החרב the dagger מבטנו out of his belly; ויצא came out. הפרשׁדנה׃ and the dirt}%
\verse{ויצא went forth אהוד המסדרונה through the porch, ויסגר and shut דלתות the doors העליה of the parlor בעדו ונעל׃ and locked}%
\verse{והוא When he יצא was gone out, ועבדיו his servants באו came; ויראו and when they saw והנה that, behold, דלתות the doors העליה of the parlor נעלות locked, ויאמרו they said, אך Surely מסיך covereth הוא he את רגליו his feet בחדר chamber. המקרה׃ in his summer}%
\verse{ויחילו And they tarried עד till בושׁ they were ashamed: והנה and, behold, איננו not פתח he opened דלתות the doors העליה of the parlor; ויקחו therefore they took את המפתח a key, ויפתחו and opened והנה and, behold, אדניהם their lord נפל fallen down ארצה on the earth. מת׃ dead}%
\verse{ואהוד נמלט escaped עד while התמהמהם they tarried, והוא עבר and passed beyond את הפסילים the quarries, וימלט and escaped השׂעירתה׃ unto Seirath.}%
\verse{ויהי And it came to pass, בבואו when he was come, ויתקע that he blew בשׁופר a trumpet בהר in the mountain אפרים of Ephraim, וירדו went down עמו with בני and the children ישׂראל of Israel מן him from ההר the mount, והוא and he לפניהם׃ before}%
\verse{ויאמר And he said אלהם unto רדפו them, Follow אחרי after כי me: for נתן hath delivered יהוה the LORD את איביכם your enemies את מואב the Moabites בידכם into your hand. וירדו And they went down אחריו after וילכדו him, and took את מעברות the fords הירדן of Jordan למואב toward Moab, ולא not נתנו and suffered אישׁ a man לעבר׃ to pass over.}%
\verse{ויכו And they slew את מואב of Moab בעת time ההיא at that כעשׂרת about ten אלפים thousand אישׁ men, כל all שׁמן lusty, וכל and all אישׁ men חיל of valor; ולא not נמלט and there escaped אישׁ׃ a man.}%
\verse{ותכנע was subdued מואב So Moab ביום day ההוא that תחת under יד the hand ישׂראל of Israel. ותשׁקט had rest הארץ And the land שׁמונים fourscore שׁנה׃ years.}%
\verseWithHeading{Shamgar}{ואחריו And after היה him was שׁמגר Shamgar בן the son ענת of Anath, ויך which slew את פלשׁתים of the Philistines שׁשׁ six מאות hundred אישׁ men במלמד goad: הבקר with an ox וישׁע delivered גם also הוא and he את ישׂראל׃ Israel.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 4
\verseWithHeading{Deborah and Barak}{ויספו again בני And the children ישׂראל of Israel לעשׂות did הרע evil בעיני in the sight יהוה of the LORD, ואהוד מת׃ was dead.}%
\verse{וימכרם sold יהוה And the LORD ביד them into the hand יבין of Jabin מלך king כנען of Canaan, אשׁר that מלך reigned בחצור in Hazor; ושׂר the captain צבאו of whose host סיסרא Sisera, והוא which יושׁב dwelt בחרשׁת in Harosheth of the Gentiles. הגוים׃}%
\verse{ויצעקו cried בני And the children ישׂראל of Israel אל unto יהוה the LORD: כי for תשׁע he had nine מאות hundred רכב chariots ברזל of iron; לו והוא he לחץ oppressed את בני the children ישׂראל of Israel. בחזקה עשׂרים and twenty שׁנה׃ years}%
\verse{ודבורה And Deborah, אשׁה a prophetess, נביאה a prophetess, אשׁת the wife לפידות of Lapidoth, היא she שׁפטה judged את ישׂראל Israel בעת time. ההיא׃ at that}%
\verse{והיא And she יושׁבת dwelt תחת under תמר the palm tree דבורה of Deborah בין between הרמה Ramah ובין בית אל and Bethel בהר in mount אפרים Ephraim: ויעלו came up אליה to בני and the children ישׂראל of Israel למשׁפט׃ her for judgment.}%
\verse{ותשׁלח And she sent ותקרא and called לברק Barak בן the son אבינעם of Abinoam מקדשׁ נפתלי of Naphtali ותאמר and said אליו unto הלא him, Hath not צוה commanded, יהוה the LORD אלהי God ישׂראל of Israel לך Go ומשׁכת and draw בהר toward mount תבור Tabor, ולקחת and take עמך with עשׂרת thee ten אלפים thousand אישׁ men מבני of the children נפתלי ומבני and of the children זבלון׃ of Zebulun?}%
\verse{ומשׁכתי And I will draw אליך unto אל thee to נחל the river קישׁון Kishon את סיסרא Sisera, שׂר the captain צבא army, יבין of Jabin's ואת רכבו with his chariots ואת המונו and his multitude; ונתתיהו and I will deliver בידך׃ him into thine hand.}%
\verse{ויאמר said אליה unto ברק And Barak אם her, If תלכי thou wilt go עמי with והלכתי me, then I will go: ואם but if לא thou wilt not תלכי go עמי with לא me, I will not אלך׃ go.}%
\verse{ותאמר And she said, הלך takest אלך and went עמך with אפס כי for לא thee: notwithstanding תהיה be תפארתך for thine honor; על הדרך the journey אשׁר that אתה thou הולך כייד into the hand אשׁה of a woman. ימכר shall sell יהוה the LORD את סיסרא Sisera ותקם arose, דבורה And Deborah ותלך עם with ברק Barak קדשׁה׃ to Kedesh.}%
\verse{ויזעק called ברק And Barak את זבולן Zebulun ואת נפתלי and Naphtali קדשׁה to Kedesh; ויעל and he went up ברגליו at his feet: עשׂרת with ten אלפי thousand אישׁ men ותעל went up עמו with דבורה׃ and Deborah}%
\verse{וחבר Now Heber הקיני נפרד had severed himself מקין מבני of the children חבב of Hobab חתן the father-in-law משׁה of Moses, ויט and pitched אהלו his tent עד unto אלון the plain בצענים of Zaanaim, אשׁר which את by קדשׁ׃ Kedesh.}%
\verse{ויגדו And they showed לסיסרא Sisera כי that עלה was gone up ברק Barak בן the son אבינעם of Abinoam הר to mount תבור׃ Tabor.}%
\verse{ויזעק gathered together סיסרא And Sisera את כל all רכבו his chariots, תשׁע nine מאות hundred רכב chariots ברזל of iron, ואת כל and all העם the people אשׁר that אתו with מחרשׁת הגויםל unto נחל the river קישׁון׃ of Kishon.}%
\verse{ותאמר said דברה And Deborah אל unto ברק Barak, קום Up; כי for זה this היום the day אשׁר in which נתן hath delivered יהוה the LORD את סיסרא Sisera בידך into thine hand: הלא is not יהוה the LORD יצא gone out לפניך before וירד went down ברק thee? So Barak מהר from mount תבור Tabor, ועשׂרת and ten אלפים thousand אישׁ men אחריו׃ after}%
\verse{ויהם discomfited יהוה And the LORD את סיסרא Sisera, ואת כל and all הרכב chariots, ואת כל and all המחנה host, לפי with the edge חרב of the sword לפני before ברק Barak; וירד lighted down סיסרא so that Sisera מעל off המרכבה chariot, וינס and fled away ברגליו׃ on his feet.}%
\verse{וברק But Barak רדף pursued אחרי after הרכב the chariots, ואחרי and after המחנה the host, עד unto חרשׁת Harosheth of the Gentiles: הגוים ויפל fell כל and all מחנה the host סיסרא of Sisera לפי upon the edge חרב of the sword; לא there was not נשׁאר left. עד אחד׃ a man}%
\verse{וסיסרא Howbeit Sisera נס fled away ברגליו on his feet אל to אהל the tent יעל of Jael אשׁת the wife חבר of Heber הקיני the Kenite: כי for שׁלום peace בין between יבין Jabin מלך the king חצור of Hazor ובין בית and the house חבר of Heber הקיני׃ the Kenite.}%
\verse{ותצא went out יעל And Jael לקראת to meet סיסרא Sisera, ותאמר and said אליו unto סורה him, Turn in, אדני my lord, סורה turn in אלי to אל not. תירא me; fear ויסר And when he had turned in אליה unto האהלה her into the tent, ותכסהו she covered בשׂמיכה׃ him with a mantle.}%
\verse{ויאמר And he said אליה unto השׁקיני to drink; נא her, Give me, I pray thee, מעט a little מים water כי for צמאתי I am thirsty. ותפתח And she opened את נאוד a bottle החלב of milk, ותשׁקהו and gave him drink, ותכסהו׃ and covered}%
\verse{ויאמר Again he said אליה unto עמד her, Stand פתח in the door האהל of the tent, והיה and it shall be, אם when אישׁ any man יבוא doth come ושׁאלך and inquire ואמר of thee, and say, הישׁ Is פה here? אישׁ there any man ואמרת that thou shalt say, אין׃ No.}%
\verse{ותקח took יעל Then Jael אשׁת wife חבר Heber's את יתד a nail האהל of the tent, ותשׂם and took את המקבת a hammer בידה in her hand, ותבוא and went אליו unto בלאט ותתקע him, and smote את היתד the nail ברקתו ותצנח and fastened בארץ it into the ground: והוא for he נרדם was fast asleep ויעף and weary. וימת׃ So he died.}%
\verse{והנה And, behold, ברק as Barak רדף pursued את סיסרא Sisera, ותצא came out יעל Jael לקראתו to meet ותאמר him, and said לו לך unto him, Come, ואראך and I will show את האישׁ thee the man אשׁר whom אתה thou מבקשׁ seekest. ויבא And when he came אליה into והנה her behold, סיסרא Sisera נפל lay מת dead, והיתד and the nail ברקתו׃}%
\verse{ויכנע subdued אלהים So God ביום day ההוא on that את יבין Jabin מלך the king כנען of Canaan לפני before בני the children ישׂראל׃ of Israel.}%
\verse{ותלך prospered, יד And the hand בני of the children ישׂראל of Israel הלוך and prevailed וקשׁה and prevailed על against יבין Jabin מלך the king כנען of Canaan, עד until אשׁר until הכריתו they had destroyed את יבין Jabin מלך king כנען׃ of Canaan.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 5
\verseWithHeading{The Song of Deborah and Barak}{ותשׁר Then sang דבורה Deborah וברק and Barak בן the son אבינעם of Abinoam ביום day, ההוא on that לאמר׃ saying,}%
\verse{בפרע for the avenging פרעות for the avenging בישׂראל of Israel, בהתנדב willingly offered themselves. עם when the people ברכו Praise יהוה׃ ye the LORD}%
\verse{שׁמעו Hear, מלכים O ye kings; האזינו give ear, רזנים O ye princes; אנכי I, ליהוה unto the LORD; אנכי I, אשׁירה will sing אזמר I will sing ליהוה to the LORD אלהי God ישׂראל׃ of Israel.}%
\verse{יהוה LORD, בצאתך when thou wentest out משׂעיר בצעדך when thou marchedst משׂדה out of the field אדום of Edom, ארץ the earth רעשׁה trembled, גם and שׁמים the heavens נטפו dropped, גם also עבים the clouds נטפו dropped מים׃ water.}%
\verse{הרים The mountains נזלו melted מפני from before יהוה the LORD, זה that סיני Sinai מפני from before יהוה the LORD אלהי God ישׂראל׃ of Israel.}%
\verse{בימי In the days שׁמגר of Shamgar בן the son ענת of Anath, בימי in the days יעל of Jael, חדלו were unoccupied, ארחות the highways והלכי and the travelers נתיבות and the travelers ילכו walked ארחות through byways. עקלקלות׃ through byways.}%
\verse{חדלו ceased, פרזון the villages בישׂראל in Israel, חדלו they ceased עד until that שׁקמתי arose, דבורה that I Deborah שׁקמתי that I arose אם a mother בישׂראל׃ in Israel.}%
\verse{יבחר They chose אלהים gods; חדשׁים new אז then לחם war שׁערים in the gates: מגן was there a shield אם יראה seen ורמח or spear בארבעים among forty אלף thousand בישׂראל׃ in Israel?}%
\verse{לבי My heart לחוקקי toward the governors ישׂראל of Israel, המתנדבים that offered themselves willingly בעם among the people. ברכו Bless יהוה׃ ye the LORD.}%
\verse{רכבי ye that ride אתנות asses, צחרות on white ישׁבי ye that sit על in מדין judgment, והלכי and walk על by דרך the way. שׂיחו׃ Speak,}%
\verse{מקול מחצצים of archers בין in משׁאבים the places of drawing water, שׁם there יתנו shall they rehearse צדקות the righteous acts יהוה of the LORD, צדקת the righteous acts פרזנו of his villages בישׂראל in Israel: אז then ירדו go down לשׁערים to the gates. עם shall the people יהוה׃ of the LORD}%
\verse{עורי Awake, עורי awake, דבורה Deborah: עורי awake, עורי awake, דברי utter שׁיר a song: קום arise, ברק Barak, ושׁבה and lead thy captivity captive, שׁביך and lead thy captivity captive, בן thou son אבינעם׃ of Abinoam.}%
\verse{אז Then ירד have dominion שׂריד he made him that remaineth לאדירים over the nobles עם among the people: יהוה the LORD ירד made me have dominion לי בגבורים׃ over the mighty.}%
\verse{מני Out of אפרים Ephraim שׁרשׁם a root בעמלק of them against Amalek; אחריך after בנימין thee, Benjamin, בעממיך among thy people; מני out of מכיר Machir ירדו came down מחקקים governors, ומזבולן משׁכים they that handle בשׁבט the pen ספר׃ of the writer.}%
\verse{ושׂרי And the princes בישׂשכר of Issachar עם with דברה Deborah; וישׂשכר even Issachar, כן and also ברק Barak: בעמק into the valley. שׁלח he was sent ברגליו on foot בפלגות For the divisions ראובן of Reuben גדלים great חקקי thoughts לב׃ of heart.}%
\verse{למה Why ישׁבת abodest בין thou among המשׁפתים the sheepfolds, לשׁמע to hear שׁרקות the bleatings עדרים of the flocks? לפלגות For the divisions ראובן of Reuben גדולים great חקרי searchings לב׃ of heart.}%
\verse{גלעד Gilead בעבר beyond הירדן Jordan: שׁכן abode ודן did Dan למה and why יגור remain אניות in ships? אשׁר Asher ישׁב continued לחוף shore, ימים on the sea ועל in מפרציו his breaches. ישׁכון׃ and abode}%
\verse{זבלון Zebulun עם a people חרף jeopardized נפשׁו their lives למות unto the death ונפתלי and Naphtali על in מרומי the high places שׂדה׃ of the field.}%
\verse{באו came מלכים The kings נלחמו fought, אז then נלחמו fought מלכי the kings כנען of Canaan בתענך in Taanach על by מי the waters מגדו of Megiddo; בצע כסף of money. לא no לקחו׃ they took}%
\verse{מן from שׁמים heaven; נלחמו They fought הכוכבים the stars ממסלותם in their courses נלחמו fought עם against סיסרא׃ Sisera.}%
\verse{נחל The river קישׁון of Kishon גרפם swept them away, נחל river, קדומים that ancient נחל the river קישׁון Kishon. תדרכי thou hast trodden down נפשׁי O my soul, עז׃ strength.}%
\verse{אז Then הלמו broken עקבי hooves סוס were the horses' מדהרות by the means of the prancings, דהרות the prancings אביריו׃ of their mighty ones.}%
\verse{אורו Curse מרוז ye Meroz, אמר said מלאך the angel יהוה of the LORD, ארו curse ye bitterly ארור curse ye bitterly ישׁביה the inhabitants כי thereof; because לא not באו they came לעזרת to the help יהוה of the LORD, לעזרת to the help יהוה of the LORD בגבורים׃ against the mighty.}%
\verse{תברך Blessed מנשׁים above women יעל shall Jael אשׁת the wife חבר of Heber הקיני the Kenite מנשׁים shall she be above women באהל in the tent. תברך׃ be, blessed}%
\verse{מים water, שׁאל He asked חלב milk; נתנה she gave בספל dish. אדירים in a lordly הקריבה she brought forth חמאה׃ butter}%
\verse{ידה her hand ליתד to the nail, תשׁלחנה She put וימינה and her right hand להלמות hammer; עמלים to the workmen's והלמה and with the hammer she smote סיסרא Sisera, מחקה she smote off ראשׁו his head, ומחצה when she had pierced וחלפה and stricken through רקתו׃}%
\verse{בין At רגליה her feet כרע he bowed, נפל he fell, שׁכב he lay down: בין at רגליה her feet כרע he bowed, נפל he fell: באשׁר where כרע he bowed, שׁם there נפל he fell down שׁדוד׃ dead.}%
\verse{בעד at החלון a window, נשׁקפה looked out ותיבב and cried אם The mother סיסרא of Sisera בעד through האשׁנב the lattice, מדוע Why בשׁשׁ is רכבו his chariot לבוא in coming? מדוע why אחרו tarry פעמי the wheels מרכבותיו׃ of his chariots?}%
\verse{חכמות Her wise שׂרותיה ladies תענינה answered אף her, yea, היא she תשׁיב returned אמריה׃}%
\verse{הלא Have they not ימצאו sped? יחלקו have they divided שׁלל the prey; רחם a damsel two; רחמתים לראשׁ to every גבר man שׁלל a prey צבעים of divers colors, לסיסרא to Sisera שׁלל a prey צבעים of divers colors רקמה of needlework, צבע of divers colors רקמתים of needlework לצוארי on both sides, for the necks שׁלל׃ of the spoil?}%
\verse{כן So יאבדו perish, כל let all אויביך thine enemies יהוה O LORD: ואהביו but them that love כצאת when he goeth forth השׁמשׁ him as the sun בגברתו in his might. ותשׁקט had rest הארץ And the land ארבעים forty שׁנה׃ years.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 6
\verseWithHeading{The Midianites Oppresses Israel}{ויעשׂו did בני And the children ישׂראל of Israel הרע evil בעיני in the sight יהוה of the LORD: ויתנם delivered יהוה and the LORD ביד them into the hand מדין of Midian שׁבע seven שׁנים׃ years.}%
\verse{ותעז prevailed יד And the hand מדין of Midian על against ישׂראל Israel: מפני because מדין of the Midianites עשׂו made להם בני the children ישׂראל of Israel את המנהרות them the dens אשׁר which בהרים in the mountains, ואת המערות and caves, ואת המצדות׃ and strongholds.}%
\verse{והיה And it was, אם when זרע had sown, ישׂראל Israel ועלה came up, מדין that the Midianites ועמלק ובני and the children קדם of the east, ועלו even they came up עליו׃ against}%
\verse{ויחנו And they encamped עליהם against וישׁחיתו them, and destroyed את יבול the increase הארץ of the earth, עד till בואך thou come עזה unto Gaza, ולא no ישׁאירו and left מחיה sustenance בישׂראל for Israel, ושׂה neither sheep, ושׁור nor ox, וחמור׃ nor ass.}%
\verse{כי For הם they ומקניהם with their cattle יעלו came up ואהליהם and their tents, יבאו and they came כדי ארבה grasshoppers לרב for multitude; ולהם ולגמליהם and their camels אין were without מספר number: ויבאו and they entered בארץ into the land לשׁחתה׃ to destroy}%
\verse{וידל impoverished ישׂראל And Israel מאד was greatly מפני because מדין of the Midianites; ויזעקו cried בני and the children ישׂראל of Israel אל unto יהוה׃ the LORD.}%
\verse{ויהי And it came to pass, כי when זעקו cried בני the children ישׂראל of Israel אל unto יהוה the LORD על because of אדות because of מדין׃ the Midianites,}%
\verse{וישׁלח sent יהוה That the LORD אישׁ נביא a prophet אל unto בני the children ישׂראל of Israel, ויאמר which said להם כה unto them, Thus אמר saith יהוה the LORD אלהי God ישׂראל of Israel, אנכי I העליתי אתכםמצרים ואציאתכם מבית out of the house עבדים׃ of bondage;}%
\verse{ואצל And I delivered אתכם מיד you out of the hand מצרים ומיד and out of the hand כל of all לחציכם that oppressed ואגרשׁ אותםפניכם from before ואתנה you, and gave לכם את ארצם׃ you their land;}%
\verse{ואמרה And I said לכם אני unto you, I יהוה the LORD אלהיכם your God; לא not תיראו fear את אלהי the gods האמרי of the Amorites, אשׁר in whose אתם ye יושׁבים dwell: בארצם land ולא but ye have not שׁמעתם obeyed בקולי׃ my voice.}%
\verseWithHeading{The Angel of Adonai Calls Gideon}{ויבא And there came מלאך an angel יהוה of the LORD, וישׁב and sat תחת under האלה an oak אשׁר which בעפרה in Ophrah, אשׁר that ליואשׁ unto Joash אבי העזרי the Abi-ezrite: וגדעון Gideon בנו and his son חבט threshed חטים wheat בגת by the winepress, להניס to hide מפני from מדין׃ the Midianites.}%
\verse{וירא appeared אליו unto מלאך And the angel יהוה of the LORD ויאמר him, and said אליו unto יהוה him, The LORD עמך with גבור thee, thou mighty man החיל׃ of valor.}%
\verse{ויאמר said אליו unto גדעון And Gideon בי him, Oh אדני my Lord, וישׁ be יהוה if the LORD עמנו with ולמה us, why מצאתנו befallen כל then is all זאת this ואיה us? and where כל all נפלאתיו his miracles אשׁר which ספרו told לנו אבותינו our fathers לאמר us of, saying, הלא Did not ממצרים העלנו bring us up יהוה the LORD ועתה but now נטשׁנו hath forsaken יהוה the LORD ויתננו us, and delivered בכף us into the hands מדין׃ of the Midianites.}%
\verse{ויפן looked אליו upon יהוה And the LORD ויאמר him, and said, לך Go in בכחך thy might, זה this והושׁעת and thou shalt save את ישׂראל Israel מכף from the hand מדין of the Midianites: הלא have not שׁלחתיך׃ I sent}%
\verse{ויאמר And he said אליו unto בי him, Oh אדני my Lord, במה wherewith אושׁיע shall I save את ישׂראל Israel? הנה behold, אלפי my family הדל poor במנשׁה in Manasseh, ואנכי and I הצעיר the least בבית house. אבי׃ in my father's}%
\verse{ויאמר said אליו unto יהוה And the LORD כי him, Surely אהיה I will be עמך with והכית thee, and thou shalt smite את מדין the Midianites כאישׁ man. אחד׃ as one}%
\verse{ויאמר And he said אליו unto אם him, If נא now מצאתי I have found חן grace בעיניך in thy sight, ועשׂית then show לי אות me a sign שׁאתה that thou מדבר talkest עמי׃ with}%
\verse{אל not נא I pray thee, תמשׁ Depart מזה hence, עד until באי I come אליך unto והצאתי thee, and bring forth את מנחתי my present, והנחתי לפניך before ויאמר thee. And he said, אנכי I אשׁב will tarry עד until שׁובך׃ thou come again.}%
\verse{וגדעון And Gideon בא went in, ויעשׂ and made ready גדי a kid, עזים a kid, ואיפת of an ephah קמח of flour: מצות and unleavened cakes הבשׂר the flesh שׂם he put בסל in a basket, והמרק the broth שׂם and he put בפרור in a pot, ויוצא and brought out אליו unto אל him under תחת him under האלה the oak, ויגשׁ׃ and presented}%
\verse{ויאמר said אליו unto מלאך And the angel האלהים of God קח him, Take את הבשׂר the flesh ואת המצות and the unleavened cakes, והנח אל upon הסלע rock, הלז this ואת המרק the broth. שׁפוך and pour out ויעשׂ And he did כן׃ so.}%
\verse{וישׁלח put forth מלאך Then the angel יהוה of the LORD את קצה the end המשׁענת of the staff אשׁר that בידו in his hand, ויגע and touched בבשׂר the flesh ובמצות and the unleavened cakes; ותעל and there rose up האשׁ fire מן out of הצור the rock, ותאכל and consumed את הבשׂר the flesh ואת המצות and the unleavened cakes. ומלאך Then the angel יהוה of the LORD הלך departed מעיניו׃ out of his sight.}%
\verse{וירא perceived גדעון And when Gideon כי that מלאך an angel יהוה of the LORD, הוא he ויאמר said, גדעון Gideon אהה Alas, אדני O Lord יהוה GOD! כי for על because כן because ראיתי I have seen מלאך an angel יהוה of the LORD פנים face אל to פנים׃ face.}%
\verse{ויאמר said לו יהוה And the LORD שׁלום unto him, Peace לך אל not: תירא unto thee; fear לא thou shalt not תמות׃ die.}%
\verse{ויבן built שׁם there גדעון Then Gideon מזבח an altar ליהוה unto the LORD, ויקרא and called לו יהוה שׁלוםד unto היום day הזה this עודנו it yet בעפרת in Ophrah אבי העזרי׃ of the Abi-ezrites.}%
\verse{ויהי And it came to pass בלילה night, ההוא the same ויאמר said לו יהוה that the LORD קח unto him, Take את פר young bullock, השׁור young bullock, אשׁר that לאביך thy father's ופר bullock השׁני even the second שׁבע of seven שׁנים years והרסת old, and throw down את מזבח the altar הבעל of Baal אשׁר that לאביך thy father ואת האשׁרה the grove אשׁר עליו by תכרת׃ hath, and cut down}%
\verse{ובנית And build מזבח an altar ליהוה unto the LORD אלהיך thy God על upon ראשׁ the top המעוז rock, הזה of this במערכה in the ordered place, ולקחת and take את הפר bullock, השׁני the second והעלית and offer עולה a burnt sacrifice בעצי with the wood האשׁרה of the grove אשׁר which תכרת׃ thou shalt cut down.}%
\verse{ויקח took גדעון Then Gideon עשׂרה ten אנשׁים men מעבדיו of his servants, ויעשׂ and did כאשׁר as דבר had said אליו unto יהוה the LORD ויהי him: and it was, כאשׁר because ירא he feared את בית household, אביו his father's ואת אנשׁי and the men העיר of the city, מעשׂות that he could not do יומם by day, ויעשׂ that he did לילה׃ by night.}%
\verseWithHeading{Gideon Destroys the Altar of Baal}{וישׁכימו arose early אנשׁי And when the men העיר of the city בבקר in the morning, והנה behold, נתץ was cast down, מזבח the altar הבעל of Baal והאשׁרה and the grove אשׁר that עליו by כרתה was cut down ואת הפר bullock השׁני it, and the second העלה was offered על upon המזבח the altar הבנוי׃ built.}%
\verse{ויאמרו And they said אישׁ one אל to רעהו another, מי Who עשׂה hath done הדבר thing? הזה this וידרשׁו And when they inquired ויבקשׁו and asked, ויאמרו they said, גדעון Gideon בן the son יואשׁ of Joash עשׂה hath done הדבר thing. הזה׃ this}%
\verse{ויאמרו said אנשׁי Then the men העיר of the city אל unto יואשׁ Joash, הוצא Bring out את בנך thy son, וימת that he may die: כי because נתץ he hath cast down את מזבח the altar הבעל of Baal, וכי and because כרת he hath cut down האשׁרה the grove אשׁר that עליו׃ by}%
\verse{ויאמר said יואשׁ And Joash לכל unto all אשׁר that עמדו stood עליו against האתם him, Will ye תריבון plead לבעל for Baal? אם if אתם will ye תושׁיעון save אותו אשׁר him? he that יריב will plead לו יומת for him, let him be put to death עד whilst הבקר morning: אם אלהים a god, הוא he ירב let him plead לו כי for himself, because נתץ hath cast down את מזבחו׃ his altar.}%
\verse{ויקרא he called לו ביום day ההוא Therefore on that ירבעל him Jerubbaal, לאמר saying, ירב plead בו הבעל Let Baal כי against him, because נתץ he hath thrown down את מזבחו׃ his altar.}%
\verse{וכל Then all מדין the Midianites ועמלק ובני and the children קדם of the east נאספו were gathered יחדו together, ויעברו and went over, ויחנו and pitched בעמק in the valley יזרעאל׃ of Jezreel.}%
\verse{ורוח But the Spirit יהוה of the LORD לבשׁה came upon את גדעון Gideon, ויתקע and he blew בשׁופר a trumpet; ויזעק was gathered אביעזר and Abi-ezer אחריו׃ after}%
\verse{ומלאכים messengers שׁלח And he sent בכל throughout all מנשׁה Manasseh; ויזעק was gathered גם also הוא who אחריו after ומלאכים messengers שׁלח him: and he sent באשׁר unto Asher, ובזבלון and unto Zebulun, ובנפתלי and unto Naphtali; ויעלו and they came up לקראתם׃ to meet}%
\verseWithHeading{Gideon Tests Adonai With the Fleece}{ויאמר said גדעון And Gideon אל unto האלהים God, אם If ישׁך מושׁיע thouwilt save בידי by mine hand, את ישׂראל Israel כאשׁר as דברת׃ thou hast said,}%
\verse{הנה Behold, אנכי I מציג will put את גזת a fleece הצמר of wool בגרן in the floor; אם if טל the dew יהיה be על on הגזה the fleece לבדה only, ועל upon כל all הארץ the earth חרב and dry וידעתי then shall I know כי that תושׁיע thou wilt save בידי by mine hand, את ישׂראל Israel כאשׁר as דברת׃ thou hast said.}%
\verse{ויהי And it was כן so: וישׁכם for he rose up early ממחרת on the morrow, ויזר and thrust את הגזה the fleece וימץ and wrung טל the dew מן on the morrow, הגזה the fleece, מלוא הספל a bowl מים׃ of water.}%
\verse{ויאמר said גדעון And Gideon אל unto האלהים God, אל Let not יחר be hot אפך thine anger בי ואדברה against me and I will speak אך but הפעם this once: אנסה let me prove, נא I pray thee, רק but הפעם this once בגזה with the fleece; יהי be נא let it now חרב dry אל upon הגזה the fleece, לבדה only ועל and upon כל all הארץ the ground יהיה let there be טל׃ dew.}%
\verse{ויעשׂ did אלהים And God כן so בלילה night: ההוא that ויהי for it was חרב dry אל upon הגזה the fleece לבדה only, ועל on כל all הארץ the ground. היה and there was טל׃ dew}%
\end{biblechapter}%
\begin{biblechapter}% Judges 7
\verseWithHeading{Gideon’s Three Hundred Men}{וישׁכם him, rose up early, ירבעל Then Jerubbaal, הוא who גדעון Gideon, וכל and all העם the people אשׁר that אתו with ויחנו and pitched על beside עין חרד the well ומחנה so that the host מדין of the Midianites היה were לו מצפון on the north side מגבעת of them, by the hill המורה of Moreh, בעמק׃ in the valley.}%
\verse{ויאמר said יהוה And the LORD אל unto גדעון Gideon, רב thee too many העם The people אשׁר that אתך with מתתי for me to give את מדין the Midianites בידם into their hands, פן lest יתפאר vaunt themselves עלי against ישׂראל Israel לאמר me, saying, ידי Mine own hand הושׁיעה׃ hath saved}%
\verse{ועתה קרא proclaim נא therefore go to, באזני in the ears העם of the people, לאמר saying, מי Whosoever ירא fearful וחרד and afraid, ישׁב let him return ויצפר and depart early מהר from mount הגלעד Gilead. וישׁב And there returned מן from mount העם the people עשׂרים twenty ושׁנים and two אלף thousand; ועשׂרת ten אלפים thousand. נשׁארו׃ and there remained}%
\verse{ויאמר said יהוה And the LORD אל unto גדעון Gideon, עוד yet העם The people רב many; הורד אותםל unto המים the water, ואצרפנו and I will try לך שׁם them for thee there: והיה and it shall be, אשׁר of whom אמר I say אליך unto זה thee, This ילך shall go אתך with הוא thee, the same ילך shall go אתך with וכל thee; and of whomsoever אשׁר thee; and of whomsoever אמר I say אליך unto זה thee, This לא shall not ילך go עמך with הוא thee, the same לא shall not ילך׃ go.}%
\verse{ויורד So he brought down את העם the people אל unto המים the water: ויאמר said יהוה and the LORD אל unto גדעון Gideon, כל Every one אשׁר that ילק lappeth בלשׁונו with his tongue, מן of המים the water כאשׁר as ילק lappeth, הכלב a dog תציג him shalt thou set אותו לבד by himself; וכל likewise every one אשׁר that יכרע boweth down על upon ברכיו his knees לשׁתות׃ to drink.}%
\verse{ויהי were מספר And the number המלקקים of them that lapped, בידם their hand אל to פיהם their mouth, שׁלשׁ three מאות hundred אישׁ men: וכל but all יתר the rest העם of the people כרעו bowed down על upon ברכיהם their knees לשׁתות to drink מים׃ water.}%
\verse{ויאמר said יהוה And the LORD אל unto גדעון Gideon, בשׁלשׁ By the three מאות hundred האישׁ men המלקקים that lapped אושׁיע will I save אתכם ונתתי you, and deliver את מדין the Midianites בידך into thine hand: וכל and let all העם the people ילכו go אישׁ every man למקמו׃ unto his place.}%
\verse{ויקחו took את צדה victuals העם So the people בידם in their hand, ואת שׁופרתיהם and their trumpets: ואת כל all אישׁ every man ישׂראל Israel שׁלח and he sent אישׁ men: לאהליו unto his tent, ובשׁלשׁ those three מאות hundred האישׁ החזיק and retained ומחנה and the host מדין of Midian היה was לו מתחת beneath בעמק׃ him in the valley.}%
\verse{ויהי And it came to pass בלילה night, ההוא the same ויאמר said אליו unto יהוה that the LORD קום him, Arise, רד get thee down במחנה unto the host; כי for נתתיו I have delivered בידך׃ it into thine hand.}%
\verse{ואם But if ירא אתה thou לרדת to go down, רד go אתה thou ופרה with Phurah נערך thy servant אל to המחנה׃ the host:}%
\verse{ושׁמעת And thou shalt hear מה what ידברו they say; ואחר and afterward תחזקנה be strengthened ידיך shall thine hands וירדת to go down במחנה unto the host. וירד Then went he down הוא ופרה with Phurah נערו his servant אל unto קצה the outside החמשׁים of the armed men אשׁר that במחנה׃ in the host.}%
\verse{ומדין And the Midianites ועמלק וכל and all בני the children קדם of the east נפלים lay בעמק along in the valley כארבה like grasshoppers לרב for multitude; ולגמליהם and their camels אין without מספר number, כחול as the sand שׁעל by שׂפת side הים the sea לרב׃ for multitude.}%
\verse{ויבא was come, גדעון And when Gideon והנה behold, אישׁ a man מספר that told לרעהו unto his fellow, חלום a dream ויאמר and said, הנה Behold, חלום a dream, חלמתי I dreamed והנה and, lo, צלול a cake לחם bread שׂערים of barley מתהפך tumbled במחנה into the host מדין of Midian, ויבא and came עד unto האהל a tent, ויכהו and smote ויפל it that it fell, ויהפכהו and overturned למעלה and overturned ונפל lay along. האהל׃ it, that the tent}%
\verse{ויען answered רעהו And his fellow ויאמר and said, אין nothing else זאת This בלתי save אם save חרב the sword גדעון of Gideon בן the son יואשׁ of Joash, אישׁ a man ישׂראל of Israel: נתן delivered האלהים hath God בידו into his hand את מדין Midian, ואת כל and all המחנה׃ the host.}%
\verse{ויהי And it was כשׁמע heard גדעון when Gideon את מספר the telling החלום of the dream, ואת שׁברו and the interpretation וישׁתחו thereof, that he worshiped, וישׁב and returned אל into מחנה the host ישׂראל of Israel, ויאמר and said, קומו Arise; כי for נתן hath delivered יהוה the LORD בידכם into your hand את מחנה the host מדין׃ of Midian.}%
\verse{ויחץ And he divided את שׁלשׁ the three מאות hundred האישׁ men שׁלשׁה three ראשׁים companies, ויתן and he put שׁופרות a trumpet ביד hand, כלם in every man's וכדים pitchers, רקים with empty ולפדים and lamps בתוך within הכדים׃ the pitchers.}%
\verse{ויאמר And he said אליהם unto ממני on תראו them, Look וכן likewise: תעשׂו me, and do והנה and, behold, אנכי when I בא come בקצה to the outside המחנה of the camp, והיה it shall be כאשׁר as אעשׂה I do, כן so תעשׂון׃ shall ye do.}%
\verse{ותקעתי When I blow בשׁופר with a trumpet, אנכי I וכל and all אשׁר that אתי with ותקעתם me, then blow בשׁופרות the trumpets גם also אתם ye סביבות on every side כל of all המחנה the camp, ואמרתם and say, ליהוה of the LORD, ולגדעון׃ and of Gideon.}%
\verse{ויבא him, came גדעון So Gideon, ומאה and the hundred אישׁ men אשׁר that אתו with בקצה unto the outside המחנה of the camp ראשׁ in the beginning האשׁמרת watch; התיכונה of the middle אך and they had but הקם newly set הקימו newly set את השׁמרים the watch: ויתקעו and they blew בשׁופרות the trumpets, ונפוץ and broke הכדים the pitchers אשׁר that בידם׃ in their hands.}%
\verse{ויתקעו blew שׁלשׁת And the three הראשׁים companies בשׁופרות the trumpets, וישׁברו and broke הכדים the pitchers, ויחזיקו and held ביד hands, שׂמאולם in their left בלפדים the lamps וביד hands ימינם in their right השׁופרות and the trumpets לתקוע to blow ויקראו and they cried, חרב The sword ליהוה of the LORD, ולגדעון׃ and of Gideon.}%
\verse{ויעמדו And they stood אישׁ every man תחתיו in his place סביב round about למחנה the camp: וירץ ran, כל and all המחנה the host ויריעו and cried, ויניסו׃ and fled.}%
\verse{ויתקעו blew שׁלשׁ And the three מאות hundred השׁופרות the trumpets, וישׂם set יהוה and the LORD את חרב sword אישׁ every man's ברעהו against his fellow, ובכל even throughout all המחנה the host: וינס fled המחנה and the host עד to בית השׁטה Beth-shittah צררתה in Zererath, עד to שׂפת the border אבל מחולה of Abel-meholah, על unto טבת׃ Tabbath.}%
\verse{ויצעק gathered themselves together אישׁ And the men ישׂראל of Israel מנפתלי ומן and out of אשׁר Asher, ומן and out of כל all מנשׁה Manasseh, וירדפו and pursued אחרי after מדין׃ the Midianites.}%
\verse{ומלאכים messengers שׁלח sent גדעון And Gideon בכל throughout all הר mount אפרים Ephraim, לאמר saying, רדו Come down לקראת against מדין the Midianites, ולכדו and take להם את המים before them the waters עד unto בית ברה Beth-barah ואת הירדן and Jordan. ויצעק gathered themselves together, כל Then all אישׁ the men אפרים of Ephraim וילכדו and took את המים the waters עד unto בית ברה Beth-barah ואת הירדן׃ and Jordan.}%
\verse{וילכדו And they took שׁני two שׂרי princes מדין of the Midianites, את ערב Oreb ואת זאב and Zeeb; ויהרגו and they slew את עורב Oreb בצור upon the rock עורב Oreb, ואת זאב and Zeeb הרגו they slew ביקב at the winepress זאב of Zeeb, וירדפו and pursued אל and pursued מדין Midian, וראשׁ the heads ערב of Oreb וזאב and Zeeb הביאו and brought אל to גדעון Gideon מעבר on the other side לירדן׃ Jordan.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 8
\verseWithHeading{Gideon Pursues Zebah and Zalmunna}{ויאמרו said אליו unto אישׁ And the men אפרים of Ephraim מה him, Why הדבר הזה us thus, עשׂית hast thou served לנו לבלתי us not, קראות that thou calledst לנו כי when הלכת thou wentest להלחם to fight במדין with the Midianites? ויריבון And they did chide אתו with בחזקה׃ him sharply.}%
\verse{ויאמר And he said אליהם unto מה them, What עשׂיתי have I done עתה now ככם הלוא in comparison of you? not טוב better עללות the gleaning of the grapes אפרים of Ephraim מבציר than the vintage אביעזר׃ of Abi-ezer?}%
\verse{בידכם into your hands נתן hath delivered אלהים God את שׂרי the princes מדין of Midian, את ערב Oreb ואת זאב and Zeeb: ומה and what יכלתי was I able עשׂות to do ככם אז in comparison of you? Then רפתה was abated רוחם their anger מעליו toward בדברו him, when he had said הדבר הזה׃ that.}%
\verse{ויבא came גדעון And Gideon הירדנה to Jordan, עבר passed over, הוא he, ושׁלשׁ and the three מאות hundred האישׁ men אשׁר that אתו with עיפים him, faint, ורדפים׃ yet pursuing}%
\verse{ויאמר And he said לאנשׁי סכות of Succoth, תנו Give, נא I pray you, ככרות loaves לחם of bread לעם unto the people אשׁר that ברגלי follow כי me; for עיפים faint, הם they ואנכי and I רדף am pursuing אחרי after זבח Zebah וצלמנע and Zalmunna, מלכי kings מדין׃ of Midian.}%
\verse{ויאמר said, שׂרי And the princes סכות of Succoth הכף the hands זבח of Zebah וצלמנע and Zalmunna עתה now בידך in thine hand, כי that נתן we should give לצבאך unto thine army? לחם׃ bread}%
\verse{ויאמר said, גדעון And Gideon לכן Therefore בתת hath delivered יהוה when the LORD את זבח Zebah ואת צלמנע and Zalmunna בידי into mine hand, ודשׁתי then I will tear את בשׂרכם your flesh את with קוצי the thorns המדבר of the wilderness ואת and with הברקנים׃ briers.}%
\verse{ויעל And he went up משׁם thence פנואל to Penuel, וידבר and spoke אליהם unto כזאת them likewise: ויענו answered אותו אנשׁי and the men פנואל of Penuel כאשׁר him as ענו had answered אנשׁי the men סכות׃ of Succoth}%
\verse{ויאמר And he spoke גם also לאנשׁי unto the men פנואל of Penuel, לאמר saying, בשׁובי When I come again בשׁלום in peace, אתץ I will break down את המגדל tower. הזה׃ this}%
\verse{וזבח Now Zebah וצלמנע and Zalmunna בקרקר in Karkor, ומחניהם and their hosts עמם with כחמשׁת them, about fifteen עשׂר them, about fifteen אלף thousand כל all הנותרים that were left מכל of all מחנה the hosts בני of the children קדם of the east: והנפלים for there fell מאה a hundred ועשׂרים and twenty אלף thousand אישׁ men שׁלף that drew חרב׃ sword.}%
\verse{ויעל went up גדעון And Gideon דרך by the way השׁכוני of them that dwelt באהלים in tents מקדם on the east לנבח of Nobah ויגבהה and Jogbehah, ויך and smote את המחנה the host: והמחנה for the host היה was בטח׃ secure.}%
\verse{וינוסו fled, זבח And when Zebah וצלמנע and Zalmunna וירדף he pursued אחריהם after וילכד them, and took את שׁני the two מלכי kings מדין of Midian, את זבח Zebah ואת צלמנע and Zalmunna, וכל all המחנה the host. החריד׃ and discomfited}%
\verse{וישׁב returned גדעון And Gideon בן the son יואשׁ of Joash מן from המלחמה battle מלמעלה before החרס׃ the sun}%
\verse{וילכד And caught נער a young man מאנשׁי סכות of Succoth, וישׁאלהו and inquired ויכתב of him: and he described אליו unto את שׂרי him the princes סכות of Succoth, ואת זקניה and the elders שׁבעים thereof, threescore and seventeen ושׁבעה thereof, threescore and seventeen אישׁ׃ of the men}%
\verse{ויבא And he came אל unto אנשׁי the men סכות of Succoth, ויאמר and said, הנה Behold זבח Zebah וצלמנע and Zalmunna, אשׁר with whom חרפתם ye did upbraid אותי לאמר me, saying, הכף the hands זבח of Zebah וצלמנע and Zalmunna עתה now בידך in thine hand, כי that נתן we should give לאנשׁיך unto thy men היעפים לחם׃ bread}%
\verse{ויקח And he took את זקני the elders העיר of the city, ואת קוצי and thorns המדבר of the wilderness ואת הברקנים and briers, וידע and with them he taught בהם את אנשׁי the men סכות׃ of Succoth.}%
\verse{ואת מגדל the tower פנואל of Penuel, נתץ And he beat down ויהרג and slew את אנשׁי the men העיר׃ of the city.}%
\verse{ויאמר Then said אל he unto זבח Zebah ואל צלמנע and Zalmunna, איפה What manner האנשׁים of men אשׁר whom הרגתם ye slew בתבור at Tabor? ויאמרו And they answered, כמוך As thou כמוהם אחד so they; each one כתאר resembled בני the children המלך׃ of a king.}%
\verse{ויאמר And he said, אחי my brethren, בני the sons אמי of my mother: הם They חי liveth, יהוה the LORD לו if החיתם אותםא I would not הרגתי slay אתכם׃}%
\verse{ויאמר And he said ליתר unto Jether בכורו his firstborn, קום Up, הרג slay אותם ולא not שׁלף drew הנער them. But the youth חרבו his sword: כי for ירא he feared, כי because עודנו he yet נער׃ a youth.}%
\verse{ויאמר said, זבח Then Zebah וצלמנע and Zalmunna קום Rise אתה thou, ופגע and fall בנו כי upon us: for כאישׁ as the man גבורתו his strength. ויקם arose, גדעון And Gideon ויהרג and slew את זבח Zebah ואת צלמנע and Zalmunna, ויקח and took away את השׂהרנים the ornaments אשׁר that בצוארי necks. גמליהם׃ on their camels'}%
\verse{ויאמרו said אישׁ Then the men ישׂראל of Israel אל unto גדעון Gideon, משׁל Rule בנו גם thou over us, both אתה thou, גם and בנך thy son, גם also: בן and thy son's בנך son כי for הושׁעתנו thou hast delivered מיד us from the hand מדין׃ of Midian.}%
\verse{ויאמר said אלהם unto גדעון And Gideon לא will not אמשׁל rule אני them, I בכם ולא over you, neither ימשׁל rule בני shall my son בכם יהוה over you: the LORD ימשׁל׃ shall rule}%
\verse{ויאמר said אלהם unto גדעון And Gideon אשׁאלה them, I would desire מכם of שׁאלה a request ותנו you, that ye would give לי אישׁ me every man נזם the earrings שׁללו of his prey. כי (For נזמי earrings, זהב they had golden להם כי because ישׁמעאלים Ishmaelites.) הם׃ they}%
\verse{ויאמרו And they answered, נתון נתןיפרשׂו And they spread את השׂמלה a garment, וישׁליכו and did cast שׁמה therein אישׁ every man נזם the earrings שׁללו׃ of his prey.}%
\verse{ויהי was משׁקל And the weight נזמי earrings הזהב of the golden אשׁר that שׁאל he requested אלף a thousand ושׁבע and seven מאות hundred זהב of gold; לבד beside מן beside השׂהרנים ornaments, והנטפות and collars, ובגדי raiment הארגמן and purple שׁעל that on מלכי the kings מדין of Midian, ולבד and beside מן and beside הענקות the chains אשׁר that בצוארי necks. גמליהם׃ about their camels'}%
\verse{ויעשׂ made אותו גדעון And Gideon לאפוד an ephod ויצג thereof, and put אותו בעירו it in his city, בעפרה in Ophrah: ויזנו went thither a whoring כל and all ישׂראל Israel אחריו after שׁם went thither a whoring ויהי it: which thing became לגדעון unto Gideon, ולביתו and to his house. למוקשׁ׃ a snare}%
\verse{ויכנע subdued מדין Thus was Midian לפני before בני the children ישׂראל of Israel, ולא no יספו more. לשׂאת so that they lifted up ראשׁם their heads ותשׁקט was in quietness הארץ And the country ארבעים forty שׁנה years בימי in the days גדעון׃ of Gideon.}%
\verseWithHeading{The Death of Gideon}{וילך went ירבעל And Jerubbaal בן the son יואשׁ of Joash וישׁב and dwelt בביתו׃ in his own house.}%
\verse{ולגדעון And Gideon היו had שׁבעים threescore and ten בנים sons יצאי begotten: ירכו of his body כי for נשׁים wives. רבות many היו׃ he had}%
\verse{ופילגשׁו And his concubine אשׁר that בשׁכם in Shechem, ילדה bore לו גם also היא she בן him a son, וישׂם he called את שׁמו whose name אבימלך׃ Abimelech.}%
\verse{וימת died גדעון And Gideon בן the son יואשׁ of Joash בשׂיבה old age, טובה in a good ויקבר and was buried בקבר in the sepulcher יואשׁ of Joash אביו his father, בעפרה in Ophrah אבי העזרי׃ of the Abi-ezrites.}%
\verse{ויהי And it came to pass, כאשׁר as soon as מת was dead, גדעון Gideon וישׁובו turned again, בני that the children ישׂראל of Israel ויזנו and went a whoring אחרי after הבעלים Baalim, וישׂימו and made להם בעל ברית Baal-berith לאלהים׃ their god.}%
\verse{ולא not זכרו remembered בני And the children ישׂראל of Israel את יהוה the LORD אלהיהם their God, המציל who had delivered אותם מיד them out of the hands כל of all איביהם their enemies מסביב׃ on every side:}%
\verse{ולא Neither עשׂו showed חסד they kindness עם to בית the house ירבעל of Jerubbaal, גדעון Gideon, ככל according to all הטובה the goodness אשׁר which עשׂה he had showed עם unto ישׂראל׃ Israel.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 9
\verseWithHeading{Abimelech Attempts to Become King}{וילך went אבימלך And Abimelech בן the son ירבעל of Jerubbaal שׁכמה to Shechem אל unto אחי brethren, אמו his mother's וידבר and communed אליהם with ואל them, and with כל all משׁפחת the family בית of the house אבי father, אמו of his mother's לאמר׃ saying,}%
\verse{דברו Speak, נא I pray you, באזני in the ears כל of all בעלי the men שׁכם of Shechem, מה Whether טוב better לכם המשׁל reign בכם שׁבעים threescore and ten אישׁ persons, כל for you, either that all בני the sons ירבעל of Jerubbaal, אם over you, or משׁל reign בכם אישׁ אחד that one וזכרתם over you? remember כי also that עצמכם your bone ובשׂרכם and your flesh. אני׃ I}%
\verse{וידברו spoke אחי brethren אמו And his mother's עליו of באזני him in the ears כל of all בעלי the men שׁכם of Shechem את כל all הדברים words: האלה these ויט inclined to follow לבם and their hearts אחרי inclined to follow אבימלך Abimelech; כי for אמרו they said, אחינו our brother. הוא׃ He}%
\verse{ויתנו And they gave לו שׁבעים him threescore and ten כסף of silver מבית out of the house בעל ברית of Baal-berith, וישׂכר hired בהם אבימלך wherewith Abimelech אנשׁים ריקים vain ופחזים and light וילכו which followed אחריו׃ which followed}%
\verse{ויבא And he went בית house אביו unto his father's עפרתה at Ophrah, ויהרג and slew את אחיו his brethren בני the sons ירבעל of Jerubbaal, שׁבעים threescore and ten אישׁ persons, על upon אבן stone: אחת one ויותר was left; יותם notwithstanding yet Jotham בן son ירבעל of Jerubbaal הקטן the youngest כי for נחבא׃ he hid himself.}%
\verse{ויאספו gathered together, כל And all בעלי the men שׁכם of Shechem וכל and all בית מלואילכו and went, וימליכו and made את אבימלך Abimelech למלך king, עם by אלון the plain מצב of the pillar אשׁר that בשׁכם׃ in Shechem.}%
\verse{ויגדו And when they told ליותם to Jotham, וילך he went ויעמד and stood בראשׁ in the top הר of mount גרזים Gerizim, וישׂא and lifted up קולו his voice, ויקרא and cried, ויאמר and said להם שׁמעו unto them, Hearken אלי unto בעלי me, ye men שׁכם of Shechem, וישׁמע may hearken אליכם unto אלהים׃ that God}%
\verse{הלוך went forth הלכו went forth העצים The trees למשׁח to anoint עליהם over מלך a king ויאמרו them; and they said לזית unto the olive tree, מלוכה Reign עלינו׃ thou over}%
\verse{ויאמר said להם הזית But the olive tree החדלתי unto them, Should I leave את דשׁני my fatness, אשׁר wherewith בי יכבדו by me they honor אלהים God ואנשׁים and man, והלכתי and go לנוע to be promoted על over העצים׃ the trees?}%
\verse{ויאמרו said העצים And the trees לתאנה to the fig tree, לכי Come את thou, מלכי reign עלינו׃ over}%
\verse{ותאמר said להם התאנה But the fig tree החדלתי unto them, Should I forsake את מתקי my sweetness, ואת תנובתי fruit, הטובה and my good והלכתי and go לנוע to be promoted על over העצים׃ the trees?}%
\verse{ויאמרו Then said העצים the trees לגפן unto the vine, לכי Come את thou, מלוכי reign עלינו׃ over}%
\verse{ותאמר said להם הגפן And the vine החדלתי unto them, Should I leave את תירושׁי my wine, המשׂמח which cheereth אלהים God ואנשׁים and man, והלכתי and go לנוע to be promoted על over העצים׃ the trees?}%
\verse{ויאמרו Then said כל all העצים the trees אל unto האטד the bramble, לך Come אתה thou, מלך reign עלינו׃ over}%
\verse{ויאמר said האטד And the bramble אל unto העצים the trees, אם If באמת in truth אתם ye משׁחים anoint אתי למלך me king עליכם over באו you, come חסו put your trust בצלי in my shadow: ואם and if אין not, תצא come out אשׁ let fire מן of האטד the bramble, ותאכל and devour את ארזי the cedars הלבנון׃ of Lebanon.}%
\verse{ועתה Now אם therefore, if באמת truly ובתמים and sincerely, עשׂיתם ye have done ותמליכו אתבימלך ואם and if טובה עשׂיתם ye have dealt עם with ירבעל Jerubbaal ועם ביתו and his house, ואם כגמול unto him according to the deserving ידיו of his hands; עשׂיתם׃ and have done}%
\verse{אשׁר (For נלחם fought אבי my father עליכם for וישׁלך you, and adventured את נפשׁו his life מנגד far, ויצל and delivered אתכם מיד you out of the hand מדין׃ of Midian:}%
\verse{ואתם And ye קמתם are risen up על against בית house אבי my father's היום this day, ותהרגו and have slain את בניו his sons, שׁבעים threescore and ten אישׁ persons, על upon אבן stone, אחת one ותמליכו king את אבימלך and have made Abimelech, בן the son אמתו of his maidservant, על over בעלי the men שׁכם of Shechem, כי because אחיכם your brother;) הוא׃ he}%
\verse{ואם If באמת truly ובתמים and sincerely עשׂיתם ye then have dealt עם with ירבעל Jerubbaal ועם and with ביתו his house היום day, הזה this שׂמחו rejoice באבימלך ye in Abimelech, וישׂמח rejoice גם also הוא׃ and let him}%
\verse{ואם But if אין not, תצא come out אשׁ let fire מאבימלך Abimelech. ותאכל and devour את בעלי the men שׁכם of Shechem, ואת בית and the house מלוא of Millo; ותצא come out אשׁ and let fire מבעלי from the men שׁכם of Shechem, ומבית and from the house מלוא of Millo, ותאכל and devour את אבימלך׃}%
\verse{וינס ran away, יותם And Jotham ויברח and fled, וילך and went בארה to Beer, וישׁב and dwelt שׁם there, מפני for fear אבימלך of Abimelech אחיו׃ his brother.}%
\verseWithHeading{The Downfall of Shechem and Abimelech}{וישׂר had reigned אבימלך When Abimelech על over ישׂראל Israel, שׁלשׁ three שׁנים׃ years}%
\verse{וישׁלח sent אלהים Then God רוח spirit רעה an evil בין between אבימלך Abimelech ובין בעלי and the men שׁכם of Shechem; ויבגדו dealt treacherously בעלי and the men שׁכם of Shechem באבימלך׃ with Abimelech:}%
\verse{לבוא might come, חמס That the cruelty שׁבעים to the threescore and ten בני sons ירבעל of Jerubbaal ודמם and their blood לשׂום be laid על upon אבימלך Abimelech אחיהם their brother, אשׁר which הרג slew אותם ועל them; and upon בעלי the men שׁכם of Shechem, אשׁר which חזקו אתדיו להרג him in the killing את אחיו׃ of his brethren.}%
\verse{וישׂימו set לו בעלי And the men שׁכם of Shechem מארבים liers in wait על for him in ראשׁי the top ההרים of the mountains, ויגזלו and they robbed את כל all אשׁר that יעבר came along עליהם by בדרך that way ויגד them: and it was told לאבימלך׃ Abimelech.}%
\verse{ויבא came געל And Gaal בן the son עבד of Ebed ואחיו with his brethren, ויעברו and went over בשׁכם to Shechem: ויבטחו put their confidence בו בעלי and the men שׁכם׃ of Shechem}%
\verse{ויצאו And they went out השׂדה into the fields, ויבצרו and gathered את כרמיהם their vineyards, וידרכו and trod ויעשׂו and made הלולים merry, ויבאו and went into בית the house אלהיהם of their god, ויאכלו and did eat וישׁתו and drink, ויקללו and cursed את אבימלך׃ Abimelech.}%
\verse{ויאמר said, געל And Gaal בן the son עבד of Ebed מי Who אבימלך Abimelech, ומי and who שׁכם Shechem, כי that נעבדנו we should serve הלא him? not בן the son ירבעל of Jerubbaal? וזבל and Zebul פקידו his officer? עבדו serve את אנשׁי the men חמור of Hamor אבי the father שׁכם ומדוע for why נעבדנו serve אנחנו׃ should we}%
\verse{ומי יתןת העם people הזה this בידי were under my hand! ואסירה then would I remove את אבימלך Abimelech. ויאמר And he said לאבימלך to Abimelech, רבה Increase צבאך thine army, וצאה׃ and come out.}%
\verse{וישׁמע heard זבל And when Zebul שׂר the ruler העיר of the city את דברי the words געל of Gaal בן the son עבד of Ebed, ויחר was kindled. אפו׃ his anger}%
\verse{וישׁלח And he sent מלאכים messengers אל unto אבימלך Abimelech בתרמה privily, לאמר saying, הנה Behold, געל Gaal בן the son עבד of Ebed ואחיו and his brethren באים be come שׁכמה to Shechem; והנם and, behold, צרים they fortify את העיר the city עליך׃ against}%
\verse{ועתה Now קום therefore up לילה by night, אתה thou והעם and the people אשׁר that אתך with וארב thee, and lie in wait בשׂדה׃ in the field:}%
\verse{והיה And it shall be, בבקר in the morning, כזרח is up, השׁמשׁ as soon as the sun תשׁכים thou shalt rise early, ופשׁטת and set על upon העיר the city: והנה and, behold, הוא he והעם and the people אשׁר that אתו with יצאים him come out אליך against ועשׂית thee, then mayest thou do לו כאשׁר to them as תמצא thou shalt find ידך׃ occasion.}%
\verse{ויקם rose up, אבימלך And Abimelech וכל and all העם the people אשׁר that עמו with לילה him, by night, ויארבו and they laid wait על against שׁכם Shechem ארבעה in four ראשׁים׃ companies.}%
\verse{ויצא went out, געל And Gaal בן the son עבד of Ebed ויעמד and stood פתח in the entering שׁער of the gate העיר of the city: ויקם rose up, אבימלך and Abimelech והעם and the people אשׁר that אתו with מן him, from המארב׃ lying in wait.}%
\verse{וירא saw געל And when Gaal את העם the people, ויאמר he said אל to זבל Zebul, הנה Behold, עם there come people down יורד there come people down מראשׁי from the top ההרים of the mountains. ויאמר said אליו unto זבל And Zebul את צל the shadow ההרים of the mountains אתה him, Thou ראה seest כאנשׁים׃ as men.}%
\verse{ויסף again עוד געל And Gaal לדבר spoke ויאמר and said, הנה See עם there come people down יורדים there come people down מעם by טבור the middle הארץ of the land, וראשׁ company אחד and another בא come מדרך אלון the plain מעוננים׃ of Meonenim.}%
\verse{ויאמר Then said אליו unto זבל Zebul איה him, Where אפוא now פיך thy mouth, אשׁר wherewith תאמר thou saidst, מי Who אבימלך Abimelech, כי that נעבדנו we should serve הלא him? not זה this העם the people אשׁר that מאסתה thou hast despised? בו צא go out, נא I pray עתה now, והלחם׃ and fight}%
\verse{ויצא went out געל And Gaal לפני before בעלי the men שׁכם of Shechem, וילחם and fought באבימלך׃ with Abimelech.}%
\verse{וירדפהו chased אבימלך And Abimelech וינס him, and he fled מפניו before ויפלו were overthrown חללים wounded, רבים him, and many עד unto פתח the entering השׁער׃ of the gate.}%
\verse{וישׁב dwelt אבימלך And Abimelech בארומה at Arumah: ויגרשׁ thrust out זבל and Zebul את געל Gaal ואת אחיו and his brethren, משׁבת that they should not dwell בשׁכם׃ in Shechem.}%
\verse{ויהי And it came to pass ממחרת on the morrow, ויצא went out העם that the people השׂדה into the field; ויגדו and they told לאבימלך׃ Abimelech.}%
\verse{ויקח And he took את העם the people, ויחצם and divided לשׁלשׁה them into three ראשׁים companies, ויארב and laid wait בשׂדה in the field, וירא and looked, והנה and, behold, העם the people יצא come forth מן out of העיר the city; ויקם and he rose up עליהם against ויכם׃ them, and smote}%
\verse{ואבימלך And Abimelech, והראשׁים and the company אשׁר that עמו with פשׁטו him, rushed forward, ויעמדו and stood פתח in the entering שׁער of the gate העיר of the city: ושׁני and the two הראשׁים companies פשׁטו ran על upon כל all אשׁר that בשׂדה in the fields, ויכום׃ and slew}%
\verse{ואבימלך And Abimelech נלחם fought בעיר against the city כל all היום day; ההוא that וילכד and he took את העיר the city, ואת העם the people אשׁר that בה הרג and slew ויתץ therein, and beat down את העיר the city, ויזרעה and sowed מלח׃ it with salt.}%
\verse{וישׁמעו heard כל And when all בעלי the men מגדל of the tower שׁכם of Shechem ויבאו they entered אל into צריח a hold בית of the house אל of the god ברית׃ Berith.}%
\verse{ויגד And it was told לאבימלך Abimelech, כי that התקבצו were gathered together. כל all בעלי the men מגדל of the tower שׁכם׃ of Shechem}%
\verse{ויעל got him up אבימלך And Abimelech הר to mount צלמון Zalmon, הוא he וכל and all העם the people אשׁר that אתו with ויקח took אבימלך him; and Abimelech את הקרדמות an axe בידו in his hand, ויכרת and cut down שׂוכת a bough עצים from the trees, וישׂאה and took וישׂם it, and laid על on שׁכמו his shoulder, ויאמר and said אל unto העם the people אשׁר that עמו with מה him, What ראיתם ye have seen עשׂיתי me do, מהרו make haste, עשׂו do כמוני׃ as I}%
\verse{ויכרתו cut down גם likewise כל And all העם the people אישׁ every man שׂוכה his bough, וילכו and followed אחרי and followed אבימלך Abimelech, וישׂימו and put על to הצריח the hold, ויציתו and set עליהם upon את הצריח the hold באשׁ on fire וימתו died גם also, כל them; so that all אנשׁי the men מגדל of the tower שׁכם of Shechem כאלף about a thousand אישׁ men ואשׁה׃ and women.}%
\verse{וילך Then went אבימלך Abimelech אל to תבץ Thebez, ויחן and encamped בתבץ against Thebez, וילכדה׃ and took}%
\verse{ומגדל tower עז a strong היה But there was בתוך within העיר the city, וינסו fled שׁמה and thither כל all האנשׁים והנשׁים and women, וכל and all בעלי they העיר of the city, ויסגרו and shut בעדם to ויעלו them, and got them up על to גג the top המגדל׃ of the tower.}%
\verse{ויבא came אבימלך And Abimelech עד unto המגדל the tower, וילחם and fought בו ויגשׁ against it, and went hard עד unto פתח the door המגדל of the tower לשׂרפו to burn באשׁ׃ it with fire.}%
\verse{ותשׁלך cast אשׁה woman אחת And a certain פלח a piece רכב of a millstone על upon ראשׁ head, אבימלך Abimelech's ותרץ and all to broke את גלגלתו׃ his skull.}%
\verse{ויקרא Then he called מהרה hastily אל unto הנער the young man נשׂא his armorbearer, כליו his armorbearer, ויאמר and said לו שׁלף unto him, Draw חרבך thy sword, ומותתני and slay פן me, that יאמרו men say לי אשׁה not of me, A woman הרגתהו slew וידקרהו thrust him through, נערו him. And his young man וימת׃ and he died.}%
\verse{ויראו saw אישׁ And when the men ישׂראל of Israel כי that מת was dead, אבימלך Abimelech וילכו they departed אישׁ every man למקמו׃ unto his place.}%
\verse{וישׁב rendered אלהים Thus God את רעת the wickedness אבימלך of Abimelech, אשׁר which עשׂה he did לאביו unto his father, להרג in slaying את שׁבעים his seventy אחיו׃ brethren:}%
\verse{ואת כל And all רעת the evil אנשׁי of the men שׁכם of Shechem השׁיב render אלהים did God בראשׁם upon their heads: ותבא them came אליהם and upon קללת the curse יותם of Jotham בן the son ירבעל׃ of Jerubbaal.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 10
\verseWithHeading{The Philistines and Ammonites Afflict the Israelites}{ויקם there arose אחרי And after אבימלך Abimelech להושׁיע to defend את ישׂראל Israel תולע Tola בן the son פואה of Puah, בן the son דודו of Dodo, אישׁ a man ישׂשכר of Issachar; והוא and he ישׁב dwelt בשׁמיר in Shamir בהר in mount אפרים׃ Ephraim.}%
\verse{וישׁפט And he judged את ישׂראל Israel עשׂרים twenty ושׁלשׁ and three שׁנה years, וימת and died, ויקבר and was buried בשׁמיר׃ in Shamir.}%
\verse{ויקם him arose אחריו And after יאיר Jair, הגלעדי a Gileadite, וישׁפט and judged את ישׂראל Israel עשׂרים twenty ושׁתים and two שׁנה׃ years.}%
\verse{ויהי And he had לו שׁלשׁים thirty בנים sons רכבים that rode על on שׁלשׁים thirty עירים ass colts, ושׁלשׁים and they had thirty עירים להםהם יקראו which are called חות יאיר Havoth-jair עד unto היום day, הזה this אשׁר which בארץ in the land הגלעד׃ of Gilead.}%
\verse{וימת died, יאיר And Jair ויקבר and was buried בקמון׃ in Camon.}%
\verse{ויספו again בני And the children ישׂראל of Israel לעשׂות did הרע evil בעיני in the sight יהוה of the LORD, ויעבדו and served את הבעלים Baalim, ואת העשׁתרות and Ashtaroth, ואת אלהי and the gods ארם of Syria, ואת אלהי and the gods צידון of Zidon, ואת אלהי and the gods מואב of Moab, ואת אלהי and the gods בני of the children עמון of Ammon, ואת אלהי and the gods פלשׁתים of the Philistines, ויעזבו and forsook את יהוה the LORD, ולא not עבדוהו׃ and served}%
\verse{ויחר was hot אף And the anger יהוה of the LORD בישׂראל against Israel, וימכרם and he sold ביד them into the hands פלשׁתים of the Philistines, וביד and into the hands בני of the children עמון׃ of Ammon.}%
\verse{וירעצו they vexed וירצצו and oppressed את בני the children ישׂראל of Israel: בשׁנה year ההיא And that שׁמנה eighteen עשׂרה eighteen שׁנה years, את כל all בני the children ישׂראל of Israel אשׁר that בעבר on the other side הירדן Jordan בארץ in the land האמרי of the Amorites, אשׁר which בגלעד׃ in Gilead.}%
\verse{ויעברו passed over בני Moreover the children עמון of Ammon את הירדן Jordan להלחם to fight גם also ביהודה against Judah, ובבנימין and against Benjamin, ובבית and against the house אפרים of Ephraim; ותצר distressed. לישׂראל so that Israel מאד׃ was sore}%
\verse{ויזעקו cried בני And the children ישׂראל of Israel אל unto יהוה the LORD, לאמר saying, חטאנו We have sinned לך וכי against thee, both because עזבנו we have forsaken את אלהינו our God, ונעבד and also served את הבעלים׃ Baalim.}%
\verse{ויאמר said יהוה And the LORD אל בני the children ישׂראל of Israel, הלא not ממצרים ומן and from האמרי the Amorites, ומן from בני the children עמון of Ammon, ומן and from פלשׁתים׃ the Philistines?}%
\verse{וצידונים The Zidonians ועמלק also, and the Amalekites, ומעון and the Maonites, לחצו did oppress אתכם ותצעקו you; and ye cried אלי to ואושׁיעה me, and I delivered אתכם מידם׃ you out of their hand.}%
\verse{ואתם Yet ye עזבתם have forsaken אותי ותעבדו me, and served אלהים gods: אחרים other לכן wherefore לא you no אוסיף more. להושׁיע I will deliver אתכם׃}%
\verse{לכו Go וזעקו and cry אל unto האלהים the gods אשׁר which בחרתם ye have chosen; בם המה let them יושׁיעו deliver לכם בעת you in the time צרתכם׃ of your tribulation.}%
\verse{ויאמרו said בני And the children ישׂראל of Israel אל unto יהוה the LORD, חטאנו We have sinned: עשׂה do אתה thou לנו ככל unto us whatsoever הטוב seemeth good בעיניך unto אך us only, הצילנו thee; deliver נא we pray thee, היום day. הזה׃ this}%
\verse{ויסירו And they put away את אלהי gods הנכר the strange מקרבם from among ויעבדו them, and served את יהוה the LORD: ותקצר was grieved נפשׁו and his soul בעמל for the misery ישׂראל׃ of Israel.}%
\verse{ויצעקו were gathered together, בני Then the children עמון of Ammon ויחנו and encamped בגלעד in Gilead. ויאספו assembled themselves together, בני And the children ישׂראל of Israel ויחנו and encamped במצפה׃ in Mizpeh.}%
\verse{ויאמרו said העם And the people שׂרי princes גלעד of Gilead אישׁ one אל to רעהו another, מי What האישׁ man אשׁר that יחל will begin להלחם to fight בבני against the children עמון of Ammon? יהיה he shall be לראשׁ head לכל over all ישׁבי the inhabitants גלעד׃ of Gilead.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 11
\verseWithHeading{Jephthah}{ויפתח Now Jephthah הגלעדי the Gileadite היה was גבור a mighty man חיל of valor, והוא and he בן the son אשׁה of a harlot: זונה of a harlot: ויולד begot גלעד and Gilead את יפתח׃ Jephthah.}%
\verse{ותלד bore אשׁת wife גלעד And Gilead's לו בנים him sons; ויגדלו grew up, בני sons האשׁה and his wife's ויגרשׁו and they thrust out את יפתח Jephthah, ויאמרו and said לו לא unto him, Thou shalt not תנחל inherit בבית house; אבינו in our father's כי for בן the son אשׁה woman. אחרת of a strange אתה׃ thou}%
\verse{ויברח fled יפתח Then Jephthah מפני from אחיו his brethren, וישׁב and dwelt בארץ in the land טוב of Tob: ויתלקטו and there were gathered אל to יפתח Jephthah, אנשׁים ריקים vain ויצאו and went out עמו׃ with}%
\verse{ויהי And it came to pass מימים in process of time, וילחמו made war בני that the children עמון of Ammon עם against ישׂראל׃ Israel.}%
\verse{ויהי And it was כאשׁר so, that when נלחמו made war בני the children עמון of Ammon עם against ישׂראל Israel, וילכו went זקני the elders גלעד of Gilead לקחת to fetch את יפתח Jephthah מארץ out of the land טוב׃ of Tob:}%
\verse{ויאמרו And they said ליפתח unto Jephthah, לכה Come, והייתה and be לנו לקצין our captain, ונלחמה that we may fight בבני with the children עמון׃ of Ammon.}%
\verse{ויאמר said יפתח And Jephthah לזקני unto the elders גלעד of Gilead, הלא Did not אתם ye שׂנאתם hate אותי ותגרשׁוני me, and expel מבית house? אבי me out of my father's ומדוע and why באתם are ye come אלי unto עתה me now כאשׁר when צר׃ ye are in distress?}%
\verse{ויאמרו said זקני And the elders גלעד of Gilead אל unto יפתח Jephthah, לכן Therefore עתה thee now, שׁבנו we turn again אליך to והלכת that thou mayest go עמנו with ונלחמת us, and fight בבני against the children עמון of Ammon, והיית and be לנו לראשׁ our head לכל over all ישׁבי the inhabitants גלעד׃ of Gilead.}%
\verse{ויאמר said יפתח And Jephthah אל unto זקני the elders גלעד of Gilead, אם If משׁיבים אתם ye אותי להלחם to fight בבני against the children עמון of Ammon, ונתן deliver יהוה and the LORD אותם לפני them before אנכי me, shall I אהיה be לכם לראשׁ׃ your head?}%
\verse{ויאמרו said זקני And the elders גלעד of Gilead אל unto יפתח Jephthah, יהוה The LORD יהיה be שׁמע witness בינותינו between אם us, if לא not כדברך according to thy words. כן so נעשׂה׃ we do}%
\verse{וילך went יפתח Then Jephthah עם with זקני the elders גלעד of Gilead, וישׂימו made העם and the people אותו עליהם over לראשׁ him head ולקצין and captain וידבר uttered יפתח them: and Jephthah את כל all דבריו his words לפני before יהוה the LORD במצפה׃ in Mizpeh.}%
\verse{וישׁלח sent יפתח And Jephthah מלאכים messengers אל unto מלך the king בני of the children עמון of Ammon, לאמר saying, מה What לי ולך כי hast thou to do with me, that באת thou art come אלי against להלחם me to fight בארצי׃ in my land?}%
\verse{ויאמר answered מלך And the king בני of the children עמון of Ammon אל unto מלאכי the messengers יפתח of Jephthah, כי Because לקח took away ישׂראל Israel את ארצי my land, בעלותו when they came up ממצרים מארנוןעד even unto היבק Jabbok, ועד and unto הירדן Jordan: ועתה now השׁיבה אתהןשׁלום׃ peaceably.}%
\verse{ויוסף again עוד again יפתח And Jephthah וישׁלח sent מלאכים messengers אל unto מלך the king בני of the children עמון׃ of Ammon:}%
\verse{ויאמר And said לו כה unto him, Thus אמר saith יפתח Jephthah, לא took not away לקח took not away ישׂראל Israel את ארץ the land מואב of Moab, ואת ארץ nor the land בני of the children עמון׃ of Ammon:}%
\verse{כי But בעלותם came up ממצרים וילך and walked ישׂראל when Israel במדבר through the wilderness עד unto ים sea, סוף the Red ויבא and came קדשׁה׃ to Kadesh;}%
\verse{וישׁלח sent ישׂראל Then Israel מלאכים messengers אל unto מלך the king אדום of Edom, לאמר saying, אעברה pass נא Let me, I pray thee, בארצך through thy land: ולא would not שׁמע hearken מלך but the king אדום of Edom וגם And in like manner אל unto מלך the king מואב of Moab: שׁלח they sent ולא not אבה but he would וישׁב abode ישׂראל and Israel בקדשׁ׃ in Kadesh.}%
\verse{וילך Then they went along במדבר through the wilderness, ויסב and compassed את ארץ the land אדום of Edom, ואת ארץ and the land מואב of Moab, ויבא and came ממזרח שׁמשׁארץ of the land מואב of Moab, ויחנון and pitched בעבר on the other side ארנון of Arnon, ולא not באו but came בגבול within the border מואב of Moab: כי for ארנון Arnon גבול the border מואב׃ of Moab.}%
\verse{וישׁלח sent ישׂראל And Israel מלאכים messengers אל unto סיחון Sihon מלך king האמרי of the Amorites, מלך the king חשׁבון of Heshbon; ויאמר said לו ישׂראל and Israel נעברה unto him, Let us pass, נא we pray thee, בארצך through thy land עד into מקומי׃ my place.}%
\verse{ולא not האמין trusted סיחון But Sihon את ישׂראל Israel עבר to pass בגבלו through his coast: ויאסף סיחון but Sihon את כלמו ויחנו and pitched ביהצה in Jahaz, וילחם and fought עם against ישׂראל׃ Israel.}%
\verse{ויתן delivered יהוה And the LORD אלהי God ישׂראל of Israel את סיחון Sihon ואת כל and all עמו his people ביד into the hand ישׂראל of Israel, ויכום and they smote ויירשׁ possessed ישׂראל them: so Israel את כל all ארץ the land האמרי of the Amorites, יושׁב the inhabitants הארץ country. ההיא׃ of that}%
\verse{ויירשׁו And they possessed את כל all גבול the coasts האמרי of the Amorites, מארנון ועד even unto היבק Jabbok, ומן and from המדבר the wilderness ועד even unto הירדן׃ Jordan.}%
\verse{ועתה So now יהוה the LORD אלהי God ישׂראל of Israel הורישׁ hath dispossessed את האמרי the Amorites מפני from before עמו his people ישׂראל Israel, ואתה and shouldest thou תירשׁנו׃ possess}%
\verse{הלא Wilt not את אשׁר that which יורישׁך thou possess כמושׁ Chemosh אלהיך thy god אותו תירשׁ giveth thee to possess? ואת כלשׁר הורישׁ shall drive out יהוה the LORD אלהינו our God מפנינו from before אותו נירשׁ׃ us, them will we possess.}%
\verse{ועתה And now הטוב any thing better טוב any thing better אתה thou מבלק בן the son צפור of Zippor, מלך king מואב of Moab? הרוב did he ever strive רב did he ever strive עם against ישׂראל Israel, אם or נלחם did he ever fight נלחם׃ did he ever fight}%
\verse{בשׁבת dwelt ישׂראל While Israel בחשׁבון in Heshbon ובבנותיה and her towns, ובערעור and in Aroer ובבנותיה and her towns, ובכל and in all הערים the cities אשׁר that על along by ידי the coasts ארנון of Arnon, שׁלשׁ three מאות hundred שׁנה years? ומדוע why לא therefore did ye not הצלתם recover בעת time? ההיא׃ within that}%
\verse{ואנכי Wherefore I לא have not חטאתי sinned לך ואתה against thee, but thou עשׂה doest אתי רעה me wrong להלחם to war בי ישׁפט the Judge יהוה against me: the LORD השׁפט be judge היום this day בין between בני the children ישׂראל of Israel ובין בני and the children עמון׃ of Ammon.}%
\verse{ולא not שׁמע hearkened מלך Howbeit the king בני of the children עמון of Ammon אל unto דברי the words יפתח of Jephthah אשׁר which שׁלח he sent אליו׃ he sent}%
\verseWithHeading{Jephthah Makes a Vow}{ותהי came על upon יפתח Jephthah, רוח Then the Spirit יהוה of the LORD ויעבר and he passed over את הגלעד Gilead, ואת מנשׁה and Manasseh, ויעבר and passed over את מצפה Mizpeh גלעד of Gilead, וממצפה גלעד of Gilead עבר he passed over בני the children עמון׃ of Ammon.}%
\verse{וידר vowed יפתח And Jephthah נדר a vow ליהוה unto the LORD, ויאמר and said, אם If נתון thou shalt without fail deliver תתן thou shalt without fail deliver את בני the children עמון of Ammon בידי׃ into mine hands,}%
\verse{והיה Then it shall be, היוצא cometh forth אשׁר that whatsoever יצא מדלתי of the doors ביתי of my house לקראתי to meet בשׁובי me, when I return בשׁלום in peace מבני from the children עמון of Ammon, והיה shall surely be ליהוה the LORD's, והעליתהו and I will offer it up עולה׃ for a burnt offering.}%
\verse{ויעבר passed over יפתח So Jephthah אל unto בני the children עמון of Ammon להלחם to fight בם ויתנם delivered יהוה against them; and the LORD בידו׃ them into his hands.}%
\verse{ויכם And he smote מערוער ועד even till בואך thou come מנית to Minnith, עשׂרים twenty עיר cities, ועד and unto אבל the plain כרמים of the vineyards, מכה slaughter. גדולה great מאד with a very ויכנעו were subdued בני Thus the children עמון of Ammon מפני before בני the children ישׂראל׃ of Israel.}%
\verse{ויבא came יפתח And Jephthah המצפה to Mizpeh אל unto ביתו his house, והנה and, behold, בתו his daughter יצאת came out לקראתו to meet בתפים him with timbrels ובמחלות and with dances: ורק היא and she יחידה only child; אין her he had neither לו ממנו beside בן son או nor בת׃ daughter.}%
\verse{ויהי And it came to pass, כראותו when he saw אותה ויקרע her, that he rent את בגדיו his clothes, ויאמר and said, אהה Alas, בתי my daughter! הכרע thou hast brought me very low, הכרעתני thou hast brought me very low, ואת and thou היית art בעכרי one of them that trouble ואנכי me: for I פציתי have opened פי my mouth אל unto יהוה the LORD, ולא אוכלשׁוב׃ go back.}%
\verse{ותאמר And she said אליו unto אבי him, My father, פציתה thou hast opened את פיך thy mouth אל unto יהוה the LORD, עשׂה do לי כאשׁר to me according to that which יצא hath proceeded out מפיך of thy mouth; אחרי forasmuch אשׁר forasmuch עשׂה hath taken לך יהוה as the LORD נקמות vengeance מאיביך for thee of thine enemies, מבני of the children עמון׃ of Ammon.}%
\verse{ותאמר And she said אל unto אביה her father, יעשׂה be done לי הדבר thing הזה Let this הרפה for me: let me alone ממני for me: let me alone שׁנים two חדשׁים months, ואלכה that I may go up וירדתי and down על upon ההרים the mountains, ואבכה and bewail על and bewail בתולי my virginity, אנכי I ורעיתי׃}%
\verse{ויאמר And he said, לכי Go. וישׁלח אותהׁני two חדשׁים months: ותלך went היא and she ורעותיה with her companions, ותבך and bewailed על and bewailed בתוליה her virginity על upon ההרים׃ the mountains.}%
\verse{ויהי And it came to pass מקץ at the end שׁנים of two חדשׁים months, ותשׁב that she returned אל unto אביה her father, ויעשׂ who did לה את נדרו with her to his vow אשׁר which נדר he had vowed: והיא and she לא no ידעה knew אישׁ man. ותהי And it was חק a custom בישׂראל׃ in Israel,}%
\verse{מימים days ימימה תלכנה went בנות the daughters ישׂראל of Israel לתנות to lament לבת the daughter יפתח of Jephthah הגלעדי the Gileadite ארבעת four ימים בשׁנה׃ in a year.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 12
\verseWithHeading{Tribal Conflict Between Gilead and Ephraim}{ויצעק gathered themselves together, אישׁ And the men אפרים of Ephraim ויעבר and went צפונה northward, ויאמרו and said ליפתח unto Jephthah, מדוע Wherefore עברת passedst thou over להלחם to fight בבני against the children עמון of Ammon, ולנו לא and didst not קראת call ללכת us to go עמך with ביתך thine house נשׂרף thee? we will burn עליך upon באשׁ׃ thee with fire.}%
\verse{ויאמר said יפתח And Jephthah אליהם unto אישׁ at great strife ריב הייתי were אני them, I ועמי and my people ובני with the children עמון of Ammon; מאד ואזעק and when I called אתכם ולא me not הושׁעתם you, ye delivered אותי מידם׃ out of their hands.}%
\verse{ואראה And when I saw כי that אינך not, מושׁיע ye delivered ואשׂימה I put נפשׁי my life בכפי in my hands, ואעברה and passed over אל against בני the children עמון >> r="#000000">of Ammon, ויתנם delivered יהוה and the LORD בידי them into my hand: ולמה wherefore עליתם then are ye come up אלי unto היום day, הזה me this להלחם׃ to fight}%
\verse{ויקבץ gathered together יפתח Then Jephthah את כל all אנשׁי the men גלעד of Gilead, וילחם and fought את with אפרים Ephraim: ויכו smote אנשׁי and the men גלעד of Gilead את אפרים Ephraim, כי because אמרו they said, פליטי fugitives אפרים of Ephraim אתם Ye גלעד Gileadites בתוך among אפרים the Ephraimites, בתוך among מנשׁה׃ the Manassites.}%
\verse{וילכד took גלעד And the Gileadites את מעברות the passages הירדן of Jordan לאפרים before the Ephraimites: והיה and it was כי that when יאמרו said, פליטי which were escaped אפרים those Ephraimites אעברה Let me go over; ויאמרו said לו אנשׁי גלעד of Gilead האפרתי an Ephraimite? אתה unto him, thou ויאמר If he said, לא׃ Nay;}%
\verse{ויאמרו Then said לו אמר they unto him, Say נא now שׁבלת Shibboleth: ויאמר and he said סבלת Sibboleth: ולא for he could not יכין frame לדבר to pronounce כן right. ויאחזו Then they took אותו וישׁחטוהו him, and slew אל him at מעברות the passages הירדן of Jordan: ויפל and there fell בעת time ההיא at that מאפרים ארבעים forty ושׁנים and two אלף׃ thousand.}%
\verse{וישׁפט judged יפתח And Jephthah את ישׂראל Israel שׁשׁ six שׁנים years. וימת Then died יפתח Jephthah הגלעדי the Gileadite, ויקבר and was buried בערי in the cities גלעד׃ of Gilead.}%
\verseWithHeading{Ibzan, Elon, and Abdon}{וישׁפט judged אחריו And after את ישׂראל Israel. אבצן him Ibzan מבית לחם׃}%
\verse{ויהי And he had לו שׁלשׁים thirty בנים sons, ושׁלשׁים and thirty בנות daughters, שׁלח he sent החוצה abroad, ושׁלשׁים thirty בנות daughters הביא and took in לבניו for his sons. מן from החוץ abroad וישׁפט And he judged את ישׂראל Israel שׁבע seven שׁנים׃ years.}%
\verse{וימת Then died אבצן Ibzan, ויקבר and was buried בבית לחם׃ at Bethlehem.}%
\verse{וישׁפט judged אחריו And after את ישׂראל Israel; אילון him Elon, הזבולני a Zebulonite, וישׁפט and he judged את ישׂראל Israel עשׂר ten שׁנים׃ years.}%
\verse{וימת died, אלון הזבולני the Zebulonite ויקבר and was buried באילון in Aijalon בארץ in the country זבולן׃}%
\verse{וישׁפט judged אחריו And after את ישׂראל Israel. עבדון him Abdon בן the son הלל of Hillel, הפרעתוני׃ a Pirathonite,}%
\verse{ויהי And he had לו ארבעים forty בנים sons ושׁלשׁים and thirty בני nephews, בנים nephews, רכבים that rode על on שׁבעים threescore and ten עירם ass colts: וישׁפט and he judged את ישׂראל Israel שׁמנה eight שׁנים׃ years.}%
\verse{וימת died, עבדון And Abdon בן the son הלל of Hillel הפרעתוני the Pirathonite ויקבר and was buried בפרעתון in Pirathon בארץ in the land אפרים of Ephraim, בהר in the mount העמלקי׃ of the Amalekites.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 13
\verseWithHeading{Samson’s Parents}{ויספו again בני And the children ישׂראל of Israel לעשׂות did הרע evil בעיני in the sight יהוה of the LORD; ויתנם delivered יהוה and the LORD ביד them into the hand פלשׁתים of the Philistines ארבעים forty שׁנה׃ years.}%
\verse{ויהי And there was אישׁ man אחד a certain מצרעה ממשׁפחת of the family הדני of the Danites, ושׁמו whose name מנוח Manoah; ואשׁתו and his wife עקרה barren, ולא not. ילדה׃ and bore}%
\verse{וירא appeared מלאך And the angel יהוה of the LORD אל unto האשׁה the woman, ויאמר and said אליה unto הנה her, Behold נא now, את thou עקרה barren, ולא not: ילדת and bearest והרית but thou shalt conceive, וילדת and bear בן׃ a son.}%
\verse{ועתה Now השׁמרי therefore beware, נא I pray thee, ואל not תשׁתי and drink יין wine ושׁכר nor strong drink, ואל not תאכלי and eat כל any טמא׃ unclean}%
\verse{כי For, הנך lo, הרה thou shalt conceive, וילדת and bear בן a son; ומורה razor לא and no יעלה shall come על on ראשׁו his head: כי for נזיר a Nazarite אלהים unto God יהיה shall be הנער the child מן from הבטן the womb: והוא and he יחל shall begin להושׁיע to deliver את ישׂראל Israel מיד out of the hand פלשׁתים׃ of the Philistines.}%
\verse{ותבא came האשׁה Then the woman ותאמר and told לאישׁה her husband, לאמר saying, אישׁ A man האלהים of God בא came אלי unto ומראהו me, and his countenance כמראה like the countenance מלאך of an angel האלהים of God, נורא terrible: מאד very ולא him not שׁאלתיהו but I asked אי whence מזה הוא he ואת שׁמו he me his name: לא neither הגיד׃ told}%
\verse{ויאמר But he said לי הנך unto me, Behold, הרה thou shalt conceive, וילדת and bear בן a son; ועתה and now אל no תשׁתי drink יין wine ושׁכר nor strong drink, ואל neither תאכלי eat כל any טמאה unclean כי for נזיר a Nazarite אלהים to God יהיה shall be הנער the child מן from הבטן the womb עד to יום the day מותו׃ of his death.}%
\verse{ויעתר entreated מנוח Then Manoah אל entreated יהוה the LORD, ויאמר and said, בי O אדוני my Lord, אישׁ let the man האלהים of God אשׁר which שׁלחת thou didst send יבוא come נא עוד again אלינו unto ויורנו us, and teach מה us what נעשׂה we shall do לנער unto the child היולד׃ that shall be born.}%
\verse{וישׁמע hearkened האלהים And God בקול to the voice מנוח of Manoah; ויבא came מלאך and the angel האלהים of God עוד again אל unto האשׁה the woman והיא as she יושׁבת sat בשׂדה in the field: ומנוח but Manoah אישׁה her husband אין not עמה׃ with}%
\verse{ותמהר made haste, האשׁה her husband, ותרץ and ran, ותגד and showed לאישׁה And the woman ותאמר and said אליו unto הנה him, Behold, נראה hath appeared אלי unto האישׁ the man אשׁר me, that בא came ביום me the day. אלי׃ unto}%
\verse{ויקם arose, וילך and went מנוח And Manoah אחרי after אשׁתו his wife, ויבא and came אל to האישׁ the man, ויאמר and said לו האתה unto him, thou האישׁ the man אשׁר that דברת spakest אל unto האשׁה the woman? ויאמר And he said, אני׃ I}%
\verse{ויאמר said, מנוח And Manoah עתה Now יבא come to pass. דבריך let thy words מה How יהיה shall we order משׁפט shall we order הנער the child, ומעשׂהו׃ and shall we do}%
\verse{ויאמר said מלאך And the angel יהוה אל unto מנוח Manoah, מכל אשׁר that אמרתי I said אל unto האשׁה the woman תשׁמר׃ let her beware.}%
\verse{מכל of any אשׁר that יצא cometh מגפן of the vine, היין wine לא She may not תאכל eat ויין ושׁכר or strong drink, אל neither תשׁת let her drink וכל any טמאה unclean אל nor תאכל eat כל all אשׁר that צויתיה I commanded תשׁמר׃ her let her observe.}%
\verse{ויאמר said מנוח And Manoah אל unto מלאך the angel יהוה of the LORD, נעצרה let us detain נא I pray thee, אותך ונעשׂה thee, until we shall have made ready לפניך for גדי a kid עזים׃ a kid}%
\verse{ויאמר said מלאך And the angel יהוה of the LORD אל unto מנוח Manoah, אם Though תעצרני thou detain לא me, I will not אכל eat בלחמך of thy bread: ואם and if תעשׂה thou wilt offer עלה a burnt offering, ליהוה it unto the LORD. תעלנה thou must offer כי For לא not ידע knew מנוח Manoah כי that מלאך an angel יהוה of the LORD. הוא׃ he}%
\verse{ויאמר said מנוח And Manoah אל unto מלאך the angel יהוה of the LORD, מי What שׁמך thy name, כי that when יבא come to pass דבריך thy sayings וכבדנוך׃ we may do thee honor?}%
\verse{ויאמר said לו מלאך And the angel יהוה of the LORD למה unto him, Why זה thou thus תשׁאל askest לשׁמי after my name, והוא seeing it פלאי׃ secret?}%
\verse{ויקח took מנוח So Manoah את גדי a kid העזים a kid ואת המנחה a meat offering, ויעל and offered על upon הצור a rock ליהוה unto the LORD: ומפלא wondrously; לעשׂות and did ומנוח and Manoah ואשׁתו and his wife ראים׃ looked on.}%
\verse{ויהי For it came to pass, בעלות went up הלהב when the flame מעל from off המזבח the altar, השׁמימה toward heaven ויעל ascended מלאך that the angel יהוה of the LORD בלהב in the flame המזבח of the altar. ומנוח And Manoah ואשׁתו and his wife ראים looked on ויפלו and fell על on פניהם their faces ארצה׃ to the ground.}%
\verse{ולא did no יסף more עוד מלאך But the angel יהוה of the LORD להראה appear אל to מנוח Manoah ואל and to אשׁתו his wife. אז Then ידע knew מנוח Manoah כי that מלאך an angel יהוה of the LORD. הוא׃ he}%
\verse{ויאמר said מנוח And Manoah אל unto אשׁתו his wife, מות נמותי because אלהים God. ראינו׃ we have seen}%
\verse{ותאמר said לו אשׁתו But his wife לו unto him, If חפץ were pleased יהוה the LORD להמיתנו to kill לא us, he would not לקח have received מידנו at our hands, עלה a burnt offering ומנחה and a meat offering ולא neither הראנו would he have showed את כל us all אלה these וכעת would as at this time לא nor השׁמיענו have told כזאת׃ us as these.}%
\verse{ותלד bore האשׁה And the woman בן a son, ותקרא and called את שׁמו his name שׁמשׁון Samson: ויגדל grew, הנער and the child ויברכהו blessed יהוה׃ and the LORD}%
\verse{ותחל began רוח And the Spirit יהוה of the LORD לפעמו to move במחנה him at times in the camp דן of Dan בין between צרעה Zorah ובין אשׁתאל׃ and Eshtaol.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 14
\verseWithHeading{Samson Marries}{וירד went down שׁמשׁון And Samson תמנתה to Timnath, וירא and saw אשׁה a woman בתמנתה in Timnath מבנות of the daughters פלשׁתים׃ of the Philistines.}%
\verse{ויעל And he came up, ויגד and told לאביו his father ולאמו and his mother, ויאמר and said, אשׁה a woman ראיתי I have seen בתמנתה in Timnath מבנות of the daughters פלשׁתים of the Philistines: ועתה now קחו therefore get אותה לי לאשׁה׃ her for me to wife.}%
\verse{ויאמר said לו אביו Then his father ואמו and his mother האין unto him, never בבנות among the daughters אחיך of thy brethren, ובכל or among all עמי my people, אשׁה a woman כי that אתה thou הולך goest לקחת to take אשׁה a wife מפלשׁתים הערלים of the uncircumcised ויאמר said שׁמשׁון And Samson אל unto אביו his father, אותה קח Get לי כי her for me; for היא she ישׁרה pleaseth me well. בעיני׃ pleaseth me well.}%
\verse{ואביו But his father ואמו and his mother לא not ידעו knew כי that מיהוה היא it כי that תאנה an occasion הוא he מבקשׁ sought מפלשׁתים the Philistines ובעת time ההיא for at that פלשׁתים משׁלים had dominion בישׂראל׃ over Israel.}%
\verse{וירד שׁמשׁוןאביו and his father ואמו and his mother, תמנתה to Timnath, ויבאו and came עד to כרמי the vineyards תמנתה of Timnath: והנה and, behold, כפיר a young אריות lion שׁאג roared לקראתו׃ against}%
\verse{ותצלח came mightily עליו upon רוח And the Spirit יהוה of the LORD וישׁסעהו him, and he rent כשׁסע him as he would have rent הגדי a kid, ומאומה and nothing אין and nothing בידו in his hand: ולא not הגיד but he told לאביו his father ולאמו or his mother את אשׁר what עשׂה׃ he had done.}%
\verse{וירד And he went down, וידבר and talked לאשׁה with the woman; ותישׁר בעיניׁמשׁון׃}%
\verse{וישׁב he returned מימים לקחתה to take ויסר her, and he turned aside לראות to see את מפלת the carcass האריה of the lion: והנה and, behold, עדת a swarm דבורים of bees בגוית in the carcass האריה of the lion. ודבשׁ׃ and honey}%
\verse{וירדהו And he took אל thereof in כפיו his hands, וילך and went on הלוך and went on ואכל eating, וילך and came אל to אביו his father ואל אמו and mother, ויתן and he gave להם ויאכלו them, and they did eat: ולא not הגיד but he told להם כי them that מגוית out of the carcass האריה of the lion. רדה he had taken הדבשׁ׃ the honey}%
\verse{וירד went down אביהו So his father אל unto האשׁה the woman: ויעשׂ made שׁם there שׁמשׁון and Samson משׁתה a feast; כי for כן so יעשׂו to do. הבחורים׃ used the young men}%
\verse{ויהי And it came to pass, כראותם when they saw אותו ויקחו him, that they brought שׁלשׁים thirty מרעים companions ויהיו to be אתו׃ with}%
\verse{ויאמר said להם שׁמשׁון And Samson אחודה put forth נא unto them, I will now לכם חידה a riddle אם unto you: if הגד ye can certainly declare תגידו ye can certainly declare אותה לי שׁבעת it me within the seven ימי days המשׁתה of the feast, ומצאתם and find out, ונתתי then I will give לכם שׁלשׁים you thirty סדינים sheets ושׁלשׁים and thirty חלפת change בגדים׃ of garments:}%
\verse{ואם But if לא ye cannot תוכלו ye cannot להגיד declare לי ונתתם give אתם me, then shall ye לי שׁלשׁים me thirty סדינים sheets ושׁלשׁים and thirty חליפות change בגדים of garments. ויאמרו And they said לו חודה unto him, Put forth חידתך thy riddle, ונשׁמענה׃ that we may hear}%
\verse{ויאמר And he said להם מהאכל יצא came forth מאכל meat, ומעז and out of the strong יצא came forth מתוק sweetness. ולא not יכלו And they could להגיד expound החידה the riddle. שׁלשׁת in three ימים׃ days}%
\verse{ויהי And it came to pass ביום day, השׁביעי on the seventh ויאמרו that they said לאשׁת wife, שׁמשׁון unto Samson's פתי Entice את אישׁך thy husband, ויגד that he may declare לנו את החידה unto us the riddle, פן lest נשׂרף we burn אותך ואתית house אביך thee and thy father's באשׁ with fire: הלירשׁנו us to take that we have? קראתם have ye called לנו הלא׃ not}%
\verse{ותבך wept אשׁת wife שׁמשׁון And Samson's עליו before ותאמר him, and said, רק Thou dost but שׂנאתני hate ולא me not: אהבתני me, and lovest החידה a riddle חדת thou hast put forth לבני unto the children עמי of my people, ולי לא and hast not הגדתה told ויאמר me. And he said לה הנה unto her, Behold, לאבי my father ולאמי nor my mother, לא I have not הגדתי told ולך אגיד׃ and shall I tell}%
\verse{ותבך And she wept עליו before שׁבעת him the seven הימים days, אשׁר while היה lasted: להם המשׁתה their feast ויהי and it came to pass ביום day, השׁביעי on the seventh ויגד that he told לה כי her, because הציקתהו she lay sore ותגד upon him: and she told החידה the riddle לבני to the children עמה׃ of her people.}%
\verse{ויאמרו said לו אנשׁי And the men העיר of the city ביום day השׁביעי unto him on the seventh בטרם before יבא went down, החרסה the sun מה What מתוק sweeter מדבשׁ than honey? ומה and what עז stronger מארי than a lion? ויאמר And he said להם לולא unto them, If חרשׁתם ye had not plowed בעגלתי with my heifer, לא ye had not מצאתם found out חידתי׃ my riddle.}%
\verse{ותצלח came עליו upon רוח And the Spirit יהוה of the LORD וירד him, and he went down אשׁקלון to Ashkelon, ויך and slew מהם שׁלשׁים thirty אישׁ men ויקח them, and took את חליצותם their spoil, ויתן and gave החליפות change of garments למגידי unto them which expounded החידה the riddle. ויחר was kindled, אפו And his anger ויעל and he went up בית house. אביהו׃ to his father's}%
\verse{ותהי was אשׁת wife שׁמשׁון But Samson's למרעהו to his companion, אשׁר whom רעה׃ he had used as his friend.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 15
\verseWithHeading{Samson Defeats the Philistines}{ויהי But it came to pass מימים within a while בימי after, in the time קציר harvest, חטים of wheat ויפקד visited שׁמשׁון that Samson את אשׁתו his wife בגדי with a kid; עזים with a kid; ויאמר and he said, אבאה I will go in אל to אשׁתי my wife החדרה into the chamber. ולא would not נתנו suffer אביה But her father לבוא׃ him to go in.}%
\verse{ויאמר said, אביה And her father אמר אמרתיי that שׂנא thou hadst utterly hated שׂנאתה thou hadst utterly hated ואתננה her; therefore I gave למרעך her to thy companion: הלא not אחתה sister הקטנה her younger טובה fairer ממנה than תהי she? take נא her, I pray thee, לך תחתיה׃ instead}%
\verse{ויאמר said להם שׁמשׁון And Samson נקיתי shall I be more blameless הפעם concerning them, Now מפלשׁתים כי though עשׂה do אני I עמם do רעה׃ them a displeasure.}%
\verse{וילך went שׁמשׁון And Samson וילכד and caught שׁלשׁ three מאות hundred שׁועלים foxes, ויקח and took לפדים firebrands, ויפן and turned זנב tail אל to זנב tail, וישׂם and put לפיד firebrand אחד a בין between שׁני two הזנבות tails. בתוך׃ in the midst}%
\verse{ויבער And when he had set אשׁ on fire, בלפידים the brands וישׁלח he let go בקמות into the standing corn פלשׁתים of the Philistines, ויבער and burnt up מגדישׁ both the shocks, ועד and also קמה the standing corn, ועד with כרם the vineyards זית׃ olives.}%
\verse{ויאמרו said, פלשׁתים Then the Philistines מי Who עשׂה hath done זאת this? ויאמרו And they answered, שׁמשׁון Samson, חתן the son-in-law התמני of the Timnite, כי because לקח he had taken את אשׁתו his wife, ויתנה and given למרעהו her to his companion. ויעלו came up, פלשׁתים And the Philistines וישׂרפו and burnt אותה ואתביה her and her father באשׁ׃ with fire.}%
\verse{ויאמר said להם שׁמשׁון And Samson אם unto them, Though תעשׂון ye have done כזאת this, כי yet אם yet נקמתי will I be avenged בכם ואחר of you, and after that אחדל׃ I will cease.}%
\verse{ויך And he smote אותם שׁוק them hip על and ירך thigh מכה slaughter: גדולה with a great וירד and he went down וישׁב and dwelt בסעיף in the top סלע of the rock עיטם׃ Etam.}%
\verse{ויעלו went up, פלשׁתים Then the Philistines ויחנו and pitched ביהודה in Judah, וינטשׁו and spread themselves בלחי׃ in Lehi.}%
\verse{ויאמרו said, אישׁ And the men יהודה of Judah למה Why עליתם are ye come up עלינו against ויאמרו us? And they answered, לאסור To bind את שׁמשׁון Samson עלינו are we come up, לעשׂות to do לו כאשׁר to him as עשׂה׃ he hath done}%
\verse{וירדו went שׁלשׁת Then three אלפים thousand אישׁ men מיהודה אל to סעיף the top סלע of the rock עיטם Etam, ויאמרו and said לשׁמשׁון to Samson, הלא thou not ידעת Knowest כי that משׁלים rulers בנו פלשׁתים the Philistines ומה over us? what זאת this עשׂית thou hast done לנו ויאמר unto us? And he said להם כאשׁר unto them, As עשׂו they did לי כן unto me, so עשׂיתי׃ have I done}%
\verse{ויאמרו And they said לו לאסרך to bind ירדנו unto him, We are come down לתתך thee, that we may deliver ביד thee into the hand פלשׁתים of the Philistines. ויאמר said להם שׁמשׁון And Samson השׁבעו unto them, Swear לי פן unto me, that תפגעון ye will not fall בי אתם׃ upon me yourselves.}%
\verse{ויאמרו And they spoke לו לאמר unto him, saying, לא No; כי but אסר we will bind thee fast, נאסרך we will bind thee fast, ונתנוך and deliver בידם thee into their hand: והמת but surely we will not kill לא נמיתךיאסרהו thee. And they bound בשׁנים him with two עבתים cords, חדשׁים new ויעלוהו and brought him up מן from הסלע׃ the rock.}%
\verse{הוא when he בא came עד unto לחי Lehi, ופלשׁתים the Philistines הריעו shouted לקראתו against ותצלח came mightily עליו upon רוח him: and the Spirit יהוה of the LORD ותהיינה became העבתים him, and the cords אשׁר that על upon זרועותיו his arms כפשׁתים as flax אשׁר that בערו was burnt באשׁ with fire, וימסו loosed אסוריו and his bands מעל from off ידיו׃ his hands.}%
\verse{וימצא And he found לחי jawbone חמור of an ass, טריה a new וישׁלח and put forth ידו his hand, ויקחה and took ויך it, and slew בה אלף a thousand אישׁ׃ men}%
\verse{ויאמר said, שׁמשׁון And Samson בלחי With the jawbone החמור of an ass, חמור חמרתיםלחי with the jaw החמור of an ass הכיתי have I slain אלף a thousand אישׁ׃ men.}%
\verse{ויהי And it came to pass, ככלתו when he had made an end לדבר of speaking, וישׁלך that he cast away הלחי the jawbone מידו out of his hand, ויקרא and called למקום place ההוא that רמת לחי׃ Ramath-lehi.}%
\verse{ויצמא מאדיקרא and called אל on יהוה the LORD, ויאמר and said, אתה Thou נתת hast given ביד into the hand עבדך of thy servant: את התשׁועה deliverance הגדלה great הזאת this ועתה and now אמות shall I die בצמא for thirst, ונפלתי and fall ביד into the hand הערלים׃ of the uncircumcised?}%
\verse{ויבקע cleaved אלהים But God את המכתשׁ a hollow place אשׁר that בלחי in the jaw, ויצאו and there came ממנו therefrom; מים water וישׁת and when he had drunk, ותשׁב came again, רוחו his spirit ויחי and he revived: על wherefore כן wherefore קרא he called שׁמה the name עין הקורא thereof En-hakkore, אשׁר which בלחי in Lehi עד unto היום day. הזה׃ this}%
\verse{וישׁפט And he judged את ישׂראל Israel בימי in the days פלשׁתים of the Philistines עשׂרים twenty שׁנה׃ years.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 16
\verseWithHeading{Samson and Delilah}{וילך Then went שׁמשׁון Samson עזתה to Gaza, וירא and saw שׁם there אשׁה a harlot, זונה a harlot, ויבא and went in אליה׃ unto}%
\verse{לעזתים the Gazites, לאמר saying, בא is come שׁמשׁון Samson הנה hither. ויסבו And they compassed ויארבו in, and laid wait לו כל for him all הלילה night בשׁער in the gate העיר of the city, ויתחרשׁו and were quiet כל all הלילה the night, לאמר saying, עד when אור it is day, הבקר In the morning, והרגנהו׃ we shall kill}%
\verse{וישׁכב lay שׁמשׁון And Samson עד till חצי midnight, הלילה midnight, ויקם and arose בחצי at midnight, הלילה at midnight, ויאחז and took בדלתות the doors שׁער of the gate העיר of the city, ובשׁתי and the two המזוזות posts, ויסעם and went away עם with them, bar and all, הבריח with them, bar and all, וישׂם and put על upon כתפיו his shoulders, ויעלם and carried them up אל to ראשׁ the top ההר of a hill אשׁר that על before פני before חברון׃ Hebron.}%
\verse{ויהי And it came to pass אחרי afterward, כן afterward, ויאהב that he loved אשׁה a woman בנחל in the valley שׂרק of Sorek, ושׁמה whose name דלילה׃ Delilah.}%
\verse{ויעלו came up אליה unto סרני And the lords פלשׁתים of the Philistines ויאמרו her, and said לה פתי unto her, Entice אותו וראי him, and see במה wherein כחו strength גדול his great ובמה and by what נוכל we may prevail לו ואסרנהו against him, that we may bind לענתו him to afflict ואנחנו him: and we נתן will give לך אישׁ thee every one אלף of us eleven hundred ומאה of us eleven hundred כסף׃ of silver.}%
\verse{ותאמר said דלילה And Delilah אל to שׁמשׁון Samson, הגידה Tell נא me, I pray thee, לי במה wherein כחך strength גדול thy great ובמה and wherewith תאסר thou mightest be bound לענותך׃ to afflict}%
\verse{ויאמר said אליה unto שׁמשׁון And Samson אם her, If יאסרני they bind בשׁבעה me with seven יתרים withes לחים green אשׁר that לא were never חרבו dried, וחליתי then shall I be weak, והייתי and be כאחד as another האדם׃ man.}%
\verse{ויעלו brought up לה סרני Then the lords פלשׁתים of the Philistines שׁבעה to her seven יתרים withes לחים green אשׁר which לא had not חרבו been dried, ותאסרהו׃ and she bound}%
\verse{והארב Now men lying in wait, ישׁב abiding לה בחדר with her in the chamber. ותאמר And she said אליו unto פלשׁתים him, The Philistines עליך upon שׁמשׁון thee, Samson. וינתק And he broke את היתרים the withes, כאשׁר as ינתק is broken פתיל a thread הנערת of tow בהריחו when it toucheth אשׁ the fire. ולא was not נודע known. כחו׃ So his strength}%
\verse{ותאמר said דלילה And Delilah אל unto שׁמשׁון Samson, הנה Behold, התלת thou hast mocked בי ותדבר me, and told אלי me, and told כזבים עתה now הגידה tell נא me, I pray thee, לי במה wherewith תאסר׃ thou mightest be bound.}%
\verse{ויאמר And he said אליה unto אם her, If אסור they bind me fast יאסרוני they bind me fast בעבתים ropes חדשׁים with new אשׁר that לא never נעשׂה were occupied, בהם מלאכה were occupied, וחליתי then shall I be weak, והייתי and be כאחד as another האדם׃ man.}%
\verse{ותקח therefore took דלילה Delilah עבתים ropes, חדשׁים new ותאסרהו and bound בהם ותאמר him therewith, and said אליו unto פלשׁתים him, The Philistines עליך upon שׁמשׁון thee, Samson. והארב And liers in wait ישׁב abiding בחדר in the chamber. וינתקם And he broke מעל them from off זרעתיו his arms כחוט׃ like a thread.}%
\verse{ותאמר said דלילה And Delilah אל unto שׁמשׁון Samson, עד הנהתלת thou hast mocked בי ותדבר me, and told אלי me, and told כזבים me lies: הגידה tell לי במה me wherewith תאסר thou mightest be bound. ויאמר And he said אליה unto אם her, If תארגי thou weavest את שׁבע the seven מחלפות locks ראשׁי of my head עם with המסכת׃ the web.}%
\verse{ותתקע And she fastened ביתד with the pin, ותאמר and said אליו unto פלשׁתים him, The Philistines עליך upon שׁמשׁון thee, Samson. וייקץ And he awaked משׁנתו out of his sleep, ויסע and went away את היתד with the pin הארג of the beam, ואת המסכת׃ and with the web.}%
\verse{ותאמר And she said אליו unto איך him, How תאמר canst thou say, אהבתיך I love ולבך thee, when thine heart אין not אתי with זה שׁלשׁ me these three פעמים times, התלת me? thou hast mocked בי ולא and hast not הגדת told לי במה me wherein כחך strength גדול׃ thy great}%
\verse{ויהי And it came to pass, כי when הציקה she pressed לו בדבריה with her words, כל him daily הימים him daily ותאלצהו and urged ותקצר was vexed נפשׁו him, that his soul למות׃ unto death;}%
\verse{ויגד That he told לה את כל her all לבו his heart, ויאמר and said לה מורה a razor לא unto her, There hath not עלה come על upon ראשׁי mine head; כי for נזיר a Nazarite אלהים unto God אני I מבטן womb: אמי from my mother's אם if גלחתי I be shaven, וסר will go ממני womb: כחי then my strength וחליתי me, and I shall become weak, והייתי and be ככל like any האדם׃ man.}%
\verse{ותרא saw דלילה And when Delilah כי that הגיד he had told לה את כל her all לבו his heart, ותשׁלח she sent ותקרא and called לסרני for the lords פלשׁתים of the Philistines, לאמר saying, עלו Come up הפעם this once, כי for הגיד he hath showed לה את כל me all לבו his heart. ועלו came up אליה unto סרני Then the lords פלשׁתים of the Philistines ויעלו her, and brought הכסף money בידם׃ in their hand.}%
\verse{ותישׁנהו And she made him sleep על upon ברכיה her knees; ותקרא and she called לאישׁ for a man, ותגלח and she caused him to shave off את שׁבע the seven מחלפות locks ראשׁו of his head; ותחל and she began לענותו to afflict ויסר went כחו him, and his strength מעליו׃ from}%
\verse{ותאמר And she said, פלשׁתים The Philistines עליך upon שׁמשׁון thee, Samson. ויקץ And he awoke משׁנתו out of his sleep, ויאמר and said, אצא I will go out כפעם as at other times before, בפעם as at other times before, ואנער and shake myself. והוא And he לא not ידע knew כי that יהוה the LORD סר was departed מעליו׃ from}%
\verse{ויאחזוהו took פלשׁתים But the Philistines וינקרו him, and put out את עיניו his eyes, ויורידו אותוזתה to Gaza, ויאסרוהו and bound בנחשׁתים him with fetters of brass; ויהי and he did טוחן grind בבית house. האסירים׃ in the prison}%
\verse{ויחל began שׂער Howbeit the hair ראשׁו of his head לצמח to grow again כאשׁר after גלח׃ he was shaven.}%
\verse{וסרני Then the lords פלשׁתים of the Philistines נאספו gathered them together לזבח for to offer זבח sacrifice גדול a great לדגון unto Dagon אלהיהם their god, ולשׂמחה and to rejoice: ויאמרו for they said, נתן hath delivered אלהינו Our god בידנו into our hand. את שׁמשׁון Samson אויבינו׃ our enemy}%
\verse{ויראו saw אתו העם And when the people ויהללו him, they praised את אלהיהם their god: כי for אמרו they said, נתן hath delivered אלהינו Our god בידנו into our hands את אויבנו our enemy, ואת מחריב and the destroyer ארצנו of our country, ואשׁר which הרבה אתללינו׃}%
\verse{ויהי And it came to pass, כי when טוב were merry, לבם their hearts ויאמרו that they said, קראו Call לשׁמשׁון for Samson, וישׂחק that he may make us sport. לנו ויקראו And they called לשׁמשׁון for Samson מבית האסיריםיצחק and he made them sport: לפניהם and he made them sport: ויעמידו and they set אותו בין him between העמודים׃ the pillars.}%
\verse{ויאמר said שׁמשׁון And Samson אל unto הנער the lad המחזיק that held בידו him by the hand, הניחה אותיהימשׁני אתעמדים the pillars אשׁר whereupon הבית the house נכון standeth, עליהם whereupon ואשׁען that I may lean עליהם׃ upon}%
\verse{והבית Now the house מלא was full האנשׁים והנשׁים and women; ושׁמה there; כל and all סרני the lords פלשׁתים of the Philistines ועל and upon הגג the roof כשׁלשׁת about three אלפים thousand אישׁ of men ואשׁה and women, הראים that beheld בשׂחוק made sport. שׁמשׁון׃ while Samson}%
\verseWithHeading{Samson’s Revenge}{ויקרא called שׁמשׁון And Samson אל unto יהוה GOD, ויאמר and said, אדני O Lord יהוה the LORD, זכרני remember נא me, I pray thee, וחזקני and strengthen נא me, I pray thee, אך only הפעם once, הזה this האלהים O God, ואנקמה נקםחת משׁתי for my two עיני eyes. מפלשׁתים׃}%
\verse{וילפת took hold שׁמשׁון And Samson את שׁני of the two עמודי pillars התוך middle אשׁר which הבית the house נכון stood, עליהם upon ויסמך which it was borne up, עליהם and on אחד of the one בימינו with his right hand, ואחד and of the other בשׂמאלו׃ with his left.}%
\verse{ויאמר said, שׁמשׁון And Samson תמות die נפשׁי Let me עם with פלשׁתים the Philistines. ויט And he bowed בכח himself with might; ויפל fell הבית and the house על upon הסרנים the lords, ועל and upon כל all העם the people אשׁר that בו ויהיו were המתים therein. So the dead אשׁר which המית he slew במותו at his death רבים more מאשׁר than which המית he slew בחייו׃ in his life.}%
\verse{וירדו came down, אחיו Then his brethren וכל and all בית the house אביהו of his father וישׂאו and took אתו ויעלו him, and brought up, ויקברו and buried אותו בין him between צרעה Zorah ובין אשׁתאל and Eshtaol בקבר in the burial place מנוח of Manoah אביו his father. והוא And he שׁפט judged את ישׂראל Israel עשׂרים twenty שׁנה׃ years.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 17
\verseWithHeading{Micah’s Idolatry}{ויהי And there was אישׁ a man מהר of mount אפרים Ephraim, ושׁמו whose name מיכיהו׃}%
\verse{ויאמר And he said לאמו unto his mother, אלף ומאהכסף of silver אשׁר that לקח were taken לך ואתי from thee, about which thou אלית cursedst, וגם of also אמרת and spakest באזני in mine ears, הנה behold, הכסף the silver אתי with אני me; I לקחתיו took ותאמר said, אמו it. And his mother ברוך Blessed בני my son. ליהוה׃ of the LORD,}%
\verse{וישׁב And when he had restored את אלף the eleven hundred ומאה the eleven hundred הכסף of silver לאמו to his mother, ותאמר said, אמו his mother הקדשׁ הקדשׁתית הכסף the silver ליהוה unto the LORD מידי from my hand לבני for my son, לעשׂות to make פסל a graven image ומסכה and a molten image: ועתה now אשׁיבנו׃ therefore I will restore}%
\verse{וישׁב Yet he restored את הכסף the money לאמו unto his mother; ותקח took אמו and his mother מאתים two hundred כסף of silver, ותתנהו and gave לצורף them to the founder, ויעשׂהו who made פסל thereof a graven image ומסכה and a molten image: ויהי and they were בבית in the house מיכיהו׃}%
\verse{והאישׁ And the man מיכה Micah לו בית had a house אלהים of gods, ויעשׂ and made אפוד an ephod, ותרפים and teraphim, וימלא אתד אחד one מבניו of his sons, ויהי who became לו לכהן׃ his priest.}%
\verse{בימים days ההם In those אין no מלך king בישׂראל in Israel, אישׁ every man הישׁר right בעיניו in his own eyes. יעשׂה׃ did}%
\verse{ויהי And there was נער a young man מבית לחם יהודה of Judah, ממשׁפחת of the family יהודה והוא who לוי a Levite, והוא and he גר sojourned שׁם׃ there.}%
\verse{וילך departed האישׁ And the man מהעיר out of the city מבית לחם יהודהגור to sojourn באשׁר where ימצא he could find ויבא and he came הר to mount אפרים Ephraim עד to בית the house מיכה of Micah, לעשׂות as he journeyed. דרכו׃ as he journeyed.}%
\verse{ויאמר said לו מיכה And Micah מאין תבוא comest ויאמר thou? And he said אליו unto לוי a Levite אנכי him, I מבית לחם יהודהאנכי and I הלך go לגור to sojourn באשׁר where אמצא׃ I may find}%
\verse{ויאמר said לו מיכה And Micah שׁבה unto him, Dwell עמדי with והיה me, and be לי לאב unto me a father ולכהן and a priest, ואנכי and I אתן will give לך עשׂרת thee ten כסף of silver לימים by the year, וערך and a suit בגדים of apparel, ומחיתך and thy victuals. וילך went in. הלוי׃ So the Levite}%
\verse{ויואל was content הלוי And the Levite לשׁבת to dwell את with האישׁ the man; ויהי was הנער and the young man לו כאחד unto him as one מבניו׃ of his sons.}%
\verse{וימלא מיכה And Micah את ידלוי the Levite; ויהי became לו הנער and the young man לכהן his priest, ויהי and was בבית in the house מיכה׃ of Micah.}%
\verse{ויאמר Then said מיכה Micah, עתה Now ידעתי know כי I that ייטיב will do me good, יהוה the LORD לי כי seeing היה I have לי הלוי a Levite לכהן׃ to priest.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 18
\verseWithHeading{The Tribe of Dan Seeks Territory}{בימים days ההם In those אין no מלך king בישׂראל in Israel: ובימים days ההם and in those שׁבט the tribe הדני of the Danites מבקשׁ sought לו נחלה them an inheritance לשׁבת to dwell כי in; for לא had not נפלה fallen לו עד unto היום day ההוא that בתוך unto them among שׁבטי the tribes ישׂראל of Israel. בנחלה׃ inheritance}%
\verse{וישׁלחו sent בני And the children דן of Dan ממשׁפחתם of their family חמשׁה five אנשׁים men מקצותם from their coasts, אנשׁים men בני חיל of valor, מצרעה ומאשׁתאלרגל to spy out את הארץ the land, ולחקרה and to search ויאמרו it; and they said אלהם unto לכו them, Go, חקרו search את הארץ the land: ויבאו who when they came הר to mount אפרים Ephraim, עד to בית the house מיכה of Micah, וילינו they lodged שׁם׃ there.}%
\verse{המה When they עם by בית the house מיכה of Micah, והמה they הכירו knew את קול the voice הנער of the young man הלוי the Levite: ויסורו and they turned in שׁם thither, ויאמרו and said לו מי unto him, Who הביאך brought הלם thee hither? ומה and what אתה thou עשׂה makest בזה in this ומה and what לך פה׃ hast thou here?}%
\verse{ויאמר And he said אלהם unto כזה them, Thus וכזה עשׂה dealeth לי מיכה Micah וישׂכרני with me, and hath hired ואהי me, and I am לו לכהן׃ his priest.}%
\verse{ויאמרו And they said לו שׁאל unto him, Ask counsel, נא we pray thee, באלהים of God, ונדעה that we may know התצליח shall be prosperous. דרכנו whether our way אשׁר which אנחנו we הלכים go עליה׃}%
\verse{ויאמר said להם הכהן And the priest לכו unto them, Go לשׁלום in peace: נכח before יהוה the LORD דרככם your way אשׁר wherein תלכו׃ ye go.}%
\verse{וילכו departed, חמשׁת Then the five האנשׁים men ויבאו and came לישׁה to Laish, ויראו and saw את העם the people אשׁר that בקרבה therein, יושׁבת how they dwelt לבטח careless, כמשׁפט after the manner צדנים of the Zidonians, שׁקט quiet ובטח and secure; ואין and no מכלים that might put to shame דבר in thing; בארץ in the land, יורשׁ magistrate עצר magistrate ורחקים far המה and they מצדנים ודבר business אין and had no להם עם with אדם׃ man.}%
\verse{ויבאו And they came אל unto אחיהם their brethren צרעה to Zorah ואשׁתאל and Eshtaol: ויאמרו said להם אחיהם and their brethren מה unto them, What אתם׃ ye?}%
\verse{ויאמרו And they said, קומה Arise, ונעלה that we may go up עליהם against כי them: for ראינו we have seen את הארץ the land, והנה and, behold, טובה good: מאד it very ואתם and ye מחשׁים still? אל be not slothful תעצלו be not slothful ללכת to go, לבא to enter לרשׁת to possess את הארץ׃ the land.}%
\verse{כבאכם When ye go, תבאו ye shall come אל unto עם a people בטח secure, והארץ land: רחבת and to a large ידים it into your hands; כי for נתנה hath given אלהים God בידכם מקום a place אשׁר where אין no שׁם מחסור want כל of any דבר thing אשׁר that בארץ׃ in the earth.}%
\verse{ויסעו And there went משׁם from thence ממשׁפחת of the family הדני of the Danites, מצרעה ומאשׁתאלׁשׁ six מאות hundred אישׁ men חגור appointed כלי with weapons מלחמה׃ of war.}%
\verse{ויעלו And they went up, ויחנו and pitched בקרית יערים in Kirjath-jearim, ביהודה in Judah: על wherefore כן wherefore קראו they called למקום place ההוא that מחנה דן Mahaneh-dan עד unto היום day: הזה this הנה behold, אחרי behind קרית יערים׃ Kirjath-jearim.}%
\verse{ויעברו And they passed משׁם thence הר unto mount אפרים Ephraim, ויבאו and came עד unto בית the house מיכה׃ of Micah.}%
\verse{ויענו Then answered חמשׁת the five האנשׁים ההלכים that went לרגל to spy out את הארץ the country לישׁ of Laish, ויאמרו and said אל unto אחיהם their brethren, הידעתם Do ye know כי that ישׁ there is בבתים houses האלה in these אפוד an ephod, ותרפים and teraphim, ופסל and a graven image, ומסכה and a molten image? ועתה now דעו therefore consider מה what תעשׂו׃ ye have to do.}%
\verse{ויסורו And they turned שׁמה thitherward, ויבאו and came אל to בית the house הנער of the young man הלוי the Levite, בית unto the house מיכה of Micah, וישׁאלו and saluted לו לשׁלום׃ and saluted}%
\verse{ושׁשׁ And the six מאות hundred אישׁ men חגורים appointed כלי with their weapons מלחמתם of war, נצבים stood פתח by the entering השׁער of the gate. אשׁר which מבני of the children דן׃ of Dan,}%
\verse{ויעלו that went חמשׁת And the five האנשׁים men ההלכים לרגל to spy out את הארץ the land באו came in שׁמה thither, לקחו took את הפסלאת האפוד and the ephod, ואת התרפים and the teraphim, ואת המסכה and the molten image: והכהן and the priest נצב stood פתח in the entering השׁער of the gate ושׁשׁ with the six מאות hundred האישׁ men החגור appointed כלי with weapons המלחמה׃ of war.}%
\verse{ואלה And these באו went בית house, מיכה into Micah's ויקחו and fetched את פסל the carved image, האפוד the ephod, ואת התרפים and the teraphim, ואת המסכה and the molten image. ויאמר Then said אליהם unto הכהן the priest מה them, What אתם ye? עשׂים׃ do}%
\verse{ויאמרו And they said לו החרשׁ unto him, Hold thy peace, שׂים lay ידך thine hand על upon פיך thy mouth, ולך and go עמנו with והיה us, and be לנו לאב to us a father ולכהן and a priest: הטוב better היותך for thee to be כהן a priest לבית unto the house אישׁ man, אחד of one או or היותך that thou be כהן a priest לשׁבט unto a tribe ולמשׁפחה and a family בישׂראל׃ in Israel?}%
\verse{וייטב was glad, לב heart הכהן And the priest's ויקח and he took את האפוד the ephod, ואת התרפים and the teraphim, ואת הפסל and the graven image, ויבא and went בקרב in the midst העם׃ of the people.}%
\verse{ויפנו So they turned וילכו and departed, וישׂימו and put את הטף the little ones ואת המקנה and the cattle ואת הכבודה and the carriage לפניהם׃ before}%
\verse{המה when they הרחיקו were a good way מבית from the house מיכה of Micah, והאנשׁים אשׁר that בבתים in the houses אשׁר עם near בית house מיכה to Micah's נזעקו were gathered together, וידביקו and overtook את בני the children דן׃ of Dan.}%
\verse{ויקראו And they cried אל unto בני the children דן of Dan. ויסבו And they turned פניהם their faces, ויאמרו and said למיכה unto Micah, מה What לך כי aileth thee, that נזעקת׃ thou comest with such a company?}%
\verse{ויאמר And he said, את אלהי my gods אשׁר which עשׂיתי I made, לקחתם Ye have taken away ואת הכהן and the priest, ותלכו and ye are gone away: ומה and what לי עוד have I more? ומה and what זה this תאמרו ye say אלי unto מה׃ me, What}%
\verse{ויאמרו said אליו unto בני And the children דן of Dan אל him, Let not תשׁמע be heard קולך thy voice עמנו among פן us, lest יפגעו run בכם אנשׁים fellows מרי angry נפשׁ angry ואספתה upon thee, and thou lose נפשׁך thy life, ונפשׁ with the lives ביתך׃ of thy household.}%
\verse{וילכו went בני And the children דן of Dan לדרכם their way: וירא saw מיכה and when Micah כי that חזקים too strong המה they ממנו for ויפן him, he turned וישׁב and went back אל unto ביתו׃ his house.}%
\verse{והמה And they לקחו took את אשׁר which עשׂה had made, מיכה Micah ואת הכהן and the priest אשׁר which היה he had, לו ויבאו and came על unto לישׁ Laish, על unto עם a people שׁקט at quiet ובטח and secure: ויכו and they smote אותם לפי them with the edge חרב of the sword, ואת העיר the city שׂרפו and burnt באשׁ׃ with fire.}%
\verse{ואין And no מציל deliverer, כי because רחוקה far היא it מצידון ודבר business אין and they had no להם עם with אדם man; והיא and it בעמק was in the valley אשׁר that לבית רחוב by Beth-rehob. ויבנו And they built את העיר a city, וישׁבו׃ and dwelt}%
\verse{ויקראו And they called שׁם the name העיר of the city דן Dan, בשׁם after the name דן of Dan אביהם their father, אשׁר who יולד was born לישׂראל unto Israel: ואולם howbeit לישׁ Laish שׁם the name העיר of the city לראשׁנה׃ at the first.}%
\verse{ויקימו set up להם בני And the children דן of Dan את הפסל the graven image: ויהונתן בן the son גרשׁם בן the son מנשׁה of Manasseh, הוא he ובניו and his sons היו were כהנים priests לשׁבט to the tribe הדני of Dan עד until יום the day גלות of the captivity הארץ׃ of the land.}%
\verse{וישׂימו And they set them up להם את פסל graven image, מיכה Micah's אשׁר which עשׂה he made, כל all ימי the time היות was בית that the house האלהים of God בשׁלה׃ in Shiloh.}%
\end{biblechapter}%
\begin{biblechapter}% Judges 19
\verseWithHeading{The Concubine and the Levite}{ויהי And it came to pass בימים days, ההם in those ומלך king אין when no בישׂראל in Israel, ויהי that there was אישׁ a certain לוי Levite גר sojourning בירכתי on the side הר of mount אפרים Ephraim, ויקח who took לו אשׁה to him a concubine פילגשׁ to him a concubine מבית לחם יהודה׃}%
\verse{ותזנה played the whore עליו against פילגשׁו And his concubine ותלך him, and went away מאתו אל him unto בית house אביה her father's אל to בית לחם יהודהתהי and was שׁם there ימים whole ארבעה four חדשׁים׃ months.}%
\verse{ויקם arose, אישׁה And her husband וילך and went אחריה after לדבר her, to speak על friendly לבה friendly להשׁיבו unto her, to bring her again, ונערו having his servant עמו with וצמד him, and a couple חמרים of asses: ותביאהו and she brought בית house: אביה him into her father's ויראהו saw אבי and when the father הנערה of the damsel וישׂמח him, he rejoiced לקראתו׃ to meet}%
\verse{ויחזק retained בו חתנו And his father-in-law, אבי father, הנערה the damsel's וישׁב him; and he abode אתו with שׁלשׁת him three ימים days: ויאכלו so they did eat וישׁתו and drink, וילינו and lodged שׁם׃ there.}%
\verse{ויהי And it came to pass ביום day, הרביעי on the fourth וישׁכימו when they arose early בבקר in the morning, ויקם that he rose up ללכת to depart: ויאמר said אבי father הנערה and the damsel's אל unto חתנו his son-in-law, סעד Comfort לבך thine heart פת with a morsel לחם of bread, ואחר and afterward תלכו׃ go your way.}%
\verse{וישׁבו And they sat down, ויאכלו and did eat שׁניהם both יחדו of them together: וישׁתו and drink ויאמר had said אבי father הנערה for the damsel's אל unto האישׁ the man, הואל Be content, נא I pray thee, ולין and tarry all night, ויטב be merry. לבך׃ and let thine heart}%
\verse{ויקם rose up האישׁ And when the man ללכת to depart, ויפצר urged בו חתנו his father-in-law וישׁב again. וילן him: therefore he lodged שׁם׃ there}%
\verse{וישׁכם And he arose early בבקר in the morning ביום day החמישׁי on the fifth ללכת to depart: ויאמר said, אבי father הנערה and the damsel's סעד Comfort נא I pray thee. לבבך thine heart, והתמהמהו And they tarried עד until נטות afternoon, היום afternoon, ויאכלו and they did eat שׁניהם׃ both}%
\verse{ויקם rose up האישׁ And when the man ללכת to depart, הוא he, ופילגשׁו and his concubine, ונערו and his servant, ויאמר said לו חתנו his father-in-law, אבי father, הנערה the damsel's הנה unto him, Behold, נא now רפה draweth היום the day לערב toward evening, לינו tarry all night: נא I pray you הנה behold, חנות groweth to an end, היום the day לין lodge פה here, וייטב may be merry; לבבך that thine heart והשׁכמתם get you early מחר and tomorrow לדרככם on your way, והלכת that thou mayest go לאהלך׃ home.}%
\verse{ולא not אבה would האישׁ But the man ללון tarry that night, ויקם but he rose up וילך and departed, ויבא and came עד over against נכח over against יבוס Jebus, היא which ירושׁלם Jerusalem; ועמו and with צמד him two חמורים asses חבושׁים saddled, ופילגשׁו his concubine עמו׃ also with}%
\verse{הם when they עם by יבוס Jebus, והיום the day רד spent; מאד was far ויאמר said הנער and the servant אל unto אדניו his master, לכה Come, נא I pray thee, ונסורה and let us turn in אל into עיר city היבוסי of the Jebusites, הזאת this ונלין׃ and lodge}%
\verse{ויאמר said אליו unto אדניו And his master לא him, We will not נסור turn aside אל hither into עיר the city נכרי of a stranger, אשׁר that לא not מבני of the children ישׂראל of Israel; הנה ועברנו we will pass over עד to גבעה׃ Gibeah.}%
\verse{ויאמר And he said לנערו unto his servant, לך Come, ונקרבה and let us draw near באחד to one המקמות of these places ולנו to lodge all night, בגבעה in Gibeah, או or ברמה׃ in Ramah.}%
\verse{ויעברו And they passed on וילכו and went their way; ותבא went down להם השׁמשׁ and the sun אצל upon them by הגבעה Gibeah, אשׁר which לבנימן׃ to Benjamin.}%
\verse{ויסרו And they turned aside שׁם thither, לבוא to go in ללון to lodge בגבעה in Gibeah: ויבא and when he went in, וישׁב he sat him down ברחוב in a street העיר of the city: ואין for no אישׁ man מאסף that took אותם הביתה them into his house ללון׃ to lodging.}%
\verse{והנה And, behold, אישׁ man זקן an old בא there came מן from מעשׂהו his work מן out of השׂדה the field בערב at even, והאישׁ but the men מהר also of mount אפרים Ephraim; והוא and he גר sojourned בגבעה in Gibeah: ואנשׁי המקום of the place בני ימיני׃ Benjamites.}%
\verse{וישׂא And when he had lifted up עיניו his eyes, וירא he saw את האישׁ man הארח a wayfaring ברחב in the street העיר of the city: ויאמר said, האישׁ man הזקן and the old אנה תלך goest ומאין thou? and whence תבוא׃ comest}%
\verse{ויאמר And he said אליו unto עברים passing אנחנו him, We מבית לחם יהודהד toward ירכתי the side הר of mount אפרים Ephraim; משׁם from thence אנכי I: ואלך and I went עד to בית לחם יהודהאת בית to the house יהוה of the LORD; אני but I הלך going ואין and there no אישׁ man מאסף that receiveth אותי הביתה׃ me to house.}%
\verse{וגם both תבן straw גם and מספוא provender ישׁ Yet there is לחמורינו for our asses; וגם also לחם bread ויין and wine ישׁ and there is לי ולאמתך for me, and for thy handmaid, ולנער and for the young man עם with עבדיך thy servants: אין no מחסור want כל of any דבר׃ thing.}%
\verse{ויאמר said, האישׁ man הזקן And the old שׁלום Peace לך רק with thee; howsoever כל all מחסורך thy wants עלי upon רק me; only ברחוב in the street. אל not תלן׃ lodge}%
\verse{ויביאהו So he brought לביתו him into his house, ויבול and gave provender לחמורים unto the asses: וירחצו and they washed רגליהם their feet, ויאכלו and did eat וישׁתו׃ and drink.}%
\verse{המה as they מיטיבים אתבם והנה behold, אנשׁי the men העיר of the city, אנשׁי certain בני sons בליעל of Belial, נסבו אתבית of the house, מתדפקים beat על at הדלת the door, ויאמרו and spoke אל to האישׁ the man בעל the master הבית thine house, הזקן the old man, לאמר saying, הוצא Bring forth את האישׁשׁר that בא came אל into ביתך ונדענו׃ that we may know}%
\verse{ויצא went out אליהם unto האישׁ And the man, בעל the master הבית of the house, ויאמר them, and said אלהם unto אל them, Nay, אחי my brethren, אל do not wickedly; תרעו do not wickedly; נא I pray you, אחרי seeing that אשׁר seeing that בא is come האישׁ man הזה this אל into ביתי mine house, אל not תעשׂו do את הנבלה folly. הזאת׃ this}%
\verse{הנה Behold, בתי my daughter הבתולה a maiden, ופילגשׁהו and his concubine; אוציאה them I will bring out נא now, אותם וענו and humble אותם ועשׂו ye them, and do להם הטוב good בעיניכם with them what seemeth ולאישׁ man הזה unto you: but unto this לא not תעשׂו do דבר a thing. הנבלה vile הזאת׃ so}%
\verse{ולא not אבו would האנשׁים לשׁמע hearken לו ויחזק took האישׁ But the men בפילגשׁו his concubine, ויצא and brought her forth אליהם unto החוץ and brought her forth וידעו them; and they knew אותה ויתעללו her, and abused בה כל her all הלילה the night עד until הבקר the morning: וישׁלחוה they let her go. בעלות began to spring, השׁחר׃ and when the day}%
\verse{ותבא Then came האשׁה the woman לפנות in the dawning הבקר of the day, ותפל and fell down פתח at the door בית house האישׁ of the man's אשׁר where אדוניה her lord שׁם where עד till האור׃ it was light.}%
\verse{ויקם rose up אדניה And her lord בבקר in the morning, ויפתח and opened דלתות the doors הבית of the house, ויצא and went out ללכת to go לדרכו his way: והנה and, behold, האשׁה the woman פילגשׁו his concubine נפלת was fallen down פתח the door הבית of the house, וידיה and her hands על upon הסף׃ the threshold.}%
\verse{ויאמר And he said אליה unto קומי her, Up, ונלכה and let us be going. ואין But none ענה answered. ויקחה Then the man took על her upon החמור an ass, ויקם rose up, האישׁ and the man וילך and got למקמו׃ him unto his place.}%
\verse{ויבא And when he was come אל into ביתו his house, ויקח he took את המאכלת a knife, ויחזק and laid hold בפילגשׁו on his concubine, וינתחה and divided לעצמיה her, with her bones, לשׁנים into twelve עשׂר into twelve נתחים pieces, וישׁלחה and sent בכל her into all גבול the coasts ישׂראל׃ of Israel.}%
\verse{והיה And it was כל so, that all הראה that saw ואמר it said, לא There was no נהיתה deed done ולא nor נראתה seen כזאת such למיום from the day עלות came up בני that the children ישׂראל of Israel מארץ out of the land מצרים of Egypt עד unto היום day: הזה this שׂימו consider לכם עליה of עצו it, take advice, ודברו׃ and speak}%
\end{biblechapter}%
\begin{biblechapter}% Judges 20
\verseWithHeading{The Punishment of Benjamin}{ויצאו went out, כל Then all בני the children ישׂראל of Israel ותקהל was gathered together העדה and the congregation כאישׁ man, אחד as one למדן ועד even to באר שׁבע Beer-sheba, וארץ with the land הגלעד of Gilead, אל unto יהוה the LORD המצפה׃ in Mizpeh.}%
\verse{ויתיצבו presented themselves פנות And the chief כל of all העם the people, כל of all שׁבטי the tribes ישׂראל of Israel, בקהל in the assembly עם of the people האלהים of God, ארבע four מאות hundred אלף thousand אישׁ footmen רגלי footmen שׁלף that drew חרב׃ sword.}%
\verse{וישׁמעו heard בני (Now the children בנימן of Benjamin כי that עלו were gone up בני the children ישׂראל of Israel המצפה to Mizpeh.) ויאמרו Then said בני the children ישׂראל of Israel, דברו Tell איכה how נהיתה was הרעה wickedness? הזאת׃ this}%
\verse{ויען answered האישׁ the husband הלוי אישׁאשׁה of the woman הנרצחה that was slain, ויאמר and said, הגבעתה into Gibeah אשׁר that לבנימן to Benjamin, באתי I came אני I ופילגשׁי and my concubine, ללון׃ to lodge.}%
\verse{ויקמו rose עלי against בעלי And the men הגבעה of Gibeah ויסבו עלי upon את הביתילה me by night, אותי דמו thought להרג to have slain ואת פילגשׁי me: and my concubine ענו have they forced, ותמת׃ that she is dead.}%
\verse{ואחז And I took בפילגשׁי my concubine, ואנתחה and cut her in pieces, ואשׁלחה and sent בכל her throughout all שׂדה the country נחלת of the inheritance ישׂראל of Israel: כי for עשׂו they have committed זמה lewdness ונבלה and folly בישׂראל׃ in Israel.}%
\verse{הנה Behold, כלכם ye all בני children ישׂראל of Israel; הבו give לכם דבר your advice ועצה and counsel. הלם׃ here}%
\verse{ויקם arose כל And all העם the people כאישׁ man, אחד as one לאמר saying, לא We will not נלך go אישׁ any לאהלו to his tent, ולא neither נסור turn אישׁ will we any לביתו׃ into his house.}%
\verse{ועתה But now זה this הדבר the thing אשׁר which נעשׂה we will do לגבעה to Gibeah; עליה against בגורל׃ by lot}%
\verse{ולקחנו And we will take עשׂרה ten אנשׁים men למאה of a hundred לכל throughout all שׁבטי the tribes ישׂראל of Israel, ומאה and a hundred לאלף of a thousand, ואלף and a thousand לרבבה out of ten thousand, לקחת to fetch צדה victual לעם for the people, לעשׂות that they may do, לבואם when they come לגבע to Gibeah בנימן of Benjamin, ככל according to all הנבלה the folly אשׁר that עשׂה they have wrought בישׂראל׃ in Israel.}%
\verse{ויאסף were gathered כל So all אישׁ the men ישׂראל of Israel אל against העיר the city, כאישׁ man. אחד as one חברים׃ knit together}%
\verse{וישׁלחו sent שׁבטי And the tribes ישׂראל of Israel אנשׁים בכל through all שׁבטי the tribe בנימן of Benjamin, לאמר saying, מה What הרעה wickedness הזאת this אשׁר that נהיתה׃ is done}%
\verse{ועתה Now תנו therefore deliver את האנשׁים the men, בני the children בליעל of Belial, אשׁר which בגבעה in Gibeah, ונמיתם that we may put them to death, ונבערה and put away רעה evil מישׂראל of Israel: ולא not אבו would בנימן of Benjamin לשׁמע hearken בקול to the voice אחיהם of their brethren בני But the children ישׂראל׃}%
\verse{ויאספו gathered themselves together בני But the children בנימן of Benjamin מן out of הערים the cities הגבעתה unto Gibeah, לצאת to go out למלחמה to battle עם against בני the children ישׂראל׃ of Israel.}%
\verse{ויתפקדו were numbered בני And the children בנימן of Benjamin ביום time ההוא at that מהערים out of the cities עשׂרים twenty ושׁשׁה and six אלף thousand אישׁ men שׁלף that drew חרב sword, לבד beside מישׁבי the inhabitants הגבעה of Gibeah, התפקדו which were numbered שׁבע seven מאות hundred אישׁ men. בחור׃ chosen}%
\verse{מכל every one העם people הזה this שׁבע seven מאות hundred אישׁ men בחור אטרד ימינול זה every one קלע could sling באבן stones אל at השׂערה a hair ולא and not יחטא׃ miss.}%
\verse{ואישׁ And the men ישׂראל of Israel, התפקדו were numbered לבד beside מבנימן ארבע four מאות hundred אלף thousand אישׁ men שׁלף that drew חרב sword: כל all זה these אישׁ men מלחמה׃ of war.}%
\verse{ויקמו arose, ויעלו and went up בית אל to the house of God, וישׁאלו and asked counsel באלהים of God, ויאמרו and said, בני And the children ישׂראל of Israel מי Which יעלה of us shall go up לנו בתחלה first למלחמה to the battle עם against בני the children בנימן of Benjamin? ויאמר said, יהוה And the LORD יהודה Judah בתחלה׃ first.}%
\verse{ויקומו rose up בני And the children ישׂראל of Israel בבקר in the morning, ויחנו and encamped על against הגבעה׃ Gibeah.}%
\verse{ויצא went out אישׁ And the men ישׂראל of Israel למלחמה to battle עם against בנימן Benjamin; ויערכו put themselves in array אתם אישׁ and the men ישׂראל of Israel מלחמה to fight אל them at הגבעה׃ Gibeah.}%
\verse{ויצאו came forth בני And the children בנימן of Benjamin מן out of הגבעה Gibeah, וישׁחיתו and destroyed down בישׂראל of the Israelites ביום day ההוא that שׁנים and two ועשׂרים twenty אלף thousand אישׁ men. ארצה׃ to the ground}%
\verse{ויתחזק encouraged themselves, העם And the people אישׁ the men ישׂראל of Israel ויספו again לערך and set מלחמה their battle במקום in the place אשׁר where ערכו in array שׁם where ביום day. הראשׁון׃ the first}%
\verse{ויעלו went up בני (And the children ישׂראל of Israel ויבכו and wept לפני before יהוה the LORD עד until הערב even, וישׁאלו and asked counsel ביהוה of the LORD, לאמר saying, האוסיף again לגשׁת Shall I go up למלחמה to battle עם against בני the children בנימן of Benjamin אחי my brother? ויאמר said, יהוה And the LORD עלו Go up אליו׃ against}%
\verse{ויקרבו came near בני And the children ישׂראל of Israel אל against בני the children בנימן of Benjamin ביום day. השׁני׃ the second}%
\verse{ויצא went forth בנימן And Benjamin לקראתם against מן them out of הגבעה Gibeah ביום day, השׁני the second וישׁחיתו and destroyed בבני of the children ישׂראל of Israel עוד again שׁמנת eighteen עשׂר eighteen אלף thousand אישׁ men; ארצה down to the ground כל all אלה these שׁלפי drew חרב׃ the sword.}%
\verse{ויעלו went up, כל Then all בני the children ישׂראל of Israel, וכל and all העם the people, ויבאו and came בית אל unto the house of God, ויבכו and wept, וישׁבו and sat שׁם there לפני before יהוה the LORD, ויצומו and fasted ביום day ההוא that עד until הערב even, ויעלו and offered עלות burnt offerings ושׁלמים and peace offerings לפני before יהוה׃ the LORD.}%
\verse{וישׁאלו inquired בני And the children ישׂראל of Israel ביהוה of the LORD, ושׁם there ארון (for the ark ברית of the covenant האלהים of God בימים days, ההם׃ in those}%
\verse{ופינחס And Phinehas, בן the son אלעזר of Eleazar, בן the son אהרן of Aaron, עמד stood לפניו before בימים days,) ההם it in those לאמר saying, האוסף again עוד Shall I yet לצאת go out למלחמה to battle עם against בני the children בנימן of Benjamin אחי my brother, אם or אחדל shall I cease? ויאמר said, יהוה And the LORD עלו Go up; כי for מחר tomorrow אתננו I will deliver בידך׃ them into thine hand.}%
\verse{וישׂם set ישׂראל And Israel ארבים liers in wait אל round about הגבעה Gibeah. סביב׃ round about}%
\verse{ויעלו went up בני And the children ישׂראל of Israel אל against בני the children בנימן of Benjamin ביום day, השׁלישׁי on the third ויערכו and put themselves in array אל against הגבעה Gibeah, כפעם as at other times. בפעם׃ as at other times.}%
\verse{ויצאו went out בני And the children בנימן of Benjamin לקראת against העם the people, הנתקו were drawn away מן from העיר the city; ויחלו and they began להכות to smite מהעם of the people, חללים kill, כפעם as at other times, בפעם as at other times, במסלות in the highways, אשׁר of which אחת one עלה goeth up בית אל to the house of God, ואחת and the other גבעתה to Gibeah בשׂדה in the field, כשׁלשׁים about thirty אישׁ men בישׂראל׃ of Israel.}%
\verse{ויאמרו said, בני And the children בנימן of Benjamin נגפים smitten down הם They לפנינו before כבראשׁנה us, as at the first. ובני But the children ישׂראל of Israel אמרו said, ננוסה Let us flee, ונתקנהו and draw מן them from העיר the city אל unto המסלות׃ the highways.}%
\verse{וכל And all אישׁ the men ישׂראל of Israel קמו rose up ממקומו out of their place, ויערכו and put themselves in array בבעל תמר at Baal-tamar: וארב and the liers in wait ישׂראל of Israel מגיח came forth ממקמו out of their places, ממערה out of the meadows גבע׃}%
\verse{ויבאו And there came מנגד against לגבעה Gibeah עשׂרת ten אלפים thousand אישׁ men בחור מכל out of all ישׂראל Israel, והמלחמה and the battle כבדה was sore: והם but they לא not ידעו knew כי that נגעת near עליהם הרעה׃ evil}%
\verse{ויגף smote יהוה And the LORD את בנימן Benjamin לפני before ישׂראל Israel: וישׁחיתו destroyed בני and the children ישׂראל of Israel בבנימן of the Benjamites ביום day ההוא that עשׂרים twenty וחמשׁה and five אלף thousand ומאה and a hundred אישׁ men: כל all אלה these שׁלף drew חרב׃ the sword.}%
\verse{ויראו saw בני So the children בנימן of Benjamin כי that נגפו they were smitten: ויתנו gave אישׁ for the men ישׂראל of Israel מקום place לבנימן to the Benjamites, כי because בטחו they trusted אל unto הארב the liers in wait אשׁר which שׂמו they had set אל beside הגבעה׃ Gibeah.}%
\verse{והארב And the liers in wait החישׁו hasted, ויפשׁטו and rushed אל upon הגבעה Gibeah; וימשׁך drew along, הארב and the liers in wait ויך and smote את כל all העיר the city לפי with the edge חרב׃ of the sword.}%
\verse{והמועד an appointed sign היה Now there was לאישׁ between the men ישׂראל of Israel עם הארב and the liers in wait, הרב that they should make a great להעלותם rise up משׂאת flame העשׁן with smoke מן out of העיר׃ the city.}%
\verse{ויהפך retired אישׁ And when the men ישׂראל of Israel במלחמה in the battle, ובנימן Benjamin החל began להכות to smite חללים kill באישׁ of the men ישׂראל of Israel כשׁלשׁים about thirty אישׁ persons: כי for אמרו they said, אך Surely נגוף are smitten down נגף are smitten down הוא they לפנינו before כמלחמה battle. הראשׁנה׃ us, as the first}%
\verse{והמשׂאת But when the flame החלה began לעלות to arise up מן out of העיר the city עמוד with a pillar עשׁן of smoke, ויפן looked בנימן the Benjamites אחריו behind והנה them, and, behold, עלה ascended up כליל the flame העיר of the city השׁמימה׃ to heaven.}%
\verse{ואישׁ And when the men ישׂראל of Israel הפך turned again, ויבהל were amazed: אישׁ the men בנימן of Benjamin כי for ראה they saw כי that נגעה was come עליו upon הרעה׃ evil}%
\verse{ויפנו Therefore they turned לפני before אישׁ the men ישׂראל of Israel אל unto דרך the way המדבר of the wilderness; והמלחמה but the battle הדביקתהו overtook ואשׁר them; and them which מהערים out of the cities משׁחיתים they destroyed אותו בתוכו׃ in the midst}%
\verse{כתרו אתנימן הרדיפהו chased מנוחה with ease הדריכהו them, trod them down עד over against נכח over against הגבעה Gibeah ממזרח שׁמשׁ׃}%
\verse{ויפלו And there fell מבנימן שׁמנה eighteen עשׂר eighteen אלף thousand אישׁ men; את כל all אלה these אנשׁי men חיל׃ of valor.}%
\verse{ויפנו And they turned וינסו and fled המדברה toward the wilderness אל unto סלע the rock הרמון of Rimmon: ויעללהו and they gleaned במסלות of them in the highways חמשׁת five אלפים thousand אישׁ men; וידביקו and pursued hard אחריו after עד them unto גדעם Gidom, ויכו and slew ממנו of אלפים two thousand אישׁ׃ men}%
\verse{ויהי were כל So that all הנפלים which fell מבנימן עשׂרים twenty וחמשׁה and five אלף thousand אישׁ men שׁלף that drew חרב the sword; ביום day ההוא that את כל all אלה these אנשׁי men חיל׃ of valor.}%
\verse{ויפנו turned וינסו and fled המדברה to the wilderness אל unto סלע the rock הרמון Rimmon, שׁשׁ But six מאות hundred אישׁ men וישׁבו and abode בסלע in the rock רמון Rimmon ארבעה four חדשׁים׃ months.}%
\verse{ואישׁ And the men ישׂראל of Israel שׁבו turned again אל upon בני the children בנימן of Benjamin, ויכום and smote לפי them with the edge חרב of the sword, מעיר of city, מתם as well the men עד as בהמה the beast, עד כל and all הנמצא that came to hand: גם also כל all הערים the cities הנמצאות that they came to. שׁלחו they set באשׁ׃ on fire}%
\end{biblechapter}%
\begin{biblechapter}% Judges 21
\verseWithHeading{A Decision Is Made About the Tribe of Benjamin}{ואישׁ Now the men ישׂראל of Israel נשׁבע had sworn במצפה in Mizpeh, לאמר saying, אישׁ any ממנו of לא There shall not יתן us give בתו his daughter לבנימן unto Benjamin לאשׁה׃ to wife.}%
\verse{ויבא came העם And the people בית אל to the house of God, וישׁבו and abode שׁם there עד till הערב even לפני before האלהים God, וישׂאו and lifted up קולם their voices, ויבכו and wept בכי sore; גדול׃ sore;}%
\verse{ויאמרו And said, למה why יהוה O LORD אלהי God ישׂראל of Israel, היתה come to pass זאת is this בישׂראל in Israel, להפקד lacking היום that there should be today מישׂראל שׁבט tribe אחד׃ one}%
\verse{ויהי And it came to pass ממחרת on the morrow, וישׁכימו rose early, העם that the people ויבנו and built שׁם there מזבח an altar, ויעלו and offered עלות burnt offerings ושׁלמים׃ and peace offerings.}%
\verse{ויאמרו said, בני And the children ישׂראל of Israel מי Who אשׁר that לא came not up עלה came not up בקהל with the congregation מכל among all שׁבטי the tribes ישׂראל of Israel אל unto יהוה the LORD? כי For השׁבועה oath הגדולה a great היתה they had made לאשׁר concerning לא him that came not up עלה him that came not up אל to יהוה the LORD המצפה to Mizpeh, לאמר saying, מות יומת׃}%
\verse{וינחמו repented בני And the children ישׂראל of Israel אל them for בנימן Benjamin אחיו their brother, ויאמרו and said, נגדע cut off היום this day. שׁבט tribe אחד There is one מישׂראל׃}%
\verse{מה How נעשׂה shall we do להם לנותרים for them that remain, לנשׁים for wives ואנחנו seeing we נשׁבענו have sworn ביהוה by the LORD לבלתי that we will not תת give להם מבנותינו them of our daughters לנשׁים׃ to wives?}%
\verse{ויאמרו And they said, מי What אחד one משׁבטי of the tribes ישׂראל of Israel אשׁר that לא came not up עלה came not up אל to יהוה the LORD? המצפה to Mizpeh והנה And, behold, לא none בא there came אישׁ none אל to המחנה the camp מיבישׁ גלעדל to הקהל׃ the assembly.}%
\verse{ויתפקד were numbered, העם For the people והנה and, behold, אין none שׁם there. אישׁ none מיושׁבי of the inhabitants יבשׁ גלעד׃}%
\verse{וישׁלחו sent שׁם hither העדה And the congregation שׁנים twelve עשׂר twelve אלף thousand אישׁ men מבני החיליצוו and commanded אותם לאמר them, saying, לכו Go והכיתם and smite את יושׁבי the inhabitants יבשׁ גלעדפי with the edge חרב of the sword, והנשׁים with the women והטף׃ and the children.}%
\verse{וזה And this הדבר the thing אשׁר that תעשׂו ye shall do, כל every זכר male, וכל and every אשׁה woman ידעת that hath lain משׁכב that hath lain זכר by man. תחרימו׃ Ye shall utterly destroy}%
\verse{וימצאו And they found מיושׁבי among the inhabitants יבישׁ גלעדרבע four מאות hundred נערה young בתולה virgins, אשׁר that לא no ידעה had known אישׁ man למשׁכב by lying זכר with any male: ויביאו and they brought אותם אל them unto המחנה the camp שׁלה to Shiloh, אשׁר which בארץ in the land כנען׃ of Canaan.}%
\verse{וישׁלחו sent כל And the whole העדה congregation וידברו to speak אל to בני the children בנימן of Benjamin אשׁר that בסלע in the rock רמון Rimmon, ויקראו and to call להם שׁלום׃ peaceably}%
\verse{וישׁב came again בנימן And Benjamin בעת time; ההיא at that ויתנו and they gave להם הנשׁים them wives אשׁר which חיו they had saved alive מנשׁי of the women יבשׁ גלעדלא them not. מצאו they sufficed להם כן׃ and yet so}%
\verse{והעם And the people נחם repented them לבנימן for Benjamin, כי because עשׂה had made יהוה that the LORD פרץ a breach בשׁבטי in the tribes ישׂראל׃ of Israel.}%
\verse{ויאמרו said, זקני Then the elders העדה of the congregation מה How נעשׂה shall we do לנותרים for them that remain, לנשׁים for wives כי seeing נשׁמדה are destroyed מבנימן אשׁה׃ the women}%
\verse{ויאמרו And they said, ירשׁת an inheritance פליטה for them that be escaped לבנימן of Benjamin, ולא be not ימחה destroyed שׁבט that a tribe מישׂראל׃}%
\verse{ואנחנו Howbeit we לא not נוכל may לתת give להם נשׁים them wives מבנותינו of our daughters: כי for נשׁבעו have sworn, בני the children ישׂראל of Israel לאמר saying, ארור Cursed נתן he that giveth אשׁה a wife לבנימן׃ to Benjamin.}%
\verse{ויאמרו Then they said, הנה Behold, חג a feast יהוה of the LORD בשׁלו in Shiloh מימים ימימהשׁר which מצפונה on the north side לבית אל מזרחה on the east side השׁמשׁ on the east side למסלה of the highway העלה that goeth up מבית אל שׁכמה to Shechem, ומנגב and on the south ללבונה׃ of Lebonah.}%
\verse{ויצו Therefore they commanded את בני the children בנימן of Benjamin, לאמר saying, לכו Go וארבתם and lie in wait בכרמים׃ in the vineyards;}%
\verse{וראיתם And see, והנה and, behold, אם if יצאו come out בנות the daughters שׁילו of Shiloh לחול to dance במחלות in dances, ויצאתם then come ye out מן of הכרמים the vineyards, וחטפתם and catch לכם אישׁ you every man אשׁתו his wife מבנות of the daughters שׁילו of Shiloh, והלכתם and go ארץ to the land בנימן׃ of Benjamin.}%
\verse{והיה And it shall be, כי when יבאו come אבותם their fathers או or אחיהם their brethren לרוב us to complain, אלינו unto ואמרנו that we will say אליהם unto חנונו them, Be favorable אותם כי unto them for our sakes: because לא not לקחנו we reserved אישׁ to each man אשׁתו his wife במלחמה in the war: כי for לא did not אתם ye נתתם give להם כעת unto them at this time, תאשׁמו׃ ye should be guilty.}%
\verse{ויעשׂו did כן so, בני And the children בנימן of Benjamin וישׂאו and took נשׁים wives, למספרם according to their number, מן of המחללות them that danced, אשׁר whom גזלו they caught: וילכו and they went וישׁובו and returned אל unto נחלתם their inheritance, ויבנו and repaired את הערים the cities, וישׁבו׃ and dwelt}%
\verse{ויתהלכו departed משׁם thence בני And the children ישׂראל of Israel בעת time, ההיא at that אישׁ every man לשׁבטו to his tribe ולמשׁפחתו and to his family, ויצאו and they went out משׁם from thence אישׁ every man לנחלתו׃ to his inheritance.}%
\verse{בימים days ההם In those אין no מלך king בישׂראל in Israel: אישׁ every man הישׁר right בעיניו in his own eyes. יעשׂה׃ did}%
\end{biblechapter}%
