\biblebook{Lamentations}
\begin{biblechapter}% Lamentations 1
\verseWithHeading{The Desolate City}{איכה How ישׁבה sit בדד solitary, העיר doth the city רבתי full עם of people! היתה is she become כאלמנה as a widow! רבתי she great בגוים among the nations, שׂרתי princess במדינות among the provinces, היתה is she become למס׃ tributary!}%
\verse{בכו תבכהלילה in the night, ודמעתה and her tears על on לחיה her cheeks: אין she hath none לה מנחם to comfort מכל among all אהביה her lovers כל all רעיה her friends בגדו have dealt treacherously בה היו with her, they are become לה לאיבים׃ her enemies.}%
\verse{גלתה is gone into captivity יהודה Judah מעני because of affliction, ומרב and because of great עבדה servitude: היא she ישׁבה dwelleth בגוים among the heathen, לא no מצאה she findeth מנוח rest: כל all רדפיה her persecutors השׂיגוה overtook בין her between המצרים׃ the straits.}%
\verse{דרכי The ways ציון of Zion אבלות do mourn, מבלי because none באי come מועד to the solemn feasts: כל all שׁעריה her gates שׁוממין are desolate: כהניה her priests נאנחים sigh, בתולתיה her virgins נוגות are afflicted, והיא and she מר׃}%
\verse{היו are צריה Her adversaries לראשׁ the chief, איביה her enemies שׁלו prosper; כי for יהוה the LORD הוגה hath afflicted על her for רב the multitude פשׁעיה of her transgressions: עולליה her children הלכו are gone שׁבי into captivity לפני before צר׃ the enemy.}%
\verse{ויצא is departed: מן And from בת the daughter ציון כל all הדרה her beauty היו are become שׂריה her princes כאילים like harts לא no מצאו find מרעה pasture, וילכו and they are gone בלא without כח strength לפני before רודף׃ the pursuer.}%
\verse{זכרה remembered ירושׁלם Jerusalem ימי in the days עניה of her affliction ומרודיה and of her miseries כל all מחמדיה her pleasant things אשׁר that היו she had מימי in the days קדם of old, בנפל fell עמה when her people ביד into the hand צר of the enemy, ואין and none עוזר did help לה ראוה saw צרים her: the adversaries שׂחקו her, did mock על at משׁבתה׃ her sabbaths.}%
\verse{חטא hath grievously sinned; חטאה hath grievously sinned; ירושׁלם Jerusalem על therefore כן therefore לנידה removed: היתה she is כל all מכבדיה that honored הזילוה כי her, because ראו they have seen ערותה her nakedness: גם yea, היא she נאנחה sigheth, ותשׁב and turneth אחור׃ backward.}%
\verse{טמאתה Her filthiness בשׁוליה in her skirts; לא not זכרה she remembereth אחריתה her last end; ותרד therefore she came down פלאים wonderfully: אין she had no מנחם comforter. לה ראה behold יהוה O LORD, את עניי my affliction: כי for הגדיל hath magnified אויב׃ the enemy}%
\verse{ידו his hand פרשׂ hath spread out צר The adversary על upon כל all מחמדיה her pleasant things: כי for ראתה she hath seen גוים the heathen באו entered מקדשׁה into her sanctuary, אשׁר whom צויתה thou didst command לא they should not יבאו enter בקהל׃ into thy congregation.}%
\verse{כל All עמה her people נאנחים sigh, מבקשׁים they seek לחם bread; נתנו they have given מחמודיהם their pleasant things באכל for meat להשׁיב to relieve נפשׁ the soul: ראה see, יהוה O LORD, והביטה and consider; כי for הייתי I am become זוללה׃ vile.}%
\verse{לוא nothing אליכם to כל you, all עברי ye that pass by? דרך ye that pass by? הביטו behold, וראו and see אם if ישׁ there be מכאוב any sorrow כמכאבי like unto my sorrow, אשׁר which עולל is done לי אשׁר unto me, wherewith הוגה hath afflicted יהוה the LORD ביום in the day חרון of his fierce אפו׃ anger.}%
\verse{ממרום שׁלח hath he sent אשׁ fire בעצמתי into my bones, וירדנה and it prevaileth against פרשׂ them: he hath spread רשׁת a net לרגלי for my feet, השׁיבני he hath turned אחור me back: נתנני he hath made שׁממה כל all היום the day. דוה׃ faint}%
\verse{נשׂקד is bound על The yoke פשׁעי of my transgressions בידו by his hand: ישׂתרגו they are wreathed, עלו come up על upon צוארי my neck: הכשׁיל to fall, כחי he hath made my strength נתנני hath delivered אדני the Lord בידי me into hands, לא אוכלום׃ to rise up.}%
\verse{סלה hath trodden under foot כל all אבירי my mighty אדני The Lord בקרבי in the midst קרא of me: he hath called עלי against מועד an assembly לשׁבר me to crush בחורי my young men: גת in a winepress. דרך hath trodden אדני the Lord לבתולת the virgin, בת the daughter יהודה׃ of Judah,}%
\verse{על For אלה these אני I בוכיה weep; עיני mine eye, עיני mine eye ירדה runneth down מים with water, כי because רחק is far ממני from מנחם the comforter משׁיב that should relieve נפשׁי my soul היו are בני me: my children שׁוממים desolate, כי because גבר prevailed. אויב׃ the enemy}%
\verse{פרשׂה spreadeth forth ציון Zion בידיה her hands, אין none מנחם to comfort לה צוה hath commanded יהוה her: the LORD ליעקב concerning Jacob, סביביו round about צריו his adversaries היתה is ירושׁלם him: Jerusalem לנדה as a menstruous woman ביניהם׃ among}%
\verse{צדיק is righteous; הוא יהוהי for פיהו against his commandment: מריתי I have rebelled שׁמעו hear, נא I pray you, כל all עמים people, וראו and behold מכאבי my sorrow: בתולתי my virgins ובחורי and my young men הלכו are gone בשׁבי׃ into captivity.}%
\verse{קראתי I called למאהבי for my lovers, המה they רמוני deceived כהני me: my priests וזקני and mine elders בעיר in the city, גועו gave up the ghost כי while בקשׁו they sought אכל their meat למו וישׁיבו to relieve את נפשׁם׃ their souls.}%
\verse{ראה Behold, יהוה O LORD; כי for צר I in distress: לי מעי my bowels חמרמרו are troubled; נהפך is turned לבי mine heart בקרבי within כי me; for מרו מריתיחוץ abroad שׁכלה bereaveth, חרב the sword בבית at home כמות׃ as death.}%
\verse{שׁמעו They have heard כי that נאנחה sigh: אני I אין none מנחם to comfort לי כל me: all איבי mine enemies שׁמעו have heard רעתי of my trouble; שׂשׂו they are glad כי that אתה thou עשׂית hast done הבאת thou wilt bring יום the day קראת thou hast called, ויהיו and they shall be כמוני׃ like unto me.}%
\verse{תבא come כל Let all רעתם their wickedness לפניך before ועולל thee; and do למו כאשׁר unto them, as עוללת thou hast done לי על unto me for כל all פשׁעי my transgressions: כי for רבות many, אנחתי my sighs ולבי and my heart דוי׃ faint.}%
\end{biblechapter}%
\begin{biblechapter}% Lamentations 2
\verseWithHeading{The Lord is Angry}{איכה How יעיב covered באפו with a cloud in his anger, אדני hath the Lord את בת the daughter ציון of Zion השׁליך cast down משׁמים from heaven ארץ unto the earth תפארת the beauty ישׂראל of Israel, ולא not זכר and remembered הדם his footstool רגליו his footstool ביום in the day אפו׃ of his anger!}%
\verse{בלע hath swallowed up אדני The Lord לא and hath not חמל pitied: את כל all נאות the habitations יעקב of Jacob, הרס he hath thrown down בעברתו in his wrath מבצרי the strongholds בת of the daughter יהודה of Judah; הגיע he hath brought down לארץ to the ground: חלל he hath polluted ממלכה the kingdom ושׂריה׃ and the princes}%
\verse{גדע He hath cut off בחרי in fierce אף anger כל all קרן the horn ישׂראל of Israel: השׁיב he hath drawn אחור back ימינו his right hand מפני from before אויב the enemy, ויבער and he burned ביעקב against Jacob כאשׁ fire, להבה like a flaming אכלה devoureth סביב׃ round about.}%
\verse{דרך He hath bent קשׁתו his bow כאויב like an enemy: נצב he stood ימינו with his right hand כצר as an adversary, ויהרג and slew כל all מחמדי pleasant עין to the eye באהל in the tabernacle בת of the daughter ציון of Zion: שׁפך he poured out כאשׁ like fire. חמתו׃ his fury}%
\verse{היה was אדני The Lord כאויב as an enemy: בלע he hath swallowed up ישׂראל Israel, בלע he hath swallowed up כל all ארמנותיה her palaces: שׁחת he hath destroyed מבצריו his strongholds, וירב and hath increased בבת in the daughter יהודה of Judah תאניה mourning ואניה׃ and lamentation.}%
\verse{ויחמס And he hath violently taken away כגן as a garden: שׂכו his tabernacle, שׁחת he hath destroyed מועדו his places of the assembly: שׁכח to be forgotten יהוה the LORD בציון in Zion, מועד hath caused the solemn feasts ושׁבת and sabbaths וינאץ and hath despised בזעם in the indignation אפו of his anger מלך the king וכהן׃ and the priest.}%
\verse{זנח hath cast off אדני The Lord מזבחו his altar, נאר he hath abhorred מקדשׁו his sanctuary, הסגיר he hath given up ביד into the hand אויב of the enemy חומת the walls ארמנותיה of her palaces; קול a noise נתנו they have made בבית in the house יהוה of the LORD, כיום as in the day מועד׃ of a solemn feast.}%
\verse{חשׁב hath purposed יהוה The LORD להשׁחית to destroy חומת the wall בת of the daughter ציון of Zion: נטה he hath stretched out קו a line, לא he hath not השׁיב withdrawn ידו his hand מבלע from destroying: ויאבל to lament; חל therefore he made the rampart וחומה and the wall יחדו together. אמללו׃ they languished}%
\verse{טבעו are sunk בארץ into the ground; שׁעריה Her gates אבד he hath destroyed ושׁבר and broken בריחיה her bars: מלכה her king ושׂריה and her princes בגוים among the Gentiles: אין no תורה the law גם also נביאיה her prophets לא no מצאו find חזון vision מיהוה׃}%
\verse{ישׁבו sit לארץ upon the ground, ידמו keep silence: זקני The elders בת of the daughter ציון of Zion העלו they have cast up עפר dust על upon ראשׁם their heads; חגרו they have girded שׂקים themselves with sackcloth: הורידו hang down לארץ to the ground. ראשׁן their heads בתולת the virgins ירושׁלם׃ of Jerusalem}%
\verse{כלו do fail בדמעות with tears, עיני Mine eyes חמרמרו are troubled, מעי my bowels נשׁפך is poured לארץ upon the earth, כבדי my liver על for שׁבר the destruction בת of the daughter עמי of my people; בעטף swoon עולל because the children ויונק and the sucklings ברחבות in the streets קריה׃ of the city.}%
\verse{לאמתם to their mothers, יאמרו They say איה Where דגן corn ויין and wine? בהתעטפם when they swooned כחלל as the wounded ברחבות in the streets עיר of the city, בהשׁתפך was poured out נפשׁם when their soul אל into חיק bosom. אמתם׃ their mothers'}%
\verse{מה What thing אעידך shall I take to witness מה for thee? what thing אדמה shall I liken לך הבת to thee, O daughter ירושׁלם of Jerusalem? מה what אשׁוה shall I equal לך ואנחמך to thee, that I may comfort בתולת thee, O virgin בת daughter ציון of Zion? כי for גדול great כים like the sea: שׁברך thy breach מי who ירפא׃ can heal}%
\verse{נביאיך Thy prophets חזו have seen לך שׁוא vain ותפל and foolish things ולא for thee: and they have not גלו discovered על discovered עונך thine iniquity, להשׁיב to turn away שׁביתך thy captivity; ויחזו but have seen לך משׂאות burdens שׁוא for thee false ומדוחים׃ and causes of banishment.}%
\verse{ספקו clap עליך at כפים hands כל All עברי that pass by דרך that pass by שׁרקו thee; they hiss וינעו and wag ראשׁם their head על at בת the daughter ירושׁלם of Jerusalem, הזאת this העיר the city שׁיאמרו that call כלילת The perfection יפי of beauty, משׂושׂ The joy לכל of the whole הארץ׃ earth?}%
\verse{פצו have opened עליך against פיהם their mouth כל All אויביך thine enemies שׁרקו thee: they hiss ויחרקו and gnash שׁן the teeth: אמרו they say, בלענו We have swallowed up: אך certainly זה this היום the day שׁקוינהו that we looked for; מצאנו we have found, ראינו׃ we have seen}%
\verse{עשׂה hath done יהוה The LORD אשׁר which זמם he had devised; בצע he hath fulfilled אמרתו his word אשׁר that צוה he had commanded מימי in the days קדם of old: הרס he hath thrown down, ולא and hath not חמל pitied: וישׂמח to rejoice עליך over אויב and he hath caused enemy הרים thee, he hath set up קרן the horn צריך׃ of thine adversaries.}%
\verse{צעק cried לבם Their heart אל unto אדני the Lord, חומת O wall בת of the daughter ציון of Zion, הורידי run down כנחל like a river דמעה let tears יומם day ולילה and night: אל thyself no תתני give פוגת rest; לך אל let not תדם cease. בת the apple עינך׃ of thine eye}%
\verse{קומי Arise, רני cry out בליל in the night: לראשׁ in the beginning אשׁמרות of the watches שׁפכי pour out כמים like water לבך thine heart נכח before פני the face אדני of the Lord: שׂאי lift up אליו toward כפיך thy hands על him for נפשׁ the life עולליך of thy young children, העטופים that faint ברעב for hunger בראשׁ in the top כל of every חוצות׃ street.}%
\verse{ראה Behold, יהוה O LORD, והביטה and consider למי to whom עוללת thou hast done כה this. אם תאכלנה eat נשׁים Shall the women פרים their fruit, עללי children טפחים of a span long? אם יהרג be slain במקדשׁ in the sanctuary אדני of the Lord? כהן shall the priest ונביא׃ and the prophet}%
\verse{שׁכבו lie לארץ on the ground חוצות in the streets: נער The young וזקן and the old בתולתי my virgins ובחורי and my young men נפלו are fallen בחרב by the sword; הרגת thou hast slain ביום in the day אפך of thine anger; טבחת thou hast killed, לא not חמלת׃ pitied.}%
\verse{תקרא Thou hast called כיום day מועד as in a solemn מגורי my terrors מסביב round about, ולא none היה escaped ביום so that in the day אף anger יהוה of the LORD's פליט escaped ושׂריד nor remained: אשׁר those that טפחתי I have swaddled ורביתי and brought up איבי hath mine enemy כלם׃ consumed.}%
\end{biblechapter}%
\begin{biblechapter}% Lamentations 3
\verseWithHeading{Israel’s Affliction}{אני I הגבר the man ראה hath seen עני affliction בשׁבט by the rod עברתו׃ of his wrath.}%
\verse{אותי נהג He hath led וילך me, and brought חשׁך darkness, ולא but not אור׃ light.}%
\verse{אך Surely בי ישׁב against me is he turned; יהפך he turneth ידו his hand כל all היום׃ the day.}%
\verse{בלה hath he made old; בשׂרי My flesh ועורי and my skin שׁבר he hath broken עצמותי׃ my bones.}%
\verse{בנה He hath built עלי against ויקף me, and compassed ראשׁ with gall ותלאה׃ and travail.}%
\verse{במחשׁכים me in dark places, הושׁיבני He hath set כמתי as dead עולם׃ of old.}%
\verse{גדר He hath hedged בעדי me about, ולא that I cannot אצא get out: הכביד he hath made my chain heavy. נחשׁתי׃ he hath made my chain heavy.}%
\verse{גם Also כי when אזעק I cry ואשׁוע and shout, שׂתם he shutteth out תפלתי׃ my prayer.}%
\verse{גדר He hath enclosed דרכי my ways בגזית with hewn stone, נתיבתי he hath made my paths crooked. עוה׃ he hath made my paths crooked.}%
\verse{דב unto me a bear ארב lying in wait, הוא He לי אריה a lion במסתרים׃ in secret places.}%
\verse{דרכי my ways, סורר He hath turned aside ויפשׁחני and pulled me in pieces: שׂמני he hath made שׁמם׃}%
\verse{דרך He hath bent קשׁתו his bow, ויציבני and set כמטרא me as a mark לחץ׃ for the arrow.}%
\verse{הביא to enter בכליותי into my reins. בני He hath caused the arrows אשׁפתו׃ of his quiver}%
\verse{הייתי I was שׂחק a derision לכל to all עמי my people; נגינתם their song כל all היום׃ the day.}%
\verse{השׂביעני He hath filled במרורים me with bitterness, הרוני he hath made me drunken לענה׃ with wormwood.}%
\verse{ויגרס He hath also broken בחצץ with gravel stones, שׁני my teeth הכפישׁני he hath covered באפר׃ me with ashes.}%
\verse{ותזנח משׁלום from peace: נפשׁי נשׁיתי I forgot טובה׃ prosperity.}%
\verse{ואמר And I said, אבד is perished נצחי My strength ותוחלתי and my hope מיהוה׃}%
\verse{זכר Remembering עניי mine affliction ומרודי and my misery, לענה the wormwood וראשׁ׃ and the gall.}%
\verse{זכור hath still in remembrance, תזכור hath still in remembrance, ותשׁיח and is humbled עלי in נפשׁי׃ My soul}%
\verse{זאת This אשׁיב I recall אל to לבי my mind, על therefore כן therefore אוחיל׃ have I hope.}%
\verse{חסדי mercies יהוה the LORD's כי that לא we are not תמנו consumed, כי because לא not. כלו fail רחמיו׃ his compassions}%
\verse{חדשׁים new לבקרים every morning: רבה great אמונתך׃ thy faithfulness.}%
\verse{חלקי my portion, יהוה The LORD אמרה saith נפשׁי my soul; על therefore כן therefore אוחיל׃ will I hope}%
\verse{טוב good יהוה The LORD לקוו unto them that wait for לנפשׁ him, to the soul תדרשׁנו׃ seeketh}%
\verse{טוב good ויחיל ודומם and quietly wait לתשׁועת for the salvation יהוה׃ of the LORD.}%
\verse{טוב good לגבר for a man כי that ישׂא he bear על the yoke בנעוריו׃ in his youth.}%
\verse{ישׁב He sitteth בדד alone וידם and keepeth silence, כי because נטל he hath borne עליו׃ upon}%
\verse{יתן He putteth בעפר in the dust; פיהו his mouth אולי if so be ישׁ there may be תקוה׃ hope.}%
\verse{יתן He giveth למכהו to him that smiteth לחי cheek ישׂבע him: he is filled full בחרפה׃ with reproach.}%
\verse{כי For לא will not יזנח cast off לעולם forever: אדני׃ the Lord}%
\verse{כי But אם though הוגה he cause grief, ורחם yet will he have compassion כרב according to the multitude חסדו׃ of his mercies.}%
\verse{כי For לא he doth not ענה afflict מלבו willingly ויגה nor grieve בני the children אישׁ׃ of men.}%
\verse{לדכא To crush תחת under רגליו his feet כל all אסירי the prisoners ארץ׃ of the earth,}%
\verse{להטות To turn aside משׁפט the right גבר of a man נגד before פני the face עליון׃ of the most High,}%
\verse{לעות To subvert אדם a man בריבו in his cause, אדני the Lord לא not. ראה׃ approveth}%
\verse{מי Who זה he אמר saith, ותהי and it cometh to pass, אדני the Lord לא not? צוה׃ commandeth}%
\verse{מפי עליון of the most High לא not תצא proceedeth הרעות evil והטוב׃ and good?}%
\verse{מה Wherefore יתאונן complain, אדם man חי doth a living גבר a man על for חטאו׃ the punishment of his sins?}%
\verse{נחפשׂה Let us search דרכינו our ways, ונחקרה and try ונשׁובה and turn again עד to יהוה׃ the LORD.}%
\verse{נשׂא Let us lift up לבבנו our heart אל with כפים hands אל unto אל God בשׁמים׃ in the heavens.}%
\verse{נחנו We פשׁענו have transgressed ומרינו and have rebelled: אתה thou לא hast not סלחת׃ pardoned.}%
\verse{סכתה Thou hast covered באף with anger, ותרדפנו and persecuted הרגת us: thou hast slain, לא thou hast not חמלת׃ pitied.}%
\verse{סכותה Thou hast covered בענן thyself with a cloud, לך מעבור should not pass through. תפלה׃ that prayer}%
\verse{סחי us the offscouring ומאוס and refuse תשׂימנו Thou hast made בקרב in the midst העמים׃ of the people.}%
\verse{פצו have opened עלינו against פיהם their mouths כל All איבינו׃ our enemies}%
\verse{פחד Fear ופחת and a snare היה is come לנו השׁאת upon us, desolation והשׁבר׃ and destruction.}%
\verse{פלגי with rivers מים of water תרד runneth down עיני Mine eye על for שׁבר the destruction בת of the daughter עמי׃ of my people.}%
\verse{עיני Mine eye נגרה trickleth down, ולא not, תדמה and ceaseth מאין without any הפגות׃ intermission,}%
\verse{עד Till ישׁקיף look down, וירא and behold יהוה the LORD משׁמים׃ from heaven.}%
\verse{עיני Mine eye עוללה affecteth לנפשׁי mine heart מכל because of all בנות the daughters עירי׃ of my city.}%
\verse{צוד chased me sore, צדוני chased me sore, כצפור like a bird, איבי Mine enemies חנם׃ without cause.}%
\verse{צמתו They have cut off בבור in the dungeon, חיי my life וידו and cast אבן׃ a stone}%
\verse{צפו flowed מים Waters על over ראשׁי mine head; אמרתי I said, נגזרתי׃ I am cut off.}%
\verse{קראתי I called upon שׁמך thy name, יהוה O LORD, מבור תחתיות׃}%
\verse{קולי my voice: שׁמעת Thou hast heard אל not תעלם hide אזנך thine ear לרוחתי at my breathing, לשׁועתי׃ at my cry.}%
\verse{קרבת Thou drewest near ביום in the day אקראך I called upon אמרת thee: thou saidst, אל not. תירא׃ Fear}%
\verse{רבת thou hast pleaded אדני O Lord, ריבי the causes נפשׁי of my soul; גאלת thou hast redeemed חיי׃ my life.}%
\verse{ראיתה thou hast seen יהוה O LORD, עותתי my wrong: שׁפטה judge משׁפטי׃ thou my cause.}%
\verse{ראיתה Thou hast seen כל all נקמתם their vengeance כל all מחשׁבתם׃ their imaginations}%
\verse{שׁמעת Thou hast heard חרפתם their reproach, יהוה O LORD, כל all מחשׁבתם their imaginations עלי׃ against}%
\verse{שׂפתי The lips קמי of those that rose up against והגיונם me, and their device עלי against כל me all היום׃ the day.}%
\verse{שׁבתם their sitting down, וקימתם and their rising up; הביטה Behold אני I מנגינתם׃ their music.}%
\verse{תשׁיב Render להם גמול unto them a recompense, יהוה O LORD, כמעשׂה according to the work ידיהם׃ of their hands.}%
\verse{תתן Give להם מגנת them sorrow לב of heart, תאלתך thy curse להם׃}%
\verse{תרדף Persecute באף them in anger ותשׁמידם and destroy מתחת from under שׁמי the heavens יהוה׃ of the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Lamentations 4
\verseWithHeading{Zion Is Punished}{איכה How יועם become dim! זהב is the gold ישׁנא changed! הכתם fine gold הטוב is the most תשׁתפכנה are poured out אבני the stones קדשׁ of the sanctuary בראשׁ in the top כל of every חוצות׃ street.}%
\verse{בני sons ציון of Zion, היקרים The precious המסלאים comparable בפז to fine gold, איכה how נחשׁבו are they esteemed לנבלי pitchers, חרשׂ as earthen מעשׂה the work ידי of the hands יוצר׃ of the potter!}%
\verse{גם Even תנין חלצו draw out שׁד the breast, היניקו they give suck גוריהן to their young ones: בת the daughter עמי of my people לאכזר cruel, כי like the ostriches ענים במדבר׃ in the wilderness.}%
\verse{דבק cleaveth לשׁון The tongue יונק of the sucking child אל to חכו the roof of his mouth בצמא for thirst: עוללים the young children שׁאלו ask לחם bread, פרשׂ breaketh אין no man להם׃}%
\verse{האכלים They that did feed למעדנים delicately נשׁמו are desolate בחוצות in the streets: האמנים they that were brought up עלי in תולע scarlet חבקו embrace אשׁפתות׃ dunghills.}%
\verse{ויגדל is greater עון For the punishment of the iniquity בת of the daughter עמי of my people מחטאת than the punishment of the sin סדם of Sodom, ההפוכה that was overthrown כמו as in רגע a moment, ולא and no חלו stayed בה ידים׃ hands}%
\verse{זכו were purer נזיריה Her Nazarites משׁלג than snow, צחו they were whiter מחלב than milk, אדמו they were more ruddy עצם in body מפנינים than rubies, ספיר of sapphire: גזרתם׃ their polishing}%
\verse{חשׁך is blacker משׁחור than a coal; תארם Their visage לא they are not נכרו known בחוצות in the streets: צפד cleaveth עורם their skin על to עצמם their bones; יבשׁ it is withered, היה it is become כעץ׃ like a stick.}%
\verse{טובים better היו are חללי slain חרב with the sword מחללי than slain רעב with hunger: שׁהם for these יזובו pine away, מדקרים stricken through מתנובת for the fruits שׂדי׃ of the field.}%
\verse{ידי The hands נשׁים women רחמניות of the pitiful בשׁלו have sodden ילדיהן their own children: היו they were לברות their meat למו בשׁבר in the destruction בת of the daughter עמי׃ of my people.}%
\verse{כלה hath accomplished יהוה The LORD את חמתו his fury; שׁפך he hath poured out חרון his fierce אפו anger, ויצת and hath kindled אשׁ a fire בציון in Zion, ותאכל and it hath devoured יסודתיה׃ the foundations}%
\verse{לא would not האמינו have believed מלכי The kings ארץ of the earth, וכל and all ישׁבי the inhabitants תבל of the world, כי that יבא should have entered צר the adversary ואויב and the enemy בשׁערי into the gates ירושׁלם׃ of Jerusalem.}%
\verse{מחטאת נביאיה of her prophets, עונות the iniquities כהניה of her priests, השׁפכים that have shed בקרבה in the midst דם the blood צדיקים׃ of the just}%
\verse{נעו They have wandered עורים blind בחוצות in the streets, נגאלו they have polluted בדם themselves with blood, בלא not יוכלו so that men could יגעו touch בלבשׁיהם׃ their garments.}%
\verse{סורו unto them, Depart טמא ye; unclean; קראו They cried למו סורו depart, סורו depart, אל not: תגעו touch כי when נצו they fled away גם and נעו wandered, אמרו they said בגוים among the heathen, לא They shall no יוסיפו more לגור׃ sojourn}%
\verse{פני The anger יהוה of the LORD חלקם hath divided לא them; he will no יוסיף more להביטם regard פני the persons כהנים of the priests, לא not נשׂאו them: they respected זקנים the elders. לא not חננו׃ they favored}%
\verse{עודינה תכלינה failed עינינו As for us, our eyes אל for עזרתנו help: הבל our vain בצפיתנו we have watched צפינו אל for גוי a nation לא could not יושׁע׃ save}%
\verse{צדו They hunt צעדינו our steps, מלכת that we cannot go ברחבתינו in our streets: קרב is near, קצינו our end מלאו are fulfilled; ימינו our days כי for בא is come. קצינו׃ our end}%
\verse{קלים swifter היו are רדפינו Our persecutors מנשׁרי than the eagles שׁמים of the heaven: על us upon ההרים the mountains, דלקנו they pursued במדבר for us in the wilderness. ארבו׃ they laid wait}%
\verse{רוח The breath אפינו of our nostrils, משׁיח the anointed יהוה of the LORD, נלכד was taken בשׁחיתותם in their pits, אשׁר of whom אמרנו we said, בצלו Under his shadow נחיה we shall live בגוים׃ among the heathen.}%
\verse{שׂישׂי Rejoice ושׂמחי בת O daughter אדום of Edom, יושׁבתי that dwellest בארץ in the land עוץ of Uz; גם also עליך unto תעבר shall pass through כוס the cup תשׁכרי thee: thou shalt be drunken, ותתערי׃ and shalt make thyself naked.}%
\verse{תם is accomplished, עונך The punishment of thine iniquity בת O daughter ציון of Zion; לא he will no יוסיף more להגלותך carry thee away into captivity: פקד he will visit עונך thine iniquity, בת O daughter אדום of Edom; גלה he will discover על he will discover חטאתיך׃ thy sins.}%
\end{biblechapter}%
\begin{biblechapter}% Lamentations 5
\verseWithHeading{A Request for Mercy}{זכר Remember, יהוה O LORD, מה what היה is come לנו הביט upon us: consider, וראה and behold את חרפתנו׃ our reproach.}%
\verse{נחלתנו Our inheritance נהפכה is turned לזרים to strangers, בתינו our houses לנכרים׃ to aliens.}%
\verse{יתומים orphans היינו We are אין and fatherless, אב and fatherless, אמתינו our mothers כאלמנות׃ as widows.}%
\verse{מימינו our water בכסף for money; שׁתינו We have drunken עצינו our wood במחיר is sold יבאו׃ is sold}%
\verse{על under צוארנו Our necks נרדפנו persecution: יגענו we labor, לא have no rest. הונח׃ have no rest.}%
\verse{מצרים נתנו We have given יד the hand אשׁור the Assyrians, לשׂבע to be satisfied לחם׃ with bread.}%
\verse{אבתינו Our fathers חטאו have sinned, אינם not; אנחנו and we עונתיהם their iniquities. סבלנו׃ have borne}%
\verse{עבדים Servants משׁלו have ruled over בנו פרק that doth deliver אין us: none מידם׃ out of their hand.}%
\verse{בנפשׁנו with our lives נביא We got לחמנו our bread מפני because of חרב the sword המדבר׃ of the wilderness.}%
\verse{עורנו Our skin כתנור like an oven נכמרו was black מפני because of זלעפות the terrible רעב׃ famine.}%
\verse{נשׁים the women בציון in Zion, ענו They ravished בתלת the maids בערי in the cities יהודה׃ of Judah.}%
\verse{שׂרים Princes בידם by their hand: נתלו are hanged up פני the faces זקנים of elders לא were not נהדרו׃ honored.}%
\verse{בחורים the young men טחון to grind, נשׂאו They took ונערים and the children בעץ under the wood. כשׁלו׃ fell}%
\verse{זקנים The elders משׁער from the gate, שׁבתו have ceased בחורים the young men מנגינתם׃ from their music.}%
\verse{שׁבת is ceased; משׂושׂ The joy לבנו of our heart נהפך is turned לאבל into mourning. מחלנו׃ our dance}%
\verse{נפלה is fallen עטרת The crown ראשׁנו our head: אוי woe נא לנו כי unto us, that חטאנו׃ we have sinned!}%
\verse{על For זה this היה is דוה faint; לבנו our heart על for אלה these חשׁכו are dim. עינינו׃ our eyes}%
\verse{על Because of הר the mountain ציון of Zion, שׁשׁמם is desolate, שׁועלים the foxes הלכו׃ walk}%
\verse{אתה Thou, יהוה לעולם forever; תשׁב remainest כסאך thy throne לדר from generation ודור׃ to generation.}%
\verse{למה Wherefore לנצח us forever, תשׁכחנו dost thou forget תעזבנו forsake לארך us so long ימים׃ time?}%
\verse{השׁיבנו Turn יהוה thee, O LORD, אליך thou us unto ונשׁוב and we shall be turned; חדשׁ renew ימינו our days כקדם׃ as of old.}%
\verse{כי אםאס thou hast utterly rejected מאסתנו קצפת us; thou art very wroth עלינו against עד מאד׃}%
\end{biblechapter}%
