\biblebook{Nehemiah}
\begin{biblechapter}% Nehemiah 1
\verseWithHeading{Nehemiah’s Prayer for the People}{דברי The words נחמיה of Nehemiah בן the son חכליה of Hachaliah. ויהי And it came to pass בחדשׁ in the month כסלו Chisleu, שׁנת year, עשׂרים in the twentieth ואני as I הייתי was בשׁושׁן in Shushan הבירה׃ the palace,}%
\verse{ויבא came, חנני That Hanani, אחד one מאחי of my brethren, הוא he ואנשׁים and men מיהודה ואשׁאלם and I asked על them concerning היהודים the Jews הפליטה that had escaped, אשׁר which נשׁארו were left מן of my brethren, השׁבי the captivity, ועל and concerning ירושׁלם׃ Jerusalem.}%
\verse{ויאמרו And they said לי הנשׁארים unto me, The remnant אשׁר that נשׁארו are left מן of השׁבי the captivity שׁם there במדינה in the province ברעה affliction גדלה in great ובחרפה and reproach: וחומת the wall ירושׁלם of Jerusalem מפרצת also broken down, ושׁעריה and the gates נצתו thereof are burned באשׁ׃ with fire.}%
\verse{ויהי And it came to pass, כשׁמעי when I heard את הדברים words, האלה these ישׁבתי that I sat down ואבכה and wept, ואתאבלה and mourned ימים days, ואהי and fasted, צם and fasted, ומתפלל and prayed לפני before אלהי the God השׁמים׃ of heaven,}%
\verse{ואמר And said, אנא I beseech יהוה אלהי God השׁמים of heaven, האל God, הגדול the great והנורא and terrible שׁמר that keepeth הברית covenant וחסד and mercy לאהביו for them that love ולשׁמרי him and observe מצותיו׃ his commandments:}%
\verse{תהי be נא now אזנך Let thine ear קשׁבת attentive, ועיניך and thine eyes פתוחות open, לשׁמע that thou mayest hear אל that thou mayest hear תפלת the prayer עבדך of thy servant, אשׁר which אנכי I מתפלל pray לפניך before היום thee now, יומם day ולילה and night, על for בני the children ישׂראל of Israel עבדיך thy servants, ומתודה and confess על and confess חטאות the sins בני of the children ישׂראל of Israel, אשׁר which חטאנו we have sinned לך ואני against thee: both I ובית house אבי and my father's חטאנו׃ have sinned.}%
\verse{חבל חבלנוך ולא against thee, and have not שׁמרנו kept את המצות the commandments, ואת החקים nor the statutes, ואת המשׁפטים nor the judgments, אשׁר which צוית thou commandedst את משׁה Moses. עבדך׃ thy servant}%
\verse{זכר Remember, נא I beseech את הדבר thee, the word אשׁר that צוית thou commandedst את משׁה Moses, עבדך thy servant לאמר saying, אתם ye תמעלו transgress, אני I אפיץ אתכםעמים׃ among the nations:}%
\verse{ושׁבתם But ye turn אלי unto ושׁמרתם me, and keep מצותי my commandments, ועשׂיתם and do אתם אם them; though יהיה there were נדחכם of you cast out בקצה unto the uttermost part השׁמים of the heaven, משׁם them from thence, אקבצם will I gather והבואתים and will bring אל them unto המקום the place אשׁר that בחרתי I have chosen לשׁכן to set את שׁמי my name שׁם׃ there.}%
\verse{והם Now these עבדיך thy servants ועמך and thy people, אשׁר whom פדית thou hast redeemed בכחך power, הגדול by thy great ובידך hand. החזקה׃ and by thy strong}%
\verse{אנא I beseech אדני O Lord, תהי be נא thee, let now אזנך thine ear קשׁבת attentive אל to תפלת the prayer עבדך of thy servant, ואל and to תפלת the prayer עבדיך of thy servants, החפצים who desire ליראה to fear את שׁמך thy name: והצליחה and prosper, נא I pray לעבדך thee, thy servant היום this day, ותנהו and grant לרחמים him mercy לפני in the sight האישׁ man. הזה of this ואני For I הייתי was משׁקה cupbearer. למלך׃ the king's}%
\end{biblechapter}%
\begin{biblechapter}% Nehemiah 2
\verseWithHeading{Nehemiah Sent}{ויהי And it came to pass בחדשׁ in the month ניסן Nisan, שׁנת year עשׂרים in the twentieth לארתחשׁסתא of Artaxerxes המלך the king, יין wine לפניו before ואשׂא him: and I took up את היין the wine, ואתנה and gave למלך unto the king. ולא Now I had not הייתי been רע sad לפניו׃ in his presence.}%
\verse{ויאמר said לי המלך Wherefore the king מדוע unto me, Why פניך thy countenance רעים sad, ואתה seeing thou אינך not חולה sick? אין nothing זה this כי but אם but רע sorrow לב of heart. ואירא afraid, הרבה sore מאד׃ Then I was very}%
\verse{ואמר And said למלך unto the king, המלך Let the king לעולם forever: יחיה live מדוע why לא should not ירעו be sad, פני my countenance אשׁר when העיר the city, בית the place קברות sepulchers, אבתי of my fathers' חרבה ושׁעריה and the gates אכלו thereof are consumed באשׁ׃ with fire?}%
\verse{ויאמר said לי המלך Then the king על unto me, For מה what זה what אתה dost thou מבקשׁ make request? ואתפלל So I prayed אל to אלהי the God השׁמים׃ of heaven.}%
\verse{ואמר And I said למלך unto the king, אם If על it please המלך the king, טוב it please ואם and if ייטב have found favor עבדך thy servant לפניך in thy sight, אשׁר that תשׁלחני thou wouldest send אל me unto יהודה Judah, אל unto עיר the city קברות sepulchers, אבתי of my fathers' ואבננה׃ that I may build}%
\verse{ויאמר said לי המלך And the king והשׁגל unto me, (the queen יושׁבת also sitting אצלו by עד him,) For מתי how long יהיה be? מהלכך shall thy journey ומתי and when תשׁוב wilt thou return? וייטב לפנימלך the king וישׁלחני to send ואתנה me; and I set לו זמן׃ him a time.}%
\verse{ואומר Moreover I said למלך unto the king, אם If על it please המלך the king, טוב it please אגרות let letters יתנו be given לי על me to פחוות the governors עבר beyond הנהר the river, אשׁר that יעבירוני they may convey me over עד till אשׁר till אבוא I come אל into יהודה׃ Judah;}%
\verse{ואגרת And a letter אל unto אסף Asaph שׁמר the keeper הפרדס forest, אשׁר that למלך of the king's אשׁר which יתן he may give לי עצים me timber לקרות to make beams את שׁערי for the gates הבירה of the palace אשׁר that לבית to the house, ולחומת and for the wall העיר of the city, ולבית and for the house אשׁר אבוא I shall enter אליו into. ויתן granted לי המלך And the king כיד hand אלהי of my God הטובה me, according to the good עלי׃ upon}%
\verse{ואבוא Then I came אל to פחוות the governors עבר beyond הנהר the river, ואתנה and gave להם את אגרות letters. המלך them the king's וישׁלח had sent עמי with המלך Now the king שׂרי captains חיל of the army ופרשׁים׃ and horsemen}%
\verse{וישׁמע heard סנבלט When Sanballat החרני the Horonite, וטוביה and Tobiah העבד the servant, העמני the Ammonite, וירע it grieved להם רעה them exceedingly גדלה them exceedingly אשׁר that בא there was come אדם a man לבקשׁ to seek טובה the welfare לבני of the children ישׂראל׃ of Israel.}%
\verseWithHeading{Nehemiah Inspects the Walls and Decides to Restore Them}{ואבוא So I came אל to ירושׁלם Jerusalem, ואהי and was שׁם there ימים days. שׁלשׁה׃ three}%
\verse{ואקום And I arose לילה in the night, אני I ואנשׁים men מעט and some few עמי with ולא me; neither הגדתי told לאדם I man מה what אלהי my God נתן had put אל in לבי my heart לעשׂות to do לירושׁלם at Jerusalem: ובהמה beast אין neither עמי with כי me, save אם me, save הבהמה the beast אשׁר that אני I רכב׃ rode}%
\verse{ואצאה And I went out בשׁער by the gate הגיא of the valley, לילה by night ואל even before פני even before עין well, התנין the dragon ואל and to שׁער port, האשׁפת the dung ואהי and viewed שׂבר and viewed בחומת the walls ירושׁלם of Jerusalem, אשׁר which המפרוצים were broken down, ושׁעריה and the gates אכלו thereof were consumed באשׁ׃ with fire.}%
\verse{ואעבר Then I went on אל to שׁער the gate העין of the fountain, ואל and to ברכת pool: המלך the king's ואין but no מקום place לבהמה for the beast לעבר me to pass. תחתי׃ under}%
\verse{ואהי and viewed עלה בנחל by the brook, לילה in the night ואהי שׂבר and viewed בחומה the wall, ואשׁוב and turned back, ואבוא and entered בשׁער by the gate הגיא of the valley, ואשׁוב׃ and returned.}%
\verse{והסגנים And the rulers לא not ידעו knew אנה whither הלכתי I went, ומה or what אני I עשׂה did; וליהודים to the Jews, ולכהנים nor to the priests, ולחרים nor to the nobles, ולסגנים nor to the rulers, וליתר nor to the rest עשׂה that did המלאכה the work. עד had I as כן yet לא neither הגדתי׃ told}%
\verse{ואומר Then said אלהם I unto אתם them, Ye ראים see הרעה the distress אשׁר that אנחנו we בה אשׁר in, how ירושׁלם Jerusalem חרבה ושׁעריה and the gates נצתו thereof are burned באשׁ with fire: לכו come, ונבנה and let us build up את חומת the wall ירושׁלם of Jerusalem, ולא no נהיה that we be עוד more חרפה׃ a reproach.}%
\verse{ואגיד Then I told להם את יד them of the hand אלהי of my God אשׁר which היא which טובה was good עלי upon ואף me; as also דברי words המלך the king's אשׁר that אמר he had spoken לי ויאמרו unto me. And they said, נקום Let us rise up ובנינו and build. ויחזקו So they strengthened ידיהם their hands לטובה׃ for good}%
\verse{וישׁמע heard סנבלט But when Sanballat החרני the Horonite, וטביה and Tobiah העבד the servant, העמוני the Ammonite, וגשׁם and Geshem הערבי the Arabian, וילעגו they laughed us to scorn, לנו ויבזו and despised עלינו and despised ויאמרו us, and said, מה What הדבר thing הזה this אשׁר that אתם ye עשׂים do? העל against המלך the king? אתם will ye מרדים׃ rebel}%
\verse{ואשׁיב אותםבר ואומר I them, and said להם אלהי unto them, The God השׁמים of heaven, הוא he יצליח will prosper לנו ואנחנו us; therefore we עבדיו his servants נקום will arise ובנינו and build: ולכם אין but ye have no חלק portion, וצדקה nor right, וזכרון nor memorial, בירושׁלם׃ in Jerusalem.}%
\end{biblechapter}%
\begin{biblechapter}% Nehemiah 3
\verseWithHeading{Organization of the Work}{ויקם rose up אלישׁיב Then Eliashib הכהן priest הגדול the high ואחיו with his brethren הכהנים the priests, ויבנו and they built את שׁער gate; הצאן the sheep המה they קדשׁוהו sanctified ויעמידו it, and set up דלתתיו the doors ועד of it; even unto מגדל the tower המאה of Meah קדשׁוהו they sanctified עד it, unto מגדל the tower חננאל׃ of Hananeel.}%
\verse{ועל unto ידו And next בנו him built אנשׁי ירחו of Jericho. ועל to ידו And next בנה them built זכור Zaccur בן the son אמרי׃ of Imri.}%
\verse{ואת שׁער gate הדגים But the fish בנו build, בני did the sons הסנאה of Hassenaah המה who קרוהו laid the beams ויעמידו thereof, and set up דלתתיו the doors מנעוליו thereof, the locks ובריחיו׃ thereof, and the bars}%
\verse{ועל unto ידם And next החזיק them repaired מרמות Meremoth בן the son אוריה of Urijah, בן the son הקוץ of Koz. ועל unto ידם And next החזיק them repaired משׁלם Meshullam בן the son ברכיה of Berechiah, בן the son משׁיזבאל of Meshezabeel. ועל unto ידם And next החזיק them repaired צדוק Zadok בן the son בענא׃}%
\verse{ועל unto ידם And next החזיקו repaired; התקועים them the Tekoites ואדיריהם but their nobles לא not הביאו put צורם their necks בעבדת to the work אדניהם׃ of their Lord.}%
\verse{ואת שׁער gate הישׁנה Moreover the old החזיקו repaired יוידע Jehoiada בן the son פסח of Paseah, ומשׁלם and Meshullam בן the son בסודיה of Besodeiah; המה they קרוהו laid the beams ויעמידו thereof, and set up דלתתיו the doors ומנעליו thereof, and the locks ובריחיו׃ thereof, and the bars}%
\verse{ועל unto ידם And next החזיק them repaired מלטיה Melatiah הגבעני the Gibeonite, וידון and Jadon המרנתי the Meronothite, אנשׁי the men גבעון of Gibeon, והמצפה and of Mizpah, לכסא unto the throne פחת of the governor עבר on this side הנהר׃ the river.}%
\verse{על unto ידו Next החזיק him repaired עזיאל Uzziel בן the son חרהיה of Harhaiah, צורפים of the goldsmiths. ועל unto ידו Next החזיק him also repaired חנניה Hananiah בן the son הרקחים of the apothecaries, ויעזבו and they fortified ירושׁלם Jerusalem עד unto החומה wall. הרחבה׃ the broad}%
\verse{ועל unto ידם And next החזיק them repaired רפיה Rephaiah בן the son חור of Hur, שׂר the ruler חצי of the half פלך part ירושׁלם׃ of Jerusalem.}%
\verse{ועל unto ידם And next החזיק them repaired ידיה Jedaiah בן the son חרומף of Harumaph, ונגד even over against ביתו his house. ועל unto ידו And next החזיק him repaired חטושׁ Hattush בן the son חשׁבניה׃ of Hashabniah.}%
\verse{מדה piece, שׁנית the other החזיק repaired מלכיה Malchijah בן the son חרם of Harim, וחשׁוב and Hashub בן the son פחת מואב of Pahath-moab, ואת מגדל and the tower התנורים׃ of the furnaces.}%
\verse{ועל unto ידו And next החזיק him repaired שׁלום Shallum בן the son הלוחשׁ of Halohesh, שׂר the ruler חצי of the half פלך part ירושׁלם of Jerusalem, הוא he ובנותיו׃ and his daughters.}%
\verse{את שׁער gate הגיא The valley החזיק repaired חנון Hanun, וישׁבי and the inhabitants זנוח of Zanoah; המה they בנוהו built ויעמידו it, and set up דלתתיו the doors מנעליו thereof, the locks ובריחיו thereof, and the bars ואלף thereof, and a thousand אמה cubits בחומה on the wall עד unto שׁער gate. השׁפות׃ the dung}%
\verse{ואת שׁער gate האשׁפות But the dung החזיק repaired מלכיה Malchiah בן the son רכב of Rechab, שׂר the ruler פלך of part בית הכרם of Beth-haccerem; הוא he יבננו built ויעמיד it, and set up דלתתיו the doors מנעליו thereof, the locks ובריחיו׃ thereof, and the bars}%
\verse{ואת שׁער But the gate העין of the fountain החזיק repaired שׁלון Shallun בן the son כל חזה of Col-hozeh, שׂר the ruler פלך of part המצפה of Mizpah; הוא he יבננו built ויטללנו it, and covered ויעמידו it, and set up דלתתיו the doors מנעליו thereof, the locks ובריחיו thereof, and the bars ואת חומת thereof, and the wall ברכת of the pool השׁלח of Siloah לגן garden, המלך by the king's ועד and unto המעלות the stairs היורדות that go down מעיר from the city דויד׃ of David.}%
\verse{אחריו After החזיק him repaired נחמיה Nehemiah בן the son עזבוק of Azbuk, שׂר the ruler חצי of the half פלך part בית צור of Beth-zur, עד unto נגד over against קברי the sepulchers דויד of David, ועד and to הברכה the pool העשׂויה that was made, ועד and unto בית the house הגברים׃ of the mighty.}%
\verse{אחריו After החזיקו him repaired הלוים the Levites, רחום Rehum בן the son בני of Bani. על unto ידו Next החזיק him repaired חשׁביה Hashabiah, שׂר the ruler חצי of the half פלך part קעילה of Keilah, לפלכו׃ in his part.}%
\verse{אחריו After החזיקו him repaired אחיהם their brethren, בוי Bavai בן the son חנדד of Henadad, שׂר the ruler חצי of the half פלך part קעילה׃ of Keilah.}%
\verse{ויחזק him repaired על to ידו And next עזר Ezer בן the son ישׁוע of Jeshua, שׂר the ruler המצפה of Mizpah, מדה piece שׁנית another מנגד over against עלת the going up הנשׁק to the armory המקצע׃ at the turning}%
\verse{אחריו After החרה earnestly החזיק repaired ברוך him Baruch בן the son זבי of Zabbai מדה piece, שׁנית the other מן from המקצוע the turning עד unto פתח the door בית of the house אלישׁיב of Eliashib הכהן priest. הגדול׃ the high}%
\verse{אחריו After החזיק him repaired מרמות Meremoth בן the son אוריה of Urijah בן the son הקוץ of Koz מדה piece, שׁנית another מפתח from the door בית of the house אלישׁיב of Eliashib ועד even to תכלית the end בית of the house אלישׁיב׃ of Eliashib.}%
\verse{ואחריו And after החזיקו him repaired הכהנים the priests, אנשׁי the men הככר׃ of the plain.}%
\verse{אחריו After החזיק him repaired בנימן Benjamin וחשׁוב and Hashub נגד over against ביתם their house. אחריו After החזיק him repaired עזריה Azariah בן the son מעשׂיה of Maaseiah בן the son ענניה of Ananiah אצל by ביתו׃ his house.}%
\verse{אחריו After החזיק him repaired בנוי Binnui בן the son חנדד of Henadad מדה piece, שׁנית another מבית from the house עזריה of Azariah עד unto המקצוע the turning ועד even unto הפנה׃ the corner.}%
\verse{פלל Palal בן the son אוזי of Uzai, מנגד over against המקצוע the turning והמגדל and the tower היוצא which lieth out מבית house, המלך from the king's העליון high אשׁר that לחצר by the court המטרה of the prison. אחריו After פדיה him Pedaiah בן the son פרעשׁ׃ of Parosh.}%
\verse{והנתינים Moreover the Nethinims היו dwelt ישׁבים dwelt בעפל in Ophel, עד unto נגד over against שׁער gate המים the water למזרח toward the east, והמגדל and the tower היוצא׃ that lieth out.}%
\verse{אחריו After החזיקו repaired התקעים them the Tekoites מדה piece, שׁנית another מנגד over against המגדל tower הגדול the great היוצא that lieth out, ועד even unto חומת the wall העפל׃ of Ophel.}%
\verse{מעל שׁער gate הסוסים the horse החזיקו repaired הכהנים the priests, אישׁ every one לנגד over against ביתו׃ his house.}%
\verse{אחריו After החזיק them repaired צדוק Zadok בן the son אמר of Immer נגד over against ביתו his house. ואחריו After החזיק him repaired שׁמעיה also Shemaiah בן the son שׁכניה of Shechaniah, שׁמר the keeper שׁער gate. המזרח׃ of the east}%
\verse{אחרי After החזיק him repaired חנניה Hananiah בן the son שׁלמיה of Shelemiah, וחנון and Hanun בן son צלף of Zalaph, השׁשׁי the sixth מדה piece. שׁני another אחריו After החזיק him repaired משׁלם Meshullam בן the son ברכיה of Berechiah נגד over against נשׁכתו׃ his chamber.}%
\verse{אחרי After החזיק him repaired מלכיה Malchiah בן son הצרפי the goldsmith's עד unto בית the place הנתינים of the Nethinims, והרכלים and of the merchants, נגד over against שׁער the gate המפקד Miphkad, ועד and to עלית the going up הפנה׃ of the corner.}%
\verse{ובין And between עלית the going up הפנה of the corner לשׁער gate הצאן unto the sheep החזיקו repaired הצרפים the goldsmiths והרכלים׃ and the merchants.}%
\end{biblechapter}%
\begin{biblechapter}% Nehemiah 4
\verseWithHeading{Opposition Begins}{ויהי But it came to pass, כאשׁר that when שׁמע heard סנבלט Sanballat כי that אנחנו we בונים built את החומה the wall, ויחר he was wroth, לו ויכעס and took great indignation, הרבה and took great indignation, וילעג and mocked על and mocked היהודים׃ the Jews.}%
\verse{ויאמר And he spoke לפני before אחיו his brethren וחיל שׁמרון of Samaria, ויאמר and said, מה What היהודים Jews? האמללים these feeble עשׂים do היעזבו will they fortify להם which היזבחו themselves? will they sacrifice? היכלו will they make an end ביום in a day? היחיו will they revive את האבנים the stones מערמות out of the heaps העפר of the rubbish והמה שׂרופות׃ are burned?}%
\verse{וטוביה Now Tobiah העמני the Ammonite אצלו by ויאמר him, and he said, גם Even אשׁר that which הם they בונים build, אם if יעלה go up, שׁועל a fox ופרץ he shall even break down חומת wall. אבניהם׃ their stone}%
\verse{שׁמע Hear, אלהינו O our God; כי for היינו we are בוזה despised: והשׁב and turn חרפתם their reproach אל upon ראשׁם their own head, ותנם and give לבזה them for a prey בארץ in the land שׁביה׃ of captivity:}%
\verse{ואל not תכס עלונם their iniquity, וחטאתם their sin מלפניך from before אל and let not תמחה be blotted out כי thee: for הכעיסו they have provoked to anger לנגד before הבונים׃ the builders.}%
\verse{ונבנה So built את החומה we the wall; ותקשׁר was joined together כל and all החומה the wall עד unto חציה the half ויהי had לב a mind לעם thereof: for the people לעשׂות׃ to work.}%
\verseWithHeading{Opposition Stopped}{ויהי But it came to pass, כאשׁר when שׁמע heard סנבלט Sanballat, וטוביה and Tobiah, והערבים and the Arabians, והעמנים and the Ammonites, והאשׁדודים and the Ashdodites, כי that עלתה were made up, ארוכה were made up, לחמות the walls ירושׁלם of Jerusalem כי that החלו began הפרצים the breaches להסתם to be stopped, ויחר wroth, להם מאד׃ then they were very}%
\verse{ויקשׁרו And conspired כלם all יחדו of them together לבוא to come להלחם to fight בירושׁלם against Jerusalem, ולעשׂות and to hinder לו תועה׃ and to hinder}%
\verse{ונתפלל Nevertheless we made our prayer אל unto אלהינו our God, ונעמיד and set משׁמר a watch עליהם against יומם them day ולילה and night, מפניהם׃ because of}%
\verse{ויאמר said, יהודה And Judah כשׁל is decayed, כח The strength הסבל of the bearers of burdens והעפר rubbish; הרבה and much ואנחנו so that we לא are not נוכל able לבנות to build בחומה׃ the wall.}%
\verse{ויאמרו said, צרינו And our adversaries לא They shall not ידעו know, ולא neither יראו see, עד till אשׁר till נבוא we come אל in תוכם the midst among והרגנום them, and slay והשׁבתנו to cease. את המלאכה׃ them, and cause the work}%
\verse{ויהי And it came to pass, כאשׁר that when באו them came, היהודים the Jews הישׁבים which dwelt אצלם by ויאמרו they said לנו עשׂר unto us ten פעמים times, מכל המקמות places אשׁר whence תשׁובו ye shall return עלינו׃ unto}%
\verse{ואעמיד Therefore set מתחתיות למקום places מאחרי behind לחומה the wall, בצחחיים ואעמיד I even set את העם the people למשׁפחות after their families עם with חרבתיהם their swords, רמחיהם their spears, וקשׁתתיהם׃ and their bows.}%
\verse{וארא And I looked, ואקום and rose up, ואמר and said אל unto החרים the nobles, ואל and to הסגנים the rulers, ואל and to יתר the rest העם of the people, אל Be not תיראו ye afraid מפניהם of את אדני the Lord, הגדול great והנורא and terrible, זכרו them: remember והלחמו and fight על for אחיכם your brethren, בניכם your sons, ובנתיכם and your daughters, נשׁיכם your wives, ובתיכם׃ and your houses.}%
\verse{ויהי And it came to pass, כאשׁר when שׁמעו heard אויבינו our enemies כי that נודע it was known לנו ויפר האלהים unto us, and God את עצתםנשׁוב that we returned כלנו all אל of us to החומה the wall, אישׁ every one אל unto מלאכתו׃ his work.}%
\verse{ויהי And it came to pass מן from היום time ההוא that חצי forth, the half נערי of my servants עשׂים wrought במלאכה in the work, וחצים and the other half מחזיקים of them held והרמחים both the spears, המגנים the shields, והקשׁתות and the bows, והשׁרינים and the habergeons; והשׂרים and the rulers אחרי behind כל all בית the house יהודה׃ of Judah.}%
\verse{הבונים They which built בחומה on the wall, והנשׂאים and they that bore בסבל burdens, עמשׂים with those that laded, באחת with one ידו of his hands עשׂה wrought במלאכה in the work, ואחת and with the other מחזקת held השׁלח׃ a weapon.}%
\verse{והבונים For the builders, אישׁ every one חרבו had his sword אסורים girded על by מתניו his side, ובונים and built. והתוקע And he that sounded בשׁופר the trumpet אצלי׃ by}%
\verse{ואמר And I said אל unto החרים the nobles, ואל and to הסגנים the rulers, ואל and to יתר the rest העם of the people, המלאכה The work הרבה great ורחבה and large, ואנחנו and we נפרדים are separated על upon החומה the wall, רחוקים far אישׁ one מאחיו׃ from another.}%
\verse{במקום place אשׁר In what תשׁמעו ye hear את קול the sound השׁופר of the trumpet, שׁמה ye thither תקבצו resort אלינו unto אלהינו us: our God ילחם׃ shall fight}%
\verse{ואנחנו So we עשׂים labored במלאכה in the work: וחצים and half מחזיקים of them held ברמחים the spears מעלות from the rising השׁחר of the morning עד till צאת appeared. הכוכבים׃ the stars}%
\verse{גם Likewise בעת time ההיא at the same אמרתי said לעם I unto the people, אישׁ Let every one ונערו with his servant ילינו lodge בתוך within ירושׁלם Jerusalem, והיו they may be לנו הלילה that in the night משׁמר a guard והיום on the day. מלאכה׃ to us, and labor}%
\verse{ואין So neither אני I, ואחי nor my brethren, ונערי nor my servants, ואנשׁי המשׁמר of the guard אשׁר which אחרי followed אין me, none אנחנו of us פשׁטים put off בגדינו our clothes, אישׁ nor the men שׁלחו המים׃ for washing.}%
\end{biblechapter}%
\begin{biblechapter}% Nehemiah 5
\verseWithHeading{Nehemiah Deals with Strife}{ותהי And there was צעקת cry העם of the people ונשׁיהם and of their wives גדולה a great אל against אחיהם their brethren היהודים׃ the Jews.}%
\verse{וישׁ For there were אשׁר that אמרים said, בנינו our sons, ובנתינו and our daughters, אנחנו We, רבים many: ונקחה therefore we take up דגן corn ונאכלה that we may eat, ונחיה׃ and live.}%
\verse{וישׁ also there were אשׁר that אמרים said, שׂדתינו our lands, וכרמינו vineyards, ובתינו and houses, אנחנו We ערבים have mortgaged ונקחה that we might buy דגן corn, ברעב׃ because of the dearth.}%
\verse{וישׁ There were אשׁר also that אמרים said, לוינו We have borrowed כסף money למדת tribute, המלך for the king's שׂדתינו our lands וכרמינו׃ and vineyards.}%
\verse{ועתה Yet now כבשׂר our flesh אחינו of our brethren, בשׂרנו as the flesh כבניהם our children בנינו as their children: והנה and, lo, אנחנו we כבשׁים bring into bondage את בנינו our sons ואת בנתינו and our daughters לעבדים to be servants, וישׁ are מבנתינו and of our daughters נכבשׁות brought unto bondage ואין neither לאל in our power ידנו in our power ושׂדתינו men have our lands וכרמינו and vineyards. לאחרים׃ for other}%
\verse{ויחר angry לי מאד And I was very כאשׁר when שׁמעתי I heard את זעקתם their cry ואת הדברים words. האלה׃ and these}%
\verse{וימלך Then I consulted לבי myself, עלי with ואריבה and I rebuked את החרים the nobles, ואת הסגנים and the rulers, ואמרה and said להם משׁא exact אישׁ every one באחיו of his brother. אתם unto them, Ye נשׁאים ואתן And I set עליהם against קהלה assembly גדולה׃ a great}%
\verse{ואמרה And I said להם אנחנו unto them, We קנינו have redeemed את אחינו our brethren היהודים the Jews, הנמכרים which were sold לגוים unto the heathen; כדי after our ability בנו וגם even אתם and will ye תמכרו sell את אחיכם your brethren? ונמכרו or shall they be sold לנו ויחרישׁו unto us? Then held they their peace, ולא nothing מצאו and found דבר׃}%
\verse{ויאמר Also I said, לא It not טוב good הדבר good אשׁר that אתם ye עשׂים do: הלוא ought ye not ביראת in the fear אלהינו of our God תלכו to walk מחרפת because of the reproach הגוים of the heathen אויבינו׃ our enemies?}%
\verse{וגם likewise, אני I אחי my brethren, ונערי and my servants, נשׁים might exact בהם כסף of them money ודגן and corn: נעזבה you, let us leave off נא I pray את המשׁא usury. הזה׃ this}%
\verse{השׁיבו Restore, נא I pray להם כהיום you, to them, even this day, שׂדתיהם their lands, כרמיהם their vineyards, זיתיהם their oliveyards, ובתיהם and their houses, ומאת also the hundredth הכסף of the money, והדגן and of the corn, התירושׁ the wine, והיצהר and the oil, אשׁר that אתם ye נשׁים׃ exact}%
\verse{ויאמרו Then said נשׁיב they, We will restore ומהם לא nothing נבקשׁ and will require כן them; so נעשׂה will we do כאשׁר as אתה thou אומר sayest. ואקרא Then I called את הכהנים the priests, ואשׁביעם and took an oath לעשׂות of them, that they should do כדבר promise. הזה׃ according to this}%
\verse{גם Also חצני my lap, נערתי I shook ואמרה and said, ככה So ינער shake out האלהים God את כל every האישׁ man אשׁר that לא not יקים performeth את הדבר promise, הזה this מביתו from his house, ומיגיעו and from his labor, וככה even thus יהיה be נעור he shaken out, ורק and emptied. ויאמרו said, כל And all הקהל the congregation אמן Amen, ויהללו and praised את יהוהיעשׂ did העם And the people כדבר promise. הזה׃ according to this}%
\verseWithHeading{Nehemiah Denies His Allotment}{גם Moreover מיום from the time אשׁר that צוה I was appointed אתי להיות to be פחם their governor בארץ in the land יהודה of Judah, משׁנת year עשׂרים from the twentieth ועד even unto שׁנת year שׁלשׁים and thirtieth ושׁתים the two לארתחשׁסתא of Artaxerxes המלך the king, שׁנים years, שׁתים twelve עשׂרה , twelve אני I ואחי and my brethren לחם the bread הפחה of the governor. לא have not אכלתי׃ eaten}%
\verse{והפחות governors הראשׁנים But the former אשׁר that לפני before הכבידו me were chargeable על unto העם the people, ויקחו and had taken מהם בלחם them bread ויין and wine, אחר beside כסף of silver; שׁקלים shekels ארבעים forty גם yea, נעריהם even their servants שׁלטו bore rule על over העם the people: ואני I, לא not עשׂיתי did כן but so מפני because יראת of the fear אלהים׃ of God.}%
\verse{וגם Yea, also במלאכת in the work החומה wall, הזאת of this החזקתי I continued ושׂדה we any land: לא neither קנינו bought וכל and all נערי my servants קבוצים gathered שׁם thither על unto המלאכה׃ the work.}%
\verse{והיהודים of the Jews והסגנים and rulers, מאה a hundred וחמשׁים and fifty אישׁ and fifty והבאים beside those that came אלינו unto מן us from among הגוים the heathen אשׁר that סביבתינו about על Moreover at שׁלחני׃ my table}%
\verse{ואשׁר Now which היה was נעשׂה prepared ליום daily אחד daily שׁור ox אחד one צאן sheep; שׁשׁ six בררות choice וצפרים also fowls נעשׂו were prepared לי ובין for me, and once עשׂרת in ten ימים days בכל of all יין sorts of wine: להרבה store ועם yet for all זה this לחם I the bread הפחה of the governor, לא not בקשׁתי required כי because כבדה was heavy העבדה the bondage על upon העם people. הזה׃ this}%
\verse{זכרה Think לי אלהי upon me, my God, לטובה for good, כל to all אשׁר that עשׂיתי I have done על for העם people. הזה׃ this}%
\end{biblechapter}%
\begin{biblechapter}% Nehemiah 6
\verseWithHeading{Enemies Foiled}{ויהי Now it came to pass, כאשׁר when נשׁמע heard לסנבלט Sanballat, וטוביה and Tobiah, ולגשׁם and Geshem הערבי the Arabian, וליתר and the rest איבינו of our enemies, כי that בניתי I had built את החומה the wall, ולא and there was no נותר left בה פרץ breach גם therein; (though עד at העת time ההיא that דלתות the doors לא I had not העמדתי set up בשׁערים׃ upon the gates;)}%
\verse{וישׁלח sent סנבלט That Sanballat וגשׁם and Geshem אלי unto לאמר me, saying, לכה Come, ונועדה let us meet יחדו together בכפירים in the villages בבקעת in the plain אונו of Ono. והמה But they חשׁבים thought לעשׂות to do לי רעה׃ me mischief.}%
\verse{ואשׁלחה And I sent עליהם unto מלאכים messengers לאמר them, saying, מלאכה work, גדולה a great אני I עשׂה doing ולא אוכלרדת come down: למה why תשׁבת cease, המלאכה should the work כאשׁר whilst ארפה I leave וירדתי it, and come down אליכם׃ to}%
\verse{וישׁלחו Yet they sent אלי unto כדבר sort; הזה after this ארבע me four פעמים times ואשׁיב and I answered אותם כדבר manner. הזה׃ them after the same}%
\verse{וישׁלח Then sent אלי unto סנבלט Sanballat כדבר manner הזה me in like פעם time חמישׁית the fifth את נערו his servant ואגרת letter פתוחה with an open בידו׃ in his hand;}%
\verse{כתוב Wherein written, בה בגוים among the heathen, נשׁמע It is reported וגשׁמו and Gashmu אמר saith אתה thou והיהודים and the Jews חשׁבים think למרוד to rebel: על for כן which cause אתה thou בונה buildest החומה the wall, ואתה that thou הוה mayest be להם למלך their king, כדברים words. האלה׃ according to these}%
\verse{וגם And thou hast also נביאים prophets העמדת appointed לקרא to preach עליך of בירושׁלם thee at Jerusalem, לאמר saying, מלך a king ביהודה in Judah: ועתה and now ישׁמע shall it be reported למלך to the king כדברים words. האלה according to these ועתה now לכה Come ונועצה therefore, and let us take counsel יחדו׃ together.}%
\verse{ואשׁלחה Then I sent אליו unto לאמר him, saying, לא no נהיה There are כדברים things האלה such אשׁר done as אתה thou אומר sayest, כי but מלבך them out of thine own heart. אתה thou בודאם׃ feignest}%
\verse{כי For כלם they all מיראים אותנואמר saying, ירפו shall be weakened ידיהם Their hands מן from המלאכה the work, ולא that it be not תעשׂה done. ועתה Now חזק therefore, strengthen את ידי׃ my hands.}%
\verse{ואני Afterward I באתי came בית unto the house שׁמעיה of Shemaiah בן the son דליה of Delaiah בן the son מהיטבאל of Mehetabeel, והוא who עצור shut up; ויאמר and he said, נועד Let us meet together אל in בית the house האלהים of God, אל within תוך within ההיכל the temple, ונסגרה and let us shut דלתות the doors ההיכל of the temple: כי for באים they will come להרגך to slay ולילה thee; yea, in the night באים will they come להרגך׃ to slay}%
\verse{ואמרה And I said, האישׁ כמוני as I יברח flee? ומי and who כמוני אשׁר that, יבוא would go אל into ההיכל the temple וחי to save his life? לא I will not אבוא׃ go in.}%
\verse{ואכירה I perceived והנה And, lo, לא had not אלהים that God שׁלחו sent כי him; but הנבואה this prophecy דבר that he pronounced עלי against וטוביה me: for Tobiah וסנבלט and Sanballat שׂכרו׃ had hired}%
\verse{למען Therefore שׂכור hired, הוא he למען that אירא I should be afraid, ואעשׂה and do כן so, וחטאתי and sin, והיה and they might have להם לשׁם report, רע for an evil למען that יחרפוני׃ they might reproach}%
\verse{זכרה think אלהי My God, לטוביה thou upon Tobiah ולסנבלט and Sanballat כמעשׂיו their works, אלה according to these וגם and לנועדיה Noadiah, הנביאה on the prophetess וליתר and the rest הנביאים of the prophets, אשׁר that היו would have מיראים put me in fear. אותי׃}%
\verseWithHeading{The Wall is Completed}{ותשׁלם was finished החומה So the wall בעשׂרים in the twenty וחמשׁה and fifth לאלול of Elul, לחמשׁים in fifty ושׁנים and two יום׃ days.}%
\verse{ויהי And it came to pass, כאשׁר that when שׁמעו heard כל all אויבינו our enemies ויראו us saw כל and all הגוים the heathen אשׁר that סביבתינו about ויפלו cast down מאד they were much בעיניהם in their own eyes: וידעו for they perceived כי that מאת of אלהינו our God. נעשׂתה was wrought המלאכה work הזאת׃ this}%
\verse{גם Moreover בימים days ההם in those מרבים many חרי the nobles יהודה of Judah אגרתיהם letters הולכות sent על unto טוביה Tobiah, ואשׁר לטוביה and of Tobiah באות came אליהם׃ unto}%
\verse{כי For רבים many ביהודה in Judah בעלי sworn שׁבועה sworn לו כי unto him, because חתן the son in law הוא he לשׁכניה of Shechaniah בן the son ארח of Arah; ויהוחנן Johanan בנו and his son לקח had taken את בת the daughter משׁלם of Meshullam בן the son ברכיה׃ of Berechiah.}%
\verse{גם Also טובתיו his good deeds היו they reported אמרים they reported לפני before ודברי my words היו me, and uttered מוציאים me, and uttered לו אגרות letters שׁלח sent טוביה to him. Tobiah ליראני׃ to put me in fear.}%
\end{biblechapter}%
\begin{biblechapter}% Nehemiah 7
\verse{ויהי Now it came to pass, כאשׁר when נבנתה was built, החומה the wall ואעמיד and I had set up הדלתות the doors, ויפקדו were appointed, השׁוערים and the porters והמשׁררים and the singers והלוים׃ and the Levites}%
\verse{ואצוה charge את חנני Hanani, אחי That I gave my brother ואת חנניה and Hananiah שׂר the ruler הבירה of the palace, על over ירושׁלם Jerusalem: כי for הוא he כאישׁ man, אמת a faithful וירא and feared את האלהים God מרבים׃ above many.}%
\verse{ויאמר And I said להם לא unto them, Let not יפתחו be opened שׁערי the gates ירושׁלם of Jerusalem עד until חם השׁמשׁ the sun ועד and while הם they עמדים stand יגיפו by, let them shut הדלתות the doors, ואחזו and bar והעמיד and appoint משׁמרות watches ישׁבי of the inhabitants ירושׁלם of Jerusalem, אישׁ every one במשׁמרו in his watch, ואישׁ and every one נגד over against ביתו׃ his house.}%
\verse{והעיר Now the city רחבת large ידים large וגדולה and great: והעם but the people מעט few בתוכה therein, ואין not בתים and the houses בנוים׃ built.}%
\verseWithHeading{Lists of the Exiles Who Returned}{ויתן put אלהי And my God אל into לבי mine heart ואקבצה to gather together את החרים the nobles, ואת הסגנים and the rulers, ואת העם and the people, להתיחשׂ that they might be reckoned by genealogy. ואמצא And I found ספר a register היחשׂ of the genealogy העולים of them which came up בראשׁונה at the first, ואמצא and found כתוב׃ written}%
\verse{אלה These בני the children המדינה of the province, העלים that went up משׁבי out of the captivity, הגולה of those that had been carried awa, אשׁר whom הגלה had carried away, נבוכדנצר Nebuchadnezzar מלך the king בבל of Babylon וישׁובו and came again לירושׁלם to Jerusalem וליהודה and to Judah, אישׁ every one לעירו׃ unto his city;}%
\verse{הבאים Who came עם with זרבבל Zerubbabel, ישׁוע Jeshua, נחמיה Nehemiah, עזריה Azariah, רעמיה Raamiah, נחמני Nahamani, מרדכי Mordecai, בלשׁן Bilshan, מספרת Mispereth, בגוי Bigvai, נחום Nehum, בענה Baanah. מספר The number, אנשׁי of the men עם of the people ישׂראל׃ of Israel}%
\verse{בני The children פרעשׁ of Parosh, אלפים two thousand מאה a hundred ושׁבעים seventy ושׁנים׃ and two.}%
\verse{בני The children שׁפטיה of Shephatiah, שׁלשׁ three מאות hundred שׁבעים seventy ושׁנים׃ and two.}%
\verse{בני The children ארח of Arah, שׁשׁ six מאות hundred חמשׁים fifty ושׁנים׃ and two.}%
\verse{בני The children פחת מואב of Pahath-moab, לבני of the children ישׁוע of Jeshua ויואב and Joab, אלפים two thousand ושׁמנה and eight מאות hundred שׁמנה eighteen. עשׂר׃ eighteen.}%
\verse{בני The children עילם of Elam, אלף a thousand מאתים two hundred חמשׁים fifty וארבעה׃ and four.}%
\verse{בני The children זתוא of Zattu, שׁמנה eight מאות hundred ארבעים forty וחמשׁה׃ and five.}%
\verse{בני The children זכי of Zaccai, שׁבע seven מאות hundred ושׁשׁים׃ and threescore.}%
\verse{בני The children בנוי of Binnui, שׁשׁ six מאות hundred ארבעים forty ושׁמנה׃ and eight.}%
\verse{בני The children בבי of Bebai, שׁשׁ six מאות hundred עשׂרים twenty ושׁמנה׃ and eight.}%
\verse{בני The children עזגד of Azgad, אלפים two thousand שׁלשׁ three מאות hundred עשׂרים twenty ושׁנים׃ and two.}%
\verse{בני The children אדניקם of Adonikam, שׁשׁ six מאות hundred שׁשׁים threescore ושׁבעה׃ and seven.}%
\verse{בני The children בגוי of Bigvai, אלפים two thousand שׁשׁים threescore ושׁבעה׃ and seven.}%
\verse{בני The children עדין of Adin, שׁשׁ six מאות hundred חמשׁים fifty וחמשׁה׃ and five.}%
\verse{בני The children אטר of Ater לחזקיה of Hezekiah, תשׁעים ninety ושׁמנה׃ and eight.}%
\verse{בני The children חשׁם of Hashum, שׁלשׁ three מאות hundred עשׂרים twenty ושׁמנה׃ and eight.}%
\verse{בני The children בצי of Bezai, שׁלשׁ three מאות hundred עשׂרים twenty וארבעה׃ and four.}%
\verse{בני The children חריף of Hariph, מאה a hundred שׁנים and twelve. עשׂר׃ and twelve.}%
\verse{בני The children גבעון of Gibeon, תשׁעים ninety וחמשׁה׃ and five.}%
\verse{אנשׁי בית לחם of Bethlehem ונטפה and Netophah, מאה a hundred שׁמנים fourscore ושׁמנה׃ and eight.}%
\verse{אנשׁי The men ענתות of Anathoth, מאה a hundred עשׂרים twenty ושׁמנה׃ and eight.}%
\verse{אנשׁי The men בית עזמות of Beth-azmaveth, ארבעים forty ושׁנים׃ and two.}%
\verse{אנשׁי The men קרית יערים of Kirjath-jearim, כפירה Chephirah, ובארות and Beeroth, שׁבע seven מאות hundred ארבעים forty ושׁלשׁה׃ and three.}%
\verse{אנשׁי The men הרמה of Ramah וגבע and Geba, שׁשׁ six מאות hundred עשׂרים twenty ואחד׃ and one.}%
\verse{אנשׁי The men מכמס of Michmas, מאה a hundred ועשׂרים and twenty ושׁנים׃ and two.}%
\verse{אנשׁי The men בית אל of Bethel והעי and Ai, מאה a hundred עשׂרים twenty ושׁלשׁה׃ and three.}%
\verse{אנשׁי The men נבו Nebo, אחר of the other חמשׁים fifty ושׁנים׃ and two.}%
\verse{בני The children עילם Elam, אחר of the other אלף a thousand מאתים two hundred חמשׁים fifty וארבעה׃ and four.}%
\verse{בני The children חרם of Harim, שׁלשׁ three מאות hundred ועשׂרים׃ and twenty.}%
\verse{בני The children ירחו of Jericho, שׁלשׁ three מאות hundred ארבעים forty וחמשׁה׃ and five.}%
\verse{בני The children לד of Lod, חדיד Hadid, ואונו and Ono, שׁבע seven מאות hundred ועשׂרים twenty ואחד׃ and one.}%
\verse{בני The children סנאה of Senaah, שׁלשׁת three אלפים thousand תשׁע nine מאות hundred ושׁלשׁים׃ and thirty.}%
\verse{הכהנים The priests: בני the children ידעיה of Jedaiah, לבית of the house ישׁוע of Jeshua, תשׁע nine מאות hundred שׁבעים seventy ושׁלשׁה׃ and three.}%
\verse{בני The children אמר of Immer, אלף a thousand חמשׁים fifty ושׁנים׃ and two.}%
\verse{בני The children פשׁחור of Pashur, אלף a thousand מאתים two hundred ארבעים forty ושׁבעה׃ and seven.}%
\verse{בני The children חרם of Harim, אלף a thousand שׁבעה and seventeen. עשׂר׃ and seventeen.}%
\verse{הלוים The Levites: בני the children ישׁוע of Jeshua, לקדמיאל of Kadmiel, לבני of the children להודוה of Hodevah, שׁבעים seventy וארבעה׃ and four.}%
\verse{המשׁררים The singers: בני the children אסף of Asaph, מאה a hundred ארבעים forty ושׁמנה׃ and eight.}%
\verse{השׁערים The porters: בני the children שׁלום of Shallum, בני the children אטר of Ater, בני the children טלמן of Talmon, בני the children עקוב of Akkub, בני the children חטיטא of Hatita, בני the children שׁבי of Shobai, מאה a hundred שׁלשׁים thirty ושׁמנה׃ and eight.}%
\verse{הנתינים The Nethinims: בני the children צחא of Ziha, בני the children חשׂפא of Hashupha, בני the children טבעות׃ of Tabbaoth,}%
\verse{בני The children קירס of Keros, בני the children סיעא of Sia, בני the children פדון׃ of Padon,}%
\verse{בני The children לבנה of Lebana, בני the children חגבה of Hagaba, בני the children שׁלמי׃ of Shalmai,}%
\verse{בני The children חנן of Hanan, בני the children גדל of Giddel, בני the children גחר׃ of Gahar,}%
\verse{בני The children ראיה of Reaiah, בני the children רצין of Rezin, בני the children נקודא׃ of Nekoda,}%
\verse{בני The children גזם of Gazzam, בני the children עזא of Uzza, בני the children פסח׃ of Phaseah,}%
\verse{בני The children בסי of Besai, בני the children מעונים of Meunim, בני the children נפושׁסים׃}%
\verse{בני The children בקבוק of Bakbuk, בני the children חקופא of Hakupha, בני the children חרחור׃ of Harhur,}%
\verse{בני The children בצלית of Bazlith, בני the children מחידא of Mehida, בני the children חרשׁא׃ of Harsha,}%
\verse{בני The children ברקוס of Barkos, בני the children סיסרא of Sisera, בני the children תמח׃ of Tamah,}%
\verse{בני The children נציח of Neziah, בני the children חטיפא׃ of Hatipha.}%
\verse{בני The children עבדי servants: שׁלמה of Solomon's בני the children סוטי of Sotai, בני the children סופרת of Sophereth, בני the children פרידא׃ of Perida,}%
\verse{בני The children יעלא of Jaala, בני the children דרקון of Darkon, בני the children גדל׃ of Giddel,}%
\verse{בני The children שׁפטיה of Shephatiah, בני the children חטיל of Hattil, בני the children פכרת הצביים of Pochereth בני the children אמון׃ of Amon.}%
\verse{כל All הנתינים the Nethinims, ובני and the children עבדי servants, שׁלמה of Solomon's שׁלשׁ three מאות hundred תשׁעים ninety ושׁנים׃ and two.}%
\verse{ואלה And these העולים they which went up מתל מלח תל חרשׁא Tel-haresha, כרוב Cherub, אדון Addon, ואמר and Immer: ולא not יכלו but they could להגיד show בית house, אבותם their father's וזרעם nor their seed, אם whether מישׂראל הם׃ they}%
\verse{בני The children דליה of Delaiah, בני the children טוביה of Tobiah, בני the children נקודא of Nekoda, שׁשׁ six מאות hundred וארבעים forty ושׁנים׃ and two.}%
\verse{ומן And of הכהנים the priests: בני the children חביה of Habaiah, בני the children הקוץ of Koz, בני the children ברזלי of Barzillai, אשׁר which לקח took מבנות of the daughters ברזלי of Barzillai הגלעדי the Gileadite אשׁה to wife, ויקרא and was called על after שׁמם׃ their name.}%
\verse{אלה These בקשׁו sought כתבם their register המתיחשׂים those that were reckoned by genealogy, ולא but it was not נמצא found: ויגאלו therefore were they, as polluted, מן put from הכהנה׃ the priesthood.}%
\verse{ויאמר said התרשׁתא And the Tirshatha להם אשׁר unto them, that לא they should not יאכלו eat מקדשׁ הקדשׁיםד till עמד there stood הכהן a priest לאורים with Urim ותומים׃ and Thummim.}%
\verse{כל The whole הקהל congregation כאחד together ארבע forty רבוא forty אלפים and two thousand שׁלשׁ three מאות hundred ושׁשׁים׃ and threescore,}%
\verse{מלבד עבדיהם their menservants ואמהתיהם and their maidservants, אלה of whom שׁבעת seven אלפים thousand שׁלשׁ three מאות hundred שׁלשׁים thirty ושׁבעה and seven: ולהם משׁררים singing ומשׁררות men and singing מאתים and they had two hundred וארבעים forty וחמשׁה׃ and five}%
\verse{סוסיהם Their horses, שׁבע seven מאות hundred שׁלשׁים thirty ושׁשׁה and six: פרדיהם their mules, מאתים two hundred ארבעים forty וחמשׁה׃ and five:}%
\verse{גמלים camels, ארבע four מאות hundred שׁלשׁים thirty וחמשׁה and five: חמרים asses. שׁשׁת six אלפים thousand שׁבע seven מאות hundred ועשׂרים׃ and twenty}%
\verse{ומקצת ראשׁי of the chief האבות of the fathers נתנו gave למלאכה unto the work. התרשׁתא The Tirshatha נתן gave לאוצר to the treasure זהב of gold, דרכמנים drams אלף a thousand מזרקות basins, חמשׁים fifty כתנות garments. כהנים priests' שׁלשׁים and thirty וחמשׁ five מאות׃ hundred}%
\verse{ומראשׁי האבות of the fathers נתנו gave לאוצר to the treasure המלאכה of the work זהב of gold, דרכמונים drams שׁתי twenty thousand רבות twenty thousand וכסף of silver. מנים pound אלפים and two thousand ומאתים׃ and two hundred}%
\verse{ואשׁר And which נתנו gave שׁארית the rest העם of the people זהב of gold, דרכמונים drams שׁתי twenty רבוא thousand וכסף of silver, מנים pound אלפים and two thousand וכתנת garments. כהנים priests' שׁשׁים and threescore ושׁבעה׃ and seven}%
\verse{וישׁבו dwelt הכהנים So the priests, והלוים and the Levites, והשׁוערים and the porters, והמשׁררים and the singers, ומן and of העם the people, והנתינים and the Nethinims, וכל and all ישׂראל Israel, בעריהם in their cities; ויגע came, החדשׁ month השׁביעי and when the seventh ובני the children ישׂראל of Israel בעריהם׃ in their cities.}%
\end{biblechapter}%
\begin{biblechapter}% Nehemiah 8
\verseWithHeading{Ezra Reads the Law to the People}{ויאספו gathered themselves together כל And all העם the people כאישׁ man אחד as one אל into הרחוב the street אשׁר that לפני before שׁער gate; המים the water ויאמרו and they spoke לעזרא unto Ezra הספר the scribe להביא to bring את ספר the book תורת of the law משׁה of Moses, אשׁר which צוה had commanded יהוה the LORD את ישׂראל׃ to Israel.}%
\verse{ויביא brought עזרא And Ezra הכהן the priest את התורה the law לפני before הקהל the congregation מאישׁ both of men ועד אשׁה and women, וכל and all מבין with understanding, לשׁמע that could hear ביום day אחד upon the first לחדשׁ month. השׁביעי׃ of the seventh}%
\verse{ויקרא And he read בו לפני therein before הרחוב the street אשׁר that לפני before שׁער gate המים the water מן from האור the morning עד until מחצית midday, היום midday, נגד before האנשׁים the men והנשׁים and the women, והמבינים and those that could understand; ואזני and the ears כל of all העם the people אל unto ספר the book התורה׃ of the law.}%
\verse{ויעמד stood עזרא And Ezra הספר the scribe על upon מגדל a pulpit עץ of wood, אשׁר which עשׂו they had made לדבר for the purpose; ויעמד him stood אצלו and beside מתתיה Mattithiah, ושׁמע and Shema, ועניה and Anaiah, ואוריה and Urijah, וחלקיה and Hilkiah, ומעשׂיה and Maaseiah, על on ימינו his right hand; ומשׂמאלו and on his left hand, פדיה Pedaiah, ומישׁאל and Mishael, ומלכיה and Malchiah, וחשׁם and Hashum, וחשׁבדנה and Hashbadana, זכריה Zechariah, משׁלם׃ Meshullam.}%
\verse{ויפתח opened עזרא And Ezra הספר the book לעיני in the sight כל of all העם the people; כי (for מעל above כל all העם the people;) היה he was וכפתחו and when he opened עמדו stood up: כל it, all העם׃ the people}%
\verse{ויברך blessed עזרא And Ezra את יהוה the LORD, האלהים God. הגדול the great ויענו answered, כל And all העם the people אמן Amen, אמן Amen, במעל with lifting up ידיהם their hands: ויקדו and they bowed וישׁתחו their heads, and worshiped ליהוה the LORD אפים with faces ארצה׃ to the ground.}%
\verse{וישׁוע Also Jeshua, ובני and Bani, ושׁרביה and Sherebiah, ימין Jamin, עקוב Akkub, שׁבתי Shabbethai, הודיה Hodijah, מעשׂיה Maaseiah, קליטא Kelita, עזריה Azariah, יוזבד Jozabad, חנן Hanan, פלאיה Pelaiah, והלוים and the Levites, מבינים to understand את העם caused the people לתורה the law: והעם and the people על in עמדם׃}%
\verse{ויקראו So they read בספר in the book בתורת in the law האלהים of God מפרשׁ distinctly, ושׂום and gave שׂכל the sense, ויבינו and caused to understand במקרא׃ the reading.}%
\verse{ויאמר said נחמיה And Nehemiah, הוא which התרשׁתא the Tirshatha, ועזרא and Ezra הכהן the priest הספר the scribe, והלוים and the Levites המבינים that taught את העם the people, לכל unto all העם the people, היום day קדשׁ holy הוא This ליהוה unto the LORD אלהיכם your God; אל not, תתאבלו mourn ואל nor תבכו weep. כי For בוכים wept, כל all העם the people כשׁמעם when they heard את דברי the words התורה׃ of the law.}%
\verse{ויאמר Then he said להם לכו unto them, Go אכלו your way, eat משׁמנים the fat, ושׁתו and drink ממתקים the sweet, ושׁלחו and send מנות portions לאין unto them for whom nothing נכון is prepared: לו כי for קדושׁ holy היום day לאדנינו ואל neither תעצבו be ye sorry; כי for חדות the joy יהוה of the LORD היא מעזכם׃ is your strength.}%
\verse{והלוים So the Levites מחשׁים stilled לכל all העם the people, לאמר saying, הסו Hold your peace, כי for היום the day קדשׁ holy; ואל neither תעצבו׃ be ye grieved.}%
\verse{וילכו went כל And all העם the people לאכל their way to eat, ולשׁתות and to drink, ולשׁלח and to send מנות portions, ולעשׂות and to make שׂמחה mirth, גדולה great כי because הבינו they had understood בדברים the words אשׁר that הודיעו׃ were declared}%
\verseWithHeading{Festival of Booths}{וביום day השׁני And on the second נאספו were gathered together ראשׁי the chief האבות of the fathers לכל of all העם the people, הכהנים the priests, והלוים and the Levites, אל unto עזרא Ezra הספר the scribe, ולהשׂכיל even to understand אל even to understand דברי the words התורה׃ of the law.}%
\verse{וימצאו And they found כתוב written בתורה in the law אשׁר which צוה had commanded יהוה the LORD ביד by משׁה Moses, אשׁר that ישׁבו should dwell בני the children ישׂראל of Israel בסכות in booths בחג in the feast בחדשׁ month: השׁביעי׃ of the seventh}%
\verse{ואשׁר And that ישׁמיעו they should publish ויעבירו and proclaim קול and proclaim בכל in all עריהם their cities, ובירושׁלם and in Jerusalem, לאמר saying, צאו Go forth ההר unto the mount, והביאו and fetch עלי branches, זית olive ועלי branches, עץ and pine שׁמן and pine ועלי branches, הדס and myrtle ועלי branches, תמרים and palm ועלי and branches עץ trees, עבת of thick לעשׂת to make סכת booths, ככתוב׃ as written.}%
\verse{ויצאו went forth, העם So the people ויביאו and brought ויעשׂו and made להם סכות themselves booths, אישׁ every one על upon גגו the roof of his house, ובחצרתיהם and in their courts, ובחצרות and in the courts בית of the house האלהים of God, וברחוב and in the street שׁער gate, המים of the water וברחוב and in the street שׁער of the gate אפרים׃ of Ephraim.}%
\verse{ויעשׂו made כל And all הקהל the congregation השׁבים of them that were come again מן out of השׁבי the captivity סכות booths, וישׁבו and sat בסכות under the booths: כי for לא had not עשׂו done מימי since the days ישׁוע of Jeshua בן the son נון of Nun כן so. בני the children ישׂראל of Israel עד unto היום day ההוא that ותהי And there was שׂמחה gladness. גדולה great מאד׃ very}%
\verse{ויקרא he read בספר in the book תורת of the law האלהים of God. יום Also day ביום by day, מן from היום day הראשׁון the first עד unto היום day, האחרון the last ויעשׂו And they kept חג the feast שׁבעת seven ימים days; וביום day השׁמיני and on the eighth עצרת a solemn assembly, כמשׁפט׃ according unto the manner.}%
\end{biblechapter}%
\begin{biblechapter}% Nehemiah 9
\verseWithHeading{The Nation Confesses}{וביום day עשׂרים Now in the twenty וארבעה and fourth לחדשׁ month הזה of this נאספו were assembled בני the children ישׂראל of Israel בצום with fasting, ובשׂקים and with sackclothes, ואדמה and earth עליהם׃ upon}%
\verse{ויבדלו separated themselves זרע And the seed ישׂראל of Israel מכל from all בני strangers, נכר strangers, ויעמדו and stood ויתודו and confessed על and confessed חטאתיהם their sins, ועונות and the iniquities אבתיהם׃ of their fathers.}%
\verse{ויקומו And they stood up על in עמדם their place, ויקראו and read בספר in the book תורת of the law יהוה of the LORD אלהיהם their God רבעית fourth היום part of the day; ורבעית and fourth מתודים part they confessed, ומשׁתחוים and worshiped ליהוה the LORD אלהיהם׃ their God.}%
\verse{ויקם Then stood up על upon מעלה the stairs, הלוים of the Levites, ישׁוע Jeshua, ובני and Bani, קדמיאל Kadmiel, שׁבניה Shebaniah, בני Bunni, שׁרביה Sherebiah, בני Bani, כנני Chenani, ויזעקו and cried בקול voice גדול with a loud אל unto יהוה the LORD אלהיהם׃ their God.}%
\verse{ויאמרו said, הלוים Then the Levites, ישׁוע Jeshua, וקדמיאל and Kadmiel, בני Bani, חשׁבניה Hashabniah, שׁרביה Sherebiah, הודיה Hodijah, שׁבניה Shebaniah, פתחיה Pethahiah, קומו Stand up ברכו bless את יהוה the LORD אלהיכם your God מן forever העולם forever עד and העולם ever: ויברכו and blessed שׁם name, כבודך be thy glorious ומרומם which is exalted על above כל all ברכה blessing ותהלה׃ and praise.}%
\verse{אתה Thou, הוא thou, יהוה LORD לבדך alone; את עשׂית hast made את השׁמים heaven, שׁמי the heaven השׁמים of heavens, וכל with all צבאם their host, הארץ the earth, וכל and all אשׁר that עליה therein, הימים the seas, וכל and all אשׁר that בהם ואתה thou מחיה preservest את כלם them all; וצבא and the host השׁמים of heaven לך משׁתחוים׃ worshipeth}%
\verse{אתה Thou הוא יהוה the LORD האלהים the God, אשׁר who בחרת didst choose באברם Abram, והוצאתו and broughtest him forth מאור כשׂדים of the Chaldees, ושׂמת and gavest שׁמו him the name אברהם׃ of Abraham;}%
\verse{ומצאת And foundest את לבבו his heart נאמן faithful לפניך before וכרות thee, and madest עמו with הברית a covenant לתת him to give את ארץ the land הכנעני of the Canaanites, החתי the Hittites, האמרי the Amorites, והפרזי and the Perizzites, והיבוסי and the Jebusites, והגרגשׁי and the Girgashites, לתת to give לזרעו to his seed, ותקם and hast performed את דבריך thy words; כי for צדיק righteous: אתה׃ thou}%
\verse{ותרא And didst see את עני the affliction אבתינו of our fathers במצרים in Egypt, ואת זעקתם their cry שׁמעת and heardest על by ים sea; סוף׃ the Red}%
\verse{ותתן And showedst אתת signs ומפתים and wonders בפרעה upon Pharaoh, ובכל and on all עבדיו his servants, ובכל and on all עם the people ארצו of his land: כי for ידעת thou knewest כי that הזידו they dealt proudly עליהם against ותעשׂ them. So didst thou get לך שׁם thee a name, כהיום day. הזה׃ as this}%
\verse{והים the sea בקעת And thou didst divide לפניהם before ויעברו them, so that they went through בתוך the midst הים of the sea ביבשׁה on the dry land; ואת רדפיהם and their persecutors השׁלכת thou threwest במצולת into the deeps, כמו as אבן a stone במים waters. עזים׃ into the mighty}%
\verse{ובעמוד pillar; ענן by a cloudy הנחיתם Moreover thou leddest יומם them in the day ובעמוד by a pillar אשׁ of fire, לילה and in the night להאיר to give them light להם אתדרך in the way אשׁר wherein ילכו׃ they should go.}%
\verse{ועל also upon הר mount סיני Sinai, ירדת Thou camest down ודבר and spakest עמהם with משׁמים them from heaven, ותתן and gavest להם משׁפטים judgments, ישׁרים them right ותורות laws, אמת and true חקים statutes ומצות and commandments: טובים׃ good}%
\verse{ואת שׁבת sabbath, קדשׁך unto them thy holy הודעת And madest known להם ומצוות them precepts, וחקים statutes, ותורה and laws, צוית and commandedst להם ביד by the hand משׁה of Moses עבדך׃ thy servant:}%
\verse{ולחם them bread משׁמים from heaven נתתה And gavest להם לרעבם for their hunger, ומים water מסלע for them out of the rock הוצאת and broughtest forth להם לצמאם for their thirst, ותאמר and promisedst להם לבוא them that they should go in לרשׁת to possess את הארץ the land אשׁר which נשׂאת אתדך לתת׃ to give}%
\verse{והם But they ואבתינו and our fathers הזידו dealt proudly, ויקשׁו and hardened את ערפם their necks, ולא not שׁמעו and hearkened אל to מצותיך׃ thy commandments,}%
\verse{וימאנו And refused לשׁמע to obey, ולא neither זכרו were mindful נפלאתיך of thy wonders אשׁר that עשׂית thou didst עמהם among ויקשׁו them; but hardened את ערפם their necks, ויתנו appointed ראשׁ a captain לשׁוב to return לעבדתם to their bondage: במרים and in their rebellion ואתה but thou אלוה a God סליחות ready to pardon, חנון gracious ורחום and merciful, ארך slow אפים to anger, ורב and of great וחסד kindness, ולא them not. עזבתם׃ and forsookest}%
\verse{אף Yea, כי when עשׂו they had made להם עגל calf, מסכה them a molten ויאמרו and said, זה This אלהיך thy God אשׁר that העלך brought thee up ממצרים ויעשׂו and had wrought נאצות provocations; גדלות׃ great}%
\verse{ואתה Yet thou ברחמיך mercies הרבים in thy manifold לא them not עזבתם forsookest במדבר in the wilderness: את עמוד the pillar הענן of the cloud לא not סר departed מעליהם from ביומם them by day, להנחתם to lead בהדרך them in the way; ואת עמוד neither the pillar האשׁ of fire בלילה by night, להאיר to show them light, להם ואת הדרך and the way אשׁר wherein ילכו׃ they should go.}%
\verse{ורוחך spirit הטובה also thy good נתת Thou gavest להשׂכילם to instruct ומנך thy manna לא not מנעת them, and withheldest מפיהם from their mouth, ומים them water נתתה and gavest להם לצמאם׃ for their thirst.}%
\verse{וארבעים Yea, forty שׁנה years כלכלתם didst thou sustain במדבר them in the wilderness, לא nothing; חסרו they lacked שׂלמתיהם their clothes לא waxed not old, בלו waxed not old, ורגליהם and their feet לא not. בצקו׃ swelled}%
\verse{ותתן Moreover thou gavest להם ממלכות them kingdoms ועממים and nations, ותחלקם and didst divide לפאה them into corners: ויירשׁו so they possessed את ארץ the land סיחון of Sihon, ואת ארץ and the land מלך of the king חשׁבון of Heshbon, ואת ארץ and the land עוג of Og מלך king הבשׁן׃ of Bashan.}%
\verse{ובניהם Their children הרבית also multipliedst כככבי thou as the stars השׁמים of heaven, ותביאם and broughtest אל them into הארץ the land, אשׁר concerning which אמרת thou hadst promised לאבתיהם to their fathers, לבוא that they should go in לרשׁת׃ to possess}%
\verse{ויבאו went in הבנים So the children ויירשׁו and possessed את הארץ the land, ותכנע and thou subduedst לפניהם before את ישׁבי them the inhabitants הארץ of the land, הכנענים the Canaanites, ותתנם and gavest בידם them into their hands, ואת מלכיהם with their kings, ואת עממי and the people הארץ of the land, לעשׂות that they might do בהם כרצונם׃ with them as they would.}%
\verse{וילכדו And they took ערים cities, בצרות strong ואדמה land, שׁמנה and a fat ויירשׁו and possessed בתים houses מלאים full כל of all טוב goods, ברות wells חצובים digged, כרמים vineyards, וזיתים and oliveyards, ועץ trees מאכל and fruit לרב in abundance: ויאכלו so they did eat, וישׂבעו and were filled, וישׁמינו and became fat, ויתעדנו and delighted themselves בטובך goodness. הגדול׃ in thy great}%
\verse{וימרו Nevertheless they were disobedient, וימרדו and rebelled בך וישׁלכו against thee, and cast את תורתך thy law אחרי behind גום their backs, ואת נביאיך thy prophets הרגו and slew אשׁר which העידו testified בם להשׁיבם against them to turn אליך them to ויעשׂו thee, and they wrought נאצות provocations. גדולת׃ great}%
\verse{ותתנם Therefore thou deliveredst ביד them into the hand צריהם of their enemies, ויצרו להם ובעת them: and in the time צרתם of their trouble, יצעקו when they cried אליך unto ואתה thee, thou משׁמים from heaven; תשׁמע heardest וכרחמיך mercies הרבים and according to thy manifold תתן thou gavest להם מושׁיעים them saviors, ויושׁיעום who saved מיד them out of the hand צריהם׃ of their enemies.}%
\verse{וכנוח But after they had rest, להם ישׁובו again לעשׂות they did רע evil לפניך before ותעזבם thee: therefore leftest ביד thou them in the hand איביהם of their enemies, וירדו so that they had the dominion בהם וישׁובו over them: yet when they returned, ויזעקוך and cried ואתה unto thee, thou משׁמים from heaven; תשׁמע heardest ותצילם didst thou deliver כרחמיך them according to thy mercies; רבות and many עתים׃ times}%
\verse{ותעד And testifiedst בהם להשׁיבם against them, that thou mightest bring them again אל unto תורתך thy law: והמה yet they הזידו dealt proudly, ולא not שׁמעו and hearkened למצותיך unto thy commandments, ובמשׁפטיך against thy judgments, חטאו but sinned בם אשׁר (which יעשׂה do, אדם if a man וחיה he shall live בהם ויתנו כתף the shoulder, סוררת וערפם their neck, הקשׁו and hardened ולא and would not שׁמעו׃ hear.}%
\verse{ותמשׁך didst thou forbear עליהם didst thou forbear שׁנים years רבות Yet many ותעד them, and testifiedst בם ברוחך against them by thy spirit ביד in נביאיך thy prophets: ולא yet would they not האזינו give ear: ותתנם therefore gavest ביד thou them into the hand עמי of the people הארצת׃ of the lands.}%
\verse{וברחמיך mercies' הרבים Nevertheless for thy great לא sake thou didst not עשׂיתם utterly consume כלה utterly consume ולא them, nor עזבתם forsake כי them; for אל God. חנון a gracious ורחום and merciful אתה׃ thou}%
\verse{ועתה Now אלהינו therefore, our God, האל God, הגדול the great, הגבור the mighty, והנורא and the terrible שׁומר who keepest הברית covenant והחסד and mercy, אל let not ימעט seem little לפניך before את כל all התלאה the trouble אשׁר thee, that מצאתנו hath come upon למלכינו us, on our kings, לשׂרינו on our princes, ולכהנינו and on our priests, ולנביאנו and on our prophets, ולאבתינו and on our fathers, ולכל and on all עמך thy people, מימי since the time מלכי of the kings אשׁור of Assyria עד unto היום day. הזה׃ this}%
\verse{ואתה Howbeit thou צדיק just על in כל all הבא that is brought עלינו upon כי us; for אמת right, עשׂית thou hast done ואנחנו but we הרשׁענו׃ have done wickedly:}%
\verse{ואת מלכינו Neither have our kings, שׂרינו our princes, כהנינו our priests, ואבתינו our fathers, לא nor עשׂו kept תורתך thy law, ולא nor הקשׁיבו hearkened אל unto מצותיך thy commandments ולעדותיך and thy testimonies, אשׁר wherewith העידת׃ thou didst testify}%
\verse{והם For they במלכותם thee in their kingdom, ובטובך goodness הרב and in thy great אשׁר that נתת thou gavest להם ובארץ land הרחבה them, and in the large והשׁמנה and fat אשׁר which נתת thou gavest לפניהם before לא have not עבדוך served ולא them, neither שׁבו turned ממעלליהם works. הרעים׃ they from their wicked}%
\verse{הנה Behold, אנחנו we היום this day, עבדים servants והארץ and the land אשׁר that נתתה thou gavest לאבתינו unto our fathers לאכל to eat את פריה the fruit ואת טובה thereof and the good הנה thereof, behold, אנחנו we עבדים servants עליה׃ in}%
\verse{ותבואתה increase מרבה And it yieldeth much למלכים unto the kings אשׁר whom נתתה thou hast set עלינו over בחטאותינו us because of our sins: ועל over גויתינו our bodies, משׁלים also they have dominion ובבהמתנו and over our cattle, כרצונם at their pleasure, ובצרה distress. גדולה in great אנחנו׃ and we}%
\verse{ובכל And because of all זאת this אנחנו we כרתים make אמנה a sure וכתבים and write ועל seal החתום seal שׂרינו and our princes, לוינו Levites, כהנינו׃ priests,}%
\end{biblechapter}%
\begin{biblechapter}% Nehemiah 10
\verseWithHeading{The Names of Those Who Signed the Covenant}{ועל החתומיםחמיה Nehemiah, התרשׁתא the Tirshatha, בן the son חכליה of Hachaliah, וצדקיה׃ and Zidkijah,}%
\verse{שׂריה Seraiah, עזריה Azariah, ירמיה׃ Jeremiah,}%
\verse{פשׁחור Pashur, אמריה Amariah, מלכיה׃ Malchijah,}%
\verse{חטושׁ Hattush, שׁבניה Shebaniah, מלוך׃ Malluch,}%
\verse{חרם Harim, מרמות Meremoth, עבדיה׃ Obadiah,}%
\verse{דניאל Daniel, גנתון Ginnethon, ברוך׃ Baruch,}%
\verse{משׁלם Meshullam, אביה Abijah, מימן׃ Mijamin,}%
\verse{מעזיה Maaziah, בלגי Bilgai, שׁמעיה Shemaiah: אלה these הכהנים׃ the priests.}%
\verse{והלוים And the Levites: וישׁוע both Jeshua בן the son אזניה of Azaniah, בנוי Binnui מבני of the sons חנדד of Henadad, קדמיאל׃ Kadmiel;}%
\verse{ואחיהם And their brethren, שׁבניה Shebaniah, הודיה קליטא Kelita, פלאיה Pelaiah, חנן׃ Hanan,}%
\verse{מיכא Micha, רחוב Rehob, חשׁביה׃ Hashabiah,}%
\verse{זכור Zaccur, שׁרביה Sherebiah, שׁבניה׃ Shebaniah,}%
\verse{הודיה בני Bani, בנינו׃ Beninu.}%
\verse{ראשׁי The chief העם of the people; פרעשׁ Parosh, פחת מואב Pahath-moab, עילם Elam, זתוא Zatthu, בני׃ Bani,}%
\verse{בני Bunni, עזגד Azgad, בבי׃ Bebai,}%
\verse{אדניה Adonijah, בגוי Bigvai, עדין׃ Adin,}%
\verse{אטר Ater, חזקיה Hizkijah, עזור׃ Azzur,}%
\verse{הודיה חשׁם Hashum, בצי׃ Bezai,}%
\verse{חריף Hariph, ענתות Anathoth, נובי׃ Nebai,}%
\verse{מגפיעשׁ Magpiash, משׁלם Meshullam, חזיר׃ Hezir,}%
\verse{משׁיזבאל Meshezabeel, צדוק Zadok, ידוע׃ Jaddua,}%
\verse{פלטיה Pelatiah, חנן Hanan, עניה׃ Anaiah,}%
\verse{הושׁע Hoshea, חנניה Hananiah, חשׁוב׃ Hashub,}%
\verse{הלוחשׁ Hallohesh, פלחא Pileha, שׁובק׃ Shobek,}%
\verse{רחום Rehum, חשׁבנה Hashabnah, מעשׂיה׃ Maaseiah,}%
\verse{ואחיה And Ahijah, חנן Hanan, ענן׃ Anan,}%
\verse{מלוך Malluch, חרם Harim, בענה׃ Baanah.}%
\verseWithHeading{A Summary of the Covenant}{ושׁאר And the rest העם of the people, הכהנים the priests, הלוים the Levites, השׁוערים the porters, המשׁררים the singers, הנתינים the Nethinims, וכל and all הנבדל they that had separated themselves מעמי from the people הארצות of the lands אל unto תורת the law האלהים of God, נשׁיהם their wives, בניהם their sons, ובנתיהם and their daughters, כל every יודע one having knowledge, מבין׃ and having understanding;}%
\verse{מחזיקים They cleaved על to אחיהם their brethren, אדיריהם their nobles, ובאים and entered באלה into a curse, ובשׁבועה and into an oath, ללכת to walk בתורת law, האלהים in God's אשׁר which נתנה was given ביד by משׁה Moses עבד the servant האלהים of God, ולשׁמור and to observe ולעשׂות and do את כל all מצות the commandments יהוה of the LORD אדנינו ומשׁפטיו and his judgments וחקיו׃ and his statutes;}%
\verse{ואשׁר And that לא we would not נתן give בנתינו our daughters לעמי unto the people הארץ of the land, ואת בנתיהם their daughters לא nor נקח take לבנינו׃ for our sons:}%
\verse{ועמי And the people הארץ of the land המביאים bring את המקחות ware וכל or any שׁבר victuals ביום day השׁבת on the sabbath למכור to sell, לא we would not נקח buy מהם בשׁבת them on the sabbath, וביום day: קדשׁ or on the holy ונטשׁ and we would leave את השׁנה year, השׁביעית the seventh ומשׁא and the exaction כל of every יד׃ debt.}%
\verse{והעמדנו Also we made עלינו for מצות ordinances לתת us, to charge עלינו ourselves שׁלשׁית with the third part השׁקל of a shekel בשׁנה yearly לעבדת for the service בית of the house אלהינו׃ of our God;}%
\verse{ללחם המערכתמנחת meat offering, התמיד and for the continual ולעולת burnt offering, התמיד and for the continual השׁבתות of the sabbaths, החדשׁים of the new moons, למועדים for the set feasts, ולקדשׁים and for the holy ולחטאות and for the sin offerings לכפר to make an atonement על for ישׂראל Israel, וכל and all מלאכת the work בית of the house אלהינו׃ of our God.}%
\verse{והגורלות הפלנול for קרבן offering, העצים the wood הכהנים the priests, הלוים the Levites, והעם and the people, להביא to bring לבית into the house אלהינו of our God, לבית after the houses אבתינו of our fathers, לעתים at times מזמנים appointed שׁנה year בשׁנה by year, לבער to burn על upon מזבח the altar יהוה of the LORD אלהינו our God, ככתוב as written בתורה׃ in the law:}%
\verse{ולהביא And to bring את בכורי the firstfruits אדמתנו of our ground, ובכורי and the firstfruits כל of all פרי fruit כל of all עץ trees, שׁנה year בשׁנה by year, לבית unto the house יהוה׃ of the LORD:}%
\verse{ואת בכרות Also the firstborn בנינו of our sons, ובהמתינו and of our cattle, ככתוב as written בתורה in the law, ואת בכוריקרינו of our herds וצאנינו and of our flocks, להביא to bring לבית to the house אלהינו of our God, לכהנים unto the priests המשׁרתים that minister בבית in the house אלהינו׃ of our God:}%
\verse{ואת ראשׁית the firstfruits עריסתינו of our dough, ותרומתינו and our offerings, ופרי and the fruit כל of all עץ manner of trees, תירושׁ of wine ויצהר and of oil, נביא And we should bring לכהנים unto the priests, אל to לשׁכות the chambers בית of the house אלהינו of our God; ומעשׂר and the tithes אדמתנו of our ground ללוים unto the Levites, והם that the same הלוים Levites המעשׂרים might have the tithes בכל in all ערי the cities עבדתנו׃ of our tillage.}%
\verse{והיה shall be הכהן And the priest בן the son אהרן of Aaron עם with הלוים the Levites, בעשׂר take tithes: הלוים when the Levites והלוים and the Levites יעלו shall bring up את מעשׂר the tithe המעשׂר of the tithes לבית unto the house אלהינו of our God, אל to הלשׁכות the chambers, לבית house. האוצר׃ into the treasure}%
\verse{כי For אל unto הלשׁכות the chambers, יביאו shall bring בני the children ישׂראל of Israel ובני and the children הלוי of Levi את תרומת the offering הדגן of the corn, התירושׁ of the new wine, והיצהר and the oil, ושׁם where כלי the vessels המקדשׁ of the sanctuary, והכהנים and the priests המשׁרתים that minister, והשׁוערים and the porters, והמשׁררים and the singers: ולא and we will not נעזב forsake את בית the house אלהינו׃ of our God.}%
\end{biblechapter}%
\begin{biblechapter}% Nehemiah 11
\verseWithHeading{The Leaders and Servants in Jerusalem}{וישׁבו dwelt שׂרי And the rulers העם of the people בירושׁלם at Jerusalem: ושׁאר the rest העם of the people הפילו also cast גורלות lots, להביא to bring אחד one מן of העשׂרה ten לשׁבת to dwell בירושׁלם in Jerusalem עיר city, הקדשׁ the holy ותשׁע and nine הידות parts בערים׃ in cities.}%
\verse{ויברכו blessed העם And the people לכל all האנשׁים the men, המתנדבים that willingly offered themselves לשׁבת to dwell בירושׁלם׃ at Jerusalem.}%
\verse{ואלה Now these ראשׁי the chief המדינה of the province אשׁר that ישׁבו dwelt בירושׁלם in Jerusalem: ובערי but in the cities יהודה of Judah ישׁבו dwelt אישׁ every one באחזתו in his possession בעריהם in their cities, ישׂראל Israel, הכהנים the priests, והלוים and the Levites, והנתינים and the Nethinims, ובני and the children עבדי servants. שׁלמה׃ of Solomon's}%
\verse{ובירושׁלם And at Jerusalem ישׁבו dwelt מבני of the children יהודה of Judah, ומבני and of the children בנימן of Benjamin. מבני the son יהודה of Judah; עתיה Athaiah בן the son עזיה of Uzziah, בן the son זכריה of Zechariah, בן the son אמריה of Amariah, בן the son שׁפטיה of Shephatiah, בן of the children מהללאל of Mahalaleel, מבני פרץ׃ of Perez;}%
\verse{ומעשׂיה And Maaseiah בן the son ברוך of Baruch, בן the son כל חזה of Col-hozeh, בן the son חזיה of Hazaiah, בן the son עדיה of Adaiah, בן the son יויריב of Joiarib, בן the son זכריה of Zechariah, בן the son השׁלני׃}%
\verse{כל All בני the sons פרץ of Perez הישׁבים that dwelt בירושׁלם at Jerusalem ארבע four מאות hundred שׁשׁים threescore ושׁמנה and eight אנשׁי חיל׃ valiant}%
\verse{ואלה And these בני the sons בנימן of Benjamin; סלא Sallu בן the son משׁלם of Meshullam, בן the son יועד of Joed, בן the son פדיה of Pedaiah, בן the son קוליה of Kolaiah, בן the son מעשׂיה of Maaseiah, בן the son איתיאל of Ithiel, בן the son ישׁעיה׃ of Jesaiah.}%
\verse{ואחריו And after גבי him Gabbai, סלי Sallai, תשׁע nine מאות hundred עשׂרים twenty ושׁמנה׃ and eight.}%
\verse{ויואל And Joel בן the son זכרי of Zichri פקיד their overseer: עליהם their overseer: ויהודה and Judah בן the son הסנואה of Senuah על over העיר the city. משׁנה׃ second}%
\verse{מן Of הכהנים the priests: ידעיה Jedaiah בן the son יויריב of Joiarib, יכין׃ Jachin.}%
\verse{שׂריה Seraiah בן the son חלקיה of Hilkiah, בן the son משׁלם of Meshullam, בן the son צדוק of Zadok, בן the son מריות of Meraioth, בן the son אחיטוב of Ahitub, נגד the ruler בית of the house האלהים׃ of God.}%
\verse{ואחיהם And their brethren עשׂי that did המלאכה the work לבית of the house שׁמנה eight מאות hundred עשׂרים twenty ושׁנים and two: ועדיה and Adaiah בן the son ירחם of Jeroham, בן the son פלליה of Pelaliah, בן the son אמצי of Amzi, בן the son זכריה of Zechariah, בן the son פשׁחור of Pashur, בן the son מלכיה׃ of Malchiah,}%
\verse{ואחיו And his brethren, ראשׁים chief לאבות of the fathers, מאתים two hundred ארבעים forty ושׁנים and two: ועמשׁסי and Amashai בן the son עזראל of Azareel, בן the son אחזי of Ahasai, בן the son משׁלמות of Meshillemoth, בן the son אמר׃ of Immer,}%
\verse{ואחיהם And their brethren, גבורי mighty men חיל of valor, מאה a hundred עשׂרים twenty ושׁמנה and eight: ופקיד and their overseer עליהם and their overseer זבדיאל Zabdiel, בן the son הגדולים׃ of the great}%
\verse{ומן Also of הלוים the Levites: שׁמעיה Shemaiah בן the son חשׁוב of Hashub, בן the son עזריקם of Azrikam, בן the son חשׁביה of Hashabiah, בן the son בוני׃ of Bunni;}%
\verse{ושׁבתי And Shabbethai ויוזבד and Jozabad, על the oversight המלאכה business החיצנה of the outward לבית of the house האלהים of God. מראשׁי of the chief הלוים׃ of the Levites,}%
\verse{ומתניה And Mattaniah בן the son מיכה of Micha, בן the son זבדי of Zabdi, בן the son אסף of Asaph, ראשׁ the principal התחלה to begin יהודה the thanksgiving לתפלה in prayer: ובקבקיה and Bakbukiah משׁנה the second מאחיו among his brethren, ועבדא and Abda בן the son שׁמוע of Shammua, בן the son גלל of Galal, בן the son ידיתון׃ of Jeduthun.}%
\verse{כל All הלוים the Levites בעיר city הקדשׁ in the holy מאתים two hundred שׁמנים fourscore וארבעה׃ and four.}%
\verse{והשׁוערים Moreover the porters, עקוב Akkub, טלמון Talmon, ואחיהם and their brethren השׁמרים that kept בשׁערים the gates, מאה a hundred שׁבעים seventy ושׁנים׃ and two.}%
\verse{ושׁאר And the residue ישׂראל of Israel, הכהנים of the priests, הלוים the Levites, בכל in all ערי the cities יהודה of Judah, אישׁ every one בנחלתו׃ in his inheritance.}%
\verse{והנתינים But the Nethinims ישׁבים dwelt בעפל in Ophel: וציחא and Ziha וגשׁפא and Gispa על over הנתינים׃ the Nethinims.}%
\verse{ופקיד The overseer הלוים also of the Levites בירושׁלם at Jerusalem עזי Uzzi בן the son בני of Bani, בן the son חשׁביה of Hashabiah, בן the son מתניה of Mattaniah, בן the son מיכא of Micha. מבני אסף of Asaph, המשׁררים the singers לנגד over מלאכת the business בית of the house האלהים׃ of God.}%
\verse{כי For מצות commandment המלך the king's עליהם concerning ואמנה them, that a certain portion על should be for המשׁררים the singers, דבר due יום for every day. ביומו׃ for every day.}%
\verse{ופתחיה And Pethahiah בן the son משׁיזבאל of Meshezabeel, מבני of the children זרח of Zerah בן the son יהודה of Judah, ליד hand המלך at the king's לכל in all דבר matters לעם׃ concerning the people.}%
\verseWithHeading{Villages Outside of Jerusalem}{ואל And for החצרים the villages, בשׂדתם with their fields, מבני of the children יהודה of Judah ישׁבו dwelt בקרית הארבע at Kirjath-arba, ובנתיה and the villages ובדיבן thereof, and at Dibon, ובנתיה and the villages וביקבצאל thereof, and at Jekabzeel, וחצריה׃ and the villages}%
\verse{ובישׁוע And at Jeshua, ובמולדה and at Moladah, ובבית פלט׃ and at Beth-phelet,}%
\verse{ובחצר שׁועל And at Hazar-shual, ובבאר שׁבע and at Beer-sheba, ובנתיה׃ and the villages}%
\verse{ובצקלג And at Ziklag, ובמכנה and at Mekonah, ובבנתיה׃ and in the villages}%
\verse{ובעין רמון And at En-rimmon, ובצרעה and at Zareah, ובירמות׃ and at Jarmuth,}%
\verse{זנח Zanoah, עדלם Adullam, וחצריהם and their villages, לכישׁ at Lachish, ושׂדתיה and the fields עזקה thereof, at Azekah, ובנתיה and the villages ויחנו thereof. And they dwelt מבאר שׁבע עד unto גיא the valley הנם׃ of Hinnom.}%
\verse{ובני The children בנימן also of Benjamin מגבע מכמשׂ at Michmash, ועיה and Aija, ובית אל and Bethel, ובנתיה׃ and their villages,}%
\verse{ענתות at Anathoth, נב Nob, ענניה׃ Ananiah,}%
\verse{חצור Hazor, רמה Ramah, גתים׃ Gittaim,}%
\verse{חדיד Hadid, צבעים Zeboim, נבלט׃ Neballat,}%
\verse{לד Lod, ואונו and Ono, גי the valley החרשׁים׃}%
\verse{ומן And of הלוים the Levites מחלקות divisions יהודה Judah, לבנימין׃ in Benjamin.}%
\end{biblechapter}%
\begin{biblechapter}% Nehemiah 12
\verseWithHeading{The Leaders and Servants in Jerusalem}{ואלה Now these הכהנים the priests והלוים and the Levites אשׁר that עלו went up עם with זרבבל Zerubbabel בן the son שׁאלתיאל of Shealtiel, וישׁוע and Jeshua: שׂריה Seraiah, ירמיה Jeremiah, עזרא׃ Ezra,}%
\verse{אמריה Amariah, מלוך Malluch, חטושׁ׃ Hattush,}%
\verse{שׁכניה Shechaniah, רחם Rehum, מרמת׃ Meremoth,}%
\verse{עדוא Iddo, גנתוי Ginnetho, אביה׃ Abijah,}%
\verse{מימין Miamin, מעדיה Maadiah, בלגה׃ Bilgah,}%
\verse{שׁמעיה Shemaiah, ויויריב and Joiarib, ידעיה׃ Jedaiah,}%
\verse{סלו Sallu, עמוק Amok, חלקיה Hilkiah, ידעיה Jedaiah. אלה These ראשׁי the chief הכהנים of the priests ואחיהם and of their brethren בימי in the days ישׁוע׃ of Jeshua.}%
\verse{והלוים Moreover the Levites: ישׁוע Jeshua, בנוי Binnui, קדמיאל Kadmiel, שׁרביה Sherebiah, יהודה Judah, מתניה Mattaniah, על over הידות the thanksgiving, הוא he ואחיו׃ and his brethren.}%
\verse{ובקבקיה Also Bakbukiah וענו and Unni, אחיהם their brethren, לנגדם over against למשׁמרות׃ them in the watches.}%
\verse{וישׁוע And Jeshua הוליד begot את יויקים Joiakim, ויויקים Joiakim הוליד also begot את אלישׁיב Eliashib, ואלישׁיב and Eliashib את יוידע׃ Joiada,}%
\verse{ויוידע And Joiada הוליד begot את יונתן Jonathan, ויונתן and Jonathan הוליד begot את ידוע׃ Jaddua.}%
\verse{ובימי And in the days יויקים of Joiakim היו were כהנים priests, ראשׁי the chief האבות of the fathers: לשׂריה of Seraiah, מריה Meraiah; לירמיה of Jeremiah, חנניה׃ Hananiah;}%
\verse{לעזרא Of Ezra, משׁלם Meshullam; לאמריה of Amariah, יהוחנן׃ Jehohanan;}%
\verse{למלוכי Of Melicu, יונתן Jonathan; לשׁבניה of Shebaniah, יוסף׃ Joseph;}%
\verse{לחרם Of Harim, עדנא Adna; למריות of Meraioth, חלקי׃ Helkai;}%
\verse{לעדיא Of Iddo, זכריה Zechariah; לגנתון of Ginnethon, משׁלם׃ Meshullam;}%
\verse{לאביה Of Abijah, זכרי Zichri; למנימין of Miniamin, למועדיה of Moadiah, פלטי׃ Piltai;}%
\verse{לבלגה Of Bilgah, שׁמוע Shammua; לשׁמעיה of Shemaiah, יהונתן׃ Jehonathan;}%
\verse{וליויריב And of Joiarib, מתני Mattenai; לידעיה of Jedaiah, עזי׃ Uzzi;}%
\verse{לסלי Of Sallai, קלי Kallai; לעמוק of Amok, עבר׃ Eber;}%
\verse{לחלקיה Of Hilkiah, חשׁביה Hashabiah; לידעיה of Jedaiah, נתנאל׃ Nethaneel.}%
\verse{הלוים The Levites בימי in the days אלישׁיב of Eliashib, יוידע Joiada, ויוחנן and Johanan, וידוע and Jaddua, כתובים recorded ראשׁי chief אבות of the fathers: והכהנים also the priests, על to מלכות the reign דריושׁ of Darius הפרסי׃ the Persian.}%
\verse{בני The sons לוי of Levi, ראשׁי the chief האבות of the fathers, כתובים written על in ספר the book דברי of the chronicles, הימים the days ועד even until ימי יוחנן of Johanan בן the son אלישׁיב׃ of Eliashib.}%
\verse{וראשׁי And the chief הלוים of the Levites: חשׁביה Hashabiah, שׁרביה Sherebiah, וישׁוע and Jeshua בן the son קדמיאל of Kadmiel, ואחיהם with their brethren לנגדם over against להלל them, to praise להודות to give thanks, במצות according to the commandment דויד of David אישׁ the man האלהים of God, משׁמר ward לעמת over against משׁמר׃ ward.}%
\verse{מתניה Mattaniah, ובקבקיה and Bakbukiah, עבדיה Obadiah, משׁלם Meshullam, טלמון Talmon, עקוב Akkub, שׁמרים keeping שׁוערים porters משׁמר the ward באספי at the thresholds השׁערים׃ of the gates.}%
\verse{אלה These בימי in the days יויקים of Joiakim בן the son ישׁוע of Jeshua, בן the son יוצדק of Jozadak, ובימי and in the days נחמיה of Nehemiah הפחה the governor, ועזרא and of Ezra הכהן the priest, הסופר׃ the scribe.}%
\verseWithHeading{Dedication of the Wall of Jerusalem}{ובחנכת And at the dedication חומת of the wall ירושׁלם of Jerusalem בקשׁו they sought את הלוים the Levites מכל out of all מקומתם their places, להביאם to bring לירושׁלם them to Jerusalem, לעשׂת to keep חנכה the dedication ושׂמחה with gladness, ובתודות both with thanksgivings, ובשׁיר and with singing, מצלתים cymbals, נבלים psalteries, ובכנרות׃ and with harps.}%
\verse{ויאספו gathered themselves together, בני And the sons המשׁררים of the singers ומן both out of הככר the plain country סביבות round about ירושׁלם Jerusalem, ומן and from חצרי the villages נטפתי׃ of Netophathi;}%
\verse{ומבית הגלגל of Gilgal, ומשׂדות and out of the fields גבע of Geba ועזמות and Azmaveth: כי for חצרים them villages בנו had built להם המשׁררים the singers סביבות round about ירושׁלם׃ Jerusalem.}%
\verse{ויטהרו purified themselves, הכהנים And the priests והלוים and the Levites ויטהרו and purified את העם the people, ואת השׁערים and the gates, ואת החומה׃ and the wall.}%
\verse{ואעלה Then I brought up את שׂרי the princes יהודה of Judah מעל upon לחומה the wall, ואעמידה and appointed שׁתי two תודת thanks, גדולת great ותהלכת לימין on the right hand מעל upon לחומה the wall לשׁער gate: האשׁפת׃ toward the dung}%
\verse{וילך them went אחריהם And after הושׁעיה Hoshaiah, וחצי and half שׂרי of the princes יהודה׃ of Judah,}%
\verse{ועזריה And Azariah, עזרא Ezra, ומשׁלם׃ and Meshullam,}%
\verse{יהודה Judah, ובנימן and Benjamin, ושׁמעיה and Shemaiah, וירמיה׃ and Jeremiah,}%
\verse{ומבני the son הכהנים בחצצרות with trumpets; זכריה Zechariah בן the son יונתן of Jonathan, בן the son שׁמעיה of Shemaiah, בן the son מתניה of Mattaniah, בן the son מיכיה of Michaiah, בן the son זכור of Zaccur, בן אסף׃ of Asaph:}%
\verse{ואחיו And his brethren, שׁמעיה Shemaiah, ועזראל and Azarael, מללי Milalai, גללי Gilalai, מעי Maai, נתנאל Nethaneel, ויהודה and Judah, חנני Hanani, בכלי instruments שׁיר with the musical דויד of David אישׁ the man האלהים of God, ועזרא and Ezra הסופר the scribe לפניהם׃ before}%
\verse{ועל And at שׁער gate, העין the fountain ונגדם which was over against עלו them, they went up על by מעלות the stairs עיר of the city דויד of David, במעלה at the going up לחומה of the wall, מעל above לבית the house דויד of David, ועד even unto שׁער gate המים the water מזרח׃ eastward.}%
\verse{והתודה thanks השׁנית And the other ההולכת went למואל over against ואני and I אחריה after וחצי them, and the half העם of the people מעל upon להחומה the wall, מעל from beyond למגדל the tower התנורים of the furnaces ועד even unto החומה wall; הרחבה׃ the broad}%
\verse{ומעל and above לשׁער the gate אפרים of Ephraim, ועל and above שׁער gate, הישׁנה ועלׁער gate, הדגים the fish ומגדל and the tower חננאל of Hananeel, ומגדל and the tower המאה of Meah, ועד even unto שׁער gate: הצאן the sheep ועמדו and they stood still בשׁער gate. המטרה׃ in the prison}%
\verse{ותעמדנה So stood שׁתי the two התודת thanks בבית in the house האלהים of God, ואני and I, וחצי and the half הסגנים of the rulers עמי׃ with}%
\verse{והכהנים And the priests; אליקים Eliakim, מעשׂיה Maaseiah, מנימין Miniamin, מיכיה Michaiah, אליועיני Elioenai, זכריה Zechariah, חנניה Hananiah, בחצצרות׃ with trumpets;}%
\verse{ומעשׂיה And Maaseiah, ושׁמעיה and Shemaiah, ואלעזר and Eleazar, ועזי and Uzzi, ויהוחנן and Jehohanan, ומלכיה and Malchijah, ועילם and Elam, ועזר and Ezer. וישׁמיעו sang loud, המשׁררים And the singers ויזרחיה with Jezrahiah הפקיד׃ overseer.}%
\verse{ויזבחו they offered ביום day ההוא Also that זבחים sacrifices, גדולים great וישׂמחו and rejoiced: כי for האלהים God שׂמחם had made them rejoice שׂמחה joy: גדולה with great וגם also הנשׁים the wives והילדים and the children שׂמחו rejoiced: ותשׁמע was heard שׂמחת so that the joy ירושׁלם of Jerusalem מרחוק׃ even afar off.}%
\verseWithHeading{Celebration at the Temple}{ויפקדו appointed ביום time ההוא And at that אנשׁים were some על over הנשׁכות the chambers לאוצרות for the treasures, לתרומות for the offerings, לראשׁית for the firstfruits, ולמעשׂרות and for the tithes, לכנוס to gather בהם לשׂדי into them out of the fields הערים of the cities מנאות the portions התורה of the law לכהנים for the priests וללוים and Levites: כי for שׂמחת rejoiced יהודה Judah על for הכהנים the priests ועל and for הלוים the Levites העמדים׃ that waited.}%
\verse{וישׁמרו kept משׁמרת the ward אלהיהם of their God, ומשׁמרת and the ward הטהרה of the purification, והמשׁררים And both the singers והשׁערים and the porters כמצות according to the commandment דויד of David, שׁלמה of Solomon בנו׃ his son.}%
\verse{כי For בימי in the days דויד of David ואסף and Asaph מקדם of old ראשׁ chief המשׁררים of the singers, ושׁיר and songs תהלה of praise והדות and thanksgiving לאלהים׃ unto God.}%
\verse{וכל And all ישׂראל Israel בימי in the days זרבבל of Zerubbabel, ובימי and in the days נחמיה of Nehemiah, נתנים gave מניות the portions המשׁררים of the singers והשׁערים and the porters, דבר his portion: יום every day ביומו every day ומקדשׁים and they sanctified ללוים unto the Levites; והלוים and the Levites מקדשׁים sanctified לבני unto the children אהרן׃ of Aaron.}%
\end{biblechapter}%
\begin{biblechapter}% Nehemiah 13
\verseWithHeading{Israel Separates Itself}{ביום day ההוא On that נקרא they read בספר in the book משׁה of Moses באזני in the audience העם of the people; ונמצא and therein was found כתוב written, בו אשׁר that לא should not יבוא come עמני the Ammonite ומאבי and the Moabite בקהל into the congregation האלהים of God עד forever; עולם׃ forever;}%
\verse{כי Because לא not קדמו they met את בני the children ישׂראל of Israel בלחם with bread ובמים and with water, וישׂכר but hired עליו against את בלעם Balaam לקללו them, that he should curse ויהפך turned אלהינו them: howbeit our God הקללה the curse לברכה׃ into a blessing.}%
\verse{ויהי Now it came to pass, כשׁמעם when they had heard את התורה the law, ויבדילו that they separated כל all ערב the mixed multitude. מישׂראל׃}%
\verseWithHeading{Nehemiah Brings Reform}{ולפני And before מזה this, אלישׁיב Eliashib הכהן the priest, נתון having the oversight בלשׁכת of the chamber בית of the house אלהינו of our God, קרוב allied לטוביה׃ unto Tobiah:}%
\verse{ויעשׂ And he had prepared לו לשׁכה chamber, גדולה for him a great ושׁם where היו formerly לפנים formerly נתנים they laid את המנחה the meat offerings, הלבונה the frankincense, והכלים and the vessels, ומעשׂר and the tithes הדגן of the corn, התירושׁ the new wine, והיצהר and the oil, מצות which was commanded הלוים to the Levites, והמשׁררים and the singers, והשׁערים and the porters; ותרומת and the offerings הכהנים׃ of the priests.}%
\verse{ובכל But in all זה this לא not הייתי was בירושׁלם I at Jerusalem: כי for בשׁנת year שׁלשׁים and thirtieth ושׁתים in the two לארתחשׁסתא of Artaxerxes מלך king בבל of Babylon באתי came אל I unto המלך the king, ולקץ and after ימים certain days נשׁאלתי obtained I leave מן of המלך׃ the king:}%
\verse{ואבוא And I came לירושׁלם to Jerusalem, ואבינה and understood ברעה of the evil אשׁר that עשׂה did אלישׁיב Eliashib לטוביה for Tobiah, לעשׂות in preparing לו נשׁכה him a chamber בחצרי in the courts בית of the house האלהים׃ of God.}%
\verse{וירע And it grieved לי מאד me sore: ואשׁליכה therefore I cast forth את כל all כלי stuff בית the household טוביה of Tobiah החוץ out מן of הלשׁכה׃ the chamber.}%
\verse{ואמרה Then I commanded, ויטהרו and they cleansed הלשׁכות the chambers: ואשׁיבה brought I again שׁם and thither כלי the vessels בית of the house האלהים of God, את with המנחה the meat offering והלבונה׃ and the frankincense.}%
\verse{ואדעה And I perceived כי that מניות the portions הלוים of the Levites לא had not נתנה been given ויברחו were fled אישׁ every one לשׂדהו to his field. הלוים for the Levites והמשׁררים and the singers, עשׂי that did המלאכה׃ the work,}%
\verse{ואריבה Then contended את I with הסגנים the rulers, ואמרה and said, מדוע Why נעזב forsaken? בית is the house האלהים of God ואקבצם And I gathered them together, ואעמדם and set על them in עמדם׃ their place.}%
\verse{וכל all יהודה Judah הביאו Then brought מעשׂר the tithe הדגן of the corn והתירושׁ and the new wine והיצהר and the oil לאוצרות׃ unto the treasuries.}%
\verse{ואוצרה And I made treasurers על over אוצרות the treasuries, שׁלמיה Shelemiah הכהן the priest, וצדוק and Zadok הסופר the scribe, ופדיה Pedaiah: מן and of הלוים the Levites, ועל and next to ידם and next to חנן them Hanan בן the son זכור of Zaccur, בן the son מתניה of Mattaniah: כי for נאמנים faithful, נחשׁבו they were counted ועליהם and their office לחלק to distribute לאחיהם׃ unto their brethren.}%
\verse{זכרה Remember לי אלהי me, O my God, על concerning זאת this, ואל and wipe not out תמח and wipe not out חסדי my good deeds אשׁר that עשׂיתי I have done בבית for the house אלהי of my God, ובמשׁמריו׃ and for the offices}%
\verseWithHeading{Nehemiah Begins Sabbath Reforms}{בימים days ההמה In those ראיתי saw ביהודה I in Judah דרכים treading גתות wine presses בשׁבת on the sabbath, ומביאים and bringing in הערמות sheaves, ועמסים and lading על and lading החמרים asses; ואף as also יין wine, ענבים grapes, ותאנים and figs, וכל and all משׂא burdens, ומביאים which they brought into ירושׁלם Jerusalem ביום day: השׁבת on the sabbath ואעיד and I testified ביום in the day מכרם wherein they sold ציד׃ victuals.}%
\verse{והצרים men of Tyre ישׁבו There dwelt בה מביאים also therein, which brought דאג fish, וכל and all מכר manner of ware, ומכרים and sold בשׁבת on the sabbath לבני unto the children יהודה of Judah, ובירושׁלם׃ and in Jerusalem.}%
\verse{ואריבה Then I contended את with חרי the nobles יהודה of Judah, ואמרה and said להם מה unto them, What הדבר thing הרע evil הזה this אשׁר that אתם ye עשׂים do, ומחללים and profane את יום day? השׁבת׃ the sabbath}%
\verse{הלוא not כה thus, עשׂו Did אבתיכם your fathers ויבא bring אלהינו and did not our God עלינו upon את כל all הרעה evil הזאת this ועל us, and upon העיר city? הזאת this ואתם yet ye מוסיפים חרון wrath על upon ישׂראל Israel לחלל by profaning את השׁבת׃ the sabbath.}%
\verse{ויהי And it came to pass, כאשׁר that when צללו began to be dark שׁערי the gates ירושׁלם of Jerusalem לפני before השׁבת the sabbath, ואמרה I commanded ויסגרו should be shut, הדלתות that the gates ואמרה and charged אשׁר that לא they should not יפתחום be opened עד till אחר after השׁבת the sabbath: ומנערי and of my servants העמדתי set על I at השׁערים the gates, לא there should no יבוא be brought in משׂא burden ביום day. השׁבת׃ on the sabbath}%
\verse{וילינו lodged הרכלים So the merchants ומכרי and sellers כל of all ממכר kind of ware מחוץ without לירושׁלם Jerusalem פעם once ושׁתים׃ or twice.}%
\verse{ואעידה Then I testified בהם ואמרה against them, and said אליהם unto מדוע them, Why אתם ye לנים lodge נגד about החומה the wall? אם if תשׁנו ye do again, יד hands אשׁלח I will lay בכם מן on you. From העת time ההיא that לא they no באו forth came בשׁבת׃ on the sabbath.}%
\verse{ואמרה And I commanded ללוים the Levites אשׁר that יהיו they should מטהרים cleanse themselves, ובאים and they should come שׁמרים keep השׁערים the gates, לקדשׁ to sanctify את יום day. השׁבת the sabbath גם also, זאת this זכרה Remember לי אלהי me, O my God, וחוסה and spare עלי and spare כרב me according to the greatness חסדך׃ of thy mercy.}%
\verseWithHeading{Mixed Marriages are Condemned}{גם also בימים days ההם In those ראיתי saw את היהודים I Jews השׁיבו had married נשׁים wives אשׁדודיות of Ashdod, עמוניות of Ammon, מואביות׃ of Moab:}%
\verse{ובניהם And their children חצי half מדבר spoke אשׁדודית in the speech of Ashdod, ואינם not מכירים and could לדבר speak יהודית in the Jews' language, וכלשׁון but according to the language עם of each ועם׃ people.}%
\verse{ואריב And I contended עמם with ואקללם them, and cursed ואכה them, and smote מהם אנשׁים certain ואמרטם them, and plucked off their hair, ואשׁביעם and made them swear באלהים by God, אם Ye shall not תתנו give בנתיכם your daughters לבניהם unto their sons, ואם nor תשׂאו take מבנתיהם their daughters לבניכם׃ unto your sons,}%
\verse{הלוא Did not על by אלה these things? חטא sin שׁלמה Solomon מלך king ישׂראל of Israel ובגוים nations הרבים yet among many לא there no היה was מלך king כמהו like him, ואהוב beloved לאלהיו of his God, היה who was ויתנהו made אלהים and God מלך him king על over כל all ישׂראל Israel: גם nevertheless אותו החטיאו cause to sin. הנשׁים women הנכריות׃ even him did outlandish}%
\verse{ולכם הנשׁמע Shall we then hearken לעשׂת unto you to do את כל all הרעה evil, הגדולה great הזאת this למעל to transgress באלהינו against our God להשׁיב in marrying נשׁים wives? נכריות׃ strange}%
\verse{ומבני the son יוידע of Joiada, בן אלישׁיב of Eliashib הכהן priest, הגדול the high חתן son in law לסנבלט to Sanballat החרני the Horonite: ואבריחהו therefore I chased מעלי׃ him from}%
\verse{זכרה Remember להם אלהי them, O my God, על because גאלי they have defiled הכהנה the priesthood, וברית and the covenant הכהנה of the priesthood, והלוים׃ and of the Levites.}%
\verse{וטהרתים Thus cleansed מכל נכר strangers, ואעמידה and appointed משׁמרות the wards לכהנים of the priests וללוים and the Levites, אישׁ every one במלאכתו׃ in his business;}%
\verse{ולקרבן offering, העצים And for the wood בעתים at times מזמנות appointed, ולבכורים and for the firstfruits. זכרה Remember לי אלהי me, O my God, לטובה׃ for good.}%
\end{biblechapter}%
