\biblebook{Obadiah}
\begin{biblechapter}% Obadiah 1
\verseWithHeading{God’s Judgment on Edom}{חזון The vision עבדיה of Obadiah. כה Thus אמר saith אדני the Lord יהוה the LORD, לאדום concerning Edom; שׁמועה a rumor שׁמענו We have heard מאת יהוה GOD וציר and an ambassador בגוים among the heathen, שׁלח is sent קומו Arise ונקומה ye, and let us rise up עליה against למלחמה׃ her in battle.}%
\verse{הנה Behold, קטן thee small נתתיך I have made בגוים among the heathen: בזוי despised. אתה thou מאד׃ art greatly}%
\verse{זדון The pride לבך of thine heart השׁיאך hath deceived שׁכני thee, thou that dwellest בחגוי in the clefts סלע of the rock, מרום high; שׁבתו אמר that saith בלבו in his heart, מי Who יורדני shall bring me down ארץ׃ to the ground?}%
\verse{אם Though תגביה thou exalt כנשׁר as the eagle, ואם and though בין among כוכבים the stars, שׂים thou set קנך thy nest משׁם thence אורידך will I bring thee down, נאם saith יהוה׃ the LORD.}%
\verse{אם If גנבים thieves באו came לך אם to thee, if שׁודדי robbers לילה by night, איך (how נדמיתה art thou cut off!) הלוא would they not יגנבו have stolen דים till they had enough? אם if בצרים the grape gatherers באו came לך הלוא to thee, would they not ישׁאירו leave עללות׃ grapes?}%
\verse{איך How נחפשׂו searched out! עשׂו are of Esau נבעו sought up! מצפניו׃ are his hidden things}%
\verse{עד thee to הגבול the border: שׁלחוך have brought כל All אנשׁי בריתך of thy confederacy השׁיאוך with thee have deceived יכלו thee, prevailed לך אנשׁי שׁלמך that were at peace לחמך against thee; thy bread ישׂימו have laid מזור a wound תחתיך under אין thee: none תבונה׃ understanding}%
\verse{הלוא Shall I not ביום day, ההוא in that נאם saith יהוה the LORD, והאבדתי even destroy חכמים the wise מאדום ותבונה and understanding מהר out of the mount עשׂו׃ of Esau?}%
\verse{וחתו shall be dismayed, גבוריך And thy mighty תימן O Teman, למען to the end that יכרת may be cut off אישׁ every one מהר of the mount עשׂו of Esau מקטל׃ by slaughter.}%
\verseWithHeading{Edom’s Treachery against Judah}{מחמס אחיך against thy brother יעקב Jacob תכסך shall cover בושׁה shame ונכרת thee, and thou shalt be cut off לעולם׃ forever.}%
\verse{ביום In the day עמדך that thou stoodest מנגד on the other side, ביום in the day שׁבות carried away captive זרים that the strangers חילו his forces, ונכרים and foreigners באו entered into שׁערו his gates, ועל upon ירושׁלם Jerusalem, ידו and cast גורל lots גם even אתה thou כאחד as one מהם׃}%
\verse{ואל But thou shouldest not תרא have looked ביום on the day אחיך of thy brother ביום in the day נכרו that he became a stranger; ואל neither תשׂמח shouldest thou have rejoiced לבני over the children יהודה of Judah ביום in the day אבדם of their destruction; ואל neither תגדל shouldest thou have spoken proudly פיך shouldest thou have spoken proudly ביום in the day צרה׃ of distress.}%
\verse{אל Thou shouldest not תבוא have entered בשׁער into the gate עמי of my people ביום in the day אידם of their calamity; אל shouldest not תרא have looked גם yea, אתה thou ברעתו on their affliction ביום in the day אידו of their calamity, ואל nor תשׁלחנה have laid בחילו on their substance ביום in the day אידו׃ of their calamity;}%
\verse{ואל Neither תעמד shouldest thou have stood על in הפרק the crossway, להכרית to cut off את פליטיו those of his that did escape; ואל neither תסגר shouldest thou have delivered up שׂרידיו those of his that did remain ביום in the day צרה׃ of distress.}%
\verseWithHeading{Future Judgment of the Nations}{כי For קרוב near יום the day יהוה of the LORD על upon כל all הגוים the heathen: כאשׁר as עשׂית thou hast done, יעשׂה it shall be done לך גמלך unto thee: thy reward ישׁוב shall return בראשׁך׃ upon thine own head.}%
\verse{כי For כאשׁר as שׁתיתם ye have drunk על upon הר mountain, קדשׁי my holy ישׁתו drink כל shall all הגוים the heathen תמיד continually, ושׁתו yea, they shall drink, ולעו and they shall swallow down, והיו and they shall be כלוא as though they had not היו׃ been.}%
\verse{ובהר But upon mount ציון Zion תהיה shall be פליטה deliverance, והיה and there shall be קדשׁ holiness; וירשׁו shall possess בית and the house יעקב of Jacob את מורשׁיהם׃ their possessions.}%
\verse{והיה shall be בית And the house יעקב of Jacob אשׁ a fire, ובית and the house יוסף of Joseph להבה a flame, ובית and the house עשׂו of Esau לקשׁ for stubble, ודלקו and they shall kindle בהם ואכלום in them, and devour ולא them; and there shall not יהיה be שׂריד remaining לבית of the house עשׂו of Esau; כי for יהוה the LORD דבר׃ hath spoken}%
\verseWithHeading{Future Blessing for Israel}{וירשׁו shall possess הנגב And the south את הר the mount עשׂו of Esau; והשׁפלה and the plain את פלשׁתים the Philistines: וירשׁו and they shall possess את שׂדה the fields אפרים of Ephraim, ואת שׂדה and the fields שׁמרון of Samaria: ובנימן and Benjamin את הגלעד׃ Gilead.}%
\verse{וגלת And the captivity החל host הזה of this לבני of the children ישׂראל of Israel אשׁר that כנענים of the Canaanites, עד unto צרפת Zarephath; וגלת and the captivity ירושׁלם of Jerusalem, אשׁר which בספרד in Sepharad, ירשׁו shall possess את ערי the cities הנגב׃ of the south.}%
\verse{ועלו shall come up מושׁעים And saviors בהר on mount ציון Zion לשׁפט to judge את הר the mount עשׂו of Esau; והיתה shall be ליהוה the LORD's. המלוכה׃ and the kingdom}%
\end{biblechapter}%
