\biblebook{Proverbs}
\begin{biblechapter}% Proverbs 1
\verseWithHeading{Prologue}{משׁלי The proverbs שׁלמה of Solomon בן the son דוד of David, מלך king ישׂראל׃ of Israel;}%
\verse{לדעת To know חכמה wisdom ומוסר and instruction; להבין to perceive אמרי the words בינה׃ of understanding;}%
\verse{לקחת To receive מוסר the instruction השׂכל of wisdom, צדק justice, ומשׁפט and judgment, ומישׁרים׃ and equity;}%
\verse{לתת To give לפתאים to the simple, ערמה subtlety לנער to the young man דעת knowledge ומזמה׃ and discretion.}%
\verse{ישׁמע will hear, חכם A wise ויוסף and will increase לקח learning; ונבון and a man of understanding תחבלות unto wise counsels: יקנה׃ shall attain}%
\verse{להבין To understand משׁל a proverb, ומליצה and the interpretation; דברי the words חכמים of the wise, וחידתם׃ and their dark sayings.}%
\verse{יראת The fear יהוה of the LORD ראשׁית the beginning דעת of knowledge: חכמה wisdom ומוסר and instruction. אוילים fools בזו׃ despise}%
\verseWithHeading{Lecture Against Gang Behavior}{שׁמע hear בני My son, מוסר the instruction אביך of thy father, ואל not תטשׁ and forsake תורת the law אמך׃ of thy mother:}%
\verse{כי For לוית an ornament חן of grace הם they לראשׁך unto thy head, וענקים and chains לגרגרתיך׃ about thy neck.}%
\verse{בני My son, אם if יפתוך entice חטאים sinners אל thou not. תבא׃ thee, consent}%
\verse{אם If יאמרו they say, לכה Come אתנו with נארבה us, let us lay wait לדם for blood, נצפנה let us lurk privily לנקי for the innocent חנם׃ without cause:}%
\verse{נבלעם Let us swallow them up כשׁאול as the grave; חיים alive ותמימים and whole, כיורדי as those that go down בור׃ into the pit:}%
\verse{כל all הון substance, יקר precious נמצא We shall find נמלא we shall fill בתינו our houses שׁלל׃ with spoil:}%
\verse{גורלך thy lot תפיל Cast in בתוכנו among כיס purse: אחד one יהיה have לכלנו׃ us; let us all}%
\verse{בני My son, אל not תלך walk בדרך thou in the way אתם with מנע them; refrain רגלך thy foot מנתיבתם׃ from their path:}%
\verse{כי For רגליהם their feet לרע to evil, ירוצו run וימהרו and make haste לשׁפך to shed דם׃ blood.}%
\verse{כי Surely חנם in vain מזרה is spread הרשׁת the net בעיני in the sight כל of any בעל bird. כנף׃ bird.}%
\verse{והם And they לדמם for their blood; יארבו lay wait יצפנו they lurk privily לנפשׁתם׃ for their lives.}%
\verse{כן So ארחות the ways כל of every one בצע that is greedy בצע of gain; את נפשׁ the life בעליו of the owners יקח׃ taketh away}%
\verseWithHeading{The Call of Lady Wisdom}{חכמות Wisdom בחוץ without; תרנה crieth ברחבות in the streets: תתן she uttereth קולה׃ her voice}%
\verse{בראשׁ in the chief המיות place of concourse, תקרא She crieth בפתחי in the openings שׁערים of the gates: בעיר in the city אמריה her words, תאמר׃ she uttereth}%
\verse{עד מתיתים ye simple ones, תאהבו will ye love פתי simplicity? ולצים and the scorners לצון in their scorning, חמדו delight להם וכסילים and fools ישׂנאו hate דעת׃ knowledge?}%
\verse{תשׁובו Turn לתוכחתי you at my reproof: הנה behold, אביעה I will pour out לכם רוחי my spirit אודיעה unto you, I will make known דברי my words אתכם׃}%
\verse{יען Because קראתי I have called, ותמאנו and ye refused; נטיתי I have stretched out ידי my hand, ואין and no man מקשׁיב׃ regarded;}%
\verse{ותפרעו But ye have set at naught כל all עצתי my counsel, ותוכחתי of my reproof: לא none אביתם׃ and would}%
\verse{גם also אני I באידכם at your calamity; אשׂחק will laugh אלעג I will mock בבא cometh; פחדכם׃ when your fear}%
\verse{בבא cometh כשׁאוה פחדכם When your fear ואידכם and your destruction כסופה as a whirlwind; יאתה cometh בבא cometh עליכם upon צרה when distress וצוקה׃ and anguish}%
\verse{אז Then יקראנני shall they call upon ולא me, but I will not אענה answer; ישׁחרנני they shall seek me early, ולא but they shall not ימצאנני׃ find}%
\verse{תחת כיׂנאו they hated דעת knowledge, ויראת the fear יהוה of the LORD: לא and did not בחרו׃ choose}%
\verse{לא none אבו They would לעצתי of my counsel: נאצו they despised כל all תוכחתי׃ my reproof.}%
\verse{ויאכלו Therefore shall they eat מפרי of the fruit דרכם of their own way, וממעצתיהם with their own devices. ישׂבעו׃ and be filled}%
\verse{כי For משׁובת the turning away פתים of the simple תהרגם shall slay ושׁלות them, and the prosperity כסילים of fools תאבדם׃ shall destroy}%
\verse{ושׁמע But whoso hearkeneth לי ישׁכן unto me shall dwell בטח safely, ושׁאנן and shall be quiet מפחד from fear רעה׃ of evil.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 2
\verseWithHeading{The Benefits of Wisdom}{בני My son, אם if תקח thou wilt receive אמרי my words, ומצותי my commandments תצפן and hide אתך׃ with}%
\verse{להקשׁיב So that thou incline לחכמה unto wisdom, אזנך thine ear תטה apply לבך thine heart לתבונה׃ to understanding;}%
\verse{כי Yea, אם if לבינה after knowledge, תקרא thou criest לתבונה for understanding; תתן liftest up קולך׃ thy voice}%
\verse{אם If תבקשׁנה thou seekest ככסף her as silver, וכמטמונים for her as hid treasures; תחפשׂנה׃ and searchest}%
\verse{אז Then תבין shalt thou understand יראת the fear יהוה of the LORD, ודעת the knowledge אלהים of God. תמצא׃ and find}%
\verse{כי For יהוה the LORD יתן giveth חכמה wisdom: מפיו out of his mouth דעת knowledge ותבונה׃ and understanding.}%
\verse{וצפן He layeth up לישׁרים for the righteous: תושׁיה sound wisdom מגן a buckler להלכי to them that walk תם׃ uprightly.}%
\verse{לנצר He keepeth ארחות the paths משׁפט of judgment, ודרך the way חסידו of his saints. ישׁמר׃ and preserveth}%
\verse{אז Then תבין shalt thou understand צדק righteousness, ומשׁפט and judgment, ומישׁרים and equity; כל every מעגל path. טוב׃ good}%
\verse{כי When תבוא entereth חכמה wisdom בלבך into thine heart, ודעת and knowledge לנפשׁך unto thy soul; ינעם׃ is pleasant}%
\verse{מזמה Discretion תשׁמר shall preserve עליך shall preserve תבונה thee, understanding תנצרכה׃ shall keep}%
\verse{להצילך To deliver מדרך thee from the way רע of the evil מאישׁ , from the man מדבר that speaketh תהפכות׃ froward things;}%
\verse{העזבים Who leave ארחות the paths ישׁר of uprightness, ללכת to walk בדרכי in the ways חשׁך׃ of darkness;}%
\verse{השׂמחים Who rejoice לעשׂות to do רע evil, יגילו delight in בתהפכות the frowardness רע׃ of the wicked;}%
\verse{אשׁר Whose ארחתיהם ways עקשׁים crooked, ונלוזים and froward במעגלותם׃ in their paths:}%
\verse{להצילך To deliver מאשׁה זרהנכריה from the stranger אמריה with her words; החליקה׃ flattereth}%
\verse{העזבת Which forsaketh אלוף the guide נעוריה of her youth, ואת ברית the covenant אלהיה of her God. שׁכחה׃ and forgetteth}%
\verse{כי For שׁחה inclineth אל unto מות death, ביתה her house ואל unto רפאים the dead. מעגלתיה׃ and her paths}%
\verse{כל באיה that go unto לא neither ישׁובון her return again, ולא ישׂיגו take they hold ארחות of the paths חיים׃ of life.}%
\verse{למען That תלך thou mayest walk בדרך in the way טובים of good וארחות the paths צדיקים of the righteous. תשׁמר׃ and keep}%
\verse{כי For ישׁרים the upright ישׁכנו shall dwell in ארץ the land, ותמימים and the perfect יותרו׃ shall remain}%
\verse{ורשׁעים But the wicked מארץ from the earth, יכרתו shall be cut off ובוגדים and the transgressors יסחו shall be rooted ממנה׃ from the earth,}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 3
\verseWithHeading{Wisdom and Piety}{בני My son, תורתי my law; אל not תשׁכח forget ומצותי my commandments: יצר keep לבך׃ but let thine heart}%
\verse{כי For ארך length ימים of days, ושׁנות and long חיים life, ושׁלום and peace, יוסיפו׃ shall they add}%
\verse{חסד mercy ואמת and truth אל Let not יעזבך forsake קשׁרם thee: bind על them about גרגרותיך thy neck; כתבם write על them upon לוח the table לבך׃ of thine heart:}%
\verse{ומצא So shalt thou find חן favor ושׂכל understanding טוב and good בעיני in the sight אלהים of God ואדם׃ and man.}%
\verse{בטח Trust אל in יהוה the LORD בכל with all לבך thine heart; ואל unto בינתך thine own understanding. אל not תשׁען׃ and lean}%
\verse{בכל In all דרכיך thy ways דעהו acknowledge והוא him, and he יישׁר shall direct ארחתיך׃ thy paths.}%
\verse{אל not תהי Be חכם wise בעיניך in thine own eyes: ירא fear את יהוהסור and depart מרע׃ from evil.}%
\verse{רפאות health תהי It shall be לשׁרך to thy navel, ושׁקוי and marrow לעצמותיך׃ to thy bones.}%
\verse{כבד Honor את יהוה the LORD מהונך with thy substance, ומראשׁית and with the firstfruits כל of all תבואתך׃ thine increase:}%
\verse{וימלאו be filled אסמיך So shall thy barns שׂבע with plenty, ותירושׁ with new wine. יקביך and thy presses יפרצו׃ shall burst out}%
\verseWithHeading{Happy Is the One Who Finds Wisdom}{מוסר the chastening יהוה of the LORD; בני My son, אל not תמאס despise ואל neither תקץ be weary בתוכחתו׃ of his correction:}%
\verse{כי For את אשׁר whom יאהב loveth יהוה the LORD יוכיח he correcteth; וכאב even as a father את בן the son ירצה׃ he delighteth.}%
\verse{אשׁרי Happy אדם the man מצא findeth חכמה wisdom, ואדם and the man יפיק getteth תבונה׃ understanding.}%
\verse{כי For טוב of it better סחרה the merchandise מסחר than the merchandise כסף of silver, ומחרוץ thereof than fine gold. תבואתה׃ and the gain}%
\verse{יקרה more precious היא She מפניים than rubies: וכל and all חפציך the things thou canst desire לא are not ישׁוו׃ to be compared}%
\verse{ארך Length ימים of days בימינה in her right hand; בשׂמאולה in her left hand עשׁר riches וכבוד׃ and honor.}%
\verse{דרכיה Her ways דרכי ways נעם of pleasantness, וכל and all נתיבותיה her paths שׁלום׃ peace.}%
\verse{עץ a tree חיים of life היא She למחזיקים to them that lay hold בה ותמכיה that retaineth מאשׁר׃ upon her: and happy}%
\verseWithHeading{Role of Wisdom in Creation and Society}{יהוה The LORD בחכמה by wisdom יסד hath founded ארץ the earth; כונן hath he established שׁמים the heavens. בתבונה׃ by understanding}%
\verse{בדעתו By his knowledge תהומות the depths נבקעו are broken up, ושׁחקים and the clouds ירעפו drop down טל׃ the dew.}%
\verse{בני My son, אל let not ילזו them depart מעיניך from thine eyes: נצר keep תשׁיה sound wisdom ומזמה׃ and discretion:}%
\verse{ויהיו So shall they be חיים life לנפשׁך unto thy soul, וחן and grace לגרגרתיך׃ to thy neck.}%
\verse{אז Then תלך shalt thou walk לבטח safely, דרכך in thy way ורגלך and thy foot לא shall not תגוף׃ stumble.}%
\verse{אם When תשׁכב thou liest down, לא thou shalt not תפחד be afraid: ושׁכבת yea, thou shalt lie down, וערבה shall be sweet. שׁנתך׃ and thy sleep}%
\verse{אל Be not תירא afraid מפחד fear, פתאם of sudden ומשׁאת neither of the desolation רשׁעים of the wicked, כי when תבא׃ it cometh.}%
\verse{כי For יהוה the LORD יהיה shall be בכסלך thy confidence, ושׁמר and shall keep רגלך thy foot מלכד׃ from being taken.}%
\verse{אל not תמנע Withhold טוב good מבעליו from them to whom it is due, בהיות when it is לאל in the power ידיך of thine hand לעשׂות׃ to do}%
\verse{אל not תאמר Say לרעיך unto thy neighbor, לך Go, ושׁוב and come again, ומחר and tomorrow אתן I will give; וישׁ when thou hast אתך׃ it by}%
\verse{אל not תחרשׁ evil על against רעך thy neighbor, רעה Devise והוא seeing he יושׁב dwelleth לבטח securely אתך׃ by}%
\verse{אל not תרוב Strive עם with אדם a man חנם without cause, אם if לא thee no גמלך he have done רעה׃ harm.}%
\verse{אל thou not תקנא Envy באישׁ the oppressor, חמס the oppressor, ואל none תבחר and choose בכל none דרכיו׃ of his ways.}%
\verse{כי For תועבת abomination יהוה to the LORD: נלוז the froward ואת ישׁרים the righteous. סודו׃ but his secret}%
\verse{מארת The curse יהוה of the LORD בבית in the house רשׁע of the wicked: ונוה the habitation צדיקים of the just. יברך׃ but he blesseth}%
\verse{אם Surely ללצים scorneth הוא he יליץ the scorners: ולעניים יתן but he giveth חן׃ grace}%
\verse{כבוד glory: חכמים The wise ינחלו shall inherit וכסילים of fools. מרים shall be the promotion קלון׃ but shame}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 4
\verseWithHeading{The Father’s Wisdom}{שׁמעו Hear, בנים ye children, מוסר the instruction אב of a father, והקשׁיבו and attend לדעת to know בינה׃ understanding.}%
\verse{כי For לקח doctrine, טוב you good נתתי I give לכם תורתי my law. אל ye not תעזבו׃ forsake}%
\verse{כי For בן son, הייתי I was לאבי my father's רך tender ויחיד and only לפני in the sight אמי׃ of my mother.}%
\verse{וירני He taught ויאמר me also, and said לי יתמך retain דברי my words: לבך unto me, Let thine heart שׁמר keep מצותי my commandments, וחיה׃ and live.}%
\verse{קנה Get חכמה wisdom, קנה get בינה understanding: אל not; תשׁכח forget ואל neither תט decline מאמרי from the words פי׃ of my mouth.}%
\verse{אל her not, תעזבה Forsake ותשׁמרך and she shall preserve אהבה thee: love ותצרך׃ her, and she shall keep}%
\verse{ראשׁית the principal thing; חכמה Wisdom קנה get חכמה wisdom: ובכל and with all קנינך thy getting קנה get בינה׃ understanding.}%
\verse{סלסלה Exalt ותרוממך her, and she shall promote תכבדך thee: she shall bring thee to honor, כי when תחבקנה׃ thou dost embrace}%
\verse{תתן She shall give לראשׁך to thine head לוית an ornament חן of grace: עטרת a crown תפארת of glory תמגנך׃ shall she deliver}%
\verseWithHeading{The Right Path}{שׁמע Hear, בני O my son, וקח and receive אמרי my sayings; וירבו shall be many. לך שׁנות and the years חיים׃ of thy life}%
\verse{בדרך thee in the way חכמה of wisdom; הרתיך I have taught הדרכתיך I have led במעגלי paths. ישׁר׃ thee in right}%
\verse{בלכתך When thou goest, לא shall not יצר be straitened; צעדך thy steps ואם and when תרוץ thou runnest, לא thou shalt not תכשׁל׃ stumble.}%
\verse{החזק Take fast hold במוסר of instruction; אל let not תרף go: נצרה keep כי her; for היא she חייך׃ thy life.}%
\verse{בארח into the path רשׁעים of the wicked, אל not תבא Enter ואל not תאשׁר and go בדרך in the way רעים׃ of evil}%
\verse{פרעהו Avoid אל not תעבר it, pass בו שׂטה by it, turn מעליו from ועבור׃ it, and pass away.}%
\verse{כי For לא not, ישׁנו they sleep אם except לא except ירעו they have done mischief; ונגזלה is taken away, שׁנתם and their sleep אם unless לא unless יכשׁולו׃ they cause to fall.}%
\verse{כי For לחמו they eat לחם the bread רשׁע of wickedness, ויין the wine חמסים of violence. ישׁתו׃ and drink}%
\verse{וארח But the path צדיקים of the just כאור light, נגה as the shining הולך more and more ואור that shineth עד unto נכון the perfect היום׃ day.}%
\verse{דרך The way רשׁעים of the wicked כאפלה as darkness: לא not ידעו they know במה at what יכשׁלו׃ they stumble.}%
\verseWithHeading{Staying the Course}{בני My son, לדברי to my words; הקשׁיבה attend לאמרי unto my sayings. הט incline אזנך׃ thine ear}%
\verse{אל Let them not יליזו depart מעיניך from thine eyes; שׁמרם keep בתוך them in the midst לבבך׃ of thine heart.}%
\verse{כי For חיים life הם they למצאיהם unto those that find ולכל to all בשׂרו their flesh. מרפא׃ them, and health}%
\verse{מכל with all משׁמר diligence; נצר Keep לבך thy heart כי for ממנו with all תוצאות it the issues חיים׃ of life.}%
\verse{הסר Put away ממך from עקשׁות thee a froward פה mouth, ולזות and perverse שׂפתים lips הרחק put far ממך׃ from}%
\verse{עיניך Let thine eyes לנכח right on, יביטו look ועפעפיך and let thine eyelids יישׁרו look straight נגדך׃ before}%
\verse{פלס Ponder מעגל the path רגלך of thy feet, וכל and let all דרכיך thy ways יכנו׃ be established.}%
\verse{אל not תט Turn ימין to the right hand ושׂמאול nor to the left: הסר remove רגלך thy foot מרע׃ from evil.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 5
\verse{בני My son, לחכמתי unto my wisdom, הקשׁיבה attend לתבונתי to my understanding: הט bow אזנך׃ thine ear}%
\verse{לשׁמר That thou mayest regard מזמות discretion, ודעת knowledge. שׂפתיך and thy lips ינצרו׃ may keep}%
\verse{כי For נפת a honeycomb, תטפנה drop שׂפתי the lips זרה of a strange woman וחלק smoother משׁמן than oil: חכה׃ and her mouth}%
\verse{ואחריתה But her end מרה is bitter כלענה as wormwood, חדה sharp כחרב sword. פיות׃ as a twoedged}%
\verse{רגליה Her feet ירדות go down מות to death; שׁאול on hell. צעדיה her steps יתמכו׃ take hold}%
\verse{ארח the path חיים of life, פן Lest תפלס thou shouldest ponder נעו are movable, מעגלתיה her ways לא thou canst not תדע׃ know}%
\verseWithHeading{Do Not Commit Adultery Against Wisdom}{ועתה me now בנים therefore, O ye children, שׁמעו Hear לי ואל not תסורו and depart מאמרי from the words פי׃ of my mouth.}%
\verse{הרחק מעליה from דרכך ואלקרב her, and come not nigh אל פתח the door ביתה׃ of her house:}%
\verse{פן Lest תתן thou give לאחרים unto others, הודך thine honor ושׁנתיך and thy years לאכזרי׃ unto the cruel:}%
\verse{פן Lest ישׂבעו be filled זרים strangers כחך with thy wealth; ועצביך and thy labors בבית in the house נכרי׃ of a stranger;}%
\verse{ונהמת And thou mourn באחריתך at the last, בכלות are consumed, בשׂרך when thy flesh ושׁארך׃ and thy body}%
\verse{ואמרת And say, איך How שׂנאתי have I hated מוסר instruction, ותוכחת reproof; נאץ despised לבי׃ and my heart}%
\verse{ולא And have not שׁמעתי obeyed בקול the voice מורי of my teachers, ולמלמדי to them that instructed לא nor הטיתי inclined אזני׃ mine ear}%
\verse{כמעט almost הייתי I was בכל in all רע evil בתוך in the midst קהל of the congregation ועדה׃ and assembly.}%
\verse{שׁתה Drink מים waters מבורך out of thine own cistern, ונזלים and running waters מתוך out of בארך׃ thine own well.}%
\verse{יפוצו be dispersed מעינתיך Let thy fountains חוצה abroad, ברחבות in the streets. פלגי rivers מים׃ of waters}%
\verse{יהיו Let them be לך לבדך only ואין thine own, and not לזרים strangers' אתך׃ with}%
\verse{יהי be מקורך Let thy fountain ברוך blessed: ושׂמח and rejoice מאשׁת with the wife נעורך׃ of thy youth.}%
\verse{אילת hind אהבים the loving ויעלת roe; חן and pleasant דדיה let her breasts ירוך satisfy בכל thee at all עת times; באהבתה with her love. תשׁגה and be thou ravished תמיד׃ always}%
\verse{ולמה And why תשׁגה be ravished בני wilt thou, my son, בזרה with a strange woman, ותחבק and embrace חק the bosom נכריה׃ of a stranger?}%
\verse{כי For נכח before עיני the eyes יהוה דרכי the ways אישׁ of man וכל all מעגלתיו his goings. מפלס׃ and he pondereth}%
\verse{עוונותיו His own iniquities ילכדנו shall take את הרשׁע the wicked ובחבלי himself, and he shall be holden חטאתו of his sins. יתמך׃ with the cords}%
\verse{הוא He ימות shall die באין without מוסר instruction; וברב and in the greatness אולתו of his folly ישׁגה׃ he shall go astray.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 6
\verseWithHeading{Against Pledges}{בני My son, אם if ערבת thou be surety לרעך for thy friend, תקעת thou hast stricken לזר with a stranger, כפיך׃ thy hand}%
\verse{נוקשׁת Thou art snared באמרי with the words פיך of thy mouth, נלכדת thou art taken באמרי with the words פיך׃ of thy mouth.}%
\verse{עשׂה Do זאת this אפוא now, בני my son, והנצל and deliver thyself, כי when באת thou art come בכף into the hand רעך of thy friend; לך go, התרפס humble thyself, ורהב and make sure רעיך׃ thy friend.}%
\verse{אל not תתן Give שׁנה sleep לעיניך to thine eyes, ותנומה nor slumber לעפעפיך׃ to thine eyelids.}%
\verse{הנצל Deliver thyself כצבי as a roe מיד from the hand וכצפור and as a bird מיד from the hand יקושׁ׃ of the fowler.}%
\verseWithHeading{Against Sloth}{לך Go אל to נמלה the ant, עצל thou sluggard; ראה consider דרכיה her ways, וחכם׃ and be wise:}%
\verse{אשׁר Which אין having no לה קצין guide, שׁטר overseer, ומשׁל׃ or ruler,}%
\verse{תכין Provideth בקיץ in the summer, לחמה her meat אגרה gathereth בקציר in the harvest. מאכלה׃ her food}%
\verse{עד מתי when עצל O sluggard? תשׁכב wilt thou sleep, מתי תקום wilt thou arise משׁנתך׃ out of thy sleep?}%
\verse{מעט a little שׁנות sleep, מעט a little תנומות slumber, מעט a little חבק folding ידים of the hands לשׁכב׃ to sleep:}%
\verse{ובא come כמהלך as one that traveleth, ראשׁך So shall thy poverty ומחסרך and thy want כאישׁ man. מגן׃ as an armed}%
\verseWithHeading{Against Worthlessness}{אדם person, בליעל A naughty אישׁ man, און a wicked הולך walketh עקשׁות with a froward פה׃ mouth.}%
\verse{קרץ He winketh בעינו with his eyes, מלל he speaketh ברגלו with his feet, מרה he teacheth באצבעתיו׃ with his fingers;}%
\verse{תהפכות Frowardness בלבו in his heart, חרשׁ he deviseth רע mischief בכל continually; עת continually; מדנים discord. ישׁלח׃ he soweth}%
\verse{על כןתאם suddenly; יבוא come אידו shall his calamity פתע suddenly ישׁבר shall he be broken ואין without מרפא׃ remedy.}%
\verseWithHeading{What Adonai Hates}{שׁשׁ six הנה These שׂנא hate: יהוה doth the LORD ושׁבע yea, seven תועבות an abomination נפשׁו׃ unto him:}%
\verse{עינים look, רמות A proud לשׁון tongue, שׁקר a lying וידים and hands שׁפכות that shed דם blood, נקי׃ innocent}%
\verse{לב A heart חרשׁ that deviseth מחשׁבות imaginations, און wicked רגלים feet ממהרות that be swift לרוץ in running לרעה׃ to mischief,}%
\verse{יפיח speaketh כזבים lies, עד witness שׁקר A false ומשׁלח and he that soweth מדנים discord בין among אחים׃ brethren.}%
\verseWithHeading{Commandment and Instruction as Guardians}{נצר keep בני My son, מצות commandment, אביך thy father's ואל not תטשׁ and forsake תורת the law אמך׃ of thy mother:}%
\verse{קשׁרם Bind על upon לבך thine heart, תמיד them continually ענדם tie על them about גרגרתך׃ thy neck.}%
\verse{בהתהלכך When thou goest, תנחה it shall lead אתך בשׁכבך thee;when thou sleepest, תשׁמר it shall keep עליך it shall keep והקיצות thee; and thou awakest, היא it תשׂיחך׃ shall talk}%
\verse{כי For נר a lamp; מצוה the commandment ותורה and the law אור light; ודרך the way חיים of life: תוכחות and reproofs מוסר׃ of instruction}%
\verse{לשׁמרך To keep מאשׁת woman, רע thee from the evil מחלקת from the flattery לשׁון of the tongue נכריה׃ of a strange woman.}%
\verse{אל not תחמד Lust יפיה after her beauty בלבבך in thine heart; ואל neither תקחך let her take בעפעפיה׃ thee with her eyelids.}%
\verse{כי For בעד by means אשׁה woman זונה of a whorish עד to ככר a piece לחם of bread: ואשׁת and the adulteress אישׁ and the adulteress נפשׁ life. יקרה for the precious תצוד׃ will hunt}%
\verseWithHeading{Warning Against Relations with a Married Woman}{היחתה take אישׁ Can a man אשׁ fire בחיקו in his bosom, ובגדיו and his clothes לא not תשׂרפנה׃ be burned?}%
\verse{אם Can יהלך go אישׁ one על upon הגחלים hot coals, ורגליו and his feet לא not תכוינה׃ be burned?}%
\verse{כן So הבא he that goeth in אל to אשׁת wife; רעהו his neighbor's לא her shall not ינקה be innocent. כל whosoever הנגע׃ toucheth}%
\verse{לא do not יבוזו despise לגנב a thief, כי if יגנוב he steal למלא to satisfy נפשׁו his soul כי when ירעב׃ he is hungry;}%
\verse{ונמצא But he be found, ישׁלם he shall restore שׁבעתים sevenfold; את כל all הון the substance ביתו of his house. יתן׃ he shall give}%
\verse{נאף whoso committeth adultery אשׁה with a woman חסר lacketh לב understanding: משׁחית it destroyeth נפשׁו his own soul. הוא he יעשׂנה׃ doeth}%
\verse{נגע A wound וקלון and dishonor ימצא shall he get; וחרפתו and his reproach לא shall not תמחה׃ be wiped away.}%
\verse{כי For קנאה jealousy חמת the rage גבר of a man: ולא therefore he will not יחמול spare ביום in the day נקם׃ of vengeance.}%
\verse{לא He will not ישׂא regard פני regard כל any כפר ransom; ולא neither יאבה will he rest content, כי though תרבה thou givest many שׁחד׃ gifts.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 7
\verseWithHeading{Warning Against the Strange Woman}{בני My son, שׁמר keep אמרי my words, ומצותי my commandments תצפן and lay up אתך׃ with}%
\verse{שׁמר Keep מצותי my commandments, וחיה and live; ותורתי and my law כאישׁון as the apple עיניך׃ of thine eye.}%
\verse{קשׁרם Bind על them upon אצבעתיך thy fingers, כתבם write על them upon לוח the table לבך׃ of thine heart.}%
\verse{אמר Say לחכמה unto wisdom, אחתי my sister; את Thou ומדע kinswoman: לבינה understanding תקרא׃ and call}%
\verse{לשׁמרך That they may keep מאשׁה woman, זרה thee from the strange מנכריה from the stranger אמריה with her words. החליקה׃ flattereth}%
\verse{כי For בחלון at the window ביתי of my house בעד through אשׁנבי my casement, נשׁקפתי׃ I looked}%
\verse{וארא And beheld בפתאים among the simple ones, אבינה I discerned בבנים among the youths, נער a young man חסר void לב׃ of understanding,}%
\verse{עבר Passing בשׁוק through the street אצל near פנה her corner; ודרך the way ביתה to her house, יצעד׃ and he went}%
\verse{בנשׁף In the twilight, בערב in the evening, יום in the evening, באישׁון in the black לילה night: ואפלה׃ and dark}%
\verse{והנה And, behold, אשׁה him a woman לקראתו there met שׁית the attire זונה of a harlot, ונצרת and subtle לב׃ of heart.}%
\verse{המיה loud היא (She וסררת and stubborn; בביתה in her house: לא not ישׁכנו abide רגליה׃ her feet}%
\verse{פעם Now בחוץ without, פעם now ברחבות in the streets, ואצל at כל every פנה corner.) תארב׃ and lieth in wait}%
\verse{והחזיקה So she caught בו ונשׁקה him, and kissed לו העזה him, with an impudent פניה face ותאמר׃ said}%
\verse{זבחי שׁלמיםלי with היום me; this day שׁלמתי have I paid נדרי׃ my vows.}%
\verse{על כןצאתי came I forth לקראתך to meet לשׁחר thee, diligently to seek פניך thy face, ואמצאך׃ and I have found}%
\verse{מרבדים with coverings of tapestry, רבדתי I have decked ערשׂי my bed חטבות with carved אטון with fine linen מצרים׃ of Egypt.}%
\verse{נפתי I have perfumed משׁכבי my bed מר with myrrh, אהלים aloes, וקנמון׃ and cinnamon.}%
\verse{לכה Come, נרוה let us take our fill דדים of love עד until הבקר the morning: נתעלסה let us solace ourselves באהבים׃ with loves.}%
\verse{כי For אין not האישׁ the goodman בביתו at home, הלך he is gone בדרך journey: מרחוק׃ a long}%
\verse{צרור a bag הכסף of money לקח He hath taken בידו with ליום at the day הכסא appointed. יבא him, will come ביתו׃ home}%
\verse{הטתו she caused him to yield, ברב With her much לקחה fair speech בחלק with the flattering שׂפתיה of her lips תדיחנו׃ she forced}%
\verse{הולך He goeth אחריה after פתאם her straightway, כשׁור as an ox אל to טבח the slaughter, יבוא goeth וכעכס of the stocks; אל to מוסר the correction אויל׃ or as a fool}%
\verse{עד Till יפלח strike through חץ a dart כבדו his liver; כמהר hasteth צפור as a bird אל to פח the snare, ולא not ידע and knoweth כי that בנפשׁו for his life. הוא׃ it}%
\verseWithHeading{Reiteration of the Warning Against a Strange Woman}{ועתה unto me now בנים therefore, O ye children, שׁמעו Hearken לי והקשׁיבו and attend לאמרי to the words פי׃ of my mouth.}%
\verse{אל Let not ישׂט decline אל to דרכיה her ways, לבך thine heart אל go not astray תתע go not astray בנתיבותיה׃ in her paths.}%
\verse{כי For רבים many חללים wounded: הפילה she hath cast down ועצמים strong כל yea, many הרגיה׃ have been slain}%
\verse{דרכי the way שׁאול to hell, ביתה Her house ירדות going down אל to חדרי the chambers מות׃ of death.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 8
\verseWithHeading{Wisdom Calls}{הלא Doth not חכמה wisdom תקרא cry? ותבונה and understanding תתן put forth קולה׃ her voice?}%
\verse{בראשׁ in the top מרומים of high places, עלי by דרך the way בית in the places נתיבות of the paths. נצבה׃ She standeth}%
\verse{ליד at שׁערים the gates, לפי at the entry קרת of the city, מבוא at the coming פתחים in at the doors. תרנה׃ She crieth}%
\verse{אליכם Unto אישׁים you, O men, אקרא I call; וקולי and my voice אל to בני the sons אדם׃ of man.}%
\verse{הבינו understand פתאים O ye simple, ערמה wisdom: וכסילים and, ye fools, הבינו be ye of an understanding לב׃ heart.}%
\verse{שׁמעו Hear; כי for נגידים of excellent things; אדבר I will speak ומפתח and the opening שׂפתי of my lips מישׁרים׃ right things.}%
\verse{כי For אמת truth; יהגה shall speak חכי my mouth ותועבת an abomination שׂפתי to my lips. רשׁע׃ and wickedness}%
\verse{בצדק in righteousness; כל All אמרי the words פי of my mouth אין nothing בהם נפתל froward ועקשׁ׃ or perverse}%
\verse{כלם They all נכחים plain למבין to him that understandeth, וישׁרים and right למצאי to them that find דעת׃ knowledge.}%
\verse{קחו Receive מוסרי my instruction, ואל and not כסף silver; ודעת and knowledge מחרוץ נבחר׃}%
\verse{כי For טובה better חכמה wisdom מפנינים than rubies; וכל and all חפצים the things that may be desired לא are not ישׁוו׃ to be compared}%
\verse{אני I חכמה wisdom שׁכנתי dwell ערמה with prudence, ודעת knowledge מזמות of witty inventions. אמצא׃ and find out}%
\verse{יראת The fear יהוה of the LORD שׂנאת to hate רע evil: גאה pride, וגאון and arrogance, ודרך way, רע and the evil ופי mouth, תהפכות and the froward שׂנאתי׃ do I hate.}%
\verse{לי עצה Counsel ותושׁיה mine, and sound wisdom: אני I בינה understanding; לי גבורה׃ I have strength.}%
\verse{בי מלכים By me kings ימלכו reign, ורוזנים and princes יחקקו decree צדק׃ justice.}%
\verse{בי שׂרים By me princes ישׂרו rule, ונדיבים and nobles, כל all שׁפטי the judges צדק׃ of the earth.}%
\verse{אני I אהביה love אהב them that love ומשׁחרי me; and those that seek me early ימצאנני׃ shall find}%
\verse{עשׁר Riches וכבוד and honor אתי with הון riches עתק me; durable וצדקה׃ and righteousness.}%
\verse{טוב better פריי My fruit מחרוץ than gold, ומפז yea, than fine gold; ותבואתי and my revenue מכסף silver. נבחר׃ than choice}%
\verse{בארח in the way צדקה of righteousness, אהלך I lead בתוך in the midst נתיבות of the paths משׁפט׃ of judgment:}%
\verse{להנחיל me to inherit אהבי That I may cause those that love ישׁ substance; ואצרתיהם their treasures. אמלא׃ and I will fill}%
\verseWithHeading{Wisdom at Creation}{יהוה The LORD קנני possessed ראשׁית me in the beginning דרכו of his way, קדם before מפעליו his works מאז׃ of old.}%
\verse{מעולם from everlasting, נסכתי I was set up מראשׁ from the beginning, מקדמי or ever ארץ׃ the earth}%
\verse{באין When no תהמות depths, חוללתי I was brought forth; באין when no מעינות fountains נכבדי abounding מים׃ with water.}%
\verse{בטרם Before הרים the mountains הטבעו were settled, לפני before גבעות the hills חוללתי׃}%
\verse{עד While as yet לא he had not עשׂה made ארץ the earth, וחוצות nor the fields, וראשׁ nor the highest part עפרות of the dust תבל׃ of the world.}%
\verse{בהכינו When he prepared שׁמים the heavens, שׁם there: אני I בחוקו when he set חוג a compass על upon פני the face תהום׃ of the depth:}%
\verse{באמצו When he established שׁחקים the clouds ממעל above: בעזוז when he strengthened עינות the fountains תהום׃ of the deep:}%
\verse{בשׂומו When he gave לים to the sea חקו his decree, ומים that the waters לא should not יעברו pass פיו his commandment: בחוקו when he appointed מוסדי ארץ׃ of the earth:}%
\verse{ואהיה Then I was אצלו by אמון him, one brought up ואהיה and I was שׁעשׁעים delight, יום daily יום daily משׂחקת rejoicing לפניו before בכל always עת׃ always}%
\verse{משׂחקת Rejoicing בתבל in the habitable part ארצו of his earth; ושׁעשׁעי and my delights את with בני the sons אדם׃ of men.}%
\verseWithHeading{Benefits of Following Wisdom}{ועתה Now בנים unto me, O ye children: שׁמעו therefore hearken לי ואשׁרי for blessed דרכי my ways. ישׁמרו׃ keep}%
\verse{שׁמעו Hear מוסר instruction, וחכמו and be wise, ואל it not. תפרעו׃ and refuse}%
\verse{אשׁרי Blessed אדם the man שׁמע that heareth לי לשׁקד me, watching על at דלתתי my gates, יום daily יום daily לשׁמר waiting מזוזת at the posts פתחי׃ of my doors.}%
\verse{כי For מצאי whoso findeth מצאי me findeth חיים life, ויפק and shall obtain רצון favor מיהוה׃}%
\verse{וחטאי But he that sinneth against חמס me wrongeth נפשׁו his own soul: כל all משׂנאי they that hate אהבו me love מות׃ death.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 9
\verseWithHeading{Wisdom’s Banquet}{חכמות Wisdom בנתה hath built ביתה her house, חצבה she hath hewn out עמודיה pillars: שׁבעה׃ her seven}%
\verse{טבחה She hath killed טבחה her beasts; מסכה she hath mingled יינה her wine; אף she hath also ערכה furnished שׁלחנה׃ her table.}%
\verse{שׁלחה She hath sent forth נערתיה her maidens: תקרא she crieth על upon גפי the highest places מרמי the highest places קרת׃ of the city,}%
\verse{מי Whoso פתי simple, יסר let him turn הנה in hither: חסר him that wanteth לב understanding, אמרה׃ she saith}%
\verse{לכו Come, לחמו eat בלחמי of my bread, ושׁתו and drink ביין of the wine מסכתי׃ I have mingled.}%
\verse{עזבו Forsake פתאים the foolish, וחיו and live; ואשׁרו and go בדרך in the way בינה׃ of understanding.}%
\verse{יסר He that reproveth לץ a scorner לקח getteth לו קלון to himself shame: ומוכיח and he that rebuketh לרשׁע a wicked מומו׃ himself a blot.}%
\verse{אל not תוכח Reprove לץ a scorner, פן lest ישׂנאך he hate הוכח thee: rebuke לחכם a wise man, ויאהבך׃ and he will love}%
\verse{תן Give לחכם to a wise ויחכם wiser: עוד and he will be yet הודע teach לצדיק a just ויוסף and he will increase לקח׃ in learning.}%
\verseWithHeading{Foolishness’ Banquet}{תחלת the beginning חכמה of wisdom: יראת The fear יהוה ודעת and the knowledge קדשׁים of the holy בינה׃ understanding.}%
\verse{כי For בי ירבו shall be multiplied, ימיך by me thy days ויוסיפו shall be increased. לך שׁנות and the years חיים׃ of thy life}%
\verse{אם If חכמת thou be wise, חכמת thou shalt be wise לך ולצת for thyself: but thou scornest, לבדך thou alone תשׂא׃ shalt bear}%
\verse{אשׁת woman כסילות A foolish המיה clamorous: פתיות simple, ובל nothing. ידעה and knoweth מה׃ nothing.}%
\verse{וישׁבה For she sitteth לפתח at the door ביתה of her house, על on כסא a seat מרמי in the high places קרת׃ of the city,}%
\verse{לקרא To call לעברי passengers דרך passengers המישׁרים who go right ארחותם׃ on their ways:}%
\verse{מי Whoso פתי simple, יסר let him turn הנה in hither: וחסר and him that wanteth לב understanding, ואמרה׃ she saith}%
\verse{מים waters גנובים Stolen ימתקו are sweet, ולחם and bread סתרים in secret ינעם׃ is pleasant.}%
\verse{ולא not ידע But he knoweth כי that רפאים the dead שׁם there; בעמקי שׁאול of hell. קראיה׃ her guests}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 10
\verseWithHeading{Proverbs of Solomon}{משׁלי The proverbs שׁלמה of Solomon. בן son חכם A wise ישׂמח maketh a glad אב father: ובן son כסיל but a foolish תוגת the heaviness אמו׃ of his mother.}%
\verse{לא nothing: יועילו profit אוצרות Treasures רשׁע of wickedness וצדקה but righteousness תציל delivereth ממות׃ from death.}%
\verse{לא will not ירעיב to famish: יהוה The LORD נפשׁ suffer the soul צדיק of the righteous והות the substance רשׁעים of the wicked. יהדף׃ but he casteth away}%
\verse{ראשׁ He becometh poor עשׂה that dealeth כף hand: רמיה a slack ויד but the hand חרוצים of the diligent תעשׁיר׃ maketh rich.}%
\verse{אגר He that gathereth בקיץ in summer בן son: משׂכיל a wise נרדם he that sleepeth בקציר in harvest בן a son מבישׁ׃ that causeth shame.}%
\verse{ברכות Blessings לראשׁ upon the head צדיק of the just: ופי the mouth רשׁעים of the wicked. יכסה covereth חמס׃ but violence}%
\verse{זכר The memory צדיק of the just לברכה blessed: ושׁם but the name רשׁעים of the wicked ירקב׃ shall rot.}%
\verse{חכם The wise לב in heart יקח will receive מצות commandments: ואויל fool שׂפתים but a prating ילבט׃ shall fall.}%
\verse{הולך He that walketh בתם uprightly ילך walketh בטח surely: ומעקשׁ but he that perverteth דרכיו his ways יודע׃ shall be known.}%
\verse{קרץ He that winketh עין with the eye יתן causeth עצבת sorrow: ואויל fool שׂפתים but a prating ילבט׃ shall fall.}%
\verse{מקור a well חיים of life: פי The mouth צדיק of a righteous ופי the mouth רשׁעים of the wicked. יכסה covereth חמס׃ but violence}%
\verse{שׂנאה Hatred תעורר stirreth up מדנים strifes: ועל covereth כל all פשׁעים sins. תכסה covereth אהבה׃ but love}%
\verse{בשׂפתי In the lips נבון of him that hath understanding תמצא is found: חכמה wisdom ושׁבט but a rod לגו for the back חסר of him that is void לב׃ of understanding.}%
\verse{חכמים Wise יצפנו lay up דעת knowledge: ופי but the mouth אויל of the foolish מחתה destruction. קרבה׃ near}%
\verse{הון wealth עשׁיר The rich man's קרית city: עזו his strong מחתת the destruction דלים of the poor רישׁם׃ their poverty.}%
\verse{פעלת The labor צדיק of the righteous לחיים to life: תבואת the fruit רשׁע of the wicked לחטאת׃ to sin.}%
\verse{ארח He the way לחיים of life שׁומר that keepeth מוסר instruction: ועוזב but he that refuseth תוכחת reproof מתעה׃ erreth.}%
\verse{מכסה He that hideth שׂנאה hatred שׂפתי lips, שׁקר lying ומוצא and he that uttereth דבה a slander, הוא כסיל׃ a fool.}%
\verse{ברב In the multitude דברים of words לא not יחדל there wanteth פשׁע sin: וחשׂך but he that refraineth שׂפתיו his lips משׂכיל׃ wise.}%
\verse{כסף silver: נבחר choice לשׁון The tongue צדיק of the just לב the heart רשׁעים of the wicked כמעט׃ little worth.}%
\verse{שׂפתי The lips צדיק of the righteous ירעו feed רבים many: ואוילים but fools בחסר for want לב of wisdom. ימותו׃ die}%
\verse{ברכת The blessing יהוה of the LORD, היא it תעשׁיר maketh rich, ולא no יוסף and he addeth עצב sorrow עמה׃ with}%
\verse{כשׂחוק as sport לכסיל to a fool עשׂות to do זמה mischief: וחכמה hath wisdom. לאישׁ but a man תבונה׃ of understanding}%
\verse{מגורת The fear רשׁע of the wicked, היא it תבואנו shall come upon ותאות him: but the desire צדיקים of the righteous יתן׃ shall be granted.}%
\verse{כעבור passeth, סופה As the whirlwind ואין no רשׁע so the wicked וצדיק but the righteous יסוד foundation. עולם׃ an everlasting}%
\verse{כחמץ As vinegar לשׁנים to the teeth, וכעשׁן and as smoke לעינים to the eyes, כן so העצל the sluggard לשׁלחיו׃ to them that send}%
\verse{יראת The fear יהוה of the LORD תוסיף prolongeth ימים days: ושׁנות but the years רשׁעים of the wicked תקצרנה׃ shall be shortened.}%
\verse{תוחלת The hope צדיקים of the righteous שׂמחה gladness: ותקות but the expectation רשׁעים of the wicked תאבד׃ shall perish.}%
\verse{מעוז strength לתם to the upright: דרך The way יהוה of the LORD ומחתה but destruction לפעלי to the workers און׃ of iniquity.}%
\verse{צדיק The righteous לעולם shall never בל shall never ימוט be removed: ורשׁעים but the wicked לא shall not ישׁכנו inhabit ארץ׃ the earth.}%
\verse{פי The mouth צדיק of the just ינוב bringeth forth חכמה wisdom: ולשׁון tongue תהפכות but the froward תכרת׃ shall be cut out.}%
\verse{שׂפתי The lips צדיק of the righteous ידעון know רצון what is acceptable: ופי but the mouth רשׁעים of the wicked תהפכות׃ frowardness.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 11
\verse{מאזני balance מרמה A false תועבת abomination יהוה to the LORD: ואבן weight שׁלמה but a just רצונו׃ his delight.}%
\verse{בא cometh, זדון pride ויבא then cometh קלון shame: ואת but with צנועים the lowly חכמה׃ wisdom.}%
\verse{תמת The integrity ישׁרים of the upright תנחם shall guide וסלף them: but the perverseness בוגדים of transgressors ושׁדם׃ shall destroy}%
\verse{לא not יועיל profit הון Riches ביום in the day עברה of wrath: וצדקה but righteousness תציל delivereth ממות׃ from death.}%
\verse{צדקת The righteousness תמים of the perfect תישׁר shall direct דרכו his way: וברשׁעתו by his own wickedness. יפל shall fall רשׁע׃ but the wicked}%
\verse{צדקת The righteousness ישׁרים of the upright תצילם shall deliver ובהות in naughtiness. בגדים them: but transgressors ילכדו׃ shall be taken}%
\verse{במות dieth, אדם man רשׁע When a wicked תאבד shall perish: תקוה expectation ותוחלת and the hope אונים of unjust אבדה׃ perisheth.}%
\verse{צדיק The righteous מצרה out of trouble, נחלץ is delivered ויבא cometh רשׁע and the wicked תחתיו׃ in his stead.}%
\verse{בפה with mouth חנף A hypocrite ישׁחת destroyeth רעהו his neighbor: ובדעת but through knowledge צדיקים shall the just יחלצו׃ be delivered.}%
\verse{בטוב When it goeth well צדיקים with the righteous, תעלץ rejoiceth: קריה the city ובאבד perish, רשׁעים and when the wicked רנה׃ shouting.}%
\verse{בברכת By the blessing ישׁרים of the upright תרום is exalted: קרת the city ובפי by the mouth רשׁעים of the wicked. תהרס׃ but it is overthrown}%
\verse{בז despiseth לרעהו his neighbor: חסר He that is void לב of wisdom ואישׁ but a man תבונות of understanding יחרישׁ׃ holdeth his peace.}%
\verse{הולך רכילגלה revealeth סוד secrets: ונאמן but he that is of a faithful רוח spirit מכסה concealeth דבר׃ the matter.}%
\verse{באין Where no תחבלות counsel יפל fall: עם the people ותשׁועה safety. ברב but in the multitude יועץ׃ of counselors}%
\verse{רע shall smart ירוע shall smart כי ערב He that is surety זר for a stranger ושׂנא and he that hateth תקעים suretiship בוטח׃ is sure.}%
\verse{אשׁת woman חן A gracious תתמך retaineth כבוד honor: ועריצים and strong יתמכו retain עשׁר׃ riches.}%
\verse{גמל doeth good נפשׁו to his own soul: אישׁ man חסד The merciful ועכר troubleth שׁארו his own flesh. אכזרי׃ but cruel}%
\verse{רשׁע The wicked עשׂה worketh פעלת work: שׁקר a deceitful וזרע but to him that soweth צדקה righteousness שׂכר reward. אמת׃ a sure}%
\verse{כן As צדקה righteousness לחיים to life: ומרדף so he that pursueth רעה evil למותו׃ to his own death.}%
\verse{תועבת abomination יהוה to the LORD: עקשׁי They that are of a froward לב heart ורצונו his delight. תמימי but upright דרך׃ in way}%
\verse{יד hand ליד in hand, לא shall not ינקה be unpunished: רע the wicked וזרע but the seed צדיקים of the righteous נמלט׃ shall be delivered.}%
\verse{נזם a jewel זהב of gold באף snout, חזיר in a swine's אשׁה woman יפה a fair וסרת which is without טעם׃ discretion.}%
\verse{תאות The desire צדיקים of the righteous אך only טוב good: תקות the expectation רשׁעים of the wicked עברה׃ wrath.}%
\verse{ישׁ There is מפזר that scattereth, ונוסף increaseth; עוד and yet וחושׂך and that withholdeth מישׁר more than is meet, אך למחסור׃ but to poverty.}%
\verse{נפשׁ soul ברכה The liberal תדשׁן shall be made fat: ומרוה and he that watereth גם also הוא himself. יורא׃ shall be watered}%
\verse{מנע He that withholdeth בר corn, יקבהו לאום the people וברכה him: but blessing לראשׁ upon the head משׁביר׃ of him that selleth}%
\verse{שׁחר He that diligently seeketh טוב good יבקשׁ procureth רצון favor: ודרשׁ but he that seeketh רעה mischief, תבואנו׃ it shall come}%
\verse{בוטח that trusteth בעשׁרו in his riches הוא He יפל shall fall: וכעלה as a branch. צדיקים but the righteous יפרחו׃ shall flourish}%
\verse{עוכר He that troubleth ביתו his own house ינחל shall inherit רוח the wind: ועבד servant אויל and the fool לחכם to the wise לב׃ of heart.}%
\verse{פרי The fruit צדיק of the righteous עץ a tree חיים of life; ולקח and he that winneth נפשׂות souls חכם׃ wise.}%
\verse{הן Behold, צדיק the righteous בארץ in the earth: ישׁלם shall be recompensed אף much more כי much more רשׁע the wicked וחוטא׃ and the sinner.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 12
\verse{אהב Whoso loveth מוסר instruction אהב loveth דעת knowledge: ושׂנא but he that hateth תוכחת reproof בער׃}%
\verse{טוב A good יפיק obtaineth רצון favor מיהוה ואישׁ but a man מזמות of wicked devices ירשׁיע׃ will he condemn.}%
\verse{לא shall not יכון be established אדם A man ברשׁע by wickedness: ושׁרשׁ but the root צדיקים of the righteous בל shall not ימוט׃ be moved.}%
\verse{אשׁת woman חיל A virtuous עטרת a crown בעלה to her husband: וכרקב as rottenness בעצמותיו in his bones. מבישׁה׃ but she that maketh ashamed}%
\verse{מחשׁבות The thoughts צדיקים of the righteous משׁפט right: תחבלות the counsels רשׁעים of the wicked מרמה׃ deceit.}%
\verse{דברי The words רשׁעים of the wicked ארב to lie in wait דם for blood: ופי but the mouth ישׁרים of the upright יצילם׃ shall deliver}%
\verse{הפוך are overthrown, רשׁעים The wicked ואינם and not: ובית but the house צדיקים of the righteous יעמד׃ shall stand.}%
\verse{לפי according שׂכלו to his wisdom: יהלל shall be commended אישׁ A man ונעוה but he that is of a perverse לב heart יהיה shall be לבוז׃ despised.}%
\verse{טוב better נקלה despised, ועבד and hath a servant, לו ממתכבד than he that honoreth himself, וחסר and lacketh לחם׃ bread.}%
\verse{יודע regardeth צדיק A righteous נפשׁ the life בהמתו of his beast: ורחמי but the tender mercies רשׁעים of the wicked אכזרי׃ cruel.}%
\verse{עבד He that tilleth אדמתו his land ישׂבע shall be satisfied לחם with bread: ומרדף but he that followeth ריקים vain חסר void לב׃ of understanding.}%
\verse{חמד desireth רשׁע The wicked מצוד the net רעים of evil ושׁרשׁ but the root צדיקים of the righteous יתן׃ yieldeth}%
\verse{בפשׁע by the transgression שׂפתים of lips: מוקשׁ is snared רע The wicked ויצא shall come out מצרה of trouble. צדיק׃ but the just}%
\verse{מפרי by the fruit פי of mouth: אישׁ A man ישׂבע shall be satisfied טוב with good וגמול and the recompense ידי hands אדם of a man's ישׁוב׃ shall be rendered}%
\verse{דרך The way אויל of a fool ישׁר right בעיניו in his own eyes: ושׁמע but he that hearkeneth לעצה unto counsel חכם׃ wise.}%
\verse{אויל A fool's ביום is presently יודע known: כעסו wrath וכסה covereth קלון shame. ערום׃ but a prudent}%
\verse{יפיח speaketh אמונה truth יגיד showeth forth צדק righteousness: ועד witness שׁקרים but a false מרמה׃ deceit.}%
\verse{ישׁ There is בוטה that speaketh כמדקרות like the piercings חרב of a sword: ולשׁון but the tongue חכמים of the wise מרפא׃ health.}%
\verse{שׂפת The lip אמת of truth תכון shall be established לעד forever: ועד but ארגיעה לשׁון tongue שׁקר׃ a lying}%
\verse{מרמה Deceit בלב in the heart חרשׁי of them that imagine רע evil: וליעצי but to the counselors שׁלום of peace שׂמחה׃ joy.}%
\verse{לא יאנה happen לצדיק to the just: כל און evil ורשׁעים but the wicked מלאו shall be filled רע׃ with mischief.}%
\verse{תועבת abomination יהוה to the LORD: שׂפתי lips שׁקר Lying ועשׂי but they that deal אמונה truly רצונו׃ his delight.}%
\verse{אדם man ערום A prudent כסה concealeth דעת knowledge: ולב but the heart כסילים of fools יקרא proclaimeth אולת׃ foolishness.}%
\verse{יד The hand חרוצים of the diligent תמשׁול shall bear rule: ורמיה but the slothful תהיה shall be למס׃ under tribute.}%
\verse{דאגה Heaviness בלב in the heart אישׁ of man ישׁחנה maketh it stoop: ודבר word טוב but a good ישׂמחנה׃ maketh it glad.}%
\verse{יתר more excellent מרעהו than his neighbor: צדיק The righteous ודרך but the way רשׁעים of the wicked תתעם׃ seduceth}%
\verse{לא not יחרך roasteth רמיה The slothful צידו that which he took in hunting: והון but the substance אדם man יקר precious. חרוץ׃ of a diligent}%
\verse{בארח In the way צדקה of righteousness חיים life; ודרך and the pathway נתיבה and the pathway אל no מות׃ death.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 13
\verse{בן son חכם A wise מוסר instruction: אב his father's ולץ but a scorner לא not שׁמע heareth גערה׃ rebuke.}%
\verse{מפרי by the fruit פי of mouth: אישׁ A man יאכל shall eat טוב good ונפשׁ but the soul בגדים of the transgressors חמס׃ violence.}%
\verse{נצר He that keepeth פיו his mouth שׁמר keepeth נפשׁו his life: פשׂק he that openeth wide שׂפתיו his lips מחתה׃ shall have destruction.}%
\verse{מתאוה desireth, ואין and nothing: נפשׁו The soul עצל of the sluggard ונפשׁ but the soul חרצים of the diligent תדשׁן׃ shall be made fat.}%
\verse{דבר lying: שׁקר lying: ישׂנא hateth צדיק A righteous ורשׁע but a wicked יבאישׁ is loathsome, ויחפיר׃ and cometh to shame.}%
\verse{צדקה Righteousness תצר keepeth תם upright דרך in the way: ורשׁעה but wickedness תסלף overthroweth חטאת׃ the sinner.}%
\verse{ישׁ There is מתעשׁר that maketh himself rich, ואין yet nothing: כל yet nothing: מתרושׁשׁ that maketh himself poor, והון riches. רב׃ yet great}%
\verse{כפר The ransom נפשׁ life אישׁ of a man's עשׁרו his riches: ורשׁ but the poor לא not שׁמע heareth גערה׃ rebuke.}%
\verse{אור The light צדיקים of the righteous ישׂמח rejoiceth: ונר but the lamp רשׁעים of the wicked ידעך׃ shall be put out.}%
\verse{רק Only בזדון by pride יתן cometh מצה contention: ואת but with נועצים the well advised חכמה׃ wisdom.}%
\verse{הון Wealth מהבל by vanity ימעט shall be diminished: וקבץ but he that gathereth על by יד labor ירבה׃ shall increase.}%
\verse{תוחלת Hope ממשׁכה deferred מחלה sick: לב maketh the heart ועץ a tree חיים of life. תאוה but the desire באה׃ cometh,}%
\verse{בז Whoso despiseth לדבר the word יחבל shall be destroyed: לו וירא that feareth מצוה the commandment הוא but he ישׁלם׃ shall be rewarded.}%
\verse{תורת The law חכם of the wise מקור a fountain חיים of life, לסור to depart ממקשׁי from the snares מות׃ of death.}%
\verse{שׂכל understanding טוב Good יתן giveth חן favor: ודרך but the way בגדים of transgressors איתן׃ hard.}%
\verse{כל Every ערום prudent יעשׂה dealeth בדעת with knowledge: וכסיל but a fool יפרשׂ layeth open אולת׃ folly.}%
\verse{מלאך messenger רשׁע A wicked יפל falleth ברע into mischief: וציר ambassador אמונים but a faithful מרפא׃ health.}%
\verse{רישׁ Poverty וקלון and shame פורע him that refuseth מוסר instruction: ושׁומר but he that regardeth תוכחת reproof יכבד׃ shall be honored.}%
\verse{תאוה The desire נהיה accomplished תערב is sweet לנפשׁ to the soul: ותועבת but abomination כסילים to fools סור to depart מרע׃ from evil.}%
\verse{הלוך He that walketh את with חכמים wise וחכם shall be wise: ורעה but a companion כסילים of fools ירוע׃ shall be destroyed.}%
\verse{חטאים sinners: תרדף pursueth רעה Evil ואת צדיקים but to the righteous ישׁלם shall be repaid. טוב׃ good}%
\verse{טוב A good ינחיל leaveth an inheritance בני to his children's בנים children: וצפון laid up לצדיק for the just. חיל and the wealth חוטא׃ of the sinner}%
\verse{רב Much אכל food ניר the tillage ראשׁים of the poor: וישׁ but there is נספה destroyed בלא for want משׁפט׃ of judgment.}%
\verse{חושׂך He that spareth שׁבטו his rod שׂונא hateth בנו his son: ואהבו but he that loveth שׁחרו him quickly. מוסר׃ him chasteneth}%
\verse{צדיק The righteous אכל eateth לשׂבע to the satisfying נפשׁו of his soul: ובטן but the belly רשׁעים of the wicked תחסר׃ shall want.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 14
\verse{חכמות Every wise נשׁים woman בנתה buildeth ביתה her house: ואולת but the foolish בידיה with her hands. תהרסנו׃ plucketh it down}%
\verse{הולך He that walketh בישׁרו in his uprightness ירא יהוה the LORD: ונלוז but perverse דרכיו in his ways בוזהו׃ despiseth}%
\verse{בפי In the mouth אויל of the foolish חטר a rod גאוה of pride: ושׂפתי but the lips חכמים of the wise תשׁמורם׃ shall preserve}%
\verse{באין Where no אלפים oxen אבוס the crib בר clean: ורב but much תבואות increase בכח by the strength שׁור׃ of the ox.}%
\verse{עד witness אמונים A faithful לא will not יכזב lie: ויפיח will utter כזבים lies. עד witness שׁקר׃ but a false}%
\verse{בקשׁ seeketh לץ A scorner חכמה wisdom, ואין and not: ודעת but knowledge לנבון unto him that understandeth. נקל׃ easy}%
\verse{לך Go מנגד from the presence לאישׁ man, כסיל of a foolish ובל not ידעת when thou perceivest שׂפתי the lips דעת׃ of knowledge.}%
\verse{חכמת The wisdom ערום of the prudent הבין to understand דרכו his way: ואולת but the folly כסילים of fools מרמה׃ deceit.}%
\verse{אולים Fools יליץ make a mock אשׁם at sin: ובין but among ישׁרים the righteous רצון׃ favor.}%
\verse{לב The heart יודע knoweth מרת נפשׁו his own ובשׂמחתו with his joy. לא doth not יתערב intermeddle זר׃ and a stranger}%
\verse{בית The house רשׁעים of the wicked ישׁמד shall be overthrown: ואהל but the tabernacle ישׁרים of the upright יפריח׃ shall flourish.}%
\verse{ישׁ There is דרך a way ישׁר which seemeth right לפני unto אישׁ a man, ואחריתה but the end דרכי thereof the ways מות׃ of death.}%
\verse{גם Even בשׂחוק in laughter יכאב is sorrowful; לב the heart ואחריתה and the end שׂמחה of that mirth תוגה׃ heaviness.}%
\verse{מדרכיו with his own ways: ישׂבע shall be filled סוג The backslider לב in heart ומעליו from אישׁ man טוב׃ and a good}%
\verse{פתי The simple יאמין believeth לכל every דבר word: וערום but the prudent יבין looketh well לאשׁרו׃ to his going.}%
\verse{חכם A wise ירא feareth, וסר and departeth מרע from evil: וכסיל but the fool מתעבר rageth, ובוטח׃ and is confident.}%
\verse{קצר soon אפים angry יעשׂה dealeth אולת foolishly: ואישׁ and a man מזמות of wicked devices ישׂנא׃ is hated.}%
\verse{נחלו inherit פתאים The simple אולת folly: וערומים but the prudent יכתרו are crowned דעת׃ with knowledge.}%
\verse{שׁחו bow רעים The evil לפני before טובים the good; ורשׁעים and the wicked על at שׁערי the gates צדיק׃ of the righteous.}%
\verse{גם even לרעהו of his own neighbor: ישׂנא is hated רשׁ The poor ואהבי friends. עשׁיר but the rich רבים׃ many}%
\verse{בז He that despiseth לרעהו his neighbor חוטא sinneth: ומחונן but he that hath mercy on עניים אשׁריו׃ happy}%
\verse{הלוא Do they not יתעו err חרשׁי that devise רע evil? וחסד but mercy ואמת and truth חרשׁי to them that devise טוב׃ good.}%
\verse{בכל In all עצב labor יהיה there is מותר profit: ודבר but the talk שׂפתים of the lips אך only למחסור׃ to penury.}%
\verse{עטרת The crown חכמים of the wise עשׁרם their riches: אולת the foolishness כסילים of fools אולת׃ folly.}%
\verse{מציל delivereth נפשׁות souls: עד witness אמת A true ויפח speaketh כזבים lies. מרמה׃ but a deceitful}%
\verse{ביראת In the fear יהוה מבטח confidence: עז strong ולבניו and his children יהיה shall have מחסה׃ a place of refuge.}%
\verse{יראת The fear יהוה of the LORD מקור a fountain חיים of life, לסור to depart ממקשׁי from the snares מות׃ of death.}%
\verse{ברב In the multitude עם of people הדרת honor: מלך the king's ובאפס but in the want לאם of people מחתת the destruction רזון׃ of the prince.}%
\verse{ארך slow אפים to wrath רב of great תבונה understanding: וקצר but hasty רוח of spirit מרים exalteth אולת׃ folly.}%
\verse{חיי the life בשׂרים of the flesh: לב heart מרפא A sound ורקב the rottenness עצמות of the bones. קנאה׃ but envy}%
\verse{עשׁק He that oppresseth דל the poor חרף reproacheth עשׂהו his Maker: ומכבדו but he that honoreth חנן him hath mercy אביון׃ on the poor.}%
\verse{ברעתו in his wickedness: ידחה is driven away רשׁע The wicked וחסה hath hope במותו in his death. צדיק׃ but the righteous}%
\verse{בלב in the heart נבון of him that hath understanding: תנוח resteth חכמה Wisdom ובקרב but in the midst כסילים of fools תודע׃ is made known.}%
\verse{צדקה Righteousness תרומם exalteth גוי a nation: וחסד a reproach לאמים to any people. חטאת׃ but sin}%
\verse{רצון favor מלך The king's לעבד servant: משׂכיל toward a wise ועברתו but his wrath תהיה is מבישׁ׃ him that causeth shame.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 15
\verse{מענה answer רך A soft ישׁיב turneth away חמה wrath: ודבר words עצב but grievous יעלה stir up אף׃ anger.}%
\verse{לשׁון The tongue חכמים of the wise תיטיב aright: דעת useth knowledge ופי but the mouth כסילים of fools יביע poureth out אולת׃ foolishness.}%
\verse{בכל in every מקום place, עיני The eyes יהוה of the LORD צפות beholding רעים the evil וטובים׃ and the good.}%
\verse{מרפא A wholesome לשׁון tongue עץ a tree חיים of life: וסלף but perverseness בה שׁבר therein a breach ברוח׃ in the spirit.}%
\verse{אויל A fool ינאץ despiseth מוסר instruction: אביו his father's ושׁמר but he that regardeth תוכחת reproof יערם׃ is prudent.}%
\verse{בית In the house צדיק of the righteous חסן treasure: רב much ובתבואת but in the revenues רשׁע of the wicked נעכרת׃ is trouble.}%
\verse{שׂפתי The lips חכמים of the wise יזרו disperse דעת knowledge: ולב but the heart כסילים of the foolish לא not כן׃ so.}%
\verse{זבח The sacrifice רשׁעים of the wicked תועבת an abomination יהוה to the LORD: ותפלת but the prayer ישׁרים of the upright רצונו׃ his delight.}%
\verse{תועבת an abomination יהוה unto the LORD: דרך The way רשׁע of the wicked ומרדף him that followeth after צדקה righteousness. יאהב׃ but he loveth}%
\verse{מוסר Correction רע grievous לעזב unto him that forsaketh ארח the way: שׂונא he that hateth תוכחת reproof ימות׃ shall die.}%
\verse{שׁאול Hell ואבדון and destruction נגד before יהוה אף how much more then כי how much more then לבות the hearts בני of the children אדם׃ of men?}%
\verse{לא not יאהב loveth לץ A scorner הוכח one that reproveth לו אל unto חכמים the wise. לא him: neither ילך׃ will he go}%
\verse{לב heart שׂמח A merry ייטב maketh a cheerful פנים countenance: ובעצבת but by sorrow לב of the heart רוח the spirit נכאה׃ is broken.}%
\verse{לב The heart נבון of him that hath understanding יבקשׁ seeketh דעת knowledge: ופני כסילים of fools ירעה feedeth on אולת׃ foolishness.}%
\verse{כל All ימי the days עני of the afflicted רעים evil: וטוב but he that is of a merry לב heart משׁתה feast. תמיד׃ a continual}%
\verse{טוב Better מעט little ביראת with the fear יהוה of the LORD מאוצר treasure רב than great ומהומה׃ and trouble}%
\verse{טוב Better ארחת a dinner ירק of herbs ואהבה love שׁם where משׁור ox אבוס is, than a stalled ושׂנאה׃ and hatred}%
\verse{אישׁ man חמה A wrathful יגרה stirreth up מדון strife: וארך but slow אפים to anger ישׁקיט appeaseth ריב׃ strife.}%
\verse{דרך The way עצל of the slothful כמשׂכת as a hedge חדק of thorns: וארח but the way ישׁרים of the righteous סללה׃ made plain.}%
\verse{בן son חכם A wise ישׂמח maketh a glad אב father: וכסיל but a foolish אדם man בוזה despiseth אמו׃ his mother.}%
\verse{אולת Folly שׂמחה joy לחסר to destitute לב of wisdom: ואישׁ but a man תבונה of understanding יישׁר uprightly. לכת׃ walketh}%
\verse{הפר are disappointed: מחשׁבות purposes באין Without סוד counsel וברב but in the multitude יועצים of counselors תקום׃ they are established.}%
\verse{שׂמחה hath joy לאישׁ A man במענה by the answer פיו of his mouth: ודבר and a word בעתו in due season, מה how טוב׃ good}%
\verse{ארח The way חיים of life למעלה above למשׂכיל to the wise, למען that סור he may depart משׁאול from hell מטה׃ beneath.}%
\verse{בית the house גאים of the proud: יסח will destroy יהוה The LORD ויצב but he will establish גבול the border אלמנה׃ of the widow.}%
\verse{תועבת an abomination יהוה to the LORD: מחשׁבות The thoughts רע of the wicked וטהרים but of the pure אמרי words. נעם׃ pleasant}%
\verse{עכר troubleth ביתו his own house; בוצע He that is greedy בצע of gain ושׂונא but he that hateth מתנת gifts יחיה׃ shall live.}%
\verse{לב The heart צדיק of the righteous יהגה studieth לענות to answer: ופי but the mouth רשׁעים of the wicked יביע poureth out רעות׃ evil things.}%
\verse{רחוק far יהוה The LORD מרשׁעים from the wicked: ותפלת the prayer צדיקים of the righteous. ישׁמע׃ but he heareth}%
\verse{מאור The light עינים of the eyes ישׂמח rejoiceth לב the heart: שׁמועה report טובה a good תדשׁן fat. עצם׃ maketh the bones}%
\verse{אזן The ear שׁמעת that heareth תוכחת the reproof חיים of life בקרב among חכמים the wise. תלין׃ abideth}%
\verse{פורע He that refuseth מוסר instruction מואס despiseth נפשׁו his own soul: ושׁומע but he that heareth תוכחת reproof קונה getteth לב׃ understanding.}%
\verse{יראת The fear יהוה of the LORD מוסר the instruction חכמה of wisdom; ולפני and before כבוד honor ענוה׃ humility.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 16
\verse{לאדם in man, מערכי The preparations לב of the heart ומיהוה מענה and the answer לשׁון׃ of the tongue,}%
\verse{כל All דרכי the ways אישׁ of a man זך clean בעיניו in his own eyes; ותכן weigheth רוחות the spirits. יהוה׃}%
\verse{גל Commit אל unto יהוה the LORD, מעשׂיך thy works ויכנו shall be established. מחשׁבתיך׃ and thy thoughts}%
\verse{כל all פעל hath made יהוה The LORD למענהו for himself: וגם yea, even רשׁע the wicked ליום for the day רעה׃ of evil.}%
\verse{תועבת an abomination יהוה to the LORD: כל Every one גבה proud לב in heart יד hand ליד in hand, לא he shall not ינקה׃ be unpunished.}%
\verse{בחסד By mercy ואמת and truth יכפר is purged: עון iniquity וביראת and by the fear יהוה of the LORD סור depart מרע׃ from evil.}%
\verse{ברצות please יהוה the LORD, דרכי ways אישׁ When a man's גם he maketh even אויביו his enemies ישׁלם to be at peace אתו׃ with}%
\verse{טוב Better מעט a little בצדקה with righteousness מרב than great תבואות revenues בלא without משׁפט׃ right.}%
\verse{לב heart אדם A man's יחשׁב deviseth דרכו his way: ויהוה but the LORD יכין directeth צעדו׃ his steps.}%
\verse{קסם A divine sentence על in שׂפתי the lips מלך of the king: במשׁפט in judgment. לא not ימעל transgresseth פיו׃ his mouth}%
\verse{פלס weight ומאזני and balance משׁפט A just ליהוה the LORD's: מעשׂהו his work. כל all אבני the weights כיס׃ of the bag}%
\verse{תועבת an abomination מלכים to kings עשׂות to commit רשׁע wickedness: כי for בצדקה by righteousness. יכון is established כסא׃ the throne}%
\verse{רצון the delight מלכים of kings; שׂפתי lips צדק Righteous ודבר him that speaketh ישׁרים right. יאהב׃ and they love}%
\verse{חמת The wrath מלך of a king מלאכי messengers מות of death: ואישׁ man חכם but a wise יכפרנה׃ will pacify}%
\verse{באור In the light פני countenance מלך of the king's חיים life; ורצונו and his favor כעב as a cloud מלקושׁ׃ of the latter rain.}%
\verse{קנה to get חכמה wisdom מה How much טוב better מחרוץ וקנות and to get בינה understanding נבחר rather to be chosen מכסף׃}%
\verse{מסלת The highway ישׁרים of the upright סור to depart מרע from evil: שׁמר preserveth נפשׁו his soul. נצר he that keepeth דרכו׃ his way}%
\verse{לפני before שׁבר destruction, גאון Pride ולפני before כשׁלון a fall. גבה and a haughty רוח׃ spirit}%
\verse{טוב Better שׁפל of a humble רוח spirit את with עניים מחלק than to divide שׁלל the spoil את with גאים׃ the proud.}%
\verse{משׂכיל עלבר ימצא shall find טוב good: ובוטח and whoso trusteth ביהוה in the LORD, אשׁריו׃ happy}%
\verse{לחכם The wise לב in heart יקרא shall be called נבון prudent: ומתק and the sweetness שׂפתים of the lips יסיף increaseth לקח׃ learning.}%
\verse{מקור a wellspring חיים of life שׂכל Understanding בעליו unto him that hath ומוסר it: but the instruction אולים of fools אולת׃ folly.}%
\verse{לב The heart חכם of the wise ישׂכיל teacheth פיהו his mouth, ועל to שׂפתיו his lips. יסיף and addeth לקח׃ learning}%
\verse{צוף a honeycomb, דבשׁ a honeycomb, אמרי words נעם Pleasant מתוק sweet לנפשׁ to the soul, ומרפא and health לעצם׃ to the bones.}%
\verse{ישׁ There is דרך a way ישׁר right לפני that seemeth אישׁ unto a man, ואחריתה but the end דרכי thereof the ways מות׃ of death.}%
\verse{נפשׁ He עמל that laboreth עמלה laboreth לו כי for himself; for אכף craveth עליו it of פיהו׃ his mouth}%
\verse{אישׁ man בליעל An ungodly כרה diggeth up רעה evil: ועל and in שׂפתיו his lips כאשׁ fire. צרבת׃ as a burning}%
\verse{אישׁ man תהפכות A froward ישׁלח soweth מדון strife: ונרגן and a whisperer מפריד separateth אלוף׃ chief friends.}%
\verse{אישׁ man חמס A violent יפתה enticeth רעהו his neighbor, והוליכו and leadeth בדרך him into the way לא not טוב׃ good.}%
\verse{עצה He shutteth עיניו his eyes לחשׁב to devise תהפכות froward things: קרץ moving שׂפתיו his lips כלה he bringeth evil to pass. רעה׃ he bringeth evil to pass.}%
\verse{עטרת head a crown תפארת of glory, שׂיבה The hoary בדרך in the way צדקה of righteousness. תמצא׃ it be found}%
\verse{טוב better ארך slow אפים to anger מגבור than the mighty; ומשׁל and he that ruleth ברוחו his spirit מלכד than he that taketh עיר׃ a city.}%
\verse{בחיק into the lap; יוטל is cast את הגורל The lot ומיהוה כל but the whole משׁפטו׃ disposing}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 17
\verse{טוב Better פת morsel, חרבה a dry ושׁלוה and quietness בה מבית therewith, than a house מלא full זבחי of sacrifices ריב׃ strife.}%
\verse{עבד servant משׂכיל A wise ימשׁל shall have rule בבן over a son מבישׁ that causeth shame, ובתוך among אחים the brethren. יחלק and shall have part נחלה׃ of the inheritance}%
\verse{מצרף The refining pot לכסף for silver, וכור and the furnace לזהב for gold: ובחן trieth לבות the hearts. יהוה׃ but the LORD}%
\verse{מרע A wicked doer מקשׁיב giveth heed על to שׂפת lips; און false שׁקר a liar מזין giveth ear על to לשׁון tongue. הות׃ a naughty}%
\verse{לעג Whoso mocketh לרשׁ the poor חרף reproacheth עשׂהו his Maker: שׂמח he that is glad לאיד at calamities לא shall not ינקה׃ be unpunished.}%
\verse{עטרת the crown זקנים of old men; בני Children's בנים children ותפארת and the glory בנים of children אבותם׃ their fathers.}%
\verse{לא not נאוה becometh לנבל a fool: שׂפת speech יתר Excellent אף much less כי much less לנדיב a prince. שׂפת lips שׁקר׃ do lying}%
\verse{אבן stone חן a precious השׁחד A gift בעיני in the eyes בעליו of him that hath אל it: whithersoever כל אשׁרפנה it turneth, ישׂכיל׃ it prospereth.}%
\verse{מכסה He that covereth פשׁע a transgression מבקשׁ seeketh אהבה love; ושׁנה but he that repeateth בדבר a matter מפריד separateth אלוף׃ friends.}%
\verse{תחת entereth more גערה A reproof במבין into a wise מהכות stripes כסיל into a fool. מאה׃ man than a hundred}%
\verse{אך only מרי rebellion: יבקשׁ seeketh רע An evil ומלאך messenger אכזרי therefore a cruel ישׁלח׃ shall be sent}%
\verse{פגושׁ of her whelps meet דב Let a bear שׁכול robbed באישׁ a man, ואל rather כסיל than a fool באולתו׃ in his folly.}%
\verse{משׁיב Whoso rewardeth רעה evil תחת for טובה good, לא shall not תמישׁ depart רעה evil מביתו׃ from his house.}%
\verse{פוטר when one letteth out מים water: ראשׁית The beginning מדון of strife ולפני before התגלע it be meddled with. הריב contention, נטושׁ׃ therefore leave off}%
\verse{מצדיק He that justifieth רשׁע the wicked, ומרשׁיע and he that condemneth צדיק the just, תועבת abomination יהוה גם even שׁניהם׃ they both}%
\verse{למה זהחיר a price ביד in the hand כסיל of a fool לקנות to get חכמה wisdom, ולב heart אין׃ seeing no}%
\verse{בכל at all עת times, אהב loveth הרע A friend ואח and a brother לצרה for adversity. יולד׃ is born}%
\verse{אדם A man חסר void לב of understanding תוקע striketh כף hands, ערב becometh ערבה surety לפני in the presence רעהו׃ of his friend.}%
\verse{אהב He loveth פשׁע transgression אהב that loveth מצה strife: מגביה he that exalteth פתחו his gate מבקשׁ seeketh שׁבר׃ destruction.}%
\verse{עקשׁ He that hath a froward לב heart לא no ימצא findeth טוב good: ונהפך and he that hath a perverse בלשׁונו tongue יפול falleth ברעה׃ into mischief.}%
\verse{ילד He that begetteth כסיל a fool לתוגה to his sorrow: לו ולא hath no ישׂמח joy. אבי and the father נבל׃ of a fool}%
\verse{לב heart שׂמח A merry ייטב doeth good גהה a medicine: ורוח spirit נכאה but a broken תיבשׁ drieth גרם׃ the bones.}%
\verse{שׁחד a gift מחיק out of the bosom רשׁע A wicked יקח taketh להטות to pervert ארחות the ways משׁפט׃ of judgment.}%
\verse{את before פני before מבין him that hath understanding; חכמה Wisdom ועיני but the eyes כסיל of a fool בקצה in the ends ארץ׃ of the earth.}%
\verse{כעס a grief לאביו to his father, בן son כסיל A foolish וממר and bitterness ליולדתו׃ to her that bore}%
\verse{גם Also ענושׁ to punish לצדיק the just לא not טוב good, להכות to strike נדיבים princes על for ישׁר׃ equity.}%
\verse{חושׂך spareth אמריו his words: יודע דעתקר רוחישׁ a man תבונה׃ of understanding}%
\verse{גם Even אויל a fool, מחרישׁ when he holdeth his peace, חכם wise: יחשׁב is counted אטם he that shutteth שׂפתיו his lips נבון׃ a man of understanding.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 18
\verse{לתאוה Through desire יבקשׁ seeketh נפרד a man, having separated himself, בכל with all תושׁיה wisdom. יתגלע׃ intermeddleth}%
\verse{לא hath no יחפץ delight כסיל A fool בתבונה in understanding, כי but אם but בהתגלות may discover itself. לבו׃ that his heart}%
\verse{בבוא cometh, רשׁע When the wicked בא cometh גם also בוז contempt, ועם and with קלון ignominy חרפה׃ reproach.}%
\verse{מים waters, עמקים deep דברי The words פי mouth אישׁ of a man's נחל brook. נבע a flowing מקור the wellspring חכמה׃ of wisdom}%
\verse{שׂאת to accept פני the person רשׁע of the wicked, לא not טוב good להטות to overthrow צדיק the righteous במשׁפט׃ in judgment.}%
\verse{שׂפתי lips כסיל A fool's יבאו enter בריב into contention, ופיו and his mouth למהלמות for strokes. יקרא׃ calleth}%
\verse{פי mouth כסיל A fool's מחתה his destruction, לו ושׂפתיו and his lips מוקשׁ the snare נפשׁו׃ of his soul.}%
\verse{דברי The words נרגן of a talebearer כמתלהמים as wounds, והם and they ירדו go down חדרי into the innermost parts בטן׃ of the belly.}%
\verse{גם also מתרפה that is slothful במלאכתו in his work אח is brother הוא He לבעל to him that is a great משׁחית׃}%
\verse{מגדל tower: עז a strong שׁם The name יהוה בו ירוץ runneth צדיק the righteous ונשׂגב׃ into it, and is safe.}%
\verse{הון wealth עשׁיר The rich man's קרית city, עזו his strong וכחומה wall נשׂגבה and as a high במשׂכיתו׃ in his own conceit.}%
\verse{לפני Before שׁבר destruction יגבה is haughty, לב the heart אישׁ of man ולפני and before כבוד honor ענוה׃ humility.}%
\verse{משׁיב He that answereth דבר a matter בטרם before ישׁמע he heareth אולת folly היא it לו וכלמה׃ and shame}%
\verse{רוח The spirit אישׁ of a man יכלכל will sustain מחלהו his infirmity; ורוח spirit נכאה but a wounded מי who ישׂאנה׃ can bear?}%
\verse{לב The heart נבון of the prudent יקנה getteth דעת knowledge; ואזן and the ear חכמים of the wise תבקשׁ seeketh דעת׃ knowledge.}%
\verse{מתן gift אדם A man's ירחיב maketh room לו ולפני him before גדלים great ינחנו׃ for him, and bringeth}%
\verse{צדיק just; הראשׁון first בריבו in his own cause יבא cometh רעהו וחקרו׃ and searcheth}%
\verse{מדינים causeth contentions ישׁבית to cease, הגורל The lot ובין between עצומים the mighty. יפריד׃ and parteth}%
\verse{אח A brother נפשׁע offended מקרית city: עז than a strong ומדונים כבריח like the bars ארמון׃ of a castle.}%
\verse{מפרי with the fruit פי of his mouth; אישׁ A man's תשׂבע shall be satisfied בטנו belly תבואת with the increase שׂפתיו of his lips ישׂבע׃ shall he be filled.}%
\verse{מות Death וחיים and life ביד in the power לשׁון of the tongue: ואהביה and they that love יאכל it shall eat פריה׃ the fruit}%
\verse{מצא findeth אשׁה a wife מצא findeth טוב a good ויפק and obtaineth רצון favor מיהוה׃}%
\verse{תחנונים entreaties; ידבר useth רשׁ The poor ועשׁיר but the rich יענה answereth עזות׃ roughly.}%
\verse{אישׁ A man רעים friends להתרעע must show himself friendly: וישׁ and there is אהב a friend דבק sticketh closer מאח׃ than a brother.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 19
\verse{טוב Better רשׁ the poor הולך that walketh בתמו in his integrity, מעקשׁ than perverse שׂפתיו in his lips, והוא כסיל׃ and is a fool.}%
\verse{גם Also, בלא without דעת knowledge, נפשׁ the soul לא not טוב good; ואץ and he that hasteth ברגלים with feet חוטא׃ sinneth.}%
\verse{אולת The foolishness אדם of man תסלף perverteth דרכו his way: ועל against יהוה the LORD. יזעף fretteth לבו׃ and his heart}%
\verse{הון Wealth יסיף maketh רעים friends; רבים many ודל but the poor מרעהו from his neighbor. יפרד׃ is separated}%
\verse{עד witness שׁקרים A false לא shall not ינקה be unpunished, ויפיח and speaketh כזבים lies לא shall not ימלט׃ escape.}%
\verse{רבים Many יחלו will entreat פני the favor נדיב of the prince: וכל and every man הרע a friend לאישׁ to him that giveth gifts. מתן׃ to him that giveth gifts.}%
\verse{כל All אחי the brethren רשׁ of the poor שׂנאהו do hate אף him: how much more כי him: how much more מרעהו do his friends רחקו go far ממנו from מרדף him? he pursueth אמרים words, לא wanting המה׃ they}%
\verse{קנה He that getteth לב wisdom אהב loveth נפשׁו his own soul: שׁמר he that keepeth תבונה understanding למצא shall find טוב׃ good.}%
\verse{עד witness שׁקרים A false לא shall not ינקה be unpunished, ויפיח and speaketh כזבים lies יאבד׃ shall perish.}%
\verse{לא is not נאוה seemly לכסיל for a fool; תענוג Delight אף much less כי for לעבד a servant משׁל to have rule בשׂרים׃ over princes.}%
\verse{שׂכל The discretion אדם of a man האריך deferreth אפו his anger; ותפארתו and his glory עבר to pass על over פשׁע׃ a transgression.}%
\verse{נהם as the roaring ככפיר of a lion; זעף wrath מלך The king's וכטל as dew על upon עשׂב the grass. רצונו׃ but his favor}%
\verse{הות the calamity לאביו of his father: בן son כסיל A foolish ודלף dropping. טרד a continual מדיני and the contentions אשׁה׃ of a wife}%
\verse{בית House והון and riches נחלת the inheritance אבות of fathers: ומיהוה אשׁה wife משׂכלת׃ and a prudent}%
\verse{עצלה Slothfulness תפיל casteth תרדמה into a deep sleep; ונפשׁ soul רמיה and an idle תרעב׃ shall suffer hunger.}%
\verse{שׁמר He that keepeth מצוה the commandment שׁמר keepeth נפשׁו his own soul; בוזה he that despiseth דרכיו his ways יומת׃ shall die.}%
\verse{מלוה lendeth יהוה unto the LORD; חונן He that hath pity דל upon the poor וגמלו and that which he hath given ישׁלם׃ will he pay him again.}%
\verse{יסר Chasten בנך thy son כי while ישׁ there is תקוה hope, ואל for המיתו his crying. אל and let not תשׂא spare נפשׁך׃ thy soul}%
\verse{גרל חמהשׂא shall suffer ענשׁ punishment: כי for אם if תציל thou deliver ועוד yet תוסף׃ thou must do it again.}%
\verse{שׁמע Hear עצה counsel, וקבל and receive מוסר instruction, למען that תחכם thou mayest be wise באחריתך׃ in thy latter}%
\verse{רבות many מחשׁבות devices בלב heart; אישׁ in a man's ועצת nevertheless the counsel יהוה היא that תקום׃ shall stand.}%
\verse{תאות The desire אדם of a man חסדו his kindness: וטוב better רשׁ and a poor מאישׁ man כזב׃ than a liar.}%
\verse{יראת The fear יהוה of the LORD לחיים to life: ושׂבע satisfied; ילין and shall abide בל he shall not יפקד be visited רע׃ with evil.}%
\verse{טמן hideth עצל A slothful ידו his hand בצלחת in bosom, גם so much אל as bring it to his mouth again. פיהו לא and will not ישׁיבנה׃}%
\verse{לץ a scorner, תכה Smite ופתי and the simple יערם will beware: והוכיח and reprove לנבון one that hath understanding, יבין he will understand דעת׃ knowledge.}%
\verse{משׁדד He that wasteth אב father, יבריח chaseth away אם mother, בן a son מבישׁ that causeth shame, ומחפיר׃ and bringeth reproach.}%
\verse{חדל Cease, בני my son, לשׁמע to hear מוסר the instruction לשׁגות to err מאמרי from the words דעת׃ of knowledge.}%
\verse{עד witness בליעל An ungodly יליץ scorneth משׁפט judgment: ופי and the mouth רשׁעים of the wicked יבלע devoureth און׃ iniquity.}%
\verse{נכונו are prepared ללצים for scorners, שׁפטים Judgments ומהלמות and stripes לגו for the back כסילים׃ of fools.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 20
\verse{לץ a mocker, היין Wine המה raging: שׁכר strong drink וכל and whosoever שׁגה is deceived בו לא thereby is not יחכם׃ wise.}%
\verse{נהם as the roaring ככפיר of a lion: אימת The fear מלך of a king מתעברו provoketh him to anger חוטא sinneth נפשׁו׃ his own soul.}%
\verse{כבוד an honor לאישׁ for a man שׁבת to cease מריב from strife: וכל but every אויל fool יתגלע׃ will be meddling.}%
\verse{מחרף by reason of the cold; עצל The sluggard לא will not יחרשׁ plow ישׁאל shall he beg בקציר in harvest, ואין׃ and nothing.}%
\verse{מים water; עמקים deep עצה Counsel בלב in the heart אישׁ of man ואישׁ but a man תבונה of understanding ידלנה׃ will draw it out.}%
\verse{רב Most אדם men יקרא will proclaim אישׁ every one חסדו his own goodness: ואישׁ man אמונים but a faithful מי who ימצא׃ can find?}%
\verse{מתהלך walketh בתמו in his integrity: צדיק The just אשׁרי blessed בניו his children אחריו׃ after}%
\verse{מלך A king יושׁב that sitteth על in כסא the throne דין of judgment מזרה scattereth away בעיניו with his eyes. כל all רע׃ evil}%
\verse{מי Who יאמר can say, זכיתי clean, לבי I have made my heart טהרתי I am pure מחטאתי׃ from my sin?}%
\verse{אבן ואבןיפה divers measures, ואיפה divers measures, תועבת abomination יהוה to the LORD. גם of them alike שׁניהם׃ both}%
\verse{גם Even במעלליו by his doings, יתנכר is known נער a child אם whether זך pure, ואם and whether ישׁר right. פעלו׃ his work}%
\verse{אזן ear, שׁמעת The hearing ועין eye, ראה and the seeing יהוה the LORD עשׂה hath made גם even שׁניהם׃ both}%
\verse{אל not תאהב Love שׁנה sleep, פן lest תורשׁ thou come to poverty; פקח open עיניך thine eyes, שׂבע thou shalt be satisfied לחם׃ with bread.}%
\verse{רע naught, רע naught, יאמר saith הקונה the buyer: ואזל but when he is gone his way, לו אז then יתהלל׃ he boasteth.}%
\verse{ישׁ There is זהב gold, ורב and a multitude פנינים of rubies: וכלי jewel. יקר a precious שׂפתי but the lips דעת׃ of knowledge}%
\verse{לקח Take בגדו his garment כי that ערב is surety זר a stranger: ובעד of him for נכרים a strange woman. חבלהו׃ and take a pledge}%
\verse{ערב לאישׁ to a man; לחם Bread שׁקר of deceit ואחר but afterwards ימלא shall be filled פיהו his mouth חצץ׃ with gravel.}%
\verse{מחשׁבות purpose בעצה by counsel: תכון is established ובתחבלות and with good advice עשׂה make מלחמה׃ war.}%
\verse{גולה revealeth סוד secrets: הולך He that goeth about רכיל a talebearer ולפתה with him that flattereth שׂפתיו with his lips. לא not תתערב׃ therefore meddle}%
\verse{מקלל Whoso curseth אביו his father ואמו or his mother, ידעך shall be put out נרו his lamp באישׁון in obscure חשׁך׃ darkness.}%
\verse{נחלה An inheritance מבחלת בראשׁנה at the beginning; ואחריתה but the end לא thereof shall not תברך׃ be blessed.}%
\verse{אל not תאמר Say אשׁלמה thou, I will recompense רע evil; קוה wait ליהוה on the LORD, וישׁע׃ and he shall save}%
\verse{תועבת an abomination יהוה unto the LORD; אבן ואבןמאזני balance מרמה and a false לא not טוב׃ good.}%
\verse{מיהוה מצעדי goings גבר Man's ואדם can a man מה how יבין then understand דרכו׃ his own way?}%
\verse{מוקשׁ a snare אדם to the man ילע devoureth קדשׁ holy, ואחר and after נדרים vows לבקר׃ to make inquiry.}%
\verse{מזרה scattereth רשׁעים the wicked, מלך king חכם A wise וישׁב and bringeth עליהם over אופן׃ the wheel}%
\verse{נר the candle יהוה of the LORD, נשׁמת The spirit אדם of man חפשׂ searching כל all חדרי the inward parts בטן׃ of the belly.}%
\verse{חסד Mercy ואמת and truth יצרו preserve מלך the king: וסעד is upheld בחסד by mercy. כסאו׃ and his throne}%
\verse{תפארת The glory בחורים of young men כחם their strength: והדר and the beauty זקנים of old men שׂיבה׃ the gray head.}%
\verse{חברות The blueness פצע of a wound תמריק ברע evil: ומכות so stripes חדרי the inward parts בטן׃ of the belly.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 21
\verse{פלגי the rivers מים of water: לב heart מלך The king's ביד in the hand יהוה of the LORD, על it whithersoever כל אשׁרחפץ he will. יטנו׃ he turneth}%
\verse{כל Every דרך way אישׁ of a man ישׁר right בעיניו in his own eyes: ותכן pondereth לבות the hearts. יהוה׃ but the LORD}%
\verse{עשׂה To do צדקה justice ומשׁפט and judgment נבחר more acceptable ליהוה to the LORD מזבח׃ than sacrifice.}%
\verse{רום עינים look, ורחב and a proud לב heart, נר the plowing רשׁעים of the wicked, חטאת׃ sin.}%
\verse{מחשׁבות The thoughts חרוץ of the diligent אך only למותר to plenteousness; וכל but of every one אץ hasty אך only למחסור׃ to want.}%
\verse{פעל The getting אוצרות of treasures בלשׁון tongue שׁקר by a lying הבל a vanity נדף tossed to and fro מבקשׁי of them that seek מות׃ death.}%
\verse{שׁד The robbery רשׁעים of the wicked יגורם shall destroy כי them; because מאנו they refuse לעשׂות to do משׁפט׃ judgment.}%
\verse{הפכפך froward דרך The way אישׁ of man וזר and strange: וזך but the pure, ישׁר right. פעלו׃ his work}%
\verse{טוב better לשׁבת to dwell על in פנת a corner גג of the housetop, מאשׁת woman מדינים than with a brawling ובית house. חבר׃ in a wide}%
\verse{נפשׁ The soul רשׁע of the wicked אותה desireth רע evil: לא findeth no יחן favor בעיניו in his eyes. רעהו׃ his neighbor}%
\verse{בענשׁ is punished, לץ When the scorner יחכם is made wise: פתי the simple ובהשׂכיל is instructed, לחכם and when the wise יקח he receiveth דעת׃ knowledge.}%
\verse{משׂכיל wisely considereth צדיק The righteous לבית the house רשׁע of the wicked: מסלף overthroweth רשׁעים the wicked לרע׃ for wickedness.}%
\verse{אטם Whoso stoppeth אזנו his ears מזעקת at the cry דל of the poor, גם he also הוא himself, יקרא shall cry ולא but shall not יענה׃ be heard.}%
\verse{מתן A gift בסתר in secret יכפה pacifieth אף anger: ושׁחד and a reward בחק in the bosom חמה wrath. עזה׃ strong}%
\verse{שׂמחה joy לצדיק to the just עשׂות to do משׁפט judgment: ומחתה but destruction לפעלי to the workers און׃ of iniquity.}%
\verse{אדם The man תועה that wandereth מדרך out of the way השׂכל of understanding בקהל in the congregation רפאים of the dead. ינוח׃ shall remain}%
\verse{אישׁ man: מחסור a poor אהב He that loveth שׂמחה pleasure אהב he that loveth יין wine ושׁמן and oil לא shall not יעשׁיר׃ be rich.}%
\verse{כפר a ransom לצדיק for the righteous, רשׁע The wicked ותחת for ישׁרים the upright. בוגד׃ and the transgressor}%
\verse{טוב better שׁבת to dwell בארץ in the wilderness, מדבר in the wilderness, מאשׁת woman. מדונים וכעס׃ and an angry}%
\verse{אוצר treasure נחמד to be desired ושׁמן and oil בנוה in the dwelling חכם of the wise; וכסיל but a foolish אדם man יבלענו׃ spendeth it up.}%
\verse{רדף He that followeth צדקה after righteousness וחסד and mercy ימצא findeth חיים life, צדקה righteousness, וכבוד׃ and honor.}%
\verse{עיר the city גברים of the mighty, עלה scaleth חכם A wise וירד and casteth down עז the strength מבטחה׃ of the confidence}%
\verse{שׁמר Whoso keepeth פיו his mouth ולשׁונו and his tongue שׁמר keepeth מצרות from troubles. נפשׁו׃ his soul}%
\verse{זד Proud יהיר haughty לץ scorner שׁמו his name, עושׂה who dealeth בעברת wrath. זדון׃ in proud}%
\verse{תאות The desire עצל of the slothful תמיתנו killeth כי him; for מאנו refuse ידיו his hands לעשׂות׃ to labor.}%
\verse{כל all היום the day התאוה He coveteth תאוה greedily וצדיק long: but the righteous יתן giveth ולא not. יחשׂך׃ and spareth}%
\verse{זבח The sacrifice רשׁעים of the wicked תועבה abomination: אף how much more, כי how much more, בזמה it with a wicked mind? יביאנו׃ he bringeth}%
\verse{עד witness כזבים A false יאבד shall perish: ואישׁ but the man שׁומע that heareth לנצח constantly. ידבר׃ speaketh}%
\verse{העז hardeneth אישׁ man רשׁע A wicked בפניו his face: וישׁר but the upright, הוא he יכין דרכיו׃ his way.}%
\verse{אין no חכמה wisdom ואין nor תבונה understanding ואין nor עצה counsel לנגד against יהוה׃}%
\verse{סוס The horse מוכן prepared ליום against the day מלחמה of battle: וליהוה of the LORD. התשׁועה׃ but safety}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 22
\verse{נבחר rather to be chosen שׁם A name מעשׁר riches, רב than great מכסף rather than silver ומזהב and gold. חן favor טוב׃ loving}%
\verse{עשׁיר The rich ורשׁ and poor נפגשׁו meet together: עשׂה the maker כלם of them all. יהוה׃ the LORD}%
\verse{ערום A prudent ראה foreseeth רעה the evil, ויסתר and hideth himself: ופתיים but the simple עברו pass on, ונענשׁו׃ and are punished.}%
\verse{עקב By ענוה humility יראת the fear יהוה of the LORD עשׁר riches, וכבוד and honor, וחיים׃ and life.}%
\verse{צנים Thorns פחים snares בדרך in the way עקשׁ of the froward: שׁומר he that doth keep נפשׁו his soul ירחק shall be far מהם׃}%
\verse{חנך Train up לנער a child על in פי he should go: דרכו the way גם and כי when יזקין he is old, לא he will not יסור depart ממנה׃ from}%
\verse{עשׁיר The rich ברשׁים over the poor, ימשׁול ruleth ועבד servant לוה and the borrower לאישׁ to the lender. מלוה׃ to the lender.}%
\verse{זורע He that soweth עולה iniquity יקצור shall reap און ושׁבט and the rod עברתו of his anger יכלה׃ shall fail.}%
\verse{טוב that hath a bountiful עין eye הוא He יברך shall be blessed; כי for נתן he giveth מלחמו of his bread לדל׃ to the poor.}%
\verse{גרשׁ Cast out לץ the scorner, ויצא shall go out; מדון and contention וישׁבת shall cease. דין yea, strife וקלון׃ and reproach}%
\verse{אהב He that loveth טהור לב pureness חן the grace שׂפתיו of his lips רעהו his friend. מלך׃ the king}%
\verse{עיני The eyes יהוה of the LORD נצרו preserve דעת knowledge, ויסלף and he overthroweth דברי the words בגד׃ of the transgressor.}%
\verse{אמר saith, עצל The slothful ארי a lion בחוץ without, בתוך in רחבות the streets. ארצח׃ I shall be slain}%
\verse{שׁוחה pit: עמקה a deep פי The mouth זרות of strange women זעום he that is abhorred יהוה of the LORD יפול shall fall שׁם׃}%
\verse{אולת Foolishness קשׁורה bound בלב in the heart נער of a child; שׁבט the rod מוסר of correction ירחיקנה shall drive it far ממנו׃ from}%
\verse{עשׁק He that oppresseth דל the poor להרבות to increase לו נתן his he that giveth לעשׁיר to the rich, אך surely למחסור׃ to want.}%
\verseWithHeading{Words of the Wise}{הט Bow down אזנך thine ear, ושׁמע and hear דברי the words חכמים of the wise, ולבך thine heart תשׁית and apply לדעתי׃ unto my knowledge.}%
\verse{כי For נעים a pleasant כי thing if תשׁמרם thou keep בבטנך them within יכנו be fitted יחדו thee; they shall withal על in שׂפתיך׃ thy lips.}%
\verse{להיות may be ביהוה in the LORD, מבטחך That thy trust הודעתיך I have made known היום to thee this day, אף even אתה׃ to thee.}%
\verse{הלא Have not כתבתי I written לך שׁלשׁום במעצות in counsels ודעת׃ and knowledge,}%
\verse{להודיעך That I might make thee know קשׁט the certainty אמרי of the words אמת of truth; להשׁיב that thou mightest answer אמרים the words אמת of truth לשׁלחיך׃ to them that send}%
\verse{אל not תגזל Rob דל the poor, כי because דל poor: הוא he ואל neither תדכא oppress עני the afflicted בשׁער׃ in the gate:}%
\verse{כי For יהוה יריב will plead ריבם their cause, וקבע and spoil את קבעיהם of those that spoiled נפשׁ׃ the soul}%
\verse{אל Make no תתרע friendship את with בעל man; אף an angry ואת and with אישׁ man חמות a furious לא thou shalt not תבוא׃ go:}%
\verse{פן Lest תאלף thou learn ארחתו his ways, ולקחת and get מוקשׁ a snare לנפשׁך׃ to thy soul.}%
\verse{אל not תהי Be בתקעי thou of them that strike כף hands, בערבים of them that are sureties משׁאות׃ for debts.}%
\verse{אם If אין thou hast nothing לך לשׁלם to pay, למה why יקח should he take away משׁכבך thy bed מתחתיך׃ from under}%
\verse{אל not תסג Remove גבול landmark, עולם the ancient אשׁר which עשׂו have set. אבותיך׃ thy fathers}%
\verse{חזית Seest אישׁ thou a man מהיר diligent במלאכתו in his business? לפני before מלכים kings; יתיצב he shall stand בל he shall not יתיצב stand לפני before חשׁכים׃ mean}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 23
\verse{כי When תשׁב thou sittest ללחום to eat את with מושׁל a ruler, בין consider diligently תבין consider diligently את אשׁר what לפניך׃ before}%
\verse{ושׂמת And put שׂכין a knife בלעך to thy throat, אם if בעל a man נפשׁ given to appetite. אתה׃ thou}%
\verse{אל Be not תתאו desirous למטעמותיו of his dainties: והוא for they לחם meat. כזבים׃ deceitful}%
\verse{אל not תיגע Labor להעשׁיר to be rich: מבינתך thine own wisdom. חדל׃ cease}%
\verse{התעוף Wilt thou set עיניך thine eyes בו ואיננו upon that which is not? כי for עשׂה certainly make יעשׂה certainly make לו כנפים themselves wings; כנשׁר as an eagle ועיף they fly away השׁמים׃ toward heaven.}%
\verse{אל thou not תלחם Eat את לחם the bread רע of an evil עין eye, ואל neither תתאו desire למטעמתיו׃ thou his dainty meats:}%
\verse{כי For כמו as שׁער he thinketh בנפשׁו in his heart, כן so הוא he: אכל Eat ושׁתה and drink, יאמר saith לך ולבו he to thee; but his heart בל not עמך׃ with thee.}%
\verse{פתך The morsel אכלת thou hast eaten תקיאנה shalt thou vomit up, ושׁחת and lose דבריך words. הנעימים׃ thy sweet}%
\verse{באזני in the ears כסיל of a fool: אל not תדבר Speak כי for יבוז he will despise לשׂכל the wisdom מליך׃ of thy words.}%
\verse{אל not תסג Remove גבול landmark; עולם the old ובשׂדי into the fields יתומים of the fatherless: אל not תבא׃ and enter}%
\verse{כי For גאלם their redeemer חזק mighty; הוא he יריב shall plead את ריבם their cause אתך׃ with}%
\verse{הביאה Apply למוסר unto instruction, לבך thine heart ואזנך and thine ears לאמרי to the words דעת׃ of knowledge.}%
\verse{אל not תמנע Withhold מנער from the child: מוסר correction כי for תכנו thou beatest בשׁבט him with the rod, לא he shall not ימות׃ die.}%
\verse{אתה Thou בשׁבט him with the rod, תכנו shalt beat ונפשׁו his soul משׁאול from hell. תציל׃ and shalt deliver}%
\verse{בני My son, אם if חכם be wise, לבך thine heart ישׂמח shall rejoice, לבי my heart גם even אני׃ mine.}%
\verse{ותעלזנה shall rejoice, כליותי Yea, my reins בדבר speak שׂפתיך when thy lips מישׁרים׃ right things.}%
\verse{אל Let not יקנא envy לבך thine heart בחטאים sinners: כי but אם ביראת in the fear יהוה of the LORD כל all היום׃ the day}%
\verse{כי For אם surely ישׁ there is אחרית an end; ותקותך and thine expectation לא shall not תכרת׃ be cut off.}%
\verse{שׁמע Hear אתה thou, בני my son, וחכם and be wise, ואשׁר and guide בדרך in the way. לבך׃ thine heart}%
\verse{אל not תהי Be בסבאי among winebibbers; יין among winebibbers; בזללי among riotous eaters בשׂר׃ of flesh:}%
\verse{כי For סבא the drunkard וזולל and the glutton יורשׁ shall come to poverty: וקרעים with rags. תלבישׁ shall clothe נומה׃ and drowsiness}%
\verse{שׁמע Hearken לאביך unto thy father זה that ילדך begot ואל not תבוז thee, and despise כי when זקנה she is old. אמך׃ thy mother}%
\verse{אמת the truth, קנה Buy ואל not; תמכר and sell חכמה wisdom, ומוסר and instruction, ובינה׃ and understanding.}%
\verse{גול shall greatly rejoice: יגול shall greatly rejoice: אבי The father צדיק of the righteous יולד and he that begetteth חכם a wise וישׂמח shall have joy בו׃}%
\verse{ישׂמח shall be glad, אביך Thy father ואמך and thy mother ותגל thee shall rejoice. יולדתך׃ and she that bore}%
\verse{תנה give בני My son, לבך me thine heart, לי ועיניך and let thine eyes דרכי my ways. תרצנה׃}%
\verse{כי For שׁוחה ditch; עמקה a deep זונה a whore ובאר pit. צרה woman a narrow נכריה׃ and a strange}%
\verse{אף also היא She כחתף as a prey, תארב lieth in wait ובוגדים the transgressors באדם among men. תוסף׃ and increaseth}%
\verse{למי Who אוי hath woe? למי who אבוי hath sorrow? למי who מדונים למי who שׂיח hath babbling? למי who פצעים hath wounds חנם without cause? למי who חכללות hath redness עינים׃ of eyes?}%
\verse{למאחרים They that tarry long על at היין the wine; לבאים they that go לחקר to seek ממסך׃ mixed wine.}%
\verse{אל not תרא Look יין thou upon the wine כי when יתאדם it is red, כי when יתן it giveth בכיס עינו his color יתהלך it moveth itself במישׁרים׃ aright.}%
\verse{אחריתו At the last כנחשׁ like a serpent, ישׁך it biteth וכצפעני like an adder. יפרשׁ׃ and stingeth}%
\verse{עיניך Thine eyes יראו shall behold זרות strange women, ולבך and thine heart ידבר shall utter תהפכות׃ perverse things.}%
\verse{והיית Yea, thou shalt be כשׁכב as he that lieth down בלב in the midst ים of the sea, וכשׁכב or as he that lieth בראשׁ upon the top חבל׃ of a mast.}%
\verse{הכוני They have stricken בל me, I was not חליתי sick; הלמוני they have beaten בל not: ידעתי me, I felt מתי when אקיץ shall I awake? אוסיף it yet אבקשׁנו I will seek עוד׃ again.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 24
\verse{אל Be not תקנא thou envious באנשׁי men, רעה against evil ואל neither תתאו desire להיות to be אתם׃ with}%
\verse{כי For שׁד destruction, יהגה studieth לבם their heart ועמל of mischief. שׂפתיהם and their lips תדברנה׃ talk}%
\verse{בחכמה Through wisdom יבנה built; בית is a house ובתבונה and by understanding יתכונן׃ it is established:}%
\verse{ובדעת And by knowledge חדרים shall the chambers ימלאו be filled כל with all הון riches. יקר precious ונעים׃ and pleasant}%
\verse{גבר man חכם A wise בעוז strong; ואישׁ yea, a man דעת of knowledge מאמץ increaseth כח׃ strength.}%
\verse{כי For בתחבלות by wise counsel תעשׂה thou shalt make לך מלחמה thy war: ותשׁועה safety. ברב and in multitude יועץ׃ of counselors}%
\verse{ראמות too high לאויל for a fool: חכמות Wisdom בשׁער in the gate. לא not יפתח he openeth פיהו׃ his mouth}%
\verse{מחשׁב He that deviseth להרע to do evil לו בעל person. מזמות a mischievous יקראו׃ shall be called}%
\verse{זמת The thought אולת of foolishness חטאת sin: ותועבת an abomination לאדם to men. לץ׃ and the scorner}%
\verse{התרפית thou faint ביום in the day צרה of adversity, צר small. כחכה׃ thy strength}%
\verse{הצל to deliver לקחים drawn למות unto death, ומטים and ready להרג to be slain; אם If תחשׂוך׃ thou forbear}%
\verse{כי If תאמר thou sayest, הן Behold, לא not; ידענו we knew זה it הלא doth not תכן he that pondereth לבות הוא doth he יבין consider ונצר and he that keepeth נפשׁך thy soul, הוא ידע know והשׁיב and shall he render לאדם to man כפעלו׃ according to his works?}%
\verse{אכל eat בני My son, דבשׁ thou honey, כי because טוב good; ונפת and the honeycomb, מתוק sweet על to חכך׃ thy taste:}%
\verse{כן So דעה the knowledge חכמה of wisdom לנפשׁך unto thy soul: אם when מצאת thou hast found וישׁ then there shall be אחרית a reward, ותקותך and thy expectation לא shall not תכרת׃ be cut off.}%
\verse{אל not תארב רשׁע O wicked לנוה against the dwelling צדיק of the righteous; אל תשׁדד spoil רבצו׃ his resting place:}%
\verse{כי For שׁבע seven יפול falleth צדיק a just וקם times, and riseth up again: ורשׁעים but the wicked יכשׁלו shall fall ברעה׃ into mischief.}%
\verse{בנפל falleth, אויביך when thine enemy אל not תשׂמח Rejoice ובכשׁלו when he stumbleth: אל and let not יגל be glad לבך׃ thine heart}%
\verse{פן Lest יראה see יהוה ורע and it displease בעיניו , and it displease והשׁיב him, and he turn away מעליו from אפו׃ his wrath}%
\verse{אל neither תתחר במרעים because of evil אל תקנא be thou envious ברשׁעים׃ at the wicked;}%
\verse{כי For לא no תהיה there shall be אחרית reward לרע to the evil נר the candle רשׁעים of the wicked ידעך׃ shall be put out.}%
\verse{ירא fear את יהוה thou the LORD בני My son, ומלך and the king: עם with שׁונים them that are given to change: אל not תתערב׃ meddle}%
\verse{כי For פתאם suddenly; יקום shall rise אידם their calamity ופיד the ruin שׁניהם of them both? מי and who יודע׃ knoweth}%
\verse{גם also אלה These לחכמים to the wise. הכר to have respect פנים of persons במשׁפט in judgment. בל not טוב׃ good}%
\verse{אמר He that saith לרשׁע unto the wicked, צדיק righteous; אתה Thou יקבהו curse, עמים him shall the people יזעמוהו shall abhor לאמים׃ nations}%
\verse{ולמוכיחים But to them that rebuke ינעם shall be delight, ועליהם upon תבוא shall come ברכת blessing טוב׃ and a good}%
\verse{שׂפתים lips ישׁק shall kiss משׁיב that giveth דברים answer. נכחים׃ a right}%
\verse{הכן Prepare בחוץ without, מלאכתך thy work ועתדה and make it fit בשׂדה for thyself in the field; לך אחר and afterwards ובנית build ביתך׃ thine house.}%
\verse{אל not תהי Be עד a witness חנם without cause; ברעך against thy neighbor והפתית and deceive בשׂפתיך׃ with thy lips.}%
\verse{אל not, תאמר Say כאשׁר to him as עשׂה I will do לי כן so אעשׂה he hath done לו אשׁיב to me: I will render לאישׁ to the man כפעלו׃ according to his work.}%
\verse{על by שׂדה the field אישׁ of the slothful, עצל of the slothful, עברתי I went ועל and by כרם the vineyard אדם of the man חסר void לב׃ of understanding;}%
\verse{והנה And, lo, עלה grown over כלו it was all קמשׂנים nettles כסו had covered פניו the face חרלים with thorns, וגדר wall אבניו thereof, and the stone נהרסה׃ thereof was broken down.}%
\verse{ואחזה saw, אנכי Then I אשׁית considered well: לבי considered well: ראיתי I looked לקחתי upon received מוסר׃ instruction.}%
\verse{מעט a little שׁנות sleep, מעט a little תנומות slumber, מעט a little חבק folding ידים of the hands לשׁכב׃ to sleep:}%
\verse{ובא come מתהלך one that traveleth; רישׁך So shall thy poverty ומחסריך and thy want כאישׁ man. מגן׃ as an armed}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 25
\verseWithHeading{More Proverbs of Solomon}{גם also אלה These משׁלי proverbs שׁלמה of Solomon, אשׁר which העתיקו copied out. אנשׁי the men חזקיה of Hezekiah מלך king יהודה׃ of Judah}%
\verse{כבד the glory אלהים of God הסתר to conceal דבר a thing: וכבד but the honor מלכים of kings חקר to search out דבר׃ a matter.}%
\verse{שׁמים The heaven לרום for height, וארץ and the earth לעמק for depth, ולב and the heart מלכים of kings אין unsearchable. חקר׃ unsearchable.}%
\verse{הגו Take away סיגים the dross מכסף from the silver, ויצא and there shall come forth לצרף for the refiner. כלי׃ a vessel}%
\verse{הגו Take away רשׁע the wicked לפני before מלך the king, ויכון shall be established בצדק in righteousness. כסאו׃ and his throne}%
\verse{אל not תתהדר לפני in the presence מלך of the king, ובמקום in the place גדלים of great אל תעמד׃ and stand}%
\verse{כי For טוב better אמר that it be said לך עלה unto thee, Come up הנה hither; מהשׁפילך than that thou shouldest be put lower לפני in the presence נדיב of the prince אשׁר whom ראו have seen. עיניך׃ thine eyes}%
\verse{אל תצארב to strive, מהר hastily פן lest מה what תעשׂה to do באחריתה in the end בהכלים אתךעך׃ thereof, when thy neighbor}%
\verse{ריבך Debate ריב thy cause את with רעך thy neighbor וסוד a secret אחר to another: אל not תגל׃ and discover}%
\verse{פן Lest יחסדך put thee to shame, שׁמע he that heareth ודבתך and thine infamy לא turn not away. תשׁוב׃ turn not away.}%
\verse{תפוחי apples זהב of gold במשׂכיות in pictures כסף of silver. דבר A word דבר spoken על fitly אפניו׃ fitly}%
\verse{נזם an earring זהב of gold, וחלי and an ornament כתם of fine gold, מוכיח reprover חכם a wise על upon אזן ear. שׁמעת׃ an obedient}%
\verse{כצנת As the cold שׁלג of snow ביום in the time קציר of harvest, ציר messenger נאמן a faithful לשׁלחיו to them that send ונפשׁ the soul אדניו of his masters. ישׁיב׃ him: for he refresheth}%
\verse{נשׂיאים clouds ורוח and wind וגשׁם rain. אין without אישׁ Whoso מתהלל boasteth himself במתת gift שׁקר׃ of a false}%
\verse{בארך By long אפים forbearing יפתה persuaded, קצין is a prince ולשׁון tongue רכה and a soft תשׁבר breaketh גרם׃ the bone.}%
\verse{דבשׁ honey? מצאת Hast thou found אכל eat דיך so much as is sufficient פן for thee, lest תשׂבענו thou be filled והקאתו׃ therewith, and vomit}%
\verse{הקר Withdraw רגלך thy foot מבית house; רעך from thy neighbor's פן lest ישׂבעך he be weary ושׂנאך׃ of thee, and hate}%
\verse{מפיץ a maul, וחרב and a sword, וחץ arrow. שׁנון and a sharp אישׁ A man ענה that beareth ברעהו against his neighbor עד witness שׁקר׃ false}%
\verse{שׁן tooth, רעה a broken ורגל and a foot מועדת out of joint. מבטח Confidence בוגד in an unfaithful man ביום in time צרה׃ of trouble}%
\verse{מעדה he that taketh away בגד a garment ביום weather, קרה in cold חמץ vinegar על upon נתר niter, ושׁר so he that singeth בשׁרים songs על to לב heart. רע׃ a heavy}%
\verse{אם If רעב שׂנאך thine enemy האכלהו to eat; לחם give him bread ואם and if צמא he be thirsty, השׁקהו to drink: מים׃ give him water}%
\verse{כי For גחלים coals of fire אתה thou חתה shalt heap על upon ראשׁו his head, ויהוה ישׁלם׃ shall reward}%
\verse{רוח wind צפון The north תחולל driveth away גשׁם rain: ופנים countenance נזעמים so an angry לשׁון tongue. סתר׃ a backbiting}%
\verse{טוב better שׁבת to dwell על in פנת the corner גג of the housetop, מאשׁת מדוניםבית house. חבר׃ and in a wide}%
\verse{מים waters קרים cold על to נפשׁ soul, עיפה a thirsty ושׁמועה news טובה so good מארץ מרחק׃}%
\verse{מעין fountain, נרפשׂ a troubled ומקור spring. משׁחת and a corrupt צדיק A righteous מט man falling down לפני before רשׁע׃ the wicked}%
\verse{אכל to eat דבשׁ honey: הרבות much לא not טוב good וחקר so to search כבדם their own glory כבוד׃ glory.}%
\verse{עיר a city פרוצה broken down, אין no חומה walls. אישׁ He אשׁר that אין without מעצר rule over לרוחו׃ his own spirit}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 26
\verse{כשׁלג As snow בקיץ in summer, וכמטר and as rain בקציר in harvest, כן so לא is not נאוה seemly לכסיל for a fool. כבוד׃ honor}%
\verse{כצפור As the bird לנוד by wandering, כדרור as the swallow לעוף by flying, כן so קללת the curse חנם causeless לא shall not תבא׃ come.}%
\verse{שׁוט A whip לסוס for the horse, מתג a bridle לחמור for the ass, ושׁבט and a rod לגו back. כסילים׃ for the fool's}%
\verse{אל not תען Answer כסיל a fool כאולתו according to his folly, פן lest תשׁוה be like לו גם also אתה׃ thou}%
\verse{ענה Answer כסיל a fool כאולתו according to his folly, פן lest יהיה he be חכם wise בעיניו׃ in his own conceit.}%
\verse{מקצה cutteth off רגלים the feet, חמס damage. שׁתה drinketh שׁלח He that sendeth דברים a message ביד by the hand כסיל׃ of a fool}%
\verse{דליו are not equal: שׁקים The legs מפסח of the lame ומשׁל so a parable בפי in the mouth כסילים׃ of fools.}%
\verse{כצרור As he that bindeth אבן a stone במרגמה in a sling, כן so נותן he that giveth לכסיל to a fool. כבוד׃ honor}%
\verse{חוח a thorn עלה goeth up ביד into the hand שׁכור of a drunkard, ומשׁל so a parable בפי in the mouth כסילים׃ of fools.}%
\verse{רב The great מחולל that formed כל all ושׂכר both rewardeth כסיל the fool, ושׂכר and rewardeth עברים׃ transgressors.}%
\verse{ככלב As a dog שׁב returneth על to קאו his vomit, כסיל a fool שׁונה returneth באולתו׃ to his folly.}%
\verse{ראית Seest אישׁ thou a man חכם wise בעיניו in his own conceit? תקוה more hope לכסיל of a fool ממנו׃ than of}%
\verse{אמר saith, עצל The slothful שׁחל a lion בדרך in the way; ארי a lion בין in הרחבות׃ the streets.}%
\verse{הדלת the door תסוב turneth על upon צירה his hinges, ועצל so the slothful על upon מטתו׃ his bed.}%
\verse{טמן hideth עצל The slothful ידו his hand בצלחת in bosom; נלאה it grieveth להשׁיבה him to bring it again אל to פיו׃ his mouth.}%
\verse{חכם wiser עצל The sluggard בעיניו in his own conceit משׁבעה than seven משׁיבי men that can render טעם׃ a reason.}%
\verse{מחזיק to him, one that taketh באזני by the ears. כלב a dog עבר He that passeth by, מתעבר meddleth על with ריב strife לא׃ not}%
\verse{כמתלהלה As a mad הירה who casteth זקים firebrands, חצים arrows, ומות׃ and death,}%
\verse{כן So אישׁ the man רמה deceiveth את רעהו his neighbor, ואמר and saith, הלא Am not משׂחק in sport? אני׃ I}%
\verse{באפס Where no עצים wood תכבה goeth out: אשׁ is, the fire ובאין so where no נרגן talebearer, ישׁתק ceaseth. מדון׃ the strife}%
\verse{פחם coals לגחלים to burning coals, ועצים and wood לאשׁ to fire; ואישׁ man מדונים לחרחר to kindle ריב׃ strife.}%
\verse{דברי The words נרגן of a talebearer כמתלהמים as wounds, והם and they ירדו go down חדרי into the innermost parts בטן׃ of the belly.}%
\verse{כסף with silver סיגים dross. מצפה covered על חרשׂ a potsherd שׂפתים lips דלקים Burning ולב heart רע׃ and a wicked}%
\verse{בשׂפתו with his lips, ינכר dissembleth שׂונא He that hateth ובקרבו within ישׁית and layeth up מרמה׃ deceit}%
\verse{כי When יחנן fair, קולו he speaketh אל him not: תאמן believe בו כי for שׁבע seven תועבות abominations בלבו׃ in his heart.}%
\verse{תכסה is covered שׂנאה hatred במשׁאון by deceit, תגלה shall be showed רעתו his wickedness בקהל׃ before the congregation.}%
\verse{כרה Whoso diggeth שׁחת a pit בה יפל shall fall וגלל therein: and he that rolleth אבן a stone, אליו upon תשׁוב׃ it will return}%
\verse{לשׁון tongue שׁקר A lying ישׂנא hateth דכיו afflicted ופה mouth חלק by it; and a flattering יעשׂה worketh מדחה׃ ruin.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 27
\verse{אל תתהלליום of tomorrow; מחר of tomorrow; כי for לא not תדע thou knowest מה what ילד may bring forth. יום׃ a day}%
\verse{יהללך praise זר Let another man ולא thee, and not פיך thine own mouth; נכרי a stranger, ואל and not שׂפתיך׃ thine own lips.}%
\verse{כבד heavy, אבן A stone ונטל weighty; החול and the sand וכעס wrath אויל but a fool's כבד heavier משׁניהם׃ than them both.}%
\verse{אכזריות cruel, חמה Wrath ושׁטף outrageous; אף and anger ומי but who יעמד able to stand לפני before קנאה׃ envy?}%
\verse{טובה better תוכחת rebuke מגלה Open מאהבה love. מסתרת׃ than secret}%
\verse{נאמנים Faithful פצעי the wounds אוהב of a friend; ונעתרות deceitful. נשׁיקות but the kisses שׂונא׃ of an enemy}%
\verse{נפשׁ soul שׂבעה The full תבוס loatheth נפת a honeycomb; ונפשׁ soul רעבה but to the hungry כל every מר bitter thing מתוק׃ is sweet.}%
\verse{כצפור As a bird נודדת that wandereth מן from קנה her nest, כן so אישׁ a man נודד that wandereth ממקומו׃ from his place.}%
\verse{שׁמן Ointment וקטרת and perfume ישׂמח rejoice לב the heart: ומתק so the sweetness רעהו of a man's friend מעצת נפשׁ׃}%
\verse{רעך Thine own friend, ורעה אביך and thy father's אל not; תעזב forsake ובית house אחיך into thy brother's אל neither תבוא go ביום in the day אידך of thy calamity: טוב better שׁכן a neighbor קרוב near מאח than a brother רחוק׃ far off.}%
\verse{חכם be wise, בני My son, ושׂמח לבי and make my heart ואשׁיבה חרפי him that reproacheth דבר׃}%
\verse{ערום A prudent ראה foreseeth רעה the evil, נסתר hideth himself; פתאים the simple עברו pass on, נענשׁו׃ are punished.}%
\verse{קח Take בגדו his garment כי that ערב is surety זר for a stranger, ובעד of him for נכריה a strange woman. חבלהו׃ and take a pledge}%
\verse{מברך He that blesseth רעהו his friend בקול voice, גדול with a loud בבקר in the morning, השׁכים rising early קללה a curse תחשׁב׃ it shall be counted}%
\verse{דלף dropping טורד A continual ביום day סגריר in a very rainy ואשׁת woman מדונים נשׁתוה׃ are alike.}%
\verse{צפניה Whosoever hideth צפן her hideth רוח the wind, ושׁמן and the ointment ימינו of his right hand, יקרא׃ betrayeth}%
\verse{ברזל Iron בברזל iron; יחד sharpeneth ואישׁ so a man יחד sharpeneth פני the countenance רעהו׃ of his friend.}%
\verse{נצר Whoso keepeth תאנה the fig tree יאכל shall eat פריה the fruit ושׁמר thereof: so he that waiteth אדניו on his master יכבד׃ shall be honored.}%
\verse{כמים As in water הפנים face לפנים to face, כן so לב the heart האדם of man לאדם׃ to man.}%
\verse{שׁאול Hell ואבדה and destruction לא are never תשׂבענה full; ועיני so the eyes האדם of man לא are never תשׂבענה׃ satisfied.}%
\verse{מצרף the refining pot לכסף for silver, וכור and the furnace לזהב for gold; ואישׁ so a man לפי to מהללו׃ to his praise.}%
\verse{אם Though תכתושׁ thou shouldest bray את האויל a fool במכתשׁ in a mortar בתוך among הריפות wheat בעלי with a pestle, לא will not תסור depart מעליו from אולתו׃ his foolishness}%
\verse{ידע תדעני the state צאנך of thy flocks, שׁית look לבך look לעדרים׃ well to thy herds.}%
\verse{כי For לא not לעולם forever: חסן riches ואם and doth נזר the crown לדור to every generation? דור׃ to every generation?}%
\verse{גלה appeareth, חציר The hay ונראה showeth itself, דשׁא and the tender grass ונאספו are gathered. עשׂבות and herbs הרים׃ of the mountains}%
\verse{כבשׂים The lambs ללבושׁך for thy clothing, ומחיר the price שׂדה of the field. עתודים׃ and the goats}%
\verse{ודי enough חלב milk עזים And goats' ללחמך for thy food, ללחם for the food ביתך of thy household, וחיים and the maintenance לנערותיך׃ for thy maidens.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 28
\verse{נסו flee ואין when no רדף man pursueth: רשׁע The wicked וצדיקים but the righteous ככפיר as a lion. יבטח׃ are bold}%
\verse{בפשׁע For the transgression ארץ of a land רבים many שׂריה the princes ובאדם thereof: but by a man מבין of understanding ידע knowledge כן the state יאריך׃ shall be prolonged.}%
\verse{גבר man רשׁ A poor ועשׁק that oppresseth דלים the poor מטר rain סחף a sweeping ואין which leaveth no לחם׃ food.}%
\verse{עזבי They that forsake תורה the law יהללו praise רשׁע the wicked: ושׁמרי but such as keep תורה the law יתגרו׃ contend}%
\verse{אנשׁי רע Evil לא not יבינו understand משׁפט judgment: ומבקשׁי but they that seek יהוה יבינו understand כל׃ all}%
\verse{טוב Better רשׁ the poor הולך that walketh בתמו in his uprightness, מעקשׁ than perverse דרכים ways, והוא though he עשׁיר׃ rich.}%
\verse{נוצר Whoso keepeth תורה the law בן son: מבין a wise ורעה but he that is a companion זוללים of riotous יכלים shameth אביו׃ his father.}%
\verse{מרבה increaseth הונו his substance, בנשׁך He that by usury ובתרבית and unjust gain לחונן it for him that will pity דלים the poor. יקבצנו׃ he shall gather}%
\verse{מסיר He that turneth away אזנו his ear משׁמע from hearing תורה the law, גם even תפלתו his prayer תועבה׃ abomination.}%
\verse{משׁגה to go astray ישׁרים Whoso causeth the righteous בדרך way, רע in an evil בשׁחותו into his own pit: הוא himself יפול he shall fall ותמימים but the upright ינחלו in possession. טוב׃ shall have good}%
\verse{חכם wise בעיניו in his own conceit; אישׁ man עשׁיר The rich ודל but the poor מבין that hath understanding יחקרנו׃ searcheth him out.}%
\verse{בעלץ do rejoice, צדיקים When righteous רבה great תפארת glory: ובקום rise, רשׁעים but when the wicked יחפשׂ is hidden. אדם׃ a man}%
\verse{מכסה He that covereth פשׁעיו his sins לא shall not יצליח prosper: ומודה but whoso confesseth ועזב and forsaketh ירחם׃ shall have mercy.}%
\verse{אשׁרי Happy אדם the man מפחד that feareth תמיד always: ומקשׁה but he that hardeneth לבו his heart יפול shall fall ברעה׃ into mischief.}%
\verse{ארי lion, נהם a roaring ודב bear; שׁוקק and a ranging משׁל ruler רשׁע a wicked על over עם people. דל׃ the poor}%
\verse{נגיד The prince חסר that wanteth תבונות understanding ורב also a great מעשׁקות oppressor: שׂנאי he that hateth בצע covetousness יאריך shall prolong ימים׃ days.}%
\verse{אדם A man עשׁק that doeth violence בדם to the blood נפשׁ of person עד to בור the pit; ינוס shall flee אל let no יתמכו׃ man stay}%
\verse{הולך Whoso walketh תמים uprightly יושׁע shall be saved: ונעקשׁ but perverse דרכים ways יפול shall fall באחת׃ at once.}%
\verse{עבד He that tilleth אדמתו his land ישׂבע shall have plenty לחם of bread: ומרדף but he that followeth רקים after vain ישׂבע enough. רישׁ׃ shall have poverty}%
\verse{אישׁ man אמונות A faithful רב shall abound ברכות with blessings: ואץ but he that maketh haste להעשׁיר to be rich לא shall not ינקה׃ be innocent.}%
\verse{הכר To have respect פנים of persons לא not טוב good: ועל for for פת a piece לחם of bread יפשׁע will transgress. גבר׃ man}%
\verse{נבהל that hasteth להון to be rich אישׁ He רע an evil עין eye, ולא not ידע and considereth כי that חסר poverty יבאנו׃ shall come upon}%
\verse{מוכיח He that rebuketh אדם a man אחרי afterwards חן more favor ימצא shall find ממחליק than he that flattereth לשׁון׃ with the tongue.}%
\verse{גוזל Whoso robbeth אביו his father ואמו or his mother, ואמר and saith, אין no פשׁע transgression; חבר the companion הוא the same לאישׁ of a destroyer. משׁחית׃ of a destroyer.}%
\verse{רחב He that is of a proud נפשׁ heart יגרה stirreth up מדון strife: ובוטח but he that putteth his trust על in יהוה the LORD ידשׁן׃ shall be made fat.}%
\verse{בוטח that trusteth בלבו in his own heart הוא He כסיל is a fool: והולך but whoso walketh בחכמה wisely, הוא he ימלט׃ shall be delivered.}%
\verse{נותן He that giveth לרשׁ unto the poor אין shall not מחסור lack: ומעלים but he that hideth עיניו his eyes רב shall have many מארות׃ a curse.}%
\verse{בקום rise, רשׁעים When the wicked יסתר hide themselves: אדם men ובאבדם but when they perish, ירבו increase. צדיקים׃ the righteous}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 29
\verse{אישׁ He, תוכחות that being often reproved מקשׁה hardeneth ערף neck, פתע shall suddenly ישׁבר be destroyed, ואין and that without מרפא׃ remedy.}%
\verse{ברבות are in authority, צדיקים When the righteous ישׂמח rejoice: העם the people ובמשׁל beareth rule, רשׁע but when the wicked יאנח mourn. עם׃ the people}%
\verse{אישׁ Whoso אהב loveth חכמה wisdom ישׂמח rejoiceth אביו his father: ורעה but he that keepeth company זונות with harlots יאבד spendeth הון׃ substance.}%
\verse{מלך The king במשׁפט by judgment יעמיד establisheth ארץ the land: ואישׁ but he תרומות that receiveth gifts יהרסנה׃ overthroweth}%
\verse{גבר A man מחליק that flattereth על that flattereth רעהו his neighbor רשׁת a net פורשׂ spreadeth על for פעמיו׃ his feet.}%
\verse{בפשׁע In the transgression אישׁ man רע of an evil מוקשׁ a snare: וצדיק but the righteous ירון doth sing ושׂמח׃ and rejoice.}%
\verse{ידע considereth צדיק The righteous דין the cause דלים of the poor: רשׁע the wicked לא not יבין regardeth דעת׃ to know}%
\verse{אנשׁי men לצון Scornful יפיחו bring a city into a snare: קריה bring a city into a snare: וחכמים but wise ישׁיבו turn away אף׃ wrath.}%
\verse{אישׁ man חכם a wise נשׁפט contendeth את with אישׁ man, אויל a foolish ורגז whether he rage ושׂחק or laugh, ואין no נחת׃ rest.}%
\verse{אנשׁי דמיםשׂנאו hate תם the upright: וישׁרים but the just יבקשׁו seek נפשׁו׃ his soul.}%
\verse{כל all רוחו his mind: יוציא uttereth כסיל A fool וחכם but a wise באחור it in till afterwards. ישׁבחנה׃ keepeth}%
\verse{משׁל If a ruler מקשׁיב hearken על to דבר lies, שׁקר lies, כל all משׁרתיו his servants רשׁעים׃ wicked.}%
\verse{רשׁ The poor ואישׁ man תככים and the deceitful נפגשׁו meet together: מאיר lighteneth עיני their eyes. שׁניהם both יהוה׃ the LORD}%
\verse{מלך The king שׁופט judgeth באמת that faithfully דלים the poor, כסאו his throne לעד forever. יכון׃ shall be established}%
\verse{שׁבט The rod ותוכחת and reproof יתן give חכמה wisdom: ונער but a child משׁלח left מבישׁ to shame. אמו׃ bringeth his mother}%
\verse{ברבות are multiplied, רשׁעים When the wicked ירבה increaseth: פשׁע transgression וצדיקים but the righteous במפלתם their fall. יראו׃ shall see}%
\verse{יסר Correct בנך thy son, ויניחך and he shall give thee rest; ויתן yea, he shall give מעדנים delight לנפשׁך׃ unto thy soul.}%
\verse{באין Where no חזון vision, יפרע perish: עם the people ושׁמר but he that keepeth תורה the law, אשׁרהו׃ happy}%
\verse{בדברים by words: לא will not יוסר be corrected עבד A servant כי for though יבין he understand ואין he will not מענה׃ answer.}%
\verse{חזית Seest אישׁ thou a man אץ hasty בדבריו in his words? תקוה more hope לכסיל of a fool ממנו׃ than of}%
\verse{מפנק He that delicately bringeth up מנער from a child עבדו his servant ואחריתו at the length. יהיה shall have him become מנון׃ son}%
\verse{אישׁ man אף An angry יגרה stirreth up מדון strife, ובעל man חמה and a furious רב aboundeth פשׁע׃ in transgression.}%
\verse{גאות pride אדם A man's תשׁפילנו shall bring him low: ושׁפל the humble רוח in spirit. יתמך shall uphold כבוד׃ but honor}%
\verse{חולק Whoso is partner עם with גנב a thief שׂונא hateth נפשׁו his own soul: אלה cursing, ישׁמע he heareth ולא not. יגיד׃ and betrayeth}%
\verse{חרדת The fear אדם of man יתן bringeth מוקשׁ a snare: ובוטח but whoso putteth his trust ביהוה in the LORD ישׂגב׃ shall be safe.}%
\verse{רבים Many מבקשׁים seek פני favor; מושׁל the ruler's ומיהוה משׁפט judgment אישׁ׃ but man's}%
\verse{תועבת an abomination צדיקים to the just: אישׁ man עול An unjust ותועבת abomination רשׁע to the wicked. ישׁר and upright דרך׃ in the way}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 30
\verseWithHeading{The Oracle of Agur}{דברי The words אגור of Agur בן the son יקה of Jakeh, המשׂא the prophecy: נאם spoke הגבר the man לאיתיאל unto Ithiel, לאיתיאל even unto Ithiel ואכל׃ and Ucal,}%
\verse{כי Surely בער more brutish אנכי I מאישׁ than man, ולא and have not בינת the understanding אדם׃ of a man.}%
\verse{ולא I neither למדתי learned חכמה wisdom, ודעת the knowledge קדשׁים of the holy. אדע׃ nor have}%
\verse{מי Who עלה hath ascended up שׁמים into heaven, וירד or descended? מי who אסף hath gathered רוח the wind בחפניו in his fists? מי who צרר hath bound מים the waters בשׂמלה in a garment? מי who הקים hath established כל all אפסי the ends ארץ of the earth? מה what שׁמו his name, ומה and what שׁם name, בנו his son's כי if תדע׃ thou canst tell?}%
\verse{כל Every אמרת word אלוה of God צרופה pure: מגן a shield הוא he לחסים׃ unto them that put their trust}%
\verse{אל thou not תוסף Add על unto דבריו his words, פן lest יוכיח he reprove בך ונכזבת׃ thee, and thou be found a liar.}%
\verse{שׁתים Two שׁאלתי have I required מאתך אל me not תמנע thee; deny ממני of בטרם before אמות׃ I die:}%
\verse{שׁוא me vanity ודבר and lies: כזב and lies: הרחק Remove far ממני from ראשׁ poverty ועשׁר nor riches; אל me neither תתן give לי הטריפני feed לחם me with food חקי׃ convenient}%
\verse{פן Lest אשׂבע I be full, וכחשׁתי and deny ואמרתי and say, מי Who יהוה ופן or lest אורשׁ I be poor, וגנבתי and steal, ותפשׂתי and take שׁם the name אלהי׃ of my God}%
\verse{אל not תלשׁן Accuse עבד a servant אל unto אדנו his master, פן lest יקללך he curse ואשׁמת׃ thee, and thou be found guilty.}%
\verse{דור a generation אביו their father, יקלל curseth ואת אמו their mother. לא and doth not יברך׃ bless}%
\verse{דור a generation טהור pure בעיניו in their own eyes, ומצאתו from their filthiness. לא and is not רחץ׃ washed}%
\verse{דור a generation, מה O how רמו lofty עיניו are their eyes! ועפעפיו and their eyelids ינשׂאו׃ are lifted up.}%
\verse{דור a generation, חרבות swords, שׁניו whose teeth ומאכלות knives, מתלעתיו and their jaw teeth לאכל to devour עניים the poor מארץ from off the earth, ואביונים and the needy מאדם׃ from men.}%
\verse{לעלוקה The horseleach שׁתי hath two בנות daughters, הב Give, הב give. שׁלושׁ הנהא are never תשׂבענה satisfied, ארבע four לא not, אמרו say הון׃ enough:}%
\verse{שׁאול The grave; ועצר and the barren רחם womb; ארץ the earth לא is not שׂבעה filled מים with water; ואשׁ and the fire לא not, אמרה saith הון׃ enough.}%
\verse{עין The eye תלעג mocketh לאב at father, ותבוז and despiseth ליקהת to obey אם mother, יקרוה shall pick it out, ערבי the ravens נחל of the valley ויאכלוה shall eat בני and the young נשׁר׃ eagles}%
\verse{שׁלשׁה There be three המה נפלאו are too wonderful ממני for וארבע me, yea, four לא not: ידעתים׃ which I know}%
\verse{דרך The way הנשׁר of an eagle בשׁמים in the air; דרך the way נחשׁ of a serpent עלי upon צור a rock; דרך the way אניה of a ship בלב in the midst ים of the sea; ודרך and the way גבר of a man בעלמה׃ with a maid.}%
\verse{כן Such דרך the way אשׁה woman; מנאפת of an adulterous אכלה she eateth, ומחתה and wipeth פיה her mouth, ואמרה and saith, לא no פעלתי I have done און׃ wickedness.}%
\verse{תחת For שׁלושׁ three רגזה is disquieted, ארץ the earth ותחת and for ארבע four לא it cannot תוכל it cannot שׂאת׃ bear:}%
\verse{תחת For עבד a servant כי when ימלוך he reigneth; ונבל and a fool כי when ישׂבע he is filled לחם׃ with meat;}%
\verse{תחת For שׂנואה an odious כי when תבעל she is married; ושׁפחה and a handmaid כי that תירשׁ is heir גברתה׃ to her mistress.}%
\verse{ארבעה There be four הם but they קטני little ארץ upon the earth, והמה חכמים exceeding wise: מחכמים׃ exceeding wise:}%
\verse{הנמלים The ants עם a people לא not עז strong, ויכינו yet they prepare בקיץ in the summer; לחמם׃ their meat}%
\verse{שׁפנים The conies עם folk, לא a feeble עצום a feeble וישׂימו yet make בסלע in the rocks; ביתם׃ they their houses}%
\verse{מלך king, אין have no לארבה The locusts ויצא yet go they forth חצץ of them by bands; כלו׃ all}%
\verse{שׂממית The spider בידים with her hands, תתפשׂ taketh hold והיא בהיכלי palaces. מלך׃ and is in kings'}%
\verse{שׁלשׁה There be three המה מיטיבי well, צעד which go וארבעה yea, four מיטבי are comely לכת׃ in going:}%
\verse{לישׁ A lion גבור strongest בבהמה among beasts, ולא and turneth not away ישׁוב and turneth not away מפני for כל׃ any;}%
\verse{זרזיר מתניםו also; תישׁ a he goat ומלך and a king, אלקום עמו׃ against}%
\verse{אם If נבלת thou hast done foolishly בהתנשׂא in lifting up thyself, ואם or if זמות thou hast thought evil, יד thine hand לפה׃ upon thy mouth.}%
\verse{כי Surely מיץ the churning חלב of milk יוציא bringeth forth חמאה butter, ומיץ and the wringing אף of the nose יוציא bringeth forth דם blood: ומיץ so the forcing אפים of wrath יוציא bringeth forth ריב׃ strife.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 31
\verseWithHeading{The Oracle of King Lemuel}{דברי The words למואל Lemuel, מלך of king משׂא the prophecy אשׁר that יסרתו taught אמו׃ his mother}%
\verse{מה What, ברי my son? ומה and what, בר the son בטני of my womb? ומה and what, בר the son נדרי׃ of my vows?}%
\verse{אל not תתן Give לנשׁים unto women, חילך thy strength ודרכיך nor thy ways למחות to that which destroyeth מלכין׃ kings.}%
\verse{אל not למלכים for kings, למואל O Lemuel, אל not למלכים for kings שׁתו to drink יין wine; ולרוזנים for princes או nor שׁכר׃ strong drink:}%
\verse{פן Lest ישׁתה they drink, וישׁכח and forget מחקק the law, וישׁנה and pervert דין the judgment כל of any בני of the afflicted. עני׃ of the afflicted.}%
\verse{תנו Give שׁכר strong drink לאובד unto him that is ready to perish, ויין and wine למרי unto those that be of heavy נפשׁ׃ hearts.}%
\verse{ישׁתה Let him drink, וישׁכח and forget רישׁו his poverty, ועמלו his misery לא no יזכר and remember עוד׃ more.}%
\verse{פתח Open פיך thy mouth לאלם for the dumb אל in דין the cause כל of all בני such as are appointed חלוף׃ to destruction.}%
\verse{פתח Open פיך thy mouth, שׁפט judge צדק righteously, ודין and plead the cause עני of the poor ואביון׃ and needy.}%
\verseWithHeading{An Excellent Woman}{אשׁת woman? חיל a virtuous מי Who ימצא can find ורחק far מפנינים above rubies. מכרה׃ for her price}%
\verse{בטח doth safely trust בה לב The heart בעלה of her husband ושׁלל of spoil. לא in her, so that he shall have no יחסר׃ need}%
\verse{גמלתהו She will do טוב him good ולא and not רע evil כל all ימי the days חייה׃ of her life.}%
\verse{דרשׁה She seeketh צמר wool, ופשׁתים and flax, ותעשׂ and worketh בחפץ willingly כפיה׃ with her hands.}%
\verse{היתה She is כאניות ships; סוחר like the merchants' ממרחק from afar. תביא she bringeth לחמה׃ her food}%
\verse{ותקם She riseth בעוד also while it is yet לילה night, ותתן and giveth טרף meat לביתה to her household, וחק and a portion לנערתיה׃ to her maidens.}%
\verse{זממה She considereth שׂדה a field, ותקחהו and buyeth מפרי it: with the fruit כפיה of her hands נטע she planteth כרם׃ a vineyard.}%
\verse{חגרה She girdeth בעוז with strength, מתניה her loins ותאמץ and strengtheneth זרעותיה׃ her arms.}%
\verse{טעמה She perceiveth כי that טוב good: סחרה her merchandise לא goeth not out יכבה goeth not out בליל by night. נרה׃ her candle}%
\verse{ידיה her hands שׁלחה She layeth בכישׁור to the spindle, וכפיה and her hands תמכו hold פלך׃ the distaff.}%
\verse{כפה her hand פרשׂה She stretcheth out לעני to the poor; וידיה her hands שׁלחה yea, she reacheth forth לאביון׃ to the needy.}%
\verse{לא She is not תירא afraid לביתה for her household: משׁלג of the snow כי for כל all ביתה her household לבשׁ clothed שׁנים׃ with scarlet.}%
\verse{מרבדים herself coverings of tapestry; עשׂתה She maketh לה שׁשׁ silk וארגמן and purple. לבושׁה׃ her clothing}%
\verse{נודע is known בשׁערים in the gates, בעלה Her husband בשׁבתו when he sitteth עם among זקני the elders ארץ׃ of the land.}%
\verse{סדין fine linen, עשׂתה She maketh ותמכר and selleth וחגור girdles נתנה and delivereth לכנעני׃ unto the merchant.}%
\verse{עז Strength והדר and honor לבושׁה her clothing; ותשׂחק and she shall rejoice ליום in time אחרון׃ to come.}%
\verse{פיה her mouth פתחה She openeth בחכמה with wisdom; ותורת the law חסד of kindness. על and in לשׁונה׃ her tongue}%
\verse{צופיה She looketh well הליכות to the ways ביתה of her household, ולחם the bread עצלות of idleness. לא not תאכל׃ and eateth}%
\verse{קמו arise up, בניה Her children ויאשׁרוה and call her blessed; בעלה her husband ויהללה׃ and he praiseth}%
\verse{רבות Many בנות daughters עשׂו have done חיל virtuously, ואת but thou עלית excellest על excellest כלנה׃ them all.}%
\verse{שׁקר deceitful, החן Favor והבל vain: היפי and beauty אשׁה a woman יראת feareth יהוה היא she תתהלל׃ shall be praised.}%
\verse{תנו Give לה מפרי her of the fruit ידיה of her hands; ויהללוה praise בשׁערים her in the gates. מעשׂיה׃ and let her own works}%
\end{biblechapter}%
