\biblebook{Psalm}
\begin{biblechapter}% Psalm 1
\verseWithHeading{The Ways of the Righteous and the Wicked}{אשׁרי Blessed האישׁ the man אשׁר that לא not הלך walketh בעצת in the counsel רשׁעים of the ungodly, ובדרך in the way חטאים of sinners, לא nor עמד standeth ובמושׁב in the seat לצים of the scornful. לא nor ישׁב׃ sitteth}%
\verse{כי אםתורת in the law יהוה חפצו his delight ובתורתו and in his law יהגה doth he meditate יומם day ולילה׃ and night.}%
\verse{והיה And he shall be כעץ like a tree שׁתול planted על by פלגי the rivers מים of water, אשׁר that פריו his fruit יתן bringeth forth בעתו in his season; ועלהו his leaf לא also shall not יבול wither; וכל and whatsoever אשׁר and whatsoever יעשׂה he doeth יצליח׃ shall prosper.}%
\verse{לא not כן so: הרשׁעים The ungodly כי but אם but כמץ like the chaff אשׁר which תדפנו driveth away. רוח׃ the wind}%
\verse{על כןא shall not יקמו stand רשׁעים the ungodly במשׁפט in the judgment, וחטאים nor sinners בעדת in the congregation צדיקים׃ of the righteous.}%
\verse{כי For יודע knoweth יהוה the LORD דרך the way צדיקים of the righteous: ודרך but the way רשׁעים of the ungodly תאבד׃ shall perish.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 2
\verseWithHeading{The Messiah’s Reign}{למה Why רגשׁו rage, גוים do the heathen ולאמים and the people יהגו imagine ריק׃ a vain thing?}%
\verse{יתיצבו set themselves, מלכי The kings ארץ of the earth ורוזנים and the rulers נוסדו take counsel יחד together, על against יהוה the LORD, ועל and against משׁיחו׃ his anointed,}%
\verse{ננתקה אתוסרותימו ונשׁליכה and cast away ממנו from עבתימו׃ their cords}%
\verse{יושׁב He that sitteth בשׁמים in the heavens ישׂחק shall laugh: אדני the Lord ילעג׃ shall have them in derision.}%
\verse{אז Then ידבר shall he speak אלימו unto באפו them in his wrath, ובחרונו them in his sore displeasure. יבהלמו׃ and vex}%
\verse{ואני Yet have I נסכתי set מלכי my king על upon ציון of Zion. הר hill קדשׁי׃ my holy}%
\verse{אספרה אל unto חק the decree: יהוה אמר hath said אלי בני my Son; אתה me, Thou אני have I היום this day ילדתיך׃ begotten}%
\verse{שׁאל Ask ממני of ואתנה me, and I shall give גוים the heathen נחלתך thine inheritance, ואחזתך thy possession. אפסי and the uttermost parts ארץ׃ of the earth}%
\verse{תרעם Thou shalt break בשׁבט them with a rod ברזל of iron; ככלי vessel. יוצר like a potter's תנפצם׃ thou shalt dash them in pieces}%
\verse{ועתה now מלכים therefore, O ye kings: השׂכילו Be wise הוסרו be instructed, שׁפטי ye judges ארץ׃ of the earth.}%
\verse{עבדו Serve את יהוה the LORD ביראה with fear, וגילו and rejoice ברעדה׃ with trembling.}%
\verse{נשׁקו Kiss בר the Son, פן lest יאנף he be angry, ותאבדו and ye perish דרך the way, כי when יבער is kindled כמעט but a little. אפו his wrath אשׁרי Blessed כל all חוסי׃ they that put their trust}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 3
\verseWithHeading{A Call to Adonai in Distress}{מזמור A Psalm לדוד of David, בברחו when he fled מפני from אבשׁלום Absalom בנו׃ his son. יהוה LORD, מה how רבו are they increased צרי that trouble רבים me! many קמים they that rise up עלי׃ against}%
\verse{רבים Many אמרים which say לנפשׁי of my soul, אין no ישׁועתה help לו באלהים for him in God. סלה׃ Selah.}%
\verse{ואתה But thou, יהוה O LORD, מגן a shield בעדי for כבודי me; my glory, ומרים and the lifter up ראשׁי׃ of mine head.}%
\verse{קולי with my voice, אל unto יהוה אקרא I cried ויענני and he heard מהר hill. קדשׁו me out of his holy סלה׃ Selah.}%
\verse{אני I שׁכבתי laid me down ואישׁנה and slept; הקיצותי I awaked; כי for יהוה the LORD יסמכני׃ sustained}%
\verse{לא I will not אירא be afraid מרבבות of ten thousands עם of people, אשׁר that סביב me round about. שׁתו have set עלי׃ against}%
\verse{קומה Arise, יהוה O LORD; הושׁיעני save אלהי me, O my God: כי for הכית thou hast smitten את כל all איבי mine enemies לחי the cheek bone; שׁני the teeth רשׁעים of the ungodly. שׁברת׃ thou hast broken}%
\verse{ליהוה unto the LORD: הישׁועה Salvation על upon עמך thy people. ברכתך thy blessing סלה׃ Selah.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 4
\verseWithHeading{Safety in Adonai}{למנצח To the chief Musician בנגינות on Neginoth, מזמור A Psalm לדוד׃ of David. בקראי me when I call, ענני Hear אלהי O God צדקי of my righteousness: בצר me in distress; הרחבת thou hast enlarged לי חנני have mercy ושׁמע upon me, and hear תפלתי׃ my prayer.}%
\verse{בני O ye sons אישׁ of men, עד how long מה how long כבודי my glory לכלמה into shame? תאהבון will ye love ריק vanity, תבקשׁו seek after כזב leasing? סלה׃ Selah.}%
\verse{ודעו But know כי that הפלה hath set apart יהוה the LORD חסיד him that is godly לו יהוה for himself: the LORD ישׁמע will hear בקראי when I call אליו׃ unto}%
\verse{רגזו Stand in awe, ואל not: תחטאו and sin אמרו commune בלבבכם with your own heart על upon משׁכבכם your bed, ודמו and be still. סלה׃ Selah.}%
\verse{זבחו Offer זבחי the sacrifices צדק of righteousness, ובטחו and put your trust אל in יהוה׃ the LORD.}%
\verse{רבים many אמרים that say, מי Who יראנו will show טוב us good? נסה lift thou up עלינו upon אור the light פניך of thy countenance יהוה׃ LORD,}%
\verse{נתתה Thou hast put שׂמחה gladness בלבי in my heart, מעת more than in the time דגנם their corn ותירושׁם and their wine רבו׃ increased.}%
\verse{בשׁלום in peace, יחדו I will both אשׁכבה lay me down ואישׁן and sleep: כי for אתה thou, יהוה LORD, לבדד only לבטח in safety. תושׁיבני׃ makest me dwell}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 5
\verseWithHeading{A Prayer for Guidance and Protection}{למנצח To the chief Musician אל upon הנחילות Nehiloth, מזמור A Psalm לדוד׃ of David. אמרי to my words, האזינה Give ear יהוה O LORD, בינה consider הגיגי׃ my meditation.}%
\verse{הקשׁיבה Hearken לקול unto the voice שׁועי of my cry, מלכי my King, ואלהי and my God: כי for אליך unto אתפלל׃ thee will I pray.}%
\verse{יהוה O LORD; בקר in the morning, תשׁמע shalt thou hear קולי My voice בקר in the morning אערך will I direct לך ואצפה׃ unto thee, and will look up.}%
\verse{כי For לא not אל a God חפץ that hath pleasure רשׁע in wickedness: אתה thou לא neither יגרך dwell רע׃ shall evil}%
\verse{לא shall not יתיצבו stand הוללים The foolish לנגד עיניך thy sight: שׂנאת thou hatest כל all פעלי workers און׃ of iniquity.}%
\verse{תאבד Thou shalt destroy דברי them that speak כזב leasing: אישׁ man. דמים the bloody ומרמה and deceitful יתעב will abhor יהוה׃ the LORD}%
\verse{ואני But as for me, ברב in the multitude חסדך of thy mercy: אבוא I will come ביתך thy house אשׁתחוה will I worship אל toward היכל temple. קדשׁך thy holy ביראתך׃ in thy fear}%
\verse{יהוה me, O LORD, נחני Lead בצדקתך in thy righteousness למען because of שׁוררי mine enemies; הושׁר straight לפני before my face. דרכך׃ make thy way}%
\verse{כי For אין no בפיהו in their mouth; נכונה faithfulness קרבם their inward part הוות very wickedness; קבר sepulcher; פתוח an open גרונם their throat לשׁונם with their tongue. יחליקון׃ they flatter}%
\verse{האשׁימם Destroy אלהים thou them, O God; יפלו let them fall ממעצותיהם by their own counsels; ברב in the multitude פשׁעיהם of their transgressions; הדיחמו cast them out כי for מרו׃ they have rebelled}%
\verse{וישׂמחו in thee rejoice: כל But let all חוסי those that put their trust בך לעולם let them ever ירננו shout for joy, ותסך because thou defendest עלימו them: ויעלצו be joyful בך אהבי let them also that love שׁמך׃ thy name}%
\verse{כי For אתה thou, תברך wilt bless צדיק the righteous; יהוה LORD, כצנה him as a shield. רצון with favor תעטרנו׃ wilt thou compass}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 6
\verseWithHeading{An Appeal for Forgiveness and Deliverance}{למנצח To the chief Musician בנגינות on Neginoth על upon השׁמינית Sheminith, מזמור A Psalm לדוד׃ of David. יהוה O LORD, אל me not באפך in thine anger, תוכיחני rebuke ואל neither בחמתך me in thy hot displeasure. תיסרני׃ chasten}%
\verse{חנני Have mercy יהוה כי for אמלל weak: אני I רפאני heal יהוה כי me; for נבהלו are vexed. עצמי׃ my bones}%
\verse{ונפשׁי My soul נבהלה vexed: מאד is also sore ואת but thou, יהוה O LORD, עד how long? מתי׃ how long?}%
\verse{שׁובה Return, יהוה O LORD, חלצה deliver נפשׁי my soul: הושׁיעני oh save למען חסדך׃}%
\verse{כי For אין no במות in death זכרך remembrance בשׁאול of thee: in the grave מי who יודה׃ shall give thee thanks?}%
\verse{יגעתי I am weary באנחתי with my groaning; אשׂחה to swim; בכל all לילה the night מטתי make I my bed בדמעתי with my tears. ערשׂי my couch אמסה׃ I water}%
\verse{עשׁשׁה is consumed מכעס because of grief; עיני Mine eye עתקה it waxeth old בכל because of all צוררי׃ mine enemies.}%
\verse{סורו Depart ממני from כל me, all פעלי ye workers און of iniquity; כי for שׁמע hath heard יהוה the LORD קול the voice בכיי׃ of my weeping.}%
\verse{שׁמע hath heard יהוה The LORD תחנתי my supplication; יהוה the LORD תפלתי my prayer. יקח׃ will receive}%
\verse{יבשׁו be ashamed ויבהלו vexed: מאד and sore כל Let all איבי mine enemies ישׁבו let them return יבשׁו be ashamed רגע׃ suddenly.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 7
\verseWithHeading{Prayer for Deliverance from Enemies}{שׁגיון Shiggaion לדוד of David, אשׁר which שׁר he sang ליהוה unto the LORD, על concerning דברי the words כושׁ of Cush בן־ימיני׃ יהוה O LORD אלהי my God, בך חסיתי in thee do I put my trust: הושׁיעני save מכל me from all רדפי them that persecute והצילני׃ me, and deliver}%
\verse{פן Lest יטרף he tear כאריה like a lion, נפשׁי my soul פרק rending in pieces, ואין while none מציל׃ to deliver.}%
\verse{יהוה O LORD אלהי my God, אם if עשׂיתי I have done זאת this; אם if ישׁ there be עול iniquity בכפי׃ in my hands;}%
\verse{אם If גמלתי I have rewarded שׁולמי unto him that was at peace רע evil ואחלצה with me; (yea, I have delivered צוררי is mine enemy:) ריקם׃ him that without cause}%
\verse{ירדף persecute אויב Let the enemy נפשׁי my soul, וישׂג and take וירמס yea, let him tread down לארץ upon the earth, חיי my life וכבודי mine honor לעפר in the dust. ישׁכן and lay סלה׃ Selah.}%
\verse{קומה Arise, יהוה O LORD, באפך in thine anger, הנשׂא lift up thyself בעברות because of the rage צוררי of mine enemies: ועורה and awake אלי משׁפט me the judgment צוית׃ thou hast commanded.}%
\verse{ועדת So shall the congregation לאמים of the people תסובבך compass thee about: ועליה for their sakes למרום thou on high. שׁובה׃ therefore return}%
\verse{יהוה The LORD ידין shall judge עמים the people: שׁפטני judge יהוה me, O LORD, כצדקי according to my righteousness, וכתמי and according to mine integrity עלי׃ in}%
\verse{יגמר come to an end; נא Oh רע let the wickedness רשׁעים of the wicked ותכונן but establish צדיק the just: ובחן trieth לבות וכליות and reins. אלהים God צדיק׃ for the righteous}%
\verse{מגני My defense על of אלהים God, מושׁיע which saveth ישׁרי the upright לב׃ in heart.}%
\verse{אלהים God שׁופט judgeth צדיק the righteous, ואל and God זעם is angry בכל every יום׃ day.}%
\verse{אם If לא not, ישׁוב he turn חרבו his sword; ילטושׁ he will whet קשׁתו his bow, דרך he hath bent ויכוננה׃ and made it ready.}%
\verse{ולו הכין He hath also prepared כלי for him the instruments מות of death; חציו his arrows לדלקים against the persecutors. יפעל׃ he ordaineth}%
\verse{הנה Behold, יחבל he travaileth און with iniquity, והרה and hath conceived עמל mischief, וילד and brought forth שׁקר׃ falsehood.}%
\verse{בור a pit, כרה He made ויחפרהו and digged ויפל it, and is fallen בשׁחת into the ditch יפעל׃ he made.}%
\verse{ישׁוב shall return עמלו His mischief בראשׁו upon his own head, ועל upon קדקדו his own pate. חמסו and his violent dealing ירד׃ shall come down}%
\verse{אודה I will praise יהוה the LORD כצדקו according to his righteousness: ואזמרה and will sing praise שׁם to the name יהוה of the LORD עליון׃ most high.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 8
\verseWithHeading{Adonai’s Glory in Creation}{למנצח To the chief Musician על upon הגתית Gittith, מזמור A Psalm לדוד׃ of David. יהוה O LORD אדנינו our Lord, מה how אדיר excellent שׁמך thy name בכל in all הארץ the earth! אשׁר who תנה hast set הודך thy glory על above השׁמים׃ the heavens.}%
\verse{מפי עוללים of babes וינקים and sucklings יסדת hast thou ordained עז strength למען because of צורריך thine enemies, להשׁבית that thou mightest still אויב the enemy ומתנקם׃ and the avenger.}%
\verse{כי When אראה I consider שׁמיך thy heavens, מעשׂי the work אצבעתיך of thy fingers, ירח the moon וכוכבים and the stars, אשׁר which כוננתה׃ thou hast ordained;}%
\verse{מה What אנושׁ is man, כי that תזכרנו thou art mindful ובן of him? and the son אדם of man, כי that תפקדנו׃ thou visitest}%
\verse{ותחסרהו מעטאלהים than the angels, וכבוד him with glory והדר and honor. תעטרהו׃ and hast crowned}%
\verse{תמשׁילהו Thou madest him to have dominion במעשׂי over the works ידיך of thy hands; כל all שׁתה thou hast put תחת under רגליו׃ his feet:}%
\verse{צנה sheep ואלפים and oxen, כלם All וגם yea, בהמות and the beasts שׂדי׃ of the field;}%
\verse{צפור The fowl שׁמים of the air, ודגי and the fish הים of the sea, עבר passeth through ארחות the paths ימים׃ of the seas.}%
\verse{יהוה O LORD אדנינו our Lord, מה how אדיר excellent שׁמך thy name בכל in all הארץ׃ the earth!}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 9
\verseWithHeading{Praise for Adonai’s Justice}{למנצח To the chief Musician עלמות upon לבן Muth-labben, מזמור A Psalm לדוד׃ of David. אודה I will praise יהוה O LORD, בכל with my whole לבי heart; אספרה I will show forth כל all נפלאותיך׃ thy marvelous works.}%
\verse{אשׂמחה I will be glad ואעלצה and rejoice בך אזמרה in thee: I will sing praise שׁמך to thy name, עליון׃ O thou most High.}%
\verse{בשׁוב are turned אויבי When mine enemies אחור back, יכשׁלו they shall fall ויאבדו and perish מפניך׃ at thy presence.}%
\verse{כי For עשׂית thou hast maintained משׁפטי my right ודיני and my cause; ישׁבת thou satest לכסא in the throne שׁופט judging צדק׃ right.}%
\verse{גערת Thou hast rebuked גוים the heathen, אבדת thou hast destroyed רשׁע the wicked, שׁמם their name מחית thou hast put out לעולם forever ועד׃ and ever.}%
\verse{האויב O thou enemy, תמו are come to a perpetual end: חרבות destructions לנצח are come to a perpetual end: וערים cities; נתשׁת and thou hast destroyed אבד is perished זכרם their memorial המה׃ with them.}%
\verse{ויהוה But the LORD לעולם forever: ישׁב shall endure כונן he hath prepared למשׁפט for judgment. כסאו׃ his throne}%
\verse{והוא And he ישׁפט shall judge תבל the world בצדק in righteousness, ידין he shall minister judgment לאמים to the people במישׁרים׃ in uprightness.}%
\verse{ויהי also will be יהוה The LORD משׂגב a refuge לדך for the oppressed, משׂגב a refuge לעתות in times בצרה׃ of trouble.}%
\verse{ויבטחו will put their trust בך יודעי And they that know שׁמך thy name כי in thee: for לא hast not עזבת forsaken דרשׁיך them that seek יהוה׃ thou, LORD,}%
\verse{זמרו Sing praises ליהוה to the LORD, ישׁב which dwelleth ציון in Zion: הגידו declare בעמים among the people עלילותיו׃ his doings.}%
\verse{כי When דרשׁ he maketh inquisition דמים for blood, אותם זכר he remembereth לא not שׁכח them: he forgetteth צעקת the cry עניים׃}%
\verse{חננני Have mercy יהוה upon me, O LORD; ראה consider עניי my trouble משׂנאי of them that hate מרוממי me, thou that liftest me up משׁערי from the gates מות׃ of death:}%
\verse{למען That אספרה I may show forth כל all תהלתיך thy praise בשׁערי in the gates בת of the daughter ציון of Zion: אגילה I will rejoice בישׁועתך׃ in thy salvation.}%
\verse{טבעו are sunk down גוים The heathen בשׁחת in the pit עשׂו they made: ברשׁת in the net זו which טמנו they hid נלכדה taken. רגלם׃ is their own foot}%
\verse{נודע is known יהוה The LORD משׁפט the judgment עשׂה he executeth: בפעל in the work כפיו of his own hands. נוקשׁ רשׁע the wicked הגיון Higgaion. סלה׃ Selah.}%
\verse{ישׁובו shall be turned רשׁעים The wicked לשׁאולה into hell, כל all גוים the nations שׁכחי that forget אלהים׃ God.}%
\verse{כי For לא shall not לנצח always ישׁכח be forgotten: אביון the needy תקות the expectation ענוים תאבד shall perish לעד׃ forever.}%
\verse{קומה Arise, יהוה אל let not יעז prevail: אנושׁ ישׁפטו be judged גוים let the heathen על in פניך׃ thy sight.}%
\verse{שׁיתה Put יהוה מורה them in fear, להם themselves ידעו may know גוים the nations אנושׁ men. המה סלה׃ Selah.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 10
\verseWithHeading{A Prayer for God to Throw down the Wicked}{למה Why יהוה O LORD? תעמד standest ברחוק thou afar off, תעלים hidest לעתות thou in times בצרה׃ of trouble?}%
\verse{בגאות in pride רשׁע The wicked ידלק doth persecute עני the poor: יתפשׂו let them be taken במזמות in the devices זו that חשׁבו׃ they have imagined.}%
\verse{כי For הלל boasteth רשׁע the wicked על of תאות desire, נפשׁו his heart's ובצע the covetous, ברך and blesseth נאץ abhorreth. יהוה׃ the LORD}%
\verse{רשׁע The wicked, כגבה through the pride אפו of his countenance, בל will not ידרשׁ seek אין not אלהים God כל in all מזמותיו׃ his thoughts.}%
\verse{יחילו grievous; דרכו His ways בכל are always עת are always מרום far above משׁפטיך thy judgments מנגדו out of his sight: כל all צורריו his enemies, יפיח׃ he puffeth}%
\verse{אמר He hath said בלבו in his heart, בל I shall not אמוט be moved: לדר for never ודר אשׁרא ברע׃ in adversity.}%
\verse{אלה of cursing פיהו His mouth מלא is full ומרמות and deceit ותך and fraud: תחת under לשׁונו his tongue עמל mischief ואון׃ and vanity.}%
\verse{ישׁב He sitteth במארב in the lurking places חצרים of the villages: במסתרים in the secret places יהרג doth he murder נקי the innocent: עיניו his eyes לחלכה against the poor. יצפנו׃ are privily set}%
\verse{יארב He lieth in wait במסתר secretly כאריה as a lion בסכה in his den: יארב he lieth in wait לחטוף to catch עני the poor: יחטף he doth catch עני the poor, במשׁכו when he draweth ברשׁתו׃ him into his net.}%
\verse{ודכה He croucheth, ישׁח humbleth ונפל may fall בעצומיו by his strong ones. חלכאים׃ himself, that the poor}%
\verse{אמר He hath said בלבו in his heart, שׁכח hath forgotten: אל God הסתיר he hideth פניו his face; בל he will never ראה see לנצח׃ he will never}%
\verse{קומה Arise, יהוה O LORD; אל O God, נשׂא lift up ידך thine hand: אל not תשׁכח forget עניים׃}%
\verse{על מהאץ contemn רשׁע doth the wicked אלהים God? אמר he hath said בלבו in his heart, לא Thou wilt not תדרשׁ׃ require}%
\verse{ראתה Thou hast seen כי for אתה thou עמל mischief וכעס and spite, תביט beholdest לתת to requite בידך with thy hand: עליך himself unto יעזב committeth חלכה the poor יתום of the fatherless. אתה thee; thou היית art עוזר׃ the helper}%
\verse{שׁבר Break זרוע thou the arm רשׁע of the wicked ורע and the evil תדרושׁ seek out רשׁעו his wickedness בל none. תמצא׃ thou find}%
\verse{יהוה The LORD מלך King עולם forever ועד and ever: אבדו are perished גוים the heathen מארצו׃ out of his land.}%
\verse{תאות the desire ענוים of the humble: שׁמעת thou hast heard יהוה LORD, תכין thou wilt prepare לבם their heart, תקשׁיב to hear: אזנך׃ thou wilt cause thine ear}%
\verse{לשׁפט To judge יתום the fatherless ודך and the oppressed, בל may no יוסיף more עוד more לערץ oppress. אנושׁ מן of הארץ׃ the earth}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 11
\verseWithHeading{Confidence in Adonai’s Righteousness}{למנצח To the chief Musician, לדוד of David. ביהוה In the LORD חסיתי put I my trust: איך how תאמרו say לנפשׁי ye to my soul, נודו Flee הרכם to your mountain? צפור׃ a bird}%
\verse{כי For, הנה lo, הרשׁעים the wicked ידרכון bend קשׁת bow, כוננו they make ready חצם their arrow על upon יתר the string, לירות shoot במו that they may privily אפל that they may privily לישׁרי at the upright לב׃ in heart.}%
\verse{כי If השׁתות the foundations יהרסון be destroyed, צדיק can the righteous מה what פעל׃ do?}%
\verse{יהוה The LORD בהיכל temple, קדשׁו in his holy יהוה the LORD's בשׁמים in heaven: כסאו throne עיניו his eyes יחזו behold, עפעפיו his eyelids יבחנו try, בני the children אדם׃ of men.}%
\verse{יהוה The LORD צדיק the righteous: יבחן trieth ורשׁע but the wicked ואהב and him that loveth חמס violence שׂנאה hateth. נפשׁו׃ his soul}%
\verse{ימטר he shall rain על Upon רשׁעים the wicked פחים snares, אשׁ fire וגפרית and brimstone, ורוח tempest: זלעפות and a horrible מנת the portion כוסם׃ of their cup.}%
\verse{כי For צדיק the righteous יהוה LORD צדקות righteousness; אהב loveth ישׁר the upright. יחזו doth behold פנימו׃ his countenance}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 12
\verseWithHeading{Human Faithlessness and God’s Faithfulness}{למנצח To the chief Musician על upon השׁמינית Sheminith, מזמור A Psalm לדוד׃ of David. הושׁיעה Help, יהוה LORD; כי for גמר ceaseth; חסיד the godly man כי for פסו fail אמונים the faithful מבני from among the children אדם׃ of men.}%
\verse{שׁוא vanity ידברו They speak אישׁ every one את with רעהו his neighbor: שׂפת lips חלקות flattering בלב with a double heart ולב with a double heart ידברו׃ do they speak.}%
\verse{יכרת shall cut off יהוה The LORD כל all שׂפתי lips, חלקות flattering לשׁון the tongue מדברת that speaketh גדלות׃ proud things:}%
\verse{אשׁר Who אמרו have said, ללשׁננו With our tongue נגביר will we prevail; שׂפתינו our lips אתנו מי our own: who אדון׃ lord}%
\verse{משׁד עניים of the poor, מאנקת for the sighing אביונים of the needy, עתה now אקום will I arise, יאמר saith יהוה the LORD; אשׁית I will set בישׁע in safety יפיח׃ puffeth}%
\verse{אמרות The words יהוה of the LORD אמרות words: טהרות pure כסף silver צרוף tried בעליל in a furnace לארץ of earth, מזקק purified שׁבעתים׃ seven}%
\verse{אתה Thou יהוה them, O LORD, תשׁמרם shalt keep תצרנו thou shalt preserve מן them from הדור generation זו this לעולם׃ forever.}%
\verse{סביב on every side, רשׁעים The wicked יתהלכון walk כרם are exalted. זלות when the vilest לבני אדם׃ men}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 13
\verseWithHeading{Trust in the Salvation of Adonai}{למנצח To the chief Musician, מזמור A Psalm לדוד׃ of David. עד how long אנה how long יהוה me, O LORD? תשׁכחני wilt thou forget נצח forever? עד אנהסתיר wilt thou hide את פניך thy face ממני׃ from}%
\verse{עד how long אנה how long אשׁית shall I take עצות counsel בנפשׁי in my soul, יגון sorrow בלבבי in my heart יומם daily? עד אנהרום be exalted איבי shall mine enemy עלי׃ over}%
\verse{הביטה Consider ענני hear יהוה me, O LORD אלהי my God: האירה lighten עיני mine eyes, פן lest אישׁן I sleep המות׃ the death;}%
\verse{פן Lest יאמר say, איבי mine enemy יכלתיו I have prevailed against צרי him; those that trouble יגילו me rejoice כי when אמוט׃ I am moved.}%
\verse{ואני But I בחסדך in thy mercy; בטחתי have trusted יגל shall rejoice לבי my heart בישׁועת׃ in thy salvation.}%
\verse{אשׁירה I will sing ליהוה unto the LORD, כי because גמל he hath dealt bountifully עלי׃ with}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 14
\verseWithHeading{The Folly of the Godless and God’s Final Triumph}{למנצח To the chief Musician, לדוד of David. אמר hath said נבל The fool בלבו in his heart, אין no אלהים God. השׁחיתו They are corrupt, התעיבו they have done abominable עלילה works, אין none עשׂה that doeth טוב׃ good.}%
\verse{יהוה The LORD משׁמים from heaven השׁקיף looked down על upon בני the children אדם of men, לראות to see הישׁ if there were משׂכיל any that did understand, דרשׁ seek את אלהים׃ God.}%
\verse{הכל They are all סר gone aside, יחדו they are together נאלחו become filthy: אין none עשׂה that doeth טוב good, אין no, גם not אחד׃ one.}%
\verse{הלא no ידעו knowledge? כל Have all פעלי the workers און of iniquity אכלי who eat up עמי my people אכלו they eat לחם bread, יהוה upon the LORD. לא not קראו׃ and call}%
\verse{שׁם There פחדו were they in great fear: פחד were they in great fear: כי for אלהים God בדור in the generation צדיק׃ of the righteous.}%
\verse{עצת the counsel עני of the poor, תבישׁו Ye have shamed כי because יהוה the LORD מחסהו׃ his refuge.}%
\verse{מי יתןציון ישׁועת the salvation ישׂראל of Israel בשׁוב bringeth back יהוה when the LORD שׁבות the captivity עמו of his people, יגל shall rejoice, יעקב Jacob ישׂמח shall be glad. ישׂראל׃ Israel}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 15
\verseWithHeading{A Description of Those Who May Dwell with Adonai}{מזמור A Psalm לדוד of David. יהוה LORD, מי who יגור shall abide באהלך in thy tabernacle? מי who ישׁכן shall dwell בהר hill? קדשׁך׃ in thy holy}%
\verse{הולך He that walketh תמים uprightly, ופעל and worketh צדק righteousness, ודבר and speaketh אמת the truth בלבבו׃ in his heart.}%
\verse{לא not רגל backbiteth על with לשׁנו his tongue, לא nor עשׂה doeth לרעהו to his neighbor, רעה evil וחרפה a reproach לא nor נשׂא taketh up על against קרבו׃ his neighbor.}%
\verse{נבזה is contemned; בעיניו In whose eyes נמאס a vile person ואת יראיהוה the LORD. יכבד but he honoreth נשׁבע sweareth להרע to hurt, ולא not. ימר׃ and changeth}%
\verse{כספו his money לא nor נתן בנשׁך to usury, ושׁחד reward על against נקי the innocent. לא shall never לקח taketh עשׂה He that doeth אלה these לא ימוט be moved. לעולם׃ shall never}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 16
\verseWithHeading{Confidence in Adonai}{מכתם Michtam לדוד of David. שׁמרני Preserve אל me, O God: כי for חסיתי׃ in thee do I put my trust.}%
\verse{אמרת thou hast said ליהוה unto the LORD, אדני my Lord: אתה Thou טובתי my goodness בל not עליך׃ to}%
\verse{לקדושׁים to the saints אשׁר that בארץ in the earth, המה ואדירי and the excellent, כל in whom all חפצי׃ my delight.}%
\verse{ירבו shall be multiplied עצבותם Their sorrows אחר another מהרו hasten בל will I not אסיך offer, נסכיהם their drink offerings מדם of blood ובל nor אשׂא take up את שׁמותם their names על into שׂפתי׃ my lips.}%
\verse{יהוה The LORD מנת חלקי of mine inheritance וכוסי and of my cup: אתה thou תומיך maintainest גורלי׃ my lot.}%
\verse{חבלים The lines נפלו are fallen לי בנעמים unto me in pleasant אף yea, נחלת heritage. שׁפרה a goodly עלי׃ I have}%
\verse{אברך I will bless את יהוה the LORD, אשׁר who יעצני hath given me counsel: אף also לילות me in the night seasons. יסרוני instruct כליותי׃ my reins}%
\verse{שׁויתי I have set יהוה the LORD לנגדי before תמיד always כי me: because מימיני at my right hand, בל I shall not אמוט׃ be moved.}%
\verse{לכן Therefore שׂמח is glad, לבי my heart ויגל rejoiceth: כבודי and my glory אף also בשׂרי my flesh ישׁכן shall rest לבטח׃ in hope.}%
\verse{כי For לא thou wilt not תעזב leave נפשׁי my soul לשׁאול in hell; לא neither תתן wilt thou suffer חסידך thine Holy One לראות to see שׁחת׃ corruption.}%
\verse{תודיעני Thou wilt show ארח me the path חיים of life: שׂבע fullness שׂמחות of joy; את פניך thy presence נעמות pleasures בימינך at thy right hand נצח׃ forevermore.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 17
\verseWithHeading{Prayer for Vindication and Protection}{תפלה A Prayer לדוד of David. שׁמעה Hear יהוה O LORD צדק the right, הקשׁיבה attend רנתי unto my cry, האזינה give ear תפלתי unto my prayer, בלא not שׂפתי lips. מרמה׃ out of feigned}%
\verse{מלפניך from thy presence; משׁפטי Let my sentence יצא come forth עיניך let thine eyes תחזינה behold מישׁרים׃ the things that are equal.}%
\verse{בחנת Thou hast proved לבי mine heart; פקדת thou hast visited לילה in the night; צרפתני thou hast tried בל nothing; תמצא me, shalt find זמתי I am purposed בל shall not יעבר transgress. פי׃ my mouth}%
\verse{לפעלות Concerning the works אדם of men, בדבר by the word שׂפתיך of thy lips אני I שׁמרתי have kept ארחות the paths פריץ׃ of the destroyer.}%
\verse{תמך Hold up אשׁרי my goings במעגלותיך in thy paths, בל not. נמוטו slip פעמי׃ my footsteps}%
\verse{אני I קראתיך have called upon כי thee, for תענני thou wilt hear אל me, O God: הט incline אזנך thine ear לי שׁמע unto me, and hear אמרתי׃ my speech.}%
\verse{הפלה חסדיךושׁיע O thou that savest חוסים them which put their trust ממתקוממים from those that rise up בימינך׃ by thy right hand}%
\verse{שׁמרני Keep כאישׁון me as the apple בת me as the apple עין of the eye, בצל me under the shadow כנפיך of thy wings, תסתירני׃ hide}%
\verse{מפני רשׁעים the wicked זו that שׁדוני oppress איבי enemies, בנפשׁ me, my deadly יקיפו compass me about. עלי׃ compass me about.}%
\verse{חלבמו their own fat: סגרו They are enclosed in פימו with their mouth דברו they speak בגאות׃ proudly.}%
\verse{אשׁרינו us in our steps: עתה They have now סבבוני compassed עיניהם their eyes ישׁיתו they have set לנטות bowing down בארץ׃ to the earth;}%
\verse{דמינו Like כאריה as a lion יכסוף is greedy לטרוף of his prey, וככפיר and as it were a young lion ישׁב lurking במסתרים׃ in secret places.}%
\verse{קומה Arise, יהוה O LORD, קדמה disappoint פניו disappoint הכריעהו him, cast him down: פלטה deliver נפשׁי my soul מרשׁע from the wicked, חרבך׃ thy sword:}%
\verse{ממתים from men ידך thy hand, יהוה O LORD, ממתים מחלד of the world, חלקם their portion בחיים in life, וצפינך תמלא thou fillest בטנם and whose belly ישׂבעו they are full בנים of children, והניחו יתרם the rest לעולליהם׃ of their to their babes.}%
\verse{אני As for me, בצדק in righteousness: אחזה I will behold פניך thy face אשׂבעה I shall be satisfied, בהקיץ when I awake, תמונתך׃ with thy likeness.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 18
\verseWithHeading{Praise to God for His Deliverance}{למנצח To the chief Musician, לעבד the servant יהוה of the LORD, לדוד of David, אשׁר who דבר spoke ליהוה unto the LORD את דברי the words השׁירה song הזאת of this ביום in the day הציל delivered יהוה the LORD אותו מכף him from the hand כל of all איביו his enemies, ומיד and from the hand שׁאול׃ of Saul: ויאמר And he said, ארחמך I will love יהוה thee, O LORD, חזקי׃ my strength.}%
\verse{יהוה The LORD סלעי my rock, ומצודתי and my fortress, ומפלטי and my deliverer; אלי my God, צורי my strength, אחסה in whom I will trust; בו מגני my buckler, וקרן and the horn ישׁעי of my salvation, משׂגבי׃ my high tower.}%
\verse{מהלל to be praised: אקרא I will call יהוה upon the LORD, ומן from איבי mine enemies. אושׁע׃ so shall I be saved}%
\verse{אפפוני compassed חבלי The sorrows מות of death ונחלי me, and the floods בליעל of ungodly men יבעתוני׃ made me afraid.}%
\verse{חבלי The sorrows שׁאול of hell סבבוני compassed me about: קדמוני prevented מוקשׁי the snares מות׃ of death}%
\verse{בצר In my distress לי אקרא I called upon יהוה the LORD, ואל unto אלהי my God: אשׁוע and cried ישׁמע he heard מהיכלו out of his temple, קולי my voice ושׁועתי and my cry לפניו before תבוא came באזניו׃ him, into his ears.}%
\verse{ותגעשׁ shook ותרעשׁ and trembled; הארץ Then the earth ומוסדי the foundations הרים also of the hills ירגזו moved ויתגעשׁו and were shaken, כי because חרה׃ he was wroth.}%
\verse{עלה There went up עשׁן a smoke באפו out of his nostrils, ואשׁ and fire מפיו out of his mouth תאכל devoured: גחלים coals בערו were kindled ממנו׃ out of his mouth}%
\verse{ויט He bowed שׁמים the heavens וירד also, and came down: וערפל and darkness תחת under רגליו׃ his feet.}%
\verse{וירכב And he rode על upon כרוב a cherub, ויעף and did fly: וידא yea, he did fly על upon כנפי the wings רוח׃ of the wind.}%
\verse{ישׁת He made חשׁך סתרו his secret place; סביבותיו round about סכתו his pavilion חשׁכת darkness מים waters עבי thick clouds שׁחקים׃ of the skies.}%
\verse{מנגה נגדו before עביו him his thick clouds עברו passed, ברד hail וגחלי and coals אשׁ׃ of fire.}%
\verse{וירעם also thundered בשׁמים in the heavens, יהוה The LORD ועליון and the Highest יתן gave קלו his voice; ברד hail וגחלי and coals אשׁ׃ of fire.}%
\verse{וישׁלח Yea, he sent out חציו his arrows, ויפיצם and scattered וברקים lightnings, רב them; and he shot out ויהמם׃ and discomfited}%
\verse{ויראו were seen, אפיקי Then the channels מים of waters ויגלו were discovered מוסדות and the foundations תבל of the world מגערתך at thy rebuke, יהוה O LORD, מנשׁמת at the blast רוח of the breath אפך׃ of thy nostrils.}%
\verse{ישׁלח He sent ממרום from above, יקחני he took ימשׁני me, he drew ממים רבים׃}%
\verse{יצילני He delivered מאיבי enemy, עז me from my strong ומשׂנאי and from them which hated כי me: for אמצו they were too strong ממני׃ enemy,}%
\verse{יקדמוני They prevented ביום me in the day אידי of my calamity: ויהי was יהוה but the LORD למשׁען׃ my stay.}%
\verse{ויוציאני He brought me forth למרחב also into a large place; יחלצני he delivered כי me, because חפץ׃ he delighted}%
\verse{יגמלני rewarded יהוה The LORD כצדקי me according to my righteousness; כבר according to the cleanness ידי of my hands ישׁיב׃ hath he recompensed}%
\verse{כי For שׁמרתי I have kept דרכי the ways יהוה of the LORD, ולא and have not רשׁעתי wickedly departed מאלהי׃}%
\verse{כי For כל all משׁפטיו his judgments לנגדי before וחקתיו his statutes לא me, and I did not אסיר put away מני׃ from}%
\verse{ואהי I was תמים also upright עמו before ואשׁתמר him, and I kept myself מעוני׃ from mine iniquity.}%
\verse{וישׁב recompensed יהוה Therefore hath the LORD לי כצדקי me according to my righteousness, כבר according to the cleanness ידי of my hands לנגד in עיניו׃ his eyesight.}%
\verse{עם With חסיד the merciful תתחסד thou wilt show thyself merciful; עם with גבר man תמים an upright תתמם׃ thou wilt show thyself upright;}%
\verse{עם With נבר the pure תתברר thou wilt show thyself pure; ועם and with עקשׁ the froward תתפתל׃ thou wilt show thyself froward.}%
\verse{כי For אתה thou wilt עם people; עני the afflicted תושׁיע save ועינים looks. רמות high תשׁפיל׃ but wilt bring down}%
\verse{כי For אתה thou תאיר wilt light נרי my candle: יהוה the LORD אלהי my God יגיה will enlighten חשׁכי׃ my darkness.}%
\verse{כי For בך ארץ by thee I have run through גדוד a troop; ובאלהי and by my God אדלג have I leaped over שׁור׃ a wall.}%
\verse{האל God, תמים perfect: דרכו his way אמרת the word יהוה of the LORD צרופה is tried: מגן a buckler הוא he לכל to all החסים׃ those that trust}%
\verse{כי For מי who אלוה God מבלעדי save יהוה the LORD? ומי or who צור a rock זולתי save אלהינו׃ our God?}%
\verse{האל God המאזרני that girdeth חיל me with strength, ויתן and maketh תמים perfect. דרכי׃ my way}%
\verse{משׁוה He maketh רגלי my feet כאילות like hinds' ועל me upon במתי my high places. יעמידני׃ and setteth}%
\verse{מלמד He teacheth ידי my hands למלחמה to war, ונחתה is broken קשׁת so that a bow נחושׁה of steel זרועתי׃ by mine arms.}%
\verse{ותתן Thou hast also given לי מגן me the shield ישׁעך of thy salvation: וימינך and thy right hand תסעדני hath holden me up, וענותך תרבני׃ hath made me great.}%
\verse{תרחיב Thou hast enlarged צעדי my steps תחתי under ולא did not מעדו slip. קרסלי׃ me, that my feet}%
\verse{ארדוף I have pursued אויבי mine enemies, ואשׂיגם and overtaken ולא them: neither אשׁוב did I turn again עד till כלותם׃ they were consumed.}%
\verse{אמחצם I have wounded ולא them that they were not יכלו able קום to rise: יפלו they are fallen תחת under רגלי׃ my feet.}%
\verse{ותאזרני For thou hast girded חיל me with strength למלחמה unto the battle: תכריע thou hast subdued קמי me those that rose up תחתי׃ under}%
\verse{ואיבי of mine enemies; נתתה Thou hast also given לי ערף me the necks ומשׂנאי them that hate אצמיתם׃ that I might destroy}%
\verse{ישׁועו They cried, ואין but none מושׁיע to save על unto יהוה the LORD, ולא them not. ענם׃ but he answered}%
\verse{ואשׁחקם Then did I beat them small כעפר as the dust על before פני before רוח the wind: כטיט as the dirt חוצות in the streets. אריקם׃ I did cast them out}%
\verse{תפלטני Thou hast delivered מריבי me from the strivings עם of the people; תשׂימני thou hast made לראשׁ me the head גוים of the heathen: עם a people לא I have not ידעתי known יעבדוני׃ shall serve}%
\verse{לשׁמע אזןשׁמעו of me, they shall obey לי בני me: the strangers נכר me: the strangers יכחשׁו׃ shall submit}%
\verse{בני נכרבלו shall fade away, ויחרגו and be afraid ממסגרותיהם׃ their close places.}%
\verse{חי liveth; יהוה The LORD וברוך and blessed צורי my rock; וירום be exalted. אלוהי ישׁעי׃ of my salvation}%
\verse{האל God הנותן that avengeth נקמות that avengeth לי וידבר me, and subdueth עמים the people תחתי׃ under}%
\verse{מפלטי He delivereth מאיבי me from mine enemies: אף yea, מן me from mine enemies: קמי those that rise up תרוממני thou liftest me up מאישׁ man. חמס me from the violent תצילני׃ against me: thou hast delivered}%
\verse{על כןודך will I give thanks בגוים among the heathen, יהוה unto thee, O LORD, ולשׁמך unto thy name. אזמרה׃ and sing praises}%
\verse{מגדל ישׁועות deliverance מלכו giveth he to his king; ועשׂה and showeth חסד mercy למשׁיחו to his anointed, לדוד to David, ולזרעו and to his seed עד forevermore. עולם׃ forevermore.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 19
\verseWithHeading{Adonai’s Creation and Law}{למנצח To the chief Musician, מזמור A Psalm לדוד׃ of David. השׁמים The heavens מספרים declare כבוד the glory אל of God; ומעשׂה his handiwork. ידיו his handiwork. מגיד showeth הרקיע׃ and the firmament}%
\verse{יום Day ליום unto day יביע uttereth אמר speech, ולילה and night ללילה unto night יחוה showeth דעת׃ knowledge.}%
\verse{אין no אמר speech ואין nor דברים language, בלי is not נשׁמע heard. קולם׃ their voice}%
\verse{בכל through all הארץ the earth, יצא is gone out קום Their line ובקצה to the end תבל of the world. מליהם and their words לשׁמשׁ for the sun, שׂם In them hath he set אהל׃ a tabernacle}%
\verse{והוא Which כחתן as a bridegroom יצא coming out מחפתו of his chamber, ישׂישׂ rejoiceth כגבור as a strong man לרוץ to run ארח׃ a race.}%
\verse{מקצה from the end השׁמים of the heaven, מוצאו His going forth ותקופתו and his circuit על unto קצותם the ends ואין of it: and there is nothing נסתר hid מחמתו׃ from the heat}%
\verse{תורת The law יהוה of the LORD תמימה perfect, משׁיבת converting נפשׁ the soul: עדות the testimony יהוה of the LORD נאמנה sure, מחכימת making wise פתי׃ the simple.}%
\verse{פקודי The statutes יהוה of the LORD ישׁרים right, משׂמחי rejoicing לב the heart: מצות the commandment יהוה of the LORD ברה pure, מאירת enlightening עינים׃ the eyes.}%
\verse{יראת The fear יהוה of the LORD טהורה clean, עומדת enduring לעד forever: משׁפטי the judgments יהוה of the LORD אמת true צדקו righteous יחדו׃ altogether.}%
\verse{הנחמדים More to be desired מזהב than gold, ומפז fine gold: רב yea, than much ומתוקים sweeter מדבשׁ also than honey ונפת and the honeycomb. צופים׃ and the honeycomb.}%
\verse{גם Moreover עבדך by them is thy servant נזהר warned: בהם בשׁמרם in keeping עקב reward. רב׃ of them great}%
\verse{שׁגיאות errors? מי Who יבין can understand מנסתרות thou me from secret נקני׃ cleanse}%
\verse{גם also מזדים from presumptuous חשׂך Keep back עבדך thy servant אל let them not ימשׁלו have dominion בי אז over me: then איתם shall I be upright, ונקיתי and I shall be innocent מפשׁע transgression. רב׃ from the great}%
\verse{יהיו be לרצון acceptable אמרי Let the words פי of my mouth, והגיון and the meditation לבי of my heart, לפניך in thy sight, יהוה O LORD, צורי my strength, וגאלי׃ and my redeemer.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 20
\verseWithHeading{God’s Blessing on the King}{למנצח To the chief Musician, מזמור A Psalm לדוד׃ of David. יענך hear יהוה The LORD ביום thee in the day צרה of trouble; ישׂגבך defend שׁם the name אלהי of the God יעקב׃ of Jacob}%
\verse{ישׁלח Send עזרך thee help מקדשׁ from the sanctuary, ומציון יסעדך׃ and strengthen}%
\verse{יזכר Remember כל all מנחתך thy offerings, ועולתך thy burnt sacrifice; ידשׁנה and accept סלה׃ Selah.}%
\verse{יתן Grant לך כלבבך thee according to thine own heart, וכל all עצתך thy counsel. ימלא׃ and fulfill}%
\verse{נרננה We will rejoice בישׁועתך in thy salvation, ובשׁם and in the name אלהינו of our God נדגל we will set up banners: ימלא fulfill יהוה the LORD כל all משׁאלותיך׃ thy petitions.}%
\verse{עתה Now ידעתי know כי I that הושׁיע saveth יהוה the LORD משׁיחו his anointed; יענהו he will hear משׁמי heaven קדשׁו him from his holy בגברות strength ישׁע with the saving ימינו׃ of his right hand.}%
\verse{אלה Some ברכב in chariots, ואלה and some בסוסים in horses: ואנחנו but we בשׁם the name יהוה of the LORD אלהינו our God. נזכיר׃ will remember}%
\verse{המה They כרעו are brought down ונפלו and fallen: ואנחנו but we קמנו are risen, ונתעודד׃ and stand upright.}%
\verse{יהוה LORD: הושׁיעה Save, המלך let the king יעננו hear ביום us when קראנו׃ we call.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 21
\verseWithHeading{Joy in the Salvation of Adonai}{למנצח To the chief Musician, מזמור A Psalm לדוד׃ of David. יהוה O LORD; בעזך in thy strength, ישׂמח shall joy מלך The king ובישׁועתך and in thy salvation מה how יגיל shall he rejoice! מאד׃ greatly}%
\verse{תאות desire, לבו him his heart's נתתה Thou hast given לו וארשׁת the request שׂפתיו of his lips. בל and hast not מנעת withheld סלה׃ Selah.}%
\verse{כי For תקדמנו thou preventest ברכות him with the blessings טוב of goodness: תשׁית thou settest לראשׁו on his head. עטרת a crown פז׃ of pure gold}%
\verse{חיים life שׁאל He asked ממך of נתתה thee, thou gavest לו ארך him, length ימים of days עולם forever ועד׃ and ever.}%
\verse{גדול great כבודו His glory בישׁועתך in thy salvation: הוד honor והדר and majesty תשׁוה hast thou laid עליו׃ upon}%
\verse{כי For תשׁיתהו thou hast made ברכות him most blessed לעד forever: תחדהו thou hast made him exceeding glad בשׂמחה thou hast made him exceeding glad את with פניך׃ thy countenance.}%
\verse{כי For המלך the king בטח trusteth ביהוה in the LORD, ובחסד and through the mercy עליון of the most High בל he shall not ימוט׃ be moved.}%
\verse{תמצא shall find out ידך Thine hand לכל all איביך thine enemies: ימינך thy right hand תמצא shall find out שׂנאיך׃ those that hate}%
\verse{תשׁיתמו Thou shalt make כתנור oven אשׁ them as a fiery לעת in the time פניך of thine anger: יהוה the LORD באפו in his wrath, יבלעם shall swallow them up ותאכלם shall devour אשׁ׃ and the fire}%
\verse{פרימו Their fruit מארץ from the earth, תאבד shalt thou destroy וזרעם and their seed מבני from among the children אדם׃ of men.}%
\verse{כי For נטו they intended עליך against רעה evil חשׁבו thee: they imagined מזמה a mischievous device, בל they are not יוכלו׃ able}%
\verse{כי Therefore תשׁיתמו shalt thou make שׁכם them turn their back, במיתריך upon thy strings תכונן thou shalt make ready על against פניהם׃ the face}%
\verse{רומה Be thou exalted, יהוה LORD, בעזך in thine own strength: נשׁירה will we sing ונזמרה and praise גבורתך׃ thy power.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 22
\verseWithHeading{Suffering and Waiting for Deliverance}{למנצח To the chief Musician על upon אילת Aijeleth השׁחר Shahar, מזמור A Psalm לדוד׃ of David. אלי My God, אלי my God, למה why עזבתני hast thou forsaken רחוק me? far מישׁועתי from helping דברי me, the words שׁאגתי׃ of my roaring?}%
\verse{אלהי O my God, אקרא I cry יומם in the daytime, ולא not; תענה but thou hearest ולילה and in the night season, ולא and am not דומיה׃ silent.}%
\verse{ואתה But thou קדושׁ holy, יושׁב that inhabitest תהלות the praises ישׂראל׃ of Israel.}%
\verse{בך בטחו trusted אבתינו Our fathers בטחו in thee: they trusted, ותפלטמו׃ and thou didst deliver}%
\verse{אליך unto זעקו They cried ונמלטו thee, and were delivered: בך בטחו they trusted ולא in thee, and were not בושׁו׃ confounded.}%
\verse{ואנכי But I תולעת a worm, ולא and no אישׁ man; חרפת a reproach אדם of men, ובזוי and despised עם׃ of the people.}%
\verse{כל All ראי they that see ילעגו me laugh לי יפטירו me to scorn: they shoot out בשׂפה the lip, יניעו they shake ראשׁ׃ the head,}%
\verse{גל He trusted אל on יהוה the LORD יפלטהו he would deliver יצילהו him: let him deliver כי him, seeing חפץ׃ he delighted}%
\verse{כי But אתה thou גחי he that took מבטן me out of the womb: מבטיחי thou didst make me hope על upon שׁדי breasts. אמי׃ my mother's}%
\verse{עליך upon השׁלכתי I was cast מרחם thee from the womb: מבטן belly. אמי from my mother's אלי my God אתה׃ thou}%
\verse{אל Be not תרחק far ממני from כי me; for צרה trouble קרובה near; כי for אין none עוזר׃ to help.}%
\verse{סבבוני have compassed פרים bulls רבים Many אבירי me: strong בשׁן of Bashan כתרוני׃ have beset me round.}%
\verse{פצו They gaped עלי upon פיהם me their mouths, אריה lion. טרף a ravening ושׁאג׃ and a roaring}%
\verse{כמים like water, נשׁפכתי I am poured out והתפרדו are out of joint: כל and all עצמותי my bones היה is לבי my heart כדונג like wax; נמס it is melted בתוך in the midst מעי׃ of my bowels.}%
\verse{יבשׁ is dried up כחרשׂ like a potsherd; כחי My strength ולשׁוני and my tongue מדבק cleaveth מלקוחי to my jaws; ולעפר me into the dust מות of death. תשׁפתני׃ and thou hast brought}%
\verse{כי For סבבוני have compassed כלבים dogs עדת me: the assembly מרעים of the wicked הקיפוני have enclosed כארי ידי my hands ורגלי׃ and my feet.}%
\verse{אספר I may tell כל all עצמותי my bones: המה they יביטו look יראו׃ stare}%
\verse{יחלקו They part בגדי my garments להם ועל upon לבושׁי my vesture. יפילו among them, and cast גורל׃ lots}%
\verse{ואתה thou יהוה from me, O LORD: אל But be not תרחק far אילותי O my strength, לעזרתי thee to help חושׁה׃ haste}%
\verse{הצילה Deliver מחרב from the sword; נפשׁי my soul מיד from the power כלב of the dog. יחידתי׃ my darling}%
\verse{הושׁיעני Save מפי mouth: אריה me from the lion's ומקרני me from the horns רמים of the unicorns. עניתני׃ for thou hast heard}%
\verse{אספרה I will declare שׁמך thy name לאחי unto my brethren: בתוך in the midst קהל of the congregation אהללך׃ will I praise}%
\verse{יראי Ye that fear יהוה the LORD, הללוהו praise כל him; all זרע ye the seed יעקב of Jacob, כבדוהו glorify וגורו him; and fear ממנו him; and fear כל him, all זרע ye the seed ישׂראל׃ of Israel.}%
\verse{כי For לא he hath not בזה despised ולא nor שׁקץ abhorred ענות the affliction עני of the afflicted; ולא neither הסתיר hath he hid פניו his face ממנו from ובשׁועו him; but when he cried אליו unto שׁמע׃ him, he heard.}%
\verse{מאתך תהלתי My praise בקהל congregation: רב thee in the great נדרי my vows אשׁלם I will pay נגד before יראיו׃ them that fear}%
\verse{יאכלו shall eat ענוים The meek וישׂבעו and be satisfied: יהללו they shall praise יהוה the LORD דרשׁיו that seek יחי shall live לבבכם him: your heart לעד׃ forever.}%
\verse{יזכרו shall remember וישׁבו and turn אל unto יהוה the LORD: כל All אפסי the ends ארץ of the world וישׁתחוו shall worship לפניך before כל and all משׁפחות the kindreds גוים׃ of the nations}%
\verse{כי For ליהוה the LORD's: המלוכה the kingdom ומשׁל and he the governor בגוים׃ among the nations.}%
\verse{אכלו shall eat וישׁתחוו and worship: כל All דשׁני fat ארץ upon earth לפניו before יכרעו shall bow כל all יורדי they that go down עפר to the dust ונפשׁו his own soul. לא him: and none חיה׃ can keep alive}%
\verse{זרע A seed יעבדנו shall serve יספר him; it shall be accounted לאדני to the Lord לדור׃ for a generation.}%
\verse{יבאו They shall come, ויגידו and shall declare צדקתו his righteousness לעם unto a people נולד that shall be born, כי that עשׂה׃ he hath done}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 23
\verseWithHeading{Adonai the Shepherd}{מזמור A Psalm לדוד of David. יהוה The LORD רעי my shepherd; לא I shall not אחסר׃ want.}%
\verse{בנאות pastures: דשׁא in green ירביצני He maketh me to lie down על me beside מי waters. מנחות the still ינהלני׃ he leadeth}%
\verse{נפשׁי my soul: ישׁובב He restoreth ינחני he leadeth במעגלי me in the paths צדק of righteousness למען שׁמו׃}%
\verse{גם Yea, כי though אלך I walk בגיא through the valley צלמות of the shadow of death, לא no אירא I will fear רע evil: כי for אתה thou עמדי with שׁבטך me; thy rod ומשׁענתך and thy staff המה they ינחמני׃ comfort}%
\verse{תערך Thou preparest לפני before שׁלחן a table נגד me in the presence צררי of mine enemies: דשׁנת thou anointest בשׁמן with oil; ראשׁי my head כוסי my cup רויה׃ runneth over.}%
\verse{אך Surely טוב goodness וחסד and mercy ירדפוני shall follow כל me all ימי the days חיי of my life: ושׁבתי and I will dwell בבית in the house יהוה of the LORD לארך forever. ימים׃ forever.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 24
\verseWithHeading{The King of Glory}{לדוד of David. מזמור A Psalm ליהוה the LORD's, הארץ The earth ומלואה and the fullness תבל thereof; the world, וישׁבי׃ and they that dwell}%
\verse{כי For הוא he על it upon ימים the seas, יסדה hath founded ועל it upon נהרות the floods. יכוננה׃ and established}%
\verse{מי Who יעלה shall ascend בהר into the hill יהוה of the LORD? ומי or who יקום shall stand במקום place? קדשׁו׃ in his holy}%
\verse{נקי He that hath clean כפים hands, ובר and a pure לבב heart; אשׁר who לא hath not נשׂא lifted up לשׁוא unto vanity, נפשׁי his soul ולא nor נשׁבע sworn למרמה׃ deceitfully.}%
\verse{ישׂא He shall receive ברכה the blessing מאת יהוה the LORD, וצדקה and righteousness מאלהי ישׁעו׃ of his salvation.}%
\verse{זה This דור the generation דרשׁו of them that seek מבקשׁי him, that seek פניך thy face, יעקב O Jacob. סלה׃ Selah.}%
\verse{שׂאו Lift up שׁערים O ye gates; ראשׁיכם your heads, והנשׂאו and be ye lift up, פתחי doors; עולם ye everlasting ויבוא shall come in. מלך and the King הכבוד׃ of glory}%
\verse{מי Who זה this מלך King הכבוד of glory? יהוה The LORD עזוז strong וגבור and mighty, יהוה the LORD גבור mighty מלחמה׃ in battle.}%
\verse{שׂאו Lift up שׁערים O ye gates; ראשׁיכם your heads, ושׂאו even lift up, פתחי doors; עולם ye everlasting ויבא shall come in. מלך and the King הכבוד׃ of glory}%
\verse{מי Who הוא is זה this מלך King הכבוד of glory? יהוה The LORD צבאות of hosts, הוא he מלך the King הכבוד of glory. סלה׃ Selah.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 25
\verseWithHeading{A Prayer for Guidance, Deliverance, and Forgiveness}{לדוד of David. אליך Unto יהוה thee, O LORD, נפשׁי my soul. אשׂא׃ do I lift up}%
\verse{אלהי O my God, בך בטחתי I trust אל in thee: let me not אבושׁה be ashamed, אל let not יעלצו triumph איבי׃ mine enemies}%
\verse{גם Yea, כל let none קויך that wait לא let none יבשׁו on thee be ashamed: יבשׁו let them be ashamed הבוגדים which transgress ריקם׃ without cause.}%
\verse{דרכיך me thy ways, יהוה O LORD; הודיעני Show ארחותיך me thy paths. למדני׃ teach}%
\verse{הדריכני Lead באמתך me in thy truth, ולמדני and teach כי me: for אתה thou אלהי the God ישׁעי of my salvation; אותך קויתי on thee do I wait כל all היום׃ the day.}%
\verse{זכר Remember, רחמיך thy tender mercies יהוה O LORD, וחסדיך and thy lovingkindnesses; כי for מעולם ever of old. המה׃ they}%
\verse{חטאות the sins נעורי of my youth, ופשׁעי nor my transgressions: אל not תזכר Remember כחסדך according to thy mercy זכר remember לי אתה thou למען טובךהוה׃ O LORD.}%
\verse{טוב Good וישׁר and upright יהוה the LORD: על therefore כן therefore יורה will he teach חטאים sinners בדרך׃ in the way.}%
\verse{ידרך will he guide ענוים The meek במשׁפט in judgment: וילמד will he teach ענוים and the meek דרכו׃ his way.}%
\verse{כל All ארחות the paths יהוה of the LORD חסד mercy ואמת and truth לנצרי unto such as keep בריתו his covenant ועדתיו׃ and his testimonies.}%
\verse{למען שׁמךהוה O LORD, וסלחת pardon לעוני mine iniquity; כי for רב great. הוא׃ it}%
\verse{מי What זה he that האישׁ man ירא feareth יהוה the LORD? יורנו him shall he teach בדרך in the way יבחר׃ he shall choose.}%
\verse{נפשׁו His soul בטוב at ease; תלין shall dwell וזרעו and his seed יירשׁ shall inherit ארץ׃ the earth.}%
\verse{סוד The secret יהוה of the LORD ליראיו with them that fear ובריתו them his covenant. להודיעם׃ him; and he will show}%
\verse{עיני Mine eyes תמיד ever אל toward יהוה the LORD; כי for הוא he יוציא shall pluck my feet out מרשׁת of the net. רגלי׃ shall pluck my feet out}%
\verse{פנה Turn אלי thee unto וחנני me, and have mercy כי upon me; for יחיד desolate ועני and afflicted. אני׃ I}%
\verse{צרות The troubles לבבי of my heart הרחיבו are enlarged: ממצוקותי of my distresses. הוציאני׃ bring thou me out}%
\verse{ראה Look upon עניי mine affliction ועמלי and my pain; ושׂא and forgive לכל all חטאותי׃ my sins.}%
\verse{ראה Consider אויבי mine enemies; כי for רבו they are many; ושׂנאת and they hate חמס me with cruel שׂנאוני׃ hatred.}%
\verse{שׁמרה O keep נפשׁי my soul, והצילני and deliver אל me: let me not אבושׁ be ashamed; כי for חסיתי׃ I put my trust}%
\verse{תם Let integrity וישׁר and uprightness יצרוני preserve כי me; for קויתיך׃ I wait on}%
\verse{פדה Redeem אלהים O God, את ישׂראל Israel, מכל out of all צרותיו׃ his troubles.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 26
\verseWithHeading{A Prayer for Vindication}{לדוד of David. שׁפטני Judge יהוה me, O LORD; כי for אני I בתמי in mine integrity: הלכתי have walked וביהוה also in the LORD; בטחתי I have trusted לא I shall not אמעד׃ slide.}%
\verse{בחנני Examine יהוה me, O LORD, ונסני and prove צרופה me; try כליותי my reins ולבי׃ and my heart.}%
\verse{כי For חסדך thy lovingkindness לנגד before עיני mine eyes: והתהלכתי and I have walked באמתך׃ in thy truth.}%
\verse{לא I have not ישׁבתי sat עם with מתי persons, שׁוא vain ועם with נעלמים dissemblers. לא neither אבוא׃ will I go in}%
\verse{שׂנאתי I have hated קהל the congregation מרעים of evildoers; ועם with רשׁעים the wicked. לא and will not אשׁב׃ sit}%
\verse{ארחץ I will wash בנקיון in innocency: כפי mine hands ואסבבה so will I compass את מזבחך thine altar, יהוה׃ O LORD:}%
\verse{לשׁמע That I may publish בקול with the voice תודה of thanksgiving, ולספר and tell כל of all נפלאותיך׃ thy wondrous works.}%
\verse{יהוה LORD, אהבתי I have loved מעון the habitation ביתך of thy house, ומקום and the place משׁכן dwelleth. כבודך׃ where thine honor}%
\verse{אל not תאסף Gather עם with חטאים sinners, נפשׁי my soul ועם with אנשׁי men: דמים bloody חיי׃ nor my life}%
\verse{אשׁר In whose בידיהם hands זמה mischief, וימינם and their right hand מלאה is full שׁחד׃ of bribes.}%
\verse{ואני But as for me, בתמי in mine integrity: אלך I will walk פדני redeem וחנני׃ me, and be merciful}%
\verse{רגלי My foot עמדה standeth במישׁור in an even place: במקהלים in the congregations אברך will I bless יהוה׃ the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 27
\verseWithHeading{A Declaration of Trust}{לדוד of David. יהוה The LORD אורי my light וישׁעי and my salvation; ממי whom אירא shall I fear? יהוה the LORD מעוז the strength חיי of my life; ממי of whom אפחד׃ shall I be afraid?}%
\verse{בקרב came עלי upon מרעים When the wicked, לאכל me to eat up את בשׂרי my flesh, צרי mine enemies ואיבי and my foes, לי המה they כשׁלו stumbled ונפלו׃ and fell.}%
\verse{אם Though תחנה should encamp עלי against מחנה a host לא shall not יירא fear: לבי me, my heart אם though תקום should rise עלי against מלחמה war בזאת me, in this אני I בוטח׃ confident.}%
\verse{אחת One שׁאלתי have I desired מאת יהוה the LORD, אותה אבקשׁ that will I seek after; שׁבתי that I may dwell בבית in the house יהוה of the LORD כל all ימי the days חיי of my life, לחזות to behold בנעם the beauty יהוה of the LORD, ולבקר and to inquire בהיכלו׃ in his temple.}%
\verse{כי For יצפנני he shall hide בסכה me in his pavilion: ביום in the time רעה of trouble יסתרני shall he hide בסתר in the secret אהלו of his tabernacle בצור upon a rock. ירוממני׃ me; he shall set me up}%
\verse{ועתה And now ירום be lifted up ראשׁי shall mine head על above איבי mine enemies סביבותי round about ואזבחה me: therefore will I offer באהלו in his tabernacle זבחי sacrifices תרועה of joy; אשׁירה I will sing, ואזמרה yea, I will sing praises ליהוה׃ unto the LORD.}%
\verse{שׁמע Hear, יהוה O LORD, קולי with my voice: אקרא I cry וחנני have mercy וענני׃ also upon me, and answer}%
\verse{לך אמר said לבי my heart בקשׁו Seek פני ye my face; את פניך unto thee, Thy face, יהוה LORD, אבקשׁ׃ will I seek.}%
\verse{אל not תסתר Hide פניך thy face ממני from אל me not, תט me; put not thy servant away באף in anger: עבדך עזרתי my help; היית thou hast been אל neither תטשׁני leave ואל תעזבני forsake אלהי me, O God ישׁעי׃ of my salvation.}%
\verse{כי When אבי my father ואמי and my mother עזבוני forsake ויהוה me, then the LORD יאספני׃ will take me up.}%
\verse{הורני Teach יהוה O LORD, דרכך me thy way, ונחני and lead בארח path, מישׁור me in a plain למען because of שׁוררי׃ mine enemies.}%
\verse{אל me not תתנני Deliver בנפשׁ over unto the will צרי of mine enemies: כי for קמו are risen up בי עדי witnesses שׁקר false ויפח against me, and such as breathe out חמס׃ cruelty.}%
\verse{לולא unless האמנתי I had believed לראות to see בטוב the goodness יהוה of the LORD בארץ in the land חיים׃ of the living.}%
\verse{קוה Wait אל on יהוה the LORD: חזק be of good courage, ויאמץ and he shall strengthen לבך thine heart: וקוה wait, אל I say, on יהוה׃ the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 28
\verseWithHeading{A Prayer for Help, and Joy in Its Answer}{לדוד of David. אליך Unto יהוה O LORD אקרא thee will I cry, צורי my rock; אל be not תחרשׁ silent ממני to פן me: lest, תחשׁה thou be silent ממני to ונמשׁלתי עםורדי them that go down בור׃ into the pit.}%
\verse{שׁמע Hear קול the voice תחנוני of my supplications, בשׁועי when I cry אליך unto בנשׂאי thee, when I lift up ידי my hands אל toward דביר oracle. קדשׁך׃ thy holy}%
\verse{אל me not תמשׁכני Draw עם away with רשׁעים the wicked, ועם and with פעלי the workers און of iniquity, דברי which speak שׁלום peace עם to רעיהם their neighbors, ורעה but mischief בלבבם׃ in their hearts.}%
\verse{תן Give להם כפעלם them according to their deeds, וכרע and according to the wickedness מעלליהם of their endeavors: כמעשׂה them after the work ידיהם of their hands; תן give להם השׁב render גמולם to them their desert. להם׃}%
\verse{כי Because לא not יבינו they regard אל they regard פעלת the works יהוה of the LORD, ואל מעשׂה nor the operation ידיו of his hands, יהרסם he shall destroy ולא them, and not יבנם׃ build them up.}%
\verse{ברוך Blessed יהוה the LORD, כי because שׁמע he hath heard קול the voice תחנוני׃ of my supplications.}%
\verse{יהוה The LORD עזי my strength ומגני and my shield; בו בטח trusted לבי my heart ונעזרתי in him, and I am helped: ויעלז greatly rejoiceth; לבי therefore my heart ומשׁירי and with my song אהודנו׃ will I praise}%
\verse{יהוה The LORD עז their strength, למו ומעוז strength ישׁועות the saving משׁיחו of his anointed. הוא׃ and he}%
\verse{הושׁיעה Save את עמך thy people, וברך and bless את נחלתך thine inheritance: ורעם feed ונשׂאם them also, and lift them up עד forever. העולם׃ forever.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 29
\verseWithHeading{Praise to God for His Glory and Strength}{מזמור A Psalm לדוד of David. הבו Give ליהוה unto the LORD, בני אליםבו give ליהוה unto the LORD כבוד glory ועז׃ and strength.}%
\verse{הבו Give ליהוה unto the LORD כבוד the glory שׁמו due unto his name; השׁתחוו worship ליהוה the LORD בהדרת in the beauty קדשׁ׃ of holiness.}%
\verse{קול The voice יהוה of the LORD על upon המים the waters: אל the God הכבוד of glory הרעים thundereth: יהוה the LORD על upon מים waters. רבים׃ many}%
\verse{קול The voice יהוה of the LORD בכח powerful; קול the voice יהוה of the LORD בהדר׃ full of majesty.}%
\verse{קול The voice יהוה of the LORD שׁבר breaketh ארזים the cedars; וישׁבר breaketh יהוה yea, the LORD את ארזי the cedars הלבנון׃ of Lebanon.}%
\verse{וירקידם He maketh them also to skip כמו like עגל a calf; לבנון Lebanon ושׂרין and Sirion כמו like בן a young ראמים׃ unicorn.}%
\verse{קול The voice יהוה of the LORD חצב divideth להבות the flames אשׁ׃ of fire.}%
\verse{קול The voice יהוה of the LORD יחיל shaketh מדבר the wilderness; יחיל shaketh יהוה the LORD מדבר the wilderness קדשׁ׃ of Kadesh.}%
\verse{קול The voice יהוה of the LORD יחולל to calve, אילות maketh the hinds ויחשׂף and discovereth יערות the forests: ובהיכלו and in his temple כלו doth every one אמר speak of כבוד׃ glory.}%
\verse{יהוה The LORD למבול upon the flood; ישׁב sitteth וישׁב sitteth יהוה yea, the LORD מלך King לעולם׃ forever.}%
\verse{יהוה The LORD עז strength לעמו unto his people; יתן will give יהוה the LORD יברך will bless את עמו his people בשׁלום׃ with peace.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 30
\verseWithHeading{Thanksgiving for Answered Prayer}{מזמור A Psalm שׁיר Song חנכת the dedication הבית of the house לדוד׃ of David. ארוממך I will extol יהוה thee, O LORD; כי for דליתני thou hast lifted me up, ולא and hast not שׂמחת to rejoice איבי made my foes}%
\verse{יהוה O LORD אלהי my God, שׁועתי I cried אליך unto ותרפאני׃ thee, and thou hast healed}%
\verse{יהוה O LORD, העלית thou hast brought up מן from שׁאול the grave: נפשׁי my soul חייתני thou hast kept me alive, מיורדי בור׃}%
\verse{זמרו Sing ליהוה unto the LORD, חסידיו O ye saints והודו of his, and give thanks לזכר at the remembrance קדשׁו׃ of his holiness.}%
\verse{כי For רגע a moment; באפו his anger חיים life: ברצונו in his favor בערב for a night, ילין may endure בכי weeping ולבקר in the morning. רנה׃ but joy}%
\verse{ואני I אמרתי said, בשׁלוי And in my prosperity בל אמוט be moved. לעולם׃}%
\verse{יהוה LORD, ברצונך by thy favor העמדתה to stand להררי thou hast made my mountain עז strong: הסתרת thou didst hide פניך thy face, הייתי I was נבהל׃ troubled.}%
\verse{אליך to יהוה thee, O LORD; אקרא I cried ואל and unto אדני the LORD אתחנן׃ I made supplication.}%
\verse{מה What בצע profit בדמי in my blood, ברדתי when I go down אל to שׁחת the pit? היודך praise עפר Shall the dust היגיד thee? shall it declare אמתך׃ thy truth?}%
\verse{שׁמע Hear, יהוה O LORD, וחנני and have mercy יהוה upon me: LORD, היה be עזר׃ thou my helper.}%
\verse{הפכת Thou hast turned מספדי for me my mourning למחול into dancing: לי פתחת thou hast put off שׂקי my sackcloth, ותאזרני and girded שׂמחה׃ me with gladness;}%
\verse{למען To the end that יזמרך may sing praise כבוד glory ולא to thee, and not ידם be silent. יהוה O LORD אלהי my God, לעולם unto thee forever. אודך׃ I will give thanks}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 31
\verseWithHeading{Adonai Is a Fortress}{למנצח To the chief Musician, מזמור A Psalm לדוד׃ of David. בך יהוה In thee, O LORD, חסיתי do I put my trust; אל let me never אבושׁה be ashamed: לעולם let me never בצדקתך me in thy righteousness. פלטני׃ deliver}%
\verse{הטה Bow down אלי to אזנך thine ear מהרה me speedily: הצילני me; deliver היה be לי לצור rock, מעוז thou my strong לבית for a house מצודות of defense להושׁיעני׃ to save}%
\verse{כי For סלעי my rock ומצודתי and my fortress; אתה thou ולמען שׁמךנחני lead ותנהלני׃ me, and guide}%
\verse{תוציאני Pull me out מרשׁת of the net זו that טמנו they have laid privily לי כי for me: for אתה thou מעוזי׃ my strength.}%
\verse{בידך Into thine hand אפקיד I commit רוחי my spirit: פדיתה thou hast redeemed אותי יהוה me, O LORD אל God אמת׃ of truth.}%
\verse{שׂנאתי I have hated השׁמרים them that regard הבלי vanities: שׁוא lying ואני but I אל in יהוה the LORD. בטחתי׃ trust}%
\verse{אגילה I will be glad ואשׂמחה and rejoice בחסדך in thy mercy: אשׁר for ראית thou hast considered את עניי my trouble; ידעת thou hast known בצרות in adversities; נפשׁי׃ my soul}%
\verse{ולא And hast not הסגרתני shut me up ביד into the hand אויב of the enemy: העמדת thou hast set במרחב in a large room. רגלי׃ my feet}%
\verse{חנני Have mercy יהוה upon me, O LORD, כי for צר לי עשׁשׁה is consumed בכעס with grief, עיני mine eye נפשׁי my soul ובטני׃ and my belly.}%
\verse{כי For כלו is spent ביגון with grief, חיי my life ושׁנותי and my years באנחה with sighing: כשׁל faileth בעוני because of mine iniquity, כחי my strength ועצמי and my bones עשׁשׁו׃ are consumed.}%
\verse{מכל among all צררי mine enemies, הייתי I was חרפה a reproach ולשׁכני among my neighbors, מאד but especially ופחד and a fear למידעי to mine acquaintance: ראי they that did see בחוץ me without נדדו fled ממני׃ among all}%
\verse{נשׁכחתי I am forgotten כמת as a dead man מלב out of mind: הייתי I am ככלי vessel. אבד׃ like a broken}%
\verse{כי For שׁמעתי I have heard דבת the slander רבים of many: מגור fear מסביב on every side: בהוסדם while they took counsel יחד together עלי against לקחת to take away נפשׁי my life. זממו׃ me, they devised}%
\verse{ואני But I עליך in בטחתי trusted יהוה thee, O LORD: אמרתי I said, אלהי my God. אתה׃ Thou}%
\verse{בידך in thy hand: עתתי My times הצילני deliver מיד me from the hand אויבי of mine enemies, ומרדפי׃ and from them that persecute}%
\verse{האירה to shine פניך Make thy face על upon עבדך thy servant: הושׁיעני save בחסדך׃ me for thy mercies'}%
\verse{יהוה O LORD; אל Let me not אבושׁה be ashamed, כי for קראתיך I have called upon יבשׁו be ashamed, רשׁעים thee: let the wicked ידמו let them be silent לשׁאול׃ in the grave.}%
\verse{תאלמנה be put to silence; שׂפתי lips שׁקר Let the lying הדברות which speak על against צדיק the righteous. עתק grievous things בגאוה proudly ובוז׃ and contemptuously}%
\verse{מה how רב great טובך thy goodness, אשׁר which צפנת thou hast laid up ליראיך for them that fear פעלת thee; thou hast wrought לחסים for them that trust בך נגד in thee before בני the sons אדם׃ of men!}%
\verse{תסתירם Thou shalt hide בסתר them in the secret פניך of thy presence מרכסי from the pride אישׁ of man: תצפנם thou shalt keep them secretly בסכה in a pavilion מריב from the strife לשׁנות׃ of tongues.}%
\verse{ברוך Blessed יהוה כי for הפליא he hath showed me his marvelous kindness חסדו he hath showed me his marvelous kindness לי בעיר city. מצור׃ in a strong}%
\verse{ואני For I אמרתי said בחפזי in my haste, נגרזתי I am cut off מנגד from before עיניך thine eyes: אכן nevertheless שׁמעת thou heardest קול the voice תחנוני of my supplications בשׁועי when I cried אליך׃ unto}%
\verse{אהבו O love את יהוה the LORD, כל all חסידיו ye his saints: אמונים the faithful, נצר preserveth יהוה the LORD ומשׁלם rewardeth על and plentifully יתר and plentifully עשׂה doer. גאוה׃ the proud}%
\verse{חזקו Be of good courage, ויאמץ and he shall strengthen לבבכם your heart, כל all המיחלים ye that hope ליהוה׃ in the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 32
\verseWithHeading{Thanksgiving for Forgiveness of Sins}{לדוד of David, משׂכיל Maschil. אשׁרי Blessed נשׂוי forgiven, פשׁע transgression כסוי covered. חטאה׃ sin}%
\verse{אשׁרי Blessed אדם the man לא not יחשׁב imputeth יהוה unto whom the LORD לו עון iniquity, ואין no ברוחו and in whose spirit רמיה׃ guile.}%
\verse{כי When החרשׁתי I kept silence, בלו waxed old עצמי my bones בשׁאגתי through my roaring כל all היום׃ the day}%
\verse{כי For יומם day ולילה and night תכבד was heavy עלי upon ידך thy hand נהפך is turned לשׁדי me: my moisture בחרבני into the drought קיץ of summer. סלה׃ Selah.}%
\verse{חטאתי my sin אודיעך I acknowledged ועוני unto thee, and mine iniquity לא have I not כסיתי hid. אמרתי I said, אודה I will confess עלי my transgressions פשׁעי my transgressions ליהוה unto the LORD; ואתה and thou נשׂאת forgavest עון the iniquity חטאתי of my sin. סלה׃ Selah.}%
\verse{על For זאת this יתפלל pray כל shall every חסיד one that is godly אליך unto לעת thee in a time מצא when thou mayest be found: רק surely לשׁטף in the floods מים waters רבים of great אליו unto לא they shall not יגיעו׃ come nigh}%
\verse{אתה Thou סתר my hiding place; לי מצר me from trouble; תצרני thou shalt preserve רני with songs פלט of deliverance. תסובבני thou shalt compass me about סלה׃ Selah.}%
\verse{אשׂכילך I will instruct ואורך thee and teach בדרך thee in the way זו which תלך thou shalt go: איעצה עליךיני׃ thee with mine eye.}%
\verse{אל ye not תהיו Be כסוס as the horse, כפרד as the mule, אין have no הבין understanding: במתג with bit ורסן and bridle, עדיו whose mouth לבלום must be held in בל lest קרב they come near אליך׃ unto thee.}%
\verse{רבים Many מכאובים sorrows לרשׁע to the wicked: והבוטח but he that trusteth ביהוה in the LORD, חסד mercy יסובבנו׃ shall compass him about.}%
\verse{שׂמחו Be glad ביהוה in the LORD, וגילו and rejoice, צדיקים ye righteous: והרנינו and shout for joy, כל all ישׁרי upright לב׃ in heart.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 33
\verseWithHeading{Praise to Adonai for His Character and Creation}{רננו Rejoice צדיקים O ye righteous: ביהוה in the LORD, לישׁרים for the upright. נאוה is comely תהלה׃ praise}%
\verse{הודו Praise ליהוה the LORD בכנור with harp: בנבל unto him with the psaltery עשׂור an instrument of ten strings. זמרו׃ sing}%
\verse{שׁירו Sing לו שׁיר song; חדשׁ unto him a new היטיבו skilfully נגן play בתרועה׃ with a loud noise.}%
\verse{כי For ישׁר right; דבר the word יהוה of the LORD וכל and all מעשׂהו his works באמונה׃ in truth.}%
\verse{אהב He loveth צדקה righteousness ומשׁפט and judgment: חסד of the goodness יהוה of the LORD. מלאה is full הארץ׃ the earth}%
\verse{בדבר By the word יהוה of the LORD שׁמים were the heavens נעשׂו made; וברוח of them by the breath פיו of his mouth. כל and all צבאם׃ the host}%
\verse{כנס He gathereth כנד together as a heap: מי the waters הים of the sea נתן he layeth up באצרות in storehouses. תהומות׃ the depth}%
\verse{ייראו fear מיהוה כל Let all הארץ the earth ממנו of יגורו stand in awe כל let all ישׁבי the inhabitants תבל׃ of the world}%
\verse{כי For הוא he אמר spoke, ויהי and it was הוא he צוה commanded, ויעמד׃ and it stood fast.}%
\verse{יהוה The LORD הפיר to naught: עצת bringeth the counsel גוים of the heathen הניא of none effect. מחשׁבות he maketh the devices עמים׃ of the people}%
\verse{עצת The counsel יהוה of the LORD לעולם forever, תעמד standeth מחשׁבות the thoughts לבו of his heart לדר to all ודר׃ generations.}%
\verse{אשׁרי Blessed הגוי the nation אשׁר whose יהוה the LORD; אלהיו God העם the people בחר he hath chosen לנחלה׃ for his own inheritance.}%
\verse{משׁמים from heaven; הביט looketh יהוה The LORD ראה he beholdeth את כל all בני the sons האדם׃ of men.}%
\verse{ממכון שׁבתו of his habitation השׁגיח he looketh אל upon כל all ישׁבי the inhabitants הארץ׃ of the earth.}%
\verse{היצר He fashioneth יחד alike; לבם their hearts המבין he considereth אל he considereth כל all מעשׂיהם׃ their works.}%
\verse{אין There is no המלך king נושׁע saved ברב by the multitude חיל of a host: גבור a mighty man לא is not ינצל delivered ברב by much כח׃ strength.}%
\verse{שׁקר a vain thing הסוס A horse לתשׁועה for safety: וברב by his great חילו strength. לא neither ימלט׃ shall he deliver}%
\verse{הנה Behold, עין the eye יהוה of the LORD אל upon יראיו them that fear למיחלים him, upon them that hope לחסדו׃ in his mercy;}%
\verse{להציל To deliver ממות from death, נפשׁם their soul ולחיותם and to keep them alive ברעב׃ in famine.}%
\verse{נפשׁנו Our soul חכתה waiteth ליהוה for the LORD: עזרנו our help ומגננו and our shield. הוא׃ he}%
\verse{כי For בו ישׂמח shall rejoice לבנו our heart כי in him, because בשׁם name. קדשׁו in his holy בטחנו׃ we have trusted}%
\verse{יהי be חסדך Let thy mercy, יהוה O LORD, עלינו upon כאשׁר us, according as יחלנו׃ we hope}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 34
\verseWithHeading{Thanksgiving for Adonai’s Deliverance}{לדוד of David, בשׁנותו when he changed את טעמו his behavior לפני before אבימלך Abimelech; ויגרשׁהו who drove him away, וילך׃ and he departed. אברכה I will bless את יהוה the LORD בכל at all עת times: תמיד continually תהלתו his praise בפי׃ in my mouth.}%
\verse{ביהוה in the LORD: תתהלל shall make her boast נפשׁי My soul ישׁמעו shall hear ענוים the humble וישׂמחו׃ and be glad.}%
\verse{גדלו O magnify ליהוה the LORD אתי with ונרוממה me, and let us exalt שׁמו his name יחדו׃ together.}%
\verse{דרשׁתי I sought את יהוה the LORD, וענני and he heard ומכל me from all מגורותי my fears. הצילני׃ me, and delivered}%
\verse{הביטו They looked אליו unto ונהרו him, and were lightened: ופניהם and their faces אל were not יחפרו׃ ashamed.}%
\verse{זה This עני poor קרא man cried, ויהוה and the LORD שׁמע heard ומכל him out of all צרותיו his troubles. הושׁיעו׃ and saved}%
\verse{חנה encampeth מלאך The angel יהוה of the LORD סביב round about ליראיו them that fear ויחלצם׃ him, and delivereth}%
\verse{טעמו O taste וראו and see כי that טוב good: יהוה the LORD אשׁרי blessed הגבר the man יחסה׃ trusteth}%
\verse{יראו O fear את יהוה the LORD, קדשׁיו ye his saints: כי for אין no מחסור want ליראיו׃ to them that fear}%
\verse{כפירים The young lions רשׁו do lack, ורעבו and suffer hunger: ודרשׁי but they that seek יהוה the LORD לא shall not יחסרו want כל any טוב׃ good}%
\verse{לכו Come, בנים ye children, שׁמעו hearken לי יראת you the fear יהוה of the LORD. אלמדכם׃ unto me: I will teach}%
\verse{מי What האישׁ man החפץ desireth חיים life, אהב loveth ימים days, לראות that he may see טוב׃ good?}%
\verse{נצר Keep לשׁונך thy tongue מרע from evil, ושׂפתיך and thy lips מדבר from speaking מרמה׃ guile.}%
\verse{סור Depart מרע from evil, ועשׂה and do טוב good; בקשׁ seek שׁלום peace, ורדפהו׃ and pursue}%
\verse{עיני The eyes יהוה of the LORD אל upon צדיקים the righteous, ואזניו and his ears אל unto שׁועתם׃ their cry.}%
\verse{פני The face יהוה of the LORD בעשׂי against them that do רע evil, להכרית to cut off מארץ of them from the earth. זכרם׃ the remembrance}%
\verse{צעקו cry, ויהוה and the LORD שׁמע heareth, ומכל them out of all צרותם their troubles. הצילם׃ and delivereth}%
\verse{קרוב nigh יהוה The LORD לנשׁברי unto them that are of a broken לב heart; ואת דכאי such as be of a contrite רוח spirit. יושׁיע׃ and saveth}%
\verse{רבות Many רעות the afflictions צדיק of the righteous: ומכלם him out of them all. יצילנו delivereth יהוה׃ but the LORD}%
\verse{שׁמר He keepeth כל all עצמותיו his bones: אחת one מהנה of them לא not נשׁברה׃ is broken.}%
\verse{תמותת shall slay רשׁע the wicked: רעה Evil ושׂנאי and they that hate צדיק the righteous יאשׁמו׃ shall be desolate.}%
\verse{פודה redeemeth יהוה The LORD נפשׁ the soul עבדיו of his servants: ולא and none יאשׁמו in him shall be desolate. כל and none החסים׃ of them that trust}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 35
\verseWithHeading{A Prayer for Rescue from Enemies}{לדוד of David. ריבה Plead יהוה O LORD, את יריבי them that strive לחם with me: fight against את לחמי׃ them that fight against}%
\verse{החזק Take hold מגן of shield וצנה and buckler, וקומה and stand up בעזרתי׃ for mine help.}%
\verse{והרק Draw out חנית also the spear, וסגר and stop לקראת against רדפי them that persecute אמר me: say לנפשׁי unto my soul, ישׁעתך thy salvation. אני׃ I}%
\verse{יבשׁו Let them be confounded ויכלמו and put to shame מבקשׁי that seek after נפשׁי my soul: יסגו let them be turned אחור back ויחפרו and brought to confusion חשׁבי that devise רעתי׃ my hurt.}%
\verse{יהיו Let them be כמץ as chaff לפני before רוח the wind: ומלאך and let the angel יהוה of the LORD דוחה׃ chase}%
\verse{יהי be דרכם Let their way חשׁך dark וחלקלקות and slippery: ומלאך and let the angel יהוה of the LORD רדפם׃ persecute}%
\verse{כי For חנם without cause טמנו have they hid לי שׁחת a pit, רשׁתם for me their net חנם without cause חפרו they have digged לנפשׁי׃ for my soul.}%
\verse{תבואהו come upon שׁואה Let destruction לא him at unawares; ידע him at unawares; ורשׁתו and let his net אשׁר that טמן he hath hid תלכדו catch בשׁואה himself: into that very destruction יפל׃ let him fall.}%
\verse{ונפשׁי And my soul תגיל shall be joyful ביהוה in the LORD: תשׂישׂ it shall rejoice בישׁועתו׃ in his salvation.}%
\verse{כל All עצמותי my bones תאמרנה shall say, יהוה LORD, מי who כמוך like unto thee, מציל which deliverest עני the poor מחזק from him that is too strong ממנו from him that is too strong ועני him, yea, the poor ואביון and the needy מגזלו׃ from him that spoileth}%
\verse{יקומון did rise up; עדי witnesses חמס False אשׁר that לא not. ידעתי I knew ישׁאלוני׃ they laid to my charge}%
\verse{ישׁלמוני They rewarded רעה me evil תחת for טובה good שׁכול the spoiling לנפשׁי׃ of my soul.}%
\verse{ואני But as for me, בחלותם when they were sick, לבושׁי my clothing שׂק sackcloth: עניתי I humbled בצום with fasting; נפשׁי my soul ותפלתי and my prayer על into חיקי mine own bosom. תשׁוב׃ returned}%
\verse{כרע as though my friend כאח brother: לי התהלכתי I behaved myself כאבל as one that mourneth אם mother. קדר heavily, שׁחותי׃ I bowed down}%
\verse{ובצלעי But in mine adversity שׂמחו they rejoiced, ונאספו and gathered themselves together: נאספו gathered themselves together עלי against נכים the abjects ולא not; ידעתי me, and I knew קרעו they did tear ולא not: דמו׃ and ceased}%
\verse{בחנפי With hypocritical לעגי mockers מעוג in feasts, חרק they gnashed עלי upon שׁנימו׃ me with their teeth.}%
\verse{אדני Lord, כמה how long תראה wilt thou look on? השׁיבה rescue נפשׁי my soul משׁאיהם from their destructions, מכפירים from the lions. יחידתי׃ my darling}%
\verse{אודך I will give thee thanks בקהל congregation: רב in the great בעם people. עצום thee among much אהללך׃ I will praise}%
\verse{אל Let not ישׂמחו rejoice לי איבי them that are mine enemies שׁקר wrongfully שׂנאי that hate חנם me without a cause. יקרצו over me: let them wink עין׃ with the eye}%
\verse{כי For לא not שׁלום peace: ידברו they speak ועל against רגעי quiet ארץ in the land. דברי matters מרמות deceitful יחשׁבון׃ but they devise}%
\verse{וירחיבו Yea, they opened עלי wide against פיהם their mouth אמרו me, said, האח Aha, האח aha, ראתה hath seen עינינו׃ our eye}%
\verse{ראיתה thou hast seen, יהוה O LORD: אל keep not תחרשׁ silence: אדני O Lord, אל be not תרחק far ממני׃ from}%
\verse{העירה Stir up thyself, והקיצה and awake למשׁפטי to my judgment, אלהי my God ואדני and my Lord. לריבי׃ unto my cause,}%
\verse{שׁפטני Judge כצדקך according to thy righteousness; יהוה me, O LORD אלהי my God, ואל and let them not ישׂמחו׃ rejoice}%
\verse{אל Let them not יאמרו say בלבם in their hearts, האח Ah, נפשׁנו so would we have it: אל let them not יאמרו say, בלענוהו׃ We have swallowed him up.}%
\verse{יבשׁו Let them be ashamed ויחפרו and brought to confusion יחדו together שׂמחי that rejoice רעתי at mine hurt: ילבשׁו let them be clothed בשׁת with shame וכלמה and dishonor המגדילים that magnify עלי׃ against}%
\verse{ירנו Let them shout for joy, וישׂמחו and be glad, חפצי that favor צדקי my righteous cause: ויאמרו yea, let them say תמיד continually, יגדל be magnified, יהוה Let the LORD החפץ which hath pleasure שׁלום in the prosperity עבדו׃ of his servant.}%
\verse{ולשׁוני And my tongue תהגה shall speak צדקך of thy righteousness כל all היום the day תהלתך׃ of thy praise}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 36
\verseWithHeading{Human Wickedness and God’s Love}{למנצח To the chief Musician, לעבד the servant יהוה of the LORD. לדוד׃ of David נאם saith פשׁע The transgression לרשׁע of the wicked בקרב within לבי my heart, אין no פחד fear אלהים of God לנגד before עיניו׃ his eyes.}%
\verse{כי For החליק he flattereth אליו he flattereth בעיניו himself in his own eyes, למצא be found עונו until his iniquity לשׂנא׃ to be hateful.}%
\verse{דברי The words פיו of his mouth און iniquity ומרמה and deceit: חדל he hath left off להשׂכיל to be wise, להיטיב׃ to do good.}%
\verse{און mischief יחשׁב He deviseth על upon משׁכבו his bed; יתיצב he setteth himself על in דרך a way לא not טוב good; רע evil. לא not ימאס׃ he abhorreth}%
\verse{יהוה O LORD, בהשׁמים in the heavens; חסדך Thy mercy, אמונתך thy faithfulness עד unto שׁחקים׃ the clouds.}%
\verse{צדקתך Thy righteousness כהררי mountains; אל like the great משׁפטך thy judgments תהום deep: רבה a great אדם man ובהמה and beast. תושׁיע thou preservest יהוה׃ O LORD,}%
\verse{מה How יקר excellent חסדך thy lovingkindness, אלהים O God! ובני therefore the children אדם of men בצל under the shadow כנפיך of thy wings. יחסיון׃ put their trust}%
\verse{ירוין They shall be abundantly satisfied מדשׁן with the fatness ביתך of thy house; ונחל of the river עדניך of thy pleasures. תשׁקם׃ and thou shalt make them drink}%
\verse{כי For עמך with מקור thee the fountain חיים of life: באורך in thy light נראה shall we see אור׃ light.}%
\verse{משׁך O continue חסדך thy lovingkindness לידעיך unto them that know וצדקתך thee; and thy righteousness לישׁרי to the upright לב׃ in heart.}%
\verse{אל Let not תבואני come against רגל the foot גאוה of pride ויד the hand רשׁעים of the wicked אל me, and let not תנדני׃ remove}%
\verse{שׁם There נפלו fallen: פעלי are the workers און of iniquity דחו they are cast down, ולא and shall not יכלו be able קום׃ to rise.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 37
\verseWithHeading{The Protection of the Righteous and the Destruction of the Wicked}{לדוד of David. אל neither תתחר במרעים because of evildoers, אל תקנא be thou envious בעשׂי against the workers עולה׃ of iniquity.}%
\verse{כי For כחציר like the grass, מהרה they shall soon ימלו be cut down וכירק as the green דשׁא herb. יבולון׃ and wither}%
\verse{בטח Trust ביהוה in the LORD, ועשׂה and do טוב good; שׁכן shalt thou dwell ארץ in the land, ורעה thou shalt be fed. אמונה׃ and verily}%
\verse{והתענג Delight thyself על also in יהוה the LORD; ויתן and he shall give לך משׁאלת thee the desires לבך׃ of thine heart.}%
\verse{גול Commit על unto יהוה the LORD; דרכך thy way ובטח trust עליו also in והוא him; and he יעשׂה׃ shall bring to pass.}%
\verse{והוציא And he shall bring forth כאור as the light, צדקך thy righteousness ומשׁפטך and thy judgment כצהרים׃ as the noonday.}%
\verse{דום Rest ליהוה in the LORD, והתחולל and wait patiently לו אל for him: fret not thyself תתחר for him: fret not thyself במצליח because of him who prospereth דרכו in his way, באישׁ because of the man עשׂה who bringeth wicked devices to pass. מזמות׃ who bringeth wicked devices to pass.}%
\verse{הרף Cease מאף from anger, ועזב and forsake חמה wrath: אל fret not thyself תתחר fret not thyself אך in any wise להרע׃ to do evil.}%
\verse{כי For מרעים evildoers יכרתון shall be cut off: וקוי but those that wait upon יהוה the LORD, המה they יירשׁו shall inherit ארץ׃ the earth.}%
\verse{ועוד For yet מעט a little ואין not רשׁע while, and the wicked והתבוננת yea, thou shalt diligently consider על : yea, thou shalt diligently consider מקומו his place, ואיננו׃ and it not}%
\verse{וענוים But the meek יירשׁו shall inherit ארץ the earth; והתענגו and shall delight themselves על in רב the abundance שׁלום׃ of peace.}%
\verse{זמם plotteth רשׁע The wicked לצדיק against the just, וחרק and gnasheth עליו upon שׁניו׃ him with his teeth.}%
\verse{אדני The Lord ישׂחק shall laugh לו כי at him: for ראה he seeth כי that יבא is coming. יומו׃ his day}%
\verse{חרב the sword, פתחו have drawn out רשׁעים The wicked ודרכו and have bent קשׁתם their bow, להפיל to cast down עני the poor ואביון and needy, לטבוח to slay ישׁרי such as be of upright דרך׃ conversation.}%
\verse{חרבם Their sword תבוא shall enter בלבם into their own heart, וקשׁתותם and their bows תשׁברנה׃ shall be broken.}%
\verse{טוב man hath better מעט A little לצדיק that a righteous מהמון than the riches רשׁעים wicked. רבים׃ of many}%
\verse{כי For זרועות the arms רשׁעים of the wicked תשׁברנה shall be broken: וסומך upholdeth צדיקים the righteous. יהוה׃ but the LORD}%
\verse{יודע knoweth יהוה The LORD ימי the days תמימם of the upright: ונחלתם and their inheritance לעולם forever. תהיה׃ shall be}%
\verse{לא They shall not יבשׁו be ashamed בעת time: רעה in the evil ובימי and in the days רעבון of famine ישׂבעו׃ they shall be satisfied.}%
\verse{כי But רשׁעים the wicked יאבדו shall perish, ואיבי and the enemies יהוה of the LORD כיקר as the fat כרים of lambs: כלו they shall consume; בעשׁן into smoke כלו׃ shall they consume away.}%
\verse{לוה borroweth, רשׁע The wicked ולא not ישׁלם and payeth וצדיק again: but the righteous חונן showeth mercy, ונותן׃ and giveth.}%
\verse{כי For מברכיו blessed יירשׁו of him shall inherit ארץ the earth; ומקלליו and cursed יכרתו׃ of him shall be cut off.}%
\verse{מיהוה מצעדי The steps גבר of a man כוננו are ordered ודרכו in his way. יחפץ׃ and he delighteth}%
\verse{כי Though יפל he fall, לא he shall not יוטל be utterly cast down: כי for יהוה the LORD סומך upholdeth ידו׃ his hand.}%
\verse{נער young, הייתי I have been גם and זקנתי am old; ולא yet have I not ראיתי seen צדיק the righteous נעזב forsaken, וזרעו nor his seed מבקשׁ begging לחם׃ bread.}%
\verse{כל היוםונן merciful, ומלוה and lendeth; וזרעו and his seed לברכה׃ blessed.}%
\verse{סור Depart מרע evil, ועשׂה and do טוב good; ושׁכן and dwell לעולם׃ forevermore.}%
\verse{כי For יהוה the LORD אהב loveth משׁפט judgment, ולא not יעזב and forsaketh את חסידיו his saints; לעולם forever: נשׁמרו they are preserved וזרע but the seed רשׁעים of the wicked נכרת׃ shall be cut off.}%
\verse{צדיקים The righteous יירשׁו shall inherit ארץ the land, וישׁכנו and dwell לעד forever. עליה׃ therein}%
\verse{פי The mouth צדיק of the righteous יהגה speaketh חכמה wisdom, ולשׁונו and his tongue תדבר talketh משׁפט׃ of judgment.}%
\verse{תורת The law אלהיו of his God בלבו in his heart; לא none תמעד shall slide. אשׁריו׃ of his steps}%
\verse{צופה watcheth רשׁע The wicked לצדיק the righteous, ומבקשׁ and seeketh להמיתו׃ to slay}%
\verse{יהוה The LORD לא will not יעזבנו leave בידו him in his hand, ולא nor ירשׁיענו condemn בהשׁפטו׃ him when he is judged.}%
\verse{קוה Wait אל on יהוה the LORD, ושׁמר and keep דרכו his way, וירוממך and he shall exalt לרשׁת thee to inherit ארץ the land: בהכרת are cut off, רשׁעים when the wicked תראה׃ thou shalt see}%
\verse{ראיתי I have seen רשׁע the wicked עריץ in great power, ומתערה and spreading himself כאזרח bay tree. רענן׃ like a green}%
\verse{ויעבר Yet he passed away, והנה and, lo, איננו he not: ואבקשׁהו yea, I sought ולא him, but he could not נמצא׃ be found.}%
\verse{שׁמר Mark תם the perfect וראה and behold ישׁר the upright: כי for אחרית the end לאישׁ of man שׁלום׃ peace.}%
\verse{ופשׁעים But the transgressors נשׁמדו shall be destroyed יחדו together: אחרית the end רשׁעים of the wicked נכרתה׃ shall be cut off.}%
\verse{ותשׁועת But the salvation צדיקים of the righteous מיהוה מעוזם their strength בעת in the time צרה׃ of trouble.}%
\verse{ויעזרם shall help יהוה And the LORD ויפלטם them, and deliver יפלטם them: he shall deliver מרשׁעים them from the wicked, ויושׁיעם and save כי them, because חסו׃ they trust}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 38
\verseWithHeading{A Prayer of Repentance}{מזמור A Psalm לדוד of David, להזכיר׃ to bring to remembrance. יהוה O LORD, אל me not בקצפך in thy wrath: תוכיחני rebuke ובחמתך me in thy hot displeasure. תיסרני׃ neither chasten}%
\verse{כי For חציך thine arrows נחתו stick fast בי ותנחת presseth me sore. עלי presseth me sore. ידך׃ in me, and thy hand}%
\verse{אין no מתם soundness בבשׂרי in my flesh מפני because זעמך of thine anger; אין neither שׁלום rest בעצמי in my bones מפני because חטאתי׃ of my sin.}%
\verse{כי For עונתי mine iniquities עברו are gone over ראשׁי mine head: כמשׂא burden כבד as a heavy יכבדו they are too heavy ממני׃ for}%
\verse{הבאישׁו stink נמקו are corrupt חבורתי My wounds מפני because אולתי׃ of my foolishness.}%
\verse{נעויתי I am troubled; שׁחתי I am bowed down עד greatly; מאד greatly; כל all היום the day קדר mourning הלכתי׃ I go}%
\verse{כי For כסלי my loins מלאו are filled נקלה with a loathsome ואין and no מתם soundness בבשׂרי׃ in my flesh.}%
\verse{נפוגותי I am feeble ונדכיתי broken: עד and sore מאד and sore שׁאגתי I have roared מנהמת by reason of the disquietness לבי׃ of my heart.}%
\verse{אדני Lord, נגדך before כל all תאותי my desire ואנחתי thee; and my groaning ממך from לא is not נסתרה׃ hid}%
\verse{לבי My heart סחרחר panteth, עזבני faileth כחי my strength ואור me: as for the light עיני of mine eyes, גם also הם it אין is gone אתי׃ from}%
\verse{אהבי My lovers ורעי and my friends מנגד נגעיעמדו stand aloof וקרובי and my kinsmen מרחק afar off. עמדו׃ stand}%
\verse{וינקשׁו lay snares מבקשׁי They also that seek after נפשׁי my life ודרשׁי and they that seek רעתי my hurt דברו speak הוות mischievous things, ומרמות deceits כל all היום the day יהגו׃ and imagine}%
\verse{ואני But I, כחרשׁ as a deaf לא not; אשׁמע heard וכאלם and as a dumb man לא not יפתח openeth פיו׃ his mouth.}%
\verse{ואהי Thus I was כאישׁ as a man אשׁר that לא not, שׁמע heareth ואין no בפיו and in whose mouth תוכחות׃ reproofs.}%
\verse{כי For לך יהוה הוחלתי do I hope: אתה thou תענה wilt hear, אדני O Lord אלהי׃ my God.}%
\verse{כי For אמרתי I said, פן lest ישׂמחו they should rejoice לי במוט slippeth, רגלי over me: when my foot עלי against הגדילו׃ they magnify}%
\verse{כי For אני I לצלע to halt, נכון ready ומכאובי and my sorrow נגדי before תמיד׃ continually}%
\verse{כי For עוני mine iniquity; אגיד I will declare אדאג I will be sorry מחטאתי׃ for my sin.}%
\verse{ואיבי But mine enemies חיים lively, עצמו they are strong: ורבו are multiplied. שׂנאי and they that hate שׁקר׃ me wrongfully}%
\verse{ומשׁלמי They also that render רעה evil תחת for טובה good ישׂטנוני are mine adversaries; תחת because רדופי I follow טוב׃}%
\verse{אל me not, תעזבני Forsake יהוה O LORD: אלהי O my God, אל be not תרחק far ממני׃ from}%
\verse{חושׁה Make haste לעזרתי to help אדני me, O Lord תשׁועתי׃ my salvation.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 39
\verseWithHeading{The Brevity of Human Life}{למנצח To the chief Musician, לידיתון to Jeduthun, מזמור A Psalm לדוד׃ of David. אמרתי I said, אשׁמרה I will take heed דרכי to my ways, מחטוא בלשׁוני with my tongue: אשׁמרה I will keep לפי my mouth מחסום with a bridle, בעד רשׁע the wicked לנגדי׃ is before}%
\verse{נאלמתי I was dumb דומיה with silence, החשׁיתי I held my peace, מטוב from good; וכאבי and my sorrow נעכר׃ was stirred.}%
\verse{חם was hot לבי My heart בקרבי within בהגיגי me, while I was musing תבער burned: אשׁ the fire דברתי spoke בלשׁוני׃ I with my tongue,}%
\verse{הודיעני make me to know יהוה LORD, קצי mine end, ומדת and the measure ימי of my days, מה what היא it אדעה I may know מה how חדל frail אני׃ I}%
\verse{הנה Behold, טפחות a handbreadth; נתתה thou hast made ימי my days וחלדי and mine age כאין as nothing נגדך before אך thee: verily כל every הבל vanity. כל altogether אדם man נצב at his best state סלה׃ Selah.}%
\verse{אך Surely בצלם in a vain show: יתהלך walketh אישׁ every man אך surely הבל in vain: יהמיון they are disquieted יצבר he heapeth up ולא not ידע and knoweth מי who אספם׃ shall gather}%
\verse{ועתה And now, מה what קויתי wait אדני Lord, תוחלתי I for? my hope לך היא׃}%
\verse{מכל me from all פשׁעי my transgressions: הצילני Deliver חרפת the reproach נבל of the foolish. אל me not תשׂימני׃ make}%
\verse{נאלמתי I was dumb, לא not אפתח I opened פי my mouth; כי because אתה thou עשׂית׃ didst}%
\verse{הסר Remove מעלי away from נגעך thy stroke מתגרת by the blow ידך of thine hand. אני me: I כליתי׃ am consumed}%
\verse{בתוכחות When thou with rebukes על for עון iniquity, יסרת dost correct אישׁ man ותמס to consume away כעשׁ like a moth: חמודו thou makest his beauty אך surely הבל vanity. כל every אדם man סלה׃ Selah.}%
\verse{שׁמעה Hear תפלתי my prayer, יהוה O LORD, ושׁועתי unto my cry; האזינה and give ear אל at דמעתי my tears: אל hold not thy peace תחרשׁ hold not thy peace כי for גר a stranger אנכי I עמך with תושׁב thee, a sojourner, ככל as all אבותי׃ my fathers}%
\verse{השׁע ממניאבליגה me, that I may recover strength, בטרם before אלך I go ואינני׃ hence, and be no}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 40
\verseWithHeading{God’s Faithfulness and Deliverance}{למנצח To the chief Musician, לדוד of David. מזמור׃ A Psalm קוה I waited קויתי patiently יהוה for the LORD; ויט and he inclined אלי unto וישׁמע me, and heard שׁועתי׃ my cry.}%
\verse{ויעלני He brought me up מבור pit, שׁאון also out of a horrible מטיט clay, היון out of the miry ויקם and set על upon סלע a rock, רגלי my feet כונן established אשׁרי׃ my goings.}%
\verse{ויתן And he hath put בפי in my mouth, שׁיר song חדשׁ a new תהלה praise לאלהינו unto our God: יראו shall see רבים many וייראו and fear, ויבטחו and shall trust ביהוה׃ in the LORD.}%
\verse{אשׁרי Blessed הגבר that man אשׁר that שׂם maketh יהוה the LORD מבטחו his trust, ולא not פנה and respecteth אל and respecteth רהבים the proud, ושׂטי nor such as turn aside כזב׃ to lies.}%
\verse{רבות Many, עשׂית hast done, אתה thou יהוה O LORD אלהי my God, נפלאתיך thy wonderful works ומחשׁבתיך and thy thoughts אלינו to אין usward: they cannot ערך be reckoned up in order אליך unto thee: אגידה I would declare ואדברה and speak עצמו they are more מספר׃ than can be numbered.}%
\verse{זבח Sacrifice ומנחה and offering לא thou didst not חפצת desire; אזנים mine ears כרית hast thou opened: לי עולה burnt offering וחטאה and sin offering לא hast thou not שׁאלת׃ required.}%
\verse{אז Then אמרתי said הנה I, Lo, באתי I come: במגלת in the volume ספר of the book כתוב written עלי׃ of}%
\verse{לעשׂות to do רצונך thy will, אלהי O my God: חפצתי I delight ותורתך yea, thy law בתוך within מעי׃ my heart.}%
\verse{בשׂרתי I have preached צדק righteousness בקהל congregation: רב in the great הנה lo, שׂפתי my lips, לא I have not אכלא refrained יהוה O LORD, אתה thou ידעת׃ knowest.}%
\verse{צדקתך thy righteousness לא I have not כסיתי hid בתוך within לבי my heart; אמונתך thy faithfulness ותשׁועתך and thy salvation: אמרתי I have declared לא I have not כחדתי concealed חסדך thy lovingkindness ואמתך and thy truth לקהל congregation. רב׃ from the great}%
\verse{אתה thou יהוה me, O LORD: לא not תכלא Withhold רחמיך thy tender mercies ממני from חסדך let thy lovingkindness ואמתך and thy truth תמיד continually יצרוני׃ preserve}%
\verse{כי For אפפו have compassed me about: עלי have compassed me about: רעות evils עד innumerable אין מספרשׂיגוני have taken hold עונתי mine iniquities ולא upon me, so that I am not יכלתי able לראות to look up; עצמו they are more משׂערות than the hairs ראשׁי of mine head: ולבי therefore my heart עזבני׃ faileth}%
\verse{רצה Be pleased, יהוה O LORD, להצילני to deliver יהוה me: O LORD, לעזרתי to help חושׁה׃ make haste}%
\verse{יבשׁו Let them be ashamed ויחפרו and confounded יחד together מבקשׁי that seek after נפשׁי my soul לספותה to destroy יסגו it; let them be driven אחור backward ויכלמו and put to shame חפצי that wish רעתי׃ me evil.}%
\verse{ישׁמו Let them be desolate על for עקב a reward בשׁתם of their shame האמרים that say לי האח unto me, Aha, האח׃ aha.}%
\verse{ישׂישׂו thee rejoice וישׂמחו and be glad בך כל Let all מבקשׁיך those that seek יאמרו say תמיד continually, יגדל be magnified. יהוה The LORD אהבי in thee: let such as love תשׁועתך׃ thy salvation}%
\verse{ואני But I עני poor ואביון and needy; אדני the Lord יחשׁב thinketh לי עזרתי my help ומפלטי and my deliverer; אתה upon me: thou אלהי O my God. אל make no תאחר׃ tarrying,}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 41
\verseWithHeading{Thanksgiving for God’s Provision in Time of Sickness}{למנצח To the chief Musician, מזמור A Psalm לדוד׃ of David. אשׁרי Blessed משׂכיל he that considereth אל he that considereth דל the poor: ביום him in time רעה of trouble. ימלטהו will deliver יהוה׃ the LORD}%
\verse{יהוה The LORD ישׁמרהו will preserve ויחיהו him, and keep him alive; יאשׁר he shall be blessed בארץ upon the earth: ואל and thou wilt not תתנהו deliver בנפשׁ him unto the will איביו׃ of his enemies.}%
\verse{יהוה The LORD יסעדנו will strengthen על him upon ערשׂ the bed דוי of languishing: כל all משׁכבו his bed הפכת thou wilt make בחליו׃ in his sickness.}%
\verse{אני I אמרתי said, יהוה LORD, חנני be merciful רפאה unto me: heal נפשׁי my soul; כי for חטאתי׃ I have sinned}%
\verse{אויבי Mine enemies יאמרו speak רע evil לי מתי of me, When ימות shall he die, ואבד perish? שׁמו׃ and his name}%
\verse{ואם And if בא he come לראות to see שׁוא vanity: ידבר he speaketh לבו his heart יקבץ gathereth און iniquity לו יצא to itself; he goeth לחוץ abroad, ידבר׃ he telleth}%
\verse{יחד together עלי against יתלחשׁו me whisper כל All שׂנאי that hate עלי me: against יחשׁבו me do they devise רעה׃ my hurt.}%
\verse{דבר disease, בליעל An evil יצוק cleaveth fast בו ואשׁר unto him: and that שׁכב he lieth לא no יוסיף more. לקום׃ he shall rise up}%
\verse{גם Yea, אישׁ mine own familiar friend, שׁלומי mine own familiar friend, אשׁר in whom בטחתי I trusted, בו אוכל which did eat לחמי of my bread, הגדיל hath lifted up עלי against עקב׃ heel}%
\verse{ואתה But thou, יהוה חנני be merciful והקימני unto me, and raise me up, ואשׁלמה that I may requite להם׃}%
\verse{בזאת By this ידעתי I know כי that חפצת thou favorest בי כי me, because לא doth not יריע triumph איבי mine enemy עלי׃ over}%
\verse{ואני And as for me, בתמי me in mine integrity, תמכת thou upholdest בי ותציבני and settest לפניך me before thy face לעולם׃ forever.}%
\verse{ברוך Blessed יהוה the LORD אלהי God ישׂראל of Israel מהעולם from everlasting, ועד and to העולם everlasting. אמן Amen, ואמן׃ and Amen.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 42
\verseWithHeading{Hope in God in the Midst of Despair}{למנצח To the chief Musician, משׂכיל Maschil, לבני for the sons קרח׃ of Korah. כאיל As the hart תערג panteth על after אפיקי brooks, מים the water כן so נפשׁי my soul תערג panteth אליך after אלהים׃ thee, O God.}%
\verse{צמאה thirsteth נפשׁי My soul לאלהים for God, לאל God: חי for the living מתי when אבוא shall I come ואראה and appear פני before אלהים׃ God?}%
\verse{היתה have been לי דמעתי My tears לחם my meat יומם day ולילה and night, באמר say אלי unto כל while they continually היום while they continually איה me, Where אלהיך׃ thy God?}%
\verse{אלה these אזכרה When I remember ואשׁפכה I pour out עלי in נפשׁי my soul כי me: for אעבר I had gone בסך with the multitude, אדדם I went עד with them to בית the house אלהים of God, בקול with the voice רנה of joy ותודה and praise, המון with a multitude חוגג׃ that kept holyday.}%
\verse{מה Why תשׁתוחחי art thou cast down, נפשׁי O my soul? ותהמי and art thou disquieted עלי in הוחילי me? hope לאלהים thou in God: כי for עוד I shall yet אודנו praise ישׁועות him the help פניו׃ of his countenance.}%
\verse{אלהי O my God, עלי within נפשׁי my soul תשׁתוחח is cast down על me: therefore כן me: therefore אזכרך will I remember מארץ thee from the land ירדן of Jordan, וחרמונים and of the Hermonites, מהר from the hill מצער׃ Mizar.}%
\verse{תהום Deep אל unto תהום deep קורא calleth לקול at the noise צנוריך of thy waterspouts: כל all משׁבריך thy waves וגליך and thy billows עלי over עברו׃ are gone}%
\verse{יומם in the daytime, יצוה will command יהוה the LORD חסדו his lovingkindness ובלילה and in the night שׁירה his song עמי with תפלה me, my prayer לאל unto the God חיי׃ of my life.}%
\verse{אומרה I will say לאל unto God סלעי my rock, למה Why שׁכחתני hast thou forgotten למה me? why קדר I mourning אלך go בלחץ because of the oppression אויב׃ of the enemy?}%
\verse{ברצח with a sword בעצמותי in my bones, חרפוני reproach צוררי mine enemies באמרם me; while they say אלי unto כל daily היום daily איה me, Where אלהיך׃ thy God?}%
\verse{מה Why תשׁתוחחי art thou cast down, נפשׁי O my soul? ומה and why תהמי art thou disquieted עלי within הוחילי me? hope לאלהים thou in God: כי for עוד I shall yet אודנו praise ישׁועת him, the health פני of my countenance, ואלהי׃ and my God.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 43
\verseWithHeading{A Prayer for Rescue}{שׁפטני Judge אלהים me, O God, וריבה and plead ריבי my cause מגוי nation: לא against an ungodly חסיד against an ungodly מאישׁ man. מרמה me from the deceitful ועולה and unjust תפלטני׃ O deliver}%
\verse{כי For אתה thou אלהי the God מעוזי of my strength: למה why זנחתני dost thou cast me off? למה why קדר I mourning אתהלך go בלחץ because of the oppression אויב׃ of the enemy?}%
\verse{שׁלח O send out אורך thy light ואמתך and thy truth: המה let them ינחוני lead יביאוני me; let them bring אל me unto הר hill, קדשׁך thy holy ואל and to משׁכנותיך׃ thy tabernacles.}%
\verse{ואבואה Then will I go אל unto מזבח the altar אלהים of God, אל unto אל God שׂמחת my exceeding גילי joy: ואודך will I praise בכנור yea, upon the harp אלהים thee, O God אלהי׃ my God.}%
\verse{מה Why תשׁתוחחי art thou cast down, נפשׁי O my soul? ומה and why תהמי art thou disquieted עלי within הוחילי me? hope לאלהים in God: כי for עוד I shall yet אודנו praise ישׁועת him, the health פני of my countenance, ואלהי׃ and my God.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 44
\verseWithHeading{Present Defeat and Past Deliverance}{למנצח To the chief Musician לבני for the sons קרח of Korah, משׂכיל׃ Maschil. אלהים O God, באזנינו with our ears, שׁמענו We have heard אבותינו our fathers ספרו have told לנו פעל us, work פעלת thou didst בימיהם in their days, בימי in the times קדם׃ of old.}%
\verse{אתה thou ידך with thy hand, גוים the heathen הורשׁת didst drive out ותטעם and plantedst תרע them; thou didst afflict לאמים the people, ותשׁלחם׃ and cast them out.}%
\verse{כי For לא they got not בחרבם by their own sword, ירשׁו in possession ארץ the land וזרועם did their own arm לא neither הושׁיעה save למו כי them: but ימינך thy right hand, וזרועך and thine arm, ואור and the light פניך of thy countenance, כי because רציתם׃ thou hadst a favor}%
\verse{אתה Thou הוא מלכי art my King, אלהים O God: צוה command ישׁועות deliverances יעקב׃ for Jacob.}%
\verse{בך צרינו our enemies: ננגח Through thee will we push down בשׁמך through thy name נבוס will we tread them under קמינו׃ that rise up against}%
\verse{כי For לא I will not בקשׁתי in my bow, אבטח trust וחרבי shall my sword לא neither תושׁיעני׃ save}%
\verse{כי But הושׁעתנו thou hast saved מצרינו us from our enemies, ומשׂנאינו that hated הבישׁות׃ and hast put them to shame}%
\verse{באלהים In God הללנו we boast כל all היום the day ושׁמך thy name לעולם forever. נודה long, and praise סלה׃ Selah.}%
\verse{אף But זנחת thou hast cast off, ותכלימנו and put us to shame; ולא and goest not forth תצא and goest not forth בצבאותינו׃ with our armies.}%
\verse{תשׁיבנו Thou makest us to turn אחור back מני from צר the enemy: ומשׂנאינו and they which hate שׁסו׃ us spoil}%
\verse{תתננו Thou hast given כצאן us like sheep מאכל for meat; ובגוים us among the heathen. זריתנו׃ and hast scattered}%
\verse{תמכר Thou sellest עמך thy people בלא for naught, הון for naught, ולא and dost not רבית increase במחיריהם׃ by their price.}%
\verse{תשׂימנו Thou makest חרפה us a reproach לשׁכנינו to our neighbors, לעג a scorn וקלס and a derision לסביבותינו׃ to them that are round about}%
\verse{תשׂימנו Thou makest משׁל us a byword בגוים among the heathen, מנוד a shaking ראשׁ of the head בל among the people. אמים׃}%
\verse{כל continually היום continually כלמתי My confusion נגדי before ובשׁת me, and the shame פני of my face כסתני׃ hath covered}%
\verse{מקול מחרף of him that reproacheth ומגדף and blasphemeth; מפני by reason אויב of the enemy ומתנקם׃ and avenger.}%
\verse{כל All זאת this באתנו is come upon ולא us; yet have we not שׁכחנוך forgotten ולא thee, neither שׁקרנו have we dealt falsely בבריתך׃ in thy covenant.}%
\verse{לא is not נסוג turned אחור back, לבנו Our heart ותט declined אשׁרינו neither have our steps מני from ארחך׃ thy way;}%
\verse{כי Though דכיתנו thou hast sore broken במקום us in the place תנים of dragons, ותכס and covered עלינו and covered בצלמות׃ us with the shadow of death.}%
\verse{אם If שׁכחנו we have forgotten שׁם the name אלהינו of our God, ונפרשׂ or stretched out כפינו our hands לאל god; זר׃ to a strange}%
\verse{הלא Shall not אלהים God יחקר search זאת this כי out? for הוא he ידע knoweth תעלמות the secrets לב׃ of the heart.}%
\verse{כי Yea, עליך for הרגנו thy sake are we killed כל all היום the day נחשׁבנו long; we are counted כצאן as sheep טבחה׃ for the slaughter.}%
\verse{עורה Awake, למה why תישׁן sleepest אדני thou, O Lord? הקיצה arise, אל not תזנח cast לנצח׃ off forever.}%
\verse{למה Wherefore פניך thou thy face, תסתיר hidest תשׁכח forgettest ענינו our affliction ולחצנו׃ and our oppression?}%
\verse{כי For שׁחה is bowed down לעפר to the dust: נפשׁנו our soul דבקה cleaveth לארץ unto the earth. בטננו׃ our belly}%
\verse{קומה Arise עזרתה for our help, לנו ופדנו and redeem למען חסדך׃}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 45
\verseWithHeading{Celebration of a Royal Wedding}{למנצח To the chief Musician על upon שׁשׁנים Shoshannim, לבני for the sons קרח of Korah, משׂכיל Maschil, שׁיר A Song ידידת׃ of loves. רחשׁ is inditing לבי My heart דבר matter: טוב a good אני I מעשׂי of the things which I have made אמר speak למלך touching the king: לשׁוני my tongue עט the pen סופר writer. מהיר׃ of a ready}%
\verse{יפיפית Thou art fairer מבני than the children אדם of men: הוצק is poured חן grace בשׂפתותיך into thy lips: על therefore כן therefore ברכך hath blessed אלהים God לעולם׃ thee forever.}%
\verse{חגור Gird חרבך thy sword על upon ירך thigh, גבור O mighty, הודך with thy glory והדרך׃ and thy majesty.}%
\verse{והדרך And in thy majesty צלח prosperously רכב ride על because דבר because אמת of truth וענוה and meekness צדק righteousness; ותורך shall teach נוראות thee terrible ימינך׃ and thy right hand}%
\verse{חציך Thine arrows שׁנונים sharp עמים the people תחתיך under יפלו fall בלב in the heart אויבי enemies; המלך׃ of the king's}%
\verse{כסאך Thy throne, אלהים O God, עולם forever ועד and ever: שׁבט the scepter מישׁר a right שׁבט scepter. מלכותך׃ of thy kingdom}%
\verse{אהבת Thou lovest צדק righteousness, ותשׂנא and hatest רשׁע wickedness: על therefore כן therefore משׁחך hath anointed אלהים God, אלהיך thy God, שׁמן thee with the oil שׂשׂון of gladness מחבריך׃ above thy fellows.}%
\verse{מר of myrrh, ואהלות and aloes, קציעות cassia, כל All בגדתיך thy garments מן out of היכלי palaces, שׁן the ivory מני שׂמחוך׃ they have made thee glad.}%
\verse{בנות daughters מלכים Kings' ביקרותיך among thy honorable נצבה did stand שׁגל the queen לימינך women: upon thy right hand בכתם in gold אופיר׃ of Ophir.}%
\verse{שׁמעי Hearken, בת O daughter, וראי and consider, והטי and incline אזנך thine ear; ושׁכחי forget עמך also thine own people, ובית house; אביך׃ and thy father's}%
\verse{ויתאו greatly desire המלך So shall the king יפיך thy beauty: כי for הוא he אדניך thy Lord; והשׁתחוי׃ and worship}%
\verse{ובת And the daughter צר of Tyre במנחה with a gift; פניך thy favor. יחלו shall entreat עשׁירי the rich עם׃ among the people}%
\verse{כל all כבודה glorious בת daughter מלך The king's פנימה within: ממשׁבצות of wrought זהב gold. לבושׁה׃ her clothing}%
\verse{לרקמות in raiment of needlework: תובל She shall be brought למלך unto the king בתולות the virgins אחריה that follow רעותיה her companions מובאות׃ her shall be brought}%
\verse{תובלנה shall they be brought: בשׂמחת With gladness וגיל and rejoicing תבאינה they shall enter בהיכל palace. מלך׃ into the king's}%
\verse{תחת Instead אבתיך of thy fathers יהיו shall be בניך thy children, תשׁיתמו whom thou mayest make לשׂרים princes בכל in all הארץ׃ the earth.}%
\verse{אזכירה to be remembered שׁמך I will make thy name בכל in all דר generations: ודר generations: על therefore כן therefore עמים shall the people יהודך praise לעלם thee forever ועד׃ and ever.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 46
\verseWithHeading{God Provides for and Protects His People}{למנצח To the chief Musician לבני for the sons קרח of Korah, על upon עלמות Alamoth. שׁיר׃ A Song אלהים God לנו מחסה our refuge ועז and strength, עזרה help בצרות in trouble. נמצא present מאד׃ a very}%
\verse{על כןא will not נירא we fear, בהמיר be removed, ארץ though the earth ובמוט be carried הרים and though the mountains בלב into the midst ימים׃ of the sea;}%
\verse{יהמו thereof roar יחמרו be troubled, מימיו the waters ירעשׁו shake הרים the mountains בגאותו with the swelling סלה׃ thereof. Selah.}%
\verse{נהר a river, פלגיו the streams ישׂמחו whereof shall make glad עיר the city אלהים of God, קדשׁ the holy משׁכני of the tabernacles עליון׃ of the most High.}%
\verse{אלהים God בקרבה in the midst בל of her; she shall not תמוט be moved: יעזרה shall help אלהים God לפנות her, right בקר׃ early.}%
\verse{המו raged, גוים The heathen מטו were moved: ממלכות the kingdoms נתן he uttered בקולו his voice, תמוג melted. ארץ׃ the earth}%
\verse{יהוה The LORD צבאות of hosts עמנו with משׂגב our refuge. לנו אלהי us; the God יעקב of Jacob סלה׃ Selah.}%
\verse{לכו Come, חזו behold מפעלות the works יהוה of the LORD, אשׁר what שׂם he hath made שׁמות desolations בארץ׃ in the earth.}%
\verse{משׁבית to cease מלחמות He maketh wars עד unto קצה the end הארץ of the earth; קשׁת the bow, ישׁבר he breaketh וקצץ and cutteth חנית the spear עגלות the chariot ישׂרף in sunder; he burneth באשׁ׃ in the fire.}%
\verse{הרפו Be still, ודעו and know כי that אנכי I אלהים God: ארום I will be exalted בגוים among the heathen, ארום I will be exalted בארץ׃ in the earth.}%
\verse{יהוה The LORD צבאות of hosts עמנו with משׂגב our refuge. לנו אלהי us; the God יעקב of Jacob סלה׃ Selah.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 47
\verseWithHeading{God Is King over All the Earth}{למנצח To the chief Musician, לבני for the sons קרח of Korah. מזמור׃ A Psalm כל all העמים ye people; תקעו O clap כף your hands, הריעו shout לאלהים unto God בקול with the voice רנה׃ of triumph.}%
\verse{כי For יהוה the LORD עליון most high נורא terrible; מלך King גדול a great על over כל all הארץ׃ the earth.}%
\verse{ידבר He shall subdue עמים the people תחתינו under ולאמים us, and the nations תחת under רגלינו׃ our feet.}%
\verse{יבחר He shall choose לנו את נחלתנו our inheritance את גאון for us, the excellency יעקב of Jacob אשׁר whom אהב he loved. סלה׃ Selah.}%
\verse{עלה is gone up אלהים God בתרועה with a shout, יהוה the LORD בקול with the sound שׁופר׃ of a trumpet.}%
\verse{זמרו Sing praises אלהים to God, זמרו sing praises: זמרו sing praises למלכנו unto our King, זמרו׃ sing praises.}%
\verse{כי For מלך the King כל of all הארץ the earth: אלהים God זמרו sing ye praises משׂכיל׃ with understanding.}%
\verse{מלך reigneth אלהים God על over גוים the heathen: אלהים God ישׁב sitteth על upon כסא the throne קדשׁו׃ of his holiness.}%
\verse{נדיבי The princes עמים of the people נאספו are gathered together, עם the people אלהי of the God אברהם of Abraham: כי for לאלהים unto God: מגני the shields ארץ of the earth מאד he is greatly נעלה׃ exalted.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 48
\verseWithHeading{The Greatness of God in Zion}{שׁיר A Song מזמור Psalm לבני for the sons קרח׃ of Korah. גדול Great יהוה the LORD, ומהלל to be praised מאד and greatly בעיר in the city אלהינו of our God, הר the mountain קדשׁו׃ of his holiness.}%
\verse{יפה Beautiful נוף for situation, משׂושׂ the joy כל of the whole הארץ earth, הר mount ציון Zion, ירכתי the sides צפון of the north, קרית the city מלך King. רב׃ of the great}%
\verse{אלהים God בארמנותיה in her palaces נודע is known למשׂגב׃ for a refuge.}%
\verse{כי For, הנה lo, המלכים the kings נועדו were assembled, עברו they passed by יחדו׃ together.}%
\verse{המה They ראו saw כן so תמהו they marveled; נבהלו they were troubled, נחפזו׃ hasted away.}%
\verse{רעדה Fear אחזתם took hold upon שׁם them there, חיל pain, כיולדה׃ as of a woman in travail.}%
\verse{ברוח wind. קדים with an east תשׁבר Thou breakest אניות the ships תרשׁישׁ׃ of Tarshish}%
\verse{כאשׁר As שׁמענו we have heard, כן so ראינו have we seen בעיר in the city יהוה צבאות of hosts, בעיר in the city אלהינו of our God: אלהים God יכוננה will establish עד it forever. עולם it forever. סלה׃ Selah.}%
\verse{דמינו We have thought אלהים O God, חסדך of thy lovingkindness, בקרב in the midst היכלך׃ of thy temple.}%
\verse{כשׁמך According to thy name, אלהים O God, כן so תהלתך thy praise על unto קצוי the ends ארץ of the earth: צדק of righteousness. מלאה is full ימינך׃ thy right hand}%
\verse{ישׂמח rejoice, הר Let mount ציון Zion תגלנה be glad, בנות let the daughters יהודה of Judah למען because of משׁפטיך׃ thy judgments.}%
\verse{סבו Walk about ציון Zion, והקיפוה and go round about ספרו her: tell מגדליה׃ the towers}%
\verse{שׁיתו לבכםחילה her bulwarks, פסגו consider ארמנותיה her palaces; למען that תספרו ye may tell לדור to the generation אחרון׃ following.}%
\verse{כי For זה this אלהים God אלהינו our God עולם forever ועד and ever: הוא he ינהגנו will be our guide על unto מות׃}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 49
\verseWithHeading{Wealth and the Fate of the Wicked}{למנצח To the chief Musician, לבני for the sons קרח of Korah. מזמור׃ A Psalm שׁמעו Hear זאת this, כל all העמים people; האזינו give ear, כל all ישׁבי inhabitants חלד׃ of the world:}%
\verse{גם Both בני אדם low גם and בני אישׁ high, יחד together. עשׁיר rich ואביון׃ and poor,}%
\verse{פי My mouth ידבר shall speak חכמות of wisdom; והגות and the meditation לבי of my heart תבונות׃ of understanding.}%
\verse{אטה I will incline למשׁל to a parable: אזני mine ear אפתח I will open בכנור upon the harp. חידתי׃ my dark saying}%
\verse{למה Wherefore אירא should I fear בימי in the days רע of evil, עון the iniquity עקבי יסובני׃ shall compass me about?}%
\verse{הבטחים They that trust על in חילם their wealth, וברב in the multitude עשׁרם of their riches; יתהללו׃ and boast themselves}%
\verse{אח his brother, לא None פדה can by any means יפדה redeem אישׁ לא nor יתן give לאלהים to God כפרו׃ a ransom}%
\verse{ויקר precious, פדיון (For the redemption נפשׁם of their soul וחדל and it ceaseth לעולם׃ forever:)}%
\verse{ויחי live עוד That he should still לנצח forever, לא not יראה see השׁחת׃ corruption.}%
\verse{כי For יראה he seeth חכמים wise men ימותו die, יחד likewise כסיל the fool ובער and the brutish person יאבדו perish, ועזבו and leave לאחרים to others. חילם׃ their wealth}%
\verse{קרבם Their inward thought בתימו their houses לעולם forever, משׁכנתם their dwelling places לדר to all generations; ודר to all generations; קראו בשׁמותם after their own names. עלי they call אדמות׃ lands}%
\verse{ואדם Nevertheless man ביקר in honor בל not: ילין abideth נמשׁל he is like כבהמות the beasts נדמו׃ perish.}%
\verse{זה This דרכם their way כסל their folly: למו ואחריהם yet their posterity בפיהם their sayings. ירצו approve סלה׃ Selah.}%
\verse{כצאן Like sheep לשׁאול in the grave; שׁתו they are laid מות death ירעם shall feed וירדו shall have dominion בם ישׁרים on them; and the upright לבקר over them in the morning; וצירם לבלות shall consume שׁאול in the grave מזבל׃ from their dwelling.}%
\verse{אך But אלהים God יפדה will redeem נפשׁי my soul מיד from the power שׁאול of the grave: כי for יקחני he shall receive סלה׃ me. Selah.}%
\verse{אל Be not תירא thou afraid כי when יעשׁר is made rich, אישׁ one כי when ירבה is increased; כבוד the glory ביתו׃ of his house}%
\verse{כי For when לא nothing במותו he dieth יקח he shall carry הכל nothing לא shall not ירד descend אחריו after כבודו׃ away: his glory}%
\verse{כי Though נפשׁו his soul: בחייו while he lived יברך he blessed ויודך and will praise כי thee, when תיטיב׃ thou doest well}%
\verse{תבוא He shall go עד to דור the generation אבותיו of his fathers; עד they shall never נצח לאראו see אור׃ light.}%
\verse{אדם Man ביקר in honor, ולא not, יבין and understandeth נמשׁל is like כבהמות the beasts נדמו׃ perish.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 50
\verseWithHeading{An Oracle Concerning Sacrifices}{מזמור A Psalm לאסף of Asaph. אל The mighty אלהים God, יהוה the LORD, דבר hath spoken, ויקרא and called ארץ the earth ממזרח from the rising שׁמשׁ of the sun עד unto מבאו׃ the going down}%
\verse{מציון מכלל the perfection יפי of beauty, אלהים God הופיע׃ hath shined.}%
\verse{יבא shall come, אלהינו Our God ואל and shall not יחרשׁ keep silence: אשׁ a fire לפניו before תאכל shall devour וסביביו round about נשׂערה tempestuous מאד׃ him, and it shall be very}%
\verse{יקרא He shall call אל to השׁמים the heavens מעל from above, ואל and to הארץ the earth, לדין that he may judge עמו׃ his people.}%
\verse{אספו לי חסידירתי unto me; those that have made בריתי a covenant עלי with me by זבח׃ sacrifice.}%
\verse{ויגידו shall declare שׁמים And the heavens צדקו his righteousness: כי for אלהים God שׁפט judge הוא himself. סלה׃ Selah.}%
\verse{שׁמעה Hear, עמי O my people, ואדברה and I will speak; ישׂראל O Israel, ואעידה and I will testify בך אלהים God, אלהיך thy God. אנכי׃ against thee: I}%
\verse{לא I will not על thee for זבחיך thy sacrifices אוכיחך reprove ועולתיך or thy burnt offerings, לנגדי before תמיד׃ continually}%
\verse{לא no אקח I will take מביתך out of thy house, פר bullock ממכלאתיך out of thy folds. עתודים׃ he goats}%
\verse{כי For לי כל every חיתו beast יער of the forest בהמות mine, the cattle בהררי hills. אלף׃ upon a thousand}%
\verse{ידעתי I know כל all עוף the fowls הרים of the mountains: וזיז and the wild beasts שׂדי of the field עמדי׃ mine.}%
\verse{אם If ארעב I were hungry, לא I would not אמר tell לך כי thee: for לי תבל the world ומלאה׃ mine, and the fullness}%
\verse{האוכל Will I eat בשׂר the flesh אבירים of bulls, ודם the blood עתודים of goats? אשׁתה׃ or drink}%
\verse{זבח Offer לאלהים unto God תודה thanksgiving; ושׁלם and pay לעליון unto the most High: נדריך׃ thy vows}%
\verse{וקראני And call upon ביום me in the day צרה of trouble: אחלצך I will deliver ותכבדני׃ thee, and thou shalt glorify}%
\verse{ולרשׁע But unto the wicked אמר saith, אלהים God מה What לך לספר hast thou to do to declare חקי my statutes, ותשׂא or thou shouldest take בריתי my covenant עלי in פיך׃ thy mouth?}%
\verse{ואתה Seeing thou שׂנאת hatest מוסר instruction, ותשׁלך and castest דברי my words אחריך׃ behind}%
\verse{אם When ראית thou sawest גנב a thief, ותרץ then thou consentedst עמו with ועם with מנאפים adulterers. חלקך׃ him, and hast been partaker}%
\verse{פיך thy mouth שׁלחת Thou givest ברעה to evil, ולשׁונך and thy tongue תצמיד frameth מרמה׃ deceit.}%
\verse{תשׁב Thou sittest באחיך against thy brother; תדבר speakest בבן son. אמך thine own mother's תתן thou slanderest דפי׃ thou slanderest}%
\verse{אלה These עשׂית hast thou done, והחרשׁתי and I kept silence; דמית thou thoughtest היות that I was אהיה altogether כמוך as thyself: אוכיחך I will reprove ואערכה thee, and set in order לעיניך׃ before thine eyes.}%
\verse{בינו consider נא Now זאת this, שׁכחי ye that forget אלוה God, פן lest אטרף I tear in pieces, ואין and none מציל׃ to deliver.}%
\verse{זבח Whoso offereth תודה praise יכבדנני glorifieth ושׂם me: and to him that ordereth דרך conversation אראנו will I show בישׁע the salvation אלהים׃ of God.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 51
\verseWithHeading{A Prayer of Repentance and Plea for Mercy}{למנצח To the chief Musician, מזמור A Psalm לדוד׃ of David, בבוא came אליו unto נתן when Nathan הנביא the prophet כאשׁר him, after בא he had gone in אל to בת־שׁבע׃ Bathsheba. חנני Have mercy אלהים upon me, O God, כחסדך according to thy lovingkindness: כרב according unto the multitude רחמיך of thy tender mercies מחה blot out פשׁעי׃ my transgressions.}%
\verse{הרבה me throughly כבסני Wash מעוני from mine iniquity, ומחטאתי me from my sin. טהרני׃ and cleanse}%
\verse{כי For פשׁעי my transgressions: אני I אדע acknowledge וחטאתי and my sin נגדי before תמיד׃ ever}%
\verse{לך לבדך Against thee, thee only, חטאתי have I sinned, והרע evil בעיניך in thy sight: עשׂיתי and done למען that תצדק thou mightest be justified בדברך when thou speakest, תזכה be clear בשׁפטך׃ when thou judgest.}%
\verse{הן Behold, בעוון in iniquity; חוללתי I was shapen ובחטא and in sin יחמתני conceive אמי׃ did my mother}%
\verse{הן Behold, אמת truth חפצת thou desirest בטחות in the inward parts: ובסתם and in the hidden חכמה wisdom. תודיעני׃ thou shalt make me to know}%
\verse{תחטאני Purge באזוב me with hyssop, ואטהר and I shall be clean: תכבסני wash ומשׁלג than snow. אלבין׃ me, and I shall be whiter}%
\verse{תשׁמיעני Make me to hear שׂשׂון joy ושׂמחה and gladness; תגלנה may rejoice. עצמות the bones דכית׃ thou hast broken}%
\verse{הסתר Hide פניך thy face מחטאי from my sins, וכל all עונתי mine iniquities. מחה׃ and blot out}%
\verse{לב heart, טהור in me a clean ברא Create לי אלהים O God; ורוח spirit נכון a right חדשׁ and renew בקרבי׃ within}%
\verse{אל me not תשׁליכני Cast מלפניך away from thy presence; ורוח spirit קדשׁך thy holy אל not תקח and take ממני׃ away from thy presence;}%
\verse{השׁיבה Restore לי שׂשׂון unto me the joy ישׁעך of thy salvation; ורוח spirit. נדיבה תסמכני׃ and uphold}%
\verse{אלמדה will I teach פשׁעים transgressors דרכיך thy ways; וחטאים and sinners אליך unto ישׁובו׃ shall be converted}%
\verse{הצילני Deliver מדמים me from bloodguiltiness, אלהים O God, אלהי thou God תשׁועתי of my salvation: תרנן shall sing aloud לשׁוני my tongue צדקתך׃ of thy righteousness.}%
\verse{אדני O Lord, שׂפתי thou my lips; תפתח open ופי and my mouth יגיד shall show forth תהלתך׃ thy praise.}%
\verse{כי For לא not תחפץ thou desirest זבח sacrifice; ואתנה else would I give עולה in burnt offering. לא not תרצה׃ thou delightest}%
\verse{זבחי The sacrifices אלהים of God רוח spirit: נשׁברה a broken לב heart, נשׁבר a broken ונדכה and a contrite אלהים O God, לא thou wilt not תבזה׃ despise.}%
\verse{היטיבה Do good ברצונך in thy good pleasure את ציון unto Zion: תבנה build חומות thou the walls ירושׁלם׃ of Jerusalem.}%
\verse{אז Then תחפץ shalt thou be pleased זבחי with the sacrifices צדק of righteousness, עולה with burnt offering וכליל and whole burnt offering: אז then יעלו shall they offer על upon מזבחך thine altar. פרים׃ bullocks}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 52
\verseWithHeading{God’s Judgment on the Wicked and Love for the Faithful}{למנצח To the chief Musician, משׂכיל Maschil, לדוד׃ of David, בבוא came דואג when Doeg האדמי the Edomite ויגד and told לשׁאול Saul, ויאמר and said לו בא is come דוד unto him, David אל to בית the house אחימלך׃ of Ahimelech. מה Why תתהלל boastest thou thyself ברעה in mischief, הגבור O mighty man? חסד the goodness אל of God כל continually. היום׃ continually.}%
\verse{הוות mischiefs; תחשׁב deviseth לשׁונך Thy tongue כתער razor, מלטשׁ like a sharp עשׂה working רמיה׃ deceitfully.}%
\verse{אהבת Thou lovest רע evil מטוב more than good; שׁקר lying מדבר rather than to speak צדק righteousness. סלה׃ Selah.}%
\verse{אהבת Thou lovest כל all דברי words, בלע devouring לשׁון tongue. מרמה׃ O deceitful}%
\verse{גם shall likewise אל God יתצך destroy לנצח thee forever, יחתך he shall take thee away, ויסחך and pluck thee out מאהל of dwelling place, ושׁרשׁך and root מארץ thee out of the land חיים of the living. סלה׃ Selah.}%
\verse{ויראו also shall see, צדיקים The righteous וייראו and fear, ועליו at ישׂחקו׃ and shall laugh}%
\verse{הנה Lo, הגבר the man לא not ישׂים made אלהים God מעוזו his strength; ויבטח but trusted ברב in the abundance עשׁרו of his riches, יעז strengthened בהותו׃ himself in his wickedness.}%
\verse{ואני But I כזית olive tree רענן like a green בבית in the house אלהים of God: בטחתי I trust בחסד in the mercy אלהים of God עולם forever ועד׃ and ever.}%
\verse{אודך I will praise לעולם thee forever, כי because עשׂית thou hast done ואקוה and I will wait on שׁמך thy name; כי for טוב good נגד before חסידיך׃ thy saints.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 53
\verseWithHeading{The Folly of the Godless and Salvation for Israel}{למנצח To the chief Musician על upon מחלת Mahalath, משׂכיל Maschil, לדוד׃ of David. אמר hath said נבל The fool בלבו in his heart, אין no אלהים God. השׁחיתו Corrupt והתעיבו are they, and have done abominable עול iniquity: אין none עשׂה that doeth טוב׃ good.}%
\verse{אלהים God משׁמים from heaven השׁקיף looked down על upon בני the children אדם of men, לראות to see הישׁ if there were משׂכיל that did understand, דרשׁ that did seek את אלהים׃ God.}%
\verse{כלו Every one סג of them is gone back: יחדו they are altogether נאלחו become filthy; אין none עשׂה that doeth טוב good, אין no, גם not אחד׃ one.}%
\verse{הלא no ידעו knowledge? פעלי Have the workers און of iniquity אכלי who eat up עמי my people אכלו they eat לחם bread: אלהים God. לא they have not קראו׃ called upon}%
\verse{שׁם There פחדו were they in great fear, פחד were they in great fear, לא no היה was: פחד fear כי for אלהים God פזר hath scattered עצמות the bones חנך of him that encampeth הבשׁתה thee: thou hast put to shame, כי because אלהים God מאסם׃ hath despised}%
\verse{מי יתןציון ישׁעות the salvation ישׂראל of Israel בשׁוב bringeth back אלהים When God שׁבות the captivity עמו of his people, יגל shall rejoice, יעקב Jacob ישׂמח shall be glad. ישׂראל׃ Israel}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 54
\verseWithHeading{Answered Prayer for Deliverance from Adversaries}{למנצח To the chief Musician בנגינת on Neginoth, משׂכיל Maschil, לדוד׃ of David, בבוא came הזיפים when the Ziphims ויאמרו and said לשׁאול to Saul, הלא Doth not דוד David מסתתר hide himself עמנו׃ with אלהים me, O God, בשׁמך by thy name, הושׁיעני us? Save ובגבורתך me by thy strength. תדינני׃ and judge}%
\verse{אלהים O God; שׁמע Hear תפלתי my prayer, האזינה give ear לאמרי to the words פי׃ of my mouth.}%
\verse{כי For זרים strangers קמו are risen up עלי against ועריצים me, and oppressors בקשׁו seek after נפשׁי my soul: לא they have not שׂמו set אלהים God לנגדם before סלה׃ them. Selah.}%
\verse{הנה Behold, אלהים God עזר mine helper: לי אדני the Lord בסמכי with them that uphold נפשׁי׃ my soul.}%
\verse{ישׁוב He shall reward הרע evil לשׁררי unto mine enemies: באמתך in thy truth. הצמיתם׃ cut them off}%
\verse{בנדבה I will freely אזבחה sacrifice לך אודה unto thee: I will praise שׁמך thy name, יהוה O LORD; כי for טוב׃ good.}%
\verse{כי For מכל me out of all צרה trouble: הצילני he hath delivered ובאיבי upon mine enemies. ראתה hath seen עיני׃ and mine eye}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 55
\verseWithHeading{Betrayal of a Friend and Trust in God}{למנצח To the chief Musician בנגינת on Neginoth, משׂכיל Maschil, לדוד׃ of David. האזינה Give ear אלהים O God; תפלתי to my prayer, ואל and hide not thyself תתעלם and hide not thyself מתחנתי׃ from my supplication.}%
\verse{הקשׁיבה Attend לי וענני unto me, and hear אריד me: I mourn בשׂיחי in my complaint, ואהימה׃ and make a noise;}%
\verse{מקול אויב of the enemy, מפני because עקת of the oppression רשׁע of the wicked: כי for ימיטו they cast עלי upon און iniquity ובאף me, and in wrath ישׂטמוני׃ they hate}%
\verse{לבי My heart יחיל is sore pained בקרבי within ואימות me: and the terrors מות of death נפלו are fallen עלי׃ upon}%
\verse{יראה Fearfulness ורעד and trembling יבא are come בי ותכסני hath overwhelmed פלצות׃ upon me, and horror}%
\verse{ואמר And I said, מי Oh that יתן I had לי אבר wings כיונה like a dove! אעופה would I fly away, ואשׁכנה׃ and be at rest.}%
\verse{הנה Lo, ארחיק far off, נדד would I wander אלין remain במדבר in the wilderness. סלה׃ Selah.}%
\verse{אחישׁה I would hasten מפלט my escape לי מרוח from the windy סעה storm מסער׃ tempest.}%
\verse{בלע Destroy, אדני O Lord, פלג divide לשׁונם their tongues: כי for ראיתי I have seen חמס violence וריב and strife בעיר׃ in the city.}%
\verse{יומם Day ולילה and night יסובבה they go about על it upon חומתיה the walls ואון thereof: mischief ועמל also and sorrow בקרבה׃ in the midst}%
\verse{הוות Wickedness בקרבה in the midst ולא not ימישׁ depart מרחבה her streets. תך thereof: deceit ומרמה׃ and guile}%
\verse{כי For לא not אויב an enemy יחרפני reproached ואשׂא me; then I could have borne לא neither משׂנאי he that hated עלי against הגדיל me did magnify ואסתר me; then I would have hid myself ממנו׃ from}%
\verse{ואתה But thou, אנושׁ a man כערכי mine equal, אלופי my guide, ומידעי׃ and mine acquaintance.}%
\verse{אשׁר יחדו together, נמתיק We took sweet סוד counsel בבית unto the house אלהים of God נהלך walked ברגשׁ׃ in company.}%
\verse{ישׁימות עלימו upon ירדו them, let them go down שׁאול into hell: חיים quick כי for רעות wickedness במגורם in their dwellings, בקרבם׃ among}%
\verse{אני As for me, I אל upon אלהים God; אקרא will call ויהוה and the LORD יושׁיעני׃ shall save}%
\verse{ערב Evening, ובקר and morning, וצהרים and at noon, אשׂיחה will I pray, ואהמה and cry aloud: וישׁמע and he shall hear קולי׃ my voice.}%
\verse{פדה He hath delivered בשׁלום in peace נפשׁי my soul מקרב from the battle לי כי against me: for ברבים many היו there were עמדי׃ with}%
\verse{ישׁמע shall hear, אל God ויענם and afflict וישׁב them, even he that abideth קדם of old. סלה Selah. אשׁר Because אין they have no חליפות changes, למו ולא not יראו therefore they fear אלהים׃ God.}%
\verse{שׁלח He hath put forth ידיו his hands בשׁלמיו against such as be at peace חלל with him: he hath broken בריתו׃ his covenant.}%
\verse{חלקו were smoother מחמאת than butter, פיו of his mouth וקרב but war לבו in his heart: רכו were softer דבריו his words משׁמן than oil, והמה yet they פתחות׃ drawn swords.}%
\verse{השׁלך Cast על upon יהוה the LORD, יהבך thy burden והוא and he יכלכלך shall sustain לא thee: he shall never יתן suffer לעולם thee: he shall never מוט to be moved. לצדיק׃ the righteous}%
\verse{ואתה But thou, אלהים O God, תורדם shalt bring them down לבאר into the pit שׁחת of destruction: אנשׁי דמים bloody ומרמה and deceitful לא shall not יחצו live out half ימיהם their days; ואני but I אבטח׃ will trust}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 56
\verseWithHeading{Prayer for Deliverance and Confidence in God}{למנצח To the chief Musician על upon יונת אלםחקים לדוד of David, מכתם Michtam באחז took אתו פלשׁתים when the Philistines בגת׃ him in Gath. חנני Be merciful אלהים unto me, O God: כי for שׁאפני would swallow me up; אנושׁ man כל daily היום daily לחם he fighting ילחצני׃ oppresseth}%
\verse{שׁאפו swallow up: שׁוררי Mine enemies כל would daily היום would daily כי for רבים many לחמים that fight לי מרום׃ against me, O thou most High.}%
\verse{יום What time אירא I am afraid, אני I אליך in אבטח׃ will trust}%
\verse{באלהים In God אהלל I will praise דברו his word, באלהים in God בטחתי I have put my trust; לא I will not אירא fear מה what יעשׂה can do בשׂר׃ flesh}%
\verse{כל Every היום day דברי my words: יעצבו they wrest עלי against כל all מחשׁבתם their thoughts לרע׃ me for evil.}%
\verse{יגורו They gather themselves together, יצפינו they hide המה themselves, they עקבי my steps, ישׁמרו mark כאשׁר when קוו they wait for נפשׁי׃ my soul.}%
\verse{על by און iniquity? פלט למו באף in anger עמים the people, הורד cast down אלהים׃ O God.}%
\verse{נדי my wanderings: ספרתה Thou tellest אתה thou שׂימה put דמעתי my tears בנאדך into thy bottle: הלא not בספרתך׃ in thy book?}%
\verse{אז then ישׁובו turn אויבי shall mine enemies אחור back: ביום When אקרא I cry זה this ידעתי I know; כי for אלהים׃ God}%
\verse{באלהים In God אהלל will I praise דבר word: ביהוה in the LORD אהלל will I praise דבר׃ word.}%
\verse{באלהים In God בטחתי have I put my trust: לא I will not אירא be afraid מה what יעשׂה can do אדם׃ man}%
\verse{עלי upon אלהים me, O God: נדריך Thy vows אשׁלם I will render תודת׃ praises}%
\verse{כי For הצלת thou hast delivered נפשׁי my soul ממות from death: הלא not רגלי my feet מדחי from falling, להתהלך that I may walk לפני before אלהים God באור in the light החיים׃ of the living.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 57
\verseWithHeading{Prayer for Rescue from Enemies}{למנצח To the chief Musician, אל תשׁחת Al-taschith, לדוד of David, מכתם Michtam בברחו when he fled מפני from שׁאול Saul במערה׃ in the cave. חנני Be merciful אלהים unto me, O God, חנני be merciful כי unto me: for בך חסיה trusteth נפשׁי my soul ובצל in thee: yea, in the shadow כנפיך of thy wings אחסה will I make my refuge, עד until יעבר be overpast. הוות׃ calamities}%
\verse{אקרא I will cry לאלהים unto God עליון most high; לאל unto God גמר that performeth עלי׃ for}%
\verse{ישׁלח He shall send משׁמים from heaven, ויושׁיעני and save חרף me the reproach שׁאפי of him that would swallow me up. סלה Selah. ישׁלח shall send forth אלהים God חסדו his mercy ואמתו׃ and his truth.}%
\verse{נפשׁי My soul בתוך among לבאם lions: אשׁכבה I lie להטים them that are set on fire, בני the sons אדם of men, שׁניהם whose teeth חנית spears וחצים and arrows, ולשׁונם and their tongue חרב sword. חדה׃ a sharp}%
\verse{רומה Be thou exalted, על above השׁמים the heavens; אלהים O God, על above כל all הארץ the earth. כבודך׃ thy glory}%
\verse{רשׁת a net הכינו They have prepared לפעמי for my steps; כפף is bowed down: נפשׁי my soul כרו they have digged לפני before שׁיחה a pit נפלו whereof they are fallen בתוכה me, into the midst סלה׃ Selah.}%
\verse{נכון is fixed, לבי My heart אלהים O God, נכון is fixed: לבי my heart אשׁירה I will sing ואזמרה׃ and give praise.}%
\verse{עורה Awake up, כבודי my glory; עורה awake, הנבל psaltery וכנור and harp: אעירה I will awake שׁחר׃ early.}%
\verse{אודך I will praise בעמים among the people: אדני thee, O Lord, אזמרך I will sing בל אמים׃ unto thee among the nations.}%
\verse{כי For גדל great עד unto שׁמים the heavens, חסדך thy mercy ועד unto שׁחקים the clouds. אמתך׃ and thy truth}%
\verse{רומה Be thou exalted, על above שׁמים the heavens: אלהים O God, על above כל all הארץ the earth. כבודך׃ thy glory}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 58
\verseWithHeading{Judgment on the Wicked}{למנצח To the chief Musician, אל תשׁחת Al-taschith, לדוד of David. מכתם׃ Michtam האמנם Do ye indeed אלם O congregation? צדק righteousness, תדברון speak מישׁרים uprightly, תשׁפטו do ye judge בני O ye sons אדם׃ of men?}%
\verse{אף Yea, בלב in heart עולת wickedness; תפעלון ye work בארץ in the earth. חמס the violence ידיכם of your hands תפלסון׃ ye weigh}%
\verse{זרו are estranged רשׁעים The wicked מרחם from the womb: תעו they go astray מבטן as soon as they be born, דברי speaking כזב׃ lies.}%
\verse{חמת Their poison למו כדמות like חמת the poison נחשׁ of a serpent: כמו like פתן adder חרשׁ the deaf יאטם stoppeth אזנו׃ her ear;}%
\verse{אשׁר Which לא will not ישׁמע hearken לקול to the voice מלחשׁים of charmers, חובר charming חברים charming מחכם׃ never so wisely.}%
\verse{אלהים O God, הרס Break שׁנימו their teeth, בפימו in their mouth: מלתעות the great teeth כפירים of the young lions, נתץ break out יהוה׃ O LORD.}%
\verse{ימאסו Let them melt away כמו as מים waters יתהלכו run למו ידרך continually: he bendeth חצו his arrows, כמו let them be as יתמללו׃ cut in pieces.}%
\verse{כמו As שׁבלול a snail תמס melteth, יהלך let pass away: נפל the untimely birth אשׁת of a woman, בל they may not חזו see שׁמשׁ׃ the sun.}%
\verse{בטרם Before יבינו can feel סירתיכם your pots אטד the thorns, כמו both חי living, כמו and in חרון wrath. ישׂערנו׃ he shall take them away as with a whirlwind,}%
\verse{ישׂמח shall rejoice צדיק The righteous כי when חזה he seeth נקם the vengeance: פעמיו his feet ירחץ he shall wash בדם in the blood הרשׁע׃ of the wicked.}%
\verse{ויאמר shall say, אדם So that a man אך Verily פרי a reward לצדיק for the righteous: אך verily ישׁ he is אלהים a God שׁפטים that judgeth בארץ׃ in the earth.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 59
\verseWithHeading{A Prayer for Protection}{למנצח To the chief Musician, אל תשׁחת Al-taschith, לדוד of David; מכתם Michtam בשׁלח sent, שׁאול when Saul וישׁמרו and they watched את הבית the house להמיתו׃ to kill הצילני him. Deliver מאיבי me from mine enemies, אלהי O my God: ממתקוממי me from them that rise up against תשׂגבני׃ defend}%
\verse{הצילני Deliver מפעלי me from the workers און of iniquity, ומאנשׁי דמים me from bloody הושׁיעני׃ and save}%
\verse{כי For, הנה lo, ארבו they lie in wait לנפשׁי for my soul: יגורו are gathered עלי against עזים the mighty לא me; not פשׁעי my transgression, ולא nor חטאתי my sin, יהוה׃ O LORD.}%
\verse{בלי without עון fault: ירוצון They run ויכוננו and prepare themselves עורה awake לקראתי to help וראה׃ me, and behold.}%
\verse{ואתה Thou יהוה therefore, O LORD אלהים God צבאות of hosts, אלהי the God ישׂראל of Israel, הקיצה awake לפקד to visit כל all הגוים the heathen: אל be not תחן merciful כל to any בגדי transgressors. און wicked סלה׃ Selah.}%
\verse{ישׁובו They return לערב at evening: יהמו they make a noise ככלב like a dog, ויסובבו and go round about עיר׃ the city.}%
\verse{הנה Behold, יביעון they belch out בפיהם with their mouth: חרבות swords בשׂפתותיהם in their lips: כי for מי who, שׁמע׃ doth hear?}%
\verse{ואתה But thou, יהוה O LORD, תשׂחק shalt laugh למו תלעג in derision. לכל at them; thou shalt have all גוים׃ the heathen}%
\verse{עזו his strength אליך upon אשׁמרה will I wait כי thee: for אלהים God משׂגבי׃ my defense.}%
\verse{אלהי The God חסדו of my mercy יקדמני shall prevent אלהים me: God יראני shall let me see בשׁררי׃ upon mine enemies.}%
\verse{אל them not, תהרגם Slay פן lest ישׁכחו forget: עמי my people הניעמו scatter בחילך them by thy power; והורידמו and bring them down, מגננו our shield. אדני׃ O Lord}%
\verse{חטאת the sin פימו of their mouth דבר the words שׂפתימו of their lips וילכדו let them even be taken בגאונם in their pride: ומאלה and for cursing ומכחשׁ and lying יספרו׃ they speak.}%
\verse{כלה Consume בחמה in wrath, כלה consume ואינמו that they not וידעו and let them know כי that אלהים God משׁל ruleth ביעקב in Jacob לאפסי unto the ends הארץ of the earth. סלה׃ Selah.}%
\verse{וישׁובו let them return; לערב And at evening יהמו let them make a noise ככלב like a dog, ויסובבו and go round about עיר׃ the city.}%
\verse{המה Let them ינועון wander up and down לאכל for meat, אם if לא they be not ישׂבעו satisfied. וילינו׃ and grudge}%
\verse{ואני But I אשׁיר will sing עזך of thy power; וארנן yea, I will sing aloud 'ecבקר in the morning: חסדך of thy mercy כי for היית thou hast been משׂגב my defense לי ומנוס and refuge ביום in the day צר׃ of my trouble.}%
\verse{עזי thee, O my strength, אליך Unto אזמרה will I sing: כי for אלהים God משׂגבי my defense, אלהי the God חסדי׃ of my mercy.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 60
\verseWithHeading{A Lament After a Defeat and a Prayer for Restoration}{למנצח To the chief Musician על upon שׁושׁן עדות Shushan-eduth, מכתם Michtam לדוד of David, ללמד׃ to teach; בהצותו when he strove with את ארם נהרים Aram-naharaim ואת ארם צובה and with Aram-zobah, וישׁב returned, יואב when Joab ויך and smote את אדום of Edom בגיא in the valley מלח of salt שׁנים twelve עשׂר twelve אלף׃ thousand. אלהים O God, זנחתנו thou hast cast us off, פרצתנו thou hast scattered אנפת us, thou hast been displeased; תשׁובב O turn}%
\verse{הרעשׁתה to tremble; ארץ Thou hast made the earth פצמתה thou hast broken רפה it: heal שׁבריה the breaches כי thereof; for מטה׃ it shaketh.}%
\verse{הראיתה Thou hast showed עמך thy people קשׁה hard השׁקיתנו things: thou hast made us to drink יין the wine תרעלה׃ of astonishment.}%
\verse{נתתה Thou hast given ליראיך to them that fear נס a banner להתנוסס thee, that it may be displayed מפני because קשׁט of the truth. סלה׃ Selah.}%
\verse{למען That יחלצון may be delivered; ידידיך thy beloved הושׁיעה save ימינך thy right hand, ועננו׃}%
\verse{אלהים God דבר hath spoken בקדשׁו in his holiness; אעלזה I will rejoice, אחלקה I will divide שׁכם Shechem, ועמק the valley סכות of Succoth. אמדד׃ and mete out}%
\verse{לי גלעד Gilead ולי מנשׁה mine, and Manasseh ואפרים mine; Ephraim מעוז also the strength ראשׁי of mine head; יהודה Judah מחקקי׃ my lawgiver;}%
\verse{מואב Moab סיר my washpot; רחצי my washpot; על over אדום Edom אשׁליך will I cast out נעלי my shoe: עלי thou because פלשׁת Philistia, התרעעי׃ triumph}%
\verse{מי Who יבלני will bring עיר city? מצור me the strong מי who נחני will lead עד me into אדום׃ Edom?}%
\verse{הלא not אתה thou, אלהים O God, זנחתנו hadst cast us off? ולא didst not תצא go out אלהים and O God, בצבאותינו׃ with our armies?}%
\verse{הבה Give לנו עזרת us help מצר from trouble: ושׁוא for vain תשׁועת the help אדם׃ of man.}%
\verse{באלהים Through God נעשׂה we shall do חיל valiantly: והוא for he יבוס shall tread down צרינו׃ our enemies.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 61
\verseWithHeading{Confidence in God’s Protection}{למנצח To the chief Musician על upon נגינת Neginah, לדוד׃ of David. שׁמעה Hear אלהים O God; רנתי my cry, הקשׁיבה attend unto תפלתי׃ my prayer.}%
\verse{מקצה הארץ of the earth אליך unto אקרא will I cry בעטף is overwhelmed: לבי thee, when my heart בצור me to the rock ירום is higher ממני than תנחני׃ lead}%
\verse{כי For היית thou hast been מחסה a shelter לי מגדל tower עז for me, a strong מפני from אויב׃ the enemy.}%
\verse{אגורה I will abide באהלך in thy tabernacle עולמים forever: אחסה I will trust בסתר in the covert כנפיך of thy wings. סלה׃ Selah.}%
\verse{כי For אתה thou, אלהים O God, שׁמעת hast heard לנדרי my vows: נתת thou hast given ירשׁת the heritage יראי of those that fear שׁמך׃ thy name.}%
\verse{ימים life: על ימילך the king's תוסיף Thou wilt prolong שׁנותיו his years כמו as דר many generations. ודר׃ many generations.}%
\verse{ישׁב He shall abide עולם forever: לפני before אלהים God חסד mercy ואמת and truth, מן O prepare ינצרהו׃ may preserve}%
\verse{כן So אזמרה will I sing praise שׁמך unto thy name לעד forever, לשׁלמי perform נדרי my vows. יום יום׃}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 62
\verseWithHeading{Confidence in God’s Salvation}{למנצח To the chief Musician, על to ידותון Jeduthun, מזמור A Psalm לדוד׃ of David. אך Truly אל upon אלהים God: דומיה waiteth נפשׁי my soul ממנו from ישׁועתי׃ him my salvation.}%
\verse{אך only הוא He צורי my rock וישׁועתי and my salvation; משׂגבי my defense; לא I shall not אמוט moved. רבה׃ be greatly}%
\verse{עד אנההותתו will ye imagine mischief על against אישׁ a man? תרצחו ye shall be slain כלכם all כקיר wall נטוי of you: as a bowing גדר fence. הדחויה׃ a tottering}%
\verse{אך They only משׂאתו from his excellency: יעצו consult להדיח to cast down ירצו they delight כזב in lies: בפיו with their mouth, יברכו they bless ובקרבם inwardly. יקללו but they curse סלה׃ Selah.}%
\verse{אך thou only לאלהים upon God; דומי wait נפשׁי My soul, כי for ממנו from תקותי׃ my expectation}%
\verse{אך only הוא He צורי my rock וישׁועתי and my salvation: משׂגבי my defense; לא I shall not אמוט׃ be moved.}%
\verse{על In אלהים God ישׁעי my salvation וכבודי and my glory: צור the rock עזי of my strength, מחסי my refuge, באלהים׃ in God.}%
\verse{בטחו Trust בו בכל in him at all עת times; עם people, שׁפכו pour out לפניו before לבבכם your heart אלהים him: God מחסה a refuge לנו סלה׃ for us. Selah.}%
\verse{אך Surely הבל vanity, בני men of low degree אדם men of low degree כזב a lie: בני men of high degree אישׁ men of high degree במאזנים in the balance, לעלות to be laid המה they מהבל than vanity. יחד׃ altogether}%
\verse{אל not תבטחו Trust בעשׁק in oppression, ובגזל in robbery: אל and become not vain תהבלו and become not vain חיל riches כי if ינוב increase, אל not תשׁיתו set לב׃ your heart}%
\verse{אחת once; דבר hath spoken אלהים God שׁתים twice זו this; שׁמעתי have I heard כי that עז power לאלהים׃ unto God.}%
\verse{ולך אדני Also unto thee, O Lord, חסד mercy: כי for אתה thou תשׁלם renderest לאישׁ to every man כמעשׂהו׃ according to his work.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 63
\verseWithHeading{Longing for God}{מזמור A Psalm לדוד of David, בהיותו when he was במדבר in the wilderness יהודה׃ of Judah. אלהים O God, אלי my God; אתה thou אשׁחרך early will I seek צמאה thirsteth לך נפשׁי thee: my soul כמה longeth לך בשׂרי for thee, my flesh בארץ land, ציה for thee in a dry ועיף and thirsty בלי where no מים׃ water}%
\verse{כן so בקדשׁ thee in the sanctuary. חזיתיך I have seen לראות To see עזך thy power וכבודך׃ and thy glory,}%
\verse{כי Because טוב better חסדך thy lovingkindness מחיים than life, שׂפתי my lips ישׁבחונך׃ shall praise}%
\verse{כן Thus אברכך will I bless בחיי thee while I live: בשׁמך in thy name. אשׂא I will lift up כפי׃ my hands}%
\verse{כמו as חלב marrow ודשׁן and fatness; תשׂבע shall be satisfied נפשׁי My soul ושׂפתי lips: רננות with joyful יהלל shall praise פי׃ and my mouth}%
\verse{אם When זכרתיך I remember על thee upon יצועי my bed, באשׁמרות on thee in the watches. אהגה׃ meditate}%
\verse{כי Because היית thou hast been עזרתה my help, לי ובצל therefore in the shadow כנפיך of thy wings ארנן׃ will I rejoice.}%
\verse{דבקה followeth hard נפשׁי My soul אחריך after בי תמכה upholdeth ימינך׃ thee: thy right hand}%
\verse{והמה But those לשׁואה to destroy יבקשׁו seek נפשׁי my soul, יבאו shall go בתחתיות into the lower parts הארץ׃ of the earth.}%
\verse{יגירהו They shall fall על by ידי by חרב the sword: מנת a portion שׁעלים for foxes. יהיו׃ they shall be}%
\verse{והמלך But the king ישׂמח shall rejoice באלהים in God; יתהלל by him shall glory: כל every one הנשׁבע that sweareth בו כי but יסכר shall be stopped. פי the mouth דוברי of them that speak שׁקר׃ lies}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 64
\verseWithHeading{A Plea for Divine Retribution}{למנצח To the chief Musician, מזמור A Psalm לדוד׃ of David. שׁמע Hear אלהים O God, קולי my voice, בשׂיחי in my prayer: מפחד from fear אויב of the enemy. תצר preserve חיי׃ my life}%
\verse{תסתירני Hide מסוד me from the secret counsel מרעים of the wicked; מרגשׁת from the insurrection פעלי of the workers און׃ of iniquity:}%
\verse{אשׁר Who שׁננו whet כחרב like a sword, לשׁונם their tongue דרכו bend חצם their arrows, דבר words: מר׃ bitter}%
\verse{לירות That they may shoot במסתרים in secret תם at the perfect: פתאם suddenly ירהו do they shoot ולא not. ייראו׃ at him, and fear}%
\verse{יחזקו They encourage למו דבר matter: רע themselves an evil יספרו they commune לטמון of laying snares privily; מוקשׁים of laying snares privily; אמרו they say, מי Who יראה׃ shall see}%
\verse{יחפשׂו They search out עולת iniquities; תמנו they accomplish חפשׂ a diligent search: מחפשׂ a diligent search: וקרב both the inward אישׁ of every one ולב and the heart, עמק׃ deep.}%
\verse{וירם shall shoot אלהים But God חץ at them an arrow; פתאום suddenly היו shall they be מכותם׃ wounded.}%
\verse{ויכשׁילוהו to fall עלימו upon לשׁונם So they shall make their own tongue יתנדדו them shall flee away. כל themselves: all ראה׃ that see}%
\verse{וייראו shall fear, כל And all אדם men ויגידו and shall declare פעל the work אלהים of God; ומעשׂהו of his doing. השׂכילו׃ for they shall wisely consider}%
\verse{ישׂמח shall be glad צדיק The righteous ביהוה וחסה and shall trust בו ויתהללו shall glory. כל in him; and all ישׁרי the upright לב׃ in heart}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 65
\verseWithHeading{Thanksgiving for God’s Provision}{למנצח To the chief Musician, מזמור A Psalm לדוד of David. שׁיר׃ Song לך דמיה waiteth תהלה Praise אלהים for thee, O God, בציון in Zion: ולך ישׁלם be performed. נדר׃ and unto thee shall the vow}%
\verse{שׁמע O thou that hearest תפלה prayer, עדיך unto כל thee shall all בשׂר flesh יבאו׃ come.}%
\verse{דברי עונתברו prevail מני against פשׁעינו me: our transgressions, אתה thou תכפרם׃ shalt purge them away.}%
\verse{אשׁרי Blessed תבחר thou choosest, ותקרב and causest to approach ישׁכן he may dwell חצריך in thy courts: נשׂבעה we shall be satisfied בטוב with the goodness ביתך of thy house, קדשׁ of thy holy היכלך׃ temple.}%
\verse{נוראות terrible things בצדק in righteousness תעננו wilt thou answer אלהי us, O God ישׁענו of our salvation; מבטח the confidence כל of all קצוי the ends ארץ of the earth, וים the sea: רחקים׃ and of them that are afar off}%
\verse{מכין setteth fast הרים the mountains; בכחו Which by his strength נאזר girded בגבורה׃ with power:}%
\verse{משׁביח Which stilleth שׁאון the noise ימים of the seas, שׁאון the noise גליהם of their waves, והמון and the tumult לאמים׃ of the people.}%
\verse{וייראו are afraid ישׁבי They also that dwell קצות in the uttermost parts מאותתיך at thy tokens: מוצאי thou makest the outgoings בקר of the morning וערב and evening תרנין׃ to rejoice.}%
\verse{פקדת Thou visitest הארץ the earth, ותשׁקקה and waterest רבת it: thou greatly תעשׁרנה enrichest פלג it with the river אלהים of God, מלא is full מים of water: תכין thou preparest דגנם them corn, כי when כן thou hast so תכינה׃ provided}%
\verse{תלמיה Thou waterest the ridges רוה thereof abundantly: נחת thou settlest גדודיה the furrows ברביבים with showers: תמגגנה thereof: thou makest it soft צמחה the springing תברך׃ thou blessest}%
\verse{עטרת Thou crownest שׁנת the year טובתך with thy goodness; ומעגליך and thy paths ירעפון drop דשׁן׃ fatness.}%
\verse{ירעפו They drop נאות the pastures מדבר of the wilderness: וגיל rejoice גבעות and the little hills תחגרנה׃ on every side.}%
\verse{לבשׁו are clothed כרים The pastures הצאן with flocks; ועמקים the valleys יעטפו also are covered over בר with corn; יתרועעו they shout for joy, אף they also ישׁירו׃ sing.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 66
\verseWithHeading{Thanksgiving to God for His Works}{למנצח To the chief Musician, שׁיר A Song מזמור Psalm. הריעו Make a joyful noise לאלהים unto God, כל all הארץ׃ ye lands:}%
\verse{זמרו Sing forth כבוד the honor שׁמו of his name: שׂימו make כבוד glorious. תהלתו׃ his praise}%
\verse{אמרו Say לאלהים unto God, מה How נורא terrible מעשׂיך thy works! ברב through the greatness עזך of thy power יכחשׁו submit לך איביך׃ shall thine enemies}%
\verse{כל All הארץ the earth ישׁתחוו shall worship לך ויזמרו thee, and shall sing לך יזמרו unto thee; they shall sing שׁמך thy name. סלה׃ Selah.}%
\verse{לכו Come וראו and see מפעלות the works אלהים of God: נורא terrible עלילה doing על toward בני the children אדם׃ of men.}%
\verse{הפך He turned ים the sea ליבשׁה into dry בנהר through the flood יעברו they went ברגל on foot: שׁם there נשׂמחה׃ did we rejoice}%
\verse{משׁל He ruleth בגבורתו by his power עולם forever; עיניו his eyes בגוים the nations: תצפינה behold הסוררים the rebellious אל let not ירימו exalt למו סלה׃ themselves. Selah.}%
\verse{ברכו O bless עמים ye people, אלהינו our God, והשׁמיעו to be heard: קול and make the voice תהלתו׃ of his praise}%
\verse{השׂם Which holdeth נפשׁנו our soul בחיים in life, ולא not נתן and suffereth למוט to be moved. רגלנו׃ our feet}%
\verse{כי For בחנתנו hast proved אלהים thou, O God, צרפתנו us: thou hast tried כצרף is tried. כסף׃ us, as silver}%
\verse{הבאתנו Thou broughtest במצודה us into the net; שׂמת thou laidst מועקה affliction במתנינו׃ upon our loins.}%
\verse{הרכבת to ride אנושׁ Thou hast caused men לראשׁנו over our heads; באנו we went באשׁ through fire ובמים and through water: ותוציאנו but thou broughtest us out לרויה׃ into a wealthy}%
\verse{אבוא I will go ביתך into thy house בעולות with burnt offerings: אשׁלם I will pay לך נדרי׃ thee my vows,}%
\verse{אשׁר Which פצו have uttered, שׂפתי my lips ודבר hath spoken, פי and my mouth בצר׃ when I was in trouble.}%
\verse{עלות unto thee burnt sacrifices מחים of fatlings, אעלה I will offer לך עם with קטרת the incense אילים of rams; אעשׂה I will offer בקר bullocks עם with עתודים goats. סלה׃ Selah.}%
\verse{לכו Come שׁמעו hear, ואספרה and I will declare כל all יראי ye that fear אלהים God, אשׁר what עשׂה he hath done לנפשׁי׃ for my soul.}%
\verse{אליו unto פי him with my mouth, קראתי I cried ורומם and he was extolled תחת with לשׁוני׃ my tongue.}%
\verse{און iniquity אם If ראיתי I regard בלבי in my heart, לא will not ישׁמע hear אדני׃ the Lord}%
\verse{אכן verily שׁמע hath heard אלהים God הקשׁיב he hath attended בקול to the voice תפלתי׃ of my prayer.}%
\verse{ברוך Blessed אלהים God, אשׁר which לא hath not הסיר turned away תפלתי my prayer, וחסדו nor his mercy מאתי׃ from}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 67
\verseWithHeading{A Prayer of Blessing}{למנצח To the chief Musician בנגינת מזמור A Psalm שׁיר׃ Song. אלהים God יחננו be merciful ויברכנו unto us, and bless יאר to shine פניו us; cause his face אתנו upon סלה׃ us; Selah.}%
\verse{לדעת may be known בארץ upon earth, דרכך That thy way בכל among all גוים nations. ישׁועתך׃ thy saving health}%
\verse{יודוך praise עמים Let the people אלהים thee, O God; יודוך praise עמים the people כלם׃ let all}%
\verse{ישׂמחו be glad וירננו and sing for joy: לאמים O let the nations כי for תשׁפט thou shalt judge עמים the people מישׁור righteously, ולאמים the nations בארץ upon earth. תנחם and govern סלה׃ Selah.}%
\verse{יודוך praise עמים Let the people אלהים thee, O God; יודוך praise עמים the people כלם׃ let all}%
\verse{ארץ shall the earth נתנה yield יבולה her increase; יברכנו shall bless אלהים God, אלהינו׃ our own God,}%
\verse{יברכנו shall bless אלהים God וייראו shall fear אתו כל us; and all אפסי the ends ארץ׃ of the earth}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 68
\verseWithHeading{Praise to God for Providing Victory}{למנצח To the chief Musician, לדוד of David. מזמור A Psalm שׁיר׃ Song יקום arise, אלהים Let God יפוצו be scattered: אויביו let his enemies וינוסו him flee משׂנאיו let them also that hate מפניו׃ before}%
\verse{כהנדף is driven away, עשׁן As smoke תנדף drive away: כהמס melteth דונג as wax מפני before אשׁ the fire, יאבדו perish רשׁעים let the wicked מפני at the presence אלהים׃ of God.}%
\verse{וצדיקים But let the righteous ישׂמחו be glad; יעלצו let them rejoice לפני before אלהים God: וישׂישׂו rejoice. בשׂמחה׃ yea, let them exceedingly}%
\verse{שׁירו Sing לאלהים unto God, זמרו sing praises שׁמו to his name: סלו extol לרכב him that rideth בערבות upon the heavens ביה JAH, שׁמו by his name ועלזו and rejoice לפניו׃ before}%
\verse{אבי A father יתומים of the fatherless, ודין and a judge אלמנות of the widows, אלהים God במעון habitation. קדשׁו׃ in his holy}%
\verse{אלהים God מושׁיב setteth יחידים the solitary ביתה in families: מוציא he bringeth out אסירים those which are bound בכושׁרות with chains: אך but סוררים the rebellious שׁכנו dwell צחיחה׃ in a dry}%
\verse{אלהים O God, בצאתך when thou wentest forth לפני before עמך thy people, בצעדך when thou didst march בישׁימון through the wilderness; סלה׃ Selah:}%
\verse{ארץ The earth רעשׁה shook, אף also שׁמים the heavens נטפו dropped מפני at the presence אלהים of God: זה itself סיני Sinai מפני at the presence אלהים of God, אלהי the God ישׂראל׃ of Israel.}%
\verse{גשׁם rain, נדבות a plentiful תניף didst send אלהים Thou, O God, נחלתך thine inheritance, ונלאה when it was weary. אתה whereby thou כוננתה׃ didst confirm}%
\verse{חיתך Thy congregation ישׁבו hath dwelt בה תכין hast prepared בטובתך of thy goodness לעני for the poor. אלהים׃ therein: thou, O God,}%
\verse{אדני The Lord יתן gave אמר the word: המבשׂרות of those that published צבא the company רב׃ great}%
\verse{מלכי Kings צבאות of armies ידדון did flee apace: ידדון did flee apace: ונות and she that tarried בית at home תחלק divided שׁלל׃ the spoil.}%
\verse{אם Though תשׁכבון ye have lain בין among שׁפתים the pots, כנפי the wings יונה of a dove נחפה covered בכסף with silver, ואברותיה with yellow בירקרק and her feathers חרוץ׃ gold.}%
\verse{בפרשׂ scattered שׁדי When the Almighty מלכים kings בה תשׁלג in it, it was as snow בצלמון׃ in Salmon.}%
\verse{הר The hill אלהים of God הר the hill בשׁן of Bashan; הר hill גבננים a high הר the hill בשׁן׃ of Bashan.}%
\verse{למה Why תרצדון leap הרים hills? גבננים ye, ye high ההר the hill חמד desireth אלהים God לשׁבתו to dwell אף in; yea, יהוה the LORD ישׁכן will dwell לנצח׃ forever.}%
\verse{רכב The chariots אלהים of God רבתים twenty thousand, אלפי thousands שׁנאן of angels: אדני the Lord בם סיני among them, Sinai, בקדשׁ׃ in the holy}%
\verse{עלית Thou hast ascended למרום on high, שׁבית thou hast led captivity captive: שׁבי thou hast led captivity captive: לקחת thou hast received מתנות gifts באדם for men; ואף also, סוררים yea, the rebellious לשׁכן might dwell יה that the LORD אלהים׃ God}%
\verse{ברוך Blessed אדני the Lord, יום daily יום daily יעמס loadeth לנו האל us the God ישׁועתנו of our salvation. סלה׃ Selah.}%
\verse{האל our God לנו אל the God למושׁעות of salvation; וליהוה and unto GOD אדני the Lord למות from death. תוצאות׃ the issues}%
\verse{אך But אלהים God ימחץ shall wound ראשׁ the head איביו of his enemies, קדקד scalp שׂער the hairy מתהלך of such a one as goeth on still באשׁמיו׃ in his trespasses.}%
\verse{אמר said, אדני The Lord מבשׁן אשׁיב I will bring again אשׁיב I will bring again ממצלות from the depths ים׃ of the sea:}%
\verse{למען That תמחץ may be dipped רגלך thy foot בדם in the blood לשׁון the tongue כלביך of thy dogs מאיבים of enemies, מנהו׃ of enemies,}%
\verse{ראו They have seen הליכותיך thy goings, אלהים O God; הליכות the goings אלי of my God, מלכי my King, בקדשׁ׃ in the sanctuary.}%
\verse{קדמו went before, שׁרים The singers אחר after; נגנים the players on instruments בתוך among עלמות the damsels תופפות׃ playing with timbrels.}%
\verse{במקהלות in the congregations, ברכו Bless אלהים ye God יהוה the Lord, ממקור from the fountain ישׂראל׃ of Israel.}%
\verse{שׁם There בנימן Benjamin צעיר little רדם their ruler, שׂרי the princes יהודה of Judah רגמתם their council, שׂרי the princes זבלון of Zebulun, שׂרי the princes נפתלי׃ of Naphtali.}%
\verse{צוה hath commanded אלהיך Thy God עזך thy strength: עוזה strengthen, אלהים O God, זו that which פעלת׃ thou hast wrought}%
\verse{מהיכלך על at ירושׁלם Jerusalem לך יובילו bring מלכים shall kings שׁי׃ presents}%
\verse{גער Rebuke חית the company קנה of spearmen, עדת the multitude אבירים of the bulls, בעגלי with the calves עמים of the people, מתרפס submit himself ברצי with pieces כסף of silver: בזר scatter עמים thou the people קרבות in war. יחפצו׃ delight}%
\verse{יאתיו shall come out חשׁמנים Princes מני of מצרים Egypt; כושׁ Ethiopia תריץ shall soon stretch out ידיו her hands לאלהים׃ unto God.}%
\verse{ממלכות ye kingdoms הארץ of the earth; שׁירו Sing לאלהים unto God, זמרו O sing praises אדני unto the Lord; סלה׃ Selah:}%
\verse{לרכב To him that rideth בשׁמי upon the heavens שׁמי of heavens, קדם of old; הן lo, יתן he doth send out בקולו his voice, קול voice. עז׃ a mighty}%
\verse{תנו Ascribe עז ye strength לאלהים unto God: על over ישׂראל Israel, גאותו his excellency ועזו and his strength בשׁחקים׃ in the clouds.}%
\verse{נורא terrible אלהים O God, ממקדשׁיך out of thy holy places: אל the God ישׂראל of Israel הוא he נתן that giveth עז strength ותעצמות and power לעם unto people. ברוך Blessed אלהים׃ God.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 69
\verseWithHeading{A Plea for Deliverance from Persecution}{למנצח To the chief Musician על upon שׁושׁנים Shoshannim, לדוד׃ of David. הושׁיעני Save אלהים me, O God; כי for באו are come in מים the waters עד unto נפשׁ׃ soul.}%
\verse{טבעתי I sink ביון mire, מצולה in deep ואין where no מעמד standing: באתי I am come במעמקי into deep מים waters, ושׁבלת where the floods שׁטפתני׃ overflow}%
\verse{יגעתי I am weary בקראי of my crying: נחר is dried: גרוני my throat כלו fail עיני mine eyes מיחל while I wait לאלהי׃ for my God.}%
\verse{רבו are more משׂערות than the hairs ראשׁי of mine head: שׂנאי They that hate חנם me without a cause עצמו are mighty: מצמיתי they that would destroy איבי me, mine enemies שׁקר wrongfully, אשׁר which לא גזלתיז then אשׁיב׃ I restored}%
\verse{אלהים O God, אתה thou ידעת knowest לאולתי my foolishness; ואשׁמותי and my sins ממך from לא are not נכחדו׃ hid}%
\verse{אל Let not יבשׁו be ashamed בי קויך them that wait on אדני thee, O Lord יהוה GOD צבאות of hosts, אל for my sake: let not יכלמו thee be confounded בי מבקשׁיך those that seek אלהי for my sake, O God ישׂראל׃ of Israel.}%
\verse{כי Because עליך for thy sake נשׂאתי I have borne חרפה reproach; כסתה hath covered כלמה shame פני׃ my face.}%
\verse{מוזר a stranger הייתי I am become לאחי unto my brethren, ונכרי and an alien לבני children. אמי׃ unto my mother's}%
\verse{כי For קנאת the zeal ביתך of thine house אכלתני hath eaten me up; וחרפות and the reproaches חורפיך of them that reproached נפלו thee are fallen עלי׃ upon}%
\verse{ואבכה When I wept, בצום with fasting, נפשׁי my soul ותהי that was לחרפות׃ to my reproach.}%
\verse{ואתנה I made לבושׁי also my garment; שׂק sackcloth ואהי and I became להם למשׁל׃ a proverb}%
\verse{ישׂיחו speak בי ישׁבי They that sit שׁער in the gate ונגינות against me; and I the song שׁותי of the drunkards. שׁכר׃ of the drunkards.}%
\verse{ואני But as for me, תפלתי my prayer לך יהוה unto thee, O LORD, עת time: רצון an acceptable אלהים O God, ברב in the multitude חסדך of thy mercy ענני hear באמת me, in the truth ישׁעך׃ of thy salvation.}%
\verse{הצילני Deliver מטיט me out of the mire, ואל and let me not אטבעה sink: אנצלה let me be delivered משׂנאי from them that hate וממעמקי me, and out of the deep מים׃ waters.}%
\verse{אל Let not תשׁטפני overflow שׁבלת the waterflood מים the waterflood ואל me, neither תבלעני swallow me up, מצולה let the deep ואל and let not תאטר shut עלי upon באר the pit פיה׃ her mouth}%
\verse{ענני Hear יהוה me, O LORD; כי for טוב good: חסדך thy lovingkindness כרב me according to the multitude רחמיך of thy tender mercies. פנה turn אלי׃ unto}%
\verse{ואל not תסתר And hide פניך thy face מעבדך from thy servant; כי for צר לי מהרנני׃ hear}%
\verse{קרבה Draw nigh אל unto נפשׁי my soul, גאלה redeem למען me because of איבי mine enemies. פדני׃ it: deliver}%
\verse{אתה Thou ידעת hast known חרפתי my reproach, ובשׁתי and my shame, וכלמתי and my dishonor: נגדך before כל all צוררי׃ mine adversaries}%
\verse{חרפה Reproach שׁברה hath broken לבי my heart; ואנושׁה and I am full of heaviness: ואקוה and I looked לנוד to take pity, ואין but none; ולמנחמים and for comforters, ולא none. מצאתי׃ but I found}%
\verse{ויתנו They gave בברותי for my meat; ראשׁ me also gall ולצמאי and in my thirst ישׁקוני to drink. חמץ׃ they gave me vinegar}%
\verse{יהי become שׁלחנם Let their table לפניהם before לפח a snare ולשׁלומים them: and for welfare, למוקשׁ׃ a trap.}%
\verse{תחשׁכנה be darkened, עיניהם Let their eyes מראות that they see not; ומתניהם and make their loins תמיד continually המעד׃ to shake.}%
\verse{שׁפך Pour out עליהם upon זעמך thine indignation וחרון them, and let thy wrathful אפך anger ישׂיגם׃ take hold}%
\verse{תהי be טירתם Let their habitation נשׁמה desolate; באהליהם in their tents. אל none יהי let ישׁב׃ dwell}%
\verse{כי For אתה thou אשׁר whom הכית hast smitten; רדפו they persecute ואל to מכאוב the grief חלליך of those whom thou hast wounded. יספרו׃ and they talk}%
\verse{תנה Add עון iniquity על unto עונם their iniquity: ואל and let them not יבאו come בצדקתך׃ into thy righteousness.}%
\verse{ימחו Let them be blotted מספר out of the book חיים of the living, ועם with צדיקים the righteous. אל and not יכתבו׃ be written}%
\verse{ואני But I עני poor וכואב and sorrowful: ישׁועתך let thy salvation, אלהים O God, תשׂגבני׃ set me up on high.}%
\verse{אהללה I will praise שׁם the name אלהים of God בשׁיר with a song, ואגדלנו and will magnify בתודה׃ him with thanksgiving.}%
\verse{ותיטב also shall please ליהוה the LORD משׁור better than an ox פר bullock מקרן that hath horns מפריס׃ and hooves.}%
\verse{ראו shall see ענוים The humble ישׂמחו be glad: דרשׁי that seek אלהים God. ויחי shall live לבבכם׃ and your heart}%
\verse{כי For שׁמע heareth אל heareth אביונים the poor, יהוה the LORD ואת אסיריו his prisoners. לא not בזה׃ and despiseth}%
\verse{יהללוהו praise שׁמים Let the heaven וארץ and earth ימים him, the seas, וכל and every thing רמשׂ׃ that moveth}%
\verse{כי For אלהים God יושׁיע will save ציון Zion, ויבנה and will build ערי the cities יהודה of Judah: וישׁבו that they may dwell שׁם there, וירשׁוה׃ and have it in possession.}%
\verse{וזרע The seed עבדיו also of his servants ינחלוה shall inherit ואהבי it: and they that love שׁמו his name ישׁכנו׃ shall dwell}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 70
\verseWithHeading{A Prayer for Deliverance from Enemies}{למנצח To the chief Musician, לדוד of David, להזכיר׃ to bring to remembrance. אלהים O God, להצילני to deliver יהוה me, O LORD. לעזרתי to help חושׁה׃ me; make haste}%
\verse{יבשׁו Let them be ashamed ויחפרו and confounded מבקשׁי that seek after נפשׁי my soul: יסגו let them be turned אחור backward, ויכלמו and put to confusion, חפצי that desire רעתי׃ my hurt.}%
\verse{ישׁובו Let them be turned back על for עקב a reward בשׁתם of their shame האמרים that say, האח Aha, האח׃ aha.}%
\verse{ישׂישׂו thee rejoice וישׂמחו and be glad בך כל Let all מבקשׁיך those that seek ויאמרו say תמיד continually, יגדל be magnified. אלהים Let God אהבי in thee: and let such as love ישׁועתך׃ thy salvation}%
\verse{ואני But I עני poor ואביון and needy: אלהים unto me, O God: חושׁה make haste לי עזרי my help ומפלטי and my deliverer; אתה thou יהוה O LORD, אל make no תאחר׃ tarrying.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 71
\verseWithHeading{A Prayer to God the Rock of Refuge}{בך יהוה In thee, O LORD, חסיתי do I put my trust: אל let me never אבושׁה be put to confusion. לעולם׃ let me never}%
\verse{בצדקתך me in thy righteousness, תצילני Deliver ותפלטני and cause me to escape: הטה incline אלי unto אזנך thine ear והושׁיעני׃ me, and save}%
\verse{היה Be לי לצור thou my strong מעון habitation, לבוא resort: תמיד whereunto I may continually צוית thou hast given commandment להושׁיעני to save כי me; for סלעי my rock ומצודתי and my fortress. אתה׃ thou}%
\verse{אלהי me, O my God, פלטני Deliver מיד out of the hand רשׁע of the wicked, מכף out of the hand מעול of the unrighteous וחומץ׃ and cruel}%
\verse{כי For אתה thou תקותי my hope, אדני O Lord יהוה מבטחי my trust מנעורי׃ from my youth.}%
\verse{עליך By נסמכתי thee have I been holden up מבטן from the womb: ממעי bowels: אמי me out of my mother's אתה thou גוזי art he that took בך תהלתי my praise תמיד׃ continually}%
\verse{כמופת as a wonder הייתי I am לרבים unto many; ואתה but thou מחסי refuge. עז׃ my strong}%
\verse{ימלא be filled פי Let my mouth תהלתך thy praise כל all היום the day. תפארתך׃ thy honor}%
\verse{אל me not תשׁליכני לעת in the time זקנה of old age; ככלות faileth. כחי when my strength אל תעזבני׃ forsake}%
\verse{כי For אמרו speak אויבי mine enemies לי ושׁמרי against me; and they that lay wait for נפשׁי my soul נועצו take counsel יחדו׃ together,}%
\verse{לאמר Saying, אלהים God עזבו hath forsaken רדפו him: persecute ותפשׂוהו and take כי him; for אין none מציל׃ to deliver}%
\verse{אלהים O God, אל be not תרחק far ממני from אלהי me: O my God, לעזרתי for my help. חישׁה׃}%
\verse{יבשׁו Let them be confounded יכלו consumed שׂטני that are adversaries נפשׁי to my soul; יעטו let them be covered חרפה reproach וכלמה and dishonor מבקשׁי that seek רעתי׃ my hurt.}%
\verse{ואני But I תמיד continually, איחל will hope והוספתי thee more and more. על כלהלתך׃ and will yet praise}%
\verse{פי My mouth יספר shall show forth צדקתך thy righteousness כל all היום the day; תשׁועתך thy salvation כי for לא not ידעתי I know ספרות׃ the numbers}%
\verse{אבוא I will go בגברות in the strength אדני of the Lord יהוה GOD: אזכיר I will make mention צדקתך of thy righteousness, לבדך׃ of thine only.}%
\verse{אלהים O God, למדתני thou hast taught מנעורי me from my youth: ועד and hitherto הנה and hitherto אגיד have I declared נפלאותיך׃ thy wondrous works.}%
\verse{וגם Now also עד when זקנה I am old ושׂיבה and grayheaded, אלהים O God, אל me not; תעזבני forsake עד until אגיד I have showed זרועך thy strength לדור unto generation, לכל to every one יבוא is to come. גבורתך׃ thy power}%
\verse{וצדקתך Thy righteousness אלהים also, O God, עד very מרום high, אשׁר who עשׂית hast done גדלות great things: אלהים O God, מי who כמוך׃ like unto thee!}%
\verse{אשׁר which הראיתנו hast showed צרות troubles, רבות me great ורעות and sore תשׁוב me again, תחיינו shalt quicken ומתהמות from the depths הארץ of the earth. תשׁוב again תעלני׃ and shalt bring me up}%
\verse{תרב Thou shalt increase גדלתי my greatness, ותסב me on every side. תנחמני׃ and comfort}%
\verse{גם will also אני I אודך praise בכלי thee with the psaltery, נבל thee with the psaltery, אמתך thy truth, אלהי O my God: אזמרה unto thee will I sing לך בכנור with the harp, קדושׁ O thou Holy One ישׂראל׃ of Israel.}%
\verse{תרננה shall greatly rejoice שׂפתי My lips כי when אזמרה I sing לך ונפשׁי unto thee; and my soul, אשׁר which פדית׃ thou hast redeemed.}%
\verse{גם also לשׁוני My tongue כל all היום the day תהגה shall talk צדקתך of thy righteousness כי long: for בשׁו they are confounded, כי for חפרו they are brought unto shame, מבקשׁי that seek רעתי׃ my hurt.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 72
\verseWithHeading{A Prayer for the Prosperity of God’s Anointed King}{לשׁלמה for Solomon. אלהים O God, משׁפטיך thy judgments, למלך the king תן Give וצדקתך and thy righteousness לבן son. מלך׃ unto the king's}%
\verse{ידין He shall judge עמך thy people בצדק with righteousness, וענייך and thy poor במשׁפט׃ with judgment.}%
\verse{ישׂאו shall bring הרים The mountains שׁלום peace לעם to the people, וגבעות and the little hills, בצדקה׃ by righteousness.}%
\verse{ישׁפט He shall judge עניי the poor עם of the people, יושׁיע he shall save לבני the children אביון of the needy, וידכא and shall break in pieces עושׁק׃ the oppressor.}%
\verse{ייראוך They shall fear עם thee as long שׁמשׁ as the sun ולפני endure, ירח and moon דור throughout all generations. דורים׃ throughout all generations.}%
\verse{ירד He shall come down כמטר like rain על upon גז the mown grass: כרביבים as showers זרזיף water ארץ׃ the earth.}%
\verse{יפרח flourish; בימיו In his days צדיק shall the righteous ורב and abundance שׁלום of peace עד so long בלי endureth. ירח׃ as the moon}%
\verse{וירד He shall have dominion מים also from sea עד to ים sea, ומנהר and from the river עד unto אפסי the ends ארץ׃ of the earth.}%
\verse{לפניו before יכרעו shall bow ציים They that dwell in the wilderness ואיביו him; and his enemies עפר the dust. ילחכו׃ shall lick}%
\verse{מלכי The kings תרשׁישׁ of Tarshish ואיים and of the isles מנחה presents: ישׁיבו shall bring מלכי the kings שׁבא of Sheba וסבא and Seba אשׁכר gifts. יקריבו׃ shall offer}%
\verse{וישׁתחוו shall fall down לו כל Yea, all מלכים kings כל before him: all גוים nations יעבדוהו׃ shall serve}%
\verse{כי For יציל he shall deliver אביון the needy משׁוע when he crieth; ועני the poor ואין also, and that hath no עזר׃ helper.}%
\verse{יחס עלל the poor ואביון and needy, ונפשׁות the souls אביונים of the needy. יושׁיע׃ and shall save}%
\verse{מתוך from deceit ומחמס and violence: יגאל He shall redeem נפשׁם their soul וייקר and precious דמם shall their blood בעיניו׃ be in his sight.}%
\verse{ויחי And he shall live, ויתן and to him shall be given לו מזהב of the gold שׁבא of Sheba: ויתפלל prayer בעדו also shall be made for תמיד him continually; כל daily היום daily יברכנהו׃ shall he be praised.}%
\verse{יהי There shall be פסת a handful בר of corn בארץ in the earth בראשׁ upon the top הרים of the mountains; ירעשׁ thereof shall shake כלבנון like Lebanon: פריו the fruit ויציצו shall flourish מעיר and of the city כעשׂב like grass הארץ׃ of the earth.}%
\verse{יהי shall endure שׁמו His name לעולם forever: לפני as long שׁמשׁ as the sun: ינין shall be continued שׁמו his name ויתברכו and shall be blessed בו כל in him: all גוים nations יאשׁרוהו׃ shall call him blessed.}%
\verse{ברוך Blessed יהוה the LORD אלהים God, אלהי the God ישׂראל of Israel, עשׂה doeth נפלאות wondrous things. לבדו׃ who only}%
\verse{וברוך And blessed שׁם name כבודו his glorious לעולם forever: וימלא be filled כבודו his glory; את כל and let the whole הארץ earth אמן Amen, ואמן׃ and Amen.}%
\verse{כלו are ended. תפלות The prayers דוד of David בן the son ישׁי׃ of Jesse}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 73
\verseWithHeading{The Wicked and the Righteous Contrasted}{מזמור A Psalm לאסף of Asaph. אך Truly טוב good לישׂראל to Israel, אלהים God לברי to such as are of a clean לבב׃ heart.}%
\verse{ואני But as for me, כמעט were almost נטוי gone; רגלי my feet כאין had well nigh שׁפכה slipped. אשׁרי׃ my steps}%
\verse{כי For קנאתי I was envious בהוללים at the foolish, שׁלום the prosperity רשׁעים of the wicked. אראה׃ I saw}%
\verse{כי For אין no חרצבות bands למותם in their death: ובריא firm. אולם׃ but their strength}%
\verse{בעמל in trouble אנושׁ men; אינמו They not ועם like אדם men. לא neither ינגעו׃ are they plagued}%
\verse{לכן Therefore ענקתמו compasseth them about as a chain; גאוה pride יעטף covereth שׁית them a garment. חמס׃ violence}%
\verse{יצא stand out מחלב with fatness: עינמו Their eyes עברו they have more משׂכיות could wish. לבב׃ than heart}%
\verse{ימיקו They are corrupt, וידברו and speak ברע wickedly עשׁק oppression: ממרום loftily. ידברו׃ they speak}%
\verse{שׁתו They set בשׁמים against the heavens, פיהם their mouth ולשׁונם and their tongue תהלך walketh בארץ׃ through the earth.}%
\verse{לכן Therefore ישׁיב return עמו his people הלם hither: ומי and waters מלא of a full ימצו׃ are wrung out}%
\verse{ואמרו And they say, איכה How ידע know? אל doth God וישׁ and is there דעה knowledge בעליון׃ in the most High?}%
\verse{הנה Behold, אלה these רשׁעים the ungodly, ושׁלוי who prosper עולם in the world; השׂגו they increase חיל׃ riches.}%
\verse{אך Verily ריק vain, זכיתי I have cleansed לבבי my heart וארחץ and washed בנקיון in innocency. כפי׃ my hands}%
\verse{ואהי long have I been נגוע plagued, כל For all היום the day ותוכחתי and chastened לבקרים׃ every morning.}%
\verse{אם If אמרתי I say, אספרה I will speak כמו thus; הנה behold, דור the generation בניך of thy children. בגדתי׃ I should offend}%
\verse{ואחשׁבה When I thought לדעת to know זאת this, עמל too painful היא it בעיני׃ for}%
\verse{עד Until אבוא I went אל into מקדשׁי the sanctuary אל of God; אבינה understood לאחריתם׃ I their end.}%
\verse{אך Surely בחלקות them in slippery places: תשׁית thou didst set למו הפלתם thou castedst them down למשׁואות׃ into destruction.}%
\verse{איך How היו are לשׁמה they into desolation, כרגע as in a moment! ספו they are utterly consumed תמו they are utterly consumed מן with בלהות׃ terrors.}%
\verse{כחלום As a dream מהקיץ when awaketh; אדני O Lord, בעיר when thou awakest, צלמם their image. תבזה׃ thou shalt despise}%
\verse{כי Thus יתחמץ was grieved, לבבי my heart וכליותי in my reins. אשׁתונן׃ and I was pricked}%
\verse{ואני I, בער So foolish ולא and ignorant: אדע and ignorant: בהמות a beast הייתי I was עמך׃ before}%
\verse{ואני Nevertheless I תמיד continually עמך with אחזת thee: thou hast holden ביד hand. ימיני׃ by my right}%
\verse{בעצתך me with thy counsel, תנחני Thou shalt guide ואחר and afterward כבוד me glory. תקחני׃ receive}%
\verse{מי Whom לי בשׁמים have I in heaven ועמך beside לא and none חפצתי I desire בארץ׃ upon earth}%
\verse{כלה faileth: שׁארי My flesh ולבבי and my heart צור the strength לבבי of my heart, וחלקי and my portion אלהים God לעולם׃ forever.}%
\verse{כי For, הנה lo, רחקיך they that are far יאבדו from thee shall perish: הצמתה thou hast destroyed כל all זונה them that go a whoring ממך׃ from}%
\verse{ואני But it קרבת for me to draw near אלהים to God: לי טוב good שׁתי I have put באדני in the Lord יהוה GOD, מחסי my trust לספר that I may declare כל all מלאכותיך׃ thy works.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 74
\verseWithHeading{A Lament in Time of National Defeat}{משׂכיל Maschil לאסף of Asaph. למה why אלהים O God, זנחת hast thou cast off לנצח forever? יעשׁן smoke אפך doth thine anger בצאן against the sheep מרעיתך׃ of thy pasture?}%
\verse{זכר Remember עדתך thy congregation, קנית thou hast purchased קדם of old; גאלת thou hast redeemed; שׁבט the rod נחלתך of thine inheritance, הר mount ציון Zion, זה this שׁכנת׃ wherein thou hast dwelt.}%
\verse{הרימה Lift up פעמיך thy feet למשׁאות desolations; נצח unto the perpetual כל all הרע hath done wickedly אויב the enemy בקדשׁ׃ in the sanctuary.}%
\verse{שׁאגו roar צרריך Thine enemies בקרב in the midst מועדך of thy congregations; שׂמו they set up אותתם their ensigns אתות׃ signs.}%
\verse{יודע was famous כמביא according as he had lifted up למעלה according as he had lifted up בסבך upon the thick עץ trees. קרדמות׃ axes}%
\verse{ועת פתוחיה the carved work יחד thereof at once בכשׁיל with axes וכילפת and hammers. יהלמון׃ they break down}%
\verse{שׁלחו They have cast באשׁ fire מקדשׁך into thy sanctuary, לארץ to the ground. חללו they have defiled משׁכן the dwelling place שׁמך׃ of thy name}%
\verse{אמרו They said בלבם in their hearts, נינם Let us destroy יחד them together: שׂרפו they have burned up כל all מועדי the synagogues אל of God בארץ׃ in the land.}%
\verse{אותתינו our signs: לא not ראינו We see אין no עוד more נביא any prophet: ולא neither אתנו among ידע us any that knoweth עד how long. מה׃ how long.}%
\verse{עד how long מתי how long אלהים O God, יחרף reproach? צר shall the adversary ינאץ blaspheme אויב shall the enemy שׁמך thy name לנצח׃ forever?}%
\verse{למה Why תשׁיב withdrawest ידך thou thy hand, וימינך even thy right hand? מקרב out of חוקך thy bosom. כלה׃ pluck}%
\verse{ואלהים For God מלכי my King מקדם of old, פעל working ישׁועות salvation בקרב in the midst הארץ׃ of the earth.}%
\verse{אתה Thou פוררת didst divide בעזך by thy strength: ים the sea שׁברת thou didst break ראשׁי the heads תנינים of the dragons על in המים׃ the waters.}%
\verse{אתה Thou רצצת didst break ראשׁי the heads לויתן of leviathan תתננו in pieces, gavest מאכל him meat לעם to the people לציים׃ inhabiting the wilderness.}%
\verse{אתה Thou בקעת didst cleave מעין the fountain ונחל and the flood: אתה thou הובשׁת driedst up נהרות rivers. איתן׃ mighty}%
\verse{לך יום The day אף also לך לילה thine, the night אתה thine: thou הכינות hast prepared מאור the light ושׁמשׁ׃ and the sun.}%
\verse{אתה Thou הצבת hast set כל all גבולות the borders ארץ of the earth: קיץ summer וחרף and winter. אתה thou יצרתם׃ hast made}%
\verse{זכר Remember זאת this, אויב the enemy חרף hath reproached, יהוה O LORD, ועם people נבל and the foolish נאצו have blasphemed שׁמך׃ thy name.}%
\verse{אל not תתן O deliver לחית unto the multitude נפשׁ the soul תורך of thy turtledove חית the congregation ענייך of thy poor אל not תשׁכח forget לנצח׃ forever.}%
\verse{הבט Have respect לברית unto the covenant: כי for מלאו are full מחשׁכי the dark places ארץ of the earth נאות of the habitations חמס׃ of cruelty.}%
\verse{אל O let not ישׁב return דך the oppressed נכלם ashamed: עני let the poor ואביון and needy יהללו praise שׁמך׃ thy name.}%
\verse{קומה Arise, אלהים O God, ריבה plead thine own cause: ריבך plead thine own cause: זכר remember חרפתך reproacheth מני how נבל the foolish man כל thee daily. היום׃ thee daily.}%
\verse{אל not תשׁכח Forget קול the voice צרריך of thine enemies: שׁאון the tumult קמיך of those that rise up against עלה thee increaseth תמיד׃ continually.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 75
\verseWithHeading{Thanksgiving for God’s Future Help}{למנצח To the chief Musician, אל תשׁחת Al-taschith, מזמור A Psalm לאסף of Asaph. שׁיר׃ Song הודינו do we give thanks, לך אלהים Unto thee, O God, הודינו do we give thanks: וקרוב is near שׁמך for thy name ספרו declare. נפלאותיך׃ thy wondrous works}%
\verse{כי When אקח I shall receive מועד the congregation אני I מישׁרים uprightly. אשׁפט׃ will judge}%
\verse{נמגים thereof are dissolved: ארץ The earth וכל and all ישׁביה the inhabitants אנכי I תכנתי bear up עמודיה the pillars סלה׃ of it. Selah.}%
\verse{אמרתי I said להוללים unto the fools, אל תהלולרשׁעים and to the wicked, אל תרימורן׃ the horn:}%
\verse{אל תרימומרום on high: קרנכם your horn תדברו speak בצואר neck. עתק׃ a stiff}%
\verse{כי For לא neither ממוצא from the east, וממערב nor from the west, ולא nor ממדבר from the south. הרים׃ promotion}%
\verse{כי But אלהים God שׁפט the judge: זה one, ישׁפיל he putteth down וזה up another. ירים׃ and setteth}%
\verse{כי For כוס a cup, ביד in the hand יהוה of the LORD ויין and the wine חמר is red; מלא it is full מסך of mixture; ויגר and he poureth out מזה of the same: אך but שׁמריה the dregs ימצו shall wring out, ישׁתו drink כל thereof, all רשׁעי the wicked ארץ׃ of the earth}%
\verse{ואני But I אגיד will declare לעלם forever; אזמרה I will sing praises לאלהי to the God יעקב׃ of Jacob.}%
\verse{וכל All קרני the horns רשׁעים of the wicked אגדע also will I cut off; תרוממנה shall be exalted. קרנות the horns צדיק׃ of the righteous}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 76
\verseWithHeading{Praise to God for His Rescue of Israel}{למנצח To the chief Musician בנגינת on Neginoth, מזמור A Psalm לאסף of Asaph. שׁיר׃ Song נודע known: ביהודה In Judah אלהים God בישׂראל in Israel. גדול great שׁמו׃ his name}%
\verse{ויהי also is בשׁלם In Salem סכו his tabernacle, ומעונתו and his dwelling place בציון׃ in Zion.}%
\verse{שׁמה There שׁבר broke רשׁפי he the arrows קשׁת of the bow, מגן the shield, וחרב and the sword, ומלחמה and the battle. סלה׃ Selah.}%
\verse{נאור more glorious אתה Thou אדיר excellent מהררי than the mountains טרף׃ of prey.}%
\verse{אשׁתוללו are spoiled, אבירי לבמו they have slept שׁנתם their sleep: ולא and none מצאו have found כל and none אנשׁי חיל of might ידיהם׃ their hands.}%
\verse{מגערתך אלהי O God יעקב of Jacob, נרדם are cast into a dead sleep. ורכב both the chariot וסוס׃ and horse}%
\verse{אתה Thou, נורא to be feared: אתה thou, ומי and who יעמד may stand לפניך in thy sight מאז when once אפך׃ thou art angry?}%
\verse{משׁמים heaven; השׁמעת to be heard דין Thou didst cause judgment ארץ the earth יראה feared, ושׁקטה׃ and was still,}%
\verse{בקום arose למשׁפט to judgment, אלהים When God להושׁיע to save כל all ענוי the meek ארץ of the earth. סלה׃ Selah.}%
\verse{כי Surely חמת the wrath אדם of man תודך shall praise שׁארית thee: the remainder חמת of wrath תחגר׃ shalt thou restrain.}%
\verse{נדרו Vow, ושׁלמו and pay ליהוה unto the LORD אלהיכם your God: כל let all סביביו that be round about יובילו him bring שׁי presents למורא׃ unto him that ought to be feared.}%
\verse{יבצר He shall cut off רוח the spirit נגידים of princes: נורא terrible למלכי to the kings ארץ׃ of the earth.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 77
\verseWithHeading{Remembering God’s Help for Israel}{למנצח To the chief Musician, על to ידיתון Jeduthun, לאסף of Asaph. מזמור׃ A Psalm קולי with my voice, אל unto אלהים God ואצעקה I cried קולי with my voice; אל unto אלהים God והאזין and he gave ear אלי׃ unto}%
\verse{ביום In the day צרתי of my trouble אדני the Lord: דרשׁתי I sought ידי my sore לילה in the night, נגרה ran ולא not: תפוג and ceased מאנה refused הנחם to be comforted. נפשׁי׃ my soul}%
\verse{אזכרה I remembered אלהים God, ואהמיה and was troubled: אשׂיחה I complained, ותתעטף was overwhelmed. רוחי and my spirit סלה׃ Selah.}%
\verse{אחזת Thou holdest שׁמרות waking: עיני mine eyes נפעמתי I am so troubled ולא that I cannot אדבר׃ speak.}%
\verse{חשׁבתי I have considered ימים the days מקדם of old, שׁנות the years עולמים׃ of ancient times.}%
\verse{אזכרה I call to remembrance נגינתי my song בלילה in the night: עם with לבבי mine own heart: אשׂיחה I commune ויחפשׂ made diligent search. רוחי׃ and my spirit}%
\verse{הלעולמים forever? יזנח cast off אדני Will the Lord ולא no יסיף more? לרצות and will he be favorable עוד׃ more?}%
\verse{האפס clean gone לנצח forever? חסדו Is his mercy גמר fail אמר doth promise לדר forevermore? ודר׃ forevermore?}%
\verse{השׁכח forgotten חנות to be gracious? אל Hath God אם קפץ shut up באף hath he in anger רחמיו his tender mercies? סלה׃ Selah.}%
\verse{ואמר And I said, חלותי my infirmity: היא This שׁנות the years ימין of the right hand עליון׃ of the most High.}%
\verse{אזכיר I will remember מעללי the works יה of the LORD: כי surely אזכרה I will remember מקדם of old. פלאך׃ thy wonders}%
\verse{והגיתי I will meditate בכל also of all פעלך thy work, ובעלילותיך of thy doings. אשׂיחה׃ and talk}%
\verse{אלהים O God, בקדשׁ in the sanctuary: דרכך Thy way, מי who אל a God גדול great כאלהים׃ as God?}%
\verse{אתה Thou האל the God עשׂה that doest פלא wonders: הודעת thou hast declared בעמים among the people. עזך׃ thy strength}%
\verse{גאלת redeemed בזרוע Thou hast with arm עמך thy people, בני the sons יעקב of Jacob ויוסף and Joseph. סלה׃ Selah.}%
\verse{ראוך saw מים The waters אלהים thee, O God, ראוך saw מים the waters יחילו thee; they were afraid: אף also ירגזו were troubled. תהמות׃ the depths}%
\verse{זרמו poured out מים water: עבות The clouds קול a sound: נתנו sent out שׁחקים the skies אף also חצציך יתהלכו׃ went abroad.}%
\verse{קול The voice רעמך of thy thunder בגלגל in the heaven: האירו lightened ברקים the lightnings תבל the world: רגזה trembled ותרעשׁ and shook. הארץ׃ the earth}%
\verse{בים in the sea, דרכך Thy way ושׁביליך and thy path במים waters, רבים in the great ועקבותיך and thy footsteps לא are not נדעו׃ known.}%
\verse{נחית Thou leddest כצאן like a flock עמך thy people ביד by the hand משׁה of Moses ואהרן׃ and Aaron.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 78
\verseWithHeading{God’s Faithfulness in Israel’s History}{משׂכיל Maschil לאסף of Asaph. האזינה Give ear, עמי O my people, תורתי my law: הטו incline אזנכם your ears לאמרי to the words פי׃ of my mouth.}%
\verse{אפתחה I will open במשׁל in a parable: פי my mouth אביעה I will utter חידות dark sayings מני of קדם׃ old:}%
\verse{אשׁר Which שׁמענו we have heard ונדעם and known, ואבותינו and our fathers ספרו׃ have told}%
\verse{לא We will not נכחד hide מבניהם from their children, לדור to the generation אחרון to come מספרים showing תהלות the praises יהוה of the LORD, ועזוזו and his strength, ונפלאותיו and his wonderful works אשׁר that עשׂה׃ he hath done.}%
\verse{ויקם For he established עדות a testimony ביעקב in Jacob, ותורה a law שׂם and appointed בישׂראל in Israel, אשׁר which צוה he commanded את אבותינו our fathers, להודיעם that they should make them known לבניהם׃ to their children:}%
\verse{למען That ידעו might know דור the generation אחרון to come בנים the children יולדו should be born; יקמו should arise ויספרו and declare לבניהם׃ to their children:}%
\verse{וישׂימו That they might set באלהים in God, כסלם their hope ולא and not ישׁכחו forget מעללי the works אל of God, ומצותיו his commandments: ינצרו׃ but keep}%
\verse{ולא And might not יהיו be כאבותם as their fathers, דור generation; סורר a stubborn ומרה and rebellious דור a generation לא was not הכין set not their heart aright, לבו ולאאמנה steadfast את with אל God. רוחו׃ and whose spirit}%
\verse{בני The children אפרים of Ephraim, נושׁקי armed, רומי carrying קשׁת bows, הפכו turned back ביום in the day קרב׃ of battle.}%
\verse{לא not שׁמרו They kept ברית the covenant אלהים of God, ובתורתו in his law; מאנו and refused ללכת׃ to walk}%
\verse{וישׁכחו And forgot עלילותיו his works, ונפלאותיו and his wonders אשׁר that הראם׃ he had showed}%
\verse{נגד he in the sight אבותם of their fathers, עשׂה did פלא Marvelous things בארץ in the land מצרים of Egypt, שׂדה the field צען׃ of Zoan.}%
\verse{בקע He divided ים the sea, ויעבירם and caused them to pass through; ויצב to stand מים and he made the waters כמו as נד׃ a heap.}%
\verse{וינחם also he led בענן them with a cloud, יומם In the daytime וכל and all הלילה the night באור with a light אשׁ׃ of fire.}%
\verse{יבקע He cleaved צרים the rocks במדבר in the wilderness, וישׁק and gave drink כתהמות depths. רבה׃ as the great}%
\verse{ויוצא נוזליםסלע of the rock, ויורד to run down כנהרות like rivers. מים׃ and caused waters}%
\verse{ויוסיפו more עוד yet לחטא And they sinned לו למרות against him by provoking עליון the most High בציה׃ in the wilderness.}%
\verse{וינסו And they tempted אל God בלבבם in their heart לשׁאל by asking אכל meat לנפשׁם׃ for their lust.}%
\verse{וידברו Yea, they spoke באלהים against God; אמרו they said, היוכל Can אל God לערך furnish שׁלחן a table במדבר׃ in the wilderness?}%
\verse{הן Behold, הכה he smote צור the rock, ויזובו gushed out, מים that the waters ונחלים and the streams ישׁטפו overflowed; הגם also? לחם bread יוכל can תת he give אם יכין can he provide שׁאר flesh לעמו׃ for his people?}%
\verse{לכן Therefore שׁמע heard יהוה the LORD ויתעבר and was wroth: ואשׁ so a fire נשׂקה was kindled ביעקב against Jacob, וגם also אף and anger עלה came up בישׂראל׃ against Israel;}%
\verse{כי Because לא not האמינו they believed באלהים in God, ולא not בטחו and trusted בישׁועתו׃ in his salvation:}%
\verse{ויצו Though he had commanded שׁחקים the clouds ממעל from above, ודלתי the doors שׁמים of heaven, פתח׃ and opened}%
\verse{וימטר And had rained down עליהם upon מן manna לאכל them to eat, ודגן them of the corn שׁמים of heaven. נתן׃ and had given}%
\verse{לחם food: אבירים angels' אכל did eat אישׁ Man צידה them meat שׁלח he sent להם לשׂבע׃ to the full.}%
\verse{יסע to blow קדים He caused an east wind בשׁמים in the heaven: וינהג he brought בעזו and by his power תימן׃ in the south wind.}%
\verse{וימטר He rained עליהם also upon כעפר them as dust, שׁאר flesh וכחול like as the sand ימים of the sea: עוף fowls כנף׃ and feathered}%
\verse{ויפל And he let fall בקרב in the midst מחנהו of their camp, סביב round about למשׁכנתיו׃ their habitations.}%
\verse{ויאכלו So they did eat, וישׂבעו filled: מאד and were well ותאותם them their own desire; יבא for he gave להם׃}%
\verse{לא They were not זרו estranged מתאותם from their lust. עוד yet אכלם But while their meat בפיהם׃ in their mouths,}%
\verse{ואף The wrath אלהים of God עלה came בהם ויהרג upon them, and slew במשׁמניהם the fattest ובחורי the chosen ישׂראל of Israel. הכריע׃ of them, and smote down}%
\verse{בכל For all זאת this חטאו they sinned עוד still, ולא not האמינו and believed בנפלאותיו׃ for his wondrous works.}%
\verse{ויכל did he consume בהבל in vanity, ימיהם Therefore their days ושׁנותם and their years בבהלה׃ in trouble.}%
\verse{אם When הרגם he slew ודרשׁוהו them, then they sought ושׁבו him: and they returned ושׁחרו and inquired early אל׃ after God.}%
\verse{ויזכרו And they remembered כי that אלהים God צורם their rock, ואל God עליון and the high גאלם׃ their redeemer.}%
\verse{ויפתוהו Nevertheless they did flatter בפיהם him with their mouth, ובלשׁונם unto him with their tongues. יכזבו׃ and they lied}%
\verse{ולבם For their heart לא was not right נכון was not right עמו with ולא him, neither נאמנו were they steadfast בבריתו׃ in his covenant.}%
\verse{והוא But he, רחום full of compassion, יכפר forgave עון iniquity, ולא not: ישׁחית and destroyed והרבה yea, many להשׁיב a time turned he his anger away, אפו a time turned he his anger away, ולא and did not יעיר stir up כל all חמתו׃ his wrath.}%
\verse{ויזכר For he remembered כי that בשׂר flesh; המה they רוח a wind הולך that passeth away, ולא and cometh not again. ישׁוב׃ and cometh not again.}%
\verse{כמה How oft ימרוהו did they provoke במדבר him in the wilderness, יעציבוהו grieve בישׁימון׃ him in the desert!}%
\verse{וישׁובו Yea, they turned back וינסו and tempted אל God, וקדושׁ the Holy One ישׂראל of Israel. התוו׃ and limited}%
\verse{לא not זכרו They remembered את ידו his hand, יום the day אשׁר when פדם he delivered מני them from צר׃ the enemy.}%
\verse{אשׁר How שׂם he had wrought במצרים in Egypt, אתותיו his signs ומופתיו and his wonders בשׂדה in the field צען׃ of Zoan:}%
\verse{ויהפך And had turned לדם into blood; יאריהם their rivers ונזליהם and their floods, בל that they could not ישׁתיון׃ drink.}%
\verse{ישׁלח He sent בהם ערב divers sorts of flies ויאכלם among them, which devoured וצפרדע them; and frogs, ותשׁחיתם׃ which destroyed}%
\verse{ויתן He gave לחסיל unto the caterpillar, יבולם also their increase ויגיעם and their labor לארבה׃ unto the locust.}%
\verse{יהרג He destroyed בברד with hail, גפנם their vines ושׁקמותם and their sycamore trees בחנמל׃ with frost.}%
\verse{ויסגר He gave up לברד also to the hail, בעירם their cattle ומקניהם and their flocks לרשׁפים׃ to hot thunderbolts.}%
\verse{ישׁלח He cast בם חרון upon them the fierceness אפו of his anger, עברה wrath, וזעם and indignation, וצרה and trouble, משׁלחת by sending מלאכי angels רעים׃ evil}%
\verse{יפלס He made נתיב a way לאפו to his anger; לא not חשׂך he spared ממות from death, נפשׁם their soul וחיתם their life לדבר over to the pestilence; הסגיר׃ but gave}%
\verse{ויך And smote כל all בכור the firstborn במצרים in Egypt; ראשׁית the chief אונים of strength באהלי in the tabernacles חם׃ of Ham:}%
\verse{ויסע to go forth כצאן like sheep, עמו But made his own people וינהגם and guided כעדר like a flock. במדבר׃ them in the wilderness}%
\verse{וינחם And he led לבטח them on safely, ולא not: פחדו so that they feared ואת אויביהם their enemies. כסה overwhelmed הים׃ but the sea}%
\verse{ויביאם And he brought אל them to גבול the border קדשׁו of his sanctuary, הר mountain, זה this קנתה had purchased. ימינו׃ his right hand}%
\verse{ויגרשׁ He cast out מפניהם also before גוים the heathen ויפילם them, and divided בחבל by line, נחלה them an inheritance וישׁכן to dwell באהליהם in their tents. שׁבטי and made the tribes ישׂראל׃ of Israel}%
\verse{וינסו Yet they tempted וימרו and provoked את אלהים God, עליון the most high ועדותיו his testimonies: לא not שׁמרו׃ and kept}%
\verse{ויסגו But turned back, ויבגדו and dealt unfaithfully כאבותם like their fathers: נהפכו they were turned aside כקשׁת bow. רמיה׃ like a deceitful}%
\verse{ויכעיסוהו For they provoked him to anger בבמותם with their high places, ובפסיליהם with their graven images. יקניאוהו׃ and moved him to jealousy}%
\verse{שׁמע heard אלהים When God ויתעבר he was wroth, וימאס abhorred מאד and greatly בישׂראל׃ Israel:}%
\verse{ויטשׁ So that he forsook משׁכן the tabernacle שׁלו of Shiloh, אהל the tent שׁכן he placed באדם׃ among men;}%
\verse{ויתן And delivered לשׁבי into captivity, עזו his strength ותפארתו and his glory ביד hand. צר׃ into the enemy's}%
\verse{ויסגר He gave לחרב also unto the sword; עמו his people ובנחלתו with his inheritance. התעבר׃ and was wroth}%
\verse{בחוריו their young men; אכלה consumed אשׁ The fire ובתולתיו and their maidens לא were not הוללו׃ given to marriage.}%
\verse{כהניו Their priests בחרב by the sword; נפלו fell ואלמנתיו and their widows לא made no תבכינה׃ lamentation.}%
\verse{ויקץ awaked כישׁן as one out of sleep, אדני Then the Lord כגבור like a mighty man מתרונן that shouteth מיין׃ by reason of wine.}%
\verse{ויך And he smote צריו his enemies אחור in the hinder parts: חרפת reproach. עולם them to a perpetual נתן׃ he put}%
\verse{וימאס Moreover he refused באהל the tabernacle יוסף of Joseph, ובשׁבט the tribe אפרים of Ephraim: לא not בחר׃ and chose}%
\verse{ויבחר But chose את שׁבט the tribe יהודה of Judah, את הר the mount ציון Zion אשׁר which אהב׃ he loved.}%
\verse{ויבן And he built כמו like רמים high מקדשׁו his sanctuary כארץ like the earth יסדה which he hath established לעולם׃ forever.}%
\verse{ויבחר He chose בדוד David עבדו also his servant, ויקחהו and took ממכלאת צאן׃}%
\verse{מאחר עלות the ewes great with young הביאו he brought לרעות him to feed ביעקב Jacob עמו his people, ובישׂראל and Israel נחלתו׃ his inheritance.}%
\verse{וירעם So he fed כתם them according to the integrity לבבו of his heart; ובתבונות them by the skilfulness כפיו of his hands. ינחם׃ and guided}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 79
\verseWithHeading{A Lament for Jerusalem after Its Destruction}{מזמור A Psalm לאסף of Asaph. אלהים O God, באו are come גוים the heathen בנחלתך into thine inheritance; טמאו have they defiled; את היכל temple קדשׁך thy holy שׂמו they have laid את ירושׁלם Jerusalem לעיים׃ on heaps.}%
\verse{נתנו have they given את נבלת The dead bodies עבדיך of thy servants מאכל meat לעוף unto the fowls השׁמים of the heaven, בשׂר the flesh חסידיך of thy saints לחיתו unto the beasts ארץ׃ of the earth.}%
\verse{שׁפכו have they shed דמם Their blood כמים like water סביבות round about ירושׁלם Jerusalem; ואין and none קובר׃ to bury}%
\verse{היינו We are become חרפה a reproach לשׁכנינו to our neighbors, לעג a scorn וקלס and derision לסביבותינו׃ to them that are round about}%
\verse{עד מההוה LORD? תאנף wilt thou be angry לנצח forever? תבער burn כמו like אשׁ fire? קנאתך׃ shall thy jealousy}%
\verse{שׁפך Pour out חמתך thy wrath אל upon הגוים the heathen אשׁר that לא have not ידעוך known ועל thee, and upon ממלכות the kingdoms אשׁר that בשׁמך upon thy name. לא קראו׃ called}%
\verse{כי For אכל they have devoured את יעקב Jacob, ואת נוהו his dwelling place. השׁמו׃ and laid waste}%
\verse{אל not תזכר O remember לנו עונת iniquities: ראשׁנים against us former מהר יקדמונו prevent רחמיך let thy tender mercies כי us: for דלונו we are brought very low. מאד׃ we are brought very low.}%
\verse{עזרנו Help אלהי us, O God ישׁענו of our salvation, על for דבר for כבוד the glory שׁמך of thy name: והצילנו and deliver וכפר us, and purge away על us, and purge away חטאתינו our sins, למען שׁמך׃}%
\verse{למה Wherefore יאמרו say, הגוים should the heathen איה Where אלהיהם their God? יודע let him be known בגיים among the heathen לעינינו in our sight נקמת the revenging דם of the blood עבדיך of thy servants השׁפוך׃ shed.}%
\verse{תבוא come לפניך before אנקת Let the sighing אסיר כגדל thee; according to the greatness זרועך of thy power הותר preserve בני thou those that are appointed תמותה׃ to die;}%
\verse{והשׁב And render לשׁכנינו unto our neighbors שׁבעתים sevenfold אל into חיקם their bosom חרפתם their reproach, אשׁר wherewith חרפוך they have reproached אדני׃ thee, O Lord.}%
\verse{ואנחנו So we עמך thy people וצאן and sheep מרעיתך of thy pasture נודה will give thee thanks לך לעולם forever: לדר to all generations. ודר to all generations. נספר we will show forth תהלתך׃ thy praise}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 80
\verseWithHeading{A Prayer to Restore Israel}{למנצח To the chief Musician אל upon שׁשׁנים עדות Shoshannim-eduth, לאסף of Asaph. מזמור׃ A Psalm רעה O Shepherd ישׂראל of Israel, האזינה Give ear, נהג thou that leadest כצאן like a flock; יוסף Joseph ישׁב thou that dwellest הכרובים the cherubims, הופיעה׃ shine forth.}%
\verse{לפני Before אפרים Ephraim ובנימן and Benjamin ומנשׁה and Manasseh עוררה stir up את גבורתך thy strength, ולכה and come לישׁעתה׃ save}%
\verse{אלהים O God, השׁיבנו Turn us again, והאר to shine; פניך and cause thy face ונושׁעה׃ and we shall be saved.}%
\verse{יהוה O LORD אלהים God צבאות of hosts, עד how long מתי how long עשׁנת wilt thou be angry בתפלת against the prayer עמך׃ of thy people?}%
\verse{האכלתם Thou feedest לחם them with the bread דמעה of tears; ותשׁקמו to drink בדמעות and givest them tears שׁלישׁ׃ in great measure.}%
\verse{תשׂימנו Thou makest מדון us a strife לשׁכנינו unto our neighbors: ואיבינו and our enemies ילעגו׃ laugh}%
\verse{אלהים O God צבאות of hosts, השׁיבנו Turn us again, והאר to shine; פניך and cause thy face ונושׁעה׃ and we shall be saved.}%
\verse{גפן a vine ממצרים תסיע Thou hast brought תגרשׁ thou hast cast out גוים the heathen, ותטעה׃ and planted}%
\verse{פנית Thou preparedst לפניה before ותשׁרשׁ it, and didst cause it to take deep root, שׁרשׁיה it, and didst cause it to take deep root, ותמלא and it filled ארץ׃ the land.}%
\verse{כסו were covered הרים The hills צלה with the shadow וענפיה of it, and the boughs ארזי cedars. אל׃ thereof the goodly}%
\verse{תשׁלח She sent out קצירה her boughs עד unto ים the sea, ואל unto נהר the river. יונקותיה׃ and her branches}%
\verse{למה Why פרצת hast thou broken down גדריה her hedges, וארוה do pluck כל so that all עברי they which pass by דרך׃ the way}%
\verse{יכרסמנה doth waste חזיר The boar מיער out of the wood וזיז it, and the wild beast שׂדי of the field ירענה׃ doth devour}%
\verse{אלהים thee, O God צבאות of hosts: שׁוב Return, נא we beseech הבט look down משׁמים from heaven, וראה and behold, ופקד and visit גפן vine; זאת׃ this}%
\verse{וכנה And the vineyard אשׁר which נטעה hath planted, ימינך thy right hand ועל בן and the branch אמצתה׃ thou madest strong}%
\verse{שׂרפה burned באשׁ with fire, כסוחה cut down: מגערת at the rebuke פניך of thy countenance. יאבדו׃ they perish}%
\verse{תהי be ידך Let thy hand על upon אישׁ the man ימינך of thy right hand, על upon בן the son אדם of man אמצת׃ thou madest strong}%
\verse{ולא So will not נסוג we go back ממך from תחינו thee: quicken ובשׁמך upon thy name. נקרא׃ us, and we will call}%
\verse{יהוה O LORD אלהים God צבאות of hosts, השׁיבנו Turn us again, האר to shine; פניך cause thy face ונושׁעה׃ and we shall be saved.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 81
\verseWithHeading{An Appeal from God to Israel}{למנצח To the chief Musician על upon הגתית Gittith, לאסף׃ of Asaph. הרנינו Sing aloud לאלהים unto God עוזנו our strength: הריעו make a joyful noise לאלהי unto the God יעקב׃ of Jacob.}%
\verse{שׂאו Take זמרה a psalm, ותנו and bring תף hither the timbrel, כנור harp נעים the pleasant עם with נבל׃ the psaltery.}%
\verse{תקעו Blow up בחדשׁ in the new moon, שׁופר the trumpet בכסה in the time appointed, ליום day. חגנו׃ on our solemn feast}%
\verse{כי For חק a statute לישׂראל for Israel, הוא this משׁפט a law לאלהי of the God יעקב׃ of Jacob.}%
\verse{עדות a testimony, ביהוסף in Joseph שׂמו This he ordained בצאתו when he went out על through ארץ the land מצרים of Egypt: שׂפת a language לא not. ידעתי I understood אשׁמע׃ I heard}%
\verse{הסירותי I removed מסבל from the burden: שׁכמו his shoulder כפיו his hands מדוד from the pots. תעברנה׃ were delivered}%
\verse{בצרה in trouble, קראת Thou calledst ואחלצך and I delivered אענך thee; I answered בסתר thee in the secret place רעם of thunder: אבחנך I proved על thee at מי the waters מריבה of Meribah. סלה׃ Selah.}%
\verse{שׁמע Hear, עמי O my people, ואעידה and I will testify בך ישׂראל unto thee: O Israel, אם if תשׁמע׃ thou wilt hearken}%
\verse{לא There shall no יהיה be בך אל god זר strange ולא in thee; neither תשׁתחוה shalt thou worship לאל god. נכר׃ any strange}%
\verse{אנכי I יהוה the LORD אלהיך thy God, המעלך which brought מארץ thee out of the land מצרים of Egypt: הרחב wide, פיך open thy mouth ואמלאהו׃ and I will fill}%
\verse{ולא would not שׁמע hearken עמי But my people לקולי to my voice; וישׂראל and Israel לא none אבה׃ would}%
\verse{ואשׁלחהו So I gave them up בשׁרירות lust: לבם unto their own hearts' ילכו they walked במועצותיהם׃ in their own counsels.}%
\verse{לו Oh that עמי my people שׁמע had hearkened לי ישׂראל unto me, Israel בדרכי in my ways! יהלכו׃ had walked}%
\verse{כמעט I should soon אויביהם their enemies, אכניע have subdued ועל against צריהם their adversaries. אשׁיב and turned ידי׃ my hand}%
\verse{משׂנאי The haters יהוה of the LORD יכחשׁו should have submitted לו ויהי should have endured עתם themselves unto him: but their time לעולם׃ forever.}%
\verse{ויאכילהו He should have fed מחלב them also with the finest חטה of the wheat: ומצור out of the rock דבשׁ and with honey אשׂביעך׃ should I have satisfied}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 82
\verseWithHeading{God Commands Justice}{מזמור A Psalm לאסף of Asaph. אלהים God נצב standeth בעדת in the congregation אל of the mighty; בקרב among אלהים the gods. ישׁפט׃ he judgeth}%
\verse{עד מתישׁפטו will ye judge עול unjustly, ופני the persons רשׁעים of the wicked? תשׂאו and accept סלה׃ Selah.}%
\verse{שׁפטו Defend דל the poor ויתום and fatherless: עני to the afflicted ורשׁ and needy. הצדיקו׃ do justice}%
\verse{פלטו Deliver דל the poor ואביון and needy: מיד out of the hand רשׁעים of the wicked. הצילו׃ rid}%
\verse{לא not, ידעו They know ולא neither יבינו will they understand; בחשׁכה in darkness: יתהלכו they walk on ימוטו are out of course. כל all מוסדי ארץ׃ of the earth}%
\verse{אני I אמרתי have said, אלהים Ye gods; אתם of you ובני children עליון of the most High. כלכם׃ and all}%
\verse{אכן But כאדם like men, תמותון ye shall die וכאחד like one השׂרים of the princes. תפלו׃ and fall}%
\verse{קומה Arise, אלהים O God, שׁפטה judge הארץ the earth: כי for אתה thou תנחל shalt inherit בכל all הגוים׃ nations.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 83
\verseWithHeading{A Request to Act against Israel’s Neighbors}{שׁיר A Song מזמור Psalm לאסף׃ of Asaph. אלהים O God: אל hold not thy peace, דמי לך אל and be not תחרשׁ hold not thy peace, ואל תשׁקט still, אל׃ O God.}%
\verse{כי For, הנה lo, אויביך thine enemies יהמיון make a tumult: ומשׂנאיך and they that hate נשׂאו thee have lifted up ראשׁ׃ the head.}%
\verse{על against עמך thy people, יערימו They have taken crafty סוד counsel ויתיעצו and consulted על against צפוניך׃ thy hidden ones.}%
\verse{אמרו They have said, לכו Come, ונכחידם and let us cut them off מגוי from a nation; ולא may be no יזכר in remembrance. שׁם that the name ישׂראל of Israel עוד׃ more}%
\verse{כי For נועצו they have consulted together לב with one consent: יחדו with one consent: עליך against ברית they are confederate יכרתו׃ they are confederate}%
\verse{אהלי The tabernacles אדום of Edom, וישׁמעאלים and the Ishmaelites; מואב of Moab, והגרים׃ and the Hagarenes;}%
\verse{גבל Gebal, ועמון and Ammon, ועמלק and Amalek; פלשׁת the Philistines עם with ישׁבי the inhabitants צור׃ of Tyre;}%
\verse{גם also אשׁור Assur נלוה is joined עמם with היו them: they have זרוע helped לבני the children לוט of Lot. סלה׃ Selah.}%
\verse{עשׂה Do להם כמדין unto them as the Midianites; כסיסרא as Sisera, כיבין as Jabin, בנחל at the brook קישׁון׃ of Kison:}%
\verse{נשׁמדו perished בעין דאר at Endor: היו they became דמן dung לאדמה׃ for the earth.}%
\verse{שׁיתמו Make נדיבמו their nobles כערב like Oreb, וכזאב and like Zeeb: וכזבח as Zebah, וכצלמנע and as Zalmunna: כל yea, all נסיכמו׃ their princes}%
\verse{אשׁר Who אמרו said, נירשׁה Let us take לנו את נאות to ourselves the houses אלהים׃ of God}%
\verse{אלהי O my God, שׁיתמו make כגלגל them like a wheel; כקשׁ as the stubble לפני before רוח׃ the wind.}%
\verse{כאשׁ As the fire תבער burneth יער a wood, וכלהבה and as the flame תלהט setteth the mountains on fire; הרים׃ setteth the mountains on fire;}%
\verse{כן So תרדפם persecute בסערך them with thy tempest, ובסופתך with thy storm. תבהלם׃ and make them afraid}%
\verse{מלא Fill פניהם their faces קלון with shame; ויבקשׁו that they may seek שׁמך thy name, יהוה׃ O LORD.}%
\verse{יבשׁו Let them be confounded ויבהלו and troubled עדי forever; עד forever; ויחפרו yea, let them be put to shame, ויאבדו׃ and perish:}%
\verse{וידעו That may know כי that אתה thou, שׁמך whose name יהוה JEHOVAH, לבדך alone עליון the most high על over כל all הארץ׃ the earth.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 84
\verseWithHeading{The Joy of Worshiping in the Temple}{למנצח To the chief Musician על upon הגתית Gittith, לבני for the sons קרח of Korah. מזמור׃ A Psalm מה How ידידות amiable משׁכנותיך thy tabernacles, יהוה O LORD צבאות׃ of hosts!}%
\verse{נכספה longeth, וגם yea even כלתה fainteth נפשׁי My soul לחצרות for the courts יהוה of the LORD: לבי my heart ובשׂרי and my flesh ירננו crieth out אל for אל God. חי׃ the living}%
\verse{גם Yea, צפור the sparrow מצאה hath found בית a house, ודרור and the swallow קן a nest לה אשׁר for herself, where שׁתה she may lay אפרחיה her young, את מזבחותיך thine altars, יהוה O LORD צבאות of hosts, מלכי my King, ואלהי׃ and my God.}%
\verse{אשׁרי Blessed יושׁבי they that dwell ביתך in thy house: עוד they will be still יהללוך praising סלה׃ thee. Selah.}%
\verse{אשׁרי Blessed אדם the man עוז whose strength לו בך מסלות the ways בלבבם׃ in thee; in whose heart}%
\verse{עברי passing בעמק through the valley הבכא of Baca מעין it a well; ישׁיתוהו make גם also ברכות the pools. יעטה filleth מורה׃ the rain}%
\verse{ילכו They go מחיל from strength אל to חיל strength, יראה appeareth אל before אלהים God. בציון׃ in Zion}%
\verse{יהוה O LORD אלהים God צבאות of hosts, שׁמעה hear תפלתי my prayer: האזינה give ear, אלהי O God יעקב of Jacob. סלה׃ Selah.}%
\verse{מגננו our shield, ראה Behold, אלהים O God והבט and look upon פני the face משׁיחך׃ of thine anointed.}%
\verse{כי For טוב better יום a day בחצריך in thy courts מאלף than a thousand. בחרתי I had rather הסתופף be a doorkeeper בבית in the house אלהי of my God, מדור than to dwell באהלי in the tents רשׁע׃ of wickedness.}%
\verse{כי For שׁמשׁ a sun ומגן and shield: יהוה the LORD אלהים God חן grace וכבוד and glory: יתן will give יהוה the LORD לא no ימנע will he withhold טוב good להלכים from them that walk בתמים׃ uprightly.}%
\verse{יהוה O LORD צבאות of hosts, אשׁרי blessed אדם the man בטח׃ that trusteth}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 85
\verseWithHeading{Hope in God’s Future Help}{למנצח To the chief Musician, לבני for the sons קרח of Korah. מזמור׃ A Psalm רצית thou hast been favorable יהוה LORD, ארצך unto thy land: שׁבת thou hast brought back שׁבות the captivity יעקב׃ of Jacob.}%
\verse{נשׂאת Thou hast forgiven עון the iniquity עמך of thy people, כסית thou hast covered כל all חטאתם their sin. סלה׃ Selah.}%
\verse{אספת Thou hast taken away כל all עברתך thy wrath: השׁיבות thou hast turned מחרון from the fierceness אפך׃ of thine anger.}%
\verse{שׁובנו Turn אלהי us, O God ישׁענו of our salvation, והפר us to cease. כעסך and cause thine anger עמנו׃ toward}%
\verse{הלעולם with us forever? תאנף Wilt thou be angry בנו תמשׁך wilt thou draw out אפך thine anger לדר to all generations? ודר׃ to all generations?}%
\verse{הלא not אתה Wilt thou תשׁוב us again: תחינו revive ועמך that thy people ישׂמחו׃ may rejoice}%
\verse{הראנו Show יהוה O LORD, חסדך us thy mercy, וישׁעך us thy salvation. תתן׃ and grant}%
\verse{אשׁמעה I will hear מה what ידבר will speak: האל God יהוה the LORD כי for ידבר he will speak שׁלום peace אל unto עמו his people, ואל and to חסידיו his saints: ואל but let them not ישׁובו turn again לכסלה׃ to folly.}%
\verse{אך Surely קרוב nigh ליראיו them that fear ישׁעו his salvation לשׁכן may dwell כבוד him; that glory בארצנו׃ in our land.}%
\verse{חסד Mercy ואמת and truth נפגשׁו are met together; צדק righteousness ושׁלום and peace נשׁקו׃ have kissed}%
\verse{אמת Truth מארץ out of the earth; תצמח shall spring וצדק and righteousness משׁמים from heaven. נשׁקף׃ shall look down}%
\verse{גם Yea, יהוה the LORD יתן shall give הטוב good; וארצנו and our land תתן shall yield יבולה׃ her increase.}%
\verse{צדק Righteousness לפניו before יהלך shall go וישׂם him; and shall set לדרך in the way פעמיו׃ of his steps.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 86
\verseWithHeading{A Prayer for Help against Ruthless Men}{תפלה A Prayer לדוד of David. הטה Bow down יהוה O LORD אזנך thine ear, ענני hear כי me: for עני poor ואביון and needy. אני׃ I}%
\verse{שׁמרה Preserve נפשׁי my soul; כי for חסיד holy: אני I הושׁע save עבדך thy servant אתה O thou אלהי my God, הבוטח that trusteth אליך׃ in thee.}%
\verse{חנני Be merciful אדני unto me, O Lord: כי for אליך unto אקרא I cry כל thee daily. היום׃ thee daily.}%
\verse{שׂמח Rejoice נפשׁ the soul עבדך of thy servant: כי for אליך unto אדני thee, O Lord, נפשׁי my soul. אשׂא׃ do I lift up}%
\verse{כי For אתה thou, אדני Lord, טוב good, וסלח and ready to forgive; ורב and plenteous חסד in mercy לכל unto all קראיך׃ them that call upon}%
\verse{האזינה Give ear, יהוה O LORD, תפלתי unto my prayer; והקשׁיבה and attend בקול to the voice תחנונותי׃ of my supplications.}%
\verse{ביום In the day צרתי of my trouble אקראך I will call upon כי thee: for תענני׃ thou wilt answer}%
\verse{אין none כמוך like unto thee, באלהים Among the gods אדני O Lord; ואין neither כמעשׂיך׃ like unto thy works.}%
\verse{כל All גוים nations אשׁר whom עשׂית thou hast made יבואו shall come וישׁתחוו and worship לפניך before אדני thee, O Lord; ויכבדו and shall glorify לשׁמך׃ thy name.}%
\verse{כי For גדול great, אתה thou ועשׂה and doest נפלאות wondrous things: אתה thou אלהים God לבדך׃ alone.}%
\verse{הורני Teach יהוה O LORD; דרכך me thy way, אהלך I will walk באמתך in thy truth: יחד unite לבבי my heart ליראה to fear שׁמך׃ thy name.}%
\verse{אודך I will praise אדני thee, O Lord אלהי my God, בכל with all לבבי my heart: ואכבדה and I will glorify שׁמך thy name לעולם׃ forevermore.}%
\verse{כי For חסדך thy mercy גדול great עלי toward והצלת me: and thou hast delivered נפשׁי my soul משׁאול hell. תחתיה׃ from the lowest}%
\verse{אלהים O God, זדים the proud קמו are risen עלי against ועדת me, and the assemblies עריצים of violent בקשׁו have sought after נפשׁי my soul; ולא and have not שׂמוך set לנגדם׃ thee before}%
\verse{ואתה But thou, אדני O Lord, אל a God רחום full of compassion, וחנון and gracious, ארך longsuffering, אפים longsuffering, ורב and plenteous חסד in mercy ואמת׃ and truth.}%
\verse{פנה O turn אלי unto וחנני me, and have mercy תנה upon me; give עזך thy strength לעבדך unto thy servant, והושׁיעה and save לבן the son אמתך׃ of thine handmaid.}%
\verse{עשׂה עמיות me a token לטובה for good; ויראו me may see שׂנאי that they which hate ויבשׁו and be ashamed: כי because אתה thou, יהוה LORD, עזרתני hast helped ונחמתני׃ me, and comforted}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 87
\verseWithHeading{Foreign Nations Come to Worship in Jerusalem}{לבני for the sons קרח of Korah. מזמור A Psalm שׁיר Song יסודתו His foundation בהררי mountains. קדשׁ׃ in the holy}%
\verse{אהב loveth יהוה The LORD שׁערי the gates ציון of Zion מכל more than all משׁכנות the dwellings יעקב׃ of Jacob.}%
\verse{נכבדות Glorious things מדבר are spoken בך עיר of thee, O city האלהים of God. סלה׃ Selah.}%
\verse{אזכיר I will make mention רהב of Rahab ובבל and Babylon לידעי to them that know הנה me: behold פלשׁת Philistia, וצור and Tyre, עם with כושׁ Ethiopia; זה this ילד was born שׁם׃ there.}%
\verse{ולציון And of Zion יאמר it shall be said, אישׁ This ואישׁ and that man ילד was born בה והוא himself יכוננה shall establish עליון׃ in her: and the highest}%
\verse{יהוה The LORD יספר shall count, בכתוב when he writeth up עמים the people, זה this ילד was born שׁם there. סלה׃ Selah.}%
\verse{ושׁרים As well the singers כחללים as the players on instruments כל all מעיני׃ my springs}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 88
\verseWithHeading{A Prayer for Help in Despair}{שׁיר A Song מזמור Psalm לבני for the sons קרח of Korah, למנצח to the chief Musician על upon מחלת Mahalath לענות Leannoth, משׂכיל Maschil להימן of Heman האזרחי׃ the Ezrahite. יהוה O LORD אלהי God ישׁועתי of my salvation, יום day צעקתי I have cried בלילה night נגדך׃ before}%
\verse{תבוא come לפניך before תפלתי Let my prayer הטה thee: incline אזנך thine ear לרנתי׃ unto my cry;}%
\verse{כי For שׂבעה is full ברעות of troubles: נפשׁי my soul וחיי and my life לשׁאול unto the grave. הגיעו׃ draweth nigh}%
\verse{נחשׁבתי I am counted עם with יורדי them that go down בור into the pit: הייתי I am כגבר as a man אין no איל׃ strength:}%
\verse{במתים among the dead, חפשׁי Free כמו like חללים the slain שׁכבי that lie קבר in the grave, אשׁר whom לא no זכרתם thou rememberest עוד more: והמה and they מידך from thy hand. נגזרו׃ are cut off}%
\verse{שׁתני Thou hast laid בבור pit, תחתיות me in the lowest במחשׁכים in darkness, במצלות׃ in the deeps.}%
\verse{עלי upon סמכה lieth hard חמתך Thy wrath וכל with all משׁבריך thy waves. ענית me, and thou hast afflicted סלה׃ Selah.}%
\verse{הרחקת Thou hast put away מידעי mine acquaintance ממני far from שׁתני me; thou hast made תועבות me an abomination למו כלא unto them: shut up, ולא and I cannot אצא׃ come forth.}%
\verse{עיני Mine eye דאבה mourneth מני by reason of עני affliction: קראתיך I have called יהוה LORD, בכל daily יום daily שׁטחתי upon thee, I have stretched out אליך unto thee. כפי׃ my hands}%
\verse{הלמתים to the dead? תעשׂה Wilt thou show פלא wonders אם רפאים shall the dead יקומו arise יודוך praise סלה׃ thee? Selah.}%
\verse{היספר be declared בקבר in the grave? חסדך Shall thy lovingkindness אמונתך thy faithfulness באבדון׃ in destruction?}%
\verse{היודע be known בחשׁך in the dark? פלאך Shall thy wonders וצדקתך and thy righteousness בארץ in the land נשׁיה׃ of forgetfulness?}%
\verse{ואני thee have I אליך But unto יהוה O LORD; שׁועתי cried, ובבקר and in the morning תפלתי shall my prayer תקדמך׃ prevent}%
\verse{למה why יהוה LORD, תזנח castest thou off נפשׁי my soul? תסתיר hidest פניך thou thy face ממני׃ from}%
\verse{עני afflicted אני I וגוע and ready to die מנער from youth up: נשׂאתי I suffer אמיך thy terrors אפונה׃ I am distracted.}%
\verse{עלי over עברו goeth חרוניך Thy fierce wrath בעותיך me; thy terrors צמתותני׃ have cut me off.}%
\verse{סבוני They came round about כמים like water; כל me daily היום me daily הקיפו they compassed me about עלי they compassed me about יחד׃ together.}%
\verse{הרחקת hast thou put far ממני from אהב Lover ורע and friend מידעי me, mine acquaintance מחשׁך׃ into darkness.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 89
\verseWithHeading{Remembering the Covenant with David, and Sorrow for Lost Blessings}{משׂכיל Maschil לאיתן of Ethan האזרחי׃ the Ezrahite. חסדי of the mercies יהוה of the LORD עולם forever: אשׁירה I will sing לדר to all generations. ודר to all generations. אודיע will I make known אמונתך thy faithfulness בפי׃ with my mouth}%
\verse{כי For אמרתי I have said, עולם forever: חסד Mercy יבנה shall be built up שׁמים in the very heavens. תכן shalt thou establish אמונתך׃ thy faithfulness}%
\verse{כרתי I have made ברית a covenant לבחירי with my chosen, נשׁבעתי I have sworn לדוד unto David עבדי׃ my servant,}%
\verse{עד forever, עולם forever, אכין will I establish זרעך Thy seed ובניתי and build up לדר to all generations. ודור >> r="#000000">to all generations. כסאך thy throne סלה׃ Selah.}%
\verse{ויודו shall praise שׁמים And the heavens פלאך thy wonders, יהוה O LORD: אף also אמונתך thy faithfulness בקהל in the congregation קדשׁים׃ of the saints.}%
\verse{כי For מי who בשׁחק in the heaven יערך can be compared ליהוה unto the LORD? ידמה can be likened ליהוה unto the LORD? בבני among the sons אלים׃ of the mighty}%
\verse{אל God נערץ to be feared בסוד in the assembly קדשׁים of the saints, רבה is greatly ונורא and to be had in reverence על of כל all סביביו׃ about}%
\verse{יהוה O LORD אלהי God צבאות of hosts, מי who כמוך like unto thee? חסין a strong יה LORD ואמונתך or to thy faithfulness סביבותיך׃ round about}%
\verse{אתה Thou מושׁל rulest בגאות the raging הים of the sea: בשׂוא thereof arise, גליו when the waves אתה thou תשׁבחם׃ stillest}%
\verse{אתה Thou דכאת hast broken כחלל in pieces, as one that is slain; רהב Rahab בזרוע arm. עזך with thy strong פזרת thou hast scattered אויביך׃ thine enemies}%
\verse{לך שׁמים The heavens אף also לך ארץ thine, the earth תבל thine: the world ומלאה and the fullness אתה thereof, thou יסדתם׃ hast founded}%
\verse{צפון The north וימין and the south אתה thou בראתם hast created תבור them: Tabor וחרמון and Hermon בשׁמך in thy name. ירננו׃ shall rejoice}%
\verse{לך זרוע arm: עם גבורה Thou hast a mighty תעז strong ידך is thy hand, תרום high ימינך׃ is thy right hand.}%
\verse{צדק Justice ומשׁפט and judgment מכון the habitation כסאך of thy throne: חסד mercy ואמת and truth יקדמו shall go פניך׃ before thy face.}%
\verse{אשׁרי Blessed העם the people יודעי that know תרועה the joyful sound: יהוה O LORD, באור in the light פניך of thy countenance. יהלכון׃ they shall walk,}%
\verse{בשׁמך In thy name יגילון shall they rejoice כל all היום the day: ובצדקתך and in thy righteousness ירומו׃ shall they be exalted.}%
\verse{כי For תפארת the glory עזמו of their strength: אתה thou וברצנך and in thy favor תרים shall be exalted. קרננו׃ our horn}%
\verse{כי For ליהוה the LORD מגננו our defense; ולקדושׁ and the Holy One ישׂראל of Israel מלכנו׃ our king.}%
\verse{אז Then דברת thou spakest בחזון in vision לחסידיך to thy holy one, ותאמר and saidst, שׁויתי I have laid עזר help על upon גבור mighty; הרימותי I have exalted בחור chosen מעם׃ out of the people.}%
\verse{מצאתי I have found דוד David עבדי my servant; בשׁמן oil קדשׁי with my holy משׁחתיו׃ have I anointed}%
\verse{אשׁר whom ידי my hand תכון shall be established: עמו With אף also זרועי mine arm תאמצנו׃ shall strengthen}%
\verse{לא shall not ישׁא exact אויב The enemy בו ובן the son עולה of wickedness לא upon him; nor יעננו׃ afflict}%
\verse{וכתותי And I will beat down מפניו before his face, צריו his foes ומשׂנאיו them that hate אגוף׃ and plague}%
\verse{ואמונתי But my faithfulness וחסדי and my mercy עמו with ובשׁמי him: and in my name תרום be exalted. קרנו׃ shall his horn}%
\verse{ושׂמתי I will set בים also in the sea, ידו his hand ובנהרות in the rivers. ימינו׃ and his right hand}%
\verse{הוא He יקראני shall cry אבי my father, אתה unto me, Thou אלי my God, וצור and the rock ישׁועתי׃ of my salvation.}%
\verse{אף Also אני I בכור him firstborn, אתנהו will make עליון higher למלכי than the kings ארץ׃ of the earth.}%
\verse{לעולם for him forevermore, אשׁמור will I keep לו חסדי My mercy ובריתי and my covenant נאמנת׃ shall stand fast}%
\verse{ושׂמתי also will I make לעד forever, זרעו His seed וכסאו and his throne כימי as the days שׁמים׃ of heaven.}%
\verse{אם If יעזבו forsake בניו his children תורתי my law, ובמשׁפטי in my judgments; לא not ילכון׃ and walk}%
\verse{אם If חקתי my statutes, יחללו they break ומצותי my commandments; לא not ישׁמרו׃ and keep}%
\verse{ופקדתי Then will I visit בשׁבט with the rod, פשׁעם their transgression ובנגעים with stripes. עונם׃ and their iniquity}%
\verse{וחסדי Nevertheless my lovingkindness לא will I not אפיר utterly take מעמו from ולא him, nor אשׁקר suffer my faithfulness to fail. באמונתי׃ suffer my faithfulness to fail.}%
\verse{לא will I not אחלל break, בריתי My covenant ומוצא the thing that is gone out שׂפתי of my lips. לא nor אשׁנה׃ alter}%
\verse{אחת Once נשׁבעתי have I sworn בקדשׁי by my holiness אם that I will not לדוד unto David. אכזב׃ lie}%
\verse{זרעו His seed לעולם forever, יהיה shall endure וכסאו and his throne כשׁמשׁ as the sun נגדי׃ before}%
\verse{כירח as the moon, יכון It shall be established עולם forever ועד witness בשׁחק in heaven. נאמן and a faithful סלה׃ Selah.}%
\verse{ואתה But thou זנחת hast cast off ותמאס and abhorred, התעברת thou hast been wroth עם with משׁיחך׃ thine anointed.}%
\verse{נארתה Thou hast made void ברית the covenant עבדך of thy servant: חללת thou hast profaned לארץ to the ground. נזרו׃ his crown}%
\verse{פרצת Thou hast broken down כל all גדרתיו his hedges; שׂמת thou hast brought מבצריו his strongholds מחתה׃ to ruin.}%
\verse{שׁסהו spoil כל All עברי that pass by דרך the way היה him: he is חרפה a reproach לשׁכניו׃ to his neighbors.}%
\verse{הרימות Thou hast set up ימין the right hand צריו of his adversaries; השׂמחת to rejoice. כל thou hast made all אויביו׃ his enemies}%
\verse{אף Thou hast also תשׁיב turned צור the edge חרבו of his sword, ולא and hast not הקימתו made him to stand במלחמה׃ in the battle.}%
\verse{השׁבת to cease, מטהרו וכסאו his throne לארץ down to the ground. מגרתה׃ and cast}%
\verse{הקצרת hast thou shortened: ימי The days עלומיו of his youth העטית thou hast covered עליו thou hast covered בושׁה him with shame. סלה׃ Selah.}%
\verse{עד מההוה LORD? תסתר wilt thou hide thyself לנצח forever? תבער burn כמו like אשׁ fire? חמתך׃ shall thy wrath}%
\verse{זכר Remember אני my מה how חלד short על time is: wherefore מה time is: wherefore שׁוא in vain? בראת hast thou made כל all בני men אדם׃ men}%
\verse{מי What גבר man יחיה liveth, ולא and shall not יראה see מות death? ימלט shall he deliver נפשׁו his soul מיד from the hand שׁאול of the grave? סלה׃ Selah.}%
\verse{איה where חסדיך lovingkindnesses, הראשׁנים thy former אדני Lord, נשׁבעת thou swarest לדוד unto David באמונתך׃ in thy truth?}%
\verse{זכר Remember, אדני Lord, חרפת the reproach עבדיך of thy servants; שׂאתי I do bear בחיקי in my bosom כל all רבים the mighty עמים׃ people;}%
\verse{אשׁר Wherewith חרפו have reproached, אויביך thine enemies יהוה O LORD; אשׁר wherewith חרפו they have reproached עקבות the footsteps משׁיחך׃ of thine anointed.}%
\verse{ברוך Blessed יהוה the LORD לעולם forevermore. אמן Amen, ואמן׃ and Amen.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 90
\verseWithHeading{God’s Eternity and Human Frailty}{תפלה A Prayer למשׁה of Moses אישׁ the man האלהים of God. אדני Lord, מעון our dwelling place אתה thou היית hast been לנו בדר in all generations. ודר׃ in all generations.}%
\verse{בטרם Before הרים the mountains ילדו were brought forth, ותחולל or ever thou hadst formed ארץ the earth ותבל and the world, ומעולם even from everlasting עד to עולם everlasting, אתה thou אל׃ God.}%
\verse{תשׁב Thou turnest אנושׁ man עד to דכא destruction; ותאמר and sayest, שׁובו Return, בני ye children אדם׃ of men.}%
\verse{כי For אלף a thousand שׁנים years בעיניך in thy sight כיום as yesterday אתמול as yesterday כי when יעבר it is past, ואשׁמורה and a watch בלילה׃ in the night.}%
\verse{זרמתם Thou carriest them away as with a flood; שׁנה a sleep: יהיו they are בבקר in the morning כחציר like grass יחלף׃ groweth up.}%
\verse{בבקר In the morning יציץ it flourisheth, וחלף and groweth up; לערב in the evening ימולל it is cut down, ויבשׁ׃ and withereth.}%
\verse{כי For כלינו we are consumed באפך by thine anger, ובחמתך and by thy wrath נבהלנו׃ are we troubled.}%
\verse{שׁת Thou hast set עונתינו our iniquities לנגדך before עלמנו thee, our secret למאור in the light פניך׃ of thy countenance.}%
\verse{כי For כל all ימינו our days פנו are passed away בעברתך in thy wrath: כלינו we spend שׁנינו our years כמו as הגה׃ a tale}%
\verse{ימי The days שׁנותינו of our years בהם שׁבעים threescore years and ten; שׁנה threescore years and ten; ואם and if בגבורת by reason of strength שׁמונים fourscore שׁנה years, ורהבם yet their strength עמל labor ואון and sorrow; כי for גז cut off, חישׁ it is soon ונעפה׃ and we fly away.}%
\verse{מי Who יודע knoweth עז the power אפך of thine anger? וכיראתך even according to thy fear, עברתך׃ thy wrath.}%
\verse{למנות to number ימינו our days, כן So הודע teach ונבא that we may apply לבב hearts חכמה׃ unto wisdom.}%
\verse{שׁובה Return, יהוה O LORD, עד how long? מתי how long? והנחם and let it repent על thee concerning עבדיך׃ thy servants.}%
\verse{שׂבענו O satisfy בבקר us early חסדך with thy mercy; ונרננה that we may rejoice ונשׂמחה and be glad בכל all ימינו׃ our days.}%
\verse{שׂמחנו Make us glad כימות according to the days עניתנו thou hast afflicted שׁנות us, the years ראינו we have seen רעה׃ evil.}%
\verse{יראה appear אל unto עבדיך thy servants, פעלך Let thy work והדרך and thy glory על unto בניהם׃ their children.}%
\verse{ויהי be נעם And let the beauty אדני of the LORD אלהינו our God עלינו upon ומעשׂה thou the work ידינו of our hands כוננה us: and establish עלינו upon ומעשׂה us; yea, the work ידינו of our hands כוננהו׃ establish}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 91
\verseWithHeading{God’s Protection in Times of Crisis}{ישׁב He that dwelleth בסתר in the secret place עליון of the most High בצל under the shadow שׁדי of the Almighty. יתלונן׃ shall abide}%
\verse{אמר I will say ליהוה of the LORD, מחסי my refuge ומצודתי and my fortress: אלהי my God; אבטח׃ in him will I trust.}%
\verse{כי Surely הוא he יצילך shall deliver מפח thee from the snare יקושׁ of the fowler, מדבר הוות׃}%
\verse{באברתו thee with his feathers, יסך He shall cover לך ותחת and under כנפיו his wings תחסה shalt thou trust: צנה shield וסחרה and buckler. אמתו׃ his truth}%
\verse{לא Thou shalt not תירא be afraid מפחד for the terror לילה by night; מחץ for the arrow יעוף flieth יומם׃ by day;}%
\verse{מדבר באפל in darkness; יהלך walketh מקטב for the destruction ישׁוד wasteth צהרים׃ at noonday.}%
\verse{יפל shall fall מצדך at thy side, אלף A thousand ורבבה and ten thousand מימינך at thy right hand; אליך come nigh לא it shall not יגשׁ׃ come nigh}%
\verse{רק Only בעיניך with thine eyes תביט shalt thou behold ושׁלמת the reward רשׁעים of the wicked. תראה׃ and see}%
\verse{כי Because אתה thou יהוה the LORD, מחסי my refuge, עליון the most High, שׂמת hast made מעונך׃ thy habitation;}%
\verse{לא There shall no תאנה befall אליך befall רעה evil ונגע shall any plague לא thee, neither יקרב come nigh באהלך׃ thy dwelling.}%
\verse{כי For מלאכיו he shall give his angels יצוה charge לך לשׁמרך over thee, to keep בכל thee in all דרכיך׃ thy ways.}%
\verse{על in כפים hands, ישׂאונך They shall bear thee up פן lest תגף thou dash באבן against a stone. רגלך׃ thy foot}%
\verse{על upon שׁחל the lion ופתן and adder: תדרך Thou shalt tread תרמס shalt thou trample under feet כפיר the young lion ותנין׃ and the dragon}%
\verse{כי Because בי חשׁק he hath set his love ואפלטהו upon me, therefore will I deliver אשׂגבהו him: I will set him on high, כי because ידע he hath known שׁמי׃ my name.}%
\verse{יקראני He shall call upon ואענהו me, and I will answer עמו with אנכי him: I בצרה him in trouble; אחלצהו I will deliver ואכבדהו׃ him, and honor}%
\verse{ארך With long ימים life אשׂביעהו will I satisfy ואראהו him, and show בישׁועתי׃ him my salvation.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 92
\verseWithHeading{Thanksgiving to Adonai for Victory}{מזמור A Psalm שׁיר Song ליום day. השׁבת׃ for the sabbath טוב good להדות to give thanks ליהוה unto the LORD, ולזמר and to sing praises לשׁמך unto thy name, עליון׃ O most High:}%
\verse{להגיד To show forth בבקר in the morning, חסדך thy lovingkindness ואמונתך and thy faithfulness בלילות׃ every night,}%
\verse{עלי Upon עשׂור an instrument of ten strings, ועלי and upon נבל the psaltery; עלי with הגיון a solemn sound. בכנור׃ upon the harp}%
\verse{כי For שׂמחתני hast made me glad יהוה thou, LORD, בפעלך through thy work: במעשׂי in the works ידיך of thy hands. ארנן׃ I will triumph}%
\verse{מה how גדלו great מעשׂיך are thy works! יהוה O LORD, מאד are very עמקו deep. מחשׁבתיך׃ thy thoughts}%
\verse{אישׁ man בער A brutish לא not; ידע knoweth וכסיל doth a fool לא neither יבין understand את זאת׃ this.}%
\verse{בפרח spring רשׁעים When the wicked כמו as עשׂב the grass, ויציצו do flourish; כל and when all פעלי the workers און of iniquity להשׁמדם that they shall be destroyed עדי forever: עד׃ forever:}%
\verse{ואתה But thou, מרום high לעלם forevermore. יהוה׃ LORD,}%
\verse{כי For, הנה lo, איביך thine enemies, יהוה O LORD, כי for, הנה lo, איביך thine enemies יאבדו shall perish; יתפרדו shall be scattered. כל all פעלי the workers און׃ of iniquity}%
\verse{ותרם shalt thou exalt כראים like a unicorn: קרני >> r="#000000">But my horn בלתי I shall be anointed בשׁמן oil. רענן׃ with fresh}%
\verse{ותבט also shall see עיני Mine eye בשׁורי on mine enemies, בקמים that rise up עלי against מרעים of the wicked תשׁמענה shall hear אזני׃ mine ears}%
\verse{צדיק The righteous כתמר like the palm tree: יפרח shall flourish כארז like a cedar בלבנון in Lebanon. ישׂגה׃ he shall grow}%
\verse{שׁתולים Those that be planted בבית in the house יהוה of the LORD בחצרות in the courts אלהינו of our God. יפריחו׃ shall flourish}%
\verse{עוד They shall still ינובון bring forth fruit בשׂיבה in old age; דשׁנים fat ורעננים and flourishing; יהיו׃ they shall be}%
\verse{להגיד To show כי that ישׁר upright: יהוה the LORD צורי my rock, ולא and no עלתה׃ unrighteousness}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 93
\verseWithHeading{Adonai Is King Over All the Earth}{יהוה The LORD מלך reigneth, גאות with majesty; לבשׁ he is clothed לבשׁ is clothed יהוה the LORD עז with strength, התאזר he hath girded himself: אף also תכון is established, תבל the world בל that it cannot תמוט׃ be moved.}%
\verse{נכון established כסאך Thy throne מאז of old: מעולם from everlasting. אתה׃ thou}%
\verse{נשׂאו have lifted up, נהרות The floods יהוה O LORD, נשׂאו have lifted up נהרות the floods קולם their voice; ישׂאו lift up נהרות the floods דכים׃ their waves.}%
\verse{מקלות than the noise מים waters, רבים of many אדירים mightier משׁברי waves ים of the sea. אדיר the mighty במרום on high יהוה׃ The LORD}%
\verse{עדתיך Thy testimonies נאמנו sure: מאד are very לביתך thine house, נאוה becometh קדשׁ holiness יהוה O LORD, לארך forever. ימים׃ forever.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 94
\verseWithHeading{A Prayer for Retribution against Oppressors}{אל God, נקמות to whom vengeance יהוה O LORD אל belongeth; O God, נקמות to whom vengeance הופיע׃ belongeth, show thyself.}%
\verse{הנשׂא Lift up thyself, שׁפט thou judge הארץ of the earth: השׁב render גמול a reward על to גאים׃ the proud.}%
\verse{עד how long מתי how long רשׁעים shall the wicked, יהוה LORD, עד how long מתי how long רשׁעים shall the wicked יעלזו׃ triumph?}%
\verse{יביעו shall they utter ידברו speak עתק hard things? יתאמרו boast themselves? כל all פעלי the workers און׃ of iniquity}%
\verse{עמך thy people, יהוה O LORD, ידכאו They break in pieces ונחלתך thine heritage. יענו׃ and afflict}%
\verse{אלמנה the widow וגר and the stranger, יהרגו They slay ויתומים the fatherless. ירצחו׃ and murder}%
\verse{ויאמרו Yet they say, לא shall not יראה see, יה The LORD ולא neither יבין regard אלהי shall the God יעקב׃ of Jacob}%
\verse{בינו Understand, בערים ye brutish בעם among the people: וכסילים and fools, מתי when תשׂכילו׃ will ye be wise?}%
\verse{הנטע He that planted אזן the ear, הלא shall he not ישׁמע hear? אם יצר he that formed עין the eye, הלא shall he not יביט׃ see?}%
\verse{היסר He that chastiseth גוים the heathen, הלא shall not יוכיח he correct? המלמד he that teacheth אדם man דעת׃ knowledge,}%
\verse{יהוה The LORD ידע knoweth מחשׁבות the thoughts אדם of man, כי that המה they הבל׃ vanity.}%
\verse{אשׁרי Blessed הגבר the man אשׁר whom תיסרנו thou chastenest, יה O LORD, ומתורתך him out of thy law; תלמדנו׃ and teachest}%
\verse{להשׁקיט That thou mayest give him rest לו מימי from the days רע of adversity, עד until יכרה be digged לרשׁע for the wicked. שׁחת׃ the pit}%
\verse{כי For לא will not יטשׁ cast off יהוה the LORD עמו his people, ונחלתו his inheritance. לא neither יעזב׃ will he forsake}%
\verse{כי But עד unto צדק righteousness: ישׁוב shall return משׁפט judgment ואחריו shall follow כל and all ישׁרי the upright לב׃ in heart}%
\verse{מי Who יקום will rise up לי עם for me against מרעים the evildoers? מי who יתיצב will stand up לי עם for me against פעלי the workers און׃ of iniquity?}%
\verse{לולי Unless יהוה the LORD עזרתה my help, לי כמעט had almost שׁכנה dwelt דומה in silence. נפשׁי׃ my soul}%
\verse{אם When אמרתי I said, מטה slippeth; רגלי My foot חסדך thy mercy, יהוה O LORD, יסעדני׃ held me up.}%
\verse{ברב In the multitude שׂרעפי of my thoughts בקרבי within תנחומיך me thy comforts ישׁעשׁעו delight נפשׁי׃ my soul.}%
\verse{היחברך have fellowship כסא Shall the throne הוות of iniquity יצר with thee, which frameth עמל mischief עלי by חק׃ a law?}%
\verse{יגודו They gather themselves together על against נפשׁ the soul צדיק of the righteous, ודם blood. נקי the innocent ירשׁיעו׃ and condemn}%
\verse{ויהי is יהוה But the LORD לי למשׂגב my defense; ואלהי and my God לצור the rock מחסי׃ of my refuge.}%
\verse{וישׁב And he shall bring עליהם upon את אונם them their own iniquity, וברעתם in their own wickedness; יצמיתם and shall cut them off יצמיתם shall cut them off. יהוה the LORD אלהינו׃ our God}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 95
\verseWithHeading{A Call to Worship and Obey}{לכו O come, נרננה let us sing ליהוה unto the LORD: נריעה let us make a joyful noise לצור to the rock ישׁענו׃ of our salvation.}%
\verse{נקדמה Let us come before פניו his presence בתודה with thanksgiving, בזמרות unto him with psalms. נריע׃ and make a joyful noise}%
\verse{כי For אל God, גדול a great יהוה the LORD ומלך King גדול and a great על above כל all אלהים׃ gods.}%
\verse{אשׁר בידו In his hand מחקרי the deep places ארץ of the earth: ותועפות the strength הרים׃ of the hills}%
\verse{אשׁר לו הים The sea והוא his, and he עשׂהו made ויבשׁת the dry ידיו it: and his hands יצרו׃ formed}%
\verse{באו O come, נשׁתחוה let us worship ונכרעה and bow down: נברכה let us kneel לפני before יהוה the LORD עשׂנו׃ our maker.}%
\verse{כי For הוא he אלהינו our God; ואנחנו and we עם the people מרעיתו of his pasture, וצאן and the sheep ידו of his hand. היום Today אם if בקלו his voice, תשׁמעו׃ ye will hear}%
\verse{אל not תקשׁו Harden לבבכם your heart, כמריבה as in the provocation, כיום as the day מסה of temptation במדבר׃ in the wilderness:}%
\verse{אשׁר When נסוני tempted אבותיכם your fathers בחנוני me, proved גם me, and ראו saw פעלי׃ my work.}%
\verse{ארבעים Forty שׁנה years אקוט long was I grieved בדור with generation, ואמר and said, עם a people תעי that do err לבב in their heart, הם It והם and they לא have not ידעו known דרכי׃ my ways:}%
\verse{אשׁר Unto whom נשׁבעתי I swore באפי in my wrath אם that יבאון they should not enter אל into מנוחתי׃ my rest.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 96
\verseWithHeading{Adonai the King Comes in Judgment}{שׁירו O sing ליהוה unto the LORD שׁיר song: חדשׁ a new שׁירו sing ליהוה unto the LORD, כל all הארץ׃ the earth.}%
\verse{שׁירו Sing ליהוה unto the LORD, ברכו bless שׁמו his name; בשׂרו show forth מיום from day ליום to day. ישׁועתו׃ his salvation}%
\verse{ספרו Declare בגוים among the heathen, כבודו his glory בכל among all העמים people. נפלאותיו׃ his wonders}%
\verse{כי For גדול great, יהוה the LORD ומהלל to be praised: מאד and greatly נורא to be feared הוא he על above כל all אלהים׃ gods.}%
\verse{כי For כל all אלהי the gods העמים of the nations אלילים idols: ויהוה but the LORD שׁמים the heavens. עשׂה׃ made}%
\verse{הוד Honor והדר and majesty לפניו before עז him: strength ותפארת and beauty במקדשׁו׃ in his sanctuary.}%
\verse{הבו Give ליהוה unto the LORD, משׁפחות O ye kindreds עמים of the people, הבו give ליהוה unto the LORD כבוד glory ועז׃ and strength.}%
\verse{הבו Give ליהוה unto the LORD כבוד the glory שׁמו his name: שׂאו bring מנחה an offering, ובאו and come לחצרותיו׃ into his courts.}%
\verse{השׁתחוו O worship ליהוה the LORD בהדרת in the beauty קדשׁ of holiness: חילו fear מפניו before כל him, all הארץ׃ the earth.}%
\verse{אמרו Say בגוים among the heathen יהוה the LORD מלך reigneth: אף also תכון shall be established תבל the world בל that it shall not תמוט be moved: ידין he shall judge עמים the people במישׁרים׃ righteously.}%
\verse{ישׂמחו rejoice, השׁמים Let the heavens ותגל be glad; הארץ and let the earth ירעם roar, הים let the sea ומלאו׃ and the fullness}%
\verse{יעלז be joyful, שׂדי Let the field וכל and all אשׁר that בו אז therein: then ירננו rejoice כל shall all עצי the trees יער׃ of the wood}%
\verse{לפני Before יהוה the LORD: כי for בא he cometh, כי for בא he cometh לשׁפט to judge הארץ the earth: ישׁפט he shall judge תבל the world בצדק with righteousness, ועמים and the people באמונתו׃ with his truth.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 97
\verseWithHeading{Adonai’s Glorious Reign}{יהוה The LORD מלך reigneth; תגל rejoice; הארץ let the earth ישׂמחו be glad איים of isles רבים׃ let the multitude}%
\verse{ענן Clouds וערפל and darkness סביביו round about צדק him: righteousness ומשׁפט and judgment מכון the habitation כסאו׃ of his throne.}%
\verse{אשׁ A fire לפניו before תלך goeth ותלהט him, and burneth up סביב round about. צריו׃ his enemies}%
\verse{האירו enlightened ברקיו His lightnings תבל the world: ראתה saw, ותחל and trembled. הארץ׃ the earth}%
\verse{הרים The hills כדונג like wax נמסו melted מלפני at the presence יהוה of the LORD, מלפני at the presence אדון of the Lord כל of the whole הארץ׃ earth.}%
\verse{הגידו declare השׁמים The heavens צדקו his righteousness, וראו see כל and all העמים the people כבודו׃ his glory.}%
\verse{יבשׁו Confounded כל be all עבדי they that serve פסל graven images, המתהללים that boast themselves באלילים of idols: השׁתחוו worship לו כל him, all אלהים׃ gods.}%
\verse{שׁמעה heard, ותשׂמח and was glad; ציון Zion ותגלנה rejoiced בנות and the daughters יהודה of Judah למען because of משׁפטיך thy judgments, יהוה׃ O LORD.}%
\verse{כי For אתה thou, יהוה LORD, עליון high על above כל all הארץ the earth: מאד far נעלית thou art exalted על above כל all אלהים׃ gods.}%
\verse{אהבי Ye that love יהוה the LORD, שׂנאו hate רע evil: שׁמר he preserveth נפשׁות the souls חסידיו of his saints; מיד them out of the hand רשׁעים of the wicked. יצילם׃ he delivereth}%
\verse{אור Light זרע is sown לצדיק for the righteous, ולישׁרי for the upright לב in heart. שׂמחה׃ and gladness}%
\verse{שׂמחו Rejoice צדיקים ye righteous; ביהוה in the LORD, והודו and give thanks לזכר at the remembrance קדשׁו׃ of his holiness.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 98
\verseWithHeading{Praise to Adonai for His Salvation and Judgment}{מזמור A Psalm. שׁירו O sing ליהוה unto the LORD שׁיר song; חדשׁ a new כי for נפלאות marvelous things: עשׂה he hath done הושׁיעה hath gotten him the victory. לו ימינו his right hand, וזרוע arm, קדשׁו׃ and his holy}%
\verse{הודיע hath made known יהוה The LORD ישׁועתו his salvation: לעיני in the sight הגוים of the heathen. גלה hath he openly showed צדקתו׃ his righteousness}%
\verse{זכר He hath remembered חסדו his mercy ואמונתו and his truth לבית toward the house ישׂראל of Israel: ראו have seen כל all אפסי the ends ארץ of the earth את ישׁועת the salvation אלהינו׃ of our God.}%
\verse{הריעו Make a joyful noise ליהוה unto the LORD, כל all הארץ the earth: פצחו make a loud noise, ורננו and rejoice, וזמרו׃ and sing praise.}%
\verse{זמרו Sing ליהוה unto the LORD בכנור with the harp; בכנור with the harp, וקול and the voice זמרה׃ of a psalm.}%
\verse{בחצצרות With trumpets וקול and sound שׁופר of cornet הריעו make a joyful noise לפני before המלך the King. יהוה׃ the LORD,}%
\verse{ירעם roar, הים Let the sea ומלאו and the fullness תבל thereof; the world, וישׁבי׃ and they that dwell}%
\verse{נהרות Let the floods ימחאו clap כף hands: יחד together הרים let the hills ירננו׃ be joyful}%
\verse{לפני Before יהוה the LORD; כי for בא he cometh לשׁפט to judge הארץ the earth: ישׁפט shall he judge תבל the world, בצדק with righteousness ועמים and the people במישׁרים׃ with equity.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 99
\verseWithHeading{Adonai Is a Holy King}{יהוה The LORD מלך reigneth; ירגזו tremble: עמים let the people ישׁב he sitteth כרובים the cherubims; תנוט be moved. הארץ׃ let the earth}%
\verse{יהוה The LORD בציון in Zion; גדול great ורם high הוא and he על above כל all העמים׃ the people.}%
\verse{יודו Let them praise שׁמך name; גדול thy great ונורא and terrible קדושׁ holy. הוא׃ it}%
\verse{ועז strength מלך The king's משׁפט judgment; אהב also loveth אתה thou כוננת dost establish מישׁרים equity, משׁפט judgment וצדקה and righteousness ביעקב in Jacob. אתה thou עשׂית׃ executest}%
\verse{רוממו Exalt יהוה ye the LORD אלהינו our God, והשׁתחוו and worship להדם at his footstool; רגליו at his footstool; קדושׁ holy. הוא׃ he}%
\verse{משׁה Moses ואהרן and Aaron בכהניו among his priests, ושׁמואל and Samuel בקראי among them that call upon שׁמו his name; קראים they called אל upon יהוה the LORD, והוא and he יענם׃ answered}%
\verse{בעמוד pillar: ענן them in the cloudy ידבר He spoke אליהם unto שׁמרו they kept עדתיו his testimonies, וחק and the ordinance נתן׃ he gave}%
\verse{יהוה אלהינו our God: אתה thou עניתם Thou answeredst אל a God נשׂא that forgavest היית wast להם ונקם them, though thou tookest vengeance על of עלילותם׃ their inventions.}%
\verse{רוממו Exalt יהוה the LORD אלהינו our God, והשׁתחוו and worship להר hill; קדשׁו at his holy כי for קדושׁ holy. יהוה the LORD אלהינו׃ our God}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 100
\verseWithHeading{Worship God with Joy}{מזמור A Psalm לתודה of praise. הריעו Make a joyful noise ליהוה unto the LORD, כל all הארץ׃ ye lands.}%
\verse{עבדו Serve את יהוה the LORD בשׂמחה with gladness: באו come לפניו before his presence ברננה׃ with singing.}%
\verse{דעו Know כי ye that יהוה the LORD הוא he אלהים God: הוא he עשׂנו hath made ולא us, and not אנחנו we ourselves; עמו his people, וצאן and the sheep מרעיתו׃ of his pasture.}%
\verse{באו Enter שׁעריו into his gates בתודה with thanksgiving, חצרתיו into his courts בתהלה with praise: הודו be thankful לו ברכו unto him, bless שׁמו׃ his name.}%
\verse{כי For טוב good; יהוה the LORD לעולם everlasting; חסדו his mercy ועד to דר all generations. ודר all generations. אמונתו׃ and his truth}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 101
\verseWithHeading{A Promise to Act with Integrity}{לדוד of David. מזמור A Psalm חסד of mercy ומשׁפט and judgment: אשׁירה I will sing לך יהוה unto thee, O LORD, אזמרה׃ will I sing.}%
\verse{אשׂכילה I will behave myself wisely בדרך way. תמים in a perfect מתי O when תבוא wilt thou come אלי unto אתהלך me? I will walk בתם with a perfect לבבי heart. בקרב within ביתי׃ my house}%
\verse{לא no אשׁית I will set לנגד before עיני mine eyes: דבר thing בליעל wicked עשׂה the work סטים of them that turn aside; שׂנאתי I hate לא shall not ידבק׃ cleave}%
\verse{לבב heart עקשׁ A froward יסור shall depart ממני from רע a wicked לא me: I will not אדע׃ know}%
\verse{מלושׁני slandereth בסתר Whoso privily רעהו his neighbor, אותו אצמית him will I cut off: גבה him that hath a high עינים look ורחב and a proud לבב heart אתו לא will not אוכל׃ I suffer.}%
\verse{עיני Mine eyes בנאמני upon the faithful ארץ of the land, לשׁבת that they may dwell עמדי with הלך me: he that walketh בדרך way, תמים in a perfect הוא he ישׁרתני׃ shall serve}%
\verse{לא shall not ישׁב dwell בקרב within ביתי my house: עשׂה He that worketh רמיה deceit דבר he that telleth שׁקרים lies לא shall not יכון tarry לנגד in עיני׃ my sight.}%
\verse{לבקרים I will early אצמית destroy כל all רשׁעי the wicked ארץ of the land; להכרית that I may cut off מעיר from the city יהוה of the LORD. כל all פעלי doers און׃ wicked}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 102
\verseWithHeading{A Plea for Personal and National Help}{תפלה A Prayer לעני of the afflicted, כי when יעטף he is overwhelmed, ולפני before יהוה the LORD. ישׁפך and poureth out שׂיחו׃ his complaint יהוה O LORD, שׁמעה Hear תפלתי my prayer, ושׁועתי and let my cry אליך unto תבוא׃ come}%
\verse{אל not תסתר Hide פניך thy face ממני from ביום me in the day צר I am in trouble; לי הטה incline אלי unto אזנך thine ear ביום me: in the day אקרא I call מהר ענני׃ answer}%
\verse{כי For כלו are consumed בעשׁן like smoke, ימי my days ועצמותי and my bones כמו a hearth. קד נחרו׃ are burned}%
\verse{הוכה is smitten, כעשׂב like grass; ויבשׁ and withered לבי My heart כי so שׁכחתי that I forget מאכל to eat לחמי׃ my bread.}%
\verse{מקול אנחתי of my groaning דבקה cleave עצמי my bones לבשׂרי׃ to my skin.}%
\verse{דמיתי I am like לקאת a pelican מדבר of the wilderness: הייתי I am ככוס like an owl חרבות׃ of the desert.}%
\verse{שׁקדתי I watch, ואהיה and am כצפור as a sparrow בודד alone על upon גג׃ the house top.}%
\verse{כל me all היום the day; חרפוני reproach אויבי Mine enemies מהוללי they that are mad בי נשׁבעו׃ against me are sworn}%
\verse{כי For אפר ashes כלחם like bread, אכלתי I have eaten ושׁקוי my drink בבכי with weeping, מסכתי׃ and mingled}%
\verse{מפני זעמך of thine indignation וקצפך and thy wrath: כי for נשׂאתני thou hast lifted me up, ותשׁליכני׃ and cast me down.}%
\verse{ימי My days כצל like a shadow נטוי that declineth; ואני and I כעשׂב like grass. איבשׁ׃ am withered}%
\verse{ואתה But thou, יהוה O LORD, לעולם forever; תשׁב shalt endure וזכרך and thy remembrance לדר unto all generations. ודר׃ unto all generations.}%
\verse{אתה Thou תקום shalt arise, תרחם have mercy upon ציון Zion: כי for עת the time לחננה to favor כי her, yea, בא is come. מועד׃ the set time,}%
\verse{כי For רצו take pleasure in עבדיך thy servants את אבניה her stones, ואת עפרה the dust יחננו׃ and favor}%
\verse{וייראו shall fear גוים So the heathen את שׁם the name יהוה of the LORD, וכל and all מלכי the kings הארץ of the earth את כבודך׃ thy glory.}%
\verse{כי When בנה shall build up יהוה the LORD ציון Zion, נראה he shall appear בכבודו׃ in his glory.}%
\verse{פנה אלפלת the prayer הערער of the destitute, ולא and not בזה despise את תפלתם׃ their prayer.}%
\verse{תכתב shall be written זאת This לדור for the generation אחרון to come: ועם and the people נברא which shall be created יהלל shall praise יה׃ the LORD.}%
\verse{כי For השׁקיף he hath looked down ממרום from the height קדשׁו of his sanctuary; יהוה did the LORD משׁמים from heaven אל behold ארץ the earth; הביט׃ behold}%
\verse{לשׁמע To hear אנקת the groaning אסיר of the prisoner; לפתח to loose בני those that are appointed תמותה׃ to death;}%
\verse{לספר To declare בציון in Zion, שׁם the name יהוה of the LORD ותהלתו and his praise בירושׁלם׃ in Jerusalem;}%
\verse{בהקבץ are gathered עמים When the people יחדו together, וממלכות and the kingdoms, לעבד to serve את יהוה׃ the LORD.}%
\verse{ענה He weakened בדרך in the way; כחו my strength קצר he shortened ימי׃ my days.}%
\verse{אמר I said, אלי O my God, אל me not תעלני take בחצי away in the midst ימי of my days: בדור throughout all generations. דורים throughout all generations. שׁנותיך׃ thy years}%
\verse{לפנים Of old הארץ of the earth: יסדת hast thou laid the foundation ומעשׂה the work ידיך of thy hands. שׁמים׃ and the heavens}%
\verse{המה They יאבדו shall perish, ואתה but thou תעמד shalt endure: וכלם yea, all כבגד like a garment; יבלו of them shall wax old כלבושׁ as a vesture תחליפם shalt thou change ויחלפו׃ them, and they shall be changed:}%
\verse{ואתה But thou הוא the same, ושׁנותיך and thy years לא shall have no יתמו׃ end.}%
\verse{בני The children עבדיך of thy servants ישׁכונו shall continue, וזרעם and their seed לפניך before יכון׃ shall be established}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 103
\verseWithHeading{Thanksgiving for Adonai’s Compassion}{לדוד of David. ברכי Bless נפשׁי O my soul: את יהוה the LORD, וכל and all קרבי that is within את שׁם name. קדשׁו׃ me, his holy}%
\verse{ברכי Bless נפשׁי O my soul, את יהוה the LORD, ואל not תשׁכחי and forget כל all גמוליו׃ his benefits:}%
\verse{הסלח Who forgiveth לכל all עונכי thine iniquities; הרפא who healeth לכל all תחלאיכי׃ thy diseases;}%
\verse{הגואל Who redeemeth משׁחת from destruction; חייכי thy life המעטרכי who crowneth חסד thee with lovingkindness ורחמים׃ and tender mercies;}%
\verse{המשׂביע Who satisfieth בטוב with good עדיך thy mouth תתחדשׁ is renewed כנשׁר like the eagle's. נעוריכי׃ thy youth}%
\verse{עשׂה executeth צדקות righteousness יהוה The LORD ומשׁפטים and judgment לכל for all עשׁוקים׃ that are oppressed.}%
\verse{יודיע He made known דרכיו his ways למשׁה unto Moses, לבני unto the children ישׂראל of Israel. עלילותיו׃ his acts}%
\verse{רחום merciful וחנון and gracious, יהוה The LORD ארך slow אפים to anger, ורב and plenteous חסד׃ in mercy.}%
\verse{לא He will not לנצח always יריב chide: ולא neither לעולם forever. יטור׃ will he keep}%
\verse{לא He hath not כחטאינו with us after our sins; עשׂה dealt לנו ולא nor כעונתינו us according to our iniquities. גמל rewarded עלינו׃ rewarded}%
\verse{כי For כגבה is high שׁמים as the heaven על above הארץ the earth, גבר great חסדו is his mercy על toward יראיו׃ them that fear}%
\verse{כרחק far hath he removed מזרח as the east ממערב is from the west, הרחיק ממנו is from the west, את פשׁעינו׃ our transgressions}%
\verse{כרחם pitieth אב Like as a father על pitieth בנים children, רחם pitieth יהוה the LORD על pitieth יראיו׃ them that fear}%
\verse{כי For הוא he ידע knoweth יצרנו our frame; זכור he remembereth כי that עפר dust. אנחנו׃ we}%
\verse{אנושׁ כחציר as grass: ימיו his days כציץ as a flower השׂדה of the field, כן so יציץ׃ he flourisheth.}%
\verse{כי For רוח the wind עברה passeth בו ואיננו over it, and it is gone; ולא it no יכירנו thereof shall know עוד more. מקומו׃ and the place}%
\verse{וחסד But the mercy יהוה of the LORD מעולם from everlasting ועד to עולם everlasting על upon יראיו them that fear וצדקתו him, and his righteousness לבני unto children's בנים׃ children;}%
\verse{לשׁמרי To such as keep בריתו his covenant, ולזכרי and to those that remember פקדיו his commandments לעשׂותם׃ to do}%
\verse{יהוה The LORD בשׁמים in the heavens; הכין hath prepared כסאו his throne ומלכותו and his kingdom בכל over all. משׁלה׃ ruleth}%
\verse{ברכו Bless יהוה the LORD, מלאכיו ye his angels, גברי that excel כח in strength, עשׂי that do דברו his commandments, לשׁמע hearkening בקול unto the voice דברו׃ of his word.}%
\verse{ברכו Bless יהוה ye the LORD, כל all צבאיו his hosts; משׁרתיו ministers עשׂי of his, that do רצונו׃ his pleasure.}%
\verse{ברכו Bless יהוה the LORD, כל all מעשׂיו his works בכל in all מקמות places ממשׁלתו of his dominion: ברכי bless נפשׁי O my soul. את יהוה׃ the LORD,}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 104
\verseWithHeading{Praise to Adonai for His Creation and Providence}{ברכי Bless נפשׁי O my soul. את יהוה the LORD, יהוה O LORD אלהי my God, גדלת great; מאד thou art very הוד with honor והדר and majesty. לבשׁת׃ thou art clothed}%
\verse{עטה Who coverest אור with light כשׂלמה as a garment: נוטה who stretchest out שׁמים the heavens כיריעה׃ like a curtain:}%
\verse{המקרה Who layeth the beams במים in the waters: עליותיו of his chambers השׂם who maketh עבים the clouds רכובו his chariot: המהלך who walketh על upon כנפי the wings רוח׃ of the wind:}%
\verse{עשׂה Who maketh מלאכיו his angels רוחות spirits; משׁרתיו his ministers אשׁ fire: להט׃ a flaming}%
\verse{יסד ארץ of the earth, על מכוניה the foundations בל it should not תמוט be removed עולם forever. ועד׃ forever.}%
\verse{תהום it with the deep כלבושׁ as a garment: כסיתו Thou coveredst על above הרים the mountains. יעמדו stood מים׃ the waters}%
\verse{מן At גערתך thy rebuke ינוסון they fled; מן at קול the voice רעמך of thy thunder יחפזון׃ they hasted away.}%
\verse{יעלו They go up הרים by the mountains; ירדו they go down בקעות by the valleys אל unto מקום the place זה which יסדת thou hast founded להם׃}%
\verse{גבול a bound שׂמת Thou hast set בל that they may not יעברון pass over; בל that they turn not again ישׁובון that they turn not again לכסות to cover הארץ׃ the earth.}%
\verse{המשׁלח He sendeth מעינים the springs בנחלים into the valleys, בין among הרים the hills. יהלכון׃ run}%
\verse{ישׁקו They give drink כל to every חיתו beast שׂדי of the field: ישׁברו quench פראים the wild asses צמאם׃ their thirst.}%
\verse{עליהם By עוף them shall the fowls השׁמים of the heaven ישׁכון have their habitation, מבין among עפאים the branches. יתנו sing קול׃ sing}%
\verse{משׁקה He watereth הרים the hills מעליותיו from his chambers: מפרי with the fruit מעשׂיך of thy works. תשׂבע is satisfied הארץ׃ the earth}%
\verse{מצמיח to grow חציר He causeth the grass לבהמה for the cattle, ועשׂב and herb לעבדת for the service האדם of man: להוציא that he may bring forth לחם food מן out of הארץ׃ the earth;}%
\verse{ויין And wine ישׂמח maketh glad לבב the heart אנושׁ of man, להצהיל to shine, פנים to make face משׁמן oil ולחם and bread לבב heart. אנושׁ man's יסעד׃ strengtheneth}%
\verse{ישׂבעו are full עצי The trees יהוה of the LORD ארזי the cedars לבנון of Lebanon, אשׁר which נטע׃ he hath planted;}%
\verse{אשׁר שׁםפרים the birds יקננו make their nests: חסידה the stork, ברושׁים the fir trees ביתה׃ her house.}%
\verse{הרים hills הגבהים The high ליעלים for the wild goats; סלעים the rocks מחסה a refuge לשׁפנים׃ for the conies.}%
\verse{עשׂה He appointed ירח the moon למועדים for seasons: שׁמשׁ the sun ידע knoweth מבואו׃ his going down.}%
\verse{תשׁת Thou makest חשׁך darkness, ויהי and it is לילה night: בו תרמשׂ do creep כל wherein all חיתו the beasts יער׃ of the forest}%
\verse{הכפירים The young lions שׁאגים roar לטרף after their prey, ולבקשׁ and seek מאל אכלם׃ their meat}%
\verse{תזרח ariseth, השׁמשׁ The sun יאספון they gather themselves together, ואל in מעונתם their dens. ירבצון׃ and lay them down}%
\verse{יצא goeth forth אדם Man לפעלו unto his work ולעבדתו and to his labor עדי ערב׃ the evening.}%
\verse{מה how רבו manifold מעשׂיך are thy works! יהוה O LORD, כלם them all: בחכמה in wisdom עשׂית hast thou made מלאה is full הארץ the earth קנינך׃ of thy riches.}%
\verse{זה this הים sea, גדול great ורחב and wide ידים and wide שׁם wherein רמשׂ things creeping ואין innumerable, מספר innumerable, חיות beasts. קטנות both small עם and גדלות׃ great}%
\verse{שׁם There אניות the ships: יהלכון go לויתן that leviathan, זה thou יצרת hast made לשׂחק׃ to play}%
\verse{כלם all אליך upon ישׂברון These wait לתת thee; that thou mayest give אכלם their meat בעתו׃ in due season.}%
\verse{תתן thou givest להם ילקטון them they gather: תפתח thou openest ידך thine hand, ישׂבעון they are filled טוב׃ with good.}%
\verse{תסתיר Thou hidest פניך thy face, יבהלון they are troubled: תסף thou takest away רוחם their breath, יגועון they die, ואל to עפרם their dust. ישׁובון׃ and return}%
\verse{תשׁלח Thou sendest forth רוחך thy spirit, יבראון they are created: ותחדשׁ and thou renewest פני the face אדמה׃ of the earth.}%
\verse{יהי shall endure כבוד The glory יהוה of the LORD לעולם forever: ישׂמח shall rejoice יהוה the LORD במעשׂיו׃ in his works.}%
\verse{המביט He looketh לארץ on the earth, ותרעד and it trembleth: יגע he toucheth בהרים the hills, ויעשׁנו׃ and they smoke.}%
\verse{אשׁירה I will sing ליהוה unto the LORD בחיי as long as I live: אזמרה I will sing praise לאלהי to my God בעודי׃ while I have my being.}%
\verse{יערב him shall be sweet: עליו of שׂיחי My meditation אנכי I אשׂמח will be glad ביהוה׃ in the LORD.}%
\verse{יתמו be consumed חטאים Let the sinners מן out of הארץ the earth, ורשׁעים and let the wicked עוד more. אינם be no ברכי Bless נפשׁי O my soul. את יהוה thou the LORD, הללו Praise יה׃ ye the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 105
\verseWithHeading{Praise to Adonai for His Work on Behalf of Israel}{הודו O give thanks ליהוה unto the LORD; קראו call בשׁמו upon his name: הודיעו make known בעמים among the people. עלילותיו׃ his deeds}%
\verse{שׁירו Sing לו זמרו unto him, sing psalms לו שׂיחו unto him: talk בכל ye of all נפלאותיו׃ his wondrous works.}%
\verse{התהללו Glory בשׁם name: קדשׁו ye in his holy ישׂמח of them rejoice לב let the heart מבקשׁי that seek יהוה׃ the LORD.}%
\verse{דרשׁו Seek יהוה the LORD, ועזו and his strength: בקשׁו seek פניו his face תמיד׃ evermore.}%
\verse{זכרו Remember נפלאותיו his marvelous works אשׁר that עשׂה he hath done; מפתיו his wonders, ומשׁפטי and the judgments פיו׃ of his mouth;}%
\verse{זרע O ye seed אברהם of Abraham עבדו his servant, בני ye children יעקב of Jacob בחיריו׃ his chosen.}%
\verse{הוא He יהוה the LORD אלהינו our God: בכל in all הארץ the earth. משׁפטיו׃ his judgments}%
\verse{זכר He hath remembered לעולם forever, בריתו his covenant דבר the word צוה he commanded לאלף to a thousand דור׃ generations.}%
\verse{אשׁר Which כרת he made את with אברהם Abraham, ושׁבועתו and his oath לישׂחק׃ unto Isaac;}%
\verse{ויעמידה And confirmed ליעקב the same unto Jacob לחק for a law, לישׂראל to Israel ברית covenant: עולם׃ an everlasting}%
\verse{לאמר Saying, לך אתן Unto thee will I give את ארץ the land כנען of Canaan, חבל the lot נחלתכם׃ of your inheritance:}%
\verse{בהיותם When they were מתי a few men מספר in number; כמעט yea, very few, וגרים׃ and strangers}%
\verse{ויתהלכו When they went מגוי from one nation אל to גוי another, מממלכה from kingdom אל to עם people; אחר׃ another}%
\verse{לא no הניח אדם man לעשׁקם to do them wrong: ויוכח yea, he reproved עליהם for their sakes; מלכים׃ kings}%
\verse{אל not תגעו Touch במשׁיחי mine anointed, ולנביאי and do my prophets אל no תרעו׃ harm.}%
\verse{ויקרא Moreover he called for רעב a famine על upon הארץ the land: כל the whole מטה staff לחם of bread. שׁבר׃ he broke}%
\verse{שׁלח He sent לפניהם before אישׁ a man לעבד for a servant: נמכר was sold יוסף׃ them, Joseph,}%
\verse{ענו they hurt בכבל with fetters: רגליו Whose feet ברזל in iron: באה was laid נפשׁו׃ he}%
\verse{עד Until עת the time בא came: דברו that his word אמרת the word יהוה of the LORD צרפתהו׃ tried}%
\verse{שׁלח sent מלך The king ויתירהו and loosed משׁל him; the ruler עמים of the people, ויפתחהו׃ and let him go free.}%
\verse{שׂמו He made אדון him lord לביתו of his house, ומשׁל and ruler בכל of all קנינו׃ his substance:}%
\verse{לאסר To bind שׂריו his princes בנפשׁו at his pleasure; וזקניו and teach his senators יחכם׃ wisdom.}%
\verse{ויבא also came into ישׂראל Israel מצרים Egypt; ויעקב and Jacob גר sojourned בארץ in the land חם׃ of Ham.}%
\verse{ויפר And he increased את עמו his people מאד greatly; ויעצמהו and made them stronger מצריו׃ than their enemies.}%
\verse{הפך He turned לבם their heart לשׂנא to hate עמו his people, להתנכל to deal subtlely בעבדיו׃ with his servants.}%
\verse{שׁלח He sent משׁה Moses עבדו his servant; אהרן Aaron אשׁר whom בחר׃ he had chosen.}%
\verse{שׂמו They showed בם דברי his signs אתותיו his signs ומפתים among them, and wonders בארץ in the land חם׃ of Ham.}%
\verse{שׁלח He sent חשׁך darkness, ויחשׁך and made it dark; ולא not מרו and they rebelled את דברוו׃ against his word.}%
\verse{הפך He turned את מימיהם their waters לדם into blood, וימת and slew את דגתם׃ their fish.}%
\verse{שׁרץ brought forth frogs in abundance, ארצם Their land צפרדעים brought forth frogs in abundance, בחדרי in the chambers מלכיהם׃ of their kings.}%
\verse{אמר He spoke, ויבא and there came ערב divers sorts of flies, כנים lice בכל in all גבולם׃ their coasts.}%
\verse{נתן He gave גשׁמיהם for rain, ברד them hail אשׁ fire להבות flaming בארצם׃ in their land.}%
\verse{ויך He smote גפנם their vines ותאנתם also and their fig trees; וישׁבר and broke עץ the trees גבולם׃ of their coasts.}%
\verse{אמר He spoke, ויבא came, ארבה and the locusts וילק and caterpillars, ואין and that without מספר׃ number,}%
\verse{ויאכל And did eat up כל all עשׂב the herbs בארצם in their land, ויאכל and devoured פרי the fruit אדמתם׃ of their ground.}%
\verse{ויך He smote כל also all בכור the firstborn בארצם in their land, ראשׁית the chief לכל of all אונם׃ their strength.}%
\verse{ויוציאם He brought them forth בכסף also with silver וזהב and gold: ואין and not בשׁבטיו among their tribes. כושׁל׃ one feeble}%
\verse{שׂמח was glad מצרים Egypt בצאתם when they departed: כי for נפל of them fell פחדם the fear עליהם׃ upon}%
\verse{פרשׂ He spread ענן a cloud למסך for a covering; ואשׁ and fire להאיר to give light לילה׃ in the night.}%
\verse{שׁאל asked, ויבא and he brought שׂלו quails, ולחם them with the bread שׁמים of heaven. ישׂביעם׃ and satisfied}%
\verse{פתח He opened צור the rock, ויזובו gushed out; מים and the waters הלכו they ran בציות in the dry places נהר׃ a river.}%
\verse{כי For זכר he remembered את דבר promise, קדשׁו his holy את אברהם Abraham עבדו׃ his servant.}%
\verse{ויוצא And he brought forth עמו his people בשׂשׂון with joy, ברנה with gladness: את בחיריו׃ his chosen}%
\verse{ויתן And gave להם ארצות them the lands גוים of the heathen: ועמל the labor לאמים of the people; יירשׁו׃ and they inherited}%
\verse{בעבור That ישׁמרו they might observe חקיו his statutes, ותורתיו his laws. ינצרו and keep הללו Praise יה׃ ye the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 106
\verseWithHeading{Praise to Adonai for His Faithfulness in Israel’s History}{הללויה Praise הודו O give thanks ליהוה unto the LORD; כי for טוב good: כי for לעולם forever. חסדו׃ his mercy}%
\verse{מי Who ימלל can utter גבורות the mighty acts יהוה of the LORD? ישׁמיע can show forth כל all תהלתו׃ his praise?}%
\verse{אשׁרי Blessed שׁמרי they that keep משׁפט judgment, עשׂה he that doeth צדקה righteousness בכל at all עת׃ times.}%
\verse{זכרני Remember יהוה me, O LORD, ברצון with the favor עמך thy people: פקדני O visit בישׁועתך׃ me with thy salvation;}%
\verse{לראות That I may see בטובת the good בחיריך of thy chosen, לשׂמח that I may rejoice בשׂמחת in the gladness גויך of thy nation, להתהלל that I may glory עם with נחלתך׃ thine inheritance.}%
\verse{חטאנו We have sinned עם with אבותינו our fathers, העוינו we have committed iniquity, הרשׁענו׃ we have done wickedly.}%
\verse{אבותינו Our fathers במצרים in Egypt; לא not השׂכילו understood נפלאותיך thy wonders לא not זכרו they remembered את רב the multitude חסדיך of thy mercies; וימרו but provoked על at ים the sea, בים sea. סוף׃ at the Red}%
\verse{ויושׁיעם Nevertheless he saved למען שׁמוהודיע to be known. את גבורתו׃ that he might make his mighty power}%
\verse{ויגער He rebuked בים sea סוף the Red ויחרב also, and it was dried up: ויוליכם so he led בתהמות them through the depths, כמדבר׃ as through the wilderness.}%
\verse{ויושׁיעם And he saved מיד them from the hand שׂונא of him that hated ויגאלם and redeemed מיד them from the hand אויב׃ of the enemy.}%
\verse{ויכסו covered מים And the waters צריהם their enemies: אחד one מהם לא there was not נותר׃ them left.}%
\verse{ויאמינו Then believed בדבריו they his words; ישׁירו they sang תהלתו׃ his praise.}%
\verse{מהרו They soon שׁכחו forgot מעשׂיו his works; לא not חכו they waited לעצתו׃ for his counsel:}%
\verse{ויתאוו תאוהמדבר in the wilderness, וינסו and tempted אל God בישׁימון׃ in the desert.}%
\verse{ויתן And he gave להם שׁאלתם them their request; וישׁלח but sent רזון leanness בנפשׁם׃ into their soul.}%
\verse{ויקנאו They envied למשׁה Moses במחנה also in the camp, לאהרן Aaron קדושׁ the saint יהוה׃ of the LORD.}%
\verse{תפתח opened ארץ The earth ותבלע and swallowed up דתן Dathan, ותכס and covered על and covered עדת the company אבירם׃ of Abiram.}%
\verse{ותבער was kindled אשׁ And a fire בעדתם in their company; להבה the flame תלהט burned up רשׁעים׃ the wicked.}%
\verse{יעשׂו They made עגל a calf בחרב in Horeb, וישׁתחוו and worshiped למסכה׃ the molten image.}%
\verse{וימירו Thus they changed את כבודם their glory בתבנית into the similitude שׁור of an ox אכל that eateth עשׂב׃ grass.}%
\verse{שׁכחו They forgot אל God מושׁיעם their savior, עשׂה which had done גדלות great things במצרים׃ in Egypt;}%
\verse{נפלאות Wondrous works בארץ in the land חם of Ham, נוראות terrible things על by ים sea. סוף׃ the Red}%
\verse{ויאמר Therefore he said להשׁמידם that he would destroy לולי them, had not משׁה Moses בחירו his chosen עמד stood בפרץ him in the breach, לפניו before להשׁיב to turn away חמתו his wrath, מהשׁחית׃ lest he should destroy}%
\verse{וימאסו Yea, they despised בארץ land, חמדה the pleasant לא not האמינו they believed לדברו׃ his word:}%
\verse{וירגנו But murmured באהליהם in their tents, לא not שׁמעו hearkened בקול unto the voice יהוה׃ of the LORD.}%
\verse{וישׂא Therefore he lifted up ידו his hand להם להפיל against them, to overthrow אותם במדבר׃ them in the wilderness:}%
\verse{ולהפיל To overthrow זרעם their seed בגוים also among the nations, ולזרותם and to scatter בארצות׃ them in the lands.}%
\verse{ויצמדו They joined themselves לבעל פעור also unto Baalpeor, ויאכלו and ate זבחי the sacrifices מתים׃ of the dead.}%
\verse{ויכעיסו Thus they provoked to anger במעלליהם with their inventions: ותפרץ broke in בם מגפה׃ and the plague}%
\verse{ויעמד Then stood up פינחס Phinehas, ויפלל and executed judgment: ותעצר was stayed. המגפה׃ and the plague}%
\verse{ותחשׁב And that was counted לו לצדקה unto him for righteousness לדר unto all generations ודר unto all generations עד forevermore. עולם׃ forevermore.}%
\verse{ויקציפו They angered על also at מי the waters מריבה of strife, וירע so that it went ill למשׁה with Moses בעבורם׃ for their sakes:}%
\verse{כי Because המרו they provoked את רוחו his spirit, ויבטא so that he spoke unadvisedly בשׂפתיו׃ with his lips.}%
\verse{לא They did not השׁמידו destroy את העמים the nations, אשׁר concerning whom אמר commanded יהוה להם׃}%
\verse{ויתערבו But were mingled בגוים among the heathen, וילמדו and learned מעשׂיהם׃ their works.}%
\verse{ויעבדו And they served את עצביהם their idols: ויהיו which were להם למוקשׁ׃ a snare}%
\verse{ויזבחו Yea, they sacrificed את בניהם their sons ואת בנותיהם and their daughters לשׁדים׃ unto devils,}%
\verse{וישׁפכו And shed דם blood, נקי innocent דם the blood בניהם of their sons ובנותיהם and of their daughters, אשׁר whom זבחו they sacrificed לעצבי unto the idols כנען of Canaan: ותחנף was polluted הארץ and the land בדמים׃ with blood.}%
\verse{ויטמאו Thus were they defiled במעשׂיהם with their own works, ויזנו and went a whoring במעלליהם׃ with their own inventions.}%
\verse{ויחר kindled אף Therefore was the wrath יהוה of the LORD בעמו against his people, ויתעב insomuch that he abhorred את נחלתו׃ his own inheritance.}%
\verse{ויתנם And he gave ביד them into the hand גוים of the heathen; וימשׁלו them ruled בהם שׂנאיהם׃ and they that hated}%
\verse{וילחצום also oppressed אויביהם Their enemies ויכנעו them, and they were brought into subjection תחת under ידם׃ their hand.}%
\verse{פעמים times רבות Many יצילם did he deliver והמה them; but they ימרו provoked בעצתם with their counsel, וימכו and were brought low בעונם׃ for their iniquity.}%
\verse{וירא Nevertheless he regarded בצר their affliction, להם בשׁמעו when he heard את רנתם׃ their cry:}%
\verse{ויזכר And he remembered להם בריתו for them his covenant, וינחם and repented כרב according to the multitude חסדו׃ of his mercies.}%
\verse{ויתן He made אותם לרחמים them also to be pitied לפני of כל all שׁוביהם׃ those that carried them captives.}%
\verse{הושׁיענו Save יהוה us, O LORD אלהינו our God, וקבצנו and gather מן us from among הגוים the heathen, להדות to give thanks לשׁם name, קדשׁך unto thy holy להשׁתבח to triumph בתהלתך׃ in thy praise.}%
\verse{ברוך Blessed יהוה the LORD אלהי God ישׂראל of Israel מן from העולם everlasting ועד to העולם everlasting: ואמר say, כל and let all העם the people אמן Amen. הללו Praise יה׃ ye the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 107
\verseWithHeading{Thanksgiving to Adonai for His of Deliverance}{הדו O give thanks ליהוה unto the LORD, כי for טוב good: כי for לעולם forever. חסדו׃ his mercy}%
\verse{יאמרו say גאולי Let the redeemed יהוה of the LORD אשׁר whom גאלם he hath redeemed מיד from the hand צר׃ of the enemy;}%
\verse{ומארצות them out of the lands, קבצם And gathered ממזרח from the east, וממערב and from the west, מצפון from the north, ומים׃ and from the south.}%
\verse{תעו They wandered במדבר in the wilderness בישׁימון in a solitary דרך way; עיר city מושׁב to dwell לא no מצאו׃ they found}%
\verse{רעבים Hungry גם and צמאים thirsty, נפשׁם their soul בהם תתעטף׃ fainted}%
\verse{ויצעקו Then they cried אל unto יהוה בצר in their trouble, להם ממצוקותיהם them out of their distresses. יצילם׃ he delivered}%
\verse{וידריכם And he led them forth בדרך way, ישׁרה by the right ללכת that they might go אל to עיר a city מושׁב׃ of habitation.}%
\verse{יודו Oh that would praise ליהוה the LORD חסדו his goodness, ונפלאותיו and his wonderful works לבני to the children אדם׃ of men!}%
\verse{כי For השׂביע he satisfieth נפשׁ soul, שׁקקה the longing ונפשׁ soul רעבה the hungry מלא and filleth טוב׃ with goodness.}%
\verse{ישׁבי Such as sit חשׁך in darkness וצלמות and in the shadow of death, אסירי bound עני in affliction וברזל׃ and iron;}%
\verse{כי Because המרו they rebelled against אמרי the words אל of God, ועצת the counsel עליון of the most High: נאצו׃ and contemned}%
\verse{ויכנע Therefore he brought down בעמל with labor; לבם their heart כשׁלו they fell down, ואין and none עזר׃ to help.}%
\verse{ויזעקו Then they cried אל unto יהוה בצר in their trouble, להם ממצקותיהם them out of their distresses. יושׁיעם׃ he saved}%
\verse{יוציאם He brought them out מחשׁך of darkness וצלמות and the shadow of death, ומוסרותיהם their bands ינתק׃ and broke}%
\verse{יודו Oh that would praise ליהוה the LORD חסדו his goodness, ונפלאותיו and his wonderful works לבני to the children אדם׃ of men!}%
\verse{כי For שׁבר he hath broken דלתות the gates נחשׁת of brass, ובריחי the bars ברזל of iron גדע׃ and cut}%
\verse{אולים Fools מדרך because פשׁעם of their transgression, ומעונתיהם and because of their iniquities, יתענו׃ are afflicted.}%
\verse{כל all manner אכל of meat; תתעב abhorreth נפשׁם Their soul ויגיעו and they draw near עד unto שׁערי the gates מות׃ of death.}%
\verse{ויזעקו Then they cry אל unto יהוה בצר in their trouble, להם ממצקותיהם them out of their distresses. יושׁיעם׃ he saveth}%
\verse{ישׁלח He sent דברו his word, וירפאם and healed וימלט them, and delivered משׁחיתותם׃ from their destructions.}%
\verse{יודו Oh that would praise ליהוה the LORD חסדו his goodness, ונפלאותיו and his wonderful works לבני to the children אדם׃ of men!}%
\verse{ויזבחו And let them sacrifice זבחי the sacrifices תודה of thanksgiving, ויספרו and declare מעשׂיו his works ברנה׃ with rejoicing.}%
\verse{יורדי They that go down הים to the sea באניות in ships, עשׂי that do מלאכה business במים waters; רבים׃ in great}%
\verse{המה These ראו see מעשׂי the works יהוה of the LORD, ונפלאותיו and his wonders במצולה׃ in the deep.}%
\verse{ויאמר For he commandeth, ויעמד and raiseth רוח wind, סערה the stormy ותרומם which lifteth up גליו׃ the waves}%
\verse{יעלו They mount up שׁמים to the heaven, ירדו they go down תהומות again to the depths: נפשׁם their soul ברעה because of trouble. תתמוגג׃ is melted}%
\verse{יחוגו They reel to and fro, וינועו and stagger כשׁכור like a drunken man, וכל חכמתםתבלע׃}%
\verse{ויצעקו Then they cry אל unto יהוה בצר in their trouble, להם וממצוקתיהם of their distresses. יוציאם׃ and he bringeth them out}%
\verse{יקם He maketh סערה the storm לדממה a calm, ויחשׁו thereof are still. גליהם׃ so that the waves}%
\verse{וישׂמחו Then are they glad כי because ישׁתקו they be quiet; וינחם so he bringeth אל them unto מחוז haven. חפצם׃ their desired}%
\verse{יודו Oh that would praise ליהוה the LORD חסדו his goodness, ונפלאותיו and his wonderful works לבני to the children אדם׃ of men!}%
\verse{וירממוהו Let them exalt בקהל him also in the congregation עם of the people, ובמושׁב him in the assembly זקנים of the elders. יהללוהו׃ and praise}%
\verse{ישׂם He turneth נהרות rivers למדבר into a wilderness, ומצאי and the watersprings מים and the watersprings לצמאון׃ into dry ground;}%
\verse{ארץ land פרי A fruitful למלחה into barrenness, מרעת for the wickedness ישׁבי׃ of them that dwell}%
\verse{ישׂם He turneth מדבר the wilderness לאגם into a standing מים water, וארץ ground ציה and dry למצאי into watersprings. מים׃ into watersprings.}%
\verse{ויושׁב he maketh the hungry to dwell, שׁם And there רעבים he maketh the hungry to dwell, ויכוננו that they may prepare עיר a city מושׁב׃ for habitation;}%
\verse{ויזרעו And sow שׂדות the fields, ויטעו and plant כרמים vineyards, ויעשׂו which may yield פרי fruits תבואה׃ of increase.}%
\verse{ויברכם He blesseth וירבו them also, so that they are multiplied מאד greatly; ובהמתם their cattle לא and suffereth not ימעיט׃ to decrease.}%
\verse{וימעטו Again, they are minished וישׁחו and brought low מעצר through oppression, רעה affliction, ויגון׃ and sorrow.}%
\verse{שׁפך He poureth בוז contempt על upon נדיבים princes, ויתעם and causeth them to wander בתהו in the wilderness, לא no דרך׃ way.}%
\verse{וישׂגב אביוןעוני from affliction, וישׂם and maketh כצאן like a flock. משׁפחות׃ families}%
\verse{יראו shall see ישׁרים The righteous וישׂמחו and rejoice: וכל and all עולה iniquity קפצה shall stop פיה׃ her mouth.}%
\verse{מי Whoso חכם wise, וישׁמר and will observe אלה these ויתבוננו even they shall understand חסדי the lovingkindness יהוה׃ of the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 108
\verseWithHeading{Prayer to Adonai for Victory over Enemies}{שׁיר A Song מזמור Psalm לדוד׃ of David. נכון is fixed; לבי my heart אלהים O God, אשׁירה I will sing ואזמרה and give praise, אף even כבודי׃ with my glory.}%
\verse{עורה Awake, הנבל psaltery וכנור and harp: אעירה I will awake שׁחר׃ early.}%
\verse{אודך I will praise בעמים among the people: יהוה thee, O LORD, ואזמרך and I will sing praises בל אמים׃ unto thee among the nations.}%
\verse{כי For גדול great מעל above שׁמים the heavens: חסדך thy mercy ועד unto שׁחקים the clouds. אמתך׃ and thy truth}%
\verse{רומה Be thou exalted, על above שׁמים the heavens: אלהים O God, ועל above כל all הארץ the earth; כבודך׃ and thy glory}%
\verse{למען That יחלצון may be delivered: ידידיך thy beloved הושׁיעה save ימינך thy right hand, וענני׃ and answer}%
\verse{אלהים God דבר hath spoken בקדשׁו in his holiness; אעלזה I will rejoice, אחלקה I will divide שׁכם Shechem, ועמק the valley סכות of Succoth. אמדד׃ and mete out}%
\verse{לי גלעד Gilead לי מנשׁה mine; Manasseh ואפרים mine; Ephraim מעוז also the strength ראשׁי of mine head; יהודה Judah מחקקי׃ my lawgiver;}%
\verse{מואב Moab סיר my washpot; רחצי my washpot; על over אדום Edom אשׁליך will I cast out נעלי my shoe; עלי over פלשׁת Philistia אתרועע׃ will I triumph.}%
\verse{מי Who יבלני will bring עיר city? מבצר me into the strong מי who נחני will lead עד me into אדום׃ Edom?}%
\verse{הלא not אלהים O God, זנחתנו hast cast us off? ולא and wilt not תצא go forth אלהים thou, O God, בצבאתינו׃ with our hosts?}%
\verse{הבה Give לנו עזרת us help מצר from trouble: ושׁוא for vain תשׁועת the help אדם׃ of man.}%
\verse{באלהים Through God נעשׂה we shall do חיל valiantly: והוא for he יבוס shall tread down צרינו׃ our enemies.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 109
\verseWithHeading{A Prayer for Help against Enemies}{למנצח To the chief Musician, לדוד of David. מזמור A Psalm אלהי thy peace, O God תהלתי of my praise; אל not תחרשׁ׃ Hold}%
\verse{כי For פי the mouth רשׁע of the wicked ופי and the mouth מרמה of the deceitful עלי against פתחו are opened דברו me: they have spoken אתי against לשׁון tongue. שׁקר׃ me with a lying}%
\verse{ודברי also with words שׂנאה of hatred; סבבוני They compassed me about וילחמוני and fought against חנם׃ me without a cause.}%
\verse{תחת For אהבתי my love ישׂטנוני they are my adversaries: ואני but I תפלה׃ prayer.}%
\verse{וישׂימו עליעה me evil תחת for טובה good, ושׂנאה and hatred תחת for אהבתי׃ my love.}%
\verse{הפקד Set עליו over רשׁע thou a wicked man ושׂטן him: and let Satan יעמד stand על at ימינו׃ his right hand.}%
\verse{בהשׁפטו When he shall be judged, יצא let him be condemned: רשׁע let him be condemned: ותפלתו his prayer תהיה become לחטאה׃ sin.}%
\verse{יהיו be ימיו Let his days מעטים few; פקדתו his office. יקח take אחר׃ let another}%
\verse{יהיו be בניו Let his children יתומים fatherless, ואשׁתו and his wife אלמנה׃ a widow.}%
\verse{ונוע be continually vagabonds, ינועו be continually vagabonds, בניו Let his children ושׁאלו and beg: ודרשׁו let them seek מחרבותיהם׃ also out of their desolate places.}%
\verse{ינקשׁ catch נושׁה Let the extortioner לכל all אשׁר that לו ויבזו spoil זרים he hath; and let the strangers יגיעו׃ his labor.}%
\verse{אל none יהי Let there be לו משׁך to extend חסד mercy ואל unto him: neither יהי let there be חונן any to favor ליתומיו׃ his fatherless children.}%
\verse{יהי be אחריתו Let his posterity להכרית cut off; בדור in the generation אחר following ימח be blotted out. שׁמם׃ let their name}%
\verse{יזכר be remembered עון Let the iniquity אבתיו of his fathers אל with יהוה the LORD; וחטאת the sin אמו of his mother אל and let not תמח׃ be blotted out.}%
\verse{יהיו Let them be נגד before יהוה the LORD תמיד continually, ויכרת that he may cut off מארץ of them from the earth. זכרם׃ the memory}%
\verse{יען Because אשׁר that לא not זכר he remembered עשׂות to show חסד mercy, וירדף but persecuted אישׁ man, עני the poor ואביון and needy ונכאה לבב in heart. למותת׃ that he might even slay}%
\verse{ויאהב As he loved קללה cursing, ותבואהו so let it come ולא not חפץ unto him: as he delighted בברכה in blessing, ותרחק so let it be far ממנו׃ from}%
\verse{וילבשׁ As he clothed קללה himself with cursing כמדו like as with his garment, ותבא so let it come כמים like water, בקרבו into his bowels וכשׁמן and like oil בעצמותיו׃ into his bones.}%
\verse{תהי Let it be לו כבגד unto him as the garment יעטה covereth ולמזח him, and for a girdle תמיד continually. יחגרה׃ wherewith he is girded}%
\verse{זאת this פעלת the reward שׂטני of mine adversaries מאת יהוה the LORD, והדברים and of them that speak רע evil על against נפשׁי׃ my soul.}%
\verse{ואתה thou יהוה me, O GOD אדני the Lord, עשׂה But do אתי for למען שׁמךי because טוב good, חסדך thy mercy הצילני׃ deliver}%
\verse{כי For עני poor ואביון and needy, אנכי I ולבי and my heart חלל is wounded בקרבי׃ within}%
\verse{כצל like the shadow כנטותו when it declineth: נהלכתי I am gone ננערתי I am tossed up and down כארבה׃ as the locust.}%
\verse{ברכי My knees כשׁלו are weak מצום through fasting; ובשׂרי and my flesh כחשׁ faileth משׁמן׃ of fatness.}%
\verse{ואני I הייתי became חרפה also a reproach להם יראוני unto them: they looked upon יניעון me they shaked ראשׁם׃ their heads.}%
\verse{עזרני Help יהוה me, O LORD אלהי my God: הושׁיעני O save כחסדך׃ me according to thy mercy:}%
\verse{וידעו they may know כי That ידך thy hand; זאת that this אתה thou, יהוה LORD, עשׂיתה׃ hast done}%
\verse{יקללו curse, המה Let them ואתה thou: תברך but bless קמו when they arise, ויבשׁו let them be ashamed; ועבדך but let thy servant ישׂמח׃ rejoice.}%
\verse{ילבשׁו be clothed שׂוטני Let mine adversaries כלמה with shame, ויעטו and let them cover כמעיל as with a mantle. בשׁתם׃ themselves with their own confusion,}%
\verse{אודה praise יהוה the LORD מאד I will greatly בפי with my mouth; ובתוך him among רבים the multitude. אהללנו׃ yea, I will praise}%
\verse{כי For יעמד he shall stand לימין at the right hand אביון of the poor, להושׁיע to save משׁפטי from those that condemn נפשׁו׃ his soul.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 110
\verseWithHeading{Adonai Gives Authority to His Messiah}{לדוד of David. מזמור A Psalm נאם said יהוה The LORD לאדני unto my Lord, שׁב Sit לימיני thou at my right hand, עד until אשׁית I make איביך thine enemies הדם thy footstool. לרגליך׃ thy footstool.}%
\verse{מטה the rod עזך of thy strength ישׁלח shall send יהוה The LORD מציון רדה rule בקרב thou in the midst איביך׃ of thine enemies.}%
\verse{עמך Thy people נדבת willing ביום in the day חילך of thy power, בהדרי in the beauties קדשׁ of holiness מרחם from the womb משׁחר of the morning: לך טל thou hast the dew ילדתיך׃ of thy youth.}%
\verse{נשׁבע hath sworn, יהוה The LORD ולא and will not ינחם repent, אתה Thou כהן a priest לעולם forever על after דברתי the order מלכי צדק׃ of Melchizedek.}%
\verse{אדני The Lord על at ימינך thy right hand מחץ shall strike through ביום in the day אפו of his wrath. מלכים׃ kings}%
\verse{ידין He shall judge בגוים among the heathen, מלא he shall fill גויות with the dead bodies; מחץ he shall wound ראשׁ the heads על over ארץ countries. רבה׃ many}%
\verse{מנחל of the brook בדרך in the way: ישׁתה He shall drink על therefore כן therefore ירים shall he lift up ראשׁ׃ the head.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 111
\verseWithHeading{Praise to God for His Work and Commands}{הללו Praise יה ye the LORD. אודה I will praise יהוה the LORD בכל with whole לבב heart, בסוד in the assembly ישׁרים of the upright, ועדה׃ and the congregation.}%
\verse{גדלים great, מעשׂי The works יהוה of the LORD דרושׁים sought out לכל of all חפציהם׃ them that have pleasure}%
\verse{הוד honorable והדר and glorious: פעלו His work וצדקתו and his righteousness עמדת endureth לעד׃ forever.}%
\verse{זכר to be remembered: עשׂה He hath made לנפלאתיו his wonderful works חנון gracious ורחום and full of compassion. יהוה׃ the LORD}%
\verse{טרף meat נתן He hath given ליראיו unto them that fear יזכר be mindful לעולם him: he will ever בריתו׃ of his covenant.}%
\verse{כח the power מעשׂיו of his works, הגיד He hath showed לעמו his people לתת that he may give להם נחלת them the heritage גוים׃ of the heathen.}%
\verse{מעשׂי The works ידיו of his hands אמת verity ומשׁפט and judgment; נאמנים sure. כל all פקודיו׃ his commandments}%
\verse{סמוכים They stand fast לעד forever לעולם and ever, עשׂוים done באמת in truth וישׁר׃ and uprightness.}%
\verse{פדות redemption שׁלח He sent לעמו unto his people: צוה he hath commanded לעולם forever: בריתו his covenant קדושׁ holy ונורא and reverend שׁמו׃ his name.}%
\verse{ראשׁית the beginning חכמה of wisdom: יראת The fear יהוה of the LORD שׂכל understanding טוב a good לכל have all עשׂיהם they that do תהלתו his praise עמדת endureth לעד׃ forever.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 112
\verseWithHeading{The Path of the Righteous and the Path of the Wicked}{הללו Praise יה ye the LORD. אשׁרי Blessed אישׁ the man ירא feareth את יהוה the LORD, במצותיו in his commandments. חפץ delighteth מאד׃ greatly}%
\verse{גבור mighty בארץ upon earth: יהיה shall be זרעו His seed דור the generation ישׁרים of the upright יברך׃ shall be blessed.}%
\verse{הון Wealth ועשׁר and riches בביתו in his house: וצדקתו and his righteousness עמדת endureth לעד׃ forever.}%
\verse{זרח there ariseth בחשׁך in the darkness: אור light לישׁרים Unto the upright חנון gracious, ורחום and full of compassion, וצדיק׃ and righteous.}%
\verse{טוב A good אישׁ man חונן showeth favor, ומלוה and lendeth: יכלכל he will guide דבריו his affairs במשׁפט׃ with discretion.}%
\verse{כי Surely לעולם forever: לא he shall not ימוט be moved לזכר remembrance. עולם in everlasting יהיה shall be צדיק׃ the righteous}%
\verse{משׁמועה tidings: רעה of evil לא He shall not יירא be afraid נכון is fixed, לבו his heart בטח trusting ביהוה׃ in the LORD.}%
\verse{סמוך established, לבו His heart לא he shall not יירא be afraid, עד until אשׁר until יראה he see בצריו׃ upon his enemies.}%
\verse{פזר He hath dispersed, נתן he hath given לאביונים to the poor; צדקתו his righteousness עמדת endureth לעד forever; קרנו his horn תרום shall be exalted בכבוד׃ with honor.}%
\verse{רשׁע The wicked יראה shall see וכעס and be grieved; שׁניו with his teeth, יחרק he shall gnash ונמס and melt away: תאות the desire רשׁעים of the wicked תאבד׃ shall perish.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 113
\verseWithHeading{God’s Majesty and Care for the Needy}{הללו Praise יה ye the LORD. הללו Praise, עבדי O ye servants יהוה of the LORD, הללו praise את שׁם the name יהוה׃ of the LORD.}%
\verse{יהי be שׁם the name יהוה of the LORD מברך Blessed מעתה from this time forth ועד and forevermore. עולם׃ and forevermore.}%
\verse{ממזרח שׁמשׁ of the sun עד unto מבואו the going down מהלל to be praised. שׁם name יהוה׃ of the same the LORD's}%
\verse{רם high על above כל all גוים nations, יהוה The LORD על above השׁמים the heavens. כבודו׃ his glory}%
\verse{מי Who כיהוה like unto the LORD אלהינו our God, המגביהי on high, לשׁבת׃ who dwelleth}%
\verse{המשׁפילי Who humbleth לראות to behold בשׁמים in heaven, ובארץ׃ and in the earth!}%
\verse{מקימי He raiseth up מעפר out of the dust, דל the poor מאשׁפת out of the dunghill; ירים lifteth אביון׃ the needy}%
\verse{להושׁיבי That he may set עם with נדיבים princes, עם with נדיבי the princes עמו׃ of his people.}%
\verse{מושׁיבי woman to keep עקרת He maketh the barren הבית house, אם mother הבנים of children. שׂמחה a joyful הללו Praise יה׃ ye the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 114
\verseWithHeading{Praise to God for His Works During the Exodus}{בצאת went out ישׂראל When Israel ממצרים בית the house יעקב of Jacob מעם from a people לעז׃ of strange language;}%
\verse{היתה was יהודה Judah לקדשׁו his sanctuary, ישׂראל Israel ממשׁלותיו׃ his dominion.}%
\verse{הים The sea ראה saw וינס and fled: הירדן Jordan יסב was driven לאחור׃ back.}%
\verse{ההרים The mountains רקדו skipped כאילים like rams, גבעות the little hills כבני like lambs. צאן׃ like lambs.}%
\verse{מה What לך הים thee, O thou sea, כי that תנוס thou fleddest? הירדן thou Jordan, תסב thou wast driven לאחור׃ back?}%
\verse{ההרים Ye mountains, תרקדו ye skipped כאילים like rams; גבעות ye little hills, כבני like lambs? צאן׃ like lambs?}%
\verse{מלפני at the presence אדון of the Lord, חולי Tremble, ארץ thou earth, מלפני at the presence אלוה of the God יעקב׃ of Jacob;}%
\verse{ההפכי Which turned הצור the rock אגם a standing מים water, חלמישׁ the flint למעינו into a fountain מים׃ of waters.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 115
\verseWithHeading{Dead Idols and the Living God}{לא Not לנו יהוה unto us, O LORD, לא not לנו כי unto us, but לשׁמך unto thy name תן give כבוד glory, על for חסדך thy mercy, על for אמתך׃ thy truth's}%
\verse{למה Wherefore יאמרו say, הגוים should the heathen איה Where נא now אלהיהם׃ their God?}%
\verse{ואלהינו But our God בשׁמים in the heavens: כל whatsoever אשׁר whatsoever חפץ he hath pleased. עשׂה׃ he hath done}%
\verse{עצביהם Their idols כסף silver וזהב and gold, מעשׂה the work ידי hands. אדם׃ of men's}%
\verse{פה They have mouths, להם ולא not: ידברו but they speak עינים eyes להם ולא not: יראו׃ have they, but they see}%
\verse{אזנים They have ears, להם ולא not: ישׁמעו but they hear אף noses להם ולא not: יריחון׃ have they, but they smell}%
\verse{ידיהם They have hands, ולא not: ימישׁון but they handle רגליהם feet ולא not: יהלכו have they, but they walk לא neither יהגו speak בגרונם׃ they through their throat.}%
\verse{כמוהם like unto them; יהיו them are עשׂיהם They that make כל every one אשׁר that בטח׃ trusteth}%
\verse{ישׂראל O Israel, בטח trust ביהוה thou in the LORD: עזרם their help ומגנם and their shield. הוא׃ he}%
\verse{בית O house אהרן of Aaron, בטחו trust ביהוה in the LORD: עזרם their help ומגנם and their shield. הוא׃ he}%
\verse{יראי Ye that fear יהוה the LORD, בטחו trust ביהוה in the LORD: עזרם their help ומגנם and their shield. הוא׃ he}%
\verse{יהוה The LORD זכרנו hath been mindful יברך of us: he will bless יברך he will bless את בית the house ישׂראל of Israel; יברך he will bless את בית the house אהרן׃ of Aaron.}%
\verse{יברך He will bless יראי them that fear יהוה the LORD, הקטנים small עם and הגדלים׃ great.}%
\verse{יסף shall increase you more and more, יהוה The LORD עליכם shall increase you more and more, עליכם ועלניכם׃ you and your children.}%
\verse{ברוכים blessed אתם Ye ליהוה of the LORD עשׂה which made שׁמים heaven וארץ׃ and earth.}%
\verse{השׁמים The heaven, שׁמים the heavens, ליהוה the LORD's: והארץ but the earth נתן hath he given לבני to the children אדם׃ of men.}%
\verse{לא not המתים The dead יהללו praise יה the LORD, ולא neither כל any ירדי that go down דומה׃ into silence.}%
\verse{ואנחנו But we נברך will bless יה the LORD מעתה from this time forth ועד and forevermore. עולם and forevermore. הללו Praise יה׃ the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 116
\verseWithHeading{Thanksgiving for God’s Deliverance}{אהבתי I love כי because ישׁמע he hath heard יהוה the LORD, את קולי my voice תחנוני׃ my supplications.}%
\verse{כי Because הטה he hath inclined אזנו his ear לי ובימי as long as I live. אקרא׃ unto me, therefore will I call upon}%
\verse{אפפוני compassed חבלי The sorrows מות of death ומצרי me, and the pains שׁאול of hell מצאוני got hold upon צרה trouble ויגון and sorrow. אמצא׃ me: I found}%
\verse{ובשׁם I upon the name יהוה of the LORD; אקרא Then called אנה I beseech יהוה O LORD, מלטה thee, deliver נפשׁי׃ my soul.}%
\verse{חנון Gracious יהוה the LORD, וצדיק and righteous; ואלהינו yea, our God מרחם׃ merciful.}%
\verse{שׁמר preserveth פתאים the simple: יהוה The LORD דלותי I was brought low, ולי יהושׁיע׃ and he helped}%
\verse{שׁובי Return נפשׁי O my soul; למנוחיכי כי for יהוה the LORD גמל hath dealt bountifully עליכי׃ with}%
\verse{כי For חלצת thou hast delivered נפשׁי my soul ממות from death, את עיני mine eyes מן from death, דמעה tears, את רגלי my feet מדחי׃ from falling.}%
\verse{אתהלך I will walk לפני before יהוה the LORD בארצות in the land החיים׃ of the living.}%
\verse{האמנתי I believed, כי therefore אדבר spoken: אני have I עניתי afflicted: מאד׃ I was greatly}%
\verse{אני I אמרתי said בחפזי in my haste, כל All האדם men כזב׃ liars.}%
\verse{מה What אשׁיב shall I render ליהוה unto the LORD כל all תגמולוהי his benefits עלי׃ toward}%
\verse{כוס the cup ישׁועות of salvation, אשׂא I will take ובשׁם upon the name יהוה of the LORD. אקרא׃ and call}%
\verse{נדרי my vows ליהוה unto the LORD אשׁלם I will pay נגדה in the presence נא now לכל of all עמו׃ his people.}%
\verse{יקר Precious בעיני in the sight יהוה of the LORD המותה the death לחסידיו׃ of his saints.}%
\verse{אנה truly יהוה O LORD, כי אני I עבדך thy servant; אני I עבדך thy servant, בן the son אמתך of thine handmaid: פתחת thou hast loosed למוסרי׃ my bonds.}%
\verse{לך אזבח I will offer זבח to thee the sacrifice תודה of thanksgiving, ובשׁם upon the name יהוה of the LORD. אקרא׃ and will call}%
\verse{נדרי my vows ליהוה unto the LORD אשׁלם I will pay נגדה in the presence נא now לכל of all עמו׃ his people,}%
\verse{בחצרות In the courts בית house, יהוה of the LORD's בתוככי in the midst ירושׁלם of thee, O Jerusalem. הללו Praise יה׃ ye the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 117
\verseWithHeading{Let All Peoples Praise Adonai}{הללו O praise את יהוה the LORD, כל all גוים ye nations: שׁבחוהו praise כל him, all האמים׃ ye people.}%
\verse{כי For גבר is great עלינו toward חסדו his merciful kindness ואמת us: and the truth יהוה of the LORD לעולם forever. הללו Praise יה׃ ye the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 118
\verseWithHeading{Praise to God for His Loyal Love}{הודו O give thanks ליהוה unto the LORD; כי for טוב good: כי because לעולם forever. חסדו׃ his mercy}%
\verse{יאמר say, נא now ישׂראל Let Israel כי that לעולם forever. חסדו׃ his mercy}%
\verse{יאמרו say, נא now בית Let the house אהרן of Aaron כי that לעולם forever. חסדו׃ his mercy}%
\verse{יאמרו say, נא Let them now יראי that fear יהוה the LORD כי that לעולם forever. חסדו׃ his mercy}%
\verse{מן in המצר distress: קראתי I called upon יה the LORD ענני answered במרחב me, in a large place. יה׃ the LORD}%
\verse{יהוה The LORD לי לא on my side; I will not אירא fear: מה what יעשׂה do לי אדם׃ can man}%
\verse{יהוה The LORD לי בעזרי taketh my part with them that help ואני me: therefore shall I אראה see בשׂנאי׃ upon them that hate}%
\verse{טוב better לחסות to trust ביהוה in the LORD מבטח than to put confidence באדם׃ in man.}%
\verse{טוב better לחסות to trust ביהוה in the LORD מבטח than to put confidence בנדיבים׃ in princes.}%
\verse{כל All גוים nations סבבוני compassed me about: בשׁם but in the name יהוה of the LORD כי אמילם׃ will I destroy}%
\verse{סבוני They compassed me about; גם yea, סבבוני they compassed me about: בשׁם but in the name יהוה of the LORD כי אמילם׃ I will destroy}%
\verse{סבוני They compassed me about כדבורים like bees; דעכו they are quenched כאשׁ as the fire קוצים of thorns: בשׁם for in the name יהוה of the LORD כי אמילם׃ I will destroy}%
\verse{דחה דחיתנינפל at me that I might fall: ויהוה but the LORD עזרני׃ helped}%
\verse{עזי my strength וזמרת and song, יה The LORD ויהי and is become לי לישׁועה׃ my salvation.}%
\verse{קול The voice רנה of rejoicing וישׁועה and salvation באהלי in the tabernacles צדיקים of the righteous: ימין the right hand יהוה of the LORD עשׂה doeth חיל׃ valiantly.}%
\verse{ימין The right hand יהוה of the LORD רוממה ימין the right hand יהוה of the LORD עשׂה doeth חיל׃ valiantly.}%
\verse{לא I shall not אמות die, כי but אחיה live, ואספר and declare מעשׂי the works יה׃ of the LORD.}%
\verse{יסר hath chastened me sore: יסרני hath chastened me sore: יה The LORD ולמות unto death. לא but he hath not נתנני׃ given me over}%
\verse{פתחו Open לי שׁערי to me the gates צדק of righteousness: אבא I will go בם אודה into them, I will praise יה׃ the LORD:}%
\verse{זה This השׁער gate ליהוה of the LORD, צדיקים into which the righteous יבאו׃ shall enter.}%
\verse{אודך I will praise כי thee: for עניתני thou hast heard ותהי me, and art become לי לישׁועה׃ my salvation.}%
\verse{אבן The stone מאסו refused הבונים the builders היתה is become לראשׁ the head פנה׃ of the corner.}%
\verse{מאת יהוה the LORD's היתה is זאת This היא it נפלאת marvelous בעינינו׃ in our eyes.}%
\verse{זה This היום the day עשׂה hath made; יהוה the LORD נגילה we will rejoice ונשׂמחה׃ and be glad}%
\verse{אנא I beseech יהוה thee, O LORD: הושׁיעה Save נא now, אנא I beseech יהוה O LORD, הצליחה prosperity. נא׃ thee, send now}%
\verse{ברוך Blessed הבא he that cometh בשׁם in the name יהוה of the LORD: ברכנוכם we have blessed מבית you out of the house יהוה׃ of the LORD.}%
\verse{אל God יהוה the LORD, ויאר לנו אסרו bind חג the sacrifice בעבתים with cords, עד unto קרנות the horns המזבח׃ of the altar.}%
\verse{אלי my God, אתה Thou ואודך and I will praise אלהי thee: my God, ארוממך׃ I will exalt}%
\verse{הודו O give thanks ליהוה unto the LORD; כי for טוב good: כי for לעולם forever. חסדו׃ his mercy}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 119
\verseWithHeading{Aleph}{אשׁרי Blessed תמימי the undefiled דרך in the way, ההלכים who walk בתורת in the law יהוה׃}%
\verse{אשׁרי Blessed נצרי they that keep עדתיו his testimonies, בכל him with the whole לב heart. ידרשׁוהו׃ seek}%
\verse{אף They also לא no פעלו do עולה iniquity: בדרכיו in his ways. הלכו׃ they walk}%
\verse{אתה Thou צויתה hast commanded פקדיך thy precepts לשׁמר to keep מאד׃ diligently.}%
\verse{אחלי O that יכנו were directed דרכי my ways לשׁמר to keep חקיך׃ thy statutes!}%
\verse{אז Then לא shall I not אבושׁ be ashamed, בהביטי when I have respect אל unto כל all מצותיך׃ thy commandments.}%
\verse{אודך I will praise בישׁר thee with uprightness לבב of heart, בלמדי when I shall have learned משׁפטי judgments. צדקך׃ thy righteous}%
\verse{את חקיך thy statutes: אשׁמר I will keep אל me not תעזבני O forsake עד utterly. מאד׃ utterly.}%
\verseWithHeading{Beth}{במה Wherewithal יזכה cleanse נער shall a young man את ארחו his way? לשׁמר by taking heed כדברך׃ according to thy word.}%
\verse{בכל With my whole לבי heart דרשׁתיך have I sought אל thee: O let me not תשׁגני wander ממצותיך׃ from thy commandments.}%
\verse{בלבי in mine heart, צפנתי have I hid אמרתך Thy word למען that לא I might not אחטא׃ sin}%
\verse{ברוך Blessed אתה thou, יהוה למדני teach חקיך׃ me thy statutes.}%
\verse{בשׂפתי With my lips ספרתי have I declared כל all משׁפטי the judgments פיך׃ of thy mouth.}%
\verse{בדרך in the way עדותיך of thy testimonies, שׂשׂתי I have rejoiced כעל as כל in all הון׃ riches.}%
\verse{בפקדיך in thy precepts, אשׂיחה I will meditate ואביטה and have respect ארחתיך׃ unto thy ways.}%
\verse{בחקתיך in thy statutes: אשׁתעשׁע I will delight myself לא I will not אשׁכח forget דברך׃ thy word.}%
\verseWithHeading{Gimel}{גמל Deal bountifully על with עבדך thy servant, אחיה I may live, ואשׁמרה and keep דברך׃ thy word.}%
\verse{גל Open עיני thou mine eyes, ואביטה that I may behold נפלאות wondrous things מתורתך׃ out of thy law.}%
\verse{גר a stranger אנכי I בארץ in the earth: אל not תסתר hide ממני from מצותיך׃ thy commandments}%
\verse{גרסה breaketh נפשׁי My soul לתאבה for the longing אל unto משׁפטיך thy judgments בכל at all עת׃ times.}%
\verse{גערת Thou hast rebuked זדים the proud ארורים cursed, השׁגים which do err ממצותיך׃ from thy commandments.}%
\verse{גל Remove מעלי from חרפה me reproach ובוז and contempt; כי for עדתיך thy testimonies. נצרתי׃ I have kept}%
\verse{גם also ישׁבו did sit שׂרים Princes בי נדברו speak עבדך against me: thy servant ישׂיח did meditate בחקיך׃ in thy statutes.}%
\verse{גם also עדתיך Thy testimonies שׁעשׁעי my delight אנשׁי עצתי׃ my counselors.}%
\verseWithHeading{Daleth}{דבקה cleaveth לעפר unto the dust: נפשׁי My soul חיני quicken כדברך׃ thou me according to thy word.}%
\verse{דרכי my ways, ספרתי I have declared ותענני and thou heardest למדני me: teach חקיך׃ me thy statutes.}%
\verse{דרך the way פקודיך of thy precepts: הבינני Make me to understand ואשׂיחה so shall I talk בנפלאותיך׃ of thy wondrous works.}%
\verse{דלפה melteth נפשׁי My soul מתוגה for heaviness: קימני strengthen כדברך׃ thou me according unto thy word.}%
\verse{דרך me the way שׁקר of lying: הסר Remove ממני from ותורתך and grant me thy law חנני׃ graciously.}%
\verse{דרך the way אמונה of truth: בחרתי I have chosen משׁפטיך thy judgments שׁויתי׃ have I laid}%
\verse{דבקתי I have stuck בעדותיך unto thy testimonies: יהוה אל put me not תבישׁני׃ to shame.}%
\verse{דרך the way מצותיך of thy commandments, ארוץ I will run כי when תרחיב thou shalt enlarge לבי׃ my heart.}%
\verseWithHeading{Heʼ}{הורני Teach יהוה דרך the way חקיך of thy statutes; ואצרנה and I shall keep עקב׃ it the end.}%
\verse{הבינני Give me understanding, ואצרה and I shall keep תורתך thy law; ואשׁמרנה yea, I shall observe בכל it with whole לב׃ heart.}%
\verse{הדריכני Make me to go בנתיב in the path מצותיך of thy commandments; כי for בו חפצתי׃ therein do I delight.}%
\verse{הט Incline לבי my heart אל unto עדותיך thy testimonies, ואל and not אל to בצע׃ covetousness.}%
\verse{העבר Turn away עיני mine eyes מראות from beholding שׁוא vanity; בדרכך thou me in thy way. חיני׃ quicken}%
\verse{הקם Establish לעבדך unto thy servant, אמרתך thy word אשׁר who ליראתך׃ to thy fear.}%
\verse{העבר Turn away חרפתי my reproach אשׁר which יגרתי I fear: כי for משׁפטיך thy judgments טובים׃ good.}%
\verse{הנה Behold, תאבתי I have longed לפקדיך after thy precepts: בצדקתך me in thy righteousness. חיני׃ quicken}%
\verseWithHeading{Waw}{ויבאני come חסדך Let thy mercies יהוה תשׁועתך thy salvation, כאמרתך׃ according to thy word.}%
\verse{ואענה to answer חרפי him that reproacheth דבר So shall I have wherewith כי me: for בטחתי I trust בדברך׃ in thy word.}%
\verse{ואל not תצל And take מפי out of my mouth; דבר the word אמת of truth עד utterly מאד utterly כי for למשׁפטך in thy judgments. יחלתי׃ I have hoped}%
\verse{ואשׁמרה So shall I keep תורתך thy law תמיד continually לעולם forever ועד׃ and ever.}%
\verse{ואתהלכה And I will walk ברחבה at liberty: כי for פקדיך thy precepts. דרשׁתי׃ I seek}%
\verse{ואדברה I will speak בעדתיך of thy testimonies נגד also before מלכים kings, ולא and will not אבושׁ׃ be ashamed.}%
\verse{ואשׁתעשׁע And I will delight myself במצותיך in thy commandments, אשׁר which אהבתי׃ I have loved.}%
\verse{ואשׂא also will I lift up כפי My hands אל unto מצותיך thy commandments, אשׁר which אהבתי I have loved; ואשׂיחה and I will meditate בחקיך׃ in thy statutes.}%
\verseWithHeading{Zayin}{זכר Remember דבר the word לעבדך unto thy servant, על upon אשׁר which יחלתני׃ thou hast caused me to hope.}%
\verse{זאת This נחמתי my comfort בעניי in my affliction: כי for אמרתך thy word חיתני׃ hath quickened}%
\verse{זדים The proud הליצני in derision: עד have had me greatly מאד have had me greatly מתורתך from thy law. לא have I not נטיתי׃ declined}%
\verse{זכרתי I remembered משׁפטיך thy judgments מעולם of old, יהוה ואתנחם׃ and have comforted myself.}%
\verse{זלעפה Horror אחזתני hath taken hold upon מרשׁעים me because of the wicked עזבי that forsake תורתך׃ thy law.}%
\verse{זמרות my songs היו have been לי חקיך Thy statutes בבית in the house מגורי׃ of my pilgrimage.}%
\verse{זכרתי I have remembered בלילה in the night, שׁמך thy name, יהוה ואשׁמרה and have kept תורתך׃ thy law.}%
\verse{זאת This היתה I had, לי כי because פקדיך thy precepts. נצרתי׃ I kept}%
\verseWithHeading{Heth}{חלקי my portion, יהוה אמרתי I have said לשׁמר that I would keep דבריך׃ thy words.}%
\verse{חליתי I entreated פניך thy favor בכל with whole לב heart: חנני be merciful כאמרתך׃ unto me according to thy word.}%
\verse{חשׁבתי I thought דרכי on my ways, ואשׁיבה and turned רגלי my feet אל unto עדתיך׃ thy testimonies.}%
\verse{חשׁתי I made haste, ולא not התמהמהתי and delayed לשׁמר to keep מצותיך׃ thy commandments.}%
\verse{חבלי The bands רשׁעים of the wicked עודני have robbed תורתך thy law. לא me: I have not שׁכחתי׃ forgotten}%
\verse{חצות לילהקום I will rise להודות to give thanks לך על unto thee because of משׁפטי judgments. צדקך׃ thy righteous}%
\verse{חבר a companion אני I לכל of all אשׁר that יראוך fear ולשׁמרי thee, and of them that keep פקודיך׃ thy precepts.}%
\verse{חסדך of thy mercy: יהוה מלאה is full הארץ The earth, חקיך me thy statutes. למדני׃ teach}%
\verseWithHeading{Teth}{טוב well עשׂית Thou hast dealt עם with עבדך thy servant, יהוה כדברך׃ according unto thy word.}%
\verse{טוב me good טעם judgment ודעת and knowledge: למדני Teach כי for במצותיך thy commandments. האמנתי׃ I have believed}%
\verse{טרם Before אענה was afflicted אני I שׁגג I went astray: ועתה but now אמרתך thy word. שׁמרתי׃ have I kept}%
\verse{טוב good, אתה Thou ומטיב and doest good; למדני teach חקיך׃ me thy statutes.}%
\verse{טפלו have forged עלי against שׁקר a lie זדים The proud אני me: I בכל with whole לב heart. אצר will keep פקודיך׃ thy precepts}%
\verse{טפשׁ is as fat כחלב as grease; לבם Their heart אני I תורתך in thy law. שׁעשׁעתי׃ delight}%
\verse{טוב good לי כי for me that עניתי I have been afflicted; למען that אלמד I might learn חקיך׃ thy statutes.}%
\verse{טוב better לי תורת The law פיך of thy mouth מאלפי unto me than thousands זהב of gold וכסף׃ and silver.}%
\verseWithHeading{Yod}{ידיך Thy hands עשׂוני have made ויכוננוני me and fashioned הבינני me: give me understanding, ואלמדה that I may learn מצותיך׃ thy commandments.}%
\verse{יראיך They that fear יראוני when they see וישׂמחו thee will be glad כי me; because לדברך in thy word. יחלתי׃ I have hoped}%
\verse{ידעתי I know, יהוה כי that צדק right, משׁפטיך thy judgments ואמונה and thou in faithfulness עניתני׃ hast afflicted}%
\verse{יהי be נא Let, I pray thee, חסדך thy merciful kindness לנחמני for my comfort, כאמרתך according to thy word לעבדך׃ unto thy servant.}%
\verse{יבאוני come רחמיך Let thy tender mercies ואחיה unto me, that I may live: כי for תורתך thy law שׁעשׁעי׃ my delight.}%
\verse{יבשׁו be ashamed; זדים Let the proud כי for שׁקר with me without a cause: עותוני they dealt perversely אני I אשׂיח will meditate בפקודיך׃ in thy precepts.}%
\verse{ישׁובו thee turn לי יראיך Let those that fear וידעו unto me, and those that have known עדתיך׃ thy testimonies.}%
\verse{יהי be לבי Let my heart תמים sound בחקיך in thy statutes; למען that לא I be not אבושׁ׃ ashamed.}%
\verseWithHeading{Kaph}{כלתה fainteth לתשׁועתך for thy salvation: נפשׁי My soul לדברך in thy word. יחלתי׃ I hope}%
\verse{כלו fail עיני Mine eyes לאמרתך for thy word, לאמר saying, מתי When תנחמני׃ wilt thou comfort}%
\verse{כי For הייתי I am become כנאד like a bottle בקיטור in the smoke; חקיך thy statutes. לא do I not שׁכחתי׃ forget}%
\verse{כמה How many ימי the days עבדך of thy servant? מתי when תעשׂה wilt thou execute ברדפי on them that persecute משׁפט׃ judgment}%
\verse{כרו have digged לי זדים The proud שׁיחות pits אשׁר for me, which לא not כתורתך׃ after thy law.}%
\verse{כל All מצותיך thy commandments אמונה faithful: שׁקר me wrongfully; רדפוני they persecute עזרני׃ help}%
\verse{כמעט They had almost כלוני consumed בארץ me upon earth; ואני but I לא not עזבתי forsook פקודיך׃ thy precepts.}%
\verse{כחסדך me after thy lovingkindness; חיני Quicken ואשׁמרה so shall I keep עדות the testimony פיך׃ of thy mouth.}%
\verseWithHeading{Lamed}{לעולם Forever, יהוה דברך thy word נצב is settled בשׁמים׃ in heaven.}%
\verse{לדר unto all generations: ודר unto all generations: אמונתך Thy faithfulness כוננת thou hast established ארץ the earth, ותעמד׃ and it abideth.}%
\verse{למשׁפטיך according to thine ordinances: עמדו They continue היום this day כי for הכל all עבדיך׃ thy servants.}%
\verse{לולי Unless תורתך thy law שׁעשׁעי my delights, אז I should then אבדתי have perished בעניי׃ in mine affliction.}%
\verse{לעולם לאשׁכח forget פקודיך thy precepts: כי for בם חייתני׃ with them thou hast quickened}%
\verse{לך אני I הושׁיעני thine, save כי me; for פקודיך thy precepts. דרשׁתי׃ I have sought}%
\verse{לי קוו have waited רשׁעים The wicked לאבדני for me to destroy עדתיך thy testimonies. אתבונן׃ me: I will consider}%
\verse{לכל of all תכלה perfection: ראיתי I have seen קץ an end רחבה broad. מצותך thy commandment מאד׃ exceeding}%
\verseWithHeading{Mem}{מה O how אהבתי love תורתך I thy law! כל all היום the day. היא it שׂיחתי׃ my meditation}%
\verse{מאיבי than mine enemies: תחכמני hast made me wiser מצותך Thou through thy commandments כי for לעולם ever היא׃ they}%
\verse{מכל than all מלמדי my teachers: השׂכלתי I have more understanding כי for עדותיך thy testimonies שׂיחה׃ my meditation.}%
\verse{מזקנים more than the ancients, אתבונן I understand כי because פקודיך thy precepts. נצרתי׃ I keep}%
\verse{מכל from every ארח way, רע evil כלאתי I have refrained רגלי my feet למען that אשׁמר I might keep דברך׃ thy word.}%
\verse{ממשׁפטיך from thy judgments: לא I have not סרתי departed כי for אתה thou הורתני׃ hast taught}%
\verse{מה How נמלצו sweet are לחכי unto my taste! אמרתך thy words מדבשׁ than honey לפי׃ to my mouth!}%
\verse{מפקודיך אתבונן I get understanding: על therefore כן therefore שׂנאתי I hate כל every ארח way. שׁקר׃ false}%
\verseWithHeading{Nun}{נר a lamp לרגלי unto my feet, דברך Thy word ואור and a light לנתיבתי׃ unto my path.}%
\verse{נשׁבעתי I have sworn, ואקימה and I will perform לשׁמר that I will keep משׁפטי judgments. צדקך׃ thy righteous}%
\verse{נעניתי I am afflicted עד very much: מאד very much: יהוה חיני quicken כדברך׃ according unto thy word.}%
\verse{נדבות the freewill offerings פי of my mouth, רצה Accept, נא I beseech thee, יהוה ומשׁפטיך me thy judgments. למדני׃ and teach}%
\verse{נפשׁי My soul בכפי in my hand: תמיד continually ותורתך thy law. לא yet do I not שׁכחתי׃ forget}%
\verse{נתנו have laid רשׁעים The wicked פח a snare לי ומפקודיך from thy precepts. לא not תעיתי׃ for me: yet I erred}%
\verse{נחלתי have I taken as a heritage עדותיך Thy testimonies לעולם forever: כי for שׂשׂון the rejoicing לבי of my heart. המה׃ they}%
\verse{נטיתי I have inclined לבי mine heart לעשׂות to perform חקיך thy statutes לעולם always, עקב׃ the end.}%
\verseWithHeading{Samek}{סעפים thoughts: שׂנאתי I hate ותורתך but thy law אהבתי׃ do I love.}%
\verse{סתרי my hiding place ומגני and my shield: אתה Thou לדברך in thy word. יחלתי׃ I hope}%
\verse{סורו Depart ממני from מרעים me, ye evildoers: ואצרה for I will keep מצות the commandments אלהי׃ of my God.}%
\verse{סמכני Uphold כאמרתך me according unto thy word, ואחיה that I may live: ואל and let me not תבישׁני be ashamed משׂברי׃ of my hope.}%
\verse{סעדני Hold thou me up, ואושׁעה and I shall be safe: ואשׁעה and I will have respect בחקיך unto thy statutes תמיד׃ continually.}%
\verse{סלית Thou hast trodden down כל all שׁוגים them that err מחקיך from thy statutes: כי for שׁקר falsehood. תרמיתם׃ their deceit}%
\verse{סגים dross: השׁבת Thou puttest away כל all רשׁעי the wicked ארץ of the earth לכן therefore אהבתי I love עדתיך׃ thy testimonies.}%
\verse{סמר trembleth מפחדך for fear בשׂרי My flesh וממשׁפטיך of thy judgments. יראתי׃ of thee; and I am afraid}%
\verseWithHeading{ʽAyin}{עשׂיתי I have done משׁפט judgment וצדק and justice: בל me not תניחני לעשׁקי׃ to mine oppressors.}%
\verse{ערב Be surety עבדך for thy servant לטוב for good: אל let not יעשׁקני oppress זדים׃ the proud}%
\verse{עיני Mine eyes כלו fail לישׁועתך for thy salvation, ולאמרת and for the word צדקך׃ of thy righteousness.}%
\verse{עשׂה Deal עם with עבדך thy servant כחסדך according unto thy mercy, וחקיך me thy statutes. למדני׃ and teach}%
\verse{עבדך thy servant; אני I הבינני give me understanding, ואדעה that I may know עדתיך׃ thy testimonies.}%
\verse{עת time לעשׂות to work: ליהוה הפרו they have made void תורתך׃ thy law.}%
\verse{על כןהבתי I love מצותיך thy commandments מזהב above gold; ומפז׃ yea, above fine gold.}%
\verse{על כןל I esteem all פקודי precepts כל all ישׁרתי right; כל every ארח way. שׁקר false שׂנאתי׃ I hate}%
\verseWithHeading{Pe}{פלאות wonderful: עדותיך Thy testimonies על therefore כן therefore נצרתם keep נפשׁי׃ doth my soul}%
\verse{פתח The entrance דבריך of thy words יאיר giveth light; מבין it giveth understanding פתיים׃ unto the simple.}%
\verse{פי my mouth, פערתי I opened ואשׁאפה and panted: כי for למצותיך for thy commandments. יאבתי׃ I longed}%
\verse{פנה Look אלי thou upon וחנני me, and be merciful כמשׁפט unto me, as thou usest to do לאהבי unto those that love שׁמך׃ thy name.}%
\verse{פעמי my steps הכן Order באמרתך in thy word: ואל and let not תשׁלט have dominion בי כל any און׃ iniquity}%
\verse{פדני Deliver מעשׁק me from the oppression אדם of man: ואשׁמרה so will I keep פקודיך׃ thy precepts.}%
\verse{פניך Make thy face האר to shine בעבדך upon thy servant; ולמדני and teach את חקיך׃ me thy statutes.}%
\verse{פלגי Rivers מים of waters ירדו run down עיני mine eyes, על because לא not שׁמרו they keep תורתך׃ thy law.}%
\verseWithHeading{Tsade}{צדיק Righteous אתה thou, יהוה וישׁר and upright משׁפטיך׃ thy judgments.}%
\verse{צוית thou hast commanded צדק righteous עדתיך Thy testimonies ואמונה faithful. מאד׃ and very}%
\verse{צמתתני hath consumed קנאתי My zeal כי me, because שׁכחו have forgotten דבריך thy words. צרי׃ mine enemies}%
\verse{צרופה pure: אמרתך Thy word מאד very ועבדך therefore thy servant אהבה׃ loveth}%
\verse{צעיר small אנכי I ונבזה and despised: פקדיך thy precepts. לא do not שׁכחתי׃ I forget}%
\verse{צדקתך Thy righteousness צדק righteousness, לעולם an everlasting ותורתך and thy law אמת׃ the truth.}%
\verse{צר Trouble ומצוק and anguish מצאוני have taken hold on מצותיך me: thy commandments שׁעשׁעי׃ my delights.}%
\verse{צדק The righteousness עדותיך of thy testimonies לעולם everlasting: הבינני give me understanding, ואחיה׃ and I shall live.}%
\verseWithHeading{Qoph}{קראתי I cried בכל with whole לב heart; ענני hear יהוה חקיך thy statutes. אצרה׃ I will keep}%
\verse{קראתיך I cried הושׁיעני unto thee; save ואשׁמרה me, and I shall keep עדתיך׃ thy testimonies.}%
\verse{קדמתי I prevented בנשׁף the dawning ואשׁועה of the morning, and cried: לדבריך in thy word. יחלתי׃ I hoped}%
\verse{קדמו prevent עיני Mine eyes אשׁמרות the watches, לשׂיח that I might meditate באמרתך׃ in thy word.}%
\verse{קולי my voice שׁמעה Hear כחסדך according unto thy lovingkindness: יהוה כמשׁפטך me according to thy judgment. חיני׃ quicken}%
\verse{קרבו They draw nigh רדפי that follow after זמה mischief: מתורתך from thy law. רחקו׃ they are far}%
\verse{קרוב near, אתה Thou יהוה וכל and all מצותיך thy commandments אמת׃ truth.}%
\verse{קדם of old ידעתי I have known מעדתיך כי that לעולם them forever. יסדתם׃ thou hast founded}%
\verseWithHeading{Resh}{ראה Consider עניי mine affliction, וחלצני and deliver כי me: for תורתך thy law. לא I do not שׁכחתי׃ forget}%
\verse{ריבה Plead ריבי my cause, וגאלני and deliver לאמרתך me according to thy word. חיני׃ me: quicken}%
\verse{רחוק far מרשׁעים from the wicked: ישׁועה Salvation כי for חקיך thy statutes. לא not דרשׁו׃ they seek}%
\verse{רחמיך thy tender mercies, רבים Great יהוה כמשׁפטיך me according to thy judgments. חיני׃ quicken}%
\verse{רבים Many רדפי my persecutors וצרי and mine enemies; מעדותיך from thy testimonies. לא do I not נטיתי׃ decline}%
\verse{ראיתי I beheld בגדים the transgressors, ואתקוטטה and was grieved; אשׁר because אמרתך thy word. לא not שׁמרו׃ they kept}%
\verse{ראה Consider כי how פקודיך thy precepts: אהבתי I love יהוה כחסדך according to thy lovingkindness. חיני׃ quicken}%
\verse{ראשׁ the beginning: דברך Thy word אמת true ולעולם forever. כל and every one משׁפט judgments צדקך׃ of thy righteous}%
\verseWithHeading{Sin/Shin}{שׂרים Princes רדפוני have persecuted חנם me without a cause: ומדבריך of thy word. פחד standeth in awe לבי׃ but my heart}%
\verse{שׂשׂ rejoice אנכי I על at אמרתך thy word, כמוצא as one that findeth שׁלל spoil. רב׃ great}%
\verse{שׁקר lying: שׂנאתי I hate ואתעבה and abhor תורתך thy law אהבתי׃ do I love.}%
\verse{שׁבע Seven ביום times a day הללתיך do I praise על thee because of משׁפטי judgments. צדקך׃ thy righteous}%
\verse{שׁלום peace רב Great לאהבי have they which love תורתך thy law: ואין and nothing למו מכשׁול׃ shall offend}%
\verse{שׂברתי I have hoped לישׁועתך for thy salvation, יהוה ומצותיך thy commandments. עשׂיתי׃ and done}%
\verse{שׁמרה hath kept נפשׁי My soul עדתיך thy testimonies; ואהבם and I love מאד׃ them exceedingly.}%
\verse{שׁמרתי I have kept פקודיך thy precepts ועדתיך and thy testimonies: כי for כל all דרכי my ways נגדך׃ before}%
\verseWithHeading{Taw}{תקרב come near רנתי Let my cry לפניך before יהוה כדברך according to thy word. הבינני׃ give me understanding}%
\verse{תבוא come תחנתי Let my supplication לפניך before כאמרתך me according to thy word. הצילני׃ thee: deliver}%
\verse{תבענה shall utter שׂפתי My lips תהלה praise, כי when תלמדני thou hast taught חקיך׃ me thy statutes.}%
\verse{תען shall speak לשׁוני My tongue אמרתך of thy word: כי for כל all מצותיך thy commandments צדק׃ righteousness.}%
\verse{תהי Let ידך thine hand לעזרני help כי me; for פקודיך thy precepts. בחרתי׃ I have chosen}%
\verse{תאבתי I have longed לישׁועתך for thy salvation, יהוה ותורתך and thy law שׁעשׁעי׃ my delight.}%
\verse{תחי live, נפשׁי Let my soul ותהללך and it shall praise ומשׁפטך thee; and let thy judgments יעזרני׃ help}%
\verse{תעיתי I have gone astray כשׂה sheep; אבד like a lost בקשׁ seek עבדך thy servant; כי for מצותיך thy commandments. לא I do not שׁכחתי׃ forget}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 120
\verseWithHeading{Prayer for Deliverance from Enemies}{שׁיר A Song המעלות of degrees. אל unto יהוה the LORD, בצרתה In my distress לי קראתי I cried ויענני׃ and he heard}%
\verse{יהוה O LORD, הצילה Deliver נפשׁי my soul, משׂפת lips, שׁקר from lying מלשׁון tongue. רמיה׃ from a deceitful}%
\verse{מה What יתן shall be given לך ומה unto thee? or what יסיף shall be done לך לשׁון tongue? רמיה׃ unto thee, thou false}%
\verse{חצי arrows גבור of the mighty, שׁנונים Sharp עם with גחלי coals רתמים׃ of juniper.}%
\verse{אויה Woe לי כי is me, that גרתי I sojourn משׁך in Mesech, שׁכנתי I dwell עם in אהלי the tents קדר׃ of Kedar!}%
\verse{רבת hath long שׁכנה dwelt לה נפשׁי My soul עם with שׂונא him that hateth שׁלום׃ peace.}%
\verse{אני I שׁלום peace: וכי but when אדבר I speak, המה they למלחמה׃ for war.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 121
\verseWithHeading{Trust in God’s Protection}{שׁיר A Song למעלות of degrees. אשׂא I will lift up עיני mine eyes אל unto ההרים the hills, מאין from whence יבא cometh עזרי׃ my help.}%
\verse{עזרי My help מעם from יהוה the LORD, עשׂה which made שׁמים heaven וארץ׃ and earth.}%
\verse{אל He will not יתן suffer למוט to be moved: רגלך thy foot אל thee will not ינום slumber. שׁמרך׃ he that keepeth}%
\verse{הנה Behold, לא shall neither ינום slumber ולא nor יישׁן sleep. שׁומר he that keepeth ישׂראל׃ Israel}%
\verse{יהוה The LORD שׁמרך thy keeper: יהוה the LORD צלך thy shade על upon יד hand. ימינך׃ thy right}%
\verse{יומם thee by day, השׁמשׁ The sun לא shall not יככה smite וירח nor the moon בלילה׃ by night.}%
\verse{יהוה The LORD ישׁמרך shall preserve מכל thee from all רע evil: ישׁמר he shall preserve את נפשׁך׃ thy soul.}%
\verse{יהוה The LORD ישׁמר shall preserve צאתך thy going out ובואך and thy coming in מעתה from this time forth, ועד and even forevermore. עולם׃ and even forevermore.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 122
\verseWithHeading{Jerusalem the Site of God’s Presence}{שׁיר A Song המעלות of degrees לדוד of David. שׂמחתי I was glad באמרים when they said לי בית into the house יהוה of the LORD. נלך׃ unto me, Let us go}%
\verse{עמדות stand היו shall רגלינו Our feet בשׁעריך within thy gates, ירושׁלם׃ O Jerusalem.}%
\verse{ירושׁלם Jerusalem הבנויה is built כעיר as a city שׁחברה that is compact לה יחדו׃ together:}%
\verse{שׁשׁם עלו go up, שׁבטים the tribes שׁבטי the tribes יה of the LORD, עדות unto the testimony לישׂראל of Israel, להדות to give thanks לשׁם unto the name יהוה׃ of the LORD.}%
\verse{כי For שׁמה there ישׁבו are set כסאות thrones למשׁפט of judgment, כסאות the thrones לבית of the house דויד׃ of David.}%
\verse{שׁאלו Pray שׁלום for the peace ירושׁלם of Jerusalem: ישׁליו they shall prosper אהביך׃ that love}%
\verse{יהי be שׁלום Peace בחילך within thy walls, שׁלוה prosperity בארמנותיך׃ within thy palaces.}%
\verse{למען For אחי my brethren ורעי and companions' אדברה say, נא sakes, I will now שׁלום׃ Peace}%
\verse{למען Because of בית the house יהוה of the LORD אלהינו our God אבקשׁה I will seek טוב׃ thy good.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 123
\verseWithHeading{Prayer for Adonai’s Action in the Face of Scorn}{שׁיר A Song המעלות of degrees. אליך Unto נשׂאתי thee lift I up את עיני mine eyes, הישׁבי O thou that dwellest בשׁמים׃ in the heavens.}%
\verse{הנה Behold, כעיני as the eyes עבדים of servants אל unto יד the hand אדוניהם of their masters, כעיני as the eyes שׁפחה of a maiden אל unto יד the hand גברתה of her mistress; כן so עינינו our eyes אל upon יהוה the LORD אלהינו our God, עד until שׁיחננו׃ that he have mercy upon}%
\verse{חננו Have mercy upon יהוה us, O LORD, חננו have mercy upon כי us: for רב we are exceedingly שׂבענו filled בוז׃ with contempt.}%
\verse{רבת is exceedingly שׂבעה filled לה נפשׁנו Our soul הלעג with the scorning השׁאננים of those that are at ease, הבוז with the contempt לגאיונים׃ of the proud.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 124
\verseWithHeading{Thanksgiving for Adonai’s Help}{שׁיר A Song המעלות of degrees לדוד of David. לולי If יהוה the LORD שׁהיה who was לנו יאמר say; נא on our side, now ישׂראל׃ may Israel}%
\verse{לולי If יהוה the LORD שׁהיה who was לנו בקום rose up עלינו against אדם׃ on our side, when men}%
\verse{אזי Then חיים quick, בלעונו they had swallowed us up בחרות was kindled אפם׃ when their wrath}%
\verse{אזי Then המים the waters שׁטפונו had overwhelmed נחלה us, the stream עבר had gone על over נפשׁנו׃ our soul:}%
\verse{אזי Then עבר had gone על over נפשׁנו our soul. המים waters הזידונים׃ the proud}%
\verse{ברוך Blessed יהוה the LORD, שׁלא who hath not נתננו given טרף us a prey לשׁניהם׃ to their teeth.}%
\verse{נפשׁנו Our soul כצפור as a bird נמלטה is escaped מפח out of the snare יוקשׁים of the fowlers: הפח the snare נשׁבר is broken, ואנחנו and we נמלטנו׃ are escaped.}%
\verse{עזרנו Our help בשׁם in the name יהוה of the LORD, עשׂה who made שׁמים heaven וארץ׃ and earth.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 125
\verseWithHeading{Confidence in Adonai’s Protection}{שׁיר A Song המעלות of degrees. הבטחים They that trust ביהוה in the LORD כהר as mount ציון Zion, לא cannot ימוט be removed, לעולם forever. ישׁב׃ abideth}%
\verse{ירושׁלם Jerusalem, הרים As the mountains סביב round about לה ויהוה so the LORD סביב round about לעמו his people מעתה from henceforth ועד even forever. עולם׃ even forever.}%
\verse{כי For לא shall not ינוח rest שׁבט the rod הרשׁע of the wicked על upon גורל the lot הצדיקים of the righteous; למען lest לא lest ישׁלחו put forth הצדיקים the righteous בעולתה unto iniquity. ידיהם׃ their hands}%
\verse{היטיבה יהוה O LORD, לטובים unto good, ולישׁרים and upright בלבותם׃ in their hearts.}%
\verse{והמטים As for such as turn aside עקלקלותם unto their crooked ways, יוליכם shall lead them forth יהוה the LORD את with פעלי the workers האון of iniquity: שׁלום peace על upon ישׂראל׃ Israel.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 126
\verseWithHeading{A Prayer for Restoration}{שׁיר A Song המעלות of degrees. בשׁוב turned again יהוה When the LORD את שׁיבת the captivity ציון of Zion, היינו we were כחלמים׃ like them that dream.}%
\verse{אז Then ימלא filled שׂחוק with laughter, פינו was our mouth ולשׁוננו and our tongue רנה with singing: אז then יאמרו said בגוים they among the heathen, הגדיל great things יהוה The LORD לעשׂות hath done עם for אלה׃ them.}%
\verse{הגדיל great things יהוה The LORD לעשׂות hath done עמנו for היינו us; we are שׂמחים׃ glad.}%
\verse{שׁובה Turn again יהוה O LORD, את שׁבותנו our captivity, כאפיקים as the streams בנגב׃ in the south.}%
\verse{הזרעים They that sow בדמעה in tears ברנה in joy. יקצרו׃ shall reap}%
\verse{הלוך ילךבכה and weepeth, נשׂא bearing משׁך precious הזרע seed, בא shall doubtless come again יבוא shall doubtless come again ברנה with rejoicing, נשׂא bringing אלמתיו׃ his sheaves}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 127
\verseWithHeading{A Prayer for Protection and Prosperity}{שׁיר A Song המעלות of degrees לשׁלמה for Solomon. אם it: except יהוה the LORD לא it: except יבנה build בית the house, שׁוא in vain עמלו they labor בוניו that build בו אם יהוה the LORD לא ישׁמר keep עיר the city, שׁוא in vain. שׁקד waketh שׁומר׃ the watchman}%
\verse{שׁוא vain לכם משׁכימי early, קום for you to rise up מאחרי late, שׁבת to sit up אכלי to eat לחם the bread העצבים of sorrows: כן so יתן he giveth לידידו his beloved שׁנא׃ sleep.}%
\verse{הנה Lo, נחלת a heritage יהוה of the LORD: בנים children שׂכר reward. פרי the fruit הבטן׃ of the womb}%
\verse{כחצים As arrows ביד in the hand גבור of a mighty כן man; so בני children הנעורים׃ of the youth.}%
\verse{אשׁרי Happy הגבר the man אשׁר that מלא full את אשׁפתו hath his quiver מהם לא them: they shall not יבשׁו be ashamed, כי but ידברו they shall speak את with אויבים the enemies בשׁער׃ in the gate.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 128
\verseWithHeading{Blessed Is Everyone Who Fears Adonai}{שׁיר A Song המעלות of degrees. אשׁרי Blessed כל every one ירא that feareth יהוה ההלך that walketh בדרכיו׃ in his ways.}%
\verse{יגיע the labor כפיך of thine hands: כי For תאכל thou shalt eat אשׁריך happy וטוב׃ thou and well}%
\verse{אשׁתך Thy wife כגפן vine פריה as a fruitful בירכתי by the sides ביתך of thine house: בניך thy children כשׁתלי plants זיתים like olive סביב round about לשׁלחנך׃ thy table.}%
\verse{הנה Behold, כי that כן thus יברך be blessed גבר shall the man ירא that feareth יהוה׃ the LORD.}%
\verse{יברכך shall bless יהוה The LORD מציון וראה and thou shalt see בטוב the good ירושׁלם of Jerusalem כל all ימי the days חייך׃ of thy life.}%
\verse{וראה Yea, thou shalt see בנים thy children's לבניך children, שׁלום peace על upon ישׂראל׃ Israel.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 129
\verseWithHeading{Victory Over the Enemies of Zion}{שׁיר A Song המעלות of degrees. רבת Many a time צררוני have they afflicted מנעורי me from my youth, יאמר say: נא now ישׂראל׃ may Israel}%
\verse{רבת Many a time צררוני have they afflicted מנעורי me from my youth: גם yet לא they have not יכלו׃ prevailed}%
\verse{על upon גבי my back: חרשׁו The plowers חרשׁים plowed האריכו they made long למענותם׃ their furrows.}%
\verse{יהוה The LORD צדיק righteous: קצץ he hath cut asunder עבות the cords רשׁעים׃ of the wicked.}%
\verse{יבשׁו be confounded ויסגו and turned אחור back כל Let them all שׂנאי that hate ציון׃ Zion.}%
\verse{יהיו Let them be כחציר as the grass גגות the housetops, שׁקדמת before שׁלף it groweth up: יבשׁ׃ which withereth}%
\verse{שׁלא not מלא filleth כפו his hand; קוצר Wherewith the mower וחצנו sheaves his bosom. מעמר׃ nor he that bindeth}%
\verse{ולא Neither אמרו say, העברים do they which go by ברכת The blessing יהוה of the LORD אליכם upon ברכנו you: we bless אתכם בשׁם you in the name יהוה׃ of the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 130
\verseWithHeading{Hope for the Redemption of Adonai}{שׁיר A Song המעלות of degrees. ממעמקים קראתיך have I cried יהוה׃ unto thee, O LORD.}%
\verse{אדני Lord, שׁמעה hear בקולי my voice: תהיינה be אזניך let thine ears קשׁבות attentive לקול to the voice תחנוני׃ of my supplications.}%
\verse{אם If עונות iniquities, תשׁמר shouldest mark יה thou, LORD, אדני O Lord, מי who יעמד׃ shall stand?}%
\verse{כי But עמך with הסליחה forgiveness למען thee, that תורא׃ thou mayest be feared.}%
\verse{קויתי I wait for יהוה the LORD, קותה doth wait, נפשׁי my soul ולדברו and in his word הוחלתי׃ do I hope.}%
\verse{נפשׁי My soul לאדני for the Lord משׁמרים more than they that watch לבקר for the morning: שׁמרים they that watch לבקר׃ for the morning.}%
\verse{יחל hope ישׂראל Let Israel אל in יהוה the LORD: כי for עם with יהוה the LORD החסד mercy, והרבה him plenteous עמו and with פדות׃ redemption.}%
\verse{והוא And he יפדה shall redeem את ישׂראל Israel מכל from all עונתיו׃ his iniquities.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 131
\verseWithHeading{Calm Trust in Adonai}{שׁיר A Song המעלות of degrees לדוד of David. יהוה LORD, לא is not גבה haughty, לבי my heart ולא nor רמו lofty: עיני mine eyes ולא neither הלכתי do I exercise בגדלות myself in great matters, ובנפלאות or in things too high ממני׃ for}%
\verse{אם לאׁויתי I have behaved ודוממתי and quieted נפשׁי myself, כגמל as a child that is weaned עלי of אמו his mother: כגמל even as a weaned child. עלי נפשׁי׃ my soul}%
\verse{יחל hope ישׂראל Let Israel אל in יהוה the LORD מעתה from henceforth ועד and forever. עולם׃ and forever.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 132
\verseWithHeading{Adonai Dwells in Zion}{שׁיר A Song המעלות of degrees. זכור remember יהוה LORD, לדוד David, את כל all ענותו׃ his afflictions:}%
\verse{אשׁר How נשׁבע he swore ליהוה unto the LORD, נדר vowed לאביר unto the mighty יעקב׃ of Jacob;}%
\verse{אם Surely אבא I will not come באהל into the tabernacle ביתי of my house, אם nor אעלה go up על into ערשׂ my bed; יצועי׃ my bed;}%
\verse{אם I will not אתן give שׁנת sleep לעיני to mine eyes, לעפעפי to mine eyelids, תנומה׃ slumber}%
\verse{עד Until אמצא I find out מקום a place ליהוה for the LORD, משׁכנות a habitation לאביר for the mighty יעקב׃ of Jacob.}%
\verse{הנה Lo, שׁמענוה we heard באפרתה of it at Ephratah: מצאנוה we found בשׂדי it in the fields יער׃ of the wood.}%
\verse{נבואה We will go למשׁכנותיו into his tabernacles: נשׁתחוה we will worship להדם at his footstool. רגליו׃ at his footstool.}%
\verse{קומה Arise, יהוה O LORD, למנוחתך into thy rest; אתה thou, וארון and the ark עזך׃ of thy strength.}%
\verse{כהניך Let thy priests ילבשׁו be clothed צדק with righteousness; וחסידיך and let thy saints ירננו׃ shout for joy.}%
\verse{בעבור sake דוד David's עבדך For thy servant אל turn not away תשׁב turn not away פני the face משׁיחך׃ of thine anointed.}%
\verse{נשׁבע hath sworn יהוה The LORD לדוד unto David; אמת truth לא he will not ישׁוב turn ממנה from מפרי בטנך of thy body אשׁית will I set לכסא׃ upon thy throne.}%
\verse{אם If ישׁמרו will keep בניך thy children בריתי my covenant ועדתי and my testimony זו אלמדם I shall teach גם shall also בניהם them, their children עדי forevermore. עד ישׁבו sit לכסא׃ upon thy throne}%
\verse{כי For בחר hath chosen יהוה the LORD בציון Zion; אוה he hath desired למושׁב׃ for his habitation.}%
\verse{זאת This מנוחתי my rest עדי forever: עד פה here אשׁב will I dwell; כי for אותיה׃ I have desired}%
\verse{צידה her provision: ברך אברךביוניה her poor אשׂביע I will satisfy לחם׃ with bread.}%
\verse{וכהניה her priests אלבישׁ I will also clothe ישׁע with salvation: וחסידיה and her saints רנן shall shout aloud for joy. ירננו׃ shall shout aloud for joy.}%
\verse{שׁם There אצמיח to bud: קרן will I make the horn לדוד of David ערכתי I have ordained נר a lamp למשׁיחי׃ for mine anointed.}%
\verse{אויביו His enemies אלבישׁ will I clothe בשׁת with shame: ועליו but upon יציץ flourish. נזרו׃ himself shall his crown}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 133
\verseWithHeading{The People of God Dwell in Unity}{שׁיר A Song המעלות of degrees לדוד of David. הנה Behold, מה how טוב good ומה and how נעים pleasant שׁבת to dwell אחים for brethren גם יחד׃}%
\verse{כשׁמן ointment הטוב like the precious על upon הראשׁ the head, ירד that ran down על upon הזקן the beard, זקן beard: אהרן Aaron's שׁירד that went down על to פי the skirts מדותיו׃ of his garments;}%
\verse{כטל As the dew חרמון of Hermon, שׁירד that descended על upon הררי the mountains ציון of Zion: כי for שׁם there צוה commanded יהוה the LORD את הברכה the blessing, חיים life עד forevermore. העולם׃ forevermore.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 134
\verseWithHeading{Praising Adonai in the Temple at Night}{שׁיר A Song המעלות of degrees. הנה Behold, ברכו bless את יהוה ye the LORD, כל all עבדי servants יהוה of the LORD, העמדים stand בבית in the house יהוה of the LORD. בלילות׃ which by night}%
\verse{שׂאו Lift up ידכם your hands קדשׁ the sanctuary, וברכו and bless את יהוה׃ the LORD.}%
\verse{יברכך bless יהוה The LORD מציון עשׂה that made שׁמים heaven וארץ׃ and earth}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 135
\verseWithHeading{Praise to God for His Power and Redemption}{הללו Praise יה ye the LORD. הללו Praise את שׁם ye the name יהוה of the LORD; הללו praise עבדי O ye servants יהוה׃ of the LORD.}%
\verse{שׁעמדים בבית in the house יהוה of the Lord, בחצרות in the courts בית of the house אלהינו׃ of our God,}%
\verse{הללו Praise יה the LORD; כי for טוב good: יהוה the LORD זמרו sing praises לשׁמו unto his name; כי for נעים׃ pleasant.}%
\verse{כי For יעקב Jacob בחר hath chosen לו יה the LORD ישׂראל unto himself, Israel לסגלתו׃ for his peculiar treasure.}%
\verse{כי For אני I ידעתי know כי that גדול great, יהוה the LORD ואדנינו מכל above all אלהים׃ gods.}%
\verse{כל and all אשׁר חפץ pleased, יהוה the LORD עשׂה did בשׁמים he in heaven, ובארץ and in earth, בימים in the seas, וכל תהומות׃ deep places.}%
\verse{מעלה to ascend נשׂאים He causeth the vapors מקצה from the ends הארץ of the earth; ברקים lightnings למטר for the rain; עשׂה he maketh מוצא he bringeth רוח the wind מאוצרותיו׃ out of his treasuries.}%
\verse{שׁהכה בכורי the firstborn מצרים of Egypt, מאדם both of man עד and בהמה׃ beast.}%
\verse{שׁלח sent אתות tokens ומפתים and wonders בתוככי into the midst מצרים of thee, O Egypt, בפרעה upon Pharaoh, ובכל and upon all עבדיו׃ his servants.}%
\verse{שׁהכה גוים nations, רבים great והרג and slew מלכים kings; עצומים׃ mighty}%
\verse{לסיחון Sihon מלך king האמרי of the Amorites, ולעוג and Og מלך king הבשׁן of Bashan, ולכל and all ממלכות the kingdoms כנען׃ of Canaan:}%
\verse{ונתן And gave ארצם their land נחלה a heritage, נחלה a heritage לישׂראל unto Israel עמו׃ his people.}%
\verse{יהוה O LORD, שׁמך Thy name, לעולם forever; יהוה O LORD, זכרך thy memorial, לדר throughout all generations. ודר׃ throughout all generations.}%
\verse{כי For ידין will judge יהוה the LORD עמו his people, ועל concerning עבדיו his servants. יתנחם׃ and he will repent himself}%
\verse{עצבי The idols הגוים of the heathen כסף silver וזהב and gold, מעשׂה the work ידי hands. אדם׃ of men's}%
\verse{פה They have mouths, להם ולא not; ידברו but they speak עינים eyes להם ולא not; יראו׃ have they, but they see}%
\verse{אזנים They have ears, להם ולא not; יאזינו but they hear אף neither אין neither ישׁ is there רוח breath בפיהם׃ in their mouths.}%
\verse{כמוהם like unto them: יהיו them are עשׂיהם They that make כל every one אשׁר that בטח׃ trusteth}%
\verse{בית O house ישׂראל of Israel: ברכו Bless את יהוה the LORD, בית O house אהרן of Aaron: ברכו bless את יהוה׃ the LORD,}%
\verse{בית O house הלוי of Levi: ברכו Bless את יהוה the LORD, יראי ye that fear יהוה the LORD, ברכו bless את יהוה׃ the LORD.}%
\verse{ברוך Blessed יהוה be the LORD מציון שׁכן which dwelleth ירושׁלם at Jerusalem. הללו Praise יה׃ ye the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 136
\verseWithHeading{Praise to God for His Creation and Deliverance}{הודו O give thanks ליהוה unto the LORD; כי for טוב good: כי for לעולם forever. חסדו׃ his mercy}%
\verse{הודו O give thanks לאלהי unto the God האלהים of gods: כי for לעולם forever. חסדו׃ his mercy}%
\verse{הודו O give thanks לאדני to the Lord האדנים of lords: כי for לעולם forever. חסדו׃ his mercy}%
\verse{לעשׂה doeth נפלאות wonders: גדלות great לבדו To him who alone כי for לעולם forever. חסדו׃ his mercy}%
\verse{לעשׂה made השׁמים the heavens: בתבונה To him that by wisdom כי for לעולם forever. חסדו׃ his mercy}%
\verse{לרקע To him that stretched out הארץ the earth על above המים the waters: כי for לעולם forever. חסדו׃ his mercy}%
\verse{לעשׂה To him that made אורים lights: גדלים great כי for לעולם forever: חסדו׃ his mercy}%
\verse{את השׁמשׁ The sun לממשׁלת to rule ביום by day: כי for לעולם forever: חסדו׃ his mercy}%
\verse{את הירח The moon וכוכבים and stars לממשׁלות to rule בלילה by night: כי for לעולם forever. חסדו׃ his mercy}%
\verse{למכה To him that smote מצרים Egypt בבכוריהם in their firstborn: כי for לעולם forever: חסדו׃ his mercy}%
\verse{ויוצא And brought out ישׂראל Israel מתוכם from among כי them: for לעולם forever: חסדו׃ his mercy}%
\verse{ביד hand, חזקה With a strong ובזרוע arm: נטויה and with a stretched out כי for לעולם forever. חסדו׃ his mercy}%
\verse{לגזר To him which divided ים sea סוף the Red לגזרים into parts: כי for לעולם forever: חסדו׃ his mercy}%
\verse{והעביר to pass through ישׂראל And made Israel בתוכו the midst כי of it: for לעולם forever: חסדו׃ his mercy}%
\verse{ונער פרעה Pharaoh וחילו and his host בים sea: סוף in the Red כי for לעולם forever. חסדו׃ his mercy}%
\verse{למוליך To him which led עמו his people במדבר through the wilderness: כי for לעולם forever. חסדו׃ his mercy}%
\verse{למכה To him which smote מלכים kings: גדלים great כי for לעולם forever: חסדו׃ his mercy}%
\verse{ויהרג And slew מלכים kings: אדירים famous כי for לעולם forever: חסדו׃ his mercy}%
\verse{לסיחון Sihon מלך king האמרי of the Amorites: כי for לעולם forever: חסדו׃ his mercy}%
\verse{ולעוג And Og מלך the king הבשׁן of Bashan: כי for לעולם forever: חסדו׃ his mercy}%
\verse{ונתן And gave ארצם their land לנחלה for a heritage: כי for לעולם forever: חסדו׃ his mercy}%
\verse{נחלה a heritage לישׂראל unto Israel עבדו his servant: כי for לעולם forever. חסדו׃ his mercy}%
\verse{שׁבשׁפלנו us in our low estate: זכר Who remembered לנו כי for לעולם forever: חסדו׃ his mercy}%
\verse{ויפרקנו And hath redeemed מצרינו us from our enemies: כי for לעולם forever. חסדו׃ his mercy}%
\verse{נתן Who giveth לחם food לכל to all בשׂר flesh: כי for לעולם forever. חסדו׃ his mercy}%
\verse{הודו O give thanks לאל unto the God השׁמים of heaven: כי for לעולם forever. חסדו׃ his mercy}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 137
\verseWithHeading{Lament During the Babylonian Exile}{על By נהרות the rivers בבל of Babylon, שׁם there ישׁבנו we sat down, גם yea, בכינו we wept, בזכרנו when we remembered את ציון׃ Zion.}%
\verse{על upon ערבים the willows בתוכה in the midst תלינו We hanged כנרותינו׃ our harps}%
\verse{כי For שׁם there שׁאלונו required שׁובינו they that carried us away captive דברי of us a song; שׁיר of us a song; ותוללינו and they that wasted שׂמחה us mirth, שׁירו Sing לנו משׁיר us of the songs ציון׃ of Zion.}%
\verse{איך How נשׁיר shall we sing את שׁיר song יהוה the LORD's על in אדמת land? נכר׃ a strange}%
\verse{אם If אשׁכחך I forget ירושׁלם thee, O Jerusalem, תשׁכח forget ימיני׃ let my right hand}%
\verse{תדבק cleave לשׁוני thee, let my tongue לחכי to the roof of my mouth; אם If לא I do not אזכרכי remember אם if לא not אעלה I prefer את ירושׁלם Jerusalem על above ראשׁ my chief שׂמחתי׃ joy.}%
\verse{זכר Remember, יהוה O LORD, לבני the children אדום of Edom את יום in the day ירושׁלם of Jerusalem; האמרים who said, ערו Raze ערו raze עד to היסוד׃ the foundation}%
\verse{בת O daughter בבל of Babylon, השׁדודה who art to be destroyed; אשׁרי happy שׁישׁלם , that rewardeth לך את גמולך thee as שׁגמלת׃ thou hast served}%
\verse{אשׁרי Happy שׁיאחז , that taketh ונפץ and dasheth את עלליך thy little ones אל against הסלע׃ the stones.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 138
\verseWithHeading{Thanksgiving for Adonai’s Goodness}{לדוד of David. אודך I will praise בכל thee with my whole לבי heart: נגד before אלהים the gods אזמרך׃ will I sing praise}%
\verse{אשׁתחוה I will worship אל toward היכל temple, קדשׁך thy holy ואודה and praise את שׁמך thy name על and for חסדך thy lovingkindness ועל above אמתך thy truth: כי for הגדלת thou hast magnified על כל all שׁמך thy name. אמרתך׃ thy word}%
\verse{ביום In the day קראתי when I cried ותענני thou answeredst תרהבני me, strengthenedst בנפשׁי in my soul. עז׃ me strength}%
\verse{יודוך shall praise יהוה thee, O LORD, כל All מלכי the kings ארץ of the earth כי when שׁמעו they hear אמרי the words פיך׃ of thy mouth.}%
\verse{וישׁירו Yea, they shall sing בדרכי in the ways יהוה of the LORD: כי for גדול great כבוד the glory יהוה׃ of the LORD.}%
\verse{כי Though רם high, יהוה the LORD ושׁפל unto the lowly: יראה yet hath he respect וגבה but the proud ממרחק afar off. יידע׃ he knoweth}%
\verse{אם Though אלך I walk בקרב in the midst צרה of trouble, תחיני thou wilt revive על against אף the wrath איבי of mine enemies, תשׁלח me: thou shalt stretch forth ידך thine hand ותושׁיעני shall save ימינך׃ and thy right hand}%
\verse{יהוה The LORD יגמר will perfect בעדי concerneth יהוה O LORD, חסדך me: thy mercy, לעולם forever: מעשׂי the works ידיך of thine own hands. אל not תרף׃ forsake}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 139
\verseWithHeading{The Knowledge of God}{למנצח To the chief Musician, לדוד of David. מזמור A Psalm יהוה O LORD, חקרתני thou hast searched ותדע׃ me, and known}%
\verse{אתה Thou ידעת knowest שׁבתי my downsitting וקומי and mine uprising, בנתה thou understandest לרעי my thought מרחוק׃ afar off.}%
\verse{ארחי my path ורבעי and my lying down, זרית Thou compassest וכל all דרכי my ways. הסכנתה׃ and art acquainted}%
\verse{כי For אין not מלה a word בלשׁוני in my tongue, הן lo, יהוה O LORD, ידעת thou knowest כלה׃ it altogether.}%
\verse{אחור me behind וקדם and before, צרתני Thou hast beset ותשׁת and laid עלי upon כפכה׃ thine hand}%
\verse{פלאיה too wonderful דעת knowledge ממני for נשׂגבה me; it is high, לא אוכל׃}%
\verse{אנה Whither אלך shall I go מרוחך from thy spirit? ואנה or whither מפניך from thy presence? אברח׃ shall I flee}%
\verse{אם If אסק I ascend up שׁמים into heaven, שׁם there: אתה thou ואציעה if I make my bed שׁאול in hell, הנך׃}%
\verse{אשׂא I take כנפי the wings שׁחר of the morning, אשׁכנה dwell באחרית in the uttermost parts ים׃ of the sea;}%
\verse{גם Even שׁם there ידך shall thy hand תנחני lead ותאחזני shall hold ימינך׃ me, and thy right hand}%
\verse{ואמר If I say, אך Surely חשׁך the darkness ישׁופני shall cover ולילה me; even the night אור shall be light בעדני׃ about}%
\verse{גם Yea, חשׁך the darkness לא not יחשׁיך hideth ממך from ולילה thee; but the night כיום as the day: יאיר shineth כחשׁיכה the darkness כאורה׃ and the light}%
\verse{כי For אתה thou קנית hast possessed כליתי my reins: תסכני thou hast covered בבטן womb. אמי׃ me in my mother's}%
\verse{אודך I will praise על thee; for כי thee; for נוראות I am fearfully נפליתי wonderfully נפלאים made: marvelous מעשׂיך thy works; ונפשׁי and my soul ידעת knoweth מאד׃ right well.}%
\verse{לא was not נכחד hid עצמי My substance ממך from אשׁר thee, when עשׂיתי I was made בסתר in secret, רקמתי curiously wrought בתחתיות in the lowest parts ארץ׃ of the earth.}%
\verse{גלמי my substance, yet being unperfect; ראו did see עיניך Thine eyes ועל and in ספרך thy book כלם all יכתבו were written, ימים in continuance יצרו were fashioned, ולא when none אחד׃ when none}%
\verse{ולי מה How יקרו precious רעיך also are thy thoughts אל unto me, O God! מה how עצמו great ראשׁיהם׃ is the sum}%
\verse{אספרם I should count מחול than the sand: ירבון them, they are more in number הקיצתי when I awake, ועודי I am still עמך׃ with thee.}%
\verse{אם Surely תקטל thou wilt slay אלוה O God: רשׁע the wicked, ואנשׁי דמים me therefore, ye bloody סורו depart מני׃ from}%
\verse{אשׁר For יאמרך they speak למזמה against thee wickedly, נשׂא take לשׁוא in vain. עריך׃ thine enemies}%
\verse{הלוא Do not משׂנאיך I hate יהוה them, O LORD, אשׂנא that hate ובתקוממיך with those that rise up against אתקוטט׃ thee? and am not I grieved}%
\verse{תכלית them with perfect שׂנאה hatred: שׂנאתים I hate לאויבים them mine enemies. היו׃ I count}%
\verse{חקרני Search אל me, O God, ודע and know לבבי my heart: בחנני try ודע me, and know שׂרעפי׃ my thoughts:}%
\verse{וראה And see אם if דרך way עצב wicked בי ונחני in me, and lead בדרך me in the way עולם׃ everlasting.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 140
\verseWithHeading{Prayer for Help in the Face of Enemies}{למנצח To the chief Musician, מזמור A Psalm לדוד׃ of David. חלצני Deliver יהוה me, O LORD, מאדם man: רע from the evil מאישׁ man; חמסים me from the violent תנצרני׃ preserve}%
\verse{אשׁר Which חשׁבו imagine רעות mischiefs בלב in heart; כל continually יום continually יגורו are they gathered together מלחמות׃ war.}%
\verse{שׁננו They have sharpened לשׁונם their tongues כמו like נחשׁ a serpent; חמת poison עכשׁוב adders' תחת under שׂפתימו their lips. סלה׃ Selah.}%
\verse{שׁמרני Keep יהוה me, O LORD, מידי from the hands רשׁע of the wicked; מאישׁ חמסיםנצרני preserve אשׁר who חשׁבו have purposed לדחות to overthrow פעמי׃ my goings.}%
\verse{טמנו have hid גאים The proud פח a snare לי וחבלים for me, and cords; פרשׂו they have spread רשׁת a net ליד by the wayside; מעגל by the wayside; מקשׁים gins שׁתו they have set לי סלה׃ for me. Selah.}%
\verse{אמרתי I said ליהוה unto the LORD, אלי my God: אתה Thou האזינה hear יהוה O LORD. קול the voice תחנוני׃ of my supplications,}%
\verse{יהוה O GOD אדני the Lord, עז the strength ישׁועתי of my salvation, סכתה thou hast covered לראשׁי my head ביום in the day נשׁק׃ of battle.}%
\verse{אל not, תתן Grant יהוה O LORD, מאויי the desires רשׁע of the wicked: זממו his wicked device; אל not תפק further ירומו they exalt סלה׃ themselves. Selah.}%
\verse{ראשׁ the head מסבי of those that compass me about, עמל let the mischief שׂפתימו of their own lips יכסומו׃ cover}%
\verse{ימיטו fall עליהם upon גחלים Let burning coals באשׁ into the fire; יפלם them: let them be cast במהמרות into deep pits, בל not יקומו׃ that they rise}%
\verse{אישׁ an evil speaker לשׁון an evil speaker בל Let not יכון be established בארץ in the earth: אישׁ man חמס the violent רע evil יצודנו shall hunt למדחפת׃ to overthrow}%
\verse{ידעת I know כי that יעשׂה will maintain יהוה the LORD דין the cause עני of the afflicted, משׁפט the right אבינים׃ of the poor.}%
\verse{אך Surely צדיקים the righteous יודו shall give thanks לשׁמך unto thy name: ישׁבו shall dwell ישׁרים the upright את פניך׃ in thy presence.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 141
\verseWithHeading{Prayer for God’s Help in Maintaining Integrity}{מזמור A Psalm לדוד of David. יהוה LORD, קראתיך I cry חושׁה unto thee: make haste לי האזינה unto me; give ear קולי unto my voice, בקראי׃ when I cry}%
\verse{תכון be set forth תפלתי Let my prayer קטרת thee incense; לפניך before משׂאת the lifting up כפי of my hands מנחת sacrifice. ערב׃ the evening}%
\verse{שׁיתה Set יהוה O LORD, שׁמרה a watch, לפי before my mouth; נצרה keep על keep דל the door שׂפתי׃ of my lips.}%
\verse{אל not תט Incline לבי my heart לדבר thing, רע to evil להתעולל to practice עללות works ברשׁע wicked את with אישׁים פעלי that work און iniquity: ובל and let me not אלחם eat במנעמיהם׃ of their dainties.}%
\verse{יהלמני smite צדיק Let the righteous חסד me; a kindness: ויוכיחני and let him reprove שׁמן oil, ראשׁ me; an excellent אל shall not יני break ראשׁי my head: כי for עוד yet ותפלתי my prayer ברעותיהם׃ also in their calamities.}%
\verse{נשׁמטו are overthrown בידי places, סלע in stony שׁפטיהם When their judges ושׁמעו they shall hear אמרי my words; כי for נעמו׃ they are sweet.}%
\verse{כמו as when פלח one cutteth ובקע and cleaveth בארץ upon the earth. נפזרו are scattered עצמינו Our bones לפי mouth, שׁאול׃ at the grave's}%
\verse{כי But אליך unto יהוה אדני the Lord: עיני mine eyes בכה חסיתי in thee is my trust; אל תער leave not my soul destitute. נפשׁי׃}%
\verse{שׁמרני Keep מידי me from פח the snares יקשׁו they have laid לי ומקשׁות for me, and the gins פעלי of the workers און׃ of iniquity.}%
\verse{יפלו fall במכמריו into their own nets, רשׁעים Let the wicked יחד withal אנכי that I עד whilst אעבור׃ escape.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 142
\verseWithHeading{A Prayer for Deliverance from Pursuers}{משׂכיל Maschil לדוד of David; בהיותו when he was במערה in the cave. תפלה׃ A Prayer קולי with my voice; אל unto יהוה the LORD אזעק I cried קולי with my voice אל unto יהוה the LORD אתחנן׃ did I make my supplication.}%
\verse{אשׁפך I poured out לפניו before שׂיחי my complaint צרתי him my trouble. לפניו before אגיד׃ him; I showed}%
\verse{בהתעטף was overwhelmed עלי within רוחי When my spirit ואתה me, then thou ידעת knewest נתיבתי my path. בארח In the way זו wherein אהלך I walked טמנו have they privily laid פח׃ a snare}%
\verse{הביט I looked ימין on right hand, וראה and beheld, ואין but no man לי מכיר that would know אבד failed מנוס me: refuge ממני failed אין me; no man דורשׁ cared לנפשׁי׃ for my soul.}%
\verse{זעקתי I cried אליך unto יהוה thee, O LORD: אמרתי I said, אתה Thou מחסי my refuge חלקי my portion בארץ in the land החיים׃ of the living.}%
\verse{הקשׁיבה Attend אל unto רנתי my cry; כי for דלותי מאדצילני deliver מרדפי me from my persecutors; כי for אמצו they are stronger ממני׃ me from my persecutors;}%
\verse{הוציאה ממסגר of prison, נפשׁי להודות that I may praise את שׁמך thy name: בי יכתרו shall compass me about; צדיקים the righteous כי for תגמל thou shalt deal bountifully עלי׃ with}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 143
\verseWithHeading{A Prayer for Rescue from Enemies}{מזמור A Psalm לדוד of David. יהוה O LORD, שׁמע Hear תפלתי my prayer, האזינה give ear אל to תחנוני my supplications: באמנתך in thy faithfulness ענני answer בצדקתך׃ me, in thy righteousness.}%
\verse{ואל not תבוא And enter במשׁפט into judgment את with עבדך thy servant: כי for לא shall no יצדק be justified. לפניך in thy sight כל man חי׃ living}%
\verse{כי For רדף hath persecuted אויב the enemy נפשׁי my soul; דכא he hath smitten לארץ down to the ground; חיתי my life הושׁיבני he hath made me to dwell במחשׁכים in darkness, כמתי dead. עולם׃ as those that have been long}%
\verse{ותתעטף overwhelmed עלי within רוחי Therefore is my spirit בתוכי within ישׁתומם me is desolate. לבי׃ me; my heart}%
\verse{זכרתי I remember ימים the days מקדם of old; הגיתי I meditate בכל on all פעלך thy works; במעשׂה on the work ידיך of thy hands. אשׂוחח׃ I muse}%
\verse{פרשׂתי I stretch forth ידי my hands אליך unto thee: נפשׁי my soul כארץ land. עיפה after thee, as a thirsty לך סלה׃ Selah.}%
\verse{מהר ענני Hear יהוה O LORD: כלתה faileth: רוחי my spirit אל not תסתר hide פניך thy face ממני from ונמשׁלתי me, lest I be like עם unto ירדי them that go down בור׃ into the pit.}%
\verse{השׁמיעני Cause me to hear בבקר in the morning; חסדך thy lovingkindness כי for בך בטחתי in thee do I trust: הודיעני cause me to know דרך the way זו wherein אלך I should walk; כי for אליך unto נשׂאתי I lift up נפשׁי׃ my soul}%
\verse{הצילני Deliver מאיבי from mine enemies: יהוה me, O LORD, אליך כסתי׃}%
\verse{למדני Teach לעשׂות me to do רצונך thy will; כי for אתה thou אלוהי רוחך thy spirit טובה good; תנחני lead בארץ me into the land מישׁור׃ of uprightness.}%
\verse{למען שׁמךהוה me, O LORD, תחיני Quicken בצדקתך for thy righteousness' תוציא sake bring my soul out מצרה of trouble. נפשׁי׃ sake bring my soul out}%
\verse{ובחסדך And of thy mercy תצמית cut off איבי mine enemies, והאבדת and destroy כל all צררי them that afflict נפשׁי my soul: כי for אני I עבדך׃ thy servant.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 144
\verseWithHeading{A Prayer for National Safety}{לדוד of David. ברוך Blessed יהוה the LORD צורי my strength, המלמד which teacheth ידי my hands לקרב to war, אצבעותי my fingers למלחמה׃ to fight:}%
\verse{חסדי My goodness, ומצודתי and my fortress; משׂגבי my high tower, ומפלטי and my deliverer; לי מגני my shield, ובו חסיתי and in whom I trust; הרודד who subdueth עמי my people תחתי׃ under}%
\verse{יהוה מה what אדם man, ותדעהו that thou takest knowledge בן of him! the son אנושׁ ותחשׁבהו׃ that thou makest account}%
\verse{אדם Man להבל to vanity: דמה is like ימיו his days כצל as a shadow עובר׃ that passeth away.}%
\verse{יהוה O LORD, הט Bow שׁמיך thy heavens, ותרד and come down: גע touch בהרים the mountains, ויעשׁנו׃ and they shall smoke.}%
\verse{ברוק Cast forth ברק lightning, ותפיצם and scatter שׁלח them: shoot out חציך thine arrows, ותהמם׃ and destroy}%
\verse{שׁלח Send ידיך thine hand ממרום from above; פצני rid והצילני me, and deliver ממים waters, רבים me out of great מיד from the hand בני children; נכר׃ of strange}%
\verse{אשׁר Whose פיהם mouth דבר speaketh שׁוא vanity, וימינם and their right hand ימין a right hand שׁקר׃ of falsehood.}%
\verse{אלהים unto thee, O God: שׁיר I will sing חדשׁ a new אשׁירה song לך בנבל upon a psaltery עשׂור an instrument of ten strings אזמרה׃ will I sing praises}%
\verse{הנותן that giveth תשׁועה salvation למלכים unto kings: הפוצה who delivereth את דוד David עבדו his servant מחרב sword. רעה׃ from the hurtful}%
\verse{פצני Rid והצילני me, and deliver מיד me from the hand בני children, נכר of strange אשׁר whose פיהם mouth דבר speaketh שׁוא vanity, וימינם and their right hand ימין a right hand שׁקר׃ of falsehood:}%
\verse{אשׁר That בנינו our sons כנטעים as plants מגדלים grown up בנעוריהם in their youth; בנותינו our daughters כזוית as corner stones, מחטבות polished תבנית the similitude היכל׃ of a palace:}%
\verse{מזוינו our garners מלאים full, מפיקים affording מזן אלן צאוננו our sheep מאליפות may bring forth thousands מרבבות and ten thousands בחוצותינו׃ in our streets:}%
\verse{אלופינו our oxen מסבלים strong to labor; אין no פרץ breaking in, ואין nor יוצאת going out ואין ; that no צוחה complaining ברחבתינו׃ in our streets.}%
\verse{אשׁרי Happy העם people, שׁככה that is in such a case: לו אשׁרי happy העם people, שׁיהוה אלהיו׃ whose God}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 145
\verseWithHeading{A Song of God’s Majesty and Love}{תהלה of praise. לדוד David's ארוממך I will extol אלוהי המלך O king; ואברכה and I will bless שׁמך thy name לעולם forever ועד׃ and ever.}%
\verse{בכל Every יום day אברכך will I bless ואהללה thee; and I will praise שׁמך thy name לעולם forever ועד׃ and ever.}%
\verse{גדול Great יהוה the LORD, ומהלל to be praised; מאד and greatly ולגדלתו and his greatness אין unsearchable. חקר׃ unsearchable.}%
\verse{דור One generation לדור to another, ישׁבח shall praise מעשׂיך thy works וגבורתיך thy mighty acts. יגידו׃ and shall declare}%
\verse{הדר honor כבוד of the glorious הודך of thy majesty, ודברי works. נפלאותיך and of thy wondrous אשׂיחה׃ I will speak}%
\verse{ועזוז of the might נוראתיך of thy terrible acts: יאמרו And shall speak וגדולתיך thy greatness. אספרנה׃ and I will declare}%
\verse{זכר the memory רב of thy great טובך goodness, יביעו They shall abundantly utter וצדקתך of thy righteousness. ירננו׃ and shall sing}%
\verse{חנון gracious, ורחום and full of compassion; יהוה The LORD ארך slow אפים to anger, וגדל and of great חסד׃ mercy.}%
\verse{טוב good יהוה The LORD לכל to all: ורחמיו and his tender mercies על over כל all מעשׂיו׃ his works.}%
\verse{יודוך shall praise יהוה thee, O LORD; כל All מעשׂיך thy works וחסידיך and thy saints יברכוכה׃ shall bless}%
\verse{כבוד of the glory מלכותך of thy kingdom, יאמרו They shall speak וגבורתך of thy power; ידברו׃ and talk}%
\verse{להודיע To make known לבני to the sons האדם of men גבורתיו his mighty acts, וכבוד and the glorious הדר majesty מלכותו׃ of his kingdom.}%
\verse{מלכותך Thy kingdom מלכות kingdom, כל an everlasting עלמים an everlasting וממשׁלתך and thy dominion בכל throughout all דור generations. ודור׃ generations.}%
\verse{סומך upholdeth יהוה The LORD לכל all הנפלים that fall, וזוקף and raiseth up לכל all הכפופים׃ bowed down.}%
\verse{עיני The eyes כל of all אליך upon ישׂברו wait ואתה thee; and thou נותן givest להם אתכלם them their meat בעתו׃ in due season.}%
\verse{פותח Thou openest את ידך thine hand, ומשׂביע and satisfiest לכל of every חי living thing. רצון׃ the desire}%
\verse{צדיק righteous יהוה The LORD בכל in all דרכיו his ways, וחסיד and holy בכל in all מעשׂיו׃ his works.}%
\verse{קרוב nigh יהוה The LORD לכל unto all קראיו them that call לכל upon him, to all אשׁר that יקראהו call באמת׃ upon him in truth.}%
\verse{רצון the desire יראיו of them that fear יעשׂה He will fulfill ואת שׁועתם their cry, ישׁמע him: he also will hear ויושׁיעם׃ and will save}%
\verse{שׁומר preserveth יהוה The LORD את כל all אהביו them that love ואת כל him: but all הרשׁעים the wicked ישׁמיד׃ will he destroy.}%
\verse{תהלת the praise יהוה of the LORD: ידבר shall speak פי My mouth ויברך bless כל and let all בשׂר flesh שׁם name קדשׁו his holy לעולם forever ועד׃ and ever.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 146
\verseWithHeading{Praise to Adonai for His Help}{הללו Praise יה ye the LORD. הללי Praise נפשׁי O my soul. את יהוה׃ the LORD,}%
\verse{אהללה will I praise יהוה the LORD: בחיי While I live אזמרה I will sing praises לאלהי unto my God בעודי׃ while I have any being.}%
\verse{אל Put not תבטחו your trust בנדיבים in princes, בבן in the son אדם of man, שׁאין in whom no לו תשׁועה׃ help.}%
\verse{תצא goeth forth, רוחו His breath ישׁב he returneth לאדמתו to his earth; ביום day ההוא in that very אבדו perish. עשׁתנתיו׃ his thoughts}%
\verse{אשׁרי Happy שׁאל יעקב of Jacob בעזרו for his help, שׂברו whose hope על in יהוה the LORD אלהיו׃ his God:}%
\verse{עשׂה Which made שׁמים heaven, וארץ and earth, את הים the sea, ואת כל and all אשׁר that בם השׁמר therein which keepeth אמת truth לעולם׃ forever:}%
\verse{עשׂה Which executeth משׁפט judgment לעשׁוקים for the oppressed: נתן which giveth לחם food לרעבים to the hungry. יהוה The LORD מתיר looseth אסורים׃ the prisoners:}%
\verse{יהוה The LORD פקח openeth עורים the blind: יהוה the LORD זקף raiseth כפופים them that are bowed down: יהוה the LORD אהב loveth צדיקים׃ the righteous:}%
\verse{יהוה The LORD שׁמר preserveth את גרים the strangers; יתום the fatherless ואלמנה and widow: יעודד he relieveth ודרך but the way רשׁעים of the wicked יעות׃ he turneth upside down.}%
\verse{ימלך shall reign יהוה The LORD לעולם forever, אלהיך thy God, ציון O Zion, לדר unto all generations. ודר unto all generations. הללו Praise יה׃ ye the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 147
\verseWithHeading{Praise to Adonai for His Providence}{הללו Praise יה ye the LORD: כי for טוב good זמרה to sing praises אלהינו unto our God; כי for נעים pleasant; נאוה is comely. תהלה׃ praise}%
\verse{בונה doth build up ירושׁלם Jerusalem: יהוה The LORD נדחי the outcasts ישׂראל of Israel. יכנס׃ he gathereth together}%
\verse{הרפא He healeth לשׁבורי the broken לב in heart, ומחבשׁ and bindeth up לעצבותם׃ their wounds.}%
\verse{מונה He telleth מספר the number לכוכבים of the stars; לכלם them all שׁמות by names. יקרא׃ he calleth}%
\verse{גדול Great אדונינו our Lord, ורב and of great כח power: לתבונתו his understanding אין infinite. מספר׃ infinite.}%
\verse{מעודד lifteth up ענוים the meek: יהוה The LORD משׁפיל he casteth the wicked down רשׁעים he casteth the wicked down עדי ארץ׃ the ground.}%
\verse{ענו Sing ליהוה unto the LORD בתודה with thanksgiving; זמרו sing praise לאלהינו unto our God: בכנור׃ upon the harp}%
\verse{המכסה Who covereth שׁמים the heaven בעבים with clouds, המכין who prepareth לארץ for the earth, מטר rain המצמיח to grow הרים upon the mountains. חציר׃ who maketh grass}%
\verse{נותן He giveth לבהמה to the beast לחמה his food, לבני to the young ערב ravens אשׁר which יקראו׃ cry.}%
\verse{לא not בגבורת in the strength הסוס of the horse: יחפץ He delighteth לא he taketh not pleasure בשׁוקי in the legs האישׁ of a man. ירצה׃ he taketh not pleasure}%
\verse{רוצה taketh pleasure in יהוה The LORD את יראיו them that fear את המיחלים him, in those that hope לחסדו׃ in his mercy.}%
\verse{שׁבחי Praise ירושׁלם O Jerusalem; את יהוה the LORD, הללי praise אלהיך thy God, ציון׃ O Zion.}%
\verse{כי For חזק he hath strengthened בריחי the bars שׁעריך of thy gates; ברך he hath blessed בניך thy children בקרבך׃ within}%
\verse{השׂם He maketh גבולך thy borders, שׁלום peace חלב thee with the finest חטים of the wheat. ישׂביעך׃ filleth}%
\verse{השׁלח He sendeth forth אמרתו his commandment ארץ earth: עד very מהרה swiftly. ירוץ runneth דברו׃ his word}%
\verse{הנתן He giveth שׁלג snow כצמר like wool: כפור the hoarfrost כאפר like ashes. יפזר׃ he scattereth}%
\verse{משׁליך He casteth forth קרחו his ice כפתים like morsels: לפני before קרתו his cold? מי who יעמד׃ can stand}%
\verse{ישׁלח He sendeth out דברו his word, וימסם and melteth ישׁב to blow, רוחו them: he causeth his wind יזלו flow. מים׃ the waters}%
\verse{מגיד He showeth דברו his word ליעקב unto Jacob, חקיו his statutes ומשׁפטיו and his judgments לישׂראל׃ unto Israel.}%
\verse{לא He hath not עשׂה dealt כן so לכל with any גוי nation: ומשׁפטים and judgments, בל they have not ידעום known הללו them. Praise יה׃ ye the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 148
\verseWithHeading{Let All Creation Praise Adonai}{הללו Praise יה ye the LORD. הללו Praise את יהוה ye the LORD מן from השׁמים the heavens: הללוהו praise במרומים׃ him in the heights.}%
\verse{הללוהו Praise כל ye him, all מלאכיו his angels: הללוהו praise כל ye him, all צבאו׃ his hosts.}%
\verse{הללוהו Praise שׁמשׁ ye him, sun וירח and moon: הללוהו praise כל him, all כוכבי ye stars אור׃ of light.}%
\verse{הללוהו Praise שׁמי him, ye heavens השׁמים of heavens, והמים and ye waters אשׁר that מעל above השׁמים׃ the heavens.}%
\verse{יהללו Let them praise את שׁם the name יהוה of the LORD: כי for הוא he צוה commanded, ונבראו׃ and they were created.}%
\verse{ויעמידם He hath also established לעד them forever לעולם and ever: חק a decree נתן he hath made ולא which shall not יעבור׃ pass.}%
\verse{הללו Praise את יהוה the LORD מן from הארץ the earth, תנינים ye dragons, וכל and all תהמות׃ deeps:}%
\verse{אשׁ Fire, וברד and hail; שׁלג snow, וקיטור and vapor; רוח wind סערה stormy עשׂה fulfilling דברו׃ his word:}%
\verse{ההרים Mountains, וכל and all גבעות hills; עץ trees, פרי fruitful וכל and all ארזים׃ cedars:}%
\verse{החיה Beasts, וכל and all בהמה cattle; רמשׂ creeping things, וצפור fowl: כנף׃ and flying}%
\verse{מלכי Kings ארץ of the earth, וכל and all לאמים people; שׂרים princes, וכל and all שׁפטי judges ארץ׃ of the earth:}%
\verse{בחורים Both young men, וגם and בתולות maidens; זקנים old men, עם and נערים׃ children:}%
\verse{יהללו Let them praise את שׁם the name יהוה of the LORD: כי for נשׂגב is excellent; שׁמו his name לבדו alone הודו his glory על above ארץ the earth ושׁמים׃ and heaven.}%
\verse{וירם He also exalteth קרן the horn לעמו of his people, תהלה the praise לכל of all חסידיו his saints; לבני of the children ישׂראל of Israel, עם a people קרבו near הללו unto him. Praise יה׃ ye the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 149
\verseWithHeading{Praise to God for His Future Judgment}{הללו Praise יה ye the LORD. שׁירו Sing ליהוה unto the LORD שׁיר song, חדשׁ a new תהלתו his praise בקהל in the congregation חסידים׃ of saints.}%
\verse{ישׂמח rejoice ישׂראל Let Israel בעשׂיו in him that made בני him: let the children ציון of Zion יגילו be joyful במלכם׃ in their King.}%
\verse{יהללו Let them praise שׁמו his name במחול in the dance: בתף unto him with the timbrel וכנור and harp. יזמרו׃ let them sing praises}%
\verse{כי For רוצה taketh pleasure יהוה the LORD בעמו in his people: יפאר he will beautify ענוים the meek בישׁועה׃ with salvation.}%
\verse{יעלזו be joyful חסידים Let the saints בכבוד in glory: ירננו let them sing aloud על upon משׁכבותם׃ their beds.}%
\verse{רוממות אל of God בגרונם in their mouth, וחרב sword פיפיות and a twoedged בידם׃ in their hand;}%
\verse{לעשׂות To execute נקמה vengeance בגוים upon the heathen, תוכחת punishments בל אמים׃ upon the people;}%
\verse{לאסר To bind מלכיהם their kings בזקים with chains, ונכבדיהם and their nobles בכבלי with fetters ברזל׃ of iron;}%
\verse{לעשׂות To execute בהם משׁפט upon them the judgment כתוב written: הדר honor הוא this לכל have all חסידיו his saints. הללו Praise יה׃ ye the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Psalm 150
\verseWithHeading{Let Everything Praise Adonai}{הללו Praise יה ye the LORD. הללו Praise אל God בקדשׁו in his sanctuary: הללוהו praise ברקיע him in the firmament עזו׃ of his power.}%
\verse{הללוהו Praise בגבורתיו him for his mighty acts: הללוהו praise כרב him according to his excellent גדלו׃ greatness.}%
\verse{הללוהו Praise בתקע him with the sound שׁופר of the trumpet: הללוהו praise בנבל him with the psaltery וכנור׃ and harp.}%
\verse{הללוהו Praise בתף him with the timbrel ומחול and dance: הללוהו praise במנים him with stringed instruments ועוגב׃ and organs.}%
\verse{הללוהו Praise בצלצלי cymbals: שׁמע him upon the loud הללוהו praise בצלצלי cymbals. תרועה׃ him upon the high sounding}%
\verse{כל Let every הנשׁמה thing that hath breath תהלל praise יה the LORD. הללו Praise יה׃ ye the LORD.}%
\end{biblechapter}%
