\biblebook{Zechariah}
\begin{biblechapter}% Zechariah 1
\verse{בחדשׁ month, השׁמיני In the eighth בשׁנת year שׁתים in the second לדריושׁ of Darius, היה came דבר the word יהוה of the LORD אל unto זכריה Zechariah, בן the son ברכיה of Berechiah, בן the son עדו of Iddo הנביא the prophet, לאמר׃ saying,}%
\verse{קצף hath been sore displeased יהוה The LORD על with אבותיכם your fathers. קצף׃ hath been sore displeased}%
\verse{ואמרת Therefore say אלהם thou unto כה them, Thus אמר saith יהוה the LORD צבאות of hosts; שׁובו Turn אלי ye unto נאם me, saith יהוה the LORD צבאות of hosts, ואשׁוב and I will turn אליכם unto אמר you, saith יהוה the LORD צבאות׃ of hosts.}%
\verse{אל ye not תהיו Be כאבתיכם as your fathers, אשׁר whom קראו have cried, אליהם unto הנביאים prophets הראשׁנים the former לאמר saying, כה Thus אמר saith יהוה the LORD צבאות of hosts; שׁובו Turn נא ye now מדרכיכם ways, הרעים from your evil ומעליליכם doings: הרעים and your evil ולא but they did not שׁמעו hear, ולא nor הקשׁיבו hearken אלי unto נאם me, saith יהוה׃ the LORD.}%
\verse{אבותיכם Your fathers, איה where הם they? והנבאים and the prophets, הלעולם forever? יחיו׃ do they live}%
\verse{אך But דברי my words וחקי and my statutes, אשׁר which צויתי I commanded את עבדי my servants הנביאים the prophets, הלוא did they not השׂיגו take hold אבתיכם of your fathers? וישׁובו and they returned ויאמרו and said, כאשׁר Like as זמם thought יהוה the LORD צבאות of hosts לעשׂות to do לנו כדרכינו unto us, according to our ways, וכמעללינו and according to our doings, כן so עשׂה hath he dealt אתנו׃ with}%
\verseWithHeading{Zechariah’s First Vision}{ביום day עשׂרים and twentieth וארבעה Upon the four לעשׁתי of the eleventh עשׂר of the eleventh חדשׁ month, הוא which חדשׁ the month שׁבט Sebat, בשׁנת year שׁתים in the second לדריושׁ of Darius, היה came דבר the word יהוה of the LORD אל unto זכריה Zechariah, בן the son ברכיהו of Berechiah, בן the son עדוא of Iddo הנביא the prophet, לאמר׃ saying,}%
\verse{ראיתי I saw הלילה by night, והנה and behold אישׁ a man רכב riding על upon סוס horse, אדם a red והוא and he עמד stood בין among ההדסים the myrtle trees אשׁר that במצלה in the bottom; ואחריו and behind סוסים horses, אדמים him red שׂרקים speckled, ולבנים׃ and white.}%
\verse{ואמר Then said מה what אלה these? אדני I, O my lord, ויאמר with me said אלי unto המלאך And the angel הדבר that talked בי אני me, I אראך will show מה thee what המה אלה׃ these}%
\verse{ויען answered האישׁ And the man העמד that stood בין among ההדסים the myrtle trees ויאמר and said, אלה These אשׁר whom שׁלח hath sent יהוה the LORD להתהלך to walk to and fro בארץ׃ through the earth.}%
\verse{ויענו And they answered את מלאך the angel יהוה העמד that stood בין among ההדסים the myrtle trees, ויאמרו and said, התהלכנו We have walked to and fro בארץ through the earth, והנה and, behold, כל all הארץ the earth ישׁבת sitteth still, ושׁקטת׃ and is at rest.}%
\verse{ויען answered מלאך Then the angel יהוה of the LORD ויאמר and said, יהוה O LORD צבאות of hosts, עד how long מתי how long אתה wilt thou לא not תרחם have mercy on את ירושׁלם Jerusalem ואת ערי and on the cities יהודה of Judah, אשׁר against which זעמתה thou hast had indignation זה these שׁבעים threescore and ten שׁנה׃ years?}%
\verse{ויען answered יהוה And the LORD את המלאך the angel הדבר that talked בי דברים words טובים with me good דברים words. נחמים׃ comfortable}%
\verse{ויאמר with me said אלי unto המלאך So the angel הדבר that communed בי קרא me, Cry לאמר thou, saying, כה Thus אמר saith יהוה the LORD צבאות of hosts; קנאתי I am jealous לירושׁלם for Jerusalem ולציון and for Zion קנאה jealousy. גדולה׃ with a great}%
\verse{וקצף sore displeased גדול am very אני And I קצף sore displeased על with הגוים the heathen השׁאננים at ease: אשׁר for אני I קצפתי displeased, מעט was but a little והמה and they עזרו helped לרעה׃ forward the affliction.}%
\verse{לכן Therefore כה thus אמר saith יהוה the LORD; שׁבתי I am returned לירושׁלם to Jerusalem ברחמים with mercies: ביתי my house יבנה shall be built בה נאם in it, saith יהוה the LORD צבאות of hosts, וקוה ינטה shall be stretched forth על upon ירושׁלם׃ Jerusalem.}%
\verse{עוד yet, קרא Cry לאמר saying, כה Thus אמר saith יהוה the LORD צבאות of hosts; עוד shall yet תפוצינה be spread abroad; ערי My cities מטוב through prosperity ונחם comfort יהוה and the LORD עוד shall yet את ציון Zion, ובחר choose עוד and shall yet בירושׁלם׃ Jerusalem.}%
\verseWithHeading{Zechariah’s Second Vision}{ואשׂא Then lifted I up את עיני mine eyes, וארא and saw, והנה and behold ארבע four קרנות׃ horns.}%
\verse{ואמר And I said אל unto המלאך the angel הדבר that talked בי מה with me, What אלה these? ויאמר אלילה me, These הקרנות the horns אשׁר which זרו have scattered את יהודה Judah, את ישׂראל Israel, וירושׁלם׃ and Jerusalem.}%
\verse{ויראני showed יהוה And the LORD ארבעה me four חרשׁים׃ carpenters.}%
\verse{ואמר Then said מה I, What אלה these באים come לעשׂות to do? ויאמר And he spoke, לאמר saying, אלה These הקרנות the horns אשׁר which זרו have scattered את יהודה Judah, כפי אישׁ man לא so that no נשׂא did lift up ראשׁו his head: ויבאו are come אלה but these להחריד to frighten אתם לידות them, to cast out את קרנות the horns הגוים of the Gentiles, הנשׂאים which lifted up קרן horn אל over ארץ the land יהודה of Judah לזרותה׃ to scatter}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 2
\verseWithHeading{Zechariah’s Third Vision}{ואשׂא I lifted up עיני mine eyes וארא again, and looked, והנה and behold אישׁ a man ובידו in his hand. חבל line מדה׃ with a measuring}%
\verse{ואמר Then said אנה I, Whither אתה thou? הלך goest ויאמר And he said אלי unto למד me, To measure את ירושׁלם Jerusalem, לראות to see כמה what רחבה the breadth וכמה thereof, and what ארכה׃ the length}%
\verse{והנה And, behold, המלאך the angel הדבר that talked בי יצא with me went forth, ומלאך angel אחר and another יצא went out לקראתו׃ to meet}%
\verse{ויאמר And said אלו unto רץ him, Run, דבר speak אל to הנער young man, הלז this לאמר saying, פרזות towns without walls תשׁב shall be inhabited ירושׁלם Jerusalem מרב for the multitude אדם of men ובהמה and cattle בתוכה׃ therein:}%
\verse{ואני For I, אהיה will be לה נאם saith יהוה חומת unto her a wall אשׁ of fire סביב round about, ולכבוד the glory אהיה and will be בתוכה׃ in the midst}%
\verse{הוי Ho, הוי ho, ונסו and flee מארץ from the land צפון of the north, נאם saith יהוה the LORD: כי for כארבע as the four רוחות winds השׁמים of the heaven, פרשׂתי אתכםאם saith יהוה׃ the LORD.}%
\verse{הוי O ציון Zion, המלטי Deliver thyself, יושׁבת that dwellest בת the daughter בבל׃ of Babylon.}%
\verse{כי For כה thus אמר saith יהוה the LORD צבאות of hosts; אחר After כבוד the glory שׁלחני hath he sent אל me unto הגוים the nations השׁללים which spoiled אתכם כי you: for הנגע he that toucheth בכם נגע you toucheth בבבת the apple עינו׃ of his eye.}%
\verse{כי For, הנני מניף I will shake את ידי mine hand עליהם upon והיו them, and they shall be שׁלל a spoil לעבדיהם to their servants: וידעתם and ye shall know כי that יהוה the LORD צבאות of hosts שׁלחני׃ hath sent}%
\verse{רני Sing ושׂמחי and rejoice, בת O daughter ציון of Zion: כי for, הנני בא I come, ושׁכנתי and I will dwell בתוכך in the midst נאם of thee, saith יהוה׃ the LORD.}%
\verse{ונלוו shall be joined גוים nations רבים And many אל to יהוה the LORD ביום day, ההוא in that והיו and shall be לי לעם my people: ושׁכנתי and I will dwell בתוכך in the midst וידעת of thee, and thou shalt know כי that יהוה the LORD צבאות of hosts שׁלחני hath sent אליך׃ me unto}%
\verse{ונחל shall inherit יהוה And the LORD את יהודה Judah חלקו his portion על in אדמת land, הקדשׁ the holy ובחר and shall choose עוד again. בירושׁלם׃ Jerusalem}%
\verse{הס Be silent, כל O all בשׂר flesh, מפני before יהוה the LORD: כי for נעור he is raised up ממעון habitation. קדשׁו׃ out of his holy}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 3
\verseWithHeading{Zechariah’s Fourth Vision}{ויראני And he showed את יהושׁע me Joshua הכהן priest הגדול the high עמד standing לפני before מלאך the angel יהוה of the LORD, והשׂטן and Satan עמד standing על at ימינו his right hand לשׂטנו׃ to resist}%
\verse{ויאמר said יהוה And the LORD אל unto השׂטן Satan, יגער rebuke יהוה The LORD בך השׂטן thee, O Satan; ויגער rebuke יהוה even the LORD בך הבחר that hath chosen בירושׁלם Jerusalem הלוא thee: not זה this אוד a brand מצל plucked מאשׁ׃ out of the fire?}%
\verse{ויהושׁע Now Joshua היה was לבשׁ clothed בגדים garments, צואים with filthy ועמד and stood לפני before המלאך׃ the angel.}%
\verse{ויען And he answered ויאמר and spoke אל unto העמדים those that stood לפניו before לאמר him, saying, הסירו Take away הבגדים garments הצאים the filthy מעליו from ויאמר him he said, אליו him. And unto ראה Behold, העברתי to pass מעליך from עונך I have caused thine iniquity והלבשׁ thee, and I will clothe אתך מחלצות׃ thee with change of raiment.}%
\verse{ואמר And I said, ישׂימו Let them set צניף miter טהור a fair על upon ראשׁו his head. וישׂימו So they set הצניף miter הטהור a fair על upon ראשׁו his head, וילבשׁהו and clothed בגדים him with garments. ומלאך And the angel יהוה of the LORD עמד׃ stood by.}%
\verse{ויעד protested מלאך And the angel יהוה ביהושׁע unto Joshua, לאמר׃ saying,}%
\verse{כה Thus אמר saith יהוה the LORD צבאות of hosts; אם If בדרכי in my ways, תלך thou wilt walk ואם and if את משׁמרתי my charge, תשׁמר thou wilt keep וגם shalt also אתה then thou תדין judge את ביתי my house, וגם and shalt also תשׁמר keep את חצרי my courts, ונתתי and I will give לך מהלכים thee places to walk בין among העמדים that stand האלה׃ these}%
\verse{שׁמע Hear נא now, יהושׁע O Joshua הכהן priest, הגדול the high אתה thou, ורעיך and thy fellows הישׁבים that sit לפניך before כי thee: for אנשׁי men מופת wondered המה they כי at: for, הנני מביא I will bring forth את עבדי my servant צמח׃ the BRANCH.}%
\verse{כי For הנה behold האבן the stone אשׁר that נתתי I have laid לפני before יהושׁע Joshua; על upon אבן stone אחת one שׁבעה seven עינים eyes: הנני מפתח I will engrave פתחה the graving נאם thereof, saith יהוה the LORD צבאות of hosts, ומשׁתי and I will remove את עון the iniquity הארץ land ההיא of that ביום day. אחד׃ in one}%
\verse{ביום day, ההוא In that נאם saith יהוה the LORD צבאות of hosts, תקראו shall ye call אישׁ every man לרעהו his neighbor אל under תחת under גפן the vine ואל and under תחת and under תאנה׃ the fig tree.}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 4
\verseWithHeading{Zechariah’s Fifth Vision}{וישׁב with me came again, המלאך And the angel הדבר that talked בי ויעירני and waked כאישׁ me, as a man אשׁר that יעור is wakened משׁנתו׃ out of his sleep,}%
\verse{ויאמר And said אלי unto מה me, What אתה thou? ראה seest ויאמר And I said, ראיתי I have looked, והנה and behold מנורת a candlestick זהב gold, כלה all וגלה with a bowl על upon ראשׁה the top ושׁבעה of it, and his seven נרתיה lamps עליה thereon, שׁבעה and seven ושׁבעה to the seven מוצקות pipes לנרות lamps, אשׁר which על upon ראשׁה׃ the top}%
\verse{ושׁנים And two זיתים olive trees עליה by אחד it, one מימין upon the right הגלה of the bowl, ואחד and the other על upon שׂמאלה׃ the left}%
\verse{ואען So I answered ואמר and spoke אל to המלאך the angel הדבר that talked בי לאמר with me, saying, מה What אלה these, אדני׃ my lord?}%
\verse{ויען with me answered המלאך Then the angel הדבר that talked בי ויאמר and said אלי unto הלוא thou not ידעת me, Knowest מה what המה אלה these ואמר be? And I said, לא No, אדני׃ my lord.}%
\verse{ויען Then he answered ויאמר and spoke אלי unto לאמר me, saying, זה This דבר the word יהוה of the LORD אל unto זרבבל Zerubbabel, לאמר saying, לא Not בחיל by might, ולא nor בכח by power, כי but אם but ברוחי by my spirit, אמר saith יהוה the LORD צבאות׃ of hosts.}%
\verse{מי Who אתה thou, הר mountain? הגדול O great לפני before זרבבל Zerubbabel למישׁר a plain: והוציא and he shall bring forth את האבן the headstone הראשׁה the headstone תשׁאות shoutings, חן Grace, חן׃ grace}%
\verse{ויהי came דבר Moreover the word יהוה of the LORD אלי unto לאמר׃ me, saying,}%
\verse{ידי The hands זרבבל of Zerubbabel יסדו הבית house; הזה of this וידיו his hands תבצענה shall also finish וידעת it; and thou shalt know כי that יהוה the LORD צבאות of hosts שׁלחני hath sent אליכם׃ me unto}%
\verse{כי For מי who בז ליום the day קטנות of small things? ושׂמחו for they shall rejoice, וראו and shall see את האבן the plummet הבדיל the plummet ביד in the hand זרבבל of Zerubbabel שׁבעה seven; אלה those עיני the eyes יהוה of the LORD, המה they משׁוטטים which run to and fro בכל through the whole הארץ׃ earth.}%
\verse{ואען Then answered ואמר I, and said אליו unto מה him, What שׁני two הזיתים olive trees האלה these על upon ימין the right המנורה of the candlestick ועל and upon שׂמאולה׃ the left}%
\verse{ואען And I answered שׁנית again, ואמר and said אליו unto מה him, What שׁתי two שׁבלי branches הזיתים olive אשׁר which ביד through שׁני the two צנתרות pipes הזהב golden המריקים empty מעליהם out of הזהב׃ the golden}%
\verse{ויאמר me and said, אלי לאמר And I said, הלוא thou not ידעת Knowest מה what אלה these ואמר לא No, אדני׃ my lord.}%
\verse{ויאמר Then said אלה he, These שׁני the two בני anointed ones, היצהר anointed ones, העמדים that stand על by אדון the Lord כל of the whole הארץ׃ earth.}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 5
\verseWithHeading{Zechariah’s Sixth Vision}{ואשׁוב Then I turned, ואשׂא and lifted up עיני mine eyes, ואראה and looked, והנה and behold מגלה roll. עפה׃ a flying}%
\verse{ויאמר And he said אלי unto מה me, What אתה thou? ראה seest ואמר And I answered, אני I ראה see מגלה roll; עפה a flying ארכה the length עשׂרים thereof twenty באמה cubits, ורחבה and the breadth עשׂר thereof ten באמה׃ cubits.}%
\verse{ויאמר Then said אלי he unto זאת me, This האלה the curse היוצאת that goeth forth על over פני the face כל of the whole הארץ earth: כי for כל every one הגנב that stealeth מזה on this side כמוה according to it; נקה shall be cut off וכל and every הנשׁבע one that sweareth מזה on that side כמוה according to it. נקה׃ shall be cut off}%
\verse{הוצאתיה I will bring it forth, נאם saith יהוה צבאות of hosts, ובאה and it shall enter אל into בית the house הגנב of the thief, ואל and into בית the house הנשׁבע of him that sweareth בשׁמי by my name: לשׁקר falsely ולנה and it shall remain בתוך in the midst ביתו of his house, וכלתו and shall consume ואת עציו it with the timber ואת אבניו׃ thereof and the stones}%
\verseWithHeading{Zechariah’s Seventh Vision}{ויצא with me went forth, המלאך Then the angel הדבר that talked בי ויאמר and said אלי unto שׂא me, Lift up נא now עיניך thine eyes, וראה and see מה what היוצאת that goeth forth. הזאת׃ this}%
\verse{ואמר And I said, מה What היא it? ויאמר And he said, זאת This האיפה an ephah היוצאת that goeth forth. ויאמר He said זאת moreover, This עינם their resemblance בכל through all הארץ׃ the earth.}%
\verse{והנה And, behold, ככר a talent עפרת of lead: נשׂאת there was lifted up וזאת and this אשׁה woman אחת a יושׁבת that sitteth בתוך in the midst האיפה׃ of the ephah.}%
\verse{ויאמר And he said, זאת This הרשׁעה wickedness. וישׁלך And he cast אתה אל it into תוך the midst האיפה of the ephah; וישׁלך and he cast את אבן the weight העפרת of lead אל upon פיה׃ the mouth}%
\verse{ואשׂא Then lifted I up עיני mine eyes, וארא and looked, והנה and, behold, שׁתים two נשׁים women, יוצאות there came out ורוח and the wind בכנפיהם in their wings; ולהנה for they כנפים had wings ככנפי like the wings החסידה of a stork: ותשׂאנה and they lifted up את האיפה the ephah בין between הארץ the earth ובין השׁמים׃ and the heaven.}%
\verse{ואמר Then said אל I to המלאך the angel הדבר that talked בי אנה with me, Whither המה do these מולכות bear את האיפה׃ the ephah?}%
\verse{ויאמר And he said אלי unto לבנות me, To build לה בית it a house בארץ in the land שׁנער of Shinar: והוכן and it shall be established, והניחה שׁם there על upon מכנתה׃ her own base.}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 6
\verseWithHeading{Zechariah’s Eighth Vision}{ואשׁב And I turned, ואשׂא and lifted up עיני mine eyes, ואראה and looked, והנה and, behold, ארבע מרכבותצאות there came four chariots out מבין from between שׁני two ההרים mountains; וההרים and the mountains הרי mountains נחשׁת׃ of brass.}%
\verse{במרכבה chariot הראשׁנה In the first סוסים horses; אדמים red ובמרכבה chariot השׁנית and in the second סוסים horses; שׁחרים׃ black}%
\verse{ובמרכבה chariot השׁלשׁית And in the third סוסים horses; לבנים white ובמרכבה chariot הרבעית and in the fourth סוסים horses. ברדים grizzled אמצים׃ and bay}%
\verse{ואען Then I answered ואמר and said אל unto המלאך the angel הדבר that talked בי מה with me, What אלה these, אדני׃ my lord?}%
\verse{ויען answered המלאך And the angel ויאמר and said אלי unto אלה me, These ארבע the four רחות spirits השׁמים of the heavens, יוצאות which go forth מהתיצב from standing על before אדון the Lord כל of all הארץ׃ the earth.}%
\verse{אשׁר which בה הסוסים horses השׁחרים The black יצאים therein go forth אל into ארץ country; צפון the north והלבנים and the white יצאו go forth אל after אחריהם after והברדים them; and the grizzled יצאו go forth אל toward ארץ country. התימן׃ the south}%
\verse{והאמצים And the bay יצאו went forth, ויבקשׁו and sought ללכת to go להתהלך that they might walk to and fro בארץ through the earth: ויאמר and he said, לכו Get you hence, התהלכו walk to and fro בארץ through the earth. ותתהלכנה So they walked to and fro בארץ׃ through the earth.}%
\verse{ויזעק Then cried אתי וידבר he upon me, and spoke אלי unto לאמר me, saying, ראה Behold, היוצאים these that go אל toward ארץ country צפון the north הניחו have quieted את רוחי my spirit בארץ country. צפון׃ in the north}%
\verseWithHeading{The Branch is Crowned}{ויהי came דבר And the word יהוה of the LORD אלי unto לאמר׃ me, saying,}%
\verse{לקוח Take מאת הגולה the captivity, מחלדי ומאתוביה Tobijah, ומאת ידעיה Jedaiah, ובאת are come אתה thou ביום day, ההוא the same ובאת and come בית into the house יאשׁיה of Josiah בן the son צפניה of Zephaniah; אשׁר which באו and go מבבל׃}%
\verse{ולקחת Then take כסף silver וזהב and gold, ועשׂית and make עטרות crowns, ושׂמת and set בראשׁ upon the head יהושׁע of Joshua בן the son יהוצדק of Josedech, הכהן priest; הגדול׃ the high}%
\verse{ואמרת And speak אליו unto לאמר him, saying, כה Thus אמר speaketh יהוה the LORD צבאות of hosts, לאמר saying, הנה Behold אישׁ the man צמח The BRANCH; שׁמו whose name ומתחתיו out of his place, יצמח and he shall grow up ובנה and he shall build את היכל the temple יהוה׃ of the LORD:}%
\verse{והוא Even he יבנה shall build את היכל the temple יהוה of the LORD; והוא and he ישׂא shall bear הוד the glory, וישׁב and shall sit ומשׁל and rule על upon כסאו his throne; והיה and he shall be כהן a priest על upon כסאו his throne: ועצת and the counsel שׁלום of peace תהיה shall be בין between שׁניהם׃ them both.}%
\verse{והעטרת And the crowns תהיה shall be לחלם to Helem, ולטוביה and to Tobijah, ולידעיה and to Jedaiah, ולחן and to Hen בן the son צפניה of Zephaniah, לזכרון for a memorial בהיכל in the temple יהוה׃ of the LORD.}%
\verse{ורחוקים And they far off יבאו shall come ובנו and build בהיכל in the temple יהוה of the LORD, וידעתם and ye shall know כי that יהוה the LORD צבאות of hosts שׁלחני hath sent אליכם me unto והיה you. And shall come to pass, אם if שׁמוע ye will diligently obey תשׁמעון ye will diligently obey בקול the voice יהוה of the LORD אלהיכם׃ your God.}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 7
\verse{ויהי And it came to pass בשׁנת year ארבע in the fourth לדריושׁ Darius, המלך of king היה came דבר the word יהוה of the LORD אל unto זכריה Zechariah בארבעה in the fourth לחדשׁ month, התשׁעי of the ninth בכסלו׃ in Chisleu;}%
\verse{וישׁלח When they had sent בית אל unto the house of God שׂר אצר Sherezer ורגם מלך and Regem-melech, ואנשׁיו לחלות to pray את פני before יהוה׃}%
\verse{לאמר to speak אל unto הכהנים the priests אשׁר which לבית in the house יהוה of the LORD צבאות of hosts, ואל and to הנביאים the prophets, לאמר saying, האבכה Should I weep בחדשׁ month, החמשׁי in the fifth הנזר separating myself, כאשׁר as עשׂיתי I have done זה these כמה so many שׁנים׃ years?}%
\verse{ויהי Then came דבר the word יהוה of the LORD צבאות of hosts אלי unto לאמר׃ me, saying,}%
\verse{אמר Speak אל unto כל all עם the people הארץ of the land, ואל and to הכהנים the priests, לאמר saying, כי When צמתם ye fasted וספוד and mourned בחמישׁי in the fifth ובשׁביעי and seventh וזה even those שׁבעים seventy שׁנה years, הצום did ye at all fast צמתני did ye at all fast אני׃ unto me, to me?}%
\verse{וכי And when תאכלו ye did eat, וכי and when תשׁתו ye did drink, הלוא did not אתם ye האכלים eat ואתם השׁתים׃ and drink}%
\verse{הלוא not את הדברים the words אשׁר which קרא hath cried יהוה the LORD ביד by הנביאים prophets, הראשׁנים the former בהיות was ירושׁלם when Jerusalem ישׁבת inhabited ושׁלוה and in prosperity, ועריה and the cities סביבתיה thereof round about והנגב the south והשׁפלה and the plain? ישׁב׃ her, when inhabited}%
\verse{ויהי came דבר And the word יהוה of the LORD אל unto זכריה Zechariah, לאמר׃ saying,}%
\verse{כה Thus אמר speaketh יהוה the LORD צבאות of hosts, לאמר saying, משׁפט judgment, אמת true שׁפטו Execute וחסד mercy ורחמים and compassions עשׂו and show אישׁ every man את to אחיו׃ his brother:}%
\verse{ואלמנה the widow, ויתום nor the fatherless, גר the stranger, ועני nor the poor; אל not תעשׁקו And oppress ורעת evil אישׁ evil אחיו against his brother אל and let none תחשׁבו of you imagine בלבבכם׃ in your heart.}%
\verse{וימאנו But they refused להקשׁיב to hearken, ויתנו and pulled כתף the shoulder, סררת away ואזניהם their ears, הכבידו and stopped משׁמוע׃ that they should not hear.}%
\verse{ולבם their hearts שׂמו Yea, they made שׁמיר an adamant stone, משׁמוע lest they should hear את התורה the law, ואת הדברים and the words אשׁר which שׁלח hath sent יהוה the LORD צבאות of hosts ברוחו in his spirit ביד by הנביאים prophets: הראשׁנים the former ויהי therefore came קצף wrath גדול a great מאת יהוה the LORD צבאות׃ of hosts.}%
\verse{ויהי Therefore it is come to pass, כאשׁר as קרא he cried, ולא and they would not שׁמעו hear; כן so יקראו they cried, ולא and I would not אשׁמע hear, אמר saith יהוה the LORD צבאות׃ of hosts:}%
\verse{ואסערם But I scattered them with a whirlwind על among כל all הגוים the nations אשׁר whom לא not. ידעום they knew והארץ Thus the land נשׁמה desolate. אחריהם after מעבר them, that no man passed through ומשׁב nor returned: וישׂימו for they laid ארץ land חמדה the pleasant לשׁמה׃ was desolate}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 8
\verseWithHeading{Adonai’s Promise to Jerusalem}{ויהי came דבר Again the word יהוה of the LORD צבאות of hosts לאמר׃ saying,}%
\verse{כה Thus אמר saith יהוה the LORD צבאות of hosts; קנאתי I was jealous לציון for Zion קנאה jealousy, גדולה with great וחמה fury. גדולה for her with great קנאתי׃ and I was jealous}%
\verse{כה Thus אמר saith יהוה the LORD; שׁבתי I am returned אל unto ציון Zion, ושׁכנתי and will dwell בתוך in the midst ירושׁלם of Jerusalem: ונקראה shall be called ירושׁלם and Jerusalem עיר a city האמת of truth; והר and the mountain יהוה of the LORD צבאות of hosts הר mountain. הקדשׁ׃ the holy}%
\verse{כה Thus אמר saith יהוה the LORD צבאות of hosts; עד There shall yet ישׁבו dwell זקנים old men וזקנות and old women ברחבות in the streets ירושׁלם of Jerusalem, ואישׁ and every man משׁענתו with his staff בידו in his hand מרב for very ימים׃ age.}%
\verse{ורחבות And the streets העיר of the city ימלאו shall be full ילדים of boys וילדות and girls משׂחקים playing ברחבתיה׃ in the streets}%
\verse{כה Thus אמר saith יהוה צבאות of hosts; כי If יפלא it be marvelous בעיני in the eyes שׁארית of the remnant העם people הזה of this בימים days, ההם in these גם should it also בעיני in mine eyes? יפלא be marvelous נאם saith יהוה צבאות׃ of hosts.}%
\verse{כה Thus אמר saith יהוה the LORD צבאות of hosts; הנני מושׁיע I will save את עמי my people מארץ country, מזרח from the east ומארץ country; מבוא and from the west השׁמשׁ׃ and from the west}%
\verse{והבאתי And I will bring אתם ושׁכנו them, and they shall dwell בתוך in the midst ירושׁלם of Jerusalem: והיו and they shall be לי לעם my people, ואני and I אהיה will be להם לאלהים their God, באמת in truth ובצדקה׃ and in righteousness.}%
\verse{כה Thus אמר saith יהוה the LORD צבאות of hosts; תחזקנה be strong, ידיכם Let your hands השׁמעים ye that hear בימים days האלה in these את הדברים words האלה these מפי by the mouth הנביאים of the prophets, אשׁר which ביום in the day יסד was laid, בית the foundation of the house יהוה of the LORD צבאות of hosts ההיכל that the temple להבנות׃ might be built.}%
\verse{כי For לפני before הימים days ההם these שׂכר hire האדם for man, לא no נהיה there was ושׂכר any hire הבהמה for beast; איננה nor וליוצא to him that went out ולבא or came in אין neither שׁלום peace מן because of הצר the affliction: ואשׁלח for I set את כל all האדם men אישׁ every one ברעהו׃ against his neighbor.}%
\verse{ועתה But now לא not כימים days, הראשׁנים as in the former אני I לשׁארית unto the residue העם people הזה of this נאם saith יהוה the LORD צבאות׃ of hosts.}%
\verse{כי For זרע the seed השׁלום prosperous; הגפן the vine תתן shall give פריה her fruit, והארץ and the ground תתן shall give את יבולה her increase, והשׁמים and the heavens יתנו shall give טלם their dew; והנחלתי to possess את שׁארית and I will cause the remnant העם people הזה of this את כל all אלה׃ these}%
\verse{והיה And it shall come to pass, כאשׁר as הייתם ye were קללה a curse בגוים among the heathen, בית O house יהודה of Judah, ובית and house ישׂראל of Israel; כן so אושׁיע will I save אתכם והייתם you, and ye shall be ברכה a blessing: אל not, תיראו fear תחזקנה be strong. ידיכם׃ let your hands}%
\verse{כי For כה thus אמר saith יהוה the LORD צבאות of hosts; כאשׁר As זממתי I thought להרע to punish לכם בהקציף אבתיכם you, when your fathers אתי אמר saith יהוה the LORD צבאות of hosts, ולא not: נחמתי׃ and I repented}%
\verse{כן So שׁבתי again זממתי have I thought בימים days האלה in these להיטיב to do well את ירושׁלם unto Jerusalem ואת בית and to the house יהודה of Judah: אל ye not. תיראו׃ fear}%
\verse{אלה These הדברים the things אשׁר that תעשׂו ye shall do; דברו Speak אמת the truth אישׁ ye every man את to רעהו his neighbor; אמת of truth ומשׁפט the judgment שׁלום and peace שׁפטו execute בשׁעריכם׃ in your gates:}%
\verse{ואישׁ אתעת evil רעהו against his neighbor; אל no תחשׁבו of you imagine בלבבכם in your hearts ושׁבעת oath: שׁקר false אל תאהבו and love כי for את כל all אלה these אשׁר that שׂנאתי I hate, נאם saith יהוה׃ the LORD.}%
\verse{ויהי came דבר And the word יהוה of the LORD צבאות of hosts אלי unto לאמר׃ me, saying,}%
\verse{כה Thus אמר saith יהוה the LORD צבאות of hosts; צום The fast הרביעי of the fourth וצום and the fast החמישׁי of the fifth, וצום and the fast השׁביעי of the seventh, וצום and the fast העשׂירי of the tenth, יהיה shall be לבית to the house יהודה of Judah לשׂשׂון joy ולשׂמחה and gladness, ולמעדים feasts; טובים and cheerful והאמת the truth והשׁלום and peace. אהבו׃ therefore love}%
\verse{כה Thus אמר saith יהוה the LORD צבאות of hosts; עד yet אשׁר that יבאו there shall come עמים people, וישׁבי and the inhabitants ערים cities: רבות׃ of many}%
\verse{והלכו shall go ישׁבי And the inhabitants אחת of one אל to אחת another, לאמר saying, נלכה will go הלוך לחלות to pray את פני before יהוה the LORD, ולבקשׁ and to seek את יהוה the LORD צבאות of hosts: אלכה גם also. אני׃ I}%
\verse{ובאו shall come עמים people רבים Yea, many וגוים nations עצומים and strong לבקשׁ to seek את יהוה the LORD צבאות of hosts בירושׁלם in Jerusalem, ולחלות and to pray את פני before יהוה׃ the LORD.}%
\verse{כה Thus אמר saith יהוה the LORD צבאות of hosts; בימים days ההמה In those אשׁר that יחזיקו shall take hold עשׂרה ten אנשׁים men מכל out of all לשׁנות languages הגוים of the nations, והחזיקו even shall take hold בכנף of the skirt אישׁ of him יהודי that is a Jew, לאמר saying, נלכה We will go עמכם with כי you: for שׁמענו we have heard אלהים God עמכם׃ with}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 9
\verse{משׂא The burden דבר of the word יהוה of the LORD בארץ in the land חדרך of Hadrach, ודמשׂק and Damascus מנחתו the rest כי thereof: when ליהוה toward the LORD. עין the eyes אדם of man, וכל as of all שׁבטי the tribes ישׂראל׃ of Israel,}%
\verse{וגם also חמת And Hamath תגבל shall border בה צר thereby; Tyrus, וצידון and Zidon, כי though חכמה wise. מאד׃ it be very}%
\verse{ותבן did build צר And Tyrus מצור herself a stronghold, לה ותצבר and heaped up כסף silver כעפר as the dust, וחרוץ and fine gold כטיט as the mire חוצות׃ of the streets.}%
\verse{הנה Behold, אדני the Lord יורשׁנה will cast her out, והכה and he will smite בים in the sea; חילה her power והיא and she באשׁ with fire. תאכל׃ shall be devoured}%
\verse{תרא shall see אשׁקלון Ashkelon ותירא and fear; ועזה Gaza ותחיל , and be very sorrowful, מאד also and be very sorrowful, ועקרון and Ekron; כי for הבישׁ shall be ashamed; מבטה her expectation ואבד shall perish מלך and the king מעזה ואשׁקלון and Ashkelon לא shall not תשׁב׃ be inhabited.}%
\verse{וישׁב shall dwell ממזר And a bastard באשׁדוד in Ashdod, והכרתי and I will cut off גאון the pride פלשׁתים׃ of the Philistines.}%
\verse{והסרתי And I will take away דמיו his blood מפיו out of his mouth, ושׁקציו and his abominations מבין from between שׁניו his teeth: ונשׁאר but he that remaineth, גם even הוא he, לאלהינו for our God, והיה and he shall be כאלף as a governor ביהודה in Judah, ועקרון and Ekron כיבוסי׃ as a Jebusite.}%
\verse{וחניתי And I will encamp לביתי about mine house מצבה because of the army, מעבר because of him that passeth by, ומשׁב and because of him that returneth: ולא and no יעבר shall pass עליהם through עוד them any more: נגשׂ oppressor כי for עתה now ראיתי have I seen בעיני׃ with mine eyes.}%
\verse{גילי Rejoice מאד greatly, בת O daughter ציון of Zion; הריעי shout, בת O daughter ירושׁלם of Jerusalem: הנה behold, מלכך thy King יבוא cometh לך צדיק just, ונושׁע and having salvation; הוא unto thee: he עני lowly, ורכב and riding על upon חמור an ass, ועל and upon עיר a colt בן the foal אתנות׃ of an ass.}%
\verse{והכרתי And I will cut off רכב the chariot מאפרים וסוס and the horse מירושׁלם ונכרתה shall be cut off: קשׁת bow מלחמה and the battle ודבר and he shall speak שׁלום peace לגוים unto the heathen: ומשׁלו and his dominion מים from sea עד to ים sea, ומנהר and from the river עד to אפסי the ends ארץ׃ of the earth.}%
\verse{גם also, את As for thee בדם by the blood בריתך of thy covenant שׁלחתי I have sent forth אסיריך thy prisoners מבור out of the pit אין wherein no מים׃ water.}%
\verse{שׁובו Turn לבצרון you to the stronghold, אסירי ye prisoners התקוה of hope: גם even היום today מגיד do I declare משׁנה double אשׁיב׃ I will render}%
\verse{כי When דרכתי I have bent לי יהודה Judah קשׁת the bow מלאתי for me, filled אפרים with Ephraim, ועוררתי and raised בניך up thy sons, ציון O Zion, על against בניך thy sons, יון O Greece, ושׂמתיך and made כחרב thee as the sword גבור׃ of a mighty man.}%
\verse{ויהוה And the LORD עליהם over יראה shall be seen ויצא shall go forth כברק as the lightning: חצו them, and his arrow ואדני and the Lord יהוה GOD בשׁופר the trumpet, יתקע shall blow והלך and shall go בסערות with whirlwinds תימן׃ of the south.}%
\verse{יהוה צבאות of hosts יגן shall defend עליהם shall defend ואכלו them; and they shall devour, וכבשׁו and subdue אבני with slingstones; קלע with slingstones; ושׁתו and they shall drink, המו make a noise כמו as through יין wine; ומלאו and they shall be filled כמזרק like bowls, כזויות as the corners מזבח׃ of the altar.}%
\verse{והושׁיעם shall save יהוה And the LORD אלהיהם their God ביום day ההוא them in that כצאן as the flock עמו of his people: כי for אבני they the stones נזר of a crown, מתנוססות lifted up as an ensign על upon אדמתו׃ his land.}%
\verse{כי For מה how טובו great his goodness, ומה and how יפיו great his beauty! דגן corn בחורים shall make the young men ותירושׁ and new wine ינובב cheerful, בתלות׃ the maids.}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 10
\verseWithHeading{Adonai Gives Rain}{שׁאלו Ask מיהוה מטר rain בעת in the time מלקושׁ of the latter rain; יהוה עשׂה shall make חזיזים bright clouds, ומטר of rain, גשׁם them showers יתן and give להם לאישׁ to every one עשׂב grass בשׂדה׃ in the field.}%
\verse{כי For התרפים the idols דברו have spoken און vanity, והקוסמים and the diviners חזו have seen שׁקר a lie, וחלמות dreams; השׁוא false ידברו and have told הבל in vain: ינחמון they comfort על therefore כן therefore נסעו they went their way כמו as צאן a flock, יענו they were troubled, כי because אין no רעה׃ shepherd.}%
\verse{על against הרעים the shepherds, חרה was kindled אפי Mine anger ועל העתודים the goats: אפקוד hath visited כי for פקד יהוה the LORD צבאות of hosts את עדרו his flock את בית the house יהודה of Judah, ושׂם and hath made אותם כסוס horse הודו them as his goodly במלחמה׃ in the battle.}%
\verse{ממנו Out of פנה the corner, ממנו out of יתד him the nail, ממנו out of קשׁת bow, מלחמה him the battle ממנו out of יצא him came forth כל him every נוגשׂ oppressor יחדו׃ together.}%
\verse{והיו And they shall be כגברים as mighty בוסים which tread down בטיט in the mire חוצות of the streets במלחמה in the battle: ונלחמו and they shall fight, כי because יהוה the LORD עמם with והבישׁו רכבי them, and the riders סוסים׃ on horses}%
\verse{וגברתי And I will strengthen את בית the house יהודה of Judah, ואת בית the house יוסף of Joseph, אושׁיע and I will save והושׁבותים and I will bring them again כי for רחמתים I have mercy upon והיו them: and they shall be כאשׁר as though לא I had not זנחתים cast them off: כי for אני I יהוה the LORD אלהיהם their God, ואענם׃ and will hear}%
\verse{והיו shall be כגבור like a mighty אפרים And Ephraim ושׂמח shall rejoice לבם and their heart כמו as through יין wine: ובניהם yea, their children יראו shall see ושׂמחו יגל shall rejoice לבם their heart ביהוה׃ in the LORD.}%
\verse{אשׁרקה I will hiss להם ואקבצם for them, and gather כי them; for פדיתים I have redeemed ורבו them: and they shall increase כמו as רבו׃ they have increased.}%
\verse{ואזרעם And I will sow בעמים them among the people: ובמרחקים me in far countries; יזכרוני and they shall remember וחיו and they shall live את with בניהם their children, ושׁבו׃ and turn again.}%
\verse{והשׁיבותים I will bring them again מארץ also out of the land מצרים of Egypt, ומאשׁור אקבצם and gather ואל them into ארץ the land גלעד of Gilead ולבנון and Lebanon; אביאם and I will bring ולא and shall not ימצא׃ be found}%
\verse{ועבר And he shall pass בים through the sea צרה with affliction, והכה and shall smite בים in the sea, גלים the waves והבישׁו shall dry up: כל and all מצולות the deeps יאר of the river והורד shall be brought down, גאון and the pride אשׁור of Assyria ושׁבט and the scepter מצרים of Egypt יסור׃ shall depart away.}%
\verse{וגברתים And I will strengthen ביהוה them in the LORD; ובשׁמו in his name, יתהלכו and they shall walk up and down נאם saith יהוה׃ the LORD.}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 11
\verse{פתח Open לבנון O Lebanon, דלתיך thy doors, ותאכל may devour אשׁ that the fire בארזיך׃ thy cedars.}%
\verse{הילל Howl, ברושׁ fir tree; כי for נפל is fallen; ארז the cedar אשׁר because אדרים the mighty שׁדדו are spoiled: הילילו howl, אלוני O ye oaks בשׁן of Bashan; כי for ירד is come down. יער the forest הבצור׃}%
\verse{קול a voice יללת of the howling הרעים of the shepherds; כי for שׁדדה is spoiled: אדרתם their glory קול a voice שׁאגת of the roaring כפירים of young lions; כי for שׁדד is spoiled. גאון the pride הירדן׃ of Jordan}%
\verseWithHeading{The Two Shepherds}{כה Thus אמר saith יהוה the LORD אלהי my God; רעה Feed את צאן the flock ההרגה׃ of the slaughter;}%
\verse{אשׁר Whose קניהן possessors יהרגן slay ולא them, and hold themselves not guilty: יאשׁמו them, and hold themselves not guilty: ומכריהן and they that sell יאמר them say, ברוך Blessed יהוה the LORD; ואעשׁר for I am rich: ורעיהם and their own shepherds לא them not. יחמול pity עליהן׃ pity}%
\verse{כי For לא I will no אחמול pity עוד more על pity ישׁבי the inhabitants הארץ of the land, נאם saith יהוה the LORD: והנה but, lo, אנכי I ממציא will deliver את האדם the men אישׁ every one ביד hand, רעהו into his neighbor's וביד and into the hand מלכו of his king: וכתתו and they shall smite את הארץ the land, ולא I will not אציל deliver מידם׃ and out of their hand}%
\verse{וארעה And I will feed את צאן the flock ההרגה of slaughter, לכן עניי you, O poor הצאן of the flock. ואקח And I took לי שׁני unto me two מקלות staves; לאחד the one קראתי I called נעם Beauty, ולאחד and the other קראתי I called חבלים Bands; וארעה and I fed את הצאן׃ the flock.}%
\verse{ואכחד also I cut off את שׁלשׁת Three הרעים shepherds בירח month; אחד in one ותקצר loathed נפשׁי and my soul בהם וגם also נפשׁם them, and their soul בחלה׃ abhorred}%
\verse{ואמר Then said לא I, I will not ארעה feed אתכם המתה you: that that dieth, תמות let it die; והנכחדת and that that is to be cut off, תכחד let it be cut off; והנשׁארות and let the rest תאכלנה eat אשׁה every one את בשׂר the flesh רעותה׃ of another.}%
\verse{ואקח And I took את מקלי my staff, את נעם Beauty, ואגדע אתוהפיר that I might break את בריתי my covenant אשׁר which כרתי I had made את with כל all העמים׃ the people.}%
\verse{ותפר And it was broken ביום day: ההוא in that וידעו me knew כן and so עניי the poor הצאן of the flock השׁמרים that waited upon אתי כי that דבר the word יהוה of the LORD. הוא׃ it}%
\verse{ואמר And I said אליהם unto אם them, If טוב בעיניכם ye think good, הבו give שׂכרי my price; ואם and if לא not, חדלו forbear. וישׁקלו So they weighed for את שׂכרי my price שׁלשׁים thirty כסף׃ of silver.}%
\verse{ויאמר said יהוה אלי unto השׁליכהו me, Cast אל it unto היוצר the potter: אדר a goodly היקר price אשׁר that יקרתי I was prised at מעליהם of ואקחה them. And I took שׁלשׁים the thirty הכסף of silver, ואשׁליך and cast אתו בית in the house יהוה אל them to היוצר׃ the potter}%
\verse{ואגדע Then I cut asunder את מקלי staff, השׁני mine other את החבלים Bands, להפר that I might break את האחוה the brotherhood בין between יהודה Judah ובין ישׂראל׃ and Israel.}%
\verse{ויאמר said יהוה And the LORD אלי unto עוד unto thee yet קח me, Take לך כלי the instruments רעה shepherd. אולי׃ of a foolish}%
\verse{כי For, הנה lo, אנכי I מקים will raise רעה up a shepherd בארץ in the land, הנכחדות those that be cut off, לא shall not יפקד visit הנער לא neither יבקשׁ shall seek והנשׁברת that that is broken, לא nor ירפא heal הנצבה that that standeth still: לא nor יכלכל feed ובשׂר the flesh הבריאה of the fat, יאכל but he shall eat ופרסיהן and tear their claws in pieces. יפרק׃ and tear their claws in pieces.}%
\verse{הוי Woe רעי shepherd האליל to the idol עזבי that leaveth הצאן the flock! חרב the sword על upon זרועו his arm, ועל and upon עין eye: ימינו his right זרעו his arm יבושׁ shall be clean dried up, תיבשׁ shall be clean dried up, ועין eye ימינו and his right כהה shall be utterly darkened. תכהה׃ shall be utterly darkened.}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 12
\verseWithHeading{The Day of Adonai}{משׂא The burden דבר of the word יהוה of the LORD על for ישׂראל Israel, נאם saith יהוה the LORD, נטה which stretcheth forth שׁמים the heavens, ויסד and layeth the foundation ארץ of the earth, ויצר and formeth רוח the spirit אדם of man בקרבו׃ within}%
\verse{הנה Behold, אנכי I שׂם will make את ירושׁלם Jerusalem סף a cup רעל of trembling לכל unto all העמים the people סביב round about, וגם both על against יהודה Judah יהיה when they shall be במצור in the siege על against ירושׁלם׃ Jerusalem.}%
\verse{והיה ביום day ההוא And in that אשׂים will I make את ירושׁלם Jerusalem אבן stone מעמסה a burdensome לכל for all העמים people: כל all עמסיה that burden שׂרוט themselves with it shall be cut in pieces, ישׂרטו themselves with it shall be cut in pieces, ונאספו be gathered together עליה against כל though all גויי the people הארץ׃ of the earth}%
\verse{ביום day, ההוא In that נאם saith יהוה the LORD, אכה I will smite כל every סוס horse בתמהון with astonishment, ורכבו and his rider בשׁגעון with madness: ועל upon בית the house יהודה of Judah, אפקח and I will open את עיני mine eyes וכל every סוס horse העמים of the people אכה and will smite בעורון׃ with blindness.}%
\verse{ואמרו shall say אלפי And the governors יהודה of Judah בלבם in their heart, אמצה my strength לי ישׁבי The inhabitants ירושׁלם of Jerusalem ביהוה in the LORD צבאות of hosts אלהיהם׃ their God.}%
\verse{ביום day ההוא In that אשׂים will I make את אלפי the governors יהודה of Judah ככיור like a hearth אשׁ of fire בעצים among the wood, וכלפיד and like a torch אשׁ of fire בעמיר in a sheaf; ואכלו and they shall devour על on ימין the right hand ועל and on שׂמאול the left: את כל all העמים the people סביב round about, וישׁבה shall be inhabited ירושׁלם and Jerusalem עוד again תחתיה in her own place, בירושׁלם׃ in Jerusalem.}%
\verse{והושׁיע also shall save יהוה The LORD את אהלי the tents יהודה of Judah בראשׁנה first, למען that לא do not תגדל magnify תפארת the glory בית of the house דויד of David ותפארת and the glory ישׁב of the inhabitants ירושׁלם of Jerusalem על against יהודה׃ Judah.}%
\verse{ביום day ההוא In that יגן defend יהוה shall the LORD בעד defend יושׁב the inhabitants ירושׁלם of Jerusalem; והיה shall be הנכשׁל and he that is feeble בהם ביום day ההוא among them at that כדויד as David; ובית and the house דויד of David כאלהים as God, כמלאך as the angel יהוה of the LORD לפניהם׃ before}%
\verse{והיה And it shall come to pass ביום day, ההוא in that אבקשׁ I will seek להשׁמיד to destroy את כל all הגוים the nations הבאים that come על against ירושׁלם׃ Jerusalem.}%
\verse{ושׁפכתי And I will pour על upon בית the house דויד of David, ועל and upon יושׁב the inhabitants ירושׁלם of Jerusalem, רוח the spirit חן of grace ותחנונים and of supplications: והביטו and they shall look אלי upon את אשׁר me whom דקרו they have pierced, וספדו and they shall mourn עליו for כמספד על for היחיד only והמר and shall be in bitterness עליו for כהמר him, as one that is in bitterness על for הבכור׃ firstborn.}%
\verse{ביום day ההוא In that יגדל shall there be a great המספד mourning בירושׁלם in Jerusalem, כמספד as the mourning הדד רמון of Hadadrimmon בבקעת in the valley מגדון׃ of Megiddon.}%
\verse{וספדה shall mourn, הארץ And the land משׁפחות every family משׁפחות every family לבד apart; משׁפחת the family בית of the house דויד of David לבד apart, ונשׁיהם and their wives לבד apart; משׁפחת the family בית of the house נתן of Nathan לבד apart, ונשׁיהם and their wives לבד׃ apart;}%
\verse{משׁפחת The family בית of the house לוי of Levi לבד apart, ונשׁיהם and their wives לבד apart; משׁפחת the family השׁמעי of Shimei לבד apart, ונשׁיהם and their wives לבד׃ apart;}%
\verse{כל All המשׁפחות the families הנשׁארות that remain, משׁפחת every family משׁפחת every family לבד apart, ונשׁיהם and their wives לבד׃ apart.}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 13
\verse{ביום day ההוא In that יהיה there shall be מקור a fountain נפתח opened לבית to the house דויד of David ולישׁבי and to the inhabitants ירושׁלם of Jerusalem לחטאת for sin ולנדה׃ and for uncleanness.}%
\verse{והיה And it shall come to pass ביום day, ההוא in that נאם saith יהוה the LORD צבאות of hosts, אכרית I will cut off את שׁמות the names העצבים of the idols מן out of הארץ the land, ולא and they shall no יזכרו be remembered: עוד more וגם and also את הנביאים I will cause the prophets ואת רוח spirit הטמאה and the unclean אעביר to pass מן out of הארץ׃ the land.}%
\verse{והיה And it shall come to pass, כי when ינבא prophesy, אישׁ any עוד shall yet ואמרו him shall say אליו unto אביו then his father ואמו and his mother ילדיו that begot לא him, Thou shalt not תחיה live; כי for שׁקר lies דברת thou speakest בשׁם in the name יהוה of the LORD: ודקרהו him shall thrust him through אביהו and his father ואמו and his mother ילדיו that begot בהנבאו׃ when he prophesieth.}%
\verse{והיה And it shall come to pass ביום day, ההוא in that יבשׁו shall be ashamed הנביאים the prophets אישׁ every one מחזינו of his vision, בהנבאתו when he hath prophesied; ולא neither ילבשׁו shall they wear אדרת garment שׂער a rough למען to כחשׁ׃ deceive:}%
\verse{ואמר But he shall say, לא no נביא prophet, אנכי I אישׁ a husbandman; עבד אדמהנכי I כי for אדם man הקנני taught me to keep cattle מנעורי׃ from my youth.}%
\verse{ואמר And shall say אליו unto מה him, What המכות wounds האלה these בין in ידיך thine hands? ואמר Then he shall answer, אשׁר with which הכיתי I was wounded בית the house מאהבי׃ of my friends.}%
\verseWithHeading{The Sword Shall Strike the Shepherd}{חרב O sword, עורי Awake, על against רעי ועל and against גבר the man עמיתי my fellow, נאם saith יהוה צבאות of hosts: הך smite את הרעה my shepherd, ותפוצין shall be scattered: הצאן and the sheep והשׁבתי and I will turn ידי mine hand על upon הצערים׃ the little ones.}%
\verse{והיה And it shall come to pass, בכל in all הארץ the land, נאם saith יהוה the LORD, פי parts שׁנים two בה יכרתו therein shall be cut off יגועו die; והשׁלשׁית but the third יותר׃ shall be left}%
\verse{והבאתי And I will bring את השׁלשׁית the third part באשׁ through the fire, וצרפתים and will refine כצרף is refined, את הכסף them as silver ובחנתים and will try כבחן is tried: את הזהב them as gold הוא they יקרא shall call בשׁמי on my name, ואני and I אענה will hear אתו אמרתי them: I will say, עמי my people: הוא It והוא and they יאמר shall say, יהוה The LORD אלהי׃ my God.}%
\end{biblechapter}%
\begin{biblechapter}% Zechariah 14
\verseWithHeading{Zechariah’s Apocalyptic Vision}{הנה Behold, יום the day בא cometh, ליהוה of the LORD וחלק shall be divided שׁללך and thy spoil בקרבך׃ in the midst}%
\verse{ואספתי For I will gather את כל all הגוים nations אל against ירושׁלם Jerusalem למלחמה to battle; ונלכדה shall be taken, העיר and the city ונשׁסו rifled, הבתים and the houses והנשׁים and the women תשׁגלנה ravished; ויצא shall go forth חצי and half העיר of the city בגולה into captivity, ויתר and the residue העם of the people לא shall not יכרת be cut off מן from העיר׃ the city.}%
\verse{ויצא go forth, יהוה ונלחם and fight against בגוים nations, ההם those כיום as when הלחמו he fought ביום in the day קרב׃ of battle.}%
\verse{ועמדו shall stand רגליו And his feet ביום day ההוא in that על upon הר the mount הזתים of Olives, אשׁר which על before פני before ירושׁלם Jerusalem מקדם on the east, ונבקע shall cleave הר and the mount הזיתים of Olives מחציו in the midst מזרחה thereof toward the east וימה and toward the west, גיא valley; גדולה great מאד a very ומשׁ shall remove חצי and half ההר of the mountain צפונה toward the north, וחציו and half נגבה׃ of it toward the south.}%
\verse{ונסתם And ye shall flee גיא the valley הרי of the mountains; כי for יגיע shall reach גי the valley הרים of the mountains אל unto אצל Azal: ונסתם yea, ye shall flee, כאשׁר like as נסתם ye fled מפני from before הרעשׁ the earthquake בימי in the days עזיה of Uzziah מלך king יהודה of Judah: ובא shall come, יהוה and the LORD אלהי my God כל all קדשׁים the saints עמך׃ with}%
\verse{והיה And it shall come to pass ביום day, ההוא in that לא shall not יהיה be אור the light יקרות clear, יקפאון׃ dark:}%
\verse{והיה But it shall be יום day אחד one הוא which יודע shall be known ליהוה to the LORD, לא not יום day, ולא nor לילה night: והיה but it shall come to pass, לעת time ערב at evening יהיה it shall be אור׃ light.}%
\verse{והיה And it shall be ביום day, ההוא in that יצאו shall go out מים waters חיים living מירושׁלם חצים half אל of them toward הים sea, הקדמוני the former וחצים and half אל of them toward הים sea: האחרון the hinder בקיץ in summer ובחרף and in winter יהיה׃ shall it be.}%
\verse{והיה shall be יהוה And the LORD למלך king על over כל all הארץ the earth: ביום day ההוא in that יהיה shall there be יהוה LORD, אחד one ושׁמו and his name אחד׃ one.}%
\verse{יסוב shall be turned כל All הארץ the land כערבה as a plain מגבע לרמון to Rimmon נגב south ירושׁלם of Jerusalem: וראמה and it shall be lifted up, וישׁבה and inhabited תחתיה in her place, למשׁער gate בנימן from Benjamin's עד unto מקום the place שׁער gate, הראשׁון of the first עד unto שׁער gate, הפנים the corner ומגדל and the tower חננאל of Hananeel עד unto יקבי winepresses. המלך׃ the king's}%
\verse{וישׁבו And shall dwell בה וחרם utter destruction; לא no יהיה in it, and there shall be עוד more וישׁבה inhabited. ירושׁלם but Jerusalem לבטח׃ shall be safely}%
\verse{וזאת And this תהיה shall be המגפה the plague אשׁר wherewith יגף will smite יהוה the LORD את כל all העמים the people אשׁר that צבאו have fought על against ירושׁלם Jerusalem; המק shall consume away בשׂרו Their flesh והוא while they עמד stand על upon their feet, רגליו upon their feet, ועיניו and their eyes תמקנה shall consume away בחריהן in their holes, ולשׁונו and their tongue תמק shall consume away בפיהם׃ in their mouth.}%
\verse{והיה And it shall come to pass ביום day, ההוא in that תהיה shall be מהומת tumult יהוה from the LORD רבה a great בהם והחזיקו among them; and they shall lay hold אישׁ every one יד on the hand רעהו of his neighbor, ועלתה shall rise up ידו and his hand על against יד the hand רעהו׃ of his neighbor.}%
\verse{וגם also יהודה And Judah תלחם shall fight בירושׁלם at Jerusalem; ואסף shall be gathered together, חיל and the wealth כל of all הגוים the heathen סביב round about זהב gold, וכסף and silver, ובגדים and apparel, לרב abundance. מאד׃ in great}%
\verse{וכן And so תהיה shall be מגפת the plague הסוס of the horse, הפרד of the mule, הגמל of the camel, והחמור and of the ass, וכל and of all הבהמה the beasts אשׁר that יהיה shall be במחנות tents, ההמה in these כמגפה plague. הזאת׃ as this}%
\verse{והיה And it shall come to pass, כל every one הנותר that is left מכל of all הגוים the nations הבאים which came על against ירושׁלם Jerusalem ועלו shall even go up מדי from שׁנה year בשׁנה to year להשׁתחות to worship למלך the King, יהוה the LORD צבאות of hosts, ולחג and to keep את חג the feast הסכות׃ of tabernacles.}%
\verse{והיה And it shall be, אשׁר whoso לא will not יעלה come up מאת משׁפחות the families הארץ of the earth אל unto ירושׁלם Jerusalem להשׁתחות to worship למלך the King, יהוה the LORD צבאות of hosts, ולא no עליהם even upon יהיה them shall be הגשׁם׃ rain.}%
\verse{ואם And if משׁפחת the family מצרים of Egypt לא go not up, תעלה go not up, ולא not, באה and come ולא that no עליהם תהיה there shall be המגפה the plague, אשׁר wherewith יגף will smite יהוה the LORD את הגוים the heathen אשׁר that לא come not up יעלו come not up לחג to keep את חג the feast הסכות׃ of tabernacles.}%
\verse{זאת This תהיה shall be חטאת the punishment מצרים of Egypt, וחטאת and the punishment כל of all הגוים nations אשׁר that לא come not up יעלו come not up לחג to keep את חג the feast הסכות׃ of tabernacles.}%
\verse{ביום day ההוא In that יהיה shall there be על upon מצלות the bells הסוס of the horses, קדשׁ HOLINESS ליהוה UNTO THE LORD; והיה shall be הסירות and the pots בבית house יהוה in the LORD's כמזרקים like the bowls לפני before המזבח׃ the altar.}%
\verse{והיה shall be כל Yea, every סיר pot בירושׁלם in Jerusalem וביהודה and in Judah קדשׁ holiness ליהוה unto the LORD צבאות of hosts: ובאו shall come כל and all הזבחים they that sacrifice ולקחו and take מהם ובשׁלו them, and seethe בהם ולא no יהיה there shall be כנעני the Canaanite עוד more בבית in the house יהוה of the LORD צבאות of hosts. ביום day ההוא׃ therein: and in that}%
\end{biblechapter}%
