\biblebook{Zephaniah}
\begin{biblechapter}% Zephaniah 1
\verseWithHeading{Judah’s Impending Judgment}{דבר The word יהוה of the LORD אשׁר which היה came אל unto צפניה Zephaniah בן the son כושׁי בן the son גדליה of Gedaliah, בן the son אמריה of Amariah, בן the son חזקיה of Hizkiah, בימי in the days יאשׁיהו of Josiah בן the son אמון of Amon, מלך king יהודה׃ of Judah.}%
\verse{אסף אסףל all מעל פניאדמה the land, נאם saith יהוה׃ the LORD.}%
\verse{אסף אדם man ובהמה and beast; אסף עוף the fowls השׁמים of the heaven, ודגי and the fishes הים of the sea, והמכשׁלות and the stumblingblocks את with הרשׁעים the wicked; והכרתי and I will cut off את האדם man מעל פניאדמה the land, נאם saith יהוה׃ the LORD.}%
\verse{ונטיתי I will also stretch out ידי mine hand על upon יהודה Judah, ועל and upon כל all יושׁבי the inhabitants ירושׁלם of Jerusalem; והכרתי and I will cut off מן from המקום place, הזה this את שׁאר the remnant הבעל of Baal את שׁם the name הכמרים of the Chemarims עם with הכהנים׃ the priests;}%
\verse{ואת המשׁתחוים And them that worship על upon הגגות the housetops; לצבא the host השׁמים of heaven ואת המשׁתחוים and them that worship הנשׁבעים that swear ליהוה by the LORD, והנשׁבעים and that swear במלכם׃ by Malcham;}%
\verse{ואת הנסוגים And them that are turned back מאחרי from יהוה ואשׁר and that לא have not בקשׁו sought את יהוהלא nor דרשׁהו׃ inquired}%
\verse{הס Hold thy peace מפני at the presence אדני of the Lord יהוה GOD: כי for קרוב at hand: יום the day יהוה of the LORD כי for הכין hath prepared יהוה the LORD זבח a sacrifice, הקדישׁ he hath bid קראיו׃ his guests.}%
\verse{והיה And it shall come to pass ביום in the day זבח sacrifice, יהוה of the LORD's ופקדתי עלשׂרים the princes, ועל בני children, המלך and the king's ועל כל and all הלבשׁים such as are clothed מלבושׁ apparel. נכרי׃ with strange}%
\verse{ופקדתי על on כל all הדולג those that leap על המפתן the threshold, ביום day ההוא In the same הממלאים which fill בית houses אדניהם their masters' חמס with violence ומרמה׃ and deceit.}%
\verse{והיה And it shall come to pass ביום day, ההוא in that נאם saith יהוה the LORD, קול the noise צעקה of a cry משׁער gate, הדגים from the fish ויללה and a howling מן gate, המשׁנה the second, ושׁבר crashing גדול and a great מהגבעות׃ from the hills.}%
\verse{הילילו Howl, ישׁבי ye inhabitants המכתשׁ of Maktesh, כי for נדמה are cut down; כל all עם people כנען the merchant נכרתו are cut off. כל all נטילי they that bear כסף׃ silver}%
\verse{והיה And it shall come to pass בעת time, ההיא at that אחפשׂ I will search את ירושׁלם Jerusalem בנרות with candles, ופקדתי and punish על and punish האנשׁים the men הקפאים that are settled על on שׁמריהם their lees: האמרים that say בלבבם in their heart, לא will not ייטיב do good, יהוה The LORD ולא neither ירע׃ will he do evil.}%
\verse{והיה shall become חילם Therefore their goods למשׁסה a booty, ובתיהם and their houses לשׁממה a desolation: ובנו they shall also build בתים houses, ולא but not ישׁבו inhabit ונטעו and they shall plant כרמים vineyards, ולא but not ישׁתו drink את יינם׃ the wine}%
\verse{קרוב near, יום day יהוה of the LORD הגדול The great קרוב near, ומהר and hasteth מאד greatly, קול the voice יום of the day יהוה of the LORD: מר bitterly. צרח shall cry שׁם there גבור׃ the mighty man}%
\verse{יום day עברה of wrath, היום a day ההוא That יום a day צרה of trouble ומצוקה and distress, יום a day שׁאה of wasteness ומשׁואה and desolation, יום a day חשׁך of darkness ואפלה and gloominess, יום a day ענן of clouds וערפל׃ and thick darkness,}%
\verse{יום A day שׁופר of the trumpet ותרועה and alarm על against הערים cities, הבצרות the fenced ועל and against הפנות towers. הגבהות׃ the high}%
\verse{והצרתי And I will bring distress לאדם upon men, והלכו that they shall walk כעורים like blind men, כי because ליהוה against the LORD: חטאו they have sinned ושׁפך shall be poured out דמם and their blood כעפר as dust, ולחמם and their flesh כגללים׃ as the dung.}%
\verse{גם nor כספם their silver גם זהבם their gold לא יוכל shall be able להצילם to deliver ביום them in the day עברת wrath; יהוה of the LORD's ובאשׁ by the fire קנאתו of his jealousy: תאכל shall be devoured כל but the whole הארץ land כי for כלה riddance אך even נבהלה a speedy יעשׂה he shall make את כל all ישׁבי them that dwell הארץ׃ in the land.}%
\end{biblechapter}%
\begin{biblechapter}% Zephaniah 2
\verseWithHeading{Judgment of Judah’s Enemies}{התקושׁשׁו Gather yourselves together, וקושׁו yea, gather together, הגוי O nation לא not נכסף׃ desired;}%
\verse{בטרם Before לדת bring forth, חק the decree כמץ as the chaff, עבר pass יום the day בטרם before לא יבוא come עליכם upon חרון the fierce אף anger יהוה of the LORD בטרם you, before לא יבוא come עליכם upon יום the day אף anger יהוה׃ of the LORD's}%
\verse{בקשׁו Seek את יהוה ye the LORD, כל all ענוי ye meek הארץ of the earth, אשׁר which משׁפטו his judgment; פעלו have wrought בקשׁו seek צדק righteousness, בקשׁו seek ענוה meekness: אולי it may be תסתרו ye shall be hid ביום in the day אף anger. יהוה׃ of the LORD's}%
\verse{כי For עזה Gaza עזובה forsaken, תהיה shall be ואשׁקלון and Ashkelon לשׁממה a desolation: אשׁדוד Ashdod בצהרים at the noon day, יגרשׁוה they shall drive out ועקרון and Ekron תעקר׃ shall be rooted up.}%
\verse{הוי Woe ישׁבי unto the inhabitants חבל coast, הים of the sea גוי the nation כרתים of the Cherethites! דבר the word יהוה of the LORD עליכם against כנען you; O Canaan, ארץ the land פלשׁתים of the Philistines, והאבדתיך I will even destroy מאין thee, that there shall be no יושׁב׃ inhabitant.}%
\verse{והיתה shall be חבל coast הים And the sea נות dwellings כרת cottages רעים for shepherds, וגדרות and folds צאן׃ for flocks.}%
\verse{והיה shall be חבל And the coast לשׁארית for the remnant בית of the house יהודה of Judah; עליהם thereupon: ירעון they shall feed בבתי in the houses אשׁקלון of Ashkelon בערב in the evening: ירבצון shall they lie down כי for יפקדם shall visit יהוה the LORD אלהיהם their God ושׁב them, and turn away שׁבותם׃ their captivity.}%
\verse{שׁמעתי I have heard חרפת the reproach מואב of Moab, וגדופי and the revilings בני of the children עמון of Ammon, אשׁר whereby חרפו they have reproached את עמי my people, ויגדילו and magnified על against גבולם׃ their border.}%
\verse{לכן Therefore חי live, אני I נאם saith יהוה the LORD צבאות of hosts, אלהי the God ישׂראל of Israel, כי Surely מואב Moab כסדם as Sodom, תהיה shall be ובני and the children עמון of Ammon כעמרה as Gomorrah, ממשׁק the breeding חרול of nettles, ומכרה מלח and saltpits, ושׁממה desolation: עד and a perpetual עולם and a perpetual שׁארית the residue עמי of my people יבזום shall spoil ויתר them, and the remnant גוי of my people ינחלום׃ shall possess}%
\verse{זאת This להם תחת shall they have for גאונם their pride, כי because חרפו they have reproached ויגדלו and magnified על against עם the people יהוה צבאות׃ of hosts.}%
\verse{נורא terrible יהוה The LORD עליהם unto כי them: for רזה he will famish את כל all אלהי the gods הארץ of the earth; וישׁתחוו and shall worship לו אישׁ him, every one ממקומו from his place, כל all איי the isles הגוים׃ of the heathen.}%
\verse{גם also, אתם Ye כושׁים Ethiopians חללי slain חרבי by my sword. המה׃ ye}%
\verse{ויט And he will stretch out ידו his hand על against צפון the north, ויאבד and destroy את אשׁור Assyria; וישׂם and will make את נינוה Nineveh לשׁממה a desolation, ציה dry כמדבר׃ like a wilderness.}%
\verse{ורבצו shall lie down בתוכה in the midst עדרים And flocks כל of her, all חיתו the beasts גוי of the nations: גם both קאת the cormorant גם and קפד the bittern בכפתריה in the upper lintels ילינו shall lodge קול of it; voice ישׁורר shall sing בחלון in the windows; חרב desolation בסף in the thresholds: כי for ארזה the cedar work. ערה׃ he shall uncover}%
\verse{זאת This העיר city העליזה the rejoicing היושׁבת that dwelt לבטח carelessly, האמרה that said בלבבה in her heart, אני I ואפסי and none עוד beside איך me: how היתה is she become לשׁמה a desolation, מרבץ to lie down לחיה a place for beasts כל in! every one עובר that passeth עליה by ישׁרק her shall hiss, יניע wag ידו׃ his hand.}%
\end{biblechapter}%
\begin{biblechapter}% Zephaniah 3
\verseWithHeading{Judgment of Jerusalem and the Nations}{הוי Woe מראה ונגאלה and polluted, העיר city! היונה׃ to the oppressing}%
\verse{לא not שׁמעה She obeyed בקול the voice; לא not לקחה she received מוסר correction; ביהוה in the LORD; לא not בטחה she trusted אל to אלהיה her God. לא she drew not near קרבה׃ she drew not near}%
\verse{שׂריה Her princes בקרבה within אריות lions; שׁאגים her roaring שׁפטיה her judges זאבי wolves; ערב evening לא they gnaw not the bones גרמו they gnaw not the bones לבקר׃ till the morrow.}%
\verse{נביאיה Her prophets פחזים light אנשׁי בגדות treacherous כהניה her priests חללו have polluted קדשׁ the sanctuary, חמסו they have done violence תורה׃ to the law.}%
\verse{יהוה LORD צדיק The just בקרבה in the midst לא thereof; he will not יעשׂה do עולה iniquity: בבקר every morning בבקר every morning משׁפטו his judgment יתן doth he bring לאור to light, לא not; נעדר he faileth ולא no יודע knoweth עול but the unjust בשׁת׃ shame.}%
\verse{הכרתי I have cut off גוים the nations: נשׁמו are desolate; פנותם their towers החרבתי חוצותםבלי that none עובר passeth by: נצדו are destroyed, עריהם their cities מבלי so that there is no אישׁ man, מאין that there is none יושׁב׃ inhabitant.}%
\verse{אמרתי I said, אך Surely תיראי thou wilt fear אותי תקחי me, thou wilt receive מוסר instruction; ולא should not יכרת be cut off, מעונה כל howsoever אשׁר howsoever פקדתי עליהכן them: but השׁכימו they rose early, השׁחיתו corrupted כל all עלילותם׃ their doings.}%
\verse{לכן Therefore חכו wait לי נאם ye upon me, saith יהוה the LORD, ליום until the day קומי that I rise up לעד to the prey: כי for משׁפטי my determination לאסף to gather גוים the nations, לקבצי that I may assemble ממלכות the kingdoms, לשׁפך to pour עליהם upon זעמי them mine indignation, כל all חרון my fierce אפי anger: כי for באשׁ with the fire קנאתי of my jealousy. תאכל shall be devoured כל all הארץ׃ the earth}%
\verse{כי For אז then אהפך will I turn אל to עמים the people שׂפה language, ברורה a pure לקרא call כלם that they may all בשׁם upon the name יהוה of the LORD, לעבדו to serve שׁכם consent. אחד׃ him with one}%
\verse{מעבר לנהרי the rivers כושׁ of Ethiopia עתרי my suppliants, בת the daughter פוצי of my dispersed, יובלון shall bring מנחתי׃ mine offering.}%
\verse{ביום day ההוא In that לא shalt thou not תבושׁי be ashamed מכל for all עלילתיך thy doings, אשׁר wherein פשׁעת thou hast transgressed בי כי against me: for אז then אסיר I will take away מקרבך out of the midst עליזי of thee them that rejoice גאותך in thy pride, ולא and thou shalt no תוספי more לגבהה be haughty עוד more בהר mountain. קדשׁי׃ because of my holy}%
\verse{והשׁארתי I will also leave בקרבך in the midst עם people, עני of thee an afflicted ודל and poor וחסו and they shall trust בשׁם in the name יהוה׃ of the LORD.}%
\verse{שׁארית The remnant ישׂראל of Israel לא shall not יעשׂו do עולה iniquity, ולא nor ידברו speak כזב lies; ולא neither ימצא be found בפיהם in their mouth: לשׁון tongue תרמית shall a deceitful כי for המה they ירעו shall feed ורבצו and lie down, ואין and none מחריד׃ shall make afraid.}%
\verse{רני Sing, בת O daughter ציון of Zion; הריעו shout, ישׂראל O Israel; שׂמחי be glad ועלזי and rejoice בכל with all לב the heart, בת O daughter ירושׁלם׃ of Jerusalem.}%
\verse{הסיר hath taken away יהוה The LORD משׁפטיך thy judgments, פנה he hath cast out איבך thine enemy: מלך the king ישׂראל of Israel, יהוה the LORD, בקרבך in the midst לא of thee: thou shalt not תיראי רע evil עוד׃ any more.}%
\verse{ביום day ההוא In that יאמר it shall be said לירושׁלם to Jerusalem, אל thou not: תיראי Fear ציון Zion, אל Let not ירפו be slack. ידיך׃ thine hands}%
\verse{יהוה The LORD אלהיך thy God בקרבך in the midst גבור of thee mighty; יושׁיע he will save, ישׂישׂ he will rejoice עליך over בשׂמחה thee with joy; יחרישׁ he will rest באהבתו in his love, יגיל he will joy עליך over ברנה׃ thee with singing.}%
\verse{נוגי sorrowful ממועד for the solemn assembly, אספתי I will gather ממך for the solemn assembly, היו are משׂאת it a burden. עליה of חרפה׃ thee, the reproach}%
\verse{הנני עשׂה I will undo את כל all מעניך that afflict בעת time ההיא at that והושׁעתי thee: and I will save את הצלעה her that halteth, והנדחה her that was driven out; אקבץ and gather ושׂמתים and I will get לתהלה them praise ולשׁם and fame בכל in every הארץ land בשׁתם׃ where they have been put to shame.}%
\verse{בעת time ההיא At that אביא will I bring אתכם ובעת you even in the time קבצי that I gather אתכם כי you: for אתן I will make אתכם לשׁם you a name ולתהלה and a praise בכל among all עמי people הארץ of the earth, בשׁובי when I turn back את שׁבותיכם your captivity לעיניכם before your eyes, אמר saith יהוה׃ the LORD.}%
\end{biblechapter}%
