\biblebook{Jonah}
\begin{biblechapter}% Jonah 1
\verseWithHeading{Jonah Disobeys Adonai}{And the word of Adonai came to Jonah the son of Amittai, saying,}%
\verse{“Get up! Go to the great city Nineveh and cry out against her, because their evil has come up before me.”\lebnote{“to my face”}}%
\verse{But Jonah set out to flee toward Tarshish from the presence of\lnCSZ{} Adonai. And he went down to Joppa and found a merchant ship going to Tarshish, and paid her fare, and went on board her to go with them toward Tarshish from the presence of\lnCSZ{} Adonai.}%
\verse{And Adonai hurled a great wind upon the sea, and it was a great storm on the sea, and the merchant ship was in danger of breaking up.\lebnote{“threatened to be broken up”}}%
\verse{And the mariners were afraid, and each cried out to his god. And they threw the contents\lebnote{“objects”} that were in the merchant ship into the sea to lighten it for them. And meanwhile Jonah went down into the hold of the vessel and lay down and fell asleep.}%
\verse{And the captain\lebnote{Or “commander”} of the ship approached him and said to him, “Why are you sound asleep?\lebnote{“in a deep sleep”} Get up! Call on your god! Perhaps your god\lebnote{“the god,” with the article indicating previous reference} will take notice of us and we won’t perish!”}%
\verse{And they said to one another,\lebnote{“each to his friend/companion”} “Come, let us cast lots so that we may know on whose account this disaster has come on us!” And they cast lots, and the lot fell on Jonah.}%
\verse{So they said to him, “Please tell us whoever is responsible that this disaster has come upon us! What is your occupation? And from where do you come? What is your country? And from which people are you?”}%
\verse{And he said to them, “I am a Hebrew, and I fear Adonai, the God of heaven, who made the sea and the dry land.”}%
\verse{Then the men were greatly afraid,\lebnote{“were afraid with a great fear”} and they said to him, “What is this you have done?” because they\lebnote{“the men” but this is redundant in English} knew that he was fleeing from the presence of\lebnote{“from before the face of”} Adonai (because he had told them).}%
\verse{So they said to him, “What shall we do to you so that the sea may quiet down for us?” because the sea was growing more and more tempestuous.\lnCTA{}}%
\verse{And he said to them, “Pick me up and hurl me into the sea so that the sea may quiet down for you, because I know that on account of me this great storm has come upon you all.”}%
\verse{But the men rowed hard to bring the ship\lebnote{the direct object is supplied from context} back to the dry land, and they could not do so because the sea was growing more and more tempestuous\lnCTA{} against them.}%
\verse{So they cried out to Adonai, and they said, “O Adonai! Please do not let us perish because of this man’s life, and do not make us guilty of innocent blood,\lebnote{“do not give innocent blood on us”} because you, O Adonai, did what you wanted.”}%
\verse{And they picked Jonah up and hurled him into the sea, and the sea ceased from its raging.}%
\verse{So the men feared Adonai greatly,\lebnote{“with a great fear”} and they offered a sacrifice to Adonai and made\lebnote{“vowed”} vows.}%
\verseWithHeading{Jonah Is Swallowed by a Fish and Prays to Adonai}{\lebnote{Jonah 1:17–2:10 in the English Bible is 2:1–11 in the Hebrew Bible} And Adonai provided a large fish to swallow up Jonah, and Jonah was in the belly of the fish three days and three nights.}%
\end{biblechapter}%
\begin{biblechapter}% Jonah 2
\verse{And Jonah prayed to Adonai his God from the belly of the fish}%
\verse{and said, “I called from my distress\lebnote{“from distress for me”} to Adonai, and he answered me; from the belly of Sheol\lebnote{“Sheol” is a Hebrew term for the place where the dead reside (i.e. the Underworld)} I cried for help — you heard my voice.}%
\verse{And you threw me into the deep, into the heart of the seas, and the sea currents surrounded me; all your breakers and your surging waves passed over me.}%
\verse{And I said, ‘I am banished\lebnote{Or “expelled,” or “driven away”} from your sight; how\lebnote{Or “however, I will continue to look on your holy temple”} will I continue to look\lebnote{“will I do again to look,” meaning “will I continue to look”} on your holy temple?’\lebnote{“on the temple of your holiness”}}%
\verse{The waters encompassed me up to my neck; the deep surrounded me; seaweed was wrapped around my head.}%
\verse{I went down to the foundations of the mountains; the Underworld — its bars were around me forever. But you brought up my life from the pit, Adonai my God.}%
\verse{When my life was ebbing away from me, I remembered Adonai, and my prayer came to you, to your holy temple.\lebnote{“to the temple of your holiness”}}%
\verse{Those who worship vain idols forsake their loyal love.\lebnote{This could mean (1) they forsake the loyal love they should show to God or (2) they forfeit \textit{forsake} the loyal love that God would have shown to them}}%
\verse{But I, with a voice of thanksgiving, will sacrifice to you; I will fulfill what I have vowed. Deliverance\lebnote{Or “salvation”} belongs to Adonai!”}%
\verse{And Adonai spoke to the fish, and it vomited Jonah out on the dry land.}%
\end{biblechapter}%
\begin{biblechapter}% Jonah 3
\verseWithHeading{The People of Nineveh Repent at Jonah’s Proclamation}{And the word of Adonai came to Jonah a second time, saying,}%
\verse{“Get up! Go to Nineveh, the great city, and proclaim to it the message that I am telling you.”}%
\verse{So Jonah got up\lebnote{Or “set out”} and went to Nineveh according to the word of Adonai. Now Nineveh was an extraordinarily great city\lebnote{“a great city to God” or “a great city to \textit{the} gods,” a disputed phrase which may refer to God’s estimate or ownership of Nineveh, to the presence of many idols, or an idiom referring to the size of Nineveh (this translation takes the last view)} — a journey of three days across.\lebnote{This phrase may also refer to a journey on which business was done, so that “three days” is the total elapsed time}}%
\verse{And Jonah began to go into the city a journey of one day, and he cried out and said, “Forty more days and Nineveh will be demolished!”\lebnote{Or “overthrown”}}%
\verse{And the people of Nineveh believed in God, and they proclaimed a fast and put on sackcloth — from the greatest of them to the least important.\lebnote{“to the smallest of them”}}%
\verseWithHeading{The King’s Proclamation}{And the news reached the king of Nineveh, and he rose from his throne and removed his royal robe, put on sackcloth, and sat in the ashes.}%
\verse{And he had a proclamation made, and said, “In Nineveh, by a decree of the king and his nobles: “No human being or animal, no herd or flock, shall taste anything! They must not eat, and they must not drink water!}%
\verse{And the human beings and the animals must be covered with sackcloth! And they must call forcefully to God, and each must turn from his evil way and from the violence that is in his\lebnote{Hebrew “their”} hands.}%
\verse{Who knows? God may relent and change his mind and turn from his blazing anger\lebnote{“from the heat of his anger”} so that\lebnote{Hebrew “and”} we will not perish.”}%
\verse{And God saw their deeds — that they turned from their evil ways — and God changed his mind about the evil that he had said he would bring upon them, and he did not do it.\lebnote{supplied from English context}}%
\end{biblechapter}%
\begin{biblechapter}% Jonah 4
\verseWithHeading{Jonah Is Angry at Adonai’s Compassion}{And this\lebnote{Hebrew “it”} was greatly displeasing\lebnote{“was displeasing \textit{with} great displeasure”} to Jonah, and he became furious.\lebnote{“it was hot for him”}}%
\verse{And he prayed to Adonai and said, “O Adonai, was this not what I said\lebnote{“my word”} while I was in my homeland? Therefore I originally fled\lebnote{“I did the first time to flee”} to Tarshish, because I knew that you are a gracious and compassionate God, slow to anger and having great steadfast love,\lebnote{“and great of steadfast love”} and one who relents concerning calamity.\lebnote{That is, calamity sent by God as judgment}}%
\verse{And so then, Adonai, please take my life from me, because for me death is better than life!”}%
\verse{And Adonai said, “Is it right for you to be angry?”\lebnote{“Rightfully is it hot for you”; some take this to mean “Are you so very angry?” (\textit{Targum} Jonah 4:4; see also JPS, NEB, NET)}}%
\verse{And Jonah went out from the city and sat down east of the city, and he made for himself a shelter there. And he sat under it in the shade, waiting to see\lebnote{“until he would see”} what would happen with the city.}%
\verse{And Adonai God appointed a plant,\lebnote{Probably a castor oil plant, though some have suggested some type of gourd plant} and he made it grow up over Jonah to be a shade over his head, to save him from his discomfort. And Jonah was very glad\lebnote{“was glad with great joy”} about the plant.}%
\verse{So God appointed a worm at daybreak\lebnote{“at the coming up of the dawn”} the next day, and it attacked the plant, and it withered.}%
\verse{And when the sun rose,\lebnote{“and it happened at the rising of the sun”} God appointed a scorching east wind, and the sun beat down on Jonah’s head and he grew faint. And he asked that he could die\lebnote{“and he asked his soul to die”} and said, “My death is better than my life!”}%
\verse{So God said to Jonah, “Is it right for you to be angry\lebnote{“rightfully is it hot for you”; some take this to mean “Are you so very angry?” (\textit{Targum} Jonah 4:4; see also JPS, NEB, NET)} about the plant?” And he said, “It is right for me to be angry enough to die!”\lebnote{“Rightfully it is hot for me until death”}}%
\verse{But Adonai said, “You are troubled about the plant, for which you did not labor nor cause it to grow. It grew up in a night and it perished in a night!\lebnote{“Which was a son of a night and \textit{as} a son of a night it perished”}}%
\verse{And should I not be concerned about Nineveh, the great city, in which there are\lebnote{“which there are in it”} more than one hundred and twenty thousand\lebnote{“than two ten myriad” (12 x 10,000)} people who do not know right from left,\lebnote{“\textit{hand}” is often supplied, but it is not clear just what deficiency is meant by this expression, which occurs only here in biblical Hebrew} plus many animals?”}%
\end{biblechapter}%
