\biblebook{Proverbs}
\begin{biblechapter}% Proverbs 1
\verseWithHeading{Prologue}{Proverbs of Solomon, son of David, king of Israel:}%
\verse{To know wisdom and instruction, to understand sayings of understanding,}%
\verse{to gain insightful instruction,\lebnote{“the instruction of insight”} righteousness and justice and equity,}%
\verse{to give shrewdness\lebnote{Or “cleverness,” or “prudence”} to the simple, knowledge and purpose\lebnote{Or “plan”} to the young,}%
\verse{may the wise hear and increase learning, and the one who understands gain direction,\lebnote{Or “guidance”}}%
\verse{to understand a proverb and an expression, words of wisdom and their riddles.}%
\verse{Fear of Adonai is the beginning of knowledge; wisdom and instruction, fools despise.}%
\verseWithHeading{Lecture Against Gang Behavior}{My child, may you keep\lebnote{Or “obey,” or “hear”} your father’s instruction, and do not reject your mother’s teachings,}%
\verse{for they are a garland of favor for your head, and pendants for your neck.}%
\verse{My child, if sinners entice you, do not consent.}%
\verse{If they say, “Come with us! We shall lie in wait for blood; we shall ambush the innocent without cause.\lebnote{Or “wantonly,” or “for nothing”}}%
\verse{Like Sheol,\lebnote{A term for the place where the dead reside, i.e., the Underworld} we will swallow them alive and whole, like those who descend to the pit.\lebnote{“like the descenders of a pit”}}%
\verse{We shall find all precious wealth, we shall fill our houses with booty,}%
\verse{you shall throw your lot in our midst, there will be one purse for all of us.”}%
\verse{My child, do not walk in their way.\lebnote{Or “on their road”} Keep your foot from their paths,}%
\verse{for their feet run to evil, and they hurry to shed blood,}%
\verse{for “in vain\lebnote{Or “without cause,” or “for nothing”} is the net scattered, in the sight of any winged bird.”\lebnote{“bird of wing”}}%
\verse{They lie in wait for their own blood. They ambush their own lives.}%
\verse{Thus are the ways of all who are greedy for gain — it will take the life of its possessors.}%
\verseWithHeading{The Call of Lady Wisdom}{Wisdom calls out in the streets, in the squares she raises her voice.}%
\verse{On a busy corner\lebnote{“head of commotion,” or “head of stirring, bustling”} she cries out, at the entrances of the gates in the city, she speaks her sayings:}%
\verse{“How long,\lebnote{“Until when”} O simple ones, will you love simplicity? And how long will scoffers delight in their scoffing, and fools hate knowledge?}%
\verse{May you turn to my argument!\lebnote{Or “turn at my reproach”} Behold, I shall pour out my spirit upon you; I will make my words known to you.}%
\verse{Because I called out and you refused me, I stretched out my hand, yet there is none who heeds.}%
\verse{You have ignored all my counsel, and my reproof you are not willing to accept.}%
\verse{I will also laugh at your calamity;\lebnote{Or “downfall”} I will mock when panic\lebnote{Or “dread, fear”} comes upon you.}%
\verse{When your panic comes like a storm, and your calamity arrives like a whirlwind, when distress and anguish come upon you,}%
\verse{then they will call\lebnote{Or “cry out \textit{to}”} me, but I will not answer; they will seek me diligently but not find me.}%
\verse{Since they hated knowledge, and did not choose the fear of Adonai,}%
\verse{they were not willing to accept my counsel, they despised all my reproof.}%
\verse{They shall eat from the fruit of their way, and they shall be sated from their own schemes,}%
\verse{for the waywardness of the simple ones will kill them, and the complacency of fools will destroy them.}%
\verse{Whoever listens to me will dwell in security and rest securely\lebnote{Or “be at ease, peace”} from dread and disaster.”}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 2
\verseWithHeading{The Benefits of Wisdom}{My child, if you will receive\lebnote{Or “take, seize”} my sayings, and hide my commands with you,}%
\verse{in order to incline your ear toward wisdom, then you shall apply your heart to understanding.}%
\verse{For if you cry out for understanding,\lebnote{Or “call to understanding”} if you lift\lebnote{Or “carry, give”} your voice for insight,}%
\verse{if you seek her like silver and search her out like treasure,\lebnote{“like the treasures”}}%
\verse{then you will understand the fear of Adonai, and the knowledge of God you will find.}%
\verse{For Adonai will give wisdom; from his mouth come knowledge and understanding.}%
\verse{For the upright, he stores\lebnote{Or “hides, keeps”} sound judgment, a shield for those who walk uprightly,}%
\verse{in order to guard paths of justice and keep\lebnote{Or “preserve, guard, watch”} the way of his faithful ones.}%
\verse{Then you will understand righteousness and justice and uprightness — every good course —}%
\verse{for wisdom will enter your heart, and knowledge will be pleasing to your soul.\lebnote{Or “inner self”}}%
\verse{Discretion will watch over you; understanding will protect you,}%
\verse{in order to deliver you from the way of evil,\lebnote{Or “evildoer”} from a man who speaks devious things —}%
\verse{those who forsake the paths of uprightness to walk in ways of darkness,}%
\verse{those who are happy to do evil, for they delight in the deviousness of evil,}%
\verse{who are crooked in their ways,\lebnote{“that \textit{in} their ways are crooked”} and devious in their paths;}%
\verse{in order to deliver you from a strange woman, from a foreign woman who flatters\lebnote{Or “makes smooth”} with her sayings,}%
\verse{she who forsakes the partner of her youth and has forgotten the covenant of her God,}%
\verse{for her house sinks to death, and to the dead\lebnote{Or “Rephaim”} are her paths.}%
\verse{Of all who go to her, none shall return, nor do they reach paths of life.}%
\verse{So that you will walk on the road of those who are good, and the paths of those who are righteous you shall keep.}%
\verse{For those who are upright will dwell in the land,\lebnote{Or “earth”} and those who are blameless\lebnote{Or “innocent”} will remain in it.}%
\verse{And those who are wicked will be cut off from the land, and those who are treacherous will be uprooted from it.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 3
\verseWithHeading{Wisdom and Piety}{My child, do not forget my instruction, and may your heart guard my commands.}%
\verse{For length of days, years of life, and peace they\lebnote{That is, “my commands” (3:1)} shall add to you.}%
\verse{May loyal love\lebnote{Or “kindness, mercy”} and truth not forsake you; bind them around your neck, write them upon your heart.}%
\verse{And you shall find favor and good sense\lebnote{Or “understanding, prudence”} in the eyes of God and humankind.}%
\verse{Trust Adonai\lebnote{“trust toward Adonai”} with all your heart; do not lean toward your own understanding.}%
\verse{In all your ways acknowledge him, and he will straighten your paths.}%
\verse{Do not be wise in your own eyes; fear Adonai and retreat from evil.}%
\verse{There shall be healing for your flesh,\lebnote{“navel”} and refreshment for your body.}%
\verse{Honor Adonai from your substance,\lebnote{Or “wealth, property”} and from the firstfruits of all that will come to you,}%
\verse{and your barns shall be full of plenty, and your vats shall burst with new wine.}%
\verseWithHeading{Happy Is the One Who Finds Wisdom}{Do not despise the discipline of Adonai, my child. Do not be weary of his reproof}%
\verse{because whomever Adonai will love, he will rebuke, as a father delights in his son.}%
\verse{Happy is the one\lebnote{“a man”} who finds wisdom, and one who obtains understanding.}%
\verse{For her\lebnote{Or “its”; both “wisdom” and “understanding” are feminine nouns and can be read as either “she” or “it”} income is better than the income of silver, and her revenue than that of gold.}%
\verse{She is more precious than rubies, and all your desires shall not compare with her.}%
\verse{Length of days is in her right hand; in her left hand are riches and honor.}%
\verse{Her ways are ways of pleasantness, and all her paths are peace.}%
\verse{She is a tree of life for those who seize her; those who take hold of her are considered happy.}%
\verseWithHeading{Role of Wisdom in Creation and Society}{Adonai in wisdom founded the earth; he established the heavens in understanding.}%
\verse{With his knowledge, depths broke open, and clouds dropped dew.}%
\verse{My child, may they\lebnote{Grammatically, “they” most likely refers to clouds (3:20). In context, “they” may refer to wisdom and understanding (3:13).} not escape from your sight;\lebnote{“eyes”} may you keep sound wisdom and prudence.}%
\verse{They shall be life for your soul\lebnote{Or “inner self”} and adornment\lebnote{Or “favor”} for your neck.}%
\verse{Then you will walk in the confidence of your ways, and your foot will not stumble.}%
\verse{If you sit down, you will not panic,\lnBTH{} and if you lie down, then your sleep shall be sweet.}%
\verse{Do not be afraid of sudden panic,\lnBTH{} or the storm of wickedness that will come.}%
\verse{Adonai will be your confidence\lebnote{“in your confidence”} and guard your foot from capture.}%
\verse{Do not withhold good from its owner\lebnote{Or “lord”} when it is in the power of your hand to do.}%
\verse{Do not say to your neighbor, “Go and return and tomorrow I will give it,” when it is with you.\lebnote{“and there is with you”}}%
\verse{Do not plot harm against your neighbor who dwells in confidence beside you.}%
\verse{Do not quarrel with anyone without cause, when he did not do you harm.}%
\verse{Do not envy a man of violence, and do not choose any of his ways,}%
\verse{for he who is perverse is an abomination of Adonai, but those who are upright are his confidence.}%
\verse{The curse of Adonai is on the house of the wicked, and the abode of the righteous ones he blesses.}%
\verse{With those who scorn, he is scornful, but to those who are humble, he gives favor.}%
\verse{They will inherit the honor of the wise, but stubborn fools, disgrace.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 4
\verseWithHeading{The Father’s Wisdom}{Children, listen to the instruction of a father, and be attentive in order to know insight.}%
\verse{For I have given you good instruction; do not forsake my teaching.}%
\verse{When I was a son to my father, tender and alone before my mother,}%
\verse{he taught me and said to me, “May your heart hold fast to my words; guard my commandments and live.}%
\verse{Get wisdom and insight; do not forget and do not turn from the sayings of my mouth.}%
\verse{Do not forsake her — then she will guard you; love her and she will keep you.}%
\verse{The beginning of wisdom: Get wisdom! With all that is in your possession, gain insight.}%
\verse{Cherish her and she will exalt you; she will honor you if you embrace her.}%
\verse{She will give a garland for your head; she shall bestow a crown of glory upon you.”}%
\verseWithHeading{The Right Path}{Listen, my child, take my sayings, and the years of your life shall be great.}%
\verse{In the way of wisdom I have instructed you; I have led you in the path of uprightness.}%
\verse{When you walk,\lebnote{“In your walking”} your step will not be hampered, and if you run, you will not stumble.}%
\verse{Seize the instruction! Do not let go! Guard her, for she is your life.}%
\verse{In the path of the wicked do not go; do not walk in the way of those who do evil.}%
\verse{Avoid it; do not transgress it; turn away from it and pass by.}%
\verse{For they will not sleep if they have not done wrong, and they are robbed of their sleep if they do not cause stumbling.}%
\verse{For they ate the bread of wickedness, and they drank the wine of violence.}%
\verse{But the path of the righteous ones is like the light of dawn, leading and shining until the day is full.\lebnote{“\textit{is} correct”}}%
\verse{The way of the wicked ones is like deep darkness; they do not know what they stumble over.}%
\verseWithHeading{Staying the Course}{My child, be attentive to my words; to my sayings incline your ear.}%
\verse{May they not escape from your sight;\lebnote{“eyes”} keep them in the midst of your heart.}%
\verse{For they are life to those who find them and healing to the entire body.\lebnote{“for all his flesh”}}%
\verse{With all vigilance, keep your heart, for from it comes the source\lebnote{Or “extremity”} of life.}%
\verse{Remove from yourself deceitful speech,\lebnote{“crookedness of mouth”} and abolish devious talk\lebnote{“deviousness of lips”} from yourself.}%
\verse{May your eyes look forward\lebnote{Or “opposite”} and your gaze be straight before you.}%
\verse{May the path of your foot be balanced and all your ways be sure.}%
\verse{Do not swerve right or left; remove your foot from evil.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 5
\verse{My child, be attentive to my wisdom, and to my understanding incline your ear;}%
\verse{in order to keep prudence, and knowledge will guard your lips.}%
\verse{For the lips of the strange woman will drip honey, and smoother than oil is her mouth.\lebnote{Or “palate”}}%
\verse{But her end is bitter as the wormwood, sharp as a two-edged sword.}%
\verse{Her feet go down to death; her steps take hold of Sheol.\lebnote{A term for the place where the dead reside, i.e., the Underworld}}%
\verse{She does not observe\lebnote{Or “examine, weigh”} the path of life; her ways wander, and she does not know it.}%
\verseWithHeading{Do Not Commit Adultery Against Wisdom}{Now, O children, listen to me; do not depart from the sayings of my mouth.}%
\verse{Keep your paths far from her, and do not go near to the door of her house,}%
\verse{lest you give your honor to the others, and your years to the merciless,}%
\verse{lest strangers take their fill of your strength, and your labors go to the house of a foreigner,}%
\verse{and you groan at your end, when your flesh and body are consumed,}%
\verse{and say “How I hated discipline, and I despised reproof!”}%
\verse{and “I did not listen to the voice of my teachers, and I did not incline my ear to my instructors!}%
\verse{I was almost at utter\lebnote{Or “all, every, whole”} ruin\lebnote{Or “evil”} in the midst of the assembly and congregation.”}%
\verse{Drink water from your own cistern and flowing waters from inside your own well.}%
\verse{Shall your springs be scattered outward? In the streets, shall there be streams of water?}%
\verse{May they be yours alone, and not for strangers who are with you.}%
\verse{May your fountain be blessed, and rejoice in the wife\lebnote{Or “woman”} of your youth.}%
\verse{She is a deer of love and a doe of grace; may her breasts satisfy you always;\lebnote{“at all time”} by her love may you be intoxicated continually.}%
\verse{Why should you be intoxicated, my child, by a strange woman, and embrace the bosom of a foreigner?}%
\verse{For before the eyes of Adonai are human ways,\lebnote{“the ways of man/humankind”} and all his paths he examines.}%
\verse{His iniquities shall ensnare him, the evildoer, and in the vanity of his sin he shall be caught.}%
\verse{He shall die for lack of\lebnote{“with there is no”} discipline, and in the greatness of his folly he shall be lost.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 6
\verseWithHeading{Against Pledges}{My child, if you have pledged to your neighbor, if you have bound yourself\lebnote{“palms of your hands”} to the stranger,}%
\verse{if you are snared by the sayings of your mouth, if you are caught by the sayings of your mouth,}%
\verse{do this, then, my child, and save yourself, for you have come into the palm of your neighbor’s hand:\lebnote{“the palm of the hand of your neighbor”} Go, humble yourself, plead with your neighbor.}%
\verse{Do not give sleep to your eyes, or slumber to your eyelids.}%
\verse{Save yourself like a gazelle from a hand, or like a bird from the hand of a fowler.}%
\verseWithHeading{Against Sloth}{Go to the ant, lazy! Consider its ways and be wise.}%
\verse{It has no chief, officer, or ruler.}%
\verse{In the summer, it prepares its food; in the harvest, it gathers its sustenance.}%
\verse{How long will you lie down, lazy? When will you rise up from your sleep?}%
\verse{A little sleep, a little slumber, a little folding of the hands for rest —}%
\verse{like a robber\lebnote{“one who walks,” that is, a vagabond} shall your poverty come, and what you lack like an armed man.}%
\verseWithHeading{Against Worthlessness}{A worthless man, an evil man, goes around with deceitful speech.\lebnote{“crookedness of mouth”}}%
\verse{Winking in his eye, shuffling in his foot, pointing in his fingers,}%
\verse{perversion in his heart, he devises evil; at all times he will send out discord.\lnBTI{}}%
\verse{Upon such a man,\lebnote{“thus”} suddenly shall his calamity come; in a moment he will be damaged and there is no healing.\lebnote{Or “repair”}}%
\verseWithHeading{What Adonai Hates}{There are six things Adonai hates, and seven things are abominations of his soul:\lebnote{Or “inner self”}}%
\verse{haughty eyes, a lying tongue, and hands that shed innocent blood,}%
\verse{a devising heart, plans of deception, feet that hurry to run to evil,\lebnote{Hebrew “the evil”}}%
\verse{a false witness who breathes lies and sends out discord\lnBTI{} between brothers.}%
\verseWithHeading{Commandment and Instruction as Guardians}{My child, keep the commandment of your father, and do not disregard the instruction of your mother.}%
\verse{Bind them on your heart continually; tie them upon your neck.}%
\verse{When you walk,\lebnote{“In your walking”} she\lebnote{That is, “commandment” and “instruction”} will lead you, When you lie down,\lebnote{“In your lying down”} she will watch over you, and when you awake, she will converse with you.}%
\verse{For like a lamp is a commandment, and instruction is light, and the way of life\lebnote{Hebrew “lives”} is the reproof of discipline,}%
\verse{in order to preserve you from an evil woman,\lebnote{Or “wife”} from the smoothness of the tongue of an adulteress.\lebnote{“a foreign woman”}}%
\verse{Do not desire her beauty in your heart; may she not capture you with her eyelashes.}%
\verse{For the price of a woman, a prostitute,\lebnote{Or “whore”} is the price of a loaf of bread, but the woman belonging to a man\lebnote{“the woman of a man”} hunts precious life.}%
\verseWithHeading{Warning Against Relations with a Married Woman}{Can a man carry fire in his lap, and his clothes not burn?}%
\verse{If a man walks upon the hot coals, will his feet not be burned?}%
\verse{Thus, he who goes to the wife of his neighbor, any who touches her shall not go unpunished.}%
\verse{People do not despise a thief when he steals to fill himself\lebnote{Or “soul,” or “inner self”} when he is hungry.}%
\verse{But if he is found, he will pay sevenfold, every possession of his house he shall give.}%
\verse{He who commits adultery with a woman lacks sense,\lebnote{“heart”} he destroys himself\lebnote{Or “his soul,” or “his life”} who does it.}%
\verse{A wound and dishonor he will find, and his disgrace will not be wiped out.}%
\verse{For jealousy is the fury of a husband, and he will not show restraint on the day of revenge.}%
\verse{He will not accept any compensation,\lebnote{“\textit{the} face of any compensation”} and he will not be willing, though the bribe is large.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 7
\verseWithHeading{Warning Against the Strange Woman}{My child, guard my sayings; store my commandments with you.}%
\verse{Keep my commands and live, and my teaching like the apple of your eye.\lebnote{A single word meaning “pupil of the eye”}}%
\verse{Bind them on your fingers; write them on the tablet of your heart.}%
\verse{Say to wisdom,\lebnote{Hebrew “the wisdom”} “you are my sister,” and you shall call insight,\lebnote{Hebrew “the insight”} “intimate friend.”\lebnote{“one who is known.” To “know” is often a euphemism for intercourse. Therefore “intimate friend” may also be read “lover.”}}%
\verse{In order to guard yourself from an adulteress,\lebnote{“a strange woman”} from the foreigner who makes her words smooth.\lebnote{“causes to be smooth her words”}}%
\verse{For at the window of my house, through my lattice, I looked down.}%
\verse{And I saw among the simple, I observed among the youth, a young man lacking sense,\lebnote{“heart”}}%
\verse{passing on the street at\lnBTJ{} her corner, and he takes the road to her house,}%
\verse{at twilight, at the day’s evening, in the midst of night and the darkness.}%
\verse{Then behold! A woman comes to meet him with the garment of a prostitute\lebnote{Or “whore”} and a secret heart.\lebnote{“secret of heart”}}%
\verse{She is loud and stubborn; her feet do not stay at her house.}%
\verse{Now in the street, now in the square, at\lnBTJ{} every corner she lies in wait.}%
\verse{She took hold of\lebnote{Or “strengthened”} him and kissed him. Her face was impudent, and she said to him,}%
\verse{“Sacrifices of peace offerings are upon me; today\lebnote{“the day”} I completed my vows.}%
\verse{So\lebnote{Hebrew “thus”} I have come out to meet you, to seek your face, and I have found you.}%
\verse{With coverings I have adorned my couch, spreads of the linen of Egypt;}%
\verse{I have perfumed my bed with myrrh, aloes, and cinnamon.}%
\verse{Come, let us take our fill of love making, until the morning let us delight in love.}%
\verse{For there is no man\lebnote{Or “husband”} in his home; he has gone on a long journey.\lebnote{“a journey from far”}}%
\verse{The bag of money he took in his hand, for on the day of the full moon he will come home.”}%
\verse{She persuades him with the greatness of her teachings; with her smooth lips she compels him.}%
\verse{He goes after her suddenly; like an ox to the slaughter he goes, and like a stag to the instruction of a fool,}%
\verse{until an arrow pierces his entrails,\lebnote{“heaviness,” often referring to the liver} like a bird rushing into a snare, but he does not know that it will cost him his life.\lebnote{“it is against his life”}}%
\verseWithHeading{Reiteration of the Warning Against a Strange Woman}{And now, my children, listen to me, and be attentive to the sayings of my mouth.}%
\verse{May your heart not turn aside to her ways; do not stray into her path.}%
\verse{For many slain she has laid low, and countless\lebnote{Or “strong men”} are all of her killings.}%
\verse{The ways of Sheol\lebnote{A term for the place where the dead reside, i.e., the Underworld} are her house, descending to chambers of death.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 8
\verseWithHeading{Wisdom Calls}{Does not wisdom call, and understanding raise its voice?}%
\verse{Atop the heights\lebnote{“At a head of the heights”} beside the road, at the crossroads she stands.}%
\verse{Beside gates, before towns, at the entrance of doors, she cries out:}%
\verse{“To you, O people,\lebnote{“men”} I call, and my cry is to the children of humankind.}%
\verse{Learn prudence, O simple ones; fools, learn intelligence.\lebnote{“heart”}}%
\verse{Listen! For noble things I will speak, and upright things from the opening of my lips.}%
\verse{My mouth will utter truth, and wickedness is an abomination to my lips.}%
\verse{All sayings of my mouth are in righteousness; none of them are twisted and crooked.}%
\verse{All of them are straight to him who understands, and upright to those who find knowledge.}%
\verse{Take my teaching and not silver; may you choose knowledge rather than choice gold.}%
\verse{For wisdom is better than jewels, and all desires shall not compare with her.}%
\verse{I, wisdom, live with prudence, and I find knowledge and discretion.}%
\verse{The fear of Adonai is hatred of evil, pride, and arrogance and an evil way. And I hate a mouth of perversity.}%
\verse{Advice and sound judgment\lebnote{“To/for me are advice and sound judgment”} are mine; I am understanding, strength is mine.\lebnote{“to/for me is strength”}}%
\verse{By me kings reign, and rulers decree righteousness.}%
\verse{By me rulers rule, and nobles — all judges of righteousness.}%
\verse{I love those who love me,\lebnote{“her lovers”} and those who seek me diligently shall find me.}%
\verse{Fortune and glory are with me, enduring wealth and righteousness.}%
\verse{My fruit is better than gold,\lebnote{“fine gold”} even refined gold, and my yield than choice silver.}%
\verse{In the way of righteousness I walk, in the midst of paths of justice,}%
\verse{in order to endow those who love me with wealth, and I will fill their treasuries.}%
\verseWithHeading{Wisdom at Creation}{“Adonai possessed\lebnote{Or “gained,” or “acquired”} me, the first of his ways, before his acts of old.\lebnote{“from before” or “from then”}}%
\verse{From eternity, I was set up from the first, from the beginning\lebnote{Hebrew “beginnings”} of the earth.}%
\verse{When there were no depths, I was brought forth, when there were no springs of abounding\lebnote{“made heavy”} water.}%
\verse{Before mountains had been shaped, before hills, I was brought forth.}%
\verse{When he had not yet made earth and fields, or the first dust of the world,}%
\verse{when he established\lebnote{“in his establishing”} the heavens, there I was, when he drew\lebnote{“in his drawing”} a circle upon the face of the deep,}%
\verse{when he made\lebnote{“in his making”} skies from above, when he founded fountains of the deep,}%
\verse{when he assigned\lebnote{“in his assigning”} his limits to the sea, that waters shall not transgress his command,\lebnote{“mouth”} when he marked\lebnote{“in his marking”} the foundations of the earth,}%
\verse{I was beside\lebnote{“at the place of”} him, a master workman, and I was delighting day by day, rejoicing before him always,\lebnote{“in all \textit{of} time”}}%
\verse{rejoicing in the world of his earth, and my delight was with the children of humankind.}%
\verseWithHeading{Benefits of Following Wisdom}{“And now, children, listen to me; happy are those who will keep my ways.}%
\verse{Hear teaching and be wise; do not neglect it.}%
\verse{Happy is the person who listens to me, in order to keep watch at my doors day by day, in order to guard the frames of my entrances.}%
\verse{For he who finds me is he who finds\lebnote{Hebrew “those who find”} life, and he obtains favor from Adonai.}%
\verse{But he who misses me injures himself.\lebnote{Or “soul,” or “inner self”} All those who hate me love death.”}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 9
\verseWithHeading{Wisdom’s Banquet}{Wisdom has built her house; she has hewn her seven pillars.}%
\verse{She has slaughtered her slaughtering, mixed her wine, and also set her table.}%
\verse{She has sent her servant girls,\lebnote{Or “young maidens”} she calls upon the wings of the high places of town,}%
\verse{“Whoever is simple, let him turn here.” As for the one who lacks sense,\lnBTK{} she says to him,}%
\verse{“Come, eat with my bread; drink with the wine I have mixed.}%
\verse{Lay aside simplicity and live; walk in the way of understanding.”}%
\verse{He who corrects a scoffer gains abuse for himself, and he who rebukes the wicked gets hurt.\lebnote{“abuse him”}}%
\verse{Do not rebuke a scoffer, lest he hate you; rebuke the wise and he will love you.}%
\verse{Give to a wise one and he will become more wise; teach\lebnote{“make known to”} a righteous one and he will increase learning.}%
\verseWithHeading{Foolishness’ Banquet}{The start of wisdom is fear of Adonai, and knowledge of the Holy One,\lebnote{Or “holy ones”} insight.}%
\verse{For by me your days shall increase, and years of life shall multiply for you.}%
\verse{If you are wise, you are wise for yourself, and if you scoff, alone you shall bear it.}%
\verse{A woman of foolishness is loud, simple, and does not know it.\lebnote{“what”}}%
\verse{She sits at the door of her house, upon a throne at the high places of town,}%
\verse{in order to call to those who pass by the road, those who go straight on their way:}%
\verse{“Whoever is simple, may he turn here!” As for he who lacks sense,\lnBTK{} she says to him,}%
\verse{“Stolen waters are sweet, and bread of secrecy is pleasant.”}%
\verse{But he does not know that the dead\lebnote{Or “Rephaim”} are there, in the depths of Sheol\lebnote{A term for the place where the dead reside, i.e., the Underworld} are her guests.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 10
\verseWithHeading{Proverbs of Solomon}{The proverbs of Solomon: A wise child makes a father glad, but a foolish child grieves his mother.}%
\verse{Treasures of wickedness do not profit, but righteousness delivers from death.}%
\verse{Adonai will not cause a righteous person\lebnote{Or “soul,” or “inner self”} to go hungry, but the craving of the wicked he will thwart.}%
\verse{A slack hand causes poverty, but the hand of the diligent enriches.}%
\verse{He who gathers in the summer is a child who is prudent; he who sleeps at the harvest is a child who brings shame.}%
\verse{Blessings belong to the head of the righteous, but the mouth of the wicked conceals violence.}%
\verse{The memory of righteousness is like a blessing, but the name of the wicked will rot.}%
\verse{A heart of wisdom will heed commandments, but a babbling fool will come to ruin.}%
\verse{Whoever walks in integrity\lebnote{Hebrew “in the integrity”} will walk securely, but whoever follows perversity, his ways will be made known.}%
\verse{The winking of an eye causes\lebnote{Or “gives, sets up”} trouble, and the foolishness of lips comes to ruin.}%
\verse{A fountain of life is a mouth of righteousness, and a mouth of wickedness conceals violence.}%
\verse{Hatred stirs up strife, but love covers over all offenses.}%
\verse{On the lips of one who has understanding, wisdom is found, but a rod is for the back of one who lacks sense.\lebnote{“heart”}}%
\verse{Those who are wise lay up knowledge, but to the mouth of the fool, ruin draws near.}%
\verse{The wealth of the rich is the city of his strength; the ruin of the poor is their poverty.}%
\verse{The wage of the righteous leads to life; the gain of the wicked to sin.}%
\verse{On the path to life is he who guards instruction, but he who rejects rebuke goes astray.}%
\verse{He who conceals hatred has lips of deceit, and he who utters slander — he is a fool.}%
\verse{In many words, transgression is not lacking, but he who restrains his lips is prudent.}%
\verse{Choice silver is a tongue of righteousness, a heart\lebnote{Or “mind”} of wickedness is of little worth.\lebnote{“like a little”}}%
\verse{Lips of righteousness feed many, but fools die for lack of sense.\lebnote{“in the lack of heart, they die”}}%
\verse{The blessing of Adonai makes one rich, and he does not increase sorrow with it.}%
\verse{It is like a sport for a fool to do wrong, wisdom for a person of understanding.}%
\verse{The dread of the wicked will come upon him, but the desire of the righteous will be granted.}%
\verse{With the passing of the tempest, there is no wickedness, but the righteous have a foundation forever.}%
\verse{Like vinegar to the tooth and like smoke to the eyes, thus is the lazy to one who employs\lebnote{Or “sends”} him.}%
\verse{The fear of Adonai adds life, but the years of the wicked are shortened.}%
\verse{The hope of the righteous is gladness, but the expectation of the wicked comes to nothing.}%
\verse{A stronghold for the upright is the way of Adonai, but ruin belongs to evildoers.\lebnote{“them who do evil”}}%
\verse{The righteous one is forever; he will not be removed. But the wicked will not remain in the land.\lebnote{Or “earth”}}%
\verse{The mouth of the righteous brings forth wisdom, but a tongue of perversity will be cut off.}%
\verse{The lips of the righteous know the acceptable, but the mouth of the wicked, the perverse.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 11
\verse{Balances of deceit are an abomination of Adonai, but an accurate weight\lebnote{“a full stone” or “a pure stone”} delights him.}%
\verse{Pride comes, then disgrace comes, but wisdom is with the humble.}%
\verse{The integrity of the upright guides them, but the crookedness of the treacherous destroys them.}%
\verse{Wealth does not profit on the day of wrath, but righteousness will deliver from death.}%
\verse{The righteousness of the blameless will keep his ways straight, but the wicked will fall by his wickedness.}%
\verse{The righteousness of the upright will save them, but by a scheme the treacherous will be taken captive.}%
\verse{With the death of a wicked person, hope will die, and the expectation of the godless perishes.}%
\verse{The righteous is delivered from trouble, but the wicked enters into it.}%
\verse{With a mouth, the godless shall destroy his neighbor, but by knowledge the righteous are delivered.}%
\verse{When good is with the righteous, the city rejoices, and with the perishing of the wicked, jubilation.}%
\verse{By the blessing of the upright, a city will be exalted, but by the mouth of the wicked, it will be overthrown.}%
\verse{He who lacks sense\lebnote{“heart”} belittles his neighbor, but a person of intelligence will remain silent.}%
\verse{A gossip walks about telling a secret, but the trustworthy in spirit keeps the matter.}%
\verse{Where there is no guidance, a nation\lebnote{Or “people”} shall fall, but there is safety in an abundance of counsel.}%
\verse{He will suffer trouble when he loans to a stranger, but he who refuses a pledge is safe.}%
\verse{A woman of grace receives honor, but the ruthless gets wealth.}%
\verse{A person of kindness rewards himself,\lebnote{Or “soul,” or “inner self”} but a cruel person harms his own flesh.}%
\verse{The wicked earns\lebnote{Or “does, makes”} deceptive gain,\lebnote{“gain of deception”} but he who sows righteousness, a true reward.\lebnote{“a reward of truth”}}%
\verse{He who is steadfast in righteousness is to life as\lebnote{Hebrew “and”} he who pursues evil is to death.}%
\verse{An abomination of Adonai are the crooked of heart,\lebnote{“crooked ones of heart”; “heart” may also be translated “mind”} but his delight are those with blameless ways.\lebnote{“blameless ones of ways”}}%
\verse{Rest assured,\lebnote{“Hand to hand”} the wicked will not go unpunished, but the offspring\lebnote{Or “seed”} of the righteous will escape.}%
\verse{A ring of gold in the snout of a pig is a woman who is beautiful but without discretion.}%
\verse{The desire of the righteous is only good, but the expectation of the wicked, wrath.}%
\verse{There is one who gives yet grows richer,\lebnote{“adding more”} but he who withholds what is right\lebnote{“from what is right”} only finds need.}%
\verse{A person of blessing will be enriched, and he who gives water also will be refreshed.}%
\verse{He who withholds grain, the people curse him, but a blessing is for the head of him who sells.}%
\verse{He who diligently seeks good seeks favor, but he who inquires of evil, it will come to him.}%
\verse{He who trusts in his wealth is he who will fall, but like a green leaf the righteous will flourish.}%
\verse{He who brings trouble to his household, he will inherit wind,\lebnote{Or “breath, spirit”} and a fool will serve the wise of heart.}%
\verse{The fruit of righteousness is a tree of life, and he who captures souls\lebnote{Or “persons,” or “inner selves”} is wise.}%
\verse{If the righteous on earth will be repaid, how much more\lebnote{“more for”} the wicked and sinner.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 12
\verse{He who loves discipline loves knowledge, but he who hates rebuke is stupid.}%
\verse{The good obtains favor from Adonai, but anyone who schemes, he condemns.}%
\verse{A person will not be established by wickedness, but the root of the righteous will not be moved.}%
\verse{A woman of strength\lebnote{Or “honor”} is the crown of her master,\lebnote{Or “lord, owner”} but like rot in his bones is she who brings shame.}%
\verse{The thoughts of the righteous are\lebnote{Hebrew “is”} just; the advice of the wicked is treacherous.}%
\verse{The words of the wicked are an ambush of blood, but the mouth of the upright delivers them.}%
\verse{The wicked are overthrown and are no more,\lebnote{“there is no them”} but the house of the righteous shall stand.}%
\verse{For his mouth of good sense, a man will be recommended, but he who is of perverse mind\lnBTL{} will be despised.}%
\verse{It is better to be lowly and a servant to someone\lebnote{“for him,” “\textit{belonging} to him”} than self-glorifying and lacking food.}%
\verse{The righteous knows the life\lebnote{Or “soul,” or “inner self”} of his animal, but the compassion\lebnote{Hebrew “compassions”} of the wicked is cruel.}%
\verse{He who works his land will have plenty of food, but he who follows worthless things lacks sense.\lnBTL{}}%
\verse{The wicked covets the proceeds of evil, but the root of the righteous bears fruit.\lebnote{“gives, sets up”}}%
\verse{By the transgression of lips, evil is ensnared, but the righteous escapes from trouble.}%
\verse{From the fruit of the mouth of a man, he is filled with good, and the reward of a man’s labor\lebnote{“the hands of a man”} will return to him.}%
\verse{The way of a fool is upright in his own eyes, but he who listens to advice is wise.}%
\verse{As for a fool, on that very day\lebnote{“on the day”} he makes his anger known, but he who ignores an insult is prudent.}%
\verse{He who will speak truth will reveal righteousness, but the witness of falsehood, deceit.}%
\verse{There is one who speaks rashly, like the thrust of a sword, but the tongue of the wise brings healing.}%
\verse{A lip of truth endures forever, but a tongue of deception lasts only a moment.}%
\verse{Deceit is in the heart\lnBTM{} of those who plan\lnBTN{} evil, but to those who plan\lnBTN{} peace, there is joy.}%
\verse{No evil will happen\lebnote{“All evil will not happen”} to the righteous, but the wicked are filled with trouble.}%
\verse{An abomination of Adonai are lips of deceit, but they who act faithfully are his delight.}%
\verse{A clever person conceals knowledge, but the heart\lnBTM{} of a fool\lebnote{Hebrew “fools”} announces folly.}%
\verse{The hand of the diligent ones will rule, but the lazy will belong to forced labor.}%
\verse{Anxiety in the heart\lnBTM{} of a man will weigh him down, but a good word will cheer him.}%
\verse{A righteous person will seek out his neighbor, but the way of the wicked will lead them astray.}%
\verse{The lazy will not roast his game, but diligence is the precious wealth of a man.}%
\verse{On the road of righteousness is life, and on the way of the path, may there be no death.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 13
\verse{A wise child hears the discipline of a father, but a scoffer does not listen to a rebuke.}%
\verse{From the fruit of the mouth of a man, he shall eat what is good, but the desire\lebnote{Or “soul”} of the treacherous, wrongdoing.}%
\verse{He who keeps his mouth guards his life;\lnBTO{} he who opens his lips, ruin belongs to him.}%
\verse{The soul\lebnote{Or “life”} of the lazy craves, but there is nothing, but the person of diligence is richly supplied.}%
\verse{The righteous hates a word of falsehood; the wicked will bring shame and disgrace.\lebnote{“he will bring disgrace”}}%
\verse{Righteousness will guard the upright of way, but wickedness will overthrow sin.}%
\verse{There is one who acts rich but has nothing;\lebnote{“there is no”} another who pretends to be poor but has wealth.}%
\verse{The ransom of the life of a man is his wealth, but the poor does not receive a threat.}%
\verse{The light of the righteous will rejoice, but the lamp of the wicked will die out.}%
\verse{Only by insolence is strife set up, and wisdom is with those who take advice.}%
\verse{Wealth gained from haste\lebnote{Or “vanity”} will dwindle, but he who gathers little by little\lebnote{“upon the hand”} will increase it.}%
\verse{Hope that is deferred makes the heart\lebnote{Or “mind”} sick, but a desire fulfilled is a tree of life.}%
\verse{He who despises a word will bring destruction on himself, but he who respects a commandment will be rewarded.}%
\verse{The teaching of the wise is a fountain of life, in order to avoid the snares of death.}%
\verse{Good sense grants favor, but the way of the faithless is coarse.}%
\verse{Anyone who is clever will act with intelligence, but the fool will display folly.}%
\verse{A messenger of wickedness will fall into trouble, but an envoy of the faithful brings healing.}%
\verse{Poverty and disgrace belong to him who ignores instruction, but he who guards reproof will be honored.}%
\verse{A desire fulfilled\lebnote{“made to be,” “brought about”} will be sweet to the soul, but an abomination of fools is turning from evil.}%
\verse{Walk with the wise and be wise, but as for the companion of fools, he will suffer harm.}%
\verse{Misfortune will pursue sinners, but the righteous will be rewarded with prosperity.\lebnote{“it will reward prosperity”}}%
\verse{He who is good will leave an inheritance to his grandchildren,\lebnote{“sons of sons”} and stored up for the righteous is the wealth of a sinner.}%
\verse{There is much food in the field of the poor, but it is swept away\lebnote{“there is sweeping away”} by injustice.\lebnote{“no justice”}}%
\verse{He who withholds his rod hates his child, but he who loves him gives him\lebnote{“visits him”} discipline.}%
\verse{The righteous eats to satisfy his life,\lnBTO{} but the belly of the wicked will lack.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 14
\verse{The wisest of women\lebnote{“wise ones of women”} builds her house, but the foolish tears it down with her hands.}%
\verse{He who walks in uprightness fears Adonai, but he who is devious in his ways displeases him.}%
\verse{In the mouth of a fool is the rod of pride, but the lips of the wise preserve them.}%
\verse{When there are no\lebnote{“In there is no”} oxen the manger is empty, but an abundance of crops comes by the strength of an ox.\lebnote{Hebrew “bull”}}%
\verse{A faithful witness does not lie, but he who breathes out falsehood is a witness of deceit.}%
\verse{A scoffer seeks wisdom, but there is none, but knowledge comes easily to him who understands.\lebnote{“to one who understands it comes easy”}}%
\verse{Leave the presence of a foolish man,\lebnote{“from the presence \textit{belonging} to a man of foolishness”} for you will not come to know words of knowledge.\lebnote{“lips of knowledge”}}%
\verse{The wisdom of the clever is understanding his ways, but the folly of fools is deceit.}%
\verse{Fools mock\lebnote{Hebrew “he mocks”} the guilt offering, but among the upright, it is favorable.}%
\verse{The heart\lnBTP{} knows the bitterness of its soul,\lebnote{Or “life,” or “inner self”} but in its joy, it will not share itself with a stranger.}%
\verse{The house of the wicked will be destroyed, but the tent of the upright will flourish.}%
\verse{There is a way that seems upright to\lebnote{“to the face of”} a man, but its end is the way of death.}%
\verse{Even in laughter, a heart may be sad, and the end of joy may be grief.}%
\verse{From his ways, the perverse of heart will be satisfied, and from his own,\lebnote{“from upon him”} so shall a good man.}%
\verse{The simple will believe every word, but the clever will consider his step.}%
\verse{The wise is cautious and turns from evil, but the fool throws off restraint and is confident.}%
\verse{He who is short of temper\lebnote{“nostril”} will act foolishly, and the man who schemes will be hated.}%
\verse{The simple are adorned with folly, but the clever are crowned with knowledge.}%
\verse{The evil bow down before the good, and the wicked at the gates of the righteous.}%
\verse{The poor is disliked even by his neighbor, but the lovers of the rich are many.}%
\verse{He who despises his neighbor is a sinner, but he who has mercy on the poor blesses him.}%
\verse{Have they not erred, those who plan evil? But loyalty and faithfulness belong to those who plan good.}%
\verse{In all toil, there is profit, but the talk\lebnote{“word, matter, thing”} of lips leads only to poverty.}%
\verse{The crown of the wise is their wealth; the folly of fools is folly.}%
\verse{He who saves lives\lebnote{Or “souls,” or “inner selves”} is a witness of truth, but he who utters lies is a betrayer.}%
\verse{In the fear of Adonai, there is confidence of strength, and for his children, there will be refuge.}%
\verse{The fear of Adonai is a fountain of life, in order to turn from the snares of death.}%
\verse{In the multitude of people is the glory of the king, but without a population, a prince is ruined.}%
\verse{He who is slow to anger has great understanding, but the hasty of spirit\lebnote{Or “breath”} exalts folly.}%
\verse{A heart\lnBTP{} of tranquility\lebnote{Or “he who heals”} is life to the flesh, but causes bones of passion to rot.}%
\verse{He who oppresses the poor insults him who made him, but he who has mercy on the poor honors him.}%
\verse{By his evildoing, the wicked will be overthrown, and the righteous will find refuge in his death.}%
\verse{In the heart\lnBTP{} of him who has understanding, wisdom rests, but even in the midst of fools it becomes known.}%
\verse{Righteousness will exalt a nation, but sin is a reproach to a people.}%
\verse{The favor of a king is for the servant who deals wisely, but his wrath will be on him who acts shamefully.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 15
\verse{A soft answer will turn away wrath, but a word of trouble will stir anger.\lebnote{“nostril”}}%
\verse{The tongue of the wise will dispense knowledge, but the mouth of fools will pour out folly.}%
\verse{In every place, the eyes of Adonai keep watch over the evil and the good.}%
\verse{Gentleness\lebnote{Or “Healing”} of tongue is a tree of life, but perverseness in it\lebnote{That is, the tongue} causes a break in spirit.}%
\verse{A fool will despise the instruction of his father, but he who guards reproof is prudent.}%
\verse{In the house of the righteous there is much treasure, but the income of the wicked brings trouble.}%
\verse{The lips of the wise will spread knowledge, but the heart\lnBTQ{} of fools, not so.}%
\verse{The sacrifice of the wicked is an abomination of Adonai, but the prayer of the upright is his delight.}%
\verse{An abomination of Adonai is the way of the wicked, but he who pursues righteousness he will love.}%
\verse{Severe discipline belongs to him who forsakes the way; he who hates a rebuke will die.}%
\verse{Sheol\lebnote{A term for the place where the dead reside, i.e., the Underworld} and Abaddon\lebnote{Poetic synonym for “Sheol.” Only mentioned in the OT in relation to Sheol, the grave, or death.} are before Adonai, how much more\lebnote{“more for”} the hearts of the children of men!}%
\verse{A scoffer does not like his rebuke;\lebnote{“rebuke \textit{belonging} to him”} to the wise he will not go.}%
\verse{A heart\lnBTQ{} of gladness will make good countenance,\lebnote{“faces”} but in sorrow of heart\lnBTQ{} a spirit is broken.}%
\verse{The heart of him who understands will seek knowledge, but the faces of fools, they\lebnote{Hebrew “he/it”} will feed on folly.}%
\verse{All the days of the poor are hard, but goodness of heart\lnBTQ{} is a continuous feast.}%
\verse{Better is little with the fear of Adonai than great treasure and trouble with it.\lnBTR{}}%
\verse{Better is a dinner of vegetables when\lebnote{Hebrew “and”} love is there than a fattened ox and hatred with it.\lnBTR{}}%
\verse{A man who is hot-tempered will stir up strife, but he who is slow to anger,\lebnote{“nostrils”} he will calm contention.}%
\verse{The way of the lazy is like a hedge of thorns,\lebnote{Hebrew “thorn”} but the path of the upright is a highway.}%
\verse{A child of wisdom will make a father glad, but a foolish person,\lebnote{“foolishness of humankind”} he despises his mother.}%
\verse{Folly is a joy to him who lacks sense,\lebnote{“heart”} and a person of understanding will walk upright.\lebnote{“he will walk upright walking”}}%
\verse{Plans go wrong when there is no counsel, but with many advisors it will succeed.}%
\verse{Joy belongs to a man with answers in his mouth, and a word in its time, how good it is!}%
\verse{The path of life leads upward for him who has insight, in order to turn away from Sheol below.}%
\verse{The house of the proud, Adonai will tear it down, but he will maintain the property line\lebnote{“boundary”} of the widow.}%
\verse{Plans of evil are an abomination of Adonai, but gracious words are pure.}%
\verse{He who makes trouble for his house is he who is greedy for unjust gain, but he who hates bribes will live.}%
\verse{A heart\lnBTQ{} of righteousness will ponder the answer, but a mouth of wickedness will pour out deceit.}%
\verse{Adonai is far from the wicked, but the prayers of the righteous he will hear.}%
\verse{From the light of the eyes, the heart\lnBTQ{} will rejoice, and good news will enliven the bones.\lebnote{Hebrew “bone”}}%
\verse{The ear of him who listens to admonitions of life, in the midst of the wise it will lodge.}%
\verse{He who ignores instruction despises himself,\lebnote{Or “soul,” or “inner self”} but he who hears admonition gains heart.\lnBTQ{}}%
\verse{Fear of Adonai is the instruction of the wise, and before honor comes humility.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 16
\verse{To mortals belong the plans of the heart, but from Adonai comes the answer of the tongue.}%
\verse{All the ways of a man are pure in his own eyes, but Adonai weighs the spirit.\lebnote{Or “breath”}}%
\verse{Commit your work to Adonai, and your plans will be established.}%
\verse{All Adonai has made is for his\lebnote{Or “its”} purpose, and even the wicked for the day of trouble.}%
\verse{An abomination of Adonai are all who are arrogant of heart;\lnBTS{} rest assured,\lebnote{“hand to hand”} he will not go unpunished.}%
\verse{By loyalty and faithfulness, iniquity will be covered over,\lebnote{Or “atoned for”} and by fear of Adonai one turns\lebnote{“turning”} from evil.}%
\verse{When the ways of a man are pleasing to Adonai, even his enemies he will cause to make peace with him.}%
\verse{Better is little with righteousness than great income with no justice.}%
\verse{The mind\lebnote{“heart”} of a person will plan his ways, and Adonai will direct his steps.}%
\verse{A decision is upon the lips of a king; in judgment his mouth will not sin.}%
\verse{A balance and scales of justice belong to Adonai; all the weights of the bag are his work.}%
\verse{An abomination to kings is doing evil, for by righteousness the throne will be established.}%
\verse{The delight of kings are the lips of righteousness, and he who speaks what is upright he will love.}%
\verse{The wrath of a king is a messenger of death, but one who is wise will appease\lebnote{Or “atone, cover over”} it.}%
\verse{In the light of the face of the king there is life, and his favor is like a cloud of spring rain.}%
\verse{Getting wisdom: how much\lebnote{“what”} better than gold! And getting understanding: it is chosen over\lebnote{“from”} silver.}%
\verse{The highway of the upright, it turns from evil, he who guards himself\lnBTT{} keeps his way.}%
\verse{Before destruction comes pride, and before a fall, a haughty spirit.\lebnote{Hebrew “a haughty of spirit”}}%
\verse{Better a lowly spirit with the poor than dividing the spoil with the proud.}%
\verse{He who is attentive to a matter will find goodness, and he who trusts in Adonai, his own happiness.}%
\verse{The wise of heart\lnBTS{} is called perceptive, but he who is pleasant of lips will increase persuasiveness.}%
\verse{A fountain of life is wisdom for its owner,\lebnote{Or “master”} but the instruction\lebnote{Or “punishment”} of fools is folly.}%
\verse{The heart\lnBTS{} of the wise will make his mouth judicious, and upon his lips, it\lebnote{Or “he”} will add persuasiveness.}%
\verse{Pleasant sayings are a honeycomb, sweetness to the soul and healing to the bones.}%
\verse{There is a way that seems upright to\lebnote{“to the face of”} a man, but its end is the way of death.}%
\verse{The life\lnBTT{} of a worker works for him, for his hunger\lebnote{“mouth”} urges him.}%
\verse{A man of wickedness concocts evil, and his lips are like a scorching fire.}%
\verse{A person of perversity will spread dissent, and he who whispers separates a close friend.}%
\verse{A person of violence will entice his neighbor and cause him to walk on a way that is not good.}%
\verse{He who winks his eyes does so in order to plan perverse things; he who purses his lips will bring evil to pass.}%
\verse{A crown of glory is gray hair; by a righteous life it is gained.}%
\verse{He who is slow to anger\lebnote{“nostrils”} is better than him who is mighty, and he who controls his spirit than him who captures a city.}%
\verse{The lot will be cast into the lap, but all of its decisions are from Adonai.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 17
\verse{Better a dry morsel and quiet with it than a house filled with feasts of strife.}%
\verse{A slave who deals wisely will rule over a child who acts shamefully, and in the midst of brothers he will share his inheritance.}%
\verse{A crucible is for the silver, and a furnace is for the gold, but Adonai will test hearts.}%
\verse{He who does evil listens to lips of wickedness, and the liar gives heed to the tongue of mischief.}%
\verse{He who mocks the poor insults him who made him; he who rejoices at calamity will not go unpunished.}%
\verse{The crown of the elderly are grandchildren,\lebnote{“sons of sons”} and the glory of children is their fathers.}%
\verse{Fine speech\lebnote{“A lip of fine\textit{ness}”} is not becoming a fool, still less\lebnote{“only for”} is false speech\lebnote{“lip of deceit”} for a ruler.}%
\verse{The bribe is a stone of magic in the eyes of its owner;\lebnote{Or “master”} everywhere\lebnote{“to all which”} he will turn, he will prosper.}%
\verse{He who forgives an affront fosters love, but he who waits on a matter will alienate a friend.}%
\verse{A rebuke strikes him who understands deeper than one hundred blows to a fool.}%
\verse{An evil person will seek only rebellion, and a cruel messenger will be sent against him.}%
\verse{May a man meet a she-bear robbed of offspring and not a fool in his folly.}%
\verse{For he who returns evil for good, evil will not depart from his house.}%
\verse{Like the release of water is the beginning of strife; before it breaks out, stop the quarrel.}%
\verse{He who justifies the wicked and he who condemns the righteous, the two of them are both abominations of Adonai.}%
\verse{Why is this? A price in the hand of a fool, in order to buy wisdom where\lebnote{Hebrew “and”} there is no sense.\lnBTU{}}%
\verse{The friend loves at all times\lebnote{Hebrew “time”}, but a brother is born for adversity.}%
\verse{A person who lacks sense\lnBTU{} pledges;\lebnote{“pledges a hand”} he becomes security before his neighbor.}%
\verse{He who loves transgression loves strife; he who builds his high thresholds seeks destruction.}%
\verse{He who is crooked of heart\lebnote{Or “mind”} will not find goodness, and he who is perverse, by his tongue he will fall into calamity.}%
\verse{He who begets a fool, there is trouble for him; the father of a fool will not rejoice.}%
\verse{A cheerful heart is good medicine, but a downcast spirit will dry out bones.}%
\verse{The wicked will accept a bribe from the lap, in order to pervert the ways of justice.}%
\verse{He who understands sets his face toward wisdom, but the eyes of a fool, to the end of the earth.\lebnote{Or “land”}}%
\verse{A grief to his father is the child of a fool, and bitterness to her who bore him.}%
\verse{Also, imposing a fine on the righteous is not good, nor to flog nobles for uprightness.}%
\verse{He who spares his sayings knows knowledge, and a cool spirit is a man of understanding.}%
\verse{Even a fool who keeps silent shall be considered wise;\lebnote{“wise, he shall be considered”} he who closes his lips is intelligent.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 18
\verse{He who is selfish seeks a craving;\lebnote{Or “boundary”} against all sound judgment he shows contempt.}%
\verse{A fool will not take pleasure in understanding, but in expressing his heart.\lnBTV{}}%
\verse{With the coming of wickedness comes contempt also, and with dishonor, disgrace.}%
\verse{Deep waters are words of the mouth of a man; a gushing stream is a fountain of wisdom.}%
\verse{Being partial to faces of evil is not good, nor to subvert the righteous at the judgment.}%
\verse{The lips of a fool will bring strife, and his mouth calls out for a flogging.}%
\verse{The mouth of a fool is ruin to him, and his lips are a snare to his soul.\lebnote{Or “life,” or “inner self”}}%
\verse{The words of a whisper are like delicious morsels, and they themselves go down to inner parts of the body.}%
\verse{Even he who is slack in his work, he is brother to a master\lebnote{Or “owner”} of destruction.}%
\verse{A tower of strength is the name of Adonai; into him\lebnote{Or “it”} the righteous will run and be safe.}%
\verse{The wealth of the rich is his strong city,\lebnote{“a city of his strength”} and like a wall, it is high in his imagination.}%
\verse{In the presence of destruction, the heart\lnBTV{} of a man will be haughty, but in the presence of honor, humble.}%
\verse{He who returns a word before he will hear, folly itself belongs to him as well as\lebnote{Hebrew “and”} shame.}%
\verse{The spirit of a man will endure his sickness, but a broken spirit, who may bear it?}%
\verse{An intelligent mind\lebnote{“heart”} will acquire knowledge, and the ear of the wise will seek knowledge.}%
\verse{The gift of a person will open doors for him, and before the great, it gives him access.}%
\verse{The first in his dispute is deemed righteous, but his neighbor will come and examine him.}%
\verse{The lot will put an end to disputes, and between powerful contenders it will decide.}%
\verse{A brother who is offended is worse than a city of strength, and quarrels are like the bars of a fortification.}%
\verse{From the fruit of a man’s mouth, his stomach will be satisfied, as for the yield of his lips, it will satisfy.}%
\verse{Death and life are in the power\lebnote{“hand”} of the tongue, and those who love her\lebnote{That is, the power of the tongue} will eat of her fruit.}%
\verse{He who finds a wife\lebnote{Or “woman”} finds good, and he will obtain favor from Adonai.}%
\verse{The poor may speak entreaties, but the rich will answer roughly.}%
\verse{A man of many friends will come to ruin, but there is a friend who sticks closer than a brother.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 19
\verse{Better a poor person walking in integrity than one who is perverse in his speech\lebnote{“in his lips”} and is a fool.}%
\verse{Also, a life\lnBTW{} without knowledge is not good, and he who moves quickly with his feet misses the mark.\lebnote{Or “sins”}}%
\verse{As for the folly of humankind, its way leads to ruin, and against Adonai his heart\lnBTX{} will rage.}%
\verse{Wealth adds many friends, but the poor will be left by his friends.}%
\verse{A witness of falsehood will not go unpunished, and he who breathes lies will not escape.}%
\verse{Many will seek favor before the generous, and everyone is the friend of a man of gifts.\lebnote{Hebrew “gift”}}%
\verse{All the brothers of the poor, if they hate him, how much more will his friends keep away from him. He pursues them with words, and they are gone.\lebnote{Or “\textit{when} he pursues words \textit{and} not them”}}%
\verse{He who acquires wisdom loves himself;\lnBTW{} he who guards understanding loves to find good.}%
\verse{A false witness will not go unpunished, and he who breathes lies will perish.}%
\verse{For a fool living in luxury is not fitting, any more than it is for a slave to rule over princes.}%
\verse{The understanding of a person makes him slow to his anger,\lebnote{“nostril”} and his glory overlooks offense.}%
\verse{The rage of a king growls like a lion, but his favor is like dew on the grass.}%
\verse{A foolish child is a ruin to his father, and the quarreling of a woman\lnBTY{} is a continuous dripping.}%
\verse{A house and wealth are an inheritance from fathers, but from Adonai comes a woman\lnBTY{} who is prudent.}%
\verse{Laziness will bring on a deep sleep, and a person\lebnote{Or “soul,” or “life”} of idleness will suffer hunger.}%
\verse{He who guards commandments guards his life;\lnBTW{} he who is careless of his ways will be killed.}%
\verse{He who lends to Adonai is he who is kind to the poor, and his benefits he will repay to him.}%
\verse{Discipline your child, for there is hope, but on his destruction do not set your desire.\lnBTW{}}%
\verse{A hot-tempered person pays a penalty; if you rescue him, you will do it yet again.}%
\verse{Listen to advice and accept instruction so that you will gain wisdom for your future.\lebnote{“after \textit{things}”}}%
\verse{Many plans are in the heart\lnBTX{} of a man, but the purpose of Adonai will be established.}%
\verse{The craving of a man is his steadfast loyalty, and it is better to be poor than a liar.\lebnote{“man of lying”}}%
\verse{Fear of Adonai leads to life; he who is filled with it will rest — he will not suffer harm.}%
\verse{A lazy person buries his hand in the dish, and even to his mouth he will not bring it back.}%
\verse{The scoffer you shall strike, and the simple, may they learn prudence, and reprove the intelligent and he will gain knowledge.}%
\verse{He who does violence to a father, he who chases away a mother, is a child who causes shame and brings reproach.}%
\verse{Cease to listen to instruction, my child, and you will stray\lebnote{“\textit{in order} to stray”} from sayings of knowledge.}%
\verse{A worthless witness will mock justice, and the mouth of the wicked will devour iniquity.}%
\verse{Judgments\lebnote{Or “Punishments,” or “Condemnations”} are prepared for the scoffers, and flogging for the back of fools.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 20
\verse{Wine is a mocker, strong drink a brawler, and any who go astray by it are not wise.}%
\verse{Roaring like the lion is the dreaded anger of a king; he who provokes him forfeits his life.\lebnote{Or “soul,” or “inner self”}}%
\verse{It is honorable for the man to refrain from strife, but every fool will be quick to quarrel.}%
\verse{The lazy person will not plow in season; he will expect at the harvest, but there will be nothing.}%
\verse{Deep waters are like purpose in the heart\lnBTZ{} of a man, and a man of understanding will draw it\lebnote{Or “her”; referring to “purpose”} out.}%
\verse{Many a person will proclaim his loyalty for himself, but a man who is trustworthy, who can find?}%
\verse{He who walks in his integrity is righteous; happy are his children who follow him.}%
\verse{A king who sits on the throne of judgment winnows all evil with his eyes.}%
\verse{Who will say “I have made my heart\lnBTZ{} clean; I am pure from my sin”?}%
\verse{Stone and stone, measure and measure, both of them\lnBUA{} are an abomination of Adonai.}%
\verse{Even by his acts, a young man\lebnote{Or “young boy,” or “adolescent”} will make himself known, whether his acts are pure and upright.}%
\verse{The ear that hears and the eye that sees, Adonai has made them both.\lnBUA{}}%
\verse{Do not love sleep, lest you become poor; open your eyes and have plenty bread.}%
\verse{“Bad, bad,” the buyer will say, but when one goes to him, then he will boast.}%
\verse{There is gold and many costly stones, but precious jewels are lips of knowledge.}%
\verse{Take his garment, for he has given security to a stranger, and on behalf of a foreigner — take it as pledge.}%
\verse{Bread gained by deceit is sweet for the man, but afterward, his mouth will be filled with gravel.}%
\verse{A plan\lebnote{Hebrew “plans”} will be established by advice, and with guidance make war.}%
\verse{He reveals a secret, he who walks about with gossip, and do not associate with a babbler’s lips.}%
\verse{He who curses his father and his mother, his lamp will be extinguished in the midst of darkness.\lebnote{“in a pupil of darkness”}}%
\verse{An inheritance acquired hastily at the beginning\lebnote{“at the first”} will not be blessed at its end.}%
\verse{Do not say “I will repay evil”; wait for Adonai and he will deliver you.}%
\verse{An abomination of Adonai is a stone and a weight,\lebnote{“a stone and a stone”} and scales of falsehood are not good.}%
\verse{Away from Adonai are the steps of a strong man, and how will humankind understand his ways?}%
\verse{It is a snare to humankind to say rashly “It is holy,” and after vows, to scrutinize.}%
\verse{A wise king winnows the wicked, and he will drive a wheel over them.}%
\verse{The lamp of Adonai is the spirit\lebnote{Or “soul”} of humankind, he who searches every innermost part.\lnBUB{}}%
\verse{Loyalty and faithfulness will preserve a king, and he is upheld with the righteousness of his throne.}%
\verse{The glory of young men is their strength,\lebnote{Or “strengthens them”} but the beauty of the aged is gray hair.}%
\verse{The blows of a wound will cleanse evil, as will\lebnote{Hebrew “and”} beatings of the innermost part.\lnBUB{}}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 21
\verse{Streams of water are the heart\lebnote{Or “mind”} of a king in the hand of Adonai; wherever\lebnote{“upon all that”} he will desire, he will turn.}%
\verse{Every way of a man is upright in his own eyes, but Adonai weighs hearts.\lebnote{Or “minds”}}%
\verse{Doing righteousness and justice is more acceptable to Adonai than sacrifice.}%
\verse{Haughtiness of the eyes and pride of heart, the lamp of the wicked are sin.}%
\verse{The plans of the diligent only lead to abundance, but all who are hasty, only to want.}%
\verse{He who makes treasure by a lying tongue is a fleeting vapor and seeker\lebnote{Hebrew “seekers”} of death.}%
\verse{The violence of the wicked will sweep them away, for they refuse to do justice.}%
\verse{Crooked is the way of a man and a foreigner, but the pure is upright in his conduct.}%
\verse{Better to dwell on the corner of a roof than to share a house with a woman\lebnote{Or “wife”} of contention.}%
\verse{The soul\lebnote{Or “soul, life, throat”} of the wicked desires evil; his neighbor will not find mercy in his eyes.}%
\verse{With the punishment of a scoffer, the simple will become wise, and with the instruction of the wise, he will obtain knowledge.}%
\verse{The righteous observes the house of the wicked; he throws the wicked to ruin.}%
\verse{He who closes his ear from the cry of the poor, he also will cry out and not be heard.}%
\verse{A gift in secret\lebnote{Hebrew “the secret”} will avert anger,\lebnote{“nostril”} and a concealed bribe,\lebnote{“bribe in the bosom”} strong wrath.}%
\verse{It is a joy to the righteous to do justice, but dismay to those who do evil.}%
\verse{Whoever wanders from the way of understanding, in the assembly of the dead\lebnote{Or “Rephaim”} he will rest.}%
\verse{A man of want is he who loves pleasure; he who loves wine and oil\lebnote{Or “fat”} will not become rich.}%
\verse{A ransom for the righteous is the wicked, and the faithless instead of the upright.}%
\verse{Better to live in a land of wilderness than with a wife of quarrels and provocation.}%
\verse{Precious treasure and oil are in the house of the wise, but the foolish person\lebnote{Or “foolish of mankind”} will devour them.\lebnote{“it”}}%
\verse{He who pursues righteousness and kindness will find life, righteousness, and honor.}%
\verse{To a city of warriors, the wise ascends, and he will bring down the stronghold — its object of trust.}%
\verse{He who guards his mouth and his tongue, he guards his life\lebnote{Or “soul,” or “inner self”} from danger.}%
\verse{The proud, haughty one: “scoffer” is his name; he acts with arrogance of pride.}%
\verse{The craving of a lazy person will kill him, for his hands refuse to work.\lebnote{Or “make”}}%
\verse{All day\lebnote{“All the day”} he craves a craving, but the righteous will give and not hold back.}%
\verse{The sacrifice of the wicked is an abomination; how much more when he brings it in divisiveness!}%
\verse{A false witness will perish, but a man who listens will testify with\lebnote{Hebrew “to the”} success.}%
\verse{A wicked man is strong in his countenance,\lebnote{“his face”} but the upright will appoint his paths.}%
\verse{There is no wisdom, nor understanding, nor counsel to oppose Adonai.}%
\verse{A horse is prepared for the day of battle, but to Adonai belongs the victory.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 22
\verse{A reputation\lebnote{“name”} is to be chosen rather than\lebnote{“from”} great riches; favor is better than silver and gold.\lebnote{“from silver and from gold, favor \textit{is} better”}}%
\verse{Rich and poor have much in common; Adonai is the maker of all of them.}%
\verse{The clever sees danger and hides, but the simple go on and suffer.}%
\verse{The reward of humility is the fear of Adonai — wealth and honor and life.}%
\verse{Thorns and snares are in the way of the perverse; he who guards himself\lebnote{Or “soul,” or “life”} will keep away from them.}%
\verse{Train the child concerning his way;\lebnote{“on the mouth of his way”} even when he is old, he will not stray from it.}%
\verse{The rich will rule over the poor, and the borrower is a slave of the lender.\lebnote{“the borrower \textit{belonging} to a man”}}%
\verse{He who sows injustice will reap calamity, and the rod of his anger will fail.}%
\verse{He who is generous\lebnote{“the generous of eye”} will be blessed, for he gives to the poor from his own bread.}%
\verse{Drive out a scoffer and strife will go out; quarrel and abuse will cease.}%
\verse{He who loves purity of heart and hasgracious speech,\lebnote{“grace of his lips”} his friend is the king.}%
\verse{The eyes of Adonai keep watch over knowledge, but he will overthrow the words of the faithless.}%
\verse{A lazy person says “A lion in the street! In the middle of the highway, I shall be killed!”}%
\verse{A deep pit is the mouth of an adulteress,\lebnote{“a strange woman”} he with whom Adonai is angry will fall there.}%
\verse{Folly is bound up in the heart\lnBUC{} of a boy;\lebnote{Or “young man,” or “adolescent”} the rod of discipline will drive it\lebnote{That is, folly} from him.}%
\verse{He who oppresses the poor in order to enrich himself, or gives to the rich, will come to poverty.\lebnote{“only loss”}}%
\verseWithHeading{Words of the Wise}{Incline your ear and hear the words of the wise; you shall apply your heart\lnBUC{} to my teaching.}%
\verse{For it is pleasant if you guard them within you;\lebnote{“in your belly/womb”} together they will be ready upon your lips.}%
\verse{In order for your trust to be in Adonai, I have made them known to you today\lebnote{“the day”} — even\lebnote{Or “only”} you.}%
\verse{Have I not written for you thirty sayings with admonitions and knowledge?}%
\verse{In order to show you what is right — sayings of truth — in order to return a true saying to him who sent you.}%
\verse{Do not rob the poor because he is poor, and do not crush the afflicted at the gate;}%
\verse{For Adonai will plead their case and despoil those who despoil them of life.\lnBUD{}}%
\verse{Do not befriend an owner\lebnote{Or “master”} of anger,\lebnote{“nostril”} and with a man of wrath you shall not associate;}%
\verse{lest you learn his way and become entangled in a snare to yourself.\lnBUD{}}%
\verse{Do not be with those who give a pledge\lebnote{“strike a hand”} by becoming\lebnote{“in the becomings of”} surety.}%
\verse{If there is nothing for you to pay,\lebnote{Or “complete”} why will he take your bed from under you?}%
\verse{Do not remove an ancient boundary marker which your ancestors\lebnote{Or “fathers”} made.}%
\verse{A man who is skillful in his work, you shall see: before kings, he will serve; he will not serve before the commoners.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 23
\verse{When you sit to eat with a ruler, you shall surely observe what is before you,}%
\verse{and you shall put a knife to your throat if you have a big appetite.\lebnote{“lord/master of life”}}%
\verse{Do not desire his delicacies, for\lebnote{Hebrew “and”} it is food of deception.}%
\verse{Do not tire in order to become rich; out of your understanding, may you desist.}%
\verse{Your eyes will alight\lebnote{“cause to fly”} on it, but there is nothing to it, for suddenly it will make for itself wings like an eagle and it will be exhausted in the heavens.}%
\verse{Do not eat the bread of the stingy,\lebnote{“stingy of eye”} and do not desire his delicacies.}%
\verse{For, like hair in his throat,\lebnote{“soul,” or “inner self”} so it is.\lebnote{Or “is he”} “Eat and drink!” he will say to you, but his heart will not be with you.}%
\verse{Your morsel you have eaten, you will vomit it up, and you will waste your pleasant words.}%
\verse{In the ears of a fool do not speak, for he will despise the wisdom of your words.}%
\verse{Do not remove an ancient boundary marker, and on the fields of orphans do not encroach;}%
\verse{For their redeemer is strong, he himself will plead their cause against you.}%
\verse{Apply your heart\lnBUE{} to instruction, and your ear to sayings of knowledge.}%
\verse{Do not withhold discipline from a child, if you will beat him with the rod, he will not die.}%
\verse{As for you, with the rod you shall beat him, and his life\lebnote{Or “soul,” or “inner self”} you will save from Sheol.\lebnote{A term for the place where the dead reside, i.e., the Underworld}}%
\verse{My child, if your heart\lnBUE{} is wise, my heart will be glad — even me!}%
\verse{And my insides\lebnote{“kidneys”} will rejoice when your lips speak what is upright.}%
\verse{May your heart\lnBUE{} not envy the sinners, but live in fear of Adonai all day.\lebnote{“all the day”}}%
\verse{Surely\lebnote{“For if”} there is a future, and your hope will not be cut off.}%
\verse{You, my child, hear and be wise, and direct your heart\lnBUE{} on the road.}%
\verse{Do not be among drinkers of wine, among gluttonous eaters of their meat.\lebnote{“meat for them”}}%
\verse{For the drunkard and gluttonous, they will become poor, and with rags, drowsiness will clothe them.}%
\verse{Listen to your father — he who gave you life,\lebnote{“caused you to be born”} and do not despise your mother when\lebnote{Or “for, because”} she is old.}%
\verse{Buy truth and do not sell it, wisdom and instruction and understanding.}%
\verse{The father of the righteous will surely rejoice; he who bears a wise person will be happy with him.}%
\verse{May your father and your mother be glad, and may she who bore you rejoice.}%
\verse{My child, may you give your heart\lnBUE{} to me, and may your eyes delight in my ways.}%
\verse{For a deep pit is a prostitute,\lebnote{Or “whore”} and a narrow well is an adulteress.\lebnote{“a foreign woman”}}%
\verse{She is also like a robber lying in wait, and the faithless among mankind she increases.}%
\verse{To whom is woe? To whom is sorrow? To whom are quarrels? To whom is complaint? To whom are wounds without cause? To whom is redness of the eyes?}%
\verse{To those who linger over wine, to those who come to try mixed wine.}%
\verse{Do not look at wine when it is red, when it sparkles\lebnote{“gives its eye”} on the cup, going down smoothly.}%
\verse{In the end,\lebnote{“end him”} it will bite like a serpent, and it will sting like an adder.}%
\verse{Your eyes will see strange things, and your heart\lnBUE{} will speak perverse things.}%
\verse{And you will be like him who lies down in the heart of the sea, and like him who lies down on top of a mast.}%
\verse{“They struck me; I was not hurt. They beat me; I did not know it. When I will awake, I will continue; I will seek it again.”}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 24
\verse{Do not envy men of evil, and do not desire to be with them.}%
\verse{For their minds will devise violence, and their lips will speak mischief.}%
\verse{By wisdom a house is built, and by understanding it is established.}%
\verse{And by knowledge, rooms are filled with all riches, precious and pleasant.}%
\verse{The warrior of wisdom is in strength, and a man of knowledge is strong in power.}%
\verse{For with wise guidance you shall make war for yourself, and victory is in an abundance of counsel.}%
\verse{Wisdom is too high for fools; at the gate he will not open his mouth.}%
\verse{He who plans to do evil for\lebnote{Or “to”} himself, they will call him “master\lebnote{Or “owner”} of mischief.”}%
\verse{Devising folly is a sin, and an abomination to humankind is a scoffer.}%
\verse{If you faint on the day of adversity, little is your strength.}%
\verse{Rescue those who are led away to the death and those who stagger to the slaughter. If you hold back,}%
\verse{if you say, “Look, we do not know this,” does not he who weighs hearts perceive it? And he who keeps your soul,\lnBUF{} he knows and will repay humankind according to his deeds.}%
\verse{My child, eat honey, for it is good, and the dripping of the honeycomb is sweet to your taste.}%
\verse{Thus know wisdom for the sake of your soul,\lnBUF{} if you find it, then there is a future, and your hope will not be cut off.}%
\verse{Do not lie in wait like an outlaw against the home of the righteous; do not do violence to his dwelling place.}%
\verse{For seven times the righteous will fall, but he will rise, but the wicked will be overthrown by calamity.}%
\verse{While your enemies are falling, do not rejoice; when he trips himself, may your heart\lnBUG{} not be glad}%
\verse{lest Adonai see and it be evil in his eyes, and turn his anger away from him.}%
\verse{Do not fret because of the evildoers; do not envy the wicked.}%
\verse{For there will not be a future for the evil; the lamp of the wicked will die out.}%
\verse{Fear Adonai, my son, and the king; with those who change, do not associate.}%
\verse{For suddenly their disaster will come, and the ruin of both of them,\lebnote{“\textit{the} two of them”} who knows?}%
\verse{These sayings are also for the wise: Partiality\lebnote{“acknowledge faces”} in judgment is not good.}%
\verse{Whoever says to the guilty, “You are righteous,” the people will curse him; the nations will abhor him.}%
\verse{But they who rebuke will have delight, and upon them blessings of goodness will come.}%
\verse{He will kiss the lips, he who gives an honest answer.}%
\verse{Prepare your work in the street and get it ready for yourself in the field; afterward, then\lebnote{Hebrew “and”} you shall build your house.}%
\verse{Do not be a witness without cause against your neighbor nor deceive with your lips.}%
\verse{Do not say, “Just as he has done to me, so shall I do to him; I will pay back the man according to his deed.”}%
\verse{I passed by the field of a lazy person, and over the vineyard of a person lacking sense;\lebnote{“heart”}}%
\verse{and behold, it was overgrown — all of it was covered with thorns, its surface with nettles, and its stone wall\lebnote{“a wall of his/its stones”} was broken down.}%
\verse{Then I myself saw and my heart\lnBUG{} considered; I looked, and I took hold of instruction:}%
\verse{A little sleep, a little slumber, a little folding of the hands for rest,}%
\verse{and your poverty will come running, and your lack like an armed warrior.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 25
\verseWithHeading{More Proverbs of Solomon}{These are also proverbs of Solomon which officials of Hezekiah king of Judah copied:}%
\verse{The glory of God\lebnote{Or “gods”} conceals things, but the glory of kings searches out things.}%
\verse{As heaven is to height and the earth is to depth, so is the heart\lebnote{Or “mind”} of kings — there is no searching.}%
\verse{Remove the dross from silver, and it will become a vessel for the smith.}%
\verse{Remove the wicked before a king, and his throne will be established in righteousness.\lebnote{Hebrew “in the righteousness”}}%
\verse{Do not promote yourself before the king, and in the place of the great ones do not stand.}%
\verse{For it is better that he say to you, “Ascend here,” than he humble you before a noble. What your eyes have seen,}%
\verse{do not hastily bring out to court, for\lebnote{Or “lest”} what will you do at its end, when your neighbor puts you to shame?}%
\verse{Argue your argument with your neighbor himself, the secret of another do not disclose,}%
\verse{lest he who hears shame you and your ill repute will not end.}%
\verse{Apples of gold in a setting of silver is a matter spoken at\lebnote{Hebrew “on”} its proper time.}%
\verse{A ring of gold and an ornament of fine gold is a rebuke of the wise to the ear of a listener.}%
\verse{Like the cold of snow on a day\lebnote{Or “at a season”} of harvest is a faithful messenger to those who send him, and the soul\lnBUH{} of his master is refreshed.\lebnote{Or “returned”}}%
\verse{Like clouds and wind when there is no rain, so too is a man who boasts in a gift of deception.}%
\verse{With patience\lebnote{“With length of face”} a ruler may be persuaded, and a soft tongue will break a bone.\lebnote{Or “strength”}}%
\verse{If you find honey, eat what is sufficient for you, lest you have your fill of it and vomit it out.}%
\verse{Make your foot scarce in the house of your neighbor, lest he become weary of you and hate you.}%
\verse{Like a club and sword and a sharp arrow is a man who bears false witness against his neighbor.}%
\verse{A bad tooth and a lame foot is the trust of a faithless person in a time\lebnote{“on a day”} of trouble.}%
\verse{Like one who removes a garment on a cold day, or like vinegar on natron,\lebnote{A mineral salt found on dry lake beds often used as a preservative.} is he who sings songs to a heavy heart.}%
\verse{If your enemy is hungry, feed him bread, and if thirsty, let him drink water.}%
\verse{For coals of fire you will heap upon his head, and Adonai will reward you.}%
\verse{The wind of the north produces rain, and a backbiting tongue, angry faces.}%
\verse{Better to live upon the corner of a roof than with a woman\lebnote{Or “wife”} of contention and in a shared house.}%
\verse{Like cold water\lebnote{Hebrew “waters”} upon a weary soul,\lnBUH{} so too is good news from a distant place.}%
\verse{Like a muddied spring or a polluted fountain is the righteous who gives way before the wicked.}%
\verse{To eat much honey is not good, nor is seeking one’s honor\lebnote{Hebrew “their honor”} honorable.}%
\verse{A breached city where there is no wall is like a man who has no\lebnote{“there is no”} self-control for his spirit.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 26
\verse{Like snow in the summer and like rain at the harvest, so honor is not fitting for a fool.}%
\verse{Like the sparrow is to fluttering and like the swallow is to flying, so an undeserved curse does not go forth.}%
\verse{A whip for the horse, a bridle for the donkey, and a rod for the back of fools.}%
\verse{Do not answer a fool according to his folly lest you become like him — even you.}%
\verse{Answer a fool according to his folly, or else he will be wise in his own eyes.}%
\verse{Like cutting off feet or drinking violence, so is he who sends messages in the hand of a fool.}%
\verse{Like legs that hang limp from a lame person, so is a proverb in the mouth of fools.}%
\verse{Like binding a stone in a sling, so is giving honor to a fool.}%
\verse{Like a thorn that goes up in the hand of a drunkard, so is a proverb in the mouth of fools.}%
\verse{Like an archer who wounds everyone, so is he who hires a fool or he who hires passersby.}%
\verse{Like a dog returning to his vomit is a fool reverting to his folly.}%
\verse{Do you see a man wise in his own eyes? There is more hope for a fool than for him.}%
\verse{A lazy person says “A lion is in the road! A lion among the streets!”}%
\verse{The door turns on its hinge, and a lazy person on his bed.}%
\verse{A lazy person buries his hands in the dish; he is too tired to return it to his mouth.}%
\verse{A lazy person is wiser in his eyes than seven who answer discreetly.}%
\verse{Grabbing onto the ears of a dog passing by is one who meddles in a quarrel that is not his own.}%
\verse{Like a maniac who shoots firebrands, arrows, and death,}%
\verse{so is a man who deceives his neighbor, but says “Am I not joking?”}%
\verse{For lack of wood, a fire goes out, and where there is no whisperer, quarreling will cease.}%
\verse{As charcoal is to hot embers and wood is to fire, so a man of quarrels is to kindling strife.}%
\verse{The words of a whisperer are like delicious morsels, and they go down to the inner parts of the body.}%
\verse{Like impure silver\lebnote{“silver of impurities”} which overlays an earthen vessel, so are smooth lips and an evil heart\lebnote{Or “mind”}.}%
\verse{On his lips, an enemy will pretend, but inside\lebnote{“in his midst”} he will harbor deceit.}%
\verse{When he makes his voice gracious, do not believe him, for seven abominations are in his heart.}%
\verse{Though hatred is covered with guile, its evil will be exposed in the assembly.}%
\verse{He who digs a pit, in it he will fall, and he who rolls a stone, on him it will come back.}%
\verse{A tongue of deceit hates its victim, and a flattering mouth makes ruin.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 27
\verse{Do not boast about tomorrow,\lebnote{“on the day of tomorrow”} for you do not know what the day will bring.}%
\verse{May another praise you and not your own mouth, a stranger and not your own lips.}%
\verse{Heavy is a stone and weighty is sand, but the provocation of a fool is heavier than both of them.}%
\verse{Cruel is wrath and overwhelming is anger, but who will stand before jealousy?}%
\verse{Better a rebuke that is open than a love that is hidden.}%
\verse{The wounds of a friend mean well, but the kisses of an enemy are profane.}%
\verse{An appetite\lnBUI{} that is sated spurns honey, but to an appetite\lnBUI{} that is ravenous, all bitterness is sweet.}%
\verse{Like a bird that strays from its nest, so is a man who strays from his place.}%
\verse{Perfume and incense will gladden a heart, and the pleasantness of one’s friend is personal advice.\lebnote{“because of advice of a person”}}%
\verse{As for your friend and a friend of your father, do not forsake them, and the house of your brother, do not enter on the day of your calamity. Better is a close neighbor than a distant brother.}%
\verse{Be wise, my child, and make my heart glad, and I will answer him who reproaches me with a word.}%
\verse{When the clever sees danger, he hides; the simple go on and suffer.}%
\verse{Take his garment, for he gives surety to a stranger, and to an adulteress\lebnote{“a foreign woman”} — so take his pledge.}%
\verse{He who blesses his neighbor with a loud voice early in the morning, a curse will be reckoned to him.}%
\verse{Dripping constantly on a day of heavy rain and a woman\lebnote{Or “wife”} of contention are alike.}%
\verse{In restraining her, he restrains wind,\lebnote{Or “breath, or “spirit”} and his right hand will grasp oil\lebnote{Or “fat”}.}%
\verse{As iron sharpens\lebnote{Or “is united with”} iron, so one man sharpens another.\lebnote{“a man sharpens \textit{the} faces of his friend”}}%
\verse{He who tends a fig tree will eat its fruit, and he who guards his master\lebnote{Or “lord”} will be honored.}%
\verse{As the waters reflect face to face,\lebnote{“the faces to the faces”} so the heart of a person reflects the person.}%
\verse{Sheol\lebnote{A term for the place where the dead reside, i.e., the Underworld} and Abaddon\lebnote{Poetic synonym for “Sheol.” Only mentioned in the OT in relation to Sheol, the grave, or death.} will not be satisfied, and the eyes of a person will not be satisfied either.}%
\verse{A crucible is for the silver, and a furnace for the gold, but a man is tested by the mouth of him who praises him.}%
\verse{If you crush a fool in the mortar with the pestle along with\lebnote{Or “in the midst of”} the crushed grain, it will not drive folly from upon him.}%
\verse{You will surely know the condition\lebnote{“faces”} of your flock; your heart\lebnote{Or “mind”} attends to the herds.}%
\verse{For riches are not forever, nor a crown for generation after generation.}%
\verse{When the grass is gone, then green growth will appear, and the herbs of the mountains will be gathered.}%
\verse{Lambs will be your clothing, and goats the price of the field.}%
\verse{And there will be enough goats’ milk for your food, for the food of your household and the nourishment\lebnote{“life”} of your maidservants.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 28
\verse{The wicked flee, but no one pursues, but the righteous is bold like a lion.}%
\verse{By the rebellion of a land, her rulers increase, but by a person of intelligence who knows justice, it will last.}%
\verse{A man who is poor and oppresses the impoverished is a beating rain that leaves\lebnote{“and there is”} no food.}%
\verse{Those who forsake instruction will praise the wicked, but they who guard instruction will struggle against them.}%
\verse{Men of evil do not understand justice, but seekers of Adonai understand completely.\lebnote{Or “everything”}}%
\verse{Better to be poor and walking in one’s integrity than to be crooked of ways when one is rich.}%
\verse{He who keeps instruction is a child of understanding, but the companion of gluttons will shame his father.}%
\verse{He who augments his wealth with interest and with usury gathers it for him who is kind to the poor.}%
\verse{He who turns his ear from listening to instruction, even his prayer is an abomination.}%
\verse{He who misleads the upright onto the way of evil, into his pits he will fall. But as for the blameless, they will inherit good.}%
\verse{A man of wealth is wise in his own eyes, but the intelligent poor sees through him.}%
\verse{When the righteous triumphs, great is the glory, but with the rising of the wicked, a person will be hidden.}%
\verse{He who conceals his transgression will not prosper, but he who confesses and forsakes will obtain mercy.}%
\verse{Happy is the person who fears continuously, but he who is stubborn of heart,\lebnote{Or “mind”} will fall into calamity.}%
\verse{Like a roaring lion and a charging bear is a wicked ruler over a poor people.}%
\verse{A ruler who lacks understanding is\lebnote{Hebrew “and”} a cruel oppressor, but those who hate unjust gain will have long\lebnote{Or “prolong”} days.}%
\verse{A person who is burdened with the blood of another,\lebnote{“life”} until death he will flee; do not take hold of him.}%
\verse{He who walks in integrity will be safe, but he who takes crooked paths will fall in one.}%
\verse{He who tills his ground will have plenty bread, but he who follows fantasies will have plenty of poverty.}%
\verse{A man of faithfulness has abundant blessings, but he who hurries to become rich will not go unpunished.}%
\verse{Showing partiality\lebnote{“Recognizing faces”} is not good, and over a morsel of bread, a strong man will do wrong.}%
\verse{He who hurries for wealth is a man with an evil eye,\lebnote{“a man of evil of eye”} but he does not know that poverty will come upon him.}%
\verse{He who rebukes a person will afterward find more favor than he who flatters with the tongue.}%
\verse{He who robs his father and his mother and says, “There is no crime,” is partner to a man who corrupts.}%
\verse{The greedy person\lebnote{“wide soul”} will stir up strife, but he who trusts in Adonai will be enriched.}%
\verse{He who trusts in his own heart\lebnote{Or “mind,” or “sense”} is a fool, but he who walks in wisdom will be saved.}%
\verse{For he who gives to the poor, there is no lacking, but for he\lebnote{Hebrew “those”} who turns his eyes, there are many curses.}%
\verse{With the rising of the wicked, humankind will be hidden, and with their perishing, the righteous will multiply.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 29
\verse{A man who is reproved, being stubborn of neck — suddenly he will be broken, and there will not be healing.}%
\verse{When the righteous are numerous, the people will rejoice, but when the wicked are ruling, people will groan.}%
\verse{A man who loves wisdom will make his parents glad, but the friend of prostitutes\lebnote{Or “whores”} will squander his wealth.}%
\verse{By justice a king gives stability to a land, but a man of bribes will ruin it.}%
\verse{A strong man who flatters his neighbor is spreading a net for his feet.}%
\verse{In transgression, an evil man is a snare, but the righteous will sing and rejoice.\lebnote{Hebrew “he will rejoice”}}%
\verse{The righteous knows the case of the poor, but the wicked does not understand knowledge.}%
\verse{Men of scoffing set a city aflame, but the wise turn away wrath.}%
\verse{If a wise man goes to court with a foolish man, then there is rankling and ridicule,\lebnote{Hebrew “there is ridicule”} but there is no relief.}%
\verse{Men of blood hate the blameless, and they seek the life of the upright.\lnBUJ{}}%
\verse{A fool gives all his breath,\lebnote{Or “spirit”} but the wise holds back in quiet.}%
\verse{A ruler listening to a word of falsehood, all his officials are wicked.}%
\verse{The poor and a man of oppression have this in common: Adonai gives light to the eyes of them both.\lebnote{“light of the eyes of the two of them”}}%
\verse{A king who judges with truthfulness to the poor, his throne will be established forever.}%
\verse{As for a rod and reproof, they\lebnote{Hebrew “it”} will give wisdom, but a neglected child is disgraced by his mother.}%
\verse{With the increase of the wicked, transgression will increase, but the righteous will look on his downfall.}%
\verse{Discipline your children, and they will give you rest, and they will give delight to your soul.\lebnote{Or “life,” or “inner self”}}%
\verse{When there is no prophecy, the people cast off restraint, but as for he who guards instruction, happiness is his.}%
\verse{By words, a servant is not disciplined, for he will understand, but there is no giving heed.}%
\verse{You see a man who is hasty in his words: there is more hope for a fool than him.}%
\verse{He who pampers his servant from childhood, arrogance will be his end.\lebnote{“and its/his end will be arrogance”}}%
\verse{A man of anger will stir strife, and the owner of anger, much transgression.}%
\verse{The pride of a person will bring him humiliation, and the lowly of spirit will obtain honor.}%
\verse{Being a partner with a thief is hating one’s life;\lnBUJ{} a curse he will hear, but not disclose.}%
\verse{The fear of a person will lay a snare, but he who trusts in Adonai will be secure.}%
\verse{Many are those who seek the favor\lebnote{“faces”} of a ruler, but from Adonai one obtains justice.\lebnote{“\textit{comes} justice for a man”}}%
\verse{A man of injustice is an abomination to the righteous, but the upright\lebnote{“upright of way”} is an abomination to the wicked.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 30
\verseWithHeading{The Oracle of Agur}{The words of Agur, son of Yakeh, the oracle. Thus says the man to Ithiel, to Ithiel, and Ucal:\lebnote{Or “I am weary, O God; I am weary, O God, and worn out”}}%
\verse{Certainly I am more stupid than a man, and the understanding of humankind is not for me.}%
\verse{And I have not learned wisdom, nor will I know knowledge of the Holy One.\lebnote{Or “holy ones”}}%
\verse{Who has ascended to heaven and come down? Who has gathered the wind in the hollow of his hand? Who has wrapped water in a\lebnote{Hebrew “the”} garment? Who has established all the ends of the earth? What is his name and what is the name of his child? For surely you know.}%
\verse{Every word of God is flawless; he is a shield for him who takes refuge in him.}%
\verse{Do not add to his words lest he rebuke you and you be found a liar.}%
\verse{Two things I ask from you; do not deny me before I die:}%
\verse{Keep falsehood and a lying word\lebnote{“word of a lie”} far from me; do not give me poverty or wealth; provide me with food only for my need.}%
\verse{Or else I will be satisfied and will deny him and say “Who is Adonai?” Or\lebnote{Or “Lest”} I will be poor and will steal and profane the name of my God.}%
\verse{Do not slander a servant to his master, or else he will curse you and you will be guilty.}%
\verse{There is a generation that will curse its father, and its mother it will not bless.}%
\verse{There is a generation that is pure in its own eyes, but from its filthiness it will not be cleansed.}%
\verse{There is a generation — how lofty are their eyes! And their eyelids they will lift.}%
\verse{There is a generation whose teeth are swords, and its jawbones, knives, in order to devour the poor from the earth\lnBUK{} and the needy from humankind.}%
\verse{For the leech, there are two daughters; “Give, give!” they cry. As for three of these, they are not satisfied; as for four, they do not say enough.}%
\verse{Sheol\lebnote{A term for the place where the dead reside, i.e., the Underworld} and barrenness of womb, the land is not satisfied with water, and fire does not say “enough!”}%
\verse{The eye that mocks a father and scorns the obedience of a mother — the ravens of the valley will peck it out, and the offspring of vultures will eat it.}%
\verse{Three of these are too wonderful for me, and four, I do not understand them:}%
\verse{the way of the eagle in the sky, the way of a snake on a rock, the way of a ship in the heart of the sea, and the way of a man with a young woman.}%
\verse{This is the way of a woman committing adultery: she eats and wipes her mouth, and says “I have not done wrong.”}%
\verse{Under three things the earth\lnBUK{} trembles, and under four, it is not able to bear up:}%
\verse{under a slave when he becomes king, and a fool when he is satisfied with food;}%
\verse{under an unloved woman when she gets married, and a maid when she succeeds her mistress.}%
\verse{There are four small things on the earth, and they are exceedingly wise:\lebnote{“wise \textit{ones} from wise \textit{ones}”}}%
\verse{The ants are a people who are not strong, yet they prepare their food in the summer;}%
\verse{the badgers are a people who are not mighty, yet they set their house on the rock;}%
\verse{there is no king for the locust, yet it marches in rank;}%
\verse{a lizard you can seize with hands, yet it is in palaces of kings.}%
\verse{There are three things that are magnificent of stride, and four that are magnificent when moving:}%
\verse{a mighty lion among the beasts,\lebnote{Hebrew “beasts”} but he will not turn back from any face;\lebnote{“faces of all”}}%
\verse{a strutting rooster or he-goat, and a king whose army is with him.}%
\verse{If you have been foolish by exalting yourself, and if you have devised evil, put your hand to your mouth.}%
\verse{For pressing milk produces curd, and pressing the nose produces blood, so pressing anger\lebnote{“nostrils”} produces strife.}%
\end{biblechapter}%
\begin{biblechapter}% Proverbs 31
\verseWithHeading{The Oracle of King Lemuel}{The words of Lemuel, the king — an oracle that his mother taught him:}%
\verse{What, my son? And what, my son in my womb? And what, son of my vows?}%
\verse{Do not give your strength to the women, and your ways to destroy kings.}%
\verse{It is not for the kings, O Lemuel; drinking wine is not for the kings, nor is strong drink for rulers.}%
\verse{Or else he will drink and forget what has been decreed, and he will pervert the rights\lnBUL{} of all the afflicted.\lebnote{“sons of affliction”}}%
\verse{Give strong drink to him who is perishing, and wine to those in bitter distress.\lebnote{“to bitterness of soul”}}%
\verse{He will drink and forget his poverty, and his misery he will not remember any more.}%
\verse{Open your mouth for the mute, for the rights\lnBUL{} of all the needy.\lebnote{“the sons of the needy”}}%
\verse{Open your mouth, judge righteousness, and defend the poor and needy.}%
\verseWithHeading{An Excellent Woman}{A woman\lebnote{Or “wife”} of excellence,\lnBUM{} who will find? For her worth is far more than precious jewels.\lebnote{Prov 31:10–31 is an acrostic poem}}%
\verse{The heart of her husband\lnBUN{} trusts in her, and gain he will not lack.}%
\verse{She does him good, but not harm all the days of her life.}%
\verse{She seeks wool and flax, and she works with the diligence of her hands.\lebnote{“her palms”}}%
\verse{She is like the ships of a merchant; from far off she brings her food;\lebnote{Or “bread”}}%
\verse{And she arises while it is still night, and gives food to her household, and tasks to her servant girls.}%
\verse{She considers a field and buys it, from the fruit of her hand\lebnote{“her palm”} she plants a vineyard.}%
\verse{She girds her waist in strength, and makes her arms strong.}%
\verse{She perceives that her merchandise is good; her lamp does not go out in the night.}%
\verse{Her hands she puts onto the distaff,\lebnote{A stick or spindle onto which wool or flax is wound in preparation for spinning} and her palms hold a spindle.}%
\verse{Her palm she opens to the poor, and her hand she reaches out to the needy.}%
\verse{She does not fear for her house when it snows, for her entire household\lebnote{“all her house, household”} is clothed in crimson.}%
\verse{She makes for herself coverings; her clothing is fine linen and purple.}%
\verse{Her husband\lnBUN{} is known at the gates, in his seat among the elders of the land.}%
\verse{Linen garments she makes and sells, and she supplies sashes for the merchants.}%
\verse{Strength and dignity are her clothing, and she laughs at the future.\lebnote{“to the day/time coming after”}}%
\verse{She opens her mouth with wisdom, and instruction of kindness is upon her tongue.}%
\verse{She looks after the ways of her household, and the bread of idleness she will not eat.}%
\verse{Her children rise and consider her happy, her husband\lnBUN{} also, and he praises her;}%
\verse{“Many daughters have done excellence,\lnBUM{} but you surpass\lebnote{“you ascend over”} all of them.”}%
\verse{Charm\lebnote{Hebrew “The charm”} is deceit and beauty\lebnote{Hebrew “the beauty”} is vain;\lebnote{Or “vapor,” or “emptiness,” or “breath”} but a woman who fears Adonai shall be praised.}%
\verse{Give to her from the fruit of her hand, and may they praise her works in the city gates.}%
\end{biblechapter}%
