\biblebook{1 Timothy}

\begin{biblechapter} % 1 Timothy 1
\verseWithHeading{Greeting} Paul, an apostle of Christ Jesus according to the command of God our Savior and of Christ Jesus our hope,
\verse to Timothy, \textit{my} true child in the faith. Grace, mercy, \textit{and} peace from God the Father and Christ Jesus our Lord.
\verseWithHeading{Instructions for Timothy in Ephesus} Just as I urged you \textit{when I} traveled to Macedonia, remain in Ephesus, so that you may instruct certain people not to teach other doctrine,
\verse and not to pay attention to myths and endless genealogies, which cause useless speculations rather than God’s plan \textit{that is} by faith.
\verse But the goal of our instruction is love from a pure heart and a good conscience and a faith without hypocrisy,
\verse from which some have deviated, \textit{and} have turned away into fruitless discussion,
\verse wanting to be teachers of the law, \textit{although they} do not understand either \textit{the things} which they are saying or \textit{the things} concerning which they are speaking confidently.
\verse But we know that the law \textit{is} good, if anyone makes use of it lawfully,
\verse knowing this, that the law \textit{is} not given for a righteous \textit{person} but for the lawless and rebellious, for the ungodly and sinners, for the unholy and totally worldly, for the one who kills his father and the one who kills his mother, for murderers,
\verse sexually immoral people, homosexuals, kidnappers, liars, perjurers, and \textit{whatever} else is opposed to sound teaching,
\verse according to the glorious gospel of the blessed God that I was entrusted with.
\verseWithHeading{Paul’s Thankfulness for the Mercy Shown to Him} \textit{I give thanks} to the one who strengthens me, Christ Jesus our Lord, because he considered me faithful, placing \textit{me} into ministry,
\verse \textit{although I} was formerly a blasphemer and a persecutor and a violent man, but I was shown mercy because I acted ignorantly in unbelief,
\verse and the grace of our Lord abounded with the faith and love \textit{that are} in Christ Jesus.
\verse The saying \textit{is} trustworthy and worthy of all acceptance: Christ Jesus came into the world to save sinners, of whom I am the foremost.
\verse But because of this I was shown mercy, in order that in me foremost, Christ Jesus might demonstrate his total patience, for an example for those who are going to believe in him for eternal life.
\verse Now to the King of the ages, immortal, invisible, to the only God, \textit{be} honor and glory \textit{forever and ever}. Amen.
\verseWithHeading{Paul’s Charge to Timothy} I am setting before you this instruction, Timothy \textit{my} child, in accordance with the prophecies spoken long ago about you, in order that by them you may fight the good fight,
\verse having faith and a good conscience, which some, \textit{because they} have rejected \textit{these}, have suffered shipwreck concerning their faith,
\verse among whom are Hymenaeus and Alexander, whom I have handed over to Satan, in order that they may be taught not to blaspheme.
\end{biblechapter}

\begin{biblechapter} % 1 Timothy 2
\verseWithHeading{Instructions to Pray for All People} Therefore, I urge first of all \textit{that} petitions, prayers, requests, \textit{and} thanksgiving be made on behalf of all people,
\verse on behalf of kings and all those who are in authority, in order that we may live a tranquil and quiet life in all godliness and dignity.
\verse This \textit{is} good and acceptable before God our Savior,
\verse who wants all people to be saved and to come to a knowledge of the truth.
\verse For \textit{there is} one God and one mediator between God and human beings, the man Christ Jesus,
\verse who gave himself a ransom for all, the testimony at the proper time,
\verse for which I was appointed a herald and an apostle—I am speaking the truth, \textit{I am} not lying—a teacher of the Gentiles in faith and truth.
\verse Therefore I want the men in every place to pray, lifting up holy hands without anger and dispute.
\verseWithHeading{Instructions for Women} Likewise also the women should adorn themselves in appropriate clothing, with modesty and self-control, not with braided hair and gold jewelry or pearls or expensive clothing,
\verse but with good deeds which are fitting for women who profess godliness.
\verse A woman must learn in quietness with all submission.
\verse But I do not permit a woman to teach or to exercise authority over a man, but \textit{to remain quiet}.
\verse For Adam was formed first, then Eve,
\verse and Adam was not deceived, but the woman, \textit{because she} was deceived, came into transgression.
\verse But she will be saved through the bearing of children, if she continues in faith and love and holiness with self-control.
\end{biblechapter}

\begin{biblechapter} % 1 Timothy 3
\verseWithHeading{Qualifications for Overseers} The saying \textit{is} trustworthy: if anyone aspires to supervision, he desires a good work.
\verse Therefore the overseer must be irreproachable, the husband of one wife, temperate, self-controlled, respectable, hospitable, skillful in teaching,
\verse not addicted to wine, not a violent person, but gentle, peaceable, not loving money,
\verse managing his own household well, having children in submission with all dignity
\verse (but if someone does not know how to manage his own household, how will he take care of the church of God?),
\verse not newly converted, lest he become conceited \textit{and} fall into the condemnation of the devil.
\verse But \textit{he} must also have a good testimony from those outside, in order that he may not fall into disgrace and the trap of the devil.
\verseWithHeading{Qualifications for Deacons} Deacons likewise \textit{must be} dignified, not insincere, not devoted to much wine, not fond of dishonest gain,
\verse holding the mystery of the faith with a clear conscience,
\verse and these also must be tested first; then let them serve \textit{if they} are above reproach.
\verse The wives likewise \textit{must be} dignified, not slanderous, temperate, faithful in all \textit{things}.
\verse Deacons must be husbands of one wife, managing \textit{their} children and their own households well.
\verse For those who have served well acquire a good standing for themselves, and great boldness in the faith \textit{that is} in Christ Jesus.
\verseWithHeading{The Mystery of Godliness Described} I am writing these \textit{things} to you, hoping to come to you in a short time.
\verse But if I am delayed, \textit{I am writing} in order that you may know how one must conduct oneself in the household of God, which is the church of the living God, the pillar and mainstay of the truth.
\verse And most certainly, great is the mystery of godliness: Who was revealed in the flesh, 
was vindicated by the Spirit, 
was seen by angels, 
was proclaimed among the Gentiles, 
was believed on in the world, 
was taken up in glory.
\end{biblechapter}

\begin{biblechapter} % 1 Timothy 4
\verseWithHeading{The Coming Apostasy} Now the Spirit explicitly says that in the last times some will depart from the faith, paying attention to deceitful spirits and teachings of demons,
\verse by the hypocrisy of liars, who are seared in their own conscience,
\verse who forbid marrying \textit{and insist on} abstaining from foods that God created for sharing in with thankfulness by those who believe and who know the truth,
\verse because everything created by God \textit{is} good and nothing \textit{is to be} rejected \textit{if it is} received with thankfulness,
\verse for it is made holy by the word of God and prayer.
\verseWithHeading{The Good Servant of Christ} \textit{By} teaching these \textit{things} to the brothers, you will be a good servant of Christ Jesus, trained in the words of the faith and of the good teaching that you have followed faithfully.
\verse But reject those \textit{worthless myths told by elderly women}, and train yourself for godliness.
\verse For the training of the body is \textit{somewhat} profitable, but godliness is profitable for everything, \textit{because it} holds promise for the present life and for the \textit{life} to come.
\verse The statement \textit{is} trustworthy and deserving of complete acceptance.
\verse For to this \textit{end} we labor and suffer reproach, because we have put our hope in \textit{the} living God, who is the Savior of all people, especially of believers.
\verse Command these \textit{things} and teach \textit{them}.
\verse Let no one look down on your youth, but be an example for the believers in word, in conduct, in love, in faith, in purity.
\verse Until I come, pay attention to the \textit{public} reading, to exhortation, to teaching.
\verse Do not neglect the gift \textit{that is} in you, that was granted to you through prophecy with the laying on of hands by the council of elders.
\verse Practice these \textit{things}. Be \textit{diligent} in these \textit{things}, in order that your progress may be evident to everyone.
\verse Fix \textit{your} attention on yourself and on your teaching. Continue in them, for \textit{by} doing this you will save both yourself and those who hear you.
\end{biblechapter}

\begin{biblechapter} % 1 Timothy 5
\verseWithHeading{Instructions About Widows} Do not rebuke an older man, but appeal to \textit{him} as a father, younger men as brothers,
\verse older women as mothers, younger women as sisters, with all purity.
\verse Honor widows \textit{who are} truly widows.
\verse But if any widow has children or grandchildren, they must learn to show profound respect for their own household first, and to pay back recompense to their parents, for this is pleasing in the sight of God.
\verse But the widow \textit{who is one} truly, and is left alone, has put her hope in God and continues in her petitions and prayers night and day.
\verse But the one who lives for sensual pleasure is dead \textit{even though she} lives.
\verse And command these \textit{things}, in order that they may be irreproachable.
\verse But if someone does not provide for his own \textit{relatives}, and especially the members of his household, he has denied the faith and is worse than an unbeliever.
\verse Let a widow be put on the list \textit{if she} is not less than sixty years \textit{old}, the wife of one husband,
\verse being well-attested by good works, if she has brought up children, if she has shown hospitality, if she has washed the feet of the saints, if she has helped those who are oppressed, if she has devoted herself to every good work.
\verse But refuse younger widows, for whenever their physical desires lead them away from Christ, they want to marry,
\verse \textit{thus} incurring condemnation because they have broken their former pledge.
\verse And at the same time also, going around from house to house, they learn \textit{to be} idle, and not only idle, but also gossipy and busybodies, saying the things \textit{that are} not necessary.
\verse Therefore I want younger \textit{widows} to marry, to bear children, to manage a household, to give the adversary no opportunity for reproach.
\verse For already some have turned away \textit{and followed} after Satan.
\verse If any believing woman has widows, she must help them, and the church must not be burdened, in order that it may help those \textit{who are} truly widows.
\verseWithHeading{Honoring Worthy Elders and Dealing With Sinners} The elders who lead well must be considered worthy of double honor, especially those who labor by speaking and teaching.
\verse For the scripture says, “You must not muzzle an ox \textit{while it} is threshing,” and “The worker \textit{is} worthy of his wages.”
\verse Do not accept an accusation against an elder except on the evidence of two or three witnesses.
\verse Reprove those who sin in the presence of all, in order that the rest also may experience fear.
\verse I testify solemnly before God and Christ Jesus and the elect angels that you observe these \textit{things} without prejudice, doing nothing according to partiality.
\verse Lay hands on no one hastily, and do not participate in the sins of others. Keep yourself pure.
\verse (No longer drink only water, but use a little wine for your stomach and your frequent illnesses.)
\verse The sins of some people are evident, preceding \textit{them} to judgment, but for some also they follow after \textit{them}.
\verse Likewise also good works \textit{are} evident, and those considered otherwise are not able to be hidden.
\end{biblechapter}

\begin{biblechapter} % 1 Timothy 6
\verseWithHeading{Slaves and Masters} All those who are under the yoke as slaves must regard their own masters as worthy of all honor, lest the name of God and the teaching be slandered.
\verse And those who have believing masters must not look down on \textit{them} because they are brothers, but rather they must serve, because those who benefit by their service are believers and dearly loved.
\verseWithHeading{False Teachers and the Love of Money} Teach and encourage these \textit{things}.
\verse If anyone teaches other doctrine and does not devote himself to the sound words of our Lord Jesus Christ and the teaching \textit{that is} in accordance with godliness,
\verse he is conceited, understanding nothing, but having a morbid interest concerning controversies and disputes about words, from which come envy, strife, slanders, evil suspicions,
\verse constant wrangling by people of depraved mind and deprived of the truth, who consider godliness to be a means of gain.
\verse But godliness with contentment is a great means of gain.
\verse For we have brought nothing into the world, so that neither can we bring anything out.
\verse But \textit{if we} have food and clothing, with these \textit{things} we will be content.
\verse But those who want to be rich fall into temptation and a trap and many foolish and harmful desires, which plunge those people into ruin and destruction.
\verseWithHeading{Renewed Charge to Timothy} But you, O man of God, flee from these \textit{things}, and pursue righteousness, godliness, faith, love, patient endurance, gentleness.
\verse Fight the good fight of the faith; take hold of the eternal life to which you were called, and confessed the good confession in the presence of many witnesses.
\verse I command you, in the sight of God who gives life to all \textit{things} and Christ Jesus who testified the good confession before Pontius Pilate,
\verse \textit{that} you observe the commandment without fault, irreproachable until the appearing of our Lord Jesus Christ,
\verse which he will make known in his own time, the blessed and only Sovereign, the King of those who reign as kings and Lord of those who rule as lords,
\verse the \textit{one who} alone possesses immortality, who lives \textit{in} unapproachable light, whom no human being has seen nor is able to see, to whom \textit{be} honor and eternal power. Amen.
\verseWithHeading{Instructions to the Rich} Command those \textit{who are} rich in this present age not to be proud and not to put their hope in the uncertainty of riches, but in God, who provides us all \textit{things} richly for enjoyment,
\verse to do good, to be rich in good works, to be generous, sharing freely,
\verse storing up for themselves a good foundation for the future, in order that they may take hold of \textit{what is} truly life.
\verseWithHeading{Final Charge and Benediction} O Timothy, guard what has been entrusted \textit{to you}. Turn away from pointless empty talk and contradictions of \textit{what is} falsely called knowledge,
\verse which some, \textit{by} professing \textit{it}, have deviated concerning the faith. Grace \textit{be} with you \textit{all}.
\end{biblechapter}

