\biblebook{John}

\begin{biblechapter} % John 1
\verseWithHeading{The Prologue to John’s Gospel} In the beginning was the Word, and the Word was with God, and the Word was God.
\verse This one was in the beginning with God.
\verse All \textit{things} came into being through him, and apart from him not one \textit{thing} came into being that has come into being.
\verse In him was life, and the life was the light of humanity.
\verse And the light shines in the darkness, and the darkness did not overcome it.
\verse A man came, sent from God, \textit{whose name was} John.
\verse This one came for a witness, in order that he could testify about the light, so that all would believe through him.
\verse That one was not the light, but \textit{came} in order that he could testify about the light.
\verse The true light, who gives light to every person, was coming into the world.
\verse He was in the world, and the world came into being through him, and the world did not recognize him.
\verse He came to his own \textit{things}, and his own \textit{people} did not receive him.
\verse But as many as received him—to those who believe in his name—he gave to them authority to become children of God,
\verse who were born not of blood, nor of the will of the flesh, nor of the will of a husband, but of God.
\verse And the Word became flesh and took up residence among us, and we saw his glory, glory as of the one and only from the Father, full of grace and truth.
\verse John testified about him and cried out, saying, “This one was \textit{he about} whom I said, ‘The one who comes after me is ahead of me, because he existed before me.’ ”
\verse For from his fullness we have all received, and grace after grace.
\verse For the law was given through Moses; grace and truth came about through Jesus Christ.
\verse No one has seen God at any time; the one and only, God, the one who is in the bosom of the Father—that one has made \textit{him} known.
\verseWithHeading{John the Baptist Testifies to Jesus} And this is the testimony of John, when the Jews sent priests and Levites from Jerusalem so that they could ask him, “Who are you?”
\verse And he confessed—and he did not deny, and confessed—“I am not the Christ!”
\verse And they asked him, “Then who \textit{are} you? Are you Elijah?” And he said, “I am not!” “Are you the Prophet?” And he answered, “No!”
\verse Then they said to him, “Who are you, so that we can give an answer to those who sent us? What do you say about yourself?”
\verse He said,
\verse just as Isaiah the prophet said.”
\verse (And they had been sent from the Pharisees.)
\verse John answered them, saying, “I baptize with water. In your midst stands \textit{one} whom you do not know—
\verse the one who comes after me, of whom I am not worthy \textit{to untie} the strap of his sandal!”
\verse These \textit{things} took place in Bethany on the other side of the Jordan, where John was baptizing.
\verse On the next day he saw Jesus coming to him and said, “Look! The Lamb of God who takes away the sin of the world!
\verse This one is \textit{the one} about whom I said, ‘After me is coming a man who is ahead of me, because he existed before me.’
\verse And I did not know him, but in order that he could be revealed to Israel, because of this I came baptizing with water.”
\verse And John testified, saying, “I have seen the Spirit descending like a dove from heaven and remaining upon him.
\verse And I did not know him, but the one who sent me to baptize with water, that one said to me, ‘\textit{The one} upon whom you see the Spirit descending and remaining upon him—this one is the one who baptizes with the Holy Spirit.’
\verse And I have seen and testify that this one is the Chosen One of God.
\verseWithHeading{Two of John’s Disciples Follow Jesus} On the next day again John was standing \textit{there}, and two of his disciples,
\verse and looking at Jesus \textit{as he} was walking by, he said, “Look! The Lamb of God!”
\verse And the two disciples heard him speaking, and they followed Jesus.
\verse And Jesus, turning around and seeing them following \textit{him}, said to them, “What do you seek?” And they said to him, “Rabbi” (which means \textit{when} translated “Teacher”), “where are you staying?”
\verse He said to them, “Come and you will see!” So they came and saw where he was staying, and they stayed with him that day (it was about the tenth hour).
\verseWithHeading{Andrew Declares Jesus to be the Messiah} Andrew, the brother of Simon Peter, was one of the two who heard John and followed him.
\verse This one first found his own brother Simon and said to him, “We have found the Messiah!” (which is translated “Christ”).
\verse He brought him to Jesus. Looking at him, Jesus said, “You are Simon the son of John. You will be called Cephas” (which is interpreted “Peter”).
\verseWithHeading{Jesus Calls Philip and Nathanael} On the next day he wanted to depart for Galilee, and he found Philip. And Jesus said to him, “Follow me!”
\verse (Now Philip was from Bethsaida, the town of Andrew and Peter.)
\verse Philip found Nathanael and said to him, “We have found \textit{the one} whom Moses wrote \textit{about} in the law, and the prophets wrote \textit{about}—Jesus son of Joseph from Nazareth!”
\verse And Nathanael said to him, “Can anything good come out of Nazareth?” Philip said to him, “Come and see!”
\verse Jesus saw Nathanael coming toward him and said about him, “Look! \textit{A true Israelite} in whom is no deceit!”
\verse Nathanael said to him, “From where do you know me?” Jesus answered and said to him, “Before Philip called you, \textit{when you} were under the fig tree, I saw you.”
\verse Nathanael answered him, “Rabbi, you are the Son of God! You are the king of Israel!”
\verse Jesus answered and said to him, “Because I said to you that I saw you under the fig tree, do you believe? You will see greater \textit{things} than these!”
\verse And he said to him, “Truly, truly I say to all of you, you will see heaven opened and the angels of God ascending and descending upon the Son of Man.”
\end{biblechapter}

\begin{biblechapter} % John 2
\verseWithHeading{The Wedding at Cana: Water Turned into Wine} And on the third day, there was a wedding at Cana in Galilee, and the mother of Jesus was there.
\verse And both Jesus and his disciples were invited to the wedding.
\verse And \textit{when the} wine ran out, the mother of Jesus said to him, “They have no wine!”
\verse And Jesus said to her, “What \textit{does your concern have to do with me}, woman? My hour has not yet come.”
\verse His mother said to the servants, “Whatever he says to you, do \textit{it}!”
\verse Now six stone water jars were set there, in accordance with the ceremonial cleansing of the Jews, each holding two or three measures.
\verse Jesus said to them, “Fill the water jars with water.” And they filled them to the brim.
\verse And he said to them, “Now draw \textit{some} out and take \textit{it} to the head steward. So they took \textit{it}.
\verse Now when the head steward tasted the water which had become wine and did not know where it was from—but the servants who had drawn the water knew—the head steward summoned the bridegroom
\verse and said to him, “\textit{Everyone} serves the good wine first, and whenever they are drunk, the inferior. You have kept the good wine until now!”
\verse This beginning of signs Jesus performed at Cana in Galilee, and revealed his glory, and his disciples believed in him.
\verseWithHeading{Jesus’ First Journey to Jerusalem} After this he went down to Capernaum, and his mother and brothers and his disciples, and they stayed there \textit{a few} days.
\verse And the Passover of the Jews was near, and Jesus went up to Jerusalem.
\verseWithHeading{The Cleansing of the Temple} And he found in the temple \textit{courts} those who were selling oxen and sheep and doves, and the money changers seated.
\verse And he made a whip of cords \textit{and} drove \textit{them} all out of the temple \textit{courts}, both the sheep and the oxen, and he poured out the coins of the money changers and overturned their tables.
\verse And to the ones selling the doves he said, “Take these \textit{things} away from here! Do not make my Father’s house \textit{a marketplace}!”
\verse His disciples remembered that it is written, “Zeal for your house will consume me.”
\verse So the Jews answered and said to him, “What sign do you show to us, because you are doing these \textit{things}?”
\verse Jesus answered and said to them, “Destroy this temple, and in three days I will raise it up!”
\verse Then the Jews said, “This temple has been under construction forty-six years, and will you raise it up in three days?”
\verse But he was speaking about the temple of his body.
\verse So when he was raised from the dead, his disciples remembered that he had said this, and they believed the scripture and the saying that Jesus had spoken.
\verseWithHeading{Jesus at the Passover} Now while he was in Jerusalem at the Passover, during the feast, many believed in his name \textit{because they} saw his signs which he was doing.
\verse But Jesus himself did not entrust himself to them, because he knew all \textit{people},
\verse and because he did not \textit{need} anyone \textit{to testify} about man, for he himself knew what was in man.
\end{biblechapter}

\begin{biblechapter} % John 3
\verseWithHeading{A Meeting with Nicodemus} Now there was a man of the Pharisees \textit{whose name was} Nicodemus, a ruler of the Jews.
\verse This man came to him at night and said to him, “Rabbi, we know that \textit{you are} a teacher who has come from God, for no one is able to perform these signs that you are performing unless God were with him.”
\verse Jesus answered and said to him, “Truly, truly I say to you, unless someone is born from above, he is not able to see the kingdom of God.”
\verse Nicodemus said to him, “How can a man be born \textit{when he} is an old man? He is not able to enter into his mother’s womb for the second time and be born, \textit{can he}?”
\verse Jesus answered, “Truly, truly I say to you, unless someone is born of water and spirit, he is not able to enter into the kingdom of God.
\verse What is born of the flesh is flesh, and what is born of the Spirit is spirit.
\verse Do not be astonished that I said to you, ‘It is necessary for you to be born from above.’
\verse The wind blows wherever it wishes, and you hear the sound of it, but you do not know where it comes from and where it is going. So is everyone who is born of the Spirit.”
\verse Nicodemus answered and said to him, “How can these \textit{things} be?”
\verse Jesus answered and said to him, “Are you the teacher of Israel, and you do not understand these \textit{things}?
\verse Truly, truly I say to you, we speak what we know, and we testify \textit{about} what we have seen, and you do not accept our testimony!
\verse If I tell you earthly things and you do not believe, how will you believe if I tell you heavenly things?
\verse And no one has ascended into heaven except the one who descended from heaven—the Son of Man.
\verse And just as Moses lifted up the snake in the wilderness, thus it is necessary \textit{that} the Son of Man be lifted up,
\verse so that everyone who believes in him will have eternal life.”
\verseWithHeading{God’s Love for the World} For in this way God loved the world, so that he gave his one and only Son, in order that everyone who believes in him will not perish, but will have eternal life.
\verse For God did not send his Son into the world in order that he should judge the world, but in order that the world should be saved through him.
\verse The one who believes in him is not judged, but the one who does not believe has already been judged, because he has not believed in the name of the one and only Son of God.
\verse And this is the judgment: that the light has come into the world, and people loved the darkness rather than the light, because their deeds were evil.
\verse For everyone who practices evil hates the light and does not come to the light, lest his deeds be exposed.
\verse But the one who practices the truth comes to the light, in order that his deeds may be revealed, that they are done in God.
\verseWithHeading{Additional Testimony by John the Baptist About Jesus} After these \textit{things} Jesus and his disciples came into Judean territory, and there he spent time with them and was baptizing.
\verse Now John was also baptizing at Aenon near Salim, because water was plentiful there, and they were coming and were being baptized.
\verse (For John had not yet been thrown into prison.)
\verse So a dispute occurred on the part of John’s disciples with a Jew concerning purification.
\verse And they came to John and said to him, “Rabbi, he who was with you on the other side of the Jordan, about whom you testified—look, this one is baptizing, and all are coming to him!”
\verse John answered and said, “A man can receive not one \textit{thing} unless it is granted to him from heaven!
\verse You yourselves testify about me that I said, ‘I am not the Christ, but I am sent before that one.’
\verse The one who has the bride is the bridegroom. But the friend of the bridegroom, who stands and hears him, rejoices \textit{greatly} because of the bridegroom’s voice. So this joy of mine is complete.
\verse It is necessary for that one to increase, but \textit{for} me to decrease.”
\verse The one who comes from above is over all. The one who is from the earth is from the earth and speaks from the earth; the one who comes from heaven is over all.
\verse What he has seen and heard, this he testifies, and no one accepts his testimony.
\verse The one who accepts his testimony has attested that God is true.
\verse For \textit{the one} whom God sent speaks the words of God, for he does not give the Spirit by measure.
\verse The Father loves the Son and has given all \textit{things} into his hand.
\verse The one who believes in the Son has eternal life, but the one who disobeys the Son will not see life—but the wrath of God remains on him.
\end{biblechapter}

\begin{biblechapter} % John 4
\verseWithHeading{The Samaritan Woman at Jacob’s Well} Now when Jesus knew that the Pharisees had heard that Jesus was making and baptizing more disciples than John
\verse (although Jesus himself was not baptizing, but his disciples),
\verse he left Judea and departed again for Galilee.
\verse And it was necessary \textit{for} him to go through Samaria.
\verse Now he came to a town of Samaria called Sychar, near the piece of land that Jacob had given to his son Joseph.
\verse And Jacob’s well was there, so Jesus, \textit{because he} had become tired from the journey, simply sat down at the well. It was about the sixth hour.
\verse A woman of Samaria came to draw water. Jesus said to her, “Give me \textit{water} to drink.”
\verse (For his disciples had gone away into the town so that they could buy food.)
\verse So the Samaritan woman said to him, “How do you, being a Jew, ask from me \textit{water} to drink, \textit{since I} am a Samaritan woman?” (For Jews have no dealings with Samaritans.)
\verse Jesus answered and said to her, “If you had known the gift of God and who it is who says to you, ‘Give me \textit{water} to drink,’ you would have asked him, and he would have given you living water.”
\verse The woman said to him, “Sir, you have no bucket and the well is deep! From where then do you get this living water?
\verse You are not greater than our father Jacob, \textit{are you}, who gave us the well and drank from it himself, and his sons and his livestock?”
\verse Jesus answered and said to her, “Everyone who drinks of this water will be thirsty again.
\verse But whoever drinks of this water which I will give to him will never be thirsty for eternity, but the water which I will give to him will become in him a well of water springing up to eternal life.”
\verse The woman said to him, “Sir, give me this water, so that I will not be thirsty or come here to draw \textit{water}!”
\verse He said to her, “Go, call your husband and come here.”
\verse The woman answered and said to him, “I do not have a husband.” Jesus said to her, “You have said rightly, ‘I do not have a husband,’
\verse for you have had five husbands, and \textit{the one} whom you have now is not your husband; this you have said truthfully!”
\verse The woman said to him, “Sir, I see that you are a prophet.
\verse Our fathers worshiped on this mountain, and you \textit{people} say that in Jerusalem is the place where it is necessary to worship.”
\verse Jesus said to her, “Believe me, woman, that an hour is coming when neither on this mountain nor in Jerusalem will you worship the Father.
\verse You worship what you do not know. We worship what we know, because salvation is from the Jews.
\verse But an hour is coming—and now is \textit{here}—when the true worshipers will worship the Father in spirit and truth, for indeed the Father seeks such \textit{people} \textit{to be} his worshipers.
\verse God \textit{is} spirit, and the ones who worship him must worship in spirit and truth.”
\verse The woman said to him, “I know that Messiah is coming” (the one called Christ); “whenever that one comes, he will proclaim all \textit{things} to us.”
\verse Jesus said to her, “I, the one speaking to you, am \textit{he}.
\verseWithHeading{The Disciples and the Harvest} And at this \textit{point} his disciples came, and they were astonished that he was speaking with a woman. However, no one said, “What do you seek?” or “Why are you speaking with her?”
\verse So the woman left her water jar and went away into the town and said to the people,
\verse “Come, see a man who told me everything I have ever done! Perhaps this one is the Christ?”
\verse They went out from the town and were coming to him.
\verse In the meanwhile the disciples were asking him, saying, “Rabbi, eat \textit{something}!”
\verse But he said to them, “I have food to eat that you do not know about.”
\verse So the disciples began to say to one another, “No one brought him \textit{anything} to eat, \textit{did they}?”
\verse Jesus said to them, “My food is that I do the will of the one who sent me and complete his work.
\verse Do you not say, ‘There are yet four months and the harvest comes’? Behold, I say to you, lift up your eyes and look at the fields, that they are white for harvest already.
\verse The one who reaps receives wages and gathers fruit for eternal life, in order that the one who sows and the one who reaps can rejoice together.
\verse For in this \textit{instance} the saying is true, ‘It is one who sows and another who reaps.’
\verse I sent you to reap what you did not work for; others have worked, and you have entered into their work.”
\verseWithHeading{The Samaritans and the Savior of the World} Now from that town many of the Samaritans believed in him because of the word of the woman who testified, “He told me everything that I have done.”
\verse So when the Samaritans came to him, they began asking him to stay with them. And he stayed there two days.
\verse And many more believed because of his word,
\verse And they were saying to the woman, “No longer because of \textit{what you said} do we believe, for we ourselves have heard, and we know that this one is truly the Savior of the world!”
\verseWithHeading{Return to Galilee} And after the two days he departed from there into Galilee.
\verse For Jesus himself testified that a prophet has no honor in his own homeland.
\verse So when he came to Galilee, the Galileans welcomed him, \textit{because they} had seen all \textit{the things} he had done in Jerusalem at the feast (for they themselves had also come to the feast).
\verseWithHeading{A Royal Official’s Son Is Healed} Now he came again to Cana in Galilee, where he had made the water wine. And \textit{there} was at Capernaum a certain royal official whose son was sick.
\verse This man, \textit{when he} heard that Jesus had come from Judea into Galilee, went to him and asked that he come down and heal his son, for he was about to die.
\verse So Jesus said to him, “Unless you \textit{people} see signs and wonders, you will never believe!”
\verse The royal official said to him, “Sir, come down before my child dies!”
\verse Jesus said to him, “Go, your son will live.” The man believed the word that Jesus spoke to him, and he departed.
\verse Now \textit{as} he was going down, his slaves met him, saying that his child was alive.
\verse So he inquired from them the hour at which he had gotten better. Then they said to him, “Yesterday at the seventh hour the fever left him.”
\verse So the father knew that \textit{it was} that same hour at which Jesus said to him, “Your son will live,” and he himself believed, and his whole household.
\verse Now this \textit{is} again a second sign Jesus performed \textit{when he} came from Judea into Galilee.
\end{biblechapter}

\begin{biblechapter} % John 5
\verseWithHeading{A Paralytic Is Healed} After these \textit{things} \textit{there} was a feast of the Jews, and Jesus went up to Jerusalem.
\verse Now there is in Jerusalem near the Sheep Gate a pool called in Aramaic \textit{Bethzatha}, which has five porticoes.
\verse In these were lying a large number of those who were sick, blind, lame, paralyzed.
\verse And a certain man was there who had \textit{been} thirty-eight years in his sickness.
\verse Jesus, \textit{when he} saw this one lying \textit{there} and knew that he had \textit{been sick} a long time already, said to him, “Do you want to become well?”
\verse The one who was sick answered him, “Sir, I do not have anyone that, whenever the water is stirred up, could put me into the pool. But \textit{while} I am coming, another goes down before me.”
\verse Jesus said to him, “Get up! Pick up your mat and walk!”
\verse And immediately the man became well and picked up his mat and began to walk. (Now it was the Sabbath on that day.)
\verse So the Jews were saying to the one who had been healed, “It is the Sabbath, and it is not permitted for you to pick up the mat!”
\verse But he answered them, “The one who made me well—that one said to me, ‘Pick up your mat and walk!’ ”
\verse So they asked him, “Who is the man who said to you, ‘Pick up \textit{your mat} and walk?’ ”
\verse But the one who was healed did not know who it was, for Jesus had withdrawn \textit{while} a crowd was in the place.
\verseWithHeading{Equal with God} After these \textit{things} Jesus found him at the temple and said to him, “Look, you have become well! Sin no longer, lest something worse happen to you.”
\verse The man went and reported to the Jews that Jesus was the one who made him well.
\verse And on account of this the Jews began to persecute Jesus, because he was doing these \textit{things} on the Sabbath.
\verse But he answered them, “My Father is working until now, and I am working.”
\verse So on account of this the Jews were seeking even more to kill him, because he not only was breaking the Sabbath, but also was calling God his own Father, \textit{thus} making himself equal with God.
\verseWithHeading{The Authority of the Son} So Jesus answered and said to them, “Truly, truly I say to you, the Son can do nothing from himself except what he sees the Father doing. For whatever that one does, these \textit{things} also the Son does likewise.
\verse For the Father loves the Son and shows him everything that he himself is doing. And greater works than these he will show him, so that you will be astonished.
\verse For just as the Father raises the dead and makes \textit{them} alive, thus also the Son makes alive whomever he wishes.
\verse For the Father does not judge anyone, but he has given all judgment to the Son,
\verse in order that all \textit{people} will honor the Son, just as they honor the Father. The one who does not honor the Son does not honor the Father who sent him.
\verse Truly, truly I say to you that the one who hears my word and who believes the one who sent me has eternal life, and does not come into judgment, but has passed from death into life.
\verse “Truly, truly I say to you, that an hour is coming—and now is \textit{here}—when the dead will hear the voice of the Son of God, and the ones who hear will live.
\verse For just as the Father has life in himself, thus also he has granted to the Son to have life in himself.
\verse And he has granted him authority to carry out judgment, because he is the Son of Man.
\verse “Do not be astonished \textit{at} this, because an hour is coming in which all those in the tombs will hear his voice
\verse and they will come out—those who have done good \textit{things} to a resurrection of life, but those who have practiced evil \textit{things} to a resurrection of judgment.
\verse I am able to do nothing from myself. Just as I hear, I judge, and my judgment is just, because I do not seek my \textit{own} will, but the will of the one who sent me.
\verseWithHeading{Further Testimony About the Son} “If I testify about myself, my testimony is not true.
\verse There is another who testifies about me, and I know that the testimony which he testifies about me is true.
\verse You have sent to John and he has testified to the truth.
\verse (And I do not receive testimony from people, but I say these \textit{things} in order that you may be saved.)
\verse That one was the lamp \textit{which was} burning and shining, and you wanted to rejoice for an hour in his light.
\verse “But I have a testimony greater than John’s, for the works which the Father has given to me that I should complete them—the very works which I am doing—\textit{these} testify about me, that the Father has sent me.
\verse And the Father who sent me, that one has testified about me. You have neither heard his voice at any time nor seen his form.
\verse And you do not have his word residing in yourselves, because the one whom that one sent, in this one you do not believe.
\verse You search the scriptures because you think that you have eternal life in them, and it is these that testify about me.
\verse And you are not willing to come to me so that you may have life.
\verse “I do not accept glory from people,
\verse but I know you, that you do not have the love of God in yourselves.
\verse I have come in my Father’s name, and you do not accept me. If another should come in his own name, you would accept that one!
\verse How are you able to believe, \textit{if you} accept glory from one another, and do not seek the glory \textit{which is} from the only God?
\verse Do not think that I will accuse you before the Father! The one who accuses you is Moses, in whom you have put your hope!
\verse For if you had believed Moses, you would believe me, for that one wrote about me.
\verse But if you do not believe that one’s writings, how will you believe my words?”
\end{biblechapter}

\begin{biblechapter} % John 6
\verseWithHeading{The Feeding of Five Thousand} After these \textit{things} Jesus went away to the other side of the sea of Galilee (that is, Tiberias).
\verse And a large crowd was following him because they were observing the signs that he was doing on those who were sick.
\verse So Jesus went up on the mountain and sat down there with his disciples.
\verse (Now the Passover, the feast of the Jews, was near.)
\verse \textit{Then Jesus, when he looked up}  and saw that a large crowd was coming to him, said to Philip, “Where can we buy bread so that these \textit{people} can eat?”
\verse (Now he said this to test him, because he knew what he was going to do.)
\verse Philip replied to him, “Two hundred denarii \textit{worth of} bread would not be enough for them, in order that each one could receive a little.”
\verse One of his disciples, Andrew the brother of Simon Peter, said to him,
\verse “Here is a boy who has five barley loaves and two fish, but what are these for so many \textit{people}?”
\verse Jesus said, “Make the people recline.” (Now \textit{there} was a lot of grass in the place.) So the men reclined, approximately five thousand \textit{in} number.
\verse Then Jesus took the bread, and \textit{after he} had given thanks, he distributed \textit{it} to those who were reclining—likewise also of the fish, as much as they wanted.
\verse And when they were satisfied, he said to his disciples, “Gather the remaining fragments so that nothing is lost.”
\verse So they gathered \textit{them}, and filled twelve baskets with fragments from the five barley loaves which were left over by those who had eaten.
\verse Now \textit{when} the people saw the sign that he performed, they began to say, “This one is truly the Prophet who is to come into the world!”
\verse Then Jesus, \textit{because he} knew that they were about to come and seize him in order to make \textit{him} king, withdrew again up the mountain \textit{by} himself alone.
\verseWithHeading{Jesus Walks on the Water} Now when evening came, his disciples went down to the sea.
\verse And getting into a boat, they began to go to the other side of the sea, to Capernaum. And it had already become dark, and Jesus had not yet come to them.
\verse And the sea began to be stirred up \textit{because} a strong wind was blowing.
\verse Then \textit{when they} had rowed about twenty-five or thirty stadia, they saw Jesus walking on the sea and coming near the boat, and they were afraid.
\verse But he said to them, “\textit{It} is I! Do not be afraid!”
\verse So they were wanting to take him into the boat, and immediately the boat came to the land to which they were going.
\verseWithHeading{Discourse About the Bread of Life} On the next day, the crowd that was on the other side of the sea saw that other boats were not there (except one), and that Jesus had not entered with his disciples into the boat, but his disciples had departed alone.
\verse Other boats from Tiberias came near the place where they had eaten the bread \textit{after} the Lord had given thanks.
\verse So when the crowd saw that Jesus was not there, nor his disciples, they themselves got into the boats and came to Capernaum seeking Jesus.
\verse And \textit{when they} found him on the other side of the sea, they said to him, “Rabbi, when did you get here?”
\verse Jesus replied to them and said, “Truly, truly I say to you, you seek me not because you saw signs, but because you ate of the loaves and were satisfied!
\verse Do not work for the food that perishes, but the food that remains to eternal life, which the Son of Man will give to you. For God the Father has set his seal on this one.”
\verse So they said to him, “What shall we do that we can accomplish the works of God?”
\verse Jesus answered and said to them, “This is the work of God: that you believe in \textit{the one} whom that one sent.”
\verse So they said to him, “Then what sign will you perform, so that we can see \textit{it} and believe you? What will you do?
\verse Our fathers ate the manna in the wilderness, just as it is written, ‘He gave them bread from heaven to eat.’
\verse Then Jesus said to them, “Truly, truly I say to you, Moses did not give you bread from heaven, but my Father is giving you the true bread from heaven!
\verse For the bread of God is the one who comes down from heaven and gives life to the world.”
\verse So they said to him, “Sir, always give us this bread!”
\verse Jesus said to them, “I am the bread of life. The one who comes to me will never be hungry, and the one who believes in me will never be thirsty again.
\verse But I said to you that you have seen me and do not believe.
\verse Everyone whom the Father gives to me will come to me, and the one who comes to me I will never throw out,
\verse because I have come down from heaven not that I should do my will, but the will of the one who sent me.
\verse Now this is the will of the one who sent me: that everyone whom he has given me, I would not lose \textit{any} of them, but raise them up on the last day.
\verse For this is the will of my Father, that everyone who looks at the Son and believes in him would have eternal life, and I will raise him up on the last day.”
\verse Now the Jews began to grumble about him because he said, “I am the bread that came down from heaven,”
\verse and they were saying, “Is this one not Jesus the son of Joseph, whose father and mother we know? How does he now say, ‘I have come down from heaven’?”
\verse Jesus answered and said to them, “Do not grumble \textit{among yourselves}!
\verse No one is able to come to me unless the Father who sent me draws him, and I will raise him up on the last day.
\verse It is written in the prophets, ‘And they will all be taught by God.’ Everyone who hears from the Father and learns comes to me.
\verse (Not that anyone has seen the Father except the one who is from God—this one has seen the Father.)
\verse Truly, truly I say to you, the one who believes has eternal life.
\verse I am the bread of life.
\verse Your fathers ate the manna in the wilderness and they died.
\verse This is the bread that comes down from heaven so that someone may eat from it and not die.
\verse I am the living bread that came down from heaven. If anyone eats from this bread, he will live \textit{forever}. And the bread that I will give for the life of the world is my flesh.”
\verse So the Jews began to quarrel \textit{among themselves}, saying, “How can this man give us his flesh to eat?”
\verse Then Jesus said to them, “Truly, truly I say to you, unless you eat the flesh of the Son of Man and drink his blood, you do not have life in yourselves!
\verse The one who eats my flesh and drinks my blood has eternal life, and I will raise him up on the last day.
\verse For my flesh is true food, and my blood is true drink.
\verse The one who eats my flesh and drinks my blood resides in me and I in him.
\verse Just as the living Father sent me, and I live because of the Father, \textit{so} also the one who eats me—that one will live because of me.
\verse This is the bread that came down from heaven, not as the fathers ate and died. The one who eats this bread will live \textit{forever}.”
\verseWithHeading{Many of Jesus’ Disciples Offended by His Teaching} He said these \textit{things} \textit{while} teaching in the synagogue in Capernaum.
\verse Thus many of his disciples, \textit{when they} heard \textit{it}, said, “This saying is hard! Who can understand it?”
\verse But Jesus, \textit{because he} knew within himself that his disciples were grumbling about this, said to them, “Does this cause you to be offended?
\verse Then \textit{what} if you see the Son of Man ascending where he was before?
\verse The Spirit is the one who gives life; the flesh profits nothing. The words that I have spoken to you are spirit and are life.
\verse But there are some of you who do not believe.” (For Jesus knew from the beginning who they were who did not believe, and who it was who would betray him.)
\verse And he said, “Because of this I said to you that no one can come to me unless it has been granted to him by the Father.”
\verseWithHeading{Peter’s Confession} For this \textit{reason} many of his disciples \textit{drew back} and were not walking with him any longer.
\verse So Jesus said to the twelve, “You do not want to go away also, \textit{do you}?”
\verse Simon Peter answered him, “Lord, to whom would we go? You have the words of eternal life.
\verse And we have believed, and have come to know, that you are the Holy One of God.”
\verse Jesus replied to them, “Did I not choose you, the twelve, and one of you is the devil?”
\verse (Now he was speaking about Judas \textit{son} of Simon Iscariot, because this one—one of the twelve—was going to betray him.)
\end{biblechapter}

\begin{biblechapter} % John 7
\verseWithHeading{Jesus’ Brothers Do Not Believe in Him} And after these \textit{things} Jesus was going about in Galilee. For he did not want to go about in Judea, because the Jews were seeking to kill him.
\verse Now the feast of the Jews—the \textit{feast of} Tabernacles—was near.
\verse So his brothers said to him, “Depart from here and go to Judea, so that your disciples also can see your works that you are doing.
\verse For no one does anything in secret and \textit{yet} he himself desires to be \textit{publicly recognized}. If you are doing these \textit{things}, reveal yourself to the world!”
\verse (For not even his brothers believed in him.)
\verseWithHeading{Jesus at the Feast of Tabernacles} So Jesus said to them, “My time has not yet come, but your time is always ready.
\verse The world cannot hate you, but it hates me, because I am testifying about it, that its deeds are evil.
\verse You go up to the feast. I am not going up to this feast, because my time is not yet completed.
\verse And \textit{when he} had said these \textit{things}, he remained in Galilee.
\verse But when his brothers had gone up to the feast, then he also went up, not openly, but (as it were) in secret.
\verse So the Jews were looking for him at the feast, and were saying, “Where is he?”
\verse And there was a lot of grumbling concerning him among the crowds; some were saying, “He is a good \textit{man},” but others were saying, “No, but he deceives the crowd.”
\verse However, no one was speaking openly about him for fear of the Jews.
\verse \textit{Now when the feast was already half over}, Jesus went to the temple \textit{courts} and began to teach.
\verse Then the Jews were astonished, saying, “How does this man \textit{possess knowledge}, \textit{because he} has not been taught?”
\verse So Jesus answered them and said, “My teaching is not mine, but \textit{is} from the one who sent me.
\verse If anyone wants to do his will, he will know about my teaching, whether it is from God or I am speaking from myself.
\verse The one who speaks from himself seeks his own glory. But the one who seeks the glory of the one who sent him—this one is true, and there is no unrighteousness in him.
\verse Has not Moses given you the law, and none of you carries out the law? Why do you seek to kill me?”
\verse The crowd replied, “You have a demon! Who is seeking to kill you?”
\verse Jesus answered and said to them, “I performed one work, and you are all astonished.
\verse Because of this Moses has given you circumcision (not that it is from Moses, but from the fathers), and you circumcise a man on the Sabbath.
\verse If a man receives circumcision on the Sabbath so that the law of Moses would not be broken, are you angry with me because I made a whole man well on the Sabbath?
\verse Do not judge according to outward appearance, but judge according to righteous judgment!”
\verseWithHeading{Is Jesus the Christ?} Then some of the inhabitants of Jerusalem began to say, “Is this not \textit{the one} whom they are seeking to kill?
\verse And behold, he is speaking openly and they are saying nothing to him! Can it be that the rulers truly know that this man is the Christ?
\verse Yet we know where this man is from, but the Christ, whenever he comes—no one knows where he is from!”
\verse Then Jesus cried out in the temple \textit{courts}, teaching and saying, “You both know me and you know where I am from! And I have not come from myself, but the one who sent me is true, whom you do not know.
\verse I know him, because I am from him and he sent me.”
\verse So they were seeking to seize him, and no one laid a hand on him, because his hour had not yet come.
\verse But from the crowd many believed in him and were saying, “Whenever the Christ comes, he will not perform more signs than this man has done, \textit{will he}?”
\verse The Pharisees heard the crowd murmuring these \textit{things} about him, and the chief priests and the Pharisees sent officers in order \textit{to take him into custody }.
\verse Then Jesus said, “Yet a little time I am with you, and I am going to the one who sent me.
\verse You will seek me and will not find \textit{me}, and where I am, you cannot come.”
\verse So the Jews said to one another, “Where \textit{is} this one going to go, that we will not find him? He is not going to go to the Dispersion among the Greeks and teach the Greeks, \textit{is he}?
\verse What is this saying that he said, ‘You will seek me and will not find \textit{me}, and where I am, you cannot come’?”
\verseWithHeading{The Promise of the Spirit} Now on the last day of the feast—the great \textit{day}—Jesus stood and cried out, saying, “If anyone is thirsty, let him come to me, and let him drink,
\verse the one who believes in me. Just as the scripture said, ‘Out of his belly will flow rivers of living water.’ ”
\verse Now he said this concerning the Spirit, whom those who believed in him were about to receive. For the Spirit was not yet \textit{given}, because Jesus had not yet been glorified.)
\verseWithHeading{Different Opinions About Jesus} Then, \textit{when they} heard these words, \textit{some} from the crowd began to say, “This man is truly the Prophet!”
\verse Others were saying, “This man is the Christ!” But others were saying, “No, for the Christ does not come from Galilee, \textit{does he}?
\verse Has not the scripture said that the Christ comes from the descendants of David, and from Bethlehem, the village where David was?”
\verse So there was a division in the crowd because of him.
\verse And some of them were wanting to seize him, but no one laid hands on him.
\verse So the officers came to the chief priests and Pharisees. And they said to them, “\textit{Why} did you not bring him?”
\verse The officers replied, “Never has a man spoken like this!”
\verse Then the Pharisees replied to them, “You have not also been deceived, \textit{have you}?
\verse \textit{None} of the rulers or of the Pharisees have believed in him, \textit{have they}?
\verse But this crowd who does not know the law is accursed!”
\verse Nicodemus, the one who came to him previously—who was one of them—said to them,
\verse “Our law does not condemn a man unless it first hears from him and knows what he is doing, \textit{does it}?”
\verse They answered and said to him, “You are not also from Galilee, \textit{are you}? Investigate and see that a prophet does not arise from Galilee!” 〚
\verse And each one went to his \textit{own} house.
\end{biblechapter}

\begin{biblechapter} % John 8
\verseWithHeading{A Woman Caught in Adultery} But Jesus went to the Mount of Olives.
\verse Now early in the morning he came again to the temple \textit{courts}. And all the people were coming, and he sat down \textit{and} began to teach them.
\verse Now the scribes and the Pharisees brought to him a woman caught in adultery. And standing her in \textit{their} midst,
\verse they said to him, testing \textit{him}, “Teacher, this woman was caught in the very act of committing adultery!
\verse Now in the law, Moses commanded us to stone such women. So what do you say?”
\verse (Now they were saying this to test him, so that they would have \textit{an occasion} to bring charges against him.) But Jesus, bending down, began to write with \textit{his} finger on the ground, taking no notice.
\verse And when they persisted in asking him, straightening up he said to them, “The \textit{one} of you without sin, let him throw the first stone at her!”
\verse And bending down again, he wrote on the ground.
\verse Now \textit{when they} heard \textit{it}, being convicted by their conscience, they began to depart, one by one, beginning with the older ones, and Jesus was left alone—and the woman who was in \textit{their} midst.
\verse So Jesus, straightening up and seeing no one except the woman, said to her, “Where are those accusers of yours? Does no one condemn you?”
\verse And she said, “No one, Lord.” So Jesus said, “Neither do I condemn you. Go, and sin no more.”〛
\verseWithHeading{Jesus, the Light of the World} Then Jesus spoke to them again, saying, “I am the light of the world! The one who follows me will never walk in darkness, but will have the light of life.”
\verse So the Pharisees said to him, “You testify concerning yourself! Your testimony is not true.”
\verse Jesus answered and said to them, “Even if I testify concerning myself, my testimony is true, because I know where I have come from and where I am going. But you do not know where I have come from or where I am going.
\verse You judge according to externals; I do not judge anyone.
\verse But even if I judge, my judgment is true, because I am not alone, but I and the Father who sent me.
\verse And even in your law it is written that the testimony of two men is true.
\verse I am the one who testifies concerning myself, and the Father who sent me testifies concerning me.”
\verse So they were saying to him, “Where is your father?” Jesus replied, “You know neither me nor my Father! If you had known me, you would have known my Father also.”
\verse He spoke these words by the treasury \textit{while} teaching in the temple \textit{courts}, and no one seized him, because his hour had not yet come.
\verseWithHeading{Jesus Predicts His Death} So he said to them again, “I am going away, and you will seek me and will die in your sin. Where I am going you cannot come!”
\verse Then the Jews began to say, “Perhaps he will kill himself, because he is saying, ‘Where I am going you cannot come.’ ”
\verse And he said to them, “You are from below; I am from above. You are from this world; I am not from this world.
\verse Thus I said to you that you will die in your sins. For if you do not believe that I am \textit{he}, you will die in your sins.”
\verse So they began to say to him, “Who are you?” Jesus said to them, “\textit{What} I have been saying to you \textit{from} the beginning.
\verse I have many \textit{things} to say and to judge concerning you, but the one who sent me is true, and \textit{the things} which I heard from him, these \textit{things} I say to the world.”
\verse (They did not know that he was speaking to them about the Father.)
\verse Then Jesus said, “When you lift up the Son of Man, then you will recognize that I am \textit{he}, and I do nothing from myself, but just as the Father taught me, I say these \textit{things}.
\verse And the one who sent me is with me. He has not left me alone, because I always do the things that are pleasing to him.”
\verse \textit{While} he was saying these \textit{things}, many believed in him.
\verseWithHeading{The Truth Will Set You Free} Then Jesus said to those Jews who had believed him, “If you continue in my word you are truly my disciples,
\verse and you will know the truth, and the truth will set you free.”
\verse They replied to him, “We are descendants of Abraham and have not been enslaved to anyone at any time. How do you say, ‘You will become free’?”
\verse Jesus replied to them, “Truly, truly I say to you, that everyone who commits sin is a slave of sin.
\verse And the slave does not remain in the household \textit{forever}; the son remains \textit{forever}.
\verse So if the son sets you free, you will be truly free.
\verse I know that you are descendants of Abraham. But you are seeking to kill me, because my word makes no progress among you.
\verse I speak \textit{the things} that I have seen with the Father; so also you do \textit{the things} that you have heard from the Father.”
\verseWithHeading{The Priority of Jesus Over Abraham} They answered and said to him, “Abraham is our father!” Jesus said to them, “If you are children of Abraham, do the deeds of Abraham!
\verse But now you are seeking to kill me, a man who spoke to you the truth which I heard from God. This Abraham did not do.
\verse You are doing the deeds of your father!”
\verse They said to him, “We were not born from sexual immorality! We have one father, God!”
\verse Jesus said to them, “If God were your father, you would love me, for I have come forth from God and have come. For I have not come from myself, but that one sent me.
\verse \textit{Why} do you not understand my way of speaking? Because you are not able to listen to my message.
\verse You are of your father the devil, and you want to do the desires of your father! That one was a murderer from the beginning, and does not stand firm in the truth, because truth is not in him. Whenever he speaks the lie, he speaks from his own \textit{nature}, because he is a liar and the father \textit{of lies}.
\verse But because I am telling the truth, you do not believe me.
\verse Who among you convicts me concerning sin? If I am telling the truth, \textit{why} do you not believe me?
\verse The Jews answered and said to him, “Do we not correctly say that you are a Samaritan and have a demon?”
\verse Jesus replied, “I do not have a demon, but I honor my Father, and you dishonor me!
\verse But I do not seek my \textit{own} glory. There is one who seeks and judges!
\verse Truly, truly I say to you, if anyone keeps my word, he will never experience death \textit{forever}.”
\verse The Jews said to him, “Now we know that you have a demon! Abraham and the prophets died, and you say, ‘If anyone keeps my word, he will never taste death \textit{forever}.’
\verse You are not greater than our father Abraham who died, \textit{are you}? And the prophets died! Who do you make yourself \textit{to be}?”
\verse Jesus replied, “If I glorify myself, my glory is nothing. The one who glorifies me is my Father, \textit{about} whom you say, ‘He is our God.’
\verse And you have not known him, but I know him. And if I were to say that I do not know him, I would be a liar like you! But I know him and I keep his word.
\verse Abraham your father rejoiced that he would see my day, and he saw \textit{it} and was glad.”
\verse So the Jews said to him, “\textit{You are} not yet fifty years \textit{old}, and have you seen Abraham?”
\verse Jesus said to them, “Truly, truly I say to you, before Abraham was, I am!”
\verse Then they picked up stones in order to throw \textit{them} at him. But Jesus was hidden and went out of the temple \textit{courts}.
\end{biblechapter}

\begin{biblechapter} % John 9
\verseWithHeading{A Man Born Blind Is Given Sight} And \textit{as he} went away, he saw a man blind from birth.
\verse And his disciples asked him, saying, “Rabbi, who sinned, this man or his parents, that he should be born blind?”
\verse Jesus replied, “Neither this man sinned nor his parents, but \textit{it happened} so that the works of God could be revealed in him.
\verse It is necessary \textit{for} us to do the deeds of the one who sent me while it is day; night is coming, when no one can work!
\verse While I am in the world, I am the light of the world.”
\verse \textit{When he} had said these \textit{things}, he spat on the ground and made clay with the saliva, and smeared the clay on his eyes.
\verse And he said to him, “Go, wash in the pool of Siloam” (which is translated “sent”). So he went and washed and came back seeing.
\verse Then the neighbors and those who saw him previously (because he was a beggar) began to say, “Is this man not the one who used to sit and beg?”
\verse Others were saying, “It is this man”; others were saying, “No, but he is like him.” That one was saying, “I am \textit{he}!”
\verse So they began to say to him, “How were your eyes opened?”
\verse He replied, “The man who is called Jesus made clay and smeared \textit{it} on my eyes and said to me, ‘Go to Siloam and wash!’ So I went, and I washed, \textit{and} I received sight.”
\verse And they said to him, “Where is that man?” He said, “I do not know.”
\verseWithHeading{The Reaction of the Pharisees to the Healing} They brought him—the one formerly blind—to the Pharisees.
\verse (Now the day on which Jesus made the clay and opened his eyes was the Sabbath.)
\verse So the Pharisees also were asking him again how he received sight. And he said to them, “He put clay on my eyes, and I washed, and I see.”
\verse So some of the Pharisees were saying, “This man is not from God, because he does not observe the Sabbath!” Others were saying, “How can a man \textit{who is} a sinner perform such signs?” And there was a division among them.
\verse So they said to the blind man again, “What do you say about him, because he opened your eyes?” And he said, “He is a prophet.”
\verse So the Jews did not believe concerning him that he had been blind and received sight, until they summoned the parents of the one who received sight.
\verse And they asked them, saying, “Is this man your son, whom you say was born blind? Then how does he now see?”
\verse So his parents answered and said, “We know that this man is our son, and that he was born blind.
\verse But how he now sees we do not know, or who opened his eyes we do not know. Ask him! \textit{He is a mature adult}; he will speak for himself!”
\verse (His parents said these \textit{things} because they were afraid of the Jews, for the Jews had already decided that if anyone should confess him \textit{to be} Christ, he would be expelled from the synagogue.
\verse Because of this his parents said, “\textit{He is a mature adult}; ask him.”)
\verse So they summoned the man who had been blind for the second time and said to him, “Give glory to God! We know that this man is a sinner!”
\verse Then that man replied, “Whether he is a sinner I do not know. One \textit{thing} I know—that \textit{although I} was blind, now I see!”
\verse So they said to him, “What did he do to you? How did he open your eyes?”
\verse He replied to them, “I told you already and you did not listen! Why do you want to hear \textit{it} again? You do not want to become his disciples also, \textit{do you}?”
\verse They reviled him and said, “You are his disciple! But we are disciples of Moses!
\verse We know that God has spoken to Moses, but we do not know where this man is from.”
\verse The man answered and said to them, “For the remarkable thing is this, that you do not know where he is from, and he opened my eyes!
\verse We know that God does not listen to sinners, but if someone is devout and does his will, he listens to this one.
\verse From \textit{time immemorial} it has not been heard that someone opened the eyes of one born blind.
\verse If this man were not from God, he would not be able to do anything!”
\verse They answered and said to him, “You were born completely in sin, and are you attempting to teach us?” And they threw him out.
\verseWithHeading{Jesus as the Son of Man} Jesus heard that they had thrown him out, and finding him, he said, “Do you believe in the Son of Man?”
\verse He answered and said, “And who is \textit{he}, sir, that I may believe in him?”
\verse Jesus said to him, “You have both seen him, and he is the one who is speaking with you.” 〚
\verse And he said, “I believe, Lord!” and he worshiped him.
\verse And Jesus said,〛 “For judgment I have come into this world, so that those who do not see may see, and those who see may become blind!”
\verse \textit{Some} of the Pharisees who were with him heard these \textit{things} and said to him, “We are not also blind, \textit{are we}?”
\verse Jesus said to them, “If you were blind, you would not have sin. But now you say, ‘We see,’ your sin remains.
\end{biblechapter}

\begin{biblechapter} % John 10
\verseWithHeading{Jesus as the Good Shepherd} “Truly, truly I say to you, the one who does not enter through the door into the fold of the sheep, but climbs up at some other place—that one is a thief and a robber.
\verse But the one who enters through the door is the shepherd of the sheep.
\verse For this one the doorkeeper opens, and the sheep hear his voice, and he calls his own sheep by name and leads them out.
\verse Whenever he sends out all his own, he goes before them, and the sheep follow him because they know his voice.
\verse And they will never follow a stranger, but will flee from him, because they do not know the voice of strangers.”
\verse Jesus told them this parable, but they did not understand what it was that he was saying to them.
\verse Then Jesus said to them again, “Truly, truly, I say to you, I am the door of the sheep.
\verse All those who came before me are thieves and robbers, but the sheep do not listen to them.
\verse I am the door. If anyone enters through me, he will be saved, and will come in and will go out and will find pasture.
\verse The thief \textit{comes only} so that he can steal and kill and destroy; I have come so that they may have life, and have \textit{it} abundantly.
\verse “I am the good shepherd. The good shepherd lays down his life for the sheep.
\verse The hired hand, who is not the shepherd, whose own the sheep are not, sees the wolf approaching and abandons the sheep and runs away—and the wolf seizes them and scatters \textit{them}—
\verse because he is a hired hand and \textit{he is not concerned} about the sheep.
\verse “I am the good shepherd, and I know my \textit{own}, and my \textit{own} know me,
\verse just as the Father knows me and I know the Father, and I lay down my life for the sheep.
\verse And I have other sheep which are not from this fold. I must bring these also, and they will hear my voice, and they will become one flock—one shepherd.
\verse Because of this the Father loves me, because I lay down my life so that I may take possession of it again.
\verse No one takes it from me, but I lay it down \textit{voluntarily}. I have authority to lay it down, and I have authority to take possession of it again. This commandment I received from my Father.”
\verse Again there was a division among the Jews because of these words.
\verse And many of them were saying, “He has a demon and is out of his mind! Why do you listen to him?”
\verse Others were saying, “These are not the words of one who is possessed by a demon! A demon is not able to open the eyes of the blind, \textit{is it}?”
\verseWithHeading{Jesus at the Feast of the Dedication} Then the feast of the Dedication took place in Jerusalem. It was winter,
\verse and Jesus was walking in the temple in the Portico of Solomon.
\verse So the Jews surrounded him and began to say to him, “\textit{How long will you keep us in suspense}? If you are the Christ, tell us plainly!”
\verse Jesus answered them, “I told you and you do not believe! The deeds that I do in the name of my Father, these testify about me.
\verse But you do not believe, because you are not of my sheep!
\verse My sheep listen to my voice, and I know them, and they follow me.
\verse And I give them eternal life, and they will never perish \textit{forever}, and no one will seize them out of my hand.
\verse My Father, who has given \textit{them} to me, is greater than all, and no one can seize \textit{them} from the Father’s hand.
\verse The Father and I are one.”
\verse Then the Jews picked up stones again so that they could stone him.
\verse Jesus answered them, “I have shown you many good deeds from the Father. For which one of them are you going to stone me?”
\verse The Jews answered him, “We are not going to stone you concerning a good deed, but concerning blasphemy, and because you, \textit{although you} are a man, make yourself \textit{to be} God!”
\verse Jesus answered them, “Is it not written in your law, ‘I said, “You are gods” ’?
\verse If he called them ‘gods’ to whom the word of God came—and the scripture cannot be broken—
\verse do you say about \textit{he} whom the Father set apart and sent into the world, ‘You are blaspheming,’ because I said, ‘I am the Son of God’?
\verse If I do not do the deeds of my Father, do not believe me.
\verse But if I am doing \textit{them}, even if you do not believe me, believe the deeds, so that you may know and understand that the Father \textit{is} in me and I \textit{am} in the Father.”
\verse So they were seeking again to seize him, and he departed out of their hand.
\verse And he went away again on the other side of the Jordan, to the place where John was baptizing at an earlier time, and he stayed there.
\verse And many came to him and began to say, “John performed no sign, but everything John said about this man was true!”
\verse And many believed in him there.
\end{biblechapter}

\begin{biblechapter} % John 11
\verseWithHeading{Lazarus Dies} Now a certain man was sick, Lazarus from Bethany, the village of Mary and her sister Martha.
\verse (Now it was Mary who anointed the Lord with perfumed oil and wiped his feet with her hair, whose brother Lazarus was sick.)
\verse So the sisters sent \textit{word} to him, saying, “Lord, behold, \textit{the one} whom you love is sick.”
\verse And \textit{when he} heard \textit{it}, Jesus said, “This sickness is not to death, but for the glory of God, in order that the Son of God may be glorified through it.”
\verse (Now Jesus loved Martha and her sister and Lazarus.)
\verse So when he heard that he was sick, then he remained in the place \textit{where} he was two days.
\verse Then after this he said to the disciples, “Let us go to Judea again.”
\verse The disciples said to him, “Rabbi, the Jews were seeking just now to stone you, and are you going there again?”
\verse Jesus replied, Are \textit{there} not twelve hours in the day? If anyone walks around in the daylight, he does not stumble, because he sees the light of this world.
\verse But if anyone walks around in the night, he stumbles, because the light is not in him.
\verse He said these \textit{things}, and after this he said to them, “Our friend Lazarus has fallen asleep, but I am going so that I can awaken him.”
\verse So the disciples said to him, “Lord, if he has fallen asleep, he will get well.”
\verse (Now Jesus had been speaking about his death, but they thought that he was speaking about \textit{real sleep}.)
\verse So Jesus then said to them plainly, “Lazarus has died,
\verse and I am glad \textit{for your sake} that I was not there, so that you may believe. But let us go to him.”
\verse Then Thomas (the one who is called Didymus) said to his fellow disciples, “Let us go also, so that we may die with him.”
\verseWithHeading{Jesus the Resurrection and the Life} So \textit{when he} arrived, Jesus found he had already \textit{been} four days in the tomb.
\verse (Now Bethany was near Jerusalem, about fifteen stadia.
\verse So many of the Jews came to Martha and Mary in order to console them concerning their brother.)
\verse Now Martha, when she heard that Jesus was coming, went to meet him, but Mary was sitting in the house.
\verse So Martha said to Jesus, “Lord, if you had been here, my brother would not have died.
\verse Even now I know that whatever you ask God, God will grant you.”
\verse Jesus said to her, “Your brother will rise again.”
\verse Martha said to him, “I know that he will rise again in the resurrection at the last day.”
\verse Jesus said to her, “I am the resurrection and the life. The one who believes in me, even if he dies, will live,
\verse and everyone who lives and believes in me will never die \textit{forever}. Do you believe this?”
\verse She said to him, “Yes, Lord, I have believed that you are the Christ, the Son of God, who comes into the world.”
\verseWithHeading{Jesus Weeps} And \textit{when she} had said this, she went and called her sister Mary privately, saying, “The Teacher is here and is calling for you.”
\verse So that one, when she heard \textit{it}, got up quickly and went to him.
\verse (Now Jesus has not yet come into the village, but was still in the place where Martha went to meet him.)
\verse So the Jews who were with her in the house and were consoling her, \textit{when they} saw Mary—that she stood up quickly and went out—followed her, \textit{because they} thought that she was going to the tomb in order to weep there.
\verse Then Mary, when she came where Jesus was \textit{and} saw him, fell at his feet, saying to him, “Lord, if you had been here, my brother would not have died.”
\verse Then Jesus, when he saw her weeping and the Jews who came with her weeping, was deeply moved in spirit and was troubled within himself.
\verse And he said, “Where have you laid him?” They said to him, “Lord, come and see.”
\verse Jesus wept.
\verse So the Jews were saying, “See how he loved him!”
\verse But some of them said, “Was not this man who opened the eyes of the blind able to do \textit{something} so that this man also would not have died?”
\verseWithHeading{Lazarus Is Raised} Then Jesus, deeply moved within himself again, came to the tomb. Now it was a cave, and a stone was lying on it.
\verse Jesus said, “Take away the stone.” Martha, the sister of the one who had died, said to him, “Lord, he is stinking already, because it has been four days.”
\verse Jesus said to her, “Did I not say to you that if you believed, you would see the glory of God?”
\verse So they took away the stone. And Jesus lifted up his eyes above and said, “Father, I give thanks to you that you hear me.
\verse And I know that you always hear me, but for the sake of the crowd standing around I said \textit{it}, so that they may believe that you sent me.”
\verse And \textit{when he} had said these \textit{things}, he cried out with a loud voice, “Lazarus, come out!”
\verse The one who had died came out, his feet and his hands bound with strips of cloth, and his face wrapped with a facecloth. Jesus said to them, “Untie him and let him go.”
\verseWithHeading{The Jewish Leaders Plot to Kill Jesus} Then many of the Jews who had come with Mary and saw \textit{the things} which he did believed in him.
\verse But some of them went to the Pharisees and told them \textit{the things} which Jesus had done.
\verse So the chief priests and the Pharisees called together the Sanhedrin and said, “What are we doing? For this man is performing many signs!
\verse If we allow him \textit{to go on} in this way, everyone will believe in him, and the Romans will come and take away both our place and our nation.”
\verse But a certain one of them, Caiaphas (who was high priest in that year), said to them, “You do not know anything at all!
\verse Nor do you consider that it is profitable for you that one man should die for the people, and the whole nation not perish.”
\verse (Now he did not say this from himself, but being high priest in that year, he prophesied that Jesus was going to die for the nation,
\verse and not for the nation only, but also that the children of God who are scattered would be gathered into one.)
\verse So from that day they resolved that they should kill him.
\verse So Jesus was no longer walking openly among the Jews, but went away from there to the region near the wilderness, to a city called Ephraim, and there he stayed with the disciples.
\verse Now the Passover of the Jews was near, and many went up to Jerusalem from the \textit{surrounding} country before the Passover, so that they could purify themselves.
\verse So they were looking for Jesus, and were speaking with one another \textit{while} standing in the temple \textit{courts}, “What do you think? That he will not come to the feast?”
\verse (Now the chief priests and the Pharisees had given orders that if anyone knew where he was, they should report \textit{it}, in order that they could arrest him.)
\end{biblechapter}

\begin{biblechapter} % John 12
\verseWithHeading{Jesus Is Anointed at Bethany} Then, six days before the Passover, Jesus came to Bethany, where Lazarus was, whom Jesus had raised from the dead.
\verse So they made him a dinner there, and Martha was serving, but Lazarus was one of the ones reclining at table with him.
\verse Then Mary took a pound of ointment of very valuable genuine nard \textit{and} anointed the feet of Jesus, and wiped his feet with her hair. And the house was filled with the fragrance of the ointment.
\verse But Judas Iscariot, one of his disciples (the one who was going to betray him) said,
\verse “\textit{Why} was this ointment not sold for three hundred denarii and given to the poor?”
\verse (Now he said this not because \textit{he was concerned} about the poor, but because he was a thief, and having the money box, he used to steal what was put into \textit{it}.)
\verse So Jesus said, “Leave her alone, so that she may keep it for the day of my preparation for burial.
\verse For you have the poor with you always, but you do not always have me.”
\verseWithHeading{The Decision to Kill Lazarus} Now a large crowd of Jews found out that he was there, and they came, not only because of Jesus, but so that they could see Lazarus also, whom he raised from the dead.
\verse So the chief priests decided that they would kill Lazarus also,
\verse because on account of him many of the Jews were going and believing in Jesus.
\verseWithHeading{The Triumphal Entry} On the next day the large crowd who had come to the feast, \textit{when they} heard that Jesus was coming to Jerusalem,
\verse took the branches of palm trees and went out to meet him, and began crying out,
\verse So Jesus found a young donkey \textit{and} sat on it, just as it is written,
\verse “Do not be afraid, daughter of Zion! 
Behold, your king is coming, 
seated on the foal of a donkey!”
\verse (His disciples did not understand these \textit{things} at first, but when Jesus was glorified, then they remembered that these \textit{things} had been written about him and they did these \textit{things} to him.)
\verse So the crowd who was with him when he called Lazarus out of the tomb and raised him from the dead were continuing to testify.
\verse Because of this also the crowd went to meet him, for they had heard \textit{that} he had performed this sign.
\verse So the Pharisees said to one another, “You see that you are accomplishing nothing! Behold, the world has gone after him.”
\verseWithHeading{Greeks Seeking Jesus} Now some Greeks were among those who had gone up in order to worship at the feast.
\verse So these approached Philip, who was from Bethsaida in Galilee, and began asking him saying, “Sir, we want to see Jesus.”
\verse Philip went and told Andrew. Andrew and Philip went and told Jesus.
\verse And Jesus answered them, saying, “The hour has come that the Son of Man will be glorified.
\verse Truly, truly I say to you, unless a grain of wheat falls into the earth \textit{and} dies, it remains \textit{by} itself alone. But if it dies, it bears much fruit.
\verse The one who loves his life loses it, and the one who hates his life in this world preserves it for eternal life.
\verse If anyone serves me, he must follow me, and where I am, there my servant will be also. If anyone serves me, the Father will honor him.
\verseWithHeading{Jesus Predicts His Death} “Now my soul is troubled, and what shall I say? ‘Father, deliver me from this hour’? But for this \textit{reason} I have come to this hour!
\verse Father, glorify your name!” Then a voice came from heaven, “I have both glorified \textit{it}, and I will glorify \textit{it} again.”
\verse Now the crowd that stood \textit{there} and heard \textit{it} said \textit{it had thundered}. Others were saying, “An angel has spoken to him!”
\verse Jesus answered and said, “This voice has not happened for my sake, but \textit{for} your sake.
\verse Now is the judgment of this world! Now the ruler of this world will be thrown out!
\verse And I, when I am lifted up from the earth, will draw all \textit{people} to myself.”
\verse (Now he said this to indicate by what sort of death he was going to die.)
\verse Then the crowd replied to him, “We have heard from the law that the Christ remains \textit{forever}! And how do you say that the Son of Man must be lifted up? Who is this Son of Man?”
\verse So Jesus said to them, “Yet a little time the light is with you! Walk while you have the light, so that the darkness does not overtake you! And the one who walks in the darkness does not know where he is going.
\verse While you have the light, believe in the light, in order that you may become sons of light.” Jesus said these \textit{things}, and \textit{then} he went away \textit{and} was hidden from them.
\verseWithHeading{The People Still Refuse to Believe} But as many signs \textit{as} he had performed before them, they did not believe in him,
\verse in order that the word of the prophet Isaiah would be fulfilled, who said,
\verse For this \textit{reason} they were not able to believe, because again Isaiah said,
\verse “He has blinded their eyes 
and hardened their hearts, 
lest they see with \textit{their} eyes 
and understand with \textit{their} hearts 
and turn, and I heal them.”
\verse Isaiah said these \textit{things} because he saw his glory, and he spoke about him.
\verse Yet despite that, even many of the rulers believed in him, but because of the Pharisees they did not confess \textit{it}, so that they would not be expelled from the synagogue.
\verse For they loved the praise of men more than praise from God.
\verseWithHeading{Jesus’ Final Public Appeal} But Jesus cried out and said, “The one who believes in me does not believe in me, but in the one who sent me,
\verse and the one who sees me sees the one who sent me.
\verse I have come \textit{as} a light into the world, in order that everyone who believes in me will not remain in the darkness.
\verse And if anyone hears my words and does not observe \textit{them}, I will not judge him. For I have not come to judge the world, but to save the world.
\verse The one who rejects me and does not accept my words has one who judges him; the word that I have spoken will judge him on the last day.
\verse For I have not spoken from myself, but the Father himself who sent me \textit{has commanded me} what I should say and what I should speak.
\verse And I know that his commandment is eternal life. So \textit{the things} that I say, just as the Father said to me, thus I say.”
\end{biblechapter}

\begin{biblechapter} % John 13
\verseWithHeading{Jesus Washes His Disciples’ Feet} Now before the feast of Passover, Jesus, knowing that his hour had come that he would depart from this world to the Father, \textit{and} having loved \textit{his} own in the world, loved them to the end.
\verse And \textit{as} a dinner was taking place, \textit{when} the devil had already put into the heart of Judas \textit{son} of Simon Iscariot that he should betray him,
\verse \textit{because he} knew that the Father had given him all \textit{things} into \textit{his} hands, and that he had come forth from God and was going away to God,
\verse he got up from the dinner and took off \textit{his} outer clothing, and taking a towel, tied \textit{it} around himself.
\verse Then he poured water into the washbasin and began to wash the feet of the disciples, and to wipe \textit{them} dry with the towel \textit{which he had tied around himself}.
\verse Then he came to Simon Peter. He said to him, “Lord, are you going to wash my feet?”
\verse Jesus answered and said to him, “What I am doing you do not understand now, but you will understand after these \textit{things}.”
\verse Peter said to him, “You will never wash my feet \textit{forever}!” Jesus replied to him, “Unless I wash you, you do not have a share with me.”
\verse Simon Peter said to him, “Lord, not my feet only, but also \textit{my} hands and \textit{my} head!”
\verse Jesus said to him, “The one who has bathed \textit{only needs} to wash \textit{his} feet, but is completely clean. And you are clean, but not all \textit{of you}.”
\verse (For he knew the one who would betray him; because of this he said, “Not all \textit{of you} are clean.”)
\verse So when he had washed their feet and taken his outer clothing and reclined at table again, he said to them, “Do you understand what I have done for you?
\verse You call me ‘Teacher’ and ‘Lord,’ and you speak correctly, for I am.
\verse If then I—\textit{your} Lord and Teacher—wash your feet, you also ought to wash one another’s feet.
\verse For I have given you an example, that just as I have done for you, you also do.
\verse Truly, truly I say to you, a slave is not greater than his master, nor a messenger greater than the one who sent him.
\verse If you understand these \textit{things}, you are blessed if you do them.
\verse “I am not speaking about all of you. I know whom I have chosen, but in order that the scripture would be fulfilled, ‘The one who eats my bread has lifted up his heel against me.’
\verse From now \textit{on} I am telling you before \textit{it} happens, in order that when \textit{it} happens you may believe that I am \textit{he}.
\verse Truly, truly I say to you, the one who receives anyone I send receives me, and the one who receives me receives the one who sent me.”
\verseWithHeading{Jesus Predicts Judas’ Betrayal} \textit{When he} had said these \textit{things}, Jesus was troubled in spirit and testified and said, “Truly, truly I say to you that one of you will betray me.”
\verse The disciples began looking at one another, uncertain about whom he was speaking.
\verse One of his disciples—the one whom Jesus loved—was reclining \textit{close beside} Jesus.
\verse So Simon Peter gestured for this one to inquire who it was about whom he was speaking.
\verse He leaned back accordingly against Jesus’ chest \textit{and} said to him, “Lord, who is it?”
\verse Jesus replied, “It is he to whom I dip the piece of bread and give \textit{it} to him.” Then \textit{after} dipping the piece of bread, he gave \textit{it}﻿ to Judas \textit{son} of Simon Iscariot.
\verse And after the piece of bread, then Satan entered into him. Then Jesus said to him, “What you are doing, do quickly!”
\verse (Now no one of those reclining at table knew for what \textit{reason} he said this to him.
\verse For some were thinking because Judas had the money box, Jesus was telling him, “Purchase \textit{what we need} for the feast,” or that he should give something to the poor.)
\verse So \textit{after he} had taken the piece of bread, he went out immediately. And it was night.
\verseWithHeading{Jesus Predicts Peter’s Denial} Then, when he had gone out, Jesus said, “Now the Son of Man is glorified, and God is glorified in him.
\verse If God is glorified in him, God will also glorify him in himself, and will glorify him immediately.
\verse Children, yet a little \textit{time} I am with you. You will seek me and just as I said to the Jews, “Where I am going you cannot come,” now I say also to you.
\verse “A new commandment I give to you: that you love one another—just as I have loved you, that you also love one another.
\verse By this everyone will know that you are my disciples—if you have love for one another.”
\verse Simon Peter said to him, “Lord, where are you going?” Jesus replied, “Where I am going you cannot follow me now, but you will follow later.”
\verse Peter said to him, “Lord, why am I not able to follow you now? I will lay down my life for you!”
\verse Jesus replied, “Will you lay down your life for me? Truly, truly I say to you, the rooster will not crow until you have denied me three times!
\end{biblechapter}

\begin{biblechapter} % John 14
\verseWithHeading{Jesus’ Farewell Discourse} “Do not let your hearts be troubled. You believe in God; believe also in me.
\verse In my Father’s house there are many dwelling places; but if not, I would have told you, because I am going away to prepare a place for you.
\verse And if I go and prepare a place for you, I will come again and receive you to myself, so that where I am, you may be also.
\verse And you know the way where I am going.”
\verse Thomas said to him, “Lord, we do not know where you are going. How are we able to know the way?”
\verse Jesus said to him, “I am the way, and the truth, and the life. No one comes to the Father except through me.
\verse If you had known me, you would have known my Father also. From now on you know him and have seen him.”
\verse Philip said to him, “Lord, show us the Father, and it is enough for us.”
\verse Jesus said to him, “Am I with you so long a time and you have not known me, Philip? The one who has seen me has seen the Father! How can you say, ‘Show us the Father?’
\verse Do you not believe that I \textit{am} in the Father and the Father is in me? The words that I say to you I do not speak from myself, but the Father residing in me does his works.
\verse Believe me that I \textit{am} in the Father and the Father \textit{is} in me; but if not, believe because of the works themselves.
\verse Truly, truly I say to you, the one who believes in me, the works that I am doing he will do also, and he will do greater \textit{works} than these because I am going to the Father.
\verse And \textit{whatever} you ask in my name, I will do this, in order that the Father may be glorified in the Son.
\verse If you ask me anything in my name, I will do \textit{it}.
\verseWithHeading{Jesus Promises the Holy Spirit} “If you love me, you will keep my commandments.
\verse And I will ask the Father, and he will give you another Advocate, in order that he may be with you \textit{forever}—
\verse the Spirit of truth, whom the world is not able to receive, because it does not see him or know \textit{him}. You know him, because he resides with you and will be in you.
\verse “I will not leave you \textit{as} orphans; I am coming to you.
\verse Yet a little \textit{time} and the world will see me no longer, but you will see me; because I live, you also will live.
\verse On that day you will know that I \textit{am} in my Father, and you \textit{are} in me, and I \textit{am} in you.
\verse The one who has my commandments and keeps them—that one is the one who loves me. And the one who loves me will be loved by my Father, and I will love him and will reveal myself to him.”
\verse Judas (not Iscariot) said to him, “Lord, why is it that you are going to reveal yourself to us and not to the world?”
\verse Jesus answered and said to him, “If anyone loves me he will keep my word, and my Father will love him, and we will come to him and \textit{will take up residence with him}.
\verse The one who does not love me does not keep my words, and the word that you hear is not mine, but the Father’s who sent me.
\verse These \textit{things} I have spoken to you \textit{while} residing with you.
\verse But the Advocate, the Holy Spirit, whom the Father will send in my name—that one will teach you all \textit{things}, and will remind you of everything that I said to you.
\verse “Peace I leave with you; my peace I give to you—not as the world gives, I give to you. Do not let your hearts be troubled, and do not let \textit{them} be afraid.
\verse You have heard that I said to you, ‘I am going away, and I am coming to you.’ If you loved me, you would have rejoiced that I am going to the Father, because the Father is greater than I \textit{am}.
\verse And now I have told you before \textit{it} happens, so that when it happens, you may believe.
\verse I will no longer speak much with you, for the ruler of the world is coming, and he has \textit{no power} \textit{over} me.
\verse But so that the world may know that I love the Father, and just as the Father has commanded me, thus I am doing. Get up, let us go from here!
\end{biblechapter}

\begin{biblechapter} % John 15
\verseWithHeading{The Vine and the Branches} “I am the true vine, and my Father is the vinedresser.
\verse Every branch that does not bear fruit in me, he removes it, and every \textit{branch} that bears fruit, he prunes it in order that it may bear more fruit.
\verse You are already clean because of the word that I have spoken to you.
\verse Remain in me, and I in you. Just as the branch is not able to bear fruit from itself unless it remains in the vine, so neither \textit{can} you, unless you remain in me.
\verse “I am the vine; you \textit{are} the branches. The one who remains in me and I in him—this one bears much fruit, for apart from me you are not able to do anything.
\verse If anyone does not remain in me, he is thrown out as a branch, and dries up, and they gather them and throw \textit{them} into the fire, and they are burned.
\verse If you remain in me and my words remain in you, ask whatever you want and it will be done for you.
\verse My Father is glorified by this: that you bear much fruit, and prove to be my disciples.
\verse “Just as the Father has loved me, I also have loved you. Remain in my love.
\verse If you keep my commandments, you will remain in my love, just as I have kept my Father’s commandments and remain in his love.
\verse I have spoken these \textit{things} to you in order that my joy may be in you, and your joy may be made complete.
\verse This is my commandment: that you love one another just as I have loved you.
\verse No one has greater love than this: that someone lay down his life for his friends.
\verse You are my friends if you do what I command you.
\verse No longer do I call you slaves, because the slave does not know what his master is doing. But I have called you friends, because everything that I have heard from my Father I have revealed to you.
\verse You did not choose me, but I chose you and appointed you that you should go and bear fruit, and your fruit should remain, in order that whatever you ask the Father in my name he will give you.
\verse These \textit{things} I command you: that you love one another.
\verseWithHeading{The World’s Hatred for the Disciples} If the world hates you, you know that it has hated me before \textit{it hated} you.
\verse If you were of the world, the world would love its own. But because you are not of the world, but I chose you out of the world, for this \textit{reason} the world hates you.
\verse Remember the word that I said to you: ‘A slave is not greater than his master.’ If they persecuted me, they will also persecute you. If they kept my word, they will keep yours also.
\verse But they will do all these \textit{things} to you on account of my name, because they do not know the one who sent me.
\verse If I had not come and spoken to them, they would not have sin. But now they do not have a valid excuse for their sin.
\verse The one who hates me hates my Father also.
\verse If I had not done among them the works that no one else did, they would not have sin. But now they have both seen and hated both me and my Father.
\verse But \textit{this happened} so that the word that is written in their law would be fulfilled, ‘They hated me without a reason.’
\verse “When the Advocate comes, whom I will send to you from the Father—the Spirit of truth, who proceeds from the Father—that one will testify about me.
\verse And you also will testify, because you have been with me from the beginning.
\end{biblechapter}

\begin{biblechapter} % John 16
\verseWithHeading{Persecution of Disciples Predicted} “I have said these \textit{things} to you so that you will not fall away.
\verse \textit{They will expel you from the synagogue}, but an hour is coming that everyone who kills you will think they are offering service to God.
\verse And they will do these \textit{things} because they do not know the Father or me.
\verse But I have said these \textit{things} to you so that when their hour comes, you may remember that I told you about them.
\verseWithHeading{Jesus’ Departure and the Coming of the Holy Spirit} “And I have not said these \textit{things} to you from the beginning, because I was with you.
\verse But now I am going away to the one who sent me, and none of you is asking me, ‘Where are you going?’
\verse But because I have said these \textit{things} to you, sorrow has filled your hearts.
\verse But I tell you the truth, it is better for you that I go away. For if I do not go away, the Advocate will not come to you; but if I go, I will send him to you.
\verse And \textit{when he} comes, he will convict the world concerning sin and concerning righteousness and concerning judgment:
\verse concerning sin, because they do not believe in me,
\verse and concerning righteousness, because I am going away to the Father and you will see me no more,
\verse I still have many \textit{things} to say to you, but you are not able to bear \textit{them} now.
\verse But when he—the Spirit of truth—comes, he will guide you into all the truth. For he will not speak from himself, but whatever he hears he will speak, and he will proclaim to you the things to come.
\verse He will glorify me, because he will take from what \textit{is} mine and will proclaim \textit{it} to you.
\verse Everything that the Father has is mine. For this \textit{reason} I said that he takes from what \textit{is} mine and will proclaim \textit{it} to you.
\verseWithHeading{Jesus Predicts His Return to the Disciples} “A little \textit{while} and you will see me no more, and again a little \textit{while} and you will see me.
\verse So some of his disciples said to one another, “What is this that he is saying to us, ‘A little \textit{while} and you will not see me, and again a little \textit{while} and you will see me,’ and ‘Because I am going away to the Father’?”
\verse So they kept on saying, “What is this that he is saying, ‘A little \textit{while}’? We do not understand what he is speaking about!”
\verse Jesus knew that they were wanting to ask him, and he said to them, “Are you deliberating with one another about this—that I said, ‘A little \textit{while}, and you will not see me, and again a little \textit{while} and you will see me’?
\verse Truly, truly I say to you, that you will weep and lament, but the world will rejoice; you will become sorrowful, but your sorrow will change to joy.
\verse A woman, when she gives birth, experiences pain because her hour has come. But when \textit{her} child is born, she no longer remembers the affliction, on account of the joy that a human being has been born into the world.
\verse So you also are experiencing sorrow now, but I will see you again, and your hearts will rejoice, and no one will take away your joy from you.
\verse And on that day you will ask me nothing. Truly, truly I say to you, whatever you ask the Father in my name, he will give you.
\verse Until now you have asked nothing in my name. Ask and you will receive, so that your joy may be complete.
\verseWithHeading{Jesus’ Victory Over the World} “I have said these \textit{things} to you in figurative sayings. An hour is coming when I will speak to you in figurative sayings no longer, but I will tell you plainly about the Father.
\verse On that day you will ask in my name, and I do not say to you that I will ask the Father on your behalf.
\verse For the Father himself loves you, because you have loved me and have believed that I came from God.
\verse I have gone out from the Father and have come into the world; again, I am leaving the world and am going to the Father.”
\verse His disciples said, “Behold, now you are speaking \textit{plainly} and are telling \textit{us} no figurative saying!
\verse Now we know that you know everything and do not \textit{need for} anyone to ask you \textit{questions}. By this we believe that you have come from God.”
\verse Jesus replied to them, “Now do you believe?
\verse Behold, an hour is coming—and has come—that you will be scattered each one to his own \textit{home}, and you will leave me alone. And I am not alone, because the Father is with me.
\verse I have said these \textit{things} to you so that in me you may have peace. In the world you have affliction, but have courage! I have conquered the world.”
\end{biblechapter}

\begin{biblechapter} % John 17
\verseWithHeading{Jesus Prays to be Glorified} Jesus said these \textit{things}, and lifting up his eyes to heaven he said, “Father, the hour has come! Glorify your Son, in order that your Son may glorify you—
\verse just as you have given him authority over all flesh, in order that he would give eternal life to them—everyone whom you have given him.
\verse Now this is eternal life: that they know you, the only true God, and Jesus Christ, whom you have sent.
\verse I have glorified you on earth \textit{by} completing the work that you have given me \textit{to do}.
\verse And now, Father, you glorify me \textit{at your side} with the glory that I had \textit{at your side} before the world existed.
\verseWithHeading{Jesus Prays for His Disciples} “I have revealed your name to the men whom you gave me out of the world. They were yours, and you have given them to me, and they have kept your word.
\verse Now they understand that all \textit{the things} that you have given me are from you,
\verse because the words that you gave to me I have given to them, and they received \textit{them} and know truly that I have come from you, and they have believed that you have sent me.
\verse I am asking on behalf of them. I am not asking on behalf of the world, but on behalf of those whom you have given me, because they are yours,
\verse and all my \textit{things} are yours, and your \textit{things} \textit{are} mine, and I have been glorified in them.
\verse And I am no longer in the world, and they are in the world, and I am coming to you. Holy Father, keep them in your name, which you have given to me, so that they may be one, just as we \textit{are}.
\verse When I was with them, I kept them in your name, which you have given to me, and guarded \textit{them}, and none of them has perished except the son of destruction, in order that the scripture would be fulfilled.
\verse And now I am coming to you, and I am saying these \textit{things} in the world so that they may have my joy completed in themselves.
\verse I have given them your word, and the world has hated them, because they are not of the world just as I am not of the world.
\verse I do not ask that you take them out of the world, but that you protect them from the evil one.
\verse They are not of the world, just as I am not of the world.
\verse Sanctify them in the truth—your word is truth.
\verse Just as you sent me into the world, I also have sent them into the world.
\verse And for them I sanctify myself, so that \textit{they} themselves also may be sanctified in the truth.
\verseWithHeading{Jesus Prays for the Unity of Believers} “And I do not ask on behalf of these only, but also on behalf of those who believe in me through their word,
\verse that they all may be one, just as you, Father, \textit{are} in me and I \textit{am} in you, that they also may be in us, in order that the world may believe that you sent me.
\verse And the glory that you have given to me, I have given to them, in order that they may be one, just as we \textit{are} one—
\verse I in them, and you in me, in order that they may be completed in one, so that the world may know that you sent me, and you have loved them just as you have loved me.
\verse “Father, \textit{those} whom you have given to me—I want that those also may be with me where I am, in order that they may see my glory that you have given me because you loved me before the foundation of the world.
\verse Righteous Father, although the world does not know you, yet I have known you, and these \textit{men} have come to know that you sent me.
\verse And I made known to them your name, and will make \textit{it} known, in order that the love \textit{with} which you loved me may be in them, and I \textit{may be} in them.”
\end{biblechapter}

\begin{biblechapter} % John 18
\verseWithHeading{Jesus Is Betrayed and Arrested} \textit{When} Jesus had said these \textit{things}, he went out with his disciples to the other side of the ravine of the Kidron, where \textit{there} was a garden into which he and his disciples entered.
\verse (Now Judas, the one who betrayed him, also knew about the place, because Jesus often gathered there with his disciples.)
\verse So Judas, taking the cohort and officers from the chief priests and from the Pharisees, came there with lanterns and torches and weapons.
\verse Then Jesus, \textit{because he} knew all the things that were coming upon him, went out and said to them, “Who are you looking for?”
\verse They replied to him, “Jesus the Nazarene.” He said to them, “I am \textit{he}.” (Now Judas, the one who betrayed him, was also standing with them.)
\verse So when he said to them, “I am \textit{he},” \textit{they drew back} and fell to the ground.
\verse Then he asked them again, “Who are you looking for?” And they said, “Jesus the Nazarene.”
\verse Jesus replied, “I said to you that I am \textit{he}! So if you are looking for me, let these \textit{men} go,”
\verse in order that the word that he had spoken would be fulfilled: “Those whom you have given to me—I have not lost anyone of them.”
\verse Then Simon Peter, who had a sword, drew it and struck the slave of the high priest and cut off his right ear. (Now the name of the slave was Malchus.)
\verse So Jesus said to Peter, “Put the sword into \textit{its} sheath! The cup that the Father has given me—shall I not drink it?”
\verseWithHeading{Jesus Taken to Annas} Then the cohort and the military tribune and the officers of the Jews seized Jesus and tied him up,
\verse and brought \textit{him} to Annas first, for he was the father-in-law of Caiaphas, who was high priest that year.
\verse (Now it was Caiaphas who had advised the Jews that it was better \textit{that} one man die for the people.)
\verseWithHeading{Peter Denies Jesus the First Time} So Simon Peter and another disciple followed Jesus. (Now that disciple was known to the high priest, and entered with Jesus into the courtyard of the high priest.)
\verse But Peter was standing by the door outside. So the other disciple \textit{who was} known to the high priest went out and spoke to the doorkeeper and brought Peter in.
\verse Then the female slave \textit{who was} the doorkeeper said to Peter, “You are not also \textit{one} of the disciples of this man, \textit{are you}?” He said, “I am not!”
\verse (Now the slaves and the officers were standing there, having made a charcoal fire because it was cold, and they were warming themselves. And Peter was also standing there with them and warming himself.)
\verseWithHeading{Jesus Before Annas} So the high priest questioned Jesus about his disciples and about his teaching.
\verse Jesus replied to him, “I have spoken openly to the world. I always taught in the synagogue and in the temple \textit{courts} where all the Jews assemble, and I have said nothing in secret.
\verse Why are you asking me? Ask those who heard what I have said to them! Behold, these \textit{people} know what I said.”
\verse Now \textit{when} he had said these \textit{things}, one of the officers who was standing by gave a slap in the face to Jesus, saying, “Do you reply to the high priest in this way?”
\verse Jesus replied to him, “If I have spoken wrongly, testify about what \textit{is} wrong! But if \textit{I have spoken} correctly, why do you strike me?”
\verse Then Annas sent him, tied up, to Caiaphas the high priest.
\verseWithHeading{Peter Denies Jesus the Second and Third Times} Now Simon Peter was standing there and warming himself. So they said to him, “You are not also \textit{one} of his disciples, \textit{are you}?” He denied \textit{it} and said, “I am not!”
\verse One of the slaves of the high priest, who was related to the one whose ear Peter had cut off, said, “Did I not see you in the garden with him?”
\verse So Peter denied \textit{it} again, and immediately a rooster crowed.
\verseWithHeading{Jesus Brought Before Pilate} Then they brought Jesus from Caiaphas to the governor’s residence. Now it was early, and they did not enter into the governor’s residence so that they would not be defiled, but could eat the Passover.
\verse So Pilate came outside to them and said, “What accusation do you bring against this man?”
\verse They answered and said to him, “If this man were not doing evil, we would not have handed him over to you!”
\verse So Pilate said to them, “You take him and judge him according to your law!” The Jews said to him, “It is not permitted for us to kill anyone,”
\verse in order that the word of Jesus would be fulfilled that he had spoken, indicating by what sort of death he was going to die.
\verseWithHeading{Pilate Questions Jesus} Then Pilate entered again into the governor’s residence and summoned Jesus and said to him, “Are you the king of the Jews?”
\verse Jesus replied, “Do you say this from yourself, or have others said \textit{this} to you about me?”
\verse Pilate replied, “I am not a Jew, \textit{am I}? Your people and the chief priests handed you over to me! What have you done?”
\verse Jesus replied, “My kingdom is not of this world. If my kingdom were of this world, my servants would be fighting so that I would not be handed over to the Jews. But now my kingdom is not from here.”
\verse Then Pilate said to him, “So then you are a king!” Jesus replied, “You say that I am a king. For this \textit{reason} I was born, and for this reason I have come into the world: in order that I can testify to the truth. Everyone who is of the truth hears my voice.”
\verse Pilate said to him, “What is truth?”
\verse And \textit{when he} had said this, he went out again to the Jews and said to them, “I find no basis for an accusation against him.
\verse But it is your custom that I release for you one \textit{prisoner} at the Passover. So do you want \textit{me} to release for you the king of the Jews?”
\end{biblechapter}

\begin{biblechapter} % John 19
\verseWithHeading{Pilate Attempts to Release Jesus} So then Pilate took Jesus and had him flogged.
\verse And the soldiers wove a crown of thorns and placed \textit{it} on his head, and put a purple robe on him,
\verse and were coming up to him and saying, “Hail, king of the Jews!” and were giving him slaps in the face.
\verse And Pilate came outside again and said to them, “Behold, I am bringing him outside to you, so that you will know that I find no basis for an accusation against him.”
\verse Then Jesus came outside wearing the crown of thorns and the purple robe, and he said to them, “Behold the man!”
\verse So when they saw him, the chief priests and the officers shouted, saying, “Crucify! Crucify!” Pilate said to them, “You take him and crucify \textit{him}! For I do not find a basis for an accusation against him.”
\verse The Jews replied to him, “We have a law, and according to the law he ought to die, because he made himself out to be the Son of God!”
\verse So when Pilate heard this statement, he was even more afraid,
\verse and he entered into the governor’s residence again and said to Jesus, “Where are you from?” But Jesus did not give him an answer.
\verse So Pilate said to him, “Will you not speak to me? Do you not know that I have authority to release you, and I have authority to crucify you?”
\verse Jesus replied to him, “You would not have any authority over me unless it was given to you from above. For this \textit{reason} the one who handed me over to you has greater sin.”
\verse From this \textit{point on} Pilate was seeking to release him, but the Jews shouted, saying, “If you release this man, you are not a friend of Caesar! Everyone who makes himself out to be a king opposes Caesar!”
\verse So Pilate, \textit{when he} heard these words, brought Jesus outside and sat down on the judgment seat, in the place called The Stone Pavement (but \textit{Gabbatha} in Aramaic).
\verse (Now it was the day of preparation of the Passover; it was about the sixth hour.) And he said to the Jews, “Behold your king!”
\verse Then those shouted, “Away \textit{with him}! Away \textit{with him}! Crucify him!” Pilate said to them, “Shall I crucify your king?” The chief priests replied, “We do not have a king except Caesar!”
\verse So then he handed him over to them in order that he could be crucified.
\verseWithHeading{Jesus Is Crucified} So they took Jesus,
\verse and carrying for himself the cross, he went out to the \textit{place} called The Place of a Skull (which is called \textit{Golgotha} in Aramaic),
\verse where they crucified him, and with him two others, \textit{one on each side}, and Jesus in the middle.
\verse And Pilate also wrote a notice and placed \textit{it} on the cross, and it was written: “Jesus the Nazarene, the king of the Jews.”
\verse So many of the Jews read this notice, because the place where Jesus was crucified was near the city. And it was written in Aramaic, in Latin, \textit{and} in Greek.
\verse Then the chief priests of the Jews said to Pilate, “Do not write ‘The king of the Jews,’ but, ‘He said, I am king of the Jews.’ ”
\verse Then the soldiers, when they had crucified Jesus, took his clothing and made four shares—for each soldier a share—and the tunic. (Now the tunic was seamless, woven from the top \textit{in a single piece}.)
\verse So they said to one another, “Let us not tear it apart, but cast lots for it, \textit{to see} whose it will be,” so that the scripture would be fulfilled that says,
\verse Now his mother and the sister of his mother, Mary the \textit{wife} of Clopas, and Mary Magdalene were standing near the cross of Jesus.
\verse So Jesus, seeing \textit{his} mother and the disciple whom he loved standing there, said to \textit{his} mother, “Woman, behold your son!”
\verse Then he said to the disciple, “Behold your mother!” And from that hour the disciple took her into his own \textit{home}.
\verseWithHeading{Jesus Dies on the Cross} After this, Jesus, knowing that now at last everything was completed, in order that the scripture would be fulfilled, said, “I am thirsty.”
\verse A jar full of sour wine was standing there, so \textit{they} put a sponge full of the sour wine on a \textit{branch of} hyssop \textit{and} brought \textit{it} to his mouth.
\verse Then when he had received the sour wine, Jesus said, “It is completed,” and bowing \textit{his} head, he gave up \textit{his} spirit.
\verse Then the Jews, because it was the day of preparation, so that the bodies would not remain on the cross on the Sabbath (for that Sabbath was an important day), asked Pilate that their legs could be broken and they could be taken away.
\verse So the soldiers came and broke the legs of the first and of the other who had been crucified with him.
\verse But \textit{when they} came to Jesus, after they saw he was already dead, they did not break his legs.
\verse But one of the soldiers pierced his side with a spear, and blood and water came out immediately.
\verse And the one who has seen \textit{it} has testified, and his testimony is true, and that person knows that he is telling the truth, so that you also may believe.
\verse For these \textit{things} happened in order that the scripture would be fulfilled: “Not a bone of his will be broken.”
\verse And again another scripture says, “They will look on \textit{the one} whom they have pierced.”
\verseWithHeading{Jesus Is Buried} And after these \textit{things}, Joseph \textit{who was} from Arimathea, who was a disciple of Jesus (but a secret one for fear of the Jews), asked Pilate that he might take away the body of Jesus. And Pilate allowed \textit{it}, so he came and took away his body.
\verse And Nicodemus—the one who had come to him formerly at night—also came, bringing a mixture of myrrh and aloes \textit{weighing} about a hundred pounds.
\verse So they took the body of Jesus and wrapped it in \textit{strips of} linen cloth with the fragrant spices, as is the Jews’ custom to prepare for burial.
\verse Now \textit{there} was a garden at the place where he was crucified, and in the garden a new tomb in which no one was yet buried.
\verse So there, on account of the day of preparation of the Jews, because the tomb was close by, they buried Jesus.
\end{biblechapter}

\begin{biblechapter} % John 20
\verseWithHeading{Jesus Is Raised} Now on the first \textit{day} of the week, Mary Magdalene came to the tomb early, \textit{while it} was still dark, and saw the stone had been taken away from the tomb.
\verse So she ran and came to Simon Peter and to the other disciple whom Jesus loved and said to them, “They have taken away the Lord from the tomb and we do not know where they have put him!”
\verse Then Peter and the other disciple went out and were going to the tomb.
\verse And the two were running together, and the other disciple ran ahead, faster than Peter, and came to the tomb first.
\verse And bending over to look, he saw the \textit{strips of} linen cloth lying there, though he did not go in.
\verse Then Simon Peter also came following him, and he went into the tomb and saw the \textit{strips of} linen cloth lying there,
\verse and the facecloth that was on his head—not lying with the \textit{strips of} linen cloth, but folded up separately in one place.
\verse So then the other disciple who had come to the tomb first also went in, and he saw and believed.
\verse (For they did not yet know the scripture that it was necessary for him to rise from the dead.)
\verseWithHeading{Jesus Appears to Mary Magdalene} Then the disciples went away again to their \textit{own homes}.
\verse But Mary stood outside at the tomb, weeping. Then, while she was weeping, she bent over to look into the tomb,
\verse and she saw two angels in white, seated one at the head and one at the feet where the body of Jesus had been lying.
\verse And they said to her, “Woman, why are you weeping?” She said to them, “They have taken away my Lord, and I do not know where they have put him!”
\verse \textit{When she} had said these \textit{things}, she turned \textit{around} and saw Jesus standing there, and she did not know that it was Jesus.
\verse Jesus said to her, “Woman, why are you weeping? Who are you looking for?” She thought that it was the gardener, \textit{and} said to him, “Sir, if you have carried him away, tell me where you have put him, and I will take him.”
\verse Jesus said to her, “Mary.” She turned around \textit{and} said to him in Aramaic, “\textit{Rabboni}” (which means “Teacher”).
\verse Jesus said to her, “Do not touch me, for I have not yet ascended to the Father. But go to my brothers and tell them, ‘I am ascending to my Father and your Father, and my God and your God.’ ”
\verse Mary Magdalene came \textit{and} announced to the disciples, “I have seen the Lord,” and he had said these \textit{things} to her.
\verseWithHeading{Jesus Appears to the Disciples} Now \textit{when it} was evening on that day—the first \textit{day} of the week—and the doors had been shut where the disciples were because of fear of the Jews, Jesus came and stood in their midst and said to them, “Peace to you.”
\verse And \textit{when he} had said this, he showed \textit{his} hands and \textit{his} side to them. Then the disciples rejoiced \textit{when they} saw the Lord.
\verse So Jesus said to them again, “Peace to you. As the Father has sent me, I also send you.”
\verse And \textit{when he} had said this, he breathed on \textit{them} and said to them, “Receive the Holy Spirit.
\verse If you forgive the sins of any, they are forgiven them. If you retain \textit{the sins} of any, they are retained.”
\verseWithHeading{Thomas Doubts But Later Believes} Now Thomas, one of the twelve, who was called Didymus, was not with them when Jesus came.
\verse So the other disciples said to him, “We have seen the Lord!” But he said to them, “Unless I see in his hands the mark of the nails, and put my finger into the mark of the nails, and put my hand into his side, I will never believe!”
\verse And after eight days his disciples were again inside, and Thomas with them. \textit{Although} the doors had been shut, Jesus came and stood in their midst and said, “Peace to you.”
\verse Then he said to Thomas, “Place your finger here and see my hands, and place your hand and put \textit{it} into my side. And do not be unbelieving, but believing!”
\verse Thomas answered and said to him, “My Lord and my God!”
\verse Jesus said to him, “Because you have seen me, have you believed? Blessed \textit{are} those who have not seen and have believed.”
\verseWithHeading{Why This Book Was Written} Now Jesus also performed many other signs in the presence of the disciples which are not recorded in this book,
\verse but these \textit{things} are recorded in order that you may believe that Jesus is the Christ, the Son of God, and that \textit{by} believing you may have life in his name.
\end{biblechapter}

\begin{biblechapter} % John 21
\verseWithHeading{Jesus Appears to the Disciples in Galilee} After these \textit{things} Jesus revealed himself again to the disciples by the Sea of Tiberias. Now he revealed \textit{himself} in this way:
\verse Simon Peter and Thomas (who was called Didymus) and Nathanael from Cana in Galilee and the \textit{sons} of Zebedee and two others of his disciples were together.
\verse Simon Peter said to them, “I am going fishing!” They said to him, “We also are coming with you.” They went out and got into the boat, and during that night they caught nothing.
\verse Now \textit{when it} was already early morning, Jesus stood on the beach. However, the disciples did not know that it was Jesus.
\verse So Jesus said to them, “Children, you do not have any fish to eat, \textit{do you}? They answered him, “No.”
\verse And he said to them, “Throw the net on the right side of the boat, and you will find \textit{some}.” So they threw \textit{it}, and were no longer able to haul it in from the large number of the fish.
\verse Then that disciple whom Jesus loved said to Peter, “It is the Lord!” So Simon Peter, \textit{when he} heard that it was the Lord, tied around himself his outer garment (for he was naked) and threw himself into the sea.
\verse But the other disciples came in the boat, dragging the net of fish, because they were not far from the land, but about two hundred cubits away.
\verse So when they got out on the land, they saw a charcoal fire laid \textit{there}, and a fish lying on \textit{it}, and bread.
\verse Jesus said to them, “Bring some of the fish that you have just now caught.”
\verse So Simon Peter got into \textit{the boat} and hauled the net to the land, full of large fish—one hundred fifty-three—and \textit{although there} were so many, the net was not torn.
\verse Jesus said to them, “Come, eat breakfast!” But none of the disciples dared to ask him, “Who are you?” \textit{because they} knew that it was the Lord.
\verse Jesus came and took the bread and gave \textit{it} to them, and the fish likewise.
\verse This \textit{was} now \textit{the} third \textit{time} Jesus was revealed to the disciples \textit{after he} had been raised from the dead.
\verseWithHeading{Peter Is Restored Three Times} Now when they had eaten breakfast, Jesus said to Simon Peter, “Simon \textit{son} of John, do you love me more than these?” He said to him, “Yes, Lord, you know that I love you.” He said to him, “Feed my lambs!”
\verse He said to him again a second time, “Simon \textit{son} of John, do you love me?” He said to him, “Yes, Lord, you know that I love you.” He said to him, “Shepherd my sheep!”
\verse He said to him a third \textit{time}, “Simon \textit{son} of John, do you love me?” Peter was distressed because he said to him a third \textit{time}, “Do you love me?” and he said to him, “Lord, you know everything! You know that I love you!” Jesus said to him, “Feed my sheep!
\verse Truly, truly I say to you, when you were young, you tied \textit{your clothes} around yourself and walked wherever you wanted. But when you grow old, you will stretch out your hands, and someone else will tie you up and carry \textit{you} where you do not want \textit{to go}.
\verse (Now he said this to indicate by what kind of death he would glorify God.) And \textit{after he} had said this, he said to him, “Follow me!”
\verseWithHeading{Peter and the Other Disciple Jesus Loved} Peter turned \textit{and} saw the disciple whom Jesus loved following \textit{them} (who also leaned back on his chest at the dinner and said, “Lord, who is the one betraying you?”)
\verse So \textit{when he} saw this one, Peter said to Jesus, “Lord, but what about this one?”
\verse Jesus said to him, “If I want him to remain until I come, what \textit{is that} to you? You follow me!”
\verse So this saying went out to the brothers that that disciple would not die. But Jesus did not say to him that he would not die, but “If I want him to remain until I come, what \textit{is that} to you?”
\verseWithHeading{A Concluding Word of Testimony} This is the disciple who is testifying about these \textit{things}, and who has written down these \textit{things}. And we know that his testimony is true.
\verse Now there are also many other \textit{things} that Jesus did, which—if they were written down one after the other—I suppose not even the world itself could contain the books that would be written.
\end{biblechapter}

