\biblebook{Luke}

\begin{biblechapter} % Luke 1
\verseWithHeading{The Preface to Luke’s Gospel} Since many have attempted to compile an account concerning the events that have been fulfilled among us,
\verse just as those who were eyewitnesses and servants of the word from the beginning passed on to us,
\verse it seemed best to me also—because I\lebnote{:|NP|:﻿*Here “because” is supplied as a component of the participle (“have followed”) which is understood as causal} have followed all things carefully from the beginning—to write them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} down in orderly sequence for you, most excellent Theophilus,
\verse so that you may know the certainty concerning the things about which you were taught.
\verseWithHeading{The Prediction of John the Baptist’s Birth} It happened that in the days of Herod, king of Judea, there was a certain priest, Zechariah by name, of the division of Abijah. And \textit{he had a wife}\lebnote{Literally “a wife to him”} from the daughters of Aaron, and her name was Elizabeth.
\verse And they were both righteous in the sight of God, living blamelessly in all the commandments and regulations of the Lord.
\verse And \textit{they did not have}\lebnote{Literally “there was not to them”} a child, because Elizabeth was barren. And they were both advanced \textit{in years}.\lebnote{Literally “in their days”}
\verse And it happened that while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal infinitive (“was serving as priest”)} he was serving as priest before God in the order of his division,
\verse according to the custom of the priesthood he was chosen by lot to enter into the temple of the Lord to burn incense.
\verse And the whole crowd of the people were praying outside at the hour of the incense offering.
\verse And an angel of the Lord appeared to him, standing at the right side of the altar of incense.
\verse And Zechariah was terrified when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw the angel,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} and fear fell upon him.
\verse But the angel said to him,
\verse “Do not be afraid, Zechariah, 
because your prayer has been heard, 
and your wife Elizabeth will bear you a son, 
and you will call his name John.
\verse And \textit{you will experience joy and exultation},\lebnote{Literally “joy and exultation will be to you”} 
and many will rejoice at his birth.
\verse For he will be great in the sight of the Lord, 
and he must never drink wine or beer, 
and he will be filled with the Holy Spirit 
while he is\lebnote{:|NP|:﻿*Here the phrase “while he is,” including the verb, is understood in Greek and is supplied in the translation} still in his mother’s womb.
\verse And he will turn many of the sons of Israel 
to the Lord their God.
\verse And Zechariah said to the angel, “By what will I know this? For I am an old man, and my wife is advanced \textit{in years}!”\lebnote{Literally “in her days”}
\verse And the angel answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to him, “I am Gabriel, who stands in the presence of God, and I was sent to speak to you and to announce to you this good news.
\verse And behold, you will be silent and not able to speak until the day these things take place, \textit{because}\lebnote{Literally “in return for which”} you did not believe my words, which will be fulfilled in their time.”
\verse And the people were waiting for Zechariah, and began to wonder\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to wonder”)} when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal infinitive (“was delayed”)} he was delayed in the temple.
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“came out”) which is understood as temporal} came out he was not able to speak to them, and they realized that he had seen a vision in the temple. And he kept making signs to them, and remained unable to speak.
\verse And it happened that when the days of his service came to an end, he went away to his home.
\verse Now after these days, his wife Elizabeth conceived, and she kept herself in seclusion for five months, saying,
\verse “Thus the Lord has done for me in the days in which he has concerned himself with me,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} to take away my disgrace among people.”
\verseWithHeading{The Prediction of Jesus’ Birth} Now in the sixth month, the angel Gabriel was sent from God to a town of Galilee \textit{named}\lebnote{Literally “to which the name”} Nazareth,
\verse to a virgin legally promised in marriage to a man \textit{named}\lebnote{Literally “to whom the name”} Joseph of the house of David. And the name of the virgin was Mary.
\verse And he came to her and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came”) has been translated as a finite verb} said, “Greetings, favored one! The Lord is with you.”
\verse But she was greatly perplexed at the statement, and was pondering what sort of greeting this might be.
\verse And the angel said to her,
\verse “Do not be afraid, Mary, for you have found favor with God.
\verse And behold, you will conceive in the womb and will give birth to a son, 
and you will call his name Jesus.
\verse This one will be great, and he will be called the Son of the Most High, 
and the Lord God will give him the throne of his father David.
\verse And Mary said to the angel, “How will this be, since I have not had sexual relations with a man?”
\verse And the angel answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to her, “The Holy Spirit will come upon you, 
and the power of the Most High will overshadow you. 
Therefore also the one to be born will be called holy, the Son of God.
\verse And behold, your relative Elizabeth—she also has conceived a son in her old age, and this is the sixth month for her who was called barren.
\verse For \textit{nothing will be impossible with God}.”\lebnote{Literally “every thing will not be impossible with God”}
\verse So Mary said, “Behold, the Lord’s female slave! May it happen to me according to your word.” And the angel departed from her.
\verseWithHeading{Mary Visits Elizabeth} Now in those days Mary set out and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“set out”) has been translated as a finite verb} traveled with haste into the hill country, to a town of Judah,
\verse and entered into the house of Zechariah, and greeted Elizabeth.
\verse And it happened that when Elizabeth heard the greeting of Mary, the baby in her womb leaped and Elizabeth was filled with the Holy Spirit.
\verse And she cried out with a loud shout and said, “Blessed are you among women, 
and blessed is the fruit of your womb!
\verse And why is this granted to me, that the mother of my Lord should come to me?
\verse For behold, when the sound of your greeting came to my ears, the baby in my womb leaped for joy!
\verse And blessed is she who believed that there will be a fulfillment to what was spoken to her from the Lord!”
\verseWithHeading{Mary’s Hymn of Praise to God} And Mary said,
\verse “My soul exalts the Lord,
\verse and my spirit has rejoiced greatly in God my Savior,
\verse because he has looked upon the humble state of his female slave, 
for behold, from now on all generations will consider me blessed,
\verse because the Mighty One has done great things for me, 
and holy is his name.
\verse And his mercy is for generation after generation 
to those who fear him.
\verse He has done a mighty deed with his arm; 
he has dispersed the proud in the thoughts of their hearts.
\verse He has brought down rulers from their thrones, 
and has exalted the lowly.
\verse He has filled those who are hungry with good things, 
and those who are rich he has sent away empty-handed.
\verse He has helped Israel his servant, 
remembering his mercy,
\verse And Mary stayed with her about three months, and returned to her home.
\verseWithHeading{The Birth of John the Baptist} Now the time came for Elizabeth that she should give birth, and she gave birth to a son.
\verse And her neighbors and relatives heard that \textit{the Lord had shown his great mercy to her},\lebnote{Literally “the Lord had made great his mercy with her”} and they rejoiced with her.
\verse And it happened that on the eighth day they came to circumcise the child, and they were wanting to name him \textit{after}\lebnote{Literally “in the name of”} his father Zechariah.
\verse And his mother answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “No, but he will be named John.”
\verse And they said to her, “There is no one of your relatives who is called by this name.”
\verse So they made signs to his father asking what he wanted him to be named,
\verse and he asked for a writing tablet and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“asked for”) has been translated as a finite verb} wrote, saying, “John is his name.” And they were all astonished.
\verse And his mouth and his tongue were opened immediately, and he began to speak,\lebnote{:|NP|:*The imperfect tense has been translated as ingressive here (“began to speak”)} praising God.
\verse And fear came on all those who lived near them, and in all the hill country of Judea all these events were discussed.
\verse And all those who heard kept these things\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} in their hearts, saying, “What then will this child be? For indeed the hand of the Lord was with him!”
\verseWithHeading{The Praise and Prophecy of Zechariah} And his father Zechariah was filled with the Holy Spirit and prophesied, saying,
\verse “Blessed be the Lord, the God of Israel, 
because he has visited to help and \textit{has redeemed}\lebnote{Literally “has done redemption for”} his people,
\verse and has raised up a horn of salvation for us 
in the house of his servant David,
\verse just as he spoke through the mouth of his holy prophets from earliest times—
\verse salvation from our enemies and from the hand of all those who hate us,
\verse to show mercy to our fathers 
and to remember his holy covenant,
\verse the oath that he swore to Abraham our father, 
to grant us
\verse that we, being rescued from the hand of our enemies, 
could serve him without fear
\verse in holiness and righteousness 
before him all our days.
\verse And so you, child, will be called the prophet of the Most High, 
for you will go on before the Lord to prepare his ways,
\verse to give knowledge of salvation to his people 
by the forgiveness of their sins,
\verse because of the merciful compassion\lebnote{Or “heart”} of our God 
by which the dawn will visit to help us from on high,
\verse to give light to those who sit in darkness and in the shadow of death, 
to direct our feet into the way of peace.”
\verse And the child kept growing and becoming strong in spirit, and was in the wilderness until the day of his public appearance to Israel.
\end{biblechapter}

\begin{biblechapter} % Luke 2
\verseWithHeading{The Birth of Jesus Christ} Now it happened that in those days a decree went out from Caesar\lebnote{Or “the emperor”} Augustus to register all the empire.
\verse (This first registration took place when\lebnote{:|NP|:﻿*Or perhaps “before”; here “when” is supplied as a component of the temporal genitive absolute participle (“was governor”)} Quirinius was governor of Syria.)
\verse And everyone went to be registered, each one to his own town.
\verse So Joseph also went up from Galilee, from the town of Nazareth, to Judea, to the city of David which is called Bethlehem, because he was of the house and family line of David,
\verse to be registered together with Mary, who was legally promised in marriage to him and\lebnote{:|NP|:﻿*Here “and” is supplied in keeping with English style} was pregnant.
\verse And it happened that while they were there, \textit{the time came}\lebnote{Literally “the days were completed”} for her to give birth.
\verse And she gave birth to her firstborn son, and wrapped him in strips of cloth and laid him in a manger, because there was no place for them in the inn.
\verseWithHeading{The Shepherds and the Angels} And there were shepherds in the same region, living out of doors and keeping watch, guarding over their flock by night.
\verse And an angel of the Lord stood near them, and the glory of the Lord shone around them, and \textit{they were terribly frightened}.\lebnote{Literally “they were afraid with great fear”}
\verse And the angel said to them, “Do not be afraid, for behold, I bring good news to you of great joy which will be for all the people:
\verse that today a Savior, who is Christ the Lord, was born for you in the city of David.
\verse And this will be the sign for you: you will find the baby wrapped in strips of cloth and lying in a manger.”
\verse And suddenly there was with the angel a multitude of the heavenly army, praising God and saying,
\verse “Glory to God in the highest, 
and on earth peace 
among people \textit{with whom he is pleased}!”\lebnote{Literally “of good pleasure”}
\verse And it happened that when the angels had departed from them into heaven, the shepherds began to say\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to say”)} to one another, “Let us go now to Bethlehem and see this thing that has happened, which the Lord has revealed to us!”
\verse And they went hurrying and found both Mary and Joseph, and the baby who was lying in the manger.
\verse And when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} they made known the statement that had been told to them about this child.
\verse And all who heard it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} were astonished concerning what had been said to them by the shepherds.
\verse But Mary treasured up all these words, pondering them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} in her heart.
\verse And the shepherds returned, glorifying and praising God for all that they had heard and seen, just as it had been told to them.
\verse And when eight days were completed \textit{so that he could be circumcised},\lebnote{Literally “to circumcise him”} he was named Jesus, his name that he was called by the angel before he was conceived in the womb.
\verseWithHeading{The Presentation of Jesus at the Temple} And when the days of their purification were completed according to the law of Moses, they brought him up to Jerusalem to present him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} to the Lord
\verse (just as it is written in the law of the Lord, “Every male that opens the womb will be called holy to the Lord”)\lebnote{An allusion to Exod 13:2, 12, 15}
\verse and to offer a sacrifice according to what was stated in the law of the Lord, “a pair of turtledoves or two young pigeons.”\lebnote{:|NP|:A quotation from Lev 5:11; 12:8}
\verseWithHeading{The Prophecy of Simeon} And behold, there was a man in Jerusalem \textit{whose name was}\lebnote{Literally “to whom the name”} Simeon, and this man was righteous and devout, looking forward to the consolation of Israel, and the Holy Spirit was upon him.
\verse And it had been revealed to him by the Holy Spirit that he would not see death before he would see the Lord’s Christ.\lebnote{Or “Messiah”}
\verse And he came in the Spirit into the temple, and when the parents brought in the child Jesus so that they could do for him according to what was customary under the law,
\verse he took him in his\lebnote{Literally “the”; the Greek article is used here as a possessive pronoun} arms and praised God and said,
\verse “Now dismiss your slave in peace, Lord, 
according to your word.
\verse For my eyes have seen your salvation
\verse that you have prepared in the presence of all the peoples,
\verse a light for revelation to the Gentiles, 
and glory to your people Israel.”
\verse And his father and mother were astonished at what was said about him.
\verse And Simeon blessed them and said to his mother Mary, “Behold, this child is appointed for the fall and rise of many in Israel, and for a sign that is opposed\lebnote{Or “rejected”}—
\verse and a sword will pierce your own soul also, so that the thoughts of many hearts will be revealed!”
\verseWithHeading{The Testimony of Anna} And there was a prophetess, Anna the daughter of Phanuel of the tribe of Asher (she was advanced \textit{in years},\lebnote{Literally “with many days”} having lived with her husband seven years \textit{after her marriage},\lebnote{Literally “from her virginity”}
\verse and herself as a widow up to eighty-four years)\lebnote{Or “eighty-four years as a widow”} who did not depart from the temple with fastings and prayers, serving night and day.
\verse And at that same hour she approached and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“approached”) has been translated as a finite verb} began to give thanks\lebnote{The imperfect tense has been translated as ingressive here (“began to give thanks”)} to God, and to speak about him to all those who were waiting for the redemption of Jerusalem.
\verse And when they had completed everything according to the law of the Lord, they returned to Galilee, to their own town of Nazareth.
\verse And the child was growing and becoming strong, filled with wisdom, and the favor of God was upon him.
\verseWithHeading{Jesus in the Temple at Twelve Years Old} And his parents went every year to Jerusalem for the feast of the Passover.
\verse And when he was twelve years old, they went up according to the custom of the feast.
\verse And after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“were completed”) which is understood as temporal} the days were completed, while they were returning, the boy Jesus stayed behind in Jerusalem. And his parents did not know it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse but believing him to be in the group of travelers, they went a day’s journey. And they began searching for\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began searching for”)} him among their\lebnote{Literally “the”; the Greek article is used here as a possessive pronoun} relatives and their\lebnote{Literally “the”; the Greek article is used here as a possessive pronoun} acquaintances,
\verse and when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“find”) which is understood as temporal} did not find him,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} they returned to Jerusalem to search for him.
\verse And it happened that after three days they found him in the temple courts,\lebnote{:|NP|:*Here “courts” is supplied to distinguish this area from the interior of the temple building itself} sitting in the midst of the teachers and listening to them and asking them questions.
\verse And all who heard him were amazed at his insight and his\lebnote{:|NP|:﻿*This is an understood repetition of “his” due to English style} answers.
\verse And when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw him, they were astounded and his mother said to him, “Child, why have you done this to us? Look, your father and I have been searching for you anxiously!”
\verse And he said to them, “\textit{Why}\lebnote{Literally “what is it that”} were you searching for me? Did you not know that it was necessary for me to be in the house\lebnote{Or “things” (= business)} of my Father?”
\verse And they did not understand the statement that he spoke to them.
\verse And he went down with them and came to Nazareth, and was submitting to them. And his mother treasured all these things in her heart.
\verse And Jesus was advancing in wisdom and stature and in favor with God and with people.
\end{biblechapter}

\begin{biblechapter} % Luke 3
\verseWithHeading{John the Baptist Begins His Ministry} Now in the fifteenth year of the reign of Tiberius Caesar,\lebnote{Or “the emperor Tiberius”} when Pontius Pilate was governor of Judea, and Herod was tetrarch of Galilee, and his brother Philip was tetrarch of the region of Iturea and Trachonitis, and Lysanias was tetrarch of Abilene,
\verse in the time of the high priest Annas and Caiaphas, the word of God came to John the son of Zechariah in the wilderness.
\verse And he went into all the surrounding region of the Jordan, preaching a baptism of repentance for the forgiveness of sins,
\verse as it is written in the book of the words of the prophet Isaiah,
\verse “The voice of one crying out in the wilderness, 
‘Prepare the way of the Lord, 
make his paths straight!
\verse Every valley will be filled, 
and every mountain and hill will be leveled, 
and the crooked will become straight, 
and the rough road will become\lebnote{:|NP|:﻿*Here “will become” is an implied repetition of the verb earlier in the verse} smooth,
\verse Therefore he was saying to the crowds that came out to be baptized by him, “Offspring of vipers! Who warned you to flee from the coming wrath?
\verse Therefore produce fruit worthy of repentance! And do not begin to say to yourselves, ‘We have Abraham as father.’ For I say to you that God is able to raise up children for Abraham from these stones!
\verse And even now the ax is positioned at the root of the trees; therefore every tree not producing good fruit is cut down and thrown into the fire.”
\verse And the crowds were asking him, saying, “What then should we do?”
\verse And he answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to them, “The one who has two tunics must share with the one who does not have one,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} and the one who has food must do likewise.”
\verse And tax collectors also came to be baptized, and they said to him, “Teacher, what should we do?”
\verse And he said to them, “Collect no more than \textit{what you are ordered to}.”\lebnote{Literally “what is ordered to you”}
\verse And those who served in the army were also asking him, saying, “What should we also do?” And he said to them, “Extort from no one, and do not blackmail anyone,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} and be content with your pay.”
\verse And while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“were waiting expectantly”)} the people were waiting expectantly and all were pondering in their hearts concerning John, whether perhaps he might be the Christ,\lebnote{Or “Messiah”}
\verse John answered them all, saying, “I baptize you with water, but the one who is more powerful than I am is coming, of whom I am not worthy to untie the strap of his sandals. He will baptize you with the Holy Spirit and fire.
\verse His winnowing shovel is in his hand, to clean out his threshing floor and to gather the wheat into his storehouse, but he will burn up the chaff with unquenchable fire.”
\verse So with many other exhortations also he proclaimed good news to the people.
\verse But Herod the tetrarch, who had been reproved by him concerning Herodias, his brother’s wife, and concerning all the evil deeds that Herod had done,
\verse added this also to them all: he also locked up John in prison.
\verseWithHeading{The Baptism of Jesus} Now it happened that when all the people were baptized, Jesus also was baptized, and while he\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“was praying”)} was praying, heaven was opened,
\verse and the Holy Spirit descended on him in bodily form like a dove, and a voice came from heaven, “You are my beloved Son; with you I am well pleased.”
\verseWithHeading{The Genealogy of Jesus Christ} And Jesus, when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“began”) which is understood as temporal} began his ministry,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} was himself about thirty years old, being the son (as it was believed) of Joseph the son of Eli,
\verse the son of Matthat, the son of Levi, the son of Melchi, the son of Jannai, the son of Joseph,
\verse the son of Mattathias, the son of Amos, the son of Nahum, the son of Esli, the son of Naggai,
\verse the son of Maath, the son of Mattathias, the son of Semein, the son of Josech, the son of Joda,
\verse the son of Joanan, the son of Rhesa, the son of Zerubbabel, the son of Shealtiel, the son of Neri,
\verse the son of Melchi, the son of Addi, the son of Cosam, the son of Elmadam, the son of Er,
\verse the son of Joshua, the son of Eliezer, the son of Jorim, the son of Matthat, the son of Levi,
\verse the son of Simeon, the son of Judah, the son of Joseph, the son of Jonam, the son of Eliakim,
\verse the son of Melea, the son of Menna, the son of Mattatha, the son of Nathan, the son of David,
\verse the son of Jesse, the son of Obed, the son of Boaz, the son of Sala, the son of Nahshon,
\verse the son of Amminadab, the son of Admin, the son of Arni, the son of Hezron, the son of Perez, the son of Judah,
\verse the son of Jacob, the son of Isaac, the son of Abraham, the son of Terah, the son of Nahor,
\verse the son of Serug, the son of Reu, the son of Peleg, the son of Eber, the son of Shelah,\lebnote{Greek “Sala”}
\verse the son of Cainan, the son of Arphaxad, the son of Shem, the son of Noah, the son of Lamech,
\verse the son of Methuselah, the son of Enoch, the son of Jared, the son of Mahalaleel, the son of Cainan,
\verse the son of Enosh, the son of Seth, the son of Adam, the son of God.
\end{biblechapter}

\begin{biblechapter} % Luke 4
\verseWithHeading{The Temptation of Jesus} And Jesus, full of the Holy Spirit, returned from the Jordan and was led by the Spirit in the wilderness
\verse forty days, being tempted by the devil. And he ate nothing during those days, and when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“were completed”)} they were completed, he was hungry.
\verse So the devil said to him, “If you are the Son of God, order this stone that it become bread!”
\verse And Jesus replied to him, “It is written, ‘Man will not live on bread alone.’ ”\lebnote{:|NP|:A quotation from Deut 8:3; most manuscripts add “but by every word of God” here}
\verse And he led him up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“led … up”) has been translated as a finite verb} showed him all the kingdoms of the world in a moment of time.
\verse And the devil said to him, “I will give you all this domain and their glory, because it has been handed over to me, and I can give it to whomever I want.
\verse So if you will worship before me, all this will be yours.”
\verse And Jesus answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to him, “It is written, ‘You shall worship the Lord your God, and serve only him.’ ”\lebnote{:|NP|:A quotation from Deut 6:13}
\verse And he brought him to Jerusalem, and had him stand on the highest point of the temple and said to him, “If you are the Son of God, throw yourself down from here,
\verse for it is written, ‘He will command his angels concerning you, 
to protect you,’\lebnote{:|NP|:A quotation from Ps 91:11}
\verse and ‘on their hands they will lift you up, 
lest you strike your foot against a stone.’ ”\lebnote{:|NP|:A quotation from Ps 91:12}
\verse And Jesus answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to him, “It is said, ‘You are not to put to the test the Lord your God.’ ”\lebnote{:|NP|:A quotation from Deut 6:16}
\verse And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“had completed”) which is understood as temporal} the devil had completed every temptation, he departed from him until a favorable time.\lebnote{Or “for a while”}
\verseWithHeading{Public Ministry in Galilee} And Jesus returned in the power of the Spirit to Galilee, and news about him went out throughout all the surrounding region.
\verse And he began to teach\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to teach”)} in their synagogues, and\lebnote{:|NP|:﻿*Here the participle (“was praised”) is translated as a finite verb because of English style} was praised by all.
\verseWithHeading{Rejected at Nazareth} And he came to Nazareth, where \textit{he had been brought up},\lebnote{Literally “he was having been brought up”} and according to \textit{his custom}\lebnote{Literally “what he was accustomed to for him”} he entered into the synagogue on the day of the Sabbath and stood up to read.
\verse And the scroll of the prophet Isaiah was given to him, and unrolling the scroll he found the place where it was written,
\verse “The Spirit of the Lord is upon me, 
because of which he has anointed me 
to proclaim good news to the poor. 
He has sent me 
to proclaim release to the captives, 
and recovery of sight to the blind, 
to send out in freedom those who are oppressed,
\verse to proclaim the favorable year of the Lord.”\lebnote{:|NP|:A quotation from Isa 61:1–2, with one line from Isa 58:6}
\verse And he rolled up the scroll and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“rolled up”) has been translated as a finite verb} gave it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} back to the attendant and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“gave … back”) has been translated as a finite verb} sat down. And the eyes of everyone in the synagogue were looking intently at him.
\verse And he began to say to them, “Today this scripture has been fulfilled in your hearing.”
\verse And they were all speaking well of him, and were astonished at the gracious words that were coming out of his mouth. And they were saying, “Is this man not the son of Joseph?”
\verse And he said to them, “Doubtless you will tell me this parable: ‘Physician, heal yourself!’ Whatever we have heard that took place in Capernaum, do here in your hometown also!”
\verse And he said, “Truly I say to you that no prophet is acceptable in his own hometown.
\verse But in truth I say to you, there were many widows in Israel in the days of Elijah, when the sky was shut for three years and six months while a great famine took place over all the land.
\verse And Elijah was sent to none of them, but only to Zarephath in the region of Sidon, to a woman who was a widow.
\verse And there were many lepers in Israel in the time of the prophet Elisha, and none of them was made clean except Naaman the Syrian.”
\verse And all those in the synagogue were filled with anger when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“heard”) which is understood as temporal} heard these things.
\verse And they stood up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“stood up”) has been translated as a finite verb} forced him out of the town and brought him up to the edge of the hill on which their town was built, so that they could throw him down the cliff.
\verse But he passed through their midst and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“passed”) has been translated as a finite verb} went on his way.
\verseWithHeading{Jesus Teaches and Heals Many in Capernaum} And he came down to Capernaum, a town of Galilee, and was teaching them on the Sabbath.
\verse And they were astounded at his teaching, because \textit{he spoke}\lebnote{Literally “his word was”} with authority.
\verse And in the synagogue there was a man who had the spirit of an unclean demon,\lebnote{Or “an unclean demonic spirit”} and he cried out with a loud voice,
\verse “Ha! \textit{Leave us alone},\lebnote{Literally “what to us and to you”} Jesus the Nazarene! Have you come to destroy us? I know who you are—the Holy One of God!”
\verse And Jesus rebuked him, saying, “Be silent and come out of him!” And after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“throwing … down”) which is understood as temporal} throwing him down in their midst, the demon came out of him without hurting him at all.
\verse And amazement came upon them all, and they began to talk\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to talk”)} with one another, saying, “What word\lebnote{Or “command”} is this? For he commands the unclean spirits with authority and power, and they come out!”
\verse And news about him went out into every place of the surrounding region.
\verse And after he\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“set out”) which is understood as temporal} set out from the synagogue, he went into Simon’s house. And Simon’s mother-in-law was afflicted with a high fever, and they asked him on behalf of her.
\verse And he stood over her and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“stood”) has been translated as a finite verb} rebuked the fever, and it left her. And immediately she got up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“got up”) has been translated as a finite verb} began to serve\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to serve”)} them.
\verse Now as\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the temporal genitive absolute participle (“was setting”)} the sun was setting, all who had those who were sick with various diseases brought them to him, and placing his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} hands on every one of them, he healed them.
\verse And demons also were coming out of many, crying out and saying, “You are the Son of God!” And he rebuked them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} and did not permit them to speak, because they knew that he was the Christ.\lebnote{Or “Messiah”}
\verse And when it\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“was”)} was day, he departed and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“departed”) has been translated as a finite verb} went to an isolated place. And the crowds were seeking him, and came to him and were trying to prevent him from departing from them.
\verse But he said to them, “It is necessary for me to proclaim the good news of the kingdom of God to the other towns also, because I was sent for this purpose.
\verse And he was preaching in the synagogues of Judea.
\end{biblechapter}

\begin{biblechapter} % Luke 5
\verseWithHeading{Jesus Calls His First Disciples} Now it happened that while the crowd was pressing around him and hearing the word of God, he was standing beside the lake of Gennesaret,\lebnote{Another name for the Sea of Galilee}
\verse and he saw two boats there beside the lake, but the fishermen had gotten out of them and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“had gotten out”) has been translated as a finite verb} were washing their nets.
\verse And he got into one of the boats, which was Simon’s, and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“got”) has been translated as a finite verb} asked him to put out from the land a little. And he sat down and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“sat down”) has been translated as a finite verb} began to teach\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to teach”)} the crowds from the boat.
\verse And when he stopped speaking, he said to Simon, “Put out into the deep water and let down your nets for a catch.”
\verse And Simon answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “Master, although we\lebnote{:|NP|:﻿*Here “although” is supplied as a component of the participle (“worked hard”) which is understood as concessive} worked hard through the whole night, we caught nothing. But at your word I will let down the nets.”
\verse And when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“did”) which is understood as temporal} did this, they caught a very large number of fish, and their nets began to tear.\lebnote{:|NP|:*The imperfect tense has been translated as ingressive here (“began to tear”)}
\verse And they signaled to their partners in the other boat to come and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“come”) has been translated as a finite verb} help them, and they came and filled both the boats so that they began to sink.
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} Simon Peter fell down at Jesus’ knees, saying, “Depart from me, Lord, because I am \textit{a sinful man}!”\lebnote{Literally “a man, a sinner”}
\verse For amazement had seized him and all those who were with him at the catch of fish that they had caught,
\verse and so also were James and John, the sons of Zebedee, who were business partners with Simon. And Jesus said to Simon, “Do not be afraid! From now on you will be catching people!”
\verse And after they\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“brought”) which is understood as temporal} brought their\lebnote{Literally “the”; the Greek article is used here as a possessive pronoun} boats to the land, they left everything and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“left”) has been translated as a finite verb} followed him.
\verseWithHeading{A Leper Cleansed} And it happened that while he was in one of the towns, \textit{there was}\lebnote{Literally “behold”} a man \textit{covered with leprosy}.\lebnote{Literally “full of leprosy”} And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw Jesus, he fell down on his face and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“fell down”) has been translated as a finite verb} begged him, saying, “Lord, if you are willing, you are able to make me clean.”
\verse And extending his\lebnote{Literally “the”; the Greek article is used here as a possessive pronoun} hand he touched him, saying, “I am willing; be clean.” And immediately the leprosy went away from him.
\verse And he ordered him, “Tell no one, but go and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“go”) has been translated as a finite verb} show yourself to the priest and bring the offering\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} for your cleansing just as Moses commanded, for a testimony to them.
\verse But the report about him spread even more, and large crowds were gathering to hear him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} and to be healed of their illnesses.
\verse But he himself was withdrawing in the wilderness and praying.
\verseWithHeading{A Paralytic Healed} And it happened that on one of the days as he was teaching, Pharisees and teachers of the law were sitting there who \textit{had come}\lebnote{Literally “were having come”} from every village of Galilee and Judea and from Jerusalem, and the power of the Lord was there in order for him to heal.
\verse And behold, men came carrying on a stretcher a man who was paralyzed, and they were seeking to bring him in and place him\lebnote{Some manuscripts Some manuscripts include the pronoun “him” after “place”}﻿\lebnote{Since Greek routinely omits direct objects when they are clear from context, the pronoun is not necessary here in the Greek text, but it must be supplied in the English translation} before him.
\verse And when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“find”) which is understood as temporal} did not find a way to bring him in because of the crowd, they went up on the roof and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“went up”) has been translated as a finite verb} let him down through the roof tiles with the stretcher into the midst of them, in front of Jesus.
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw their faith, he said, “Friend, your sins are forgiven you.”
\verse And the scribes and the Pharisees began to reason, saying, “Who is this man who speaks blasphemies? Who is able to forgive sins except God alone?”
\verse But Jesus, perceiving their thoughts, answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to them, “Why are you reasoning in your hearts?
\verse Which is easier to say, ‘Your sins are forgiven you,’ or to say, ‘Get up and walk?’
\verse But in order that you may know that the Son of Man has authority on earth to forgive sins,” he said to the one who was paralyzed, “I say to you, ‘Get up and pick up your stretcher and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“pick up”) has been translated as a finite verb} go to your home.’ ”
\verse And immediately he stood up before them, picked up what he had been lying on, and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participles (“stood up” and “picked up”) have been translated as finite verbs} went away to his home, glorifying God.
\verse And amazement seized them all, and they began to glorify\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to glorify”)} God. And they were filled with fear, saying, “We have seen wonderful things today!”
\verseWithHeading{Levi Called to Follow Jesus} And after these things, he went out and saw a tax collector \textit{named}\lebnote{Literally “by name”} Levi sitting at the tax booth, and he said to him, “Follow me!”
\verse And leaving everything behind, he got up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“got up”) has been translated as a finite verb} began to follow\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to follow”)} him.
\verse And Levi gave a great banquet for him in his house, and there was a large crowd of tax collectors and others who were reclining for the meal with them.
\verse And the Pharisees and their scribes began to complain\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to complain”)} to his disciples, saying, “Why do you eat and drink with the tax collectors and sinners?”
\verse And Jesus answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to them, “Those who are healthy do not have need of a physician, but those \textit{who are sick}.\lebnote{Literally “having badly”}
\verse I have not come to call the righteous but sinners to repentance.”
\verseWithHeading{On Fasting} And they said to him, “The disciples of John fast often and make prayers—likewise also the disciples\lebnote{:|NP|:﻿*The word “disciples” is not in the Greek text but is implied} of the Pharisees—but yours are eating and drinking!”
\verse So he\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} said\lebnote{Some manuscripts have “So Jesus said”} to them, “You are not able to make the \textit{bridegroom’s attendants}\lebnote{Literally “sons of the bridal chamber”} fast as long as the bridegroom is with them, are you?\lebnote{:|NP|:*The negative construction in Greek anticipates a negative answer here, indicated in the translation by the phrase “are you”}
\verse But days will come, and when the bridegroom is taken away from them, then they will fast in those days.”
\verse And he also told a parable to them: “No one tears a patch from a new garment and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“tears”) has been translated as a finite verb} puts it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} on an old garment. Otherwise, he will have torn the new also, and the old will not match the patch that is from the new.
\verse And no one pours new wine into old wineskins. Otherwise, the new wine will burst the wineskins, and it will be spilled and the wineskins will be destroyed.
\verse But new wine must be put into new wineskins.
\verse And no one after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“drinking”) which is understood as temporal} drinking old wine\lebnote{:|NP|:﻿*The word “wine” is not in the Greek text but is implied} wants new, because he says, ‘The old is just fine!’ ”
\end{biblechapter}

\begin{biblechapter} % Luke 6
\verseWithHeading{Plucking Grain on the Sabbath} Now it happened that on a Sabbath he went through the grain fields, and his disciples were picking and eating the heads of grain, rubbing them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} in their\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} hands.
\verse But some of the Pharisees said, “Why are you doing what is not permitted on the Sabbath?
\verse And Jesus answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to them, “Have you not read this, what David did when he and those who were with him were hungry—
\verse how he entered into the house of God and took the bread of the presentation, which it is not permitted to eat (except the priests alone), and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took”) has been translated as a finite verb} ate it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} and gave it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} to those with him?”
\verse And he said to them, “The Son of Man is Lord of the Sabbath.”
\verseWithHeading{A Man with a Withered Hand Healed} Now it happened that on another Sabbath he entered into the synagogue and was teaching, and a man was there, and his right hand was withered.
\verse So the scribes and the Pharisees were watching closely\lebnote{Some manuscripts have “were watching him closely”} to see if he would heal on the Sabbath, in order that they could find a reason\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} to accuse him.
\verse But he knew their thoughts and said to the man who had the withered hand, “Get up and stand in the middle,” and he got up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“get up”) has been translated as a finite verb} stood there.
\verse And Jesus said to them, “I ask you whether it is permitted on the Sabbath to do good or to do evil, to save a life or to destroy it?”\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“looking around”) which is understood as temporal} looking around at them all, he said to him, “Stretch out your hand,” and he did, and his hand was restored.
\verse But they were filled with fury, and began discussing\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began discussing”)} with one another what they might do to Jesus.
\verseWithHeading{The Selection of the Twelve Apostles} Now it happened that in these days he went away to the mountain to pray, and was spending the whole night in prayer to God.
\verse And when day came, he summoned his disciples and chose from them twelve, whom he also named apostles:
\verse Simon (whom he also named Peter) and his brother Andrew, and James, and John, and Philip, and Bartholomew,
\verse and Matthew, and Thomas, and James the son of Alphaeus, and Simon who was called the Zealot,
\verse and Judas the son of James, and Judas Iscariot, who became a traitor.
\verseWithHeading{The Sermon on the Plain: The Beatitudes} And he came down with them and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came down”) has been translated as a finite verb} stood on a level place, and a large crowd of his disciples and a great multitude of people from all of Judea and Jerusalem and the seacoast district of Tyre and Sidon,
\verse who came to hear him and to be healed of their diseases, and those who were troubled by unclean spirits were cured.
\verse And the whole crowd was seeking to touch him, because power was going out from him and healing them all.
\verse And he lifted up his eyes to his disciples and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“lifted up”) has been translated as a finite verb} said,
\verse “Blessed are the poor, 
because yours is the kingdom of God.
\verse Blessed are those who are hungry now, 
because you will be satisfied. 
Blessed are those who weep now, 
Because you will laugh.
\verse Blessed are you when people hate you, and when they exclude you and revile you and spurn your name as evil on account of the Son of Man.
\verseWithHeading{The Sermon on the Plain: Woes} “But woe to you who are rich, 
because you have received your comfort.
\verse Woe to you who are satisfied now, 
because you will be hungry. 
Woe, you who laugh now, 
because you will mourn and weep.
\verse Woe whenever all people speak well of you, 
for their fathers used to do the same things to the false prophets.
\verseWithHeading{The Sermon on the Plain: Love for Enemies} “But to you who are listening I say: Love your enemies, do good to those who hate you,
\verse bless those who curse you, pray for those who mistreat you.
\verse To the one who strikes you on the cheek, offer the other also, and from the one who takes away your cloak, do not withhold your tunic also.
\verse Give to everyone who asks you, and from the one who takes away your things, do not ask for them back.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And just as you want people \textit{to do}\lebnote{Literally “would do”} to you, do \textit{the same}\lebnote{Literally “likewise”} to them.
\verse “And if you love those who love you, what kind of credit is that to you? For even sinners love those who love them!
\verse And if\lebnote{Some manuscripts have “For even if”} you do good to those who do good to you, what kind of credit is that to you? Even the sinners do the same!
\verse And if you lend to those from whom you expect to receive back, what kind of credit is that to you? Even sinners lend to sinners, so that they may get back an equal amount!
\verse But love your enemies, and do good, and lend expecting back nothing, and your reward will be great, and you will be sons of the Most High, because he is kind to the ungrateful and wicked.
\verse Be merciful, just as your Father is merciful!\lebnote{Some manuscripts have “also is merciful”}
\verseWithHeading{The Sermon on the Plain: On Judging Others} “And do not judge, and you will never be judged. And do not condemn, and you will never be condemned. Pardon, and you will be pardoned.
\verse Give, and it will be given to you, a good measure—pressed down, shaken, overflowing—they will pour out into your lap. For with the measure by which you measure out, it will be measured out to you in return.”
\verse And he also told them a parable: “Surely a blind person cannot lead the blind, can he?\lebnote{:|NP|:*The negative construction in Greek anticipates a negative answer here, indicated in the translation by the phrase “can he”} Will they not both fall into a pit?
\verse A disciple is not superior to his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} teacher, but everyone, when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“is fully trained”) which is understood as temporal} is fully trained, will be like his teacher.
\verse And why do you see the speck that is in your brother’s eye, but do not notice the beam of wood that is in your own eye?
\verse How are you able to say to your brother, “Brother, allow me to remove the speck that is in your eye,” while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the participle (“see”) which is understood as temporal} you yourself do not see the beam of wood in your own eye? Hypocrite! First remove the beam of wood from your own eye, and then you will see clearly to remove the speck that is in your brother’s eye!
\verseWithHeading{The Sermon on the Plain: Trees and Their Fruit} “For there is no good tree that produces bad fruit, nor on the other hand a bad tree that produces good fruit,
\verse for each tree is known by its own fruit. For figs are not gathered from thorn plants, nor are grapes harvested from thorn bushes.
\verse The good person out of the good treasury of his heart brings forth good, and the evil person out of his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} evil treasury\lebnote{:|NP|:﻿*The word “treasury” here is an understood repetition from earlier in the verse} brings forth evil. For out of the abundance of the heart his mouth speaks.
\verse “And why do you call me ‘Lord, Lord,’ and do not do what I tell you?\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verseWithHeading{The Sermon on the Plain: Two Houses and Two Foundations} “Everyone who comes to me and listens to my words and does them—I will show you what he is like:
\verse he is like a man building a house, who dug and went down deep and laid the foundation on the rock. And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“came”)} a flood came, the river burst against that house and was not able to shake it, because it had been built well.
\verse But the one who hears my words\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} and does not do them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} is like a man who built a house on the ground without a foundation, which the river burst against, and immediately it collapsed—and the collapse of that house was great!”
\end{biblechapter}

\begin{biblechapter} % Luke 7
\verseWithHeading{A Centurion’s Slave Healed} After he had finished all his statements in the hearing of the people, he entered into Capernaum.
\verse Now a certain centurion’s slave, who was esteemed by him, \textit{was sick}\lebnote{Literally “was having badly”} and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“was having”) has been translated as a finite verb} was about to die.
\verse So when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“heard”) which is understood as temporal} heard about Jesus, he sent Jewish elders to him, asking him that he would come and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“come”) has been translated as a finite verb} cure his slave.
\verse And when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“came”) which is understood as temporal} came to Jesus, they began imploring\lebnote{The imperfect tense has been translated as ingressive here (“began imploring”)} him earnestly, saying, “He is worthy that you grant this for him,
\verse because he loves our nation and he himself built the synagogue for us.”
\verse So Jesus went with them. Now by this time he was not far away from the house, and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“away”) has been translated as a finite verb} the centurion sent friends, saying to him, “Lord, do not trouble yourself, for I am not worthy that you should come in under my roof.
\verse For this reason neither did I consider myself worthy to come to you. But say the word and my slave must be healed.
\verse For I also am a man placed under authority, who has soldiers under me, and I say to this one, ‘Go!’ and he goes, and to another one, ‘Come!’ and he comes, and to my slave, ‘Do this!’ and he does it.”\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“heard”) which is understood as temporal} Jesus heard these things, he marveled at him, and turning around to the crowd that was following him, he said, “I tell you, not even in Israel have I found such great faith!”
\verse And when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“returned”) which is understood as temporal} returned to the house, those who had been sent found the slave healthy.
\verseWithHeading{A Widow’s Son Raised} And it happened that on the next day he went to a town called Nain, and his disciples and a large crowd went with him.
\verse And as he approached the gate of the town, behold, a man who had died was being carried out, his mother’s only son, and she was a widow. And a large crowd from the town was with her.
\verse And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} the Lord saw her, he had compassion for her and said to her, “Do not weep!”
\verse And he came up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came up”) has been translated as a finite verb} touched the bier, and those who were carrying it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} stopped. And he said, “Young man, I say to you, get up!”
\verse And the dead man sat up and began to speak, and he gave him to his mother.
\verse And fear seized them all, and they began to glorify\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to glorify”)} God, saying, “A great prophet has appeared among us!” and “God has visited to help his people!”
\verse And this report about him went out in the whole of Judea and in all the surrounding region.
\verseWithHeading{A Question from John the Baptist} And his disciples reported to John about all these things. And summoning a certain two of his disciples, John
\verse sent them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} to the Lord,\lebnote{A number of significant manuscripts read “Jesus”} saying, “Are you the one who is to come, or should we look for another?”
\verse And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“came”) which is understood as temporal} the men came to him, they said, “John the Baptist sent us to you, saying, ‘Are you the one who is to come, or should we look for another?’ ”
\verse In that hour he healed many people of diseases and suffering and evil spirits, and he granted sight to many blind people.
\verse And he answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to them, “Go and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“go”) has been translated as a finite verb} tell John what you have seen and heard: the blind receive sight, the lame walk, lepers are cleansed, the deaf hear;\lebnote{Some manuscripts have “and the deaf hear”} the dead are raised, the poor have good news announced to them.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And whoever is not offended by me is blessed.”
\verse And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“had departed”)} the messengers of John had departed, he began to speak to the crowds concerning John: “What did you go out into the wilderness to see? A reed shaken by the wind?
\verse But what did you go out to see? A man dressed in soft clothing? Behold, those who are in splendid clothing and luxury are in the royal palaces.
\verse But what did you go out to see? A prophet? Yes, I tell you, and even more than a prophet!
\verse It is this man about whom it is written: ‘Behold, I am sending my messenger before your face, 
who will prepare your way before you.’\lebnote{:|NP|:A quotation from Mal 3:1; cf. Mark 1:2; Matt 11:10}
\verse I tell you, there is no one greater among those born of women than John, but the one who is least in the kingdom of God is greater than he.
\verse (And all the people, when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“heard”) which is understood as temporal} heard this\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation}—even the tax collectors—affirmed the righteousness of God, because they\lebnote{:|NP|:﻿*Here “because” is supplied as a component of the participle (“had been baptized”) which is understood as causal} had been baptized with the baptism of John,
\verse but the Pharisees and the legal experts rejected the purpose of God for themselves, because they\lebnote{:|NP|:﻿*Here “because” is supplied as a component of the participle (“been baptized”) which is understood as causal} had not been baptized by him.)
\verse “To what then shall I compare the people of this generation, and what are they like?
\verse They are like children sitting in the marketplace and calling out to one another, who say, ‘We played the flute for you and you did not dance; 
we sang a lament and you did not weep.’
\verse For John the Baptist has come not eating bread or drinking wine, and you say, ‘He has a demon!’
\verse The Son of Man has come eating and drinking, and you say, ‘Behold, a man who is a glutton and a drunkard, a friend of tax collectors and sinners!’
\verse And wisdom is vindicated by all her children.”
\verseWithHeading{A Sinful Woman Anoints Jesus’ Feet} Now one of the Pharisees asked him to eat with him, and he entered into the house of the Pharisee and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“entered”) has been translated as a finite verb} reclined at the table.
\verse And behold, a woman in the town who was a sinner, when she\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“learned”) which is understood as temporal} learned that he was dining in the Pharisee’s house, brought an alabaster flask of perfumed oil,
\verse and standing behind him at his feet weeping, she began to wet his feet with her tears and was wiping them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} with the hair of her head and was kissing his feet and anointing them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} with the perfumed oil.
\verse Now when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} the Pharisee who invited him saw this,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} he spoke to himself, saying, “If this man were a prophet, he would have known who and what kind of woman this is who is touching him, that she is a sinner.”
\verse And Jesus answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to him, “Simon, I have something to say to you.” And he said, “Teacher, say it.”\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse “There were two debtors who owed a certain creditor. One owed five hundred denarii and the other fifty.
\verse When\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“able”)} they were not able to repay him,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} he forgave the debts\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} of both. Now which of them will love him more?”
\verse Simon answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “I suppose that it is the one to whom he forgave more.” And he said to him, “You have judged correctly.”
\verse And turning toward the woman, he said to Simon, “Do you see this woman? I entered into your house. You did not give me water for my feet, but she wet my feet with her tears and wiped them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} with her hair.
\verse You did not give me a kiss, but from the time I entered, she has not stopped kissing my feet.
\verse You did not anoint my head with olive oil, but she anointed my feet with perfumed oil.
\verse \textit{For this reason}\lebnote{Literally “on account of which”} I tell you, her sins—which were many—have been forgiven, for she loved much. But the one to whom little is forgiven loves little.”
\verse And he said to her, “Your sins are forgiven.”
\verse And those who were reclining at the table with him began to say among themselves, “Who is this who even forgives sins?”
\verse And he said to the woman, “Your faith has saved you. Go in peace.”
\end{biblechapter}

\begin{biblechapter} % Luke 8
\verseWithHeading{Some Women Accompany Jesus} And it happened that \textit{afterward}\lebnote{Literally “in what follows”} also he was going about from one town and village to another preaching and proclaiming the good news concerning the kingdom of God, and the twelve were with him,
\verse and some women who had been healed of evil spirits and diseases: Mary (who was called Magdalene), from whom seven demons had gone out,
\verse and Joanna the wife of Chuza (Herod’s household manager), and Susanna, and many others who were helping to support them from their possessions.
\verseWithHeading{The Parable of the Sower} And while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“was gathering”)} a large crowd was gathering and they were going to him from town after town, he spoke by means of a parable:
\verse “The sower went out to sow his seed, and while he was sowing, \textit{some seed}\lebnote{Literally “some of which”} fell on the side of the path and was trampled under foot, and the birds of the sky devoured it.
\verse And other seed fell on the rock, and when it\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“came up”) which is understood as temporal} came up, it withered, because it did not have moisture.
\verse And other seed fell in the midst of the thorn plants, and the thorn plants grew up with it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“grew up with”) has been translated as a finite verb} choked it.
\verse And other seed fell on the good soil, and when it\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“came up”) which is understood as temporal} came up, it produced a hundred times as much grain.” As he\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the participle (“said”) which is understood as temporal} said these things, he called out, “The one who has ears to hear, let him hear!”
\verseWithHeading{The Reason for the Parables} And his disciples asked him what this parable meant.
\verse And he said, “To you it has been given to know the mysteries of the kingdom of God, but to the rest they are in parables, so that ‘Seeing they may not see, 
and hearing they may not understand.’\lebnote{:|NP|:A quotation from Isa 6:9}
\verseWithHeading{The Parable of the Sower Interpreted} Now the parable means this: the seed is the word of God,
\verse and those beside the path are the ones who have heard. Then the devil comes and takes away the word from their heart, so that they may not believe and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“believe”) has been translated as a finite verb} be saved.
\verse And those on the rock are those who receive the word with joy when they hear it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} and these do not have enough root, who believe for a time and in a time of testing fall away.
\verse And the seed that fell into the thorn plants—these are the ones who hear and as they\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the participle (“go along”) which is understood as temporal} go along are choked by the worries and riches and pleasures of life, and they do not bear fruit to maturity.
\verse But the seed on the good soil—these are the ones who, after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“hearing”) which is understood as temporal} hearing the word, hold fast to it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} with a noble and good heart, and bear fruit with patient endurance.
\verseWithHeading{The Parable of the Lamp} “And no one, after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“lighting”) which is understood as temporal} lighting a lamp, covers it with a jar or puts it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} under a bed, but puts it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} on a lampstand, so that those who come in can see the light.
\verse For nothing is secret that will not become evident, and nothing hidden that will never be known and come to light.
\verse Therefore consider how you listen, for whoever has, to him more will be given, and whoever does not have, even what he thinks that he has will be taken away from him.”
\verseWithHeading{Jesus’ Mother and Brothers} Now his mother and brothers came to him, and they were not able to meet with him because of the crowd.
\verse And it was reported to him, “Your mother and your brothers are standing outside wanting to see you.”
\verse But he answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to them, “These are my mother and my brothers—the ones who hear the word of God and do it.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verseWithHeading{Calming of a Storm} Now it happened that on one of the days both he and his disciples got into a boat, and he said to them, “Let us cross over to the other side of the lake.” And they set sail,
\verse and as\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the temporal genitive absolute participle (“were sailing”)} they were sailing, he fell asleep. And a storm of wind came down on the lake, and they were being swamped and were in danger.
\verse And they came and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came”) has been translated as a finite verb} woke him up, saying, “Master, master! We are perishing!” So he got up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“got up”) has been translated as a finite verb} rebuked the wind and the billowing waves of water and they ceased, and it became calm.
\verse And he said to them, “Where is your faith?” But they were afraid and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“were afraid”) has been translated as a finite verb} were astonished, saying to one another, “Who then is this, that he commands even the winds and the water and they obey him?”
\verseWithHeading{A Demon-possessed Gerasene Healed} And they sailed to the region of the Gerasenes, which is opposite Galilee.
\verse And as\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the participle (“got out”) which is understood as temporal} he got out on the land, a certain man from the town met him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} who had demons and for a considerable time had not worn clothes and did not live in a house, but among the tombs.
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw Jesus, he cried out, fell down before him, and said with a loud voice, “\textit{What do I have to do with you},\lebnote{Literally “what to me and to you”} Jesus, Son of the Most High God? I beg you, do not torment me!”
\verse For he had commanded the unclean spirit to come out of the man. (For it had seized him many times, and he was bound with chains and shackles and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“was bound”) has been translated as a finite verb} was guarded, and breaking the bonds he would be driven by the demon into the deserted places.)
\verse So Jesus asked him, “What is your name?” And he said, “Legion,” because many demons had entered into him.
\verse And they began imploring\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began imploring”)} him that he would not order them to depart into the abyss.
\verse Now there was a large herd of pigs feeding there on the hill, and they implored him that he would permit them to enter into those pigs. And he permitted them.
\verse So the demons came out of the man and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came out”) has been translated as a finite verb} entered into the pigs, and the herd rushed headlong down the steep slope into the lake and were drowned.
\verse And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} the herdsmen saw what had happened, they fled and reported it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} in the town and in the countryside.
\verse So they went out to see what had happened, and they came to Jesus and found the man from whom the demons had gone out sitting there clothed and in his right mind, at the feet of Jesus, and they were afraid.
\verse And those who had seen it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} reported to them how the man who had been demon-possessed had been healed.
\verse And all the people of the surrounding region of the Gerasenes asked him to depart from them, because they had been seized with great fear. So he got into the boat and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“got”) has been translated as a finite verb} returned.
\verse And the man from who the demons had gone out was begging him to stay with him, but he sent him away, saying,
\verse “Return to your home and tell all that God has done for you.” And he went away, proclaiming throughout the whole town all that Jesus had done for him.
\verseWithHeading{A Woman Healed and a Daughter Raised} Now when Jesus returned, the crowd welcomed him, because they were all waiting for him.
\verse And behold, a man \textit{who was named}\lebnote{Literally “to whom the name”} Jairus came, and this man was a ruler of the synagogue. And he fell down at the feet of Jesus and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“fell down”) has been translated as a finite verb} began imploring\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began imploring”)} him to come to his house,
\verse because \textit{he had}\lebnote{Literally “there was to him”} an only daughter, about twelve years old, and she was dying.
\verse Now as he was going, the crowds were pressing against him.
\verse And a woman who was \textit{suffering from hemorrhages}\lebnote{Literally, “with a flow of blood”} for twelve years (who, although she\lebnote{:|NP|:﻿*Here “although” is supplied as a component of the participle (“had spent”) which is understood as concessive} had spent all her\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} assets on physicians, was not able to be healed by anyone)
\verse came up behind him and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came up”) has been translated as a finite verb} touched the edge of his cloak, and immediately \textit{her hemorrhaging}\lebnote{Literally “the flow of her blood”} stopped.
\verse And Jesus said, “Who is the one who touched me?” And when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“denied”)} all denied it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} Peter said, “Master, the crowds are pressing you hard and crowding you!”\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse But Jesus said, “Someone touched me, because I know power has gone out from me.”
\verse And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} the woman saw that she did not escape notice, she came trembling and falling down before him. In the presence of all the people, she told for what reason she had touched him, and that she was healed immediately.
\verse While\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“speaking”)} he was still speaking, someone came from the synagogue ruler’s house, saying, “Your daughter is dead! Trouble the Teacher no longer!”
\verse But Jesus, when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“heard”) which is understood as temporal} heard this,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} replied to him, “Do not be afraid! Only believe, and she will be healed.”
\verse Now when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“came”) which is understood as temporal} came to the house, he did not allow anyone to enter with him except Peter and John and James and the father and mother of the child.
\verse And they were all weeping and mourning for her, but he said, “Do not weep! For she is not dead, but is sleeping.”
\verse And they began laughing\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began laughing”)} at him, because they\lebnote{:|NP|:﻿*Here “because” is supplied as a component of the participle (“knew”) which is understood as causal} knew that she was dead.
\verse But he took her hand and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took”) has been translated as a finite verb} called, saying, “Child, get up.”
\verse And her spirit returned, and she got up immediately, and he ordered something\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} to be given to her to eat.
\verse And her parents were astonished, but he ordered them to tell no one what had happened.
\end{biblechapter}

\begin{biblechapter} % Luke 9
\verseWithHeading{The Twelve Commissioned and Sent Out} And summoning the twelve, he gave them power and authority over all the demons and to cure diseases,
\verse and he sent them out to proclaim the kingdom of God and to heal the sick.
\verse And he said to them, “Take along nothing for the journey—neither a staff, nor a traveler’s bag, nor bread, nor money, nor to have two tunics apiece.
\verse And into whatever house you enter, stay there and depart from there.
\verse And as for all those who do not welcome you—when you\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“depart”) which is understood as temporal} depart from that town, shake off the dust from your feet for a testimony against them.”
\verse So they departed and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“departed”) has been translated as a finite verb} went throughout the villages, proclaiming the good news and healing everywhere.
\verseWithHeading{Herod Perplexed About Jesus} Now Herod the tetrarch heard about all that was happening, and he was greatly perplexed, because it was said by some that John has been raised from the dead,
\verse and by some that Elijah had appeared, and others that some prophet of ancient times had risen.
\verse And Herod said, “John I beheaded, but who is this about whom I hear such things?” And he was wanting to see him.
\verseWithHeading{The Feeding of Five Thousand} And when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“returned”) which is understood as temporal} returned, the apostles described to him all that they had done. And he took them along and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took … along”) has been translated as a finite verb} withdrew privately to a town called Bethsaida.
\verse But when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“found out”) which is understood as temporal} the crowds found out, they followed him, and welcoming them, he began to speak\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to speak”)} to them about the kingdom of God, and he cured those who had need of healing.
\verse Now the day began to be far spent, and the twelve came up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came up”) has been translated as a finite verb} said to him, “Send away the crowd so that they can go into the surrounding villages and farms to obtain lodging and find provisions, because we are here in a desolate place.
\verse But he said to them, “You give them something to eat!” And they said, “\textit{We have no}\lebnote{Literally “there is not to us”} more than five loaves and two fish, unless perhaps we go and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“go”) has been translated as a finite verb} purchase food for all these people.”
\verse (For there were about five thousand men.) So he said to his disciples, “Have them sit down in groups of about fifty each.”
\verse And they did so, and had them all sit down.
\verse And taking the five loaves and the two fish, and\lebnote{:|NP|:﻿*Here “and” is supplied before the participle (“looking up”) in keeping with English style} looking up to heaven, he gave thanks and broke them and began giving\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began giving”)} them\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to speak”)} to the disciples to set before the crowd.
\verse And they all ate and were satisfied, and what was left over was picked up by them—twelve baskets of broken pieces.
\verseWithHeading{Peter’s Confession} And it happened that while he was praying alone, the disciples were with him. And he asked them, saying, “Who do the crowds say that I am?”
\verse And they answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “John the Baptist, but others, Elijah, and others, that one of the ancient prophets has risen.”
\verse And he said to them, “But who do you say that I am?” And Peter answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “The Christ of God.”
\verseWithHeading{Jesus Predicts His Death and Resurrection} But he warned and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“warned”) has been translated as a finite verb} commanded them to tell this to no one,
\verse saying, “It is necessary for the Son of Man to suffer many things and to be rejected by the elders and chief priests and scribes, and to be killed, and to be raised on the third day.
\verseWithHeading{Taking Up One’s Cross to Follow Jesus} And he said to them all, “If anyone wants to come after me, he must deny himself and take up his cross every day and follow me.
\verse For whoever wants to save his life will lose it, but whoever loses his life on account of me, this person will save it.
\verse For what is a person benefited if he\lebnote{:|NP|:﻿*Here “if” is supplied as a component of the participle (“gains”) which is understood as conditional} gains the whole world but loses or forfeits himself?
\verse For whoever is ashamed of me and my words, the Son of Man will be ashamed of this person when he comes in his glory and the glory\lebnote{:|NP|:﻿*Here “glory” is an understood repetition of the same word earlier in this verse} of the Father and of the holy angels.
\verse But I tell you truly, there are some of those standing here who will never experience death until they see the kingdom of God.”
\verseWithHeading{The Transfiguration} Now it happened that about eight days after these words, he took along Peter and John and James and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took along”) has been translated as a finite verb} went up on the mountain to pray.
\verse And as he was praying, the appearance of his face became different, and his clothing became white, gleaming like lightning.\lebnote{Or “became brilliant as light”; or “became dazzling white”}
\verse And behold, two men were talking with him, who were Moses and Elijah,
\verse who appeared in glory and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“appeared”) has been translated as a finite verb} were speaking about his departure which he was about to fulfill\lebnote{Or “to accomplish”} in Jerusalem.
\verse Now Peter and those with him were \textit{very sleepy},\lebnote{Literally “burdened with sleep”} but when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“became fully awake”) which is understood as temporal} became fully awake, they saw his glory and the two men who were standing with him.
\verse And it happened that as they were going away from him, Peter said to Jesus, “Master, it is good for us to be here. And let us make three shelters, one for you and one for Moses and one for Elijah,” not knowing what he was saying.
\verse And while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“saying”)} he was saying these things, a cloud came and overshadowed them, and they were afraid as they entered into the cloud.
\verse And a voice came from the cloud, saying, “This is my Son, my Chosen One. Listen to him!”
\verse And after the voice had occurred, Jesus was found alone. And they kept silent and told no one in those days anything of what they had seen.
\verseWithHeading{A Demon-possessed Boy Healed} Now it happened that on the next day, when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“had come down”)} they had come down from the mountain, a large crowd met him.
\verse And behold, a man from the crowd cried out, saying, “Teacher, I beg you to look with concern on my son, because he is my only son!
\verse And behold, a spirit seizes him and suddenly he screams, and it convulses him with foam and rarely withdraws from him, battering him severely.
\verse And I begged your disciples that they would expel it, and they were not able to do so.”
\verse So Jesus answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “O unbelieving and perverted generation! \textit{How long}\lebnote{Literally “until when”} will I be with you and put up with you? Bring your son here!”
\verse And while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“approaching”)} he was still approaching, the demon threw him down and convulsed him.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} But Jesus rebuked the unclean spirit and healed the boy, and gave him back to his father.
\verse And they were all astounded at the impressiveness of God.
\verseWithHeading{Jesus Predicts His Suffering} But while they\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“marveling”)} were all marveling at all the things that he was doing, he said to his disciples,
\verse “\textit{You take these words to heart},\lebnote{Literally “you put these words into your ears”} for the Son of Man is about to be betrayed into the hands of men.”
\verseWithHeading{The Question About Who Is Greatest} And an argument developed among them as to who of them might be greatest.
\verse But Jesus, because he\lebnote{:|NP|:﻿*Here “because” is supplied as a component of the participle (“knew”) which is understood as causal} knew the thoughts of their hearts, took hold of a child and had him stand beside him
\verse and said to them, “Whoever welcomes this child in my name welcomes me, and whoever welcomes me welcomes the one who sent me. For the one who is least among you all—this one is great.”
\verseWithHeading{Whoever Is Not Against Us Is for Us} And John answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “Master, we saw someone expelling demons in your name, and we tried to prevent him, because he does not follow in company with us.”
\verse But Jesus said to him, “Do not prevent him,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} because whoever is not against you is for you.”
\verseWithHeading{Jesus Rejected in a Samaritan Village} Now it happened that when the days were approaching \textit{for him to be taken up},\lebnote{Literally “of his taking up”} he set his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} face to go to Jerusalem.
\verse And he sent messengers before \textit{him},\lebnote{Literally “his face”} and they went and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“went”) has been translated as a finite verb} entered into a village of the Samaritans in order to prepare for him.
\verse And they did not welcome him because \textit{he was determined to go}\lebnote{Literally “his face was going”} to Jerusalem.
\verse Now when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} the disciples James and John saw it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} they said, “Lord, do you want us to call fire to come down from heaven and consume them?”
\verse But he turned around and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“turned around”) has been translated as a finite verb} rebuked them,
\verse and they proceeded to another village.
\verseWithHeading{Would-be Followers} And as\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the temporal genitive absolute participle (“were going”)} they were traveling on the road, someone said to him, “I will follow you wherever you go!”
\verse And Jesus said to him, “Foxes have dens and birds of the sky have nests, but the Son of Man has no place to lay his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} head.”
\verse And he said to another, “Follow me!” But he said, “Lord, first allow me to go and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“go”) has been translated as an infinitive} bury my father.”
\verse But he said to him, “Leave the dead to bury their own dead! But you go and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“go”) has been translated as a finite verb} proclaim the kingdom of God.”
\verse And another person also said, “I will follow you, Lord, but first allow me to say farewell to those in my house.”
\verse But Jesus said,\lebnote{Some manuscripts have “said to him”} “No one who puts his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} hand on the plow and looks back is fit for the kingdom of God!”
\end{biblechapter}

\begin{biblechapter} % Luke 10
\verseWithHeading{The Seventy-Two Appointed and Sent Out} And after these things, the Lord also\lebnote{Some manuscripts omit “also”} appointed seventy-two others and sent them out two by two before \textit{him}\lebnote{Literally “his presence”} into every town and place where he was about to go.
\verse And he said to them, “The harvest is plentiful, but the workers are few. Therefore ask the Lord of the harvest that he send out workers into his harvest.
\verse Go! Behold, I am sending you out like lambs in the midst of wolves!
\verse Do not carry a money bag or a traveler’s bag or sandals, and greet no one along the road.
\verse And into whatever house you enter, first say, “Peace be to this household!”
\verse And if a son of peace is there, your peace will rest on him. But if not, it will return to you.
\verse And remain in the same house, eating and drinking \textit{whatever they provide},\lebnote{Literally “the things from them”} for the worker is worthy of his pay. Do not move from house to house.
\verse And into whatever town you enter and they welcome you, eat \textit{whatever is}\lebnote{Literally “the things”} set before you,
\verse and heal the sick in it, and say to them, “The kingdom of God has come near to you.”
\verse But into whatever town you enter and they do not welcome you, go out into its streets and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“go out”) has been translated as a finite verb} say,
\verse “Even the dust of your town that clings to our feet we wipe off against you! Nevertheless know this: that the kingdom of God has come near!”\lebnote{Or “has come”}
\verse I tell you that it will be more bearable on that day for Sodom than for that town!
\verse Woe to you, Chorazin! Woe to you, Bethsaida! For if the miracles that were done in you had been done in Tyre and Sidon, they would have repented long ago, sitting in sackcloth and ashes!
\verse But it will be more bearable for Tyre and for Sidon in the judgment than for you!
\verse And you, Capernaum, will you be exalted to heaven? No! You will be brought down to Hades!
\verse The one who listens to you listens to me, and the one who rejects you rejects me. But the one who rejects me rejects the one who sent me.”
\verse And the seventy-two returned with joy, saying, “Lord, even the demons are subject to us in your name!”
\verse So he said to them, “I saw Satan falling like lightning from heaven.
\verse Behold, I have given you the authority to tread on snakes and scorpions, and over all the power of the enemy, and nothing will ever harm you.
\verse Nevertheless, do not rejoice in this, that the spirits are subject to you, but rejoice that your names are inscribed in heaven.”
\verseWithHeading{Jesus Rejoices and Prays} At that same time he rejoiced in the Holy Spirit and said, “I praise you, Father, Lord of heaven and earth, that you have hidden these things from the wise and intelligent and have revealed them to young children. Yes, Father, for this was pleasing before you.
\verse All things have been handed over to me by my Father, and no one knows who the Son is except the Father and who the Father is except the Son, and anyone to whom the Son wants to reveal him.”\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And turning to the disciples, he said privately, “Blessed are the eyes that see the things which you see!
\verse For I tell you that many prophets and kings desired to see the things which you see, and did not see them,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} and to hear the things which you hear, and did not hear them.”\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verseWithHeading{The Parable of the Good Samaritan} And behold, a certain legal expert stood up to test him, saying, “Teacher, what must I do so that I will inherit eternal life?”
\verse And he said to him, “What is written in the law? How do you read it?”\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And he answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “You shall love the Lord your God from all your heart, and with all your soul, and with all your strength, and with all your mind,\lebnote{:|NP|:A quotation from Deut 6:5} and your neighbor as yourself.”\lebnote{:|NP|:A quotation from Lev 19:18}
\verse And he said to him, “You have answered correctly. Do this and you will live.”
\verse But he, wanting to justify himself, said to Jesus, “And who is my neighbor?”
\verse And\lebnote{Some manuscripts omit “and”} Jesus replied and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“replied”) has been translated as a finite verb} said, “A certain man was going down from Jerusalem to Jericho, and fell into the hands of robbers, who both stripped him and beat him.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} After\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“inflicting blows on”) which is understood as temporal} inflicting blows on him,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} they went away, leaving him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} half dead.
\verse Now by coincidence a certain priest was going down on that road, and when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw him, he passed by on the opposite side.
\verse And in the same way also a Levite, when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“came”) which is understood as temporal} came down to the place\lebnote{Some manuscripts have “who happened by the place, when he came up to him”} and saw him,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} passed by on the opposite side.
\verse But a certain Samaritan who was traveling came up to him and, when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw him,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} had compassion.
\verse And he came up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came up”) has been translated as a finite verb} bandaged his wounds, pouring on olive oil and wine, and he put him on his own animal and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“put”) has been translated as a finite verb} brought him to an inn and took care of him.
\verse And on the next day, he took out two denarii and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took out”) has been translated as a finite verb} gave them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation}﻿\lebnote{Some manuscripts have “he took out and gave two denarii”} to the innkeeper, and said, “Take care of him, and whatever you spend in addition, I will repay to you when I return.
\verse Which of these three do you suppose became a neighbor of the man who fell among the robbers?”
\verse So he said, “The one who showed mercy to him.” And Jesus said to him, “You go and do likewise.”
\verseWithHeading{Martha and Mary} Now as they traveled along, he entered into a certain village. And a certain woman \textit{named}\lebnote{Literally “by name”} Martha welcomed him.\lebnote{Most manuscripts add some form of the location such as “into her house” but there is considerable variation in the exact wording, so the shorter reading is preferred}
\verse And \textit{she had}\lebnote{Literally “this was”} a sister named Mary, who also sat at the feet of Jesus\lebnote{Some manuscripts have “of the Lord”} and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“sat”) has been translated as a finite verb} was listening to his teaching.
\verse But Martha was distracted with much preparation, so she approached and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“approached”) has been translated as a finite verb} said, “Lord, is it not a concern to you that my sister has left me alone to make preparations? Then tell her that she should help me!”
\verse But the Lord answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to her, “Martha, Martha, you are anxious and troubled about many things!
\verse But few things are necessary, or only one thing,\lebnote{Some manuscripts have “But one thing is necessary”} for Mary has chosen the better part, which will not be taken away from her.”
\end{biblechapter}

\begin{biblechapter} % Luke 11
\verseWithHeading{How to Pray} And it happened that while he was in a certain place praying, when he stopped a certain one of his disciples said to him, “Lord, teach us to pray, just as John also taught his disciples.”
\verse And he said to them, “When you pray, say,
\verse “Father, 
may your name be treated as holy. 
May your kingdom come.
\verse Give us each day our daily bread.
\verseWithHeading{Ask, Seek, Knock} And he said to them, “Who of you will have a friend, and will go to him at midnight and say to him, ‘Friend, lend me three loaves,
\verse because a friend of mine has come to me on a journey, and I do not have anything to set before him.’
\verse And that one will answer from inside and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“will answer”) has been translated as a finite verb} say, ‘Do not cause me trouble! The door has already been shut and my children are with me in bed! I am not able to get up to give you anything.’\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse I tell you, even if he does not give him anything\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} after he\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“gets up”) which is understood as temporal} gets up because he is his friend, at any rate because of his impudence\lebnote{Or “shamelessness”; some translate as “persistence” based on the context, though this is not the normal meaning of the word} he will get up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“will get up”) has been translated as a finite verb} give him whatever he needs.
\verse And I tell you, ask and it will be given to you; seek and you will find; knock and it will be opened for you.
\verse For everyone who asks receives, and the one who seeks finds, and to the one who knocks it will be opened.
\verse But what father from among you, if his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} son will ask for a fish, instead of a fish will give him a snake?
\verse Or also, if he will ask for an egg, will give him a scorpion?
\verse Therefore if you, although you\lebnote{:|NP|:﻿*Here “although” is supplied as a component of the participle (“are”) which is understood as concessive} are evil, know how to give good gifts to your children, how much more will the Father from heaven give the Holy Spirit to those who ask him?”
\verseWithHeading{A House Divided Cannot Stand} And he was expelling a mute demon.\lebnote{Some manuscripts have “a demon, and it was mute”} Now it happened that when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“came out”)} the demon came out, the man who had been mute spoke, and the crowds were astonished.
\verse But some of them said, “By Beelzebul the ruler of demons he expels demons!”
\verse And others, in order to\lebnote{:|NP|:﻿*Here “in order to” is supplied as a component of the participle (“test”) which is understood as purpose} test him,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} were demanding from him a sign from heaven.
\verse But he, knowing their thoughts, said to them, “Every kingdom divided against itself is laid waste, and \textit{a divided household}\lebnote{Literally “house against house”} falls.
\verse So if Satan also is divided against himself, how will his kingdom stand? For you say that I expel demons by Beelzebul.
\verse But if I expel demons by Beelzebul, by whom do your sons expel them?\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} For this reason they will be your judges!
\verse But if I expel demons by the finger of God, then the kingdom of God has come upon you!
\verse When a strong man, fully armed, guards his own palace, his possessions are \textit{safe}.\lebnote{Literally “in peace”}
\verse But when a stronger man attacks him and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“attacks”) has been translated as a finite verb} conquers him, he takes away his full armor in which he trusted and distributes his plunder.
\verse The one who is not with me is against me, and the one who does not gather with me scatters.
\verseWithHeading{An Unclean Spirit Returns} “Whenever an unclean spirit has gone out of a person, it travels through waterless places searching for rest, and does not find it.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} It says,\lebnote{Some manuscripts have “Then it says”} ‘I will return to my house from which I came out.’
\verse And when it\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“arrives”) which is understood as temporal} arrives it finds the house\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} swept and put in order.
\verse Then it goes and brings along seven other spirits more evil than itself, and they go in and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“go in”) has been translated as a finite verb} live there. And the last state of that person becomes worse than the first!”
\verse Now it happened that as he said these things, a certain woman from the crowd raised her voice and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“raised”) has been translated as a finite verb} said to him, “Blessed is the womb that bore you, and the breasts at which you nursed!”
\verse But he said, “On the contrary, blessed are those who hear the word of God and follow it!”\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verseWithHeading{The Sign of Jonah} And as\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the temporal genitive absolute participle (“were increasing”)} the crowds were increasing, he began to say, “This generation is an evil generation! It demands a sign, and no sign will be given to it except the sign of Jonah!
\verse For just as Jonah became a sign to the Ninevites, so also the Son of Man will be to this generation.
\verse The queen of the south will rise up at the judgment with the people of this generation and condemn them, because she came from the ends of the earth to hear the wisdom of Solomon, and behold, something\lebnote{:|NP|:﻿*Here “something” is supplied in the translation because the adjective is neuter gender} greater than Solomon is here!
\verse The people of Nineveh will stand up at the judgment with this generation and condemn it, because they repented at the proclamation of Jonah, and behold, something\lebnote{:|NP|:﻿*Here “something” is supplied in the translation because the adjective is neuter gender} greater than Jonah is here!
\verseWithHeading{Light and Darkness} “No one after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“lighting”) which is understood as temporal} lighting a lamp puts it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} in a cellar or under a bushel basket, but on a lampstand, so that those who come in can see the light.
\verse Your eye is the lamp of the body. When your eye is sincere, your whole body is full of light also. But when it is evil, your body is dark also.
\verse Therefore pay careful attention that the light in you is not darkness!
\verse If therefore your whole body is full of light, not having any part dark, it will be completely full of light, as when the lamp with its light gives light to you.”
\verseWithHeading{Pharisees and Legal Experts Denounced} And as he was speaking, a Pharisee asked him \textit{to have a meal}\lebnote{Literally “that he would have a meal”} with him, and he went in and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“went in”) has been translated as a finite verb} reclined at table.
\verse And the Pharisee, when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} was astonished that he did not first wash before the meal.
\verse But the Lord said to him, “Now you Pharisees cleanse the outside of the cup and of the dish, but your inside is full of greediness and wickedness.
\verse Fools! Did not the one who made the outside make the inside also?
\verse But give as charitable giving the things that are within, and behold, everything is clean for you.
\verse “But woe to you, Pharisees, because you pay a tenth of mint and rue and every garden herb, and neglect justice and love for God! But it was necessary to do these things without neglecting those things also.
\verse Woe to you, Pharisees, because you love the best seat in the synagogues and the greetings in the marketplaces!
\verse Woe to you, because you are like unmarked graves, and the people who walk over them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} do not know it!\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And one of the legal experts answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to him, “Teacher, when you\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“say”) which is understood as temporal} say these things, you insult us also!”
\verse So he said, “Woe to you also, legal experts, because you load people with burdens hard to bear, and you yourselves do not touch the burdens with one of your fingers!
\verse Woe to you, because you build the tombs of the prophets, and your fathers killed them!
\verse As a result you are witnesses, and you approve of the deeds of your fathers, because they killed them and you build their tombs!\lebnote{A large number of later manuscripts add the words “their tombs” here, with variations of wording; although the words are not likely to be original, it is necessary to supply them in keeping with English style}
\verse For this reason also the wisdom of God said, ‘I will send to them prophets and apostles, and some of them they will kill and persecute,’
\verse so that the blood of all the prophets that has been shed from the foundation of the world may be required of this generation,
\verse from the blood of Abel to the blood of Zechariah, who perished between the altar and the temple building.\lebnote{Literally “the house,” here a reference to the temple} Yes, I tell you, it will be required of this generation!
\verse Woe to you, legal experts, because you have taken away the key to knowledge! You did not enter yourselves, and you hindered those who were entering!”
\verse And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“departed”)} he departed from there, the scribes and the Pharisees began to be terribly hostile, and to question him closely about many things,
\verse plotting to catch him with reference to something \textit{he might say}.\lebnote{Literally “from his mouth”}
\end{biblechapter}

\begin{biblechapter} % Luke 12
\verseWithHeading{Warning Against Hypocrisy} During \textit{this time}\lebnote{Literally “which time”} when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“had gathered together”)} a crowd of many thousands had gathered together, so that they were trampling one another, he began to say to his disciples first, “Beware for yourselves of the leaven of the Pharisees, which is hypocrisy.
\verse But nothing is concealed that will not be revealed, and secret that will not be made known.
\verse Therefore everything that you have said in the dark will be heard in the light, and what \textit{you have whispered}\lebnote{Literally “you have spoken to the ear”} in the inner rooms will be proclaimed on the housetops.
\verseWithHeading{Fear God Rather Than People} “And I tell you, my friends, do not be afraid of those who kill the body, and after these things do not have anything more to do.
\verse But I will show you whom you should fear: fear the one who has authority, after the killing, to throw you\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} into hell! Yes, I tell you, fear this one!
\verse Are not five sparrows sold for two pennies? And not one of them is forgotten in the sight of God.
\verse But even the hairs of your head are all numbered! Do not be afraid; you are worth more than many sparrows.
\verseWithHeading{Acknowledgement of Christ and Persecution of Disciples} “And I tell you, everyone who acknowledges me before people, the Son of Man also will acknowledge him before the angels of God,
\verse but the one who denies me before people will be denied before the angels of God.
\verse And everyone who speaks a word against the Son of Man, it will be forgiven him, but to the one who blasphemes against the Holy Spirit, it will not be forgiven.
\verse But when they bring you before the synagogues and the rulers and the authorities, do not be anxious how or what you should speak in your own defense or what you should say,
\verse for the Holy Spirit will teach you in that same hour what it is necessary to say.”
\verseWithHeading{The Parable of the Rich Landowner Who Was a Fool} Now someone from the crowd said to him, “Teacher, tell my brother to divide the inheritance with me!”
\verse But he said to him, “Man, who made me a judge or an arbitrator over you?”
\verse And he said to them, “Watch out and guard yourselves from all greediness, because not even when someone has an abundance \textit{does}\lebnote{Literally “is”} his life consist of his possessions.”
\verse And he told a parable to them, saying, “The land of a certain rich man yielded an abundant harvest.
\verse And he reasoned to himself, saying, ‘What should I do? For I do not have anywhere I can gather in my crops.’
\verse And he said, ‘I will do this: I will tear down my barns and build larger ones, and I will gather in there all my grain and possessions.
\verse And I will say to my soul, “Soul, you have many possessions stored up for many years. Relax, eat, drink, celebrate!” ’
\verse But God said to him, ‘Fool! This night your life\lebnote{The same Greek word can be translated “soul” or “life” depending on the context} is demanded from you, and the things which you have prepared—whose will they be?’
\verse So is the one who stores up treasure for himself, and who is not rich toward God!”
\verseWithHeading{Anxiety} And he said to his disciples, “For this reason I tell you, do not be anxious for your\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} life, what you will eat, or for your\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} body, what you will wear.
\verse For life is more than food, and the body more than clothing.
\verse Consider the ravens, that they neither sow nor reap; to them there is neither storeroom nor barn, and God feeds them. How much more are you worth than the birds?
\verse And which of you by\lebnote{:|NP|:﻿*Here “by” is supplied as a component of the participle (“being anxious”) which is understood as means} being anxious is able to add an hour\lebnote{Or “a cubit” (the literal meaning); most scholars understand this to refer figuratively to an “hour” of life here, though some take it as a literal measurement of height} to his life span?
\verse If then you are not even able to do a very little thing, why are you anxious about the rest?
\verse Consider the lilies, how they grow: they do not toil or spin, but I say to you, not even Solomon in all his glory was dressed like one of these.
\verse But if God clothes the grass in the field in this way, although it\lebnote{:|NP|:﻿*Here “although” is supplied as a component of the participle (“is”) which is understood as concessive} is here today and tomorrow is thrown into the oven, how much more will he do so for\lebnote{:|NP|:﻿*The phrase “will he do so for” is not in the Greek text but is implied} you, you of little faith?
\verse And you, do not consider what you will eat and what you will drink, and do not be anxious.
\verse For all the nations of the world seek after these things, and your Father knows that you need these things.
\verse But seek his kingdom and these things will be added to you.
\verse “Do not be afraid, little flock, because your Father is well pleased to give you the kingdom.
\verse Sell your possessions and give charitable gifts. Make for yourselves money bags that do not wear out, an inexhaustible treasure in heaven\lebnote{Or “in the heavens”} where thief does not approach or moth destroy.
\verse For where your treasure is, there your heart will be also.
\verseWithHeading{On the Alert for the Master’s Return} “\textit{You must be prepared for action}\lebnote{Literally “your loins must be girded”} and your\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} lamps burning.
\verse And you, be like people who are waiting for their master when he returns from the wedding feast,\lebnote{Or perhaps simply “feast”} so that when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“comes back”)} comes back and knocks, they can open the door\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} for him immediately.
\verse Blessed are those slaves whom the master will find on the alert when he returns! Truly I say to you that he will dress himself for service and have them recline at the table and will come by and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“will come by”) has been translated as a finite verb} serve them.
\verse Even if he should come back in the second or in the third watch of the night and find them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} like this, blessed are they!
\verse But understand this, that if the master of the house had known what hour the thief was coming, he would not have left his house to be broken into.
\verse You also must be ready, because the Son of Man is coming at an hour that you do not think he will come.”\lebnote{:|NP|:*The words “he will come” are not in the Greek text but are implied}
\verseWithHeading{A Faithful Slave and an Unfaithful Slave} And Peter said, “Lord, are you telling this parable for us, or also for everyone?”
\verse And the Lord said, “Who then is the faithful wise manager whom the master will put in charge over his servants to give them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} their\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} food allowance at the right time?
\verse Blessed is that slave whom his master will find so doing when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“comes back”) which is understood as temporal} comes back.
\verse Truly I say to you that he will put him in charge of all his possessions.
\verse But if that slave should say \textit{to himself},\lebnote{Literally “in his heart”} ‘My master is taking a long time to return,’ and he begins to beat the male slaves and the female slaves and to eat and drink and get drunk,
\verse the master of that slave will come on a day that he does not expect and at an hour that he does not know, and will cut him in two and assign his place with the unbelievers.
\verse And that slave who knew the will of his master and did not prepare or do according to his will \textit{will be given a severe beating}.\lebnote{Literally “will be beaten much”}
\verse But the one who did not know and did things deserving blows \textit{will be given a light beating}.\lebnote{Literally “will be beaten a few times”} And from everyone to whom much has been given, much will be demanded, and from him to whom they entrusted much, they will ask him for even more.
\verseWithHeading{Not Peace, But a Sword of Divisiveness} “I have come to bring fire on the earth, and how I wish that it had been kindled already!
\verse But I have a baptism to be baptized with, and how I am distressed until it is accomplished!
\verse Do you think that I have come to grant peace on the earth? No, I tell you, but rather division!
\verse For from now on there will be five in one household, divided three against two and two against three.
\verse They will be divided, father against son and son against father, mother against daughter and daughter against mother, mother-in-law against her daughter-in-law and daughter-in-law against mother-in-law.”
\verseWithHeading{The Signs of the Times} And he also said to the crowds, “When you see a cloud coming up in the west, you say at once, ‘A rainstorm is coming,’ and so it happens.
\verse And when you see the south wind blowing, you say, ‘There will be burning heat,’ and it happens.
\verse Hypocrites! You know how to evaluate the appearance of the earth and the sky, but how is it you do not know how to evaluate this present time?
\verseWithHeading{Settle Accounts Quickly} And why do you not also judge for yourselves what is right?
\verse For as you are going with your accuser before the magistrate, make an effort to come to a settlement with him on the way, so that he will not drag you to the judge, and the judge will hand you over to the bailiff, and the bailiff will throw you into prison.
\verse I tell you, you will never get out of there until you have paid back even the last cent!”\lebnote{Literally “lepton,” a small copper coin worth 1/128 of a denarius}
\end{biblechapter}

\begin{biblechapter} % Luke 13
\verseWithHeading{Repent or Perish} Now at the same time some had come to tell him about the Galileans whose blood Pilate had mixed with their sacrifices.
\verse And he answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to them, “Do you think that these Galileans were sinners worse than all the Galileans, because they suffered these things?
\verse No, I tell you, but unless you repent you will all perish as well!
\verse Or those eighteen on whom the tower in Siloam fell and killed them—do you think that they were sinners worse than all the people who live in Jerusalem?
\verse No, I tell you, but unless you repent, you will all perish as well!”
\verseWithHeading{The Parable of the Barren Fig Tree} And he told this parable: “A certain man had a fig tree planted in his vineyard, and he came looking for fruit on it and did not find any.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse So he said to the gardener, ‘Behold, \textit{for three years}\lebnote{Literally “three years from which”} I have come looking for fruit on this fig tree and did not find any.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} Cut it down!\lebnote{Some manuscripts have “Therefore cut it down!”} Why should it even exhaust the soil?’
\verse But he answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to him, ‘Sir, leave it alone this year also, until I dig around it and put manure on it.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And if indeed it produces fruit in the coming year, so much the better,\lebnote{:|NP|:*The phrase “so much the better” is not in the Greek text but is implied} but if not, you can cut it down.’ ”
\verseWithHeading{A Woman with a Disabling Spirit Healed} Now he was teaching in one of the synagogues on the Sabbath.
\verse And behold, a woman was there\lebnote{:|NP|:﻿*The phrase “was there” is not in the Greek text but is supplied in keeping with English style} who had a spirit \textit{that had disabled her}\lebnote{Literally “of weakness”} for eighteen years, and she was bent over and not able to straighten herself up completely.\lebnote{Or “at all”}
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw her, Jesus summoned her\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} and said to her, “Woman, you are freed from your disability!”
\verse And he placed his\lebnote{Literally “the”; the Greek article is used here as a possessive pronoun} hands on her, and immediately she straightened up and glorified God.
\verse But the ruler of the synagogue, indignant because Jesus had healed on the Sabbath, answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to the crowd, “There are six days on which it is necessary to work. Therefore come and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“come”) has been translated as a finite verb} be healed on them, and not on the day of the Sabbath!”
\verse But the Lord answered and said to him, “Hypocrites! Does not each one of you untie his ox or his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} donkey from the feeding trough on the Sabbath and lead it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} away to water it?\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And this woman, who is a daughter of Abraham, whom Satan bound \textit{eighteen}\lebnote{Literally “ten and eight”} long years—is it not necessary that she be released from this bond on the day of the Sabbath?”
\verse And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“said”)} he said these things, all those who opposed him were humiliated, and the whole crowd was rejoicing at all the splendid things that were being done by him.
\verseWithHeading{The Parable of the Mustard Seed} Therefore he said, “What is the kingdom of God like, and to what shall I compare it?
\verse It is like a mustard seed that a man took and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took”) has been translated as a finite verb} sowed in his own garden, and it grew and became a tree, and the birds of the sky nested in its branches.”
\verseWithHeading{The Parable of the Yeast} And again he said, “To what shall I compare the kingdom of God?
\verse It is like yeast that a woman took and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took”) has been translated as a finite verb} hid in\lebnote{Some manuscripts have “put into”} three measures of wheat flour until the whole batch was leavened.”
\verseWithHeading{The Narrow Door} And he was going throughout towns and villages, teaching and making his journey toward Jerusalem.
\verse And someone said to him, “Lord, \textit{are there only}\lebnote{Literally “if”} a few who are saved?” And he said to them,
\verse “Make every effort to enter through the narrow door, because many, I tell you, will seek to enter and will not be able to,
\verse when once the master of the house has gotten up and shut the door, and you begin to stand outside and knock on the door, saying, ‘Lord, open the door\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} for us!’ And he will answer and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answer”) has been translated as a finite verb} say to you, ‘I do not know where you are from!’
\verse Then you will begin to say, ‘We ate and drank in your presence, and you taught in our streets!’
\verse And he will reply, saying to you, ‘I do not know where you are from! Go away from me, all you \textit{evildoers}!’\lebnote{Literally “workers of unrighteousness”}
\verse In that place there will be weeping and gnashing of teeth, when you see Abraham and Isaac and Jacob and all the prophets in the kingdom of God, but yourselves thrown outside!
\verse And they will come from east and west, and from north and south, and will recline at the table in the kingdom of God.
\verse And behold, some are last who will be first, and some are first who will be last.”
\verseWithHeading{The Lament over Jerusalem} At that same hour some Pharisees came up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came up”) has been translated as a finite verb} said to him, “Go out and depart from here, because Herod wants to kill you!”
\verse And he said to them, “Go and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“go”) has been translated as a finite verb} tell that fox, ‘Behold, I am expelling demons and performing healings today and tomorrow, and on the third day I will complete my work.’\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse Nevertheless, it is necessary for me to be on the way today and tomorrow and on the next day, because it is not possible for a prophet to perish outside Jerusalem.
\verse “Jerusalem, Jerusalem, the one who kills the prophets and stones those who are sent to her! How many times I wanted to gather your children together \textit{the way}\lebnote{Literally “in the manner in which”} a hen gathers\lebnote{:|NP|:﻿*Here the term “gathers” is not in the Greek text but is implied} her own brood under her\lebnote{Literally “the”; the Greek article is used here as a possessive pronoun} wings, and you were not willing!
\verse Behold, your house has been left to you! And I tell you, you will never see me until the time\lebnote{:|NP|:﻿*The words “the time” are not in the Greek text but are implied} will come when you say, ‘Blessed is the one who comes in the name of the Lord!’ ”\lebnote{:|NP|:A quotation from Ps 118:26}
\end{biblechapter}

\begin{biblechapter} % Luke 14
\verseWithHeading{A Man Suffering from Edema Healed} And it happened that when he came to the house of a certain one of the leaders of the Pharisees on a Sabbath to eat \textit{a meal},\lebnote{Literally “bread”} they were watching him closely.
\verse And behold, a certain man was in front of him, suffering from edema.
\verse And Jesus answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to the legal experts and Pharisees, saying, “Is it permitted to heal on the Sabbath, or not?”
\verse But they remained silent. And he took hold of him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took hold of”) has been translated as a finite verb} healed him, and sent him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} away.
\verse And he said to them, “Who among you, if your\lebnote{:|NP|:﻿*The words “if your” are not in the Greek text but are implied} son or your ox falls into a well\lebnote{Or “cistern”} on the day of the Sabbath, will not immediately pull him out?”
\verse And they were not able to make a reply to these things.
\verseWithHeading{The Parable of the Guests at the Wedding Feast} Now he told a parable to those who had been invited when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“noticed”) which is understood as temporal} noticed how they were choosing for themselves the places of honor, saying to them,
\verse “When you are invited by someone to a wedding feast,\lebnote{Or perhaps simply “a feast”} do not recline at the table in the place of honor, lest someone more distinguished than you has been invited by him,
\verse and the one who invited you \textit{both}\lebnote{Literally “and him”} will come and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“will come”) has been translated as a finite verb} say to you, ‘Give the place to this person,’ and then with shame you will begin to take the last place.
\verse But when you are invited, go and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“go”) has been translated as a finite verb} recline at the table in the last place, so that when the one who invited you comes, he will say to you, ‘Friend, move up higher.’ Then it will be an honor to you in the presence of all those who are reclining at the table with you.
\verse For everyone who exalts himself will be humbled, and the one who humbles himself will be exalted.”
\verseWithHeading{The Parable of the Great Banquet} And he also said to the one who had invited him, “When you give a dinner or a banquet, do not invite your friends or your brothers or your relatives or wealthy neighbors, lest they also invite you in return, and repayment come to you.
\verse But whenever you give a banquet, invite the poor, the crippled, the lame, the blind,
\verse and you will be blessed, because they are not able to repay you. For it will be paid back to you at the resurrection of the righteous.”
\verse Now when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“heard”) which is understood as temporal} one of those reclining at the table with him heard these things, he said to him, “Blessed is \textit{everyone who}\lebnote{Literally “whoever”} will eat bread in the kingdom of God!”
\verse But he said to him, “A certain man was giving a large banquet and invited many.
\verse And he sent his slave at the hour of the banquet to say to those who have been invited, ‘Come, because now it is ready!’
\verse And they all \textit{alike}\lebnote{Literally “by one”} began to excuse themselves. The first said to him, ‘I have purchased a field, and \textit{I must}\lebnote{Literally “I have necessity”} go out to look at it. I ask you, consider me excused.’
\verse And another said, ‘I have purchased five yoke of oxen, and I am going to examine them. I ask you, consider me excused.’
\verse And another said, ‘I have married a wife, and for this reason I am not able to come.’
\verse And the slave came and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came”) has been translated as a finite verb} reported these things to his master. Then the master of the house became angry and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“became angry”) has been translated as a finite verb} said to his slave, ‘Go out quickly into the streets and alleys of the city and bring in here the poor and crippled and blind and lame!’
\verse And the slave said, ‘Sir, what you ordered has been done, and there is still room.’
\verse And the master said to the slave, ‘Go out into the highways and hedges and press them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} to come in, so that my house will be filled!
\verse For I say to you that none of those persons who were invited will taste my banquet!’ ”
\verseWithHeading{The Cost of Discipleship} Now large crowds were going along with him, and he turned around and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“turned around”) has been translated as a finite verb} said to them,
\verse “If anyone comes to me and does not hate his own father and mother and wife and children and brothers and sisters, and furthermore, even his own life, \textit{he cannot be}\lebnote{Literally “he is not able to be”} my disciple.
\verse Whoever does not carry his own cross and \textit{follow}\lebnote{Literally “come after”} me \textit{cannot be}\lebnote{Literally “is not able to be”} my disciple.
\verse For which of you, wanting to build a tower, does not first sit down and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“sit down”) has been translated as a finite verb} calculate the cost to see if he has enough\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} \textit{to complete it}?\lebnote{Literally “for completion”}
\verse \textit{Otherwise}\lebnote{Literally “so that lest”} after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the temporal genitive absolute participle (“has laid”)} he has laid the foundation and is not able to finish it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} all who see it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} will begin to ridicule him,
\verse saying, ‘This man began to build and was not able to finish!’
\verse Or what king, going out to engage another king in battle, does not sit down first and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“sit down”) has been translated as a finite verb} deliberate whether he is able with ten thousand to oppose the one coming against him with twenty thousand.
\verse But if not, while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“is”)} the other is still far away, he sends an ambassador and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“asks”) has been translated as a finite verb} asks \textit{for terms of}\lebnote{Literally “the things with reference to”} peace.
\verse In the same way, therefore, every one of you who does not renounce all his own possessions \textit{cannot be}\lebnote{Literally “is not able to be”} my disciple.
\verse “Now salt is good, but if salt becomes tasteless, with what will it be made salty?
\verse It is usable neither for the soil nor for the manure pile; they throw it out. The one who has ears to hear, let him hear!”
\end{biblechapter}

\begin{biblechapter} % Luke 15
\verseWithHeading{The Parable of the Lost Sheep} Now all the tax collectors and the sinners were drawing near to hear him.
\verse And both the Pharisees and the scribes were complaining, saying, “This man welcomes sinners and eats with them!”
\verse So he told them this parable, saying,
\verse “What man of you, having a hundred sheep and losing one of them, does not leave the ninety-nine in the grassland and go after the one that was lost until he finds it?
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“has found”) which is understood as temporal} has found it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} he places it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} on his shoulders, rejoicing.
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“returns”) which is understood as temporal} returns to his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} home, he calls together his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} friends and neighbors, saying to them, ‘Rejoice with me, because I have found my sheep that was lost!’
\verse I tell you that in the same way, there will be more joy in heaven over one sinner who repents than over ninety-nine righteous people who have no need of repentance.
\verseWithHeading{The Parable of the Lost Coin} Or what woman who has ten drachmas, if she loses one drachma, does not light a lamp and sweep the house and search carefully until she finds it?\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And when she\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“has found”) which is understood as temporal} has found it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} she calls together her\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} friends and neighbors, saying, ‘Rejoice with me, because I have found the drachma that I had lost!’
\verse In the same way, I tell you, there is joy in the presence of the angels of God over one sinner who repents.”
\verseWithHeading{The Parable of the Lost Son} And he said, “A certain man had two sons.
\verse And the younger of them said to his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} father, ‘Father, give me the share of the property that is coming to me.’ So he divided his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} assets between them.
\verse And after not many days, the younger son gathered everything and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“gathered”) has been translated as a finite verb} went on a journey to a distant country, and there he squandered his wealth by\lebnote{:|NP|:﻿*Here “by” is supplied as a component of the adverbial participle of manner (“living”)} living wastefully.
\verse And after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the temporal genitive absolute participle (“had spent”)} he had spent everything, there was a severe famine throughout that country, and he began to be in need.
\verse And he went and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“went”) has been translated as a finite verb} hired himself out to one of the citizens of that country, and he sent him into his fields to tend pigs.
\verse And he was longing to fill his stomach with\lebnote{Some manuscripts have “to stuff himself with”} the carob pods that the pigs were eating, and no one was giving anything\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} to him.
\verse “But when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“came”) which is understood as temporal} came to himself, he said, ‘How many of my father’s hired workers have an abundance \textit{of food},\lebnote{Literally “of bread”} and I am dying here from hunger!
\verse I will set out and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“set out”) has been translated as a finite verb} go to my father and will say to him, ‘Father, I have sinned against heaven and \textit{in your sight}!\lebnote{Literally “in the sight of you”}
\verse I am no longer worthy to be called your son! Make me like one of your hired workers.’
\verse And he set out and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“set out”) has been translated as a finite verb} came to his own father. But while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“away”)} he was still a long way away, his father saw him and had compassion, and ran \textit{and embraced him}\lebnote{Literally “fell on his neck”} and kissed him.
\verse And his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} son said to him, ‘Father, I have sinned against heaven and \textit{in your sight}!\lebnote{Literally “in the sight of you”} I am no longer worthy to be called your son!’
\verse But his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} father said to his slaves, ‘Quickly bring out the best robe and put it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} on him, and put a ring on his finger\lebnote{Literally “hand,” but this is a metonymy of whole (“hand”) for part (“finger”)} and sandals on his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} feet!
\verse And bring the fattened calf—kill it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} and let us eat and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“eat”) has been translated as a finite verb} celebrate,
\verse because this son of mine was dead, and is alive again! He was lost and is found!’ And they began to celebrate.
\verse “Now his older son was in the field, and when he came and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came”) has been translated as a finite verb} approached the house, he heard music and dancing.
\verse And he summoned one of the slaves and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“summoned”) has been translated as a finite verb} asked what these things meant.
\verse And he said to him, ‘Your brother has come, and your father has killed the fattened calf because he has gotten him back healthy.’
\verse But he became angry and did not want to go in. So his father came out and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came out”) has been translated as a finite verb} began to implore\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to implore”)} him.
\verse But he answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to his father, ‘Behold, so many years I have served you, and have never disobeyed your command! And you never gave me a young goat so that I could celebrate with my friends!
\verse But when this son of yours returned—who has consumed your assets with prostitutes—you killed the fattened calf for him!’
\verse But he said to him, ‘Child, you are always with me, and \textit{everything I have belongs to you}.\lebnote{Literally “all my things are yours”}
\verse But it was necessary to celebrate and to rejoice, because this brother of yours was dead, and is alive, and was lost, and is found!’ ”
\end{biblechapter}

\begin{biblechapter} % Luke 16
\verseWithHeading{The Parable of the Dishonest Manager} And he also said to the disciples, “A certain man was rich, who had a manager. And charges were brought to him that this person was squandering his possessions.
\verse And he summoned him and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“summoned”) has been translated as a finite verb} said to him, ‘What is this I hear about you? Give the account of your management, because you can no longer manage.’
\verse And the manager said to himself, ‘What should I do, because my master is taking away the management from me? I am not strong enough to dig; I am ashamed to beg.
\verse I know what I should do, so that when I am removed from the management they will welcome me into their homes!’
\verse And he summoned each one of his own master’s debtors and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“summoned”) has been translated as a finite verb} said to the first, ‘How much do you owe my master?’
\verse And he said, ‘A hundred measures of olive oil.’ So he said to him, ‘Take your promissory note and sit down quickly and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“sit down”) has been translated as a finite verb} write fifty.’
\verse Then he said to another, ‘And how much do you owe?’ And he said, ‘A hundred measures of wheat.’ He said to him, ‘Take your promissory note and write eighty.’
\verse And the master praised the dishonest manager, because he had acted shrewdly. For the sons of this age are shrewder than the sons of light with regard to their own generation.\lebnote{Or “kind”}
\verse And I tell you, make friends for yourselves by means of unrighteous wealth, so that when it runs out they will welcome you into the eternal dwellings.
\verse “The one who is faithful in very little is also faithful in much, and the one who is dishonest in very little is also dishonest in much.
\verse If then you have not been faithful with unrighteous wealth, who will entrust to you the true riches?\lebnote{:|NP|:*The word “riches” is not in the Greek text but is implied}
\verse And if you have not been faithful with what belongs to another, who will give you your own?
\verse No domestic slave is able to serve two masters, for either he will hate the one and love the other, or he will be devoted to one and will despise the other. You are not able to serve God and money.”
\verseWithHeading{Hypocrisy, Law, and the Kingdom of God} Now the Pharisees, who were lovers of money, heard all these things, and they ridiculed him.
\verse And he said to them, “You are the ones who justify themselves in the sight of men, but God knows your hearts! For what is considered exalted among men is an abomination in the sight of God.
\verse “The law and the prophets were until John; from that time on the kingdom of God has been proclaimed, and everyone is urgently pressed\lebnote{The verb is translated here as a passive; some English versions translate the verb as active (“forces their way into it”)} into it.
\verse But it is easier for heaven and earth to pass away than for one stroke of a letter of the law to become invalid.
\verseWithHeading{On Divorce} “Everyone who divorces his wife and marries another commits adultery, and the one who marries a woman divorced from her husband commits adultery.
\verseWithHeading{The Rich Man and Lazarus} “Now a certain man was rich, and dressed in purple cloth and fine linen, feasting sumptuously every day.
\verse And a certain poor man \textit{named}\lebnote{Literally “by name”} Lazarus, covered with sores, lay at his gate,
\verse and was longing to be filled with what fell from the table of the rich man. But even the dogs came and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came”) has been translated as a finite verb} licked his sores.
\verse Now it happened that the poor man died, and he was carried away by the angels to \textit{Abraham’s side}.\lebnote{Literally “the bosom of Abraham”} And the rich man also died and was buried.
\verse And in Hades he lifted up his eyes as he\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the participle (“was”) which is understood as temporal} was in torment and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“lifted up”) has been translated as a finite verb} saw Abraham from a distance, and Lazarus \textit{at his side}.\lebnote{Literally “in his bosom”}
\verse And he called out and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“called out”) has been translated as a finite verb} said, ‘Father Abraham, have mercy on me, and send Lazarus so that he could dip the tip of his finger in water and cool my tongue, because I am suffering pain in this flame!’
\verse But Abraham said, ‘Child, remember that you received your good things during your life, and Lazarus likewise bad things. But now he is comforted here, but you are suffering pain.
\verse And in addition to all these things, a great chasm has been established between us and you, so that those who want to cross over from here to you are not able to do so,\lebnote{:|NP|:*The words “to do so” are not in the Greek text but are implied} nor can they cross over from there to us.’
\verse So he said, ‘Then I ask you, father, that you send him to my father’s house,
\verse for I have five brothers, so that he could warn them, in order that they also should not come to this place of torment!’
\verse But Abraham said, ‘They have Moses and the prophets; they must listen to them.’
\verse And he said, ‘No, father Abraham, but if someone from the dead goes to them, they will repent!’
\verse But he said to him, ‘If they do not listen to Moses and the prophets, neither will they be convinced if someone rises from the dead.’ ”
\end{biblechapter}

\begin{biblechapter} % Luke 17
\verseWithHeading{Sin, Forgiveness, Faith, and Service} And he said to his disciples, “It is impossible for causes for stumbling not to come, but woe to him through whom they come!
\verse It would be better for him if \textit{a millstone}\lebnote{Literally “a stone belonging to a mill”} is placed around his neck and he is thrown into the sea than that he causes one of these little ones to sin.
\verse “Be concerned about yourselves! If your brother sins, rebuke him, and if he repents, forgive him.
\verse And if he sins against you seven times in a day, and seven times he returns to you saying, ‘I repent,’ you must forgive him.”
\verse And the apostles said to the Lord, “Increase our faith!”
\verse So the Lord said, “If you have faith like a mustard seed, you could say to this mulberry tree, ‘Be uprooted and planted in the sea,’ and it would obey you.
\verse “And which of you who has a slave plowing or shepherding sheep\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} who comes in from the field will say to him, ‘Come here at once and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“come here”) has been translated as a finite verb} recline at the table’?
\verse Will he not rather say to him, ‘Prepare something that I may eat, and dress yourself to serve me while I eat and drink, and after these things you will eat and drink.’
\verse He will not \textit{be grateful}\lebnote{Literally “have gratitude”} to the slave because he did what was ordered, will he?\lebnote{:|NP|:*The negative construction in Greek anticipates a negative answer here, indicated by “will he”}
\verse Thus you also, when you have done all the \textit{things you were ordered to do},\lebnote{Literally “things that were ordered to you”} say, ‘We are unworthy slaves; we have done what we were obligated to do.’ ”
\verseWithHeading{Ten Lepers Cleansed} And it happened that while traveling toward Jerusalem, he was passing through \textit{the region between}\lebnote{Literally “through the midst”} Samaria and Galilee.
\verse And as\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the temporal genitive absolute participle (“was entering”)} he was entering into a certain village, ten men met him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation}﻿\lebnote{Some manuscripts explicitly state “him”}—lepers, who stood at a distance.
\verse And they raised their voices, saying, “Jesus, Master, have mercy on us!”
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} he said to them, “Go and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“go”) has been translated as a finite verb} show yourselves to the priests.” And it happened that as they were going, they were cleansed.
\verse But one of them, when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw that he was healed, turned back, praising God with a loud voice.
\verse And he fell on his face at his feet, giving thanks to him. And he was a Samaritan.
\verse So Jesus answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “Were not ten cleansed? And where are the nine?
\verse Was no one found to turn back and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“turn back”) has been translated as an infinitive} give praise to God except this foreigner?”
\verse And he said to him, “Get up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“get up”) has been translated as a finite verb} go your way. Your faith has saved you.”
\verseWithHeading{The Coming of the Kingdom of God} Now when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“was asked”) which is understood as temporal} was asked by the Pharisees when the kingdom of God would come, he answered them and said, “The kingdom of God does not come with \textit{things that can be observed},\lebnote{Literally “observation”}
\verse nor will they say, ‘Behold, here it is!’ or ‘There!’ For behold, the kingdom of God is in your midst.”
\verseWithHeading{The Coming of the Son of Man} And he said to the disciples, “Days will come when you will desire to see one of the days of the Son of Man, and you will not see it.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And they will say to you, ‘Behold, there!’ ‘Behold, here!’\lebnote{Some manuscripts have “ ‘Behold, there!’ or ‘Behold, here!’ ”} Do not go out or run after them!\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse For just as the lightning shines forth, flashing from one place under heaven to another place under heaven, so the Son of Man will be in his day.
\verse But first it is necessary for him to suffer many things, and to be rejected by this generation.
\verse And just as it was in the days of Noah, so also it will be in the days of the Son of Man—
\verse they were eating, they were drinking, they were marrying, they were being given in marriage, until the day Noah entered into the ark, and the flood came and destroyed them all.
\verse Likewise, just as it was in the days of Lot—they were eating, they were drinking, they were buying, they were selling, they were planting, they were building.
\verse But on the day that Lot went out from Sodom, it rained fire and sulphur from heaven and destroyed them all.
\verse It will be \textit{just the same}\lebnote{Literally “according to the same”} on the day that the Son of Man is revealed.
\verse On that day, whoever is on the housetop and his goods are in the house must not come down to take them away. And likewise the one who is in the field must not turn back
\verse Remember Lot’s wife!
\verse Whoever seeks to preserve his life will lose it, but whoever loses it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} will keep it.
\verse I tell you that in that night there will be two in one bed; one will be taken and the other will be left.
\verse There will be two women\lebnote{:|NP|:﻿*Here “women” is supplied because the form (“two”) is feminine gender in Greek} grinding at the same place; one will be taken and the other will be left.”\lebnote{A few manuscripts add v. 36 (with some variations): “There will be two in the field; one will be taken and the other will be left.”}
\verse And they answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to him, “Where, Lord?” So he said to them, “Where the dead body is, there also the vultures will be gathered.”
\end{biblechapter}

\begin{biblechapter} % Luke 18
\verseWithHeading{The Parable of the Unjust Judge} And he told them a parable to show that they must always pray and not be discouraged,
\verse saying, “There was a certain judge in a certain town who did not fear God and did not respect people.
\verse And there was a widow in that town, and she kept coming to him, saying, ‘Grant me justice against my adversary!’
\verse And he was not willing for a time, but after these things he said to himself, ‘Even if I do not fear God or respect people,
\verse yet because this widow is causing trouble for me, I will grant her justice, so that she does not wear me down in the end by her\lebnote{:|NP|:﻿*Here “by” is supplied as a component of the participle (“coming back”) which is understood as means} coming back!’ ”
\verse And the Lord said, “Listen to what the unrighteous judge is saying!
\verse And will not God surely \textit{see to it that justice is done}\lebnote{Literally “carry out the giving of justice”} to his chosen ones who cry out to him day and night, and will he delay toward them?
\verse I tell you that \textit{he will see to it that justice is done}\lebnote{Literally “he will carry out the giving of justice”} for them soon! Nevertheless, when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“comes”) which is understood as temporal} the Son of Man comes, then will he find faith on earth?”
\verseWithHeading{The Parable of the Pharisee and the Tax Collector} And he also told this parable to some who trusted in themselves that they were righteous, and looked down on \textit{everyone else}:\lebnote{Literally “the rest”}
\verse “Two men went up to the temple to pray, one a Pharisee and the other a tax collector.
\verse The Pharisee stood and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“stood”) has been translated as a finite verb} prayed these things with reference to himself: ‘God, I give thanks to you that I am not like other people—swindlers, unrighteous people, adulterers, or even like this tax collector!
\verse I fast twice a week; I give a tenth of all that I get.’
\verse But the tax collector, standing far away, did not want even to raise his eyes to heaven, but was beating his breast, saying, ‘God, be merciful to me, a sinner!’
\verse I tell you, this man went down to his house justified rather than that one! For everyone who exalts himself will be humbled, but the one who humbles himself will be exalted.”
\verseWithHeading{Little Children Brought to Jesus} Now they were bringing even their\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} babies to him so that he could touch them. But when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} the disciples saw it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} they rebuked them.
\verse But Jesus called them to himself, saying, “Allow the children to come to me, and do not forbid them, \textit{for to such belongs}\lebnote{Literally “for of such is”} the kingdom of God.
\verse Truly I say to you, whoever does not welcome the kingdom of God like a young child will never enter into it.”
\verseWithHeading{A Rich Young Man} And a certain ruler asked him, saying, “Good Teacher, by\lebnote{:|NP|:﻿*Here “by” is supplied as a component of the participle (“doing”) which is understood as means} doing what will I inherit eternal life?”
\verse And Jesus said to him, “Why do you call me good? No one is good except God alone.
\verse You know the commandments: ‘Do not commit adultery, do not murder, do not steal, do not give false testimony, honor your father and mother.’ ”\lebnote{:|NP|:A quotation from Exod 20:12–16; Deut 5:16–20}
\verse And he said, “All these I have observed from my\lebnote{Some manuscripts omit “my”} youth.”
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“heard”) which is understood as temporal} heard this,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} Jesus said to him, “\textit{You still lack one thing}:\lebnote{Literally “one still lacking to you”} Sell all that you have, and distribute the proceeds\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} to the poor—and you will have treasure in heaven—and come, follow me.”
\verse But when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“heard”) which is understood as temporal} heard these things he became very sad, because he was extremely wealthy.
\verse And Jesus took notice of him\lebnote{Some manuscripts add “becoming very sad”} and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took notice of”) has been translated as a finite verb} said, “How \textit{difficult it is for}\lebnote{Literally “with difficulty”} those who possess wealth to enter into the kingdom of God!
\verse For it is easier for a camel to go through the eye of a needle than for a rich person to enter into the kingdom of God.
\verse So those who heard this\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} said, “And who can be saved?”
\verse But he said, “What is impossible with men is possible with God.”
\verse And Peter said, “Behold, we have left \textit{all that is ours}\lebnote{Literally “our own”} and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“have left”) has been translated as a finite verb} followed you.”
\verse And he said to them, “Truly I say to you that there is no one who has left house or wife or brothers or parents or children on account of the kingdom of God,
\verse who will not receive many times more in this time and in the age to come, eternal life.”
\verseWithHeading{Jesus Predicts His Death and Resurrection a Third Time} And taking aside the twelve, he said to them, “Behold, we are going up to Jerusalem, and all the things that are written by the prophets with reference to the Son of Man will be accomplished.
\verse For he will be handed over to the Gentiles and will be mocked and mistreated and spit on,
\verse and after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“flogging”) which is understood as temporal} flogging him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} they will kill him, and on the third day he will rise.”
\verse And they understood none of these things, and this saying was concealed from them, and they did not comprehend the things that were said.
\verseWithHeading{A Blind Man Healed at Jericho} Now it happened that as he drew near to Jericho, a certain blind man was sitting on the side of the road begging.
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“heard”) which is understood as temporal} heard a crowd going by, he inquired what this meant.
\verse And they told him, “Jesus the Nazarene is passing by.”
\verse And he called out, saying, “Jesus, Son of David, have mercy on me!”
\verse And those who were in front rebuked him, that he should be silent, but he cried out \textit{even more loudly},\lebnote{Literally “by much more”} “Son of David, have mercy on me!”
\verse So Jesus stopped and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“stopped”) has been translated as a finite verb} ordered him to be brought to him. And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“approached”)} he approached, he asked him,
\verse “What do you want me to do for you?” And he said, “Lord, that I may regain my sight.
\verse And Jesus said to him, “Regain your sight! Your faith has saved you.”
\verse And immediately he regained his sight and began to follow\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to follow”)} him, glorifying God. And all the people, when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} gave praise to God.
\end{biblechapter}

\begin{biblechapter} % Luke 19
\verseWithHeading{Jesus and Zacchaeus} And he entered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“entered”) has been translated as a finite verb} traveled through Jericho.
\verse And \textit{there was}\lebnote{Literally “behold”} a man \textit{named}\lebnote{Literally “called by name”} Zacchaeus, and he was a chief tax collector, and he was rich.
\verse And he was seeking to see Jesus—who he was—and he was not able to as a result of the crowd, because he was short in stature.
\verse And he ran on ahead and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“ran on”) has been translated as a finite verb} climbed up into a sycamore tree so that he could see him, because he was going to go through that way.
\verse And when he came to the place, Jesus looked up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“looked up”) has been translated as a finite verb} said to him, “Zacchaeus, come down quickly, because it is necessary for me to stay at your house today!”
\verse And he came down quickly and welcomed him joyfully.
\verse And when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} they all began to complain,\lebnote{:|NP|:*The imperfect tense has been translated as ingressive here (“began to complain”)} saying, “He has gone in to find lodging with a man who is a sinner!”
\verse And Zacchaeus stopped and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“stopped”) has been translated as a finite verb} said to the Lord, “Behold, half of my possessions, Lord, I am giving to the poor, and if I have extorted anything from anyone, I am paying it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} back four times as much!”
\verse And Jesus said to him, “Today salvation has come to this house, because he too is a son of Abraham.
\verse For the Son of Man came to seek and to save those who are lost.”
\verseWithHeading{The Parable of the Ten Minas} Now while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“were listening to”)} they were listening to these things, he went on and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“went on”) has been translated as a finite verb} told a parable, because he was near Jerusalem and they thought that the kingdom of God was going to appear immediately.
\verse Therefore he said, “A certain nobleman traveled to a distant country to receive for himself a kingdom and to return.
\verse And summoning ten of his own slaves, he gave them ten minas\lebnote{A Greek monetary unit equal to 100 drachmas} and said to them, ‘Do business \textit{until I come back}.’\lebnote{Literally “in which time I am coming back”}
\verse But his citizens hated him, and sent a delegation after him, saying, ‘We do not want this man to be king over us!’
\verse And it happened that when he returned after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“receiving”) which is understood as temporal} receiving the kingdom, he ordered these slaves to whom he had given the money to be summoned to him, so that he could know what they had gained by trading.
\verse So the first arrived, saying, ‘Sir, your mina has made ten minas more!’
\verse And he said to him, ‘Well done, good slave! Because you have been faithful in a very small thing, \textit{have authority}\lebnote{Literally “be having authority”} over ten cities.’
\verse And the second came, saying, ‘Sir, your mina has made five minas.’
\verse So he said to this one also, ‘And you be over five cities.’
\verse And another came, saying, ‘Sir, behold your mina, which I had put away for safekeeping in a piece of cloth.
\verse For I was afraid of you, because you are a severe man—you withdraw what you did not deposit, and you reap what you did not sow!’
\verse He said to him, ‘\textit{By your own words}\lebnote{Literally “from your own mouth”} I will judge you, wicked slave! You knew that I am a severe man, withdrawing what I did not deposit and reaping what I did not sow.
\verse And why did you not give my money to the bank, and I, when I\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“returned”) which is understood as temporal} returned, would have collected it with interest?’
\verse And to the bystanders he said, ‘Take away from him the mina and give it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} to the one who has the ten minas!’
\verse And they said to him, ‘Sir, he has ten minas.’
\verse ‘I tell you that to everyone who has, more will be given. But from the one who does not have, even what he has will be taken away.
\verse But these enemies of mine who did not want me to be king over them—bring them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} here and slaughter them in my presence!’ ”
\verseWithHeading{The Triumphal Entry} And after he\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“had said”) which is understood as temporal} had said these things, he traveled on ahead, going up to Jerusalem.
\verse And it happened that when he drew near to Bethphage and Bethany, to the hill\lebnote{Or “mountain”} called the Mount of Olives, he sent two of the disciples,
\verse saying, ‘Go into the village in front of you, in which as you\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the participle (“enter”) which is understood as temporal} enter you will find a colt tied, on which \textit{no person has ever}\lebnote{Literally “no one of men ever”} sat, and untie it and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“untie”) has been translated as a finite verb} bring it.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And if anyone asks you, ‘Why are you untying it?’\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} you will say this: ‘The Lord has need of it.’ ”
\verse So those who were sent went and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“went”) has been translated as a finite verb} found it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} just as he had told them.
\verse And as\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the temporal genitive absolute participle (“were untying”)} they were untying the colt, its owners said to them, ‘Why are you untying the colt?’
\verse So they said, ‘The Lord has need of it.’
\verse And they brought it to Jesus, and throwing their cloaks on the colt, they put Jesus on it.
\verse And as\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the temporal genitive absolute participle (“was going along”)} he was going along, they were spreading out their cloaks on the road.
\verse Now as\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the temporal genitive absolute participle (“was drawing near”)} he was drawing near by this time to the descent from the Mount of Olives, the whole crowd of the disciples began rejoicing to praise God with a loud voice for all the miracles that they had seen,
\verse saying, “Blessed is the king, 
the one who comes in the name of the Lord!\lebnote{:|NP|:A quotation from Ps 118:26, with “the king” added as a clarification} 
Peace in heaven 
and glory in the highest!”
\verse And some of the Pharisees from the crowd said to him, “Teacher, rebuke your disciples!”
\verse And he answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “I tell you that\lebnote{Some manuscripts omit “that”} if these keep silent, the stones will cry out!”
\verseWithHeading{Jesus Weeps over Jerusalem} And when he approached and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“approached”) has been translated as a finite verb} saw the city, he wept over it,
\verse saying, “If you had known on this day—even you—the things that make for peace! But now they are hidden from your eyes.
\verse For days will come upon you and your enemies will put up an embankment\lebnote{Or “a palisade” (the term can refer to either a wooden or an earthen barricade)} against you, and will surround you and press you hard from all directions.
\verse And they will raze you to the ground, you and your children within you, and will not leave a stone upon a stone within you, \textit{because}\lebnote{Literally “in return for which”} you did not recognize the time of your visitation.”
\verseWithHeading{The Cleansing of the Temple} And he entered into the temple courts\lebnote{:|NP|:﻿*Here “courts” is supplied to distinguish this area from the interior of the temple building itself} and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“entered”) has been translated as a finite verb} began to drive out those who were selling,
\verse saying to them, “It is written, ‘And my house will be a house of prayer,’\lebnote{:|NP|:A quotation from Isa 56:7} but you have made it a cave of robbers!”
\verse And he was teaching every day in the temple courts,\lebnote{:|NP|:*Here “courts” is supplied to distinguish this area from the interior of the temple building itself} and the chief priests and the scribes and the most prominent men of the people were seeking to destroy him.
\verse And they did not find anything they could do, because all the people were paying close attention to what they were hearing from him.
\end{biblechapter}

\begin{biblechapter} % Luke 20
\verseWithHeading{Jesus’ Authority Challenged} And it happened that on one of the days while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“was teaching”)} he was teaching the people in the temple courts\lebnote{:|NP|:﻿*Here “courts” is supplied to distinguish this area from the interior of the temple building itself} and proclaiming the gospel, the chief priests and the scribes approached together with the elders
\verse and said, saying to him, “Tell us, by what authority you are doing these things, or who is the one who gave you this authority?
\verse And he answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to them, “I also will ask you a question, and you tell me:
\verse The baptism of John—was it from heaven or from men?
\verse And they discussed this\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} with one another, saying, “If we say ‘From heaven,’ he will say, ‘Why did you not believe him?’
\verse But if we say, ‘From men,’ all the people will stone us to death, because they are convinced that John was a prophet.”
\verse And they replied that they did not know where it was from.
\verse And Jesus said to them, “Neither will I tell you by what authority I am doing these things.”
\verseWithHeading{The Parable of the Tenant Farmers in the Vineyard} And he began to tell the people this parable: “A man\lebnote{Some manuscripts have “A certain man”} planted a vineyard, and leased it to tenant farmers, and went on a journey for a long time.
\verse And at the proper time he sent a slave to the tenant farmers, so that they would give him some of the fruit of the vineyard. But the tenant farmers sent him away empty-handed after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“beating”) which is understood as temporal} beating him.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And he proceeded to send another slave, but they beat and dishonored that one also, and\lebnote{:|NP|:﻿*Here “and” is supplied because the two previous participles (“beat” and “dishonored”) have been translated as finite verbs} sent him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} away empty-handed.
\verse And he proceeded to send a third, but they wounded and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“wounded”) has been translated as a finite verb} threw out this one also.
\verse So the owner of the vineyard said, ‘What should I do? I will send my beloved son; perhaps they will respect him.’
\verse But when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} the tenant farmers saw him, they began to reason\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to reason”)} with one another, saying, ‘This is the heir. Let us kill him so that the inheritance will become ours!’
\verse And they threw him out of the vineyard and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“threw”) has been translated as a finite verb} killed him.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} What then will the owner of the vineyard do to them?
\verse He will come and destroy those tenant farmers and give the vineyard to others.” And when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“heard”) which is understood as temporal} heard this,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} they said, “May this never happen!”
\verse But he looked intently at them and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“looked intently at”) has been translated as a finite verb} said, “What then is this that is written: ‘The stone which the builders rejected, 
this has become \textit{the cornerstone}.’\lebnote{Literally “the head of the corner”}
\verse Everyone who falls on that stone will be broken to pieces, and the one on whom it falls—it will crush him!”
\verse And the scribes and the chief priests sought to lay their\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} hands on him at that same hour, and they were afraid of the people, for they knew that he had told this parable with reference to them.
\verseWithHeading{Paying Taxes to Caesar} And they watched him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} closely and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“watched closely”) has been translated as a finite verb} sent spies who pretended they were upright, in order that they could catch him in a statement, so that they could hand him over to the authority and the jurisdiction of the governor.
\verse And they asked him, saying, “Teacher, we know that you speak and teach rightly, and do not \textit{show partiality},\lebnote{Literally “receive face”} but teach the way of God in truth.
\verse Is it permitted for us to pay taxes\lebnote{Or “the tribute tax”} to Caesar or not?”
\verse But seeing through their craftiness, he said to them,
\verse “Show me a denarius! Whose image and inscription does it have?” And they answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said,\lebnote{Some manuscripts have “And they said”} “Caesar’s.”
\verse So he said to them, “Well then, give to Caesar the things of Caesar, and to God the things of God!”
\verse And they were not able to catch him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation}﻿\lebnote{Some manuscripts explicitly state “him”} in a statement in the sight of the people, and astonished at his answer, they became silent.
\verseWithHeading{A Question About Marriage and the Resurrection} Now some of the Sadducees—who deny \textit{that there is a resurrection}\lebnote{Literally “resurrection not to exist”}—came up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“came up”) has been translated as a finite verb} asked him,
\verse saying, “Teacher, Moses wrote for us if someone’s brother dies having a wife, and this man is childless, that his brother should take the wife and \textit{father}\lebnote{Literally “raise up”} descendants for his brother.
\verse Now there were seven brothers, and the first took a wife and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took”) has been translated as a finite verb} died childless,
\verse and the second,
\verse and the third took her, and likewise also the seven did not leave children and died.
\verse Finally the woman also died.
\verse Therefore in the resurrection, the woman—whose wife will she be? For the seven had her as wife.”
\verse And Jesus said to them, “The sons of this age marry and are given in marriage,
\verse but those who are considered worthy to attain to that age and to the resurrection from the dead neither marry nor are given in marriage,
\verse for they are not even able to die any longer, because they are like the angels and are sons of God, because they\lebnote{:|NP|:﻿*Here “because” is supplied as a component of the participle (“are”) which is understood as causal} are sons of the resurrection.
\verse But that the dead are raised, even Moses revealed in the passage about\lebnote{:|NP|:﻿*The words “the passage about” are not in the Greek text but are implied; here a common form of rabbinic citation is being used to refer to an Old Testament passage} the bush, when he calls the Lord the God of Abraham and the God of Isaac and the God of Jacob.
\verse Now he is not God of the dead, but of the living, for all live to him!”
\verse And some of the scribes answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “Teacher, you have spoken well.”
\verse For they no longer dared to ask him anything.
\verseWithHeading{David’s Son and Lord} But he said to them, “In what sense do they say that the Christ is David’s son?
\verse For David himself says in the book of Psalms,
\verse ‘The Lord said to my Lord, 
“Sit at my right hand,
\verse David therefore calls him ‘Lord,’ and how is he his son?”
\verseWithHeading{Warning to Beware of the Scribes} And while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“were listening”)} all the people were listening, he said to the disciples,\lebnote{Some manuscripts have “to his disciples”}
\verse “Beware of the scribes, who like walking around in long robes and who love greetings in the marketplaces and the best seats in the synagogues and the places of honor at banquets,
\verse who devour the houses of widows and pray lengthy prayers for the sake of appearance. These will receive more severe condemnation!”
\end{biblechapter}

\begin{biblechapter} % Luke 21
\verseWithHeading{A Poor Widow’s Offering} And he looked up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“looked up”) has been translated as a finite verb} saw the rich putting their gifts into the contribution box,
\verse and he saw a certain poor widow putting in there two small copper coins.\lebnote{This coin was the \textit{lepton}, worth 1/128 of a denarius}
\verse And he said, “Truly I say to you that this poor widow put in more than all of them.
\verse For these all put gifts\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} into the offering out of their abundance, but this woman out of her poverty put in all the means of subsistence that she had.”
\verseWithHeading{The Destruction of the Temple Predicted} And while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“were speaking”)} some were speaking about the temple, that it was adorned with precious stones and votive offerings, he said,
\verse “As for these things that you see—days will come in which not one stone will be left on another stone that will not be thrown down!”
\verseWithHeading{Signs of the End of the Age} And they asked him, saying, “Teacher, when therefore will these things happen, and what will be the sign when these things are about to take place?”
\verse And he said, Watch out that you are not deceived! For many will come in my name, saying, ‘I am he,’ and ‘The time is near!’ Do not go after them!
\verse And when you hear about wars and insurrections, do not be terrified, for these things must happen first, but the end will not be at once.”
\verse Then he said to them, “nation will rise up against nation and kingdom against kingdom.
\verse There will be great earthquakes and famines and plagues in various places. There will be terrible sights and great signs from heaven.
\verseWithHeading{Persecution of Disciples Predicted} “But before all these things, they will lay their hands on you and will persecute you,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} handing you\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} over to the synagogues and prisons. You will be brought before\lebnote{:|NP|:﻿*Here the participle (“be brought before”) is translated as a finite verb because of English style} kings and governors because of my name.
\verse This will turn out to you for a time of witness.
\verse \textit{Therefore make up your minds}\lebnote{Literally “therefore place in your hearts”} not to prepare in advance to speak in your own defense,
\verse for I will give you a mouth and wisdom that all your opponents will not be able to resist or contradict.
\verse And you will be handed over even by parents and brothers and relatives and friends, and they will put to death some of you.
\verse And you will be hated by all because of my name.
\verse Even a hair of your head will never perish!
\verse By your patient endurance you will gain your lives.
\verseWithHeading{The Desolation of Jerusalem} “But when you see Jerusalem surrounded by armies, then know that its desolation has come near.
\verse Then those in Judea must flee to the mountains, and those \textit{inside it}\lebnote{Literally “in the midst of it”} must depart, and those in the fields must not enter into it,
\verse because these are days of vengeance, so that all the things that are written can be fulfilled.
\verse Woe to those \textit{who are pregnant}\lebnote{Literally “who have in the womb”} and to those who are nursing their babies\lebnote{:|NP|:﻿*The words “their babies” are not in the Greek text but are supplied as a necessary clarification} in those days! For there will be great distress on the earth and wrath against this people,
\verse and they will fall by the edge of the sword, and will be led captive into all the nations,\lebnote{The same Greek word, occurring three times in this verse, can be translated “nations” or “Gentiles” depending on the context} and Jerusalem will be trampled down by the Gentiles until the times of the Gentiles are fulfilled.
\verseWithHeading{The Arrival of the Son of Man} “And there will be signs in the sun and moon and stars, and on the earth distress of nations in perplexity from the noise of the sea and its surging,
\verse people fainting from fear and expectation of what is coming on the inhabited earth, for the powers of the heavens will be shaken.\lebnote{An allusion to Isa 34:4}
\verse And then they will see the Son of Man arriving in a cloud\lebnote{An allusion to Dan 7:13} with power and great glory.
\verse But when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“begin”)} these things begin to happen, stand up straight and raise your heads, because your redemption is drawing near!”
\verseWithHeading{The Parable of the Fig Tree} And he told them a parable: “Look at the fig tree and all the trees.
\verse When they put out foliage,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} now you see for yourselves and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“see”) has been translated as a finite verb} know that by this time the summer is near.
\verse So also you, when you see these things happening, know\lebnote{Or “you know”} that the kingdom of God is near.
\verse Truly I say to you that this generation will never pass away until all things take place!
\verse Heaven and earth will pass away, but my words will never pass away.
\verseWithHeading{Be Alert} “But take care for yourselves, lest your hearts are weighed down with dissipation and drunkenness and the worries of daily life, and that day come upon you suddenly
\verse like a trap. For it will come upon all who reside on the face of the whole earth.
\verse But be alert at all times, praying that you may have strength to escape all these things that are going to happen, and to stand before the Son of Man.”
\verse So throughout the days he was teaching in the temple courts,\lebnote{:|NP|:*Here “courts” is supplied to distinguish this area from the interior of the temple building itself} and throughout the nights he was going out and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“was going out”) has been translated as a finite verb} spending the night on the hill that is called the Mount of Olives.
\verse And all the people were getting up very early in the morning to come\lebnote{:|NP|:﻿*The words “to come” are not in the Greek text but are implied} to him in the temple courts\lebnote{:|NP|:﻿*Here “courts” is supplied to distinguish this area from the interior of the temple building itself} to listen to him.
\end{biblechapter}

\begin{biblechapter} % Luke 22
\verseWithHeading{The Chief Priests and Scribes Plot to Kill Jesus} Now the feast of Unleavened Bread (which is called Passover) was drawing near.
\verse And the chief priests and the scribes were seeking how they could destroy him, because they were afraid of the people.
\verseWithHeading{Judas Arranges to Betray Jesus} And Satan entered into Judas, the one called Iscariot, who was of the number of the twelve.
\verse And he went away and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“went away”) has been translated as a finite verb} discussed with the chief priests and officers of the temple guard how he could betray him to them.
\verse And they were delighted, and came to an agreement with him to give him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} money.
\verse And he agreed, and began looking\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began looking”)} for a favorable opportunity to betray him to them apart from the crowd.
\verseWithHeading{Jesus’ Final Passover with the Disciples} And the day of the feast of Unleavened Bread came, on which it was necessary for the Passover lamb to be sacrificed.
\verse And he sent Peter and John, saying, “Go and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“go”) has been translated as a finite verb} prepare the Passover for us, so that we may eat it.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse So they said to him, “Where do you want us to prepare it?”\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse And he said to them, “Behold, when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“have entered”)} you have entered into the city, a man carrying a jar of water will meet you. Follow him into the house which he enters.
\verse And you will say to the master of the house, ‘The Teacher says to you, “Where is the guest room where I may eat the Passover with my disciples?” ’
\verse And he will show you a large furnished\lebnote{Or perhaps “paved” or “panelled”} upstairs room. Make preparations there.”
\verse So they went and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“went”) has been translated as a finite verb} found everything\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} just as he had told them, and they prepared the Passover.
\verseWithHeading{The Lord’s Supper} And when the hour came, he reclined at the table, and the apostles with him.
\verse And he said to them, “\textit{I have earnestly desired}\lebnote{Literally “I have desired with desire”} to eat this Passover with you before I suffer.
\verse For I tell you that I will not eat it until it is fulfilled in the kingdom of God.”
\verse And he took in hand a cup, and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took in hand”) has been translated as a finite verb} after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the temporal participle (“giving thanks”)} giving thanks he said, “Take this and share it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} among yourselves.
\verse For I tell you,\lebnote{Some manuscripts have “I tell you that”} from now on I will not drink of the product of the vine until the kingdom of God comes.”
\verse And he took bread, and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took”) has been translated as a finite verb} after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the temporal participle (“giving thanks”)} giving thanks, he broke it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} and gave it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} to them, saying, “This is my body which is given for you. Do this in remembrance of me.”
\verse And in the same way the cup after they had eaten, saying, “This cup is the new covenant in my blood which is poured out for you.
\verse “But behold, the hand of the one who is betraying me is with me on the table!
\verse For the Son of Man is going according to what has been determined, but woe to that man by whom he is betrayed!”
\verse And they began to debate with one another who then of them it could be who was going to do this.
\verseWithHeading{A Dispute About Who Is Greatest} And a dispute also occurred among them as to which of them was recognized as being greatest.
\verse So he said to them, “The kings of the Gentiles\lebnote{The same Greek word can be translated “nations” or “Gentiles” depending on the context} lord it over them, and those who have authority over them are called benefactors.
\verse But you are not to be like this! But the one who is greatest among you must become like the youngest, and the one who leads like the one who serves.
\verse For who is greater, the one who reclines at the table or the one who serves? Is it not the one who reclines at the table? But I am in your midst as the one who serves.
\verse “And you are the ones who have remained\lebnote{Or “ones who have continued”} with me in my trials,
\verse and I confer on you a kingdom, just as my Father conferred on me,
\verse that you may eat and drink at my table in my kingdom, and you will sit on thrones judging the twelve tribes of Israel.
\verseWithHeading{Jesus Predicts Peter’s Denial} “Simon, Simon, behold, Satan has demanded to sift you like wheat,
\verse but I have prayed for you, that your faith may not fail. And you, when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“have turned back”) which is understood as temporal} once you have turned back,\lebnote{Or “have turned around”} strengthen your brothers.”
\verse But he said to him, “Lord, I am ready to go with you both to prison and to death!”
\verse And he said, “I tell you, Peter, the rooster will not crow today until you have denied three times that you know me!”
\verseWithHeading{The Two Swords} And he said to them, “When I sent you out without a money bag and a traveler’s bag and sandals, you did not lack anything, did you?”\lebnote{:|NP|:*The negative construction in Greek anticipates a negative answer here, indicated in the translation by “did you”} And they said, “Nothing.”
\verse And he said to them, “But now the one who has a money bag must take it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} and likewise a traveler’s bag. And the one who does not have a sword must sell his cloak and buy one.
\verse For I tell you that this that is written must be fulfilled in me: ‘And he was counted with the criminals.’\lebnote{:|NP|:A quotation from Isa 53:12} For indeed, what is written\lebnote{:|NP|:﻿*The phrase “what is written” is not in the Greek text but is an understood repetition of the similar phrase at the beginning of the verse} about me \textit{is being fulfilled}.”\lebnote{Literally “is having an end”}
\verse So they said, “Lord, behold, here are two swords!” And he said to them, “It is adequate.”
\verseWithHeading{The Prayer in Gethsemane} And he went away and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“went away”) has been translated as a finite verb} proceeded, according to his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} custom, to the Mount of Olives, and the disciples also followed him.
\verse And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“came”) which is understood as temporal} he came to the place, he said to them, “Pray that you will not enter into temptation.”
\verse And he withdrew from them about a stone’s throw and \textit{knelt down}\lebnote{Literally “bent his knees”} and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“knelt down”; literally “bent his knees”) has been translated as a finite verb} began to pray,\lebnote{:|NP|:*The imperfect tense has been translated as ingressive here (“began to pray”)}
\verse saying, “Father, if you are willing, take away this cup from me. Nevertheless, not my will but yours be done.”
\verse \lebnote{And an angel from heaven appeared to him, strengthening him.
\verse And being in anguish, he began praying\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began praying”)} more fervently and his sweat became like drops of blood falling down to the ground.}\lebnote{A number of early and important Greek manuscripts lack verses 43 and 44}
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“got up”) which is understood as temporal} got up from the prayer and\lebnote{:|NP|:﻿*Here “and” is supplied because the participle (“came”) has been translated as a finite verb in keeping with English style} came to the disciples, he found them sleeping from sorrow,
\verse and he said to them, “Why are you sleeping? Get up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“get up”) has been translated as a finite verb} pray that you will not enter into temptation!”
\verseWithHeading{The Betrayal and Arrest of Jesus} While\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“speaking”)} he was still speaking, behold, there came a crowd, and the one named Judas, one of the twelve, leading them. And he approached Jesus to kiss him.
\verse But Jesus said to him, “Judas, are you betraying the Son of Man with a kiss?”
\verse And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} those around him saw what was about to happen, they said, “Lord, should we strike with the sword?”
\verse And a certain one of them struck the slave of the high priest and cut off his right ear.
\verse But Jesus answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “\textit{Stop! No more of this!}”\lebnote{Literally “leave off to this”} And he touched his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} ear and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“touched”) has been translated as a finite verb} healed him.
\verse And Jesus said to the chief priests and officers of the temple and elders who had come out against him, “Have you come out with swords and clubs, as against a robber?
\verse Every day when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the temporal genitive absolute participle (“was”)} I was with you in the temple courts,\lebnote{:|NP|:*Here “courts” is supplied to distinguish this area from the interior of the temple building itself} you did not stretch out your\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} hands against me! But this is your hour and the domain of darkness!”
\verseWithHeading{Jesus Before the Sanhedrin} And they arrested him and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“arrested”) has been translated as a finite verb} led him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} away and brought him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} into the house of the high priest. But Peter was following at a distance.
\verse And when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“had kindled”) which is understood as temporal} had kindled a fire in the middle of the courtyard and sat down together, Peter sat down among them.
\verse And a certain female slave, seeing him sitting in the light and looking intently at him, said, “This man also was with him!”
\verse But he denied it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} saying, “Woman, I do not know him!”
\verse And after a short time another person saw him and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“saw”) has been translated as a finite verb} said, “You also are one of them!” But Peter said, “Man, I am not!”
\verse And after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the temporal genitive absolute participle (“had passed”)} about one hour had passed, someone else was insisting, saying, “In truth this man also was with him, because he is also a Galilean!”
\verse But Peter said, “Man, I do not know what you are talking about!” And immediately, while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“was speaking”)} he was still speaking, a rooster crowed.
\verse And the Lord turned around and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“turned around”) has been translated as a finite verb} looked intently at Peter. And Peter remembered the word of the Lord,\lebnote{Some manuscripts have “the Lord’s statement”} how he said to him, “Before the rooster crows today, you will deny me three times.”
\verse And he went outside and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“went”) has been translated as a finite verb} wept bitterly.
\verse And the men who were guarding him began to mock\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to mock”)} him while\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“beat”) which is understood as temporal} they beat him,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse and after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“blindfolding”) which is understood as temporal} blindfolding him they repeatedly asked\lebnote{:|NP|:﻿*This imperfect verb is translated as iterative (“repeatedly asked”)} him,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} saying, “Prophesy! Who is the one who struck you?”
\verse And they were saying many other things against him, reviling him.\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verseWithHeading{Jesus Before the Sanhedrin} And when day came, the council of elders of the people gathered, both chief priests and scribes, and they led him away to their Sanhedrin,\lebnote{Or “council”}
\verse saying, “If you are the Christ, tell us!” But he said to them, “If I tell you, you will never believe,
\verse and if I ask you,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} you will never answer!
\verse But from now on the Son of Man will be seated at the right hand of the power of God.”
\verse So they all said, “Are you then the Son of God?” And he said to them, “You say that I am.”
\verse And they said, “Why do we have need of further testimony? For we ourselves have heard it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} from his mouth!”
\end{biblechapter}

\begin{biblechapter} % Luke 23
\verseWithHeading{Jesus Brought Before Pilate} And the whole assembly of them rose up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“rose up”) has been translated as a finite verb} brought him before Pilate.
\verse And they began to accuse him, saying, “We have found this man misleading our nation and forbidding us\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} to pay taxes to Caesar, and saying he himself is Christ, a king!”
\verse And Pilate asked him, saying, “Are you the king of the Jews?” And he answered him and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said, “You say so.”
\verse So Pilate said to the chief priests and the crowds, “I find no basis for an accusation against this man.”
\verse But they insisted, saying, “He incites the people, teaching throughout the whole of Judea and beginning from Galilee as far as here.”
\verseWithHeading{Jesus Brought Before Herod} Now when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“heard”) which is understood as temporal} Pilate heard this,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} he asked if the man was a Galilean.
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“found out”) which is understood as temporal} found out that he was from the jurisdiction of Herod, he sent him over to Herod, who was also in Jerusalem in those days.
\verse And when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} Herod saw Jesus, he was very glad, for he had been wanting to see him for a long time, because he had heard about him and was hoping to see some miracle performed by him.
\verse So he questioned him \textit{at considerable length},\lebnote{Literally “with many words”} but he answered nothing to him.
\verse And the chief priests and the scribes were standing there vehemently accusing him.
\verse And Herod with his soldiers also treated him with contempt, and after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“mocking”) which is understood as temporal} mocking him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} and\lebnote{:|NP|:﻿*Here “and” is supplied to connect the two participles (“mocking” and “dressing”) in keeping with English style} dressing him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} in glistening clothing, he sent him back to Pilate.
\verse And both Herod and Pilate became friends with one another on that same day, \textit{for they had previously been enemies of one another}.\lebnote{Literally “for they had previously existed being at enmity with each other”}
\verseWithHeading{Pilate Releases Barabbas} So Pilate called together the chief priests and the rulers and the people
\verse and\lebnote{:|NP|:﻿*Here “and” is supplied because the participle in the previous verse (“called together”) has been translated as a finite verb} said to them, “You brought me this man as one who was misleading the people, and behold, when I\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“examined”) which is understood as temporal} examined him\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} before you, I found nothing in this man as basis for the accusation which \textit{you are making}\lebnote{Literally “you are accusing”} against him.
\verse But neither did Herod, because he sent him back to us. And behold, nothing deserving death \textit{has been done}\lebnote{Literally “is having been done”} by him.
\verse Therefore I will punish him and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“punish”) has been translated as a finite verb} release him.”\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} \lebnote{Many of the most important Greek manuscripts lack v. 17, “Now he was obligated to release for them at the feast one person.”}
\verse But they all cried out in unison, saying, “Take this man away, and release for us Barabbas!”
\verse (who had been thrown in prison because of a certain insurrection that had taken place in the city, and for murder).
\verse And Pilate, wanting to release Jesus, addressed them\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation}﻿\lebnote{Some manuscripts explicitly state “them”} again,
\verse but they kept crying out, saying, “Crucify! Crucify him!”
\verse So he said to them a third time, “Why? What wrong has this man done? I found no basis for an accusation \textit{deserving death}\lebnote{Literally “of death”} in him. Therefore I will punish him and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“punish”) has been translated as a finite verb} release him.”\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation}
\verse But they were urgent, demanding with loud cries that he be crucified. And their cries prevailed.
\verse And Pilate decided that their demand should be granted.
\verse And he released the one who had been thrown into prison because of insurrection and murder, whom they were asking for, but Jesus he handed over to their will.
\verseWithHeading{Jesus Is Crucified} And as they led him away, they seized Simon, a certain man of Cyrene, who was coming from the country, and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“seized”) has been translated as a finite verb} placed the cross on him, to carry it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} behind Jesus.
\verse And a great crowd of the people were following him, and of women who were mourning and lamenting him.
\verse But turning to them, Jesus said, “Daughters of Jerusalem, do not weep for me, but weep for yourselves and for your children!
\verse For behold, days are coming in which they will say, ‘Blessed are the barren, and the wombs that did not give birth, and the breasts that did not nurse!’
\verse Then they will begin to say to the mountains, ‘Fall on us!’ and to the hills, ‘Cover us!’
\verse For if they do these things \textit{when the wood is green},\lebnote{Literally “in the green wood”} what will happen \textit{when it is dry}?”\lebnote{Literally “in the dry”}
\verse And two other criminals were also led away to be executed with him.
\verse And when they came to the place that is called “The Skull,” there they crucified him, and the criminals, the one on his right and the other on his left.
\verse \lebnote{But Jesus said, “Father, forgive them, for they do not know what they are doing.”}\lebnote{Many important manuscripts lack v. 34a, “But Jesus said, ‘Father, forgive them, for they do not know what they are doing.’ ”} And they cast lots to divide his clothes.
\verse And the people stood there watching, but the rulers also ridiculed him,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} saying, “He saved others; let him save himself, if this man is the Christ\lebnote{Or “Messiah”} of God, the Chosen One!”
\verse And the soldiers also mocked him, coming up and\lebnote{:|NP|:﻿*Here “and” is supplied to connect the two participles (“coming up” and “offering”) in keeping with English style} offering him sour wine
\verse and saying, “If you are the king of the Jews, save yourself!”
\verse And there was also an inscription over him, “This is the king of the Jews.”
\verse And one of the criminals who were hanged there reviled him, saying, “Are you not the Christ? Save yourself—and us!”
\verse But the other answered and\lebnote{:|NP|:﻿*Here “and” is supplied to connect the two participles (“answered” and “rebuked”) in keeping with English style} rebuked him, saying, “Do you not even fear God, because you are undergoing the same condemnation?
\verse And we indeed justly, for we are receiving \textit{what we deserve}\lebnote{Literally “things worthy”} for what we have done. But this man has done nothing wrong!”
\verse And he said, “Jesus, remember me when you come into your kingdom!”
\verse And he said to him, “Truly I say to you, today you will be with me in paradise.”
\verseWithHeading{Jesus Dies on the Cross} And by this time it was about the sixth hour, and darkness came over the whole land until the ninth hour
\verse because\lebnote{:|NP|:﻿*Here “because” is supplied as a component of the participle (“failed”) which is understood as causal} the light of the sun failed. And the curtain of the temple was torn apart down the middle.
\verse And Jesus, calling out with a loud voice, said, “Father, into your hands I entrust my spirit!”\lebnote{:|NP|:A quotation from Ps 31:5} And after he\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“said”) which is understood as temporal} said this, he expired.
\verse Now when\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} the centurion saw what had happened, he began to praise\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to praise”)} God, saying, “Certainly this man was righteous!”
\verse And all the crowds that had come together for this spectacle, when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“saw”) which is understood as temporal} saw the things that had happened, returned home\lebnote{:|NP|:﻿*The word “home” is not in the Greek text, but is implied} beating their\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} breasts.
\verse And all his acquaintances, and the women who had followed him from Galilee who saw these things, stood at a distance.
\verseWithHeading{Jesus Is Buried} And behold, a man \textit{named}\lebnote{Literally “by name”} Joseph, who was a member of the council,\lebnote{Or “a member of the Sanhedrin”} a good\lebnote{Some manuscripts have “and a good”} and righteous man
\verse (this man was not consenting to their plan and deed), from Arimathea, \textit{a Judean town},\lebnote{Literally “a town of the Jews”} who was looking forward to the kingdom of God.
\verse This man approached Pilate and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“approached”) has been translated as a finite verb} asked for the body of Jesus.
\verse And he took it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} down and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“wrapped”) has been translated as a finite verb} wrapped it in a linen cloth and placed him in a tomb cut into the rock where no one had ever been placed.
\verse And it was the day of preparation, and the Sabbath was drawing near.
\verse And the women who had been accompanying him from Galilee followed and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“followed”) has been translated as a finite verb} saw the tomb and how his body was placed.
\verse And they returned and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“returned”) has been translated as a finite verb} prepared fragrant spices and perfumes, and on the Sabbath they rested according to the commandment.
\end{biblechapter}

\begin{biblechapter} % Luke 24
\verseWithHeading{Jesus Is Raised} Now on the first day of the week, at very early dawn, they came back to the tomb bringing the fragrant spices which they had prepared.
\verse And they found the stone had been rolled away from the tomb,
\verse but when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“went in”) which is understood as temporal} went in, they did not find the body.\lebnote{Some manuscripts add “of the Lord Jesus”}
\verse And it happened that while they were perplexed about this, behold, two men in gleaming clothing stood near them.
\verse And as\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the temporal genitive absolute participles (“were” and “bowed”)} they were terrified and bowed their faces to the ground, they said to them, “Why are you looking for the living among the dead?
\verse He is not here, but has been raised! Remember how he spoke to you while he\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the participle (“was”) which is understood as temporal} was still in Galilee,
\verse saying that the Son of Man must be delivered into the hands of men who are sinners, and be crucified, and on the third day rise?”
\verse And they remembered his words,
\verse and when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“returned”) which is understood as temporal} returned from the tomb, they reported all these things to the eleven and to all the rest.
\verse Now Mary Magdalene and Joanna and Mary the mother of James and the others with them were telling these things to the apostles.
\verse And these words appeared to them as nonsense, and they refused to believe them.
\verse But Peter got up and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“got up”) has been translated as a finite verb} ran to the tomb, and bending over to look, he saw only the strips of linen cloth, and he went away to his home\lebnote{Or “wondering to himself,” if the prepositional phrase modifies the following participle} wondering what had happened.
\verseWithHeading{Jesus Encountered on the Road to Emmaus} And behold, on that same day, two of them were traveling to a village \textit{named}\lebnote{Literally “to which the name”} Emmaus that was sixty stadia\lebnote{A “stade” or “stadium” (plur. “stadia”) is about 607 ft (187 m)} distant from Jerusalem,
\verse and they were conversing with one another about all these things that had happened.
\verse And it happened that while they were conversing, and discussing, Jesus himself also approached and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“approached”) has been translated as a finite verb} began to go along with\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began to go along with”)} them,
\verse but their eyes were prevented from recognizing him.
\verse And he said to them, “What are these matters that you are discussing with one another as you\lebnote{:|NP|:﻿*Here “as” is supplied as a component of the participle (“walking along”) which is understood as temporal} are walking along?” And they stood still, looking sad.
\verse And one of them, \textit{named}\lebnote{Literally “by name”} Cleopas, answered and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“answered”) has been translated as a finite verb} said to him, “Are you the only one living near Jerusalem and not knowing the things that have happened in it in these days?”
\verse And he said to them, “What things?” So they said to him, “The things concerning Jesus the Nazarene, a man who was a prophet, powerful in deed and word before God and all the people,
\verse and how our chief priests and rulers handed him over to a sentence of death, and crucified him.
\verse But we were hoping that he was the one who was going to redeem Israel. But in addition to all these things, this is the third day \textit{since}\lebnote{Literally “he spends from which”} these things took place.
\verse But also some women from among us astonished us, who were at the tomb early in the morning,
\verse and when they\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“find”) which is understood as temporal} did not find his body, they came back saying they had seen even a vision of angels, who said that he was alive!
\verse And some of those with us went out to the tomb and found it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} like this, just as the women had also said, but him they did not see.”
\verse And he said to them, “O foolish and slow in heart to believe in all that the prophets have spoken!
\verse Was it not necessary that the Christ suffer these things and enter into his glory?”
\verse And beginning from Moses and from all the prophets, he interpreted to them the things concerning himself in all the scriptures.
\verse And they drew near to the village where they were going, and he acted as though he was going farther.
\verse And they urged him strongly, saying, “Stay with us, because it is getting toward evening, and by this time the day is far spent.” And he went in to stay with them.
\verse And it happened that when he reclined at the table with them, he took the bread and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took”) has been translated as a finite verb} gave thanks, and after\lebnote{:|NP|:﻿*Here “after” is supplied as a component of the participle (“breaking”) which is understood as temporal} breaking it,\lebnote{:|NP|:*Here the direct object is supplied from context in the English translation} he gave it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} to them.
\verse And their eyes were opened, and they recognized him, and he became invisible to them.
\verse And they said to one another, “Were not our hearts burning within us while he was speaking with us on the road, while he was explaining the scriptures to us?”
\verse And they got up that same hour and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“got up”) has been translated as a finite verb} returned to Jerusalem and found the eleven and those with them assembled,
\verse saying, “The Lord has really been raised, and has appeared to Simon!”
\verse And they began describing\lebnote{:|NP|:﻿*The imperfect tense has been translated as ingressive here (“began describing”)} \textit{what happened}\lebnote{Literally “the things”} on the road, and how he was recognized by them in the breaking of the bread.
\verseWithHeading{Jesus Appears to His Disciples} And while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“were saying”)} they were saying these things, he himself stood there among them.\lebnote{Some manuscripts add “and said to them, ‘Peace to you!’ ”}
\verse But they were startled and became terrified, and\lebnote{:|NP|:﻿*Here “and” is supplied because the two previous participles (“startled” and “terrified”) have been translated as finite verbs} thought they had seen a ghost.
\verse And he said to them, “Why are you frightened? And for what reason do doubts arise in your hearts?
\verse Look at my hands and my feet, that I am I myself! Touch me and see, because a ghost does not have flesh and bones, as \textit{you see that I have}.”\lebnote{Literally “you see me having”}
\verse And when he\lebnote{:|NP|:﻿*Here “when” is supplied as a component of the participle (“had said”) which is understood as temporal} had said this, he showed them his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} hands and his\lebnote{:|NP|:﻿*Literally “the”; the Greek article is used here as a possessive pronoun} feet.
\verse And while\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the temporal genitive absolute participle (“disbelieving”)} they were still disbelieving because of joy and were marveling, he said to them, “Do you have anything \textit{to eat}\lebnote{Literally “eatable”} here?”
\verse So they gave him a piece of broiled fish,
\verse and he took it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“took”) has been translated as a finite verb} ate it\lebnote{:|NP|:﻿*Here the direct object is supplied from context in the English translation} in front of them.
\verseWithHeading{Jesus Commissions His Disciples} And he said to them, “These are my words that I spoke to you while I\lebnote{:|NP|:﻿*Here “while” is supplied as a component of the participle (“was”) which is understood as temporal} was still with you, that everything that is written about me in the law of Moses and the prophets and psalms must be fulfilled.”
\verse Then he opened their minds to understand the scriptures,
\verse and said to them, “Thus it is written that the Christ would suffer and would rise from the dead on the third day,
\verse and repentance and the forgiveness\lebnote{Some manuscripts have “repentance for the forgiveness”} of sins would be proclaimed in his name to all the nations,\lebnote{The same Greek word can be translated “nations” or “Gentiles” depending on the context} beginning from Jerusalem.
\verse You are witnesses of these things.
\verse And behold, I am sending out\lebnote{Some manuscripts have “am sending”} what was promised by my Father upon you, but you stay in the city until you are clothed with power from on high.”
\verseWithHeading{The Ascension} And he led them out as far as Bethany, and lifting up his hands, he blessed them.
\verse And it happened that while he was blessing them, he parted from them and was taken up into heaven.
\verse And they worshiped him and\lebnote{:|NP|:﻿*Here “and” is supplied because the previous participle (“worshiped”) has been translated as a finite verb} returned to Jerusalem with great joy.
\verse And they were \textit{continually}\lebnote{Literally “through everything”} in the temple courts\lebnote{:|NP|:﻿*Here “courts” is supplied to distinguish this area from the interior of the temple building itself} praising God.
\end{biblechapter}

