\biblebook{Philemon}

\begin{biblechapter} % Philemon 1
\verseWithHeading{Greeting} Paul, a prisoner of Christ Jesus, and Timothy our brother, to Philemon, our dear friend and fellow worker,
\verse and to Apphia our sister, and to Archippus our fellow soldier, and to the church in your house.
\verse Grace to you and peace from God our Father and the Lord Jesus Christ.
\verseWithHeading{Thanksgiving and Prayer for Philemon} I thank my God, always making mention of you in my prayers,
\verse because I\lebnote{:|NP|:﻿*Here “because” is supplied as a component of the participle (“hear about”) which is understood as causal} hear about your love and faith which you have toward the Lord Jesus and for all the saints.
\verse I pray that the fellowship of your faith may become effective in the knowledge of every good thing that is in us for Christ.
\verse For I have great joy and encouragement from your love, because the hearts of the saints have been refreshed through you, brother.
\verseWithHeading{Paul’s Appeal for Onesimus} Therefore, although I\lebnote{:|NP|:﻿*Here “although” is supplied as a component of the participle (“have”) which is understood as concessive} have great confidence in Christ to order you to do what is proper,
\verse instead I appeal to you because of love, since I\lebnote{:|NP|:﻿*Here “since” is supplied as a component of the participle (“am”) which is understood as causal} am such a one as Paul, now an old man and also a prisoner of Christ Jesus.
\verse I am appealing to you concerning my child whom I became the father of during my imprisonment, Onesimus.
\verse Once he was useless to you, but now he is useful to you\lebnote{Some manuscripts have “both to you”} and to me,
\verse whom I have sent back to you himself, that is, my heart,
\verse whom I wanted to keep with me, in order that he might serve me on behalf of you during my imprisonment for the gospel.
\verse But apart from your consent, I wanted to do nothing, in order that your good deed might be not as according to necessity, but according to your own free will.
\verse For perhaps because of this, he was separated from you for a time, in order that you might have him back forever,
\verse no longer as a slave, but more than a slave, a beloved brother, especially to me, but how much more to you, both in the flesh and in the Lord.
\verseWithHeading{Paul’s Request to Philemon} If therefore you consider me a partner, receive him as you would me.
\verse But if in anything he has caused you loss or owes you anything, \textit{charge this to my account}\lebnote{Literally “this to me charge to my account”}.
\verse I, Paul, write this with my own hand: I will pay it back, lest I mention to you that \textit{you owe me even your very self besides}\lebnote{Literally “even yourself to me you owe besides”}.
\verse Yes, brother, I ought to have some benefit of you in the Lord; refresh my heart in Christ.
\verse Confident of your obedience, I am writing to you because I\lebnote{:|NP|:﻿*Here “because” is supplied as a component of the participle (“know”) which is understood as causal} know that you will do even beyond what I say.
\verse At the same time also, prepare a guest room for me, for I hope that through your prayers I will be restored to you.
\verseWithHeading{Final Greetings and Benediction} Epaphras, my fellow prisoner in Christ Jesus, greets you,
\verse and so do Mark, Aristarchus, Demas, and Luke, my fellow workers.
\verse The grace of the Lord Jesus Christ be with your spirit.
\end{biblechapter}

