\biblebook{Titus}

\begin{biblechapter} % Titus 1
\verseWithHeading{Greeting} Paul, a slave of God and an apostle of Jesus Christ for the faith of the chosen of God and the knowledge of the truth that \textit{is} according to godliness,
\verse in the hope of eternal life which God, who does not lie, promised before eternal ages,
\verse but at the proper time has disclosed his message in the proclamation \textit{with} which I was entrusted according to the command of God our Savior,
\verse to Titus, \textit{my} true child according to a common faith. Grace and peace from God the Father and Christ Jesus our Savior.
\verseWithHeading{Instructions to Titus in Crete} On account of this, I left you behind in Crete, in order that what remains may be set in order and you may appoint elders in every town, as I ordered you.
\verse If anyone is blameless, \textit{the} husband of one wife, having faithful children, not \textit{accused of dissipation} or rebellious.
\verse For it is necessary \textit{for} the overseer to be blameless as God’s steward, not self-willed, not quick-tempered, not addicted to wine, not violent, not greedy for dishonest gain,
\verse but hospitable, loving what is good, prudent, just, devout, self-controlled,
\verse holding fast to the faithful message according to the teaching, in order that he may be able both to exhort with sound instruction and to reprove those who speak against \textit{it}.
\verse For there are many rebellious \textit{people}, idle talkers and deceivers, especially those of the circumcision,
\verse whom it is necessary to silence, whoever are ruining whole families \textit{by} teaching \textit{things} which must not be \textit{taught} for the sake of dishonest gain.
\verse A certain one of them, \textit{one of} their own prophets, has said, “Cretans \textit{are} always liars, evil beasts, lazy gluttons.”
\verse This testimony is true, for which reason reprove them severely, in order that they may be sound in the faith,
\verse not paying attention to Jewish myths and commandments of people who turn away from the truth.
\verse To the pure all \textit{things} \textit{are} pure, but to those who are defiled and unbelieving nothing \textit{is} pure, but both their mind and conscience are defiled.
\verse They profess to know God, but by their deeds they deny \textit{him}, \textit{because they} are detestable and disobedient, and unfit for every good deed.
\end{biblechapter}

\begin{biblechapter} % Titus 2
\verseWithHeading{Sound Instruction and Consistent Behavior} But you, speak \textit{the things} which are fitting for sound instruction.
\verse Older men are to be temperate, worthy of respect, self-controlled, sound in faith, in love, \textit{and} in endurance.
\verse Older women likewise \textit{are to be} reverent in \textit{their} behavior, not slanderous, not enslaved to much wine, teaching what is good,
\verse in order that they may encourage the young women to love their husbands \textit{and to} love their children,
\verse \textit{to be} self-controlled, pure, busy at home, good, being subject to their own husbands, in order that the word of God may not be slandered.
\verse Likewise, exhort the young men to be self-controlled,
\verse concerning everything showing yourself \textit{to be} an example of good deeds, in your teaching \textit{demonstrating} soundness, dignity,
\verse \textit{a} sound message beyond reproach, in order that \textit{an opponent} may be put to shame, \textit{because he} has nothing bad to say concerning us.
\verse Slaves must be subject to their own masters in everything, to be well-pleasing, not to talk back,
\verse not stealing, but demonstrating all good faith, in order that they may do credit to the teaching of God our Savior in everything.
\verse For the grace of God has appeared, bringing salvation to all people,
\verse training us in order that, denying impiety and worldly desires, we may live self-controlled and righteously and godly in the present age,
\verse looking forward to the blessed hope and the glorious appearing of our great God and Savior Jesus Christ,
\verse who gave himself for us, in order that he might redeem us from all lawlessness and purify for himself a people for his own possession, zealous for good deeds.
\verse Speak these \textit{things} and exhort and reprove with all authority. Let no one disregard you.
\end{biblechapter}

\begin{biblechapter} % Titus 3
\verseWithHeading{Conduct Toward All People} Remind them to be subject to the rulers \textit{and} to the authorities, to obey, to be prepared for every good work,
\verse to speak evil of no one, to be peaceable, gentle, showing all courtesy to all people.
\verse For we also were once foolish, disobedient, led astray, enslaved to various desires and pleasures, spending our lives in wickedness and envy, despicable, hating one another.
\verse But when the kindness and love for mankind of God our Savior appeared,
\verse he saved us, not by deeds of righteousness that we have done, but because of his mercy, through the washing of regeneration and renewal by the Holy Spirit,
\verse whom he poured out on us abundantly through Jesus Christ our Savior,
\verse so that, having been justified by his grace, we may become heirs according to the hope of eternal life.
\verseWithHeading{Final Instructions to Titus} The saying \textit{is} trustworthy, and I want you to insist concerning these \textit{things}, so that those who have believed in God may be careful to engage in good deeds. These things are good and beneficial for people.
\verse But avoid foolish controversies and genealogies and contentions and quarrels about the law, for they are useless and fruitless.
\verse Reject a divisive person after a first and second admonition,
\verse knowing that such a person is perverted and is sinning, being self-condemned.
\verse When I send Artemas or Tychicus to you, make haste to come to me in Nicopolis, for I have decided to spend the winter there.
\verse Diligently send on their way Zenas the lawyer and Apollos, so that \textit{they may lack nothing}.
\verse But also our \textit{people} must learn to engage in good deeds for necessary needs, so that they will not be unfruitful.
\verse All those with me greet you. Greet those who love us in the faith. Grace \textit{be} with all of you.
\end{biblechapter}

